### 2012-03-01 LCAforActiveSonarImaging/lca.tex
Branching was stupid. Reverting.

% Note that the a4paper option is mainly intended so that authors in
% countries using A4 can easily print to A4 and see how their papers will
% look in print - the typesetting of the document will not typically be
% affected with changes in paper size (but the bottom and side margins will).
% Use the testflow package mentioned above to verify correct handling of
% both paper sizes by the user's LaTeX system.
%
% Also note that the "draftcls" or "draftclsnofoot", not "draft", option
% should be used if it is desired that the figures are to be displayed in
% draft mode.
%
\documentclass[12pt,journal,compsoc]{ieee/IEEEtran}
% The Computer Society requires 12pt.
% If IEEEtran.cls has not been installed into the LaTeX system files,
% manually specify the path to it like:
% \documentclass[10pt,journal,compsoc]{../sty/IEEEtran}

% For Computer Society journals, IEEEtran defaults to the use of 
% Palatino/Palladio as is done in IEEE Computer Society journals.
% To go back to Times Roman, you can use this code:
%\renewcommand{\rmdefault}{ptm}\selectfont

% Some very useful LaTeX packages include:
% (uncomment the ones you want to load)

% *** MISC UTILITY PACKAGES ***
%
%\usepackage{ifpdf}
% Heiko Oberdiek's ifpdf.sty is very useful if you need conditional
% compilation based on whether the output is pdf or dvi.
% usage:
% \ifpdf
%   % pdf code
% \else
%   % dvi code
% \fi
% The latest version of ifpdf.sty can be obtained from:
% http://www.ctan.org/tex-archive/macros/latex/contrib/oberdiek/
% Also, note that IEEEtran.cls V1.7 and later provides a builtin
% \ifCLASSINFOpdf conditional that works the same way.
% When switching from latex to pdflatex and vice-versa, the compiler may
% have to be run twice to clear warning/error messages.

% *** CITATION PACKAGES ***
%
\ifCLASSOPTIONcompsoc
  % IEEE Computer Society needs nocompress option
  % requires cite.sty v4.0 or later (November 2003)
  % \usepackage[nocompress]{cite}
\else
  % normal IEEE
  % \usepackage{cite}
\fi
% cite.sty was written by Donald Arseneau
% V1.6 and later of IEEEtran pre-defines the format of the cite.sty package
% \cite{} output to follow that of IEEE. Loading the cite package will
% result in citation numbers being automatically sorted and properly
% "compressed/ranged". e.g., [1], [9], [2], [7], [5], [6] without using
% cite.sty will become [1], [2], [5]--[7], [9] using cite.sty. cite.sty's
% \cite will automatically add leading space, if needed. Use cite.sty's
% noadjust option (cite.sty V3.8 and later) if you want to turn this off.
% cite.sty is already installed on most LaTeX systems. Be sure and use
% version 4.0 (2003-05-27) and later if using hyperref.sty. cite.sty does
% not currently provide for hyperlinked citations.
% The latest version can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/cite/
% The documentation is contained in the cite.sty file itself.
%
% Note that some packages require special options to format as the Computer
% Society requires. In particular, Computer Society  papers do not use
% compressed citation ranges as is done in typical IEEE papers
% (e.g., [1]-[4]). Instead, they list every citation separately in order
% (e.g., [1], [2], [3], [4]). To get the latter we need to load the cite
% package with the nocompress option which is supported by cite.sty v4.0
% and later. Note also the use of a CLASSOPTION conditional provided by
% IEEEtran.cls V1.7 and later.

% *** GRAPHICS RELATED PACKAGES ***
%
\ifCLASSINFOpdf
  % \usepackage[pdftex]{graphicx}
  % declare the path(s) where your graphic files are
  % \graphicspath{{../pdf/}{../jpeg/}}
  % and their extensions so you won't have to specify these with
  % every instance of \includegraphics
  % \DeclareGraphicsExtensions{.pdf,.jpeg,.png}
\else
  % or other class option (dvipsone, dvipdf, if not using dvips). graphicx
  % will default to the driver specified in the system graphics.cfg if no
  % driver is specified.
  % \usepackage[dvips]{graphicx}
  % declare the path(s) where your graphic files are
  % \graphicspath{{../eps/}}
  % and their extensions so you won't have to specify these with
  % every instance of \includegraphics
  % \DeclareGraphicsExtensions{.eps}
\fi
% graphicx was written by David Carlisle and Sebastian Rahtz. It is
% required if you want graphics, photos, etc. graphicx.sty is already
% installed on most LaTeX systems. The latest version and documentation can
% be obtained at: 
% http://www.ctan.org/tex-archive/macros/latex/required/graphics/
% Another good source of documentation is "Using Imported Graphics in
% LaTeX2e" by Keith Reckdahl which can be found as epslatex.ps or
% epslatex.pdf at: http://www.ctan.org/tex-archive/info/
%
% latex, and pdflatex in dvi mode, support graphics in encapsulated
% postscript (.eps) format. pdflatex in pdf mode supports graphics
% in .pdf, .jpeg, .png and .mps (metapost) formats. Users should ensure
% that all non-photo figures use a vector format (.eps, .pdf, .mps) and
% not a bitmapped formats (.jpeg, .png). IEEE frowns on bitmapped formats
% which can result in "jaggedy"/blurry rendering of lines and letters as
% well as large increases in file sizes.
%
% You can find documentation about the pdfTeX application at:
% http://www.tug.org/applications/pdftex

%\usepackage{ps4pdf}
% dvi->ps workflow is required to use such packages as psfrag.sty and
% pstricks.sty. However, Rolf Niepraschk's ps4pdf.sty provides a way to
% apply psfrag/pstricks effects to .eps figures and then get the resultant
% figures in .pdf form. Thus, providing an easier way for migrating from
% .eps to .pdf figures. After ps4pdf.sty loads, if:
% 1. producing .dvi output: the output file will consist ONLY of the
%    figures (or other constructs encased within \PSforPDF commands)
% 2. producing .pdf output: pdflatex will look in the filename-pics.pdf
%    file, where filename is the basename of the tex document, for the
%    graphics (or other constructs encased within \PSforPDF commands).
%    NOTE: If you ever change your figures, you must remember to remake
%    the filename-pics.pdf file.
%
% This way you can do a:
% 
% latex filename
% dvips -Ppdf -o filename-pics.ps filename.dvi
% ps2pdf filename-pics.ps filename-pics.pdf
% 
% to produce a filename-pics.pdf graphics container that contains
% .pdf versions of the graphics with psfrag, pstricks, etc. features.
% Note that you will not typically be able to view the figures in 
% filename-pics.ps because of an offset. However, you will be able to
% view them in filename-pics.pdf. Also, note that when ps4pdf is in effect
% with .dvi output, you may get harmless over/under full box warnings - 
% ignore them. 
% Then, run pdflatex:
% 
% pdflatex filename
% 
% to use pdflatex to make PDF output, automatically using the figures in
% filename-pics.pdf. Alternatively, you could use dvips -i option to
% obtain separate .pdf files for each figure:
%
% dvips -Ppdf -i -E -o fig filename
%
% then convert each figure to pdf via a command such as epstopdf and then
% use pdflatex with these pdf figures and then to dispense with ps4pdf.
%
% Remember to rerun through latex/dvips/ps2pdf if you ever change your
% figures so that filename-pics.pdf gets updated.
% ps4pdf requires David Kastrup's preview-latex and a recent LaTeX system
% (circa 2001 or later). The ps4pdf package and documentation can be
% obtained at: http://www.ctan.org/tex-archive/macros/latex/contrib/ps4pdf/
% The preview-latex package and documentation can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/preview/
%
% provide a bogus \PSforPDF, even when not loading pd4pdf. This way we can
% stop loading ps4pdf.sty if we choose to make separate .pdf versions of
% each of our figures.
\providecommand{\PSforPDF}[1]{#1}
% Note that in order for ps4pdf to work, all commands related to psfrag,
% pstricks, etc. must be called within the PSforPDF command. This applies
% even when *loading* via \usepackage psfrag.sty, etc.

%\PSforPDF{\usepackage{psfrag}}
% psfrag.sty was written by Craig Barratt, Michael C. Grant, and
% David Carlisle. It allows you to substitute LaTeX commands for text in
% imported EPS graphic files. In this way, LaTeX symbols can be placed into
% graphics that have been generated by other applications. You must use
% latex->dvips->ps2pdf workflow (not direct pdf output from pdflatex) if
% you wish to use this capability because it works via some PostScript
% tricks. Alternatively, the graphics could be processed as separate files
% via psfrag and dvips, then converted to PDF for inclusion in the main file
% which uses pdflatex. ps4pdf.sty (above) provides a way of doing this all
% at once within the main file.
% Docs are in "The PSfrag System" by Michael C. Grant and David Carlisle.
% There is also some information about using psfrag in "Using Imported
% Graphics in LaTeX2e" by Keith Reckdahl which documents the graphicx
% package (see above). The psfrag package and documentation can be obtained
% at: http://www.ctan.org/tex-archive/macros/latex/contrib/psfrag/
% 
% Note that the current version of psfrag does not "turn itself off" when
% running under pdf output. This will result in a harmless warning
% about a non-PDF \special. However, to silence this, a bogus psfrag
% command can be provided instead of loading psfrag.sty when PDF output
% is being used. Thus, a more complex alternative conditional loading scheme
% can be employed instead of the straightforword way above:
%
%\ifCLASSINFOpdf
% if outputting PDF, do not use or load psfrag.sty as current versions
% output a non-PDF special that generates a harmless, but annoying warning.
% Instead, we provide a bogus \psfrag command that does nothing with
% its arguments. This is a tad tricky because \psfrag can have up to six
% arguments four of which are optional: \psfrag{}[][][][]{}
% Code based on that in psfrag.sty
%\makeatletter
%\def\psfrag{\@ifstar{\@BOGUSpsfraga}{\@BOGUSpsfraga}}
%\def\@BOGUSpsfraga{\begingroup
%   \@makeother\"\@makeother\*\@makeother\!\@makeother\~%
%   \@makeother\:\@makeother\\\@makeother\%\@makeother\#%
%   \@makeother\ \@BOGUSpsfragb}
%\def\@BOGUSpsfragb#1{\endgroup
%                \@ifnextchar [{\@BOGUSpsfragc}%
%                              {\@BOGUSpsfrag}}
%\def\@BOGUSpsfragc[#1]{\@ifnextchar [{\@BOGUSpsfragd}%
%                                     {\@BOGUSpsfrag}}
%\def\@BOGUSpsfragd[#1]{\@ifnextchar [{\@BOGUSpsfrage}%
%                                     {\@BOGUSpsfrag}}
%\def\@BOGUSpsfrage[#1]{\@ifnextchar [{\@BOGUSpsfragf}%
%                                     {\@BOGUSpsfrag}}
%\def\@BOGUSpsfragf[#1]{\@BOGUSpsfrag}
%\def\@BOGUSpsfrag#1{\ignorespaces}
%\makeatother
%\else
% using dvi output, load psfrag, but funnel it through PSforPDF
% as required by ps4pdf.sty
%\PSforPDF{\usepackage{psfrag}}
%\fi

% *** MATH PACKAGES ***
%
%\usepackage[cmex10]{amsmath}
% A popular package from the American Mathematical Society that provides
% many useful and powerful commands for dealing with mathematics. If using
% it, be sure to load this package with the cmex10 option to ensure that
% only type 1 fonts will utilized at all point sizes. Without this option,
% it is possible that some math symbols, particularly those within
% footnotes, will be rendered in bitmap form which will result in a
% document that can not be IEEE Xplore compliant!
%
% Also, note that the amsmath package sets \interdisplaylinepenalty to 10000
% thus preventing page breaks from occurring within multiline equations. Use:
%\interdisplaylinepenalty=2500
% after loading amsmath to restore such page breaks as IEEEtran.cls normally
% does. amsmath.sty is already installed on most LaTeX systems. The latest
% version and documentation can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/required/amslatex/math/

% *** SPECIALIZED LIST PACKAGES ***
%\usepackage{acronym}
% acronym.sty was written by Tobias Oetiker. This package provides tools for
% managing documents with large numbers of acronyms. (You don't *have* to
% use this package - unless you have a lot of acronyms, you may feel that
% such package management of them is bit of an overkill.)
% Do note that the acronym environment (which lists acronyms) will have a
% problem when used under IEEEtran.cls because acronym.sty relies on the
% description list environment - which IEEEtran.cls has customized for
% producing IEEE style lists. A workaround is to declared the longest
% label width via the IEEEtran.cls \IEEEiedlistdecl global control:
%
% \renewcommand{\IEEEiedlistdecl}{\IEEEsetlabelwidth{SONET}}
% \begin{acronym}
%
% \end{acronym}
% \renewcommand{\IEEEiedlistdecl}{\relax}% remember to reset \IEEEiedlistdecl
%
% instead of using the acronym environment's optional argument.
% The latest version and documentation can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/acronym/

%\usepackage{algorithmic}
% algorithmic.sty was written by Peter Williams and Rogerio Brito.
% This package provides an algorithmic environment fo describing algorithms.
% You can use the algorithmic environment in-text or within a figure
% environment to provide for a floating algorithm. Do NOT use the algorithm
% floating environment provided by algorithm.sty (by the same authors) or
% algorithm2e.sty (by Christophe Fiorio) as IEEE does not use dedicated
% algorithm float types and packages that provide these will not provide
% correct IEEE style captions. The latest version and documentation of
% algorithmic.sty can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/algorithms/
% There is also a support site at:
% http://algorithms.berlios.de/index.html
% Also of interest may be the (relatively newer and more customizable)
% algorithmicx.sty package by Szasz Janos:
% http://www.ctan.org/tex-archive/macros/latex/contrib/algorithmicx/

% *** ALIGNMENT PACKAGES ***
%
%\usepackage{array}
% Frank Mittelbach's and David Carlisle's array.sty patches and improves
% the standard LaTeX2e array and tabular environments to provide better
% appearance and additional user controls. As the default LaTeX2e table
% generation code is lacking to the point of almost being broken with
% respect to the quality of the end results, all users are strongly
% advised to use an enhanced (at the very least that provided by array.sty)
% set of table tools. array.sty is already installed on most systems. The
% latest version and documentation can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/required/tools/

%\usepackage{mdwmath}
%\usepackage{mdwtab}
% Also highly recommended is Mark Wooding's extremely powerful MDW tools,
% especially mdwmath.sty and mdwtab.sty which are used to format equations
% and tables, respectively. The MDWtools set is already installed on most
% LaTeX systems. The lastest version and documentation is available at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/mdwtools/

% IEEEtran contains the IEEEeqnarray family of commands that can be used to
% generate multiline equations as well as matrices, tables, etc., of high
% quality.

%\usepackage{eqparbox}
% Also of notable interest is Scott Pakin's eqparbox package for creating
% (automatically sized) equal width boxes - aka "natural width parboxes".
% Available at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/eqparbox/

% *** SUBFIGURE PACKAGES ***
%\ifCLASSOPTIONcompsoc
%\usepackage[tight,normalsize,sf,SF]{subfigure}
%\else
%\usepackage[tight,footnotesize]{subfigure}
%\fi
% subfigure.sty was written by Steven Douglas Cochran. This package makes it
% easy to put subfigures in your figures. e.g., "Figure 1a and 1b". For IEEE
% work, it is a good idea to load it with the tight package option to reduce
% the amount of white space around the subfigures. Computer Society papers
% use a larger font and \sffamily font for their captions, hence the
% additional options needed under compsoc mode. subfigure.sty is already
% installed on most LaTeX systems. The latest version and documentation can
% be obtained at:
% http://www.ctan.org/tex-archive/obsolete/macros/latex/contrib/subfigure/
% subfigure.sty has been superceeded by subfig.sty.

%\ifCLASSOPTIONcompsoc
%  \usepackage[caption=false]{caption}
%  \usepackage[font=normalsize,labelfont=sf,textfont=sf]{subfig}
%\else
%  \usepackage[caption=false]{caption}
%  \usepackage[font=footnotesize]{subfig}
%\fi
% subfig.sty, also written by Steven Douglas Cochran, is the modern
% replacement for subfigure.sty. However, subfig.sty requires and
% automatically loads Axel Sommerfeldt's caption.sty which will override
% IEEEtran.cls handling of captions and this will result in nonIEEE style
% figure/table captions. To prevent this problem, be sure and preload
% caption.sty with its "caption=false" package option. This is will preserve
% IEEEtran.cls handing of captions. Version 1.3 (2005/06/28) and later 
% (recommended due to many improvements over 1.2) of subfig.sty supports
% the caption=false option directly:
%\ifCLASSOPTIONcompsoc
%  \usepackage[caption=false,font=normalsize,labelfont=sf,textfont=sf]{subfig}
%\else
%  \usepackage[caption=false,font=footnotesize]{subfig}
%\fi
%
% The latest version and documentation can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/subfig/
% The latest version and documentation of caption.sty can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/caption/

% *** FLOAT PACKAGES ***
%
%\usepackage{fixltx2e}
% fixltx2e, the successor to the earlier fix2col.sty, was written by
% Frank Mittelbach and David Carlisle. This package corrects a few problems
% in the LaTeX2e kernel, the most notable of which is that in current
% LaTeX2e releases, the ordering of single and double column floats is not
% guaranteed to be preserved. Thus, an unpatched LaTeX2e can allow a
% single column figure to be placed prior to an earlier double column
% figure. The latest version and documentation can be found at:
% http://www.ctan.org/tex-archive/macros/latex/base/

%\usepackage{stfloats}
% stfloats.sty was written by Sigitas Tolusis. This package gives LaTeX2e
% the ability to do double column floats at the bottom of the page as well
% as the top. (e.g., "\begin{figure*}[!b]" is not normally possible in
% LaTeX2e). It also provides a command:
%\fnbelowfloat
% to enable the placement of footnotes below bottom floats (the standard
% LaTeX2e kernel puts them above bottom floats). This is an invasive package
% which rewrites many portions of the LaTeX2e float routines. It may not work
% with other packages that modify the LaTeX2e float routines. The latest
% version and documentation can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/sttools/
% Documentation is contained in the stfloats.sty comments as well as in the
% presfull.pdf file. Do not use the stfloats baselinefloat ability as IEEE
% does not allow \baselineskip to stretch. Authors submitting work to the
% IEEE should note that IEEE rarely uses double column equations and
% that authors should try to avoid such use. Do not be tempted to use the
% cuted.sty or midfloat.sty packages (also by Sigitas Tolusis) as IEEE does
% not format its papers in such ways.

%\ifCLASSOPTIONcaptionsoff
%  \usepackage[nomarkers]{endfloat}
% \let\MYoriglatexcaption\caption
% \renewcommand{\caption}[2][\relax]{\MYoriglatexcaption[#2]{#2}}
%\fi
% endfloat.sty was written by James Darrell McCauley and Jeff Goldberg.
% This package may be useful when used in conjunction with IEEEtran.cls'
% captionsoff option. Some IEEE journals/societies require that submissions
% have lists of figures/tables at the end of the paper and that
% figures/tables without any captions are placed on a page by themselves at
% the end of the document. If needed, the draftcls IEEEtran class option or
% \CLASSINPUTbaselinestretch interface can be used to increase the line
% spacing as well. Be sure and use the nomarkers option of endfloat to
% prevent endfloat from "marking" where the figures would have been placed
% in the text. The two hack lines of code above are a slight modification of
% that suggested by in the endfloat docs (section 8.3.1) to ensure that
% the full captions always appear in the list of figures/tables - even if
% the user used the short optional argument of \caption[]{}.
% IEEE papers do not typically make use of \caption[]'s optional argument,
% so this should not be an issue. A similar trick can be used to disable
% captions of packages such as subfig.sty that lack options to turn off
% the subcaptions:
% For subfig.sty:
% \let\MYorigsubfloat\subfloat
% \renewcommand{\subfloat}[2][\relax]{\MYorigsubfloat[]{#2}}
% For subfigure.sty:
% \let\MYorigsubfigure\subfigure
% \renewcommand{\subfigure}[2][\relax]{\MYorigsubfigure[]{#2}}
% However, the above trick will not work if both optional arguments of
% the \subfloat/subfig command are used. Furthermore, there needs to be a
% description of each subfigure *somewhere* and endfloat does not add
% subfigure captions to its list of figures. Thus, the best approach is to
% avoid the use of subfigure captions (many IEEE journals avoid them anyway)
% and instead reference/explain all the subfigures within the main caption.
% The latest version of endfloat.sty and its documentation can obtained at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/endfloat/
%
% The IEEEtran \ifCLASSOPTIONcaptionsoff conditional can also be used
% later in the document, say, to conditionally put the References on a 
% page by themselves.

% *** PDF, URL AND HYPERLINK PACKAGES ***
%
%\usepackage{url}
% url.sty was written by Donald Arseneau. It provides better support for
% handling and breaking URLs. url.sty is already installed on most LaTeX
% systems. The latest version can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/misc/
% Read the url.sty source comments for usage information. Basically,
% \url{my_url_here}.

% NOTE: PDF thumbnail features are not required in IEEE papers
%       and their use requires extra complexity and work.
%\ifCLASSINFOpdf
%  \usepackage[pdftex]{thumbpdf}
%\else
%  \usepackage[dvips]{thumbpdf}
%\fi
% thumbpdf.sty and its companion Perl utility were written by Heiko Oberdiek.
% It allows the user a way to produce PDF documents that contain fancy
% thumbnail images of each of the pages (which tools like acrobat reader can
% utilize). This is possible even when using dvi->ps->pdf workflow if the
% correct thumbpdf driver options are used. thumbpdf.sty incorporates the
% file containing the PDF thumbnail information (filename.tpm is used with
% dvips, filename.tpt is used with pdftex, where filename is the base name of
% your tex document) into the final ps or pdf output document. An external
% utility, the thumbpdf *Perl script* is needed to make these .tpm or .tpt
% thumbnail files from a .ps or .pdf version of the document (which obviously
% does not yet contain pdf thumbnails). Thus, one does a:
% 
% thumbpdf filename.pdf 
%
% to make a filename.tpt, and:
%
% thumbpdf --mode dvips filename.ps
%
% to make a filename.tpm which will then be loaded into the document by
% thumbpdf.sty the NEXT time the document is compiled (by pdflatex or
% latex->dvips->ps2pdf). Users must be careful to regenerate the .tpt and/or
% .tpm files if the main document changes and then to recompile the
% document to incorporate the revised thumbnails to ensure that thumbnails
% match the actual pages. It is easy to forget to do this!
% 
% Unix systems come with a Perl interpreter. However, MS Windows users
% will usually have to install a Perl interpreter so that the thumbpdf
% script can be run. The Ghostscript PS/PDF interpreter is also required.
% See the thumbpdf docs for details. The latest version and documentation
% can be obtained at.
% http://www.ctan.org/tex-archive/support/thumbpdf/
% Be sure and use only version 3.8 (2005/07/06) or later of thumbpdf as
% earlier versions will not work properly with recent versions of pdfTeX
% (1.20a and later).

% NOTE: PDF hyperlink and bookmark features are not required in IEEE
%       papers and their use requires extra complexity and work.
% *** IF USING HYPERREF BE SURE AND CHANGE THE EXAMPLE PDF ***
% *** TITLE/SUBJECT/AUTHOR/KEYWORDS INFO BELOW!!           ***
\newcommand\MYhyperrefoptions{bookmarks=true,bookmarksnumbered=true,
pdfpagemode={UseOutlines},plainpages=false,pdfpagelabels=true,
colorlinks=true,linkcolor={black},citecolor={black},pagecolor={black},
urlcolor={black},
pdftitle={Bare Demo of IEEEtran.cls for Computer Society Journals},%<!CHANGE!
pdfsubject={Typesetting},%<!CHANGE!
pdfauthor={Michael D. Shell},%<!CHANGE!
pdfkeywords={Computer Society, IEEEtran, journal, LaTeX, paper,
             template}}%<^!CHANGE!
%\ifCLASSINFOpdf
%\usepackage[\MYhyperrefoptions,pdftex]{hyperref}
%\else
%\usepackage[\MYhyperrefoptions,breaklinks=true,dvips]{hyperref}
%\usepackage{breakurl}
%\fi
% One significant drawback of using hyperref under DVI output is that the
% LaTeX compiler cannot break URLs across lines or pages as can be done
% under pdfLaTeX's PDF output via the hyperref pdftex driver. This is
% probably the single most important capability distinction between the
% DVI and PDF output. Perhaps surprisingly, all the other PDF features
% (PDF bookmarks, thumbnails, etc.) can be preserved in
% .tex->.dvi->.ps->.pdf workflow if the respective packages/scripts are
% loaded/invoked with the correct driver options (dvips, etc.). 
% As most IEEE papers use URLs sparingly (mainly in the references), this
% may not be as big an issue as with other publications.
%
% That said, recently Vilar Camara Neto introduced his breakurl.sty
% package which permits hyperref to easily break URLs even in dvi
% mode. Note that breakurl, unlike most other packages, must be loaded
% AFTER hyperref. The latest version of breakurl and its documentation can
% be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/breakurl/
% breakurl.sty is not for use under pdflatex pdf mode. Versions 1.10 
% (September 23, 2005) and later are recommened to avoid bugs in earlier
% releases.
%
% The advanced features offer by hyperref.sty are not required for IEEE
% submission, so users should weigh these features against the added
% complexity of use. Users who wish to use hyperref *must* ensure that
% their hyperref version is 6.72u or later *and* IEEEtran.cls is version
% 1.6b or later.
% The package options above demonstrate how to enable PDF bookmarks
% (a type of table of contents viewable in Acrobat Reader) as well as
% PDF document information (title, subject, author and keywords) that is
% viewable in Acrobat reader's Document_Properties menu. PDF document
% information is also used extensively to automate the cataloging of PDF
% documents. The above set of options ensures that hyperlinks will not be
% colored in the text and thus will not be visible in the printed page,
% but will be active on "mouse over". USING COLORS OR OTHER HIGHLIGHTING
% OF HYPERLINKS CAN RESULT IN DOCUMENT REJECTION BY THE IEEE, especially if
% these appear on the "printed" page. IF IN DOUBT, ASK THE RELEVANT
% SUBMISSION EDITOR. You may need to add the option hypertexnames=false if
% you used duplicate equation numbers, etc., but this should not be needed
% in normal IEEE work.
% The latest version of hyperref and its documentation can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/hyperref/

% *** Do not adjust lengths that control margins, column widths, etc. ***
% *** Do not use packages that alter fonts (such as pslatex).         ***
% There should be no need to do such things with IEEEtran.cls V1.6 and later.
% (Unless specifically asked to do so by the journal or conference you plan
% to submit to, of course. )

% correct bad hyphenation here
\hyphenation{op-tical net-works semi-conduc-tor}

\begin{document}
%
% paper title
% can use linebreaks \\ within to get better formatting as desired
\title{Bare Advanced Demo of IEEEtran.cls\\ for Computer Society Journals}
%
%
% author names and IEEE memberships
% note positions of commas and nonbreaking spaces ( ~ ) LaTeX will not break
% a structure at a ~ so this keeps an author's name from being broken across
% two lines.
% use \thanks{} to gain access to the first footnote area
% a separate \thanks must be used for each paragraph as LaTeX2e's \thanks
% was not built to handle multiple paragraphs
%
%
%\IEEEcompsocitemizethanks is a special \thanks that produces the bulleted
% lists the Computer Society journals use for "first footnote" author
% affiliations. Use \IEEEcompsocthanksitem which works much like \item
% for each affiliation group. When not in compsoc mode,
% \IEEEcompsocitemizethanks becomes like \thanks and
% \IEEEcompsocthanksitem becomes a line break with idention. This
% facilitates dual compilation, although admittedly the differences in the
% desired content of \author between the different types of papers makes a
% one-size-fits-all approach a daunting prospect. For instance, compsoc 
% journal papers have the author affiliations above the "Manuscript
% received ..."  text while in non-compsoc journals this is reversed. Sigh.

\author{Michael~Shell,~\IEEEmembership{Member,~IEEE,}
        John~Doe,~\IEEEmembership{Fellow,~OSA,}
        and~Jane~Doe,~\IEEEmembership{Life~Fellow,~IEEE}% <-this % stops a space
\IEEEcompsocitemizethanks{\IEEEcompsocthanksitem M. Shell is with the Department
of Electrical and Computer Engineering, Georgia Institute of Technology, Atlanta,
GA, 30332.\protect\\
% note need leading \protect in front of \\ to get a newline within \thanks as
% \\ is fragile and will error, could use \hfil\break instead.
E-mail: see http://www.michaelshell.org/contact.html
\IEEEcompsocthanksitem J. Doe and J. Doe are with Anonymous University.}% <-this % stops a space
\thanks{Manuscript received April 19, 2005; revised January 11, 2007.}}

% note the % following the last \IEEEmembership and also \thanks - 
% these prevent an unwanted space from occurring between the last author name
% and the end of the author line. i.e., if you had this:
% 
% \author{....lastname \thanks{...} \thanks{...} }
%                     ^------------^------------^----Do not want these spaces!
%
% a space would be appended to the last name and could cause every name on that
% line to be shifted left slightly. This is one of those "LaTeX things". For
% instance, "\textbf{A} \textbf{B}" will typeset as "A B" not "AB". To get
% "AB" then you have to do: "\textbf{A}\textbf{B}"
% \thanks is no different in this regard, so shield the last } of each \thanks
% that ends a line with a % and do not let a space in before the next \thanks.
% Spaces after \IEEEmembership other than the last one are OK (and needed) as
% you are supposed to have spaces between the names. For what it is worth,
% this is a minor point as most people would not even notice if the said evil
% space somehow managed to creep in.

% The paper headers
\markboth{Journal of \LaTeX\ Class Files,~Vol.~6, No.~1, January~2007}%
{Shell \MakeLowercase{\textit{et al.}}: Bare Advanced Demo of IEEEtran.cls for Journals}
% The only time the second header will appear is for the odd numbered pages
% after the title page when using the twoside option.
% 
% *** Note that you probably will NOT want to include the author's ***
% *** name in the headers of peer review papers.                   ***
% You can use \ifCLASSOPTIONpeerreview for conditional compilation here if
% you desire.

% The publisher's ID mark at the bottom of the page is less important with
% Computer Society journal papers as those publications place the marks
% outside of the main text columns and, therefore, unlike regular IEEE
% journals, the available text space is not reduced by their presence.
% If you want to put a publisher's ID mark on the page you can do it like
% this:
%\IEEEpubid{0000--0000/00\$00.00~\copyright~2007 IEEE}
% or like this to get the Computer Society new two part style.
%\IEEEpubid{\makebox[\columnwidth]{\hfill 0000--0000/00/\$00.00~\copyright~2007 IEEE}%
%\hspace{\columnsep}\makebox[\columnwidth]{Published by the IEEE Computer Society\hfill}}
% Remember, if you use this you must call \IEEEpubidadjcol in the second
% column for its text to clear the IEEEpubid mark (Computer Society jorunal
% papers don't need this extra clearance.)

% use for special paper notices
%\IEEEspecialpapernotice{(Invited Paper)}

% for Computer Society papers, we must declare the abstract and index terms
% PRIOR to the title within the \IEEEcompsoctitleabstractindextext IEEEtran
% command as these need to go into the title area created by \maketitle.
\IEEEcompsoctitleabstractindextext{%
\begin{abstract}
%\boldmath
The abstract goes here.
\end{abstract}
% IEEEtran.cls defaults to using nonbold math in the Abstract.
% This preserves the distinction between vectors and scalars. However,
% if the journal you are submitting to favors bold math in the abstract,
% then you can use LaTeX's standard command \boldmath at the very start
% of the abstract to achieve this. Many IEEE journals frown on math
% in the abstract anyway. In particular, the Computer Society does
% not want either math or citations to appear in the abstract.

% Note that keywords are not normally used for peerreview papers.
\begin{IEEEkeywords}
Computer Society, IEEEtran, journal, \LaTeX, paper, template.
\end{IEEEkeywords}}

% make the title area
\maketitle

% To allow for easy dual compilation without having to reenter the
% abstract/keywords data, the \IEEEcompsoctitleabstractindextext text will
% not be used in maketitle, but will appear (i.e., to be "transported")
% here as \IEEEdisplaynotcompsoctitleabstractindextext when compsoc mode
% is not selected <OR> if conference mode is selected - because compsoc
% conference papers position the abstract like regular (non-compsoc)
% papers do!
\IEEEdisplaynotcompsoctitleabstractindextext
% \IEEEdisplaynotcompsoctitleabstractindextext has no effect when using
% compsoc under a non-conference mode.

% For peer review papers, you can put extra information on the cover
% page as needed:
% \ifCLASSOPTIONpeerreview
% \begin{center} \bfseries EDICS Category: 3-BBND \end{center}
% \fi
%
% For peerreview papers, this IEEEtran command inserts a page break and
% creates the second title. It will be ignored for other modes.
\IEEEpeerreviewmaketitle

\section{Introduction}
% Computer Society journal papers do something a tad strange with the very
% first section heading (almost always called "Introduction"). They place it
% ABOVE the main text! IEEEtran.cls currently does not do this for you.
% However, You can achieve this effect by making LaTeX jump through some
% hoops via something like:
%
%\ifCLASSOPTIONcompsoc
%  \noindent\raisebox{2\baselineskip}[0pt][0pt]%
%  {\parbox{\columnwidth}{\section{Introduction}\label{sec:introduction}%
%  \global\everypar=\everypar}}%
%  \vspace{-1\baselineskip}\vspace{-\parskip}\par
%\else
%  \section{Introduction}\label{sec:introduction}\par
%\fi
%
% Admittedly, this is a hack and may well be fragile, but seems to do the
% trick for me. Note the need to keep any \label that may be used right
% after \section in the above as the hack puts \section within a raised box.

% The very first letter is a 2 line initial drop letter followed
% by the rest of the first word in caps (small caps for compsoc).
% 
% form to use if the first word consists of a single letter:
% \IEEEPARstart{A}{demo} file is ....
% 
% form to use if you need the single drop letter followed by
% normal text (unknown if ever used by IEEE):
% \IEEEPARstart{A}{}demo file is ....
% 
% Some journals put the first two words in caps:
% \IEEEPARstart{T}{his demo} file is ....
% 
% Here we have the typical use of a "T" for an initial drop letter
% and "HIS" in caps to complete the first word.
\IEEEPARstart{T}{his} demo file is intended to serve as a ``starter file''
for IEEE Computer Society journal papers produced under \LaTeX\ using
IEEEtran.cls version 1.7 and later.
% You must have at least 2 lines in the paragraph with the drop letter
% (should never be an issue)
I wish you the best of success.

\hfill mds

\hfill January 11, 2007

\subsection{Subsection Heading Here}
Subsection text here.

% needed in second column of first page if using \IEEEpubid
%\IEEEpubidadjcol

\subsubsection{Subsubsection Heading Here}
Subsubsection text here.

% An example of a floating figure using the graphicx package.
% Note that \label must occur AFTER (or within) \caption.
% For figures, \caption should occur after the \includegraphics.
% Note that IEEEtran v1.7 and later has special internal code that
% is designed to preserve the operation of \label within \caption
% even when the captionsoff option is in effect. However, because
% of issues like this, it may be the safest practice to put all your
% \label just after \caption rather than within \caption{}.
%
% Reminder: the "draftcls" or "draftclsnofoot", not "draft", class
% option should be used if it is desired that the figures are to be
% displayed while in draft mode.
%
%\begin{figure}[!t]
%\centering
%\includegraphics[width=2.5in]{myfigure}
% where an .eps filename suffix will be assumed under latex, 
% and a .pdf suffix will be assumed for pdflatex; or what has been declared
% via \DeclareGraphicsExtensions.
%\caption{Simulation Results}
%\label{fig_sim}
%\end{figure}

% Note that IEEE typically puts floats only at the top, even when this
% results in a large percentage of a column being occupied by floats.
% However, the Computer Society has been known to put floats at the bottom.

% An example of a double column floating figure using two subfigures.
% (The subfig.sty package must be loaded for this to work.)
% The subfigure \label commands are set within each subfloat command, the
% \label for the overall figure must come after \caption.
% \hfil must be used as a separator to get equal spacing.
% The subfigure.sty package works much the same way, except \subfigure is
% used instead of \subfloat.
%
%\begin{figure*}[!t]
%\centerline{\subfloat[Case I]\includegraphics[width=2.5in]{subfigcase1}%
%\label{fig_first_case}}
%\hfil
%\subfloat[Case II]{\includegraphics[width=2.5in]{subfigcase2}%
%\label{fig_second_case}}}
%\caption{Simulation results}
%\label{fig_sim}
%\end{figure*}
%
% Note that often IEEE papers with subfigures do not employ subfigure
% captions (using the optional argument to \subfloat), but instead will
% reference/describe all of them (a), (b), etc., within the main caption.

% An example of a floating table. Note that, for IEEE style tables, the 
% \caption command should come BEFORE the table. Table text will default to
% \footnotesize as IEEE normally uses this smaller font for tables.
% The \label must come after \caption as always.
%
%\begin{table}[!t]
%% increase table row spacing, adjust to taste
%\renewcommand{\arraystretch}{1.3}
% if using array.sty, it might be a good idea to tweak the value of
% \extrarowheight as needed to properly center the text within the cells
%\caption{An Example of a Table}
%\label{table_example}
%\centering
%% Some packages, such as MDW tools, offer better commands for making tables
%% than the plain LaTeX2e tabular which is used here.
%\begin{tabular}{|c||c|}
%\hline
%One & Two\\
%\hline
%Three & Four\\
%\hline
%\end{tabular}
%\end{table}

% Note that IEEE does not put floats in the very first column - or typically
% anywhere on the first page for that matter. Also, in-text middle ("here")
% positioning is not used. Most IEEE journals use top floats exclusively.
% However, Computer Society journals sometimes do use bottom floats - bear
% this in mind when choosing appropriate optional arguments for the
% figure/table environments.
% Note that, LaTeX2e, unlike IEEE journals, places footnotes above bottom
% floats. This can be corrected via the \fnbelowfloat command of the
% stfloats package.

\section{Conclusion}
The conclusion goes here.

% if have a single appendix:
%\appendix[Proof of the Zonklar Equations]
% or
%\appendix  % for no appendix heading
% do not use \section anymore after \appendix, only \section*
% is possibly needed

% use appendices with more than one appendix
% then use \section to start each appendix
% you must declare a \section before using any
% \subsection or using \label (\appendices by itself
% starts a section numbered zero.)
%

\appendices
\section{Proof of the First Zonklar Equation}
Appendix one text goes here.

% you can choose not to have a title for an appendix
% if you want by leaving the argument blank
\section{}
Appendix two text goes here.

% use section* for acknowledgement
\ifCLASSOPTIONcompsoc
  % The Computer Society usually uses the plural form
  \section*{Acknowledgments}
\else
  % regular IEEE prefers the singular form
  \section*{Acknowledgment}
\fi

The authors would like to thank...

% Can use something like this to put references on a page
% by themselves when using endfloat and the captionsoff option.
\ifCLASSOPTIONcaptionsoff
  \newpage
\fi

% trigger a \newpage just before the given reference
% number - used to balance the columns on the last page
% adjust value as needed - may need to be readjusted if
% the document is modified later
%\IEEEtriggeratref{8}
% The "triggered" command can be changed if desired:
%\IEEEtriggercmd{\enlargethispage{-5in}}

% references section

% can use a bibliography generated by BibTeX as a .bbl file
% BibTeX documentation can be easily obtained at:
% http://www.ctan.org/tex-archive/biblio/bibtex/contrib/doc/
% The IEEEtran BibTeX style support page is at:
% http://www.michaelshell.org/tex/ieeetran/bibtex/
%\bibliographystyle{IEEEtran}
% argument is your BibTeX string definitions and bibliography database(s)
%\bibliography{IEEEabrv,../bib/paper}
%
% <OR> manually copy in the resultant .bbl file
% set second argument of \begin to the number of references
% (used to reserve space for the reference number labels box)
\begin{thebibliography}{1}

\bibitem{IEEEhowto:kopka}
H.~Kopka and P.~W. Daly, \emph{A Guide to {\LaTeX}}, 3rd~ed.\hskip 1em plus
  0.5em minus 0.4em\relax Harlow, England: Addison-Wesley, 1999.

\end{thebibliography}

% biography section
% 
% If you have an EPS/PDF photo (graphicx package needed) extra braces are
% needed around the contents of the optional argument to biography to prevent
% the LaTeX parser from getting confused when it sees the complicated
% \includegraphics command within an optional argument. (You could create
% your own custom macro containing the \includegraphics command to make things
% simpler here.)
%\begin{biography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{mshell}}]{Michael Shell}
% or if you just want to reserve a space for a photo:

\begin{IEEEbiography}{Michael Shell}
Biography text here.
\end{IEEEbiography}

% if you will not have a photo at all:
\begin{IEEEbiographynophoto}{John Doe}
Biography text here.
\end{IEEEbiographynophoto}

% insert where needed to balance the two columns on the last page with
% biographies
%\newpage

\begin{IEEEbiographynophoto}{Jane Doe}
Biography text here.
\end{IEEEbiographynophoto}

% You can push biographies down or up by placing
% a \vfill before or after them. The appropriate
% use of \vfill depends on what kind of text is
% on the last page and whether or not the columns
% are being equalized.

%\vfill

% Can be used to pull up biographies so that the bottom of the last one
% is flush with the other column.
%\enlargethispage{-5in}

% that's all folks
\end{document}

### 2012-03-02 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2012:03:01	20:12:52	(Thu 1.Mar)
la2pdf commit - 2012:03:01	22:12:40	(Thu 1.Mar)
la2pdf commit - 2012:03:02	13:12:53	(Fri 2.Mar)

@@
 %
-\documentclass[12pt,journal,compsoc]{ieee/IEEEtran}
+\documentclass[12pt,journal]{ieee/IEEEtran}
 % The Computer Society requires 12pt.
@@
 %
-% paper title
-% can use linebreaks \\ within to get better formatting as desired
-\title{Bare Advanced Demo of IEEEtran.cls\\ for Computer Society Journals}
-%
-%
-% author names and IEEE memberships
-% note positions of commas and nonbreaking spaces ( ~ ) LaTeX will not break
-% a structure at a ~ so this keeps an author's name from being broken across
-% two lines.
-% use \thanks{} to gain access to the first footnote area
-% a separate \thanks must be used for each paragraph as LaTeX2e's \thanks
-% was not built to handle multiple paragraphs
-%
-%
-%\IEEEcompsocitemizethanks is a special \thanks that produces the bulleted
-% lists the Computer Society journals use for "first footnote" author
-% affiliations. Use \IEEEcompsocthanksitem which works much like \item
-% for each affiliation group. When not in compsoc mode,
-% \IEEEcompsocitemizethanks becomes like \thanks and
-% \IEEEcompsocthanksitem becomes a line break with idention. This
-% facilitates dual compilation, although admittedly the differences in the
-% desired content of \author between the different types of papers makes a
-% one-size-fits-all approach a daunting prospect. For instance, compsoc 
-% journal papers have the author affiliations above the "Manuscript
-% received ..."  text while in non-compsoc journals this is reversed. Sigh.
+\title{Low Complexity Adaptive Beamformer\\ for Active Sonar Imaging}
 
-\author{Michael~Shell,~\IEEEmembership{Member,~IEEE,}
-        John~Doe,~\IEEEmembership{Fellow,~OSA,}
-        and~Jane~Doe,~\IEEEmembership{Life~Fellow,~IEEE}% <-this % stops a space
-\IEEEcompsocitemizethanks{\IEEEcompsocthanksitem M. Shell is with the Department
-of Electrical and Computer Engineering, Georgia Institute of Technology, Atlanta,
-GA, 30332.\protect\\
-% note need leading \protect in front of \\ to get a newline within \thanks as
-% \\ is fragile and will error, could use \hfil\break instead.
-E-mail: see http://www.michaelshell.org/contact.html
-\IEEEcompsocthanksitem J. Doe and J. Doe are with Anonymous University.}% <-this % stops a space
-\thanks{Manuscript received April 19, 2005; revised January 11, 2007.}}
+\author{Jo~Inge~Buskenes, %
+        Andreas~Austeng, %
+        Carl-Inge~Columbo~Nilsen%
+\IEEEcompsocitemizethanks{\IEEEcompsocthanksitem All authors are with the Department
+of Informatics, University of Oslo, Norway.}% <-this % stops a space
 
-% note the % following the last \IEEEmembership and also \thanks - 
-% these prevent an unwanted space from occurring between the last author name
-% and the end of the author line. i.e., if you had this:
-% 
-% \author{....lastname \thanks{...} \thanks{...} }
-%                     ^------------^------------^----Do not want these spaces!
-%
-% a space would be appended to the last name and could cause every name on that
-% line to be shifted left slightly. This is one of those "LaTeX things". For
-% instance, "\textbf{A} \textbf{B}" will typeset as "A B" not "AB". To get
-% "AB" then you have to do: "\textbf{A}\textbf{B}"
-% \thanks is no different in this regard, so shield the last } of each \thanks
-% that ends a line with a % and do not let a space in before the next \thanks.
-% Spaces after \IEEEmembership other than the last one are OK (and needed) as
-% you are supposed to have spaces between the names. For what it is worth,
-% this is a minor point as most people would not even notice if the said evil
-% space somehow managed to creep in.
+% \thanks{Manuscript received April 19, 2005; revised January 11, 2007.}
+}
 
 % The paper headers
-\markboth{Journal of \LaTeX\ Class Files,~Vol.~6, No.~1, January~2007}%
-{Shell \MakeLowercase{\textit{et al.}}: Bare Advanced Demo of IEEEtran.cls for Journals}
-% The only time the second header will appear is for the odd numbered pages
-% after the title page when using the twoside option.
-% 
-% *** Note that you probably will NOT want to include the author's ***
-% *** name in the headers of peer review papers.                   ***
-% You can use \ifCLASSOPTIONpeerreview for conditional compilation here if
-% you desire.
+\markboth{IEEE Journal of Oceanic Engineering}%
+{Low Complexity Adaptive Beamformer for Active Sonar Imaging}
 

### 2012-03-02 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2012:03:02	13:45:34	(Fri 2.Mar)
la2pdf commit - 2012:03:02	14:21:32	(Fri 2.Mar)

@@
-% Note that the a4paper option is mainly intended so that authors in
-% countries using A4 can easily print to A4 and see how their papers will
-% look in print - the typesetting of the document will not typically be
-% affected with changes in paper size (but the bottom and side margins will).
-% Use the testflow package mentioned above to verify correct handling of
-% both paper sizes by the user's LaTeX system.
-%
-% Also note that the "draftcls" or "draftclsnofoot", not "draft", option
-% should be used if it is desired that the figures are to be displayed in
-% draft mode.
-%
-\documentclass[12pt,journal]{ieee/IEEEtran}
-% The Computer Society requires 12pt.
-% If IEEEtran.cls has not been installed into the LaTeX system files,
-% manually specify the path to it like:
-% \documentclass[10pt,journal,compsoc]{../sty/IEEEtran}
+%%%%%%%%%%%%%%%%                                ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
+% CONDITIONALS %
+%%%%%%%%%%%%%%%%                                ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
 
-% For Computer Society journals, IEEEtran defaults to the use of 
-% Palatino/Palladio as is done in IEEE Computer Society journals.
-% To go back to Times Roman, you can use this code:
-%\renewcommand{\rmdefault}{ptm}\selectfont
+\newif\ifPeerReview \PeerReviewtrue             % Whether to create the PeerReview version or
+                                                % Journal version
+\newif\ifFlatArchive\FlatArchivefalse           % Whether archive is flat (messy) or contain 
+                                                % subfolders for graphics etc.
 
-% Some very useful LaTeX packages include:
-% (uncomment the ones you want to load)
+%%%%%%%%%%%%                                    ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
+% IEEEtran %
+%%%%%%%%%%%%                                    ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
 
-% *** MISC UTILITY PACKAGES ***
-%
-%\usepackage{ifpdf}
-% Heiko Oberdiek's ifpdf.sty is very useful if you need conditional
-% compilation based on whether the output is pdf or dvi.
-% usage:
-% \ifpdf
-%   % pdf code
-% \else
-%   % dvi code
-% \fi
-% The latest version of ifpdf.sty can be obtained from:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/oberdiek/
-% Also, note that IEEEtran.cls V1.7 and later provides a builtin
-% \ifCLASSINFOpdf conditional that works the same way.
-% When switching from latex to pdflatex and vice-versa, the compiler may
-% have to be run twice to clear warning/error messages.
+\ifPeerReview
+\documentclass[12pt,journal,onecolumn]{IEEEtran}
+\else
+\documentclass[journal]{IEEEtran}
+\fi
 
-% *** CITATION PACKAGES ***
-%
-\ifCLASSOPTIONcompsoc
-  % IEEE Computer Society needs nocompress option
-  % requires cite.sty v4.0 or later (November 2003)
-  % \usepackage[nocompress]{cite}
-\else
-  % normal IEEE
-  % \usepackage{cite}
-\fi
-% cite.sty was written by Donald Arseneau
-% V1.6 and later of IEEEtran pre-defines the format of the cite.sty package
-% \cite{} output to follow that of IEEE. Loading the cite package will
-% result in citation numbers being automatically sorted and properly
-% "compressed/ranged". e.g., [1], [9], [2], [7], [5], [6] without using
-% cite.sty will become [1], [2], [5]--[7], [9] using cite.sty. cite.sty's
-% \cite will automatically add leading space, if needed. Use cite.sty's
-% noadjust option (cite.sty V3.8 and later) if you want to turn this off.
-% cite.sty is already installed on most LaTeX systems. Be sure and use
-% version 4.0 (2003-05-27) and later if using hyperref.sty. cite.sty does
-% not currently provide for hyperlinked citations.
-% The latest version can be obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/cite/
-% The documentation is contained in the cite.sty file itself.
-%
-% Note that some packages require special options to format as the Computer
-% Society requires. In particular, Computer Society  papers do not use
-% compressed citation ranges as is done in typical IEEE papers
-% (e.g., [1]-[4]). Instead, they list every citation separately in order
-% (e.g., [1], [2], [3], [4]). To get the latter we need to load the cite
-% package with the nocompress option which is supported by cite.sty v4.0
-% and later. Note also the use of a CLASSOPTION conditional provided by
-% IEEEtran.cls V1.7 and later.
+%%%%%%%%%%%%                                    ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
+% PACKAGES %
+%%%%%%%%%%%%                                    ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
+
+% \input{preamble}
 
@@
 \ifCLASSINFOpdf
-  % \usepackage[pdftex]{graphicx}
-  % declare the path(s) where your graphic files are
-  % \graphicspath{{../pdf/}{../jpeg/}}
-  % and their extensions so you won't have to specify these with
-  % every instance of \includegraphics
-  % \DeclareGraphicsExtensions{.pdf,.jpeg,.png}
+   \usepackage[dvips]{graphicx}
+   \ifFlatArchive\else
+      \graphicspath{./gfx/}
+   \fi
 \else
-  % or other class option (dvipsone, dvipdf, if not using dvips). graphicx
-  % will default to the driver specified in the system graphics.cfg if no
-  % driver is specified.
-  % \usepackage[dvips]{graphicx}
-  % declare the path(s) where your graphic files are
-  % \graphicspath{{../eps/}}
-  % and their extensions so you won't have to specify these with
-  % every instance of \includegraphics
-  % \DeclareGraphicsExtensions{.eps}
+   \usepackage[dvips]{graphicx}
+   \ifFlatArchive\else
+      \graphicspath{./gfx/}
    \fi
-% graphicx was written by David Carlisle and Sebastian Rahtz. It is
-% required if you want graphics, photos, etc. graphicx.sty is already
-% installed on most LaTeX systems. The latest version and documentation can
-% be obtained at: 
-% http://www.ctan.org/tex-archive/macros/latex/required/graphics/
-% Another good source of documentation is "Using Imported Graphics in
-% LaTeX2e" by Keith Reckdahl which can be found as epslatex.ps or
-% epslatex.pdf at: http://www.ctan.org/tex-archive/info/
-%
-% latex, and pdflatex in dvi mode, support graphics in encapsulated
-% postscript (.eps) format. pdflatex in pdf mode supports graphics
-% in .pdf, .jpeg, .png and .mps (metapost) formats. Users should ensure
-% that all non-photo figures use a vector format (.eps, .pdf, .mps) and
-% not a bitmapped formats (.jpeg, .png). IEEE frowns on bitmapped formats
-% which can result in "jaggedy"/blurry rendering of lines and letters as
-% well as large increases in file sizes.
-%
-% You can find documentation about the pdfTeX application at:
-% http://www.tug.org/applications/pdftex
+\fi
 
-%\usepackage{ps4pdf}
-% dvi->ps workflow is required to use such packages as psfrag.sty and
-% pstricks.sty. However, Rolf Niepraschk's ps4pdf.sty provides a way to
-% apply psfrag/pstricks effects to .eps figures and then get the resultant
-% figures in .pdf form. Thus, providing an easier way for migrating from
-% .eps to .pdf figures. After ps4pdf.sty loads, if:
-% 1. producing .dvi output: the output file will consist ONLY of the
-%    figures (or other constructs encased within \PSforPDF commands)
-% 2. producing .pdf output: pdflatex will look in the filename-pics.pdf
-%    file, where filename is the basename of the tex document, for the
-%    graphics (or other constructs encased within \PSforPDF commands).
-%    NOTE: If you ever change your figures, you must remember to remake
-%    the filename-pics.pdf file.
-%
-% This way you can do a:
-% 
-% latex filename
-% dvips -Ppdf -o filename-pics.ps filename.dvi
-% ps2pdf filename-pics.ps filename-pics.pdf
-% 
-% to produce a filename-pics.pdf graphics container that contains
-% .pdf versions of the graphics with psfrag, pstricks, etc. features.
-% Note that you will not typically be able to view the figures in 
-% filename-pics.ps because of an offset. However, you will be able to
-% view them in filename-pics.pdf. Also, note that when ps4pdf is in effect
-% with .dvi output, you may get harmless over/under full box warnings - 
-% ignore them. 
-% Then, run pdflatex:
-% 
-% pdflatex filename
-% 
-% to use pdflatex to make PDF output, automatically using the figures in
-% filename-pics.pdf. Alternatively, you could use dvips -i option to
-% obtain separate .pdf files for each figure:
-%
-% dvips -Ppdf -i -E -o fig filename
-%
-% then convert each figure to pdf via a command such as epstopdf and then
-% use pdflatex with these pdf figures and then to dispense with ps4pdf.
-%
-% Remember to rerun through latex/dvips/ps2pdf if you ever change your
-% figures so that filename-pics.pdf gets updated.
-% ps4pdf requires David Kastrup's preview-latex and a recent LaTeX system
-% (circa 2001 or later). The ps4pdf package and documentation can be
-% obtained at: http://www.ctan.org/tex-archive/macros/latex/contrib/ps4pdf/
-% The preview-latex package and documentation can be obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/preview/
-%
-% provide a bogus \PSforPDF, even when not loading pd4pdf. This way we can
-% stop loading ps4pdf.sty if we choose to make separate .pdf versions of
-% each of our figures.
-\providecommand{\PSforPDF}[1]{#1}
-% Note that in order for ps4pdf to work, all commands related to psfrag,
-% pstricks, etc. must be called within the PSforPDF command. This applies
-% even when *loading* via \usepackage psfrag.sty, etc.
+\usepackage[cmex10]{amsmath}                    % cmex10 option to be IEEE explore compliant
+\interdisplaylinepenalty=2500                   % Allows multiline equations to be broken
 
-%\PSforPDF{\usepackage{psfrag}}
-% psfrag.sty was written by Craig Barratt, Michael C. Grant, and
-% David Carlisle. It allows you to substitute LaTeX commands for text in
-% imported EPS graphic files. In this way, LaTeX symbols can be placed into
-% graphics that have been generated by other applications. You must use
-% latex->dvips->ps2pdf workflow (not direct pdf output from pdflatex) if
-% you wish to use this capability because it works via some PostScript
-% tricks. Alternatively, the graphics could be processed as separate files
-% via psfrag and dvips, then converted to PDF for inclusion in the main file
-% which uses pdflatex. ps4pdf.sty (above) provides a way of doing this all
-% at once within the main file.
-% Docs are in "The PSfrag System" by Michael C. Grant and David Carlisle.
-% There is also some information about using psfrag in "Using Imported
-% Graphics in LaTeX2e" by Keith Reckdahl which documents the graphicx
-% package (see above). The psfrag package and documentation can be obtained
-% at: http://www.ctan.org/tex-archive/macros/latex/contrib/psfrag/
-% 
-% Note that the current version of psfrag does not "turn itself off" when
-% running under pdf output. This will result in a harmless warning
-% about a non-PDF \special. However, to silence this, a bogus psfrag
-% command can be provided instead of loading psfrag.sty when PDF output
-% is being used. Thus, a more complex alternative conditional loading scheme
-% can be employed instead of the straightforword way above:
-%
-%\ifCLASSINFOpdf
-% if outputting PDF, do not use or load psfrag.sty as current versions
-% output a non-PDF special that generates a harmless, but annoying warning.
-% Instead, we provide a bogus \psfrag command that does nothing with
-% its arguments. This is a tad tricky because \psfrag can have up to six
-% arguments four of which are optional: \psfrag{}[][][][]{}
-% Code based on that in psfrag.sty
-%\makeatletter
-%\def\psfrag{\@ifstar{\@BOGUSpsfraga}{\@BOGUSpsfraga}}
-%\def\@BOGUSpsfraga{\begingroup
-%   \@makeother\"\@makeother\*\@makeother\!\@makeother\~%
-%   \@makeother\:\@makeother\\\@makeother\%\@makeother\#%
-%   \@makeother\ \@BOGUSpsfragb}
-%\def\@BOGUSpsfragb#1{\endgroup
-%                \@ifnextchar [{\@BOGUSpsfragc}%
-%                              {\@BOGUSpsfrag}}
-%\def\@BOGUSpsfragc[#1]{\@ifnextchar [{\@BOGUSpsfragd}%
-%                                     {\@BOGUSpsfrag}}
-%\def\@BOGUSpsfragd[#1]{\@ifnextchar [{\@BOGUSpsfrage}%
-%                                     {\@BOGUSpsfrag}}
-%\def\@BOGUSpsfrage[#1]{\@ifnextchar [{\@BOGUSpsfragf}%
-%                                     {\@BOGUSpsfrag}}
-%\def\@BOGUSpsfragf[#1]{\@BOGUSpsfrag}
-%\def\@BOGUSpsfrag#1{\ignorespaces}
-%\makeatother
-%\else
-% using dvi output, load psfrag, but funnel it through PSforPDF
-% as required by ps4pdf.sty
-%\PSforPDF{\usepackage{psfrag}}
-%\fi
+% \RequirePackage{amssymb}
 
-% *** MATH PACKAGES ***
-%
-%\usepackage[cmex10]{amsmath}
-% A popular package from the American Mathematical Society that provides
-% many useful and powerful commands for dealing with mathematics. If using
-% it, be sure to load this package with the cmex10 option to ensure that
-% only type 1 fonts will utilized at all point sizes. Without this option,
-% it is possible that some math symbols, particularly those within
-% footnotes, will be rendered in bitmap form which will result in a
-% document that can not be IEEE Xplore compliant!
-%
-% Also, note that the amsmath package sets \interdisplaylinepenalty to 10000
-% thus preventing page breaks from occurring within multiline equations. Use:
-%\interdisplaylinepenalty=2500
-% after loading amsmath to restore such page breaks as IEEEtran.cls normally
-% does. amsmath.sty is already installed on most LaTeX systems. The latest
-% version and documentation can be obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/required/amslatex/math/
+\RequirePackage{array}
 
-% *** SPECIALIZED LIST PACKAGES ***
-%\usepackage{acronym}
-% acronym.sty was written by Tobias Oetiker. This package provides tools for
-% managing documents with large numbers of acronyms. (You don't *have* to
-% use this package - unless you have a lot of acronyms, you may feel that
-% such package management of them is bit of an overkill.)
-% Do note that the acronym environment (which lists acronyms) will have a
-% problem when used under IEEEtran.cls because acronym.sty relies on the
-% description list environment - which IEEEtran.cls has customized for
-% producing IEEE style lists. A workaround is to declared the longest
-% label width via the IEEEtran.cls \IEEEiedlistdecl global control:
-%
-% \renewcommand{\IEEEiedlistdecl}{\IEEEsetlabelwidth{SONET}}
-% \begin{acronym}
-%
-% \end{acronym}
-% \renewcommand{\IEEEiedlistdecl}{\relax}% remember to reset \IEEEiedlistdecl
-%
-% instead of using the acronym environment's optional argument.
-% The latest version and documentation can be obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/acronym/
-
-%\usepackage{algorithmic}
-% algorithmic.sty was written by Peter Williams and Rogerio Brito.
-% This package provides an algorithmic environment fo describing algorithms.
-% You can use the algorithmic environment in-text or within a figure
-% environment to provide for a floating algorithm. Do NOT use the algorithm
-% floating environment provided by algorithm.sty (by the same authors) or
-% algorithm2e.sty (by Christophe Fiorio) as IEEE does not use dedicated
-% algorithm float types and packages that provide these will not provide
-% correct IEEE style captions. The latest version and documentation of
-% algorithmic.sty can be obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/algorithms/
-% There is also a support site at:
-% http://algorithms.berlios.de/index.html
-% Also of interest may be the (relatively newer and more customizable)
-% algorithmicx.sty package by Szasz Janos:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/algorithmicx/
-
-% *** ALIGNMENT PACKAGES ***
-%
-%\usepackage{array}
-% Frank Mittelbach's and David Carlisle's array.sty patches and improves
-% the standard LaTeX2e array and tabular environments to provide better
-% appearance and additional user controls. As the default LaTeX2e table
-% generation code is lacking to the point of almost being broken with
-% respect to the quality of the end results, all users are strongly
-% advised to use an enhanced (at the very least that provided by array.sty)
-% set of table tools. array.sty is already installed on most systems. The
-% latest version and documentation can be obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/required/tools/
-
-%\usepackage{mdwmath}
-%\usepackage{mdwtab}
-% Also highly recommended is Mark Wooding's extremely powerful MDW tools,
-% especially mdwmath.sty and mdwtab.sty which are used to format equations
-% and tables, respectively. The MDWtools set is already installed on most
-% LaTeX systems. The lastest version and documentation is available at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/mdwtools/
-
-% IEEEtran contains the IEEEeqnarray family of commands that can be used to
-% generate multiline equations as well as matrices, tables, etc., of high
-% quality.
-
-%\usepackage{eqparbox}
-% Also of notable interest is Scott Pakin's eqparbox package for creating
-% (automatically sized) equal width boxes - aka "natural width parboxes".
-% Available at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/eqparbox/
-
-% *** SUBFIGURE PACKAGES ***
-%\ifCLASSOPTIONcompsoc
-%\usepackage[tight,normalsize,sf,SF]{subfigure}
-%\else
-%\usepackage[tight,footnotesize]{subfigure}
-%\fi
+\ifCLASSOPTIONcompsoc
+\usepackage[tight,normalsize,sf,SF]{subfigure}
+\else
+\usepackage[tight,footnotesize]{subfigure}
+\fi
 % subfigure.sty was written by Steven Douglas Cochran. This package makes it

### 2012-03-02 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2012:03:02	15:06:39	(Fri 2.Mar)

@@
 \ifPeerReview
-\documentclass[12pt,journal,onecolumn]{IEEEtran}
+\documentclass[12pt,journal,captionsoff,onecolumn]{IEEEtran}
+\newcommand\CLASSINPUTbaselinestretch{1.66}
 \else
@@
 \ifCLASSINFOpdf
-   \usepackage[dvips]{graphicx}
-   \ifFlatArchive\else
+   \usepackage[dvips]{graphicx}                 % Might not work. Use 'latex' instead of 
+   \ifFlatArchive\else                          % 'pdflatex'
       \graphicspath{./gfx/}
@@
 \ifCLASSOPTIONcompsoc
-\usepackage[tight,normalsize,sf,SF]{subfigure}
+   \usepackage[caption=false,font=normalsize,labelfont=sf,textfont=sf]{subfig}
 \else
-\usepackage[tight,footnotesize]{subfigure}
+   \usepackage[caption=false,font=footnotesize]{subfig}
 \fi
-% subfigure.sty was written by Steven Douglas Cochran. This package makes it
-% easy to put subfigures in your figures. e.g., "Figure 1a and 1b". For IEEE
-% work, it is a good idea to load it with the tight package option to reduce
-% the amount of white space around the subfigures. Computer Society papers
-% use a larger font and \sffamily font for their captions, hence the
-% additional options needed under compsoc mode. subfigure.sty is already
-% installed on most LaTeX systems. The latest version and documentation can
-% be obtained at:
-% http://www.ctan.org/tex-archive/obsolete/macros/latex/contrib/subfigure/
-% subfigure.sty has been superceeded by subfig.sty.
+\ifCLASSOPTIONcaptionsoff                       % IEEE promoted hack to turn off captions from the 
+   \let\MYorigsubfloat\subfloat                 % subfloat package should the captionsoff option
+   \renewcommand{\subfloat}[2][\relax]{\MYorigsubfloat[]{#2}} % be specified.
+\fi
 
-%\ifCLASSOPTIONcompsoc
-%  \usepackage[caption=false]{caption}
-%  \usepackage[font=normalsize,labelfont=sf,textfont=sf]{subfig}
-%\else
-%  \usepackage[caption=false]{caption}
-%  \usepackage[font=footnotesize]{subfig}
-%\fi
-% subfig.sty, also written by Steven Douglas Cochran, is the modern
-% replacement for subfigure.sty. However, subfig.sty requires and
-% automatically loads Axel Sommerfeldt's caption.sty which will override
-% IEEEtran.cls handling of captions and this will result in nonIEEE style
-% figure/table captions. To prevent this problem, be sure and preload
-% caption.sty with its "caption=false" package option. This is will preserve
-% IEEEtran.cls handing of captions. Version 1.3 (2005/06/28) and later 
-% (recommended due to many improvements over 1.2) of subfig.sty supports
-% the caption=false option directly:
-%\ifCLASSOPTIONcompsoc
-%  \usepackage[caption=false,font=normalsize,labelfont=sf,textfont=sf]{subfig}
-%\else
-%  \usepackage[caption=false,font=footnotesize]{subfig}
-%\fi
-%
-% The latest version and documentation can be obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/subfig/
-% The latest version and documentation of caption.sty can be obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/caption/
+\ifCLASSOPTIONcaptionsoff                       % Places float at the end of the document when the
+  \usepackage[nomarkers]{endfloat}              % captionsoff options is specified to IEEEtrans.cls
+  \let\MYoriglatexcaption\caption               % (PeerReview mode)
+  \renewcommand{\caption}[2][\relax]{\MYoriglatexcaption[#2]{#2}}
+\fi
 
-% *** FLOAT PACKAGES ***
-%
-%\usepackage{fixltx2e}
-% fixltx2e, the successor to the earlier fix2col.sty, was written by
-% Frank Mittelbach and David Carlisle. This package corrects a few problems
-% in the LaTeX2e kernel, the most notable of which is that in current
-% LaTeX2e releases, the ordering of single and double column floats is not
-% guaranteed to be preserved. Thus, an unpatched LaTeX2e can allow a
-% single column figure to be placed prior to an earlier double column
-% figure. The latest version and documentation can be found at:
-% http://www.ctan.org/tex-archive/macros/latex/base/
+\usepackage{fixltx2e}                           % Fix some twocolumn float problems
 
-%\usepackage{stfloats}
-% stfloats.sty was written by Sigitas Tolusis. This package gives LaTeX2e
-% the ability to do double column floats at the bottom of the page as well
-% as the top. (e.g., "\begin{figure*}[!b]" is not normally possible in
-% LaTeX2e). It also provides a command:
-%\fnbelowfloat
-% to enable the placement of footnotes below bottom floats (the standard
-% LaTeX2e kernel puts them above bottom floats). This is an invasive package
-% which rewrites many portions of the LaTeX2e float routines. It may not work
-% with other packages that modify the LaTeX2e float routines. The latest
-% version and documentation can be obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/sttools/
-% Documentation is contained in the stfloats.sty comments as well as in the
-% presfull.pdf file. Do not use the stfloats baselinefloat ability as IEEE
-% does not allow \baselineskip to stretch. Authors submitting work to the
-% IEEE should note that IEEE rarely uses double column equations and
-% that authors should try to avoid such use. Do not be tempted to use the
-% cuted.sty or midfloat.sty packages (also by Sigitas Tolusis) as IEEE does
-% not format its papers in such ways.
+%\usepackage{stfloats}                          % Allows: \begin{figure*}[!b]
+                                                % (double column figures on top/bottom)
 
-%\ifCLASSOPTIONcaptionsoff
-%  \usepackage[nomarkers]{endfloat}
-% \let\MYoriglatexcaption\caption
-% \renewcommand{\caption}[2][\relax]{\MYoriglatexcaption[#2]{#2}}
-%\fi
-% endfloat.sty was written by James Darrell McCauley and Jeff Goldberg.
 % This package may be useful when used in conjunction with IEEEtran.cls'
@@
 % For subfig.sty:
+
 % \let\MYorigsubfloat\subfloat
 % \renewcommand{\subfloat}[2][\relax]{\MYorigsubfloat[]{#2}}
+
 % For subfigure.sty:

### 2012-03-02 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2012:03:02	16:44:35	(Fri 2.Mar)

@@
 
-\newif\ifPeerReview \PeerReviewtrue             % Whether to create the PeerReview version or
+\newif\ifPeerReview \PeerReviewfalse            % Whether to create the PeerReview version or
                                                 % Journal version
@@
                                                 % subfolders for graphics etc.
+\newif\ifFloatAtEnd\FloatAtEndfalse             % Available in PeerReview mode:
+                                                % Place floats at end of document?
 
@@
 \documentclass[12pt,journal,captionsoff,onecolumn]{IEEEtran}
-\newcommand\CLASSINPUTbaselinestretch{1.66}
+\newcommand\CLASSINPUTbaselinestretch{1.66}     % http://theoval.cmp.uea.ac.uk/~nlct/latex/thesis/node17.html
 \else
@@
 
+\ifFloatAtEnd
 \ifCLASSOPTIONcaptionsoff                       % Places float at the end of the document when the
@@
   \renewcommand{\caption}[2][\relax]{\MYoriglatexcaption[#2]{#2}}
+\fi
 \fi
@@
 
-% This package may be useful when used in conjunction with IEEEtran.cls'
-% captionsoff option. Some IEEE journals/societies require that submissions
-% have lists of figures/tables at the end of the paper and that
-% figures/tables without any captions are placed on a page by themselves at
-% the end of the document. If needed, the draftcls IEEEtran class option or
-% \CLASSINPUTbaselinestretch interface can be used to increase the line
-% spacing as well. Be sure and use the nomarkers option of endfloat to
-% prevent endfloat from "marking" where the figures would have been placed
-% in the text. The two hack lines of code above are a slight modification of
-% that suggested by in the endfloat docs (section 8.3.1) to ensure that
-% the full captions always appear in the list of figures/tables - even if
-% the user used the short optional argument of \caption[]{}.
-% IEEE papers do not typically make use of \caption[]'s optional argument,
-% so this should not be an issue. A similar trick can be used to disable
-% captions of packages such as subfig.sty that lack options to turn off
-% the subcaptions:
-% For subfig.sty:
-
-% \let\MYorigsubfloat\subfloat
-% \renewcommand{\subfloat}[2][\relax]{\MYorigsubfloat[]{#2}}
-
-% For subfigure.sty:
-% \let\MYorigsubfigure\subfigure
-% \renewcommand{\subfigure}[2][\relax]{\MYorigsubfigure[]{#2}}
-% However, the above trick will not work if both optional arguments of
-% the \subfloat/subfig command are used. Furthermore, there needs to be a
-% description of each subfigure *somewhere* and endfloat does not add
-% subfigure captions to its list of figures. Thus, the best approach is to
-% avoid the use of subfigure captions (many IEEE journals avoid them anyway)
-% and instead reference/explain all the subfigures within the main caption.
-% The latest version of endfloat.sty and its documentation can obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/endfloat/
-%
-% The IEEEtran \ifCLASSOPTIONcaptionsoff conditional can also be used
-% later in the document, say, to conditionally put the References on a 
-% page by themselves.
-
-% *** PDF, URL AND HYPERLINK PACKAGES ***
-%
-%\usepackage{url}
-% url.sty was written by Donald Arseneau. It provides better support for
-% handling and breaking URLs. url.sty is already installed on most LaTeX
-% systems. The latest version can be obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/misc/
-% Read the url.sty source comments for usage information. Basically,
-% \url{my_url_here}.
-
-% NOTE: PDF thumbnail features are not required in IEEE papers
-%       and their use requires extra complexity and work.
-%\ifCLASSINFOpdf
-%  \usepackage[pdftex]{thumbpdf}
-%\else
-%  \usepackage[dvips]{thumbpdf}
-%\fi
-% thumbpdf.sty and its companion Perl utility were written by Heiko Oberdiek.
-% It allows the user a way to produce PDF documents that contain fancy
-% thumbnail images of each of the pages (which tools like acrobat reader can
-% utilize). This is possible even when using dvi->ps->pdf workflow if the
-% correct thumbpdf driver options are used. thumbpdf.sty incorporates the
-% file containing the PDF thumbnail information (filename.tpm is used with
-% dvips, filename.tpt is used with pdftex, where filename is the base name of
-% your tex document) into the final ps or pdf output document. An external
-% utility, the thumbpdf *Perl script* is needed to make these .tpm or .tpt
-% thumbnail files from a .ps or .pdf version of the document (which obviously
-% does not yet contain pdf thumbnails). Thus, one does a:
-% 
-% thumbpdf filename.pdf 
-%
-% to make a filename.tpt, and:
-%
-% thumbpdf --mode dvips filename.ps
-%
-% to make a filename.tpm which will then be loaded into the document by
-% thumbpdf.sty the NEXT time the document is compiled (by pdflatex or
-% latex->dvips->ps2pdf). Users must be careful to regenerate the .tpt and/or
-% .tpm files if the main document changes and then to recompile the
-% document to incorporate the revised thumbnails to ensure that thumbnails
-% match the actual pages. It is easy to forget to do this!
-% 
-% Unix systems come with a Perl interpreter. However, MS Windows users
-% will usually have to install a Perl interpreter so that the thumbpdf
-% script can be run. The Ghostscript PS/PDF interpreter is also required.
-% See the thumbpdf docs for details. The latest version and documentation
-% can be obtained at.
-% http://www.ctan.org/tex-archive/support/thumbpdf/
-% Be sure and use only version 3.8 (2005/07/06) or later of thumbpdf as
-% earlier versions will not work properly with recent versions of pdfTeX
-% (1.20a and later).
+\usepackage{url}                                % Support for handling and breaking URLs
 
@@
 %       papers and their use requires extra complexity and work.
-% *** IF USING HYPERREF BE SURE AND CHANGE THE EXAMPLE PDF ***
-% *** TITLE/SUBJECT/AUTHOR/KEYWORDS INFO BELOW!!           ***
 \newcommand\MYhyperrefoptions{bookmarks=true,bookmarksnumbered=true,
@@
 urlcolor={black},
-pdftitle={Bare Demo of IEEEtran.cls for Computer Society Journals},%<!CHANGE!
-pdfsubject={Typesetting},%<!CHANGE!
-pdfauthor={Michael D. Shell},%<!CHANGE!
-pdfkeywords={Computer Society, IEEEtran, journal, LaTeX, paper,
-             template}}%<^!CHANGE!
-%\ifCLASSINFOpdf
-%\usepackage[\MYhyperrefoptions,pdftex]{hyperref}
-%\else
-%\usepackage[\MYhyperrefoptions,breaklinks=true,dvips]{hyperref}
-%\usepackage{breakurl}
-%\fi
-% One significant drawback of using hyperref under DVI output is that the
-% LaTeX compiler cannot break URLs across lines or pages as can be done
-% under pdfLaTeX's PDF output via the hyperref pdftex driver. This is
-% probably the single most important capability distinction between the
-% DVI and PDF output. Perhaps surprisingly, all the other PDF features
-% (PDF bookmarks, thumbnails, etc.) can be preserved in
-% .tex->.dvi->.ps->.pdf workflow if the respective packages/scripts are
-% loaded/invoked with the correct driver options (dvips, etc.). 
-% As most IEEE papers use URLs sparingly (mainly in the references), this
-% may not be as big an issue as with other publications.
-%
-% That said, recently Vilar Camara Neto introduced his breakurl.sty
-% package which permits hyperref to easily break URLs even in dvi
-% mode. Note that breakurl, unlike most other packages, must be loaded
-% AFTER hyperref. The latest version of breakurl and its documentation can
-% be obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/breakurl/
-% breakurl.sty is not for use under pdflatex pdf mode. Versions 1.10 
-% (September 23, 2005) and later are recommened to avoid bugs in earlier
-% releases.
-%
-% The advanced features offer by hyperref.sty are not required for IEEE
-% submission, so users should weigh these features against the added
-% complexity of use. Users who wish to use hyperref *must* ensure that
-% their hyperref version is 6.72u or later *and* IEEEtran.cls is version
-% 1.6b or later.
-% The package options above demonstrate how to enable PDF bookmarks
-% (a type of table of contents viewable in Acrobat Reader) as well as
-% PDF document information (title, subject, author and keywords) that is
-% viewable in Acrobat reader's Document_Properties menu. PDF document
-% information is also used extensively to automate the cataloging of PDF
-% documents. The above set of options ensures that hyperlinks will not be
-% colored in the text and thus will not be visible in the printed page,
-% but will be active on "mouse over". USING COLORS OR OTHER HIGHLIGHTING
-% OF HYPERLINKS CAN RESULT IN DOCUMENT REJECTION BY THE IEEE, especially if
-% these appear on the "printed" page. IF IN DOUBT, ASK THE RELEVANT
-% SUBMISSION EDITOR. You may need to add the option hypertexnames=false if
-% you used duplicate equation numbers, etc., but this should not be needed
-% in normal IEEE work.
-% The latest version of hyperref and its documentation can be obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/hyperref/
-
-% *** Do not adjust lengths that control margins, column widths, etc. ***
-% *** Do not use packages that alter fonts (such as pslatex).         ***
-% There should be no need to do such things with IEEEtran.cls V1.6 and later.
-% (Unless specifically asked to do so by the journal or conference you plan
-% to submit to, of course. )
+pdftitle={Low Complexity Adaptive Beamformer for Active Sonar Imaging},
+pdfsubject={},
+pdfauthor={Jo Inge Buskenes},
+pdfkeywords={adaptive beamforming, beamforming, complexity, sonar, active}}%
+\ifCLASSINFOpdf
+   \usepackage[\MYhyperrefoptions,pdftex]{hyperref}
+\else
+   \usepackage[\MYhyperrefoptions,breaklinks=true,dvips]{hyperref}
+   \usepackage{breakurl}                        % Allows 'dvips' driver to break links
+\fi
 

### 2012-03-05 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2012:03:05	12:13:43	(Mon 5.Mar)

@@
 pdfpagemode={UseOutlines},plainpages=false,pdfpagelabels=true,
-colorlinks=true,linkcolor={black},citecolor={black},pagecolor={black},
-urlcolor={black},
+colorlinks=true,linkcolor={black},citecolor={black},urlcolor={black},
 pdftitle={Low Complexity Adaptive Beamformer for Active Sonar Imaging},

### 2012-03-05 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2012:03:05	12:50:32	(Mon 5.Mar)

@@
                                                 % Place floats at end of document?
+\newif\ifTODO\TODOtrue                          % Use todo notes?
 
@@
 
-%%%%%%%%%%%%                                    ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
-% PACKAGES %
-%%%%%%%%%%%%                                    ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%                  ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
+% IEEE ''APPROVED'' PACKAGES %
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%                  ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
 
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%                         ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
+% ADDITIONAL PACKAGES %       
+%%%%%%%%%%%%%%%%%%%%%%%                         ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
+
+\RequirePackage[table,dvipsnames,svgnames]{xcolor}
+
+\ifTODO
+   \definecolor{lightblue}{rgb}{0.8,0.8,1}
+   \usepackage[backgroundcolor=LightBlue]{todonotes}
+\else
+   \usepackage[disable]{todonotes} 
+\fi
+
 % correct bad hyphenation here
-\hyphenation{op-tical net-works semi-conduc-tor}
+% \hyphenation{op-tical net-works semi-conduc-tor}
+
+%%%%%%%%%%%%%%%%%%                              ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
+% DOCUMENT START %
+%%%%%%%%%%%%%%%%%%                              ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
 
@@
 
-% The publisher's ID mark at the bottom of the page is less important with
-% Computer Society journal papers as those publications place the marks
-% outside of the main text columns and, therefore, unlike regular IEEE
-% journals, the available text space is not reduced by their presence.
-% If you want to put a publisher's ID mark on the page you can do it like
-% this:
+% Publishers ID mark:
 %\IEEEpubid{0000--0000/00\$00.00~\copyright~2007 IEEE}
-% or like this to get the Computer Society new two part style.
-%\IEEEpubid{\makebox[\columnwidth]{\hfill 0000--0000/00/\$00.00~\copyright~2007 IEEE}%
-%\hspace{\columnsep}\makebox[\columnwidth]{Published by the IEEE Computer Society\hfill}}
-% Remember, if you use this you must call \IEEEpubidadjcol in the second
-% column for its text to clear the IEEEpubid mark (Computer Society jorunal
-% papers don't need this extra clearance.)
 
@@
 \begin{abstract}
-%\boldmath
-The abstract goes here.
+The angular resolution and contrast in active sonar images depend on the beamformer's ability to receive signals from directions of interest, while suppressing noise and interference emanating from other directions.
+For sonar arrays, this is achieved by applying weights to the array channels.
+
+While classical beamformers use predefined windows, adaptive beamformers estimate the optimal window by analytical evaluation of the data.
+The minimum variance (MV) beamformer, for instance, calculates the set of weights that minimises the variance of the beamformer's output.
+
+We have implemented a Low Complexity Adaptive (LCA) beamformer, which adaptively selects a window from a predefined set.
+The set is comprised of windows that are typical solutions found by the Minimum Variance method.
+The LCA beamformer was tested using simulated and experimental data from the Kongsberg Maritime HISAS 1030 sonar.
+On a simulated scene with speckle, highlight and shadow, the beamformer offered better lateral edge definition compared to the MV beamformer, and speckle intensity and shape comparable to DAS and MV beamformers.
+These results were verified by the experimental data.
+
+An attactive trait of the LCA is its low computational complexity; while the MV beamformer is of O($M^3$), the LCA method is of O($MW$), with $M$ being the number of channels and $W$ the number of windows.
+We made the LCA perform like the MV method using a well designed set of 30 windows.
+Hence, unless the array is very small, the proposed method will perform like the MV beamformer or better, and at a fraction of the computational cost.
 \end{abstract}
-% IEEEtran.cls defaults to using nonbold math in the Abstract.
-% This preserves the distinction between vectors and scalars. However,
-% if the journal you are submitting to favors bold math in the abstract,
-% then you can use LaTeX's standard command \boldmath at the very start
-% of the abstract to achieve this. Many IEEE journals frown on math
-% in the abstract anyway. In particular, the Computer Society does
-% not want either math or citations to appear in the abstract.
 
-% Note that keywords are not normally used for peerreview papers.
+% Keywords (normally not used for peer reviews)
+\ifPeerReview\else
 \begin{IEEEkeywords}
-Computer Society, IEEEtran, journal, \LaTeX, paper, template.
-\end{IEEEkeywords}}
+Beamforming, adaptive beamforming, sonar, active, complexity.
+\end{IEEEkeywords}
+\fi}
+% \fi
 
@@
 
-% To allow for easy dual compilation without having to reenter the
-% abstract/keywords data, the \IEEEcompsoctitleabstractindextext text will
-% not be used in maketitle, but will appear (i.e., to be "transported")
-% here as \IEEEdisplaynotcompsoctitleabstractindextext when compsoc mode
-% is not selected <OR> if conference mode is selected - because compsoc
-% conference papers position the abstract like regular (non-compsoc)
-% papers do!
+% This command fixes abstract positioning for compsoc articles:
 \IEEEdisplaynotcompsoctitleabstractindextext
-% \IEEEdisplaynotcompsoctitleabstractindextext has no effect when using
-% compsoc under a non-conference mode.
 
-% For peer review papers, you can put extra information on the cover
-% page as needed:
+% (Optional) Add some extra info on cover page of peer review papers:
 % \ifCLASSOPTIONpeerreview
@@
 % \fi
-%
-% For peerreview papers, this IEEEtran command inserts a page break and
-% creates the second title. It will be ignored for other modes.
+
+% Insert page break and insert second title (peer review mode)
 \IEEEpeerreviewmaketitle
@@
 
-% The very first letter is a 2 line initial drop letter followed
-% by the rest of the first word in caps (small caps for compsoc).
-% 
-% form to use if the first word consists of a single letter:
-% \IEEEPARstart{A}{demo} file is ....
-% 
-% form to use if you need the single drop letter followed by
-% normal text (unknown if ever used by IEEE):
-% \IEEEPARstart{A}{}demo file is ....
-% 
-% Some journals put the first two words in caps:
-% \IEEEPARstart{T}{his demo} file is ....
-% 
-% Here we have the typical use of a "T" for an initial drop letter
-% and "HIS" in caps to complete the first word.
-\IEEEPARstart{T}{his} demo file is intended to serve as a ``starter file''
+\IEEEPARstart{A}{daptive} beamformers have only recently been introduced in active sonar imaging.\todo[inline]{break} This is because , as demo file is intended to serve as a ``starter file''
 for IEEE Computer Society journal papers produced under \LaTeX\ using

### 2012-03-05 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2012:03:05	13:28:57	(Mon 5.Mar)
la2pdf commit - 2012:03:05	14:00:44	(Mon 5.Mar)

@@
 
-\newif\ifPeerReview \PeerReviewfalse            % Whether to create the PeerReview version or
+\newif\ifPeerReview\PeerReviewfalse             % Whether to create the PeerReview version or
                                                 % Journal version
@@
    \fi
+\fi
+
+\RequirePackage[table,dvipsnames,svgnames]{xcolor}
+
+\ifTODO
+   \setlength{\marginparwidth}{1cm}
+%    \setlength{\marginparpush}{1cm}
+   \setlength\textwidth{\columnwidth}
+   \addtolength\marginparsep{1.5cm}
+   \definecolor{lightblue}{rgb}{0.8,0.8,1}
+   \usepackage[backgroundcolor=LightBlue]{todonotes}
+\else
+   \usepackage[disable]{todonotes} 
 \fi
@@
 
-\RequirePackage[table,dvipsnames,svgnames]{xcolor}
-
-\ifTODO
-   \definecolor{lightblue}{rgb}{0.8,0.8,1}
-   \usepackage[backgroundcolor=LightBlue]{todonotes}
-\else
-   \usepackage[disable]{todonotes} 
-\fi
-
 % correct bad hyphenation here
@@
 %%%%%%%%%%%%%%%%%%                              ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
-
+  \RequirePackage{layout}  
 \begin{document}
+\layout
 %
@@
 
-\IEEEPARstart{A}{daptive} beamformers have only recently been introduced in active sonar imaging.\todo[inline]{break} This is because , as demo file is intended to serve as a ``starter file''
+\IEEEPARstart{A}{daptive} beamformers have only recently been introduced in active sonar imaging.{\todo{break at something}} This is because , as demo file is intended to serve as a ``starter file''
 for IEEE Computer Society journal papers produced under \LaTeX\ using

### 2012-03-05 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2012:03:05	14:46:57	(Mon 5.Mar)

@@
 %    \setlength{\marginparpush}{1cm}
-   \setlength\textwidth{\columnwidth}
-   \addtolength\marginparsep{1.5cm}
-   \definecolor{lightblue}{rgb}{0.8,0.8,1}
-   \usepackage[backgroundcolor=LightBlue]{todonotes}
+%    \setlength\marginpar{1.5cm}
+   \addtolength\marginparsep{0.1cm}
+%    \definecolor{lightblue}{rgb}{0.8,0.8,1}
+%    \newlength\textwidthold
+%    \setlength\textwidthold{\textwidth}
+%    \setlength\textwidth{\columnwidth}
+   \usepackage{./todonotes}
+%    \setlength\textwidth\textwidthold
 \else
@@
 
-\IEEEPARstart{A}{daptive} beamformers have only recently been introduced in active sonar imaging.{\todo{break at something}} This is because , as demo file is intended to serve as a ``starter file''
+\IEEEPARstart{A}{daptive} beamformers have only recently been introduced in active sonar imaging.\todo{break at something} This is because , as demo file is intended to serve as a ``starter file''
 for IEEE Computer Society journal papers produced under \LaTeX\ using
 IEEEtran.cls version 1.7 and later.
-% You must have at least 2 lines in the paragraph with the drop letter
-% (should never be an issue)
-I wish you the best of success.
+
+We have implemented a Low Complexity Adaptive (LCA) beamformer, which adaptively selects a window from a predefined set.
+The set is comprised of windows that are typical solutions found by the Minimum Variance method.
+The LCA beamformer was tested using simulated and experimental data from the Kongsberg Maritime HISAS 1030 sonar.
+On a simulated scene with speckle, highlight and shadow, the beamformer offered better lateral edge definition compared to the MV beamformer, and speckle intensity and shape comparable to DAS and MV beamformers.
+These results were verified by the experimental data.
+
+An attactive trait of the LCA is its low computational complexity; while the MV beamformer is of O($M^3$), the LCA method is of O($MW$), with $M$ being the number of channels and $W$ the number of windows.
+We made the LCA perform like the MV method using a well designed set of 30 windows.
+Hence, unless the array is very small, the proposed method will perform like the MV beamformer or better, and at a fraction of the computational cost.
+
+We have implemented a Low Complexity Adaptive (LCA) beamformer, which adaptively selects a window from a predefined set.
+The set is comprised of windows that are typical solutions found by the Minimum Variance method.
+The LCA beamformer was tested using simulated and experimental data from the Kongsberg Maritime HISAS 1030 sonar.
+On a simulated scene with speckle, highlight and shadow, the beamformer offered better lateral edge definition compared to the MV beamformer, and speckle intensity and shape comparable to DAS and MV beamformers.
+These results were verified by the experimental data.
+
+An attactive trait of the LCA\todo{break at something} is its low computational complexity; while the MV beamformer is of O($M^3$), the LCA method is of O($MW$), with $M$ being the number of channels and $W$ the number of windows.
+We made the LCA perform like the MV method using a well designed set of 30 windows.
+Hence, unless the array is very small, the proposed method will perform like the MV beamformer or better, and at a fraction of the computational cost.
+
+We have implemented a Low Complexity Adaptive (LCA) beamformer, which adaptively selects a window from a predefined set.
+The set is comprised of windows that are typical solutions found by the Minimum Variance method.
+The LCA beamformer was tested using simulated and experimental data from the Kongsberg Maritime HISAS 1030 sonar.
+On a simulated scene with speckle, highlight and shadow, the beamformer offered better lateral edge definition compared to the MV beamformer, and speckle intensity and shape comparable to DAS and MV beamformers.
+These results were verified by the experimental data.
+
+An attactive trait of the LCA is its low computational complexity; while the MV beamformer is of O($M^3$), the LCA method is of O($MW$), with $M$ being the number of channels and $W$ the number of windows.
+We made the LCA perform like the MV method using a well designed set of 30 windows.
+Hence, unless the array is very small, the proposed method will perform like the MV beamformer or better, and at a fraction of the computational cost.
 

### 2012-03-07 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2012:03:05	15:17:06	(Mon 5.Mar)
la2pdf commit - 2012:03:05	15:47:24	(Mon 5.Mar)
la2pdf commit - 2012:03:05	16:19:35	(Mon 5.Mar)
la2pdf commit - 2012:03:07	11:23:25	(Wed 7.Mar)

@@
 
-% \input{preamble}
-
-% *** GRAPHICS RELATED PACKAGES ***
-%
 \ifCLASSINFOpdf
@@
 \RequirePackage[table,dvipsnames,svgnames]{xcolor}
-
-\ifTODO
-   \setlength{\marginparwidth}{1cm}
-%    \setlength{\marginparpush}{1cm}
-%    \setlength\marginpar{1.5cm}
-   \addtolength\marginparsep{0.1cm}
-%    \definecolor{lightblue}{rgb}{0.8,0.8,1}
-%    \newlength\textwidthold
-%    \setlength\textwidthold{\textwidth}
-%    \setlength\textwidth{\columnwidth}
-   \usepackage{./todonotes}
-%    \setlength\textwidth\textwidthold
-\else
-   \usepackage[disable]{todonotes} 
-\fi
 
@@
 
+\newcounter{todoidx}
+% \setcounter{todoidx}
+
+\ifTODO
+   \definecolor{todobackground}{rgb}{0.95,0.95,0.95}
+   \setlength\marginparsep{3pt}
+   \setlength\marginparwidth{42pt}
+   \newlength\marginparwidthsmall
+   \setlength\marginparwidthsmall{\marginparwidth}
+   \addtolength\marginparwidthsmall{-7pt}
+   \newcommand\todo[1]{%
+      \addtocounter{todoidx}{1}%
+      {\color{Red}\fbox{\bf\thetodoidx{}}}%
+      \marginpar{%
+         {\vspace*{-10pt}\color{Red}\fbox{\bf\thetodoidx{}}}\\%
+         \fcolorbox{red}{todobackground}{\parbox{\marginparwidthsmall}{#1}}}}
+\else
+%    \usepackage[disable]{./todonotes} 
+   \newcommand\todo[1]{}
+\fi
+
 % correct bad hyphenation here
@@
 %%%%%%%%%%%%%%%%%%                              ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
-  \RequirePackage{layout}  
+
 \begin{document}
-\layout
-%
+
 \title{Low Complexity Adaptive Beamformer\\ for Active Sonar Imaging}
@@
 
-\IEEEPARstart{A}{daptive} beamformers have only recently been introduced in active sonar imaging.\todo{break at something} This is because , as demo file is intended to serve as a ``starter file''
-for IEEE Computer Society journal papers produced under \LaTeX\ using
-IEEEtran.cls version 1.7 and later.
+\IEEEPARstart{T}{o} form images from a modern phased array sonar system the received waveform is usually recorded, and then processed by a digital beamformer.
+Estimating one pixel at a time, the beamformer adjusts the arrays spatial response to ensure the maximum gain in the direction of the pixel of interest, while maintaining maximum suppression at all other angles.
+
+The imaging capabilities of a modern phased array sonar system depend on physical attributes such element response and array geometry, the transmitted signal, as well as the beamforming method being used on transmission and reception.
+Beamforming is the concept of applying delays and weights to the sensors channels to steer the arrays response to points of interest.
+
+\begin{itemize}
+\item Why beamforming?
+\item Adaptive beamformers
+\item Challenges in active sonar systems. Correlated noise, robustification kills the adaptive potential. Quite computationally intensive.
+\item LCA
+\end{itemize}
+
+, and maximum suppression of while ensuring that the beamformer digitally  before each of the pixels are estimated one at a time.
+The resolution and contrast of such a system will depend on the systems spatial response, which ideally should be narrow  be very sharp in the desired direction its ability to achieve  fundamental principle of forming a sonar image is to record the received wavefield,
+
+image quality of a phased array sonar imaging systems depend on  the choice of weights to apply to each of the sensors are crucial.
+
+A modern phased array imaging system may be thought of as a spatial filter. To achieve the best possible performance, the
+
+resolution and contrast
+
+Adaptive beamformers have only recently been introduced in active sonar imaging.
+For a while they were considered unsuited for this purpose because the backscattered signal is largely correlated with the
 
@@
 
-% needed in second column of first page if using \IEEEpubid
-%\IEEEpubidadjcol
-
-\subsubsection{Subsubsection Heading Here}
-Subsubsection text here.
-
-% An example of a floating figure using the graphicx package.
-% Note that \label must occur AFTER (or within) \caption.
-% For figures, \caption should occur after the \includegraphics.
-% Note that IEEEtran v1.7 and later has special internal code that
-% is designed to preserve the operation of \label within \caption
-% even when the captionsoff option is in effect. However, because
-% of issues like this, it may be the safest practice to put all your
-% \label just after \caption rather than within \caption{}.
-%
-% Reminder: the "draftcls" or "draftclsnofoot", not "draft", class
-% option should be used if it is desired that the figures are to be
-% displayed while in draft mode.
-%
 %\begin{figure}[!t]
@@
 %\end{figure}
-
-% Note that IEEE typically puts floats only at the top, even when this
-% results in a large percentage of a column being occupied by floats.
-% However, the Computer Society has been known to put floats at the bottom.
 
@@
 
+\bibliography{../../Library/library}
+
 % biography section

### 2012-03-07 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2012:03:07	12:26:29	(Wed 7.Mar)

@@
 \item Challenges in active sonar systems. Correlated noise, robustification kills the adaptive potential. Quite computationally intensive.
-\item LCA
+\item LCA is like DAS, only we test several windows and choose the one that yield the lowest beamformer output power. Ultrasound \cite{Synnevag2008}, active sonar 
 \end{itemize}
@@
 
-% can use a bibliography generated by BibTeX as a .bbl file
-% BibTeX documentation can be easily obtained at:
-% http://www.ctan.org/tex-archive/biblio/bibtex/contrib/doc/
-% The IEEEtran BibTeX style support page is at:
-% http://www.michaelshell.org/tex/ieeetran/bibtex/
-%\bibliographystyle{IEEEtran}
-% argument is your BibTeX string definitions and bibliography database(s)
-%\bibliography{IEEEabrv,../bib/paper}
+\bibliographystyle{IEEEtran}
+\bibliography{../../Library/library}
+
+% End up doing this:
+% \begin{thebibliography}{1}
 % 
-% <OR> manually copy in the resultant .bbl file
-% set second argument of \begin to the number of references
-% (used to reserve space for the reference number labels box)
-\begin{thebibliography}{1}
-
-\bibitem{IEEEhowto:kopka}
-H.~Kopka and P.~W. Daly, \emph{A Guide to {\LaTeX}}, 3rd~ed.\hskip 1em plus
-  0.5em minus 0.4em\relax Harlow, England: Addison-Wesley, 1999.
-
-\end{thebibliography}
-
-\bibliography{../../Library/library}
+% \bibitem{IEEEhowto:kopka}
+% H.~Kopka and P.~W. Daly, \emph{A Guide to {\LaTeX}}, 3rd~ed.\hskip 1em plus
+%   0.5em minus 0.4em\relax Harlow, England: Addison-Wesley, 1999.
+% 
+% \end{thebibliography}
 
@@
 
-\begin{IEEEbiography}{Michael Shell}
-Biography text here.
+\vfill
+
+\begin{IEEEbiography}{Noname dude}
+He doesn't exist.
 \end{IEEEbiography}
@@
 % if you will not have a photo at all:
-\begin{IEEEbiographynophoto}{John Doe}
-Biography text here.
+\begin{IEEEbiographynophoto}{Jo Inge Buskenes}
+Not in at the moment.
 \end{IEEEbiographynophoto}
 
-% insert where needed to balance the two columns on the last page with
-% biographies
 %\newpage
 
-\begin{IEEEbiographynophoto}{Jane Doe}
-Biography text here.
-\end{IEEEbiographynophoto}
-
-% You can push biographies down or up by placing
-% a \vfill before or after them. The appropriate
-% use of \vfill depends on what kind of text is
-% on the last page and whether or not the columns
-% are being equalized.
+% \begin{IEEEbiographynophoto}{Jo Inge Buskenes}
+% Not in at the moment.
+% \end{IEEEbiographynophoto}
 
@@
 
-% that's all folks
 \end{document}

### 2012-03-07 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2012:03:07	14:10:10	(Wed 7.Mar)

@@
 \item Challenges in active sonar systems. Correlated noise, robustification kills the adaptive potential. Quite computationally intensive.
-\item LCA is like DAS, only we test several windows and choose the one that yield the lowest beamformer output power. Ultrasound \cite{Synnevag2008}, active sonar 
+\item LCA is like DAS, only we test several windows and choose the one that yield the lowest beamformer output power.
+Ultrasound \cite{Synnevag2008}, active sonar \cite{Buskenes2011}\cite{Blomberg2011}\cite{Blomberg2012}
+\item Outline: Start with Capon's response when means for robustification are added.
+\end{itemize}
+
+Outline:
+\begin{itemize}
+\item Capon's resonse 
+\item Choice of window functions makes little difference.
+\item Steering and mainlobewidths have outer bounds.
+\item Beamspace?
+\item Chosen window plots - what may they tell us? Variance intensity values when using various windows.
+\item Assymmetric windows?
 \end{itemize}

### 2012-03-07 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2012:03:07	14:59:18	(Wed 7.Mar)

@@
 
+%%%%%%%%%%                                      ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
+% MACROS %       
+%%%%%%%%%%                                      ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
+
+\newcommand\Grey[1]{{\color{Grey}#1}}
+\newcommand\Red[1]{{\color{Red}#1}}
+\newcommand\Blue[1]{{\color{Blue}#1}}
+\newcommand\DarkBlue[1]{{\color{DarkBlue}#1}}
+\newcommand\LightBlue[1]{{\color{LightBlue}#1}}
+\newcommand\Brown[1]{{\color{Brown}#1}}
+\newcommand\Green[1]{{\color{Green}#1}}
+\newcommand\SeaGreen[1]{{\color{SeaGreen}#1}}
+\newcommand\Yellow[1]{{\color{yellow}#1}}
+\newcommand\Orange[1]{{\color{orange}#1}}
+
+\newcommand\nn{\nonumber\\}
+
+\newcommand\nmat[1]{\begin{matrix}#1\end{matrix}}
+\newcommand\bmat[1]{\begin{bmatrix}#1\end{bmatrix}}
+\newcommand\case[1]{\begin{cases}#1\end{cases}}
+\newcommand\textbox[2]{\footnotesize\text{\parbox{#1}{\centering\emph{#2}}}}
+
+\newcommand\rand{\text{rand}}
+\newcommand\randn{\text{randn}}
+\newcommand\rect{\text{rect}}
+\newcommand\sinc{\text{sinc}}
+\newcommand\tr{\text{tr}}
+\newcommand\adj{\text{adj}}
+
+% \newcommand\max{\text{max}}
+% \newcommand\min{\text{min}}
+
+\newcommand\qqquad{\quad\qquad}
+\newcommand\qqqquad{\qquad\qquad}
+
+\renewcommand\l[1]{\left#1}
+\renewcommand\r[1]{\right#1}
+
+% {\text{\parbox{1.5cm}{\centering volume hyper- sphere}}}
+
+%Keyword colouring:
+\newcommand\kw[1]{#1}
+\newcommand\parm[1]{#1}%\color{Black}#1\color{Black}}
+
+\newcommand\of[1]{\scriptstyle(\parm{#1})\displaystyle}
+\newcommand\df[1]{\scriptstyle[\parm{#1}]\displaystyle}
+\newcommand\var[3]{#1_\text{#2}\of{#3}}
+
+\newcommand\diag{\text{diag}}
+
+% \raisebox{lift}[extend-above-baseline][extend-below-baseline]{text}
+\newcommand\mt[1]{\text{\emph{#1}}} %mt = mathtext
+\newcommand\mathnorm{\textstyle}
+\newcommand\mathbig[1]{\displaystyle#1\mathnorm}
+\newcommand\mathsmall[1]{\scriptstyle#1\mathnorm}
+\newcommand\mathtiny[1]{\scriptscriptstyle#1\mathnorm}
+\newcommand\sfrac[2]{\scriptstyle\raisebox{0.25pt}[0pt][0pt]{$\frac{#1}{#2}$}\mathnorm}
+\newcommand\nfrac[2]{\textstyle\frac{#1}{#2}\displaystyle}
+
+\newcommand\sumu[1]{\sum\limits^{#1}\,}
+\newcommand\suml[1]{\sum\limits_{#1}\,}
+\newcommand\sumb[2]{\sum\limits_{#1}^{#2}\,}
+
+\newcommand\produ[1]{\prod\limits^{#1}\,}
+\newcommand\prodl[1]{\prod\limits_{#1}\,}
+\newcommand\prodb[2]{\prod\limits_{#1}^{#2}\,}
+
+%Math macros:
+\newcommand\diff[2]{\frac{\kw{d}\,\textstyle #1\scriptstyle}{\kw{d\parm{#2}}}\displaystyle}
+\newcommand\ddiff[2]{\frac{\kw{d^2}\,\displaystyle #1\scriptstyle}{\kw{d\parm{#2}}^2}\displaystyle}
+
+\renewcommand\d[1]{\scriptstyle\kw{\,d\parm{#1}}\displaystyle}
+
+% These commands are mutually exclusive. Remember to "renew" in v2.
+\newcommand\intb[4]{\int\limits_{#3}^{#4} #1 \d{#2}} % \int{exp}{var}{from}{to}
+\newcommand\intl[3]{\int\limits_{#3} #1 \d{#2}} % \int{exp}{var}{for all}
+\newcommand\intu[2]{\int #1 \d{#2}} % \int{exp}{var}{for all}
+
+\newcommand\T{^{\scriptscriptstyle T}}
+\renewcommand\H{^{\scriptscriptstyle H}}
+
+\renewcommand\vec[1]{\boldsymbol{#1}}
+\newcommand\mat[1]{\boldsymbol{#1}}
+
+\renewcommand*\a{\vec a}
+\renewcommand*\i{\vec i}
+\renewcommand*\k{\vec k}
+\newcommand*\n{\vec n}
+\newcommand*\p{\vec p}
+\newcommand*\s{\vec s}
+\newcommand*\w{\vec w}
+\newcommand*\x{\vec x}
+\newcommand*\y{\vec y}
+
+\newcommand*\A{\mat A}
+\newcommand*\B{\mat B}
+\newcommand*\C{\mat C}
+\newcommand*\E{\mat E}
+% \renewcommand*\H{\mat H}
+\renewcommand*\P{\mat P}
+\newcommand*\eP{\mat{\hat P}}
+\newcommand*\R{\mat R}
+\newcommand*\Ri{\R^{-1}}
+\newcommand*\eR{\mat{\hat R}}
+\newcommand*\W{\mat W}
+\newcommand*\X{\mat X}
+\newcommand*\Y{\mat Y}
+
+\renewcommand*\L{\mat \Lambda}
+\newcommand*\U{\mat U}
+% \renewcommand*\t{\mathtiny{^T}}
+% \newcommand*\h{\mathtiny{^H}}
+\renewcommand*\t{^T}
+\newcommand*\h{^H}
+
+\newcommand\D{\vec\nabla} %Del: Vector differential operator - nabla
+\newcommand\Dx{\vec\nabla\times}
+\newcommand\Dd{\vec\nabla\cdot}
+
 % correct bad hyphenation here
@@
 
-\IEEEPARstart{T}{o} form images from a modern phased array sonar system the received waveform is usually recorded, and then processed by a digital beamformer.
-Estimating one pixel at a time, the beamformer adjusts the arrays spatial response to ensure the maximum gain in the direction of the pixel of interest, while maintaining maximum suppression at all other angles.
+\IEEEPARstart{T}{o} form images from a modern phased array sonar system the received wavefield is usually recorded, and then postprocessed by a digital beamformer.
+The beamformer applies delays and weights to the sensor channels, the beamformer adjusts the arrays spatial response to focus at one pixel at a time.
+such that signals emanating from regions of interest add constructively, while ensuring that noise and interference from other angles do not.
 
@@
 \item LCA is like DAS, only we test several windows and choose the one that yield the lowest beamformer output power.
-Ultrasound \cite{Synnevag2008}, active sonar \cite{Buskenes2011}\cite{Blomberg2011}\cite{Blomberg2012}
+Ultrasound \cite{Synnevag2008}, active sonar \cite{Buskenes2011}\cite{Blomberg2011}\cite{Blomberg2012}\todo{who was first?}
 \item Outline: Start with Capon's response when means for robustification are added.
@@
 \begin{itemize}
-\item Capon's resonse 
+\item Capon's resonse when applying robustification
 \item Choice of window functions makes little difference.
@@
 \end{itemize}
+
+\begin{align}
+z[n] = \sumb{m=0}{M-1} w_m[n]^*x_m[n-\Delta_m] = \w[n]\H\x_m[n-\Delta_m]
+\end{align}
 

### 2012-03-07 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2012:03:07	18:46:58	(Wed 7.Mar)

@@
 \begin{abstract}
+
+\begin{itemize}
+\item Dynamic weights $\Rightarrow$ greater contrast/resolution potential
+\item But: Active system $\Rightarrow$ noise and signal correlated $\Rightarrow$ adaptive beamformers no longer robust.
+Computationally complex O($M^3$), and although subspace/beamspace methods exist,
+\item Subarray averaging decorrelate the noise, but in the process we sacrifice resolution.
+\end{itemize}
+
 The angular resolution and contrast in active sonar images depend on the beamformer's ability to receive signals from directions of interest, while suppressing noise and interference emanating from other directions.
@@
 \section{Introduction}
-% Computer Society journal papers do something a tad strange with the very
-% first section heading (almost always called "Introduction"). They place it
-% ABOVE the main text! IEEEtran.cls currently does not do this for you.
-% However, You can achieve this effect by making LaTeX jump through some
-% hoops via something like:
-%
-%\ifCLASSOPTIONcompsoc
-%  \noindent\raisebox{2\baselineskip}[0pt][0pt]%
-%  {\parbox{\columnwidth}{\section{Introduction}\label{sec:introduction}%
-%  \global\everypar=\everypar}}%
-%  \vspace{-1\baselineskip}\vspace{-\parskip}\par
-%\else
-%  \section{Introduction}\label{sec:introduction}\par
-%\fi
-%
-% Admittedly, this is a hack and may well be fragile, but seems to do the
-% trick for me. Note the need to keep any \label that may be used right
-% after \section in the above as the hack puts \section within a raised box.
 
+\begin{itemize}
+\item Sell it!
+\item As briefly as possible, introduce beamforming.
+\item Adaptive beamformers
+\item Challenges in active sonar systems. Correlated noise, robustification kills the adaptive potential. Quite computationally intensive.
+\item LCA is like DAS, only we test several windows and choose the one that yield the lowest beamformer output power.
+Ultrasound \cite{Synnevag2008}, active sonar \cite{Buskenes2011}\cite{Blomberg2011}\cite{Blomberg2012}\todo{who was first?}
+\item Outline: Start with Capon's response when means for robustification are added.
+\end{itemize}
+
+\ \\
 \IEEEPARstart{T}{o} form images from a modern phased array sonar system the received wavefield is usually recorded, and then postprocessed by a digital beamformer.
@@
 \begin{align}
-z[n] = \sumb{m=0}{M-1} w_m[n]^*x_m[n-\Delta_m] = \w[n]\H\x_m[n-\Delta_m]
+z[n] = \sumb{m=0}{M-1} w_m[n]^*x_m[n-\Delta_m] = \w\H[n]\x[n-\Delta_m]
 \end{align}
+
+\section{Methods}
+
+Basically, we are working with a practical implementation of the Capon beamformer that computes a set of weights $\vec w$ for every single sample $n$ by solving:
+\begin{gather*}
+\vec w[n] = \frac{\hat{\mat R}\,\!^{-1}[n]\vec a}{\vec a\H\hat{\mat R}\,\!^{-1}[n]\vec a} = \frac{\text{\raisebox{1.9pt}{$\vec\chi$}}[n]}{\vec a\H\text{\raisebox{1.9pt}{$\vec\chi$}}[n]}
+\end{gather*}%
+where
+% \newcommand\X{\text{\raisebox{2pt}{$\vec\chi$}}}
+\begin{gather*}
+\text{\raisebox{1.9pt}{$\vec\chi$}} = \hat{\mat R}\,\!^{-1}\vec a \qquad\qquad\text{is the solution to}\qquad\qquad \hat{\mat R}\text{\raisebox{1.9pt}{$\vec\chi$}} = \vec a.
+\end{gather*}
+Here $\vec a$ is a \emph{steering vector} that if set to $\vec 1$ steers towards broadside, and $\hat{\mat R}$ is an estimate of the spatial covariance matrix, often computed as
+\begin{gather*}
+r_{ij}[n] = \frac{1}{K(2N_{\text{avg}}+1)}\sumb{k=0}{K-1} \sumb{n'=n-N_{\text{avg}}}{n+N_{\text{avg}}} x_{i+k}[n']\,x_{j+k}[n'],
+\end{gather*}
+where $r_{ij}$ is the $(i,j)$'th element of the covariance matrix, $K$ is the number of subarrays, $2N_{\text{avg}}+1$ is the number of samples to perform time averaging over, and $\vec x_i[n']$ is the data recorded by the $i$'th sensor at sample $n'$.
+\emph{Diagonal loading} is usually also applied to constrain the adaptability and ensure numerical stability, but this operation is insignificant in the context of performance and will be ignored.
 

### 2012-03-08 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2012:03:08	12:34:44	(Thu 8.Mar)

@@
 \item As briefly as possible, introduce beamforming.
-\item Adaptive beamformers
-\item Challenges in active sonar systems. Correlated noise, robustification kills the adaptive potential. Quite computationally intensive.
+\item Adaptive beamformer potential
+\item Why active sonar make it hard. Correlated noise, robustification kills the adaptive potential. Quite computationally intensive.
 \item LCA is like DAS, only we test several windows and choose the one that yield the lowest beamformer output power.
-Ultrasound \cite{Synnevag2008}, active sonar \cite{Buskenes2011}\cite{Blomberg2011}\cite{Blomberg2012}\todo{who was first?}
+Ultrasound \cite{Synnevag2008}, active sonar \cite{Buskenes2011}\cite{Blomberg2011}\cite{Blomberg2012}\todo{who was that guy in the 80ties?}
 \item Outline: Start with Capon's response when means for robustification are added.
@@
 
-\begin{itemize}
-\item Why beamforming?
-\item Adaptive beamformers
-\item Challenges in active sonar systems. Correlated noise, robustification kills the adaptive potential. Quite computationally intensive.
-\item LCA is like DAS, only we test several windows and choose the one that yield the lowest beamformer output power.
+% \begin{itemize}
+% \item Why beamforming?
+% \item Adaptive beamformers
+% \item Challenges in active sonar systems. Correlated noise, robustification kills the adaptive potential. Quite computationally intensive.
+% \item LCA is like DAS, only we test several windows and choose the one that yield the lowest beamformer output power.
 Ultrasound \cite{Synnevag2008}, active sonar \cite{Buskenes2011}\cite{Blomberg2011}\cite{Blomberg2012}\todo{who was first?}
-\item Outline: Start with Capon's response when means for robustification are added.
-\end{itemize}
-
-Outline:
-\begin{itemize}
-\item Capon's resonse when applying robustification
-\item Choice of window functions makes little difference.
-\item Steering and mainlobewidths have outer bounds.
-\item Beamspace?
-\item Chosen window plots - what may they tell us? Variance intensity values when using various windows.
-\item Assymmetric windows?
-\end{itemize}
-
-\begin{align}
-z[n] = \sumb{m=0}{M-1} w_m[n]^*x_m[n-\Delta_m] = \w\H[n]\x[n-\Delta_m]
-\end{align}
-
-\section{Methods}
-
-Basically, we are working with a practical implementation of the Capon beamformer that computes a set of weights $\vec w$ for every single sample $n$ by solving:
-\begin{gather*}
-\vec w[n] = \frac{\hat{\mat R}\,\!^{-1}[n]\vec a}{\vec a\H\hat{\mat R}\,\!^{-1}[n]\vec a} = \frac{\text{\raisebox{1.9pt}{$\vec\chi$}}[n]}{\vec a\H\text{\raisebox{1.9pt}{$\vec\chi$}}[n]}
-\end{gather*}%
-where
-% \newcommand\X{\text{\raisebox{2pt}{$\vec\chi$}}}
-\begin{gather*}
-\text{\raisebox{1.9pt}{$\vec\chi$}} = \hat{\mat R}\,\!^{-1}\vec a \qquad\qquad\text{is the solution to}\qquad\qquad \hat{\mat R}\text{\raisebox{1.9pt}{$\vec\chi$}} = \vec a.
-\end{gather*}
-Here $\vec a$ is a \emph{steering vector} that if set to $\vec 1$ steers towards broadside, and $\hat{\mat R}$ is an estimate of the spatial covariance matrix, often computed as
-\begin{gather*}
-r_{ij}[n] = \frac{1}{K(2N_{\text{avg}}+1)}\sumb{k=0}{K-1} \sumb{n'=n-N_{\text{avg}}}{n+N_{\text{avg}}} x_{i+k}[n']\,x_{j+k}[n'],
-\end{gather*}
-where $r_{ij}$ is the $(i,j)$'th element of the covariance matrix, $K$ is the number of subarrays, $2N_{\text{avg}}+1$ is the number of samples to perform time averaging over, and $\vec x_i[n']$ is the data recorded by the $i$'th sensor at sample $n'$.
+% \item Outline: Start with Capon's response when means for robustification are added.
+% \end{itemize}
+% 
+% 
+% Outline:
+% \begin{itemize}
+% \item Capon's resonse when applying robustification
+% \item Choice of window functions makes little difference.
+% \item Steering and mainlobewidths have outer bounds.
+% \item Beamspace?
+% \item Chosen window plots - what may they tell us? Variance intensity values when using various windows.
+% \item Assymmetric windows?
+% \end{itemize}
+% 
+% 
+% \begin{align}
+% z[n] = \sumb{m=0}{M-1} w_m[n]^*x_m[n-\Delta_m] = \w\H[n]\x[n-\Delta_m]
+% \end{align}
+% 
+% 
+% \section{Methods}
+% 
+% Basically, we are working with a practical implementation of the Capon beamformer that computes a set of weights $\vec w$ for every single sample $n$ by solving:
+% \begin{gather*}
+% \vec w[n] = \frac{\hat{\mat R}\,\!^{-1}[n]\vec a}{\vec a\H\hat{\mat R}\,\!^{-1}[n]\vec a} = \frac{\text{\raisebox{1.9pt}{$\vec\chi$}}[n]}{\vec a\H\text{\raisebox{1.9pt}{$\vec\chi$}}[n]}
+% \end{gather*}%
+% where
+% % \newcommand\X{\text{\raisebox{2pt}{$\vec\chi$}}}
+% \begin{gather*}
+% \text{\raisebox{1.9pt}{$\vec\chi$}} = \hat{\mat R}\,\!^{-1}\vec a \qquad\qquad\text{is the solution to}\qquad\qquad \hat{\mat R}\text{\raisebox{1.9pt}{$\vec\chi$}} = \vec a.
+% \end{gather*}
+% Here $\vec a$ is a \emph{steering vector} that if set to $\vec 1$ steers towards broadside, and $\hat{\mat R}$ is an estimate of the spatial covariance matrix, often computed as
+% \begin{gather*}
+% r_{ij}[n] = \frac{1}{K(2N_{\text{avg}}+1)}\sumb{k=0}{K-1} \sumb{n'=n-N_{\text{avg}}}{n+N_{\text{avg}}} x_{i+k}[n']\,x_{j+k}[n'],
+% \end{gather*}
+% where $r_{ij}$ is the $(i,j)$'th element of the covariance matrix, $K$ is the number of subarrays, $2N_{\text{avg}}+1$ is the number of samples to perform time averaging over, and $\vec x_i[n']$ is the data recorded by the $i$'th sensor at sample $n'$.
 \emph{Diagonal loading} is usually also applied to constrain the adaptability and ensure numerical stability, but this operation is insignificant in the context of performance and will be ignored.
-
-, and maximum suppression of while ensuring that the beamformer digitally  before each of the pixels are estimated one at a time.
+% 
+% , and maximum suppression of while ensuring that the beamformer digitally  before each of the pixels are estimated one at a time.
 The resolution and contrast of such a system will depend on the systems spatial response, which ideally should be narrow  be very sharp in the desired direction its ability to achieve  fundamental principle of forming a sonar image is to record the received wavefield,
-
-image quality of a phased array sonar imaging systems depend on  the choice of weights to apply to each of the sensors are crucial.
-
-A modern phased array imaging system may be thought of as a spatial filter. To achieve the best possible performance, the
-
-resolution and contrast
-
-Adaptive beamformers have only recently been introduced in active sonar imaging.
+% 
+% image quality of a phased array sonar imaging systems depend on  the choice of weights to apply to each of the sensors are crucial. 
+% 
+% A modern phased array imaging system may be thought of as a spatial filter. To achieve the best possible performance, the 
+% 
+% resolution and contrast 
+% 
+% Adaptive beamformers have only recently been introduced in active sonar imaging.
 For a while they were considered unsuited for this purpose because the backscattered signal is largely correlated with the
-
-We have implemented a Low Complexity Adaptive (LCA) beamformer, which adaptively selects a window from a predefined set.
+% 
+% 
+% 
+% We have implemented a Low Complexity Adaptive (LCA) beamformer, which adaptively selects a window from a predefined set.
 The set is comprised of windows that are typical solutions found by the Minimum Variance method.
@@
 These results were verified by the experimental data.
-
-An attactive trait of the LCA is its low computational complexity; while the MV beamformer is of O($M^3$), the LCA method is of O($MW$), with $M$ being the number of channels and $W$ the number of windows.
+% 
+% An attactive trait of the LCA is its low computational complexity; while the MV beamformer is of O($M^3$), the LCA method is of O($MW$), with $M$ being the number of channels and $W$ the number of windows.
 We made the LCA perform like the MV method using a well designed set of 30 windows.
 Hence, unless the array is very small, the proposed method will perform like the MV beamformer or better, and at a fraction of the computational cost.
-
-We have implemented a Low Complexity Adaptive (LCA) beamformer, which adaptively selects a window from a predefined set.
+% 
+% We have implemented a Low Complexity Adaptive (LCA) beamformer, which adaptively selects a window from a predefined set.
 The set is comprised of windows that are typical solutions found by the Minimum Variance method.
@@
 These results were verified by the experimental data.
-
-An attactive trait of the LCA\todo{break at something} is its low computational complexity; while the MV beamformer is of O($M^3$), the LCA method is of O($MW$), with $M$ being the number of channels and $W$ the number of windows.
+% 
+% An attactive trait of the LCA\todo{break at something} is its low computational complexity; while the MV beamformer is of O($M^3$), the LCA method is of O($MW$), with $M$ being the number of channels and $W$ the number of windows.
 We made the LCA perform like the MV method using a well designed set of 30 windows.
 Hence, unless the array is very small, the proposed method will perform like the MV beamformer or better, and at a fraction of the computational cost.
-
-We have implemented a Low Complexity Adaptive (LCA) beamformer, which adaptively selects a window from a predefined set.
+% 
+% We have implemented a Low Complexity Adaptive (LCA) beamformer, which adaptively selects a window from a predefined set.
 The set is comprised of windows that are typical solutions found by the Minimum Variance method.
@@
 These results were verified by the experimental data.
-
-An attactive trait of the LCA is its low computational complexity; while the MV beamformer is of O($M^3$), the LCA method is of O($MW$), with $M$ being the number of channels and $W$ the number of windows.
+% 
+% An attactive trait of the LCA is its low computational complexity; while the MV beamformer is of O($M^3$), the LCA method is of O($MW$), with $M$ being the number of channels and $W$ the number of windows.
 We made the LCA perform like the MV method using a well designed set of 30 windows.
 Hence, unless the array is very small, the proposed method will perform like the MV beamformer or better, and at a fraction of the computational cost.
-
-\hfill mds
-
-\hfill January 11, 2007
-
-\subsection{Subsection Heading Here}
-Subsection text here.
+% 
+% \hfill mds
+%  
+% \hfill January 11, 2007
+% 
+% \subsection{Subsection Heading Here}
+% Subsection text here.
 
@@
 
+%%%%%%%%%%%%%%%%%%                              ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
+% DOCUMENT APPENDICES %
+%%%%%%%%%%%%%%%%%%                              ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
+
 \appendices
-\section{Proof of the First Zonklar Equation}
-Appendix one text goes here.
-
-% you can choose not to have a title for an appendix
-% if you want by leaving the argument blank
-\section{}
-Appendix two text goes here.
 
@@
 \ifCLASSOPTIONcompsoc
-  % The Computer Society usually uses the plural form
   \section*{Acknowledgments}
 \else
-  % regular IEEE prefers the singular form
   \section*{Acknowledgment}

### 2012-03-08 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2012:03:08	13:18:56	(Thu 8.Mar)
la2pdf commit - 2012:03:08	14:02:17	(Thu 8.Mar)
la2pdf commit - 2012:03:08	14:33:11	(Thu 8.Mar)

@@
 
+\usepackage{tikz}
+\usetikzlibrary{shapes,snakes}
+\usepackage{amsmath,amssymb}
+
+\newenvironment{outline}
+{\begin{itemize}}
+{\end{itemize}}
+
+%    \definecolor{todobackground}{rgb}{0.95,0.95,0.95}
+%    \setlength\marginparsep{3pt}
+%    \setlength\marginparwidth{42pt}
+%    \newlength\marginparwidthsmall
+%    \setlength\marginparwidthsmall{\marginparwidth}
+%    \addtolength\marginparwidthsmall{-7pt}
+%    \newcommand\todo[1]{%
+%       \addtocounter{todoidx}{1}%
+%       {\color{Red}\fbox{\bf\thetodoidx{}}}%
+%       \marginpar{%
+%          {\vspace*{-10pt}\color{Red}\fbox{\bf\thetodoidx{}}}\\%
+%          \fcolorbox{red}{todobackground}{\parbox{\marginparwidthsmall}{#1}}}}
+%
+
 % correct bad hyphenation here
@@
 \item Dynamic weights $\Rightarrow$ greater contrast/resolution potential
-\item But: Active system $\Rightarrow$ noise and signal correlated $\Rightarrow$ adaptive beamformers no longer robust.
-Computationally complex O($M^3$), and although subspace/beamspace methods exist,
+\item But:
+\begin{itemize}
+\item Active system $\Rightarrow$ noise and signal correlated $\Rightarrow$ adaptive beamformers no longer robust.
 \item Subarray averaging decorrelate the noise, but in the process we sacrifice resolution.
+\item Computationally complex O($M^3$) due to inversion of a spatial covariance matrix.
+LCA is only O($MW$).
+Even if Capon subspace/beamspace versions exist, they remain more complex than the LCA, and is not as ideally suited for implementation of SIMD hardware (GPUs, DSPs, ...).
+\end{itemize}
+\item LCA is inherently robust, is ``parameter free'' once setup with a well designed window set, and can often be implemented with minor modifications to already existing DAS beamformers.
 \end{itemize}
 
-The angular resolution and contrast in active sonar images depend on the beamformer's ability to receive signals from directions of interest, while suppressing noise and interference emanating from other directions.
-For sonar arrays, this is achieved by applying weights to the array channels.
-
-While classical beamformers use predefined windows, adaptive beamformers estimate the optimal window by analytical evaluation of the data.
-The minimum variance (MV) beamformer, for instance, calculates the set of weights that minimises the variance of the beamformer's output.
-
-We have implemented a Low Complexity Adaptive (LCA) beamformer, which adaptively selects a window from a predefined set.
-The set is comprised of windows that are typical solutions found by the Minimum Variance method.
-The LCA beamformer was tested using simulated and experimental data from the Kongsberg Maritime HISAS 1030 sonar.
-On a simulated scene with speckle, highlight and shadow, the beamformer offered better lateral edge definition compared to the MV beamformer, and speckle intensity and shape comparable to DAS and MV beamformers.
-These results were verified by the experimental data.
-
-An attactive trait of the LCA is its low computational complexity; while the MV beamformer is of O($M^3$), the LCA method is of O($MW$), with $M$ being the number of channels and $W$ the number of windows.
-We made the LCA perform like the MV method using a well designed set of 30 windows.
-Hence, unless the array is very small, the proposed method will perform like the MV beamformer or better, and at a fraction of the computational cost.
 \end{abstract}
@@
 \item Adaptive beamformer potential
-\item Why active sonar make it hard. Correlated noise, robustification kills the adaptive potential. Quite computationally intensive.
+\item Why adaptive beamformers struggle in active sonar systems.
+Correlated noise, robustification kills the adaptive potential.
+Quite computationally intensive.
+Constraints must be applied in one way or another - parameters must be tuned.
 \item LCA is like DAS, only we test several windows and choose the one that yield the lowest beamformer output power.
 Ultrasound \cite{Synnevag2008}, active sonar \cite{Buskenes2011}\cite{Blomberg2011}\cite{Blomberg2012}\todo{who was that guy in the 80ties?}
-\item Outline: Start with Capon's response when means for robustification are added.
+\item Outline: 
+\begin{itemize}
+\item Methods.
+\item See what happens to Capon's response when means for robustification are added.
+\item Setup the window set for LCA to match this.
+\item Notice that the LCA performs like Capon, even with a small window database.
+\item Start with Kaiser, see that other window functions also work well.
+\end{itemize}
 \end{itemize}
 
-\ \\
-\IEEEPARstart{T}{o} form images from a modern phased array sonar system the received wavefield is usually recorded, and then postprocessed by a digital beamformer.
+\section{Methods}
+
+\begin{align}
+z[n] = \sumb{m=0}{M-1} w_m[n]^*x_m[n-\Delta_m] = \w\H[n]\X_\Delta[n]
+\end{align}
+where
+\begin{align}
+\w[n] = \bmat{w_0[n]\\w_1[n]\\\vdots\\w_{M-1}[n]} \qquad \text{and} \X_\Delta[n] = 
+\end{align}
+
+\section{Conclusion}
+
+\begin{itemize}
+\item Something similar to the abstract.
+\end{itemize}
+
+% \ \\
+% \IEEEPARstart{T}{o} form images from a modern phased array sonar system the received wavefield is usually recorded, and then postprocessed by a digital beamformer.
 The beamformer applies delays and weights to the sensor channels, the beamformer adjusts the arrays spatial response to focus at one pixel at a time.
 such that signals emanating from regions of interest add constructively, while ensuring that noise and interference from other angles do not.
-
-The imaging capabilities of a modern phased array sonar system depend on physical attributes such element response and array geometry, the transmitted signal, as well as the beamforming method being used on transmission and reception.
+% 
+% The imaging capabilities of a modern phased array sonar system depend on physical attributes such element response and array geometry, the transmitted signal, as well as the beamforming method being used on transmission and reception.
 Beamforming is the concept of applying delays and weights to the sensors channels to steer the arrays response to points of interest.
 
-% \begin{itemize}
-% \item Why beamforming?
-% \item Adaptive beamformers
-% \item Challenges in active sonar systems. Correlated noise, robustification kills the adaptive potential. Quite computationally intensive.
-% \item LCA is like DAS, only we test several windows and choose the one that yield the lowest beamformer output power.
-Ultrasound \cite{Synnevag2008}, active sonar \cite{Buskenes2011}\cite{Blomberg2011}\cite{Blomberg2012}\todo{who was first?}
-% \item Outline: Start with Capon's response when means for robustification are added.
-% \end{itemize}
 % 

### 2012-03-08 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2012:03:08	15:07:35	(Thu 8.Mar)
la2pdf commit - 2012:03:08	16:06:40	(Thu 8.Mar)

@@
 % \newcommand\max{\text{max}}
-% \newcommand\min{\text{min}}
+\newcommand\argmin{\text{argmin}}
 
@@
 \newcommand*\X{\mat X}
+\newcommand*\Xd{\X_{\!\Delta}}
 \newcommand*\Y{\mat Y}
@@
 \begin{align}
-z[n] = \sumb{m=0}{M-1} w_m[n]^*x_m[n-\Delta_m] = \w\H[n]\X_\Delta[n]
+z[n] = \sumb{m=0}{M-1} w_m[n]^*x_m[n-\Delta_m] = \w\H[n]\Xd[n]
 \end{align}
@@
 \begin{align}
-\w[n] = \bmat{w_0[n]\\w_1[n]\\\vdots\\w_{M-1}[n]} \qquad \text{and} \X_\Delta[n] = 
+\w[n] = \bmat{w_0[n]\\w_1[n]\\\vdots\\w_{M-1}[n]} \quad \text{and} \quad\Xd[n] = \bmat{x_0[n]\\x_1[n]\\\vdots\\x_{M-1}[n]}.
 \end{align}
+
+Minimum variance distortionless resonse
+\begin{align}
+\underset{\w[n]}{\argmin}\, E\{ |z[n]|^2 \} = \underset{\w[n]}{\argmin}\, \w\H[n]\R[n]\w[n], 
+\end{align}
+subject to
+\begin{align}
+\w\H[n]\a = 1,
+\end{align}
+where
+\begin{align}
+\R[n] = E\{ \x\x\H \}.
+\end{align}s<
+If $\eR$ is the estimate of $\R$, the solution to the MVDR beamformer is
+\begin{gather*}
+\vec w[n] = \frac{\hat{\mat R}\,\!^{-1}[n]\vec a}{\vec a\H\hat{\mat R}\,\!^{-1}[n]\vec a} = \frac{\text{\raisebox{1.9pt}{$\vec\chi$}}[n]}{\vec a\H\text{\raisebox{1.9pt}{$\vec\chi$}}[n]}
+\end{gather*}
+A robust estimate of $\eR$ is found by
+\begin{itemize}
+\item Spatial averaging to decorrelate noise from signal. \emph{Subarray averaging} used here.
+\item Temporal averaging to ensure valid speckle statistics.
+\item \emph{Diagonal loading} to ensure  numerical stability prior to inversion.
+\item Choice of robustification parameters (C-I) \cite{}
+\end{itemize}
+
+\section{Results}
 

### 2012-03-08 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2012:03:08	16:38:22	(Thu 8.Mar)
la2pdf commit - 2012:03:08	17:11:31	(Thu 8.Mar)
la2pdf commit - 2012:03:08	17:44:40	(Thu 8.Mar)
la2pdf commit - 2012:03:08	20:59:03	(Thu 8.Mar)
la2pdf commit - 2012:03:08	22:55:01	(Thu 8.Mar)

@@
 
+% \RequirePackage[utf8]{inputenc}%              % Set input encoding (optionally latin1)
+% \RequirePackage[T1]{fontenc}%                 % Set font encoding
+
 \begin{document}
@@
 \item As briefly as possible, introduce beamforming.
-\item Adaptive beamformer potential
+\item Adaptive beamformer's potential lies in its ability to suppress interference power
 \item Why adaptive beamformers struggle in active sonar systems.
@@
 Constraints must be applied in one way or another - parameters must be tuned.
+\begin{itemize}
+\item Increases variance of speckle, spatial compunting needed \cite{Vignon2009}. Subarray averaging applied here \cite{Synnevag2007}
+\end{itemize}
 \item LCA is like DAS, only we test several windows and choose the one that yield the lowest beamformer output power.
@@
 \item Methods.
-\item See what happens to Capon's response when means for robustification are added.
+\item See what happens to Capon's response %\cite{Synnevag2007}  when means for robustification are added.
 \item Setup the window set for LCA to match this.
@@
 
-Minimum variance distortionless resonse
+Minimum variance distortionless resonse \cite{Capon1969}
 \begin{align}
@@
 \item \emph{Diagonal loading} to ensure  numerical stability prior to inversion.
-\item Choice of robustification parameters (C-I) \cite{}
+\item Choice of robustification parameters \cite{Synnevag2007}
 \end{itemize}
@@
 \section{Results}
+
+\begin{figure}[!t]\centering
+% \includegraphics[width=2.5in]{}
+\caption{Simulation Results}
+\label{fig_sim}
+\end{figure}
 

### 2012-03-09 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2012:03:09	01:55:32	(Fri 9.Mar)

@@
          \fcolorbox{red}{todobackground}{\parbox{\marginparwidthsmall}{#1}}}}
+
+   \newcommand\todopar[1]{\fcolorbox{red}{white}{\parbox{0.97\linewidth}{#1}}}
 \else
@@
 \begin{itemize}
-\item Increases variance of speckle, spatial compunting needed \cite{Vignon2009}. Subarray averaging applied here \cite{Synnevag2007}
+\item Noise and signal is correlated. Spatial averaging required. 
+\item Increases variance of speckle (not only in active systems?), spatial compunting \cite{Vignon2009} (US) or time averaging?.
+Subarray averaging applied here \cite{Synnevag2007} (US) \todo{Is it bad to mix ultrasound/ sonar refs?}
 \end{itemize}
@@
 \end{itemize}
+\todopar{Out of the blue idea: Weigh CF with LCA? This guy did it with MVDR \cite{Wang2009}. Probably you know already... Okay, back on track.}
 
 \section{Methods}
+
+\todopar{Threw some formulas in to get a feel of them}
 
@@
 \R[n] = E\{ \x\x\H \}.
-\end{align}s<
+\end{align}
 If $\eR$ is the estimate of $\R$, the solution to the MVDR beamformer is
@@
 
+Here $\vec a$ is a \emph{steering vector} that if set to $\vec 1$ steers towards broadside, and $\hat{\mat R}$ is an estimate of the spatial covariance matrix, often computed as
+\begin{gather*}
+r_{ij}[n] = \frac{1}{K(2N_{\text{avg}}+1)}\sumb{k=0}{K-1} \sumb{n'=n-N_{\text{avg}}}{n+N_{\text{avg}}} x_{i+k}[n']\,x_{j+k}[n'],
+\end{gather*}
+where $r_{ij}$ is the $(i,j)$'th element of the covariance matrix, $K$ is the number of subarrays, $2N_{\text{avg}}+1$ is the number of samples to perform time averaging over, and $\vec x_i[n']$ is the data recorded by the $i$'th sensor at sample $n'$.
+
+\emph{Diagonal loading} chosen as \cite{Synnevag2007}
+\begin{align}
+\varepsilon = \Delta \cdot \tr\{\eR[n]\}
+\end{align}
+
 \section{Results}
 
-\begin{figure}[!t]\centering
-% \includegraphics[width=2.5in]{}
-\caption{Simulation Results}
-\label{fig_sim}
-\end{figure}
+% \begin{figure}[!t]\centering
+% % \includegraphics[width=2.5in]{}
+% \caption{Simulation Results}
+% \label{fig_sim}
+% \end{figure}
+
+\input{gfx/gfx_input}
+% \begin{figure*}[!t]
+% \centerline{\subfloat[Case I]\includegraphics[width=2.5in]{subfigcase1}%
+% \label{fig_first_case}}
+% \hfil
+% \subfloat[Case II]{\includegraphics[width=2.5in]{subfigcase2}%
+% \label{fig_second_case}}}
+% \caption{Simulation results}
+% \label{fig_sim}
+% \end{figure*}
 
@@
 % \end{gather*}
-% Here $\vec a$ is a \emph{steering vector} that if set to $\vec 1$ steers towards broadside, and $\hat{\mat R}$ is an estimate of the spatial covariance matrix, often computed as
-% \begin{gather*}
-% r_{ij}[n] = \frac{1}{K(2N_{\text{avg}}+1)}\sumb{k=0}{K-1} \sumb{n'=n-N_{\text{avg}}}{n+N_{\text{avg}}} x_{i+k}[n']\,x_{j+k}[n'],
-% \end{gather*}
-% where $r_{ij}$ is the $(i,j)$'th element of the covariance matrix, $K$ is the number of subarrays, $2N_{\text{avg}}+1$ is the number of samples to perform time averaging over, and $\vec x_i[n']$ is the data recorded by the $i$'th sensor at sample $n'$.
-\emph{Diagonal loading} is usually also applied to constrain the adaptability and ensure numerical stability, but this operation is insignificant in the context of performance and will be ignored.
+
 % 
@@
 %
-% We have implemented a Low Complexity Adaptive (LCA) beamformer, which adaptively selects a window from a predefined set.
-The set is comprised of windows that are typical solutions found by the Minimum Variance method.
-The LCA beamformer was tested using simulated and experimental data from the Kongsberg Maritime HISAS 1030 sonar.
-On a simulated scene with speckle, highlight and shadow, the beamformer offered better lateral edge definition compared to the MV beamformer, and speckle intensity and shape comparable to DAS and MV beamformers.
-These results were verified by the experimental data.
-% 
-% An attactive trait of the LCA is its low computational complexity; while the MV beamformer is of O($M^3$), the LCA method is of O($MW$), with $M$ being the number of channels and $W$ the number of windows.
-We made the LCA perform like the MV method using a well designed set of 30 windows.
-Hence, unless the array is very small, the proposed method will perform like the MV beamformer or better, and at a fraction of the computational cost.
-% 
-% We have implemented a Low Complexity Adaptive (LCA) beamformer, which adaptively selects a window from a predefined set.
-The set is comprised of windows that are typical solutions found by the Minimum Variance method.
-The LCA beamformer was tested using simulated and experimental data from the Kongsberg Maritime HISAS 1030 sonar.
-On a simulated scene with speckle, highlight and shadow, the beamformer offered better lateral edge definition compared to the MV beamformer, and speckle intensity and shape comparable to DAS and MV beamformers.
-These results were verified by the experimental data.
-% 
-% An attactive trait of the LCA\todo{break at something} is its low computational complexity; while the MV beamformer is of O($M^3$), the LCA method is of O($MW$), with $M$ being the number of channels and $W$ the number of windows.
-We made the LCA perform like the MV method using a well designed set of 30 windows.
-Hence, unless the array is very small, the proposed method will perform like the MV beamformer or better, and at a fraction of the computational cost.
-% 
-% We have implemented a Low Complexity Adaptive (LCA) beamformer, which adaptively selects a window from a predefined set.
-The set is comprised of windows that are typical solutions found by the Minimum Variance method.
-The LCA beamformer was tested using simulated and experimental data from the Kongsberg Maritime HISAS 1030 sonar.
-On a simulated scene with speckle, highlight and shadow, the beamformer offered better lateral edge definition compared to the MV beamformer, and speckle intensity and shape comparable to DAS and MV beamformers.
-These results were verified by the experimental data.
-% 
-% An attactive trait of the LCA is its low computational complexity; while the MV beamformer is of O($M^3$), the LCA method is of O($MW$), with $M$ being the number of channels and $W$ the number of windows.
-We made the LCA perform like the MV method using a well designed set of 30 windows.
-Hence, unless the array is very small, the proposed method will perform like the MV beamformer or better, and at a fraction of the computational cost.
-% 
-% \hfill mds
-%  
-% \hfill January 11, 2007
-% 
-% \subsection{Subsection Heading Here}
-% Subsection text here.
 

### 2012-03-13 SiPS2012/sips2012.tex
Added sips article files

% Template for SiPS-2010 paper; to be used with:
%          spconf.sty  - ICASSP/ICIP LaTeX style file, and
%          IEEEbib.bst - IEEE bibliography style file.
% --------------------------------------------------------------------------
\documentclass{article}
\usepackage{spconf,amsmath,epsfig}

% Example definitions.
% --------------------
\def\x{{\mathbf x}}
\def\L{{\cal L}}

% Title.
% ------
\title{Implementing Adaptive Beamforming on the GPU for Real Time Ultrasound Imaging}
%
% Single address.
% ---------------
\name{J. P. \AA{}sen$^1$, J. I. Buskenes$^2$, C.-I. C. Nilsen$^2$, A. Austeng$^2$ and S. Holm$^{1,2}$}
\address{$^1$Mi Lab, Norwegian University of Science and Technology, Trondheim, Norway \\ $^2$Department of Informatics, University of Oslo, Oslo, Norway}
%
% For example:
% ------------
%\address{School\\
%	Department\\
%	Address}
%
% Two addresses (uncomment and modify for two-address case).
% ----------------------------------------------------------
%\twoauthors
%  {A. Author-one, B. Author-two\sthanks{Thanks to XYZ agency for funding.}}
%	{School A-B\\
%	Department A-B\\
%	Address A-B}
%  {C. Author-three, D. Author-four\sthanks{The fourth author performed the work
%	while at ...}}
%	{School C-D\\
%	Department C-D\\
%	Address C-D}
%
\begin{document}
%\ninept
%
\maketitle
%
\begin{abstract}

\end{abstract}

\section{Introduction}\label{sec:intro}

%\begin{abstract}
%The abstract should appear at the top of the left-hand column of text, about
%0.5 inch (12 mm) below the title area and no more than 3.125 inches (80 mm) in
%length.  Leave a 0.5 inch (12 mm) space between the end of the abstract and the
%beginning of the main text.  The abstract should contain about 100 to 150
%words, and should be identical to the abstract text submitted electronically
%along with the paper cover sheet.  All manuscripts must be in English, printed
%in black ink.
%\end{abstract}
%%
%\begin{keywords}
%One, two, three, four, five
%\end{keywords}
%%
%\section{Introduction}
%\label{sec:intro}
%
%These guidelines include complete descriptions of the fonts, spacing, and
%related information for producing your proceedings manuscripts. Please follow
%them and if you have any questions, direct them to Conference Management
%Services, Inc.: Phone +1-979-846-6800 or Fax +1-979-846-6900 or email
%to papers@sips2010.org.
%
%\section{Formatting your paper}
%\label{sec:format}
%
%All printed material, including text, illustrations, and charts, must be kept
%within a print area of 7 inches (178 mm) wide by 9 inches (229 mm) high. Do
%not write or print anything outside the print area. The top margin must be 1
%inch (25 mm), except for the title page, and the left margin must be 0.75 inch
%(19 mm).  All {\it text} must be in a two-column format. Columns are to be 3.39
%inches (86 mm) wide, with a 0.24 inch (6 mm) space between them. Text must be
%fully justified.
%
%\section{PAGE TITLE SECTION}
%\label{sec:pagestyle}
%
%The paper title (on the first page) should begin 1.38 inches (35 mm) from the
%top edge of the page, centered, completely capitalized, and in Times 14-point,
%boldface type.  The authors' name(s) and affiliation(s) appear below the title
%in capital and lower case letters.  Papers with multiple authors and
%affiliations may require two or more lines for this information.
%
%\section{TYPE-STYLE AND FONTS}
%\label{sec:typestyle}
%
%To achieve the best rendering both in the proceedings and from the CD-ROM, we
%strongly encourage you to use Times-Roman font.  In addition, this will give
%the proceedings a more uniform look.  Use a font that is no smaller than nine
%point type throughout the paper, including figure captions.
%
%In nine point type font, capital letters are 2 mm high.  If you use the
%smallest point size, there should be no more than 3.2 lines/cm (8 lines/inch)
%vertically.  This is a minimum spacing; 2.75 lines/cm (7 lines/inch) will make
%the paper much more readable.  Larger type sizes require correspondingly larger
%vertical spacing.  Please do not double-space your paper.  True-Type 1 fonts
%are preferred.
%
%The first paragraph in each section should not be indented, but all the
%following paragraphs within the section should be indented as these paragraphs
%demonstrate.
%
%\section{MAJOR HEADINGS}
%\label{sec:majhead}
%
%Major headings, for example, "1. Introduction", should appear in all capital
%letters, bold face if possible, centered in the column, with one blank line
%before, and one blank line after. Use a period (".") after the heading number,
%not a colon.
%
%\subsection{Subheadings}
%\label{ssec:subhead}
%
%Subheadings should appear in lower case (initial word capitalized) in
%boldface.  They should start at the left margin on a separate line.
% 
%\subsubsection{Sub-subheadings}
%\label{sssec:subsubhead}
%
%Sub-subheadings, as in this paragraph, are discouraged. However, if you
%must use them, they should appear in lower case (initial word
%capitalized) and start at the left margin on a separate line, with paragraph
%text beginning on the following line.  They should be in italics.
%
%\section{PRINTING YOUR PAPER}
%\label{sec:print}
%
%Print your properly formatted text on high-quality, 8.5 x 11-inch white printer
%paper. A4 paper is also acceptable, but please leave the extra 0.5 inch (12 mm)
%empty at the BOTTOM of the page and follow the top and left margins as
%specified.  If the last page of your paper is only partially filled, arrange
%the columns so that they are evenly balanced if possible, rather than having
%one long column.
%
%In LaTeX, to start a new column (but not a new page) and help balance the
%last-page column lengths, you can use the command ``$\backslash$pagebreak'' as
%demonstrated on this page (see the LaTeX source below).
%
%\section{PAGE NUMBERING}
%\label{sec:page}
%
%Please do {\bf not} paginate your paper.  Page numbers, session numbers, and
%conference identification will be inserted when the paper is included in the
%proceedings.
%
%\section{ILLUSTRATIONS, GRAPHS, AND PHOTOGRAPHS}
%\label{sec:illust}
%
%Illustrations must appear within the designated margins.  They may span the two
%columns.  If possible, position illustrations at the top of columns, rather
%than in the middle or at the bottom.  Caption and number every illustration.
%All halftone illustrations must be clear black and white prints.  Do not use
%any colors in illustrations.
%
%Since there are many ways, often incompatible, of including images (e.g., with
%experimental results) in a LaTeX document, below is an example of how to do
%this \cite{Lamp86}.
%
%% Below is an example of how to insert images. Delete the ``\vspace'' line,
%% uncomment the preceding line ``\centerline...'' and replace ``imageX.ps''
%% with a suitable PostScript file name.
%% -------------------------------------------------------------------------
%\begin{figure}[htb]
%
%\begin{minipage}[b]{1.0\linewidth}
%  \centering
%% \centerline{\epsfig{figure=image1.ps,width=8.5cm}}
%  \vspace{2.0cm}
%  \centerline{(a) Result 1}\medskip
%\end{minipage}
%%
%\begin{minipage}[b]{.48\linewidth}
%  \centering
%% \centerline{\epsfig{figure=image3.ps,width=4.0cm}}
%  \vspace{1.5cm}
%  \centerline{(b) Results 3}\medskip
%\end{minipage}
%\hfill
%\begin{minipage}[b]{0.48\linewidth}
%  \centering
%% \centerline{\epsfig{figure=image4.ps,width=4.0cm}}
%  \vspace{1.5cm}
%  \centerline{(c) Result 4}\medskip
%\end{minipage}
%%
%\caption{Example of placing a figure with experimental results.}
%\label{fig:res}
%%
%\end{figure}
%
%% To start a new column (but not a new page) and help balance the last-page
%% column length use \vfill\pagebreak.
%% -------------------------------------------------------------------------
%\vfill
%\pagebreak
%
%
%\section{FOOTNOTES}
%\label{sec:foot}
%
%Use footnotes sparingly (or not at all!) and place them at the bottom of the
%column on the page on which they are referenced. Use Times 9-point type,
%single-spaced. To help your readers, avoid using footnotes altogether and
%include necessary peripheral observations in the text (within parentheses, if
%you prefer, as in this sentence).
%
%
%\section{COPYRIGHT FORMS}
%\label{sec:copyright}
%
%You must include your fully completed, signed IEEE copyright release form when
%you submit your paper. We {\bf must} have this form before your paper can be
%published in the proceedings.  The copyright form is available as a Word file,
%a PDF file, and an HTML file. You can also use the form sent with your author
%kit.
%
%\section{REFERENCES}
%\label{sec:ref}
%
%List and number all bibliographical references at the end of the paper.  The references can be numbered in alphabetic order or in order of appearance in the document.  When referring to them in the text, type the corresponding reference number in square brackets as shown at the end of this sentence \cite{C2}.
%
%% References should be produced using the bibtex program from suitable
%% BiBTeX files (here: strings, refs, manuals). The IEEEbib.bst bibliography
%% style file from IEEE produces unsorted bibliography list.
%% -------------------------------------------------------------------------
%\bibliographystyle{IEEEbib}
%\bibliography{strings,refs}

\end{document}

### 2012-03-13 SiPS2012/sips2012.tex
started writing the sips article

@@
 % --------------------
-\def\x{{\mathbf x}}
-\def\L{{\cal L}}
+%\def\x{{\mathbf x}}
+%\def\L{{\cal L}}
+
+%%% My commands
+\newcommand{\mat}[1]{\mathbf{#1}}
+\renewcommand{\vec}[1]{\mathbf{#1}}
 
@@
 \section{Introduction}\label{sec:intro}
+Introduction to Capon adaptive beamforming, ultrasound imaging and GPU.
+
+Delay and sum beamformer (DAS):
+\begin{align}
+z[n] = \sum_{m = 0}^{M-1}\vec{w}_m^*\vec{x}_m[n - \Delta_m[n]] = \vec{w}^H\vec{x}[n],
+\end{align}
+where $M$ is the number of elements.
+
+Capon weights: 
+\begin{align}
+\vec{w}[n] = \frac{\mat{R}[n]^{-1}\vec{a}}{\vec{a}^H\mat{R}[n]^{-1}\vec{a}},
+\end{align}
+where $\vec{a} = \vec{1}$ when $\vec{x}$ are pre-delayed.
+
+Estimation of $\mat{R}[n]$: $\mat{\hat{R}}[n] = \vec{x}[n]\vec{x}[n]^H$.
+
+In order to get a well conditioned $\mat{\hat{R}}$, avoid signal cancellation and DAS-like speckle, $\mat{\hat{R}}$ has to be averaged over $L$ subarrays and $N_{avg}$ time samples.
+\begin{align}
+\mat{\hat{R}} = \frac{\sum_{n_{avg}=-N_{avg}}^{N_{avg}} \sum_{l=0}^{M-L+1} \vec{x}_l[n+n_{avg}]\vec{x}_l[n+n_{avg}]^H}{(2N_{avg} + 1)(M-L+1)},
+\end{align}
+where $\vec{x}_l$ is the $l^\text{th}$ subarray $[x_l[n], \dotso x_{l+L}[n]]$.
+
+\section{Method}\label{sec:meth}
+\subsection{Building the Sample Covariance Matrix}
+Building $R$ using a sliding window approach across the subarrays.
+We have a limited amount of fast, near-core memory on the GPU.
+On NVIDIA architecture this memory is known as shared memory, and is restricted to 48KB per compute block.
+The maximum block size of 32x32, further restrict how large arrays we can handle inside a single compute block.
+
+\subsection{Solving Small Linear Systems}
+Gauss Jordan elimination to solve $\mat{R}^{-1}\vec{1}$.
+
+\section{Results}\label{sec:res}
+
+Present images before and after Capon weights has been applied. Present graph showing running times for Cython-Capon v.s. CUDA-Capon for different choices of parameters.
+
+\section{Discussion}\label{sec:dis}
+
+\section{Conclusion}\label{sec:con}
 

### 2012-03-15 SiPS2012/sips2012.tex
updated version of sips abstract. Sips is now terminated.

@@
 %\def\L{{\cal L}}
+
+%%% My packages
+\usepackage{amsfonts}
 
@@
 \section{Introduction}\label{sec:intro}
-Introduction to Capon adaptive beamforming, ultrasound imaging and GPU.
+Introduction to Capon adaptive beamforming, ultrasound imaging and GPU/CUDA.
 
@@
 
-Capon weights: 
+Capon weights are found by solving the following minimization problem:
+\begin{align}
+&\min_{\vec{w}} E\{|z[n]|^2\} = \vec{w}^H\mat{R}\vec{w}\\
+&\text{subjected to } \vec{w}^H\vec{1} = 1,
+\end{align}
+where $\mat{R} = E\{\vec{x}[n]\vec{x}[n]^H\}$.
+
+The solution to this minimization problem is
 \begin{align}
@@
 \end{align}
-where $\vec{a} = \vec{1}$ when $\vec{x}$ are pre-delayed.
+where $\vec{a} = \vec{1}$ when $\vec{x}$ are pre-delayed and $\vec{w} \in \mathbb{C}^M$.
 
@@
 
-In order to get a well conditioned $\mat{\hat{R}}$, avoid signal cancellation and DAS-like speckle, $\mat{\hat{R}}$ has to be averaged over $L$ subarrays and $N_{avg}$ time samples.
+In order to get a well conditioned $\mat{\hat{R}}$, avoid signal cancellation and to get DAS-like speckle, $\mat{\hat{R}}$ has to be averaged over $L\le M/2$ subarrays and $N_{avg}$ time samples.
 \begin{align}
-\mat{\hat{R}} = \frac{\sum_{n_{avg}=-N_{avg}}^{N_{avg}} \sum_{l=0}^{M-L+1} \vec{x}_l[n+n_{avg}]\vec{x}_l[n+n_{avg}]^H}{(2N_{avg} + 1)(M-L+1)},
+\mat{\hat{R}} = \frac{1}{(2N_{avg} + 1)K}\sum_{n'=n-N_{avg}}^{n+N_{avg}} \sum_{l=0}^{K} \vec{x}_l[n']\vec{x}_l[n']^H,
 \end{align}
-where $\vec{x}_l$ is the $l^\text{th}$ subarray $[x_l[n], \dotso x_{l+L}[n]]$.
+where $K = M-L+1$, $Y_{avg}\sim \tau/T_s$ ($\tau$ is the imaging pulse length and $T_s$ is the sampling period.) and $\vec{x}_l$ is the $l^\text{th}$ subarray $[x_l[n], \dotso x_{l+L}[n]]$.
+
+Building and calculating the inverse of $\hat{R}$ is computational demanding, limiting the methods accessibility. Following the innovation in GPU computing, this has now started to change.
 
@@
 \subsection{Building the Sample Covariance Matrix}
-Building $R$ using a sliding window approach across the subarrays.
+Building $\hat{\mat{R}}$ using a sliding window approach across $K$ and $N_{avg}$.
 We have a limited amount of fast, near-core memory on the GPU.
@@
 The maximum block size of 32x32, further restrict how large arrays we can handle inside a single compute block.
+We therefore restrict $L$ to a maxima of $32$ elements.
+Since $L \le M/2$, $M \le 64$.
+We can not afford to hold the full $\hat{\mat{R}}$ ($M = L$) when $\hat{\mat{R}} > 32$, because of limitied amount of memory.
+...
+Explain in detail how \texttt{buildR} is implemented in a kernel.
+...
+Give more details on how much shared memory we have, and how it can be divided between compute blocks occupying one stream multiprocessor (SM).
 
-\subsection{Solving Small Linear Systems}
-Gauss Jordan elimination to solve $\mat{R}^{-1}\vec{1}$.
+\subsection{Solving Multiple Small Linear Systems}
+Most focus on large matrices in the literature.
+We have used an unreleased GPU implementation of Gauss Jordan (GJ) elimination (by NVIDIA) to solve $\mat{R}\vec{b} = \vec{1}$, where $\vec{b} = \mat{R}^{-1}\vec{1}$.
+Can include details on GPU implementation of $\mat{U}^H\mat{D}\mat{U}$, but this has not proved to be faster than NVIDIA's GJ implementation.
+However, the complexity for solving with $\mat{U}^H\mat{D}\mat{U}$ decomposition is supposed to be $1/3$ of GJ.
 
@@
 
-Present images before and after Capon weights has been applied. Present graph showing running times for Cython-Capon v.s. CUDA-Capon for different choices of parameters.
+Present images before and after Capon weights has been applied.
+Comment on the effect of selecting different parameters.
+Present graph showing running times for Cython-Capon v.s.
+CUDA-Capon for different choices of parameters.
+
+(A prerequisite for this is to get the GPU-processing up and running on simulated data before Easter!! Maybe also a Matlab plug-in.)
 
 \section{Discussion}\label{sec:dis}
+Discuss results presented in Section \ref{sec:res}.
 
 \section{Conclusion}\label{sec:con}
+...
 

### 2012-04-11 GPUCapon/gpu_capon.tex
la2pdf commit - 2012:04:11	14:41:02	(Wed11.Apr)

--- 2012-03-09 LCAforActiveSonarImaging/lca.tex
@@
 \fi
+
+% \RequirePackage[latin1]{inputenc}%              % Set input encoding (optionally latin1)
+% \RequirePackage[T1]{fontenc}%                 % Set font encoding
+% \usepackage[norsk]{babel}
 
@@
 
+\usepackage[maxfloats=25]{morefloats}
 \newcounter{todoidx}
@@
    \definecolor{todobackground}{rgb}{0.95,0.95,0.95}
-   \setlength\marginparsep{3pt}
-   \setlength\marginparwidth{42pt}
+   \setlength\marginparsep{1pt}
+   \setlength\marginparwidth{35pt}
    \newlength\marginparwidthsmall
@@
          {\vspace*{-10pt}\color{Red}\fbox{\bf\thetodoidx{}}}\\%
-         \fcolorbox{red}{todobackground}{\parbox{\marginparwidthsmall}{#1}}}}
+         \fcolorbox{red}{todobackground}{\parbox{\marginparwidthsmall}{\scriptsize #1}}}}
 
@@
 \fi
+
+\newenvironment{narrow}[2]{%
+\begin{list}{}{%
+\setlength{\topsep}{0pt}%
+\setlength{\leftmargin}{#1}%
+\setlength{\rightmargin}{#2}%
+\setlength{\listparindent}{\parindent}%
+\setlength{\itemindent}{\parindent}%
+\setlength{\parsep}{\parskip}}%
+\item[]}{\end{list}}
+
+\usepackage{float}
 
@@
 \newcommand*\eR{\mat{\hat R}}
+\newcommand*\eRi{\hat{\mat R}\,\!^{-1}}
+\newcommand*\Navg{N_\text{avg}}
 \newcommand*\W{\mat W}
@@
 
-% \RequirePackage[utf8]{inputenc}%              % Set input encoding (optionally latin1)
-% \RequirePackage[T1]{fontenc}%                 % Set font encoding
+% \usepackage{yfonts}
 
@@
 
-\title{Low Complexity Adaptive Beamformer\\ for Active Sonar Imaging}
+\title{A GPU Implementation of Capon for an Active Sonar System}
 
 \author{Jo~Inge~Buskenes, %
-        Andreas~Austeng, %
-        Carl-Inge~Columbo~Nilsen%
+        Jon~Petter~Aasen, %
+        Carl-Inge~Colombo~Nilsen%
 \IEEEcompsocitemizethanks{\IEEEcompsocthanksitem All authors are with the Department
@@
 \markboth{IEEE Journal of Oceanic Engineering}%
-{Low Complexity Adaptive Beamformer for Active Sonar Imaging}
+{A GPU Implementation of Capon for an Active Sonar System}
 
@@
 \begin{itemize}
-\item Dynamic weights $\Rightarrow$ greater contrast/resolution potential
-\item But:
-\begin{itemize}
-\item Active system $\Rightarrow$ noise and signal correlated $\Rightarrow$ adaptive beamformers no longer robust.
-\item Subarray averaging decorrelate the noise, but in the process we sacrifice resolution.
-\item Computationally complex O($M^3$) due to inversion of a spatial covariance matrix.
-LCA is only O($MW$).
-Even if Capon subspace/beamspace versions exist, they remain more complex than the LCA, and is not as ideally suited for implementation of SIMD hardware (GPUs, DSPs, ...).
-\end{itemize}
-\item LCA is inherently robust, is ``parameter free'' once setup with a well designed window set, and can often be implemented with minor modifications to already existing DAS beamformers.
+\item Todo.
 \end{itemize}
@@
 \begin{IEEEkeywords}
-Beamforming, adaptive beamforming, sonar, active, complexity.
+Beamforming, adaptive beamforming, sonar, active, complexity. \todo{not entirely true?}
 \end{IEEEkeywords}
@@
 
+\todopar{Case:
 \begin{itemize}
-\item Sell it!
+\item UFFC(?)
+\item Ultrasound. Arrays up to 128 channels. Processing power vital. What's the best way to implement 
+\item Other articles evaluates Capon image quality, and some have shown that speedups can be achieved by reducing the complexity of the traditional Capon algorithm or by implementing specific versions on a GPU.
+\item This article compare a few promising ways to implement Capon in general, in light of general complexity and suitability for implementation on a CPU and GPU.
+\end{itemize}
+}
+
+\begin{itemize}
 \item As briefly as possible, introduce beamforming.
@@
 Constraints must be applied in one way or another - parameters must be tuned.
+\item The three main ways to implement Capon \cite{Capon1969} is \todo{not entirely true?}
 \begin{itemize}
-\item Noise and signal is correlated. Spatial averaging required. 
-\item Increases variance of speckle (not only in active systems?), spatial compunting \cite{Vignon2009} (US) or time averaging?.
-Subarray averaging applied here \cite{Synnevag2007} (US) \todo{Is it bad to mix ultrasound/ sonar refs?}
+\item Traditional way: Build $\eR$ and solve $\w = \frac{\Ri\a}{\a\H\Ri\a}$. 
+\item Iterative methods: Woodbury, Conjugated gradients
+\item Beamspace
+\item LCA (or other methods, as reference)
 \end{itemize}
-\item LCA is like DAS, only we test several windows and choose the one that yield the lowest beamformer output power.
-Ultrasound \cite{Synnevag2008}, active sonar \cite{Buskenes2011}\cite{Blomberg2011}\cite{Blomberg2012}\todo{who was that guy in the 80ties?}
-\item Outline: 
+\item Choice of robustification parameters $L,M,\Delta$ largely impacts what the ``optimal'' way of implementing Capon 
+\item In most imaging applications 
+Outline:
 \begin{itemize}
-\item Methods.
-\item See what happens to Capon's response %\cite{Synnevag2007}  when means for robustification are added.
-\item Setup the window set for LCA to match this.
-\item Notice that the LCA performs like Capon, even with a small window database.
-\item Start with Kaiser, see that other window functions also work well.
+\item Speeding up: Build $\eR$ and solve $\w = \frac{\Ri\a}{\a\H\Ri\a}$. 
+\item Iterative methods: Woodbury (Sherman Morrison), Conjugated gradients - show performance of GPU performance.
+\item Woodbury: Must take the average over many samples in space and time - lots of (small) outerproducts. Can be used if not a lot of averaging is required.
+\item Beamspace
+\item LCA (or other methods, as reference)
 \end{itemize}
 \end{itemize}
-\todopar{Out of the blue idea: Weigh CF with LCA? This guy did it with MVDR \cite{Wang2009}. Probably you know already... Okay, back on track.}
 
 \section{Methods}
-
-\todopar{Threw some formulas in to get a feel of them}
 
@@
 
-\section{Results}
+\subsection{Traditional approach}
 
-% \begin{figure}[!t]\centering
-% % \includegraphics[width=2.5in]{}
-% \caption{Simulation Results}
-% \label{fig_sim}
-% \end{figure}
+Let us first consider the \emph{computation of $\hat{\mat R}$}.
+Notice that when implementing the formula directly a lot of summations and multiplications will be repeated unnecessary.
+To illustrate this, consider for instance a case with $M=6$ sensors that are grouped into subarrays of length $L=3$ where all but one element overlap.
+Let us further assume that we need to average over 7 time samples, i.e.
+that $N_{\text{avg}} = 3$.
 
-\input{gfx/gfx_input}
-% \begin{figure*}[!t]
-% \centerline{\subfloat[Case I]\includegraphics[width=2.5in]{subfigcase1}%
-% \label{fig_first_case}}
-% \hfil
-% \subfloat[Case II]{\includegraphics[width=2.5in]{subfigcase2}%
-% \label{fig_second_case}}}
-% \caption{Simulation results}
-% \label{fig_sim}
-% \end{figure*}
+These procedures are commutative, so the order in which we perform them does not matter.
+But for the sake of this example let us start by performing the time averaging.
+Assuming that $\vec x_i$ and $\vec x_j$ have a zero mean, a time averaged covariance coefficient $r_{ij}$ in $\hat{\mat R}$ may be obtained by (omitting the normalisation):
+\begin{align*}
+r_{ij}[n] &= \sumb{n'=n-N_{\text{avg}}}{n+N_{\text{avg}}} x_i[n']\,x_j[n'],
+\end{align*}
+which can be easily rewritten in a sliding window fashion
+\begin{align*}
+r_{ij}[n] &= r_{ij}[n-1] - x_i[n\text{-}N_\text{avg}\text{-}1]\,x_j[n\text{-}N_\text{avg}\text{-}1]
+                         + x_i[n\text{+}N_\text{avg}]\,x_j[n\text{+}N_\text{avg}].
+\end{align*}
+Hence, whenever $N_\text{avg}>1$ the sliding window approach will demand less instructions at the expense of a slightly higher memory consumption.
+The next thing to realise is that not all $r_{ij}[n]$ need to be computed.
+Since $r_{ji} = r^*_{ij}$, and the subarrays only span $L$ elements, we only compute an upper triangular band of R (marked \Green{green})
+\begin{figure}[H]
+% \begin{narrow}{-3.5cm}{-3.5cm}
+\includegraphics[width=\linewidth]{gfx/capon_build_R_full.eps}
+\caption{Building $\hat{\mat R}$: Symmetry and element coverage}\label{buildingRfull}
+% \end{narrow}
+\end{figure}
+The final step is to average all the subarray matrices to form $\hat{\mat R}$ by 
+{\renewcommand*{\arraystretch}{1.8}
+\begin{align*}
+\hat{\mat R}_\text{sub} &= {\color{Blue}\mat R_{\text{sub},1} + \mat R_{\text{sub},2} + \mat R_{\text{sub},3} + \mat R_{\text{sub},4}} \\
+&= \left[\;
+\begin{matrix}
+\color{Green}r_{00}+r_{11}+r_{22} & \color{Red}r_{01}+r_{12}+r_{23} & r_{02}+r_{13}+r_{24} \\
+                                       & \color{Green}r_{11}+r_{22}+r_{33} & \color{Red}r_{12}+r_{23}+r_{34} \\
+                                       &                                        & \color{Green}r_{22}+r_{33}+r_{44}
+\end{matrix}
+\;\right]
+\end{align*}}%
+Note that consecutive elements along the diagonals overlap on all but one covariance coefficient, hence a sliding window may also be used here.
+The subarray size $L$ is usually quite a bit larger than the time averaging window, so there is more to gain from implementing the sliding window here.\\\\
+
+{\bf Method complexity:}\\
+\begin{tabular}{l l}
+No optimizations            & $2M^2\,(2\Navg+1)$ \\
+Banded symmetric            & $L(2M-L+1)\,(2\Navg+1)$ \\
+Sliding, banded, symmetric  & $1.5L(2M-L+1)$ \\
+\end{tabular}\\\\
+
+\subsubsection{Computing $\text{\raisebox{1.9pt}{$\vec\chi$}} = \hat{\mat R}\,\!^{-1}\vec a$}
+
+There are two ways of finding $\text{\raisebox{1.9pt}{$\vec\chi$}} = \hat{\mat R}\,\!^{-1}\vec a$, either by solving the equation directly or by inverting $\hat{\mat R}$ first.
+Solving the equation is less computationally intensive than inverting $\hat{\mat R}$, and seems like the obvious choice.
+However, with iterative methods for updating $\eRi$ for each range pixel the situation is changed...
+
+\subsection{Iterative methods}
+
+\begin{itemize}
+\item Woodbury
+\item Conjugated gradients
+\end{itemize}
+
+\subsection{Beamspace}
+
+\begin{itemize}
+\item Some minimum of background, and good references.
+\end{itemize}
+
+\section{Results/Discussion}
+
+\begin{itemize}
+\item Compare computational complexity vs. memory constraints for the various methods. Supporting plots.
+\item Relate this to whether the favourable architecture is a CPU or a GPU \todo{check ``when to gpu article''}
+\item Run benchmarks on some sonar dataset(s)
+\end{itemize}
 
@@
 \begin{itemize}
-\item Something similar to the abstract.
+\item When to use what.
 \end{itemize}
@@
 %
-%\begin{figure*}[!t]
-%\centerline{\subfloat[Case I]\includegraphics[width=2.5in]{subfigcase1}%
-%\label{fig_first_case}}
-%\hfil
-%\subfloat[Case II]{\includegraphics[width=2.5in]{subfigcase2}%
-%\label{fig_second_case}}}
-%\caption{Simulation results}
-%\label{fig_sim}
-%\end{figure*}
+% \begin{figure*}[!t]
+% \centerline{\subfloat[Case I]\includegraphics[width=2.5in]{subfigcase1}%
+% \label{fig_first_case}}
+% \hfil
+% \subfloat[Case II]{\includegraphics[width=2.5in]{subfigcase2}%
+% \label{fig_second_case}}}
+% \caption{Simulation results}
+% \label{fig_sim}
+% \end{figure*}
 %
@@
 
+\markboth{}%
+{}
+% \input{gfx/gfx_input}
+
 \end{document}

### 2012-04-12 GPUCapon/gpu_capon.tex
la2pdf commit - 2012:04:12	14:52:28	(Thu12.Apr)

@@
 
+The use of phased arrays in modern sonar imaging systems allow each sensor to be independently delayed and weighted 
+Delays and weights are applied to each of the sensor channels to ensure that signals emanating from a pixel of interest add constructively, while noise and interference from other regions add destructively.
+This method of adjusting the spatial focus of the antenna is commonly referred to as a beamformer.
+A
+
+Beamformer algorithms are distinction
+
+The Minimum Variance One of the most well established adaptive
+
+One of the main issues with adaptive beamformers is their computational complexity, and .
+
+Most modern active sonar imaging systems make use of phased arrays to transmit and receive wavefields.
+On reception, each pixel in the image is estimated by digitally steering the arrays response to the corresponding region in space, a concept called beamforming.
+This is achieved by applying a suitable set of delays and weights to each of the sensor channels to ensure that signals coming from the region of interest add constructively, while attempting to make sure noise and interference from all other regions cancel out.
+
+If a beamformer only employ a static set of weights and delays, it is referred to as a conventional one.
+Adaptive beamformers, on the other hand, attempts to find an optimal spatial response that best fits the incoming wavefield.
+The most common approach is perform a rough adjustment using a static set of delays, and compute a new set of weights at each time instant.
+
+Conventional beamformers employ a static set of weights and delays,
+
 \todopar{Case:
 \begin{itemize}
-\item UFFC(?)
-\item Ultrasound. Arrays up to 128 channels. Processing power vital. What's the best way to implement 
+\item JOE
+\item Sonar. Channel count up to 32/64 ish channels?
 \item Other articles evaluates Capon image quality, and some have shown that speedups can be achieved by reducing the complexity of the traditional Capon algorithm or by implementing specific versions on a GPU.

### 2012-04-23 SiPS2012/sips2012.tex
Done some wrighting on the sips/ius/uffc article

@@
 
+Finally, $\mat{\hat{R}}$ is loaded with diagonal factor $\epsilon$, $\mat{\hat{R}} = \mat{\hat{R}} + \epsilon\mat{I}$.
+A weighting proportional to the output power is beneficial, and the trace of $\mat{\hat{R}}$, $\epsilon = d*tr\{\mat{\hat{R}}\}$, has previously shown good results in the literature (cite).
+
 Building and calculating the inverse of $\hat{R}$ is computational demanding, limiting the methods accessibility. Following the innovation in GPU computing, this has now started to change.
@@
 The maximum block size of 32x32, further restrict how large arrays we can handle inside a single compute block.
-We therefore restrict $L$ to a maxima of $32$ elements.
+It is therefore natural to restrict $L$ to a maxima of $32$ elements.
 Since $L \le M/2$, $M \le 64$.
-We can not afford to hold the full $\hat{\mat{R}}$ ($M = L$) when $\hat{\mat{R}} > 32$, because of limitied amount of memory.
+We can not afford to hold the full $\hat{\mat{R}}$ ($M = L$) when $\hat{\mat{R}} > 32$, because of limited amount of memory.
 ...
@@
 Give more details on how much shared memory we have, and how it can be divided between compute blocks occupying one stream multiprocessor (SM).
+
+Discuss the bottleneck adaptive diagonal loading causes.
+
+Time averaging takes a lot of resources.
+It has been shown that time averaging can maintain both resolution and delay-and-sum-like speckle.
+The same speckle statistics can be obtain with small subarrays, but then we loose resolution.
+Small sub-arrays benefits from being less computational demanding.
 
@@
 However, the complexity for solving with $\mat{U}^H\mat{D}\mat{U}$ decomposition is supposed to be $1/3$ of GJ.
+
+In the final journal article we could discuss Conjugated gradient and Woodbury and the benefits they bring.
 

### 2012-04-23 SiPS2012/sips2012.tex
Added a section about beamspace prosessing to the sips/ius/uffc article

@@
 \end{align}
-where $\vec{a} = \vec{1}$ when $\vec{x}$ are pre-delayed and $\vec{w} \in \mathbb{C}^M$.
+where $\vec{a} = \vec{1}$ when $\vec{x}$ are pre-delayed to broadside,  and $\vec{w} \in \mathbb{C}^M$.
 
@@
 Building and calculating the inverse of $\hat{R}$ is computational demanding, limiting the methods accessibility. Following the innovation in GPU computing, this has now started to change.
+
+Nilsen et.
+al has proposed to apply beamspace processing to further reduce the computations required.
+With beamspace we mean transforming the channel data from element space to a fan of beams covering the imaging sector, $\vec{x}_{BS} = \mat{B}x$.
+The steering vectors in $\mat{B}$, the so-called Buttler matrix, determines each beams direction in space.
+We can therefore easily remove beams where no interference is present by removing rows in $\mat{B}$.
+For a focused system like medical ultrasound, where the received signal is concentrated in a narrow band around broadside, there is a small percentage of the beams that contains almost all of the energy.
+Nilsen et.
+al has shown that as little as three beams still produced results comparable with applying the full Capon estimation in element space.
 
@@
 
-Discuss the bottleneck adaptive diagonal loading causes.
+Discuss the bottleneck adaptive diagonal loading causes. Possible solutions includes, use trace from previous image, use pre-calculated array power, constant weight.
 

### 2012-04-24 SiPS2012/sips2012.tex
added a section about the cuda compute model in the sips/ius/uffc article

@@
 Small sub-arrays benefits from being less computational demanding.
+
+\subsubsection{CUDA Compute Model}
+The CUDA compute model is based on execution of a kernel across a grid of compute threads.
+Each position in the grid holds a block of threads.
+Each block is further divided into groups of 32 threads, known as a warp.
+The maximum number of threads inside a block is restricted to 512, hence 16 warps.
+From a CUDA perspective the GPU consist of one or more stream multi processors (SM), where each SM is capable of 32 or more, depending on the architecture, concurrent multiplications and/or additions.
+The CUDA 2.x compute capability (CC) has available 48 KB of near-core shared memory per SM that can be utilized by resident blocks.
+The maximum number of resident blocks per SM is 8, but in addition the maximum number of resident warps per SM is restricted to 48.
+Hence, with a block size of 512 only $48/(512/32) = 3$ blocks can occupy the SM at once.
+In addition, if one thread where to use 64B of shared memory the number of allowed resident blocks with a block size of 512 would be reduced to 1.
+Up to a curtain point it is beneficial to have as many warps as possible occupying the SM at once to hide latency from accessing memory and executing time consuming mathematical functions.
+The SM has for CC 2.x 32K 32-bit long registers.
+When implementing algorithms on the GPU it is important to keep data as close to the core as possible.
+Utilizing available registers and shared memory to its maximum is therefore important.
+
+\subsubsection{Optimizing kernel launch}
 

### 2012-04-26 GPUCapon/gpu_capon.tex
...

--- 2012-04-11 GPUCapon/gpu_capon.tex
@@
 
+Delays and weights are applied to each of the sensor channels to ensure that signals emanating from a pixel of interest add constructively, while noise and interference from other regions add destructively.
+This method of adjusting the spatial focus of the antenna is commonly referred to as a beamformer.
+A
+
+Beamformer algorithms are distinction
+
+The Minimum Variance One of the most well established adaptive
+
+One of the main issues with adaptive beamformers is their computational complexity, and .
+
+Most modern active sonar imaging systems record and process the incoming wavefield through the use of phased arrays.
+A pixel value in the image is estimated by digitally steering the arrays response to the corresponding region in space.
+This concept is called beamforming and involves applying a suitable set of delays and weights to each of the sensor channels to cause constrictive interference of the signals coming from the region of interest, and destructive interference of anything else.
+Ideally, we want a spatial response
+
+If a beamformer only employ a static set of weights and delays, it is referred to as a conventional one.
+Applying , which  Various weight sets, or windows, may be added, Adaptive beamformers, on the other hand, attempts to find an optimal spatial response that best fits the incoming wavefield.
+The most common approach is perform a rough adjustment using a static set of delays, and compute a new set of weights at each time instant.
+
 \todopar{Case:
 \begin{itemize}
-\item UFFC(?)
-\item Ultrasound. Arrays up to 128 channels. Processing power vital. What's the best way to implement 
+\item JOE
+\item Sonar. Channel count up to 32/64 ish channels?
 \item Other articles evaluates Capon image quality, and some have shown that speedups can be achieved by reducing the complexity of the traditional Capon algorithm or by implementing specific versions on a GPU.

### 2012-04-26 ECUA2012/gpu_capon_ecua2012.tex
la2pdf commit - 2012:04:26	15:36:11	(Thu26.Apr)

--- 2012-04-26 GPUCapon/gpu_capon.tex
@@
 
-%%%%%%%%%%%%                                    ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
-% IEEEtran %
-%%%%%%%%%%%%                                    ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
+%%%%%%%%%%%%%%                                  ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
+% ECUA style %
+%%%%%%%%%%%%%%                                  ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
 
-\ifPeerReview
-\documentclass[12pt,journal,captionsoff,onecolumn]{IEEEtran}
-\newcommand\CLASSINPUTbaselinestretch{1.66}     % http://theoval.cmp.uea.ac.uk/~nlct/latex/thesis/node17.html
-\else
-\documentclass[journal]{IEEEtran}
-\fi
-
-% \RequirePackage[latin1]{inputenc}%              % Set input encoding (optionally latin1)
-% \RequirePackage[T1]{fontenc}%                 % Set font encoding
-% \usepackage[norsk]{babel}
+\documentclass[10pt,a4paper]{article}
+\usepackage{template/ecua}
 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%                  ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
-% IEEE ''APPROVED'' PACKAGES %
+% ECUA ''APPROVED'' PACKAGES %
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%                  ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
 
-\ifCLASSINFOpdf
-   \usepackage[dvips]{graphicx}                 % Might not work. Use 'latex' instead of 
-   \ifFlatArchive\else                          % 'pdflatex'
-      \graphicspath{./gfx/}
-   \fi
-\else
-   \usepackage[dvips]{graphicx}
-   \ifFlatArchive\else
-      \graphicspath{./gfx/}
-   \fi
-\fi
-
-\RequirePackage[table,dvipsnames,svgnames]{xcolor}
-
-\usepackage[cmex10]{amsmath}                    % cmex10 option to be IEEE explore compliant
-\interdisplaylinepenalty=2500                   % Allows multiline equations to be broken
-
-% \RequirePackage{amssymb}
-
-\RequirePackage{array}
-
-\ifCLASSOPTIONcompsoc
-   \usepackage[caption=false,font=normalsize,labelfont=sf,textfont=sf]{subfig}
-\else
-   \usepackage[caption=false,font=footnotesize]{subfig}
-\fi
-\ifCLASSOPTIONcaptionsoff                       % IEEE promoted hack to turn off captions from the 
-   \let\MYorigsubfloat\subfloat                 % subfloat package should the captionsoff option
-   \renewcommand{\subfloat}[2][\relax]{\MYorigsubfloat[]{#2}} % be specified.
-\fi
-
-\ifFloatAtEnd
-\ifCLASSOPTIONcaptionsoff                       % Places float at the end of the document when the
-  \usepackage[nomarkers]{endfloat}              % captionsoff options is specified to IEEEtrans.cls
-  \let\MYoriglatexcaption\caption               % (PeerReview mode)
-  \renewcommand{\caption}[2][\relax]{\MYoriglatexcaption[#2]{#2}}
-\fi
-\fi
-
-\usepackage{fixltx2e}                           % Fix some twocolumn float problems
-
-%\usepackage{stfloats}                          % Allows: \begin{figure*}[!b]
-                                                % (double column figures on top/bottom)
-
-\usepackage{url}                                % Support for handling and breaking URLs
-
-% NOTE: PDF hyperlink and bookmark features are not required in IEEE
-%       papers and their use requires extra complexity and work.
-\newcommand\MYhyperrefoptions{bookmarks=true,bookmarksnumbered=true,
-pdfpagemode={UseOutlines},plainpages=false,pdfpagelabels=true,
-colorlinks=true,linkcolor={black},citecolor={black},urlcolor={black},
-pdftitle={Low Complexity Adaptive Beamformer for Active Sonar Imaging},
-pdfsubject={},
-pdfauthor={Jo Inge Buskenes},
-pdfkeywords={adaptive beamforming, beamforming, complexity, sonar, active}}%
-\ifCLASSINFOpdf
-   \usepackage[\MYhyperrefoptions,pdftex]{hyperref}
-\else
-   \usepackage[\MYhyperrefoptions,breaklinks=true,dvips]{hyperref}
-   \usepackage{breakurl}                        % Allows 'dvips' driver to break links
-\fi
+% load some other useful packages
+\usepackage{cite} % improved formatting for citation lists
+\usepackage{graphicx}
 
@@
 
-\usepackage[maxfloats=25]{morefloats}
+\usepackage[table,dvipsnames,svgnames]{xcolor}
 \newcounter{todoidx}
-% \setcounter{todoidx}
 
@@
 
-% \usepackage{yfonts}
+\title{A GPU Implementation of Capon for an Active Sonar System}
+
+\author{%
+\begin{tabular}{p{30mm}p{125mm}}
+Jo~Inge~Buskenes & Department of Informatics, University of Oslo, Norway \\
+Jon~Petter~Aasen & Department of Informatics, University of Oslo, Norway \\
+Carl-Inge~Colombo~Nilsen & Department of Informatics, University of Oslo, Norway \\
+\end{tabular}
+}
 
@@
 
-\title{A GPU Implementation of Capon for an Active Sonar System}
-
-\author{Jo~Inge~Buskenes, %
-        Jon~Petter~Aasen, %
-        Carl-Inge~Colombo~Nilsen%
-\IEEEcompsocitemizethanks{\IEEEcompsocthanksitem All authors are with the Department
-of Informatics, University of Oslo, Norway.}% <-this % stops a space
-
-% \thanks{Manuscript received April 19, 2005; revised January 11, 2007.}
-}
-
-% The paper headers
-\markboth{IEEE Journal of Oceanic Engineering}%
-{A GPU Implementation of Capon for an Active Sonar System}
-
-% Publishers ID mark:
-%\IEEEpubid{0000--0000/00\$00.00~\copyright~2007 IEEE}
-
-% use for special paper notices
-%\IEEEspecialpapernotice{(Invited Paper)}
-
-% for Computer Society papers, we must declare the abstract and index terms
-% PRIOR to the title within the \IEEEcompsoctitleabstractindextext IEEEtran
-% command as these need to go into the title area created by \maketitle.
-\IEEEcompsoctitleabstractindextext{%
-\begin{abstract}
-
-\begin{itemize}
-\item Todo.
-\end{itemize}
-
-\end{abstract}
-
-% Keywords (normally not used for peer reviews)
-\ifPeerReview\else
-\begin{IEEEkeywords}
-Beamforming, adaptive beamforming, sonar, active, complexity. \todo{not entirely true?}
-\end{IEEEkeywords}
-\fi}
-% \fi
-
-% make the title area
 \maketitle
 
-% This command fixes abstract positioning for compsoc articles:
-\IEEEdisplaynotcompsoctitleabstractindextext
+\section{Introduction}
 
-% (Optional) Add some extra info on cover page of peer review papers:
-% \ifCLASSOPTIONpeerreview
-% \begin{center} \bfseries EDICS Category: 3-BBND \end{center}
-% \fi
-
-% Insert page break and insert second title (peer review mode)
-\IEEEpeerreviewmaketitle
-
-\section{Introduction}
+When forming images from an active sonar system employing a phased array, the lateral \gls{SNR} and resolution will largely depend on the choice of weights applied to each of the array channels. 
+The Capon beamformer has been around for quite som time now. It is adaptive
 
@@
 
-\subsubsection{Computing $\text{\raisebox{1.9pt}{$\vec\chi$}} = \hat{\mat R}\,\!^{-1}\vec a$}
+% \subsubsection{Computing $\text{\raisebox{1.9pt}{$\vec\chi$}} = \hat{\mat R}\,\!^{-1}\vec a$}
 
@@
 
-\appendices
-
-% use section* for acknowledgement
-\ifCLASSOPTIONcompsoc
 \section*{Acknowledgments}
-\else
-  \section*{Acknowledgment}
-\fi
 
@@
 
-% Can use something like this to put references on a page
-% by themselves when using endfloat and the captionsoff option.
-\ifCLASSOPTIONcaptionsoff
-  \newpage
-\fi
-
-% trigger a \newpage just before the given reference
-% number - used to balance the columns on the last page
-% adjust value as needed - may need to be readjusted if
-% the document is modified later
-%\IEEEtriggeratref{8}
-% The "triggered" command can be changed if desired:
-%\IEEEtriggercmd{\enlargethispage{-5in}}
-
-% references section
-
-\bibliographystyle{IEEEtran}
+\bibliographystyle{./ecua}
 \bibliography{../../Library/library}
 
-% End up doing this:
-% \begin{thebibliography}{1}
-% 
-% \bibitem{IEEEhowto:kopka}
-% H.~Kopka and P.~W. Daly, \emph{A Guide to {\LaTeX}}, 3rd~ed.\hskip 1em plus
-%   0.5em minus 0.4em\relax Harlow, England: Addison-Wesley, 1999.
-% 
-% \end{thebibliography}
-
-% biography section
-% 
-% If you have an EPS/PDF photo (graphicx package needed) extra braces are
-% needed around the contents of the optional argument to biography to prevent
-% the LaTeX parser from getting confused when it sees the complicated
-% \includegraphics command within an optional argument. (You could create
-% your own custom macro containing the \includegraphics command to make things
-% simpler here.)
-%\begin{biography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{mshell}}]{Michael Shell}
-% or if you just want to reserve a space for a photo:
-
-\vfill
-
-\begin{IEEEbiography}{Noname dude}
-He doesn't exist.
-\end{IEEEbiography}
-
-% if you will not have a photo at all:
-\begin{IEEEbiographynophoto}{Jo Inge Buskenes}
-Not in at the moment.
-\end{IEEEbiographynophoto}
-
-%\newpage
-
-% \begin{IEEEbiographynophoto}{Jo Inge Buskenes}
-% Not in at the moment.
-% \end{IEEEbiographynophoto}
-
-%\vfill
-
-% Can be used to pull up biographies so that the bottom of the last one
-% is flush with the other column.
-%\enlargethispage{-5in}
-
-\markboth{}%
-{}
-% \input{gfx/gfx_input}
-
 \end{document}

### 2012-05-02 ECUA2012/gpu_capon_ecua2012.tex
la2pdf commit - 2012:05:02	16:47:50	(Wed 2.May)

@@
 
+\usepackage{glossaries}
 \usepackage[table,dvipsnames,svgnames]{xcolor}
@@
    \setlength\marginparwidthsmall{\marginparwidth}
-   \addtolength\marginparwidthsmall{-7pt}
+   \addtolength\marginparwidthsmall{+6mm}
    \newcommand\todo[1]{%
@@
 
+\newcommand\defeq{\overset{\underset{\mathrm{def}}{}}{=}}
+
 %Math macros:
@@
 
+\newcommand\1{\vec 1}
+\newcommand\I{\mat I}
 \renewcommand*\a{\vec a}
@@
 
+%%%%%%%%%%%%                                    ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
+% GLOSSARY %
+%%%%%%%%%%%%                                    ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
+
+\makeglossaries
+\input{../../Library/glossary}
+
 %%%%%%%%%%%%%%%%%%                              ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
@@
 \author{%
-\begin{tabular}{p{30mm}p{125mm}}
+\begin{tabular}{p{45mm}p{125mm}}
 Jo~Inge~Buskenes & Department of Informatics, University of Oslo, Norway \\
-Jon~Petter~Aasen & Department of Informatics, University of Oslo, Norway \\
+Jon~Petter~Aasen & Mi Lab, Norwegian University of Science and Technology, Norway \\
 Carl-Inge~Colombo~Nilsen & Department of Informatics, University of Oslo, Norway \\
+Andreas-Austeng & Department of Informatics, University of Oslo, Norway
 \end{tabular}
@@
 
-When forming images from an active sonar system employing a phased array, the lateral \gls{SNR} and resolution will largely depend on the choice of weights applied to each of the array channels. 
-The Capon beamformer has been around for quite som time now. It is adaptive
+% Images from a phased array active sonar system are formed in the post-processing by focusing on one pixel at a time.
+This is achieved by delaying and weighting each of the array's channels, providing coarse and fine adjustments to the region of focus, respectively.
+This is commonly referred to as beamforming.
+Assuming far-field conditions, the delays are usually chosen to steer the pixel of interest to broadside, and the weights chosen to reject the noise impinging on the array from non-broadside directions.
+% 
+% The algorithm responsible for carrying out this task is referred to as a beamformer.
+Depending on whether data is evaluated when choosing a suitable set of delays and weights, beamformers can be branded either conventional or adaptive.
+A good example of the former is the The Delay-and-Sum (DAS) beamformer, which - as the name implies - simply delay the sensor channels appropriately prior to summing them up.
+Applying various weight sets, or windows, to the data prior to summation allows the lateral \gls{SNR} to be improved, but at the inevitable expense of a deteriorated lateral resolution.
+\todo{aka.
+the curse of physics.}
+% 
+% The Minimum Variance Distortionless Response (MVDR) (or Capon) beamformer, have the potential to overcome this limitation.
+While ensuring unit gain in the look direction, it computes the set of weights that minimizes the energy accumulated by the array from other directions.
+In other words, whenever the distribution of noise and interference is primarily focused around specific angles, an array response with a high degree of suppression at these angles are chosen, resulting in an image with a lateral \gls{SNR} and resolution superior to anything the \gls{DAS} can come up with.
 
-Delays and weights are applied to each of the sensor channels to ensure that signals emanating from a pixel of interest add constructively, while noise and interference from other regions add destructively.
-This method of adjusting the spatial focus of the antenna is commonly referred to as a beamformer.
-A
+The image of modern phased array active sonar systems suffer from poor resolution and \gls{SNR} laterally.
+This is because the lateral extent of the array is too small compared to the wavelength of the received signal, and while it is possible to trade resolution for improved \gls{SNR} by applying various weightsets to the array channels, one always end up with a compromise between the two.
+By introducing adaptive techniques, however, it is possible to resolve lateral details that would otherwise have been missed.
 
-Beamformer algorithms are distinction
+The \gls{MVDR} (or Capon) beamformer is one such technique.
+While ensuring unit gain in the look direction, it computes the set of weights that minimizes the energy accumulated by the array from other directions.
+Whenever the distribution of noise and interference is primarily focused around specific angles, an array response with a high degree of suppression at these angles are chosen, resulting in an image with much improved lateral detail level.
 
-The Minimum Variance One of the most well established adaptive
+However, despite its inherent potential the \gls{MVDR} beamformer has yet to see widespread adoption in the active sonar community.
+There may be several reasons for this.
+First, the method is not inherently robust, and will suffer from a phenomena called signal cancellation in active systems.
+Second, in its original form, the computational complexity is cubic with the number of channels, O(M$^3$), while \gls{DAS} is only at O($M$).
 
-One of the main issues with adaptive beamformers is their computational complexity, and .
+Perhaps surprisingly, we will show that the suggested means for making the \gls{MVDR} behave well in active systems will also make it less computationally complex.
+Furthermore, by implementing it on a modern \gls{GPU} we have gained a speedup of 2-3 orders of magnitude compared to optimized single thread Matlab and C implementations.
 
-Most modern active sonar imaging systems record and process the incoming wavefield through the use of phased arrays.
-A pixel value in the image is estimated by digitally steering the arrays response to the corresponding region in space.
-This concept is called beamforming and involves applying a suitable set of delays and weights to each of the sensor channels to cause constrictive interference of the signals coming from the region of interest, and destructive interference of anything else.
-Ideally, we want a spatial response
+% - Who's done stuff on GPU beamforming before?
+% 
+% - Complex data
+% -
 
-If a beamformer only employ a static set of weights and delays, it is referred to as a conventional one.
-Applying , which  Various weight sets, or windows, may be added, Adaptive beamformers, on the other hand, attempts to find an optimal spatial response that best fits the incoming wavefield.
-The most common approach is perform a rough adjustment using a static set of delays, and compute a new set of weights at each time instant.
-
-\todopar{Case:
-\begin{itemize}
-\item JOE
-\item Sonar. Channel count up to 32/64 ish channels?
-\item Other articles evaluates Capon image quality, and some have shown that speedups can be achieved by reducing the complexity of the traditional Capon algorithm or by implementing specific versions on a GPU.
-\item This article compare a few promising ways to implement Capon in general, in light of general complexity and suitability for implementation on a CPU and GPU.
-\end{itemize}
-}
-
-\begin{itemize}
-\item As briefly as possible, introduce beamforming.
-\item Adaptive beamformer's potential lies in its ability to suppress interference power
-\item Why adaptive beamformers struggle in active sonar systems.
-Correlated noise, robustification kills the adaptive potential.
-Quite computationally intensive.
-Constraints must be applied in one way or another - parameters must be tuned.
-\item The three main ways to implement Capon \cite{Capon1969} is \todo{not entirely true?}
-\begin{itemize}
-\item Traditional way: Build $\eR$ and solve $\w = \frac{\Ri\a}{\a\H\Ri\a}$. 
-\item Iterative methods: Woodbury, Conjugated gradients
-\item Beamspace
-\item LCA (or other methods, as reference)
-\end{itemize}
-\item Choice of robustification parameters $L,M,\Delta$ largely impacts what the ``optimal'' way of implementing Capon 
-\item In most imaging applications 
-Outline:
-\begin{itemize}
-\item Speeding up: Build $\eR$ and solve $\w = \frac{\Ri\a}{\a\H\Ri\a}$. 
-\item Iterative methods: Woodbury (Sherman Morrison), Conjugated gradients - show performance of GPU performance.
-\item Woodbury: Must take the average over many samples in space and time - lots of (small) outerproducts. Can be used if not a lot of averaging is required.
-\item Beamspace
-\item LCA (or other methods, as reference)
-\end{itemize}
-\end{itemize}
-
+\newpage
 \section{Methods}
 
+Consider an $M$ element phased array.
+Let the sample recorded by the $m$'th channel at time instant $n$ be represented as $x_m[n]$, and the accompanying weights and delays be  $w_m[n]$ and $\Delta_m$.
+The beamformer output $z[n]$ is then given as
 \begin{align}
-z[n] = \sumb{m=0}{M-1} w_m[n]^*x_m[n-\Delta_m] = \w\H[n]\Xd[n]
+z[n] = \w\H[n]\X[n] = \bmat{w_0[n]\\w_1[n]\\\vdots\\w_{M-1}[n]}^H \bmat{x_0[n-\Delta_0]\\x_1[n-\Delta_1]\\\vdots\\x_{M-1}[n-\Delta_{M-1}]}.
 \end{align}
-where
+The \gls{MVDR} beamformer \cite{Capon1969} finds the set of weights that minimizes the output power $z[n]$, while ensuring unity gain in the look direction.
+Assuming predelayed data, the solution to this optimization problem is given as
+\begin{gather*}
+\vec w[n] = \frac{\eRi[n]\1}{\1\T\eRi[n]\1} = \frac{\vec\beta[n]}{\1\T\vec\beta[n]},
+\end{gather*}
+where $\1$ is a row vector of ones, and $\eR$ is an estimate of the spatial covariance matrix.
+To ensure robustness in an active sonar setting, we perform subarray averaging in order to decorrelate noise from the signal, and some degree of temporal averaging to retain \gls{DAS}-like speckle statistics.
+The $(i,j)$'th element of the sample covariance matrix, $\hat r_{ij}$, is then given as
+\begin{gather*}
+r_{ij}[n] = \frac{1}{(M-L+1)(2N_{\text{avg}}+1)}\sumb{k=0}{M-L} \sumb{n'=n-N_{\text{avg}}}{n+N_{\text{avg}}} x_{i+k}[n']\,x_{j+k}[n'],
+\end{gather*}
+where $r_{ij}$ is the $(i,j)$'th element of the covariance matrix, $L$ is the subarray size, and $2N_{\text{avg}}+1$ is the number of temporal samples to perform averaging over.
+Finally, to make sure that $\eR$ is well conditioned, we add a fraction $d$ of the total power of $\eR$ to its diagonal\cite{Synnevag2007},
 \begin{align}
-\w[n] = \bmat{w_0[n]\\w_1[n]\\\vdots\\w_{M-1}[n]} \quad \text{and} \quad\Xd[n] = \bmat{x_0[n]\\x_1[n]\\\vdots\\x_{M-1}[n]}.
-\end{align}
-
-Minimum variance distortionless resonse \cite{Capon1969}
-\begin{align}
-\underset{\w[n]}{\argmin}\, E\{ |z[n]|^2 \} = \underset{\w[n]}{\argmin}\, \w\H[n]\R[n]\w[n], 
-\end{align}
-subject to
-\begin{align}
-\w\H[n]\a = 1,
-\end{align}
-where
-\begin{align}
-\R[n] = E\{ \x\x\H \}.
-\end{align}
-If $\eR$ is the estimate of $\R$, the solution to the MVDR beamformer is
-\begin{gather*}
-\vec w[n] = \frac{\hat{\mat R}\,\!^{-1}[n]\vec a}{\vec a\H\hat{\mat R}\,\!^{-1}[n]\vec a} = \frac{\text{\raisebox{1.9pt}{$\vec\chi$}}[n]}{\vec a\H\text{\raisebox{1.9pt}{$\vec\chi$}}[n]}
-\end{gather*}
-A robust estimate of $\eR$ is found by
-\begin{itemize}
-\item Spatial averaging to decorrelate noise from signal. \emph{Subarray averaging} used here.
-\item Temporal averaging to ensure valid speckle statistics.
-\item \emph{Diagonal loading} to ensure  numerical stability prior to inversion.
-\item Choice of robustification parameters \cite{Synnevag2007}
-\end{itemize}
-
-Here $\vec a$ is a \emph{steering vector} that if set to $\vec 1$ steers towards broadside, and $\hat{\mat R}$ is an estimate of the spatial covariance matrix, often computed as
-\begin{gather*}
-r_{ij}[n] = \frac{1}{K(2N_{\text{avg}}+1)}\sumb{k=0}{K-1} \sumb{n'=n-N_{\text{avg}}}{n+N_{\text{avg}}} x_{i+k}[n']\,x_{j+k}[n'],
-\end{gather*}
-where $r_{ij}$ is the $(i,j)$'th element of the covariance matrix, $K$ is the number of subarrays, $2N_{\text{avg}}+1$ is the number of samples to perform time averaging over, and $\vec x_i[n']$ is the data recorded by the $i$'th sensor at sample $n'$.
-
-\emph{Diagonal loading} chosen as \cite{Synnevag2007}
-\begin{align}
-\varepsilon = \Delta \cdot \tr\{\eR[n]\}
+\eR = \eR + \I \cdot d \cdot \tr\{\eR[n]\}.
 \end{align}
@@
 \begin{align*}
-r_{ij}[n] &= r_{ij}[n-1] - x_i[n\text{-}N_\text{avg}\text{-}1]\,x_j[n\text{-}N_\text{avg}\text{-}1]
+\hat r_{ij}[n] &= r_{ij}[n-1] - x_i[n\text{-}N_\text{avg}\text{-}1]\,x_j[n\text{-}N_\text{avg}\text{-}1]
                          + x_i[n\text{+}N_\text{avg}]\,x_j[n\text{+}N_\text{avg}].
@@
 
-\subsection{Iterative methods}
-
-\begin{itemize}
-\item Woodbury
-\item Conjugated gradients
-\end{itemize}
-
-\subsection{Beamspace}
-
-\begin{itemize}
-\item Some minimum of background, and good references.
-\end{itemize}
-
 \section{Results/Discussion}
@@
 \begin{itemize}
-\item Compare computational complexity vs. memory constraints for the various methods. Supporting plots.
-\item Relate this to whether the favourable architecture is a CPU or a GPU \todo{check ``when to gpu article''}
-\item Run benchmarks on some sonar dataset(s)
+\item Execution time plots. For given of M/L, plot for different Yavg (note negligible diagonal loading).
+\item An image? Holmengrå?
 \end{itemize}
@@
 \item When to use what.
+\end{itemize}
+
+\todopar{Case:
+\begin{itemize}
+\item JOE
+\item Sonar. Channel count up to 32/64 ish channels?
+\item Other articles evaluates Capon image quality, and some have shown that speedups can be achieved by reducing the complexity of the traditional Capon algorithm or by implementing specific versions on a GPU.
+\item This article compare a few promising ways to implement Capon in general, in light of general complexity and suitability for implementation on a CPU and GPU.
+\end{itemize}
+}
+
+\begin{itemize}
+\item As briefly as possible, introduce beamforming.
+\item Adaptive beamformer's potential lies in its ability to suppress interference power
+\item Why adaptive beamformers struggle in active sonar systems.
+Correlated noise, robustification kills the adaptive potential.
+Quite computationally intensive.
+Constraints must be applied in one way or another - parameters must be tuned.
+\item The three main ways to implement Capon \cite{Capon1969} is \todo{not entirely true?}
+\begin{itemize}
+\item Traditional way: Build $\eR$ and solve $\w = \frac{\Ri\a}{\a\H\Ri\a}$. 
+\item Iterative methods: Woodbury, Conjugated gradients
+\item Beamspace
+\item LCA (or other methods, as reference)
+\end{itemize}
+\item Choice of robustification parameters $L,M,\Delta$ largely impacts what the ``optimal'' way of implementing Capon 
+\item In most imaging applications 
+Outline:
+\begin{itemize}
+\item Speeding up: Build $\eR$ and solve $\w = \frac{\Ri\a}{\a\H\Ri\a}$. 
+\item Iterative methods: Woodbury (Sherman Morrison), Conjugated gradients - show performance of GPU performance.
+\item Woodbury: Must take the average over many samples in space and time - lots of (small) outerproducts. Can be used if not a lot of averaging is required.
+\item Beamspace
+\item LCA (or other methods, as reference)
+\end{itemize}
 \end{itemize}
@@
 
-\bibliographystyle{./ecua}
+\bibliographystyle{./template/ecua}
 \bibliography{../../Library/library}

### 2012-05-03 ECUA2012/gpu_capon_ecua2012.tex
la2pdf commit - 2012:05:03	13:14:22	(Thu 3.May)

@@
 
-The image of modern phased array active sonar systems suffer from poor resolution and \gls{SNR} laterally.
+The image of modern phased array active sonar systems suffer from poor lateral resolution and \gls{SNR}.
 This is because the lateral extent of the array is too small compared to the wavelength of the received signal, and while it is possible to trade resolution for improved \gls{SNR} by applying various weightsets to the array channels, one always end up with a compromise between the two.
@@
 
-However, despite its inherent potential the \gls{MVDR} beamformer has yet to see widespread adoption in the active sonar community.
+However, despite its inherent potential, the \gls{MVDR} beamformer has yet to see widespread adoption in the active sonar community.
 There may be several reasons for this.
@@
 Perhaps surprisingly, we will show that the suggested means for making the \gls{MVDR} behave well in active systems will also make it less computationally complex.
-Furthermore, by implementing it on a modern \gls{GPU} we have gained a speedup of 2-3 orders of magnitude compared to optimized single thread Matlab and C implementations.
+Furthermore, by implementing it on a modern \gls{GPU} we have gained a speedup of 2-3 orders of magnitude compared to optimized single thread implementations in C and Matlab.
 
@@
 \end{align}
-The \gls{MVDR} beamformer \cite{Capon1969} finds the set of weights that minimizes the output power $z[n]$, while ensuring unity gain in the look direction.
+The \gls{MVDR} beamformer \cite{Capon1969} finds the set of weights that minimizes the output power, $\min E\{|z[n]|^2\}$, while ensuring unity gain in the look direction.
 Assuming predelayed data, the solution to this optimization problem is given as
-\begin{gather*}
+\begin{gather}
 \vec w[n] = \frac{\eRi[n]\1}{\1\T\eRi[n]\1} = \frac{\vec\beta[n]}{\1\T\vec\beta[n]},
-\end{gather*}
+\end{gather}
 where $\1$ is a row vector of ones, and $\eR$ is an estimate of the spatial covariance matrix.
@@
 The $(i,j)$'th element of the sample covariance matrix, $\hat r_{ij}$, is then given as
-\begin{gather*}
+\begin{gather}
 r_{ij}[n] = \frac{1}{(M-L+1)(2N_{\text{avg}}+1)}\sumb{k=0}{M-L} \sumb{n'=n-N_{\text{avg}}}{n+N_{\text{avg}}} x_{i+k}[n']\,x_{j+k}[n'],
-\end{gather*}
+\end{gather}
 where $r_{ij}$ is the $(i,j)$'th element of the covariance matrix, $L$ is the subarray size, and $2N_{\text{avg}}+1$ is the number of temporal samples to perform averaging over.
@@
 
-\subsection{Traditional approach}
+\section{Implementation}
+
+A massively parallel hardware such as \glspl{GPU} have a potential for accelerating a great many signal processing tasks, but fully exploiting this potential is not always trivial.
+Modern \gls{GPU} chips may contain hundreds of processing cores, but the amount of memory available locally for core is very limited.
+There is also a considerable latency involved in moving data between the \gls{CPU} and these cores, but as the data transfer is asynchronous this latency is largely hidden as long as the threads running on each core are computationally intensive.
+
+thread = computationally intense and low memory consumption.
+consecutive threads should have linear memory access patterns
+all threads should have the same single line of execution
+
+down to these cores and back to the \gls{CPU} is also when copying data to and from the \gls{GPU} are also rather high, so each core should be   performing calculations there, and copying the data back to the \gls{CPU}
+
+ but in order to fully exploit this potential we must be aware of its limitations.
+
+When adapting an algorithm to a massively parallel hardware such as \glspl{GPU}, one must be familiar with its limitations in order to that the potential that lies within these chips can not be fully realised
+
+When trying to fit any algorithm to massively parallel hardware such as \glspl{GPU}, one must be careful to select a proper decomposition scheme for the device at hand.
+Modern \glspl{GPU} have several hundred computing cores, but only a limited amount of  are  various decomposition schemes should be contemplated before performing the implecareful decomposition of the
 

### 2012-05-03 ECUA2012/gpu_capon_ecua2012.tex
la2pdf commit - 2012:05:03	13:56:31	(Thu 3.May)

@@
 
+\todo{Get some references in here...}
 % - Who's done stuff on GPU beamforming before?
@@
 all threads should have the same single line of execution
+
+\section{Results}
+
+Benchmarks were carried out on a test system comprised of a quad-core Intel Xeon E5620 running at 2.4GHz, 64Gb of \gls{RAM}, and an nVidia Quadro 6000 card. This \gls{GPU}
+
+To test the \gls{CUDA} implementation of the \gls{MVDR} beamformer, we have chosen a dataset from the $M=32$ element Kongsberg Maritime HISAS1030 sonar~\cite{Hansen2009}. \todo{Data???}.
 

### 2012-05-03 ECUA2012/gpu_capon_ecua2012.tex
la2pdf commit - 2012:05:03	14:30:15	(Thu 3.May)

@@
 
-A massively parallel hardware such as \glspl{GPU} have a potential for accelerating a great many signal processing tasks, but fully exploiting this potential is not always trivial.
-Modern \gls{GPU} chips may contain hundreds of processing cores, but the amount of memory available locally for core is very limited.
-There is also a considerable latency involved in moving data between the \gls{CPU} and these cores, but as the data transfer is asynchronous this latency is largely hidden as long as the threads running on each core are computationally intensive.
+Massively parallel hardware such as \glspl{GPU} is  have a potential for accelerating a great many signal processing tasks, but fully exploiting their potential is not always trivial.
+Modern \gls{GPU} chips may contain hundreds of processing cores, but the penalty of such a design is that the amount of memory available locally for each core is very limited.
+There is also a considerable latency involved when moving data between the \gls{CPU} and these cores, so assessing various memory access schemes is crucial to achieve optimal performance.
 
-thread = computationally intense and low memory consumption.
-consecutive threads should have linear memory access patterns
+Hence, the programming strategy should be to decompose the algorithm into as many threads as possible, each being computationally intense and be light on the memory consumption and access.
+Consecutive threads should also access consecutive memory regions, to enable coalesced data transfers, and have the same line of execution.
+
+If we evaluate the \gls{MVDR} beamformer, the two computationally dominating steps is to
+
+\begin{itemize}
+ \item asf
+\end{itemize}
+
 all threads should have the same single line of execution
@@
 
-Benchmarks were carried out on a test system comprised of a quad-core Intel Xeon E5620 running at 2.4GHz, 64Gb of \gls{RAM}, and an nVidia Quadro 6000 card. This \gls{GPU}
+Benchmarks were carried out on a test system comprised of a quad-core Intel Xeon E5620 running at 2.4GHz, 64Gb of \gls{RAM}, and an nVidia Quadro 6000 card.
+This is a \gls{GPU} with 448 \gls{CUDA} cores, 6Gb of onboard \gls{RAM},
 

### 2012-05-03 ECUA2012/gpu_capon_ecua2012.tex
la2pdf commit - 2012:05:03	15:00:49	(Thu 3.May)

@@
 Consider an $M$ element phased array.
-Let the sample recorded by the $m$'th channel at time instant $n$ be represented as $x_m[n]$, and the accompanying weights and delays be  $w_m[n]$ and $\Delta_m$.
+Let the sample recorded by the $m$'th channel at time instant $n$ be represented as $x_m[n]$, and the accompanying weights and delays be  $w_m[n]$ and $\Delta_m[n]$.
 The beamformer output $z[n]$ is then given as
@@
 The \gls{MVDR} beamformer \cite{Capon1969} finds the set of weights that minimizes the output power, $\min E\{|z[n]|^2\}$, while ensuring unity gain in the look direction.
-Assuming predelayed data, the solution to this optimization problem is given as
+The solution to this optimization problem is given as
 \begin{gather}
-\vec w[n] = \frac{\eRi[n]\1}{\1\T\eRi[n]\1} = \frac{\vec\beta[n]}{\1\T\vec\beta[n]},
+\vec w[n] = \frac{\Ri[n]\1}{\1\T\Ri[n]\1},
 \end{gather}
-where $\1$ is a row vector of ones, and $\eR$ is an estimate of the spatial covariance matrix.
+where $\1$ is a row vector of ones, and $\R$ is the spatial covariance matrix. 
+\begin{gather}
+R[n] = \frac{1}{2N_\text{avg}+1} \sumb{n'=n-N_\text{avg}}{n+N_\text{avg}} \x[n']\x\H['n]
+\end{gather}
+
 To ensure robustness in an active sonar setting, we perform subarray averaging in order to decorrelate noise from the signal, and some degree of temporal averaging to retain \gls{DAS}-like speckle statistics.
@@
 \end{itemize}
+
+\begin{gather}
+\vec w[n] = \frac{\Ri[n]\1}{\1\T\Ri[n]\1} = \frac{\vec\beta[n]}{\1\T\vec\beta[n]},
+\end{gather}
 

### 2012-05-03 ECUA2012/gpu_capon_ecua2012.tex
la2pdf commit - 2012:05:03	15:49:59	(Thu 3.May)

@@
 
-The image of modern phased array active sonar systems suffer from poor lateral resolution and \gls{SNR}.
-This is because the lateral extent of the array is too small compared to the wavelength of the received signal, and while it is possible to trade resolution for improved \gls{SNR} by applying various weightsets to the array channels, one always end up with a compromise between the two.
+Modern phased array active sonar systems are often limited by its lateral resolution and \gls{SNR} of the system.
+This is because of the the lateral extent of the array is too small compared to the wavelength of the received signal, and while it is possible to trade resolution for improved \gls{SNR} by applying various weightsets to the array channels, one always end up with a compromise between the two.
 By introducing adaptive techniques, however, it is possible to resolve lateral details that would otherwise have been missed.
@@
 \begin{align}
-z[n] = \w\H[n]\X[n] = \bmat{w_0[n]\\w_1[n]\\\vdots\\w_{M-1}[n]}^H \bmat{x_0[n-\Delta_0]\\x_1[n-\Delta_1]\\\vdots\\x_{M-1}[n-\Delta_{M-1}]}.
+z[n] = \w\H[n]\x[n] = \bmat{w_0[n]\\w_1[n]\\\vdots\\w_{M-1}[n]}^H \bmat{x_0[n-\Delta_0]\\x_1[n-\Delta_1]\\\vdots\\x_{M-1}[n-\Delta_{M-1}]}.
 \end{align}
@@
 where $\1$ is a row vector of ones, and $\R$ is the spatial covariance matrix.
+Since $\R$ is unknown, it must be estimated.
+First, we split the segmenting the array into subarrays where all but one element overlap, and call the output vectors
 \begin{gather}
-R[n] = \frac{1}{2N_\text{avg}+1} \sumb{n'=n-N_\text{avg}}{n+N_\text{avg}} \x[n']\x\H['n]
+\x_l[n] = \bmat{x_l[n-\Delta_l] & x_{l+1}[n-\Delta_l] & \dots & x_{l+L-1}[n-\Delta_{l+L-1}]}\T.
+\end{gather}
+
+\begin{gather}
+\tilde\R[n] = \frac{1}{M-L+1} \sumb{i=0}{M-L+1} \x_{i+k}[n]\x_{i+k}\H[n]
+\end{gather}
+\begin{gather}
+\tilde\R[n] = \frac{1}{2N_\text{avg}+1} \sumb{n'=n-N_\text{avg}}{n+N_\text{avg}} \x[n']\x\H['n]
 \end{gather}

### 2012-05-03 ECUA2012/gpu_capon_ecua2012.tex
la2pdf commit - 2012:05:03	16:21:27	(Thu 3.May)

@@
 
-Modern phased array active sonar systems are often limited by its lateral resolution and \gls{SNR} of the system.
-This is because of the the lateral extent of the array is too small compared to the wavelength of the received signal, and while it is possible to trade resolution for improved \gls{SNR} by applying various weightsets to the array channels, one always end up with a compromise between the two.
+Modern phased array active sonar systems are often limited by their lateral resolution and \gls{SNR}, which is due to limited lateral extent of the array the spatial extent of the array sample the extent of the array is too small compared to the wavelength of the received signal, and while it is possible to trade resolution for improved \gls{SNR} by applying various weightsets to the array channels, one always end up with a compromise between the two.
 By introducing adaptive techniques, however, it is possible to resolve lateral details that would otherwise have been missed.
@@
 Since $\R$ is unknown, it must be estimated.
-First, we split the segmenting the array into subarrays where all but one element overlap, and call the output vectors
+We will do this by performing some degree of both spatial and temporal averaging.
+We start with the former, by segmenting the array into $M-L-1$ subarrays of length $L$, where all but one element overlap.
+The data vector $x_l[n]$ from subarray $l$ is then given as
 \begin{gather}
@@
 \end{gather}
-
+Then the average of all these subarrays are computed,
 \begin{gather}
@@
 \end{gather}
+Finally, 
 \begin{gather}

### 2012-05-03 ECUA2012/gpu_capon_ecua2012.tex
la2pdf commit - 2012:05:03	16:55:24	(Thu 3.May)

@@
 Since $\R$ is unknown, it must be estimated.
-We will do this by performing some degree of both spatial and temporal averaging.
+We will do this by performing some degree of spatial averaging to avoid signal cancellation, and temporal averaging to achieve \gls{DAS}-like speckle statistics.
 We start with the former, by segmenting the array into $M-L-1$ subarrays of length $L$, where all but one element overlap.
-The data vector $x_l[n]$ from subarray $l$ is then given as
+The data vector $x_l[n]$ from subarray $l$ is given as
 \begin{gather}
-\x_l[n] = \bmat{x_l[n-\Delta_l] & x_{l+1}[n-\Delta_l] & \dots & x_{l+L-1}[n-\Delta_{l+L-1}]}\T.
+\x_l[n] = \bmat{x_l[n-\Delta_l] & x_{l+1}[n-\Delta_l] & \dots & x_{l+L-1}[n-\Delta_{l+L-1}]}\T,
 \end{gather}
-Then the average of all these subarrays are computed,
+and the spatially averaged sample covariance matrix $\tilde R$ is given as
 \begin{gather}
-\tilde\R[n] = \frac{1}{M-L+1} \sumb{i=0}{M-L+1} \x_{i+k}[n]\x_{i+k}\H[n]
+\tilde\R[n] = \frac{1}{M-L+1} \sumb{i=0}{M-L+1} \x_{i+k}[n]\x_{i+k}\H[n].
 \end{gather}
-Finally, 
+Finally, we take the average of $2N_\text{avg}+1$ temporal samples to form the final sample covariance matrix $\eR$,
 \begin{gather}
-\tilde\R[n] = \frac{1}{2N_\text{avg}+1} \sumb{n'=n-N_\text{avg}}{n+N_\text{avg}} \x[n']\x\H['n]
+\eR[n] = \frac{1}{2N_\text{avg}+1} \sumb{n'=n-N_\text{avg}}{n+N_\text{avg}} \tilde\R[n'].
 \end{gather}
-
-To ensure robustness in an active sonar setting, we perform subarray averaging in order to decorrelate noise from the signal, and some degree of temporal averaging to retain \gls{DAS}-like speckle statistics.
-The $(i,j)$'th element of the sample covariance matrix, $\hat r_{ij}$, is then given as
-\begin{gather}
-r_{ij}[n] = \frac{1}{(M-L+1)(2N_{\text{avg}}+1)}\sumb{k=0}{M-L} \sumb{n'=n-N_{\text{avg}}}{n+N_{\text{avg}}} x_{i+k}[n']\,x_{j+k}[n'],
-\end{gather}
-where $r_{ij}$ is the $(i,j)$'th element of the covariance matrix, $L$ is the subarray size, and $2N_{\text{avg}}+1$ is the number of temporal samples to perform averaging over.
 Finally, to make sure that $\eR$ is well conditioned, we add a fraction $d$ of the total power of $\eR$ to its diagonal\cite{Synnevag2007},
@@
 
+% To ensure robustness in an active sonar setting, we perform subarray averaging in order to decorrelate noise from the signal, and some degree of temporal averaging to retain \gls{DAS}-like speckle statistics.
+The $(i,j)$'th element of the sample covariance matrix, $\hat r_{ij}$, is then given as
+% \begin{gather}
+% r_{ij}[n] = \frac{1}{(M-L+1)(2N_{\text{avg}}+1)}\sumb{k=0}{M-L} \sumb{n'=n-N_{\text{avg}}}{n+N_{\text{avg}}} x_{i+k}[n']\,x_{j+k}[n'],
+% \end{gather}
+% where $r_{ij}$ is the $(i,j)$'th element of the covariance matrix, $L$ is the subarray size, and $2N_{\text{avg}}+1$ is the number of temporal samples to perform averaging over.
+
 \section{Implementation}
 
-Massively parallel hardware such as \glspl{GPU} is  have a potential for accelerating a great many signal processing tasks, but fully exploiting their potential is not always trivial.
-Modern \gls{GPU} chips may contain hundreds of processing cores, but the penalty of such a design is that the amount of memory available locally for each core is very limited.
+Massively parallel hardware such as \glspl{GPU} have a potential for accelerating many signal processing tasks, but fully exploiting their potential is not always trivial.
+Modern \gls{GPU} chips may contain hundreds of processing cores, but the penalty of this design is that the amount of memory available locally for each core is very limited.
 There is also a considerable latency involved when moving data between the \gls{CPU} and these cores, so assessing various memory access schemes is crucial to achieve optimal performance.

### 2012-05-03 ECUA2012/gpu_capon_ecua2012.tex
la2pdf commit - 2012:05:03	17:31:18	(Thu 3.May)
la2pdf commit - 2012:05:03	19:52:08	(Thu 3.May)

@@
 \begin{align}
-z[n] = \w\H[n]\x[n] = \bmat{w_0[n]\\w_1[n]\\\vdots\\w_{M-1}[n]}^H \bmat{x_0[n-\Delta_0]\\x_1[n-\Delta_1]\\\vdots\\x_{M-1}[n-\Delta_{M-1}]}.
+z[n] = \w\H[n]\x[n] = \bmat{w_0[n]\\w_1[n]\\\vdots\\w_{M-1}[n]}^H \bmat{x_0[n-\Delta_0]\\x_1[n-\Delta_1]\\\vdots\\x_{M-1}[n-\Delta_{M-1}]}.\label{z}
 \end{align}
@@
 \begin{gather}
-\vec w[n] = \frac{\Ri[n]\1}{\1\T\Ri[n]\1},
+\vec w[n] = \frac{\Ri[n]\1}{\1\T\Ri[n]\1},\label{weights}
 \end{gather}
@@
 \end{gather}
-and the spatially averaged sample covariance matrix $\tilde R$ is given as
+and the spatially averaged sample covariance matrix, $\tilde R$, is given as
 \begin{gather}
-\tilde\R[n] = \frac{1}{M-L+1} \sumb{i=0}{M-L+1} \x_{i+k}[n]\x_{i+k}\H[n].
+\tilde\R[n] = \frac{1}{M-L+1} \sumb{l=0}{M-L+1} \x_l[n]\x_l\H[n].\label{spatialR}
 \end{gather}
@@
 \begin{align}
-\eR = \eR + \I \cdot d \cdot \tr\{\eR[n]\}.
+\eR[n] = \eR[n] + \I \cdot \frac{d}{L} \cdot \tr\{\eR[n]\}.\label{finalR}
 \end{align}
+To sum it up, for each range line, and for each sample $n$ in this line:
+\begin{enumerate}
+\item Compute the spatial covariance matrix $\eR[n]$ (equation \ref{spatialR}-\ref{finalR})
+\item Compute $\eRi[n]\1$ (from equation \ref{weights})
+\item Insert the result from step (2) into equation \ref{weights} to find the weights
+\item Insert the result from step (3) into equation \ref{z} to acquire the pixel value.
+\end{enumerate}
 
@@
 
-If we evaluate the \gls{MVDR} beamformer, the two computationally dominating steps is to
-
 \begin{itemize}
- \item asf
+ \item 
 \end{itemize}

### 2012-05-03 ECUA2012/gpu_capon_ecua2012.tex
la2pdf commit - 2012:05:03	21:23:13	(Thu 3.May)

@@
 \end{align}
-To sum it up, for each range line, and for each sample $n$ in this line:
+To sum it up, for each range line, and for each sample $n$ in this line we must
 \begin{enumerate}
-\item Compute the spatial covariance matrix $\eR[n]$ (equation \ref{spatialR}-\ref{finalR})
-\item Compute $\eRi[n]\1$ (from equation \ref{weights})
-\item Insert the result from step (2) into equation \ref{weights} to find the weights
+\item Compute the spatial covariance matrix $\eR[n]$ (equation \ref{spatialR}-\ref{finalR}).
+\item Compute $\eRi[n]\1$ (from equation \ref{weights}).
+\item Insert the result from step (2) into equation \ref{weights} to find the weights.
 \item Insert the result from step (3) into equation \ref{z} to acquire the pixel value.
@@
 Massively parallel hardware such as \glspl{GPU} have a potential for accelerating many signal processing tasks, but fully exploiting their potential is not always trivial.
-Modern \gls{GPU} chips may contain hundreds of processing cores, but the penalty of this design is that the amount of memory available locally for each core is very limited.
-There is also a considerable latency involved when moving data between the \gls{CPU} and these cores, so assessing various memory access schemes is crucial to achieve optimal performance.
+Modern \gls{GPU} chips contain several hundred processing cores, but the amount of memory available locally for each core is very limited.
+There is also a considerable latency involved when accessing on-board memory, so assessing various memory access schemes is crucial to achieve optimal performance.
 
-Hence, the programming strategy should be to decompose the algorithm into as many threads as possible, each being computationally intense and be light on the memory consumption and access.
+Hence, the programming strategy should be to decompose the algorithm into as many threads as possible, each being computationally intense and be light on memory access and consumption.
 Consecutive threads should also access consecutive memory regions, to enable coalesced data transfers, and have the same line of execution.
-
-\begin{itemize}
- \item 
-\end{itemize}
 
@@
 
-Benchmarks were carried out on a test system comprised of a quad-core Intel Xeon E5620 running at 2.4GHz, 64Gb of \gls{RAM}, and an nVidia Quadro 6000 card.
-This is a \gls{GPU} with 448 \gls{CUDA} cores, 6Gb of onboard \gls{RAM},
+Benchmarks were carried out on a test system comprised of a quad-core Intel Xeon E5620, 64Gb of \gls{RAM}, and an nVidia Quadro 6000 card.
+This is a \gls{GPU} with 448 \gls{CUDA} cores and 6Gb of onboard \gls{RAM}, capable of roughly 1Tflops.
 

### 2012-05-03 ECUA2012/gpu_capon_ecua2012.tex
la2pdf commit - 2012:05:03	21:58:21	(Thu 3.May)

@@
 
+Massively parallel hardware such as \glspl{GPU} have a potential for accelerating many signal processing tasks, and the \gls{MVDR} beamformer is no exception.
+Each pixel in the image can be processed independently, and
+
+Modern \gls{GPU} chips contain several hundred processing cores, but the amount of memory available locally for each core is very limited.\todo{repetition} There is also a considerable latency involved when accessing on-board memory, but these memory transfers are asynchronous and may be hidden.
+
 \todo{Get some references in here...}
@@
 \end{align}
-To sum it up, for each range line, and for each sample $n$ in this line we must
-\begin{enumerate}
-\item Compute the spatial covariance matrix $\eR[n]$ (equation \ref{spatialR}-\ref{finalR}).
-\item Compute $\eRi[n]\1$ (from equation \ref{weights}).
-\item Insert the result from step (2) into equation \ref{weights} to find the weights.
-\item Insert the result from step (3) into equation \ref{z} to acquire the pixel value.
-\end{enumerate}
 
@@
 
+By the methods just presented, we observe that for each range line, and for each sample $n$, the following steps must be carried out:
+\begin{enumerate}
+\item \emph{Compute the spatial covariance matrix} $\eR[n]$ (equation \ref{spatialR}-\ref{finalR}).
+\item Compute $\eRi[n]\1$ (from equation \ref{weights}).
+\item Substitute the solution for $\eRi[n]\1$ into equation \ref{weights} to find the weights.
+\item Apply the weights to each data vector $\x_l[n]$, and sum them all up (as in equation \ref{z}) to acquire the pixel value $z[n]$.
+\end{enumerate}
+
 Massively parallel hardware such as \glspl{GPU} have a potential for accelerating many signal processing tasks, but fully exploiting their potential is not always trivial.
-Modern \gls{GPU} chips contain several hundred processing cores, but the amount of memory available locally for each core is very limited.
-There is also a considerable latency involved when accessing on-board memory, so assessing various memory access schemes is crucial to achieve optimal performance.
+Modern \gls{GPU} chips contain several hundred processing cores, but the amount of memory available locally for each core is very limited.\todo{repetition} There is also a considerable latency involved when accessing on-board memory, but these memory transfers are asynchronous and may be hidden.
 
-Hence, the programming strategy should be to decompose the algorithm into as many threads as possible, each being computationally intense and be light on memory access and consumption.
+decompose the algorithm into as many threads as possible, each being computationally intense and be light on memory access and consumption.
 Consecutive threads should also access consecutive memory regions, to enable coalesced data transfers, and have the same line of execution.
+
+How may this be decomposed? Already decomposed at pixel-level.
+
+We implement the 
+Our implementation of th
+
+When
 

### 2012-05-03 ECUA2012/gpu_capon_ecua2012.tex
la2pdf commit - 2012:05:03	22:28:25	(Thu 3.May)

@@
 
+CUDA...
+
 Massively parallel hardware such as \glspl{GPU} have a potential for accelerating many signal processing tasks, and the \gls{MVDR} beamformer is no exception.
@@
 
-By the methods just presented, we observe that for each range line, and for each sample $n$, the following steps must be carried out:
+A modern \gls{GPU} is an hardware architecture comprised of several hundred computing cores, each having access to a very limited amount of local memory.
+This suggests that the algorithm should be decomposed into as many lightweight threads as possible, and using a single thread per pixel is not lightweight enough.
+Therefore, we seek to devise a clever decomposition scheme for the methods just presented.
+In particular, we want to distinguish between the following steps:
 \begin{enumerate}
-\item \emph{Compute the spatial covariance matrix} $\eR[n]$ (equation \ref{spatialR}-\ref{finalR}).
-\item Compute $\eRi[n]\1$ (from equation \ref{weights}).
-\item Substitute the solution for $\eRi[n]\1$ into equation \ref{weights} to find the weights.
-\item Apply the weights to each data vector $\x_l[n]$, and sum them all up (as in equation \ref{z}) to acquire the pixel value $z[n]$.
+\item \emph{Computing the spatial covariance matrix} $\eR[n]$ (equation \ref{spatialR}-\ref{finalR}).
+In this step we launch a thread per diagonal in the upper half triangular of $\eR$, and let finished threads wrap around and start on diagonals on the lower half triangular.
+This way we process one row of $\eR$ at the time, and both reads and writes are kept coalesced.
+\item \emph{Computing} $\eRi[n]\1$ (from equation \ref{weights}).
+While intuition may suggest that this step is carried out by inverting $\eR$, it is wiser to solve the linear equation $\R\beta = \1$ for $\beta$ instead, which gives us $\beta = \eRi\1$ directly.
+We have tested various solvers for this task, but achieved the best performance by using an unofficial batch solver from nVidia.
+After we optimized step 1, this remains at the most computationally intensive task.
+\item \emph{Computing the subarray weights} by substituting $\eRi[n]\1$ into equation \ref{weights}.
+\item \emph{Acquire the pixel value $z[n]$} by applying the weights to each data vector $\x_l[n]$, and sum them all up (as in equation \ref{z}).
 \end{enumerate}
+
+Our implementation is based on launching Since these steps have a different nature,
+
+\subsection{Computing $\eR[n]$}
 

### 2012-05-04 ECUA2012/gpu_capon_ecua2012.tex
la2pdf commit - 2012:05:04	11:23:46	(Fri 4.May)

@@
 
+\newpage
 \section{Implementation}
@@
 This suggests that the algorithm should be decomposed into as many lightweight threads as possible, and using a single thread per pixel is not lightweight enough.
-Therefore, we seek to devise a clever decomposition scheme for the methods just presented.
-In particular, we want to distinguish between the following steps:
+Therefore, we seek to devise decomposition scheme for the methods just presented.
+In particular, the solutions we came up with for the different steps in the algorithm, listed in their natural order of execution, was:
 \begin{enumerate}
 \item \emph{Computing the spatial covariance matrix} $\eR[n]$ (equation \ref{spatialR}-\ref{finalR}).
-In this step we launch a thread per diagonal in the upper half triangular of $\eR$, and let finished threads wrap around and start on diagonals on the lower half triangular.
-This way we process one row of $\eR$ at the time, and both reads and writes are kept coalesced.
+Here we launch a thread per diagonal in the upper half triangular of $\eR$, and let finished threads wrap around and start on diagonals on the lower half triangular.
+This way we process one row of $\eR$, and manage to keep both data reads and writes coalesced.
 \item \emph{Computing} $\eRi[n]\1$ (from equation \ref{weights}).
-While intuition may suggest that this step is carried out by inverting $\eR$, it is wiser to solve the linear equation $\R\beta = \1$ for $\beta$ instead, which gives us $\beta = \eRi\1$ directly.
+While intuition may suggest that this step is carried out by inverting $\eR$, it is better to solve the linear equation $\R\beta = \1$ for $\beta$ instead, which gives us $\beta = \eRi\1$ directly.
 We have tested various solvers for this task, but achieved the best performance by using an unofficial batch solver from nVidia.
-After we optimized step 1, this remains at the most computationally intensive task.
+This is by far the most computationally intensive task, and a key area of focus for further improvements.
 \item \emph{Computing the subarray weights} by substituting $\eRi[n]\1$ into equation \ref{weights}.
@@
 
-Our implementation is based on launching Since these steps have a different nature,
-
-\subsection{Computing $\eR[n]$}
-
-Massively parallel hardware such as \glspl{GPU} have a potential for accelerating many signal processing tasks, but fully exploiting their potential is not always trivial.
-Modern \gls{GPU} chips contain several hundred processing cores, but the amount of memory available locally for each core is very limited.\todo{repetition} There is also a considerable latency involved when accessing on-board memory, but these memory transfers are asynchronous and may be hidden.
-
-decompose the algorithm into as many threads as possible, each being computationally intense and be light on memory access and consumption.
-Consecutive threads should also access consecutive memory regions, to enable coalesced data transfers, and have the same line of execution.
-
-How may this be decomposed? Already decomposed at pixel-level.
-
-We implement the 
-Our implementation of th
-
-When
-
-\begin{gather}
-\vec w[n] = \frac{\Ri[n]\1}{\1\T\Ri[n]\1} = \frac{\vec\beta[n]}{\1\T\vec\beta[n]},
-\end{gather}
-
-all threads should have the same single line of execution
-
 \section{Results}
 
-Benchmarks were carried out on a test system comprised of a quad-core Intel Xeon E5620, 64Gb of \gls{RAM}, and an nVidia Quadro 6000 card.
+To test the \gls{CUDA} implementation of the \gls{MVDR} beamformer, we have chosen a dataset from the $M=32$ element Kongsberg Maritime HISAS1030 sonar~\cite{Hansen2009}, \todo{Data???} for
+
+The test system was a quad-core Intel Xeon E5620, 64Gb of \gls{RAM}, and an nVidia Quadro 6000 card.
 This is a \gls{GPU} with 448 \gls{CUDA} cores and 6Gb of onboard \gls{RAM}, capable of roughly 1Tflops.
-
-To test the \gls{CUDA} implementation of the \gls{MVDR} beamformer, we have chosen a dataset from the $M=32$ element Kongsberg Maritime HISAS1030 sonar~\cite{Hansen2009}. \todo{Data???}.
 

### 2012-05-04 ECUA2012/gpu_capon_ecua2012.tex
la2pdf commit - 2012:05:04	14:29:37	(Fri 4.May)

@@
 \item \emph{Computing the spatial covariance matrix} $\eR[n]$ (equation \ref{spatialR}-\ref{finalR}).
-Here we launch a thread per diagonal in the upper half triangular of $\eR$, and let finished threads wrap around and start on diagonals on the lower half triangular.
+Here we create threads that slide along the diagonals of $\eR$, and let finished threads wrap around and start on diagonals on the lower half triangular.
 This way we process one row of $\eR$, and manage to keep both data reads and writes coalesced.
@@
 
-To test the \gls{CUDA} implementation of the \gls{MVDR} beamformer, we have chosen a dataset from the $M=32$ element Kongsberg Maritime HISAS1030 sonar~\cite{Hansen2009}, \todo{Data???} for
+We have tested the \gls{CUDA} implementation of the \gls{MVDR} beamformer on experimental dat from $M=32$ element Kongsberg Maritime HISAS1030 sonar~\cite{Hansen2009}.
 
@@
 This is a \gls{GPU} with 448 \gls{CUDA} cores and 6Gb of onboard \gls{RAM}, capable of roughly 1Tflops.
+
+\begin{figure}[!t]
+\centering
+\includegraphics{gfx/benchmark.png}
+\caption{Capon benchmarks.}
+\label{fig_sim}
+\end{figure}
+
+% A sidescan image of the 1500 dwt oil tanker wreck Holmengraa is shown in \Fig{holmengraa}.
+It is about 68\,m long and 9\,m wide, and lies on a slanted seabed at 77\,m depth~\cite{holmengraa}.
+The sidescan image was created using data from the HISAS 1030 sonar, which is rather unsuited for this purpose because of its large opening angle.
+This, and the fact that the wreck was imaged at a range of about 105\,m makes the image quality poor, but sufficient to compare our beamformers.
+
+% In the Holmengraa image we note again that the LCA produces a cleaner shadow and better edge definition. The MV method performed almost identically to LCA in this case, and was omitted.
 

### 2012-05-04 ECUA2012/gpu_capon_ecua2012.tex
la2pdf commit - 2012:05:04	15:04:34	(Fri 4.May)

@@
 \newpage
+\begin{figure}[!t]
+\centering
+\includegraphics[width=\linewidth]{gfx/img_holmengraa.pdf}
+\caption{HISAS sidescan sonar (SSS) image of the shipwreck Holmengraa.}\label{holmengraa}
+\end{figure}
+
 \section{Implementation}
@@
 
-We have tested the \gls{CUDA} implementation of the \gls{MVDR} beamformer on experimental dat from $M=32$ element Kongsberg Maritime HISAS1030 sonar~\cite{Hansen2009}.
+We have tested \gls{GPU} implementation of the \gls{MVDR} beamformer on experimental data from $M=32$ element Kongsberg Maritime HISAS1030 sonar~\cite{Hansen2009}.
+The sonar was attached to the HUGIN \gls{AUV}, and the imaged object was the 1500 dwt oil tanker wreck Holmengraa (fig.
+\ref{holmengraa}).
+Holmengraa is roughly 68\,m long and 9\,m wide, and lies on a slanted seabed at 77\,m depth.
 
@@
 \centering
-\includegraphics{gfx/benchmark.png}
-\caption{Capon benchmarks.}
-\label{fig_sim}
+\includegraphics[width=\linewidth]{gfx/benchmark.pdf}
+\caption{Capon benchmarks.}\label{benchmarks}
 \end{figure}
@@
 % A sidescan image of the 1500 dwt oil tanker wreck Holmengraa is shown in \Fig{holmengraa}.
-It is about 68\,m long and 9\,m wide, and lies on a slanted seabed at 77\,m depth~\cite{holmengraa}.
+depth~\cite{holmengraa}.
 The sidescan image was created using data from the HISAS 1030 sonar, which is rather unsuited for this purpose because of its large opening angle.

### 2012-05-04 SiPS2012/sips2012.tex
la2pdf commit - 2012:04:26	15:36:11	(Thu26.Apr)
Update in sips notes

@@
 % ------
-\title{Implementing Adaptive Beamforming on the GPU for Real Time Ultrasound Imaging}
+\title{Implementing Capon Adaptive Beamforming on the GPU for Real Time Ultrasound Imaging}
 %
@@
 
-\section{Method}\label{sec:meth}
-\subsection{Building the Sample Covariance Matrix}
+\subsubsection{GPU Compute Model}
+In this section we briefly discuss the principle behind GPU computing.
+CUDA by Nvidia has been our choice of GPU developer environment.
+However, even though CUDA has been used to implement the Capon beamformer, the challenges and proposed methods described in this paper will apply for other massive parallel implementation environments like e.g.
+OpenCL.
+The CUDA compute model is based on execution of a kernel across a grid of compute threads.
+Each position in the grid holds a block of threads.
+Each block is further divided into groups of 32 threads, known as a warp.
+The maximum number of threads inside a block is restricted to 512, hence 16 warps.
+From a CUDA perspective the GPU consist of one or more stream multi processors (SM), where each SM is capable of 32 or more, depending on the architecture, concurrent multiplications and/or additions.
+The CUDA 2.x compute capability (CC) has available 48 KB of near-core shared memory per SM that can be utilized by resident blocks.
+The maximum number of resident blocks per SM is 8, but in addition the maximum number of resident warps per SM is restricted to 48.
+Hence, with a block size of 512 only $48/(512/32) = 3$ blocks can occupy the SM at once.
+In addition, if one thread where to use 64B of shared memory the number of allowed resident blocks with a block size of 512 would be reduced to 1.
+Up to a curtain point it is beneficial to have as many warps as possible occupying the SM at once to hide latency from accessing memory.
+The SM has for CC 2.x 32K 32-bit long registers.
+When implementing algorithms on the GPU it is important to keep data as close to the core as possible.
+Utilizing available registers and shared memory to its maximum is therefore important.
+
+\subsubsection{Optimizing kernel launch}
+Set the size of grid adaptively based on $L$, $M$ and $Yavg$.
+
+It is better to slide in space than time. $K$ is typically much larger than $2*Yavg + 1$.
+
+\section{Parallel Capon}\label{sec:meth}
+\subsection{Building Multiple Sample Covariance Matrices}
 Building $\hat{\mat{R}}$ using a sliding window approach across $K$ and $N_{avg}$.
@@
 
-\subsubsection{CUDA Compute Model}
-The CUDA compute model is based on execution of a kernel across a grid of compute threads.
-Each position in the grid holds a block of threads.
-Each block is further divided into groups of 32 threads, known as a warp.
-The maximum number of threads inside a block is restricted to 512, hence 16 warps.
-From a CUDA perspective the GPU consist of one or more stream multi processors (SM), where each SM is capable of 32 or more, depending on the architecture, concurrent multiplications and/or additions.
-The CUDA 2.x compute capability (CC) has available 48 KB of near-core shared memory per SM that can be utilized by resident blocks.
-The maximum number of resident blocks per SM is 8, but in addition the maximum number of resident warps per SM is restricted to 48.
-Hence, with a block size of 512 only $48/(512/32) = 3$ blocks can occupy the SM at once.
-In addition, if one thread where to use 64B of shared memory the number of allowed resident blocks with a block size of 512 would be reduced to 1.
-Up to a curtain point it is beneficial to have as many warps as possible occupying the SM at once to hide latency from accessing memory and executing time consuming mathematical functions.
-The SM has for CC 2.x 32K 32-bit long registers.
-When implementing algorithms on the GPU it is important to keep data as close to the core as possible.
-Utilizing available registers and shared memory to its maximum is therefore important.
-
-\subsubsection{Optimizing kernel launch}
-
 \subsection{Solving Multiple Small Linear Systems}
@@
 CUDA-Capon for different choices of parameters.
+Comment on different choices of parameters change the running times.
+Temporal averaging for instance, lowers the efficiency since $2*Y_{avg}$ more operations has to be performed per $\mat{\hat{R}}$
 
 (A prerequisite for this is to get the GPU-processing up and running on simulated data before Easter!! Maybe also a Matlab plug-in.)
+
+\subsection{Compute Beamformer Output}
 

### 2012-05-04 ECUA2012/gpu_capon_ecua2012.tex

@@
 
-CUDA...
+\begin{itemize}
+\item Adaptive beamformer's potential lies in its ability to suppress interference power
+\item Why adaptive beamformers struggle in active sonar systems.
+Correlated noise, robustification kills the adaptive potential.
+Quite computationally intensive.
+Constraints must be applied in one way or another - parameters must be tuned.
+\item CUDA?
+\end{itemize}
 
-Massively parallel hardware such as \glspl{GPU} have a potential for accelerating many signal processing tasks, and the \gls{MVDR} beamformer is no exception.
-Each pixel in the image can be processed independently, and
-
-Modern \gls{GPU} chips contain several hundred processing cores, but the amount of memory available locally for each core is very limited.\todo{repetition} There is also a considerable latency involved when accessing on-board memory, but these memory transfers are asynchronous and may be hidden.
-
-\todo{Get some references in here...}
+REFERENCES!!! (all over..)
 % - Who's done stuff on GPU beamforming before?
@@
 
+TODO:
+\begin{itemize}
+\item Choice of robustification parameters $L,M,\Delta$ largely impacts what the ``optimal'' way of implementing Capon 
+\item Introduce CUDA, or ditch it?
+\item Doesn't sell well that we used nVidias solver, especially when that's the bottleneck in the design...
+\end{itemize}
+
 \section{Results}
 
-We have tested \gls{GPU} implementation of the \gls{MVDR} beamformer on experimental data from $M=32$ element Kongsberg Maritime HISAS1030 sonar~\cite{Hansen2009}.
+We have tested \gls{GPU} implementation of the \gls{MVDR} beamformer on experimental data from $M=32$ element Kongsberg Maritime HISAS1030 sonar~\cite{Hansen2009}\todo{specify SAS somewhere here..}.
 The sonar was attached to the HUGIN \gls{AUV}, and the imaged object was the 1500 dwt oil tanker wreck Holmengraa (fig.
@@
 Holmengraa is roughly 68\,m long and 9\,m wide, and lies on a slanted seabed at 77\,m depth.
-
-The test system was a quad-core Intel Xeon E5620, 64Gb of \gls{RAM}, and an nVidia Quadro 6000 card.
-This is a \gls{GPU} with 448 \gls{CUDA} cores and 6Gb of onboard \gls{RAM}, capable of roughly 1Tflops.
 
@@
 
-% A sidescan image of the 1500 dwt oil tanker wreck Holmengraa is shown in \Fig{holmengraa}.
-depth~\cite{holmengraa}.
-The sidescan image was created using data from the HISAS 1030 sonar, which is rather unsuited for this purpose because of its large opening angle.
-This, and the fact that the wreck was imaged at a range of about 105\,m makes the image quality poor, but sufficient to compare our beamformers.
+Benchmarks\\
+The test system was a quad-core Intel Xeon E5620, 64Gb of \gls{RAM}, and an nVidia Quadro 6000 card.
+This is a \gls{GPU} with 448 \gls{CUDA} cores and 6Gb of onboard \gls{RAM}, capable of roughly 1Tflops.
 
-% In the Holmengraa image we note again that the LCA produces a cleaner shadow and better edge definition. The MV method performed almost identically to LCA in this case, and was omitted.
-
-down to these cores and back to the \gls{CPU} is also when copying data to and from the \gls{GPU} are also rather high, so each core should be   performing calculations there, and copying the data back to the \gls{CPU}
-
- but in order to fully exploit this potential we must be aware of its limitations.
-
-When adapting an algorithm to a massively parallel hardware such as \glspl{GPU}, one must be familiar with its limitations in order to that the potential that lies within these chips can not be fully realised
-
-When trying to fit any algorithm to massively parallel hardware such as \glspl{GPU}, one must be careful to select a proper decomposition scheme for the device at hand.
-Modern \glspl{GPU} have several hundred computing cores, but only a limited amount of  are  various decomposition schemes should be contemplated before performing the implecareful decomposition of the
-
-Let us first consider the \emph{computation of $\hat{\mat R}$}.
-Notice that when implementing the formula directly a lot of summations and multiplications will be repeated unnecessary.
-To illustrate this, consider for instance a case with $M=6$ sensors that are grouped into subarrays of length $L=3$ where all but one element overlap.
-Let us further assume that we need to average over 7 time samples, i.e.
-that $N_{\text{avg}} = 3$.
-
-These procedures are commutative, so the order in which we perform them does not matter.
-But for the sake of this example let us start by performing the time averaging.
-Assuming that $\vec x_i$ and $\vec x_j$ have a zero mean, a time averaged covariance coefficient $r_{ij}$ in $\hat{\mat R}$ may be obtained by (omitting the normalisation):
-\begin{align*}
-r_{ij}[n] &= \sumb{n'=n-N_{\text{avg}}}{n+N_{\text{avg}}} x_i[n']\,x_j[n'],
-\end{align*}
-which can be easily rewritten in a sliding window fashion
-\begin{align*}
-\hat r_{ij}[n] &= r_{ij}[n-1] - x_i[n\text{-}N_\text{avg}\text{-}1]\,x_j[n\text{-}N_\text{avg}\text{-}1]
-                         + x_i[n\text{+}N_\text{avg}]\,x_j[n\text{+}N_\text{avg}].
-\end{align*}
-Hence, whenever $N_\text{avg}>1$ the sliding window approach will demand less instructions at the expense of a slightly higher memory consumption.
-The next thing to realise is that not all $r_{ij}[n]$ need to be computed.
-Since $r_{ji} = r^*_{ij}$, and the subarrays only span $L$ elements, we only compute an upper triangular band of R (marked \Green{green})
-\begin{figure}[H]
-% \begin{narrow}{-3.5cm}{-3.5cm}
-\includegraphics[width=\linewidth]{gfx/capon_build_R_full.eps}
-\caption{Building $\hat{\mat R}$: Symmetry and element coverage}\label{buildingRfull}
-% \end{narrow}
-\end{figure}
-The final step is to average all the subarray matrices to form $\hat{\mat R}$ by 
-{\renewcommand*{\arraystretch}{1.8}
-\begin{align*}
-\hat{\mat R}_\text{sub} &= {\color{Blue}\mat R_{\text{sub},1} + \mat R_{\text{sub},2} + \mat R_{\text{sub},3} + \mat R_{\text{sub},4}} \\
-&= \left[\;
-\begin{matrix}
-\color{Green}r_{00}+r_{11}+r_{22} & \color{Red}r_{01}+r_{12}+r_{23} & r_{02}+r_{13}+r_{24} \\
-                                       & \color{Green}r_{11}+r_{22}+r_{33} & \color{Red}r_{12}+r_{23}+r_{34} \\
-                                       &                                        & \color{Green}r_{22}+r_{33}+r_{44}
-\end{matrix}
-\;\right]
-\end{align*}}%
-Note that consecutive elements along the diagonals overlap on all but one covariance coefficient, hence a sliding window may also be used here.
-The subarray size $L$ is usually quite a bit larger than the time averaging window, so there is more to gain from implementing the sliding window here.\\\\
-
-{\bf Method complexity:}\\
-\begin{tabular}{l l}
-No optimizations            & $2M^2\,(2\Navg+1)$ \\
-Banded symmetric            & $L(2M-L+1)\,(2\Navg+1)$ \\
-Sliding, banded, symmetric  & $1.5L(2M-L+1)$ \\
-\end{tabular}\\\\
-
-% \subsubsection{Computing $\text{\raisebox{1.9pt}{$\vec\chi$}} = \hat{\mat R}\,\!^{-1}\vec a$}
-
-There are two ways of finding $\text{\raisebox{1.9pt}{$\vec\chi$}} = \hat{\mat R}\,\!^{-1}\vec a$, either by solving the equation directly or by inverting $\hat{\mat R}$ first.
-Solving the equation is less computationally intensive than inverting $\hat{\mat R}$, and seems like the obvious choice.
-However, with iterative methods for updating $\eRi$ for each range pixel the situation is changed...
-
-\section{Results/Discussion}
+\section{Discussion}
 
 \begin{itemize}
-\item Execution time plots. For given of M/L, plot for different Yavg (note negligible diagonal loading).
-\item An image? Holmengrå?
+\item Relate Capon image to DAS image
+\item Discuss the benchmarks
+\item Need for speed: HUGIN 4 banks of 32 elements, can be processed faster than the ping repetition rate, with margins to spare.
 \end{itemize}
@@
 
-\todopar{Case:
-\begin{itemize}
-\item JOE
-\item Sonar. Channel count up to 32/64 ish channels?
-\item Other articles evaluates Capon image quality, and some have shown that speedups can be achieved by reducing the complexity of the traditional Capon algorithm or by implementing specific versions on a GPU.
-\item This article compare a few promising ways to implement Capon in general, in light of general complexity and suitability for implementation on a CPU and GPU.
-\end{itemize}
-}
+% A sidescan image of the 1500 dwt oil tanker wreck Holmengraa is shown in \Fig{holmengraa}.
+depth~\cite{holmengraa}.
+The sidescan image was created using data from the HISAS 1030 sonar, which is rather unsuited for this purpose because of its large opening angle.
+This, and the fact that the wreck was imaged at a range of about 105\,m makes the image quality poor, but sufficient to compare our beamformers.
 
-\begin{itemize}
-\item As briefly as possible, introduce beamforming.
-\item Adaptive beamformer's potential lies in its ability to suppress interference power
-\item Why adaptive beamformers struggle in active sonar systems.
+% % In the Holmengraa image we note again that the LCA produces a cleaner shadow and better edge definition. The MV method performed almost identically to LCA in this case, and was omitted.
+% 
+% 
+% 
+% down to these cores and back to the \gls{CPU} is also when copying data to and from the \gls{GPU} are also rather high, so each core should be   performing calculations there, and copying the data back to the \gls{CPU}
+% 
+%  but in order to fully exploit this potential we must be aware of its limitations.  
+% 
+% When adapting an algorithm to a massively parallel hardware such as \glspl{GPU}, one must be familiar with its limitations in order to that the potential that lies within these chips can not be fully realised
+% 
+% When trying to fit any algorithm to massively parallel hardware such as \glspl{GPU}, one must be careful to select a proper decomposition scheme for the device at hand.
+Modern \glspl{GPU} have several hundred computing cores, but only a limited amount of  are  various decomposition schemes should be contemplated before performing the implecareful decomposition of the
+% 
+% 
+% Let us first consider the \emph{computation of $\hat{\mat R}$}.
+Notice that when implementing the formula directly a lot of summations and multiplications will be repeated unnecessary.
+To illustrate this, consider for instance a case with $M=6$ sensors that are grouped into subarrays of length $L=3$ where all but one element overlap.
+Let us further assume that we need to average over 7 time samples, i.e.
+that $N_{\text{avg}} = 3$.
+% 
+% These procedures are commutative, so the order in which we perform them does not matter.
+But for the sake of this example let us start by performing the time averaging.
+Assuming that $\vec x_i$ and $\vec x_j$ have a zero mean, a time averaged covariance coefficient $r_{ij}$ in $\hat{\mat R}$ may be obtained by (omitting the normalisation):
+% \begin{align*}
+% r_{ij}[n] &= \sumb{n'=n-N_{\text{avg}}}{n+N_{\text{avg}}} x_i[n']\,x_j[n'],
+% \end{align*}
+% which can be easily rewritten in a sliding window fashion
+% \begin{align*}
+% \hat r_{ij}[n] &= r_{ij}[n-1] - x_i[n\text{-}N_\text{avg}\text{-}1]\,x_j[n\text{-}N_\text{avg}\text{-}1]
+%                          + x_i[n\text{+}N_\text{avg}]\,x_j[n\text{+}N_\text{avg}].
+% \end{align*}
+% Hence, whenever $N_\text{avg}>1$ the sliding window approach will demand less instructions at the expense of a slightly higher memory consumption.
+The next thing to realise is that not all $r_{ij}[n]$ need to be computed.
+Since $r_{ji} = r^*_{ij}$, and the subarrays only span $L$ elements, we only compute an upper triangular band of R (marked \Green{green})
+% \begin{figure}[H]
+% % \begin{narrow}{-3.5cm}{-3.5cm}
+% \includegraphics[width=\linewidth]{gfx/capon_build_R_full.eps}
+% \caption{Building $\hat{\mat R}$: Symmetry and element coverage}\label{buildingRfull}
+% % \end{narrow}
+% \end{figure}
+% The final step is to average all the subarray matrices to form $\hat{\mat R}$ by 
+% {\renewcommand*{\arraystretch}{1.8}
+% \begin{align*}
+% \hat{\mat R}_\text{sub} &= {\color{Blue}\mat R_{\text{sub},1} + \mat R_{\text{sub},2} + \mat R_{\text{sub},3} + \mat R_{\text{sub},4}} \\
+% &= \left[\;
+% \begin{matrix}
+% \color{Green}r_{00}+r_{11}+r_{22} & \color{Red}r_{01}+r_{12}+r_{23} & r_{02}+r_{13}+r_{24} \\
+%                                        & \color{Green}r_{11}+r_{22}+r_{33} & \color{Red}r_{12}+r_{23}+r_{34} \\
+%                                        &                                        & \color{Green}r_{22}+r_{33}+r_{44}
+% \end{matrix}
+% \;\right]
+% \end{align*}}%
+% Note that consecutive elements along the diagonals overlap on all but one covariance coefficient, hence a sliding window may also be used here.
+The subarray size $L$ is usually quite a bit larger than the time averaging window, so there is more to gain from implementing the sliding window here.\\\\
+% 
+% {\bf Method complexity:}\\
+% \begin{tabular}{l l}
+% No optimizations            & $2M^2\,(2\Navg+1)$ \\
+% Banded symmetric            & $L(2M-L+1)\,(2\Navg+1)$ \\
+% Sliding, banded, symmetric  & $1.5L(2M-L+1)$ \\
+% \end{tabular}\\\\
+
+% \subsubsection{Computing $\text{\raisebox{1.9pt}{$\vec\chi$}} = \hat{\mat R}\,\!^{-1}\vec a$}
+% 
+% There are two ways of finding $\text{\raisebox{1.9pt}{$\vec\chi$}} = \hat{\mat R}\,\!^{-1}\vec a$, either by solving the equation directly or by inverting $\hat{\mat R}$ first.
+Solving the equation is less computationally intensive than inverting $\hat{\mat R}$, and seems like the obvious choice.
+However, with iterative methods for updating $\eRi$ for each range pixel the situation is changed...
+% 
+% 
+% \section{Results/Discussion}
+% 
+% 
+% 
+% 
+% 
+% \todopar{Case:
+% \begin{itemize}
+% \item JOE
+% \item Sonar. Channel count up to 32/64 ish channels?
+% \item Other articles evaluates Capon image quality, and some have shown that speedups can be achieved by reducing the complexity of the traditional Capon algorithm or by implementing specific versions on a GPU.
+% \item This article compare a few promising ways to implement Capon in general, in light of general complexity and suitability for implementation on a CPU and GPU.
+% \end{itemize}
+% }
+% 
+% \begin{itemize}
+% \item As briefly as possible, introduce beamforming.
+% \item Adaptive beamformer's potential lies in its ability to suppress interference power
+% \item Why adaptive beamformers struggle in active sonar systems.
 Correlated noise, robustification kills the adaptive potential.
@@
 Constraints must be applied in one way or another - parameters must be tuned.
-\item The three main ways to implement Capon \cite{Capon1969} is \todo{not entirely true?}
-\begin{itemize}
-\item Traditional way: Build $\eR$ and solve $\w = \frac{\Ri\a}{\a\H\Ri\a}$. 
-\item Iterative methods: Woodbury, Conjugated gradients
-\item Beamspace
-\item LCA (or other methods, as reference)
-\end{itemize}
-\item Choice of robustification parameters $L,M,\Delta$ largely impacts what the ``optimal'' way of implementing Capon 
-\item In most imaging applications 
-Outline:
-\begin{itemize}
-\item Speeding up: Build $\eR$ and solve $\w = \frac{\Ri\a}{\a\H\Ri\a}$. 
-\item Iterative methods: Woodbury (Sherman Morrison), Conjugated gradients - show performance of GPU performance.
-\item Woodbury: Must take the average over many samples in space and time - lots of (small) outerproducts. Can be used if not a lot of averaging is required.
-\item Beamspace
-\item LCA (or other methods, as reference)
-\end{itemize}
-\end{itemize}
+% \item The three main ways to implement Capon \cite{Capon1969} is \todo{not entirely true?}
+% \begin{itemize}
+% \item Traditional way: Build $\eR$ and solve $\w = \frac{\Ri\a}{\a\H\Ri\a}$. 
+% \item Iterative methods: Woodbury, Conjugated gradients
+% \item Beamspace
+% \item LCA (or other methods, as reference)
+% \end{itemize}
+% \item Choice of robustification parameters $L,M,\Delta$ largely impacts what the ``optimal'' way of implementing Capon 
+% \item In most imaging applications 
+% Outline:
+% \begin{itemize}
+% \item Speeding up: Build $\eR$ and solve $\w = \frac{\Ri\a}{\a\H\Ri\a}$. 
+% \item Iterative methods: Woodbury (Sherman Morrison), Conjugated gradients - show performance of GPU performance.
+% \item Woodbury: Must take the average over many samples in space and time - lots of (small) outerproducts. Can be used if not a lot of averaging is required.
+% \item Beamspace
+% \item LCA (or other methods, as reference)
+% \end{itemize}
+% \end{itemize}
+%
 
@@
 % stfloats package.
-
-\section{Conclusion}
-The conclusion goes here.
-
-% if have a single appendix:
-%\appendix[Proof of the Zonklar Equations]
-% or
-%\appendix  % for no appendix heading
-% do not use \section anymore after \appendix, only \section*
-% is possibly needed
-
-% use appendices with more than one appendix
-% then use \section to start each appendix
-% you must declare a \section before using any
-% \subsection or using \label (\appendices by itself
-% starts a section numbered zero.)
 % 
+% 
+% 
+% \section{Conclusion}
+% The conclusion goes here.
+% 
+% 
+% 
+% 
+% 
+% % if have a single appendix:
+% %\appendix[Proof of the Zonklar Equations]
+% % or
+% %\appendix  % for no appendix heading
+% % do not use \section anymore after \appendix, only \section*
+% % is possibly needed
+% 
+% % use appendices with more than one appendix
+% % then use \section to start each appendix
+% % you must declare a \section before using any
+% % \subsection or using \label (\appendices by itself
+% % starts a section numbered zero.)
+% %
 
@@
 
-\section*{Acknowledgments}
+% 
+\section{Acknowledgments}
 

### 2012-05-04 ECUA2012/gpu_capon_ecua2012.tex
Scripts and images for ECUA was missing. Added.

@@
 \end{gather}
-where $\1$ is a row vector of ones, and $\R$ is the spatial covariance matrix.
+where $\1$ is a row vector of ones that represents broadside steering, and $\R$ is the spatial covariance matrix.
 Since $\R$ is unknown, it must be estimated.
@@
 This suggests that the algorithm should be decomposed into as many lightweight threads as possible, and using a single thread per pixel is not lightweight enough.
-Therefore, we seek to devise decomposition scheme for the methods just presented.
+Therefore, we seek to devise a way to decompose the methods just presented.
 In particular, the solutions we came up with for the different steps in the algorithm, listed in their natural order of execution, was:

### 2012-05-04 ECUA2012/gpu_capon_ecua2012.tex
la2pdf commit - 2012:05:04	16:08:07	(Fri 4.May)

@@
 
-We have tested \gls{GPU} implementation of the \gls{MVDR} beamformer on experimental data from $M=32$ element Kongsberg Maritime HISAS1030 sonar~\cite{Hansen2009}\todo{specify SAS somewhere here..}.
+We have tested \gls{GPU} implementation of the \gls{MVDR} beamformer on experimental data from the $M=32$ element Kongsberg Maritime HISAS1030 sonar~\cite{Hansen2009}\todo{specify SAS somewhere here..}.
 The sonar was attached to the HUGIN \gls{AUV}, and the imaged object was the 1500 dwt oil tanker wreck Holmengraa (fig.
@@
 Holmengraa is roughly 68\,m long and 9\,m wide, and lies on a slanted seabed at 77\,m depth.
+The Capon image
 

### 2012-05-04 ECUA2012/gpu_capon_ecua2012.tex
Done some writing on the implementation part. Good luck!

--- 2012-05-04 ECUA2012/gpu_capon_ecua2012.tex
@@
 \makeglossaries
-\input{../../Library/glossary}
+%\input{../../Library/glossary}
+% windows hack for JP
+\renewcommand\gls[1]{#1}
 
@@
 \newpage
-\section{Methods}
+\section{Methods}\label{methods}
 
@@
 \section{Implementation}
-
-A modern \gls{GPU} is an hardware architecture comprised of several hundred computing cores, each having access to a very limited amount of local memory.
-This suggests that the algorithm should be decomposed into as many lightweight threads as possible, and using a single thread per pixel is not lightweight enough.
-Therefore, we seek to devise a way to decompose the methods just presented.
+The paradigm of \gls{GPU} computing is formalized as the execution of thousand of threads, where each thread runs a copy of a common function, also know as a kernel.
+Threads are organized in blocks, and blocks further into a grid of blocks.
+The threads index in this grid usually decides the data to be process and the instruction path to go along.
+A modern \gls{GPU} is an hardware architecture comprised of several hundred computing cores, each having access to a very limited amount of local memory an registers.
+This suggests that the algorithm should be decomposed into as many lightweight threads as possible.
+For a typical array configuration, using a single thread per pixel for Capon is not lightweight enough.
+Therefore, we seek to devise a way to decompose the methods presented in section \ref{methods}.
 In particular, the solutions we came up with for the different steps in the algorithm, listed in their natural order of execution, was:
@@
 \item \emph{Computing the spatial covariance matrix} $\eR[n]$ (equation \ref{spatialR}-\ref{finalR}).
-Here we create threads that slide along the diagonals of $\eR$, and let finished threads wrap around and start on diagonals on the lower half triangular.
-This way we process one row of $\eR$, and manage to keep both data reads and writes coalesced.
+A group of threads are created that slide along the diagonals of $\eR$.
+In this way we can exploit the fact that entries on the diagonals overlaps across subarrays and time.
+%Finished threads can also wrap around and start on diagonals on the lower half triangular. 
+In this way we process one row of $\eR$ per kernel iteration, and manage to keep the numbers of both data reads and writes at a minimum.
 \item \emph{Computing} $\eRi[n]\1$ (from equation \ref{weights}).
 While intuition may suggest that this step is carried out by inverting $\eR$, it is better to solve the linear equation $\R\beta = \1$ for $\beta$ instead, which gives us $\beta = \eRi\1$ directly.
-We have tested various solvers for this task, but achieved the best performance by using an unofficial batch solver from nVidia.
-This is by far the most computationally intensive task, and a key area of focus for further improvements.
-\item \emph{Computing the subarray weights} by substituting $\eRi[n]\1$ into equation \ref{weights}.
-\item \emph{Acquire the pixel value $z[n]$} by applying the weights to each data vector $\x_l[n]$, and sum them all up (as in equation \ref{z}).
+We have tested various solvers for this task, both in-house and proprietary implementations, but achieved the best performance by using an unofficial batch solver from nVidia.
+Inverting $\eR$ is by far the most computationally intensive task, and a key area of focus for further improvements.
+\item \emph{Computing the subarray weights} by substituting $\beta = \eRi[n]\1$ into equation \ref{weights}, $\vec w[n] = \frac{\beta}{\1\T\beta}$.
+A group of L threads per pixel was used to reduce shared memory pressure and to obtain coalesced reads and a minimum of writes.
+\emph{Finally we acquire the pixel value $z[n]$} by applying the weights to each data vector $\x_l[n]$, and sum them all up (as in equation \ref{z}).
+When applying weights, there are two approaches from an implementation point of view.
+The subarray data can, as here, be reduced to coincide in length with the subarray weights, or the weights can be extended to M in size and applied directly on $\x[n]$ and summed.
 \end{enumerate}
+For the GPU, the largest bottleneck is the transfer of memory from CPU-side (Host) to the GPU-side (Device).
+In the upcoming results, Host to Device transfer are not taken into account.
+The reasoning behind this is the fact that Capon beamforming requires already delayed data.
+For efficiency, this should also be carried out on the GPU, leaving the challenging task of overcoming the limited Host-Device memory bandwidth to those who implement the delay part of delay-and-sum.
 
@@
 \end{itemize}
+
+\newpage
 

### 2012-05-04 ECUA2012/gpu_capon_ecua2012.tex
.....

--- 2012-05-04 ECUA2012/gpu_capon_ecua2012.tex
@@
 
-We have tested \gls{GPU} implementation of the \gls{MVDR} beamformer on experimental data from $M=32$ element Kongsberg Maritime HISAS1030 sonar~\cite{Hansen2009}\todo{specify SAS somewhere here..}.
+We have tested \gls{GPU} implementation of the \gls{MVDR} beamformer on experimental data from 32 element Kongsberg Maritime HISAS1030 sonar~\cite{Hansen2009}\todo{specify SAS somewhere here..}.
 The sonar was attached to the HUGIN \gls{AUV}, and the imaged object was the 1500 dwt oil tanker wreck Holmengraa (fig.
@@
 Holmengraa is roughly 68\,m long and 9\,m wide, and lies on a slanted seabed at 77\,m depth.
+The Capon image was created with the parameters $L=16$, $N_\text{avg}=3$, and $d=1\%$.
 

### 2012-05-06 ECUA2012/gpu_capon_ecua2012.tex
la2pdf commit - 2012:05:06	21:04:09	(Sun 6.May)
la2pdf commit - 2012:05:06	22:22:52	(Sun 6.May)

@@
 
+The feasibility of using the Capon beamformer in active sonar imaging has been demonstrated in e.g. \cite{}\todo{...}.
+E.g.
+
+GPU ultrasound (\cite{So2011}
+
+Complex data, all robustification techniques.
+
 \begin{itemize}
@@
 
+The principle of active sonar imaging is to transmit pulses that reflect off some interesting surface, and try to estimate some of its properties by inspecting the received echoes.
+When a receiver array is used, data from each sensor is coherently combined to   When an array of sensors  sensors, which  and then use them to estimateThe signal returns are often recorded using phased arrays, which enable
+A common active sonar system transmits and receives encoded signals using phased arrays.
+Such arrays allow the  to transmit encoded signals that using a phased array usually operates by transmitting encoded signals towards some surface of interest, and record the returns using a .
+The signal returns are then recorded by
+An active sonar system transmit signals 
 Consider an $M$ element phased array.
@@
 Since $\R$ is unknown, it must be estimated.
-We will do this by performing some degree of spatial averaging to avoid signal cancellation, and temporal averaging to achieve \gls{DAS}-like speckle statistics.
-We start with the former, by segmenting the array into $M-L-1$ subarrays of length $L$, where all but one element overlap.
-The data vector $x_l[n]$ from subarray $l$ is given as
+We will do this by computing the sample covariance matrix, $\eR$.
+which will require some degree of spatial averaging to avoid signal cancellation, and temporal averaging to achieve \gls{DAS}-like speckle statistics.
+Spatial averaging\todo{not quite there yet} is achieved by segmenting the array into $M-L-1$ subarrays of length $L$, where all but one element overlap.
+Let $x_l[n]$ represent the datavector from subarray $l$,
 \begin{gather}
-\x_l[n] = \bmat{x_l[n-\Delta_l] & x_{l+1}[n-\Delta_l] & \dots & x_{l+L-1}[n-\Delta_{l+L-1}]}\T,
+\x_l[n] = \bmat{x_l[n-\Delta_l] & x_{l+1}[n-\Delta_l] & \dots & x_{l+L-1}[n-\Delta_{l+L-1}]}\T.
 \end{gather}
-and the spatially averaged sample covariance matrix, $\tilde R$, is given as
+The spatially averaged sample covariance matrix, $\tilde R$, is given as
 \begin{gather}

### 2012-05-06 ECUA2012/gpu_capon_ecua2012.tex
la2pdf commit - 2012:05:06	23:08:43	(Sun 6.May)

@@
 
-The feasibility of using the Capon beamformer in active sonar imaging has been demonstrated in e.g. \cite{}\todo{...}.
+The feasibility of using the Capon beamformer in active sonar imaging has been demonstrated in e.g. \cite{Blomberg2012a,Blomberg2011,Dursun2009}.
 E.g.
 
-GPU ultrasound (\cite{So2011}
+GPU ultrasound (\cite{So2011,Chen2011})
 
@@
 
-The principle of active sonar imaging is to transmit pulses that reflect off some interesting surface, and try to estimate some of its properties by inspecting the received echoes.
-When a receiver array is used, data from each sensor is coherently combined to   When an array of sensors  sensors, which  and then use them to estimateThe signal returns are often recorded using phased arrays, which enable
-A common active sonar system transmits and receives encoded signals using phased arrays.
-Such arrays allow the  to transmit encoded signals that using a phased array usually operates by transmitting encoded signals towards some surface of interest, and record the returns using a .
-The signal returns are then recorded by
-An active sonar system transmit signals 
-Consider an $M$ element phased array.
+An active sonar imaging system usually transmits encoded signals signals towards a surface interest, and record the resulting reflections using receiver arrays.
+The receiver outputs may then be coherently combined in the post-processing to focus on a point in the image.
+This process is called beamforming, and the combination process involves assigning suitable delays and weights to the array's channels.
+
+Now consider an $M$ element phased array.
 Let the sample recorded by the $m$'th channel at time instant $n$ be represented as $x_m[n]$, and the accompanying weights and delays be  $w_m[n]$ and $\Delta_m[n]$.

### 2012-05-06 ECUA2012/gpu_capon_ecua2012.tex
la2pdf commit - 2012:05:06	23:41:43	(Sun 6.May)

@@
 REFERENCES!!! (all over..)
+\todo{Capon or MVDR?}
 % - Who's done stuff on GPU beamforming before?
@@
 
-An active sonar imaging system usually transmits encoded signals signals towards a surface interest, and record the resulting reflections using receiver arrays.
-The receiver outputs may then be coherently combined in the post-processing to focus on a point in the image.
-This process is called beamforming, and the combination process involves assigning suitable delays and weights to the array's channels.
+An active sonar imaging system usually transmits encoded signals signals towards a surface interest, and record the resulting reflections using an array of receivers.
+This is allows the receiver outputs to be coherently combined in the post-processing to focus on specific points in the image.
+The combination process involves assigning suitable delays and weights to the array's channels, and is usually referred to as beamforming.
 
@@
 which will require some degree of spatial averaging to avoid signal cancellation, and temporal averaging to achieve \gls{DAS}-like speckle statistics.
-Spatial averaging\todo{not quite there yet} is achieved by segmenting the array into $M-L-1$ subarrays of length $L$, where all but one element overlap.
-Let $x_l[n]$ represent the datavector from subarray $l$,
+
+\emph{Spatial averaging} is achieved by segmenting the array into $M-L-1$ subarrays of length $L$, where all but one element overlap. If we let $x_l[n]$ represent the datavector from subarray $l$,
 \begin{gather}
-\x_l[n] = \bmat{x_l[n-\Delta_l] & x_{l+1}[n-\Delta_l] & \dots & x_{l+L-1}[n-\Delta_{l+L-1}]}\T.
+\x_l[n] = \bmat{x_l[n-\Delta_l] & x_{l+1}[n-\Delta_l] & \dots & x_{l+L-1}[n-\Delta_{l+L-1}]}\T,
 \end{gather}
-The spatially averaged sample covariance matrix, $\tilde R$, is given as
+then the a ``subarray averaged'' sample covariance matrix, $\eR_\text{s}$, is given as
 \begin{gather}
-\tilde\R[n] = \frac{1}{M-L+1} \sumb{l=0}{M-L+1} \x_l[n]\x_l\H[n].\label{spatialR}
+\eR_\text{s}[n] = \frac{1}{M-L+1} \sumb{l=0}{M-L+1} \x_l[n]\x_l\H[n].\label{spatialR}
 \end{gather}
-Finally, we take the average of $2N_\text{avg}+1$ temporal samples to form the final sample covariance matrix $\eR$,
+
+\emph{Temporal averaging} can now be introduced by taking the average of $2N_\text{avg}+1$ temporal samples of $\eR_\text{s}[n]$. Let the result be called $\eR_\text{t,s}[n]$,
 \begin{gather}
-\eR[n] = \frac{1}{2N_\text{avg}+1} \sumb{n'=n-N_\text{avg}}{n+N_\text{avg}} \tilde\R[n'].
+\eR_\text{t,s}[n] = \frac{1}{2N_\text{avg}+1} \sumb{n'=n-N_\text{avg}}{n+N_\text{avg}} \eR_\text{s}[n'].
 \end{gather}
-Finally, to make sure that $\eR$ is well conditioned, we add a fraction $d$ of the total power of $\eR$ to its diagonal\cite{Synnevag2007},
+Finally, to ensure a well conditioned sample covariance matrix, we add a fraction $d$ of the total power of $\eR_\text{t,s}[n]$ to its diagonal\cite{Synnevag2007},
 \begin{align}
-\eR[n] = \eR[n] + \I \cdot \frac{d}{L} \cdot \tr\{\eR[n]\}.\label{finalR}
+\eR[n] = \eR_\text{t,s}[n] + \I \cdot \frac{d}{L} \cdot \tr\{\eR_\text{t,s}[n]\}.\label{finalR}
 \end{align}
@@
 
-We have tested \gls{GPU} implementation of the \gls{MVDR} beamformer on experimental data from 32 element Kongsberg Maritime HISAS1030 sonar~\cite{Hansen2009}\todo{specify SAS somewhere here..}.
-The sonar was attached to the HUGIN \gls{AUV}, and the imaged object was the 1500 dwt oil tanker wreck Holmengraa (fig.
-\ref{holmengraa}).
+We have tested \gls{GPU} implementation of the \gls{MVDR} beamformer on experimental data from 32 element Kongsberg Maritime HISAS1030 synthetic aperture sonar~\cite{Hansen2009}.
+The sonar was attached to the HUGIN \gls{AUV} and used in strip-map mode, and the imaged object was the 1500 dwt oil tanker wreck Holmengraa (Figure \ref{holmengraa}).
 Holmengraa is roughly 68\,m long and 9\,m wide, and lies on a slanted seabed at 77\,m depth.
-The Capon image was created with the parameters $L=16$, $N_\text{avg}=3$, and $d=1\%$.
+The Capon image was created with the parameters $L=16$, $N_\text{avg}=3$, and $d=1\%$, which proved to be a reasonable selection for this scenario.
+
+Figure \ref{holmengraa} demonstrates the potential of the Capon beamformer.
 

### 2012-05-07 ECUA2012/gpu_capon_ecua2012.tex
la2pdf commit - 2012:05:07	09:46:53	(Mon 7.May)

@@
 
-Figure \ref{holmengraa} demonstrates the potential of the Capon beamformer.
+The Holmengraa demonstrates the Capon beamformer's ability to suppress interference and achieve better detail resolution.
+Compared the \gls{DAS} beamformer, the ship edges come out as sharper, and the shadows are less noisy.
+\todo{DAS faster, get it in somehow..}
 
@@
 
-Benchmarks\\
-The test system was a quad-core Intel Xeon E5620, 64Gb of \gls{RAM}, and an nVidia Quadro 6000 card.
-This is a \gls{GPU} with 448 \gls{CUDA} cores and 6Gb of onboard \gls{RAM}, capable of roughly 1Tflops.
+The Capon parameters influence the complexity of the algorithm. Runtime scale proportionally with the number of image pixels, $M$\todo{prolly not accurate} and $N_\text{avg}$, but not with $L$. To
+
+Figure \ref{benchmarks}. Here the number of channels was set to $M=32$, and
+
+test system was a quad-core Intel Xeon E5620, 64Gb of \gls{RAM}, and an nVidia Quadro 6000 card. This is a \gls{GPU} with 448 \gls{CUDA} cores and 6Gb of onboard \gls{RAM}, capable of roughly 1Tflops.
 

### 2012-05-07 ECUA2012/gpu_capon_ecua2012.tex
la2pdf commit - 2012:05:07	11:19:12	(Mon 7.May)

@@
 
-The \gls{MVDR} (or Capon) beamformer is one such technique.
+The Capon (or \gls{MVDR}) beamformer is one such technique.
 While ensuring unit gain in the look direction, it computes the set of weights that minimizes the energy accumulated by the array from other directions.
@@
 
-However, despite its inherent potential, the \gls{MVDR} beamformer has yet to see widespread adoption in the active sonar community.
+However, despite its inherent potential, the Capon beamformer has yet to see widespread adoption in the active sonar community.
 There may be several reasons for this.
@@
 
-Perhaps surprisingly, we will show that the suggested means for making the \gls{MVDR} behave well in active systems will also make it less computationally complex.
+Perhaps surprisingly, we will show that the suggested means for making the Capon behave well in active systems will also make it less computationally complex.
 Furthermore, by implementing it on a modern \gls{GPU} we have gained a speedup of 2-3 orders of magnitude compared to optimized single thread implementations in C and Matlab.
@@
 
-An active sonar imaging system usually transmits encoded signals signals towards a surface interest, and record the resulting reflections using an array of receivers.
-This is allows the receiver outputs to be coherently combined in the post-processing to focus on specific points in the image.
-The combination process involves assigning suitable delays and weights to the array's channels, and is usually referred to as beamforming.
+In a typical active sonar imaging scenario encoded signals are transmitted towards a surface interest, and the resulting reflections are recorded using an array of receivers.
+This is favourable setup as it allows the receiver outputs to be coherently combined in the post-processing to focus on specific points in the image.
+The combination process involves assigning suitable delays and weights to the array's channels, and is commonly referred to as beamforming.
 
@@
 \end{align}
-The \gls{MVDR} beamformer \cite{Capon1969} finds the set of weights that minimizes the output power, $\min E\{|z[n]|^2\}$, while ensuring unity gain in the look direction.
+The Capon beamformer \cite{Capon1969} finds the set of weights that minimizes the output power, $\min E\{|z[n]|^2\}$, while ensuring unity gain in the look direction.
 The solution to this optimization problem is given as
@@
 Since $\R$ is unknown, it must be estimated.
-We will do this by computing the sample covariance matrix, $\eR$.
-which will require some degree of spatial averaging to avoid signal cancellation, and temporal averaging to achieve \gls{DAS}-like speckle statistics.
+We will do this by computing the sample covariance matrix, $\eR$, which will require some degree of spatial averaging to avoid signal cancellation, and temporal averaging to achieve \gls{DAS}-like speckle statistics.
 
@@
 
-We have tested \gls{GPU} implementation of the \gls{MVDR} beamformer on experimental data from 32 element Kongsberg Maritime HISAS1030 synthetic aperture sonar~\cite{Hansen2009}.
+We have tested \gls{GPU} implementation of the Capon beamformer on experimental data from 32 element Kongsberg Maritime HISAS1030 synthetic aperture sonar~\cite{Hansen2009}.
 The sonar was attached to the HUGIN \gls{AUV} and used in strip-map mode, and the imaged object was the 1500 dwt oil tanker wreck Holmengraa (Figure \ref{holmengraa}).
@@
 The Capon image was created with the parameters $L=16$, $N_\text{avg}=3$, and $d=1\%$, which proved to be a reasonable selection for this scenario.
-
-The Holmengraa demonstrates the Capon beamformer's ability to suppress interference and achieve better detail resolution.
-Compared the \gls{DAS} beamformer, the ship edges come out as sharper, and the shadows are less noisy.
-\todo{DAS faster, get it in somehow..}
 
@@
 
+To test the runtime speed of our \gls{GPU} implementation of Capon, we have compared it to an optimized single thread $C$ and $Matlab$ implementation (Figure \ref{benchmarks}).
+The test system was a quad-core Intel Xeon E5620, with 64Gb of \gls{RAM}, and an nVidia Quadro 6000 card.
+This is a \gls{GPU} with 448 \gls{CUDA} cores and 6Gb of onboard \gls{RAM}, capable of roughly 1Tflops.
+We kept the channel count at $M=32$, and processed a $640\times 86$ pixel image with $d=2\%$ diagonal loading.
+
+\section{Discussion}
+
+Figure \ref{holmengraa} demonstrates the Capon beamformer's ability to suppress interference and achieve better detail resolution.
+Compared the \gls{DAS} beamformer, the ship edges come out as sharper, and the shadows are less noisy.
+But while \gls{DAS} processed this 1.3Mpx image in milliseconds, our Matlab implementation of Capon needed minutes.
+
+Figure \ref{benchmarks} shows how a \gls{GPU} can be used to improve matters.
+The test image here was 55kpx, hence much smaller than the Holmengraa image, but the relative performance increase seen when accelerating the algorithm with a \gls{GPU} is still accurate.
+
 The Capon parameters influence the complexity of the algorithm. Runtime scale proportionally with the number of image pixels, $M$\todo{prolly not accurate} and $N_\text{avg}$, but not with $L$. To
@@
 Figure \ref{benchmarks}. Here the number of channels was set to $M=32$, and
-
-test system was a quad-core Intel Xeon E5620, 64Gb of \gls{RAM}, and an nVidia Quadro 6000 card. This is a \gls{GPU} with 448 \gls{CUDA} cores and 6Gb of onboard \gls{RAM}, capable of roughly 1Tflops.
-
-\section{Discussion}
 

### 2012-05-07 ECUA2012/gpu_capon_ecua2012.tex
la2pdf commit - 2012:05:07	11:52:36	(Mon 7.May)

@@
 
-We have tested \gls{GPU} implementation of the Capon beamformer on experimental data from 32 element Kongsberg Maritime HISAS1030 synthetic aperture sonar~\cite{Hansen2009}.
+We have tested our \gls{GPU} implementation of the Capon beamformer on experimental data from the 32 element Kongsberg Maritime HISAS1030 synthetic aperture sonar~\cite{Hansen2009}.
 The sonar was attached to the HUGIN \gls{AUV} and used in strip-map mode, and the imaged object was the 1500 dwt oil tanker wreck Holmengraa (Figure \ref{holmengraa}).
@@
 \includegraphics[width=\linewidth]{gfx/benchmark.pdf}
-\caption{Capon benchmarks.}\label{benchmarks}
+\caption{Capon benchmarks. An 55\,kpx image processed with $M=32$ channels and $d=2\%$ diagonal loading. }\label{benchmarks}
 \end{figure}
 
-To test the runtime speed of our \gls{GPU} implementation of Capon, we have compared it to an optimized single thread $C$ and $Matlab$ implementation (Figure \ref{benchmarks}).
+To test the runtime speed of our \gls{GPU} implementation of Capon, we have compared it to an optimized single thread C and Matlab implementation (Figure \ref{benchmarks}).
 The test system was a quad-core Intel Xeon E5620, with 64Gb of \gls{RAM}, and an nVidia Quadro 6000 card.
 This is a \gls{GPU} with 448 \gls{CUDA} cores and 6Gb of onboard \gls{RAM}, capable of roughly 1Tflops.
-We kept the channel count at $M=32$, and processed a $640\times 86$ pixel image with $d=2\%$ diagonal loading.
+We kept the channel count at $M=32$, and processed a 55\,kpixel image with $d=2\%$ diagonal loading.
 
@@
 Figure \ref{benchmarks} shows how a \gls{GPU} can be used to improve matters.
-The test image here was 55kpx, hence much smaller than the Holmengraa image, but the relative performance increase seen when accelerating the algorithm with a \gls{GPU} is still accurate.
+The test image used here was 55kpx,   and the \gls{} hence the runtimes are seen are   much smaller than the Holmengraa image, but the relative performance increase seen when accelerating the algorithm with a \gls{GPU} is still accurate.
 

### 2012-05-07 ECUA2012/gpu_capon_ecua2012.tex
la2pdf commit - 2012:05:07	12:23:20	(Mon 7.May)

@@
 \includegraphics[width=\linewidth]{gfx/benchmark.pdf}
-\caption{Capon benchmarks. An 55\,kpx image processed with $M=32$ channels and $d=2\%$ diagonal loading. }\label{benchmarks}
+\caption{Capon benchmarks. \protect\newline A 55\,kpx image processed with $M=32$ channels and $d=2\%$ diagonal loading.}\label{benchmarks}
 \end{figure}
@@
 Figure \ref{holmengraa} demonstrates the Capon beamformer's ability to suppress interference and achieve better detail resolution.
-Compared the \gls{DAS} beamformer, the ship edges come out as sharper, and the shadows are less noisy.
+Compared the \gls{DAS} beamformer, the ship's edges come out as sharper, and the shadows are less noisy.
 But while \gls{DAS} processed this 1.3Mpx image in milliseconds, our Matlab implementation of Capon needed minutes.
 
-Figure \ref{benchmarks} shows how a \gls{GPU} can be used to improve matters.
-The test image used here was 55kpx,   and the \gls{} hence the runtimes are seen are   much smaller than the Holmengraa image, but the relative performance increase seen when accelerating the algorithm with a \gls{GPU} is still accurate.
+As Figure \ref{benchmarks} illustrates, using a \gls{GPU} improve matters.
+and the \gls{} hence the runtimes are seen are   much smaller than the Holmengraa image, but the relative performance increase seen when accelerating the algorithm with a \gls{GPU} is still accurate.
 

### 2012-05-07 ECUA2012/gpu_capon_ecua2012.tex
la2pdf commit - 2012:05:07	12:53:44	(Mon 7.May)

@@
 This is a \gls{GPU} with 448 \gls{CUDA} cores and 6Gb of onboard \gls{RAM}, capable of roughly 1Tflops.
-We kept the channel count at $M=32$, and processed a 55\,kpixel image with $d=2\%$ diagonal loading.
+We kept the channel count at $M=32$ and added $d=2\%$ diagonal loading.
+The test image was a 55\,kpixels sectorscan.
 
@@
 Compared the \gls{DAS} beamformer, the ship's edges come out as sharper, and the shadows are less noisy.
-But while \gls{DAS} processed this 1.3Mpx image in milliseconds, our Matlab implementation of Capon needed minutes.
+But while \gls{DAS} processed this 1.3\,Mpixel image in milliseconds, our Matlab implementation of Capon needed minutes.
 
-As Figure \ref{benchmarks} illustrates, using a \gls{GPU} improve matters.
-and the \gls{} hence the runtimes are seen are   much smaller than the Holmengraa image, but the relative performance increase seen when accelerating the algorithm with a \gls{GPU} is still accurate.
+As Figure \ref{benchmarks} illustrates, using a \gls{GPU} improves matters.
+If we look at the total runtimes for a subarray size of $L=16$ and $2N_\text{avg}+1=5$ pixels of temporal averaging, the \gls{GPU} implementation is able to process the test image in roughly 40\,ms.
+For the same scenario the C implementation clocked in at roughly 3\,s and Matlab at 17\,s, which corresponds to a speedup factor of 75 and 435, respectively.
 
-The Capon parameters influence the complexity of the algorithm. Runtime scale proportionally with the number of image pixels, $M$\todo{prolly not accurate} and $N_\text{avg}$, but not with $L$. To
-
-Figure \ref{benchmarks}. Here the number of channels was set to $M=32$, and
+Interesting to note is also that solving $\eRi\1$ remain the main bottleneck in the \gls{GPU} implementation, and for larger $L$'s it dominates.
+However, this is a general Gauss Jordan solver which does not exploit symmetry properties of $\eR$, so creating e.g.
+a batch based Cholesky or UHDU solver should cut the runtimes quite a bit further.
+Both the C and Matlab implementation in this benchmark take this symmetry into account, which explains why the complexity curve of these implementation is roughly proportional with $L$, while the \gls{GPU} implementation is not.
 

### 2012-05-07 ECUA2012/gpu_capon_ecua2012.tex
la2pdf commit - 2012:05:07	13:35:48	(Mon 7.May)

@@
 \makeglossaries
-%\input{../../Library/glossary}
+\input{../../Library/glossary}
 % windows hack for JP
-\renewcommand\gls[1]{#1}
+% \renewcommand\gls[1]{#1}
 
@@
 
-\title{A GPU Implementation of Capon for an Active Sonar System}
+\title{A GPU Implementation of \gls{MVDR} for an Active Sonar System}
 
@@
 % 
-% The Minimum Variance Distortionless Response (MVDR) (or Capon) beamformer, have the potential to overcome this limitation.
+% The Minimum Variance Distortionless Response (MVDR) (or \gls{MVDR}) beamformer, have the potential to overcome this limitation.
 While ensuring unit gain in the look direction, it computes the set of weights that minimizes the energy accumulated by the array from other directions.
@@
 
-The Capon (or \gls{MVDR}) beamformer is one such technique.
+The \gls{MVDR} (or Capon) beamformer is one such technique.
 While ensuring unit gain in the look direction, it computes the set of weights that minimizes the energy accumulated by the array from other directions.
@@
 
-However, despite its inherent potential, the Capon beamformer has yet to see widespread adoption in the active sonar community.
+However, despite its inherent potential, the \gls{MVDR} beamformer has yet to see widespread adoption in the active sonar community.
 There may be several reasons for this.
@@
 
-Perhaps surprisingly, we will show that the suggested means for making the Capon behave well in active systems will also make it less computationally complex.
+Perhaps surprisingly, we will show that the suggested means for making the \gls{MVDR} behave well in active systems will also make it less computationally complex.
 Furthermore, by implementing it on a modern \gls{GPU} we have gained a speedup of 2-3 orders of magnitude compared to optimized single thread implementations in C and Matlab.
 
-The feasibility of using the Capon beamformer in active sonar imaging has been demonstrated in e.g. \cite{Blomberg2012a,Blomberg2011,Dursun2009}.
+The feasibility of using the \gls{MVDR} beamformer in active sonar imaging has been demonstrated in e.g. \cite{Blomberg2012a,Blomberg2011,Dursun2009}.
 E.g.
@@
 REFERENCES!!! (all over..)
-\todo{Capon or MVDR?}
+\todo{\gls{MVDR} or MVDR?}
 % - Who's done stuff on GPU beamforming before?
@@
 \end{align}
-The Capon beamformer \cite{Capon1969} finds the set of weights that minimizes the output power, $\min E\{|z[n]|^2\}$, while ensuring unity gain in the look direction.
+The \gls{MVDR} beamformer \cite{Capon1969} finds the set of weights that minimizes the output power, $\min E\{|z[n]|^2\}$, while ensuring unity gain in the look direction.
 The solution to this optimization problem is given as
@@
 \section{Implementation}
+
 The paradigm of \gls{GPU} computing is formalized as the execution of thousand of threads, where each thread runs a copy of a common function, also know as a kernel.
@@
 The threads index in this grid usually decides the data to be process and the instruction path to go along.
-A modern \gls{GPU} is an hardware architecture comprised of several hundred computing cores, each having access to a very limited amount of local memory an registers.
+
+A modern \gls{GPU} is a hardware architecture comprised of several hundred computing cores, each running a thread that runs a copy of a common function referred to as a kernel.
+This is a design that shows great promise in terms of computing power, but the tradeoff is that each computing core is left with a rather limited amount of local memory, and distributing data to the computing cores and collecting the result is a processes with some inherent latency.
+
+data level of parallelism comes with some challenges though.
+First,  Data must distruted to the threads that need it, and Data must be distributed to   A fundamental limitation with such a design is that each computing core is left with a minimum of local memory, and
+
+having access to a very limited amount of local memory an registers.
 This suggests that the algorithm should be decomposed into as many lightweight threads as possible.
-For a typical array configuration, using a single thread per pixel for Capon is not lightweight enough.
+For a typical array configuration, using a single thread per pixel for \gls{MVDR} is not lightweight enough.
 Therefore, we seek to devise a way to decompose the methods presented in section \ref{methods}.
@@
 In the upcoming results, Host to Device transfer are not taken into account.
-The reasoning behind this is the fact that Capon beamforming requires already delayed data.
+The reasoning behind this is the fact that \gls{MVDR} beamforming requires already delayed data.
 For efficiency, this should also be carried out on the GPU, leaving the challenging task of overcoming the limited Host-Device memory bandwidth to those who implement the delay part of delay-and-sum.
@@
 \begin{itemize}
-\item Choice of robustification parameters $L,M,\Delta$ largely impacts what the ``optimal'' way of implementing Capon 
+\item Choice of robustification parameters $L,M,\Delta$ largely impacts what the ``optimal'' way of implementing \gls{MVDR} 
 \item Introduce CUDA, or ditch it?
@@
 
-We have tested our \gls{GPU} implementation of the Capon beamformer on experimental data from the 32 element Kongsberg Maritime HISAS1030 synthetic aperture sonar~\cite{Hansen2009}.
+We have tested our \gls{GPU} implementation of the \gls{MVDR} beamformer on experimental data from the 32 element Kongsberg Maritime HISAS1030 synthetic aperture sonar~\cite{Hansen2009}.
 The sonar was attached to the HUGIN \gls{AUV} and used in strip-map mode, and the imaged object was the 1500 dwt oil tanker wreck Holmengraa (Figure \ref{holmengraa}).
 Holmengraa is roughly 68\,m long and 9\,m wide, and lies on a slanted seabed at 77\,m depth.
-The Capon image was created with the parameters $L=16$, $N_\text{avg}=3$, and $d=1\%$, which proved to be a reasonable selection for this scenario.
+The \gls{MVDR} image was created with the parameters $L=16$, $N_\text{avg}=3$, and $d=1\%$, which proved to be a reasonable selection for this scenario.
 
@@
 \includegraphics[width=\linewidth]{gfx/benchmark.pdf}
-\caption{Capon benchmarks. \protect\newline A 55\,kpx image processed with $M=32$ channels and $d=2\%$ diagonal loading.}\label{benchmarks}
+\caption{\gls{MVDR} benchmarks. \protect\newline A 55\,kpx image processed with $M=32$ channels and $d=2\%$ diagonal loading.}\label{benchmarks}
 \end{figure}
 
-To test the runtime speed of our \gls{GPU} implementation of Capon, we have compared it to an optimized single thread C and Matlab implementation (Figure \ref{benchmarks}).
+To test the runtime speed of our \gls{GPU} implementation of \gls{MVDR}, we have compared it to an optimized single thread C and Matlab implementation (Figure \ref{benchmarks}).
 The test system was a quad-core Intel Xeon E5620, with 64Gb of \gls{RAM}, and an nVidia Quadro 6000 card.
@@
 
-Figure \ref{holmengraa} demonstrates the Capon beamformer's ability to suppress interference and achieve better detail resolution.
+Figure \ref{holmengraa} demonstrates the \gls{MVDR} beamformer's ability to suppress interference and achieve better detail resolution.
 Compared the \gls{DAS} beamformer, the ship's edges come out as sharper, and the shadows are less noisy.
-But while \gls{DAS} processed this 1.3\,Mpixel image in milliseconds, our Matlab implementation of Capon needed minutes.
+But while \gls{DAS} processed this 1.3\,Mpixel image in milliseconds, our Matlab implementation of \gls{MVDR} needed minutes.
 
@@
 \begin{itemize}
-\item Relate Capon image to DAS image
-\item Discuss the benchmarks
 \item Need for speed: HUGIN 4 banks of 32 elements, can be processed faster than the ping repetition rate, with margins to spare.
@@
 
-\begin{itemize}
-\item When to use what.
-\end{itemize}
+The \gls{MVDR} beamformer is an algorithm well suited for implementation on a \gls{GPU}.
+Each pixel in the image can be processed independently, and the steps within each pixel will also benefit from parallelisation.
+We implemented the \gls{MVDR} bench Compared to an optimized C and Matlab implementation, our \gls{GPU}
+
+We achieved speedups Our \gls{GPU} implementation 
+Using \glspl{GPU} to accelerate the is a
 
@@
 % \item Sonar. Channel count up to 32/64 ish channels?
-% \item Other articles evaluates Capon image quality, and some have shown that speedups can be achieved by reducing the complexity of the traditional Capon algorithm or by implementing specific versions on a GPU.
-% \item This article compare a few promising ways to implement Capon in general, in light of general complexity and suitability for implementation on a CPU and GPU.
+% \item Other articles evaluates \gls{MVDR} image quality, and some have shown that speedups can be achieved by reducing the complexity of the traditional \gls{MVDR} algorithm or by implementing specific versions on a GPU.
+% \item This article compare a few promising ways to implement \gls{MVDR} in general, in light of general complexity and suitability for implementation on a CPU and GPU.
 % \end{itemize}
@@
 Constraints must be applied in one way or another - parameters must be tuned.
-% \item The three main ways to implement Capon \cite{Capon1969} is \todo{not entirely true?}
+% \item The three main ways to implement \gls{MVDR} \cite{\gls{MVDR}1969} is \todo{not entirely true?}
 % \begin{itemize}
@@
 % \end{itemize}
-% \item Choice of robustification parameters $L,M,\Delta$ largely impacts what the ``optimal'' way of implementing Capon 
+% \item Choice of robustification parameters $L,M,\Delta$ largely impacts what the ``optimal'' way of implementing \gls{MVDR} 
 % \item In most imaging applications 
@@
 % \begin{itemize}
-% \item Capon's resonse when applying robustification
+% \item \gls{MVDR}'s resonse when applying robustification
 % \item Choice of window functions makes little difference.
@@
 % 
-% Basically, we are working with a practical implementation of the Capon beamformer that computes a set of weights $\vec w$ for every single sample $n$ by solving:
+% Basically, we are working with a practical implementation of the \gls{MVDR} beamformer that computes a set of weights $\vec w$ for every single sample $n$ by solving:
 % \begin{gather*}

### 2012-05-07 ECUA2012/gpu_capon_ecua2012.tex
la2pdf commit - 2012:05:07	14:06:07	(Mon 7.May)

@@
 The \gls{MVDR} (or Capon) beamformer is one such technique.
-While ensuring unit gain in the look direction, it computes the set of weights that minimizes the energy accumulated by the array from other directions.
+While ensuring unit gain in the look direction, it computes the set of weights that minimizes the power accumulated by the array from other directions.
 Whenever the distribution of noise and interference is primarily focused around specific angles, an array response with a high degree of suppression at these angles are chosen, resulting in an image with much improved lateral detail level.
@@
 
-The paradigm of \gls{GPU} computing is formalized as the execution of thousand of threads, where each thread runs a copy of a common function, also know as a kernel.
-Threads are organized in blocks, and blocks further into a grid of blocks.
-The threads index in this grid usually decides the data to be process and the instruction path to go along.
+Beamforming is a process well suited for acceleration by parallelisation, because each pixel can be processed idependently.
 
-A modern \gls{GPU} is a hardware architecture comprised of several hundred computing cores, each running a thread that runs a copy of a common function referred to as a kernel.
-This is a design that shows great promise in terms of computing power, but the tradeoff is that each computing core is left with a rather limited amount of local memory, and distributing data to the computing cores and collecting the result is a processes with some inherent latency.
+A modern \gls{GPU} is a hardware architecture comprised of several hundred computing cores, each running a thread that executes a copy of a common function called a kernel.
+However, there are some inherent drawbacks with such a design: Each core is left with a rather limited amount of local memory, and distributing data to the computing cores and collecting the result is a processes with some inherent latency.
 
-data level of parallelism comes with some challenges though.
-First,  Data must distruted to the threads that need it, and Data must be distributed to   A fundamental limitation with such a design is that each computing core is left with a minimum of local memory, and
-
-having access to a very limited amount of local memory an registers.
-This suggests that the algorithm should be decomposed into as many lightweight threads as possible.
-For a typical array configuration, using a single thread per pixel for \gls{MVDR} is not lightweight enough.
+This suggests that the algorithm should be decomposed into as many lightweight threads as possible, which are both computationally intensive and light on the memory consumption.
+For a typical array configuration, using a single thread per pixel is not lightweight enough.
 Therefore, we seek to devise a way to decompose the methods presented in section \ref{methods}.
@@
 Inverting $\eR$ is by far the most computationally intensive task, and a key area of focus for further improvements.
-\item \emph{Computing the subarray weights} by substituting $\beta = \eRi[n]\1$ into equation \ref{weights}, $\vec w[n] = \frac{\beta}{\1\T\beta}$.
+\item \emph{Computing the beamformer output} $z[n]$.
+by substituting $\beta = \eRi[n]\1$ into equation \ref{weights}, $\vec w[n] = \frac{\beta}{\1\T\beta}$.
 A group of L threads per pixel was used to reduce shared memory pressure and to obtain coalesced reads and a minimum of writes.

### 2012-05-07 ECUA2012/gpu_capon_ecua2012.tex
la2pdf commit - 2012:05:07	14:38:49	(Mon 7.May)

@@
 
-Modern phased array active sonar systems are often limited by their lateral resolution and \gls{SNR}, which is due to limited lateral extent of the array the spatial extent of the array sample the extent of the array is too small compared to the wavelength of the received signal, and while it is possible to trade resolution for improved \gls{SNR} by applying various weightsets to the array channels, one always end up with a compromise between the two.
+Modern phased array active sonar systems are often limited by their lateral resolution and \gls{SNR}.
+This is due to the limited lateral extent of the array, and while it is possible to trade resolution for improved \gls{SNR} by applying various weightsets to the array channels, one always end up with a compromise between the two.
 By introducing adaptive techniques, however, it is possible to resolve lateral details that would otherwise have been missed.
@@
 While ensuring unit gain in the look direction, it computes the set of weights that minimizes the power accumulated by the array from other directions.
-Whenever the distribution of noise and interference is primarily focused around specific angles, an array response with a high degree of suppression at these angles are chosen, resulting in an image with much improved lateral detail level.
+This ability to suppress interference power improves the image detail level and \gls{SNR}, and the effect can be dramatic if the noise and interference is primarily focused around specific angles.The feasibility of using the \gls{MVDR} beamformer in active sonar imaging has been demonstrated in e.g.
+\cite{Blomberg2012a,Blomberg2011,Dursun2009}.
 
@@
 Second, in its original form, the computational complexity is cubic with the number of channels, O(M$^3$), while \gls{DAS} is only at O($M$).
+This makes it impractical in many situations, and even infeasible in high channel systems.
+\todo{mention disadvantage of tuning parameters?}
 
-Perhaps surprisingly, we will show that the suggested means for making the \gls{MVDR} behave well in active systems will also make it less computationally complex.
-Furthermore, by implementing it on a modern \gls{GPU} we have gained a speedup of 2-3 orders of magnitude compared to optimized single thread implementations in C and Matlab.
+We will show that the suggested means for making the \gls{MVDR} behave well in active systems will also make it less computationally complex.
+Furthermore, by implementing it on a modern \gls{GPU} we have gained a speedup of around 2 orders of magnitude compared to an optimized single thread implementations in C.
 
-The feasibility of using the \gls{MVDR} beamformer in active sonar imaging has been demonstrated in e.g. \cite{Blomberg2012a,Blomberg2011,Dursun2009}.
-E.g.
-
-GPU ultrasound (\cite{So2011,Chen2011})
-
-Complex data, all robustification techniques.
-
-\begin{itemize}
-\item Adaptive beamformer's potential lies in its ability to suppress interference power
-\item Why adaptive beamformers struggle in active sonar systems.
-Correlated noise, robustification kills the adaptive potential.
-Quite computationally intensive.
-Constraints must be applied in one way or another - parameters must be tuned.
-\item CUDA?
-\end{itemize}
-
-REFERENCES!!! (all over..)
-\todo{\gls{MVDR} or MVDR?}
-% - Who's done stuff on GPU beamforming before?
-% 
-% - Complex data
-% -
+Yay or nay? $\Rightarrow$ GPU ultrasound (\cite{So2011,Chen2011}). Complex data, all robustification techniques.
 
@@
 
-Beamforming is a process well suited for acceleration by parallelisation, because each pixel can be processed idependently.
+Beamforming is a process well suited for acceleration by parallelisation, because each pixel can be processed idependently. \todo{blend it in}
 

### 2012-05-07 ECUA2012/gpu_capon_ecua2012.tex
la2pdf commit - 2012:05:07	15:09:00	(Mon 7.May)

@@
 
-\title{A GPU Implementation of \gls{MVDR} for an Active Sonar System}
+\title{A GPU Implementation of MVDR for an Active Sonar System}
 
@@
 
-The \gls{MVDR} (or Capon) beamformer is one such technique.
+The Minimum Variance Distortionless Response (MVDR, or Capon) beamformer is one such technique.
 While ensuring unit gain in the look direction, it computes the set of weights that minimizes the power accumulated by the array from other directions.
-This ability to suppress interference power improves the image detail level and \gls{SNR}, and the effect can be dramatic if the noise and interference is primarily focused around specific angles.The feasibility of using the \gls{MVDR} beamformer in active sonar imaging has been demonstrated in e.g.
-\cite{Blomberg2012a,Blomberg2011,Dursun2009}.
+This ability to suppress interference power improves the image detail level and \gls{SNR}, and the effect can be dramatic if the noise and interference is primarily focused around specific angles.
+The feasibility of using the \gls{MVDR} beamformer in active sonar imaging has been demonstrated in e.g.~\cite{Blomberg2012a,Blomberg2011,Dursun2009}.
 
@@
 Second, in its original form, the computational complexity is cubic with the number of channels, O(M$^3$), while \gls{DAS} is only at O($M$).
-This makes it impractical in many situations, and even infeasible in high channel systems.
+This makes it impractical in many situations, and even infeasible in high channel systems~\cite{So2011}.
 \todo{mention disadvantage of tuning parameters?}
@@
 
-Yay or nay? $\Rightarrow$ GPU ultrasound (\cite{So2011,Chen2011}). Complex data, all robustification techniques.
+% Yay or nay? $\Rightarrow$ GPU ultrasound (\cite{So2011,Chen2011}). Complex data, all robustification techniques.
 
@@
 
-Beamforming is a process well suited for acceleration by parallelisation, because each pixel can be processed idependently. \todo{blend it in}
-
+When processing images it might be well worth having a look at the \gls{GPU}, as this is a task it was made for.
 A modern \gls{GPU} is a hardware architecture comprised of several hundred computing cores, each running a thread that executes a copy of a common function called a kernel.
@@
 This suggests that the algorithm should be decomposed into as many lightweight threads as possible, which are both computationally intensive and light on the memory consumption.
-For a typical array configuration, using a single thread per pixel is not lightweight enough.
+Beamforming fits well into this scheme, as each pixel can be processed idependently.
+However, for a typical array configuration, using a single thread per pixel is not lightweight enough.
 Therefore, we seek to devise a way to decompose the methods presented in section \ref{methods}.
@@
 A group of threads are created that slide along the diagonals of $\eR$.
-In this way we can exploit the fact that entries on the diagonals overlaps across subarrays and time.
+In this way we exploit the fact that entries on the diagonals overlap across subarrays and time, and keep the numbers of both data reads and writes at a minimum.
 %Finished threads can also wrap around and start on diagonals on the lower half triangular. 
-In this way we process one row of $\eR$ per kernel iteration, and manage to keep the numbers of both data reads and writes at a minimum.
+% In this way we process one row of $\eR$ per kernel iteration, and manage to keep the numbers of both data reads and writes at a minimum.
 \item \emph{Computing} $\eRi[n]\1$ (from equation \ref{weights}).
 While intuition may suggest that this step is carried out by inverting $\eR$, it is better to solve the linear equation $\R\beta = \1$ for $\beta$ instead, which gives us $\beta = \eRi\1$ directly.
-We have tested various solvers for this task, both in-house and proprietary implementations, but achieved the best performance by using an unofficial batch solver from nVidia.
+We have tested various solvers for this task, both in-house and proprietary implementations, and achieved the best performance by using an unofficial batch solver from nVidia.
 Inverting $\eR$ is by far the most computationally intensive task, and a key area of focus for further improvements.

### 2012-05-07 ECUA2012/gpu_capon_ecua2012.tex
la2pdf commit - 2012:05:07	16:47:08	(Mon 7.May)

@@
 
-\title{A GPU Implementation of MVDR for an Active Sonar System}
+\title{GPU-based Adaptive Beamforming for Active Sonar Imaging}
 
@@
 Inverting $\eR$ is by far the most computationally intensive task, and a key area of focus for further improvements.
-\item \emph{Computing the beamformer output} $z[n]$.
-by substituting $\beta = \eRi[n]\1$ into equation \ref{weights}, $\vec w[n] = \frac{\beta}{\1\T\beta}$.
+\item \emph{Computing the beamformer output} $z[n]$ by substituting $\beta = \eRi[n]\1$ into equation \ref{weights}, $\vec w[n] = \frac{\beta}{\1\T\beta}$.
 A group of L threads per pixel was used to reduce shared memory pressure and to obtain coalesced reads and a minimum of writes.
@@
 To test the runtime speed of our \gls{GPU} implementation of \gls{MVDR}, we have compared it to an optimized single thread C and Matlab implementation (Figure \ref{benchmarks}).
+For this test we used a 55\,kpixels sectorscan image.
 The test system was a quad-core Intel Xeon E5620, with 64Gb of \gls{RAM}, and an nVidia Quadro 6000 card.
@@
 We kept the channel count at $M=32$ and added $d=2\%$ diagonal loading.
-The test image was a 55\,kpixels sectorscan.
 

### 2012-05-07 ECUA2012/gpu_capon_ecua2012.tex
la2pdf commit - 2012:05:07	17:36:43	(Mon 7.May)

@@
 
+When designing an active sonar system one usually finds the image quality limited by the laterial resolution and contrast.
+This is a physical limitation related to the array geometry and the frequency of the transmitted signal.
+
 The Minimum Variance Distortionless Response (MVDR, or Capon) beamformer is one such technique.
@@
 Both the C and Matlab implementation in this benchmark take this symmetry into account, which explains why the complexity curve of these implementation is roughly proportional with $L$, while the \gls{GPU} implementation is not.
+
+HUGIN: 1.8m/s - range 250m \\
+fs = 100kHz*30\%*$\frac{4}{3}$ = 40kHz \\
+$N_{range px} = \frac{2\,250m}{1500m/s}40kHz = 13.3kpx$ \\
+$\times 32$ beams = 426kpx \\\\
+
+PRF = $\frac{1500m/s}{2 250m} = 3Hz$ \\
+TP = 215px 3Hz = 1.28Mpx/s \\
+4 arrays: 1.28Mpx/s * 4 = 5.12Mpx/s
 

### 2012-05-07 ECUA2012/gpu_capon_ecua2012.tex
la2pdf commit - 2012:05:07	18:07:08	(Mon 7.May)

@@
 
-\usepackage{tikz}
-\usetikzlibrary{shapes,snakes}
+% \usepackage{tikz}
+% \usetikzlibrary{shapes,snakes}
 \usepackage{amsmath,amssymb}
@@
 When designing an active sonar system one usually finds the image quality limited by the laterial resolution and contrast.
-This is a physical limitation related to the array geometry and the frequency of the transmitted signal.
+This is a physical limitation related to the lateral array geometry and the frequency of the transmitted signal.
 
@@
 This ability to suppress interference power improves the image detail level and \gls{SNR}, and the effect can be dramatic if the noise and interference is primarily focused around specific angles.
-The feasibility of using the \gls{MVDR} beamformer in active sonar imaging has been demonstrated in e.g.~\cite{Blomberg2012a,Blomberg2011,Dursun2009}.
+The feasibility of using the \gls{MVDR} beamformer in active sonar imaging has been demonstrated in e.g.~\cite{Blomberg2012a,Blomberg2011,Dursun2009,Lo2004}.
 
@@
 \newpage
-\begin{figure}[!t]
-\centering
-\includegraphics[width=\linewidth]{gfx/img_holmengraa.pdf}
-\caption{HISAS sidescan sonar (SSS) image of the shipwreck Holmengraa.}\label{holmengraa}
-\end{figure}
-
 \section{Implementation}
@@
 
+\begin{figure}[!t]
+\centering
+\includegraphics[width=\linewidth]{gfx/img_holmengraa.pdf}
+\caption{HISAS sidescan sonar (SSS) image of the shipwreck Holmengraa.}\label{holmengraa}
+\end{figure}
+
 TODO:
@@
 \newpage
-
 \section{Results}
@@
 
-% A sidescan image of the 1500 dwt oil tanker wreck Holmengraa is shown in \Fig{holmengraa}.
-depth~\cite{holmengraa}.
-The sidescan image was created using data from the HISAS 1030 sonar, which is rather unsuited for this purpose because of its large opening angle.
-This, and the fact that the wreck was imaged at a range of about 105\,m makes the image quality poor, but sufficient to compare our beamformers.
-
-% % In the Holmengraa image we note again that the LCA produces a cleaner shadow and better edge definition. The MV method performed almost identically to LCA in this case, and was omitted.
-% 
-% 
-% 
-% down to these cores and back to the \gls{CPU} is also when copying data to and from the \gls{GPU} are also rather high, so each core should be   performing calculations there, and copying the data back to the \gls{CPU}
-% 
-%  but in order to fully exploit this potential we must be aware of its limitations.  
-% 
-% When adapting an algorithm to a massively parallel hardware such as \glspl{GPU}, one must be familiar with its limitations in order to that the potential that lies within these chips can not be fully realised
-% 
-% When trying to fit any algorithm to massively parallel hardware such as \glspl{GPU}, one must be careful to select a proper decomposition scheme for the device at hand.
-Modern \glspl{GPU} have several hundred computing cores, but only a limited amount of  are  various decomposition schemes should be contemplated before performing the implecareful decomposition of the
-% 
-% 
-% Let us first consider the \emph{computation of $\hat{\mat R}$}.
-Notice that when implementing the formula directly a lot of summations and multiplications will be repeated unnecessary.
-To illustrate this, consider for instance a case with $M=6$ sensors that are grouped into subarrays of length $L=3$ where all but one element overlap.
-Let us further assume that we need to average over 7 time samples, i.e.
-that $N_{\text{avg}} = 3$.
-% 
-% These procedures are commutative, so the order in which we perform them does not matter.
-But for the sake of this example let us start by performing the time averaging.
-Assuming that $\vec x_i$ and $\vec x_j$ have a zero mean, a time averaged covariance coefficient $r_{ij}$ in $\hat{\mat R}$ may be obtained by (omitting the normalisation):
-% \begin{align*}
-% r_{ij}[n] &= \sumb{n'=n-N_{\text{avg}}}{n+N_{\text{avg}}} x_i[n']\,x_j[n'],
-% \end{align*}
-% which can be easily rewritten in a sliding window fashion
-% \begin{align*}
-% \hat r_{ij}[n] &= r_{ij}[n-1] - x_i[n\text{-}N_\text{avg}\text{-}1]\,x_j[n\text{-}N_\text{avg}\text{-}1]
-%                          + x_i[n\text{+}N_\text{avg}]\,x_j[n\text{+}N_\text{avg}].
-% \end{align*}
-% Hence, whenever $N_\text{avg}>1$ the sliding window approach will demand less instructions at the expense of a slightly higher memory consumption.
-The next thing to realise is that not all $r_{ij}[n]$ need to be computed.
-Since $r_{ji} = r^*_{ij}$, and the subarrays only span $L$ elements, we only compute an upper triangular band of R (marked \Green{green})
-% \begin{figure}[H]
-% % \begin{narrow}{-3.5cm}{-3.5cm}
-% \includegraphics[width=\linewidth]{gfx/capon_build_R_full.eps}
-% \caption{Building $\hat{\mat R}$: Symmetry and element coverage}\label{buildingRfull}
-% % \end{narrow}
-% \end{figure}
-% The final step is to average all the subarray matrices to form $\hat{\mat R}$ by 
-% {\renewcommand*{\arraystretch}{1.8}
-% \begin{align*}
-% \hat{\mat R}_\text{sub} &= {\color{Blue}\mat R_{\text{sub},1} + \mat R_{\text{sub},2} + \mat R_{\text{sub},3} + \mat R_{\text{sub},4}} \\
-% &= \left[\;
-% \begin{matrix}
-% \color{Green}r_{00}+r_{11}+r_{22} & \color{Red}r_{01}+r_{12}+r_{23} & r_{02}+r_{13}+r_{24} \\
-%                                        & \color{Green}r_{11}+r_{22}+r_{33} & \color{Red}r_{12}+r_{23}+r_{34} \\
-%                                        &                                        & \color{Green}r_{22}+r_{33}+r_{44}
-% \end{matrix}
-% \;\right]
-% \end{align*}}%
-% Note that consecutive elements along the diagonals overlap on all but one covariance coefficient, hence a sliding window may also be used here.
-The subarray size $L$ is usually quite a bit larger than the time averaging window, so there is more to gain from implementing the sliding window here.\\\\
-% 
-% {\bf Method complexity:}\\
-% \begin{tabular}{l l}
-% No optimizations            & $2M^2\,(2\Navg+1)$ \\
-% Banded symmetric            & $L(2M-L+1)\,(2\Navg+1)$ \\
-% Sliding, banded, symmetric  & $1.5L(2M-L+1)$ \\
-% \end{tabular}\\\\
-
-% \subsubsection{Computing $\text{\raisebox{1.9pt}{$\vec\chi$}} = \hat{\mat R}\,\!^{-1}\vec a$}
-% 
-% There are two ways of finding $\text{\raisebox{1.9pt}{$\vec\chi$}} = \hat{\mat R}\,\!^{-1}\vec a$, either by solving the equation directly or by inverting $\hat{\mat R}$ first.
-Solving the equation is less computationally intensive than inverting $\hat{\mat R}$, and seems like the obvious choice.
-However, with iterative methods for updating $\eRi$ for each range pixel the situation is changed...
-% 
-% 
-% \section{Results/Discussion}
-% 
-% 
-% 
-% 
-% 
-% \todopar{Case:
-% \begin{itemize}
-% \item JOE
-% \item Sonar. Channel count up to 32/64 ish channels?
-% \item Other articles evaluates \gls{MVDR} image quality, and some have shown that speedups can be achieved by reducing the complexity of the traditional \gls{MVDR} algorithm or by implementing specific versions on a GPU.
-% \item This article compare a few promising ways to implement \gls{MVDR} in general, in light of general complexity and suitability for implementation on a CPU and GPU.
-% \end{itemize}
-% }
-% 
-% \begin{itemize}
-% \item As briefly as possible, introduce beamforming.
-% \item Adaptive beamformer's potential lies in its ability to suppress interference power
-% \item Why adaptive beamformers struggle in active sonar systems.
-Correlated noise, robustification kills the adaptive potential.
-Quite computationally intensive.
-Constraints must be applied in one way or another - parameters must be tuned.
-% \item The three main ways to implement \gls{MVDR} \cite{\gls{MVDR}1969} is \todo{not entirely true?}
-% \begin{itemize}
-% \item Traditional way: Build $\eR$ and solve $\w = \frac{\Ri\a}{\a\H\Ri\a}$. 
-% \item Iterative methods: Woodbury, Conjugated gradients
-% \item Beamspace
-% \item LCA (or other methods, as reference)
-% \end{itemize}
-% \item Choice of robustification parameters $L,M,\Delta$ largely impacts what the ``optimal'' way of implementing \gls{MVDR} 
-% \item In most imaging applications 
-% Outline:
-% \begin{itemize}
-% \item Speeding up: Build $\eR$ and solve $\w = \frac{\Ri\a}{\a\H\Ri\a}$. 
-% \item Iterative methods: Woodbury (Sherman Morrison), Conjugated gradients - show performance of GPU performance.
-% \item Woodbury: Must take the average over many samples in space and time - lots of (small) outerproducts. Can be used if not a lot of averaging is required.
-% \item Beamspace
-% \item LCA (or other methods, as reference)
-% \end{itemize}
-% \end{itemize}
-%
-
-% \ \\
-% \IEEEPARstart{T}{o} form images from a modern phased array sonar system the received wavefield is usually recorded, and then postprocessed by a digital beamformer.
-The beamformer applies delays and weights to the sensor channels, the beamformer adjusts the arrays spatial response to focus at one pixel at a time.
-such that signals emanating from regions of interest add constructively, while ensuring that noise and interference from other angles do not.
-% 
-% The imaging capabilities of a modern phased array sonar system depend on physical attributes such element response and array geometry, the transmitted signal, as well as the beamforming method being used on transmission and reception.
-Beamforming is the concept of applying delays and weights to the sensors channels to steer the arrays response to points of interest.
-
-% 
-% 
-% Outline:
-% \begin{itemize}
-% \item \gls{MVDR}'s resonse when applying robustification
-% \item Choice of window functions makes little difference.
-% \item Steering and mainlobewidths have outer bounds.
-% \item Beamspace?
-% \item Chosen window plots - what may they tell us? Variance intensity values when using various windows.
-% \item Assymmetric windows?
-% \end{itemize}
-% 
-% 
-% \begin{align}
-% z[n] = \sumb{m=0}{M-1} w_m[n]^*x_m[n-\Delta_m] = \w\H[n]\x[n-\Delta_m]
-% \end{align}
-% 
-% 
-% \section{Methods}
-% 
-% Basically, we are working with a practical implementation of the \gls{MVDR} beamformer that computes a set of weights $\vec w$ for every single sample $n$ by solving:
-% \begin{gather*}
-% \vec w[n] = \frac{\hat{\mat R}\,\!^{-1}[n]\vec a}{\vec a\H\hat{\mat R}\,\!^{-1}[n]\vec a} = \frac{\text{\raisebox{1.9pt}{$\vec\chi$}}[n]}{\vec a\H\text{\raisebox{1.9pt}{$\vec\chi$}}[n]}
-% \end{gather*}%
-% where
-% % \newcommand\X{\text{\raisebox{2pt}{$\vec\chi$}}}
-% \begin{gather*}
-% \text{\raisebox{1.9pt}{$\vec\chi$}} = \hat{\mat R}\,\!^{-1}\vec a \qquad\qquad\text{is the solution to}\qquad\qquad \hat{\mat R}\text{\raisebox{1.9pt}{$\vec\chi$}} = \vec a.
-% \end{gather*}
-
-% 
-% , and maximum suppression of while ensuring that the beamformer digitally  before each of the pixels are estimated one at a time.
-The resolution and contrast of such a system will depend on the systems spatial response, which ideally should be narrow  be very sharp in the desired direction its ability to achieve  fundamental principle of forming a sonar image is to record the received wavefield,
-% 
-% image quality of a phased array sonar imaging systems depend on  the choice of weights to apply to each of the sensors are crucial. 
-% 
-% A modern phased array imaging system may be thought of as a spatial filter. To achieve the best possible performance, the 
-% 
-% resolution and contrast 
-% 
-% Adaptive beamformers have only recently been introduced in active sonar imaging.
-For a while they were considered unsuited for this purpose because the backscattered signal is largely correlated with the
-% 
-% 
-%
-
-%\begin{figure}[!t]
-%\centering
-%\includegraphics[width=2.5in]{myfigure}
-% where an .eps filename suffix will be assumed under latex, 
-% and a .pdf suffix will be assumed for pdflatex; or what has been declared
-% via \DeclareGraphicsExtensions.
-%\caption{Simulation Results}
-%\label{fig_sim}
-%\end{figure}
-
-% An example of a double column floating figure using two subfigures.
-% (The subfig.sty package must be loaded for this to work.)
-% The subfigure \label commands are set within each subfloat command, the
-% \label for the overall figure must come after \caption.
-% \hfil must be used as a separator to get equal spacing.
-% The subfigure.sty package works much the same way, except \subfigure is
-% used instead of \subfloat.
-%
-% \begin{figure*}[!t]
-% \centerline{\subfloat[Case I]\includegraphics[width=2.5in]{subfigcase1}%
-% \label{fig_first_case}}
-% \hfil
-% \subfloat[Case II]{\includegraphics[width=2.5in]{subfigcase2}%
-% \label{fig_second_case}}}
-% \caption{Simulation results}
-% \label{fig_sim}
-% \end{figure*}
-%
-% Note that often IEEE papers with subfigures do not employ subfigure
-% captions (using the optional argument to \subfloat), but instead will
-% reference/describe all of them (a), (b), etc., within the main caption.
-
-% An example of a floating table. Note that, for IEEE style tables, the 
-% \caption command should come BEFORE the table. Table text will default to
-% \footnotesize as IEEE normally uses this smaller font for tables.
-% The \label must come after \caption as always.
-%
-%\begin{table}[!t]
-%% increase table row spacing, adjust to taste
-%\renewcommand{\arraystretch}{1.3}
-% if using array.sty, it might be a good idea to tweak the value of
-% \extrarowheight as needed to properly center the text within the cells
-%\caption{An Example of a Table}
-%\label{table_example}
-%\centering
-%% Some packages, such as MDW tools, offer better commands for making tables
-%% than the plain LaTeX2e tabular which is used here.
-%\begin{tabular}{|c||c|}
-%\hline
-%One & Two\\
-%\hline
-%Three & Four\\
-%\hline
-%\end{tabular}
-%\end{table}
-
-% Note that IEEE does not put floats in the very first column - or typically
-% anywhere on the first page for that matter. Also, in-text middle ("here")
-% positioning is not used. Most IEEE journals use top floats exclusively.
-% However, Computer Society journals sometimes do use bottom floats - bear
-% this in mind when choosing appropriate optional arguments for the
-% figure/table environments.
-% Note that, LaTeX2e, unlike IEEE journals, places footnotes above bottom
-% floats. This can be corrected via the \fnbelowfloat command of the
-% stfloats package.
-% 
-% 
-% 
-% \section{Conclusion}
-% The conclusion goes here.
-% 
-% 
-% 
-% 
-% 
-% % if have a single appendix:
-% %\appendix[Proof of the Zonklar Equations]
-% % or
-% %\appendix  % for no appendix heading
-% % do not use \section anymore after \appendix, only \section*
-% % is possibly needed
-% 
-% % use appendices with more than one appendix
-% % then use \section to start each appendix
-% % you must declare a \section before using any
-% % \subsection or using \label (\appendices by itself
-% % starts a section numbered zero.)
-% %
-
 %%%%%%%%%%%%%%%%%%                              ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
@@
 %
-\section{Acknowledgments}
+
+{\Large\bf\MakeUppercase Acknowledgments}
 

### 2012-05-07 ECUA2012/gpu_capon_ecua2012.tex
la2pdf commit - 2012:05:07	18:38:49	(Mon 7.May)
la2pdf commit - 2012:05:07	19:09:07	(Mon 7.May)
la2pdf commit - 2012:05:07	19:40:12	(Mon 7.May)

@@
 
-Modern phased array active sonar systems are often limited by their lateral resolution and \gls{SNR}.
-This is due to the limited lateral extent of the array, and while it is possible to trade resolution for improved \gls{SNR} by applying various weightsets to the array channels, one always end up with a compromise between the two.
-By introducing adaptive techniques, however, it is possible to resolve lateral details that would otherwise have been missed.
+In a typical active sonar imaging scenario encoded signals are transmitted towards a surface of interest, and the resulting reflections are recorded using an array of receivers.
+The receiver outputs can then be coherently combined in the post-processing to focus on specific points in the image.
+The combination process involves assigning suitable delays and weights to the array's channels, and is commonly referred to as beamforming.
 
-When designing an active sonar system one usually finds the image quality limited by the laterial resolution and contrast.
-This is a physical limitation related to the lateral array geometry and the frequency of the transmitted signal.
+In conventional beamforming the array sensors are assigned a static set of weights.
+This allows lateral resolution to be traded for improved noise suppression (contrast), but one always ends up with a compromise between the two.
+Adaptive beamformers target this limitation by allowing the weights to change to better fit the dynamic nature of the incoming wavefield.
 
@@
 
-%
+% \titleformat{<command>}[<shape>]{<format}{<label>}{<sep>}{<before>}[<after>]
 
-{\Large\bf\MakeUppercase Acknowledgments}
+% \titleformat{\section}[hang]{\bf}
+% {\thesection.\enspace}%\thesubsubsection}%  {\footnotesize \enspace \emph{Sec.}  }
+%    {0pt}{\MakeUppercase}[]
+% \titlespacing*{\section}{0pt}{2\lineheight}{\lineheight}
 
-The authors would like to thank...
+% {-2ex plus -.5ex minus -.2ex}{1.0ex plus .2ex minus .2ex}
+
+% \renewcommand\section[1]{{\Large\bf\MakeUppercase #1}}
+% \renewcommand\section*[1]{{\Large\bf\MakeUppercase #1}}
+
+\section*{Acknowledgments}
+
+The authors would like to thank nVidia for providing us their unofficial linear equation batch solver, and Kongsberg Maritime and the Norwegian Defence Research Establishment (FFI) for providing us with experimental data.
 

### 2012-05-07 ECUA2012/gpu_capon_ecua2012.tex
la2pdf commit - 2012:05:07	20:10:32	(Mon 7.May)

@@
 In a typical active sonar imaging scenario encoded signals are transmitted towards a surface of interest, and the resulting reflections are recorded using an array of receivers.
-The receiver outputs can then be coherently combined in the post-processing to focus on specific points in the image.
-The combination process involves assigning suitable delays and weights to the array's channels, and is commonly referred to as beamforming.
+These receivers may be independently delayed and weighted to adjust the array's point of focus, a concept commonly referred to as beamforming.
 
@@
 While ensuring unit gain in the look direction, it computes the set of weights that minimizes the power accumulated by the array from other directions.
-This ability to suppress interference power improves the image detail level and \gls{SNR}, and the effect can be dramatic if the noise and interference is primarily focused around specific angles.
-The feasibility of using the \gls{MVDR} beamformer in active sonar imaging has been demonstrated in e.g.~\cite{Blomberg2012a,Blomberg2011,Dursun2009,Lo2004}.
+Its ability to suppress interference power and improve image quality in active sonar systems has been demonstrated in e.g.~\cite{Blomberg2012a,Blomberg2011,Dursun2009,Lo2004}.
 
@@
 There may be several reasons for this.
-First, the method is not inherently robust, and will suffer from a phenomena called signal cancellation in active systems.
-Second, in its original form, the computational complexity is cubic with the number of channels, O(M$^3$), while \gls{DAS} is only at O($M$).
-This makes it impractical in many situations, and even infeasible in high channel systems~\cite{So2011}.
-\todo{mention disadvantage of tuning parameters?}
+For one, the method is not inherently robust, and will suffer from a phenomena called signal cancellation in active systems.
+Another reason is that in its original form, the computational complexity is cubic with the number of channels, O(M$^3$), while conventional beamformers are usually only at O($M$).
 
 We will show that the suggested means for making the \gls{MVDR} behave well in active systems will also make it less computationally complex.
-Furthermore, by implementing it on a modern \gls{GPU} we have gained a speedup of around 2 orders of magnitude compared to an optimized single thread implementations in C.
+Furthermore, by implementing it on a modern \gls{GPU} we have gained a speedup of around 2 orders of magnitude compared to an optimized single thread implementation in C.
 
@@
 \section{Methods}\label{methods}
+
+The receiver outputs can then be coherently combined in the post-processing to focus on specific points in the image.
+The combination process involves assigning suitable delays and weights to the array's channels, and is commonly referred to as beamforming.
 
@@
 
+In a sonar processing chain, using one or more \glspl{GPU} can help offload the CPU.
+
 We achieved speedups Our \gls{GPU} implementation 

### 2012-05-07 ECUA2012/gpu_capon_ecua2012.tex
la2pdf commit - 2012:05:07	21:12:22	(Mon 7.May)

@@
 In a typical active sonar imaging scenario encoded signals are transmitted towards a surface of interest, and the resulting reflections are recorded using an array of receivers.
-These receivers may be independently delayed and weighted to adjust the array's point of focus, a concept commonly referred to as beamforming.
+The receiver outputs are then coherently combined in the post-processing to focus on specific points in the image.
+This involves assigning suitable delays and weights to the array's channels, a concept commonly referred to as beamforming.
 
@@
 
-The receiver outputs can then be coherently combined in the post-processing to focus on specific points in the image.
-The combination process involves assigning suitable delays and weights to the array's channels, and is commonly referred to as beamforming.
-
-In a typical active sonar imaging scenario encoded signals are transmitted towards a surface interest, and the resulting reflections are recorded using an array of receivers.
-This is favourable setup as it allows the receiver outputs to be coherently combined in the post-processing to focus on specific points in the image.
-The combination process involves assigning suitable delays and weights to the array's channels, and is commonly referred to as beamforming.
+The fundamental problem of sonar imaging is to estimate each pixel's individual contribution to the final recorded echo.
+If the data is digitally stored, we can perform this task by focusing on one pixel at a time in the post-processing.
+Since the geometric relationship of the array and image is known, one usually start by selecting delays that mitigates the difference in transmitter-pixel-receiver propagation time.
+the delays are usually simply chosen to .
 

### 2012-05-07 ECUA2012/gpu_capon_ecua2012.tex
la2pdf commit - 2012:05:07	21:42:24	(Mon 7.May)

@@
 
-The fundamental problem of sonar imaging is to estimate each pixel's individual contribution to the final recorded echo.
-If the data is digitally stored, we can perform this task by focusing on one pixel at a time in the post-processing.
+The fundamental task of sonar imaging is to estimate each pixel's individual contribution to the final recorded echo.
+If the data is digitally stored, we can do this by focusing on one pixel at a time in the post-processing.
 Since the geometric relationship of the array and image is known, one usually start by selecting delays that mitigates the difference in transmitter-pixel-receiver propagation time.
-the delays are usually simply chosen to .
 
-Now consider an $M$ element phased array.
-Let the sample recorded by the $m$'th channel at time instant $n$ be represented as $x_m[n]$, and the accompanying weights and delays be  $w_m[n]$ and $\Delta_m[n]$.
-The beamformer output $z[n]$ is then given as
+Consider an $M$ element phased array.
+Let the sample recorded by the $m$'th channel at time instant $n$ be represented as $x_m[n]$, and the accompanying weights be  $w_m[n]$ and $\Delta_m[n]$.
+To simplify the upcoming notation, let sample $x_m[n]$ delayed by a factor $\Delta_m$ be written as
 \begin{align}
-z[n] = \w\H[n]\x[n] = \bmat{w_0[n]\\w_1[n]\\\vdots\\w_{M-1}[n]}^H \bmat{x_0[n-\Delta_0]\\x_1[n-\Delta_1]\\\vdots\\x_{M-1}[n-\Delta_{M-1}]}.\label{z}
+x^\Delta_m = x_m[n-\Delta_m[n]].
+\end{align}
+The beamformer output $z[n]$ is then simply the weighted sum of all the delayed data samples:
+\begin{align}
+z[n] = \w\H[n]\x[n] = \bmat{w_0[n]\\w_1[n]\\\vdots\\w_{M-1}[n]}^H \bmat{x^\Delta_0[n]\\x^\Delta_1[n]\\\vdots\\x^\Delta_{M-1}[n]}.\label{z}
 \end{align}
@@
 \end{gather}
-where $\1$ is a row vector of ones that represents broadside steering, and $\R$ is the spatial covariance matrix.
+where $\1$ is a row vector of ones that represents broadside steering, and $\R=E\{\x[n]\x[n]\H\}$ is the spatial covariance matrix.
 Since $\R$ is unknown, it must be estimated.
-We will do this by computing the sample covariance matrix, $\eR$, which will require some degree of spatial averaging to avoid signal cancellation, and temporal averaging to achieve \gls{DAS}-like speckle statistics.
+We will do this by computing the sample covariance matrix, $\eR$.
+In this computation we will perform some degree of spatial averaging to avoid signal cancellation, and temporal averaging to achieve \gls{DAS}-like speckle statistics.
 
-\emph{Spatial averaging} is achieved by segmenting the array into $M-L-1$ subarrays of length $L$, where all but one element overlap. If we let $x_l[n]$ represent the datavector from subarray $l$,
+\emph{Spatial averaging} is achieved by grouping array's channels into $M-L-1$ subarrays of length $L$, where all but one element overlap.
+If we let $x_l[n]$ represent the datavector from subarray $l$,
 \begin{gather}
-\x_l[n] = \bmat{x_l[n-\Delta_l] & x_{l+1}[n-\Delta_l] & \dots & x_{l+L-1}[n-\Delta_{l+L-1}]}\T,
+\x_l[n] = \bmat{x^\Delta_l[n] & x^\Delta_{l+1}[n] & \dots & x^\Delta_{l+L-1}[n]}\T,
 \end{gather}
@@
 \end{gather}
-Finally, to ensure a well conditioned sample covariance matrix, we add a fraction $d$ of the total power of $\eR_\text{t,s}[n]$ to its diagonal\cite{Synnevag2007},
+Finally, to ensure a well conditioned sample covariance matrix, we add a fraction $d$ of the total power of $\eR_\text{t,s}[n]$ to its diagonal\cite{Synnevag2007}:
 \begin{align}

### 2012-05-07 ECUA2012/gpu_capon_ecua2012.tex
la2pdf commit - 2012:05:07	22:13:08	(Mon 7.May)

@@
 Let the sample recorded by the $m$'th channel at time instant $n$ be represented as $x_m[n]$, and the accompanying weights be  $w_m[n]$ and $\Delta_m[n]$.
-To simplify the upcoming notation, let sample $x_m[n]$ delayed by a factor $\Delta_m$ be written as
+To simplify the upcoming notation, let sample $x_m[n]$ delayed by a factor $\Delta_m$ be written as:
 \begin{align}
@@
 The \gls{MVDR} beamformer \cite{Capon1969} finds the set of weights that minimizes the output power, $\min E\{|z[n]|^2\}$, while ensuring unity gain in the look direction.
-The solution to this optimization problem is given as
+The solution to this optimization problem is given as:
 \begin{gather}
@@
 We will do this by computing the sample covariance matrix, $\eR$.
-In this computation we will perform some degree of spatial averaging to avoid signal cancellation, and temporal averaging to achieve \gls{DAS}-like speckle statistics.
+In this computation we will perform some degree of \emph{spatial averaging} to avoid signal cancellation, \emph{temporal averaging} to achieve \gls{DAS}-like speckle statistics, and \emph{diagonal loading} to ensure a numerically well conditioned $\eR$~\cite{Synnevag2009a}.
 
@@
 \end{gather}
-then the a ``subarray averaged'' sample covariance matrix, $\eR_\text{s}$, is given as
+then the a ``subarray averaged'' sample covariance matrix, $\eR_\text{s}$, is given as:
 \begin{gather}
@@
 \begin{align}
-\eR[n] = \eR_\text{t,s}[n] + \I \cdot \frac{d}{L} \cdot \tr\{\eR_\text{t,s}[n]\}.\label{finalR}
+\eR[n] = \eR_\text{t,s}[n] + \I \cdot \frac{d}{L} \cdot \tr\{\eR_\text{t,s}[n]\},\label{finalR}
 \end{align}
-
-% To ensure robustness in an active sonar setting, we perform subarray averaging in order to decorrelate noise from the signal, and some degree of temporal averaging to retain \gls{DAS}-like speckle statistics.
-The $(i,j)$'th element of the sample covariance matrix, $\hat r_{ij}$, is then given as
-% \begin{gather}
-% r_{ij}[n] = \frac{1}{(M-L+1)(2N_{\text{avg}}+1)}\sumb{k=0}{M-L} \sumb{n'=n-N_{\text{avg}}}{n+N_{\text{avg}}} x_{i+k}[n']\,x_{j+k}[n'],
-% \end{gather}
-% where $r_{ij}$ is the $(i,j)$'th element of the covariance matrix, $L$ is the subarray size, and $2N_{\text{avg}}+1$ is the number of temporal samples to perform averaging over.
+where $\I$ is an identity matrix and $\tr\{\cdot\}$ means matrix trace operation.
 
@@
 
-When processing images it might be well worth having a look at the \gls{GPU}, as this is a task it was made for.
+Because each pixel in the image can be processed independently, the \gls{MVDR} beamformer is well suited for acceleration by parallel hardware \cite{Nilsen2009}.
+Of particular interest these days is perhaps the \gls{GPU}, which is tailored for image processing, are available off-the-shelf at reasonable prices, and are found in most desktop computers already.
 A modern \gls{GPU} is a hardware architecture comprised of several hundred computing cores, each running a thread that executes a copy of a common function called a kernel.
@@
 
-The \gls{MVDR} beamformer is an algorithm well suited for implementation on a \gls{GPU}.
-Each pixel in the image can be processed independently, and the steps within each pixel will also benefit from parallelisation.
-We implemented the \gls{MVDR} bench Compared to an optimized C and Matlab implementation, our \gls{GPU}
-
 In a sonar processing chain, using one or more \glspl{GPU} can help offload the CPU.

### 2012-05-07 ECUA2012/gpu_capon_ecua2012.tex
la2pdf commit - 2012:05:07	22:49:27	(Mon 7.May)

@@
 %%%%%%%%%%                                      ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
+
+\newcommand\Eq[1]{Equation (\ref{#1})}
+\newcommand\Fig[1]{Figure \ref{#1}}
 
@@
 
-Because each pixel in the image can be processed independently, the \gls{MVDR} beamformer is well suited for acceleration by parallel hardware \cite{Nilsen2009}.
-Of particular interest these days is perhaps the \gls{GPU}, which is tailored for image processing, are available off-the-shelf at reasonable prices, and are found in most desktop computers already.
+Because each pixel in the image can be processed independently, the \gls{MVDR} beamformer is well suited for parallel acceleration.
+Of particular interest is the \gls{GPU}~\cite{Nilsen2009}, which is tailored for image processing, are available off-the-shelf at reasonable prices, and are found in most desktop computers already.
 A modern \gls{GPU} is a hardware architecture comprised of several hundred computing cores, each running a thread that executes a copy of a common function called a kernel.
@@
 
-This suggests that the algorithm should be decomposed into as many lightweight threads as possible, which are both computationally intensive and light on the memory consumption.
-Beamforming fits well into this scheme, as each pixel can be processed idependently.
+This suggests that the algorithm should be decomposed into as many lightweight threads as possible, which are both computationally intensive and light on the memory access and consumption.
 However, for a typical array configuration, using a single thread per pixel is not lightweight enough.
-Therefore, we seek to devise a way to decompose the methods presented in section \ref{methods}.
+Therefore, we the methods presented section \ref{methods} had to be decomposed further.
 In particular, the solutions we came up with for the different steps in the algorithm, listed in their natural order of execution, was:
 \begin{enumerate}
-\item \emph{Computing the spatial covariance matrix} $\eR[n]$ (equation \ref{spatialR}-\ref{finalR}).
+\item \emph{Computing the spatial covariance matrix} $\eR[n]$ (\ref{spatialR}-\ref{finalR}).
 A group of threads are created that slide along the diagonals of $\eR$.
-In this way we exploit the fact that entries on the diagonals overlap across subarrays and time, and keep the numbers of both data reads and writes at a minimum.
+In this way we exploited the fact that entries on the diagonals overlap across subarrays and time, and keep the numbers of both data reads and writes at a minimum.
 %Finished threads can also wrap around and start on diagonals on the lower half triangular. 
 % In this way we process one row of $\eR$ per kernel iteration, and manage to keep the numbers of both data reads and writes at a minimum.
-\item \emph{Computing} $\eRi[n]\1$ (from equation \ref{weights}).
+\item \emph{Computing} $\eRi[n]\1$ (from (\ref{weights})).
 While intuition may suggest that this step is carried out by inverting $\eR$, it is better to solve the linear equation $\R\beta = \1$ for $\beta$ instead, which gives us $\beta = \eRi\1$ directly.
@@
 Inverting $\eR$ is by far the most computationally intensive task, and a key area of focus for further improvements.
-\item \emph{Computing the beamformer output} $z[n]$ by substituting $\beta = \eRi[n]\1$ into equation \ref{weights}, $\vec w[n] = \frac{\beta}{\1\T\beta}$.
-A group of L threads per pixel was used to reduce shared memory pressure and to obtain coalesced reads and a minimum of writes.
-\emph{Finally we acquire the pixel value $z[n]$} by applying the weights to each data vector $\x_l[n]$, and sum them all up (as in equation \ref{z}).
+\item \emph{Computing the beamformer output} $z[n]$ by substituting $\beta = \eRi[n]\1$ into (\ref{weights}), and (\ref{weights}) into (\ref{z}).
+A group of $L$ threads per pixel was used to reduce local memory pressure and to obtain coalesced reads and a minimum of writes.
 When applying weights, there are two approaches from an implementation point of view.
-The subarray data can, as here, be reduced to coincide in length with the subarray weights, or the weights can be extended to M in size and applied directly on $\x[n]$ and summed.
+The subarray data can, as here, be reduced to coincide in length with the subarray weights, or the weights can be extended to $M$ in size and applied directly on $\x[n]$ and summed.
 \end{enumerate}
-For the GPU, the largest bottleneck is the transfer of memory from CPU-side (Host) to the GPU-side (Device).
-In the upcoming results, Host to Device transfer are not taken into account.
-The reasoning behind this is the fact that \gls{MVDR} beamforming requires already delayed data.
-For efficiency, this should also be carried out on the GPU, leaving the challenging task of overcoming the limited Host-Device memory bandwidth to those who implement the delay part of delay-and-sum.
+
+In the upcoming results, we have beamformed pre-delayed data.
+This is a large dataset, and the latency experienced when copying it from the \gls{CPU} side to the \gls{GPU} was significant.
+However, in a practical system only the channel data should be transferred as the \gls{GPU} can effortlessly also perform simple delaying of data.
+The remaining latency can be easily hidden as long as the \gls{GPU} is kept busy, hence the data transfer times will not be reported in the upcoming results.
 
@@
 We have tested our \gls{GPU} implementation of the \gls{MVDR} beamformer on experimental data from the 32 element Kongsberg Maritime HISAS1030 synthetic aperture sonar~\cite{Hansen2009}.
-The sonar was attached to the HUGIN \gls{AUV} and used in strip-map mode, and the imaged object was the 1500 dwt oil tanker wreck Holmengraa (Figure \ref{holmengraa}).
+The sonar was attached to the HUGIN \gls{AUV} and used in strip-map mode, and the imaged object was the 1500 dwt oil tanker wreck Holmengraa (\Fig{holmengraa}).
 Holmengraa is roughly 68\,m long and 9\,m wide, and lies on a slanted seabed at 77\,m depth.
@@
 
-To test the runtime speed of our \gls{GPU} implementation of \gls{MVDR}, we have compared it to an optimized single thread C and Matlab implementation (Figure \ref{benchmarks}).
+To test the runtime speed of our \gls{GPU} implementation of \gls{MVDR}, we have compared it to an optimized single thread C and Matlab implementation (\Fig{benchmarks}).
 For this test we used a 55\,kpixels sectorscan image.

### 2012-05-07 ECUA2012/gpu_capon_ecua2012.tex
la2pdf commit - 2012:05:07	23:20:50	(Mon 7.May)

@@
 \end{align}
-The \gls{MVDR} beamformer \cite{Capon1969} finds the set of weights that minimizes the output power, $\min E\{|z[n]|^2\}$, while ensuring unity gain in the look direction.
+If the weights in (\ref{z}) were static, this would be referred to as the \gls{DAS} beamformer.
+The \gls{MVDR} beamformer~\cite{Capon1969}, however, finds the set of weights that minimizes the output power, $\min E\{|z[n]|^2\}$, while ensuring unity gain in the look direction.
 The solution to this optimization problem is given as:
@@
 This is a large dataset, and the latency experienced when copying it from the \gls{CPU} side to the \gls{GPU} was significant.
-However, in a practical system only the channel data should be transferred as the \gls{GPU} can effortlessly also perform simple delaying of data.
-The remaining latency can be easily hidden as long as the \gls{GPU} is kept busy, hence the data transfer times will not be reported in the upcoming results.
+However, in a practical system only the channel data should be transferred as the delaying of data can effortlessly be performed by the \gls{GPU}.
+The remaining latency can be hidden as long as the \gls{GPU} is kept busy, hence the data transfer times will not be reported in the upcoming results.
+
+In this paper we have targeted an nVidia \gls{GPU} using the C interface to nVidia's \gls{CUDA} framework~\cite{NvidiaCuda}.
 
@@
 
-TODO:
-\begin{itemize}
-\item Choice of robustification parameters $L,M,\Delta$ largely impacts what the ``optimal'' way of implementing \gls{MVDR} 
-\item Introduce CUDA, or ditch it?
-\item Doesn't sell well that we used nVidias solver, especially when that's the bottleneck in the design...
-\end{itemize}
-
 \newpage
@@
 
-We have tested our \gls{GPU} implementation of the \gls{MVDR} beamformer on experimental data from the 32 element Kongsberg Maritime HISAS1030 synthetic aperture sonar~\cite{Hansen2009}.
-The sonar was attached to the HUGIN \gls{AUV} and used in strip-map mode, and the imaged object was the 1500 dwt oil tanker wreck Holmengraa (\Fig{holmengraa}).
+We have tested our \gls{GPU} implementation of the \gls{MVDR} beamformer on experimental data from the 32 element Kongsberg Maritime HISAS1030~\cite{Hansen2009}.
+The sonar was attached to the HUGIN \gls{AUV} and used in sidescan mode, and the imaged object was the 1500 dwt oil tanker wreck Holmengraa (\Fig{holmengraa}).
 Holmengraa is roughly 68\,m long and 9\,m wide, and lies on a slanted seabed at 77\,m depth.
@@
 \includegraphics[width=\linewidth]{gfx/benchmark.pdf}
-\caption{\gls{MVDR} benchmarks. \protect\newline A 55\,kpx image processed with $M=32$ channels and $d=2\%$ diagonal loading.}\label{benchmarks}
+\caption{\gls{MVDR} benchmarks. Forming a 55\,kpixels image from a $M=32$ channel array.}\label{benchmarks}
 \end{figure}
@@
 This is a \gls{GPU} with 448 \gls{CUDA} cores and 6Gb of onboard \gls{RAM}, capable of roughly 1Tflops.
-We kept the channel count at $M=32$ and added $d=2\%$ diagonal loading.
 

### 2012-05-07 ECUA2012/gpu_capon_ecua2012.tex
la2pdf commit - 2012:05:07	23:56:40	(Mon 7.May)

@@
 Because each pixel in the image can be processed independently, the \gls{MVDR} beamformer is well suited for parallel acceleration.
-Of particular interest is the \gls{GPU}~\cite{Nilsen2009}, which is tailored for image processing, are available off-the-shelf at reasonable prices, and are found in most desktop computers already.
-A modern \gls{GPU} is a hardware architecture comprised of several hundred computing cores, each running a thread that executes a copy of a common function called a kernel.
+A technology of particular interest here are \glspl{GPU}~\cite{Nilsen2009}, which tailored for image processing, available off-the-shelf at reasonable prices, and are found in most desktop computers already.
+The \gls{GPU} is a hardware architecture comprised of several hundred computing cores, each running a thread that executes a copy of a common function called a kernel.
 However, there are some inherent drawbacks with such a design: Each core is left with a rather limited amount of local memory, and distributing data to the computing cores and collecting the result is a processes with some inherent latency.
@@
 To test the runtime speed of our \gls{GPU} implementation of \gls{MVDR}, we have compared it to an optimized single thread C and Matlab implementation (\Fig{benchmarks}).
-For this test we used a 55\,kpixels sectorscan image.
+Processing times for a 55\,kpixels sectorscan test image are shown for varying subarray lengths $L$ and varying amounts of temporal averaging $N_\text{avg}$.
 The test system was a quad-core Intel Xeon E5620, with 64Gb of \gls{RAM}, and an nVidia Quadro 6000 card.
@@
 If we look at the total runtimes for a subarray size of $L=16$ and $2N_\text{avg}+1=5$ pixels of temporal averaging, the \gls{GPU} implementation is able to process the test image in roughly 40\,ms.
+This translates to a processing capacity of 1.1\,Mpixels/s.
 For the same scenario the C implementation clocked in at roughly 3\,s and Matlab at 17\,s, which corresponds to a speedup factor of 75 and 435, respectively.
 
-Interesting to note is also that solving $\eRi\1$ remain the main bottleneck in the \gls{GPU} implementation, and for larger $L$'s it dominates.
-However, this is a general Gauss Jordan solver which does not exploit symmetry properties of $\eR$, so creating e.g.
-a batch based Cholesky or UHDU solver should cut the runtimes quite a bit further.
-Both the C and Matlab implementation in this benchmark take this symmetry into account, which explains why the complexity curve of these implementation is roughly proportional with $L$, while the \gls{GPU} implementation is not.
+Interesting to note is also that solving $\eRi\1$ remains the main bottleneck in the \gls{GPU} implementation, and it gets even worse for larger $L$'s.
+However, this is a general Gauss Jordan solver which does not exploit symmetry properties of $\eR$, so creating e.g.\ a batch based Cholesky or UHDU solver should improve the runtimes further.
+Both the C and Matlab implementation in this benchmark take this symmetry into account, which partially explains why the complexity curve of these implementation have a different curve than the \gls{GPU} implementation.
 
-HUGIN: 1.8m/s - range 250m \\
-fs = 100kHz*30\%*$\frac{4}{3}$ = 40kHz \\
-$N_{range px} = \frac{2\,250m}{1500m/s}40kHz = 13.3kpx$ \\
-$\times 32$ beams = 426kpx \\\\
-
-PRF = $\frac{1500m/s}{2 250m} = 3Hz$ \\
-TP = 215px 3Hz = 1.28Mpx/s \\
-4 arrays: 1.28Mpx/s * 4 = 5.12Mpx/s
-
-\begin{itemize}
-\item Need for speed: HUGIN 4 banks of 32 elements, can be processed faster than the ping repetition rate, with margins to spare.
-\end{itemize}
+% HUGIN: 1.8m/s - range 250m \\
+% fs = 100kHz*30\%*$\frac{4}{3}$ = 40kHz \\
+% $N_{range px} = \frac{2\,250m}{1500m/s}40kHz = 13.3kpx$ \\
+% $\times 32$ beams = 426kpx \\\\
+% 
+% PRF = $\frac{1500m/s}{2 250m} = 3Hz$ \\
+% TP = 215px 3Hz = 1.28Mpx/s \\
+% 4 arrays: 1.28Mpx/s * 4 = 5.12Mpx/s
+% 
+% 
+% \begin{itemize}
+% \item Need for speed: HUGIN 4 banks of 32 elements, can be processed faster than the ping repetition rate, with margins to spare.
+% \end{itemize}
 
 \section{Conclusion}
+
+The \gls{MVDR} beamformer is an algorithm well suited for implementation on a \gls{GPU}. Each pixel in the image can be processed independently, and the steps within each pixel can also
 

### 2012-05-07 ECUA2012/gpu_capon_ecua2012.tex
la2pdf commit - 2012:05:08	00:38:47	(Tue 8.May)

@@
 There may be several reasons for this.
-For one, the method is not inherently robust, and will suffer from a phenomena called signal cancellation in active systems.
-Another reason is that in its original form, the computational complexity is cubic with the number of channels, O(M$^3$), while conventional beamformers are usually only at O($M$).
+For one, the method is not inherently robust, and will suffer from a phenomenon called signal cancellation in active systems.
+Another reason is that in its original form, the computational complexity is cubic with the number of channels, O(M$^3$), while conventional beamformers are at O($M$).
 
 We will show that the suggested means for making the \gls{MVDR} behave well in active systems will also make it less computationally complex.
-Furthermore, by implementing it on a modern \gls{GPU} we have gained a speedup of around 2 orders of magnitude compared to an optimized single thread implementation in C.
+Furthermore, by implementing it on a modern \gls{GPU} we have gained a speedup of around 2 orders of magnitude compared to an optimized single-thread implementation in C.
 
@@
 If the data is digitally stored, we can do this by focusing on one pixel at a time in the post-processing.
-Since the geometric relationship of the array and image is known, one usually start by selecting delays that mitigates the difference in transmitter-pixel-receiver propagation time.
+Since the geometric relationship of the array and image is known, one usually start by selecting delays that compensates for the difference in transmitter-pixel-receiver propagation time.
 
@@
 Let the sample recorded by the $m$'th channel at time instant $n$ be represented as $x_m[n]$, and the accompanying weights be  $w_m[n]$ and $\Delta_m[n]$.
-To simplify the upcoming notation, let sample $x_m[n]$ delayed by a factor $\Delta_m$ be written as:
+To simplify the upcoming notation, let sample $x_m[n]$ delayed by $\Delta_m$ samples be written as:
 \begin{align}
@@
 
-\emph{Spatial averaging} is achieved by grouping array's channels into $M-L-1$ subarrays of length $L$, where all but one element overlap.
+\emph{Spatial averaging} is achieved by grouping array's channels into $M-L+1$ subarrays of length $L$, where all but one element overlap.
 If we let $x_l[n]$ represent the datavector from subarray $l$,
@@
 \begin{gather}
-\eR_\text{s}[n] = \frac{1}{M-L+1} \sumb{l=0}{M-L+1} \x_l[n]\x_l\H[n].\label{spatialR}
+\eR_\text{s}[n] = \frac{1}{M-L+1} \sumb{l=0}{M-L} \x_l[n]\x_l\H[n].\label{spatialR}
 \end{gather}
@@
 Because each pixel in the image can be processed independently, the \gls{MVDR} beamformer is well suited for parallel acceleration.
-A technology of particular interest here are \glspl{GPU}~\cite{Nilsen2009}, which tailored for image processing, available off-the-shelf at reasonable prices, and are found in most desktop computers already.
+A technology of particular interest here are \glspl{GPU}~\cite{Nilsen2009}, which are tailored for image processing, available off-the-shelf at reasonable prices, and are found in most desktop computers already.
 The \gls{GPU} is a hardware architecture comprised of several hundred computing cores, each running a thread that executes a copy of a common function called a kernel.
-However, there are some inherent drawbacks with such a design: Each core is left with a rather limited amount of local memory, and distributing data to the computing cores and collecting the result is a processes with some inherent latency.
+However, there are some inherent drawbacks with such a design: Each core is left with a rather limited amount of local memory, and distributing data to the computing cores and collecting the result is a process with some inherent latency.
 
@@
 
-To test the runtime speed of our \gls{GPU} implementation of \gls{MVDR}, we have compared it to an optimized single thread C and Matlab implementation (\Fig{benchmarks}).
+To test the runtime speed of our \gls{GPU} implementation of \gls{MVDR}, we have compared it to an optimized single-thread C and Matlab implementation (\Fig{benchmarks}).
 Processing times for a 55\,kpixels sectorscan test image are shown for varying subarray lengths $L$ and varying amounts of temporal averaging $N_\text{avg}$.
@@
 If we look at the total runtimes for a subarray size of $L=16$ and $2N_\text{avg}+1=5$ pixels of temporal averaging, the \gls{GPU} implementation is able to process the test image in roughly 40\,ms.
-This translates to a processing capacity of 1.1\,Mpixels/s.
+This translates to a processing throughput of 1.1\,Mpixels/s.
 For the same scenario the C implementation clocked in at roughly 3\,s and Matlab at 17\,s, which corresponds to a speedup factor of 75 and 435, respectively.
@@
 However, this is a general Gauss Jordan solver which does not exploit symmetry properties of $\eR$, so creating e.g.\ a batch based Cholesky or UHDU solver should improve the runtimes further.
-Both the C and Matlab implementation in this benchmark take this symmetry into account, which partially explains why the complexity curve of these implementation have a different curve than the \gls{GPU} implementation.
+Both the C and Matlab implementation in this benchmark take this symmetry into account, which partially explains why the complexity curve of these implementations have a different complexity curve than the \gls{GPU} implementation.
 
@@
 
-The \gls{MVDR} beamformer is an algorithm well suited for implementation on a \gls{GPU}. Each pixel in the image can be processed independently, and the steps within each pixel can also
+The \gls{MVDR} beamformer is an algorithm well suited for implementation on a \gls{GPU}.
+Pixels in the image can be processed independently, and this also partially applies to the steps within each pixel.
 

### 2012-05-07 ECUA2012/gpu_capon_ecua2012.tex
la2pdf commit - 2012:05:08	01:10:52	(Tue 8.May)

@@
 Let the sample recorded by the $m$'th channel at time instant $n$ be represented as $x_m[n]$, and the accompanying weights be  $w_m[n]$ and $\Delta_m[n]$.
-To simplify the upcoming notation, let sample $x_m[n]$ delayed by $\Delta_m$ samples be written as:
+To simplify the upcoming notation, let sample $x_m[n]$ delayed by $\Delta_m[n]$ be written as:
 \begin{align}
@@
 
-\emph{Spatial averaging} is achieved by grouping array's channels into $M-L+1$ subarrays of length $L$, where all but one element overlap.
+\emph{Spatial averaging} is achieved by grouping the array channels into $M-L+1$ subarrays of length $L$, where all but one element overlap.
 If we let $x_l[n]$ represent the datavector from subarray $l$,
@@
 \end{gather}
-then the a ``subarray averaged'' sample covariance matrix, $\eR_\text{s}$, is given as:
+then the ``subarray averaged'' sample covariance matrix, $\eR_\text{s}$, is given as:
 \begin{gather}
@@
 However, for a typical array configuration, using a single thread per pixel is not lightweight enough.
-Therefore, we the methods presented section \ref{methods} had to be decomposed further.
+Therefore, the methods presented section \ref{methods} had to be decomposed further.
 In particular, the solutions we came up with for the different steps in the algorithm, listed in their natural order of execution, was:
@@
 
-We have tested our \gls{GPU} implementation of the \gls{MVDR} beamformer on experimental data from the 32 element Kongsberg Maritime HISAS1030~\cite{Hansen2009}.
-The sonar was attached to the HUGIN \gls{AUV} and used in sidescan mode, and the imaged object was the 1500 dwt oil tanker wreck Holmengraa (\Fig{holmengraa}).
+We have tested our \gls{GPU} implementation of the \gls{MVDR} beamformer on two experimental dataset from the 32 element Kongsberg Maritime HISAS1030 sonar~\cite{Hansen2009}.
+The sonar was attached to the HUGIN \gls{AUV}, and used in sidescan mode to image the 1500 dwt oil tanker wreck Holmengraa (\Fig{holmengraa}).
 Holmengraa is roughly 68\,m long and 9\,m wide, and lies on a slanted seabed at 77\,m depth.
@@
 
-To test the runtime speed of our \gls{GPU} implementation of \gls{MVDR}, we have compared it to an optimized single-thread C and Matlab implementation (\Fig{benchmarks}).
-Processing times for a 55\,kpixels sectorscan test image are shown for varying subarray lengths $L$ and varying amounts of temporal averaging $N_\text{avg}$.
+The other dataset was a 55\,kpixels sectorscan test image.
+We used it to compare the runtimes of our \gls{GPU} implementation of \gls{MVDR}, and an optimized single-thread C and Matlab implementation (\Fig{benchmarks}).
+Processing times for a are shown for varying subarray lengths $L$ and varying amounts of temporal averaging $N_\text{avg}$.
 The test system was a quad-core Intel Xeon E5620, with 64Gb of \gls{RAM}, and an nVidia Quadro 6000 card.
-This is a \gls{GPU} with 448 \gls{CUDA} cores and 6Gb of onboard \gls{RAM}, capable of roughly 1Tflops.
+This is a \gls{GPU} with 448 \gls{CUDA} cores and 6Gb of onboard \gls{RAM}, capable of performing roughly 1Tflops.
 
@@
 This translates to a processing throughput of 1.1\,Mpixels/s.
-For the same scenario the C implementation clocked in at roughly 3\,s and Matlab at 17\,s, which corresponds to a speedup factor of 75 and 435, respectively.
+For the same scenario the C implementation clocked in at roughly 3\,s and Matlab at 17\,s, which is 75 and 435 times more than the \gls{GPU} implementation, respectively.
 
 Interesting to note is also that solving $\eRi\1$ remains the main bottleneck in the \gls{GPU} implementation, and it gets even worse for larger $L$'s.
-However, this is a general Gauss Jordan solver which does not exploit symmetry properties of $\eR$, so creating e.g.\ a batch based Cholesky or UHDU solver should improve the runtimes further.
-Both the C and Matlab implementation in this benchmark take this symmetry into account, which partially explains why the complexity curve of these implementations have a different complexity curve than the \gls{GPU} implementation.
+However, this is a general Gauss Jordan solver which does not exploit symmetry properties of $\eR$, so creating e.g.\ a batch based Cholesky solver should improve the runtimes further.
+Both the C and Matlab implementation in this benchmark take this symmetry into account, which partially explains why the complexity curves of these implementations are different from that of the \gls{GPU} implementation.
 

### 2012-05-07 ECUA2012/gpu_capon_ecua2012.tex
la2pdf commit - 2012:05:08	01:45:47	(Tue 8.May)

@@
 
-We have tested our \gls{GPU} implementation of the \gls{MVDR} beamformer on two experimental dataset from the 32 element Kongsberg Maritime HISAS1030 sonar~\cite{Hansen2009}.
+We have tested our \gls{GPU} implementation of the \gls{MVDR} beamformer on two experimental datasets from the 32 element Kongsberg Maritime HISAS1030 sonar~\cite{Hansen2009}.
 The sonar was attached to the HUGIN \gls{AUV}, and used in sidescan mode to image the 1500 dwt oil tanker wreck Holmengraa (\Fig{holmengraa}).
@@
 The other dataset was a 55\,kpixels sectorscan test image.
-We used it to compare the runtimes of our \gls{GPU} implementation of \gls{MVDR}, and an optimized single-thread C and Matlab implementation (\Fig{benchmarks}).
-Processing times for a are shown for varying subarray lengths $L$ and varying amounts of temporal averaging $N_\text{avg}$.
+We used it to benchmark our \gls{MVDR} implementation, and compared the result with the runtimes of an optimized single-thread C and Matlab implementation (\Fig{benchmarks}).
+Processing times are shown for varying subarray lengths $L$ and varying amounts of temporal averaging $N_\text{avg}$.
 The test system was a quad-core Intel Xeon E5620, with 64Gb of \gls{RAM}, and an nVidia Quadro 6000 card.
@@
 Compared the \gls{DAS} beamformer, the ship's edges come out as sharper, and the shadows are less noisy.
-But while \gls{DAS} processed this 1.3\,Mpixel image in milliseconds, our Matlab implementation of \gls{MVDR} needed minutes.
+But while \gls{DAS} processed this 1.3\,Mpixel image in milliseconds, our Matlab implementation of \gls{MVDR} beamformer needed minutes.
 
@@
 This translates to a processing throughput of 1.1\,Mpixels/s.
-For the same scenario the C implementation clocked in at roughly 3\,s and Matlab at 17\,s, which is 75 and 435 times more than the \gls{GPU} implementation, respectively.
+For the same scenario the C implementation clocked in at roughly 3\,s and Matlab at 17\,s, which is 75 and 435 times slower than the \gls{GPU} implementation, respectively.
 
@@
 
-The \gls{MVDR} beamformer is an algorithm well suited for implementation on a \gls{GPU}.
-Pixels in the image can be processed independently, and this also partially applies to the steps within each pixel.
+The \gls{MVDR} beamformer is an algorithm well suited for implementation on a \gls{GPU}. Each pixels in the image can be processed independently, and the various steps within each pixel .
 

### 2012-05-08 ECUA2012/gpu_capon_ecua2012.tex
la2pdf commit - 2012:05:08	02:22:49	(Tue 8.May)

@@
 
-The \gls{MVDR} beamformer is an algorithm well suited for implementation on a \gls{GPU}. Each pixels in the image can be processed independently, and the various steps within each pixel .
-
-In a sonar processing chain, using one or more \glspl{GPU} can help offload the CPU.
-
-We achieved speedups Our \gls{GPU} implementation 
-Using \glspl{GPU} to accelerate the is a
+The \gls{MVDR} beamformer is an algorithm well suited for implementation on a \gls{GPU}.
+This is because the computations are independent on a pixel level, and partially also within each pixel.
+We were able to achieve speedup factors of 2-3 orders of magnitude by implementing our active sonar capable \gls{MVDR} implementation on a high-end \gls{GPU}.
+In sonar processing chains where the \gls{CPU} strain is considerable, simply being able to offloading some computationally intensive tasks to an idling \gls{GPU} might be just what is needed.
 

### 2012-05-08 ECUA2012/gpu_capon_ecua2012.tex
la2pdf commit - 2012:05:08	03:32:59	(Tue 8.May)

@@
 
+\section*{Abstract}
+
+Two limiting factors of modern phased array sonar imaging systems are the angular resolution and contrast.
+Adaptive beamformers target these limitatitons by weighting sensors dynamically to fit the impinging wavefield.
+
+The Minimum Variance (MV) adaptive beamformer has recently been introduced in active sonar imaging.
+While displaying great potential for creating high quality images, it is computationally intensive and relies on robustification.
+These are important reasons why the MV beamformer is not in widespread use today.
+
+We propose means for adapting the MV beamformer and its robustification techniques to fit a modern Graphics Processing Unit (GPU) architecture. These devices have in recent years
+emerged as a flexible and highly potent parallel technology well suited for many DSP tasks, and are available off-the-shelf at reasonable prices.
+
+We benchmarked optimized implementations of the MV beamformer in Matlab and C, running on a quad-core Intel Xeon E5620, and in CUDA running on a mid-range nVidia card.
+The beamformers operated on data from the 32x4 element Kongsberg Martime HISAS1030 sonar, using a conservative set of robustification parameters.
+We found that the GPU rendered more than 1Mpixels/s, representing a speedup of more than two orders of magnitude over C, and three orders of magnitude over Matlab.
+If the HISAS1030 was attached to a platform traveling at 1.8m/s, creating full-coverage sectorscan images in realtime would be possible with this performance.
+The CPU would then be free to take on the rest of the beamforming as well.
+
+\newpage
 \maketitle
@@
 If we look at the total runtimes for a subarray size of $L=16$ and $2N_\text{avg}+1=5$ pixels of temporal averaging, the \gls{GPU} implementation is able to process the test image in roughly 40\,ms.
-This translates to a processing throughput of 1.1\,Mpixels/s.
+This translates to a processing throughput of 1.4\,Mpixels/s.
 For the same scenario the C implementation clocked in at roughly 3\,s and Matlab at 17\,s, which is 75 and 435 times slower than the \gls{GPU} implementation, respectively.
@@
 Both the C and Matlab implementation in this benchmark take this symmetry into account, which partially explains why the complexity curves of these implementations are different from that of the \gls{GPU} implementation.
+
+If the HISAS1030 was attached to a platform moving at 1.8m/s, for which the maximum range is 250\,m~\cite{Hansen2010}, the pulse repetition frequency could be set to 3\,Hz.
+Then the throughput required for creating critically sampled 32-beam sectorscan images in realtime would be roughly 1\,Mpixels/s.
+This is something our \gls{GPU} implementation could handle alone, leaving the \gls{CPU} free to take on other assignments.
 
@@
 We were able to achieve speedup factors of 2-3 orders of magnitude by implementing our active sonar capable \gls{MVDR} implementation on a high-end \gls{GPU}.
-In sonar processing chains where the \gls{CPU} strain is considerable, simply being able to offloading some computationally intensive tasks to an idling \gls{GPU} might be just what is needed.
+This performance is sufficient for computing critically sampled full-coverage sectorsscan images from the HISAS1030 sonar in realtime.
+Furthermore, offloading tasks to the \gls{GPU} might be desirable in itself since sonar processing chains are often \gls{CPU} intensive.
 
@@
 
+\newpage
 \bibliographystyle{./template/ecua}

### 2012-05-08 ECUA2012/gpu_capon_ecua2012.tex
la2pdf commit - 2012:05:08	04:15:29	(Tue 8.May)

@@
 
+\newpage
 Interesting to note is also that solving $\eRi\1$ remains the main bottleneck in the \gls{GPU} implementation, and it gets even worse for larger $L$'s.
@@
 If the HISAS1030 was attached to a platform moving at 1.8m/s, for which the maximum range is 250\,m~\cite{Hansen2010}, the pulse repetition frequency could be set to 3\,Hz.
-Then the throughput required for creating critically sampled 32-beam sectorscan images in realtime would be roughly 1\,Mpixels/s.
-This is something our \gls{GPU} implementation could handle alone, leaving the \gls{CPU} free to take on other assignments.
+Furthermore, the HISAS1030's bandwidth is 30\,kHz, so we set this to be the complex sampling frequency, and beamform $M=32$ lateral beams.
+The throughput required for realtime processing of these images would then be 1\,Mpixels/s.
+As we have seen, this is something our \gls{GPU} implementation could handle, leaving the \gls{CPU} free to take on other assignments.
 
@@
 
-The authors would like to thank nVidia for providing us their unofficial linear equation batch solver, and Kongsberg Maritime and the Norwegian Defence Research Establishment (FFI) for providing us with experimental data.
+The authors would like to thank nVidia for providing them their unofficial linear equation batch solver, and Kongsberg Maritime and the Norwegian Defence Research Establishment (FFI) for providing experimental data.
 

### 2012-05-08 ECUA2012/gpu_capon_ecua2012.tex
la2pdf commit - 2012:05:08	04:55:41	(Tue 8.May)

@@
 
-We have tested our \gls{GPU} implementation of the \gls{MVDR} beamformer on two experimental datasets from the 32 element Kongsberg Maritime HISAS1030 sonar~\cite{Hansen2009}.
-The sonar was attached to the HUGIN \gls{AUV}, and used in sidescan mode to image the 1500 dwt oil tanker wreck Holmengraa (\Fig{holmengraa}).
+We have tested our \gls{GPU} implementation of the \gls{MVDR} beamformer on two different experimental datasets from the 32 element Kongsberg Maritime HISAS1030 sonar~\cite{Hansen2009}.
+The sonar was attached to the HUGIN \gls{AUV}, and one one of the datasets used in sidescan mode to image the 1500 dwt oil tanker wreck Holmengraa (\Fig{holmengraa}).
 Holmengraa is roughly 68\,m long and 9\,m wide, and lies on a slanted seabed at 77\,m depth.
@@
 If the HISAS1030 was attached to a platform moving at 1.8m/s, for which the maximum range is 250\,m~\cite{Hansen2010}, the pulse repetition frequency could be set to 3\,Hz.
-Furthermore, the HISAS1030's bandwidth is 30\,kHz, so we set this to be the complex sampling frequency, and beamform $M=32$ lateral beams.
+We further assume a critical complex sampling frequency equal to the HISAS1030's bandwidth of 30\,kHz, and beamform $M=32$ lateral beams.
 The throughput required for realtime processing of these images would then be 1\,Mpixels/s.
-As we have seen, this is something our \gls{GPU} implementation could handle, leaving the \gls{CPU} free to take on other assignments.
+As we have seen, this is something our \gls{GPU} implementation can handle if $L$ is not too large.%, leaving the \gls{CPU} free to take on other assignments.
 
@@
 The \gls{MVDR} beamformer is an algorithm well suited for implementation on a \gls{GPU}.
-This is because the computations are independent on a pixel level, and partially also within each pixel.
+This is because the computations involved are independent on a pixel level, and partially also within each pixel.
 We were able to achieve speedup factors of 2-3 orders of magnitude by implementing our active sonar capable \gls{MVDR} implementation on a high-end \gls{GPU}.
 This performance is sufficient for computing critically sampled full-coverage sectorsscan images from the HISAS1030 sonar in realtime.
-Furthermore, offloading tasks to the \gls{GPU} might be desirable in itself since sonar processing chains are often \gls{CPU} intensive.
+Furthermore, even if such processing speeds are not required, it might be advantageous to relieve the \gls{CPU} from some of its workload.
+After all, these chips are designed to cooperate.
 

### 2012-05-08 ECUA2012/gpu_capon_ecua2012.tex
ECUA contribution almost ready.

@@
 \includegraphics[width=\linewidth]{gfx/benchmark.pdf}
-\caption{\gls{MVDR} benchmarks. Forming a 55\,kpixels image from a $M=32$ channel array.}\label{benchmarks}
+\caption{\gls{MVDR} benchmarks. Forming a 55\,kpixel image from a $M=32$ channel array.}\label{benchmarks}
 \end{figure}
 
-The other dataset was a 55\,kpixels sectorscan test image.
+The other dataset was a 55\,kpixel sectorscan test image.
 We used it to benchmark our \gls{MVDR} implementation, and compared the result with the runtimes of an optimized single-thread C and Matlab implementation (\Fig{benchmarks}).
@@
 Furthermore, even if such processing speeds are not required, it might be advantageous to relieve the \gls{CPU} from some of its workload.
-After all, these chips are designed to cooperate.
+After all, these chips were designed to cooperate.
 

### 2012-05-08 ECUA2012/gpu_capon_ecua2012.tex
la2pdf commit - 2012:05:08	10:14:07	(Tue 8.May)
la2pdf commit - 2012:05:08	10:46:47	(Tue 8.May)

@@
 
-The Minimum Variance (MV) adaptive beamformer has recently been introduced in active sonar imaging.
+The Minimum Variance Distortionless Response (MVDR) adaptive beamformer has recently been introduced in active sonar imaging.
 While displaying great potential for creating high quality images, it is computationally intensive and relies on robustification.
@@
 
-The Minimum Variance Distortionless Response (MVDR, or Capon) beamformer is one such technique.
+The MVDR adaptive beamformer (also known as the Minimum Variance or Capon beamformer) is one such technique.
 While ensuring unit gain in the look direction, it computes the set of weights that minimizes the power accumulated by the array from other directions.
@@
 \end{gather}
-
 \emph{Temporal averaging} can now be introduced by taking the average of $2N_\text{avg}+1$ temporal samples of $\eR_\text{s}[n]$. Let the result be called $\eR_\text{t,s}[n]$,
@@
 We have tested various solvers for this task, both in-house and proprietary implementations, and achieved the best performance by using an unofficial batch solver from nVidia.
-Inverting $\eR$ is by far the most computationally intensive task, and a key area of focus for further improvements.
 \item \emph{Computing the beamformer output} $z[n]$ by substituting $\beta = \eRi[n]\1$ into (\ref{weights}), and (\ref{weights}) into (\ref{z}).
@@
 \newpage
-Interesting to note is also that solving $\eRi\1$ remains the main bottleneck in the \gls{GPU} implementation, and it gets even worse for larger $L$'s.
+Interesting to note is also that solving $\eRi\1$ remains the main bottleneck in the \gls{GPU} implementation, and the complexity depends on the subarray size $L$'s.
+This means the overall complexity of the \gls{MVDR} beamformer is actually reduced when we apply subarray averaging.
 However, this is a general Gauss Jordan solver which does not exploit symmetry properties of $\eR$, so creating e.g.\ a batch based Cholesky solver should improve the runtimes further.

### 2012-05-08 ECUA2012/gpu_capon_ecua2012.tex
la2pdf commit - 2012:05:08	11:22:19	(Tue 8.May)

@@
 While displaying great potential for creating high quality images, it is computationally intensive and relies on robustification.
-These are important reasons why the MV beamformer is not in widespread use today.
+These are important reasons why the MVDR beamformer is not in widespread use today.
 
-We propose means for adapting the MV beamformer and its robustification techniques to fit a modern Graphics Processing Unit (GPU) architecture. These devices have in recent years
+We propose means for adapting the MVDR beamformer and its robustification techniques to fit a modern Graphics Processing Unit (GPU) architecture. These devices have in recent years
 emerged as a flexible and highly potent parallel technology well suited for many DSP tasks, and are available off-the-shelf at reasonable prices.
@@
 \end{gather}
-then the ``subarray averaged'' sample covariance matrix, $\eR_\text{s}$, is given as:
+then a ``subarray averaged'' sample covariance matrix, $\eR_\text{s}$, is given as:
 \begin{gather}
@@
 However, for a typical array configuration, using a single thread per pixel is not lightweight enough.
-Therefore, the methods presented section \ref{methods} had to be decomposed further.
+Therefore, the methods presented in section \ref{methods} had to be decomposed further.
 In particular, the solutions we came up with for the different steps in the algorithm, listed in their natural order of execution, was:
@@
 \item \emph{Computing the spatial covariance matrix} $\eR[n]$ (\ref{spatialR}-\ref{finalR}).
-A group of threads are created that slide along the diagonals of $\eR$.
-In this way we exploited the fact that entries on the diagonals overlap across subarrays and time, and keep the numbers of both data reads and writes at a minimum.
+A group of threads is created that slide along the diagonals of $\eR$.
+In this way we exploit the fact that entries on the diagonals overlap across subarrays and time, and keep the number of data reads and writes at a minimum.
 %Finished threads can also wrap around and start on diagonals on the lower half triangular. 
@@
 
-We have tested our \gls{GPU} implementation of the \gls{MVDR} beamformer on two different experimental datasets from the 32 element Kongsberg Maritime HISAS1030 sonar~\cite{Hansen2009}.
-The sonar was attached to the HUGIN \gls{AUV}, and one one of the datasets used in sidescan mode to image the 1500 dwt oil tanker wreck Holmengraa (\Fig{holmengraa}).
+We have tested our \gls{GPU} implementation of the \gls{MVDR} beamformer on two different experimental datasets from the 32 element Kongsberg Maritime HISAS1030 sonar~\cite{Hansen2009}, which was attached to the HUGIN \gls{AUV}.
+In the first dataset, the sonar was used in sidescan mode to image the 1500 dwt oil tanker wreck Holmengraa (\Fig{holmengraa}).
 Holmengraa is roughly 68\,m long and 9\,m wide, and lies on a slanted seabed at 77\,m depth.
@@
 This means the overall complexity of the \gls{MVDR} beamformer is actually reduced when we apply subarray averaging.
-However, this is a general Gauss Jordan solver which does not exploit symmetry properties of $\eR$, so creating e.g.\ a batch based Cholesky solver should improve the runtimes further.
-Both the C and Matlab implementation in this benchmark take this symmetry into account, which partially explains why the complexity curves of these implementations are different from that of the \gls{GPU} implementation.
 
-If the HISAS1030 was attached to a platform moving at 1.8m/s, for which the maximum range is 250\,m~\cite{Hansen2010}, the pulse repetition frequency could be set to 3\,Hz.
+The nVidia batch solver is a general Gauss Jordan solver which does not exploit symmetry properties of $\eR$.
+Therefore the implementation can be further optimized by implementing e.g.\ a batch based Cholesky solver.
+Both the C and Matlab implementation in this benchmark are Cholesky based, which partially explains why the complexity curves of these implementations differ.
+
+If the HISAS1030 sonar was attached to a platform moving at 1.8\,m/s, for which the maximum range is 250\,m~\cite{Hansen2010}, the pulse repetition frequency could be set to 3\,Hz.
 We further assume a critical complex sampling frequency equal to the HISAS1030's bandwidth of 30\,kHz, and beamform $M=32$ lateral beams.

### 2012-05-08 ECUA2012/gpu_capon_ecua2012.tex
la2pdf commit - 2012:05:08	11:57:00	(Tue 8.May)

@@
 
-The Minimum Variance Distortionless Response (MVDR) adaptive beamformer has recently been introduced in active sonar imaging.
+The Minimum Variance Distortionless Response (MV) adaptive beamformer has recently been introduced in active sonar imaging.
 While displaying great potential for creating high quality images, it is computationally intensive and relies on robustification.
-These are important reasons why the MVDR beamformer is not in widespread use today.
+These are important reasons why the MV beamformer is not in widespread use today.
 
-We propose means for adapting the MVDR beamformer and its robustification techniques to fit a modern Graphics Processing Unit (GPU) architecture. These devices have in recent years
-emerged as a flexible and highly potent parallel technology well suited for many DSP tasks, and are available off-the-shelf at reasonable prices.
+We propose means for adapting the MV beamformer and its robustification techniques to fit a modern Graphics Processing Unit (GPU) architecture.
+These devices have in recent years emerged as a flexible and highly potent parallel technology well suited for many DSP tasks, and are available off-the-shelf at reasonable prices.
 
@@
 If the HISAS1030 was attached to a platform traveling at 1.8m/s, creating full-coverage sectorscan images in realtime would be possible with this performance.
-The CPU would then be free to take on the rest of the beamforming as well.
+This would also free the CPU to take on other parts of beamforming as well.
 
@@
 % 
-% The Minimum Variance Distortionless Response (MVDR) (or \gls{MVDR}) beamformer, have the potential to overcome this limitation.
+% The Minimum Variance Distortionless Response (MVDR) (or \gls{MV}) beamformer, have the potential to overcome this limitation.
 While ensuring unit gain in the look direction, it computes the set of weights that minimizes the energy accumulated by the array from other directions.
@@
 
-The MVDR adaptive beamformer (also known as the Minimum Variance or Capon beamformer) is one such technique.
+The MV adaptive beamformer (also known as the Minimum Variance or Capon beamformer) is one such technique.
 While ensuring unit gain in the look direction, it computes the set of weights that minimizes the power accumulated by the array from other directions.
@@
 
-However, despite its inherent potential, the \gls{MVDR} beamformer has yet to see widespread adoption in the active sonar community.
+However, despite its inherent potential, the \gls{MV} beamformer has yet to see widespread adoption in the active sonar community.
 There may be several reasons for this.
@@
 
-We will show that the suggested means for making the \gls{MVDR} behave well in active systems will also make it less computationally complex.
+We will show that the suggested means for making the \gls{MV} behave well in active systems will also make it less computationally complex.
 Furthermore, by implementing it on a modern \gls{GPU} we have gained a speedup of around 2 orders of magnitude compared to an optimized single-thread implementation in C.
@@
 If the weights in (\ref{z}) were static, this would be referred to as the \gls{DAS} beamformer.
-The \gls{MVDR} beamformer~\cite{Capon1969}, however, finds the set of weights that minimizes the output power, $\min E\{|z[n]|^2\}$, while ensuring unity gain in the look direction.
+The \gls{MV} beamformer~\cite{Capon1969}, however, finds the set of weights that minimizes the output power, $\min E\{|z[n]|^2\}$, while ensuring unity gain in the look direction.
 The solution to this optimization problem is given as:
@@
 
-Because each pixel in the image can be processed independently, the \gls{MVDR} beamformer is well suited for parallel acceleration.
+Because each pixel in the image can be processed independently, the \gls{MV} beamformer is well suited for parallel acceleration.
 A technology of particular interest here are \glspl{GPU}~\cite{Nilsen2009}, which are tailored for image processing, available off-the-shelf at reasonable prices, and are found in most desktop computers already.
@@
 \item \emph{Computing the spatial covariance matrix} $\eR[n]$ (\ref{spatialR}-\ref{finalR}).
-A group of threads is created that slide along the diagonals of $\eR$.
+We create a group of threads that slide along the diagonals of $\eR$.
 In this way we exploit the fact that entries on the diagonals overlap across subarrays and time, and keep the number of data reads and writes at a minimum.
@@
 \item \emph{Computing the beamformer output} $z[n]$ by substituting $\beta = \eRi[n]\1$ into (\ref{weights}), and (\ref{weights}) into (\ref{z}).
-A group of $L$ threads per pixel was used to reduce local memory pressure and to obtain coalesced reads and a minimum of writes.
+A group of $L$ threads per pixel is used to reduce local memory pressure and to obtain coalesced reads and a minimum of writes.
 When applying weights, there are two approaches from an implementation point of view.
@@
 
-We have tested our \gls{GPU} implementation of the \gls{MVDR} beamformer on two different experimental datasets from the 32 element Kongsberg Maritime HISAS1030 sonar~\cite{Hansen2009}, which was attached to the HUGIN \gls{AUV}.
+We have tested our \gls{GPU} implementation of the \gls{MV} beamformer on two different experimental datasets from the 32 element Kongsberg Maritime HISAS1030 sonar~\cite{Hansen2009}, which was attached to the HUGIN \gls{AUV}.
 In the first dataset, the sonar was used in sidescan mode to image the 1500 dwt oil tanker wreck Holmengraa (\Fig{holmengraa}).
 Holmengraa is roughly 68\,m long and 9\,m wide, and lies on a slanted seabed at 77\,m depth.
-The \gls{MVDR} image was created with the parameters $L=16$, $N_\text{avg}=3$, and $d=1\%$, which proved to be a reasonable selection for this scenario.
+The \gls{MV} image was created with the parameters $L=16$, $N_\text{avg}=3$, and $d=1\%$, which proved to be a reasonable selection for this scenario.
 
@@
 \includegraphics[width=\linewidth]{gfx/benchmark.pdf}
-\caption{\gls{MVDR} benchmarks. Forming a 55\,kpixel image from a $M=32$ channel array.}\label{benchmarks}
+\caption{\gls{MV} benchmarks. Forming a 55\,kpixel image from a $M=32$ channel array.}\label{benchmarks}
 \end{figure}
@@
 The other dataset was a 55\,kpixel sectorscan test image.
-We used it to benchmark our \gls{MVDR} implementation, and compared the result with the runtimes of an optimized single-thread C and Matlab implementation (\Fig{benchmarks}).
+We used it to benchmark our \gls{MV} implementation, and compared the result with the runtimes of an optimized single-thread C and Matlab implementation (\Fig{benchmarks}).
 Processing times are shown for varying subarray lengths $L$ and varying amounts of temporal averaging $N_\text{avg}$.
@@
 
-Figure \ref{holmengraa} demonstrates the \gls{MVDR} beamformer's ability to suppress interference and achieve better detail resolution.
+Figure \ref{holmengraa} demonstrates the \gls{MV} beamformer's ability to suppress interference and achieve better detail resolution.
 Compared the \gls{DAS} beamformer, the ship's edges come out as sharper, and the shadows are less noisy.
-But while \gls{DAS} processed this 1.3\,Mpixel image in milliseconds, our Matlab implementation of \gls{MVDR} beamformer needed minutes.
+But while \gls{DAS} processed this 1.3\,Mpixel image in milliseconds, our Matlab implementation of \gls{MV} beamformer needed minutes.
 
@@
 Interesting to note is also that solving $\eRi\1$ remains the main bottleneck in the \gls{GPU} implementation, and the complexity depends on the subarray size $L$'s.
-This means the overall complexity of the \gls{MVDR} beamformer is actually reduced when we apply subarray averaging.
+This means the overall complexity of the \gls{MV} beamformer is actually reduced when we apply subarray averaging.
 
@@
 We further assume a critical complex sampling frequency equal to the HISAS1030's bandwidth of 30\,kHz, and beamform $M=32$ lateral beams.
-The throughput required for realtime processing of these images would then be 1\,Mpixels/s.
-As we have seen, this is something our \gls{GPU} implementation can handle if $L$ is not too large.%, leaving the \gls{CPU} free to take on other assignments.
+The throughput required for realtime processing of such sectorscan images would then be 1\,Mpixels/s.
+As we have seen, this is something our \gls{GPU} implementation can handle if $L$ is not too large.
+Naturally, this means also means that realtime sidescan imaging can easily be accomplished.
+%, leaving the \gls{CPU} free to take on other assignments.
 
@@
 
-The \gls{MVDR} beamformer is an algorithm well suited for implementation on a \gls{GPU}.
+The \gls{MV} beamformer is an algorithm well suited for implementation on a \gls{GPU}.
 This is because the computations involved are independent on a pixel level, and partially also within each pixel.
-We were able to achieve speedup factors of 2-3 orders of magnitude by implementing our active sonar capable \gls{MVDR} implementation on a high-end \gls{GPU}.
-This performance is sufficient for computing critically sampled full-coverage sectorsscan images from the HISAS1030 sonar in realtime.
+We were able to achieve speedup factors of 2-3 orders of magnitude by implementing our active sonar capable \gls{MV} implementation on a high-end \gls{GPU}.
+This performance is sufficient for computing critically sampled full-coverage sectorsscan and sidescan images from the HISAS1030 sonar in realtime.
 Furthermore, even if such processing speeds are not required, it might be advantageous to relieve the \gls{CPU} from some of its workload.
@@
 
-\newpage
 \bibliographystyle{./template/ecua}

### 2012-05-14 HRUIonGPU/jrnl_article.tex
Added TUFFC article on Capon GPU.

%% bare_jrnl.tex
%% V1.3
%% 2007/01/11
%% by Michael Shell
%% see http://www.michaelshell.org/
%% for current contact information.
%%
%% This is a skeleton file demonstrating the use of IEEEtran.cls
%% (requires IEEEtran.cls version 1.7 or later) with an IEEE journal paper.
%%
%% Support sites:
%% http://www.michaelshell.org/tex/ieeetran/
%% http://www.ctan.org/tex-archive/macros/latex/contrib/IEEEtran/
%% and
%% http://www.ieee.org/

% *** Authors should verify (and, if needed, correct) their LaTeX system  ***
% *** with the testflow diagnostic prior to trusting their LaTeX platform ***
% *** with production work. IEEE's font choices can trigger bugs that do  ***
% *** not appear when using other class files.                            ***
% The testflow support page is at:
% http://www.michaelshell.org/tex/testflow/

%%*************************************************************************
%% Legal Notice:
%% This code is offered as-is without any warranty either expressed or
%% implied; without even the implied warranty of MERCHANTABILITY or
%% FITNESS FOR A PARTICULAR PURPOSE! 
%% User assumes all risk.
%% In no event shall IEEE or any contributor to this code be liable for
%% any damages or losses, including, but not limited to, incidental,
%% consequential, or any other damages, resulting from the use or misuse
%% of any information contained here.
%%
%% All comments are the opinions of their respective authors and are not
%% necessarily endorsed by the IEEE.
%%
%% This work is distributed under the LaTeX Project Public License (LPPL)
%% ( http://www.latex-project.org/ ) version 1.3, and may be freely used,
%% distributed and modified. A copy of the LPPL, version 1.3, is included
%% in the base LaTeX documentation of all distributions of LaTeX released
%% 2003/12/01 or later.
%% Retain all contribution notices and credits.
%% ** Modified files should be clearly indicated as such, including  **
%% ** renaming them and changing author support contact information. **
%%
%% File list of work: IEEEtran.cls, IEEEtran_HOWTO.pdf, bare_adv.tex,
%%                    bare_conf.tex, bare_jrnl.tex, bare_jrnl_compsoc.tex
%%*************************************************************************

% Note that the a4paper option is mainly intended so that authors in
% countries using A4 can easily print to A4 and see how their papers will
% look in print - the typesetting of the document will not typically be
% affected with changes in paper size (but the bottom and side margins will).
% Use the testflow package mentioned above to verify correct handling of
% both paper sizes by the user's LaTeX system.
%
% Also note that the "draftcls" or "draftclsnofoot", not "draft", option
% should be used if it is desired that the figures are to be displayed in
% draft mode.
%
\documentclass[journal]{IEEEtran}
%
% If IEEEtran.cls has not been installed into the LaTeX system files,
% manually specify the path to it like:
% \documentclass[journal]{../sty/IEEEtran}

% Some very useful LaTeX packages include:
% (uncomment the ones you want to load)

% *** MISC UTILITY PACKAGES ***
%
%\usepackage{ifpdf}
% Heiko Oberdiek's ifpdf.sty is very useful if you need conditional
% compilation based on whether the output is pdf or dvi.
% usage:
% \ifpdf
%   % pdf code
% \else
%   % dvi code
% \fi
% The latest version of ifpdf.sty can be obtained from:
% http://www.ctan.org/tex-archive/macros/latex/contrib/oberdiek/
% Also, note that IEEEtran.cls V1.7 and later provides a builtin
% \ifCLASSINFOpdf conditional that works the same way.
% When switching from latex to pdflatex and vice-versa, the compiler may
% have to be run twice to clear warning/error messages.

% *** CITATION PACKAGES ***
%
%\usepackage{cite}
% cite.sty was written by Donald Arseneau
% V1.6 and later of IEEEtran pre-defines the format of the cite.sty package
% \cite{} output to follow that of IEEE. Loading the cite package will
% result in citation numbers being automatically sorted and properly
% "compressed/ranged". e.g., [1], [9], [2], [7], [5], [6] without using
% cite.sty will become [1], [2], [5]--[7], [9] using cite.sty. cite.sty's
% \cite will automatically add leading space, if needed. Use cite.sty's
% noadjust option (cite.sty V3.8 and later) if you want to turn this off.
% cite.sty is already installed on most LaTeX systems. Be sure and use
% version 4.0 (2003-05-27) and later if using hyperref.sty. cite.sty does
% not currently provide for hyperlinked citations.
% The latest version can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/cite/
% The documentation is contained in the cite.sty file itself.

% *** GRAPHICS RELATED PACKAGES ***
%
\ifCLASSINFOpdf
  % \usepackage[pdftex]{graphicx}
  % declare the path(s) where your graphic files are
  % \graphicspath{{../pdf/}{../jpeg/}}
  % and their extensions so you won't have to specify these with
  % every instance of \includegraphics
  % \DeclareGraphicsExtensions{.pdf,.jpeg,.png}
\else
  % or other class option (dvipsone, dvipdf, if not using dvips). graphicx
  % will default to the driver specified in the system graphics.cfg if no
  % driver is specified.
  % \usepackage[dvips]{graphicx}
  % declare the path(s) where your graphic files are
  % \graphicspath{{../eps/}}
  % and their extensions so you won't have to specify these with
  % every instance of \includegraphics
  % \DeclareGraphicsExtensions{.eps}
\fi
% graphicx was written by David Carlisle and Sebastian Rahtz. It is
% required if you want graphics, photos, etc. graphicx.sty is already
% installed on most LaTeX systems. The latest version and documentation can
% be obtained at: 
% http://www.ctan.org/tex-archive/macros/latex/required/graphics/
% Another good source of documentation is "Using Imported Graphics in
% LaTeX2e" by Keith Reckdahl which can be found as epslatex.ps or
% epslatex.pdf at: http://www.ctan.org/tex-archive/info/
%
% latex, and pdflatex in dvi mode, support graphics in encapsulated
% postscript (.eps) format. pdflatex in pdf mode supports graphics
% in .pdf, .jpeg, .png and .mps (metapost) formats. Users should ensure
% that all non-photo figures use a vector format (.eps, .pdf, .mps) and
% not a bitmapped formats (.jpeg, .png). IEEE frowns on bitmapped formats
% which can result in "jaggedy"/blurry rendering of lines and letters as
% well as large increases in file sizes.
%
% You can find documentation about the pdfTeX application at:
% http://www.tug.org/applications/pdftex

% *** MATH PACKAGES ***
%
%\usepackage[cmex10]{amsmath}
% A popular package from the American Mathematical Society that provides
% many useful and powerful commands for dealing with mathematics. If using
% it, be sure to load this package with the cmex10 option to ensure that
% only type 1 fonts will utilized at all point sizes. Without this option,
% it is possible that some math symbols, particularly those within
% footnotes, will be rendered in bitmap form which will result in a
% document that can not be IEEE Xplore compliant!
%
% Also, note that the amsmath package sets \interdisplaylinepenalty to 10000
% thus preventing page breaks from occurring within multiline equations. Use:
%\interdisplaylinepenalty=2500
% after loading amsmath to restore such page breaks as IEEEtran.cls normally
% does. amsmath.sty is already installed on most LaTeX systems. The latest
% version and documentation can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/required/amslatex/math/

% *** SPECIALIZED LIST PACKAGES ***
%
%\usepackage{algorithmic}
% algorithmic.sty was written by Peter Williams and Rogerio Brito.
% This package provides an algorithmic environment fo describing algorithms.
% You can use the algorithmic environment in-text or within a figure
% environment to provide for a floating algorithm. Do NOT use the algorithm
% floating environment provided by algorithm.sty (by the same authors) or
% algorithm2e.sty (by Christophe Fiorio) as IEEE does not use dedicated
% algorithm float types and packages that provide these will not provide
% correct IEEE style captions. The latest version and documentation of
% algorithmic.sty can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/algorithms/
% There is also a support site at:
% http://algorithms.berlios.de/index.html
% Also of interest may be the (relatively newer and more customizable)
% algorithmicx.sty package by Szasz Janos:
% http://www.ctan.org/tex-archive/macros/latex/contrib/algorithmicx/

% *** ALIGNMENT PACKAGES ***
%
%\usepackage{array}
% Frank Mittelbach's and David Carlisle's array.sty patches and improves
% the standard LaTeX2e array and tabular environments to provide better
% appearance and additional user controls. As the default LaTeX2e table
% generation code is lacking to the point of almost being broken with
% respect to the quality of the end results, all users are strongly
% advised to use an enhanced (at the very least that provided by array.sty)
% set of table tools. array.sty is already installed on most systems. The
% latest version and documentation can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/required/tools/

%\usepackage{mdwmath}
%\usepackage{mdwtab}
% Also highly recommended is Mark Wooding's extremely powerful MDW tools,
% especially mdwmath.sty and mdwtab.sty which are used to format equations
% and tables, respectively. The MDWtools set is already installed on most
% LaTeX systems. The lastest version and documentation is available at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/mdwtools/

% IEEEtran contains the IEEEeqnarray family of commands that can be used to
% generate multiline equations as well as matrices, tables, etc., of high
% quality.

%\usepackage{eqparbox}
% Also of notable interest is Scott Pakin's eqparbox package for creating
% (automatically sized) equal width boxes - aka "natural width parboxes".
% Available at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/eqparbox/

% *** SUBFIGURE PACKAGES ***
%\usepackage[tight,footnotesize]{subfigure}
% subfigure.sty was written by Steven Douglas Cochran. This package makes it
% easy to put subfigures in your figures. e.g., "Figure 1a and 1b". For IEEE
% work, it is a good idea to load it with the tight package option to reduce
% the amount of white space around the subfigures. subfigure.sty is already
% installed on most LaTeX systems. The latest version and documentation can
% be obtained at:
% http://www.ctan.org/tex-archive/obsolete/macros/latex/contrib/subfigure/
% subfigure.sty has been superceeded by subfig.sty.

%\usepackage[caption=false]{caption}
%\usepackage[font=footnotesize]{subfig}
% subfig.sty, also written by Steven Douglas Cochran, is the modern
% replacement for subfigure.sty. However, subfig.sty requires and
% automatically loads Axel Sommerfeldt's caption.sty which will override
% IEEEtran.cls handling of captions and this will result in nonIEEE style
% figure/table captions. To prevent this problem, be sure and preload
% caption.sty with its "caption=false" package option. This is will preserve
% IEEEtran.cls handing of captions. Version 1.3 (2005/06/28) and later 
% (recommended due to many improvements over 1.2) of subfig.sty supports
% the caption=false option directly:
%\usepackage[caption=false,font=footnotesize]{subfig}
%
% The latest version and documentation can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/subfig/
% The latest version and documentation of caption.sty can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/caption/

% *** FLOAT PACKAGES ***
%
%\usepackage{fixltx2e}
% fixltx2e, the successor to the earlier fix2col.sty, was written by
% Frank Mittelbach and David Carlisle. This package corrects a few problems
% in the LaTeX2e kernel, the most notable of which is that in current
% LaTeX2e releases, the ordering of single and double column floats is not
% guaranteed to be preserved. Thus, an unpatched LaTeX2e can allow a
% single column figure to be placed prior to an earlier double column
% figure. The latest version and documentation can be found at:
% http://www.ctan.org/tex-archive/macros/latex/base/

%\usepackage{stfloats}
% stfloats.sty was written by Sigitas Tolusis. This package gives LaTeX2e
% the ability to do double column floats at the bottom of the page as well
% as the top. (e.g., "\begin{figure*}[!b]" is not normally possible in
% LaTeX2e). It also provides a command:
%\fnbelowfloat
% to enable the placement of footnotes below bottom floats (the standard
% LaTeX2e kernel puts them above bottom floats). This is an invasive package
% which rewrites many portions of the LaTeX2e float routines. It may not work
% with other packages that modify the LaTeX2e float routines. The latest
% version and documentation can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/sttools/
% Documentation is contained in the stfloats.sty comments as well as in the
% presfull.pdf file. Do not use the stfloats baselinefloat ability as IEEE
% does not allow \baselineskip to stretch. Authors submitting work to the
% IEEE should note that IEEE rarely uses double column equations and
% that authors should try to avoid such use. Do not be tempted to use the
% cuted.sty or midfloat.sty packages (also by Sigitas Tolusis) as IEEE does
% not format its papers in such ways.

%\ifCLASSOPTIONcaptionsoff
%  \usepackage[nomarkers]{endfloat}
% \let\MYoriglatexcaption\caption
% \renewcommand{\caption}[2][\relax]{\MYoriglatexcaption[#2]{#2}}
%\fi
% endfloat.sty was written by James Darrell McCauley and Jeff Goldberg.
% This package may be useful when used in conjunction with IEEEtran.cls'
% captionsoff option. Some IEEE journals/societies require that submissions
% have lists of figures/tables at the end of the paper and that
% figures/tables without any captions are placed on a page by themselves at
% the end of the document. If needed, the draftcls IEEEtran class option or
% \CLASSINPUTbaselinestretch interface can be used to increase the line
% spacing as well. Be sure and use the nomarkers option of endfloat to
% prevent endfloat from "marking" where the figures would have been placed
% in the text. The two hack lines of code above are a slight modification of
% that suggested by in the endfloat docs (section 8.3.1) to ensure that
% the full captions always appear in the list of figures/tables - even if
% the user used the short optional argument of \caption[]{}.
% IEEE papers do not typically make use of \caption[]'s optional argument,
% so this should not be an issue. A similar trick can be used to disable
% captions of packages such as subfig.sty that lack options to turn off
% the subcaptions:
% For subfig.sty:
% \let\MYorigsubfloat\subfloat
% \renewcommand{\subfloat}[2][\relax]{\MYorigsubfloat[]{#2}}
% For subfigure.sty:
% \let\MYorigsubfigure\subfigure
% \renewcommand{\subfigure}[2][\relax]{\MYorigsubfigure[]{#2}}
% However, the above trick will not work if both optional arguments of
% the \subfloat/subfig command are used. Furthermore, there needs to be a
% description of each subfigure *somewhere* and endfloat does not add
% subfigure captions to its list of figures. Thus, the best approach is to
% avoid the use of subfigure captions (many IEEE journals avoid them anyway)
% and instead reference/explain all the subfigures within the main caption.
% The latest version of endfloat.sty and its documentation can obtained at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/endfloat/
%
% The IEEEtran \ifCLASSOPTIONcaptionsoff conditional can also be used
% later in the document, say, to conditionally put the References on a 
% page by themselves.

% *** PDF, URL AND HYPERLINK PACKAGES ***
%
%\usepackage{url}
% url.sty was written by Donald Arseneau. It provides better support for
% handling and breaking URLs. url.sty is already installed on most LaTeX
% systems. The latest version can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/misc/
% Read the url.sty source comments for usage information. Basically,
% \url{my_url_here}.

% *** Do not adjust lengths that control margins, column widths, etc. ***
% *** Do not use packages that alter fonts (such as pslatex).         ***
% There should be no need to do such things with IEEEtran.cls V1.6 and later.
% (Unless specifically asked to do so by the journal or conference you plan
% to submit to, of course. )

% correct bad hyphenation here
\hyphenation{op-tical net-works semi-conduc-tor}

%%% My packages
\usepackage{amsmath,epsfig}
\usepackage{amsfonts}

%%% My commands
\newcommand{\mat}[1]{\mathbf{#1}}
\renewcommand{\vec}[1]{\mathbf{#1}}

\begin{document}
%
% paper title
% can use linebreaks \\ within to get better formatting as desired
\title{Implementing Capon Adaptive Beamforming on the GPU for Real Time Ultrasound Imaging}
%
%
% author names and IEEE memberships
% note positions of commas and nonbreaking spaces ( ~ ) LaTeX will not break
% a structure at a ~ so this keeps an author's name from being broken across
% two lines.
% use \thanks{} to gain access to the first footnote area
% a separate \thanks must be used for each paragraph as LaTeX2e's \thanks
% was not built to handle multiple paragraphs
%

\author{J. P. \AA{}sen, J. I. Buskenes, C.-I. C. Nilsen, A. Austeng and S. Holm%\IEEEmembership{Member,~IEEE,}% <-this % stops a space
\thanks{M. Shell is with the Department
of Electrical and Computer Engineering, Georgia Institute of Technology, Atlanta,
GA, 30332 USA e-mail: (see http://www.michaelshell.org/contact.html).}% <-this % stops a space
\thanks{J. Doe and J. Doe are with Anonymous University.}% <-this % stops a space
\thanks{Manuscript received April 19, 2005; revised January 11, 2007.}}

% note the % following the last \IEEEmembership and also \thanks - 
% these prevent an unwanted space from occurring between the last author name
% and the end of the author line. i.e., if you had this:
% 
% \author{....lastname \thanks{...} \thanks{...} }
%                     ^------------^------------^----Do not want these spaces!
%
% a space would be appended to the last name and could cause every name on that
% line to be shifted left slightly. This is one of those "LaTeX things". For
% instance, "\textbf{A} \textbf{B}" will typeset as "A B" not "AB". To get
% "AB" then you have to do: "\textbf{A}\textbf{B}"
% \thanks is no different in this regard, so shield the last } of each \thanks
% that ends a line with a % and do not let a space in before the next \thanks.
% Spaces after \IEEEmembership other than the last one are OK (and needed) as
% you are supposed to have spaces between the names. For what it is worth,
% this is a minor point as most people would not even notice if the said evil
% space somehow managed to creep in.

% The paper headers
\markboth{Transactions on Ultrasonics and Ferroelectrics and Frequency Control,~Vol.~x, No.~y, January~2013?}%
{\AA{}sen \MakeLowercase{\textit{et al.}}: Implementing Capon Adaptive Beamforming on the GPU for Real Time Ultrasound Imaging}
% The only time the second header will appear is for the odd numbered pages
% after the title page when using the twoside option.
% 
% *** Note that you probably will NOT want to include the author's ***
% *** name in the headers of peer review papers.                   ***
% You can use \ifCLASSOPTIONpeerreview for conditional compilation here if
% you desire.

% If you want to put a publisher's ID mark on the page you can do it like
% this:
%\IEEEpubid{0000--0000/00\$00.00~\copyright~2007 IEEE}
% Remember, if you use this you must call \IEEEpubidadjcol in the second
% column for its text to clear the IEEEpubid mark.

% use for special paper notices
%\IEEEspecialpapernotice{(Invited Paper)}

% make the title area
\maketitle

\begin{abstract}
%\boldmath
The abstract goes here.
\end{abstract}
% IEEEtran.cls defaults to using nonbold math in the Abstract.
% This preserves the distinction between vectors and scalars. However,
% if the journal you are submitting to favors bold math in the abstract,
% then you can use LaTeX's standard command \boldmath at the very start
% of the abstract to achieve this. Many IEEE journals frown on math
% in the abstract anyway.

% Note that keywords are not normally used for peerreview papers.
\begin{IEEEkeywords}
Adaptive beamforming, Capon beamformer, Graphics Processing Unit, Parallel Computing, Ultrasound Imaging.
\end{IEEEkeywords}

% For peer review papers, you can put extra information on the cover
% page as needed:
% \ifCLASSOPTIONpeerreview
% \begin{center} \bfseries EDICS Category: 3-BBND \end{center}
% \fi
%
% For peerreview papers, this IEEEtran command inserts a page break and
% creates the second title. It will be ignored for other modes.
\IEEEpeerreviewmaketitle

\section{Introduction}
% The very first letter is a 2 line initial drop letter followed
% by the rest of the first word in caps.
% 
% form to use if the first word consists of a single letter:
% \IEEEPARstart{A}{demo} file is ....
% 
% form to use if you need the single drop letter followed by
% normal text (unknown if ever used by IEEE):
% \IEEEPARstart{A}{}demo file is ....
% 
% Some journals put the first two words in caps:
% \IEEEPARstart{T}{his demo} file is ....
% 
% Here we have the typical use of a "T" for an initial drop letter
% and "HIS" in caps to complete the first word.

\IEEEPARstart{I}{ntroduction} to Capon adaptive beamforming, ultrasound imaging and GPU/CUDA.

Delay and sum beamformer (DAS):
\begin{align}
z[n] = \sum_{m = 0}^{M-1}\vec{w}_m^*\vec{x}_m[n - \Delta_m[n]] = \vec{w}^H\vec{x}[n],
\end{align}
where $M$ is the number of elements.

Capon weights are found by solving the following minimization problem:
\begin{align}
&\min_{\vec{w}} E\{|z[n]|^2\} = \vec{w}^H\mat{R}\vec{w}\\
&\text{subjected to } \vec{w}^H\vec{1} = 1,
\end{align}
where $\mat{R} = E\{\vec{x}[n]\vec{x}[n]^H\}$.

The solution to this minimization problem is
\begin{align}
\vec{w}[n] = \frac{\mat{R}[n]^{-1}\vec{a}}{\vec{a}^H\mat{R}[n]^{-1}\vec{a}},
\end{align}
where $\vec{a} = \vec{1}$ when $\vec{x}$ are pre-delayed to broadside,  and $\vec{w} \in \mathbb{C}^M$.

Estimation of $\mat{R}[n]$: $\mat{\hat{R}}[n] = \vec{x}[n]\vec{x}[n]^H$.

In order to get a well conditioned $\mat{\hat{R}}$, avoid signal cancellation and to get DAS-like speckle, $\mat{\hat{R}}$ has to be averaged over $L\le M/2$ subarrays and $N_{avg}$ time samples.
\begin{align}
\mat{\hat{R}} = \frac{1}{(2N_{avg} + 1)K}\sum_{n'=n-N_{avg}}^{n+N_{avg}} \sum_{l=0}^{K} \vec{x}_l[n']\vec{x}_l[n']^H,
\end{align}
where $K = M-L+1$, $Y_{avg}\sim \tau/T_s$ ($\tau$ is the imaging pulse length and $T_s$ is the sampling period.) and $\vec{x}_l$ is the $l^\text{th}$ subarray $[x_l[n], \dotso x_{l+L}[n]]$.

Finally, $\mat{\hat{R}}$ is loaded with diagonal factor $\epsilon$, $\mat{\hat{R}} = \mat{\hat{R}} + \epsilon\mat{I}$. A weighting proportional to the output power is beneficial, and the trace of $\mat{\hat{R}}$, $\epsilon = d*tr\{\mat{\hat{R}}\}$, has previously shown good results in the literature (cite).

Building and calculating the inverse of $\hat{R}$ is computational demanding, limiting the methods accessibility. Following the innovation in GPU computing, this has now started to change.

Nilsen et. al has proposed to apply beamspace processing to further reduce the computations required. With beamspace we mean transforming the channel data from element space to a fan of beams covering the imaging sector, $\vec{x}_{BS} = \mat{B}x$.  The steering vectors in $\mat{B}$, the so-called Butler matrix, determines each beams direction in space. We can therefore easily remove beams where no interference is present by removing rows in $\mat{B}$. For a focused system like medical ultrasound, where the received signal is concentrated in a narrow band around broadside, there is a small percentage of the beams that contains almost all of the energy. Nilsen et. al has shown that as little as three beams still produced results comparable with applying the full Capon estimation in element space.

\subsection{GPU Compute Model}
In this section we briefly discuss the principle behind GPU computing. CUDA by Nvidia has been our choice of GPU developer environment. However, even though CUDA has been used to implement the Capon beamformer, the challenges and proposed methods described in this paper will apply for other massive parallel implementation environments like e.g. OpenCL. The CUDA compute model is based on execution of a kernel across a grid of compute threads. Each position in the grid holds a block of threads. Each block is further divided into groups of 32 threads, known as a warp. The maximum number of threads inside a block is restricted to 512, hence 16 warps. From a CUDA perspective the GPU consist of one or more stream multi processors (SM), where each SM is capable of 32 or more, depending on the architecture, concurrent multiplications and/or additions. The CUDA 2.x compute capability (CC) has available 48 KB of near-core shared memory per SM that can be utilized by resident blocks. The maximum number of resident blocks per SM is 8, but in addition the maximum number of resident warps per SM is restricted to 48. Hence, with a block size of 512 only $48/(512/32) = 3$ blocks can occupy the SM at once. In addition, if one thread where to use 64B of shared memory the number of allowed resident blocks with a block size of 512 would be reduced to 1. Up to a curtain point it is beneficial to have as many warps as possible occupying the SM at once to hide latency from accessing memory. The SM has for CC 2.x 32K 32-bit long registers. When implementing algorithms on the GPU it is important to keep data as close to the core as possible. Utilizing available registers and shared memory to its maximum is therefore important.

\subsection{Optimizing kernel launch}
Set the size of grid adaptively based on $L$, $M$ and $Yavg$.

It is better to slide in space than time. $K$ is typically much larger than $2*Yavg + 1$.

% needed in second column of first page if using \IEEEpubid
%\IEEEpubidadjcol

\section{Parallel Capon}\label{sec:meth}
\subsection{Building Multiple Sample Covariance Matrices}
Building $\hat{\mat{R}}$ using a sliding window approach across $K$ and $N_{avg}$. We have a limited amount of fast, near-core memory on the GPU. On NVIDIA architecture this memory is known as shared memory, and is restricted to 48KB per compute block. The maximum block size of 32x32, further restrict how large arrays we can handle inside a single compute block. It is therefore natural to restrict $L$ to a maxima of $32$ elements. Since $L \le M/2$, $M \le 64$. We can not afford to hold the full $\hat{\mat{R}}$ ($M = L$) when $\hat{\mat{R}} > 32$, because of limited amount of memory. ... Explain in detail how \texttt{buildR} is implemented in a kernel. ... Give more details on how much shared memory we have, and how it can be divided between compute blocks occupying one stream multiprocessor (SM).

Discuss the bottleneck adaptive diagonal loading causes. Possible solutions includes, use trace from previous image, use pre-calculated array power, constant weight.

Time averaging takes a lot of resources. It has been shown that time averaging can maintain both resolution and delay-and-sum-like speckle. The same speckle statistics can be obtain with small subarrays, but then we loose resolution. Small sub-arrays benefits from being less computational demanding.

%\subsubsection{CUDA Compute Model}
%The CUDA compute model is based on execution of a kernel across a grid of compute threads. Each position in the grid holds a block of threads. Each block is further divided into groups of 32 threads, known as a warp. The maximum number of threads inside a block is restricted to 512, hence 16 warps. From a CUDA perspective the GPU consist of one or more stream multi processors (SM), %where each SM is capable of 32 or more, depending on the architecture, concurrent multiplications and/or additions. The CUDA 2.x compute capability (CC) has available 48 KB of near-core shared memory per SM that can be utilized by resident blocks. The maximum number of resident blocks per SM is 8, but in addition the maximum number of resident warps per SM is restricted to 48. Hence, %with a block size of 512 only $48/(512/32) = 3$ blocks can occupy the SM at once. In addition, if one thread where to use 64B of shared memory the number of allowed resident blocks with a block size of 512 would be reduced to 1. Up to a curtain point 
%it is beneficial to have as many warps as possible occupying the SM at once to hide latency from accessing memory and executing time consuming mathematical functions. The SM has for CC 2.x 32K 32-bit long registers. When implementing algorithms on the GPU it is important to keep data as close to the core as possible. Utilizing available registers and shared memory to its maximum is %therefore important.

\subsection{Solving Multiple Small Linear Systems}
Most focus on large matrices in the literature.
We have used an unreleased GPU implementation of Gauss Jordan (GJ) elimination (by NVIDIA) to solve $\mat{R}\vec{b} = \vec{1}$, where $\vec{b} = \mat{R}^{-1}\vec{1}$. Can include details on GPU implementation of $\mat{U}^H\mat{D}\mat{U}$, but this has not proved to be faster than NVIDIA's GJ implementation. However, the complexity for solving with $\mat{U}^H\mat{D}\mat{U}$ decomposition is supposed to be $1/3$ of GJ.

In the final journal article we could discuss Conjugated gradient and Woodbury and the benefits they bring.

\section{Results}\label{sec:res}

Present images before and after Capon weights has been applied. Comment on the effect of selecting different parameters. Present graph showing running times for Cython-Capon v.s. CUDA-Capon for different choices of parameters. Comment on different choices of parameters change the running times. Temporal averaging for instance, lowers the efficiency since $2*Y_{avg}$ more operations has to be performed per $\mat{\hat{R}}$

(A prerequisite for this is to get the GPU-processing up and running on simulated data before Easter!! Maybe also a Matlab plug-in.)

\subsection{Compute Beamformer Output}

\section{Discussion}\label{sec:dis}
Discuss results presented in Section \ref{sec:res}.

\section{Conclusion}\label{sec:con}
...

% An example of a floating figure using the graphicx package.
% Note that \label must occur AFTER (or within) \caption.
% For figures, \caption should occur after the \includegraphics.
% Note that IEEEtran v1.7 and later has special internal code that
% is designed to preserve the operation of \label within \caption
% even when the captionsoff option is in effect. However, because
% of issues like this, it may be the safest practice to put all your
% \label just after \caption rather than within \caption{}.
%
% Reminder: the "draftcls" or "draftclsnofoot", not "draft", class
% option should be used if it is desired that the figures are to be
% displayed while in draft mode.
%
%\begin{figure}[!t]
%\centering
%\includegraphics[width=2.5in]{myfigure}
% where an .eps filename suffix will be assumed under latex, 
% and a .pdf suffix will be assumed for pdflatex; or what has been declared
% via \DeclareGraphicsExtensions.
%\caption{Simulation Results}
%\label{fig_sim}
%\end{figure}

% Note that IEEE typically puts floats only at the top, even when this
% results in a large percentage of a column being occupied by floats.

% An example of a double column floating figure using two subfigures.
% (The subfig.sty package must be loaded for this to work.)
% The subfigure \label commands are set within each subfloat command, the
% \label for the overall figure must come after \caption.
% \hfil must be used as a separator to get equal spacing.
% The subfigure.sty package works much the same way, except \subfigure is
% used instead of \subfloat.
%
%\begin{figure*}[!t]
%\centerline{\subfloat[Case I]\includegraphics[width=2.5in]{subfigcase1}%
%\label{fig_first_case}}
%\hfil
%\subfloat[Case II]{\includegraphics[width=2.5in]{subfigcase2}%
%\label{fig_second_case}}}
%\caption{Simulation results}
%\label{fig_sim}
%\end{figure*}
%
% Note that often IEEE papers with subfigures do not employ subfigure
% captions (using the optional argument to \subfloat), but instead will
% reference/describe all of them (a), (b), etc., within the main caption.

% An example of a floating table. Note that, for IEEE style tables, the 
% \caption command should come BEFORE the table. Table text will default to
% \footnotesize as IEEE normally uses this smaller font for tables.
% The \label must come after \caption as always.
%
%\begin{table}[!t]
%% increase table row spacing, adjust to taste
%\renewcommand{\arraystretch}{1.3}
% if using array.sty, it might be a good idea to tweak the value of
% \extrarowheight as needed to properly center the text within the cells
%\caption{An Example of a Table}
%\label{table_example}
%\centering
%% Some packages, such as MDW tools, offer better commands for making tables
%% than the plain LaTeX2e tabular which is used here.
%\begin{tabular}{|c||c|}
%\hline
%One & Two\\
%\hline
%Three & Four\\
%\hline
%\end{tabular}
%\end{table}

% Note that IEEE does not put floats in the very first column - or typically
% anywhere on the first page for that matter. Also, in-text middle ("here")
% positioning is not used. Most IEEE journals use top floats exclusively.
% Note that, LaTeX2e, unlike IEEE journals, places footnotes above bottom
% floats. This can be corrected via the \fnbelowfloat command of the
% stfloats package.

%\section{Conclusion}
%The conclusion goes here.

% if have a single appendix:
%\appendix[Proof of the Zonklar Equations]
% or
%\appendix  % for no appendix heading
% do not use \section anymore after \appendix, only \section*
% is possibly needed

% use appendices with more than one appendix
% then use \section to start each appendix
% you must declare a \section before using any
% \subsection or using \label (\appendices by itself
% starts a section numbered zero.)
%

%\appendices
%\section{Proof of the First Zonklar Equation}
%Appendix one text goes here.

% you can choose not to have a title for an appendix
% if you want by leaving the argument blank
%\section{}
%Appendix two text goes here.

% use section* for acknowledgement
\section*{Acknowledgment}

The authors would like to thank... \cite{Synnevag2009}

% Can use something like this to put references on a page
% by themselves when using endfloat and the captionsoff option.
\ifCLASSOPTIONcaptionsoff
  \newpage
\fi

% trigger a \newpage just before the given reference
% number - used to balance the columns on the last page
% adjust value as needed - may need to be readjusted if
% the document is modified later
%\IEEEtriggeratref{8}
% The "triggered" command can be changed if desired:
%\IEEEtriggercmd{\enlargethispage{-5in}}

% references section

% can use a bibliography generated by BibTeX as a .bbl file
% BibTeX documentation can be easily obtained at:
% http://www.ctan.org/tex-archive/biblio/bibtex/contrib/doc/
% The IEEEtran BibTeX style support page is at:
% http://www.michaelshell.org/tex/ieeetran/bibtex/
\bibliographystyle{IEEEtran}
% argument is your BibTeX string definitions and bibliography database(s)
%\bibliography{IEEEabrv,../bib/paper}
\bibliography{mybib}
%
% <OR> manually copy in the resultant .bbl file
% set second argument of \begin to the number of references
% (used to reserve space for the reference number labels box)
%\begin{thebibliography}{1}

%\bibitem{IEEEhowto:kopka}
%H.~Kopka and P.~W. Daly, \emph{A Guide to \LaTeX}, 3rd~ed.\hskip 1em plus
%  0.5em minus 0.4em\relax Harlow, England: Addison-Wesley, 1999.

%\end{thebibliography}

% biography section
% 
% If you have an EPS/PDF photo (graphicx package needed) extra braces are
% needed around the contents of the optional argument to biography to prevent
% the LaTeX parser from getting confused when it sees the complicated
% \includegraphics command within an optional argument. (You could create
% your own custom macro containing the \includegraphics command to make things
% simpler here.)
%\begin{biography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{mshell}}]{Michael Shell}
% or if you just want to reserve a space for a photo:

% insert where needed to balance the two columns on the last page with
% biographies
\newpage

\begin{IEEEbiography}{Jon Petter \AA{}sen}
Biography text here.
\end{IEEEbiography}

\begin{IEEEbiography}{Jo Inge Buskenes}
Biography text here.
\end{IEEEbiography}

\begin{IEEEbiography}{Carl-Inge Colombo Nilsen}
Biography text here.
\end{IEEEbiography}

% insert where needed to balance the two columns on the last page with
% biographies
\newpage

\begin{IEEEbiography}{Andreas Austeng}
Biography text here.
\end{IEEEbiography}

\begin{IEEEbiography}{Sverre Holm}
Biography text here.
\end{IEEEbiography}

% You can push biographies down or up by placing
% a \vfill before or after them. The appropriate
% use of \vfill depends on what kind of text is
% on the last page and whether or not the columns
% are being equalized.

%\vfill

% Can be used to pull up biographies so that the bottom of the last one
% is flush with the other column.
%\enlargethispage{-5in}

% that's all folks
\end{document}

### 2012-05-14 HRUIonGPU/jrnl_article.tex
Added sections about beamspace and iterative capon

@@
 % can use linebreaks \\ within to get better formatting as desired
-\title{Implementing Capon Adaptive Beamforming on the GPU for Real Time Ultrasound Imaging}
+\title{Implementing Capon Adaptive Beamforming on the GPU for Real Time Cardiac Ultrasound Imaging}
 %
@@
 \markboth{Transactions on Ultrasonics and Ferroelectrics and Frequency Control,~Vol.~x, No.~y, January~2013?}%
-{\AA{}sen \MakeLowercase{\textit{et al.}}: Implementing Capon Adaptive Beamforming on the GPU for Real Time Ultrasound Imaging}
+{\AA{}sen \MakeLowercase{\textit{et al.}}: Capon Adaptive Beamforming on the GPU}
 % The only time the second header will appear is for the odd numbered pages
@@
 \begin{IEEEkeywords}
-Adaptive beamforming, Capon beamformer, Graphics Processing Unit, Parallel Computing, Ultrasound Imaging.
+Adaptive beamforming, Capon beamformer, Graphics Processing Unit, GPGPU, Parallel Computing, Ultrasound Imaging.
 \end{IEEEkeywords}
@@
 
+Capon beamforming has recently shown promising results in several studies.
+Although it improves the lateral resolution, widespread use has been limited by its computational complexity.
+Luckily, the calculation of the Capon weights is well suited for a GPU (Graphics Processing Unit) implementation, since the calculation is both independent across pixels, and partially intra-pixel.
+
+In this paper we introduce the first GPU implementation of the Capon beamformer capable of processing a 70 degree sector cardiac image from a 64 element phased array at interactive frame rates using both spatial and temporal smoothing.
+Both types of smoothing are important for maintaining delay-and-sum-like speckle and at the same time keep the high resolution property.
+This is achieved using complex base band data, which reduces the size of each range line to a minimum.
+In addition, complex data results in complex weights; hence this is a correct implementation of Capon beamforming with all degrees of freedom preserved.
+
 Delay and sum beamformer (DAS):
@@
 where $M$ is the number of elements.
+
+\subsection{Capon Beamforming}
 
@@
 The solution to this minimization problem is
-\begin{align}
+\begin{align}\label{eq:w}
 \vec{w}[n] = \frac{\mat{R}[n]^{-1}\vec{a}}{\vec{a}^H\mat{R}[n]^{-1}\vec{a}},
@@
 Building and calculating the inverse of $\hat{R}$ is computational demanding, limiting the methods accessibility. Following the innovation in GPU computing, this has now started to change.
-
-Nilsen et.
-al has proposed to apply beamspace processing to further reduce the computations required.
-With beamspace we mean transforming the channel data from element space to a fan of beams covering the imaging sector, $\vec{x}_{BS} = \mat{B}x$.
-The steering vectors in $\mat{B}$, the so-called Butler matrix, determines each beams direction in space.
-We can therefore easily remove beams where no interference is present by removing rows in $\mat{B}$.
-For a focused system like medical ultrasound, where the received signal is concentrated in a narrow band around broadside, there is a small percentage of the beams that contains almost all of the energy.
-Nilsen et.
-al has shown that as little as three beams still produced results comparable with applying the full Capon estimation in element space.
 
@@
 \subsection{Solving Multiple Small Linear Systems}
-Most focus on large matrices in the literature.
-We have used an unreleased GPU implementation of Gauss Jordan (GJ) elimination (by NVIDIA) to solve $\mat{R}\vec{b} = \vec{1}$, where $\vec{b} = \mat{R}^{-1}\vec{1}$.
+The focus has mostly been on large systems in the literature, and GPU libraries usually lack a comprehensive collection of batched solvers.
+The reason for this is two-folded, first the GPU needs to solve thousand of small systems in order to beat the low memory latency of the CPU, second there is a range of system dimensions that does not fit well with the GPU architecture, where only a small amount of near-core memory is available.
+Because of this, it has been proven hard to optimize a manageable set of solvers that provides speedups compared with the CPU in all cases.
+
+For the results presented in this paper, we have used an unreleased GPU implementation of Gauss Jordan (GJ) elimination (by NVIDIA) to solve $\mat{R}\vec{b} = \vec{1}$, where $\vec{b} = \mat{R}^{-1}\vec{1}$.
 Can include details on GPU implementation of $\mat{U}^H\mat{D}\mat{U}$, but this has not proved to be faster than NVIDIA's GJ implementation.
-However, the complexity for solving with $\mat{U}^H\mat{D}\mat{U}$ decomposition is supposed to be $1/3$ of GJ.
+However, the complexity for solving with $\mat{U}^H\mat{D}\mat{U}$ decomposition is supposed to be $1/2$ of GJ.
 
 In the final journal article we could discuss Conjugated gradient and Woodbury and the benefits they bring.
+
+\subsection{Compute Beamformer Output}
+
+\section{Real Time Performance for $L > 16$}
+
+\subsection{Beamspace Processing}
+Nilsen et.
+al has proposed to apply beamspace processing to further reduce the required computations.
+With beamspace we mean transforming the channel data from element space to a fan of beams covering the imaging sector, $\vec{x}_{BS} = \mat{B}x$.
+The steering vectors in $\mat{B}$, the so-called Butler matrix, determines each beams direction in space.
+We can therefore easily remove beams where no interference is present by removing rows in $\mat{B}$.
+For a focused system like a cardiac ultrasound scanner, where the received signal is concentrated in a narrow band around broadside, there is a small percentage of the beams that contains almost all the energy.
+Nilsen et.
+al concluded that as little as three beams around broadside where adequate to produce results comparable to applying Capon beamforming in element space.
+The beamspace version of Capon is derived by replacing the covariance matrix in Eq.
+\ref{eq:w} with the beamspace covariance matrix $\mat{R}_{BS} = \vec{x}_{BS}\vec{x}_{BS}^H$.
+
+\subsection{Iterative Capon}
+The idea behind iterative Capon is to perform Capon using small subarrays in several iterations until there is only two subarrays to sum.
+In this way we fix the size of L, hence the complexity of solving $\mat{R}\vec{b} = \vec{1}$.
+However, for each iteration we re-obtain L-1 degrees of freedom until all are retrieved after $K = M-L+1$ iterations.
+Since subarray averaging has been applied to decorrelated the signal, we might loose this effect when reestablishing all degrees of freedom.
+The number of iterations should therefore be restricted to $K/2$ to equal our standard choice of $L = M/2$ when it comes to degrees of freedom.
 
@@
 Comment on the effect of selecting different parameters.
-Present graph showing running times for Cython-Capon v.s.
+Present graph showing executing times for Cython-Capon v.s.
 CUDA-Capon for different choices of parameters.
@@
 
-(A prerequisite for this is to get the GPU-processing up and running on simulated data before Easter!! Maybe also a Matlab plug-in.)
-
-\subsection{Compute Beamformer Output}
+Present images, both simulated and in-vivo, processed using beamspace Capon and iterative Capon. Comment on the differences between the two and on how much they deviates from the full method.
+Show beampatterns for the two compared with the full method. Show executing times for the two methods compared with the full method on the GPU.
 
@@
 Discuss results presented in Section \ref{sec:res}.
+
+For a typical cardiac ultrasound image of 80*400 pixels (70 degree sector, 15cm range) acquired using a 2.5MHz, $M=64$ element phased array, the result is 10fps (subarray length $L=M/2$, temporal smoothing over 3 samples).
+If we do a 2-element pre-beamforming to get the channel count down to 32, the frame rate increases to 44fps.
+For a 32 element phased array we need less beams to cover the sector (40*400 pixels), hence with the same parameters the frame rate increases to 87fps.
+The bottleneck of the Capon algorithm is to find the inverse covariance matrix for all pixels.
+This is an $O(L^3)$ operation, and results show that when L is larger than 16, real time performance is not possible for cardiac imaging with current high-end hardware.
+The target GPU has been Nvidia’s Quadro 6000, capable of 1Tflops.
 

### 2012-05-23 HRUIonGPU/jrnl_article.tex

@@
 \author{J. P. \AA{}sen, J. I. Buskenes, C.-I. C. Nilsen, A. Austeng and S. Holm%\IEEEmembership{Member,~IEEE,}% <-this % stops a space
-\thanks{M. Shell is with the Department
-of Electrical and Computer Engineering, Georgia Institute of Technology, Atlanta,
-GA, 30332 USA e-mail: (see http://www.michaelshell.org/contact.html).}% <-this % stops a space
-\thanks{J. Doe and J. Doe are with Anonymous University.}% <-this % stops a space
-\thanks{Manuscript received April 19, 2005; revised January 11, 2007.}}
+%\thanks{M. Shell is with the Department
+%of Electrical and Computer Engineering, Georgia Institute of Technology, Atlanta,
+%GA, 30332 USA e-mail: (see http://www.michaelshell.org/contact.html).}% <-this % stops a space
+%\thanks{J. Doe and J. Doe are with Anonymous University.}% <-this % stops a space
+%\thanks{Manuscript received April 19, 2005; revised January 11, 2007.}}
+}
 
@@
 
-\IEEEPARstart{I}{ntroduction} to Capon adaptive beamforming, ultrasound imaging and GPU/CUDA.
+%\IEEEPARstart{I}{ntroduction} to Capon adaptive beamforming, ultrasound imaging and GPU/CUDA.
 
-Capon beamforming has recently shown promising results in several studies.
-Although it improves the lateral resolution, widespread use has been limited by its computational complexity.
-Luckily, the calculation of the Capon weights is well suited for a GPU (Graphics Processing Unit) implementation, since the calculation is both independent across pixels, and partially intra-pixel.
+\IEEEPARstart{W}{hen} sensing the world around us with an array of elements, we perform a spatio-temporal filtering operation known as beamforming.
+The conventional form of beamforming is delay-and-sum (DAS), where the signal from each element, $\vec{x}_m[n]$, is delayed to broadside with an appropriate delay $\Delta_m[n]$.
+The beamformer output, $z[n]$, is then the sum of all $M$ channels weighted with a factor $w_m$,
+\begin{align}
+z[n] = \sum_{m = 0}^{M-1}w_m^*x_m[n - \Delta_m[n]] = \vec{w}^H\vec{x}[n].
+\end{align}
+The weight $\vec{w}$, is usually applied to trade resolution for a better signal-to-noise (SNR) ratio. Hence, we increase the main lobe width of the system while gaining a lower side lobe level.
 
+The basic idea of the Capon beamformer is to calculate the weights on-the-fly, based on the data at hand, and on a selected optimization criteria.
+For a base-band data vector $\vec{x}_m \in \mathbb{C}^M$, we now have a data-dependent weight, $\vec{w}(\vec{x}[n])$ that when applied on $\vec{x}$, is the best approximation to the selected criteria.
+Using the Capon criterion, the adaptive weights are found by solving the following minimization problem:
+\begin{align}
+&\min_{\vec{w}} E\{|z[n]|^2\} = \vec{w}^H\mat{R}\vec{w}\label{eq:minz}\\
+&\text{subjected to } \vec{w}^H\vec{1} = 1,\label{eq:constraint}
+\end{align}
+where $\mat{R} = E\{\vec{x}[n]\vec{x}[n]^H\}$ is the signal covariance matrix.
+Thus, we minimize the output power, while maintaining unit gain in the look direction.
+The constraint in Eq.
+\ref{eq:constraint} is there to avoid the trivial solution $\vec{w} = 0$, and a distorted respons at broadside.
+The solution to the minimization problem in Eq.
+\ref{eq:minz} is
+\begin{align}\label{eq:w}
+\vec{w}[n] = \frac{\mat{R}[n]^{-1}\vec{a}}{\vec{a}^H\mat{R}[n]^{-1}\vec{a}},
+\end{align}
+where $\vec{a} = \vec{1}$ when $\vec{x}$ are pre-delayed to broadside.
+In practice, the covariance matrix has to be approximated by the sample covariance matrix, $\mat{\hat{R}}[n] = \vec{x}[n]\vec{x}[n]^H$.
+In passive systems, assuming stationary statistics, the sample covariance matrix is typically averaged over all available samples to improve on the estimate.
+For an active, pulsed system like ultrasound imaging, the available data from which $\mat{R}$ can be estimated is quite limited.
+The resulting sample covariance matrix might therefore be an inaccurate estimate, and in addition poorly conditioned.
+Another challenge with applying Capon to active systems is the effect of signal cancellation, where high correlation between signal and interference favors weights with opposite phase.
+In this way, the beamformer output can be zero at broadside (with signal pressent), even with the distortion-less criterion applied.
+
+In order to get a well conditioned $\mat{\hat{R}}$, avoid signal cancellation and to get DAS-like speckle, $\mat{\hat{R}}$ has to be averaged over $L\le M/2$ subarrays and $N$ time samples.
+\begin{align}
+\mat{\hat{R}} = \frac{1}{NK}\sum_{n'=n-N_{avg}}^{n+N_{avg}} \sum_{l=0}^{K} \vec{x}_l[n']\vec{x}_l[n']^H,
+\end{align}
+where $K = M-L+1$, $N = 2N_{avg} + 1\sim \tau/T_s$ ($\tau$ is the pulse length and $T_s$ is the sampling period) and $\vec{x}_l$ is the $l^\text{th}$ subarray $[x_l[n], \dotso x_{l+L}[n]]$.
+Finally, $\mat{\hat{R}}$ is loaded with diagonal factor $\epsilon$, $\mat{\hat{R}} = \mat{\hat{R}} + \epsilon\mat{I}$.
+A weighting proportional to the output power is beneficial, and the trace of $\mat{\hat{R}}$, $\epsilon = d*tr\{\mat{\hat{R}}\}$, has previously shown good results in the literature (cite).
+
+Building and calculating the inverse of $\hat{R}$ is computational demanding, limiting the methods accessibility. Following the innovation in GPU computing, this has now started to change. 
 In this paper we introduce the first GPU implementation of the Capon beamformer capable of processing a 70 degree sector cardiac image from a 64 element phased array at interactive frame rates using both spatial and temporal smoothing.
@@
 In addition, complex data results in complex weights; hence this is a correct implementation of Capon beamforming with all degrees of freedom preserved.
+For high channel count systems we propose two ways of reducing computational complexity while maintaining
 
-Delay and sum beamformer (DAS):
-\begin{align}
-z[n] = \sum_{m = 0}^{M-1}\vec{w}_m^*\vec{x}_m[n - \Delta_m[n]] = \vec{w}^H\vec{x}[n],
-\end{align}
-where $M$ is the number of elements.
-
-\subsection{Capon Beamforming}
-
-Capon weights are found by solving the following minimization problem:
-\begin{align}
-&\min_{\vec{w}} E\{|z[n]|^2\} = \vec{w}^H\mat{R}\vec{w}\\
-&\text{subjected to } \vec{w}^H\vec{1} = 1,
-\end{align}
-where $\mat{R} = E\{\vec{x}[n]\vec{x}[n]^H\}$.
-
-The solution to this minimization problem is
-\begin{align}\label{eq:w}
-\vec{w}[n] = \frac{\mat{R}[n]^{-1}\vec{a}}{\vec{a}^H\mat{R}[n]^{-1}\vec{a}},
-\end{align}
-where $\vec{a} = \vec{1}$ when $\vec{x}$ are pre-delayed to broadside,  and $\vec{w} \in \mathbb{C}^M$.
-
-Estimation of $\mat{R}[n]$: $\mat{\hat{R}}[n] = \vec{x}[n]\vec{x}[n]^H$.
-
-In order to get a well conditioned $\mat{\hat{R}}$, avoid signal cancellation and to get DAS-like speckle, $\mat{\hat{R}}$ has to be averaged over $L\le M/2$ subarrays and $N_{avg}$ time samples.
-\begin{align}
-\mat{\hat{R}} = \frac{1}{(2N_{avg} + 1)K}\sum_{n'=n-N_{avg}}^{n+N_{avg}} \sum_{l=0}^{K} \vec{x}_l[n']\vec{x}_l[n']^H,
-\end{align}
-where $K = M-L+1$, $Y_{avg}\sim \tau/T_s$ ($\tau$ is the imaging pulse length and $T_s$ is the sampling period.) and $\vec{x}_l$ is the $l^\text{th}$ subarray $[x_l[n], \dotso x_{l+L}[n]]$.
-
-Finally, $\mat{\hat{R}}$ is loaded with diagonal factor $\epsilon$, $\mat{\hat{R}} = \mat{\hat{R}} + \epsilon\mat{I}$.
-A weighting proportional to the output power is beneficial, and the trace of $\mat{\hat{R}}$, $\epsilon = d*tr\{\mat{\hat{R}}\}$, has previously shown good results in the literature (cite).
-
-Building and calculating the inverse of $\hat{R}$ is computational demanding, limiting the methods accessibility. Following the innovation in GPU computing, this has now started to change.
-
-\subsection{GPU Compute Model}
+\section{GPU Compute Model}
+This turned out to be to technical?
 In this section we briefly discuss the principle behind GPU computing.
@@
 \section{Parallel Capon}\label{sec:meth}
+Capon beamforming has recently shown promising results in several studies.
+Although it improves the lateral resolution, widespread use has been limited by its computational complexity.
+However, the calculation of Capon weights is well suited for a GPU (Graphics Processing Unit) implementation, since the calculation is both independent across pixels, and partially intra-pixel.
+
 \subsection{Building Multiple Sample Covariance Matrices}

### 2012-05-23 HRUIonGPU/jrnl_article.tex
done some writing on the HRUIonGPU article

@@
 %\IEEEPARstart{I}{ntroduction} to Capon adaptive beamforming, ultrasound imaging and GPU/CUDA.
+\IEEEPARstart{C}{apon} adaptive beamforming applied in a medical ultrasound setting has recently shown promising results in several studies \cite{Synnevag2007, Austeng2008, Vignon2008, Viola, Mehdizadeh2012}.
+Although the method introduce some interesting trade-offs, like maintaining image resolution combined with improved framerate, smaller probes, or improved penetration \cite{Synnevag2009}, widespread use has been limited by the tremendous computational complexity associated with the method \cite{So2011, Chen2011a, Chen2011}.
+In this paper we introduce the first implementation of the Capon Beamformer exhibiting real-time performance when applied on typical cardiac data sets.
+To accomplish this task we make use of the massive parallel processing power found in modern graphics processing units (GPU's).
 
-\IEEEPARstart{W}{hen} sensing the world around us with an array of elements, we perform a spatio-temporal filtering operation known as beamforming.
-The conventional form of beamforming is delay-and-sum (DAS), where the signal from each element, $\vec{x}_m[n]$, is delayed to broadside with an appropriate delay $\Delta_m[n]$.
+\section{Background}
+\subsection{Capon Beamforming}
+When sensing the world around us with an array of elements, we perform a spatio-temporal filtering operation known as beamforming.
+The conventional form of beamforming is delay-and-sum (DAS), where the signal recorded on each element, $x_m[n]$, is delayed to broadside with an appropriate delay $\Delta_m[n]$.
 The beamformer output, $z[n]$, is then the sum of all $M$ channels weighted with a factor $w_m$,
@@
 \end{align}
-The weight $\vec{w}$, is usually applied to trade resolution for a better signal-to-noise (SNR) ratio. Hence, we increase the main lobe width of the system while gaining a lower side lobe level.
+In conventional beamforming, the weight $w_m \in \mathbb{R}$ is usually applied to trade resolution for a better signal-to-noise ratio (SNR).
+Hence, by increasing the main lobe width of the system while gaining lower side lobe levels.
 
-The basic idea of the Capon beamformer is to calculate the weights on-the-fly, based on the data at hand, and on a selected optimization criteria.
-For a base-band data vector $\vec{x}_m \in \mathbb{C}^M$, we now have a data-dependent weight, $\vec{w}(\vec{x}[n])$ that when applied on $\vec{x}$, is the best approximation to the selected criteria.
-Using the Capon criterion, the adaptive weights are found by solving the following minimization problem:
+The basic idea of an adaptive beamformer is to calculate the weights on-the-fly, based on the data at hand, and on a selected optimization criteria in such a way that we minimize interference impinging from other directions than the point of focus.
+For a complex data vector $\vec{x}[n] \in \mathbb{C}^M$, we now have data-dependent weights, $\vec{w}(\vec{x}[n])$, that when applied on $\vec{x}[n]$ is the best approximation to achieving the selected criteria.
+With the Capon beamformer, adaptive weights are found by solving the following minimization problem \cite{Capon1969}:
 \begin{align}
-&\min_{\vec{w}} E\{|z[n]|^2\} = \vec{w}^H\mat{R}\vec{w}\label{eq:minz}\\
-&\text{subjected to } \vec{w}^H\vec{1} = 1,\label{eq:constraint}
+&\min_{\vec{w}} E\{|z[n]|^2\} = \min_{\vec{w}}\vec{w}^H\mat{R}\vec{w}\label{eq:minz}\\
+&\text{subjected to } \vec{w}^H\vec{a} = 1,\label{eq:constraint}
 \end{align}
-where $\mat{R} = E\{\vec{x}[n]\vec{x}[n]^H\}$ is the signal covariance matrix.
+where $\mat{R} = E\{\vec{x}[n]\vec{x}[n]^H\}$ is the spatial covariance matrix, and $\vec{a}$ is the far-field steering direction.
 Thus, we minimize the output power, while maintaining unit gain in the look direction.
 The constraint in Eq.
-\ref{eq:constraint} is there to avoid the trivial solution $\vec{w} = 0$, and a distorted respons at broadside.
+\ref{eq:constraint} is there to avoid the trivial solution $\vec{w} = 0$, and a distorted respons in the direction of $\vec{a}$.
 The solution to the minimization problem in Eq.
-\ref{eq:minz} is
+\ref{eq:minz} and \ref{eq:constraint} is
 \begin{align}\label{eq:w}
@@
 where $\vec{a} = \vec{1}$ when $\vec{x}$ are pre-delayed to broadside.
+
 In practice, the covariance matrix has to be approximated by the sample covariance matrix, $\mat{\hat{R}}[n] = \vec{x}[n]\vec{x}[n]^H$.
-In passive systems, assuming stationary statistics, the sample covariance matrix is typically averaged over all available samples to improve on the estimate.
+In passive systems, assuming stationary statistics, the sample covariance matrix is typically averaged over all available samples to improve on the estimate \cite{Krima}.
 For an active, pulsed system like ultrasound imaging, the available data from which $\mat{R}$ can be estimated is quite limited.
-The resulting sample covariance matrix might therefore be an inaccurate estimate, and in addition poorly conditioned.
-Another challenge with applying Capon to active systems is the effect of signal cancellation, where high correlation between signal and interference favors weights with opposite phase.
-In this way, the beamformer output can be zero at broadside (with signal pressent), even with the distortion-less criterion applied.
-
-In order to get a well conditioned $\mat{\hat{R}}$, avoid signal cancellation and to get DAS-like speckle, $\mat{\hat{R}}$ has to be averaged over $L\le M/2$ subarrays and $N$ time samples.
+The resulting sample covariance matrix might be both inaccurate, and poorly conditioned.
+Another challenge with applying Capon to active systems is the effect of signal cancellation, where high correlation between signal and interference favours weights that in the end cancel out the signal \cite{Synnevag2009a}.
+In this way, the beamformer output can be zero at broadside, even when there is signal present and the distortion-less criterion is applied.
+In order to get a well conditioned $\mat{\hat{R}}$, avoid signal cancellation and to get DAS-like speckle, $\mat{\hat{R}}$ has to be averaged over $L\le M/2$ long subarrays and $N$ time samples \cite{Synnevag2007, Synnevag2007a}.
 \begin{align}
@@
 Finally, $\mat{\hat{R}}$ is loaded with diagonal factor $\epsilon$, $\mat{\hat{R}} = \mat{\hat{R}} + \epsilon\mat{I}$.
-A weighting proportional to the output power is beneficial, and the trace of $\mat{\hat{R}}$, $\epsilon = d*tr\{\mat{\hat{R}}\}$, has previously shown good results in the literature (cite).
+A weighting proportional to the output power is often used to reduce the need of parameter adjustments, and the trace of $\mat{\hat{R}}$, $\epsilon = d*tr\{\mat{\hat{R}}\}$ has previously shown good results in the literature \cite{Synnevag2007}.
 
@@
 
-\section{GPU Compute Model}
+\subsection{GPU Compute Model}
 This turned out to be to technical?
@@
 \section{Parallel Capon}\label{sec:meth}
-Capon beamforming has recently shown promising results in several studies.
-Although it improves the lateral resolution, widespread use has been limited by its computational complexity.
+Although Capon adaptive beamforming improves the lateral resolution and contrast, widespread use has been limited by its computational complexity.
 However, the calculation of Capon weights is well suited for a GPU (Graphics Processing Unit) implementation, since the calculation is both independent across pixels, and partially intra-pixel.
+
+The algorithm can be broken down into three main steps. For each data vector $\vec{x}$ in the image:
+\begin{enumerate}
+\item Build the sample covariance matrix $\mat{\hat{R}}$
+\item Solve the linear system of equations $\mat{\hat{R}}\vec{b} =\vec{ 1}$
+\item Calculate $\vec{w} = \vec{b}/|\vec{b}|$
+\end{enumerate}
+The computational complexity of these three steps, with $L = M/2$, is $O(L^3)$, $O(L^3)$, and $O(L)$ respectively.
+For larger $L$'s, the first two steps quickly becomes unmanageable, and to achieve real time performance it is therefore important to find novel solutions to these steps.
 
@@
 Nilsen et.
-al has proposed to apply beamspace processing to further reduce the required computations.
+al \cite{Nilsen2009} has proposed to apply beamspace processing to further reduce the required computations.
 With beamspace we mean transforming the channel data from element space to a fan of beams covering the imaging sector, $\vec{x}_{BS} = \mat{B}x$.
@@
 %\bibliography{IEEEabrv,../bib/paper}
-\bibliography{mybib}
+\bibliography{mybib} % link to mendeley organized bibtex-file
 %

### 2012-05-23 HRUIonGPU/jrnl_article.tex
done some more writing on the article. to hell with the exam.

@@
 \IEEEPARstart{C}{apon} adaptive beamforming applied in a medical ultrasound setting has recently shown promising results in several studies \cite{Synnevag2007, Austeng2008, Vignon2008, Viola, Mehdizadeh2012}.
-Although the method introduce some interesting trade-offs, like maintaining image resolution combined with improved framerate, smaller probes, or improved penetration \cite{Synnevag2009}, widespread use has been limited by the tremendous computational complexity associated with the method \cite{So2011, Chen2011a, Chen2011}.
-In this paper we introduce the first implementation of the Capon Beamformer exhibiting real-time performance when applied on typical cardiac data sets.
-To accomplish this task we make use of the massive parallel processing power found in modern graphics processing units (GPU's).
+Although the method introduce some interesting trade-offs in addition to just improved resolution, like maintaining image resolution combined with higher framerates, smaller probes, or deeper penetration \cite{Synnevag2009}, widespread use has been limited by the tremendous computational complexity associated with the method \cite{So2011}.
+In this paper we introduce the first implementation of the Capon beamformer exhibiting real-time performance when applied on typical cardiac data sets.
+To accomplish this task we make use of the massive parallel processing power found in modern graphics processing units (GPU's), combined with novel methods to reduce the computational complexity as the problem size grows.
+Chen et al.
+\cite{Chen2011a, Chen2011} have previously implemented the Capon beamformer on a GPU, however their implementation was based on real data, where the Capon beamformer assumes complex.
+Using real data will results in symmetric windows, and even if the number of samples they managed to process per second was high, both the number of floating point operations per sample and the imaging region is reduced compared with use of complex data.
+It is therefore crucial to use complex data within the Capon beamformer framework, both for reducing the amount of data per image, and to allow for asymmetric windows.
+
+In the next section we present background information about the Capon beamformer and the GPU computing model.
+In Section \ref{sec:meth} we go in detail about the parallel implementation of Capon beamforming, before we in Section \ref{sec:meth2} introduce methods to apply when the problem size increases.
+Benchmark results and resulting images are presented in Section \ref{sec:res}, and finally we discuss the results in Section \ref{sec:dis}, and draw conclusions in Section \ref{sec:con}.
 
@@
 \end{align}
-where $\vec{a} = \vec{1}$ when $\vec{x}$ are pre-delayed to broadside.
+where $\vec{a} = \vec{1}$ when $\vec{x}$ is pre-delayed to broadside.
 
@@
 For an active, pulsed system like ultrasound imaging, the available data from which $\mat{R}$ can be estimated is quite limited.
-The resulting sample covariance matrix might be both inaccurate, and poorly conditioned.
+The resulting sample covariance matrix might therefore be both inaccurate, and poorly conditioned.
 Another challenge with applying Capon to active systems is the effect of signal cancellation, where high correlation between signal and interference favours weights that in the end cancel out the signal \cite{Synnevag2009a}.
@@
 \begin{align}
-\mat{\hat{R}} = \frac{1}{NK}\sum_{n'=n-N_{avg}}^{n+N_{avg}} \sum_{l=0}^{K} \vec{x}_l[n']\vec{x}_l[n']^H,
+\mat{\hat{R}} = \frac{1}{NK}\sum_{n'=n-N_{avg}}^{n+N_{avg}} \sum_{l=0}^{K} \vec{x}_l[n']\vec{x}_l[n']^H,\label{eq:R}
 \end{align}
 where $K = M-L+1$, $N = 2N_{avg} + 1\sim \tau/T_s$ ($\tau$ is the pulse length and $T_s$ is the sampling period) and $\vec{x}_l$ is the $l^\text{th}$ subarray $[x_l[n], \dotso x_{l+L}[n]]$.
-Finally, $\mat{\hat{R}}$ is loaded with diagonal factor $\epsilon$, $\mat{\hat{R}} = \mat{\hat{R}} + \epsilon\mat{I}$.
-A weighting proportional to the output power is often used to reduce the need of parameter adjustments, and the trace of $\mat{\hat{R}}$, $\epsilon = d*tr\{\mat{\hat{R}}\}$ has previously shown good results in the literature \cite{Synnevag2007}.
+Finally, $\mat{\hat{R}}$ is loaded with diagonal factor $\epsilon$, 
+\begin{align}
+\mat{\hat{R}} = \mat{\hat{R}} + \epsilon\mat{I}
+\end{align}
+A weighting proportional to the output power is often used to reduce the need of parameter adjustments, and the trace of $\mat{\hat{R}}$, 
+\begin{align}
+\epsilon = d*tr\{\mat{\hat{R}}\},
+\end{align}
+has previously shown good results in the literature \cite{Synnevag2007}.
 
-Building and calculating the inverse of $\hat{R}$ is computational demanding, limiting the methods accessibility. Following the innovation in GPU computing, this has now started to change. 
-In this paper we introduce the first GPU implementation of the Capon beamformer capable of processing a 70 degree sector cardiac image from a 64 element phased array at interactive frame rates using both spatial and temporal smoothing.
+%Building and calculating the inverse of $\hat{R}$ is computational demanding, limiting the methods accessibility. Following the innovation in GPU computing, this has now started to change. 
+%In this paper we introduce the first GPU implementation of the Capon beamformer capable of processing a 70 degree sector cardiac image from a 64 element phased array at interactive frame rates using both spatial and temporal smoothing.
 Both types of smoothing are important for maintaining delay-and-sum-like speckle and at the same time keep the high resolution property.
@@
 \subsection{GPU Compute Model}
-This turned out to be to technical?
 In this section we briefly discuss the principle behind GPU computing.
-CUDA by Nvidia has been our choice of GPU developer environment.
-However, even though CUDA has been used to implement the Capon beamformer, the challenges and proposed methods described in this paper will apply for other massive parallel implementation environments like e.g.
-OpenCL.
-The CUDA compute model is based on execution of a kernel across a grid of compute threads.
+CUDA by Nvidia has been our choice of GPU developer environment \cite{Nvidia2011}.
+However, even though CUDA has been used to implement the Capon beamformer, the challenges and proposed methods described in this paper applies for other massive parallel implementation environments like e.g.
+OpenCL as well.
+The CUDA compute model is based on execution of a kernel-function across a grid of compute threads.
 Each position in the grid holds a block of threads.
-Each block is further divided into groups of 32 threads, known as a warp.
-The maximum number of threads inside a block is restricted to 512, hence 16 warps.
-From a CUDA perspective the GPU consist of one or more stream multi processors (SM), where each SM is capable of 32 or more, depending on the architecture, concurrent multiplications and/or additions.
-The CUDA 2.x compute capability (CC) has available 48 KB of near-core shared memory per SM that can be utilized by resident blocks.
+%Each block is further divided into groups of 32 threads, known as a warp. The maximum number of threads inside a block is restricted to 512, hence 16 warps. 
+From a CUDA perspective the GPU consist of one or more stream multi processors (SM's), where each SM is capable of 32 or more, depending on the architecture, concurrent multiplications and/or additions.
+A group of blocks is scheduled to each SM until all blocks have been processed.
+More important, the SM has a limited amount of near-core memory and registers available.
+Each thread therefore has to be as light on memory as possible and has to perform enough instructions per used byte in order to hide memory latency.
+This pose a real challenge as the target problem has to map to the parallel framework both on a grid, block and memory consumption level in order to benefits from parallel acceleration.
+
+%When implementing algorithms on the GPU it is important to keep data as close to the core as possible. Utilizing available registers and shared memory to its maximum is therefore important.
+
+%More important, the CUDA 2.x compute capability (CC) has available 48 KB of near-core shared memory per SM that can be utilized by resident blocks.
 The maximum number of resident blocks per SM is 8, but in addition the maximum number of resident warps per SM is restricted to 48.
@@
 
-\subsection{Optimizing kernel launch}
-Set the size of grid adaptively based on $L$, $M$ and $Yavg$.
-
-It is better to slide in space than time. $K$ is typically much larger than $2*Yavg + 1$.
+%\subsection{Optimizing kernel launch}
+%Set the size of grid adaptively based on $L$, $M$ and $Yavg$.
+%It is better to slide in space than time. $K$ is typically much larger than $2*Yavg + 1$.
 
@@
 \section{Parallel Capon}\label{sec:meth}
-Although Capon adaptive beamforming improves the lateral resolution and contrast, widespread use has been limited by its computational complexity.
-However, the calculation of Capon weights is well suited for a GPU (Graphics Processing Unit) implementation, since the calculation is both independent across pixels, and partially intra-pixel.
-
-The algorithm can be broken down into three main steps. For each data vector $\vec{x}$ in the image:
+As mentioned, the widespread use of Capon adaptive beamforming has been limited by its computational complexity.
+However, the calculation of Capon weights is well suited for a GPU implementation, since the calculations are both independent across pixels, and partially intra-pixel.
+On a top level it is trivial to recognise that one grid cell of threads should process one pixel in the final image.
+Further, on a pixel-level, the algorithm can be broken down into three main steps.
+For each data vector $\vec{x}$ in the image:
 \begin{enumerate}
@@
 \item Solve the linear system of equations $\mat{\hat{R}}\vec{b} =\vec{ 1}$
-\item Calculate $\vec{w} = \vec{b}/|\vec{b}|$
+\item Calculate $\vec{w} = \vec{b}/|\vec{b}|$ and the amplitude $\vec{w}^H\vec{x}$
 \end{enumerate}
@@
 \subsection{Building Multiple Sample Covariance Matrices}
-Building $\hat{\mat{R}}$ using a sliding window approach across $K$ and $N_{avg}$.
+The straight forward sum in Eq.
+\ref{eq:R} has complexity $O(L^2KN) = O(L^3)$ if $L$ is large.
+Note that N typically is small compared to L and M.
+For smaller $L$ the construction of $\mat{\hat{R}}$ is therefore dominated by $M$.
+The quadratic complexity for large $L$ can be improved by exploiting the overlap from element to element on the diagonals of $\mat{\hat{R}}$ (ref figure).
+This overlap is present both across subarrays and time.
+
+%Building $\hat{\mat{R}}$ using a sliding window approach across $K$ and $N_{avg}$.
 We have a limited amount of fast, near-core memory on the GPU.
@@
 
-Discuss the bottleneck adaptive diagonal loading causes. Possible solutions includes, use trace from previous image, use pre-calculated array power, constant weight.
+%Discuss the bottleneck adaptive diagonal loading causes. Possible solutions includes, use trace from previous image, use pre-calculated array power, constant weight.
 
-Time averaging takes a lot of resources.
+%Time averaging takes a lot of resources.
 It has been shown that time averaging can maintain both resolution and delay-and-sum-like speckle.
@@
 
-\section{Real Time Performance for $L > 16$}
+\section{Real Time Performance for $L > 16$}\label{sec:meth2}
 

### 2012-05-24 HRUIonGPU/jrnl_article.tex
article writing jp

@@
 OpenCL as well.
-The CUDA compute model is based on execution of a kernel-function across a grid of compute threads.
-Each position in the grid holds a block of threads.
+
+The CUDA compute model is based on execution of a kernel-function across a grid of compute threads. Each position in the grid holds a block of threads.
 %Each block is further divided into groups of 32 threads, known as a warp. The maximum number of threads inside a block is restricted to 512, hence 16 warps. 
@@
 \section{Parallel Capon}\label{sec:meth}
-As mentioned, the widespread use of Capon adaptive beamforming has been limited by its computational complexity.
+As mentioned, widespread use of Capon adaptive beamforming has been limited by its computational complexity.
 However, the calculation of Capon weights is well suited for a GPU implementation, since the calculations are both independent across pixels, and partially intra-pixel.
 On a top level it is trivial to recognise that one grid cell of threads should process one pixel in the final image.
-Further, on a pixel-level, the algorithm can be broken down into three main steps.
+Continuing, on a pixel-level, the algorithm can be broken down into three main steps that to some extent can be further divided into a small set of parallel processes.
 For each data vector $\vec{x}$ in the image:

### 2012-06-01 FFI-Simulator/ffi-simulator.tex
Adding FFI simuator documentation

%===============%                               %~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~%
% DOCUMENTCLASS %                                See full option description in "mytemplate.cls"
%===============%                               %~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~%

\documentclass[
   article                                      % Document type (non-standard)
 , 12pt                                         % Text size
%  , draft
%  , final                                        % Quality
 , xelatex                                      % Use the XeLaTeX compiler
%  , biblatex                                     % Use 'biblatex' for references (best by far)
 , bibtex                                       % Use 'bibtex' for references (oldschool :)
%  , movie
%  , notodos                                      % Disable todos
 , layout
%  , defaultformat                                % Attempt to use default formatting options
%  , glossary                                     % Use a glossary
]{common/mytemplate}

% \bibliography{references}
\hypersetup{
   bookmarksopen=true
 , bookmarksopenlevel=2
}
% Load the references.bib file
% \bibliography{references}

\begin{document}
% \setlength{\headrulewidth}{0.0pt}

%~~ TitlePage ~~%                               %~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~%
\thispagestyle{empty}
\pagenumbering{arabic} %Normal numbers

\begin{center}

{\Large Design Suggestion}\\[0.2cm]
{\itshape for a}\\[0.2cm]
{\Large GPU Based Sonar Template Simulator}\\[2cm]
{\itshape by}\\[0.2cm]
\begin{minipage}{0.35\textwidth}\small\centering
Jo Inge \textsc{Buskenes}\\
\textnormal{\color{SeaGreen}joibu (at) ifi.uio.no}
\end{minipage}
\begin{minipage}{0.35\textwidth}\small\centering
Jon Petter \textsc{Åsen}\\
\textnormal{\color{SeaGreen}jpaasen (at) ifi.uio.no}
\end{minipage}\\[2cm]
{\itshape on the behalf of}\\[0.2cm]
\begin{minipage}[t]{0.49\textwidth}\centering
  \centering\includegraphics[height=3cm]{gfx/FFILogo.png}\\
  \textsc{The Norwegian Defence Research Establishment (FFI)}
\end{minipage}\\[0.5cm]
\begin{minipage}[t]{0.4\textwidth}\small\centering
Øivind \textsc{Midtgård}\\
\textnormal{\color{SeaGreen}oivind.midtgaard (at) ffi.no}
\end{minipage}
\begin{minipage}[t]{0.4\textwidth}\small\centering
Herman \textsc{Midelfart}\\
\textnormal{\color{SeaGreen}herman.midelfart (at) ffi.no}
\end{minipage}\\[0.3cm]
\begin{minipage}{0.32\textwidth}\small\centering
Roy Edgar \textsc{Hansen}\\
\textnormal{\color{SeaGreen}roy-edgar.hansen (at) ffi.no}
\end{minipage}
\vfill
\small
\today
\end{center}

\clearpage
\pagestyle{fancy}
%~~ Sections ~~%                                %~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~%

\section{Introduction}

Template matching is a common technique used when classifying objects in SAS images. The principle is to isolate an image segment containing an object of interest, correlate it with a set of template images, and assign it to the class of the template yielding the highest correlation coefficient. The challenge is to come up with a suitable set of template images, considering the vast span of possible seafloor and object characteristics.

FFI target this challenge by using a sonar simulator that can take a seafloor model or data as input, in which various target objects are postitioned in various ways. For each object configuration the scene is rendered and the result added to the template set. But while producing templates on-the-fly is a very flexible solution, it is also computationally demanding.

This is why we are investiging the feasibility of using a GPU to speed up the simulations.

\begin{figure}[b]
\graphicsAI[drawing,width=\linewidth]{gfx/specs.svg}
\caption{Simulator specifications.}\label{specs}
\end{figure}

\section{Specification}

Most of the specifications received from FFI are illustrated in Figure \ref{specs}.

\section{Design Suggestion}

The way we see it, there are two possible approaches to implementing the mentioned simulator (figure \ref{implementation}). One can either

\begin{figure}[b]
\begin{narrow}{-2.5cm}{-2.5cm}
\graphicsAI[drawing,width=\linewidth]{gfx/simulator.svg}
\end{narrow}
\caption{Two possible implementation strategies.}\label{implementation}
\end{figure}

\begin{itemize0}
\item use OpenGL to produce depth-maps of the scene, and CUDA to illuminate it as a sonar would, or
\item use OpenGL to render the full scene with lighting.
\end{itemize0}

In either case the simulator will take a seafloor and target model as input, and generate a simulated sonar image. To offload the CPU further, the GPU can then proceed to compute the correlation between the template and the isolated image segment, and an image mask. This should be fairly easy to implement, will compute efficiently on the GPU, and allows us to avoid sending each template back to the CPU over the limited bandwidth CPU$\leftrightarrow$GPU bus. Note that these steps will have to be carried out in CUDA regardless of the approach selected.

\subsection{Using OpenGL only for the simulations?}

When we wish to set up a 3D scene, move objects and cameras around, and render it using a GPU, it is better to use e.g. OpenGL than CUDA. OpenGL was made for this specific task, is cross-platform, feature-rich and well-matured. By looking at the graphics quality and frame-rates displayed by modern OpenGL-based games it should also be fairly straight-forward to get the performance we need out of this library.

Our original idea was to use OpenGL to produce depth-maps of the object and (optionally) the seafloor, and then use CUDA to compute how a sonar array would illuminate this scene. However, if we can treat sound as ``light'', we believe it to be better to use OpenGL simulate the entire scene and generate the required 2D views and perspectives of it. The advantage of doing ``everything'' in OpenGL is pretty obvious, we may then support whatever OpenGL supports with little extra effort. The gaming industry demands speed, so such a simulator should also be very fast.

We have compared the two methods in Table \ref{comparison}.

\begin{table}[H]\centering
\begin{tabular}[c]{l l l}\hline
\rowcolor{tabBlue}                  & CUDA/OpenGL    & OpenGL only \\\hline
Development time (estimate)         & 6 weeks        & 4 weeks     \\
Development complexity              & Moderate       & Low \\
Amount of code                      & More           & Less \\
Maintainability                     & Moderate       & Moderate \\
Code readability                    & Moderate       & Moderate \\
Signal model                        & Arbitrary      & Light rays  \\
Sound$\rightarrow$light calibration & Not necessary  & Necessary(?) \\
Multipath support                   & Possible       & Possible \\
Speckle support                     & Possible       & Hard \\
Diffraction support                 & Possible       & Hard \\
ATI/nVidia support                  & nVidia only    & Both \\
\end{tabular}
\vspace{5pt}\caption{Comparison of implementation approaches}\label{comparison}
\end{table}

\section{Final thoughts}

We propose using OpenGL to simulate the scene, and model the array as lightsources. This will make it unsuitable for simulating speckle and diffraction effects, but a ``simple'' simulator should probably not take such effects into account anyhow. Penetrable objects can be simulated in either case, although we are not sure whether the

\end{document}

### 2012-06-01 HRUIonGPU/jrnl_article.tex

@@
 The basic idea of an adaptive beamformer is to calculate the weights on-the-fly, based on the data at hand, and on a selected optimization criteria in such a way that we minimize interference impinging from other directions than the point of focus.
-For a complex data vector $\vec{x}[n] \in \mathbb{C}^M$, we now have data-dependent weights, $\vec{w}(\vec{x}[n])$, that when applied on $\vec{x}[n]$ is the best approximation to achieving the selected criteria.
+For a complex data vector $\vec{x}[n] \in \mathbb{C}^M$, we have for an adaptive beamformer a data-dependent set of weights, $\vec{w}(\vec{x}[n])$, that when applied on $\vec{x}[n]$ is the best approximation to achieving the selected criteria.
 With the Capon beamformer, adaptive weights are found by solving the following minimization problem \cite{Capon1969}:
@@
 \begin{align}
-\mat{\hat{R}} = \mat{\hat{R}} + \epsilon\mat{I}
+\mat{\hat{R}} = \mat{\hat{R}} + \epsilon\mat{I}.
 \end{align}
@@
 Each thread therefore has to be as light on memory as possible and has to perform enough instructions per used byte in order to hide memory latency.
-This pose a real challenge as the target problem has to map to the parallel framework both on a grid, block and memory consumption level in order to benefits from parallel acceleration.
+This pose a real challenge as the target problem then has to map to the parallel framework both on a grid, block and memory consumption level in order to benefits from parallel acceleration.
 
@@
 The computational complexity of these three steps, with $L = M/2$, is $O(L^3)$, $O(L^3)$, and $O(L)$ respectively.
-For larger $L$'s, the first two steps quickly becomes unmanageable, and to achieve real time performance it is therefore important to find novel solutions to these steps.
+For larger $L$'s, the first two steps quickly becomes unmanageable, and real time performance can only be achieved by finding novel solutions to these two steps.
 

### 2012-06-01 FFI-Simulator/ffi-simulator.tex
added some text in ffi designe doc

@@
 
-\subsection{Using OpenGL only for the simulations?}
+In addition we see that each approach shares the same initial steps.
+The risk of selecting a OpenGL-only implementation is therefore reduced, since a CUDA implementation for the middle step can be selected based on experience with the initial steps in OpenGL.
+
+\subsection{Using only OpenGL for the simulator?}
 
@@
 The gaming industry demands speed, so such a simulator should also be very fast.
+A CUDA implementation will benefit from more flexibility when it comes to simulating sound propagation in a non-ideal medium, and translucent objects.
+With more complex models both the code complexity and developing time will naturally increase.
+However, we expect that such features will be more realizable in CUDA than OpenGL.
 
@@
 We propose using OpenGL to simulate the scene, and model the array as lightsources.
-This will make it unsuitable for simulating speckle and diffraction effects, but a ``simple'' simulator should probably not take such effects into account anyhow.
-Penetrable objects can be simulated in either case, although we are not sure whether the
+This will make it unsuitable for simulating speckle and diffraction effects, but a "simple'' simulator should probably not take such effects into account anyhow.
+Penetrable objects can be simulated in either case, although we are not sure whether the OpenGL framework can support this with ease inside its framework limits.
+An OpenGL implementation will be highly portable, and believed to support the functionality need to generate a good enough template image.
 

### 2012-06-01 FFI-Simulator/ffi-simulator.tex
la2pdf commit - 2012:06:01	14:29:30	(Fri 1.Jun)

--- 2012-06-01 FFI-Simulator/ffi-simulator.tex
@@
 
-\subsection{Using OpenGL only for the simulations?}
+In addition we see that each approach shares the same initial steps.
+The risk of choosing an OpenGL-only implementation is therefore reduced, since we may easily switch to the CUDA implementation if an OpenGL prototype fails to produce the desired results.
 
-When we wish to set up a 3D scene, move objects and cameras around, and render it using a GPU, it is better to use e.g.
-OpenGL than CUDA.
-OpenGL was made for this specific task, is cross-platform, feature-rich and well-matured.
-By looking at the graphics quality and frame-rates displayed by modern OpenGL-based games it should also be fairly straight-forward to get the performance we need out of this library.
+\subsection{Using only OpenGL for the simulator?}
 
-Our original idea was to use OpenGL to produce depth-maps of the object and (optionally) the seafloor, and then use CUDA to compute how a sonar array would illuminate this scene.
-However, if we can treat sound as ``light'', we believe it to be better to use OpenGL simulate the entire scene and generate the required 2D views and perspectives of it.
-The advantage of doing ``everything'' in OpenGL is pretty obvious, we may then support whatever OpenGL supports with little extra effort.
-The gaming industry demands speed, so such a simulator should also be very fast.
-
-We have compared the two methods in Table \ref{comparison}.
-
-\begin{table}[H]\centering
+\begin{table}[b]\centering
 \begin{tabular}[c]{l l l}\hline
 \rowcolor{tabBlue}                  & CUDA/OpenGL    & OpenGL only \\\hline
-Development time (estimate)         & 6 weeks        & 4 weeks     \\
+Development time (per person)       & 6 weeks        & 4 weeks     \\
 Development complexity              & Moderate       & Low \\
@@
 Diffraction support                 & Possible       & Hard \\
+Complex image                       & Easy           & Hard \\
 ATI/nVidia support                  & nVidia only    & Both \\
@@
 
+When we wish to set up a 3D scene, move objects and cameras around, and render it using a GPU, it is better to use e.g.
+OpenGL than CUDA.
+OpenGL was made for this specific task, is cross-platform, feature-rich and well-matured.
+By looking at the graphics quality and frame-rates displayed by modern OpenGL-based games it should also be fairly straight-forward to get the performance we need out of this library.
+
+Our original idea was to use OpenGL to produce depth-maps of the object and (optionally) the seafloor, and then use CUDA to compute how a sonar array would illuminate this scene.
+However, since were only interested in pixel intensities, and sound may be treated as ``light'' with reasonable accuracy, we believe that OpenGL can be used to simulate the entire scene.
+This way we may support whatever OpenGL supports with little extra effort.
+The gaming industry demands speed, so such a simulator should also be very fast.
+
+The main advantage of the CUDA implementation lies in its flexibility.
+If desired, it can be modified to handle sound propagation in non-ideal mediums, penetrable objects, signal multipath, diffraction and phase-dependent effects such as speckle.
+However, this will also lead to increased code complexity and development time, and  will naturally increase.
+However, we expect that such features will be more realizable in CUDA than OpenGL.
+
+We have compared the two methods in Table \ref{comparison}.
+
 \section{Final thoughts}
@@
 We propose using OpenGL to simulate the scene, and model the array as lightsources.
-This will make it unsuitable for simulating speckle and diffraction effects, but a ``simple'' simulator should probably not take such effects into account anyhow.
-Penetrable objects can be simulated in either case, although we are not sure whether the
+This will make it unsuitable for simulating speckle and diffraction effects, but a "simple'' simulator should probably not take such effects into account anyhow.
+Penetrable objects can be simulated in either case, although we are not sure whether the OpenGL framework can support this with ease.
+An OpenGL implementation will be highly portable, and believed to support the functionality needed to generate a good enough template image.
+
+We aim to start development in July.
+The goal is to have an OpenGL prototype ready within a couple weeks, and from there decide which approach to go for.
+We propose monthly progress reports and meetings to discuss further
+development.
 

### 2012-06-04 HRUIonGPU/jrnl_article.tex
done some writing on tuffc article

@@
 Although the method introduce some interesting trade-offs in addition to just improved resolution, like maintaining image resolution combined with higher framerates, smaller probes, or deeper penetration \cite{Synnevag2009}, widespread use has been limited by the tremendous computational complexity associated with the method \cite{So2011}.
-In this paper we introduce the first implementation of the Capon beamformer exhibiting real-time performance when applied on typical cardiac data sets.
+In this paper we introduce the first implementation of the Capon beamformer exhibiting real-time performance when applied on a typical cardiac data set.
 To accomplish this task we make use of the massive parallel processing power found in modern graphics processing units (GPU's), combined with novel methods to reduce the computational complexity as the problem size grows.
-Chen et al.
+Earlier attempts on real time Capon beamforming have been focused around radar and passive sonar using intricate systolic array processors \cite{McWhirter1989, Moonen1993, Sinha2002}.
+A systolic array processor is custom made hardware, and even if the computational power might be higher for selected problems compared with a GPU, the cost related to development and flexibility is much higher.
+For medical ultrasound, Chen et al.
 \cite{Chen2011a, Chen2011} have previously implemented the Capon beamformer on a GPU, however their implementation was based on real data, where the Capon beamformer assumes complex.
@@
 It is therefore crucial to use complex data within the Capon beamformer framework, both for reducing the amount of data per image, and to allow for asymmetric windows.
+All results in this paper is reported with use of critically sampled complex data, both laterally and in range.
 
@@
 \end{align}
-In conventional beamforming, the weight $w_m \in \mathbb{R}$ is usually applied to trade resolution for a better signal-to-noise ratio (SNR).
+In conventional beamforming, the weighting $\vec{w} \in \mathbb{R}$ is usually applied to trade resolution for a better signal-to-noise ratio (SNR).
 Hence, by increasing the main lobe width of the system while gaining lower side lobe levels.
@@
 \end{align}
-where $\mat{R} = E\{\vec{x}[n]\vec{x}[n]^H\}$ is the spatial covariance matrix, and $\vec{a}$ is the far-field steering direction.
+where $\mat{R} = E\{\vec{x}[n]\vec{x}[n]^H\}$ is the spatial covariance matrix, and $\vec{a} = \left[ \begin{array}{cccc} e^{-jk_{\tilde{x}}\tilde{x}_0}& e^{-jk_{\tilde{x}}\tilde{x}_1}& ...& e^{-jk_{\tilde{x}}\tilde{x}_{M-1}}\end{array}\right] ^T$ is a far-field steering vector towards angle $\theta = \arcsin(\frac{\lambda k_{\tilde{x}}}{2\pi})$.
 Thus, we minimize the output power, while maintaining unit gain in the look direction.
-The constraint in Eq.
-\ref{eq:constraint} is there to avoid the trivial solution $\vec{w} = 0$, and a distorted respons in the direction of $\vec{a}$.
-The solution to the minimization problem in Eq.
-\ref{eq:minz} and \ref{eq:constraint} is
+The constraint in (\ref{eq:constraint}) is there to avoid the trivial solution $\vec{w} = 0$, and a distorted respons in the direction of $\vec{a}$.
+The solution to the minimization problem in (\ref{eq:minz}) and (\ref{eq:constraint}) is
 \begin{align}\label{eq:w}
@@
 \end{align}
-where $\vec{a} = \vec{1}$ when $\vec{x}$ is pre-delayed to broadside.
+where $\vec{a} = [1_0 1_1... 1_{M-1}]^T = \vec{1}$ when $\vec{x}$ is pre-delayed to broadside.
 
@@
 \begin{align}
-\mat{\hat{R}} = \frac{1}{NK}\sum_{n'=n-N_{avg}}^{n+N_{avg}} \sum_{l=0}^{K} \vec{x}_l[n']\vec{x}_l[n']^H,\label{eq:R}
+\mat{\hat{R}}[n] = \frac{1}{NK}\sum_{n'=n-N_{avg}}^{n+N_{avg}} \sum_{l=0}^{K-1} \vec{x}_l[n']\vec{x}_l[n']^H,\label{eq:R}
 \end{align}
@@
 \begin{align}
-\epsilon = d*tr\{\mat{\hat{R}}\},
+\epsilon &= d*tr\{\mat{\hat{R}}\}, \\
+d &= 1/L
 \end{align}
-has previously shown good results in the literature \cite{Synnevag2007}.
+introduced by Synnev\aa{}g et al. \cite{Synnevag2007} has been applied in much of the resent literature on Capon beamforming for medical ultrasound imaging \cite{Nilsen2009, Wang2009, Mehdizadeh2012}.
 
@@
 \subsection{GPU Compute Model}
-In this section we briefly discuss the principle behind GPU computing.
+In this section we briefly discuss the principle behind GPU computing, laying the basis for further discussion on parallel implementation strategies.
 CUDA by Nvidia has been our choice of GPU developer environment \cite{Nvidia2011}.
-However, even though CUDA has been used to implement the Capon beamformer, the challenges and proposed methods described in this paper applies for other massive parallel implementation environments like e.g.
-OpenCL as well.
+However, even though CUDA has been used to implement the Capon beamformer, the challenges and proposed methods described in this paper applies for other massive parallel implementation environments as well, like e.g.
+OpenCL.
 
@@
 Each thread therefore has to be as light on memory as possible and has to perform enough instructions per used byte in order to hide memory latency.
-This pose a real challenge as the target problem then has to map to the parallel framework both on a grid, block and memory consumption level in order to benefits from parallel acceleration.
+This pose a real challenge as the target problem has to map to the parallel framework both on a grid, block and memory consumption level in order to benefits from parallel acceleration.
 
@@
 However, the calculation of Capon weights is well suited for a GPU implementation, since the calculations are both independent across pixels, and partially intra-pixel.
-On a top level it is trivial to recognise that one grid cell of threads should process one pixel in the final image.
+Based on memory dependencies it is trivial to recognise that one block of threads should process at least one pixel in the final image.
 Continuing, on a pixel-level, the algorithm can be broken down into three main steps that to some extent can be further divided into a small set of parallel processes.
-For each data vector $\vec{x}$ in the image:
+That is, for each data vector $\vec{x}$ in the image:
 \begin{enumerate}
-\item Build the sample covariance matrix $\mat{\hat{R}}$
+\item Calculate the sample covariance matrix $\mat{\hat{R}}$ as in (\ref{eq:R})
 \item Solve the linear system of equations $\mat{\hat{R}}\vec{b} =\vec{ 1}$
-\item Calculate $\vec{w} = \vec{b}/|\vec{b}|$ and the amplitude $\vec{w}^H\vec{x}$
+\item Calculate $\vec{w} = \vec{b}/\vec{1}^H\vec{b}$ and the amplitude $\vec{w}^H\vec{x}$
 \end{enumerate}
 The computational complexity of these three steps, with $L = M/2$, is $O(L^3)$, $O(L^3)$, and $O(L)$ respectively.
-For larger $L$'s, the first two steps quickly becomes unmanageable, and real time performance can only be achieved by finding novel solutions to these two steps.
+For larger $L$'s, the first two steps quickly becomes unmanageable, and real time performance can only be achieved by exploring novel solutions to reduce complexity.
 
 \subsection{Building Multiple Sample Covariance Matrices}
-The straight forward sum in Eq.
-\ref{eq:R} has complexity $O(L^2KN) = O(L^3)$ if $L$ is large.
-Note that N typically is small compared to L and M.
-For smaller $L$ the construction of $\mat{\hat{R}}$ is therefore dominated by $M$.
-The quadratic complexity for large $L$ can be improved by exploiting the overlap from element to element on the diagonals of $\mat{\hat{R}}$ (ref figure).
-This overlap is present both across subarrays and time.
+The straight forward sum in (\ref{eq:R}) has complexity $O(L^2KN) = O(L^3)$ if $L$ is a small fraction of $M$.
+This is actually the same as for inverting a $L \times L$ matrix.
+However, little attention has been devoted to this step in the literature until now.
+Note that for a typical array configuration, N is usually small compared to L and M.
+Hence, for smaller $L$ the construction of $\mat{\hat{R}}$ is dominated by $M$, since $K = M - L + 1$.
+
+The quadratic complexity with $L$ and linear complexity with $N$ can be reduced by realizing that the calculation in (\ref{eq:R}) overlap both across subarrays and in time.
+The $(i,j)$ element of $\mat{\hat{R}}[n]$ can be calculated from element $(i-1, j-1), \text{for } i,j > 0$ by subtracting $x_{i-1}x_{j-1}^*$ and adding $x_{i+K-1}x_{j+K-1}^*$.
+The same is true in time where $\mat{\hat{R}}[n]$ can be calculated from $\mat{\hat{R}}[n-1]$ by subtracting $\mat{\hat{R}}[n-N_{avg}-1]$ and adding $\mat{\hat{R}}[n+N_{avg}]$.
+Both sliding across subarrays and in time can be realized simultaneously if we hold $\vec{x}\vec{x}^H[n]$ averaged over $N$ time samples in memory and calculate $\mat{\hat{R}}$ averaged over subarrays from that buffer (figure).
+However, since $N$ typically is small, and we have a limited amount of memory per thread on the GPU, we have chosen to focus on sliding across subarrays only.
+In that case, the complexity of calculating $\mat{\hat{R}}$ is reduced to $O(L^2)$.
+
+Taking symmetry into account, only the upper half of $\mat{\hat{R}}$ needs to be calculated.
+However, the way we have decided to organize the GPU compute threads per $\mat{\hat{R}}$, the lower triangle can be calculated with little overhead on the GPU.
+This also means that we do not relay on a symmetric solver to solve $\mat{\hat{R}}\vec{b} = \vec{1}$, even if that could have saved some instructions.
+The memory layout for the upper triangle only is also cumbersome to work with.
+
+For subarray averaging, the first row and column of $\mat{\hat{R}}$ has to be fully calculated, and for time averaging $\mat{\hat{R}}[0]$ has to be fully calculated.
+
+Since $L$ usually is larger than $N$, we have
 
@@
 al concluded that as little as three beams around broadside where adequate to produce results comparable to applying Capon beamforming in element space.
-The beamspace version of Capon is derived by replacing the covariance matrix in Eq.
-\ref{eq:w} with the beamspace covariance matrix $\mat{R}_{BS} = \vec{x}_{BS}\vec{x}_{BS}^H$.
+The beamspace version of Capon is derived by replacing the covariance matrix in (\ref{eq:w}) with the beamspace covariance matrix $\mat{R}_{BS} = \vec{x}_{BS}\vec{x}_{BS}^H$, where $\mat{B} \in \mathbb{C}^{N_b,M}$, and $N_b$ is the number of selected beams.
+Note that when subarray averaging is applied, the transformation must be done per subarray, yielding subarrays that no longer overlap in space.
+Hence, the amount of data is change from $M$ to $KN_b$ per sample.
 
@@
 \bibliography{mybib} % link to mendeley organized bibtex-file
-%
+%\bibliography{IEEEabrv,mybib} 
 % <OR> manually copy in the resultant .bbl file

### 2012-06-05 HRUIonGPU/jrnl_article.tex
done some writing on the tuffc article

@@
 A weighting proportional to the output power is often used to reduce the need of parameter adjustments, and the trace of $\mat{\hat{R}}$, 
-\begin{align}
+\begin{align}\label{eq:diag}
 \epsilon &= d*tr\{\mat{\hat{R}}\}, \\
@@
 
-The CUDA compute model is based on execution of a kernel-function across a grid of compute threads. Each position in the grid holds a block of threads.
+The CUDA compute model is based on execution of a kernel-function across a grid of compute threads.
+Each position in the grid holds a block of threads, with maximal size of 1024 threads, that can be easily synchronized if needed.
 %Each block is further divided into groups of 32 threads, known as a warp. The maximum number of threads inside a block is restricted to 512, hence 16 warps. 
@@
 Each thread therefore has to be as light on memory as possible and has to perform enough instructions per used byte in order to hide memory latency.
-This pose a real challenge as the target problem has to map to the parallel framework both on a grid, block and memory consumption level in order to benefits from parallel acceleration.
+Global read and writes on the GPU, in addition to CPU-GPU transfers have high penalty, and should be minimized as much as possible.
+All of these constraints combined pose a real challenge as the target problem has to map on to the parallel framework both on a grid, block and memory consumption level in order to benefits from parallel acceleration.
 
@@
 
-\subsection{Building Multiple Sample Covariance Matrices}
+\subsection{Calculating Multiple Sample Covariance Matrices}
 The straight forward sum in (\ref{eq:R}) has complexity $O(L^2KN) = O(L^3)$ if $L$ is a small fraction of $M$.
@@
 Hence, for smaller $L$ the construction of $\mat{\hat{R}}$ is dominated by $M$, since $K = M - L + 1$.
+From (\ref{eq:R}) it is obvious that calculation of one element in $\mat{\hat{R}}$ is independent of all the others.
+A natural level of granularity would then be to assign one thread to each element of $\mat{\hat{R}}$.
+This will minimize global reads and writes needed per thread, but unfortunately the maximum number of threads per block puts an upper limit on $L$, $L_{max}=32$, if we want to avoid block-to-block synchronization.
 
@@
 In that case, the complexity of calculating $\mat{\hat{R}}$ is reduced to $O(L^2)$.
+Another aspect is that sliding will break independence in the sliding dimension.
+With our choice of calculating $\mat{\hat{R}}$, calculations are still independent across time, but not across subarrays.
+We therefore have selected to use on thread per column in $\mat{\hat{R}}$, and for small $L$'s several covariance matrices are calculated per thread block.
+Maximum supported $L$ is with this approach increased to 512, however for small $L$ high $M$ configurations, memory consumption per thread will be to high to achieve full occupancy of the GPU.
 
@@
 This also means that we do not relay on a symmetric solver to solve $\mat{\hat{R}}\vec{b} = \vec{1}$, even if that could have saved some instructions.
-The memory layout for the upper triangle only is also cumbersome to work with.
+The memory layout for the upper triangle only is also believed to be cumbersome.
 
-For subarray averaging, the first row and column of $\mat{\hat{R}}$ has to be fully calculated, and for time averaging $\mat{\hat{R}}[0]$ has to be fully calculated.
+Diagonal loading as a constant value can be added with marginal cost, but the adaptive diagonal loading factor in (\ref{eq:diag}) is a different story.
+In this case, the trace of $\mat{\hat{R}}$ has to be known before diagonal elements are written back to global memory.
+Since one row for all $\mat{\hat{R}}$'s is written to global memory per kernel iteration, the solution was to hold the diagonal elements in a local buffer until the trace was accumulated, and then do scattered writes of the diagonal elements to global memory.
 
-Since $L$ usually is larger than $N$, we have
+%For subarray averaging, the first row and column of $\mat{\hat{R}}$ has to be fully calculated, and for time averaging $\mat{\hat{R}}[0]$ has to be fully calculated.
+
+%Since $L$ usually is larger than $N$, we have
 
@@
 \subsection{Solving Multiple Small Linear Systems}
-The focus has mostly been on large systems in the literature, and GPU libraries usually lack a comprehensive collection of batched solvers.
-The reason for this is two-folded, first the GPU needs to solve thousand of small systems in order to beat the low memory latency of the CPU, second there is a range of system dimensions that does not fit well with the GPU architecture, where only a small amount of near-core memory is available.
-Because of this, it has been proven hard to optimize a manageable set of solvers that provides speedups compared with the CPU in all cases.
+When it comes to solving systems of linear equations on the GPU the focus has mostly been on large systems in the literature, and GPU libraries usually lack a comprehensive collection of batched solvers.
+The reason for this is two-folded, first the GPU needs to solve thousand of small systems in order to beat the low memory latency of the CPU, second there is a range of system dimensions that does not fit well with the GPU architecture, where only a small amount of near-core memory registers are available.
+Because of this, it has been proven hard to optimize a manageable set of solvers that provides speedups compared with the CPU in every cases.
 
-For the results presented in this paper, we have used an unreleased GPU implementation of Gauss Jordan (GJ) elimination (by NVIDIA) to solve $\mat{R}\vec{b} = \vec{1}$, where $\vec{b} = \mat{R}^{-1}\vec{1}$.
-Can include details on GPU implementation of $\mat{U}^H\mat{D}\mat{U}$, but this has not proved to be faster than NVIDIA's GJ implementation.
+For the results presented in this paper, we have used an unreleased GPU implementation of Gauss Jordan (GJ) elimination (by NVIDIA) to solve $\mat{R}\vec{b} = \vec{1}$, where $\vec{b} = \mat{R}^{-1}\vec{1}$, but we will go through an implementation of an $\mat{U}^H\mat{D}\mat{U}$-based solver.
+...
+
+%Can include details on GPU implementation of $\mat{U}^H\mat{D}\mat{U}$, but this has not proved to be faster than NVIDIA's GJ implementation.
 However, the complexity for solving with $\mat{U}^H\mat{D}\mat{U}$ decomposition is supposed to be $1/2$ of GJ.
 
-In the final journal article we could discuss Conjugated gradient and Woodbury and the benefits they bring.
+%In the final journal article we could discuss Conjugated gradient and Woodbury and the benefits they bring.
+Other solvers that provides different benefits includes Conjugated gradient and the use of the Woodbury matrix identity.
 
 \subsection{Compute Beamformer Output}
+One thread per weight ($L$ threads). Let one thread sum the solution vector $\vec{b}$, or let several threads cooperate to form the sum using a binary reduction sum.
+Let one thread read one element of $\vec{b}$, scale it with the inverse sum, and optionally write it back to global memory.
+The final output can be formed by either reduce the data down to length $L$, or increase the weight vector to length $M$.
+Since we have decided to have $L$ threads per output, the data is reduced to length $L$ by calculating $x_{r} = \sum_{k=r}^{r+K-1}x_{k}$ for $r \in [0, L-1]$.
+Each thread now computes $z_r = w_r^*x_r$, and finally one thread computes and outputs $z = \sum_{r=0}^{L-1}$ to global memory.
 

### 2012-06-05 HRUIonGPU/jrnl_article.tex
tuffc capon on speed

@@
 Note that for a typical array configuration, N is usually small compared to L and M.
-Hence, for smaller $L$ the construction of $\mat{\hat{R}}$ is dominated by $M$, since $K = M - L + 1$.
+Hence, for a small $L$ the construction of $\mat{\hat{R}}$ is dominated by $M$, since $K = M - L + 1$.
 From (\ref{eq:R}) it is obvious that calculation of one element in $\mat{\hat{R}}$ is independent of all the others.
@@
 
-For the results presented in this paper, we have used an unreleased GPU implementation of Gauss Jordan (GJ) elimination (by NVIDIA) to solve $\mat{R}\vec{b} = \vec{1}$, where $\vec{b} = \mat{R}^{-1}\vec{1}$, but we will go through an implementation of an $\mat{U}^H\mat{D}\mat{U}$-based solver.
+For the results presented in this paper, we have used an unreleased GPU implementation of Gauss Jordan (GJ) elimination (by NVIDIA) to solve $\mat{R}\vec{b} = \vec{1}$, where $\vec{b} = \mat{R}^{-1}\vec{1}$.
+However we will go through an implementation of an $\mat{U}^H\mat{D}\mat{U}$-decomposition-based solver to point out challenges when implementing batched solving of small linear systems on the GPU.
 ...
+(Unsure if this should be included).
 
@@
 However, the complexity for solving with $\mat{U}^H\mat{D}\mat{U}$ decomposition is supposed to be $1/2$ of GJ.
+
+For the smallest systems a direct solver using one thread per system is preferable. For this purpose we have implemented Cramer's rule for $3 \times 3$ (, $2 \times 2$?  and $4 \times 4$?) matrices.
 
@@
 \subsection{Compute Beamformer Output}
-One thread per weight ($L$ threads). Let one thread sum the solution vector $\vec{b}$, or let several threads cooperate to form the sum using a binary reduction sum.
+(Still just notes) One thread per weight ($L$ threads). Let one thread sum the solution vector $\vec{b}$, or let several threads cooperate to form the sum using a binary reduction sum.
 Let one thread read one element of $\vec{b}$, scale it with the inverse sum, and optionally write it back to global memory.
@@
 Each thread now computes $z_r = w_r^*x_r$, and finally one thread computes and outputs $z = \sum_{r=0}^{L-1}$ to global memory.
+The last sum can also be performed using a binary reduction sum across several threads for increased speed.
 
@@
 Since subarray averaging has been applied to decorrelated the signal, we might loose this effect when reestablishing all degrees of freedom.
-The number of iterations should therefore be restricted to $K/2$ to equal our standard choice of $L = M/2$ when it comes to degrees of freedom.
+The number of iterations should therefore be restricted to $K/2$ to equal the normal choice of $L = M/2$ when it comes to degrees of freedom.
 
@@
 Show beampatterns for the two compared with the full method. Show executing times for the two methods compared with the full method on the GPU.
+
+Describe the base band (IQ-demodulation) process that the data has been undergoing.
 

### 2012-06-07 FFI-Simulator/ffi-simulator.tex
...

--- 2012-06-01 FFI-Simulator/ffi-simulator.tex
@@
 
+\newpage
+
+\section{Ideas for paper}
+
+\begin{itemize}
+\item With effects such as layover, shadows, limited resolution (detail retention capability), and complex object and surface shapes, computing the actual scenario based on the sonar image is hard.
+\item Better to perform fast, recursive simulations instead?
+\item Adaptive images in a video makes it much easier to understand what is going on. Need much better erfora
+\item Litits: massive storage, fixed templates, amplitude/noise sensitive
+\item Alternatives: Pattern recognition (hard to extract decent features, selection, fixed templates) and model-based (massive computational requirements)
+\item Why models? Hard to get accurate models from limted amount dat, targets not always availagme, variations in target/terrain.
+\item CAD models. Sufficient complexity? Materials? Transparency, characterisation, multilauers.
+\item Accuracy vs. comp. speed.
+\end{itemize}
+
 \end{document}

### 2012-06-07 FFI-Simulator/ffi-simulator.tex
...

--- 2012-06-01 FFI-Simulator/ffi-simulator.tex
@@
 
-\begin{table}[b]\centering
+When we wish to set up a 3D scene, move objects and cameras around, and render it using a GPU, it is better to use e.g.
+OpenGL than CUDA.
+OpenGL was made for this specific task, is cross-platform, feature-rich and well-matured.
+By looking at the graphics quality and frame-rates displayed by modern OpenGL-based games it should also be fairly straight-forward to get the performance we need out of this library.
+
+Our original idea was to use OpenGL to produce depth-maps of the object and (optionally) the seafloor, and then use CUDA to compute how a sonar array would illuminate this scene.
+However, since were only interested in pixel intensities, and sound may be treated as ``light'' with reasonable accuracy, we believe that OpenGL can be used to simulate the entire scene.
+This way we may support whatever OpenGL supports with little extra effort.
+The gaming industry demands speed, so such a simulator should also be very fast.
+
+The main advantage of the CUDA implementation lies in its flexibility.
+If desired, it can be modified to handle sound propagation in non-ideal mediums, penetrable objects, signal multipath, diffraction and phase-dependent effects such as speckle.
+However, this will also lead to increased code complexity and development time, and  will naturally increase.
+However, we expect that such features will be more realizable in CUDA than OpenGL.
+
+We have compared the two methods in Table \ref{comparison}.
+
+\begin{table}[t]\centering
 \begin{tabular}[c]{l l l}\hline
@@
 
-When we wish to set up a 3D scene, move objects and cameras around, and render it using a GPU, it is better to use e.g.
-OpenGL than CUDA.
-OpenGL was made for this specific task, is cross-platform, feature-rich and well-matured.
-By looking at the graphics quality and frame-rates displayed by modern OpenGL-based games it should also be fairly straight-forward to get the performance we need out of this library.
-
-Our original idea was to use OpenGL to produce depth-maps of the object and (optionally) the seafloor, and then use CUDA to compute how a sonar array would illuminate this scene.
-However, since were only interested in pixel intensities, and sound may be treated as ``light'' with reasonable accuracy, we believe that OpenGL can be used to simulate the entire scene.
-This way we may support whatever OpenGL supports with little extra effort.
-The gaming industry demands speed, so such a simulator should also be very fast.
-
-The main advantage of the CUDA implementation lies in its flexibility.
-If desired, it can be modified to handle sound propagation in non-ideal mediums, penetrable objects, signal multipath, diffraction and phase-dependent effects such as speckle.
-However, this will also lead to increased code complexity and development time, and  will naturally increase.
-However, we expect that such features will be more realizable in CUDA than OpenGL.
-
-We have compared the two methods in Table \ref{comparison}.
-
 \section{Final thoughts}

### 2012-06-11 UAM2011/Article/article.tex
More files... Git argued a bit there..

%===============%                               %~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~%
% DOCUMENTCLASS %                                See full option description in "mytemplate.cls"
%===============%                               %~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~%

\documentclass[
   UAM                                          % Document type (non-standard)
 , 12pt                                         % Text size
%  , draft
%  , final                                        % Quality
 , xelatex                                      % Use the XeLaTeX compiler
%  , biblatex                                     % Use 'biblatex' for references (best by far)
 , bibtex                                       % Use 'bibtex' for references (oldschool :)
%  , movie
%  , notodos                                      % Disable todos
 , layout
%  , defaultformat                                % Attempt to use default formatting options
%  , glossary                                     % Use a glossary
]{common/mytemplate}

% \bibliography{references}
\hypersetup{
   bookmarksopen=true
 , bookmarksopenlevel=2
}
% Load the references.bib file
% \bibliography{references}

\newlength\oldparindent
\setlength\oldparindent{\parindent}

\newlength\figwidth
\setlength\figwidth{0.7\linewidth}

% \setlength\floatsep{\baselineskip}
\setcounter{topnumber}{1}
% \setlength\baselineskip{12pt}
% \selectfont
\begin{document}

\pagestyle{plain}
% \layout
% \printinunitsof{cm}
% \prntlen{\baselineskip}
% \the\textwidth
% \the\linewidth
% \prntlen{\textwidth}\\
% \prntlen\linewidth\\
% \prntlen\oddsidemargin\\
% \prntlen\oddsidemargin
% \newpage

\title{A Low Complexity Adaptive Beamformer\\for Active Sonar Imaging}%
%
\author{J.I.Buskenes\firstAddress, A.Austeng\firstAddress, C.-I.C.Nilsen\firstAddress}%
%
\begin{contact}
  \firstAddress Dept. of Informatics, P.Box 1080 Blindern, N-0316 OSLO
\end{contact}%
%
\begin{contact}
Contact author: Jo Inge Buskenes\\
Dept. of Informatics, Univ. of Oslo, P.Box 1080 Blindern, N-0316 OSLO\\
\href{mailto:joibu@ifi.uio.no}{joibu@ifi.uio.no}
\end{contact}%
%
\begin{abstract}
The angular resolution and contrast in active sonar images depend on the beamformer's ability to receive signals from directions of interest, while suppressing noise and interference emanating from other directions. For sonar arrays, this is achieved by applying weights to the array channels.

While classical beamformers use predefined windows, adaptive beamformers estimate the optimal window by analytical evaluation of the data. The minimum variance (MV) beamformer, for instance, calculates the set of weights that minimises the variance of the beamformer's output.

We have implemented a Low Complexity Adaptive (LCA) beamformer, which adaptively selects a window from a predefined set. The set is comprised of windows that are typical solutions found by the Minimum Variance method. The LCA beamformer was tested using simulated and experimental data from the Kongsberg Maritime HISAS 1030 sonar. On a simulated scene with speckle, highlight and shadow, the beamformer offered better lateral edge definition compared to the MV beamformer, and speckle intensity and shape comparable to DAS and MV beamformers. These results were verified by the experimental data.

An attactive trait of the LCA is its low computational complexity; while the MV beamformer is of O($M^3$), the LCA method is of O($MW$), with $M$ being the number of channels and $W$ the number of windows. We made the LCA perform like the MV method using a well designed set of 30 windows. Hence, unless the array is very small, the proposed method will perform like the MV beamformer or better, and at a fraction of the computational cost.
\end{abstract}%
%
\keywords{Beamforming, Capon, Minimum variance, Low Complexity, Sonar}

% \titlepage

\newpage
% \the\lineheight
\section{Introduction}

In active sonar array imaging, an acoustic wave is transmitted and the received echoes are usually recorded using an array of sensors. Each of these may be independently delayed and weighted such that signals emanating from directions of interest are added constructively, while noise and interference from other directions add destructively. Additionally, a window may be applied to the sensor channels to further adjust the arrays spatial response. This process is known as beamforming.% The process of combining the signals from each sensor is known as beamforming.

Traditional beamformers such as Delay-and-Sum (DAS) weight the sensors using a predefined window. However, since the optimal spatial response is likely to change with time, it is often preferred to use beamformers that continously adapt the weights to the dynamic data. This usually involves solving some optimization problem.

The Minimum Variance (MV) beamformer selects the window that minimizes the beamformer's output power~\cite{cap69}. It has recently been investigated for active sonar imaging in e.g.~\cite{saf09}, and in ultrasound imaging in e.g.~\cite{syn07}. We suggest using a Low Complexity Adaptive (LCA) beamformer, which for each pixel applies a set of predefined windows and selects one of them based on the MV criterion. Using the LCA beamformer results in increased robustness and improved edge definition compared to the MV and DAS beamformers.

% When the wavefield is sampled using an array of sensors, the data from each sensor may be 
% 
% Traditional beamformers such as Delay-and-Sum (DAS) apply predefined windows to all incoming data. However, due to the non-stationary nature of sonar data, the optimal window for any given time instant will generally differ from the next. This is where adaptive beamformers thrive, because they compute the optimal window coefficients for the data at each time instant. The choice of optimisation criteria is what mainly differentiates the various adaptive beamformers. The Minimum Variance (MV) beamformer, for instance, selects the set of weights that minimizes the beamformer output power for any given time instant~\cite{cap69}.

\section{Background}

% Adaptive beamformers usually compute the window coefficients by estimating and inverting a spatial covariance matrix. There are two inherent problems to this procedure. First, for improved estimation accuracy the data is either averaged in space or time, and the significance of the covariance matrix can be reduced using a regularisation parameter~\cite{Carl Inge}. These are all attempts to constrain the beamformer to ensure the window coefficience are not over-adapted to the data. Second, inverting the covariance matrix has a computational complexity of O($M^3$)~\cite{Carl Inge}. For larger arrays the computational burden becomes significant, and the question arises whether this processing power can be better utilised.

The MV beamformer computes the window coefficients by estimating and inverting a spatial covariance matrix. For improved estimation accuracy the data is averaged in space or time, or both. In addition, a certain amount of diagonal loading is often added to the covariance matrix before inversion. These actions, while leading to statistical robustness, tend to deteriorate the beamformer's performance. Also, the inversion of the covariance matrix has a computational complexity of O($M^3$), where $M$ is the number of elements. The computational burden makes the MV beamformer for large arrays less attractive, and sometimes even infeasible.

Based on the proposed method by Synnevåg~\cite{syn11}, we have implemented the LCA beamformer that keeps the minimum variance optimisation criterion but reduces the solution space to a discrete set of predefined windows. This reduces the computational complexity to O($MW$), where $W$ is the number of windows in the set. We use the Kaiser window function because it allows us to design a wide range of windows with different mainlobe widths and sidelobe suppression by adjusting the tradeoff parameter $\beta$~\cite{kai66}. In addition we apply steering to each of the windows and constrain the window design to ensure unit gain in the look direction.

The following will show that the proposed method performs similarly to or better than the MV method.

\begin{figure}[tp]\centering%
\subfigure{%
\graphicsAI[width=\linewidth]{gfx/img_das_lca_mv.pdf}}
\subfigure{%
\graphicsAI[width=\linewidth]{gfx/img_das_lca_mv_mean100.pdf}}%
\caption{A simulated circular object buried in speckle is imaged from above.\newline\parbox[t]{1.50cm}{Top:\newline Bottom:}\parbox[t]{6.8cm}{Image of a single speckle realisation.\newline Mean image of 100 speckle realisations.}}\label{speckle}%%
\end{figure}%
\begin{figure}[t]\centering%
\graphicsAI[width=\linewidth]{gfx/img_mean_cut.pdf}%
\caption{Lateral cut in the mean image at range 41.1\,m (highlight) and 44.5\,m (shadow).}\label{cutmean}%
\end{figure}
% \begin{figure}[t]\centering
% % \graphicsAI[drawing,width=\linewidth]{gfx/the_box.svg}\\
% \includegraphics[width=\linewidth]{gfx/selected_windows.pdf}
% \caption{Selected windows}\label{selected_windows}
% \end{figure}
% \begin{figure}[t]\centering
% \includegraphics[width=\linewidth]{gfx/selected_windows_angle.pdf}
% \caption{Selected windows - Colored by angle}\label{selected_windows_angle}
% \includegraphics[width=\linewidth]{gfx/selected_windows_sym_angle.pdf}
% \caption{Selected windows - Colored symmetrically by angle}\label{selected_windows_sym}
% \end{figure}
\begin{figure}[t]\centering
\graphicsAI[width=\linewidth]{gfx/img_holmengraa.pdf}%
\caption{HISAS sidescan sonar (SSS) image of the shipwreck Holmengraa.}\label{holmengraa}%
\end{figure}

% Adaptive beamformers usually compute the window coefficients by estimating and inverting a spatial covariance matrix. There are two inherent problems to this procedure. First, for improved estimation accuracy the data is either averaged in space or time, and the significance of the covariance matrix can be reduced using a regularisation parameter~\cite{Carl Inge}. These are all attempts to constrain the beamformer to ensure the window coefficience are not over-adapted to the data. Second, inverting the covariance matrix has a computational complexity of O($M^3$)~\cite{Carl Inge}. For larger arrays the computational burden becomes significant, and the question arises whether this processing power can be better utilised. 
% 
% Based on the proposed method by Synnevåg~\cite{syn11}, we have implemented a Low Complexity Adaptive (LCA) beamformer that keeps the minimum variance optimisation criteria but reduces the solution space to a discrete set of predefined windows. We use the Kaiser window function because it allows us to design a wide range of windows with different mainlobe widths and sidelobe suppression by adjusting the tradeoff parameter $\beta$. In addition we apply steering $\varphi$ to each of the windows and and constrain the window design to ensure unit gain in the look direction.

\section{Experimental Setup}

For the LCA to perform well it is important to let it have a diverse selection of windows to choose from. This selection is based on observations of the solutions found by the MV beamformer. We obtained good results in our experiments by letting the LCA beamformer select among 5 Kaiser windows with uniformly distributed $\beta$'s in the range $[0.05, 0.5]$ and a rectangular window. Each window was steered in 5 different directions, uniformly distributed within 80\% of the mainlobe width of the rectangular window, which is the most narrow. This adds to a total of 30 windows.

To make the choice of window less susceptible to pixel value uncertainties, the LCA beamformer was set to apply the most frequently selected window in an 11 pixel range region to the center pixel in that region. In physical terms, this is a region of approximately 20\,cm. The MV beamformer estimated the covariance matrix as described in~\cite{syn07} by averaging over 16 subarrays and 11 range pixels, and applying 3\% diagonal loading.

% Furthermore, the algorithm maps well to single-instruction-multiple-data (SIMD) hardware such as graphics processing units (GPUs). We show that even a modest amount of optimisation work has resulted in a factor 10 speed-up of this algorithm when implemented on a GPU as opposed to on a CPU.

% \newpage
\section{Results and Discussion}

We have tested the beamformers using data aquired by the Kongsberg Maritime HISAS1030 sonar~\cite{han09}, and on simulations of the same sonar with 100 realisations of speckle.

The simulations contained a few highlighted cylindric regions with 2\,m diameter and a constant intensity 15.4\,dB over the average speckle level. We focused on an object centered at 41\,m range and -3 degrees azimuth angle. The images produced by the LCA, MV and unweighted DAS for a single speckle realisation and a mean image of all 100 realisations are shown in \Fig{speckle}. Each image was normalised by their respective speckle level, and the dynamic range was clamped at \{-30,~15\}\,dB. The physical extent of the of the simulated object was superimposed for reference. %\ref{saf09}

In \Fig{speckle} we observe that both the adaptive beamformers produce images with a more clearly defined shadow. The edges are sharper and the shadow more distinct. This is because the adaptive beamformers have better sidelobe suppression, and thus allow less signal energy to leak from the speckle region into the shadow region. The same effects can also be observed laterally around the highlight region of the mean image; DAS is more prone to smear energy into the speckle region than the adaptive beamformers are.

\Fig{cutmean} displays two lateral cuts in the mean image, one through the highlight at 41.1\,m range and one through the shadow at 44.5\,m range. Each cut was computed as the mean of a 40\,cm range band centered at the respective ranges. The transition region between highlight and shadow is shortest for the LCA, which is advantageous since it results in a more accurate representation of the object size. The slightly inferior performance of the MV beamformer is due to the subarray averaging, which makes the effective array smaller~\cite{syn07}.

By studying the LCA performance on the 100 speckle realisations, we found the choice of window to be highly dependent on the phenomena being imaged. In speckle regions LCA favored narrow spatial responses with minor steering, while wide and fully steered responses were often selected in shadow regions. There was also a clear distinction between the responses selected in sidelobe regions and in speckle regions.

% \Fig{hist} illustrates how often a particular window is chosen in general. There are 10 $\phi$'s for each of the 10 $\beta$'s. The first 10 values are rectangular windows. Value 5 is a non-steered rectangular window, and we see that this is most frequently selected. At higher steering angles the windows are rarely selected, and thus we might get better results by tightening our steering boundaries. At value 15 we find inverted Kaiser windows with $\beta = 0.05$, and then $\beta$ is increased in increments of 0.505 for every 10th value. At value 65 the $\beta$ value is 2.025, and these and the remaining windows are hardly selected.

A sidescan image of the 1500 dwt oil tanker wreck Holmengraa is shown in \Fig{holmengraa}. It is about 68\,m long and 9\,m wide, and lies on a slanted seabed at 77\,m depth~\cite{holmengraa}. The sidescan image was created using data from the HISAS 1030 sonar, which is rather unsuited for this purpose because of its large opening angle. This, and the fact that the wreck was imaged at a range of about 105\,m makes the image quality poor, but sufficient to compare our beamformers.

In the Holmengraa image we note again that the LCA produces a cleaner shadow and better edge definition. The MV method performed almost identically to LCA in this case, and was omitted.

\section{Conclusion}

The LCA method performs similar to the MV method, but at a fraction of the computational cost. The key to achieve this success lies in the design of the window set. Windows that yield vagely steered narrow responses are preferred in speckle regions, while wide and steered responses are typically preferred in highlight and shadow regions. A window set of 6 different responses each steered in 5 different directions proved sufficient in our experiment to match the performance of the MV beamformer.

The tendency of the LCA to perform similarly to the MV beamformer seems to indicate that a full solution space is rarely needed in real world scenarios.

% Nature, while being stochastic, acts within predictable bounds\todo{hehe, jada, jeg skal fjerne..} that can be well adapted to by using beamformers with fewer degrees of freedom than what the MV method provides.

% While the MV beamformer is expected to take advantage of its full solution space to design a spatial response with zero sensitivity at the interferers angles, while LCA will be limted to the spatial responses that we selected. In short, whenever a scenario comes along that favors a spatial response we did not think of, the MV will triumph. On the other hand, increasing the LCA solution space will at some point leave it with nearly the same freedom as the MV, without suffering from statistical instability.
% \todo[inline]{mer innhold..}

% While e found that the tendency of the LCA to perform similarly to the MV beamformer is consistent in most scenarios, except when it is of essence that strong interferers close to object being imaged are present\todo{referanse?}. In such cases the MV beamformer is expected to take advantage of its full solution space to design a spatial response with zero sensitivity at the interferers angles, while LCA will be limted to the spatial responses that we selected. In short, whenever a scenario comes along that favors a spatial response we did not think of, the MV will triumph. On the other hand, increasing the LCA solution space will at some point leave it with nearly the same freedom as the MV, without suffering from statistical instability.
% \todo[inline]{mer innhold..}

\newpage
\bibliography{references}

% If biblatex (file is loaded in preamble):
% \printreferences

\end{document}

### 2012-06-11 UAM2011/Article/removed_contents.tex
More files... Git argued a bit there..

Let us consider an array with $M$ sensors, where the $m$'th element records a signal $x_m[n]$ at instant $n$. If we apply delays $\Delta_m$ and weights $w_m[n]$ to each of these channels, the beamformer output $z[n]$ is given as:
%
\begin{flalign}
z[n] = \sumb{m=0}{M-1} w_m[n]\,x_m[n-\Delta_m] = \w\H[n]\,\X[n],
\end{flalign}
%
where $\w\H[n] = \bmat{w_1[n] & \dots & w_{M-1}[n]}\T$ and $\X = \bmat{w_1[n-\Delta_1] & \dots & w_{M-1}[n-\Delta_{M-1}] }$. The delays chosen such that signals emanating from the direction of interest sum coherently, while noise and interference from other directions do not. The delays are usually derived from geometry alone. The choice of window, however, promote narrow mainlobe and suppressed sidelobe levels. DAS use predefined windows that allows for a suitable compromise between mainlobe width and sidelobe levels. Adaptive beamformers, however, compute the set of weights based on analytical evaluation of the data.
As mentioned the MV beamformer compute the set of weights that miminizes the beamformer output power for each time instant,
\begin{flalign}
\min E\big\{ |z[n]|^2 \big\} = \w\H[n]\, \R[n]\, \w \qquad \text{subject to} \qquad \w\H[n] \a = 1.
\end{flalign}
The contraint is added to ensure a unit gain in the look direction.

% 
% \begin{align}
% \w[n] = \frac{\Ri[n]\,\a}{\a\H\,\Ri[n]\,\a}
% \end{align}

For the LCMV beamformer we have created an assortion of Kaiser window functions. The Kaiser function allows us to design a wide range of windows with different mainlobe widths and sidelobe suppression by adjusting the tradeoff parameter $\beta$. In addition we apply steering $\varphi$ and constrain the window to unit gain in the look direction. We achieved good results by choosing 5 uniformly distributed $\beta$'s in the range $[0.05, 0.5]$, and 5 uniformly distributed $\phi$'s in the range $-0.8\frac{\lambda}{D}, 0.8\frac{\lambda}{D}$.

For statistical robustness the LCMV beamformer was set apply the window with the best performance in a 11 pixel region in range of the image to the center pixel in that region. Likewise, the MV beamformer was set up with range averaging using a window with 11 values, and with 3\% regularisation.

### 2012-06-18 HRUIonGPU/jrnl_article.tex
Forgotten updates to uffc article. capon on speed

@@
 Although the method introduce some interesting trade-offs in addition to just improved resolution, like maintaining image resolution combined with higher framerates, smaller probes, or deeper penetration \cite{Synnevag2009}, widespread use has been limited by the tremendous computational complexity associated with the method \cite{So2011}.
-In this paper we introduce the first implementation of the Capon beamformer exhibiting real-time performance when applied on a typical cardiac data set.
+In this paper we introduce the first implementation of the Capon beamformer exhibiting real-time performance when applied in a typical cardiac setting.
 To accomplish this task we make use of the massive parallel processing power found in modern graphics processing units (GPU's), combined with novel methods to reduce the computational complexity as the problem size grows.
-Earlier attempts on real time Capon beamforming have been focused around radar and passive sonar using intricate systolic array processors \cite{McWhirter1989, Moonen1993, Sinha2002}.
-A systolic array processor is custom made hardware, and even if the computational power might be higher for selected problems compared with a GPU, the cost related to development and flexibility is much higher.
+Earlier attempts on real time Capon beamforming have been focused around radar and passive sonar systems using intricate systolic array processors \cite{McWhirter1989, Moonen1993, Sinha2002}.
+A systolic array processor is custom made hardware, and even if the computational throughput might be higher for selected problems compared with a GPU, the cost related to development and flexibility is considered much higher.
 For medical ultrasound, Chen et al.
 \cite{Chen2011a, Chen2011} have previously implemented the Capon beamformer on a GPU, however their implementation was based on real data, where the Capon beamformer assumes complex.
-Using real data will results in symmetric windows, and even if the number of samples they managed to process per second was high, both the number of floating point operations per sample and the imaging region is reduced compared with use of complex data.
-It is therefore crucial to use complex data within the Capon beamformer framework, both for reducing the amount of data per image, and to allow for asymmetric windows.
+Using real data will results in symmetric windows, and even if the number of samples they managed to process per second was high, both the number of floating point operations per sample and the imaging region is reduced using real data compared with use of complex data.
+It is crucial to use complex data within the Capon beamformer framework, both for reducing the amount of data per image, and to allow for asymmetric windows.
 All results in this paper is reported with use of critically sampled complex data, both laterally and in range.
@@
 In the next section we present background information about the Capon beamformer and the GPU computing model.
-In Section \ref{sec:meth} we go in detail about the parallel implementation of Capon beamforming, before we in Section \ref{sec:meth2} introduce methods to apply when the problem size increases.
+In Section \ref{sec:meth} we go in detail about the parallel implementation of the Capon beamformer, before in Section \ref{sec:meth2} we introduce methods to apply as the problem size increases.
 Benchmark results and resulting images are presented in Section \ref{sec:res}, and finally we discuss the results in Section \ref{sec:dis}, and draw conclusions in Section \ref{sec:con}.
@@
 \end{align}
-In conventional beamforming, the weighting $\vec{w} \in \mathbb{R}$ is usually applied to trade resolution for a better signal-to-noise ratio (SNR).
+In conventional beamforming, the weighting $\vec{w}$ is usually real, and applied to trade resolution for a better signal-to-noise ratio (SNR).
 Hence, by increasing the main lobe width of the system while gaining lower side lobe levels.
@@
 All of these constraints combined pose a real challenge as the target problem has to map on to the parallel framework both on a grid, block and memory consumption level in order to benefits from parallel acceleration.
+In the next section, we will describe how this mapping should be done for Capon beamforming of ultrasound data.
 

### 2012-06-18 HRUIonGPU/jrnl_article.tex

@@
 % the caption=false option directly:
-%\usepackage[caption=false,font=footnotesize]{subfig}
+\usepackage[caption=false,font=footnotesize]{subfig}
 %
@@
 % For subfig.sty:
-% \let\MYorigsubfloat\subfloat
-% \renewcommand{\subfloat}[2][\relax]{\MYorigsubfloat[]{#2}}
+%\let\MYorigsubfloat\subfloat
+%\renewcommand{\subfloat}[2][\relax]{\MYorigsubfloat[]{#2}}
 % For subfigure.sty:
@@
 
+(Go in detail about the implementation using sliding DFT).
+
 \subsection{Iterative Capon}
@@
 
+\subsection{Toeplitz-based solver} 
+(cite new article). Discuss it.
+
 \section{Results}\label{sec:res}
+
+\begin{figure*}[!t]
+\centerline{\subfloat[]{\includegraphics[width=3.5in]{gfx/benchmark_us_32.png} \label{fig:benchUS32}}
+\hfil
+\subfloat[]{\includegraphics[width=3.5in]{gfx/benchmark_us_64.png}%
+\label{fig:benchUS64}}}
+\caption{Benchmark of GPU Capon beamforming for a cardiac image covering a 70 degrees sector.
+Execution time is plotted as a function of subarray length $L$.
+The plot shows execution times for the three steps listed in Section \ref{sec:meth}.
+Calculating covariance matrices in red, solver in green, and the final beamformer sum in blue.
+For each step, there is one line for three different choices of temporal averaging, $N_{avg} \in [0, 1, 2]$.
+(a) A 32 elements array.
+(b) A 64 elements array.}
+\label{fig:bench}
+\end{figure*}
 
@@
 Describe the base band (IQ-demodulation) process that the data has been undergoing.
+
+How has the data been acquired?
 

### 2012-06-18 UAM2011/Article/article2.tex

--- 2012-06-11 UAM2011/Article/article.tex
@@
    UAM                                          % Document type (non-standard)
- , 12pt                                         % Text size
 %  , draft
 %  , final                                        % Quality
- , xelatex                                      % Use the XeLaTeX compiler
+%  , xelatex                                      % Use the XeLaTeX compiler
 %  , biblatex                                     % Use 'biblatex' for references (best by far)
@@
 \begin{document}
+% \fontsize{12pt}{12pt}\selectfont
 
+\layout
 \pagestyle{plain}
@@
 \begin{contact}
-Contact author: Jo Inge Buskenes\\
+Jo Inge Buskenes\\
 Dept. of Informatics, Univ. of Oslo, P.Box 1080 Blindern, N-0316 OSLO\\
@@
 For sonar arrays, this is achieved by applying weights to the array channels.
-
 While classical beamformers use predefined windows, adaptive beamformers estimate the optimal window by analytical evaluation of the data.
@@
 
-An attactive trait of the LCA is its low computational complexity; while the MV beamformer is of O($M^3$), the LCA method is of O($MW$), with $M$ being the number of channels and $W$ the number of windows.
-We made the LCA perform like the MV method using a well designed set of 30 windows.
-Hence, unless the array is very small, the proposed method will perform like the MV beamformer or better, and at a fraction of the computational cost.
+An attactive trait of the LCA is its low computational complexity; while the MV beamformer is of O($M^3$), the LCA is of O($MW$).
+Here $M$ is the number of channels, $W$ the number of windows.
+Hence the proposed method performs like a MV beamformer or better, but at a fraction of the computational cost.
 \end{abstract}%
@@
 
-In active sonar array imaging, an acoustic wave is transmitted and the received echoes are usually recorded using an array of sensors.
+In active sonar array imaging, an acoustic wave is transmitted and the received echoes are recorded using an array of sensors.
 Each of these may be independently delayed and weighted such that signals emanating from directions of interest are added constructively, while noise and interference from other directions add destructively.
@@
 Traditional beamformers such as Delay-and-Sum (DAS) weight the sensors using a predefined window.
-However, since the optimal spatial response is likely to change with time, it is often preferred to use beamformers that continously adapt the weights to the dynamic data.
-This usually involves solving some optimization problem.
+Adaptive beamformers find an optimal window for each pixel in the resulting image.
+These windows are usually found by solving an optimization problem based on the data.
 
 The Minimum Variance (MV) beamformer selects the window that minimizes the beamformer's output power~\cite{cap69}.
-It has recently been investigated for active sonar imaging in e.g.~\cite{saf09}, and in ultrasound imaging in e.g.~\cite{syn07}.
-We suggest using a Low Complexity Adaptive (LCA) beamformer, which for each pixel applies a set of predefined windows and selects one of them based on the MV criterion.
+It has previously been investigated for active sonar imaging in e.g.
+\cite{saf09}.
+We suggest a Low Complexity Adaptive (LCA) beamformer, which for each pixel applies a set of predefined windows and selects one of them based on the MV criterion.
 Using the LCA beamformer results in increased robustness and improved edge definition compared to the MV and DAS beamformers.
+
+% 959 27 818
 
@@
 In addition, a certain amount of diagonal loading is often added to the covariance matrix before inversion.
-These actions, while leading to statistical robustness, tend to deteriorate the beamformer's performance.
+These actions, while leading to statistical robustness, often also deteriorate the beamformers performance.
 Also, the inversion of the covariance matrix has a computational complexity of O($M^3$), where $M$ is the number of elements.
@@
 
-Based on the proposed method by Synnevåg~\cite{syn11}, we have implemented the LCA beamformer that keeps the minimum variance optimisation criterion but reduces the solution space to a discrete set of predefined windows.
+Based on the proposed method by Synnevaag~\cite{syn08}, we have implemented the LCA beamformer that keeps the minimum variance
+optimisation criterion but reduces the solution space to a discrete set of predefined windows.
 This reduces the computational complexity to O($MW$), where $W$ is the number of windows in the set.
-We use the Kaiser window function because it allows us to design a wide range of windows with different mainlobe widths and sidelobe suppression by adjusting the tradeoff parameter $\beta$~\cite{kai66}.
+We use the Kaiser window function because it allows us to design a wide range of windows with different mainlobe widths and sidelobe suppression by adjusting the tradeoff parameter $\beta$.
 In addition we apply steering to each of the windows and constrain the window design to ensure unit gain in the look direction.
@@
 \subfigure{%
-\graphicsAI[width=\linewidth]{gfx/img_das_lca_mv.pdf}}
-\subfigure{%
-\graphicsAI[width=\linewidth]{gfx/img_das_lca_mv_mean100.pdf}}%
-\caption{A simulated circular object buried in speckle is imaged from above.\newline\parbox[t]{1.50cm}{Top:\newline Bottom:}\parbox[t]{6.8cm}{Image of a single speckle realisation.\newline Mean image of 100 speckle realisations.}}\label{speckle}%%
+\graphicsAI[width=\linewidth]{gfx/img_das_lca_mv.pdf}%
+}
+\subfigure{
+\graphicsAI[width=\linewidth]{gfx/img_das_lca_mv_mean100.pdf}%
+}\label{speckle}%
+% \caption{\parbox[t]{1.50cm}{Top:\protect\newline Bottom:}\parbox[t]{6.8cm}{Image of a single speckle realisation.\protect\newline Mean image of 100 speckle realisations.}}%
 \end{figure}%
@@
 \graphicsAI[width=\linewidth]{gfx/img_mean_cut.pdf}%
-\caption{Lateral cut in the mean image at range 41.1\,m (highlight) and 44.5\,m (shadow).}\label{cutmean}%
+\caption{Cut through shadow and highlight region of the mean image.}\label{cutmean}%
 \end{figure}
@@
 \graphicsAI[width=\linewidth]{gfx/img_holmengraa.pdf}%
-\caption{HISAS sidescan sonar (SSS) image of the shipwreck Holmengraa.}\label{holmengraa}%
+\caption{HISAS Synthetic Aperture Sonar (SAS) image of the shipwreck Holmengraa}\label{holmengraa}%
 \end{figure}
@@
 % 
-% Based on the proposed method by Synnevåg~\cite{syn11}, we have implemented a Low Complexity Adaptive (LCA) beamformer that keeps the minimum variance optimisation criteria but reduces the solution space to a discrete set of predefined windows.
+% Based on the proposed method by Synnevåg~\cite{syn08}, we have implemented a Low Complexity Adaptive (LCA) beamformer that keeps the minimum variance optimisation criteria but reduces the solution space to a discrete set of predefined windows.
 We use the Kaiser window function because it allows us to design a wide range of windows with different mainlobe widths and sidelobe suppression by adjusting the tradeoff parameter $\beta$.
@@
 This selection is based on observations of the solutions found by the MV beamformer.
-We obtained good results in our experiments by letting the LCA beamformer select among 5 Kaiser windows with uniformly distributed $\beta$'s in the range $[0.05, 0.5]$ and a rectangular window.
-Each window was steered in 5 different directions, uniformly distributed within 80\% of the mainlobe width of the rectangular window, which is the most narrow.
-This adds to a total of 30 windows.
+We obtained good results in these experiments by letting the LCA beamformer select among 10 Kaiser windows with uniformly distributed parameters ($\beta$) in the range $[0.05, 0.5]$ and a rectangular window.
+Each window is steered in 10 different directions, uniformly distributed within 80\% of the mainlobe width caused by the rectangular window, which is the most narrow.
+This adds to a total of 110 windows.
 
-To make the choice of window less susceptible to pixel value uncertainties, the LCA beamformer was set to apply the most frequently selected window in an 11 pixel range region to the center pixel in that region.
-In physical terms, this is a region of approximately 20\,cm.
-The MV beamformer estimated the covariance matrix as described in~\cite{syn07} by averaging over 16 subarrays and 11 range pixels, and applying 3\% diagonal loading.
+For statistical robustness the LCA beamformer was set to apply the most frequently selected window in an 11 pixel range region to the center pixel in that region.
+The MV beamformer estimated the covariance matrix as described in~\cite{syn08} by averaging over 16 subarrays and 11 range pixels, and applying 3\% diagonal loading.
 
@@
 
-We have tested the beamformers using data aquired by the Kongsberg Maritime HISAS1030 sonar~\cite{han09}, and on simulations of the same sonar with 100 realisations of speckle.
-
+% We have tested the beamformers using data aquired by the Kongsberg Maritime HISAS1030 sonar~\cite{Hansen,Callow,et.al}, and on simulations of the same sonar with 100 realisations of speckle.
 The simulations contained a few highlighted cylindric regions with 2\,m diameter and a constant intensity 15.4\,dB over the average speckle level.
-We focused on an object centered at 41\,m range and -3 degrees azimuth angle.
-The images produced by the LCA, MV and unweighted DAS for a single speckle realisation and a mean image of all 100 realisations are shown in \Fig{speckle}.
-Each image was normalised by their respective speckle level, and the dynamic range was clamped at \{-30,~15\}\,dB.
-The physical extent of the of the simulated object was superimposed for reference.
+We focus on an object centered at 41\,m range and -3 degrees azimuth angle.
+The images produced for a single speckle realisation, and the mean image of all 100 realisations in \Fig{speckle}.
+Each image was normalised by their respective speckle level, and the dynamic range was clamped at \{-30, 15\}\,dB.
+The physical extent of the of the simulated object is superimposed for reference.
 %\ref{saf09}
 
-In \Fig{speckle} we observe that both the adaptive beamformers produce images with a more clearly defined shadow.
-The edges are sharper and the shadow more distinct.
-This is because the adaptive beamformers have better sidelobe suppression, and thus allow less signal energy to leak from the speckle region into the shadow region.
-The same effects can also be observed laterally around the highlight region of the mean image; DAS is more prone to smear energy into the speckle region than the adaptive beamformers are.
+We see that both the adaptive beamformers have far better angular sidelobe suppression than DAS, and produce a more clearly defined shadow.
+This is also observed in \fig{cutmean}, which displays a cut through the shadow and highlight region of the image.
+The transition region between highlight and shadow is shortest for the LCA, implying that its angular resolution is better than that of the MV beamformer.
+This is advantageous since it causes less leakage of highlight into shadow areas, thereby resulting in a more accurate representation of the object size.
+The slightly inferior MV resolution is due to the subarray averaging, which makes the effective array smaller.
 
-\Fig{cutmean} displays two lateral cuts in the mean image, one through the highlight at 41.1\,m range and one through the shadow at 44.5\,m range.
-Each cut was computed as the mean of a 40\,cm range band centered at the respective ranges.
-The transition region between highlight and shadow is shortest for the LCA, which is advantageous since it results in a more accurate representation of the object size.
-The slightly inferior performance of the MV beamformer is due to the subarray averaging, which makes the effective array smaller~\cite{syn07}.
+We have found that the tendency of the LCA to perform similarly to the MV beamformer is consistent in most scenarios, except when it is of essence that strong interferers close to object being imaged are present.
+In such cases the MV beamformer is expected to take advantage of its full solution space to design a spatial response with zero sensitivity at the interferers angles, while LCA will be limted to the spatial responses that we selected.
+In short, whenever a scenario comes along that favors a spatial response we did not think of, the MV will triumph.
+On the other hand, increasing the LCA solution space will at some point leave it with nearly the same freedom as the MV, without suffering from statistical instability.
 
 By studying the LCA performance on the 100 speckle realisations, we found the choice of window to be highly dependent on the phenomena being imaged.
-In speckle regions LCA favored narrow spatial responses with minor steering, while wide and fully steered responses were often selected in shadow regions.
+In speckle regions LCA favored narrow spatial responses with minor steering, while in shadow regions a wide and fully steered responses were often selected.
 There was also a clear distinction between the responses selected in sidelobe regions and in speckle regions.
@@
 
-A sidescan image of the 1500 dwt oil tanker wreck Holmengraa is shown in \Fig{holmengraa}.
-It is about 68\,m long and 9\,m wide, and lies on a slanted seabed at 77\,m depth~\cite{holmengraa}.
-The sidescan image was created using data from the HISAS 1030 sonar, which is rather unsuited for this purpose because of its large opening angle.
-This, and the fact that the wreck was imaged at a range of about 105\,m makes the image quality poor, but sufficient to compare our beamformers.
-
-In the Holmengraa image we note again that the LCA produces a cleaner shadow and better edge definition. The MV method performed almost identically to LCA in this case, and was omitted.
+In \Fig{holmengraa} the wreck Holmengraa is imaged using the LCA and DAS beamformer.
+It is 26\,m long and 6\,m wide, and lies on a slanted seabed at 77\, depth.
+Again, the LCA produces a cleaner shadow and better edge definition.
+The MV method performed almost identically to LCA in this case, and was omitted.
 
@@
 
-The LCA method performs similar to the MV method, but at a fraction of the computational cost.
-The key to achieve this success lies in the design of the window set.
-Windows that yield vagely steered narrow responses are preferred in speckle regions, while wide and steered responses are typically preferred in highlight and shadow regions.
-A window set of 6 different responses each steered in 5 different directions proved sufficient in our experiment to match the performance of the MV beamformer.
+The LCA method delivers performance similar to the MV method, but at a fraction of the computational cost.
+This is verified in our simulations of speckle, and on experimental data.If the windows are With a careful window design, If the window database is designed wisely,  a window database to suit
 
-The tendency of the LCA to perform similarly to the MV beamformer seems to indicate that a full solution space is rarely needed in real world scenarios.
-
-% Nature, while being stochastic, acts within predictable bounds\todo{hehe, jada, jeg skal fjerne..} that can be well adapted to by using beamformers with fewer degrees of freedom than what the MV method provides.
-
-% While the MV beamformer is expected to take advantage of its full solution space to design a spatial response with zero sensitivity at the interferers angles, while LCA will be limted to the spatial responses that we selected.
-In short, whenever a scenario comes along that favors a spatial response we did not think of, the MV will triumph.
-On the other hand, increasing the LCA solution space will at some point leave it with nearly the same freedom as the MV, without suffering from statistical instability.
-% \todo[inline]{mer innhold..}
-
-% While e found that the tendency of the LCA to perform similarly to the MV beamformer is consistent in most scenarios, except when it is of essence that strong interferers close to object being imaged are present\todo{referanse?}.
-In such cases the MV beamformer is expected to take advantage of its full solution space to design a spatial response with zero sensitivity at the interferers angles, while LCA will be limted to the spatial responses that we selected.
-In short, whenever a scenario comes along that favors a spatial response we did not think of, the MV will triumph.
-On the other hand, increasing the LCA solution space will at some point leave it with nearly the same freedom as the MV, without suffering from statistical instability.
-% \todo[inline]{mer innhold..}
-
-\newpage
-\bibliography{references}
+% \printreferences
+% If bibtex:
+\bibliography{references2}
 

### 2012-06-18 HRUIonGPU/jrnl_article.tex
added pointtarget image

--- 2012-06-18 HRUIonGPU/jrnl_article.tex
@@
 Although the method introduce some interesting trade-offs in addition to just improved resolution, like maintaining image resolution combined with higher framerates, smaller probes, or deeper penetration \cite{Synnevag2009}, widespread use has been limited by the tremendous computational complexity associated with the method \cite{So2011}.
-In this paper we introduce the first implementation of the Capon beamformer exhibiting real-time performance when applied in a typical cardiac setting.
+In this paper we introduce the first implementation of the Capon beamformer exhibiting real-time performance when applied in a typical cardiac ultrasound setting.
 To accomplish this task we make use of the massive parallel processing power found in modern graphics processing units (GPU's), combined with novel methods to reduce the computational complexity as the problem size grows.
@@
 
-In practice, the covariance matrix has to be approximated by the sample covariance matrix, $\mat{\hat{R}}[n] = \vec{x}[n]\vec{x}[n]^H$.
+In practice, the covariance matrix has to be approximated by the sample covariance matrix, 
+\begin{align}
+\mat{\hat{R}}[n] = \vec{x}[n]\vec{x}[n]^H.
+\end{align} 
 In passive systems, assuming stationary statistics, the sample covariance matrix is typically averaged over all available samples to improve on the estimate \cite{Krima}.
@@
 introduced by Synnev\aa{}g et al. \cite{Synnevag2007} has been applied in much of the resent literature on Capon beamforming for medical ultrasound imaging \cite{Nilsen2009, Wang2009, Mehdizadeh2012}.
+
+\begin{figure}
+\includegraphics[width=3.5in]{gfx/beamforming_mv_lowres.png}%
+\label{fig:mv}
+\caption{}
+\label{fig:mv}
+\end{figure}
 

### 2012-06-24 HRUIofMovingObjects/jrnl_article.tex

--- 2012-06-18 HRUIonGPU/jrnl_article.tex
@@
 % the caption=false option directly:
-\usepackage[caption=false,font=footnotesize]{subfig}
+%\usepackage[caption=false,font=footnotesize]{subfig}
 %
@@
 % For subfig.sty:
-%\let\MYorigsubfloat\subfloat
-%\renewcommand{\subfloat}[2][\relax]{\MYorigsubfloat[]{#2}}
+% \let\MYorigsubfloat\subfloat
+% \renewcommand{\subfloat}[2][\relax]{\MYorigsubfloat[]{#2}}
 % For subfigure.sty:
@@
 % can use linebreaks \\ within to get better formatting as desired
-\title{Implementing Capon Adaptive Beamforming on the GPU for Real Time Cardiac Ultrasound Imaging}
+\title{Capon Adaptive Beamforming Applied on Moving Objects - An Analysis of Local Shift Invariance}
 %
@@
 
-\author{J. P. \AA{}sen, J. I. Buskenes, C.-I. C. Nilsen, A. Austeng and S. Holm%\IEEEmembership{Member,~IEEE,}% <-this % stops a space
-%\thanks{M. Shell is with the Department
-%of Electrical and Computer Engineering, Georgia Institute of Technology, Atlanta,
-%GA, 30332 USA e-mail: (see http://www.michaelshell.org/contact.html).}% <-this % stops a space
-%\thanks{J. Doe and J. Doe are with Anonymous University.}% <-this % stops a space
-%\thanks{Manuscript received April 19, 2005; revised January 11, 2007.}}
-}
+\author{J. P. \AA{}sen, C.-I. C. Nilsen, A. Austeng and S. Holm%\IEEEmembership{Member,~IEEE,}% <-this % stops a space
+\thanks{M. Shell is with the Department
+of Electrical and Computer Engineering, Georgia Institute of Technology, Atlanta,
+GA, 30332 USA e-mail: (see http://www.michaelshell.org/contact.html).}% <-this % stops a space
+\thanks{J. Doe and J. Doe are with Anonymous University.}% <-this % stops a space
+\thanks{Manuscript received April 19, 2005; revised January 11, 2007.}}
 
@@
 \markboth{Transactions on Ultrasonics and Ferroelectrics and Frequency Control,~Vol.~x, No.~y, January~2013?}%
-{\AA{}sen \MakeLowercase{\textit{et al.}}: Capon Adaptive Beamforming on the GPU}
+{\AA{}sen \MakeLowercase{\textit{et al.}}: The Capon Adaptive Beamformer Applied on Moving Objects }
 % The only time the second header will appear is for the odd numbered pages
@@
 \begin{IEEEkeywords}
-Adaptive beamforming, Capon beamformer, Graphics Processing Unit, GPGPU, Parallel Computing, Ultrasound Imaging.
+Adaptive beamforming, Capon beamformer, Shift Invariance, Ultrasound Imaging.
 \end{IEEEkeywords}
@@
 
-%\IEEEPARstart{I}{ntroduction} to Capon adaptive beamforming, ultrasound imaging and GPU/CUDA.
-\IEEEPARstart{C}{apon} adaptive beamforming applied in a medical ultrasound setting has recently shown promising results in several studies \cite{Synnevag2007, Austeng2008, Vignon2008, Viola, Mehdizadeh2012}.
-Although the method introduce some interesting trade-offs in addition to just improved resolution, like maintaining image resolution combined with higher framerates, smaller probes, or deeper penetration \cite{Synnevag2009}, widespread use has been limited by the tremendous computational complexity associated with the method \cite{So2011}.
-In this paper we introduce the first implementation of the Capon beamformer exhibiting real-time performance when applied on a typical cardiac data set.
-To accomplish this task we make use of the massive parallel processing power found in modern graphics processing units (GPU's), combined with novel methods to reduce the computational complexity as the problem size grows.
-Earlier attempts on real time Capon beamforming have been focused around radar and passive sonar using intricate systolic array processors \cite{McWhirter1989, Moonen1993, Sinha2002}.
-A systolic array processor is custom made hardware, and even if the computational power might be higher for selected problems compared with a GPU, the cost related to development and flexibility is much higher.
-For medical ultrasound, Chen et al.
-\cite{Chen2011a, Chen2011} have previously implemented the Capon beamformer on a GPU, however their implementation was based on real data, where the Capon beamformer assumes complex.
-Using real data will results in symmetric windows, and even if the number of samples they managed to process per second was high, both the number of floating point operations per sample and the imaging region is reduced compared with use of complex data.
-It is therefore crucial to use complex data within the Capon beamformer framework, both for reducing the amount of data per image, and to allow for asymmetric windows.
-All results in this paper is reported with use of critically sampled complex data, both laterally and in range.
+\IEEEPARstart{I}{ntroduction} to Capon adaptive beamforming and temporal artifacts.
 
-In the next section we present background information about the Capon beamformer and the GPU computing model.
-In Section \ref{sec:meth} we go in detail about the parallel implementation of Capon beamforming, before we in Section \ref{sec:meth2} introduce methods to apply when the problem size increases.
-Benchmark results and resulting images are presented in Section \ref{sec:res}, and finally we discuss the results in Section \ref{sec:dis}, and draw conclusions in Section \ref{sec:con}.
+With the introduction of a fast implementation of the Capon adaptive beamformer, a stream of images can now be processed at interactive frame rates.
+A natural question then is how a moving object behaves throughout the stream of images.
+An important property of an ultrasound imaging system is its local shift invariance.
+An object’s appearance should be preserved when moved slightly, or if the ultrasound probe moves.
+The Capon beamformer has previously only been investigated intra-frame; it is therefore unknown how it works when applied in a real-time ultrasound setting.
 
-\section{Background}
-\subsection{Capon Beamforming}
-When sensing the world around us with an array of elements, we perform a spatio-temporal filtering operation known as beamforming.
-The conventional form of beamforming is delay-and-sum (DAS), where the signal recorded on each element, $x_m[n]$, is delayed to broadside with an appropriate delay $\Delta_m[n]$.
-The beamformer output, $z[n]$, is then the sum of all $M$ channels weighted with a factor $w_m$,
+In this paper we analyze the shift invariance property of the Capon beamformer by studying simulated and in vivo images of moving objects.
+We show that when imaging moving point scatterers using the Capon beamformer with the same beam spacing as used for delay-and-sum, artifacts arise in the image, manifested as oscillating point scatteres.
+(TODO: Process csound data to see if we need over sampling on in vivo data, How often do we have bright point scatterers in vivo?).
+
+Delay and sum beamformer (DAS):
 \begin{align}
-z[n] = \sum_{m = 0}^{M-1}w_m^*x_m[n - \Delta_m[n]] = \vec{w}^H\vec{x}[n].
+z[n] = \sum_{m = 0}^{M-1}\vec{w}_m^*\vec{x}_m[n - \Delta_m[n]] = \vec{w}^H\vec{x}[n],
 \end{align}
-In conventional beamforming, the weighting $\vec{w} \in \mathbb{R}$ is usually applied to trade resolution for a better signal-to-noise ratio (SNR).
-Hence, by increasing the main lobe width of the system while gaining lower side lobe levels.
+where $M$ is the number of elements.
 
-The basic idea of an adaptive beamformer is to calculate the weights on-the-fly, based on the data at hand, and on a selected optimization criteria in such a way that we minimize interference impinging from other directions than the point of focus.
-For a complex data vector $\vec{x}[n] \in \mathbb{C}^M$, we have for an adaptive beamformer a data-dependent set of weights, $\vec{w}(\vec{x}[n])$, that when applied on $\vec{x}[n]$ is the best approximation to achieving the selected criteria.
-With the Capon beamformer, adaptive weights are found by solving the following minimization problem \cite{Capon1969}:
+Capon weights are found by solving the following minimization problem:
 \begin{align}
-&\min_{\vec{w}} E\{|z[n]|^2\} = \min_{\vec{w}}\vec{w}^H\mat{R}\vec{w}\label{eq:minz}\\
-&\text{subjected to } \vec{w}^H\vec{a} = 1,\label{eq:constraint}
+&\min_{\vec{w}} E\{|z[n]|^2\} = \vec{w}^H\mat{R}\vec{w}\\
+&\text{subjected to } \vec{w}^H\vec{1} = 1,
 \end{align}
-where $\mat{R} = E\{\vec{x}[n]\vec{x}[n]^H\}$ is the spatial covariance matrix, and $\vec{a} = \left[ \begin{array}{cccc} e^{-jk_{\tilde{x}}\tilde{x}_0}& e^{-jk_{\tilde{x}}\tilde{x}_1}& ...& e^{-jk_{\tilde{x}}\tilde{x}_{M-1}}\end{array}\right] ^T$ is a far-field steering vector towards angle $\theta = \arcsin(\frac{\lambda k_{\tilde{x}}}{2\pi})$.
-Thus, we minimize the output power, while maintaining unit gain in the look direction.
-The constraint in (\ref{eq:constraint}) is there to avoid the trivial solution $\vec{w} = 0$, and a distorted respons in the direction of $\vec{a}$.
-The solution to the minimization problem in (\ref{eq:minz}) and (\ref{eq:constraint}) is
-\begin{align}\label{eq:w}
+where $\mat{R} = E\{\vec{x}[n]\vec{x}[n]^H\}$.
+
+The solution to this minimization problem is
+\begin{align}
 \vec{w}[n] = \frac{\mat{R}[n]^{-1}\vec{a}}{\vec{a}^H\mat{R}[n]^{-1}\vec{a}},
 \end{align}
-where $\vec{a} = [1_0 1_1... 1_{M-1}]^T = \vec{1}$ when $\vec{x}$ is pre-delayed to broadside.
+where $\vec{a} = \vec{1}$ when $\vec{x}$ are pre-delayed to broadside,  and $\vec{w} \in \mathbb{C}^M$.
 
-In practice, the covariance matrix has to be approximated by the sample covariance matrix, $\mat{\hat{R}}[n] = \vec{x}[n]\vec{x}[n]^H$.
-In passive systems, assuming stationary statistics, the sample covariance matrix is typically averaged over all available samples to improve on the estimate \cite{Krima}.
-For an active, pulsed system like ultrasound imaging, the available data from which $\mat{R}$ can be estimated is quite limited.
-The resulting sample covariance matrix might therefore be both inaccurate, and poorly conditioned.
-Another challenge with applying Capon to active systems is the effect of signal cancellation, where high correlation between signal and interference favours weights that in the end cancel out the signal \cite{Synnevag2009a}.
-In this way, the beamformer output can be zero at broadside, even when there is signal present and the distortion-less criterion is applied.
-In order to get a well conditioned $\mat{\hat{R}}$, avoid signal cancellation and to get DAS-like speckle, $\mat{\hat{R}}$ has to be averaged over $L\le M/2$ long subarrays and $N$ time samples \cite{Synnevag2007, Synnevag2007a}.
+Estimation of $\mat{R}[n]$: $\mat{\hat{R}}[n] = \vec{x}[n]\vec{x}[n]^H$.
+
+In order to get a well conditioned $\mat{\hat{R}}$, avoid signal cancellation and to get DAS-like speckle, $\mat{\hat{R}}$ has to be averaged over $L\le M/2$ subarrays and $N_{avg}$ time samples.
 \begin{align}
-\mat{\hat{R}}[n] = \frac{1}{NK}\sum_{n'=n-N_{avg}}^{n+N_{avg}} \sum_{l=0}^{K-1} \vec{x}_l[n']\vec{x}_l[n']^H,\label{eq:R}
+\mat{\hat{R}} = \frac{1}{(2N_{avg} + 1)K}\sum_{n'=n-N_{avg}}^{n+N_{avg}} \sum_{l=0}^{K} \vec{x}_l[n']\vec{x}_l[n']^H,
 \end{align}
-where $K = M-L+1$, $N = 2N_{avg} + 1\sim \tau/T_s$ ($\tau$ is the pulse length and $T_s$ is the sampling period) and $\vec{x}_l$ is the $l^\text{th}$ subarray $[x_l[n], \dotso x_{l+L}[n]]$.
-Finally, $\mat{\hat{R}}$ is loaded with diagonal factor $\epsilon$, 
-\begin{align}
-\mat{\hat{R}} = \mat{\hat{R}} + \epsilon\mat{I}.
-\end{align}
-A weighting proportional to the output power is often used to reduce the need of parameter adjustments, and the trace of $\mat{\hat{R}}$, 
-\begin{align}\label{eq:diag}
-\epsilon &= d*tr\{\mat{\hat{R}}\}, \\
-d &= 1/L
-\end{align}
-introduced by Synnev\aa{}g et al. \cite{Synnevag2007} has been applied in much of the resent literature on Capon beamforming for medical ultrasound imaging \cite{Nilsen2009, Wang2009, Mehdizadeh2012}.
+where $K = M-L+1$, $Y_{avg}\sim \tau/T_s$ ($\tau$ is the imaging pulse length and $T_s$ is the sampling period.) and $\vec{x}_l$ is the $l^\text{th}$ subarray $[x_l[n], \dotso x_{l+L}[n]]$.
 
-%Building and calculating the inverse of $\hat{R}$ is computational demanding, limiting the methods accessibility. Following the innovation in GPU computing, this has now started to change. 
-%In this paper we introduce the first GPU implementation of the Capon beamformer capable of processing a 70 degree sector cardiac image from a 64 element phased array at interactive frame rates using both spatial and temporal smoothing.
-Both types of smoothing are important for maintaining delay-and-sum-like speckle and at the same time keep the high resolution property.
-This is achieved using complex base band data, which reduces the size of each range line to a minimum.
-In addition, complex data results in complex weights; hence this is a correct implementation of Capon beamforming with all degrees of freedom preserved.
-For high channel count systems we propose two ways of reducing computational complexity while maintaining
+Finally, $\mat{\hat{R}}$ is loaded with diagonal factor $\epsilon$, $\mat{\hat{R}} = \mat{\hat{R}} + \epsilon\mat{I}$.
+A weighting proportional to the output power is beneficial, and the trace of $\mat{\hat{R}}$, $\epsilon = d*tr\{\mat{\hat{R}}\}$, has previously shown good results in the literature (cite).
 
-\subsection{GPU Compute Model}
-In this section we briefly discuss the principle behind GPU computing, laying the basis for further discussion on parallel implementation strategies.
-CUDA by Nvidia has been our choice of GPU developer environment \cite{Nvidia2011}.
-However, even though CUDA has been used to implement the Capon beamformer, the challenges and proposed methods described in this paper applies for other massive parallel implementation environments as well, like e.g.
-OpenCL.
+Building and calculating the inverse of $\hat{R}$ is computational demanding, limiting the methods accessibility. Following the innovation in GPU computing, this has now started to change.
 
-The CUDA compute model is based on execution of a kernel-function across a grid of compute threads.
-Each position in the grid holds a block of threads, with maximal size of 1024 threads, that can be easily synchronized if needed.
-%Each block is further divided into groups of 32 threads, known as a warp. The maximum number of threads inside a block is restricted to 512, hence 16 warps. 
-From a CUDA perspective the GPU consist of one or more stream multi processors (SM's), where each SM is capable of 32 or more, depending on the architecture, concurrent multiplications and/or additions.
-A group of blocks is scheduled to each SM until all blocks have been processed.
-More important, the SM has a limited amount of near-core memory and registers available.
-Each thread therefore has to be as light on memory as possible and has to perform enough instructions per used byte in order to hide memory latency.
-Global read and writes on the GPU, in addition to CPU-GPU transfers have high penalty, and should be minimized as much as possible.
-All of these constraints combined pose a real challenge as the target problem has to map on to the parallel framework both on a grid, block and memory consumption level in order to benefits from parallel acceleration.
-
-%When implementing algorithms on the GPU it is important to keep data as close to the core as possible. Utilizing available registers and shared memory to its maximum is therefore important.
-
-%More important, the CUDA 2.x compute capability (CC) has available 48 KB of near-core shared memory per SM that can be utilized by resident blocks.
-The maximum number of resident blocks per SM is 8, but in addition the maximum number of resident warps per SM is restricted to 48.
-Hence, with a block size of 512 only $48/(512/32) = 3$ blocks can occupy the SM at once.
-In addition, if one thread where to use 64B of shared memory the number of allowed resident blocks with a block size of 512 would be reduced to 1.
-Up to a curtain point it is beneficial to have as many warps as possible occupying the SM at once to hide latency from accessing memory.
-The SM has for CC 2.x 32K 32-bit long registers.
-When implementing algorithms on the GPU it is important to keep data as close to the core as possible.
-Utilizing available registers and shared memory to its maximum is therefore important.
-
-%\subsection{Optimizing kernel launch}
-%Set the size of grid adaptively based on $L$, $M$ and $Yavg$.
-%It is better to slide in space than time. $K$ is typically much larger than $2*Yavg + 1$.
+Nilsen et.
+al has proposed to apply beamspace processing to further reduce the computations required.
+With beamspace we mean transforming the channel data from element space to a fan of beams covering the imaging sector, $\vec{x}_{BS} = \mat{B}x$.
+The steering vectors in $\mat{B}$, the so-called Butler matrix, determines each beams direction in space.
+We can therefore easily remove beams where no interference is present by removing rows in $\mat{B}$.
+For a focused system like medical ultrasound, where the received signal is concentrated in a narrow band around broadside, there is a small percentage of the beams that contains almost all of the energy.
+Nilsen et.
+al has shown that as little as three beams still produced results comparable with applying the full Capon estimation in element space.
 
@@
 
-\section{Parallel Capon}\label{sec:meth}
-As mentioned, widespread use of Capon adaptive beamforming has been limited by its computational complexity.
-However, the calculation of Capon weights is well suited for a GPU implementation, since the calculations are both independent across pixels, and partially intra-pixel.
-Based on memory dependencies it is trivial to recognise that one block of threads should process at least one pixel in the final image.
-Continuing, on a pixel-level, the algorithm can be broken down into three main steps that to some extent can be further divided into a small set of parallel processes.
-That is, for each data vector $\vec{x}$ in the image:
-\begin{enumerate}
-\item Calculate the sample covariance matrix $\mat{\hat{R}}$ as in (\ref{eq:R})
-\item Solve the linear system of equations $\mat{\hat{R}}\vec{b} =\vec{ 1}$
-\item Calculate $\vec{w} = \vec{b}/\vec{1}^H\vec{b}$ and the amplitude $\vec{w}^H\vec{x}$
-\end{enumerate}
-The computational complexity of these three steps, with $L = M/2$, is $O(L^3)$, $O(L^3)$, and $O(L)$ respectively.
-For larger $L$'s, the first two steps quickly becomes unmanageable, and real time performance can only be achieved by exploring novel solutions to reduce complexity.
+\section{Temporal Artifacts}
 
-\subsection{Calculating Multiple Sample Covariance Matrices}
-The straight forward sum in (\ref{eq:R}) has complexity $O(L^2KN) = O(L^3)$ if $L$ is a small fraction of $M$.
-This is actually the same as for inverting a $L \times L$ matrix.
-However, little attention has been devoted to this step in the literature until now.
-Note that for a typical array configuration, N is usually small compared to L and M.
-Hence, for a small $L$ the construction of $\mat{\hat{R}}$ is dominated by $M$, since $K = M - L + 1$.
-From (\ref{eq:R}) it is obvious that calculation of one element in $\mat{\hat{R}}$ is independent of all the others.
-A natural level of granularity would then be to assign one thread to each element of $\mat{\hat{R}}$.
-This will minimize global reads and writes needed per thread, but unfortunately the maximum number of threads per block puts an upper limit on $L$, $L_{max}=32$, if we want to avoid block-to-block synchronization.
+\section{Causes}\label{sec:meth}
 
-The quadratic complexity with $L$ and linear complexity with $N$ can be reduced by realizing that the calculation in (\ref{eq:R}) overlap both across subarrays and in time.
-The $(i,j)$ element of $\mat{\hat{R}}[n]$ can be calculated from element $(i-1, j-1), \text{for } i,j > 0$ by subtracting $x_{i-1}x_{j-1}^*$ and adding $x_{i+K-1}x_{j+K-1}^*$.
-The same is true in time where $\mat{\hat{R}}[n]$ can be calculated from $\mat{\hat{R}}[n-1]$ by subtracting $\mat{\hat{R}}[n-N_{avg}-1]$ and adding $\mat{\hat{R}}[n+N_{avg}]$.
-Both sliding across subarrays and in time can be realized simultaneously if we hold $\vec{x}\vec{x}^H[n]$ averaged over $N$ time samples in memory and calculate $\mat{\hat{R}}$ averaged over subarrays from that buffer (figure).
-However, since $N$ typically is small, and we have a limited amount of memory per thread on the GPU, we have chosen to focus on sliding across subarrays only.
-In that case, the complexity of calculating $\mat{\hat{R}}$ is reduced to $O(L^2)$.
-Another aspect is that sliding will break independence in the sliding dimension.
-With our choice of calculating $\mat{\hat{R}}$, calculations are still independent across time, but not across subarrays.
-We therefore have selected to use on thread per column in $\mat{\hat{R}}$, and for small $L$'s several covariance matrices are calculated per thread block.
-Maximum supported $L$ is with this approach increased to 512, however for small $L$ high $M$ configurations, memory consumption per thread will be to high to achieve full occupancy of the GPU.
+\subsection{Local Shift Variance}
 
-Taking symmetry into account, only the upper half of $\mat{\hat{R}}$ needs to be calculated.
-However, the way we have decided to organize the GPU compute threads per $\mat{\hat{R}}$, the lower triangle can be calculated with little overhead on the GPU.
-This also means that we do not relay on a symmetric solver to solve $\mat{\hat{R}}\vec{b} = \vec{1}$, even if that could have saved some instructions.
-The memory layout for the upper triangle only is also believed to be cumbersome.
+\section{Cures}
 
-Diagonal loading as a constant value can be added with marginal cost, but the adaptive diagonal loading factor in (\ref{eq:diag}) is a different story.
-In this case, the trace of $\mat{\hat{R}}$ has to be known before diagonal elements are written back to global memory.
-Since one row for all $\mat{\hat{R}}$'s is written to global memory per kernel iteration, the solution was to hold the diagonal elements in a local buffer until the trace was accumulated, and then do scattered writes of the diagonal elements to global memory.
+\subsection{Multiple Parallel Recieve Lines}
 
-%For subarray averaging, the first row and column of $\mat{\hat{R}}$ has to be fully calculated, and for time averaging $\mat{\hat{R}}[0]$ has to be fully calculated.
-
-%Since $L$ usually is larger than $N$, we have
-
-%Building $\hat{\mat{R}}$ using a sliding window approach across $K$ and $N_{avg}$.
-We have a limited amount of fast, near-core memory on the GPU.
-On NVIDIA architecture this memory is known as shared memory, and is restricted to 48KB per compute block.
-The maximum block size of 32x32, further restrict how large arrays we can handle inside a single compute block.
-It is therefore natural to restrict $L$ to a maxima of $32$ elements.
-Since $L \le M/2$, $M \le 64$.
-We can not afford to hold the full $\hat{\mat{R}}$ ($M = L$) when $\hat{\mat{R}} > 32$, because of limited amount of memory.
-...
-Explain in detail how \texttt{buildR} is implemented in a kernel.
-...
-Give more details on how much shared memory we have, and how it can be divided between compute blocks occupying one stream multiprocessor (SM).
-
-%Discuss the bottleneck adaptive diagonal loading causes. Possible solutions includes, use trace from previous image, use pre-calculated array power, constant weight.
-
-%Time averaging takes a lot of resources.
-It has been shown that time averaging can maintain both resolution and delay-and-sum-like speckle.
-The same speckle statistics can be obtain with small subarrays, but then we loose resolution.
-Small sub-arrays benefits from being less computational demanding.
-
-%\subsubsection{CUDA Compute Model}
-%The CUDA compute model is based on execution of a kernel across a grid of compute threads.
-Each position in the grid holds a block of threads.
-Each block is further divided into groups of 32 threads, known as a warp.
-The maximum number of threads inside a block is restricted to 512, hence 16 warps.
-From a CUDA perspective the GPU consist of one or more stream multi processors (SM), %where each SM is capable of 32 or more, depending on the architecture, concurrent multiplications and/or additions.
-The CUDA 2.x compute capability (CC) has available 48 KB of near-core shared memory per SM that can be utilized by resident blocks.
-The maximum number of resident blocks per SM is 8, but in addition the maximum number of resident warps per SM is restricted to 48.
-Hence, %with a block size of 512 only $48/(512/32) = 3$ blocks can occupy the SM at once.
-In addition, if one thread where to use 64B of shared memory the number of allowed resident blocks with a block size of 512 would be reduced to 1.
-Up to a curtain point
-%it is beneficial to have as many warps as possible occupying the SM at once to hide latency from accessing memory and executing time consuming mathematical functions.
-The SM has for CC 2.x 32K 32-bit long registers.
-When implementing algorithms on the GPU it is important to keep data as close to the core as possible.
-Utilizing available registers and shared memory to its maximum is %therefore important.
-
-\subsection{Solving Multiple Small Linear Systems}
-When it comes to solving systems of linear equations on the GPU the focus has mostly been on large systems in the literature, and GPU libraries usually lack a comprehensive collection of batched solvers.
-The reason for this is two-folded, first the GPU needs to solve thousand of small systems in order to beat the low memory latency of the CPU, second there is a range of system dimensions that does not fit well with the GPU architecture, where only a small amount of near-core memory registers are available.
-Because of this, it has been proven hard to optimize a manageable set of solvers that provides speedups compared with the CPU in every cases.
-
-For the results presented in this paper, we have used an unreleased GPU implementation of Gauss Jordan (GJ) elimination (by NVIDIA) to solve $\mat{R}\vec{b} = \vec{1}$, where $\vec{b} = \mat{R}^{-1}\vec{1}$.
-However we will go through an implementation of an $\mat{U}^H\mat{D}\mat{U}$-decomposition-based solver to point out challenges when implementing batched solving of small linear systems on the GPU.
-...
-(Unsure if this should be included).
-
-%Can include details on GPU implementation of $\mat{U}^H\mat{D}\mat{U}$, but this has not proved to be faster than NVIDIA's GJ implementation.
-However, the complexity for solving with $\mat{U}^H\mat{D}\mat{U}$ decomposition is supposed to be $1/2$ of GJ.
-
-For the smallest systems a direct solver using one thread per system is preferable. For this purpose we have implemented Cramer's rule for $3 \times 3$ (, $2 \times 2$?  and $4 \times 4$?) matrices.
-
-%In the final journal article we could discuss Conjugated gradient and Woodbury and the benefits they bring.
-Other solvers that provides different benefits includes Conjugated gradient and the use of the Woodbury matrix identity.
-
-\subsection{Compute Beamformer Output}
-(Still just notes) One thread per weight ($L$ threads). Let one thread sum the solution vector $\vec{b}$, or let several threads cooperate to form the sum using a binary reduction sum.
-Let one thread read one element of $\vec{b}$, scale it with the inverse sum, and optionally write it back to global memory.
-The final output can be formed by either reduce the data down to length $L$, or increase the weight vector to length $M$.
-Since we have decided to have $L$ threads per output, the data is reduced to length $L$ by calculating $x_{r} = \sum_{k=r}^{r+K-1}x_{k}$ for $r \in [0, L-1]$.
-Each thread now computes $z_r = w_r^*x_r$, and finally one thread computes and outputs $z = \sum_{r=0}^{L-1}$ to global memory.
-The last sum can also be performed using a binary reduction sum across several threads for increased speed.
-
-\section{Real Time Performance for $L > 16$}\label{sec:meth2}
-
-\subsection{Beamspace Processing}
-Nilsen et.
-al \cite{Nilsen2009} has proposed to apply beamspace processing to further reduce the required computations.
-With beamspace we mean transforming the channel data from element space to a fan of beams covering the imaging sector, $\vec{x}_{BS} = \mat{B}x$.
-The steering vectors in $\mat{B}$, the so-called Butler matrix, determines each beams direction in space.
-We can therefore easily remove beams where no interference is present by removing rows in $\mat{B}$.
-For a focused system like a cardiac ultrasound scanner, where the received signal is concentrated in a narrow band around broadside, there is a small percentage of the beams that contains almost all the energy.
-Nilsen et.
-al concluded that as little as three beams around broadside where adequate to produce results comparable to applying Capon beamforming in element space.
-The beamspace version of Capon is derived by replacing the covariance matrix in (\ref{eq:w}) with the beamspace covariance matrix $\mat{R}_{BS} = \vec{x}_{BS}\vec{x}_{BS}^H$, where $\mat{B} \in \mathbb{C}^{N_b,M}$, and $N_b$ is the number of selected beams.
-Note that when subarray averaging is applied, the transformation must be done per subarray, yielding subarrays that no longer overlap in space.
-Hence, the amount of data is change from $M$ to $KN_b$ per sample.
-
-(Go in detail about the implementation using sliding DFT).
-
-\subsection{Iterative Capon}
-The idea behind iterative Capon is to perform Capon using small subarrays in several iterations until there is only two subarrays to sum.
-In this way we fix the size of L, hence the complexity of solving $\mat{R}\vec{b} = \vec{1}$.
-However, for each iteration we re-obtain L-1 degrees of freedom until all are retrieved after $K = M-L+1$ iterations.
-Since subarray averaging has been applied to decorrelated the signal, we might loose this effect when reestablishing all degrees of freedom.
-The number of iterations should therefore be restricted to $K/2$ to equal the normal choice of $L = M/2$ when it comes to degrees of freedom.
-
-\subsection{Toeplitz-based solver} 
-(cite new article). Discuss it.
+\subsection{Phase Rotation}
 
 \section{Results}\label{sec:res}
-
-\begin{figure*}[!t]
-\centerline{\subfloat[]{\includegraphics[width=3.5in]{gfx/benchmark_us_32.png} \label{fig:benchUS32}}
-\hfil
-\subfloat[]{\includegraphics[width=3.5in]{gfx/benchmark_us_64.png}%
-\label{fig:benchUS64}}}
-\caption{Benchmark of GPU Capon beamforming for a cardiac image covering a 70 degrees sector.
-Execution time is plotted as a function of subarray length $L$.
-The plot shows execution times for the three steps listed in Section \ref{sec:meth}.
-Calculating covariance matrices in red, solver in green, and the final beamformer sum in blue.
-For each step, there is one line for three different choices of temporal averaging, $N_{avg} \in [0, 1, 2]$.
-(a) A 32 elements array.
-(b) A 64 elements array.}
-\label{fig:bench}
-\end{figure*}
-
-Present images before and after Capon weights has been applied.
-Comment on the effect of selecting different parameters.
-Present graph showing executing times for Cython-Capon v.s.
-CUDA-Capon for different choices of parameters.
-Comment on different choices of parameters change the running times.
-Temporal averaging for instance, lowers the efficiency since $2*Y_{avg}$ more operations has to be performed per $\mat{\hat{R}}$
-
-Present images, both simulated and in-vivo, processed using beamspace Capon and iterative Capon. Comment on the differences between the two and on how much they deviates from the full method.
-Show beampatterns for the two compared with the full method. Show executing times for the two methods compared with the full method on the GPU.
-
-Describe the base band (IQ-demodulation) process that the data has been undergoing.
-
-How has the data been acquired?
 
@@
 
-For a typical cardiac ultrasound image of 80*400 pixels (70 degree sector, 15cm range) acquired using a 2.5MHz, $M=64$ element phased array, the result is 10fps (subarray length $L=M/2$, temporal smoothing over 3 samples).
-If we do a 2-element pre-beamforming to get the channel count down to 32, the frame rate increases to 44fps.
-For a 32 element phased array we need less beams to cover the sector (40*400 pixels), hence with the same parameters the frame rate increases to 87fps.
-The bottleneck of the Capon algorithm is to find the inverse covariance matrix for all pixels.
-This is an $O(L^3)$ operation, and results show that when L is larger than 16, real time performance is not possible for cardiac imaging with current high-end hardware.
-The target GPU has been Nvidia’s Quadro 6000, capable of 1Tflops.
+The figure below shows a lateral shift invariance plots for a well sampled delay-and-sum image (a), the same sampling with Capon processing (b), and 4x over sampling processed with Capon.
+The shift invariance plots the beamprofile (TODO: Make the shift invariance plot).
+The results show that when using the Capon beamformer we have to over sample with a factor four to remove the blinking artifact.
+This can be done without affecting frame rate by using the available parallel receive-lines found on most ultrasound systems.
+May also use phase rotation.
+
+TODO: Add citations – Hergum/Bjåstad Local shift invariance and MLA’s… Is there any publications stating that Capon do not work on moving objects?
 
@@
 %\bibliography{IEEEabrv,../bib/paper}
-\bibliography{mybib} % link to mendeley organized bibtex-file
-%\bibliography{IEEEabrv,mybib} 
+\bibliography{mybib}
+%
 % <OR> manually copy in the resultant .bbl file
@@
 \begin{IEEEbiography}{Jon Petter \AA{}sen}
-Biography text here.
-\end{IEEEbiography}
-
-\begin{IEEEbiography}{Jo Inge Buskenes}
 Biography text here.

### 2012-06-24 HRUIonGPU/jrnl_article.tex
added figures to document

@@
 %\boldmath
-The abstract goes here.
+The abstract goes here. The abstract goes here. The abstract goes here. The abstract goes here. The abstract goes here. 
+The abstract goes here. The abstract goes here. The abstract goes here. The abstract goes here. The abstract goes here. 
+The abstract goes here. The abstract goes here. The abstract goes here. The abstract goes here. The abstract goes here. 
+The abstract goes here. The abstract goes here. The abstract goes here. The abstract goes here. The abstract goes here.  
+The abstract goes here. The abstract goes here. The abstract goes here. The abstract goes here. The abstract goes here. 
+The abstract goes here. The abstract goes here. The abstract goes here. The abstract goes here. The abstract goes here. 
+The abstract goes here. The abstract goes here. The abstract goes here. The abstract goes here. The abstract goes here. 
+The abstract goes here. The abstract goes here. The abstract goes here. The abstract goes here. The abstract goes here. 
+The abstract goes here. The abstract goes here. The abstract goes here. The abstract goes here. The abstract goes here. 
+The abstract goes here. The abstract goes here. The abstract goes here. The abstract goes here. The abstract goes here.  
+The abstract goes here. The abstract goes here. The abstract goes here. The abstract goes here. The abstract goes here. 
+The abstract goes here. The abstract goes here. The abstract goes here. The abstract goes here. The abstract goes here. 
 \end{abstract}
@@
 \section{Background}
+
+\begin{figure*}[t!]
+\centerline{\subfloat[]{\includegraphics[width=3in]{gfx/beamforming_mv_lowres.png} \label{fig:mvbf}}
+\hfil
+\subfloat[]{\includegraphics[width=3in]{gfx/gpu_layout_lowres.png}%
+\label{fig:gpulayout}}}
+\caption{a) Beamforming with adaptive weights $\vec{w}[n]$.
+Impinging signal is demodulated and aligned in phase by applying steering and focusing delays $\Delta_m[n]$.
+Capon adaptive weights are then calculated based on the in-phase data, and finally the output is formed by a weighted coherent sum.
+b) Calculation of Capon adaptive weights mapped to GPU architecture.
+A recorded echo across the array for a given range and angle (selected cell) is assigned to a group of $L$ threads that can run independently of all other groups.
+Note that several samples can be processed per block of threads if $L$ is small.
+The task of these $L$ threads is to calculate the Capon beamformer output in the three depicted steps.}
+\label{fig:fig1}
+\end{figure*}
+
 \subsection{Capon Beamforming}
@@
 
-The basic idea of an adaptive beamformer is to calculate the weights on-the-fly, based on the data at hand, and on a selected optimization criteria in such a way that we minimize interference impinging from other directions than the point of focus.
+The basic idea of an adaptive beamformer, as shown in Figure \ref{fig:mvbf}, is to calculate the weights on-the-fly, based on the data at hand, and on a selected optimization criteria in such a way that we minimize interference impinging from other directions than the point of focus.
 For a complex data vector $\vec{x}[n] \in \mathbb{C}^M$, we have for an adaptive beamformer a data-dependent set of weights, $\vec{w}(\vec{x}[n])$, that when applied on $\vec{x}[n]$ is the best approximation to achieving the selected criteria.
@@
 
-\begin{figure}
-\includegraphics[width=3.5in]{gfx/beamforming_mv_lowres.png}%
-\label{fig:mv}
-\caption{}
-\label{fig:mv}
-\end{figure}
+%\begin{figure}
+%\includegraphics[width=3in]{gfx/beamforming_mv_lowres.png}
+%\caption{}
+%\label{fig:mv}
+%\end{figure}
 
@@
 The CUDA compute model is based on execution of a kernel-function across a grid of compute threads.
-Each position in the grid holds a block of threads, with maximal size of 1024 threads, that can be easily synchronized if needed.
+As depicted in Figure \ref{fig:gpulayout}, each position in the grid holds a block of threads, with maximal size of 1024 threads, that can be easily synchronized if needed.
 %Each block is further divided into groups of 32 threads, known as a warp. The maximum number of threads inside a block is restricted to 512, hence 16 warps. 
@@
 For larger $L$'s, the first two steps quickly becomes unmanageable, and real time performance can only be achieved by exploring novel solutions to reduce complexity.
+
+(The three steps have, for modularity, been implemented using three different kernels. Joining all kernels into one could possible lead to an increase in speed.)
+
+%\begin{figure}
+%\includegraphics[width=3in]{gfx/gpu_layout_lowres.png}%
+%\caption{}
+%\label{fig:gpulayout}
+%\end{figure}
 
@@
 \begin{figure*}[!t]
-\centerline{\subfloat[]{\includegraphics[width=3.5in]{gfx/benchmark_us_32.png} \label{fig:benchUS32}}
+\centerline{\subfloat[]{\includegraphics[width=3.5in]{gfx/benchmark_us_32_lr.png} \label{fig:benchUS32}}
 \hfil
-\subfloat[]{\includegraphics[width=3.5in]{gfx/benchmark_us_64.png}%
+\subfloat[]{\includegraphics[width=3.5in]{gfx/benchmark_us_64_lr.png}%
 \label{fig:benchUS64}}}
@@
 (b) A 64 elements array.}
+\label{fig:bench}
+\end{figure*}
+
+\begin{figure*}[!t]
+\centerline{\subfloat[]{\includegraphics[width=3.5in]{gfx/benchmark_us_32_lr.png} \label{fig:benchUS32}}
+\hfil
+\subfloat[]{\includegraphics[width=3.5in]{gfx/benchmark_us_64_lr.png}%
+\label{fig:benchUS64}}}
+\caption{(TODO: Add the right figure) Benchmark of GPU Beamspace Capon beamforming for a cardiac image covering a 70 degrees sector.
+Execution time is plotted as a function of number of elements $M$.
+The plot shows execution times for the three steps listed in Section \ref{sec:meth}.
+Calculating covariance matrices in red, solver in green, and the final beamformer sum in blue.
+For each step, there is one line for three different choices of temporal averaging, $N_{avg} \in [0, 1, 2]$.
+(a) 3 beams.
+(b) 5 beams.}
 \label{fig:bench}
@@
 
+We expect that a large speedup had the Capon beamformer been implemented on the CPU using threads and SIMD (single instruction multiple data) instructions.
+
+(results)?
 For a typical cardiac ultrasound image of 80*400 pixels (70 degree sector, 15cm range) acquired using a 2.5MHz, $M=64$ element phased array, the result is 10fps (subarray length $L=M/2$, temporal smoothing over 3 samples).

### 2012-06-24 UAM2011/Article/template/example.tex

%===============%                               %~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~%
% DOCUMENTCLASS %                                See full option description in "mytemplate.cls"
%===============%                               %~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~%

\documentclass[
%  , draft
%  , final                                        % Quality
%   xelatex                                      % Use the XeLaTeX compiler?
%  , biblatex                                     % Use 'biblatex' for references (best by far)
  bibtex                                       % Use 'bibtex' for references (oldschool :)
]{uam}

% If biblatex is used: Load those references loaded in preamble:
% \bibliography{references}

\begin{document}

\pagestyle{plain}

% Note. Keep the %'s on the title page to ensure that no extra paragraphs are added.
%       We don't want Latex to apply magic rubber'lengths here, we want control.
%       Eventually I'll need to redo the \titlepage command, but not now...

\title{Keep on Rockin'}
%
\author{J.I.Buskenes\firstAddress, A.Austeng\firstAddress, C.-I.C.Nilsen\secondAddress}
%
\begin{contact}
  \firstAddress Institute of Informatics, P.Box 1080 Blindern, N-0316 OSLO \\
  \secondAddress No street, $\infty\infty\infty\infty$ NEVER EVER LAND
\end{contact}
%
\begin{contact}
Jo Inge Buskenes \\
Dept. of Informatics, Univ. of Oslo, P.Box 1080 Blindern, N-0316 OSLO \\
\href{mailto:joibu@ifi.uio.no}{joibu@ifi.uio.no}
\end{contact}
%
\begin{abstract}
Mauris a dui ac velit lacinia sodales sit amet at libero. Nam lobortis ante non velit tincidunt vestibulum. Nunc eu ipsum quis ante fermentum tincidunt. Donec ac lorem et tortor vulputate elementum. Pellentesque vel velit sed neque sagittis pretium. Duis posuere sagittis arcu sed imperdiet. Morbi sed ligula a tellus pulvinar tincidunt. Pellentesque lacinia urna vel velit lobortis fermentum. Ut eu metus in nulla pharetra lacinia. Nullam lacus velit, viverra sed aliquet id, laoreet ut enim. Suspendisse dictum sem sit amet lorem tincidunt mollis. In vestibulum, dui id faucibus aliquam, purus libero consequat leo, quis condimentum sem quam ut dui. Mauris ut vestibulum augue. In porttitor orci vel nulla aliquam pretium. Donec nunc erat, ultrices ac tincidunt nec, condimentum eget elit. Aenean lacinia purus eu tellus tincidunt blandit. Suspendisse potenti. In rutrum urna eget dolor interdum non tristique lorem egestas. In hac habitasse platea dictumst.
\end{abstract}
%
\keywords{Beamforming, Adaptive beamforming, Capon, High resolution, Complexity, Sonar}

\newpage
\section{Contents}

Lots of cryptic stuff...

\section*{Acknowledgements}
\addcontentsline{toc}{section}{Acknowledgements}

Yeah whatever... \cite{JohnDoe10}

% I use a custom command here since biblatex is different to bibtex in syntax here.
% Feel free to mail me complaints. :)
% \printreferences

% If bibtex:
\bibliography{references}

\the\abovecaptionskip
\the\belowcaptionskip
% If biblatex (file is loaded in preamble):
% \printreferences

\end{document}

### 2012-06-26 HRUIonGPU/jrnl_article.tex
Started incorparated comments from sverre++

@@
 %
-%\usepackage{cite}
+\usepackage{cite}
 % cite.sty was written by Donald Arseneau
@@
 \begin{IEEEkeywords}
-Adaptive beamforming, Capon beamformer, Graphics Processing Unit, GPGPU, Parallel Computing, Ultrasound Imaging.
+Adaptive Beamforming, Capon Beamformer, Minimum Variance Beamforming, Graphics Processing Unit, GPGPU, Parallel Computing, Ultrasound Imaging.
 \end{IEEEkeywords}
@@
 \IEEEPARstart{C}{apon} adaptive beamforming applied in a medical ultrasound setting has recently shown promising results in several studies \cite{Synnevag2007, Austeng2008, Vignon2008, Viola, Mehdizadeh2012}.
-Although the method introduce some interesting trade-offs in addition to just improved resolution, like maintaining image resolution combined with higher framerates, smaller probes, or deeper penetration \cite{Synnevag2009}, widespread use has been limited by the tremendous computational complexity associated with the method \cite{So2011}.
+Although the method introduces some interesting trade-offs in addition to just improved resolution, like maintaining image resolution in combination with higher framerates, smaller probes, or deeper penetration \cite{Synnevag2009}, widespread use has been limited by the tremendous computational complexity associated with the method \cite{So2011}.
 In this paper we introduce the first implementation of the Capon beamformer exhibiting real-time performance when applied in a typical cardiac ultrasound setting.
-To accomplish this task we make use of the massive parallel processing power found in modern graphics processing units (GPU's), combined with novel methods to reduce the computational complexity as the problem size grows.
-Earlier attempts on real time Capon beamforming have been focused around radar and passive sonar systems using intricate systolic array processors \cite{McWhirter1989, Moonen1993, Sinha2002}.
-A systolic array processor is custom made hardware, and even if the computational throughput might be higher for selected problems compared with a GPU, the cost related to development and flexibility is considered much higher.
+To accomplish this task we make use of the massive, and easy accessed, parallel processing power found in modern graphics processing units (GPU's), combined with novel methods to reduce the computational complexity as the problem size grows.
+
+Previous attempts on implementing a parallel Capon beamformer have mainly been focused around radar and passive sonar systems using custom made systolic array processors \cite{McWhirter1989, Moonen1993, Sinha2002}.
+%A systolic array processor is custom made hardware, and even if the computational throughput might be higher for selected problems compared with a GPU, the cost related to development and flexibility is considered much higher.
 For medical ultrasound, Chen et al.
-\cite{Chen2011a, Chen2011} have previously implemented the Capon beamformer on a GPU, however their implementation was based on real data, where the Capon beamformer assumes complex.
-Using real data will results in symmetric windows, and even if the number of samples they managed to process per second was high, both the number of floating point operations per sample and the imaging region is reduced using real data compared with use of complex data.
+\cite{Chen2011a, Chen2011} have implemented a Capon beamformer on the GPU that produced real apodization weights from real data.
+Since a real function is symmetric in frequency space the Capon beamformer is not allowed to shift the main lobe under this constraint, and therefore the full potential is not reached.
+Shifting the main lobe is an important property of the Capon beamformer in order for it to improve edge definitions and contrast in an ultrasound image.
+Another aspect of using complex data is the benefits of a lowered sampling rate, keeping the number of samples for a complete image at a minimum.
+All results in this paper is reported with the use of baseband data.
+
+%Using real data will results in symmetric windows, and even if the number of samples they managed to process per second was high, both the number of floating point operations per sample and the imaging region is reduced using real data compared with use of complex data.
 It is crucial to use complex data within the Capon beamformer framework, both for reducing the amount of data per image, and to allow for asymmetric windows.
 All results in this paper is reported with use of critically sampled complex data, both laterally and in range.
+(ikke helt kritisk samplet, ikke si det, bare at det er i baseband).
+
+In previous work (cite JOE) we have applied the GPU-based Capon beamformer in an active sonar system.
+For sonar the real time requirement is much lower than for ultrasound, which means that most array configurations can be handled by a straight forward implementation.
+%(Referer til JOE-artikkelen, få med beskrivelse av hva som blir anerledes for ultralyd.)
 
@@
 
-\begin{figure*}[t!]
-\centerline{\subfloat[]{\includegraphics[width=3in]{gfx/beamforming_mv_lowres.png} \label{fig:mvbf}}
-\hfil
-\subfloat[]{\includegraphics[width=3in]{gfx/gpu_layout_lowres.png}%
-\label{fig:gpulayout}}}
-\caption{a) Beamforming with adaptive weights $\vec{w}[n]$.
-Impinging signal is demodulated and aligned in phase by applying steering and focusing delays $\Delta_m[n]$.
-Capon adaptive weights are then calculated based on the in-phase data, and finally the output is formed by a weighted coherent sum.
-b) Calculation of Capon adaptive weights mapped to GPU architecture.
-A recorded echo across the array for a given range and angle (selected cell) is assigned to a group of $L$ threads that can run independently of all other groups.
-Note that several samples can be processed per block of threads if $L$ is small.
-The task of these $L$ threads is to calculate the Capon beamformer output in the three depicted steps.}
-\label{fig:fig1}
-\end{figure*}
-
 \subsection{Capon Beamforming}
-When sensing the world around us with an array of elements, we perform a spatio-temporal filtering operation known as beamforming.
+
+Bruk riktig terminologi. $N_{avg}$ er K. Antall sub-arrayer må hete noe annet.
+
+Presis på hvilke formler vi skal implementere.
+
+De skal forstå MV etter denne seksjonen.
+
+Ton ned introduksjonen, og sett opp akkurat de ligningene vi implementerer.
+
+(Skriv om den introduserende setningen, start rett på linje to, Kombinere data fra mange sensorer, Leseren kan Ultralyd ikke gpu)
+
 The conventional form of beamforming is delay-and-sum (DAS), where the signal recorded on each element, $x_m[n]$, is delayed to broadside with an appropriate delay $\Delta_m[n]$.
@@
 \begin{align}
-z[n] = \sum_{m = 0}^{M-1}w_m^*x_m[n - \Delta_m[n]] = \vec{w}^H\vec{x}[n].
+z[n] = \sum_{m = 0}^{M-1}w_m^*x_m[n - \Delta_m[n]] = \vec{w}^H\vec{x}[n]. \label{eq:z}
 \end{align}
@@
 For a complex data vector $\vec{x}[n] \in \mathbb{C}^M$, we have for an adaptive beamformer a data-dependent set of weights, $\vec{w}(\vec{x}[n])$, that when applied on $\vec{x}[n]$ is the best approximation to achieving the selected criteria.
+
+(Vi skal implementer dette. vis bare vektene. Dropp ligning 2, siter, bruk R-hat med en gang. Få med den endelige feilen )
+
+\begin{figure}
+\centerline{\includegraphics[width=3in]{gfx/beamforming_mv_lowres.png}}
+\caption{Beamforming with adaptive weights $\vec{w}[n]$.
+Impinging signal is demodulated and aligned in phase by applying steering and focusing delays $\Delta_m[n]$.
+Capon adaptive weights are then calculated based on the in-phase data, and finally the output is formed by a weighted coherent sum.}
+\label{fig:mvbf}
+\end{figure}
+
 With the Capon beamformer, adaptive weights are found by solving the following minimization problem \cite{Capon1969}:
@@
 
+\begin{figure}
+\centerline{\includegraphics[width=3in]{gfx/gpu_layout_lowres.png}}
+\caption{Calculation of Capon adaptive weights mapped to GPU architecture.
+A recorded echo across the array for a given range and angle (selected cell) is assigned to a group of $L$ threads that can run independently of all other groups.
+Note that several samples can be processed per block of threads if $L$ is small.
+The task of these $L$ threads is to calculate the Capon beamformer output in the three depicted steps.}
+\label{fig:gpulayout}
+\end{figure}
+% Få med at vi processerer hele bildet på en gang. Forklar forskjellen med å prosessere en beam av gangen. Hva med MLA?
+% Illustrer tråder og tider for hvordan processeringen går.
+% Vi snakker om mottak-stråleformeren.
+
 %\begin{figure}
@@
 \subsection{GPU Compute Model}
+
+(Ta med mer banale ting, dette er GPU for ultralydfolk)
+
+Etter denne så må de forstå GPU og hvorfor MV passer.
+
 In this section we briefly discuss the principle behind GPU computing, laying the basis for further discussion on parallel implementation strategies.
@@
 As mentioned, widespread use of Capon adaptive beamforming has been limited by its computational complexity.
-However, the calculation of Capon weights is well suited for a GPU implementation, since the calculations are both independent across pixels, and partially intra-pixel.
-Based on memory dependencies it is trivial to recognise that one block of threads should process at least one pixel in the final image.
-Continuing, on a pixel-level, the algorithm can be broken down into three main steps that to some extent can be further divided into a small set of parallel processes.
+However, if we analyse the beamformer in (\ref{eq:z}) in combination with (\ref{eq:w}) and (\ref{eq:R}), we see that each calculation of a set of Capon weights is independent across the output $z[n]$.
+As shown in Figure \ref{fig:gpulayout}, a first level of granularity is therefore to divide the image into a grid, where the output in each cell can be calculated independently of all others.
+Further, based on memory dependencies and to hold memory consumption per thread at a minimum, one should recognise that one block of threads should process at least one output in the final image.
+Continuing with the calculation of one weight vector, the algorithm can be broken down into three main steps that to some extent can be further divided into a small set of parallel processes.
 That is, for each data vector $\vec{x}$ in the image:
-\begin{enumerate}
-\item Calculate the sample covariance matrix $\mat{\hat{R}}$ as in (\ref{eq:R})
-\item Solve the linear system of equations $\mat{\hat{R}}\vec{b} =\vec{ 1}$
-\item Calculate $\vec{w} = \vec{b}/\vec{1}^H\vec{b}$ and the amplitude $\vec{w}^H\vec{x}$
-\end{enumerate}
-The computational complexity of these three steps, with $L = M/2$, is $O(L^3)$, $O(L^3)$, and $O(L)$ respectively.
-For larger $L$'s, the first two steps quickly becomes unmanageable, and real time performance can only be achieved by exploring novel solutions to reduce complexity.
-
-(The three steps have, for modularity, been implemented using three different kernels. Joining all kernels into one could possible lead to an increase in speed.)
+\begin{align*}
+\begin{array}{l}
+\text{1) Calculate the sample covariance matrix } \mat{\hat{R}} \text{ as in (\ref{eq:R})}\\
+\text{2) Solve the linear system of equations } \mat{\hat{R}}\vec{b} =\vec{ 1}\\
+\text{3) Calculate } \vec{w} = \vec{b}/\vec{1}^H\vec{b} \text{ and the amplitude } \vec{w}^H\vec{x}
+\end{array}
+\end{align*}
+%\begin{enumerate}
+%\item Calculate the sample covariance matrix $\mat{\hat{R}}$ as in (\ref{eq:R})
+%\end{enumerate}
+%\begin{enumerate}
+%\item[2)] Solve the linear system of equations $\mat{\hat{R}}\vec{b} =\vec{ 1}$
+%\end{enumerate}
+%\begin{enumerate}
+%\item[3)] Calculate $\vec{w} = \vec{b}/\vec{1}^H\vec{b}$ and the amplitude $\vec{w}^H\vec{x}$
+%\end{enumerate}
+The computational complexity of these three steps is $O(L^3)$, $O(L^3)$, and $O(L)$ respectively if $L = M/2$.
+For large $L$'s, the first two steps quickly becomes unmanageable, and real time performance can only be achieved by exploring novel solutions to reduce complexity.
+
+%(The three steps have, for modularity, been implemented using three different kernels. Joining all kernels into one could possible lead to an increase in speed.)
 
@@
 \subsection{Calculating Multiple Sample Covariance Matrices}
-The straight forward sum in (\ref{eq:R}) has complexity $O(L^2KN) = O(L^3)$ if $L$ is a small fraction of $M$.
+
+The straight forward sum in (\ref{eq:R}) has complexity $O(L^2KN) = O(L^3)$ if the number of subarrays $K$ is small.
 This is actually the same as for inverting a $L \times L$ matrix.
 However, little attention has been devoted to this step in the literature until now.
-Note that for a typical array configuration, N is usually small compared to L and M.
-Hence, for a small $L$ the construction of $\mat{\hat{R}}$ is dominated by $M$, since $K = M - L + 1$.
+Note that for a typical array configuration we typically smooth more in space than in range.
+Hence, N is usually small compared to $K$.
+For short subarrays the construction of $\mat{\hat{R}}$ is dominated by the number of elements $M$, since $K = M - L + 1$.
 From (\ref{eq:R}) it is obvious that calculation of one element in $\mat{\hat{R}}$ is independent of all the others.
 A natural level of granularity would then be to assign one thread to each element of $\mat{\hat{R}}$.
-This will minimize global reads and writes needed per thread, but unfortunately the maximum number of threads per block puts an upper limit on $L$, $L_{max}=32$, if we want to avoid block-to-block synchronization.
+This will minimize global reads and writes needed per thread, but unfortunately the maximum number of threads per block puts an upper limit on subarray length $L$, $L_{max}=32$, if we want to avoid block-to-block synchronization.
 
@@
 
+Skriv om utfordringene om å lage en solver på GPU. Ikke dra inn UHDU.
+
 For the smallest systems a direct solver using one thread per system is preferable. For this purpose we have implemented Cramer's rule for $3 \times 3$ (, $2 \times 2$?  and $4 \times 4$?) matrices.
@@
 
-\subsection{Iterative Capon}
-The idea behind iterative Capon is to perform Capon using small subarrays in several iterations until there is only two subarrays to sum.
+%\subsection{Iterative Capon}
+%The idea behind iterative Capon is to perform Capon using small subarrays in several iterations until there is only two subarrays to sum.
 In this way we fix the size of L, hence the complexity of solving $\mat{R}\vec{b} = \vec{1}$.
@@
 
-\subsection{Toeplitz-based solver} 
-(cite new article). Discuss it.
-
 \section{Results}\label{sec:res}
+
+% stolpediagram, med de ulike metodenes kjøretid oppehverandre.
 
@@
 \label{fig:benchUS64}}}
-\caption{Benchmark of GPU Capon beamforming for a cardiac image covering a 70 degrees sector.
+\caption{(Change into bar plot?) Benchmark of GPU Capon beamforming for a cardiac image covering a 70 degrees sector.
 Execution time is plotted as a function of subarray length $L$.
@@
 \begin{figure*}[!t]
-\centerline{\subfloat[]{\includegraphics[width=3.5in]{gfx/benchmark_us_32_lr.png} \label{fig:benchUS32}}
+\centerline{\subfloat[]{\includegraphics[width=3.5in]{gfx/benchmark_us_32_lr.png} \label{fig:benchBS3}}
 \hfil
 \subfloat[]{\includegraphics[width=3.5in]{gfx/benchmark_us_64_lr.png}%
-\label{fig:benchUS64}}}
-\caption{(TODO: Add the right figure) Benchmark of GPU Beamspace Capon beamforming for a cardiac image covering a 70 degrees sector.
+\label{fig:benchBS5}}}
+\caption{(TODO: Add the right figure.
+Change it into a bar plot?) Benchmark of GPU Beamspace Capon beamforming for a cardiac image covering a 70 degrees sector.
 Execution time is plotted as a function of number of elements $M$.
@@
 (b) 5 beams.}
+\label{fig:benchBS}
+\end{figure*}
+
+\begin{figure*}[!t]
+\centerline{\subfloat[]{\includegraphics[width=2in]{gfx/mv_phantom.png} \label{fig:phantomDAS}}
+\hfil
+\subfloat[]{\includegraphics[width=2in]{gfx/mv_phantom.png} \label{fig:phantomMV}}
+\hfil
+\subfloat[]{\includegraphics[width=2in]{gfx/mv_phantom.png} \label{fig:phantomBS}}}
+\caption{(TODO: Add the right figures) Have graph showing cyst contrast as a function of different parameters instead?
+Cyst and point targets phantom a) Delay-and-sum, uniform weights.
+b) Capon ($M=?, L=?, N=?$) c) Beamspace Capon ($M=?, L=?, N=?, N_b=?$).}
+\label{fig:bench}
+\end{figure*}
+
+\begin{figure*}[!t]
+\centerline{\subfloat[]{\includegraphics[width=2in]{gfx/us_image.png} \label{fig:phantomDAS}}
+\hfil
+\subfloat[]{\includegraphics[width=2in]{gfx/us_image.png} \label{fig:phantomMV}}
+\hfil
+\subfloat[]{\includegraphics[width=2in]{gfx/us_image.png} \label{fig:phantomBS}}}
+\caption{(TODO: Add the right figures) Have graph showing cavity-wall contrast as a function of parameters instead?
+In-vivo cardiac ultrasound image a) Delay-and-sum, uniform weights.
+b) Capon ($M=?, L=?, N=?$) c) Beamspace Capon ($M=?, L=?, N=?, N_b=?$).}
 \label{fig:bench}
@@
 
-Present images, both simulated and in-vivo, processed using beamspace Capon and iterative Capon. Comment on the differences between the two and on how much they deviates from the full method.
-Show beampatterns for the two compared with the full method. Show executing times for the two methods compared with the full method on the GPU.
+Present images, both simulated and in-vivo, processed using beamspace Capon. Comment on the differences between the two and on how much they deviates from the full method.
+%Show beampatterns for the two compared with the full method. 
+Show executing times for the two methods compared with the full method on the GPU.
 
@@
 How has the data been acquired?
+
+Ta med en serie med bilde. Må vise tydelig at det gir forbedring.
 
@@
 The target GPU has been Nvidia’s Quadro 6000, capable of 1Tflops.
+
+Toeplitz-based solver (cite new article). Discuss it.
 

### 2012-06-26 ECUA2012/gpu_capon_ecua2012.tex

@@
 Carl-Inge~Colombo~Nilsen & Department of Informatics, University of Oslo, Norway \\
-Andreas-Austeng & Department of Informatics, University of Oslo, Norway
+Andreas~Austeng & Department of Informatics, University of Oslo, Norway
 \end{tabular}

### 2012-07-11 HRUIonGPU/jrnl_article.tex
Update of HRUI article

@@
 %
-%\usepackage{url}
+\usepackage{url}
 % url.sty was written by Donald Arseneau. It provides better support for
@@
 %\IEEEPARstart{I}{ntroduction} to Capon adaptive beamforming, ultrasound imaging and GPU/CUDA.
-\IEEEPARstart{C}{apon} adaptive beamforming applied in a medical ultrasound setting has recently shown promising results in several studies \cite{Synnevag2007, Austeng2008, Vignon2008, Viola, Mehdizadeh2012}.
+\IEEEPARstart{C}{apon} adaptive beamforming \cite{Capon1969} applied in a medical ultrasound setting has recently shown promising results in several studies \cite{Synnevag2007, Austeng2008, Vignon2008, Viola, Mehdizadeh2012}.
 Although the method introduces some interesting trade-offs in addition to just improved resolution, like maintaining image resolution in combination with higher framerates, smaller probes, or deeper penetration \cite{Synnevag2009}, widespread use has been limited by the tremendous computational complexity associated with the method \cite{So2011}.
@@
 For medical ultrasound, Chen et al.
-\cite{Chen2011a, Chen2011} have implemented a Capon beamformer on the GPU that produced real apodization weights from real data.
-Since a real function is symmetric in frequency space the Capon beamformer is not allowed to shift the main lobe under this constraint, and therefore the full potential is not reached.
-Shifting the main lobe is an important property of the Capon beamformer in order for it to improve edge definitions and contrast in an ultrasound image.
-Another aspect of using complex data is the benefits of a lowered sampling rate, keeping the number of samples for a complete image at a minimum.
+\cite{Chen2011a, Chen2011} have implemented a Capon beamformer on the GPU that produced real apodization weights based on real data.
+Since a real function is symmetric in frequency space the Capon beamformer is not allowed to shift the main lobe under this constraint, and therefore its full potential is not reached.
+Shifting the main lobe is an important property of the Capon beamformer in order for it to improve edge definitions and contrast throughout an ultrasound image.
+Another aspect of using complex data is that eventhough the number of aritmetic operations per sample is increased we can now sample at the system bandwidth, keeping the number of samples for a full cardiac imaging sector at a minimum.
 All results in this paper is reported with the use of baseband data.
@@
 
-In previous work (cite JOE) we have applied the GPU-based Capon beamformer in an active sonar system.
-For sonar the real time requirement is much lower than for ultrasound, which means that most array configurations can be handled by a straight forward implementation.
+% denne må utelates i IUS-artikkelen
+In recent work (cite JOE) we have applied a GPU-based Capon beamformer in an active sonar system.
+For sonar the real time requirement is much lower than for ultrasound, which means that most common array configurations can be handled using the GPU without taking any additional measures.
+For ultrasound, and especially cardiac ultrasound, frame rate requirements increase to a level where even a high-end GPU can not do the processing required for a standard implementation of Capon beamforming.
+Since the execution time for a Capon beamformer highly depends on the number of elements in the given array, we need to find ways of reducing the computational complexity for high channel count ultrasound systems.
+Different approaches to low complexity Capon beamforming have recently been proposed \cite{Asl2012, Jensen2012}, but in this paper we have implemented the beamspace approach described by Nilsen and Hafizovic \cite{Nilsen2009} on the GPU to increase frame rates for high channel count systems.
 %(Referer til JOE-artikkelen, få med beskrivelse av hva som blir anerledes for ultralyd.)
@@
 In the next section we present background information about the Capon beamformer and the GPU computing model.
-In Section \ref{sec:meth} we go in detail about the parallel implementation of the Capon beamformer, before in Section \ref{sec:meth2} we introduce methods to apply as the problem size increases.
+In Section \ref{sec:meth} we go in detail about the parallel implementation of the Capon beamformer, before in Section \ref{sec:meth2} we introduce the method applied when the problem size increases.
 Benchmark results and resulting images are presented in Section \ref{sec:res}, and finally we discuss the results in Section \ref{sec:dis}, and draw conclusions in Section \ref{sec:con}.
@@
 
-Bruk riktig terminologi. $N_{avg}$ er K. Antall sub-arrayer må hete noe annet.
-
-Presis på hvilke formler vi skal implementere.
-
-De skal forstå MV etter denne seksjonen.
-
-Ton ned introduksjonen, og sett opp akkurat de ligningene vi implementerer.
-
-(Skriv om den introduserende setningen, start rett på linje to, Kombinere data fra mange sensorer, Leseren kan Ultralyd ikke gpu)
+\textbf{Bruk riktig terminologi. $N_{avg}$ er K. Antall sub-arrayer må hete noe annet. F.eks $N_L$}
+
+\textbf{Presis på hvilke formler vi skal implementere.}
+
+\textbf{De skal forstå MV etter denne seksjonen.}
+
+\textbf{Ton ned introduksjonen, og sett opp akkurat de ligningene vi implementerer.}
+
+\textbf{(Skriv om den introduserende setningen, start rett på linje to, Kombinere data fra mange sensorer, Leseren kan Ultralyd ikke gpu)}
 
 The conventional form of beamforming is delay-and-sum (DAS), where the signal recorded on each element, $x_m[n]$, is delayed to broadside with an appropriate delay $\Delta_m[n]$.
-The beamformer output, $z[n]$, is then the sum of all $M$ channels weighted with a factor $w_m$,
+The beamformer output, $z[n]$, is then the sum of all $M$ channels weighted with a per-element factor $w_m$,
 \begin{align}
-z[n] = \sum_{m = 0}^{M-1}w_m^*x_m[n - \Delta_m[n]] = \vec{w}^H\vec{x}[n]. \label{eq:z}
+z[n] = \sum_{m = 0}^{M-1}w_m^*x_m[n - \Delta_m[n]] = \vec{w}^H\vec{x}^{\Delta}[n]. \label{eq:z}
 \end{align}
@@
 
-The basic idea of an adaptive beamformer, as shown in Figure \ref{fig:mvbf}, is to calculate the weights on-the-fly, based on the data at hand, and on a selected optimization criteria in such a way that we minimize interference impinging from other directions than the point of focus.
-For a complex data vector $\vec{x}[n] \in \mathbb{C}^M$, we have for an adaptive beamformer a data-dependent set of weights, $\vec{w}(\vec{x}[n])$, that when applied on $\vec{x}[n]$ is the best approximation to achieving the selected criteria.
-
-(Vi skal implementer dette. vis bare vektene. Dropp ligning 2, siter, bruk R-hat med en gang. Få med den endelige feilen )
+The basic idea of the Capon adaptive beamformer, as shown in Figure \ref{fig:mvbf}, is to calculate a set of weights on-the-fly, based on the data at hand, in such a way that we minimize the beamformer output power while maintaining unit gain at broadside.
+The result being that interference impinging from other directions than the point of focus is suppressed.
+
+%\textbf{(Vi skal implementer dette. vis bare vektene. Dropp ligning 2, siter, bruk R-hat med en gang. Få med den endelige feilen )}
 
@@
 \centerline{\includegraphics[width=3in]{gfx/beamforming_mv_lowres.png}}
-\caption{Beamforming with adaptive weights $\vec{w}[n]$.
-Impinging signal is demodulated and aligned in phase by applying steering and focusing delays $\Delta_m[n]$.
-Capon adaptive weights are then calculated based on the in-phase data, and finally the output is formed by a weighted coherent sum.}
+\caption{Capon adaptive beamforming.
+Impinging signal is demodulated, and aligned in phase by applying steering and focusing delays $\Delta_m[n]$.
+Adaptive weights, $\vec{w}[n]$, are then calculated based on in-phase data, and finally the output is formed by a weighted coherent sum.}
 \label{fig:mvbf}
@@
 
-With the Capon beamformer, adaptive weights are found by solving the following minimization problem \cite{Capon1969}:
+The adaptive weights are therefore found by solving the following minimization problem:
 \begin{align}
@@
 The resulting sample covariance matrix might therefore be both inaccurate, and poorly conditioned.
-Another challenge with applying Capon to active systems is the effect of signal cancellation, where high correlation between signal and interference favours weights that in the end cancel out the signal \cite{Synnevag2009a}.
+Another challenge with applying Capon to active systems is the effect of signal cancellation, where high correlation between signal and interference favours weights that in the end cancel out the signal \cite{Reddy1987}.
 In this way, the beamformer output can be zero at broadside, even when there is signal present and the distortion-less criterion is applied.
@@
 \end{align}
-introduced by Synnev\aa{}g et al. \cite{Synnevag2007} has been applied in much of the resent literature on Capon beamforming for medical ultrasound imaging \cite{Nilsen2009, Wang2009, Mehdizadeh2012}.
+introduced by Synnev\aa{}g et al.
+\cite{Synnevag2007} has been applied in much of the resent literature on Capon beamforming for medical ultrasound imaging \cite{Nilsen2009, Wang2009, Mehdizadeh2012}.
+Recently we have also found that this way of doing diagonal loading also has been applied to HF and VHF antenna processing \cite{Featherstone1997}.
+
+Finally we calculate the beamformer output by combining the results from all subarrays.
+\begin{align}
+z[n] = \sum_{l=0}^{l=N_L-1} .
+\end{align}
 
 \begin{figure}
-\centerline{\includegraphics[width=3in]{gfx/gpu_layout_lowres.png}}
+\centerline{\includegraphics[width=3in]{gfx/gpu_layout_vertical_lr.png}}
 \caption{Calculation of Capon adaptive weights mapped to GPU architecture.
@@
 
-(Ta med mer banale ting, dette er GPU for ultralydfolk)
-
-Etter denne så må de forstå GPU og hvorfor MV passer.
+\textbf{(Ta med mer banale ting, dette er GPU for ultralydfolk)}
+
+\textbf{(Etter denne så må de forstå GPU og hvorfor MV passer.)}
 

### 2012-07-16 HRUIonGPU/jrnl_article.tex

@@
 
-\subsection{Capon Beamforming}
-
-\textbf{Bruk riktig terminologi. $N_{avg}$ er K. Antall sub-arrayer må hete noe annet. F.eks $N_L$}
-
-\textbf{Presis på hvilke formler vi skal implementere.}
-
-\textbf{De skal forstå MV etter denne seksjonen.}
-
-\textbf{Ton ned introduksjonen, og sett opp akkurat de ligningene vi implementerer.}
-
-\textbf{(Skriv om den introduserende setningen, start rett på linje to, Kombinere data fra mange sensorer, Leseren kan Ultralyd ikke gpu)}
+%\subsection{Capon Beamforming}
+
+%\textbf{Bruk riktig terminologi. $N_{avg}$ er K. Antall sub-arrayer må hete noe annet. F.eks $N_L$}
+
+%\textbf{Presis på hvilke formler vi skal implementere.}
+
+%\textbf{De skal forstå MV etter denne seksjonen.}
+
+%\textbf{Ton ned introduksjonen, og sett opp akkurat de ligningene vi implementerer.}
+
+%\textbf{(Skriv om den introduserende setningen, start rett på linje to, Kombinere data fra mange sensorer, Leseren kan Ultralyd ikke gpu)}
 
@@
 The basic idea of the Capon adaptive beamformer, as shown in Figure \ref{fig:mvbf}, is to calculate a set of weights on-the-fly, based on the data at hand, in such a way that we minimize the beamformer output power while maintaining unit gain at broadside.
-The result being that interference impinging from other directions than the point of focus is suppressed.
-
+The result being that interference impinging from other directions than the point of focus is suppressed \cite{Synnevag2007}.
 %\textbf{(Vi skal implementer dette. vis bare vektene. Dropp ligning 2, siter, bruk R-hat med en gang. Få med den endelige feilen )}
+
+%Formally stated, the adaptive weights are found by solving the following minimization problem:
+%\begin{align}
+%&\min_{\vec{w}} E\{|z[n]|^2\} = \min_{\vec{w}}\vec{w}^H\mat{R}\vec{w}\label{eq:minz}\\
+%&\text{subjected to } \vec{w}^H\vec{a} = 1,\label{eq:constraint}
+%\end{align}
+%where $\mat{R} = E\{\vec{x}[n]\vec{x}[n]^H\}$ is the spatial covariance matrix, and $\vec{a} = \left[ \begin{array}{cccc} e^{-jk_{\tilde{x}}\tilde{x}_0}& e^{-jk_{\tilde{x}}\tilde{x}_1}& ...& e^{-jk_{\tilde{x}}\tilde{x}_{M-1}}\end{array}\right] ^T$ is a far-field steering vector towards angle $\theta = \arcsin(\frac{\lambda k_{\tilde{x}}}{2\pi})$.
+%The constraint in (\ref{eq:constraint}) is there to avoid the trivial solution $\vec{w} = 0$, and a distorted respons in the direction of $\vec{a}$. 
+%The solution to the minimization problem in (\ref{eq:minz}) and (\ref{eq:constraint}) is
 
@@
 
-The adaptive weights are therefore found by solving the following minimization problem:
-\begin{align}
-&\min_{\vec{w}} E\{|z[n]|^2\} = \min_{\vec{w}}\vec{w}^H\mat{R}\vec{w}\label{eq:minz}\\
-&\text{subjected to } \vec{w}^H\vec{a} = 1,\label{eq:constraint}
+The solution to this optimization problem is given by the following expression:
+\begin{align}\label{eq:w}
+\vec{w}[n] = \frac{\mat{\hat{R}}[n]^{-1}\vec{a}}{\vec{a}^H\mat{\hat{R}}[n]^{-1}\vec{a}},
 \end{align}
-where $\mat{R} = E\{\vec{x}[n]\vec{x}[n]^H\}$ is the spatial covariance matrix, and $\vec{a} = \left[ \begin{array}{cccc} e^{-jk_{\tilde{x}}\tilde{x}_0}& e^{-jk_{\tilde{x}}\tilde{x}_1}& ...& e^{-jk_{\tilde{x}}\tilde{x}_{M-1}}\end{array}\right] ^T$ is a far-field steering vector towards angle $\theta = \arcsin(\frac{\lambda k_{\tilde{x}}}{2\pi})$.
-Thus, we minimize the output power, while maintaining unit gain in the look direction.
-The constraint in (\ref{eq:constraint}) is there to avoid the trivial solution $\vec{w} = 0$, and a distorted respons in the direction of $\vec{a}$.
-The solution to the minimization problem in (\ref{eq:minz}) and (\ref{eq:constraint}) is
-\begin{align}\label{eq:w}
-\vec{w}[n] = \frac{\mat{R}[n]^{-1}\vec{a}}{\vec{a}^H\mat{R}[n]^{-1}\vec{a}},
-\end{align}
-where $\vec{a} = [1_0 1_1... 1_{M-1}]^T = \vec{1}$ when $\vec{x}$ is pre-delayed to broadside.
-
-In practice, the covariance matrix has to be approximated by the sample covariance matrix, 
-\begin{align}
-\mat{\hat{R}}[n] = \vec{x}[n]\vec{x}[n]^H.
-\end{align} 
-In passive systems, assuming stationary statistics, the sample covariance matrix is typically averaged over all available samples to improve on the estimate \cite{Krima}.
-For an active, pulsed system like ultrasound imaging, the available data from which $\mat{R}$ can be estimated is quite limited.
+where $\vec{a} = [1_0, 1_1, ..., 1_{M-1}]^T = \vec{1}$ when $\vec{x}$ is pre-delayed to broadside, and $\mat{\hat{R}}$ is the sample covariance matrix that can be calculated as $\vec{x}[n]\vec{x}[n]^H$.
+%\begin{align}
+%\mat{\hat{R}}[n] = \vec{x}[n]\vec{x}[n]^H.
+%\end{align} 
+Note that this calculation only use a single sample. Typically the sample covariance matrix has to be averaged over several samples in order to ensure numerical stability and statistical robustness.
+
+In passive systems, assuming stationary statistics, the sample covariance matrix is typically averaged over a long sequence of samples\cite{Krima}.
+For an active, pulsed system like ultrasound imaging, the available data from which $\mat{\hat{R}}$ can be estimated is quite limited.
 The resulting sample covariance matrix might therefore be both inaccurate, and poorly conditioned.
 Another challenge with applying Capon to active systems is the effect of signal cancellation, where high correlation between signal and interference favours weights that in the end cancel out the signal \cite{Reddy1987}.
-In this way, the beamformer output can be zero at broadside, even when there is signal present and the distortion-less criterion is applied.
-In order to get a well conditioned $\mat{\hat{R}}$, avoid signal cancellation and to get DAS-like speckle, $\mat{\hat{R}}$ has to be averaged over $L\le M/2$ long subarrays and $N$ time samples \cite{Synnevag2007, Synnevag2007a}.
+In this way, the beamformer output can be to low at broadside, even when there is signal present and the distortion-less criterion is applied.
+
+In order to get a well conditioned $\mat{\hat{R}}$, avoid signal cancellation and to get DAS-like speckle, $\mat{\hat{R}}$ has to be averaged over $L\le M/2$ long subarrays and $N_K \sim \tau/T_s$ time samples \cite{Synnevag2007, Synnevag2007a}.
+Where $\tau$ is the pulse length in seconds and $T_s$ is the sampling period.
+We then get
 \begin{align}
-\mat{\hat{R}}[n] = \frac{1}{NK}\sum_{n'=n-N_{avg}}^{n+N_{avg}} \sum_{l=0}^{K-1} \vec{x}_l[n']\vec{x}_l[n']^H,\label{eq:R}
+\mat{\breve{R}}[n] = \frac{1}{N_LN_K}\sum_{n'=n-K}^{n+K} \sum_{l=0}^{N_L-1} \vec{x}_l^{\Delta}[n']\vec{x}_l^{\Delta}[n']^H,\label{eq:R}
 \end{align}
-where $K = M-L+1$, $N = 2N_{avg} + 1\sim \tau/T_s$ ($\tau$ is the pulse length and $T_s$ is the sampling period) and $\vec{x}_l$ is the $l^\text{th}$ subarray $[x_l[n], \dotso x_{l+L}[n]]$.
-Finally, $\mat{\hat{R}}$ is loaded with diagonal factor $\epsilon$, 
+where $N_L = M-L+1$, $N_K = 2K + 1$, and $\vec{x}_l^{\Delta}$ is the $l^\text{th}$ subarray $[x_l^{\Delta}[n], \dotso, x_{l+L-1}^{\Delta}[n]]$.
+The dimension of the covariance matrix will after this smoothing be $L \times L$.
+Hence, the size of $\vec{w}[n]$ and $\vec{a}$ in (\ref{eq:w}) needs to be $L \times 1$ when the covariance matrix in (\ref{eq:R}) is used.
+Finally, the covariance matrix is loaded with a diagonal factor $\epsilon$ to ensure numerical stability, 
 \begin{align}
-\mat{\hat{R}} = \mat{\hat{R}} + \epsilon\mat{I}.
+\mat{\hat{R}} = \mat{\breve{R}} + \epsilon\mat{I}.
 \end{align}
-A weighting proportional to the output power is often used to reduce the need of parameter adjustments, and the trace of $\mat{\hat{R}}$, 
+A weighting proportional to the output power is often used to reduce the need of parameter adjustments, and the trace of $\mat{\breve{R}}$, 
 \begin{align}\label{eq:diag}
-\epsilon &= d*tr\{\mat{\hat{R}}\}, \\
-d &= 1/L
+\epsilon &= d*tr\{\mat{\breve{R}}\}, \\
+d &\approx 1/L
 \end{align}
@@
 
-Finally we calculate the beamformer output by combining the results from all subarrays.
+The beamformer output is in the end calculated by combining all $N_L$ subarrays weighted with the same set of adaptive weights.
 \begin{align}
-z[n] = \sum_{l=0}^{l=N_L-1} .
+z[n] &= \vec{w}[n]^H \sum_{l=0}^{N_L-1} \vec{x}_l^{\Delta}[n] \label{eq:z_mv}\\
+&= (\vec{w}[n]^H \ast{} [1_0, ..., 1_{N_L-1}])\vec{x}^{\Delta}[n] \label{eq:z_mv2}.
 \end{align}
+Note how we can either select to sum all subarrays first in (\ref{eq:z_mv}), or accumulate the right weighting per element in (\ref{eq:z_mv2}).
 

### 2012-07-17 Info/OutlineHints/outline_hints.tex
Added article outline tips.

--- 2012-04-11 GPUCapon/gpu_capon.tex
@@
 
-\title{A GPU Implementation of Capon for an Active Sonar System}
+\title{Outline hints}
 
-\author{Jo~Inge~Buskenes, %
-        Jon~Petter~Aasen, %
-        Carl-Inge~Colombo~Nilsen%
+\author{Authors%
 \IEEEcompsocitemizethanks{\IEEEcompsocthanksitem All authors are with the Department
@@
 % The paper headers
-\markboth{IEEE Journal of Oceanic Engineering}%
-{A GPU Implementation of Capon for an Active Sonar System}
+\markboth{Some journal}%
+{Outline hints}
 
@@
 \begin{IEEEkeywords}
-Beamforming, adaptive beamforming, sonar, active, complexity. \todo{not entirely true?}
+Outline hints
 \end{IEEEkeywords}
@@
 
+\section{Prerequisites}
+
+\begin{itemize}
+\item Research articles. Take note of journals aim for these. Perhaps take impact factor into account.
+\item For a given journal, make note of typical article and section lengths.
+\item Notice language and outline of such articles as well.
+\end{itemize}
+
+IMRAD style:
+
+\section{Abstract}
+
+\begin{itemize}
+\item Past tense.
+\end{itemize}
+
 \section{Introduction}
 
-\todopar{Case:
 \begin{itemize}
-\item UFFC(?)
-\item Ultrasound. Arrays up to 128 channels. Processing power vital. What's the best way to implement 
-\item Other articles evaluates Capon image quality, and some have shown that speedups can be achieved by reducing the complexity of the traditional Capon algorithm or by implementing specific versions on a GPU.
-\item This article compare a few promising ways to implement Capon in general, in light of general complexity and suitability for implementation on a CPU and GPU.
+\item Establish audience, and outline what they need to know.
+\item Purpose: What the audience need to know in order to understand the findings of the paper.
+\item Reason: Why is the research in this field necessary, and where does this individual contribution fit in?
+\item Active voice.
+\item Present tense, as focus will be put on the problem at hand, and respect should be paid to established knowledge.
 \end{itemize}
-}
 
+\begin{enumerate}
+\item Nature and scope of the problem, with all possible clarity. Motivate!!! Why the subject was chosen, why it is important.
+\item Brief review of present litterature for orientation purposes. 
+\item Method of investigation? Optionally, why was it chosen? Mediate our way of thinking!
+\item Principal results. Some overlap with abstract ok.
+\item Principal conclustions suggested by the results (don't keep the reader in suspense). Some overlap with abstract ok.
+\end{enumerate}
 \begin{itemize}
-\item As briefly as possible, introduce beamforming.
-\item Adaptive beamformer's potential lies in its ability to suppress interference power
-\item Why adaptive beamformers struggle in active sonar systems.
-Correlated noise, robustification kills the adaptive potential.
-Quite computationally intensive.
-Constraints must be applied in one way or another - parameters must be tuned.
-\item The three main ways to implement Capon \cite{Capon1969} is \todo{not entirely true?}
-\begin{itemize}
-\item Traditional way: Build $\eR$ and solve $\w = \frac{\Ri\a}{\a\H\Ri\a}$. 
-\item Iterative methods: Woodbury, Conjugated gradients
-\item Beamspace
-\item LCA (or other methods, as reference)
-\end{itemize}
-\item Choice of robustification parameters $L,M,\Delta$ largely impacts what the ``optimal'' way of implementing Capon 
-\item In most imaging applications 
-Outline:
-\begin{itemize}
-\item Speeding up: Build $\eR$ and solve $\w = \frac{\Ri\a}{\a\H\Ri\a}$. 
-\item Iterative methods: Woodbury (Sherman Morrison), Conjugated gradients - show performance of GPU performance.
-\item Woodbury: Must take the average over many samples in space and time - lots of (small) outerproducts. Can be used if not a lot of averaging is required.
-\item Beamspace
-\item LCA (or other methods, as reference)
-\end{itemize}
+\item Funnel effect desirable: Importance of overall topic, summarize knowledge about an aspect of it, identify an uresolved question of that aspect, say how paper answers that question.
 \end{itemize}
@@
 
-\begin{align}
-z[n] = \sumb{m=0}{M-1} w_m[n]^*x_m[n-\Delta_m] = \w\H[n]\Xd[n]
-\end{align}
-where
-\begin{align}
-\w[n] = \bmat{w_0[n]\\w_1[n]\\\vdots\\w_{M-1}[n]} \quad \text{and} \quad\Xd[n] = \bmat{x_0[n]\\x_1[n]\\\vdots\\x_{M-1}[n]}.
-\end{align}
-
-Minimum variance distortionless resonse \cite{Capon1969}
-\begin{align}
-\underset{\w[n]}{\argmin}\, E\{ |z[n]|^2 \} = \underset{\w[n]}{\argmin}\, \w\H[n]\R[n]\w[n], 
-\end{align}
-subject to
-\begin{align}
-\w\H[n]\a = 1,
-\end{align}
-where
-\begin{align}
-\R[n] = E\{ \x\x\H \}.
-\end{align}
-If $\eR$ is the estimate of $\R$, the solution to the MVDR beamformer is
-\begin{gather*}
-\vec w[n] = \frac{\hat{\mat R}\,\!^{-1}[n]\vec a}{\vec a\H\hat{\mat R}\,\!^{-1}[n]\vec a} = \frac{\text{\raisebox{1.9pt}{$\vec\chi$}}[n]}{\vec a\H\text{\raisebox{1.9pt}{$\vec\chi$}}[n]}
-\end{gather*}
-A robust estimate of $\eR$ is found by
 \begin{itemize}
-\item Spatial averaging to decorrelate noise from signal. \emph{Subarray averaging} used here.
-\item Temporal averaging to ensure valid speckle statistics.
-\item \emph{Diagonal loading} to ensure  numerical stability prior to inversion.
-\item Choice of robustification parameters \cite{Synnevag2007}
+\item Describe and defend method sufficently to allow experiment to be repeated.
+\item Even though most readers will skip it since method is stated in introduction, results will not be taken seriously if method is insufficiently documented.
+\item Use subheadings to group methods, try to make them match the ones used in Results.
+\item Like a cookbook: ``How'', and ``how much'' should be precisely answered.
+\item Statistical knowledge - of established knowledge - should not be included. If deemed important, cite relevant litterature.
+\item When referencing, state the findings, but omit the reasons behind the findings.
+\item Passive voice appropriate, of use of ``we''.
+\item Past tense.
 \end{itemize}
 
-Here $\vec a$ is a \emph{steering vector} that if set to $\vec 1$ steers towards broadside, and $\hat{\mat R}$ is an estimate of the spatial covariance matrix, often computed as
-\begin{gather*}
-r_{ij}[n] = \frac{1}{K(2N_{\text{avg}}+1)}\sumb{k=0}{K-1} \sumb{n'=n-N_{\text{avg}}}{n+N_{\text{avg}}} x_{i+k}[n']\,x_{j+k}[n'],
-\end{gather*}
-where $r_{ij}$ is the $(i,j)$'th element of the covariance matrix, $K$ is the number of subarrays, $2N_{\text{avg}}+1$ is the number of samples to perform time averaging over, and $\vec x_i[n']$ is the data recorded by the $i$'th sensor at sample $n'$.
-
-\emph{Diagonal loading} chosen as \cite{Synnevag2007}
-\begin{align}
-\varepsilon = \Delta \cdot \tr\{\eR[n]\}
-\end{align}
-
-\subsection{Traditional approach}
-
-Let us first consider the \emph{computation of $\hat{\mat R}$}.
-Notice that when implementing the formula directly a lot of summations and multiplications will be repeated unnecessary.
-To illustrate this, consider for instance a case with $M=6$ sensors that are grouped into subarrays of length $L=3$ where all but one element overlap.
-Let us further assume that we need to average over 7 time samples, i.e.
-that $N_{\text{avg}} = 3$.
-
-These procedures are commutative, so the order in which we perform them does not matter.
-But for the sake of this example let us start by performing the time averaging.
-Assuming that $\vec x_i$ and $\vec x_j$ have a zero mean, a time averaged covariance coefficient $r_{ij}$ in $\hat{\mat R}$ may be obtained by (omitting the normalisation):
-\begin{align*}
-r_{ij}[n] &= \sumb{n'=n-N_{\text{avg}}}{n+N_{\text{avg}}} x_i[n']\,x_j[n'],
-\end{align*}
-which can be easily rewritten in a sliding window fashion
-\begin{align*}
-r_{ij}[n] &= r_{ij}[n-1] - x_i[n\text{-}N_\text{avg}\text{-}1]\,x_j[n\text{-}N_\text{avg}\text{-}1]
-                         + x_i[n\text{+}N_\text{avg}]\,x_j[n\text{+}N_\text{avg}].
-\end{align*}
-Hence, whenever $N_\text{avg}>1$ the sliding window approach will demand less instructions at the expense of a slightly higher memory consumption.
-The next thing to realise is that not all $r_{ij}[n]$ need to be computed.
-Since $r_{ji} = r^*_{ij}$, and the subarrays only span $L$ elements, we only compute an upper triangular band of R (marked \Green{green})
-\begin{figure}[H]
-% \begin{narrow}{-3.5cm}{-3.5cm}
-\includegraphics[width=\linewidth]{gfx/capon_build_R_full.eps}
-\caption{Building $\hat{\mat R}$: Symmetry and element coverage}\label{buildingRfull}
-% \end{narrow}
-\end{figure}
-The final step is to average all the subarray matrices to form $\hat{\mat R}$ by 
-{\renewcommand*{\arraystretch}{1.8}
-\begin{align*}
-\hat{\mat R}_\text{sub} &= {\color{Blue}\mat R_{\text{sub},1} + \mat R_{\text{sub},2} + \mat R_{\text{sub},3} + \mat R_{\text{sub},4}} \\
-&= \left[\;
-\begin{matrix}
-\color{Green}r_{00}+r_{11}+r_{22} & \color{Red}r_{01}+r_{12}+r_{23} & r_{02}+r_{13}+r_{24} \\
-                                       & \color{Green}r_{11}+r_{22}+r_{33} & \color{Red}r_{12}+r_{23}+r_{34} \\
-                                       &                                        & \color{Green}r_{22}+r_{33}+r_{44}
-\end{matrix}
-\;\right]
-\end{align*}}%
-Note that consecutive elements along the diagonals overlap on all but one covariance coefficient, hence a sliding window may also be used here.
-The subarray size $L$ is usually quite a bit larger than the time averaging window, so there is more to gain from implementing the sliding window here.\\\\
-
-{\bf Method complexity:}\\
-\begin{tabular}{l l}
-No optimizations            & $2M^2\,(2\Navg+1)$ \\
-Banded symmetric            & $L(2M-L+1)\,(2\Navg+1)$ \\
-Sliding, banded, symmetric  & $1.5L(2M-L+1)$ \\
-\end{tabular}\\\\
-
-\subsubsection{Computing $\text{\raisebox{1.9pt}{$\vec\chi$}} = \hat{\mat R}\,\!^{-1}\vec a$}
-
-There are two ways of finding $\text{\raisebox{1.9pt}{$\vec\chi$}} = \hat{\mat R}\,\!^{-1}\vec a$, either by solving the equation directly or by inverting $\hat{\mat R}$ first.
-Solving the equation is less computationally intensive than inverting $\hat{\mat R}$, and seems like the obvious choice.
-However, with iterative methods for updating $\eRi$ for each range pixel the situation is changed...
-
-\subsection{Iterative methods}
+\section{Results}
 
 \begin{itemize}
-\item Woodbury
-\item Conjugated gradients
+\item Digest results until only the vital pieces of information remains. Make what is important stand out.
+\item Drop variables that are neglible, and be sure to mention it.
+\item If negative aspects are meaningful, include them.
+\item Strive for clarity! This is the new piece of information offered to the world. Make sure it sees it.
+\item Results define everything. The introduction and methods explain why and how they come into existence, discusstion explains what they mean.
+\item Avoid redundancy. State data in figures/tables, not in text.
+\item Although dominated by passive voice, add some active voice to the mix too.
+\item Past tense on own results.
 \end{itemize}
 
-\subsection{Beamspace}
+\begin{enumerate}
+\item Overall description of the experiment, providing ``big picture'' without repeating technicalities of methods section.
+\item Results, with crystal clarity.
+\end{enumerate}
 
-\begin{itemize}
-\item Some minimum of background, and good references.
-\end{itemize}
+\section{Discussion}
 
-\section{Results/Discussion}
-
-\begin{itemize}
-\item Compare computational complexity vs. memory constraints for the various methods. Supporting plots.
-\item Relate this to whether the favourable architecture is a CPU or a GPU \todo{check ``when to gpu article''}
-\item Run benchmarks on some sonar dataset(s)
-\end{itemize}
+\begin{enumerate}
+\item Present principles, relationships, and generalizations shown by the Results. Discuss, not recapitulate.
+\item Point out exceptions, lack of correlation, or unsettled points.
+\item Show how results and interpretations agree/contrast with previously published work, and with own expectations.
+\item Discuss theoretical implications, as well as possible practical applications. This will largely affect how significant the article will be percieved.
+\item State conclusions are clearly as possible.
+\item Summarize evidence for each conclusion. Beware of causality leaps!
+\item Discussion pairs Introduction. Be sure to answer any question asked there. Don't recapitulate, but extend from where introduction left off.
+\item Be sure to emphasize that this is a ``small truth'', not the ``whole truth''.
+\item Active voice.
+\end{enumerate}
 
@@
 
+\begin{enumerate}
+\item Take a step back. Summarize experiment, relating it to the big picture.
+\item Discuss further work
+\item Did the results support / reject the hypothesis stated in the introduction?
+\item If the used should remember just one sentence from your article, state it here!
+\end{enumerate}
+
+\section{Revision}
+
 \begin{itemize}
-\item When to use what.
+\item Information selection wise?
+\item Information accurate?
+\item Consistency?
+\item Organization logical?
+\item Clear wording?
+\item Points stated brielfy, simply, directly?
+\item Grammar, spelling, word use, correct?
+\item Check figures and tables for errors/consistency.
+\item IEEE compliant?
 \end{itemize}
-
-% \ \\
-% \IEEEPARstart{T}{o} form images from a modern phased array sonar system the received wavefield is usually recorded, and then postprocessed by a digital beamformer.
-The beamformer applies delays and weights to the sensor channels, the beamformer adjusts the arrays spatial response to focus at one pixel at a time.
-such that signals emanating from regions of interest add constructively, while ensuring that noise and interference from other angles do not.
-% 
-% The imaging capabilities of a modern phased array sonar system depend on physical attributes such element response and array geometry, the transmitted signal, as well as the beamforming method being used on transmission and reception.
-Beamforming is the concept of applying delays and weights to the sensors channels to steer the arrays response to points of interest.
-
-% 
-% 
-% Outline:
-% \begin{itemize}
-% \item Capon's resonse when applying robustification
-% \item Choice of window functions makes little difference.
-% \item Steering and mainlobewidths have outer bounds.
-% \item Beamspace?
-% \item Chosen window plots - what may they tell us? Variance intensity values when using various windows.
-% \item Assymmetric windows?
-% \end{itemize}
-% 
-% 
-% \begin{align}
-% z[n] = \sumb{m=0}{M-1} w_m[n]^*x_m[n-\Delta_m] = \w\H[n]\x[n-\Delta_m]
-% \end{align}
-% 
-% 
-% \section{Methods}
-% 
-% Basically, we are working with a practical implementation of the Capon beamformer that computes a set of weights $\vec w$ for every single sample $n$ by solving:
-% \begin{gather*}
-% \vec w[n] = \frac{\hat{\mat R}\,\!^{-1}[n]\vec a}{\vec a\H\hat{\mat R}\,\!^{-1}[n]\vec a} = \frac{\text{\raisebox{1.9pt}{$\vec\chi$}}[n]}{\vec a\H\text{\raisebox{1.9pt}{$\vec\chi$}}[n]}
-% \end{gather*}%
-% where
-% % \newcommand\X{\text{\raisebox{2pt}{$\vec\chi$}}}
-% \begin{gather*}
-% \text{\raisebox{1.9pt}{$\vec\chi$}} = \hat{\mat R}\,\!^{-1}\vec a \qquad\qquad\text{is the solution to}\qquad\qquad \hat{\mat R}\text{\raisebox{1.9pt}{$\vec\chi$}} = \vec a.
-% \end{gather*}
-
-% 
-% , and maximum suppression of while ensuring that the beamformer digitally  before each of the pixels are estimated one at a time.
-The resolution and contrast of such a system will depend on the systems spatial response, which ideally should be narrow  be very sharp in the desired direction its ability to achieve  fundamental principle of forming a sonar image is to record the received wavefield,
-% 
-% image quality of a phased array sonar imaging systems depend on  the choice of weights to apply to each of the sensors are crucial. 
-% 
-% A modern phased array imaging system may be thought of as a spatial filter. To achieve the best possible performance, the 
-% 
-% resolution and contrast 
-% 
-% Adaptive beamformers have only recently been introduced in active sonar imaging.
-For a while they were considered unsuited for this purpose because the backscattered signal is largely correlated with the
-% 
-% 
-%
-
-%\begin{figure}[!t]
-%\centering
-%\includegraphics[width=2.5in]{myfigure}
-% where an .eps filename suffix will be assumed under latex, 
-% and a .pdf suffix will be assumed for pdflatex; or what has been declared
-% via \DeclareGraphicsExtensions.
-%\caption{Simulation Results}
-%\label{fig_sim}
-%\end{figure}
-
-% An example of a double column floating figure using two subfigures.
-% (The subfig.sty package must be loaded for this to work.)
-% The subfigure \label commands are set within each subfloat command, the
-% \label for the overall figure must come after \caption.
-% \hfil must be used as a separator to get equal spacing.
-% The subfigure.sty package works much the same way, except \subfigure is
-% used instead of \subfloat.
-%
-% \begin{figure*}[!t]
-% \centerline{\subfloat[Case I]\includegraphics[width=2.5in]{subfigcase1}%
-% \label{fig_first_case}}
-% \hfil
-% \subfloat[Case II]{\includegraphics[width=2.5in]{subfigcase2}%
-% \label{fig_second_case}}}
-% \caption{Simulation results}
-% \label{fig_sim}
-% \end{figure*}
-%
-% Note that often IEEE papers with subfigures do not employ subfigure
-% captions (using the optional argument to \subfloat), but instead will
-% reference/describe all of them (a), (b), etc., within the main caption.
-
-% An example of a floating table. Note that, for IEEE style tables, the 
-% \caption command should come BEFORE the table. Table text will default to
-% \footnotesize as IEEE normally uses this smaller font for tables.
-% The \label must come after \caption as always.
-%
-%\begin{table}[!t]
-%% increase table row spacing, adjust to taste
-%\renewcommand{\arraystretch}{1.3}
-% if using array.sty, it might be a good idea to tweak the value of
-% \extrarowheight as needed to properly center the text within the cells
-%\caption{An Example of a Table}
-%\label{table_example}
-%\centering
-%% Some packages, such as MDW tools, offer better commands for making tables
-%% than the plain LaTeX2e tabular which is used here.
-%\begin{tabular}{|c||c|}
-%\hline
-%One & Two\\
-%\hline
-%Three & Four\\
-%\hline
-%\end{tabular}
-%\end{table}
-
-% Note that IEEE does not put floats in the very first column - or typically
-% anywhere on the first page for that matter. Also, in-text middle ("here")
-% positioning is not used. Most IEEE journals use top floats exclusively.
-% However, Computer Society journals sometimes do use bottom floats - bear
-% this in mind when choosing appropriate optional arguments for the
-% figure/table environments.
-% Note that, LaTeX2e, unlike IEEE journals, places footnotes above bottom
-% floats. This can be corrected via the \fnbelowfloat command of the
-% stfloats package.
-
-\section{Conclusion}
-The conclusion goes here.
-
-% if have a single appendix:
-%\appendix[Proof of the Zonklar Equations]
-% or
-%\appendix  % for no appendix heading
-% do not use \section anymore after \appendix, only \section*
-% is possibly needed
-
-% use appendices with more than one appendix
-% then use \section to start each appendix
-% you must declare a \section before using any
-% \subsection or using \label (\appendices by itself
-% starts a section numbered zero.)
-%
 
@@
 
-% trigger a \newpage just before the given reference
-% number - used to balance the columns on the last page
-% adjust value as needed - may need to be readjusted if
-% the document is modified later
-%\IEEEtriggeratref{8}
-% The "triggered" command can be changed if desired:
-%\IEEEtriggercmd{\enlargethispage{-5in}}
-
-% references section
-
-\bibliographystyle{IEEEtran}
-\bibliography{../../Library/library}
-
-% End up doing this:
-% \begin{thebibliography}{1}
-% 
-% \bibitem{IEEEhowto:kopka}
-% H.~Kopka and P.~W. Daly, \emph{A Guide to {\LaTeX}}, 3rd~ed.\hskip 1em plus
-%   0.5em minus 0.4em\relax Harlow, England: Addison-Wesley, 1999.
-% 
-% \end{thebibliography}
-
-% biography section
-% 
-% If you have an EPS/PDF photo (graphicx package needed) extra braces are
-% needed around the contents of the optional argument to biography to prevent
-% the LaTeX parser from getting confused when it sees the complicated
-% \includegraphics command within an optional argument. (You could create
-% your own custom macro containing the \includegraphics command to make things
-% simpler here.)
-%\begin{biography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{mshell}}]{Michael Shell}
-% or if you just want to reserve a space for a photo:
-
-\vfill
-
-\begin{IEEEbiography}{Noname dude}
-He doesn't exist.
-\end{IEEEbiography}
-
-% if you will not have a photo at all:
-\begin{IEEEbiographynophoto}{Jo Inge Buskenes}
-Not in at the moment.
-\end{IEEEbiographynophoto}
-
-%\newpage
-
-% \begin{IEEEbiographynophoto}{Jo Inge Buskenes}
-% Not in at the moment.
-% \end{IEEEbiographynophoto}
-
-%\vfill
-
-% Can be used to pull up biographies so that the bottom of the last one
-% is flush with the other column.
-%\enlargethispage{-5in}
-
-\markboth{}%
-{}
-% \input{gfx/gfx_input}
+% \bibliographystyle{IEEEtran}
+% \bibliography{../../Library/library}
 

### 2012-07-30 HRUIonGPU/jrnl_article.tex
first update of tuffc article after the summer vecation

@@
 %\boldmath
-The abstract goes here. The abstract goes here. The abstract goes here. The abstract goes here. The abstract goes here. 
-The abstract goes here. The abstract goes here. The abstract goes here. The abstract goes here. The abstract goes here. 
-The abstract goes here. The abstract goes here. The abstract goes here. The abstract goes here. The abstract goes here. 
-The abstract goes here. The abstract goes here. The abstract goes here. The abstract goes here. The abstract goes here.  
-The abstract goes here. The abstract goes here. The abstract goes here. The abstract goes here. The abstract goes here. 
-The abstract goes here. The abstract goes here. The abstract goes here. The abstract goes here. The abstract goes here. 
-The abstract goes here. The abstract goes here. The abstract goes here. The abstract goes here. The abstract goes here. 
-The abstract goes here. The abstract goes here. The abstract goes here. The abstract goes here. The abstract goes here. 
-The abstract goes here. The abstract goes here. The abstract goes here. The abstract goes here. The abstract goes here. 
-The abstract goes here. The abstract goes here. The abstract goes here. The abstract goes here. The abstract goes here.  
-The abstract goes here. The abstract goes here. The abstract goes here. The abstract goes here. The abstract goes here. 
-The abstract goes here. The abstract goes here. The abstract goes here. The abstract goes here. The abstract goes here. 
+An amazing abstract goes here. An amazing abstract goes here. An amazing abstract goes here. An amazing abstract goes here.
+An amazing abstract goes here. An amazing abstract goes here. An amazing abstract goes here. An amazing abstract goes here.
+An amazing abstract goes here. An amazing abstract goes here. An amazing abstract goes here. An amazing abstract goes here.
+An amazing abstract goes here. An amazing abstract goes here. An amazing abstract goes here. An amazing abstract goes here.
+An amazing abstract goes here. An amazing abstract goes here. An amazing abstract goes here. An amazing abstract goes here.
+An amazing abstract goes here. An amazing abstract goes here. An amazing abstract goes here. An amazing abstract goes here.
+An amazing abstract goes here. An amazing abstract goes here. An amazing abstract goes here. An amazing abstract goes here.
+An amazing abstract goes here. An amazing abstract goes here. An amazing abstract goes here. An amazing abstract goes here.
+An amazing abstract goes here. An amazing abstract goes here. An amazing abstract goes here. An amazing abstract goes here.
+An amazing abstract goes here. An amazing abstract goes here. An amazing abstract goes here. An amazing abstract goes here.
 \end{abstract}
@@
 Since the execution time for a Capon beamformer highly depends on the number of elements in the given array, we need to find ways of reducing the computational complexity for high channel count ultrasound systems.
-Different approaches to low complexity Capon beamforming have recently been proposed \cite{Asl2012, Jensen2012}, but in this paper we have implemented the beamspace approach described by Nilsen and Hafizovic \cite{Nilsen2009} on the GPU to increase frame rates for high channel count systems.
+Different approaches to low complexity Capon beamforming have recently been proposed \cite{Asl2012, Jensen2012}, but in this paper we have selected to implemented the beamspace approach described by Nilsen and Hafizovic \cite{Nilsen2009} on the GPU to maintain high frame rates as the channel count increases.
 %(Referer til JOE-artikkelen, få med beskrivelse av hva som blir anerledes for ultralyd.)
@@
 \end{align}
-where $\vec{a} = [1_0, 1_1, ..., 1_{M-1}]^T = \vec{1}$ when $\vec{x}$ is pre-delayed to broadside, and $\mat{\hat{R}}$ is the sample covariance matrix that can be calculated as $\vec{x}[n]\vec{x}[n]^H$.
+where $\vec{a} = [1_0, 1_1, ..., 1_{M-1}]^T = \vec{1}_M$ when $\vec{x}$ is pre-delayed to broadside, and $\mat{\hat{R}}$ is the sample covariance matrix calculated as $\vec{x}[n]\vec{x}[n]^H$. 
 %\begin{align}
@@
 In order to get a well conditioned $\mat{\hat{R}}$, avoid signal cancellation and to get DAS-like speckle, $\mat{\hat{R}}$ has to be averaged over $L\le M/2$ long subarrays and $N_K \sim \tau/T_s$ time samples \cite{Synnevag2007, Synnevag2007a}.
-Where $\tau$ is the pulse length in seconds and $T_s$ is the sampling period.
-We then get
+Here, $\tau$ is the pulse length in seconds and $T_s$ is the sampling period.
+Continuing, we therefore calculate
 \begin{align}
@@
 \end{align}
-where $N_L = M-L+1$, $N_K = 2K + 1$, and $\vec{x}_l^{\Delta}$ is the $l^\text{th}$ subarray $[x_l^{\Delta}[n], \dotso, x_{l+L-1}^{\Delta}[n]]$.
+where $N_L = M-L+1$, $N_K = 2K + 1$ is the number of subarrays, and $\vec{x}_l^{\Delta}$ is the $l^\text{th}$ subarray $[x_l^{\Delta}[n], \dotso, x_{l+L-1}^{\Delta}[n]]$.
 The dimension of the covariance matrix will after this smoothing be $L \times L$.
-Hence, the size of $\vec{w}[n]$ and $\vec{a}$ in (\ref{eq:w}) needs to be $L \times 1$ when the covariance matrix in (\ref{eq:R}) is used.
+Hence, the size of $\vec{w}[n]$ and $\vec{a}$ in (\ref{eq:w}) has to be $L \times 1$ when the covariance matrix in (\ref{eq:R}) is used.
 Finally, the covariance matrix is loaded with a diagonal factor $\epsilon$ to ensure numerical stability, 
-\begin{align}
+\begin{align}\label{eq:diag}
 \mat{\hat{R}} = \mat{\breve{R}} + \epsilon\mat{I}.
@@
 A weighting proportional to the output power is often used to reduce the need of parameter adjustments, and the trace of $\mat{\breve{R}}$, 
-\begin{align}\label{eq:diag}
+\begin{align}\label{eq:diag_adapt}
 \epsilon &= d*tr\{\mat{\breve{R}}\}, \\
@@
 \begin{align}
-z[n] &= \vec{w}[n]^H \sum_{l=0}^{N_L-1} \vec{x}_l^{\Delta}[n] \label{eq:z_mv}\\
-&= (\vec{w}[n]^H \ast{} [1_0, ..., 1_{N_L-1}])\vec{x}^{\Delta}[n] \label{eq:z_mv2}.
+z[n] &= \frac{1}{N_L}\vec{w}[n]^H \sum_{l=0}^{N_L-1} \vec{x}_l^{\Delta}[n] \label{eq:z_mv}\\
+&= (\vec{w}[n]^H \ast{} \vec{1}_{N_L}/N_L)\vec{x}^{\Delta}[n] \label{eq:z_mv2}.
 \end{align}
 Note how we can either select to sum all subarrays first in (\ref{eq:z_mv}), or accumulate the right weighting per element in (\ref{eq:z_mv2}).
+
+% Få med at vi processerer hele bildet på en gang. Forklar forskjellen med å prosessere en beam av gangen. Hva med MLA?
+% Illustrer tråder og tider for hvordan processeringen går.
+% Vi snakker om mottak-stråleformeren.
+
+%\begin{figure}
+%\includegraphics[width=3in]{gfx/beamforming_mv_lowres.png}
+%\caption{}
+%\label{fig:mv}
+%\end{figure}
+
+%Building and calculating the inverse of $\hat{R}$ is computational demanding, limiting the methods accessibility. Following the innovation in GPU computing, this has now started to change. 
+%In this paper we introduce the first GPU implementation of the Capon beamformer capable of processing a 70 degree sector cardiac image from a 64 element phased array at interactive frame rates using both spatial and temporal smoothing.
+Both types of smoothing are important for maintaining delay-and-sum-like speckle and at the same time keep the high resolution property.
+This is achieved using complex base band data, which reduces the size of each range line to a minimum.
+In addition, complex data results in complex weights; hence this is a correct implementation of Capon beamforming with all degrees of freedom preserved.
+For high channel count systems we propose two ways of reducing computational complexity while maintaining
+
+\subsection{GPU Compute Model}
+
+%\textbf{(Ta med mer banale ting, dette er GPU for ultralydfolk)}
+
+%\textbf{(Etter denne så må de forstå GPU og hvorfor MV passer.)}
+
+In this section we briefly discuss the principle behind GPU computing, laying the basis for further discussion on parallel implementation strategies.
+CUDA by Nvidia has been our choice of GPU developer environment \cite{Nvidia2011}.
+However, even though CUDA has been used to implement the Capon beamformer, the challenges and proposed methods described in this paper apply for other massive parallel implementation environments as well, like e.g.
+OpenCL.
+The key point is that GPU's can accelerate small concurrent tasks, like beamforming one pixel, much better than a CPU.
+Where the CPU has few cores and logic to process complex threads, the GPU has many cores, but each thread needs to be simple in order to benefit from the extra computational power.
+
+The CUDA compute model is based on execution of a kernel function across a grid of compute threads, where the kernel function contains the work to be done by each thread.
+As depicted in Figure \ref{fig:gpulayout}, each position in the grid holds a block of threads, with maximal size of 1024 threads, that can be synchronized with little cost if needed.
+%Each block is further divided into groups of 32 threads, known as a warp. The maximum number of threads inside a block is restricted to 512, hence 16 warps. 
+From a CUDA perspective the GPU consist of one or more stream multi processors (SM's), where each SM is capable of 32 or more, depending on the architecture, concurrent multiplications and/or additions.
+A group of blocks is scheduled to each SM until all blocks have been processed.
+
+A SM has available a limited amount of near-core memory and registers.
+Each thread therefore has to be as light on memory as possible and has to perform enough instructions per used byte in order to hide memory latency.
+Global read and writes on the GPU, in addition to CPU-GPU transfers have high penalty, and should be minimized as much as possible.
+All of these constraints combined pose a real challenge as the target problem has to map on to the parallel framework both on a grid, block and memory consumption level in order to benefits from parallel acceleration.
+In the next section, we will describe how this mapping should be done for Capon beamforming of ultrasound data.
 
@@
 \end{figure}
-% Få med at vi processerer hele bildet på en gang. Forklar forskjellen med å prosessere en beam av gangen. Hva med MLA?
-% Illustrer tråder og tider for hvordan processeringen går.
-% Vi snakker om mottak-stråleformeren.
-
-%\begin{figure}
-%\includegraphics[width=3in]{gfx/beamforming_mv_lowres.png}
-%\caption{}
-%\label{fig:mv}
-%\end{figure}
-
-%Building and calculating the inverse of $\hat{R}$ is computational demanding, limiting the methods accessibility. Following the innovation in GPU computing, this has now started to change. 
-%In this paper we introduce the first GPU implementation of the Capon beamformer capable of processing a 70 degree sector cardiac image from a 64 element phased array at interactive frame rates using both spatial and temporal smoothing.
-Both types of smoothing are important for maintaining delay-and-sum-like speckle and at the same time keep the high resolution property.
-This is achieved using complex base band data, which reduces the size of each range line to a minimum.
-In addition, complex data results in complex weights; hence this is a correct implementation of Capon beamforming with all degrees of freedom preserved.
-For high channel count systems we propose two ways of reducing computational complexity while maintaining
-
-\subsection{GPU Compute Model}
-
-\textbf{(Ta med mer banale ting, dette er GPU for ultralydfolk)}
-
-\textbf{(Etter denne så må de forstå GPU og hvorfor MV passer.)}
-
-In this section we briefly discuss the principle behind GPU computing, laying the basis for further discussion on parallel implementation strategies.
-CUDA by Nvidia has been our choice of GPU developer environment \cite{Nvidia2011}.
-However, even though CUDA has been used to implement the Capon beamformer, the challenges and proposed methods described in this paper applies for other massive parallel implementation environments as well, like e.g.
-OpenCL.
-
-The CUDA compute model is based on execution of a kernel-function across a grid of compute threads.
-As depicted in Figure \ref{fig:gpulayout}, each position in the grid holds a block of threads, with maximal size of 1024 threads, that can be easily synchronized if needed.
-%Each block is further divided into groups of 32 threads, known as a warp. The maximum number of threads inside a block is restricted to 512, hence 16 warps. 
-From a CUDA perspective the GPU consist of one or more stream multi processors (SM's), where each SM is capable of 32 or more, depending on the architecture, concurrent multiplications and/or additions.
-A group of blocks is scheduled to each SM until all blocks have been processed.
-More important, the SM has a limited amount of near-core memory and registers available.
-Each thread therefore has to be as light on memory as possible and has to perform enough instructions per used byte in order to hide memory latency.
-Global read and writes on the GPU, in addition to CPU-GPU transfers have high penalty, and should be minimized as much as possible.
-All of these constraints combined pose a real challenge as the target problem has to map on to the parallel framework both on a grid, block and memory consumption level in order to benefits from parallel acceleration.
-In the next section, we will describe how this mapping should be done for Capon beamforming of ultrasound data.
 
@@
 As mentioned, widespread use of Capon adaptive beamforming has been limited by its computational complexity.
-However, if we analyse the beamformer in (\ref{eq:z}) in combination with (\ref{eq:w}) and (\ref{eq:R}), we see that each calculation of a set of Capon weights is independent across the output $z[n]$.
+However, if we analyse the beamformer in (\ref{eq:z_mv}) in combination with (\ref{eq:w}) and (\ref{eq:R}), we see that each calculation of a set of Capon weights is independent across the output $z[n]$.
 As shown in Figure \ref{fig:gpulayout}, a first level of granularity is therefore to divide the image into a grid, where the output in each cell can be calculated independently of all others.
@@
 Continuing with the calculation of one weight vector, the algorithm can be broken down into three main steps that to some extent can be further divided into a small set of parallel processes.
-That is, for each data vector $\vec{x}$ in the image:
+That is, for each data vector $\vec{x}^\Delta$ in the image:
 \begin{align*}
-\begin{array}{l}
-\text{1) Calculate the sample covariance matrix } \mat{\hat{R}} \text{ as in (\ref{eq:R})}\\
-\text{2) Solve the linear system of equations } \mat{\hat{R}}\vec{b} =\vec{ 1}\\
-\text{3) Calculate } \vec{w} = \vec{b}/\vec{1}^H\vec{b} \text{ and the amplitude } \vec{w}^H\vec{x}
-\end{array}
+%\begin{array}{l}
+\text{1) } &\text{Calculate the sample covariance matrix } \mat{\hat{R}} \text{ as in } \text{(\ref{eq:diag})}\\
+\text{2) } &\text{Solve the linear system of equations } \vec{b} =\mat{\hat{R}}^{-1}\vec{1}_L\\
+\text{3) } &\text{Calculate } \vec{w} = \vec{b}/\vec{1}_{L}^H\vec{b} \text{ and the amplitude as in (\ref{eq:z_mv})}
+%\end{array}
 \end{align*}
@@
 %\end{enumerate}
-The computational complexity of these three steps is $O(L^3)$, $O(L^3)$, and $O(L)$ respectively if $L = M/2$.
-For large $L$'s, the first two steps quickly becomes unmanageable, and real time performance can only be achieved by exploring novel solutions to reduce complexity.
+The computational complexity of these three steps is $O(L^3)$, $O(L^3)$, and $O(L)$ respectively.
+As $L$ increases, the first two steps quickly becomes unmanageable, and real time performance can only be achieved by exploring novel solutions to reduce complexity.
 
@@
 
-The straight forward sum in (\ref{eq:R}) has complexity $O(L^2KN) = O(L^3)$ if the number of subarrays $K$ is small.
+The straight forward sum in (\ref{eq:R}) has complexity $O(L^2N_LN_K) = O(L^3)$ if $N_L$, the number of subarrays, is small.
 This is actually the same as for inverting a $L \times L$ matrix.
 However, little attention has been devoted to this step in the literature until now.
-Note that for a typical array configuration we typically smooth more in space than in range.
-Hence, N is usually small compared to $K$.
-For short subarrays the construction of $\mat{\hat{R}}$ is dominated by the number of elements $M$, since $K = M - L + 1$.
+Note that for a typical array configuration we typically do more spacial then temporal smoothing.
+Hence, $N_K$ is usually small compared to $N_L$.
+For short subarrays the construction of $\mat{\hat{R}}$ is dominated by the number of elements $M$, since $N_L = M - L + 1$.
 From (\ref{eq:R}) it is obvious that calculation of one element in $\mat{\hat{R}}$ is independent of all the others.
 A natural level of granularity would then be to assign one thread to each element of $\mat{\hat{R}}$.
-This will minimize global reads and writes needed per thread, but unfortunately the maximum number of threads per block puts an upper limit on subarray length $L$, $L_{max}=32$, if we want to avoid block-to-block synchronization.
-
-The quadratic complexity with $L$ and linear complexity with $N$ can be reduced by realizing that the calculation in (\ref{eq:R}) overlap both across subarrays and in time.
-The $(i,j)$ element of $\mat{\hat{R}}[n]$ can be calculated from element $(i-1, j-1), \text{for } i,j > 0$ by subtracting $x_{i-1}x_{j-1}^*$ and adding $x_{i+K-1}x_{j+K-1}^*$.
-The same is true in time where $\mat{\hat{R}}[n]$ can be calculated from $\mat{\hat{R}}[n-1]$ by subtracting $\mat{\hat{R}}[n-N_{avg}-1]$ and adding $\mat{\hat{R}}[n+N_{avg}]$.
-Both sliding across subarrays and in time can be realized simultaneously if we hold $\vec{x}\vec{x}^H[n]$ averaged over $N$ time samples in memory and calculate $\mat{\hat{R}}$ averaged over subarrays from that buffer (figure).
-However, since $N$ typically is small, and we have a limited amount of memory per thread on the GPU, we have chosen to focus on sliding across subarrays only.
+This will minimize global reads and writes needed per thread, but unfortunately the maximum number of threads per block puts an upper limit on the subarray length, $L_{max}=32$, if we want to avoid block-to-block synchronization.
+
+The quadratic complexity with $L$ and linear complexity with $N_K$ can be reduced by realizing that the calculation in (\ref{eq:R}) overlap both across subarrays and in time.
+The $(i,j)$ element of $\mat{\hat{R}}[n]$ can be calculated from element $(i-1, j-1), \text{for } i,j > 0$ by subtracting $x_{i-1}x_{j-1}^*$ and adding $x_{i+N_L-1}x_{j+N_L-1}^*$.
+The same is true in time where $\mat{\hat{R}}[n]$ can be calculated from $\mat{\hat{R}}[n-1]$ by subtracting $\mat{\hat{R}}[n-K-1]$ and adding $\mat{\hat{R}}[n+K]$.
+Both sliding across subarrays and in time can be realized simultaneously if we hold a $N_K$-average in time of the outer products $\vec{x}^{\Delta}[n]\vec{x}^{\Delta}[n]^H$ in memory and calculate $\mat{\hat{R}}$ averaged over subarrays from that buffer (figure?).
+However, since $N_K$ typically is small, and we have a limited amount of memory per thread on the GPU, we have chosen to focus on sliding across subarrays only.
 In that case, the complexity of calculating $\mat{\hat{R}}$ is reduced to $O(L^2)$.
@@
 With our choice of calculating $\mat{\hat{R}}$, calculations are still independent across time, but not across subarrays.
-We therefore have selected to use on thread per column in $\mat{\hat{R}}$, and for small $L$'s several covariance matrices are calculated per thread block.
+We therefore have selected to use one thread per column in $\mat{\hat{R}}$, and for small $L$'s several covariance matrices are calculated per thread block.
 Maximum supported $L$ is with this approach increased to 512, however for small $L$ high $M$ configurations, memory consumption per thread will be to high to achieve full occupancy of the GPU.
@@
 However, the way we have decided to organize the GPU compute threads per $\mat{\hat{R}}$, the lower triangle can be calculated with little overhead on the GPU.
-This also means that we do not relay on a symmetric solver to solve $\mat{\hat{R}}\vec{b} = \vec{1}$, even if that could have saved some instructions.
+This also means that we do not relay on a symmetric solver to solve $\mat{\hat{R}}\vec{b} = \vec{1}_L$, even if that could have saved some instructions.
 The memory layout for the upper triangle only is also believed to be cumbersome.
@@
 Diagonal loading as a constant value can be added with marginal cost, but the adaptive diagonal loading factor in (\ref{eq:diag}) is a different story.
-In this case, the trace of $\mat{\hat{R}}$ has to be known before diagonal elements are written back to global memory.
+The trace of $\mat{\hat{R}}$ has to be known before diagonal elements are written back to global memory.
 Since one row for all $\mat{\hat{R}}$'s is written to global memory per kernel iteration, the solution was to hold the diagonal elements in a local buffer until the trace was accumulated, and then do scattered writes of the diagonal elements to global memory.
@@
 \subsection{Solving Multiple Small Linear Systems}
-When it comes to solving systems of linear equations on the GPU the focus has mostly been on large systems in the literature, and GPU libraries usually lack a comprehensive collection of batched solvers.
-The reason for this is two-folded, first the GPU needs to solve thousand of small systems in order to beat the low memory latency of the CPU, second there is a range of system dimensions that does not fit well with the GPU architecture, where only a small amount of near-core memory registers are available.
+In the literature, when it comes to solving systems of linear equations on the GPU the focus has mostly been on large systems, and GPU libraries usually lack a comprehensive collection of batched solvers.
+The reason for this is two-fold, first the GPU needs to solve thousand of small systems in order to beat the low memory latency of the CPU, second there is a range of system dimensions that does not fit well with the GPU architecture, where only a small amount of near-core memory and registers are available.
 Because of this, it has been proven hard to optimize a manageable set of solvers that provides speedups compared with the CPU in every cases.
 
-For the results presented in this paper, we have used an unreleased GPU implementation of Gauss Jordan (GJ) elimination (by NVIDIA) to solve $\mat{R}\vec{b} = \vec{1}$, where $\vec{b} = \mat{R}^{-1}\vec{1}$.
-However we will go through an implementation of an $\mat{U}^H\mat{D}\mat{U}$-decomposition-based solver to point out challenges when implementing batched solving of small linear systems on the GPU.
+For the results presented in this paper, we have used an unreleased GPU implementation of Gauss Jordan (GJ) elimination (by NVIDIA) to solve $\mat{\hat{R}}\vec{b} = \vec{1}_L$, where $\vec{b} = \mat{\hat{R}}^{-1}\vec{1}_L$.
+For smaller systems, $L < 10$, a direct solver using one thread per system is preferable.
+For this purpose we have implemented Cramer's rule for $2 \times 2$, $3 \times 3$ and $4 \times 4$ matrices.
+This has been benchmarked against the latest MKL libraries.
+
+%However we will go through an implementation of an $\mat{U}^H\mat{D}\mat{U}$-decomposition-based solver to point out challenges when implementing batched solving of small linear systems on the GPU.
 ...
@@
 
-Skriv om utfordringene om å lage en solver på GPU. Ikke dra inn UHDU.
-
-For the smallest systems a direct solver using one thread per system is preferable. For this purpose we have implemented Cramer's rule for $3 \times 3$ (, $2 \times 2$?  and $4 \times 4$?) matrices.
+%Skriv om utfordringene om å lage en solver på GPU. Ikke dra inn UHDU.
 
 %In the final journal article we could discuss Conjugated gradient and Woodbury and the benefits they bring.
-Other solvers that provides different benefits includes Conjugated gradient and the use of the Woodbury matrix identity.
+%Other solvers that provides different benefits includes Conjugated gradient and the use of the Woodbury matrix identity.
 
 \subsection{Compute Beamformer Output}
-(Still just notes) One thread per weight ($L$ threads). Let one thread sum the solution vector $\vec{b}$, or let several threads cooperate to form the sum using a binary reduction sum.
-Let one thread read one element of $\vec{b}$, scale it with the inverse sum, and optionally write it back to global memory.
-The final output can be formed by either reduce the data down to length $L$, or increase the weight vector to length $M$.
-Since we have decided to have $L$ threads per output, the data is reduced to length $L$ by calculating $x_{r} = \sum_{k=r}^{r+K-1}x_{k}$ for $r \in [0, L-1]$.
-Each thread now computes $z_r = w_r^*x_r$, and finally one thread computes and outputs $z = \sum_{r=0}^{L-1}$ to global memory.
+To calculate the beamformer output we have implement the formula in (\ref{eq:z_mv}) with one thread per element-weight ($L$ threads per weight vector).
+After the solution vector $\vec{b}$ has been placed in shared memory, one of the $L$ threads finds the sum of $\vec{b}$.
+Another option is to let several threads cooperate to form the sum using a binary reduction sum.
+However, in the following benchmarks the first approach has been used.
+Each of the $L$ threads then read one element of $\vec{b}$, scale it with the inverse sum just found to get the weight vector $\vec{w}[n]$.
+
+The final output, as shown in (\ref{eq:z_mv}) and (\ref{eq:z_mv2}), can be formed by either reducing the data down to length $L$, or increase the weight vector to length $M$.
+Since we have decided to have $L$ threads per output, the data is reduced to length $L$ by calculating
+\begin{align}
+\breve{x}_{r}^{\Delta} = \sum_{k=r}^{r+N_L-1}x_{k}^{\Delta} \text{ for } r \in [0, L-1].
+\end{align}
+Each thread now computes $z_r = w_r^*x_r$, and finally one thread computes and outputs $z = \sum_{r=0}^{L-1} z_r$ to global memory.
 The last sum can also be performed using a binary reduction sum across several threads for increased speed.
@@
 \subsection{Beamspace Processing}
-Nilsen et.
-al \cite{Nilsen2009} has proposed to apply beamspace processing to further reduce the required computations.
-With beamspace we mean transforming the channel data from element space to a fan of beams covering the imaging sector, $\vec{x}_{BS} = \mat{B}x$.
+Nilsen et al.
+\cite{Nilsen2009} have proposed to apply beamspace processing to further reduce the required computations.
+With beamspace we mean transforming the channel data from element space to a fan of beams covering the sector illuminated by transmit beams.
+The transformation is written as
+\begin{align}
+\vec{x}_{BS} = \mat{B}\vec{x}^{\Delta}.
+\end{align} 
 The steering vectors in $\mat{B}$, the so-called Butler matrix, determines each beams direction in space.
@@
 al concluded that as little as three beams around broadside where adequate to produce results comparable to applying Capon beamforming in element space.
-The beamspace version of Capon is derived by replacing the covariance matrix in (\ref{eq:w}) with the beamspace covariance matrix $\mat{R}_{BS} = \vec{x}_{BS}\vec{x}_{BS}^H$, where $\mat{B} \in \mathbb{C}^{N_b,M}$, and $N_b$ is the number of selected beams.
+
+The beamspace version of Capon is derived by replacing the covariance matrix in (\ref{eq:w}) with the beamspace covariance matrix $\mat{\hat{R}}_{BS} = f(\vec{x}_{BS}\vec{x}_{BS}^H)$, where $\mat{B} \in \mathbb{C}^{N_b,M}$, and $N_b$ is the number of selected beams.
 Note that when subarray averaging is applied, the transformation must be done per subarray, yielding subarrays that no longer overlap in space.
-Hence, the amount of data is change from $M$ to $KN_b$ per sample.
+Hence, the amount of data is change from $M$ to $N_LN_b$ per sample.
 
 (Go in detail about the implementation using sliding DFT).
+
+(Go in detail about what kind of modifications we need to do to each of the three steps)
 
@@
 For each step, there is one line for three different choices of temporal averaging, $N_{avg} \in [0, 1, 2]$.
-(a) A 32 elements array.
-(b) A 64 elements array.}
+(a) A 32 element array.
+(b) A 64 element array.}
 \label{fig:bench}
@@
 We expect that a large speedup had the Capon beamformer been implemented on the CPU using threads and SIMD (single instruction multiple data) instructions.
+
+Solving linear systems on the GPU is not believed to be much faster than a CPU implementation, but there is two key points.
+First the CPU is offloaded, and second the time it takes to transfer all covariance matrices back to the CPU makes the choice of doing the solving on the GPU simple to make.
 
@@
 
-The authors would like to thank... \cite{Synnevag2009}
+The authors would like to thank...
 

### 2012-07-31 HRUIonGPU/jrnl_article.tex
small update on article

@@
 \begin{align}\label{eq:diag}
-\mat{\hat{R}} = \mat{\breve{R}} + \epsilon\mat{I}.
+\mat{\hat{R}} =  f(\vec{x}^{\Delta}[n]) = \mat{\breve{R}} + \epsilon\mat{I}.
 \end{align}
@@
 z[n] &= \frac{1}{N_L}\vec{w}[n]^H \sum_{l=0}^{N_L-1} \vec{x}_l^{\Delta}[n] \label{eq:z_mv}\\
-&= (\vec{w}[n]^H \ast{} \vec{1}_{N_L}/N_L)\vec{x}^{\Delta}[n] \label{eq:z_mv2}.
+&= 1/N_L \text{CONV}_m(\vec{w}[n], \vec{1}_{N_L})^H\vec{x}^{\Delta}[n] \label{eq:z_mv2},
 \end{align}
+%where (this matrix is wrong)
+%\begin{align}
+%\mat{C} = \left(
+%\begin{matrix}
+%\vec{1}_1 & \vec{1}_2 & \cdots & \vec{1}_{N_L} & \vec{0}_1 & \cdots & \vec{0}_{N_L-1} & \vec{0}_{N_L}\\
+%\vec{0}_{N_L-1} & \vec{0}_{N_L-2} & \cdots & \vec{0}_{0} & \vec{1}_{N_L-1} & \cdots & \vec{1}_{2} & \vec{1}_{1}\\
+%\end{matrix}
+%\right)^T
+%\end{align}
 Note how we can either select to sum all subarrays first in (\ref{eq:z_mv}), or accumulate the right weighting per element in (\ref{eq:z_mv2}).
@@
 For this purpose we have implemented Cramer's rule for $2 \times 2$, $3 \times 3$ and $4 \times 4$ matrices.
-This has been benchmarked against the latest MKL libraries.
+All solvers have been benchmarked against the latest MKL libraries.
 
@@
 The final output, as shown in (\ref{eq:z_mv}) and (\ref{eq:z_mv2}), can be formed by either reducing the data down to length $L$, or increase the weight vector to length $M$.
-Since we have decided to have $L$ threads per output, the data is reduced to length $L$ by calculating
+Since we have decided to have $L$ threads per output, the data is reduced to length $L$ by letting thread $r$ calculate
 \begin{align}
@@
 
-The beamspace version of Capon is derived by replacing the covariance matrix in (\ref{eq:w}) with the beamspace covariance matrix $\mat{\hat{R}}_{BS} = f(\vec{x}_{BS}\vec{x}_{BS}^H)$, where $\mat{B} \in \mathbb{C}^{N_b,M}$, and $N_b$ is the number of selected beams.
+The beamspace version of Capon is derived by replacing the covariance matrix in (\ref{eq:w}) with the beamspace covariance matrix $\mat{\hat{R}}_{BS} = f(\vec{x}_{BS})$, where $\mat{B} \in \mathbb{C}^{N_b,M}$, and $N_b$ is the number of selected beams.
 Note that when subarray averaging is applied, the transformation must be done per subarray, yielding subarrays that no longer overlap in space.

### 2012-08-01 HRUIonGPU/jrnl_article.tex
small update

@@
 \begin{align}\label{eq:w}
-\vec{w}[n] = \frac{\mat{\hat{R}}[n]^{-1}\vec{a}}{\vec{a}^H\mat{\hat{R}}[n]^{-1}\vec{a}},
+\vec{w}[n] = \frac{\mat{\hat{R}}[n]^{-1}\vec{a}}{\vec{a}^H\mat{\hat{R}}[n]^{-1}\vec{a}} = \frac{\vec{b}}{\vec{a}^H\vec{b}},
 \end{align}
@@
 z[n] &= \frac{1}{N_L}\vec{w}[n]^H \sum_{l=0}^{N_L-1} \vec{x}_l^{\Delta}[n] \label{eq:z_mv}\\
-&= 1/N_L \text{CONV}_m(\vec{w}[n], \vec{1}_{N_L})^H\vec{x}^{\Delta}[n] \label{eq:z_mv2},
+&= 1/N_L \, \text{CONV}_m(\vec{w}[n], \vec{1}_{N_L})^H\vec{x}^{\Delta}[n] \label{eq:z_mv2},
 \end{align}
@@
 As shown in Figure \ref{fig:gpulayout}, a first level of granularity is therefore to divide the image into a grid, where the output in each cell can be calculated independently of all others.
-Further, based on memory dependencies and to hold memory consumption per thread at a minimum, one should recognise that one block of threads should process at least one output in the final image.
+Further, based on memory dependencies and to hold memory consumption per thread at a minimum, one should recognise that a group of threads should process one output in the final image.
 Continuing with the calculation of one weight vector, the algorithm can be broken down into three main steps that to some extent can be further divided into a small set of parallel processes.
@@
 
-The straight forward sum in (\ref{eq:R}) has complexity $O(L^2N_LN_K) = O(L^3)$ if $N_L$, the number of subarrays, is small.
+The straight forward sum in (\ref{eq:R}) has complexity $O(L^2N_LN_K) = O(L^3)$ since $N_L$, the number of subarrays, is a function of $L$.
 This is actually the same as for inverting a $L \times L$ matrix.
-However, little attention has been devoted to this step in the literature until now.
+However little attention has been devoted to this step in the literature before.
 Note that for a typical array configuration we typically do more spacial then temporal smoothing.
 Hence, $N_K$ is usually small compared to $N_L$.
-For short subarrays the construction of $\mat{\hat{R}}$ is dominated by the number of elements $M$, since $N_L = M - L + 1$.
+For short subarrays the construction of $\mat{\hat{R}}$ is dominated by the number of elements $M$ ($N_L = M - L + 1$), but since $L$ is selected in most cases to be between $M/2$ and $M/4$ a large $M$ implies a large $L$.
+
 From (\ref{eq:R}) it is obvious that calculation of one element in $\mat{\hat{R}}$ is independent of all the others.
@@
 The same is true in time where $\mat{\hat{R}}[n]$ can be calculated from $\mat{\hat{R}}[n-1]$ by subtracting $\mat{\hat{R}}[n-K-1]$ and adding $\mat{\hat{R}}[n+K]$.
-Both sliding across subarrays and in time can be realized simultaneously if we hold a $N_K$-average in time of the outer products $\vec{x}^{\Delta}[n]\vec{x}^{\Delta}[n]^H$ in memory and calculate $\mat{\hat{R}}$ averaged over subarrays from that buffer (figure?).
+Both sliding across subarrays and in time can be realized simultaneously if we hold a $N_K$-average in time of the outer products $\vec{x}^{\Delta}[n]\vec{x}^{\Delta}[n]^H$ in memory and calculate $\mat{\hat{R}}$ averaged over subarrays from that buffer (ref JOE).
 However, since $N_K$ typically is small, and we have a limited amount of memory per thread on the GPU, we have chosen to focus on sliding across subarrays only.
 In that case, the complexity of calculating $\mat{\hat{R}}$ is reduced to $O(L^2)$.
-Another aspect is that sliding will break independence in the sliding dimension.
+Important to notice is that sliding will break independence in the sliding dimension.
 With our choice of calculating $\mat{\hat{R}}$, calculations are still independent across time, but not across subarrays.
-We therefore have selected to use one thread per column in $\mat{\hat{R}}$, and for small $L$'s several covariance matrices are calculated per thread block.
+We therefore have selected to use one thread per column in $\mat{\hat{R}}$, and for small $L$'s several covariance matrices are calculated per thread block (Figure \ref{fig:gpulayout}).
 Maximum supported $L$ is with this approach increased to 512, however for small $L$ high $M$ configurations, memory consumption per thread will be to high to achieve full occupancy of the GPU.
@@
 \subsection{Solving Multiple Small Linear Systems}
-In the literature, when it comes to solving systems of linear equations on the GPU the focus has mostly been on large systems, and GPU libraries usually lack a comprehensive collection of batched solvers.
+In the literature, when it comes to solving systems of linear equations on the GPU the focus has mostly been on large systems, and GPU libraries therefore often lack a comprehensive collection of batched solvers.
 The reason for this is two-fold, first the GPU needs to solve thousand of small systems in order to beat the low memory latency of the CPU, second there is a range of system dimensions that does not fit well with the GPU architecture, where only a small amount of near-core memory and registers are available.
 Because of this, it has been proven hard to optimize a manageable set of solvers that provides speedups compared with the CPU in every cases.
+It is therefore not given that solving on the GPU is faster than solving on the CPU in general.
+However, there is two arguments for doing solving on the GPU.
+First the CPU, which already runs a lot of algorithms, is offloaded.
+Second the transfer of several thousand covariance matrices from CPU-side to GPU-side is spared.
 
@@
 For this purpose we have implemented Cramer's rule for $2 \times 2$, $3 \times 3$ and $4 \times 4$ matrices.
-All solvers have been benchmarked against the latest MKL libraries.
+All solvers have been benchmarked against the latest Intel MKL libraries.
 
@@
 
-\section{Real Time Performance for $L > 16$}\label{sec:meth2}
-
-\subsection{Beamspace Processing}
+\section{Beamspace Processing}
 Nilsen et al.
@@
 \section{Results}\label{sec:res}
-
+Describe the computer system (CPU, GPU...). Describe the data that are processed (field II, Channel data from GE Vingmed).
 % stolpediagram, med de ulike metodenes kjøretid oppehverandre.

### 2012-08-15 HRUIonGPU/jrnl_article.tex

@@
 %\IEEEPARstart{I}{ntroduction} to Capon adaptive beamforming, ultrasound imaging and GPU/CUDA.
-\IEEEPARstart{C}{apon} adaptive beamforming \cite{Capon1969} applied in a medical ultrasound setting has recently shown promising results in several studies \cite{Synnevag2007, Austeng2008, Vignon2008, Viola, Mehdizadeh2012}.
-Although the method introduces some interesting trade-offs in addition to just improved resolution, like maintaining image resolution in combination with higher framerates, smaller probes, or deeper penetration \cite{Synnevag2009}, widespread use has been limited by the tremendous computational complexity associated with the method \cite{So2011}.
-In this paper we introduce the first implementation of the Capon beamformer exhibiting real-time performance when applied in a typical cardiac ultrasound setting.
-To accomplish this task we make use of the massive, and easy accessed, parallel processing power found in modern graphics processing units (GPU's), combined with novel methods to reduce the computational complexity as the problem size grows.
-
-Previous attempts on implementing a parallel Capon beamformer have mainly been focused around radar and passive sonar systems using custom made systolic array processors \cite{McWhirter1989, Moonen1993, Sinha2002}.
+\IEEEPARstart{C}{apon} adaptive beamforming \cite{Capon1969} applied to medical ultrasound imaging has recently shown promising results in several studies \cite{Synnevag2007, Austeng2008, Vignon2008, Viola, Mehdizadeh2012}.
+It has been shown that the method introduces some interesting trade-offs in addition to just improved resolution, like maintaining image resolution in combination with higher framerates, smaller probes, or deeper penetration \cite{Synnevag2009}.
+Despite these benefits, widespread use has been limited by the immense computational complexity associated with the method \cite{So2011}.
+In this paper we introduce the first implementation of the Capon beamformer exhibiting real-time performance when applied in a typical cardiac ultrasound imaging setting.
+To accomplish this task we make use of the massive, and easily accessed, parallel processing power found in modern Graphics Processing Units (GPU's), combined with novel methods to reduce the computational complexity as the problem size grows.
+
+Previous work on implementing a parallel Capon beamformer have mainly been focused around radar and passive sonar systems using custom made systolic array processors \cite{McWhirter1989, Moonen1993, Sinha2002}.
 %A systolic array processor is custom made hardware, and even if the computational throughput might be higher for selected problems compared with a GPU, the cost related to development and flexibility is considered much higher.
@@
 \cite{Chen2011a, Chen2011} have implemented a Capon beamformer on the GPU that produced real apodization weights based on real data.
-Since a real function is symmetric in frequency space the Capon beamformer is not allowed to shift the main lobe under this constraint, and therefore its full potential is not reached.
+Under this constraint the Capon beamformer is not allowed to shift the main lobe, and therefore its full potential is not reached.
 Shifting the main lobe is an important property of the Capon beamformer in order for it to improve edge definitions and contrast throughout an ultrasound image.
-Another aspect of using complex data is that eventhough the number of aritmetic operations per sample is increased we can now sample at the system bandwidth, keeping the number of samples for a full cardiac imaging sector at a minimum.
+Another aspect of using complex data is that even though the number of aritmetic operations per sample is increased we can now sample at the system bandwidth, keeping the number of samples for a full cardiac imaging sector at a minimum.
 All results in this paper is reported with the use of baseband data.
@@
 In recent work (cite JOE) we have applied a GPU-based Capon beamformer in an active sonar system.
-For sonar the real time requirement is much lower than for ultrasound, which means that most common array configurations can be handled using the GPU without taking any additional measures.
-For ultrasound, and especially cardiac ultrasound, frame rate requirements increase to a level where even a high-end GPU can not do the processing required for a standard implementation of Capon beamforming.
-Since the execution time for a Capon beamformer highly depends on the number of elements in the given array, we need to find ways of reducing the computational complexity for high channel count ultrasound systems.
-Different approaches to low complexity Capon beamforming have recently been proposed \cite{Asl2012, Jensen2012}, but in this paper we have selected to implemented the beamspace approach described by Nilsen and Hafizovic \cite{Nilsen2009} on the GPU to maintain high frame rates as the channel count increases.
+For sonar the real-time requirement is much lower than for medical ultrasound imaging, which means that most common array configurations can be handled using the GPU without taking any additional measures.
+For medical ultrasound imaging, and especially cardiac ultrasound imaging, frame rate requirements increase to a level where even a high-end GPU can not do the processing required for a standard implementation of Capon beamforming.
+Since the execution time for a Capon beamformer depends highly on the number of elements in the given array, it is important to find ways of reducing the computational complexity for high-channel-count systems.
+Different approaches to low-complexity Capon beamforming have recently been proposed \cite{Asl2012, Jensen2012}, and in this paper we have selected to implement the beamspace transformation described by Nilsen and Hafizovic \cite{Nilsen2009} on the GPU to maintain high framerates as the channel count increases.
 %(Referer til JOE-artikkelen, få med beskrivelse av hva som blir anerledes for ultralyd.)
@@
 In the next section we present background information about the Capon beamformer and the GPU computing model.
-In Section \ref{sec:meth} we go in detail about the parallel implementation of the Capon beamformer, before in Section \ref{sec:meth2} we introduce the method applied when the problem size increases.
+In Section \ref{sec:meth} we go in detail about the parallel implementation of the Capon beamformer, before in Section \ref{sec:bs} we introduce the method applied when the problem size increases.
 Benchmark results and resulting images are presented in Section \ref{sec:res}, and finally we discuss the results in Section \ref{sec:dis}, and draw conclusions in Section \ref{sec:con}.
@@
 
-The conventional form of beamforming is delay-and-sum (DAS), where the signal recorded on each element, $x_m[n]$, is delayed to broadside with an appropriate delay $\Delta_m[n]$.
+The conventional form of array beamforming is delay-and-sum (DAS).
+Here, the signal $x_m[n]$ recorded at element number $m \in [0,M-1]$ from time $n = 0$ to $N-1$ is delayed to broadside with an appropriate delay $\Delta_m[n]$.
 The beamformer output, $z[n]$, is then the sum of all $M$ channels weighted with a per-element factor $w_m$,
@@
 \end{align}
-In conventional beamforming, the weighting $\vec{w}$ is usually real, and applied to trade resolution for a better signal-to-noise ratio (SNR).
-Hence, by increasing the main lobe width of the system while gaining lower side lobe levels.
-
-The basic idea of the Capon adaptive beamformer, as shown in Figure \ref{fig:mvbf}, is to calculate a set of weights on-the-fly, based on the data at hand, in such a way that we minimize the beamformer output power while maintaining unit gain at broadside.
-The result being that interference impinging from other directions than the point of focus is suppressed \cite{Synnevag2007}.
+In conventional beamforming, the weighting $\vec{w}$ is usually real, and is applied to trade resolution for a better signal-to-noise ratio (SNR).
+
+As shown in Figure \ref{fig:mvbf}, the basic idea of an adaptive beamformer is to form a weight vector based on the data at hand.
+For Capon adaptive beamforming the weights are found that minimize the beamformer output power while maintaining unit gain at broadside.
+The result is that interference impinging from other directions than the point of focus is suppressed \cite{Synnevag2007}.
 %\textbf{(Vi skal implementer dette. vis bare vektene. Dropp ligning 2, siter, bruk R-hat med en gang. Få med den endelige feilen )}
@@
 \centerline{\includegraphics[width=3in]{gfx/beamforming_mv_lowres.png}}
-\caption{Capon adaptive beamforming.
+\caption{Adaptive beamforming.
 Impinging signal is demodulated, and aligned in phase by applying steering and focusing delays $\Delta_m[n]$.
@@
 
-The solution to this optimization problem is given by the following expression:
+The Capon weight vector is given by the following expression:
 \begin{align}\label{eq:w}
@@
 \end{align}
-where $\vec{a} = [1_0, 1_1, ..., 1_{M-1}]^T = \vec{1}_M$ when $\vec{x}$ is pre-delayed to broadside, and $\mat{\hat{R}}$ is the sample covariance matrix calculated as $\vec{x}[n]\vec{x}[n]^H$. 
+where $\vec{a} = [1_0, 1_1, ..., 1_{M-1}]^T = \vec{1}_M$ when $\vec{x}$ is pre-delayed, $\mat{\hat{R}}$ is the sample covariance matrix calculated from the data vector $\vec{x}^\Delta[n]$, and $\vec{b} = \mat{\hat{R}}[n]^{-1}\vec{a}$.
 %\begin{align}
@@
 %\end{align} 
-Note that this calculation only use a single sample. Typically the sample covariance matrix has to be averaged over several samples in order to ensure numerical stability and statistical robustness.
+%Typically the sample covariance matrix has to be averaged over several samples in order to ensure numerical stability and statistical robustness.
 
 In passive systems, assuming stationary statistics, the sample covariance matrix is typically averaged over a long sequence of samples\cite{Krima}.
-For an active, pulsed system like ultrasound imaging, the available data from which $\mat{\hat{R}}$ can be estimated is quite limited.
+For an active, pulsed system like medical ultrasound imaging, the available data from which $\mat{\hat{R}}$ can be estimated is quite limited.
 The resulting sample covariance matrix might therefore be both inaccurate, and poorly conditioned.
-Another challenge with applying Capon to active systems is the effect of signal cancellation, where high correlation between signal and interference favours weights that in the end cancel out the signal \cite{Reddy1987}.
-In this way, the beamformer output can be to low at broadside, even when there is signal present and the distortion-less criterion is applied.
-
-In order to get a well conditioned $\mat{\hat{R}}$, avoid signal cancellation and to get DAS-like speckle, $\mat{\hat{R}}$ has to be averaged over $L\le M/2$ long subarrays and $N_K \sim \tau/T_s$ time samples \cite{Synnevag2007, Synnevag2007a}.
-Here, $\tau$ is the pulse length in seconds and $T_s$ is the sampling period.
-Continuing, we therefore calculate
+Another challenge with applying Capon to active systems is the effect of signal cancellation, where high correlation between signal and interference favors weights that cancel out the signal \cite{Reddy1987}.
+In this way, the beamformer output can be too low at broadside, even when there is signal present and the distortionless-response criterion is applied.
+
+In order to get a well-conditioned $\mat{\hat{R}}$, avoid signal cancellation, and to get DAS-like speckle, $\mat{\hat{R}}$ has to be averaged over $L\le M/2$ long subarrays and $N_K \sim T_p/T_s$ time samples \cite{Synnevag2007, Synnevag2007a}.
+Where, $T_p$ is the pulse length in seconds and $T_s$ is the sampling period.
+We therefore calculate
 \begin{align}
@@
 \end{align}
-where $N_L = M-L+1$, $N_K = 2K + 1$ is the number of subarrays, and $\vec{x}_l^{\Delta}$ is the $l^\text{th}$ subarray $[x_l^{\Delta}[n], \dotso, x_{l+L-1}^{\Delta}[n]]$.
-The dimension of the covariance matrix will after this smoothing be $L \times L$.
+where  $N_K = 2K + 1$ is the number of samples included in time, $N_L = M-L+1$ is the number of subarrays, and $\vec{x}_l^{\Delta}$ is the $l^\text{th}$ subarray $[x_l^{\Delta}[n], \dotso, x_{l+L-1}^{\Delta}[n]]$.
+The dimension of the covariance matrix after this smoothing is $L \times L$.
 Hence, the size of $\vec{w}[n]$ and $\vec{a}$ in (\ref{eq:w}) has to be $L \times 1$ when the covariance matrix in (\ref{eq:R}) is used.
@@
 \begin{align}\label{eq:diag_adapt}
-\epsilon &= d*tr\{\mat{\breve{R}}\}, \\
-d &\approx 1/L
+\epsilon &= d \times tr\{\mat{\breve{R}}\}, \\
+d &= 1/L
 \end{align}
@@
 \cite{Synnevag2007} has been applied in much of the resent literature on Capon beamforming for medical ultrasound imaging \cite{Nilsen2009, Wang2009, Mehdizadeh2012}.
-Recently we have also found that this way of doing diagonal loading also has been applied to HF and VHF antenna processing \cite{Featherstone1997}.
+This form of adaptive diagonal loading has also been applied to HF and VHF antenna processing \cite{Featherstone1997}.
 
@@
 %\end{align}
+where $\text{CONV}_m$ is a non-truncated convolution across elements.
 Note how we can either select to sum all subarrays first in (\ref{eq:z_mv}), or accumulate the right weighting per element in (\ref{eq:z_mv2}).
@@
 OpenCL.
-The key point is that GPU's can accelerate small concurrent tasks, like beamforming one pixel, much better than a CPU.
+The key point is that GPUs can accelerate small concurrent tasks, like beamforming one pixel, much better than a CPU.
 Where the CPU has few cores and logic to process complex threads, the GPU has many cores, but each thread needs to be simple in order to benefit from the extra computational power.
 
-The CUDA compute model is based on execution of a kernel function across a grid of compute threads, where the kernel function contains the work to be done by each thread.
+The CUDA compute model is based on execution of a kernel function across a grid of compute threads, where the kernel function describe the work to be done by each thread.
 As depicted in Figure \ref{fig:gpulayout}, each position in the grid holds a block of threads, with maximal size of 1024 threads, that can be synchronized with little cost if needed.
 %Each block is further divided into groups of 32 threads, known as a warp. The maximum number of threads inside a block is restricted to 512, hence 16 warps. 
-From a CUDA perspective the GPU consist of one or more stream multi processors (SM's), where each SM is capable of 32 or more, depending on the architecture, concurrent multiplications and/or additions.
+From a CUDA perspective the GPU consist of one or more stream multi processors (SMs), where each SM is capable of 32 or more, depending on the architecture, concurrent multiplications and/or additions.
 A group of blocks is scheduled to each SM until all blocks have been processed.
@@
 Each thread therefore has to be as light on memory as possible and has to perform enough instructions per used byte in order to hide memory latency.
-Global read and writes on the GPU, in addition to CPU-GPU transfers have high penalty, and should be minimized as much as possible.
+Global reads and writes on the GPU, in addition to CPU-GPU transfers, have high penalty, and should be minimized as much as possible.
 All of these constraints combined pose a real challenge as the target problem has to map on to the parallel framework both on a grid, block and memory consumption level in order to benefits from parallel acceleration.
-In the next section, we will describe how this mapping should be done for Capon beamforming of ultrasound data.
+In the next section, we will describe how this mapping should be done for Capon beamforming of cardiac ultrasound data.
 
@@
 As mentioned, widespread use of Capon adaptive beamforming has been limited by its computational complexity.
-However, if we analyse the beamformer in (\ref{eq:z_mv}) in combination with (\ref{eq:w}) and (\ref{eq:R}), we see that each calculation of a set of Capon weights is independent across the output $z[n]$.
+However, if we analyze the beamformer in (\ref{eq:z_mv}) in combination with (\ref{eq:w}) and (\ref{eq:R}), we see that each calculation of a set of Capon weights is independent across the outputs $z[n]$.
 As shown in Figure \ref{fig:gpulayout}, a first level of granularity is therefore to divide the image into a grid, where the output in each cell can be calculated independently of all others.
 Further, based on memory dependencies and to hold memory consumption per thread at a minimum, one should recognise that a group of threads should process one output in the final image.
-Continuing with the calculation of one weight vector, the algorithm can be broken down into three main steps that to some extent can be further divided into a small set of parallel processes.
+Continuing with the calculation of one weight vector, the algorithm can be broken down into three main steps that, to some extent, can be further divided into a small set of parallel processes.
 That is, for each data vector $\vec{x}^\Delta$ in the image:
@@
 The computational complexity of these three steps is $O(L^3)$, $O(L^3)$, and $O(L)$ respectively.
-As $L$ increases, the first two steps quickly becomes unmanageable, and real time performance can only be achieved by exploring novel solutions to reduce complexity.
+As $L$ increases, the first two steps quickly become unmanageable, and real-time performance can only be achieved by exploring novel solutions to reduce complexity.
 
@@
 
-The straight forward sum in (\ref{eq:R}) has complexity $O(L^2N_LN_K) = O(L^3)$ since $N_L$, the number of subarrays, is a function of $L$.
-This is actually the same as for inverting a $L \times L$ matrix.
+The straightforward sum in (\ref{eq:R}) has complexity $O(L^2N_LN_K) = O(L^3)$ since $N_L$, the number of subarrays, is a function of $L$.
+This is actually the same as for inverting an $L \times L$ matrix.
 However little attention has been devoted to this step in the literature before.
-Note that for a typical array configuration we typically do more spacial then temporal smoothing.
+Note that for a typical array configuration we do more spacial smoothing than temporal smoothing.
 Hence, $N_K$ is usually small compared to $N_L$.
-For short subarrays the construction of $\mat{\hat{R}}$ is dominated by the number of elements $M$ ($N_L = M - L + 1$), but since $L$ is selected in most cases to be between $M/2$ and $M/4$ a large $M$ implies a large $L$.
+The construction of $\mat{\hat{R}}$ is dominated by the number of elements $M$ ($N_L = M - L + 1$) for short subarrays, but since $L$ is selected in most cases to be between $M/2$ and $M/4$ a large $M$ implies a large $L$.
 
@@
 The quadratic complexity with $L$ and linear complexity with $N_K$ can be reduced by realizing that the calculation in (\ref{eq:R}) overlap both across subarrays and in time.
-The $(i,j)$ element of $\mat{\hat{R}}[n]$ can be calculated from element $(i-1, j-1), \text{for } i,j > 0$ by subtracting $x_{i-1}x_{j-1}^*$ and adding $x_{i+N_L-1}x_{j+N_L-1}^*$.
-The same is true in time where $\mat{\hat{R}}[n]$ can be calculated from $\mat{\hat{R}}[n-1]$ by subtracting $\mat{\hat{R}}[n-K-1]$ and adding $\mat{\hat{R}}[n+K]$.
+The $(i,j)$ element of $\mat{\hat{R}}[n]$ can be calculated from element $(i-1, j-1), \text{for } i,j \in [0, L-1]$  as
+\begin{align}
+\mat{\breve{R}}_{i,j}[n] &=  \mat{\breve{R}}_{i-1,j-1}[n] - \vec{x}_{i-1}^{\Delta}[n]\vec{x}_{j-1}^{\Delta}[n]^H \nonumber \\
+&+ \vec{x}_{i+N_L-1}^{\Delta}[n]\vec{x}_{j+N_L-1}^{\Delta}[n]^H
+\end{align}
+The same is true in time where
+\begin{align}
+\mat{\breve{R}}[n] = \mat{\breve{R}}[n-1] - \mat{\breve{R}}[n-K-1] + \mat{\breve{R}}[n-K]
+\end{align}
 Both sliding across subarrays and in time can be realized simultaneously if we hold a $N_K$-average in time of the outer products $\vec{x}^{\Delta}[n]\vec{x}^{\Delta}[n]^H$ in memory and calculate $\mat{\hat{R}}$ averaged over subarrays from that buffer (ref JOE).
@@
 Important to notice is that sliding will break independence in the sliding dimension.
-With our choice of calculating $\mat{\hat{R}}$, calculations are still independent across time, but not across subarrays.
-We therefore have selected to use one thread per column in $\mat{\hat{R}}$, and for small $L$'s several covariance matrices are calculated per thread block (Figure \ref{fig:gpulayout}).
+With our choice of calculating $\mat{\hat{R}}$, calculations are still independent across time, but not across elements.
+We therefore have selected to use one thread per column in $\mat{\hat{R}}$, and for small $L$s several covariance matrices are calculated per thread block (Figure \ref{fig:gpulayout}).
 Maximum supported $L$ is with this approach increased to 512, however for small $L$ high $M$ configurations, memory consumption per thread will be to high to achieve full occupancy of the GPU.
@@
 
-Diagonal loading as a constant value can be added with marginal cost, but the adaptive diagonal loading factor in (\ref{eq:diag}) is a different story.
-The trace of $\mat{\hat{R}}$ has to be known before diagonal elements are written back to global memory.
-Since one row for all $\mat{\hat{R}}$'s is written to global memory per kernel iteration, the solution was to hold the diagonal elements in a local buffer until the trace was accumulated, and then do scattered writes of the diagonal elements to global memory.
+Diagonal loading as a constant value can be added with marginal cost.
+For the adaptive diagonal loading factor in (\ref{eq:diag}), the trace of $\mat{\hat{R}}$ has to be known before diagonal elements are written back to global memory.
+Since one row for all covariance matrices is written to global memory per kernel iteration, the diagonal elements need to be held in a local memory until the trace is accumulated.
+When it is found, scattered writes of the diagonal elements to global memory has to be performed.
 
@@
 In the literature, when it comes to solving systems of linear equations on the GPU the focus has mostly been on large systems, and GPU libraries therefore often lack a comprehensive collection of batched solvers.
-The reason for this is two-fold, first the GPU needs to solve thousand of small systems in order to beat the low memory latency of the CPU, second there is a range of system dimensions that does not fit well with the GPU architecture, where only a small amount of near-core memory and registers are available.
+The reason for this is two-fold, first the GPU needs to solve thousands of small systems in order to beat the low memory latency of the CPU, second there is a range of system dimensions that does not fit well with the GPU architecture, where only a small amount of near-core memory and registers are available.
 Because of this, it has been proven hard to optimize a manageable set of solvers that provides speedups compared with the CPU in every cases.
@@
 
-For the results presented in this paper, we have used an unreleased GPU implementation of Gauss Jordan (GJ) elimination (by NVIDIA) to solve $\mat{\hat{R}}\vec{b} = \vec{1}_L$, where $\vec{b} = \mat{\hat{R}}^{-1}\vec{1}_L$.
+For the results presented in this paper, we have used an unreleased GPU implementation of Gauss-Jordan (GJ) elimination (by Nvidia) to solve $\mat{\hat{R}}\vec{b} = \vec{1}_L$, where $\vec{b} = \mat{\hat{R}}^{-1}\vec{1}_L$.
 For smaller systems, $L < 10$, a direct solver using one thread per system is preferable.
@@
 
-\section{Beamspace Processing}
+\section{Beamspace Processing}\label{sec:bs}
 Nilsen et al.
@@
 \end{align} 
-The steering vectors in $\mat{B}$, the so-called Butler matrix, determines each beams direction in space.
+The steering vectors in $\mat{B}$, the so-called Butler matrix, determine each beam's direction in space.
 We can therefore easily remove beams where no interference is present by removing rows in $\mat{B}$.
 For a focused system like a cardiac ultrasound scanner, where the received signal is concentrated in a narrow band around broadside, there is a small percentage of the beams that contains almost all the energy.
-Nilsen et.
-al concluded that as little as three beams around broadside where adequate to produce results comparable to applying Capon beamforming in element space.
+Nilsen et al.
+concluded that as little as three beams around broadside were adequate to produce results comparable to applying Capon beamforming in element space.
 

### 2012-08-15 HRUIonGPU/jrnl_article.tex
tuffc article

@@
 The quadratic complexity with $L$ and linear complexity with $N_K$ can be reduced by realizing that the calculation in (\ref{eq:R}) overlap both across subarrays and in time.
-The $(i,j)$ element of $\mat{\hat{R}}[n]$ can be calculated from element $(i-1, j-1), \text{for } i,j \in [0, L-1]$  as
+Element $(i,j)$ of $\mat{\hat{R}}[n]$ can be calculated from element $(i-1, j-1), \text{for } i,j \in [1, L-1]$ as
 \begin{align}
-\mat{\breve{R}}_{i,j}[n] &=  \mat{\breve{R}}_{i-1,j-1}[n] - \vec{x}_{i-1}^{\Delta}[n]\vec{x}_{j-1}^{\Delta}[n]^H \nonumber \\
-&+ \vec{x}_{i+N_L-1}^{\Delta}[n]\vec{x}_{j+N_L-1}^{\Delta}[n]^H
+\mat{\breve{R}}_{i,j}[n] &=  \mat{\breve{R}}_{i-1,j-1}[n]  + \sum_{n'=n-K}^{n+K} (x_{i+N_L-1}^{\Delta}[n']x_{j+N_L-1}^{\Delta}[n']^* \nonumber \\
+ &- x_{i-1}^{\Delta}[n']x_{j-1}^{\Delta}[n']^H) \label{eq:sliding}
 \end{align}
@@
 \begin{align}
-\mat{\breve{R}}[n] = \mat{\breve{R}}[n-1] - \mat{\breve{R}}[n-K-1] + \mat{\breve{R}}[n-K]
+\mat{\breve{R}}[n] &= \mat{\breve{R}}[n-1] + \sum_{l=0}^{N_L-1} (\vec{x}^{\Delta}_l[n+K]\vec{x}^{\Delta}_l[n+K]^H \nonumber \\
+&- \vec{x}^{\Delta}[n-K-1]\vec{x}^{\Delta}[n-K-1]^H).
 \end{align}
-Both sliding across subarrays and in time can be realized simultaneously if we hold a $N_K$-average in time of the outer products $\vec{x}^{\Delta}[n]\vec{x}^{\Delta}[n]^H$ in memory and calculate $\mat{\hat{R}}$ averaged over subarrays from that buffer (ref JOE).
-However, since $N_K$ typically is small, and we have a limited amount of memory per thread on the GPU, we have chosen to focus on sliding across subarrays only.
+%- \mat{\breve{R}}[n-K-1] + \mat{\breve{R}}[n-K].
+Both sliding across subarrays and in time can be realized simultaneously if we hold a $N_K$-average in time of the outer products $\vec{x}^{\Delta}[n]\vec{x}^{\Delta}[n]^H$ in memory and calculate $\mat{\hat{R}}$ averaged over subarrays from this buffer (ref JOE).
+However, since $N_K$ typically is small, and we have a limited amount of memory per thread on the GPU, we have chosen to focus on sliding across subarrays only (\ref{eq:sliding}).
 In that case, the complexity of calculating $\mat{\hat{R}}$ is reduced to $O(L^2)$.
 Important to notice is that sliding will break independence in the sliding dimension.
-With our choice of calculating $\mat{\hat{R}}$, calculations are still independent across time, but not across elements.
-We therefore have selected to use one thread per column in $\mat{\hat{R}}$, and for small $L$s several covariance matrices are calculated per thread block (Figure \ref{fig:gpulayout}).
+With our choice of calculating $\mat{\hat{R}}$, calculations are still independent across time, but not along diagonals in $\mat{\hat{R}}$.
+We therefore have selected to use one thread per diagonal in the upper triangle of $\mat{\hat{R}}$, and for small $L$s several covariance matrices are calculated per thread block (Figure \ref{fig:gpulayout}).
 Maximum supported $L$ is with this approach increased to 512, however for small $L$ high $M$ configurations, memory consumption per thread will be to high to achieve full occupancy of the GPU.
@@
 Comment on different choices of parameters change the running times.
-Temporal averaging for instance, lowers the efficiency since $2*Y_{avg}$ more operations has to be performed per $\mat{\hat{R}}$
+Temporal averaging for instance, lowers the efficiency since $2Y_{avg}$ more operations has to be performed per $\mat{\hat{R}}$
 
@@
 If we do a 2-element pre-beamforming to get the channel count down to 32, the frame rate increases to 44fps.
-For a 32 element phased array we need less beams to cover the sector (40*400 pixels), hence with the same parameters the frame rate increases to 87fps.
+For a 32 element phased array we need less beams to cover the sector ($40 \times 400$ pixels), hence with the same parameters the frame rate increases to 87fps.
 The bottleneck of the Capon algorithm is to find the inverse covariance matrix for all pixels.
@@
 % biographies
-\newpage
+%\newpage
 

### 2012-09-11 HRUIonGPU/conf_article.tex
started on IUS article

--- 2012-08-15 HRUIonGPU/jrnl_article.tex
@@
-%% bare_jrnl.tex
+%% bare_conf.tex
 %% V1.3
@@
 %% by Michael Shell
-%% see http://www.michaelshell.org/
+%% See:
+%% http://www.michaelshell.org/
 %% for current contact information.
@@
 %% This is a skeleton file demonstrating the use of IEEEtran.cls
-%% (requires IEEEtran.cls version 1.7 or later) with an IEEE journal paper.
+%% (requires IEEEtran.cls version 1.7 or later) with an IEEE conference paper.
 %%
@@
 %% http://www.ieee.org/
-
-% *** Authors should verify (and, if needed, correct) their LaTeX system  ***
-% *** with the testflow diagnostic prior to trusting their LaTeX platform ***
-% *** with production work. IEEE's font choices can trigger bugs that do  ***
-% *** not appear when using other class files.                            ***
-% The testflow support page is at:
-% http://www.michaelshell.org/tex/testflow/
 
@@
 
+% *** Authors should verify (and, if needed, correct) their LaTeX system  ***
+% *** with the testflow diagnostic prior to trusting their LaTeX platform ***
+% *** with production work. IEEE's font choices can trigger bugs that do  ***
+% *** not appear when using other class files.                            ***
+% The testflow support page is at:
+% http://www.michaelshell.org/tex/testflow/
+
 % Note that the a4paper option is mainly intended so that authors in
@@
 %
-\documentclass[journal]{IEEEtran}
+\documentclass[conference]{IEEEtran}
+% Add the compsoc option for Computer Society conferences.
 %
@@
 % manually specify the path to it like:
-% \documentclass[journal]{../sty/IEEEtran}
+% \documentclass[conference]{../sty/IEEEtran}
 
@@
 
-%\ifCLASSOPTIONcaptionsoff
-%  \usepackage[nomarkers]{endfloat}
-% \let\MYoriglatexcaption\caption
-% \renewcommand{\caption}[2][\relax]{\MYoriglatexcaption[#2]{#2}}
-%\fi
-% endfloat.sty was written by James Darrell McCauley and Jeff Goldberg.
-% This package may be useful when used in conjunction with IEEEtran.cls'
-% captionsoff option. Some IEEE journals/societies require that submissions
-% have lists of figures/tables at the end of the paper and that
-% figures/tables without any captions are placed on a page by themselves at
-% the end of the document. If needed, the draftcls IEEEtran class option or
-% \CLASSINPUTbaselinestretch interface can be used to increase the line
-% spacing as well. Be sure and use the nomarkers option of endfloat to
-% prevent endfloat from "marking" where the figures would have been placed
-% in the text. The two hack lines of code above are a slight modification of
-% that suggested by in the endfloat docs (section 8.3.1) to ensure that
-% the full captions always appear in the list of figures/tables - even if
-% the user used the short optional argument of \caption[]{}.
-% IEEE papers do not typically make use of \caption[]'s optional argument,
-% so this should not be an issue. A similar trick can be used to disable
-% captions of packages such as subfig.sty that lack options to turn off
-% the subcaptions:
-% For subfig.sty:
-%\let\MYorigsubfloat\subfloat
-%\renewcommand{\subfloat}[2][\relax]{\MYorigsubfloat[]{#2}}
-% For subfigure.sty:
-% \let\MYorigsubfigure\subfigure
-% \renewcommand{\subfigure}[2][\relax]{\MYorigsubfigure[]{#2}}
-% However, the above trick will not work if both optional arguments of
-% the \subfloat/subfig command are used. Furthermore, there needs to be a
-% description of each subfigure *somewhere* and endfloat does not add
-% subfigure captions to its list of figures. Thus, the best approach is to
-% avoid the use of subfigure captions (many IEEE journals avoid them anyway)
-% and instead reference/explain all the subfigures within the main caption.
-% The latest version of endfloat.sty and its documentation can obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/endfloat/
-%
-% The IEEEtran \ifCLASSOPTIONcaptionsoff conditional can also be used
-% later in the document, say, to conditionally put the References on a 
-% page by themselves.
-
 % *** PDF, URL AND HYPERLINK PACKAGES ***
@@
 \title{Implementing Capon Adaptive Beamforming on the GPU for Real Time Cardiac Ultrasound Imaging}
-%
-%
-% author names and IEEE memberships
-% note positions of commas and nonbreaking spaces ( ~ ) LaTeX will not break
-% a structure at a ~ so this keeps an author's name from being broken across
-% two lines.
-% use \thanks{} to gain access to the first footnote area
-% a separate \thanks must be used for each paragraph as LaTeX2e's \thanks
-% was not built to handle multiple paragraphs
-%
 
-\author{J. P. \AA{}sen, J. I. Buskenes, C.-I. C. Nilsen, A. Austeng and S. Holm%\IEEEmembership{Member,~IEEE,}% <-this % stops a space
-%\thanks{M. Shell is with the Department
-%of Electrical and Computer Engineering, Georgia Institute of Technology, Atlanta,
-%GA, 30332 USA e-mail: (see http://www.michaelshell.org/contact.html).}% <-this % stops a space
-%\thanks{J. Doe and J. Doe are with Anonymous University.}% <-this % stops a space
-%\thanks{Manuscript received April 19, 2005; revised January 11, 2007.}}
+% author names and affiliations
+% use a multiple column layout for up to three different
+% affiliations
+\author{
+\IEEEauthorblockN{
+Jon Petter \AA{}sen\IEEEauthorrefmark{1},
+Jo Inge Buskenes\IEEEauthorrefmark{2},
+Carl-Inge C. Nilsen\IEEEauthorrefmark{2},
+Andreas Austeng\IEEEauthorrefmark{2},
+Sverre Holm\IEEEauthorrefmark{1}\IEEEauthorrefmark{2}
+}
+\IEEEauthorblockA{
+\IEEEauthorrefmark{1} Mi Lab, Norwegian University of Science and Technology, Trondheim, Norway,\\
+\IEEEauthorrefmark{2} Department of Informatics, University of Oslo, Oslo, Norway
+}
 }
 
-% note the % following the last \IEEEmembership and also \thanks - 
-% these prevent an unwanted space from occurring between the last author name
-% and the end of the author line. i.e., if you had this:
+% conference papers do not typically use \thanks and this command
+% is locked out in conference mode. If really needed, such as for
+% the acknowledgment of grants, issue a \IEEEoverridecommandlockouts
+% after \documentclass
+
+% for over three affiliations, or if they all won't fit within the width
+% of the page, use this alternative format:
 % 
-% \author{....lastname \thanks{...} \thanks{...} }
-%                     ^------------^------------^----Do not want these spaces!
-%
-% a space would be appended to the last name and could cause every name on that
-% line to be shifted left slightly. This is one of those "LaTeX things". For
-% instance, "\textbf{A} \textbf{B}" will typeset as "A B" not "AB". To get
-% "AB" then you have to do: "\textbf{A}\textbf{B}"
-% \thanks is no different in this regard, so shield the last } of each \thanks
-% that ends a line with a % and do not let a space in before the next \thanks.
-% Spaces after \IEEEmembership other than the last one are OK (and needed) as
-% you are supposed to have spaces between the names. For what it is worth,
-% this is a minor point as most people would not even notice if the said evil
-% space somehow managed to creep in.
-
-% The paper headers
-\markboth{Transactions on Ultrasonics and Ferroelectrics and Frequency Control,~Vol.~x, No.~y, January~2013?}%
-{\AA{}sen \MakeLowercase{\textit{et al.}}: Capon Adaptive Beamforming on the GPU}
-% The only time the second header will appear is for the odd numbered pages
-% after the title page when using the twoside option.
-% 
-% *** Note that you probably will NOT want to include the author's ***
-% *** name in the headers of peer review papers.                   ***
-% You can use \ifCLASSOPTIONpeerreview for conditional compilation here if
-% you desire.
-
-% If you want to put a publisher's ID mark on the page you can do it like
-% this:
-%\IEEEpubid{0000--0000/00\$00.00~\copyright~2007 IEEE}
-% Remember, if you use this you must call \IEEEpubidadjcol in the second
-% column for its text to clear the IEEEpubid mark.
+%\author{\IEEEauthorblockN{Michael Shell\IEEEauthorrefmark{1},
+%Homer Simpson\IEEEauthorrefmark{2},
+%James Kirk\IEEEauthorrefmark{3}, 
+%Montgomery Scott\IEEEauthorrefmark{3} and
+%Eldon Tyrell\IEEEauthorrefmark{4}}
+%\IEEEauthorblockA{\IEEEauthorrefmark{1}School of Electrical and Computer Engineering\\
+%Georgia Institute of Technology,
+%Atlanta, Georgia 30332--0250\\ Email: see http://www.michaelshell.org/contact.html}
+%\IEEEauthorblockA{\IEEEauthorrefmark{2}Twentieth Century Fox, Springfield, USA\\
+%Email: homer@thesimpsons.com}
+%\IEEEauthorblockA{\IEEEauthorrefmark{3}Starfleet Academy, San Francisco, California 96678-2391\\
+%Telephone: (800) 555--1212, Fax: (888) 555--1212}
+%\IEEEauthorblockA{\IEEEauthorrefmark{4}Tyrell Inc., 123 Replicant Street, Los Angeles, California 90210--4321}}
 
@@
 \begin{abstract}
-%\boldmath
-An amazing abstract goes here. An amazing abstract goes here. An amazing abstract goes here. An amazing abstract goes here.
-An amazing abstract goes here. An amazing abstract goes here. An amazing abstract goes here. An amazing abstract goes here.
-An amazing abstract goes here. An amazing abstract goes here. An amazing abstract goes here. An amazing abstract goes here.
-An amazing abstract goes here. An amazing abstract goes here. An amazing abstract goes here. An amazing abstract goes here.
-An amazing abstract goes here. An amazing abstract goes here. An amazing abstract goes here. An amazing abstract goes here.
-An amazing abstract goes here. An amazing abstract goes here. An amazing abstract goes here. An amazing abstract goes here.
-An amazing abstract goes here. An amazing abstract goes here. An amazing abstract goes here. An amazing abstract goes here.
-An amazing abstract goes here. An amazing abstract goes here. An amazing abstract goes here. An amazing abstract goes here.
-An amazing abstract goes here. An amazing abstract goes here. An amazing abstract goes here. An amazing abstract goes here.
-An amazing abstract goes here. An amazing abstract goes here. An amazing abstract goes here. An amazing abstract goes here.
+\boldmath
+Recent results \cite{Synnevag2007, Austeng2008, Vignon2008, Viola, Mehdizadeh2012} suggest that the Capon beamformer can provide increased lateral resolution to a medical ultrasound image.
+In this paper, the high computational complexity of the Capon beamformer is targeted with the use of Graphics Processing Units (GPUs), and we show in-vivio results with Capon beamforming applied on a full cardiac cycle in addition to simulations.
+With GPUs we are able to process a 70 degree sector cardiac image from a 64 element phased array at interactive frame rates using both spatial and temporal smoothing.
+
+For a typical cardiac ultrasound image of $80\times400$ pixels (70 degree sector, 15cm range) acquired using a 2.5MHz, M=64 element phased array, we obtain 10fps (subarray length L=M/2, temporal smoothing over 3 samples).
+If we do a 2-element pre-beamforming to get the channel count down to 32, the frame rate increases to 44fps.
+For a 32 element phased array we need less beams to cover the sector ($40\times400$ pixels), hence with the same parameters the frame rate increases to 87fps.
+The bottleneck of the Capon algorithm is to find the inverse covariance matrix for all pixels.
+This is an $O(L^3)$ operation, and results show that when L is larger than 16 (M = 64), real time performance is not possible for cardiac imaging with current high-end hardware.
+The target GPU has been Nvidia’s Quadro 6000, capable of 1Tflops.
+
 \end{abstract}
@@
 % This preserves the distinction between vectors and scalars. However,
-% if the journal you are submitting to favors bold math in the abstract,
+% if the conference you are submitting to favors bold math in the abstract,
 % then you can use LaTeX's standard command \boldmath at the very start
-% of the abstract to achieve this. Many IEEE journals frown on math
-% in the abstract anyway.
+% of the abstract to achieve this. Many IEEE journals/conferences frown on
+% math in the abstract anyway.
 
-% Note that keywords are not normally used for peerreview papers.
-\begin{IEEEkeywords}
-Adaptive Beamforming, Capon Beamformer, Minimum Variance Beamforming, Graphics Processing Unit, GPGPU, Parallel Computing, Ultrasound Imaging.
-\end{IEEEkeywords}
+% no keywords
 
@@
 \section{Introduction}
-% The very first letter is a 2 line initial drop letter followed
-% by the rest of the first word in caps.
-% 
-% form to use if the first word consists of a single letter:
-% \IEEEPARstart{A}{demo} file is ....
-% 
-% form to use if you need the single drop letter followed by
-% normal text (unknown if ever used by IEEE):
-% \IEEEPARstart{A}{}demo file is ....
-% 
-% Some journals put the first two words in caps:
-% \IEEEPARstart{T}{his demo} file is ....
-% 
-% Here we have the typical use of a "T" for an initial drop letter
-% and "HIS" in caps to complete the first word.
-
-%\IEEEPARstart{I}{ntroduction} to Capon adaptive beamforming, ultrasound imaging and GPU/CUDA.
-\IEEEPARstart{C}{apon} adaptive beamforming \cite{Capon1969} applied to medical ultrasound imaging has recently shown promising results in several studies \cite{Synnevag2007, Austeng2008, Vignon2008, Viola, Mehdizadeh2012}.
-It has been shown that the method introduces some interesting trade-offs in addition to just improved resolution, like maintaining image resolution in combination with higher framerates, smaller probes, or deeper penetration \cite{Synnevag2009}.
+Capon adaptive beamforming \cite{Capon1969} applied to medical ultrasound imaging has recently shown promising results in several studies \cite{Synnevag2007, Austeng2008, Vignon2008, Viola, Mehdizadeh2012}.
+For instance it has been introduced some interesting trade-offs in addition to just improved resolution, like maintaining image resolution in combination with higher framerates, smaller probes, or deeper penetration \cite{Synnevag2009}.
 Despite these benefits, widespread use has been limited by the immense computational complexity associated with the method \cite{So2011}.
 In this paper we introduce the first implementation of the Capon beamformer exhibiting real-time performance when applied in a typical cardiac ultrasound imaging setting.
-To accomplish this task we make use of the massive, and easily accessed, parallel processing power found in modern Graphics Processing Units (GPU's), combined with novel methods to reduce the computational complexity as the problem size grows.
+To accomplish this task we make use of the massive, and easily accessed, parallel processing power found in modern Graphics Processing Units (GPUs), combined with novel methods to reduce the computational complexity as the problem size grows.
 
-Previous work on implementing a parallel Capon beamformer have mainly been focused around radar and passive sonar systems using custom made systolic array processors \cite{McWhirter1989, Moonen1993, Sinha2002}.
-%A systolic array processor is custom made hardware, and even if the computational throughput might be higher for selected problems compared with a GPU, the cost related to development and flexibility is considered much higher.
+Previous work on implementing a parallel Capon beamformer have mainly been focused around radar and passive sonar systems using custom made systolic array processors \cite{McWhirter1989, Sinha2002}.
 For medical ultrasound, Chen et al.
@@
 
-%Using real data will results in symmetric windows, and even if the number of samples they managed to process per second was high, both the number of floating point operations per sample and the imaging region is reduced using real data compared with use of complex data.
-It is crucial to use complex data within the Capon beamformer framework, both for reducing the amount of data per image, and to allow for asymmetric windows.
-All results in this paper is reported with use of critically sampled complex data, both laterally and in range.
-(ikke helt kritisk samplet, ikke si det, bare at det er i baseband).
-
-% denne må utelates i IUS-artikkelen
-In recent work (cite JOE) we have applied a GPU-based Capon beamformer in an active sonar system.
-For sonar the real-time requirement is much lower than for medical ultrasound imaging, which means that most common array configurations can be handled using the GPU without taking any additional measures.
-For medical ultrasound imaging, and especially cardiac ultrasound imaging, frame rate requirements increase to a level where even a high-end GPU can not do the processing required for a standard implementation of Capon beamforming.
-Since the execution time for a Capon beamformer depends highly on the number of elements in the given array, it is important to find ways of reducing the computational complexity for high-channel-count systems.
-Different approaches to low-complexity Capon beamforming have recently been proposed \cite{Asl2012, Jensen2012}, and in this paper we have selected to implement the beamspace transformation described by Nilsen and Hafizovic \cite{Nilsen2009} on the GPU to maintain high framerates as the channel count increases.
-%(Referer til JOE-artikkelen, få med beskrivelse av hva som blir anerledes for ultralyd.)
-
-In the next section we present background information about the Capon beamformer and the GPU computing model.
+%In the next section we present background information about the Capon beamformer and the GPU computing model.
 In Section \ref{sec:meth} we go in detail about the parallel implementation of the Capon beamformer, before in Section \ref{sec:bs} we introduce the method applied when the problem size increases.
@@
 \section{Background}
-
-%\subsection{Capon Beamforming}
-
-%\textbf{Bruk riktig terminologi. $N_{avg}$ er K. Antall sub-arrayer må hete noe annet. F.eks $N_L$}
-
-%\textbf{Presis på hvilke formler vi skal implementere.}
-
-%\textbf{De skal forstå MV etter denne seksjonen.}
-
-%\textbf{Ton ned introduksjonen, og sett opp akkurat de ligningene vi implementerer.}
-
-%\textbf{(Skriv om den introduserende setningen, start rett på linje to, Kombinere data fra mange sensorer, Leseren kan Ultralyd ikke gpu)}
-
+\subsection{Capon Adaptive Beamforming}
 The conventional form of array beamforming is delay-and-sum (DAS).
@@
 
-The Capon weight vector is given by the following expression:
-\begin{align}\label{eq:w}
-\vec{w}[n] = \frac{\mat{\hat{R}}[n]^{-1}\vec{a}}{\vec{a}^H\mat{\hat{R}}[n]^{-1}\vec{a}} = \frac{\vec{b}}{\vec{a}^H\vec{b}},
-\end{align}
-where $\vec{a} = [1_0, 1_1, ..., 1_{M-1}]^T = \vec{1}_M$ when $\vec{x}$ is pre-delayed, $\mat{\hat{R}}$ is the sample covariance matrix calculated from the data vector $\vec{x}^\Delta[n]$, and $\vec{b} = \mat{\hat{R}}[n]^{-1}\vec{a}$.
-%\begin{align}
-%\mat{\hat{R}}[n] = \vec{x}[n]\vec{x}[n]^H.
-%\end{align} 
-%Typically the sample covariance matrix has to be averaged over several samples in order to ensure numerical stability and statistical robustness.
-
-In passive systems, assuming stationary statistics, the sample covariance matrix is typically averaged over a long sequence of samples\cite{Krima}.
-For an active, pulsed system like medical ultrasound imaging, the available data from which $\mat{\hat{R}}$ can be estimated is quite limited.
-The resulting sample covariance matrix might therefore be both inaccurate, and poorly conditioned.
-Another challenge with applying Capon to active systems is the effect of signal cancellation, where high correlation between signal and interference favors weights that cancel out the signal \cite{Reddy1987}.
-In this way, the beamformer output can be too low at broadside, even when there is signal present and the distortionless-response criterion is applied.
-
-In order to get a well-conditioned $\mat{\hat{R}}$, avoid signal cancellation, and to get DAS-like speckle, $\mat{\hat{R}}$ has to be averaged over $L\le M/2$ long subarrays and $N_K \sim T_p/T_s$ time samples \cite{Synnevag2007, Synnevag2007a}.
-Where, $T_p$ is the pulse length in seconds and $T_s$ is the sampling period.
-We therefore calculate
+To calculate the Capon beamformer output we need three equations. First the sample covariance matrix has to be estimated,
 \begin{align}
@@
 where  $N_K = 2K + 1$ is the number of samples included in time, $N_L = M-L+1$ is the number of subarrays, and $\vec{x}_l^{\Delta}$ is the $l^\text{th}$ subarray $[x_l^{\Delta}[n], \dotso, x_{l+L-1}^{\Delta}[n]]$.
-The dimension of the covariance matrix after this smoothing is $L \times L$.
-Hence, the size of $\vec{w}[n]$ and $\vec{a}$ in (\ref{eq:w}) has to be $L \times 1$ when the covariance matrix in (\ref{eq:R}) is used.
-Finally, the covariance matrix is loaded with a diagonal factor $\epsilon$ to ensure numerical stability, 
+Note that the dimension of the covariance matrix is $L \times L$.
+The covariance matrix is also loaded with a diagonal factor $\epsilon$ to ensure numerical stability,
 \begin{align}\label{eq:diag}
-\mat{\hat{R}} =  f(\vec{x}^{\Delta}[n]) = \mat{\breve{R}} + \epsilon\mat{I}.
+\mat{\hat{R}} = \mat{\breve{R}} + \epsilon\mat{I}.
 \end{align}
@@
 \end{align}
-introduced by Synnev\aa{}g et al.
-\cite{Synnevag2007} has been applied in much of the resent literature on Capon beamforming for medical ultrasound imaging \cite{Nilsen2009, Wang2009, Mehdizadeh2012}.
-This form of adaptive diagonal loading has also been applied to HF and VHF antenna processing \cite{Featherstone1997}.
+has been applied in much of the resent literature on Capon beamforming for medical ultrasound imaging \cite{Synnevag2007, Mehdizadeh2012}.
 
-The beamformer output is in the end calculated by combining all $N_L$ subarrays weighted with the same set of adaptive weights.
+Second, we calculate the Capon weight vector as
+\begin{align}\label{eq:w}
+\vec{w}[n] = \frac{\mat{\hat{R}}[n]^{-1}\vec{a}}{\vec{a}^H\mat{\hat{R}}[n]^{-1}\vec{a}} = \frac{\vec{b}}{\vec{a}^H\vec{b}},
+\end{align}
+where $\vec{a} = [1_0, 1_1, ..., 1_{L-1}]^T = \vec{1}_L$ when $\vec{x}$ is pre-delayed.
+
+And last, the beamformer output is calculated by combining all $N_L$ subarrays weighted with the same set of adaptive weights,
 \begin{align}
-z[n] &= \frac{1}{N_L}\vec{w}[n]^H \sum_{l=0}^{N_L-1} \vec{x}_l^{\Delta}[n] \label{eq:z_mv}\\
-&= 1/N_L \, \text{CONV}_m(\vec{w}[n], \vec{1}_{N_L})^H\vec{x}^{\Delta}[n] \label{eq:z_mv2},
+z[n] &= \frac{1}{N_L}\vec{w}[n]^H \sum_{l=0}^{N_L-1} \vec{x}_l^{\Delta}[n]. \label{eq:z_mv}
 \end{align}
-%where (this matrix is wrong)
+
+%The Capon weight vector is given by the following expression:
+%\begin{align}\label{eq:w}
+%\vec{w}[n] = \frac{\mat{\hat{R}}[n]^{-1}\vec{a}}{\vec{a}^H\mat{\hat{R}}[n]^{-1}\vec{a}} = \frac{\vec{b}}{\vec{a}^H\vec{b}},
+%\end{align}
+%where $\vec{a} = [1_0, 1_1, ..., 1_{M-1}]^T = \vec{1}_M$ when $\vec{x}$ is pre-delayed, $\mat{\hat{R}}$ is the sample covariance matrix calculated from %the data vector $\vec{x}^\Delta[n]$, and $\vec{b} = \mat{\hat{R}}[n]^{-1}\vec{a}$.
+
+%In order to get a well-conditioned $\mat{\hat{R}}$, avoid signal cancellation, and to get DAS-like speckle, $\mat{\hat{R}}$ has to be averaged over $L\le M/2$ long subarrays and $N_K \sim T_p/T_s$ time samples \cite{Synnevag2007, Synnevag2007a}.
+Where, $T_p$ is the pulse length in seconds and $T_s$ is the sampling period.
+We therefore calculate
 %\begin{align}
-%\mat{C} = \left(
-%\begin{matrix}
-%\vec{1}_1 & \vec{1}_2 & \cdots & \vec{1}_{N_L} & \vec{0}_1 & \cdots & \vec{0}_{N_L-1} & \vec{0}_{N_L}\\
-%\vec{0}_{N_L-1} & \vec{0}_{N_L-2} & \cdots & \vec{0}_{0} & \vec{1}_{N_L-1} & \cdots & \vec{1}_{2} & \vec{1}_{1}\\
-%\end{matrix}
-%\right)^T
+%\mat{\breve{R}}[n] = \frac{1}{N_LN_K}\sum_{n'=n-K}^{n+K} \sum_{l=0}^{N_L-1} \vec{x}_l^{\Delta}[n']\vec{x}_l^{\Delta}[n']^H,\label{eq:R}
 %\end{align}
-where $\text{CONV}_m$ is a non-truncated convolution across elements.
-Note how we can either select to sum all subarrays first in (\ref{eq:z_mv}), or accumulate the right weighting per element in (\ref{eq:z_mv2}).
+%where  $N_K = 2K + 1$ is the number of samples included in time, $N_L = M-L+1$ is the number of subarrays, and $\vec{x}_l^{\Delta}$ is the $l^\text{th}$ subarray $[x_l^{\Delta}[n], \dotso, x_{l+L-1}^{\Delta}[n]]$.
+The dimension of the covariance matrix after this smoothing is $L \times L$.
+Hence, the size of $\vec{w}[n]$ and $\vec{a}$ in (\ref{eq:w}) has to be $L \times 1$ when the covariance matrix in (\ref{eq:R}) is used.
+%Finally, the covariance matrix is loaded with a diagonal factor $\epsilon$ to ensure numerical stability, 
+%\begin{align}\label{eq:diag}
+%\mat{\hat{R}} = \mat{\breve{R}} + \epsilon\mat{I}.
+%\end{align}
+%A weighting proportional to the output power is often used to reduce the need of parameter adjustments, and the trace of $\mat{\breve{R}}$, 
+%\begin{align}\label{eq:diag_adapt}
+%\epsilon &= d \times tr\{\mat{\breve{R}}\}, \\
+%d &= 1/L
+%\end{align}
+%has been applied in much of the resent literature on Capon beamforming for medical ultrasound imaging \cite{Synnevag2007, Wang2009, Mehdizadeh2012}.
 
-% Få med at vi processerer hele bildet på en gang. Forklar forskjellen med å prosessere en beam av gangen. Hva med MLA?
-% Illustrer tråder og tider for hvordan processeringen går.
-% Vi snakker om mottak-stråleformeren.
-
-%\begin{figure}
-%\includegraphics[width=3in]{gfx/beamforming_mv_lowres.png}
-%\caption{}
-%\label{fig:mv}
-%\end{figure}
-
-%Building and calculating the inverse of $\hat{R}$ is computational demanding, limiting the methods accessibility. Following the innovation in GPU computing, this has now started to change. 
-%In this paper we introduce the first GPU implementation of the Capon beamformer capable of processing a 70 degree sector cardiac image from a 64 element phased array at interactive frame rates using both spatial and temporal smoothing.
-Both types of smoothing are important for maintaining delay-and-sum-like speckle and at the same time keep the high resolution property.
-This is achieved using complex base band data, which reduces the size of each range line to a minimum.
-In addition, complex data results in complex weights; hence this is a correct implementation of Capon beamforming with all degrees of freedom preserved.
-For high channel count systems we propose two ways of reducing computational complexity while maintaining
+%The beamformer output is in the end calculated by combining all $N_L$ subarrays weighted with the same set of adaptive weights,
+%\begin{align}
+%z[n] &= \frac{1}{N_L}\vec{w}[n]^H \sum_{l=0}^{N_L-1} \vec{x}_l^{\Delta}[n]. \label{eq:z_mv}
+%\end{align}
 
 \subsection{GPU Compute Model}
-
-%\textbf{(Ta med mer banale ting, dette er GPU for ultralydfolk)}
-
-%\textbf{(Etter denne så må de forstå GPU og hvorfor MV passer.)}
-
-In this section we briefly discuss the principle behind GPU computing, laying the basis for further discussion on parallel implementation strategies.
 CUDA by Nvidia has been our choice of GPU developer environment \cite{Nvidia2011}.
-However, even though CUDA has been used to implement the Capon beamformer, the challenges and proposed methods described in this paper apply for other massive parallel implementation environments as well, like e.g.
-OpenCL.
-The key point is that GPUs can accelerate small concurrent tasks, like beamforming one pixel, much better than a CPU.
-Where the CPU has few cores and logic to process complex threads, the GPU has many cores, but each thread needs to be simple in order to benefit from the extra computational power.
-
 The CUDA compute model is based on execution of a kernel function across a grid of compute threads, where the kernel function describe the work to be done by each thread.
@@
 
-%When implementing algorithms on the GPU it is important to keep data as close to the core as possible. Utilizing available registers and shared memory to its maximum is therefore important.
-
-%More important, the CUDA 2.x compute capability (CC) has available 48 KB of near-core shared memory per SM that can be utilized by resident blocks.
-The maximum number of resident blocks per SM is 8, but in addition the maximum number of resident warps per SM is restricted to 48.
-Hence, with a block size of 512 only $48/(512/32) = 3$ blocks can occupy the SM at once.
-In addition, if one thread where to use 64B of shared memory the number of allowed resident blocks with a block size of 512 would be reduced to 1.
-Up to a curtain point it is beneficial to have as many warps as possible occupying the SM at once to hide latency from accessing memory.
-The SM has for CC 2.x 32K 32-bit long registers.
-When implementing algorithms on the GPU it is important to keep data as close to the core as possible.
-Utilizing available registers and shared memory to its maximum is therefore important.
-
-%\subsection{Optimizing kernel launch}
-%Set the size of grid adaptively based on $L$, $M$ and $Yavg$.
-%It is better to slide in space than time. $K$ is typically much larger than $2*Yavg + 1$.
-
-% needed in second column of first page if using \IEEEpubid
-%\IEEEpubidadjcol
-
 \section{Parallel Capon}\label{sec:meth}
-As mentioned, widespread use of Capon adaptive beamforming has been limited by its computational complexity.
-However, if we analyze the beamformer in (\ref{eq:z_mv}) in combination with (\ref{eq:w}) and (\ref{eq:R}), we see that each calculation of a set of Capon weights is independent across the outputs $z[n]$.
+If we analyze the beamformer in (\ref{eq:z_mv}) in combination with (\ref{eq:w}) and (\ref{eq:R}), we see that each calculation of a set of Capon weights is independent across the outputs $z[n]$.
 As shown in Figure \ref{fig:gpulayout}, a first level of granularity is therefore to divide the image into a grid, where the output in each cell can be calculated independently of all others.
@@
 \end{align*}
-%\begin{enumerate}
-%\item Calculate the sample covariance matrix $\mat{\hat{R}}$ as in (\ref{eq:R})
-%\end{enumerate}
-%\begin{enumerate}
-%\item[2)] Solve the linear system of equations $\mat{\hat{R}}\vec{b} =\vec{ 1}$
-%\end{enumerate}
-%\begin{enumerate}
-%\item[3)] Calculate $\vec{w} = \vec{b}/\vec{1}^H\vec{b}$ and the amplitude $\vec{w}^H\vec{x}$
-%\end{enumerate}
+
 The computational complexity of these three steps is $O(L^3)$, $O(L^3)$, and $O(L)$ respectively.
-As $L$ increases, the first two steps quickly become unmanageable, and real-time performance can only be achieved by exploring novel solutions to reduce complexity.
-
-%(The three steps have, for modularity, been implemented using three different kernels. Joining all kernels into one could possible lead to an increase in speed.)
-
-%\begin{figure}
-%\includegraphics[width=3in]{gfx/gpu_layout_lowres.png}%
-%\caption{}
-%\label{fig:gpulayout}
-%\end{figure}
 
 \subsection{Calculating Multiple Sample Covariance Matrices}
-
 The straightforward sum in (\ref{eq:R}) has complexity $O(L^2N_LN_K) = O(L^3)$ since $N_L$, the number of subarrays, is a function of $L$.
@@
 Hence, $N_K$ is usually small compared to $N_L$.
-The construction of $\mat{\hat{R}}$ is dominated by the number of elements $M$ ($N_L = M - L + 1$) for short subarrays, but since $L$ is selected in most cases to be between $M/2$ and $M/4$ a large $M$ implies a large $L$.
+For short subarrays the construction of $\mat{\hat{R}}$ is dominated by the number of elements $M$ ($N_L = M - L + 1$), but since $L$  in most cases is selected to be between $M/4$ and $M/2$, many array elements implies large subarrays.
 
-From (\ref{eq:R}) it is obvious that calculation of one element in $\mat{\hat{R}}$ is independent of all the others.
+From (\ref{eq:R}) it is obvious that calculation of one element in $\mat{\hat{R}}$ is independent of all the other elements.
 A natural level of granularity would then be to assign one thread to each element of $\mat{\hat{R}}$.
-This will minimize global reads and writes needed per thread, but unfortunately the maximum number of threads per block puts an upper limit on the subarray length, $L_{max}=32$, if we want to avoid block-to-block synchronization.
+This will minimize global reads and writes needed per thread, but unfortunately the maximum number of threads per block puts an upper limit on the subarray length, $L_{max}=32$, since we want to avoid block-to-block synchronization.
 
@@
 %- \mat{\breve{R}}[n-K-1] + \mat{\breve{R}}[n-K].
-Both sliding across subarrays and in time can be realized simultaneously if we hold a $N_K$-average in time of the outer products $\vec{x}^{\Delta}[n]\vec{x}^{\Delta}[n]^H$ in memory and calculate $\mat{\hat{R}}$ averaged over subarrays from this buffer (ref JOE).
+Both sliding across subarrays and in time can be realized simultaneously if we hold a $N_K$-average in time of the outer products $\vec{x}^{\Delta}[n]\vec{x}^{\Delta}[n]^H$ in memory and calculate $\mat{\hat{R}}$ averaged over subarrays from this buffer.
 However, since $N_K$ typically is small, and we have a limited amount of memory per thread on the GPU, we have chosen to focus on sliding across subarrays only (\ref{eq:sliding}).
@@
 However, the way we have decided to organize the GPU compute threads per $\mat{\hat{R}}$, the lower triangle can be calculated with little overhead on the GPU.
-This also means that we do not relay on a symmetric solver to solve $\mat{\hat{R}}\vec{b} = \vec{1}_L$, even if that could have saved some instructions.
-The memory layout for the upper triangle only is also believed to be cumbersome.
+This also means that we do not relay on a upper triangular solver to solve $\mat{\hat{R}}\vec{b} = \vec{1}_L$, even if that could have saved some instructions.
 
@@
 
-%For subarray averaging, the first row and column of $\mat{\hat{R}}$ has to be fully calculated, and for time averaging $\mat{\hat{R}}[0]$ has to be fully calculated.
-
-%Since $L$ usually is larger than $N$, we have
-
-%Building $\hat{\mat{R}}$ using a sliding window approach across $K$ and $N_{avg}$.
-We have a limited amount of fast, near-core memory on the GPU.
-On NVIDIA architecture this memory is known as shared memory, and is restricted to 48KB per compute block.
-The maximum block size of 32x32, further restrict how large arrays we can handle inside a single compute block.
-It is therefore natural to restrict $L$ to a maxima of $32$ elements.
-Since $L \le M/2$, $M \le 64$.
-We can not afford to hold the full $\hat{\mat{R}}$ ($M = L$) when $\hat{\mat{R}} > 32$, because of limited amount of memory.
-...
-Explain in detail how \texttt{buildR} is implemented in a kernel.
-...
-Give more details on how much shared memory we have, and how it can be divided between compute blocks occupying one stream multiprocessor (SM).
-
-%Discuss the bottleneck adaptive diagonal loading causes. Possible solutions includes, use trace from previous image, use pre-calculated array power, constant weight.
-
-%Time averaging takes a lot of resources.
-It has been shown that time averaging can maintain both resolution and delay-and-sum-like speckle.
-The same speckle statistics can be obtain with small subarrays, but then we loose resolution.
-Small sub-arrays benefits from being less computational demanding.
-
-%\subsubsection{CUDA Compute Model}
-%The CUDA compute model is based on execution of a kernel across a grid of compute threads.
-Each position in the grid holds a block of threads.
-Each block is further divided into groups of 32 threads, known as a warp.
-The maximum number of threads inside a block is restricted to 512, hence 16 warps.
-From a CUDA perspective the GPU consist of one or more stream multi processors (SM), %where each SM is capable of 32 or more, depending on the architecture, concurrent multiplications and/or additions.
-The CUDA 2.x compute capability (CC) has available 48 KB of near-core shared memory per SM that can be utilized by resident blocks.
-The maximum number of resident blocks per SM is 8, but in addition the maximum number of resident warps per SM is restricted to 48.
-Hence, %with a block size of 512 only $48/(512/32) = 3$ blocks can occupy the SM at once.
-In addition, if one thread where to use 64B of shared memory the number of allowed resident blocks with a block size of 512 would be reduced to 1.
-Up to a curtain point
-%it is beneficial to have as many warps as possible occupying the SM at once to hide latency from accessing memory and executing time consuming mathematical functions.
-The SM has for CC 2.x 32K 32-bit long registers.
-When implementing algorithms on the GPU it is important to keep data as close to the core as possible.
-Utilizing available registers and shared memory to its maximum is %therefore important.
-
 \subsection{Solving Multiple Small Linear Systems}
-In the literature, when it comes to solving systems of linear equations on the GPU the focus has mostly been on large systems, and GPU libraries therefore often lack a comprehensive collection of batched solvers.
+In the literature, when it comes to solving systems of linear equations on the GPU the focus has mostly been on large systems.
+GPU libraries therefore often lack a comprehensive collection of batched solvers.
 The reason for this is two-fold, first the GPU needs to solve thousands of small systems in order to beat the low memory latency of the CPU, second there is a range of system dimensions that does not fit well with the GPU architecture, where only a small amount of near-core memory and registers are available.
@@
 However, there is two arguments for doing solving on the GPU.
-First the CPU, which already runs a lot of algorithms, is offloaded.
+First the CPU, which might already runs a lot of algorithms, is offloaded.
 Second the transfer of several thousand covariance matrices from CPU-side to GPU-side is spared.
 
-For the results presented in this paper, we have used an unreleased GPU implementation of Gauss-Jordan (GJ) elimination (by Nvidia) to solve $\mat{\hat{R}}\vec{b} = \vec{1}_L$, where $\vec{b} = \mat{\hat{R}}^{-1}\vec{1}_L$.
-For smaller systems, $L < 10$, a direct solver using one thread per system is preferable.
-For this purpose we have implemented Cramer's rule for $2 \times 2$, $3 \times 3$ and $4 \times 4$ matrices.
-All solvers have been benchmarked against the latest Intel MKL libraries.
-
-%However we will go through an implementation of an $\mat{U}^H\mat{D}\mat{U}$-decomposition-based solver to point out challenges when implementing batched solving of small linear systems on the GPU.
-...
-(Unsure if this should be included).
-
-%Can include details on GPU implementation of $\mat{U}^H\mat{D}\mat{U}$, but this has not proved to be faster than NVIDIA's GJ implementation.
-However, the complexity for solving with $\mat{U}^H\mat{D}\mat{U}$ decomposition is supposed to be $1/2$ of GJ.
-
-%Skriv om utfordringene om å lage en solver på GPU. Ikke dra inn UHDU.
-
-%In the final journal article we could discuss Conjugated gradient and Woodbury and the benefits they bring.
-%Other solvers that provides different benefits includes Conjugated gradient and the use of the Woodbury matrix identity.
+For the results presented in this paper, we have used an unreleased GPU implementation of Gauss-Jordan (GJ) elimination (by Nvidia) to solve $\mat{\hat{R}}\vec{b} = \vec{1}_L$.
 
@@
 
-The final output, as shown in (\ref{eq:z_mv}) and (\ref{eq:z_mv2}), can be formed by either reducing the data down to length $L$, or increase the weight vector to length $M$.
+The final output can be formed by either reducing the data down to length $L$ (as shown in (\ref{eq:z_mv})), or by increasing the weight vector to length $M$.
 Since we have decided to have $L$ threads per output, the data is reduced to length $L$ by letting thread $r$ calculate
@@
 
-\section{Beamspace Processing}\label{sec:bs}
-Nilsen et al.
-\cite{Nilsen2009} have proposed to apply beamspace processing to further reduce the required computations.
-With beamspace we mean transforming the channel data from element space to a fan of beams covering the sector illuminated by transmit beams.
-The transformation is written as
-\begin{align}
-\vec{x}_{BS} = \mat{B}\vec{x}^{\Delta}.
-\end{align} 
-The steering vectors in $\mat{B}$, the so-called Butler matrix, determine each beam's direction in space.
-We can therefore easily remove beams where no interference is present by removing rows in $\mat{B}$.
-For a focused system like a cardiac ultrasound scanner, where the received signal is concentrated in a narrow band around broadside, there is a small percentage of the beams that contains almost all the energy.
-Nilsen et al.
-concluded that as little as three beams around broadside were adequate to produce results comparable to applying Capon beamforming in element space.
-
-The beamspace version of Capon is derived by replacing the covariance matrix in (\ref{eq:w}) with the beamspace covariance matrix $\mat{\hat{R}}_{BS} = f(\vec{x}_{BS})$, where $\mat{B} \in \mathbb{C}^{N_b,M}$, and $N_b$ is the number of selected beams.
-Note that when subarray averaging is applied, the transformation must be done per subarray, yielding subarrays that no longer overlap in space.
-Hence, the amount of data is change from $M$ to $N_LN_b$ per sample.
-
-(Go in detail about the implementation using sliding DFT).
-
-(Go in detail about what kind of modifications we need to do to each of the three steps)
-
-%\subsection{Iterative Capon}
-%The idea behind iterative Capon is to perform Capon using small subarrays in several iterations until there is only two subarrays to sum.
-In this way we fix the size of L, hence the complexity of solving $\mat{R}\vec{b} = \vec{1}$.
-However, for each iteration we re-obtain L-1 degrees of freedom until all are retrieved after $K = M-L+1$ iterations.
-Since subarray averaging has been applied to decorrelated the signal, we might loose this effect when reestablishing all degrees of freedom.
-The number of iterations should therefore be restricted to $K/2$ to equal the normal choice of $L = M/2$ when it comes to degrees of freedom.
-
-\section{Results}\label{sec:res}
-Describe the computer system (CPU, GPU...). Describe the data that are processed (field II, Channel data from GE Vingmed).
-% stolpediagram, med de ulike metodenes kjøretid oppehverandre.
-
 \begin{figure*}[!t]
-\centerline{\subfloat[]{\includegraphics[width=3.5in]{gfx/benchmark_us_32_lr.png} \label{fig:benchUS32}}
+\centerline{\subfloat[]{\includegraphics[width=3.9in]{gfx/benchmark_us_32_lr.png} \label{fig:benchUS32}}
 \hfil
-\subfloat[]{\includegraphics[width=3.5in]{gfx/benchmark_us_64_lr.png}%
+\subfloat[]{\includegraphics[width=3.9in]{gfx/benchmark_us_64_lr.png}%
 \label{fig:benchUS64}}}
-\caption{(Change into bar plot?) Benchmark of GPU Capon beamforming for a cardiac image covering a 70 degrees sector.
+\caption{Benchmark of GPU Capon beamforming for a cardiac image covering a 70 degrees sector.
 Execution time is plotted as a function of subarray length $L$.
@@
 Calculating covariance matrices in red, solver in green, and the final beamformer sum in blue.
-For each step, there is one line for three different choices of temporal averaging, $N_{avg} \in [0, 1, 2]$.
+For each step, there is one line for three different choices of temporal averaging, $K \in [0, 1, 2]$.
 (a) A 32 element array.
@@
 
-\begin{figure*}[!t]
-\centerline{\subfloat[]{\includegraphics[width=3.5in]{gfx/benchmark_us_32_lr.png} \label{fig:benchBS3}}
-\hfil
-\subfloat[]{\includegraphics[width=3.5in]{gfx/benchmark_us_64_lr.png}%
-\label{fig:benchBS5}}}
-\caption{(TODO: Add the right figure.
-Change it into a bar plot?) Benchmark of GPU Beamspace Capon beamforming for a cardiac image covering a 70 degrees sector.
-Execution time is plotted as a function of number of elements $M$.
-The plot shows execution times for the three steps listed in Section \ref{sec:meth}.
-Calculating covariance matrices in red, solver in green, and the final beamformer sum in blue.
-For each step, there is one line for three different choices of temporal averaging, $N_{avg} \in [0, 1, 2]$.
-(a) 3 beams.
-(b) 5 beams.}
-\label{fig:benchBS}
-\end{figure*}
+\section{Benchmarks and Videos}
+Figure \ref{fig:bench} shows the results from benchmarking the described implementation on a Nvidia Quadro 6000 graphics card (14 SMs, and 1 TFlops single precision, 6 GB global memory, and 144 GB/s memory bandwidth).
+The problem sizes are calculated based on a 32 and 64 elements array with center frequency 2.5 MHz.
+By assuming a system bandwidth of 80\% we need a sampling frequency of 2 MHz.
+If we image down to a depth of 15 cm we get approximately 400 samples per range line.
+To correctly sample a 70 degrees sector with a 64 and 32 elements array ($\lambda/2$ pitch) we need at least 80 and 40 beams respectively.
+The presented execution times does not include transferring of data from CPU-side to the GPU.
+What is included is all calculations, memory transfers from global to near-core GPU memory, and near core memory accessing.
 
-\begin{figure*}[!t]
-\centerline{\subfloat[]{\includegraphics[width=2in]{gfx/mv_phantom.png} \label{fig:phantomDAS}}
-\hfil
-\subfloat[]{\includegraphics[width=2in]{gfx/mv_phantom.png} \label{fig:phantomMV}}
-\hfil
-\subfloat[]{\includegraphics[width=2in]{gfx/mv_phantom.png} \label{fig:phantomBS}}}
-\caption{(TODO: Add the right figures) Have graph showing cyst contrast as a function of different parameters instead?
-Cyst and point targets phantom a) Delay-and-sum, uniform weights.
-b) Capon ($M=?, L=?, N=?$) c) Beamspace Capon ($M=?, L=?, N=?, N_b=?$).}
-\label{fig:bench}
-\end{figure*}
+In the attached videos you will find a cardiac loop (add ref to video), and a simulated phantom (add ref to videos) processed with different types of settings given to the Capon beamformer.
 
-\begin{figure*}[!t]
-\centerline{\subfloat[]{\includegraphics[width=2in]{gfx/us_image.png} \label{fig:phantomDAS}}
-\hfil
-\subfloat[]{\includegraphics[width=2in]{gfx/us_image.png} \label{fig:phantomMV}}
-\hfil
-\subfloat[]{\includegraphics[width=2in]{gfx/us_image.png} \label{fig:phantomBS}}}
-\caption{(TODO: Add the right figures) Have graph showing cavity-wall contrast as a function of parameters instead?
-In-vivo cardiac ultrasound image a) Delay-and-sum, uniform weights.
-b) Capon ($M=?, L=?, N=?$) c) Beamspace Capon ($M=?, L=?, N=?, N_b=?$).}
-\label{fig:bench}
-\end{figure*}
-
-Present images before and after Capon weights has been applied.
-Comment on the effect of selecting different parameters.
-Present graph showing executing times for Cython-Capon v.s.
-CUDA-Capon for different choices of parameters.
-Comment on different choices of parameters change the running times.
-Temporal averaging for instance, lowers the efficiency since $2Y_{avg}$ more operations has to be performed per $\mat{\hat{R}}$
-
-Present images, both simulated and in-vivo, processed using beamspace Capon. Comment on the differences between the two and on how much they deviates from the full method.
-%Show beampatterns for the two compared with the full method. 
-Show executing times for the two methods compared with the full method on the GPU.
-
-Describe the base band (IQ-demodulation) process that the data has been undergoing.
-
-How has the data been acquired?
-
-Ta med en serie med bilde. Må vise tydelig at det gir forbedring.
-
-\section{Discussion}\label{sec:dis}
-Discuss results presented in Section \ref{sec:res}.
-
-We expect that a large speedup had the Capon beamformer been implemented on the CPU using threads and SIMD (single instruction multiple data) instructions.
-
-Solving linear systems on the GPU is not believed to be much faster than a CPU implementation, but there is two key points.
-First the CPU is offloaded, and second the time it takes to transfer all covariance matrices back to the CPU makes the choice of doing the solving on the GPU simple to make.
-
-(results)?
-For a typical cardiac ultrasound image of 80*400 pixels (70 degree sector, 15cm range) acquired using a 2.5MHz, $M=64$ element phased array, the result is 10fps (subarray length $L=M/2$, temporal smoothing over 3 samples).
-If we do a 2-element pre-beamforming to get the channel count down to 32, the frame rate increases to 44fps.
-For a 32 element phased array we need less beams to cover the sector ($40 \times 400$ pixels), hence with the same parameters the frame rate increases to 87fps.
-The bottleneck of the Capon algorithm is to find the inverse covariance matrix for all pixels.
-This is an $O(L^3)$ operation, and results show that when L is larger than 16, real time performance is not possible for cardiac imaging with current high-end hardware.
-The target GPU has been Nvidia’s Quadro 6000, capable of 1Tflops.
-
-Toeplitz-based solver (cite new article). Discuss it.
-
-\section{Conclusion}\label{sec:con}
-...
-
+\section{Discussion}
 % An example of a floating figure using the graphicx package.
@@
 % anywhere on the first page for that matter. Also, in-text middle ("here")
-% positioning is not used. Most IEEE journals use top floats exclusively.
-% Note that, LaTeX2e, unlike IEEE journals, places footnotes above bottom
-% floats. This can be corrected via the \fnbelowfloat command of the
-% stfloats package.
+% positioning is not used. Most IEEE journals/conferences use top floats
+% exclusively. Note that, LaTeX2e, unlike IEEE journals/conferences, places
+% footnotes above bottom floats. This can be corrected via the \fnbelowfloat
+% command of the stfloats package.
 
@@
 
-% if have a single appendix:
-%\appendix[Proof of the Zonklar Equations]
-% or
-%\appendix  % for no appendix heading
-% do not use \section anymore after \appendix, only \section*
-% is possibly needed
-
-% use appendices with more than one appendix
-% then use \section to start each appendix
-% you must declare a \section before using any
-% \subsection or using \label (\appendices by itself
-% starts a section numbered zero.)
-%
-
-%\appendices
-%\section{Proof of the First Zonklar Equation}
-%Appendix one text goes here.
-
-% you can choose not to have a title for an appendix
-% if you want by leaving the argument blank
-%\section{}
-%Appendix two text goes here.
+% conference papers do not normally have an appendix
 
@@
 \section*{Acknowledgment}
-
-The authors would like to thank...
-
-% Can use something like this to put references on a page
-% by themselves when using endfloat and the captionsoff option.
-\ifCLASSOPTIONcaptionsoff
-  \newpage
-\fi
+The authors would like to thank Tore Bj\aa{}stad at GE Vingmed Ultrasound for providing cardiac data, and Nvidia for the batched solver written in CUDA.
 
@@
 % argument is your BibTeX string definitions and bibliography database(s)
-%\bibliography{IEEEabrv,../bib/paper}
-\bibliography{mybib} % link to mendeley organized bibtex-file
-%\bibliography{IEEEabrv,mybib} 
+\bibliography{mybib}
+%
 % <OR> manually copy in the resultant .bbl file
@@
 
-% biography section
-% 
-% If you have an EPS/PDF photo (graphicx package needed) extra braces are
-% needed around the contents of the optional argument to biography to prevent
-% the LaTeX parser from getting confused when it sees the complicated
-% \includegraphics command within an optional argument. (You could create
-% your own custom macro containing the \includegraphics command to make things
-% simpler here.)
-%\begin{biography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{mshell}}]{Michael Shell}
-% or if you just want to reserve a space for a photo:
-
-% insert where needed to balance the two columns on the last page with
-% biographies
-%\newpage
-
-\begin{IEEEbiography}{Jon Petter \AA{}sen}
-Biography text here.
-\end{IEEEbiography}
-
-\begin{IEEEbiography}{Jo Inge Buskenes}
-Biography text here.
-\end{IEEEbiography}
-
-\begin{IEEEbiography}{Carl-Inge Colombo Nilsen}
-Biography text here.
-\end{IEEEbiography}
-
-% insert where needed to balance the two columns on the last page with
-% biographies
-%\newpage
-
-\begin{IEEEbiography}{Andreas Austeng}
-Biography text here.
-\end{IEEEbiography}
-
-\begin{IEEEbiography}{Sverre Holm}
-Biography text here.
-\end{IEEEbiography}
-
-% You can push biographies down or up by placing
-% a \vfill before or after them. The appropriate
-% use of \vfill depends on what kind of text is
-% on the last page and whether or not the columns
-% are being equalized.
-
-%\vfill
-
-% Can be used to pull up biographies so that the bottom of the last one
-% is flush with the other column.
-%\enlargethispage{-5in}
-
 % that's all folks

### 2012-09-11 HRUIonGPU/jrnl_article.tex
started on IUS article

@@
 In this paper we introduce the first implementation of the Capon beamformer exhibiting real-time performance when applied in a typical cardiac ultrasound imaging setting.
-To accomplish this task we make use of the massive, and easily accessed, parallel processing power found in modern Graphics Processing Units (GPU's), combined with novel methods to reduce the computational complexity as the problem size grows.
+To accomplish this task we make use of the massive, and easily accessed, parallel processing power found in modern Graphics Processing Units (GPUs), combined with novel methods to reduce the computational complexity as the problem size grows.
 
@@
 However, there is two arguments for doing solving on the GPU.
-First the CPU, which already runs a lot of algorithms, is offloaded.
+First the CPU, which might already runs a lot of algorithms, is offloaded.
 Second the transfer of several thousand covariance matrices from CPU-side to GPU-side is spared.

### 2012-09-20 HRUIonGPU/gfx/all_plots.py
added plot function for plotting ius2012 figures

@@ -1,0 +1,138 @@
+'''
+Created on Mar 22, 2011
+
+@author: me
+'''
+
+#import framework.calc as calc
+#from framework.calc import db
+#from framework.mynumpy import size, arange, log10, pi, linspace, mean, cos, sin, arcsin, unique, zeros
+from numpy import size, arange, log10, pi, linspace, mean, cos, sin, arcsin, unique, zeros
+#from framework.mynumpy import mode_
+#import framework.mynumpy as np
+import numpy as np
+from scipy import io
+#from framework.
+from pylab import legend,grid, plot, imshow, ion, ioff, figure, savefig, cm, xlim, ylim, clim, xticks, subplot, subplots_adjust, get_current_fig_manager, setp, colorbar, xlabel, ylabel,show, gca
+#import mypylab as pl
+#from mpl_toolkits.axes_grid1 import ImageGrid
+#from framework.gfuncs import error
+
+GFX_PATH='/home/me/Work/Phd/Code/gfx'
+
+
+
+def plotBenchmarkGPU(filename, M, Nx, Ny):
+   
+   timings = np.loadtxt(filename)
+   
+   parameters = []
+   
+   min_L = 2
+   max_L = M/2
+
+   for L in range(min_L, max_L):
+      for Yavg in [0,1,2]:
+         parameters.append((M, L, Yavg, 0.2, Nx, Ny))
+
+   yavg1 = timings[0::3,:]
+   yavg2 = timings[1::3,:]
+   yavg3 = timings[2::3,:]
+   yavg1_sum = yavg1.sum(1)
+   yavg2_sum = yavg2.sum(1)
+   yavg3_sum = yavg3.sum(1)
+   
+   print yavg1.shape
+   print yavg2.shape
+   print yavg3.shape
+   
+   fig = figure()
+   ax = fig.add_subplot(1,1,1)
+   cax = ax.plot(range(min_L,max_L+1), yavg1[:,0], 'r')
+   cax = ax.plot(range(min_L,max_L+1), yavg1[:,1], 'g')
+   cax = ax.plot(range(min_L,max_L+1), yavg1[:,2], 'b')
+   cax = ax.plot(range(min_L,max_L+1), yavg1_sum, 'k')
+   
+   cax = ax.plot([0,0.001],[0,0], 'k')
[…]

### 2012-09-20 HRUIonGPU/conf_article.tex

@@
 \boldmath
-Recent results \cite{Synnevag2007, Austeng2008, Vignon2008, Viola, Mehdizadeh2012} suggest that the Capon beamformer can provide increased lateral resolution to a medical ultrasound image.
-In this paper, the high computational complexity of the Capon beamformer is targeted with the use of Graphics Processing Units (GPUs), and we show in-vivio results with Capon beamforming applied on a full cardiac cycle in addition to simulations.
-With GPUs we are able to process a 70 degree sector cardiac image from a 64 element phased array at interactive frame rates using both spatial and temporal smoothing.
+Recent results \cite{Synnevag2007, Austeng2008, Vignon2008, Viola, Mehdizadeh2012} suggest that the Capon beamformer can provide increased lateral resolution when applied in medical ultrasound imaging.
+In this paper, the high computational complexity of the Capon beamformer is targeted with the use of Graphics Processing Units (GPUs).
+We also present in-vivio results with Capon beamforming applied on a full cardiac cycle in addition to simulations.
+With GPUs we are able to process a 70 degrees sector cardiac image from a 64 element phased array at interactive frame rates using both spatial and temporal smoothing.
 
@@
 For a 32 element phased array we need less beams to cover the sector ($40\times400$ pixels), hence with the same parameters the frame rate increases to 87fps.
-The bottleneck of the Capon algorithm is to find the inverse covariance matrix for all pixels.
+After optimization the bottleneck of the Capon algorithm is to find the inverse covariance matrix for all pixels.
 This is an $O(L^3)$ operation, and results show that when L is larger than 16 (M = 64), real time performance is not possible for cardiac imaging with current high-end hardware.
@@
 In this paper we introduce the first implementation of the Capon beamformer exhibiting real-time performance when applied in a typical cardiac ultrasound imaging setting.
-To accomplish this task we make use of the massive, and easily accessed, parallel processing power found in modern Graphics Processing Units (GPUs), combined with novel methods to reduce the computational complexity as the problem size grows.
+To accomplish this task we make use of the massive, and easily accessed, parallel processing power found in modern Graphics Processing Units (GPUs).
 
@@
 
-To calculate the Capon beamformer output we need three equations. First the sample covariance matrix has to be estimated,
+To calculate the Capon beamformer output we need three equations. First the sample covariance matrix has to be calculated,
 \begin{align}
@@
 Note that the dimension of the covariance matrix is $L \times L$.
-The covariance matrix is also loaded with a diagonal factor $\epsilon$ to ensure numerical stability,
+The covariance matrix is usually loaded with a diagonal factor $\epsilon$ to ensure numerical stability,
 \begin{align}\label{eq:diag}
@@
 \subsection{Calculating Multiple Sample Covariance Matrices}
-The straightforward sum in (\ref{eq:R}) has complexity $O(L^2N_LN_K) = O(L^3)$ since $N_L$, the number of subarrays, is a function of $L$.
+The straightforward sum in (\ref{eq:R}) has complexity $O(L^2N_LN_K)$.
+Note that for a typical array configuration ($M \in [32, 128]$) we do more spacial smoothing than temporal smoothing.
+Hence, $N_K$ is usually small compared to the number of subarrays ($N_L$).
+If we let $L = M/\alpha$, $L^2N_L$ becomes $(\alpha - 1)L^3 + L^2$, $\alpha \in [1, M]$.
+A common choice for $\alpha$ in medical ultrasound imaging is between two and four.
+In this interval, calculating the sample covariance matrix is cubic with $L$.
 This is actually the same as for inverting an $L \times L$ matrix.
-However little attention has been devoted to this step in the literature before.
-Note that for a typical array configuration we do more spacial smoothing than temporal smoothing.
-Hence, $N_K$ is usually small compared to $N_L$.
-For short subarrays the construction of $\mat{\hat{R}}$ is dominated by the number of elements $M$ ($N_L = M - L + 1$), but since $L$  in most cases is selected to be between $M/4$ and $M/2$, many array elements implies large subarrays.
+However little attention has been devoted to this step in the literature.
 
 From (\ref{eq:R}) it is obvious that calculation of one element in $\mat{\hat{R}}$ is independent of all the other elements.
-A natural level of granularity would then be to assign one thread to each element of $\mat{\hat{R}}$.
+A natural level of granularity would then be to assign one thread to each element of $\mat{\hat{R}}$ \cite{Chen2011}.
 This will minimize global reads and writes needed per thread, but unfortunately the maximum number of threads per block puts an upper limit on the subarray length, $L_{max}=32$, since we want to avoid block-to-block synchronization.
 
-The quadratic complexity with $L$ and linear complexity with $N_K$ can be reduced by realizing that the calculation in (\ref{eq:R}) overlap both across subarrays and in time.
+The cubic complexity with $L$ and linear complexity with $N_K$ can be reduced by realizing that the calculation in (\ref{eq:R}) overlap both across subarrays and in time.
 Element $(i,j)$ of $\mat{\hat{R}}[n]$ can be calculated from element $(i-1, j-1), \text{for } i,j \in [1, L-1]$ as
@@
 \end{align}
-The same is true in time where
-\begin{align}
-\mat{\breve{R}}[n] &= \mat{\breve{R}}[n-1] + \sum_{l=0}^{N_L-1} (\vec{x}^{\Delta}_l[n+K]\vec{x}^{\Delta}_l[n+K]^H \nonumber \\
-&- \vec{x}^{\Delta}[n-K-1]\vec{x}^{\Delta}[n-K-1]^H).
-\end{align}
+%The same is true in time where
+%\begin{align}
+%\mat{\breve{R}}[n] &= \mat{\breve{R}}[n-1] + \sum_{l=0}^{N_L-1} (\vec{x}^{\Delta}_l[n+K]\vec{x}^{\Delta}_l[n+K]^H \nonumber \\
+%&- \vec{x}^{\Delta}[n-K-1]\vec{x}^{\Delta}[n-K-1]^H).
+%\end{align}
 %- \mat{\breve{R}}[n-K-1] + \mat{\breve{R}}[n-K].
@@
 With our choice of calculating $\mat{\hat{R}}$, calculations are still independent across time, but not along diagonals in $\mat{\hat{R}}$.
-We therefore have selected to use one thread per diagonal in the upper triangle of $\mat{\hat{R}}$, and for small $L$s several covariance matrices are calculated per thread block (Figure \ref{fig:gpulayout}).
+We have therefore selected to use one thread per diagonal in the upper triangle of $\mat{\hat{R}}$.
+And for small subarrays, several covariance matrices are calculated per thread block (Figure \ref{fig:gpulayout}).
 Maximum supported $L$ is with this approach increased to 512, however for small $L$ high $M$ configurations, memory consumption per thread will be to high to achieve full occupancy of the GPU.
@@
 However, the way we have decided to organize the GPU compute threads per $\mat{\hat{R}}$, the lower triangle can be calculated with little overhead on the GPU.
-This also means that we do not relay on a upper triangular solver to solve $\mat{\hat{R}}\vec{b} = \vec{1}_L$, even if that could have saved some instructions.
+This also means that we do not rely on a upper triangular solver to solve $\mat{\hat{R}}\vec{b} = \vec{1}_L$, even if that could have saved some instructions.
 
@@
 \subsection{Solving Multiple Small Linear Systems}
-In the literature, when it comes to solving systems of linear equations on the GPU the focus has mostly been on large systems.
+In the literature, when it comes to solving systems of linear equations on the GPU, focus has mainly been on large systems.
 GPU libraries therefore often lack a comprehensive collection of batched solvers.
@@
 
-For the results presented in this paper, we have used an unreleased GPU implementation of Gauss-Jordan (GJ) elimination (by Nvidia) to solve $\mat{\hat{R}}\vec{b} = \vec{1}_L$.
+For the results presented in this paper, we have used an unreleased GPU implementation of batched Gauss-Jordan (GJ) elimination (by Nvidia) to solve $\mat{\hat{R}}\vec{b} = \vec{1}_L$.
 
 \subsection{Compute Beamformer Output}
-To calculate the beamformer output we have implement the formula in (\ref{eq:z_mv}) with one thread per element-weight ($L$ threads per weight vector).
+The formula in (\ref{eq:z_mv}) has been implemented with one thread per element-weight ($L$ threads per weight vector).
 After the solution vector $\vec{b}$ has been placed in shared memory, one of the $L$ threads finds the sum of $\vec{b}$.
-Another option is to let several threads cooperate to form the sum using a binary reduction sum.
-However, in the following benchmarks the first approach has been used.
 Each of the $L$ threads then read one element of $\vec{b}$, scale it with the inverse sum just found to get the weight vector $\vec{w}[n]$.
@@
 Each thread now computes $z_r = w_r^*x_r$, and finally one thread computes and outputs $z = \sum_{r=0}^{L-1} z_r$ to global memory.
-The last sum can also be performed using a binary reduction sum across several threads for increased speed.
 
@@
 \caption{Benchmark of GPU Capon beamforming for a cardiac image covering a 70 degrees sector.
-Execution time is plotted as a function of subarray length $L$.
+Execution time is plotted as a function of subarray length $L \in [1, M/2]$.
 The plot shows execution times for the three steps listed in Section \ref{sec:meth}.
@@
 In the attached videos you will find a cardiac loop (add ref to video), and a simulated phantom (add ref to videos) processed with different types of settings given to the Capon beamformer.
+Cardiac data was acquired using a 64 elements and 3.5 MHz phased array.
+Simulations was done using a 64 elements and 2.5 MHz phased array.
 
 \section{Discussion}
+From Figure \ref{fig:benchUS64} we see that for $L = M/2 = 32$ and $K=1$, total execution time is 100 ms (10 fps).
+Taking into account that we use complex data, but less samples per image, this is similar to the performance reported by Chen et al.
+\cite{Chen2011}.
+For the 32 element array in Figure (\ref{fig:benchUS32}), execution time is reduced to 11.5 ms (87 fps) using the same settings ($L=M/2, K = 1$).
+Doubling this number can be interpreted as pre-beamforming a 64 elements array down to 32.
+The frame rate for a 64 elements array is then increased from 10 to 44 fps.
+
+A major part of the total running time comes from solving systems of linear equations.
+Further research should therefore focus on making this contribution smaller.
+Either by implementing faster solvers, or reducing the sample covariance matrix size in a novel way (cite BS and other low complexity methods?).
+When doing such a reduction, including the pre-beamforming described above, one should investigate the loss of contrast and resolution compared with full Capon beamforming.
+
+As for now, we have shown that the Capon beamformer can be implemented in a GPU framework.
+And if we as in \cite{Chen2011} define real-time as 10 fps, all execution times  in Figure (\ref{fig:bench}) are real time.
+If we on the other hand define real-time as the time it takes to acquire the image.
+The image formed using a 32 elements array, has a real time requirement of 125 fps.
+For the 64 elements array real time is 62.5 fps.
+There is some distance to cover before that goal is reached.
+
+%Do we need more beams?
+
+% cite field II
+
 % An example of a floating figure using the graphicx package.
@@
 \section*{Acknowledgment}
-The authors would like to thank Tore Bj\aa{}stad at GE Vingmed Ultrasound for providing cardiac data, and Nvidia for the batched solver written in CUDA.
+The authors would like to thank Tore Bj\aa{}stad at GE Vingmed Ultrasound for providing cardiac data, and Nvidia for their batched solver written in CUDA.
 

### 2012-09-20 HRUIonGPU/conf_article.tex
Revision of ius2012 article after comments by Sverre

@@
 \renewcommand{\vec}[1]{\mathbf{#1}}
+\newcommand{\degree}{\ensuremath{^\circ}}
 
@@
 In this paper, the high computational complexity of the Capon beamformer is targeted with the use of Graphics Processing Units (GPUs).
-We also present in-vivio results with Capon beamforming applied on a full cardiac cycle in addition to simulations.
-With GPUs we are able to process a 70 degrees sector cardiac image from a 64 element phased array at interactive frame rates using both spatial and temporal smoothing.
+We also present in-vivo results with Capon beamforming applied on a full cardiac cycle in addition to simulations.
+With GPUs we are able to process a $70\degree$ sector cardiac image from a 64 element phased array at interactive frame rates using both spatial and temporal smoothing.
 
-For a typical cardiac ultrasound image of $80\times400$ pixels (70 degree sector, 15cm range) acquired using a 2.5MHz, M=64 element phased array, we obtain 10fps (subarray length L=M/2, temporal smoothing over 3 samples).
-If we do a 2-element pre-beamforming to get the channel count down to 32, the frame rate increases to 44fps.
-For a 32 element phased array we need less beams to cover the sector ($40\times400$ pixels), hence with the same parameters the frame rate increases to 87fps.
+For a typical cardiac ultrasound image of $80\times400$ pixels ($70\degree$ sector, 15cm range) acquired using a 2.5MHz, M=64 element phased array, we obtain 10 fps (subarray length L=M/2, temporal smoothing over 3 samples).
+If we do a 2-element pre-beamforming to get the channel count down to 32, the frame rate increases to 44 fps.
+For a 32 element phased array we need less beams to cover the sector ($40\times400$ pixels), hence with the same parameters the frame rate increases to 87 fps.
 After optimization the bottleneck of the Capon algorithm is to find the inverse covariance matrix for all pixels.
-This is an $O(L^3)$ operation, and results show that when L is larger than 16 (M = 64), real time performance is not possible for cardiac imaging with current high-end hardware.
-The target GPU has been Nvidia’s Quadro 6000, capable of 1Tflops.
+This is an $O(L^3)$ operation, and when L is larger than 16 (M = 64), real time performance is not possible for cardiac imaging with current high-end hardware.
+The target GPU has been Nvidia’s Quadro 6000, capable of 1 Tflops.
 
@@
 Capon adaptive beamforming \cite{Capon1969} applied to medical ultrasound imaging has recently shown promising results in several studies \cite{Synnevag2007, Austeng2008, Vignon2008, Viola, Mehdizadeh2012}.
-For instance it has been introduced some interesting trade-offs in addition to just improved resolution, like maintaining image resolution in combination with higher framerates, smaller probes, or deeper penetration \cite{Synnevag2009}.
+For instance, some interesting trade-offs in addition to just improved resolution were introduced, like maintaining image resolution in combination with higher framerates, smaller probes, or deeper penetration \cite{Synnevag2009}.
 Despite these benefits, widespread use has been limited by the immense computational complexity associated with the method \cite{So2011}.
@@
 
-Previous work on implementing a parallel Capon beamformer have mainly been focused around radar and passive sonar systems using custom made systolic array processors \cite{McWhirter1989, Sinha2002}.
+Previous work on implementing a parallel Capon beamformer has mainly been focused around radar and passive sonar systems using custom made systolic array processors \cite{McWhirter1989, Sinha2002}.
 For medical ultrasound, Chen et al.
 \cite{Chen2011a, Chen2011} have implemented a Capon beamformer on the GPU that produced real apodization weights based on real data.
-Under this constraint the Capon beamformer is not allowed to shift the main lobe, and therefore its full potential is not reached.
-Shifting the main lobe is an important property of the Capon beamformer in order for it to improve edge definitions and contrast throughout an ultrasound image.
+Under this constraint the Capon beamformer is not allowed to micro steer the main lobe, and therefore its full potential is not reached.
+Micro steering of the main lobe is an important property of the Capon beamformer in order for it to improve edge definitions and contrast throughout an ultrasound image.
 Another aspect of using complex data is that even though the number of aritmetic operations per sample is increased we can now sample at the system bandwidth, keeping the number of samples for a full cardiac imaging sector at a minimum.
-All results in this paper is reported with the use of baseband data.
+All results in this paper are reported with the use of baseband data.
 
@@
 The conventional form of array beamforming is delay-and-sum (DAS).
-Here, the signal $x_m[n]$ recorded at element number $m \in [0,M-1]$ from time $n = 0$ to $N-1$ is delayed to broadside with an appropriate delay $\Delta_m[n]$.
+Here, the signal $x_m[n]$ recorded at element number $m \in [0,M-1]$ at time $n$ is delayed to broadside with an appropriate delay $\Delta_m[n]$.
 The beamformer output, $z[n]$, is then the sum of all $M$ channels weighted with a per-element factor $w_m$,
@@
 CUDA by Nvidia has been our choice of GPU developer environment \cite{Nvidia2011}.
-The CUDA compute model is based on execution of a kernel function across a grid of compute threads, where the kernel function describe the work to be done by each thread.
+The CUDA compute model is based on execution of a kernel function across a grid of compute threads, where the kernel function describes the work to be done by each thread.
 As depicted in Figure \ref{fig:gpulayout}, each position in the grid holds a block of threads, with maximal size of 1024 threads, that can be synchronized with little cost if needed.
 %Each block is further divided into groups of 32 threads, known as a warp. The maximum number of threads inside a block is restricted to 512, hence 16 warps. 
-From a CUDA perspective the GPU consist of one or more stream multi processors (SMs), where each SM is capable of 32 or more, depending on the architecture, concurrent multiplications and/or additions.
+From a CUDA perspective the GPU consists of one or more stream multi processors (SMs), where each SM is capable of 32 or more, depending on the architecture, concurrent multiplications and/or additions.
 A group of blocks is scheduled to each SM until all blocks have been processed.
@@
 All of these constraints combined pose a real challenge as the target problem has to map on to the parallel framework both on a grid, block and memory consumption level in order to benefits from parallel acceleration.
-In the next section, we will describe how this mapping should be done for Capon beamforming of cardiac ultrasound data.
+In the next section, we will describe how this mapping can be done for Capon beamforming of cardiac ultrasound data.
 
@@
 The straightforward sum in (\ref{eq:R}) has complexity $O(L^2N_LN_K)$.
-Note that for a typical array configuration ($M \in [32, 128]$) we do more spacial smoothing than temporal smoothing.
+Note that for a typical array configuration ($M \in [32, 128]$) we do more spatial smoothing than temporal smoothing.
 Hence, $N_K$ is usually small compared to the number of subarrays ($N_L$).
 If we let $L = M/\alpha$, $L^2N_L$ becomes $(\alpha - 1)L^3 + L^2$, $\alpha \in [1, M]$.
-A common choice for $\alpha$ in medical ultrasound imaging is between two and four.
+A common choice for $\alpha$ is between two and four (cite!!!).
 In this interval, calculating the sample covariance matrix is cubic with $L$.
@@
 
-The cubic complexity with $L$ and linear complexity with $N_K$ can be reduced by realizing that the calculation in (\ref{eq:R}) overlap both across subarrays and in time.
+The cubic complexity with $L$ and linear complexity with $N_K$ can be reduced by realizing that the calculation in (\ref{eq:R}) overlaps both across subarrays and in time.
 Element $(i,j)$ of $\mat{\hat{R}}[n]$ can be calculated from element $(i-1, j-1), \text{for } i,j \in [1, L-1]$ as
@@
 %- \mat{\breve{R}}[n-K-1] + \mat{\breve{R}}[n-K].
-Both sliding across subarrays and in time can be realized simultaneously if we hold a $N_K$-average in time of the outer products $\vec{x}^{\Delta}[n]\vec{x}^{\Delta}[n]^H$ in memory and calculate $\mat{\hat{R}}$ averaged over subarrays from this buffer.
+Both sliding across subarrays and in time can be realized simultaneously if we hold an $N_K$-average in time of the outer products $\vec{x}^{\Delta}[n]\vec{x}^{\Delta}[n]^H$ in memory and calculate $\mat{\hat{R}}$ averaged over subarrays from this buffer.
 However, since $N_K$ typically is small, and we have a limited amount of memory per thread on the GPU, we have chosen to focus on sliding across subarrays only (\ref{eq:sliding}).
@@
 Important to notice is that sliding will break independence in the sliding dimension.
-With our choice of calculating $\mat{\hat{R}}$, calculations are still independent across time, but not along diagonals in $\mat{\hat{R}}$.
+With our way of calculating $\mat{\hat{R}}$, calculations are still independent across time, but not along diagonals in $\mat{\hat{R}}$.
 We have therefore selected to use one thread per diagonal in the upper triangle of $\mat{\hat{R}}$.
 And for small subarrays, several covariance matrices are calculated per thread block (Figure \ref{fig:gpulayout}).
-Maximum supported $L$ is with this approach increased to 512, however for small $L$ high $M$ configurations, memory consumption per thread will be to high to achieve full occupancy of the GPU.
+Maximum supported $L$ is with this approach increased to 512, however for small $L$ high $M$ configurations, memory consumption per thread will be too high to achieve full occupancy of the GPU.
 
@@
 However, the way we have decided to organize the GPU compute threads per $\mat{\hat{R}}$, the lower triangle can be calculated with little overhead on the GPU.
-This also means that we do not rely on a upper triangular solver to solve $\mat{\hat{R}}\vec{b} = \vec{1}_L$, even if that could have saved some instructions.
+This also means that we do not rely on an upper triangular solver to solve $\mat{\hat{R}}\vec{b} = \vec{1}_L$, even if that could have saved some instructions.
 
@@
 Since one row for all covariance matrices is written to global memory per kernel iteration, the diagonal elements need to be held in a local memory until the trace is accumulated.
-When it is found, scattered writes of the diagonal elements to global memory has to be performed.
+When it is found, scattered writes of the diagonal elements to global memory have to be performed.
 
@@
 The reason for this is two-fold, first the GPU needs to solve thousands of small systems in order to beat the low memory latency of the CPU, second there is a range of system dimensions that does not fit well with the GPU architecture, where only a small amount of near-core memory and registers are available.
-Because of this, it has been proven hard to optimize a manageable set of solvers that provides speedups compared with the CPU in every cases.
+Because of this, it has proven hard to optimize a manageable set of solvers that provides speedups compared with the CPU in every case.
 It is therefore not given that solving on the GPU is faster than solving on the CPU in general.
-However, there is two arguments for doing solving on the GPU.
+However, there are two arguments for doing solving on the GPU.
 First the CPU, which might already runs a lot of algorithms, is offloaded.
@@
 After the solution vector $\vec{b}$ has been placed in shared memory, one of the $L$ threads finds the sum of $\vec{b}$.
-Each of the $L$ threads then read one element of $\vec{b}$, scale it with the inverse sum just found to get the weight vector $\vec{w}[n]$.
+Each of the $L$ threads then reads one element of $\vec{b}$ and scales it with the inverse sum just found to get the weight vector $\vec{w}[n]$.
 
@@
 \label{fig:benchUS64}}}
-\caption{Benchmark of GPU Capon beamforming for a cardiac image covering a 70 degrees sector.
+\caption{Benchmark of GPU Capon beamforming for a cardiac image covering a $70\degree$ sector.
 Execution time is plotted as a function of subarray length $L \in [1, M/2]$.
-The plot shows execution times for the three steps listed in Section \ref{sec:meth}.
-Calculating covariance matrices in red, solver in green, and the final beamformer sum in blue.
-For each step, there is one line for three different choices of temporal averaging, $K \in [0, 1, 2]$.
+The plot shows execution times for the three steps listed in Section \ref{sec:meth}: Calculation of covariance matrices in red, solver in green, and the final beamformer sum in blue.
+For each step, there is one line per choice of temporal averaging, $K \in [0, 1, 2]$.
 (a) A 32 element array.
@@
 \section{Benchmarks and Videos}
-Figure \ref{fig:bench} shows the results from benchmarking the described implementation on a Nvidia Quadro 6000 graphics card (14 SMs, and 1 TFlops single precision, 6 GB global memory, and 144 GB/s memory bandwidth).
-The problem sizes are calculated based on a 32 and 64 elements array with center frequency 2.5 MHz.
-By assuming a system bandwidth of 80\% we need a sampling frequency of 2 MHz.
+Figure \ref{fig:bench} shows the results from benchmarking the implementation described in Section \ref{sec:meth} on a Nvidia Quadro 6000 graphics card (14 SMs, 1 TFlops single precision, 6 GB global memory, and 144 GB/s memory bandwidth).
+The problem sizes are calculated based on 32 and 64 element arrays with center frequency 2.5 MHz.
+By assuming a system bandwidth of 80\% we need a complex sampling frequency of 2 MHz.
 If we image down to a depth of 15 cm we get approximately 400 samples per range line.
-To correctly sample a 70 degrees sector with a 64 and 32 elements array ($\lambda/2$ pitch) we need at least 80 and 40 beams respectively.
-The presented execution times does not include transferring of data from CPU-side to the GPU.
-What is included is all calculations, memory transfers from global to near-core GPU memory, and near core memory accessing.
+To correctly sample a $70\degree$ sector with 64 and 32 element arrays ($\lambda/2$ pitch) we need at least 80 and 40 beams respectively.
+The presented execution times do not include transfer of data from CPU-side to the GPU.
+What is included are all calculations, memory transfers from global to near-core GPU memory, and near core memory access.
 
-In the attached videos you will find a cardiac loop (add ref to video), and a simulated phantom (add ref to videos) processed with different types of settings given to the Capon beamformer.
-Cardiac data was acquired using a 64 elements and 3.5 MHz phased array.
-Simulations was done using a 64 elements and 2.5 MHz phased array.
+In the attached videos one will find a cardiac loop (add ref to video), and a simulated phantom (add ref to videos) processed with different types of settings given to the Capon beamformer.
+Cardiac data was acquired using a 64 element and 3.5 MHz phased array.
+Simulations were done using a 64 element 2.5 MHz phased array.
 
@@
 For the 32 element array in Figure (\ref{fig:benchUS32}), execution time is reduced to 11.5 ms (87 fps) using the same settings ($L=M/2, K = 1$).
-Doubling this number can be interpreted as pre-beamforming a 64 elements array down to 32.
-The frame rate for a 64 elements array is then increased from 10 to 44 fps.
+Doubling this number can be interpreted as pre-beamforming a 64 element array down to 32.
+The frame rate for a 64 element array is then increased from 10 to 44 fps.
 
-A major part of the total running time comes from solving systems of linear equations.
+A major part of the total execution time comes from solving systems of linear equations.
 Further research should therefore focus on making this contribution smaller.
-Either by implementing faster solvers, or reducing the sample covariance matrix size in a novel way (cite BS and other low complexity methods?).
+This can be done either by implementing faster solvers, or reducing the sample covariance matrix size in a novel way (ref?).
 When doing such a reduction, including the pre-beamforming described above, one should investigate the loss of contrast and resolution compared with full Capon beamforming.
@@
 And if we as in \cite{Chen2011} define real-time as 10 fps, all execution times  in Figure (\ref{fig:bench}) are real time.
-If we on the other hand define real-time as the time it takes to acquire the image.
-The image formed using a 32 elements array, has a real time requirement of 125 fps.
-For the 64 elements array real time is 62.5 fps.
-There is some distance to cover before that goal is reached.
+If we on the other hand define real-time as the time it takes to acquire a given image.
+The image formed using a 32 and 64 element array has a real time requirement of 125 fps and 62.5 fps respectively.
+There is some distance to cover before that goal is reached for all values of $L \le M/2$.
+However we see, by the last definition, that real time is possible for $M=64, L < 16$, and for $M = 32, L < 13$.
+
+(Should comment on videos)
 
@@
 % argument is your BibTeX string definitions and bibliography database(s)
-\bibliography{mybib}
+\bibliography{mybib_conf}
 %

### 2012-09-26 FFI-Simulator/ffi-documentation.tex

% !TEX TS-program = pdflatex
% !TEX encoding = UTF-8 Unicode

% This is a simple template for a LaTeX document using the "article" class.
% See "book", "report", "letter" for other types of document.

\documentclass[11pt]{article} % use larger type; default would be 10pt

\usepackage[utf8]{inputenc} % set input encoding (not needed with XeLaTeX)

%%% Examples of Article customizations
% These packages are optional, depending whether you want the features they provide.
% See the LaTeX Companion or other references for full information.

%%% PAGE DIMENSIONS
\usepackage{geometry} % to change the page dimensions
\geometry{a4paper} % or letterpaper (US) or a5paper or....
% \geometry{margin=2in} % for example, change the margins to 2 inches all round
% \geometry{landscape} % set up the page for landscape
%   read geometry.pdf for detailed page layout information

\usepackage{graphicx} % support the \includegraphics command and options

% \usepackage[parfill]{parskip} % Activate to begin paragraphs with an empty line rather than an indent

%%% PACKAGES
\usepackage{booktabs} % for much better looking tables
\usepackage{array} % for better arrays (eg matrices) in maths
\usepackage{paralist} % very flexible & customisable lists (eg. enumerate/itemize, etc.)
\usepackage{verbatim} % adds environment for commenting out blocks of text & for better verbatim
\usepackage{subfig} % make it possible to include more than one captioned figure/table in a single float
% These packages are all incorporated in the memoir class to one degree or another...

%%% HEADERS & FOOTERS
\usepackage{fancyhdr} % This should be set AFTER setting up the page geometry
\pagestyle{fancy} % options: empty , plain , fancy
\renewcommand{\headrulewidth}{0pt} % customise the layout...
\lhead{}\chead{}\rhead{}
\lfoot{}\cfoot{\thepage}\rfoot{}

%%% SECTION TITLE APPEARANCE
\usepackage{sectsty}
\allsectionsfont{\sffamily\mdseries\upshape} % (See the fntguide.pdf for font help)
% (This matches ConTeXt defaults)

%%% ToC (table of contents) APPEARANCE
\usepackage[nottoc,notlof,notlot]{tocbibind} % Put the bibliography in the ToC
\usepackage[titles,subfigure]{tocloft} % Alter the style of the Table of Contents
\renewcommand{\cftsecfont}{\rmfamily\mdseries\upshape}
\renewcommand{\cftsecpagefont}{\rmfamily\mdseries\upshape} % No bold!

%%% END Article customizations

%%% The "real" document content comes below...

\title{Documentation FFI Sonar Simulator}
\author{Jo Inge Buskenes and Jon Petter \AA{}sen}
%\date{} % Activate to display a given date or no date (if empty),
         % otherwise the current date is printed

\begin{document}
\maketitle

\section{Loading 3D Models}

Describe how models should be loaded. Framework and formate etc.

\section{Calculation by Rasterization}
\subsection{From 3D Models to Rasterized Points}\label{sec:modelsToPoints}
The 3D models and sea floor has to be rasterized on to a view plane in order to calculate output intensity using CUDA.

We plan to use openGL to construct depth and reflection coefficient buffers at each along-track position. For this we need to set up a camera in a scene that mimic the sonar array imaging objects on a sea floor. The output from this step is a depth and a reflection coefficient buffer for each along-track position. Investigation has to be done to see if information is redundant along the track. We might also need to calculate per-pixel normals for use in the next step.

\subsection{Calculating Intensity}\label{sec:calcIntensity}
For each pixel, load normal, range, and reflection coefficient. Calculating the contribution from this location is then done by calculating the reciprocal doubled range, and weight it with the reflection coefficient and the normal component towards the along-track position. Saving these calculations to a buffer, we can now find the slant-range pixel by summing all values closer than half the sampling distance for the given slant-range value.

\subsection{Parallel Design Pattern for the GPU}
It is natural that we loop over along-track positions. At each position we generate the listed buffers in Section \ref{sec:modelsToPoints} and holds them in GPU memory. The buffers is then fed to a custom CUDA kernel that calculates the reflected intensity per buffer location. All buffers should be placed in texture memory so that resolution can be adjusted at any step. If we render each buffer listed in Section \ref{sec:modelsToPoints} at resolution $M \times N$, the step in Section \ref{sec:calcIntensity} can easily be rendered at $2M \times 2N$ by interpolation. Finally we arrive to what is believed to be the less parallel part of this approach, the sum operation. A kernel has to be launched that for each slant-range accumulates the total contribution in the given intensity buffer. We might want to use a library kernel for this if it exist. If we are going to code this step, one idea could be as follow: Assign a region to a block of treads, find min and max, and let one thread search for one ore more slant-range values\footnote{Need to do some more thinking here}.

\subsection{Concluding Remarks}
Uses a lot of memory, many calculations. Flexible. Will support point-based objects and sea floor out of the box.

\section{Geometric Calculation of Output Intensity}
Inspired by the Field II ultrasound simulator, where array responses are found by considering geometry, the same concept can be used for reflection off object geometry. Geometry in our case can be viewed as arrays, where the emitted energy is reflected energy from a point source at the along-track position.

Given a list of triangle patches, were the three corners of each triangle has a corresponding reflection coefficient, normal and position/depth, the reflection intensity can be found by simple geometric considerations. First we should sort triangles in depth, so that GPU-threads in a block, process vertices that are close to each other in range. This to maximize throughput of writing to global memory. One thread loads a triangle and iterates from min to max depth using the slant-range sampling distance as step length. For each step the thread calculates the intersection between the triangle and a sphere with radius given by the current depth. The intersection length is weighted with both the reciprocal depth, the reflection coefficient and the triangle normal \footnote{We should only use triangles with equal normals in each corner. Hence they are all equal to the average triangle normal. Might not be compliant.}.

\subsection{Concluding Remarks}
Uses little memory, few calculations. Will not support point-based objects and sea floor out of the box.

\end{document}

### 2012-10-05 HRUIonGPU/gfx/all_plots.py
Before final revision of ius article

@@ -102,13 +102,13 @@ def barBenchmarkGPU(filename, M, Nx, Ny):
    
    fig = figure()
    ax = fig.add_subplot(1,1,1)
-   cax1 = ax.bar(x_range, yavg1[:,0], width, color='0.25')
-   cax4 = ax.bar(x_range, yavg1[:,1], width, bottom=yavg1[:,0], color='0.5')
-   cax5 = ax.bar(x_range, yavg1[:,2], width, bottom=(yavg1[:,0] + yavg1[:,1]), color='0.75')
+   cax5 = ax.bar(x_range, yavg1[:,2], width, color='0.25')
+   cax4 = ax.bar(x_range, yavg1[:,1], width, bottom=yavg1[:,2], color='0.5')
+   cax1 = ax.bar(x_range, yavg1[:,0], width, bottom=(yavg1[:,1] + yavg1[:,2]), color='0.75')
    
-   cax2 = ax.bar(x_range + 1/3.0, yavg2[:,0], width/3.0, color='0.25', hatch='//')
-   cax3 = ax.bar(x_range + 2/3.0, yavg3[:,0], width/3.0, color='0.25', hatch='\\')
-   
+   cax2 = ax.bar(x_range + 1/3.0, yavg2[:,0], width/3.0, bottom=(yavg1[:,1] + yavg1[:,2]), color='0.75', hatch='/')
+   cax3 = ax.bar(x_range + 2/3.0, yavg3[:,0], width/3.0, bottom=(yavg1[:,1] + yavg1[:,2]), color='0.75', hatch='o')
+
    #cax = ax.plot([0,0.001],[0,0], 'k')
    #cax = ax.plot([0,0.001],[0,0], ':k')
    #cax = ax.plot([0,0.001],[0,0], '--k')
@@ -127,7 +127,7 @@ def barBenchmarkGPU(filename, M, Nx, Ny):
    #cax = ax.plot(range(min_L,max_L+1), yavg3_sum, '--k')
 
    if M > 32:
-      x_range = arange(min_L, max_L, 2)
+      x_range = arange(min_L, max_L+1, 2)
    
    ax.set_title('M=%d, %d beams, %d samples in range'%(M,Nx,Ny), fontsize='large')
    xlabel('L')

### 2012-10-08 HRUIonGPU/conf_article.tex

@@
 \IEEEauthorblockA{
-\IEEEauthorrefmark{1} Mi Lab, Norwegian University of Science and Technology, Trondheim, Norway,\\
+\IEEEauthorrefmark{1} MI-Lab, Norwegian University of Science and Technology, Trondheim, Norway,\\
 \IEEEauthorrefmark{2} Department of Informatics, University of Oslo, Oslo, Norway
@@
 
-For a typical cardiac ultrasound image of $80\times400$ pixels ($70\degree$ sector, 15cm range) acquired using a 2.5MHz, M=64 element phased array, we obtain 10 fps (subarray length L=M/2, temporal smoothing over 3 samples).
+For a typical cardiac ultrasound image of $80\times400$ pixels ($70\degree$ sector, 15 cm range) acquired using a 2.5 MHz, M=64 element phased array, we obtain 10 fps (subarray length L=M/2, temporal smoothing over 3 samples).
 If we do a 2-element pre-beamforming to get the channel count down to 32, the frame rate increases to 44 fps.
@@
 \begin{figure}
-\centerline{\includegraphics[width=3in]{gfx/beamforming_mv_lowres.png}}
+\centerline{\includegraphics[width=3in]{gfx/beamforming_mv.pdf}}
 \caption{Adaptive beamforming.
@@
 where $\vec{a} = [1_0, 1_1, ..., 1_{L-1}]^T = \vec{1}_L$ when $\vec{x}$ is pre-delayed. 
-
 And last, the beamformer output is calculated by combining all $N_L$ subarrays weighted with the same set of adaptive weights,
@@
 The CUDA compute model is based on execution of a kernel function across a grid of compute threads, where the kernel function describes the work to be done by each thread.
-As depicted in Figure \ref{fig:gpulayout}, each position in the grid holds a block of threads, with maximal size of 1024 threads, that can be synchronized with little cost if needed.
+As depicted in Figure \ref{fig:gpulayout}, each position in the grid holds a block of threads, with maximal size of 1024 threads, that supports low-cost synchronization if needed.
 %Each block is further divided into groups of 32 threads, known as a warp. The maximum number of threads inside a block is restricted to 512, hence 16 warps. 
-From a CUDA perspective the GPU consists of one or more stream multi processors (SMs), where each SM is capable of 32 or more, depending on the architecture, concurrent multiplications and/or additions.
-A group of blocks is scheduled to each SM until all blocks have been processed.
+From a CUDA perspective the GPU consists of one or more stream multi processors (SMs), where each SM performs single-instruction-multiple-data (SIMD) operations.
+A group of blocks is scheduled to each SM until all blocks in the grid have been processed.
 
-A SM has available a limited amount of near-core memory and registers.
+An SM has available a limited amount of near-core memory and registers.
 Each thread therefore has to be as light on memory as possible and has to perform enough instructions per used byte in order to hide memory latency.
@@
 All of these constraints combined pose a real challenge as the target problem has to map on to the parallel framework both on a grid, block and memory consumption level in order to benefits from parallel acceleration.
-In the next section, we will describe how this mapping can be done for Capon beamforming of cardiac ultrasound data.
+In the next section, we will describe how this mapping can be done for the Capon beamformer.
 
@@
 \section{Parallel Capon}\label{sec:meth}
-If we analyze the beamformer in (\ref{eq:z_mv}) in combination with (\ref{eq:w}) and (\ref{eq:R}), we see that each calculation of a set of Capon weights is independent across the outputs $z[n]$.
+If we analyze (\ref{eq:z_mv}) in combination with (\ref{eq:w}) and (\ref{eq:R}), we see that each calculation of a set of Capon weights is independent across the output $z[n]$.
 As shown in Figure \ref{fig:gpulayout}, a first level of granularity is therefore to divide the image into a grid, where the output in each cell can be calculated independently of all others.
@@
 The straightforward sum in (\ref{eq:R}) has complexity $O(L^2N_LN_K)$.
-Note that for a typical array configuration ($M \in [32, 128]$) we do more spatial smoothing than temporal smoothing.
+Note that for a typical array configuration ($M \in [32, 128]$) we do more spatial than temporal smoothing.
 Hence, $N_K$ is usually small compared to the number of subarrays ($N_L$).
 If we let $L = M/\alpha$, $L^2N_L$ becomes $(\alpha - 1)L^3 + L^2$, $\alpha \in [1, M]$.
-A common choice for $\alpha$ is between two and four (cite!!!).
+A common choice for $\alpha$ is between two and four \cite{Synnevag2007, Mehdizadeh2012, Synnevag2007a, Austeng2008}.
 In this interval, calculating the sample covariance matrix is cubic with $L$.
@@
 From (\ref{eq:R}) it is obvious that calculation of one element in $\mat{\hat{R}}$ is independent of all the other elements.
-A natural level of granularity would then be to assign one thread to each element of $\mat{\hat{R}}$ \cite{Chen2011}.
-This will minimize global reads and writes needed per thread, but unfortunately the maximum number of threads per block puts an upper limit on the subarray length, $L_{max}=32$, since we want to avoid block-to-block synchronization.
+A natural level of granularity would then be to assign one thread to each element of $\mat{\hat{R}}$ as in \cite{Chen2011}.
+This will minimize global reads and writes per thread, but unfortunately the maximum number of threads per block puts an upper limit on the subarray length we then can process, $L_{max}=32$, in order to avoid block-to-block synchronization.
 
-The cubic complexity with $L$ and linear complexity with $N_K$ can be reduced by realizing that the calculation in (\ref{eq:R}) overlaps both across subarrays and in time.
+The cubic complexity with $L$ can be reduced by realizing that the calculation in (\ref{eq:R}) overlaps across subarrays.
 Element $(i,j)$ of $\mat{\hat{R}}[n]$ can be calculated from element $(i-1, j-1), \text{for } i,j \in [1, L-1]$ as
@@
 %- \mat{\breve{R}}[n-K-1] + \mat{\breve{R}}[n-K].
-Both sliding across subarrays and in time can be realized simultaneously if we hold an $N_K$-average in time of the outer products $\vec{x}^{\Delta}[n]\vec{x}^{\Delta}[n]^H$ in memory and calculate $\mat{\hat{R}}$ averaged over subarrays from this buffer.
-However, since $N_K$ typically is small, and we have a limited amount of memory per thread on the GPU, we have chosen to focus on sliding across subarrays only (\ref{eq:sliding}).
-In that case, the complexity of calculating $\mat{\hat{R}}$ is reduced to $O(L^2)$.
+%Both sliding across subarrays and in time can be realized simultaneously if we hold an $N_K$-average in time of the outer products $\vec{x}^{\Delta}[n]\vec{x}^{\Delta}[n]^H$ in memory and calculate $\mat{\hat{R}}$ averaged over subarrays from this buffer.
+However, since $N_K$ typically is small, and we have a limited amount of memory per thread on the GPU, we have chosen to focus on sliding across subarrays only.
+The complexity of calculating $\mat{\hat{R}}$ is now reduced to $O(L^2)$.
 Important to notice is that sliding will break independence in the sliding dimension.
-With our way of calculating $\mat{\hat{R}}$, calculations are still independent across time, but not along diagonals in $\mat{\hat{R}}$.
+The calculation in (\ref{eq:sliding}) is still independent across time, but not along diagonals in $\mat{\hat{R}}$.
 We have therefore selected to use one thread per diagonal in the upper triangle of $\mat{\hat{R}}$.
@@
 For the adaptive diagonal loading factor in (\ref{eq:diag}), the trace of $\mat{\hat{R}}$ has to be known before diagonal elements are written back to global memory.
-Since one row for all covariance matrices is written to global memory per kernel iteration, the diagonal elements need to be held in a local memory until the trace is accumulated.
+Since one row for all covariance matrices is written to global memory per kernel iteration, the diagonal elements need to be held in near-core memory until the trace is accumulated.
 When it is found, scattered writes of the diagonal elements to global memory have to be performed.
@@
 The formula in (\ref{eq:z_mv}) has been implemented with one thread per element-weight ($L$ threads per weight vector).
-After the solution vector $\vec{b}$ has been placed in shared memory, one of the $L$ threads finds the sum of $\vec{b}$.
+After the solution vector $\vec{b}$ has been placed in near-core memory, one of the $L$ threads finds the sum of $\vec{b}$.
 Each of the $L$ threads then reads one element of $\vec{b}$ and scales it with the inverse sum just found to get the weight vector $\vec{w}[n]$.
@@
 \begin{figure*}[!t]
-\centerline{\subfloat[]{\includegraphics[width=3.9in]{gfx/benchmark_us_32_lr.png} \label{fig:benchUS32}}
+\centerline{\subfloat[]{\includegraphics[width=3.49in]{gfx/benchmark_bar_32_40_400.png} \label{fig:benchUS32}}
 \hfil
-\subfloat[]{\includegraphics[width=3.9in]{gfx/benchmark_us_64_lr.png}%
+\subfloat[]{\includegraphics[width=3.49in]{gfx/benchmark_bar_64_80_400.png}%
 \label{fig:benchUS64}}}
@@
 Execution time is plotted as a function of subarray length $L \in [1, M/2]$.
-The plot shows execution times for the three steps listed in Section \ref{sec:meth}: Calculation of covariance matrices in red, solver in green, and the final beamformer sum in blue.
-For each step, there is one line per choice of temporal averaging, $K \in [0, 1, 2]$.
+The figures show execution times for the three steps listed in Section \ref{sec:meth}: Calculation of covariance matrices in light gray, solver in gray, and the final beamformer sum in dark gray.
+For the calculation of covariance matrices, there is one bar per choice of temporal averaging, $K \in [0, 1, 2]$.
 (a) A 32 element array.
@@
 \section{Benchmarks and Videos}
-Figure \ref{fig:bench} shows the results from benchmarking the implementation described in Section \ref{sec:meth} on a Nvidia Quadro 6000 graphics card (14 SMs, 1 TFlops single precision, 6 GB global memory, and 144 GB/s memory bandwidth).
+Figure \ref{fig:bench} shows the results from benchmarking the implementation described in Section \ref{sec:meth} on an Nvidia Quadro 6000 graphics card (14 SMs, 1 TFlops single precision, 6 GB global memory, and 144 GB/s memory bandwidth).
 The problem sizes are calculated based on 32 and 64 element arrays with center frequency 2.5 MHz.
@@
 
-In the attached videos one will find a cardiac loop (add ref to video), and a simulated phantom (add ref to videos) processed with different types of settings given to the Capon beamformer.
+At (cite) one will find videos of a cardiac loop, 
+%(add ref to video)
+and a simulated phantom 
+%(add ref to videos) 
+processed with different types of settings given to the Capon beamformer.
 Cardiac data was acquired using a 64 element and 3.5 MHz phased array.
@@
 Further research should therefore focus on making this contribution smaller.
-This can be done either by implementing faster solvers, or reducing the sample covariance matrix size in a novel way (ref?).
-When doing such a reduction, including the pre-beamforming described above, one should investigate the loss of contrast and resolution compared with full Capon beamforming.
-
+This can be done either by implementing faster solvers, or reducing the sample covariance matrix size in a novel way \cite{Nilsen2009}.
+%When doing such a reduction, including the pre-beamforming described above, one should investigate the loss of contrast and resolution compared with full Capon beamforming.
 As for now, we have shown that the Capon beamformer can be implemented in a GPU framework.
 And if we as in \cite{Chen2011} define real-time as 10 fps, all execution times  in Figure (\ref{fig:bench}) are real time.
-If we on the other hand define real-time as the time it takes to acquire a given image.
-The image formed using a 32 and 64 element array has a real time requirement of 125 fps and 62.5 fps respectively.
+If we on the other hand define real-time as the time it takes to acquire a given image, the image formed using a 32 and 64 element array has a real time requirement of 125 fps and 62.5 fps respectively.
 There is some distance to cover before that goal is reached for all values of $L \le M/2$.
-However we see, by the last definition, that real time is possible for $M=64, L < 16$, and for $M = 32, L < 13$.
-
-(Should comment on videos)
+However we see that real time, by the last definition, is possible for $M=64, L < 16$, and for $M = 32, L < 13$.
+(Should comment on videos? Ran out of space)
 

### 2012-10-09 HRUIonGPU/jrnl_article.tex

@@
 
-The quadratic complexity with $L$ and linear complexity with $N_K$ can be reduced by realizing that the calculation in (\ref{eq:R}) overlap both across subarrays and in time.
+The cubic complexity with $L$ and linear complexity with $N_K$ can be reduced by realizing that the calculation in (\ref{eq:R}) overlap both across subarrays and in time.
 Element $(i,j)$ of $\mat{\hat{R}}[n]$ can be calculated from element $(i-1, j-1), \text{for } i,j \in [1, L-1]$ as
@@
 However, the way we have decided to organize the GPU compute threads per $\mat{\hat{R}}$, the lower triangle can be calculated with little overhead on the GPU.
-This also means that we do not relay on a symmetric solver to solve $\mat{\hat{R}}\vec{b} = \vec{1}_L$, even if that could have saved some instructions.
+This also means that we do not rely on a symmetric solver to solve $\mat{\hat{R}}\vec{b} = \vec{1}_L$, even if that could have saved some instructions.
 The memory layout for the upper triangle only is also believed to be cumbersome.

### 2012-11-12 ASA2013/asa2013.tex
la2pdf commit - 2012:11:12	16:21:51	(Mon12.Nov)

--- 2012-06-01 FFI-Simulator/ffi-simulator.tex
@@
 
-\begin{center}
-
-{\Large Design Suggestion}\\[0.2cm]
-{\itshape for a}\\[0.2cm]
-{\Large GPU Based Sonar Template Simulator}\\[2cm]
-{\itshape by}\\[0.2cm]
-\begin{minipage}{0.35\textwidth}\small\centering
-Jo Inge \textsc{Buskenes}\\
-\textnormal{\color{SeaGreen}joibu (at) ifi.uio.no}
-\end{minipage}
-\begin{minipage}{0.35\textwidth}\small\centering
-Jon Petter \textsc{Åsen}\\
-\textnormal{\color{SeaGreen}jpaasen (at) ifi.uio.no}
-\end{minipage}\\[2cm]
-{\itshape on the behalf of}\\[0.2cm]
-\begin{minipage}[t]{0.49\textwidth}\centering
-  \centering\includegraphics[height=3cm]{gfx/FFILogo.png}\\
-  \textsc{The Norwegian Defence Research Establishment (FFI)}
-\end{minipage}\\[0.5cm]
-\begin{minipage}[t]{0.4\textwidth}\small\centering
-Øivind \textsc{Midtgård}\\
-\textnormal{\color{SeaGreen}oivind.midtgaard (at) ffi.no}
-\end{minipage}
-\begin{minipage}[t]{0.4\textwidth}\small\centering
-Herman \textsc{Midelfart}\\
-\textnormal{\color{SeaGreen}herman.midelfart (at) ffi.no}
-\end{minipage}\\[0.3cm]
-\begin{minipage}{0.32\textwidth}\small\centering
-Roy Edgar \textsc{Hansen}\\
-\textnormal{\color{SeaGreen}roy-edgar.hansen (at) ffi.no}
-\end{minipage}
-\vfill
-\small
-\today
-\end{center}
-
-\clearpage
-\pagestyle{fancy}
-%~~ Sections ~~%                                %~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~%
+\begin{abstract}
+hello
+\end{abstract}
 
@@
 
-Template matching is a common technique used when classifying objects in SAS images.
-The principle is to isolate an image segment containing an object of interest, correlate it with a set of template images, and assign it to the class of the template yielding the highest correlation coefficient.
-The challenge is to come up with a suitable set of template images, considering the vast span of possible seafloor and object characteristics.
-
-FFI target this challenge by using a sonar simulator that can take a seafloor model or data as input, in which various target objects are postitioned in various ways.
-For each object configuration the scene is rendered and the result added to the template set.
-But while producing templates on-the-fly is a very flexible solution, it is also computationally demanding.
-
-This is why we are investiging the feasibility of using a GPU to speed up the simulations.
-
-\begin{figure}[b]
-\graphicsAI[drawing,width=\linewidth]{gfx/specs.svg}
-\caption{Simulator specifications.}\label{specs}
-\end{figure}
-
-\section{Specification}
-
-Most of the specifications received from FFI are illustrated in Figure \ref{specs}.
-
-\section{Design Suggestion}
-
-The way we see it, there are two possible approaches to implementing the mentioned simulator (figure \ref{implementation}). One can either
-
-\begin{figure}[b]
-\begin{narrow}{-2.5cm}{-2.5cm}
-\graphicsAI[drawing,width=\linewidth]{gfx/simulator.svg}
-\end{narrow}
-\caption{Two possible implementation strategies.}\label{implementation}
-\end{figure}
-
-\begin{itemize0}
-\item use OpenGL to produce depth-maps of the scene, and CUDA to illuminate it as a sonar would, or
-\item use OpenGL to render the full scene with lighting.
-\end{itemize0}
-
-In either case the simulator will take a seafloor and target model as input, and generate a simulated sonar image.
-To offload the CPU further, the GPU can then proceed to compute the correlation between the template and the isolated image segment, and an image mask.
-This should be fairly easy to implement, will compute efficiently on the GPU, and allows us to avoid sending each template back to the CPU over the limited bandwidth CPU$\leftrightarrow$GPU bus.
-Note that these steps will have to be carried out in CUDA regardless of the approach selected.
-
-In addition we see that each approach shares the same initial steps.
-The risk of choosing an OpenGL-only implementation is therefore reduced, since we may easily switch to the CUDA implementation if an OpenGL prototype fails to produce the desired results.
-
-\subsection{Using only OpenGL for the simulator?}
-
-\begin{table}[b]\centering
-\begin{tabular}[c]{l l l}\hline
-\rowcolor{tabBlue}                  & CUDA/OpenGL    & OpenGL only \\\hline
-Development time (per person)       & 6 weeks        & 4 weeks     \\
-Development complexity              & Moderate       & Low \\
-Amount of code                      & More           & Less \\
-Maintainability                     & Moderate       & Moderate \\
-Code readability                    & Moderate       & Moderate \\
-Signal model                        & Arbitrary      & Light rays  \\
-Sound$\rightarrow$light calibration & Not necessary  & Necessary(?) \\
-Multipath support                   & Possible       & Possible \\
-Speckle support                     & Possible       & Hard \\
-Diffraction support                 & Possible       & Hard \\
-Complex image                       & Easy           & Hard \\
-ATI/nVidia support                  & nVidia only    & Both \\
-\end{tabular}
-\vspace{5pt}\caption{Comparison of implementation approaches}\label{comparison}
-\end{table}
-
-When we wish to set up a 3D scene, move objects and cameras around, and render it using a GPU, it is better to use e.g.
-OpenGL than CUDA.
-OpenGL was made for this specific task, is cross-platform, feature-rich and well-matured.
-By looking at the graphics quality and frame-rates displayed by modern OpenGL-based games it should also be fairly straight-forward to get the performance we need out of this library.
-
-Our original idea was to use OpenGL to produce depth-maps of the object and (optionally) the seafloor, and then use CUDA to compute how a sonar array would illuminate this scene.
-However, since were only interested in pixel intensities, and sound may be treated as ``light'' with reasonable accuracy, we believe that OpenGL can be used to simulate the entire scene.
-This way we may support whatever OpenGL supports with little extra effort.
-The gaming industry demands speed, so such a simulator should also be very fast.
-
-The main advantage of the CUDA implementation lies in its flexibility.
-If desired, it can be modified to handle sound propagation in non-ideal mediums, penetrable objects, signal multipath, diffraction and phase-dependent effects such as speckle.
-However, this will also lead to increased code complexity and development time, and  will naturally increase.
-However, we expect that such features will be more realizable in CUDA than OpenGL.
-
-We have compared the two methods in Table \ref{comparison}.
-
-\section{Final thoughts}
-
-We propose using OpenGL to simulate the scene, and model the array as lightsources.
-This will make it unsuitable for simulating speckle and diffraction effects, but a "simple'' simulator should probably not take such effects into account anyhow.
-Penetrable objects can be simulated in either case, although we are not sure whether the OpenGL framework can support this with ease.
-An OpenGL implementation will be highly portable, and believed to support the functionality needed to generate a good enough template image.
-
-We aim to start development in July.
-The goal is to have an OpenGL prototype ready within a couple weeks, and from there decide which approach to go for.
-We propose monthly progress reports and meetings to discuss further
-development.
-
 \end{document}

### 2012-11-14 ASA2013/asa2013.tex
la2pdf commit - 2012:11:14	18:43:24	(Wed14.Nov)
la2pdf commit - 2012:11:14	19:33:49	(Wed14.Nov)

@@
 \begin{abstract}
-hello
+The use of adaptive beamformers in active sonar use of a Minimum Variance (MV) beamforming have demostrated
+
+The quality of active sonar images can be improved by the use of a Minimum Variance (MV) beamformer.
+
+Several studies have shown that using a Minimum Variance (MV) beamformer when forming active sonar images
+
+The Minimum Variance Distortionless Reponse (MVDR) beamformer has lately been applied to active sonar imaging.
+It displays a potential for improving the image quality beyond the limits of conventional methods, but rely on the estimation and inversion of a spatial covariance matrix.
+This translates to a complexity that is cubic with the number of elements $M$, O($M^3$), major complexity involved,  its use is discouraged due to its major computational complexity and need for robustification.
+This is because it involves the estimation and inversion of a spatial covariance matrix,
+
+Its ability to be It is reported to improve .
+In many scenarios it offers Its  can be applied to active sonar imaging, and  to image quality of active sonar systems   beamformer have in recent studies displayed shown to improve image quality in active sonar systems.
+
+  is an adaptive technique with the potential to improve active sonar imaging. One such method is the Minimum Variance Distorionless
+
+Their  beamforming active sonar imaging 
+Adaptive beamformers have recently been introduced in active sonar imaging 
+- If algorithm is slow - What technology will help me?
+- What skills are needed?
+- What tools are available and what's the cost?
+- What acceleration factor are we looking at?
+- How can I get started?
+
+- What we did. Which tools? Was it hard? Speedup?
 \end{abstract}

### 2012-11-14 ASA2013/asa2013.tex
la2pdf commit - 2012:11:14	20:38:41	(Wed14.Nov)

@@
 \begin{abstract}
-The use of adaptive beamformers in active sonar use of a Minimum Variance (MV) beamforming have demostrated
+The Minimum Variance Distortionless Reponse (MVDR) beamformer has lately been applied to active sonar imaging.
+While displaying an ability to improve the image quality beyond the limits of conventional beamformers, it relies on the estimation and inversion of a spatial covariance matrix for each the image pixels.
+This may discourage its use, since robustification methods must be introduced to ensure numerical and statistical validity, and the computational complexity is cubic with the number of elements, O($M^3$).
 
-The quality of active sonar images can be improved by the use of a Minimum Variance (MV) beamformer.
+We will show that for systems with less than $M$=32 channels, the estimation process is the main bottleneck.
+In this case, we suggest arithmetic optimizations for the estimation step, and explain how the the MVDR can be mapped to a high-end GPU to yield processing rates of more than 1\,Mpx/s.
+When $M$ exceeds 32, the inversion step dominates, and
 
-Several studies have shown that using a Minimum Variance (MV) beamformer when forming active sonar images
+This represents a speedup of 2-3 orders of magnitude over a moderately optimized C and Matlab reference implementation. For higher channel systems, we employ MVDR beamspace processing on the GPU
 
-The Minimum Variance Distortionless Reponse (MVDR) beamformer has lately been applied to active sonar imaging.
-It displays a potential for improving the image quality beyond the limits of conventional methods, but rely on the estimation and inversion of a spatial covariance matrix.
-This translates to a complexity that is cubic with the number of elements $M$, O($M^3$), major complexity involved,  its use is discouraged due to its major computational complexity and need for robustification.
+ With a channel count that exceeds 32, we show that we switch to using frequency based methods we run on a some algorithmic optimizations and porting of the  to a GPU
+
+Depending on the number of channels in the system, we target these issues in 3 different ways.
+
+  The computational complexity involved, as well as the need to introduce robustification techniques to make  This may discurage its use, since the estimation process requires  process The computational complexity involved, and the means required to make the well suited,
+
+Obtaining a robust estimate This translates to a complexity that is cubic with the number of elements $M$, O($M^3$), major complexity involved,  its use is discouraged due to its major computational complexity and need for robustification.
 This is because it involves the estimation and inversion of a spatial covariance matrix,
@@
 
-\section{Introduction}
+% \section{Introduction}
 

### 2012-11-14 ASA2013/asa2013.tex
la2pdf commit - 2012:11:14	21:10:31	(Wed14.Nov)

--- 2012-11-12 ASA2013/asa2013.tex
@@
 \begin{abstract}
-hello
+The Minimum Variance Distortionless Reponse (MVDR) beamformer has lately been applied to active sonar imaging.
+While displaying an ability to improve the image quality beyond the limits of conventional beamformers, it relies on the estimation and inversion of a spatial covariance matrix for each the image pixels.
+This may discourage its use, since robustification methods must be introduced to ensure numerical and statistical validity\todo{denne henger}, and the computational complexity is cubic with the number of elements, O($M^3$).
+
+We will show that for systems with less than 32 channels, the estimation process is the main bottleneck.
+In this case, we suggest arithmetic optimizations for the estimation step, and explain how the MVDR can be mapped to a high-end GPU to yield processing rates of more than 1\,Mpx/s.
+When $M$ exceeds 32, we show that frequency domain processing can be employed instead, which allows a comparable performance to be maintained, at a neglible reduction in image quality.
+Compared to a reference C and Matlab implementation, the GPU implementation offer a speedup of 2-3 orders of magnitude, but is not a trivial design.
+
+An alternative to the MVDR is the Low Complexity Adaptive beamformer (LCA).
+It does not compute a weightset, but merely computes the beamformer output for a range of predefined weight sets, and selects the one that best fulfills the MVDR optimization criteria.
+We will show that with little effort the LCA can produce images comparable to the MVDR, with hardly any development cost or complexity, and in a way that is extremely well suited for the GPU.
+
+\ \\
+Total: 258 words (200 max)
+\ \\
+Questions they wanted answered:
+\begin{itemize}
+\item If algorithm is slow - What technology will help me?
+\item What skills are needed?
+\item What tools are available and what's the cost?
+\item What acceleration factor are we looking at?
+\item How can I get started?
+\item What we did. Which tools? Was it hard? Speedup?
+\end{itemize}
 \end{abstract}
 
-\section{Introduction}
+% \section{Introduction}
 

### 2012-11-15 HRUIonGPU/jrnl_article.tex

@@
 % can use linebreaks \\ within to get better formatting as desired
-\title{Implementing Capon Adaptive Beamforming on the GPU for Real Time Cardiac Ultrasound Imaging}
+\title{Implementing Capon Beamforming on the GPU for Real Time Cardiac Ultrasound Imaging}
 %
@@
 Since we have decided to have $L$ threads per output, the data is reduced to length $L$ by letting thread $r$ calculate
-\begin{align}
+\begin{align}\label{eq:subsum}
 \breve{x}_{r}^{\Delta} = \sum_{k=r}^{r+N_L-1}x_{k}^{\Delta} \text{ for } r \in [0, L-1].
@@
 Nilsen et al.
-\cite{Nilsen2009} have proposed to apply beamspace processing to further reduce the required computations.
+\cite{Nilsen2009} have proposed to apply beamspace processing to further reduce the required computations for the Capon beamformer.
 With beamspace we mean transforming the channel data from element space to a fan of beams covering the sector illuminated by transmit beams.
@@
 \begin{align}
-\vec{x}_{BS} = \mat{B}\vec{x}^{\Delta}.
+\vec{x}_{BS}[n] &= \mat{B}\vec{x}^{\Delta}[n], \ \ \ \mat{B} \in C^{M,M}\\
+[\mat{B}]_{p,q} &= \frac{1}{\sqrt{M}}e^{-j 2 \pi p q / M}.
 \end{align} 
 The steering vectors in $\mat{B}$, the so-called Butler matrix, determine each beam's direction in space.
-We can therefore easily remove beams where no interference is present by removing rows in $\mat{B}$.
-For a focused system like a cardiac ultrasound scanner, where the received signal is concentrated in a narrow band around broadside, there is a small percentage of the beams that contains almost all the energy.
+We can therefore easily remove beams where no interference is present by removing rows from $\mat{B}$.
+
+For a focused system like cardiac ultrasound imaging, where the received signal is concentrated in a narrow band around broadside, there is a small percentage of the beams that contains almost all the energy.
 Nilsen et al.
 concluded that as little as three beams around broadside were adequate to produce results comparable to applying Capon beamforming in element space.
-
-The beamspace version of Capon is derived by replacing the covariance matrix in (\ref{eq:w}) with the beamspace covariance matrix $\mat{\hat{R}}_{BS} = f(\vec{x}_{BS})$, where $\mat{B} \in \mathbb{C}^{N_b,M}$, and $N_b$ is the number of selected beams.
+Hence, the inversion steps is found to be constant with respect to the number of array elements.
+
+The beamspace version of the Capon beamformer is derived by replacing the covariance matrix in (\ref{eq:w}) with the beamspace covariance matrix $\mat{\hat{R}}_{BS} = f(\vec{x}_{BS}) = f(\mat{B}\vec{x})$, where $\mat{B} \in \mathbb{C}^{N_b,L}$, and $N_b$ is the number of selected beams.
 Note that when subarray averaging is applied, the transformation must be done per subarray, yielding subarrays that no longer overlap in space.
-Hence, the amount of data is change from $M$ to $N_LN_b$ per sample.
-
-(Go in detail about the implementation using sliding DFT).
-
-(Go in detail about what kind of modifications we need to do to each of the three steps)
+If we want to apply the transformation before the covariance matrix is constructing, the amount of data is change from $M$ to $N_LN_b$ per sample.
+The implementation calculating covariance matrices needs to be adjusted to handle non-overlapping subarrays.
+BuildR complexity is changed to $O(N_b^2L)$, so constructing the covariance matrices will be faster in beamspace as long as $N_b < \sqrt{L}$.
+However it will involve large changes to the processing pipeline described in Section \ref{sec:meth}.
+
+In both cases we need to solve using a transformed steering vector
+\begin{align}
+\vec{a}_{BS} &= \mat{B}\vec{1}_L = \sqrt{L}\vec{e}_1,
+\end{align}
+where $\vec{e}_i$ has value one in the i$^{th}$ position and is zero everywhere else.
+
+Another choice is to transform the covariance matrix to beamspace. 
+\begin{align}
+\mat{\hat{R}}_{BS} &= \mat{B}\mat{\hat{R}}\mat{B}^H\\
+\end{align}
+By this approach we go from solving $\mat{\hat{R}}\vec{b} = \vec{1}_L$, which is an $L \times L$ system, to solving the $N_b \times N_b$ system $\mat{\hat{R}}_{BS}\vec{b}_{BS} = (\sqrt{L}\vec{e}_1)$.
+Now, one can either choose to transform the subarray sums, as in (\ref{eq:subsum}), to beamspace or transform the result, $\vec{b}_{BS}$, back to element space.
+
+In our implementation we have tried to minimize the number of adjustments that had to be done in order to integrate beamspace processing into the GPU-Capon processing pipeline.
+The beamspace-results presented in this paper is therefore produced by transforming covariance matrices to beamspace, and after solving, transforming the result back to element space.
+
+%(Go in detail about the implementation using sliding DFT).
+
+%(Go in detail about what kind of modifications we need to do to each of the three steps)
+
+%Note that $\mat{B}\mat{B}^H = \mat{I}$ and $\mat{B}^H\mat{B} \ne \mat{I}$ when $N_b < L$.
+
+%Another way of computing the beamspace transformation is to transform the steering vector $\vec{a}$ and covariance vector $\mat{\hat{R}}$ beforing solving the following system of linear equations: $(\mat{B}\mat{\hat{R}}\mat{B}^H)^{-1}(\mat{B}a) = b_{BS}$.
 

### 2012-11-15 ASA2013/asa2013.tex
la2pdf commit - 2012:11:15	12:24:43	(Thu15.Nov)

@@
 While displaying an ability to improve the image quality beyond the limits of conventional beamformers, it relies on the estimation and inversion of a spatial covariance matrix for each the image pixels.
-This may discourage its use, since robustification methods must be introduced to ensure numerical and statistical validity\todo{denne henger}, and the computational complexity is cubic with the number of elements, O($M^3$).
+This may discourage its use, since robustification methods must be introduced to ensure numerical and statistical validity\todo{denne tror jeg vi kutter}, and the computational complexity is cubic with the number of channels, O($M^3$).
 
-We will show that for systems with less than 32 channels, the estimation process is the main bottleneck.
-In this case, we suggest arithmetic optimizations for the estimation step, and explain how the MVDR can be mapped to a high-end GPU to yield processing rates of more than 1\,Mpx/s.
+For systems with less than 32 channels, we suggest arithmetic optimizations for the estimation step, and explain how the MVDR can be mapped to a high-end GPU to yield processing rates of more than 1\,Mpx/s.
 When $M$ exceeds 32, we show that frequency domain processing can be employed instead, which allows a comparable performance to be maintained, at a neglible reduction in image quality.
-Compared to a reference C and Matlab implementation, the GPU implementation offer a speedup of 2-3 orders of magnitude, but is not a trivial design.
+Compared to our reference C and Matlab implementation, the GPU implementation offered a speedup of 2-3 orders of magnitude, but at the expense of a non-trivial design.
 
-An alternative to the MVDR is the Low Complexity Adaptive beamformer (LCA).
-It does not compute a weightset, but merely computes the beamformer output for a range of predefined weight sets, and selects the one that best fulfills the MVDR optimization criteria.
-We will show that with little effort the LCA can produce images comparable to the MVDR, with hardly any development cost or complexity, and in a way that is extremely well suited for the GPU.
+A very good alternative to the MVDR is the Low Complexity Adaptive beamformer (LCA).
+It does not compute a weightset, but merely computes the beamformer output for each of a each predefined set of weights, and selects the one that best fulfills the MVDR criteria.
+The images it produces are comparable to the MVDR, without the development cost or complexity of the MVDR, and it is perfectly suited for a GPU.
 
 \ \\
-Total: 258 words (200 max)
+Total: 239 words (200 max)
+
+\ \\
+Noen tanker jeg har selv nå:
+\begin{itemize}
+\item Dette blir mye beamforming, beamforming, beamforming. Mulig jeg må riste av meg ``JOE formen'' og fokusere litt mer på implementasjonsaspektet.
+\item Det er mye ord, og litt muntlig. Det går seg til så snart jeg har bestemt meg for innhold.
+\item Det mangler noe om bruksområder.
+\item Må passe på å ikke selge LCA så bra at MVDR ser helt håpløs ut(?)
+\item 200 ord er ingenting. :)
+\end{itemize}
+
 \ \\
@@
 \begin{itemize}
+\item This session will feature work on adapting computationally demanding modeling and signal processing tasks to take advantage of various parallel architectures, such as cloud computing, computing clusters, multicore CPUs, multi-core graphic processing units (GPUs), DSP chips, and FPGAs.
 \item If algorithm is slow - What technology will help me?

### 2012-11-15 ASA2013/asa2013.tex
la2pdf commit - 2012:11:15	12:56:24	(Thu15.Nov)

@@
 \begin{abstract}
-The Minimum Variance Distortionless Reponse (MVDR) beamformer has lately been applied to active sonar imaging.
-While displaying an ability to improve the image quality beyond the limits of conventional beamformers, it relies on the estimation and inversion of a spatial covariance matrix for each the image pixels.
-This may discourage its use, since robustification methods must be introduced to ensure numerical and statistical validity\todo{denne tror jeg vi kutter}, and the computational complexity is cubic with the number of channels, O($M^3$).
+The MVDR adaptive beamformer has lately been shown to improve image quality in active sonar systems, but relies on the estimation and inversion of a spatial covariance for each image pixel.
+This translates to a computational complexity that is cubic with number of channels.
 

### 2012-11-15 ASA2013/asa2013.tex
la2pdf commit - 2012:11:15	13:27:01	(Thu15.Nov)

@@
 \begin{abstract}
-The MVDR adaptive beamformer has lately been shown to improve image quality in active sonar systems, but relies on the estimation and inversion of a spatial covariance for each image pixel.
-This translates to a computational complexity that is cubic with number of channels.
+The MVDR adaptive beamformer has lately been shown to improve image quality in active sonar systems, but at a major computational expense.
+This is because the estimation and inversion of a spatial covariance matrix is required for each image pixel.
+We target this problem by altering and mapping the MVDR algorithm GPU, and suggest three different solutions depending on the system size.
 
-For systems with less than 32 channels, we suggest arithmetic optimizations for the estimation step, and explain how the MVDR can be mapped to a high-end GPU to yield processing rates of more than 1\,Mpx/s.
-When $M$ exceeds 32, we show that frequency domain processing can be employed instead, which allows a comparable performance to be maintained, at a neglible reduction in image quality.
-Compared to our reference C and Matlab implementation, the GPU implementation offered a speedup of 2-3 orders of magnitude, but at the expense of a non-trivial design.
+For systems with less than 32 channels, we suggest arithmetic optimizations for the estimation step, and show how a careful mapping to a high-end GPU to yield processing rates of more than 1\,Mpx/s.
+When the system exceeds this size, we show that frequency domain processing can be employed instead, which allows a comparable performance to be maintained, at a neglible reduction in image quality.
+These GPU implementations consistently reduced the runtime by 2-3 orders of magnitude compared to our reference C and Matlab implementations.
 
-A very good alternative to the MVDR is the Low Complexity Adaptive beamformer (LCA).
+Should this this still be insufficient, we suggest having a look that the LCA beamformer.
 It does not compute a weightset, but merely computes the beamformer output for each of a each predefined set of weights, and selects the one that best fulfills the MVDR criteria.
@@
 \ \\
-Total: 239 words (200 max)
+Total: 215 words (200 max)
 
@@
 \begin{itemize}
-\item Dette blir mye beamforming, beamforming, beamforming. Mulig jeg må riste av meg ``JOE formen'' og fokusere litt mer på implementasjonsaspektet.
-\item Det er mye ord, og litt muntlig. Det går seg til så snart jeg har bestemt meg for innhold.
 \item Det mangler noe om bruksområder.

### 2012-11-15 ASA2013/asa2013.tex
la2pdf commit - 2012:11:15	14:32:28	(Thu15.Nov)
la2pdf commit - 2012:11:15	16:40:44	(Thu15.Nov)

@@
 This is because the estimation and inversion of a spatial covariance matrix is required for each image pixel.
-We target this problem by altering and mapping the MVDR algorithm GPU, and suggest three different solutions depending on the system size.
+We target this problem by altering and mapping the MVDR algorithm to a GPU, and suggest three different solutions depending on the system size.
 
-For systems with less than 32 channels, we suggest arithmetic optimizations for the estimation step, and show how a careful mapping to a high-end GPU to yield processing rates of more than 1\,Mpx/s.
-When the system exceeds this size, we show that frequency domain processing can be employed instead, which allows a comparable performance to be maintained, at a neglible reduction in image quality.
+For systems with less than 32 channels, we suggest arithmetic optimizations for the estimation step, and show how a GPU can be used to yield processing rates of more than 1\,Mpx/s.
+When the system exceeds this size, we show that frequency domain processing can be employed, which allows a comparable performance to be maintained, at a neglible reduction in image quality.
 These GPU implementations consistently reduced the runtime by 2-3 orders of magnitude compared to our reference C and Matlab implementations.

### 2012-11-15 ASA2013/asa2013.tex
la2pdf commit - 2012:11:15	17:14:28	(Thu15.Nov)

@@
 
-For systems with less than 32 channels, we suggest arithmetic optimizations for the estimation step, and show how a GPU can be used to yield processing rates of more than 1\,Mpx/s.
-When the system exceeds this size, we show that frequency domain processing can be employed, which allows a comparable performance to be maintained, at a neglible reduction in image quality.
+For systems with relatively few channels, we suggest arithmetic optimizations for the estimation step, and show how a GPU can be used to yield processing rates of more than 1\,Mpx/s.
+For larger systems  we show that frequency domain processing is preferable.
+This allows a comparable performance to be maintained at a negligible reduction in image quality.
 These GPU implementations consistently reduced the runtime by 2-3 orders of magnitude compared to our reference C and Matlab implementations.
 
-Should this this still be insufficient, we suggest having a look that the LCA beamformer.
-It does not compute a weightset, but merely computes the beamformer output for each of a each predefined set of weights, and selects the one that best fulfills the MVDR criteria.
-The images it produces are comparable to the MVDR, without the development cost or complexity of the MVDR, and it is perfectly suited for a GPU.
+For even larger systems we suggest employing the LCA beamformer.
+It does not compute a weightset, but merely computes the beamformer output for each of a predefined set of weights, and selects the one that best fulfils the MVDR criterion.
+The LCA creates images with a quality comparable to MVDR, and it is perfectly suited for a GPU.
 

### 2012-11-15 ASA2013/asa2013.tex
la2pdf commit - 2012:11:15	17:45:53	(Thu15.Nov)

@@
 \begin{abstract}
-The MVDR adaptive beamformer has lately been shown to improve image quality in active sonar systems, but at a major computational expense.
+The MVDR adaptive beamformer has been shown to improve image quality in active sonar systems, but at a major computational expense.
 This is because the estimation and inversion of a spatial covariance matrix is required for each image pixel.
@@
 
-For systems with relatively few channels, we suggest arithmetic optimizations for the estimation step, and show how a GPU can be used to yield processing rates of more than 1\,Mpx/s.
+For systems with relatively few channels, we suggest arithmetic optimizations for the estimation step, and show how a GPU can be used to yield image creation rates of more than 1\,Mpx/s.
 For larger systems we show that frequency domain processing is preferable.
-This allows a comparable performance to be maintained at a negligible reduction in image quality.
+This promotes high processing rates at a negligible reduction in image quality.
 These GPU implementations consistently reduced the runtime by 2-3 orders of magnitude compared to our reference C and Matlab implementations.

### 2012-11-15 ASA2013/asa2013.tex
la2pdf commit - 2012:11:15	18:18:24	(Thu15.Nov)

@@
 
+\title{Adapting the MVDR to a GPU for active sonar systems of different sizes.}
+
 \begin{abstract}
-The MVDR adaptive beamformer has been shown to improve image quality in active sonar systems, but at a major computational expense.
-This is because the estimation and inversion of a spatial covariance matrix is required for each image pixel.
-We target this problem by altering and mapping the MVDR algorithm to a GPU, and suggest three different solutions depending on the system size.
+The MVDR beamformer has been shown to improve active sonar image quality compared to conventional methods.
+Unfortunately, it is also significantly more computationally expensive because a spatial covariance matrix must be estimated and inverted for each image pixel.
+We target this challenge by altering and mapping MVDR to a GPU, and suggest three different solutions depending on the system size.
 

### 2012-11-26 ASA2013/asa2013.tex

@@
 
-\title{Adapting the MVDR to a GPU for active sonar systems of different sizes.}
+{\Large\bf Adapting the MVDR beamformer to a GPU for differently sized active sonar imaging systems.}
 
@@
 For even larger systems we suggest employing the LCA beamformer.
-It does not compute a weightset, but merely computes the beamformer output for each of a predefined set of weights, and selects the one that best fulfils the MVDR criterion.
+It does not calculate a weightset, but merely computes the beamformer output for each of a predefined set of weights, and selects the one that best fulfils the MVDR criterion.
 The LCA creates images with a quality comparable to MVDR, and it is perfectly suited for a GPU.
@@
 \ \\
-Total: 215 words (200 max)
+Total: 192 words (200 max)
 

### 2012-11-26 HRUIonGPU/jrnl_article.tex

@@
 
+Derivation of sliding beamspace transformation. Given a $M$ long data vector divided into $N_L$ subarrays, the beamspace transformation of the $l^{th}$ subarray is given as:
+\begin{align}
+x_{BS,l,k} = \frac{1}{\sqrt{L}}\sum_{p=i}^{l+L-1}x_p e^{-j2\pi k(p-l)/L} , k \in [0, N_b-1].
+\end{align}
+Manipulating this expression we find that
+\begin{align}
+x_{BS,l,k} &= \frac{1}{\sqrt{L}}((x_{BS,l-1,k} - x_{l-1})e^{-j2\pi k/L} \\ &+ x_{i+L-1}e^{-j2\pi k(L-1)/L}) \\
+&= \frac{1}{\sqrt{L}}(x_{BS,l-1,k} - x_{l-1} + x_{i+L-1})e^{-j2\pi k/L}
+\end{align}
+For $l=0$ a full computation has to be performed.
+Instructions can be saved if we utilize the overlap across subarrays.
+However, since the number of beams usually is small we will end up with a large memory per thread ratio ($M > N_b$).
+Using one GPU-thread per beam is therefore not feasible for small $N_b$ large $M$ configurations.
+The solution is to combine sliding with full computations.
+A number of thread groups target different sets of subarrays which they transform using the sliding beamspace algorithm.
+In that way, number of threads per sample is increased and thus the memory per thread ratio is decreased.
+Notice how subarrays no longer overlap after the beamspace transformation.
+The amount of data is increased from $M$ to $N_LN_b$ per sample.
+This means that we no longer can exploit spatial overlap when calculating the sample covariance matrix, but since $N_b$ is small we can now use one thread per element.
+
+How do we calculate the number of possible subarrays $N_L$ with overlap $L-s$ in a data vector of length $M$.
+\begin{align}
+N_L = floor(\frac{M}{s}) - floor(\frac{L-1}{s})
+\end{align}
+
 %(Go in detail about the implementation using sliding DFT).

### 2012-12-09 HRUIonGPU/jrnl_article.tex
Started fineshing up capon-gpu article

--- 2012-11-15 HRUIonGPU/jrnl_article.tex
@@
 %\boldmath
-An amazing abstract goes here. An amazing abstract goes here. An amazing abstract goes here. An amazing abstract goes here.
-An amazing abstract goes here. An amazing abstract goes here. An amazing abstract goes here. An amazing abstract goes here.
-An amazing abstract goes here. An amazing abstract goes here. An amazing abstract goes here. An amazing abstract goes here.
-An amazing abstract goes here. An amazing abstract goes here. An amazing abstract goes here. An amazing abstract goes here.
-An amazing abstract goes here. An amazing abstract goes here. An amazing abstract goes here. An amazing abstract goes here.
-An amazing abstract goes here. An amazing abstract goes here. An amazing abstract goes here. An amazing abstract goes here.
-An amazing abstract goes here. An amazing abstract goes here. An amazing abstract goes here. An amazing abstract goes here.
-An amazing abstract goes here. An amazing abstract goes here. An amazing abstract goes here. An amazing abstract goes here.
-An amazing abstract goes here. An amazing abstract goes here. An amazing abstract goes here. An amazing abstract goes here.
-An amazing abstract goes here. An amazing abstract goes here. An amazing abstract goes here. An amazing abstract goes here.
+...
 \end{abstract}
@@
 %\IEEEPARstart{I}{ntroduction} to Capon adaptive beamforming, ultrasound imaging and GPU/CUDA.
-\IEEEPARstart{C}{apon} adaptive beamforming \cite{Capon1969} applied to medical ultrasound imaging has recently shown promising results in several studies \cite{Synnevag2007, Austeng2008, Vignon2008, Viola, Mehdizadeh2012}.
-It has been shown that the method introduces some interesting trade-offs in addition to just improved resolution, like maintaining image resolution in combination with higher framerates, smaller probes, or deeper penetration \cite{Synnevag2009}.
+\IEEEPARstart{C}{apon} beamforming \cite{Capon1969} applied to medical ultrasound imaging has recently shown promising results in several studies \cite{Synnevag2007, Austeng2008, Vignon2008, Viola, Mehdizadeh2012}.
+Some interesting trade-offs in addition to just improved resolution has been introduced, like maintaining image resolution in combination with higher framerates, smaller probes, or deeper penetration \cite{Synnevag2009}.
 Despite these benefits, widespread use has been limited by the immense computational complexity associated with the method \cite{So2011}.
-In this paper we introduce the first implementation of the Capon beamformer exhibiting real-time performance when applied in a typical cardiac ultrasound imaging setting.
+In this paper we introduce an implementation of the Capon beamformer exhibiting real-time performance when applied in a typical cardiac ultrasound imaging setting.
 To accomplish this task we make use of the massive, and easily accessed, parallel processing power found in modern Graphics Processing Units (GPUs), combined with novel methods to reduce the computational complexity as the problem size grows.
 
-Previous work on implementing a parallel Capon beamformer have mainly been focused around radar and passive sonar systems using custom made systolic array processors \cite{McWhirter1989, Moonen1993, Sinha2002}.
-%A systolic array processor is custom made hardware, and even if the computational throughput might be higher for selected problems compared with a GPU, the cost related to development and flexibility is considered much higher.
+Previous work on parallel implementations of the Capon beamformer have mainly been focused around radar and passive sonar systems using custom make systolic array processors \cite{McWhirter1989, Moonen1993, Sinha2002}.
 For medical ultrasound, Chen et al.
-\cite{Chen2011a, Chen2011} have implemented a Capon beamformer on the GPU that produced real apodization weights based on real data.
-Under this constraint the Capon beamformer is not allowed to shift the main lobe, and therefore its full potential is not reached.
-Shifting the main lobe is an important property of the Capon beamformer in order for it to improve edge definitions and contrast throughout an ultrasound image.
-Another aspect of using complex data is that even though the number of aritmetic operations per sample is increased we can now sample at the system bandwidth, keeping the number of samples for a full cardiac imaging sector at a minimum.
+\cite{Chen2011a, Chen2011} implemented a Capon beamformer on the GPU that produced real apodization weights based on real data.
+Under this constraint the Capon beamformer is not allowed to micro-steer the main lobe, and therefore its full potential is not reached.
+Micro-steering of the main lobe is an important property of the Capon beamformer in order for it to improve structural details and edge definitions throughout an ultrasound image.
+Another aspect of using complex data is that even though the number of arithmetic operations per sample is increased we can sample at the system bandwidth, keeping the number of samples for a full cardiac imaging sector at a minimum.
 All results in this paper is reported with the use of baseband data.
@@
 % denne må utelates i IUS-artikkelen
-In recent work (cite JOE) we have applied a GPU-based Capon beamformer in an active sonar system.
+In recent work \cite{asen2012}(cite JOE or ECUA?) we have applied a GPU-based Capon beamformer in both an active sonar system and for cardiac ultrasound imaging.
 For sonar the real-time requirement is much lower than for medical ultrasound imaging, which means that most common array configurations can be handled using the GPU without taking any additional measures.
@@
 Since the execution time for a Capon beamformer depends highly on the number of elements in the given array, it is important to find ways of reducing the computational complexity for high-channel-count systems.
-Different approaches to low-complexity Capon beamforming have recently been proposed \cite{Asl2012, Jensen2012}, and in this paper we have selected to implement the beamspace transformation described by Nilsen and Hafizovic \cite{Nilsen2009} on the GPU to maintain high framerates as the channel count increases.
+Different approaches to low-complexity Capon beamforming have therefore recently been proposed \cite{Synnevag2011, Asl2012, Jensen2012, Kim}.
+In this paper we have selected to implement the beamspace transformation described by Nilsen and Hafizovic \cite{Nilsen2009} on the GPU to maintain high framerates as the channel count increases.
 %(Referer til JOE-artikkelen, få med beskrivelse av hva som blir anerledes for ultralyd.)
@@
 
-%\subsection{Capon Beamforming}
+\subsection{Capon Beamforming}
 
@@
 
-The conventional form of array beamforming is delay-and-sum (DAS).
-Here, the signal $x_m[n]$ recorded at element number $m \in [0,M-1]$ from time $n = 0$ to $N-1$ is delayed to broadside with an appropriate delay $\Delta_m[n]$.
+The conventional form of array beamforming in medical ultrasound imaging is delay-and-sum (DAS).
+Here, the signal $x_m[n]$ recorded at element number $m \in [0,M-1]$ at time $n$ is delayed to broadside with an appropriate delay $\Delta_m[n]$.
 The beamformer output, $z[n]$, is then the sum of all $M$ channels weighted with a per-element factor $w_m$,
@@
 \end{align}
-In conventional beamforming, the weighting $\vec{w}$ is usually real, and is applied to trade resolution for a better signal-to-noise ratio (SNR).
-
-As shown in Figure \ref{fig:mvbf}, the basic idea of an adaptive beamformer is to form a weight vector based on the data at hand.
-For Capon adaptive beamforming the weights are found that minimize the beamformer output power while maintaining unit gain at broadside.
+The weight vector or window $\vec{w}$ is usually real, and is applied to trade resolution for a better signal-to-noise ratio (SNR).
+Even if different windows can be used in different areas of the images they have traditionally been predefined.
+
+As shown in Figure \ref{fig:mvbf}, the basic idea of an adaptive beamformer is to form a complex weight vector based on the received data.
+For Capon beamforming the weights are found that minimize the beamformer output power while maintaining unit gain at broadside.
 The result is that interference impinging from other directions than the point of focus is suppressed \cite{Synnevag2007}.
@@
 \begin{figure}
-\centerline{\includegraphics[width=3in]{gfx/beamforming_mv_lowres.png}}
-\caption{Adaptive beamforming.
+\centerline{\includegraphics[width=3in]{gfx/beamforming_mv.pdf}}
+\caption{Capon beamforming.
 Impinging signal is demodulated, and aligned in phase by applying steering and focusing delays $\Delta_m[n]$.
-Adaptive weights, $\vec{w}[n]$, are then calculated based on in-phase data, and finally the output is formed by a weighted coherent sum.}
+Adaptive weights, $\vec{w}[n]$, are then calculated based on in-phase data, and finally the output is formed by a Capon-weighted coherent sum.}
 \label{fig:mvbf}
@@
 
-The Capon weight vector is given by the following expression:
-\begin{align}\label{eq:w}
-\vec{w}[n] = \frac{\mat{\hat{R}}[n]^{-1}\vec{a}}{\vec{a}^H\mat{\hat{R}}[n]^{-1}\vec{a}} = \frac{\vec{b}}{\vec{a}^H\vec{b}},
-\end{align}
-where $\vec{a} = [1_0, 1_1, ..., 1_{M-1}]^T = \vec{1}_M$ when $\vec{x}$ is pre-delayed, $\mat{\hat{R}}$ is the sample covariance matrix calculated from the data vector $\vec{x}^\Delta[n]$, and $\vec{b} = \mat{\hat{R}}[n]^{-1}\vec{a}$.
-%\begin{align}
-%\mat{\hat{R}}[n] = \vec{x}[n]\vec{x}[n]^H.
-%\end{align} 
-%Typically the sample covariance matrix has to be averaged over several samples in order to ensure numerical stability and statistical robustness.
-
-In passive systems, assuming stationary statistics, the sample covariance matrix is typically averaged over a long sequence of samples\cite{Krima}.
-For an active, pulsed system like medical ultrasound imaging, the available data from which $\mat{\hat{R}}$ can be estimated is quite limited.
-The resulting sample covariance matrix might therefore be both inaccurate, and poorly conditioned.
-Another challenge with applying Capon to active systems is the effect of signal cancellation, where high correlation between signal and interference favors weights that cancel out the signal \cite{Reddy1987}.
-In this way, the beamformer output can be too low at broadside, even when there is signal present and the distortionless-response criterion is applied.
-
-In order to get a well-conditioned $\mat{\hat{R}}$, avoid signal cancellation, and to get DAS-like speckle, $\mat{\hat{R}}$ has to be averaged over $L\le M/2$ long subarrays and $N_K \sim T_p/T_s$ time samples \cite{Synnevag2007, Synnevag2007a}.
-Where, $T_p$ is the pulse length in seconds and $T_s$ is the sampling period.
-We therefore calculate
+The Capon beamformer can be summerized in three main steps. First a sample covariance matrix has to be calculated, 
 \begin{align}
-\mat{\breve{R}}[n] = \frac{1}{N_LN_K}\sum_{n'=n-K}^{n+K} \sum_{l=0}^{N_L-1} \vec{x}_l^{\Delta}[n']\vec{x}_l^{\Delta}[n']^H,\label{eq:R}
-\end{align}
-where  $N_K = 2K + 1$ is the number of samples included in time, $N_L = M-L+1$ is the number of subarrays, and $\vec{x}_l^{\Delta}$ is the $l^\text{th}$ subarray $[x_l^{\Delta}[n], \dotso, x_{l+L-1}^{\Delta}[n]]$.
-The dimension of the covariance matrix after this smoothing is $L \times L$.
-Hence, the size of $\vec{w}[n]$ and $\vec{a}$ in (\ref{eq:w}) has to be $L \times 1$ when the covariance matrix in (\ref{eq:R}) is used.
-Finally, the covariance matrix is loaded with a diagonal factor $\epsilon$ to ensure numerical stability, 
+\mat{\breve{R}}[n] = \frac{1}{N_LN_K}\sum_{n'=n-K}^{n+K} \sum_{l=0}^{N_L-1} (\vec{x}_l^{\Delta}[n'])(\vec{x}_l^{\Delta}[n'])^H,\label{eq:R}
+\end{align}
+where  $N_K = 2K + 1$ is the number of samples included in time, $\vec{x}_l^{\Delta}$ is the $l^\text{th}$ subarray of length $L$ $[x_l^{\Delta}[n], \dotso, x_{l+L-1}^{\Delta}[n]]$ and $N_L = M-L+1$ is the number of subarrays.
+Note that the covariance matrix dimension is $L \times L$.
+The covariance matrix is then loaded with a diagonal factor $\epsilon$ to ensure numerical stability and increased robustness,
 \begin{align}\label{eq:diag}
-\mat{\hat{R}} =  f(\vec{x}^{\Delta}[n]) = \mat{\breve{R}} + \epsilon\mat{I}.
+\mat{\hat{R}}[n] =  f(\vec{x}^{\Delta}[n]) = \mat{\breve{R}}[n] + \epsilon[n]\mat{I}.
 \end{align}
@@
 \begin{align}\label{eq:diag_adapt}
-\epsilon &= d \times tr\{\mat{\breve{R}}\}, \\
-d &= 1/L
-\end{align}
-introduced by Synnev\aa{}g et al.
-\cite{Synnevag2007} has been applied in much of the resent literature on Capon beamforming for medical ultrasound imaging \cite{Nilsen2009, Wang2009, Mehdizadeh2012}.
-This form of adaptive diagonal loading has also been applied to HF and VHF antenna processing \cite{Featherstone1997}.
-
-The beamformer output is in the end calculated by combining all $N_L$ subarrays weighted with the same set of adaptive weights.
+\epsilon[n] &= d \times \frac{tr\{\mat{\breve{R}}[n]\}}{L}
+\end{align}
+has been applied in much of the resent literature on Capon beamforming for medical ultrasound imaging \cite{Synnevag2007, Nilsen2009, Wang2009, Mehdizadeh2012}.
+This form of adaptive diagonal loading has also been applied to HF and VHF antenna processing \cite{Featherstone1997b}.
+
+Second, a linear system of equations has to be solved
+\begin{align}\label{eq:b}
+\vec{b}[n] = \mat{\hat{R}}[n]^{-1}\vec{1}_L,
+\end{align}
+where $\vec{1}_L = [1_0, 1_1, ..., 1_{L-1}]^T$. Finally the Capon weight vector is formed based on (\ref{eq:b}) as
+\begin{align}\label{eq:w}
+\vec{w}[n] = \frac{\vec{b}[n]}{\vec{1}^H\vec{b}[n]},
+\end{align}
+and the beamformer output is calculated by combining all $N_L$ subarrays weighted with the same set of adaptive weights.
 \begin{align}
@@
 where $\text{CONV}_m$ is a non-truncated convolution across elements.
-Note how we can either select to sum all subarrays first in (\ref{eq:z_mv}), or accumulate the right weighting per element in (\ref{eq:z_mv2}).
+Note how we can choose to first sum all subarrays (\ref{eq:z_mv}), or accumulate the right weighting per element (\ref{eq:z_mv2}).
+
+In passive systems, assuming stationary statistics, the sample covariance matrix is typically averaged over a long sequence of samples\cite{Krima}.
+For an active, pulsed system like medical ultrasound imaging, the available data from which $\mat{\hat{R}}$ can be estimated is quite limited.
+The resulting sample covariance matrix might therefore be both inaccurate, and poorly conditioned.
+Another challenge with applying Capon to active systems is the effect of signal cancellation, where high correlation between signal and interference favors weights that cancel out the signal \cite{Reddy1987}.
+In this way, the beamformer output can be too low at broadside, even when there is signal present and the distortionless-response criterion is applied.
+
+In order to get a well-conditioned $\mat{\hat{R}}$, avoid signal cancellation, and to get DAS-like speckle, $\mat{\hat{R}}$ has to be averaged over $L\le M/2$ long subarrays and $N_K \sim T_p/T_s$ time samples \cite{Synnevag2007, Synnevag2007a}.
+Where, $T_p$ is the pulse length in seconds and $T_s$ is the sampling period.
 
@@
 
-The CUDA compute model is based on execution of a kernel function across a grid of compute threads, where the kernel function describe the work to be done by each thread.
-As depicted in Figure \ref{fig:gpulayout}, each position in the grid holds a block of threads, with maximal size of 1024 threads, that can be synchronized with little cost if needed.
-%Each block is further divided into groups of 32 threads, known as a warp. The maximum number of threads inside a block is restricted to 512, hence 16 warps. 
+The CUDA compute model is based on execution of a kernel function across a large set of threads, where the kernel function describe the work to be done in each thread.
+As depicted in Figure \ref{fig:gpulayout}, threads are organized in a grid of thread-blocks, with maximal size of 1024 threads per block.
+Threads inside a block supports low-cost synchronized if needed.
 From a CUDA perspective the GPU consist of one or more stream multi processors (SMs), where each SM is capable of 32 or more, depending on the architecture, concurrent multiplications and/or additions.
@@
 \section{Parallel Capon}\label{sec:meth}
-As mentioned, widespread use of Capon adaptive beamforming has been limited by its computational complexity.
-However, if we analyze the beamformer in (\ref{eq:z_mv}) in combination with (\ref{eq:w}) and (\ref{eq:R}), we see that each calculation of a set of Capon weights is independent across the outputs $z[n]$.
+If we analyze the beamformer in (\ref{eq:z_mv}) in combination with (\ref{eq:b}) and (\ref{eq:R}), we see that each calculation of a set of Capon weights is independent across the outputs $z[n]$.
 As shown in Figure \ref{fig:gpulayout}, a first level of granularity is therefore to divide the image into a grid, where the output in each cell can be calculated independently of all others.
@@
 \text{1) } &\text{Calculate the sample covariance matrix } \mat{\hat{R}} \text{ as in } \text{(\ref{eq:diag})}\\
-\text{2) } &\text{Solve the linear system of equations } \vec{b} =\mat{\hat{R}}^{-1}\vec{1}_L\\
+\text{2) } &\text{Solve the linear system of equations in (\ref{eq:b}})\\ %\vec{b} =\mat{\hat{R}}^{-1}\vec{1}_L\\
 \text{3) } &\text{Calculate } \vec{w} = \vec{b}/\vec{1}_{L}^H\vec{b} \text{ and the amplitude as in (\ref{eq:z_mv})}
@@
 For the results presented in this paper, we have used an unreleased GPU implementation of Gauss-Jordan (GJ) elimination (by Nvidia) to solve $\mat{\hat{R}}\vec{b} = \vec{1}_L$, where $\vec{b} = \mat{\hat{R}}^{-1}\vec{1}_L$.
-For smaller systems, $L < 10$, a direct solver using one thread per system is preferable.
+%For smaller systems, $L < 10$, a direct solver using one thread per system is preferable.
 For this purpose we have implemented Cramer's rule for $2 \times 2$, $3 \times 3$ and $4 \times 4$ matrices.
@@
 \end{align}
-Each thread now computes $z_r = w_r^*x_r$, and finally one thread computes and outputs $z = \sum_{r=0}^{L-1} z_r$ to global memory.
+Each thread now computes $z_r = w_r^*x_r$, and finally one thread computes and outputs $z = \frac{1}{N_L}\sum_{r=0}^{L-1} z_r$ to global memory.
 The last sum can also be performed using a binary reduction sum across several threads for increased speed.
@@
 Note that when subarray averaging is applied, the transformation must be done per subarray, yielding subarrays that no longer overlap in space.
-If we want to apply the transformation before the covariance matrix is constructing, the amount of data is change from $M$ to $N_LN_b$ per sample.
-The implementation calculating covariance matrices needs to be adjusted to handle non-overlapping subarrays.
-BuildR complexity is changed to $O(N_b^2L)$, so constructing the covariance matrices will be faster in beamspace as long as $N_b < \sqrt{L}$.
-However it will involve large changes to the processing pipeline described in Section \ref{sec:meth}.
-
+If we want to apply the transformation before the covariance matrix is constructing, the amount of data is change from $M$ to $N_LN_b$ per sample, however all remaining steps will get lower complexity.
+
+Another choice is to transform the element space covariance matrix to beamspace. 
+\begin{align}
+\mat{\hat{R}}_{BS} &= \mat{B}\mat{\hat{R}}\mat{B}^H
+\end{align}
+The pros using this approach is that minor changes has to be made to the pipeline described in Section \ref{sec:meth}, but it will only speed up the solving step which is why we have implemented the first approach.
+%By this approach we go from solving $\mat{\hat{R}}\vec{b} = \vec{1}_L$, which is an $L \times L$ system, to solving the $N_b \times N_b$ system $\mat{\hat{R}}_{BS}\vec{b}_{BS} = (\sqrt{L}\vec{e}_1)$.
+Now, one can either choose to transform the subarray sums, as in (\ref{eq:subsum}), to beamspace or transform the result, $\vec{b}_{BS}$, back to element space.
 In both cases we need to solve using a transformed steering vector
@@
 
-Another choice is to transform the covariance matrix to beamspace. 
+%The implementation calculating covariance matrices therefore needs to be adjusted to handle non-overlapping subarrays.
+The complexity of constructing the covariance matrix is changed to $O(N_b^2L)$, so constructing the covariance matrices will be faster in beamspace as long as $N_b < \sqrt{L}$.
+%Derivation of sliding beamspace transformation. 
+Given a $M$ long data vector divided into $N_L$ subarrays, the beamspace transformation of the $l^{th}$ subarray is given as:
+\begin{align}\label{eq:bs_formula}
+x_{BS,l,k} = \frac{1}{\sqrt{L}}\sum_{p=i}^{l+L-1}x_p e^{-j2\pi k(p-l)/L} , k \in [0, N_b-1].
+\end{align}
+Manipulating this expression we find that
+\begin{align}\label{eq:sliding_bs}
+x_{BS,l,k} &= \frac{1}{\sqrt{L}}((x_{BS,l-1,k} - x_{l-1})e^{-j2\pi k/L} \\ &+ x_{i+L-1}e^{-j2\pi k(L-1)/L}) \\
+&= \frac{1}{\sqrt{L}}(x_{BS,l-1,k} - x_{l-1} + x_{i+L-1})e^{-j2\pi k/L}
+\end{align}
+For $l=0$ a full computation has to be performed (\ref{eq:bs_formula}). We call this the sliding beamspace transformation.
+
+From (\ref{eq:sliding_bs}) it is obvious that instructions can be saved if we utilize the overlap across subarrays.
+However, since the number of beams usually is small we will end up with a large memory per thread ratio ($M > N_b$).
+Using one GPU-thread per beam is therefore not feasible for small $N_b$ large $M$ configurations.
+The solution is to combine sliding with full computations.
+As shown in Fig.
+(???), a number of thread groups target different sets of subarrays which they transform using the sliding beamspace algorithm.
+In that way, the number of threads per sample is increased and thus the memory per thread ratio is decreased.
+
+As mentioned the amount of data is increased from $M$ to $N_LN_b$ per sample in beamspace.
+This means that we no longer can exploit spatial overlap when calculating the sample covariance matrix, but since $N_b$ is small we can now use one thread per element.
+A kernel similar to the one described by Chen et al.
+\cite{Chen2011} has been implemented.
+After solving, the beamspace weight vector is calculated as
 \begin{align}
-\mat{\hat{R}}_{BS} &= \mat{B}\mat{\hat{R}}\mat{B}^H\\
-\end{align}
-By this approach we go from solving $\mat{\hat{R}}\vec{b} = \vec{1}_L$, which is an $L \times L$ system, to solving the $N_b \times N_b$ system $\mat{\hat{R}}_{BS}\vec{b}_{BS} = (\sqrt{L}\vec{e}_1)$.
-Now, one can either choose to transform the subarray sums, as in (\ref{eq:subsum}), to beamspace or transform the result, $\vec{b}_{BS}$, back to element space.
-
-In our implementation we have tried to minimize the number of adjustments that had to be done in order to integrate beamspace processing into the GPU-Capon processing pipeline.
-The beamspace-results presented in this paper is therefore produced by transforming covariance matrices to beamspace, and after solving, transforming the result back to element space.
+\vec{w}_{BS} = \frac{\vec{b}_{BS}}{\sqrt{L}\vec{e}_1\vec{b}_{BS}}
+\end{align}  
+and applied to the transformed data summed across subarrays.
+
+%How do we calculate the number of possible subarrays $N_L$ with overlap $L-s$ in a data vector of length $M$.
+%\begin{align}
+%N_L = floor(\frac{M}{s}) - floor(\frac{L-1}{s})
+%\end{align}
+
+The traditional way of applying adaptive diagonal loading is to load the sample covariance matrix with the average energy per element, hence $trace\{\mat{\hat{R}}\}/L$.
+If adaptive diagonal loading is applied after data has been transformed into beamspace, a high percentage of the total energy is mapped to a small subset of beams.
+Thus the average energy per beam is usually higher than the average energy per element, especially if the system uses focused transmit beams.
+This means that less diagonal loading has to be applied in beamspace than in element space ($d_{BS} \ll d_{element}$).
 
@@
 \begin{figure*}[!t]
-\centerline{\subfloat[]{\includegraphics[width=3.5in]{gfx/benchmark_us_32_lr.png} \label{fig:benchUS32}}
+\centerline{\subfloat[]{\includegraphics[width=3.5in]{gfx/benchmark_bar_32_40_400.png} \label{fig:benchUS32}}
 \hfil
-\subfloat[]{\includegraphics[width=3.5in]{gfx/benchmark_us_64_lr.png}%
+\subfloat[]{\includegraphics[width=3.5in]{gfx/benchmark_bar_64_80_400.png}%
 \label{fig:benchUS64}}}
@@
 \begin{figure*}[!t]
-\centerline{\subfloat[]{\includegraphics[width=3.5in]{gfx/benchmark_us_32_lr.png} \label{fig:benchBS3}}
+\centerline{\subfloat[]{\includegraphics[width=3.5in]{gfx/benchmark_bar_bs_M=64_Nx=80_Ny=400_Nb=3.png} \label{fig:benchBS3}}
 \hfil
-\subfloat[]{\includegraphics[width=3.5in]{gfx/benchmark_us_64_lr.png}%
+\subfloat[]{\includegraphics[width=3.5in]{gfx/benchmark_bar_bs_M=64_Nx=80_Ny=400_Nb=3.png}%
 \label{fig:benchBS5}}}
@@
 Toeplitz-based solver (cite new article). Discuss it.
+
+How does beamspace perform when using plane wave imaging...?
 

### 2012-12-11 HRUIonGPU/jrnl_article.tex

@@
 
-Previous work on parallel implementations of the Capon beamformer have mainly been focused around radar and passive sonar systems using custom make systolic array processors \cite{McWhirter1989, Moonen1993, Sinha2002}.
+Previous work on parallel implementations of the Capon beamformer have mainly been focused around radar and passive sonar systems using custom made systolic array processors \cite{McWhirter1989, Moonen1993, Sinha2002}.
 For medical ultrasound, Chen et al.
@@
 In this paper we have selected to implement the beamspace transformation described by Nilsen and Hafizovic \cite{Nilsen2009} on the GPU to maintain high framerates as the channel count increases.
+The beamspace transformation matches well with the focused beams used in cardiac ultrasound imaging.
 %(Referer til JOE-artikkelen, få med beskrivelse av hva som blir anerledes for ultralyd.)
@@
 In the next section we present background information about the Capon beamformer and the GPU computing model.
-In Section \ref{sec:meth} we go in detail about the parallel implementation of the Capon beamformer, before in Section \ref{sec:bs} we introduce the method applied when the problem size increases.
+In Section \ref{sec:meth} we go in detail about the parallel implementation of the Capon beamformer, before in Section \ref{sec:bs} we introduce the beamspace method and how it has been implemented on the GPU.
 Benchmark results and resulting images are presented in Section \ref{sec:res}, and finally we discuss the results in Section \ref{sec:dis}, and draw conclusions in Section \ref{sec:con}.
@@
 
-As shown in Figure \ref{fig:mvbf}, the basic idea of an adaptive beamformer is to form a complex weight vector based on the received data.
+As shown in Fig.
+\ref{fig:mvbf}, the basic idea of an adaptive beamformer is to form a complex weight vector based on the received data.
 For Capon beamforming the weights are found that minimize the beamformer output power while maintaining unit gain at broadside.
@@
 \begin{align}\label{eq:b}
-\vec{b}[n] = \mat{\hat{R}}[n]^{-1}\vec{1}_L,
-\end{align}
-where $\vec{1}_L = [1_0, 1_1, ..., 1_{L-1}]^T$. Finally the Capon weight vector is formed based on (\ref{eq:b}) as
+\vec{b}[n] = \mat{\hat{R}}[n]^{-1}\vec{a},
+\end{align}
+where the steering vector $\vec{a} = \vec{1}_L = [1_0, 1_1, ..., 1_{L-1}]^T$ since $\vec{x}^\Delta$ is pre-delayed. Finally the Capon weight vector is formed based on (\ref{eq:b}) as
 \begin{align}\label{eq:w}
@@
 The key point is that GPUs can accelerate small concurrent tasks, like beamforming one pixel, much better than a CPU.
-Where the CPU has few cores and logic to process complex threads, the GPU has many cores, but each thread needs to be simple in order to benefit from the extra computational power.
+Where the CPU has few cores and logic to process complex threads, the GPU has many cores, but each thread needs to be simple in order to benefit from the GPU's computational power.
 
 The CUDA compute model is based on execution of a kernel function across a large set of threads, where the kernel function describe the work to be done in each thread.
-As depicted in Figure \ref{fig:gpulayout}, threads are organized in a grid of thread-blocks, with maximal size of 1024 threads per block.
+As depicted in Fig.
+\ref{fig:gpulayout}, threads are organized in a grid of thread-blocks, with maximal size of 1024 threads per block.
 Threads inside a block supports low-cost synchronized if needed.
@@
 Each thread therefore has to be as light on memory as possible and has to perform enough instructions per used byte in order to hide memory latency.
+It is also important that the problem can be divided into enough threads to hide instruction latency.
 Global reads and writes on the GPU, in addition to CPU-GPU transfers, have high penalty, and should be minimized as much as possible.
-All of these constraints combined pose a real challenge as the target problem has to map on to the parallel framework both on a grid, block and memory consumption level in order to benefits from parallel acceleration.
+All of these constraints combined pose a real challenge as the target problem has to map on to the parallel framework both on a grid, block, thread and memory consumption level in order to benefits from parallel acceleration.
 In the next section, we will describe how this mapping should be done for Capon beamforming of cardiac ultrasound data.
@@
 If we analyze the beamformer in (\ref{eq:z_mv}) in combination with (\ref{eq:b}) and (\ref{eq:R}), we see that each calculation of a set of Capon weights is independent across the outputs $z[n]$.
-As shown in Figure \ref{fig:gpulayout}, a first level of granularity is therefore to divide the image into a grid, where the output in each cell can be calculated independently of all others.
+As shown in Fig.
+\ref{fig:gpulayout}, a first level of granularity is therefore to divide the image into a grid, where the output in each cell can be calculated independently of all others.
 Further, based on memory dependencies and to hold memory consumption per thread at a minimum, one should recognise that a group of threads should process one output in the final image.
@@
 
-\subsection{Calculating Multiple Sample Covariance Matrices}
-
-The straightforward sum in (\ref{eq:R}) has complexity $O(L^2N_LN_K) = O(L^3)$ since $N_L$, the number of subarrays, is a function of $L$.
+\subsection{Calculating Multiple Sample Covariance Matrices}\label{sec:calcR}
+
+%The straightforward sum in (\ref{eq:R}) has complexity $O(L^2N_LN_K) = O(L^3)$ since $N_L$, the number of subarrays, is a function of $L$.
 This is actually the same as for inverting an $L \times L$ matrix.
@@
 
-From (\ref{eq:R}) it is obvious that calculation of one element in $\mat{\hat{R}}$ is independent of all the others.
-A natural level of granularity would then be to assign one thread to each element of $\mat{\hat{R}}$.
-This will minimize global reads and writes needed per thread, but unfortunately the maximum number of threads per block puts an upper limit on the subarray length, $L_{max}=32$, if we want to avoid block-to-block synchronization.
+From (\ref{eq:R}) it is obvious that calculation of one element in $\mat{\hat{R}}$ is independent of all the other elements.
+A natural level of granularity would then be to assign one thread to each element of $\mat{\hat{R}}$ as in \cite{Chen2011}.
+This will minimize global reads and writes needed per thread, but unfortunately the maximum number of threads per block puts an upper limit on the subarray length, $L_{max}=32$, since we want to avoid block-to-block synchronization.
 
 The cubic complexity with $L$ and linear complexity with $N_K$ can be reduced by realizing that the calculation in (\ref{eq:R}) overlap both across subarrays and in time.
-Element $(i,j)$ of $\mat{\hat{R}}[n]$ can be calculated from element $(i-1, j-1), \text{for } i,j \in [1, L-1]$ as
+Element $(i,j)$ of $\mat{\breve{R}}[n]$ can be calculated from element $(i-1, j-1), \text{for } i,j \in [1, L-1]$ as
 \begin{align}
-\mat{\breve{R}}_{i,j}[n] &=  \mat{\breve{R}}_{i-1,j-1}[n]  + \sum_{n'=n-K}^{n+K} (x_{i+N_L-1}^{\Delta}[n']x_{j+N_L-1}^{\Delta}[n']^* \nonumber \\
+\mat{\breve{R}}_{i,j}[n] &=  \mat{\breve{R}}_{i-1,j-1}[n]  \nonumber \\
+&+ \sum_{n'=n-K}^{n+K} (x_{i+N_L-1}^{\Delta}[n']x_{j+N_L-1}^{\Delta}[n']^* \nonumber \\
 &- x_{i-1}^{\Delta}[n']x_{j-1}^{\Delta}[n']^H) \label{eq:sliding}
@@
 %- \mat{\breve{R}}[n-K-1] + \mat{\breve{R}}[n-K].
-Both sliding across subarrays and in time can be realized simultaneously if we hold a $N_K$-average in time of the outer products $\vec{x}^{\Delta}[n]\vec{x}^{\Delta}[n]^H$ in memory and calculate $\mat{\hat{R}}$ averaged over subarrays from this buffer (ref JOE).
-However, since $N_K$ typically is small, and we have a limited amount of memory per thread on the GPU, we have chosen to focus on sliding across subarrays only (\ref{eq:sliding}).
+Both sliding across subarrays and in time can be realized simultaneously if we hold a $N_K$-average in time of the outer products $\vec{x}^{\Delta}[n]\vec{x}^{\Delta}[n]^H$ in memory and calculate $\mat{\hat{R}}$ averaged over subarrays from this buffer (cite JOE?).
+However, since $N_K$ typically is small, and we have a limited amount of memory available per thread, we have chosen to focus on sliding across subarrays only (\ref{eq:sliding}).
 In that case, the complexity of calculating $\mat{\hat{R}}$ is reduced to $O(L^2)$.
-Important to notice is that sliding will break independence in the sliding dimension.
+Notice how sliding will break independence along the sliding dimension.
 With our choice of calculating $\mat{\hat{R}}$, calculations are still independent across time, but not along diagonals in $\mat{\hat{R}}$.
-We therefore have selected to use one thread per diagonal in the upper triangle of $\mat{\hat{R}}$, and for small $L$s several covariance matrices are calculated per thread block (Figure \ref{fig:gpulayout}).
+We therefore have selected to use one thread per diagonal in the upper triangle of $\mat{\hat{R}}$, and for small values of $L$, several covariance matrices are calculated per thread block (Fig.
+\ref{fig:gpulayout}).
 Maximum supported $L$ is with this approach increased to 512, however for small $L$ high $M$ configurations, memory consumption per thread will be to high to achieve full occupancy of the GPU.
@@
 This also means that we do not rely on a symmetric solver to solve $\mat{\hat{R}}\vec{b} = \vec{1}_L$, even if that could have saved some instructions.
-The memory layout for the upper triangle only is also believed to be cumbersome.
 
@@
 The beamspace version of the Capon beamformer is derived by replacing the covariance matrix in (\ref{eq:w}) with the beamspace covariance matrix $\mat{\hat{R}}_{BS} = f(\vec{x}_{BS}) = f(\mat{B}\vec{x})$, where $\mat{B} \in \mathbb{C}^{N_b,L}$, and $N_b$ is the number of selected beams.
-Note that when subarray averaging is applied, the transformation must be done per subarray, yielding subarrays that no longer overlap in space.
-If we want to apply the transformation before the covariance matrix is constructing, the amount of data is change from $M$ to $N_LN_b$ per sample, however all remaining steps will get lower complexity.
+The transformation must be done per subarray, yielding subarrays that no longer overlap in space.
+If the transformation is applied before the covariance matrix is constructing, the amount of input data is change from $M$ to $N_LN_b$ elements per sample, however all remaining steps will get lower complexity.
 
@@
 \end{align}
-The pros using this approach is that minor changes has to be made to the pipeline described in Section \ref{sec:meth}, but it will only speed up the solving step which is why we have implemented the first approach.
+The pros using this approach is that minor changes has to be made to the pipeline described in Section \ref{sec:meth}, but it will only speed up the solving step if the solution $\vec{b}_{BS}$ is transformed back to element space.
+In this paper we will describe how the first approach has been implemented on the GPU.
 %By this approach we go from solving $\mat{\hat{R}}\vec{b} = \vec{1}_L$, which is an $L \times L$ system, to solving the $N_b \times N_b$ system $\mat{\hat{R}}_{BS}\vec{b}_{BS} = (\sqrt{L}\vec{e}_1)$.
 Now, one can either choose to transform the subarray sums, as in (\ref{eq:subsum}), to beamspace or transform the result, $\vec{b}_{BS}$, back to element space.
-In both cases we need to solve using a transformed steering vector
+In both cases solving has to be performed against a transformed steering vector
 \begin{align}
@@
 \end{align}
-where $\vec{e}_i$ has value one in the i$^{th}$ position and is zero everywhere else.
+where $\vec{e}_i$ has value one in the i$^{th}$ position and is zero elsewhere.
 
@@
 \end{align}
-Manipulating this expression we find that
+Manipulating this expression we find that for $l \in [1, L-1]$
 \begin{align}\label{eq:sliding_bs}
-x_{BS,l,k} &= \frac{1}{\sqrt{L}}((x_{BS,l-1,k} - x_{l-1})e^{-j2\pi k/L} \\ &+ x_{i+L-1}e^{-j2\pi k(L-1)/L}) \\
-&= \frac{1}{\sqrt{L}}(x_{BS,l-1,k} - x_{l-1} + x_{i+L-1})e^{-j2\pi k/L}
-\end{align}
-For $l=0$ a full computation has to be performed (\ref{eq:bs_formula}). We call this the sliding beamspace transformation.
-
-From (\ref{eq:sliding_bs}) it is obvious that instructions can be saved if we utilize the overlap across subarrays.
-However, since the number of beams usually is small we will end up with a large memory per thread ratio ($M > N_b$).
-Using one GPU-thread per beam is therefore not feasible for small $N_b$ large $M$ configurations.
-The solution is to combine sliding with full computations.
+x_{BS,l,k} &= \frac{1}{\sqrt{L}}((x_{BS,l-1,k} - x_{l-1})e^{-j2\pi k/L} \nonumber \\ &+ x_{i+L-1}e^{-j2\pi k(L-1)/L}) \nonumber \\
+&= \frac{1}{\sqrt{L}}(x_{BS,l-1,k} - x_{l-1} + x_{i+L-1})e^{-j2\pi k/L}.
+\end{align}
+We call this the sliding beamspace transformation (SBS), and it is equal to a normalized sliding DFT. For $l=0$ a full computation has to be performed using (\ref{eq:bs_formula}).
+
+From (\ref{eq:sliding_bs}) it is obvious that instructions can be saved if we utilize the overlap across subarrays, with an increase in task parallel granularity (As discussed in Section \ref{sec:calcR}).
+With SBS we can therefore select to have one thread per beam (index $k$) that calculates this beam for all subarrays.
+However, since the number of beams usually is small we will end up with a large memory per thread ratio ($M \gg N_b$).
+Using one thread per beam is therefore not feasible for small $N_b$ large $M$ configurations.
+The solution is to combine SBS with full computations.
 As shown in Fig.
-(???), a number of thread groups target different sets of subarrays which they transform using the sliding beamspace algorithm.
+(???), a number of thread groups target different sets of subarrays which they transform using the SBS algorithm.
 In that way, the number of threads per sample is increased and thus the memory per thread ratio is decreased.
@@
 As mentioned the amount of data is increased from $M$ to $N_LN_b$ per sample in beamspace.
-This means that we no longer can exploit spatial overlap when calculating the sample covariance matrix, but since $N_b$ is small we can now use one thread per element.
+This means that we no longer can exploit spatial overlap when calculating the sample covariance matrix, but since $N_b$ is small we can now allow to have one thread per element in $\mat{\hat{R}}_{BS}$.
 A kernel similar to the one described by Chen et al.
-\cite{Chen2011} has been implemented.
-After solving, the beamspace weight vector is calculated as
+\cite{Chen2011}, has therefore been implemented.
+For small $N_b$ values, several covariance matrices are calculated per thread block to achieve high occupancy of the SMs.
+
+After solving $\vec{b}_{BS} = \sqrt{L}\mat{\hat{R}}_{BS}^{-1}\vec{e}_1$, the beamspace weight vector is calculated as 
 \begin{align}
@@
 \end{align}  
-and applied to the transformed data summed across subarrays.
+and applied to the beamspace data summed across subarrays.
 
@@
 
-The traditional way of applying adaptive diagonal loading is to load the sample covariance matrix with the average energy per element, hence $trace\{\mat{\hat{R}}\}/L$.
+As presented in (\ref{eq:diag}), adaptive diagonal loading is performed by loading the sample covariance matrix with the average energy per element, hence $trace\{\mat{\hat{R}}\}/L$.
 If adaptive diagonal loading is applied after data has been transformed into beamspace, a high percentage of the total energy is mapped to a small subset of beams.
-Thus the average energy per beam is usually higher than the average energy per element, especially if the system uses focused transmit beams.
+Thus the average energy per beam is higher than the average energy per element, if the system uses focused transmit beams.
 This means that less diagonal loading has to be applied in beamspace than in element space ($d_{BS} \ll d_{element}$).
@@
 Ta med en serie med bilde. Må vise tydelig at det gir forbedring.
+
+Add a comparison between beamspace and full capon using small L.
 

### 2012-12-12 HRUIonGPU/jrnl_article.tex
Added some nice figures

@@
 For sonar the real-time requirement is much lower than for medical ultrasound imaging, which means that most common array configurations can be handled using the GPU without taking any additional measures.
-For medical ultrasound imaging, and especially cardiac ultrasound imaging, frame rate requirements increase to a level where even a high-end GPU can not do the processing required for a standard implementation of Capon beamforming.
-Since the execution time for a Capon beamformer depends highly on the number of elements in the given array, it is important to find ways of reducing the computational complexity for high-channel-count systems.
+For medical ultrasound imaging, and especially cardiac ultrasound imaging, frame rate requirements increase to a level where even a high-end GPU can not do the processing required for a standard implementation of the Capon beamformer.
+Since the execution time highly depends on the number of elements in the given array, it is important to find ways of reducing computational complexity for high-channel-count systems.
 Different approaches to low-complexity Capon beamforming have therefore recently been proposed \cite{Synnevag2011, Asl2012, Jensen2012, Kim}.
-In this paper we have selected to implement the beamspace transformation described by Nilsen and Hafizovic \cite{Nilsen2009} on the GPU to maintain high framerates as the channel count increases.
-The beamspace transformation matches well with the focused beams used in cardiac ultrasound imaging.
+In this paper we have selected to implement the beamspace transformation, described by Nilsen and Hafizovic \cite{Nilsen2009}, on the GPU to maintain high framerates as the channel count increases.
+The beamspace transformation matches well with the focused transmit beams used in cardiac ultrasound imaging.
 %(Referer til JOE-artikkelen, få med beskrivelse av hva som blir anerledes for ultralyd.)
@@
 \end{align}
-where the steering vector $\vec{a} = \vec{1}_L = [1_0, 1_1, ..., 1_{L-1}]^T$ since $\vec{x}^\Delta$ is pre-delayed. Finally the Capon weight vector is formed based on (\ref{eq:b}) as
+where the steering vector $\vec{a} = \vec{1}_L = [1_0, 1_1, ..., 1_{L-1}]^T$ since $\vec{x}^\Delta$ is pre-delayed. Finally the Capon weight vector is formed from (\ref{eq:b}) as
 \begin{align}\label{eq:w}
@@
 
-In passive systems, assuming stationary statistics, the sample covariance matrix is typically averaged over a long sequence of samples\cite{Krima}.
+In passive systems, assuming stationary statistics, the sample covariance matrix in (\ref{eq:R}) is typically averaged over a long sequence of samples\cite{Krima}.
 For an active, pulsed system like medical ultrasound imaging, the available data from which $\mat{\hat{R}}$ can be estimated is quite limited.
@@
 Where, $T_p$ is the pulse length in seconds and $T_s$ is the sampling period.
+For baseband data, one pulse length ($1.5\lambda$) is around 3 samples ($K=1$).
 
@@
 
-A SM has available a limited amount of near-core memory and registers.
+An SM has available a limited amount of near-core memory and registers.
 Each thread therefore has to be as light on memory as possible and has to perform enough instructions per used byte in order to hide memory latency.
@@
 Further, based on memory dependencies and to hold memory consumption per thread at a minimum, one should recognise that a group of threads should process one output in the final image.
-Continuing with the calculation of one weight vector, the algorithm can be broken down into three main steps that, to some extent, can be further divided into a small set of parallel processes.
+Continuing with the calculation of one weight vector, the algorithm can be broken down into three main steps that, to some extent, can be divided further into a set of parallel processes.
 That is, for each data vector $\vec{x}^\Delta$ in the image:
@@
 
-From (\ref{eq:R}) it is obvious that calculation of one element in $\mat{\hat{R}}$ is independent of all the other elements.
-A natural level of granularity would then be to assign one thread to each element of $\mat{\hat{R}}$ as in \cite{Chen2011}.
+From (\ref{eq:R}) it is obvious that calculation of one element in $\mat{\hat{R}}$ is independent of all other elements.
+A natural level of granularity would then be to assign one thread to each element of $\mat{\hat{R}}$ as done by \cite{Chen2011}.
 This will minimize global reads and writes needed per thread, but unfortunately the maximum number of threads per block puts an upper limit on the subarray length, $L_{max}=32$, since we want to avoid block-to-block synchronization.
@@
 &+ \sum_{n'=n-K}^{n+K} (x_{i+N_L-1}^{\Delta}[n']x_{j+N_L-1}^{\Delta}[n']^* \nonumber \\
-&- x_{i-1}^{\Delta}[n']x_{j-1}^{\Delta}[n']^H) \label{eq:sliding}
+&- x_{i-1}^{\Delta}[n']x_{j-1}^{\Delta}[n']^H). \label{eq:sliding}
 \end{align}
@@
 The transformation must be done per subarray, yielding subarrays that no longer overlap in space.
-If the transformation is applied before the covariance matrix is constructing, the amount of input data is change from $M$ to $N_LN_b$ elements per sample, however all remaining steps will get lower complexity.
+If the transformation is applied before the covariance matrix is constructed, the amount of input data is change from $M$ to $N_LN_b$ elements per sample, however all remaining steps will get lower complexity if $N_b < L$.
 
@@
 With SBS we can therefore select to have one thread per beam (index $k$) that calculates this beam for all subarrays.
-However, since the number of beams usually is small we will end up with a large memory per thread ratio ($M \gg N_b$).
+However, since the number of beams usually is small ($N_b \ll M$) we will end up with a large memory per thread ratio.
 Using one thread per beam is therefore not feasible for small $N_b$ large $M$ configurations.
@@
 Thus the average energy per beam is higher than the average energy per element, if the system uses focused transmit beams.
-This means that less diagonal loading has to be applied in beamspace than in element space ($d_{BS} \ll d_{element}$).
+This means that less diagonal loading (smaller $d$) has to be applied in beamspace than in element space.
 
@@
 \section{Results}\label{sec:res}
+
 Describe the computer system (CPU, GPU...). Describe the data that are processed (field II, Channel data from GE Vingmed).
@@
 \label{fig:benchUS64}}}
-\caption{(Change into bar plot?) Benchmark of GPU Capon beamforming for a cardiac image covering a 70 degrees sector.
+\caption{Benchmark of GPU Capon beamforming for a cardiac image covering a $70\degree$ wide $15.4$cm deep sector.
 Execution time is plotted as a function of subarray length $L$.
 The plot shows execution times for the three steps listed in Section \ref{sec:meth}.
-Calculating covariance matrices in red, solver in green, and the final beamformer sum in blue.
-For each step, there is one line for three different choices of temporal averaging, $N_{avg} \in [0, 1, 2]$.
-(a) A 32 element array.
-(b) A 64 element array.}
+Calculating covariance matrices with three different choices of temporal averaging, solver, and the final beamformer sum.
+For larger arrays, more transmit beams have to be emitted for sufficient lateral sampling.
+(a) Shows the benchmark for a 32 element array.
+(b) Shows the benchmark for a 64 element array.}
 \label{fig:bench}
@@
 \hfil
-\subfloat[]{\includegraphics[width=3.5in]{gfx/benchmark_bar_bs_M=64_Nx=80_Ny=400_Nb=3.png}%
+\subfloat[]{\includegraphics[width=3.5in]{gfx/benchmark_bar_bs_M=96_Nx=120_Ny=400_Nb=3.png}%
 \label{fig:benchBS5}}}
-\caption{(TODO: Add the right figure.
-Change it into a bar plot?) Benchmark of GPU Beamspace Capon beamforming for a cardiac image covering a 70 degrees sector.
-Execution time is plotted as a function of number of elements $M$.
-The plot shows execution times for the three steps listed in Section \ref{sec:meth}.
-Calculating covariance matrices in red, solver in green, and the final beamformer sum in blue.
-For each step, there is one line for three different choices of temporal averaging, $N_{avg} \in [0, 1, 2]$.
-(a) 3 beams.
-(b) 5 beams.}
+\caption{Benchmark of GPU beamspace Capon beamforming for a cardiac image covering a $70 \degree$ and $15.4$cm deep sector.
+Execution time is plotted as a function of subarray length $L$.
+(a) Shows the benchmark for a 64 elements array.
+(b) Shows the benchmark for a 96 elements array.}
 \label{fig:benchBS}
@@
 \begin{figure*}[!t]
-\centerline{\subfloat[]{\includegraphics[width=2in]{gfx/mv_phantom.png} \label{fig:phantomDAS}}
+\centerline{\subfloat[]{\includegraphics[width=2in]{gfx/das.png} \label{fig:phantomDAS}}
 \hfil
-\subfloat[]{\includegraphics[width=2in]{gfx/mv_phantom.png} \label{fig:phantomMV}}
+\subfloat[]{\includegraphics[width=2in]{gfx/capon.png} \label{fig:phantomMV}}
 \hfil
-\subfloat[]{\includegraphics[width=2in]{gfx/mv_phantom.png} \label{fig:phantomBS}}}
-\caption{(TODO: Add the right figures) Have graph showing cyst contrast as a function of different parameters instead?
-Cyst and point targets phantom a) Delay-and-sum, uniform weights.
-b) Capon ($M=?, L=?, N=?$) c) Beamspace Capon ($M=?, L=?, N=?, N_b=?$).}
-\label{fig:bench}
+\subfloat[]{\includegraphics[width=2in]{gfx/capon_bs.png} \label{fig:phantomBS}}}
+\caption{Simulated phantom containing one bright and one dark "cyst" and two point targets in speckle.
+The simulation was done using Field II with a 2.5MHz, $\lambda/2$ and 64 elements array.
+a) Delay-and-sum with uniform weights.
+b) Capon ($L=32, K=2, d=0.1$) c) Beamspace Capon ($L=32, K=2, N_b=3, d=0.01$).}
+\label{fig:phantom}
 \end{figure*}
@@
 \begin{figure*}[!t]
-\centerline{\subfloat[]{\includegraphics[width=2in]{gfx/us_image.png} \label{fig:phantomDAS}}
+\centerline{\subfloat[]{\includegraphics[width=2in]{gfx/das_invivo.png} \label{fig:invivoDAS}}
 \hfil
-\subfloat[]{\includegraphics[width=2in]{gfx/us_image.png} \label{fig:phantomMV}}
+\subfloat[]{\includegraphics[width=2in]{gfx/capon_invivo.png} \label{fig:invivoMV}}
 \hfil
-\subfloat[]{\includegraphics[width=2in]{gfx/us_image.png} \label{fig:phantomBS}}}
-\caption{(TODO: Add the right figures) Have graph showing cavity-wall contrast as a function of parameters instead?
-In-vivo cardiac ultrasound image a) Delay-and-sum, uniform weights.
-b) Capon ($M=?, L=?, N=?$) c) Beamspace Capon ($M=?, L=?, N=?, N_b=?$).}
-\label{fig:bench}
+\subfloat[]{\includegraphics[width=2in]{gfx/capon_bs_invivo.png} \label{fig:invivoBS}}}
+\caption{\textit{In-vivo} harmonic cardiac ultrasound image acquired using a 3.5MHz, 64 elements phased array.
+a) Delay-and-sum with uniform weights.
+b) Capon ($L=32, K=2, d=0.1$) c) Beamspace Capon ($L=32, K=2, N_b=3, d=0.01$).}
+\label{fig:invivo}
 \end{figure*}
+
+Fig.
+\ref{fig:bench} shows the results from benchmarking the implementation described in Section \ref{sec:meth} on an Nvidia Quadro 6000 graphics card (14 SMs, 1 TFlops single precision, 6 GB global memory, and 144 GB/s memory bandwidth).
+The problem sizes are calculated based on 32 and 64 element arrays with center frequency 2.5 MHz.
+By assuming a system bandwidth of 80\% we need a complex sampling frequency of 2 MHz.
+If we image down to a depth of 15 cm we get approximately 400 samples per range line.
+To correctly sample a $70\degree$ sector with 64 and 32 element arrays ($\lambda/2$ pitch), we need at least 80 and 40 beams, respectively.
+The presented execution times do not include transfer of data from CPU-side to the GPU.
+What is included are all calculations, memory transfers from global to near-core GPU memory, and near core memory access.
+
+There are two videos attached to this paper. One cardiac loop, 
+%(add ref to video)
+and one simulation, both 
+%(add ref to videos) 
+processed with the following Capon beamformer settings: $L = 32$, $K = 2$, $d=0.01$.
+These settings have previously been shown to provide high resolution while maintaining DAS-like speckle \cite{Synnevag2007a} in ultrasound images.
+Cardiac data was acquired using a 64 element and 3.5 MHz phased array.
+Simulations were done using a 64 element 2.5 MHz phased array.
+For the simulation, 16x lateral over-sampling has been used, in order to avoid under sampling of the two bright point scatterers.
+The cardiac recording only has $30\%$ lateral over sampling.
 
@@
 Discuss results presented in Section \ref{sec:res}.
+
+From Fig.
+\ref{fig:benchUS64} we see that for $L = M/2 = 32$ and $K=1$, total execution time is 100 ms (10 fps).
+Taking into account that we use complex data, but less samples per image, this is similar to the performance reported by Chen et al.
+\cite{Chen2011}.
+For the 32 element array in Fig.
+(\ref{fig:benchUS32}), execution time is reduced to 11.5 ms (87 fps) using the same settings ($L=M/2, K = 1$).
+Doubling this number can be interpreted as pre-beamforming a 64 element array down to 32.
+The frame rate for a 64 element array is then increased from 10 to 44 fps.
+Note that lateral oversampling will increase the total execution time with a factor equal the amount of oversampling.
+
+A major part of the total execution time comes from solving systems of linear equations.
+Further research should therefore focus on making this contribution smaller.
+This can be done either by implementing faster solvers, or reducing the sample covariance matrix size in a novel way \cite{Nilsen2009}.
+%When doing such a reduction, including the pre-beamforming described above, one should investigate the loss of contrast and resolution compared with full Capon beamforming.
+As for now, we have shown that the Capon beamformer can be implemented in a GPU framework.
+And if we as in \cite{Chen2011} define real-time as 10 fps, all execution times  in Fig.
+(\ref{fig:bench}) are real time.
+If we on the other hand define real-time as the time it takes to acquire a given image, the image formed using a 32 and 64 element array has a real time requirement of 125 fps and 62.5 fps respectively.
+There is some distance to cover before that goal is reached for all values of $L \le M/2$.
+However we see that real time, by the last definition, is possible for $M=64, L < 16$, and for $M = 32, L < 13$.
+Another option to reach real-time execution is to use several GPUs simultaneously.
+Note that multi-line acquisition will increase the frame rate requirement accordingly.
+
+In the attached video of a simulated phantom one can see that Capon beamforming highly resolves moving point targets.
+The beamformer is shift invariant as long as the lateral sampling is increased.
+No major improvement in contrast is observed.
+In the cardiac recording, structures are thinner when the Capon beamformer is applied, however we observe a loss in contrast due to signal cancellation.
+Future work should focus on solving this issue.
 

### 2012-12-12 HRUIonGPU/gfx/all_plots.py

@@ -106,11 +106,22 @@ def barBenchmarkGPU(filename, M, Nx, Ny, fromL, toL):
    
    cax2 = ax.bar(x_range + 1/3.0, yavg2[:,0], width/3.0, bottom=(yavg1[:,1] + yavg1[:,2]), color='0.75', hatch='/')
    cax3 = ax.bar(x_range + 2/3.0, yavg3[:,0], width/3.0, bottom=(yavg1[:,1] + yavg1[:,2]), color='0.75', hatch='o')
+   
+   # calc acquisition rate for the given number of beams if depth equals 15.4cm.
+   acq_speed = 0.2*Nx
+   cax6 = ax.plot([0, max_L+1], [acq_speed, acq_speed], '-', color='0.2')
 
    #cax = ax.plot([0,0.001],[0,0], 'k')
    #cax = ax.plot([0,0.001],[0,0], ':k')
    #cax = ax.plot([0,0.001],[0,0], '--k')
-   legend((cax1[0], cax2[0], cax3[0], cax4[0], cax5[0]),('Calculate covariance matrices (K=0)', 'Calculate covariance matrices (K=1)', 'Calculate covariance matrices (K=2)', 'Nvidia solver','Calculate beamformer output'), loc='upper left')
+   legend((cax1[0], cax2[0], cax3[0], cax4[0], cax5[0], cax6[0]),
+          ('Calculate covariance matrices (K=0)', 
+           'Calculate covariance matrices (K=1)', 
+           'Calculate covariance matrices (K=2)', 
+           'Nvidia solver',
+           'Calculate beamformer output',
+           'Real time requirement'), 
+          loc='upper left')
    #legend(('Calc covariance matrices','Nvidia solver','Calc beamformer output','Total', 'K = 0','K = 1','K = 2'), loc='upper left')
 #   gca().add_artist(l1)
    
@@ -160,12 +171,19 @@ def barBeamspaceBenchmarkGPU(filename, M, Nb, Nx, Ny, fromL, toL):
    width = 1
    x_range = arange(min_L,max_L+1)
    
+   K = 2
+   yavg = yavg3
+   
    fig = figure()
    ax = fig.add_subplot(1,1,1)
-   cax4 = ax.bar(x_range, yavg1[:,3], width, color='0.25')
[…]

### 2012-12-12 HRUIonGPU/gfx/make_plots.py

@@ -9,3 +9,4 @@ bar('timer_values_64_80_400.txt', 64, 80, 400, 2, 32)
 bar('timer_values_32_40_400.txt', 32, 40, 400, 2, 16)
 
 barBS('timer_values_M=64_Nx=80_Ny=400_L=2-32_Nb=3_slidingBS.txt', 64, 3, 80, 400, 3, 32)
+barBS('timer_values_M=96_Nx=120_Ny=400_L=3-48_Nb=3_slidingBS.txt', 96, 3, 120, 400, 3, 48)

### 2012-12-14 HRUIonGPU/jrnl_article.tex
back on writing for uffc

@@
 \begin{figure*}[!t]
-\centerline{\subfloat[]{\includegraphics[width=2in]{gfx/das.png} \label{fig:phantomDAS}}
+\centerline{\subfloat[]{\includegraphics[width=2.3in]{gfx/das.png} \label{fig:phantomDAS}}
 \hfil
-\subfloat[]{\includegraphics[width=2in]{gfx/capon.png} \label{fig:phantomMV}}
+\subfloat[]{\includegraphics[width=2.3in]{gfx/capon.png} \label{fig:phantomMV}}
 \hfil
-\subfloat[]{\includegraphics[width=2in]{gfx/capon_bs.png} \label{fig:phantomBS}}}
+\subfloat[]{\includegraphics[width=2.3in]{gfx/capon_bs.png} \label{fig:phantomBS}}}
 \caption{Simulated phantom containing one bright and one dark "cyst" and two point targets in speckle.
-The simulation was done using Field II with a 2.5MHz, $\lambda/2$ and 64 elements array.
+The simulation was done using Field II with a 2.5MHz, $\lambda/2$ and 64 elements phased array.
+Dynamic range is -30dB to 30dB with mean value represented as 0dB.
 a) Delay-and-sum with uniform weights.
@@
 \begin{figure*}[!t]
-\centerline{\subfloat[]{\includegraphics[width=2in]{gfx/das_invivo.png} \label{fig:invivoDAS}}
+\centerline{\subfloat[]{\includegraphics[width=2.3in]{gfx/das_invivo.png} \label{fig:invivoDAS}}
 \hfil
-\subfloat[]{\includegraphics[width=2in]{gfx/capon_invivo.png} \label{fig:invivoMV}}
+\subfloat[]{\includegraphics[width=2.3in]{gfx/capon_invivo.png} \label{fig:invivoMV}}
 \hfil
-\subfloat[]{\includegraphics[width=2in]{gfx/capon_bs_invivo.png} \label{fig:invivoBS}}}
+\subfloat[]{\includegraphics[width=2.3in]{gfx/capon_bs_invivo.png} \label{fig:invivoBS}}}
 \caption{\textit{In-vivo} harmonic cardiac ultrasound image acquired using a 3.5MHz, 64 elements phased array.
+Number of receive beams for the $70\degree$ sector is 98.
+Dynamic range is -20dB to 20dB with mean value represented as 0dB.
 a) Delay-and-sum with uniform weights.

### 2012-12-17 HRUIonGPU/jrnl_article.tex
done some writing on the uffc article

@@
 In this paper we introduce an implementation of the Capon beamformer exhibiting real-time performance when applied in a typical cardiac ultrasound imaging setting.
-To accomplish this task we make use of the massive, and easily accessed, parallel processing power found in modern Graphics Processing Units (GPUs), combined with novel methods to reduce the computational complexity as the problem size grows.
+To accomplish this task we make use of the massive parallel processing power found in modern Graphics Processing Units (GPUs), combined with novel methods to reduce the computational complexity as the problem size grows.
 
@@
 All results in this paper is reported with the use of baseband data.
+Meaning that the sampling frequency is close to the system bandwidth.
 
@@
 \begin{align}\label{eq:b}
-\vec{b}[n] = \mat{\hat{R}}[n]^{-1}\vec{a},
-\end{align}
-where the steering vector $\vec{a} = \vec{1}_L = [1_0, 1_1, ..., 1_{L-1}]^T$ since $\vec{x}^\Delta$ is pre-delayed. Finally the Capon weight vector is formed from (\ref{eq:b}) as
+\vec{b}[n] = \mat{\hat{R}}[n]^{-1}\vec{a} \in \mathbb{C}^L,
+\end{align}
+where the steering vector $\vec{a} = \vec{1}_L = [1_0, 1_1, ..., 1_{L-1}]^T$ since $\vec{x}^\Delta$ is pre-delayed.
+
+Finally the Capon weight vector is formed from (\ref{eq:b}) as
 \begin{align}\label{eq:w}
-\vec{w}[n] = \frac{\vec{b}[n]}{\vec{1}^H\vec{b}[n]},
-\end{align}
-and the beamformer output is calculated by combining all $N_L$ subarrays weighted with the same set of adaptive weights.
+\vec{w}[n] = \frac{\vec{b}[n]}{\vec{1}^H\vec{b}[n]} \in \mathbb{C}^L,
+\end{align}
+and the beamformer output is calculated by combining all $N_L$ subarrays weighted with the same set of adaptive weights. Formally stated as
 \begin{align}
@@
 %\end{enumerate}
-The computational complexity of these three steps is $O(L^3)$, $O(L^3)$, and $O(L)$ respectively.
+The computational complexity of these three steps is $O(N_KN_LL^2)$, $O(L^3)$, and $O(L)$ respectively.
 As $L$ increases, the first two steps quickly become unmanageable, and real-time performance can only be achieved by exploring novel solutions to reduce complexity.
@@
 However, since $N_K$ typically is small, and we have a limited amount of memory available per thread, we have chosen to focus on sliding across subarrays only (\ref{eq:sliding}).
-In that case, the complexity of calculating $\mat{\hat{R}}$ is reduced to $O(L^2)$.
-Notice how sliding will break independence along the sliding dimension.
+In that case, the complexity of calculating $\mat{\hat{R}}$ is reduced to $O(N_KL^2)$.
+Note that sliding will break independence along the sliding dimension.
 With our choice of calculating $\mat{\hat{R}}$, calculations are still independent across time, but not along diagonals in $\mat{\hat{R}}$.
@@
 However, there is two arguments for doing solving on the GPU.
-First the CPU, which might already runs a lot of algorithms, is offloaded.
+First the CPU, which might already runs a lot of ultrasound related algorithms, is offloaded.
 Second the transfer of several thousand covariance matrices from CPU-side to GPU-side is spared.
 
-For the results presented in this paper, we have used an unreleased GPU implementation of Gauss-Jordan (GJ) elimination (by Nvidia) to solve $\mat{\hat{R}}\vec{b} = \vec{1}_L$, where $\vec{b} = \mat{\hat{R}}^{-1}\vec{1}_L$.
-%For smaller systems, $L < 10$, a direct solver using one thread per system is preferable.
+For the results presented in this paper, we have used an GPU implementation of Gauss-Jordan (GJ) elimination made by Nvidia to solve $\mat{\hat{R}}\vec{b} = \vec{1}_L$, where $\vec{b} = \mat{\hat{R}}^{-1}\vec{1}_L$.
+The solver is open for registered CUDA developers.%For smaller systems, $L < 10$, a direct solver using one thread per system is preferable.
 For this purpose we have implemented Cramer's rule for $2 \times 2$, $3 \times 3$ and $4 \times 4$ matrices.
@@
 \begin{align}
-\vec{x}_{BS}[n] &= \mat{B}\vec{x}^{\Delta}[n], \ \ \ \mat{B} \in C^{M,M}\\
+\vec{x}_{BS}[n] &= \mat{B}\vec{x}^{\Delta}[n], \ \ \ \mat{B} \in \mathbb{C}^{N,M}\\
 [\mat{B}]_{p,q} &= \frac{1}{\sqrt{M}}e^{-j 2 \pi p q / M}.
@@
 The beamspace version of the Capon beamformer is derived by replacing the covariance matrix in (\ref{eq:w}) with the beamspace covariance matrix $\mat{\hat{R}}_{BS} = f(\vec{x}_{BS}) = f(\mat{B}\vec{x})$, where $\mat{B} \in \mathbb{C}^{N_b,L}$, and $N_b$ is the number of selected beams.
-The transformation must be done per subarray, yielding subarrays that no longer overlap in space.
-If the transformation is applied before the covariance matrix is constructed, the amount of input data is change from $M$ to $N_LN_b$ elements per sample, however all remaining steps will get lower complexity if $N_b < L$.
+The transformation must be performed per subarray, yielding subarrays that no longer overlap in space.
+Therefore, if the transformation is applied before the covariance matrix is constructed, the amount of input data is change from $M$ to $N_LN_b$ elements per sample, however all remaining steps will get lower complexity when $N_b < L$.
 
@@
 \begin{align}
-\mat{\hat{R}}_{BS} &= \mat{B}\mat{\hat{R}}\mat{B}^H
+\mat{\hat{R}}_{BS} &= \mat{B}\mat{\hat{R}}\mat{B}^H.
 \end{align}
 The pros using this approach is that minor changes has to be made to the pipeline described in Section \ref{sec:meth}, but it will only speed up the solving step if the solution $\vec{b}_{BS}$ is transformed back to element space.
-In this paper we will describe how the first approach has been implemented on the GPU.
+In this paper we will focus on the first approach, and how it has been implemented on the GPU.
 %By this approach we go from solving $\mat{\hat{R}}\vec{b} = \vec{1}_L$, which is an $L \times L$ system, to solving the $N_b \times N_b$ system $\mat{\hat{R}}_{BS}\vec{b}_{BS} = (\sqrt{L}\vec{e}_1)$.
 Now, one can either choose to transform the subarray sums, as in (\ref{eq:subsum}), to beamspace or transform the result, $\vec{b}_{BS}$, back to element space.
-In both cases solving has to be performed against a transformed steering vector
+In both cases solving has to be performed against a beamspace steering vector
 \begin{align}
@@
 \end{align}
-where $\vec{e}_i$ has value one in the i$^{th}$ position and is zero elsewhere.
+where $\vec{e}_i$ has value one in the i$^{th}$ position and zero elsewhere.
 
@@
 This means that we no longer can exploit spatial overlap when calculating the sample covariance matrix, but since $N_b$ is small we can now allow to have one thread per element in $\mat{\hat{R}}_{BS}$.
-A kernel similar to the one described by Chen et al.
-\cite{Chen2011}, has therefore been implemented.
+A kernel similar to the one described in \cite{Chen2011}, has therefore been implemented.
 For small $N_b$ values, several covariance matrices are calculated per thread block to achieve high occupancy of the SMs.
@@
 \begin{align}
-\vec{w}_{BS} = \frac{\vec{b}_{BS}}{\sqrt{L}\vec{e}_1\vec{b}_{BS}}
+\vec{w}_{BS} = \frac{\vec{b}_{BS}}{\sqrt{L}\vec{e}_1\vec{b}_{BS}} \in \mathbb{C}^{N_b}
 \end{align}  
@@
 This means that less diagonal loading (smaller $d$) has to be applied in beamspace than in element space.
+
+To summarize, by transforming data to beamspace we reduce the complexity of all the subsequent steps.
+Calculating the covariance matrix is now $O(N_KN_LN_b^2)$, the inversion step is $O(N_b^3)$, and calculating weights and the beamformer output are both $O(N_b)$.
+The added complexity of SBS is $O(N_bL)$.
 
@@
 
-\section{Results}\label{sec:res}
-
-Describe the computer system (CPU, GPU...). Describe the data that are processed (field II, Channel data from GE Vingmed).
+\section{Benchmarks}\label{sec:res}
+
+%Describe the computer system (CPU, GPU...). Describe the data that are processed (field II, Channel data from GE Vingmed).
 % stolpediagram, med de ulike metodenes kjøretid oppehverandre.
@@
 \label{fig:benchUS64}}}
-\caption{Benchmark of GPU Capon beamforming for a cardiac image covering a $70\degree$ wide $15.4$cm deep sector.
+\caption{Benchmark of GPU Capon beamforming for a cardiac image covering a $70\degree$ wide $15$cm deep sector.
 Execution time is plotted as a function of subarray length $L$.
@@
 \label{fig:benchBS5}}}
-\caption{Benchmark of GPU beamspace Capon beamforming for a cardiac image covering a $70 \degree$ and $15.4$cm deep sector.
+\caption{(Move BS trans below calcR) Benchmark of GPU beamspace Capon beamforming for a cardiac image covering a $70 \degree$ and $15$cm deep sector.
 Execution time is plotted as a function of subarray length $L$.
@@
 Fig.
-\ref{fig:bench} shows the results from benchmarking the implementation described in Section \ref{sec:meth} on an Nvidia Quadro 6000 graphics card (14 SMs, 1 TFlops single precision, 6 GB global memory, and 144 GB/s memory bandwidth).
+\ref{fig:bench} shows the results from benchmarking the implementation described in Section \ref{sec:meth} on an Nvidia Quadro 6000 graphics card (14 SMs, 1 TFlops theoretical single precision, 6 GB global memory, and 144 GB/s memory bandwidth).
 The problem sizes are calculated based on 32 and 64 element arrays with center frequency 2.5 MHz.
@@
 To correctly sample a $70\degree$ sector with 64 and 32 element arrays ($\lambda/2$ pitch), we need at least 80 and 40 beams, respectively.
-The presented execution times do not include transfer of data from CPU-side to the GPU.
+The presented execution times do not include transfer of data from CPU-side to the GPU, since data are assumed to be pre-delayed and available in GPU memory.
 What is included are all calculations, memory transfers from global to near-core GPU memory, and near core memory access.
+
+In Fig.
+\ref{fig:benchBS} we present similar benchmarks for the beamspace version of the Capon beamformer, but now for a 64 and 96 elements probe.
+For the latter, number of transmit beams is increased to 120.
+For all benchmarks there is a line indicating the time it takes to acquire the underlying image.
+Hence, below this line we are capable of real time processing.
+
+\section{Simulated phantom}
+Present images, both simulated and in-vivo, processed using beamspace Capon. Comment on the differences between the two and on how much they deviates from the full method.
+
+Present images before and after Capon weights has been applied.
+Comment on the effect of selecting different parameters.
+Present graph showing executing times for Cython-Capon v.s.
+CUDA-Capon for different choices of parameters.
+Comment on different choices of parameters change the running times.
+Temporal averaging for instance, lowers the efficiency since $2Y_{avg}$ more operations has to be performed per $\mat{\hat{R}}$
+
+\section{\textit{Ion-vivo} cardiac data}
+How has the data been acquired?
 
@@
 
-Present images before and after Capon weights has been applied.
-Comment on the effect of selecting different parameters.
-Present graph showing executing times for Cython-Capon v.s.
-CUDA-Capon for different choices of parameters.
-Comment on different choices of parameters change the running times.
-Temporal averaging for instance, lowers the efficiency since $2Y_{avg}$ more operations has to be performed per $\mat{\hat{R}}$
-
-Present images, both simulated and in-vivo, processed using beamspace Capon. Comment on the differences between the two and on how much they deviates from the full method.
-%Show beampatterns for the two compared with the full method. 
-Show executing times for the two methods compared with the full method on the GPU.
-
-Describe the base band (IQ-demodulation) process that the data has been undergoing.
-
-How has the data been acquired?
-
-Ta med en serie med bilde. Må vise tydelig at det gir forbedring.
-
+\section{Trading resolution for speed}
 Add a comparison between beamspace and full capon using small L.

### 2012-12-20 HRUIonGPU/jrnl_article.tex
Done some writing

@@
 \end{align}
-We call this the sliding beamspace transformation (SBS), and it is equal to a normalized sliding DFT. For $l=0$ a full computation has to be performed using (\ref{eq:bs_formula}).
+We call this the sliding beamspace transformation (SBS), and it is equal to a normalized sliding DFT \cite{Lyons2003}. For $l=0$ a full computation has to be performed using (\ref{eq:bs_formula}).
 
@@
 \label{fig:benchBS5}}}
-\caption{(Move BS trans below calcR) Benchmark of GPU beamspace Capon beamforming for a cardiac image covering a $70 \degree$ and $15$cm deep sector.
+\caption{Benchmark of GPU beamspace Capon beamforming for a cardiac image covering a $70 \degree$ and $15$cm deep sector.
 Execution time is plotted as a function of subarray length $L$.
@@
 \end{figure*}
+
+Fig.
+\ref{fig:bench} shows the results from benchmarking the implementation described in Section \ref{sec:meth} on an Nvidia Quadro 6000 graphics card (14 SMs, 1 TFlops theoretical single precision, 6 GB global memory, and 144 GB/s memory bandwidth).
+The problem sizes are calculated based on 32 and 64 element phased arrays with center frequency 2.5 MHz.
+By assuming a system bandwidth of 80\% we need a complex sampling frequency of 2 MHz.
+If we image down to a depth of 15 cm we get approximately 400 samples per range line.
+To correctly sample a $70\degree$ sector with 64 and 32 element arrays ($\lambda/2$ pitch), we need at least 80 and 40 beams, respectively.
+The presented execution times do not include transfer of data from CPU-side to the GPU, since data are assumed to be pre-delayed and available in GPU memory.
+What is included are all calculations, memory transfers from global to near-core GPU memory, and near core memory access.
+For $M=64$, $L=32$ and $K=2$ (Fig.
+\ref{fig:benchUS64}), a typical setup for high quality imaging, processing throughput is 10 frames per second.
+
+In Fig.
+\ref{fig:benchBS} we present similar benchmarks for the beamspace version of the Capon beamformer.
+Benchmarks are based on a 64 and 96 elements phased array respectively.
+For the latter, number of transmit beams is increased to 120.
+The figure only present results with use of temporal averaging ($K=2$).
+For $M=96$, $L=48$, $K=2$ and $N_b=3$, processing throughput is 100 frames per second.
+
+For all benchmarks there is a line indicating the time it takes to acquire the underlying image using single line acquisition. Hence, below this line we are capable of real time processing.
+
+\section{Simulated phantom and cardiac data}
 
@@
 \subfloat[]{\includegraphics[width=2.3in]{gfx/capon_bs.png} \label{fig:phantomBS}}}
-\caption{Simulated phantom containing one bright and one dark "cyst" and two point targets in speckle.
-The simulation was done using Field II with a 2.5MHz, $\lambda/2$ and 64 elements phased array.
-Dynamic range is -30dB to 30dB with mean value represented as 0dB.
+\caption{Simulated phantom containing one bright (24 dB) and one dark (-30 dB) circular area and two point targets (40 dB) in speckle.
+The simulation was done using Field II with a 2.5 MHz, $\lambda/2$ and 64 elements phased array.
+Dynamic range is -30 dB to 30 dB with mean value represented as 0 dB.
 a) Delay-and-sum with uniform weights.
@@
 \subfloat[]{\includegraphics[width=2.3in]{gfx/capon_bs_invivo.png} \label{fig:invivoBS}}}
-\caption{\textit{In-vivo} harmonic cardiac ultrasound image acquired using a 3.5MHz, 64 elements phased array.
-Number of receive beams for the $70\degree$ sector is 98.
-Dynamic range is -20dB to 20dB with mean value represented as 0dB.
+\caption{\textit{In-vivo} harmonic cardiac ultrasound image acquired using a 3.5 MHz, 64 elements phased array.
+Number of receive beams for the $70\degree$ sector is 98, and number of samples in range is 610.
+Dynamic range is -20 dB to 20 dB with mean value represented as 0 dB.
 a) Delay-and-sum with uniform weights.
@@
 
+\begin{figure*}[!t]
+\centerline{
+\subfloat[]{\includegraphics[width=3in]{gfx/simulation_slice.png} \label{fig:simulationSlice}}
+\hfil
+\subfloat[]{\includegraphics[width=3in]{gfx/invivo_slice.png} \label{fig:invivoSlice}}
+}
+\caption{Lateral slices taken from the images in Fig.
+\ref{fig:phantom} and Fig.
+\ref{fig:invivo}.
+a) From Fig.
+\ref{fig:phantom} at 93 mm range.
+Straight through the two bright point scatterers.
+b) From Fig.
+\ref{fig:invivo} at 96 mm range, going through the tip of the mitral valve.}
+\label{fig:slices}
+\end{figure*}
+
 Fig.
-\ref{fig:bench} shows the results from benchmarking the implementation described in Section \ref{sec:meth} on an Nvidia Quadro 6000 graphics card (14 SMs, 1 TFlops theoretical single precision, 6 GB global memory, and 144 GB/s memory bandwidth).
-The problem sizes are calculated based on 32 and 64 element arrays with center frequency 2.5 MHz.
-By assuming a system bandwidth of 80\% we need a complex sampling frequency of 2 MHz.
-If we image down to a depth of 15 cm we get approximately 400 samples per range line.
-To correctly sample a $70\degree$ sector with 64 and 32 element arrays ($\lambda/2$ pitch), we need at least 80 and 40 beams, respectively.
-The presented execution times do not include transfer of data from CPU-side to the GPU, since data are assumed to be pre-delayed and available in GPU memory.
-What is included are all calculations, memory transfers from global to near-core GPU memory, and near core memory access.
-
-In Fig.
-\ref{fig:benchBS} we present similar benchmarks for the beamspace version of the Capon beamformer, but now for a 64 and 96 elements probe.
-For the latter, number of transmit beams is increased to 120.
-For all benchmarks there is a line indicating the time it takes to acquire the underlying image.
-Hence, below this line we are capable of real time processing.
-
-\section{Simulated phantom}
-Present images, both simulated and in-vivo, processed using beamspace Capon. Comment on the differences between the two and on how much they deviates from the full method.
-
-Present images before and after Capon weights has been applied.
-Comment on the effect of selecting different parameters.
-Present graph showing executing times for Cython-Capon v.s.
-CUDA-Capon for different choices of parameters.
-Comment on different choices of parameters change the running times.
-Temporal averaging for instance, lowers the efficiency since $2Y_{avg}$ more operations has to be performed per $\mat{\hat{R}}$
-
-\section{\textit{Ion-vivo} cardiac data}
-How has the data been acquired?
-
-There are two videos attached to this paper. One cardiac loop, 
-%(add ref to video)
-and one simulation, both 
-%(add ref to videos) 
-processed with the following Capon beamformer settings: $L = 32$, $K = 2$, $d=0.01$.
+\ref{fig:phantom} presents a phantom simulated using Field II \cite{Jensen1992}.
+The simulation was performed using a 2.5 MHz, $\lambda/2$ pitch, 10 $\mu$m kerf, 50\% bandwidth and 10 mm high phased array.
+The transducer was exited with a $3\lambda$ long pulse and the transmit focus is at 80 mm.
+The phantom contains two circular areas, one bright at 24 dB and one dark at -30 dB, and two point targets at 40 dB.
+Speckle level is at 0 dB.
+The resulting image has been processed both using DAS with uniform weights, Capon and BS Capon.
+The presented image is the first frame in a series of images where motion is imposed on both circles, point targets and the probe.
+This simulation is presented in an attached video.
+
+Fig.
+\ref{fig:invivo} shows an \textit{in-vivo} image of the left ventricle.
+The image was acquired using a experimental scanner at MI-Lab in Trondheim using a 3.5 MHz phased array in harmonic mode.
+The image is extracted from a cardiac cycle, that can be viewed in its entirety in the attached videos.
+
+Both data sets have been processed with the following Capon beamformer settings: $L = 32$, $K = 2$, $d=0.1$.
 These settings have previously been shown to provide high resolution while maintaining DAS-like speckle \cite{Synnevag2007a} in ultrasound images.
-Cardiac data was acquired using a 64 element and 3.5 MHz phased array.
-Simulations were done using a 64 element 2.5 MHz phased array.
-For the simulation, 16x lateral over-sampling has been used, in order to avoid under sampling of the two bright point scatterers.
-The cardiac recording only has $30\%$ lateral over sampling.
+For the beamspace processing the following parameters have been used: $L=32$, $K=2$, $d=0.01$, $N_b=3$.
+
+For the simulation, 16x lateral over-sampling has been used, in order to avoid under sampling of the two bright point scatterers. The cardiac recording only has $30\%$ lateral over sampling.
 
 \section{Trading resolution for speed}
-Add a comparison between beamspace and full capon using small L.
+Add a comparison between beamspace and full capon using small L. Plot system resolution vs $L$ and $N_b$ for the two methods. ???
+Fig \ref{fig:speed_res_trade}.
+
+\begin{figure}
+\centering
+\includegraphics[width=3in]{gfx/speed_res_trade.png}
+\caption{Trade-off between resolution and speed by reducing subarray size or reducing beamspace.}
+\label{fig:speed_res_trade}
+\end{figure}
 
 \section{Discussion}\label{sec:dis}
-Discuss results presented in Section \ref{sec:res}.
-
 From Fig.
-\ref{fig:benchUS64} we see that for $L = M/2 = 32$ and $K=1$, total execution time is 100 ms (10 fps).
+\ref{fig:benchUS64} we see that for $L = M/2 = 32$ and $K=1$, the total execution time is 100 ms (10 fps).
 Taking into account that we use complex data, but less samples per image, this is similar to the performance reported by Chen et al.
@@
 ...
+
+\section{Future work}
+Plane waves...
 

### 2013-01-10 HRUIonGPU/jrnl_article.tex
Altered jrnl_article.tex for one-column view.

--- 2012-12-14 HRUIonGPU/jrnl_article.tex
@@
 %
-\documentclass[journal]{IEEEtran}
+%\documentclass[journal]{IEEEtran}
+
+% Single column review mode:
+\documentclass[12pt,journal,onecolumn]{IEEEtran}
+\newcommand\CLASSINPUTbaselinestretch{2}
+\RequirePackage{calc}
+\RequirePackage{fp}
+
+% Left margin = 1 inch + hoffset + oddsidemargin (or evensidemargin)
+% Adding the 2cm gutter width to the odd/even side margins:
+\setlength\hoffset{0pt}
+\setlength\oddsidemargin{4cm}       % 4cm margins on the left side
+\addtolength\oddsidemargin{-1in}    % Subtract the initial 1 inch
+
+\setlength\textwidth{21cm-8cm}    % A4 width (21cm) minus margins on either side
+
 %
@@
 \renewcommand{\vec}[1]{\mathbf{#1}}
-\newcommand{\degree}{\ensuremath{^\circ}}
 
@@
 %\boldmath
-...
+An amazing abstract goes here. An amazing abstract goes here. An amazing abstract goes here. An amazing abstract goes here.
+An amazing abstract goes here. An amazing abstract goes here. An amazing abstract goes here. An amazing abstract goes here.
+An amazing abstract goes here. An amazing abstract goes here. An amazing abstract goes here. An amazing abstract goes here.
+An amazing abstract goes here. An amazing abstract goes here. An amazing abstract goes here. An amazing abstract goes here.
+An amazing abstract goes here. An amazing abstract goes here. An amazing abstract goes here. An amazing abstract goes here.
+An amazing abstract goes here. An amazing abstract goes here. An amazing abstract goes here. An amazing abstract goes here.
+An amazing abstract goes here. An amazing abstract goes here. An amazing abstract goes here. An amazing abstract goes here.
+An amazing abstract goes here. An amazing abstract goes here. An amazing abstract goes here. An amazing abstract goes here.
+An amazing abstract goes here. An amazing abstract goes here. An amazing abstract goes here. An amazing abstract goes here.
+An amazing abstract goes here. An amazing abstract goes here. An amazing abstract goes here. An amazing abstract goes here.
 \end{abstract}
@@
 %\IEEEPARstart{I}{ntroduction} to Capon adaptive beamforming, ultrasound imaging and GPU/CUDA.
-\IEEEPARstart{C}{apon} beamforming \cite{Capon1969} applied to medical ultrasound imaging has recently shown promising results in several studies \cite{Synnevag2007, Austeng2008, Vignon2008, Viola, Mehdizadeh2012}.
-Some interesting trade-offs in addition to just improved resolution has been introduced, like maintaining image resolution in combination with higher framerates, smaller probes, or deeper penetration \cite{Synnevag2009}.
+\IEEEPARstart{C}{apon} adaptive beamforming \cite{Capon1969} applied to medical ultrasound imaging has recently shown promising results in several studies \cite{Synnevag2007, Austeng2008, Vignon2008, Viola, Mehdizadeh2012}.
+It has been shown that the method introduces some interesting trade-offs in addition to just improved resolution, like maintaining image resolution in combination with higher framerates, smaller probes, or deeper penetration \cite{Synnevag2009}.
 Despite these benefits, widespread use has been limited by the immense computational complexity associated with the method \cite{So2011}.
-In this paper we introduce an implementation of the Capon beamformer exhibiting real-time performance when applied in a typical cardiac ultrasound imaging setting.
+In this paper we introduce the first implementation of the Capon beamformer exhibiting real-time performance when applied in a typical cardiac ultrasound imaging setting.
 To accomplish this task we make use of the massive, and easily accessed, parallel processing power found in modern Graphics Processing Units (GPUs), combined with novel methods to reduce the computational complexity as the problem size grows.
 
-Previous work on parallel implementations of the Capon beamformer have mainly been focused around radar and passive sonar systems using custom made systolic array processors \cite{McWhirter1989, Moonen1993, Sinha2002}.
+Previous work on implementing a parallel Capon beamformer have mainly been focused around radar and passive sonar systems using custom made systolic array processors \cite{McWhirter1989, Moonen1993, Sinha2002}.
+%A systolic array processor is custom made hardware, and even if the computational throughput might be higher for selected problems compared with a GPU, the cost related to development and flexibility is considered much higher.
 For medical ultrasound, Chen et al.
-\cite{Chen2011a, Chen2011} implemented a Capon beamformer on the GPU that produced real apodization weights based on real data.
-Under this constraint the Capon beamformer is not allowed to micro-steer the main lobe, and therefore its full potential is not reached.
-Micro-steering of the main lobe is an important property of the Capon beamformer in order for it to improve structural details and edge definitions throughout an ultrasound image.
-Another aspect of using complex data is that even though the number of arithmetic operations per sample is increased we can sample at the system bandwidth, keeping the number of samples for a full cardiac imaging sector at a minimum.
+\cite{Chen2011a, Chen2011} have implemented a Capon beamformer on the GPU that produced real apodization weights based on real data.
+Under this constraint the Capon beamformer is not allowed to shift the main lobe, and therefore its full potential is not reached.
+Shifting the main lobe is an important property of the Capon beamformer in order for it to improve edge definitions and contrast throughout an ultrasound image.
+Another aspect of using complex data is that even though the number of aritmetic operations per sample is increased we can now sample at the system bandwidth, keeping the number of samples for a full cardiac imaging sector at a minimum.
 All results in this paper is reported with the use of baseband data.
@@
 % denne må utelates i IUS-artikkelen
-In recent work \cite{asen2012}(cite JOE or ECUA?) we have applied a GPU-based Capon beamformer in both an active sonar system and for cardiac ultrasound imaging.
+In recent work (cite JOE) we have applied a GPU-based Capon beamformer in an active sonar system.
 For sonar the real-time requirement is much lower than for medical ultrasound imaging, which means that most common array configurations can be handled using the GPU without taking any additional measures.
-For medical ultrasound imaging, and especially cardiac ultrasound imaging, frame rate requirements increase to a level where even a high-end GPU can not do the processing required for a standard implementation of the Capon beamformer.
-Since the execution time highly depends on the number of elements in the given array, it is important to find ways of reducing computational complexity for high-channel-count systems.
-Different approaches to low-complexity Capon beamforming have therefore recently been proposed \cite{Synnevag2011, Asl2012, Jensen2012, Kim}.
-In this paper we have selected to implement the beamspace transformation, described by Nilsen and Hafizovic \cite{Nilsen2009}, on the GPU to maintain high framerates as the channel count increases.
-The beamspace transformation matches well with the focused transmit beams used in cardiac ultrasound imaging.
+For medical ultrasound imaging, and especially cardiac ultrasound imaging, frame rate requirements increase to a level where even a high-end GPU can not do the processing required for a standard implementation of Capon beamforming.
+Since the execution time for a Capon beamformer depends highly on the number of elements in the given array, it is important to find ways of reducing the computational complexity for high-channel-count systems.
+Different approaches to low-complexity Capon beamforming have recently been proposed \cite{Asl2012, Jensen2012}, and in this paper we have selected to implement the beamspace transformation described by Nilsen and Hafizovic \cite{Nilsen2009} on the GPU to maintain high framerates as the channel count increases.
 %(Referer til JOE-artikkelen, få med beskrivelse av hva som blir anerledes for ultralyd.)
@@
 In the next section we present background information about the Capon beamformer and the GPU computing model.
-In Section \ref{sec:meth} we go in detail about the parallel implementation of the Capon beamformer, before in Section \ref{sec:bs} we introduce the beamspace method and how it has been implemented on the GPU.
+In Section \ref{sec:meth} we go in detail about the parallel implementation of the Capon beamformer, before in Section \ref{sec:bs} we introduce the method applied when the problem size increases.
 Benchmark results and resulting images are presented in Section \ref{sec:res}, and finally we discuss the results in Section \ref{sec:dis}, and draw conclusions in Section \ref{sec:con}.
@@
 
-\subsection{Capon Beamforming}
+%\subsection{Capon Beamforming}
 
@@
 
-The conventional form of array beamforming in medical ultrasound imaging is delay-and-sum (DAS).
-Here, the signal $x_m[n]$ recorded at element number $m \in [0,M-1]$ at time $n$ is delayed to broadside with an appropriate delay $\Delta_m[n]$.
+The conventional form of array beamforming is delay-and-sum (DAS).
+Here, the signal $x_m[n]$ recorded at element number $m \in [0,M-1]$ from time $n = 0$ to $N-1$ is delayed to broadside with an appropriate delay $\Delta_m[n]$.
 The beamformer output, $z[n]$, is then the sum of all $M$ channels weighted with a per-element factor $w_m$,
@@
 \end{align}
-The weight vector or window $\vec{w}$ is usually real, and is applied to trade resolution for a better signal-to-noise ratio (SNR).
-Even if different windows can be used in different areas of the images they have traditionally been predefined.
-
-As shown in Fig.
-\ref{fig:mvbf}, the basic idea of an adaptive beamformer is to form a complex weight vector based on the received data.
-For Capon beamforming the weights are found that minimize the beamformer output power while maintaining unit gain at broadside.
+In conventional beamforming, the weighting $\vec{w}$ is usually real, and is applied to trade resolution for a better signal-to-noise ratio (SNR).
+
+As shown in Figure \ref{fig:mvbf}, the basic idea of an adaptive beamformer is to form a weight vector based on the data at hand.
+For Capon adaptive beamforming the weights are found that minimize the beamformer output power while maintaining unit gain at broadside.
 The result is that interference impinging from other directions than the point of focus is suppressed \cite{Synnevag2007}.
@@
 \begin{figure}
-\centerline{\includegraphics[width=3in]{gfx/beamforming_mv.pdf}}
-\caption{Capon beamforming.
+\centerline{\includegraphics[width=3in]{gfx/beamforming_mv_lowres.png}}
+\caption{Adaptive beamforming.
 Impinging signal is demodulated, and aligned in phase by applying steering and focusing delays $\Delta_m[n]$.
-Adaptive weights, $\vec{w}[n]$, are then calculated based on in-phase data, and finally the output is formed by a Capon-weighted coherent sum.}
+Adaptive weights, $\vec{w}[n]$, are then calculated based on in-phase data, and finally the output is formed by a weighted coherent sum.}
 \label{fig:mvbf}
@@
 
-The Capon beamformer can be summerized in three main steps. First a sample covariance matrix has to be calculated, 
+The Capon weight vector is given by the following expression:
+\begin{align}\label{eq:w}
+\vec{w}[n] = \frac{\mat{\hat{R}}[n]^{-1}\vec{a}}{\vec{a}^H\mat{\hat{R}}[n]^{-1}\vec{a}} = \frac{\vec{b}}{\vec{a}^H\vec{b}},
+\end{align}
+where $\vec{a} = [1_0, 1_1, ..., 1_{M-1}]^T = \vec{1}_M$ when $\vec{x}$ is pre-delayed, $\mat{\hat{R}}$ is the sample covariance matrix calculated from the data vector $\vec{x}^\Delta[n]$, and $\vec{b} = \mat{\hat{R}}[n]^{-1}\vec{a}$.
+%\begin{align}
+%\mat{\hat{R}}[n] = \vec{x}[n]\vec{x}[n]^H.
+%\end{align} 
+%Typically the sample covariance matrix has to be averaged over several samples in order to ensure numerical stability and statistical robustness.
+
+In passive systems, assuming stationary statistics, the sample covariance matrix is typically averaged over a long sequence of samples\cite{Krima}.
+For an active, pulsed system like medical ultrasound imaging, the available data from which $\mat{\hat{R}}$ can be estimated is quite limited.
+The resulting sample covariance matrix might therefore be both inaccurate, and poorly conditioned.
+Another challenge with applying Capon to active systems is the effect of signal cancellation, where high correlation between signal and interference favors weights that cancel out the signal \cite{Reddy1987}.
+In this way, the beamformer output can be too low at broadside, even when there is signal present and the distortionless-response criterion is applied.
+
+In order to get a well-conditioned $\mat{\hat{R}}$, avoid signal cancellation, and to get DAS-like speckle, $\mat{\hat{R}}$ has to be averaged over $L\le M/2$ long subarrays and $N_K \sim T_p/T_s$ time samples \cite{Synnevag2007, Synnevag2007a}.
+Where, $T_p$ is the pulse length in seconds and $T_s$ is the sampling period.
+We therefore calculate
 \begin{align}
-\mat{\breve{R}}[n] = \frac{1}{N_LN_K}\sum_{n'=n-K}^{n+K} \sum_{l=0}^{N_L-1} (\vec{x}_l^{\Delta}[n'])(\vec{x}_l^{\Delta}[n'])^H,\label{eq:R}
+\mat{\breve{R}}[n] = \frac{1}{N_LN_K}\sum_{n'=n-K}^{n+K} \sum_{l=0}^{N_L-1} \vec{x}_l^{\Delta}[n']\vec{x}_l^{\Delta}[n']^H,\label{eq:R}
 \end{align}
-where  $N_K = 2K + 1$ is the number of samples included in time, $\vec{x}_l^{\Delta}$ is the $l^\text{th}$ subarray of length $L$ $[x_l^{\Delta}[n], \dotso, x_{l+L-1}^{\Delta}[n]]$ and $N_L = M-L+1$ is the number of subarrays.
-Note that the covariance matrix dimension is $L \times L$.
-The covariance matrix is then loaded with a diagonal factor $\epsilon$ to ensure numerical stability and increased robustness,
+where  $N_K = 2K + 1$ is the number of samples included in time, $N_L = M-L+1$ is the number of subarrays, and $\vec{x}_l^{\Delta}$ is the $l^\text{th}$ subarray $[x_l^{\Delta}[n], \dotso, x_{l+L-1}^{\Delta}[n]]$.
+The dimension of the covariance matrix after this smoothing is $L \times L$.
+Hence, the size of $\vec{w}[n]$ and $\vec{a}$ in (\ref{eq:w}) has to be $L \times 1$ when the covariance matrix in (\ref{eq:R}) is used.
+Finally, the covariance matrix is loaded with a diagonal factor $\epsilon$ to ensure numerical stability, 
 \begin{align}\label{eq:diag}
-\mat{\hat{R}}[n] =  f(\vec{x}^{\Delta}[n]) = \mat{\breve{R}}[n] + \epsilon[n]\mat{I}.
+\mat{\hat{R}} =  f(\vec{x}^{\Delta}[n]) = \mat{\breve{R}} + \epsilon\mat{I}.
 \end{align}
@@
 \begin{align}\label{eq:diag_adapt}
-\epsilon[n] &= d \times \frac{tr\{\mat{\breve{R}}[n]\}}{L}
+\epsilon &= d \times tr\{\mat{\breve{R}}\}, \\
+d &= 1/L
 \end{align}
-has been applied in much of the resent literature on Capon beamforming for medical ultrasound imaging \cite{Synnevag2007, Nilsen2009, Wang2009, Mehdizadeh2012}.
-This form of adaptive diagonal loading has also been applied to HF and VHF antenna processing \cite{Featherstone1997b}.
-
-Second, a linear system of equations has to be solved
-\begin{align}\label{eq:b}
-\vec{b}[n] = \mat{\hat{R}}[n]^{-1}\vec{a},
-\end{align}
-where the steering vector $\vec{a} = \vec{1}_L = [1_0, 1_1, ..., 1_{L-1}]^T$ since $\vec{x}^\Delta$ is pre-delayed. Finally the Capon weight vector is formed from (\ref{eq:b}) as
-\begin{align}\label{eq:w}
-\vec{w}[n] = \frac{\vec{b}[n]}{\vec{1}^H\vec{b}[n]},
-\end{align}
-and the beamformer output is calculated by combining all $N_L$ subarrays weighted with the same set of adaptive weights.
+introduced by Synnev\aa{}g et al.
+\cite{Synnevag2007} has been applied in much of the resent literature on Capon beamforming for medical ultrasound imaging \cite{Nilsen2009, Wang2009, Mehdizadeh2012}.
+This form of adaptive diagonal loading has also been applied to HF and VHF antenna processing \cite{Featherstone1997}.
+
+The beamformer output is in the end calculated by combining all $N_L$ subarrays weighted with the same set of adaptive weights.
 \begin{align}
@@
 where $\text{CONV}_m$ is a non-truncated convolution across elements.
-Note how we can choose to first sum all subarrays (\ref{eq:z_mv}), or accumulate the right weighting per element (\ref{eq:z_mv2}).
-
-In passive systems, assuming stationary statistics, the sample covariance matrix in (\ref{eq:R}) is typically averaged over a long sequence of samples\cite{Krima}.
-For an active, pulsed system like medical ultrasound imaging, the available data from which $\mat{\hat{R}}$ can be estimated is quite limited.
-The resulting sample covariance matrix might therefore be both inaccurate, and poorly conditioned.
-Another challenge with applying Capon to active systems is the effect of signal cancellation, where high correlation between signal and interference favors weights that cancel out the signal \cite{Reddy1987}.
-In this way, the beamformer output can be too low at broadside, even when there is signal present and the distortionless-response criterion is applied.
-
-In order to get a well-conditioned $\mat{\hat{R}}$, avoid signal cancellation, and to get DAS-like speckle, $\mat{\hat{R}}$ has to be averaged over $L\le M/2$ long subarrays and $N_K \sim T_p/T_s$ time samples \cite{Synnevag2007, Synnevag2007a}.
-Where, $T_p$ is the pulse length in seconds and $T_s$ is the sampling period.
-For baseband data, one pulse length ($1.5\lambda$) is around 3 samples ($K=1$).
+Note how we can either select to sum all subarrays first in (\ref{eq:z_mv}), or accumulate the right weighting per element in (\ref{eq:z_mv2}).
 
@@
 The key point is that GPUs can accelerate small concurrent tasks, like beamforming one pixel, much better than a CPU.
-Where the CPU has few cores and logic to process complex threads, the GPU has many cores, but each thread needs to be simple in order to benefit from the GPU's computational power.
-
-The CUDA compute model is based on execution of a kernel function across a large set of threads, where the kernel function describe the work to be done in each thread.
-As depicted in Fig.
-\ref{fig:gpulayout}, threads are organized in a grid of thread-blocks, with maximal size of 1024 threads per block.
-Threads inside a block supports low-cost synchronized if needed.
+Where the CPU has few cores and logic to process complex threads, the GPU has many cores, but each thread needs to be simple in order to benefit from the extra computational power.
+
+The CUDA compute model is based on execution of a kernel function across a grid of compute threads, where the kernel function describe the work to be done by each thread.
+As depicted in Figure \ref{fig:gpulayout}, each position in the grid holds a block of threads, with maximal size of 1024 threads, that can be synchronized with little cost if needed.
+%Each block is further divided into groups of 32 threads, known as a warp. The maximum number of threads inside a block is restricted to 512, hence 16 warps. 
 From a CUDA perspective the GPU consist of one or more stream multi processors (SMs), where each SM is capable of 32 or more, depending on the architecture, concurrent multiplications and/or additions.
@@
 
-An SM has available a limited amount of near-core memory and registers.
+A SM has available a limited amount of near-core memory and registers.
 Each thread therefore has to be as light on memory as possible and has to perform enough instructions per used byte in order to hide memory latency.
-It is also important that the problem can be divided into enough threads to hide instruction latency.
 Global reads and writes on the GPU, in addition to CPU-GPU transfers, have high penalty, and should be minimized as much as possible.
-All of these constraints combined pose a real challenge as the target problem has to map on to the parallel framework both on a grid, block, thread and memory consumption level in order to benefits from parallel acceleration.
+All of these constraints combined pose a real challenge as the target problem has to map on to the parallel framework both on a grid, block and memory consumption level in order to benefits from parallel acceleration.
 In the next section, we will describe how this mapping should be done for Capon beamforming of cardiac ultrasound data.
@@
 \section{Parallel Capon}\label{sec:meth}
-If we analyze the beamformer in (\ref{eq:z_mv}) in combination with (\ref{eq:b}) and (\ref{eq:R}), we see that each calculation of a set of Capon weights is independent across the outputs $z[n]$.
-As shown in Fig.
-\ref{fig:gpulayout}, a first level of granularity is therefore to divide the image into a grid, where the output in each cell can be calculated independently of all others.
+As mentioned, widespread use of Capon adaptive beamforming has been limited by its computational complexity.
+However, if we analyze the beamformer in (\ref{eq:z_mv}) in combination with (\ref{eq:w}) and (\ref{eq:R}), we see that each calculation of a set of Capon weights is independent across the outputs $z[n]$.
+As shown in Figure \ref{fig:gpulayout}, a first level of granularity is therefore to divide the image into a grid, where the output in each cell can be calculated independently of all others.
 Further, based on memory dependencies and to hold memory consumption per thread at a minimum, one should recognise that a group of threads should process one output in the final image.
-Continuing with the calculation of one weight vector, the algorithm can be broken down into three main steps that, to some extent, can be divided further into a set of parallel processes.
+Continuing with the calculation of one weight vector, the algorithm can be broken down into three main steps that, to some extent, can be further divided into a small set of parallel processes.
 That is, for each data vector $\vec{x}^\Delta$ in the image:
@@
 \text{1) } &\text{Calculate the sample covariance matrix } \mat{\hat{R}} \text{ as in } \text{(\ref{eq:diag})}\\
-\text{2) } &\text{Solve the linear system of equations in (\ref{eq:b}})\\ %\vec{b} =\mat{\hat{R}}^{-1}\vec{1}_L\\
+\text{2) } &\text{Solve the linear system of equations } \vec{b} =\mat{\hat{R}}^{-1}\vec{1}_L\\
 \text{3) } &\text{Calculate } \vec{w} = \vec{b}/\vec{1}_{L}^H\vec{b} \text{ and the amplitude as in (\ref{eq:z_mv})}
@@
 
-\subsection{Calculating Multiple Sample Covariance Matrices}\label{sec:calcR}
-
-%The straightforward sum in (\ref{eq:R}) has complexity $O(L^2N_LN_K) = O(L^3)$ since $N_L$, the number of subarrays, is a function of $L$.
+\subsection{Calculating Multiple Sample Covariance Matrices}
+
+The straightforward sum in (\ref{eq:R}) has complexity $O(L^2N_LN_K) = O(L^3)$ since $N_L$, the number of subarrays, is a function of $L$.
 This is actually the same as for inverting an $L \times L$ matrix.
@@
 
-From (\ref{eq:R}) it is obvious that calculation of one element in $\mat{\hat{R}}$ is independent of all other elements.
-A natural level of granularity would then be to assign one thread to each element of $\mat{\hat{R}}$ as done by \cite{Chen2011}.
-This will minimize global reads and writes needed per thread, but unfortunately the maximum number of threads per block puts an upper limit on the subarray length, $L_{max}=32$, since we want to avoid block-to-block synchronization.
+From (\ref{eq:R}) it is obvious that calculation of one element in $\mat{\hat{R}}$ is independent of all the others.
+A natural level of granularity would then be to assign one thread to each element of $\mat{\hat{R}}$.
+This will minimize global reads and writes needed per thread, but unfortunately the maximum number of threads per block puts an upper limit on the subarray length, $L_{max}=32$, if we want to avoid block-to-block synchronization.
 
 The cubic complexity with $L$ and linear complexity with $N_K$ can be reduced by realizing that the calculation in (\ref{eq:R}) overlap both across subarrays and in time.
-Element $(i,j)$ of $\mat{\breve{R}}[n]$ can be calculated from element $(i-1, j-1), \text{for } i,j \in [1, L-1]$ as
+Element $(i,j)$ of $\mat{\hat{R}}[n]$ can be calculated from element $(i-1, j-1), \text{for } i,j \in [1, L-1]$ as
 \begin{align}
-\mat{\breve{R}}_{i,j}[n] &=  \mat{\breve{R}}_{i-1,j-1}[n]  \nonumber \\
-&+ \sum_{n'=n-K}^{n+K} (x_{i+N_L-1}^{\Delta}[n']x_{j+N_L-1}^{\Delta}[n']^* \nonumber \\
-&- x_{i-1}^{\Delta}[n']x_{j-1}^{\Delta}[n']^H). \label{eq:sliding}
+\mat{\breve{R}}_{i,j}[n] &=  \mat{\breve{R}}_{i-1,j-1}[n]  + \sum_{n'=n-K}^{n+K} (x_{i+N_L-1}^{\Delta}[n']x_{j+N_L-1}^{\Delta}[n']^* \nonumber \\
+ &- x_{i-1}^{\Delta}[n']x_{j-1}^{\Delta}[n']^H) \label{eq:sliding}
 \end{align}
@@
 %- \mat{\breve{R}}[n-K-1] + \mat{\breve{R}}[n-K].
-Both sliding across subarrays and in time can be realized simultaneously if we hold a $N_K$-average in time of the outer products $\vec{x}^{\Delta}[n]\vec{x}^{\Delta}[n]^H$ in memory and calculate $\mat{\hat{R}}$ averaged over subarrays from this buffer (cite JOE?).
-However, since $N_K$ typically is small, and we have a limited amount of memory available per thread, we have chosen to focus on sliding across subarrays only (\ref{eq:sliding}).
+Both sliding across subarrays and in time can be realized simultaneously if we hold a $N_K$-average in time of the outer products $\vec{x}^{\Delta}[n]\vec{x}^{\Delta}[n]^H$ in memory and calculate $\mat{\hat{R}}$ averaged over subarrays from this buffer (ref JOE).
+However, since $N_K$ typically is small, and we have a limited amount of memory per thread on the GPU, we have chosen to focus on sliding across subarrays only (\ref{eq:sliding}).
 In that case, the complexity of calculating $\mat{\hat{R}}$ is reduced to $O(L^2)$.
-Notice how sliding will break independence along the sliding dimension.
+Important to notice is that sliding will break independence in the sliding dimension.
 With our choice of calculating $\mat{\hat{R}}$, calculations are still independent across time, but not along diagonals in $\mat{\hat{R}}$.
-We therefore have selected to use one thread per diagonal in the upper triangle of $\mat{\hat{R}}$, and for small values of $L$, several covariance matrices are calculated per thread block (Fig.
-\ref{fig:gpulayout}).
+We therefore have selected to use one thread per diagonal in the upper triangle of $\mat{\hat{R}}$, and for small $L$s several covariance matrices are calculated per thread block (Figure \ref{fig:gpulayout}).
 Maximum supported $L$ is with this approach increased to 512, however for small $L$ high $M$ configurations, memory consumption per thread will be to high to achieve full occupancy of the GPU.
@@
 This also means that we do not rely on a symmetric solver to solve $\mat{\hat{R}}\vec{b} = \vec{1}_L$, even if that could have saved some instructions.
+The memory layout for the upper triangle only is also believed to be cumbersome.
 
@@
 For the results presented in this paper, we have used an unreleased GPU implementation of Gauss-Jordan (GJ) elimination (by Nvidia) to solve $\mat{\hat{R}}\vec{b} = \vec{1}_L$, where $\vec{b} = \mat{\hat{R}}^{-1}\vec{1}_L$.
-%For smaller systems, $L < 10$, a direct solver using one thread per system is preferable.
+For smaller systems, $L < 10$, a direct solver using one thread per system is preferable.
 For this purpose we have implemented Cramer's rule for $2 \times 2$, $3 \times 3$ and $4 \times 4$ matrices.
@@
 \end{align}
-Each thread now computes $z_r = w_r^*x_r$, and finally one thread computes and outputs $z = \frac{1}{N_L}\sum_{r=0}^{L-1} z_r$ to global memory.
+Each thread now computes $z_r = w_r^*x_r$, and finally one thread computes and outputs $z = \sum_{r=0}^{L-1} z_r$ to global memory.
 The last sum can also be performed using a binary reduction sum across several threads for increased speed.
@@
 The beamspace version of the Capon beamformer is derived by replacing the covariance matrix in (\ref{eq:w}) with the beamspace covariance matrix $\mat{\hat{R}}_{BS} = f(\vec{x}_{BS}) = f(\mat{B}\vec{x})$, where $\mat{B} \in \mathbb{C}^{N_b,L}$, and $N_b$ is the number of selected beams.
-The transformation must be done per subarray, yielding subarrays that no longer overlap in space.
-If the transformation is applied before the covariance matrix is constructed, the amount of input data is change from $M$ to $N_LN_b$ elements per sample, however all remaining steps will get lower complexity if $N_b < L$.
-
-Another choice is to transform the element space covariance matrix to beamspace. 
-\begin{align}
-\mat{\hat{R}}_{BS} &= \mat{B}\mat{\hat{R}}\mat{B}^H
-\end{align}
-The pros using this approach is that minor changes has to be made to the pipeline described in Section \ref{sec:meth}, but it will only speed up the solving step if the solution $\vec{b}_{BS}$ is transformed back to element space.
-In this paper we will describe how the first approach has been implemented on the GPU.
-%By this approach we go from solving $\mat{\hat{R}}\vec{b} = \vec{1}_L$, which is an $L \times L$ system, to solving the $N_b \times N_b$ system $\mat{\hat{R}}_{BS}\vec{b}_{BS} = (\sqrt{L}\vec{e}_1)$.
-Now, one can either choose to transform the subarray sums, as in (\ref{eq:subsum}), to beamspace or transform the result, $\vec{b}_{BS}$, back to element space.
-In both cases solving has to be performed against a transformed steering vector
+Note that when subarray averaging is applied, the transformation must be done per subarray, yielding subarrays that no longer overlap in space.
+If we want to apply the transformation before the covariance matrix is constructing, the amount of data is change from $M$ to $N_LN_b$ per sample.
+The implementation calculating covariance matrices needs to be adjusted to handle non-overlapping subarrays.
+BuildR complexity is changed to $O(N_b^2L)$, so constructing the covariance matrices will be faster in beamspace as long as $N_b < \sqrt{L}$.
+However it will involve large changes to the processing pipeline described in Section \ref{sec:meth}.
+
+In both cases we need to solve using a transformed steering vector
 \begin{align}
@@
 \end{align}
-where $\vec{e}_i$ has value one in the i$^{th}$ position and is zero elsewhere.
-
-%The implementation calculating covariance matrices therefore needs to be adjusted to handle non-overlapping subarrays.
-The complexity of constructing the covariance matrix is changed to $O(N_b^2L)$, so constructing the covariance matrices will be faster in beamspace as long as $N_b < \sqrt{L}$.
-%Derivation of sliding beamspace transformation. 
-Given a $M$ long data vector divided into $N_L$ subarrays, the beamspace transformation of the $l^{th}$ subarray is given as:
-\begin{align}\label{eq:bs_formula}
-x_{BS,l,k} = \frac{1}{\sqrt{L}}\sum_{p=i}^{l+L-1}x_p e^{-j2\pi k(p-l)/L} , k \in [0, N_b-1].
+where $\vec{e}_i$ has value one in the i$^{th}$ position and is zero everywhere else.
+
+Another choice is to transform the covariance matrix to beamspace. 
+\begin{align}
+\mat{\hat{R}}_{BS} &= \mat{B}\mat{\hat{R}}\mat{B}^H\\
 \end{align}
-Manipulating this expression we find that for $l \in [1, L-1]$
-\begin{align}\label{eq:sliding_bs}
-x_{BS,l,k} &= \frac{1}{\sqrt{L}}((x_{BS,l-1,k} - x_{l-1})e^{-j2\pi k/L} \nonumber \\ &+ x_{i+L-1}e^{-j2\pi k(L-1)/L}) \nonumber \\
-&= \frac{1}{\sqrt{L}}(x_{BS,l-1,k} - x_{l-1} + x_{i+L-1})e^{-j2\pi k/L}.
-\end{align}
-We call this the sliding beamspace transformation (SBS), and it is equal to a normalized sliding DFT. For $l=0$ a full computation has to be performed using (\ref{eq:bs_formula}).
-
-From (\ref{eq:sliding_bs}) it is obvious that instructions can be saved if we utilize the overlap across subarrays, with an increase in task parallel granularity (As discussed in Section \ref{sec:calcR}).
-With SBS we can therefore select to have one thread per beam (index $k$) that calculates this beam for all subarrays.
-However, since the number of beams usually is small ($N_b \ll M$) we will end up with a large memory per thread ratio.
-Using one thread per beam is therefore not feasible for small $N_b$ large $M$ configurations.
-The solution is to combine SBS with full computations.
-As shown in Fig.
-(???), a number of thread groups target different sets of subarrays which they transform using the SBS algorithm.
-In that way, the number of threads per sample is increased and thus the memory per thread ratio is decreased.
-
-As mentioned the amount of data is increased from $M$ to $N_LN_b$ per sample in beamspace.
-This means that we no longer can exploit spatial overlap when calculating the sample covariance matrix, but since $N_b$ is small we can now allow to have one thread per element in $\mat{\hat{R}}_{BS}$.
-A kernel similar to the one described by Chen et al.
-\cite{Chen2011}, has therefore been implemented.
-For small $N_b$ values, several covariance matrices are calculated per thread block to achieve high occupancy of the SMs.
-
-After solving $\vec{b}_{BS} = \sqrt{L}\mat{\hat{R}}_{BS}^{-1}\vec{e}_1$, the beamspace weight vector is calculated as 
-\begin{align}
-\vec{w}_{BS} = \frac{\vec{b}_{BS}}{\sqrt{L}\vec{e}_1\vec{b}_{BS}}
-\end{align}  
-and applied to the beamspace data summed across subarrays.
-
-%How do we calculate the number of possible subarrays $N_L$ with overlap $L-s$ in a data vector of length $M$.
-%\begin{align}
-%N_L = floor(\frac{M}{s}) - floor(\frac{L-1}{s})
-%\end{align}
-
-As presented in (\ref{eq:diag}), adaptive diagonal loading is performed by loading the sample covariance matrix with the average energy per element, hence $trace\{\mat{\hat{R}}\}/L$.
-If adaptive diagonal loading is applied after data has been transformed into beamspace, a high percentage of the total energy is mapped to a small subset of beams.
-Thus the average energy per beam is higher than the average energy per element, if the system uses focused transmit beams.
-This means that less diagonal loading (smaller $d$) has to be applied in beamspace than in element space.
+By this approach we go from solving $\mat{\hat{R}}\vec{b} = \vec{1}_L$, which is an $L \times L$ system, to solving the $N_b \times N_b$ system $\mat{\hat{R}}_{BS}\vec{b}_{BS} = (\sqrt{L}\vec{e}_1)$.
+Now, one can either choose to transform the subarray sums, as in (\ref{eq:subsum}), to beamspace or transform the result, $\vec{b}_{BS}$, back to element space.
+
+In our implementation we have tried to minimize the number of adjustments that had to be done in order to integrate beamspace processing into the GPU-Capon processing pipeline.
+The beamspace-results presented in this paper is therefore produced by transforming covariance matrices to beamspace, and after solving, transforming the result back to element space.
 
@@
 \section{Results}\label{sec:res}
-
 Describe the computer system (CPU, GPU...). Describe the data that are processed (field II, Channel data from GE Vingmed).
@@
 \begin{figure*}[!t]
-\centerline{\subfloat[]{\includegraphics[width=3.5in]{gfx/benchmark_bar_32_40_400.png} \label{fig:benchUS32}}
+\centerline{\subfloat[]{\includegraphics[width=3.5in]{gfx/benchmark_us_32_lr.png} \label{fig:benchUS32}}
 \hfil
-\subfloat[]{\includegraphics[width=3.5in]{gfx/benchmark_bar_64_80_400.png}%
+\subfloat[]{\includegraphics[width=3.5in]{gfx/benchmark_us_64_lr.png}%
 \label{fig:benchUS64}}}
-\caption{Benchmark of GPU Capon beamforming for a cardiac image covering a $70\degree$ wide $15.4$cm deep sector.
+\caption{(Change into bar plot?) Benchmark of GPU Capon beamforming for a cardiac image covering a 70 degrees sector.
 Execution time is plotted as a function of subarray length $L$.
 The plot shows execution times for the three steps listed in Section \ref{sec:meth}.
-Calculating covariance matrices with three different choices of temporal averaging, solver, and the final beamformer sum.
-For larger arrays, more transmit beams have to be emitted for sufficient lateral sampling.
-(a) Shows the benchmark for a 32 element array.
-(b) Shows the benchmark for a 64 element array.}
+Calculating covariance matrices in red, solver in green, and the final beamformer sum in blue.
+For each step, there is one line for three different choices of temporal averaging, $N_{avg} \in [0, 1, 2]$.
+(a) A 32 element array.
+(b) A 64 element array.}
 \label{fig:bench}
@@
 \begin{figure*}[!t]
-\centerline{\subfloat[]{\includegraphics[width=3.5in]{gfx/benchmark_bar_bs_M=64_Nx=80_Ny=400_Nb=3.png} \label{fig:benchBS3}}
+\centerline{\subfloat[]{\includegraphics[width=3.5in]{gfx/benchmark_us_32_lr.png} \label{fig:benchBS3}}
 \hfil
-\subfloat[]{\includegraphics[width=3.5in]{gfx/benchmark_bar_bs_M=96_Nx=120_Ny=400_Nb=3.png}%
+\subfloat[]{\includegraphics[width=3.5in]{gfx/benchmark_us_64_lr.png}%
 \label{fig:benchBS5}}}
-\caption{Benchmark of GPU beamspace Capon beamforming for a cardiac image covering a $70 \degree$ and $15.4$cm deep sector.
-Execution time is plotted as a function of subarray length $L$.
-(a) Shows the benchmark for a 64 elements array.
-(b) Shows the benchmark for a 96 elements array.}
+\caption{(TODO: Add the right figure.
+Change it into a bar plot?) Benchmark of GPU Beamspace Capon beamforming for a cardiac image covering a 70 degrees sector.
+Execution time is plotted as a function of number of elements $M$.
+The plot shows execution times for the three steps listed in Section \ref{sec:meth}.
+Calculating covariance matrices in red, solver in green, and the final beamformer sum in blue.
+For each step, there is one line for three different choices of temporal averaging, $N_{avg} \in [0, 1, 2]$.
+(a) 3 beams.
+(b) 5 beams.}
 \label{fig:benchBS}
@@
 \begin{figure*}[!t]
-\centerline{\subfloat[]{\includegraphics[width=2.3in]{gfx/das.png} \label{fig:phantomDAS}}
+\centerline{\subfloat[]{\includegraphics[width=2in]{gfx/mv_phantom.png} \label{fig:phantomDAS}}
 \hfil
-\subfloat[]{\includegraphics[width=2.3in]{gfx/capon.png} \label{fig:phantomMV}}
+\subfloat[]{\includegraphics[width=2in]{gfx/mv_phantom.png} \label{fig:phantomMV}}
 \hfil
-\subfloat[]{\includegraphics[width=2.3in]{gfx/capon_bs.png} \label{fig:phantomBS}}}
-\caption{Simulated phantom containing one bright and one dark "cyst" and two point targets in speckle.
-The simulation was done using Field II with a 2.5MHz, $\lambda/2$ and 64 elements phased array.
-Dynamic range is -30dB to 30dB with mean value represented as 0dB.
-a) Delay-and-sum with uniform weights.
-b) Capon ($L=32, K=2, d=0.1$) c) Beamspace Capon ($L=32, K=2, N_b=3, d=0.01$).}
-\label{fig:phantom}
+\subfloat[]{\includegraphics[width=2in]{gfx/mv_phantom.png} \label{fig:phantomBS}}}
+\caption{(TODO: Add the right figures) Have graph showing cyst contrast as a function of different parameters instead?
+Cyst and point targets phantom a) Delay-and-sum, uniform weights.
+b) Capon ($M=?, L=?, N=?$) c) Beamspace Capon ($M=?, L=?, N=?, N_b=?$).}
+\label{fig:bench}
 \end{figure*}
@@
 \begin{figure*}[!t]
-\centerline{\subfloat[]{\includegraphics[width=2.3in]{gfx/das_invivo.png} \label{fig:invivoDAS}}
+\centerline{\subfloat[]{\includegraphics[width=2in]{gfx/us_image.png} \label{fig:phantomDAS}}
 \hfil
-\subfloat[]{\includegraphics[width=2.3in]{gfx/capon_invivo.png} \label{fig:invivoMV}}
+\subfloat[]{\includegraphics[width=2in]{gfx/us_image.png} \label{fig:phantomMV}}
 \hfil
-\subfloat[]{\includegraphics[width=2.3in]{gfx/capon_bs_invivo.png} \label{fig:invivoBS}}}
-\caption{\textit{In-vivo} harmonic cardiac ultrasound image acquired using a 3.5MHz, 64 elements phased array.
-Number of receive beams for the $70\degree$ sector is 98.
-Dynamic range is -20dB to 20dB with mean value represented as 0dB.
-a) Delay-and-sum with uniform weights.
-b) Capon ($L=32, K=2, d=0.1$) c) Beamspace Capon ($L=32, K=2, N_b=3, d=0.01$).}
-\label{fig:invivo}
+\subfloat[]{\includegraphics[width=2in]{gfx/us_image.png} \label{fig:phantomBS}}}
+\caption{(TODO: Add the right figures) Have graph showing cavity-wall contrast as a function of parameters instead?
+In-vivo cardiac ultrasound image a) Delay-and-sum, uniform weights.
+b) Capon ($M=?, L=?, N=?$) c) Beamspace Capon ($M=?, L=?, N=?, N_b=?$).}
+\label{fig:bench}
 \end{figure*}
-
-Fig.
-\ref{fig:bench} shows the results from benchmarking the implementation described in Section \ref{sec:meth} on an Nvidia Quadro 6000 graphics card (14 SMs, 1 TFlops single precision, 6 GB global memory, and 144 GB/s memory bandwidth).
-The problem sizes are calculated based on 32 and 64 element arrays with center frequency 2.5 MHz.
-By assuming a system bandwidth of 80\% we need a complex sampling frequency of 2 MHz.
-If we image down to a depth of 15 cm we get approximately 400 samples per range line.
-To correctly sample a $70\degree$ sector with 64 and 32 element arrays ($\lambda/2$ pitch), we need at least 80 and 40 beams, respectively.
-The presented execution times do not include transfer of data from CPU-side to the GPU.
-What is included are all calculations, memory transfers from global to near-core GPU memory, and near core memory access.
-
-There are two videos attached to this paper. One cardiac loop, 
-%(add ref to video)
-and one simulation, both 
-%(add ref to videos) 
-processed with the following Capon beamformer settings: $L = 32$, $K = 2$, $d=0.01$.
-These settings have previously been shown to provide high resolution while maintaining DAS-like speckle \cite{Synnevag2007a} in ultrasound images.
-Cardiac data was acquired using a 64 element and 3.5 MHz phased array.
-Simulations were done using a 64 element 2.5 MHz phased array.
-For the simulation, 16x lateral over-sampling has been used, in order to avoid under sampling of the two bright point scatterers.
-The cardiac recording only has $30\%$ lateral over sampling.
 
@@
 
-Add a comparison between beamspace and full capon using small L.
-
 \section{Discussion}\label{sec:dis}
 Discuss results presented in Section \ref{sec:res}.
-
-From Fig.
-\ref{fig:benchUS64} we see that for $L = M/2 = 32$ and $K=1$, total execution time is 100 ms (10 fps).
-Taking into account that we use complex data, but less samples per image, this is similar to the performance reported by Chen et al.
-\cite{Chen2011}.
-For the 32 element array in Fig.
-(\ref{fig:benchUS32}), execution time is reduced to 11.5 ms (87 fps) using the same settings ($L=M/2, K = 1$).
-Doubling this number can be interpreted as pre-beamforming a 64 element array down to 32.
-The frame rate for a 64 element array is then increased from 10 to 44 fps.
-Note that lateral oversampling will increase the total execution time with a factor equal the amount of oversampling.
-
-A major part of the total execution time comes from solving systems of linear equations.
-Further research should therefore focus on making this contribution smaller.
-This can be done either by implementing faster solvers, or reducing the sample covariance matrix size in a novel way \cite{Nilsen2009}.
-%When doing such a reduction, including the pre-beamforming described above, one should investigate the loss of contrast and resolution compared with full Capon beamforming.
-As for now, we have shown that the Capon beamformer can be implemented in a GPU framework.
-And if we as in \cite{Chen2011} define real-time as 10 fps, all execution times  in Fig.
-(\ref{fig:bench}) are real time.
-If we on the other hand define real-time as the time it takes to acquire a given image, the image formed using a 32 and 64 element array has a real time requirement of 125 fps and 62.5 fps respectively.
-There is some distance to cover before that goal is reached for all values of $L \le M/2$.
-However we see that real time, by the last definition, is possible for $M=64, L < 16$, and for $M = 32, L < 13$.
-Another option to reach real-time execution is to use several GPUs simultaneously.
-Note that multi-line acquisition will increase the frame rate requirement accordingly.
-
-In the attached video of a simulated phantom one can see that Capon beamforming highly resolves moving point targets.
-The beamformer is shift invariant as long as the lateral sampling is increased.
-No major improvement in contrast is observed.
-In the cardiac recording, structures are thinner when the Capon beamformer is applied, however we observe a loss in contrast due to signal cancellation.
-Future work should focus on solving this issue.
 
@@
 Toeplitz-based solver (cite new article). Discuss it.
-
-How does beamspace perform when using plane wave imaging...?
 

### 2013-01-20 ASA2013/asa2013.tex
la2pdf commit - 2013:01:16	15:35:24	(Wed16.Jan)
ASA article on the way

@@
+\newif\ifTODO\TODOtrue                        % Use todo notes?
+
 %===============%                               %~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~%
@@
 \documentclass[
-   article                                      % Document type (non-standard)
- , 12pt                                         % Text size
-%  , draft
-%  , final                                        % Quality
- , xelatex                                      % Use the XeLaTeX compiler
-%  , biblatex                                     % Use 'biblatex' for references (best by far)
- , bibtex                                       % Use 'bibtex' for references (oldschool :)
-%  , movie
-%  , notodos                                      % Disable todos
- , layout
-%  , defaultformat                                % Attempt to use default formatting options
-%  , glossary                                     % Use a glossary
-]{common/mytemplate}
+a4paper,10pt
+]{common/ica2013_2}
 
-% \bibliography{references}
-\hypersetup{
-   bookmarksopen=true
- , bookmarksopenlevel=2
-}
-% Load the references.bib file
-% \bibliography{references}
+\usepackage[utf8x]{inputenc}
+\usepackage[table,dvipsnames,svgnames]{xcolor}
+\usepackage{titlesec}
+
+\newcommand\Fig[1]{Fig. \ref{#1}}
+
+\newcommand\Grey[1]{{\color{Grey}#1}}
+\newcommand\Red[1]{{\color{Red}#1}}
+\newcommand\Blue[1]{{\color{Blue}#1}}
+\newcommand\DarkBlue[1]{{\color{DarkBlue}#1}}
+\newcommand\LightBlue[1]{{\color{LightBlue}#1}}
+\newcommand\Brown[1]{{\color{Brown}#1}}
+\newcommand\Green[1]{{\color{Green}#1}}
+\newcommand\SeaGreen[1]{{\color{SeaGreen}#1}}
+\newcommand\Yellow[1]{{\color{yellow}#1}}
+\newcommand\Orange[1]{{\color{orange}#1}}
+
+\ifTODO
+   \newcounter{todoidx}
+   \definecolor{todobackground}{rgb}{0.95,0.95,0.95}
+   \setlength\marginparsep{1pt}
+   \setlength\marginparwidth{35pt}
+   \newlength\marginparwidthsmall
+   \setlength\marginparwidthsmall{\marginparwidth}
+   \addtolength\marginparwidthsmall{-7pt}
+   \newcommand\todo[1]{%
+      \addtocounter{todoidx}{1}%
+      {\color{Red}\bf(\thetodoidx{})}%%\fbox{\bf\thetodoidx{}}}%
+      \marginpar{%
+         {\vspace*{-10pt}\color{Red}\fbox{\bf\thetodoidx{}}}\\%
+         \fcolorbox{red}{todobackground}{\parbox{\marginparwidthsmall}{\scriptsize #1}}}}
+
+   \newcommand\todopar[1]{\fcolorbox{red}{white}{\parbox{0.97\linewidth}{#1}}}
+\else
+%    \usepackage[disable]{./todonotes} 
+   \newcommand\todo[1]{}
+\fi
+
+\newcommand\nn{\nonumber\\}
+
+\newcommand\nmat[1]{\begin{matrix}#1\end{matrix}}
+\newcommand\bmat[1]{\begin{bmatrix}#1\end{bmatrix}}
+\newcommand\case[1]{\begin{cases}#1\end{cases}}
+\newcommand\textbox[2]{\footnotesize\text{\parbox{#1}{\centering\emph{#2}}}}
+
+\newcommand\rand{\text{rand}}
+\newcommand\randn{\text{randn}}
+\newcommand\rect{\text{rect}}
+\newcommand\sinc{\text{sinc}}
+\newcommand\tr{\text{tr}}
+\newcommand\adj{\text{adj}}
+
+% \newcommand\max{\text{max}}
+\newcommand\argmin[1]{\text{arg}\,\underset{#1}{\text{min}}}
+
+\newcommand\qqquad{\quad\qquad}
+\newcommand\qqqquad{\qquad\qquad}
+
+\renewcommand\l[1]{\left#1}
+\renewcommand\r[1]{\right#1}
+
+% {\text{\parbox{1.5cm}{\centering volume hyper- sphere}}}
+
+%Keyword colouring:
+\newcommand\kw[1]{#1}
+\newcommand\parm[1]{#1}%\color{Black}#1\color{Black}}
+
+\newcommand\of[1]{\scriptstyle(\parm{#1})\displaystyle}
+\newcommand\df[1]{\scriptstyle[\parm{#1}]\displaystyle}
+\newcommand\var[3]{#1_\text{#2}\of{#3}}
+
+\newcommand\diag{\text{diag}}
+
+% \raisebox{lift}[extend-above-baseline][extend-below-baseline]{text}
+\newcommand\mt[1]{\text{\emph{#1}}} %mt = mathtext
+\newcommand\mathnorm{\textstyle}
+\newcommand\mathbig[1]{\displaystyle#1\mathnorm}
+\newcommand\mathsmall[1]{\scriptstyle#1\mathnorm}
+\newcommand\mathtiny[1]{\scriptscriptstyle#1\mathnorm}
+\newcommand\sfrac[2]{\scriptstyle\raisebox{0.25pt}[0pt][0pt]{$\frac{#1}{#2}$}\mathnorm}
+\newcommand\nfrac[2]{\textstyle\frac{#1}{#2}\displaystyle}
+
+\newcommand\sumu[1]{\sum\limits^{#1}\,}
+\newcommand\suml[1]{\sum\limits_{#1}\,}
+\newcommand\sumb[2]{\sum\limits_{#1}^{#2}\,}
+
+\newcommand\produ[1]{\prod\limits^{#1}\,}
+\newcommand\prodl[1]{\prod\limits_{#1}\,}
+\newcommand\prodb[2]{\prod\limits_{#1}^{#2}\,}
+
+\newcommand\defeq{\overset{\underset{\mathrm{def}}{}}{=}}
+
+%Math macros:
+\newcommand\diff[2]{\frac{\kw{d}\,\textstyle #1\scriptstyle}{\kw{d\parm{#2}}}\displaystyle}
+\newcommand\ddiff[2]{\frac{\kw{d^2}\,\displaystyle #1\scriptstyle}{\kw{d\parm{#2}}^2}\displaystyle}
+
+\renewcommand\d[1]{\scriptstyle\kw{\,d\parm{#1}}\displaystyle}
+
+% These commands are mutually exclusive. Remember to "renew" in v2.
+\newcommand\intb[4]{\int\limits_{#3}^{#4} #1 \d{#2}} % \int{exp}{var}{from}{to}
+\newcommand\intl[3]{\int\limits_{#3} #1 \d{#2}} % \int{exp}{var}{for all}
+\newcommand\intu[2]{\int #1 \d{#2}} % \int{exp}{var}{for all}
+
+\newcommand\T{^{\scriptscriptstyle T}}
+\renewcommand\H{^{\scriptscriptstyle H}}
+
+\renewcommand\vec[1]{\boldsymbol{#1}}
+\newcommand\mat[1]{\boldsymbol{#1}}
+
+\newcommand\1{\vec 1}
+\newcommand\I{\mat I}
+\renewcommand*\a{\vec a}
+\renewcommand*\i{\vec i}
+\renewcommand*\k{\vec k}
+\newcommand*\n{\vec n}
+\newcommand*\p{\vec p}
+\newcommand*\s{\vec s}
+\newcommand*\w{\vec w}
+\newcommand*\x{\vec x}
+\newcommand*\y{\vec y}
+
+\newcommand*\A{\mat A}
+\newcommand*\B{\mat B}
+\newcommand*\C{\mat C}
+\newcommand*\E{\mat E}
+% \renewcommand*\H{\mat H}
+\renewcommand*\P{\mat P}
+\newcommand*\eP{\mat{\hat P}}
+\newcommand*\R{\mat R}
+\newcommand*\Ri{\R^{-1}}
+\newcommand*\eR{\mat{\hat R}}
+\newcommand*\eRi{\hat{\mat R}\,\!^{-1}}
+\newcommand*\Navg{N_\text{avg}}
+\newcommand*\W{\mat W}
+\newcommand*\X{\mat X}
+\newcommand*\Xd{\X_{\!\Delta}}
+\newcommand*\Y{\mat Y}
+
+\renewcommand*\L{\mat \Lambda}
+\newcommand*\U{\mat U}
+% \renewcommand*\t{\mathtiny{^T}}
+% \newcommand*\h{\mathtiny{^H}}
+\renewcommand*\t{^T}
+\newcommand*\h{^H}
+
+\newcommand\D{\vec\nabla} %Del: Vector differential operator - nabla
+\newcommand\Dx{\vec\nabla\times}
+\newcommand\Dd{\vec\nabla\cdot}
+
+\usepackage{tikz}
+\usetikzlibrary{shapes,snakes}
+\usepackage{amsmath,amssymb}
+\usepackage{glossaries}
+
+\newenvironment{outline}
+{\begin{itemize}}
+{\end{itemize}}
 
@@
 The LCA creates images with a quality comparable to MVDR, and it is perfectly suited for a GPU.
+\end{abstract}
 
-\ \\
-Total: 192 words (200 max)
+\section{Introduction}
 
-\ \\
-Noen tanker jeg har selv nå:
-\begin{itemize}
-\item Det mangler noe om bruksområder.
-\item Må passe på å ikke selge LCA så bra at MVDR ser helt håpløs ut(?)
-\item 200 ord er ingenting. :)
-\end{itemize}
+To improve the image quality active sonar imaging several data adaptive techniques have
 
-\ \\
-Questions they wanted answered:
-\begin{itemize}
-\item This session will feature work on adapting computationally demanding modeling and signal processing tasks to take advantage of various parallel architectures, such as cloud computing, computing clusters, multicore CPUs, multi-core graphic processing units (GPUs), DSP chips, and FPGAs.
-\item If algorithm is slow - What technology will help me?
-\item What skills are needed?
-\item What tools are available and what's the cost?
-\item What acceleration factor are we looking at?
-\item How can I get started?
-\item What we did. Which tools? Was it hard? Speedup?
-\end{itemize}
-\end{abstract}
+Most modern active sonar imaging systems operate by combining data from sensor arrays to , from which data is combined
+
+The Minimum Variance Distortionless Response (MVDR) beamformer is known for its potential to improve sonar image quality.
+
+Most active sonar systems today employ phased arrays to record and
+
+Adaptive beamforming have in recent years been applied to active sonar imaging by several authors.
+
+The potential that adaptive beamformers to improve active sonar image quality is  ers is that theimprove image quality by utilising the apriori knowledge present in the backscattered wavefield.
+has for improving active sonar image quality has been demonstrated in e.g.
+[].
+are known for the potential to improve image quality in to When using adaptive beamformers to improbe
+
+\section{Methods}
+
+Suppose that a backscattered wavefield is recorded by an $M$ element uniform linear array, and let the output from the $m$'th channel be given as $x_m$.
+By definition a beamformer's output $z$ is a weighted sum $z$ of all these data outputs:
+\begin{align}
+z[n] = \w\H[n]\x[n] = \bmat{w_0[n]\\w_1[n]\\\vdots\\w_{M-1}[n]}^H \bmat{x_0[n]\\x_1[n]\\\vdots\\x_{M-1}[n]},\label{z}
+\end{align}
+where $w_m$ is the weight factor assigned to channel $m$.
+With static weights this would be referred to as the conventional delay-and-sum (DAS) beamformer.
+Various weighting patterns exists here for trading lateral resolution for improved noise suppression (contrast), but one always ends up with a compromise between the two\todo{don't quite like this one}~\cite{Harris1978}.
+
+\subsection{MVDR beamformer}
+
+Adaptive beamformers target this limitation by allowing the weights to change for each pixel to better fit the dynamic nature of the incoming wavefield.
+The MVDR beamformer is one such method.
+It finds the set of complex weights that minimizes the beamformer's expected output power, $\argmin{\w}\,E\{|z[n]|^2\}$, while ensuring unity gain in the look direction~\cite{Capon1969}.
+The solution to this optimization problem is given as:
+\begin{gather}
+\vec w[n] = \frac{\Ri[n]\1}{\1\T\Ri[n]\1},\label{weights}
+\end{gather}
+where $\1$ is a row vector of ones that represents broadside steering, and $\R=E\{\x[n]\x\H[n]\}$ is the spatial covariance matrix for the full array.
+Since $\R$ is unknown, we estimate it by by computing a sample covariance matrix $\eR$.
+In this computation we will perform some degree of \emph{spatial averaging} to avoid signal cancellation, \emph{temporal averaging} to maintain true speckle statistics, and \emph{diagonal loading} to improve robustness to parameter errors ~\cite{Synnevag2009a}.
+Combined, these steps will also ensure a numberically well conditioned $\eR$\todo{should we mention variance reduction?}.
+
+If we let $x_l[n]$ represent the datavector from subarray $l$,
+\begin{gather}
+\x_l[n] = \bmat{x_l[n] & x_{l+1}[n] & \dots & x_{l+L-1}[n]}\T,
+\end{gather}
+then a sample covariance matrix formed with temporal and spatial averaging, $\eR_\text{s,t}$, is given as:
+\begin{gather}
+\eR_\text{s,t}[n] =  \frac{1}{N_K N_L} \sumb{l=0}{M-L}\sumb{n'=n-K}{n+K} \x_l[n]\x_l\H[n] \in\mathbb{C}^{L,L},\label{spatialR}
+\end{gather}
+where $N_K = 2K+1$ is the number of temporal samples to perform averaging over, and $N_L = M-L+1$ is the number of subarrays.
+
+To further improve robustness, we add a fraction $d$ of the total power of $\eR_\text{t,s}[n]$ to its diagonal~\cite{Synnevag2007}:
+\begin{align}
+\eR[n] = \eR_\text{t,s}[n] + \I \frac{d}{L} \tr\{\eR_\text{t,s}[n]\},\label{finalR}
+\end{align}
+where $\I$ is an identity matrix, $\tr\{\cdot\}$ represents the matrix trace operation, and $\tr\{\eR_\text{t,s}[n]\}$ is an estimate of the energy received by this pixel.
+
+\subsection{LCA beamformer}
+
+Note how subarray averaging led to a size reduction of $\eR$ from $\mathbb{C}^{L,L}$ to $\mathbb{C}^{M,M}$, and hence will produce an $L$-element weightset when substituted into (\ref{z}).
+This weightset is then applied to on all the subarrays, prior to computing the beamformer output as in (\ref{z}).
+Or, equivalently, we may apply the weightset to the sum of all the subarrays:
+\begin{align}
+z[n] = \w\H[n] \sumb{l=0}{M-L} \x_l[n].\label{finalZ}
+\end{align}
+\begin{figure}[!t]\centering
+\includegraphics[width=\linewidth]{gfx/implementation.eps}
+\caption{MVDR beamforming. For one of the total number of pixels in range and azimuth, $N_y$ and $N_x$,\newline
+1. an $L\times{}L$ sample covariance matrix $\eR$ is computed, \newline
+2. the term $\eR^{-1}\1$ is found using a linear equation solver,\newline
+3. and the beamformer output is computed from $z$ from (\ref{finalZ}), where $\w$ is found by substituting $\eR^{-1}\1$ into (\ref{weights}). } \label{mvdr_beamforming}
+\end{figure}
+% \begin{figure}[!t]\centering
+% \includegraphics[width=\linewidth]{gfx/algorithm_structure.eps}
+% \caption{MVDR beamforming.
+First a spatial covariance matrix is estimated from the delayed data (\ref{spatialR}-\ref{finalR}), then the weights are computed (\ref{weights}) and finally applied to the delayed channel data (\ref{z}).}
+% \label{implementation}
+% \end{figure}
+As summarized in Fig. \ref{mvdr_beamforming}, the MVDR method is applied to each pixel independently, by
+\begin{enumerate}
+\item computing the sample covariance matrix $\eR$ in (\ref{finalR}),
+\item computing $\eRi\1$ in (\ref{weights}), and
+\item computing the beamformer output $z$ in (\ref{finalZ}).
+\end{enumerate}
+
+\section{Results}
+
+\begin{figure}[!t]\centering
+\includegraphics[width=\linewidth]{gfx/implementation.eps}
+\caption{MVDR beamforming. For one of the total number of pixels in range and azimuth, $N_y$ and $N_x$,\newline
+1. an $L\times{}L$ sample covariance matrix $\eR$ is computed, \newline
+2. the term $\eR^{-1}\1$ is found using a linear equation solver,\newline
+3. and the beamformer output is computed from $z$ from (\ref{finalZ}), where $\w$ is found by substituting $\eR^{-1}\1$ into (\ref{weights}). } \label{mvdr_beamforming}
+\end{figure}
+
+% 
+% \ \\
+% Total: 192 words (200 max)
+% 
+% \ \\
+% Noen tanker jeg har selv nå:
+% \begin{itemize}
+% \item Det mangler noe om bruksområder.
+% \item Må passe på å ikke selge LCA så bra at MVDR ser helt håpløs ut(?)
+% \item 200 ord er ingenting. :)
+% \end{itemize}
+% 
+% 
+% \ \\
+% Questions they wanted answered:\cite{Kailath1985}
+% \begin{itemize}
+% \item This session will feature work on adapting computationally demanding modeling and signal processing tasks to take advantage of various parallel architectures, such as cloud computing, computing clusters, multicore CPUs, multi-core graphic processing units (GPUs), DSP chips, and FPGAs.
+% \item If algorithm is slow - What technology will help me?
+% \item What skills are needed?
+% \item What tools are available and what's the cost?
+% \item What acceleration factor are we looking at?
+% \item How can I get started?
+% \item What we did. Which tools? Was it hard? Speedup?
+% \end{itemize}
+% \end{abstract}
+
+\bibliographystyle{common/jasanum}
+\bibliography{../../Library/library}
 

### 2013-01-20 ASA2013/asa2013.tex
la2pdf commit - 2013:01:20	15:35:38	(sø.20.jan.)

--- 2012-11-26 ASA2013/asa2013.tex
@@
 Unfortunately, it is also significantly more computationally expensive because a spatial covariance matrix must be estimated and inverted for each image pixel.
-We target this challenge by altering and mapping MVDR to a GPU, and suggest three different solutions depending on the system size.
+We target this challenge by altering and mapping the MVDR beamformer to a GPU, and suggest three different solutions depending on the system size.
 
 For systems with relatively few channels, we suggest arithmetic optimizations for the estimation step, and show how a GPU can be used to yield image creation rates of more than 1\,Mpx/s.
-For larger systems we show that frequency domain processing is preferable.
-This promotes high processing rates at a negligible reduction in image quality.
+For larger systems we show that frequency domain processing is preferable, as this promotes high processing rates at a negligible reduction in image quality.
 These GPU implementations consistently reduced the runtime by 2-3 orders of magnitude compared to our reference C and Matlab implementations.
@@
 \ \\
-Total: 192 words (200 max)
+Total: 195 words (200 max)
 

### 2013-01-20 ASA2013/common/ica_num.tex

\documentclass[a4paper,10pt]{ica2013_2}
\usepackage[utf8x]{inputenc}
\usepackage{xcolor}
\usepackage{titlesec}
%\usepackage{natbib}

\begin{document}
\section{First level heading (heading 1)}
This is the standard font and layout for the individual paragraphs.  The style is called "Paragraph."  Replace this text with your text.
\subsection{Second Level Heading (Heading 2) with Each Initial Letter Capitalized (Note: 
prepositions and articles should be lowercase)}
This is the standard font and layout for the individual paragraphs.  The style is called "Paragraph."  Replace this text with your text.
\subsubsection{Third Level Heading (Heading 3) with Each Initial Letter Capitalized (Note: prepositions and articles should be lowercase)}
Below is an example to place a new equation.
\begin{equation}
 \frac{d F_1}{d\omega}=SAm_2\,\cos \omega
\end{equation}
Some references \cite{bacon2000hot,hamilton1837third, kulvait2012nonlinear} are cited here as sample refernces. These references are included in the sample bibtex file attached with this file.

Sample table below
\begin{table}[!h]
\centering
\caption{Center the caption above the Table.  Tables should have top and bottom 
rules, and a rule separating the column heads from the rest of the table only.  Do not display all grid lines}\label{table:aerodynamic_results}
 \begin{tabular}{ccc}
\hline
\textbf{Column Header Goes Here} & \textbf{Column Header Goes Here}  & \textbf{Column Header Goes Here}\\
\hline
Numerical  & 0.751  & 0.08\\
Experimental & 0.750 & 0.071\\
\hline
 \end{tabular}
\end{table}

You can also add figures in EPS (using latex) or PDF format (using pdflatex). Pictures can also be imported as PNG files if you use pdflatex. See for example Fig.~
\begin{figure}[!h]
\centering
\includegraphics[width=0.45\textwidth]{sample_fig.pdf}
\caption{This is a sample figure. You can put the caption here.}\label{fig:sample_fig}
\end{figure}

\section*{ACKNOWLEDGMENTS}
Acknowledgments go here. Note that you must use \begin{verbatim}\section*\end{verbatim} command to remove the section numbering.

The references in the next section should follow either \textit{JASAnum} or \textit{JASAAuthyear} bibtex style. If using \textit{JASAAuthyear}, use the \textit{natbib} package by uncommenting the respective command in the header of the  latex file.
\bibliographystyle{jasanum}
%\bibliographystyle{jasaauthyear}

\bibliography{template}
\end{document}

### 2013-01-20 ASA2013/asa2013.tex

--- 2013-01-20 ASA2013/asa2013.tex
@@
 The LCA creates images with a quality comparable to MVDR, and it is perfectly suited for a GPU.
+
+\ \\
+Total: 195 words (200 max)
+
+\ \\
+Noen tanker jeg har selv naa:
+\begin{itemize}
+\item Det mangler noe om bruksomraader.
+\item Maa passe paa aa ikke selge LCA saa bra at MVDR ser helt haaplos ut(?)
+\item 200 ord er ingenting. :)
+\end{itemize}
+
+\ \\
+Questions they wanted answered:
+\begin{itemize}
+\item This session will feature work on adapting computationally demanding modeling and signal processing tasks to take advantage of various parallel architectures, such as cloud computing, computing clusters, multicore CPUs, multi-core graphic processing units (GPUs), DSP chips, and FPGAs.
+\item If algorithm is slow - What technology will help me?
+\item What skills are needed?
+\item What tools are available and what's the cost?
+\item What acceleration factor are we looking at?
+\item How can I get started?
+\item What we did. Which tools? Was it hard? Speedup?
+\end{itemize}
 \end{abstract}
@@
 % \ \\
-% Noen tanker jeg har selv nå:
+% Noen tanker jeg har selv naa:
 % \begin{itemize}
-% \item Det mangler noe om bruksområder.
-% \item Må passe på å ikke selge LCA så bra at MVDR ser helt håpløs ut(?)
+% \item Det mangler noe om bruksomraader.
+% \item Maa passe paa aa ikke selge LCA saa bra at MVDR ser helt haapløs ut(?)
 % \item 200 ord er ingenting. :)

### 2013-01-21 ASA2013/asa2013.tex
Late night closure.

@@
 
+\newcommand\graphicsAI[2][]{%
+  \immediate\write18{./bin/laFigure #2 #1}%
+  \input{result}}%
+
 \newenvironment{outline}
@@
 The LCA creates images with a quality comparable to MVDR, and it is perfectly suited for a GPU.
-
-\ \\
-Total: 195 words (200 max)
-
-\ \\
-Noen tanker jeg har selv naa:
-\begin{itemize}
-\item Det mangler noe om bruksomraader.
-\item Maa passe paa aa ikke selge LCA saa bra at MVDR ser helt haaplos ut(?)
-\item 200 ord er ingenting. :)
-\end{itemize}
-
-\ \\
-Questions they wanted answered:
-\begin{itemize}
-\item This session will feature work on adapting computationally demanding modeling and signal processing tasks to take advantage of various parallel architectures, such as cloud computing, computing clusters, multicore CPUs, multi-core graphic processing units (GPUs), DSP chips, and FPGAs.
-\item If algorithm is slow - What technology will help me?
-\item What skills are needed?
-\item What tools are available and what's the cost?
-\item What acceleration factor are we looking at?
-\item How can I get started?
-\item What we did. Which tools? Was it hard? Speedup?
-\end{itemize}
 \end{abstract}
+% \ \\
+% Total: 195 words (200 max)
+% 
+% \ \\
+% Noen tanker jeg har selv naa:
+% \begin{itemize}
+% \item Det mangler noe om bruksomraader.
+% \item Maa passe paa aa ikke selge LCA saa bra at MVDR ser helt haaplos ut(?)
+% \end{itemize}
+% 
+% 
+% \ \\
+% Questions they wanted answered:
+% \begin{itemize}
+% \item This session will feature work on adapting computationally demanding modeling and signal processing tasks to take advantage of various parallel architectures, such as cloud computing, computing clusters, multicore CPUs, multi-core graphic processing units (GPUs), DSP chips, and FPGAs.
+% \item If algorithm is slow - What technology will help me?
+% \item What skills are needed?
+% \item What tools are available and what's the cost?
+% \item What acceleration factor are we looking at?
+% \item How can I get started?
+% \item What we did. Which tools? Was it hard? Speedup?
+% \end{itemize}
 
@@
 
-To improve the image quality active sonar imaging several data adaptive techniques have
+Graphics Processing Units (GPUs) is a technology tailored to perform image processing at incredible speeds.
+They are available off-the-shelf at reasonable prices, are very power efficient, and can be programmed in .
+Unlike a CPU, which employ clever means to make a few threads run as fast as possible, a GPU does little to optimize the execution speed of a single thread, but in return allow thousands of them to run simultaneously.
 
-Most modern active sonar imaging systems operate by combining data from sensor arrays to , from which data is combined
+This may seem perfectly suited for image reconstruction algorithms, because usually each pixel can be computed independently and concurrently of one another.
+While this might be true for simple techniques such as conventional beamformers, the more complex per-pixel computations present in adaptive techniques are harder for the GPU to handle.
+Through a case study of on such technique, the Minimum Variance Distortionless Response (MVDR) beamformer, we identify these challenges and explain how they were overcome.
+The end design is more than 2 orders of magnitude faster than a C implementation we started off with, it took considerable effort to get there.
 
-The Minimum Variance Distortionless Response (MVDR) beamformer is known for its potential to improve sonar image quality.
+<< PREVIOUS WORK >>
 
-Most active sonar systems today employ phased arrays to record and
+<< OUTLINE >>
 
-Adaptive beamforming have in recent years been applied to active sonar imaging by several authors.
+% MVDR - two parts - estimating R and solving
+% less 32: 
+%
 
-The potential that adaptive beamformers to improve active sonar image quality is  ers is that theimprove image quality by utilising the apriori knowledge present in the backscattered wavefield.
+% However, to a problem that is commonly met is not that the peak arithmetic performance is insufficient, but that it is difficult move data around fast enough to keep all the computing cores busy.
+We say that the algorithm is memory bound, as opposed to compute bound.
+
+% 
+% One of the hottest topics in high performance computing today is how Graphics Processing Units (GPUs) can be used to  are one of the hottest candidates for accelerating  .
+Their appeal in the gaming community makes them relatively cheap, readily available off-the-shelf, and most computers today already have one installed.
+% 
+% To improve the image quality active sonar imaging several data adaptive techniques have 
+% 
+% Most modern active sonar imaging systems operate by combining data from sensor arrays to , from which data is combined 
+% 
+% The Minimum Variance Distortionless Response (MVDR) beamformer is known for its potential to improve sonar image quality. 
+% 
+% Most active sonar systems today employ phased arrays to record and 
+% 
+% Adaptive beamforming have in recent years been applied to active sonar imaging by several authors. 
+% 
+% The potential that adaptive beamformers to improve active sonar image quality is  ers is that theimprove image quality by utilising the apriori knowledge present in the backscattered wavefield.
 has for improving active sonar image quality has been demonstrated in e.g.
@@
 
-Suppose that a backscattered wavefield is recorded by an $M$ element uniform linear array, and let the output from the $m$'th channel be given as $x_m$.
-By definition a beamformer's output $z$ is a weighted sum $z$ of all these data outputs:
+Suppose that a backscattered wavefield is recorded by an $M$ element uniform linear array, that a suitable set of delays are applied to the channels for the pixel of interest, and let the delayed output from the $m$'th channel be given as $x_m$.
+By definition a beamformer's output $z[n]$ at time instant $n$ is a weighted sum all the delayed data outputs:
 \begin{align}
@@
 
-Adaptive beamformers target this limitation by allowing the weights to change for each pixel to better fit the dynamic nature of the incoming wavefield.
+\begin{figure}[t]\centering%
+\graphicsAI[drawing,width=0.7\linewidth]{gfx/implementation.svg}%
+\caption{MVDR beamforming. For one of the total number of pixels in range and azimuth, $N_y$ and $N_x$,\newline
+1. an $L\times{}L$ sample covariance matrix $\eR$ is computed, \newline
+2. the term $\eR^{-1}\1$ is found using a linear equation solver,\newline
+3. and the beamformer output is computed from $z$ from (\ref{finalZ}), where $\w$ is found by substituting $\eR^{-1}\1$ into (\ref{weights}). }\label{mvdr_beamforming}
+\end{figure}
+Adaptive beamformers target this limitation by computing a set of weights for each pixel that to better fit the dynamic nature of the incoming wavefield.
 The MVDR beamformer is one such method.
-It finds the set of complex weights that minimizes the beamformer's expected output power, $\argmin{\w}\,E\{|z[n]|^2\}$, while ensuring unity gain in the look direction~\cite{Capon1969}.
-The solution to this optimization problem is given as:
+It computes a set of weights that minimizes the noise and signal power accumulated by the array, but with a constraint of unity gain in the desired direction.
+When the data is delayed this is formulated as follows:
+\begin{gather}
+\argmin{\w}\ E\Big\{\big|\w\H[n]\x[n]\big|^2\Big\} \qquad\text{subject to}\qquad \w\H[n]\1 = 1. \label{mvdr_def}
+% \vec w[n] = \frac{\Ri[n]\1}{\1\T\Ri[n]\1},\label{weights}
+\end{gather}
+This is a convex optimization problem with the solution:
 \begin{gather}
@@
 Since $\R$ is unknown, we estimate it by by computing a sample covariance matrix $\eR$.
-In this computation we will perform some degree of \emph{spatial averaging} to avoid signal cancellation, \emph{temporal averaging} to maintain true speckle statistics, and \emph{diagonal loading} to improve robustness to parameter errors ~\cite{Synnevag2009a}.
-Combined, these steps will also ensure a numberically well conditioned $\eR$\todo{should we mention variance reduction?}.
+In this computation we perform some degree of \emph{spatial averaging} to avoid signal cancellation, \emph{temporal averaging} to maintain true speckle statistics, and \emph{diagonal loading} to improve robustness to parameter errors ~\cite{Synnevag2009a}.
+Combined, these techniques will also ensure a numerically well conditioned $\eR$.
 
-If we let $x_l[n]$ represent the datavector from subarray $l$,
+If we let $x_l[n]$ represent the datavector from subarray $l$\todo{replace with with a drawing?},
 \begin{gather}
@@
 \end{gather}
-then a sample covariance matrix formed with temporal and spatial averaging, $\eR_\text{s,t}$, is given as:
+then a sample covariance matrix formed with temporal and spatial averaging, $\eR$, is given as:
 \begin{gather}
-\eR_\text{s,t}[n] =  \frac{1}{N_K N_L} \sumb{l=0}{M-L}\sumb{n'=n-K}{n+K} \x_l[n]\x_l\H[n] \in\mathbb{C}^{L,L},\label{spatialR}
+\eR[n] =  \frac{1}{N_K N_L} \sumb{l=0}{M-L}\sumb{n'=n-K}{n+K} \x_l[n]\x_l\H[n] \in\mathbb{C}^{L,L},\label{spatialR}
 \end{gather}
 where $N_K = 2K+1$ is the number of temporal samples to perform averaging over, and $N_L = M-L+1$ is the number of subarrays.
-
-To further improve robustness, we add a fraction $d$ of the total power of $\eR_\text{t,s}[n]$ to its diagonal~\cite{Synnevag2007}:
-\begin{align}
-\eR[n] = \eR_\text{t,s}[n] + \I \frac{d}{L} \tr\{\eR_\text{t,s}[n]\},\label{finalR}
-\end{align}
-where $\I$ is an identity matrix, $\tr\{\cdot\}$ represents the matrix trace operation, and $\tr\{\eR_\text{t,s}[n]\}$ is an estimate of the energy received by this pixel.
+Finally, a fraction of the total power of $\eR[n]$ is added to its diagonal~\cite{Synnevag2007}.
 
@@
 
-Note how subarray averaging led to a size reduction of $\eR$ from $\mathbb{C}^{L,L}$ to $\mathbb{C}^{M,M}$, and hence will produce an $L$-element weightset when substituted into (\ref{z}).
-This weightset is then applied to on all the subarrays, prior to computing the beamformer output as in (\ref{z}).
-Or, equivalently, we may apply the weightset to the sum of all the subarrays:
-\begin{align}
-z[n] = \w\H[n] \sumb{l=0}{M-L} \x_l[n].\label{finalZ}
-\end{align}
-\begin{figure}[!t]\centering
-\includegraphics[width=\linewidth]{gfx/implementation.eps}
-\caption{MVDR beamforming. For one of the total number of pixels in range and azimuth, $N_y$ and $N_x$,\newline
-1. an $L\times{}L$ sample covariance matrix $\eR$ is computed, \newline
-2. the term $\eR^{-1}\1$ is found using a linear equation solver,\newline
-3. and the beamformer output is computed from $z$ from (\ref{finalZ}), where $\w$ is found by substituting $\eR^{-1}\1$ into (\ref{weights}). } \label{mvdr_beamforming}
+The Low Complexity Adaptive (LCA) beamformer is another take on the optimization problem in (\ref{mvdr_def}).
+Instead of calculating the weights like in (\ref{weights}), it simply computes the beamformer output for each window in a collection of such, and finally applies the one that best fulfils the given optimization criterion.
+If we let $\w_p$ be the $p$'th of a total of $P$ windows, then this can be formulated as:
+\begin{gather}
+\argmin{p}\ E\Big\{\big|\w_p\H\x[n]\big|^2\Big\} \qquad\text{subject to}\qquad \w_p\H\1 = 1. \label{lca_def}
+% \vec w[n] = \frac{\Ri[n]\1}{\1\T\Ri[n]\1},\label{weights}
+\end{gather}
+Note that (\ref{mvdr_ref}) and (\ref{lca_ref}) is essentially the same criterion.
+
+<< AVERAGING >>
+
+\section{Implementation}
+
+\begin{figure}[t]\centering%
+\graphicsAI[drawing,width=0.7\linewidth]{gfx/mvdr_implementation.svg}%
+\caption{MVDR implementation.}\label{mvdr_implementation}
 \end{figure}
+
+\section{Results}
+
+\section{Discussion}
+
+% 
+% $\underset{p}{\argmin}$
+% 
+% \begin{gather}
+% \vec w[n] = \frac{\Ri[n]\1}{\1\T\Ri[n]\1},\label{weights}
+% \end{gather}
+
 % \begin{figure}[!t]\centering
@@
 % \end{figure}
-As summarized in Fig. \ref{mvdr_beamforming}, the MVDR method is applied to each pixel independently, by
-\begin{enumerate}
-\item computing the sample covariance matrix $\eR$ in (\ref{finalR}),
-\item computing $\eRi\1$ in (\ref{weights}), and
-\item computing the beamformer output $z$ in (\ref{finalZ}).
-\end{enumerate}
+% As summarized in Fig. \ref{mvdr_beamforming}, the MVDR method is applied to each pixel independently, by
+% \begin{enumerate}
+% \item computing the sample covariance matrix $\eR$ in (\ref{finalR}),
+% \item computing $\eRi\1$ in (\ref{weights}), and
+% \item computing the beamformer output $z$ in (\ref{finalZ}).
+% \end{enumerate}
 
-\section{Results}
-
-\begin{figure}[!t]\centering
-\includegraphics[width=\linewidth]{gfx/implementation.eps}
-\caption{MVDR beamforming. For one of the total number of pixels in range and azimuth, $N_y$ and $N_x$,\newline
-1. an $L\times{}L$ sample covariance matrix $\eR$ is computed, \newline
-2. the term $\eR^{-1}\1$ is found using a linear equation solver,\newline
-3. and the beamformer output is computed from $z$ from (\ref{finalZ}), where $\w$ is found by substituting $\eR^{-1}\1$ into (\ref{weights}). } \label{mvdr_beamforming}
-\end{figure}
+% 
+% \section{Results}
+% 
+% \begin{figure}[t]\centering
+% \graphicsAI[width=\linewidth]{gfx/implementation.svg}
+% \caption{MVDR beamforming. For one of the total number of pixels in range and azimuth, $N_y$ and $N_x$,\newline
+% 1. an $L\times{}L$ sample covariance matrix $\eR$ is computed, \newline
+% 2. the term $\eR^{-1}\1$ is found using a linear equation solver,\newline
+% 3. and the beamformer output is computed from $z$ from (\ref{finalZ}), where $\w$ is found by substituting $\eR^{-1}\1$ into (\ref{weights}). } \label{mvdr_beamforming}
+% \end{figure}
 

### 2013-01-21 ASA2013/asa2013.tex
la2pdf commit - 2013:01:21	21:57:56	(Mon21.Jan)

@@
 a4paper,10pt
-]{common/ica2013_2}
+]{ica2013_2}
 
@@
 
+\definecolor{tabBlue}{HTML}{AACCFF}
+
 \newenvironment{outline}
@@
 
-<< OUTLINE >>
+This article is outlined as follows: In section \ref{methods} we will present the theoretical background on MVDR operated in time and frequency domain, as well as relating this to the LCA beamformer.
+As we move along we aim to mention complexity issues and suggest means to overcome them, but implementational aspect will first be presented in \ref{implementation}.
 
@@
 
-\section{Methods}
+\section{Methods}\label{methods}
 
@@
 
-\subsection{MVDR beamformer}
+\subsection{MVDR beamforming in time domain}
 
@@
 
+\subsection{MVDR beamspace}
+
+The Low Complexity Adaptive (LCA) beamformer is another take on the optimization problem in (\ref{mvdr_def}).
+Instead of calculating the weights like in (\ref{weights}), it simply computes the beamformer output for each window in a collection of such, and finally applies the one that best fulfils the given optimization criterion.
+If we let $\w_p$ be the $p$'th of a total of $P$ windows, then this can be formulated as:
+\begin{gather}
+\argmin{p}\ E\Big\{\big|\w_p\H\x[n]\big|^2\Big\} \qquad\text{subject to}\qquad \w_p\H\1 = 1. \label{lca_def}
+% \vec w[n] = \frac{\Ri[n]\1}{\1\T\Ri[n]\1},\label{weights}
+\end{gather}
+Note that (\ref{mvdr_ref}) and (\ref{lca_ref}) is essentially the same criterion.
+
+<< AVERAGING >>
+
 \subsection{LCA beamformer}
@@
 
-\section{Implementation}
+\section{Implementation}\label{implementation}
+
+% Most of the challenges we faced when porting the MVDR to a GPU are caused by a single troublesome fact: GPU prefers threads that computationally intensive but very light on memory consumption.
+
+Throughout our work with porting the MVDR beamformer to a GPU, one issue has been particularly difficult to solve: The GPU will only perform efficiently when each computing thread is very computationally intense.
+This should be apparent by studying Table \ref{tabbandwidth}.
+It compares theoretical peak values for the memory bandwidth and arithmetic throughput for an Nvidia Quadro 6000 GPU.
+Looking at global memory, we note that it is only able to move 1 float for every 30 floating point numbers that the GPU is able to execute.
+Now, if we consider the simple equation $C=A+B$, where the ratio is 3 floats read/written for every 1 floating point operation, we realise that global memory should be avoided whenever possible.
+Using shared memory (L1 cache) and registers is a big step in the right direction, but this e
+\begin{table}[b]\centering%\normalsize
+\begin{tabular}[c]{l r r r@{:}  l}\hline
+\rowcolor{tabBlue} & \multicolumn{1}{>{\columncolor{tabBlue}}c}{\bf BW$_\text{arithmetic}$} & \multicolumn{1}{>{\columncolor{tabBlue}}c}{\bf BW$_\text{memory}$} & \multicolumn{2}{>{\columncolor{tabBlue}}c}{\bf BW$_\text{memory}$/BW$_\text{arithmetic}$} \\\hline
+Arithmetic throughput & 1.03 Tflop/s & & \hspace{30pt} 1 & 30 \\
+Global memory & & 36 Gfloats/s & \hspace{30pt} 1 & 30 \\
+Shared memory & & 257 Gfloats/s & 1 & 4 \\
+Registers & & $>$1.5 Tfloats/s & $>$3 & 2~\cite{Vasilyy}
+\end{tabular}
+\caption{Nvidia Quadro 6000; Memory bandwidths (BW) compared to the peak arithmetic bandwidth of 1.03 TB/s.}\label{tabbandwidth}
+\end{table}
 
@@
 
+\begin{figure}[t]\centering%
+\graphicsAI[drawing,width=\linewidth]{gfx/plot_holmengraa_L16_Navg1.pdf}%
+\caption{Holmengraa.}\label{mvdr_implementation}
+\end{figure}
+
 \section{Discussion}
+
+MVDR for systems with less than 32 channels. After that MVDR in beamspace.
 
@@
 
-\bibliographystyle{common/jasanum}
-\bibliography{../../Library/library}
+\bibliographystyle{jasanum}
+% \bibliography{../../Library/library}
+\bibliography{library}
 

### 2013-01-21 ASA2013/asa2013.tex
la2pdf commit - 2013:01:21	23:05:42	(Mon21.Jan)

@@
 It compares theoretical peak values for the memory bandwidth and arithmetic throughput for an Nvidia Quadro 6000 GPU.
-Looking at global memory, we note that it is only able to move 1 float for every 30 floating point numbers that the GPU is able to execute.
-Now, if we consider the simple equation $C=A+B$, where the ratio is 3 floats read/written for every 1 floating point operation, we realise that global memory should be avoided whenever possible.
-Using shared memory (L1 cache) and registers is a big step in the right direction, but this e
+Looking at global memory, we note that it is only able to move 1 float for every 30 floating point operation that the GPU is able to execute.
+Now, consider the simple equation we the simple equation $C=A+B$, where the ratio is 3 floats read/written for every 1 floating point operation, we realise that global memory should be avoided whenever possible.
+
+The apparent solution is make sure that each thread primarily access the shared memory (L1 cache) and registers, but these are scarce resources. 
 \begin{table}[b]\centering%\normalsize
@@
 
-\begin{figure}[t]\centering%
+\begin{figure}[h]\centering%
 \graphicsAI[drawing,width=\linewidth]{gfx/plot_holmengraa_L16_Navg1.pdf}%

### 2013-01-22 ASA2013/asa2013.tex
la2pdf commit - 2013:01:22	11:07:12	(Tue22.Jan)
la2pdf commit - 2013:01:22	11:39:43	(Tue22.Jan)

@@
 \begin{gather}
-\argmin{p}\ E\Big\{\big|\w_p\H\x[n]\big|^2\Big\} \qquad\text{subject to}\qquad \w_p\H\1 = 1. \label{lca_def}
+\argmin{p}\ E\Big\{\big|\w_p\H\x[n]\big|^2\Big\} \qquad\text{subject to}\qquad \w_p\H\1 = 1\label{lca_def}. 
 % \vec w[n] = \frac{\Ri[n]\1}{\1\T\Ri[n]\1},\label{weights}
@@
 \argmin{p}\ E\Big\{\big|\w_p\H\x[n]\big|^2\Big\} \qquad\text{subject to}\qquad \w_p\H\1 = 1. \label{lca_def}
-% \vec w[n] = \frac{\Ri[n]\1}{\1\T\Ri[n]\1},\label{weights}
 \end{gather}
-Note that (\ref{mvdr_ref}) and (\ref{lca_ref}) is essentially the same criterion.
+Note the similarity between (\ref{mvdr_def}) and (\ref{lca_def}).
+The LCA and MVDR uses essentially the same optimization criterion, but whereas MVDR searches a continuous solution space for the best set of weights to apply, the LCA beamformer only have $P$ solutions to select from.
+However, once the LCA have 10-20 well designed windows to choose from its performance quickly converge towards the MVDR beamformer's performance.
+The difference in computational complexity, however, is
 
@@
 
-The apparent solution is make sure that each thread primarily access the shared memory (L1 cache) and registers, but these are scarce resources. 
+The solution is make sure that each thread primarily access the shared memory (L1 cache) and registers. Unfortunately, these are scarce resources and 
 \begin{table}[b]\centering%\normalsize
@@
 \end{tabular}
-\caption{Nvidia Quadro 6000; Memory bandwidths (BW) compared to the peak arithmetic bandwidth of 1.03 TB/s.}\label{tabbandwidth}
+\caption{Nvidia Quadro 6000; Memory bandwidths (BW) compared to the peak arithmetic bandwidth of 1.03 Tflop/s.}\label{tabbandwidth}
 \end{table}
@@
 \begin{figure}[t]\centering%
-\graphicsAI[drawing,width=0.7\linewidth]{gfx/mvdr_implementation.svg}%
+\graphicsAI[drawing,width=\linewidth]{gfx/mvdr_implementation.svg}%
 \caption{MVDR implementation.}\label{mvdr_implementation}

### 2013-01-22 ASA2013/asa2013.tex
la2pdf commit - 2013:01:22	12:23:56	(Tue22.Jan)

@@
 Graphics Processing Units (GPUs) is a technology tailored to perform image processing at incredible speeds.
-They are available off-the-shelf at reasonable prices, are very power efficient, and can be programmed in .
+They are available off-the-shelf at reasonable prices, are very power efficient, and can be programmed through .
 Unlike a CPU, which employ clever means to make a few threads run as fast as possible, a GPU does little to optimize the execution speed of a single thread, but in return allow thousands of them to run simultaneously.
@@
 
-\subsection{MVDR beamforming in time domain}
+\subsection{MVDR beamforming in the time domain}
 
@@
 
-\subsection{MVDR beamspace}
+\subsection{MVDR beamforming in the frequency domain}
 
@@
 The LCA and MVDR uses essentially the same optimization criterion, but whereas MVDR searches a continuous solution space for the best set of weights to apply, the LCA beamformer only have $P$ solutions to select from.
-However, once the LCA have 10-20 well designed windows to choose from its performance quickly converge towards the MVDR beamformer's performance.
-The difference in computational complexity, however, is
+However, once the LCA have 10-20 well designed windows to choose from the its ability to create high quality images quickly approaches the quality level of the MVDR beamformer.
+
+Although LCA promotes good image quality, its two most attractive traits is perhaps its simplicity and low computational complexity.
+The LCA is essentially a delay-and-sum (DAS) beamformer where the delay step is carried out as usual, but where the weighting and summing is repeated for each of the $P$ windows.
+In its basic form its computational complexity is then of O($M\,P$).
 
@@
 
-Throughout our work with porting the MVDR beamformer to a GPU, one issue has been particularly difficult to solve: The GPU will only perform efficiently when each computing thread is very computationally intense.
+Throughout our work with porting the MVDR beamformer to a GPU, one challenge has been particularly difficult to solve: The GPU will only perform efficiently when each computing thread is very computationally intense.
 This should be apparent by studying Table \ref{tabbandwidth}.

### 2013-01-22 HRUIonGPU/jrnl_article.tex
uffc article

--- 2012-12-20 HRUIonGPU/jrnl_article.tex
@@
 Different approaches to low-complexity Capon beamforming have therefore recently been proposed \cite{Synnevag2011, Asl2012, Jensen2012, Kim}.
-In this paper we have selected to implement the beamspace transformation, described by Nilsen and Hafizovic \cite{Nilsen2009}, on the GPU to maintain high framerates as the channel count increases.
+In this paper we will make use of the beamspace transformation, described by Nilsen et al.
+\cite{Nilsen2009}, in order to maintain high framerates as the channel count increases.
 The beamspace transformation matches well with the focused transmit beams used in cardiac ultrasound imaging.
@@
 The conventional form of array beamforming in medical ultrasound imaging is delay-and-sum (DAS).
-Here, the signal $x_m[n]$ recorded at element number $m \in [0,M-1]$ at time $n$ is delayed to broadside with an appropriate delay $\Delta_m[n]$.
+Here, the signal recorded at element number $m \in [0,M-1]$ at time $n$ is delayed with an appropriate delay $\Delta_m[n]$ to focus and steer the ultrasound beam in space.
+This delayed data is denoted $x_m[n]$.
 The beamformer output, $z[n]$, is then the sum of all $M$ channels weighted with a per-element factor $w_m$,
 \begin{align}
-z[n] = \sum_{m = 0}^{M-1}w_m^*x_m[n - \Delta_m[n]] = \vec{w}^H\vec{x}^{\Delta}[n]. \label{eq:z}
+z[n] = \sum_{m = 0}^{M-1}w_m^*x_m[n] = \vec{w}^H\vec{x}[n]. \label{eq:z}
 \end{align}
 The weight vector or window $\vec{w}$ is usually real, and is applied to trade resolution for a better signal-to-noise ratio (SNR).
-Even if different windows can be used in different areas of the images they have traditionally been predefined.
+And even if different windows can be used in different areas of the images they have traditionally been predefined.
 
@@
 \begin{align}
-\mat{\breve{R}}[n] = \frac{1}{N_LN_K}\sum_{n'=n-K}^{n+K} \sum_{l=0}^{N_L-1} (\vec{x}_l^{\Delta}[n'])(\vec{x}_l^{\Delta}[n'])^H,\label{eq:R}
-\end{align}
-where  $N_K = 2K + 1$ is the number of samples included in time, $\vec{x}_l^{\Delta}$ is the $l^\text{th}$ subarray of length $L$ $[x_l^{\Delta}[n], \dotso, x_{l+L-1}^{\Delta}[n]]$ and $N_L = M-L+1$ is the number of subarrays.
+\mat{\breve{R}}[n] = \frac{1}{N_LN_K}\sum_{n'=n-K}^{n+K} \sum_{l=0}^{N_L-1} \vec{x}_l[n']\vec{x}_l[n']^H,\label{eq:R}
+\end{align}
+where  $N_K = 2K + 1$ is the number of samples included in time, $\vec{x}_l = [x_l[n], \dotso, x_{l+L-1}[n]]$ is the $l^\text{th}$ subarray of length $L$ and $N_L = M-L+1$ is the number of subarrays.
 Note that the covariance matrix dimension is $L \times L$.
@@
 \begin{align}\label{eq:diag}
-\mat{\hat{R}}[n] =  f(\vec{x}^{\Delta}[n]) = \mat{\breve{R}}[n] + \epsilon[n]\mat{I}.
+\mat{\hat{R}}[n] =  f(\vec{x}[n]) = \mat{\breve{R}}[n] + \epsilon[n]\mat{I}.
 \end{align}
@@
 has been applied in much of the resent literature on Capon beamforming for medical ultrasound imaging \cite{Synnevag2007, Nilsen2009, Wang2009, Mehdizadeh2012}.
-This form of adaptive diagonal loading has also been applied to HF and VHF antenna processing \cite{Featherstone1997b}.
+It has also been applied to HF and VHF antenna processing \cite{Featherstone1997b}.
 
@@
 \end{align}
-where the steering vector $\vec{a} = \vec{1}_L = [1_0, 1_1, ..., 1_{L-1}]^T$ since $\vec{x}^\Delta$ is pre-delayed.
+where the steering vector $\vec{a} = \vec{1}_L = [1_0, 1_1, ..., 1_{L-1}]^T$ since $\vec{x}$ is pre-delayed.
 
@@
 \begin{align}
-z[n] &= \frac{1}{N_L}\vec{w}[n]^H \sum_{l=0}^{N_L-1} \vec{x}_l^{\Delta}[n] \label{eq:z_mv}\\
-&= 1/N_L \, \text{CONV}_m(\vec{w}[n], \vec{1}_{N_L})^H\vec{x}^{\Delta}[n] \label{eq:z_mv2},
+z[n] &= \frac{1}{N_L}\vec{w}[n]^H \sum_{l=0}^{N_L-1} \vec{x}_l[n] \label{eq:z_mv}\\
+&= 1/N_L \, \text{CONV}_m(\vec{w}[n], \vec{1}_{N_L})^H\vec{x}[n] \label{eq:z_mv2},
 \end{align}
@@
 For an active, pulsed system like medical ultrasound imaging, the available data from which $\mat{\hat{R}}$ can be estimated is quite limited.
-The resulting sample covariance matrix might therefore be both inaccurate, and poorly conditioned.
+Because of this, the resulting sample covariance matrix is often found to be both inaccurate, and poorly conditioned.
 Another challenge with applying Capon to active systems is the effect of signal cancellation, where high correlation between signal and interference favors weights that cancel out the signal \cite{Reddy1987}.
@@
 All of these constraints combined pose a real challenge as the target problem has to map on to the parallel framework both on a grid, block, thread and memory consumption level in order to benefits from parallel acceleration.
-In the next section, we will describe how this mapping should be done for Capon beamforming of cardiac ultrasound data.
+In the next section, we will describe how this mapping could be done for the Capon beamforming algorithm.
 
@@
 Continuing with the calculation of one weight vector, the algorithm can be broken down into three main steps that, to some extent, can be divided further into a set of parallel processes.
-That is, for each data vector $\vec{x}^\Delta$ in the image:
+That is, for each data vector $\vec{x}$ in the image:
 \begin{align*}
@@
 The computational complexity of these three steps is $O(N_KN_LL^2)$, $O(L^3)$, and $O(L)$ respectively.
-As $L$ increases, the first two steps quickly become unmanageable, and real-time performance can only be achieved by exploring novel solutions to reduce complexity.
+As $M$ increases (and thereby $L$), the first two steps quickly become unmanageable, and real-time performance can only be achieved by exploring novel solutions to reduce complexity.
 
@@
 \mat{\breve{R}}_{i,j}[n] &=  \mat{\breve{R}}_{i-1,j-1}[n]  \nonumber \\
-&+ \sum_{n'=n-K}^{n+K} (x_{i+N_L-1}^{\Delta}[n']x_{j+N_L-1}^{\Delta}[n']^* \nonumber \\
-&- x_{i-1}^{\Delta}[n']x_{j-1}^{\Delta}[n']^H). \label{eq:sliding}
+&+ \sum_{n'=n-K}^{n+K} (x_{i+N_L-1}[n']x_{j+N_L-1}[n']^* \nonumber \\
+&- x_{i-1}[n']x_{j-1}[n']^H). \label{eq:sliding}
 \end{align}
@@
 \begin{align}
-\mat{\breve{R}}[n] &= \mat{\breve{R}}[n-1] + \sum_{l=0}^{N_L-1} (\vec{x}^{\Delta}_l[n+K]\vec{x}^{\Delta}_l[n+K]^H \nonumber \\
-&- \vec{x}^{\Delta}[n-K-1]\vec{x}^{\Delta}[n-K-1]^H).
+\mat{\breve{R}}[n] &= \mat{\breve{R}}[n-1] + \sum_{l=0}^{N_L-1} (\vec{x}_l[n+K]\vec{x}_l[n+K]^H \nonumber \\
+&- \vec{x}[n-K-1]\vec{x}[n-K-1]^H).
 \end{align}
 %- \mat{\breve{R}}[n-K-1] + \mat{\breve{R}}[n-K].
-Both sliding across subarrays and in time can be realized simultaneously if we hold a $N_K$-average in time of the outer products $\vec{x}^{\Delta}[n]\vec{x}^{\Delta}[n]^H$ in memory and calculate $\mat{\hat{R}}$ averaged over subarrays from this buffer (cite JOE?).
-However, since $N_K$ typically is small, and we have a limited amount of memory available per thread, we have chosen to focus on sliding across subarrays only (\ref{eq:sliding}).
+Both sliding across subarrays and in time can be realized simultaneously if we hold a $N_K$-average in time of the outer products $\vec{x}[n]\vec{x}[n]^H$ in memory and calculate $\mat{\hat{R}}$ averaged over subarrays from this buffer (cite JOE?).
+However, since $N_K$ typically is small, and we have a limited amount of memory available per thread, we focus on sliding across subarrays only.
 In that case, the complexity of calculating $\mat{\hat{R}}$ is reduced to $O(N_KL^2)$.
@@
 With our choice of calculating $\mat{\hat{R}}$, calculations are still independent across time, but not along diagonals in $\mat{\hat{R}}$.
-We therefore have selected to use one thread per diagonal in the upper triangle of $\mat{\hat{R}}$, and for small values of $L$, several covariance matrices are calculated per thread block (Fig.
+One thread is therefore used per diagonal in the upper triangle of $\mat{\hat{R}}$, and for small values of $L$, several covariance matrices are calculated per thread block (Fig.
 \ref{fig:gpulayout}).
@@
 Taking symmetry into account, only the upper half of $\mat{\hat{R}}$ needs to be calculated.
-However, the way we have decided to organize the GPU compute threads per $\mat{\hat{R}}$, the lower triangle can be calculated with little overhead on the GPU.
-This also means that we do not rely on a symmetric solver to solve $\mat{\hat{R}}\vec{b} = \vec{1}_L$, even if that could have saved some instructions.
-
-Diagonal loading as a constant value can be added with marginal cost.
+However, the GPU compute threads per $\mat{\hat{R}}$ are organized in such a way that the lower triangle can be calculated with little overhead.
+This also means that no special solver is needed in order to solve $\mat{\hat{R}}\vec{b} = \vec{1}_L$, even if a symmetric solver could have saved some instructions.
+
+Diagonal loading as a constant value can be added with negligible cost.
 For the adaptive diagonal loading factor in (\ref{eq:diag}), the trace of $\mat{\hat{R}}$ has to be known before diagonal elements are written back to global memory.
 Since one row for all covariance matrices is written to global memory per kernel iteration, the diagonal elements need to be held in a local memory until the trace is accumulated.
-When it is found, scattered writes of the diagonal elements to global memory has to be performed.
+When it is found, scattered writes of the scaled diagonal elements to global memory has to be performed.
 
@@
 For the results presented in this paper, we have used an GPU implementation of Gauss-Jordan (GJ) elimination made by Nvidia to solve $\mat{\hat{R}}\vec{b} = \vec{1}_L$, where $\vec{b} = \mat{\hat{R}}^{-1}\vec{1}_L$.
-The solver is open for registered CUDA developers.%For smaller systems, $L < 10$, a direct solver using one thread per system is preferable.
+The solver has been made available for registered CUDA developers.%For smaller systems, $L < 10$, a direct solver using one thread per system is preferable.
 For this purpose we have implemented Cramer's rule for $2 \times 2$, $3 \times 3$ and $4 \times 4$ matrices.
@@
 \begin{align}\label{eq:subsum}
-\breve{x}_{r}^{\Delta} = \sum_{k=r}^{r+N_L-1}x_{k}^{\Delta} \text{ for } r \in [0, L-1].
+\breve{x}_{r} = \sum_{k=r}^{r+N_L-1}x_{k} \text{ for } r \in [0, L-1].
 \end{align}
@@
 \begin{align}
-\vec{x}_{BS}[n] &= \mat{B}\vec{x}^{\Delta}[n], \ \ \ \mat{B} \in \mathbb{C}^{N,M}\\
+\vec{x}_{BS}[n] &= \mat{B}\vec{x}[n], \ \ \ \mat{B} \in \mathbb{C}^{N,M}\\
 [\mat{B}]_{p,q} &= \frac{1}{\sqrt{M}}e^{-j 2 \pi p q / M}.
@@
 %Derivation of sliding beamspace transformation. 
-Given a $M$ long data vector divided into $N_L$ subarrays, the beamspace transformation of the $l^{th}$ subarray is given as:
+Given a $M$ long data vector divided into $N_L$ subarrays, the beamspace transformation of the $l^{th}$ subarray for the $k^{th}$ beam is:
 \begin{align}\label{eq:bs_formula}
@@
 As presented in (\ref{eq:diag}), adaptive diagonal loading is performed by loading the sample covariance matrix with the average energy per element, hence $trace\{\mat{\hat{R}}\}/L$.
-If adaptive diagonal loading is applied after data has been transformed into beamspace, a high percentage of the total energy is mapped to a small subset of beams.
-Thus the average energy per beam is higher than the average energy per element, if the system uses focused transmit beams.
+If adaptive diagonal loading is applied after data have been transformed into beamspace, a high percentage of the total energy is mapped to a small subset of beams.
+Thus the average energy per beam is higher than the average energy was per element, if the system uses focused transmit beams.
 This means that less diagonal loading (smaller $d$) has to be applied in beamspace than in element space.
@@
 The presented image is the first frame in a series of images where motion is imposed on both circles, point targets and the probe.
-This simulation is presented in an attached video.
+The entire simulation is presented in an attached video.
 
@@
 The image was acquired using a experimental scanner at MI-Lab in Trondheim using a 3.5 MHz phased array in harmonic mode.
-The image is extracted from a cardiac cycle, that can be viewed in its entirety in the attached videos.
+The image is extracted from a cardiac cycle, that can be viewed in its entirety in an attached video.
 
 Both data sets have been processed with the following Capon beamformer settings: $L = 32$, $K = 2$, $d=0.1$.
-These settings have previously been shown to provide high resolution while maintaining DAS-like speckle \cite{Synnevag2007a} in ultrasound images.
+These settings have previously been shown to provide high resolution while maintaining DAS-like speckle in ultrasound images \cite{Synnevag2007a}.
 For the beamspace processing the following parameters have been used: $L=32$, $K=2$, $d=0.01$, $N_b=3$.

### 2013-01-22 ASA2013/asa2013.tex

@@
 Graphics Processing Units (GPUs) is a technology tailored to perform image processing at incredible speeds.
-They are available off-the-shelf at reasonable prices, are very power efficient, and can be programmed through .
-Unlike a CPU, which employ clever means to make a few threads run as fast as possible, a GPU does little to optimize the execution speed of a single thread, but in return allow thousands of them to run simultaneously.
+They are available off-the-shelf at reasonable prices, are very power efficient, and can be programmed to carry out a wide range of computational tasks, often displaying an order of magnitude speedups compared to comparable designs running on a CPU.
+Initially, GPUs seem perfectly suited for image reconstruction algorithms, because usually each pixel can be computed independently and concurrently of one another.
+While this might be true for simple techniques such as conventional beamformers, the more complex per-pixel computations present in adaptive techniques are harder for the GPU to handle.
 
-This may seem perfectly suited for image reconstruction algorithms, because usually each pixel can be computed independently and concurrently of one another.
-While this might be true for simple techniques such as conventional beamformers, the more complex per-pixel computations present in adaptive techniques are harder for the GPU to handle.
+%Unlike a CPU, which employ clever means to make a few threads run as fast as possible, a GPU does little to optimize the execution speed of a single thread, but in return allow thousands of them to run simultaneously.
+
 Through a case study of on such technique, the Minimum Variance Distortionless Response (MVDR) beamformer, we identify these challenges and explain how they were overcome.
+The MVDR technique display a potent potential for improving image quality over simple conventional methods, but is far more computationally complex since a spatial covariance matrix must be estimated (of O($L$)) and inverted for each image pixel.
+While most studies tend to focus on reducing the dimensionality of the convariance matrix to
+
 The end design is more than 2 orders of magnitude faster than a C implementation we started off with, it took considerable effort to get there.
-
-<< PREVIOUS WORK >>
-
+\\
+<< MOTIVATION >>
+\\
+Comparable can be found in ultrasound imaging, where Chen et.
+al have investigated porting a time domain MVDR method to a GPU \cite{Chen2009,Chen2011,Chen2011a,}.
+Unlike their work, however, we do not impose the restriction of the input data to be real, nor do we approximate the
+\\
 This article is outlined as follows: In section \ref{methods} we will present the theoretical background on MVDR operated in time and frequency domain, as well as relating this to the LCA beamformer.

### 2013-01-22 ASA2013/asa2013.tex
la2pdf commit - 2013:01:22	13:33:31	(Tue22.Jan)

@@
 They are available off-the-shelf at reasonable prices, are very power efficient, and can be programmed to carry out a wide range of computational tasks, often displaying an order of magnitude speedups compared to comparable designs running on a CPU.
-Initially, GPUs seem perfectly suited for image reconstruction algorithms, because usually each pixel can be computed independently and concurrently of one another.
-While this might be true for simple techniques such as conventional beamformers, the more complex per-pixel computations present in adaptive techniques are harder for the GPU to handle.
+Initially, GPU may seem perfectly suited for image reconstruction algorithms, because usually each pixel can be computed independently and concurrently of one another.
+While this is true for simple techniques such as conventional beamformers, the more complex per-pixel computations present in adaptive techniques are harder for the GPU to handle.
 
@@
 Through a case study of on such technique, the Minimum Variance Distortionless Response (MVDR) beamformer, we identify these challenges and explain how they were overcome.
-The MVDR technique display a potent potential for improving image quality over simple conventional methods, but is far more computationally complex since a spatial covariance matrix must be estimated (of O($L$)) and inverted for each image pixel.
-While most studies tend to focus on reducing the dimensionality of the convariance matrix to
+The MVDR is a technique that displays a great potential for improving image quality over simple conventional methods, but is far more computationally complex.
+This is because a spatial covariance matrix must be estimated (of O($M$) and inverted for each image pixel.
+For systems with less than 20-30 channels, the estimation step dominates,   %Unlike other studies, w While most studies tend to focus on reducing the dimensionality of the convariance matrix to
+
+ .
 
@@
 \\
-Comparable can be found in ultrasound imaging, where Chen et.
+Comparable can be found e.g.
+in ultrasound imaging, where Chen et.
 al have investigated porting a time domain MVDR method to a GPU \cite{Chen2009,Chen2011,Chen2011a,}.
-Unlike their work, however, we do not impose the restriction of the input data to be real, nor do we approximate the
+Unlike their work, however, we do not impose the restriction of the input data to be real, nor do we approach larger systems by enforcing Toeplitz structure of the covariance matrix.
 \\
@@
 Finally, a fraction of the total power of $\eR[n]$ is added to its diagonal~\cite{Synnevag2007}.
+
+O($N_K N_L L^2$) and the inversion of O($L^3$)
 

### 2013-01-22 ASA2013/asa2013.tex
la2pdf commit - 2013:01:22	14:40:42	(Tue22.Jan)
la2pdf commit - 2013:01:22	15:37:25	(Tue22.Jan)

@@
 \newcommand*\Y{\mat Y}
+\renewcommand*\B{\mat B}
 
@@
 
-O($N_K N_L L^2$) and the inversion of O($L^3$)
+Complexity: O($N_K N_L L^2$) and the inversion of O($L^3$)
 
@@
 
-The Low Complexity Adaptive (LCA) beamformer is another take on the optimization problem in (\ref{mvdr_def}).
-Instead of calculating the weights like in (\ref{weights}), it simply computes the beamformer output for each window in a collection of such, and finally applies the one that best fulfils the given optimization criterion.
-If we let $\w_p$ be the $p$'th of a total of $P$ windows, then this can be formulated as:
+For larger systems the inversion step will completely dominate the computation time.
+Although the dimensionality of the coveriance matrix was reduced to $L$ due to subarray averaging, we can reduce it further by moving to the frequency domain.
+If the data is narrowband enough there is a very close relation between the frequency conent and angles of arrival in the spatial domain; the zero-frequency component will correspond to broadside, and    component corresponding to broadside and increasing frequencies  This is usually handled by reducing the dimensionality of the covariance matrix to
+
 \begin{gather}
-\argmin{p}\ E\Big\{\big|\w_p\H\x[n]\big|^2\Big\} \qquad\text{subject to}\qquad \w_p\H\1 = 1\label{lca_def}. 
-% \vec w[n] = \frac{\Ri[n]\1}{\1\T\Ri[n]\1},\label{weights}
+\x_\text{BS}[n] = \B\x[n], \qquad \B\in \mathbb{C}^{N,M}, \qquad \text{where} \qquad [\B]_{p,q} = \frac{1}{\sqrt{M}} e^{-j2\pi\,p\,q/M}.\label{beamspace}
 \end{gather}
-Note that (\ref{mvdr_ref}) and (\ref{lca_ref}) is essentially the same criterion.
 
-<< AVERAGING >>
+Complexity: To frequency domain: O($N_b\,L$). Build $R_\text{BS}$, O($N_b^2\,N_L\,N_K$). Inversion: O($N_b^3$).
 
@@
 In its basic form its computational complexity is then of O($M\,P$).
-
-<< AVERAGING >>
 
@@
 
-\begin{figure}[h]\centering%
+\begin{figure}[!h]\centering%
 \graphicsAI[drawing,width=\linewidth]{gfx/plot_holmengraa_L16_Navg1.pdf}%

### 2013-01-22 ASA2013/asa2013.tex
la2pdf commit - 2013:01:22	16:21:12	(Tue22.Jan)

@@
 Graphics Processing Units (GPUs) is a technology tailored to perform image processing at incredible speeds.
-They are available off-the-shelf at reasonable prices, are very power efficient, and can be programmed to carry out a wide range of computational tasks, often displaying an order of magnitude speedups compared to comparable designs running on a CPU.
-Initially, GPU may seem perfectly suited for image reconstruction algorithms, because usually each pixel can be computed independently and concurrently of one another.
+They are available off-the-shelf at reasonable prices, are very power efficient, and can be programmed to carry out a wide range of computational tasks, often displaying an order of magnitude speedups compared to similar CPU designs.
+Initially, GPUs may seem perfectly suited for image reconstruction algorithms, because usually each pixel can be computed independently and concurrently of one another.
 While this is true for simple techniques such as conventional beamformers, the more complex per-pixel computations present in adaptive techniques are harder for the GPU to handle.

### 2013-01-22 ASA2013/asa2013.tex
la2pdf commit - 2013:01:22	16:54:53	(Tue22.Jan)

@@
 The MVDR is a technique that displays a great potential for improving image quality over simple conventional methods, but is far more computationally complex.
-This is because a spatial covariance matrix must be estimated (of O($M$) and inverted for each image pixel.
+This is because a spatial covariance matrix must be estimated and inverted for each image pixel.
 For systems with less than 20-30 channels, the estimation step dominates,   %Unlike other studies, w While most studies tend to focus on reducing the dimensionality of the convariance matrix to
@@
 \\
-This article is outlined as follows: In section \ref{methods} we will present the theoretical background on MVDR operated in time and frequency domain, as well as relating this to the LCA beamformer.
-As we move along we aim to mention complexity issues and suggest means to overcome them, but implementational aspect will first be presented in \ref{implementation}.
+This article is outlined as follows: In section \ref{methods} we will present the background on the MVDR method operated in time and frequency domain, and on the LCA beamformer.
+For each method we start with the mathematical formulation, then evaluate complexity issues, and finally give our thoughts on some implementational aspects.
+The results are collected in \ref{implementation}.
+Finally, we conclude in \ref{discssion}.
 
@@
 Various weighting patterns exists here for trading lateral resolution for improved noise suppression (contrast), but one always ends up with a compromise between the two\todo{don't quite like this one}~\cite{Harris1978}.
+Adaptive beamformers target this limitation by computing a set of weights for each pixel that to better fit the dynamic nature of the incoming wavefield.
 
@@
 \end{figure}
-Adaptive beamformers target this limitation by computing a set of weights for each pixel that to better fit the dynamic nature of the incoming wavefield.
 The MVDR beamformer is one such method.

### 2013-01-22 ASA2013/asa2013.tex
la2pdf commit - 2013:01:22	17:33:39	(Tue22.Jan)

@@
 Various weighting patterns exists here for trading lateral resolution for improved noise suppression (contrast), but one always ends up with a compromise between the two\todo{don't quite like this one}~\cite{Harris1978}.
-Adaptive beamformers target this limitation by computing a set of weights for each pixel that to better fit the dynamic nature of the incoming wavefield.
 
@@
 \end{figure}
-The MVDR beamformer is one such method.
-It computes a set of weights that minimizes the noise and signal power accumulated by the array, but with a constraint of unity gain in the desired direction.
+The MVDR beamformer target this limitation by computing a set of weights for each pixel that to better fit the dynamic nature of the incoming wavefield.
+It computes the set of weights that minimizes the noise and signal power accumulated by the array, but with the constraint of unity gain for signal coming from the pixel of interest.
 When the data is delayed this is formulated as follows:
@@
 
-If we let $x_l[n]$ represent the datavector from subarray $l$\todo{replace with with a drawing?},
+In terms of computational complexity the two important  of the MVDR method the 
+Before If we let $x_l[n]$ represent the datavector from subarray $l$\todo{replace with with a drawing?},
 \begin{gather}

### 2013-01-22 HRUIonGPU/jrnl_article.tex
finally started

@@
 %
-\documentclass[journal]{IEEEtran}
+
+%\documentclass[journal]{IEEEtran}
+
+% Single column review mode:
+\documentclass[12pt,journal,onecolumn]{IEEEtran}
+\newcommand\CLASSINPUTbaselinestretch{2}
+\RequirePackage{calc}
+\RequirePackage{fp} 
+% Left margin = 1 inch + hoffset + oddsidemargin (or evensidemargin)
+% Adding the 2cm gutter width to the odd/even side margins:
+\setlength\hoffset{0pt}
+\setlength\oddsidemargin{4cm}       % 4cm margins on the left side
+\addtolength\oddsidemargin{-1in}    % Subtract the initial 1 inch
+\setlength\textwidth{21cm-8cm}    % A4 width (21cm) minus margins on either side
+
 %
@@
 The steering vectors in $\mat{B}$, the so-called Butler matrix, determine each beam's direction in space.
-We can therefore easily remove beams where no interference is present by removing rows from $\mat{B}$.
+We can therefore remove beams where no interference is present by removing rows from $\mat{B}$.
 
@@
 \end{align}
-We call this the sliding beamspace transformation (SBS), and it is equal to a normalized sliding DFT \cite{Lyons2003}. For $l=0$ a full computation has to be performed using (\ref{eq:bs_formula}).
+We denote this the sliding beamspace transformation (SBS), and it is equal to a normalized sliding DFT \cite{Lyons2003}. For $l=0$ a full computation has to be performed using (\ref{eq:bs_formula}).
 
@@
 (a) Shows the benchmark for a 64 elements array.
-(b) Shows the benchmark for a 96 elements array.}
+(b) Shows the benchmark for a 96 elements array.
+(TODO: Change this to be a function of $N_b$)}
 \label{fig:benchBS}
@@
 
-For the simulation, 16x lateral over-sampling has been used, in order to avoid under sampling of the two bright point scatterers. The cardiac recording only has $30\%$ lateral over sampling.
+For the simulation, 16x lateral over-sampling has been used, in order to avoid under sampling of the two bright point scatterers as they move.
+The cardiac recording only has $30\%$ lateral over sampling.
 
 \section{Trading resolution for speed}
-Add a comparison between beamspace and full capon using small L. Plot system resolution vs $L$ and $N_b$ for the two methods. ???
-Fig \ref{fig:speed_res_trade}.
+As we see from Fig.
+\ref{fig:bench} and \ref{fig:benchBS}, the execution time is reduced when either $L$ or $N_b$ is reduced.
+The question is how this impact the image quality.
+In Fig.
+\ref{fig:speed_res_trade} the minimum distance between the two point targets in Fig.
+\ref{fig:phantom} (while preserving a 6dB lower level in between) is plotted against $L$ (when processed with Capon) and $N_b$ (when processed with BS Capon).
+For high values the two methods are equal, but for lower values BS Capon provides higher resolution than what is possible with small subarrays.
+For this result, $d=0.01$ has been used on both methods.
 
@@
 \includegraphics[width=3in]{gfx/speed_res_trade.png}
-\caption{Trade-off between resolution and speed by reducing subarray size or reducing beamspace.}
+\caption{Trade-off between resolution and speed by reducing subarray size or reducing beamspace dimension (based on Fig.
+\ref{fig:phantom}).
+As $L$ or $N_b$ gets smaller, the execution time is reduced (Fig.
+\ref{fig:bench} and \ref{fig:benchBS}).
+When $N_b$ is varied, $L$ is fixed on 32.}
 \label{fig:speed_res_trade}
@@
 \section{Discussion}\label{sec:dis}
+We have presented a parallel implementation of both the Capon og BS Capon beamformer.
+
 From Fig.
@@
 
-A major part of the total execution time comes from solving systems of linear equations.
-Further research should therefore focus on making this contribution smaller.
-This can be done either by implementing faster solvers, or reducing the sample covariance matrix size in a novel way \cite{Nilsen2009}.
-%When doing such a reduction, including the pre-beamforming described above, one should investigate the loss of contrast and resolution compared with full Capon beamforming.
-As for now, we have shown that the Capon beamformer can be implemented in a GPU framework.
+A major part of the total execution time comes from solving systems of linear equations. This limitation is targeted by reducing the sample covariance matrix size in a novel way \cite{Nilsen2009}.
+
+We have shown that the Capon beamformer can be implemented in a GPU framework.
 And if we as in \cite{Chen2011} define real-time as 10 fps, all execution times  in Fig.

### 2013-01-22 ASA2013/asa2013.tex
la2pdf commit - 2013:01:22	18:08:51	(Tue22.Jan)

@@
 
-Graphics Processing Units (GPUs) is a technology tailored to perform image processing at incredible speeds.
-They are available off-the-shelf at reasonable prices, are very power efficient, and can be programmed to carry out a wide range of computational tasks, often displaying an order of magnitude speedups compared to similar CPU designs.
+Graphics Processing Units (GPUs) is a technology tailored to perform image processing as fast as possible.
+They are available off-the-shelf at reasonable prices, are very power efficient, and can be programmed to carry out a wide range of computational tasks, often displaying a speed increase of an order of magnitude compared to similar CPU designs.
 Initially, GPUs may seem perfectly suited for image reconstruction algorithms, because usually each pixel can be computed independently and concurrently of one another.
@@
 
-%Unlike a CPU, which employ clever means to make a few threads run as fast as possible, a GPU does little to optimize the execution speed of a single thread, but in return allow thousands of them to run simultaneously.
-
 Through a case study of on such technique, the Minimum Variance Distortionless Response (MVDR) beamformer, we identify these challenges and explain how they were overcome.
-The MVDR is a technique that displays a great potential for improving image quality over simple conventional methods, but is far more computationally complex.
+The MVDR is a technique that displays a notable potential for improving image quality over simple conventional methods, but is far more computationally complex.
 This is because a spatial covariance matrix must be estimated and inverted for each image pixel.
-For systems with less than 20-30 channels, the estimation step dominates,   %Unlike other studies, w While most studies tend to focus on reducing the dimensionality of the convariance matrix to
-
- .
+For systems with less than 20-30 channels the estimation step dominates, while the inversion step   %Unlike other studies, w While most studies tend to focus on reducing the dimensionality of the convariance matrix to
 
@@
 \\
-Comparable can be found e.g.
+Comparable work can be found e.g.
 in ultrasound imaging, where Chen et.
-al have investigated porting a time domain MVDR method to a GPU \cite{Chen2009,Chen2011,Chen2011a,}.
+al have investigated porting a time domain MVDR method to a GPU \cite{Chen2009,Chen2011,Chen2011a}.
 Unlike their work, however, we do not impose the restriction of the input data to be real, nor do we approach larger systems by enforcing Toeplitz structure of the covariance matrix.
@@
 \end{figure}
-The MVDR beamformer target this limitation by computing a set of weights for each pixel that to better fit the dynamic nature of the incoming wavefield.
+The MVDR beamformer target this limitation by adapting the weights for each pixel to best fit the dynamic nature of the incoming wavefield.
 It computes the set of weights that minimizes the noise and signal power accumulated by the array, but with the constraint of unity gain for signal coming from the pixel of interest.

### 2013-01-22 ASA2013/asa2013.tex
la2pdf commit - 2013:01:22	18:42:31	(Tue22.Jan)

@@
 Through a case study of on such technique, the Minimum Variance Distortionless Response (MVDR) beamformer, we identify these challenges and explain how they were overcome.
-The MVDR is a technique that displays a notable potential for improving image quality over simple conventional methods, but is far more computationally complex.
+The MVDR technique displays a notable potential for improving image quality over simple conventional methods, but is far more computationally complex.
 This is because a spatial covariance matrix must be estimated and inverted for each image pixel.
-For systems with less than 20-30 channels the estimation step dominates, while the inversion step   %Unlike other studies, w While most studies tend to focus on reducing the dimensionality of the convariance matrix to
+These steps take turn in being the most complex depending on system size.
+With less than 20-30 channels the algorithm spends most of its time with the estimation step, while at larger system sizes the inversion step completely dominates.
+For large systems one should try to bypass the estimation/inversion step altogether.
+
+Therefore we suggest two different approaches for     dominates the computation time, while for larger system sizes the the inversion step   %Unlike other studies, w While most studies tend to focus on reducing the dimensionality of the convariance matrix to
 

### 2013-01-22 ASA2013/asa2013.tex
la2pdf commit - 2013:01:22	19:18:52	(Tue22.Jan)

--- 2013-01-22 ASA2013/asa2013.tex
@@
 Through a case study of on such technique, the Minimum Variance Distortionless Response (MVDR) beamformer, we identify these challenges and explain how they were overcome.
-The MVDR is a technique that displays a notable potential for improving image quality over simple conventional methods, but is far more computationally complex.
-This is because a spatial covariance matrix must be estimated and inverted for each image pixel.
-For systems with less than 20-30 channels the estimation step dominates, while the inversion step   %Unlike other studies, w While most studies tend to focus on reducing the dimensionality of the convariance matrix to
+The MVDR technique is known for its notable potential for improving image quality over simple conventional methods, but it is also far more computationally complex.
+This is because a spatial covariance matrix must be estimated and inverted for each image pixel, the former dominating for systems with less than 20-30 channels, and the latter otherwise.
 
-The end design is more than 2 orders of magnitude faster than a C implementation we started off with, it took considerable effort to get there.
-\\
-<< MOTIVATION >>
-\\
+We propose three different solutions to the complexity issue depending on system size.
+For smaller systems we use the standard time-domain MVDR method, arithmetically optimize the estimation step, and show that image creation rates of more than 1\,Mpx/s can be achieved when this algorithm is run on a GPU.
+For medium sized systems we suggest applying the MVDR in the frequency domain (beamspace) instead, which allows for comparable processing rates at a negligible reduction in image quality.
+Finally, for large systems the estimation and inversion step can be avoided completely by utilising the Low Complexity Adaptive (LCA) beamformer.
+% 
+% 
+% The end design is more than 2 orders of magnitude faster than a C implementation we started off with, it took considerable effort to get there.
+% \\
+% << MOTIVATION >>
+% \\
+
 Comparable work can be found e.g.
@@
 Unlike their work, however, we do not impose the restriction of the input data to be real, nor do we approach larger systems by enforcing Toeplitz structure of the covariance matrix.
-\\
-This article is outlined as follows: In section \ref{methods} we will present the background on the MVDR method operated in time and frequency domain, and on the LCA beamformer.
-For each method we start with the mathematical formulation, then evaluate complexity issues, and finally give our thoughts on some implementational aspects.
+% \\
+
+This article is outlined as follows: In section \ref{methods} we will present background material on the MVDR method operated in time and frequency domain, and on the LCA beamformer.
+For each method we start with the mathematical formulation, then evaluate complexity issues, and finally give our thoughts on some implementation aspects.
 The results are collected in \ref{implementation}.
 Finally, we conclude in \ref{discssion}.
-
-% MVDR - two parts - estimating R and solving
-% less 32: 
-%
-
-% However, to a problem that is commonly met is not that the peak arithmetic performance is insufficient, but that it is difficult move data around fast enough to keep all the computing cores busy.
-We say that the algorithm is memory bound, as opposed to compute bound.
-
-% 
-% One of the hottest topics in high performance computing today is how Graphics Processing Units (GPUs) can be used to  are one of the hottest candidates for accelerating  .
-Their appeal in the gaming community makes them relatively cheap, readily available off-the-shelf, and most computers today already have one installed.
-% 
-% To improve the image quality active sonar imaging several data adaptive techniques have 
-% 
-% Most modern active sonar imaging systems operate by combining data from sensor arrays to , from which data is combined 
-% 
-% The Minimum Variance Distortionless Response (MVDR) beamformer is known for its potential to improve sonar image quality. 
-% 
-% Most active sonar systems today employ phased arrays to record and 
-% 
-% Adaptive beamforming have in recent years been applied to active sonar imaging by several authors. 
-% 
-% The potential that adaptive beamformers to improve active sonar image quality is  ers is that theimprove image quality by utilising the apriori knowledge present in the backscattered wavefield.
-has for improving active sonar image quality has been demonstrated in e.g.
-[].
-are known for the potential to improve image quality in to When using adaptive beamformers to improbe
 

### 2013-01-22 ASA2013/asa2013.tex
la2pdf commit - 2013:01:22	19:53:42	(Tue22.Jan)

@@
 
-\usepackage[utf8x]{inputenc}
+% \usepackage[utf8x]{inputenc}
 \usepackage[table,dvipsnames,svgnames]{xcolor}
@@
    \setlength\marginparsep{1pt}
-   \setlength\marginparwidth{35pt}
+   \setlength\marginparwidth{50pt}
    \newlength\marginparwidthsmall
@@
 \end{abstract}
+\vspace{2cm}
+(Forsiden genereres av ASA.)
 % \ \\
@@
 
+\newpage
 \section{Introduction}
@@
 For medium sized systems we suggest applying the MVDR in the frequency domain (beamspace) instead, which allows for comparable processing rates at a negligible reduction in image quality.
-Finally, for large systems the estimation and inversion step can be avoided completely by utilising the Low Complexity Adaptive (LCA) beamformer.
+Finally, for large systems the estimation and inversion step can be avoided completely by utilising the Low Complexity Adaptive (LCA) beamformer.\todo{disse tre første avsnittene burde sikkert vært kortet ned mye men...}
 % 
@@
 
-This article is outlined as follows: In section \ref{methods} we will present background material on the MVDR method operated in time and frequency domain, and on the LCA beamformer.
+This article is outlined as follows: First we will present\todo{denne formuleringstypen begynner aa bli oppbrukt...} background material on the MVDR method operated in time and frequency domain, and on the LCA beamformer.
 For each method we start with the mathematical formulation, then evaluate complexity issues, and finally give our thoughts on some implementation aspects.
 The results are collected in \ref{implementation}.
-Finally, we conclude in \ref{discssion}.
+Finally, we conclude in section \ref{discssion}.
 

### 2013-01-22 ASA2013/asa2013.tex
la2pdf commit - 2013:01:22	20:44:26	(Tue22.Jan)

@@
 Various weighting patterns exists here for trading lateral resolution for improved noise suppression (contrast), but one always ends up with a compromise between the two\todo{don't quite like this one}~\cite{Harris1978}.
+However, there exists a class of adaptive beamformers that target this limitation by adapting the weights to the impinging wavefield.
+The MVDR method is one such technique.
 
@@
 \end{figure}
-The MVDR beamformer target this limitation by adapting the weights for each pixel to best fit the dynamic nature of the incoming wavefield.
-It computes the set of weights that minimizes the noise and signal power accumulated by the array, but with the constraint of unity gain for signal coming from the pixel of interest.
-When the data is delayed this is formulated as follows:
+The basic idea behind the MVDR method is to compute the set of weights that minimizes the noise and signal power accumulated by the array, but with the constraint of unity gain for signals coming from the pixel of interest.
+When the data is delayed this can be formulated as follows:
 \begin{gather}
@@
 
-In terms of computational complexity the two important  of the MVDR method the 
-Before If we let $x_l[n]$ represent the datavector from subarray $l$\todo{replace with with a drawing?},
+In terms of computational complexity the diagonal loading factor is negligible, and will therefore be ignored from now on.
+Now let $x_l[n]$ represent the datavector from subarray $l$\todo{replace with with a drawing?},
 \begin{gather}

### 2013-01-22 ASA2013/asa2013.tex
la2pdf commit - 2013:01:22	21:17:56	(Tue22.Jan)

@@
 
-In terms of computational complexity the diagonal loading factor is negligible, and will therefore be ignored from now on.
-Now let $x_l[n]$ represent the datavector from subarray $l$\todo{replace with with a drawing?},
+To arrive at a formulation of how $\eR$ is estimated, we first split the array into subarrays of length $L$ that overlap on all but one sensor. Let $x_l[n]$ represent the datavector from subarray $l$,
 \begin{gather}
-\x_l[n] = \bmat{x_l[n] & x_{l+1}[n] & \dots & x_{l+L-1}[n]}\T,
+\x_l[n] = \bmat{x_l[n] & x_{l+1}[n] & \dots & x_{l+L-1}[n]}\T.
 \end{gather}
-then a sample covariance matrix formed with temporal and spatial averaging, $\eR$, is given as:
+Now the sample covariance matrix formed with temporal and spatial averaging, $\eR$, is given as:
 \begin{gather}
-\eR[n] =  \frac{1}{N_K N_L} \sumb{l=0}{M-L}\sumb{n'=n-K}{n+K} \x_l[n]\x_l\H[n] \in\mathbb{C}^{L,L},\label{spatialR}
+\eR[n] =  \frac{1}{N_K N_L} \sumb{l=0}{N_L-1}\sumb{n'=n-K}{n+K} \x_l[n]\x_l\H[n] \in\mathbb{C}^{L,L},\label{spatialR}
 \end{gather}
 where $N_K = 2K+1$ is the number of temporal samples to perform averaging over, and $N_L = M-L+1$ is the number of subarrays.
-Finally, a fraction of the total power of $\eR[n]$ is added to its diagonal~\cite{Synnevag2007}.
+In the actual design a fraction of the total power of $\eR[n]$ is finally added to its diagonal~\cite{Synnevag2007}, but this step will be ignored here due to its negligible contribution to runtime.
 
-Complexity: O($N_K N_L L^2$) and the inversion of O($L^3$)
+\subsubsection{Complexity and implementation}
+
+If (\ref{spatialR}) were to be implemented directly, then we note that the estimation step has a complexity of O($N_K N_L L^2$).
+By comparison the inversion is generally of O($L^3$), which implies that whenever $N_K\,N_L > L$ then the estimation step will dominate.
+Note that this inequality will increase as we increase the number of subarrays or temporal samples to perform averaging over.
 

### 2013-01-22 ASA2013/asa2013.tex
la2pdf commit - 2013:01:22	21:50:38	(Tue22.Jan)

@@
 
-To arrive at a formulation of how $\eR$ is estimated, we first split the array into subarrays of length $L$ that overlap on all but one sensor. Let $x_l[n]$ represent the datavector from subarray $l$,
+To arrive at a formulation of how $\eR$ can be estimated, we first split the array into subarrays of length $L$ that overlap on all but one sensor.
+If we let $x_l[n]$ represent the datavector from subarray $l$,
 \begin{gather}
@@
 \end{gather}
-Now the sample covariance matrix formed with temporal and spatial averaging, $\eR$, is given as:
+then the sample covariance matrix formed with temporal and spatial averaging, $\eR$, is given as:
 \begin{gather}
@@
 
-If (\ref{spatialR}) were to be implemented directly, then we note that the estimation step has a complexity of O($N_K N_L L^2$).
-By comparison the inversion is generally of O($L^3$), which implies that whenever $N_K\,N_L > L$ then the estimation step will dominate.
-Note that this inequality will increase as we increase the number of subarrays or temporal samples to perform averaging over.
+If (\ref{spatialR}) were to be implemented directly, then we note that the estimation step would have a complexity of O($N_K N_L L^2$).
+By comparison performing a Gauss Jordan inversion is generally of O($L^3$), which implies that whenever $N_K\,N_L = N_K(M-L+1) \gg L$ then the estimation step will dominate.
+This tend to be true in active systems because it is usually required to set $L<M/2$ and $N_K\ge>3$ to MVDR yield the expected results.
+
+\begin{figure}[t]\centering%
+\graphicsAI[drawing,width=\linewidth]{gfx/mvdr_implementation.svg}%
+\caption{MVDR implementation.}\label{mvdr_implementation}
+\end{figure}
 
@@
 \end{table}
-
-\begin{figure}[t]\centering%
-\graphicsAI[drawing,width=\linewidth]{gfx/mvdr_implementation.svg}%
-\caption{MVDR implementation.}\label{mvdr_implementation}
-\end{figure}
 

### 2013-01-22 ASA2013/asa2013.tex
la2pdf commit - 2013:01:22	22:22:37	(Tue22.Jan)

@@
 
+To summarize as in Figure \ref{mvdr_beamforming}: For every pixel in the image we estimate $\eR$ (\ref{spatialR}), solve the linear equation $\eRi\1$ (\ref{weights}), the finally substitute some to get the beamformer output (\ref{mvdr_def}-\ref{weights}).
+
 \subsubsection{Complexity and implementation}
@@
 If (\ref{spatialR}) were to be implemented directly, then we note that the estimation step would have a complexity of O($N_K N_L L^2$).
-By comparison performing a Gauss Jordan inversion is generally of O($L^3$), which implies that whenever $N_K\,N_L = N_K(M-L+1) \gg L$ then the estimation step will dominate.
-This tend to be true in active systems because it is usually required to set $L<M/2$ and $N_K\ge>3$ to MVDR yield the expected results.
+By comparison, performing a Gauss Jordan inversion is generally of O($L^3$), which implies that whenever $N_K\,N_L = N_K(M-L+1) \gg L$ the estimation step will dominate.
+This tend to be true in active systems because we often want to set set $L<M/2$ and $N_K\ge>3$ for the MVDR to be robust enough to yield the expected results.
 
-\begin{figure}[t]\centering%
-\graphicsAI[drawing,width=\linewidth]{gfx/mvdr_implementation.svg}%
-\caption{MVDR implementation.}\label{mvdr_implementation}
-\end{figure}
+The good news is that this step can greatly benefit from both arithmetic optimizations and parallel computing power. By carefully evaluating (\ref{spatialR}) we can see that:
+\begin{itemize}
+\item 
+\end{itemize}
 
@@
 
+\begin{figure}[t]\centering%
+\graphicsAI[drawing,width=\linewidth]{gfx/mvdr_implementation.svg}%
+\caption{MVDR implementation.}\label{mvdr_implementation}
+\end{figure}
+
+Some key aspects of our time domain MVDR implementation is illustrated in Figure \ref{mvdr_implementation}.
+
+First we make each SM compute entire range lines of pixels, then load the subset of $\x$ that all these pixels depend upon into shared memory.
+This lets us perform temporal averaging without the need for reading additional channel data.
+
+Second we assign $L$ threads per pixel to traverse the diagonals of $\eR$.
+On the top row a full computation is carried out, then that entire row is saved back to global memory following a collective access pattern that maximize global memory bandwidth.
+For subsequent rows the threads move along the diagonals while performing iterative summations; the result from the previous element on the diagonal is updated by adding and removing the correlation coefficients that enters and exits the sum, respectively.
+To minimize memory consumption, we compute the coefficients again every time we need them.
+When a thread has finished up a diagonal, we have them wrap around to compute one of the diagonals in the lower triangular of $\eR$.
+Since $\eR$ is conjugate symmetric, the values in the leftmost column is obtained by a complex conjugate copy of the relevant value in the first row.
+Combined these steps balances the load evenly on all threads, is almost completely free of arithmetic redundancy, and consumes less memory.
+
+These challenges are very closely linked.
+The Quadro 6000 architecture is of compute capability 2.0, which means there are 48\,kB of shared memory (L1 cache) and 128\,kB of registers per SM~\cite{Nvidia2012}.
+This memory is shared by all active threads on that SM, a number that should be no less than 768~\cite{Nvidia2012a}.
+This will expose a sufficient level of data parallelism to ensure that memory latency is completely hidden (Little's law).
+By dividing the shared memory evenly on all 768 threads we find that each thread should store no more than 8 single precision complex numbers in shared memory\todo{introduce memory structure -shared for data shared between threads}, and 24 stored in registers.
+This should make it apparent why computing a single pixel per thread is a bad idea, and why we need to keep each thread as light on memory consumption as possible.
+
 % Most of the challenges we faced when porting the MVDR to a GPU are caused by a single troublesome fact: GPU prefers threads that computationally intensive but very light on memory consumption.
+
+\section{Results}
+
+\begin{figure}[!h]\centering%
+\graphicsAI[drawing,width=\linewidth]{gfx/plot_holmengraa_L16_Navg1.pdf}%
+\caption{Holmengraa.}\label{mvdr_implementation}
+\end{figure}
+
+\section{Discussion}
 
@@
 \end{table}
-
-\section{Results}
-
-\begin{figure}[!h]\centering%
-\graphicsAI[drawing,width=\linewidth]{gfx/plot_holmengraa_L16_Navg1.pdf}%
-\caption{Holmengraa.}\label{mvdr_implementation}
-\end{figure}
-
-\section{Discussion}
 

### 2013-01-22 ASA2013/asa2013.tex
la2pdf commit - 2013:01:22	22:58:06	(Tue22.Jan)

@@
 
-The good news is that this step can greatly benefit from both arithmetic optimizations and parallel computing power. By carefully evaluating (\ref{spatialR}) we can see that:
+First, we target this limitation arithmetic optimizations and parallel computing power. In particular, we may infer from (\ref{spatialR}) that:
 \begin{itemize}
-\item 
+\item The subarray covariance matrices $R_l[n] = \x_l[n]\x_l\H[n]$ are Hermitian semi-definite, which means that we can get away with computing only one triagonal in them.
+\item When summing $R_l[n]$ most of the multiplications can be computed only once, and most sums need only be updated iteratively.
 \end{itemize}
@@
 
-For larger systems the inversion step will completely dominate the computation time.
-Although the dimensionality of the coveriance matrix was reduced to $L$ due to subarray averaging, we can reduce it further by moving to the frequency domain.
-If the data is narrowband enough there is a very close relation between the frequency conent and angles of arrival in the spatial domain; the zero-frequency component will correspond to broadside, and    component corresponding to broadside and increasing frequencies  This is usually handled by reducing the dimensionality of the covariance matrix to
+As we increase system size the inversion step will become the main bottleneck.
+Although the dimensionality of the covariance matrix was reduced to $L$ due to subarray averaging, we can reduce it further by moving to the frequency domain.
+For narrowband  relation between the frequency conent and angles of arrival in the spatial domain; the zero-frequency component will correspond to broadside, and larger frequency components will correspond to increasing .
+component corresponding to broadside and increasing frequencies  This is usually handled by reducing the dimensionality of the covariance matrix to
 
@@
 The LCA and MVDR uses essentially the same optimization criterion, but whereas MVDR searches a continuous solution space for the best set of weights to apply, the LCA beamformer only have $P$ solutions to select from.
-However, once the LCA have 10-20 well designed windows to choose from the its ability to create high quality images quickly approaches the quality level of the MVDR beamformer.
+However, we have found that once the LCA has 10-20 well designed windows available the quality of the images it produces quickly approaches that of the MVDR beamformer.
 
@@
 Second we assign $L$ threads per pixel to traverse the diagonals of $\eR$.
-On the top row a full computation is carried out, then that entire row is saved back to global memory following a collective access pattern that maximize global memory bandwidth.
+On the top row a full computation is carried out, then that entire row is saved back to global memory following a collective access pattern that maximizes global memory bandwidth.
 For subsequent rows the threads move along the diagonals while performing iterative summations; the result from the previous element on the diagonal is updated by adding and removing the correlation coefficients that enters and exits the sum, respectively.

### 2013-01-22 ASA2013/asa2013.tex
la2pdf commit - 2013:01:22	23:42:09	(Tue22.Jan)

@@
 
-If (\ref{spatialR}) were to be implemented directly, then we note that the estimation step would have a complexity of O($N_K N_L L^2$).
+If (\ref{spatialR}) were to be implemented directly, we note that the estimation step would have a complexity of O($N_K N_L L^2$).
 By comparison, performing a Gauss Jordan inversion is generally of O($L^3$), which implies that whenever $N_K\,N_L = N_K(M-L+1) \gg L$ the estimation step will dominate.
-This tend to be true in active systems because we often want to set set $L<M/2$ and $N_K\ge>3$ for the MVDR to be robust enough to yield the expected results.
+This tends to be true in active systems because we often want to set $L<M/2$ and $N_K\ge>3$ for the MVDR to be robust enough to yield the expected results.
 
-First, we target this limitation arithmetic optimizations and parallel computing power. In particular, we may infer from (\ref{spatialR}) that:
+A few properties let us improve these matters notably. First, we may infer from (\ref{spatialR}) that:
 \begin{itemize}
@@
 \end{itemize}
+Second, each of the diagonals of $R_l[n]\ \forall\ l$ can be computed independently, which is good news for our GPU design.
 
@@
 
-As we increase system size the inversion step will become the main bottleneck.
-Although the dimensionality of the covariance matrix was reduced to $L$ due to subarray averaging, we can reduce it further by moving to the frequency domain.
-For narrowband  relation between the frequency conent and angles of arrival in the spatial domain; the zero-frequency component will correspond to broadside, and larger frequency components will correspond to increasing .
-component corresponding to broadside and increasing frequencies  This is usually handled by reducing the dimensionality of the covariance matrix to
+As we increase system size the inversion step will become more of a bottleneck.
+Although the dimensionality of the covariance matrix is reduced to $L$ due to subarray averaging, we can reduce it further by performing MVDR beamforming.
+The underlying concept here that for narrowband data there will be a direct relation between the Fourier transform across the array angles of arrival in the spatial domain; the zero-frequency correspond to broadside, and increasing frequencies map to monotonously increasing opening angles.
+Since the combined transmit and receive response is highly directive, why would we want to process angles where there are hardly any information?
 
+Beamspace MVDR very closely resembles time domain MVDR. The main difference is the need to Fourier transform the data:
 \begin{gather}
-\x_\text{BS}[n] = \B\x[n], \qquad \B\in \mathbb{C}^{N,M}, \qquad \text{where} \qquad [\B]_{p,q} = \frac{1}{\sqrt{M}} e^{-j2\pi\,p\,q/M}.\label{beamspace}
+\x_\text{BS}[n] = \B\x[n], \qquad \B\in \mathbb{C}^{N_b,M}, \qquad \text{where} \qquad [\B]_{p,q} = \frac{1}{\sqrt{M}} e^{-j2\pi\,p\,q/M}.\label{beamspace}
 \end{gather}
-
+$\B$ is usually referred to as 
 Complexity: To frequency domain: O($N_b\,L$). Build $R_\text{BS}$, O($N_b^2\,N_L\,N_K$). Inversion: O($N_b^3$).

### 2013-01-23 ASA2013/asa2013.tex
la2pdf commit - 2013:01:23	01:03:41	(Wed23.Jan)
la2pdf commit - 2013:01:23	01:34:18	(Wed23.Jan)

@@
 \usepackage{glossaries}
+\usepackage{common/slashbox}
 
@@
 The underlying concept here that for narrowband data there will be a direct relation between the Fourier transform across the array angles of arrival in the spatial domain; the zero-frequency correspond to broadside, and increasing frequencies map to monotonously increasing opening angles.
-Since the combined transmit and receive response is highly directive, why would we want to process angles where there are hardly any information?
+Since the combined transmit and receive response is highly directive, this method allow us to only process the angles where the signal came from.
 
-Beamspace MVDR very closely resembles time domain MVDR. The main difference is the need to Fourier transform the data:
+The beamspace MVDR formulation very closely resembles that of time domain MVDR. The main difference is the initial need to Fourier transform the data:
 \begin{gather}
@@
 \end{gather}
-$\B$ is usually referred to as 
-Complexity: To frequency domain: O($N_b\,L$). Build $R_\text{BS}$, O($N_b^2\,N_L\,N_K$). Inversion: O($N_b^3$).
+$\B$ here is usually referred to as a Butler matrix. 
+Complexity: To frequency domain: O($N_b\,L$). Build $R_\text{BS}$, O($N_b^2\,N_L\,N_K$), Inversion: O($N_b^3$).
 
@@
 \end{figure}
+
+\begin{figure}[!h]\centering%
+\graphicsAI[drawing,width=\linewidth]{gfx/benchmark_tagged.svg}%
+\caption{Holmengraa.}\label{mvdr_implementation}
+\end{figure}
+
+% \begin{table}[h]\centering%\normalsize
+% \begin{tabular}[c]{l r r r@{:}  l}\hline
+% \rowcolor{tabBlue} & \multicolumn{1}{>{\columncolor{tabBlue}}c}{\bf BW$_\text{arithmetic}$} & \multicolumn{1}{>{\columncolor{tabBlue}}c}{\bf BW$_\text{memory}$} & \multicolumn{2}{>{\columncolor{tabBlue}}c}{\bf BW$_\text{memory}$/BW$_\text{arithmetic}$} \\\hline
+%  & & & & \\
+% \end{tabular}
+% \caption{Nvidia Quadro 6000; Memory bandwidths (BW) compared to the peak arithmetic bandwidth of 1.03 Tflop/s.}\label{tabbandwidth}
+% \end{table}
 

### 2013-01-23 ASA2013/asa2013.tex
la2pdf commit - 2013:01:23	02:05:28	(Wed23.Jan)

@@
 The LCA is essentially a delay-and-sum (DAS) beamformer where the delay step is carried out as usual, but where the weighting and summing is repeated for each of the $P$ windows.
-In its basic form its computational complexity is then of O($M\,P$).
+In its basic form its computational complexity only of O($M\,P$).
 
@@
 
-Some key aspects of our time domain MVDR implementation is illustrated in Figure \ref{mvdr_implementation}.
+Our GPU implementation consists of three main kernels relating to the computing steps illustrated in Figure \ref{mvdr_implementation}.
+In the first step we estimate $\eR$, in the second we solve the linear equation $\eRi\1$, and finally we perform some substitutions to obtain the beamformer output, $z$.
+All pixels are processed before moving on to the next step.
+Our efforts here were mainly focused on estimating $\eR$, because an official batch linear equation solver is already available from Nvidia, and computing the beamformer output is a rather simple task.
 
-First we make each SM compute entire range lines of pixels, then load the subset of $\x$ that all these pixels depend upon into shared memory.
-This lets us perform temporal averaging without the need for reading additional channel data.
-
-Second we assign $L$ threads per pixel to traverse the diagonals of $\eR$.
+When computing $\eR$ we we assign $L$ threads per pixel to traverse the diagonals of $\eR$.
 On the top row a full computation is carried out, then that entire row is saved back to global memory following a collective access pattern that maximizes global memory bandwidth.
@@
 
-\begin{figure}[!h]\centering%
+We have tested our GPU implementation of the MVDR beamformer experimental datasets from the 32 element Kongsberg Maritime HISAS1030 sonar being attached to the HUGIN AUV~\cite{Hansen2009}.
+To obtain the studied object, the 1500 dwt oil tanker wreck Holmengraa, the sonar was operated in sidescan mode to image (\Fig{hlmengraa}).
+Holmengraa is roughly 68\,m long and 9\,m wide, and lies on a slanted seabed at 77\,m depth.
+The image was processed with MVDR parameters $L=16$, $N_\text{avg}=3$, and $d=1\%$, which proved to be a reasonable selection for this scenario.
+
+The computational performance of our GPU implementation is displayed in (\Fig{benchmarks}).
+Processing times are shown for varying subarray lengths $L$ and varying amounts of temporal averaging $N_\text{avg}$.
+A single-thread C and Matlab implementation is added as reference, although only moderate optimization work was carried out on these.
+The test system was a quad-core Intel Xeon E5620, with 64Gb of RAM, and an nVidia Quadro 6000 card.
+
+\begin{figure}[!t]\centering%
 \graphicsAI[drawing,width=\linewidth]{gfx/plot_holmengraa_L16_Navg1.pdf}%
-\caption{Holmengraa.}\label{mvdr_implementation}
+\caption{Holmengraa.}\label{holmengraa}
 \end{figure}
@@
 \graphicsAI[drawing,width=\linewidth]{gfx/benchmark_tagged.svg}%
-\caption{Holmengraa.}\label{mvdr_implementation}
+\caption{Holmengraa.}\label{benchmark}
 \end{figure}

### 2013-01-23 ASA2013/asa2013.tex
la2pdf commit - 2013:01:23	02:39:33	(Wed23.Jan)

@@
 \end{gather}
-$\B$ here is usually referred to as a Butler matrix. 
+The number of angles to compute is here defined by $N_b$, $\B$, usually referred to as a Butler matrix
 Complexity: To frequency domain: O($N_b\,L$). Build $R_\text{BS}$, O($N_b^2\,N_L\,N_K$), Inversion: O($N_b^3$).
@@
 Since $\eR$ is conjugate symmetric, the values in the leftmost column is obtained by a complex conjugate copy of the relevant value in the first row.
-Combined these steps balances the load evenly on all threads, is almost completely free of arithmetic redundancy, and consumes less memory.
+This procedure balances the load evenly on all threads, is almost completely free of arithmetic redundancy, and consumes very little memory.
 
-These challenges are very closely linked.
+% These challenges are very closely linked.
 The Quadro 6000 architecture is of compute capability 2.0, which means there are 48\,kB of shared memory (L1 cache) and 128\,kB of registers per SM~\cite{Nvidia2012}.
@@
 
+% The solution is make sure that each thread primarily access the shared memory (L1 cache) and registers. Unfortunately, these are scarce resources and
+
 \section{Results}
 
-We have tested our GPU implementation of the MVDR beamformer experimental datasets from the 32 element Kongsberg Maritime HISAS1030 sonar being attached to the HUGIN AUV~\cite{Hansen2009}.
-To obtain the studied object, the 1500 dwt oil tanker wreck Holmengraa, the sonar was operated in sidescan mode to image (\Fig{hlmengraa}).
+To display the imaging capabilities of the mentioned beamformers, we have processed experimental datasets from the 32 element Kongsberg Maritime HISAS1030 sonar being attached to the HUGIN AUV~\cite{Hansen2009}.
+The studied object is the 1500 dwt oil tanker wreck Holmengraa, lying at 77\,m depth outside of Horten, Norway.
+HISAS1030 is a synthetic apterture sonar, but was operated in sidescan mode here.
 Holmengraa is roughly 68\,m long and 9\,m wide, and lies on a slanted seabed at 77\,m depth.
@@
 
-Throughout our work with porting the MVDR beamformer to a GPU, one challenge has been particularly difficult to solve: The GPU will only perform efficiently when each computing thread is very computationally intense.
-This should be apparent by studying Table \ref{tabbandwidth}.
-It compares theoretical peak values for the memory bandwidth and arithmetic throughput for an Nvidia Quadro 6000 GPU.
-Looking at global memory, we note that it is only able to move 1 float for every 30 floating point operation that the GPU is able to execute.
-Now, consider the simple equation we the simple equation $C=A+B$, where the ratio is 3 floats read/written for every 1 floating point operation, we realise that global memory should be avoided whenever possible.
-
-The solution is make sure that each thread primarily access the shared memory (L1 cache) and registers. Unfortunately, these are scarce resources and 
 \begin{table}[b]\centering%\normalsize
@@
 \end{tabular}
-\caption{Nvidia Quadro 6000; Memory bandwidths (BW) compared to the peak arithmetic bandwidth of 1.03 Tflop/s.}\label{tabbandwidth}
+\caption{Nvidia Quadro 6000: Memory bandwidths (BW) compared to the peak arithmetic bandwidth of 1.03 Tflop/s.}\label{tabbandwidth}
 \end{table}
+
+One thing seems clear to us after we have learned from this work is that it is very hard to reach the peak performance of a GPU, and more often
+
+. In our this learned from this process is how that each GPU thread have very small amounts of A thing we took home from this work 
+But no matter how Throughout our work with porting the MVDR beamformer to a GPU, one challenge has been particularly difficult to solve: The GPU will only perform efficiently when each computing thread is very computationally intense.
+This should be apparent by studying Table \ref{tabbandwidth}.
+It compares theoretical peak values for the memory bandwidth and arithmetic throughput for an Nvidia Quadro 6000 GPU.
+Looking at global memory, we note that it is only able to move 1 float for every 30 floating point operation that the GPU is able to execute.
+Now, consider the simple equation we the simple equation $C=A+B$, where the ratio is 3 floats read/written for every 1 floating point operation, we realise that global memory should be avoided whenever possible.
 

### 2013-01-23 ASA2013/asa2013.tex
la2pdf commit - 2013:01:23	03:10:38	(Wed23.Jan)

@@
 \begin{gather}
-\x_\text{BS}[n] = \B\x[n], \qquad \B\in \mathbb{C}^{N_b,M}, \qquad \text{where} \qquad [\B]_{p,q} = \frac{1}{\sqrt{M}} e^{-j2\pi\,p\,q/M}.\label{beamspace}
+\x_{\text{BS},l}[n] = \B\x_l[n], \qquad \B\in \mathbb{C}^{N_b,M}, \qquad \text{where} \qquad [\B]_{p,q} = \frac{1}{\sqrt{M}} e^{-j2\pi\,p\,q/M}.\label{beamspace}
 \end{gather}
-The number of angles to compute is here defined by $N_b$, $\B$, usually referred to as a Butler matrix
-Complexity: To frequency domain: O($N_b\,L$). Build $R_\text{BS}$, O($N_b^2\,N_L\,N_K$), Inversion: O($N_b^3$).
+Here $\B$ is referred to as the Butler matrix, $l$ is the subarray index, and $N_b$ is the number of beams that should be computed.
+Note has once $\x_{\text{BS},l}[n]$ is computed, $\eR_\text{BS}$ can be found as in (\ref{spatialR}), $\w_\text{BS}$ as in (\ref{weights}), and $z$ as in (\ref{z}).
+
+By reducing the covariance size from $L$ to $N_b$ we also reduce the estimation step to O($N_K N_L N_b^2$), and the inversion step to O($N_b^3$).
+The downside is the additional O[$N_b  Complexity: To\,L$) computations required  frequency domain: O($N_b\,L$).
+Build $R_\text{BS}$, O($N_b^2\,N_L\,N_K$), Inversion: O($N_b^3$).
 

### 2013-01-23 ASA2013/asa2013.tex
la2pdf commit - 2013:01:23	03:41:09	(Wed23.Jan)

@@
 
-This article is outlined as follows: First we will present\todo{denne formuleringstypen begynner aa bli oppbrukt...} background material on the MVDR method operated in time and frequency domain, and on the LCA beamformer.
-For each method we start with the mathematical formulation, then evaluate complexity issues, and finally give our thoughts on some implementation aspects.
-The results are collected in \ref{implementation}.
-Finally, we conclude in section \ref{discssion}.
+This article is outlined as follows: First we will present the mathematical formulation and evaluate complexity aspects of the MVDR method operated in time or frequency domain, and on the LCA beamformer.
+Then we mention some of our experiences with implementing the MVDR on a GPU, before concluding with some results and final thoughts.
 
@@
 \item The subarray covariance matrices $R_l[n] = \x_l[n]\x_l\H[n]$ are Hermitian semi-definite, which means that we can get away with computing only one triagonal in them.
-\item When summing $R_l[n]$ most of the multiplications can be computed only once, and most sums need only be updated iteratively.
+\item When summing $R_l[n]$ most of the multiplications need only be computed only once, and most sums can be updated iteratively.
 \end{itemize}
@@
 By reducing the covariance size from $L$ to $N_b$ we also reduce the estimation step to O($N_K N_L N_b^2$), and the inversion step to O($N_b^3$).
-The downside is the additional O[$N_b  Complexity: To\,L$) computations required  frequency domain: O($N_b\,L$).
-Build $R_\text{BS}$, O($N_b^2\,N_L\,N_K$), Inversion: O($N_b^3$).
+Although the additional computations required by the frequency transform must be taken into consideration, the effect of adjusting far outweighs it for larger systems.
 
@@
 This procedure balances the load evenly on all threads, is almost completely free of arithmetic redundancy, and consumes very little memory.
+
+Frequency domain MVDR beamforming
 

### 2013-01-23 ASA2013/asa2013.tex
la2pdf commit - 2013:01:23	04:11:31	(Wed23.Jan)

@@
 For smaller systems we use the standard time-domain MVDR method, arithmetically optimize the estimation step, and show that image creation rates of more than 1\,Mpx/s can be achieved when this algorithm is run on a GPU.
-For medium sized systems we suggest applying the MVDR in the frequency domain (beamspace) instead, which allows for comparable processing rates at a negligible reduction in image quality.
+For medium sized systems we suggest applying the MVDR in the beamspace (frequency domain) instead, which allows for comparable processing rates at a negligible reduction in image quality.
 Finally, for large systems the estimation and inversion step can be avoided completely by utilising the Low Complexity Adaptive (LCA) beamformer.\todo{disse tre første avsnittene burde sikkert vært kortet ned mye men...}
@@
 
-Comparable work can be found e.g.
+Related work can be found e.g.
 in ultrasound imaging, where Chen et.
@@
 \begin{itemize}
-\item The subarray covariance matrices $R_l[n] = \x_l[n]\x_l\H[n]$ are Hermitian semi-definite, which means that we can get away with computing only one triagonal in them.
+\item The subarray covariance matrices $R_l[n] = \x_l[n]\x_l\H[n]$ are Hermitian semi-definite, which means that we can get away with computing only one of their triagonals.
 \item When summing $R_l[n]$ most of the multiplications need only be computed only once, and most sums can be updated iteratively.
@@
 
-As we increase system size the inversion step will become more of a bottleneck.
-Although the dimensionality of the covariance matrix is reduced to $L$ due to subarray averaging, we can reduce it further by performing MVDR beamforming.
-The underlying concept here that for narrowband data there will be a direct relation between the Fourier transform across the array angles of arrival in the spatial domain; the zero-frequency correspond to broadside, and increasing frequencies map to monotonously increasing opening angles.
-Since the combined transmit and receive response is highly directive, this method allow us to only process the angles where the signal came from.
+As we increase system size the inversion step will take over as the main bottleneck.
+We can, however, reduce the dimensionality further than the subarray size of $L$ by performing MVDR beamforming the frequency domain.
+The underlying concept here that for narrowband data there will be a direct relation between the Fourier transform across the array and angles of arrival in the spatial domain; the zero-frequency will correspond to broadside, and increasing frequencies will map to monotonously increasing opening angles.
+Since the combined transmit and receive response is highly directive, this method allow us to only process the angles where our signals are coming from~\cite{Nilsen2009a,VanTrees2002}.
+The method is named beamspace MVDR.
 
@@
 In the first step we estimate $\eR$, in the second we solve the linear equation $\eRi\1$, and finally we perform some substitutions to obtain the beamformer output, $z$.
-All pixels are processed before moving on to the next step.
+All pixels are processed by one step before we move on to the next step.
 Our efforts here were mainly focused on estimating $\eR$, because an official batch linear equation solver is already available from Nvidia, and computing the beamformer output is a rather simple task.

### 2013-01-23 ASA2013/asa2013.tex
la2pdf commit - 2013:01:23	06:50:24	(Wed23.Jan)

@@
-\newif\ifTODO\TODOtrue                        % Use todo notes?
+\newif\ifTODO\TODOfalse                        % Use todo notes?
 
@@
 \begin{document}
-% \setlength{\headrulewidth}{0.0pt}
-
-%~~ TitlePage ~~%                               %~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~%
-\thispagestyle{empty}
-\pagenumbering{arabic} %Normal numbers
-
-{\Large\bf Adapting the MVDR beamformer to a GPU for differently sized active sonar imaging systems.}
-
-\begin{abstract}
-The MVDR beamformer has been shown to improve active sonar image quality compared to conventional methods.
+% % \setlength{\headrulewidth}{0.0pt}
+% 
+% %~~ TitlePage ~~%                               %~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~%
+% \thispagestyle{empty}
+% \pagenumbering{arabic} %Normal numbers
+% 
+% {\Large\bf Adapting the MVDR beamformer to a GPU for differently sized active sonar imaging systems.}
+% 
+% \begin{abstract}
+% The MVDR beamformer has been shown to improve active sonar image quality compared to conventional methods.
 Unfortunately, it is also significantly more computationally expensive because a spatial covariance matrix must be estimated and inverted for each image pixel.
 We target this challenge by altering and mapping the MVDR beamformer to a GPU, and suggest three different solutions depending on the system size.
-
-For systems with relatively few channels, we suggest arithmetic optimizations for the estimation step, and show how a GPU can be used to yield image creation rates of more than 1\,Mpx/s.
+% 
+% For systems with relatively few channels, we suggest arithmetic optimizations for the estimation step, and show how a GPU can be used to yield image creation rates of more than 1\,Mpx/s.
 For larger systems we show that frequency domain processing is preferable, as this promotes high processing rates at a negligible reduction in image quality.
 These GPU implementations consistently reduced the runtime by 2-3 orders of magnitude compared to our reference C and Matlab implementations.
-
-For even larger systems we suggest employing the LCA beamformer.
+% 
+% For even larger systems we suggest employing the LCA beamformer.
 It does not calculate a weightset, but merely computes the beamformer output for each of a predefined set of weights, and selects the one that best fulfils the MVDR criterion.
 The LCA creates images with a quality comparable to MVDR, and it is perfectly suited for a GPU.
-\end{abstract}
-\vspace{2cm}
-(Forsiden genereres av ASA.)
-% \ \\
-% Total: 195 words (200 max)
+% \end{abstract}
+% \vspace{2cm}
+% (Forsiden genereres av ASA.)
+% % \ \\
+% % Total: 195 words (200 max)
+% % 
+% % \ \\
+% % Noen tanker jeg har selv naa:
+% % \begin{itemize}
+% % \item Det mangler noe om bruksomraader.
+% % \item Maa passe paa aa ikke selge LCA saa bra at MVDR ser helt haaplos ut(?)
+% % \end{itemize}
+% % 
+% % 
+% % \ \\
+% % Questions they wanted answered:
+% % \begin{itemize}
+% % \item This session will feature work on adapting computationally demanding modeling and signal processing tasks to take advantage of various parallel architectures, such as cloud computing, computing clusters, multicore CPUs, multi-core graphic processing units (GPUs), DSP chips, and FPGAs.
+% % \item If algorithm is slow - What technology will help me?
+% % \item What skills are needed?
+% % \item What tools are available and what's the cost?
+% % \item What acceleration factor are we looking at?
+% % \item How can I get started?
+% % \item What we did. Which tools? Was it hard? Speedup?
+% % \end{itemize}
 % 
-% \ \\
-% Noen tanker jeg har selv naa:
-% \begin{itemize}
-% \item Det mangler noe om bruksomraader.
-% \item Maa passe paa aa ikke selge LCA saa bra at MVDR ser helt haaplos ut(?)
-% \end{itemize}
-% 
-% 
-% \ \\
-% Questions they wanted answered:
-% \begin{itemize}
-% \item This session will feature work on adapting computationally demanding modeling and signal processing tasks to take advantage of various parallel architectures, such as cloud computing, computing clusters, multicore CPUs, multi-core graphic processing units (GPUs), DSP chips, and FPGAs.
-% \item If algorithm is slow - What technology will help me?
-% \item What skills are needed?
-% \item What tools are available and what's the cost?
-% \item What acceleration factor are we looking at?
-% \item How can I get started?
-% \item What we did. Which tools? Was it hard? Speedup?
-% \end{itemize}
-
-\newpage
+% \newpage
 \section{Introduction}
@@
 Graphics Processing Units (GPUs) is a technology tailored to perform image processing as fast as possible.
-They are available off-the-shelf at reasonable prices, are very power efficient, and can be programmed to carry out a wide range of computational tasks, often displaying a speed increase of an order of magnitude compared to similar CPU designs.
+They are available off-the-shelf at reasonable prices, are very power efficient, and can be programmed to carry out a wide range of computational tasks, often displaying speed increases of an order of magnitude compared to similar CPU designs.
 Initially, GPUs may seem perfectly suited for image reconstruction algorithms, because usually each pixel can be computed independently and concurrently of one another.
@@
 Instead of calculating the weights like in (\ref{weights}), it simply computes the beamformer output for each window in a collection of such, and finally applies the one that best fulfils the given optimization criterion.
-If we let $\w_p$ be the $p$'th of a total of $P$ windows, then this can be formulated as:
+If we let $\w_p$ be the $p$'th of a total of $P$ windows, then this can be formulated as~\cite{Synnevag2008}:
 \begin{gather}
@@
 
-Frequency domain MVDR beamforming
+The implementation of beamspace MVDR is very similar, but the data access pattern is different.
+In the time domain subsequent subarrays overlap on all but one element in memory, but once Fourier transformed the data does not overlap at all.
+Such a seemingly innocent detail can make all the difference in a GPU design, because efficient memory accessing is key to obtaining good performance.
 
@@
 
-To display the imaging capabilities of the mentioned beamformers, we have processed experimental datasets from the 32 element Kongsberg Maritime HISAS1030 sonar being attached to the HUGIN AUV~\cite{Hansen2009}.
-The studied object is the 1500 dwt oil tanker wreck Holmengraa, lying at 77\,m depth outside of Horten, Norway.
-HISAS1030 is a synthetic apterture sonar, but was operated in sidescan mode here.
-Holmengraa is roughly 68\,m long and 9\,m wide, and lies on a slanted seabed at 77\,m depth.
-The image was processed with MVDR parameters $L=16$, $N_\text{avg}=3$, and $d=1\%$, which proved to be a reasonable selection for this scenario.
+To display the imaging capabilities of the mentioned beamformers, we have processed experimental datasets from the 32 element Kongsberg Maritime HISAS1030 sonar mounted on the HUGIN AUV~\cite{Hansen2009}.
+HISAS1030 is a high resolution synthetic aperture sonar with an array length of 1.2\,m and operating frequency of 100\,kHz, and it was operated in sidescan mode here.
+The studied object is the 1500 dwt oil tanker wreck Holmengraa.
+It is 68\,m long and 9\,m wide, and lies at a slanted seabed at 77\,m depth outside of Horten, Norway.
+The results are shown in Figure \ref{holmengraa}.
+The MVDR images were processed with parameters $L$=16, $K$=3, and 1\% diagonal loading.
+For beamspace MVDR we set $N_b=L/2$.
+LCA was designed with a window database consisting of 5 Kaiser windows, each with different window-parameters from the range $\beta\in[0,10]$, and each steed in 5 distinct directions leaving a total of 25 windows for the LCA to select from.
 
-The computational performance of our GPU implementation is displayed in (\Fig{benchmarks}).
-Processing times are shown for varying subarray lengths $L$ and varying amounts of temporal averaging $N_\text{avg}$.
-A single-thread C and Matlab implementation is added as reference, although only moderate optimization work was carried out on these.
-The test system was a quad-core Intel Xeon E5620, with 64Gb of RAM, and an nVidia Quadro 6000 card.
+A few benchmarks of these methods are presented in Figure \ref{benchmark}.
+On our test system, a quad-core Intel Xeon E5620 with 64Gb of RAM and an nVidia Quadro 6000 card, the MVDR methods managed to create images at rates just over 1\,Mpx/s.
+A single-thread C implementation that was added as reference was outperformed by around 2 orders of magnitude.
+However, do not consider this comparison fair because while the GPU design is make efficient use of the L1 cache, similar optimisation work was not carried out on the CPU design.
 
@@
 \graphicsAI[drawing,width=\linewidth]{gfx/plot_holmengraa_L16_Navg1.pdf}%
-\caption{Holmengraa.}\label{holmengraa}
-\end{figure}
-
-\begin{figure}[!h]\centering%
-\graphicsAI[drawing,width=\linewidth]{gfx/benchmark_tagged.svg}%
-\caption{Holmengraa.}\label{benchmark}
+\caption{The 68$\times$9\,m oil tanker Holmengraa lying at a depth of 77\,m.
+The MVDR images were processed with L=16, K=3, and 1\% diagonal loading.
+For beamspace MVDR we set $N_b=L/2$.
+LCA was set up with a window database consisting of 5 Kaiser windows with the window-parameter set in the range $\beta\in[0,10]$, and then each of these were steed in 5 distinct directions leaving a total of 25 windows for the LCA to select from..}\label{holmengraa}
 \end{figure}
@@
 
+\begin{figure}[!h]\centering%
+\graphicsAI[drawing,width=.8\linewidth]{gfx/benchmark1_tagged.pdf}%
+\caption{Benchmarks for the our time-domain and frequency-domain MVDR method running on a GPU.
+The grey line marks the realtime requirement for the HUGIN AUV to perform full-coverage sectorscan images with a ping rate of 3 pings/s.}\label{benchmark}
+\end{figure}
+
 \section{Discussion}
 
-\begin{table}[b]\centering%\normalsize
-\begin{tabular}[c]{l r r r@{:}  l}\hline
-\rowcolor{tabBlue} & \multicolumn{1}{>{\columncolor{tabBlue}}c}{\bf BW$_\text{arithmetic}$} & \multicolumn{1}{>{\columncolor{tabBlue}}c}{\bf BW$_\text{memory}$} & \multicolumn{2}{>{\columncolor{tabBlue}}c}{\bf BW$_\text{memory}$/BW$_\text{arithmetic}$} \\\hline
-Arithmetic throughput & 1.03 Tflop/s & & \hspace{30pt} 1 & 30 \\
-Global memory & & 36 Gfloats/s & \hspace{30pt} 1 & 30 \\
-Shared memory & & 257 Gfloats/s & 1 & 4 \\
-Registers & & $>$1.5 Tfloats/s & $>$3 & 2~\cite{Vasilyy}
-\end{tabular}
-\caption{Nvidia Quadro 6000: Memory bandwidths (BW) compared to the peak arithmetic bandwidth of 1.03 Tflop/s.}\label{tabbandwidth}
-\end{table}
+% \begin{table}[b]\centering%\normalsize
+% \begin{tabular}[c]{l r r r@{:}  l}\hline
+% \rowcolor{tabBlue} & \multicolumn{1}{>{\columncolor{tabBlue}}c}{\bf BW$_\text{arithmetic}$} & \multicolumn{1}{>{\columncolor{tabBlue}}c}{\bf BW$_\text{memory}$} & \multicolumn{2}{>{\columncolor{tabBlue}}c}{\bf BW$_\text{memory}$/BW$_\text{arithmetic}$} \\\hline
+% Arithmetic throughput & 1.03 Tflop/s & & \hspace{30pt} 1 & 30 \\
+% Global memory & & 36 Gfloats/s & \hspace{30pt} 1 & 30 \\
+% Shared memory & & 257 Gfloats/s & 1 & 4 \\
+% Registers & & $>$1.5 Tfloats/s & $>$3 & 2~\cite{Vasilyy}
+% \end{tabular}
+% \caption{Nvidia Quadro 6000: Memory bandwidths (BW) compared to the peak arithmetic bandwidth of 1.03 Tflop/s.}\label{tabbandwidth}
+% \end{table}
 
-One thing seems clear to us after we have learned from this work is that it is very hard to reach the peak performance of a GPU, and more often
+The MVDR beamformer is a technique that is able to improve image quality over conventional methods.
+This can be seen in Figure \ref{holmengraa}, where the MVDR produces images with cleaner shadows and slightly improved edge detection.
+Contrary to common belief, it can also be made to create these images relatively fast.
+Through careful optimization work and by running the MVDR on a GPU we have achieved image creation rates of more than 1\,Mpx/s.
+This is an improvement of 2 orders of magnitude over our reference implementation in C.
+To obtain these rates we run MVDR in the time-domain for smaller systems, and in the frequency domain for larger systems.
 
-. In our this learned from this process is how that each GPU thread have very small amounts of A thing we took home from this work 
-But no matter how Throughout our work with porting the MVDR beamformer to a GPU, one challenge has been particularly difficult to solve: The GPU will only perform efficiently when each computing thread is very computationally intense.
+If this performance is insufficient, we suggest looking at the LCA method.
+It is based on the same optimisation criterion as the MVDR method, but avoids the estimation and inversion step altogether by limiting the weight's solution space to only a few predefined windows.
+As seen the Figure \ref{holmengraa}, it produces images with a quality comparable to the MVDR method, but computing the images takes only a fraction of the time.
+
+% Finally, throughout our work with porting the MVDR beamformer to a GPU, one challenge has been particularly difficult to solve: The GPU will only perform efficiently when each computing thread is very computationally intense.
 This should be apparent by studying Table \ref{tabbandwidth}.
@@
 Now, consider the simple equation we the simple equation $C=A+B$, where the ratio is 3 floats read/written for every 1 floating point operation, we realise that global memory should be avoided whenever possible.
-
-MVDR for systems with less than 32 channels. After that MVDR in beamspace.
 

### 2013-01-23 ASA2013/asa2013.tex

@@
 \begin{document}
-% % \setlength{\headrulewidth}{0.0pt}
-% 
-% %~~ TitlePage ~~%                               %~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~%
-% \thispagestyle{empty}
-% \pagenumbering{arabic} %Normal numbers
-% 
-% {\Large\bf Adapting the MVDR beamformer to a GPU for differently sized active sonar imaging systems.}
-% 
-% \begin{abstract}
-% The MVDR beamformer has been shown to improve active sonar image quality compared to conventional methods.
+% \setlength{\headrulewidth}{0.0pt}
+
+%~~ TitlePage ~~%                               %~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~%
+\thispagestyle{empty}
+\pagenumbering{arabic} %Normal numbers
+
+{\Large\bf Adapting the MVDR beamformer to a GPU for differently sized active sonar imaging systems.}
+
+\begin{abstract}
+The MVDR beamformer has been shown to improve active sonar image quality compared to conventional methods.
 Unfortunately, it is also significantly more computationally expensive because a spatial covariance matrix must be estimated and inverted for each image pixel.
 We target this challenge by altering and mapping the MVDR beamformer to a GPU, and suggest three different solutions depending on the system size.
-% 
-% For systems with relatively few channels, we suggest arithmetic optimizations for the estimation step, and show how a GPU can be used to yield image creation rates of more than 1\,Mpx/s.
+
+For systems with relatively few channels, we suggest arithmetic optimizations for the estimation step, and show how a GPU can be used to yield image creation rates of more than 1\,Mpx/s.
 For larger systems we show that frequency domain processing is preferable, as this promotes high processing rates at a negligible reduction in image quality.
 These GPU implementations consistently reduced the runtime by 2-3 orders of magnitude compared to our reference C and Matlab implementations.
-% 
-% For even larger systems we suggest employing the LCA beamformer.
+
+For even larger systems we suggest employing the LCA beamformer.
 It does not calculate a weightset, but merely computes the beamformer output for each of a predefined set of weights, and selects the one that best fulfils the MVDR criterion.
 The LCA creates images with a quality comparable to MVDR, and it is perfectly suited for a GPU.
-% \end{abstract}
-% \vspace{2cm}
-% (Forsiden genereres av ASA.)
-% % \ \\
-% % Total: 195 words (200 max)
-% % 
-% % \ \\
-% % Noen tanker jeg har selv naa:
-% % \begin{itemize}
-% % \item Det mangler noe om bruksomraader.
-% % \item Maa passe paa aa ikke selge LCA saa bra at MVDR ser helt haaplos ut(?)
-% % \end{itemize}
-% % 
-% % 
-% % \ \\
-% % Questions they wanted answered:
-% % \begin{itemize}
-% % \item This session will feature work on adapting computationally demanding modeling and signal processing tasks to take advantage of various parallel architectures, such as cloud computing, computing clusters, multicore CPUs, multi-core graphic processing units (GPUs), DSP chips, and FPGAs.
-% % \item If algorithm is slow - What technology will help me?
-% % \item What skills are needed?
-% % \item What tools are available and what's the cost?
-% % \item What acceleration factor are we looking at?
-% % \item How can I get started?
-% % \item What we did. Which tools? Was it hard? Speedup?
-% % \end{itemize}
+\end{abstract}
+\vspace{2cm}
+(Forsiden genereres av ASA.)
+% \ \\
+% Total: 195 words (200 max)
 % 
-% \newpage
+% \ \\
+% Noen tanker jeg har selv naa:
+% \begin{itemize}
+% \item Det mangler noe om bruksomraader.
+% \item Maa passe paa aa ikke selge LCA saa bra at MVDR ser helt haaplos ut(?)
+% \end{itemize}
+% 
+% 
+% \ \\
+% Questions they wanted answered:
+% \begin{itemize}
+% \item This session will feature work on adapting computationally demanding modeling and signal processing tasks to take advantage of various parallel architectures, such as cloud computing, computing clusters, multicore CPUs, multi-core graphic processing units (GPUs), DSP chips, and FPGAs.
+% \item If algorithm is slow - What technology will help me?
+% \item What skills are needed?
+% \item What tools are available and what's the cost?
+% \item What acceleration factor are we looking at?
+% \item How can I get started?
+% \item What we did. Which tools? Was it hard? Speedup?
+% \end{itemize}
+
+\newpage
 \section{Introduction}
@@
 As we increase system size the inversion step will take over as the main bottleneck.
-We can, however, reduce the dimensionality further than the subarray size of $L$ by performing MVDR beamforming the frequency domain.
+We can, however, reduce the dimensionality further than subarray averaging already has by performing MVDR beamforming in the frequency domain.
 The underlying concept here that for narrowband data there will be a direct relation between the Fourier transform across the array and angles of arrival in the spatial domain; the zero-frequency will correspond to broadside, and increasing frequencies will map to monotonously increasing opening angles.
-Since the combined transmit and receive response is highly directive, this method allow us to only process the angles where our signals are coming from~\cite{Nilsen2009a,VanTrees2002}.
-The method is named beamspace MVDR.
+Since the combined transmit and receive response is highly directive, this method allows us to only process the angles where our signals are coming from~\cite{Nilsen2009a,VanTrees2002}.
+The method is called beamspace MVDR.
 
-The beamspace MVDR formulation very closely resembles that of time domain MVDR. The main difference is the initial need to Fourier transform the data:
+The beamspace MVDR formulation very closely resembles that of the time domain MVDR. The main difference is the initial need to Fourier transform the data:
 \begin{gather}
-\x_{\text{BS},l}[n] = \B\x_l[n], \qquad \B\in \mathbb{C}^{N_b,M}, \qquad \text{where} \qquad [\B]_{p,q} = \frac{1}{\sqrt{M}} e^{-j2\pi\,p\,q/M}.\label{beamspace}
+\x_{\text{BS},l}[n] = \B\x_l[n], \qquad \B\in \mathbb{C}^{N_b,L}, \qquad \text{where} \qquad [\B]_{p,q} = \frac{1}{\sqrt{L}} e^{-j2\pi\,p\,q/L}.\label{beamspace}
 \end{gather}
@@
 
-By reducing the covariance size from $L$ to $N_b$ we also reduce the estimation step to O($N_K N_L N_b^2$), and the inversion step to O($N_b^3$).
-Although the additional computations required by the frequency transform must be taken into consideration, the effect of adjusting far outweighs it for larger systems.
+When we reduce the covariance size from $L$ to $N_b$ we also reduce the estimation step to O($N_K N_L N_b^2$), and the inversion step to O($N_b^3$).
+The downside is the additional computations required by the frequency transform, which must be weighted up against the gain of reducing $N_b$.
 
@@
 The Low Complexity Adaptive (LCA) beamformer is another take on the optimization problem in (\ref{mvdr_def}).
-Instead of calculating the weights like in (\ref{weights}), it simply computes the beamformer output for each window in a collection of such, and finally applies the one that best fulfils the given optimization criterion.
+Instead of calculating the weights like in (\ref{weights}), it simply computes the beamformer output for each window in a larger set of windows, and finally applies the one that best fulfils the given optimization criterion.
 If we let $\w_p$ be the $p$'th of a total of $P$ windows, then this can be formulated as~\cite{Synnevag2008}:
@@
 
-Our GPU implementation consists of three main kernels relating to the computing steps illustrated in Figure \ref{mvdr_implementation}.
+Our GPU implementation consists of three main kernels, one for each of the computing steps illustrated in Figure \ref{mvdr_implementation}.
 In the first step we estimate $\eR$, in the second we solve the linear equation $\eRi\1$, and finally we perform some substitutions to obtain the beamformer output, $z$.
 All pixels are processed by one step before we move on to the next step.
-Our efforts here were mainly focused on estimating $\eR$, because an official batch linear equation solver is already available from Nvidia, and computing the beamformer output is a rather simple task.
+Our efforts here were mainly focused on estimating $\eR$, because a highly optimised batch linear equation solver is already available from Nvidia, and computing the beamformer output is a rather simple task.
 
@@
 The implementation of beamspace MVDR is very similar, but the data access pattern is different.
-In the time domain subsequent subarrays overlap on all but one element in memory, but once Fourier transformed the data does not overlap at all.
-Such a seemingly innocent detail can make all the difference in a GPU design, because efficient memory accessing is key to obtaining good performance.
+In the time domain subsequent subarrays overlap on all but one element in memory, but the Fourier transformed data does not overlap at all.
+Such a seemingly innocent detail can make all the difference in a GPU design, because designing to make efficient use of memory is key to obtaining good performance.
 
@@
 For beamspace MVDR we set $N_b=L/2$.
-LCA was designed with a window database consisting of 5 Kaiser windows, each with different window-parameters from the range $\beta\in[0,10]$, and each steed in 5 distinct directions leaving a total of 25 windows for the LCA to select from.
+LCA was designed with a window database consisting of 5 Kaiser windows, each with different window-parameters from the range $\beta\in[0,10]$, and each steered in 5 distinct directions leaving a total of 25 windows for the LCA to select from.
 
 A few benchmarks of these methods are presented in Figure \ref{benchmark}.
-On our test system, a quad-core Intel Xeon E5620 with 64Gb of RAM and an nVidia Quadro 6000 card, the MVDR methods managed to create images at rates just over 1\,Mpx/s.
+On our test system with a quad-core Intel Xeon E5620, 64Gb of RAM and an nVidia Quadro 6000 card, the MVDR methods managed to create images at rates just over 1\,Mpx/s.
 A single-thread C implementation that was added as reference was outperformed by around 2 orders of magnitude.
@@
 
+\newpage
 \section{Discussion}
@@
 
-The MVDR beamformer is a technique that is able to improve image quality over conventional methods.
-This can be seen in Figure \ref{holmengraa}, where the MVDR produces images with cleaner shadows and slightly improved edge detection.
-Contrary to common belief, it can also be made to create these images relatively fast.
+The MVDR beamformer is a technique that is capable of improving image quality over conventional methods.
+This can be seen in Figure \ref{holmengraa}, where the MVDR produces images with cleaner shadows and slightly improved edge definitions.
+Contrary to common belief, it is also possible to create these images relatively fast.
 Through careful optimization work and by running the MVDR on a GPU we have achieved image creation rates of more than 1\,Mpx/s.

### 2013-01-23 HRUIonGPU/jrnl_article.tex

@@
 If adaptive diagonal loading is applied after data have been transformed into beamspace, a high percentage of the total energy is mapped to a small subset of beams.
-Thus the average energy per beam is higher than the average energy was per element, if the system uses focused transmit beams.
-This means that less diagonal loading (smaller $d$) has to be applied in beamspace than in element space.
+Thus the average energy per beam is higher than the average energy was per element, when beams containing little energy are removed.
+This means that less diagonal loading (smaller $d$) has to be applied in beamspace than in element space when $N_b < L$.
 
@@
 Execution time is plotted as a function of subarray length $L$.
+If $L$ is kept constant on $M/2$ and execution time is plotted against $N_b$, the plot would be similar to the two in Fig.
+\ref{fig:bench}.
 (a) Shows the benchmark for a 64 elements array.
-(b) Shows the benchmark for a 96 elements array.
-(TODO: Change this to be a function of $N_b$)}
+(b) Shows the benchmark for a 96 elements array.}
 \label{fig:benchBS}

### 2013-01-24 HRUIonGPU/jrnl_article.tex
uffc

@@
 For medical ultrasound, Chen et al.
-\cite{Chen2011a, Chen2011} implemented a Capon beamformer on the GPU that produced real apodization weights based on real data.
+\cite{Chen2011a, Chen2011, Chen} implemented a Capon beamformer on the GPU that produced real apodization weights based on real data.
 Under this constraint the Capon beamformer is not allowed to micro-steer the main lobe, and therefore its full potential is not reached.
@@
 where $\text{CONV}_m$ is a non-truncated convolution across elements.
-Note how we can choose to first sum all subarrays (\ref{eq:z_mv}), or accumulate the right weighting per element (\ref{eq:z_mv2}).
+Note how we can either choose to sum all subarrays (\ref{eq:z_mv}), or accumulate the right weighting per element (\ref{eq:z_mv2}).
 
 In passive systems, assuming stationary statistics, the sample covariance matrix in (\ref{eq:R}) is typically averaged over a long sequence of samples\cite{Krima}.
-For an active, pulsed system like medical ultrasound imaging, the available data from which $\mat{\hat{R}}$ can be estimated is quite limited.
+For an active, pulsed system like medical ultrasound imaging, the available data from which $\mat{\hat{R}}$ can be estimated is limited.
 Because of this, the resulting sample covariance matrix is often found to be both inaccurate, and poorly conditioned.
@@
 
-In order to get a well-conditioned $\mat{\hat{R}}$, avoid signal cancellation, and to get DAS-like speckle, $\mat{\hat{R}}$ has to be averaged over $L\le M/2$ long subarrays and $N_K \sim T_p/T_s$ time samples \cite{Synnevag2007, Synnevag2007a}.
-Where, $T_p$ is the pulse length in seconds and $T_s$ is the sampling period.
+In order to get a well-conditioned $\mat{\hat{R}}$, avoid signal cancellation, and to get DAS-like speckle, $\mat{\hat{R}}$ has to be averaged over $L\le M/2$ long subarrays and $N_K \sim T_p/T_s$ time samples \cite{Synnevag2007, Synnevag2007a}, where, $T_p$ is the pulse length in seconds and $T_s$ is the sampling period.
 For baseband data, one pulse length ($1.5\lambda$) is around 3 samples ($K=1$).
@@
 Global reads and writes on the GPU, in addition to CPU-GPU transfers, have high penalty, and should be minimized as much as possible.
-All of these constraints combined pose a real challenge as the target problem has to map on to the parallel framework both on a grid, block, thread and memory consumption level in order to benefits from parallel acceleration.
+All of these constraints combined pose a real challenge as the target problem has to map on to the GPU framework both on a grid, block, thread and memory consumption level in order to benefits from parallel acceleration.
 In the next section, we will describe how this mapping could be done for the Capon beamforming algorithm.
@@
 \text{1) } &\text{Calculate the sample covariance matrix } \mat{\hat{R}} \text{ as in } \text{(\ref{eq:diag})}\\
-\text{2) } &\text{Solve the linear system of equations in (\ref{eq:b}})\\ %\vec{b} =\mat{\hat{R}}^{-1}\vec{1}_L\\
+\text{2) } &\text{Solve the linear system of equations as in (\ref{eq:b}})\\ %\vec{b} =\mat{\hat{R}}^{-1}\vec{1}_L\\
 \text{3) } &\text{Calculate } \vec{w} = \vec{b}/\vec{1}_{L}^H\vec{b} \text{ and the amplitude as in (\ref{eq:z_mv})}
@@
 
-From (\ref{eq:R}) it is obvious that calculation of one element in $\mat{\hat{R}}$ is independent of all other elements.
-A natural level of granularity would then be to assign one thread to each element of $\mat{\hat{R}}$ as done by \cite{Chen2011}.
+From (\ref{eq:R}) it is obvious that calculating one element of $\mat{\hat{R}}$ is independent of all other elements.
+A natural level of granularity would then be to assign one thread to each element of $\mat{\hat{R}}$ \cite{Chen2011}.
 This will minimize global reads and writes needed per thread, but unfortunately the maximum number of threads per block puts an upper limit on the subarray length, $L_{max}=32$, since we want to avoid block-to-block synchronization.
-
-The cubic complexity with $L$ and linear complexity with $N_K$ can be reduced by realizing that the calculation in (\ref{eq:R}) overlap both across subarrays and in time.
+Since arrays of 96-128 elements are common in medical ultrasound imaging an implementation not limited by $L_max=32$ is preferable.
+
+To reach this goal we start by showing how the cubic complexity with $L$ and linear complexity with $N_K$ can be reduced by realizing that calculations in (\ref{eq:R}) overlap both across subarrays and in time.
 Element $(i,j)$ of $\mat{\breve{R}}[n]$ can be calculated from element $(i-1, j-1), \text{for } i,j \in [1, L-1]$ as
@@
 Since one row for all covariance matrices is written to global memory per kernel iteration, the diagonal elements need to be held in a local memory until the trace is accumulated.
-When it is found, scattered writes of the scaled diagonal elements to global memory has to be performed.
+When that is done, scattered writes of the scaled diagonal elements to global memory has to be performed.
 
@@
 First the CPU, which might already runs a lot of ultrasound related algorithms, is offloaded.
-Second the transfer of several thousand covariance matrices from CPU-side to GPU-side is spared.
-
-For the results presented in this paper, we have used an GPU implementation of Gauss-Jordan (GJ) elimination made by Nvidia to solve $\mat{\hat{R}}\vec{b} = \vec{1}_L$, where $\vec{b} = \mat{\hat{R}}^{-1}\vec{1}_L$.
+Second the transfer of several thousand covariance matrices over the PCI-Express buss, from CPU-side to GPU-side, is spared.
+
+For the results presented in this paper, we have used a GPU implementation of batched Gauss-Jordan (GJ) elimination made by Nvidia to solve $\mat{\hat{R}}\vec{b} = \vec{1}_L$, where $\vec{b} = \mat{\hat{R}}^{-1}\vec{1}_L$ for all pixels at once.
 The solver has been made available for registered CUDA developers.%For smaller systems, $L < 10$, a direct solver using one thread per system is preferable.
@@
 \subsection{Compute Beamformer Output}
-To calculate the beamformer output we have implement the formula in (\ref{eq:z_mv}) with one thread per element-weight ($L$ threads per weight vector).
+To calculate the beamformer output, equation (\ref{eq:z_mv}) has been implemented using one thread per element-weight ($L$ threads per weight vector).
 After the solution vector $\vec{b}$ has been placed in shared memory, one of the $L$ threads finds the sum of $\vec{b}$.
-Another option is to let several threads cooperate to form the sum using a binary reduction sum.
-However, in the following benchmarks the first approach has been used.
+%Another option is to let several threads cooperate to form the sum using a binary reduction sum. However, in the following benchmarks the first approach has been used. 
 Each of the $L$ threads then read one element of $\vec{b}$, scale it with the inverse sum just found to get the weight vector $\vec{w}[n]$.
@@
 The final output, as shown in (\ref{eq:z_mv}) and (\ref{eq:z_mv2}), can be formed by either reducing the data down to length $L$, or increase the weight vector to length $M$.
-Since we have decided to have $L$ threads per output, the data is reduced to length $L$ by letting thread $r$ calculate
+Since the design is $L$ threads per output, the data is reduced to length $L$ by letting thread $r$ calculate
 \begin{align}\label{eq:subsum}
@@
 Each thread now computes $z_r = w_r^*x_r$, and finally one thread computes and outputs $z = \frac{1}{N_L}\sum_{r=0}^{L-1} z_r$ to global memory.
-The last sum can also be performed using a binary reduction sum across several threads for increased speed.
+%The last sum can also be performed using a binary reduction sum across several threads for increased speed.
 
@@
 Nilsen et al.
-\cite{Nilsen2009} have proposed to apply beamspace processing to further reduce the required computations for the Capon beamformer.
+\cite{Nilsen2009} have proposed to apply beamspace processing to further reduce the required computations of Capon beamformer.
 With beamspace we mean transforming the channel data from element space to a fan of beams covering the sector illuminated by transmit beams.
-The transformation is written as
+The transformation is formally stated as
 \begin{align}
@@
 The steering vectors in $\mat{B}$, the so-called Butler matrix, determine each beam's direction in space.
-We can therefore remove beams where no interference is present by removing rows from $\mat{B}$.
+We can therefore remove beams where no interference is present by simply removing rows from $\mat{B}$.
 
@@
 
-The beamspace version of the Capon beamformer is derived by replacing the covariance matrix in (\ref{eq:w}) with the beamspace covariance matrix $\mat{\hat{R}}_{BS} = f(\vec{x}_{BS}) = f(\mat{B}\vec{x})$, where $\mat{B} \in \mathbb{C}^{N_b,L}$, and $N_b$ is the number of selected beams.
+The beamspace version of the Capon beamformer is derived by replacing the covariance matrix in (\ref{eq:w}) with the beamspace covariance matrix 
+\begin{align}
+\mat{\hat{R}}_{BS} = f(\vec{x}_{BS}) = f(\mat{B}\vec{x}), 
+\end{align}
+where $\mat{B} \in \mathbb{C}^{N_b,L}$, and $N_b$ is the number of selected beams.
 The transformation must be performed per subarray, yielding subarrays that no longer overlap in space.
@@
 
-From (\ref{eq:sliding_bs}) it is obvious that instructions can be saved if we utilize the overlap across subarrays, with an increase in task parallel granularity (As discussed in Section \ref{sec:calcR}).
-With SBS we can therefore select to have one thread per beam (index $k$) that calculates this beam for all subarrays.
-However, since the number of beams usually is small ($N_b \ll M$) we will end up with a large memory per thread ratio.
+Equation (\ref{eq:sliding_bs}) shows that instructions are saved if we utilize the overlap across subarrays, but with an increase in task parallel granularity (As discussed in Section \ref{sec:calcR}).
+The most obvious is to use one thread per beam (index $k$) that calculates this beam for all subarrays.
+However, since the number of selected beams usually is small ($N_b \ll M$) we will end up with a large memory per thread ratio.
 Using one thread per beam is therefore not feasible for small $N_b$ large $M$ configurations.
@@
 
-As mentioned the amount of data is increased from $M$ to $N_LN_b$ per sample in beamspace.
+As mentioned the amount of channel data is increased from $M$ to $N_LN_b$ per sample in beamspace.
 This means that we no longer can exploit spatial overlap when calculating the sample covariance matrix, but since $N_b$ is small we can now allow to have one thread per element in $\mat{\hat{R}}_{BS}$.
-A kernel similar to the one described in \cite{Chen2011}, has therefore been implemented.
-For small $N_b$ values, several covariance matrices are calculated per thread block to achieve high occupancy of the SMs.
+A kernel similar to the one described in \cite{Chen2011} has therefore been implemented.
+In addition, for small $N_b$ values, several covariance matrices are calculated per thread block to achieve high occupancy of the GPU.
 
@@
 \end{align}  
-and applied to the beamspace data summed across subarrays.
+and applied to the beamspace data summed across subarrays,
+\begin{align}
+z[n] = \vec{w}_{BS}^H\sum_{l=0}^{L-1}\vec{x}_{BS,l}. 
+\end{align}
 
@@
 As presented in (\ref{eq:diag}), adaptive diagonal loading is performed by loading the sample covariance matrix with the average energy per element, hence $trace\{\mat{\hat{R}}\}/L$.
-If adaptive diagonal loading is applied after data have been transformed into beamspace, a high percentage of the total energy is mapped to a small subset of beams.
+When data are transformed into beamspace, a high percentage of the total energy is mapped to a small subset of beams.
 Thus the average energy per beam is higher than the average energy was per element, when beams containing little energy are removed.
@@
 The added complexity of SBS is $O(N_bL)$.
+To maximize performance, large adjustments to the processing pipeline described in Section \ref{sec:meth} is required.
 
@@
 For larger arrays, more transmit beams have to be emitted for sufficient lateral sampling.
-(a) Shows the benchmark for a 32 element array.
-(b) Shows the benchmark for a 64 element array.}
+(a)  32 element array.
+(b)  64 element array.}
 \label{fig:bench}
@@
 \ref{fig:bench}.
-(a) Shows the benchmark for a 64 elements array.
-(b) Shows the benchmark for a 96 elements array.}
+(a) 64 elements array.
+(b) 96 elements array.}
 \label{fig:benchBS}
 \end{figure*}
-
-Fig.
-\ref{fig:bench} shows the results from benchmarking the implementation described in Section \ref{sec:meth} on an Nvidia Quadro 6000 graphics card (14 SMs, 1 TFlops theoretical single precision, 6 GB global memory, and 144 GB/s memory bandwidth).
-The problem sizes are calculated based on 32 and 64 element phased arrays with center frequency 2.5 MHz.
-By assuming a system bandwidth of 80\% we need a complex sampling frequency of 2 MHz.
-If we image down to a depth of 15 cm we get approximately 400 samples per range line.
-To correctly sample a $70\degree$ sector with 64 and 32 element arrays ($\lambda/2$ pitch), we need at least 80 and 40 beams, respectively.
+In this section we present benchmarks of the presented implementations performed on a Nvidia Quadro 6000 graphics card (14 SMs, 1 TFlops theoretical single precision, 6 GB global memory, and 144 GB/s memory bandwidth).
+Problem sizes are calculated based on 32, 64, and 96 element phased arrays with a 2.5 MHz center frequency.
+Assuming a system bandwidth of 80\% we need a complex sampling frequency of 2 MHz, and when imaging down to a depth of 15 cm we end up with approximately 400 samples per range line.
+To correctly sample a $70\degree$ sector with 32, 64 and 96 element arrays ($\lambda/2$ pitch), we need at least 40, 80 and 120 beams respectively.
 The presented execution times do not include transfer of data from CPU-side to the GPU, since data are assumed to be pre-delayed and available in GPU memory.
 What is included are all calculations, memory transfers from global to near-core GPU memory, and near core memory access.
+
+A benchmark of the implementation described in Section \ref{sec:meth} is presented in Fig.
+\ref{fig:bench}.
+Benchmarks are based on 32 and 64 elements phased arrays respectively, using 40 and 80 transmit beams.
+For each subarray size the total execution time is split into one bar for each of the three steps described in Section \ref{sec:meth}, and calculation of $\mat{\hat{R}}$ is presented for three different values of $K$.
 For $M=64$, $L=32$ and $K=2$ (Fig.
-\ref{fig:benchUS64}), a typical setup for high quality imaging, processing throughput is 10 frames per second.
-
+\ref{fig:benchUS64}), a typical setup for high quality cardiac imaging, processing throughput is 10 frames per second.
 In Fig.
-\ref{fig:benchBS} we present similar benchmarks for the beamspace version of the Capon beamformer.
-Benchmarks are based on a 64 and 96 elements phased array respectively.
-For the latter, number of transmit beams is increased to 120.
+\ref{fig:benchBS} we present similar benchmarks for the beamspace version of the Capon beamformer introduced in Section \ref{sec:bs}.
+Benchmarks are based on 64 and 96 elements phased arrays, using 80 and 120 transmit beams.
 The figure only present results with use of temporal averaging ($K=2$).
 For $M=96$, $L=48$, $K=2$ and $N_b=3$, processing throughput is 100 frames per second.
-
-For all benchmarks there is a line indicating the time it takes to acquire the underlying image using single line acquisition. Hence, below this line we are capable of real time processing.
-
-\section{Simulated phantom and cardiac data}
+For all benchmarks there is a line indicating the time it takes to acquire the underlying image using single line acquisition.
+Hence, below this line we are capable of real time processing.
+
+\section{Simulated phantom and \textit{in-vivo} cardiac data}
 
@@
 
-Fig.
-\ref{fig:phantom} presents a phantom simulated using Field II \cite{Jensen1992}.
+To investigate the image quality of both the Capon and BS Capon beamformer a phantom was simulated using Field II \cite{Jensen1992} (Fig.
+\ref{fig:phantom}).
 The simulation was performed using a 2.5 MHz, $\lambda/2$ pitch, 10 $\mu$m kerf, 50\% bandwidth and 10 mm high phased array.
-The transducer was exited with a $3\lambda$ long pulse and the transmit focus is at 80 mm.
+The transducer was exited with a $3\lambda$ long pulse with transmit focus at 80 mm.
 The phantom contains two circular areas, one bright at 24 dB and one dark at -30 dB, and two point targets at 40 dB.
@@
 Fig.
-\ref{fig:invivo} shows an \textit{in-vivo} image of the left ventricle.
-The image was acquired using a experimental scanner at MI-Lab in Trondheim using a 3.5 MHz phased array in harmonic mode.
-The image is extracted from a cardiac cycle, that can be viewed in its entirety in an attached video.
+\ref{fig:invivo} presents results of applying Capon and BS Capon on an \textit{in-vivo} image of the left ventricle.
+The image was acquired using a experimental research scanner at MI-Lab in Trondheim using a 3.5 MHz phased array in harmonic mode.
+The image is extracted from a cardiac cycle, that can be viewed in its full length in an attached video.
+Its purpose is to show that Capon beamforming in both element- and beam-space is feasible for cardiac imaging, not just in simulations.
+
+To get a better impression of the high resolution images obtained with Capon beamforming, two slices taken from the images in Fig.
+\ref{fig:phantom} and \ref{fig:invivo} is included in Fig.
+\ref{fig:slices}.
 
 Both data sets have been processed with the following Capon beamformer settings: $L = 32$, $K = 2$, $d=0.1$.
-These settings have previously been shown to provide high resolution while maintaining DAS-like speckle in ultrasound images \cite{Synnevag2007a}.
+These settings have previously been shown to provide high resolution while maintaining DAS-like speckle\cite{Synnevag2007a}.
 For the beamspace processing the following parameters have been used: $L=32$, $K=2$, $d=0.01$, $N_b=3$.
-
 For the simulation, 16x lateral over-sampling has been used, in order to avoid under sampling of the two bright point scatterers as they move.
@@
 As we see from Fig.
-\ref{fig:bench} and \ref{fig:benchBS}, the execution time is reduced when either $L$ or $N_b$ is reduced.
-The question is how this impact the image quality.
+\ref{fig:bench} and \ref{fig:benchBS}, execution time is reduced when either $L$ or $N_b$ is reduced.
+The question is how does this impact image quality.
 In Fig.
-\ref{fig:speed_res_trade} the minimum distance between the two point targets in Fig.
+\ref{fig:speed_res_trade} minimum distance between the two point targets in Fig.
 \ref{fig:phantom} (while preserving a 6dB lower level in between) is plotted against $L$ (when processed with Capon) and $N_b$ (when processed with BS Capon).
@@
 For this result, $d=0.01$ has been used on both methods.
+Why resolution decreases for BS Capon for small $N_b$s is partially explained by reduced information of interference at higher angles, and automatically increased diagonal loading.
+It is also interesting to see that Capon beamforming can perform worse than DAS for small subarrays.
 

### 2013-01-25 HRUIonGPU/jrnl_article.tex
uffc

@@
 \end{figure*}
-In this section we present benchmarks of the presented implementations performed on a Nvidia Quadro 6000 graphics card (14 SMs, 1 TFlops theoretical single precision, 6 GB global memory, and 144 GB/s memory bandwidth).
+In this section we present benchmarks of the implementations presented in Section \ref{sec:meth} and \ref{sec:bs} performed on a Nvidia Quadro 6000 graphics card (14 SMs, 1 TFlops theoretical single precision, 6 GB global memory, and 144 GB/s memory bandwidth).
 Problem sizes are calculated based on 32, 64, and 96 element phased arrays with a 2.5 MHz center frequency.
@@
 \ref{fig:benchUS64}), a typical setup for high quality cardiac imaging, processing throughput is 10 frames per second.
+
 In Fig.
@@
 For $M=96$, $L=48$, $K=2$ and $N_b=3$, processing throughput is 100 frames per second.
-For all benchmarks there is a line indicating the time it takes to acquire the underlying image using single line acquisition.
-Hence, below this line we are capable of real time processing.
-
-\section{Simulated phantom and \textit{in-vivo} cardiac data}
+For all four benchmarks there is a line indicating the time it takes to acquire the underlying image using single line acquisition.
+Hence, below this line we are capable of real time processing when a single receive line is used per transmit line.
+
+\section{Simulated phantom and \textit{in-vivo} cardiac images}
 
@@
 \ref{fig:invivo} presents results of applying Capon and BS Capon on an \textit{in-vivo} image of the left ventricle.
-The image was acquired using a experimental research scanner at MI-Lab in Trondheim using a 3.5 MHz phased array in harmonic mode.
+The image was acquired using a experimental research scanner at MI-Lab in Trondheim using a 64 elements, 3.5 MHz, and $\lambda/2$ pitched phased array in harmonic mode.
 The image is extracted from a cardiac cycle, that can be viewed in its full length in an attached video.
-Its purpose is to show that Capon beamforming in both element- and beam-space is feasible for cardiac imaging, not just in simulations.
+Its is included to show that Capon beamforming in both element- and beam-space is feasible for cardiac imaging, not just in simulations.
 
@@
 \ref{fig:slices}.
+For the phantom, the cut goes through the two point targets at 93 mm depth, and for the cardiac image it goes through the tip of the mitral valve at 96 mm depth.
 
@@
 For the beamspace processing the following parameters have been used: $L=32$, $K=2$, $d=0.01$, $N_b=3$.
-For the simulation, 16x lateral over-sampling has been used, in order to avoid under sampling of the two bright point scatterers as they move.
+For the simulated phantom, 16x lateral over-sampling has been used, in order to avoid under sampling of the two bright point scatterers as they move.
 The cardiac recording only has $30\%$ lateral over sampling.
@@
 \ref{fig:bench} and \ref{fig:benchBS}, execution time is reduced when either $L$ or $N_b$ is reduced.
+So to gain an  increase in processing throughput we can reduce $L$ or $N_b$.
 The question is how does this impact image quality.
@@
 \ref{fig:phantom} (while preserving a 6dB lower level in between) is plotted against $L$ (when processed with Capon) and $N_b$ (when processed with BS Capon).
-For high values the two methods are equal, but for lower values BS Capon provides higher resolution than what is possible with small subarrays.
+For high values the two methods are equal, but for lower values, BS Capon provides higher resolution than what is possible with small subarrays.
 For this result, $d=0.01$ has been used on both methods.
-Why resolution decreases for BS Capon for small $N_b$s is partially explained by reduced information of interference at higher angles, and automatically increased diagonal loading.
-It is also interesting to see that Capon beamforming can perform worse than DAS for small subarrays.
+Why resolution decreases for BS Capon for small $N_b$s is partially explained by reduced information of interference at higher angles, and an automatically increase in diagonal loading.
+It is also interesting to see that Capon beamforming performs worse than DAS for small subarrays.
+(Comment on beamspace).
 
@@
 \section{Discussion}\label{sec:dis}
-We have presented a parallel implementation of both the Capon og BS Capon beamformer.
+In this paper we have presented a parallel implementation of the Capon beamformer both in element- and beam-space.
+The processing rates acquired shows that it is now possible to apply BS Capon in real time in cardiac ultrasound imaging.
+From Fig.
+...
+2MLA is supported.
 
@@
 
-Toeplitz-based solver (cite new article). Discuss it.
-
 How does beamspace perform when using plane wave imaging...?
+
+MLA's has, to some extent be used to increase later sampling.
+When using a high resolution beamformer, Nyquist tells us that the sampling frequency must be increase.
+Future work will concentrate on how much over sampling is required when using the Capon beamformer whit i curtain set of parameters.
 

### 2013-01-29 HRUIonGPU/jrnl_article.tex
uffc

@@
 %\boldmath
-...
+Capon beamforming for   
 \end{abstract}
@@
 In Section \ref{sec:meth} we go in detail about the parallel implementation of the Capon beamformer, before in Section \ref{sec:bs} we introduce the beamspace method and how it has been implemented on the GPU.
-Benchmark results and resulting images are presented in Section \ref{sec:res}, and finally we discuss the results in Section \ref{sec:dis}, and draw conclusions in Section \ref{sec:con}.
+Benchmark results and resulting images are presented in Section \ref{sec:bench} and \ref{sec:images}, together with a discussion of trading image quality for speed in Section \ref{sec:trade}.
+Finally we discuss the results in Section \ref{sec:dis}, and draw conclusions in Section \ref{sec:con}.
 
@@
 This will minimize global reads and writes needed per thread, but unfortunately the maximum number of threads per block puts an upper limit on the subarray length, $L_{max}=32$, since we want to avoid block-to-block synchronization.
-Since arrays of 96-128 elements are common in medical ultrasound imaging an implementation not limited by $L_max=32$ is preferable.
+Since arrays of 96-128 elements are common in medical ultrasound imaging an implementation not limited by $L_{max}=32$ is preferable.
 
@@
 
-For the results presented in this paper, we have used a GPU implementation of batched Gauss-Jordan (GJ) elimination made by Nvidia to solve $\mat{\hat{R}}\vec{b} = \vec{1}_L$, where $\vec{b} = \mat{\hat{R}}^{-1}\vec{1}_L$ for all pixels at once.
-The solver has been made available for registered CUDA developers.%For smaller systems, $L < 10$, a direct solver using one thread per system is preferable.
+For the results presented in this paper, we have used a GPU implementation of batched Gauss-Jordan (GJ) elimination made by Nvidia to solve $\mat{\hat{R}}\vec{b} = \vec{1}_L$, where $\vec{b} = \mat{\hat{R}}^{-1}\vec{1}_L$, for all pixels at once.
+The solver has been made available online for registered CUDA developers.%For smaller systems, $L < 10$, a direct solver using one thread per system is preferable.
 For this purpose we have implemented Cramer's rule for $2 \times 2$, $3 \times 3$ and $4 \times 4$ matrices.
@@
 The final output, as shown in (\ref{eq:z_mv}) and (\ref{eq:z_mv2}), can be formed by either reducing the data down to length $L$, or increase the weight vector to length $M$.
-Since the design is $L$ threads per output, the data is reduced to length $L$ by letting thread $r$ calculate
+Since the design is using $L$ threads per output, the data is reduced to length $L$ by letting thread $r$ calculate
 \begin{align}\label{eq:subsum}
@@
 where $\mat{B} \in \mathbb{C}^{N_b,L}$, and $N_b$ is the number of selected beams.
-The transformation must be performed per subarray, yielding subarrays that no longer overlap in space.
-Therefore, if the transformation is applied before the covariance matrix is constructed, the amount of input data is change from $M$ to $N_LN_b$ elements per sample, however all remaining steps will get lower complexity when $N_b < L$.
+If the transformation is applied before the covariance matrix is constructed, one transformation must be performed per subarray, yielding subarrays that no longer overlap in space.
+The amount of input data is then change from $M$ to $N_LN_b$ elements per sample, however all remaining steps will get lower complexity when $N_b < L$.
 
@@
 
-Equation (\ref{eq:sliding_bs}) shows that instructions are saved if we utilize the overlap across subarrays, but with an increase in task parallel granularity (As discussed in Section \ref{sec:calcR}).
+Equation (\ref{eq:sliding_bs}) shows that instructions are spared if we utilize the overlap across subarrays, but with an increase in task parallel granularity (As discussed in Section \ref{sec:calcR}).
 The most obvious is to use one thread per beam (index $k$) that calculates this beam for all subarrays.
@@
 
-As mentioned the amount of channel data is increased from $M$ to $N_LN_b$ per sample in beamspace.
-This means that we no longer can exploit spatial overlap when calculating the sample covariance matrix, but since $N_b$ is small we can now allow to have one thread per element in $\mat{\hat{R}}_{BS}$.
+As mentioned, the amount of channel data is increased from $M$ to $N_LN_b$ per sample when using SBS.
+This means that the spatial overlap exploited when calculating the sample covariance matrix in element space is no longer available.
+On the other hand, since $N_b$ should be small ($N_b < 32$), one thread per element in $\mat{\hat{R}}_{BS}$ is now doable.
 A kernel similar to the one described in \cite{Chen2011} has therefore been implemented.
@@
 \end{align}  
-and applied to the beamspace data summed across subarrays,
+using on thread per beam, and applied to the beamspace data summed across subarrays,
 \begin{align}
@@
 \end{align}
+And again are several pixels calculated per thread block for small values of $N_b$ in order to achieve high GPU occupancy.
+The beamspace data summed across subarrays are calculated and outputted as part of the SBS kernel.
 
@@
 When data are transformed into beamspace, a high percentage of the total energy is mapped to a small subset of beams.
-Thus the average energy per beam is higher than the average energy was per element, when beams containing little energy are removed.
+Thus the average energy per beam is higher than the average energy was per element when beams containing little energy are removed.
 This means that less diagonal loading (smaller $d$) has to be applied in beamspace than in element space when $N_b < L$.
@@
 
-\section{Benchmarks}\label{sec:res}
+\section{Benchmarks}\label{sec:bench}
 
@@
 \begin{figure*}[!t]
-\centerline{\subfloat[]{\includegraphics[width=3.5in]{gfx/benchmark_bar_bs_M=64_Nx=80_Ny=400_Nb=3.png} \label{fig:benchBS3}}
+\centerline{\subfloat[]{\includegraphics[width=3.5in]{gfx/benchmark_bar_bs_M=64_Nx=80_Ny=400_Nb=3.png} \label{fig:benchBS64}}
 \hfil
 \subfloat[]{\includegraphics[width=3.5in]{gfx/benchmark_bar_bs_M=96_Nx=120_Ny=400_Nb=3.png}%
-\label{fig:benchBS5}}}
+\label{fig:benchBS96}}}
 \caption{Benchmark of GPU beamspace Capon beamforming for a cardiac image covering a $70 \degree$ and $15$cm deep sector.
@@
 
-\section{Simulated phantom and \textit{in-vivo} cardiac images}
+\section{Simulated phantom and \textit{in-vivo} cardiac images}\label{sec:images}
 
@@
 
-\section{Trading resolution for speed}
+\section{Trading resolution for speed}\label{sec:trade}
 As we see from Fig.
@@
 Why resolution decreases for BS Capon for small $N_b$s is partially explained by reduced information of interference at higher angles, and an automatically increase in diagonal loading.
-It is also interesting to see that Capon beamforming performs worse than DAS for small subarrays.
-(Comment on beamspace).
+It is also interesting to note that Capon beamforming in this example performs worse than DAS for small subarrays.
 
@@
 \section{Discussion}\label{sec:dis}
-In this paper we have presented a parallel implementation of the Capon beamformer both in element- and beam-space.
-The processing rates acquired shows that it is now possible to apply BS Capon in real time in cardiac ultrasound imaging.
-From Fig.
-...
-2MLA is supported.
-
+In this paper we have presented a parallel implementation of the Capon beamformer in both element- and beam-space for high resolution cardiac ultrasound imaging.
+The results show that it is now possible to apply BS Capon in real time in cardiac ultrasound imaging utilizing the power of GPU's.
+According to Fig.
+\ref{fig:benchBS} two parallel receive lines is also within reach (for both $M=64$ and $M=96$.
+$L=M/2$), but at the moment it is unknown if these extra lines is needed for increased sampling or not.
+Another concern is that lateral oversampling using parallel receive lines will increase the total execution time with a factor equal the amount of oversampling.
+
+For Capon in element space, processing rates for high quality settings is still to high to support real time cardiac ultrasound imaging.
 From Fig.
 \ref{fig:benchUS64} we see that for $L = M/2 = 32$ and $K=1$, the total execution time is 100 ms (10 fps).
-Taking into account that we use complex data, but less samples per image, this is similar to the performance reported by Chen et al.
+Taking into account that we use complex data, but less samples per image, these results are similar to the performance reported by Chen et al.
 \cite{Chen2011}.
 For the 32 element array in Fig.
-(\ref{fig:benchUS32}), execution time is reduced to 11.5 ms (87 fps) using the same settings ($L=M/2, K = 1$).
-Doubling this number can be interpreted as pre-beamforming a 64 element array down to 32.
-The frame rate for a 64 element array is then increased from 10 to 44 fps.
-Note that lateral oversampling will increase the total execution time with a factor equal the amount of oversampling.
-
-A major part of the total execution time comes from solving systems of linear equations. This limitation is targeted by reducing the sample covariance matrix size in a novel way \cite{Nilsen2009}.
-
-We have shown that the Capon beamformer can be implemented in a GPU framework.
-And if we as in \cite{Chen2011} define real-time as 10 fps, all execution times  in Fig.
-(\ref{fig:bench}) are real time.
-If we on the other hand define real-time as the time it takes to acquire a given image, the image formed using a 32 and 64 element array has a real time requirement of 125 fps and 62.5 fps respectively.
-There is some distance to cover before that goal is reached for all values of $L \le M/2$.
-However we see that real time, by the last definition, is possible for $M=64, L < 16$, and for $M = 32, L < 13$.
-Another option to reach real-time execution is to use several GPUs simultaneously.
-Note that multi-line acquisition will increase the frame rate requirement accordingly.
-
-In the attached video of a simulated phantom one can see that Capon beamforming highly resolves moving point targets.
-The beamformer is shift invariant as long as the lateral sampling is increased.
-No major improvement in contrast is observed.
-In the cardiac recording, structures are thinner when the Capon beamformer is applied, however we observe a loss in contrast due to signal cancellation.
+(\ref{fig:benchUS32}) execution time is reduced to 11.5 ms (87 fps) using the same settings ($L=M/2, K = 1$).
+
+A major part of the total execution time for element space Capon comes from solving the system of linear equations.
+This limitation is targeted by reducing the sample covariance matrix size in beamspace.
+As shown in Fig.
+\ref{fig:benchBS}, inverting the three times three beamspace covariance matrix is barely visible.
+The time taken to construct all covariance matrices is also reduced, but for few beams this is now the major contributor to the total time.
+If real time is defined as 10 fps \cite{Chen2011} all execution times in both Fig.
+\ref{fig:bench} and \ref{fig:benchBS} are real time.
+More natural for cardiac ultrasound imaging is to define real-time as the time it takes to acquire the underlying image.
+With this requirement the images formed using a 32, 64 and 96 element array has a real time requirement of 125, 62.5 and 41.7 fps respectively using single line acquisition.
+Each benchmark plot has a line indicating this real time requirement.
+With beamspace, all processing times for large subarrays are below this line.
+
+This paper further shows how to obtain a parallel Capon beamformer, and how it can be implemented in a GPU framework.
+The per-pixel granularity suggest that the beamformer is easily dividable across multiple GPUs.
+Another option to reach real-time execution for all cases presented in this paper is therefore to use several GPUs simultaneously.
+In that case, processing rates are increased with a factor equal to the number GPUs.
+However, there is usually a space-cost-energy limitation for how many GPUs that can be fitted in an ultrasound scanner.
+
+Turning to image quality we see from Fig.
+\ref{fig:phantom}, and the attached video of the simulation, that Capon beamforming highly resolves moving point targets both in element- and in a reduced beam-space.
+Our preliminary finding is that the Capon beamformer is shift invariant as long as the lateral sampling is increased enough.
+No major improvement in contrast is visually observed in these images.
+In the cardiac recording (Fig.
+\ref{fig:invivo}), structures are thinner and better resolved both when the Capon and BS Capon beamformer is applied.
+However we observe a loss in contrast, especially in the septum (Fig.
+\ref{fig:slices}).
+This loss can be partially explained by the present of highly correlated signals and errors in the assumed steering vector.
 Future work should focus on solving this issue.
 
-We expect that a large speedup had the Capon beamformer been implemented on the CPU using threads and SIMD (single instruction multiple data) instructions.
-
-Solving linear systems on the GPU is not believed to be much faster than a CPU implementation, but there is two key points.
-First the CPU is offloaded, and second the time it takes to transfer all covariance matrices back to the CPU makes the choice of doing the solving on the GPU simple to make.
-
-(results)?
-For a typical cardiac ultrasound image of 80*400 pixels (70 degree sector, 15cm range) acquired using a 2.5MHz, $M=64$ element phased array, the result is 10fps (subarray length $L=M/2$, temporal smoothing over 3 samples).
+Nilsen et.
+al suggest to do automatic selection of beams in beamspace, hence picking the beams containing most energy.
+In this paper the center beams are always picked.
+This seems to work fine, but an adaptive pick of beams should be implemented if interference at high angles is shown to contaminate the image.
+
+%We expect that a large speedup had the Capon beamformer been implemented on the CPU using threads and SIMD (single instruction multiple data) instructions.
+
+%For a typical cardiac ultrasound image of 80*400 pixels (70 degree sector, 15cm range) acquired using a 2.5MHz, $M=64$ element phased array, the result is 10fps (subarray length $L=M/2$, temporal smoothing over 3 samples).
 If we do a 2-element pre-beamforming to get the channel count down to 32, the frame rate increases to 44fps.
@@
 
-How does beamspace perform when using plane wave imaging...?
-
-MLA's has, to some extent be used to increase later sampling.
+\section{Conclusion}\label{sec:con}
+In this paper we have shown that the Capon beamformer can be accelerated using GPUs.
+For beamspace Capon beamforming processing rates supporting real time cardiac ultrasound imaging is shown to be possible.
+Image quality is investigated both in a simulated phantom and in \textit{in-vivo} cardiac data.
+Results shows that Capon beamforming is feasible for cardiac ultrasound imaging, providing images with highly resolved structures.
+
+\section{Future work}
+Study shift invariance and minimum sampling...
+
+MLA's has, to some extent been used to increase later sampling.
 When using a high resolution beamformer, Nyquist tells us that the sampling frequency must be increase.
-Future work will concentrate on how much over sampling is required when using the Capon beamformer whit i curtain set of parameters.
-
-\section{Conclusion}\label{sec:con}
-...
-
-\section{Future work}
+Future work will concentrate on how much over sampling is required when using the Capon beamformer whit i curtain set of parameters...
+
 Plane waves...

### 2013-02-14 HRUIonGPU/jrnl_article.tex
Integrated feedback from Laila

@@
 \newcommand{\degree}{\ensuremath{^\circ}}
+\newcommand{\R}{$\mat{\hat{R}}$ }
 
@@
 % The paper headers
-\markboth{Transactions on Ultrasonics and Ferroelectrics and Frequency Control,~Vol.~x, No.~y, January~2013?}%
+\markboth{Transactions on Ultrasonics and Ferroelectrics and Frequency Control,~Vol.~x, No.~y, March~2013}%
 {\AA{}sen \MakeLowercase{\textit{et al.}}: Capon Adaptive Beamforming on the GPU}
@@
 %\boldmath
-Capon beamforming for   
+Capon beamforming is associated with a high computational complexity, which limits its use as a real time array processor in many applications.
+In this paper we present an implementation of the Capon beamformer exhibiting real-time performance when applied in a typical cardiac ultrasound imaging setting.
+To accomplish this task we make use of the massive parallel processing power found in modern Graphics Processing Units (GPUs), combined with beamspace processing to reduce the computational complexity as the problem size grows.
+For  beamspace Capon beamforming processing rates supporting real time cardiac ultrasound imaging is shown to be possible, meaning that processing rates higher than the image acquisition rate are obtained for a wide range of parameters.
+Image quality is investigated both in a simulated phantom and an \textit{in-vivo} cardiac dataset.
+These results show that Capon beamforming is feasible for cardiac ultrasound imaging, providing images with sharper structures and no local shift-variance effects.
 \end{abstract}
@@
 Under this constraint the Capon beamformer is not allowed to micro-steer the main lobe, and therefore its full potential is not reached.
-Micro-steering of the main lobe is an important property of the Capon beamformer in order for it to improve structural details and edge definitions throughout an ultrasound image.
+Micro-steering of the main lobe is an important property of the Capon beamformer in order for it to improve structural details and edge definitions throughout an ultrasound image (Fig 9.
+in \cite{Synnevag2009}).
 Another aspect of using complex data is that even though the number of arithmetic operations per sample is increased we can sample at the system bandwidth, keeping the number of samples for a full cardiac imaging sector at a minimum.
-All results in this paper is reported with the use of baseband data.
-Meaning that the sampling frequency is close to the system bandwidth.
+All results in this paper are reported with the use of baseband data, hence the sampling frequency is close to the system bandwidth.
 
@@
 In recent work \cite{asen2012}(cite JOE or ECUA?) we have applied a GPU-based Capon beamformer in both an active sonar system and for cardiac ultrasound imaging.
-For sonar the real-time requirement is much lower than for medical ultrasound imaging, which means that most common array configurations can be handled using the GPU without taking any additional measures.
+For sonar the real-time requirement is much lower than for medical ultrasound imaging, which means that most array configurations can be handled in real time.
 For medical ultrasound imaging, and especially cardiac ultrasound imaging, frame rate requirements increase to a level where even a high-end GPU can not do the processing required for a standard implementation of the Capon beamformer.
-Since the execution time highly depends on the number of elements in the given array, it is important to find ways of reducing computational complexity for high-channel-count systems.
+Since execution time highly depends on the number of elements in the given array, it is important to reduce the computational complexity for high-channel-count systems.
 Different approaches to low-complexity Capon beamforming have therefore recently been proposed \cite{Synnevag2011, Asl2012, Jensen2012, Kim}.
@@
 In the next section we present background information about the Capon beamformer and the GPU computing model.
-In Section \ref{sec:meth} we go in detail about the parallel implementation of the Capon beamformer, before in Section \ref{sec:bs} we introduce the beamspace method and how it has been implemented on the GPU.
+In Section \ref{sec:meth} we go in detail about the parallel implementation of the Capon beamformer, before the beamspace method and how it has been implemented on the GPU is presented in Section \ref{sec:bs}.
 Benchmark results and resulting images are presented in Section \ref{sec:bench} and \ref{sec:images}, together with a discussion of trading image quality for speed in Section \ref{sec:trade}.
@@
 Here, the signal recorded at element number $m \in [0,M-1]$ at time $n$ is delayed with an appropriate delay $\Delta_m[n]$ to focus and steer the ultrasound beam in space.
-This delayed data is denoted $x_m[n]$.
+This delayed data is here denoted $x_m[n]$.
 The beamformer output, $z[n]$, is then the sum of all $M$ channels weighted with a per-element factor $w_m$,
@@
 The weight vector or window $\vec{w}$ is usually real, and is applied to trade resolution for a better signal-to-noise ratio (SNR).
-And even if different windows can be used in different areas of the images they have traditionally been predefined.
+Even if different windows can be used in different areas of the images they have traditionally been predefined.
 
@@
 \ref{fig:mvbf}, the basic idea of an adaptive beamformer is to form a complex weight vector based on the received data.
-For Capon beamforming the weights are found that minimize the beamformer output power while maintaining unit gain at broadside.
+For Capon beamforming the weights are found that minimize the beamformer output power while maintaining unity gain at broadside.
 The result is that interference impinging from other directions than the point of focus is suppressed \cite{Synnevag2007}.
-%\textbf{(Vi skal implementer dette. vis bare vektene. Dropp ligning 2, siter, bruk R-hat med en gang. Få med den endelige feilen )}
+It is worth noting that the micro steering caused by the complex weights is minute compared to the pre-steering and -focusing step, therefore the beamformer also works well for broadband ultrasound signals.
 
@@
 
-The Capon beamformer can be summerized in three main steps. First a sample covariance matrix has to be calculated, 
+The Capon beamformer can be summarized in three main steps. First a sample covariance matrix has to be calculated, 
 \begin{align}
@@
 The key point is that GPUs can accelerate small concurrent tasks, like beamforming one pixel, much better than a CPU.
-Where the CPU has few cores and logic to process complex threads, the GPU has many cores, but each thread needs to be simple in order to benefit from the GPU's computational power.
+Where the CPU has few cores and logic to process complex threads, the GPU has many cores, but each thread needs to be simple in order to maximize the benefit of using GPUs.
 
@@
 An SM has available a limited amount of near-core memory and registers.
-Each thread therefore has to be as light on memory as possible and has to perform enough instructions per used byte in order to hide memory latency.
-It is also important that the problem can be divided into enough threads to hide instruction latency.
+Hence, to achieve maximum GPU occupancy, each thread has to be as light on registers and memory as possible.
+It is also important that each thread performs enough instructions per used byte in order to hide memory latency, and that the problem can be divided into enough threads to hide instruction latency.
 Global reads and writes on the GPU, in addition to CPU-GPU transfers, have high penalty, and should be minimized as much as possible.
@@
 \section{Parallel Capon}\label{sec:meth}
-If we analyze the beamformer in (\ref{eq:z_mv}) in combination with (\ref{eq:b}) and (\ref{eq:R}), we see that each calculation of a set of Capon weights is independent across the outputs $z[n]$.
+If the beamformer in (\ref{eq:z_mv}) is analyzed in combination with (\ref{eq:b}) and (\ref{eq:diag}), we see that each calculation of a set of Capon weights is independent across the outputs $z[n]$.
 As shown in Fig.
-\ref{fig:gpulayout}, a first level of granularity is therefore to divide the image into a grid, where the output in each cell can be calculated independently of all others.
+\ref{fig:gpulayout}, a first level of granularity is therefore to divide the image into a grid, where the output in each cell can be calculated independently.
 Further, based on memory dependencies and to hold memory consumption per thread at a minimum, one should recognise that a group of threads should process one output in the final image.
@@
 
-\subsection{Calculating Multiple Sample Covariance Matrices}\label{sec:calcR}
+\subsection{Calculation of Multiple Sample Covariance Matrices}\label{sec:calcR}
 
@@
 
-From (\ref{eq:R}) it is obvious that calculating one element of $\mat{\hat{R}}$ is independent of all other elements.
+Calculation of one element of \R is from (\ref{eq:R}) found to be independent of calculation of all the other elements.
 A natural level of granularity would then be to assign one thread to each element of $\mat{\hat{R}}$ \cite{Chen2011}.
-This will minimize global reads and writes needed per thread, but unfortunately the maximum number of threads per block puts an upper limit on the subarray length, $L_{max}=32$, since we want to avoid block-to-block synchronization.
-Since arrays of 96-128 elements are common in medical ultrasound imaging an implementation not limited by $L_{max}=32$ is preferable.
+This will minimize global reads and writes needed per thread, but since we want to avoid block-to-block synchronization, the maximum subarray length is limited to 32 by this approach.
+Since most medical ultrasound arrays have more than or exact 64 elements, an implementation not limited by $L_{max}=32$ is preferable.
 
@@
 \end{align}
-The same is true in time where
+In time we have that
 \begin{align}
@@
 %- \mat{\breve{R}}[n-K-1] + \mat{\breve{R}}[n-K].
-Both sliding across subarrays and in time can be realized simultaneously if we hold a $N_K$-average in time of the outer products $\vec{x}[n]\vec{x}[n]^H$ in memory and calculate $\mat{\hat{R}}$ averaged over subarrays from this buffer (cite JOE?).
-However, since $N_K$ typically is small, and we have a limited amount of memory available per thread, we focus on sliding across subarrays only.
+Both sliding across subarrays and in time can be realized simultaneously if $\mat{\hat{R}}$ averaged over subarrays is calculate from an $N_K$-average in time of the outer products $\vec{x}[n]\vec{x}[n]^H$ (cite JOE?).
+However, since $N_K$ typically is small, and the GPU has a limited amount of memory available per thread, we focus on sliding across subarrays only.
 In that case, the complexity of calculating $\mat{\hat{R}}$ is reduced to $O(N_KL^2)$.
-Note that sliding will break independence along the sliding dimension.
-With our choice of calculating $\mat{\hat{R}}$, calculations are still independent across time, but not along diagonals in $\mat{\hat{R}}$.
+Note that sliding will break independence along the sliding dimension, thus, calculations are still independent across time, but not along diagonals of $\mat{\hat{R}}$.
 One thread is therefore used per diagonal in the upper triangle of $\mat{\hat{R}}$, and for small values of $L$, several covariance matrices are calculated per thread block (Fig.
@@
 The final output, as shown in (\ref{eq:z_mv}) and (\ref{eq:z_mv2}), can be formed by either reducing the data down to length $L$, or increase the weight vector to length $M$.
-Since the design is using $L$ threads per output, the data is reduced to length $L$ by letting thread $r$ calculate
+Since the design is $L$ threads per output, the data is reduced to length $L$ by letting thread $r$ calculate
 \begin{align}\label{eq:subsum}
@@
 Nilsen et al.
-\cite{Nilsen2009} have proposed to apply beamspace processing to further reduce the required computations of Capon beamformer.
-With beamspace we mean transforming the channel data from element space to a fan of beams covering the sector illuminated by transmit beams.
+\cite{Nilsen2009} have proposed to apply beamspace processing to further reduce the required computations of the Capon beamformer when applied to medical ultrasound imaging.
+Beamspace is an alternative representation of the per-element data, transformed into a fan of beams covering the sector illuminated by transmit beams.
 The transformation is formally stated as
@@
 The steering vectors in $\mat{B}$, the so-called Butler matrix, determine each beam's direction in space.
-We can therefore remove beams where no interference is present by simply removing rows from $\mat{B}$.
-
-For a focused system like cardiac ultrasound imaging, where the received signal is concentrated in a narrow band around broadside, there is a small percentage of the beams that contains almost all the energy.
+Beams where no interference is present can therefore be removed by simply removing rows from $\mat{B}$.
+
+For a focused system like cardiac ultrasound imaging, where the received signal is concentrated in a narrow band around broadside, there is a small percentage of beams containing almost all received energy.
 Nilsen et al.
@@
 where $\mat{B} \in \mathbb{C}^{N_b,L}$, and $N_b$ is the number of selected beams.
-If the transformation is applied before the covariance matrix is constructed, one transformation must be performed per subarray, yielding subarrays that no longer overlap in space.
+If the transformation is applied before the covariance matrix is constructed, one transformation is performed per subarray, yielding subarrays that no longer overlap in space.
 The amount of input data is then change from $M$ to $N_LN_b$ elements per sample, however all remaining steps will get lower complexity when $N_b < L$.
@@
 \end{align}
-The pros using this approach is that minor changes has to be made to the pipeline described in Section \ref{sec:meth}, but it will only speed up the solving step if the solution $\vec{b}_{BS}$ is transformed back to element space.
+The advantages of using this approach is that minor changes have to be made to the pipeline described in Section \ref{sec:meth}, but it will only speed up the solving step if the solution $\vec{b}_{BS}$ is transformed back to element space.
 In this paper we will focus on the first approach, and how it has been implemented on the GPU.
@@
 Equation (\ref{eq:sliding_bs}) shows that instructions are spared if we utilize the overlap across subarrays, but with an increase in task parallel granularity (As discussed in Section \ref{sec:calcR}).
-The most obvious is to use one thread per beam (index $k$) that calculates this beam for all subarrays.
+The most obvious in this situation is to use one thread per beam (index $k$) that calculates this beam for all subarrays.
 However, since the number of selected beams usually is small ($N_b \ll M$) we will end up with a large memory per thread ratio.
@@
 The solution is to combine SBS with full computations.
-As shown in Fig.
-(???), a number of thread groups target different sets of subarrays which they transform using the SBS algorithm.
+A number of thread groups target different sets of subarrays which they transform using the SBS algorithm.
 In that way, the number of threads per sample is increased and thus the memory per thread ratio is decreased.
 
-As mentioned, the amount of channel data is increased from $M$ to $N_LN_b$ per sample when using SBS.
-This means that the spatial overlap exploited when calculating the sample covariance matrix in element space is no longer available.
-On the other hand, since $N_b$ should be small ($N_b < 32$), one thread per element in $\mat{\hat{R}}_{BS}$ is now doable.
+Transforming subarrays with SBS yields subarrays that no longer overlap.
+This means that the spatial overlap previously exploited when calculating the sample covariance matrix in element space is no longer available.
+On the other hand, since $N_b$ should be small ($N_b < 32$), one thread per element in $\mat{\hat{R}}_{BS}$ is now possible.
 A kernel similar to the one described in \cite{Chen2011} has therefore been implemented.
@@
 \end{align}  
-using on thread per beam, and applied to the beamspace data summed across subarrays,
+using one thread per beam, and applied to the beamspace data summed across subarrays,
 \begin{align}
@@
 \end{align}
-And again are several pixels calculated per thread block for small values of $N_b$ in order to achieve high GPU occupancy.
-The beamspace data summed across subarrays are calculated and outputted as part of the SBS kernel.
+Several pixels are calculated per thread block for small values of $N_b$ in order to achieve high GPU occupancy. The beamspace data summed across subarrays are calculated as part of the SBS kernel.
 
@@
 When data are transformed into beamspace, a high percentage of the total energy is mapped to a small subset of beams.
-Thus the average energy per beam is higher than the average energy was per element when beams containing little energy are removed.
+Thus the average energy per beam in a reduced beamspace is higher than the average energy per element.
 This means that less diagonal loading (smaller $d$) has to be applied in beamspace than in element space when $N_b < L$.
 
-To summarize, by transforming data to beamspace we reduce the complexity of all the subsequent steps.
-Calculating the covariance matrix is now $O(N_KN_LN_b^2)$, the inversion step is $O(N_b^3)$, and calculating weights and the beamformer output are both $O(N_b)$.
+To summarize, by transforming data to beamspace the complexity of all subsequent steps is reduced.
+Calculation of the covariance matrix is now $O(N_KN_LN_b^2)$, the inversion step is $O(N_b^3)$, and calculation of weights and the beamformer output are both $O(N_b)$.
 The added complexity of SBS is $O(N_bL)$.
@@
 Execution time is plotted as a function of subarray length $L$.
-If $L$ is kept constant on $M/2$ and execution time is plotted against $N_b$, the plot would be similar to the two in Fig.
+If $L$ had been kept constant on $M/2$ and execution time was plotted against $N_b$, the plot would be similar to the two in Fig.
 \ref{fig:bench}.
@@
 \end{figure*}
-In this section we present benchmarks of the implementations presented in Section \ref{sec:meth} and \ref{sec:bs} performed on a Nvidia Quadro 6000 graphics card (14 SMs, 1 TFlops theoretical single precision, 6 GB global memory, and 144 GB/s memory bandwidth).
+In this section we present benchmarks of the implementations presented in Sections \ref{sec:meth} and \ref{sec:bs} performed on a Nvidia Quadro 6000 graphics card (14 SMs, 1 TFlops theoretical single precision, 6 GB global memory, and 144 GB/s memory bandwidth).
 Problem sizes are calculated based on 32, 64, and 96 element phased arrays with a 2.5 MHz center frequency.
-Assuming a system bandwidth of 80\% we need a complex sampling frequency of 2 MHz, and when imaging down to a depth of 15 cm we end up with approximately 400 samples per range line.
+Assuming a system bandwidth of 80\% we need a complex sampling frequency of 2 MHz, and when imaging down to a depth of 15 cm we have approximately 400 samples per range line.
 To correctly sample a $70\degree$ sector with 32, 64 and 96 element arrays ($\lambda/2$ pitch), we need at least 40, 80 and 120 beams respectively.
@@
 Benchmarks are based on 32 and 64 elements phased arrays respectively, using 40 and 80 transmit beams.
-For each subarray size the total execution time is split into one bar for each of the three steps described in Section \ref{sec:meth}, and calculation of $\mat{\hat{R}}$ is presented for three different values of $K$.
+For each subarray size, total execution time is split into one bar for each of the three steps described in Section \ref{sec:meth}, and calculation of $\mat{\hat{R}}$ is presented for three different values of $K$.
 For $M=64$, $L=32$ and $K=2$ (Fig.
@@
 The simulation was performed using a 2.5 MHz, $\lambda/2$ pitch, 10 $\mu$m kerf, 50\% bandwidth and 10 mm high phased array.
-The transducer was exited with a $3\lambda$ long pulse with transmit focus at 80 mm.
+The transducer was excited with a $3\lambda$ long pulse with transmit focus at 80 mm.
 The phantom contains two circular areas, one bright at 24 dB and one dark at -30 dB, and two point targets at 40 dB.
@@
 \ref{fig:invivo} presents results of applying Capon and BS Capon on an \textit{in-vivo} image of the left ventricle.
-The image was acquired using a experimental research scanner at MI-Lab in Trondheim using a 64 elements, 3.5 MHz, and $\lambda/2$ pitched phased array in harmonic mode.
+The image was acquired using an experimental research scanner equipped with a 64 elements, 3.5 MHz, and $\lambda/2$ pitched phased array in harmonic mode.
 The image is extracted from a cardiac cycle, that can be viewed in its full length in an attached video.
@@
 To get a better impression of the high resolution images obtained with Capon beamforming, two slices taken from the images in Fig.
-\ref{fig:phantom} and \ref{fig:invivo} is included in Fig.
+\ref{fig:phantom} and \ref{fig:invivo} are included in Fig.
 \ref{fig:slices}.
@@
 \section{Trading resolution for speed}\label{sec:trade}
-As we see from Fig.
+As seen from Fig.
 \ref{fig:bench} and \ref{fig:benchBS}, execution time is reduced when either $L$ or $N_b$ is reduced.
 So to gain an  increase in processing throughput we can reduce $L$ or $N_b$.
-The question is how does this impact image quality.
+The question is then, how does this impact image quality?
 In Fig.
-\ref{fig:speed_res_trade} minimum distance between the two point targets in Fig.
-\ref{fig:phantom} (while preserving a 6dB lower level in between) is plotted against $L$ (when processed with Capon) and $N_b$ (when processed with BS Capon).
+\ref{fig:speed_res_trade}, minimum distance between the two point targets in Fig.
+\ref{fig:phantom} (while preserving a 6dB saddle point in between the two) is plotted against $L$ (when processed with Capon) and $N_b$ (when processed with BS Capon).
 For high values the two methods are equal, but for lower values, BS Capon provides higher resolution than what is possible with small subarrays.
 For this result, $d=0.01$ has been used on both methods.
-Why resolution decreases for BS Capon for small $N_b$s is partially explained by reduced information of interference at higher angles, and an automatically increase in diagonal loading.
-It is also interesting to note that Capon beamforming in this example performs worse than DAS for small subarrays.
+Why resolution decreases for BS Capon for small $N_b$s is partially explained by reduced information of interference at higher angles, and an automatically increase in diagonal loading (See end of Section \ref{sec:bs}).
+In element space, resolution decays rapidly as a result of a reduced effective adaptive aperture.
+For $N_b=3$, BS Capon provides approximately a 1 mm increase in resolution over DAS.
+However, note that resolution for the Capon beamformer highly depends on image SNR and the choice of parameters.
+Thus, different SNR scenarios and choices of parameters will give different results.
+It is also worth noting that Capon beamforming in this example performs worse than DAS for small subarrays.
 
@@
 In this paper we have presented a parallel implementation of the Capon beamformer in both element- and beam-space for high resolution cardiac ultrasound imaging.
-The results show that it is now possible to apply BS Capon in real time in cardiac ultrasound imaging utilizing the power of GPU's.
+The results show that it is now possible to apply BS Capon in real time in cardiac ultrasound imaging by utilizing the power of GPUs.
 According to Fig.
-\ref{fig:benchBS} two parallel receive lines is also within reach (for both $M=64$ and $M=96$.
-$L=M/2$), but at the moment it is unknown if these extra lines is needed for increased sampling or not.
+\ref{fig:benchBS}, two parallel receive lines are also within reach (for both $M=64$ and $M=96$.
+$L=M/2$), but at the moment it is unclear whether these extra lines are needed for increased sampling or not.
 Another concern is that lateral oversampling using parallel receive lines will increase the total execution time with a factor equal the amount of oversampling.
-
-For Capon in element space, processing rates for high quality settings is still to high to support real time cardiac ultrasound imaging.
+Hence, four parallel receive lines is not possible to process in real time using BS Capon in combination with our system configuration.
+
+For Capon in element space, processing rates for high quality settings are still to high to support real time cardiac ultrasound imaging.
 From Fig.
@@
 \ref{fig:benchBS}, inverting the three times three beamspace covariance matrix is barely visible.
-The time taken to construct all covariance matrices is also reduced, but for few beams this is now the major contributor to the total time.
+The time taken to construct all covariance matrices is also reduced, but for few beams this is now the major contributor to the total execution time.
 If real time is defined as 10 fps \cite{Chen2011} all execution times in both Fig.
@@
 More natural for cardiac ultrasound imaging is to define real-time as the time it takes to acquire the underlying image.
-With this requirement the images formed using a 32, 64 and 96 element array has a real time requirement of 125, 62.5 and 41.7 fps respectively using single line acquisition.
+With this requirement the images formed using a 32, 64 and 96 element array have a real time requirement of 125, 62.5 and 41.7 fps respectively using single line acquisition.
 Each benchmark plot has a line indicating this real time requirement.
-With beamspace, all processing times for large subarrays are below this line.
-
-This paper further shows how to obtain a parallel Capon beamformer, and how it can be implemented in a GPU framework.
+With beamspace processing, all processing times for large subarrays are below this line.
+
+This paper further shows how to obtain a parallel formulation of the Capon and BS Capon beamformer, and how to implement them in a GPU framework.
 The per-pixel granularity suggest that the beamformer is easily dividable across multiple GPUs.
-Another option to reach real-time execution for all cases presented in this paper is therefore to use several GPUs simultaneously.
-In that case, processing rates are increased with a factor equal to the number GPUs.
+Real-time execution for all cases presented in this paper is therefore possible if several GPUs are used simultaneously.
 However, there is usually a space-cost-energy limitation for how many GPUs that can be fitted in an ultrasound scanner.
@@
 Turning to image quality we see from Fig.
-\ref{fig:phantom}, and the attached video of the simulation, that Capon beamforming highly resolves moving point targets both in element- and in a reduced beam-space.
+\ref{fig:phantom}, and the attached video of the simulation, that Capon beamforming highly resolves moving point targets both in element- and in a reduced beam-space ($N_b=3$).
 Our preliminary finding is that the Capon beamformer is shift invariant as long as the lateral sampling is increased enough.
-No major improvement in contrast is visually observed in these images.
+How large this oversampling factor needs to be and how the oversampling should be conducted is left unanswered in this paper.
+No major improvement in contrast is visually observed in these images, other than slightly rounder circles due to mainlobe micro steering.
 In the cardiac recording (Fig.
-\ref{fig:invivo}), structures are thinner and better resolved both when the Capon and BS Capon beamformer is applied.
+\ref{fig:invivo}), structures are thinner and better resolved both when the Capon and BS Capon beamformer are applied.
 However we observe a loss in contrast, especially in the septum (Fig.
 \ref{fig:slices}).
-This loss can be partially explained by the present of highly correlated signals and errors in the assumed steering vector.
+This loss can be partially explained by the presence of highly correlated signals and mismatch between the assumed steering vector and signal.
 Future work should focus on solving this issue.
@@
 Nilsen et.
-al suggest to do automatic selection of beams in beamspace, hence picking the beams containing most energy.
+al suggested to do automatic selection of beams in beamspace, hence picking the beams containing most energy.
 In this paper the center beams are always picked.
-This seems to work fine, but an adaptive pick of beams should be implemented if interference at high angles is shown to contaminate the image.
+This seems to work fine, but an adaptive picking of beams should be implemented if interference at high angles is shown to contaminate the image.
 
@@
 \section{Conclusion}\label{sec:con}
-In this paper we have shown that the Capon beamformer can be accelerated using GPUs.
-For beamspace Capon beamforming processing rates supporting real time cardiac ultrasound imaging is shown to be possible.
-Image quality is investigated both in a simulated phantom and in \textit{in-vivo} cardiac data.
-Results shows that Capon beamforming is feasible for cardiac ultrasound imaging, providing images with highly resolved structures.
-
-\section{Future work}
-Study shift invariance and minimum sampling...
-
-MLA's has, to some extent been used to increase later sampling.
+In this paper a parallel formulation and implementation of the element- and beam-space Capon beamformer have been presented, and
+both beamformers have successfully been implemented on a GPU.
+For beamspace Capon beamforming processing rates supporting real time cardiac ultrasound imaging was shown to be possible, meaning that processing rates higher than the image acquisition rate are obtained for a wide range of parameters.
+Image quality was investigated both in a simulated phantom and an \textit{in-vivo} cardiac dataset.
+These results showed that Capon beamforming is feasible for cardiac ultrasound imaging, providing images with sharper structures and no local shift-variance effects.s
+
+%\section{Future work}
+%Study shift invariance and minimum sampling...
+
+%MLA's has, to some extent been used to increase later sampling.
 When using a high resolution beamformer, Nyquist tells us that the sampling frequency must be increase.
@@
 
-Plane waves...
+%Plane waves...
 
@@
 % use section* for acknowledgement
-\section*{Acknowledgment}
-
-The authors would like to thank...
+%\section*{Acknowledgment}
+
+%The authors would like to thank...
 

### 2013-02-26 HRUIonGPU/jrnl_article.tex
Incorporated comments from Jo

@@
 % can use linebreaks \\ within to get better formatting as desired
-\title{Implementing Capon Beamforming on the GPU for Real Time Cardiac Ultrasound Imaging}
+\title{Implementing Capon Beamforming on a GPU for Real-Time Cardiac Ultrasound Imaging}
 %
@@
 %\boldmath
-Capon beamforming is associated with a high computational complexity, which limits its use as a real time array processor in many applications.
-In this paper we present an implementation of the Capon beamformer exhibiting real-time performance when applied in a typical cardiac ultrasound imaging setting.
-To accomplish this task we make use of the massive parallel processing power found in modern Graphics Processing Units (GPUs), combined with beamspace processing to reduce the computational complexity as the problem size grows.
-For  beamspace Capon beamforming processing rates supporting real time cardiac ultrasound imaging is shown to be possible, meaning that processing rates higher than the image acquisition rate are obtained for a wide range of parameters.
-Image quality is investigated both in a simulated phantom and an \textit{in-vivo} cardiac dataset.
+Capon beamforming is associated with a high computational complexity, which limits its use as a real-time array processor in many applications.
+In this paper we present an implement of the Capon beamformer exhibiting real-time performance when applied in a typical cardiac ultrasound imaging setting.
+To reach this performance we make use of the parallel processing power found in modern Graphics Processing Units (GPUs), combined with beamspace processing to reduce the computational complexity as the number of array elements increases.
+For a three dimensional beamspace, processing rates supporting real-time cardiac ultrasound imaging is shown to be possible, meaning that processing rates higher than the image acquisition rate are obtained for a wide range of parameters.
+Image quality is investigated in videos of both a simulated phantom and an \textit{in-vivo} cardiac dataset.
 These results show that Capon beamforming is feasible for cardiac ultrasound imaging, providing images with sharper structures and no local shift-variance effects.
@@
 \begin{IEEEkeywords}
-Adaptive Beamforming, Capon Beamformer, Minimum Variance Beamforming, Graphics Processing Unit, GPGPU, Parallel Computing, Ultrasound Imaging.
+Adaptive Beamforming, Capon Beamforming, Minimum Variance Beamforming, Graphics Processing Unit, GPGPU, Parallel Computing, Ultrasound Imaging.
 \end{IEEEkeywords}
@@
 \IEEEPARstart{C}{apon} beamforming \cite{Capon1969} applied to medical ultrasound imaging has recently shown promising results in several studies \cite{Synnevag2007, Austeng2008, Vignon2008, Viola, Mehdizadeh2012}.
-Some interesting trade-offs in addition to just improved resolution has been introduced, like maintaining image resolution in combination with higher framerates, smaller probes, or deeper penetration \cite{Synnevag2009}.
-Despite these benefits, widespread use has been limited by the immense computational complexity associated with the method \cite{So2011}.
-In this paper we introduce an implementation of the Capon beamformer exhibiting real-time performance when applied in a typical cardiac ultrasound imaging setting.
-To accomplish this task we make use of the massive parallel processing power found in modern Graphics Processing Units (GPUs), combined with novel methods to reduce the computational complexity as the problem size grows.
-
-Previous work on parallel implementations of the Capon beamformer have mainly been focused around radar and passive sonar systems using custom made systolic array processors \cite{McWhirter1989, Moonen1993, Sinha2002}.
+Most of this work focus on the improved lateral resolution and contrast obtained with the Capon beamformer.
+However, some interesting trade-offs has also been introduced.
+In \cite{Synnevag2009}, Synnev\aa{}g et al.
+explains how image resolution can be maintained at higher framerates, with smaller probes, or for deeper penetration.
+Despite these benefits, widespread adoption of the Capon beamformer has not been seen due to its high computational complexity \cite{So2011}.
+In this paper we present an implementation of the Capon beamformer exhibiting real-time performance when applied in a typical cardiac ultrasound imaging setting.
+To reach this performance we make use of the parallel processing power found in modern Graphics Processing Units (GPUs) (found in almost all PCs), combined with novel methods to reduce the computational complexity as the number of array elements increases.
+
+Previous work on parallel implementations of the Capon beamformer have been focused on radar and passive sonar systems using custom made systolic array processors \cite{McWhirter1989, Moonen1993, Sinha2002}.
 For medical ultrasound, Chen et al.
@@
 Under this constraint the Capon beamformer is not allowed to micro-steer the main lobe, and therefore its full potential is not reached.
-Micro-steering of the main lobe is an important property of the Capon beamformer in order for it to improve structural details and edge definitions throughout an ultrasound image (Fig 9.
+Micro-steering of the main lobe is an important property of the Capon beamformer in order for it to improve structural details and edge definitions in an ultrasound image (Fig 9.
 in \cite{Synnevag2009}).
-Another aspect of using complex data is that even though the number of arithmetic operations per sample is increased we can sample at the system bandwidth, keeping the number of samples for a full cardiac imaging sector at a minimum.
-All results in this paper are reported with the use of baseband data, hence the sampling frequency is close to the system bandwidth.
+Although complex arithmetic requires more operations per sample, complex (I/Q or baseband) data can be sampled at a lower rate, thus minimizing the number of samples needed per image.
+All results in this paper are reported with the use of baseband data, hence where the sampling frequency is close to the system bandwidth.
 
@@
 % denne må utelates i IUS-artikkelen
-In recent work \cite{asen2012}(cite JOE or ECUA?) we have applied a GPU-based Capon beamformer in both an active sonar system and for cardiac ultrasound imaging.
-For sonar the real-time requirement is much lower than for medical ultrasound imaging, which means that most array configurations can be handled in real time.
-For medical ultrasound imaging, and especially cardiac ultrasound imaging, frame rate requirements increase to a level where even a high-end GPU can not do the processing required for a standard implementation of the Capon beamformer.
-Since execution time highly depends on the number of elements in the given array, it is important to reduce the computational complexity for high-channel-count systems.
-Different approaches to low-complexity Capon beamforming have therefore recently been proposed \cite{Synnevag2011, Asl2012, Jensen2012, Kim}.
-In this paper we will make use of the beamspace transformation, described by Nilsen et al.
-\cite{Nilsen2009}, in order to maintain high framerates as the channel count increases.
+In recent work \cite{asen2012}(cite JOE or ECUA?) we have applied a GPU-based Capon beamformer for both active sonar and for cardiac ultrasound imaging.
+Sonar has a real-time requirement less strict than medical ultrasound imaging, which means that most array configurations can be handled in real-time using the GPU.
+Medical ultrasound imaging, and in particular cardiac ultrasound imaging, demands frame rates at a level where even a high-end GPU can not do the processing required for a straightforward implementation of the Capon beamformer.
+Since the execution time is a cubic with respect to the number of array elements, we need methods that reduce the computational complexity for large arrays.
+Various approaches to low-complexity Capon beamforming have therefore recently been proposed in the literature \cite{Synnevag2011, Asl2012, Jensen2012, Kim}.
+In this paper we will make use of a beamspace transformation described by Nilsen et al.
+\cite{Nilsen2009}, in order to maintain high framerates as the number of elements increases.
 The beamspace transformation matches well with the focused transmit beams used in cardiac ultrasound imaging.
@@
 
-In the next section we present background information about the Capon beamformer and the GPU computing model.
-In Section \ref{sec:meth} we go in detail about the parallel implementation of the Capon beamformer, before the beamspace method and how it has been implemented on the GPU is presented in Section \ref{sec:bs}.
+In the next section we present background information on the Capon beamformer and the GPU computing model.
+In Section \ref{sec:meth} we go in detail on the parallel implementation of the Capon beamformer, before the beamspace method, and how it has been implemented on the GPU, is presented in Section \ref{sec:bs}.
 Benchmark results and resulting images are presented in Section \ref{sec:bench} and \ref{sec:images}, together with a discussion of trading image quality for speed in Section \ref{sec:trade}.
@@
 The conventional form of array beamforming in medical ultrasound imaging is delay-and-sum (DAS).
-Here, the signal recorded at element number $m \in [0,M-1]$ at time $n$ is delayed with an appropriate delay $\Delta_m[n]$ to focus and steer the ultrasound beam in space.
+Here, the signal recorded at element number $m \in [0,M-1]$ at time $n$ is delayed with an appropriate delay $\Delta_m[n]$ to focus and steer the ultrasound beam.
 This delayed data is here denoted $x_m[n]$.
-The beamformer output, $z[n]$, is then the sum of all $M$ channels weighted with a per-element factor $w_m$,
+The beamformer output, $z[n]$, is then the sum of all $M$ elements weighted with a factor $w_m$,
 \begin{align}
@@
 \end{align}
-The weight vector or window $\vec{w}$ is usually real, and is applied to trade resolution for a better signal-to-noise ratio (SNR).
-Even if different windows can be used in different areas of the images they have traditionally been predefined.
+The weight vector, or window, $\vec{w}$ is usually real, and is applied to trade resolution for a better signal-to-noise ratio (SNR).
+Although different windows can be used in different areas of the images, $\vec{w}$ is usually predefined.
 
@@
 \ref{fig:mvbf}, the basic idea of an adaptive beamformer is to form a complex weight vector based on the received data.
-For Capon beamforming the weights are found that minimize the beamformer output power while maintaining unity gain at broadside.
-The result is that interference impinging from other directions than the point of focus is suppressed \cite{Synnevag2007}.
-It is worth noting that the micro steering caused by the complex weights is minute compared to the pre-steering and -focusing step, therefore the beamformer also works well for broadband ultrasound signals.
+The Capon beamformer produces weights that minimize the output power while maintaining unity gain in the focus direction.
+The result is that interference impinging from other directions is suppressed \cite{Synnevag2007}.
+It is worth noting that the micro-steering caused by the complex weights is minute compared to the initial steering and focusing step, therefore the beamformer also works well for broadband ultrasound signals.
 
@@
 
-The Capon beamformer can be summarized in three main steps. First a sample covariance matrix has to be calculated, 
+The Capon beamformer, following the approach in \cite{Synnevag2009}, can be divided into three main steps: estimation of a sample covariance matrix, solving a system of linear equation, and calculation of the beamformer output.
+The sample covariance matrix is estimated as
 \begin{align}
@@
 \end{align}
-where  $N_K = 2K + 1$ is the number of samples included in time, $\vec{x}_l = [x_l[n], \dotso, x_{l+L-1}[n]]$ is the $l^\text{th}$ subarray of length $L$ and $N_L = M-L+1$ is the number of subarrays.
+where  $N_K = 2K + 1$ is the number of samples included in time, $\vec{x}_l = [x_l[n], \dotso, x_{l+L-1}[n]]$ is the $l^\text{th}$ subarray of length $L$, and $N_L = M-L+1$ is the number of subarrays.
 Note that the covariance matrix dimension is $L \times L$.
-The covariance matrix is then loaded with a diagonal factor $\epsilon$ to ensure numerical stability and increased robustness,
+The covariance matrix is loaded with a diagonal factor $\epsilon$ to ensure numerical stability and increased robustness,
 \begin{align}\label{eq:diag}
@@
 \end{align}
-A weighting proportional to the output power is often used to reduce the need of parameter adjustments, and the trace of $\mat{\breve{R}}$, 
+A weight proportional to the output power is often used to reduce the need for parameter adjustments, and the trace of $\mat{\breve{R}}$, 
 \begin{align}\label{eq:diag_adapt}
-\epsilon[n] &= d \times \frac{tr\{\mat{\breve{R}}[n]\}}{L}
+\epsilon[n] &= d \times \frac{\text{trace}\{\mat{\breve{R}}[n]\}}{L},
 \end{align}
@@
 
-Second, a linear system of equations has to be solved
+Next, a linear system of equations has to be solved,
 \begin{align}\label{eq:b}
@@
 \end{align}
-and the beamformer output is calculated by combining all $N_L$ subarrays weighted with the same set of adaptive weights. Formally stated as
+and the beamformer output is calculated by combining all $N_L$ subarrays weighted with the same set of adaptive weights. This is formally stated as
 \begin{align}
 z[n] &= \frac{1}{N_L}\vec{w}[n]^H \sum_{l=0}^{N_L-1} \vec{x}_l[n] \label{eq:z_mv}\\
-&= 1/N_L \, \text{CONV}_m(\vec{w}[n], \vec{1}_{N_L})^H\vec{x}[n] \label{eq:z_mv2},
+&= 1/N_L \, (\vec{w}[n] *_{m} \vec{1}_{N_L})^H\vec{x}[n] \label{eq:z_mv2},
 \end{align}
@@
 %\end{align}
-where $\text{CONV}_m$ is a non-truncated convolution across elements.
-Note how we can either choose to sum all subarrays (\ref{eq:z_mv}), or accumulate the right weighting per element (\ref{eq:z_mv2}).
+where $*_m$ is a non-truncated convolution across elements.
+Note how we can either choose to sum all subarrays (\ref{eq:z_mv}), or accumulate a weighting per element (\ref{eq:z_mv2}).
 
 In passive systems, assuming stationary statistics, the sample covariance matrix in (\ref{eq:R}) is typically averaged over a long sequence of samples\cite{Krima}.
-For an active, pulsed system like medical ultrasound imaging, the available data from which $\mat{\hat{R}}$ can be estimated is limited.
-Because of this, the resulting sample covariance matrix is often found to be both inaccurate, and poorly conditioned.
-Another challenge with applying Capon to active systems is the effect of signal cancellation, where high correlation between signal and interference favors weights that cancel out the signal \cite{Reddy1987}.
-In this way, the beamformer output can be too low at broadside, even when there is signal present and the distortionless-response criterion is applied.
-
-In order to get a well-conditioned $\mat{\hat{R}}$, avoid signal cancellation, and to get DAS-like speckle, $\mat{\hat{R}}$ has to be averaged over $L\le M/2$ long subarrays and $N_K \sim T_p/T_s$ time samples \cite{Synnevag2007, Synnevag2007a}, where, $T_p$ is the pulse length in seconds and $T_s$ is the sampling period.
-For baseband data, one pulse length ($1.5\lambda$) is around 3 samples ($K=1$).
+Since medical ultrasound imaging is an active, short-pulsed, system, there is a limited amount of data from which $\mat{\hat{R}}$ can be estimated.
+The estimated matrix is therefore often found to be both inaccurate and poorly conditioned.
+Another challenge with active systems is the high correlation between signal and interference, resulting in signal cancellation \cite{Reddy1987}.
+Thus, the beamformer output can be lower than expected, even when there is signal present and the distortionless-response criterion is applied.
+
+In order to get a well-conditioned $\mat{\hat{R}}$, avoid signal cancellation, and to get DAS-like speckle, $\mat{\hat{R}}$ has to be averaged over $L\le M/2$ long subarrays and $N_K \sim T_p/T_s$ time samples \cite{Synnevag2007a}, where, $T_p$ is the pulse length in seconds and $T_s$ is the sampling period.
+For our baseband data, one pulse length ($1.5\lambda$) is around 3 samples ($K=1$).
 
@@
 
-In this section we briefly discuss the principle behind GPU computing, laying the basis for further discussion on parallel implementation strategies.
-CUDA by Nvidia has been our choice of GPU developer environment \cite{Nvidia2011}.
-However, even though CUDA has been used to implement the Capon beamformer, the challenges and proposed methods described in this paper apply for other massive parallel implementation environments as well, like e.g.
+In this section we give a brief discussion of GPU computing, laying the basis for further discussion on parallel implementation strategies.
+As GPU framework, we have selected CUDA by Nvidia \cite{Nvidia2011}, however the challenges and proposed methods described in this paper are valid for other GPU framworks as well, like e.g.
 OpenCL.
-The key point is that GPUs can accelerate small concurrent tasks, like beamforming one pixel, much better than a CPU.
-Where the CPU has few cores and logic to process complex threads, the GPU has many cores, but each thread needs to be simple in order to maximize the benefit of using GPUs.
-
-The CUDA compute model is based on execution of a kernel function across a large set of threads, where the kernel function describe the work to be done in each thread.
+The key point is that GPUs can accelerate large sets of small concurrent tasks, like beamforming a pixel in a large image, much better than a CPU.
+While the CPU has few cores and logic to process complex threads, the GPU has many cores, but each core needs many (preferably simple) threads to keep it occupied.
+
+The CUDA compute model is based on execution of a kernel function over a large set of threads, where the kernel function describes the work to be done in each thread.
 As depicted in Fig.
-\ref{fig:gpulayout}, threads are organized in a grid of thread-blocks, with maximal size of 1024 threads per block.
-Threads inside a block supports low-cost synchronized if needed.
-From a CUDA perspective the GPU consist of one or more stream multi processors (SMs), where each SM is capable of 32 or more, depending on the architecture, concurrent multiplications and/or additions.
+\ref{fig:gpulayout}, threads are organized in a grid of thread-blocks, with maximal size of 1024 threads per block (32x32).
+Threads inside a block supports low-cost synchronization if needed.
+From a CUDA perspective the GPU consist of one or more streaming multi-processors (SMs), where each SM is capable of 32 or more concurrent multiplications and/or additions, depending on the architecture.
 A group of blocks is scheduled to each SM until all blocks have been processed.
 
-An SM has available a limited amount of near-core memory and registers.
-Hence, to achieve maximum GPU occupancy, each thread has to be as light on registers and memory as possible.
-It is also important that each thread performs enough instructions per used byte in order to hide memory latency, and that the problem can be divided into enough threads to hide instruction latency.
-Global reads and writes on the GPU, in addition to CPU-GPU transfers, have high penalty, and should be minimized as much as possible.
-All of these constraints combined pose a real challenge as the target problem has to map on to the GPU framework both on a grid, block, thread and memory consumption level in order to benefits from parallel acceleration.
-In the next section, we will describe how this mapping could be done for the Capon beamforming algorithm.
+An SM has a limited amount of near-core memory and registers.
+Hence, to achieve maximum GPU occupancy (having the maximum number of threads residing on each SM), each thread has to be as light on register and memory consumption as possible.
+It is also important that each thread performs enough instructions per transferred byte in order to hide memory latency, and that the problem can be divided into enough threads to hide instruction latency.
+Reads and writes to and from GPU DRAM, in addition to CPU-GPU transfers, are slow and should be minimized as much as possible.
+All of these constraints combined pose a real challenge as the target problem has to map well to the GPU framework on a grid, block, thread and memory consumption level in order to benefit from parallel acceleration.
+In the next section, we will describe how this mapping can be done for the Capon beamforming algorithm.
 
@@
 \ref{fig:gpulayout}, a first level of granularity is therefore to divide the image into a grid, where the output in each cell can be calculated independently.
-Further, based on memory dependencies and to hold memory consumption per thread at a minimum, one should recognise that a group of threads should process one output in the final image.
+Further, based on memory dependencies and to hold memory consumption per thread at a minimum, a group of threads should process one output in the final image.
 Continuing with the calculation of one weight vector, the algorithm can be broken down into three main steps that, to some extent, can be divided further into a set of parallel processes.
@@
 %\begin{array}{l}
-\text{1) } &\text{Calculate the sample covariance matrix } \mat{\hat{R}} \text{ as in } \text{(\ref{eq:diag})}\\
-\text{2) } &\text{Solve the linear system of equations as in (\ref{eq:b}})\\ %\vec{b} =\mat{\hat{R}}^{-1}\vec{1}_L\\
-\text{3) } &\text{Calculate } \vec{w} = \vec{b}/\vec{1}_{L}^H\vec{b} \text{ and the amplitude as in (\ref{eq:z_mv})}
+\text{1) } &\text{Calculate the sample covariance matrix } \mat{\hat{R}} \text{ as in } \text{(\ref{eq:diag})}.\\
+\text{2) } &\text{Solve the linear system of equations as in (\ref{eq:b}}).\\ %\vec{b} =\mat{\hat{R}}^{-1}\vec{1}_L\\
+\text{3) } &\text{Calculate } \vec{w} = \vec{b}/\vec{1}_{L}^H\vec{b} \text{ and the amplitude as in (\ref{eq:z_mv})}.
 %\end{array}
@@
 %\end{enumerate}
-The computational complexity of these three steps is $O(N_KN_LL^2)$, $O(L^3)$, and $O(L)$ respectively.
+The computational complexity of these three steps is $O(N_KN_LL^2)$, $O(L^3)$, and $O(L)$, respectively.
 As $M$ increases (and thereby $L$), the first two steps quickly become unmanageable, and real-time performance can only be achieved by exploring novel solutions to reduce complexity.
@@
 
-Calculation of one element of \R is from (\ref{eq:R}) found to be independent of calculation of all the other elements.
+Calculation of one element of \R is from (\ref{eq:R}) found to be independent of calculation of other elements.
 A natural level of granularity would then be to assign one thread to each element of $\mat{\hat{R}}$ \cite{Chen2011}.
 This will minimize global reads and writes needed per thread, but since we want to avoid block-to-block synchronization, the maximum subarray length is limited to 32 by this approach.
-Since most medical ultrasound arrays have more than or exact 64 elements, an implementation not limited by $L_{max}=32$ is preferable.
+Since most medical ultrasound arrays have more than 32 elements, an implementation not limited by $L_{max}=32$ is preferable.
 
@@
 %- \mat{\breve{R}}[n-K-1] + \mat{\breve{R}}[n-K].
-Both sliding across subarrays and in time can be realized simultaneously if $\mat{\hat{R}}$ averaged over subarrays is calculate from an $N_K$-average in time of the outer products $\vec{x}[n]\vec{x}[n]^H$ (cite JOE?).
+Both sliding across subarrays and in time can be realized simultaneously if $\mat{\hat{R}}$, averaged over subarrays, is calculate from an $N_K$-average in time of the outer products $\vec{x}[n]\vec{x}[n]^H$ (cite JOE?).
 However, since $N_K$ typically is small, and the GPU has a limited amount of memory available per thread, we focus on sliding across subarrays only.
@@
 \ref{fig:gpulayout}).
-Maximum supported $L$ is with this approach increased to 512, however for small $L$ high $M$ configurations, memory consumption per thread will be to high to achieve full occupancy of the GPU.
+Maximum supported $L$ is increased to the maximum number of threads per block (currently 1024) with this approach.
+However, for small $L$ high $M$ configurations, memory consumption per thread will be to high to achieve full occupancy of the GPU.
 
 Taking symmetry into account, only the upper half of $\mat{\hat{R}}$ needs to be calculated.
-However, the GPU compute threads per $\mat{\hat{R}}$ are organized in such a way that the lower triangle can be calculated with little overhead.
+However, the GPU computes $\mat{\hat{R}}$ in such a way that the lower triangle can be found with little overhead (cite JOE).
 This also means that no special solver is needed in order to solve $\mat{\hat{R}}\vec{b} = \vec{1}_L$, even if a symmetric solver could have saved some instructions.
@@
 \subsection{Solving Multiple Small Linear Systems}
-In the literature, when it comes to solving systems of linear equations on the GPU the focus has mostly been on large systems, and GPU libraries therefore often lack a comprehensive collection of batched solvers.
-The reason for this is two-fold, first the GPU needs to solve thousands of small systems in order to beat the low memory latency of the CPU, second there is a range of system dimensions that does not fit well with the GPU architecture, where only a small amount of near-core memory and registers are available.
-Because of this, it has been proven hard to optimize a manageable set of solvers that provides speedups compared with the CPU in every cases.
+The literature has mostly focused on large systems when it comes to solving systems of linear equations on the GPU.
+GPU libraries therefore often lack a comprehensive collection of batched solvers for large sets of small systems.
+The reason for this is two-fold.
+First the GPU needs to solve thousands of small systems in order to beat the low memory latency of the CPU.
+Second there is a range of system dimensions that does not fit well with the GPU architecture, where only a small amount of near-core memory and registers are available.
+Because of this, it has been proven hard to optimize a manageable set of solvers that provides speedups compared to the CPU in every case.
 It is therefore not given that solving on the GPU is faster than solving on the CPU in general.
-However, there is two arguments for doing solving on the GPU.
-First the CPU, which might already runs a lot of ultrasound related algorithms, is offloaded.
-Second the transfer of several thousand covariance matrices over the PCI-Express buss, from CPU-side to GPU-side, is spared.
-
-For the results presented in this paper, we have used a GPU implementation of batched Gauss-Jordan (GJ) elimination made by Nvidia to solve $\mat{\hat{R}}\vec{b} = \vec{1}_L$, where $\vec{b} = \mat{\hat{R}}^{-1}\vec{1}_L$, for all pixels at once.
+However, there are two arguments for solving on the GPU.
+First the CPU, which might already run a lot of ultrasound related algorithms, is offloaded.
+Second, the transfer of several thousand covariance matrices over the PCI-Express bus, from the CPU to GPU, is avoided.
+
+For the results presented in this paper, we have used a GPU implementation of batched Gauss-Jordan (GJ) elimination made by Nvidia to solve $\mat{\hat{R}}\vec{b} = \vec{1}_L$, where $\vec{b} = \mat{\hat{R}}^{-1}\vec{1}_L$, for all pixels in one kernel launch.
 The solver has been made available online for registered CUDA developers.%For smaller systems, $L < 10$, a direct solver using one thread per system is preferable.
@@
 %Another option is to let several threads cooperate to form the sum using a binary reduction sum. However, in the following benchmarks the first approach has been used. 
-Each of the $L$ threads then read one element of $\vec{b}$, scale it with the inverse sum just found to get the weight vector $\vec{w}[n]$.
-
-The final output, as shown in (\ref{eq:z_mv}) and (\ref{eq:z_mv2}), can be formed by either reducing the data down to length $L$, or increase the weight vector to length $M$.
+Each of the $L$ threads then reads one element of $\vec{b}$, and scales it with the inverse sum to get the weight vector $\vec{w}[n]$.
+
+The final output, as shown in (\ref{eq:z_mv}) and (\ref{eq:z_mv2}), can be formed by either reducing the data down to length $L$, or increasing the weight vector to length $M$.
 Since the design is $L$ threads per output, the data is reduced to length $L$ by letting thread $r$ calculate
@@
 \begin{align}
-\vec{x}_{BS}[n] &= \mat{B}\vec{x}[n], \ \ \ \mat{B} \in \mathbb{C}^{N,M}\\
+\vec{x}_\text{BS}[n] &= \mat{B}\vec{x}[n], \ \ \ \mat{B} \in \mathbb{C}^{N,M}\\
 [\mat{B}]_{p,q} &= \frac{1}{\sqrt{M}}e^{-j 2 \pi p q / M}.
@@
 
-For a focused system like cardiac ultrasound imaging, where the received signal is concentrated in a narrow band around broadside, there is a small percentage of beams containing almost all received energy.
+For a focused system like cardiac ultrasound imaging, where the received signal is concentrated in a narrow sector around the transmit direction, a small percentage of beams contains almost all received energy.
 Nilsen et al.
-concluded that as little as three beams around broadside were adequate to produce results comparable to applying Capon beamforming in element space.
-Hence, the inversion steps is found to be constant with respect to the number of array elements.
+concluded that as little as three beams around (and including) the transmit direction were adequate to produce results comparable to applying Capon beamforming in element space.
+Hence, for  $\lambda /2$ spaced arrays the inversion steps is found to be constant with respect to the number of array elements.
 
@@
 \begin{align}
-\mat{\hat{R}}_{BS} = f(\vec{x}_{BS}) = f(\mat{B}\vec{x}), 
+\mat{\hat{R}}_\text{BS} = f(\vec{x}_\text{BS}) = f(\mat{B}\vec{x}), 
 \end{align}
@@
 If the transformation is applied before the covariance matrix is constructed, one transformation is performed per subarray, yielding subarrays that no longer overlap in space.
-The amount of input data is then change from $M$ to $N_LN_b$ elements per sample, however all remaining steps will get lower complexity when $N_b < L$.
-
-Another choice is to transform the element space covariance matrix to beamspace. 
+The amount of input data are then changed from $M$ to $N_LN_b$ elements per sample, however all remaining steps will get lower computational complexity when $N_b < L$.
+
+Another option is to transform the element space covariance matrix to beamspace. 
 \begin{align}
-\mat{\hat{R}}_{BS} &= \mat{B}\mat{\hat{R}}\mat{B}^H.
+\mat{\hat{R}}_\text{BS} &= \mat{B}\mat{\hat{R}}\mat{B}^H.
 \end{align}
@@
 \begin{align}
-\vec{a}_{BS} &= \mat{B}\vec{1}_L = \sqrt{L}\vec{e}_1,
-\end{align}
-where $\vec{e}_i$ has value one in the i$^{th}$ position and zero elsewhere.
+\vec{a}_\text{BS} &= \mat{B}\vec{1}_L = \sqrt{L}\vec{e}_1,
+\end{align}
+where $\vec{e}_i$ has the value one in the i$^{th}$ position and zeroes elsewhere.
 
@@
 \begin{align}\label{eq:bs_formula}
-x_{BS,l,k} = \frac{1}{\sqrt{L}}\sum_{p=i}^{l+L-1}x_p e^{-j2\pi k(p-l)/L} , k \in [0, N_b-1].
+x_{\text{BS},l,k} = \frac{1}{\sqrt{L}}\sum_{p=i}^{l+L-1}x_p e^{-j2\pi k(p-l)/L} , k \in [0, N_b-1].
 \end{align}
@@
 \begin{align}\label{eq:sliding_bs}
-x_{BS,l,k} &= \frac{1}{\sqrt{L}}((x_{BS,l-1,k} - x_{l-1})e^{-j2\pi k/L} \nonumber \\ &+ x_{i+L-1}e^{-j2\pi k(L-1)/L}) \nonumber \\
-&= \frac{1}{\sqrt{L}}(x_{BS,l-1,k} - x_{l-1} + x_{i+L-1})e^{-j2\pi k/L}.
+x_{\text{BS},l,k} &= \frac{1}{\sqrt{L}}((x_{\text{BS},l-1,k} - x_{l-1})e^{-j2\pi k/L} \nonumber \\ &+ x_{i+L-1}e^{-j2\pi k(L-1)/L}) \nonumber \\
+&= \frac{1}{\sqrt{L}}(x_{\text{BS},l-1,k} - x_{l-1} + x_{i+L-1})e^{-j2\pi k/L}.
 \end{align}
@@
 
-Equation (\ref{eq:sliding_bs}) shows that instructions are spared if we utilize the overlap across subarrays, but with an increase in task parallel granularity (As discussed in Section \ref{sec:calcR}).
-The most obvious in this situation is to use one thread per beam (index $k$) that calculates this beam for all subarrays.
-However, since the number of selected beams usually is small ($N_b \ll M$) we will end up with a large memory per thread ratio.
+Equation (\ref{eq:sliding_bs}) shows that instructions are saved if we utilize the overlap across subarrays, but with an increase in task level granularity (as discussed in Section \ref{sec:calcR}).
+In this situation, it might be tempting to use one thread per beam (index $k$) that calculates this beam for all subarrays.
+However, since the number of selected beams usually is small ($N_b \ll M$) we will end up with a large memory per thread ratio ($M/N_b$).
 Using one thread per beam is therefore not feasible for small $N_b$ large $M$ configurations.
 The solution is to combine SBS with full computations.
-A number of thread groups target different sets of subarrays which they transform using the SBS algorithm.
+A number of thread groups is created that targets different sets of subarrays which they transform using the SBS algorithm.
 In that way, the number of threads per sample is increased and thus the memory per thread ratio is decreased.
 
-Transforming subarrays with SBS yields subarrays that no longer overlap.
+Transforming subarrays into beamspace yields subarrays that no longer overlap.
 This means that the spatial overlap previously exploited when calculating the sample covariance matrix in element space is no longer available.
-On the other hand, since $N_b$ should be small ($N_b < 32$), one thread per element in $\mat{\hat{R}}_{BS}$ is now possible.
+On the other hand, since $N_b$ should be small ($N_b < 32$), one thread per element in $\mat{\hat{R}}_\text{BS}$ is now possible.
 A kernel similar to the one described in \cite{Chen2011} has therefore been implemented.
@@
 
-After solving $\vec{b}_{BS} = \sqrt{L}\mat{\hat{R}}_{BS}^{-1}\vec{e}_1$, the beamspace weight vector is calculated as 
+After solving $\vec{b}_\text{BS} = \sqrt{L}\mat{\hat{R}}_\text{BS}^{-1}\vec{e}_1$, the beamspace weight vector is calculated as 
 \begin{align}
-\vec{w}_{BS} = \frac{\vec{b}_{BS}}{\sqrt{L}\vec{e}_1\vec{b}_{BS}} \in \mathbb{C}^{N_b}
+\vec{w}_\text{BS} = \frac{\vec{b}_\text{BS}}{\sqrt{L}\vec{e}_1\vec{b}_\text{BS}} \in \mathbb{C}^{N_b}
 \end{align}  
@@
 \begin{align}
-z[n] = \vec{w}_{BS}^H\sum_{l=0}^{L-1}\vec{x}_{BS,l}. 
+z[n] = \vec{w}_\text{BS}^H\sum_{l=0}^{L-1}\vec{x}_{\text{BS},l}. 
 \end{align}
@@
 
-As presented in (\ref{eq:diag}), adaptive diagonal loading is performed by loading the sample covariance matrix with the average energy per element, hence $trace\{\mat{\hat{R}}\}/L$.
+As presented in (\ref{eq:diag}), adaptive diagonal loading is performed by loading the sample covariance matrix with the average energy per element, hence $\text{trace}\{\mat{\hat{R}}\}/L$.
 When data are transformed into beamspace, a high percentage of the total energy is mapped to a small subset of beams.
-Thus the average energy per beam in a reduced beamspace is higher than the average energy per element.
-This means that less diagonal loading (smaller $d$) has to be applied in beamspace than in element space when $N_b < L$.
+Thus the average energy per beam in a reduced beamspace is, for a focused system, higher than the average energy per element.
+This means that less diagonal loading (smaller $d$) has to be applied when the covariance is estimated from beams instead of elements.
 
@@
 \label{fig:benchUS64}}}
-\caption{Benchmark of GPU Capon beamforming for a cardiac image covering a $70\degree$ wide $15$cm deep sector.
+\caption{Benchmarks of GPU Capon beamforming for a cardiac image covering a $70\degree$ wide $15$ cm deep sector.
 Execution time is plotted as a function of subarray length $L$.
@@
 \label{fig:benchBS96}}}
-\caption{Benchmark of GPU beamspace Capon beamforming for a cardiac image covering a $70 \degree$ and $15$cm deep sector.
+\caption{Benchmarks of GPU beamspace Capon beamforming for a cardiac image covering a $70 \degree$ and $15$ cm deep sector.
 Execution time is plotted as a function of subarray length $L$.
@@
 To correctly sample a $70\degree$ sector with 32, 64 and 96 element arrays ($\lambda/2$ pitch), we need at least 40, 80 and 120 beams respectively.
-The presented execution times do not include transfer of data from CPU-side to the GPU, since data are assumed to be pre-delayed and available in GPU memory.
+The presented execution times do not include transfer of data from CPU- to the GPU-side, since data are assumed to be pre-delayed and available in GPU memory.
 What is included are all calculations, memory transfers from global to near-core GPU memory, and near core memory access.
@@
 Benchmarks are based on 32 and 64 elements phased arrays respectively, using 40 and 80 transmit beams.
-For each subarray size, total execution time is split into one bar for each of the three steps described in Section \ref{sec:meth}, and calculation of $\mat{\hat{R}}$ is presented for three different values of $K$.
+For each subarray size, the total execution time is split into one bar for each of the three steps described in Section \ref{sec:meth}, and the calculation of $\mat{\hat{R}}$ is presented for three different values of $K$.
 For $M=64$, $L=32$ and $K=2$ (Fig.
@@
 For all four benchmarks there is a line indicating the time it takes to acquire the underlying image using single line acquisition.
-Hence, below this line we are capable of real time processing when a single receive line is used per transmit line.
+Hence, below this line we are capable of performing real-time processing when a single receive line is used per transmit line.
 
@@
 \ref{fig:phantom}).
-The simulation was performed using a 2.5 MHz, $\lambda/2$ pitch, 10 $\mu$m kerf, 50\% bandwidth and 10 mm high phased array.
-The transducer was excited with a $3\lambda$ long pulse with transmit focus at 80 mm.
+The simulation was performed using a $\lambda/2$ pitch, 10 $\mu$m kerf, 50\% bandwidth and 10 mm high phased array.
+The transducer was excited with a 2.5 MHz and $3\lambda$ long pulse with transmit focus at 80 mm.
 The phantom contains two circular areas, one bright at 24 dB and one dark at -30 dB, and two point targets at 40 dB.
@@
 \ref{fig:invivo} presents results of applying Capon and BS Capon on an \textit{in-vivo} image of the left ventricle.
-The image was acquired using an experimental research scanner equipped with a 64 elements, 3.5 MHz, and $\lambda/2$ pitched phased array in harmonic mode.
+The image was acquired using an experimental research scanner equipped with a 64 elements and $0.65\lambda$ pitched phased array in harmonic mode (1.7 MHz at transmit and 3.5 MHz at receive).
 The image is extracted from a cardiac cycle, that can be viewed in its full length in an attached video.
-Its is included to show that Capon beamforming in both element- and beam-space is feasible for cardiac imaging, not just in simulations.
-
-To get a better impression of the high resolution images obtained with Capon beamforming, two slices taken from the images in Fig.
+It is included to show that Capon beamforming in both element- and beam-space is feasible for cardiac imaging, not just in simulations.
+
+To get a better impression of the improved lateral resolution obtained with Capon beamforming, two slices taken from the images in Fig.
 \ref{fig:phantom} and \ref{fig:invivo} are included in Fig.
@@
 
-Both data sets have been processed with the following Capon beamformer settings: $L = 32$, $K = 2$, $d=0.1$.
+Both data sets have been processed with the following Capon beamformer settings: $L = 32$, $K = 2$, and $d=0.1$.
 These settings have previously been shown to provide high resolution while maintaining DAS-like speckle\cite{Synnevag2007a}.
-For the beamspace processing the following parameters have been used: $L=32$, $K=2$, $d=0.01$, $N_b=3$.
-For the simulated phantom, 16x lateral over-sampling has been used, in order to avoid under sampling of the two bright point scatterers as they move.
-The cardiac recording only has $30\%$ lateral over sampling.
+For the beamspace processing the following parameters have been used: $L=32$, $K=2$, $d=0.01$, and $N_b=3$.
+For the simulated phantom, 16x lateral oversampling on transmit has been used, in order to avoid under sampling of the two bright point scatterers as they move.
+The cardiac recording only has $30\%$ lateral oversampling on receive.
 
@@
 In Fig.
-\ref{fig:speed_res_trade}, minimum distance between the two point targets in Fig.
-\ref{fig:phantom} (while preserving a 6dB saddle point in between the two) is plotted against $L$ (when processed with Capon) and $N_b$ (when processed with BS Capon).
+\ref{fig:speed_res_trade}, the minimum distance between the two point targets in Fig.
+\ref{fig:phantom} (while preserving a 6 dB saddle point in between the two) is plotted against $L$ (when processed with Capon) and $N_b$ (when processed with BS Capon).
 For high values the two methods are equal, but for lower values, BS Capon provides higher resolution than what is possible with small subarrays.
@@
 In element space, resolution decays rapidly as a result of a reduced effective adaptive aperture.
-For $N_b=3$, BS Capon provides approximately a 1 mm increase in resolution over DAS.
+For $N_b=3$, BS Capon provides approximately 1 mm improved resolution over DAS.
 However, note that resolution for the Capon beamformer highly depends on image SNR and the choice of parameters.
@@
 In this paper we have presented a parallel implementation of the Capon beamformer in both element- and beam-space for high resolution cardiac ultrasound imaging.
-The results show that it is now possible to apply BS Capon in real time in cardiac ultrasound imaging by utilizing the power of GPUs.
+The results show that it is now possible to apply BS Capon in real-time in cardiac ultrasound imaging by utilizing the power of GPUs.
 According to Fig.
-\ref{fig:benchBS}, two parallel receive lines are also within reach (for both $M=64$ and $M=96$.
-$L=M/2$), but at the moment it is unclear whether these extra lines are needed for increased sampling or not.
-Another concern is that lateral oversampling using parallel receive lines will increase the total execution time with a factor equal the amount of oversampling.
-Hence, four parallel receive lines is not possible to process in real time using BS Capon in combination with our system configuration.
-
-For Capon in element space, processing rates for high quality settings are still to high to support real time cardiac ultrasound imaging.
+\ref{fig:benchBS} it is also possible to process two parallel receive lines per transmit (for both $M=64$ and $M=96$.
+$L=M/2$).
+However, these extra lines might be needed for increased sampling instead of increased frame rate when using the Capon beamformers.
+If parallel receive lines are used to obtain denser sampling, acquisition time remains the same but processing times are increased with a factor equal to the amount of oversampling.
+If parallel receive lines are used for increased frame rate, the acquisition time is lowered, and the processing time remains constant.
+Hence, four parallel receive lines are not possible to process in real-time using BS Capon in combination with our system configuration.
+
+For Capon in element space, processing rates for high quality settings are still to high too support real-time cardiac ultrasound imaging.
 From Fig.
@@
 A major part of the total execution time for element space Capon comes from solving the system of linear equations.
-This limitation is targeted by reducing the sample covariance matrix size in beamspace.
-As shown in Fig.
-\ref{fig:benchBS}, inverting the three times three beamspace covariance matrix is barely visible.
-The time taken to construct all covariance matrices is also reduced, but for few beams this is now the major contributor to the total execution time.
-If real time is defined as 10 fps \cite{Chen2011} all execution times in both Fig.
-\ref{fig:bench} and \ref{fig:benchBS} are real time.
+This limitation is targeted by reducing the sample covariance matrix size in beamspace, and as shown in Fig.
+\ref{fig:benchBS}, inverting the 3x3 beamspace covariance matrix is then barely visible.
+The time it takes to construct all covariance matrices is also reduced, but for few beams this is now the major contributor to the total execution time.
+If real-time is defined as 10 fps \cite{Chen2011} all execution times in both Fig.
+\ref{fig:bench} and \ref{fig:benchBS} are real-time.
 More natural for cardiac ultrasound imaging is to define real-time as the time it takes to acquire the underlying image.
-With this requirement the images formed using a 32, 64 and 96 element array have a real time requirement of 125, 62.5 and 41.7 fps respectively using single line acquisition.
-Each benchmark plot has a line indicating this real time requirement.
+With this requirement the images formed using a 32, 64 and 96 element array have a real-time requirement of 125, 62.5 and 41.7 fps respectively using single line acquisition.
+Each benchmark plot has a line indicating this real-time requirement.
 With beamspace processing, all processing times for large subarrays are below this line.
@@
 Turning to image quality we see from Fig.
-\ref{fig:phantom}, and the attached video of the simulation, that Capon beamforming highly resolves moving point targets both in element- and in a reduced beam-space ($N_b=3$).
-Our preliminary finding is that the Capon beamformer is shift invariant as long as the lateral sampling is increased enough.
+\ref{fig:phantom}, and the attached video of the simulation, that Capon beamforming provides improved lateral resolution of moving point targets both in element- and in a reduced beam-space ($N_b=3$).
+Our preliminary finding is that the Capon beamformer is shift invariant (see video of the simulation) as long as the lateral sampling is increased enough.
 How large this oversampling factor needs to be and how the oversampling should be conducted is left unanswered in this paper.
@@
 In the cardiac recording (Fig.
-\ref{fig:invivo}), structures are thinner and better resolved both when the Capon and BS Capon beamformer are applied.
+\ref{fig:invivo}), structures are thinner and better resolved when the Capon and BS Capon beamformer are applied.
 However we observe a loss in contrast, especially in the septum (Fig.
@@
 In this paper the center beams are always picked.
-This seems to work fine, but an adaptive picking of beams should be implemented if interference at high angles is shown to contaminate the image.
+This seems to work fine, but an adaptive selection of beams should be implemented if interference at high angles is shown to contaminate the image.
 
@@
 %For a typical cardiac ultrasound image of 80*400 pixels (70 degree sector, 15cm range) acquired using a 2.5MHz, $M=64$ element phased array, the result is 10fps (subarray length $L=M/2$, temporal smoothing over 3 samples).
-If we do a 2-element pre-beamforming to get the channel count down to 32, the frame rate increases to 44fps.
+If we do a 2-element pre-beamforming to get the channel count down to 32, the
+ rate increases to 44fps.
 For a 32 element phased array we need less beams to cover the sector ($40 \times 400$ pixels), hence with the same parameters the frame rate increases to 87fps.
@@
 both beamformers have successfully been implemented on a GPU.
-For beamspace Capon beamforming processing rates supporting real time cardiac ultrasound imaging was shown to be possible, meaning that processing rates higher than the image acquisition rate are obtained for a wide range of parameters.
+For beamspace Capon beamforming processing rates supporting real-time cardiac ultrasound imaging was shown to be possible, meaning that processing rates higher than the image acquisition rate are obtained for a wide range of parameters.
 Image quality was investigated both in a simulated phantom and an \textit{in-vivo} cardiac dataset.
-These results showed that Capon beamforming is feasible for cardiac ultrasound imaging, providing images with sharper structures and no local shift-variance effects.s
+These results showed that Capon beamforming is feasible for cardiac ultrasound imaging, providing images with sharper structures and no local shift-variance effects.
 

### 2013-02-27 HRUIonGPU/jrnl_article.tex
Done some writing

@@
 However, since $N_K$ typically is small, and the GPU has a limited amount of memory available per thread, we focus on sliding across subarrays only.
-In that case, the complexity of calculating $\mat{\hat{R}}$ is reduced to $O(N_KL^2)$.
+In that case, the complexity of calculating $\mat{\hat{R}}$ is reduced to $O(N_KL^2)$ (if $L=M/2$).
 Note that sliding will break independence along the sliding dimension, thus, calculations are still independent across time, but not along diagonals of $\mat{\hat{R}}$.
@@
 \subsection{Solving Multiple Small Linear Systems}
-The literature has mostly focused on large systems when it comes to solving systems of linear equations on the GPU.
+The literature has mostly focused on large systems when solving systems of linear equations on the GPU.
 GPU libraries therefore often lack a comprehensive collection of batched solvers for large sets of small systems.
@@
 
-For the results presented in this paper, we have used a GPU implementation of batched Gauss-Jordan (GJ) elimination made by Nvidia to solve $\mat{\hat{R}}\vec{b} = \vec{1}_L$, where $\vec{b} = \mat{\hat{R}}^{-1}\vec{1}_L$, for all pixels in one kernel launch.
-The solver has been made available online for registered CUDA developers.%For smaller systems, $L < 10$, a direct solver using one thread per system is preferable.
+For the results presented in this paper, we have used a GPU implementation of batched Gauss-Jordan elimination made by Nvidia to solve $\mat{\hat{R}}\vec{b} = \vec{1}_L$, where $\vec{b} = \mat{\hat{R}}^{-1}\vec{1}_L$, for all pixels in one kernel launch.
+The solver is available online for registered CUDA developers.%For smaller systems, $L < 10$, a direct solver using one thread per system is preferable.
 For this purpose we have implemented Cramer's rule for $2 \times 2$, $3 \times 3$ and $4 \times 4$ matrices.
@@
 To calculate the beamformer output, equation (\ref{eq:z_mv}) has been implemented using one thread per element-weight ($L$ threads per weight vector).
-After the solution vector $\vec{b}$ has been placed in shared memory, one of the $L$ threads finds the sum of $\vec{b}$.
+After the solution vector $\vec{b}$ has been placed in near-core memory, one of the $L$ threads finds the sum of $\vec{b}$.
 %Another option is to let several threads cooperate to form the sum using a binary reduction sum. However, in the following benchmarks the first approach has been used. 
@@
 \end{align}
-Each thread now computes $z_r = w_r^*x_r$, and finally one thread computes and outputs $z = \frac{1}{N_L}\sum_{r=0}^{L-1} z_r$ to global memory.
+Each thread now computes $z_r = w_r^*x_r$, and finally one thread computes and outputs $z = \frac{1}{N_L}\sum_{r=0}^{L-1} z_r$ to  GPU DRAM.
 %The last sum can also be performed using a binary reduction sum across several threads for increased speed.
@@
 concluded that as little as three beams around (and including) the transmit direction were adequate to produce results comparable to applying Capon beamforming in element space.
-Hence, for  $\lambda /2$ spaced arrays the inversion steps is found to be constant with respect to the number of array elements.
+Hence, for $\lambda$-factor spaced arrays the inversion steps is found to be constant with respect to the number of array elements.
 
@@
 \end{align}
-The advantages of using this approach is that minor changes have to be made to the pipeline described in Section \ref{sec:meth}, but it will only speed up the solving step if the solution $\vec{b}_{BS}$ is transformed back to element space.
+The advantages of this approach is that minor changes have to be made to the pipeline described in Section \ref{sec:meth}, but it will only speed up the solving step if the solution $\vec{b}_{BS}$ is transformed back to element space.
 In this paper we will focus on the first approach, and how it has been implemented on the GPU.
@@
 
-Equation (\ref{eq:sliding_bs}) shows that instructions are saved if we utilize the overlap across subarrays, but with an increase in task level granularity (as discussed in Section \ref{sec:calcR}).
+Equation (\ref{eq:sliding_bs}) shows that instructions are saved if we utilize the overlap between subarrays, but with an increase in task level granularity (as discussed in Section \ref{sec:calcR}).
 In this situation, it might be tempting to use one thread per beam (index $k$) that calculates this beam for all subarrays.
@@
 This means that the spatial overlap previously exploited when calculating the sample covariance matrix in element space is no longer available.
-On the other hand, since $N_b$ should be small ($N_b < 32$), one thread per element in $\mat{\hat{R}}_\text{BS}$ is now possible.
+On the other hand, since $N_b$ should be small ($N_b < 32$), one thread per element in $\mat{\hat{R}}_\text{BS}$ is now an option.
 A kernel similar to the one described in \cite{Chen2011} has therefore been implemented.

### 2013-02-28 HRUIonGPU/gfx/all_plots.py
Added plot-function

@@ -88,9 +88,9 @@ def barBenchmarkGPU(filename, M, Nx, Ny, fromL, toL):
    min_L = fromL
    max_L = toL
 
-   yavg1 = timings[0::3,:]
-   yavg2 = timings[1::3,:]
-   yavg3 = timings[2::3,:]
+   yavg1 = timings[0::3,[2, 6, 8]]
+   yavg2 = timings[1::3,[2, 6, 8]]
+   yavg3 = timings[2::3,[2, 6, 8]]
    yavg1_sum = yavg1.sum(1)
    yavg2_sum = yavg2.sum(1)
    yavg3_sum = yavg3.sum(1)
@@ -157,22 +157,25 @@ def barBeamspaceBenchmarkGPU(filename, M, Nb, Nx, Ny, fromL, toL):
    min_L = fromL
    max_L = toL
 
-   yavg1 = timings[0::3,[1, 2, 6, 8]]
-   yavg2 = timings[1::3,[1, 2, 6, 8]]
-   yavg3 = timings[2::3,[1, 2, 6, 8]]
-   yavg1_sum = yavg1.sum(1)
-   yavg2_sum = yavg2.sum(1)
-   yavg3_sum = yavg3.sum(1)
+   #yavg1 = timings[0::3,[1, 2, 6, 8]]
+   #yavg2 = timings[1::3,[1, 2, 6, 8]]
+   #yavg3 = timings[2::3,[1, 2, 6, 8]]
+   #yavg1_sum = yavg1.sum(1)
+   #yavg2_sum = yavg2.sum(1)
+   #yavg3_sum = yavg3.sum(1)
 
-   print 'Yavg1 ', yavg1.shape
-   print 'Yavg2 ', yavg2.shape
-   print 'Yavg3 ', yavg3.shape
+   #print 'Yavg1 ', yavg1.shape
+   #print 'Yavg2 ', yavg2.shape
+   #print 'Yavg3 ', yavg3.shape
 
    width = 1
    x_range = arange(min_L,max_L+1)
+   print 'x_range ', x_range.shape
    
    K = 2
-   yavg = yavg3
+   yavg = timings[:,[1,2,6,8]]#yavg1#yavg3
+   
+   print 'Yavg ', yavg.shape
    
    fig = figure()
    ax = fig.add_subplot(1,1,1)
@@ -221,3 +224,55 @@ def barBeamspaceBenchmarkGPU(filename, M, Nb, Nx, Ny, fromL, toL):
       ylim((0, yavg.max()+1))
    
[…]

### 2013-03-03 HRUIonGPU/jrnl_article.tex

@@
 
-\begin{figure*}[!t]
-\centerline{\subfloat[]{\includegraphics[width=3.5in]{gfx/benchmark_bar_32_40_400.png} \label{fig:benchUS32}}
-\hfil
-\subfloat[]{\includegraphics[width=3.5in]{gfx/benchmark_bar_64_80_400.png}%
-\label{fig:benchUS64}}}
-\caption{Benchmarks of GPU Capon beamforming for a cardiac image covering a $70\degree$ wide $15$ cm deep sector.
+\begin{figure}
+%\centerline{\subfloat[]{\includegraphics[width=3.5in]{gfx/benchmark_bar_32_40_400.png} \label{fig:benchUS32}}
+%\hfil
+%\subfloat[]{\includegraphics[width=3.5in]{gfx/benchmark_bar_64_80_606.png}%
+%\label{fig:benchUS64}}}
+\centerline{\includegraphics[width=3.5in]{gfx/benchmark_bar_64_80_606.png}}
+\caption{Benchmark of GPU Capon beamforming for a cardiac image covering a $70\degree$ wide $15$ cm deep sector.
 Execution time is plotted as a function of subarray length $L$.
 The plot shows execution times for the three steps listed in Section \ref{sec:meth}.
-Calculating covariance matrices with three different choices of temporal averaging, solver, and the final beamformer sum.
-For larger arrays, more transmit beams have to be emitted for sufficient lateral sampling.
-(a)  32 element array.
-(b)  64 element array.}
+Calculating covariance matrices with three different choices of temporal averaging, solver, and the final beamformer sum.}
 \label{fig:bench}
-\end{figure*}
+\end{figure}
 
 \begin{figure*}[!t]
-\centerline{\subfloat[]{\includegraphics[width=3.5in]{gfx/benchmark_bar_bs_M=64_Nx=80_Ny=400_Nb=3.png} \label{fig:benchBS64}}
+\centerline{\subfloat[]{\includegraphics[width=3.5in]{gfx/benchmark_bar_bs_M=64_Nx=80_Ny=606_Nb=3.png} \label{fig:benchBS64}}
 \hfil
-\subfloat[]{\includegraphics[width=3.5in]{gfx/benchmark_bar_bs_M=96_Nx=120_Ny=400_Nb=3.png}%
+\subfloat[]{\includegraphics[width=3.5in]{gfx/benchmark_bar_bs_M=96_Nx=120_Ny=606_Nb=3.png}%
 \label{fig:benchBS96}}}
@@
 Execution time is plotted as a function of subarray length $L$.
-If $L$ had been kept constant on $M/2$ and execution time was plotted against $N_b$, the plot would be similar to the two in Fig.
+If $L$ had been kept constant on $M/2$ and execution time was plotted against $N_b$, the plot would be similar to the one in Fig.
 \ref{fig:bench}.
-(a) 64 elements array.
-(b) 96 elements array.}
+The variation in execution time is caused by adaptive kernel launch configurations.}
 \label{fig:benchBS}
@@
 \ref{fig:phantom} (while preserving a 6 dB saddle point in between the two) is plotted against $L$ (when processed with Capon) and $N_b$ (when processed with BS Capon).
-For high values the two methods are equal, but for lower values, BS Capon provides higher resolution than what is possible with small subarrays.
-For this result, $d=0.01$ has been used on both methods.
-Why resolution decreases for BS Capon for small $N_b$s is partially explained by reduced information of interference at higher angles, and an automatically increase in diagonal loading (See end of Section \ref{sec:bs}).
+For the result in Fig.
+\ref{fig:speed_res_trade}, $d=0.01$ has been used on both methods.
+
+For high $L$ and $N_b$ values the two methods are equal.
+However, for lower values, BS Capon provides higher resolution than what is possible with small subarrays.
+We see that the resolution for BS Capon gradually decreases due to an automatically increase in diagonal loading (See end of Section \ref{sec:bs}).
+If this is compensated for, BS Capon will provide resolution almost equal to Capon for all values except $N_b=1$ \cite{Nilsen2009}.
+For $N_b=1$ the resolution is worse than for DAS because of a triangular apodization caused by subarray averaging.
 In element space, resolution decays rapidly as a result of a reduced effective adaptive aperture.
-For $N_b=3$, BS Capon provides approximately 1 mm improved resolution over DAS.
+For $N_b=3$, BS Capon still provides approximately 1 mm improved resolution over DAS.
 However, note that resolution for the Capon beamformer highly depends on image SNR and the choice of parameters.
@@
 It is also worth noting that Capon beamforming in this example performs worse than DAS for small subarrays.
+For small values of $L$, there are not enough degrees of freedom in order steer zeros at each point target.
 

### 2013-03-04 HRUIonGPU/gfx/make_plots.py
Article after comments from Andreas, Jo Inge and Kjell has been integrated. Added some plot commands

@@ -12,10 +12,10 @@ from all_plots import barBeamspaceBenchmarkGPU2 as barBS2
 #barBS('timer_values_M=64_Nx=80_Ny=400_L=2-32_Nb=3_slidingBS.txt', 64, 3, 80, 400, 3, 32)
 #barBS('timer_values_M=96_Nx=120_Ny=400_L=3-48_Nb=3_slidingBS.txt', 96, 3, 120, 400, 3, 48)
 
-bar('timer_values_M=64_Nx=80_Ny=606_L=2-32.txt', 64, 80, 606, 2, 32)
+bar('timer_values_M=64_Nx=80_Ny=540_L=2-32.txt', 64, 80, 540, 2, 32)
 
-#barBS('timer_values_M=64_Nx=80_Ny=606_L=16-48_Nb=3.txt', 64, 3, 80, 606, 16, 47)
-#barBS('timer_values_M=96_Nx=120_Ny=606_L=24-72_Nb=3.txt', 96, 3, 120, 606, 24, 71)
+barBS('timer_values_M=64_Nx=80_Ny=540_L=16-48_Nb=3.txt', 64, 3, 80, 540, 16, 47)
+barBS('timer_values_M=96_Nx=120_Ny=540_L=24-72_Nb=3.txt', 96, 3, 120, 540, 24, 71)
 
-barBS2('timer_values_M=64_Nx=80_Ny=606_L=32_Nb=1-5.txt', 64, 32, 80, 606, 1, 5, False)
-barBS2('timer_values_M=96_Nx=120_Ny=606_L=48_Nb=1-5.txt', 96, 48, 120, 606, 1, 5)
+#barBS2('timer_values_M=64_Nx=80_Ny=606_L=32_Nb=1-5.txt', 64, 32, 80, 606, 1, 5, False)
+#barBS2('timer_values_M=96_Nx=120_Ny=606_L=48_Nb=1-5.txt', 96, 48, 120, 606, 1, 5)

### 2013-03-06 HRUIonGPU/jrnl_article.tex

@@
 
-%\documentclass[journal]{IEEEtran}
+\documentclass[journal]{IEEEtran}
 
 % Single column review mode:
-\documentclass[12pt,journal,onecolumn]{IEEEtran}
-\newcommand\CLASSINPUTbaselinestretch{2}
-\RequirePackage{calc}
-\RequirePackage{fp} 
+%\documentclass[12pt,journal,onecolumn]{IEEEtran}
+%\newcommand\CLASSINPUTbaselinestretch{2}
+%\RequirePackage{calc}
+%\RequirePackage{fp} 
 % Left margin = 1 inch + hoffset + oddsidemargin (or evensidemargin)
 % Adding the 2cm gutter width to the odd/even side margins:
-\setlength\hoffset{0pt}
-\setlength\oddsidemargin{4cm}       % 4cm margins on the left side
-\addtolength\oddsidemargin{-1in}    % Subtract the initial 1 inch
-\setlength\textwidth{21cm-8cm}    % A4 width (21cm) minus margins on either side
+%\setlength\hoffset{0pt}
+%\setlength\oddsidemargin{4cm}       % 4cm margins on the left side
+%\addtolength\oddsidemargin{-1in}    % Subtract the initial 1 inch
+%\setlength\textwidth{21cm-8cm}    % A4 width (21cm) minus margins on either side
 
@@
 Image quality is investigated in videos of both a simulated phantom and an \textit{in-vivo} cardiac dataset.
-These results show that Capon beamforming is feasible for cardiac ultrasound imaging, providing images with sharper structures and no local shift-variance effects.
+These results show that Capon beamforming is feasible for cardiac ultrasound imaging, providing images with sharper structures and no local shift-variance effects, both in element- and beam-space.
 \end{abstract}
@@
 % denne må utelates i IUS-artikkelen
-In recent work \cite{asen2012}(cite JOE or ECUA?) we have applied a GPU-based Capon beamformer for both active sonar and for cardiac ultrasound imaging.
+In recent work \cite{asen2012, Buskenes} we have applied a GPU-based Capon beamformer for both active sonar and for cardiac ultrasound imaging.
 Sonar has a real-time requirement less strict than medical ultrasound imaging, which means that most array configurations can be handled in real-time using the GPU.
@@
 \begin{figure}
-\centerline{\includegraphics[width=3in]{gfx/gpu_layout_vertical_lr.png}}
+\centerline{\includegraphics[width=3.4in]{gfx/gpu_layout_vertical_lr.png}}
 \caption{Calculation of Capon adaptive weights mapped to GPU architecture.
@@
 %- \mat{\breve{R}}[n-K-1] + \mat{\breve{R}}[n-K].
-Both sliding across subarrays and in time can be realized simultaneously if $\mat{\hat{R}}$, averaged over subarrays, is calculate from an $N_K$-average in time of the outer products $\vec{x}[n]\vec{x}[n]^H$ (cite JOE?).
+Both sliding across subarrays and in time can be realized simultaneously if $\mat{\hat{R}}$, averaged over subarrays, is calculate from an $N_K$-average in time of the outer products $\vec{x}[n]\vec{x}[n]^H$ \cite{Buskenes}.
 However, since $N_K$ typically is small, and the GPU has a limited amount of memory available per thread, we focus on sliding across subarrays only.
@@
 Taking symmetry into account, only the upper half of $\mat{\hat{R}}$ needs to be calculated.
-However, the GPU computes $\mat{\hat{R}}$ in such a way that the lower triangle can be found with little overhead (cite JOE).
+However, the GPU computes $\mat{\hat{R}}$ in such a way that the lower triangle can be found with little overhead \cite{Buskenes}.
 This also means that no special solver is needed in order to solve $\mat{\hat{R}}\vec{b} = \vec{1}_L$, even if a symmetric solver could have saved some instructions.
@@
 %\label{fig:benchUS64}}}
-\centerline{\includegraphics[width=3.5in]{gfx/benchmark_bar_64_80_606.png}}
+\centerline{\includegraphics[width=3.5in]{gfx/benchmark_bar_64_80_540.png}}
 \caption{Benchmark of GPU Capon beamforming for a cardiac image covering a $70\degree$ wide $15$ cm deep sector.
@@
 \begin{figure*}[!t]
-\centerline{\subfloat[]{\includegraphics[width=3.5in]{gfx/benchmark_bar_bs_M=64_Nx=80_Ny=606_Nb=3.png} \label{fig:benchBS64}}
+\centerline{\subfloat[]{\includegraphics[width=3.5in]{gfx/benchmark_bar_bs_M=64_Nx=80_Ny=540_Nb=3.png} \label{fig:benchBS64}}
 \hfil
-\subfloat[]{\includegraphics[width=3.5in]{gfx/benchmark_bar_bs_M=96_Nx=120_Ny=606_Nb=3.png}%
+\subfloat[]{\includegraphics[width=3.5in]{gfx/benchmark_bar_bs_M=96_Nx=120_Ny=540_Nb=3.png}%
 \label{fig:benchBS96}}}
@@
 In this section we present benchmarks of the implementations presented in Sections \ref{sec:meth} and \ref{sec:bs} performed on a Nvidia Quadro 6000 graphics card (14 SMs, 1 TFlops theoretical single precision, 6 GB global memory, and 144 GB/s memory bandwidth).
-Problem sizes are calculated based on 32, 64, and 96 element phased arrays with a 2.5 MHz center frequency.
-Assuming a system bandwidth of 80\% we need a complex sampling frequency of 2 MHz, and when imaging down to a depth of 15 cm we have approximately 400 samples per range line.
-To correctly sample a $70\degree$ sector with 32, 64 and 96 element arrays ($\lambda/2$ pitch), we need at least 40, 80 and 120 beams respectively.
+Problem sizes are calculated based on 64 and 96 element phased arrays with a 3.4 MHz center frequency.
+Assuming a system bandwidth of 80\% we need a complex sampling frequency of 2.7 MHz, and when imaging down to a depth of 15.4 cm we have approximately 540 samples per range line.
+To correctly sample a $70\degree$ sector with 64 and 96 element arrays ($\lambda/2$ pitch), we need at least 80 and 120 receive beams respectively.
 The presented execution times do not include transfer of data from CPU- to the GPU-side, since data are assumed to be pre-delayed and available in GPU memory.
@@
 \ref{fig:bench}.
-Benchmarks are based on 32 and 64 elements phased arrays respectively, using 40 and 80 transmit beams.
+The benchmark is based on a 64 elements phased array, using 80 receive beams.
 For each subarray size, the total execution time is split into one bar for each of the three steps described in Section \ref{sec:meth}, and the calculation of $\mat{\hat{R}}$ is presented for three different values of $K$.
 For $M=64$, $L=32$ and $K=2$ (Fig.
-\ref{fig:benchUS64}), a typical setup for high quality cardiac imaging, processing throughput is 10 frames per second.
+\ref{fig:bench}), a typical setup for high quality cardiac imaging, processing throughput is 7 frames per second.
 
@@
 The figure only present results with use of temporal averaging ($K=2$).
-For $M=96$, $L=48$, $K=2$ and $N_b=3$, processing throughput is 100 frames per second.
-For all four benchmarks there is a line indicating the time it takes to acquire the underlying image using single line acquisition.
+For $M=64$, $L=32$, $K=2$ and $N_b=3$, processing throughput is 125 frames per second (8 ms).
+For $M=96$, $L=48$, $K=2$ and $N_b=3$, processing throughput is 77 frames per second (13 ms).
+In all four benchmarks there is a line indicating the time it takes to acquire the underlying image using single line acquisition.
 Hence, below this line we are capable of performing real-time processing when a single receive line is used per transmit line.
@@
 a) From Fig.
-\ref{fig:phantom} at 93 mm range.
-Straight through the two bright point scatterers.
+\ref{fig:phantom} at 93 mm constant depth, through the two bright point scatterers.
 b) From Fig.
-\ref{fig:invivo} at 96 mm range, going through the tip of the mitral valve.}
+\ref{fig:invivo} at 96 mm range, through the tip of the mitral valve.}
 \label{fig:slices}
@@
 \ref{fig:phantom}).
-The simulation was performed using a $\lambda/2$ pitch, 10 $\mu$m kerf, 50\% bandwidth and 10 mm high phased array.
-The transducer was excited with a 2.5 MHz and $3\lambda$ long pulse with transmit focus at 80 mm.
-The phantom contains two circular areas, one bright at 24 dB and one dark at -30 dB, and two point targets at 40 dB.
+The simulation was performed using a $\lambda/2$ pitch, 10 $\mu$m kerf, 80\% bandwidth and 10 mm high phased array.
+The transducer was excited with a 3.4 MHz and $1.5\lambda$ long pulse with transmit focus at 90 mm.
+The phantom contains two circular areas, one bright at 24 dB and one dark at -30 dB, and two point targets at 40 dB initially located at 90 mm depth.
 Speckle level is at 0 dB.
@@
 \ref{fig:slices}.
-For the phantom, the cut goes through the two point targets at 93 mm depth, and for the cardiac image it goes through the tip of the mitral valve at 96 mm depth.
+For the phantom, the cut goes through the two point targets at 93 mm constant depth, and for the cardiac image it goes through the tip of the mitral valve at 96 mm constant depth.
 
@@
 As seen from Fig.
-\ref{fig:bench} and \ref{fig:benchBS}, execution time is reduced when either $L$ or $N_b$ is reduced.
-So to gain an  increase in processing throughput we can reduce $L$ or $N_b$.
+\ref{fig:bench}, execution time is reduced when $L$ is reduced.
+In Fig.
+\ref{fig:benchBS} we see that execution time is also reduced when moving to beamspace.
+So to gain an  increase in processing throughput we can reduce $L$ in element space or $N_b$ in beamspace.
 The question is then, how does this impact image quality?
@@
 We see that the resolution for BS Capon gradually decreases due to an automatically increase in diagonal loading (See end of Section \ref{sec:bs}).
-If this is compensated for, BS Capon will provide resolution almost equal to Capon for all values except $N_b=1$ \cite{Nilsen2009}.
+If this is compensated for, BS Capon will provide resolution almost equal to Capon (with $L=M/2$) for all values except $N_b=1$ \cite{Nilsen2009}.
 For $N_b=1$ the resolution is worse than for DAS because of a triangular apodization caused by subarray averaging.
@@
 It is also worth noting that Capon beamforming in this example performs worse than DAS for small subarrays.
-For small values of $L$, there are not enough degrees of freedom in order steer zeros at each point target.
+For small values of $L$, there are not enough degrees of freedom in order perfectly steer a zero at each point target.
 
@@
 If parallel receive lines are used for increased frame rate, the acquisition time is lowered, and the processing time remains constant.
-Hence, four parallel receive lines are not possible to process in real-time using BS Capon in combination with our system configuration.
+Hence, four parallel receive lines are not possible to process in real-time using BS Capon (with $N_b=3$) in combination with our system configuration.
 
@@
 From Fig.
-\ref{fig:benchUS64} we see that for $L = M/2 = 32$ and $K=1$, the total execution time is 100 ms (10 fps).
+\ref{fig:bench} we see that for $L = M/2 = 32$ and $K=1$, the total execution time is 140 ms (7 fps).
 Taking into account that we use complex data, but less samples per image, these results are similar to the performance reported by Chen et al.
 \cite{Chen2011}.
-For the 32 element array in Fig.
-(\ref{fig:benchUS32}) execution time is reduced to 11.5 ms (87 fps) using the same settings ($L=M/2, K = 1$).
 
@@
 The time it takes to construct all covariance matrices is also reduced, but for few beams this is now the major contributor to the total execution time.
-If real-time is defined as 10 fps \cite{Chen2011} all execution times in both Fig.
+If real-time is defined as 10 fps \cite{Chen2011} almost all execution times in both Fig.
 \ref{fig:bench} and \ref{fig:benchBS} are real-time.
 More natural for cardiac ultrasound imaging is to define real-time as the time it takes to acquire the underlying image.
-With this requirement the images formed using a 32, 64 and 96 element array have a real-time requirement of 125, 62.5 and 41.7 fps respectively using single line acquisition.
+With this requirement the images formed using a 64 and 96 element array have a real-time requirement of 62.5 and 41.7 fps respectively using single line acquisition.
 Each benchmark plot has a line indicating this real-time requirement.
@@
 %For a typical cardiac ultrasound image of 80*400 pixels (70 degree sector, 15cm range) acquired using a 2.5MHz, $M=64$ element phased array, the result is 10fps (subarray length $L=M/2$, temporal smoothing over 3 samples).
-If we do a 2-element pre-beamforming to get the channel count down to 32, the
- rate increases to 44fps.
+If we do a 2-element pre-beamforming to get the channel count down to 32, the rate increases to 44fps.
 For a 32 element phased array we need less beams to cover the sector ($40 \times 400$ pixels), hence with the same parameters the frame rate increases to 87fps.
@@
 both beamformers have successfully been implemented on a GPU.
-For beamspace Capon beamforming processing rates supporting real-time cardiac ultrasound imaging was shown to be possible, meaning that processing rates higher than the image acquisition rate are obtained for a wide range of parameters.
+For a three dimensional beamspace, processing rates supporting real-time cardiac ultrasound imaging was shown to be possible, meaning that processing rates higher than the image acquisition rate are obtained for a wide range of parameters.
 Image quality was investigated both in a simulated phantom and an \textit{in-vivo} cardiac dataset.
-These results showed that Capon beamforming is feasible for cardiac ultrasound imaging, providing images with sharper structures and no local shift-variance effects.
+These results showed that Capon beamforming is feasible for cardiac ultrasound imaging, providing images with sharper structures and no local shift-variance effects, both in element- and beam-space.
 
@@
 
-\begin{IEEEbiography}{Jon Petter \AA{}sen}
+\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{photos/jp.jpg}}]{Jon Petter \AA{}sen}
+(S'12) was born in Porsgrunn, Norway in 1986.
+He received the M.S degree in computer science from the University of Oslo, Norway, in 2010.
+He is currently pursuing his Ph.D degree in medical ultrasound technology at the Norwegian University of Science and Technology (NTNU) Medical Imaging Lab (MI-Lab), Trondheim, Norway.
+His research interests include adaptive ultrasound processing techniques and acceleration of ultrasound algorithms using Graphics processing units (GPUs).
+\end{IEEEbiography}
+
+\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{photos/ji.jpg}}]{Jo Inge Buskenes}
 Biography text here.
@@
 
-\begin{IEEEbiography}{Jo Inge Buskenes}
-Biography text here.
-\end{IEEEbiography}
-
-\begin{IEEEbiography}{Carl-Inge Colombo Nilsen}
-Biography text here.
+\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{photos/ci.png}}]{Carl-Inge Colombo Nilsen}
+(S’06–M’10) received the M.sc.
+and Ph.d.
+degrees in computer science from the University of Oslo, Norway, in 2005 and 2010.
+He is currently working at the University of Oslo as a postdoctoral research fellow.
+His research interests include signal and array processing for ultrasound imaging and other acoustic applications.
 \end{IEEEbiography}
@@
 
-\begin{IEEEbiography}{Andreas Austeng}
-Biography text here.
+\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{photos/aa.jpg}}]{Andreas Austeng}
+was born in Oslo, Norway, in 1970.
+He received the M.S.
+degree in physics in 1996 and the Ph.D.
+degree in computer science in 2001, both from the University of Oslo.
+since 2001, he has been working at the department of Informatics, University of Oslo, first as a postdoctoral research fellow and currently as an associate Professor.
+His research interests include signal processing for acoustical imaging.
 \end{IEEEbiography}
 
-\begin{IEEEbiography}{Sverre Holm}
-Biography text here.
+\newpage
+
+\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{photos/sh.jpg}}]{Sverre Holm}
+(M’82-SM’02) was born in Oslo, Norway, in 1954.
+He received the M.S.
+and Ph.D.
+degrees in electrical engineering from the Norwegian Institute of Technology, Trondheim, in 1978 and 1982, respectively.
+His academic experience includes Yarmouk University in Jordan (1984-1986) and the Norwegian Institute of Technology and later the University of Oslo (1989-1994) as an adjunct professor.
+Since 1995, he has been a professor of signal processing in the department of Informatics at the University of Oslo.
+In 2002, he was elected a member of the Norwegian Academy of Technological Sciences, and in 2009, he became an adjunct professor in the Norwegian University of Science and Technology (NTNU) Medical Imaging Lab, Trondheim.
+His industry experience includes GE Vingmed Ultrasound, Horten, Norway (1990-1994).
+He worked with Sonitor Technologies (2000-2005), where he developed ultrasonic technology that is the core of the indoor positioning system described in \textit{Scientific Americanin} 2008 as "a positioning system that goes where GPS can’t." He has published about 130 publications internationally, plus about 10 patents.
+He has had sabbaticals at GE Global Research, NY (1998), and the Waves and
+Acoustics Laboratory of ESPCI, Paris (2008-2009).
+He was an associate editor of the \textit{IEEE Transactions on Ultrasonics, Ferroelectrics, and Frequency Control} from 1997 to 2002.
+His research interests are medical ultrasound imaging, modeling of tissue-ultrasound interaction, and ultrasound positioning.
 \end{IEEEbiography}

### 2013-03-07 HRUIonGPU/jrnl_article.tex
uffc article

@@
 The transducer was excited with a 3.4 MHz and $1.5\lambda$ long pulse with transmit focus at 90 mm.
-The phantom contains two circular areas, one bright at 24 dB and one dark at -30 dB, and two point targets at 40 dB initially located at 90 mm depth.
-Speckle level is at 0 dB.
+The phantom contains two circular areas, one bright at 24 dB and one dark at -30 dB, and two point targets at 40 dB initially located at 93 mm depth.
+Speckle level is at 0 dB, covering an area from $\pm90\degree$ in azimuth.
 The resulting image has been processed both using DAS with uniform weights, Capon and BS Capon.
@@
 \ref{fig:invivo} presents results of applying Capon and BS Capon on an \textit{in-vivo} image of the left ventricle.
-The image was acquired using an experimental research scanner equipped with a 64 elements and $0.65\lambda$ pitched phased array in harmonic mode (1.7 MHz at transmit and 3.5 MHz at receive).
+The image was acquired using a 64 elements and $0.65\lambda$ pitched phased array in harmonic mode (1.7 MHz at transmit and 3.5 MHz at receive).
 The image is extracted from a cardiac cycle, that can be viewed in its full length in an attached video.

### 2013-03-08 HRUIonGPU/jrnl_article.tex

@@
 %\boldmath
-Capon beamforming is associated with a high computational complexity, which limits its use as a real-time array processor in many applications.
-In this paper we present an implement of the Capon beamformer exhibiting real-time performance when applied in a typical cardiac ultrasound imaging setting.
+Capon beamforming is associated with a high computational complexity, which limits its use as a real-time method in many applications.
+In this paper we present an implementation of the Capon beamformer exhibiting real-time performance when applied in a typical cardiac ultrasound imaging setting.
 To reach this performance we make use of the parallel processing power found in modern Graphics Processing Units (GPUs), combined with beamspace processing to reduce the computational complexity as the number of array elements increases.
-For a three dimensional beamspace, processing rates supporting real-time cardiac ultrasound imaging is shown to be possible, meaning that processing rates higher than the image acquisition rate are obtained for a wide range of parameters.
+For a three dimensional beamspace, processing rates supporting real-time cardiac ultrasound imaging are shown to be possible, meaning that processing rates higher than the image acquisition rate are obtained for a wide range of parameters.
 Image quality is investigated in videos of both a simulated phantom and an \textit{in-vivo} cardiac dataset.
-These results show that Capon beamforming is feasible for cardiac ultrasound imaging, providing images with sharper structures and no local shift-variance effects, both in element- and beam-space.
+These results show that Capon beamforming is feasible for cardiac ultrasound imaging, providing images with sharper structures and no local shift-variance effects, both in element- and beamspace.
 \end{abstract}
@@
 \IEEEPARstart{C}{apon} beamforming \cite{Capon1969} applied to medical ultrasound imaging has recently shown promising results in several studies \cite{Synnevag2007, Austeng2008, Vignon2008, Viola, Mehdizadeh2012}.
-Most of this work focus on the improved lateral resolution and contrast obtained with the Capon beamformer.
-However, some interesting trade-offs has also been introduced.
+Most of this work focuses on the improved lateral resolution and contrast obtained with the Capon beamformer.
+However, some interesting trade-offs have also been introduced.
 In \cite{Synnevag2009}, Synnev\aa{}g et al.
@@
 \cite{Chen2011a, Chen2011, Chen} implemented a Capon beamformer on the GPU that produced real apodization weights based on real data.
-Under this constraint the Capon beamformer is not allowed to micro-steer the main lobe, and therefore its full potential is not reached.
+However, the Capon beamformer is meant to work on complex data.
+Under the constraint of real data, the Capon beamformer is not allowed to micro-steer the main lobe, and therefore its full potential is not reached.
 Micro-steering of the main lobe is an important property of the Capon beamformer in order for it to improve structural details and edge definitions in an ultrasound image (Fig 9.
 in \cite{Synnevag2009}).
-Although complex arithmetic requires more operations per sample, complex (I/Q or baseband) data can be sampled at a lower rate, thus minimizing the number of samples needed per image.
-All results in this paper are reported with the use of baseband data, hence where the sampling frequency is close to the system bandwidth.
+Complex arithmetic also requires more operations per sample, but the sampling frequency can also be reduced.
+The number of samples needed per image is therefore minimized.
+All results in this paper are reported with the use of baseband (or I/Q) data, where the sampling frequency is close to the system bandwidth.
 
@@
 In recent work \cite{asen2012, Buskenes} we have applied a GPU-based Capon beamformer for both active sonar and for cardiac ultrasound imaging.
-Sonar has a real-time requirement less strict than medical ultrasound imaging, which means that most array configurations can be handled in real-time using the GPU.
+Sonar has a less strict real-time requirement than medical ultrasound imaging, which means that most array configurations can be handled in real time using the GPU.
 Medical ultrasound imaging, and in particular cardiac ultrasound imaging, demands frame rates at a level where even a high-end GPU can not do the processing required for a straightforward implementation of the Capon beamformer.
-Since the execution time is a cubic with respect to the number of array elements, we need methods that reduce the computational complexity for large arrays.
+Since the execution time is cubic with respect to the number of array elements, we need methods that reduce the computational complexity for large arrays.
 Various approaches to low-complexity Capon beamforming have therefore recently been proposed in the literature \cite{Synnevag2011, Asl2012, Jensen2012, Kim}.
@@
 In the next section we present background information on the Capon beamformer and the GPU computing model.
-In Section \ref{sec:meth} we go in detail on the parallel implementation of the Capon beamformer, before the beamspace method, and how it has been implemented on the GPU, is presented in Section \ref{sec:bs}.
+In Section \ref{sec:meth} we go in detail on the parallel implementation of the Capon beamformer, before the beamspace method, and its GPU implementation, is presented in Section \ref{sec:bs}.
 Benchmark results and resulting images are presented in Section \ref{sec:bench} and \ref{sec:images}, together with a discussion of trading image quality for speed in Section \ref{sec:trade}.
@@
 \begin{figure}
-\centerline{\includegraphics[width=3in]{gfx/beamforming_mv.pdf}}
+\centerline{\includegraphics[width=3in]{gfx/beamforming_mv.eps}}
 \caption{Capon beamforming.
@@
 \end{align}
-has been applied in much of the resent literature on Capon beamforming for medical ultrasound imaging \cite{Synnevag2007, Nilsen2009, Wang2009, Mehdizadeh2012}.
+has been applied in much of the recent literature on Capon beamforming for medical ultrasound imaging \cite{Synnevag2007, Nilsen2009, Wang2009, Mehdizadeh2012}.
 It has also been applied to HF and VHF antenna processing \cite{Featherstone1997b}.
@@
 In passive systems, assuming stationary statistics, the sample covariance matrix in (\ref{eq:R}) is typically averaged over a long sequence of samples\cite{Krima}.
-Since medical ultrasound imaging is an active, short-pulsed, system, there is a limited amount of data from which $\mat{\hat{R}}$ can be estimated.
+Since medical ultrasound imaging is an active, short-pulsed system, there is a limited amount of data from which $\mat{\hat{R}}$ can be estimated.
 The estimated matrix is therefore often found to be both inaccurate and poorly conditioned.
@@
 In this section we give a brief discussion of GPU computing, laying the basis for further discussion on parallel implementation strategies.
-As GPU framework, we have selected CUDA by Nvidia \cite{Nvidia2011}, however the challenges and proposed methods described in this paper are valid for other GPU framworks as well, like e.g.
+As GPU framework, we have selected CUDA by Nvidia \cite{Nvidia2013}, however the challenges and proposed methods described in this paper are valid for other GPU framworks as well, like e.g.
 OpenCL.
 The key point is that GPUs can accelerate large sets of small concurrent tasks, like beamforming a pixel in a large image, much better than a CPU.
-While the CPU has few cores and logic to process complex threads, the GPU has many cores, but each core needs many (preferably simple) threads to keep it occupied.
-
-The CUDA compute model is based on execution of a kernel function over a large set of threads, where the kernel function describes the work to be done in each thread.
+While the CPU has few cores and little logic to process complex threads, the GPU has many cores, but each core needs many (preferably simple) threads to keep it occupied.
+
+The CUDA compute model is based on the execution of a kernel function over a large set of threads, where the kernel function describes the work to be done in each thread.
 As depicted in Fig.
-\ref{fig:gpulayout}, threads are organized in a grid of thread-blocks, with maximal size of 1024 threads per block (32x32).
-Threads inside a block supports low-cost synchronization if needed.
-From a CUDA perspective the GPU consist of one or more streaming multi-processors (SMs), where each SM is capable of 32 or more concurrent multiplications and/or additions, depending on the architecture.
+\ref{fig:gpulayout}, threads are organized in a grid of thread-blocks, with maximal size of 1024 threads per block ($32\times32$).
+Threads inside a block support low-cost synchronization if needed.
+From a CUDA perspective the GPU consist of one or more streaming multi-processors (SM), where each SM is capable of 32 or more concurrent multiplications and/or additions, depending on the architecture.
 A group of blocks is scheduled to each SM until all blocks have been processed.
 
-An SM has a limited amount of near-core memory and registers.
+An SM has a limited amount of near-core, or local, memory and registers.
 Hence, to achieve maximum GPU occupancy (having the maximum number of threads residing on each SM), each thread has to be as light on register and memory consumption as possible.
 It is also important that each thread performs enough instructions per transferred byte in order to hide memory latency, and that the problem can be divided into enough threads to hide instruction latency.
-Reads and writes to and from GPU DRAM, in addition to CPU-GPU transfers, are slow and should be minimized as much as possible.
-All of these constraints combined pose a real challenge as the target problem has to map well to the GPU framework on a grid, block, thread and memory consumption level in order to benefit from parallel acceleration.
+Reads and writes to and from global GPU memory, in addition to CPU-GPU transfers, are slow and should be minimized as much as possible.
+All of these constraints combined pose a real challenge as the target problem has to map well to the GPU framework on a grid, block, thread, and memory consumption level in order to benefit from parallel acceleration.
 In the next section, we will describe how this mapping can be done for the Capon beamforming algorithm.
@@
 \begin{figure}
-\centerline{\includegraphics[width=3.4in]{gfx/gpu_layout_vertical_lr.png}}
+\centerline{\includegraphics[width=3.5in]{gfx/gpu_layout_vertical.eps}}
 \caption{Calculation of Capon adaptive weights mapped to GPU architecture.
@@
 %\end{enumerate}
-The computational complexity of these three steps is $O(N_KN_LL^2)$, $O(L^3)$, and $O(L)$, respectively.
-As $M$ increases (and thereby $L$), the first two steps quickly become unmanageable, and real-time performance can only be achieved by exploring novel solutions to reduce complexity.
+The computational complexities of these three steps are $O(N_KN_LL^2)$, $O(L^3)$, and $O(L)$, respectively.
+As $M$ (and thereby $L$) increases, the first two steps quickly become unmanageable, and real-time performance can only be achieved by exploring novel solutions to reduce complexity.
 
@@
 %- \mat{\breve{R}}[n-K-1] + \mat{\breve{R}}[n-K].
-Both sliding across subarrays and in time can be realized simultaneously if $\mat{\hat{R}}$, averaged over subarrays, is calculate from an $N_K$-average in time of the outer products $\vec{x}[n]\vec{x}[n]^H$ \cite{Buskenes}.
+Both sliding across subarrays and in time can be realized simultaneously if $\mat{\hat{R}}$, averaged over subarrays, is calculated from an $N_K$-average in time of the outer products $\vec{x}[n]\vec{x}[n]^H$ \cite{Buskenes}.
 However, since $N_K$ typically is small, and the GPU has a limited amount of memory available per thread, we focus on sliding across subarrays only.
-In that case, the complexity of calculating $\mat{\hat{R}}$ is reduced to $O(N_KL^2)$ (if $L=M/2$).
+In that case, the complexity of calculating $\mat{\hat{R}}$ is reduced to $O(N_K(L(N_L-2)+2L^2))$.
+For $L=M/2$ this reduces to $O(N_KM^2)$.
 Note that sliding will break independence along the sliding dimension, thus, calculations are still independent across time, but not along diagonals of $\mat{\hat{R}}$.
@@
 Maximum supported $L$ is increased to the maximum number of threads per block (currently 1024) with this approach.
-However, for small $L$ high $M$ configurations, memory consumption per thread will be to high to achieve full occupancy of the GPU.
+However, for small-$L$-high-$M$ configurations, memory consumption per thread will be too high to achieve full occupancy of the GPU.
 
@@
 
-Diagonal loading as a constant value can be added with negligible cost.
 For the adaptive diagonal loading factor in (\ref{eq:diag}), the trace of $\mat{\hat{R}}$ has to be known before diagonal elements are written back to global memory.
@@
 However, there are two arguments for solving on the GPU.
-First the CPU, which might already run a lot of ultrasound related algorithms, is offloaded.
-Second, the transfer of several thousand covariance matrices over the PCI-Express bus, from the CPU to GPU, is avoided.
+First the CPU, which might already run a lot of ultrasound-related algorithms, is offloaded.
+Second, the transfer of several thousand covariance matrices over the PCI-Express bus, from the CPU to the GPU, is avoided.
 
 For the results presented in this paper, we have used a GPU implementation of batched Gauss-Jordan elimination made by Nvidia to solve $\mat{\hat{R}}\vec{b} = \vec{1}_L$, where $\vec{b} = \mat{\hat{R}}^{-1}\vec{1}_L$, for all pixels in one kernel launch.
-The solver is available online for registered CUDA developers.%For smaller systems, $L < 10$, a direct solver using one thread per system is preferable.
+The solver is available online for registered CUDA developers \cite{Nvidiaa}.%For smaller systems, $L < 10$, a direct solver using one thread per system is preferable.
 For this purpose we have implemented Cramer's rule for $2 \times 2$, $3 \times 3$ and $4 \times 4$ matrices.
@@
 \subsection{Compute Beamformer Output}
-To calculate the beamformer output, equation (\ref{eq:z_mv}) has been implemented using one thread per element-weight ($L$ threads per weight vector).
+To calculate the beamformer output, equation (\ref{eq:z_mv}) has been implemented using one thread per element weight ($L$ threads per weight vector).
 After the solution vector $\vec{b}$ has been placed in near-core memory, one of the $L$ threads finds the sum of $\vec{b}$.
@@
 \end{align}
-Each thread now computes $z_r = w_r^*x_r$, and finally one thread computes and outputs $z = \frac{1}{N_L}\sum_{r=0}^{L-1} z_r$ to  GPU DRAM.
+Each thread now computes $z_r = w_r^*x_r$, and finally one thread computes and outputs $z = \frac{1}{N_L}\sum_{r=0}^{L-1} z_r$ to global memory.
 %The last sum can also be performed using a binary reduction sum across several threads for increased speed.
@@
 \begin{align}
-\vec{x}_\text{BS}[n] &= \mat{B}\vec{x}[n], \ \ \ \mat{B} \in \mathbb{C}^{N,M}\\
-[\mat{B}]_{p,q} &= \frac{1}{\sqrt{M}}e^{-j 2 \pi p q / M}.
+\vec{x}_\text{BS}[n] &= \mat{B}\vec{x}[n], \ \ \ \ \ \ \ \ \ \ \ \ \ \, \mat{B} \in \mathbb{C}^{N,M}\\
+[\mat{B}]_{p,q} &= \frac{1}{\sqrt{M}}e^{-j 2 \pi p q / M} \ \ \ N \le M.
 \end{align} 
@@
 \end{align}
-The advantages of this approach is that minor changes have to be made to the pipeline described in Section \ref{sec:meth}, but it will only speed up the solving step if the solution $\vec{b}_{BS}$ is transformed back to element space.
+The advantage of this approach is that minor changes have to be made to the pipeline described in Section \ref{sec:meth}, but it will only speed up the solving step if the solution $\vec{b}_{BS}$ is transformed back to element space.
 In this paper we will focus on the first approach, and how it has been implemented on the GPU.
@@
 In this situation, it might be tempting to use one thread per beam (index $k$) that calculates this beam for all subarrays.
-However, since the number of selected beams usually is small ($N_b \ll M$) we will end up with a large memory per thread ratio ($M/N_b$).
-Using one thread per beam is therefore not feasible for small $N_b$ large $M$ configurations.
+However, since the number of selected beams usually is small ($N_b \ll M$) we will end up with a large memory-per-thread ratio ($M/N_b$).
+Using one thread per beam is therefore not feasible for small-$N_b$-large-$M$ configurations.
 The solution is to combine SBS with full computations.
 A number of thread groups is created that targets different sets of subarrays which they transform using the SBS algorithm.
-In that way, the number of threads per sample is increased and thus the memory per thread ratio is decreased.
+In that way, the number of threads per sample is increased and thus the memory-per-thread ratio is decreased.
 
@@
 Assuming a system bandwidth of 80\% we need a complex sampling frequency of 2.7 MHz, and when imaging down to a depth of 15.4 cm we have approximately 540 samples per range line.
-To correctly sample a $70\degree$ sector with 64 and 96 element arrays ($\lambda/2$ pitch), we need at least 80 and 120 receive beams respectively.
-The presented execution times do not include transfer of data from CPU- to the GPU-side, since data are assumed to be pre-delayed and available in GPU memory.
+To correctly sample a $70\degree$ sector with 64- and 96-element arrays ($\lambda/2$ pitch), we need at least 80 and 120 receive beams respectively.
+The presented execution times do not include transfer of data from CPU- to the GPU-side, since data are assumed to be pre-delayed and available in global GPU memory.
 What is included are all calculations, memory transfers from global to near-core GPU memory, and near core memory access.
@@
 \ref{fig:bench}.
-The benchmark is based on a 64 elements phased array, using 80 receive beams.
+The benchmark is based on a 64-element phased array, using 80 receive beams.
 For each subarray size, the total execution time is split into one bar for each of the three steps described in Section \ref{sec:meth}, and the calculation of $\mat{\hat{R}}$ is presented for three different values of $K$.
 For $M=64$, $L=32$ and $K=2$ (Fig.
-\ref{fig:bench}), a typical setup for high quality cardiac imaging, processing throughput is 7 frames per second.
+\ref{fig:bench}), a typical setup for high-quality cardiac imaging, processing throughput is 7 frames per second.
 
@@
 \ref{fig:benchBS} we present similar benchmarks for the beamspace version of the Capon beamformer introduced in Section \ref{sec:bs}.
-Benchmarks are based on 64 and 96 elements phased arrays, using 80 and 120 transmit beams.
-The figure only present results with use of temporal averaging ($K=2$).
+Benchmarks are based on 64- and 96-element phased arrays, using 80 and 120 transmit beams.
+The figure only presents results with use of temporal averaging ($K=2$).
 For $M=64$, $L=32$, $K=2$ and $N_b=3$, processing throughput is 125 frames per second (8 ms).
 For $M=96$, $L=48$, $K=2$ and $N_b=3$, processing throughput is 77 frames per second (13 ms).
-In all four benchmarks there is a line indicating the time it takes to acquire the underlying image using single line acquisition.
-Hence, below this line we are capable of performing real-time processing when a single receive line is used per transmit line.
+In all four benchmarks there is a line indicating the time it takes to acquire the underlying image using one receive line per transmit.
+Hence, below this line we are capable of performing real-time processing.
 
@@
 \subfloat[]{\includegraphics[width=2.3in]{gfx/capon_bs.png} \label{fig:phantomBS}}}
-\caption{Simulated phantom containing one bright (24 dB) and one dark (-30 dB) circular area and two point targets (40 dB) in speckle.
-The simulation was done using Field II with a 2.5 MHz, $\lambda/2$ and 64 elements phased array.
-Dynamic range is -30 dB to 30 dB with mean value represented as 0 dB.
+\caption{Simulated phantom containing one bright (20 dB) and one dark (-20 dB) circular area and two point targets (30 dB) in speckle (0 dB).
+The simulation was done using Field II with a 64-element, 3.4 MHz, and $\lambda/2$ pitch phased array.
+Dynamic range is -25 dB to 25 dB with mean value represented as 0 dB.
 a) Delay-and-sum with uniform weights.
@@
 \ref{fig:phantom}).
-The simulation was performed using a $\lambda/2$ pitch, 10 $\mu$m kerf, 80\% bandwidth and 10 mm high phased array.
+The simulation was performed using a 64-element, $\lambda/2$ pitch, 10 $\mu$m kerf, 80\% bandwidth, and 10 mm high phased array.
 The transducer was excited with a 3.4 MHz and $1.5\lambda$ long pulse with transmit focus at 90 mm.
-The phantom contains two circular areas, one bright at 24 dB and one dark at -30 dB, and two point targets at 40 dB initially located at 93 mm depth.
+The phantom contains two circular areas, one bright at 20 dB and one dark at -20 dB, and two point targets at 30 dB initially located at 90 mm depth.
 Speckle level is at 0 dB, covering an area from $\pm90\degree$ in azimuth.
@@
 \ref{fig:invivo} presents results of applying Capon and BS Capon on an \textit{in-vivo} image of the left ventricle.
-The image was acquired using a 64 elements and $0.65\lambda$ pitched phased array in harmonic mode (1.7 MHz at transmit and 3.5 MHz at receive).
+The image was acquired using a 64-element and $0.65\lambda$ pitched phased array in harmonic mode (1.7 MHz at transmit and 3.4 MHz at receive).
 The image is extracted from a cardiac cycle, that can be viewed in its full length in an attached video.
-It is included to show that Capon beamforming in both element- and beam-space is feasible for cardiac imaging, not just in simulations.
+It is included to show that Capon beamforming in both element- and beamspace is feasible for cardiac imaging, not just in simulations.
 
@@
 \ref{fig:slices}.
-For the phantom, the cut goes through the two point targets at 93 mm constant depth, and for the cardiac image it goes through the tip of the mitral valve at 96 mm constant depth.
+For the phantom, the cut goes through the two point targets at 90 mm constant depth, and for the cardiac image it goes through the tip of the mitral valve at 96 mm constant depth.
 
@@
 \section{Discussion}\label{sec:dis}
-In this paper we have presented a parallel implementation of the Capon beamformer in both element- and beam-space for high resolution cardiac ultrasound imaging.
+In this paper we have presented a parallel implementation of the Capon beamformer in both element- and beamspace for high-resolution cardiac ultrasound imaging.
 The results show that it is now possible to apply BS Capon in real-time in cardiac ultrasound imaging by utilizing the power of GPUs.
@@
 
-For Capon in element space, processing rates for high quality settings are still to high too support real-time cardiac ultrasound imaging.
+For Capon in element space, processing rates for high quality settings are still too high to support real-time cardiac ultrasound imaging.
 From Fig.
@@
 
-A major part of the total execution time for element space Capon comes from solving the system of linear equations.
+A major part of the total execution time for element-space Capon comes from solving the system of linear equations.
 This limitation is targeted by reducing the sample covariance matrix size in beamspace, and as shown in Fig.
-\ref{fig:benchBS}, inverting the 3x3 beamspace covariance matrix is then barely visible.
+\ref{fig:benchBS}, inverting the $3\times3$ beamspace covariance matrix is then barely visible.
 The time it takes to construct all covariance matrices is also reduced, but for few beams this is now the major contributor to the total execution time.
@@
 This paper further shows how to obtain a parallel formulation of the Capon and BS Capon beamformer, and how to implement them in a GPU framework.
-The per-pixel granularity suggest that the beamformer is easily dividable across multiple GPUs.
+The per-pixel granularity suggests that the beamformer is easily dividable across multiple GPUs.
 Real-time execution for all cases presented in this paper is therefore possible if several GPUs are used simultaneously.
@@
 Our preliminary finding is that the Capon beamformer is shift invariant (see video of the simulation) as long as the lateral sampling is increased enough.
-How large this oversampling factor needs to be and how the oversampling should be conducted is left unanswered in this paper.
+How large this oversampling factor needs to be and how the oversampling should be conducted is outside the scope of this paper.
 No major improvement in contrast is visually observed in these images, other than slightly rounder circles due to mainlobe micro steering.
@@
 \section{Conclusion}\label{sec:con}
-In this paper a parallel formulation and implementation of the element- and beam-space Capon beamformer have been presented, and
+In this paper a parallel formulation and implementation of the element- and beamspace Capon beamformer have been presented, and
 both beamformers have successfully been implemented on a GPU.
@@
 Image quality was investigated both in a simulated phantom and an \textit{in-vivo} cardiac dataset.
-These results showed that Capon beamforming is feasible for cardiac ultrasound imaging, providing images with sharper structures and no local shift-variance effects, both in element- and beam-space.
+These results showed that Capon beamforming is feasible for cardiac ultrasound imaging, providing images with sharper structures and no local shift-variance effects, both in element- and beamspace.
 
@@
 
-\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{photos/jp.jpg}}]{Jon Petter \AA{}sen}
+\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{photos/jon_petter.eps}}]{Jon Petter \AA{}sen}
 (S'12) was born in Porsgrunn, Norway in 1986.
-He received the M.S degree in computer science from the University of Oslo, Norway, in 2010.
+He received the B.S and M.S degree in computer science from the University of Oslo, Norway, in 2008 and 2010.
 He is currently pursuing his Ph.D degree in medical ultrasound technology at the Norwegian University of Science and Technology (NTNU) Medical Imaging Lab (MI-Lab), Trondheim, Norway.
+
+He has industry experience from GE Vingmed Ultrasound, Horten, Norway, were he completed three summer internships from 2008 to 2010.
 His research interests include adaptive ultrasound processing techniques and acceleration of ultrasound algorithms using Graphics processing units (GPUs).
@@
 
-\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{photos/ji.jpg}}]{Jo Inge Buskenes}
-Biography text here.
+\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{photos/jo_inge.eps}}]{Jo Inge Buskenes}
+received the B.S.
+degree in electrical engineering from Gj\o{}vik College University, Norway, in 2007, and the M.S.
+degree in instrumentation for particle physics from the University of Oslo, Norway, in 2010.
+He is currently pursuing the Ph.D.
+degree in image reconstruction and technology at the University of Oslo.
+
+His industry experience includes the European Organization for Nuclear Research (CERN), Geneva, Switzerland (2007-2008), and the Norwegian Defence Research Establishment, Kjeller, Norway (2009).
+He has lectured in digital signal processing at the Gj\o{}vik College University (2009), and at the University of Oslo (2010-2012).
+His research interests include adaptive beamforming, digital image reconstruction, high performance computing, intelligent detector design and embedded Linux platforms.
 \end{IEEEbiography}
 
-\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{photos/ci.png}}]{Carl-Inge Colombo Nilsen}
+\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{photos/carl-inge.eps}}]{Carl-Inge Colombo Nilsen}
 (S’06–M’10) received the M.sc.
@@
 
-\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{photos/aa.jpg}}]{Andreas Austeng}
+\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{photos/andreas.eps}}]{Andreas Austeng}
 was born in Oslo, Norway, in 1970.
@@
 
-\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{photos/sh.jpg}}]{Sverre Holm}
+\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{photos/sverre.eps}}]{Sverre Holm}
 (M’82-SM’02) was born in Oslo, Norway, in 1954.
@@
 His industry experience includes GE Vingmed Ultrasound, Horten, Norway (1990-1994).
-He worked with Sonitor Technologies (2000-2005), where he developed ultrasonic technology that is the core of the indoor positioning system described in \textit{Scientific Americanin} 2008 as "a positioning system that goes where GPS can’t." He has published about 130 publications internationally, plus about 10 patents.
+He worked with Sonitor Technologies (2000-2005), where he developed ultrasonic technology that is the core of the indoor positioning system described in \textit{Scientific Americanin} 2008 as "a positioning system that goes where GPS can't." He has published about 130 publications internationally, plus about 10 patents.
 He has had sabbaticals at GE Global Research, NY (1998), and the Waves and

### 2013-03-13 HRUIonGPU/jrnl_article.tex
adjusted plots

@@
 
-For high $L$ and $N_b$ values the two methods are equal.
-However, for lower values, BS Capon provides higher resolution than what is possible with small subarrays.
-We see that the resolution for BS Capon gradually decreases due to an automatically increase in diagonal loading (See end of Section \ref{sec:bs}).
-If this is compensated for, BS Capon will provide resolution almost equal to Capon (with $L=M/2$) for all values except $N_b=1$ \cite{Nilsen2009}.
+For high values of $L$ and $N_b$ the two methods are equal, but for lower values, BS Capon provides improved resolution compared to using small subarrays.
+We see that the resolution for BS Capon gradually decreases due to an automatic increase in diagonal loading (See end of Section \ref{sec:bs}).
+If we compensated by reducing $d$, BS Capon will provide resolution almost equal to Capon (with $L=M/2$) for all values except $N_b=1$ \cite{Nilsen2009}.
 For $N_b=1$ the resolution is worse than for DAS because of a triangular apodization caused by subarray averaging.
@@
 For $N_b=3$, BS Capon still provides approximately 1 mm improved resolution over DAS.
-However, note that resolution for the Capon beamformer highly depends on image SNR and the choice of parameters.
-Thus, different SNR scenarios and choices of parameters will give different results.
+However, note that resolution of the Capon beamformer strongly depends on image SNR and the choice of parameters.
+Thus, different SNR scenarios and choices of parameters will lead to different results.
 It is also worth noting that Capon beamforming in this example performs worse than DAS for small subarrays.
-For small values of $L$, there are not enough degrees of freedom in order perfectly steer a zero at each point target.
+For small values of $L$, the amount of robustification is too high in order to perfectly steer a zero at each point target.
+The Capon beamformer therefore produce a result worse than DAS.
 
@@
 \centering
-\includegraphics[width=3in]{gfx/speed_res_trade.png}
+\includegraphics[width=3in]{gfx/speed_res_trade_d=0-01.png}
 \caption{Trade-off between resolution and speed by reducing subarray size or reducing beamspace dimension (based on Fig.

### 2013-03-13 HRUIonGPU/gfx/make_plots.py

@@ -15,7 +15,7 @@ from all_plots import barBeamspaceBenchmarkGPU2 as barBS2
 bar('timer_values_M=64_Nx=80_Ny=540_L=2-32.txt', 64, 80, 540, 2, 32)
 
 barBS('timer_values_M=64_Nx=80_Ny=540_L=16-48_Nb=3.txt', 64, 3, 80, 540, 16, 47)
-barBS('timer_values_M=96_Nx=120_Ny=540_L=24-72_Nb=3.txt', 96, 3, 120, 540, 24, 71)
+barBS('timer_values_M=96_Nx=120_Ny=540_L=24-72_Nb=3.txt', 96, 3, 120, 540, 24, 71, False)
 
 #barBS2('timer_values_M=64_Nx=80_Ny=606_L=32_Nb=1-5.txt', 64, 32, 80, 606, 1, 5, False)
 #barBS2('timer_values_M=96_Nx=120_Ny=606_L=48_Nb=1-5.txt', 96, 48, 120, 606, 1, 5)

### 2013-03-13 HRUIonGPU/gfx/make_plots.py
Added photos. Started to integrate some comments from Andreas in tuffc

@@ -14,8 +14,8 @@ from all_plots import barBeamspaceBenchmarkGPU2 as barBS2
 
 bar('timer_values_M=64_Nx=80_Ny=540_L=2-32.txt', 64, 80, 540, 2, 32)
 
-barBS('timer_values_M=64_Nx=80_Ny=540_L=16-48_Nb=3.txt', 64, 3, 80, 540, 16, 47)
-barBS('timer_values_M=96_Nx=120_Ny=540_L=24-72_Nb=3.txt', 96, 3, 120, 540, 24, 71, False)
+barBS('timer_values_M=64_Nx=80_Ny=540_L=16-48_Nb=3.txt', 64, 3, 80, 540, 16, 47, False)
+barBS('timer_values_M=96_Nx=120_Ny=540_L=24-72_Nb=3.txt', 96, 3, 120, 540, 24, 71)
 
 #barBS2('timer_values_M=64_Nx=80_Ny=606_L=32_Nb=1-5.txt', 64, 32, 80, 606, 1, 5, False)
 #barBS2('timer_values_M=96_Nx=120_Ny=606_L=48_Nb=1-5.txt', 96, 48, 120, 606, 1, 5)

### 2013-03-14 HRUIonGPU/jrnl_article.tex

@@
 
-\author{J. P. \AA{}sen, J. I. Buskenes, C.-I. C. Nilsen, A. Austeng and S. Holm%\IEEEmembership{Member,~IEEE,}% <-this % stops a space
-%\thanks{M. Shell is with the Department
-%of Electrical and Computer Engineering, Georgia Institute of Technology, Atlanta,
-%GA, 30332 USA e-mail: (see http://www.michaelshell.org/contact.html).}% <-this % stops a space
-%\thanks{J. Doe and J. Doe are with Anonymous University.}% <-this % stops a space
-%\thanks{Manuscript received April 19, 2005; revised January 11, 2007.}}
+\author{J.
+P.
+\AA{}sen, \IEEEmembership{Student Member,~IEEE,} J.
+I.
+Buskenes, C.-I.
+C.
+Nilsen, \IEEEmembership{Member,~IEEE,} A.
+Austeng, \IEEEmembership{Member,~IEEE,} and S.
+Holm, \IEEEmembership{Senior Member,~IEEE,}% <-this % stops a space
+\thanks{Manuscript received ...; accepted ...}
+\thanks{Jon Petter \AA{}sen and Sverre Holm is with the Medical Imaging Lab (MI-Lab) at the Norwegian University of Sience and Technology, Trondheim, Norway.
+Jo Inge Buskenes, Carl-Inge Colombo Nilsen, Andreas Austeng and Sverre Holm is with the Department of Informatics, University of Oslo, Oslo, Norway.
+(e-mail: jon.p.asen@ntnu.no)}
 }
@@
 % The paper headers
-\markboth{Transactions on Ultrasonics and Ferroelectrics and Frequency Control,~Vol.~x, No.~y, March~2013}%
+\markboth{Transactions on Ultrasonics and Ferroelectrics and Frequency Control,~Vol.~x, No.~y, Z~201\ae{}}%
 {\AA{}sen \MakeLowercase{\textit{et al.}}: Capon Adaptive Beamforming on the GPU}
@@
 Under the constraint of real data, the Capon beamformer is not allowed to micro-steer the main lobe, and therefore its full potential is not reached.
-Micro-steering of the main lobe is an important property of the Capon beamformer in order for it to improve structural details and edge definitions in an ultrasound image (Fig 9.
-in \cite{Synnevag2009}).
+Micro-steering of the main lobe is an important property of the Capon beamformer in order for it to improve structural details and edge definitions in an ultrasound image \cite[Fig 9.]{Synnevag2009}.
 Complex arithmetic also requires more operations per sample, but the sampling frequency can also be reduced.
@@
 Various approaches to low-complexity Capon beamforming have therefore recently been proposed in the literature \cite{Synnevag2011, Asl2012, Jensen2012, Kim}.
-In this paper we will make use of a beamspace transformation described by Nilsen et al.
-\cite{Nilsen2009}, in order to maintain high framerates as the number of elements increases.
+In this paper we will make use of a beamspace transformation described by Nilsen and Hafizovic \cite{Nilsen2009}, in order to maintain high framerates as the number of elements increases.
 The beamspace transformation matches well with the focused transmit beams used in cardiac ultrasound imaging.
@@
 This delayed data is here denoted $x_m[n]$.
-The beamformer output, $z[n]$, is then the sum of all $M$ elements weighted with a factor $w_m$,
+The beamformer output $z[n]$ is then the sum of all $M$ elements weighted with a factor $w_m$,
 \begin{align}
@@
 \end{align}
-where  $N_K = 2K + 1$ is the number of samples included in time, $\vec{x}_l = [x_l[n], \dotso, x_{l+L-1}[n]]$ is the $l^\text{th}$ subarray of length $L$, and $N_L = M-L+1$ is the number of subarrays.
+where  $N_K = 2K + 1$ is the number of samples included in time, $\vec{x}_l = [x_l[n], \dotso, x_{l+L-1}[n]]$ is the $l\text{th}$ subarray of length $L$, and $N_L = M-L+1$ is the number of subarrays.
 Note that the covariance matrix dimension is $L \times L$.
@@
 \begin{align}\label{eq:diag}
-\mat{\hat{R}}[n] =  f(\vec{x}[n]) = \mat{\breve{R}}[n] + \epsilon[n]\mat{I}.
-\end{align}
-A weight proportional to the output power is often used to reduce the need for parameter adjustments, and the trace of $\mat{\breve{R}}$, 
+\mat{\hat{R}}[n] = \mat{\breve{R}}[n] + \epsilon[n]\mat{I} = f(\vec{x}[n]).
+\end{align}
+Where $f(\cdot)$ represents the process of constructing $\mat{\hat{R}}$ from the input $\vec{x}$.
+A factor proportional to the output power is often used to reduce the need for parameter adjustments, and the trace of $\mat{\breve{R}}$, 
 \begin{align}\label{eq:diag_adapt}
@@
 For our baseband data, one pulse length ($1.5\lambda$) is around 3 samples ($K=1$).
+Typical values for $d$ is between $1/10$ and $1/100$.
 
@@
 The key point is that GPUs can accelerate large sets of small concurrent tasks, like beamforming a pixel in a large image, much better than a CPU.
-While the CPU has few cores and little logic to process complex threads, the GPU has many cores, but each core needs many (preferably simple) threads to keep it occupied.
+While the CPU has few cores and little logic to process complex threads, the GPU has many cores, but each core needs many (preferably simple) threads to keep it occupied and to achieve maximum throughput.
 
@@
 As depicted in Fig.
-\ref{fig:gpulayout}, threads are organized in a grid of thread-blocks, with maximal size of 1024 threads per block ($32\times32$).
+\ref{fig:gpulayout}, threads are organized in a grid of thread-blocks, with maximal size of 1024 threads per block (e.g.
+$32\times32$).
 Threads inside a block support low-cost synchronization if needed.
@@
 
-An SM has a limited amount of near-core, or local, memory and registers.
-Hence, to achieve maximum GPU occupancy (having the maximum number of threads residing on each SM), each thread has to be as light on register and memory consumption as possible.
-It is also important that each thread performs enough instructions per transferred byte in order to hide memory latency, and that the problem can be divided into enough threads to hide instruction latency.
-Reads and writes to and from global GPU memory, in addition to CPU-GPU transfers, are slow and should be minimized as much as possible.
+An SM has a limited amount of local (near-core or shared) memory and registers.
+Therefore, to achieve maximum GPU occupancy (defined as having the maximum number of threads residing on each SM), each thread has to consume as few registers and as little local memory as possible.
+Each thread should also perform enough instructions per transferred byte in order to hide memory latency.
+Dividing the problem into enough threads will help to hide instruction latency.
+Global GPU memory transactions, in addition to CPU-GPU transfers, are slow and must be minimized.
 All of these constraints combined pose a real challenge as the target problem has to map well to the GPU framework on a grid, block, thread, and memory consumption level in order to benefit from parallel acceleration.
@@
 A recorded echo across the array for a given range and angle (selected cell) is assigned to a group of $L$ threads that can run independently of all other groups.
-Note that several samples can be processed per block of threads if $L$ is small.
+Note that several samples can be processed per thread-block if $L$ is small.
 The task of these $L$ threads is to calculate the Capon beamformer output in the three depicted steps.}
@@
 This will minimize global reads and writes needed per thread, but since we want to avoid block-to-block synchronization, the maximum subarray length is limited to 32 by this approach.
-Since most medical ultrasound arrays have more than 32 elements, an implementation not limited by $L_{max}=32$ is preferable.
+Since most medical ultrasound arrays have 64 elements or more, an implementation not limited by $L_{max}=M/2=32$ is preferable.
 
 To reach this goal we start by showing how the cubic complexity with $L$ and linear complexity with $N_K$ can be reduced by realizing that calculations in (\ref{eq:R}) overlap both across subarrays and in time.
-Element $(i,j)$ of $\mat{\breve{R}}[n]$ can be calculated from element $(i-1, j-1), \text{for } i,j \in [1, L-1]$ as
+Element $(i,j)$ of $\mat{\breve{R}}[n]$ can be calculated from element $(i-1, j-1), \text{for } i,j \in [1, L-1]$ in a sliding window maner as
 \begin{align}
@@
 %- \mat{\breve{R}}[n-K-1] + \mat{\breve{R}}[n-K].
-Both sliding across subarrays and in time can be realized simultaneously if $\mat{\hat{R}}$, averaged over subarrays, is calculated from an $N_K$-average in time of the outer products $\vec{x}[n]\vec{x}[n]^H$ \cite{Buskenes}.
-However, since $N_K$ typically is small, and the GPU has a limited amount of memory available per thread, we focus on sliding across subarrays only.
+Sliding windows across subarrays and in time can be realized simultaneously if $\mat{\hat{R}}$, averaged over subarrays, is calculated from an $N_K$-average in time of the outer products $\vec{x}[n]\vec{x}[n]^H$ \cite{Buskenes}.
+However, since $N_K$ typically is small, and the GPU has a limited amount of local memory available per thread, we focus on subarrays only.
 In that case, the complexity of calculating $\mat{\hat{R}}$ is reduced to $O(N_K(L(N_L-2)+2L^2))$.
 For $L=M/2$ this reduces to $O(N_KM^2)$.
-Note that sliding will break independence along the sliding dimension, thus, calculations are still independent across time, but not along diagonals of $\mat{\hat{R}}$.
+The sliding-window approach will cause dependence among the iterative calculations.
+The level of task granularity is thereby increased.
+When sliding across subarrays, calculations are still independent across time, but not along diagonals of $\mat{\hat{R}}$.
 One thread is therefore used per diagonal in the upper triangle of $\mat{\hat{R}}$, and for small values of $L$, several covariance matrices are calculated per thread block (Fig.
 \ref{fig:gpulayout}).
-Maximum supported $L$ is increased to the maximum number of threads per block (currently 1024) with this approach.
+Maximum supported $L$ is now increased to the maximum number of threads per block (currently 1024).
 However, for small-$L$-high-$M$ configurations, memory consumption per thread will be too high to achieve full occupancy of the GPU.
@@
 First the GPU needs to solve thousands of small systems in order to beat the low memory latency of the CPU.
-Second there is a range of system dimensions that does not fit well with the GPU architecture, where only a small amount of near-core memory and registers are available.
+Second there is a range of system dimensions that does not fit well with the GPU architecture, where only a small amount of local memory and registers are available.
 Because of this, it has been proven hard to optimize a manageable set of solvers that provides speedups compared to the CPU in every case.
@@
 \subsection{Compute Beamformer Output}
-To calculate the beamformer output, equation (\ref{eq:z_mv}) has been implemented using one thread per element weight ($L$ threads per weight vector).
-After the solution vector $\vec{b}$ has been placed in near-core memory, one of the $L$ threads finds the sum of $\vec{b}$.
+To calculate the beamformer output, (\ref{eq:z_mv}) has been implemented using one thread per element weight ($L$ threads per weight vector).
+After the solution vector $\vec{b}$ has been placed in local memory, one of the $L$ threads finds the sum of $\vec{b}$.
 %Another option is to let several threads cooperate to form the sum using a binary reduction sum. However, in the following benchmarks the first approach has been used. 
@@
 \section{Beamspace Processing}\label{sec:bs}
-Nilsen et al.
-\cite{Nilsen2009} have proposed to apply beamspace processing to further reduce the required computations of the Capon beamformer when applied to medical ultrasound imaging.
+Nilsen and Hafizovic \cite{Nilsen2009} have proposed to apply beamspace processing to further reduce the required computations of the Capon beamformer when applied to medical ultrasound imaging.
 Beamspace is an alternative representation of the per-element data, transformed into a fan of beams covering the sector illuminated by transmit beams.
@@
 \end{align} 
-The steering vectors in $\mat{B}$, the so-called Butler matrix, determine each beam's direction in space.
+The steering vectors in $\mat{B}$, the so-called Butler or normalized DFT matrix, determine each beam's direction in space.
 Beams where no interference is present can therefore be removed by simply removing rows from $\mat{B}$.
@@
 Nilsen et al.
-concluded that as little as three beams around (and including) the transmit direction were adequate to produce results comparable to applying Capon beamforming in element space.
-Hence, for $\lambda$-factor spaced arrays the inversion steps is found to be constant with respect to the number of array elements.
+concluded that as little as three beams around (and including) the transmit direction were adequate to produce results comparable to applying Capon beamforming in element space (Section \ref{sec:meth}).
+Hence, for  equally pitched arrays the inversion steps is found to be constant with respect to the number of array elements.
 
@@
 \end{align}
-where $\mat{B} \in \mathbb{C}^{N_b,L}$, and $N_b$ is the number of selected beams.
+where $\mat{B} \in \mathbb{C}^{N_b,L}$, $N_b$ is the number of selected beams, and $f(\cdot)$ is as defined in (\ref{eq:diag}).
 If the transformation is applied before the covariance matrix is constructed, one transformation is performed per subarray, yielding subarrays that no longer overlap in space.
@@
 \end{align}
-where $\vec{e}_i$ has the value one in the i$^{th}$ position and zeroes elsewhere.
+where $\vec{e}_i$ has the value one in the $i$th position and zeroes elsewhere.
 
@@
 %Derivation of sliding beamspace transformation. 
-Given a $M$ long data vector divided into $N_L$ subarrays, the beamspace transformation of the $l^{th}$ subarray for the $k^{th}$ beam is:
+Given a $M$ long data vector divided into $N_L$ subarrays, the beamspace transformation of the $l$th subarray for the $k$th beam is:
 \begin{align}\label{eq:bs_formula}
@@
 \end{align}
-We denote this the sliding beamspace transformation (SBS), and it is equal to a normalized sliding DFT \cite{Lyons2003}. For $l=0$ a full computation has to be performed using (\ref{eq:bs_formula}).
+We denote this the sliding-window beamspace transformation (SBS), and it is equal to a normalized sliding DFT \cite{Lyons2003}.
+For $l=0$ a full computation has to be performed using (\ref{eq:bs_formula}).
 
@@
 \centerline{\includegraphics[width=3.5in]{gfx/benchmark_bar_64_80_540.png}}
-\caption{Benchmark of GPU Capon beamforming for a cardiac image covering a $70\degree$ wide $15$ cm deep sector.
+\caption{Benchmark of GPU Capon beamforming in element space for a cardiac image covering a $70\degree$ wide $15$ cm deep sector.
 Execution time is plotted as a function of subarray length $L$.
-The plot shows execution times for the three steps listed in Section \ref{sec:meth}.
-Calculating covariance matrices with three different choices of temporal averaging, solver, and the final beamformer sum.}
+The plot shows execution times for the three steps listed in Section \ref{sec:meth}: Calculating covariance matrices with three different choices of temporal averaging, solver, and the final beamformer sum.}
 \label{fig:bench}
@@
 The presented execution times do not include transfer of data from CPU- to the GPU-side, since data are assumed to be pre-delayed and available in global GPU memory.
-What is included are all calculations, memory transfers from global to near-core GPU memory, and near core memory access.
-
-A benchmark of the implementation described in Section \ref{sec:meth} is presented in Fig.
+What is included are all calculations, memory transfers from global to local GPU memory, and local memory access.
+
+A benchmark of the element-space Capon, as described in Section \ref{sec:meth}, is presented in Fig.
 \ref{fig:bench}.
@@
 The simulation was done using Field II with a 64-element, 3.4 MHz, and $\lambda/2$ pitch phased array.
-Dynamic range is -25 dB to 25 dB with mean value represented as 0 dB.
+Dynamic range is -20 dB to 20 dB with mean value represented as 0 dB.
+The image is the 15th frame in the attached video (ADD VIDEO FILENAME IN RED).
 a) Delay-and-sum with uniform weights.
@@
 Dynamic range is -20 dB to 20 dB with mean value represented as 0 dB.
+The image is the first frame in the attached video (ADD VIDEO FILENAME IN RED).
 a) Delay-and-sum with uniform weights.
@@
 
-To investigate the image quality of both the Capon and BS Capon beamformer a phantom was simulated using Field II \cite{Jensen1992} (Fig.
+To investigate the image quality of both the Capon and BS Capon beamformer a phantom was simulated using Field II \cite{Jensen1992, Jensen1996a} (Fig.
 \ref{fig:phantom}).
@@
 For the simulated phantom, 16x lateral oversampling on transmit has been used, in order to avoid under sampling of the two bright point scatterers as they move.
-The cardiac recording only has $30\%$ lateral oversampling on receive.
+The cardiac recording has no lateral oversampling on receive.
 
@@
 
-\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{photos/jon_petter.eps}}]{Jon Petter \AA{}sen}
+\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{photos/jon_petter2.eps}}]{Jon Petter \AA{}sen}
 (S'12) was born in Porsgrunn, Norway in 1986.
-He received the B.S and M.S degree in computer science from the University of Oslo, Norway, in 2008 and 2010.
-He is currently pursuing his Ph.D degree in medical ultrasound technology at the Norwegian University of Science and Technology (NTNU) Medical Imaging Lab (MI-Lab), Trondheim, Norway.
+He received the B.Sc.
+and M.Sc.
+degree in computer science from the University of Oslo, Norway, in 2008 and 2010.
+He is currently pursuing his Ph.D.
+degree in medical ultrasound technology at the Norwegian University of Science and Technology (NTNU) Medical Imaging Lab (MI-Lab), Trondheim, Norway.
 
 He has industry experience from GE Vingmed Ultrasound, Horten, Norway, were he completed three summer internships from 2008 to 2010.
-His research interests include adaptive ultrasound processing techniques and acceleration of ultrasound algorithms using Graphics processing units (GPUs).
+His research interests include adaptive ultrasound processing techniques and acceleration of ultrasound algorithms using Graphics Processing Units (GPUs).
 \end{IEEEbiography}
 
-\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{photos/jo_inge.eps}}]{Jo Inge Buskenes}
-received the B.S.
-degree in electrical engineering from Gj\o{}vik College University, Norway, in 2007, and the M.S.
+\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{photos/jo_inge2.eps}}]{Jo Inge Buskenes}
+received the B.Sc.
+degree in electrical engineering from Gj\o{}vik College University, Norway, in 2007, and the M.Sc.
 degree in instrumentation for particle physics from the University of Oslo, Norway, in 2010.
@@
 \begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{photos/carl-inge.eps}}]{Carl-Inge Colombo Nilsen}
-(S’06–M’10) received the M.sc.
-and Ph.d.
+(S’06–M’10) received the M.Sc.
+and Ph.D.
 degrees in computer science from the University of Oslo, Norway, in 2005 and 2010.
@@
 degree in computer science in 2001, both from the University of Oslo.
-since 2001, he has been working at the department of Informatics, University of Oslo, first as a postdoctoral research fellow and currently as an associate Professor.
-His research interests include signal processing for acoustical imaging.
+Since 2001, he has been working at the Department of Informatics, University of Oslo, first as a postdoctoral research fellow and currently as an associate professor.
+His research interests include signal and array processing for acoustical imaging.
 \end{IEEEbiography}
@@
 His academic experience includes Yarmouk University in Jordan (1984-1986) and the Norwegian Institute of Technology and later the University of Oslo (1989-1994) as an adjunct professor.
-Since 1995, he has been a professor of signal processing in the department of Informatics at the University of Oslo.
+Since 1995, he has been a professor of signal processing in the Department of Informatics at the University of Oslo.
 In 2002, he was elected a member of the Norwegian Academy of Technological Sciences, and in 2009, he became an adjunct professor in the Norwegian University of Science and Technology (NTNU) Medical Imaging Lab, Trondheim.

### 2013-03-20 HRUIonGPU/jrnl_article.tex
Added gfx for uffc article on gpu capon

@@
 \usepackage{amsfonts}
+\usepackage{color}
 
@@
 \newcommand{\R}{$\mat{\hat{R}}$ }
+
+%\newcommand{\Red}[1]{{\color{red}#1}}
+\newcommand\multimedia[1]{\textbf{{\color{red}#1}}}
 
@@
 Austeng, \IEEEmembership{Member,~IEEE,} and S.
-Holm, \IEEEmembership{Senior Member,~IEEE,}% <-this % stops a space
-\thanks{Manuscript received ...; accepted ...}
-\thanks{Jon Petter \AA{}sen and Sverre Holm is with the Medical Imaging Lab (MI-Lab) at the Norwegian University of Sience and Technology, Trondheim, Norway.
+Holm, \IEEEmembership{Senior Member,~IEEE,} %<-this % stops a space
+%\thanks{Manuscript received ...; accepted ...}
+\thanks{Jon Petter \AA{}sen and Sverre Holm is with the Medical Imaging Lab (MI-Lab) at the Norwegian University of Science and Technology, Trondheim, Norway.
 Jo Inge Buskenes, Carl-Inge Colombo Nilsen, Andreas Austeng and Sverre Holm is with the Department of Informatics, University of Oslo, Oslo, Norway.
@@
 To reach this performance we make use of the parallel processing power found in modern Graphics Processing Units (GPUs), combined with beamspace processing to reduce the computational complexity as the number of array elements increases.
+
 For a three dimensional beamspace, processing rates supporting real-time cardiac ultrasound imaging are shown to be possible, meaning that processing rates higher than the image acquisition rate are obtained for a wide range of parameters.
 Image quality is investigated in videos of both a simulated phantom and an \textit{in-vivo} cardiac dataset.
-These results show that Capon beamforming is feasible for cardiac ultrasound imaging, providing images with sharper structures and no local shift-variance effects, both in element- and beamspace.
+These results show that Capon beamforming is feasible for cardiac ultrasound imaging, providing images with improved lateral resolution and no local shift-variance effects, both in element- and beamspace.
 \end{abstract}
@@
 \begin{IEEEkeywords}
-Adaptive Beamforming, Capon Beamforming, Minimum Variance Beamforming, Graphics Processing Unit, GPGPU, Parallel Computing, Ultrasound Imaging.
+Adaptive beamforming, Capon beamforming, Covariance matrix, GPGPU, Graphics processing unit, Minimum variance beamforming, Parallel programming, Ultrasound imaging.
 \end{IEEEkeywords}
@@
 
-To reach this goal we start by showing how the cubic complexity with $L$ and linear complexity with $N_K$ can be reduced by realizing that calculations in (\ref{eq:R}) overlap both across subarrays and in time.
+In \cite{asen2012, Buskenes} we demonstrated how the cubic complexity with $L$ and linear complexity with $N_K$ can be reduced by realizing that calculations in (\ref{eq:R}) overlap both across subarrays and in time.
 Element $(i,j)$ of $\mat{\breve{R}}[n]$ can be calculated from element $(i-1, j-1), \text{for } i,j \in [1, L-1]$ in a sliding window maner as
@@
 \end{align}
-In time we have that
+Along the time dimension we also find that
 \begin{align}
@@
 %- \mat{\breve{R}}[n-K-1] + \mat{\breve{R}}[n-K].
-Sliding windows across subarrays and in time can be realized simultaneously if $\mat{\hat{R}}$, averaged over subarrays, is calculated from an $N_K$-average in time of the outer products $\vec{x}[n]\vec{x}[n]^H$ \cite{Buskenes}.
+Sliding windows across subarrays and in time can be realized simultaneously if $\mat{\hat{R}}$, averaged over subarrays, is calculated from an $N_K$-average in time of the outer products $\vec{x}[n]\vec{x}[n]^H$.
 However, since $N_K$ typically is small, and the GPU has a limited amount of local memory available per thread, we focus on subarrays only.
@@
 \end{align} 
-The steering vectors in $\mat{B}$, the so-called Butler or normalized DFT matrix, determine each beam's direction in space.
+The steering vectors in $\mat{B}$, the so-called Butler or the normalized discrete fourier transform matrix, determine each beam's direction in space.
 Beams where no interference is present can therefore be removed by simply removing rows from $\mat{B}$.
 
-For a focused system like cardiac ultrasound imaging, where the received signal is concentrated in a narrow sector around the transmit direction, a small percentage of beams contains almost all received energy.
-Nilsen et al.
-concluded that as little as three beams around (and including) the transmit direction were adequate to produce results comparable to applying Capon beamforming in element space (Section \ref{sec:meth}).
+For a focused system like cardiac ultrasound imaging, where the received signal is concentrated in a narrow sector around the transmit direction, a small percentage of the beams contains almost all the received energy.
+Nilsen and Hafizovic concluded that as little as three beams around (and including) the transmit direction were adequate to produce results comparable to applying Capon beamforming in element space (Section \ref{sec:meth}).
 Hence, for  equally pitched arrays the inversion steps is found to be constant with respect to the number of array elements.
@@
 To maximize performance, large adjustments to the processing pipeline described in Section \ref{sec:meth} is required.
+In the remaining sections, the methods described in this section and Section \ref{sec:meth} will be referred to as beamspace Capon (BS-Capon) and element-space Capon (ES-Capon) respectively.
+The name Capon will refer to both methods.
 
@@
 \begin{figure*}[!t]
-\centerline{\subfloat[]{\includegraphics[width=3.5in]{gfx/benchmark_bar_bs_M=64_Nx=80_Ny=540_Nb=3.png} \label{fig:benchBS64}}
+\centerline{\subfloat[]{\includegraphics[width=3.45in]{gfx/benchmark_bar_bs_M=64_Nx=80_Ny=540_Nb=3.png} \label{fig:benchBS64}}
 \hfil
-\subfloat[]{\includegraphics[width=3.5in]{gfx/benchmark_bar_bs_M=96_Nx=120_Ny=540_Nb=3.png}%
+\subfloat[]{\includegraphics[width=3.45in]{gfx/benchmark_bar_bs_M=96_Nx=120_Ny=540_Nb=3.png}%
 \label{fig:benchBS96}}}
@@
 Assuming a system bandwidth of 80\% we need a complex sampling frequency of 2.7 MHz, and when imaging down to a depth of 15.4 cm we have approximately 540 samples per range line.
-To correctly sample a $70\degree$ sector with 64- and 96-element arrays ($\lambda/2$ pitch), we need at least 80 and 120 receive beams respectively.
+To sample a $70\degree$ sector with 64- and 96-element arrays ($d=\lambda/2$ pitch) at twice the Rayleigh criterion ($\lambda/(M*d)$), we need at approximately 80 and 120 receive beams respectively.
 The presented execution times do not include transfer of data from CPU- to the GPU-side, since data are assumed to be pre-delayed and available in global GPU memory.
@@
 
-A benchmark of the element-space Capon, as described in Section \ref{sec:meth}, is presented in Fig.
+A benchmark of the ES-Capon is presented in Fig.
 \ref{fig:bench}.
@@
 For each subarray size, the total execution time is split into one bar for each of the three steps described in Section \ref{sec:meth}, and the calculation of $\mat{\hat{R}}$ is presented for three different values of $K$.
-For $M=64$, $L=32$ and $K=2$ (Fig.
-\ref{fig:bench}), a typical setup for high-quality cardiac imaging, processing throughput is 7 frames per second.
+For $M=64$, $L=32$ and $K=2$, a typical setup for high-quality cardiac imaging, processing throughput is 7 frames per second.
 
 In Fig.
-\ref{fig:benchBS} we present similar benchmarks for the beamspace version of the Capon beamformer introduced in Section \ref{sec:bs}.
-Benchmarks are based on 64- and 96-element phased arrays, using 80 and 120 transmit beams.
+\ref{fig:benchBS} we present similar benchmarks for the BS-Capon beamformer.
+Benchmarks are based on 64- and 96-element phased arrays, using 80 and 120 receive beams.
 The figure only presents results with use of temporal averaging ($K=2$).
@@
 \begin{figure*}[!t]
-\centerline{\subfloat[]{\includegraphics[width=2.3in]{gfx/das.png} \label{fig:phantomDAS}}
+\centerline{
+\subfloat[]{ \includegraphics[width=2.2in]{gfx/das.png} \label{fig:phantomDAS} }
 \hfil
-\subfloat[]{\includegraphics[width=2.3in]{gfx/capon.png} \label{fig:phantomMV}}
+\subfloat[]{ \includegraphics[width=2.2in]{gfx/capon.png} \label{fig:phantomMV} }
 \hfil
-\subfloat[]{\includegraphics[width=2.3in]{gfx/capon_bs.png} \label{fig:phantomBS}}}
+\subfloat[]{ 
+\includegraphics[width=2.2in]{gfx/capon_bs.png} \label{fig:phantomBS} 
+\hfil
+\includegraphics[width=0.35in]{gfx/colorBarPhantom2.png}
+}
+}
 \caption{Simulated phantom containing one bright (20 dB) and one dark (-20 dB) circular area and two point targets (30 dB) in speckle (0 dB).
@@
 Dynamic range is -20 dB to 20 dB with mean value represented as 0 dB.
-The image is the 15th frame in the attached video (ADD VIDEO FILENAME IN RED).
+The image is the 15th frame in the attached videos.
 a) Delay-and-sum with uniform weights.
-b) Capon ($L=32, K=2, d=0.1$) c) Beamspace Capon ($L=32, K=2, N_b=3, d=0.01$).}
+b) Element-space Capon ($L=32, K=2, d=0.1$) \multimedia{Media-Movie 1} c) Beamspace Capon ($L=32, K=2, N_b=3, d=0.01$) \multimedia{Media-Movie 2}.}
 \label{fig:phantom}
@@
 \begin{figure*}[!t]
-\centerline{\subfloat[]{\includegraphics[width=2.3in]{gfx/das_invivo.png} \label{fig:invivoDAS}}
+\centerline{
+\subfloat[]{\includegraphics[width=2.2in]{gfx/das_invivo.png} \label{fig:invivoDAS}}
 \hfil
-\subfloat[]{\includegraphics[width=2.3in]{gfx/capon_invivo.png} \label{fig:invivoMV}}
+\subfloat[]{\includegraphics[width=2.2in]{gfx/capon_invivo.png} \label{fig:invivoMV}}
 \hfil
-\subfloat[]{\includegraphics[width=2.3in]{gfx/capon_bs_invivo.png} \label{fig:invivoBS}}}
+\subfloat[]{
+\includegraphics[width=2.2in]{gfx/capon_bs_invivo.png} \label{fig:invivoBS}}
+\hfil
+\includegraphics[width=0.4in]{gfx/colorBarInvivo2.png}
+}
 \caption{\textit{In-vivo} harmonic cardiac ultrasound image acquired using a 3.5 MHz, 64 elements phased array.
@@
 Dynamic range is -20 dB to 20 dB with mean value represented as 0 dB.
-The image is the first frame in the attached video (ADD VIDEO FILENAME IN RED).
-a) Delay-and-sum with uniform weights.
-b) Capon ($L=32, K=2, d=0.1$) c) Beamspace Capon ($L=32, K=2, N_b=3, d=0.01$).}
+The image is the first frame in the attached videos.
+a) Delay-and-sum with uniform weights \multimedia{Media-Movie 3}.
+b) Element-space Capon ($L=32, K=2, d=0.1$) \multimedia{Media-Movie 4} c) Beamspace Capon ($L=32, K=2, N_b=3, d=0.01$) \multimedia{Media-Movie 5}.}
 \label{fig:invivo}
@@
 
-To investigate the image quality of both the Capon and BS Capon beamformer a phantom was simulated using Field II \cite{Jensen1992, Jensen1996a} (Fig.
+To investigate the image quality of both the ES and BS-Capon beamformer a phantom was simulated using Field II \cite{Jensen1992, Jensen1996a} (Fig.
 \ref{fig:phantom}).
@@
 The transducer was excited with a 3.4 MHz and $1.5\lambda$ long pulse with transmit focus at 90 mm.
-The phantom contains two circular areas, one bright at 20 dB and one dark at -20 dB, and two point targets at 30 dB initially located at 90 mm depth.
+The phantom contains two circular areas, one bright at 20 dB and one dark at -20 dB, and two point targets at 30 dB, located at 90 mm depth.
 Speckle level is at 0 dB, covering an area from $\pm90\degree$ in azimuth.
-The resulting image has been processed both using DAS with uniform weights, Capon and BS Capon.
+The resulting image has been processed both using DAS with uniform weights, ES and BS-Capon.
 The presented image is the first frame in a series of images where motion is imposed on both circles, point targets and the probe.
-The entire simulation is presented in an attached video.
+The entire simulation is presented in three attached videos \multimedia{Media-Movie 1 and 2}.
 
 Fig.
-\ref{fig:invivo} presents results of applying Capon and BS Capon on an \textit{in-vivo} image of the left ventricle.
+\ref{fig:invivo} presents results of applying ES and BS-Capon on an \textit{in-vivo} image of the left ventricle.
 The image was acquired using a 64-element and $0.65\lambda$ pitched phased array in harmonic mode (1.7 MHz at transmit and 3.4 MHz at receive).
-The image is extracted from a cardiac cycle, that can be viewed in its full length in an attached video.
+The image is extracted from a cardiac cycle, that can be viewed in its full length in the attached videos \multimedia{Media-Movie 3, 4 and 5} .
 It is included to show that Capon beamforming in both element- and beamspace is feasible for cardiac imaging, not just in simulations.
@@
 \ref{fig:slices}.
-For the phantom, the cut goes through the two point targets at 90 mm constant depth, and for the cardiac image it goes through the tip of the mitral valve at 96 mm constant depth.
-
-Both data sets have been processed with the following Capon beamformer settings: $L = 32$, $K = 2$, and $d=0.1$.
-These settings have previously been shown to provide high resolution while maintaining DAS-like speckle\cite{Synnevag2007a}.
-For the beamspace processing the following parameters have been used: $L=32$, $K=2$, $d=0.01$, and $N_b=3$.
-For the simulated phantom, 16x lateral oversampling on transmit has been used, in order to avoid under sampling of the two bright point scatterers as they move.
-The cardiac recording has no lateral oversampling on receive.
+For the phantom, the cut goes through the two point targets at 90 mm constant radius, and for the cardiac image it goes through the tip of the mitral valve at 96 mm constant radius.
+
+Both data sets have been processed with the following settings for the ES-Capon beamformer: $L = 32$, $K = 2$, and $d=0.1$.
+These settings have previously been shown to provide high resolution while maintaining DAS-like speckle \cite{Synnevag2007a}.
+For BS-Capon the following parameters have been used: $L=32$, $K=2$, $d=0.01$, and $N_b=3$.
+For the simulated phantom, 16x lateral oversampling on transmit has been performed in order to avoid under sampling of the two bright point scatterers as they move.
+The cardiac recording of a 70\degree{} sector has a minimal lateral sampling of 98 receive beams.
 
@@
 So to gain an  increase in processing throughput we can reduce $L$ in element space or $N_b$ in beamspace.
-The question is then, how does this impact image quality?
+The question is then, how does this impact the lateral resolution?
 In Fig.
 \ref{fig:speed_res_trade}, the minimum distance between the two point targets in Fig.
-\ref{fig:phantom} (while preserving a 6 dB saddle point in between the two) is plotted against $L$ (when processed with Capon) and $N_b$ (when processed with BS Capon).
+\ref{fig:phantom} (while preserving a 6 dB saddle point in between the two) is plotted against $L$ (when processed with ES-Capon) and $N_b$ (when processed with BS-Capon).
 For the result in Fig.
-\ref{fig:speed_res_trade}, $d=0.01$ has been used on both methods.
-
-For high values of $L$ and $N_b$ the two methods are equal, but for lower values, BS Capon provides improved resolution compared to using small subarrays.
-We see that the resolution for BS Capon gradually decreases due to an automatic increase in diagonal loading (See end of Section \ref{sec:bs}).
-If we compensated by reducing $d$, BS Capon will provide resolution almost equal to Capon (with $L=M/2$) for all values except $N_b=1$ \cite{Nilsen2009}.
+\ref{fig:speed_res_trade}, $d=0.01$ has been used on both methods in order to obtain the same resolution for $L=N_b=M/2$.
+This is considered a small value for ES-Capon and a normal value for BS-Capon (cf.
+Section \ref{sec:bs}), hence the resolution for ES-Capon is here better than what is presented in Fig.
+\ref{fig:phantom}.
+The precision of the presented measurements is 0.12 mm, dictated by the simulated frame rate, 25 frames per second (FPS), and the relative speed of the two point scatterers, 3 mm/s.
+
+For high values of $L$ and $N_b$ the two methods are equal, but for lower values, BS-Capon provides improved resolution compared to using small subarrays.
+We see that the resolution for BS-Capon gradually decreases due to an automatic increase in diagonal loading (cf.
+end of Section \ref{sec:bs}).
+If we compensated by reducing $d$, BS-Capon will provide resolution equal to ES-Capon (with $L=M/2$) for all values of $N_b$ except one \cite{Nilsen2009}.
 For $N_b=1$ the resolution is worse than for DAS because of a triangular apodization caused by subarray averaging.
-In element space, resolution decays rapidly as a result of a reduced effective adaptive aperture.
-For $N_b=3$, BS Capon still provides approximately 1 mm improved resolution over DAS.
-However, note that resolution of the Capon beamformer strongly depends on image SNR and the choice of parameters.
+
+In element space, resolution decays rapidly as a result of a reduced adaptive aperture, and for $L = N_b=3$, BS-Capon provides almost 1 mm improved resolution over DAS and ES-Capon.
+However, note that resolution of the Capon beamformers strongly depends on image SNR and the choice of parameters.
 Thus, different SNR scenarios and choices of parameters will lead to different results.
-It is also worth noting that Capon beamforming in this example performs worse than DAS for small subarrays.
-For small values of $L$, the amount of robustification is too high in order to perfectly steer a zero at each point target.
-The Capon beamformer therefore produce a result worse than DAS.
+It is also worth noting that ES-Capon beamforming in this example performs worse than DAS for small values of $L$.
+For these small subarrays, the amount of robustification is too high in order to correctly steer a zero at each point target or to accurately micro-steer the main lobe.
+Subarray averaging further impose a triangular or trapezoid apodization function, which leads to worse resolution if the introduced adaptivity can not compensate for the increase in main lobe width.
+For $L=1$ the ES-Capon is equal to DAS.
 
@@
 In this paper we have presented a parallel implementation of the Capon beamformer in both element- and beamspace for high-resolution cardiac ultrasound imaging.
-The results show that it is now possible to apply BS Capon in real-time in cardiac ultrasound imaging by utilizing the power of GPUs.
+The results show that it is now possible to apply BS-Capon in real-time in cardiac ultrasound imaging by utilizing the power of GPUs.
 According to Fig.
@@
 $L=M/2$).
-However, these extra lines might be needed for increased sampling instead of increased frame rate when using the Capon beamformers.
+However, these extra lines might be needed for increased sampling instead of increased frame rate when using the Capon beamformers in high SNR situations (Fig \ref{fig:phantom}).
 If parallel receive lines are used to obtain denser sampling, acquisition time remains the same but processing times are increased with a factor equal to the amount of oversampling.
 If parallel receive lines are used for increased frame rate, the acquisition time is lowered, and the processing time remains constant.
-Hence, four parallel receive lines are not possible to process in real-time using BS Capon (with $N_b=3$) in combination with our system configuration.
-
-For Capon in element space, processing rates for high quality settings are still too high to support real-time cardiac ultrasound imaging.
+Hence, four parallel receive lines are not possible to process in real-time using BS-Capon (with $N_b=3$) in combination with our single GPU.
+
+For ES-Capon, processing rates for high quality settings are still too high to support real-time cardiac ultrasound imaging.
 From Fig.
-\ref{fig:bench} we see that for $L = M/2 = 32$ and $K=1$, the total execution time is 140 ms (7 fps).
+\ref{fig:bench} we see that for $L = M/2 = 32$ and $K=1$, the total execution time is 140 ms (7 FPS).
 Taking into account that we use complex data, but less samples per image, these results are similar to the performance reported by Chen et al.
@@
 
-A major part of the total execution time for element-space Capon comes from solving the system of linear equations.
+A major part of the total execution time for ES-Capon comes from solving the system of linear equations.
 This limitation is targeted by reducing the sample covariance matrix size in beamspace, and as shown in Fig.
@@
 The time it takes to construct all covariance matrices is also reduced, but for few beams this is now the major contributor to the total execution time.
-If real-time is defined as 10 fps \cite{Chen2011} almost all execution times in both Fig.
+If real-time is defined as 10 FPS \cite{Chen2011} almost all execution times in both Fig.
 \ref{fig:bench} and \ref{fig:benchBS} are real-time.
 More natural for cardiac ultrasound imaging is to define real-time as the time it takes to acquire the underlying image.
-With this requirement the images formed using a 64 and 96 element array have a real-time requirement of 62.5 and 41.7 fps respectively using single line acquisition.
+With this requirement the images formed using a 64 and 96 element array have a real-time requirement of 62.5 and 41.7 FPS respectively using single line acquisition.
 Each benchmark plot has a line indicating this real-time requirement.
@@
 
-This paper further shows how to obtain a parallel formulation of the Capon and BS Capon beamformer, and how to implement them in a GPU framework.
+This paper further shows how to obtain a parallel formulation of the ES-Capon and BS-Capon beamformer, and how to implement them in a GPU framework.
 The per-pixel granularity suggests that the beamformer is easily dividable across multiple GPUs.
@@
 \ref{fig:phantom}, and the attached video of the simulation, that Capon beamforming provides improved lateral resolution of moving point targets both in element- and in a reduced beam-space ($N_b=3$).
-Our preliminary finding is that the Capon beamformer is shift invariant (see video of the simulation) as long as the lateral sampling is increased enough.
+Our preliminary finding is that the Capon beamformers are shift invariant as long as the lateral sampling is increased to a curtain level \multimedia{Media-Movie 1 and 2}.
 How large this oversampling factor needs to be and how the oversampling should be conducted is outside the scope of this paper.
-No major improvement in contrast is visually observed in these images, other than slightly rounder circles due to mainlobe micro steering.
+No major improvement in contrast is visually observed in the presented images images and videos, other than slightly rounder circles due to mainlobe micro steering.
 In the cardiac recording (Fig.
-\ref{fig:invivo}), structures are thinner and better resolved when the Capon and BS Capon beamformer are applied.
+\ref{fig:invivo}), structures are thinner and better resolved when the Capon beamformers are applied.
 However we observe a loss in contrast, especially in the septum (Fig.
@@
 
-Nilsen et.
-al suggested to do automatic selection of beams in beamspace, hence picking the beams containing most energy.
+Nilsen and Hafizovic suggested to do automatic selection of beams in beamspace, hence picking the beams containing most energy.
 In this paper the center beams are always picked.
-This seems to work fine, but an adaptive selection of beams should be implemented if interference at high angles is shown to contaminate the image.
+This seems to work fine for focused systems, but an adaptive selection of beams should be implemented if interference at high angles is shown to contaminate the image.
 
@@
 In this paper a parallel formulation and implementation of the element- and beamspace Capon beamformer have been presented, and
-both beamformers have successfully been implemented on a GPU.
+both beamformers have successfully been implemented on the GPU.
 For a three dimensional beamspace, processing rates supporting real-time cardiac ultrasound imaging was shown to be possible, meaning that processing rates higher than the image acquisition rate are obtained for a wide range of parameters.
 Image quality was investigated both in a simulated phantom and an \textit{in-vivo} cardiac dataset.
-These results showed that Capon beamforming is feasible for cardiac ultrasound imaging, providing images with sharper structures and no local shift-variance effects, both in element- and beamspace.
+These results showed that Capon beamforming is feasible for cardiac ultrasound imaging, providing images with improved lateral resolution and no local shift-variance effects, both in element- and beamspace.
 
@@
 % use section* for acknowledgement
-%\section*{Acknowledgment}
-
-%The authors would like to thank...
+\section*{Acknowledgment}
+The authors would like to thank Dr Kjell Kristoffersen, GE Vingmed Ultrasound, for valuable comments.
 
@@
 \begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{photos/jon_petter2.eps}}]{Jon Petter \AA{}sen}
-(S'12) was born in Porsgrunn, Norway in 1986.
+(S12) was born in Porsgrunn, Norway in 1986.
 He received the B.Sc.

### 2013-03-24 HRUIonGPU/jrnl_article.tex
Moved first part of beamspace section to background in tuffc article

@@
 
-\author{J.
-P.
-\AA{}sen, \IEEEmembership{Student Member,~IEEE,} J.
-I.
-Buskenes, C.-I.
-C.
-Nilsen, \IEEEmembership{Member,~IEEE,} A.
-Austeng, \IEEEmembership{Member,~IEEE,} and S.
-Holm, \IEEEmembership{Senior Member,~IEEE,} %<-this % stops a space
-%\thanks{Manuscript received ...; accepted ...}
-\thanks{Jon Petter \AA{}sen and Sverre Holm is with the Medical Imaging Lab (MI-Lab) at the Norwegian University of Science and Technology, Trondheim, Norway.
-Jo Inge Buskenes, Carl-Inge Colombo Nilsen, Andreas Austeng and Sverre Holm is with the Department of Informatics, University of Oslo, Oslo, Norway.
-(e-mail: jon.p.asen@ntnu.no)}
+\author{Jon Petter \AA{}sen, \IEEEmembership{Student Member,~IEEE,} Jo Inge Buskenes, Carl-Inge Colombo Nilsen, \IEEEmembership{Member,~IEEE,} Andreas Austeng, \IEEEmembership{Member,~IEEE,} and Sverre Holm, \IEEEmembership{Senior Member,~IEEE,} %<-this % stops a space
+\thanks{Manuscript received ...; accepted ...}
+\thanks{J. P. \AA{}sen and S. Holm is with the Medical Imaging Lab (MI-Lab) at the Norwegian University of Science and Technology, Trondheim, Norway.}
+\thanks{J. I. Buskenes, C.-I. C. Nilsen, A. Austeng and S. Holm is with the Department of Informatics, University of Oslo, Oslo, Norway. (e-mail: jon.p.asen@ntnu.no)} 
 }
@@
 Capon beamforming is associated with a high computational complexity, which limits its use as a real-time method in many applications.
-In this paper we present an implementation of the Capon beamformer exhibiting real-time performance when applied in a typical cardiac ultrasound imaging setting.
-To reach this performance we make use of the parallel processing power found in modern Graphics Processing Units (GPUs), combined with beamspace processing to reduce the computational complexity as the number of array elements increases.
-
-For a three dimensional beamspace, processing rates supporting real-time cardiac ultrasound imaging are shown to be possible, meaning that processing rates higher than the image acquisition rate are obtained for a wide range of parameters.
-Image quality is investigated in videos of both a simulated phantom and an \textit{in-vivo} cardiac dataset.
-These results show that Capon beamforming is feasible for cardiac ultrasound imaging, providing images with improved lateral resolution and no local shift-variance effects, both in element- and beamspace.
+In this paper we present an implementation of the Capon beamformer that exhibits real-time performance when applied in a typical cardiac ultrasound imaging setting.
+To achieve this performance we make use of the parallel processing power found in modern Graphics Processing Units (GPUs), combined with beamspace processing to reduce the computational complexity as the number of array elements increases.
+
+For a three dimensional beamspace we show that processing rates supporting real-time cardiac ultrasound imaging are possible, meaning that images can be processed faster than the image acquisition rate for a wide range of parameters.
+Image quality is investigated in videos of both a simulated phantom, and in an \textit{in-vivo} cardiac dataset.
+These results show that Capon beamforming is feasible for cardiac ultrasound imaging, providing images with improved lateral resolution and no local shift-variance effects, both in element-space and beamspace.
 \end{abstract}
@@
 %\IEEEPARstart{I}{ntroduction} to Capon adaptive beamforming, ultrasound imaging and GPU/CUDA.
-\IEEEPARstart{C}{apon} beamforming \cite{Capon1969} applied to medical ultrasound imaging has recently shown promising results in several studies \cite{Synnevag2007, Austeng2008, Vignon2008, Viola, Mehdizadeh2012}.
+\IEEEPARstart{C}{apon} beamforming \cite{Capon1969} has recently been applied to medical ultrasound imaging with promising results in several studies \cite{Synnevag2007, Austeng2008, Vignon2008, Viola, Mehdizadeh2012}.
 Most of this work focuses on the improved lateral resolution and contrast obtained with the Capon beamformer.
 However, some interesting trade-offs have also been introduced.
-In \cite{Synnevag2009}, Synnev\aa{}g et al.
-explains how image resolution can be maintained at higher framerates, with smaller probes, or for deeper penetration.
+In \cite{Synnevag2009}, Synnev\aa{}g \textit{et al.} explains how image resolution can be maintained at higher framerates, with smaller probes, or for deeper penetration.
 Despite these benefits, widespread adoption of the Capon beamformer has not been seen due to its high computational complexity \cite{So2011}.
 In this paper we present an implementation of the Capon beamformer exhibiting real-time performance when applied in a typical cardiac ultrasound imaging setting.
-To reach this performance we make use of the parallel processing power found in modern Graphics Processing Units (GPUs) (found in almost all PCs), combined with novel methods to reduce the computational complexity as the number of array elements increases.
+To achieve this performance we make use of the parallel processing power found in modern Graphics Processing Units (GPUs) (found in almost all PCs), combined with novel methods to reduce the computational complexity as the number of array elements increases.
 
@@
 For medical ultrasound, Chen et al.
-\cite{Chen2011a, Chen2011, Chen} implemented a Capon beamformer on the GPU that produced real apodization weights based on real data.
-However, the Capon beamformer is meant to work on complex data.
-Under the constraint of real data, the Capon beamformer is not allowed to micro-steer the main lobe, and therefore its full potential is not reached.
-Micro-steering of the main lobe is an important property of the Capon beamformer in order for it to improve structural details and edge definitions in an ultrasound image \cite[Fig 9.]{Synnevag2009}.
-Complex arithmetic also requires more operations per sample, but the sampling frequency can also be reduced.
-The number of samples needed per image is therefore minimized.
-All results in this paper are reported with the use of baseband (or I/Q) data, where the sampling frequency is close to the system bandwidth.
+\cite{Chen2011a, Chen2011, Chen} implemented a Capon beamformer on the GPU that produces real apodization weights based on real data.
+However, the Capon beamformer is meant to process complex data.
+Under the constraint of using real data, the Capon beamformer is not allowed to micro-steer the main lobe, and therefore its full potential is not reached.
+Micro-steering of the main lobe is an important property of the Capon beamformer in order for it to improve structural details and edge definitions in an ultrasound image \cite[Fig~9.]{Synnevag2009}.
+Complex arithmetic leads to more operations per sample, but also allows the sampling frequency to be reduced.
+The number of samples needed per image can therefore be minimized.
+All results in this paper are reported with the use of baseband (or I/Q) data, where the sampling frequency is near the system bandwidth.
 
@@
 % denne må utelates i IUS-artikkelen
-In recent work \cite{asen2012, Buskenes} we have applied a GPU-based Capon beamformer for both active sonar and for cardiac ultrasound imaging.
+In recent work \cite{asen2012, Buskenes, Buskenes2013} we have applied a GPU-based Capon beamformer for both active sonar imaging and for cardiac ultrasound imaging.
 Sonar has a less strict real-time requirement than medical ultrasound imaging, which means that most array configurations can be handled in real time using the GPU.
-Medical ultrasound imaging, and in particular cardiac ultrasound imaging, demands frame rates at a level where even a high-end GPU can not do the processing required for a straightforward implementation of the Capon beamformer.
+Capon beamforming in medical ultrasound imaging, and in particular cardiac ultrasound imaging, demands frame rates that not even a high-end GPU can provide.
 Since the execution time is cubic with respect to the number of array elements, we need methods that reduce the computational complexity for large arrays.
-Various approaches to low-complexity Capon beamforming have therefore recently been proposed in the literature \cite{Synnevag2011, Asl2012, Jensen2012, Kim}.
+Various approaches to low-complexity Capon beamforming for medical ultrasound imaging have therefore been proposed in the literature \cite{Synnevag2011, Asl2012, Jensen2012, Kim}.
 In this paper we will make use of a beamspace transformation described by Nilsen and Hafizovic \cite{Nilsen2009}, in order to maintain high framerates as the number of elements increases.
-The beamspace transformation matches well with the focused transmit beams used in cardiac ultrasound imaging.
+%The beamspace transformation matches well with the focused transmit beams used in cardiac ultrasound imaging. 
 %(Referer til JOE-artikkelen, få med beskrivelse av hva som blir anerledes for ultralyd.)
 
-In the next section we present background information on the Capon beamformer and the GPU computing model.
-In Section \ref{sec:meth} we go in detail on the parallel implementation of the Capon beamformer, before the beamspace method, and its GPU implementation, is presented in Section \ref{sec:bs}.
+In the next section we present background information on Capon beamforming in both element-space and beamspace and the GPU computing model.
+In Section \ref{sec:meth} we recap our parallel implementation of the Capon beamformer, before the beamspace method, and its GPU implementation, is presented in Section \ref{sec:bs}.
 Benchmark results and resulting images are presented in Section \ref{sec:bench} and \ref{sec:images}, together with a discussion of trading image quality for speed in Section \ref{sec:trade}.
@@
 
-\section{Background}
-
-\subsection{Capon Beamforming}
+\section{Background}\label{background}
+
+\subsection{Capon Beamforming}\label{sec:es-capon}
 
@@
 \end{align}
-The weight vector, or window, $\vec{w}$ is usually real, and is applied to trade resolution for a better signal-to-noise ratio (SNR).
+The weight vector, or window, $\vec{w}$ is usually real in DAS beamforming, and is applied to trade resolution for a better signal-to-noise ratio (SNR).
 Although different windows can be used in different areas of the images, $\vec{w}$ is usually predefined.
@@
 \ref{fig:mvbf}, the basic idea of an adaptive beamformer is to form a complex weight vector based on the received data.
-The Capon beamformer produces weights that minimize the output power while maintaining unity gain in the focus direction.
+The Capon beamformer produces weights that minimize the beamformer output power while maintaining unity gain in the focus direction.
 The result is that interference impinging from other directions is suppressed \cite{Synnevag2007}.
-It is worth noting that the micro-steering caused by the complex weights is minute compared to the initial steering and focusing step, therefore the beamformer also works well for broadband ultrasound signals.
+Micro-steering caused by the complex weights is minute compared to the initial steering and focusing step, therefore the beamformer also works well for broadband ultrasound signals.
 
@@
 
-The Capon beamformer, following the approach in \cite{Synnevag2009}, can be divided into three main steps: estimation of a sample covariance matrix, solving a system of linear equation, and calculation of the beamformer output.
+The Capon beamformer, following the approach in \cite{Synnevag2009}, can be divided into three main steps: estimation of a sample covariance matrix, solving a system of linear equations, and calculation of the beamformer output.
 The sample covariance matrix is estimated as
@@
 \end{align}
-where  $N_K = 2K + 1$ is the number of samples included in time, $\vec{x}_l = [x_l[n], \dotso, x_{l+L-1}[n]]$ is the $l\text{th}$ subarray of length $L$, and $N_L = M-L+1$ is the number of subarrays.
+where  $N_K = 2K + 1$ is the number of temporal samples included, $\vec{x}_l = [x_l[n], \dotso, x_{l+L-1}[n]]$ is the $l\text{th}$ subarray of length $L$, and $N_L = M-L+1$ is the number of subarrays.
 Note that the covariance matrix dimension is $L \times L$.
@@
 z[n] &= \frac{1}{N_L}\vec{w}[n]^H \sum_{l=0}^{N_L-1} \vec{x}_l[n] \label{eq:z_mv}\\
-&= 1/N_L \, (\vec{w}[n] *_{m} \vec{1}_{N_L})^H\vec{x}[n] \label{eq:z_mv2},
+&= 1/N_L \, (\vec{w}[n] * \vec{1}_{N_L})^H\vec{x}[n] \label{eq:z_mv2},
 \end{align}
@@
 %\end{align}
-where $*_m$ is a non-truncated convolution across elements.
+where $*$ is a non-truncated convolution across elements.
 Note how we can either choose to sum all subarrays (\ref{eq:z_mv}), or accumulate a weighting per element (\ref{eq:z_mv2}).
@@
 
+\subsection{Beamspace Processing}\label{bs-capon}
+Nilsen and Hafizovic \cite{Nilsen2009} have proposed to apply beamspace processing to reduce the required computations of the Capon beamformer when applied to medical ultrasound imaging.
+Beamspace is an alternative representation of the per-element data, transformed into a fan of beams covering the sector illuminated by transmit beams.
+The transformation is formally stated as
+\begin{align}
+\vec{x}_\text{BS}[n] &= \mat{B}\vec{x}[n], \ \ \ \ \ \ \ \ \ \ \ \ \ \, \mat{B} \in \mathbb{C}^{N,M} \label{eq:butler} \\ 
+[\mat{B}]_{p,q} &= \frac{1}{\sqrt{M}}e^{-j 2 \pi p q / M} \ \ \ N \le M.
+\end{align} 
+The steering vectors in $\mat{B}$, the so-called Butler or normalized DFT matrix, determine each beam's direction in space.
+Beams where no interference is present can therefore be removed by simply removing rows from $\mat{B}$.
+
+For a focused system like cardiac ultrasound imaging, where the received signal is concentrated in a narrow sector around the transmit direction, a small percentage of the beams contains almost all the received energy.
+Nilsen and Hafizovic concluded that as little as three beams around (and including) the transmit direction were adequate to produce results comparable to applying Capon beamforming in element space.
+Hence, for equally pitched arrays the inversion steps is found to be constant with respect to the number of array elements.
+
+The beamspace version of the Capon beamformer is derived by replacing the covariance matrix in (\ref{eq:w}) with the beamspace covariance matrix 
+\begin{align}
+\mat{\hat{R}}_\text{BS} = f(\vec{x}_\text{BS}) = f(\mat{B}\vec{x}), 
+\end{align}
+where $\mat{B} \in \mathbb{C}^{N_b,L}$, $N_b$ is the number of selected beams, and $f(\cdot)$ represent the covariance estimation process as defined in (\ref{eq:diag}).
+
+In the remaining sections, the methods described in this section and Section \ref{sec:es-capon} will be referred to as beamspace Capon (BS-Capon) and element-space Capon (ES-Capon) respectively.
+The name Capon will refer to both methods.
+
 \subsection{GPU Compute Model}
@@
 
-In this section we give a brief discussion of GPU computing, laying the basis for further discussion on parallel implementation strategies.
-As GPU framework, we have selected CUDA by Nvidia \cite{Nvidia2013}, however the challenges and proposed methods described in this paper are valid for other GPU framworks as well, like e.g.
+In this section we give a brief discussion of GPU computing, laying the basis for the discussions on parallel implementation strategies.
+The Capon beamformers have been implemented using the CUDA GPU framework by Nvidia \cite{Nvidia2013}, however the challenges and proposed methods described in this paper are valid for other GPU framworks as well, like e.g.
 OpenCL.
-The key point is that GPUs can accelerate large sets of small concurrent tasks, like beamforming a pixel in a large image, much better than a CPU.
-While the CPU has few cores and little logic to process complex threads, the GPU has many cores, but each core needs many (preferably simple) threads to keep it occupied and to achieve maximum throughput.
-
-The CUDA compute model is based on the execution of a kernel function over a large set of threads, where the kernel function describes the work to be done in each thread.
+Independent of framework, the key point is that GPUs can accelerate large sets of small concurrent tasks, like beamforming a pixel in a large image, much better than a CPU.
+While the CPU has few cores and lack logic to process complex threads, the GPU has many cores, but each core needs many (preferably simple) threads to keep it occupied and to thereby achieve maximum throughput.
+
+In CUDA programming a kernel function is executed over a large set of threads, where the kernel function describes the work to be done in each thread.
 As depicted in Fig.
@@
 Threads inside a block support low-cost synchronization if needed.
-From a CUDA perspective the GPU consist of one or more streaming multi-processors (SM), where each SM is capable of 32 or more concurrent multiplications and/or additions, depending on the architecture.
+From a CUDA perspective the GPU consist of one or more streaming multi-processors (SM), where each SM is capable of 32 or more concurrent multiplications or additions, or both, depending on the architecture.
 A group of blocks is scheduled to each SM until all blocks have been processed.
 
-An SM has a limited amount of local (near-core or shared) memory and registers.
-Therefore, to achieve maximum GPU occupancy (defined as having the maximum number of threads residing on each SM), each thread has to consume as few registers and as little local memory as possible.
+An SM has a limited amount of shared memory (or near-core cache) and registers.
+Therefore, to achieve maximum GPU occupancy (defined as having the maximum number of threads residing on each SM), each thread has to consume as few registers and as little shared memory as possible.
 Each thread should also perform enough instructions per transferred byte in order to hide memory latency.
@@
 
-\section{Parallel Capon}\label{sec:meth}
-If the beamformer in (\ref{eq:z_mv}) is analyzed in combination with (\ref{eq:b}) and (\ref{eq:diag}), we see that each calculation of a set of Capon weights is independent across the outputs $z[n]$.
+\section{Parallel ES-Capon}\label{sec:meth}
+In this section we will recap our parallel implementation of the ES-Capon beamformer as presented in \cite{asen2012, Buskenes}.
+This implementation was our starting point when the BS-Capon beamformer was implemented.
+
+If the beamformer in (\ref{eq:z_mv}) is analyzed in combination with the matrix equation in (\ref{eq:b}) and the covariance estimation process in (\ref{eq:diag}), we see that the calculation of a set of Capon weights is independent across the outputs $z[n]$.
 As shown in Fig.
@@
 The computational complexities of these three steps are $O(N_KN_LL^2)$, $O(L^3)$, and $O(L)$, respectively.
-As $M$ (and thereby $L$) increases, the first two steps quickly become unmanageable, and real-time performance can only be achieved by exploring novel solutions to reduce complexity.
+As $M$ (and thereby $L$) increases, the first two steps quickly become unmanageable, and real-time performance can only be achieved by exploring methods with reduced complexity.
 
@@
 
-Calculation of one element of \R is from (\ref{eq:R}) found to be independent of calculation of other elements.
+Calculation of one element of \R is from (\ref{eq:R}) found to be independent of calculation of the other elements.
 A natural level of granularity would then be to assign one thread to each element of $\mat{\hat{R}}$ \cite{Chen2011}.
@@
 Sliding windows across subarrays and in time can be realized simultaneously if $\mat{\hat{R}}$, averaged over subarrays, is calculated from an $N_K$-average in time of the outer products $\vec{x}[n]\vec{x}[n]^H$.
-However, since $N_K$ typically is small, and the GPU has a limited amount of local memory available per thread, we focus on subarrays only.
+However, since $N_K$ typically is small, and the GPU has a limited amount of shared memory available per thread, we focus on subarrays only.
 In that case, the complexity of calculating $\mat{\hat{R}}$ is reduced to $O(N_K(L(N_L-2)+2L^2))$.
 For $L=M/2$ this reduces to $O(N_KM^2)$.
+
 The sliding-window approach will cause dependence among the iterative calculations.
-The level of task granularity is thereby increased.
-When sliding across subarrays, calculations are still independent across time, but not along diagonals of $\mat{\hat{R}}$.
-One thread is therefore used per diagonal in the upper triangle of $\mat{\hat{R}}$, and for small values of $L$, several covariance matrices are calculated per thread block (Fig.
+The level of task granularity is therefore increased.
+Calculations are still independent across time, but not along diagonals of $\mat{\hat{R}}$.
+One thread is therefore assigned to each diagonal in the upper triangle of $\mat{\hat{R}}$.
+For small values of $L$, several covariance matrices are calculated per thread block (Fig.
 \ref{fig:gpulayout}).
 Maximum supported $L$ is now increased to the maximum number of threads per block (currently 1024).
-However, for small-$L$-high-$M$ configurations, memory consumption per thread will be too high to achieve full occupancy of the GPU.
+However, for small-$L$-high-$M$ configurations, memory consumption per thread will be too high to achieve full GPU occupancy.
 
@@
 For the adaptive diagonal loading factor in (\ref{eq:diag}), the trace of $\mat{\hat{R}}$ has to be known before diagonal elements are written back to global memory.
-Since one row for all covariance matrices is written to global memory per kernel iteration, the diagonal elements need to be held in a local memory until the trace is accumulated.
+Since one row for all covariance matrices is written to global memory per kernel iteration, the diagonal elements need to be held in a shared memory until the trace is accumulated.
 When that is done, scattered writes of the scaled diagonal elements to global memory has to be performed.
@@
 First the GPU needs to solve thousands of small systems in order to beat the low memory latency of the CPU.
-Second there is a range of system dimensions that does not fit well with the GPU architecture, where only a small amount of local memory and registers are available.
+Second there is a range of system dimensions that does not fit well with the GPU architecture, where only a small amount of shared memory and registers are available.
 Because of this, it has been proven hard to optimize a manageable set of solvers that provides speedups compared to the CPU in every case.
@@
 \subsection{Compute Beamformer Output}
-To calculate the beamformer output, (\ref{eq:z_mv}) has been implemented using one thread per element weight ($L$ threads per weight vector).
-After the solution vector $\vec{b}$ has been placed in local memory, one of the $L$ threads finds the sum of $\vec{b}$.
+The calculation of the beamformer output in (\ref{eq:z_mv}) has been implemented using one thread per element weight ($L$ threads per weight vector).
+After the solution vector $\vec{b}$ has been placed in shared memory, one of the $L$ threads finds the sum of $\vec{b}$.
 %Another option is to let several threads cooperate to form the sum using a binary reduction sum. However, in the following benchmarks the first approach has been used. 
@@
 
-The final output, as shown in (\ref{eq:z_mv}) and (\ref{eq:z_mv2}), can be formed by either reducing the data down to length $L$, or increasing the weight vector to length $M$.
+The final output can be formed by either reducing the data down to length $L$ as in (\ref{eq:z_mv}), or increasing the weight vector to length $M$ as in (\ref{eq:z_mv2}).
 Since the design is $L$ threads per output, the data is reduced to length $L$ by letting thread $r$ calculate
@@
 
-\section{Beamspace Processing}\label{sec:bs}
-Nilsen and Hafizovic \cite{Nilsen2009} have proposed to apply beamspace processing to further reduce the required computations of the Capon beamformer when applied to medical ultrasound imaging.
-Beamspace is an alternative representation of the per-element data, transformed into a fan of beams covering the sector illuminated by transmit beams.
-The transformation is formally stated as
-\begin{align}
-\vec{x}_\text{BS}[n] &= \mat{B}\vec{x}[n], \ \ \ \ \ \ \ \ \ \ \ \ \ \, \mat{B} \in \mathbb{C}^{N,M}\\
-[\mat{B}]_{p,q} &= \frac{1}{\sqrt{M}}e^{-j 2 \pi p q / M} \ \ \ N \le M.
-\end{align} 
-The steering vectors in $\mat{B}$, the so-called Butler or the normalized discrete fourier transform matrix, determine each beam's direction in space.
-Beams where no interference is present can therefore be removed by simply removing rows from $\mat{B}$.
-
-For a focused system like cardiac ultrasound imaging, where the received signal is concentrated in a narrow sector around the transmit direction, a small percentage of the beams contains almost all the received energy.
-Nilsen and Hafizovic concluded that as little as three beams around (and including) the transmit direction were adequate to produce results comparable to applying Capon beamforming in element space (Section \ref{sec:meth}).
-Hence, for  equally pitched arrays the inversion steps is found to be constant with respect to the number of array elements.
-
-The beamspace version of the Capon beamformer is derived by replacing the covariance matrix in (\ref{eq:w}) with the beamspace covariance matrix 
-\begin{align}
-\mat{\hat{R}}_\text{BS} = f(\vec{x}_\text{BS}) = f(\mat{B}\vec{x}), 
-\end{align}
-where $\mat{B} \in \mathbb{C}^{N_b,L}$, $N_b$ is the number of selected beams, and $f(\cdot)$ is as defined in (\ref{eq:diag}).
-If the transformation is applied before the covariance matrix is constructed, one transformation is performed per subarray, yielding subarrays that no longer overlap in space.
-The amount of input data are then changed from $M$ to $N_LN_b$ elements per sample, however all remaining steps will get lower computational complexity when $N_b < L$.
-
-Another option is to transform the element space covariance matrix to beamspace. 
+\section{Parallel BS-Capon}\label{sec:bs}
+In this section we will derive a parallel implementation of the BS-Capon beamformer, and comment on how it deviates from the implementation in Section \ref{sec:meth}.
+The three main steps described in Section \ref{sec:meth} are present here as well, but in addition we need to transform the per-element data to beamspace.
+This can be done using a FFT, but for few beams ($N_b < \log{L}$) a matrix multiply with the truncated Butler matrix is less costly.
+
+In order to reduce the complexity of the matrix inversion problem the transformation must be performed before or after the covariance matrix has been estimated.
+If the transformation is applied before, one transformation is needed per subarray.
+After this transformation the subarrays no longer overlap.
+The amount of input data are therefore changed from $M$ to $N_LN_b$ elements per sample, but all remaining steps will get lower computational complexity as long as $N_b < L$.
+
+The other option is to transform the element space covariance matrix to beamspace:
 \begin{align}
@@
 
+As mentioned, transforming subarrays into beamspace yields subarrays that no longer overlap.
+This means that the spatial overlap previously exploited when calculating the sample covariance matrix in element space is no longer available.
+On the other hand, since $N_b$ should be small ($N_b < 32$), one thread assigned to each element in $\mat{\hat{R}}_\text{BS}$ is now an option.
+A kernel similar to the one described in \cite{Chen2011} has therefore been implemented.
+In addition, for small $N_b$ values, several covariance matrices are calculated per thread block to achieve high GPU occupancy.
+
 %The implementation calculating covariance matrices therefore needs to be adjusted to handle non-overlapping subarrays.
@@
 %Derivation of sliding beamspace transformation. 
-Given a $M$ long data vector divided into $N_L$ subarrays, the beamspace transformation of the $l$th subarray for the $k$th beam is:
+We will now show how to minimize the amount of operations required for the beamspace transformation by utilizing the overlap between subarrays previously exploited when estimating the covariance matrix in element space.
+Given a $M$ long data vector divided into $N_L$ subarrays, the beamspace transformation of the $l$th subarray for the $k$th beam is from (\ref{eq:butler}) found to be:
 \begin{align}\label{eq:bs_formula}
-x_{\text{BS},l,k} = \frac{1}{\sqrt{L}}\sum_{p=i}^{l+L-1}x_p e^{-j2\pi k(p-l)/L} , k \in [0, N_b-1].
-\end{align}
-Manipulating this expression we find that for $l \in [1, L-1]$
+x_{\text{BS},l,k} &= \mat{B}_k\vec{x}_l \,\,\,\,\,\,\,\,\, k \in [0, N_b-1] \\
+&= \frac{1}{\sqrt{L}}\sum_{p=i}^{l+L-1}x_p e^{-j2\pi k(p-l)/L},
+\end{align}
+where $\mat{B}_k$ is the $k$th row of $\mat{B}$. Manipulating this expression we find that for $l \in [1, L-1]$
 \begin{align}\label{eq:sliding_bs}
@@
 
-Transforming subarrays into beamspace yields subarrays that no longer overlap.
-This means that the spatial overlap previously exploited when calculating the sample covariance matrix in element space is no longer available.
-On the other hand, since $N_b$ should be small ($N_b < 32$), one thread per element in $\mat{\hat{R}}_\text{BS}$ is now an option.
-A kernel similar to the one described in \cite{Chen2011} has therefore been implemented.
-In addition, for small $N_b$ values, several covariance matrices are calculated per thread block to achieve high occupancy of the GPU.
-
 After solving $\vec{b}_\text{BS} = \sqrt{L}\mat{\hat{R}}_\text{BS}^{-1}\vec{e}_1$, the beamspace weight vector is calculated as 
@@
 
-As presented in (\ref{eq:diag}), adaptive diagonal loading is performed by loading the sample covariance matrix with the average energy per element, hence $\text{trace}\{\mat{\hat{R}}\}/L$.
+As presented in (\ref{eq:diag}), adaptive diagonal loading is performed by loading the sample covariance matrix with the average energy per element, that is $\text{trace}\{\mat{\hat{R}}\}/L$.
 When data are transformed into beamspace, a high percentage of the total energy is mapped to a small subset of beams.
@@
 To maximize performance, large adjustments to the processing pipeline described in Section \ref{sec:meth} is required.
-In the remaining sections, the methods described in this section and Section \ref{sec:meth} will be referred to as beamspace Capon (BS-Capon) and element-space Capon (ES-Capon) respectively.
-The name Capon will refer to both methods.
 
@@
 \end{figure*}
-In this section we present benchmarks of the implementations presented in Sections \ref{sec:meth} and \ref{sec:bs} performed on a Nvidia Quadro 6000 graphics card (14 SMs, 1 TFlops theoretical single precision, 6 GB global memory, and 144 GB/s memory bandwidth).
+In this section we present benchmarks of the implementations presented in Sections \ref{sec:meth} and \ref{sec:bs} performed on a Nvidia Quadro 6000 graphics card (14 SMs, 1 TFlops theoretical single precision performance, 6 GB global memory, and 144 GB/s memory bandwidth).
 Problem sizes are calculated based on 64 and 96 element phased arrays with a 3.4 MHz center frequency.
-Assuming a system bandwidth of 80\% we need a complex sampling frequency of 2.7 MHz, and when imaging down to a depth of 15.4 cm we have approximately 540 samples per range line.
+Assuming a system bandwidth of 80\% (relative to the center frequency) we need a complex sampling frequency of 2.7 MHz, and when imaging down to a depth of 15.4 cm we have approximately 540 samples per range line.
 To sample a $70\degree$ sector with 64- and 96-element arrays ($d=\lambda/2$ pitch) at twice the Rayleigh criterion ($\lambda/(M*d)$), we need at approximately 80 and 120 receive beams respectively.
 The presented execution times do not include transfer of data from CPU- to the GPU-side, since data are assumed to be pre-delayed and available in global GPU memory.
-What is included are all calculations, memory transfers from global to local GPU memory, and local memory access.
+What is included are all calculations, memory transfers from global to shared GPU memory, and shared memory access.
 
@@
 a) Delay-and-sum with uniform weights.
-b) Element-space Capon ($L=32, K=2, d=0.1$) \multimedia{Media-Movie 1} c) Beamspace Capon ($L=32, K=2, N_b=3, d=0.01$) \multimedia{Media-Movie 2}.}
+b) Element-space Capon ($L=32, K=2, d=0.1$) \multimedia{Media-Movie 1}.
+c) Beamspace Capon ($L=32, K=2, N_b=3, d=0.01$) \multimedia{Media-Movie 2}.}
 \label{fig:phantom}
@@
 a) Delay-and-sum with uniform weights \multimedia{Media-Movie 3}.
-b) Element-space Capon ($L=32, K=2, d=0.1$) \multimedia{Media-Movie 4} c) Beamspace Capon ($L=32, K=2, N_b=3, d=0.01$) \multimedia{Media-Movie 5}.}
+b) Element-space Capon ($L=32, K=2, d=0.1$) \multimedia{Media-Movie 4}.
+c) Beamspace Capon ($L=32, K=2, N_b=3, d=0.01$) \multimedia{Media-Movie 5}.}
 \label{fig:invivo}
@@
 The image was acquired using a 64-element and $0.65\lambda$ pitched phased array in harmonic mode (1.7 MHz at transmit and 3.4 MHz at receive).
-The image is extracted from a cardiac cycle, that can be viewed in its full length in the attached videos \multimedia{Media-Movie 3, 4 and 5} .
-It is included to show that Capon beamforming in both element- and beamspace is feasible for cardiac imaging, not just in simulations.
+The image is extracted from a cardiac cycle, that can be viewed in its full length in the attached videos \multimedia{Media-Movie 3, 4 and 5}.
+It is included to show that ES-Capon and BS-Capon beamforming are both feasible for cardiac imaging, not just in simulations.
 
@@
 Both data sets have been processed with the following settings for the ES-Capon beamformer: $L = 32$, $K = 2$, and $d=0.1$.
-These settings have previously been shown to provide high resolution while maintaining DAS-like speckle \cite{Synnevag2007a}.
+These settings have previously been shown to provide improved lateral resolution while maintaining DAS-like speckle \cite{Synnevag2007a}.
 For BS-Capon the following parameters have been used: $L=32$, $K=2$, $d=0.01$, and $N_b=3$.
@@
 So to gain an  increase in processing throughput we can reduce $L$ in element space or $N_b$ in beamspace.
-The question is then, how does this impact the lateral resolution?
+In this section we investigate how adjusting $L$ for ES-Capon and $N_b$ for BS-Capon impacts the lateral resolution.
 In Fig.
@@
 end of Section \ref{sec:bs}).
-If we compensated by reducing $d$, BS-Capon will provide resolution equal to ES-Capon (with $L=M/2$) for all values of $N_b$ except one \cite{Nilsen2009}.
+If we compensate by reducing $d$, BS-Capon will provide resolution equal to ES-Capon (with $L=M/2$) for all values of $N_b$ except one \cite{Nilsen2009}.
 For $N_b=1$ the resolution is worse than for DAS because of a triangular apodization caused by subarray averaging.
 
-In element space, resolution decays rapidly as a result of a reduced adaptive aperture, and for $L = N_b=3$, BS-Capon provides almost 1 mm improved resolution over DAS and ES-Capon.
-However, note that resolution of the Capon beamformers strongly depends on image SNR and the choice of parameters.
+In element space, resolution decays rapidly since the effective adaptive aperture is proportional to $L$.
+For $L = N_b=3$, BS-Capon provides almost 1 mm improved resolution over DAS and ES-Capon.
+However, note that the resolution of the Capon beamformers strongly depends on image SNR and the choice of parameters.
 Thus, different SNR scenarios and choices of parameters will lead to different results.
@@
 \ref{fig:phantom}, and the attached video of the simulation, that Capon beamforming provides improved lateral resolution of moving point targets both in element- and in a reduced beam-space ($N_b=3$).
-Our preliminary finding is that the Capon beamformers are shift invariant as long as the lateral sampling is increased to a curtain level \multimedia{Media-Movie 1 and 2}.
+Our preliminary finding is that the Capon beamformers are local shift invariant as long as the lateral sampling is increased to a curtain level \multimedia{Media-Movie 1 and 2}.
 How large this oversampling factor needs to be and how the oversampling should be conducted is outside the scope of this paper.
@@
 
-\newpage
+%\newpage
 

### 2013-03-26 HRUIonGPU/jrnl_article.tex
Finished tuffc article on gpu capon

@@
 Previous work on parallel implementations of the Capon beamformer have been focused on radar and passive sonar systems using custom made systolic array processors \cite{McWhirter1989, Moonen1993, Sinha2002}.
-For medical ultrasound, Chen et al.
-\cite{Chen2011a, Chen2011, Chen} implemented a Capon beamformer on the GPU that produces real apodization weights based on real data.
+For medical ultrasound, Chen \textit{et al.} \cite{Chen2011a, Chen2011, Chen} implemented a Capon beamformer on the GPU that produces real apodization weights based on real data.
 However, the Capon beamformer is meant to process complex data.
@@
 In the next section we present background information on Capon beamforming in both element-space and beamspace and the GPU computing model.
-In Section \ref{sec:meth} we recap our parallel implementation of the Capon beamformer, before the beamspace method, and its GPU implementation, is presented in Section \ref{sec:bs}.
+In Section \ref{sec:meth} we recap our parallel implementation of the Capon beamformer, before the GPU implementation of the beamspace Capon beamformer is presented in Section \ref{sec:bs}.
 Benchmark results and resulting images are presented in Section \ref{sec:bench} and \ref{sec:images}, together with a discussion of trading image quality for speed in Section \ref{sec:trade}.
@@
 The result is that interference impinging from other directions is suppressed \cite{Synnevag2007}.
-Micro-steering caused by the complex weights is minute compared to the initial steering and focusing step, therefore the beamformer also works well for broadband ultrasound signals.
+The complex weights cause a micro-steering which is minute compared to the initial steering and focusing step, therefore the beamformer also works well for broadband ultrasound signals.
 
@@
 Note that the covariance matrix dimension is $L \times L$.
-The covariance matrix is loaded with a diagonal factor $\epsilon$ to ensure numerical stability and increased robustness,
+The covariance matrix is loaded with a diagonal factor $\epsilon$ to ensure numerical stability and increased robustness:
 \begin{align}\label{eq:diag}
@@
 \end{align}
-Where $f(\cdot)$ represents the process of constructing $\mat{\hat{R}}$ from the input $\vec{x}$.
+The operator $f(\cdot)$ represents the process of estimating $\mat{\hat{R}}$ from the input $\vec{x}$.
 A factor proportional to the output power is often used to reduce the need for parameter adjustments, and the trace of $\mat{\breve{R}}$, 
@@
 
-In this section we give a brief discussion of GPU computing, laying the basis for the discussions on parallel implementation strategies.
+In this section we give a brief discussion of GPU computing, laying the basis for our discussions on parallel implementation strategies.
 The Capon beamformers have been implemented using the CUDA GPU framework by Nvidia \cite{Nvidia2013}, however the challenges and proposed methods described in this paper are valid for other GPU framworks as well, like e.g.
@@
 Independent of framework, the key point is that GPUs can accelerate large sets of small concurrent tasks, like beamforming a pixel in a large image, much better than a CPU.
-While the CPU has few cores and lack logic to process complex threads, the GPU has many cores, but each core needs many (preferably simple) threads to keep it occupied and to thereby achieve maximum throughput.
+While the CPU has few cores, but logic to process complex threads, the GPU has many cores, but each core needs many (preferably simple) threads to keep it occupied and to achieve maximum throughput.
 
@@
 An SM has a limited amount of shared memory (or near-core cache) and registers.
-Therefore, to achieve maximum GPU occupancy (defined as having the maximum number of threads residing on each SM), each thread has to consume as few registers and as little shared memory as possible.
-Each thread should also perform enough instructions per transferred byte in order to hide memory latency.
-Dividing the problem into enough threads will help to hide instruction latency.
+Therefore, to achieve maximum GPU occupancy (defined as having the maximum number of threads residing on each SM), a conservative use of registers and shared memory per thread is required.
+Each thread should perform enough instructions per transferred byte in order to hide memory latency, and the problem should be divided into enough threads in order to hide instruction latency.
 Global GPU memory transactions, in addition to CPU-GPU transfers, are slow and must be minimized.
-All of these constraints combined pose a real challenge as the target problem has to map well to the GPU framework on a grid, block, thread, and memory consumption level in order to benefit from parallel acceleration.
-In the next section, we will describe how this mapping can be done for the Capon beamforming algorithm.
+All of these constraints combined pose a real challenge as the target problem has to map well to the GPU framework on both a grid, block, thread, and memory consumption level in order to benefit from parallel acceleration.
+In the next section, we will describe how this mapping can be done for the ES-Capon beamformer.
 
@@
 For the adaptive diagonal loading factor in (\ref{eq:diag}), the trace of $\mat{\hat{R}}$ has to be known before diagonal elements are written back to global memory.
-Since one row for all covariance matrices is written to global memory per kernel iteration, the diagonal elements need to be held in a shared memory until the trace is accumulated.
+Since one row for all covariance matrices is written to global memory per kernel iteration, the diagonal elements need to be held in shared memory until the trace is accumulated.
 When that is done, scattered writes of the scaled diagonal elements to global memory has to be performed.
@@
 
-For the results presented in this paper, we have used a GPU implementation of batched Gauss-Jordan elimination made by Nvidia to solve $\mat{\hat{R}}\vec{b} = \vec{1}_L$, where $\vec{b} = \mat{\hat{R}}^{-1}\vec{1}_L$, for all pixels in one kernel launch.
+For the results presented in this paper, we have used a GPU implementation of batched Gauss-Jordan elimination made by Nvidia to solve $\mat{\hat{R}}\vec{b} = \vec{1}_L$, where $\vec{b} = \mat{\hat{R}}^{-1}\vec{1}_L$, for all pixels in an image in one kernel launch.
 The solver is available online for registered CUDA developers \cite{Nvidiaa}.%For smaller systems, $L < 10$, a direct solver using one thread per system is preferable.
@@
 If the transformation is applied before, one transformation is needed per subarray.
-After this transformation the subarrays no longer overlap.
+This results in subarrays that no longer overlap.
 The amount of input data are therefore changed from $M$ to $N_LN_b$ elements per sample, but all remaining steps will get lower computational complexity as long as $N_b < L$.
@@
 
-As presented in (\ref{eq:diag}), adaptive diagonal loading is performed by loading the sample covariance matrix with the average energy per element, that is $\text{trace}\{\mat{\hat{R}}\}/L$.
+As presented in (\ref{eq:diag}), adaptive diagonal loading is performed by loading the sample covariance matrix with the average energy per element, that is $\text{trace}\{\mat{\breve{R}}\}/L$.
 When data are transformed into beamspace, a high percentage of the total energy is mapped to a small subset of beams.
-Thus the average energy per beam in a reduced beamspace is, for a focused system, higher than the average energy per element.
-This means that less diagonal loading (smaller $d$) has to be applied when the covariance is estimated from beams instead of elements.
+Therefore, if a set of high energy beams is selected, the average energy per beam will be higher than the average energy per element ($\text{trace}\{\mat{\breve{R}}\}/L < \text{trace}\{\mat{\breve{R}}_{\text{BS}}\}/N_b$).
+This means that less diagonal loading (smaller $d$) has to be applied when the covariance is estimated among beams and not elements.
 
@@
 Assuming a system bandwidth of 80\% (relative to the center frequency) we need a complex sampling frequency of 2.7 MHz, and when imaging down to a depth of 15.4 cm we have approximately 540 samples per range line.
-To sample a $70\degree$ sector with 64- and 96-element arrays ($d=\lambda/2$ pitch) at twice the Rayleigh criterion ($\lambda/(M*d)$), we need at approximately 80 and 120 receive beams respectively.
+To sample a $70\degree$ sector with 64- and 96-element arrays ($d=\lambda/2$ pitch) at twice the Rayleigh criterion ($\lambda/(2Md)$), we need approximately 80 and 120 receive beams respectively.
 The presented execution times do not include transfer of data from CPU- to the GPU-side, since data are assumed to be pre-delayed and available in global GPU memory.
@@
 
-To investigate the image quality of both the ES and BS-Capon beamformer a phantom was simulated using Field II \cite{Jensen1992, Jensen1996a} (Fig.
+To investigate the image quality of both the ES and BS-Capon beamformer, a phantom was simulated using Field II \cite{Jensen1992, Jensen1996a} (Fig.
 \ref{fig:phantom}).
@@
 Speckle level is at 0 dB, covering an area from $\pm90\degree$ in azimuth.
-The resulting image has been processed both using DAS with uniform weights, ES and BS-Capon.
+The resulting image has been processed both using DAS with uniform weights, ES- and BS-Capon.
 The presented image is the first frame in a series of images where motion is imposed on both circles, point targets and the probe.
-The entire simulation is presented in three attached videos \multimedia{Media-Movie 1 and 2}.
+The entire simulation is presented in two attached videos \multimedia{Media-Movie 1 and 2}.
 
 Fig.
-\ref{fig:invivo} presents results of applying ES and BS-Capon on an \textit{in-vivo} image of the left ventricle.
+\ref{fig:invivo} presents results of applying ES- and BS-Capon on an \textit{in-vivo} image of the left ventricle.
 The image was acquired using a 64-element and $0.65\lambda$ pitched phased array in harmonic mode (1.7 MHz at transmit and 3.4 MHz at receive).
 The image is extracted from a cardiac cycle, that can be viewed in its full length in the attached videos \multimedia{Media-Movie 3, 4 and 5}.
-It is included to show that ES-Capon and BS-Capon beamforming are both feasible for cardiac imaging, not just in simulations.
+It is included to show that both ES-Capon and BS-Capon beamforming are feasible for \textit{in-vivo} cardiac imaging, not just in simulations.
 
@@
 For the simulated phantom, 16x lateral oversampling on transmit has been performed in order to avoid under sampling of the two bright point scatterers as they move.
-The cardiac recording of a 70\degree{} sector has a minimal lateral sampling of 98 receive beams.
+The cardiac recording of a 70\degree{} sector has a minimal lateral sampling of 98 receive beams ($\lambda/2Md$).
 
@@
 \ref{fig:phantom} (while preserving a 6 dB saddle point in between the two) is plotted against $L$ (when processed with ES-Capon) and $N_b$ (when processed with BS-Capon).
-For the result in Fig.
+To produce the result in Fig.
 \ref{fig:speed_res_trade}, $d=0.01$ has been used on both methods in order to obtain the same resolution for $L=N_b=M/2$.
-This is considered a small value for ES-Capon and a normal value for BS-Capon (cf.
+We considered this to be a small value for ES-Capon and a normal value for BS-Capon (cf.
 Section \ref{sec:bs}), hence the resolution for ES-Capon is here better than what is presented in Fig.
@@
 It is also worth noting that ES-Capon beamforming in this example performs worse than DAS for small values of $L$.
-For these small subarrays, the amount of robustification is too high in order to correctly steer a zero at each point target or to accurately micro-steer the main lobe.
+For these small subarrays, the method is not able to steer zeros at each point target or accurately micro-steer the main lobe.
 Subarray averaging further impose a triangular or trapezoid apodization function, which leads to worse resolution if the introduced adaptivity can not compensate for the increase in main lobe width.
@@
 According to Fig.
-\ref{fig:benchBS} it is also possible to process two parallel receive lines per transmit (for both $M=64$ and $M=96$.
+\ref{fig:benchBS} it is also possible to process two parallel receive lines per transmit beam (for both $M=64$ and $M=96$.
 $L=M/2$).
-However, these extra lines might be needed for increased sampling instead of increased frame rate when using the Capon beamformers in high SNR situations (Fig \ref{fig:phantom}).
+These extra lines might be needed for increased sampling instead of increased frame rate when using the Capon beamformers in high SNR situations (Fig \ref{fig:phantom}).
 If parallel receive lines are used to obtain denser sampling, acquisition time remains the same but processing times are increased with a factor equal to the amount of oversampling.
 If parallel receive lines are used for increased frame rate, the acquisition time is lowered, and the processing time remains constant.
-Hence, four parallel receive lines are not possible to process in real-time using BS-Capon (with $N_b=3$) in combination with our single GPU.
+Hence, four parallel receive lines are not possible to process in real-time using BS-Capon (with $N_b=3$) in combination with our single GPU setup.
 
@@
 How large this oversampling factor needs to be and how the oversampling should be conducted is outside the scope of this paper.
-No major improvement in contrast is visually observed in the presented images images and videos, other than slightly rounder circles due to mainlobe micro steering.
+No major improvement in contrast is visually observed in the presented images and videos, other than slightly rounder circles due to mainlobe micro steering.
 In the cardiac recording (Fig.
@@
 \section*{Acknowledgment}
-The authors would like to thank Dr Kjell Kristoffersen, GE Vingmed Ultrasound, for valuable comments.
+The authors would like to thank Dr K. Kristoffersen, GE Vingmed Ultrasound, for valuable comments.
 

### 2013-03-31 HRUIonGPU/jrnl_article.tex
added new figure about sliding beamspace

@@
 
-In \cite{asen2012, Buskenes} we demonstrated how the cubic complexity with $L$ and linear complexity with $N_K$ can be reduced by realizing that calculations in (\ref{eq:R}) overlap both across subarrays and in time.
+In \cite{asen2012, Buskenes} we demonstrated how the cubic complexity with $L$ and linear complexity with $N_K$ can be reduced by realizing that calculations in (\ref{eq:R}) overlap both across subarrays.
 Element $(i,j)$ of $\mat{\breve{R}}[n]$ can be calculated from element $(i-1, j-1), \text{for } i,j \in [1, L-1]$ in a sliding window maner as
@@
 \end{align}
-Along the time dimension we also find that
-\begin{align}
-\mat{\breve{R}}[n] &= \mat{\breve{R}}[n-1] + \sum_{l=0}^{N_L-1} (\vec{x}_l[n+K]\vec{x}_l[n+K]^H \nonumber \\
-&- \vec{x}[n-K-1]\vec{x}[n-K-1]^H).
-\end{align}
+%Along the time dimension we also find that
+%\begin{align}
+%\mat{\breve{R}}[n] &= \mat{\breve{R}}[n-1] + \sum_{l=0}^{N_L-1} (\vec{x}_l[n+K]\vec{x}_l[n+K]^H \nonumber \\
+%&- \vec{x}[n-K-1]\vec{x}[n-K-1]^H).
+%\end{align}
 %- \mat{\breve{R}}[n-K-1] + \mat{\breve{R}}[n-K].
@@
 %The last sum can also be performed using a binary reduction sum across several threads for increased speed.
+
+\begin{figure}
+\centerline{\includegraphics[width=3.5in]{gfx/gpu_layout_bs.eps}}
+\caption{Calculation of the sliding beamspace transformation.
+A group of $N_b$ threads are assigned a set of subarrays which they transform to beamspace in a sliding window fashion.
+In this example $N_b=3$ and $N_L\,\text{mod}\,N_b = 0$.}
+\label{fig:gpulayoutbs}
+\end{figure}
 
@@
 \section*{Acknowledgment}
-The authors would like to thank Dr K. Kristoffersen, GE Vingmed Ultrasound, for valuable comments.
+The authors would like to thank Dr. K. Kristoffersen, GE Vingmed Ultrasound, for valuable comments.
 

### 2013-04-03 HRUIonGPU/jrnl_article.tex
Final commit after submitting gpucapon paper to uffc.

@@
 
-\documentclass[journal]{IEEEtran}
+%\documentclass[journal]{IEEEtran}
+\documentclass[11pt, draftcls, onecolumn]{IEEEtran}
 
@@
 \begin{align}
-z[n] = \sum_{m = 0}^{M-1}w_m^*x_m[n] = \vec{w}^H\vec{x}[n]. \label{eq:z}
-\end{align}
+z[n] = \sum_{m = 0}^{M-1}w_m^*x_m[n] = \vec{w}^H\vec{x}[n], \label{eq:z}
+\end{align}
+where $(\cdot)^H$ is the Hermitian transpose.
 The weight vector, or window, $\vec{w}$ is usually real in DAS beamforming, and is applied to trade resolution for a better signal-to-noise ratio (SNR).
@@
 \end{align} 
-The steering vectors in $\mat{B}$, the so-called Butler or normalized DFT matrix, determine each beam's direction in space.
+The steering vectors in $\mat{B}$, the so-called Butler or normalized discrete fourier transform (DFT) matrix, determine each beam's direction in space.
 Beams where no interference is present can therefore be removed by simply removing rows from $\mat{B}$.
@@
 OpenCL.
-Independent of framework, the key point is that GPUs can accelerate large sets of small concurrent tasks, like beamforming a pixel in a large image, much better than a CPU.
-While the CPU has few cores, but logic to process complex threads, the GPU has many cores, but each core needs many (preferably simple) threads to keep it occupied and to achieve maximum throughput.
+Independent of framework, the key point is that GPUs are designed to execute large sets of small concurrent tasks, like beamforming a pixel in a large image, much faster than a CPU.
+While the CPU has few cores, but logic to process complex threads, the GPU has many cores, but each core needs many (preferably simple) threads to keep it occupied and by that achieve maximum throughput.
 
@@
 All of these constraints combined pose a real challenge as the target problem has to map well to the GPU framework on both a grid, block, thread, and memory consumption level in order to benefit from parallel acceleration.
-In the next section, we will describe how this mapping can be done for the ES-Capon beamformer.
+In the next two sections, we will describe how this mapping can be done for the Capon beamformers.
 
 \begin{figure}
-\centerline{\includegraphics[width=3.5in]{gfx/gpu_layout_vertical.eps}}
+\centerline{\includegraphics[width=3.5in]{gfx/gpu_layout_vertical_2.eps}}
 \caption{Calculation of Capon adaptive weights mapped to GPU architecture.
@@
 The computational complexities of these three steps are $O(N_KN_LL^2)$, $O(L^3)$, and $O(L)$, respectively.
-As $M$ (and thereby $L$) increases, the first two steps quickly become unmanageable, and real-time performance can only be achieved by exploring methods with reduced complexity.
+As $M$ (and thereby $L$) increases, the first two steps quickly become unmanageable, and real-time performance in cardiac ultrasound imaging can only be achieved by exploring methods with reduced complexity.
 
@@
 A natural level of granularity would then be to assign one thread to each element of $\mat{\hat{R}}$ \cite{Chen2011}.
-This will minimize global reads and writes needed per thread, but since we want to avoid block-to-block synchronization, the maximum subarray length is limited to 32 by this approach.
+This will minimize global reads and writes needed per thread, but since we want to avoid block-to-block synchronization, the maximum supported subarray length is limited to 32 by this approach.
 Since most medical ultrasound arrays have 64 elements or more, an implementation not limited by $L_{max}=M/2=32$ is preferable.
 
-In \cite{asen2012, Buskenes} we demonstrated how the cubic complexity with $L$ and linear complexity with $N_K$ can be reduced by realizing that calculations in (\ref{eq:R}) overlap both across subarrays.
+In \cite{asen2012, Buskenes} we demonstrated how the cubic complexity with $L$ and linear complexity with $N_K$ can be reduced by realizing that calculations in (\ref{eq:R}) overlap across subarrays.
 Element $(i,j)$ of $\mat{\breve{R}}[n]$ can be calculated from element $(i-1, j-1), \text{for } i,j \in [1, L-1]$ in a sliding window maner as
@@
 \begin{figure}
-\centerline{\includegraphics[width=3.5in]{gfx/gpu_layout_bs.eps}}
-\caption{Calculation of the sliding beamspace transformation.
-A group of $N_b$ threads are assigned a set of subarrays which they transform to beamspace in a sliding window fashion.
-In this example $N_b=3$ and $N_L\,\text{mod}\,N_b = 0$.}
+\centerline{\includegraphics[width=3.2in]{gfx/gpu_layout_bs_2.eps}}
+\caption{Calculation of the sliding beamspace transform.
+A group of $N_b$ threads is assigned a set of subarrays which they transform to beamspace in a sliding window fashion.
+In this example $N_b=3$, $l$ is the current subarray and $k$ is the current beam.}
 \label{fig:gpulayoutbs}
@@
 This results in subarrays that no longer overlap.
-The amount of input data are therefore changed from $M$ to $N_LN_b$ elements per sample, but all remaining steps will get lower computational complexity as long as $N_b < L$.
+The amount of input data are therefore changed from $M$ to $N_LN_b$ elements per sample, but all remaining steps should get lower computational complexity as long as $N_b < L$.
 
@@
 In addition, for small $N_b$ values, several covariance matrices are calculated per thread block to achieve high GPU occupancy.
+The computational complexity of this step is now $O(N_KN_LN_b^2)$, which for $L=M/2$ is less than the complexity of building $\mat{\hat{R}}$ in element space as long as $N_b < \sqrt{M}$.
 
@@
 Using one thread per beam is therefore not feasible for small-$N_b$-large-$M$ configurations.
-The solution is to combine SBS with full computations.
-A number of thread groups is created that targets different sets of subarrays which they transform using the SBS algorithm.
+One solution, as depicted in Fig.
+\ref{fig:gpulayoutbs}, is to combine SBS with full computations.
+A number of sub thread groups of size $N_b$ is created that targets different sets of subarrays which they transform using the SBS algorithm.
 In that way, the number of threads per sample is increased and thus the memory-per-thread ratio is decreased.
@@
 
-To summarize, by transforming data to beamspace the complexity of all subsequent steps is reduced.
+To summarize, by transforming data to beamspace the complexity of all subsequent steps is potentially reduced.
 Calculation of the covariance matrix is now $O(N_KN_LN_b^2)$, the inversion step is $O(N_b^3)$, and calculation of weights and the beamformer output are both $O(N_b)$.

### 2013-04-03 HRUIonGPU/retningslinjer_uffc/aberrated.tex
Final commit after submitting gpucapon paper to uffc.

%\iffalse
% --- IEEE publication
	\documentclass[%
	%11pt,journal,twoside,onecolumn%%
	10pt,journal,twoside,twocolumn%
	%,draft,a4paper,
	]{IEEEtran}
	\usepackage[square,comma,compress,numbers]{natbib}
	\usepackage[breaklinks]{hyperref}
	%\usepackage{cite}
	\usepackage{hypernat}
	\usepackage[english]{babel}  % avstavning (?)
		\let\ORGforeignlanguage\foreignlanguage
		\def\foreignlanguage#1{\lowercase{\ORGforeignlanguage{#1}}} 
	\usepackage[latin1]{inputenc}
	\usepackage[dvips%
	%,draft%
	]{graphicx}
	\usepackage{amssymb}
	\usepackage{amsmath}
	\interdisplaylinepenalty=2500
	\usepackage{epsfig}
	\usepackage{color} 
	\usepackage[raggedright,tight,hang]{subfigure}
	\usepackage{booktabs}
	\usepackage{srcltx}
	\usepackage{soul}

	\newcommand{\maybespace}{\ }
	\newcommand{\mayberowbreak}{\\}
	\newcommand{\maybehalffigure}{\columnwidth}
\iffalse
	% --- double spacing for peer review
	\usepackage{setspace}
	\doublespacing
	\renewcommand{\mayberowbreak}{ }
	\newlength{\templen}
	\setlength{\templen}{.49\textwidth}
	%\renewcommand{\maybehalffigure}{.49\textwidth}
	\renewcommand{\maybehalffigure}{\templen}
\fi

%\iffalse

\newcommand{\maybehighlight}[1]{#1}
%\newcommand{\maybehighlight}[1]{\hl{#1}}

% -------- USE COMPUTER MODERN FONT, NOT THE STANDARD IEEEtran TIMES ROMAN FONT
	\renewcommand{\sfdefault}{cmss}
	\renewcommand{\rmdefault}{cmr}
	\renewcommand{\ttdefault}{cmtt}

\renewcommand{\thetable}{\Roman{table}.}

%\fi
% ----------
%

%
%
\newcommand{\panedescription}{Left: $\hh_{z_a}$ adjustment, middle: $\tau_a$ adjustment, right: no adjustment.}
\newcommand{\hh}{\ensuremath{\mathfrak{h}}}
\newcommand{\vr}{\vx}
\newcommand{\vx}{\ensuremath{\myvecx}}
\newcommand{\myvec}[1]{\vec{#1}}
\newcommand{\myvecx}{\myvec{r}}
\newcommand{\dd}{\mathrm d}
\newcommand{\h}{\protect{2\ensuremath{^\textrm{h}}}}
\newcommand{\Real}{\mathfrak{Re}}
\newcommand{\Imag}{\mathfrak{Im}}
\newcommand{\tcp}{\text{\textcircled{$+$}}}
\newcommand{\tcm}{\text{\textcircled{$-$}}}
\newcommand{\MI}{\protect{\ensuremath{\mathrm{MI}}}}
\newcommand{\HF}{\protect{\ensuremath{\mathrm{HF}}}}
\newcommand{\LF}{\protect{\ensuremath{\mathrm{LF}}}}
\newcommand{\HH}{\protect{\ensuremath{\mathrm{H}}}}
\newcommand{\LL}{\protect{\ensuremath{\mathrm{L}}}}
\renewcommand{\theenumi}{\roman{enumi}}
\renewcommand{\thefootnote}{\fnsymbol{footnote}} % use symbols for footnotes

\newcommand{\maybecoloreqn}[1]{{#1}}

\newlength{\figwidthD}

\setlength{\figwidthD}{.4\textwidth}

%\setlength{\figwidthD}{.245\textwidth}
%\newcommand{\mayberowbreak}{}

\newcommand{\insertkeywords}{%
\begin{keywords}
Ultrasound, SURF, reverberation, multiple scattering, acoustic noise suppression, transmit-beam, post-processing, wave-front aberration, inhomogeneous medium, body-wall
\end{keywords}}

\begin{document}
\title{%
\bf%
%{SURF reverberation suppression transmit-beam generation and post-processing adjustment in an aberrating medium}}
{SURF imaging beams in an aberrative medium: generation and post-processing enhancement}}
\author{Sven~Peter~{Näsholm},~\IEEEmembership{Member,~IEEE} and Bjørn~A.~J.~{Angelsen},~\IEEEmembership{Senior~Member,~IEEE}%
%\email{peter.nasholm@ntnu.no}

\thanks{Manuscript received Month DD, 2011; accepted Month DD, 201X. %}
%\thanks{%
This work was supported by the Medicine and Health program of the Research Council of Norway.}
\thanks{The research was mainly done while both authors were with the Department of Circulation and Imaging, Norwegian University of Science and Technology, Trondheim, Norway. Sven Peter Näsholm is now with the Department of Informatics, University of Oslo, Norway (e-mail: svenpn@ifi.uio.no).}
\thanks{Digital Object Identifier XX.XXXX/TUFFC.201Y.NNNN} 
}

%\markboth{}{}
\markboth{\sc IEEE Transactions on Ultrasonics, Ferroelectrics and Frequency Control, Vol.~X, no.~X, Month 201X}{\sc Näsholm and Angelsen: {SURF} imaging beams in aberrating medium}

\pubid{0000--0000/\$25.00 \copyright\ 2012 IEEE}
%\pubid{\copyright 2011}
\maketitle
%\input{aberrated_content}
%
%
%
\newcommand{\linedescriptions}{%
%
\includegraphics{thick_dashed_line_red}: time-shift adjustment  (homog.),  
\includegraphics{thin_blue_line}: time-shift (inhomog.), %
\includegraphics{thin_dashed_line_red}: filter $\hh_{z_a}$ adjustment (homog.), % 
\includegraphics{thick_line}: filter{ }$\hh_{z_a}$ adjustment (inhomog.)%
}
\newcommand{\linedescriptionscoresoverviewfields}{%
\includegraphics{thick_line}: fundamental (inhomog.),%
\raisebox{-.02cm}{\includegraphics{thin_dashed_line_red}}: SURF (inhomog..), %
\includegraphics{thin_blue_line}: fundamental (homog.), %
\raisebox{-.02cm}{\includegraphics{thick_dashed_line_red}}: SURF (homog.).%
%
}%

\newcommand{\linedescriptionssecfig}{%
	Time-shift adjusted: \includegraphics[width=.6cm]{thick_line} (homogeneous) and \includegraphics[width=.9cm]{thin_line} (inhomogeneous. %
	Filter-adjusted: \raisebox{-.03cm}{\includegraphics[width=1cm]{thick_dotted_line}} (homog.) and \raisebox{-.03cm}{\includegraphics[width=.9cm]{thin_dotted_line}} (inhomog.), %
	No adjustment: \raisebox{-.03cm}{\includegraphics[width=.8cm]{thick_dash-dot_gray_red}} (homog.) and  \includegraphics[width=.8cm]{thin_dash-dot_gray_red} (inhomog.), %
	Fundamental beam: \raisebox{-.02cm}{\includegraphics[width=.9cm]{thick_dashed_line_red}} (homog.) and \raisebox{-.02cm}{\includegraphics[width=.9cm]{thin_dashed_line_red}} (inhomog.). %
	}

{\footnotesize%
\begin{abstract}
This paper presents numerical simulations of dual-frequency second-order ultrasound field (SURF) reverberation suppression transmit-pulse complexes. Such propagation was previously studied in a homogeneous medium. Here instead the propagation path includes a strongly aberrating body-wall modeled by a sequence of delay-screens. %
The applied SURF transmit pulse complexes each consist of a high-frequency imaging 3.5\,MHz pulse combined with a low-frequency 0.5\,MHz sound speed manipulation pulse. %
%
Furthermore, the feasibility of two signal post-processing methods are investigated using the aberrated transmit SURF beams. These methods are previously shown to adjust the depth of maximum SURF reverberation suppression within a homogeneous medium. %
%
The request of the study arises because imaging situations where reverberation suppression is useful are also likely to produce pulse wave-front distortion (aberration). Such distortions could potentially produce time-delays that cancel the accumulated propagation time-delay needed for the SURF reverberation suppression technique. %
%
Results show that both the generation of synthetic SURF reverberation suppression imaging transmit-beams, and the following post-processing adjustments, are attainable even when a body-wall introduces time-delays which are larger than previously reported delays measured on human body-wall specimens.%
\end{abstract}%
}
%
%
%
%\insertkeywords

\IEEEpeerreviewmaketitle

%-------------------------------------------------------------------
\section{Introduction}
%-------------------------------------------------------------------

%XXXX also cite: 
%-- Yan and Hamilton JASA 2011 statistical model of clutter suppression in tissue harmonic imaging

\PARstart{T}{his} work concerns dual-frequency transmit-beams that are utilized for reverberation noise suppression in ultrasound image reconstruction by use of methods analyzed in \cite{nasholm2009,nasholm2011}. The developments of the current paper regard the feasibility of such beam generation within an inhomogeneous medium that generates sound-speed variations that aberrate the beam. \maybehighlight{Although the subject is connected, here we don't apply or develop aberration correction algorithms. }%

{The} quality of medical ultrasound images varies greatly between patients. Spatial inhomogeneities in compressibility and density within the ultrasound propagation path cause additive reverberation (multiple scattering) noise, as well as resolution degradation due to wave-front distortion (aberration). This is especially prominent when imaging through {e.g.} the abdominal wall or the chest wall of obese patients\maybespace\cite{hinkelman:pulse_dist_meas-98, keogh:real_or_artifact-01, mahmutyazicioglu:transabdominal-05, scanlan:artifacts-91, shmulewitz:quality_factors}. %
A recent simulation study supports the hypothesis that the image degradation due to reverberation noise is more prevalent than degradation due to aberration noise \cite{Pinton2011,Pinton2011erratum}. %

In dual-frequency second-order ultrasound field (SURF) reverberation-suppression imaging, a synthetic transmit-beam is generated from the difference between two high-frequency (HF) imaging pulses transmitted in the same direction. %
This beam is in the following simply denoted the SURF beam. %
%
Both transmissions comprise a dual-frequency band pulse-complex with the HF part used for image reconstruction, added to a low-frequency (LF) part used for material compressibility manipulation. %
%
The high frequency is typically of the order 10 times the low frequency. %
%
The compressibility is pressure-dependent due to material nonlinearity. This makes the sound-speed pressure-dependent, being higher for the compressed medium than for the expanded medium \cite{angelsen:ultrasoundbook_II}. %
%
The LF polarity is switched for the second transmission, therefore making the LF pressure experienced by the second HF pulse opposite what was experienced by the first HF pulse. %
%
%Because they are transmit in conjunction with opposite LF pulse polarity, 
{The pressure-dependent sound-speed thus causes the two HF pulses to propagate at different speeds. %
With increasing distance traveled, the difference in propagation-time required for the two HF pulses hence increases.} %
%
%
% needed in second column of first page if using \pubid
% -------
\pubidadjcol
% -------
%
%
%
%
%
{Consequently, the difference obtained when subtracting the HF pulses is zero at zero depth and grows with covered distance to attain a maximum at 180$^\circ$ phase-shift.} This is the effect that is exploited for reverberation suppression as follows. %
%
The SURF beam is the synthetic transmit-beam obtained when subtracting the propagated HF pulses at each spatial position. %
%
%
%
{The amplitude of scattered (or reflected) pulses are severely reduced, hence making the nonlinear sound-speed manipulation negligible. Therefore scattered HF parts accumulate less relative time-shift than forward-propagating ones.} %
%
%
Especially when the first scattering takes place at shallow depths, multiple scattered contributions within the received HF pulses thus give negligible contribution to the receive SURF difference HF field. 
%
%
%Following the definitions of\maybespace\cite{nasholm2009}, such reverberations are within the Class I and Class II categories and the SURF beam illustrates the reverberation suppression ability by being reduced near the transducer. The tissue harmonic imaging transmit-beam also possesses this property\maybespace\cite{kaupang2011,thomas:thi_why_does_it_work}. %
%
%
%
The reverberation suppression ability of SURF imaging is illustrated by the fact that the beam is reduced near the transducer, as was studied in \cite{nasholm2009} %
where nonlinear numerical propagation simulations in a homogeneous medium were reported. %
In addition, the combination of tissue-harmonic pulse-inversion with SURF reverberation suppression was introduced. %
%In \cite{nasholm2011}, a post-processing enhancement method for SURF reverberation-suppression imaging was presented. %

In case the dominating reverberation noise is due to a strong first scattering (reflection) taking place around a depth $z_a$, the SURF synthetic transmit field may without modification or re-transmission of the sent-out pulses, be adjusted to increase the transmit-beam suppression around $z_a$ by use of post-processing methods as described and analyzed in %
%\cite{nasholm:SURF_postpr} %
\cite{nasholm2011} for a homogeneous medium. %
%
Both multiple scattering and multiple reflection image artifacts are in the following considered as reverberation artifacts. %
%
The recent publications \cite{Hansen2011,Masoy2010,hansen2010} concern SURF dual-frequency acoustics within a wider context. %Hansen2011B,
%
Dual frequency band pulse complexes have also been used for contrast agent detection \cite{deng:dual_frequency, angelsen_hansen:ieee_proc07, bouakaz:radial_modulation, masoy:SURF_in_vivo, hansen2009, emmer2009}.
%
%
%

The purpose of the computer simulation study presented in the current paper is to evaluate the feasibility of SURF transmit reverberation suppression beam generation when a strongly aberrative body-wall is present within the wave-propagation path. Furthermore, it intends to compare these SURF beams to standard fundamental imaging transmit-beams, and in addition to test the post-processing SURF transmit field enhancement techniques described in \cite{nasholm2011} in case of aberration. %

The study is important because the body-wall of a patient where standard fundamental ultrasound imaging is aggravated due to reverberation noise, is also likely to produce propagation time-shifts due to aberration. If these time-shifts heavily distort the accumulated time-shifts needed for adequate SURF synthetic beam generation, the reverberation suppression gain of the SURF imaging method is no longer obtained. %

%
%
This paper is organized as follows: First the dual-frequency SURF imaging and fundamental imaging excitation pulses used in the numerical simulations are presented. Then a description of the body-wall model and field simulation method is given. %
%
The results section show{s} comparisons between transmit-beams from the described excitation pulses both for a homogeneous medium and after propagation through the body-wall. Beam profiles in the focal plane are derived from the transmit-beams, to quantify the effect the body-wall has on the transmit field.
%
Finally, post-processing adjusted SURF transmit-beams are constructed for some chosen example depths of decreased transmit-beam amplitude. %

%The same raw simulated transmit field data is used as for the 3.5\;MHz setup in \cite{nasholm:SURF_beams}. The additional processing is applied in order to construct $N$ SURF transmit-beams tailored to suppress multiple scattering noise from a set of $N$ near-field depths.

%-------------------------------------------------------------------
\section{Theory and Methods}\label{section:theory_methodsD}
%-------------------------------------------------------------------
%-------------------------------------------------------------------
\subsection{Excitation pulses}
The two excitation pulses utilized to generate the SURF synthetic transmit-beam and the transducer apertures are equal as for a setup applied in \cite{nasholm2009,nasholm2011}. The HF imaging part of each pulse complex has a center frequency at $f_\HH=3.5$\,MHz (50\% fractional bandwidth at $-6$\:dB) while the LF part, whose polarity is inverted for the second pulse, has a center frequency at $f_\LL=0.5$\,MHz (25\% bandwidth at $-6$\:dB). %
The HF and LF surface pressure amplitudes are adjusted to keep the Mechanical Index (MI) below 1.9 and are therefore set to $3.5$ and $0.85$\:MPa respectively. %
%
{The peak negative pressure value applied for MI calculation is read before filtering the SURF pulse complex, hence taking the amplitude contribution of the LF part into account. Therefore the highest MI is observed for the SURF complex of negative LF manipulation polarity.} %
%
The axisymmetric aperture is focused at 82\:mm with an outer radius for the HF part of $7.1$\,mm and $10$\,mm for the LF part. The HF pulse transmission is delayed {by $\tau_0$}$\: = 0.2$\,$\mu$s compared to the LF, measured at their centers. Fig.~2~in Ref.~\citenum{nasholm2009} illustrates the influence of $\tau_0$ on the transmit SURF pulse complex. %

The excitation pulse utilized to generate the comparison fundamental imaging field, is equal to the HF part of the SURF complexes described above.

%\subsection{Summasjon av två HF-pulser, som ær olika aberrerte}
%-------------------------------------------------------------------
%assuming LF not aberrated
%vad borde hænda med HF-profilerna
%
%- LF-nivån som går ned på sidorna borde ta ned sidelobene som kommer av aberrationer

%-------------------------------------------------------------------
\subsection{Body-wall model and pulse propagation simulation}
%-------------------------------------------------------------------
The applied forward wave-propagation simulation method solves the nonlinear Khokhlov-Zabolotskaya-Kuznetsov (KZK) wave equation taking attenuation and interaction between the HF and LF parts of the SURF pulse complexes into account\maybespace\cite{frijlink2008, varslot:propagation}. %
The transducer apertures modeled are rotational symmetric. %Therefore the simulations could in case of a homogeneous medium efficiently be performed using an axisymmetric two-dimensional propagation method. %
%
{Because} a non-symmetric aberrating body-wall is taken into account within this work, all simulations are performed in three spatial dimensions, hence demanding longer computation times and greater storage capacity {than for 2-D rotational symmetric simulations}. %

The body-wall model utilized was generated using a set of 13 two-dimensional filtered time-delay white-noise screens, adjusted to emulate a strongly aberrating abdominal wall. The screens are equally spaced by $\Delta z=3$\,mm thus giving a total wall thickness of $d=39$\:mm as shown in Fig.~\ref{fig:body-wall}. %
%
%
\begin{figure}[!tb]
\renewcommand{\figurename}{Fig.} % Make figure captions say Fig. instead of Figure.
  \includegraphics[width=\columnwidth]{3D_setup}
\caption{Body-wall model and simulation setup outline. The total wall thickness is $d=39$\:mm, the transmit focus is at $F$, and $\Delta z=3$\:mm is the distance separating each delay screen.}
\label{fig:body-wall}
\end{figure}%
%
%
This body-wall is similarly modeled as in \cite{varslot:aberr_nonlin_wave_prop}, however the aberrative effect is here chosen to be more severe in order to test the feasibility of the SURF beam generation methods under harsher imaging conditions. The body-wall model construction method is thoroughly described in \cite{masoy:correction-aberration} and the characteristics of the modeled body-wall, as found from simulations of back-propagation of the signal from an excited point source in focus, are displayed in Table \ref{tab:body_wall} %
%
%
\begin{table}[!bt]
\renewcommand{\tablename}{\sc TABLE} % Make figure captions say Fig. instead of Figure.
	\caption{Characteristics of the simulated body-wall\label{tab:body_wall}}
	\centering
	\begin{tabular}{l l}
		%\toprule
		%& unit & value\\
		%\midrule
		\toprule 
			Amplitude correlation-length & 1.4 mm\\
			Amplitude RMS fluctuation & 4.5 dB\\
			Time-delay correlation-length& 1.2 mm\\
			Time-delay RMS fluctuation & 450 ns\\
		\bottomrule
	\end{tabular}
\end{table}
%
At zero-pressure, the material property average values are: speed of sound 1550 m/s, density 1.06 mg/mm$^3$, and the nonlinearity parameter $\beta_n=3.9$. The attenuation is described by a power-law making it proportional to $f^{1.1}$ with the loss $\alpha_0=0.52$\,dB/cm at 1\,MHz. %
%
{Details on the simulation and attenuation implementations are given in Ref.~}\citenum{abersim_manual}. %

Both the corresponding non-aberrated SURF beam and the non-aberrated fundamental beam are simulated and used as reference beams. %
All beams are generated by detection of the temporal maximum of the pressure field at each {spatial} coordinate. %

%-------------------------------------------------------------------

\subsection{SURF signal post-processing}
%-------------------------------------------------------------------
Two different signal post-processing methods, further described and analyzed for a homogeneous medium in\maybespace\cite{nasholm2011}, are applied on the SURF field data filtered around the frequency $f_\HH = 3.5$\,MHz. %
%
They both aim to further suppress the SURF transmit-beam at some depth $z_a$, so that the ratio between the beam energy within the imaging depth region and the energy at the chosen suppression depth is increased. This way image reverberation-artifacts due to first scattering from some strongly reflecting feature at the depth $z_a$ may be suppressed. %

In short, given that the two transmit HF waves propagated in conjunction with LF manipulation pulses of opposite polarities are $s_+(\vr,t)$ and $s_-(\vr,t)$, the second HF signal is processed to form the field $\hat s_-(\vr,t)$ which is used to form the difference field
\begin{align}
	s_\Delta(\vr,t) =s_+(\vr,t)-\hat s_-(\vr,t).
\end{align}
The adjusted SURF synthetic transmit-beam is then generated from $s_\Delta(\vr,t)$. %

The first post-processing method involves a pure time-shift of the $s_-(\vr,t)$ field by $\tau_a$ so that 
\begin{align}
	\hat s_-(\vr,t)=s_-(\vr,t-\tau_a).
\end{align}

The second post-processing method involves some more general filter $\hh_{z_a}$ which makes the on-axis pulse of $s_+$ and $s_-$ equal at the depth $z_a$:
\begin{align}
	\hat s_-(\vr,t)=\hh_{z_a}\big\{s_-(\vr,t)\big\}.
\end{align}
Here this filter $\hh_{z_a}$ is determined by comparison of the two simulated fields as averaged over a laterally $0.5$\,mm wide region $S$ on-axis at the depth $z_a$ and finding the filter that makes these two average signals equal. %
%
For the demonstrated inhomogeneous medium case, the adjustment filter $\hh_{z_a}$ utilized is not generated from the actual aberrated $s_+(\vr,t)$ and $s_-(\vr,t)$ fields, but instead from the $s_+(\vr,t)$ and $s_-(\vr,t)$ emerging from the homogeneous medium propagation. This approach is chosen to in some extent better emulate how an \emph{à priori} determination of $\hh_{z_a}$ %
%
{could be performed by experimental field characterization}. %
%
The averaging over the region $S$, which was not performed in \cite{nasholm2011}, is introduced to take a finite hydrophone width into account. % 
 The determination of $\hh_{z_a}$ in a real medical imaging setup may be more cumbersome, as further discussed in \cite{nasholm2011}. %
{We remark that when receive channel-data is available, any of the proposed suppression methods may be applied, and images can be reconstructed independently for multiple choices of $z_a$. Great flexibility in reverberation suppression customization is hence enabled. }

%-------------------------------------------------------------------
\subsection{Compared transmit-beams and fields}
%-------------------------------------------------------------------
In the following is a listing of the simulated datasets utilized to generate the transmit-beams that are compared within this work: %
%
\subsubsection{Non-adjusted SURF beam\label{subsubsection:non-adjusted}} %
The HF transmit fields $s_+(\vr,t)$ and $s_-(\vr,t)$, filtered around $f_\HH=3.5$\,MHz on the same actual dataset as utilized in \cite{nasholm2009}, are subtracted without further modification to generate $s_\Delta(\vr,t)=s_+(\vr,t) - s_-(\vr,t)$. 
%
\subsubsection{Time-shift adjusted SURF beams} % 
The time-shift adjustment post-processing method is applied on the same original dataset as used to generate the non-adjusted SURF, for a number of different $z_a$ to generate the adjusted HF field $s_\Delta(\vr,t) = s_+(\vr,t) - s_-(\vr,t-\tau_a)$.
%
\subsubsection{Filter-adjusted SURF beams} %
The filter-adjustment post-processing method is applied on same original dataset as for the non-adjusted SURF, for a number of different $z_a$ to generate the adjusted HF field $s_\Delta(\vr,t) = s_+(\vr,t) - \hh_{z_a}\big\{s_-(\vr,t-\tau_a)\big\}$.
%
%
\subsubsection{Fundamental imaging beam} %
Standard fundamental imaging transmit fields without LF manipulation or SURF processing. The transmit pulse is equal to the HF part of the SURF pulse complex described above. %

%

%-------------------------------------------------------------------
\section{Results and Discussion}
%-------------------------------------------------------------------

%-------------------------------------------------------------------
%\subsection{Non-adjusted SURF fields compared to fundamental fields}
%-------------------------------------------------------------------

{The propagation effects caused by the modeled body-wall, including focus degradation, are further illustrated in Fig}~\ref{fig:fields_no_processing}, %
which shows beam profiles, transmit-beams, and axial pulses for (non-adjusted) SURF and fundamental beams. % 
%
%
%
\begin{figure}[!tb]
\renewcommand{\figurename}{Fig.} % Make figure captions say Fig. instead of Figure.
\centering
	\includegraphics[width=.92\maybehalffigure]{fig_focusprofiles_1}%
	{\smallskip}
	\\%mayberowbreak
	\includegraphics[width=.92\maybehalffigure]{fig_no_postproc_1}%
	\:\raisebox{.5cm}{\includegraphics[height=2.3cm]{colorbarnew}}\quad
	%{\smallskip}
	\\
	\includegraphics[width=.92\maybehalffigure]{newfig}%
	{\smallskip}
	\\%mayberowbreak
	\includegraphics[width=.92\maybehalffigure]{fig_no_postproc_3}%
	%
	%
	\caption[]{{Body-wall model impact on the transmit beams illustrated through comparison between non-adjusted SURF and fundamental imaging transmit fields, for homogeneous and inhomogeneous medium. %
	%
	%
	Top pane row: beam profiles in focus. On the left along the $x$ axis, on the right along the $y$ axis. %
	Second pane row from top: Transmit-beams, from left to right: SH, FH, SI($zx$), FI($zx$), SI($zy$), and FI($zy$) with the abbreviations explained in Table~}\ref{tab:short_names} %
	%
	($zx$) and ($zy$) refer to two planes perpendicular to the beam axis. %
	%
	%
	Third pane row from top: Beam-axis pulses, from left to right: SH, FH, SI, and FI. %
	Bottom two panes: beam profiles along the depth axis. Individual normalization on top and common normalization below. %
	%
	Line notations: \linedescriptionscoresoverviewfields{ }}
  \label{fig:fields_no_processing}
\end{figure}
%
%
%
%
\begin{table}[!bt]
\renewcommand{\tablename}{\sc TABLE} % Make figure captions say Fig. instead of Figure.
	\caption{\label{tab:short_names}Abbreviations used in the field descriptions of Fig.~\ref{fig:fields_no_processing} -- \ref{fig:beam_profiles}.}
	\centering
	\begin{tabular}{l c c c c}
		\toprule
			&             &\multicolumn{3}{c}{SURF} \\
			\cmidrule{3-5}
			&             & non- & time-shift & filter \\
			& fundamental & adjusted & adjusted & adjusted\\
		\midrule
			homogeneous & FH & SH & $\tau$SH & $\hh$SH\\
			inhomogeneous & FI & SI & $\tau$SI & $\hh$SI\\
		\bottomrule
	\end{tabular}
\end{table}
%
%
%
%
%-------------------------------------------------------------------
%\subsection{Adjusted SURF fields}
%-------------------------------------------------------------------
Figure \ref{fig:adjusted_beams} shows adjusted SURF beams simulated by homogeneous and inhomogeneous propagation with 6 different choices of suppression depths $z_a$, using both time-shift adjustment or filter-adjustment. The corresponding beam profiles along the $z$ axis are shown in Fig.~\ref{fig:beam_profiles}. %
%
The time-development of the SURF on-axis pulses as a function of depth are displayed in \ref{fig:axis_pulses} for the same $z_a$ set. These pulses are further inspected in Fig.~\ref{fig:time_pulses}, where their signatures at the $z_a$ depths are compared to the signatures in focus. %
%
%
%
{Side-by-side comparison of the adjusted beams for different $z_a$ is of particular significance, as it visualizes that for the selected $z_a$ depths, the adjustment methods are about as viable for inhomogeneous as for homogeneous media. }%
%

The {adjustment abilities in inhomogeneous media are further quantified by the} specific reverberation suppression beam quality ratios $Q_{z_a}$ corresponding to the shown beams. These are {displayed} for different suppression depths in Fig.~\ref{fig:scores} using time-shift adjustment and filter-adjustment as well as for non-adjusted and fundamental beams. This is done both for homogeneous medium and for propagation through the body-wall. The figure also shows general reverberation suppression quality ratios $Q$ for different time-shift adjustments for the same medium cases. %
%
%
%
%Table \ref{tab:line_notation} shows the line notations used in the figure. %
The specific quality ratio was defined in \cite{nasholm2011} as:%
%
\begin{align}
	Q_{z_a}\triangleq {\displaystyle \sum_{z=z_n}^{z_f} \sum_{r=0}^{\infty}\sum_{\theta=0}^{2\pi} E(\vx)}  \left/\   {\displaystyle \sum_{r=0}^{\infty}\sum_{\theta=0}^{2\pi} E(\vx)}\big|_{z=z_a}\right.,
	\label{eq:quality_measureD}
\end{align}
where $E(\vx)$ is the beam energy at the spatial coordinate $\vx$ described by the cylindrical coordinates $(z,r,\theta)$. The imaging depth region is within $z\in[z_n,z_f]$ and is equal to the focal region with the given aperture and imaging frequency. %
{The measure $Q_{z_a}$ illustrates how well the transmit beam is adapted to suppression of reverberations where the first scattering takes place at the depth $z = z_a$.} %
%
%
The general quality ratio was defined in \cite{nasholm2009} as:%
\begin{align}
	Q\triangleq {\displaystyle \sum_{z=z_n}^{z_f}\sum_{r=0}^{\infty}\sum_{\theta=0}^{2\pi} E(\myvecx)}\left/\ {\displaystyle \sum_{z=0}^{z_n} \sum_{r=0}^{\infty}\sum_{\theta=0}^{2\pi} E(\myvecx)}\right..
\label{eq:quality_measure_generalD}
\end{align}
%
%
{The measure $Q$ illustrates the ability of the transmit beam to suppress reverberations when the first scatterings are distributed within the near-field region $z\in[0,z_n]$.} % 
%
%
%
A related quality measure was applied in \cite{kaupang2011} for assessment of near-field echo suppression in tissue-harmonic imaging. %

%-------------------------------------------------------------------
%\section{Discussion}
%-------------------------------------------------------------------
%* Principles, relationships and generalizations\\
%* Explanation/clarification of anomalies\\

The body-wall model chosen causes severe aberration disturbance of the transmit fields. Its characteristics, as displayed in Table \ref{tab:body_wall}, may be compared to values found at body temperature in experimental measurements on human abdomen specimens done by Hinkelman \emph{et al.}\maybespace\cite{hinkelman:ab_wall_meas}, where the amplitude correlation length is within {1.3 to 2.9}\,mm, the amplitude RMS value within {2.9 to 3.5}\,dB, the time-delay correlation length within {24 to 64}\,ns, and the time-delay correlation length within {3.3 to 17}\,mm. Especially the time-delay RMS fluctuation is more severe for the aberrative model utilized here than for the cited measurements. %

The RMS magnitude of the time-delay fluctuations introduced by the utilized aberrating layer model are, as indicated in Table \ref{tab:body_wall}, $450$\,ns, while the accumulated time-delay between $s_+(\vr,t)$ and $s_-(\vr,t)$ needed to generate maximum SURF imaging gain following the model introduced in\maybespace\cite{nasholm2009}, is smaller: $\frac{1}{2f_\HH} = 142$\,ns. The SURF synthetic transmit-beam is however shown to be generated despite these great delays induced by the body-wall. Two factors explain this: %
%

{1}) The $s_+$ and $s_-$ propagate following the same path and therefore experience the same aberration time-delay at each spatial point. Therefore {this abberation delay }is canceled when forming the difference field $s_\Delta(\vr,t)=s_+(\vr,t)-s_-(\vr,t)$. %
% 
\begin{figure*}[!htb]
\renewcommand{\figurename}{Fig.} % Make figure captions say Fig. instead of Figure.
	%
	\ \includegraphics[scale=.6]{fig_4}%
	\:\includegraphics[scale=.6]{fig_12}\\%
	\includegraphics[scale=.6]{fig_28}%
	\includegraphics[scale=.6]{fig_52}\\%
	\includegraphics[scale=.6]{fig_80}%
	\:\ \includegraphics[scale=.6]{fig_104}% 
	%
	%
	\caption[]{{Adjusted SURF beam cuts for $z_a=1.5$\:cm (top left pane group), $z_a=4.5$\:cm (top right), $z_a=10.5$\:cm (middle left), $z_a=19.5$\:cm (middle right), $z_a=30.5$\:cm (bottom left), and $z_a=39$\:cm (bottom right). Abbreviation codes shown in Table }\ref{tab:short_names}{  %
	%
	%
	Beams, from left to right within each $z_a$ pane group: $\hh$SH, $\tau$SH, $\hh$S{I}($zx$), $\tau$S{I}($zx$), $\hh$S{I}($zy$), and $\tau$S{I}($zy$). %
	%
	The bottom-right colorbar indicates the 35\:dB dynamic range\maybehighlight{ that is used for all the above beam cuts.}}}
	\label{fig:adjusted_beams}%
\end{figure*}

\begin{figure*}[!htb]
\renewcommand{\figurename}{Fig.} % Make figure captions say Fig. instead of Figure.
	%
	\begin{tabular}{c c c}
	\includegraphics[scale=.6]{fig_6}%
	& %
	\includegraphics[scale=.6]{fig_14}%
	& %
	\includegraphics[scale=.6]{fig_30}%
	\\%
	\hline%
	\noalign{\smallskip}
	\includegraphics[scale=.6]{fig_54}%
	& %
	\includegraphics[scale=.6]{fig_82}%
	& %
	\includegraphics[scale=.6]{fig_106}% 
	\end{tabular}
	%
	%
	\caption[]{{On-axis SURF beam profiles as function of depth $z$ for $z_a=1.5$\:cm (top left pane group), $z_a=4.5$\:cm (top right), $z_a=10.5$\:cm (middle left), $z_a=19.5$\:cm (middle right), $z_a=30.5$\:cm (bottom left), and $z_a=39$\:cm (bottom right). 
	%
	Line descriptions: }\linedescriptions
	%
	}
	\label{fig:beam_profiles}%
\end{figure*}

\begin{figure*}[!htb]
\renewcommand{\figurename}{Fig.} % Make figure captions say Fig. instead of Figure.
	%
	\begin{tabular}{c c}
	\includegraphics[scale=.6]{fig_5}%
	& \includegraphics[scale=.6]{fig_13}%
	\\%
	\hline%
	\noalign{\smallskip}
	\includegraphics[scale=.6]{fig_29}%
	& \includegraphics[scale=.6]{fig_53}%
	\\%
	\hline%
	\noalign{\smallskip}
	\includegraphics[scale=.6]{fig_81}%
	&\ \includegraphics[scale=.6]{fig_105}% 
	\end{tabular}
	%
	%
	\caption[]{{On-axis SURF pulses as function of depth $z$ and time $t$ for $z_a=1.5$\:cm (top left pane group), $z_a=4.5$\:cm (top middle), $z_a=10.5$\:cm (top right), $z_a=19.5$\:cm (bottom left), $z_a=30.5$\:cm (bottom middle), and $z_a=39$\:cm (bottom right). Abbreviation codes shown in Table }\ref{tab:short_names}{  %
	%
	%
	Beams, from left to right within each $z_a$ pane group: $\hh$S{H}, $\tau$S{H}, $\hh$S{I}, and $\tau$S{I}. %
	%
	}}
	\label{fig:axis_pulses}%
\end{figure*}

\begin{figure*}[!htb]
\renewcommand{\figurename}{Fig.} % Make figure captions say Fig. instead of Figure.
	%
	\centering
	\begin{tabular}{c c c}
	\includegraphics[width=.31\textwidth]{fig_7}%
	& %
	\includegraphics[width=.31\textwidth]{fig_15}%
	& %
	\includegraphics[width=.31\textwidth]{fig_31}%
	\\%
	\noalign{\smallskip}
	\hline%
	\noalign{\smallskip}
	\noalign{\smallskip}
	\includegraphics[width=.31\textwidth]{fig_55}%
	& %
	\includegraphics[width=.31\textwidth]{fig_83}%
	& %
	\includegraphics[width=.31\textwidth]{fig_107}% 
	\end{tabular}
	%
	%
	\caption[]{{On-axis pressure in focus (bottom sub-panes of each pair) and at $z_a$ (top sub-panes of each pair) as function of time $t$ for $z_a=1.5$\:cm (top left pane group), $z_a=4.5$\:cm (top right), $z_a=10.5$\:cm (middle left), $z_a=19.5$\:cm (middle right), $z_a=30.5$\:cm (bottom left), and $z_a=39$\:cm (bottom right). 
	%
	Line notation as in Fig.~}\ref{fig:beam_profiles}.
	%
	}
	\label{fig:time_pulses}%
\end{figure*}
%
%
\begin{figure}[!bth]
\renewcommand{\figurename}{Fig.} % Make figure captions say Fig. instead of Figure.
\centering
	\includegraphics[width=\maybehalffigure]{fig_scorefigs_1}\\ %
	\includegraphics[width=\maybehalffigure]{fig_scorefigs_2}%
	\caption[]{%
	Reverberation suppression beam quality measures with the specific ratios $Q_{z_a}$ {as a function of $z_a$} in the {top} pane and the general $Q$ {as a function of the adjustment time-shift $\tau_a$} in the {bottom} pane. %
	%
	\linedescriptionssecfig{ }
	 }%
	\label{fig:scores}%
\end{figure}
%
%
{2}) The LF part of each SURF pulse complex experience the same aberration delay during propagation. %
%
%
Therefore the HF pulse remains within roughly the same LF pressure as when propagating without aberration. This is illustrated in Fig.~\ref{fig:total_pulse_samples}, where the total propagating SURF pulse complexes with positive and negative LF polarities are sampled at one depth within the body-wall. %

The SURF transmit-beam in the homogeneous medium has suppressed focal sidelobes compared to the fundamental transmit-beam, while the mainlobe stays unchanged. The suppression is only around $1$\,dB for the first sidelobe, while it increases to around $3$\,dB for the second sidelobe. In the inhomogeneous medium case, the sidelobes are also suppressed for SURF compared to fundamental, however about half as much as for {the} homogeneous medium. %

Figure~\ref{fig:scores} indicates that, both for homogeneous and inhomogeneous medium, the time-shift adjustment gives superior, more robust and more predictable specific reverberation suppression quality ratio $Q_{z_a}$ than the filter-shift adjustment for depths where $z_a{>}8$\,mm, while the ratio is higher for filter-adjustment for shallower depths. The time-shift adjusted beams have $Q_{z_a}$ around $17$\,dB above $Q_{z_a}$ for the non-adjusted SURF beam, both for homogeneous and inhomogeneous medium. %

The $Q_{z_a}$ ratio calculated in the homogeneous medium for the time-shift adjusted beams, the non-adjusted SURF beams and the fundamental beams align well with what was found for the same 2-D axisymmetric simulations in \cite{nasholm2011}. The filter-adjusted SURF beams do however show worse and more unstable $Q_{z_a}$ in the present 3-D simulations than in the referred work. This is due to the averaging over the region $S$ which is used here on both $s_+(\vr,t)$ and $s_-(\vr,t)$ before $\hh_{z_a}$ is calculated. Then the canceling of temporal parts of the pulses which vary the most over the region $S$, that is the edge wave parts of the pulse, are less suppressed. The amplitude of these edge waves may be decreased if apodization is applied to the surface of pulse transmission\maybespace\cite{harris:transient_field-81, tabei:JASA-03}. That way the lateral region where $\hh_{z_a}$ generates high suppression might get larger. %

For $z_a\gtrsim8$\,mm, $Q_{z_a}$\:is lower both for the homogeneous and inhomogeneous media for time-shift, compared to for filter-adjustment. %
%
Therefore use of time-shift is preferred when $z_a$ is {far from} the transducer. %
{This is also reflected in the $z_a=39$\:cm section of Fig.~}\ref{fig:adjusted_beams} both with and without the body-wall. There significant in-focus field degradation is observed for the filter-adjustment beam as opposed to for the time-shift adjusted beam. %

{%
Ultrasound images of patients whose anatomy cause noisy images, for example due to obesity, are likely to be blurred owing to both aberration and reverberation noise. }%
%

{The repetitive nature of reverberation noise makes the receive signals correlated both in the temporal and the spatial directions. This may obstruct estimation of the filter that is used in aberration correction schemes like }\cite{masoy:JASA-05}. %
%
By suppression of reverberation noise, estimation of the aberration correction filter is thus enhanced, and therefore the aberration correction itself is likely to get more accurate. %
%
Instruments that may combine aberration correction with reverberation suppression are thus, compared to instruments with aberration correction alone, likely to produce images of enhanced quality also regarding the wave-front aberration noise. %

%
%- Speculation about research that might be done in the future to build upon results in the present paper. NO remarks about what authors intend to do next:

    %* Definition of conditions under which conclusions apply\\
    %* Agreement or disagreement with previous work\\
    %* Theoretical implications\\
    %* Possible applications\\
    %* Further thought or research\\
    %* Summary/conclusion -- significance of work\\
\begin{figure}[!tb]
\renewcommand{\figurename}{Fig.} % Make figure captions say Fig. instead of Figure.
	\centering
  	\includegraphics[width=\maybehalffigure]{pulse_samples}
	\caption{Propagating SURF pulse complexes sampled in the ($zx$) plane at the depth $z=6$\,mm. Horizontal axis: time, vertical axis: lateral direction. Top and middle panes: total SURF pulse complexes with opposite LF polarities. Bottom pane: extracted HF difference pulse $s_\Delta(x,t)$.}
	\label{fig:total_pulse_samples}
\end{figure}%

%-------------------------------------------------------------------
\section{Concluding remarks}
%-------------------------------------------------------------------
This work indicates through computer simulations that generation a SURF reverberation suppression synthetic transmit-beam is feasible in spite of the propagating waves being distorted by an inhomogeneous medium which emulates a strongly aberrating body-wall. Generation of the beam is hence also likely to be possible when utilizing body-wall models producing less severe aberrations corresponding to what was measured {e.g.} in \cite{hinkelman:ab_wall_meas}. %
%

The post-processing suppression depth adjustment methods, as previously implemented in \cite{nasholm2011} for a homogeneous medium, are here shown to be feasible on the simulated aberrated fields {also} within an inhomogeneous medium. %for the purpose of being able to shadow noise due the first multiple scattering by some especially strong object or interface located at the depth $z_a$. %

%The benefit of utilizing the general operator-shift processing method instead of the time-shift based one is largest for shallow depths also . %
%
%The conclusions regarding the reverberation suppression abilities are based on synthetic SURF transmit-beams after adjustment.

%A strength of the method is that%
%

%(- Not repeat abstract, or subject\\
%- Review principal results, and where they emerged\\
%- Broad discussion about possible implications\\

Further research of interest within the field also includes \emph{in vivo} imaging through true body-walls using the investigated methods and also performance comparisons with tissue harmonic imaging, especially when utilizing pulse inversion. %A first step is to perform the tests in a water-tank using some aberrating medium within the first part of the propagation path, as well as imaging tests on aberration-emulating phantoms. %
%
%

{Non-adjusted and adjusted beams could also be generated from SURF transmit fields measured by a hydrophone in a water-tank. Beam generation with a inhomogeneous body-wall setup may be emulated using some aberrating material within the tank, \emph{e.g}\ as done by use of gel layers in }\cite{jing:JASA-07}. %
%
%
It is furthermore desirable to perform{ }numerical propagation simulations using a tool that properly handles multiple scattering (as does not {the} forward-propagation method used here) in order to further quantify the reverberation suppression ability of the methods within an aberrative medium, as well as doing comparisons to the performance of tissue harmonic imaging using the same setup. %
%

{We expect the family of SURF reverberation suppression methods to come out as versatile techniques to enhance ultrasound image reconstruction also in clinical settings.}
%For an ultrasound imaging system capable of transmitting dual-frequency SURF pulses, both 
%
%

\bibliographystyle{IEEEtran}
{\footnotesize\bibliography{SURF}}
\iffalse
{\footnotesize

% Generated by IEEEtran.bst, version: 1.12 (2007/01/11)
\begin{thebibliography}{10}
\providecommand{\url}[1]{#1}
\csname url@samestyle\endcsname
\providecommand{\newblock}{\relax}
\providecommand{\bibinfo}[2]{#2}
\providecommand{\BIBentrySTDinterwordspacing}{\spaceskip=0pt\relax}
\providecommand{\BIBentryALTinterwordstretchfactor}{4}
\providecommand{\BIBentryALTinterwordspacing}{\spaceskip=\fontdimen2\font plus
\BIBentryALTinterwordstretchfactor\fontdimen3\font minus
  \fontdimen4\font\relax}
\providecommand{\BIBforeignlanguage}[2]{{%
\expandafter\ifx\csname l@#1\endcsname\relax
\typeout{** WARNING: IEEEtran.bst: No hyphenation pattern has been}%
\typeout{** loaded for the language `#1'. Using the pattern for}%
\typeout{** the default language instead.}%
\else
\language=\csname l@#1\endcsname
\fi
#2}}
\providecommand{\BIBdecl}{\relax}
\BIBdecl

\bibitem{nasholm2009}
S.~P. Näsholm, R.~Hansen, S.-E. Måsøy, T.~F. Johansen, and B.~A.~J. Angelsen,
  ``Transmit beams adapted to reverberation noise suppression using
  dual-frequency {SURF} imaging,'' \emph{IEEE Trans.\ Ultrason.\ Ferroelectr.,\
  Freq.\ Control}, vol.~56, no.~10, pp. 2124--2133, 2009.

\bibitem{nasholm2011}
S.~P. Näsholm, R.~Hansen, and B.~A.~J. Angelsen, ``Post-processing enhancement
  of reverberation-noise suppression in dual-frequency {SURF} imaging,''
  \emph{IEEE Trans.\ Ultrason.\ Ferroelectr.,\ Freq.\ Control}, vol.~58, no.~2,
  pp. 338--348, 2011.

\bibitem{hinkelman:pulse_dist_meas-98}
L.~M. Hinkelman, T.~D. Mast, L.~A. Metlay, and R.~C. Waag, ``The effect of
  abdominal wall morphology on ultrasonic pulse distortion. {P}art {I}:
  Measurements,'' \emph{J. Acoust. Soc. Am.}, vol. 104, no.~6, pp. 3635--3649,
  1998.

\bibitem{keogh:real_or_artifact-01}
C.~F. Keogh and P.~L. Cooperberg, ``Is it real or is it an artifact,''
  \emph{Ultrasound Q.}, vol.~17, no.~4, pp. 201--210, Dec. 2001.

\bibitem{mahmutyazicioglu:transabdominal-05}
K.~Mahmutyazicioglu, H.~A. Tanriverdi, H.~Ozdemir, A.~Barut, H.~Davsanci, and
  S.~Gundogdu, ``Transabdominal pulse inversion harmonic imaging improves
  assesment of ovarian morphology in virgin patients with pcos: comparison with
  conventional b-mode sonography,'' \emph{Eur. J. Radiol.}, vol.~53, no.~2, pp.
  208--286, Feb. 2005.

\bibitem{scanlan:artifacts-91}
K.~A. Scanlan, ``Sonographic artifacts and their origins,'' \emph{Am. J.
  Roentgenol.}, vol. 156, no.~6, pp. 1267--1272, Jun. 1991.

\bibitem{shmulewitz:quality_factors}
A.~Shmulewitz, S.~A. Teefey, and B.~S. Robinson, ``Factors affecting image
  quality and diagnostic efficacy in abdominal sonography: A prospective study
  of 140 patients,'' \emph{J. Clin. Ultrasound}, vol.~21, no.~9, pp. 623--630,
  1993.

\bibitem{Pinton2011}
G.~F. Pinton, G.~E. Trahey, and J.~J. Dahl, ``Sources of image degradation in
  fundamental and harmonic ultrasound imaging using nonlinear, full-wave
  simulations,'' \emph{IEEE Trans.\ Ultrason.\ Ferroelectr.,\ Freq.\ Control},
  vol.~58, no.~4, pp. 754--765, Apr. 2011.

\bibitem{Pinton2011erratum}
------, ``Erratum: Sources of image degradation in fundamental and harmonic
  ultrasound imaging: a nonlinear, full-wave, simulation study [{A}pr.\ 11
  754--765],'' \emph{IEEE Trans.\ Ultrason.\ Ferroelectr.,\ Freq.\ Control},
  vol.~58, no.~6, pp. 1272--1283, Jun. 2011.

\bibitem{angelsen:ultrasoundbook_II}
B.~Angelsen, \emph{Ultrasound imaging. Waves, signals and signal
  processing}.\hskip 1em plus 0.5em minus 0.4em\relax Trondheim: Emantec, 2000,
  vol.~II, http://www.ultrasoundbook.com.

\bibitem{Hansen2011}
R.~Hansen, S.-E. M{\aa}s{\o}y, T.~A. Tangen, and B.~A. Angelsen, ``Nonlinear
  propagation delay and pulse distortion resulting from dual frequency band
  transmit pulse complexes,'' \emph{J.\ Acoust.\ Soc.\ Am.}, vol. 129, no.~2,
  pp. 1117--1127, 2011.

\bibitem{Masoy2010}
S.-E. M{\aa}s{\o}y, {\O}.~Standal, J.~M. Deibele, S.~P. N\"{a}sholm,
  B.~Angelsen, T.~F. Johansen, T.~A. Tangen, and R.~Hansen, ``Nonlinear
  propagation acoustics of dual--frequency wide--band excitation pulses in a
  focused ultrasound system,'' \emph{J.\ Acoust.\ Soc.\ Am.}, vol. 128, no.~5,
  pp. 2695--2703, 2010.

\bibitem{hansen2010}
R.~Hansen, S.-E. Måsøy, T.~F. Johansen, and B.~A. Angelsen, ``Utilizing dual
  frequency band transmit pulse complexes in medical ultrasound imaging,''
  \emph{J. Acoust. Soc. Am.}, vol. 127, no.~1, pp. 579--587, 2010.

\bibitem{deng:dual_frequency}
C.~X. Deng, F.~L. Lizzi, A.~Kalisz, A.~Rosado, R.~H. Silverman, and D.~J.
  Coleman, ``Study of ultrasonic contrast agents using a dual-frequency band
  technique,'' \emph{Ultrasound Med.~Biol.}, vol.~26, pp. 819--831, 2000.

\bibitem{angelsen_hansen:ieee_proc07}
B.~A.~J. Angelsen and R.~Hansen, ``{SURF} imaging -- a new method for
  ultrasound contrast agent imaging,'' in \emph{Proc. {IEEE} Ultrason. Symp.},
  New York, NY, Oct. 2007, pp. 531--541.

\bibitem{bouakaz:radial_modulation}
A.~Bouakaz, M.~Versluis, J.~Borsboom, and N.~de~Jong, ``Radial modulation of
  microbubbles for ultrasound contrast imaging,'' \emph{IEEE Trans.\ Ultrason.\
  Ferroelectr.,\ Freq.\ Control}, vol.~54, no.~11, pp. 2283--2290, Nov. 2007.

\bibitem{masoy:SURF_in_vivo}
S.-E. Måsøy, Ø.~Standal, P.~Näsholm, T.~F. Johansen, R.~Hansen, and
  B.~Angelsen, ``{SURF} imaging: In vivo demonstration of an ultrasound
  contrast agent detection technique,'' \emph{IEEE Trans.\ Ultrason.\
  Ferroelectr.,\ Freq.\ Control}, vol.~55, no.~5, pp. 1112--1121, 2008.

\bibitem{hansen2009}
R.~Hansen and B.~A.~J. Angelsen, ``{SURF} imaging for contrast agent
  detection,'' \emph{IEEE Trans.\ Ultrason.\ Ferroelectr.,\ Freq.\ Control},
  vol.~56, no.~2, pp. 280--290, 2009.

\bibitem{emmer2009}
M.~Emmer, H.~J. Vos, M.~Versluis, and N.~D. Jong, ``Radial modulation of single
  microbubbles,'' \emph{IEEE Trans.\ Ultrason.\ Ferroelectr.,\ Freq.\ Control},
  vol.~56, no.~11, pp. 2370--2379, 2009.

\bibitem{frijlink2008}
M.~Frijlink, H.~Kaupang, T.~Varslot, and S.-E. Masoy, ``Abersim: {A} simulation
  program for 3{D} nonlinear acoustic wave propagation for arbitrary pulses and
  arbitrary transducer geometries,'' in \emph{IEEE Ultrasonics Symp. Proc.},
  Nov. 2008, pp. 1282--1285.

\bibitem{varslot:propagation}
T.~Varslot and G.~Taraldsen, ``Computer simulation of forward wave propagation
  in soft tissue,'' \emph{IEEE Trans.\ Ultrason.\ Ferroelectr.,\ Freq.\
  Control}, vol.~52, no.~9, pp. 1473--1482, 2005.

\bibitem{varslot:aberr_nonlin_wave_prop}
T.~Varslot, S.-E. Måsøy, T.~F. Johansen, and B.~Angelsen, ``Aberration in
  nonlinear acoustic wave propagation,'' \emph{IEEE Trans.\ Ultrason.\
  Ferroelectr.,\ Freq.\ Control}, vol.~54, no.~3, pp. 470--479, Mar. 2007.

\bibitem{masoy:correction-aberration}
S.-E. Måsøy, T.~F. Johansen, and B.~Angelsen, ``Correction of ultrasonic wave
  aberration with a time delay and amplitude filter,'' \emph{J. Acoust. Soc.
  Am.}, vol. 113, no.~4, pp. 2009--2020, Apr. 2003.

\bibitem{abersim_manual}
H.~Kaupang, ``Abersim 2.x reference manual with tutorials,'' Norwegian
  University of Science and Technology, Trondheim, Norway, Tech. Rep., 2008.

\bibitem{kaupang2011}
H.~Høilund-Kaupang and S.-E. Måsøy, ``Transmit beamforming for optimal
  second-harmonic generation,'' \emph{IEEE Trans.\ Ultrason.\ Ferroelectr.,\
  Freq.\ Control}, vol.~58, no.~8, pp. 1559--1569, Aug. 2011.

\bibitem{hinkelman:ab_wall_meas}
L.~M. Hinkelman, D.-L. Liu, L.~A. Metlay, and R.~C. Waag, ``Measurements of
  ultrasonic pulse arrival time and energy level variations produced by
  propagation through abdominal wall,'' \emph{J. Acoust. Soc. Am.}, vol.~95,
  no.~1, pp. 530--541, 1994.

\bibitem{harris:transient_field-81}
G.~R. {Harris}, ``{Transient field of a baffled planar piston having an
  arbitrary vibration amplitude distribution },'' \emph{J. Acoust. Soc. Am.},
  vol.~70, pp. 186--204, Jul. 1981.

\bibitem{tabei:JASA-03}
M.~Tabei, T.~D. Mast, and R.~C. Waag, ``Simulation of ultrasonic focus
  aberration and correction through human tissue,'' \emph{J. Acoust. Soc. Am.},
  vol. 113, no.~2, pp. 1166--1176, 2003.

\bibitem{masoy:JASA-05}
S.-E. Måsøy, T.~Varslot, and B.~Angelsen, ``Iteration of transmit-beam
  aberration correction in medical ultrasound imaging,'' \emph{J. Acoust. Soc.
  Am.}, vol. 117, no.~1, pp. 450--461, 2005.

\bibitem{jing:JASA-07}
Y.~Jing and R.~O. Cleveland, ``Modeling the propagation of nonlinear
  three-dimensional acoustic beams in inhomogeneous media,'' \emph{J. Acoust.
  Soc. Am.}, vol. 122, no.~3, pp. 1352--1364, 2007.

\end{thebibliography}
}
\fi

\begin{IEEEbiography}[%
	%
	%
	{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{svenpn-bw}}]{Sven Peter Näsholm}\footnotesize 
was born in Örnsköldsvik, Sweden, in 1975. He received his M.Sc. degree in engineering physics in 2002 from Umeå University, Sweden. %
In 2008, he successfully defended his Ph.D. thesis entitled ``Ultrasound beams for enhanced image quality'' at the Norwegian University of Science and Technology, Trondheim, Norway. %
In 2009, he joined the Digital Signal Processing and Image Analysis group as a pos-doctoral fellow at the Department of Informatics, University of Oslo, Norway. %
His research interests are within the fields of sonar and ultrasound imaging including nonlinear effects, transducer design, field simulation, and acoustic noise suppression.
\end{IEEEbiography}
%
%
%
\begin{IEEEbiography}[%
	%
{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{bjorn-biographypic}}]{Bjørn A. J. Angelsen}\footnotesize
was born 1946 in Vestvågøy, Norway. He received a MEE in 1971 from the Norwegian University of Science and Technology, Trondheim, and a Ph.D. from the same University in 1977. His Ph.D. work was on Doppler ultrasound measurement of blood velocities and
flow in the heart and the large arteries. He is a professor of Medical Imaging at the same university since 1983. In 1977--78 he was a visiting Post Doc at University of California, Berkeley, and Stanford Research Institute, Palo Alto. He has been strongly involved in development of cardiac ultrasound imaging instruments in collaboration with Vingmed Ultrasound, now GE Vingmed Ultrasound. He has written textbooks on ultrasound cardiac Doppler measurements and theoretical ultrasound acoustics, and holds several patents in the field of ultrasound imaging.
\end{IEEEbiography}
\vfill
\end{document}

### 2013-05-15 FFI-Simulator/ffi-simulator.tex
...
la2pdf commit - 2013:05:15	18:25:05	(Wed15.May)

@@
 
-{\Large Design Suggestion}\\[0.2cm]
+{\Large Design Suggestion} v2\\[0.2cm]
 {\itshape for a}\\[0.2cm]
@@
 
-The way we see it, there are two possible approaches to implementing the mentioned simulator (figure \ref{implementation}). One can either
+In the original specification we put forth two possible approaches to implementing the mentioned simulator (figure \ref{implementation}). We envisioned to either
 
@@
 \begin{itemize0}
-\item use OpenGL to produce depth-maps of the scene, and CUDA to illuminate it as a sonar would, or
+\item use OpenGL to produce depth-maps of the scene, and CUDA/OpenCL to illuminate it as a sonar would, or
 \item use OpenGL to render the full scene with lighting.
@@
 
-In either case the simulator will take a seafloor and target model as input, and generate a simulated sonar image.
-To offload the CPU further, the GPU can then proceed to compute the correlation between the template and the isolated image segment, and an image mask.
-This should be fairly easy to implement, will compute efficiently on the GPU, and allows us to avoid sending each template back to the CPU over the limited bandwidth CPU$\leftrightarrow$GPU bus.
+In each of these cases the simulator takes a seafloor and target model as input, and generates a simulated sonar image.
+
+We believe the first alternative is the more customizable of the two, and the second being the least complex to implement.
+
+To further offload the CPU, the GPU could proceed to compute the correlation between the template and the isolated image segment, and an image mask.
+The advantages included a fairly straight forward implementation that would compute efficiently on the GPU,
+
+ and allows us to avoid sending each template back to the CPU over the limited bandwidth CPU$\leftrightarrow$GPU bus.
 Note that these steps will have to be carried out in CUDA regardless of the approach selected.
@@
 When we wish to set up a 3D scene, move objects and cameras around, and render it using a GPU, it is better to use e.g.
-OpenGL than CUDA.
+OpenGL than CUDA/OpenGL.
 OpenGL was made for this specific task, is cross-platform, feature-rich and well-matured.
@@
 If desired, it can be modified to handle sound propagation in non-ideal mediums, penetrable objects, signal multipath, diffraction and phase-dependent effects such as speckle.
-However, this will also lead to increased code complexity and development time, and  will naturally increase.
-However, we expect that such features will be more realizable in CUDA than OpenGL.
+The drawback is the added code complexity and development time.
 
@@
 
+\newpage
+
+% \section{Documentation}
+
+% \input{ffi-documentation}
 \end{document}

### 2013-05-22 FFI-Simulator/ffi-simulator.tex
la2pdf commit - 2013:05:22	11:55:24	(Wed22.May)

@@
 
-{\Large Design Suggestion} v2\\[0.2cm]
+{\Large Design Suggestion}\\[0.2cm]
 {\itshape for a}\\[0.2cm]
@@
 \small
-\today
+Revision 2, \today
 \end{center}
@@
 
+As detailed below, in this revision we have decided to opt for slightly more flexibile design to better be able to handle effects such as geometric layover.
+% This is to better be able to simulate geometrical sonar imaging effects such as layover. We will describe both the original suggestions and the updated design here.
+
+\subsection{The Initial Suggestion}
+
 In the original specification we put forth two possible approaches to implementing the mentioned simulator (figure \ref{implementation}). We envisioned to either
@@
 \begin{figure}[b]
-\begin{narrow}{-2.5cm}{-2.5cm}
+\begin{narrow}{-2.3cm}{-2.3cm}
 \graphicsAI[drawing,width=\linewidth]{gfx/simulator.svg}
@@
 
-In each of these cases the simulator takes a seafloor and target model as input, and generates a simulated sonar image.
+In each of these cases the simulator takes a seafloor and target model as input, and generates a simulated sonar image, as well as perform template correlation if desired.
+For the first step of setting up the 3D scene, move objects and cameras around, and render it using a GPU, we initially believed OpenGL to be the better choice.
+OpenGL was made for this specific task, is cross-platform, feature-rich, well-matured and tuned to peak performance to meet the heavy needs of the gaming industry.
+
+But can we accurately simulate a sonar system using a framework designed to simulate optics?
+
+\subsection{Geometrical Layover}
+
+When we wish to , it is better to use e.g. OpenGL than CUDA/OpenGL.
 
@@
 \subsection{Using only OpenGL for the simulator?}
-
-When we wish to set up a 3D scene, move objects and cameras around, and render it using a GPU, it is better to use e.g.
-OpenGL than CUDA/OpenGL.
-OpenGL was made for this specific task, is cross-platform, feature-rich and well-matured.
-By looking at the graphics quality and frame-rates displayed by modern OpenGL-based games it should also be fairly straight-forward to get the performance we need out of this library.
 

### 2013-05-22 FFI-Simulator/ffi-simulator.tex
la2pdf commit - 2013:05:23	00:48:31	(Thu23.May)

@@
 
-As detailed below, in this revision we have decided to opt for slightly more flexibile design to better be able to handle effects such as geometric layover.
+As detailed below, in this revision we have decided to add a couple alternative designs that is more flexible and able to handle effects such as geometric layover and multipath.
 % This is to better be able to simulate geometrical sonar imaging effects such as layover. We will describe both the original suggestions and the updated design here.
@@
 
-\subsection{Geometrical Layover}
+\subsection{Layover and multipath}
+
+When using OpenGL, there is
 
@@
 \subsection{Using only OpenGL for the simulator?}
+
+\begin{figure}[b]
+\begin{narrow}{-2.3cm}{-2.3cm}
+\graphicsAI[drawing,width=\linewidth]{gfx/simulation_setup.png}
+\end{narrow}
+\caption{Two possible implementation strategies.}\label{implementation}
+\end{figure}
 

### 2013-05-22 FFI-Simulator/ffi-simulator.tex
la2pdf commit - 2013:05:23	01:21:41	(Thu23.May)

@@
 
+\subsection{Simulating a sonar scenario using optics}
+
+\begin{figure}[b]
+\begin{narrow}{-2.3cm}{-2.3cm}
+\graphicsAI[drawing,width=\linewidth]{gfx/simulation_setup.png}
+\end{narrow}
+\caption{Two possible implementation strategies.}\label{simulation_setup}
+\end{figure}
+
+As a preliminary study we created a 3D model of a manta mine and a cylinder located on a flat ground surface (fig.
+\ref{simulation_setup}).
+Here we placed the camera 6 meters above the surface looking down on it.
+The light source was put at a 50\;m distance and set up without falloff, and the light hits the centre of the scene at an angle of 10$^\circ$.
+The manta mine was made according to specifications with 48\;cm height and 98\;cm ground diameter, the same applies to the cylinder.
+The distance between the two objects is 1\;m.
+
+\begin{figure}[H]
+\begin{minipage}[c]{0.49\linewidth}
+\graphicsAI[drawing,width=\linewidth]{gfx/simulation_mask.png}
+\end{minipage}\mbox{}\hfill
+\begin{minipage}[c]{0.49\linewidth}
+\graphicsAI[drawing,width=\linewidth]{gfx/simulation_object.png}
+\end{minipage}
+\graphicsAI[drawing,width=\linewidth]{gfx/simulation_optical.png}
+\caption{Optical rendering.}\label{s}
+\end{figure}
+
 \subsection{Layover and multipath}
@@
 The risk of choosing an OpenGL-only implementation is therefore reduced, since we may easily switch to the CUDA implementation if an OpenGL prototype fails to produce the desired results.
-
-\subsection{Using only OpenGL for the simulator?}
-
-\begin{figure}[b]
-\begin{narrow}{-2.3cm}{-2.3cm}
-\graphicsAI[drawing,width=\linewidth]{gfx/simulation_setup.png}
-\end{narrow}
-\caption{Two possible implementation strategies.}\label{implementation}
-\end{figure}
 

### 2013-05-23 FFI-Simulator/ffi-simulator.tex
la2pdf commit - 2013:05:23	02:16:23	(Thu23.May)

@@
 The manta mine was made according to specifications with 48\;cm height and 98\;cm ground diameter, the same applies to the cylinder.
-The distance between the two objects is 1\;m.
+The distance between the two objects was 1\;m.
 
-\begin{figure}[H]
+\begin{figure}[b]
 \begin{minipage}[c]{0.49\linewidth}
 \graphicsAI[drawing,width=\linewidth]{gfx/simulation_mask.png}
+\centering\itshape Simulation mask
 \end{minipage}\mbox{}\hfill
@@
 \graphicsAI[drawing,width=\linewidth]{gfx/simulation_object.png}
+\centering\itshape Simulated object
 \end{minipage}
 \graphicsAI[drawing,width=\linewidth]{gfx/simulation_optical.png}
+\centering\itshape Combined image\vspace{5pt}
 \caption{Optical rendering.}\label{s}
@@
 
+To simulate this scene we combine a normal optical simulation with a mask image.
+The mask image scene is set up with a powerful light source and highly diffractive surfaces, which causes everything that is hit by light to obtain a high intensity value.
+Only direct light is processed, so shadows will come out as empty.
+In the scene with the simulated object material properties are adjusted to reflect a more reasonable amount of light, and a light insensitive background was chosen.
+The two images was finally combined by colour multiplication.
+
+Are such an approach accurate enough? Or do we need to consider layover, multipath, or other effects as well?
+
 \subsection{Layover and multipath}
 
-When using OpenGL, there is
-
-When we wish to , it is better to use e.g. OpenGL than CUDA/OpenGL.
-
-We believe the first alternative is the more customizable of the two, and the second being the least complex to implement.
-
-To further offload the CPU, the GPU could proceed to compute the correlation between the template and the isolated image segment, and an image mask.
-The advantages included a fairly straight forward implementation that would compute efficiently on the GPU,
-
- and allows us to avoid sending each template back to the CPU over the limited bandwidth CPU$\leftrightarrow$GPU bus.
-Note that these steps will have to be carried out in CUDA regardless of the approach selected.
+If layover and multipath are important, then the techniques used to simulate the previous images can not be used.
+Then we believe that a pure CUDA / OpenCL implementation is the better choice, but how these should be designed depends on the effects we wish to implement.
+For layover one could launch a thread for each pixel in the image, let it find all the points in the image that contributes with direct returns to that pixel, and sum the contributions up.
+For multipath a ray tracer customized for sound is likely the better option.
 
@@
 Sound$\rightarrow$light calibration & Not necessary  & Necessary(?) \\
-Multipath support                   & Possible       & Possible \\
+Multipath support                   & Possible       & Hard \\
 Speckle support                     & Possible       & Hard \\
@@
 
-We propose using OpenGL to simulate the scene, and model the array as lightsources.
-This will make it unsuitable for simulating speckle and diffraction effects, but a "simple'' simulator should probably not take such effects into account anyhow.
-Penetrable objects can be simulated in either case, although we are not sure whether the OpenGL framework can support this with ease.
-An OpenGL implementation will be highly portable, and believed to support the functionality needed to generate a good enough template image.
+Do we need to include effects such as layover and multipath in our simulations?
+
+If the answer to this is no, then we propose using a OpenGL to simulate the scene, and model the array as lightsources.
+Geometrical effects can be added in the post-processing using CUDA / OpenCL, but extending the simulator to support other sonar specific effects will likely be hard.
+This approach with also be very portable, and should not require too much effort.
+
+On the other hand, a CUDA / OpenCL implementation can handle anything.
+But does it need to?
+We should be fairly certain what we aim for here, since the various advanced effects calls for completely different implementation strategies.
 
@@
 The goal is to have an OpenGL prototype ready within a couple weeks, and from there decide which approach to go for.
-We propose monthly progress reports and meetings to discuss further
-development.
+We propose monthly progress reports and meetings to discuss further development.
 

### 2013-05-23 FFI-Simulator/ffi-simulator.tex
la2pdf commit - 2013:05:23	14:49:40	(Thu23.May)

@@
 
-As detailed below, in this revision we have decided to add a couple alternative designs that is more flexible and able to handle effects such as geometric layover and multipath.
+As detailed below, in this revision we have decided to add a couple of alternative designs that is more flexible and able to handle effects such as geometric layover and multipath.
 % This is to better be able to simulate geometrical sonar imaging effects such as layover. We will describe both the original suggestions and the updated design here.
@@
 If layover and multipath are important, then the techniques used to simulate the previous images can not be used.
-Then we believe that a pure CUDA / OpenCL implementation is the better choice, but how these should be designed depends on the effects we wish to implement.
+A CUDA / OpenCL implementation can handle this.
+but no implementation is best at everything.
+but how these should be designed depends on the effects we wish to implement.
 For layover one could launch a thread for each pixel in the image, let it find all the points in the image that contributes with direct returns to that pixel, and sum the contributions up.
@@
 The risk of choosing an OpenGL-only implementation is therefore reduced, since we may easily switch to the CUDA implementation if an OpenGL prototype fails to produce the desired results.
+%
 

### 2013-05-23 FFI-Simulator/ffi-simulator.tex
la2pdf commit - 2013:05:23	15:21:25	(Thu23.May)

@@
 
-But can we accurately simulate a sonar system using a framework designed to simulate optics?
+But can we accurately simulate a sonar system using a framework designed to simulate optics? We did a quick simulation to get a feel for this.
 
@@
 \begin{minipage}[c]{0.49\linewidth}
-\graphicsAI[drawing,width=\linewidth]{gfx/simulation_mask.png}
+\graphicsAI[drawing,width=\linewidth]{gfx/simMask0001.png}
 \centering\itshape Simulation mask
@@
 \begin{minipage}[c]{0.49\linewidth}
-\graphicsAI[drawing,width=\linewidth]{gfx/simulation_object.png}
+\graphicsAI[drawing,width=\linewidth]{gfx/simObject0001.png}
 \centering\itshape Simulated object
 \end{minipage}
-\graphicsAI[drawing,width=\linewidth]{gfx/simulation_optical.png}
+\graphicsAI[drawing,width=\linewidth]{gfx/simScene0001.png}
 \centering\itshape Combined image\vspace{5pt}
@@
 
-Are such an approach accurate enough? Or do we need to consider layover, multipath, or other effects as well?
+Is such an approach accurate enough? Or do we need to consider layover, multipath, or other effects as well?
 
@@
 If layover and multipath are important, then the techniques used to simulate the previous images can not be used.
-A CUDA / OpenCL implementation can handle this.
-but no implementation is best at everything.
+A CUDA / OpenCL implementation can handle this, but no implementation is best at everything.
 but how these should be designed depends on the effects we wish to implement.

### 2013-05-23 FFI-Simulator/ffi-simulator.tex
la2pdf commit - 2013:05:23	18:19:46	(Thu23.May)
la2pdf commit - 2013:05:23	18:50:15	(Thu23.May)

@@
 \end{narrow}
-\caption{Two possible implementation strategies.}\label{simulation_setup}
+\caption{Simulating a sonar in a optics render. The transmitter is modelled as a light source, and a camera is placed directly above the object looking down on them.}\label{simulation_setup}
 \end{figure}
@@
 
-To simulate this scene we combine a normal optical simulation with a mask image.
-The mask image scene is set up with a powerful light source and highly diffractive surfaces, which causes everything that is hit by light to obtain a high intensity value.
-Only direct light is processed, so shadows will come out as empty.
-In the scene with the simulated object material properties are adjusted to reflect a more reasonable amount of light, and a light insensitive background was chosen.
-The two images was finally combined by colour multiplication.
+To simulate this scene we combined a normal optical simulation with a mask image.
+The mask image scene was set up with a powerful light source and highly diffractive surfaces, which caused everything that was hit by light to obtain a high intensity value.
+Only direct light was processed, so shadows came out as empty.
+In the simulated object scene materials were adjusted to reflect a more reasonable amount of light, and a light insensitive background was chosen to avoid double shadows.
+The two images were finally combined using RGB multiplication.
 
-Is such an approach accurate enough? Or do we need to consider layover, multipath, or other effects as well?
+We believe these simulations to be similar to be similar to the ones initially supplied by FFI, made in Matlab.
+But are these models accurate enough?
+Or do we need to consider layover, multipath, or other effects as well?
 
@@
 
-If layover and multipath are important, then the techniques used to simulate the previous images can not be used.
-A CUDA / OpenCL implementation can handle this, but no implementation is best at everything.
-but how these should be designed depends on the effects we wish to implement.
-For layover one could launch a thread for each pixel in the image, let it find all the points in the image that contributes with direct returns to that pixel, and sum the contributions up.
+If layover and multipath are important, then we believe the techniques used to simulate the previous images can not be used.
+A CUDA / OpenCL implementation can handle this, but the proper way to design it depends on the effects we wish to implement.
+For layover one could launch a thread for each pixel in the image, which may proceed by finding all patches in the image that contributes by direct returns to that pixel (fig \ref{single_path}).
 For multipath a ray tracer customized for sound is likely the better option.
 
-In addition we see that each approach shares the same initial steps.
-The risk of choosing an OpenGL-only implementation is therefore reduced, since we may easily switch to the CUDA implementation if an OpenGL prototype fails to produce the desired results.
-%
-
-Our original idea was to use OpenGL to produce depth-maps of the object and (optionally) the seafloor, and then use CUDA to compute how a sonar array would illuminate this scene.
-However, since were only interested in pixel intensities, and sound may be treated as ``light'' with reasonable accuracy, we believe that OpenGL can be used to simulate the entire scene.
-This way we may support whatever OpenGL supports with little extra effort.
-The gaming industry demands speed, so such a simulator should also be very fast.
+\begin{figure}[b]
+\graphicsAI[drawing,width=\linewidth]{gfx/layover.svg}
+\caption{A CUDA / OpenCL implementation able to handle single path return.}\label{single_path}
+\end{figure}
 

### 2013-05-23 FFI-Simulator/ffi-simulator.tex
la2pdf commit - 2013:05:23	19:22:24	(Thu23.May)

@@
 Only direct light was processed, so shadows came out as empty.
-In the simulated object scene materials were adjusted to reflect a more reasonable amount of light, and a light insensitive background was chosen to avoid double shadows.
+In the simulated object scene materials were adjusted to reflect a more reasonable amount of light, and a light insensitive background with some added noise was chosen to avoid double shadows.
 The two images were finally combined using RGB multiplication.
 
-We believe these simulations to be similar to be similar to the ones initially supplied by FFI, made in Matlab.
+We believe these simulations to be similar to the ones initially supplied by FFI, which were made in Matlab.
 But are these models accurate enough?
@@
 If layover and multipath are important, then we believe the techniques used to simulate the previous images can not be used.
-A CUDA / OpenCL implementation can handle this, but the proper way to design it depends on the effects we wish to implement.
+In this case we are likely better off using CUDA / OpenCL from the beginning.
 For layover one could launch a thread for each pixel in the image, which may proceed by finding all patches in the image that contributes by direct returns to that pixel (fig \ref{single_path}).
-For multipath a ray tracer customized for sound is likely the better option.
+For multipath a ray tracer altered to handle some sound effects is likely the better option.
 
@@
 \graphicsAI[drawing,width=\linewidth]{gfx/layover.svg}
-\caption{A CUDA / OpenCL implementation able to handle single path return.}\label{single_path}
+\caption{A CUDA / OpenCL implementation able to handle direct returns and arbitrary geometries. Might approximate the effects of multipath and layover fairly well.}\label{single_path}
 \end{figure}
 
-The main advantage of the CUDA implementation lies in its flexibility.
-If desired, it can be modified to handle sound propagation in non-ideal mediums, penetrable objects, signal multipath, diffraction and phase-dependent effects such as speckle.
+We believe the added flexibility of a CUDA / OpenCL based approach will also make it easier to add new features to the implementation at a later stage, should this be deemed necessary.
+It could e.g.
+be altered to handle sound propagation in non-ideal mediums, penetrable objects, signal multipath, diffraction and phase-dependent effects such as speckle.
 The drawback is the added code complexity and development time.
 
-We have compared the two methods in Table \ref{comparison}.
+We also suggest using OpenCL these days, since it is a standard promoted by major players such as Intel and AMD, and because it allows the code to be almost as portable as CUDA.
+In terms of syntax and maturity the two standards are fairly similar.
+
+We have compared the two alternatives in Table \ref{comparison}.
 

### 2013-05-23 FFI-Simulator/ffi-simulator.tex
la2pdf commit - 2013:05:23	19:54:13	(Thu23.May)

@@
 
-If the answer to this is no, then we propose using a OpenGL to simulate the scene, and model the array as lightsources.
+If the answer to this is no, then we propose using a OpenGL to simulate the scene, and model the array as light sources.
 Geometrical effects can be added in the post-processing using CUDA / OpenCL, but extending the simulator to support other sonar specific effects will likely be hard.
-This approach with also be very portable, and should not require too much effort.
+On the upside this approach will be very portable, and should not require too much effort.
 
@@
 We propose monthly progress reports and meetings to discuss further development.
+See our specification on the next page.
 
 \newpage
+
+\section{Specification}
+
+\begin{tabular}[c]{l p{0.6\linewidth}}
+Development time       & 4-6 weeks per person  \\
+Our price              & 98\;000 NOK \\
+We deliver        & 
+\end{tabular}
+\begin{itemize0}
+\item A fully functional simulator with the features requested.
+\item Support up to 6 months after delivery.
+\item Through documentation.
+\item All rights to use, alter and redistribute the code in whatever way the customer see fit.
+\end{itemize0}
+\begin{tabular}[c]{l p{0.6\linewidth}}
+We reserve the right to &
+\end{tabular}
+\begin{itemize0}
+\item Use the code and ideas obtained during the development of this simulator for research purposes, but only with the consent of FFI.
+\end{itemize0}
 

### 2013-06-05 FFI-Simulator/ffi-simulator.tex

--- 2013-05-23 FFI-Simulator/ffi-simulator.tex
@@
 
-As a preliminary study we created a 3D model of a manta mine and a cylinder located on a flat ground surface (fig.
-\ref{simulation_setup}).
+As a preliminary study we created a 3D model of a manta mine and a cylinder located on a flat ground surface (Figure  \ref{simulation_setup}).
 Here we placed the camera 6 meters above the surface looking down on it.
@@
 Only direct light was processed, so shadows came out as empty.
-In the simulated object scene materials were adjusted to reflect a more reasonable amount of light, and a light insensitive background with some added noise was chosen to avoid double shadows.
+In the scene with the simulated object the materials were adjusted to reflect a more reasonable amount of light, and a light insensitive background with some added noise was chosen to avoid double shadows.
 The two images were finally combined using RGB multiplication.
@@
 
-If layover and multipath are important, then we believe the techniques used to simulate the previous images can not be used.
-In this case we are likely better off using CUDA / OpenCL from the beginning.
-For layover one could launch a thread for each pixel in the image, which may proceed by finding all patches in the image that contributes by direct returns to that pixel (fig \ref{single_path}).
-For multipath a ray tracer altered to handle some sound effects is likely the better option.
+If effects such as layover and multipath are important the techniques used to simulate the previous images are less attractive.
+In such a case we are likely better off using CUDA/OpenCL from the beginning.
+A possible implementation for simulating layover is to launch a thread for each pixel in the image, which may proceed by finding all patches in the image that contributes by direct returns to that pixel (Figure \ref{single_path}).
+For simulating multipath a custom ray tracer is likely a better option.
 
@@
 \graphicsAI[drawing,width=\linewidth]{gfx/layover.svg}
-\caption{A CUDA / OpenCL implementation able to handle direct returns and arbitrary geometries. Might approximate the effects of multipath and layover fairly well.}\label{single_path}
+\caption{A CUDA/OpenCL implementation able to handle direct returns and arbitrary geometries. Might approximate the effects of multipath and layover fairly well.}\label{single_path}
 \end{figure}
 
-We believe the added flexibility of a CUDA / OpenCL based approach will also make it easier to add new features to the implementation at a later stage, should this be deemed necessary.
+We believe the added flexibility of a CUDA/OpenCL based approach will also make it easier to add new features to the implementation at a later stage, should this be deemed necessary.
 It could e.g.
@@
 
-We have compared the two alternatives in Table \ref{comparison}.
+We have compared the OpenGL alternative to CUDA/OpenCL in Table \ref{comparison}.
 
@@
 
-If the answer to this is no, then we propose using a OpenGL to simulate the scene, and model the array as lightsources.
-Geometrical effects can be added in the post-processing using CUDA / OpenCL, but extending the simulator to support other sonar specific effects will likely be hard.
-This approach with also be very portable, and should not require too much effort.
+If the answer to this is no, then we propose using a OpenGL to simulate the scene, and model the array as light sources.
+Geometrical effects can be added in the post-processing using CUDA/OpenCL, but extending the simulator to support other sonar specific effects will likely be hard.
+On the upside this approach will be very portable, and should not require too much effort.
 
-On the other hand, a CUDA / OpenCL implementation can handle anything.
+A CUDA/OpenCL implementation can handle anything.
 But does it need to?
@@
 We propose monthly progress reports and meetings to discuss further development.
+See our specification on the next page.
 
@@
 
-\section{Ideas for paper}
+\section{Specification}
 
-\begin{itemize}
-\item With effects such as layover, shadows, limited resolution (detail retention capability), and complex object and surface shapes, computing the actual scenario based on the sonar image is hard.
-\item Better to perform fast, recursive simulations instead?
-\item Adaptive images in a video makes it much easier to understand what is going on. Need much better erfora
-\item Litits: massive storage, fixed templates, amplitude/noise sensitive
-\item Alternatives: Pattern recognition (hard to extract decent features, selection, fixed templates) and model-based (massive computational requirements)
-\item Why models? Hard to get accurate models from limted amount dat, targets not always availagme, variations in target/terrain.
-\item CAD models. Sufficient complexity? Materials? Transparency, characterisation, multilauers.
-\item Accuracy vs. comp. speed.
-\end{itemize}
+\begin{tabular}[c]{>{\bf}l p{0.6\linewidth}}
+Development time       & 4-6 weeks per person  \\
+Our price              & 98\;000 NOK
+\end{tabular}
 
-\newpage
+\begin{tabular}[c]{>{\bf}l p{0.6\linewidth}}
+We provide        &
+\end{tabular}
+\begin{itemize0}
+\item A fully functional simulator with the specifications mutually agreed upon.
+\item A thorough documentation.
+\item The customer obtains ownership of the code, including all rights to use, alter and redistribute it.
+\end{itemize0}
+\begin{tabular}[c]{>{\bf}l p{0.6\linewidth}}
+We reserve the right to &
+\end{tabular}
+\begin{itemize0}
+\item Use the code and ideas obtained during the development of this simulator for research purposes, but we refrain from publishing anything without a full consent from the customer.
+\end{itemize0}
+
+% \section{Ideas for paper}
+% 
+% \begin{itemize}
+% \item With effects such as layover, shadows, limited resolution (detail retention capability), and complex object and surface shapes, computing the actual scenario based on the sonar image is hard.
+% \item Better to perform fast, recursive simulations instead?
+% \item Adaptive images in a video makes it much easier to understand what is going on. Need much better erfora
+% \item Litits: massive storage, fixed templates, amplitude/noise sensitive
+% \item Alternatives: Pattern recognition (hard to extract decent features, selection, fixed templates) and model-based (massive computational requirements)
+% \item Why models? Hard to get accurate models from limted amount dat, targets not always availagme, variations in target/terrain.
+% \item CAD models. Sufficient complexity? Materials? Transparency, characterisation, multilauers.
+% \item Accuracy vs. comp. speed.
+% \end{itemize}
+% 
+% \newpage
 

### 2013-06-05 HRUIonGPU/jrnl_article.tex

@@
 
-%\documentclass[journal]{IEEEtran}
-\documentclass[11pt, draftcls, onecolumn]{IEEEtran}
+\documentclass[journal]{IEEEtran}
+%\documentclass[11pt, draftcls, onecolumn]{IEEEtran}
 
@@
 Image quality is investigated in videos of both a simulated phantom, and in an \textit{in-vivo} cardiac dataset.
-These results show that Capon beamforming is feasible for cardiac ultrasound imaging, providing images with improved lateral resolution and no local shift-variance effects, both in element-space and beamspace.
+These results show that Capon beamforming is feasible for cardiac ultrasound imaging, providing images with improved lateral resolution both in element-space and beamspace.
 \end{abstract}
@@
 % denne må utelates i IUS-artikkelen
-In recent work \cite{asen2012, Buskenes, Buskenes2013} we have applied a GPU-based Capon beamformer for both active sonar imaging and for cardiac ultrasound imaging.
+In recent work we have applied a GPU-based Capon beamformer for both active sonar imaging \cite{Buskenes, Buskenes2013} and cardiac ultrasound imaging \cite{asen2012}.
 Sonar has a less strict real-time requirement than medical ultrasound imaging, which means that most array configurations can be handled in real time using the GPU.
-Capon beamforming in medical ultrasound imaging, and in particular cardiac ultrasound imaging, demands frame rates that not even a high-end GPU can provide.
-Since the execution time is cubic with respect to the number of array elements, we need methods that reduce the computational complexity for large arrays.
-Various approaches to low-complexity Capon beamforming for medical ultrasound imaging have therefore been proposed in the literature \cite{Synnevag2011, Asl2012, Jensen2012, Kim}.
-In this paper we will make use of a beamspace transformation described by Nilsen and Hafizovic \cite{Nilsen2009}, in order to maintain high framerates as the number of elements increases.
+In \cite{asen2012} and in this paper we show that Capon beamforming applied in medical ultrasound imaging, and in particular cardiac ultrasound imaging, demands frame rates that not even a cluster of high-end GPUs can provide.
+Since the execution time of Capon beamforming is cubic with respect to the number of array elements, we need methods that reduce the computational complexity for large arrays.
+Various approaches to low-complexity Capon beamforming for medical ultrasound imaging have been proposed in the literature \cite{Synnevag2011, Asl2012, Jensen2012, Kim}.
+In this paper we extend the work done in \cite{asen2012} and \cite{Buskenes2013} by utilizing a beamspace transformation.
+The beamspace transform has been proposed used in ultrasound imaging by Nilsen and Hafizovic \cite{Nilsen2009} in order to maintain high frame rates as the number of elements increases while maintaining almost the same lateral resolution and sidelobe reduction capabilities of full Capon beamforming.
 %The beamspace transformation matches well with the focused transmit beams used in cardiac ultrasound imaging. 
@@
 In the next section we present background information on Capon beamforming in both element-space and beamspace and the GPU computing model.
-In Section \ref{sec:meth} we recap our parallel implementation of the Capon beamformer, before the GPU implementation of the beamspace Capon beamformer is presented in Section \ref{sec:bs}.
-Benchmark results and resulting images are presented in Section \ref{sec:bench} and \ref{sec:images}, together with a discussion of trading image quality for speed in Section \ref{sec:trade}.
+In Section \ref{sec:meth} we give a summary of our parallel implementation of the element-space Capon beamformer, before the GPU implementation of the beamspace Capon beamformer is presented in Section \ref{sec:bs}.
+Benchmark results and resulting images for both implementations are presented in Section \ref{sec:bench} and \ref{sec:images}, together with a discussion of trading image quality for speed in Section \ref{sec:trade}.
 Finally we discuss the results in Section \ref{sec:dis}, and draw conclusions in Section \ref{sec:con}.
@@
 where $(\cdot)^H$ is the Hermitian transpose.
-The weight vector, or window, $\vec{w}$ is usually real in DAS beamforming, and is applied to trade resolution for a better signal-to-noise ratio (SNR).
+The weight vector, or window, $\vec{w}$ is usually real in DAS beamforming, and is applied to trade resolution for a lower side lobe level.
 Although different windows can be used in different areas of the images, $\vec{w}$ is usually predefined.
@@
 The result is that interference impinging from other directions is suppressed \cite{Synnevag2007}.
-The complex weights cause a micro-steering which is minute compared to the initial steering and focusing step, therefore the beamformer also works well for broadband ultrasound signals.
+As with coarse-fine beamformers \cite{Thomenius}, the complex weights cause a micro-steering which is minute compared to the initial steering and focusing step, therefore the beamformer also works well for broadband ultrasound signals.
 
@@
 
-For a focused system like cardiac ultrasound imaging, where the received signal is concentrated in a narrow sector around the transmit direction, a small percentage of the beams contains almost all the received energy.
+For a imaging system using narrow transmit and receive beams, like cardiac ultrasound imaging, the received signal will be concentrated in a narrow band in wavenumber space.
+Hence, a small percentage of the beams will contains almost all the received energy.
 Nilsen and Hafizovic concluded that as little as three beams around (and including) the transmit direction were adequate to produce results comparable to applying Capon beamforming in element space.
@@
 \section{Parallel ES-Capon}\label{sec:meth}
-In this section we will recap our parallel implementation of the ES-Capon beamformer as presented in \cite{asen2012, Buskenes}.
+In this section we will give a brief summary of our parallel implementation of the ES-Capon beamformer as presented in \cite{asen2012} and \cite{Buskenes}.
 This implementation was our starting point when the BS-Capon beamformer was implemented.
@@
 
-In \cite{asen2012, Buskenes} we demonstrated how the cubic complexity with $L$ and linear complexity with $N_K$ can be reduced by realizing that calculations in (\ref{eq:R}) overlap across subarrays.
-Element $(i,j)$ of $\mat{\breve{R}}[n]$ can be calculated from element $(i-1, j-1), \text{for } i,j \in [1, L-1]$ in a sliding window maner as
-\begin{align}
-\mat{\breve{R}}_{i,j}[n] &=  \mat{\breve{R}}_{i-1,j-1}[n]  \nonumber \\
-&+ \sum_{n'=n-K}^{n+K} (x_{i+N_L-1}[n']x_{j+N_L-1}[n']^* \nonumber \\
-&- x_{i-1}[n']x_{j-1}[n']^H). \label{eq:sliding}
-\end{align}
+In \cite{asen2012, Buskenes} we demonstrated how the cubic complexity with $L$ %and linear complexity with $N_K$ 
+can be reduced by realizing that calculations in (\ref{eq:R}) overlap across subarrays.
+%Element $(i,j)$ of $\mat{\breve{R}}[n]$ can be calculated from element $(i-1, j-1), \text{for } i,j \in [1, L-1]$ in a sliding window maner as
+%\begin{align}
+%\mat{\breve{R}}_{i,j}[n] &=  \mat{\breve{R}}_{i-1,j-1}[n]  \nonumber \\
+%&+ \sum_{n'=n-K}^{n+K} (x_{i+N_L-1}[n']x_{j+N_L-1}[n']^* \nonumber \\
+%&- x_{i-1}[n']x_{j-1}[n']^H). \label{eq:sliding}
+%\end{align}
 %Along the time dimension we also find that
@@
 %- \mat{\breve{R}}[n-K-1] + \mat{\breve{R}}[n-K].
-Sliding windows across subarrays and in time can be realized simultaneously if $\mat{\hat{R}}$, averaged over subarrays, is calculated from an $N_K$-average in time of the outer products $\vec{x}[n]\vec{x}[n]^H$.
+%Sliding windows across subarrays and in time can be realized simultaneously if $\mat{\hat{R}}$, averaged over subarrays, is calculated from an $N_K$-average in time of the outer products $\vec{x}[n]\vec{x}[n]^H$.
 However, since $N_K$ typically is small, and the GPU has a limited amount of shared memory available per thread, we focus on subarrays only.
-In that case, the complexity of calculating $\mat{\hat{R}}$ is reduced to $O(N_K(L(N_L-2)+2L^2))$.
-For $L=M/2$ this reduces to $O(N_KM^2)$.
-
-The sliding-window approach will cause dependence among the iterative calculations.
+%In that case, 
+The complexity of calculating $\mat{\hat{R}}$ is by this reduced to $O(N_K(L(N_L-2)+2L^2))$. For $L=M/2$ this reduces to $O(N_KM^2)$.
+%The sliding-window approach will cause dependence among the iterative calculations.
 The level of task granularity is therefore increased.
@@
 \ref{fig:gpulayout}).
-Maximum supported $L$ is now increased to the maximum number of threads per block (currently 1024).
-However, for small-$L$-high-$M$ configurations, memory consumption per thread will be too high to achieve full GPU occupancy.
-
-Taking symmetry into account, only the upper half of $\mat{\hat{R}}$ needs to be calculated.
+Maximum supported $L$ is also increased to the maximum number of threads per block (currently 1024).
+%However, for small-$L$-high-$M$ configurations, memory consumption per thread will be too high to achieve full GPU occupancy.
+
+%Taking symmetry into account, only the upper half of $\mat{\hat{R}}$ needs to be calculated.
 However, the GPU computes $\mat{\hat{R}}$ in such a way that the lower triangle can be found with little overhead \cite{Buskenes}.
@@
 
-For the adaptive diagonal loading factor in (\ref{eq:diag}), the trace of $\mat{\hat{R}}$ has to be known before diagonal elements are written back to global memory.
-Since one row for all covariance matrices is written to global memory per kernel iteration, the diagonal elements need to be held in shared memory until the trace is accumulated.
+%For the adaptive diagonal loading factor in (\ref{eq:diag}), the trace of $\mat{\hat{R}}$ has to be known before diagonal elements are written back to global memory.
+One row for all covariance matrices is written to global memory per kernel iteration, the diagonal elements therefore need to reside in shared memory until the trace is accumulated.
 When that is done, scattered writes of the scaled diagonal elements to global memory has to be performed.
@@
 \subsection{Solving Multiple Small Linear Systems}
-The literature has mostly focused on large systems when solving systems of linear equations on the GPU.
+%The literature has mostly focused on large systems when solving systems of linear equations on the GPU.
 GPU libraries therefore often lack a comprehensive collection of batched solvers for large sets of small systems.
@@
 Because of this, it has been proven hard to optimize a manageable set of solvers that provides speedups compared to the CPU in every case.
-It is therefore not given that solving on the GPU is faster than solving on the CPU in general.
+
+In \cite{asen2012} we discussed how it is not given that solving on the GPU is faster than solving on the CPU in general.
 However, there are two arguments for solving on the GPU.
@@
 \subsection{Compute Beamformer Output}
-The calculation of the beamformer output in (\ref{eq:z_mv}) has been implemented using one thread per element weight ($L$ threads per weight vector).
+%The calculation of the beamformer output in (\ref{eq:z_mv}) has been implemented using one thread per element weight ($L$ threads per weight vector).
 After the solution vector $\vec{b}$ has been placed in shared memory, one of the $L$ threads finds the sum of $\vec{b}$.
 %Another option is to let several threads cooperate to form the sum using a binary reduction sum. However, in the following benchmarks the first approach has been used. 
-Each of the $L$ threads then reads one element of $\vec{b}$, and scales it with the inverse sum to get the weight vector $\vec{w}[n]$.
-
+%Each of the $L$ threads then reads one element of $\vec{b}$, and scales it with the inverse sum to get the weight vector $\vec{w}[n]$. 
 The final output can be formed by either reducing the data down to length $L$ as in (\ref{eq:z_mv}), or increasing the weight vector to length $M$ as in (\ref{eq:z_mv2}).
-Since the design is $L$ threads per output, the data is reduced to length $L$ by letting thread $r$ calculate
+Since we have selected to use $L$ threads per output, the data is reduced to length $L$ by letting thread $r$ calculate
 \begin{align}\label{eq:subsum}
@@
 \end{align}
-where $\vec{e}_i$ has the value one in the $i$th position and zeroes elsewhere.
+where $\vec{e}_i$ has the value one in the $i$th position and zeroes elsewhere. Note that the factor $\sqrt{L}$ was not included in \cite{Nilsen2009}.
 
@@
 A kernel similar to the one described in \cite{Chen2011} has therefore been implemented.
-In addition, for small $N_b$ values, several covariance matrices are calculated per thread block to achieve high GPU occupancy.
 The computational complexity of this step is now $O(N_KN_LN_b^2)$, which for $L=M/2$ is less than the complexity of building $\mat{\hat{R}}$ in element space as long as $N_b < \sqrt{M}$.
@@
 %Derivation of sliding beamspace transformation. 
-We will now show how to minimize the amount of operations required for the beamspace transformation by utilizing the overlap between subarrays previously exploited when estimating the covariance matrix in element space.
+%We will now show how to minimize the amount of operations required for the beamspace transformation by utilizing the overlap between subarrays previously exploited when estimating the covariance matrix in element space.
+As the overlap between subarrays were exploited when estimating the covariance matrix in element space, the overlap can now be utilized when calculating the beamspace transform. 
 Given a $M$ long data vector divided into $N_L$ subarrays, the beamspace transformation of the $l$th subarray for the $k$th beam is from (\ref{eq:butler}) found to be:
@@
 \end{align}
+
 We denote this the sliding-window beamspace transformation (SBS), and it is equal to a normalized sliding DFT \cite{Lyons2003}.
@@
 
-Equation (\ref{eq:sliding_bs}) shows that instructions are saved if we utilize the overlap between subarrays, but with an increase in task level granularity (as discussed in Section \ref{sec:calcR}).
+Equation (\ref{eq:sliding_bs}) shows that instructions are saved if we utilize the overlap between subarrays, but with an increase in task level granularity.
 In this situation, it might be tempting to use one thread per beam (index $k$) that calculates this beam for all subarrays.
@@
 \end{align}
-Several pixels are calculated per thread block for small values of $N_b$ in order to achieve high GPU occupancy. The beamspace data summed across subarrays are calculated as part of the SBS kernel.
+%Several pixels are calculated per thread block for small values of $N_b$ in order to achieve high GPU occupancy. The beamspace data summed across subarrays are calculated as part of the SBS kernel.
 
@@
 
-\begin{figure}
-%\centerline{\subfloat[]{\includegraphics[width=3.5in]{gfx/benchmark_bar_32_40_400.png} \label{fig:benchUS32}}
-%\hfil
-%\subfloat[]{\includegraphics[width=3.5in]{gfx/benchmark_bar_64_80_606.png}%
-%\label{fig:benchUS64}}}
-\centerline{\includegraphics[width=3.5in]{gfx/benchmark_bar_64_80_540.png}}
-\caption{Benchmark of GPU Capon beamforming in element space for a cardiac image covering a $70\degree$ wide $15$ cm deep sector.
+\begin{figure*}[!t]
+\centerline{
+\subfloat[]{\includegraphics[width=3.4in]{gfx/benchmark_bar_64_80_540.png} \label{fig:benchUS64}}
+\hfil
+\subfloat[]{\includegraphics[width=3.4in]{gfx/benchmark_bar_96_120_540.png}%
+\label{fig:benchUS96}}}
+%\centerline{\includegraphics[width=3.5in]{gfx/benchmark_bar_64_80_540.png}}
+\caption{Benchmarks of GPU Capon beamforming in element space for a cardiac image covering a $70\degree$ wide $15.4$ cm deep sector.
 Execution time is plotted as a function of subarray length $L$.
-The plot shows execution times for the three steps listed in Section \ref{sec:meth}: Calculating covariance matrices with three different choices of temporal averaging, solver, and the final beamformer sum.}
+The figures show execution times for the three steps listed in Section \ref{sec:meth}: Calculating covariance matrices with three different choices of temporal averaging, solver, and the final beamformer sum.}
 \label{fig:bench}
-\end{figure}
+\end{figure*}
 
 \begin{figure*}[!t]
-\centerline{\subfloat[]{\includegraphics[width=3.45in]{gfx/benchmark_bar_bs_M=64_Nx=80_Ny=540_Nb=3.png} \label{fig:benchBS64}}
+\centerline{\subfloat[]{\includegraphics[width=3.4in]{gfx/benchmark_bar_bs_M=64_Nx=80_Ny=540_Nb=3.png} \label{fig:benchBS64}}
 \hfil
-\subfloat[]{\includegraphics[width=3.45in]{gfx/benchmark_bar_bs_M=96_Nx=120_Ny=540_Nb=3.png}%
+\subfloat[]{\includegraphics[width=3.4in]{gfx/benchmark_bar_bs_M=96_Nx=120_Ny=540_Nb=3.png}%
 \label{fig:benchBS96}}}
-\caption{Benchmarks of GPU beamspace Capon beamforming for a cardiac image covering a $70 \degree$ and $15$ cm deep sector.
+\caption{Benchmarks of GPU beamspace Capon beamforming for a cardiac image covering a $70 \degree$ and $15.4$ cm deep sector.
 Execution time is plotted as a function of subarray length $L$.
-If $L$ had been kept constant on $M/2$ and execution time was plotted against $N_b$, the plot would be similar to the one in Fig.
+If $L$ had been fixed at $M/2$ and execution time was plotted against $N_b$, the solver would have shown the same quadratic growth as in Fig.
 \ref{fig:bench}.
-The variation in execution time is caused by adaptive kernel launch configurations.}
+When $N_b$ is fixed, calculating the covariance matrix and the BS-transform is inverse proportional and proportional to $L$ respectively.
+The large jumps in execution times are caused by adaptive kernel launch configurations, which depends on $M$, $L$ and $N_b$.}
 \label{fig:benchBS}
@@
 Assuming a system bandwidth of 80\% (relative to the center frequency) we need a complex sampling frequency of 2.7 MHz, and when imaging down to a depth of 15.4 cm we have approximately 540 samples per range line.
-To sample a $70\degree$ sector with 64- and 96-element arrays ($d=\lambda/2$ pitch) at twice the Rayleigh criterion ($\lambda/(2Md)$), we need approximately 80 and 120 receive beams respectively.
+The angular sampling rate is given by twice the Rayleigh criterion \cite{Hergum2007}
+\begin{equation}\label{eq:rayleigh}
+\delta{}\theta = \lambda/(2Mp).
+\end{equation}
+To sample a $70\degree$ sector with $64$- and $96$-element arrays with pitch $p=\lambda/2$ we therefore need approximately 80 and 120 receive beams respectively.
+
 The presented execution times do not include transfer of data from CPU- to the GPU-side, since data are assumed to be pre-delayed and available in global GPU memory.
+This is a valid assumption, since all imaging scenarios presented in this paper are well below the 8.0 GB/s limit of PCI Express 2.0.
 What is included are all calculations, memory transfers from global to shared GPU memory, and shared memory access.
 
-A benchmark of the ES-Capon is presented in Fig.
+Benchmarks of the ES-Capon beamformer are presented in Fig.
 \ref{fig:bench}.
-The benchmark is based on a 64-element phased array, using 80 receive beams.
+The benchmarks are based on a 64-element and 96-elements phased array respectively using 80 receive beams.
 For each subarray size, the total execution time is split into one bar for each of the three steps described in Section \ref{sec:meth}, and the calculation of $\mat{\hat{R}}$ is presented for three different values of $K$.
-For $M=64$, $L=32$ and $K=2$, a typical setup for high-quality cardiac imaging, processing throughput is 7 frames per second.
+For $M=64$, $L=32$ and $K=2$, a typical setup for high-quality cardiac imaging, processing throughput is 8.3 frames per second (FPS).
+For $M=96$, $L=48$ and $K=2$ the processing throughput drops below $0.5$ FPS.
 
@@
 The figure only presents results with use of temporal averaging ($K=2$).
-For $M=64$, $L=32$, $K=2$ and $N_b=3$, processing throughput is 125 frames per second (8 ms).
-For $M=96$, $L=48$, $K=2$ and $N_b=3$, processing throughput is 77 frames per second (13 ms).
+For $M=64$, $L=32$, $K=2$ and $N_b=3$, processing throughput is 125 FPS (8 ms).
+For $M=96$, $L=48$, $K=2$ and $N_b=3$, processing throughput is 77 FPS (13 ms).
 In all four benchmarks there is a line indicating the time it takes to acquire the underlying image using one receive line per transmit.
@@
 \ref{fig:phantom}).
-The simulation was performed using a 64-element, $\lambda/2$ pitch, 10 $\mu$m kerf, 80\% bandwidth, and 10 mm high phased array.
+The simulation was performed using a 64-element, $p=\lambda/2$ pitch, 10 $\mu$m kerf, 80\% bandwidth, and 10 mm high phased array.
 The transducer was excited with a 3.4 MHz and $1.5\lambda$ long pulse with transmit focus at 90 mm.
@@
 The resulting image has been processed both using DAS with uniform weights, ES- and BS-Capon.
-The presented image is the first frame in a series of images where motion is imposed on both circles, point targets and the probe.
+The presented image is frame $15$ in a series of images where motion is imposed on both circles, point targets and the probe.
 The entire simulation is presented in two attached videos \multimedia{Media-Movie 1 and 2}.
@@
 \ref{fig:invivo} presents results of applying ES- and BS-Capon on an \textit{in-vivo} image of the left ventricle.
-The image was acquired using a 64-element and $0.65\lambda$ pitched phased array in harmonic mode (1.7 MHz at transmit and 3.4 MHz at receive).
+The image was acquired using a 64-element and $p=0.65\lambda$ pitched phased array in harmonic mode (1.7 MHz at transmit and 3.4 MHz at receive).
 The image is extracted from a cardiac cycle, that can be viewed in its full length in the attached videos \multimedia{Media-Movie 3, 4 and 5}.
@@
 For BS-Capon the following parameters have been used: $L=32$, $K=2$, $d=0.01$, and $N_b=3$.
-For the simulated phantom, 16x lateral oversampling on transmit has been performed in order to avoid under sampling of the two bright point scatterers as they move.
-The cardiac recording of a 70\degree{} sector has a minimal lateral sampling of 98 receive beams ($\lambda/2Md$).
+
+For the simulated phantom, modulation effects were observed when the two bright point scatterers moved across receive-beams.
+The artefact did not disappear until 16x lateral oversampling relative to (\ref{eq:rayleigh}) was used on transmit.
+The cardiac recording of a 70\degree{} sector has a minimal lateral sampling of 98 receive beams given by (\ref{eq:rayleigh}).
+Despite this, the same artefacts were not observed here.
+The image was acquired using 50 transmit beams and "Synthetic Transmit Beamforming" was used on receive \cite{Hergum2007}.
 
@@
 \ref{fig:phantom}.
-The precision of the presented measurements is 0.12 mm, dictated by the simulated frame rate, 25 frames per second (FPS), and the relative speed of the two point scatterers, 3 mm/s.
+The precision of the presented measurements is 0.12 mm, dictated by the simulated frame rate, 25 FPS, and the relative speed of the two point scatterers, 3 mm/s.
 
@@
 For $L = N_b=3$, BS-Capon provides almost 1 mm improved resolution over DAS and ES-Capon.
-However, note that the resolution of the Capon beamformers strongly depends on image SNR and the choice of parameters.
+However, note that the resolution of the Capon beamformers strongly depends on the image signal-to-noise ratio (SNR) and the choice of parameters.
 Thus, different SNR scenarios and choices of parameters will lead to different results.
@@
 \centering
-\includegraphics[width=3in]{gfx/speed_res_trade_d=0-01.png}
+\includegraphics[width=3.3in]{gfx/speed_res_trade_d=0-01.png}
 \caption{Trade-off between resolution and speed by reducing subarray size or reducing beamspace dimension (based on Fig.
@@
 \section{Discussion}\label{sec:dis}
-In this paper we have presented a parallel implementation of the Capon beamformer in both element- and beamspace for high-resolution cardiac ultrasound imaging.
-The results show that it is now possible to apply BS-Capon in real-time in cardiac ultrasound imaging by utilizing the power of GPUs.
+In this paper we have presented a parallel implementation of the Capon beamformer in beamspace for high-resolution cardiac ultrasound imaging.
+The results show that it is now possible to apply BS-Capon in real-time in cardiac ultrasound imaging by utilizing the power of a GPU.
 According to Fig.
@@
 $L=M/2$).
-These extra lines might be needed for increased sampling instead of increased frame rate when using the Capon beamformers in high SNR situations (Fig \ref{fig:phantom}).
+These extra lines might be needed for increased sampling instead of increased frame rate since we had to use oversampling in our simulation.
 If parallel receive lines are used to obtain denser sampling, acquisition time remains the same but processing times are increased with a factor equal to the amount of oversampling.
 If parallel receive lines are used for increased frame rate, the acquisition time is lowered, and the processing time remains constant.
-Hence, four parallel receive lines are not possible to process in real-time using BS-Capon (with $N_b=3$) in combination with our single GPU setup.
+Hence, four parallel receive lines are not possible to process in real-time using BS-Capon with $N_b=3$ in combination with our single GPU setup.
 
@@
 From Fig.
-\ref{fig:bench} we see that for $L = M/2 = 32$ and $K=1$, the total execution time is 140 ms (7 FPS).
+\ref{fig:benchUS64} we see that for $L = M/2 = 32$ and $K=1$, the total execution time is 120 ms (8.3 FPS).
 Taking into account that we use complex data, but less samples per image, these results are similar to the performance reported by Chen et al.
@@
 If real-time is defined as 10 FPS \cite{Chen2011} almost all execution times in both Fig.
-\ref{fig:bench} and \ref{fig:benchBS} are real-time.
+\ref{fig:benchUS64} and \ref{fig:benchBS} are real-time.
 More natural for cardiac ultrasound imaging is to define real-time as the time it takes to acquire the underlying image.
@@
 
-This paper further shows how to obtain a parallel formulation of the ES-Capon and BS-Capon beamformer, and how to implement them in a GPU framework.
+This paper further shows how to obtain a parallel formulation of the BS-Capon beamformer, and how to implement it in a GPU framework.
 The per-pixel granularity suggests that the beamformer is easily dividable across multiple GPUs.
@@
 \ref{fig:phantom}, and the attached video of the simulation, that Capon beamforming provides improved lateral resolution of moving point targets both in element- and in a reduced beam-space ($N_b=3$).
-Our preliminary finding is that the Capon beamformers are local shift invariant as long as the lateral sampling is increased to a curtain level \multimedia{Media-Movie 1 and 2}.
+Our preliminary finding is that the Capon beamformers are local shift invariant as long as the lateral sampling is increased to a certain level \multimedia{Media-Movie 1 and 2}.
 How large this oversampling factor needs to be and how the oversampling should be conducted is outside the scope of this paper.
@@
 Future work should focus on solving this issue.
+Cardiac ultrasound has not been selected as the target modality in this paper because Capon beamforming shows large improvements in image quality in particular, it has been selected because it is the most computationally demanding modality (measured in float operations per second).
+The rationale is that if BS-Capon can run in real time for cardiac ultrasound it can run in real time for all medical ultrasound modalities.
+We believe that Capon will show more improvements in other imaging modalities.
 
@@
 \section{Conclusion}\label{sec:con}
-In this paper a parallel formulation and implementation of the element- and beamspace Capon beamformer have been presented, and
-both beamformers have successfully been implemented on the GPU.
+In this paper a parallel formulation and implementation of the beamspace Capon beamformer have been presented, and
+the beamformer have successfully been implemented on the GPU.
 For a three dimensional beamspace, processing rates supporting real-time cardiac ultrasound imaging was shown to be possible, meaning that processing rates higher than the image acquisition rate are obtained for a wide range of parameters.
 Image quality was investigated both in a simulated phantom and an \textit{in-vivo} cardiac dataset.
-These results showed that Capon beamforming is feasible for cardiac ultrasound imaging, providing images with improved lateral resolution and no local shift-variance effects, both in element- and beamspace.
+These results showed that Capon beamforming is feasible for cardiac ultrasound imaging, providing images with improved lateral resolution both in element- and beamspace.
 

### 2013-09-06 HRUIonGPU/jrnl_article.tex
final submit

@@
 For a three dimensional beamspace we show that processing rates supporting real-time cardiac ultrasound imaging are possible, meaning that images can be processed faster than the image acquisition rate for a wide range of parameters.
-Image quality is investigated in videos of both a simulated phantom, and in an \textit{in-vivo} cardiac dataset.
+Image quality is investigated in an \textit{in-vivo} cardiac dataset.
 These results show that Capon beamforming is feasible for cardiac ultrasound imaging, providing images with improved lateral resolution both in element-space and beamspace.
@@
 In this paper we extend the work done in \cite{asen2012} and \cite{Buskenes2013} by utilizing a beamspace transformation.
-The beamspace transform has been proposed used in ultrasound imaging by Nilsen and Hafizovic \cite{Nilsen2009} in order to maintain high frame rates as the number of elements increases while maintaining almost the same lateral resolution and sidelobe reduction capabilities of full Capon beamforming.
+The beamspace transform has been proposed used in ultrasound imaging by Nilsen and Hafizovic \cite{Nilsen2009} in order to maintain high frame rates as the number of elements increases while maintaining almost the same lateral resolution and sidelobe reduction capabilities of element-space Capon when applied in narrow-beam systems.
 %The beamspace transformation matches well with the focused transmit beams used in cardiac ultrasound imaging. 
@@
 
-In the next section we present background information on Capon beamforming in both element-space and beamspace and the GPU computing model.
+In the next section we present background information on Capon beamforming in both element-space and beamspace.
+We also give an introduction to the GPU computing model.
 In Section \ref{sec:meth} we give a summary of our parallel implementation of the element-space Capon beamformer, before the GPU implementation of the beamspace Capon beamformer is presented in Section \ref{sec:bs}.
@@
 
-As shown in Fig.
-\ref{fig:mvbf}, the basic idea of an adaptive beamformer is to form a complex weight vector based on the received data.
+As shown in Fig.\ \ref{fig:mvbf}, the basic idea of an adaptive beamformer is to form a complex weight vector based on the received data.
 The Capon beamformer produces weights that minimize the beamformer output power while maintaining unity gain in the focus direction.
@@
 \caption{Capon beamforming.
-Impinging signal is demodulated, and aligned in phase by applying steering and focusing delays $\Delta_m[n]$.
+The impinging signal is demodulated, and aligned in phase by applying steering and focusing delays $\Delta_m[n]$.
 Adaptive weights, $\vec{w}[n]$, are then calculated based on in-phase data, and finally the output is formed by a Capon-weighted coherent sum.}
@@
 
-For a imaging system using narrow transmit and receive beams, like cardiac ultrasound imaging, the received signal will be concentrated in a narrow band in wavenumber space.
-Hence, a small percentage of the beams will contains almost all the received energy.
+For an imaging system using narrow transmit and receive beams, like cardiac ultrasound imaging, the received signal will be concentrated in a narrow band in wavenumber space.
+Hence, a small percentage of the beams will contain almost all the received energy.
 Nilsen and Hafizovic concluded that as little as three beams around (and including) the transmit direction were adequate to produce results comparable to applying Capon beamforming in element space.
@@
 In CUDA programming a kernel function is executed over a large set of threads, where the kernel function describes the work to be done in each thread.
-As depicted in Fig.
-\ref{fig:gpulayout}, threads are organized in a grid of thread-blocks, with maximal size of 1024 threads per block (e.g.
+As depicted in Fig.\ \ref{fig:gpulayout}, threads are organized in a grid of thread-blocks, with maximal size of 1024 threads per block (e.g.
 $32\times32$).
 Threads inside a block support low-cost synchronization if needed.
-From a CUDA perspective the GPU consist of one or more streaming multi-processors (SM), where each SM is capable of 32 or more concurrent multiplications or additions, or both, depending on the architecture.
-A group of blocks is scheduled to each SM until all blocks have been processed.
+From a CUDA perspective the GPU consist of one or more streaming multi-processors (SM), where each SM is capable of executing multiple arithmetic operations per clock cycle.
+%32 or more concurrent multiplications or additions, or both, depending on the architecture. A group of blocks is scheduled to each SM until all blocks have been processed.
 
@@
 If the beamformer in (\ref{eq:z_mv}) is analyzed in combination with the matrix equation in (\ref{eq:b}) and the covariance estimation process in (\ref{eq:diag}), we see that the calculation of a set of Capon weights is independent across the outputs $z[n]$.
-As shown in Fig.
-\ref{fig:gpulayout}, a first level of granularity is therefore to divide the image into a grid, where the output in each cell can be calculated independently.
+As shown in Fig.\ \ref{fig:gpulayout}, a first level of granularity is therefore to divide the image into a grid, where the output in each cell can be calculated independently.
 Further, based on memory dependencies and to hold memory consumption per thread at a minimum, a group of threads should process one output in the final image.
@@
 The final output can be formed by either reducing the data down to length $L$ as in (\ref{eq:z_mv}), or increasing the weight vector to length $M$ as in (\ref{eq:z_mv2}).
-Since we have selected to use $L$ threads per output, the data is reduced to length $L$ by letting thread $r$ calculate
+Since we selected to use $L$ threads per output, the data is reduced to length $L$ by letting thread $r$ calculate
 \begin{align}\label{eq:subsum}
@@
 %We will now show how to minimize the amount of operations required for the beamspace transformation by utilizing the overlap between subarrays previously exploited when estimating the covariance matrix in element space.
-As the overlap between subarrays were exploited when estimating the covariance matrix in element space, the overlap can now be utilized when calculating the beamspace transform. 
+The overlap between subarrays that were exploited when estimating the covariance matrix in element space, can now be utilized when calculating the beamspace transform. 
 Given a $M$ long data vector divided into $N_L$ subarrays, the beamspace transformation of the $l$th subarray for the $k$th beam is from (\ref{eq:butler}) found to be:
@@
 \end{align}
-where $\mat{B}_k$ is the $k$th row of $\mat{B}$. Manipulating this expression we find that for $l \in [1, L-1]$
-\begin{align}\label{eq:sliding_bs}
-x_{\text{BS},l,k} &= \frac{1}{\sqrt{L}}((x_{\text{BS},l-1,k} - x_{l-1})e^{-j2\pi k/L} \nonumber \\ &+ x_{i+L-1}e^{-j2\pi k(L-1)/L}) \nonumber \\
-&= \frac{1}{\sqrt{L}}(x_{\text{BS},l-1,k} - x_{l-1} + x_{i+L-1})e^{-j2\pi k/L}.
-\end{align}
-
-We denote this the sliding-window beamspace transformation (SBS), and it is equal to a normalized sliding DFT \cite{Lyons2003}.
+where $\mat{B}_k$ is the $k$th row of $\mat{B}$. %Manipulating this expression we find that for $l \in [1, L-1]$
+%\begin{align}\label{eq:sliding_bs}
+%x_{\text{BS},l,k} &= \frac{1}{\sqrt{L}}((x_{\text{BS},l-1,k} - x_{l-1})e^{-j2\pi k/L} \nonumber \\ &+ x_{i+L-1}e^{-j2\pi k(L-1)/L}) \nonumber \\
+%&= \frac{1}{\sqrt{L}}(x_{\text{BS},l-1,k} - x_{l-1} + x_{i+L-1})e^{-j2\pi k/L}.
+%\end{align}
+
+%We denote this the sliding-window beamspace transformation (SBS), and it is equal to a normalized sliding DFT \cite{Lyons2003}.
 For $l=0$ a full computation has to be performed using (\ref{eq:bs_formula}).
-
-Equation (\ref{eq:sliding_bs}) shows that instructions are saved if we utilize the overlap between subarrays, but with an increase in task level granularity.
+This formula has been implemented using a normalized sliding DFT (SDFT)\cite{Lyons2003}.
+%Equation (\ref{eq:sliding_bs}) shows that 
+The SDFT allows us to still save instructions by utilizing the overlap between subarrays, but with an increase in task level granularity.
 In this situation, it might be tempting to use one thread per beam (index $k$) that calculates this beam for all subarrays.
@@
 Using one thread per beam is therefore not feasible for small-$N_b$-large-$M$ configurations.
-One solution, as depicted in Fig.
-\ref{fig:gpulayoutbs}, is to combine SBS with full computations.
-A number of sub thread groups of size $N_b$ is created that targets different sets of subarrays which they transform using the SBS algorithm.
+One solution, as depicted in Fig.\ \ref{fig:gpulayoutbs}, is to combine SDFT with full computations.
+A number of sub thread groups of size $N_b$ is created that targets different sets of subarrays which they transform using the SDFT algorithm.
 In that way, the number of threads per sample is increased and thus the memory-per-thread ratio is decreased.
@@
 Calculation of the covariance matrix is now $O(N_KN_LN_b^2)$, the inversion step is $O(N_b^3)$, and calculation of weights and the beamformer output are both $O(N_b)$.
-The added complexity of SBS is $O(N_bL)$.
+The added complexity of SDFT is $O(N_bL)$.
 To maximize performance, large adjustments to the processing pipeline described in Section \ref{sec:meth} is required.
@@
 Execution time is plotted as a function of subarray length $L$.
-If $L$ had been fixed at $M/2$ and execution time was plotted against $N_b$, the solver would have shown the same quadratic growth as in Fig.
-\ref{fig:bench}.
+If $L$ had been fixed at $M/2$ and execution time was plotted against $N_b$, the solver would have shown the same quadratic growth as in Fig.\ \ref{fig:bench}.
 When $N_b$ is fixed, calculating the covariance matrix and the BS-transform is inverse proportional and proportional to $L$ respectively.
@@
 Assuming a system bandwidth of 80\% (relative to the center frequency) we need a complex sampling frequency of 2.7 MHz, and when imaging down to a depth of 15.4 cm we have approximately 540 samples per range line.
-The angular sampling rate is given by twice the Rayleigh criterion \cite{Hergum2007}
+The angular sampling is given by the Rayleigh criterion divided by two \cite{Hergum2007},
 \begin{equation}\label{eq:rayleigh}
-\delta{}\theta = \lambda/(2Mp).
+\delta{}\theta = \lambda/(2Mp),
 \end{equation}
-To sample a $70\degree$ sector with $64$- and $96$-element arrays with pitch $p=\lambda/2$ we therefore need approximately 80 and 120 receive beams respectively.
+where $\lambda$ is the pulse wavelength and $p$ is the element pitch. 
+To sample a $70\degree$ sector with $64$- and $96$-element arrays with $p=\lambda/2$ we therefore need approximately 80 and 120 receive beams respectively.
 
 The presented execution times do not include transfer of data from CPU- to the GPU-side, since data are assumed to be pre-delayed and available in global GPU memory.
-This is a valid assumption, since all imaging scenarios presented in this paper are well below the 8.0 GB/s limit of PCI Express 2.0.
+This is a valid assumption, since the imaging scenarios presented in this paper are well below the 8.0 GB/s limit of PCI Express 2.0.
 What is included are all calculations, memory transfers from global to shared GPU memory, and shared memory access.
-
-Benchmarks of the ES-Capon beamformer are presented in Fig.
-\ref{fig:bench}.
-The benchmarks are based on a 64-element and 96-elements phased array respectively using 80 receive beams.
+The delay step has previously been shown to take far less time than what we report for both Capon beamformers \cite{Song2012, Chen2011}.
+
+Benchmarks of the ES-Capon beamformer are presented in Fig.\ \ref{fig:bench}.
 For each subarray size, the total execution time is split into one bar for each of the three steps described in Section \ref{sec:meth}, and the calculation of $\mat{\hat{R}}$ is presented for three different values of $K$.
@@
 
-In Fig.
-\ref{fig:benchBS} we present similar benchmarks for the BS-Capon beamformer.
-Benchmarks are based on 64- and 96-element phased arrays, using 80 and 120 receive beams.
+In Fig.\ \ref{fig:benchBS} we present similar benchmarks for the BS-Capon beamformer.
 The figure only presents results with use of temporal averaging ($K=2$).
@@
 
-\section{Simulated phantom and \textit{in-vivo} cardiac images}\label{sec:images}
-
-\begin{figure*}[!t]
-\centerline{
-\subfloat[]{ \includegraphics[width=2.2in]{gfx/das.png} \label{fig:phantomDAS} }
-\hfil
-\subfloat[]{ \includegraphics[width=2.2in]{gfx/capon.png} \label{fig:phantomMV} }
-\hfil
-\subfloat[]{ 
-\includegraphics[width=2.2in]{gfx/capon_bs.png} \label{fig:phantomBS} 
-\hfil
-\includegraphics[width=0.35in]{gfx/colorBarPhantom2.png}
-}
-}
-\caption{Simulated phantom containing one bright (20 dB) and one dark (-20 dB) circular area and two point targets (30 dB) in speckle (0 dB).
+\section{In-vivo cardiac images}\label{sec:images}
+
+%\begin{figure*}[!t]
+%\centerline{
+%\subfloat[]{ \includegraphics[width=2.2in]{gfx/das.png} \label{fig:phantomDAS} }
+%\hfil
+%\subfloat[]{ \includegraphics[width=2.2in]{gfx/capon.png} \label{fig:phantomMV} }
+%\hfil
+%\subfloat[]{ 
+%\includegraphics[width=2.2in]{gfx/capon_bs.png} \label{fig:phantomBS} 
+%\hfil
+%\includegraphics[width=0.35in]{gfx/colorBarPhantom2.png}
+%}
+%}
+%\caption{Simulated phantom containing one bright (20 dB) and one dark (-20 dB) circular area and two point targets (30 dB) in speckle (0 dB).
 The simulation was done using Field II with a 64-element, 3.4 MHz, and $\lambda/2$ pitch phased array.
@@
 c) Beamspace Capon ($L=32, K=2, N_b=3, d=0.01$) \multimedia{Media-Movie 2}.}
-\label{fig:phantom}
-\end{figure*}
+%\label{fig:phantom}
+%\end{figure*}
 
@@
 The image is the first frame in the attached videos.
-a) Delay-and-sum with uniform weights \multimedia{Media-Movie 3}.
-b) Element-space Capon ($L=32, K=2, d=0.1$) \multimedia{Media-Movie 4}.
-c) Beamspace Capon ($L=32, K=2, N_b=3, d=0.01$) \multimedia{Media-Movie 5}.}
+a) Delay-and-sum with uniform weights \multimedia{Media-Movie 1}.
+b) Element-space Capon ($L=32, K=2, d=0.1$) \multimedia{Media-Movie 2}.
+c) Beamspace Capon ($L=32, K=2, N_b=3, d=0.01$) \multimedia{Media-Movie 3}.}
 \label{fig:invivo}
@@
 
-\begin{figure*}[!t]
-\centerline{
-\subfloat[]{\includegraphics[width=3in]{gfx/simulation_slice.png} \label{fig:simulationSlice}}
-\hfil
-\subfloat[]{\includegraphics[width=3in]{gfx/invivo_slice.png} \label{fig:invivoSlice}}
-}
-\caption{Lateral slices taken from the images in Fig.
+%\begin{figure*}[!t]
+%\centerline{
+%\subfloat[]{\includegraphics[width=3in]{gfx/simulation_slice.png} \label{fig:simulationSlice}}
+%\hfil
+%\subfloat[]{\includegraphics[width=3in]{gfx/invivo_slice.png} \label{fig:invivoSlice}}
+%}
+%\caption{Lateral slices taken from the images in Fig.
 \ref{fig:phantom} and Fig.
@@
 \ref{fig:invivo} at 96 mm range, through the tip of the mitral valve.}
-\label{fig:slices}
-\end{figure*}
-
-To investigate the image quality of both the ES and BS-Capon beamformer, a phantom was simulated using Field II \cite{Jensen1992, Jensen1996a} (Fig.
+%\label{fig:slices}
+%\end{figure*}
+
+%To investigate the image quality of both the ES and BS-Capon beamformer, a phantom was simulated using Field II \cite{Jensen1992, Jensen1996a} (Fig.
 \ref{fig:phantom}).
@@
 
-Fig.
-\ref{fig:invivo} presents results of applying ES- and BS-Capon on an \textit{in-vivo} image of the left ventricle.
+Fig.\ \ref{fig:invivo} presents results of applying ES- and BS-Capon on an \textit{in-vivo} image of the left ventricle.
 The image was acquired using a 64-element and $p=0.65\lambda$ pitched phased array in harmonic mode (1.7 MHz at transmit and 3.4 MHz at receive).
-The image is extracted from a cardiac cycle, that can be viewed in its full length in the attached videos \multimedia{Media-Movie 3, 4 and 5}.
-It is included to show that both ES-Capon and BS-Capon beamforming are feasible for \textit{in-vivo} cardiac imaging, not just in simulations.
-
-To get a better impression of the improved lateral resolution obtained with Capon beamforming, two slices taken from the images in Fig.
+The image frame is extracted from a cardiac cycle, that can be viewed in its full length in the attached videos \multimedia{Media-Movie 1, 2 and 3}.
+The region of interest is 70\degree{} with a minimal lateral sampling of 98 receive beams given by (\ref{eq:rayleigh}).
+The image was acquired using 50 transmit beams and "Synthetic Transmit Beamforming" was used on receive \cite{Hergum2007}.
+%It is included to show that both ES-Capon and BS-Capon beamforming are feasible for \textit{in-vivo} cardiac imaging, not just in simulations.
+
+%To get a better impression of the improved lateral resolution obtained with Capon beamforming, two slices taken from the images in Fig.
 \ref{fig:phantom} and \ref{fig:invivo} are included in Fig.
@@
 
-Both data sets have been processed with the following settings for the ES-Capon beamformer: $L = 32$, $K = 2$, and $d=0.1$.
+%Both data sets have 
+The data set has been processed with the following settings for the ES-Capon beamformer: $L = 32$, $K = 2$, and $d=0.1$.
 These settings have previously been shown to provide improved lateral resolution while maintaining DAS-like speckle \cite{Synnevag2007a}.
@@
 
-For the simulated phantom, modulation effects were observed when the two bright point scatterers moved across receive-beams.
-The artefact did not disappear until 16x lateral oversampling relative to (\ref{eq:rayleigh}) was used on transmit.
-The cardiac recording of a 70\degree{} sector has a minimal lateral sampling of 98 receive beams given by (\ref{eq:rayleigh}).
-Despite this, the same artefacts were not observed here.
-The image was acquired using 50 transmit beams and "Synthetic Transmit Beamforming" was used on receive \cite{Hergum2007}.
+%When the simulated phantom was processed with the ES-Capon beamformer, modulation effects were observed when the two bright point scatterers moved across receive-beams.
+The artifact did not disappear until 16x lateral oversampling relative to (\ref{eq:rayleigh}) was used on transmit.
+%The cardiac recording shows a 70\degree{} sector with a minimal lateral sampling of 98 receive beams given by (\ref{eq:rayleigh}). %Despite this, the same artifacts were not observed here. 
+%The image was acquired using 50 transmit beams and "Synthetic Transmit Beamforming" was used on receive \cite{Hergum2007}.
 
 \section{Trading resolution for speed}\label{sec:trade}
-As seen from Fig.
-\ref{fig:bench}, execution time is reduced when $L$ is reduced.
-In Fig.
-\ref{fig:benchBS} we see that execution time is also reduced when moving to beamspace.
+As seen from Fig.\ \ref{fig:bench} the execution time, for the solver in particular, is reduced when $L$ is reduced.
+The same is true when $N_b$ is reduced in beamspace.
 So to gain an  increase in processing throughput we can reduce $L$ in element space or $N_b$ in beamspace.
 In this section we investigate how adjusting $L$ for ES-Capon and $N_b$ for BS-Capon impacts the lateral resolution.
-In Fig.
-\ref{fig:speed_res_trade}, the minimum distance between the two point targets in Fig.
-\ref{fig:phantom} (while preserving a 6 dB saddle point in between the two) is plotted against $L$ (when processed with ES-Capon) and $N_b$ (when processed with BS-Capon).
-To produce the result in Fig.
-\ref{fig:speed_res_trade}, $d=0.01$ has been used on both methods in order to obtain the same resolution for $L=N_b=M/2$.
-We considered this to be a small value for ES-Capon and a normal value for BS-Capon (cf.
+
+In Fig.\ \ref{fig:speed_res_trade}, the minimum distance between two point targets (while preserving a 6 dB saddle point in between the two) is plotted against $L$ (when processed with ES-Capon) and $N_b$ (when processed with BS-Capon).
+To produce the result in Fig.\ \ref{fig:speed_res_trade} $K=2$ and $d=0.01$ has been used on both methods in order to obtain the same resolution for $L=N_b=M/2$.
+In Fig.\ \ref{fig:invivo}, the diagonal loading factor for BS-Capon was adjusted to make the two results equal for $L=32$ and $N_b=3$.
+%The resolution for ES-Capon is therefore better here than what is presented in Fig.
+\ref{fig:phantom}.
+%We considered this to be a small value for ES-Capon and a normal value for BS-Capon (cf.
 Section \ref{sec:bs}), hence the resolution for ES-Capon is here better than what is presented in Fig.
 \ref{fig:phantom}.
-The precision of the presented measurements is 0.12 mm, dictated by the simulated frame rate, 25 FPS, and the relative speed of the two point scatterers, 3 mm/s.
+
+The phantom used contained two point targets with 30 dB amplitude located at 90 mm depth.
+The two point targets were placed inside speckle, with 0 dB in amplitude, covering an area from $\pm90\degree$ in azimuth.
+The phantom was simulated using Field II \cite{Jensen1992, Jensen1996a}.
+The simulation was performed using a 64-element, $p=\lambda/2$ pitch, 10 $\mu$m kerf, 80\% bandwidth, and 10 mm high phased array.
+The transducer was excited with a 3.4 MHz and $1.5\lambda$ long pulse with transmit focus at 90 mm.
+The precision of the presented measurements is 0.12 mm, dictated by the simulated frame rate of 25 FPS, and the relative speed of the two point scatterers, 3 mm/s.
+To avoid having to deal with signal cancellation and errors caused by not hitting each point with a receive beam, 16x oversampling has been used on transmit.
 
@@
 \centering
-\includegraphics[width=3.3in]{gfx/speed_res_trade_d=0-01.png}
-\caption{Trade-off between resolution and speed by reducing subarray size or reducing beamspace dimension (based on Fig.
-\ref{fig:phantom}).
-As $L$ or $N_b$ gets smaller, the execution time is reduced (Fig.
-\ref{fig:bench} and \ref{fig:benchBS}).
+\includegraphics[width=3.3in]{gfx/speed_res_trade.png}
+\caption{Trade-off between resolution and speed by reducing subarray size or reducing beamspace dimension.
+As $L$ or $N_b$ gets smaller, the execution time is reduced (Fig.\ \ref{fig:bench} and \ref{fig:benchBS}).
 When $N_b$ is varied, $L$ is fixed on 32.}
@@
 \section{Discussion}\label{sec:dis}
-In this paper we have presented a parallel implementation of the Capon beamformer in beamspace for high-resolution cardiac ultrasound imaging.
+In this paper we have presented a parallel implementation of the BS-Capon beamformer for high-resolution cardiac ultrasound imaging.
 The results show that it is now possible to apply BS-Capon in real-time in cardiac ultrasound imaging by utilizing the power of a GPU.
-According to Fig.
-\ref{fig:benchBS} it is also possible to process two parallel receive lines per transmit beam (for both $M=64$ and $M=96$.
+According to Fig.\ \ref{fig:benchBS} it is also possible to process two parallel receive lines per transmit beam (for both $M=64$ and $M=96$.
 $L=M/2$).
-These extra lines might be needed for increased sampling instead of increased frame rate since we had to use oversampling in our simulation.
-If parallel receive lines are used to obtain denser sampling, acquisition time remains the same but processing times are increased with a factor equal to the amount of oversampling.
-If parallel receive lines are used for increased frame rate, the acquisition time is lowered, and the processing time remains constant.
+%However, since we had to use oversampling in our simulation, these extra lines might be needed for increased sampling instead of increasing the frame rate.
+If parallel receive lines are used to obtain denser sampling, the acquisition time remains the same but the processing time is increased with a factor equal to the amount of oversampling.
+If parallel receive lines are used to increase the frame rate, the acquisition time is lowered, and the processing time remains constant.
 Hence, four parallel receive lines are not possible to process in real-time using BS-Capon with $N_b=3$ in combination with our single GPU setup.
@@
 For ES-Capon, processing rates for high quality settings are still too high to support real-time cardiac ultrasound imaging.
-From Fig.
-\ref{fig:benchUS64} we see that for $L = M/2 = 32$ and $K=1$, the total execution time is 120 ms (8.3 FPS).
+From Fig.\ \ref{fig:benchUS64} we see that for $L = M/2 = 32$ and $K=1$, the total execution time is 110 ms (9 FPS).
 Taking into account that we use complex data, but less samples per image, these results are similar to the performance reported by Chen et al.
@@
 A major part of the total execution time for ES-Capon comes from solving the system of linear equations.
-This limitation is targeted by reducing the sample covariance matrix size in beamspace, and as shown in Fig.
-\ref{fig:benchBS}, inverting the $3\times3$ beamspace covariance matrix is then barely visible.
+This limitation is targeted by reducing the sample covariance matrix size in beamspace, and as shown in Fig.\ \ref{fig:benchBS}, inverting the $3\times3$ beamspace covariance matrix is then barely visible.
 The time it takes to construct all covariance matrices is also reduced, but for few beams this is now the major contributor to the total execution time.
-If real-time is defined as 10 FPS \cite{Chen2011} almost all execution times in both Fig.
-\ref{fig:benchUS64} and \ref{fig:benchBS} are real-time.
+If real-time is defined as 10 FPS \cite{Chen2011} almost all execution times in both Fig.\ \ref{fig:benchUS64} and \ref{fig:benchBS} are real-time.
 More natural for cardiac ultrasound imaging is to define real-time as the time it takes to acquire the underlying image.
@@
 The per-pixel granularity suggests that the beamformer is easily dividable across multiple GPUs.
-Real-time execution for all cases presented in this paper is therefore possible if several GPUs are used simultaneously.
-However, there is usually a space-cost-energy limitation for how many GPUs that can be fitted in an ultrasound scanner.
-
-Turning to image quality we see from Fig.
+It is therefore possible to achieve higher frame rates if several GPUs are used simultaneously.
+However, there is usually a space-cost-energy limitation for how many GPUs that can be assembled in an ultrasound scanner.
+
+%Turning to image quality we see from Fig.
 \ref{fig:phantom}, and the attached video of the simulation, that Capon beamforming provides improved lateral resolution of moving point targets both in element- and in a reduced beam-space ($N_b=3$).
-Our preliminary finding is that the Capon beamformers are local shift invariant as long as the lateral sampling is increased to a certain level \multimedia{Media-Movie 1 and 2}.
-How large this oversampling factor needs to be and how the oversampling should be conducted is outside the scope of this paper.
-No major improvement in contrast is visually observed in the presented images and videos, other than slightly rounder circles due to mainlobe micro steering.
-In the cardiac recording (Fig.
-\ref{fig:invivo}), structures are thinner and better resolved when the Capon beamformers are applied.
-However we observe a loss in contrast, especially in the septum (Fig.
-\ref{fig:slices}).
+Our preliminary finding is that the Capon beamformers are local shift invariant as long as the lateral sampling is increased to a certain level in simulations \multimedia{Media-Movie 1 and 2}.
+%How large this oversampling factor needs to be and how the oversampling in the end is conducted will impact the real time requirement, and .
+%Future work should focus on how large this oversampling needs to be and how it should be conducted in order to still achieve real time processing.
+%No major improvement in contrast is visually observed in the presented images and videos, other than slightly rounder circles due to mainlobe micro steering.
+
+Turning to image quality we see from Fig.\ \ref{fig:speed_res_trade} and the attached videos, that Capon beamforming provides improved lateral resolution of moving point targets both in element- and in a reduced beam-space ($N_b=3$).
+In the cardiac recording (Fig.\ \ref{fig:invivo}), structures are thinner and better resolved when the Capon beamformers are applied.
+One example is the mitral valve located at 10 cm depth and -1.5 cm width.
+On the other hand we observe a loss in contrast, especially in the septum (Fig.\ \ref{fig:invivo}, 6 cm depth, -2 cm width).
 This loss can be partially explained by the presence of highly correlated signals and mismatch between the assumed steering vector and signal.
-Future work should focus on solving this issue.
-Cardiac ultrasound has not been selected as the target modality in this paper because Capon beamforming shows large improvements in image quality in particular, it has been selected because it is the most computationally demanding modality (measured in float operations per second).
-The rationale is that if BS-Capon can run in real time for cardiac ultrasound it can run in real time for all medical ultrasound modalities.
-We believe that Capon will show more improvements in other imaging modalities.
-
-Nilsen and Hafizovic suggested to do automatic selection of beams in beamspace, hence picking the beams containing most energy.
+This is the same issue that made us use 16x oversampling when producing Fig.\ \ref{fig:speed_res_trade}.
+Future work should focus on how large this oversampling needs to be and how it should be conducted in order to still achieve real time processing.
+
+In the end we want to point out that cardiac ultrasound has not been selected as the target modality in this paper because Capon beamforming shows large improvements in image quality in particular.
+It has been selected because it is the most computationally demanding modality, measured in floating points operations per second.
+The rationale being that if BS-Capon can run in real time for cardiac ultrasound it can run in real time for all medical ultrasound modalities.
+We believe that Capon beamforming may show more improvements in other imaging modalities where detection of small point targets is important.
+
+%Nilsen and Hafizovic suggested to do automatic selection of beams in beamspace, hence picking the beams containing most energy.
 In this paper the center beams are always picked.
@@
 In this paper a parallel formulation and implementation of the beamspace Capon beamformer have been presented, and
-the beamformer have successfully been implemented on the GPU.
+the beamformer has successfully been implemented on the GPU.
 For a three dimensional beamspace, processing rates supporting real-time cardiac ultrasound imaging was shown to be possible, meaning that processing rates higher than the image acquisition rate are obtained for a wide range of parameters.
-Image quality was investigated both in a simulated phantom and an \textit{in-vivo} cardiac dataset.
+Image quality was investigated in an \textit{in-vivo} cardiac dataset.
 These results showed that Capon beamforming is feasible for cardiac ultrasound imaging, providing images with improved lateral resolution both in element- and beamspace.
@@
 \section*{Acknowledgment}
-The authors would like to thank Dr. K. Kristoffersen, GE Vingmed Ultrasound, for valuable comments.
+The authors would like to thank Prof. K. Kristoffersen, GE Vingmed Ultrasound and Norwegian University of Science and Technology, for valuable comments.
 
@@
 He has industry experience from GE Vingmed Ultrasound, Horten, Norway, were he completed three summer internships from 2008 to 2010.
-His research interests include adaptive ultrasound processing techniques and acceleration of ultrasound algorithms using Graphics Processing Units (GPUs).
+His research interests include image and signal processing, adaptive ultrasound processing techniques and acceleration of ultrasound algorithms using Graphics Processing Units (GPUs).
 \end{IEEEbiography}

### 2013-09-20 HRUIofMovingObjects/jrnl_article.tex
updated tuffc article, maybe som changes from ius as well
Started working on shift-invariance paper

@@
 \renewcommand{\vec}[1]{\mathbf{#1}}
+\newcommand{\R}{$\mat{\hat{R}}$ }
+
+  \newcommand{\img}{img/}
+
+\newcommand\multimedia[1]{\textbf{{\color{red}#1}}}
 
@@
 % can use linebreaks \\ within to get better formatting as desired
-\title{Capon Adaptive Beamforming Applied on Moving Objects - An Analysis of Local Shift Invariance}
+\title{Capon Beamforming Applied on Moving Objects - An Analysis of Local Shift Invariance}
 %
@@
 
-\author{J. P. \AA{}sen, C.-I. C. Nilsen, A. Austeng and S. Holm%\IEEEmembership{Member,~IEEE,}% <-this % stops a space
-\thanks{M. Shell is with the Department
-of Electrical and Computer Engineering, Georgia Institute of Technology, Atlanta,
-GA, 30332 USA e-mail: (see http://www.michaelshell.org/contact.html).}% <-this % stops a space
-\thanks{J. Doe and J. Doe are with Anonymous University.}% <-this % stops a space
-\thanks{Manuscript received April 19, 2005; revised January 11, 2007.}}
+\author{
+   J. P. \AA{}sen, A. Austeng and S. Holm%\IEEEmembership{Member,~IEEE,}% <-this % stops a space
+
+   \thanks{...}
+}
+%M. Shell is with the Department
+%of Electrical and Computer Engineering, Georgia Institute of Technology, Atlanta,
+%GA, 30332 USA e-mail: (see http://www.michaelshell.org/contact.html).}% <-this % stops a space
+%\thanks{J. Doe and J. Doe are with Anonymous University.}% <-this % stops a space
+%\thanks{Manuscript received April 19, 2005; revised January 11, 2007.}}
 
@@
 % The paper headers
-\markboth{Transactions on Ultrasonics and Ferroelectrics and Frequency Control,~Vol.~x, No.~y, January~2013?}%
+\markboth{Transactions on Ultrasonics and Ferroelectrics and Frequency Control,~Vol.~x, No.~y, January~201\AA{}}%
 {\AA{}sen \MakeLowercase{\textit{et al.}}: The Capon Adaptive Beamformer Applied on Moving Objects }
@@
 
-Delay and sum beamformer (DAS):
-\begin{align}
+% needed in second column of first page if using \IEEEpubid
+%\IEEEpubidadjcol
+
+\section{Background}
+
+The standard form of time domain array beamforming is the delay and sum beamformer (DAS):
+\begin{align}\label{eq:das}
 z[n] = \sum_{m = 0}^{M-1}\vec{w}_m^*\vec{x}_m[n - \Delta_m[n]] = \vec{w}^H\vec{x}[n],
 \end{align}
-where $M$ is the number of elements.
+where $M$ is the number of elements or channels in the array, $\Delta_m[n]$ is the per-element focusing and steering delay, and $\vec{w}$ is a weight vector or window function.
+The weight vector is typically selected in order to trade resolution for a lower side lobe level.
 
-Capon weights are found by solving the following minimization problem:
+\subsection{Capon Beamforming}
+The Capon beamformer produce a weight vector, like the one in (\ref{eq:das}), based on the impinging signal and an optimization criteria.
+The weights are found by solving the following minimization problem:
 \begin{align}
-&\min_{\vec{w}} E\{|z[n]|^2\} = \vec{w}^H\mat{R}\vec{w}\\
-&\text{subjected to } \vec{w}^H\vec{1} = 1,
+&\min_{\vec{w}} E\{|z[n]|^2\} \rightarrow \min_{\vec{w}} \vec{w}^H \mat{\hat{R}} \vec{w} \label{eq:capon_optimization_criteria} \\
+&\text{subjected to } \vec{w}^H\vec{a} = 1,
 \end{align}
-where $\mat{R} = E\{\vec{x}[n]\vec{x}[n]^H\}$.
+where $\mat{\hat{R}}$ is the sample covariance matrix.
+Hence, the resulting weight vector will minimizes the output power while maintaining unit gain along the steering direction $\vec{a}$.
+%The covariance matrix is usually unknown and has to be estimated from the element data.
+The covariance matrix in (\ref{eq:capon_optimization_criteria}) is therefore substituted with the sample covariance matrix.
 
-The solution to this minimization problem is
+The sample covariance matrix can be estimated in the following way \cite{Synnevag2009}:
 \begin{align}
-\vec{w}[n] = \frac{\mat{R}[n]^{-1}\vec{a}}{\vec{a}^H\mat{R}[n]^{-1}\vec{a}},
+\mat{\breve{R}} = \frac{1}{N_LN_K}\sum_{n'=n-K}^{n+K} \sum_{l=0}^{N_L-1} \vec{x}_l[n']\vec{x}_l[n']^H,
 \end{align}
-where $\vec{a} = \vec{1}$ when $\vec{x}$ are pre-delayed to broadside,  and $\vec{w} \in \mathbb{C}^M$.
+where $K = M-L+1$, $K\sim \tau/T_s$ ($\tau$ is the imaging pulse length and $T_s$ is the sampling period.) and $\vec{x}_l$ is the $l\text{th}$ subarray $[x_l[n], \dotso x_{l+L}[n]]$ of length $L$.
+Finally $\mat{\hat{R}}$ is loaded with diagonal factor for numerical stability and increased robustness,
+\begin{align}
+\epsilon &= d*\text{trace}\{\mat{\breve{R}}\}\\
+\mat{\hat{R}} &= \mat{\breve{R}} + \epsilon\mat{I}.
+\end{align}
 
-Estimation of $\mat{R}[n]$: $\mat{\hat{R}}[n] = \vec{x}[n]\vec{x}[n]^H$.
+The solution to the minimization problem in (\ref{eq:capon_optimization_criteria}) is
+\begin{align}
+\vec{w}[n] = \frac{\mat{\hat{R}}[n]^{-1}\vec{a}}{\vec{a}^H\mat{\hat{R}}[n]^{-1}\vec{a}} \in \mathbb{C}^L,
+\end{align}.
 
-In order to get a well conditioned $\mat{\hat{R}}$, avoid signal cancellation and to get DAS-like speckle, $\mat{\hat{R}}$ has to be averaged over $L\le M/2$ subarrays and $N_{avg}$ time samples.
-\begin{align}
-\mat{\hat{R}} = \frac{1}{(2N_{avg} + 1)K}\sum_{n'=n-N_{avg}}^{n+N_{avg}} \sum_{l=0}^{K} \vec{x}_l[n']\vec{x}_l[n']^H,
-\end{align}
-where $K = M-L+1$, $Y_{avg}\sim \tau/T_s$ ($\tau$ is the imaging pulse length and $T_s$ is the sampling period.) and $\vec{x}_l$ is the $l^\text{th}$ subarray $[x_l[n], \dotso x_{l+L}[n]]$.
+Note that one matrix has to be constructed and inverted for each raw data pixel in the image.
+This is indeed computational demanding, and has for long limited the methods accessibility.
+Following the innovation in GPU computing, this has now started to change.
 
-Finally, $\mat{\hat{R}}$ is loaded with diagonal factor $\epsilon$, $\mat{\hat{R}} = \mat{\hat{R}} + \epsilon\mat{I}$.
-A weighting proportional to the output power is beneficial, and the trace of $\mat{\hat{R}}$, $\epsilon = d*tr\{\mat{\hat{R}}\}$, has previously shown good results in the literature (cite).
-
-Building and calculating the inverse of $\hat{R}$ is computational demanding, limiting the methods accessibility. Following the innovation in GPU computing, this has now started to change.
-
-Nilsen et.
+%Nilsen et.
 al has proposed to apply beamspace processing to further reduce the computations required.
@@
 
-% needed in second column of first page if using \IEEEpubid
-%\IEEEpubidadjcol
+\subsection{Sampling and Shift-Invariance}
 
-\section{Temporal Artifacts}
+\begin{figure*}[!t]
+\centerline{
+\subfloat[]{\includegraphics[width=0.5\linewidth]{\img das_no_undersampling.eps}\label{fig:das_shiftinvariant}}
+\hfill{}
+\subfloat[]{\includegraphics[width=0.5\linewidth]{\img das_undersampled_2x.eps}\label{fig:das_shiftvariant}}
+}
+\caption{...}
+\label{fig:das}
+\end{figure*}
 
-\section{Causes}\label{sec:meth}
+\section{Local Shift Invariance of the Capon Beamformer}
 
-\subsection{Local Shift Variance}
+\begin{figure*}[!t]
+\centerline{
+\subfloat[]{
+\includegraphics[width=0.5\linewidth]{\img capon_L=32_K=1_d=001_normal.eps}\label{fig:capon}
+}
+\subfloat[]{
+\includegraphics[width=0.5\linewidth]{\img capon_L=32_K=1_d=1_normal.eps}
+}
+}
+\caption{...}
+\label{fig:image_capon}
+\end{figure*}
 
-\section{Cures}
+\section{Oversampling on Transmitt}
 
-\subsection{Multiple Parallel Recieve Lines}
+\begin{figure*}[!t]
+\centerline{
+\subfloat[]{
+\includegraphics[width=0.5\linewidth]{\img capon_L=32_K=1_d=001_2x.eps}
+}
+\subfloat[]{
+\includegraphics[width=0.5\linewidth]{\img capon_L=32_K=1_d=001_4x.eps}
+}
+}
+\caption{...}
+\label{fig:capon_oversampling}
+\end{figure*}
 
-\subsection{Phase Rotation}
+\section{Oversampling by Phase Rotation}
+
+\begin{figure*}[!t]
+\centerline{
+\subfloat[]{
+\includegraphics[width=0.5\linewidth]{\img capon_L=32_K=1_d=001_2x_microsteering_correct_pitch.eps}
+}
+\subfloat[]{
+\includegraphics[width=0.5\linewidth]{\img capon_L=32_K=1_d=001_4x_microsteering_correct_pitch.eps}
+}
+}
+\caption{...}
+\label{fig:benchmark_capon_bs}
+\end{figure*}
+
+\section{Oversampling by Parallel Recieve Lines}
+\begin{figure}[!t]
+\centering
+\includegraphics[width=\linewidth]{\img capon_L=32_K=1_d=001_2MLA_and_2x_microsteering_correct_pitch.eps}
+\label{fig:image_capon_bs}
+\end{figure}
+
+\section{Simulation Setup}
 
@@
 
-\begin{IEEEbiography}{Carl-Inge Colombo Nilsen}
-Biography text here.
-\end{IEEEbiography}
+%\begin{IEEEbiography}{Carl-Inge Colombo Nilsen}
+%Biography text here.
+%\end{IEEEbiography}
 

### 2013-10-11 HRUIofMovingObjects/jrnl_article.tex
Started writing moving object article

@@
 \newcommand\multimedia[1]{\textbf{{\color{red}#1}}}
+\newcommand\comment[1]{\textit{{\color{red}#1}}}
 
@@
 %\boldmath
-The abstract goes here.
+\comment{The abstract goes here.}
 \end{abstract}
@@
 
-\IEEEPARstart{I}{ntroduction} to Capon adaptive beamforming and temporal artifacts.
+\IEEEPARstart{W}{ith} the introduction of a fast implementation of the Capon adaptive beamformer, a stream of images can now be processed at interactive frame rates \cite{Asen}.
+A natural question then is how a moving object appear after Capon beamforming has been applied.
+Earlier work on Capon beamforming for medical ultrasound imaging has investigated the method's capabilities in single frames only, and temporal behaviour and effects has therefore been neglected.
 
-With the introduction of a fast implementation of the Capon adaptive beamformer, a stream of images can now be processed at interactive frame rates.
-A natural question then is how a moving object behaves throughout the stream of images.
 An important property of an ultrasound imaging system is its local shift invariance.
-An object’s appearance should be preserved when moved slightly, or if the ultrasound probe moves.
-The Capon beamformer has previously only been investigated intra-frame; it is therefore unknown how it works when applied in a real-time ultrasound setting.
+Local shift invariance means that an object’s appearance should be preserved when the object, or the ultrasound probe, is moved slightly.
+Local shift invariance is what makes techniques like doppler and flow imaging and speckle tracking possible \comment{(need citation)} .
+
+Capon beamforming has the ability to provide increased lateral resolution and contrast in an ultrasound image \cite{Synnevag2009, Chen2011}.
+However, it is also sensitive to errors like perturbations of the array element, correlated signals and errors in the assumed steering vector.
+
+\comment{Add intro to other work.}
 
@@
 We show that when imaging moving point scatterers using the Capon beamformer with the same beam spacing as used for delay-and-sum, artifacts arise in the image, manifested as oscillating point scatteres.
-(TODO: Process csound data to see if we need over sampling on in vivo data, How often do we have bright point scatterers in vivo?).
+\comment{(TODO: Process csound data to see if we need over sampling on in vivo data, How often do we have bright point scatterers in vivo?
+Might improve contrast further)}.
 
@@
 
-The standard form of time domain array beamforming is the delay and sum beamformer (DAS):
+The standard form of time domain array beamforming is the delay-and-sum beamformer (DAS):
 \begin{align}\label{eq:das}
@@
 \subsection{Capon Beamforming}
-The Capon beamformer produce a weight vector, like the one in (\ref{eq:das}), based on the impinging signal and an optimization criteria.
-The weights are found by solving the following minimization problem:
+The Capon beamformer produce a weight vector, like the one in (\ref{eq:das}), based on the impinging signal and an optimization criteria. The minimization problem is as follow:
 \begin{align}
@@
 \begin{align}
-\mat{\breve{R}} = \frac{1}{N_LN_K}\sum_{n'=n-K}^{n+K} \sum_{l=0}^{N_L-1} \vec{x}_l[n']\vec{x}_l[n']^H,
+\mat{\breve{R}}[n] = \frac{1}{N_LN_K}\sum_{n'=n-K}^{n+K} \sum_{l=0}^{N_L-1} \vec{x}_l[n']\vec{x}_l[n']^H,
 \end{align}
-where $K = M-L+1$, $K\sim \tau/T_s$ ($\tau$ is the imaging pulse length and $T_s$ is the sampling period.) and $\vec{x}_l$ is the $l\text{th}$ subarray $[x_l[n], \dotso x_{l+L}[n]]$ of length $L$.
-Finally $\mat{\hat{R}}$ is loaded with diagonal factor for numerical stability and increased robustness,
+where $K = M-L+1$ typical is proportional to the pulse length, $N_K = 2K + 1$, $N_L = M-L+1$, and $\vec{x}_l$ is the $l\text{th}$ subarray $[x_l[n], \dotso x_{l+L}[n]]$ of length $L$.
+Finally $\mat{\hat{R}}$ is loaded with a diagonal factor for numerical stability and increased robustness,
 \begin{align}
-\epsilon &= d*\text{trace}\{\mat{\breve{R}}\}\\
+\epsilon &= d*\text{trace}\{\mat{\breve{R}}\}/L\\
 \mat{\hat{R}} &= \mat{\breve{R}} + \epsilon\mat{I}.
@@
 
-Note that one matrix has to be constructed and inverted for each raw data pixel in the image.
+Note that one matrix has to be constructed and inverted for each data vector $\vec{x}$ received by the system.
 This is indeed computational demanding, and has for long limited the methods accessibility.
-Following the innovation in GPU computing, this has now started to change.
+Following the innovation in GPU computing, this has now changed.
 
@@
 \subsection{Sampling and Shift-Invariance}
+
+The lateral sampling of an ultrasound image is given by the aperture size on transmit ($D_{tx}$) and receive ($D_{rx}$) combined and the center frequency ($f_c = c/\lambda$), where $c$ is the speed of sound in the media and $\lambda$ is the wavelength.
+The required beam spacing is then:
+\begin{align}
+\Delta < \frac{\lambda}{D_{tx} + D_{rx}}.
+\end{align}
+If this requirement is not met, aliasing effects will appear in the image and the system will not be lateral shift invariant.
+
+In order to visualize the local shift-invariance of a given imaging system, we will use lateral shift invariance plots, introduced by Hergum \textit{et al} \cite{Hergum2007}.
 
@@
 \section{Results}\label{sec:res}
+Make plot of contrast between two areas in a simulation for different values of $L$.
+Make the same plot for different number of rx beams.
+Do the same for an in-vivo image.
+The rational being that contrast is also improved by denser rx-sampling not just shift-invariance.
 

### 2013-10-25 HRUIofMovingObjects/jrnl_article.tex
Finished 1st draft of introduction in Capon on moving object paper

@@
 % can use linebreaks \\ within to get better formatting as desired
-\title{Capon Beamforming Applied on Moving Objects - An Analysis of Local Shift Invariance}
+\title{Capon Beamforming Applied on Moving Objects - An Analysis of Local Shift-Invariance}
 %
@@
 \IEEEPARstart{W}{ith} the introduction of a fast implementation of the Capon adaptive beamformer, a stream of images can now be processed at interactive frame rates \cite{Asen}.
-A natural question then is how a moving object appear after Capon beamforming has been applied.
+It is therefore an interesting question how a moving object appears after Capon beamforming has been applied.
 Earlier work on Capon beamforming for medical ultrasound imaging has investigated the method's capabilities in single frames only, and temporal behaviour and effects has therefore been neglected.
 
-An important property of an ultrasound imaging system is its local shift invariance.
-Local shift invariance means that an object’s appearance should be preserved when the object, or the ultrasound probe, is moved slightly.
-Local shift invariance is what makes techniques like doppler and flow imaging and speckle tracking possible \comment{(need citation)} .
+An important property of an ultrasound imaging system is its local shift-invariance.
+Local shift-invariance means that an object’s appearance should be preserved when the object, or the ultrasound probe, is moved slightly.
+This local movement invariance is what makes ultrasound imaging feasible in general, however it is specially important for quantitative techniques that correlate time-varying features, like pulsed doppler and flow imaging, and speckle tracking of tissue and blood.
+In this paper we analyze the shift-invariance property of the Capon beamformer by studying the beamformer output when applied on simulated and \textit{in vitro} images of moving objects.
 
-Capon beamforming has the ability to provide increased lateral resolution and contrast in an ultrasound image \cite{Synnevag2009, Chen2011}.
-However, it is also sensitive to errors like perturbations of the array element, correlated signals and errors in the assumed steering vector.
+Capon beamforming has been shown to provide increased lateral resolution and contrast in ultrasound images \cite{Synnevag2007, Synnevag2009, Chen2011}.
+However, it is also known for its suboptimal performance if the sample size available for statistical estimation is limited, if correlated signals are present, and if the assumed steering vector does not match the signal propagation vector\cite{Mestre2006, Widrow1982, Wax1996, Wax1996a}.
+This has led to the development of several robust versions of the Capon beamformer using either spatial smoothing \cite{Shan1985}, diagonal loading \cite{JianLi2003} or steering vector uncertainty sets \cite{Lorenz2005, Rubsamen2013}.
+The first two, together with temporal smoothing, have been applied in earlier work on Capon beamforming for medical ultrasound imaging \cite{Synnevag2009}, and both trades resolution for robustness.
+The latter has not been applied, even though it is well known that the Capon beamformer exhibits severe signal-nulling if the steering vector does not match the signal steering vector exactly.
+This self-nulling actually contribute a lot to the method's super resolution capability in passive-narrowband applications.
 
-\comment{Add intro to other work.}
+In \cite{Cox1973}, Cox derives formulas for the resolution of the Capon beamformer (called the $\vec{k}_3$ processor), and presents in great detail how the amount of mismatch between the selected steering vector and the signal propagation vector impacts the beamformer output in simple situations.
+Cox states the following important fact about the Capon beamformer (the $\vec{k}_3$ processor):
+\begin{quote}
+"\textit{...
+a $\vec{k}_3$ processor requires more closely spaced beams than a ...
+$\vec{k}_1$-processor in order to avoid serious signal suppression effects being introduced on signals arriving from directions between the beams}",
+\end{quote}
+where $\vec{k}_1$ is the delay-and-sum (DAS) beamformer.
+In Fig.\,\ref{fig:das_capon_beams} we have plotted the steered response for DAS and the Capon beamformer using (26) and (32) in \cite{Cox1973}.
+A 64-element linear array with $\lambda/2$ element spacing, where $\lambda$ is the wavelength, is scanned over a sector equal to three times the system resolution.
+Impinging on the array is a plane wave with a propagation vector orthogonal to the array surface normal with no spatially white noise present.
+Hence, the interference-noise matrix $\sigma_0^2\mat{Q}$ is the identity.
+We have also added the effect of typical sub-array averaging and diagonal loading \cite{Asen} to the Capon beamformer output.
+The figure shows that when the DAS beamformer is used, only a 3.92 dB scallop loss \cite{Harris1978} should be observed if the source of the plane wave starts moving parallel to the array.
+However if the Capon beamformer is used with the beam density sufficient for DAS the scallop loss could be as large as $31$ dB.
 
-In this paper we analyze the shift invariance property of the Capon beamformer by studying simulated and in vivo images of moving objects.
-We show that when imaging moving point scatterers using the Capon beamformer with the same beam spacing as used for delay-and-sum, artifacts arise in the image, manifested as oscillating point scatteres.
-\comment{(TODO: Process csound data to see if we need over sampling on in vivo data, How often do we have bright point scatterers in vivo?
-Might improve contrast further)}.
+\begin{figure}[!t]
+\centerline{
+\includegraphics[width=\linewidth]{\img steerd_respons_capon_das_resolution.eps}
+}
+\caption{...}
+\label{fig:das_capon_beams}
+\end{figure}
+
+In previous literature on Capon beamforming for medical ultrasound imaging in single frames this effect has been passed over in silence by either oversampling on transmit, or carefully position point scatterers on transmit-receive beam pairs, or ignorance.
+In this paper we show that when imaging a moving point scatterer with ultrasound using the Capon beamformer with the same beam spacing used for DAS the same scallop loss is observed, manifested as oscillating point scatteres.
+The question is then, what can we do in order to improve on the situation.
 
@@
 \end{align}
-where $\mat{\hat{R}}$ is the sample covariance matrix.
-Hence, the resulting weight vector will minimizes the output power while maintaining unit gain along the steering direction $\vec{a}$.
+where $\mat{\hat{R}}$ is a sample covariance matrix.
+Hence, the resulting weight vector will minimizes the output power while maintaining unit gain in the steering direction $\vec{a}$.
 %The covariance matrix is usually unknown and has to be estimated from the element data.
@@
 where $K = M-L+1$ typical is proportional to the pulse length, $N_K = 2K + 1$, $N_L = M-L+1$, and $\vec{x}_l$ is the $l\text{th}$ subarray $[x_l[n], \dotso x_{l+L}[n]]$ of length $L$.
-Finally $\mat{\hat{R}}$ is loaded with a diagonal factor for numerical stability and increased robustness,
+Finally the matrix is loaded with a diagonal factor for numerical stability and increased robustness,
 \begin{align}
@@
 \begin{align}
-\vec{w}[n] = \frac{\mat{\hat{R}}[n]^{-1}\vec{a}}{\vec{a}^H\mat{\hat{R}}[n]^{-1}\vec{a}} \in \mathbb{C}^L,
-\end{align}.
+\vec{w}[n] = \frac{\mat{\hat{R}}[n]^{-1}\vec{a}}{\vec{a}^H\mat{\hat{R}}[n]^{-1}\vec{a}} \in \mathbb{C}^L.
+\end{align}
 
-Note that one matrix has to be constructed and inverted for each data vector $\vec{x}$ received by the system.
+Note that one matrix has to be constructed and inverted for each data vector $\vec{x}[n]$ received by the system.
 This is indeed computational demanding, and has for long limited the methods accessibility.
-Following the innovation in GPU computing, this has now changed.
+Following the innovation in GPU computing, and by applying a novel method in order to reduce the size of the matrix inversion problem \cite{Nilsen2009}, this has now changed \cite{Asen}.
 
@@
 
-The lateral sampling of an ultrasound image is given by the aperture size on transmit ($D_{tx}$) and receive ($D_{rx}$) combined and the center frequency ($f_c = c/\lambda$), where $c$ is the speed of sound in the media and $\lambda$ is the wavelength.
-The required beam spacing is then:
+The lateral sampling of an ultrasound image is given by the aperture size on transmit ($D_{tx}$) and receive ($D_{rx}$) combined \cite{Hergum2007} and the center frequency ($f_c = c/\lambda$), where $c$ is the speed of sound in the media and $\lambda$ is the wavelength.
+The required Nyquist beam spacing is then:
 \begin{align}
@@
 \end{align}
-If this requirement is not met, aliasing effects will appear in the image and the system will not be lateral shift invariant.
+If this requirement is not fulfilled, aliasing effects will appear in the image and the system will not be lateral shift invariant.
 
@@
 TODO: Add citations – Hergum/Bjåstad Local shift invariance and MLA’s… Is there any publications stating that Capon do not work on moving objects?
+
+\subsection{Impact on Real Time Performance}
+Discuss, and show with benchmarks how the different methods for making Capon beamforming shift-invariant impact the total execution time.
 

### 2013-10-25 HRUIofMovingObjects/jrnl_article.tex
Moving objects

@@
 }
-\caption{...}
+\caption{Comparison of drop in output power in between beams (half the system resolution) for delay-and-sum (DAS) and Capon beamforming.}
 \label{fig:das_capon_beams}
@@
 
-In previous literature on Capon beamforming for medical ultrasound imaging in single frames this effect has been passed over in silence by either oversampling on transmit, or carefully position point scatterers on transmit-receive beam pairs, or ignorance.
-In this paper we show that when imaging a moving point scatterer with ultrasound using the Capon beamformer with the same beam spacing used for DAS the same scallop loss is observed, manifested as oscillating point scatteres.
+In previous literature on Capon beamforming for medical ultrasound imaging this effect has been passed over in silence (in single frames) by either oversampling on transmit, or carefully position point scatterers on transmit-receive beam pairs, or ignorance.
+In this paper we show that when imaging a moving point scatterer with ultrasound and the Capon beamformer with the same beam spacing sufficient for DAS the same scallop loss is observed, manifested as oscillating point scatteres.
 The question is then, what can we do in order to improve on the situation.
+
+\comment{Next sections summary.}
 
@@
 \begin{align}
-\Delta < \frac{\lambda}{D_{tx} + D_{rx}}.
+\Delta < \frac{\lambda}{D_{tx} + D_{rx}}. \label{eq:resolution}
 \end{align}
@@
 
-In order to visualize the local shift-invariance of a given imaging system, we will use lateral shift invariance plots, introduced by Hergum \textit{et al} \cite{Hergum2007}.
+In order to visualize the local lateral shift-invariance of a given imaging system, we will use lateral shift-variance plots (LSV-plots), introduced by Hergum \textit{et al.} \cite{Hergum2007}.
+A LSV-plot is constructed for a given imaging system by imaging a laterally moving point scatterer.
+For each lateral position the root-mean-square (RMS) is calculated in range, and the resulting beam profiles are stacked in order to produce an image.
+The system is said to be local lateral shift-invariant if the plot shows straight diagonals with close to constant amplitude.
+
+Fig.\,\ref{fig:das_shiftinvariant} presents a LSV-plot for imaging system using DAS beamforming and with a beam spacing following (\ref{eq:resolution}).
 

### 2013-10-26 HRUIofMovingObjects/jrnl_article.tex
Moving objects paper: Added new LSV plots

@@
 Local shift-invariance means that an object’s appearance should be preserved when the object, or the ultrasound probe, is moved slightly.
-This local movement invariance is what makes ultrasound imaging feasible in general, however it is specially important for quantitative techniques that correlate time-varying features, like pulsed doppler and flow imaging, and speckle tracking of tissue and blood.
+This invariance for local movements is what makes ultrasound imaging feasible in general, however it is specially important for quantitative techniques that correlate time-varying features, like pulsed doppler and flow imaging, and speckle tracking of tissue and blood.
 In this paper we analyze the shift-invariance property of the Capon beamformer by studying the beamformer output when applied on simulated and \textit{in vitro} images of moving objects.
@@
 Capon beamforming has been shown to provide increased lateral resolution and contrast in ultrasound images \cite{Synnevag2007, Synnevag2009, Chen2011}.
-However, it is also known for its suboptimal performance if the sample size available for statistical estimation is limited, if correlated signals are present, and if the assumed steering vector does not match the signal propagation vector\cite{Mestre2006, Widrow1982, Wax1996, Wax1996a}.
-This has led to the development of several robust versions of the Capon beamformer using either spatial smoothing \cite{Shan1985}, diagonal loading \cite{JianLi2003} or steering vector uncertainty sets \cite{Lorenz2005, Rubsamen2013}.
-The first two, together with temporal smoothing, have been applied in earlier work on Capon beamforming for medical ultrasound imaging \cite{Synnevag2009}, and both trades resolution for robustness.
-The latter has not been applied, even though it is well known that the Capon beamformer exhibits severe signal-nulling if the steering vector does not match the signal steering vector exactly.
+The method accomplish this by adjusting the array weights based on element-to-element covariance.
+Hence, the method yields data dependent weights.
+Capon beamforming is however also known for its suboptimal performance if the sample size available for covariance estimation is limited \cite{Mestre2006}, if correlated signals are present \cite{Widrow1982}, and if the assumed steering vector does not match the signal propagation vector \cite{Wax1996, Wax1996a}.
+This has led to the development of several robust versions of the Capon beamformer with e.g.
+spatial smoothing \cite{Shan1985} or diagonal loading \cite{JianLi2003} of the covariance estimate, or by applying steering vector uncertainty sets (SVUS) \cite{Lorenz2005, Rubsamen2013}.
+The first two, together with temporal smoothing of the covariance estimate, have been applied in earlier work on Capon beamforming for medical ultrasound imaging \cite{Synnevag2009}.
+The first two trades resolution for robustness and temporal smoothing helps to preserve speckle statistics \cite{Synnevag2007a}.
+SVUS has not been applied, even though it is well known that the Capon beamformer exhibits severe signal-nulling if the signal is present during the covariance estimation process and if this estimate is later used with a steering vector that does not match the signal propagation vector exactly.
 This self-nulling actually contribute a lot to the method's super resolution capability in passive-narrowband applications.
@@
 A 64-element linear array with $\lambda/2$ element spacing, where $\lambda$ is the wavelength, is scanned over a sector equal to three times the system resolution.
-Impinging on the array is a plane wave with a propagation vector orthogonal to the array surface normal with no spatially white noise present.
+Impinging on the array is a monochromatic plane wave with a propagation vector orthogonal to the array surface normal with no spatially white noise present.
 Hence, the interference-noise matrix $\sigma_0^2\mat{Q}$ is the identity.
-We have also added the effect of typical sub-array averaging and diagonal loading \cite{Asen} to the Capon beamformer output.
+We have also added the effect of typical subarray averaging and diagonal loading \cite{Asen} to the Capon beamformer output.
 The figure shows that when the DAS beamformer is used, only a 3.92 dB scallop loss \cite{Harris1978} should be observed if the source of the plane wave starts moving parallel to the array.
@@
 
-In previous literature on Capon beamforming for medical ultrasound imaging this effect has been passed over in silence (in single frames) by either oversampling on transmit, or carefully position point scatterers on transmit-receive beam pairs, or ignorance.
+In previous literature on Capon beamforming for medical ultrasound imaging this effect has been passed over in silence (in single frames) by either oversampling on transmit, or carefully position point scatterers on transmit-receive beam pairs, or by ignorance.
 In this paper we show that when imaging a moving point scatterer with ultrasound and the Capon beamformer with the same beam spacing sufficient for DAS the same scallop loss is observed, manifested as oscillating point scatteres.
@@
 
-The sample covariance matrix can be estimated in the following way \cite{Synnevag2009}:
+The sample covariance matrix can be estimated for an active system in the following way \cite{Synnevag2009}:
 \begin{align}
@@
 \end{align}
-where $K = M-L+1$ typical is proportional to the pulse length, $N_K = 2K + 1$, $N_L = M-L+1$, and $\vec{x}_l$ is the $l\text{th}$ subarray $[x_l[n], \dotso x_{l+L}[n]]$ of length $L$.
+where $K = M-L+1$ typically is proportional to the pulse length, $N_K = 2K + 1$, $N_L = M-L+1$, and $\vec{x}_l$ is the $l\text{th}$ subarray $[x_l[n], \dotso x_{l+L}[n]]$ of length $L$.
 Finally the matrix is loaded with a diagonal factor for numerical stability and increased robustness,
@@
 Note that one matrix has to be constructed and inverted for each data vector $\vec{x}[n]$ received by the system.
-This is indeed computational demanding, and has for long limited the methods accessibility.
-Following the innovation in GPU computing, and by applying a novel method in order to reduce the size of the matrix inversion problem \cite{Nilsen2009}, this has now changed \cite{Asen}.
+This is indeed computational demanding, but following the innovation in GPU computing, and by applying a novel method in order to reduce the size of the matrix inversion problem \cite{Nilsen2009}, real time processing is now possible \cite{Asen}.
 
@@
 
-\subsection{Sampling and Shift-Invariance}
+\subsection{Lateral Sampling and Shift-Invariance}
 
-The lateral sampling of an ultrasound image is given by the aperture size on transmit ($D_{tx}$) and receive ($D_{rx}$) combined \cite{Hergum2007} and the center frequency ($f_c = c/\lambda$), where $c$ is the speed of sound in the media and $\lambda$ is the wavelength.
+The lateral sampling of an ultrasound image is given by the aperture size on transmit ($D_{tx}$) and receive ($D_{rx}$) combined \cite{Hergum2007} and the center frequency ($f_c = c/\lambda$), where $c$ is the speed of sound in the media and $\lambda_c$ is the wavelength of the center frequency.
 The required Nyquist beam spacing is then:
@@
 \end{align}
-If this requirement is not fulfilled, aliasing effects will appear in the image and the system will not be lateral shift invariant.
+If this requirement is not fulfilled, aliasing effects will appear in the image and the system will not be lateral shift-invariant.
 
@@
 For each lateral position the root-mean-square (RMS) is calculated in range, and the resulting beam profiles are stacked in order to produce an image.
-The system is said to be local lateral shift-invariant if the plot shows straight diagonals with close to constant amplitude.
+The image is displayed as a contour plot with a contour on -1, -2, -3, -6, -12 and -24 dB.
+The maximum for each beam profile is also marked with a black dot.
+The system is said to be local lateral shift-invariant if the contours are straight diagonals, hence the diagonals have constant amplitude.
+
+To further quantify the shift-invariance we have calculated the mean absolute error (MAE) of the maximum point in each beam profile versus $f(x)=x$ (P-MAE), and the negative MAE of the amplitude variation for the same set of points (A-MAE).
+The first measure quantifies the amount of geometric distortion, and the latter quantifies the gain variation or in our case the loss of signal.
+The maximum amplitude deviation along the max-point trace is therefore also highly interesting.
+We present all these numbers together with the LSV-plots.
 
@@
 \centerline{
-\subfloat[]{\includegraphics[width=0.5\linewidth]{\img das_no_undersampling.eps}\label{fig:das_shiftinvariant}}
+\subfloat[]{\includegraphics[width=0.5\linewidth]{\img das_normal2.eps}\label{fig:das_shiftinvariant}}
 \hfill{}
-\subfloat[]{\includegraphics[width=0.5\linewidth]{\img das_undersampled_2x.eps}\label{fig:das_shiftvariant}}
+\subfloat[]{\includegraphics[width=0.5\linewidth]{\img das_undersampled2_2x.eps}\label{fig:das_shiftvariant}}
 }
@@
 \subfloat[]{
-\includegraphics[width=0.5\linewidth]{\img capon_L=32_K=1_d=001_normal.eps}\label{fig:capon}
+\includegraphics[width=0.5\linewidth]{\img capon_L=32_K=1_d=001_2.eps}\label{fig:capon}
 }
 \subfloat[]{
-\includegraphics[width=0.5\linewidth]{\img capon_L=32_K=1_d=1_normal.eps}
+\includegraphics[width=0.5\linewidth]{\img capon_L=32_K=1_d=1_2.eps}
 }
@@
 
-\section{Simulation Setup}
+\section{In-vitro Images}
+Acquire channel data from a phantom containing point wire targets and cysts while moving the probe laterally. Confirm the improved shift-invariance when using phase rotation.
+
+\section{In-Vivo Images}
+Investigate if phase rotation improves contrast in in-vivo images of the heart.
 

### 2013-10-27 HRUIofMovingObjects/jrnl_article.tex
Moving objects paper: done some writing.

@@
 \usepackage{color}
+\usepackage{epstopdf}
 
@@
 \newcommand{\img}{img/}
+\epstopdfsetup{outdir=\img}
 
@@
 It is therefore an interesting question how a moving object appears after Capon beamforming has been applied.
-Earlier work on Capon beamforming for medical ultrasound imaging has investigated the method's capabilities in single frames only, and temporal behaviour and effects has therefore been neglected.
+Earlier work on Capon beamforming for medical ultrasound imaging has investigated the method's capabilities in single frames only, and temporal behavior and effects has therefore been neglected.
 
@@
 Impinging on the array is a monochromatic plane wave with a propagation vector orthogonal to the array surface normal with no spatially white noise present.
-Hence, the interference-noise matrix $\sigma_0^2\mat{Q}$ is the identity.
+Hence, the interference-noise matrix ($\sigma_0^2\mat{Q}$) is the identity.
 We have also added the effect of typical subarray averaging and diagonal loading \cite{Asen} to the Capon beamformer output.
@@
 The solution to the minimization problem in (\ref{eq:capon_optimization_criteria}) is
-\begin{align}
+\begin{align}\label{eq:capon_weights}
 \vec{w}[n] = \frac{\mat{\hat{R}}[n]^{-1}\vec{a}}{\vec{a}^H\mat{\hat{R}}[n]^{-1}\vec{a}} \in \mathbb{C}^L.
@@
 
-The lateral sampling of an ultrasound image is given by the aperture size on transmit ($D_{tx}$) and receive ($D_{rx}$) combined \cite{Hergum2007} and the center frequency ($f_c = c/\lambda$), where $c$ is the speed of sound in the media and $\lambda_c$ is the wavelength of the center frequency.
+The lateral sampling required in an ultrasound image is given by the combined aperture size on transmit ($D_{tx}$) and receive ($D_{rx}$) \cite{Hergum2007}, and the center frequency ($f_c = c/\lambda$), where $c$ is the speed of sound in the body and $\lambda_c$ is the wavelength of the center frequency.
 The required Nyquist beam spacing is then:
 \begin{align}
-\Delta < \frac{\lambda}{D_{tx} + D_{rx}}. \label{eq:resolution}
+\Delta < \frac{\lambda_c}{D_{tx} + D_{rx}}. \label{eq:resolution}
 \end{align}
-If this requirement is not fulfilled, aliasing effects will appear in the image and the system will not be lateral shift-invariant.
+If this requirement is not fulfilled, the scallop loss will be to high and the system will not be lateral shift-invariant.
 
@@
 For each lateral position the root-mean-square (RMS) is calculated in range, and the resulting beam profiles are stacked in order to produce an image.
-The image is displayed as a contour plot with a contour on -1, -2, -3, -6, -12 and -24 dB.
+The image is in this paper displayed as a contour plot with contours on -1, -2, -3, -6, -12 and -24 dB.
 The maximum for each beam profile is also marked with a black dot.
-The system is said to be local lateral shift-invariant if the contours are straight diagonals, hence the diagonals have constant amplitude.
+The system is said to be local lateral shift-invariant if the contours are diagonal, hence the diagonals have constant amplitude.
 
-To further quantify the shift-invariance we have calculated the mean absolute error (MAE) of the maximum point in each beam profile versus $f(x)=x$ (P-MAE), and the negative MAE of the amplitude variation for the same set of points (A-MAE).
-The first measure quantifies the amount of geometric distortion, and the latter quantifies the gain variation or in our case the loss of signal.
-The maximum amplitude deviation along the max-point trace is therefore also highly interesting.
+To further quantify the shift-invariance we have calculated the mean absolute error (MAE) of the maximum point in each beam profile versus $f(x)=x$, and the MAE of the amplitude variation for the same set of points.
+The first number quantifies the amount of geometric distortion, and the latter quantifies the gain variation or in our case the loss of signal.
+The maximum amplitude deviation along the max-point trace is also measured.
 We present all these numbers together with the LSV-plots.
 
-Fig.\,\ref{fig:das_shiftinvariant} presents a LSV-plot for imaging system using DAS beamforming and with a beam spacing following (\ref{eq:resolution}).
+Fig.\,\ref{fig:das_shiftinvariant} presents a LSV-plot for an imaging system using DAS beamforming with a beam spacing equal to (\ref{eq:resolution}).
+The red dotted vertical lines indicates the position of tx-rx beams.
+The point scatterer response has been simulated using Field II \comment{add citations}.
+We see that the contours are approximately diagonal and therefore we can conclude that the system is lateral shift-invariant.
+The P-MAE and A-MAE are also minimal.
+\comment{Why is the scallop loss only 0.78 dB when Fig.
+1 says it could be 3.92?}.
+In Fig.\,\ref{fig:das_shiftvariant} the beam spacing is reduced by a factor of two.
+We see that the scallop loss increases to a level where the contours no longer are diagonal.
+This loss can not be reconstructed through IQ or in-coherent interpolation, the max peak is forever lost when the point is located between two beams and the distance between the two beams are to high.
 
@@
 \section{Local Shift Invariance of the Capon Beamformer}
+Now we will investigate the lateral local shift-invariance of the Capon beamformer with typical parameters. A
 
@@
 
-\section{Oversampling on Transmitt}
+\section{Oversampling on Transmit}
+The typical way of obtain sufficient sampling in active broadband application is by  increasing the number of transmit beams until (\ref{eq:resolution}) is fulfilled.  If the number of tr
 
@@
 \section{Oversampling by Phase Rotation}
+The Capon beamformer formula in (\ref{eq:capon_weights}) gives a direct opportunity for oversampling using phase rotation.
+The steering vector $\vec{a}$ has previously been set to $\vec{1}$ since the data are pre-delayed, but it could be varied over a set of pre-defined steering vectors.
+\comment{Look at the poster for notation}
 
@@
 
-\section{Oversampling by Parallel Recieve Lines}
-\begin{figure}[!t]
-\centering
-\includegraphics[width=\linewidth]{\img capon_L=32_K=1_d=001_2MLA_and_2x_microsteering_correct_pitch.eps}
-\label{fig:image_capon_bs}
-\end{figure}
+\section{Oversampling by Parallel Receive Lines}
+In (\comment{Cite paper where they use STB + Capon if its out}) it is shown that  synthetic transmit beamforming (STB) \cite{Hergum2007} works with Capon beamforming.
+In  \cite{Asen} we also used STB in order to increase the beam density of a cardiac recording.
+STB has been shown to remove the blocky artifacts introduced by parallel receive lines (PRL) per transmit line.
+As long as some degree of motion compensation is applied (\comment{cite bastien's paper on motion compensation }) we can regard PRL as being equal to oversampling on transmit.
+It will also increase the number of covariance matrices, and thereby inversions with a factor equal to the number of parallel lines.
+%\begin{figure}[!t]
+%	\centering
+%	\includegraphics[width=\linewidth]{\img capon_L=32_K=1_d=001_2MLA_and_2x_microsteering_correct_pitch.eps}
+%	\label{fig:image_capon_bs}
+%\end{figure}
 
 \section{In-vitro Images}
-Acquire channel data from a phantom containing point wire targets and cysts while moving the probe laterally. Confirm the improved shift-invariance when using phase rotation.
+Acquire channel data from a phantom containing point wire targets and cysts while moving the probe laterally.
+Confirm the improved shift-invariance when using phase rotation.
+See if improved sampling also increase cyst contrast.
 
@@
 Discuss results presented in Section \ref{sec:res}.
+
+When signal detection (absolute value and square root) is performed before an ultrasound image is displayed the lateral bandwidth is increased.
+In order to not introduce aliasing in the image, IQ-interpolation is typically used on ultrasound systems to double the lateral beam density before detection.
+For the same reason we need to increase the lateral sampling when Capon beamforming is applied, but now on a channel data level.
+The Capon beamformer improves the lateral resolution, hence the lateral bandwidth is increased, and higher lateral sampling is required.
+
+In this paper we have shown that this oversampling can be accomplished by phase rotation and the number of transmit beams are therefore left unchanged.
 

### 2013-10-27 HRUIofMovingObjects/jrnl_article.tex
Moving objects paper: done some writing 2.0.

@@
 \newcommand\multimedia[1]{\textbf{{\color{red}#1}}}
-\newcommand\comment[1]{\textit{{\color{red}#1}}}
+\newcommand\comment[1]{\textit{{\color{red}(#1)}}}
 
@@
 \section{Local Shift Invariance of the Capon Beamformer}
-Now we will investigate the lateral local shift-invariance of the Capon beamformer with typical parameters. A
+We will now investigate the lateral local shift-invariance of the Capon beamformer with parameters equal to the once used in earlier work \comment{cite}. A  . Fig.\,\ref{fig:capon1}.
 
@@
 \subfloat[]{
-\includegraphics[width=0.5\linewidth]{\img capon_L=32_K=1_d=001_2.eps}\label{fig:capon}
+\includegraphics[width=0.5\linewidth]{\img capon_L=32_K=1_d=001_2.eps}\label{fig:capon1}
 }
 \subfloat[]{
-\includegraphics[width=0.5\linewidth]{\img capon_L=32_K=1_d=1_2.eps}
+\includegraphics[width=0.5\linewidth]{\img capon_L=32_K=1_d=1_2.eps}\label{fig:capon2}
 }

### 2013-10-31 HRUIofMovingObjects/jrnl_article.tex
some changes to moving objects paper

--- 2013-10-26 HRUIofMovingObjects/jrnl_article.tex
@@
 
-\section{Local Shift Invariance of the Capon Beamformer}
+\section{Local Lateral Shift-Invariance of the Capon Beamformer}
 
@@
 
-\section{Oversampling on Transmitt}
+\section{Methods}
+
+\subsection{Oversampling on Transmitt}
 
@@
 
-\section{Oversampling by Phase Rotation}
+\subsection{Oversampling by Phase Rotation}
 
@@
 
-\section{Oversampling by Parallel Recieve Lines}
+\subsection{Oversampling by Parallel Recieve Lines}
 \begin{figure}[!t]
@@
 
-\section{In-vitro Images}
-Acquire channel data from a phantom containing point wire targets and cysts while moving the probe laterally. Confirm the improved shift-invariance when using phase rotation.
+\section{Results}\label{sec:res}
 
-\section{In-Vivo Images}
-Investigate if phase rotation improves contrast in in-vivo images of the heart.
-
-\section{Results}\label{sec:res}
 Make plot of contrast between two areas in a simulation for different values of $L$.
@@
 The rational being that contrast is also improved by denser rx-sampling not just shift-invariance.
+
+\subsection{In-vitro Images}
+Acquire channel data from a phantom containing point wire targets and cysts while moving the probe laterally. Confirm the improved shift-invariance when using phase rotation.
+
+\subsection{In-Vivo Images}
+Investigate if phase rotation improves contrast in in-vivo images of the heart.
 

### 2013-10-31 HRUIofMovingObjects/jrnl_article.tex

--- 2013-10-27 HRUIofMovingObjects/jrnl_article.tex
@@
 
-\section{Local Shift Invariance of the Capon Beamformer}
+\section{Local Lateral Shift-Invariance of the Capon Beamformer}
 We will now investigate the lateral local shift-invariance of the Capon beamformer with parameters equal to the once used in earlier work \comment{cite}. A  . Fig.\,\ref{fig:capon1}.
@@
 \end{figure*}
+
+\section{Methods}
 
@@
 
-\section{Oversampling by Phase Rotation}
+\subsection{Oversampling by Phase Rotation}
 The Capon beamformer formula in (\ref{eq:capon_weights}) gives a direct opportunity for oversampling using phase rotation.
@@
 
-\section{Oversampling by Parallel Receive Lines}
+\subsection{Oversampling by Parallel Recieve Lines}
 In (\comment{Cite paper where they use STB + Capon if its out}) it is shown that  synthetic transmit beamforming (STB) \cite{Hergum2007} works with Capon beamforming.
@@
 
-\section{In-vitro Images}
+\section{Results}\label{sec:res}
+
 Acquire channel data from a phantom containing point wire targets and cysts while moving the probe laterally.
@@
 See if improved sampling also increase cyst contrast.
-
-\section{In-Vivo Images}
-Investigate if phase rotation improves contrast in in-vivo images of the heart.
-
-\section{Results}\label{sec:res}
 Make plot of contrast between two areas in a simulation for different values of $L$.

### 2013-10-31 HRUIofMovingObjects/jrnl_article.tex
Moving targets paper: done some writing

@@
 For each lateral position the root-mean-square (RMS) is calculated in range, and the resulting beam profiles are stacked in order to produce an image.
-The image is in this paper displayed as a contour plot with contours on -1, -2, -3, -6, -12 and -24 dB.
+The LSV-plots in this paper are displayed as a contour plot with contours on -1, -2, -3, -6, -12 and -24 dB.
+In a ultrasound image with 256 gray levels and 50 dB dynamic range 1 dB will corresponds to 5 gray levels.
+Depending on the surroundings, this is around the lowest difference a human being can detect \comment{need reference}.
+A loss larger than 1dB will therefore be visible in the image.
 The maximum for each beam profile is also marked with a black dot.
@@
 The first number quantifies the amount of geometric distortion, and the latter quantifies the gain variation or in our case the loss of signal.
-The maximum amplitude deviation along the max-point trace is also measured.
+The maximum amplitude deviation along the max-point trace is also calculated and reported.
 We present all these numbers together with the LSV-plots.
@@
 Fig.\,\ref{fig:das_shiftinvariant} presents a LSV-plot for an imaging system using DAS beamforming with a beam spacing equal to (\ref{eq:resolution}).
-The red dotted vertical lines indicates the position of tx-rx beams.
-The point scatterer response has been simulated using Field II \comment{add citations}.
-We see that the contours are approximately diagonal and therefore we can conclude that the system is lateral shift-invariant.
+The red dotted vertical lines indicates the position of tx-rx beam pairs.
+The point scatterer response has been simulated using Field II \comment{add citations} \comment{add table with simulation settings}.
+We see that the contours are approximately diagonal and therefore we can conclude that the system is locally lateral shift-invariant.
 The P-MAE and A-MAE are also minimal.
@@
 In Fig.\,\ref{fig:das_shiftvariant} the beam spacing is reduced by a factor of two.
-We see that the scallop loss increases to a level where the contours no longer are diagonal.
-This loss can not be reconstructed through IQ or in-coherent interpolation, the max peak is forever lost when the point is located between two beams and the distance between the two beams are to high.
+We see that the scallop loss (max A-MAE) increases to a level where the contours no longer are diagonal.
+The maximum peak is lost when the point is located between two beams and the distance between the two beams are too high.
 
@@
 }
-\caption{...}
+\caption{LSV-plots of DAS beamforming.
+If the contours are diagonal the imaging system is said to be lateral local shif-invariant.
+Red vertical lines are plotted where tx-rx beams are located.
+a) Sampling corresponding to (\ref{eq:resolution}).
+b) Undersampling with a factor two.}
 \label{fig:das}
@@
 \section{Local Lateral Shift-Invariance of the Capon Beamformer}
-We will now investigate the lateral local shift-invariance of the Capon beamformer with parameters equal to the once used in earlier work \comment{cite}. A  . Fig.\,\ref{fig:capon1}.
+We will now investigate the local lateral shift-invariance of the Capon beamformer with parameters equal to the once used in earlier work \comment{cite}. A  . Fig.\,\ref{fig:capon1}.
 
@@
 
-\section{Oversampling on Transmit}
+\subsection{Oversampling on Transmit}
 The typical way of obtain sufficient sampling in active broadband application is by  increasing the number of transmit beams until (\ref{eq:resolution}) is fulfilled.  If the number of tr
@@
 \subfloat[]{
-\includegraphics[width=0.5\linewidth]{\img capon_L=32_K=1_d=001_2x.eps}
+%\includegraphics[width=0.5\linewidth]{\img capon_L=32_K=1_d=001_2x.eps}
 }
 \subfloat[]{
-\includegraphics[width=0.5\linewidth]{\img capon_L=32_K=1_d=001_4x.eps}
+%\includegraphics[width=0.5\linewidth]{\img capon_L=32_K=1_d=001_4x.eps}
 }
 }
-\caption{...}
+\caption{Oversampling on transmit ...}
 \label{fig:capon_oversampling}
-\end{figure*}
-
-\subsection{Oversampling by Phase Rotation}
-The Capon beamformer formula in (\ref{eq:capon_weights}) gives a direct opportunity for oversampling using phase rotation.
-The steering vector $\vec{a}$ has previously been set to $\vec{1}$ since the data are pre-delayed, but it could be varied over a set of pre-defined steering vectors.
-\comment{Look at the poster for notation}
-
-\begin{figure*}[!t]
-	\centerline{
-		\subfloat[]{
-			\includegraphics[width=0.5\linewidth]{\img capon_L=32_K=1_d=001_2x_microsteering_correct_pitch.eps}
-		}
-		\subfloat[]{
-			\includegraphics[width=0.5\linewidth]{\img capon_L=32_K=1_d=001_4x_microsteering_correct_pitch.eps}
-		}
-	}
-	\caption{...}
-	\label{fig:benchmark_capon_bs}
 \end{figure*}
@@
 
+\subsection{Oversampling by Phase Rotation}
+The Capon beamformer formula in (\ref{eq:capon_weights}) gives a direct opportunity for oversampling using phase rotation.
+The steering vector $\vec{a}$ has previously been set to $\vec{1}$ since the data are pre-delayed, but it could be varied over a set of pre-defined steering vectors.
+\comment{Look at the poster for notation}
+
+\begin{figure*}[!t]
+	\centerline{
+%		\subfloat[]{
+%			\includegraphics[width=0.5\linewidth]{\img capon_L=32_K=1_d=001_2x_microsteering_correct_pitch.eps}
+%		}
+%		\subfloat[]{
+%			\includegraphics[width=0.5\linewidth]{\img capon_L=32_K=1_d=001_4x_microsteering_correct_pitch.eps}
+%		}
+	}
+	\caption{Oversampling using phase rotation ...}
+	\label{fig:benchmark_capon_bs}
+\end{figure*}
+
 \section{Results}\label{sec:res}
@@
 
+\begin{figure}\caption{Plot of contrast vs oversampling factor (phase rotation) in in-vivo data.}
+\end{figure}
+
 \subsection{In-vitro Images}
@@
 
+\begin{figure*}[!t]
+\centerline{
+%\subfloat[]{\includegraphics[width=0.3\linewidth]{\img das_normal2.eps}\label{fig:das_shiftinvariant}}
+\hfill{}
+%\subfloat[]{\includegraphics[width=0.3\linewidth]{\img das_normal2.eps}\label{fig:das_shiftinvariant}}
+\hfill{}
+%\subfloat[]{\includegraphics[width=0.3\linewidth]{\img das_undersampled2_2x.eps}\label{fig:das_shiftvariant}}
+}
+\caption{LSV-plots of in-vitro data. a) Image of the data. b) Normal. c) Oversampling with phase rotation}
+\label{fig:das}
+\end{figure*}
+
 \subsection{In-Vivo Images}
 Investigate if phase rotation improves contrast in in-vivo images of the heart.
+
+\begin{figure*}[!t]
+\centerline{
+%\subfloat[]{\includegraphics[width=0.3\linewidth]{\img das_normal2.eps}\label{fig:das_shiftinvariant}}
+\hfill{}
+%\subfloat[]{\includegraphics[width=0.3\linewidth]{\img das_normal2.eps}\label{fig:das_shiftinvariant}}
+\hfill{}
+%\subfloat[]{\includegraphics[width=0.3\linewidth]{\img das_undersampled2_2x.eps}\label{fig:das_shiftvariant}}
+}
+\caption{LSV-plots of in-vivo data. a) Image of the data. b) Normal. c) Oversampling with phase rotation}
+\label{fig:das}
+\end{figure*}
 

### 2013-11-02 HRUIofMovingObjects/jrnl_article.tex
Moving objects paper
Moving obj paper: done some writing

@@
 The P-MAE and A-MAE are also minimal.
-\comment{Why is the scallop loss only 0.78 dB when Fig.
-1 says it could be 3.92?}.
+The scallop loss is smaller than what was predicted by Fig.\,\ref{fig:das_capon_beams}.
+The reason for this is that Fig.\,\ref{fig:das_capon_beams} only shows the one-way response, and the system resolution is therefore calculated based on the receive aperture only.
+The beamspacing in (\ref{eq:resolution}) is derived from the fact that the two-way response is $D_{tx}+D_{rx}$ in K-space.
+However, in space the mainlobe maintains the same width from zero too zero and therefore the resolution does not change that much.
+Sampling according to (\ref{eq:resolution}) is therefore more than what is strictly needed, thereby the smaller loss.
 In Fig.\,\ref{fig:das_shiftvariant} the beam spacing is reduced by a factor of two.
@@
 \section{Local Lateral Shift-Invariance of the Capon Beamformer}
-We will now investigate the local lateral shift-invariance of the Capon beamformer with parameters equal to the once used in earlier work \comment{cite}. A  . Fig.\,\ref{fig:capon1}.
+We will now investigate the local lateral shift-invariance of the Capon beamformer with parameters equal to the once used in earlier work \comment{cite}.
+In Fig.\,\ref{fig:capon1}.
+the same point scatterer response as in Fig.\,\ref{fig:das_shiftinvariant} has been weighted with a Capon weight calculated using (\ref{eq:capon_weights}) with $L = M/2 = 32$, $K=1$ and $d=1/100$.
+The set of lateral shifts is constructed in a way that the central beam is the only one hitting the point scatterer exactly.
+We see that the scallop loss is tremendous, and not far from the value predicted by Fig\,\ref{fig:das_capon_beams}.
+The difference between .
 
@@
 \section{Methods}
+We will now discuss different approaches to improve the lateral shift-invariance when using the Capon beamformer.
 
@@
 \subfloat[]{
-%\includegraphics[width=0.5\linewidth]{\img capon_L=32_K=1_d=001_2x.eps}
+\includegraphics[width=0.5\linewidth]{\img capon_L=32_K=1_d=001_2x_oversampling_2.eps}
 }
 \subfloat[]{
-%\includegraphics[width=0.5\linewidth]{\img capon_L=32_K=1_d=001_4x.eps}
+\includegraphics[width=0.5\linewidth]{\img capon_L=32_K=1_d=001_10x_oversampling_2_zoomed.eps}
 }

### 2013-11-04 HRUIofMovingObjects/jrnl_article.tex
Moving objects article. Done some writing and added some figures.

--- 2013-10-31 HRUIofMovingObjects/jrnl_article.tex
@@
 The first two, together with temporal smoothing of the covariance estimate, have been applied in earlier work on Capon beamforming for medical ultrasound imaging \cite{Synnevag2009}.
-The first two trades resolution for robustness and temporal smoothing helps to preserve speckle statistics \cite{Synnevag2007a}.
+The first two trade resolution for robustness and temporal smoothing helps to preserve speckle statistics \cite{Synnevag2007a}.
 SVUS has not been applied, even though it is well known that the Capon beamformer exhibits severe signal-nulling if the signal is present during the covariance estimation process and if this estimate is later used with a steering vector that does not match the signal propagation vector exactly.
@@
 In this paper we show that when imaging a moving point scatterer with ultrasound and the Capon beamformer with the same beam spacing sufficient for DAS the same scallop loss is observed, manifested as oscillating point scatteres.
-The question is then, what can we do in order to improve on the situation.
+The question is then, what can we do in order to improve on the situation without affecting the imaging frame rate.
 
@@
 The P-MAE and A-MAE are also minimal.
-\comment{Why is the scallop loss only 0.78 dB when Fig.
-1 says it could be 3.92?}.
+
+The scallop loss (max A-MAE) is smaller than what was predicted by Fig.\,\ref{fig:das_capon_beams}.
+To understand this note that Fig.\,\ref{fig:das_capon_beams} shows the one-way response of a passive array, and the system resolution has therefore been calculated based on the receive aperture only.
+Whereas the beamspacing in (\ref{eq:resolution}) is derived from the fact that the two-way lateral response of an active rectangular aperture is a $D_{tx}+D_{rx}$ wide triangle in K-space \comment{Cite Hergum et al.
+Simulation in K-space}.
+Even though the lateral bandwidth is increased, the distance to the first zero remains unchanged (The Rayleigh criteria) when multiplying two sinc functions.
+Sampling according to (\ref{eq:resolution}) is therefore more than what is strictly needed in practise.
+The loss in Fig.\,\ref{fig:das_shiftinvariant} is thus smaller than what was predicted.
+
 In Fig.\,\ref{fig:das_shiftvariant} the beam spacing is reduced by a factor of two.
-We see that the scallop loss (max A-MAE) increases to a level where the contours no longer are diagonal.
-The maximum peak is lost when the point is located between two beams and the distance between the two beams are too high.
+We see that the scallop loss increases to a level where the contours no longer are diagonal.
+Hence, the imaging system is no longer locally lateral shift-invariant.
+We observe that the maximum peak is lost when the point is located between two beams and the distance between the beams are too high.
 
@@
 \section{Local Lateral Shift-Invariance of the Capon Beamformer}
-We will now investigate the local lateral shift-invariance of the Capon beamformer with parameters equal to the once used in earlier work \comment{cite}. A  . Fig.\,\ref{fig:capon1}.
+We will now investigate the local lateral shift-invariance of the Capon beamformer with parameters equal to the once used in earlier work \comment{cite}.
+In Fig.\,\ref{fig:capon1}.
+the same point scatterer response as presented in Fig.\,\ref{fig:das_shiftinvariant} has been weighted with Capon weights calculated using (\ref{eq:capon_weights}) with $L = M/2 = 32$, $K=1$ and $d=1/100$.
+The set of lateral displacements is constructed in such a way that the center beam is the only one hitting the point scatterer exactly.
+We see that the scallop loss is tremendous, and not far from the value predicted by Fig\,\ref{fig:das_capon_beams}.
+The difference between the predicted and the simulated loss is due to several factors.
+The most important one being that Fig.\,\ref{fig:das_capon_beams} shows a single frequency whereas the RMS used in Fig.\,\ref{fig:capon1} average a wideband signal in range.
+For a slice through the center of the pulse the scallop loss is below -30dB.
+
+The diagonal factor $d$ and the effective adaptive aperture size $L$ can be used to control the mainlobe width of the Capon beamformer.
+In both extreme cases, $d=inf$ \comment{add correct symbol} and $L=1$, the Capon beamformer becomes the DAS beamformer, and therefore requires a beamspacing according to (\ref{eq:resolution}).
+In Fig.\,\ref{fig:capon2} we show how the scallop loss is reduced when the amount of diagonal loading is increased.
+We observe that Fig.\,\ref{fig:capon2} contains the same pattern as present in Fig.\,\ref{fig:das_shiftvariant}.
+From Fig.\,\ref{fig:image_capon} it should be obvious that the Capon beamformer requires an increase in the lateral sampling in order to make the imaging system locally lateral shift-invariant, or the resolution has to be reduced to the same level as DAS.
+It would be unwise to do all the computation involved with Capon beamforming with a high $d$ value and end up with a resolution equal to DAS.
+However, we see that the diagonal factor can be a useful tool if resolution has to be traded for the need of oversampling.
 
@@
 \section{Methods}
+We will now discuss different ways to improve on the lateral shift-invariance when using the Capon beamformer.
+From the previous section it should now be clear that we somehow need to increase the lateral sampling.
+However, several questions still has to be answered.
+How should the oversampling be conducted, how large does the oversampling needs to be and how will the different methods affect the system frame rate?
 
 \subsection{Oversampling on Transmit}
-The typical way of obtain sufficient sampling in active broadband application is by  increasing the number of transmit beams until (\ref{eq:resolution}) is fulfilled.  If the number of tr
+The typical way of obtaining sufficient sampling in an active broadband system is by increasing the number of transmit beams until (\ref{eq:resolution}) is fulfilled.
+In Fig.\,\ref{fig:capon_oversampling_a} and Fig.\,\ref{fig:capon_oversampling_b} we have increased the lateral sampling on transmit by a factor two and ten respectively.
+As expected the scallop loss decreases with increased lateral sampling.
+Note that Fig.\,\ref{fig:capon_oversampling_b} is zoomed in order to better see the beam-to-beam variation.
+Even though increasing the number of transmit improves the lateral shift-invariance it will also reduce the imaging frame rate with a factor equal to the oversampling factor.
+It is therefore not a feasible option for real time ultrasound imaging.
+
+From Fig.\,\ref{fig:capon_oversampling_b} we see that more than ten times oversampling has to be applied when imaging a moving point scatterer with no spatially white noise and $1/100$ in diagonal loading.
+A diagonal loading factor of $1/100$ will introduce a white noise component in the sample covariance matrix $40$ dB lower than the signal present.
+Hence, the signal-to-noise ratio before the array gain is applied is $40$ dB.
+\comment{do calculations}.
 
@@
 \subfloat[]{
-%\includegraphics[width=0.5\linewidth]{\img capon_L=32_K=1_d=001_2x.eps}
+\includegraphics[width=0.5\linewidth]{\img capon_L=32_K=1_d=001_2x_oversampling_2.eps}\label{fig:capon_oversampling_a}
 }
 \subfloat[]{
-%\includegraphics[width=0.5\linewidth]{\img capon_L=32_K=1_d=001_4x.eps}
+\includegraphics[width=0.5\linewidth]{\img capon_L=32_K=1_d=001_10x_oversampling_2_zoomed.eps}\label{fig:capon_oversampling_b}
 }
@@
 The Capon beamformer formula in (\ref{eq:capon_weights}) gives a direct opportunity for oversampling using phase rotation.
-The steering vector $\vec{a}$ has previously been set to $\vec{1}$ since the data are pre-delayed, but it could be varied over a set of pre-defined steering vectors.
+The steering vector $\vec{a}$ has in previous work been set to $\vec{1}$ since the ultrasound data are pre-delayed.
+The steering vector could be varied over a set of pre-defined steering vectors.
 \comment{Look at the poster for notation}
@@
 	\centerline{
-%		\subfloat[]{
-%			\includegraphics[width=0.5\linewidth]{\img capon_L=32_K=1_d=001_2x_microsteering_correct_pitch.eps}
-%		}
-%		\subfloat[]{
-%			\includegraphics[width=0.5\linewidth]{\img capon_L=32_K=1_d=001_4x_microsteering_correct_pitch.eps}
-%		}
+		\subfloat[]{
+			\includegraphics[width=0.5\linewidth]{\img capon_L=32_K=1_d=001_4x_oversampling_2_PR_zoomed.eps}
+		}
+		\subfloat[]{
+			\includegraphics[width=0.5\linewidth]{\img capon_L=32_K=1_d=001_10x_oversampling_2_zoomed_PR.eps}
+		}
 	}

### 2013-11-04 HRUIofMovingObjects/jrnl_article.tex
Moving objects paper: Done some writing. Added some references

@@
 The red dotted vertical lines indicates the position of tx-rx beam pairs.
-The point scatterer response has been simulated using Field II \comment{add citations} \comment{add table with simulation settings}.
+The point scatterer response has been simulated using Field II \cite{Jensen1992, Jensen1996a} \comment{add table with simulation settings}.
 We see that the contours are approximately diagonal and therefore we can conclude that the system is locally lateral shift-invariant.
@@
 The scallop loss (max A-MAE) is smaller than what was predicted by Fig.\,\ref{fig:das_capon_beams}.
-To understand this note that Fig.\,\ref{fig:das_capon_beams} shows the one-way response of a passive array, and the system resolution has therefore been calculated based on the receive aperture only.
-Whereas the beamspacing in (\ref{eq:resolution}) is derived from the fact that the two-way lateral response of an active rectangular aperture is a $D_{tx}+D_{rx}$ wide triangle in K-space \comment{Cite Hergum et al.
-Simulation in K-space}.
-Even though the lateral bandwidth is increased, the distance to the first zero remains unchanged (The Rayleigh criteria) when multiplying two sinc functions.
+To understand why note that Fig.\,\ref{fig:das_capon_beams} shows the one-way response of a passive array, and the system resolution has therefore been calculated based on the receive aperture only.
+Whereas the beamspacing in (\ref{eq:resolution}) is derived from the fact that the two-way lateral response of an active rectangular aperture is a $D_{tx}+D_{rx}$ wide triangle in K-space \cite{Hergum2009}.
+Even though the lateral bandwidth is increased, the distance to the first zero (The Rayleigh criteria) remains the same when multiplying two sinc functions.
 Sampling according to (\ref{eq:resolution}) is therefore more than what is strictly needed in practise.
-The loss in Fig.\,\ref{fig:das_shiftinvariant} is thus smaller than what was predicted.
+The loss in Fig.\,\ref{fig:das_shiftinvariant} is because of this smaller than what was predicted.
 
 In Fig.\,\ref{fig:das_shiftvariant} the beam spacing is reduced by a factor of two.
-We see that the scallop loss increases to a level where the contours no longer are diagonal.
+We see how the scallop loss is now increased to a level where the contours no longer are diagonal.
 Hence, the imaging system is no longer locally lateral shift-invariant.
@@
 The diagonal factor $d$ and the effective adaptive aperture size $L$ can be used to control the mainlobe width of the Capon beamformer.
-In both extreme cases, $d=inf$ \comment{add correct symbol} and $L=1$, the Capon beamformer becomes the DAS beamformer, and therefore requires a beamspacing according to (\ref{eq:resolution}).
+In both extreme cases, $d=\infty$ and $L=1$, the Capon beamformer becomes the DAS beamformer, and will require a beamspacing according to (\ref{eq:resolution}).
 In Fig.\,\ref{fig:capon2} we show how the scallop loss is reduced when the amount of diagonal loading is increased.
 We observe that Fig.\,\ref{fig:capon2} contains the same pattern as present in Fig.\,\ref{fig:das_shiftvariant}.
-From Fig.\,\ref{fig:image_capon} it should be obvious that the Capon beamformer requires an increase in the lateral sampling in order to make the imaging system locally lateral shift-invariant, or the resolution has to be reduced to the same level as DAS.
+From Fig.\,\ref{fig:image_capon} it should be obvious that the Capon beamformer either require an increase in the lateral sampling in order to make the imaging system locally lateral shift-invariant, or the resolution has to be reduced to the same level as DAS.
 It would be unwise to do all the computation involved with Capon beamforming with a high $d$ value and end up with a resolution equal to DAS.
@@
 From the previous section it should now be clear that we somehow need to increase the lateral sampling.
-However, several questions still has to be answered.
-How should the oversampling be conducted, how large does the oversampling needs to be and how will the different methods affect the system frame rate?
+We will investigate how this oversampling should be conducted, how large it needs to be and how the different methods will affect the system frame rate.
 
@@
 The typical way of obtaining sufficient sampling in an active broadband system is by increasing the number of transmit beams until (\ref{eq:resolution}) is fulfilled.
-In Fig.\,\ref{fig:capon_oversampling_a} and Fig.\,\ref{fig:capon_oversampling_b} we have increased the lateral sampling on transmit by a factor two and ten respectively.
+In Fig.\,\ref{fig:capon_oversampling_a} and Fig.\,\ref{fig:capon_oversampling_b} we have increased the lateral sampling on transmit by a factor four and ten respectively.
 As expected the scallop loss decreases with increased lateral sampling.
@@
 \subfloat[]{
-\includegraphics[width=0.5\linewidth]{\img capon_L=32_K=1_d=001_2x_oversampling_2.eps}\label{fig:capon_oversampling_a}
+\includegraphics[width=0.5\linewidth]{\img capon_L=32_K=1_d=001_4x_oversampling_2_zoomed.eps}\label{fig:capon_oversampling_a}
 }

### 2013-11-14 HRUIofMovingObjects/jrnl_article.tex
Moving objects article: done som writing

@@
 % can use linebreaks \\ within to get better formatting as desired
-\title{Capon Beamforming Applied on Moving Objects - An Analysis of Local Shift-Invariance}
+\title{Capon Beamforming Applied on Moving Objects - An Analysis of Local Lateral Shift-Invariance}
 %
@@
 A 64-element linear array with $\lambda/2$ element spacing, where $\lambda$ is the wavelength, is scanned over a sector equal to three times the system resolution.
-Impinging on the array is a monochromatic plane wave with a propagation vector orthogonal to the array surface normal with no spatially white noise present.
+Impinging on the array is a monochromatic plane wave with a propagation vector orthogonal to the array surface with no spatially white noise present.
 Hence, the interference-noise matrix ($\sigma_0^2\mat{Q}$) is the identity.
-We have also added the effect of typical subarray averaging and diagonal loading \cite{Asen} to the Capon beamformer output.
-The figure shows that when the DAS beamformer is used, only a 3.92 dB scallop loss \cite{Harris1978} should be observed if the source of the plane wave starts moving parallel to the array.
-However if the Capon beamformer is used with the beam density sufficient for DAS the scallop loss could be as large as $31$ dB.
+The effect of typical subarray averaging and diagonal loading ($L=32$ and $d=1/100$) has also been added to the Capon beamformer output.
+A diagonal loading of $1/100$ corresponds to an signal-to-noise ratio of 40 dB before array gain is applied.
+The figure shows that when the DAS beamformer is used, only a 3.92 dB scalloping loss \cite{Harris1978} should be observed if the source of the plane wave starts moving parallel to the array.
+However if the Capon beamformer is used with the beam density sufficient for DAS the scalloping loss could be as large as $31$ dB.
 
@@
 }
-\caption{Comparison of drop in output power in between beams (half the system resolution) for delay-and-sum (DAS) and Capon beamforming.}
+\caption{Comparison of drop in output power in between beams (half the system resolution) for delay-and-sum (DAS) and Capon beamforming ($M=64$, $L=32$, $d=0.01$).
+A diagonal loading of $1/100$ corresponds to an effective signal-to-noise ratio after array gain of 40 dB.}
 \label{fig:das_capon_beams}
@@
 
-In previous literature on Capon beamforming for medical ultrasound imaging this effect has been passed over in silence (in single frames) by either oversampling on transmit, or carefully position point scatterers on transmit-receive beam pairs, or by ignorance.
-In this paper we show that when imaging a moving point scatterer with ultrasound and the Capon beamformer with the same beam spacing sufficient for DAS the same scallop loss is observed, manifested as oscillating point scatteres.
+In previous literature on Capon beamforming for medical ultrasound imaging (single frames) this effect has been passed over in silence by either oversampling on transmit, or carefully position point scatterers on transmit-receive beam pairs, or by ignorance.
+In our first work where Capon beamforming was applied on multiple frames containing moving objects \cite{Asen} we commented on the effect, but no solution was presented other than oversampling on transmit in simulations.
+In this paper we investigate the issue further, and we show that when a moving point scatterer is imaged with ultrasound and the Capon beamformer with the same beam spacing sufficient for DAS a large scalloping loss is observed, manifested as oscillating point scatteres.
 The question is then, what can we do in order to improve on the situation without affecting the imaging frame rate.
+In this paper we therefore study the scalloping loss of the Capon beamforming and investigate different methods for oversampling and how they affect the imaging frame rate and the processing complexity.
+Before that we give an introduction to Capon beamforming for medical ultrasound imaging.
 
-\comment{Next sections summary.}
+%\comment{Next sections summary.}
 
@@
 \section{Background}
-
+\subsection{Beamforming}
 The standard form of time domain array beamforming is the delay-and-sum beamformer (DAS):
@@
 
-The lateral sampling required in an ultrasound image is given by the combined aperture size on transmit ($D_{tx}$) and receive ($D_{rx}$) \cite{Hergum2007}, and the center frequency ($f_c = c/\lambda$), where $c$ is the speed of sound in the body and $\lambda_c$ is the wavelength of the center frequency.
-The required Nyquist beam spacing is then:
+The lateral sampling required in an ultrasound image is given by the combined aperture size on transmit ($D_{tx}$) and receive ($D_{rx}$), and the center frequency ($f_c = c/\lambda$), where $c$ is the speed of sound in the body and $\lambda_c$ is the wavelength of the center frequency.
+The required Nyquist beam spacing is then \cite{Hergum2007}:
 \begin{align}
@@
 \end{align}
-If this requirement is not fulfilled, the scallop loss will be to high and the system will not be lateral shift-invariant.
+If this requirement is not fulfilled the scalloping loss could become visible in the image and the system will no longer be lateral shift-invariant.
+In an ultrasound image with 256 gray levels and 50 dB dynamic range a 1 dB loss corresponds to 5 gray levels.
+This is approximately equal to the visibility threshold (Weber fraction of 2\%) for a grayscale image, and so a loss larger than 1dB could be visible to the observer.
+In our analysis a system would therefore be regarded as lateral shift-invariant if the loss is smaller than this threshold.
 
@@
 A LSV-plot is constructed for a given imaging system by imaging a laterally moving point scatterer.
-For each lateral position the root-mean-square (RMS) is calculated in range, and the resulting beam profiles are stacked in order to produce an image.
+For each lateral position the root-mean-square (RMS) is calculated in range, and the resulting beam profiles are stacked to produce an image.
 The LSV-plots in this paper are displayed as a contour plot with contours on -1, -2, -3, -6, -12 and -24 dB.
-In a ultrasound image with 256 gray levels and 50 dB dynamic range 1 dB will corresponds to 5 gray levels.
-Depending on the surroundings, this is around the lowest difference a human being can detect \comment{need reference}.
-A loss larger than 1dB will therefore be visible in the image.
 The maximum for each beam profile is also marked with a black dot.
-The system is said to be local lateral shift-invariant if the contours are diagonal, hence the diagonals have constant amplitude.
+The system is said to be local lateral shift-invariant if the contours are diagonal, hence the diagonals have a loss less than 1 dB.
 
@@
 
-The scallop loss (max A-MAE) is smaller than what was predicted by Fig.\,\ref{fig:das_capon_beams}.
+The scalloping loss (max A-MAE) in Fig.\,\ref{fig:das_shiftinvariant} is smaller than what was predicted by Fig.\,\ref{fig:das_capon_beams}.
 To understand why note that Fig.\,\ref{fig:das_capon_beams} shows the one-way response of a passive array, and the system resolution has therefore been calculated based on the receive aperture only.
 Whereas the beamspacing in (\ref{eq:resolution}) is derived from the fact that the two-way lateral response of an active rectangular aperture is a $D_{tx}+D_{rx}$ wide triangle in K-space \cite{Hergum2009}.
-Even though the lateral bandwidth is increased, the distance to the first zero (The Rayleigh criteria) remains the same when multiplying two sinc functions.
+Even though the lateral bandwidth is increased on receive, the distance to the first zero (The Rayleigh criteria) remains the same when multiplying two sinc functions.
 Sampling according to (\ref{eq:resolution}) is therefore more than what is strictly needed in practise.
@@
 In Fig.\,\ref{fig:das_shiftvariant} the beam spacing is reduced by a factor of two.
-We see how the scallop loss is now increased to a level where the contours no longer are diagonal.
+We see how the scalloping loss is increased to a level where the contours no longer are diagonal.
 Hence, the imaging system is no longer locally lateral shift-invariant.
@@
 \section{Local Lateral Shift-Invariance of the Capon Beamformer}
-We will now investigate the local lateral shift-invariance of the Capon beamformer with parameters equal to the once used in earlier work \comment{cite}.
+We will now investigate the local lateral shift-invariance of the Capon beamformer with parameters equal to the once used in earlier work \cite{Synnevag2009, Asen}.
 In Fig.\,\ref{fig:capon1}.
@@
 The set of lateral displacements is constructed in such a way that the center beam is the only one hitting the point scatterer exactly.
-We see that the scallop loss is tremendous, and not far from the value predicted by Fig\,\ref{fig:das_capon_beams}.
+We see that the scalloping loss is tremendous, and not far from the value predicted by Fig\,\ref{fig:das_capon_beams}.
 The difference between the predicted and the simulated loss is due to several factors.
 The most important one being that Fig.\,\ref{fig:das_capon_beams} shows a single frequency whereas the RMS used in Fig.\,\ref{fig:capon1} average a wideband signal in range.
-For a slice through the center of the pulse the scallop loss is below -30dB.
+For a slice through the center of the pulse the scalloping loss is below -30dB.
 
@@
 In both extreme cases, $d=\infty$ and $L=1$, the Capon beamformer becomes the DAS beamformer, and will require a beamspacing according to (\ref{eq:resolution}).
-In Fig.\,\ref{fig:capon2} we show how the scallop loss is reduced when the amount of diagonal loading is increased.
+In Fig.\,\ref{fig:capon2} we show how the scalloping loss is reduced when the amount of diagonal loading is increased.
 We observe that Fig.\,\ref{fig:capon2} contains the same pattern as present in Fig.\,\ref{fig:das_shiftvariant}.
@@
 From the previous section it should now be clear that we somehow need to increase the lateral sampling.
-We will investigate how this oversampling should be conducted, how large it needs to be and how the different methods will affect the system frame rate.
+In this section we will investigate how this oversampling should be conducted, how large it needs to be and how the different methods will affect the system frame rate and the processing complexity.
 
@@
 In Fig.\,\ref{fig:capon_oversampling_a} and Fig.\,\ref{fig:capon_oversampling_b} we have increased the lateral sampling on transmit by a factor four and ten respectively.
-As expected the scallop loss decreases with increased lateral sampling.
+As expected the scalloping loss decreases with increased lateral sampling.
 Note that Fig.\,\ref{fig:capon_oversampling_b} is zoomed in order to better see the beam-to-beam variation.
+
+From Fig.\,\ref{fig:capon_oversampling_b} we see that more than ten times oversampling has to be applied to achieve local lateral shift-invariance when imaging a moving point scatterer with no spatially white noise and $1/100$ in diagonal loading.
+Note that a diagonal loading factor of $1/100$ will introduce a deterministic white noise component in the sample covariance matrix $40$ dB lower than the signal present.
+Hence, the covariance has been estimated from data which appeared to have an SNR of 40 dB.
+
 Even though increasing the number of transmit improves the lateral shift-invariance it will also reduce the imaging frame rate with a factor equal to the oversampling factor.
 It is therefore not a feasible option for real time ultrasound imaging.
-
-From Fig.\,\ref{fig:capon_oversampling_b} we see that more than ten times oversampling has to be applied when imaging a moving point scatterer with no spatially white noise and $1/100$ in diagonal loading.
-A diagonal loading factor of $1/100$ will introduce a white noise component in the sample covariance matrix $40$ dB lower than the signal present.
-Hence, the signal-to-noise ratio before the array gain is applied is $40$ dB.
-\comment{do calculations}.
 
@@
 
-\subsection{Oversampling by Parallel Recieve Lines}
+\subsection{Oversampling by Parallel Receive Beamforming}
+Parallel receive beamforming (PRB) means that multiple rx-beams are constructed from one tx-beam.
+By reducing the number of tx-beams and relying on multiple receive beams in order to get sufficient sampling it is possible to achieve high frame rates in e.g.
+cardiac ultrasound imaging.
+
+In this paper we will not discuss the artifacts introduced by PRL and we will
+
 In (\comment{Cite paper where they use STB + Capon if its out}) it is shown that  synthetic transmit beamforming (STB) \cite{Hergum2007} works with Capon beamforming.
@@
 }
-\caption{LSV-plots of in-vivo data. a) Image of the data. b) Normal. c) Oversampling with phase rotation}
+\caption{In-vivo data. a) Image of the data. b) Capon image. c) Capon image with oversampling with phase rotation}
 \label{fig:das}

### 2013-11-16 HRUIofMovingObjects/jrnl_article.tex
Moving objects article: doen some writing

@@
 
+Phase rotation will introduced block artifacts equal to the once observed when using PRB.
+This can be improved on by implementing a phase rotation version of STB where 8PR are converted to 4PR-STB by averaging 4 phase rotated outputs from two beams rotated to the same angle.
+
 \begin{figure}\caption{Plot of contrast vs oversampling factor (phase rotation) in in-vivo data.}
@@
 May also use phase rotation.
+
+Other methods claiming super resolution would also need some degree of oversampling (Eigenspace, LCA etc.).
+
+When oversampling is applied (L=M/2, K=1, d=1/100) becomes too extreme. L=M/4 with K=0 is better.
+
+Oversampling is not only essential to keep the amplitude level of bright point scatterers it will also help to improve contrast and edge definitions.
+If oversampling is not applied the edge delineation of an moving object will fluctuate in the same way as a moving point scatterer.
 

### 2013-11-16 HRUIofMovingObjects/jrnl_article.tex
Moving object article: Done some writing. Added some figures.

@@
 \comment{The abstract goes here.} 
+..............................................................................................................
+..............................................................................................................
+..............................................................................................................
+..............................................................................................................
+..............................................................................................................
+..............................................................................................................
+..............................................................................................................
+..............................................................................................................
+..............................................................................................................
+..............................................................................................................
+..............................................................................................................
+..............................................................................................................
+..............................................................................................................
+..............................................................................................................
+..............................................................................................................
+..............................................................................................................
+..............................................................................................................
+..............................................................................................................
 \end{abstract}
@@
 The difference between the predicted and the simulated loss is due to several factors.
-The most important one being that Fig.\,\ref{fig:das_capon_beams} shows a single frequency whereas the RMS used in Fig.\,\ref{fig:capon1} average a wideband signal in range.
+The most important one being that Fig.\,\ref{fig:das_capon_beams} shows a single frequency whereas the RMS used in Fig.\,\ref{fig:capon1} average a broadband signal in range.
 For a slice through the center of the pulse the scalloping loss is below -30dB.
@@
 We observe that Fig.\,\ref{fig:capon2} contains the same pattern as present in Fig.\,\ref{fig:das_shiftvariant}.
-From Fig.\,\ref{fig:image_capon} it should be obvious that the Capon beamformer either require an increase in the lateral sampling in order to make the imaging system locally lateral shift-invariant, or the resolution has to be reduced to the same level as DAS.
+From Fig.\,\ref{fig:capon} it should be obvious that the Capon beamformer either require increased lateral sampling in order to make the imaging system locally lateral shift-invariant, or the resolution has to be reduced to the same level as DAS.
 It would be unwise to do all the computation involved with Capon beamforming with a high $d$ value and end up with a resolution equal to DAS.
 However, we see that the diagonal factor can be a useful tool if resolution has to be traded for the need of oversampling.
+The same is true for the subarray length parameter.
 
@@
 \caption{...}
-\label{fig:image_capon}
+\label{fig:capon}
 \end{figure*}
@@
 We will now discuss different ways to improve on the lateral shift-invariance when using the Capon beamformer.
-From the previous section it should now be clear that we somehow need to increase the lateral sampling.
-In this section we will investigate how this oversampling should be conducted, how large it needs to be and how the different methods will affect the system frame rate and the processing complexity.
+From the previous section it should now be clear that we somehow need to increase the lateral sampling to make the Capon imaging system lateral shift invariant.
+In this section we will investigate how this oversampling should be conducted, how large it needs to be and how the proposed methods will affect the system frame rate and the processing complexity.
 
 \subsection{Oversampling on Transmit}
-The typical way of obtaining sufficient sampling in an active broadband system is by increasing the number of transmit beams until (\ref{eq:resolution}) is fulfilled.
+The straight forward way of obtaining sufficient sampling in an active broadband system is by increasing the number of transmit beams (tx-beams) until (\ref{eq:resolution}) is fulfilled.
 In Fig.\,\ref{fig:capon_oversampling_a} and Fig.\,\ref{fig:capon_oversampling_b} we have increased the lateral sampling on transmit by a factor four and ten respectively.
 As expected the scalloping loss decreases with increased lateral sampling.
-Note that Fig.\,\ref{fig:capon_oversampling_b} is zoomed in order to better see the beam-to-beam variation.
-
-From Fig.\,\ref{fig:capon_oversampling_b} we see that more than ten times oversampling has to be applied to achieve local lateral shift-invariance when imaging a moving point scatterer with no spatially white noise and $1/100$ in diagonal loading.
+Note that Fig.\,\ref{fig:capon_oversampling_b} is zoomed compared to Fig.\,\ref{fig:capon} in order to better see the beam-to-beam variation.
+
+From Fig.\,\ref{fig:capon_oversampling_b} we see that more than ten times oversampling has to be applied to achieve local lateral shift-invariance when a moving point scatterer is imaged without noise and with the same parameters as in Fig\,\ref{fig:capon}.
+%$1/100$ in diagonal loading.
 Note that a diagonal loading factor of $1/100$ will introduce a deterministic white noise component in the sample covariance matrix $40$ dB lower than the signal present.
@@
 
-Even though increasing the number of transmit improves the lateral shift-invariance it will also reduce the imaging frame rate with a factor equal to the oversampling factor.
-It is therefore not a feasible option for real time ultrasound imaging.
+Even though increasing the number of transmits improves the lateral shift-invariance it will also reduce the imaging frame rate with a factor equal to the oversampling factor.
+The processing complexity will also increase equally.
+For these two reasons, oversampling on transmit is not a feasible option for real time ultrasound imaging.
 
@@
 \subsection{Oversampling by Parallel Receive Beamforming}
-Parallel receive beamforming (PRB) means that multiple rx-beams are constructed from one tx-beam.
-By reducing the number of tx-beams and relying on multiple receive beams in order to get sufficient sampling it is possible to achieve high frame rates in e.g.
+Parallel receive beamforming (PRB) means that multiple narrow receive beams (rx-beams) are constructed from one broad tx-beam.
+Then, by reducing the number of tx-beams and relying on multiple rx-beams in order to get sufficient sampling on receive, it is possible to achieve high frame rates in e.g.
 cardiac ultrasound imaging.
-
-In this paper we will not discuss the artifacts introduced by PRL and we will
-
+However, PRB also introduce motion artifacts in the image which to some degree can be compensated for \cite{Hergum2007} \comment{cite Bastien}.
 In (\comment{Cite paper where they use STB + Capon if its out}) it is shown that synthetic transmit beamforming (STB) \cite{Hergum2007} works with Capon beamforming.
-In  \cite{Asen} we also used STB in order to increase the beam density of a cardiac recording.
-STB has been shown to remove the blocky artifacts introduced by parallel receive lines (PRL) per transmit line.
-As long as some degree of motion compensation is applied (\comment{cite bastien's paper on motion compensation }) we can regard PRL as being equal to oversampling on transmit.
-It will also increase the number of covariance matrices, and thereby inversions with a factor equal to the number of parallel lines.
-%\begin{figure}[!t]
-%	\centering
-%	\includegraphics[width=\linewidth]{\img capon_L=32_K=1_d=001_2MLA_and_2x_microsteering_correct_pitch.eps}
-%	\label{fig:image_capon_bs}
-%\end{figure}
+In \cite{Asen} we also used STB in order to increase the beam density of a cardiac ultrasound recording.
+In this paper we will consider PRB as being equal to oversampling on transmit when it comes to image quality.
+
+In cardiac ultrasound imaging, PRB is typically used in order to keep the number of rx-beams while the number of tx-beams are reduced.
+When using PRB for oversampling we would like to keep the number of tx-beams while the number of rx-beams are increased.
+The acquisition frame rate will then not be affected.
+On the other hand PRB will also increase the number of covariance matrices, and thereby the number of matrix-inversions with a factor equal to the number of parallel receive lines.
+The processing times reported in \cite{Asen} will then also increase with a factor equal to the number of parallel receive lines.
+Since almost 10x oversampling is required in order to achieve lateral shift invariant imaging of a moving point scatterer, non of the implementations described in \cite{Asen} would be real time any more.
+
+One possible solution is to estimate the covariance matrix from multiple beams \cite{Jensen2012}, making one covariance matrix per range.
+Tailoring the method of Jensen and Austeng to cardiac ultrasound imaging it makes more sense to construct one covariance matrix per tx-beam using multiple parallel rx-beams.
+This solution would then not increase the number of inversions needed, but the time taken to construct the matrices would increase.
+We will not investigate this method further in this paper.
 
@@
 The steering vector $\vec{a}$ has in previous work been set to $\vec{1}$ since the ultrasound data are pre-delayed.
-The steering vector could be varied over a set of pre-defined steering vectors.
-\comment{Look at the poster for notation}
+However, the steering vector can also, as in narrow band applications, be varied over a set of pre-defined vectors.
+In a narrow band setting the beam can make a sweep across the whole sector by only using phase rotation.
+This is also possible in broadband applications as long as the phase rotation is limited to a fraction of the wave length.
+It is therefore possible to make a sweep between the current beam and its neighbours using only phase rotation.
+
+For a uniform linear array located along the x-axis the steering vector $\vec{a}$ should be calculated as 
+\begin{align}
+\vec{a} = 
+\begin{bmatrix}
+e^{-j\frac{2\pi}{\lambda_c}x_0\sin(\theta)} \\
+e^{-j\frac{2\pi}{\lambda_c}x_1\sin(\theta)} \\
+\vdots \\ 
+e^{-j\frac{2\pi}{\lambda_c}x_{L-1}\sin(\theta)}
+\end{bmatrix}
+\end{align}
+where $x_i$ is the element position and $\theta$ is swiped from $-\Delta/2$ too $\Delta/2$.
+
+\comment{Add phase rotation formula.}
+
+\comment{Discuss different ways of selecting angles}
+
+\comment{Phase rotation version of STB.}
 
@@
 \centerline{
-%\subfloat[]{\includegraphics[width=0.3\linewidth]{\img das_normal2.eps}\label{fig:das_shiftinvariant}}
+\subfloat[]{\includegraphics[width=0.3\linewidth]{\img phantom1_standard_processing.png}\label{fig:phantom_das}}
 \hfill{}
-%\subfloat[]{\includegraphics[width=0.3\linewidth]{\img das_normal2.eps}\label{fig:das_shiftinvariant}}
+\subfloat[]{\includegraphics[width=0.3\linewidth]{\img LSV_capon.png}\label{fig:phantom_capon_LSV}}
 \hfill{}
-%\subfloat[]{\includegraphics[width=0.3\linewidth]{\img das_undersampled2_2x.eps}\label{fig:das_shiftvariant}}
-}
-\caption{LSV-plots of in-vitro data. a) Image of the data. b) Normal. c) Oversampling with phase rotation}
+\subfloat[]{\includegraphics[width=0.3\linewidth]{\img LSV_capon.png}\label{fig:phantom_capon_LSV_8PR}}
+}
+\caption{\comment{Not final figure. Need new data}. LSV-plots of in-vitro data. a) Image of the data. b) Normal capon. c) Capon with oversampling using 8PR.}
 \label{fig:das}
@@
 \centerline{
-%\subfloat[]{\includegraphics[width=0.3\linewidth]{\img das_normal2.eps}\label{fig:das_shiftinvariant}}
+\subfloat[]{\includegraphics[width=0.5\linewidth]{\img phantom1_capon_processing_K=0_L=24_full.png}\label{fig:in-vivo_capon}}
 \hfill{}
-%\subfloat[]{\includegraphics[width=0.3\linewidth]{\img das_normal2.eps}\label{fig:das_shiftinvariant}}
-\hfill{}
+\subfloat[]{\includegraphics[width=0.5\linewidth]{\img phantom1_capon_processing_K=0_L=24_8pr_full.png}\label{fig:in-vivo_capon_8pr}}
+%\hfill{}
 %\subfloat[]{\includegraphics[width=0.3\linewidth]{\img das_undersampled2_2x.eps}\label{fig:das_shiftvariant}}
 }
-\caption{In-vivo data. a) Image of the data. b) Capon image. c) Capon image with oversampling with phase rotation}
+\caption{\comment{For the moment phantom data. Need to acquire some high quality in-vivo data.}. In-vivo data. a) Capon image. b) Capon image with oversampling using 8PR.}
 \label{fig:das}
@@
 
-\begin{IEEEbiography}{Jon Petter \AA{}sen}
-Biography text here.
+\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{photos/jon_petter2.eps}}]{Jon Petter \AA{}sen}
+(S12) was born in Porsgrunn, Norway in 1986.
+He received the B.Sc.
+and M.Sc.
+degree in computer science from the University of Oslo, Norway, in 2008 and 2010.
+He is currently pursuing his Ph.D.
+degree in medical ultrasound technology at the Norwegian University of Science and Technology (NTNU) Medical Imaging Lab (MI-Lab), Trondheim, Norway.
+
+He has industry experience from GE Vingmed Ultrasound, Horten, Norway, were he completed three summer internships from 2008 to 2010.
+His research interests include image and signal processing, adaptive ultrasound processing techniques and acceleration of ultrasound algorithms using Graphics Processing Units (GPUs).
 \end{IEEEbiography}
 
-%\begin{IEEEbiography}{Carl-Inge Colombo Nilsen}
-%Biography text here.
+%\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{photos/jo_inge2.eps}}]{Jo Inge Buskenes}
+%received the B.Sc.
+degree in electrical engineering from Gj\o{}vik College University, Norway, in 2007, and the M.Sc.
+degree in instrumentation for particle physics from the University of Oslo, Norway, in 2010.
+He is currently pursuing the Ph.D.
+degree in image reconstruction and technology at the University of Oslo.
+
+%His industry experience includes the European Organization for Nuclear Research (CERN), Geneva, Switzerland (2007-2008), and the Norwegian Defence Research Establishment, Kjeller, Norway (2009).
+He has lectured in digital signal processing at the Gj\o{}vik College University (2009), and at the University of Oslo (2010-2012).
+His research interests include adaptive beamforming, digital image reconstruction, high performance computing, intelligent detector design and embedded Linux platforms.
+%\end{IEEEbiography}
+
+%\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{photos/carl-inge.eps}}]{Carl-Inge Colombo Nilsen}
+%(S’06–M’10) received the M.Sc.
+and Ph.D.
+degrees in computer science from the University of Oslo, Norway, in 2005 and 2010.
+He is currently working at the University of Oslo as a postdoctoral research fellow.
+His research interests include signal and array processing for ultrasound imaging and other acoustic applications.
 %\end{IEEEbiography}
@@
 
-\begin{IEEEbiography}{Andreas Austeng}
-Biography text here.
+\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{photos/andreas.eps}}]{Andreas Austeng}
+was born in Oslo, Norway, in 1970.
+He received the M.S.
+degree in physics in 1996 and the Ph.D.
+degree in computer science in 2001, both from the University of Oslo.
+Since 2001, he has been working at the Department of Informatics, University of Oslo, first as a postdoctoral research fellow and currently as an associate professor.
+His research interests include signal and array processing for acoustical imaging.
 \end{IEEEbiography}
 
-\begin{IEEEbiography}{Sverre Holm}
-Biography text here.
+%\newpage
+
+\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{photos/sverre.eps}}]{Sverre Holm}
+(M’82-SM’02) was born in Oslo, Norway, in 1954.
+He received the M.S.
+and Ph.D.
+degrees in electrical engineering from the Norwegian Institute of Technology, Trondheim, in 1978 and 1982, respectively.
+His academic experience includes Yarmouk University in Jordan (1984-1986) and the Norwegian Institute of Technology and later the University of Oslo (1989-1994) as an adjunct professor.
+Since 1995, he has been a professor of signal processing in the Department of Informatics at the University of Oslo.
+In 2002, he was elected a member of the Norwegian Academy of Technological Sciences, and in 2009, he became an adjunct professor in the Norwegian University of Science and Technology (NTNU) Medical Imaging Lab, Trondheim.
+His industry experience includes GE Vingmed Ultrasound, Horten, Norway (1990-1994).
+He worked with Sonitor Technologies (2000-2005), where he developed ultrasonic technology that is the core of the indoor positioning system described in \textit{Scientific Americanin} 2008 as "a positioning system that goes where GPS can't." He has published about 130 publications internationally, plus about 10 patents.
+He has had sabbaticals at GE Global Research, NY (1998), and the Waves and
+Acoustics Laboratory of ESPCI, Paris (2008-2009).
+He was an associate editor of the \textit{IEEE Transactions on Ultrasonics, Ferroelectrics, and Frequency Control} from 1997 to 2002.
+His research interests are medical ultrasound imaging, modeling of tissue-ultrasound interaction, and ultrasound positioning.
 \end{IEEEbiography}

### 2013-11-17 HRUIofMovingObjects/jrnl_article.tex
Moving object article: done some writing

@@
 \author{
-   J. P. \AA{}sen, A. Austeng and S. Holm%\IEEEmembership{Member,~IEEE,}% <-this % stops a space
-
-   \thanks{...}
+   Jon Petter \AA{}sen, \IEEEmembership{Student Member,~IEEE,}, Andreas Austeng, \IEEEmembership{Member,~IEEE,} and Sverre Holm, \IEEEmembership{Senior Member,~IEEE,}% <-this % stops a space
+
+\thanks{Manuscript received ...; accepted ...}
+\thanks{J. P. \AA{}sen and S. Holm is with the Medical Imaging Lab (MI-Lab) at the Norwegian University of Science and Technology, Trondheim, Norway.}
+\thanks{A. Austeng and S. Holm is with the Department of Informatics, University of Oslo, Oslo, Norway. (e-mail: jon.p.asen@ntnu.no)} 
+%J. I. Buskenes, C.-I. C. Nilsen, 
 }
@@
 %\boldmath
-\comment{The abstract goes here.} 
-..............................................................................................................
-..............................................................................................................
-..............................................................................................................
-..............................................................................................................
-..............................................................................................................
-..............................................................................................................
-..............................................................................................................
-..............................................................................................................
-..............................................................................................................
-..............................................................................................................
-..............................................................................................................
-..............................................................................................................
-..............................................................................................................
-..............................................................................................................
-..............................................................................................................
-..............................................................................................................
-..............................................................................................................
-..............................................................................................................
+If an ultrasound imaging system provides a presentation of a moving object witch is not invariant to local spatial shifts the system is said to be local shift-variant.
+This can for instance happen if the axial and lateral sampling is insufficient.
+The Capon beamformer has been shown to provide increased lateral resolution in ultrasound images.
+Increased lateral resolution should demand denser lateral sampling.
+In previous literature on Capon beamforming for medical ultrasound imaging however only single frame scenarios have been simulated, and temporal behaviour and effects caused by the increased resolution and lack of oversampling has therefore been neglected.
+
+In this paper we investigate the local lateral shift-invariance of the Capon beamformer when imaging moving objects.
+We then discuss and investigate different methods in order to improve it.
+We show that insufficient lateral sampling makes an imaging system based on the Capon beamformer locally lateral shift-variant.
+We then show that the locally lateral shift-invariance can be improved by simple phase rotation without affecting the acquisition frame rate and with minor changes in processing complexity.
 \end{abstract}
@@
 It is therefore an interesting question how a moving object appears after Capon beamforming has been applied.
-Earlier work on Capon beamforming for medical ultrasound imaging has investigated the method's capabilities in single frames only, and temporal behavior and effects has therefore been neglected.
+Earlier work on Capon beamforming for medical ultrasound imaging has investigated the method's capabilities in single frames only, and temporal behaviour and effects has therefore been neglected.
 
@@
 In this paper we analyze the shift-invariance property of the Capon beamformer by studying the beamformer output when applied on simulated and \textit{in vitro} images of moving objects.
+We also show some examples of how the shift-invariance and resolution is improved in \textit{in-vivo} images.
 
@@
 SVUS has not been applied, even though it is well known that the Capon beamformer exhibits severe signal-nulling if the signal is present during the covariance estimation process and if this estimate is later used with a steering vector that does not match the signal propagation vector exactly.
-This self-nulling actually contribute a lot to the method's super resolution capability in passive-narrowband applications.
+This self-nulling is actually how the method obtains its super resolution capabilities.
 
@@
 The effect of typical subarray averaging and diagonal loading ($L=32$ and $d=1/100$) has also been added to the Capon beamformer output.
-A diagonal loading of $1/100$ corresponds to an signal-to-noise ratio of 40 dB before array gain is applied.
+A diagonal loading of $1/100$ corresponds to an signal-to-noise ratio of 40 dB before array gain is applied \comment{Is this correct?}.
 The figure shows that when the DAS beamformer is used, only a 3.92 dB scalloping loss \cite{Harris1978} should be observed if the source of the plane wave starts moving parallel to the array.
@@
 
-In previous literature on Capon beamforming for medical ultrasound imaging (single frames) this effect has been passed over in silence by either oversampling on transmit, or carefully position point scatterers on transmit-receive beam pairs, or by ignorance.
-In our first work where Capon beamforming was applied on multiple frames containing moving objects \cite{Asen} we commented on the effect, but no solution was presented other than oversampling on transmit in simulations.
-In this paper we investigate the issue further, and we show that when a moving point scatterer is imaged with ultrasound and the Capon beamformer with the same beam spacing sufficient for DAS a large scalloping loss is observed, manifested as oscillating point scatteres.
+In previous literature on Capon beamforming for medical ultrasound imaging (with single frame simulations) this effect has been passed over in silence by either oversampling on transmit, or by carefully position point scatterers exactly on transmit-receive beam pairs, or by ignorance.
+In our first work where Capon beamforming was applied on a loop of images containing moving objects \cite{Asen2012, Asen} we commented on the effect, but no solution was presented other than oversampling on transmit when simulating.
+
+In this paper we investigate the issue further, and we show that when a moving point scatterer is imaged with the Capon beamformer with the same beam spacing sufficient for DAS the predicted scalloping loss is observed, manifested as oscillating point scatteres.
 The question is then, what can we do in order to improve on the situation without affecting the imaging frame rate.
-In this paper we therefore study the scalloping loss of the Capon beamforming and investigate different methods for oversampling and how they affect the imaging frame rate and the processing complexity.
-Before that we give an introduction to Capon beamforming for medical ultrasound imaging.
+In this paper we therefore study the scalloping loss of the Capon beamforming and investigate different methods aiming at reducing this effect, and how the different methods will affect the imaging frame rate and the processing complexity.
+
+In the next section we give an introduction to Capon beamforming for medical ultrasound imaging together with a short introduction to lateral sampling and shift invariance.
+In Section \ref{sec:capon_LLSI} the local lateral shift-invariance of the Capon beamformer is investigated, before the different methods are presented in Section \ref{sec:methods}.
 
@@
 
-\section{Local Lateral Shift-Invariance of the Capon Beamformer}
+\section{Local Lateral Shift-Invariance of the Capon Beamformer}\label{sec:capon_LLSI}
 We will now investigate the local lateral shift-invariance of the Capon beamformer with parameters equal to the once used in earlier work \cite{Synnevag2009, Asen}.
@@
 
-\section{Methods}
+\section{Methods}\label{sec:methods}
 We will now discuss different ways to improve on the lateral shift-invariance when using the Capon beamformer.

### 2013-11-17 HRUIofMovingObjects/jrnl_article.tex
Moving objects article: finished all section until results.

@@
 In Section \ref{sec:capon_LLSI} the local lateral shift-invariance of the Capon beamformer is investigated, before the different methods are presented in Section \ref{sec:methods}.
+Then, in Section \ref{sec:res} and Section \ref{sec:dis} we present and discuss the results of applying the phase rotation method on \textit{in-vitro} data .
+Finally we give our conclusion in Section \ref{sec:con}.
 
@@
 \section{Background}
-\subsection{Beamforming}
-The standard form of time domain array beamforming is the delay-and-sum beamformer (DAS):
+\subsection{Standard Beamforming}
+The standard form of time domain array beamforming is the delay-and-sum beamformer (DAS) where the output $z$ is calculated as:
 \begin{align}\label{eq:das}
@@
 The weight vector is typically selected in order to trade resolution for a lower side lobe level.
+Hence, the weights are typically real and their K-space response is then symmetric.
 
@@
 This is indeed computational demanding, but following the innovation in GPU computing, and by applying a novel method in order to reduce the size of the matrix inversion problem \cite{Nilsen2009}, real time processing is now possible \cite{Asen}.
+From (\ref{eq:capon_weights} it is clear that complex data will result in complex weights.
+The weights can therefore have an asymmetric K-space response which basically means that zeros does not have to be symmetric and that the beam can be micro-steered by phase rotation.
 
@@
 In an ultrasound image with 256 gray levels and 50 dB dynamic range a 1 dB loss corresponds to 5 gray levels.
-This is approximately equal to the visibility threshold (Weber fraction of 2\%) for a grayscale image, and so a loss larger than 1dB could be visible to the observer.
-In our analysis a system would therefore be regarded as lateral shift-invariant if the loss is smaller than this threshold.
-
-In order to visualize the local lateral shift-invariance of a given imaging system, we will use lateral shift-variance plots (LSV-plots), introduced by Hergum \textit{et al.} \cite{Hergum2007}.
+This is approximately equal to the visibility threshold (Weber fraction of 2\%) for a grayscale image, and a loss larger than 1dB could therefore end up being visible to the observer.
+In our analysis a system would, based on this reasoning, be regarded as lateral shift-invariant if the loss is smaller than 1 dB.
+However, depending on light conditions and e.g.
+the selected gray level mapping, a larger loss might be acceptable.
+
+In order to visualize the local lateral shift-invariance of a given imaging system, we will make use of lateral shift-variance plots (LSV-plots), introduced by Hergum \textit{et al.} \cite{Hergum2007}.
 A LSV-plot is constructed for a given imaging system by imaging a laterally moving point scatterer.
@@
 
-To further quantify the shift-invariance we have calculated the mean absolute error (MAE) of the maximum point in each beam profile versus $f(x)=x$, and the MAE of the amplitude variation for the same set of points.
+To further quantify the shift-invariance we have calculated the mean absolute error (MAE) of the maximum point in each beam profile versus $f(x)=x$ (P-MAE), and the MAE of the amplitude variation for the same set of points (A-MAE).
 The first number quantifies the amount of geometric distortion, and the latter quantifies the gain variation or in our case the loss of signal.
@@
 The red dotted vertical lines indicates the position of tx-rx beam pairs.
-The point scatterer response has been simulated using Field II \cite{Jensen1992, Jensen1996a} \comment{add table with simulation settings}.
+The point scatterer response has been simulated using Field II \cite{Jensen1992, Jensen1996a} \comment{add table with simulation settings or just insert in text?}.
 We see that the contours are approximately diagonal and therefore we can conclude that the system is locally lateral shift-invariant.
@@
 If the contours are diagonal the imaging system is said to be lateral local shif-invariant.
-Red vertical lines are plotted where tx-rx beams are located.
+Red vertical lines are plotted where transmit-receive beam pairs are located.
 a) Sampling corresponding to (\ref{eq:resolution}).
-b) Undersampling with a factor two.}
+The system is locally lateral shift-invariant.
+b) Undersampling with a factor two.
+The system is no longer locally lateral shift-invariant.}
 \label{fig:das}
@@
 We see that the scalloping loss is tremendous, and not far from the value predicted by Fig\,\ref{fig:das_capon_beams}.
-The difference between the predicted and the simulated loss is due to several factors.
+The difference between the predicted and the simulated loss might be caused by several factors.
 The most important one being that Fig.\,\ref{fig:das_capon_beams} shows a single frequency whereas the RMS used in Fig.\,\ref{fig:capon1} average a broadband signal in range.
@@
 The diagonal factor $d$ and the effective adaptive aperture size $L$ can be used to control the mainlobe width of the Capon beamformer.
-In both extreme cases, $d=\infty$ and $L=1$, the Capon beamformer becomes the DAS beamformer, and will require a beamspacing according to (\ref{eq:resolution}).
+In both extreme cases, $d\gg1$ and $L=1$, the Capon beamformer becomes the DAS beamformer, and will require a beamspacing according to (\ref{eq:resolution}).
 In Fig.\,\ref{fig:capon2} we show how the scalloping loss is reduced when the amount of diagonal loading is increased.
@@
 }
-\caption{...}
+\caption{LSV-plots of Capon beamforming with $L=M/2=32$ and $K=1$, and with sampling corresponding to (\ref{eq:resolution}) a) The diagonal loading factor $d=1/100$.
+The system is not locally lateral shift-invariant.
+b) The diagonal loading factor $d$ is increased to 1.
+The system is still not locally lateral shift-invariant.}
 \label{fig:capon}
@@
 \section{Methods}\label{sec:methods}
-We will now discuss different ways to improve on the lateral shift-invariance when using the Capon beamformer.
-From the previous section it should now be clear that we somehow need to increase the lateral sampling to make the Capon imaging system lateral shift invariant.
+We will now discuss different ways to improve on the lateral shift-invariance when the Capon beamformer is used.
+From the previous section it should be clear that we somehow need to increase the lateral sampling to make the Capon imaging system locally lateral shift-invariant.
 In this section we will investigate how this oversampling should be conducted, how large it needs to be and how the proposed methods will affect the system frame rate and the processing complexity.
@@
 \subsection{Oversampling on Transmit}
-The straight forward way of obtaining sufficient sampling in an active broadband system is by increasing the number of transmit beams (tx-beams) until (\ref{eq:resolution}) is fulfilled.
-In Fig.\,\ref{fig:capon_oversampling_a} and Fig.\,\ref{fig:capon_oversampling_b} we have increased the lateral sampling on transmit by a factor four and ten respectively.
+The straight forward way of obtaining sufficient sampling in an active system is by increasing the number of transmit beams (tx-beams) until (\ref{eq:resolution}) is fulfilled or the LSV-plot has a uniform level along the diagonals.
+In Fig.\,\ref{fig:capon_oversampling_a} and Fig.\,\ref{fig:capon_oversampling_b} we have increased the lateral sampling on transmit by a factor of four and ten respectively.
 As expected the scalloping loss decreases with increased lateral sampling.
@@
 }
-\caption{Oversampling on transmit ...}
+\caption{Capon beamforming as described in Fig\,\ref{fig:capon} with oversampling on transmit.
+The figure is zoomed compared with Fig\,\ref{fig:capon} to better see the beam-to-beam variation.
+a) 4x oversampling.
+b) 10x oversampling.}
 \label{fig:capon_oversampling}
@@
 In \cite{Asen} we also used STB in order to increase the beam density of a cardiac ultrasound recording.
-In this paper we will consider PRB as being equal to oversampling on transmit when it comes to image quality.
+In this discussion we consider PRB as being equal to oversampling on transmit when it comes to image quality.
 
@@
 The acquisition frame rate will then not be affected.
-On the other hand PRB will also increase the number of covariance matrices, and thereby the number of matrix-inversions with a factor equal to the number of parallel receive lines.
+On the other hand more rx-beams will increase the number of covariance matrices, and thereby the number of matrix-inversions with a factor equal to the number of parallel receive lines.
 The processing times reported in \cite{Asen} will then also increase with a factor equal to the number of parallel receive lines.
@@
 
-One possible solution is to estimate the covariance matrix from multiple beams \cite{Jensen2012}, making one covariance matrix per range.
+One possible solution is to estimate the covariance matrix from multiple beams \cite{Jensen2012}, generating one covariance matrix per range.
 Tailoring the method of Jensen and Austeng to cardiac ultrasound imaging it makes more sense to construct one covariance matrix per tx-beam using multiple parallel rx-beams.
-This solution would then not increase the number of inversions needed, but the time taken to construct the matrices would increase.
+This solution will not increase the number of matrix inversions, but the time taken to construct the matrices will increase.
 We will not investigate this method further in this paper.
+\comment{Kanskje litt dumt å nevne det da...?}
 
@@
 It is therefore possible to make a sweep between the current beam and its neighbours using only phase rotation.
+Phase rotation is what the Capon beamformer use internally to micro-steer away from bright point scatterers and to enhance cyst edges.
 
@@
 
-\comment{Add phase rotation formula.}
-
-\comment{Discuss different ways of selecting angles}
-
-\comment{Phase rotation version of STB.}
+The steering angles can be spread out in many different ways.
+If the number of steering angles is $N_\theta$ we have selected a uniform distribution of $\theta$  with $N_\theta/2$ angles on each side of the original steering direction if $N_\theta$ is even, and $(N_\theta-1)/2$ if $N_\theta$ is odd.
+
+Oversampling by phase rotation introduce similar block artifacts as PRB, but on a smaller scale.
+These artifacts could be removed if we distributes our steering angles as beams are spread out for the STB technique.
+We then get a phase rotation version of STB where e.g.
+eight phase rotated beams are averaged into four.
+\comment{Move to discussion?}
 
@@
 	}
-	\caption{Oversampling using phase rotation ...}
+	\caption{Capon beamforming as described in Fig\,\ref{fig:capon} with oversampling based on phase rotation.
+The figure is zoomed compared with Fig\,\ref{fig:capon} to better see the beam-to-beam variation.
+a) 4x oversampling.
+b) 10x oversampling.}
 	\label{fig:benchmark_capon_bs}
@@
 \subsection{In-Vivo Images}
-Investigate if phase rotation improves contrast in in-vivo images of the heart.
+Investigate if phase rotation improves contrast in in-vivo images of the heart or liver.
 
@@
 
-Other methods claiming super resolution would also need some degree of oversampling (Eigenspace, LCA etc.).
+Other methods claiming super resolution would also need oversampling (Eigenspace, LCA etc.) to some extent.
 
@@
 
-Oversampling is not only essential to keep the amplitude level of bright point scatterers it will also help to improve contrast and edge definitions.
+Oversampling is not only essential to keep the amplitude level of bright point scatterers it will also helps to improve contrast and edge definitions.
 If oversampling is not applied the edge delineation of an moving object will fluctuate in the same way as a moving point scatterer.
-
-TODO: Add citations – Hergum/Bjåstad Local shift invariance and MLA’s… Is there any publications stating that Capon do not work on moving objects?
 

### 2013-11-21 HRUIofMovingObjects/jrnl_article.tex
Moving object article: Done some writing. Added new figures

@@
 \author{
-   Jon Petter \AA{}sen, \IEEEmembership{Student Member,~IEEE,}, Andreas Austeng, \IEEEmembership{Member,~IEEE,} and Sverre Holm, \IEEEmembership{Senior Member,~IEEE,}% <-this % stops a space
+   Jon Petter \AA{}sen, \IEEEmembership{Student Member,~IEEE,}, Andreas Austeng, \IEEEmembership{Member,~IEEE,} and Sverre Holm, \IEEEmembership{Senior Member,~IEEE}% <-this % stops a space
 
@@
 %\boldmath
-If an ultrasound imaging system provides a presentation of a moving object witch is not invariant to local spatial shifts the system is said to be local shift-variant.
+If an ultrasound imaging system provides a presentation of a moving object which is sensitive to local spatial shifts, the system is said to be local shift-invariant.
 This can for instance happen if the axial and lateral sampling is insufficient.
@@
 Increased lateral resolution should demand denser lateral sampling.
-In previous literature on Capon beamforming for medical ultrasound imaging however only single frame scenarios have been simulated, and temporal behaviour and effects caused by the increased resolution and lack of oversampling has therefore been neglected.
+In previous literature on Capon beamforming for medical ultrasound imaging however only single frame scenarios have been simulated, and temporal behaviour and effects caused by the increased resolution and lack of oversampling have therefore been neglected.
 
@@
 
-\IEEEPARstart{W}{ith} the introduction of a fast implementation of the Capon adaptive beamformer, a stream of images can now be processed at interactive frame rates \cite{Asen}.
+\IEEEPARstart{I}{n} the last decade several authors have applied the Capon beamformer for ultrasound imaging \cite{Synnevag2007, Vignon2008, Viola}.
+The adaptive nature of the method has been shown to greatly improve the lateral resolution in ultrasound images.
+In recent work we introduced a fast implementation of the Capon beamformer, capable of processing a stream of images at interactive frame rates \cite{Asen}.
 It is therefore an interesting question how a moving object appears after Capon beamforming has been applied.
@@
 An important property of an ultrasound imaging system is its local shift-invariance.
-Local shift-invariance means that an object’s appearance should be preserved when the object, or the ultrasound probe, is moved slightly.
-This invariance for local movements is what makes ultrasound imaging feasible in general, however it is specially important for quantitative techniques that correlate time-varying features, like pulsed doppler and flow imaging, and speckle tracking of tissue and blood.
-In this paper we analyze the shift-invariance property of the Capon beamformer by studying the beamformer output when applied on simulated and \textit{in vitro} images of moving objects.
-We also show some examples of how the shift-invariance and resolution is improved in \textit{in-vivo} images.
-
-Capon beamforming has been shown to provide increased lateral resolution and contrast in ultrasound images \cite{Synnevag2007, Synnevag2009, Chen2011}.
-The method accomplish this by adjusting the array weights based on element-to-element covariance.
+Local shift-invariance means that an object's appearance should be preserved when the object, or the ultrasound probe, is moved slightly.
+This invariance for local movements is what makes ultrasound imaging feasible in general, but it is specially important for quantitative techniques that correlate time-varying features, like pulsed doppler and flow imaging, and speckle tracking of tissue and blood.
+In this paper we analyze the lateral shift-invariance property of the Capon beamformer by studying the beamformer output when applied on simulated and \textit{in vitro} images of moving objects.
+We also show some \textit{in-vivo} examples.
+
+As mentioned, Capon beamforming has been shown to provide improved lateral resolution and edge definitions in ultrasound images \cite{Synnevag2007, Synnevag2009, Chen2011}.
+The method accomplishes this by adjusting the array weights based on element-to-element covariance.
 Hence, the method yields data dependent weights.
@@
 The first two trade resolution for robustness and temporal smoothing helps to preserve speckle statistics \cite{Synnevag2007a}.
-SVUS has not been applied, even though it is well known that the Capon beamformer exhibits severe signal-nulling if the signal is present during the covariance estimation process and if this estimate is later used with a steering vector that does not match the signal propagation vector exactly.
-This self-nulling is actually how the method obtains its super resolution capabilities.
-
-In \cite{Cox1973}, Cox derives formulas for the resolution of the Capon beamformer (called the $\vec{k}_3$ processor), and presents in great detail how the amount of mismatch between the selected steering vector and the signal propagation vector impacts the beamformer output in simple situations.
+SVUS reduces the signal-nulling effect of the Capon beamformer which appears if the signal is present during the covariance estimation process and if this estimate is later used with a steering vector that does not match the signal propagation vector exactly.
+This self-nulling is actually how the method obtains its super resolution capabilities, but the signal can be lost in the process.
+In this paper we make use of a steering vector set to reduce this self-nulling effect.
+
+In \cite{Cox1973}, Cox derives formulas for the resolution of the Capon beamformer, and presents in great detail how the amount of mismatch between the selected steering vector and the signal propagation vector impacts the beamformer output in simple situations.
 Cox states the following important fact about the Capon beamformer (the $\vec{k}_3$ processor):
@@
 Impinging on the array is a monochromatic plane wave with a propagation vector orthogonal to the array surface with no spatially white noise present.
-Hence, the interference-noise matrix ($\sigma_0^2\mat{Q}$) is the identity.
+Hence, the interference-noise matrix ($\sigma_0^2\mat{Q}$ in \cite{Cox1973}) is the identity.
 The effect of typical subarray averaging and diagonal loading ($L=32$ and $d=1/100$) has also been added to the Capon beamformer output.
-A diagonal loading of $1/100$ corresponds to an signal-to-noise ratio of 40 dB before array gain is applied \comment{Is this correct?}.
-The figure shows that when the DAS beamformer is used, only a 3.92 dB scalloping loss \cite{Harris1978} should be observed if the source of the plane wave starts moving parallel to the array.
-However if the Capon beamformer is used with the beam density sufficient for DAS the scalloping loss could be as large as $31$ dB.
+A diagonal loading of $1/100$ corresponds to an signal-to-noise ratio of 20 dB before array gain is applied.
+The figure shows that when the DAS beamformer is used with a beamspacing equal to the system resolution, only a 3.92 dB scalloping loss \cite{Harris1978} should be observed if the source of the plane wave starts moving parallel to the array.
+However if the Capon beamformer is used with the same beam density the scalloping loss could be as large as $31$ dB.
 
@@
 }
-\caption{Comparison of drop in output power in between beams (half the system resolution) for delay-and-sum (DAS) and Capon beamforming ($M=64$, $L=32$, $d=0.01$).
-A diagonal loading of $1/100$ corresponds to an effective signal-to-noise ratio after array gain of 40 dB.}
+\caption{Comparison of drop in output power in between receive beams.
+Half the system resolution is plotted to show where two beams with a spacing equal to the system resolution would intersect.
+The steered response of both delay-and-sum (DAS) and Capon beamforming ($M=64$, $L=32$, $d=0.01$) is plotted.
+A monochromatic plane wave is located at zero degrees.}
 \label{fig:das_capon_beams}
@@
 
-In previous literature on Capon beamforming for medical ultrasound imaging (with single frame simulations) this effect has been passed over in silence by either oversampling on transmit, or by carefully position point scatterers exactly on transmit-receive beam pairs, or by ignorance.
+In previous literature on Capon beamforming \comment{cite earlier work} for medical ultrasound imaging (with single frame simulations) this effect has been passed over in silence by either oversampling on transmit, or by carefully position point scatterers exactly on transmit-receive beam pairs, or by ignorance.
 In our first work where Capon beamforming was applied on a loop of images containing moving objects \cite{Asen2012, Asen} we commented on the effect, but no solution was presented other than oversampling on transmit when simulating.
@@
 In Section \ref{sec:capon_LLSI} the local lateral shift-invariance of the Capon beamformer is investigated, before the different methods are presented in Section \ref{sec:methods}.
-Then, in Section \ref{sec:res} and Section \ref{sec:dis} we present and discuss the results of applying the phase rotation method on \textit{in-vitro} data .
+Then, in Section \ref{sec:res} and Section \ref{sec:dis} we present and discuss the results of applying the most promising method on \textit{in-vitro} and \textit{in-vivo} data .
 Finally we give our conclusion in Section \ref{sec:con}.
@@
 \subsection{Capon Beamforming}
-The Capon beamformer produce a weight vector, like the one in (\ref{eq:das}), based on the impinging signal and an optimization criteria. The minimization problem is as follow:
+The Capon beamformer produce a weight vector, like the one in (\ref{eq:das}), based on the impinging signal and an optimization criteria. The minimization problem is as follow \cite{Capon1969}:
 \begin{align}
@@
 This is indeed computational demanding, but following the innovation in GPU computing, and by applying a novel method in order to reduce the size of the matrix inversion problem \cite{Nilsen2009}, real time processing is now possible \cite{Asen}.
-From (\ref{eq:capon_weights} it is clear that complex data will result in complex weights.
-The weights can therefore have an asymmetric K-space response which basically means that zeros does not have to be symmetric and that the beam can be micro-steered by phase rotation.
+From (\ref{eq:capon_weights}) it is clear that complex data will result in complex weights.
+The weights can therefore have an asymmetric K-space response which basically means that zeros does not have to be symmetric around the origin and that the beam can be micro-steered by phase rotation.
 
@@
 This is approximately equal to the visibility threshold (Weber fraction of 2\%) for a grayscale image, and a loss larger than 1dB could therefore end up being visible to the observer.
-In our analysis a system would, based on this reasoning, be regarded as lateral shift-invariant if the loss is smaller than 1 dB.
+In our analysis a system would, based on this reasoning, be regarded as locally lateral shift-invariant if the beam-too-beam loss is smaller than 1 dB.
 However, depending on light conditions and e.g.
@@
 Fig.\,\ref{fig:das_shiftinvariant} presents a LSV-plot for an imaging system using DAS beamforming with a beam spacing equal to (\ref{eq:resolution}).
-The red dotted vertical lines indicates the position of tx-rx beam pairs.
+The red dotted vertical lines indicates the position of transmit-receive beam pairs.
 The point scatterer response has been simulated using Field II \cite{Jensen1992, Jensen1996a} \comment{add table with simulation settings or just insert in text?}.
@@
 \subsection{Oversampling on Transmit}
-The straight forward way of obtaining sufficient sampling in an active system is by increasing the number of transmit beams (tx-beams) until (\ref{eq:resolution}) is fulfilled or the LSV-plot has a uniform level along the diagonals.
+The straight forward way of obtaining sufficient sampling in an active system is by increasing the number of transmit beams (tx-beams) until (\ref{eq:resolution}) is fulfilled or the LSV-plot has an uniform level along the diagonals.
 In Fig.\,\ref{fig:capon_oversampling_a} and Fig.\,\ref{fig:capon_oversampling_b} we have increased the lateral sampling on transmit by a factor of four and ten respectively.
@@
 Hence, the covariance has been estimated from data which appeared to have an SNR of 40 dB.
+\comment{correct?}
 
@@
 The processing complexity will also increase equally.
-For these two reasons, oversampling on transmit is not a feasible option for real time ultrasound imaging.
+For these two reasons, oversampling on transmit is not a feasible option for real time ultrasound imaging, and we will not discuss this method further.
 
@@
 In \cite{Asen} we also used STB in order to increase the beam density of a cardiac ultrasound recording.
-In this discussion we consider PRB as being equal to oversampling on transmit when it comes to image quality.
+In the following discussion we consider PRB as being equal to oversampling on transmit when it comes to image quality.
 
@@
 On the other hand more rx-beams will increase the number of covariance matrices, and thereby the number of matrix-inversions with a factor equal to the number of parallel receive lines.
-The processing times reported in \cite{Asen} will then also increase with a factor equal to the number of parallel receive lines.
+The processing times reported in \cite{Asen} will then also increase with the same factor.
 Since almost 10x oversampling is required in order to achieve lateral shift invariant imaging of a moving point scatterer, non of the implementations described in \cite{Asen} would be real time any more.
 
-One possible solution is to estimate the covariance matrix from multiple beams \cite{Jensen2012}, generating one covariance matrix per range.
-Tailoring the method of Jensen and Austeng to cardiac ultrasound imaging it makes more sense to construct one covariance matrix per tx-beam using multiple parallel rx-beams.
-This solution will not increase the number of matrix inversions, but the time taken to construct the matrices will increase.
+One way of reducing the number of matrix inversions is to estimate the covariance matrix from multiple beams \cite{Jensen2012}, generating one covariance matrix per range.
+Tailoring the method of Jensen and Austeng to cardiac ultrasound imaging it makes more sense, due to motion, to construct one covariance matrix per tx-beam using multiple parallel rx-beams.
+With this solution the number of matrix inversions will not changes, but the time taken to construct the matrices will increase.
 We will not investigate this method further in this paper.
-\comment{Kanskje litt dumt å nevne det da...?}
+\comment{Kanskje litt dumt å nevne det da...?} \comment{move to discussion}
 
@@
 In a narrow band setting the beam can make a sweep across the whole sector by only using phase rotation.
-This is also possible in broadband applications as long as the phase rotation is limited to a fraction of the wave length.
-It is therefore possible to make a sweep between the current beam and its neighbours using only phase rotation.
-Phase rotation is what the Capon beamformer use internally to micro-steer away from bright point scatterers and to enhance cyst edges.
-
-For a uniform linear array located along the x-axis the steering vector $\vec{a}$ should be calculated as 
+This is also possible in broadband applications as long as the phase rotation is limited to a fraction of the wave length \comment{cite fine-coarse beamformer}.
+It is therefore possible to make a sweep between the current beam and its nearest neighbours using only phase rotation.
+Phase rotation is actually what the Capon beamformer use internally to micro-steer away from bright point scatterers and to enhance edges.
+
+For an uniform linear array located along the x-axis the steering vector $\vec{a}$ should be calculated as 
 \begin{align}
@@
 \end{align}
-where $x_i$ is the element position and $\theta$ is swiped from $-\Delta/2$ too $\Delta/2$.
+where $x_i$ is the element position and $\theta$ is swiped from $-\Delta/2$ too $\Delta/2$ where $\Delta$ is as defined in (\ref{eq:resolution}).
+Note that the steering vector is calculated for a $L$ element array centered around the origin, even though the full array has $M$ elements.
+It is important that this synthetic array has the same element pitch as the full array.
 
@@
 \centerline{
-\subfloat[]{\includegraphics[width=0.3\linewidth]{\img phantom1_standard_processing.png}\label{fig:phantom_das}}
+\begin{tabular}{cc}
+\subfloat[]{\includegraphics[width=0.50\linewidth]{\img LSV_image.eps}\label{fig:phantom_das}}
 \hfill{}
-\subfloat[]{\includegraphics[width=0.3\linewidth]{\img LSV_capon.png}\label{fig:phantom_capon_LSV}}
+\subfloat[]{\includegraphics[width=0.50\linewidth]{\img LSV_das.eps}\label{fig:phantom_capon_LSV}}\\
 \hfill{}
-\subfloat[]{\includegraphics[width=0.3\linewidth]{\img LSV_capon.png}\label{fig:phantom_capon_LSV_8PR}}
-}
-\caption{\comment{Not final figure. Need new data}. LSV-plots of in-vitro data. a) Image of the data. b) Normal capon. c) Capon with oversampling using 8PR.}
+\subfloat[]{\includegraphics[width=0.50\linewidth]{\img LSV_capon.eps}\label{fig:phantom_capon_LSV}}
+\hfill{}
+\subfloat[]{\includegraphics[width=0.50\linewidth]{\img LSV_capon_10pr.eps}\label{fig:phantom_capon_LSV_8PR}}
+\end{tabular}
+}
+\caption{LSV-plots generated from \textit{in-vitro} phantom data.
+The probe was manually dragged across the phantom.
+a) First frame.
+b) DAS c) Normal capon.
+d) Capon with oversampling using 10 times phase rotation.}
 \label{fig:das}
@@
 
-The authors would like to thank...
+The authors would like to thank Anders Sornes for acquiring data.
 

### 2013-11-22 HRUIofMovingObjects/jrnl_article.tex
Moving object article: done some writing. Added some figures

@@
 In this paper we analyze the lateral shift-invariance property of the Capon beamformer by studying the beamformer output when applied on simulated and \textit{in vitro} images of moving objects.
-We also show some \textit{in-vivo} examples.
 
@@
 SVUS reduces the signal-nulling effect of the Capon beamformer which appears if the signal is present during the covariance estimation process and if this estimate is later used with a steering vector that does not match the signal propagation vector exactly.
-This self-nulling is actually how the method obtains its super resolution capabilities, but the signal can be lost in the process.
+This self-nulling is actually how the method obtains its super resolution capabilities, but the signal can also get lost in the process.
 In this paper we make use of a steering vector set to reduce this self-nulling effect.
@@
 In Fig.\,\ref{fig:das_capon_beams} we have plotted the steered response for DAS and the Capon beamformer using (26) and (32) in \cite{Cox1973}.
-A 64-element linear array with $\lambda/2$ element spacing, where $\lambda$ is the wavelength, is scanned over a sector equal to three times the system resolution.
+A 64-element linear array with $\lambda/2$ element spacing, where $\lambda$ is the wavelength, is scanned over a sector equal to three times the system resolution ($1/32$ rad).
 Impinging on the array is a monochromatic plane wave with a propagation vector orthogonal to the array surface with no spatially white noise present.
@@
 
-In previous literature on Capon beamforming \comment{cite earlier work} for medical ultrasound imaging (with single frame simulations) this effect has been passed over in silence by either oversampling on transmit, or by carefully position point scatterers exactly on transmit-receive beam pairs, or by ignorance.
+In previous literature on Capon beamforming for medical ultrasound imaging (with single frame simulations) this effect has been passed over in silence by either oversampling on transmit, or by carefully position point scatterers exactly on transmit-receive beam pairs, or by ignorance.
 In our first work where Capon beamforming was applied on a loop of images containing moving objects \cite{Asen2012, Asen} we commented on the effect, but no solution was presented other than oversampling on transmit when simulating.
-
-In this paper we investigate the issue further, and we show that when a moving point scatterer is imaged with the Capon beamformer with the same beam spacing sufficient for DAS the predicted scalloping loss is observed, manifested as oscillating point scatteres.
+The same was done by Jensen and Austeng \cite{Jensen2012}.
+
+In this paper we investigate the issue further, and we show that when a moving point scatterer is imaged with the Capon beamformer with the same beam spacing sufficient for DAS the predicted scalloping loss is observed, manifested as blinking point scatteres.
 The question is then, what can we do in order to improve on the situation without affecting the imaging frame rate.
@@
 In Section \ref{sec:capon_LLSI} the local lateral shift-invariance of the Capon beamformer is investigated, before the different methods are presented in Section \ref{sec:methods}.
-Then, in Section \ref{sec:res} and Section \ref{sec:dis} we present and discuss the results of applying the most promising method on \textit{in-vitro} and \textit{in-vivo} data .
+Then, in Section \ref{sec:res} and Section \ref{sec:dis} we present and discuss the results of applying the most promising method on \textit{in-vitro} and \textit{in-vivo} data.
 Finally we give our conclusion in Section \ref{sec:con}.
@@
 
-The sample covariance matrix can be estimated for an active system in the following way \cite{Synnevag2009}:
+The sample covariance matrix can be estimated for an active broadband system in the following way \cite{Synnevag2009}:
 \begin{align}
@@
 This is indeed computational demanding, but following the innovation in GPU computing, and by applying a novel method in order to reduce the size of the matrix inversion problem \cite{Nilsen2009}, real time processing is now possible \cite{Asen}.
-From (\ref{eq:capon_weights}) it is clear that complex data will result in complex weights.
-The weights can therefore have an asymmetric K-space response which basically means that zeros does not have to be symmetric around the origin and that the beam can be micro-steered by phase rotation.
+From (\ref{eq:capon_weights}) it is clear that complex data can lead to complex weights.
+The weights can therefore have an asymmetric K-space response which basically means that the main lobe and zeros does not have to be symmetric around the origin.
+The beam can therefore be micro-steered by phase rotation.
 
@@
 \end{align}
-If this requirement is not fulfilled the scalloping loss could become visible in the image and the system will no longer be lateral shift-invariant.
-In an ultrasound image with 256 gray levels and 50 dB dynamic range a 1 dB loss corresponds to 5 gray levels.
+If this requirement is not fulfilled the scalloping loss could become visible in the image and the system will no longer be locally lateral shift-invariant.
+In an ultrasound image with 256 gray levels and 50 dB dynamic range, a 1 dB loss corresponds to 5 gray levels.
 This is approximately equal to the visibility threshold (Weber fraction of 2\%) for a grayscale image, and a loss larger than 1dB could therefore end up being visible to the observer.
-In our analysis a system would, based on this reasoning, be regarded as locally lateral shift-invariant if the beam-too-beam loss is smaller than 1 dB.
+In our analysis a system would, based on this reasoning, be regarded as locally lateral shift-invariant if the loss in between beams is smaller than 1 dB.
 However, depending on light conditions and e.g.
@@
 A LSV-plot is constructed for a given imaging system by imaging a laterally moving point scatterer.
-For each lateral position the root-mean-square (RMS) is calculated in range, and the resulting beam profiles are stacked to produce an image.
+For each lateral position of the point scatterer the root-mean-square (RMS) is calculated in the range dimension, and the resulting beam profiles are stacked to produce an image.
 The LSV-plots in this paper are displayed as a contour plot with contours on -1, -2, -3, -6, -12 and -24 dB.
 The maximum for each beam profile is also marked with a black dot.
-The system is said to be local lateral shift-invariant if the contours are diagonal, hence the diagonals have a loss less than 1 dB.
-
-To further quantify the shift-invariance we have calculated the mean absolute error (MAE) of the maximum point in each beam profile versus $f(x)=x$ (P-MAE), and the MAE of the amplitude variation for the same set of points (A-MAE).
-The first number quantifies the amount of geometric distortion, and the latter quantifies the gain variation or in our case the loss of signal.
-The maximum amplitude deviation along the max-point trace is also calculated and reported.
+The system is said to be locally lateral shift-invariant if the contours are diagonal, hence the diagonals have a loss less than 1 dB.
+
+To further quantify the shift-invariance we have calculated the mean absolute error (MAE) of the maximum point in each beam profile versus a straight line ($f(x)=x$).
+We name this measure P-MAE.
+The MAE of the amplitudes for the same set of maximum points versus the maximum in the whole plot is also measured.
+We call this measure A-MAE.
+The first measure quantifies the amount of geometric distortion, and the latter quantifies the gain variation or in our case the loss of signal.
+The maximum amplitude deviation along the max-point trace, or the scalloping loss, is also calculated and reported.
 We present all these numbers together with the LSV-plots.
 
-Fig.\,\ref{fig:das_shiftinvariant} presents a LSV-plot for an imaging system using DAS beamforming with a beam spacing equal to (\ref{eq:resolution}).
+Figure \ref{fig:das_shiftinvariant} presents a LSV-plot for an imaging system using DAS beamforming with a beam spacing equal to (\ref{eq:resolution}).
 The red dotted vertical lines indicates the position of transmit-receive beam pairs.
-The point scatterer response has been simulated using Field II \cite{Jensen1992, Jensen1996a} \comment{add table with simulation settings or just insert in text?}.
+The point scatterer response has been simulated using Field II \cite{Jensen1992, Jensen1996a} with a probe setup mimicking the M5S probe by GE Vingmed Ultrasound, Horten, Norway.
+The probe was excited with a 2.5 MHz and x cycles long pulse \comment{Add correct number of cycles}.
 We see that the contours are approximately diagonal and therefore we can conclude that the system is locally lateral shift-invariant.
@@
 
-In Fig.\,\ref{fig:das_shiftvariant} the beam spacing is reduced by a factor of two.
+In Fig.\,\ref{fig:das_shiftvariant} the beam density is reduced by a factor of two.
 We see how the scalloping loss is increased to a level where the contours no longer are diagonal.
 Hence, the imaging system is no longer locally lateral shift-invariant.
-We observe that the maximum peak is lost when the point is located between two beams and the distance between the beams are too high.
+We observe that the maximum peak is greatly reduced when the point is located between two beams and the distance between the beams are too high.
 
@@
 }
-\caption{LSV-plots of DAS beamforming.
-If the contours are diagonal the imaging system is said to be lateral local shif-invariant.
+\caption{LSV-plots of delay-and-sum beamforming.
+If the contours are diagonal the imaging system is said to be locally lateral shift-invariant.
 Red vertical lines are plotted where transmit-receive beam pairs are located.
@@
 The system is locally lateral shift-invariant.
-b) Undersampling with a factor two.
+b) Undersampling with a factor of two.
 The system is no longer locally lateral shift-invariant.}
@@
 \section{Local Lateral Shift-Invariance of the Capon Beamformer}\label{sec:capon_LLSI}
-We will now investigate the local lateral shift-invariance of the Capon beamformer with parameters equal to the once used in earlier work \cite{Synnevag2009, Asen}.
-In Fig.\,\ref{fig:capon1}.
-the same point scatterer response as presented in Fig.\,\ref{fig:das_shiftinvariant} has been weighted with Capon weights calculated using (\ref{eq:capon_weights}) with $L = M/2 = 32$, $K=1$ and $d=1/100$.
+We will now investigate the local lateral shift-invariance of the Capon beamformer with parameters equal to the ones used in earlier work \cite{Synnevag2009, Asen}.
+In Fig.\,\ref{fig:capon1} the same point scatterer response as presented in Fig.\,\ref{fig:das_shiftinvariant} has been weighted with Capon weights calculated using (\ref{eq:capon_weights}) with $L = M/2 = 32$, $K=1$ and $d=1/100$.
 The set of lateral displacements is constructed in such a way that the center beam is the only one hitting the point scatterer exactly.
@@
 The difference between the predicted and the simulated loss might be caused by several factors.
-The most important one being that Fig.\,\ref{fig:das_capon_beams} shows a single frequency whereas the RMS used in Fig.\,\ref{fig:capon1} average a broadband signal in range.
+The most important one being that Fig.\,\ref{fig:das_capon_beams} shows a single frequency whereas the RMS operator used in Fig.\,\ref{fig:capon1} average a broadband signal in range.
 For a slice through the center of the pulse the scalloping loss is below -30dB.
@@
 The diagonal factor $d$ and the effective adaptive aperture size $L$ can be used to control the mainlobe width of the Capon beamformer.
-In both extreme cases, $d\gg1$ and $L=1$, the Capon beamformer becomes the DAS beamformer, and will require a beamspacing according to (\ref{eq:resolution}).
+In both extreme cases, $d\gg1$ and $L=1$, the Capon beamformer becomes either a triangle or uniform-weighted DAS beamformer respectively, and will require a beamspacing according to (\ref{eq:resolution}).
 In Fig.\,\ref{fig:capon2} we show how the scalloping loss is reduced when the amount of diagonal loading is increased.
-We observe that Fig.\,\ref{fig:capon2} contains the same pattern as present in Fig.\,\ref{fig:das_shiftvariant}.
-From Fig.\,\ref{fig:capon} it should be obvious that the Capon beamformer either require increased lateral sampling in order to make the imaging system locally lateral shift-invariant, or the resolution has to be reduced to the same level as DAS.
+We observe that Fig.\,\ref{fig:capon2} now contains a scaled version of the pattern presented in Fig.\,\ref{fig:das_shiftvariant}.
+From Fig.\,\ref{fig:capon} it should be obvious that the Capon beamformer either requires increased lateral sampling in order to make the imaging system locally lateral shift-invariant, or the resolution has to be reduced to the same level as DAS.
 It would be unwise to do all the computation involved with Capon beamforming with a high $d$ value and end up with a resolution equal to DAS.
@@
 }
-\caption{LSV-plots of Capon beamforming with $L=M/2=32$ and $K=1$, and with sampling corresponding to (\ref{eq:resolution}) a) The diagonal loading factor $d=1/100$.
+\caption{LSV-plots of Capon beamforming with $L=M/2=32$ and $K=1$, and with sampling corresponding to (\ref{eq:resolution}).
+a) Capon beamforming with $d=1/100$ as diagonal loading.
 The system is not locally lateral shift-invariant.
@@
 \section{Methods}\label{sec:methods}
-We will now discuss different ways to improve on the lateral shift-invariance when the Capon beamformer is used.
+We will now discuss different ways of improving the local lateral shift-invariance when the Capon beamformer is used.
 From the previous section it should be clear that we somehow need to increase the lateral sampling to make the Capon imaging system locally lateral shift-invariant.
@@
 \subsection{Oversampling on Transmit}
-The straight forward way of obtaining sufficient sampling in an active system is by increasing the number of transmit beams (tx-beams) until (\ref{eq:resolution}) is fulfilled or the LSV-plot has an uniform level along the diagonals.
+The straight forward way of obtaining sufficient sampling in an active system is by increasing the number of transmit beams (tx-beams) until the system is laterally local shift-invariant.
 In Fig.\,\ref{fig:capon_oversampling_a} and Fig.\,\ref{fig:capon_oversampling_b} we have increased the lateral sampling on transmit by a factor of four and ten respectively.
@@
 
-From Fig.\,\ref{fig:capon_oversampling_b} we see that more than ten times oversampling has to be applied to achieve local lateral shift-invariance when a moving point scatterer is imaged without noise and with the same parameters as in Fig\,\ref{fig:capon}.
+From Fig.\,\ref{fig:capon_oversampling_b} we see that more than ten times oversampling has to be applied to achieve a local lateral shift-invariant Capon beamformer when a moving point scatterer is imaged without noise and with the same parameters as in Fig\,\ref{fig:capon}.
 %$1/100$ in diagonal loading.
-Note that a diagonal loading factor of $1/100$ will introduce a deterministic white noise component in the sample covariance matrix $40$ dB lower than the signal present.
-Hence, the covariance has been estimated from data which appeared to have an SNR of 40 dB.
-\comment{correct?}
+Note that a diagonal loading factor of $1/100$ will introduce a deterministic white noise component in the sample covariance matrix $20$ dB lower than the signal present.
+Hence, the covariance has been estimated from data which appeared to have an SNR of 20 dB.
 
@@
 The processing complexity will also increase equally.
-For these two reasons, oversampling on transmit is not a feasible option for real time ultrasound imaging, and we will not discuss this method further.
+For these two reasons, oversampling on transmit is something which is desirable to avoid in real time ultrasound imaging, and therefore we will not discuss this method further.
 
@@
 cardiac ultrasound imaging.
-However, PRB also introduce motion artifacts in the image which to some degree can be compensated for \cite{Hergum2007} \comment{cite Bastien}.
+However, PRB also introduces motion artifacts in the image which to some degree can be compensated for \cite{Hergum2007} \comment{cite Bastien}.
 In (\comment{Cite paper where they use STB + Capon if its out}) it is shown that synthetic transmit beamforming (STB) \cite{Hergum2007} works with Capon beamforming.
 In \cite{Asen} we also used STB in order to increase the beam density of a cardiac ultrasound recording.
-In the following discussion we consider PRB as being equal to oversampling on transmit when it comes to image quality.
-
-In cardiac ultrasound imaging, PRB is typically used in order to keep the number of rx-beams while the number of tx-beams are reduced.
-When using PRB for oversampling we would like to keep the number of tx-beams while the number of rx-beams are increased.
+In the following discussion we consider PRB as being equal to oversampling on transmit when it comes to image quality, and the LSV-plots for PRB is therefore not included.
+
+In cardiac ultrasound imaging, PRB is typically used in order to keep the number of rx-beams high while the number of tx-beams are reduced.
+When using PRB for oversampling we would like to maintain the same number of tx-beams while the number of rx-beams are increased.
 The acquisition frame rate will then not be affected.
@@
 The processing times reported in \cite{Asen} will then also increase with the same factor.
-Since almost 10x oversampling is required in order to achieve lateral shift invariant imaging of a moving point scatterer, non of the implementations described in \cite{Asen} would be real time any more.
-
-One way of reducing the number of matrix inversions is to estimate the covariance matrix from multiple beams \cite{Jensen2012}, generating one covariance matrix per range.
-Tailoring the method of Jensen and Austeng to cardiac ultrasound imaging it makes more sense, due to motion, to construct one covariance matrix per tx-beam using multiple parallel rx-beams.
-With this solution the number of matrix inversions will not changes, but the time taken to construct the matrices will increase.
-We will not investigate this method further in this paper.
-\comment{Kanskje litt dumt å nevne det da...?} \comment{move to discussion}
+Since almost 10x oversampling is required in order to achieve lateral shift invariant imaging of a moving point scatterer, none of the implementations described in \cite{Asen} would be real time any more.
 
@@
 In a narrow band setting the beam can make a sweep across the whole sector by only using phase rotation.
-This is also possible in broadband applications as long as the phase rotation is limited to a fraction of the wave length \comment{cite fine-coarse beamformer}.
-It is therefore possible to make a sweep between the current beam and its nearest neighbours using only phase rotation.
+This is also possible in broadband applications as long as the phase rotation is limited to a fraction of the wave length (coarse-fine beamforming \cite{Thomenius}).
+It is therefore possible to make a sweep between the current beam and its two nearest neighbours using only phase rotation.
 Phase rotation is actually what the Capon beamformer use internally to micro-steer away from bright point scatterers and to enhance edges.
@@
 \end{align}
-where $x_i$ is the element position and $\theta$ is swiped from $-\Delta/2$ too $\Delta/2$ where $\Delta$ is as defined in (\ref{eq:resolution}).
+where $x_i$ is the element position and $\theta$ is swiped from $-\Delta/2$ to $\Delta/2$ where $\Delta$ is as defined in (\ref{eq:resolution}).
 Note that the steering vector is calculated for a $L$ element array centered around the origin, even though the full array has $M$ elements.
-It is important that this synthetic array has the same element pitch as the full array.
+It is important that this subarray has the same element pitch as the full array.
 
@@
 If the number of steering angles is $N_\theta$ we have selected a uniform distribution of $\theta$  with $N_\theta/2$ angles on each side of the original steering direction if $N_\theta$ is even, and $(N_\theta-1)/2$ if $N_\theta$ is odd.
-
-Oversampling by phase rotation introduce similar block artifacts as PRB, but on a smaller scale.
-These artifacts could be removed if we distributes our steering angles as beams are spread out for the STB technique.
-We then get a phase rotation version of STB where e.g.
-eight phase rotated beams are averaged into four.
-\comment{Move to discussion?}
 
@@
 \section{Results}\label{sec:res}
-
-Acquire channel data from a phantom containing point wire targets and cysts while moving the probe laterally.
+In Fig.\,\ref{fig:LSV_invitro} we present the result of applying the Capon beamformer on \textit{in-vitro} data with and without oversampling by phase rotation.
+The data was acquired using a modified version of the Vivid E9 high-end cardiac ultrasound scanner by GE Vingmed Ultrasound, Horten, Norway, equipped with a M5S transducer in harmonic mode.
+The transmit beam density was set to match (\ref{eq:resolution}).
+A tissue mimicking phantom containing wire targets and cysts was imaged, and the data was acquired while the probe was moved across the phantom manually.
+Fig.\,\ref{fig:image_das} shows the first frame out of 14 in the acquired loop.
+Movies of the whole loop processed with DAS and Capon with and without 10 times oversampling by phase rotation can be found in the supplementary material.
+
+In Fig\,\ref{fig:image_das} one point scatterer has been selected (red box), and then used to generate LSV-plots.
+Figure \ref{fig:phantom_das} shows the resulting LSV-plot when using DAS beamforming, Fig\,\ref{fig:phantom_capon_LSV} shows the same for the Capon beamformer without oversampling, and Fig\,\ref{fig:phantom_capon_LSV_10PR} presents the result of applying 10 times oversampling by phase rotation.
+As before, vertical dotted lines indicates rx-tx beam pairs.
+
+%Acquire channel data from a phantom containing point wire targets and cysts while moving the probe laterally.
 Confirm the improved shift-invariance when using phase rotation.
 See if improved sampling also increase cyst contrast.
-Make plot of contrast between two areas in a simulation for different values of $L$.
+%Make plot of contrast between two areas in a simulation for different values of $L$.
 Make the same plot for different number of rx beams.
@@
 
-Phase rotation will introduced block artifacts equal to the once observed when using PRB.
-This can be improved on by implementing a phase rotation version of STB where 8PR are converted to 4PR-STB by averaging 4 phase rotated outputs from two beams rotated to the same angle.
-
 \begin{figure}\caption{Plot of contrast vs oversampling factor (phase rotation) in in-vivo data.}
@@
 
-\subsection{In-vitro Images}
-Acquire channel data from a phantom containing point wire targets and cysts while moving the probe laterally. Confirm the improved shift-invariance when using phase rotation.
+%\subsection{In-vitro Images}
+%Acquire channel data from a phantom containing point wire targets and cysts while moving the probe laterally. Confirm the improved shift-invariance when using phase rotation.
 
@@
 \begin{tabular}{cc}
-\subfloat[]{\includegraphics[width=0.50\linewidth]{\img LSV_image.eps}\label{fig:phantom_das}}
-\hfill{}
-\subfloat[]{\includegraphics[width=0.50\linewidth]{\img LSV_das.eps}\label{fig:phantom_capon_LSV}}\\
-\hfill{}
-\subfloat[]{\includegraphics[width=0.50\linewidth]{\img LSV_capon.eps}\label{fig:phantom_capon_LSV}}
-\hfill{}
-\subfloat[]{\includegraphics[width=0.50\linewidth]{\img LSV_capon_10pr.eps}\label{fig:phantom_capon_LSV_8PR}}
+\subfloat[]{\includegraphics[width=0.45\linewidth]{\img LSV_image.eps}\label{fig:image_das}}
+\subfloat[]{\includegraphics[width=0.45\linewidth]{\img LSV_das.eps}\label{fig:phantom_das}}\\
+\subfloat[]{\includegraphics[width=0.45\linewidth]{\img LSV_capon.eps}\label{fig:phantom_capon_LSV}}
+\subfloat[]{\includegraphics[width=0.45\linewidth]{\img LSV_capon_10pr.eps}\label{fig:phantom_capon_LSV_10PR}}
 \end{tabular}
@@
 The probe was manually dragged across the phantom.
-a) First frame.
-b) DAS c) Normal capon.
-d) Capon with oversampling using 10 times phase rotation.}
-\label{fig:das}
+See the supplementary material for videos of the whole loop \comment{Add ref to videos.} a) First frame in the recorded loop.
+A red box indicates the point scatterer used to construct the LSV-plots.
+b) Delay-and-sum beamforming.
+The amplitude variation of the point scatterer is minimal.
+c) Capon beamforming without oversampling.
+In the frame where the point scatterer is at -6 degrees the receive beam direction is very close to the signal propagation vector.
+Hence, we get a high amplitude value.
+In all the other frames the signal loss is massive.
+d) Capon with 10 times oversampling using phase rotation.
+The amplitude of the point scatterer is better preserved in all frames.
+Note the the oscillation along the diagonal does not indicate shift-variance.
+It is a result of having to low frame rate and moving the probe to fast.}
+\label{fig:LSV_invitro}
 \end{figure*}
 
-\subsection{In-Vivo Images}
-Investigate if phase rotation improves contrast in in-vivo images of the heart or liver.
-
-\begin{figure*}[!t]
-\centerline{
-\subfloat[]{\includegraphics[width=0.5\linewidth]{\img phantom1_capon_processing_K=0_L=24_full.png}\label{fig:in-vivo_capon}}
-\hfill{}
-\subfloat[]{\includegraphics[width=0.5\linewidth]{\img phantom1_capon_processing_K=0_L=24_8pr_full.png}\label{fig:in-vivo_capon_8pr}}
+%\subsection{In-Vivo Images}
+%Investigate if phase rotation improves contrast in in-vivo images of the heart or liver.
+
+%\begin{figure*}[!t]
+%\centerline{
+%\subfloat[]{\includegraphics[width=0.5\linewidth]{\img phantom1_capon_processing_K=0_L=24_full.png}\label{fig:in-vivo_capon}}
+%\hfill{}
+%\subfloat[]{\includegraphics[width=0.5\linewidth]{\img phantom1_capon_processing_K=0_L=24_8pr_full.png}\label{fig:in-vivo_capon_8pr}}
 %\hfill{}
 %\subfloat[]{\includegraphics[width=0.3\linewidth]{\img das_undersampled2_2x.eps}\label{fig:das_shiftvariant}}
-}
-\caption{\comment{For the moment phantom data. Need to acquire some high quality in-vivo data.}. In-vivo data. a) Capon image. b) Capon image with oversampling using 8PR.}
-\label{fig:das}
-\end{figure*}
+%}
+%\caption{\comment{For the moment phantom data. Need to acquire some high quality in-vivo data.}. In-vivo data. a) Capon image. b) Capon image with oversampling using 8PR.}
+%\label{fig:das}
+%\end{figure*}
 
@@
 If oversampling is not applied the edge delineation of an moving object will fluctuate in the same way as a moving point scatterer.
+
+One way of reducing the number of matrix inversions in PRB is to estimate the covariance matrix from multiple beams \cite{Jensen2012}, generating one covariance matrix per range.
+Tailoring the method of Jensen and Austeng to cardiac ultrasound imaging it makes more sense, due to motion, to construct one covariance matrix per tx-beam using multiple parallel rx-beams.
+With this solution the number of matrix inversions will not changes, but the time taken to construct the matrices will increase.
+We will not investigate this method further in this paper.
+\comment{Kanskje litt dumt å nevne det da...?} \comment{move to discussion}
+
+The phase rotation method in this paper is similar to a SVUS, but where SVUS typically has a criteria for selecting the best output after applying all the steering vectors we output all of them.
+
+Oversampling by phase rotation introduce similar block artifacts as PRB, but on a smaller scale.
+These artifacts could be removed if we distributes our steering angles as beams are spread out for the STB technique.
+We then get a phase rotation version of STB where e.g.
+eight phase rotated beams are averaged into four.
+\comment{Move to discussion?}
+
+Phase rotation will introduced block artifacts equal to the once observed when using PRB.
+This can be improved on by implementing a phase rotation version of STB where 8PR are converted to 4PR-STB by averaging 4 phase rotated outputs from two beams rotated to the same angle.
 
@@
 
-The authors would like to thank Anders Sornes for acquiring data.
+The authors would like to thank Anders S\o{}rnes for the help with acquiring \textit{in-vitro} data.
 

### 2013-11-28 HRUIofMovingObjects/jrnl_article.tex
Moving objects article: Done some writing and added updated figures.

@@
 If an ultrasound imaging system provides a presentation of a moving object which is sensitive to local spatial shifts, the system is said to be local shift-invariant.
-This can for instance happen if the axial and lateral sampling is insufficient.
+This can for instance happen if the axial or lateral sampling is insufficient.
 The Capon beamformer has been shown to provide increased lateral resolution in ultrasound images.
@@
 In this paper we investigate the local lateral shift-invariance of the Capon beamformer when imaging moving objects.
-We then discuss and investigate different methods in order to improve it.
+We then discuss and investigate different methods in order to improve on it.
 We show that insufficient lateral sampling makes an imaging system based on the Capon beamformer locally lateral shift-variant.
-We then show that the locally lateral shift-invariance can be improved by simple phase rotation without affecting the acquisition frame rate and with minor changes in processing complexity.
+We then show that the locally lateral shift-invariance can be improved by oversampling based on phase rotation without affecting the acquisition frame rate and with minor changes in processing complexity.
 \end{abstract}
@@
 In Fig.\,\ref{fig:das_capon_beams} we have plotted the steered response for DAS and the Capon beamformer using (26) and (32) in \cite{Cox1973}.
-A 64-element linear array with $\lambda/2$ element spacing, where $\lambda$ is the wavelength, is scanned over a sector equal to three times the system resolution ($1/32$ rad).
-Impinging on the array is a monochromatic plane wave with a propagation vector orthogonal to the array surface with no spatially white noise present.
+A 64-element linear array with $\lambda/2$ element spacing, where $\lambda$ is the wavelength, is scanned over a sector equal to three times the system resolution ($1/32$ radians).
+Impinging on the array is a monochromatic plane wave originating from a source moving parallel to the array surface.
+The plot shows the steered response of the source when it is located at plus and minus half the system resolution.
+No spatially white noise present.
 Hence, the interference-noise matrix ($\sigma_0^2\mat{Q}$ in \cite{Cox1973}) is the identity.
@@
 A diagonal loading of $1/100$ corresponds to an signal-to-noise ratio of 20 dB before array gain is applied.
-The figure shows that when the DAS beamformer is used with a beamspacing equal to the system resolution, only a 3.92 dB scalloping loss \cite{Harris1978} should be observed if the source of the plane wave starts moving parallel to the array.
+The figure shows that a DAS beamformer, with a beamspacing equal to the system resolution, will yield a 3.92 dB scalloping loss \cite{Harris1978} if the source is moving parallel to the array.
 However if the Capon beamformer is used with the same beam density the scalloping loss could be as large as $31$ dB.
@@
 \centerline{
-\includegraphics[width=\linewidth]{\img steerd_respons_capon_das_resolution.eps}
+\includegraphics[width=\linewidth]{\img fig1.eps}
 }
 \caption{Comparison of drop in output power in between receive beams.
-Half the system resolution is plotted to show where two beams with a spacing equal to the system resolution would intersect.
-The steered response of both delay-and-sum (DAS) and Capon beamforming ($M=64$, $L=32$, $d=0.01$) is plotted.
-A monochromatic plane wave is located at zero degrees.}
+The steered response of both delay-and-sum (DAS) and Capon beamforming ($M=64$, $L=32$, $d=0.01$) is plotted for two different time instances.
+A monochromatic plane wave is located at an angle equal to half the system resolution, and is moved a distance corresponding to the system resolution for the second steered response.}
 \label{fig:das_capon_beams}
@@
 In this paper we investigate the issue further, and we show that when a moving point scatterer is imaged with the Capon beamformer with the same beam spacing sufficient for DAS the predicted scalloping loss is observed, manifested as blinking point scatteres.
-The question is then, what can we do in order to improve on the situation without affecting the imaging frame rate.
+The question is then, what can we do in order to improve on the situation without affecting the imaging frame rate and processing throughout.
 In this paper we therefore study the scalloping loss of the Capon beamforming and investigate different methods aiming at reducing this effect, and how the different methods will affect the imaging frame rate and the processing complexity.
@@
 In Section \ref{sec:capon_LLSI} the local lateral shift-invariance of the Capon beamformer is investigated, before the different methods are presented in Section \ref{sec:methods}.
-Then, in Section \ref{sec:res} and Section \ref{sec:dis} we present and discuss the results of applying the most promising method on \textit{in-vitro} and \textit{in-vivo} data.
-Finally we give our conclusion in Section \ref{sec:con}.
+Then, in Section \ref{sec:res} and Section \ref{sec:dis} we present and discuss the results of applying the most promising method on \textit{in-vitro} data.
+Finally we give our conclusions in Section \ref{sec:con}.
 
@@
 \end{align}
-
 Note that one matrix has to be constructed and inverted for each data vector $\vec{x}[n]$ received by the system.
@@
 In an ultrasound image with 256 gray levels and 50 dB dynamic range, a 1 dB loss corresponds to 5 gray levels.
-This is approximately equal to the visibility threshold (Weber fraction of 2\%) for a grayscale image, and a loss larger than 1dB could therefore end up being visible to the observer.
+This is approximately equal to the visibility threshold (Weber fraction of 2\%) for a grayscale image, and a loss larger than 1 dB could therefore end up being visible to the observer.
 In our analysis a system would, based on this reasoning, be regarded as locally lateral shift-invariant if the loss in between beams is smaller than 1 dB.
-However, depending on light conditions and e.g.
+However, depending on viewing light conditions and e.g.
 the selected gray level mapping, a larger loss might be acceptable.
@@
 A LSV-plot is constructed for a given imaging system by imaging a laterally moving point scatterer.
-For each lateral position of the point scatterer the root-mean-square (RMS) is calculated in the range dimension, and the resulting beam profiles are stacked to produce an image.
+For each lateral position of the point scatterer the root-mean-square (RMS) of the image data is calculated in range, and the resulting beam profiles are stacked to produce an image.
 The LSV-plots in this paper are displayed as a contour plot with contours on -1, -2, -3, -6, -12 and -24 dB.
@@
 We call this measure A-MAE.
-The first measure quantifies the amount of geometric distortion, and the latter quantifies the gain variation or in our case the loss of signal.
-The maximum amplitude deviation along the max-point trace, or the scalloping loss, is also calculated and reported.
+The first measure quantifies the amount of geometric distortion, and the latter quantifies the peak gain variation or in our case the scalloping loss.
+The maximum amplitude deviation along the maximum-point trace, or the maximum scalloping loss, is also calculated and reported.
 We present all these numbers together with the LSV-plots.
@@
 The point scatterer response has been simulated using Field II \cite{Jensen1992, Jensen1996a} with a probe setup mimicking the M5S probe by GE Vingmed Ultrasound, Horten, Norway.
-The probe was excited with a 2.5 MHz and x cycles long pulse \comment{Add correct number of cycles}.
-We see that the contours are approximately diagonal and therefore we can conclude that the system is locally lateral shift-invariant.
+The probe was excited with a 2.5 MHz and 1.5 cycles long pulse.
+We see that the contours are approximately diagonal and therefore we conclude that the system is locally lateral shift-invariant.
 The P-MAE and A-MAE are also minimal.
@@
 \centerline{
-\subfloat[]{\includegraphics[width=0.5\linewidth]{\img das_normal2.eps}\label{fig:das_shiftinvariant}}
+\subfloat[]{\includegraphics[width=0.5\linewidth]{\img fig2a.eps}\label{fig:das_shiftinvariant}}
 \hfill{}
-\subfloat[]{\includegraphics[width=0.5\linewidth]{\img das_undersampled2_2x.eps}\label{fig:das_shiftvariant}}
+\subfloat[]{\includegraphics[width=0.5\linewidth]{\img fig2b.eps}\label{fig:das_shiftvariant}}
 }
@@
 \subfloat[]{
-\includegraphics[width=0.5\linewidth]{\img capon_L=32_K=1_d=001_2.eps}\label{fig:capon1}
+\includegraphics[width=0.5\linewidth]{\img fig3a.eps}\label{fig:capon1}
 }
 \subfloat[]{
-\includegraphics[width=0.5\linewidth]{\img capon_L=32_K=1_d=1_2.eps}\label{fig:capon2}
+\includegraphics[width=0.5\linewidth]{\img fig3b.eps}\label{fig:capon2}
 }
@@
 cardiac ultrasound imaging.
-However, PRB also introduces motion artifacts in the image which to some degree can be compensated for \cite{Hergum2007} \comment{cite Bastien}.
-In (\comment{Cite paper where they use STB + Capon if its out}) it is shown that synthetic transmit beamforming (STB) \cite{Hergum2007} works with Capon beamforming.
+However, PRB also introduces motion artifacts in the image which to some degree can be compensated for \cite{Hergum2007, Denarie2011, Denarie2013}.
+In \cite{Rabinovich2013} it is shown that synthetic transmit beamforming (STB) \cite{Hergum2007} works with Capon beamforming.
 In \cite{Asen} we also used STB in order to increase the beam density of a cardiac ultrasound recording.

### 2013-11-29 HRUIofMovingObjects/jrnl_article.tex
Moving objects: writing and updated figures

@@
 In order to visualize the local lateral shift-invariance of a given imaging system, we will make use of lateral shift-variance plots (LSV-plots), introduced by Hergum \textit{et al.} \cite{Hergum2007}.
-A LSV-plot is constructed for a given imaging system by imaging a laterally moving point scatterer.
-For each lateral position of the point scatterer the root-mean-square (RMS) of the image data is calculated in range, and the resulting beam profiles are stacked to produce an image.
+A LSV-plot is constructed for a given imaging system by imaging a point scatterer moving laterally at constant speed.
+For each lateral position of the point scatterer the root-mean-square (RMS) of the image data is calculated in range, and the resulting beam profiles are stacked in to an image.
+For a given scan sequence there are twice as many point positions than there are receive lines.
 The LSV-plots in this paper are displayed as a contour plot with contours on -1, -2, -3, -6, -12 and -24 dB.
-The maximum for each beam profile is also marked with a black dot.
-The system is said to be locally lateral shift-invariant if the contours are diagonal, hence the diagonals have a loss less than 1 dB.
-
-To further quantify the shift-invariance we have calculated the mean absolute error (MAE) of the maximum point in each beam profile versus a straight line ($f(x)=x$).
-We name this measure P-MAE.
-The MAE of the amplitudes for the same set of maximum points versus the maximum in the whole plot is also measured.
-We call this measure A-MAE.
-The first measure quantifies the amount of geometric distortion, and the latter quantifies the peak gain variation or in our case the scalloping loss.
-The maximum amplitude deviation along the maximum-point trace, or the maximum scalloping loss, is also calculated and reported.
+The maximum for each beam profile is also marked with a black dot, and the amplitude variation among these points is presented in a subplot next to the LSV-plot.
+
+To further quantify the shift-invariance we have calculated the mean absolute error (MAE) of the maximum point position in each beam profile versus a straight line ($f(x)=x$).
+We will refer to this measure as P-MAE.
+The MAE of the maximum point's amplitude versus the data maximum is also measured.
+We refer to this measure as A-MAE.
+The first measure quantifies the amount of geometric distortion \cite{Hergum2007}, and the latter quantifies the peak gain variation or in our case the scalloping loss.
+The maximum amplitude deviation, or the maximum scalloping loss, along this point trace is also calculated.
 We present all these numbers together with the LSV-plots.
 
-Figure \ref{fig:das_shiftinvariant} presents a LSV-plot for an imaging system using DAS beamforming with a beam spacing equal to (\ref{eq:resolution}).
+An imaging system is said to be locally lateral shift-invariant if the contours are diagonal and if the amplitude deviation among the maximum points is less than 1 dB.
+Then there is minimal geometric distortion (low P-MAE) and the scalloping loss will not be visible for the observer (low A-MAE).
+%Since both effects can occur at the same time it is important to have a measure not effected by geometric distortion.
+
+Figure \ref{fig:das_shiftinvariant} presents a LSV-plot for an imaging system using the DAS beamformer with a beam spacing equal to (\ref{eq:resolution}).
 The red dotted vertical lines indicates the position of transmit-receive beam pairs.
-The point scatterer response has been simulated using Field II \cite{Jensen1992, Jensen1996a} with a probe setup mimicking the M5S probe by GE Vingmed Ultrasound, Horten, Norway.
-The probe was excited with a 2.5 MHz and 1.5 cycles long pulse.
+The point scatterer response has been simulated using Field II \cite{Jensen1992, Jensen1996a} with a probe setup equal to the M5S probe by GE Vingmed Ultrasound, Horten, Norway.
+The probe was excited with a 2.5 MHz and 1.5 cycles long pulse, and the lateral sampling density has been doubled prior to detection through coherent interpolation.
 We see that the contours are approximately diagonal and therefore we conclude that the system is locally lateral shift-invariant.
 The P-MAE and A-MAE are also minimal.
-
-The scalloping loss (max A-MAE) in Fig.\,\ref{fig:das_shiftinvariant} is smaller than what was predicted by Fig.\,\ref{fig:das_capon_beams}.
+In this, and all the following LSV-plots based on simulations, the scan sequence always has a beam at zero degrees and the point scatterer is located exactly on this beam at one time instance.
+
+The scalloping loss in Fig.\,\ref{fig:das_shiftinvariant} is smaller than what was predicted by Fig.\,\ref{fig:das_capon_beams}.
 To understand why note that Fig.\,\ref{fig:das_capon_beams} shows the one-way response of a passive array, and the system resolution has therefore been calculated based on the receive aperture only.
@@
 Hence, the imaging system is no longer locally lateral shift-invariant.
-We observe that the maximum peak is greatly reduced when the point is located between two beams and the distance between the beams are too high.
+We observe that the maximum peak is reduced when the point is located between two beams and the distance between the beams are too high.
+The loss in Fig.\,\ref{fig:das_shiftvariant} is smaller than the predicted loss in Fig\,\ref{fig:das_capon_beams} since there are only two samples per beam along the y-axis.
+Cubic interpolation of this result reveals a scalloping loss of 2.71 dB.
 
@@
 \caption{LSV-plots of delay-and-sum beamforming.
-If the contours are diagonal the imaging system is said to be locally lateral shift-invariant.
+If the contours are diagonal and the amplitude variation among the peak points is small, the imaging system is said to be locally lateral shift-invariant.
 Red vertical lines are plotted where transmit-receive beam pairs are located.
@@
 \subfloat[]{
-\includegraphics[width=0.5\linewidth]{\img capon_L=32_K=1_d=001_4x_oversampling_2_zoomed.eps}\label{fig:capon_oversampling_a}
+\includegraphics[width=0.5\linewidth]{\img fig4a.eps}\label{fig:capon_oversampling_a}
 }
 \subfloat[]{
-\includegraphics[width=0.5\linewidth]{\img capon_L=32_K=1_d=001_10x_oversampling_2_zoomed.eps}\label{fig:capon_oversampling_b}
+\includegraphics[width=0.5\linewidth]{\img fig4b.eps}\label{fig:capon_oversampling_b}
 }
@@
 		\subfloat[]{
-			\includegraphics[width=0.5\linewidth]{\img capon_L=32_K=1_d=001_4x_oversampling_2_PR_zoomed.eps}
+			\includegraphics[width=0.5\linewidth]{\img fig5a.eps}
 		}
 		\subfloat[]{
-			\includegraphics[width=0.5\linewidth]{\img capon_L=32_K=1_d=001_10x_oversampling_2_zoomed_PR.eps}
+			\includegraphics[width=0.5\linewidth]{\img fig5b.eps}
 		}
@@
 \subfloat[]{\includegraphics[width=0.45\linewidth]{\img LSV_image.eps}\label{fig:image_das}}
-\subfloat[]{\includegraphics[width=0.45\linewidth]{\img LSV_das.eps}\label{fig:phantom_das}}\\
-\subfloat[]{\includegraphics[width=0.45\linewidth]{\img LSV_capon.eps}\label{fig:phantom_capon_LSV}}
-\subfloat[]{\includegraphics[width=0.45\linewidth]{\img LSV_capon_10pr.eps}\label{fig:phantom_capon_LSV_10PR}}
+\subfloat[]{\includegraphics[width=0.45\linewidth]{\img fig7b.eps}\label{fig:phantom_das}}\\
+\subfloat[]{\includegraphics[width=0.45\linewidth]{\img fig7c.eps}\label{fig:phantom_capon_LSV}}
+\subfloat[]{\includegraphics[width=0.45\linewidth]{\img fig7d.eps}\label{fig:phantom_capon_LSV_10PR}}
 \end{tabular}

### 2013-11-30 HRUIofMovingObjects/jrnl_article.tex
Moving objects: Done some writing. Updated and added figures.

@@
 %\boldmath
-If an ultrasound imaging system provides a presentation of a moving object which is sensitive to local spatial shifts, the system is said to be local shift-invariant.
+If an ultrasound imaging system provides a presentation of a moving object which is sensitive to local spatial shifts, the system is said to be spatially locally shift-invariant.
 This can for instance happen if the axial or lateral sampling is insufficient.
@@
 We then discuss and investigate different methods in order to improve on it.
-We show that insufficient lateral sampling makes an imaging system based on the Capon beamformer locally lateral shift-variant.
-We then show that the locally lateral shift-invariance can be improved by oversampling based on phase rotation without affecting the acquisition frame rate and with minor changes in processing complexity.
+We show that insufficient lateral sampling makes an imaging system based on the Capon beamformer locally laterally shift-variant.
+We then show that the local lateral shift-invariance can be improved by oversampling based on phase rotation without affecting the acquisition frame rate and with minor changes in processing complexity.
 \end{abstract}
@@
 where $\vec{k}_1$ is the delay-and-sum (DAS) beamformer.
-In Fig.\,\ref{fig:das_capon_beams} we have plotted the steered response for DAS and the Capon beamformer using (26) and (32) in \cite{Cox1973}.
-A 64-element linear array with $\lambda/2$ element spacing, where $\lambda$ is the wavelength, is scanned over a sector equal to three times the system resolution ($1/32$ radians).
-Impinging on the array is a monochromatic plane wave originating from a source moving parallel to the array surface.
-The plot shows the steered response of the source when it is located at plus and minus half the system resolution.
-No spatially white noise present.
+In Fig.\,\ref{fig:das_capon_beams} we have plotted the two-way response for DAS and the Capon beamformer using (26) and (32) in \cite{Cox1973}.
+A 64-element linear array with $\lambda/2$ element spacing, where $\lambda$ is the wavelength, is scanned over a sector equal to two times the two-way system resolution ($1/64$ radians) using a narrow band pulse.
+Located in the farfield of the sector is an object moving parallel to the array surface.
+The plot shows the scanned output when the object is located at plus and minus half the system resolution.
+No spatially white noise is present.
 Hence, the interference-noise matrix ($\sigma_0^2\mat{Q}$ in \cite{Cox1973}) is the identity.
@@
 A diagonal loading of $1/100$ corresponds to an signal-to-noise ratio of 20 dB before array gain is applied.
-The figure shows that a DAS beamformer, with a beamspacing equal to the system resolution, will yield a 3.92 dB scalloping loss \cite{Harris1978} if the source is moving parallel to the array.
-However if the Capon beamformer is used with the same beam density the scalloping loss could be as large as $31$ dB.
+The figure shows that a DAS beamformer, with a beamspacing equal to the system resolution, could yield a 1.9 dB scalloping loss \cite{Harris1978} if the object is moving parallel to the array.
+However if the Capon beamformer is used with the same beam density the scalloping loss could be as large as $27$ dB.
 
@@
 \centerline{
-\includegraphics[width=\linewidth]{\img fig1.eps}
-}
-\caption{Comparison of drop in output power in between receive beams.
-The steered response of both delay-and-sum (DAS) and Capon beamforming ($M=64$, $L=32$, $d=0.01$) is plotted for two different time instances.
-A monochromatic plane wave is located at an angle equal to half the system resolution, and is moved a distance corresponding to the system resolution for the second steered response.}
+\includegraphics[width=\linewidth]{\img fig1_2way.eps}
+}
+\caption{Theoretical drop in output power mid-way between receive beams for delay-and-sum and Capon beamforming ($M=64$, $L=32$, $d=0.01$) applied in a farfield-narrowband setting.
+The scanned output from both beamformers is plotted for two different time instances.
+For the first instances an object is located in the farfield of the scanned sector at an angle equal to half the system resolution, and for the latter the object is moved a distance corresponding to the system resolution.}
 \label{fig:das_capon_beams}
@@
 The question is then, what can we do in order to improve on the situation without affecting the imaging frame rate and processing throughout.
-In this paper we therefore study the scalloping loss of the Capon beamforming and investigate different methods aiming at reducing this effect, and how the different methods will affect the imaging frame rate and the processing complexity.
-
-In the next section we give an introduction to Capon beamforming for medical ultrasound imaging together with a short introduction to lateral sampling and shift invariance.
+In this paper we therefore study the scalloping loss of the Capon beamforming and investigate different methods aimed at reducing it.
+We also discuss how the different methods will affect the imaging frame rate and the processing complexity.
+
+In the next section we give an introduction to Capon beamforming for medical ultrasound imaging together with a short introduction to lateral sampling and spatial shift invariance.
 In Section \ref{sec:capon_LLSI} the local lateral shift-invariance of the Capon beamformer is investigated, before the different methods are presented in Section \ref{sec:methods}.
@@
 \end{align}
-If this requirement is not fulfilled the scalloping loss could become visible in the image and the system will no longer be locally lateral shift-invariant.
+If this requirement is not fulfilled the scalloping loss could become visible in the image and the system will no longer be locally laterally shift-invariant.
 In an ultrasound image with 256 gray levels and 50 dB dynamic range, a 1 dB loss corresponds to 5 gray levels.
 This is approximately equal to the visibility threshold (Weber fraction of 2\%) for a grayscale image, and a loss larger than 1 dB could therefore end up being visible to the observer.
-In our analysis a system would, based on this reasoning, be regarded as locally lateral shift-invariant if the loss in between beams is smaller than 1 dB.
+In our analysis a system would, based on this reasoning, be regarded as locally laterally shift-invariant if the loss in between beams is smaller than 1 dB.
 However, depending on viewing light conditions and e.g.
@@
 A LSV-plot is constructed for a given imaging system by imaging a point scatterer moving laterally at constant speed.
-For each lateral position of the point scatterer the root-mean-square (RMS) of the image data is calculated in range, and the resulting beam profiles are stacked in to an image.
+For each lateral position the root-mean-square (RMS) of the image data is calculated in range, and the resulting beam profiles are stacked in to an image.
 For a given scan sequence there are twice as many point positions than there are receive lines.
@@
 
-An imaging system is said to be locally lateral shift-invariant if the contours are diagonal and if the amplitude deviation among the maximum points is less than 1 dB.
-Then there is minimal geometric distortion (low P-MAE) and the scalloping loss will not be visible for the observer (low A-MAE).
+A given imaging system will then be locally laterally shift-invariant if the contours are diagonal and if the amplitude deviation among the maximum points is less than 1 dB.
+Then there is minimal geometric distortion (low P-MAE) and the scalloping loss will hopefully not be visible for the observer (low A-MAE).
 %Since both effects can occur at the same time it is important to have a measure not effected by geometric distortion.
 
-Figure \ref{fig:das_shiftinvariant} presents a LSV-plot for an imaging system using the DAS beamformer with a beam spacing equal to (\ref{eq:resolution}).
+Figure \ref{fig:das_shiftinvariant} presents such a LSV-plot for an imaging system using the DAS beamformer with a beam spacing equal to (\ref{eq:resolution}).
 The red dotted vertical lines indicates the position of transmit-receive beam pairs.
 The point scatterer response has been simulated using Field II \cite{Jensen1992, Jensen1996a} with a probe setup equal to the M5S probe by GE Vingmed Ultrasound, Horten, Norway.
-The probe was excited with a 2.5 MHz and 1.5 cycles long pulse, and the lateral sampling density has been doubled prior to detection through coherent interpolation.
-We see that the contours are approximately diagonal and therefore we conclude that the system is locally lateral shift-invariant.
-The P-MAE and A-MAE are also minimal.
+The probe was excited with a 2.5 MHz and 1.5 cycles long pulse with a $\text{f-number}=3$ and corresponding cosine apodization on receive.
+The transmit focus in azimuth and elevation was configured to be equal and the point scatterer was moved through this focus.
+The lateral sampling density has been doubled prior to detection through coherent interpolation.
+We see that the contours are approximately diagonal (low P-MAE) and that the amplitude variation among the maximum points is minimal (low A-MAE).
+We therefore conclude that the system is locally laterally shift-invariant.
 In this, and all the following LSV-plots based on simulations, the scan sequence always has a beam at zero degrees and the point scatterer is located exactly on this beam at one time instance.
@@
 The scalloping loss in Fig.\,\ref{fig:das_shiftinvariant} is smaller than what was predicted by Fig.\,\ref{fig:das_capon_beams}.
+To understand why note that Fig.\,\ref{fig:das_capon_beams} shows the scanned response from a narrowband system while the imaging system in Fig.\,\ref{fig:das_shiftinvariant} uses a broadband pulse.
+The beam in Fig.\,\ref{fig:das_shiftinvariant} while therefore be more similar to a squared gaussian function than the squared $sinc$ functions present in Fig.\,\ref{fig:das_capon_beams}.
+The imaging system in Fig.\,\ref{fig:das_shiftinvariant} also uses an expanding cosine window on receive.
+
+%The scalloping loss in Fig.\,\ref{fig:das_shiftinvariant} is smaller than what was predicted by Fig.\,\ref{fig:das_capon_beams}.
 To understand why note that Fig.\,\ref{fig:das_capon_beams} shows the one-way response of a passive array, and the system resolution has therefore been calculated based on the receive aperture only.
@@
 In Fig.\,\ref{fig:das_shiftvariant} the beam density is reduced by a factor of two.
-We see how the scalloping loss is increased to a level where the contours no longer are diagonal.
-Hence, the imaging system is no longer locally lateral shift-invariant.
-We observe that the maximum peak is reduced when the point is located between two beams and the distance between the beams are too high.
-The loss in Fig.\,\ref{fig:das_shiftvariant} is smaller than the predicted loss in Fig\,\ref{fig:das_capon_beams} since there are only two samples per beam along the y-axis.
-Cubic interpolation of this result reveals a scalloping loss of 2.71 dB.
+We see how the geometric distortion and scalloping loss is increased to a level where the contours no longer are diagonal and constant.
+Hence, the imaging system is no longer locally laterally shift-invariant.
+We observe that the maximum peak is both geometrically distorted and attenuated when the point is located between two beams and the distance between the beams are too high.
 
@@
 In Fig.\,\ref{fig:capon1} the same point scatterer response as presented in Fig.\,\ref{fig:das_shiftinvariant} has been weighted with Capon weights calculated using (\ref{eq:capon_weights}) with $L = M/2 = 32$, $K=1$ and $d=1/100$.
-The set of lateral displacements is constructed in such a way that the center beam is the only one hitting the point scatterer exactly.
-We see that the scalloping loss is tremendous, and not far from the value predicted by Fig\,\ref{fig:das_capon_beams}.
-The difference between the predicted and the simulated loss might be caused by several factors.
+As mentioned, the set of lateral displacements is constructed in such a way that the center beam is the only one hitting the point scatterer exactly.
+The signal cancellation caused by the difference between the signal propagation vector and the assumed steering vector is therefore minimal at this location.
+From the plot we observe that the signal cancellation is high in all the other lateral positions.
+As a consequence the scalloping loss is tremendous, and not far from the predicted value in Fig\,\ref{fig:das_capon_beams}.
+%The difference between the predicted and the simulated loss might be caused by several factors.
 The most important one being that Fig.\,\ref{fig:das_capon_beams} shows a single frequency whereas the RMS operator used in Fig.\,\ref{fig:capon1} average a broadband signal in range.
@@
 In Fig.\,\ref{fig:capon2} we show how the scalloping loss is reduced when the amount of diagonal loading is increased.
+
 We observe that Fig.\,\ref{fig:capon2} now contains a scaled version of the pattern presented in Fig.\,\ref{fig:das_shiftvariant}.
-From Fig.\,\ref{fig:capon} it should be obvious that the Capon beamformer either requires increased lateral sampling in order to make the imaging system locally lateral shift-invariant, or the resolution has to be reduced to the same level as DAS.
+From Fig.\,\ref{fig:capon} it should be obvious that the Capon beamformer either requires increased lateral sampling in order to make the imaging system locally laterally shift-invariant, or the resolution has to be reduced to the same level as DAS.
 It would be unwise to do all the computation involved with Capon beamforming with a high $d$ value and end up with a resolution equal to DAS.
@@
 
-\section{Methods}\label{sec:methods}
+\section{Oversampling Methods}\label{sec:methods}
 We will now discuss different ways of improving the local lateral shift-invariance when the Capon beamformer is used.
-From the previous section it should be clear that we somehow need to increase the lateral sampling to make the Capon imaging system locally lateral shift-invariant.
+From the previous section it should be clear that we somehow need to increase the lateral sampling to make the Capon imaging system locally laterally shift-invariant.
 In this section we will investigate how this oversampling should be conducted, how large it needs to be and how the proposed methods will affect the system frame rate and the processing complexity.
@@
 \subsection{Oversampling on Transmit}
-The straight forward way of obtaining sufficient sampling in an active system is by increasing the number of transmit beams (tx-beams) until the system is laterally local shift-invariant.
+The straight forward way of obtaining sufficient sampling in an active system is by increasing the number of transmit beams (tx-beams) until the system is locally laterally shift-invariant.
 In Fig.\,\ref{fig:capon_oversampling_a} and Fig.\,\ref{fig:capon_oversampling_b} we have increased the lateral sampling on transmit by a factor of four and ten respectively.
-As expected the scalloping loss decreases with increased lateral sampling.
+As expected the geometric distortion and scalloping loss decreases with increased lateral sampling.
 Note that Fig.\,\ref{fig:capon_oversampling_b} is zoomed compared to Fig.\,\ref{fig:capon} in order to better see the beam-to-beam variation.
@@
 The processing complexity will also increase equally.
-For these two reasons, oversampling on transmit is something which is desirable to avoid in real time ultrasound imaging, and therefore we will not discuss this method further.
+For these two reasons, oversampling on transmit is something which is desirable to avoid in real time ultrasound applications, and therefore we will not discuss this method further.
 
@@
 In \cite{Asen} we also used STB in order to increase the beam density of a cardiac ultrasound recording.
-In the following discussion we consider PRB as being equal to oversampling on transmit when it comes to image quality, and the LSV-plots for PRB is therefore not included.
+In the following discussion we consider corrected PRB as being equal to oversampling on transmit when it comes to image quality.
+The LSV-plots for PRB are therefore not included.
 
@@
 The processing times reported in \cite{Asen} will then also increase with the same factor.
-Since almost 10x oversampling is required in order to achieve lateral shift invariant imaging of a moving point scatterer, none of the implementations described in \cite{Asen} would be real time any more.
+Since more than ten times oversampling is required in order to achieve lateral shift invariant imaging of a moving point scatterer, none of the implementations described in \cite{Asen} would be real time any more.
 
@@
 This is also possible in broadband applications as long as the phase rotation is limited to a fraction of the wave length (coarse-fine beamforming \cite{Thomenius}).
-It is therefore possible to make a sweep between the current beam and its two nearest neighbours using only phase rotation.
+It is therefore possible to make a sweep between a given beam and its two nearest neighbours using only phase rotation without introducing to large errors.
 Phase rotation is actually what the Capon beamformer use internally to micro-steer away from bright point scatterers and to enhance edges.
@@
 This can be improved on by implementing a phase rotation version of STB where 8PR are converted to 4PR-STB by averaging 4 phase rotated outputs from two beams rotated to the same angle.
+
+The geometric distortion remaining in Fig. 7d should be possible to fix by a phase rotation version of STB.
 

### 2013-12-01 HRUIofMovingObjects/jrnl_article.tex
writing

@@
 %\boldmath
-If an ultrasound imaging system provides a presentation of a moving object which is sensitive to local spatial shifts, the system is said to be spatially locally shift-invariant.
+If an ultrasound imaging system provides a presentation of a moving object which is sensitive to local spatial shifts, the system is said to be spatially locally shift-variant.
 This can for instance happen if the axial or lateral sampling is insufficient.
@@
 
-In this paper we investigate the local lateral shift-invariance of the Capon beamformer when imaging moving objects.
+In this paper we analyse the local lateral shift-invariance of the Capon beamformer when imaging moving objects.
 We then discuss and investigate different methods in order to improve on it.
@@
 
-\IEEEPARstart{I}{n} the last decade several authors have applied the Capon beamformer for ultrasound imaging \cite{Synnevag2007, Vignon2008, Viola}.
+\IEEEPARstart{I}{n} the last decade several authors have applied the Capon beamformer (also known as the minimum variance beamformer) for ultrasound imaging \cite{Synnevag2007, Vignon2008, Viola}.
 The adaptive nature of the method has been shown to greatly improve the lateral resolution in ultrasound images.
@@
 The first two, together with temporal smoothing of the covariance estimate, have been applied in earlier work on Capon beamforming for medical ultrasound imaging \cite{Synnevag2009}.
-The first two trade resolution for robustness and temporal smoothing helps to preserve speckle statistics \cite{Synnevag2007a}.
+The first two trade resolution for robustness and temporal smoothing helps to preserve speckle statistics for high resolution settings \cite{Synnevag2007a}.
 SVUS reduces the signal-nulling effect of the Capon beamformer which appears if the signal is present during the covariance estimation process and if this estimate is later used with a steering vector that does not match the signal propagation vector exactly.
@@
 The maximum for each beam profile is also marked with a black dot, and the amplitude variation among these points is presented in a subplot next to the LSV-plot.
+\comment{Add interp factor}
 
@@
 The lateral sampling density has been doubled prior to detection through coherent interpolation.
-We see that the contours are approximately diagonal (low P-MAE) and that the amplitude variation among the maximum points is minimal (low A-MAE).
+We see that the contours in Fig.\,\ref{fig:das_shiftinvariant} are approximately diagonal (low P-MAE) and that the amplitude variation among the maximum points is minimal (low A-MAE).
 We therefore conclude that the system is locally laterally shift-invariant.
@@
 The scalloping loss in Fig.\,\ref{fig:das_shiftinvariant} is smaller than what was predicted by Fig.\,\ref{fig:das_capon_beams}.
-To understand why note that Fig.\,\ref{fig:das_capon_beams} shows the scanned response from a narrowband system while the imaging system in Fig.\,\ref{fig:das_shiftinvariant} uses a broadband pulse.
-The beam in Fig.\,\ref{fig:das_shiftinvariant} while therefore be more similar to a squared gaussian function than the squared $sinc$ functions present in Fig.\,\ref{fig:das_capon_beams}.
+To understand why note that Fig.\,\ref{fig:das_capon_beams} shows the scanned response for a narrowband system while the imaging system in Fig.\,\ref{fig:das_shiftinvariant} uses a broadband pulse.
+The beam in Fig.\,\ref{fig:das_shiftinvariant} while therefore be more similar to a squared gaussian function than the squared $sinc$ functions depicted in Fig.\,\ref{fig:das_capon_beams}.
 The imaging system in Fig.\,\ref{fig:das_shiftinvariant} also uses an expanding cosine window on receive.
@@
 \caption{LSV-plots of delay-and-sum beamforming.
-If the contours are diagonal and the amplitude variation among the peak points is small, the imaging system is said to be locally lateral shift-invariant.
+If the contours are diagonal and the amplitude variation among the peak points (black dots) is small, the imaging system is said to be locally laterally shift-invariant.
 Red vertical lines are plotted where transmit-receive beam pairs are located.
 a) Sampling corresponding to (\ref{eq:resolution}).
-The system is locally lateral shift-invariant.
+The system is locally laterally shift-invariant.
 b) Undersampling with a factor of two.
-The system is no longer locally lateral shift-invariant.}
+The system is no longer locally laterally shift-invariant.}
 \label{fig:das}
@@
 The straight forward way of obtaining sufficient sampling in an active system is by increasing the number of transmit beams (tx-beams) until the system is locally laterally shift-invariant.
-In Fig.\,\ref{fig:capon_oversampling_a} and Fig.\,\ref{fig:capon_oversampling_b} we have increased the lateral sampling on transmit by a factor of four and ten respectively.
+In Fig.\,\ref{fig:capon_oversampling_a} and Fig.\,\ref{fig:capon_oversampling_b} we have increased the lateral sampling on transmit by a factor of four and ten \comment{Increase the sampling to 16x to get the scalloping loss below 1 dB?} respectively.
 As expected the geometric distortion and scalloping loss decreases with increased lateral sampling.
@@
 The figure is zoomed compared with Fig\,\ref{fig:capon} to better see the beam-to-beam variation.
-a) 4x oversampling.
-b) 10x oversampling.}
+a) Four times oversampling.
+b) Ten times oversampling.}
 \label{fig:capon_oversampling}
@@
 The processing times reported in \cite{Asen} will then also increase with the same factor.
-Since more than ten times oversampling is required in order to achieve lateral shift invariant imaging of a moving point scatterer, none of the implementations described in \cite{Asen} would be real time any more.
+Since more than ten times oversampling is required in order to achieve lateral shift invariant high resolution imaging of a moving point scatterer, none of the implementations described in \cite{Asen} would be real time any more.
 
@@
 In a narrow band setting the beam can make a sweep across the whole sector by only using phase rotation.
-This is also possible in broadband applications as long as the phase rotation is limited to a fraction of the wave length (coarse-fine beamforming \cite{Thomenius}).
+This is also possible in broadband applications as long as the phase rotation is limited to a fraction of the wave length (See coarse-fine beamforming in \cite{Thomenius}).
 It is therefore possible to make a sweep between a given beam and its two nearest neighbours using only phase rotation without introducing to large errors.
@@
 
-For an uniform linear array located along the x-axis the steering vector $\vec{a}$ should be calculated as 
+For an uniform linear array located along the x-axis the steering vector $\vec{a}_\theta$ should be calculated as 
 \begin{align}
-\vec{a} = 
+\vec{a}_\theta = 
 \begin{bmatrix}
@@
 Note that the steering vector is calculated for a $L$ element array centered around the origin, even though the full array has $M$ elements.
-It is important that this subarray has the same element pitch as the full array.
+It is important that this subarray has the same element pitch, $x_i - x_{i-1}$, as the full array.
 
@@
 If the number of steering angles is $N_\theta$ we have selected a uniform distribution of $\theta$  with $N_\theta/2$ angles on each side of the original steering direction if $N_\theta$ is even, and $(N_\theta-1)/2$ if $N_\theta$ is odd.
+
+Since the phase rotation method operates on the same data as before, there will be no reduction in the acquisition frame rate by applying this method.
+In addition, since the same number of covariance matrices are constructed and inverted, there is no extra cost involved here as well.
+The only additional cost will be the multiplication of the extra steering vectors with the inverted covariance matrix.
+If the equation $\mat{\hat{R}}\vec{a} = \vec{b}$, as in \cite{Asen}, is solved using Gauss-Jordan elimination, bringing matrix $[\mat{\hat{R}}  \vec{a}_{\theta_1} \vec{a}_{\theta_2} \cdots \vec{a}_{\theta_{N_\theta}}]$ into a reduced row echelon form has the same complexity as for inverting $\mat{\hat{R}}$, as long as $N_\theta < L$.
 
@@
 
-\section{Results}\label{sec:res}
+\section{In-vitro phantom data}\label{sec:res}
 In Fig.\,\ref{fig:LSV_invitro} we present the result of applying the Capon beamformer on \textit{in-vitro} data with and without oversampling by phase rotation.
-The data was acquired using a modified version of the Vivid E9 high-end cardiac ultrasound scanner by GE Vingmed Ultrasound, Horten, Norway, equipped with a M5S transducer in harmonic mode.
+The data was acquired using a modified Vivid E9 high-end cardiac ultrasound scanner by GE Vingmed Ultrasound, Horten, Norway, equipped with a M5S transducer operating in harmonic mode.
 The transmit beam density was set to match (\ref{eq:resolution}).
@@
 Fig.\,\ref{fig:image_das} shows the first frame out of 14 in the acquired loop.
-Movies of the whole loop processed with DAS and Capon with and without 10 times oversampling by phase rotation can be found in the supplementary material.
-
-In Fig\,\ref{fig:image_das} one point scatterer has been selected (red box), and then used to generate LSV-plots.
-Figure \ref{fig:phantom_das} shows the resulting LSV-plot when using DAS beamforming, Fig\,\ref{fig:phantom_capon_LSV} shows the same for the Capon beamformer without oversampling, and Fig\,\ref{fig:phantom_capon_LSV_10PR} presents the result of applying 10 times oversampling by phase rotation.
-As before, vertical dotted lines indicates rx-tx beam pairs.
+%Movies of the whole loop processed with DAS and Capon beamforming with and without ten times oversampling by phase rotation can be found in the supplementary material \multimedia{Videos}.
+
+In Fig\,\ref{fig:image_das} one point scatterer has been selected (red box \comment{convert into a thicker white box}), and then used to generate LSV-plots.
+\comment{Todo: Add lateral coherent interpolation as done for the simulations?
+As for now speckle is not fully resolved without it.}
+
+Figure \ref{fig:phantom_das} shows the resulting LSV-plot when using the DAS beamformer.
+The scalloping loss is less than 1 dB, and the system is therefore locally laterally shift-invariant.
+As a consequence point scatterers has a consistent appearance when subjected to motion \multimedia{Video DAS}.
+
+Figure \ref{fig:phantom_capon_LSV} shows the same for the Capon beamformer without oversampling.
+In the frame where the point scatterer is at -6 degrees the receive beam direction is very close to the signal propagation vector.
+Hence, we get a high amplitude value at this location.
+In almost all the other frames the signal loss is massive.
+The result is blinking point scatterers when subjected to motion \multimedia{Video Capon}.
+
+Figure \ref{fig:phantom_capon_LSV_10PR} presents the result of applying ten times oversampling by phase rotation.
+We observe how the amplitude of the point scatterer is better preserved in all frames, and the scalloping loss is not far from our 1 dB threshold.
+Note that the oscillation along the diagonal does not indicate shift-variance.
+It is a result of lack of control when the probe was manually dragged across the phantom.
+In order to get a continues straight diagonal the probe has to be dragged across the phantom with a speed which match the acquisition frame rate.
+In addition, due to the increase in lateral resolution the Capon beamformer requires smaller frame-to-frame movements than DAS.
+Looking at the video of Capon beamforming with ten times oversampling by phase rotation \multimedia{Capon + 10PR} there is no visible blinking anymore.
+We therefore conclude that the system is now locally laterally shift invariant.
 
@@
 
-\begin{figure}\caption{Plot of contrast vs oversampling factor (phase rotation) in in-vivo data.}
-\end{figure}
+%\begin{figure}\caption{Plot of contrast vs oversampling factor (phase rotation) in in-vivo data.}
+%\end{figure}
 
@@
 }
-\caption{LSV-plots generated from \textit{in-vitro} phantom data.
-The probe was manually dragged across the phantom.
-See the supplementary material for videos of the whole loop \comment{Add ref to videos.} a) First frame in the recorded loop.
+\caption{LSV-plots generated from \textit{in-vitro} phantom data where the probe has been manually dragged across the phantom.
+See the supplementary material for videos of the whole loop \multimedia{Videos}.
+Capon beamforming has been performed with the same settings as in Fig.\,\ref{fig:capon1}.
+a) First frame in the recorded loop.
 A red box indicates the point scatterer used to construct the LSV-plots.
 b) Delay-and-sum beamforming.
-The amplitude variation of the point scatterer is minimal.
 c) Capon beamforming without oversampling.
-In the frame where the point scatterer is at -6 degrees the receive beam direction is very close to the signal propagation vector.
-Hence, we get a high amplitude value.
-In all the other frames the signal loss is massive.
-d) Capon with 10 times oversampling using phase rotation.
-The amplitude of the point scatterer is better preserved in all frames.
-Note the the oscillation along the diagonal does not indicate shift-variance.
-It is a result of having to low frame rate and moving the probe to fast.}
+d) Capon with ten times oversampling using phase rotation.}
 \label{fig:LSV_invitro}
@@
 \section{Discussion}\label{sec:dis}
-Discuss results presented in Section \ref{sec:res}.
-
-When signal detection (absolute value and square root) is performed before an ultrasound image is displayed the lateral bandwidth is increased.
+In this paper we have investigated the lateral shift-invariance property of the Capon beamformer when imaging moving objects.
+In Section \ref{} it was shown that a Capon beamformer with standard settings is not locally laterally shift-invariant when subjected to motion.
+In Section \ref{} and \ref{} we presented restored
+
+%Discuss results presented in Section \ref{sec:res}.
+
+%When signal detection (absolute value and square root) is performed before an ultrasound image is displayed the lateral bandwidth is increased.
 In order to not introduce aliasing in the image, IQ-interpolation is typically used on ultrasound systems to double the lateral beam density before detection.
@@
 
-The figure below shows a lateral shift invariance plots for a well sampled delay-and-sum image (a), the same sampling with Capon processing (b), and 4x over sampling processed with Capon.
+%The figure below shows a lateral shift invariance plots for a well sampled delay-and-sum image (a), the same sampling with Capon processing (b), and 4x over sampling processed with Capon.
 The shift invariance plots the beamprofile (TODO: Make the shift invariance plot).
@@
 
-Other methods claiming super resolution would also need oversampling (Eigenspace, LCA etc.) to some extent.
+%Other methods claiming super resolution would also need oversampling (Eigenspace, LCA etc.) to some extent.
 
@@
 
-One way of reducing the number of matrix inversions in PRB is to estimate the covariance matrix from multiple beams \cite{Jensen2012}, generating one covariance matrix per range.
+%One way of reducing the number of matrix inversions in PRB is to estimate the covariance matrix from multiple beams \cite{Jensen2012}, generating one covariance matrix per range.
 Tailoring the method of Jensen and Austeng to cardiac ultrasound imaging it makes more sense, due to motion, to construct one covariance matrix per tx-beam using multiple parallel rx-beams.
@@
 
-The phase rotation method in this paper is similar to a SVUS, but where SVUS typically has a criteria for selecting the best output after applying all the steering vectors we output all of them.
-
-Oversampling by phase rotation introduce similar block artifacts as PRB, but on a smaller scale.
+%The phase rotation method in this paper is similar to a SVUS, but where SVUS typically has a criteria for selecting the best output after applying all the steering vectors we output all of them.
+
+%Oversampling by phase rotation introduce similar block artifacts as PRB, but on a smaller scale.
 These artifacts could be removed if we distributes our steering angles as beams are spread out for the STB technique.
@@
 
-Phase rotation will introduced block artifacts equal to the once observed when using PRB.
+%Phase rotation will introduced block artifacts equal to the once observed when using PRB.
 This can be improved on by implementing a phase rotation version of STB where 8PR are converted to 4PR-STB by averaging 4 phase rotated outputs from two beams rotated to the same angle.
 
-The geometric distortion remaining in Fig. 7d should be possible to fix by a phase rotation version of STB.
-
-\subsection{Impact on Real Time Performance}
-Discuss, and show with benchmarks how the different methods for making Capon beamforming shift-invariant impact the total execution time.
+%The geometric distortion remaining in Fig. 7d should be possible to fix by a phase rotation version of STB.
+
+%Empirical testing is required to find how large the oversampling needs to be.
+Typically it depends on how bright the point scatterer or edge are with respect to the background.
+Cox \cite{Cox1973} derives formula for how large the oversampling has to be for a given signal-to-noise scenario.
+It is therefore possible to measure the highest bright-point-to-speckle value and derive the level of oversampling from this value.
+
+%\subsection{Impact on Real Time Performance}
+%Discuss, and show with benchmarks how the different methods for making Capon beamforming shift-invariant impact the total execution time.
 
 \section{Conclusion}\label{sec:con}
-...
+In this paper we have investigated the lateral shift-invariance properties of the Capon beamformer when imaging moving objects.
+It was shown that Capon beamforming with the same beam density as required for delay-and-sum beforming results in blinking point scatterers when subjected to motion.
+We then discussed and investigated different methods in order to reduced this effect.
+It was then shown that the local lateral shift-invariance could be improved by oversampling based on phase rotation without affecting the acquisition frame rate and with minor changes in processing complexity.
 

### 2013-12-01 HRUIofMovingObjects/jrnl_article.tex
writing

@@
 \begin{tabular}{cc}
-\subfloat[]{\includegraphics[width=0.45\linewidth]{\img LSV_image.eps}\label{fig:image_das}}
-\subfloat[]{\includegraphics[width=0.45\linewidth]{\img fig7b.eps}\label{fig:phantom_das}}\\
-\subfloat[]{\includegraphics[width=0.45\linewidth]{\img fig7c.eps}\label{fig:phantom_capon_LSV}}
-\subfloat[]{\includegraphics[width=0.45\linewidth]{\img fig7d.eps}\label{fig:phantom_capon_LSV_10PR}}
+\subfloat[]{\includegraphics[width=0.5\linewidth]{\img LSV_image.eps}\label{fig:image_das}}
+\subfloat[]{\includegraphics[width=0.5\linewidth]{\img fig7b.eps}\label{fig:phantom_das}}\\
+\subfloat[]{\includegraphics[width=0.5\linewidth]{\img fig7c.eps}\label{fig:phantom_capon_LSV}}
+\subfloat[]{\includegraphics[width=0.5\linewidth]{\img fig7d.eps}\label{fig:phantom_capon_LSV_10PR}}
 \end{tabular}
@@
 \section{Discussion}\label{sec:dis}
+\comment{work in progress}
 In this paper we have investigated the lateral shift-invariance property of the Capon beamformer when imaging moving objects.
-In Section \ref{} it was shown that a Capon beamformer with standard settings is not locally laterally shift-invariant when subjected to motion.
-In Section \ref{} and \ref{} we presented restored
-
-%Discuss results presented in Section \ref{sec:res}.
+In Section \ref{sec:capon_LLSI} it was shown that a Capon beamformer with high resolution parameters is not locally laterally shift-invariant when subjected to motion.
+In Section \ref{sec:methods} we therefore presented methods for improved shift-invariance.
+The most promising oversampling method, with respect to improved shift-invariance, reduction in acquisition frame rate and processing complexity, was based on phase rotation.
+In Section \ref{sec:res} we presented results showing that this method also works on \textit{in-vitro} data.
+
+The phase rotation method in this paper is similar to a SVUS.
+However, where SVUSs typically has a criteria for selecting the best output after applying all the steering vectors contained in the set, oversampling with phase rotation outputs all.
+\comment{When all vectors are calculated I do not see the point of selecting only one}.
+
+Oversampling by phase rotation introduce similar block artifacts as PRB, but on a smaller scale.
+These artifacts could be removed in the same way as for PRB if we distributes our steering angles as beams are spread out for the STB technique.
+We then get a phase rotation version of STB where e.g.
+eight phase rotated beams are averaged into four.
+The geometric distortion, or stripes, see in the video \multimedia{Capon + 10PR} should be removed by this technique.
+
+This demand for oversampling does not only affect the Capon beamformer, other methods claiming super resolution, like the Eigenspace method, LCA beamformer and etc.
+\comment{add citations}, would also need oversampling.
+
+When oversampling is applied it becomes evident that the high resolution settings $L=M/2$, $K=1$ and $d=1/100$ becomes too extreme.
+Traditionally the point spread function is always wider laterally than in range.
+With the Capon beamformer this is now changed.
+A more symmetric point spread function is obtained with the parameters $L=M/4$ and $K=0$.
+Another option is to perform pulse inversion if a spiky point spread function is desirable.
+
+Oversampling is not only essential to keep the amplitude level of bright point scatterers it will also helps to improve contrast and edge definitions.
+If oversampling is not applied the edge delineation of an moving object will fluctuate in the same way as a moving point scatterer.
+
+%Empirical testing is required to find how large the oversampling needs to be.
+Typically it depends on how bright the point scatterer or edge are with respect to the background.
+Cox \cite{Cox1973} derives formula for how large the oversampling has to be for a given signal-to-noise scenario.
+It is therefore possible to measure the highest bright-point-to-speckle value and derive the level of oversampling from this value.
 
@@
 
-In this paper we have shown that this oversampling can be accomplished by phase rotation and the number of transmit beams are therefore left unchanged.
-
-%The figure below shows a lateral shift invariance plots for a well sampled delay-and-sum image (a), the same sampling with Capon processing (b), and 4x over sampling processed with Capon.
-The shift invariance plots the beamprofile (TODO: Make the shift invariance plot).
-The results show that when using the Capon beamformer we have to over sample with a factor four to remove the blinking artifact.
-This can be done without affecting frame rate by using the available parallel receive-lines found on most ultrasound systems.
-May also use phase rotation.
-
-%Other methods claiming super resolution would also need oversampling (Eigenspace, LCA etc.) to some extent.
-
-When oversampling is applied (L=M/2, K=1, d=1/100) becomes too extreme. L=M/4 with K=0 is better.
-
-Oversampling is not only essential to keep the amplitude level of bright point scatterers it will also helps to improve contrast and edge definitions.
-If oversampling is not applied the edge delineation of an moving object will fluctuate in the same way as a moving point scatterer.
-
 %One way of reducing the number of matrix inversions in PRB is to estimate the covariance matrix from multiple beams \cite{Jensen2012}, generating one covariance matrix per range.
@@
 We will not investigate this method further in this paper.
-\comment{Kanskje litt dumt å nevne det da...?} \comment{move to discussion}
-
-%The phase rotation method in this paper is similar to a SVUS, but where SVUS typically has a criteria for selecting the best output after applying all the steering vectors we output all of them.
-
-%Oversampling by phase rotation introduce similar block artifacts as PRB, but on a smaller scale.
-These artifacts could be removed if we distributes our steering angles as beams are spread out for the STB technique.
-We then get a phase rotation version of STB where e.g.
-eight phase rotated beams are averaged into four.
-\comment{Move to discussion?}
-
-%Phase rotation will introduced block artifacts equal to the once observed when using PRB.
-This can be improved on by implementing a phase rotation version of STB where 8PR are converted to 4PR-STB by averaging 4 phase rotated outputs from two beams rotated to the same angle.
-
-%The geometric distortion remaining in Fig. 7d should be possible to fix by a phase rotation version of STB.
-
-%Empirical testing is required to find how large the oversampling needs to be.
-Typically it depends on how bright the point scatterer or edge are with respect to the background.
-Cox \cite{Cox1973} derives formula for how large the oversampling has to be for a given signal-to-noise scenario.
-It is therefore possible to measure the highest bright-point-to-speckle value and derive the level of oversampling from this value.
 
@@
 We then discussed and investigated different methods in order to reduced this effect.
-It was then shown that the local lateral shift-invariance could be improved by oversampling based on phase rotation without affecting the acquisition frame rate and with minor changes in processing complexity.
+Finally we proved that the local lateral shift-invariance could be improved by oversampling based on phase rotation without affecting the acquisition frame rate and with minor changes in processing complexity.
 

### 2013-12-04 HRUIofMovingObjects/jrnl_article.tex
Moving object: writing

@@
 However if the Capon beamformer is used with the same beam density the scalloping loss could be as large as $27$ dB.
+To get the scalloping loss for the Capon beamforming smaller than 1.9 dB for this specific example the angular displacement has to reduced with a factor 27.
 
@@
 The scanned output from both beamformers is plotted for two different time instances.
-For the first instances an object is located in the farfield of the scanned sector at an angle equal to half the system resolution, and for the latter the object is moved a distance corresponding to the system resolution.}
+For the first instances an object is located in the farfield of the scanned sector at an angle equal to half the system resolution, and for the latter the object is moved an angular distance corresponding to the system resolution.}
 \label{fig:das_capon_beams}
@@
 In Section \ref{sec:capon_LLSI} it was shown that a Capon beamformer with high resolution parameters is not locally laterally shift-invariant when subjected to motion.
-In Section \ref{sec:methods} we therefore presented methods for improved shift-invariance.
+\comment{Skriv hva som skapte shift invariansen} In Section \ref{sec:methods} we therefore presented methods for improved shift-invariance.
 The most promising oversampling method, with respect to improved shift-invariance, reduction in acquisition frame rate and processing complexity, was based on phase rotation.
@@
 If oversampling is not applied the edge delineation of an moving object will fluctuate in the same way as a moving point scatterer.
+
+\comment{Discuss the findings by Cox.
+47 dB SNR equals 10x higher resolution.
+Explaine the difference betweeen one moving point scatterer and the problem of resolving two closely spaced point scatterers.}
 

### 2013-12-05 HRUIofMovingObjects/jrnl_article.tex
Moving objects: Finally all sections are completed. Let the feedback loop begin!

--- 2013-12-01 HRUIofMovingObjects/jrnl_article.tex
@@
 Increased lateral resolution should demand denser lateral sampling.
-In previous literature on Capon beamforming for medical ultrasound imaging however only single frame scenarios have been simulated, and temporal behaviour and effects caused by the increased resolution and lack of oversampling have therefore been neglected.
+In previous literature on Capon beamforming for medical ultrasound imaging however only single frame scenarios have been simulated.
+Temporal behaviour and effects caused by the increased resolution and lack of oversampling have therefore been neglected.
 
@@
 \IEEEPARstart{I}{n} the last decade several authors have applied the Capon beamformer (also known as the minimum variance beamformer) for ultrasound imaging \cite{Synnevag2007, Vignon2008, Viola}.
-The adaptive nature of the method has been shown to greatly improve the lateral resolution in ultrasound images.
-In recent work we introduced a fast implementation of the Capon beamformer, capable of processing a stream of images at interactive frame rates \cite{Asen}.
+The adaptive nature of the method has been shown to greatly improve the lateral resolution in ultrasound images, but with a high computational cost.
+In recent work we introduced an implementation of the Capon beamformer capable of processing a stream of images at interactive frame rates \cite{Asen}.
 It is therefore an interesting question how a moving object appears after Capon beamforming has been applied.
@@
 
-An important property of an ultrasound imaging system is its local shift-invariance.
-Local shift-invariance means that an object's appearance should be preserved when the object, or the ultrasound probe, is moved slightly.
-This invariance for local movements is what makes ultrasound imaging feasible in general, but it is specially important for quantitative techniques that correlate time-varying features, like pulsed doppler and flow imaging, and speckle tracking of tissue and blood.
+An important property of an ultrasound imaging system is it's local spatial shift-invariance.
+Local spatial shift-invariance means that an object's appearance should be preserved when the object, or the ultrasound probe, is moved slightly.
+This invariance to local movements is what makes ultrasound imaging feasible in general, but it is specially important for quantitative techniques that correlate time-varying features, like pulsed doppler and flow imaging, and speckle tracking of tissue and blood.
 In this paper we analyze the lateral shift-invariance property of the Capon beamformer by studying the beamformer output when applied on simulated and \textit{in vitro} images of moving objects.
@@
 The first two trade resolution for robustness and temporal smoothing helps to preserve speckle statistics for high resolution settings \cite{Synnevag2007a}.
-SVUS reduces the signal-nulling effect of the Capon beamformer which appears if the signal is present during the covariance estimation process and if this estimate is later used with a steering vector that does not match the signal propagation vector exactly.
-This self-nulling is actually how the method obtains its super resolution capabilities, but the signal can also get lost in the process.
-In this paper we make use of a steering vector set to reduce this self-nulling effect.
+SVUS helps to reduce the signal-nulling effect of the Capon beamformer which appears if the signal is present during the covariance estimation process and if this estimate is later used with a steering vector that does not match the signal propagation vector exactly.
+This self-nulling is actually how the method obtains it's super resolution capabilities, but the signal can also get lost in the process.
+In this paper we will make use of steering vector set to reduce this self-nulling effect.
 
@@
 In Fig.\,\ref{fig:das_capon_beams} we have plotted the two-way response for DAS and the Capon beamformer using (26) and (32) in \cite{Cox1973}.
-A 64-element linear array with $\lambda/2$ element spacing, where $\lambda$ is the wavelength, is scanned over a sector equal to two times the two-way system resolution ($1/64$ radians) using a narrow band pulse.
+A 64-element linear array with $\lambda/2$ element spacing, where $\lambda$ is the wavelength, is scanned over a sector equal to two times the two-way system resolution ($1/64$ radians) using narrow band excitation.
 Located in the farfield of the sector is an object moving parallel to the array surface.
@@
 A diagonal loading of $1/100$ corresponds to an signal-to-noise ratio of 20 dB before array gain is applied.
-The figure shows that a DAS beamformer, with a beamspacing equal to the system resolution, could yield a 1.9 dB scalloping loss \cite{Harris1978} if the object is moving parallel to the array.
-However if the Capon beamformer is used with the same beam density the scalloping loss could be as large as $27$ dB.
+The figure shows that a DAS beamformer, with a beamspacing equal to the system resolution, could yield a $-1.9$ dB scalloping loss \cite{Harris1978} if the object is moving parallel to the array.
+However if the Capon beamformer is used with the same beam density the scalloping loss could be as large as $-27$ dB.
+A $-1.9$ dB scalloping loss for the Capon beamforming is obtained for this specific example when the beam density is increased with a factor of 25.
 
@@
 The scanned output from both beamformers is plotted for two different time instances.
-For the first instances an object is located in the farfield of the scanned sector at an angle equal to half the system resolution, and for the latter the object is moved a distance corresponding to the system resolution.}
+For the first instances an object is located at an angle equal to half the system resolution, and for the latter the object is moved an angular distance corresponding to the system resolution.}
 \label{fig:das_capon_beams}
@@
 
-In this paper we investigate the issue further, and we show that when a moving point scatterer is imaged with the Capon beamformer with the same beam spacing sufficient for DAS the predicted scalloping loss is observed, manifested as blinking point scatteres.
+In this paper we investigate the issue further, and we show that when a moving point scatterer is imaged with the Capon beamformer with the same beam spacing sufficient for DAS the predicted scalloping loss is observed.
 The question is then, what can we do in order to improve on the situation without affecting the imaging frame rate and processing throughout.
@@
 In the next section we give an introduction to Capon beamforming for medical ultrasound imaging together with a short introduction to lateral sampling and spatial shift invariance.
-In Section \ref{sec:capon_LLSI} the local lateral shift-invariance of the Capon beamformer is investigated, before the different methods are presented in Section \ref{sec:methods}.
-Then, in Section \ref{sec:res} and Section \ref{sec:dis} we present and discuss the results of applying the most promising method on \textit{in-vitro} data.
-Finally we give our conclusions in Section \ref{sec:con}.
+In Section \ref{sec:capon_LLSI} the local lateral shift-invariance of the Capon beamformer is investigated, before different methods for improved local lateral shift-invariance are presented in Section \ref{sec:methods}.
+Then, in Section \ref{sec:res} we present the results of applying the most promising method on \textit{in-vitro} data.
+Finally we discuss our findings in and Section \ref{sec:dis} and give our conclusions in Section \ref{sec:con}.
 
@@
 
-The lateral sampling required in an ultrasound image is given by the combined aperture size on transmit ($D_{tx}$) and receive ($D_{rx}$), and the center frequency ($f_c = c/\lambda$), where $c$ is the speed of sound in the body and $\lambda_c$ is the wavelength of the center frequency.
+The lateral sampling required in an ultrasound image for a given center frequency, $f_c$, is given by the combined aperture size on transmit ($D_{tx}$) and receive ($D_{rx}$), and the wavelength of the center frequency, $\lambda_c = c/f_c$, where $c$ is the speed of sound in the body.
 The required Nyquist beam spacing is then \cite{Hergum2007}:
 \begin{align}
-\Delta < \frac{\lambda_c}{D_{tx} + D_{rx}}. \label{eq:resolution}
+\Delta < \frac{\lambda_c}{D_{tx} + D_{rx}}\frac{1}{p}, \label{eq:resolution}
 \end{align}
-If this requirement is not fulfilled the scalloping loss could become visible in the image and the system will no longer be locally laterally shift-invariant.
-In an ultrasound image with 256 gray levels and 50 dB dynamic range, a 1 dB loss corresponds to 5 gray levels.
+where the over sampling factor $p$ can be used e.g. to reduce folding of frequencies $>f_c$. A $p$ factor of one has been used throughout this paper.
+
+Equation (\ref{eq:resolution}) was used to calculate the object position or beam spacing in Fig.\,\ref{fig:das_capon_beams}.
+Therefore, if (\ref{eq:resolution}) is not fulfilled the scalloping loss could become visible in the image and the system will no longer be locally laterally shift-invariant.
+In an ultrasound image with 256 gray levels and 50 dB dynamic range, a -1 dB loss corresponds to 5 gray levels.
 This is approximately equal to the visibility threshold (Weber fraction of 2\%) for a grayscale image, and a loss larger than 1 dB could therefore end up being visible to the observer.
-In our analysis a system would, based on this reasoning, be regarded as locally laterally shift-invariant if the loss in between beams is smaller than 1 dB.
-However, depending on viewing light conditions and e.g.
+In our further analysis a system would, based on this reasoning, be regarded as locally laterally shift-invariant if the loss in between beams is less than or close to -1 dB.
+%However, depending on viewing light conditions and e.g.
 the selected gray level mapping, a larger loss might be acceptable.
@@
 The maximum for each beam profile is also marked with a black dot, and the amplitude variation among these points is presented in a subplot next to the LSV-plot.
-\comment{Add interp factor}
+The LSV-plots based on simulations has been up-interpolated to 128 samples using cubic interpolation before presentation.
 
@@
 We refer to this measure as A-MAE.
-The first measure quantifies the amount of geometric distortion \cite{Hergum2007}, and the latter quantifies the peak gain variation or in our case the scalloping loss.
-The maximum amplitude deviation, or the maximum scalloping loss, along this point trace is also calculated.
-We present all these numbers together with the LSV-plots.
-
-A given imaging system will then be locally laterally shift-invariant if the contours are diagonal and if the amplitude deviation among the maximum points is less than 1 dB.
+The first measure quantifies the amount of geometric distortion \cite{Hergum2007}, and the latter quantifies the peak gain variation or in other words the scalloping loss.
+The maximum amplitude deviation, or maximum scalloping loss, along this point trace is also calculated.
+We present all these numbers together with the LSV-plot.
+
+A given imaging system will then be locally laterally shift-invariant if the contours are diagonal and if the amplitude deviation among the maximum points is less than or close to -1 dB.
 Then there is minimal geometric distortion (low P-MAE) and the scalloping loss will hopefully not be visible for the observer (low A-MAE).
@@
 Figure \ref{fig:das_shiftinvariant} presents such a LSV-plot for an imaging system using the DAS beamformer with a beam spacing equal to (\ref{eq:resolution}).
-The red dotted vertical lines indicates the position of transmit-receive beam pairs.
+The dotted vertical lines indicates the position of transmit-receive beam pairs.
 The point scatterer response has been simulated using Field II \cite{Jensen1992, Jensen1996a} with a probe setup equal to the M5S probe by GE Vingmed Ultrasound, Horten, Norway.
@@
 We therefore conclude that the system is locally laterally shift-invariant.
-In this, and all the following LSV-plots based on simulations, the scan sequence always has a beam at zero degrees and the point scatterer is located exactly on this beam at one time instance.
+In this, and the following LSV-plots based on simulations, the scan sequence always has a beam at zero degrees and the point scatterer is located exactly on this beam at one time instance.
 
@@
 To understand why note that Fig.\,\ref{fig:das_capon_beams} shows the scanned response for a narrowband system while the imaging system in Fig.\,\ref{fig:das_shiftinvariant} uses a broadband pulse.
-The beam in Fig.\,\ref{fig:das_shiftinvariant} while therefore be more similar to a squared gaussian function than the squared $sinc$ functions depicted in Fig.\,\ref{fig:das_capon_beams}.
+The beam in Fig.\,\ref{fig:das_shiftinvariant} will therefore be more similar to a squared gaussian function than the squared $sinc$ functions depicted in Fig.\,\ref{fig:das_capon_beams}.
 The imaging system in Fig.\,\ref{fig:das_shiftinvariant} also uses an expanding cosine window on receive.
@@
 If the contours are diagonal and the amplitude variation among the peak points (black dots) is small, the imaging system is said to be locally laterally shift-invariant.
-Red vertical lines are plotted where transmit-receive beam pairs are located.
-a) Sampling corresponding to (\ref{eq:resolution}).
+Dotted vertical lines are plotted where transmit-receive beam pairs are located.
+a) Sampling density according to (\ref{eq:resolution}).
 The system is locally laterally shift-invariant.
@@
 We will now investigate the local lateral shift-invariance of the Capon beamformer with parameters equal to the ones used in earlier work \cite{Synnevag2009, Asen}.
-In Fig.\,\ref{fig:capon1} the same point scatterer response as presented in Fig.\,\ref{fig:das_shiftinvariant} has been weighted with Capon weights calculated using (\ref{eq:capon_weights}) with $L = M/2 = 32$, $K=1$ and $d=1/100$.
+In Fig.\,\ref{fig:capon1} the same point scatterer response as presented in Fig.\,\ref{fig:das_shiftinvariant} has been weighted with Capon weights calculated using (\ref{eq:capon_weights}) with $L = M/2 = 48$, $K=1$ and $d=1/100$.
 As mentioned, the set of lateral displacements is constructed in such a way that the center beam is the only one hitting the point scatterer exactly.
@@
 }
-\caption{LSV-plots of Capon beamforming with $L=M/2=32$ and $K=1$, and with sampling corresponding to (\ref{eq:resolution}).
+\caption{LSV-plots of Capon beamforming with $L=M/2=48$ and $K=1$, and with sampling corresponding to (\ref{eq:resolution}).
 a) Capon beamforming with $d=1/100$ as diagonal loading.
@@
 The straight forward way of obtaining sufficient sampling in an active system is by increasing the number of transmit beams (tx-beams) until the system is locally laterally shift-invariant.
-In Fig.\,\ref{fig:capon_oversampling_a} and Fig.\,\ref{fig:capon_oversampling_b} we have increased the lateral sampling on transmit by a factor of four and ten \comment{Increase the sampling to 16x to get the scalloping loss below 1 dB?} respectively.
+In Fig.\,\ref{fig:capon_oversampling_a} and Fig.\,\ref{fig:capon_oversampling_b} we have increased the lateral sampling on transmit by a factor of four and ten respectively.
 As expected the geometric distortion and scalloping loss decreases with increased lateral sampling.
@@
 a) Four times oversampling.
-b) Ten times oversampling.}
+b) Ten times oversampling.
+The maximum scalloping loss is still too high for the system to be called locally laterally shift invariant.}
 \label{fig:capon_oversampling}
@@
 It is important that this subarray has the same element pitch, $x_i - x_{i-1}$, as the full array.
+The steering error introduced by this steering vector for a frequency band $B$ is less than $B/2 \times \Delta/2$.
 
@@
 The only additional cost will be the multiplication of the extra steering vectors with the inverted covariance matrix.
-If the equation $\mat{\hat{R}}\vec{a} = \vec{b}$, as in \cite{Asen}, is solved using Gauss-Jordan elimination, bringing matrix $[\mat{\hat{R}}  \vec{a}_{\theta_1} \vec{a}_{\theta_2} \cdots \vec{a}_{\theta_{N_\theta}}]$ into a reduced row echelon form has the same complexity as for inverting $\mat{\hat{R}}$, as long as $N_\theta < L$.
+If the equation $\mat{\hat{R}}\vec{a} = \vec{b}$, as in \cite{Asen}, is solved using the Gauss-Jordan algorithm, we can solve the problem for all steering vectors at once by row reducing the matrix $[\mat{\hat{R}},  \vec{a}_{\theta_1}, \vec{a}_{\theta_2}, \cdots \vec{a}_{\theta_{N_\theta}}]$ to the following form $[\mat{I},  \vec{b}_{\theta_1}, \vec{b}_{\theta_2}, \cdots \vec{b}_{\theta_{N_\theta}}]$.
+This has lower complexity than inverting $\mat{\hat{R}}$, $O(L^3)$, as long as $N_\theta < L$.
 
@@
 In Fig.\,\ref{fig:LSV_invitro} we present the result of applying the Capon beamformer on \textit{in-vitro} data with and without oversampling by phase rotation.
-The data was acquired using a modified Vivid E9 high-end cardiac ultrasound scanner by GE Vingmed Ultrasound, Horten, Norway, equipped with a M5S transducer operating in harmonic mode.
-The transmit beam density was set to match (\ref{eq:resolution}).
-A tissue mimicking phantom containing wire targets and cysts was imaged, and the data was acquired while the probe was moved across the phantom manually.
+The data was acquired using a modified Vivid E9 high-end cardiac ultrasound scanner by GE Vingmed Ultrasound, Horten, Norway, equipped with a M5S transducer operating in harmonic mode with a receive frequency of 3.7 MHz.
+The transmit beam density was set to match (\ref{eq:resolution}) based on this frequency.
+Note that in harmonic imaging the generated transmit beam at the harmonic frequency is a weighted version of the beam resulting from transmitting at the same frequency \cite{Fedewa2004}.
+The lateral sampling is therefore better than what was used in simulations.
+A tissue mimicking phantom containing wire targets and anechoic cysts was imaged, and the data was acquired while the probe was moved across the phantom manually.
 Fig.\,\ref{fig:image_das} shows the first frame out of 14 in the acquired loop.
@@
 
-In Fig\,\ref{fig:image_das} one point scatterer has been selected (red box \comment{convert into a thicker white box}), and then used to generate LSV-plots.
+In Fig\,\ref{fig:image_das} one point scatterer has been selected (white box), and then used to generate LSV-plots.
 \comment{Todo: Add lateral coherent interpolation as done for the simulations?
-As for now speckle is not fully resolved without it.}
-
-Figure \ref{fig:phantom_das} shows the resulting LSV-plot when using the DAS beamformer.
+As for now speckle is not fully resolved without it.} Figure \ref{fig:phantom_das} shows the resulting LSV-plot when using the DAS beamformer.
 The scalloping loss is less than 1 dB, and the system is therefore locally laterally shift-invariant.
@@
 \begin{tabular}{cc}
-\subfloat[]{\includegraphics[width=0.5\linewidth]{\img LSV_image.eps}\label{fig:image_das}}
+\subfloat[]{\includegraphics[width=0.5\linewidth]{\img fig7a.eps}\label{fig:image_das}}
 \subfloat[]{\includegraphics[width=0.5\linewidth]{\img fig7b.eps}\label{fig:phantom_das}}\\
@@
 \caption{LSV-plots generated from \textit{in-vitro} phantom data where the probe has been manually dragged across the phantom.
-See the supplementary material for videos of the whole loop \multimedia{Videos}.
+See the supplementary material for videos of the whole loop processed with the different beamformers \multimedia{Videos}.
 Capon beamforming has been performed with the same settings as in Fig.\,\ref{fig:capon1}.
+P-MAE is not shown due to a lack of probe movement control.
 a) First frame in the recorded loop.
-A red box indicates the point scatterer used to construct the LSV-plots.
+A white box indicates the point scatterer used to construct the LSV-plots.
 b) Delay-and-sum beamforming.
@@
 \section{Discussion}\label{sec:dis}
-\comment{work in progress}
 In this paper we have investigated the lateral shift-invariance property of the Capon beamformer when imaging moving objects.
 In Section \ref{sec:capon_LLSI} it was shown that a Capon beamformer with high resolution parameters is not locally laterally shift-invariant when subjected to motion.
-In Section \ref{sec:methods} we therefore presented methods for improved shift-invariance.
-The most promising oversampling method, with respect to improved shift-invariance, reduction in acquisition frame rate and processing complexity, was based on phase rotation.
-In Section \ref{sec:res} we presented results showing that this method also works on \textit{in-vitro} data.
-
-The phase rotation method in this paper is similar to a SVUS.
+The root of cause was found to be a too low lateral sampling density.
+In Section \ref{sec:methods} we therefore presented methods for improved shift-invariance based on oversampling.
+The most promising method, with respect to improved shift-invariance, reduction in acquisition frame rate and processing complexity, was based on phase rotation.
+In Section \ref{sec:res} we presented results showing that this method also works on \textit{in-vitro} data when the probe is subjected to motion.
+
+Note that in this paper we have investigated a single array configuration and a limited parameter range for the Capon beamformer.
+The goal was not to make a reference for which oversampling factor to use in different scenarios, but rather do a detailed investigation of the scalloping loss effect.
+We hope with this investigation that future readers will be able to find out them self what oversampling factor to use for their system.
+
+If the scene contains bright point scatterers or sharp edges we recommend to measure the ratio between highest peak and the neighbouring background level.
+In his seminal paper \cite{Cox1973} Cox derives formulas for how large the oversampling has to be for a given output signal-to-noise ratio (SNR).
+E.g 47 dB is suppose to result in a ten time improvement in resolution.
+Be aware that Cox measured resolution as the saddle point between two impinging plane waves, which typically shows smaller scalloping loss than a single moving point scatterer.
+In our example more than ten times oversampling was found to be required for a 40 dB output SNR in simulations, however \textit{in-vitro} ten times oversampling was enough for the system to be locally laterally shift-invariant.
+Possibly due to a higher initial noise level before diagonal loading was applied.
+
+Relating the presented phase rotation method to earlier work we find that it is similar to a SVUS.
 However, where SVUSs typically has a criteria for selecting the best output after applying all the steering vectors contained in the set, oversampling with phase rotation outputs all.
-\comment{When all vectors are calculated I do not see the point of selecting only one}.
-
-Oversampling by phase rotation introduce similar block artifacts as PRB, but on a smaller scale.
-These artifacts could be removed in the same way as for PRB if we distributes our steering angles as beams are spread out for the STB technique.
+It can also be though of as a narrow band approach to PRB where Capon is applied on all receive beams.
+
+Oversampling by phase rotation can therefore introduce similar block artifacts as PRB, but on a smaller scale when PRB is used to lower the number of tx-beams.
+These artifacts could be removed in the same way as for PRB if we distribute our steering angles as beams are spread out for the STB technique.
 We then get a phase rotation version of STB where e.g.
 eight phase rotated beams are averaged into four.
-The geometric distortion, or stripes, see in the video \multimedia{Capon + 10PR} should be removed by this technique.
-
-This demand for oversampling does not only affect the Capon beamformer, other methods claiming super resolution, like the Eigenspace method, LCA beamformer and etc.
-\comment{add citations}, would also need oversampling.
-
-When oversampling is applied it becomes evident that the high resolution settings $L=M/2$, $K=1$ and $d=1/100$ becomes too extreme.
-Traditionally the point spread function is always wider laterally than in range.
+The geometric distortion, or stripes, seen at the bottom of the sector in the video \multimedia{Capon + 10PR} should be removed by this technique.
+
+When oversampling is applied it becomes evident that the high resolution settings $L=M/2$, $K=1$ and $d=1/100$ comes through as too extreme.
+Traditionally in ultrasound imaging the point spread function is always wider laterally than in range.
 With the Capon beamformer this is now changed.
 A more symmetric point spread function is obtained with the parameters $L=M/4$ and $K=0$.
-Another option is to perform pulse inversion if a spiky point spread function is desirable.
-
-Oversampling is not only essential to keep the amplitude level of bright point scatterers it will also helps to improve contrast and edge definitions.
-If oversampling is not applied the edge delineation of an moving object will fluctuate in the same way as a moving point scatterer.
-
-%Empirical testing is required to find how large the oversampling needs to be.
-Typically it depends on how bright the point scatterer or edge are with respect to the background.
-Cox \cite{Cox1973} derives formula for how large the oversampling has to be for a given signal-to-noise scenario.
-It is therefore possible to measure the highest bright-point-to-speckle value and derive the level of oversampling from this value.
+Another option is to perform pulse inversion if a spiky symmetric point spread function is desirable.
+
+Oversampling is not only essential to keep the amplitude level of bright point scatterers it also helps to improve contrast and edge definitions.
+If oversampling is not applied homogeneous speckle regions could get a too low average value and the edge delineation of an moving object will fluctuate in the same way as a moving point scatterer would have sparkled.
+
+Finally we want to stress the fact that this demand for oversampling does not only affect the Capon or minimum variance beamformer.
+Other methods claiming super resolution, like the Eigenspace method, LCA beamformer etc.
+\cite{Synnevag2011, Nilsen2009, Mehdizadeh2012a, Kim}, would also need oversampling when higher lateral resolution is obtained.
+The same applies for high resolution spectral estimation \cite{Ekroll2012}.
 
@@
 
-The authors would like to thank Anders S\o{}rnes for the help with acquiring \textit{in-vitro} data.
+The authors would like to thank Anders S\o{}rnes for the help with acquiring \textit{in-vitro} data. \comment{Ask Anders if hi wants to be mentioned.}
 

### 2013-12-06 ThesisJP/Introduction/introduction.tex
ThesisJP: Started writing.

% Thesis introduction
% Author: Tore.
%

General introduction to ultrasound, GPUs etc...

\section {Ultrasound imaging}

\subsection{Basic beamforming}

\subsection{Adaptive Beamforming}

\subsection{Shift invariance and sampling}

\subsection{Volume rendering}

\subsection{Simulations}

\section{General purpose programming on graphics processing units}

\section{Motivation and aims of study}
\label{sec:motivationAndAims}

\section{Thesis outline}

\section{Summary of presented work}

\subsubsection{Paper no. 1:\\
An Optimised GPU Implementation of the MVDR Beamformer for Active Sonar Imaging}

\vspace{0.3 cm}
{\it Submitted for review to IEEE Transactions on Oceanic Engineering.}
\vspace{0.3 cm}

Summary of paper...

\vspace{0.3 cm}
\emph{The work was a joint effort with PhD student ...}

\subsubsection{Paper no. 2:\\
Implementing Capon Beamforming on a GPU for Real-Time Cardiac Ultrasound Imaging}

\vspace{0.3 cm}
{\it IEEE Transactions on Ultrasonics, Ferroelectrics, and Frequency Control, vol. XY, January 2014.}
\vspace{0.3 cm}

Summary of paper...

\vspace{0.3 cm}
\emph{The work was a joint effort with PhD student ...}

\subsubsection{Paper no. 3:\\
Capon Beamforming Applied on Moving Objects - An Analysis of Local Lateral Shift-Invariance}

\vspace{0.3 cm}
{\it Submitted for review to IEEE Transactions on Ultrasonics, Ferroelectrics, and Frequency Control.}
\vspace{0.3 cm}

Summary of paper...

\subsubsection{Paper no. 4:\\
Adaptive Volume Rendering of Cardiac 3D Ultrasound Images - Utilizing Blood Pool Statistics}

\vspace{0.3 cm}
{\it Proc. SPIE Medical Imaging 2012}
\vspace{0.3 cm}

Summary of paper...

\subsubsection{Paper no. 5:\\
Huygens on Speed: Interactive Simulation of Ultrasound Pressure Fields}

\vspace{0.3 cm}
{\it Proc. IEEE Ultrasonics Symposium 2012}
\vspace{0.3 cm}

Summary of paper...

\section{Discussion}

\section{Conclusion}

\section{Future work}

\section{Publication list}

\subsubsection{Papers included in the thesis}
\begin{enumerate}
	\item J. I. Buskenes, J. P. \AA{}sen, C.-I. C. Nilsen and A. Austeng, "An Optimised GPU Implementation of the MVDR Beamformer for Active Sonar Imaging", {\it Submitted for review to IEEE Transactions on Oceanic Engineering}.
	\item 
\end{enumerate}

\subsubsection{Conference proceedings}
\begin{enumerate}
	\item 
	\item 
\end{enumerate}

\subsubsection{Presentations and abstracts}
\begin{enumerate}
	\item
\end{enumerate}

\endinput

### 2013-12-06 ThesisJP/Background/background.tex
ThesisJP: Changed section layout.

--- 2013-12-06 ThesisJP/Introduction/introduction.tex
@@
 
-\subsection{Volume rendering}
+\section{Volume rendering}
 
-\subsection{Simulations}
+\section{Simulations}
 
@@
 
-\section{Motivation and aims of study}
-\label{sec:motivationAndAims}
-
-\section{Thesis outline}
-
-\section{Summary of presented work}
-
-\subsubsection{Paper no. 1:\\
-An Optimised GPU Implementation of the MVDR Beamformer for Active Sonar Imaging}
-
-\vspace{0.3 cm}
-{\it Submitted for review to IEEE Transactions on Oceanic Engineering.}
-\vspace{0.3 cm}
-
-Summary of paper...
-
-\vspace{0.3 cm}
-\emph{The work was a joint effort with PhD student ...}
-
-\subsubsection{Paper no. 2:\\
-Implementing Capon Beamforming on a GPU for Real-Time Cardiac Ultrasound Imaging}
-
-\vspace{0.3 cm}
-{\it IEEE Transactions on Ultrasonics, Ferroelectrics, and Frequency Control, vol. XY, January 2014.}
-\vspace{0.3 cm}
-
-Summary of paper...
-
-\vspace{0.3 cm}
-\emph{The work was a joint effort with PhD student ...}
-
-\subsubsection{Paper no. 3:\\
-Capon Beamforming Applied on Moving Objects - An Analysis of Local Lateral Shift-Invariance}
-
-\vspace{0.3 cm}
-{\it Submitted for review to IEEE Transactions on Ultrasonics, Ferroelectrics, and Frequency Control.}
-\vspace{0.3 cm}
-
-Summary of paper...
-
-\subsubsection{Paper no. 4:\\
-Adaptive Volume Rendering of Cardiac 3D Ultrasound Images - Utilizing Blood Pool Statistics}
-
-\vspace{0.3 cm}
-{\it Proc. SPIE Medical Imaging 2012}
-\vspace{0.3 cm}
-
-Summary of paper...
-
-\subsubsection{Paper no. 5:\\
-Huygens on Speed: Interactive Simulation of Ultrasound Pressure Fields}
-
-\vspace{0.3 cm}
-{\it Proc. IEEE Ultrasonics Symposium 2012}
-\vspace{0.3 cm}
-
-Summary of paper...
-
-\section{Discussion}
-
-\section{Conclusion}
-
-\section{Future work}
-
-\section{Publication list}
-
-\subsubsection{Papers included in the thesis}
-\begin{enumerate}
-	\item J.
-I.
-Buskenes, J.
-P.
-\AA{}sen, C.-I.
-C.
-Nilsen and A.
-Austeng, "An Optimised GPU Implementation of the MVDR Beamformer for Active Sonar Imaging", {\it Submitted for review to IEEE Transactions on Oceanic Engineering}.
-	\item 
-\end{enumerate}
-
-\subsubsection{Conference proceedings}
-\begin{enumerate}
-	\item 
-	\item 
-\end{enumerate}
-
-\subsubsection{Presentations and abstracts}
-\begin{enumerate}
-	\item
-\end{enumerate}
-
 \endinput

### 2013-12-06 ThesisJP/Introduction/introduction.tex
ThesisJP: Changed section layout.

@@
-% Thesis introduction
-% Author: Tore.
-%
-
-General introduction to ultrasound, GPUs etc...
-
-\section {Ultrasound imaging}
-
-\subsection{Basic beamforming}
-
-\subsection{Adaptive Beamforming}
-
-\subsection{Shift invariance and sampling}
-
-\subsection{Volume rendering}
-
-\subsection{Simulations}
-
-\section{General purpose programming on graphics processing units}
-
 \section{Motivation and aims of study}
@@
 
-\vspace{0.3 cm}
 {\it Submitted for review to IEEE Transactions on Oceanic Engineering.}
@@
 
-\vspace{0.3 cm}
 {\it IEEE Transactions on Ultrasonics, Ferroelectrics, and Frequency Control, vol. XY, January 2014.}
@@
 
-\vspace{0.3 cm}
 {\it Submitted for review to IEEE Transactions on Ultrasonics, Ferroelectrics, and Frequency Control.}
@@
 
-\vspace{0.3 cm}
 {\it Proc. SPIE Medical Imaging 2012}
@@
 
-\vspace{0.3 cm}
 {\it Proc. IEEE Ultrasonics Symposium 2012}

### 2013-12-06 ThesisJP/Introduction/introduction.tex
ThesisJP: Added reference list.

@@
-\section{Motivation and aims of study}
-\label{sec:motivationAndAims}
+\section{Background and motivation}
 
-\section{Thesis outline}
+\section{Aims of study}
 
@@
 
-\subsubsection{Paper no. 1:\\
+\subsubsection{Paper no. 1 (Chapter 3):\\
 An Optimised GPU Implementation of the MVDR Beamformer for Active Sonar Imaging}
@@
 
-\subsubsection{Paper no. 2:\\
+\subsubsection{Paper no. 2 (Chapter 4):\\
 Implementing Capon Beamforming on a GPU for Real-Time Cardiac Ultrasound Imaging}
@@
 
-\subsubsection{Paper no. 3:\\
+\subsubsection{Paper no. 3 (Chapter 5):\\
 Capon Beamforming Applied on Moving Objects - An Analysis of Local Lateral Shift-Invariance}
@@
 
-\subsubsection{Paper no. 4:\\
+\subsubsection{Paper no. 4 (Chapter 6):\\
 Adaptive Volume Rendering of Cardiac 3D Ultrasound Images - Utilizing Blood Pool Statistics}
@@
 
-\subsubsection{Paper no. 5:\\
+\subsubsection{Paper no. 5 (Chapter 7):\\
 Huygens on Speed: Interactive Simulation of Ultrasound Pressure Fields}
@@
 
+\section{Future work}
+
 \section{Conclusion}
-
-\section{Future work}
 
@@
 I.
-Buskenes, J.
+Buskenes, \textbf{J.
 P.
-\AA{}sen, C.-I.
+\AA{}sen}, C.-I.
 C.
 Nilsen and A.
-Austeng, "An Optimised GPU Implementation of the MVDR Beamformer for Active Sonar Imaging", {\it Submitted for review to IEEE Transactions on Oceanic Engineering}.
-	\item 
+Austeng, "An optimised GPU implementation of the MVDR beamformer for active sonar imaging", {\it Submitted for review to IEEE Transactions on Oceanic Engineering}.
+	\item \textbf{J.
+P.
+\AA{}sen}, J.
+I.
+Buskenes, C.-I.
+C.
+Nilsen, A.
+Austeng and S.
+Holm, "Implementing Capon beamforming on a GPU for real-time cardiac ultrasound imaging", {\it IEEE Transactions on Ultrasonics, Ferroelectrics, and Frequency Control, vol.
+XY, January 2014}.
+ 	\item \textbf{J.
+P.
+\AA{}sen}, A.
+Austeng and S.
+Holm, "Capon beamforming applied on moving objects - An analysis of local lateral shift-invariance", {\it Submitted for review to IEEE Transactions on Ultrasonics, Ferroelectrics, and Frequency Control}.
+	\item \textbf{J.
+P.
+\AA{}sen}, E.
+Steen, G.
+Kiss, A.
+Thorstensen and S.
+I.
+Rabben, "Adaptive volume rendering of cardiac 3D ultrasound images - utilizing blood pool statistics", {\it Proc.
+SPIE Medical Imaging 2012}.
+	\item \textbf{J. P. \AA{}sen} and S. Holm, "Huygens on speed: Interactive simulation of ultrasound pressure fields", {\it Proc. IEEE Ultrasonics Symposium 2012}.
 \end{enumerate}
 
-\subsubsection{Conference proceedings}
+\subsubsection{Related publications not included in the thesis}
 \begin{enumerate}
-	\item 
-	\item 
-\end{enumerate}
+	\item \textbf{J.
+P.
+\AA{}sen}, J.
+I.
+Buskenes, C.-I.
+C.
+Nilsen, A.
+Austeng and S.
+Holm, "Implementing Capon Beamforming on the GPU for Real-Time Cardiac Ultrasound Imaging", {\it Proc.
+IEEE Ultrasonics Symposium 2012}.
 
-\subsubsection{Presentations and abstracts}
-\begin{enumerate}
-	\item
+	\item J.
+I.
+Buskenes, \textbf{J.
+P.
+\AA{}sen}, C.-I.
+C.
+Nilsen and A.
+Austeng, "Adapting the Minimum variance beamformer to a graphics processing unit for active sonar imaging systems", {\it Proc.
+Meetings on Acoustics 2013}.
+
+	\item G. Kiss, E. Steen, \textbf{J. P \AA{}sen} and H. Torp, "GPU volume rendering in 3D echocardiography: real-time pre-processing and ray-casting", {\it Proc. IEEE Ultrasonics Symposium 2010}.
+
+	\item G.
+Kiss, \textbf{J.
+P.
+\AA{}sen}, J.
+Bogaert, B.
+H.
+Amundsen, P.
+Claus, J.
+R.
+M.
+D'hooge and H.
+Torp, "Multi-modal cardiac image fusion and visualization on the GPU", {\it Proc.
+IEEE Ultrasonics Symposium 2011}.
+
+	\item E. S. Br\o{}nstad, \textbf{J. P. \AA{}sen}, H. Torp and G. Kiss, "Visibility driven visualization of 3D cardiac ultrasound data on the GPU", {\it Proc. IEEE Ultrasonics Symposium 2012}.
 \end{enumerate}

### 2013-12-08 HRUIofMovingObjects/jrnl_article.tex
Moving objects: done some writing.

@@
 \author{
-   Jon Petter \AA{}sen, \IEEEmembership{Student Member,~IEEE,}, Andreas Austeng, \IEEEmembership{Member,~IEEE,} and Sverre Holm, \IEEEmembership{Senior Member,~IEEE}% <-this % stops a space
+   Jon Petter \AA{}sen, \IEEEmembership{Student Member,~IEEE,} Andreas Austeng, \IEEEmembership{Member,~IEEE,} and Sverre Holm, \IEEEmembership{Senior Member,~IEEE}% <-this % stops a space
 
@@
 
-An important property of an ultrasound imaging system is it's local spatial shift-invariance.
+An important property of an ultrasound imaging system is its local spatial shift-invariance.
 Local spatial shift-invariance means that an object's appearance should be preserved when the object, or the ultrasound probe, is moved slightly.
@@
 Hence, the method yields data dependent weights.
-Capon beamforming is however also known for its suboptimal performance if the sample size available for covariance estimation is limited \cite{Mestre2006}, if correlated signals are present \cite{Widrow1982}, and if the assumed steering vector does not match the signal propagation vector \cite{Wax1996, Wax1996a}.
+Capon beamforming is however also known for its suboptimal performance if e.g.
+the sample size available for covariance estimation is limited \cite{Mestre2006}, if correlated signals are present \cite{Widrow1982}, and if the assumed steering vector does not match the signal propagation vector \cite{Wax1996, Wax1996a}.
 This has led to the development of several robust versions of the Capon beamformer with e.g.
-spatial smoothing \cite{Shan1985} or diagonal loading \cite{JianLi2003} of the covariance estimate, or by applying steering vector uncertainty sets (SVUS) \cite{Lorenz2005, Rubsamen2013}.
-The first two, together with temporal smoothing of the covariance estimate, have been applied in earlier work on Capon beamforming for medical ultrasound imaging \cite{Synnevag2009}.
-The first two trade resolution for robustness and temporal smoothing helps to preserve speckle statistics for high resolution settings \cite{Synnevag2007a}.
-SVUS helps to reduce the signal-nulling effect of the Capon beamformer which appears if the signal is present during the covariance estimation process and if this estimate is later used with a steering vector that does not match the signal propagation vector exactly.
-This self-nulling is actually how the method obtains it's super resolution capabilities, but the signal can also get lost in the process.
-In this paper we will make use of steering vector set to reduce this self-nulling effect.
+spatial smoothing \cite{Shan1985} or diagonal loading \cite{JianLi2003} of the covariance estimate, or by applying a steering vector uncertainty set \cite{Lorenz2005, Rubsamen2013}.
+Spatial smoothing and diagonal loading, together with temporal smoothing of the covariance estimate, have been applied in earlier work on Capon beamforming for medical ultrasound imaging \cite{Synnevag2009}.
+The first two trade resolution for robustness and temporal smoothing helps to preserve speckle statistics for high resolution parameters \cite{Synnevag2007a}.
+A steering vector uncertainty sets helps to reduce signal-nulling which appears if the signal of interest is present during the covariance estimation process and if this estimate is later used with a steering vector that does not match the signal propagation vector exactly.
+This self-nulling is actually how the Capon beamformer obtains its super resolution capabilities, but the signal can also get lost in the process.
+In \cite{JianLi2003} it was shown that most methods based on steering vector uncertainty sets are equivalent to diagonal loading where the loading factor depends on the steering vector uncertainty.
+In this paper we will make use of a set of steering vectors to reduce this self-nulling effect by means of over sampling.
 
@@
 The effect of typical subarray averaging and diagonal loading ($L=32$ and $d=1/100$) has also been added to the Capon beamformer output.
-A diagonal loading of $1/100$ corresponds to an signal-to-noise ratio of 20 dB before array gain is applied.
-The figure shows that a DAS beamformer, with a beamspacing equal to the system resolution, could yield a $-1.9$ dB scalloping loss \cite{Harris1978} if the object is moving parallel to the array.
-However if the Capon beamformer is used with the same beam density the scalloping loss could be as large as $-27$ dB.
-A $-1.9$ dB scalloping loss for the Capon beamforming is obtained for this specific example when the beam density is increased with a factor of 25.
+A diagonal loading of $1/100$ corresponds to a signal-to-noise ratio (SNR) of 20 dB before array gain is applied.
+The figure shows that a DAS beamformer, with a beamspacing equal to the system resolution, could yield a $1.9$ dB scalloping loss \cite{Harris1978} if the object is moving parallel to the array.
+However if the Capon beamformer is used with the same beam density the scalloping loss could be as large as $27$ dB.
+A $1.9$ dB scalloping loss for the Capon beamforming is obtained for this specific example when the beam density is increased with a factor of 25.
 
@@
 The scanned output from both beamformers is plotted for two different time instances.
-For the first instances an object is located at an angle equal to half the system resolution, and for the latter the object is moved an angular distance corresponding to the system resolution.}
+For the first instances an object is located at an angle equal to half the system resolution, and for the latter the object is moved an angular distance corresponding to the system resolution.
+If the beam density was reduced to one beam per system resolution the maximum scalloping loss is found at zero degrees.}
 \label{fig:das_capon_beams}
@@
 
-In this paper we investigate the issue further, and we show that when a moving point scatterer is imaged with the Capon beamformer with the same beam spacing sufficient for DAS the predicted scalloping loss is observed.
+In this paper we investigate the issue further.
+We show that when a moving point scatterer is imaged with the Capon beamformer with the same beam spacing sufficient for DAS, the predicted scalloping loss is observed.
 The question is then, what can we do in order to improve on the situation without affecting the imaging frame rate and processing throughout.
-In this paper we therefore study the scalloping loss of the Capon beamforming and investigate different methods aimed at reducing it.
+We try two answer this by studying the scalloping loss of the Capon beamforming and investigate different methods aimed at reducing it.
 We also discuss how the different methods will affect the imaging frame rate and the processing complexity.
 
-In the next section we give an introduction to Capon beamforming for medical ultrasound imaging together with a short introduction to lateral sampling and spatial shift invariance.
+In the next section we give an introduction to Capon beamforming for medical ultrasound imaging together with a short introduction to lateral sampling and spatial shift-invariance.
 In Section \ref{sec:capon_LLSI} the local lateral shift-invariance of the Capon beamformer is investigated, before different methods for improved local lateral shift-invariance are presented in Section \ref{sec:methods}.
 Then, in Section \ref{sec:res} we present the results of applying the most promising method on \textit{in-vitro} data.
-Finally we discuss our findings in and Section \ref{sec:dis} and give our conclusions in Section \ref{sec:con}.
+Finally we discuss our findings in Section \ref{sec:dis} and give our conclusions in Section \ref{sec:con}.
 
@@
 \subsection{Capon Beamforming}
-The Capon beamformer produce a weight vector, like the one in (\ref{eq:das}), based on the impinging signal and an optimization criteria. The minimization problem is as follow \cite{Capon1969}:
+The Capon beamformer produce a weight vector, like the one in (\ref{eq:das}), based on the impinging signal and an optimization criterion. The minimization problem is as follow \cite{Capon1969}:
 \begin{align}
@@
 
-The lateral sampling required in an ultrasound image for a given center frequency, $f_c$, is given by the combined aperture size on transmit ($D_{tx}$) and receive ($D_{rx}$), and the wavelength of the center frequency, $\lambda_c = c/f_c$, where $c$ is the speed of sound in the body.
+The lateral sampling required in an ultrasound image for a given center frequency, $f_c$, and bandwidth, $B$, is given by the combined aperture size on transmit ($D_{tx}$) and receive ($D_{rx}$), and the wavelength of the center frequency, $\lambda_c = c/f_c$, where $c$ is the speed of sound in the body.
 The required Nyquist beam spacing is then \cite{Hergum2007}:
@@
 \end{align}
-where the over sampling factor $p$ can be used e.g. to reduce folding of frequencies $>f_c$. A $p$ factor of one has been used throughout this paper.
+where the over sampling factor $p$ can be used to reduce folding of frequencies between $f_c$ and $f_c + B/2$. A $p$ factor of one has been used throughout this paper.
 
@@
 Therefore, if (\ref{eq:resolution}) is not fulfilled the scalloping loss could become visible in the image and the system will no longer be locally laterally shift-invariant.
-In an ultrasound image with 256 gray levels and 50 dB dynamic range, a -1 dB loss corresponds to 5 gray levels.
+In an ultrasound image with 256 gray levels and 50 dB dynamic range, a 1 dB loss corresponds to 5 gray levels.
 This is approximately equal to the visibility threshold (Weber fraction of 2\%) for a grayscale image, and a loss larger than 1 dB could therefore end up being visible to the observer.
-In our further analysis a system would, based on this reasoning, be regarded as locally laterally shift-invariant if the loss in between beams is less than or close to -1 dB.
+In our further analysis a system would be regarded as locally laterally shift-invariant if the scalloping loss is less than 1 dB.
+However, what eventually matters is whether the loss is visible in the image or not.
 %However, depending on viewing light conditions and e.g.
@@
 We will refer to this measure as P-MAE.
-The MAE of the maximum point's amplitude versus the data maximum is also measured.
+The MAE of the maximum points' amplitude versus the data maximum is also measured.
 We refer to this measure as A-MAE.
@@
 
-A given imaging system will then be locally laterally shift-invariant if the contours are diagonal and if the amplitude deviation among the maximum points is less than or close to -1 dB.
+A given imaging system will then be locally laterally shift-invariant if the contours are diagonal and if the amplitude deviation among the maximum points is less than 1 dB.
 Then there is minimal geometric distortion (low P-MAE) and the scalloping loss will hopefully not be visible for the observer (low A-MAE).
@@
 In Fig.\,\ref{fig:das_shiftvariant} the beam density is reduced by a factor of two.
-We see how the geometric distortion and scalloping loss is increased to a level where the contours no longer are diagonal and constant.
+We see how the geometric distortion and scalloping loss is increased to a level where the contours no longer are diagonal and the maximum trace is not constant.
 Hence, the imaging system is no longer locally laterally shift-invariant.
@@
 In Fig.\,\ref{fig:capon1} the same point scatterer response as presented in Fig.\,\ref{fig:das_shiftinvariant} has been weighted with Capon weights calculated using (\ref{eq:capon_weights}) with $L = M/2 = 48$, $K=1$ and $d=1/100$.
-As mentioned, the set of lateral displacements is constructed in such a way that the center beam is the only one hitting the point scatterer exactly.
-The signal cancellation caused by the difference between the signal propagation vector and the assumed steering vector is therefore minimal at this location.
-From the plot we observe that the signal cancellation is high in all the other lateral positions.
+As mentioned, the beam at zero degrees is the only one where the point scatterer is located in the center of the beam.
+The signal cancellation caused by the difference between the signal propagation vector and the assumed steering vector is therefore minimal when the point scatterer has zero lateral offset.
+From the plot we observe that the signal cancellation is high for all the other offsets.
 As a consequence the scalloping loss is tremendous, and not far from the predicted value in Fig\,\ref{fig:das_capon_beams}.
@@
 The diagonal factor $d$ and the effective adaptive aperture size $L$ can be used to control the mainlobe width of the Capon beamformer.
-In both extreme cases, $d\gg1$ and $L=1$, the Capon beamformer becomes either a triangle or uniform-weighted DAS beamformer respectively, and will require a beamspacing according to (\ref{eq:resolution}).
+In both extreme cases, $d\gg1$ and $L=1$, the Capon beamformer becomes either a triangle or uniform-weighted DAS beamformer respectively, and a beamspacing according to (\ref{eq:resolution}) would then be sufficient.
 In Fig.\,\ref{fig:capon2} we show how the scalloping loss is reduced when the amount of diagonal loading is increased.
@@
 %$1/100$ in diagonal loading.
-Note that a diagonal loading factor of $1/100$ will introduce a deterministic white noise component in the sample covariance matrix $20$ dB lower than the signal present.
+Note that a diagonal loading factor of $1/100$ will introduce a white noise component in the sample covariance matrix $20$ dB lower than the signal present.
 Hence, the covariance has been estimated from data which appeared to have an SNR of 20 dB.
@@
 cardiac ultrasound imaging.
-However, PRB also introduces motion artifacts in the image which to some degree can be compensated for \cite{Hergum2007, Denarie2011, Denarie2013}.
-In \cite{Rabinovich2013} it is shown that synthetic transmit beamforming (STB) \cite{Hergum2007} works with Capon beamforming.
+However, PRB also introduces geometrical distortion which to some degree can be compensated for by using synthetic transmit beamforming (STB) \cite{Hergum2007, Denarie2011, Denarie2013}.
+In \cite{Rabinovich2013} it is shown that STB works with Capon beamforming.
 In \cite{Asen} we also used STB in order to increase the beam density of a cardiac ultrasound recording.
@@
 However, the steering vector can also, as in narrow band applications, be varied over a set of pre-defined vectors.
-In a narrow band setting the beam can make a sweep across the whole sector by only using phase rotation.
-This is also possible in broadband applications as long as the phase rotation is limited to a fraction of the wave length (See coarse-fine beamforming in \cite{Thomenius}).
+In a narrow band setting the beam can make a sweep across the whole sector by using phase rotation.
+This is also possible in broadband applications as long as the phase rotation is limited to a fraction of the wave length (See coarse-fine beamforming in \cite{Thomenius} \comment{find a better citation}).
 It is therefore possible to make a sweep between a given beam and its two nearest neighbours using only phase rotation without introducing to large errors.
@@
 
-For an uniform linear array located along the x-axis the steering vector $\vec{a}_\theta$ should be calculated as 
+For a uniform linear array located along the x-axis the steering vector $\vec{a}_\theta$ should be calculated as 
 \begin{align}

### 2013-12-10 HRUIofMovingObjects/jrnl_article.tex
Moving objects: done some writing

@@
 
-\IEEEPARstart{I}{n} the last decade several authors have applied the Capon beamformer (also known as the minimum variance beamformer) for ultrasound imaging \cite{Synnevag2007, Vignon2008, Viola}.
-The adaptive nature of the method has been shown to greatly improve the lateral resolution in ultrasound images, but with a high computational cost.
+\IEEEPARstart{I}{n} the last decade several authors have applied the Capon beamformer (also known as the minimum variance beamformer) for medical ultrasound imaging \cite{Synnevag2007, Vignon2008, Viola}.
+The adaptive nature of the method has been shown to greatly improve the lateral resolution and edge sharpness in ultrasound images, but with a high computational cost.
 In recent work we introduced an implementation of the Capon beamformer capable of processing a stream of images at interactive frame rates \cite{Asen}.
@@
 
-As mentioned, Capon beamforming has been shown to provide improved lateral resolution and edge definitions in ultrasound images \cite{Synnevag2007, Synnevag2009, Chen2011}.
+As mentioned, Capon beamforming has been shown to provide improved lateral resolution and edge sharpness in ultrasound images \cite{Synnevag2007, Synnevag2009, Chen2011}.
 The method accomplishes this by adjusting the array weights based on element-to-element covariance.
@@
 The first two trade resolution for robustness and temporal smoothing helps to preserve speckle statistics for high resolution parameters \cite{Synnevag2007a}.
-A steering vector uncertainty sets helps to reduce signal-nulling which appears if the signal of interest is present during the covariance estimation process and if this estimate is later used with a steering vector that does not match the signal propagation vector exactly.
+However, the issue of having a mismatch between the assumed steering vector and the signal propagation vector has not been addressed directly.
+
+A steering vector uncertainty set helps to reduce signal-nulling which appears if the signal of interest is present during the covariance estimation process and if this estimate is later used with a steering vector that does not match the signal propagation vector exactly.
 This self-nulling is actually how the Capon beamformer obtains its super resolution capabilities, but the signal can also get lost in the process.
 In \cite{JianLi2003} it was shown that most methods based on steering vector uncertainty sets are equivalent to diagonal loading where the loading factor depends on the steering vector uncertainty.
-In this paper we will make use of a set of steering vectors to reduce this self-nulling effect by means of over sampling.
+In that way they make up for a given uncertainty by making the Capon beam wider.
+Hence, the maximal achievable resolution is decreased if the initial steering vector spacing was too coarse.
+In medical ultrasound imaging this initial set of steering vectors is govern by the number of receive beams.
+In this paper we will make use of an additional set of steering vectors to reduce the self-nulling effect by means of over sampling.
+The maximal achievable resolution is then maintained without the presence of severe self-nulling.
 
@@
 In Fig.\,\ref{fig:das_capon_beams} we have plotted the two-way response for DAS and the Capon beamformer using (26) and (32) in \cite{Cox1973}.
-A 64-element linear array with $\lambda/2$ element spacing, where $\lambda$ is the wavelength, is scanned over a sector equal to two times the two-way system resolution ($1/64$ radians) using narrow band excitation.
+A 64-element linear array with $\lambda/2$ element spacing, where $\lambda$ is the wavelength, is scanned over a sector equal to two times the required beam spacing ($1/64$ radians) using narrow band excitation.
 Located in the farfield of the sector is an object moving parallel to the array surface.
-The plot shows the scanned output when the object is located at plus and minus half the system resolution.
+The plot shows the scanned output when the object is located at plus and minus half the required beam spacing for DAS (\ref{eq:resolution}).
 No spatially white noise is present.
@@
 A diagonal loading of $1/100$ corresponds to a signal-to-noise ratio (SNR) of 20 dB before array gain is applied.
-The figure shows that a DAS beamformer, with a beamspacing equal to the system resolution, could yield a $1.9$ dB scalloping loss \cite{Harris1978} if the object is moving parallel to the array.
+The figure shows that a DAS beamformer, with the required beam spacing, could yield a $1.9$ dB scalloping loss \cite{Harris1978} if the object is moving parallel to the array.
 However if the Capon beamformer is used with the same beam density the scalloping loss could be as large as $27$ dB.
@@
 \caption{Theoretical drop in output power mid-way between receive beams for delay-and-sum and Capon beamforming ($M=64$, $L=32$, $d=0.01$) applied in a farfield-narrowband setting.
-The scanned output from both beamformers is plotted for two different time instances.
-For the first instances an object is located at an angle equal to half the system resolution, and for the latter the object is moved an angular distance corresponding to the system resolution.
-If the beam density was reduced to one beam per system resolution the maximum scalloping loss is found at zero degrees.}
+%The scanned output from both beamformers is plotted for two different time instances.
+For the first instances an object is located at an angle equal to half the required beam spacing for delay-and-sum, and for the latter the object is moved an angular distance corresponding to the system resolution.
+If the beam density was reduced to one beam per system resolution the maximum scalloping loss is found at zero degrees.
+}
 \label{fig:das_capon_beams}
@@
 \subsection{Capon Beamforming}
-The Capon beamformer produce a weight vector, like the one in (\ref{eq:das}), based on the impinging signal and an optimization criterion. The minimization problem is as follow \cite{Capon1969}:
+The Capon beamformer produce a weight vector, like the one in (\ref{eq:das}), based on the impinging signal and an optimization criterion. The problem is as follow \cite{Capon1969}:
 \begin{align}
@@
 
-The lateral sampling required in an ultrasound image for a given center frequency, $f_c$, and bandwidth, $B$, is given by the combined aperture size on transmit ($D_{tx}$) and receive ($D_{rx}$), and the wavelength of the center frequency, $\lambda_c = c/f_c$, where $c$ is the speed of sound in the body.
+The lateral sampling required in an ultrasound image for a given center frequency, $f_c$, and bandwidth, $B$, is given by the aperture size, $D$, and the wavelength of the center frequency, $\lambda_c = c/f_c$, where $c$ is the speed of sound in the body.
 The required Nyquist beam spacing is then \cite{Hergum2007}:
 \begin{align}
-\Delta < \frac{\lambda_c}{D_{tx} + D_{rx}}\frac{1}{p}, \label{eq:resolution}
+\Delta < \frac{\lambda_c}{pD}, \label{eq:resolution}
 \end{align}
-where the over sampling factor $p$ can be used to reduce folding of frequencies between $f_c$ and $f_c + B/2$. A $p$ factor of one has been used throughout this paper.
+where the over sampling factor $p$ can be used to reduce folding of frequencies between $f_c$ and $f_c + B/2$. A $p$-factor of two has been used throughout this paper.
 
@@
 
-A given imaging system will then be locally laterally shift-invariant if the contours are diagonal and if the amplitude deviation among the maximum points is less than 1 dB.
+A given imaging system will then be locally laterally shift-invariant if the contours are diagonal and if the amplitude variation among the maximum points is less than 1 dB.
 Then there is minimal geometric distortion (low P-MAE) and the scalloping loss will hopefully not be visible for the observer (low A-MAE).
@@
 We therefore conclude that the system is locally laterally shift-invariant.
-In this, and the following LSV-plots based on simulations, the scan sequence always has a beam at zero degrees and the point scatterer is located exactly on this beam at one time instance.
+In this, and the following LSV-plots based on simulations, the scan sequence always has a beam at zero degrees and the point scatterer is located in the center of this beam at one time instance.
 
@@
 In Fig.\,\ref{fig:das_shiftvariant} the beam density is reduced by a factor of two.
-We see how the geometric distortion and scalloping loss is increased to a level where the contours no longer are diagonal and the maximum trace is not constant.
+We see how the geometric distortion and scalloping loss is increased to a level where the contours no longer are diagonal and the maximum trace has larger deviations.
 Hence, the imaging system is no longer locally laterally shift-invariant.
@@
 The signal cancellation caused by the difference between the signal propagation vector and the assumed steering vector is therefore minimal when the point scatterer has zero lateral offset.
-From the plot we observe that the signal cancellation is high for all the other offsets.
+From this plot we observe that the signal cancellation is high for all the other offsets.
 As a consequence the scalloping loss is tremendous, and not far from the predicted value in Fig\,\ref{fig:das_capon_beams}.
@@
 \section{Oversampling Methods}\label{sec:methods}
-We will now discuss different ways of improving the local lateral shift-invariance when the Capon beamformer is used.
+We will now discuss different ways of improving the local lateral shift-invariance when the Capon beamformer is applied.
 From the previous section it should be clear that we somehow need to increase the lateral sampling to make the Capon imaging system locally laterally shift-invariant.
@@
 \subsection{Oversampling on Transmit}
-The straight forward way of obtaining sufficient sampling in an active system is by increasing the number of transmit beams (tx-beams) until the system is locally laterally shift-invariant.
+A straight forward way of obtaining sufficient sampling in an active imaging system is by increasing the number of transmit beams (tx-beams) until the system is locally laterally shift-invariant.
 In Fig.\,\ref{fig:capon_oversampling_a} and Fig.\,\ref{fig:capon_oversampling_b} we have increased the lateral sampling on transmit by a factor of four and ten respectively.
@@
 Then, by reducing the number of tx-beams and relying on multiple rx-beams in order to get sufficient sampling on receive, it is possible to achieve high frame rates in e.g.
-cardiac ultrasound imaging.
-However, PRB also introduces geometrical distortion which to some degree can be compensated for by using synthetic transmit beamforming (STB) \cite{Hergum2007, Denarie2011, Denarie2013}.
+3D cardiac ultrasound imaging.
+However, PRB also introduces geometrical distortion which to some degree can be compensated for by using e.g.
+synthetic transmit beamforming (STB) \cite{Hergum2007, Denarie2011, Denarie2013}.
 In \cite{Rabinovich2013} it is shown that STB works with Capon beamforming.
@@
 In a narrow band setting the beam can make a sweep across the whole sector by using phase rotation.
-This is also possible in broadband applications as long as the phase rotation is limited to a fraction of the wave length (See coarse-fine beamforming in \cite{Thomenius} \comment{find a better citation}).
-It is therefore possible to make a sweep between a given beam and its two nearest neighbours using only phase rotation without introducing to large errors.
+This is also possible in broadband applications as long as the phase rotation is less than one wave length (See coarse-fine beamforming in \cite{Thomenius}).
+It is therefore possible to make a sweep between a given beam and its two nearest neighbours using only phase rotation without introducing large errors.
 Phase rotation is actually what the Capon beamformer use internally to micro-steer away from bright point scatterers and to enhance edges.
@@
 \end{align}
-where $x_i$ is the element position and $\theta$ is swiped from $-\Delta/2$ to $\Delta/2$ where $\Delta$ is as defined in (\ref{eq:resolution}).
-Note that the steering vector is calculated for a $L$ element array centered around the origin, even though the full array has $M$ elements.
+where $x_i$ is the element position and $\theta$ is swept from $-\Delta/2$ to $\Delta/2$ where $\Delta$ is as defined in (\ref{eq:resolution}).
+Note that the steering vector is calculated for a $L$ element subarray centered around the origin, even though the full array has $M$ elements.
 It is important that this subarray has the same element pitch, $x_i - x_{i-1}$, as the full array.
-The steering error introduced by this steering vector for a frequency band $B$ is less than $B/2 \times \Delta/2$.
+%The steering error introduced by this steering vector for a frequency band $B$ is less than $B/2 \times \Delta/2$.
 
@@
 If the equation $\mat{\hat{R}}\vec{a} = \vec{b}$, as in \cite{Asen}, is solved using the Gauss-Jordan algorithm, we can solve the problem for all steering vectors at once by row reducing the matrix $[\mat{\hat{R}},  \vec{a}_{\theta_1}, \vec{a}_{\theta_2}, \cdots \vec{a}_{\theta_{N_\theta}}]$ to the following form $[\mat{I},  \vec{b}_{\theta_1}, \vec{b}_{\theta_2}, \cdots \vec{b}_{\theta_{N_\theta}}]$.
-This has lower complexity than inverting $\mat{\hat{R}}$, $O(L^3)$, as long as $N_\theta < L$.
+This is still an $O(L^3)$ operation.
 
@@
 The scalloping loss is less than 1 dB, and the system is therefore locally laterally shift-invariant.
-As a consequence point scatterers has a consistent appearance when subjected to motion \multimedia{Video DAS}.
+As a consequence point scatterers have a consistent appearance when subjected to motion \multimedia{Video DAS}.
 
@@
 Hence, we get a high amplitude value at this location.
-In almost all the other frames the signal loss is massive.
+In almost all the other frames the signal loss is massive (10 dB).
 The result is blinking point scatterers when subjected to motion \multimedia{Video Capon}.
@@
 It is a result of lack of control when the probe was manually dragged across the phantom.
-In order to get a continues straight diagonal the probe has to be dragged across the phantom with a speed which match the acquisition frame rate.
+In order to get a diagonal without peaks the probe has to be dragged across the phantom with a speed which matches the acquisition frame rate.
 In addition, due to the increase in lateral resolution the Capon beamformer requires smaller frame-to-frame movements than DAS.
@@
 In Section \ref{sec:capon_LLSI} it was shown that a Capon beamformer with high resolution parameters is not locally laterally shift-invariant when subjected to motion.
-The root of cause was found to be a too low lateral sampling density.
+The root cause was found to be a too low lateral sampling density.
 In Section \ref{sec:methods} we therefore presented methods for improved shift-invariance based on oversampling.
@@
 
-Note that in this paper we have investigated a single array configuration and a limited parameter range for the Capon beamformer.
+%Note that in this paper we have investigated a single array configuration and a limited parameter range for the Capon beamformer.
 The goal was not to make a reference for which oversampling factor to use in different scenarios, but rather do a detailed investigation of the scalloping loss effect.
@@
 
-If the scene contains bright point scatterers or sharp edges we recommend to measure the ratio between highest peak and the neighbouring background level.
-In his seminal paper \cite{Cox1973} Cox derives formulas for how large the oversampling has to be for a given output signal-to-noise ratio (SNR).
-E.g 47 dB is suppose to result in a ten time improvement in resolution.
-Be aware that Cox measured resolution as the saddle point between two impinging plane waves, which typically shows smaller scalloping loss than a single moving point scatterer.
-In our example more than ten times oversampling was found to be required for a 40 dB output SNR in simulations, however \textit{in-vitro} ten times oversampling was enough for the system to be locally laterally shift-invariant.
-Possibly due to a higher initial noise level before diagonal loading was applied.
-
-Relating the presented phase rotation method to earlier work we find that it is similar to a SVUS.
-However, where SVUSs typically has a criteria for selecting the best output after applying all the steering vectors contained in the set, oversampling with phase rotation outputs all.
-It can also be though of as a narrow band approach to PRB where Capon is applied on all receive beams.
-
-Oversampling by phase rotation can therefore introduce similar block artifacts as PRB, but on a smaller scale when PRB is used to lower the number of tx-beams.
-These artifacts could be removed in the same way as for PRB if we distribute our steering angles as beams are spread out for the STB technique.
+If faced with an imaging scene containing bright point scatterers or sharp edges, one could wonder how large the oversampling factor needs to be in order to avoid visible scalloping losses.
+In his seminal paper, Cox \cite{Cox1973} derives formulas for the maximal achievable resolution for a given output SNR.
+For instance 47 dB is found to give a ten time improvement in resolution.
+We would therefore recommend to first use the dynamic range in the image as guidance.
+The amount of diagonal loading will also dictate the maximal achievable resolution since $10\log{d}$ can be interpreted as the input SNR seen by the Capon beamformer.
+Cox measures resolution as the 3 dB saddle point between two impinging monochromatic plane waves.
+For ultrasound data only a 50 \% increase in resolution was found in \cite[Fig.~7]{Asen} for a 30 dB output SNR when measured in a similar way as Cox (6 dB saddle point).
+However, sixteen times oversampling was required to make the imaging system shift invariant.
+In this paper more than ten times oversampling was found to be required for a 40 dB output SNR (after array gain) in simulations, however for the \textit{in-vitro} data ten times oversampling was enough for the system to be locally laterally shift-invariant.
+This might be a result of improved initial sampling (as discussed in Section \ref{sec:res}) and a higher initial noise level before diagonal loading was applied.
+
+Relating the presented phase rotation method to earlier work we find that it is similar to a steering vector uncertainty set.
+However, where these set-based methods typically has a criterion for selecting the best output after applying all the steering vectors contained in the set, oversampling with phase rotation outputs all.
+It can also be thought of as a narrow band approach to parallel receive beamforming where Capon is applied on all receive beams.
+
+Oversampling by phase rotation can therefore introduce similar block artifacts as parallel receive beamforming, but on a smaller scale when the technique is used to lower the number of tx-beams.
+These artifacts could be removed in the same way as for parallel receive beamforming if we distribute our steering angles as beams are spread out for the synthetic transmit beamforming technique (STB).
 We then get a phase rotation version of STB where e.g.
@@
 With the Capon beamformer this is now changed.
-A more symmetric point spread function is obtained with the parameters $L=M/4$ and $K=0$.
+A more symmetric point spread function while preserving speckle statistics is obtained with the parameters $L=M/4$ and $K=0$ \cite{Synnevag2007a}.
 Another option is to perform pulse inversion if a spiky symmetric point spread function is desirable.
@@
 Oversampling is not only essential to keep the amplitude level of bright point scatterers it also helps to improve contrast and edge definitions.
-If oversampling is not applied homogeneous speckle regions could get a too low average value and the edge delineation of an moving object will fluctuate in the same way as a moving point scatterer would have sparkled.
+If oversampling is not applied homogeneous speckle regions could get a too low average value and the edge delineation of a moving object will fluctuate in the same way as a moving point scatterer would have sparkled.
 
@@
 \cite{Synnevag2011, Nilsen2009, Mehdizadeh2012a, Kim}, would also need oversampling when higher lateral resolution is obtained.
-The same applies for high resolution spectral estimation \cite{Ekroll2012}.
+The same applies for high resolution spectral estimation \cite{Ekroll2012}, but here it is much easier to obtain sufficient sampling initially.
 
@@
 
-The authors would like to thank Anders S\o{}rnes for the help with acquiring \textit{in-vitro} data. \comment{Ask Anders if hi wants to be mentioned.}
+The authors would like to thank Anders S\o{}rnes at GE Vingmed Ultrasound, Horten, Norway for the help with acquiring \textit{in-vitro} data.
 

### 2013-12-12 HRUIofMovingObjects/jrnl_article.tex
Moving objects: done some writing. Updated figures.

@@
 Increased lateral resolution should demand denser lateral sampling.
-In previous literature on Capon beamforming for medical ultrasound imaging however only single frame scenarios have been simulated.
+However, in previous literature on Capon beamforming for medical ultrasound imaging only single frame scenarios have been simulated.
 Temporal behaviour and effects caused by the increased resolution and lack of oversampling have therefore been neglected.
@@
 \IEEEPARstart{I}{n} the last decade several authors have applied the Capon beamformer (also known as the minimum variance beamformer) for medical ultrasound imaging \cite{Synnevag2007, Vignon2008, Viola}.
-The adaptive nature of the method has been shown to greatly improve the lateral resolution and edge sharpness in ultrasound images, but with a high computational cost.
-In recent work we introduced an implementation of the Capon beamformer capable of processing a stream of images at interactive frame rates \cite{Asen}.
-It is therefore an interesting question how a moving object appears after Capon beamforming has been applied.
+The method has been shown to improve the lateral resolution and edge sharpness in ultrasound images, but at an increased computational expense.
+In recent work we introduced an implementation of the Capon beamformer capable of processing ultrasound images at interactive frame rates \cite{Asen}.
+Now it is therefore an interesting question how a moving object appears after Capon beamforming has been applied.
 Earlier work on Capon beamforming for medical ultrasound imaging has investigated the method's capabilities in single frames only, and temporal behaviour and effects has therefore been neglected.
@@
 The first two trade resolution for robustness and temporal smoothing helps to preserve speckle statistics for high resolution parameters \cite{Synnevag2007a}.
-However, the issue of having a mismatch between the assumed steering vector and the signal propagation vector has not been addressed directly.
-
-A steering vector uncertainty set helps to reduce signal-nulling which appears if the signal of interest is present during the covariance estimation process and if this estimate is later used with a steering vector that does not match the signal propagation vector exactly.
+However, the issue of having a mismatch between the assumed steering vector and the signal propagation vector has not been addressed directly for medical ultrasound imaging.
+
+A steering vector uncertainty set has been shown to reduce signal-nulling which appears if the signal of interest is present during the covariance estimation process and if this estimate is later used with a steering vector that does not match the signal propagation vector exactly.
 This self-nulling is actually how the Capon beamformer obtains its super resolution capabilities, but the signal can also get lost in the process.
@@
 In medical ultrasound imaging this initial set of steering vectors is govern by the number of receive beams.
-In this paper we will make use of an additional set of steering vectors to reduce the self-nulling effect by means of over sampling.
+In this paper we make use of an additional set of steering vectors to reduce the self-nulling effect by means of over sampling.
 The maximal achievable resolution is then maintained without the presence of severe self-nulling.
@@
 In \cite{Cox1973}, Cox derives formulas for the resolution of the Capon beamformer, and presents in great detail how the amount of mismatch between the selected steering vector and the signal propagation vector impacts the beamformer output in simple situations.
-Cox states the following important fact about the Capon beamformer (the $\vec{k}_3$ processor):
+Cox states the following important fact about the Capon beamformer [the $\vec{k}_3$ processor]:
 \begin{quote}
@@
 Note that one matrix has to be constructed and inverted for each data vector $\vec{x}[n]$ received by the system.
-This is indeed computational demanding, but following the innovation in GPU computing, and by applying a novel method in order to reduce the size of the matrix inversion problem \cite{Nilsen2009}, real time processing is now possible \cite{Asen}.
-From (\ref{eq:capon_weights}) it is clear that complex data can lead to complex weights.
-The weights can therefore have an asymmetric K-space response which basically means that the main lobe and zeros does not have to be symmetric around the origin.
-The beam can therefore be micro-steered by phase rotation.
+%This is indeed computational demanding, but following the innovation in GPU computing, and by applying a novel method in order to reduce the size of the matrix inversion problem \cite{Nilsen2009}, real time processing is now possible \cite{Asen}.
+From (\ref{eq:capon_weights}) it is also clear that complex data [e.g.
+in-phase and quadrature data] can lead to complex weights.
+The weights can therefore have an asymmetric K-space response with a shifted main lobe and zeros at arbitrary positions.
+%The beam can therefore be micro-steered by phase rotation.
 
@@
 
-Equation (\ref{eq:resolution}) was used to calculate the object position or beam spacing in Fig.\,\ref{fig:das_capon_beams}.
+Equation (\ref{eq:resolution}) was used to calculate the object position or required beam spacing in Fig.\,\ref{fig:das_capon_beams}.
 Therefore, if (\ref{eq:resolution}) is not fulfilled the scalloping loss could become visible in the image and the system will no longer be locally laterally shift-invariant.
@@
 This is approximately equal to the visibility threshold (Weber fraction of 2\%) for a grayscale image, and a loss larger than 1 dB could therefore end up being visible to the observer.
-In our further analysis a system would be regarded as locally laterally shift-invariant if the scalloping loss is less than 1 dB.
-However, what eventually matters is whether the loss is visible in the image or not.
+%However, what eventually matters is whether the loss is visible in the image or not.%In our further analysis a system would be regarded as locally laterally shift-invariant if the scalloping loss is less than 1 dB.
 %However, depending on viewing light conditions and e.g.
@@
 For each lateral position the root-mean-square (RMS) of the image data is calculated in range, and the resulting beam profiles are stacked in to an image.
-For a given scan sequence there are twice as many point positions than there are receive lines.
+%For a given scan sequence there are twice as many point positions than there are receive lines.
 The LSV-plots in this paper are displayed as a contour plot with contours on -1, -2, -3, -6, -12 and -24 dB.
 The maximum for each beam profile is also marked with a black dot, and the amplitude variation among these points is presented in a subplot next to the LSV-plot.
-The LSV-plots based on simulations has been up-interpolated to 128 samples using cubic interpolation before presentation.
-
-To further quantify the shift-invariance we have calculated the mean absolute error (MAE) of the maximum point position in each beam profile versus a straight line ($f(x)=x$).
+The LSV-plots based on simulations has been up-interpolated to 128 samples in each direction using cubic interpolation if the number of receive beams is less than 128.
+
+To further quantify the shift-invariance we have calculated the mean absolute error (MAE) between the maximum point position in each beam profile and a straight line ($f(x)=x$).
 We will refer to this measure as P-MAE.
@@
 A given imaging system will then be locally laterally shift-invariant if the contours are diagonal and if the amplitude variation among the maximum points is less than 1 dB.
-Then there is minimal geometric distortion (low P-MAE) and the scalloping loss will hopefully not be visible for the observer (low A-MAE).
+Then there is minimal geometric distortion (low P-MAE) and the scalloping loss will not be visible for the observer (low A-MAE).
 %Since both effects can occur at the same time it is important to have a measure not effected by geometric distortion.
@@
 The point scatterer response has been simulated using Field II \cite{Jensen1992, Jensen1996a} with a probe setup equal to the M5S probe by GE Vingmed Ultrasound, Horten, Norway.
-The probe was excited with a 2.5 MHz and 1.5 cycles long pulse with a $\text{f-number}=3$ and corresponding cosine apodization on receive.
+The probe was excited with a 2.5 MHz and 1.5 cycles long pulse.
 The transmit focus in azimuth and elevation was configured to be equal and the point scatterer was moved through this focus.
-The lateral sampling density has been doubled prior to detection through coherent interpolation.
 We see that the contours in Fig.\,\ref{fig:das_shiftinvariant} are approximately diagonal (low P-MAE) and that the amplitude variation among the maximum points is minimal (low A-MAE).
@@
 
-The scalloping loss in Fig.\,\ref{fig:das_shiftinvariant} is smaller than what was predicted by Fig.\,\ref{fig:das_capon_beams}.
+%The scalloping loss in Fig.\,\ref{fig:das_shiftinvariant} is smaller than what was predicted by Fig.\,\ref{fig:das_capon_beams}.
 To understand why note that Fig.\,\ref{fig:das_capon_beams} shows the scanned response for a narrowband system while the imaging system in Fig.\,\ref{fig:das_shiftinvariant} uses a broadband pulse.
@@
 As mentioned, the beam at zero degrees is the only one where the point scatterer is located in the center of the beam.
-The signal cancellation caused by the difference between the signal propagation vector and the assumed steering vector is therefore minimal when the point scatterer has zero lateral offset.
+The signal cancellation caused by the difference between the signal propagation vector and the assumed steering vector is therefore minimal when the point scatterer is located at this position.
 From this plot we observe that the signal cancellation is high for all the other offsets.
@@
 The diagonal factor $d$ and the effective adaptive aperture size $L$ can be used to control the mainlobe width of the Capon beamformer.
-In both extreme cases, $d\gg1$ and $L=1$, the Capon beamformer becomes either a triangle or uniform-weighted DAS beamformer respectively, and a beamspacing according to (\ref{eq:resolution}) would then be sufficient.
+In both extreme cases, $d\gg1$ and $L=1$, the Capon beamformer becomes either a triangle or uniform-weighted DAS beamformer respectively, and a beam spacing according to (\ref{eq:resolution}) would then be sufficient.
 In Fig.\,\ref{fig:capon2} we show how the scalloping loss is reduced when the amount of diagonal loading is increased.
@@
 \section{Oversampling Methods}\label{sec:methods}
-We will now discuss different ways of improving the local lateral shift-invariance when the Capon beamformer is applied.
+In this section we will discuss different methods aimed at improving the local lateral shift-invariance for the Capon beamformer.
 From the previous section it should be clear that we somehow need to increase the lateral sampling to make the Capon imaging system locally laterally shift-invariant.
@@
 synthetic transmit beamforming (STB) \cite{Hergum2007, Denarie2011, Denarie2013}.
-In \cite{Rabinovich2013} it is shown that STB works with Capon beamforming.
+In \cite{Rabinovich2013} it is shown that STB works well with Capon beamforming.
 In \cite{Asen} we also used STB in order to increase the beam density of a cardiac ultrasound recording.
@@
 However, the steering vector can also, as in narrow band applications, be varied over a set of pre-defined vectors.
-In a narrow band setting the beam can make a sweep across the whole sector by using phase rotation.
-This is also possible in broadband applications as long as the phase rotation is less than one wave length (See coarse-fine beamforming in \cite{Thomenius}).
-It is therefore possible to make a sweep between a given beam and its two nearest neighbours using only phase rotation without introducing large errors.
+In a narrow band setting the beam can make a sweep across the whole imaging sector by using phase rotation.
+This is also possible in broadband applications as long as the phase rotation is less than one pulse length (See coarse-fine beamforming in \cite{Thomenius}).
+It is therefore possible to make a sweep between a given beam and its two nearest neighbours using phase rotation without introducing large errors.
 Phase rotation is actually what the Capon beamformer use internally to micro-steer away from bright point scatterers and to enhance edges.
@@
 where $x_i$ is the element position and $\theta$ is swept from $-\Delta/2$ to $\Delta/2$ where $\Delta$ is as defined in (\ref{eq:resolution}).
-Note that the steering vector is calculated for a $L$ element subarray centered around the origin, even though the full array has $M$ elements.
-It is important that this subarray has the same element pitch, $x_i - x_{i-1}$, as the full array.
+Note that the steering vector must be calculated for a $L$ element symmetric subarray, even though the full array has $M$ elements.
+This subarray should have the same element pitch, $x_i - x_{i-1}$, as the full array.
 %The steering error introduced by this steering vector for a frequency band $B$ is less than $B/2 \times \Delta/2$.
@@
 The steering angles can be spread out in many different ways.
-If the number of steering angles is $N_\theta$ we have selected a uniform distribution of $\theta$  with $N_\theta/2$ angles on each side of the original steering direction if $N_\theta$ is even, and $(N_\theta-1)/2$ if $N_\theta$ is odd.
-
-Since the phase rotation method operates on the same data as before, there will be no reduction in the acquisition frame rate by applying this method.
+If the number of steering angles is $N_\theta$ we have selected a uniform distribution of $\theta$  with $N_\theta/2$ angles on each side of the original steering direction when $N_\theta$ is even.%, and $(N_\theta-1)/2$ [including the original steering direction] if $N_\theta$ is odd.
+
+Since the phase rotation method operates on the same channel data as in Fig.\,\ref{fig:capon1}, there will be no reduction in the acquisition frame rate by applying this method.
 In addition, since the same number of covariance matrices are constructed and inverted, there is no extra cost involved here as well.
@@
 The transmit beam density was set to match (\ref{eq:resolution}) based on this frequency.
-Note that in harmonic imaging the generated transmit beam at the harmonic frequency is a weighted version of the beam resulting from transmitting at the same frequency \cite{Fedewa2004}.
-The lateral sampling is therefore better than what was used in simulations.
+Note that in harmonic imaging the generated transmit beam at the harmonic frequency will be wider and have lower sidelobes than the beam resulting from transmitting at the harmonic frequency \cite{Fedewa2004}.
+The lateral sampling is therefore increased compared with our simulations.
 A tissue mimicking phantom containing wire targets and anechoic cysts was imaged, and the data was acquired while the probe was moved across the phantom manually.
@@
 In Fig\,\ref{fig:image_das} one point scatterer has been selected (white box), and then used to generate LSV-plots.
-\comment{Todo: Add lateral coherent interpolation as done for the simulations?
-As for now speckle is not fully resolved without it.} Figure \ref{fig:phantom_das} shows the resulting LSV-plot when using the DAS beamformer.
+Figure \ref{fig:phantom_das} shows the resulting LSV-plot when using the DAS beamformer.
 The scalloping loss is less than 1 dB, and the system is therefore locally laterally shift-invariant.
@@
 
-Figure \ref{fig:phantom_capon_LSV} shows the same for the Capon beamformer without oversampling.
-In the frame where the point scatterer is at -6 degrees the receive beam direction is very close to the signal propagation vector.
+Figure \ref{fig:phantom_capon_LSV} shows the same point scatterer response processed with the Capon beamformer without oversampling.
+For the frame where the point scatterer is at -6 degrees the receive beam direction is very close to the signal propagation vector.
 Hence, we get a high amplitude value at this location.
@@
 In order to get a diagonal without peaks the probe has to be dragged across the phantom with a speed which matches the acquisition frame rate.
-In addition, due to the increase in lateral resolution the Capon beamformer requires smaller frame-to-frame movements than DAS.
+In addition, due to the increase in lateral resolution, the Capon beamformer requires smaller frame-to-frame movements than DAS.
 Looking at the video of Capon beamforming with ten times oversampling by phase rotation \multimedia{Capon + 10PR} there is no visible blinking anymore.
@@
 In Section \ref{sec:capon_LLSI} it was shown that a Capon beamformer with high resolution parameters is not locally laterally shift-invariant when subjected to motion.
-The root cause was found to be a too low lateral sampling density.
+The root cause was found to be too coarse lateral sampling.
 In Section \ref{sec:methods} we therefore presented methods for improved shift-invariance based on oversampling.
-The most promising method, with respect to improved shift-invariance, reduction in acquisition frame rate and processing complexity, was based on phase rotation.
-In Section \ref{sec:res} we presented results showing that this method also works on \textit{in-vitro} data when the probe is subjected to motion.
+The most promising method, with respect to improved shift-invariance, reduction in acquisition frame rate and processing complexity combined, was based on phase rotation.
+In Section \ref{sec:res} we presented results showing that this method also works on \textit{in-vitro} data when the ultrasound probe is subjected to motion.
+In simulations, oversampling based on phase rotation was found to reduce the scalloping loss by more than 20 dB.
+On \textit{in-vitro} data the reduction was more than 7 dB.
 
@@
 For ultrasound data only a 50 \% increase in resolution was found in \cite[Fig.~7]{Asen} for a 30 dB output SNR when measured in a similar way as Cox (6 dB saddle point).
-However, sixteen times oversampling was required to make the imaging system shift invariant.
+However, sixteen times oversampling was required to make the imaging system shift invariant in the same paper.
 In this paper more than ten times oversampling was found to be required for a 40 dB output SNR (after array gain) in simulations, however for the \textit{in-vitro} data ten times oversampling was enough for the system to be locally laterally shift-invariant.
@@
 
-Oversampling by phase rotation can therefore introduce similar block artifacts as parallel receive beamforming, but on a smaller scale when the technique is used to lower the number of tx-beams.
-These artifacts could be removed in the same way as for parallel receive beamforming if we distribute our steering angles as beams are spread out for the synthetic transmit beamforming technique (STB).
-We then get a phase rotation version of STB where e.g.
+Oversampling by phase rotation can therefore also introduce similar block artifacts as parallel receive beamforming.
+These artifacts should be possible to remove in the same way as for parallel receive beamforming by using a synthetic transmit beamforming (STB) approach.
+If we distribute our steering angles in an overlapping scheme [like rx-beams are spread out for STB] we get a phase rotation version of STB where e.g.
 eight phase rotated beams are averaged into four.
@@
 
-Oversampling is not only essential to keep the amplitude level of bright point scatterers it also helps to improve contrast and edge definitions.
+Oversampling is not only essential to keep the amplitude level of bright point scatterers, it also helps to improve contrast and edge definitions.
 If oversampling is not applied homogeneous speckle regions could get a too low average value and the edge delineation of a moving object will fluctuate in the same way as a moving point scatterer would have sparkled.

### 2013-12-14 HRUIofMovingObjects/jrnl_article.tex
Moving objects: updated based on comments from Sverre

@@
 %\boldmath
-If an ultrasound imaging system provides a presentation of a moving object which is sensitive to local spatial shifts, the system is said to be spatially locally shift-variant.
+If an ultrasound imaging system provides a presentation of a moving object which is sensitive to small spatial shifts, the system is said to be locally spatially shift-variant.
 This can for instance happen if the axial or lateral sampling is insufficient.
@@
 \IEEEPARstart{I}{n} the last decade several authors have applied the Capon beamformer (also known as the minimum variance beamformer) for medical ultrasound imaging \cite{Synnevag2007, Vignon2008, Viola}.
-The method has been shown to improve the lateral resolution and edge sharpness in ultrasound images, but at an increased computational expense.
-In recent work we introduced an implementation of the Capon beamformer capable of processing ultrasound images at interactive frame rates \cite{Asen}.
-Now it is therefore an interesting question how a moving object appears after Capon beamforming has been applied.
-Earlier work on Capon beamforming for medical ultrasound imaging has investigated the method's capabilities in single frames only, and temporal behaviour and effects has therefore been neglected.
-
-An important property of an ultrasound imaging system is its local spatial shift-invariance.
-Local spatial shift-invariance means that an object's appearance should be preserved when the object, or the ultrasound probe, is moved slightly.
-This invariance to local movements is what makes ultrasound imaging feasible in general, but it is specially important for quantitative techniques that correlate time-varying features, like pulsed doppler and flow imaging, and speckle tracking of tissue and blood.
-In this paper we analyze the lateral shift-invariance property of the Capon beamformer by studying the beamformer output when applied on simulated and \textit{in vitro} images of moving objects.
-
-As mentioned, Capon beamforming has been shown to provide improved lateral resolution and edge sharpness in ultrasound images \cite{Synnevag2007, Synnevag2009, Chen2011}.
-The method accomplishes this by adjusting the array weights based on element-to-element covariance.
-Hence, the method yields data dependent weights.
-Capon beamforming is however also known for its suboptimal performance if e.g.
-the sample size available for covariance estimation is limited \cite{Mestre2006}, if correlated signals are present \cite{Widrow1982}, and if the assumed steering vector does not match the signal propagation vector \cite{Wax1996, Wax1996a}.
+The method has been shown to improve the lateral resolution and edge sharpness \cite{Synnevag2007, Synnevag2009, Chen2011}.
+This is accomplished by adaptively adjusting the array weights based on element-to-element covariance in order to minimize interference.
+The method is, however, also highly sensitive to model errors \cite{Mestre2006, Widrow1982, Wax1996, Wax1996a}.
 This has led to the development of several robust versions of the Capon beamformer with e.g.
@@
 The first two trade resolution for robustness and temporal smoothing helps to preserve speckle statistics for high resolution parameters \cite{Synnevag2007a}.
-However, the issue of having a mismatch between the assumed steering vector and the signal propagation vector has not been addressed directly for medical ultrasound imaging.
-
-A steering vector uncertainty set has been shown to reduce signal-nulling which appears if the signal of interest is present during the covariance estimation process and if this estimate is later used with a steering vector that does not match the signal propagation vector exactly.
+Steering vector uncertainty sets, however, have not been applied for medical ultrasound imaging.
+
+% but at an increased computational expense \cite{So2011}.
+
+%In recent work we introduced an implementation of the Capon beamformer capable of processing ultrasound images at interactive frame rates \cite{Asen}.
+In the same paper we discovered that oversampling had to be applied to make the imaging system locally spatially shift-invariant.
+%It is therefore now interesting how a moving object appears after Capon beamforming has been applied.
+%Earlier work on Capon beamforming for medical ultrasound imaging has investigated the method's capabilities in single frames only, and temporal behaviour and effects have therefore been neglected.
+%
+
+A steering vector uncertainty set reduce signal-nulling which appears if the signal of interest is present during the covariance estimation process and if this estimate is later used with a steering vector that does not match the signal propagation vector exactly.
 This self-nulling is actually how the Capon beamformer obtains its super resolution capabilities, but the signal can also get lost in the process.
@@
 Hence, the maximal achievable resolution is decreased if the initial steering vector spacing was too coarse.
-In medical ultrasound imaging this initial set of steering vectors is govern by the number of receive beams.
-In this paper we make use of an additional set of steering vectors to reduce the self-nulling effect by means of over sampling.
+In medical ultrasound imaging this initial set of steering vectors is governed by the number of receive beams.
+In this paper we will make use of an additional set of steering vectors to reduce the self-nulling effect by means of oversampling.
 The maximal achievable resolution is then maintained without the presence of severe self-nulling.
 
-In \cite{Cox1973}, Cox derives formulas for the resolution of the Capon beamformer, and presents in great detail how the amount of mismatch between the selected steering vector and the signal propagation vector impacts the beamformer output in simple situations.
-Cox states the following important fact about the Capon beamformer [the $\vec{k}_3$ processor]:
+In \cite{Cox1973}, Cox presents in great detail how the amount of mismatch between the selected steering vector and the signal propagation vector impacts the beamformer output in simple situations.
+He also states the following important fact about the Capon beamformer [the $\vec{k}_3$ processor]:
 \begin{quote}
@@
 In Fig.\,\ref{fig:das_capon_beams} we have plotted the two-way response for DAS and the Capon beamformer using (26) and (32) in \cite{Cox1973}.
-A 64-element linear array with $\lambda/2$ element spacing, where $\lambda$ is the wavelength, is scanned over a sector equal to two times the required beam spacing ($1/64$ radians) using narrow band excitation.
+A 64-element linear array with $\lambda/2$ element spacing, where $\lambda$ is the wavelength, is scanned over a sector equal to two times the required beam spacing using narrow band excitation.
 Located in the farfield of the sector is an object moving parallel to the array surface.
-The plot shows the scanned output when the object is located at plus and minus half the required beam spacing for DAS (\ref{eq:resolution}).
-No spatially white noise is present.
+The plot shows the scanned output when the object is located at plus and minus half the required beam spacing for DAS [See (\ref{eq:resolution})].
+%No spatially white noise is present.
 Hence, the interference-noise matrix ($\sigma_0^2\mat{Q}$ in \cite{Cox1973}) is the identity.
@@
 The figure shows that a DAS beamformer, with the required beam spacing, could yield a $1.9$ dB scalloping loss \cite{Harris1978} if the object is moving parallel to the array.
-However if the Capon beamformer is used with the same beam density the scalloping loss could be as large as $27$ dB.
-A $1.9$ dB scalloping loss for the Capon beamforming is obtained for this specific example when the beam density is increased with a factor of 25.
+However if the Capon beamformer is used with the same beam density the loss could be as large as $27$ dB.
+This imaging system is then highly sensitive to small spatial shifts.
+A $1.9$ dB scalloping loss for the Capon beamforming is obtained for this specific example when the beam spacing or object distance is decreased with a factor of 25.
 
@@
 
-In previous literature on Capon beamforming for medical ultrasound imaging (with single frame simulations) this effect has been passed over in silence by either oversampling on transmit, or by carefully position point scatterers exactly on transmit-receive beam pairs, or by ignorance.
+In previous literature on Capon beamforming for medical ultrasound imaging only single frame scenarios have been simulated.
+The scalloping loss effect has therefore been passed over in silence by either oversampling on transmit, by carefully positioning point scatterers exactly on transmit-receive beam pairs, or by ignorance.
 In our first work where Capon beamforming was applied on a loop of images containing moving objects \cite{Asen2012, Asen} we commented on the effect, but no solution was presented other than oversampling on transmit when simulating.
@@
 In this paper we investigate the issue further.
-We show that when a moving point scatterer is imaged with the Capon beamformer with the same beam spacing sufficient for DAS, the predicted scalloping loss is observed.
+We show that when a moving point scatterer is imaged with the Capon beamformer with the same beam spacing sufficient for DAS, the predicted scalloping loss is observed both in simulations and \textit{in-vitro} data.
 The question is then, what can we do in order to improve on the situation without affecting the imaging frame rate and processing throughout.
-We try two answer this by studying the scalloping loss of the Capon beamforming and investigate different methods aimed at reducing it.
-We also discuss how the different methods will affect the imaging frame rate and the processing complexity.
-
-In the next section we give an introduction to Capon beamforming for medical ultrasound imaging together with a short introduction to lateral sampling and spatial shift-invariance.
+We try two answer this by studying the scalloping loss of the Capon beamforming and investigating different methods aimed at reducing it.
+We also discuss how the different methods will affect the acquisition frame rate and the processing complexity.
+
+In the next section we give an introduction to Capon beamforming for medical ultrasound imaging together with a short introduction to lateral sampling and shift-invariance.
 In Section \ref{sec:capon_LLSI} the local lateral shift-invariance of the Capon beamformer is investigated, before different methods for improved local lateral shift-invariance are presented in Section \ref{sec:methods}.
@@
 \subsection{Standard Beamforming}
-The standard form of time domain array beamforming is the delay-and-sum beamformer (DAS) where the output $z$ is calculated as:
+The standard form of time domain array beamforming is the DAS beamformer where the output $z$ is calculated as:
 \begin{align}\label{eq:das}
@@
 \end{align}
-where $K = M-L+1$ typically is proportional to the pulse length, $N_K = 2K + 1$, $N_L = M-L+1$, and $\vec{x}_l$ is the $l\text{th}$ subarray $[x_l[n], \dotso x_{l+L}[n]]$ of length $L$.
+where $K = M-L+1$ typically is proportional to the pulse length, $N_K = 2K + 1$, $N_L = M-L+1$, and $\vec{x}_l$ is a $L$ long subarray $[x_l[n], \dotso x_{l+L}[n]]$.
 Finally the matrix is loaded with a diagonal factor for numerical stability and increased robustness,
@@
 
-The lateral sampling required in an ultrasound image for a given center frequency, $f_c$, and bandwidth, $B$, is given by the aperture size, $D$, and the wavelength of the center frequency, $\lambda_c = c/f_c$, where $c$ is the speed of sound in the body.
+The lateral sampling required in an ultrasound image for a given center frequency, $f_c$, and bandwidth, $B$, is given by the aperture size, $D$, and the wavelength of the center frequency, $\lambda_c = c/f_c$, where $c$ is the speed of sound.
 The required Nyquist beam spacing is then \cite{Hergum2007}:
 \begin{align}
-\Delta < \frac{\lambda_c}{pD}, \label{eq:resolution}
+\Delta < \frac{\lambda_c}{qpD}, \label{eq:resolution}
 \end{align}
-where the over sampling factor $p$ can be used to reduce folding of frequencies between $f_c$ and $f_c + B/2$. A $p$-factor of two has been used throughout this paper.
-
-Equation (\ref{eq:resolution}) was used to calculate the object position or required beam spacing in Fig.\,\ref{fig:das_capon_beams}.
-Therefore, if (\ref{eq:resolution}) is not fulfilled the scalloping loss could become visible in the image and the system will no longer be locally laterally shift-invariant.
-In an ultrasound image with 256 gray levels and 50 dB dynamic range, a 1 dB loss corresponds to 5 gray levels.
-This is approximately equal to the visibility threshold (Weber fraction of 2\%) for a grayscale image, and a loss larger than 1 dB could therefore end up being visible to the observer.
+where the oversampling factor $p$ is applied to reduce folding of frequencies between $f_c$ and $f_c + B/2$.
+A $p$-factor of two has been used throughout this paper.
+The oversampling factor $q$ will be used to oversample prior to Capon beamforming, and is one if not stated otherwise.
+
+Equation (\ref{eq:resolution}) was used to calculate the object positions in Fig.\,\ref{fig:das_capon_beams}.
+Therefore, if (\ref{eq:resolution}) is not fulfilled, the resulting scalloping loss for the DAS beamformer could become visible in the image.
+The system output will then be sensitive to small spatial shifts.
+We say that the system is locally spatially shift-variant.
+Since Capon beamforming improves the lateral resolution we will focus on sensitivity to small lateral shifts, or in other words the local lateral shift-invariance property of the imaging system.%This invariance to local movements is what makes ultrasound imaging feasible in general, but it is specially important for quantitative techniques, like speckle tracking of tissue and blood, that correlate time-varying features.
+%In this paper we analyze the lateral shift-invariance property of the Capon beamformer by studying the beamformer output when applied on simulated and \textit{in vitro} images of moving objects.
+
+In an ultrasound image with 50 dB dynamic range mapped to 256 gray levels, a 1 dB loss corresponds to 5 gray levels.
+This is approximately equal to the visibility threshold [Weber fraction of 2\%] for gray scale images.
+A loss larger than 1 dB could therefore end up being visible to the observer.
 %However, what eventually matters is whether the loss is visible in the image or not.%In our further analysis a system would be regarded as locally laterally shift-invariant if the scalloping loss is less than 1 dB.
@@
 
-In order to visualize the local lateral shift-invariance of a given imaging system, we will make use of lateral shift-variance plots (LSV-plots), introduced by Hergum \textit{et al.} \cite{Hergum2007}.
+In order to visualize the local lateral shift-invariance of a given imaging system, we will make use of lateral shift-variance plots (LSV-plots) \cite{Hergum2007}.
 A LSV-plot is constructed for a given imaging system by imaging a point scatterer moving laterally at constant speed.
@@
 
-In Fig.\,\ref{fig:das_shiftvariant} the beam density is reduced by a factor of two.
-We see how the geometric distortion and scalloping loss is increased to a level where the contours no longer are diagonal and the maximum trace has larger deviations.
+In Fig.\,\ref{fig:das_shiftvariant} the beam density is reduced by a factor of two compared to Fig.\,\ref{fig:das_shiftinvariant}.
+We see how the geometric distortion increases to a level where the contours no longer are diagonal.
+The maximum scalloping loss is also larger than 1 dB.
 Hence, the imaging system is no longer locally laterally shift-invariant.
@@
 A straight forward way of obtaining sufficient sampling in an active imaging system is by increasing the number of transmit beams (tx-beams) until the system is locally laterally shift-invariant.
-In Fig.\,\ref{fig:capon_oversampling_a} and Fig.\,\ref{fig:capon_oversampling_b} we have increased the lateral sampling on transmit by a factor of four and ten respectively.
+In Fig.\,\ref{fig:capon_oversampling_a} and Fig.\,\ref{fig:capon_oversampling_b} we have increased the lateral sampling on transmit by a $q$-factor of four and ten respectively.
 As expected the geometric distortion and scalloping loss decreases with increased lateral sampling.
@@
 %$1/100$ in diagonal loading.
-Note that a diagonal loading factor of $1/100$ will introduce a white noise component in the sample covariance matrix $20$ dB lower than the signal present.
+Note again that a diagonal loading factor of $1/100$ will introduce a white noise component in the sample covariance matrix $20$ dB lower than the signal present.
 Hence, the covariance has been estimated from data which appeared to have an SNR of 20 dB.
@@
 Since more than ten times oversampling is required in order to achieve lateral shift invariant high resolution imaging of a moving point scatterer, none of the implementations described in \cite{Asen} would be real time any more.
+We therefore need a different way of oversampling our imaging sector.
 
@@
 This is also possible in broadband applications as long as the phase rotation is less than one pulse length (See coarse-fine beamforming in \cite{Thomenius}).
-It is therefore possible to make a sweep between a given beam and its two nearest neighbours using phase rotation without introducing large errors.
+The maximal steering angle is therefore given by
+\begin{align}
+\theta_{max} = \alpha p \Delta,
+\end{align}
+where $\alpha$ is the pulse length in $\lambda$'s, and $p$ and $\Delta$ is the oversampling factor and beam spacing respectively as defined in (\ref{eq:resolution}).
+For our simulations $\theta_{max} = 3\Delta$.
+It is therefore possible to make a sweep using phase rotation between a given beam and half-way to its two nearest neighbours without introducing large errors.
 Phase rotation is actually what the Capon beamformer use internally to micro-steer away from bright point scatterers and to enhance edges.
@@
 \end{align}
-where $x_i$ is the element position and $\theta$ is swept from $-\Delta/2$ to $\Delta/2$ where $\Delta$ is as defined in (\ref{eq:resolution}).
-Note that the steering vector must be calculated for a $L$ element symmetric subarray, even though the full array has $M$ elements.
+where $x_i$ is the element position and $\theta$ is swept from $-\Delta/2$ to $\Delta/2$.
+Note that the steering vector must be calculated for an $L$ element symmetric subarray, even though the full array has $M$ elements.
 This subarray should have the same element pitch, $x_i - x_{i-1}$, as the full array.
@@
 If the equation $\mat{\hat{R}}\vec{a} = \vec{b}$, as in \cite{Asen}, is solved using the Gauss-Jordan algorithm, we can solve the problem for all steering vectors at once by row reducing the matrix $[\mat{\hat{R}},  \vec{a}_{\theta_1}, \vec{a}_{\theta_2}, \cdots \vec{a}_{\theta_{N_\theta}}]$ to the following form $[\mat{I},  \vec{b}_{\theta_1}, \vec{b}_{\theta_2}, \cdots \vec{b}_{\theta_{N_\theta}}]$.
-This is still an $O(L^3)$ operation.
+This is still an $O(L^3)$ operation as long as $N_\theta$ is not much greater than $L$.
 
@@
 In addition, due to the increase in lateral resolution, the Capon beamformer requires smaller frame-to-frame movements than DAS.
-Looking at the video of Capon beamforming with ten times oversampling by phase rotation \multimedia{Capon + 10PR} there is no visible blinking anymore.
-We therefore conclude that the system is now locally laterally shift invariant.
+Looking at the video of Capon beamforming with ten times oversampling by phase rotation \multimedia{Video Capon + 10PR} the blinking is now highly reduced.
+%The system is however not fully shift invariant.
+To achieve this a higher level of oversampling, %We therefore conclude that the system is now locally laterally shift invariant.
 
@@
 In this paper we have investigated the lateral shift-invariance property of the Capon beamformer when imaging moving objects.
-In Section \ref{sec:capon_LLSI} it was shown that a Capon beamformer with high resolution parameters is not locally laterally shift-invariant when subjected to motion.
+We have shown that the Capon beamformer is less sensitive to small lateral shifts if oversampling based on phase rotation is applied.
+This without reducing the acquisition frame rate and with a minor increase in processing complexity.
+%In Section \ref{sec:capon_LLSI} it was shown that a Capon beamformer with high resolution parameters is not locally laterally shift-invariant when subjected to motion.
 The root cause was found to be too coarse lateral sampling.
-In Section \ref{sec:methods} we therefore presented methods for improved shift-invariance based on oversampling.
+%In Section \ref{sec:methods} we presented methods for improved shift-invariance based on oversampling.
 The most promising method, with respect to improved shift-invariance, reduction in acquisition frame rate and processing complexity combined, was based on phase rotation.
-In Section \ref{sec:res} we presented results showing that this method also works on \textit{in-vitro} data when the ultrasound probe is subjected to motion.
-In simulations, oversampling based on phase rotation was found to reduce the scalloping loss by more than 20 dB.
-On \textit{in-vitro} data the reduction was more than 7 dB.
+%In Section \ref{sec:res} we presented results showing that this method also works on \textit{in-vitro} data when the ultrasound probe is subjected to motion. 
+In simulations the method reduced the maximal scalloping loss by more than 20 dB. On \textit{in-vitro} data the reduction was more than 7 dB.
 
@@
 
-If faced with an imaging scene containing bright point scatterers or sharp edges, one could wonder how large the oversampling factor needs to be in order to avoid visible scalloping losses.
-In his seminal paper, Cox \cite{Cox1973} derives formulas for the maximal achievable resolution for a given output SNR.
-For instance 47 dB is found to give a ten time improvement in resolution.
+Faced with an imaging scene containing bright point scatterers or sharp edges, one could wonder how large the oversampling factor needs to be in order to avoid visible scalloping losses.
+In his seminal paper, Cox \cite{Cox1973} derives the required beam density for a given output SNR.
+For instance 47 dB requires a beam density for the Capon beamformer ten times the one given by (\ref{eq:resolution}).
 We would therefore recommend to first use the dynamic range in the image as guidance.
@@
 Cox measures resolution as the 3 dB saddle point between two impinging monochromatic plane waves.
-For ultrasound data only a 50 \% increase in resolution was found in \cite[Fig.~7]{Asen} for a 30 dB output SNR when measured in a similar way as Cox (6 dB saddle point).
+For ultrasound data only a 50 \% increase in resolution was found in \cite[Fig.~7]{Asen} for a 30 dB output SNR when measured in a similar way as Cox [6 dB saddle point].
 However, sixteen times oversampling was required to make the imaging system shift invariant in the same paper.
@@
 This might be a result of improved initial sampling (as discussed in Section \ref{sec:res}) and a higher initial noise level before diagonal loading was applied.
+\comment{Mye forskjellige tall...
+output, input SNR osv...
+Rydd opp!}
 
@@
 eight phase rotated beams are averaged into four.
-The geometric distortion, or stripes, seen at the bottom of the sector in the video \multimedia{Capon + 10PR} should be removed by this technique.
-
-When oversampling is applied it becomes evident that the high resolution settings $L=M/2$, $K=1$ and $d=1/100$ comes through as too extreme.
+The geometric distortion, or stripes, seen at the bottom of the sector in the video \multimedia{Capon + 10PR} should be possible to remove using this technique.
+
+When oversampling is applied it becomes evident that the high resolution settings $L=M/2$, $K=1$ and $d=1/100$ might be to extreme.
 Traditionally in ultrasound imaging the point spread function is always wider laterally than in range.
@@
 A more symmetric point spread function while preserving speckle statistics is obtained with the parameters $L=M/4$ and $K=0$ \cite{Synnevag2007a}.
-Another option is to perform pulse inversion if a spiky symmetric point spread function is desirable.
-
-Oversampling is not only essential to keep the amplitude level of bright point scatterers, it also helps to improve contrast and edge definitions.
+The level of required oversampling is then also reduced.
+%Another option is to perform pulse inversion if a spiky symmetric point spread function is desirable.
+
+Oversampling is not only essential to keep the amplitude level of moving point scatterers.
+It also helps to improve contrast and edge definitions.
 If oversampling is not applied homogeneous speckle regions could get a too low average value and the edge delineation of a moving object will fluctuate in the same way as a moving point scatterer would have sparkled.
@@
 Finally we want to stress the fact that this demand for oversampling does not only affect the Capon or minimum variance beamformer.
-Other methods claiming super resolution, like the Eigenspace method, LCA beamformer etc.
+Other methods claiming super resolution, like the Eigenspace method, low complexity beamformer etc.
 \cite{Synnevag2011, Nilsen2009, Mehdizadeh2012a, Kim}, would also need oversampling when higher lateral resolution is obtained.
@@
 % biographies
-%\newpage
+
+\newpage
 
@@
 %\newpage
+
+%\vspace{-4cm}
 
@@
 
+%\vspace{-4cm}
 %\newpage
@@
 
+\vfill
 % You can push biographies down or up by placing

### 2013-12-15 HRUIofMovingObjects/jrnl_article.tex
Moving objects: done some writing

@@
 %This is indeed computational demanding, but following the innovation in GPU computing, and by applying a novel method in order to reduce the size of the matrix inversion problem \cite{Nilsen2009}, real time processing is now possible \cite{Asen}.
-From (\ref{eq:capon_weights}) it is also clear that complex data [e.g.
-in-phase and quadrature data] can lead to complex weights.
+From (\ref{eq:capon_weights}) it is also clear that complex data [in-phase and quadrature data] can lead to complex weights.
 The weights can therefore have an asymmetric K-space response with a shifted main lobe and zeros at arbitrary positions.
@@
 A $p$-factor of two has been used throughout this paper.
-The oversampling factor $q$ will be used to oversample prior to Capon beamforming, and is one if not stated otherwise.
+The oversampling factor $q$ will be used to oversample prior to Capon beamforming and is one if not stated otherwise.
 
@@
 We therefore conclude that the system is locally laterally shift-invariant.
-In this, and the following LSV-plots based on simulations, the scan sequence always has a beam at zero degrees and the point scatterer is located in the center of this beam at one time instance.
+%In this, and the following LSV-plots based on simulations, the scan sequence always has a beam at zero degrees and the point scatterer is located in the center of this beam at one time instance.
 
@@
 In Fig.\,\ref{fig:capon1} the same point scatterer response as presented in Fig.\,\ref{fig:das_shiftinvariant} has been weighted with Capon weights calculated using (\ref{eq:capon_weights}) with $L = M/2 = 48$, $K=1$ and $d=1/100$.
-As mentioned, the beam at zero degrees is the only one where the point scatterer is located in the center of the beam.
+The lateral offsets are calculated in such a way that the beam at zero degrees is the only one where the point scatterer is located in the center of the beam.
 The signal cancellation caused by the difference between the signal propagation vector and the assumed steering vector is therefore minimal when the point scatterer is located at this position.
@@
 
+Oversampling is not only essential to keep the amplitude level of moving point scatterers.
+It should also helps to improve contrast and edge definitions.
+If oversampling is not applied homogeneous speckle regions could get a too low average value and the edge delineation of a moving object will fluctuate in the same way as a moving point scatterer would have sparkled.
+
 %Note that in this paper we have investigated a single array configuration and a limited parameter range for the Capon beamformer.
@@
 For instance 47 dB requires a beam density for the Capon beamformer ten times the one given by (\ref{eq:resolution}).
-We would therefore recommend to first use the dynamic range in the image as guidance.
+Since resolution depends on SNR we would therefore recommend to first use the dynamic range in the image as guidance.
 The amount of diagonal loading will also dictate the maximal achievable resolution since $10\log{d}$ can be interpreted as the input SNR seen by the Capon beamformer.
-Cox measures resolution as the 3 dB saddle point between two impinging monochromatic plane waves.
+%Cox measures resolution as the 3 dB saddle point between two impinging monochromatic plane waves.
 For ultrasound data only a 50 \% increase in resolution was found in \cite[Fig.~7]{Asen} for a 30 dB output SNR when measured in a similar way as Cox [6 dB saddle point].
@@
 This might be a result of improved initial sampling (as discussed in Section \ref{sec:res}) and a higher initial noise level before diagonal loading was applied.
-\comment{Mye forskjellige tall...
-output, input SNR osv...
-Rydd opp!}
 
@@
 
-Oversampling by phase rotation can therefore also introduce similar block artifacts as parallel receive beamforming.
-These artifacts should be possible to remove in the same way as for parallel receive beamforming by using a synthetic transmit beamforming (STB) approach.
+Oversampling by phase rotation can therefore also introduce similar block artifacts as parallel receive beamforming (PRB).
+These artifacts should be possible to remove in the same way as for PRB by using the synthetic transmit beamforming (STB) approach.
 If we distribute our steering angles in an overlapping scheme [like rx-beams are spread out for STB] we get a phase rotation version of STB where e.g.
@@
 
-When oversampling is applied it becomes evident that the high resolution settings $L=M/2$, $K=1$ and $d=1/100$ might be to extreme.
+When oversampling is applied it becomes evident that the high resolution settings used in this paper might be to extreme, both with respect to the level of required oversampling and the point's appearance.
 Traditionally in ultrasound imaging the point spread function is always wider laterally than in range.
@@
 A more symmetric point spread function while preserving speckle statistics is obtained with the parameters $L=M/4$ and $K=0$ \cite{Synnevag2007a}.
-The level of required oversampling is then also reduced.
+The level of required oversampling and the computational complexity are also reduced.
 %Another option is to perform pulse inversion if a spiky symmetric point spread function is desirable.
 
-Oversampling is not only essential to keep the amplitude level of moving point scatterers.
-It also helps to improve contrast and edge definitions.
-If oversampling is not applied homogeneous speckle regions could get a too low average value and the edge delineation of a moving object will fluctuate in the same way as a moving point scatterer would have sparkled.
-
-Finally we want to stress the fact that this demand for oversampling does not only affect the Capon or minimum variance beamformer.
+Finally we want to stress the fact that this demand for oversampling does not only affect the Capon [or minimum variance beamformer].
 Other methods claiming super resolution, like the Eigenspace method, low complexity beamformer etc.
 \cite{Synnevag2011, Nilsen2009, Mehdizadeh2012a, Kim}, would also need oversampling when higher lateral resolution is obtained.
-The same applies for high resolution spectral estimation \cite{Ekroll2012}, but here it is much easier to obtain sufficient sampling initially.
+The same applies for high resolution spectral estimation \cite{Ekroll2012}.
 
@@
 \end{IEEEbiography}
+
+\vfill{}
+
+\newpage
 

### 2013-12-16 HRUIofMovingObjects/jrnl_article.tex
Moving objects: done some writing

@@
 This has led to the development of several robust versions of the Capon beamformer with e.g.
-spatial smoothing \cite{Shan1985} or diagonal loading \cite{JianLi2003} of the covariance estimate, or by applying a steering vector uncertainty set \cite{Lorenz2005, Rubsamen2013}.
+spatial smoothing \cite{Shan1985} or diagonal loading \cite{JianLi2003} of the covariance estimate, or by applying steering vector uncertainty sets \cite{Lorenz2005, Rubsamen2013}.
 Spatial smoothing and diagonal loading, together with temporal smoothing of the covariance estimate, have been applied in earlier work on Capon beamforming for medical ultrasound imaging \cite{Synnevag2009}.
@@
 In Fig.\,\ref{fig:das_capon_beams} we have plotted the two-way response for DAS and the Capon beamformer using (26) and (32) in \cite{Cox1973}.
-A 64-element linear array with $\lambda/2$ element spacing, where $\lambda$ is the wavelength, is scanned over a sector equal to two times the required beam spacing using narrow band excitation.
+A 64-element linear array with $\lambda/2$ element spacing, where $\lambda$ is the wavelength, is scanned over a sector equal to four times the array resolution using narrow band excitation.
 Located in the farfield of the sector is an object moving parallel to the array surface.
-The plot shows the scanned output when the object is located at plus and minus half the required beam spacing for DAS [See (\ref{eq:resolution})].
+The plot shows the high precision scanned output when the object is located at plus and minus half the required beam spacing for DAS [See (\ref{eq:resolution})].
 %No spatially white noise is present.
 Hence, the interference-noise matrix ($\sigma_0^2\mat{Q}$ in \cite{Cox1973}) is the identity.
-The effect of typical subarray averaging and diagonal loading ($L=32$ and $d=1/100$) has also been added to the Capon beamformer output.
+The effect of typical subarray averaging and diagonal loading ($L=32$ and $d=1/100$) has been added to the Capon beamformer output.
 A diagonal loading of $1/100$ corresponds to a signal-to-noise ratio (SNR) of 20 dB before array gain is applied.
-The figure shows that a DAS beamformer, with the required beam spacing, could yield a $1.9$ dB scalloping loss \cite{Harris1978} if the object is moving parallel to the array.
+The figure shows that a DAS beamformer, using the required beam spacing, could yield a $1.9$ dB scalloping loss \cite{Harris1978} if the object is moving parallel to the array.
 However if the Capon beamformer is used with the same beam density the loss could be as large as $27$ dB.
-This imaging system is then highly sensitive to small spatial shifts.
+This imaging system is then highly sensitive to small spatial shifts, either by the object or the array itself.
 A $1.9$ dB scalloping loss for the Capon beamforming is obtained for this specific example when the beam spacing or object distance is decreased with a factor of 25.
@@
 We show that when a moving point scatterer is imaged with the Capon beamformer with the same beam spacing sufficient for DAS, the predicted scalloping loss is observed both in simulations and \textit{in-vitro} data.
-The question is then, what can we do in order to improve on the situation without affecting the imaging frame rate and processing throughout.
-We try two answer this by studying the scalloping loss of the Capon beamforming and investigating different methods aimed at reducing it.
+The question is then, what can we do in order to improve on the situation without affecting the acquisition frame rate and processing throughout.
+%We try two answer this by studying the scalloping loss of the Capon beamforming and investigating different methods aimed at reducing it.
 We also discuss how the different methods will affect the acquisition frame rate and the processing complexity.
 
-In the next section we give an introduction to Capon beamforming for medical ultrasound imaging together with a short introduction to lateral sampling and shift-invariance.
-In Section \ref{sec:capon_LLSI} the local lateral shift-invariance of the Capon beamformer is investigated, before different methods for improved local lateral shift-invariance are presented in Section \ref{sec:methods}.
+The next section gives an introduction to Capon beamforming for medical ultrasound imaging together with a short introduction to lateral sampling and shift-invariance.
+In Section \ref{sec:capon_LLSI} we study the scalloping loss yield by the Capon beamformer and its local lateral shift-invariance property.
+Different methods for reduced scalloping loss and by that improved local lateral shift-invariance are then presented in Section \ref{sec:methods}.
 Then, in Section \ref{sec:res} we present the results of applying the most promising method on \textit{in-vitro} data.
@@
 The probe was excited with a 2.5 MHz and 1.5 cycles long pulse.
-The transmit focus in azimuth and elevation was configured to be equal and the point scatterer was moved through this focus.
+The transmit focus in azimuth and elevation were configured to be equal and the point scatterer was moved through this point.
 We see that the contours in Fig.\,\ref{fig:das_shiftinvariant} are approximately diagonal (low P-MAE) and that the amplitude variation among the maximum points is minimal (low A-MAE).
@@
 The Capon beamformer formula in (\ref{eq:capon_weights}) gives a direct opportunity for oversampling using phase rotation.
-The steering vector $\vec{a}$ has in previous work been set to $\vec{1}$ since the ultrasound data are pre-delayed.
+The steering vector $\vec{a}$ has in previous work been set to $\vec{1}$ since ultrasound data have to be pre-delayed.
 However, the steering vector can also, as in narrow band applications, be varied over a set of pre-defined vectors.
-In a narrow band setting the beam can make a sweep across the whole imaging sector by using phase rotation.
+In a narrow band setting the beam can make a sweep across the whole imaging sector by phase steering.
 This is also possible in broadband applications as long as the phase rotation is less than one pulse length (See coarse-fine beamforming in \cite{Thomenius}).
@@
 %In Section \ref{sec:res} we presented results showing that this method also works on \textit{in-vitro} data when the ultrasound probe is subjected to motion. 
-In simulations the method reduced the maximal scalloping loss by more than 20 dB. On \textit{in-vitro} data the reduction was more than 7 dB.
-
-Oversampling is not only essential to keep the amplitude level of moving point scatterers.
+In simulations the method reduced the maximal scalloping loss by more than 20 dB.
+On \textit{in-vitro} data the reduction was more than 7 dB.
+On \textit{in-vitro} data the imaging system is almost locally laterally shift invariant.
+
+Note that oversampling is not only essential to keep the amplitude level of moving point scatterers.
 It should also helps to improve contrast and edge definitions.
@@
 For instance 47 dB requires a beam density for the Capon beamformer ten times the one given by (\ref{eq:resolution}).
-Since resolution depends on SNR we would therefore recommend to first use the dynamic range in the image as guidance.
-The amount of diagonal loading will also dictate the maximal achievable resolution since $10\log{d}$ can be interpreted as the input SNR seen by the Capon beamformer.
+Since resolution for the Capon beamformer depends on SNR we would recommend to first use the dynamic range in the image as guidance.
+The amount of diagonal loading will also dictate the maximal achievable resolution since $10\log{d}$ can be interpreted as the maximal input SNR seen by the Capon beamformer.
 %Cox measures resolution as the 3 dB saddle point between two impinging monochromatic plane waves.
@@
 
-Oversampling by phase rotation can therefore also introduce similar block artifacts as parallel receive beamforming (PRB).
-These artifacts should be possible to remove in the same way as for PRB by using the synthetic transmit beamforming (STB) approach.
+Oversampling by phase rotation can therefore introduce similar block artifacts as parallel receive beamforming (PRB).
+These artifacts should be possible to remove in the same way as for PRB by using the synthetic transmit beamforming (STB) technique.
 If we distribute our steering angles in an overlapping scheme [like rx-beams are spread out for STB] we get a phase rotation version of STB where e.g.

### 2013-12-17 HRUIofMovingObjects/jrnl_article.tex
Moving objects: writing, updated figures

@@
 A straight forward way of obtaining sufficient sampling in an active imaging system is by increasing the number of transmit beams (tx-beams) until the system is locally laterally shift-invariant.
-In Fig.\,\ref{fig:capon_oversampling_a} and Fig.\,\ref{fig:capon_oversampling_b} we have increased the lateral sampling on transmit by a $q$-factor of four and ten respectively.
+In Fig.\,\ref{fig:capon_oversampling_a} and Fig.\,\ref{fig:capon_oversampling_b} we have increased the lateral sampling on transmit by a $q$-factor of four and sixteen respectively.
 As expected the geometric distortion and scalloping loss decreases with increased lateral sampling.
@@
 
-From Fig.\,\ref{fig:capon_oversampling_b} we see that more than ten times oversampling has to be applied to achieve a local lateral shift-invariant Capon beamformer when a moving point scatterer is imaged without noise and with the same parameters as in Fig\,\ref{fig:capon}.
+From Fig.\,\ref{fig:capon_oversampling_b} we see that sixteen times oversampling is not enough to achieve a local lateral shift-invariant Capon beamformer when a moving point scatterer is imaged without noise and with the same parameters as in Fig\,\ref{fig:capon}, but we are getting closer.
 %$1/100$ in diagonal loading.
@@
 a) Four times oversampling.
-b) Ten times oversampling.
-The maximum scalloping loss is still too high for the system to be called locally laterally shift invariant.}
+b) Sixteen times oversampling.
+Even though its reduced, the maximum scalloping loss is still too high for the system to be named locally laterally shift invariant.}
 \label{fig:capon_oversampling}
@@
 The processing times reported in \cite{Asen} will then also increase with the same factor.
-Since more than ten times oversampling is required in order to achieve lateral shift invariant high resolution imaging of a moving point scatterer, none of the implementations described in \cite{Asen} would be real time any more.
+Since more than ten times oversampling is required in order to achieve lateral shift invariant high resolution imaging of a moving point scatterer \textit{in-vitro}, none of the implementations described in \cite{Asen} would be real time any more.
 We therefore need a different way of oversampling our imaging sector.
@@
 The figure is zoomed compared with Fig\,\ref{fig:capon} to better see the beam-to-beam variation.
-a) 4x oversampling.
-b) 10x oversampling.}
+a) Four times oversampling.
+b) Sixteen times oversampling.}
 	\label{fig:benchmark_capon_bs}
@@
 
-Figure \ref{fig:phantom_capon_LSV_10PR} presents the result of applying ten times oversampling by phase rotation.
+Figure \ref{fig:phantom_capon_LSV_10PR} presents the result of applying sixteen times oversampling by phase rotation.
 We observe how the amplitude of the point scatterer is better preserved in all frames, and the scalloping loss is not far from our 1 dB threshold.

### 2013-12-18 HRUIofMovingObjects/jrnl_article.tex
Moving objects. Almost there-....

@@
 \IEEEPARstart{I}{n} the last decade several authors have applied the Capon beamformer (also known as the minimum variance beamformer) for medical ultrasound imaging \cite{Synnevag2007, Vignon2008, Viola}.
-The method has been shown to improve the lateral resolution and edge sharpness \cite{Synnevag2007, Synnevag2009, Chen2011}.
+The method has been shown to improve lateral resolution and edge sharpness \cite{Synnevag2007, Synnevag2009, Chen2011} in ultrasound images.
 This is accomplished by adaptively adjusting the array weights based on element-to-element covariance in order to minimize interference.
@@
 
-A steering vector uncertainty set reduce signal-nulling which appears if the signal of interest is present during the covariance estimation process and if this estimate is later used with a steering vector that does not match the signal propagation vector exactly.
+A steering vector uncertainty set reduces signal-nulling which appears if the signal of interest is present during the covariance estimation process and if this estimate is later used with a steering vector that does not match the signal propagation vector exactly.
 This self-nulling is actually how the Capon beamformer obtains its super resolution capabilities, but the signal can also get lost in the process.
@@
 Hence, the interference-noise matrix ($\sigma_0^2\mat{Q}$ in \cite{Cox1973}) is the identity.
-The effect of typical subarray averaging and diagonal loading ($L=32$ and $d=1/100$) has been added to the Capon beamformer output.
-A diagonal loading of $1/100$ corresponds to a signal-to-noise ratio (SNR) of 20 dB before array gain is applied.
-The figure shows that a DAS beamformer, using the required beam spacing, could yield a $1.9$ dB scalloping loss \cite{Harris1978} if the object is moving parallel to the array.
+The effect of typical subarray averaging and diagonal loading ($L=32$ and $d=1/100$) have been added to the Capon beamformer output.
+We see that a DAS beamformer, using the required beam spacing, could yield a $1.9$ dB scalloping loss \cite{Harris1978} if the object is moving parallel to the array.
 However if the Capon beamformer is used with the same beam density the loss could be as large as $27$ dB.
@@
 }
-\caption{Theoretical drop in output power mid-way between receive beams for delay-and-sum and Capon beamforming ($M=64$, $L=32$, $d=0.01$) applied in a farfield-narrowband setting.
+\caption{Estimated drop in output power mid-way between receive beams for delay-and-sum and Capon beamforming ($M=64$, $L=32$, $d=0.01$) applied in a farfield-narrowband setting.
 %The scanned output from both beamformers is plotted for two different time instances.
@@
 Figure \ref{fig:das_shiftinvariant} presents such a LSV-plot for an imaging system using the DAS beamformer with a beam spacing equal to (\ref{eq:resolution}).
-The dotted vertical lines indicates the position of transmit-receive beam pairs.
+Dotted vertical lines indicate position of transmit-receive beam pairs.
 The point scatterer response has been simulated using Field II \cite{Jensen1992, Jensen1996a} with a probe setup equal to the M5S probe by GE Vingmed Ultrasound, Horten, Norway.
@@
 In Fig.\,\ref{fig:das_shiftvariant} the beam density is reduced by a factor of two compared to Fig.\,\ref{fig:das_shiftinvariant}.
-We see how the geometric distortion increases to a level where the contours no longer are diagonal.
+The geometric distortion is now increased to a level where the contours are not diagonal.
 The maximum scalloping loss is also larger than 1 dB.
 Hence, the imaging system is no longer locally laterally shift-invariant.
-We observe that the maximum peak is both geometrically distorted and attenuated when the point is located between two beams and the distance between the beams are too high.
+We observe that the maximum peak is both geometrically distorted and attenuated when the point is located between two beams and the distance between the beams is too high.
 
@@
 \section{Local Lateral Shift-Invariance of the Capon Beamformer}\label{sec:capon_LLSI}
-We will now investigate the local lateral shift-invariance of the Capon beamformer with parameters equal to the ones used in earlier work \cite{Synnevag2009, Asen}.
+We will now investigate the local lateral shift-invariance of the Capon beamformer with high resolution parameters \cite{Synnevag2009, Asen}.
 In Fig.\,\ref{fig:capon1} the same point scatterer response as presented in Fig.\,\ref{fig:das_shiftinvariant} has been weighted with Capon weights calculated using (\ref{eq:capon_weights}) with $L = M/2 = 48$, $K=1$ and $d=1/100$.
-The lateral offsets are calculated in such a way that the beam at zero degrees is the only one where the point scatterer is located in the center of the beam.
-The signal cancellation caused by the difference between the signal propagation vector and the assumed steering vector is therefore minimal when the point scatterer is located at this position.
-From this plot we observe that the signal cancellation is high for all the other offsets.
+We see how the signal cancellation caused by the difference between the signal propagation vector and the assumed steering vector increases as the point moves away from a given transmit-receive beam.
 As a consequence the scalloping loss is tremendous, and not far from the predicted value in Fig\,\ref{fig:das_capon_beams}.
@@
 b) The diagonal loading factor $d$ is increased to 1.
-The system is still not locally lateral shift-invariant.}
+The scalloping loss is reduced, but the system is still not locally lateral shift-invariant.}
 \label{fig:capon}
@@
 From the previous section it should be clear that we somehow need to increase the lateral sampling to make the Capon imaging system locally laterally shift-invariant.
-In this section we will investigate how this oversampling should be conducted, how large it needs to be and how the proposed methods will affect the system frame rate and the processing complexity.
+In this section we will investigate how this oversampling should be conducted, how large it needs to be and how the proposed methods will affect the acquisition frame rate and the processing complexity.
 
@@
 
-From Fig.\,\ref{fig:capon_oversampling_b} we see that sixteen times oversampling is not enough to achieve a local lateral shift-invariant Capon beamformer when a moving point scatterer is imaged without noise and with the same parameters as in Fig\,\ref{fig:capon}, but we are getting closer.
+From Fig.\,\ref{fig:capon_oversampling_b} we see that sixteen times oversampling reduces the scalloping loss significantly, but the maximum is still larger than 1 dB.
+A even higher level of oversampling is therefore required in order to have a completely local lateral shift-invariant Capon beamformer when a moving point scatterer is simulated without noise and with the same parameters as in Fig\,\ref{fig:capon}.
 %$1/100$ in diagonal loading.
-Note again that a diagonal loading factor of $1/100$ will introduce a white noise component in the sample covariance matrix $20$ dB lower than the signal present.
+Note that a diagonal loading factor of $1/100$ will introduce a white noise component in the sample covariance matrix $20$ dB lower than the signal present.
 Hence, the covariance has been estimated from data which appeared to have an SNR of 20 dB.
 
-Even though increasing the number of transmits improves the lateral shift-invariance it will also reduce the imaging frame rate with a factor equal to the oversampling factor.
+Even though increasing the number of transmits improves the lateral shift-invariance it will also reduce the acquisition frame rate with a factor equal to the oversampling factor.
 The processing complexity will also increase equally.
@@
 The processing times reported in \cite{Asen} will then also increase with the same factor.
-Since more than ten times oversampling is required in order to achieve lateral shift invariant high resolution imaging of a moving point scatterer \textit{in-vitro}, none of the implementations described in \cite{Asen} would be real time any more.
+Since more than sixteen times oversampling is required in order to achieve lateral-shift-invariant and high-resolution imaging of a moving point scatterer, none of the implementations described in \cite{Asen} would be real time any more.
 We therefore need a different way of oversampling our imaging sector.
@@
 For our simulations $\theta_{max} = 3\Delta$.
-It is therefore possible to make a sweep using phase rotation between a given beam and half-way to its two nearest neighbours without introducing large errors.
+It is therefore possible to make a sweep using phase rotation between a given beam and half-way to its two nearest neighbours without introducing to large errors.
 Phase rotation is actually what the Capon beamformer use internally to micro-steer away from bright point scatterers and to enhance edges.
@@
 If the equation $\mat{\hat{R}}\vec{a} = \vec{b}$, as in \cite{Asen}, is solved using the Gauss-Jordan algorithm, we can solve the problem for all steering vectors at once by row reducing the matrix $[\mat{\hat{R}},  \vec{a}_{\theta_1}, \vec{a}_{\theta_2}, \cdots \vec{a}_{\theta_{N_\theta}}]$ to the following form $[\mat{I},  \vec{b}_{\theta_1}, \vec{b}_{\theta_2}, \cdots \vec{b}_{\theta_{N_\theta}}]$.
-This is still an $O(L^3)$ operation as long as $N_\theta$ is not much greater than $L$.
+This is an $O(L^3)$ operation as long as $N_\theta$ is not much greater than $L$.
 
@@
 For the frame where the point scatterer is at -6 degrees the receive beam direction is very close to the signal propagation vector.
-Hence, we get a high amplitude value at this location.
-In almost all the other frames the signal loss is massive (10 dB).
+Hence, we see a high amplitude value at this location.
+In almost all the other frames the signal loss is massive (at most 10 dB).
 The result is blinking point scatterers when subjected to motion \multimedia{Video Capon}.
@@
 In addition, due to the increase in lateral resolution, the Capon beamformer requires smaller frame-to-frame movements than DAS.
-Looking at the video of Capon beamforming with ten times oversampling by phase rotation \multimedia{Video Capon + 10PR} the blinking is now highly reduced.
+Looking at the video of Capon beamforming with sixteen times oversampling by phase rotation \multimedia{Video Capon + 16PR} the blinking is now highly reduced.
 %The system is however not fully shift invariant.
@@
 We have shown that the Capon beamformer is less sensitive to small lateral shifts if oversampling based on phase rotation is applied.
-This without reducing the acquisition frame rate and with a minor increase in processing complexity.
+This without reducing the acquisition frame rate and with only a minor increase in processing complexity.
 %In Section \ref{sec:capon_LLSI} it was shown that a Capon beamformer with high resolution parameters is not locally laterally shift-invariant when subjected to motion.
@@
 %In Section \ref{sec:res} we presented results showing that this method also works on \textit{in-vitro} data when the ultrasound probe is subjected to motion. 
-In simulations the method reduced the maximal scalloping loss by more than 20 dB.
-On \textit{in-vitro} data the reduction was more than 7 dB.
-On \textit{in-vitro} data the imaging system is almost locally laterally shift invariant.
+In simulations the method reduced the maximal scalloping loss by 16 dB.
+On \textit{in-vitro} data the reduction was more than 7 dB, and the imaging system is now almost locally laterally shift invariant.
 
 Note that oversampling is not only essential to keep the amplitude level of moving point scatterers.
-It should also helps to improve contrast and edge definitions.
+It should also help to improve contrast and edge definitions.
 If oversampling is not applied homogeneous speckle regions could get a too low average value and the edge delineation of a moving object will fluctuate in the same way as a moving point scatterer would have sparkled.
@@
 The amount of diagonal loading will also dictate the maximal achievable resolution since $10\log{d}$ can be interpreted as the maximal input SNR seen by the Capon beamformer.
-%Cox measures resolution as the 3 dB saddle point between two impinging monochromatic plane waves.
+Fig\,\ref{fig:das_capon_beams} shows that 25 times oversampling is required for a $d$-value of $1/100$.
+When \textit{in-vitro} data contains additional noise this oversampling requirement is reduced.%Cox measures resolution as the 3 dB saddle point between two impinging monochromatic plane waves.
 For ultrasound data only a 50 \% increase in resolution was found in \cite[Fig.~7]{Asen} for a 30 dB output SNR when measured in a similar way as Cox [6 dB saddle point].
@@
 However, where these set-based methods typically has a criterion for selecting the best output after applying all the steering vectors contained in the set, oversampling with phase rotation outputs all.
-It can also be thought of as a narrow band approach to parallel receive beamforming where Capon is applied on all receive beams.
+It can also be thought of as a narrow band approach to parallel receive beamforming where Capon's method is applied on all receive beams.
 
@@
 eight phase rotated beams are averaged into four.
-The geometric distortion, or stripes, seen at the bottom of the sector in the video \multimedia{Capon + 10PR} should be possible to remove using this technique.
-
-When oversampling is applied it becomes evident that the high resolution settings used in this paper might be to extreme, both with respect to the level of required oversampling and the point's appearance.
-Traditionally in ultrasound imaging the point spread function is always wider laterally than in range.
+The geometric distortion, or stripes, seen at the bottom of the sector in the video with sixteen times oversampling \multimedia{Capon+16PR} should be possible to remove using this technique.
+
+When oversampling is applied it becomes evident that the high resolution settings used in this paper might be to extreme, both with respect to the level of required oversampling and the point scatterer's appearance.
+Traditionally in medical ultrasound imaging the point spread function is always wider laterally than in range.
 With the Capon beamformer this is now changed.
@@
 Other methods claiming super resolution, like the Eigenspace method, low complexity beamformer etc.
-\cite{Synnevag2011, Nilsen2009, Mehdizadeh2012a, Kim}, would also need oversampling when higher lateral resolution is obtained.
+\cite{Synnevag2011, Nilsen2009, Mehdizadeh2012a, Kim}, would also require oversampling when improved lateral resolution is obtained.
 The same applies for high resolution spectral estimation \cite{Ekroll2012}.
@@
 We then discussed and investigated different methods in order to reduced this effect.
-Finally we proved that the local lateral shift-invariance could be improved by oversampling based on phase rotation without affecting the acquisition frame rate and with minor changes in processing complexity.
+Finally we showed that the local lateral shift-invariance could be improved by oversampling based on phase rotation without affecting the acquisition frame rate and with minor changes in processing complexity.
 

### 2014-01-04 ThesisJP/Introduction/introduction.tex
Small layout adjustments

--- 2013-12-06 ThesisJP/Introduction/introduction.tex
@@
 
-\subsubsection{Paper no. 1:\\
+\subsubsection{Paper I:\\
 An Optimised GPU Implementation of the MVDR Beamformer for Active Sonar Imaging}
@@
 
-\subsubsection{Paper no. 2:\\
+\subsubsection{Paper II:\\
 Implementing Capon Beamforming on a GPU for Real-Time Cardiac Ultrasound Imaging}
@@
 
+\section{Future work}
+
 \section{Conclusion}
 
-\section{Future work}
-
-\section{Publication list}
-
-\subsubsection{Papers included in the thesis}
-\begin{enumerate}
-	\item J.
-I.
-Buskenes, J.
-P.
-\AA{}sen, C.-I.
-C.
-Nilsen and A.
-Austeng, "An Optimised GPU Implementation of the MVDR Beamformer for Active Sonar Imaging", {\it Submitted for review to IEEE Transactions on Oceanic Engineering}.
-	\item 
-\end{enumerate}
-
-\subsubsection{Conference proceedings}
-\begin{enumerate}
-	\item 
-	\item 
-\end{enumerate}
-
-\subsubsection{Presentations and abstracts}
-\begin{enumerate}
-	\item
-\end{enumerate}
-
 \endinput

### 2014-01-04 ThesisJP/Thesis.tex
Small layout adjustments

%\documentclass[11pt,a4paper]{book}
\documentclass[10pt,b5paper,twoside]{book}

% Line spacing
%\linespread{1.0}

% Include document setup script
\include{documentproperties}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% TITLE PAGE
\author{Jon Petter \AA{}sen}
\title{%Integrated Ultrasound Visualization and Formation
%Real-Time Adaptive Beamforming, Visualization, and Simulation 
%Real-Time Adaptive Ultrasound Algorithms
Accelerating adaptive ultrasound algorithms using graphics processing units 
\vspace{0.5cm}}
\department{Department of Circulation and Medical Imaging}

\usepackage{color}
\newcommand\todo[1]{{\textit{\color{red}(#1)}}}

\begin{document}
\date{\today}
\pagestyle{empty} \maketitle

%Norw.  sammendrag
% Norsk sammendrag
%\Chapter*{}
\newpage%
~\\
\vspace{10.0cm}
~\\
{\bf NTNU}
\\
Norwegian University of Science and Technology\\
\\
Thesis for the degree of Philosophiae Doctor\\
\\
Faculty of Medicine\\
Department of Circulation and Medical Imaging\\
\\
\copyright\,Jon Petter \AA{}sen\\
\\
ISBN XXX-XX-XXX-XXXX-X (printed version)\\
ISBN XXX-XX-XXX-XXXX-X (electronic version)\\
ISSN XX-XX\\
\\
Doctoral theses at NTNU, 20014:XX\\
\\
Printed by NTNU-trykk

~\\
\newpage
\thispagestyle{empty}%
~\\%
~\\%
~\\
\noindent\large 
Sammendrag p\aa{} norsk.
%\vspace{0.2cm}\\%
\\\\%
\normalsize
\todo{Skriv sammendrag p\aa{} norsk}

\vspace{0.3cm}
\noindent Jon Petter \AA{}sen\\
Institutt for sirkulasjon og bildediagnostikk, NTNU\\
Hovedveileder: Sverre Holm\\ 
Biveiledere: Stein-Inge Rabben, Erik Steen and Hans Torp\\
Finansieringskilde: MI-Lab

\vspace{0.2cm}
\noindent \emph{Ovennevnte avhandling er funnet verdig til \aa{} forsvares offentlig for graden Philosophiae Doctor (PhD) i medisinsk teknologi. Disputas finner sted ... .}

% ENGLISH ABSTRACT / SUMMARY
\CustChapter*{Abstract}
%\chapter*{Abstract}
\thispagestyle{empty}%
\todo{Write abstract ...}

%\setcounter{page}{3} %I don't want the norwegian summary to be counted in. (This will change page numbering since the delivered version)

% PREFACE AND ACKNOWLEDGEMENTS
\CustChapter*{Preface}%
\thispagestyle{empty}%
The present thesis is submitted in partial fulfillment of the requirements for the degree of \textit{Philosophiae Doctor} (Ph.D.) at the Faculty of Medicine at the Norwegian University of Science and Technology (NTNU). The research was funded by \textit{Medical Imaging Laboratory} (MI-Lab), and was carried out at the Department of Circulation and Medical Imaging. The main supervisor has been Professor Sverre Holm  in the Department of Informatics, University of Oslo, and Professor II at MI-Lab. Co-supervisors have been PhD Stein Inge Rabben, PhD Erik Steen and Professor Hans Torp.

\subsubsection{Acknowledgments}
\todo{Write acknowledgments ...}

\CustChapter*{List of publications}
This thesis is based on the following five papers, referred to in the text by their Roman numerals (I-V): 
\begin{enumerate}[I]
\renewcommand\labelenumi{\bfseries\theenumi}
	\item J. I. Buskenes, \textbf{J. P. \AA{}sen}, C.-I. C. Nilsen and A. Austeng, "An optimised GPU implementation of the MVDR beamformer for active sonar imaging", {\it  IEEE Transactions on Oceanic Engineering, submitted}.
	\item \textbf{J. P. \AA{}sen}, J. I. Buskenes, C.-I. C. Nilsen, A. Austeng and S. Holm, "Implementing Capon beamforming on a GPU for real-time cardiac ultrasound imaging", {\it IEEE Transactions on Ultrasonics, Ferroelectrics, and Frequency Control, vol. 61, no. 1, pp. \todo{add pages} January 2014}.
 	\item \textbf{J. P. \AA{}sen}, A. Austeng and S. Holm, "Capon beamforming and moving objects - An analysis of lateral shift-invariance", {\it IEEE Transactions on Ultrasonics, Ferroelectrics, and Frequency Control, submitted}.
	\item \textbf{J. P. \AA{}sen}, E. Steen, G. Kiss, A. Thorstensen and S. I. Rabben, "Adaptive volume rendering of cardiac 3D ultrasound images - utilizing blood pool statistics", {\it Proc. SPIE Medical Imaging 2012 \todo{add pages}}.
	\item \textbf{J. P. \AA{}sen} and S. Holm, "Huygens on speed: Interactive simulation of ultrasound pressure fields", {\it Proc. IEEE Ultrasonics Symposium 2012 \todo{add pages}}.
\end{enumerate}

\CustChapter*{Related publications}
The following papers are related to the included papers, and are referred to in the text by their Roman numerals:

\begin{enumerate}[VI]
\setcounter{enumi}{5}
\renewcommand\labelenumi{\bfseries\theenumi}

	\item \textbf{J. P. \AA{}sen}, J. I. Buskenes, C.-I. C. Nilsen, A. Austeng and S. Holm, "Implementing Capon Beamforming on the GPU for Real-Time Cardiac Ultrasound Imaging", {\it Proc. IEEE Ultrasonics Symposium 2012}.

	\item J. I. Buskenes, \textbf{J. P. \AA{}sen}, C.-I. C. Nilsen and A. Austeng, "Adapting the Minimum variance beamformer to a graphics processing unit for active sonar imaging systems", {\it Proc. Meetings on Acoustics 2013}.

	\item G. Kiss, E. Steen, \textbf{J. P \AA{}sen} and H. Torp, "GPU volume rendering in 3D echocardiography: real-time pre-processing and ray-casting", {\it Proc. IEEE Ultrasonics Symposium 2010}.

	%\item G. Kiss, \textbf{J. P. \AA{}sen}, J. Bogaert, B. H. Amundsen, P. Claus, J. R. M. D'hooge and H. Torp, "Multi-modal cardiac image fusion and visualization on the GPU", {\it Proc. IEEE Ultrasonics Symposium 2011}.

	\item E. S. Br\o{}nstad, \textbf{J. P. \AA{}sen}, H. Torp and G. Kiss, "Visibility driven visualization of 3D cardiac ultrasound data on the GPU", {\it Proc. IEEE Ultrasonics Symposium 2012}.
\end{enumerate}

\newpage\thispagestyle{empty}
\pagestyle{fancy}
%
\tableofcontents

%%%% CHAPTER 1 %%%%
\begin{bibunit}[ieeetr]
\CustChapter{Introduction} 
\runningtitle{Introduction}
\input{./Introduction/introduction}
\newpage\addcontentsline{toc}{section}{References}
\putbib[bibIntroduction]
\end{bibunit}

%%%% CHAPTER 2 %%%%
\begin{bibunit}[ieeetr]
\CustChapter{Background} 
\runningtitle{Background}
\input{./Background/background}
\newpage\addcontentsline{toc}{section}{References}
\putbib[bibBackground]
\end{bibunit}

%%%% Paper 1 %%%%
\begin{bibunit}[ieeetr]
\CustChapter{An Optimised GPU Implementation of the MVDR Beamformer for Active Sonar Imaging}
\runningtitle{An Optimised GPU Implementation of the MVDR Beamformer for Active Sonar Imaging}
\authors{
	Jo Inge Buskenes$^{1}$, 
    	Jon Petter \AA{}sen$^{2}$, 
    	Carl-Inge C. Nilsen$^{1}$ and
    	Andreas Austeng$^{1}$
} 
{
	$^{1}$Department of Informatics, University of Oslo, Oslo, Norway\\
    	$^{2}$Medical Imaging Lab (MI-Lab), Norwegian University of Science and Technology, Trondheim, Norway
}
\noindent \textit{Submitted for review to IEEE Transactions on Oceanic Engineering}
%\input{./STB/STB_manuscript}
\newpage\addcontentsline{toc}{section}{References}
\putbib[bibSTB]
\end{bibunit}

%%%% Paper 2 %%%%
\begin{bibunit}[ieeetr]
\CustChapter{Implementing Capon Beamforming on a GPU for Real-Time Cardiac Ultrasound Imaging}
\runningtitle{Implementing Capon Beamforming on a GPU for Real-Time Cardiac Ultrasound Imaging}
\authors{
	Jon Petter \AA{}sen$^{1}$,
    	Jo Inge Buskenes$^{2}$,  
    	Carl-Inge C. Nilsen$^{2}$,
    	Andreas Austeng$^{2}$ and
    	Sverre Holm$^{1,2}$
} 
{
	$^{1}$Medical Imaging Lab (MI-Lab), Norwegian University of Science and Technology, Trondheim, Norway\\
    	$^{2}$Department of Informatics, University of Oslo, Oslo, Norway
}
\noindent \textit{IEEE Transactions on Ultrasonics, Ferroelectrics, and Frequency Control, vol. XY, January 2014}
%\input{./MLAandAb/Aberration_HiFR}
\newpage\addcontentsline{toc}{section}{References}
\putbib[bibMLAandAb]
\end{bibunit}

%%%% Paper 3 %%%%
\begin{bibunit}[ieeetr]
\CustChapter{Capon Beamforming Applied on Moving Objects - An Analysis of Local Lateral Shift-Invariance}
\runningtitle{Capon Beamforming Applied on Moving Objects - An Analysis of Local Lateral Shift-Invariance}
\authors{
	Jon Petter \AA{}sen$^{1}$, 
    	Andreas Austeng$^{2}$ and
    	Sverre Holm$^{1,2}$
} 
{
	$^{1}$Medical Imaging Lab (MI-Lab), Norwegian University of Science and Technology, Trondheim, Norway\\
    	$^{2}$Department of Informatics, University of Oslo, Oslo, Norway
}
\noindent \textit{Submitted for review to IEEE Transactions on Ultrasonics, Ferroelectrics, and Frequency Control}
%\input{./STBandAb/STBandAberration}
\newpage\addcontentsline{toc}{section}{References}
\putbib[bibSTBandAb]
\end{bibunit}

%%%% Paper 4 %%%%
\begin{bibunit}[ieeetr]
\CustChapter{Adaptive Volume Rendering of Cardiac 3D Ultrasound Images - Utilizing Blood Pool Statistics}
\runningtitle{Adaptive Volume Rendering of Cardiac 3D Ultrasound Images - Utilizing Blood Pool Statistics}
\authors{
	Jon Petter \AA{}sen$^{1}$,
	Erik Steen$^2$,
	Gabriel Kiss$^{3,4}$,
	Anders Thorstensen$^{1,5}$,
	Stein Inge Rabben$^2$
}
{
$^{1}$Medical Imaging Lab (MI-Lab), Norwegian University of Science and Technology, Trondheim, Norway\\
$^{2}$ GE Vingmed Ultrasound, Horten, Norway\\
$^{3}$ Norwegian University of Science and Technology, ISB - Department of Circulation and Medical Imaging, Trondheim, Norway\\
$^{4}$ St. Olavs Hospital, Trondheim, Norway\\
$^{5}$ Department of Cardiology, St. Olavs Hospital, Trondheim, Norway
}
\noindent \textit{Proc. SPIE Medical Imaging 2012}
%\input{./SPTD/StbSinglePulseTissueDoppler}
\newpage\addcontentsline{toc}{section}{References}
\putbib[bibSPTD]
\end{bibunit}

%%%% Paper 5 %%%%
\begin{bibunit}[ieeetr]
\CustChapter{Huygens on Speed: Interactive Simulation of Ultrasound Pressure Fields}
\runningtitle{Huygens on speed: Interactive simulation of ultrasound pressure fields}
\authors{Jon Petter \AA{}sen$^1$ and Sverre Holm$^{1,2}$}
{
	$^{1}$Medical Imaging Lab (MI-Lab), Norwegian University of Science and Technology, Trondheim, Norway\\
    	$^{2}$Department of Informatics, University of Oslo, Oslo, Norway
}
\noindent \textit{Proc. IEEE Ultrasonics Symposium 2012}
%\input{./SPTD/StbSinglePulseTissueDoppler}
\newpage\addcontentsline{toc}{section}{References}
\putbib[bibSPTD]
\end{bibunit}

\appendix

\end{document}

### 2014-01-04 ThesisJP/Introduction/introduction.tex
Finished layout.

@@
-\section{Background and motivation}
+\section{Motivation}
 
@@
 
-\section{Summary of presented work}
+\section{Summary of papers}
 
-\subsubsection{Paper I:\\
-An Optimised GPU Implementation of the MVDR Beamformer for Active Sonar Imaging}
-
-{\it Submitted for review to IEEE Transactions on Oceanic Engineering.}
-\vspace{0.3 cm}
-
+\subsection{Paper I}
+\textbf{An Optimised GPU Implementation of the MVDR Beamformer for Active Sonar Imaging}\\
+J. I. Buskenes, \textbf{J. P. \AA{}sen}, C.-I. C. Nilsen and A. Austeng\\
+{\it IEEE Transactions on Oceanic Engineering, submitted.}\\\\
 Summary of paper...
 
-\vspace{0.3 cm}
-\emph{The work was a joint effort with PhD student ...}
-
-\subsubsection{Paper II:\\
-Implementing Capon Beamforming on a GPU for Real-Time Cardiac Ultrasound Imaging}
-
-{\it IEEE Transactions on Ultrasonics, Ferroelectrics, and Frequency Control, vol. XY, January 2014.}
-\vspace{0.3 cm}
-
+\subsection{Paper II}
+\textbf{Implementing Capon Beamforming on a GPU for Real-Time Cardiac Ultrasound Imaging}\\
+\textbf{J. P. \AA{}sen}, J. I. Buskenes, C.-I. C. Nilsen, A. Austeng and S. Holm\\
+{\it IEEE Transactions on Ultrasonics, Ferroelectrics, and Frequency Control, vol. 61, no. 1, pp. \todo{add pages}, Jan. 2014.}\\\\
 Summary of paper...
 
-\vspace{0.3 cm}
-\emph{The work was a joint effort with PhD student ...}
-
-\subsubsection{Paper no. 3 (Chapter 5):\\
-Capon Beamforming Applied on Moving Objects - An Analysis of Local Lateral Shift-Invariance}
-
-{\it Submitted for review to IEEE Transactions on Ultrasonics, Ferroelectrics, and Frequency Control.}
-\vspace{0.3 cm}
-
+\subsection{Paper III}
+\textbf{Capon Beamforming and Moving Objects - An Analysis of Lateral Shift-Invariance}\\
+\textbf{J. P. \AA{}sen}, A. Austeng and S. Holm\\
+{\it IEEE Transactions on Ultrasonics, Ferroelectrics, and Frequency Control, submitted.}
 Summary of paper...
 
-\subsubsection{Paper no. 4 (Chapter 6):\\
-Adaptive Volume Rendering of Cardiac 3D Ultrasound Images - Utilizing Blood Pool Statistics}
-
-{\it Proc. SPIE Medical Imaging 2012}
-\vspace{0.3 cm}
-
+\subsection{Paper IV}
+\textbf{Adaptive Volume Rendering of Cardiac 3D Ultrasound Images - Utilizing Blood Pool Statistics}\\
+\textbf{J. P. \AA{}sen}, E. Steen, G. Kiss, A. Thorstensen and S. I. Rabben\\
+{\it Proc. SPIE Medical Imaging 2012, \todo{add pages}.}\\\\
 Summary of paper...
 
-\subsubsection{Paper no. 5 (Chapter 7):\\
-Huygens on Speed: Interactive Simulation of Ultrasound Pressure Fields}
-
-{\it Proc. IEEE Ultrasonics Symposium 2012}
-\vspace{0.3 cm}
-
+\subsection{Paper V}
+\textbf{Huygens on Speed: Interactive Simulation of Ultrasound Pressure Fields}\\
+\textbf{J. P. \AA{}sen} and S. Holm
+{\it Proc. IEEE Ultrasonics Symposium 2012 \todo{add pages}.}\\\\
 Summary of paper...
@@
 
+\section{Conclusion}
+
 \section{Future work}
 
-\section{Conclusion}
-
 \endinput

### 2014-01-04 ThesisJP/Thesis.tex
Finished layout.

@@
 \newpage%
+
+\renewcommand{\thepage}{\roman{page}}% Roman numerals for page counter
+
 ~\\
@@
 \newpage
-\thispagestyle{empty}%
+%\thispagestyle{empty}%
 ~\\%
@@
 \CustChapter*{Abstract}
+\addcontentsline{toc}{chapter}{Abstract}
 %\chapter*{Abstract}
-\thispagestyle{empty}%
+%\thispagestyle{empty}%
 \todo{Write abstract ...}
@@
 \CustChapter*{Preface}%
-\thispagestyle{empty}%
+\addcontentsline{toc}{chapter}{Preface}
+%\thispagestyle{empty}%
 The present thesis is submitted in partial fulfillment of the requirements for the degree of \textit{Philosophiae Doctor} (Ph.D.) at the Faculty of Medicine at the Norwegian University of Science and Technology (NTNU).
 The research was funded by \textit{Medical Imaging Laboratory} (MI-Lab), and was carried out at the Department of Circulation and Medical Imaging.
-The main supervisor has been Professor Sverre Holm  in the Department of Informatics, University of Oslo, and Professor II at MI-Lab.
+The main supervisor has been Professor Sverre Holm  in the Department of Informatics, University of Oslo, and Professor II at MI-Lab, NTNU.
 Co-supervisors have been PhD Stein Inge Rabben, PhD Erik Steen and Professor Hans Torp.
@@
 \CustChapter*{List of publications}
+\addcontentsline{toc}{chapter}{List of publications}
+
 This thesis is based on the following five papers, referred to in the text by their Roman numerals (I-V): 
@@
 1, pp.
-\todo{add pages} January 2014}.
+\todo{add pages} Jan 2014}.
  	\item \textbf{J.
@@
 \CustChapter*{Related publications}
+\addcontentsline{toc}{chapter}{Related publications}
 The following papers are related to the included papers, and are referred to in the text by their Roman numerals:
@@
 Holm, "Implementing Capon Beamforming on the GPU for Real-Time Cardiac Ultrasound Imaging", {\it Proc.
-IEEE Ultrasonics Symposium 2012}.
+IEEE Ultrasonics Symposium 2012} \todo{add pages}.
 
@@
 Austeng, "Adapting the Minimum variance beamformer to a graphics processing unit for active sonar imaging systems", {\it Proc.
-Meetings on Acoustics 2013}.
+Meetings on Acoustics 2013} \todo{add pages?}.
 
-	\item G. Kiss, E. Steen, \textbf{J. P \AA{}sen} and H. Torp, "GPU volume rendering in 3D echocardiography: real-time pre-processing and ray-casting", {\it Proc. IEEE Ultrasonics Symposium 2010}.
+	\item G.
+Kiss, E.
+Steen, \textbf{J.
+P \AA{}sen} and H.
+Torp, "GPU volume rendering in 3D echocardiography: real-time pre-processing and ray-casting", {\it Proc.
+IEEE Ultrasonics Symposium 2010} \todo{add pages}.
 
@@
 
-	\item E. S. Br\o{}nstad, \textbf{J. P. \AA{}sen}, H. Torp and G. Kiss, "Visibility driven visualization of 3D cardiac ultrasound data on the GPU", {\it Proc. IEEE Ultrasonics Symposium 2012}.
+	\item E.
+S.
+Br\o{}nstad, \textbf{J.
+P.
+\AA{}sen}, H.
+Torp and G.
+Kiss, "Visibility driven visualization of 3D cardiac ultrasound data on the GPU", {\it Proc.
+IEEE Ultrasonics Symposium 2012} \todo{add pages}.
 \end{enumerate}
@@
 \pagestyle{fancy}
+
 %
+
+\cleardoublepage
+\addcontentsline{toc}{chapter}{Contents}
+\renewcommand*\contentsname{Contents}
 \tableofcontents
+
+\clearpage
+\setcounter{page}{1}
+\renewcommand{\thepage}{\arabic{page}}
+
+\renewcommand*\thesection{\arabic{section}}
 
@@
 \begin{bibunit}[ieeetr]
-\CustChapter{Introduction} 
+\CustChapter*{Introduction} 
+\addcontentsline{toc}{chapter}{Introduction}
 \runningtitle{Introduction}
@@
 
+\setcounter{section}{0}
+\setcounter{chapter}{1}
+
 %%%% CHAPTER 2 %%%%
 \begin{bibunit}[ieeetr]
-\CustChapter{Background} 
+\CustChapter*{Background}
+\addcontentsline{toc}{chapter}{Background} 
 \runningtitle{Background}
@@
 
+\setcounter{chapter}{0}
+
 %%%% Paper 1 %%%%
 \begin{bibunit}[ieeetr]
-\CustChapter{An Optimised GPU Implementation of the MVDR Beamformer for Active Sonar Imaging}
-\runningtitle{An Optimised GPU Implementation of the MVDR Beamformer for Active Sonar Imaging}
+\CustChapterPaper{An Optimised GPU Implementation of the MVDR Beamformer for Active Sonar Imaging}
+\runningtitle{GPU MVDR Beamformer for Active Sonar Imaging}
 \authors{
 	Jo Inge Buskenes$^{1}$, 
-    	Jon Petter \AA{}sen$^{2}$, 
+    	\textbf{Jon Petter \AA{}sen}$^{2}$, 
     	Carl-Inge C. Nilsen$^{1}$ and
@@
 }
-\noindent \textit{Submitted for review to IEEE Transactions on Oceanic Engineering}
-%\input{./STB/STB_manuscript}
-\newpage\addcontentsline{toc}{section}{References}
-\putbib[bibSTB]
+\noindent \textit{IEEE Transactions on Oceanic Engineering, submitted}
+%\newpage\input{./CaponSonarGPU/CaponSonarGPU}
+%\newpage\addcontentsline{toc}{section}{References}
+\putbib[bibCaponSonarGPU]
 \end{bibunit}
@@
 \begin{bibunit}[ieeetr]
-\CustChapter{Implementing Capon Beamforming on a GPU for Real-Time Cardiac Ultrasound Imaging}
-\runningtitle{Implementing Capon Beamforming on a GPU for Real-Time Cardiac Ultrasound Imaging}
+\CustChapterPaper{Implementing Capon Beamforming on a GPU for Real-Time Cardiac Ultrasound Imaging}
+\runningtitle{GPU Capon Beamformer for Cardiac Ultrasound Imaging}
 \authors{
-	Jon Petter \AA{}sen$^{1}$,
+	\textbf{Jon Petter \AA{}sen}$^{1}$,
     	Jo Inge Buskenes$^{2}$,  
@@
 }
-\noindent \textit{IEEE Transactions on Ultrasonics, Ferroelectrics, and Frequency Control, vol. XY, January 2014}
-%\input{./MLAandAb/Aberration_HiFR}
-\newpage\addcontentsline{toc}{section}{References}
-\putbib[bibMLAandAb]
+\noindent \textit{IEEE Transactions on Ultrasonics, Ferroelectrics, and Frequency Control, vol. 61, Jan 2014}
+%\input{./CaponCardiacGPU/CaponCardiacGPU}
+%\newpage\addcontentsline{toc}{section}{References}
+\putbib[bibCaponCardiacGPU]
 \end{bibunit}
@@
 \begin{bibunit}[ieeetr]
-\CustChapter{Capon Beamforming Applied on Moving Objects - An Analysis of Local Lateral Shift-Invariance}
-\runningtitle{Capon Beamforming Applied on Moving Objects - An Analysis of Local Lateral Shift-Invariance}
+\CustChapterPaper{Capon Beamforming and Moving Objects - An Analysis of Lateral Shift-Invariance}
+\runningtitle{Capon Beamforming and Moving Objects}
 \authors{
-	Jon Petter \AA{}sen$^{1}$, 
+	\textbf{Jon Petter \AA{}sen}$^{1}$, 
     	Andreas Austeng$^{2}$ and
@@
 }
-\noindent \textit{Submitted for review to IEEE Transactions on Ultrasonics, Ferroelectrics, and Frequency Control}
-%\input{./STBandAb/STBandAberration}
-\newpage\addcontentsline{toc}{section}{References}
-\putbib[bibSTBandAb]
+\noindent \textit{IEEE Transactions on Ultrasonics, Ferroelectrics, and Frequency Control, submitted}
+%\input{./CaponMovingObjects/CaponMovingObjects}
+%\newpage\addcontentsline{toc}{section}{References}
+\putbib[bibCaponMovingObjects]
 \end{bibunit}
@@
 \begin{bibunit}[ieeetr]
-\CustChapter{Adaptive Volume Rendering of Cardiac 3D Ultrasound Images - Utilizing Blood Pool Statistics}
-\runningtitle{Adaptive Volume Rendering of Cardiac 3D Ultrasound Images - Utilizing Blood Pool Statistics}
+\CustChapterPaper{Adaptive Volume Rendering of Cardiac 3D Ultrasound Images - Utilizing Blood Pool Statistics}
+\runningtitle{Adaptive Volume Rendering of Cardiac Ultrasound}
 \authors{
-	Jon Petter \AA{}sen$^{1}$,
+	\textbf{Jon Petter \AA{}sen}$^{1}$,
 	Erik Steen$^2$,
@@
 }
-\noindent \textit{Proc. SPIE Medical Imaging 2012}
-%\input{./SPTD/StbSinglePulseTissueDoppler}
-\newpage\addcontentsline{toc}{section}{References}
-\putbib[bibSPTD]
+\noindent \textit{Proc. SPIE Medical Imaging 2012 \todo{add pages}}
+%\input{./AdaptiveVolumeRendering/AdaptiveVolumeRendering}
+%\newpage\addcontentsline{toc}{section}{References}
+\putbib[bibAdaptiveVolumeRendering]
 \end{bibunit}
@@
 \begin{bibunit}[ieeetr]
-\CustChapter{Huygens on Speed: Interactive Simulation of Ultrasound Pressure Fields}
-\runningtitle{Huygens on speed: Interactive simulation of ultrasound pressure fields}
-\authors{Jon Petter \AA{}sen$^1$ and Sverre Holm$^{1,2}$}
+\CustChapterPaper{Huygens on Speed: Interactive Simulation of Ultrasound Pressure Fields}
+\runningtitle{Huygens on speed}
+\authors{\textbf{Jon Petter \AA{}sen}$^1$ and Sverre Holm$^{1,2}$}
 {
@@
 }
-\noindent \textit{Proc. IEEE Ultrasonics Symposium 2012}
-%\input{./SPTD/StbSinglePulseTissueDoppler}
-\newpage\addcontentsline{toc}{section}{References}
-\putbib[bibSPTD]
+\noindent \textit{Proc. IEEE Ultrasonics Symposium 2012 \todo{add pages}}
+%\input{./Hos/Hos}
+%\newpage\addcontentsline{toc}{section}{References}
+\putbib[bibHos]
 \end{bibunit}

### 2014-01-04 ThesisJP/Introduction/introduction.tex
Change from variable spacing to fixed spacing in author names

@@
 \textbf{An Optimised GPU Implementation of the MVDR Beamformer for Active Sonar Imaging}\\
-J. I. Buskenes, \textbf{J. P. \AA{}sen}, C.-I. C. Nilsen and A. Austeng\\
+J.\:I.\:Buskenes, \textbf{J.\:P.\:\AA{}sen}, C.-I.\:C.\:Nilsen and A.\:Austeng\\
 {\it IEEE Transactions on Oceanic Engineering, submitted.}\\\\
@@
 \textbf{Implementing Capon Beamforming on a GPU for Real-Time Cardiac Ultrasound Imaging}\\
-\textbf{J. P. \AA{}sen}, J. I. Buskenes, C.-I. C. Nilsen, A. Austeng and S. Holm\\
+\textbf{J.\:P.\:\AA{}sen}, J.\:I.\:Buskenes, C.-I.\:C\:Nilsen, A.\:Austeng and S.\:Holm\\
 {\it IEEE Transactions on Ultrasonics, Ferroelectrics, and Frequency Control, vol. 61, no. 1, pp. \todo{add pages}, Jan. 2014.}\\\\
@@
 \textbf{Capon Beamforming and Moving Objects - An Analysis of Lateral Shift-Invariance}\\
-\textbf{J. P. \AA{}sen}, A. Austeng and S. Holm\\
+\textbf{J.\:P.\:\AA{}sen}, A.\:Austeng and S.\:Holm\\
 {\it IEEE Transactions on Ultrasonics, Ferroelectrics, and Frequency Control, submitted.}
@@
 \textbf{Adaptive Volume Rendering of Cardiac 3D Ultrasound Images - Utilizing Blood Pool Statistics}\\
-\textbf{J. P. \AA{}sen}, E. Steen, G. Kiss, A. Thorstensen and S. I. Rabben\\
+\textbf{J.\:P.\:\AA{}sen}, E.\:Steen, G.\:Kiss, A.\:Thorstensen and S.\:I.\:Rabben\\
 {\it Proc. SPIE Medical Imaging 2012, \todo{add pages}.}\\\\
@@
 \textbf{Huygens on Speed: Interactive Simulation of Ultrasound Pressure Fields}\\
-\textbf{J. P. \AA{}sen} and S. Holm
+\textbf{J.\:P.\:\AA{}sen} and S.\:Holm\\
 {\it Proc. IEEE Ultrasonics Symposium 2012 \todo{add pages}.}\\\\

### 2014-01-04 ThesisJP/Thesis.tex
Change from variable spacing to fixed spacing in author names

@@
 \renewcommand\labelenumi{\bfseries\theenumi}
-	\item J.
-I.
-Buskenes, \textbf{J.
-P.
-\AA{}sen}, C.-I.
-C.
-Nilsen and A.
-Austeng, "An optimised GPU implementation of the MVDR beamformer for active sonar imaging", {\it  IEEE Transactions on Oceanic Engineering, submitted}.
-	\item \textbf{J.
-P.
-\AA{}sen}, J.
-I.
-Buskenes, C.-I.
-C.
-Nilsen, A.
-Austeng and S.
-Holm, "Implementing Capon beamforming on a GPU for real-time cardiac ultrasound imaging", {\it IEEE Transactions on Ultrasonics, Ferroelectrics, and Frequency Control, vol.
+	\item J.\:I.\:Buskenes, \textbf{J.\:P.\:\AA{}sen}, C.-I.\:C.\:Nilsen and A.\:Austeng, "An optimised GPU implementation of the MVDR beamformer for active sonar imaging", {\it  IEEE Transactions on Oceanic Engineering, submitted}.
+	\item \textbf{J.\:P.\:\AA{}sen}, J.\:I.\:Buskenes, C.-I.\:C.\:Nilsen, A.\:Austeng and S.\:Holm, "Implementing Capon beamforming on a GPU for real-time cardiac ultrasound imaging", {\it IEEE Transactions on Ultrasonics, Ferroelectrics, and Frequency Control, vol.
 61, no.
@@
 \todo{add pages} Jan 2014}.
- 	\item \textbf{J.
-P.
-\AA{}sen}, A.
-Austeng and S.
-Holm, "Capon beamforming and moving objects - An analysis of lateral shift-invariance", {\it IEEE Transactions on Ultrasonics, Ferroelectrics, and Frequency Control, submitted}.
-	\item \textbf{J.
-P.
-\AA{}sen}, E.
-Steen, G.
-Kiss, A.
-Thorstensen and S.
-I.
-Rabben, "Adaptive volume rendering of cardiac 3D ultrasound images - utilizing blood pool statistics", {\it Proc.
+ 	\item \textbf{J.\:P.\:\AA{}sen}, A.\:Austeng and S.\:Holm, "Capon beamforming and moving objects - An analysis of lateral shift-invariance", {\it IEEE Transactions on Ultrasonics, Ferroelectrics, and Frequency Control, submitted}.
+	\item \textbf{J.\:P.\:\AA{}sen}, E.\:Steen, G.\:Kiss, A.\:Thorstensen and S.\:I.\:Rabben, "Adaptive volume rendering of cardiac 3D ultrasound images - utilizing blood pool statistics", {\it Proc.
 SPIE Medical Imaging 2012 \todo{add pages}}.
-	\item \textbf{J. P. \AA{}sen} and S. Holm, "Huygens on speed: Interactive simulation of ultrasound pressure fields", {\it Proc. IEEE Ultrasonics Symposium 2012 \todo{add pages}}.
+	\item \textbf{J.\:P.\:\AA{}sen} and S.\:Holm, "Huygens on speed: Interactive simulation of ultrasound pressure fields", {\it Proc. IEEE Ultrasonics Symposium 2012 \todo{add pages}}.
 \end{enumerate}
@@
 
-	\item \textbf{J.
-P.
-\AA{}sen}, J.
-I.
-Buskenes, C.-I.
-C.
-Nilsen, A.
-Austeng and S.
-Holm, "Implementing Capon Beamforming on the GPU for Real-Time Cardiac Ultrasound Imaging", {\it Proc.
+	\item \textbf{J.\:P.\:\AA{}sen}, J.\:I.\:Buskenes, C.-I.\:C.\:Nilsen, A.\:Austeng and S.\:Holm, "Implementing Capon Beamforming on the GPU for Real-Time Cardiac Ultrasound Imaging", {\it Proc.
 IEEE Ultrasonics Symposium 2012} \todo{add pages}.
 
-	\item J.
-I.
-Buskenes, \textbf{J.
-P.
-\AA{}sen}, C.-I.
-C.
-Nilsen and A.
-Austeng, "Adapting the Minimum variance beamformer to a graphics processing unit for active sonar imaging systems", {\it Proc.
+	\item J.\:I.\:Buskenes, \textbf{J.\:P.\:\AA{}sen}, C.-I.\:C.\:Nilsen and A.\:Austeng, "Adapting the Minimum variance beamformer to a graphics processing unit for active sonar imaging systems", {\it Proc.
 Meetings on Acoustics 2013} \todo{add pages?}.
 
-	\item G.
-Kiss, E.
-Steen, \textbf{J.
-P \AA{}sen} and H.
-Torp, "GPU volume rendering in 3D echocardiography: real-time pre-processing and ray-casting", {\it Proc.
+	\item G.\:Kiss, E.\:Steen, \textbf{J.\:P.\:\AA{}sen} and H.\:Torp, "GPU volume rendering in 3D echocardiography: real-time pre-processing and ray-casting", {\it Proc.
 IEEE Ultrasonics Symposium 2010} \todo{add pages}.
@@
 
-	\item E.
-S.
-Br\o{}nstad, \textbf{J.
-P.
-\AA{}sen}, H.
-Torp and G.
-Kiss, "Visibility driven visualization of 3D cardiac ultrasound data on the GPU", {\it Proc.
+	\item E.\:S.\:Br\o{}nstad, \textbf{J.\:P.\:\AA{}sen}, H.\:Torp and G.\:Kiss, "Visibility driven visualization of 3D cardiac ultrasound data on the GPU", {\it Proc.
 IEEE Ultrasonics Symposium 2012} \todo{add pages}.

### 2014-01-05 ThesisJP/Introduction/introduction.tex
Added symbol list. Started writing the introduction.

@@
 \section{Motivation}
+Rapid development in computer game technology and accompanying programming languages have recently provided researchers with small personal super computers, comprised in a single graphics card (\nom{GPU}{Graphics Processing Unit}).
+The latest graphics cards\footnote{As of January 2014} from both Nvidia and AMD provides close to 6 T\nom{FLOPS}{Floating point operations per second} single precision and 1.5 TFLOPS double precision.
+For double precision this is more than the worlds most powerful supercomputer provided in 1998.
+
+This immense rise in computational power and programming flexibility gives new possibilities when designing ultrasound imaging systems.
+Algorithms which previously had to be implemented in hardware for performance reasons can now be moved to software.
+It also means that algorithms, either too complicated to implement in hardware, or though to be too computationally heavy for real-time use, becomes realizable.
+It is clear that high performance programmable processors, like the GPU, have and will make future ultrasound imaging systems more flexible, cheaper to produce, and equipped with even more cutting-edge processing.
 
 \section{Aims of study}
+The aim of this thesis has been to investigate the possibility of utilizing GPUs for advanced processing in an ultrasound imaging system.
+
+%\subsubsection{Accelerate the Capon beamformer to facilitate real time imaging}
+
+%\subsubsection{}
 
@@
 \section{Discussion}
+The main contributions of this thesis are:
+\begin{enumerate}
+\item a GPU implementation of the Capon beamformer
+\item a GPU implementation of the beamspace Capon beamformer
+\item real-time beamspace Capon beamforming for cardiac ultrasound imaging
+\item the first investigation of Capon beamforming applied on multiple frames in medical ultrasound imaging (simulated, \textit{in-vitro}, and \textit{in-vivo})
+\item an investigation of shift-invariance for the Capon beamformer
+\item a method for improved shift-invariance of the Capon beamformer
+\item a method for reduced blood-pool noise in volume rendering of cardiac ultrasound volumes
+\item a GPU implementation of the adaptive volume rendering method
+\item a GPU implementation of simple paint-like simulation tool
+\item several discussions on how to utilize the GPU for computationally intense algorithms  
+\end{enumerate}
+
+Discuss all papers.
+
+Paper1
+
+Paper2
+
+Paper3
+
+Paper4
+
+Paper5
 
@@
 
+Concluding remarks...
+
 \section{Future work}
 
+All five papers...
+
 \endinput

### 2014-01-05 ThesisJP/Background/background.tex
Small layout adjustments
Finished layout.
Some writing on the introduction.

@@
 
+\section{General-purpose computing on graphics processing units}
+
+About theoretical FLOPS: Note that these are theoretical numbers and the actual throughput typically is algorithm dependent.
+
 \section {Ultrasound imaging}
@@
 
-\subsection{Adaptive Beamforming}
+\subsection{Adaptive Beamforming}\label{sec:adaptbf}
+
+Add section about the computationally complexity. How many flops are required per rx-beam etc...
 
@@
 
-\section{Simulations}
+\subsection{Adaptive volume rendering}
 
-\section{General purpose programming on graphics processing units}
+\section{Field Simulations}
 

### 2014-01-05 ThesisJP/Introduction/introduction.tex
Some writing on the introduction.

@@
 The latest graphics cards\footnote{As of January 2014} from both Nvidia and AMD provides close to 6 T\nom{FLOPS}{Floating point operations per second} single precision and 1.5 TFLOPS double precision.
-For double precision this is more than the worlds most powerful supercomputer provided in 1998.
+For double precision this is more than the worlds most powerful supercomputer could provide in 1998.
 
 This immense rise in computational power and programming flexibility gives new possibilities when designing ultrasound imaging systems.
-Algorithms which previously had to be implemented in hardware for performance reasons can now be moved to software.
-It also means that algorithms, either too complicated to implement in hardware, or though to be too computationally heavy for real-time use, becomes realizable.
-It is clear that high performance programmable processors, like the GPU, have and will make future ultrasound imaging systems more flexible, cheaper to produce, and equipped with even more cutting-edge processing.
+Algorithms which previously had to be implemented in hardware for performance reasons can now be implemented in software.
+And algorithms, either too complicated to implement in hardware, or though to be too computationally heavy for real-time use, becomes realizable.
+It is clear that high-performance programmable processors, like the GPU, already have and will continue to make future ultrasound imaging systems more flexible, cheaper to produce, and equipped with even more cutting-edge processing.
+
+The Capon beamformer is a good example of an algorithm which has been highly studied for the last decade for the application of medical ultrasound imaging \todo{cite papers}, but where the number of calculations was though to be too much for real time processing.
+No wonder, to achieve real-time processing in e.g.
+cardiac ultrasound imaging an effective processing rate of around 100 MFLOPS is required.
+This number, as explained in Background \ref{sec:adaptbf}, also grows with larger subarrays, more image pixels, and higher framerate requirements.
+But with modern GPUs these levels of processing are actually within reach with only a single card.
 
 \section{Aims of study}
-The aim of this thesis has been to investigate the possibility of utilizing GPUs for advanced processing in an ultrasound imaging system.
+The aim of this study has been to investigate the possibility of utilizing GPUs for advanced processing in an ultrasound imaging system.
+
+My main focus have been the Capon beamformer, and the problem of making this computationally expensive algorithm available for real-time ultrasound imaging.
+However, two additional methods have also been included in the thesis.
+The first is adaptive volume rendering of cardiac ultrasound volumes and the latter is simulation of dense ultrasound pressure fields.
+Both share the property of being computationally heavy, but they also consist of many independent computations which is perfect for parallel GPU computing.
+
+Well on my way into the project, when a real-time Capon beamformer was realized, and loops of images for the first time were processed, new issues where discovered and had to be solved.
+This led me to a more theoretical study of the Capon beamformer, and how to obtain high lateral resolution while preserving the important shift-invariant property of medical ultrasound imaging.
 
@@
 Summary of paper...
+
+To facilitate study of in-vitro performance.
 
@@
 \item a method for reduced blood-pool noise in volume rendering of cardiac ultrasound volumes
-\item a GPU implementation of the adaptive volume rendering method
-\item a GPU implementation of simple paint-like simulation tool
+\item a GPU implementation of this adaptive volume rendering method
+\item a GPU implementation of simple paint-like simulation tool for rapid visualization ultrasound pressure fields
 \item several discussions on how to utilize the GPU for computationally intense algorithms  
@@
 Paper2
+Discuss why cardiac ultrasound is selected and if full Capon will be real-time for other modalities.
 

### 2014-01-08 ThesisJP/Introduction/introduction.tex
Started writing summary of first paper. Add nomenclatur files to gitignor.

@@
 \section{Motivation}
-Rapid development in computer game technology and accompanying programming languages have recently provided researchers with small personal super computers, comprised in a single graphics card (\nom{GPU}{Graphics Processing Unit}).
+A rapid development in computer game technology and accompanying programming languages have recently provided researchers with small personal super computers, comprised in a single graphics card (\nom{GPU}{Graphics Processing Unit}).
 The latest graphics cards\footnote{As of January 2014} from both Nvidia and AMD provides close to 6 T\nom{FLOPS}{Floating point operations per second} single precision and 1.5 TFLOPS double precision.
@@
 The Capon beamformer is a good example of an algorithm which has been highly studied for the last decade for the application of medical ultrasound imaging \todo{cite papers}, but where the number of calculations was though to be too much for real time processing.
-No wonder, to achieve real-time processing in e.g.
+No wonder, to achieve real-time processing for e.g.
 cardiac ultrasound imaging an effective processing rate of around 100 MFLOPS is required.
 This number, as explained in Background \ref{sec:adaptbf}, also grows with larger subarrays, more image pixels, and higher framerate requirements.
-But with modern GPUs these levels of processing are actually within reach with only a single card.
+With modern GPUs these levels of processing are available within a single card.
 
 \section{Aims of study}
-The aim of this study has been to investigate the possibility of utilizing GPUs for advanced processing in an ultrasound imaging system.
+The overall aim of this study has been to investigate the possibility of utilizing GPUs for advanced processing in an ultrasound imaging system.
 
-My main focus have been the Capon beamformer, and the problem of making this computationally expensive algorithm available for real-time ultrasound imaging.
+My main focus have been the Capon beamformer, and the problem of making this computationally intense algorithm available for real-time ultrasound imaging.
 However, two additional methods have also been included in the thesis.
@@
 Well on my way into the project, when a real-time Capon beamformer was realized, and loops of images for the first time were processed, new issues where discovered and had to be solved.
-This led me to a more theoretical study of the Capon beamformer, and how to obtain high lateral resolution while preserving the important shift-invariant property of medical ultrasound imaging.
+This led me to a more theoretical study of the Capon beamformer in \textbf{Paper III}, with special attention on how to obtain high lateral resolution while preserving the important shift-invariant property of medical ultrasound imaging.
+Shift-invariant behavior is crucial if the method is going to be applied for live scanning.
 
@@
 {\it IEEE Transactions on Oceanic Engineering, submitted.}\\\\
-Summary of paper...
+The first paper describes in details how the Capon beamformer is mapped to GPU architecture.
+Even though the paper is written within the field of active sonar imaging, the implementation described is also applicable to other active ultrasound imaging systems.
+A similar discussion for cardiac ultrasound imaging can be found in \textbf{Paper VI}.
+Active sonar imaging typically differs from medical ultrasound imaging by a lower real-time requirement, element count, center frequency, and relative bandwidth.
+
+The estimation of the spatial covariance matrix receives special attention.
+In previous literature on Capon beamforming the matrix inversion is always regarded as the most computationally complex piece.
 

### 2014-01-09 ThesisJP/Introduction/introduction.tex
Extended summary of Paper 1.

@@
 A rapid development in computer game technology and accompanying programming languages have recently provided researchers with small personal super computers, comprised in a single graphics card (\nom{GPU}{Graphics Processing Unit}).
-The latest graphics cards\footnote{As of January 2014} from both Nvidia and AMD provides close to 6 T\nom{FLOPS}{Floating point operations per second} single precision and 1.5 TFLOPS double precision.
+The latest graphics cards\footnote{As of January 2014} from both \nom{Nvidia}{Graphics card producer} and \nom{AMD}{Graphics card producer} provides close to 6 T\nom{FLOPS}{Floating point operations per second} single precision and 1.5 TFLOPS double precision.
 For double precision this is more than the worlds most powerful supercomputer could provide in 1998.
@@
 This number, as explained in Background \ref{sec:adaptbf}, also grows with larger subarrays, more image pixels, and higher framerate requirements.
-With modern GPUs these levels of processing are available within a single card.
+With modern GPUs these levels of processing are finally available within a single card.
 
@@
 
-My main focus have been the Capon beamformer, and the problem of making this computationally intense algorithm available for real-time ultrasound imaging.
-However, two additional methods have also been included in the thesis.
-The first is adaptive volume rendering of cardiac ultrasound volumes and the latter is simulation of dense ultrasound pressure fields.
+My main focus have been the Capon beamformer, and the problem of making this computationally intense algorithm available for real-time ultrasound imaging (\textbf{Paper\,I} and \textbf{II}).
+However, two additional methods were also explored.
+The first is adaptive volume rendering of cardiac ultrasound volumes (\textbf{Paper\,IV}) and the latter is simulation of dense ultrasound pressure fields (\textbf{Paper\,V}).
 Both share the property of being computationally heavy, but they also consist of many independent computations which is perfect for parallel GPU computing.
@@
 Well on my way into the project, when a real-time Capon beamformer was realized, and loops of images for the first time were processed, new issues where discovered and had to be solved.
-This led me to a more theoretical study of the Capon beamformer in \textbf{Paper III}, with special attention on how to obtain high lateral resolution while preserving the important shift-invariant property of medical ultrasound imaging.
-Shift-invariant behavior is crucial if the method is going to be applied for live scanning.
+This led me to a more theoretical study of the Capon beamformer (\textbf{Paper III}), with special attention on how to obtain high lateral resolution while preserving the important shift-invariant property of medical ultrasound imaging.
+Shift-invariant behavior is crucial if the method is ever to be applied for live scanning.
 
@@
 The first paper describes in details how the Capon beamformer is mapped to GPU architecture.
-Even though the paper is written within the field of active sonar imaging, the implementation described is also applicable to other active ultrasound imaging systems.
+Even though the paper is written within the field of active sonar imaging, the implementation described is applicable to a vast range of active imaging systems.
 A similar discussion for cardiac ultrasound imaging can be found in \textbf{Paper VI}.
@@
 The estimation of the spatial covariance matrix receives special attention.
-In previous literature on Capon beamforming the matrix inversion is always regarded as the most computationally complex piece.
+In previous literature on Capon beamforming the matrix inversion has always been regarded as the most computationally complex piece.
+In this paper we show that estimation of the sample covariance matrix is actually the most complex part when spatial and temporal smoothing is applied with common parameters.
 

### 2014-01-10 ThesisJP/Introduction/introduction.tex
Started on summary of second paper.

@@
 
-Well on my way into the project, when a real-time Capon beamformer was realized, and loops of images for the first time were processed, new issues where discovered and had to be solved.
+Well on my way into the project, when a real-time Capon beamformer\footnote{The name ''Capon beamformer'' is due to work by J.
+Capon \todo{add citation} on seismic arrays \todo{Move to background.}} was realized, and loops of images for the first time were processed, new issues where discovered and had to be solved.
 This led me to a more theoretical study of the Capon beamformer (\textbf{Paper III}), with special attention on how to obtain high lateral resolution while preserving the important shift-invariant property of medical ultrasound imaging.
@@
 The first paper describes in details how the Capon beamformer is mapped to GPU architecture.
-Even though the paper is written within the field of active sonar imaging, the implementation described is applicable to a vast range of active imaging systems.
+Even though the paper is written within the field of active \nom{sonar}{SOund Navigation And Ranging (usually under water)} imaging, the depicted implementation is applicable to a vast range of active imaging systems.
 A similar discussion for cardiac ultrasound imaging can be found in \textbf{Paper VI}.
@@
 
-The estimation of the spatial covariance matrix receives special attention.
+The estimation of the spatial covariance matrix receives special attention in this work.
 In previous literature on Capon beamforming the matrix inversion has always been regarded as the most computationally complex piece.
-In this paper we show that estimation of the sample covariance matrix is actually the most complex part when spatial and temporal smoothing is applied with common parameters.
+In this paper we show that estimation of the sample covariance matrix is actually the most complex part when spatial and temporal smoothing is included with common parameters.
+It is then shown how the arithmetic complexity of this estimation process can be reduced from cubic to square.
+Finally, we give an in-depth analysis of the arithmetic throughput on multiple platforms.
+The reported throughput of 1 Mpx/s is, as far as the authors know, enough to provide real-time processing for all known sonar scan sequences.
 
@@
 {\it IEEE Transactions on Ultrasonics, Ferroelectrics, and Frequency Control, vol. 61, no. 1, pp. \todo{add pages}, Jan. 2014.}\\\\
-Summary of paper...
+The second paper aims at achieving real-time Capon beamforming for cardiac ultrasound imaging.
+Such an implementation will facilitates further study of the methods \textit{in vivo} performance in medical ultrasound imaging.
+In \textbf{Paper I} we explored arrays with no more than 32 elements.
+A linear array for cardiac ultrasound imaging typically has 64 or more elements.
+In \textbf{Paper VI}, which is summarized in \textbf{Paper II}, it is shown that our implementation does not reach the level of throughput required for real-time Capon beamforming of cardiac ultrasound.
+The matrices that has to be inverted becomes to large, and the number frames that needs to be processed per second is to high.
 
-To facilitate study of in-vitro performance.
+To reduce the matrix size we take, in this paper, advantage of the beamspace version of the Capon beamformer (\nom{BS-Capon}{Beamspace Capon beamforming}), and implement it on the GPU.
+For parameters previously derived to give similar performance for BS-Capon and element space Capon (\nom{ES-Capon}{Element Space Capon}), we show that real-time BS-Capon beamforming is feasible for cardiac ultrasound imaging.
 

### 2014-01-11 ThesisJP/Introduction/introduction.tex
Started writing summary of third paper.

@@
 
-The Capon beamformer is a good example of an algorithm which has been highly studied for the last decade for the application of medical ultrasound imaging \todo{cite papers}, but where the number of calculations was though to be too much for real time processing.
+The Capon beamformer\footnote{The name ''Capon beamformer'' is due to work by J.
+Capon \todo{add citation} on seismic arrays \todo{Move to background.}}  is a good example of an algorithm which has been highly studied for the last decade for the application of medical ultrasound imaging \todo{cite papers}, but where the number of calculations was though to be too much for real time processing.
 No wonder, to achieve real-time processing for e.g.
@@
 
-Well on my way into the project, when a real-time Capon beamformer\footnote{The name ''Capon beamformer'' is due to work by J.
-Capon \todo{add citation} on seismic arrays \todo{Move to background.}} was realized, and loops of images for the first time were processed, new issues where discovered and had to be solved.
+Well on my way into the project, when a real-time Capon beamformer was realized, and loops of images for the first time were processed, new issues where discovered and had to be solved.
 This led me to a more theoretical study of the Capon beamformer (\textbf{Paper III}), with special attention on how to obtain high lateral resolution while preserving the important shift-invariant property of medical ultrasound imaging.
@@
 A similar discussion for cardiac ultrasound imaging can be found in \textbf{Paper VI}.
-Active sonar imaging typically differs from medical ultrasound imaging by a lower real-time requirement, element count, center frequency, and relative bandwidth.
+Active sonar imaging typically differs from medical ultrasound imaging by a lower real-time requirement and fewer array elements.
+This makes it somewhat easier to reach our goal of real-time processing.
 
-The estimation of the spatial covariance matrix receives special attention in this work.
+The estimation of the spatial covariance matrix receives special attention in this study.
 In previous literature on Capon beamforming the matrix inversion has always been regarded as the most computationally complex piece.
@@
 Finally, we give an in-depth analysis of the arithmetic throughput on multiple platforms.
-The reported throughput of 1 Mpx/s is, as far as the authors know, enough to provide real-time processing for all known sonar scan sequences.
+The reported throughput of 1 Mpx/s on a high-end GPU is, as far as the authors know, enough to provide real-time processing for all known sonar scan sequences.
+However, the number of effective FLOPS is only 5 \% of the theoretical throughput of the target GPU.
+This result shows how hard it is to reach the theoretical level of throughput, and that how close one can get is highly algorithm depended.
 
@@
 The second paper aims at achieving real-time Capon beamforming for cardiac ultrasound imaging.
-Such an implementation will facilitates further study of the methods \textit{in vivo} performance in medical ultrasound imaging.
-In \textbf{Paper I} we explored arrays with no more than 32 elements.
+Such an implementation will facilitates further study of the method's \textit{in vivo} performance in medical ultrasound imaging.
+This is something which has been sought for the past decade \todo{Add citations}.
+In \textbf{Paper I} we investigated arrays with no more than 32 elements.
 A linear array for cardiac ultrasound imaging typically has 64 or more elements.
-In \textbf{Paper VI}, which is summarized in \textbf{Paper II}, it is shown that our implementation does not reach the level of throughput required for real-time Capon beamforming of cardiac ultrasound.
-The matrices that has to be inverted becomes to large, and the number frames that needs to be processed per second is to high.
+In \textbf{Paper VI}, which is summarized in \textbf{Paper II}, it is shown that our implementation from  \textbf{Paper I} does not reach the level of throughput required for real-time Capon beamforming of cardiac ultrasound imaging.
+The matrices that have to be inverted becomes to large, and the number of frames that needs to be processed per second is to many.
 
@@
 For parameters previously derived to give similar performance for BS-Capon and element space Capon (\nom{ES-Capon}{Element Space Capon}), we show that real-time BS-Capon beamforming is feasible for cardiac ultrasound imaging.
+The reported processing throughput is able to keep up with the acquisition frame rate for a typical cardiac ultrasound imaging system with 64 and 96 element linear arrays.
+\textbf{Paper II} and \textbf{VI} also presents, for the first time, videos where the Capon beamformer has been applied on loops of medical ultrasound images, both simulations and \textit{in vivo}.
 
@@
 \textbf{J.\:P.\:\AA{}sen}, A.\:Austeng and S.\:Holm\\
-{\it IEEE Transactions on Ultrasonics, Ferroelectrics, and Frequency Control, submitted.}
-Summary of paper...
+{\it IEEE Transactions on Ultrasonics, Ferroelectrics, and Frequency Control, submitted.}\\\\
+In the third paper we study the shift-invariant property of an imaging system based on the Capon beamformer.
+As mentioned, shift-invariant imaging is essential if the method is ever to be adopted.
+The paper was written based on observations of aliasing artifacts when imaging bright point scatterers \textit{in vitro} and in simulations.
+In earlier work on Capon beamforming for medical ultrasound imaging the real-time nature of ultrasound imaging has been ignored, and a large degree of oversampling has often been applied in simulations.
 

### 2014-01-12 ThesisJP/Introduction/introduction.tex
Writing summary of paper 3.

@@
 The paper was written based on observations of aliasing artifacts when imaging bright point scatterers \textit{in vitro} and in simulations.
-In earlier work on Capon beamforming for medical ultrasound imaging the real-time nature of ultrasound imaging has been ignored, and a large degree of oversampling has often been applied in simulations.
+The points were twinkling like stars in the sky.
+In earlier work on Capon beamforming for medical ultrasound imaging the real-time nature of ultrasound imaging has been ignored, and a large degree of oversampling on transmit has often been applied without comments.
+Oversampling on transmit is not an option if the imaging modality is used to image rapidly varying objects.
+
+In \textbf{Paper III} we investigate how the Capon beamforming achieves its super resolution, and how this makes it highly sensitive to differences between the assumed steering vector and the signal propagation vector.
+Earlier work, focusing on single images, has typically referred to this effect as speed-of-sound errors, but it also comes naturally into play when objects moves from frame to frame and the beam density is insufficient.
+With a narrow band and farfield model we show that 25 times oversampling is required to obtain the same lateral gain variations as with \nom{DAS}{Delay-and-sum beamforming} beamforming.
+These calculations are done based on typical parameters for the Capon beamformer.
+For
 

### 2014-01-12 ThesisJP/Introduction/introduction.tex
Finished summary of paper 4 and 5.

@@
 {\it IEEE Transactions on Oceanic Engineering, submitted.}\\\\
-The first paper describes in details how the Capon beamformer is mapped to GPU architecture.
+The first paper describes in details how we have mapped the Capon beamformer to GPU architecture.
 Even though the paper is written within the field of active \nom{sonar}{SOund Navigation And Ranging (usually under water)} imaging, the depicted implementation is applicable to a vast range of active imaging systems.
@@
 The second paper aims at achieving real-time Capon beamforming for cardiac ultrasound imaging.
-Such an implementation will facilitates further study of the method's \textit{in vivo} performance in medical ultrasound imaging.
+Such an implementation will facilitates further study of the method's \textit{in vivo} performance for this modality.
 This is something which has been sought for the past decade \todo{Add citations}.
@@
 For parameters previously derived to give similar performance for BS-Capon and element space Capon (\nom{ES-Capon}{Element Space Capon}), we show that real-time BS-Capon beamforming is feasible for cardiac ultrasound imaging.
-The reported processing throughput is able to keep up with the acquisition frame rate for a typical cardiac ultrasound imaging system with 64 and 96 element linear arrays.
-\textbf{Paper II} and \textbf{VI} also presents, for the first time, videos where the Capon beamformer has been applied on loops of medical ultrasound images, both simulations and \textit{in vivo}.
+The reported processing throughput is able to keep up with the acquisition frame rate in a typical cardiac ultrasound imaging system equipped with 64 and 96 element linear arrays.
+\textbf{Paper II} and \textbf{VI} also presents, for the first time, videos where the ES-Capon and BS-Capon beamformer have been applied on loops of simulated and \textit{in vivo} medical ultrasound images.
 
@@
 The paper was written based on observations of aliasing artifacts when imaging bright point scatterers \textit{in vitro} and in simulations.
-The points were twinkling like stars in the sky.
+The points were observed to twinkle like stars in the sky.
 In earlier work on Capon beamforming for medical ultrasound imaging the real-time nature of ultrasound imaging has been ignored, and a large degree of oversampling on transmit has often been applied without comments.
@@
 
-In \textbf{Paper III} we investigate how the Capon beamforming achieves its super resolution, and how this makes it highly sensitive to differences between the assumed steering vector and the signal propagation vector.
+In this paper we investigate how the Capon beamforming achieves its super resolution, and how this makes it highly sensitive to differences between the assumed steering vector and the signal propagation vector.
 Earlier work, focusing on single images, has typically referred to this effect as speed-of-sound errors, but it also comes naturally into play when objects moves from frame to frame and the beam density is insufficient.
-With a narrow band and farfield model we show that 25 times oversampling is required to obtain the same lateral gain variations as with \nom{DAS}{Delay-and-sum beamforming} beamforming.
-These calculations are done based on typical parameters for the Capon beamformer.
-For
+With a narrowband and farfield model we show that the Capon beamformer suffers from beam-to-beam gain variations as large as 27 dB for parameters previously used to obtain high resolution medical ultrasound images.
+To obtain the same lateral gain variations as with \nom{DAS}{Delay-and-sum beamforming} beamforming, 25 times oversampling is shown to be required.
+For broadband nearfield simulations the same gain variations are observed.
+Finally we show that the gain variation can be reduced, hence improved shift-invariance, by oversampling on receive using phase rotation steering.
+This without affecting the acquisition frame rate and with a minor increase in computationally complexity.
+The method is successfully applied on simulations and \textit{in vitro} phantom data.
 
@@
 {\it Proc. SPIE Medical Imaging 2012, \todo{add pages}.}\\\\
-Summary of paper...
+The fourth paper propose an adaptive volume rendering method based on statistics derived from the blood pool of the left ventricle.
+Noise located in the blood pool tends to occlude cardiac tissue when rendered in 3D, and is often impossible to remove by adjusting global rendering parameters.
+This because ultrasound signals have high variability within blood as well as high variability within tissue, even when heavily smoothed.
+Delineated of the blood pool is done by a state estimation algorithm, capable of tracking the endocardium in real time.
+The final visualizations are compared, with respect to location of tissue, with the result from a state-of-the-art endocardium segmentation tool.
+The paper presents both quantitative and qualitative results supporting the method's improved capabilities of reducing tissue dropouts and spurious structures inside the blood pool, and by that maximizing the amount of visible endocardium tissue.
+The proposed method is implemented on a GPU which results in interactive frame rates.
 
@@
 {\it Proc. IEEE Ultrasonics Symposium 2012 \todo{add pages}.}\\\\
-Summary of paper...
+The fifth paper presents an interactive simulations tool capable of simulating dense pressure fields from ultrasound arrays in real time.
+Simulation tools are heavily used both by researchers and probe manufactures in order to test out new algorithms and to deduce the performance of new arrays prior to manufacturing.
+Unfortunately, simulation tools performing dense simulations are inherently slow.
+The tool is based on Huygens' principle which describes diffraction caused by a slit as the super position of several point sources located in the slit opening.
+Thus, the simulator works by accumulating the contribution from a collection of point sources in a set of observation points.
+This is the core engine of the simulator and it is, in the context of this thesis, of course running on a GPU.
+How the source and observation points are laid out is up to the client to decide.
+The simulator is provided with both a paint-like interface and a Matlab interface for precise scripting of array configurations.
+The main contributions in this paper is the increased performance of the GPU implementation compared with a CPU and Matlab (The Ultrasim toolbox) version.
+Second, the paint-like interface provides an interactive and neat way of demonstrating array beamforming principle.
+We believe this will be of great value both for array designers and teachers.
 
@@
 \item a GPU implementation of the beamspace Capon beamformer
-\item real-time beamspace Capon beamforming for cardiac ultrasound imaging
+\item that real-time beamspace Capon beamforming is achieved for cardiac ultrasound imaging
 \item the first investigation of Capon beamforming applied on multiple frames in medical ultrasound imaging (simulated, \textit{in-vitro}, and \textit{in-vivo})
-\item an investigation of shift-invariance for the Capon beamformer
+\item an investigation of shift-invariance when using the Capon beamformer
 \item a method for improved shift-invariance of the Capon beamformer
-\item a method for reduced blood-pool noise in volume rendering of cardiac ultrasound volumes
+\item a method for reduced blood-pool noise in volume renderings of cardiac ultrasound volumes
 \item a GPU implementation of this adaptive volume rendering method
 \item a GPU implementation of simple paint-like simulation tool for rapid visualization ultrasound pressure fields
-\item several discussions on how to utilize the GPU for computationally intense algorithms  
+\item several discussions on how to utilize the GPU for computationally intense algorithms for advanced ultrasound imaging 
 \end{enumerate}
@@
 Paper5
+This is not an adaptive algorithm.
+
+Mark that code for GPU-Capon and Huygens on speed have been made open source.
 

### 2014-01-12 ThesisJP/Introduction/introduction.tex
Adjusted title.

@@
 {\it IEEE Transactions on Oceanic Engineering, submitted.}\\\\
-The first paper describes in details how we have mapped the Capon beamformer to GPU architecture.
+The first paper describes in details how we mapped the Capon beamformer to GPU architecture.
 Even though the paper is written within the field of active \nom{sonar}{SOund Navigation And Ranging (usually under water)} imaging, the depicted implementation is applicable to a vast range of active imaging systems.
@@
 Mark that code for GPU-Capon and Huygens on speed have been made open source.
+It has over 200 downloads and soon 1000 views on youtube.
 

### 2014-01-15 ThesisJP/Introduction/introduction.tex
Small adjustment in summary of papers.

@@
 The reported throughput of 1 Mpx/s on a high-end GPU is, as far as the authors know, enough to provide real-time processing for all known sonar scan sequences.
-However, the number of effective FLOPS is only 5 \% of the theoretical throughput of the target GPU.
+However, the number of effective FLOPS is only 4 \% and 14 \% of the theoretical throughput of the target GPU for the solver and covariance estimator respectively.
 This result shows how hard it is to reach the theoretical level of throughput, and that how close one can get is highly algorithm depended.
@@
 Simulation tools are heavily used both by researchers and probe manufactures in order to test out new algorithms and to deduce the performance of new arrays prior to manufacturing.
-Unfortunately, simulation tools performing dense simulations are inherently slow.
-The tool is based on Huygens' principle which describes diffraction caused by a slit as the super position of several point sources located in the slit opening.
+Dense simulations are unfortunately inherently slow.
+The presented tool is based on Huygens' principle which describes diffraction caused by a slit as the super position of several point sources located inside the opening.
 Thus, the simulator works by accumulating the contribution from a collection of point sources in a set of observation points.
-This is the core engine of the simulator and it is, in the context of this thesis, of course running on a GPU.
-How the source and observation points are laid out is up to the client to decide.
-The simulator is provided with both a paint-like interface and a Matlab interface for precise scripting of array configurations.
-The main contributions in this paper is the increased performance of the GPU implementation compared with a CPU and Matlab (The Ultrasim toolbox) version.
-Second, the paint-like interface provides an interactive and neat way of demonstrating array beamforming principle.
+This is the core engine of the simulator, and it is, in the context of this thesis, of course running on a GPU.
+How the source and observation points are laid out is up to the user.
+The simulator is provided with both a paint-like interface for interactive drawing of arrays, and a Matlab interface for precise scripting of array configurations.
+The main contributions in this paper is the increased performance of the GPU implementation compared with a CPU and Matlab\footnote{The Ultrasim toolbox \todo{add citation}} version.
+Second, the paint-like interface provides a neat way of demonstrating array beamforming principle in real time.
 We believe this will be of great value both for array designers and teachers.
@@
 Paper1
+Solving is done by a batched Gauss Jordan solver implemented by Nvidia.
+We decided to focus on the sample covariance estimation, and not trying to beat Nvidia's solver.
+A natural next step would be to analyze the current solver, and figure out if any optimizations are possible.
+Since the sample covariance matrix is hermitian, a solver based on Cholesky decomposition might improve the throughput for solving.
 
@@
 Discuss why cardiac ultrasound is selected and if full Capon will be real-time for other modalities.
+
+No detailed evaluation of \textit{in vivo} performance.
+In cardiac application it has been proven hard to transfer results obtained in simulations to \textit{in vivo}.
+The improvements given is minimal with respect to the amount of calculation required.
 

### 2014-01-15 ThesisJP/Introduction/introduction.tex
Started writing discussion section.

@@
 However, the number of effective FLOPS is only 4 \% and 14 \% of the theoretical throughput of the target GPU for the solver and covariance estimator respectively.
-This result shows how hard it is to reach the theoretical level of throughput, and that how close one can get is highly algorithm depended.
 
@@
 
-Discuss all papers.
+Even with all our effort in \textbf{Paper I} we were only able to achieve a fraction of the maximum throughput of the target GPU.
+This result shows how hard it is to reach the theoretical level of throughput, and that how close one can get is highly algorithm depended.
+It might be ways of improving these numbers, but both the solver and covariance estimation step consist of several points where fine granular synchronization and serialized steps are needed.
+This will restrict even the most fine tuned implementation from reaching the maximum throughput of the GPU.
+An interesting observation can be found in Fig.\,10 in \textbf{Paper I}.
+Here we see how the throughput of the covariance estimation kernel is more than three times higher than the theoretical throughput of a modern, high-end, \nom{CPU}{Central processing unit}.
+This clearly shows the benefit of modern GPU computing.
 
-Paper1
-Solving is done by a batched Gauss Jordan solver implemented by Nvidia.
-We decided to focus on the sample covariance estimation, and not trying to beat Nvidia's solver.
-A natural next step would be to analyze the current solver, and figure out if any optimizations are possible.
-Since the sample covariance matrix is hermitian, a solver based on Cholesky decomposition might improve the throughput for solving.
+Solving was done by a batched Gauss Jordan solver implemented by Nvidia.
+We decided to focus on the covariance estimation step, and did not try to beat Nvidia at home.
+Nevertheless, with the measured throughput in mind, a natural next step would be to analyze the current solver, and figure out if any further optimization is possible.
+Since the sample covariance matrix is hermitian, a solver based on Cholesky decomposition might also help to improve the solver's throughput.
+\\\\
+Paper2
 
-Paper2
 Discuss why cardiac ultrasound is selected and if full Capon will be real-time for other modalities.
@@
 The improvements given is minimal with respect to the amount of calculation required.
+\\\\
+Paper3
+\\\\
+Paper4
+\\\\
+Paper5
 
-Paper3
-
-Paper4
-
-Paper5
 This is not an adaptive algorithm.
@@
 
+Integrated beamforming and visualization. Using adaptive beamforming to improve e.g. speckle tracking and segmentation. Use adaptive weight statistics to guide image smoothing.
+
 \endinput

### 2014-01-18 ThesisJP/Introduction/introduction.tex
Continued writing on introduction

@@
 This number, as explained in Background \ref{sec:adaptbf}, also grows with larger subarrays, more image pixels, and higher framerate requirements.
-With modern GPUs these levels of processing are finally available within a single card.
+With modern GPUs these levels of processing are finally available in a single card.
 
@@
 However, two additional methods were also explored.
-The first is adaptive volume rendering of cardiac ultrasound volumes (\textbf{Paper\,IV}) and the latter is simulation of dense ultrasound pressure fields (\textbf{Paper\,V}).
+The first is adaptive visualization of cardiac ultrasound volumes (\textbf{Paper\,IV}) and the latter is simulation of dense ultrasound pressure fields (\textbf{Paper\,V}).
 Both share the property of being computationally heavy, but they also consist of many independent computations which is perfect for parallel GPU computing.
@@
 In previous literature on Capon beamforming the matrix inversion has always been regarded as the most computationally complex piece.
-In this paper we show that estimation of the sample covariance matrix is actually the most complex part when spatial and temporal smoothing is included with common parameters.
+In this paper we show that estimation of the sample covariance matrix is actually the most complex part when spatial and temporal smoothing are included with common parameters.
 It is then shown how the arithmetic complexity of this estimation process can be reduced from cubic to square.
 Finally, we give an in-depth analysis of the arithmetic throughput on multiple platforms.
-The reported throughput of 1 Mpx/s on a high-end GPU is, as far as the authors know, enough to provide real-time processing for all known sonar scan sequences.
-However, the number of effective FLOPS is only 4 \% and 14 \% of the theoretical throughput of the target GPU for the solver and covariance estimator respectively.
+Despite our effort, the number of effective FLOPS is only 4 \% and 14 \% of the theoretical throughput of the target GPU for the solver and covariance estimator respectively.
+Still, the reported throughput of 1 Mpx/s on a high-end GPU is enough to provide real-time processing for common sonar scan sequences.
 
@@
 To reduce the matrix size we take, in this paper, advantage of the beamspace version of the Capon beamformer (\nom{BS-Capon}{Beamspace Capon beamforming}), and implement it on the GPU.
-For parameters previously derived to give similar performance for BS-Capon and element space Capon (\nom{ES-Capon}{Element Space Capon}), we show that real-time BS-Capon beamforming is feasible for cardiac ultrasound imaging.
+For parameters previously derived to give similar performance for BS-Capon and element-space Capon (\nom{ES-Capon}{Element Space Capon}), we show that real-time BS-Capon beamforming is feasible for cardiac ultrasound imaging.
 The reported processing throughput is able to keep up with the acquisition frame rate in a typical cardiac ultrasound imaging system equipped with 64 and 96 element linear arrays.
@@
 The points were observed to twinkle like stars in the sky.
-In earlier work on Capon beamforming for medical ultrasound imaging the real-time nature of ultrasound imaging has been ignored, and a large degree of oversampling on transmit has often been applied without comments.
+In earlier work on Capon beamforming for medical ultrasound imaging the real-time nature of ultrasound imaging has often been ignored, and a large degree of oversampling on transmit might have been applied without comments.
 Oversampling on transmit is not an option if the imaging modality is used to image rapidly varying objects.
@@
 The fifth paper presents an interactive simulations tool capable of simulating dense pressure fields from ultrasound arrays in real time.
-Simulation tools are heavily used both by researchers and probe manufactures in order to test out new algorithms and to deduce the performance of new arrays prior to manufacturing.
-Dense simulations are unfortunately inherently slow.
-The presented tool is based on Huygens' principle which describes diffraction caused by a slit as the super position of several point sources located inside the opening.
+Simulation tools are heavily used by researchers in order to test out new algorithms and to deduce the performance of new array designs prior to manufacturing.
+But dense simulations also involves a lot of waiting.
+With the presented simulator we try to reduce the time for calculating pressure fields by means of GPU acceleration.
+The tool is based on Huygens' principle which describes diffraction caused by a slit as the super position of several point sources located inside the opening.
 Thus, the simulator works by accumulating the contribution from a collection of point sources in a set of observation points.
-This is the core engine of the simulator, and it is, in the context of this thesis, of course running on a GPU.
-How the source and observation points are laid out is up to the user.
+How the source and observation points are laid out is up to the user to decide.
 The simulator is provided with both a paint-like interface for interactive drawing of arrays, and a Matlab interface for precise scripting of array configurations.
@@
 Second, the paint-like interface provides a neat way of demonstrating array beamforming principle in real time.
-We believe this will be of great value both for array designers and teachers.
+We believe this to be of great value both for array designers and teachers.
 
@@
 \item a GPU implementation of simple paint-like simulation tool for rapid visualization ultrasound pressure fields
-\item several discussions on how to utilize the GPU for computationally intense algorithms for advanced ultrasound imaging 
 \end{enumerate}
+
+In addition, the thesis provides several discussions on how to utilize the GPU to accelerate advanced ultrasound imaging algorithms.
+Now follows a discussion of the five papers that constitutes this thesis.
 
@@
 This result shows how hard it is to reach the theoretical level of throughput, and that how close one can get is highly algorithm depended.
-It might be ways of improving these numbers, but both the solver and covariance estimation step consist of several points where fine granular synchronization and serialized steps are needed.
+It might be ways of improving these numbers, but both the solver and covariance estimation step consist of several points where fine granular synchronization and serialized instructions are needed.
 This will restrict even the most fine tuned implementation from reaching the maximum throughput of the GPU.
 An interesting observation can be found in Fig.\,10 in \textbf{Paper I}.
-Here we see how the throughput of the covariance estimation kernel is more than three times higher than the theoretical throughput of a modern, high-end, \nom{CPU}{Central processing unit}.
+Here we see how the achieved throughput of the covariance estimation kernel is more than three times higher than the theoretical throughput of a modern, high-end, \nom{CPU}{Central processing unit}.
 This clearly shows the benefit of modern GPU computing.
@@
 Solving was done by a batched Gauss Jordan solver implemented by Nvidia.
-We decided to focus on the covariance estimation step, and did not try to beat Nvidia at home.
+In our work we focused on the covariance estimation step, and did not try to beat Nvidia at home.
 Nevertheless, with the measured throughput in mind, a natural next step would be to analyze the current solver, and figure out if any further optimization is possible.
@@
 \\\\
-Paper2
+In \textbf{Paper II} we present the first medical ultrasound loops processed with ES- and BS-Capon beamforming.
+The paper does not provide a detailed evaluation of the method's \textit{in vivo} performance.
+Yet, initially it has been proven hard to transfer results obtained in simulations to \textit{in vivo} applications.
+The improvements seen are minimal, especially with the amount of calculation required in mind.
+The method is clearly able to decrees the lateral with of lateral point scatterers and thin structures in addition to reduce sidelobes cluttering tissue edges.
+On the other hand, the overall contrast is visually not improved and sometimes worse.
+Since the Capon beamformer does not reduce the noise floor just directional noise, contrast will get worse in low SNR applications when signal cancellation occurs.
 
@@
 
-No detailed evaluation of \textit{in vivo} performance.
-In cardiac application it has been proven hard to transfer results obtained in simulations to \textit{in vivo}.
-The improvements given is minimal with respect to the amount of calculation required.
+Discuss LCA.
 \\\\
 Paper3
+Discuss how contrast might get improved since the amount of signal cancellation is reduced. 
 \\\\
 Paper4
+
+Chicken-or-egg method. If the tracking fails we will fail. View-dependent visibility.
+
+Propose an method for automatic tuning of rendering parameters. Might be to extreme... A global opacity function based on global statistics combined with improved smothing, might be the future....
 \\\\
@@
 
+Merge this section with discussions..?
+
 All five papers...
+
+Detailed investigation of why results obtained in simulations are not directly transferable to \textit{in vivo}.
 

### 2014-01-19 ThesisJP/Background/background.tex
Finished summary of paper 4 and 5.
Writing discussion.

@@
 
+Architecture. See Paper II.
+
 About theoretical FLOPS: Note that these are theoretical numbers and the actual throughput typically is algorithm dependent.
@@
 
+Small section about heart anatomy.
+
+Typical probes, scan sequences, resolution and sampling.
+
 \subsection{Basic beamforming}
+
+Delay and sum. Apodization. (See intro to Paper II or III).
+Time v.s. phase delays.
 
@@
 
+Intro to adaptive beamforming. List other variants (LCA, beamspace, Eigen space etc.).
+
+Beamspace data is typically refers to the polar grid that cardiac ultrasound data is located in prior to scan conversion.
+In combination with the Capon beamformer, beamspace refers to the K-space representation of the impinging signals (hence the \nom{FFT}{Fast fourier transform} of the channel data).
+
+Not phase aberration correction.
+
 Add section about the computationally complexity. How many flops are required per rx-beam etc...
 
-\subsection{Shift invariance and sampling}
+Add section about how to present data (max v.s mean etc.)
+
+\subsection{Shift invariance}
 
@@
 
+Get section from master theses. Ray casting and opacity functions.
+
 \subsection{Adaptive volume rendering}
+
+Visibility driven visualization.
 
@@
 
+Small chapter about different simulation tools (See hos paper).
+
 \endinput

### 2014-01-19 ThesisJP/Introduction/introduction.tex
Writing discussion.

@@
 
-\section{Discussion}
+\section{Discussion and future work}
 The main contributions of this thesis are:
@@
 \item that real-time beamspace Capon beamforming is achieved for cardiac ultrasound imaging
-\item the first investigation of Capon beamforming applied on multiple frames in medical ultrasound imaging (simulated, \textit{in-vitro}, and \textit{in-vivo})
+\item the first investigation of Capon beamforming applied on multiple frames in medical ultrasound imaging (simulated, \textit{in vitro}, and \textit{in vivo})
 \item an investigation of shift-invariance when using the Capon beamformer
@@
 This clearly shows the benefit of modern GPU computing.
-
-Solving was done by a batched Gauss Jordan solver implemented by Nvidia.
+Solving was performed by a batched Gauss Jordan solver implemented by Nvidia.
 In our work we focused on the covariance estimation step, and did not try to beat Nvidia at home.
@@
 \\\\
-In \textbf{Paper II} we present the first medical ultrasound loops processed with ES- and BS-Capon beamforming.
-The paper does not provide a detailed evaluation of the method's \textit{in vivo} performance.
-Yet, initially it has been proven hard to transfer results obtained in simulations to \textit{in vivo} applications.
-The improvements seen are minimal, especially with the amount of calculation required in mind.
-The method is clearly able to decrees the lateral with of lateral point scatterers and thin structures in addition to reduce sidelobes cluttering tissue edges.
-On the other hand, the overall contrast is visually not improved and sometimes worse.
-Since the Capon beamformer does not reduce the noise floor just directional noise, contrast will get worse in low SNR applications when signal cancellation occurs.
-
-Discuss why cardiac ultrasound is selected and if full Capon will be real-time for other modalities.
-
-Discuss LCA.
+In \textbf{Paper II} we focused on achieving real time Capon processing for cardiac ultrasound, but we also present the first medical ultrasound loops processed with ES- and BS-Capon beamforming.
+The paper does not provide any detailed evaluation of the method's \textit{in vivo} performance.
+Yet, initially is has been proven hard to transfer results obtained in simulations to \textit{in vivo} images.
+The improvements seen are minimal, especially with the amount of calculations required in mind.
+The beamformers are clearly able to decrees the lateral width of lateral point scatterers and thin structures, and to sharpen step edges.
+On the other hand, the overall contrast is visually not improved and sometimes it is even worse.
+Since the Capon beamformer only affect directional noise, it is clear that contrast will get worse in low \nom{SNR}{Signal-to-noise ratio} applications when signal cancellation occurs.
+The oversampling approach presented in \textbf{Paper III} might improve on this, but it will only reduce signal cancellation if the phase shifts across the aperture are small and close to linear.
+Another issue is that bright points with a wide lateral profile exhibits better visual contrast when surrounded by speckle noise than points with narrow lateral profiles.
+In future work a detailed investigation of why results obtained in simulations are not directly transferable to \textit{in vivo} needs to be performed.
+It will also be crucial to show larger improvements \textit{in vivio} than what is presented in \textbf{Paper II}.
+Both to justify all our computations, and to show an image which has improved resolution and contrast.
+As pointed out at the end of \textbf{Paper II}, cardiac ultrasound might not be the best modality for Capon beamforming.
+Cardiac ultrasound imaging was selected because it is the medical ultrasound modality which requires most floating points operations per second.
+An application where detection of closely spaced point targets are of importance might be better suited.
+A better suited adaptive beamformer for cardiac applications is the \nom{LCA}{Low complexity beamformer} beamformer, where the Capon optimization problem is applied on a set of predefined windows.
+This method has linear complexity and shares many properties with the Capon beamformer.
 \\\\
@@
 Discuss how contrast might get improved since the amount of signal cancellation is reduced.
+
+Oversampling on receive will increase the penalty of processing beamspace data.
+Scan conversion will also take more time, and the end result should probably have higher resolution to provide the sharp definitions provided by Capon beamforming.
 \\\\
@@
 
-\section{Conclusion}
-
-Concluding remarks...
-
-\section{Future work}
-
-Merge this section with discussions..?
-
-All five papers...
-
-Detailed investigation of why results obtained in simulations are not directly transferable to \textit{in vivo}.
-
 Integrated beamforming and visualization. Using adaptive beamforming to improve e.g. speckle tracking and segmentation. Use adaptive weight statistics to guide image smoothing.

### 2014-01-20 ThesisJP/Introduction/introduction.tex
...

@@
 
-\section{Discussion and future work}
+\section{Main contributions}
 The main contributions of this thesis are:
@@
 In addition, the thesis provides several discussions on how to utilize the GPU to accelerate advanced ultrasound imaging algorithms.
-Now follows a discussion of the five papers that constitutes this thesis.
 
+\section{Discussion and future work}
 Even with all our effort in \textbf{Paper I} we were only able to achieve a fraction of the maximum throughput of the target GPU.
@@
 It will also be crucial to show larger improvements \textit{in vivio} than what is presented in \textbf{Paper II}.
-Both to justify all our computations, and to show an image which has improved resolution and contrast.
+Both to justify all our computations, and to show an image which has both improved resolution and contrast.
 As pointed out at the end of \textbf{Paper II}, cardiac ultrasound might not be the best modality for Capon beamforming.
@@
 \\\\
-Paper3
+In \textbf{Paper3} we
+
 Discuss how contrast might get improved since the amount of signal cancellation is reduced.
@@
 Scan conversion will also take more time, and the end result should probably have higher resolution to provide the sharp definitions provided by Capon beamforming.
+
+With the shift-invariant problem solved it will be interesting to see if high resolution beamforming could increase the precision of speckle tracking.
+In that case the image is generated for post processing and does not need to be visually pleasant.
 \\\\
@@
 Propose an method for automatic tuning of rendering parameters. Might be to extreme... A global opacity function based on global statistics combined with improved smothing, might be the future....
+
+Statistics derived from the selected window by e.g Capon beamforming or the LCA beamformer might be utilized to adjust the opacity.
+It can also be used as input to image filtering.
+A adaptive window will be symmetric in homogeneous regions and left-right shifted close to tissue interfaces.
+In that way we can apply less filtering and higher opacity on edges.
+If this will be different than just looking at neighboring pixels is to be seen.
 \\\\

### 2014-01-22 ThesisJP/Introduction/introduction.tex
Written discussion of paper 3.

@@
 Oversampling on transmit is not an option if the imaging modality is used to image rapidly varying objects.
+Smooth presentation of probe motion is also important in most ultrasound imaging applications.
+Real-time user interaction is one of the key selling points for ultrasound imaging.
 
@@
 Simulation tools are heavily used by researchers in order to test out new algorithms and to deduce the performance of new array designs prior to manufacturing.
-But dense simulations also involves a lot of waiting.
+But dense simulations also involves a lot of calculations and waiting.
 With the presented simulator we try to reduce the time for calculating pressure fields by means of GPU acceleration.
@@
 How the source and observation points are laid out is up to the user to decide.
-The simulator is provided with both a paint-like interface for interactive drawing of arrays, and a Matlab interface for precise scripting of array configurations.
+Points position along a line will for example simulate an ultrasound array.
+The simulator is linked with both a paint-like interface for interactive drawing of arrays, and a Matlab interface for precise scripting of array configurations.
 The main contributions in this paper is the increased performance of the GPU implementation compared with a CPU and Matlab\footnote{The Ultrasim toolbox \todo{add citation}} version.
@@
 This result shows how hard it is to reach the theoretical level of throughput, and that how close one can get is highly algorithm depended.
-It might be ways of improving these numbers, but both the solver and covariance estimation step consist of several points where fine granular synchronization and serialized instructions are needed.
+It might be ways to improve on these numbers, but both the solver and covariance estimation step consist of several points where fine granular synchronization and serialized instructions are needed.
 This will restrict even the most fine tuned implementation from reaching the maximum throughput of the GPU.
-An interesting observation can be found in Fig.\,10 in \textbf{Paper I}.
+An interesting observation can be found in Fig.\,10 of \textbf{Paper I}.
 Here we see how the achieved throughput of the covariance estimation kernel is more than three times higher than the theoretical throughput of a modern, high-end, \nom{CPU}{Central processing unit}.
@@
 \\\\
-In \textbf{Paper II} we focused on achieving real time Capon processing for cardiac ultrasound, but we also present the first medical ultrasound loops processed with ES- and BS-Capon beamforming.
+In \textbf{Paper II} we focused on achieving real time Capon processing for cardiac ultrasound by implementing the BS-Capon beamformer on a GPU.
+However, we also present the first medical ultrasound loops processed with ES- and BS-Capon beamforming.
 The paper does not provide any detailed evaluation of the method's \textit{in vivo} performance.
 Yet, initially is has been proven hard to transfer results obtained in simulations to \textit{in vivo} images.
-The improvements seen are minimal, especially with the amount of calculations required in mind.
-The beamformers are clearly able to decrees the lateral width of lateral point scatterers and thin structures, and to sharpen step edges.
-On the other hand, the overall contrast is visually not improved and sometimes it is even worse.
+The improvements seen are minimal, especially with the number of calculations in mind.
+The beamformers are clearly able to decrees the lateral width of lateral point scatterers and thin structures \textit{in vivo}, and to sharpen step edges.
+On the other hand, the overall contrast is visually not improved and sometimes it is worse (Fig.\,6b of \textbf{Paper II}).
 Since the Capon beamformer only affect directional noise, it is clear that contrast will get worse in low \nom{SNR}{Signal-to-noise ratio} applications when signal cancellation occurs.
@@
 Another issue is that bright points with a wide lateral profile exhibits better visual contrast when surrounded by speckle noise than points with narrow lateral profiles.
-In future work a detailed investigation of why results obtained in simulations are not directly transferable to \textit{in vivo} needs to be performed.
+In future work a detailed investigation of why some results obtained in simulations are not transferable to \textit{in vivo} needs to be performed.
 It will also be crucial to show larger improvements \textit{in vivio} than what is presented in \textbf{Paper II}.
-Both to justify all our computations, and to show an image which has both improved resolution and contrast.
+Both to justify all our computations, and to show an image which has both highly improved resolution and contrast.
 As pointed out at the end of \textbf{Paper II}, cardiac ultrasound might not be the best modality for Capon beamforming.
-Cardiac ultrasound imaging was selected because it is the medical ultrasound modality which requires most floating points operations per second.
-An application where detection of closely spaced point targets are of importance might be better suited.
-A better suited adaptive beamformer for cardiac applications is the \nom{LCA}{Low complexity beamformer} beamformer, where the Capon optimization problem is applied on a set of predefined windows.
+Cardiac ultrasound imaging was selected because it is the medical ultrasound modality which requires most calculations per second.
+An application where the focus is detection of closely spaced point targets, might be better suited.
+A better suited adaptive beamformer for cardiac applications is the low-complexity beamformer (\nom{LCA}{Low complexity beamformer}) beamformer, where the Capon optimization problem is applied on a set of predefined windows.
 This method has linear complexity and shares many properties with the Capon beamformer.
 \\\\
-In \textbf{Paper3} we
-
-Discuss how contrast might get improved since the amount of signal cancellation is reduced.
-
-Oversampling on receive will increase the penalty of processing beamspace data.
-Scan conversion will also take more time, and the end result should probably have higher resolution to provide the sharp definitions provided by Capon beamforming.
-
-With the shift-invariant problem solved it will be interesting to see if high resolution beamforming could increase the precision of speckle tracking.
-In that case the image is generated for post processing and does not need to be visually pleasant.
+In \textbf{Paper3} we investigate the inherent self nulling involved with Capon beamforming, and that this effect comes naturally into play when the beamformer is subjected to linear steering vector errors caused by lateral undersampling.
+These errors are naturally reduced when the lateral sampling is increased.
+This is something which has not been talked much about in the literature, either due to ignorance or its impact on imaging frame rate.
+It is not that strange since nobody have looked at Capon beamforming on multiple consecutive frames before, especially not for real time imaging.
+Fortunately, as shown in \textbf{Paper III}, it is possible to maintain the same frame rate as with DAS by applying oversampling on receive using a set of steering vectors.
+It also turns out that we don't need to perform more matrix inversions either.
+Oversampling on receive will of course increase the penalty of applying processing, like filtering,  on scan-grid data.
+Scan conversion from a polar to Cartesian display grid will also take more time, and the end result should also have high resolution in order to preserve the un-sharpening introduced by Capon beamforming.
+The oversampling method might help at improving contrast since the amount of signal cancellation is reduced.
+It will not correct for steering vector errors caused by phase aberrations.
+At best it will correct for the linear term introduced by the aberrator.
+With improved shift-invariance it will be interesting to see if high resolution beamforming could act as a post-processing step for other algorithms.
+For example to increase the precision of speckle tracking.
+In the case where the image is generated for post processing it does not need to be visually pleasant anymore as long as we avoid signal cancellation.
+Good looking speckle, or DAS-like speckle, has been the driving force behind many of the robustification techniques seen in the literature.
 \\\\

### 2014-01-25 ThesisJP/Introduction/introduction.tex
Small add-on to the motivation section.

@@
 The Capon beamformer\footnote{The name ''Capon beamformer'' is due to work by J.
-Capon \todo{add citation} on seismic arrays \todo{Move to background.}}  is a good example of an algorithm which has been highly studied for the last decade for the application of medical ultrasound imaging \todo{cite papers}, but where the number of calculations was though to be too much for real time processing.
+Capon \todo{add citation} on seismic arrays \todo{Move to background.}}  is a good example of an algorithm which has been highly studied for the last decade for the application of medical ultrasound imaging \todo{cite papers}, but where the number of calculations was though to be too high for real-time processing to happen any time soon.
 No wonder, to achieve real-time processing for e.g.
-cardiac ultrasound imaging an effective processing rate of around 100 MFLOPS is required.
-This number, as explained in Background \ref{sec:adaptbf}, also grows with larger subarrays, more image pixels, and higher framerate requirements.
-With modern GPUs these levels of processing are finally available in a single card.
+cardiac ultrasound imaging, an effective processing rate of around 100 MFLOPS is required.
+This number, as explained in Background \ref{sec:adaptbf}, also grows with larger subarrays, an increasing number of samples, and higher frame rates.
+Nevertheless, with modern GPUs, these levels of processing are finally available in a single card.
+When researchers are exploring new algorithms for ultrasound imaging it is therefore important to keep the architecture of parallel accelerators in mind.
+If the algorithm is suppose to run in real time it needs to fit the programmable and parallel pipeline of modern ultrasound scanners.
 

### 2014-01-25 ThesisJP/Background/background.tex
Written discussion of paper 3.
Small add-on to the motivation section.
Writing. Still some sections unfinished in discussion part.

@@
 
+Add sentence about siemens and supersonic imagine scanners.
+
 \section {Ultrasound imaging}
@@
 Typical probes, scan sequences, resolution and sampling.
+
+Key selling points of ultrasound imaging: cost, safety and real-time user interaction.
+
+One sentence about speckle tracking.
 
@@
 \subsection{Adaptive Beamforming}\label{sec:adaptbf}
+
+Capon beamforming\footnote{The name ''Capon beamformer'' is due to work by J. Capon \todo{add citation} on seismic arrays \todo{Move to background.}} or minimum variance beamforming.
 
@@
 
-\section{Field Simulations}
+\section{Field simulations}
 

### 2014-01-25 ThesisJP/Introduction/introduction.tex
Writing. Still some sections unfinished in discussion part.

@@
 A rapid development in computer game technology and accompanying programming languages have recently provided researchers with small personal super computers, comprised in a single graphics card (\nom{GPU}{Graphics Processing Unit}).
-The latest graphics cards\footnote{As of January 2014} from both \nom{Nvidia}{Graphics card producer} and \nom{AMD}{Graphics card producer} provides close to 6 T\nom{FLOPS}{Floating point operations per second} single precision and 1.5 TFLOPS double precision.
+The latest graphics cards\footnote{As of January 2014.} from both \nom{Nvidia}{Graphics card producer} and \nom{AMD}{Graphics card producer} provide close to 6 T\nom{FLOPS}{Floating point operations per second} single precision and 1.5 TFLOPS double precision.
 For double precision this is more than the worlds most powerful supercomputer could provide in 1998.
 
-This immense rise in computational power and programming flexibility gives new possibilities when designing ultrasound imaging systems.
+This immense rise in computational power and improved programmability are changing how ultrasound imaging systems are designed.
 Algorithms which previously had to be implemented in hardware for performance reasons can now be implemented in software.
@@
 
-The Capon beamformer\footnote{The name ''Capon beamformer'' is due to work by J.
-Capon \todo{add citation} on seismic arrays \todo{Move to background.}}  is a good example of an algorithm which has been highly studied for the last decade for the application of medical ultrasound imaging \todo{cite papers}, but where the number of calculations was though to be too high for real-time processing to happen any time soon.
+The Capon beamforme  is a good example of an algorithm which has been highly studied for the last decade for the application of medical ultrasound imaging, but where the number of calculations was though to be too high for real-time processing to happen any time soon.
 No wonder, to achieve real-time processing for e.g.
@@
 This number, as explained in Background \ref{sec:adaptbf}, also grows with larger subarrays, an increasing number of samples, and higher frame rates.
-Nevertheless, with modern GPUs, these levels of processing are finally available in a single card.
+Nevertheless, with modern GPUs, these levels of processing are finally available within a single card.
 When researchers are exploring new algorithms for ultrasound imaging it is therefore important to keep the architecture of parallel accelerators in mind.
@@
 
-Well on my way into the project, when a real-time Capon beamformer was realized, and loops of images for the first time were processed, new issues where discovered and had to be solved.
+Well on my way into the project, when a real-time Capon beamformer was realized, and loops of images for the first time were processed at interactive and real-time frame rates, new issues where discovered and had to be solved.
 This led me to a more theoretical study of the Capon beamformer (\textbf{Paper III}), with special attention on how to obtain high lateral resolution while preserving the important shift-invariant property of medical ultrasound imaging.
@@
 In \textbf{Paper VI}, which is summarized in \textbf{Paper II}, it is shown that our implementation from  \textbf{Paper I} does not reach the level of throughput required for real-time Capon beamforming of cardiac ultrasound imaging.
-The matrices that have to be inverted becomes to large, and the number of frames that needs to be processed per second is to many.
+The matrices that have to be inverted become to large, and the frames that needs to be processed per second are to many.
 
@@
 Smooth presentation of probe motion is also important in most ultrasound imaging applications.
-Real-time user interaction is one of the key selling points for ultrasound imaging.
+Real-time user interaction is one of the key selling points for ultrasound imaging and frame rate should therefore not be lower than 10 frames per second.
 
@@
 The paper presents both quantitative and qualitative results supporting the method's improved capabilities of reducing tissue dropouts and spurious structures inside the blood pool, and by that maximizing the amount of visible endocardium tissue.
-The proposed method is implemented on a GPU which results in interactive frame rates.
+The proposed method is implemented on a GPU with interactive frame rates as the result.
 
@@
 How the source and observation points are laid out is up to the user to decide.
-Points position along a line will for example simulate an ultrasound array.
+Point sources positioned along a line will for example simulate an ultrasound array.
 The simulator is linked with both a paint-like interface for interactive drawing of arrays, and a Matlab interface for precise scripting of array configurations.
@@
 \begin{enumerate}
-\item a GPU implementation of the Capon beamformer
-\item a GPU implementation of the beamspace Capon beamformer
-\item that real-time beamspace Capon beamforming is achieved for cardiac ultrasound imaging
-\item the first investigation of Capon beamforming applied on multiple frames in medical ultrasound imaging (simulated, \textit{in vitro}, and \textit{in vivo})
-\item an investigation of shift-invariance when using the Capon beamformer
-\item a method for improved shift-invariance of the Capon beamformer
-\item a method for reduced blood-pool noise in volume renderings of cardiac ultrasound volumes
-\item a GPU implementation of this adaptive volume rendering method
-\item a GPU implementation of simple paint-like simulation tool for rapid visualization ultrasound pressure fields
+\item a GPU implementation of the Capon beamformer.
+\item a GPU implementation of the beamspace Capon beamformer.
+\item that real-time beamspace Capon beamforming is achieved for cardiac ultrasound imaging.
+\item the first investigation of Capon beamforming applied on multiple frames in medical ultrasound imaging (simulated, \textit{in vitro}, and \textit{in vivo}).
+\item an investigation of shift-invariance when the Capon beamformer is applied on consecutive frames.
+\item a method for improved shift-invariance of the Capon beamformer.
+\item a method for reduced blood-pool noise in volume renderings of cardiac ultrasound volumes.
+\item a GPU implementation of this adaptive volume rendering method.
+\item a GPU implementation of simple paint-like simulation tool for rapid visualization ultrasound pressure fields.
 \end{enumerate}
-
 In addition, the thesis provides several discussions on how to utilize the GPU to accelerate advanced ultrasound imaging algorithms.
@@
 The paper does not provide any detailed evaluation of the method's \textit{in vivo} performance.
-Yet, initially is has been proven hard to transfer results obtained in simulations to \textit{in vivo} images.
+Yet, initially is has been proven hard to transfer all results obtained in simulations to \textit{in vivo} images.
 The improvements seen are minimal, especially with the number of calculations in mind.
@@
 On the other hand, the overall contrast is visually not improved and sometimes it is worse (Fig.\,6b of \textbf{Paper II}).
-Since the Capon beamformer only affect directional noise, it is clear that contrast will get worse in low \nom{SNR}{Signal-to-noise ratio} applications when signal cancellation occurs.
-The oversampling approach presented in \textbf{Paper III} might improve on this, but it will only reduce signal cancellation if the phase shifts across the aperture are small and close to linear.
+Since the Capon beamformer only affect directional noise, it is clear that contrast will get worse in low \nom{SNR}{Signal-to-noise ratio} situations when signal cancellation occurs.
+The oversampling approach presented in \textbf{Paper III} might improve on this, but it will only reduce signal cancellation if the phase shifts across the aperture are close to linear and small.
 Another issue is that bright points with a wide lateral profile exhibits better visual contrast when surrounded by speckle noise than points with narrow lateral profiles.
-In future work a detailed investigation of why some results obtained in simulations are not transferable to \textit{in vivo} needs to be performed.
+In future work a detailed investigation of why some results, especially the contrast, obtained in simulations are not transferable to \textit{in vivo} needs to be performed.
 It will also be crucial to show larger improvements \textit{in vivio} than what is presented in \textbf{Paper II}.
 Both to justify all our computations, and to show an image which has both highly improved resolution and contrast.
-As pointed out at the end of \textbf{Paper II}, cardiac ultrasound might not be the best modality for Capon beamforming.
+As pointed out at the end of \textbf{Paper II}, cardiac ultrasound imaging might not be the best modality for Capon beamforming.
 Cardiac ultrasound imaging was selected because it is the medical ultrasound modality which requires most calculations per second.
 An application where the focus is detection of closely spaced point targets, might be better suited.
-A better suited adaptive beamformer for cardiac applications is the low-complexity beamformer (\nom{LCA}{Low complexity beamformer}) beamformer, where the Capon optimization problem is applied on a set of predefined windows.
+A better suited adaptive beamformer for cardiac applications might be the low-complexity adaptive beamformer (\nom{LCA beamformer}{Low complexity adaptive beamformer}), where the Capon optimization problem is applied on a set of predefined windows.
 This method has linear complexity and shares many properties with the Capon beamformer.
+If only minor improvements are obtained, it would be easier to ignore this fact if the algorithm is less costly.
+Finally, it would also be interesting to see how the Eigenspace-based Capon beamformer performs on cardiac images, and how its absence of a distortionless criterion will influence the image spatially and temporally.
 \\\\
-In \textbf{Paper3} we investigate the inherent self nulling involved with Capon beamforming, and that this effect comes naturally into play when the beamformer is subjected to linear steering vector errors caused by lateral undersampling.
+In \textbf{Paper3} we investigate the inherent self nulling involved with Capon beamforming, and how this effect comes naturally into play when the beamformer is subjected to linear steering vector errors caused by lateral undersampling.
 These errors are naturally reduced when the lateral sampling is increased.
-This is something which has not been talked much about in the literature, either due to ignorance or its impact on imaging frame rate.
-It is not that strange since nobody have looked at Capon beamforming on multiple consecutive frames before, especially not for real time imaging.
-Fortunately, as shown in \textbf{Paper III}, it is possible to maintain the same frame rate as with DAS by applying oversampling on receive using a set of steering vectors.
-It also turns out that we don't need to perform more matrix inversions either.
-Oversampling on receive will of course increase the penalty of applying processing, like filtering,  on scan-grid data.
-Scan conversion from a polar to Cartesian display grid will also take more time, and the end result should also have high resolution in order to preserve the un-sharpening introduced by Capon beamforming.
-The oversampling method might help at improving contrast since the amount of signal cancellation is reduced.
-It will not correct for steering vector errors caused by phase aberrations.
-At best it will correct for the linear term introduced by the aberrator.
-With improved shift-invariance it will be interesting to see if high resolution beamforming could act as a post-processing step for other algorithms.
-For example to increase the precision of speckle tracking.
-In the case where the image is generated for post processing it does not need to be visually pleasant anymore as long as we avoid signal cancellation.
-Good looking speckle, or DAS-like speckle, has been the driving force behind many of the robustification techniques seen in the literature.
+This simple observation is something which has been missing in the literature on Capon beamforming for medical ultrasound imaging.
+The reason could be ignorance or its possible negative impact on imaging frame rate.
+A lack of studies where Capon beamforming is performed on consecutive frames might also explain its absence.
+Fortunately, as shown in \textbf{Paper III}, it is possible to maintain the same frame rate as with DAS beamforming by applying oversampling on receive using a set of steering vectors (phase delays).
+It turns out that we do not need to perform more matrix inversions either.
+Oversampling on receive will, however,  increase the penalty of applying post-processing, like filtering,  on scan-grid data.
+%The time taken to convert data from a polar scan grid to a Cartesian display grid will also increase.
+The Cartesian grid should also have high resolution in order to preserve the un-sharpening introduced by Capon beamforming.
+As just mentioned in the discussion of \textbf{Paper II}, the reduced signal cancellation should prevent contrast from degrade to less than the contrast obtain with DAS.
+It will, however,  not correct for signal cancellation caused by phase aberrations.
+At best it will correct for the linear term introduced by the aberrator if the term is small.
+With improved shift-invariance it will also be interesting to see if high resolution beamforming could act as a post-processing step for other algorithms.
+For example to increase the precision of speckle and edge tracking.
+In the case where the image is generated for post processing it does not need to be visually pleasant anymore, as long as signal cancellation is avoided.
+Good looking speckle, or DAS-like speckle, has been the driving force behind many of the robustification techniques and the default parameters seen in the literature.
+Hence, more aggressive parameters might be used if the image produced is intended as input to another algorithm.
 \\\\
-Paper4
+Objections to \textbf{Paper IV} includes its inherent chicken-or-egg dilemma.
+The visualization is adjusted based on the data acquired by a user.
+The user sees the visualization and adjust scanning based on the visual input.
+This recursive behavior suggest that the method is best suited for post-processing when data has been recorded and saved.
 
-Chicken-or-egg method. If the tracking fails we will fail. View-dependent visibility.
+It is also clear that if the tracking fails the suggested method will fail.
+
+Does data need to be saved in order to perform tracking?
+
+View-dependent visibility.
 
@@
 If this will be different than just looking at neighboring pixels is to be seen.
-\\\\
-Paper5
-
-This is not an adaptive algorithm.
-
-Mark that code for GPU-Capon and Huygens on speed have been made open source.
-It has over 200 downloads and soon 1000 views on youtube.
 
 Integrated beamforming and visualization. Using adaptive beamforming to improve e.g. speckle tracking and segmentation. Use adaptive weight statistics to guide image smoothing.
+\\\\
+The fifth paper stands out from the other papers by not dealing with an adaptive processing technique.
+It is still, however, centered around the topic of GPU implementation of algorithms for ultrasound imaging.
+Ultrasound simulations are typically used for research and to pre-calculate configurations of a given scan sequence.
+The continuously growth in computational power will, in the future, make it possible to evaluate a scan sequence on the fly, removing the need for
 
+Achieving higher processing speed for an excising algorithm is not impressive as long as there is no novel analysis of the method modification    and the increase does not enables new ways of utilizing the increased speed.
+
+The simulator has been download 200 times and have close to 1000 views on youtube.
+
+Fast calculation of pressure fields can be used to automatically evaluate the pressure field of the current scan sequence. Hence, we can optimize weights etc. on the fly.
+\\\\
+The code for both the Capon beamformer and the ultrasound field simulator have been made open source and is available on github\footnote{github.com/jpaasen/hos and github.com/jpaasen/cos}. 
 \endinput

### 2014-01-25 ThesisJP/Introduction/introduction.tex
Finished the discussion. Next background and revisions...

@@
 \\\\
-Objections to \textbf{Paper IV} includes its inherent chicken-or-egg dilemma.
+%Objections to \textbf{Paper IV} includes its inherent chicken-or-egg dilemma.
 The visualization is adjusted based on the data acquired by a user.
@@
 This recursive behavior suggest that the method is best suited for post-processing when data has been recorded and saved.
-
-It is also clear that if the tracking fails the suggested method will fail.
-
-Does data need to be saved in order to perform tracking?
-
-View-dependent visibility.
-
-Propose an method for automatic tuning of rendering parameters. Might be to extreme... A global opacity function based on global statistics combined with improved smothing, might be the future....
-
-Statistics derived from the selected window by e.g Capon beamforming or the LCA beamformer might be utilized to adjust the opacity.
-It can also be used as input to image filtering.
-A adaptive window will be symmetric in homogeneous regions and left-right shifted close to tissue interfaces.
+%Does data need to be saved in order to perform tracking?
+One limitation of the algorithm described in \textbf{Paper IV} is its dependence on a successful tracking.
+If the tracking fails the proposed method will fail.
+Another issue is the view-dependent visibility, where an object could become visible or disappear while rotating or if the local statistics change a lot during the cardiac cycle.
+This means that the rendering will be come non-intuitive to interact with and unrealistic to watch.
+However, the method could be used to clean up still images of standard views.
+The paper does also propose an interesting method for automatic tuning of rendering parameters.
+Figure 3d in \textbf{Paper IV} actually shows that an optimal threshold could be found based on global statistics as well.
+Even though the final visualization has more errors, with respect to endocardium visibility, the tuning of the opacity function is fully automatic.
+An automatically tuned opacity function based on global statistics combined with improved smoothing could be the future for cardiac ultrasound imaging.
+It would also be interesting to see if statistics derived based on the window selected by the Capon or the LCA beamformer could be utilized to adjust the opacity.
+It should also be possible to use this statistic as input to image filtering.
+An adaptively selected window will be symmetric in homogeneous regions and left-right shifted close to tissue interfaces.
 In that way we can apply less filtering and higher opacity on edges.
 If this will be different than just looking at neighboring pixels is to be seen.
-
-Integrated beamforming and visualization. Using adaptive beamforming to improve e.g. speckle tracking and segmentation. Use adaptive weight statistics to guide image smoothing.
 \\\\
 The fifth paper stands out from the other papers by not dealing with an adaptive processing technique.
-It is still, however, centered around the topic of GPU implementation of algorithms for ultrasound imaging.
-Ultrasound simulations are typically used for research and to pre-calculate configurations of a given scan sequence.
-The continuously growth in computational power will, in the future, make it possible to evaluate a scan sequence on the fly, removing the need for
-
-Achieving higher processing speed for an excising algorithm is not impressive as long as there is no novel analysis of the method modification    and the increase does not enables new ways of utilizing the increased speed.
-
-The simulator has been download 200 times and have close to 1000 views on youtube.
-
-Fast calculation of pressure fields can be used to automatically evaluate the pressure field of the current scan sequence. Hence, we can optimize weights etc. on the fly.
+It is still, however, centered around the topic of implementing algorithms for ultrasound imaging on GPUs.
+Ultrasound simulations are typically used for research and to sometimes pre-calculate configurations of a given scan sequence to achieve a certain pressure in a given area.
+The continuously growth in computational power will, in the future, make it possible to evaluate a scan sequence on the fly, removing the need for lookup tables.
+Estimation of the mechanical and thermal index could also be improved.
+Achieving higher processing speed for an excising algorithm by implementing it on a new architecture is not academically relevant in its own.
+It has to combined with either a good analysis of what special adjustments that had to be made to the algorithm in order for the increase to happen or a suggestion of how to utilize the speed gained.
+These are points we have tried to answer in both \textbf{Paper I}, \textbf{II},  \textbf{III}, and \textbf{V}.
+As a curiosity I would like to mention the as of January 2014 the simulator has been download 200 times and counting.
+It also has a "whopping" 1000 views on youtube.
 \\\\
-The code for both the Capon beamformer and the ultrasound field simulator have been made open source and is available on github\footnote{github.com/jpaasen/hos and github.com/jpaasen/cos}. 
+Finally it is worth mentioning that the code for both the Capon beamformer\footnote{github.com/jpaasen/cos} and the ultrasound field simulator\footnote{github.com/jpaasen/hos} have been made open source and is available on github.
 \endinput

### 2014-02-01 ThesisJP/Introduction/introduction.tex
Small changes in the introduction section.

@@
 A rapid development in computer game technology and accompanying programming languages have recently provided researchers with small personal super computers, comprised in a single graphics card (\nom{GPU}{Graphics Processing Unit}).
-The latest graphics cards\footnote{As of January 2014.} from both \nom{Nvidia}{Graphics card producer} and \nom{AMD}{Graphics card producer} provide close to 6 T\nom{FLOPS}{Floating point operations per second} single precision and 1.5 TFLOPS double precision.
+The latest graphics cards\footnote{As of January 2014.} from both \nom{Nvidia}{Graphics card producer} and \nom{AMD}{Graphics card producer} provide close to 6 T\nom{FLOPS}{Floating point operations per second} single precision and 1.5 TFLOPS double precision computations.
 For double precision this is more than the worlds most powerful supercomputer could provide in 1998.
 
-This immense rise in computational power and improved programmability are changing how ultrasound imaging systems are designed.
+This immense rise in computational power and improved programmability are currently changing how ultrasound imaging systems are designed.
 Algorithms which previously had to be implemented in hardware for performance reasons can now be implemented in software.
-And algorithms, either too complicated to implement in hardware, or though to be too computationally heavy for real-time use, becomes realizable.
+And algorithms, either too complicated or costly to implement in hardware, and where a software implementation was though to be too computationally heavy for real-time use, becomes realizable.
 It is clear that high-performance programmable processors, like the GPU, already have and will continue to make future ultrasound imaging systems more flexible, cheaper to produce, and equipped with even more cutting-edge processing.
 
-The Capon beamforme  is a good example of an algorithm which has been highly studied for the last decade for the application of medical ultrasound imaging, but where the number of calculations was though to be too high for real-time processing to happen any time soon.
+The Capon beamformer  is a good example of an algorithm which has been highly studied for the last decade for the application of medical ultrasound imaging, but where the number of calculations was though to be too high for real-time processing to happen any time soon.
 No wonder, to achieve real-time processing for e.g.
@@
 When researchers are exploring new algorithms for ultrasound imaging it is therefore important to keep the architecture of parallel accelerators in mind.
-If the algorithm is suppose to run in real time it needs to fit the programmable and parallel pipeline of modern ultrasound scanners.
+If a new complex algorithm is suppose to run in real time it needs to fit the programmable and parallel pipeline of modern ultrasound scanners.
 
@@
 The first is adaptive visualization of cardiac ultrasound volumes (\textbf{Paper\,IV}) and the latter is simulation of dense ultrasound pressure fields (\textbf{Paper\,V}).
-Both share the property of being computationally heavy, but they also consist of many independent computations which is perfect for parallel GPU computing.
+Both share the property of being computationally heavy, but they also consist of many independent computations which makes them perfectly suited for parallel GPU acceleration.
 
 Well on my way into the project, when a real-time Capon beamformer was realized, and loops of images for the first time were processed at interactive and real-time frame rates, new issues where discovered and had to be solved.
-This led me to a more theoretical study of the Capon beamformer (\textbf{Paper III}), with special attention on how to obtain high lateral resolution while preserving the important shift-invariant property of medical ultrasound imaging.
+This led me to a more theoretical study of the Capon beamformer (\textbf{Paper III}), with special attention on how to obtain high lateral resolution while preserving the important shift-invariant property of ultrasound imaging.
 Shift-invariant behavior is crucial if the method is ever to be applied for live scanning.
@@
 Oversampling on transmit is not an option if the imaging modality is used to image rapidly varying objects.
-Smooth presentation of probe motion is also important in most ultrasound imaging applications.
-Real-time user interaction is one of the key selling points for ultrasound imaging and frame rate should therefore not be lower than 10 frames per second.
+Smooth presentation of probe motion is also important in most ultrasound imaging applications since real-time user interaction is one of the key selling points for ultrasound imaging.
+The frame rate should therefore  be higher than 10 frames per second.
 
@@
 The paper does not provide any detailed evaluation of the method's \textit{in vivo} performance.
-Yet, initially is has been proven hard to transfer all results obtained in simulations to \textit{in vivo} images.
+Yet, initially it has been proven hard to transfer all results obtained in simulations to \textit{in vivo} images.
 The improvements seen are minimal, especially with the number of calculations in mind.
@@
 Fortunately, as shown in \textbf{Paper III}, it is possible to maintain the same frame rate as with DAS beamforming by applying oversampling on receive using a set of steering vectors (phase delays).
-It turns out that we do not need to perform more matrix inversions either.
+It also turns out that more matrix inversions are not needed either.
 Oversampling on receive will, however,  increase the penalty of applying post-processing, like filtering,  on scan-grid data.
@@
 The Cartesian grid should also have high resolution in order to preserve the un-sharpening introduced by Capon beamforming.
-As just mentioned in the discussion of \textbf{Paper II}, the reduced signal cancellation should prevent contrast from degrade to less than the contrast obtain with DAS.
+As just mentioned in the discussion of \textbf{Paper II}, reduced signal cancellation should prevent contrast from degrade to less than the contrast obtain with DAS.
 It will, however,  not correct for signal cancellation caused by phase aberrations.
@@
 This means that the rendering will be come non-intuitive to interact with and unrealistic to watch.
-However, the method could be used to clean up still images of standard views.
+However, the method could still be used to clean up still images of standard cardiac views.
 The paper does also propose an interesting method for automatic tuning of rendering parameters.
-Figure 3d in \textbf{Paper IV} actually shows that an optimal threshold could be found based on global statistics as well.
-Even though the final visualization has more errors, with respect to endocardium visibility, the tuning of the opacity function is fully automatic.
-An automatically tuned opacity function based on global statistics combined with improved smoothing could be the future for cardiac ultrasound imaging.
-It would also be interesting to see if statistics derived based on the window selected by the Capon or the LCA beamformer could be utilized to adjust the opacity.
+Figure 3d in \textbf{Paper IV} actually shows that an optimal threshold could be found based on global and not only local statistics.
+Even though the visualization based on global statistics has more errors, with respect to endocardium visibility, the tuning of the opacity function is still fully automatic if an iterative search for the minima in Figure 3d is performed.
+The rendering will then be realistic, but still dependent on a successful tracking for the auto-adjustment to work.
+An automatically tuned opacity function based on global statistics combined with an improved smoothing scheme could be the future for cardiac ultrasound imaging.
+It would also be interesting to see if statistics based on the window selected by the Capon or the LCA beamformer could be utilized to adjust the opacity.
 It should also be possible to use this statistic as input to image filtering.
@@
 It is still, however, centered around the topic of implementing algorithms for ultrasound imaging on GPUs.
-Ultrasound simulations are typically used for research and to sometimes pre-calculate configurations of a given scan sequence to achieve a certain pressure in a given area.
+Ultrasound simulations are typically used for research and sometimes also to pre-calculate configurations of a given scan sequence in order to achieve a certain pressure in a given area.
 The continuously growth in computational power will, in the future, make it possible to evaluate a scan sequence on the fly, removing the need for lookup tables.
@@
 These are points we have tried to answer in both \textbf{Paper I}, \textbf{II},  \textbf{III}, and \textbf{V}.
-As a curiosity I would like to mention the as of January 2014 the simulator has been download 200 times and counting.
+In \textbf{Paper V} the increased speed makes an interactive paint program feasible.
+As of January 2014 the simulator has been download 200 times and counting.
 It also has a "whopping" 1000 views on youtube.

### 2014-02-02 ThesisJP/Introduction/introduction.tex
Feedback from Laila.

@@
 Algorithms which previously had to be implemented in hardware for performance reasons can now be implemented in software.
-And algorithms, either too complicated or costly to implement in hardware, and where a software implementation was though to be too computationally heavy for real-time use, becomes realizable.
+And algorithms, either too complicated or costly to implement in hardware, and where a software implementation was thought to be too computationally heavy for real-time use, becomes realizable.
 It is clear that high-performance programmable processors, like the GPU, already have and will continue to make future ultrasound imaging systems more flexible, cheaper to produce, and equipped with even more cutting-edge processing.
 
-The Capon beamformer  is a good example of an algorithm which has been highly studied for the last decade for the application of medical ultrasound imaging, but where the number of calculations was though to be too high for real-time processing to happen any time soon.
+The Capon beamformer  is a good example of an algorithm which has been highly studied for the last decade for the application of medical ultrasound imaging, but where the number of calculations was thought to be too high for real-time processing to happen any time soon.
 No wonder, to achieve real-time processing for e.g.
@@
 
-My main focus have been the Capon beamformer, and the problem of making this computationally intense algorithm available for real-time ultrasound imaging (\textbf{Paper\,I} and \textbf{II}).
+Main focus have been the Capon beamformer, and the problem of making this computationally intense algorithm available for real-time ultrasound imaging (\textbf{Paper\,I} and \textbf{II}).
 However, two additional methods were also explored.
 The first is adaptive visualization of cardiac ultrasound volumes (\textbf{Paper\,IV}) and the latter is simulation of dense ultrasound pressure fields (\textbf{Paper\,V}).
-Both share the property of being computationally heavy, but they also consist of many independent computations which makes them perfectly suited for parallel GPU acceleration.
+Both share the property of being computationally complex, but consist of many independent computations which makes them perfectly suited for parallel GPU acceleration.
 
 Well on my way into the project, when a real-time Capon beamformer was realized, and loops of images for the first time were processed at interactive and real-time frame rates, new issues where discovered and had to be solved.
-This led me to a more theoretical study of the Capon beamformer (\textbf{Paper III}), with special attention on how to obtain high lateral resolution while preserving the important shift-invariant property of ultrasound imaging.
+This led to a more theoretical study of the Capon beamformer (\textbf{Paper III}), with special attention on how to obtain high lateral resolution while preserving the important shift-invariant property of ultrasound imaging.
 Shift-invariant behavior is crucial if the method is ever to be applied for live scanning.
@@
 {\it IEEE Transactions on Oceanic Engineering, submitted.}\\\\
-The first paper describes in details how we mapped the Capon beamformer to GPU architecture.
+The first paper describes in details how the Capon beamformer was mapped to GPU architecture.
 Even though the paper is written within the field of active \nom{sonar}{SOund Navigation And Ranging (usually under water)} imaging, the depicted implementation is applicable to a vast range of active imaging systems.
@@
 Active sonar imaging typically differs from medical ultrasound imaging by a lower real-time requirement and fewer array elements.
-This makes it somewhat easier to reach our goal of real-time processing.
+This makes it somewhat easier to reach the goal of real-time processing.
 
 The estimation of the spatial covariance matrix receives special attention in this study.
-In previous literature on Capon beamforming the matrix inversion has always been regarded as the most computationally complex piece.
-In this paper we show that estimation of the sample covariance matrix is actually the most complex part when spatial and temporal smoothing are included with common parameters.
+In previous literature on Capon beamforming the matrix inversion has always been regarded as the most computationally complex step.
+In this paper we show that estimation of the sample covariance matrix is actually the most complex part when spatial and temporal smoothing with common parameters are included.
 It is then shown how the arithmetic complexity of this estimation process can be reduced from cubic to square.
-Finally, we give an in-depth analysis of the arithmetic throughput on multiple platforms.
-Despite our effort, the number of effective FLOPS is only 4 \% and 14 \% of the theoretical throughput of the target GPU for the solver and covariance estimator respectively.
+Finally, an in-depth analysis of the arithmetic throughput on multiple platforms is given.
+Despite our effort, the number of effective FLOPS is only 4 \% and 14 \% of the theoretical throughput of the target GPU for the matrix equation solver and covariance estimator respectively.
 Still, the reported throughput of 1 Mpx/s on a high-end GPU is enough to provide real-time processing for common sonar scan sequences.
@@
 {\it IEEE Transactions on Ultrasonics, Ferroelectrics, and Frequency Control, vol. 61, no. 1, pp. \todo{add pages}, Jan. 2014.}\\\\
-The second paper aims at achieving real-time Capon beamforming for cardiac ultrasound imaging.
-Such an implementation will facilitates further study of the method's \textit{in vivo} performance for this modality.
+The second paper aims at implementing real-time Capon beamforming for cardiac ultrasound imaging.
+Achieving this will facilitate further study of this method's \textit{in vivo} performance.
 This is something which has been sought for the past decade \todo{Add citations}.
-In \textbf{Paper I} we investigated arrays with no more than 32 elements.
+In \textbf{Paper I} arrays with no more than 32 elements were investigated.
 A linear array for cardiac ultrasound imaging typically has 64 or more elements.
 In \textbf{Paper VI}, which is summarized in \textbf{Paper II}, it is shown that our implementation from  \textbf{Paper I} does not reach the level of throughput required for real-time Capon beamforming of cardiac ultrasound imaging.
-The matrices that have to be inverted become to large, and the frames that needs to be processed per second are to many.
+The matrices that have to be inverted become to large, and the frames that need to be processed per second are too many.
 
-To reduce the matrix size we take, in this paper, advantage of the beamspace version of the Capon beamformer (\nom{BS-Capon}{Beamspace Capon beamforming}), and implement it on the GPU.
+In this paper, to reduce the matrix size, it is taken advantage of the beamspace version of the Capon beamformer (\nom{BS-Capon}{Beamspace Capon beamforming}), which is implemented on the GPU.
 For parameters previously derived to give similar performance for BS-Capon and element-space Capon (\nom{ES-Capon}{Element Space Capon}), we show that real-time BS-Capon beamforming is feasible for cardiac ultrasound imaging.
@@
 The paper was written based on observations of aliasing artifacts when imaging bright point scatterers \textit{in vitro} and in simulations.
-The points were observed to twinkle like stars in the sky.
-In earlier work on Capon beamforming for medical ultrasound imaging the real-time nature of ultrasound imaging has often been ignored, and a large degree of oversampling on transmit might have been applied without comments.
+The point scatterers were observed to twinkle like stars in the sky.
+In earlier work on Capon beamforming for medical ultrasound imaging the real-time nature of ultrasound imaging has often been ignored, and a large degree of oversampling on transmit might have been applied without being clearly stated.
 Oversampling on transmit is not an option if the imaging modality is used to image rapidly varying objects.
-Smooth presentation of probe motion is also important in most ultrasound imaging applications since real-time user interaction is one of the key selling points for ultrasound imaging.
+Smooth presentation of probe motion is equally important since real-time user interaction is one of the key selling points of ultrasound imaging.
 The frame rate should therefore be higher than 10 frames per second.
 
-In this paper we investigate how the Capon beamforming achieves its super resolution, and how this makes it highly sensitive to differences between the assumed steering vector and the signal propagation vector.
-Earlier work, focusing on single images, has typically referred to this effect as speed-of-sound errors, but it also comes naturally into play when objects moves from frame to frame and the beam density is insufficient.
-With a narrowband and farfield model we show that the Capon beamformer suffers from beam-to-beam gain variations as large as 27 dB for parameters previously used to obtain high resolution medical ultrasound images.
+Further it has been investigated how the Capon beamforming achieves its super resolution, and how this makes it highly sensitive to differences between the assumed steering vector and the signal propagation vector.
+Earlier work, focusing on single images, has typically referred to this effect as speed-of-sound errors, however, it also comes naturally into play when objects moves from frame to frame and the beam density is insufficient.
+With a narrowband and farfield model it is shown that the Capon beamformer suffers from beam-to-beam gain variations as large as 27 dB for parameters previously used to obtain high resolution medical ultrasound images.
 To obtain the same lateral gain variations as with \nom{DAS}{Delay-and-sum beamforming} beamforming, 25 times oversampling is shown to be required.
@@
 Finally we show that the gain variation can be reduced, hence improved shift-invariance, by oversampling on receive using phase rotation steering.
-This without affecting the acquisition frame rate and with a minor increase in computationally complexity.
+This is achieved without affecting the acquisition frame rate and with a minor increase in computational complexity.
 The method is successfully applied on simulations and \textit{in vitro} phantom data.
@@
 This because ultrasound signals have high variability within blood as well as high variability within tissue, even when heavily smoothed.
-Delineated of the blood pool is done by a state estimation algorithm, capable of tracking the endocardium in real time.
-The final visualizations are compared, with respect to location of tissue, with the result from a state-of-the-art endocardium segmentation tool.
-The paper presents both quantitative and qualitative results supporting the method's improved capabilities of reducing tissue dropouts and spurious structures inside the blood pool, and by that maximizing the amount of visible endocardium tissue.
+Delineation of the blood pool is done by a state estimation algorithm, capable of tracking the endocardium in real time.
+The final visualizations are compared, with respect to location of tissue, by using a state-of-the-art endocardium segmentation tool.
+The paper presents both quantitative and qualitative results supporting the method's improved capabilities of reducing tissue dropouts and spurious structures inside the blood pool, leading to a maximized amount of visible endocardium tissue.
 The proposed method is implemented on a GPU with interactive frame rates as the result.
+\todo{litt tafatt.
+m\aa sm\oe{}re p\aa litt}
 
@@
 Simulation tools are heavily used by researchers in order to test out new algorithms and to deduce the performance of new array designs prior to manufacturing.
-But dense simulations also involves a lot of calculations and waiting.
+However, dense simulations involve extensive calculation and computational time.
 With the presented simulator we try to reduce the time for calculating pressure fields by means of GPU acceleration.
-The tool is based on Huygens' principle which describes diffraction caused by a slit as the super position of several point sources located inside the opening.
+The tool is based on Huygens' principle which describes diffraction caused by a slit as the superposition of several point sources located inside the opening.
 Thus, the simulator works by accumulating the contribution from a collection of point sources in a set of observation points.
-How the source and observation points are laid out is up to the user to decide.
-Point sources positioned along a line will for example simulate an ultrasound array.
-The simulator is linked with both a paint-like interface for interactive drawing of arrays, and a Matlab interface for precise scripting of array configurations.
+How the source and observation points are laid out is for the user to decide.
+Point sources positioned along a line will for instance simulate an ultrasound array.
+The simulator is linked with both a Paint-like interface for interactive drawing of arrays, and a Matlab interface for precise scripting of array configurations.
 The main contributions in this paper is the increased performance of the GPU implementation compared with a CPU and Matlab\footnote{The Ultrasim toolbox \todo{add citation}} version.

### 2014-02-05 ThesisJP/Background/background.tex
Started adding figures in background section.

@@
 
-General introduction to ultrasound, GPUs etc...
+In this chapter the background information needed in order to understand the topics of this thesis and its papers is given.
+First, in Section \ref{sec:gpgpu} an introduction to general purpose programming on graphics processing units (\nom{GPGPU}{General purpose programming on graphics processing units}) is presented.
+Next, an introduction to ultrasound imaging is given in Section \ref{sec:ultrasound}.
 
-\section{General-purpose computing on graphics processing units}
+\section{General-purpose computing on graphics processing units}\label{sec:gpgpu}
 
@@
 
-Add sentence about siemens and supersonic imagine scanners.
+Add sentence about siemens and supersonic imagine scanners. They use GPUs? What does this mean? FPGAs and ASICs.
 
-\section {Ultrasound imaging}
+\begin{figure}
+\centering
+\includegraphics[width=0.9\textwidth]{img/free_lunsh.png}
+\caption{Illustration of several transitions currently taking place in the computer industry. \todo{Site Herb Sutter - Welcom to the jungle http://herbsutter.com/welcome-to-the-jungle/}}
+\label{fig:jungle}
+\end{figure}
+
+\begin{figure}
+\centering
+\subfigure[Threads, blocks and grid]{
+	\includegraphics[width=0.4\textwidth]{img/cuda_threads.png}
+}
+\subfigure[Kernel]{
+	\includegraphics[width=0.5\textwidth]{img/kernel.pdf}
+}
+\caption{Depicts how GPU threads are grouped into blocks and arranged in a grid. One thread runs a copy of a kernel function, in this example an element-wise matrix square operation.}
+\label{fig:gpu_grid}
+\end{figure}
+
+\section {Ultrasound imaging}\label{sec:ultrasound}
+Ultrasound imaging encompasses technology which generates images based on sound whose frequencies we can not hear. For medical ultrasound imaging frequencies above 2 MHz are typically used.
 
 Small section about heart anatomy.
+
+\begin{figure}
+\centering
+\subfigure{
+	\includegraphics[width=0.47\textwidth]{img/Diagram_of_the_human_heart.png}
+}
+\subfigure{
+	\includegraphics[width=0.47\textwidth]{img/HeartWall.png}
+}
+\caption{Overview of the human heart. Illustrations from wikipedia.org.}
+\label{fig:human_heart}
+\end{figure}
 
@@
 \subsection{Adaptive Beamforming}\label{sec:adaptbf}
+
+\begin{figure}[t!]
+\subfigure[Delay-and-sum]{
+	\includegraphics[width=0.47\textwidth]{img/scenario_das_resp2.png}
+}
+\subfigure[Capon]{
+	\includegraphics[width=0.47\textwidth]{img/scenario_mv_resp2.png}
+}
+\caption{Array beam pattern with uniform and Capon weights. An interfering source is located at 20 degrees.}
+\end{figure}
 
@@
 
+\begin{figure}
+\centering
+\subfigure[Ray-casting]{
+	\includegraphics[width=0.4\textwidth]{img/Volumeraycasting.png}
+}
+\subfigure[Opacity transfer function]{
+	\includegraphics[width=0.5\textwidth]{img/otf.png}
+}
+\caption{.}
+\label{fig:vr}
+\end{figure}
+
 \subsection{Adaptive volume rendering}

### 2014-02-05 ThesisJP/Introduction/introduction.tex
Small adjustments to the text.

@@
 \section{Motivation}
-A rapid development in computer game technology and accompanying programming languages have recently provided researchers with small personal super computers, comprised in a single graphics card (\nom{GPU}{Graphics Processing Unit}).
-The latest graphics cards\footnote{As of January 2014.} from both \nom{Nvidia}{Graphics card producer} and \nom{AMD}{Graphics card producer} provide close to 6 T\nom{FLOPS}{Floating point operations per second} single precision and 1.5 TFLOPS double precision computations.
+A rapid development in computer game technology and accompanying programming languages have recently provided researchers with small personal super computers, comprised in a single graphics card (\nom{GPU}{Graphics processing unit}).
+The latest graphics cards\footnote{As of January 2014.} from both \nom{Nvidia}{Graphics card producer} and \nom{AMD}{Graphics card producer} provide close to 6 T\nom{FLOPS}{Floating point operations per second} single precision and 1.5 TFLOPS double precision in computational power.
 For double precision this is more than the worlds most powerful supercomputer could provide in 1998.
@@
 cardiac ultrasound imaging, an effective processing rate of around 100 MFLOPS is required.
-This number, as explained in Background \ref{sec:adaptbf}, also grows with larger subarrays, an increasing number of samples, and higher frame rates.
+This number, as explained in Section \ref{sec:adaptbf}, also grows with larger subarrays, an increasing number of samples, and higher frame rates.
 Nevertheless, with modern GPUs, these levels of processing are finally available within a single card.
@@
 The first is adaptive visualization of cardiac ultrasound volumes (\textbf{Paper\,IV}) and the latter is simulation of dense ultrasound pressure fields (\textbf{Paper\,V}).
-Both share the property of being computationally complex, but consist of many independent computations which makes them perfectly suited for parallel GPU acceleration.
+Both share the property of being computationally complex, but on the other hand they consist of many independent computations which makes them perfectly suited for parallel GPU acceleration.
 
@@
 
-In this paper, to reduce the matrix size, it is taken advantage of the beamspace version of the Capon beamformer (\nom{BS-Capon}{Beamspace Capon beamforming}), which is implemented on the GPU.
-For parameters previously derived to give similar performance for BS-Capon and element-space Capon (\nom{ES-Capon}{Element Space Capon}), we show that real-time BS-Capon beamforming is feasible for cardiac ultrasound imaging.
+In this paper, to reduce the matrix size, it is taken advantage of the beamspace version of the Capon beamformer (\nom{BS-Capon}{Beamspace capon beamforming}), which is implemented on the GPU.
+For parameters previously derived to give similar performance for BS-Capon and element-space Capon (\nom{ES-Capon}{Element space capon beamforming}), we show that real-time BS-Capon beamforming is feasible for cardiac ultrasound imaging.
 The reported processing throughput is able to keep up with the acquisition frame rate in a typical cardiac ultrasound imaging system equipped with 64 and 96 element linear arrays.
@@
 However, dense simulations involve extensive calculation and computational time.
-With the presented simulator we try to reduce the time for calculating pressure fields by means of GPU acceleration.
+With the presented simulator we therefore try to reduce the time for calculating pressure fields by means of GPU acceleration.
 The tool is based on Huygens' principle which describes diffraction caused by a slit as the superposition of several point sources located inside the opening.
 Thus, the simulator works by accumulating the contribution from a collection of point sources in a set of observation points.
-How the source and observation points are laid out is for the user to decide.
+How the source and observation points are laid out is left for the user to decide.
 Point sources positioned along a line will for instance simulate an ultrasound array.

### 2014-02-06 ThesisJP/Background/background.tex
STarted writing about gpgpu.

@@
 In this chapter the background information needed in order to understand the topics of this thesis and its papers is given.
-First, in Section \ref{sec:gpgpu} an introduction to general purpose programming on graphics processing units (\nom{GPGPU}{General purpose programming on graphics processing units}) is presented.
-Next, an introduction to ultrasound imaging is given in Section \ref{sec:ultrasound}.
+First, in Section \ref{sec:gpgpu}, general-purpose programming on graphics processing units (\nom{GPGPU}{General-purpose programming on graphics processing units}) is reviewed, followed by an introduction to medical ultrasound imaging in Section \ref{sec:ultrasound}.
+Finally, the concepts of volume rendering and field simulations are presented in Section \ref{sec:volren} and \ref{sec:field}.
 
 \section{General-purpose computing on graphics processing units}\label{sec:gpgpu}
-
-Architecture. See Paper II.
-
-About theoretical FLOPS: Note that these are theoretical numbers and the actual throughput typically is algorithm dependent.
-
-Add sentence about siemens and supersonic imagine scanners. They use GPUs? What does this mean? FPGAs and ASICs.
+In a recent essay, Herb Sutter, a leading authority on software development, reviews the current state of the software and computer industry\todo{cite}.
+The essay bears the name "Welcome to the Jungle" and is a sequel to the essay "The Free Lunch is Over" from 2004.
+The two essays spins around the fact that chip manufactures hit a frequency-wall in the beginning of the 21st century.
+Until then, all mainstream computer programs where typically running in a single thread on a single core, and a programmer could expect the software to annually gain performance without touching the code.
+This was the "free-lunch" era as depicted in Fig.\,\ref{fig:jungle}.
+The increase in processing power came mostly from an ever-growing clock frequency.
+However, power consumption and heat generation where also growing with the increase in clock frequency, until the level of cooling required finally became to much in 2004.
+From that point, chip manufactures had to pursue a different approach  than just increasing the clock frequency.
+The immediate solution was to embed several cores in one CPU chip.
+This insured continuation of Moore's law, since the number of transistors per chip could continue to grow.
+For programmers, multi-core CPUs meant that computer programs now had to be multi-threaded in order to continue to gain performance.
+It was a concurrency revolution,and  the free lunch was over.
 
@@
 \end{figure}
+
+Around the same time as CPU manufactures hit the frequency wall, people started experimented with programmable shaders which recently had been introduced in the field of graphics programming.
+The graphics processing pipeline had previously been a fixed function pipeline, where e.g.
+vertices and texture coordinates where fed to the graphics processing unit (GPU) in one end and a rasterized rendered image was outputted in the other.
+What programmable shaders  introduced was the ability to replace some steps in the fixed-function pipeline with custom computations.
+This made it possible to utilize the GPU for other tasks than just rendering graphics.
+The graphics pipeline could for instance be hacked to evaluate a differential equation over grid defined by graphics primitives\todo{cite Seland et al.}.
+The rationale behind this exploration was that hidden in the GPU was processing power an order of magnitude larger than what the CPU could provide at the same time.
+Actually, at this time, the GPU had already gone multi-core, driven by the increasing demand for realistic computer games.
+Multiple cores was somewhat easier to achieve with the GPU since the rendering problem is inherently parallel.
+The pixels in the rendered image can be calculated in any order, and one pixel is therefore an concealed unit which the GPU scheduler is free to issue at any time to whatever core that is available.
+It it is also much less costly to switch between a pixel-thread for the GPU then a program thread for the CPU.
+Since there was only one program running on the GPU, the fixed function pipeline, GPU designers had a good idea of how the program would execute, and could therefore skip advance feature as predicting execution paths and different levels of data caching.
+The released silicon was used to make more cores.
+
+It is common to say that a CPU has few but advance generic cores, and the GPU has many but simple cores.
+
+Three eras at the same time: Multi-core, Heterogeneous-cores, and Cloud-cores.
+
+Early gpgpu programming required expert knowledge about graphics and the remaining fixed funtions in the pipeline. New programming languages where needed. OpenCL and CUDA. C++ AMP.
+
+Architecture and how to program. See Paper II.
+
+About theoretical FLOPS: Note that these are theoretical numbers and the actual throughput typically is algorithm dependent.
+
+Compare CPU and GPU ALUs.
+
+Add sentence about siemens and supersonic imagine scanners. They use GPUs? What does this mean? FPGAs and ASICs.
 
@@
 
-\section {Ultrasound imaging}\label{sec:ultrasound}
+\section {Medical ultrasound imaging}\label{sec:ultrasound}
 Ultrasound imaging encompasses technology which generates images based on sound whose frequencies we can not hear. For medical ultrasound imaging frequencies above 2 MHz are typically used.
@@
 
-\section{Volume rendering}
+\section{Volume rendering}\label{sec:volren}
 
@@
 
-\section{Field simulations}
+\section{Field simulations}\label{sec:field}
 

### 2014-02-08 ThesisJP/Background/background.tex
Small corrections.

@@
 The increase in processing power came mostly from an ever-growing clock frequency.
-However, power consumption and heat generation where also growing with the increase in clock frequency, until the level of cooling required finally became to much in 2004.
+However, power consumption and heat generation where also growing at the same rate, until the level of cooling required finally became to much in 2004.
 From that point, chip manufactures had to pursue a different approach  than just increasing the clock frequency.
@@
 For programmers, multi-core CPUs meant that computer programs now had to be multi-threaded in order to continue to gain performance.
-It was a concurrency revolution,and  the free lunch was over.
+It was a concurrency revolution, and  the free lunch was over.
 
@@
 
-Around the same time as CPU manufactures hit the frequency wall, people started experimented with programmable shaders which recently had been introduced in the field of graphics programming.
+Around the same time as CPU manufactures hit the frequency wall people started experimented with programmable shaders which recently had been introduced in the field of graphics programming.
 The graphics processing pipeline had previously been a fixed function pipeline, where e.g.

### 2014-02-08 ThesisJP/Introduction/introduction.tex
Small corrections.

@@
 A rapid development in computer game technology and accompanying programming languages have recently provided researchers with small personal super computers, comprised in a single graphics card (\nom{GPU}{Graphics processing unit}).
-The latest graphics cards\footnote{As of January 2014.} from both \nom{Nvidia}{Graphics card producer} and \nom{AMD}{Graphics card producer} provide close to 6 T\nom{FLOPS}{Floating point operations per second} single precision and 1.5 TFLOPS double precision in computational power.
+The latest graphics cards\footnote{As of January 2014.} from both \nom{Nvidia}{Graphics card producer} and \nom{AMD}{Graphics card producer} provide close to 6 T\nom{FLOPS}{Floating point operations per second} single precision and 1.5 TFLOPS double precision computations.
 For double precision this is more than the worlds most powerful supercomputer could provide in 1998.

### 2014-02-08 ThesisJP/Background/background.tex
Started adding references.

@@
 \section{General-purpose computing on graphics processing units}\label{sec:gpgpu}
-In a recent essay, Herb Sutter, a leading authority on software development, reviews the current state of the software and computer industry\todo{cite}.
-The essay bears the name "Welcome to the Jungle" and is a sequel to the essay "The Free Lunch is Over" from 2004.
+In a recent essay, Herb Sutter, a leading authority on software development, reviews the current state of the software and computer industry \cite{HerbSutter}.
+The essay bears the name "Welcome to the Jungle" and is a sequel to the essay "The Free Lunch is Over" from 2004 \cite{HerbSuttera}.
 The two essays spins around the fact that chip manufactures hit a frequency-wall in the beginning of the 21st century.
@@
 \subfigure[Kernel]{
-	\includegraphics[width=0.5\textwidth]{img/kernel.pdf}
+	%\includegraphics[width=0.5\textwidth]{img/kernel.pdf}
 }

### 2014-02-08 ThesisJP/Introduction/introduction.tex
Started adding references.

@@
 The second paper aims at implementing real-time Capon beamforming for cardiac ultrasound imaging.
-Achieving this will facilitate further study of this method's \textit{in vivo} performance.
-This is something which has been sought for the past decade \todo{Add citations}.
+Achieving this will facilitate further study of the method's \textit{in vivo} performance.
+This is something which has been stated as further work in many publications on Capon beamforming for medical ultrasound imaging the past decade.
 In \textbf{Paper I} arrays with no more than 32 elements were investigated.
@@
 The paper presents both quantitative and qualitative results supporting the method's improved capabilities of reducing tissue dropouts and spurious structures inside the blood pool, leading to a maximized amount of visible endocardium tissue.
-The proposed method is implemented on a GPU with interactive frame rates as the result.
-\todo{litt tafatt.
-m\aa sm\oe{}re p\aa litt}
+The proposed method is implemented on a GPU with real-time frame rates as the result.
+This results shows that with modern GPUs it is possible to add more advanced visualization to an ultrasound imaging system and still have real-time performance.
 
@@
 However, dense simulations involve extensive calculation and computational time.
-With the presented simulator we therefore try to reduce the time for calculating pressure fields by means of GPU acceleration.
+With the presented simulator we aim at reducing the time needed for calculating pressure fields by means of GPU acceleration.
 The tool is based on Huygens' principle which describes diffraction caused by a slit as the superposition of several point sources located inside the opening.
@@
 \item a GPU implementation of this adaptive volume rendering method.
-\item a GPU implementation of simple paint-like simulation tool for rapid visualization ultrasound pressure fields.
+\item a GPU implementation of simple Paint-like simulation tool for rapid visualization ultrasound pressure fields.
 \end{enumerate}
@@
 This method has linear complexity and shares many properties with the Capon beamformer.
-If only minor improvements are obtained, it would be easier to ignore this fact if the algorithm is less costly.
+If only minor improvements are obtained, it would be easier to ignore this fact if the algorithm is less computationally costly.
 Finally, it would also be interesting to see how the Eigenspace-based Capon beamformer performs on cardiac images, and how its absence of a distortionless criterion will influence the image spatially and temporally.
@@
 The reason could be ignorance or its possible negative impact on imaging frame rate.
-A lack of studies where Capon beamforming is performed on consecutive frames might also explain its absence.
+A lack of studies where Capon beamforming is performed on consecutive frames might also explain the lack of comments.
 Fortunately, as shown in \textbf{Paper III}, it is possible to maintain the same frame rate as with DAS beamforming by applying oversampling on receive using a set of steering vectors (phase delays).
@@
 Good looking speckle, or DAS-like speckle, has been the driving force behind many of the robustification techniques and the default parameters seen in the literature.
-Hence, more aggressive parameters might be used if the image produced is intended as input to another algorithm.
+Hence, more aggressive parameters might be used if the image produced is intended as input to another algorithm and not for the human eye.
 \\\\
@@
 Another issue is the view-dependent visibility, where an object could become visible or disappear while rotating or if the local statistics change a lot during the cardiac cycle.
-This means that the rendering will be come non-intuitive to interact with and unrealistic to watch.
+Even if the method successfully removes noise, this means that the rendering will become non-intuitive to interact with and unrealistic to watch.
 However, the method could still be used to clean up still images of standard cardiac views.
@@
 Even though the visualization based on global statistics has more errors, with respect to endocardium visibility, the tuning of the opacity function is still fully automatic if an iterative search for the minima in Figure 3d is performed.
-The rendering will then be realistic, but still dependent on a successful tracking for the auto-adjustment to work.
+A global opacity transfer function will assure a rendering which are realistic, but it will still be dependent on a successful tracking for the auto-adjustment to work.
 An automatically tuned opacity function based on global statistics combined with an improved smoothing scheme could be the future for cardiac ultrasound imaging.
@@
 It has to combined with either a good analysis of what special adjustments that had to be made to the algorithm in order for the increase to happen or a suggestion of how to utilize the speed gained.
-These are points we have tried to answer in both \textbf{Paper I}, \textbf{II},  \textbf{III}, and \textbf{V}.
+These are points we have tried to answer in both \textbf{Paper I}, \textbf{II},  \textbf{IV}, and \textbf{V}.
 In \textbf{Paper V} the increased speed makes an interactive paint program feasible.

### 2014-02-09 ThesisJP/Background/background.tex
Added new citation. Some small adjustments.

@@
 What programmable shaders  introduced was the ability to replace some steps in the fixed-function pipeline with custom computations.
-This made it possible to utilize the GPU for other tasks than just rendering graphics.
-The graphics pipeline could for instance be hacked to evaluate a differential equation over grid defined by graphics primitives\todo{cite Seland et al.}.
+This made it possible to utilize the GPU for other tasks than just rendering graphics \cite{Seland2007}.
 The rationale behind this exploration was that hidden in the GPU was processing power an order of magnitude larger than what the CPU could provide at the same time.
@@
 \subfigure[Kernel]{
-	%\includegraphics[width=0.5\textwidth]{img/kernel.pdf}
+	\includegraphics[width=0.5\textwidth]{img/kernel.pdf}
 }

### 2014-02-09 ThesisJP/Thesis.tex
Added symbol list. Started writing the introduction.
Adjusted title.
Writing. Still some sections unfinished in discussion part.
Started adding figures in background section.
STarted writing about gpgpu.
Started adding references.
Added new citation. Some small adjustments.

@@
 %Real-Time Adaptive Ultrasound Algorithms
-Accelerating adaptive ultrasound algorithms using graphics processing units 
+%Accelerating adaptive ultrasound algorithms using graphics processing units 
+%Accelerating adaptive ultrasound algorithms with graphics processing units 
+Accelerating adaptive ultrasound algorithms by means of general-purpose programming on graphics processing units 
+%Accelerating adaptive ultrasound algorithms by general purpose programming on graphics processing units 
+%Accelerating ultrasound algorithms with graphics processing units 
+%Accelerating ultrasound algorithms using graphics processing units 
 \vspace{0.5cm}}
@@
 
+% symbol list
+\usepackage{nomencl}
+\makenomenclature
+\renewcommand{\nomname}{Symbols and Abbreviations}
+\newcommand*{\nom}[2]{#1\nomenclature{#1}{#2}}
+
 \begin{document}
+
 \date{\today}
@@
 1, pp.
-\todo{add pages} Jan 2014}.
+76-85, Jan 2014}.
  	\item \textbf{J.\:P.\:\AA{}sen}, A.\:Austeng and S.\:Holm, "Capon beamforming and moving objects - An analysis of lateral shift-invariance", {\it IEEE Transactions on Ultrasonics, Ferroelectrics, and Frequency Control, submitted}.
 	\item \textbf{J.\:P.\:\AA{}sen}, E.\:Steen, G.\:Kiss, A.\:Thorstensen and S.\:I.\:Rabben, "Adaptive volume rendering of cardiac 3D ultrasound images - utilizing blood pool statistics", {\it Proc.
-SPIE Medical Imaging 2012 \todo{add pages}}.
-	\item \textbf{J.\:P.\:\AA{}sen} and S.\:Holm, "Huygens on speed: Interactive simulation of ultrasound pressure fields", {\it Proc. IEEE Ultrasonics Symposium 2012 \todo{add pages}}.
+SPIE Medical Imaging 2012, vol.
+8320, pp.
+832008.}.
+	\item \textbf{J.\:P.\:\AA{}sen} and S.\:Holm, "Huygens on speed: Interactive simulation of ultrasound pressure fields", {\it Proc. IEEE Ultrasonics Symposium 2012, pp. 1643-1646}.
 \end{enumerate}
@@
 	\item \textbf{J.\:P.\:\AA{}sen}, J.\:I.\:Buskenes, C.-I.\:C.\:Nilsen, A.\:Austeng and S.\:Holm, "Implementing Capon Beamforming on the GPU for Real-Time Cardiac Ultrasound Imaging", {\it Proc.
-IEEE Ultrasonics Symposium 2012} \todo{add pages}.
+IEEE Ultrasonics Symposium 2012}, pp.
+2133-2136.
 
 	\item J.\:I.\:Buskenes, \textbf{J.\:P.\:\AA{}sen}, C.-I.\:C.\:Nilsen and A.\:Austeng, "Adapting the Minimum variance beamformer to a graphics processing unit for active sonar imaging systems", {\it Proc.
-Meetings on Acoustics 2013} \todo{add pages?}.
+Meetings on Acoustics 2013} vol.
+13.
 
 	\item G.\:Kiss, E.\:Steen, \textbf{J.\:P.\:\AA{}sen} and H.\:Torp, "GPU volume rendering in 3D echocardiography: real-time pre-processing and ray-casting", {\it Proc.
-IEEE Ultrasonics Symposium 2010} \todo{add pages}.
+IEEE Ultrasonics Symposium 2010}, pp.
+193-196.
 
@@
 	\item E.\:S.\:Br\o{}nstad, \textbf{J.\:P.\:\AA{}sen}, H.\:Torp and G.\:Kiss, "Visibility driven visualization of 3D cardiac ultrasound data on the GPU", {\it Proc.
-IEEE Ultrasonics Symposium 2012} \todo{add pages}.
+IEEE Ultrasonics Symposium 2012}, pp.
+2651-2654.
 \end{enumerate}
@@
 
-\clearpage
+\cleardoublepage
+\addcontentsline{toc}{chapter}{Symbols and Abbreviations}
+\printnomenclature
+
+\cleardoublepage
 \setcounter{page}{1}
@@
 
-\renewcommand*\thesection{\arabic{section}}
-
-%%%% CHAPTER 1 %%%%
-\begin{bibunit}[ieeetr]
-\CustChapter*{Introduction} 
-\addcontentsline{toc}{chapter}{Introduction}
-\runningtitle{Introduction}
-\input{./Introduction/introduction}
-\newpage\addcontentsline{toc}{section}{References}
-\putbib[bibIntroduction]
-\end{bibunit}
-
-\setcounter{section}{0}
-\setcounter{chapter}{1}
+%\renewcommand*\thesection{\arabic{section}}
 
@@
 \begin{bibunit}[ieeetr]
-\CustChapter*{Background}
-\addcontentsline{toc}{chapter}{Background} 
+\CustChapter{Background}
 \runningtitle{Background}
@@
 \newpage\addcontentsline{toc}{section}{References}
-\putbib[bibBackground]
+\putbib[./Background/background]
 \end{bibunit}
+
+%\setcounter{section}{0}
+%\setcounter{chapter}{1}
+
+%%%% CHAPTER 1 %%%%
+\CustChapter{Introduction} 
+\runningtitle{Introduction}
+\input{./Introduction/introduction}
 
@@
 }
-\noindent \textit{IEEE Transactions on Oceanic Engineering, submitted}
+\noindent \textit{IEEE Transactions on Oceanic Engineering, submitted}.
 %\newpage\input{./CaponSonarGPU/CaponSonarGPU}
 %\newpage\addcontentsline{toc}{section}{References}
-\putbib[bibCaponSonarGPU]
+%\putbib[./CaponSonarGPU/bibCaponSonarGPU]
 \end{bibunit}
@@
 }
-\noindent \textit{IEEE Transactions on Ultrasonics, Ferroelectrics, and Frequency Control, vol. 61, Jan 2014}
+\noindent \textit{IEEE Transactions on Ultrasonics, Ferroelectrics, and Frequency Control, vol. 61, pp. 76-85, Jan 2014}.
 %\input{./CaponCardiacGPU/CaponCardiacGPU}
 %\newpage\addcontentsline{toc}{section}{References}
-\putbib[bibCaponCardiacGPU]
+%\putbib[./CaponCardiacGPU/bibCaponCardiacGPU]
 \end{bibunit}
@@
 }
-\noindent \textit{IEEE Transactions on Ultrasonics, Ferroelectrics, and Frequency Control, submitted}
+\noindent \textit{IEEE Transactions on Ultrasonics, Ferroelectrics, and Frequency Control, submitted}.
 %\input{./CaponMovingObjects/CaponMovingObjects}
 %\newpage\addcontentsline{toc}{section}{References}
-\putbib[bibCaponMovingObjects]
+%\putbib[./CaponMovingObjects/bibCaponMovingObjects]
 \end{bibunit}
@@
 }
-\noindent \textit{Proc. SPIE Medical Imaging 2012 \todo{add pages}}
+\noindent \textit{Proc. SPIE Medical Imaging 2012, vol. 8320, pp. 832008.}
 %\input{./AdaptiveVolumeRendering/AdaptiveVolumeRendering}
 %\newpage\addcontentsline{toc}{section}{References}
-\putbib[bibAdaptiveVolumeRendering]
+%\putbib[./AdaptiveVolumeRendering/bibAdaptiveVolumeRendering]
 \end{bibunit}
@@
 }
-\noindent \textit{Proc. IEEE Ultrasonics Symposium 2012 \todo{add pages}}
+\noindent \textit{Proc. IEEE Ultrasonics Symposium 2012, pp. 1643-1646}.
 %\input{./Hos/Hos}
 %\newpage\addcontentsline{toc}{section}{References}
-\putbib[bibHos]
+%\putbib[./Hos/bibHos]
 \end{bibunit}
 
-\appendix
+%\appendix
 

### 2014-02-10 ThesisJP/Background/background.tex
Continued writing background information.

@@
 First, in Section \ref{sec:gpgpu}, general-purpose programming on graphics processing units (\nom{GPGPU}{General-purpose programming on graphics processing units}) is reviewed, followed by an introduction to medical ultrasound imaging in Section \ref{sec:ultrasound}.
-Finally, the concepts of volume rendering and field simulations are presented in Section \ref{sec:volren} and \ref{sec:field}.
+Finally, the concepts of volume rendering and ultrasound field simulations are presented in Section \ref{sec:volren} and \ref{sec:field}.
 
@@
 Until then, all mainstream computer programs where typically running in a single thread on a single core, and a programmer could expect the software to annually gain performance without touching the code.
-This was the "free-lunch" era as depicted in Fig.\,\ref{fig:jungle}.
+This was the "free-lunch"-era as depicted in Fig.\,\ref{fig:jungle}.
 The increase in processing power came mostly from an ever-growing clock frequency.
 However, power consumption and heat generation where also growing at the same rate, until the level of cooling required finally became to much in 2004.
-From that point, chip manufactures had to pursue a different approach  than just increasing the clock frequency.
+From that point, chip manufactures had to pursue a different direction  than just increasing the clock frequency.
 The immediate solution was to embed several cores in one CPU chip.
@@
 For programmers, multi-core CPUs meant that computer programs now had to be multi-threaded in order to continue to gain performance.
-It was a concurrency revolution, and  the free lunch was over.
+It was a concurrency revolution, as Herb Sutter named it, and  the "free lunch" was over.
 
@@
 \includegraphics[width=0.9\textwidth]{img/free_lunsh.png}
-\caption{Illustration of several transitions currently taking place in the computer industry. \todo{Site Herb Sutter - Welcom to the jungle http://herbsutter.com/welcome-to-the-jungle/}}
+\caption{Illustration of transitions between different eras in the computer industry.
+Currently there are three ongoing transitions, multi-core, heterogeneous-core and cloud-core.
+Image courtesy of Herb Sutter (herbsutter.com).}
 \label{fig:jungle}
@@
 
-Around the same time as CPU manufactures hit the frequency wall people started experimented with programmable shaders which recently had been introduced in the field of graphics programming.
-The graphics processing pipeline had previously been a fixed function pipeline, where e.g.
-vertices and texture coordinates where fed to the graphics processing unit (GPU) in one end and a rasterized rendered image was outputted in the other.
+Around the same time as CPU manufactures hit the frequency wall, people started experimenting with programmable shaders which recently had been introduced in the field of graphics programming.
+The graphics processing pipeline had before this been a fixed function pipeline, where e.g.
+geometry and textures where fed to the graphics processing unit (GPU) in one end and a rasterized rendered image was outputted in the other.
 What programmable shaders  introduced was the ability to replace some steps in the fixed-function pipeline with custom computations.
 This made it possible to utilize the GPU for other tasks than just rendering graphics \cite{Seland2007}.
-The rationale behind this exploration was that hidden in the GPU was processing power an order of magnitude larger than what the CPU could provide at the same time.
-Actually, at this time, the GPU had already gone multi-core, driven by the increasing demand for realistic computer games.
-Multiple cores was somewhat easier to achieve with the GPU since the rendering problem is inherently parallel.
-The pixels in the rendered image can be calculated in any order, and one pixel is therefore an concealed unit which the GPU scheduler is free to issue at any time to whatever core that is available.
-It it is also much less costly to switch between a pixel-thread for the GPU then a program thread for the CPU.
-Since there was only one program running on the GPU, the fixed function pipeline, GPU designers had a good idea of how the program would execute, and could therefore skip advance feature as predicting execution paths and different levels of data caching.
-The released silicon was used to make more cores.
+The rationale behind this exploration was that hidden in the GPU was processing powers an order of magnitude larger than what the CPU could provide at the same time (See Fig.\,\ref{fig:cpu_vs_gpu}).
+Actually, at this time the GPU had already gone multi-core, driven by an ever-increasing demand for more realistic computer games.
+What also made this possible was the inherent parallel nature of the rendering problem.
+When geometry is projected and rasterized the same instruction is typically needed for a lot of data at the same time.
+This makes a special kind of instruction, Single Instruction Multiple Data (\nom{SIMD}{Single instruction multiple data (instruction)}) instruction, especially suited for this kind of problem.
+The rendering pipeline also exhibits minimal execution path branching, and data are typically used once.
+Designers of GPUs could therefore skip advance feature, found in most CPUs, like branch prediction and different levels of data caching.
+This reduced the silicon footprint of each core and made it possible to add multiple SIMD cores to the same chip before CPU designers where able to do the same.
+Today, even if more advanced caching has added to GPUs, this is still the main reasons why GPUs have higher peak performance than CPUs (as shown in Fig.\,\ref{fig:cpu_vs_gpu}).
 
-It is common to say that a CPU has few but advance generic cores, and the GPU has many but simple cores.
+The first adopters of GPU programming, or General-purpose programming on graphics processing units (GPGPU), wanted to use the GPU to accelerate computationally demanding algorithms.
+
+FPGAs (Field-programmable gate array), APU (Accelerated processing unit).
 
@@
 
+\subsection{Comparing CPU and GPU performance} 
+It is common to say that a CPU has few but advance generic cores, and the GPU has many but simple cores.
+
+It it is also much less costly to switch between a pixel-thread for the GPU then a program thread for the CPU.
+
+About theoretical FLOPS: Note that these are theoretical numbers and the actual throughput typically is algorithm dependent.
+
+Compare CPU and GPU ALUs.
+SSE, AVX ...
+The GPU has super-wide registers.
+The SM in Fermi have 32 cores or 1024 bit registers.
+The SMX in Kepler have 192 cores.
+6144 bit registers.
+SSE (Streaming SIMD Extensions) instructions have 128 bit registers.
+AVX (Advanced Vector Extensions) have 256 bit registers.
+AVX-512 has 512 bit registers.
+Single instruction, multiple data (SIMD).
+
+\begin{figure}
+\centering
+\includegraphics[width=0.9\textwidth]{img/cpu-vs-gpu.png}
+\caption{Development in theoretical peak throughput, single precision (SP) and double precision (DP), the last decade for Intel CPUs and Nvidia GPUs.
+Image courtesy of Michael Galloy (michaelgalloy.com).}
+\label{fig:cpu_vs_gpu}
+\end{figure}
+
+\subsection{Programming the GPU}
 Early gpgpu programming required expert knowledge about graphics and the remaining fixed funtions in the pipeline. New programming languages where needed. OpenCL and CUDA. C++ AMP.
@@
 
-About theoretical FLOPS: Note that these are theoretical numbers and the actual throughput typically is algorithm dependent.
+Hopefully, in the future, GPU programming will be the job of e.g.
+c++ compiler, and direct fiddling with GPU kernels should only be needed if maximum throughput is absolutely needed.
+(In the same way the compiler is able to translate c++ code into SSE and AVX instructions.)
 
-Compare CPU and GPU ALUs.
-
+\subsection{Accelerators in medical ultrasound imaging }
 Add sentence about siemens and supersonic imagine scanners. They use GPUs? What does this mean? FPGAs and ASICs.
@@
 
+Typical probes, scan sequences, resolution and sampling.
+
+Key selling points of ultrasound imaging: cost, safety and real-time user interaction.
+
+One sentence about speckle tracking.
+
+\subsection{Cardiac anatomy}
 Small section about heart anatomy.
@@
 \end{figure}
-
-Typical probes, scan sequences, resolution and sampling.
-
-Key selling points of ultrasound imaging: cost, safety and real-time user interaction.
-
-One sentence about speckle tracking.
 

### 2014-02-10 ThesisJP/Introduction/introduction.tex
Continued writing background information.

@@
 \textbf{J.\:P.\:\AA{}sen}, J.\:I.\:Buskenes, C.-I.\:C\:Nilsen, A.\:Austeng and S.\:Holm\\
-{\it IEEE Transactions on Ultrasonics, Ferroelectrics, and Frequency Control, vol. 61, no. 1, pp. \todo{add pages}, Jan. 2014.}\\\\
+{\it IEEE Transactions on Ultrasonics, Ferroelectrics, and Frequency Control, vol. 61, no. 1, pp. 76-85, Jan. 2014.}\\\\
 The second paper aims at implementing real-time Capon beamforming for cardiac ultrasound imaging.
@@
 \textbf{J.\:P.\:\AA{}sen}, E.\:Steen, G.\:Kiss, A.\:Thorstensen and S.\:I.\:Rabben\\
-{\it Proc. SPIE Medical Imaging 2012, \todo{add pages}.}\\\\
+{\it Proc. SPIE Medical Imaging 2012, vol. 8320, pp. 832008.}\\\\
 The fourth paper propose an adaptive volume rendering method based on statistics derived from the blood pool of the left ventricle.
@@
 \textbf{J.\:P.\:\AA{}sen} and S.\:Holm\\
-{\it Proc. IEEE Ultrasonics Symposium 2012 \todo{add pages}.}\\\\
+{\it Proc. IEEE Ultrasonics Symposium 2012, pp. 1643-1646.}\\\\
 The fifth paper presents an interactive simulations tool capable of simulating dense pressure fields from ultrasound arrays in real time.
@@
 An application where the focus is detection of closely spaced point targets, might be better suited.
-A better suited adaptive beamformer for cardiac applications might be the low-complexity adaptive beamformer (\nom{LCA beamformer}{Low complexity adaptive beamformer}), where the Capon optimization problem is applied on a set of predefined windows.
+A better suited adaptive beamformer for cardiac applications might be the low-complexity adaptive beamformer (\nom{LCA}{Low complexity adaptive (beamformer)} beamformer), where the Capon optimization problem is applied on a set of predefined windows.
 This method has linear complexity and shares many properties with the Capon beamformer.

### 2014-02-11 ThesisJP/Background/background.tex
Updated cpu vs gpu figure.
Write on background.

@@
 \caption{Illustration of transitions between different eras in the computer industry.
-Currently there are three ongoing transitions, multi-core, heterogeneous-core and cloud-core.
+Currently there are three ongoing transitions; multi-core, heterogeneous-core and cloud-core.
 Image courtesy of Herb Sutter (herbsutter.com).}
@@
 This made it possible to utilize the GPU for other tasks than just rendering graphics \cite{Seland2007}.
-The rationale behind this exploration was that hidden in the GPU was processing powers an order of magnitude larger than what the CPU could provide at the same time (See Fig.\,\ref{fig:cpu_vs_gpu}).
-Actually, at this time the GPU had already gone multi-core, driven by an ever-increasing demand for more realistic computer games.
-What also made this possible was the inherent parallel nature of the rendering problem.
+The rationale behind this exploration was that GPUs were already a highly available processor, and hidden behind its graphics interface was processing powers an order of magnitude larger than what the CPU could provide (Fig.\,\ref{fig:cpu_vs_gpu}).
+Actually, at the time when CPUs hit the frequency wall the GPU had already gone multi-core, driven by an ever-increasing demand for more realistic computer games.
+The rendering pipeline exhibits minimal execution path branching, and data are typically used once.
+Designers of GPUs could therefore skip advance feature, found in most CPUs, like branch prediction and different levels of data caching.
+This reduced the silicon footprint of each core and made it possible to add multiple cores to the same chip before CPU designers where able to do the same.
+Today, even if more advanced caching has been added to GPUs, this is still the main reason why GPUs have higher peak performance than CPUs (Fig.\,\ref{fig:cpu_vs_gpu}).
+Another reason is the inherent parallel nature of the rendering problem.
 When geometry is projected and rasterized the same instruction is typically needed for a lot of data at the same time.
 This makes a special kind of instruction, Single Instruction Multiple Data (\nom{SIMD}{Single instruction multiple data (instruction)}) instruction, especially suited for this kind of problem.
-The rendering pipeline also exhibits minimal execution path branching, and data are typically used once.
-Designers of GPUs could therefore skip advance feature, found in most CPUs, like branch prediction and different levels of data caching.
-This reduced the silicon footprint of each core and made it possible to add multiple SIMD cores to the same chip before CPU designers where able to do the same.
-Today, even if more advanced caching has added to GPUs, this is still the main reasons why GPUs have higher peak performance than CPUs (as shown in Fig.\,\ref{fig:cpu_vs_gpu}).
+The differences between how SIMD instructions are utilized in modern CPUs and GPUs is discussed in Section \ref{sec:cpu_vs_gpu}.
 
-The first adopters of GPU programming, or General-purpose programming on graphics processing units (GPGPU), wanted to use the GPU to accelerate computationally demanding algorithms.
+For the first adopters of GPU programming, or General-purpose programming on graphics processing units (GPGPU) as it was named, it soon became evident that offloading all types of computations to the GPU was not always a good idea.
+In most cases it is important to balance the load equally between the CPU and GPU to obtain their combined computationally power.
+Balancing typically means that highly parallel tasks should be offloaded to the GPU and serialized work should stay on the CPU.
+This paradigm of utilizing specialized cores for solving specific problems is today known as heterogeneous computing.
+The specialized core can
 
@@
 
-Three eras at the same time: Multi-core, Heterogeneous-cores, and Cloud-cores.
+Three currently ongoing transitions: Multi-core, Heterogeneous-cores, and Cloud-cores.
 
-\subsection{Comparing CPU and GPU performance} 
+This thesis is focusing on heterogeneous computing with CPUs and GPUs.
+
+\subsection{Comparing CPU and GPU performance}\label{sec:cpu_vs_gpu}
 It is common to say that a CPU has few but advance generic cores, and the GPU has many but simple cores.

### 2014-02-12 ThesisJP/Background/background.tex
updated cpu vs gpu figure.
Some writing on the background.

@@
 This made it possible to utilize the GPU for other tasks than just rendering graphics \cite{Seland2007}.
-The rationale behind this exploration was that GPUs were already a highly available processor, and hidden behind its graphics interface was processing powers an order of magnitude larger than what the CPU could provide (Fig.\,\ref{fig:cpu_vs_gpu}).
-Actually, at the time when CPUs hit the frequency wall the GPU had already gone multi-core, driven by an ever-increasing demand for more realistic computer games.
-The rendering pipeline exhibits minimal execution path branching, and data are typically used once.
+The rationale behind this exploration was that GPUs were already highly available in consumer PCs, and hidden behind its graphics interface were processing powers an order of magnitude larger than what the CPU could provide (Fig.\,\ref{fig:cpu_vs_gpu}).
+Actually, at the time when CPUs hit the frequency wall the GPU had already gone multi-core, first of all driven by an ever-increasing demand for more realistic computer games.
+Second, the rendering pipeline exhibits minimal execution path branching, and data are typically used once.
 Designers of GPUs could therefore skip advance feature, found in most CPUs, like branch prediction and different levels of data caching.
-This reduced the silicon footprint of each core and made it possible to add multiple cores to the same chip before CPU designers where able to do the same.
+These measures reduced the silicon footprint of each core and made it possible to add multiple cores to one GPU before CPU designers where able to do the same.
 Today, even if more advanced caching has been added to GPUs, this is still the main reason why GPUs have higher peak performance than CPUs (Fig.\,\ref{fig:cpu_vs_gpu}).
@@
 This makes a special kind of instruction, Single Instruction Multiple Data (\nom{SIMD}{Single instruction multiple data (instruction)}) instruction, especially suited for this kind of problem.
-The differences between how SIMD instructions are utilized in modern CPUs and GPUs is discussed in Section \ref{sec:cpu_vs_gpu}.
+The differences between how SIMD instructions are utilized in modern CPUs and GPUs are discussed in Section \ref{sec:cpu_vs_gpu}.
 
@@
 This paradigm of utilizing specialized cores for solving specific problems is today known as heterogeneous computing.
-The specialized core can
+The specialized core can be a GPU or e.g.
+a field-programmable gate array (\nom{FPGA}{Field-programmable gate array}).
+The latest trend is that specialized accelerator cores are embedded onto the CPU.
+This heterogeneous processors is often referred to as an accelerated processing unit (\nom{APU}{Accelerated processing unit}).
+The last three generations of CPUs from Intel have all had on-chip integrated graphics.
 
-FPGAs (Field-programmable gate array), APU (Accelerated processing unit).
+The final transition currently taking place in the computer industry is the introduction of cloud computing, where compute power is provided as a web-service and automatically scales on demand.
+Together, multi-core, heterogeneous cores, and cloud cores forms the "jungle" that todays software engineers have to navigate (Fig.\,\ref{fig:jungle}).
 
-Three currently ongoing transitions: Multi-core, Heterogeneous-cores, and Cloud-cores.
-
-This thesis is focusing on heterogeneous computing with CPUs and GPUs.
+This thesis focus on heterogeneous computing with one multi-core CPU and a discrete high performance GPU, a cheap system which is likely to be the specification of current and future ultrasound imaging system.
 
@@
 It is common to say that a CPU has few but advance generic cores, and the GPU has many but simple cores.
+
+ the most powerful one (HD 5200) have a peak performance of nearly one tera single precision floating point operations per second (FLOPs).
 
@@
 \centering
-\includegraphics[width=0.9\textwidth]{img/cpu-vs-gpu.png}
+\includegraphics[width=\textwidth]{img/cpu_vs_gpu.pdf}
 \caption{Development in theoretical peak throughput, single precision (SP) and double precision (DP), the last decade for Intel CPUs and Nvidia GPUs.
@@
 Typical probes, scan sequences, resolution and sampling.
+
+Sampling rates, and data size.
+%Due to the high bandwidth requirement for medical ultrasound imaging, as we will see in Section \ref{sec:ultrasound}, a cloud computing ultrasound imaging system will not feasibly for a long time.
+Because of the high data rate and processing requirements ultrasound systems where first analog systems up to the mid 90s. The first digital systems where made with custom boards and ASICs.
 

### 2014-02-14 ThesisJP/Background/background.tex
Done some writing.

@@
 The rationale behind this exploration was that GPUs were already highly available in consumer PCs, and hidden behind its graphics interface were processing powers an order of magnitude larger than what the CPU could provide (Fig.\,\ref{fig:cpu_vs_gpu}).
-Actually, at the time when CPUs hit the frequency wall the GPU had already gone multi-core, first of all driven by an ever-increasing demand for more realistic computer games.
+Actually, at the time when CPU manufacturers hit the frequency wall the GPU had already gone multi-core, first of all driven by an ever-increasing demand for more realistic computer games.
 Second, the rendering pipeline exhibits minimal execution path branching, and data are typically used once.
@@
 When geometry is projected and rasterized the same instruction is typically needed for a lot of data at the same time.
-This makes a special kind of instruction, Single Instruction Multiple Data (\nom{SIMD}{Single instruction multiple data (instruction)}) instruction, especially suited for this kind of problem.
+This makes a special kind of instruction, Single Instruction Multiple Data (\nom{SIMD}{Single instruction multiple data (instruction type)}) instruction \cite{Flynn1966}, especially suited for this kind of problem.
 The differences between how SIMD instructions are utilized in modern CPUs and GPUs are discussed in Section \ref{sec:cpu_vs_gpu}.
@@
 a field-programmable gate array (\nom{FPGA}{Field-programmable gate array}).
-The latest trend is that specialized accelerator cores are embedded onto the CPU.
+A recent trend is that specialized accelerator cores are embedded onto the CPU.
 This heterogeneous processors is often referred to as an accelerated processing unit (\nom{APU}{Accelerated processing unit}).
 The last three generations of CPUs from Intel have all had on-chip integrated graphics.
+Even if an on-chip GPU is less powerful than a discrete GPU, valuable time is saved by not having to send data across the PCI-express bus, which connects the different parts of a computer to the motherboard.
+This is the reason why the first rules of GPGPU is to minimize PCI-express traffic.
 
 The final transition currently taking place in the computer industry is the introduction of cloud computing, where compute power is provided as a web-service and automatically scales on demand.
-Together, multi-core, heterogeneous cores, and cloud cores forms the "jungle" that todays software engineers have to navigate (Fig.\,\ref{fig:jungle}).
-
+Together, multi-core, heterogeneous cores, and cloud cores forms the "jungle" that today's software engineers have to tackle (Fig.\,\ref{fig:jungle}).
 This thesis focus on heterogeneous computing with one multi-core CPU and a discrete high performance GPU, a cheap system which is likely to be the specification of current and future ultrasound imaging system.
@@
 It is common to say that a CPU has few but advance generic cores, and the GPU has many but simple cores.
+
+The SIMD registers on a GPU is longer than on a CPU and the GPU have more of them.
 
@@
 
-\subsection{Programming the GPU}
+\subsection{Programming a GPU}
 Early gpgpu programming required expert knowledge about graphics and the remaining fixed funtions in the pipeline. New programming languages where needed. OpenCL and CUDA. C++ AMP.

### 2014-02-14 ThesisJP/Introduction/introduction.tex
Done some writing.

@@
 
-Well on my way into the project, when a real-time Capon beamformer was realized, and loops of images for the first time were processed at interactive and real-time frame rates, new issues where discovered and had to be solved.
-This led to a more theoretical study of the Capon beamformer (\textbf{Paper III}), with special attention on how to obtain high lateral resolution while preserving the important shift-invariant property of ultrasound imaging.
+Well on my way into the project, when a real-time Capon beamformer was realized, and loops of images for the first time were processed at real-time frame rates, new issues where discovered and had to be solved.
+This led to a more theoretical study of the Capon beamformer in \textbf{Paper\,III}, with special attention on how to obtain high lateral resolution while preserving the important shift-invariant property of ultrasound imaging.
 Shift-invariant behavior is crucial if the method is ever to be applied for live scanning.
@@
 
-\subsection{Paper I}
+\subsection{Paper\,I}
 \textbf{An Optimised GPU Implementation of the MVDR Beamformer for Active Sonar Imaging}\\
@@
 Even though the paper is written within the field of active \nom{sonar}{SOund Navigation And Ranging (usually under water)} imaging, the depicted implementation is applicable to a vast range of active imaging systems.
-A similar discussion for cardiac ultrasound imaging can be found in \textbf{Paper VI}.
+A similar discussion for cardiac ultrasound imaging can be found in \textbf{Paper\,VI}.
 Active sonar imaging typically differs from medical ultrasound imaging by a lower real-time requirement and fewer array elements.
@@
 
-\subsection{Paper II}
+\subsection{Paper\,II}
 \textbf{Implementing Capon Beamforming on a GPU for Real-Time Cardiac Ultrasound Imaging}\\
@@
 This is something which has been stated as further work in many publications on Capon beamforming for medical ultrasound imaging the past decade.
-In \textbf{Paper I} arrays with no more than 32 elements were investigated.
+In \textbf{Paper\,I} arrays with no more than 32 elements were investigated.
 A linear array for cardiac ultrasound imaging typically has 64 or more elements.
-In \textbf{Paper VI}, which is summarized in \textbf{Paper II}, it is shown that our implementation from  \textbf{Paper I} does not reach the level of throughput required for real-time Capon beamforming of cardiac ultrasound imaging.
+In \textbf{Paper\,VI}, which is summarized in \textbf{Paper\,II}, it is shown that our implementation from  \textbf{Paper\,I} does not reach the level of throughput required for real-time Capon beamforming of cardiac ultrasound imaging.
 The matrices that have to be inverted become to large, and the frames that need to be processed per second are too many.
@@
 The reported processing throughput is able to keep up with the acquisition frame rate in a typical cardiac ultrasound imaging system equipped with 64 and 96 element linear arrays.
-\textbf{Paper II} and \textbf{VI} also presents, for the first time, videos where the ES-Capon and BS-Capon beamformer have been applied on loops of simulated and \textit{in vivo} medical ultrasound images.
+\textbf{Paper\,II} and \textbf{VI} also presents, for the first time, videos where the ES-Capon and BS-Capon beamformer have been applied on loops of simulated and \textit{in vivo} medical ultrasound images.
 
-\subsection{Paper III}
+\subsection{Paper\,III}
 \textbf{Capon Beamforming and Moving Objects - An Analysis of Lateral Shift-Invariance}\\
@@
 
-\subsection{Paper IV}
+\subsection{Paper\,IV}
 \textbf{Adaptive Volume Rendering of Cardiac 3D Ultrasound Images - Utilizing Blood Pool Statistics}\\
@@
 
-\subsection{Paper V}
+\subsection{Paper\,V}
 \textbf{Huygens on Speed: Interactive Simulation of Ultrasound Pressure Fields}\\
@@
 \item an investigation of shift-invariance when the Capon beamformer is applied on consecutive frames.
-\item a method for improved shift-invariance of the Capon beamformer.
+\item a method for improved shift-invariance of the Capon beamformer when applied to medical ultrasound imaging.
 \item a method for reduced blood-pool noise in volume renderings of cardiac ultrasound volumes.
@@
 \section{Discussion and future work}
-Even with all our effort in \textbf{Paper I} we were only able to achieve a fraction of the maximum throughput of the target GPU.
+Even with all our effort in \textbf{Paper\,I} we were only able to achieve a fraction of the maximum throughput of the target GPU.
 This result shows how hard it is to reach the theoretical level of throughput, and that how close one can get is highly algorithm depended.
@@
 This will restrict even the most fine tuned implementation from reaching the maximum throughput of the GPU.
-An interesting observation can be found in Fig.\,10 of \textbf{Paper I}.
+An interesting observation can be found in Fig.\,10 of \textbf{Paper\,I}.
 Here we see how the achieved throughput of the covariance estimation kernel is more than three times higher than the theoretical throughput of a modern, high-end, \nom{CPU}{Central processing unit}.
@@
 \\\\
-In \textbf{Paper II} we focused on achieving real time Capon processing for cardiac ultrasound by implementing the BS-Capon beamformer on a GPU.
+In \textbf{Paper\,II} we focused on achieving real time Capon processing for cardiac ultrasound by implementing the BS-Capon beamformer on a GPU.
 However, we also present the first medical ultrasound loops processed with ES- and BS-Capon beamforming.
@@
 The beamformers are clearly able to decrees the lateral width of lateral point scatterers and thin structures \textit{in vivo}, and to sharpen step edges.
-On the other hand, the overall contrast is visually not improved and sometimes it is worse (Fig.\,6b of \textbf{Paper II}).
+On the other hand, the overall contrast is visually not improved and sometimes it is worse (Fig.\,6b of \textbf{Paper\,II}).
 Since the Capon beamformer only affect directional noise, it is clear that contrast will get worse in low \nom{SNR}{Signal-to-noise ratio} situations when signal cancellation occurs.
-The oversampling approach presented in \textbf{Paper III} might improve on this, but it will only reduce signal cancellation if the phase shifts across the aperture are close to linear and small.
+The oversampling approach presented in \textbf{Paper\,III} might improve on this, but it will only reduce signal cancellation if the phase shifts across the aperture are close to linear and small.
 Another issue is that bright points with a wide lateral profile exhibits better visual contrast when surrounded by speckle noise than points with narrow lateral profiles.
 In future work a detailed investigation of why some results, especially the contrast, obtained in simulations are not transferable to \textit{in vivo} needs to be performed.
-It will also be crucial to show larger improvements \textit{in vivio} than what is presented in \textbf{Paper II}.
+It will also be crucial to show larger improvements \textit{in vivio} than what is presented in \textbf{Paper\,II}.
 Both to justify all our computations, and to show an image which has both highly improved resolution and contrast.
-As pointed out at the end of \textbf{Paper II}, cardiac ultrasound imaging might not be the best modality for Capon beamforming.
-Cardiac ultrasound imaging was selected because it is the medical ultrasound modality which requires most calculations per second.
+As pointed out at the end of \textbf{Paper\,II}, cardiac ultrasound imaging might not be the best modality for Capon beamforming.
+Cardiac ultrasound imaging was mainly selected because it is the medical ultrasound modality which requires most calculations per second.
 An application where the focus is detection of closely spaced point targets, might be better suited.
@@
 Finally, it would also be interesting to see how the Eigenspace-based Capon beamformer performs on cardiac images, and how its absence of a distortionless criterion will influence the image spatially and temporally.
+The work in \textbf{Paper\,II} made it clear that interesting findings are done when users are able to watch the temporal behavior of an algorithm.
+This led us to the work of \textbf{Paper\,III}.
 \\\\
-In \textbf{Paper3} we investigate the inherent self nulling involved with Capon beamforming, and how this effect comes naturally into play when the beamformer is subjected to linear steering vector errors caused by lateral undersampling.
+In \textbf{Paper\,III} we investigate the inherent self nulling involved with Capon beamforming, and how this effect comes naturally into play when the beamformer is subjected to linear steering vector errors caused by lateral undersampling.
 These errors are naturally reduced when the lateral sampling is increased.
@@
 A lack of studies where Capon beamforming is performed on consecutive frames might also explain the lack of comments.
-Fortunately, as shown in \textbf{Paper III}, it is possible to maintain the same frame rate as with DAS beamforming by applying oversampling on receive using a set of steering vectors (phase delays).
+Fortunately, as shown in \textbf{Paper\,III}, it is possible to maintain the same frame rate as with DAS beamforming by applying oversampling on receive using a set of steering vectors (phase delays).
 It also turns out that more matrix inversions are not needed either.
@@
 The Cartesian grid should also have high resolution in order to preserve the un-sharpening introduced by Capon beamforming.
-As just mentioned in the discussion of \textbf{Paper II}, reduced signal cancellation should prevent contrast from degrade to less than the contrast obtain with DAS.
+As just mentioned in the discussion of \textbf{Paper\,II}, reduced signal cancellation should prevent contrast from degrade to less than the contrast obtain with DAS.
 It will, however,  not correct for signal cancellation caused by phase aberrations.
@@
 \\\\
-%Objections to \textbf{Paper IV} includes its inherent chicken-or-egg dilemma.
+%Objections to \textbf{Paper\,IV} includes its inherent chicken-or-egg dilemma.
 The visualization is adjusted based on the data acquired by a user.
@@
 %Does data need to be saved in order to perform tracking?
-One limitation of the algorithm described in \textbf{Paper IV} is its dependence on a successful tracking.
+One limitation of the algorithm described in \textbf{Paper\,IV} is its dependence on a successful tracking.
 If the tracking fails the proposed method will fail.
@@
 The paper does also propose an interesting method for automatic tuning of rendering parameters.
-Figure 3d in \textbf{Paper IV} actually shows that an optimal threshold could be found based on global and not only local statistics.
+Figure 3d in \textbf{Paper\,IV} actually shows that an optimal threshold could be found based on global and not only local statistics.
 Even though the visualization based on global statistics has more errors, with respect to endocardium visibility, the tuning of the opacity function is still fully automatic if an iterative search for the minima in Figure 3d is performed.
 A global opacity transfer function will assure a rendering which are realistic, but it will still be dependent on a successful tracking for the auto-adjustment to work.
-An automatically tuned opacity function based on global statistics combined with an improved smoothing scheme could be the future for cardiac ultrasound imaging.
+An automatically tuned opacity function based on global statistics combined with an improved smoothing scheme is likely to be better suited for cardiac ultrasound imaging than a local opacity transfer function.
+A local opacity transfer function is just to extreme to be clinically robust.
 It would also be interesting to see if statistics based on the window selected by the Capon or the LCA beamformer could be utilized to adjust the opacity.
@@
 In that way we can apply less filtering and higher opacity on edges.
-If this will be different than just looking at neighboring pixels is to be seen.
+If this will be different than just looking at neighboring pixels in a smoothing filter is to be seen.
 \\\\
@@
 It is still, however, centered around the topic of implementing algorithms for ultrasound imaging on GPUs.
-Ultrasound simulations are typically used for research and sometimes also to pre-calculate configurations of a given scan sequence in order to achieve a certain pressure in a given area.
+Ultrasound simulations are typically used for research and sometimes also to pre-calculate configurations of a given scan sequence, for instance to achieve a certain pressure in a given area.
 The continuously growth in computational power will, in the future, make it possible to evaluate a scan sequence on the fly, removing the need for lookup tables.
@@
 Achieving higher processing speed for an excising algorithm by implementing it on a new architecture is not academically relevant in its own.
-It has to combined with either a good analysis of what special adjustments that had to be made to the algorithm in order for the increase to happen or a suggestion of how to utilize the speed gained.
-These are points we have tried to answer in both \textbf{Paper I}, \textbf{II},  \textbf{IV}, and \textbf{V}.
-In \textbf{Paper V} the increased speed makes an interactive paint program feasible.
+It has to be combined with either a good analysis of what special adjustments that had to be made to the algorithm in order for the increase to happen or a suggestion of how to utilize the speed gained.
+These are points we have tried to answer in both \textbf{Paper\,I}, \textbf{II},  \textbf{IV}, and \textbf{V}.
+In \textbf{Paper\,V} the increased speed made an interactive ''Paint-program'' feasible.
 As of January 2014 the simulator has been download 200 times and counting.
-It also has a "whopping" 1000 views on youtube.
+It also has close to a "whopping" 1000 views on youtube.
 \\\\

### 2014-02-14 ThesisJP/Background/background.tex
writing and added figure.

@@
 The last three generations of CPUs from Intel have all had on-chip integrated graphics.
-Even if an on-chip GPU is less powerful than a discrete GPU, valuable time is saved by not having to send data across the PCI-express bus, which connects the different parts of a computer to the motherboard.
-This is the reason why the first rules of GPGPU is to minimize PCI-express traffic.
+Even if an on-chip GPU is less powerful than a discrete GPU, valuable time is saved by not having to send data across the PCI Express bus, which connects the different parts of a computer to the motherboard.
+This is the reason why the first rules of GPGPU (with discrete GPUs) is to minimize CPU-to-GPU transfers.
 
@@
 Together, multi-core, heterogeneous cores, and cloud cores forms the "jungle" that today's software engineers have to tackle (Fig.\,\ref{fig:jungle}).
-This thesis focus on heterogeneous computing with one multi-core CPU and a discrete high performance GPU, a cheap system which is likely to be the specification of current and future ultrasound imaging system.
+This thesis focus on heterogeneous computing with one multi-core CPU and a discrete high performance GPU, a cheap and high performance system which is likely to be the specification of current and future ultrasound imaging system.
 
 \subsection{Comparing CPU and GPU performance}\label{sec:cpu_vs_gpu}
+The CPU is design to be the main control unit of a computer.
+It is therefore good at general processing.
+The GPU, on the other hand, has a history of processing graphics which involves heavy vector arithmetic.
+Even though they started out as very different processors, modern CPUs are including more and more GPU-like features, and GPUs are adding CPU features.
+In this section we will see why there is a big difference in peak performance for a CPU and GPU.
+
 It is common to say that a CPU has few but advance generic cores, and the GPU has many but simple cores.
@@
 
- the most powerful one (HD 5200) have a peak performance of nearly one tera single precision floating point operations per second (FLOPs).
+ the most powerful integrated intel GPU (HD 5200) have a peak performance of nearly one tera single precision floating point operations per second (FLOPs).
 
@@
 \label{fig:cpu_vs_gpu}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics[width=\textwidth]{img/CPU-GPU.png}
+\caption{Overview of CPU and GPU architecture.}
+\label{fig:cpu_gpu}
 \end{figure}
@@
 \section {Medical ultrasound imaging}\label{sec:ultrasound}
-Ultrasound imaging encompasses technology which generates images based on sound whose frequencies we can not hear. For medical ultrasound imaging frequencies above 2 MHz are typically used.
+Ultrasound imaging encompasses technology which generates images based on sound whose frequencies we can not hear.
+For medical ultrasound imaging frequencies between 2 MHz and 10 MHz are typically used.
+Frequency dependent attenuation force us to select a lower frequency if the object we want to image is located deep into the body.
+This is the reason why 2-4 MHz is used for cardiac imaging and a frequency around 7 MHz is used for vascular imaging.
 
@@
 }
-\caption{Array beam pattern with uniform and Capon weights. An interfering source is located at 20 degrees.}
+\caption{Array beam pattern with uniform and Capon weights.
+An interfering source is located at 20\degree.
+Notice the reduced sidelobe level for the Capon beam pattern in the direction of the interferer.}
 \end{figure}

### 2014-02-17 HRUIofMovingObjects/response_to_reviewers.tex
Started answering comments from reviewers.

\documentclass{article}

\begin{document}

Answers to the reviewers in \textit{italic}.

\section*{Reviewer: 1}

As my scores indicate, I think the paper addresses an interesting problem and does it very well. Beside two VERY minor typos, I think it can be published unaltered.
1. "In" missing at the beginning of the Introduction section.
2. In section IIB, the opening sentence, there is an 's' missing in 'produce'.

\textit{We want to thank the reviewer for an encouraging review. Typos have been corrected.}

\section*{Reviewer: 2}

Thanks to the authors for addressing the problem of the temporal aspects in minimum variance beamforming, demonstrating the extra requirements on spatial sampling, and demonstrating how to efficiently oversample. The paper is good but I would like it to go through a second review round based on the following comments:
\\\\
Page2 column2 lines 34-35, and Fig. 1, L, M and d have not been introduced (need to have read the previous literature or equation (4) to know their meaning). I would suggest that in the introduction we do not need these details.
\\\\
Equation (8) the two factors p and q are confusing. Because q is always one, I would suggest omitting it.  Let us say that the oversampling is used both for Capon and for reducing folding of frequencies over the bandwidth.
\\\\
At what depth are the LSV plots taken with respect to the focal depth? Does it matter?
\\\\
Please specify the characteristics of the M5S probe (size, number of elements) if GE permits. Is it a phased or a linear array? What is the physical beam spacing, the field of view, the number of transmit and receive beams in the base (non oversampled) configuration? What is the transmit focal depth? Are transmit and receive apodization applied? Without this information, the work is not reproductible.
\\\\
In section “IV-A. Oversampling in transmit”: please confirm (specify) that 1x1 receive beamforming is performed at this point.
\\\\
In section “IV-B. Oversampling by parallel receive beamforming”, could you explain in more details what are the differences between Capon + STB and the technique proposed by Rabinovitch (2013) – reference [27]?
\\\\
Page10 Column1 line56  write “should be removed” instead of “should be remove”
\\\\
Page 10 Column 1 lines 40-60, you state that with synthetic transmit beamforming, some of the artifacts seen in oversampling in receive by phase rotation could be eliminated. It would be interesting to see such technique implemented.
\\\\
P10 column 2 line 28-29 write “too extreme” instead of “to extreme”
\\\\
Fig 6a: specify which beamforming is applied here (I suspect delay and sum with no multiline receive?). Maybe you want to consider showing still images of Capon and Capon-oversampled as well (not necessarily on fig. 6), just in case some people won’t look at the videos.

\end{document}

### 2014-02-17 HRUIofMovingObjects/jrnl_article.tex
Started answering comments from reviewers.

@@
 %
-\documentclass[journal]{IEEEtran}
+\documentclass[draftcls]{IEEEtran}
+%\documentclass[journal]{IEEEtran}
 %
@@
 
-\newcommand\multimedia[1]{\textbf{{\color{red}#1}}}
+\newcommand\multimedia[1]{\textbf{{\color{red}[#1]}}}
 \newcommand\comment[1]{\textit{{\color{red}(#1)}}}
@@
 % can use linebreaks \\ within to get better formatting as desired
-\title{Capon Beamforming Applied on Moving Objects - An Analysis of Local Lateral Shift-Invariance}
+\title{Capon Beamforming and Moving Objects - An Analysis of Lateral Shift-Invariance}
 %
@@
 In this paper we analyse the local lateral shift-invariance of the Capon beamformer when imaging moving objects.
-We then discuss and investigate different methods in order to improve on it.
-We show that insufficient lateral sampling makes an imaging system based on the Capon beamformer locally laterally shift-variant.
-We then show that the local lateral shift-invariance can be improved by oversampling based on phase rotation without affecting the acquisition frame rate and with minor changes in processing complexity.
+We show that insufficient lateral sampling makes an imaging system based on the Capon beamformer laterally shift-variant.
+Different methods for oversampling on transmit and receive are then discussed and investigated in order to improve on it.
+It is shown that lateral shift-invariance can be improved by oversampling based on phase rotation on receive.
+This without affecting the acquisition frame rate and with a minor change in processing complexity.
 \end{abstract}
@@
 We see that a DAS beamformer, using the required beam spacing, could yield a $1.9$ dB scalloping loss \cite{Harris1978} if the object is moving parallel to the array.
-However if the Capon beamformer is used with the same beam density the loss could be as large as $27$ dB.
+However, if the Capon beamformer is used with the same beam density the loss could be as large as $27$ dB.
 This imaging system is then highly sensitive to small spatial shifts, either by the object or the array itself.
@@
 In previous literature on Capon beamforming for medical ultrasound imaging only single frame scenarios have been simulated.
-The scalloping loss effect has therefore been passed over in silence by either oversampling on transmit, by carefully positioning point scatterers exactly on transmit-receive beam pairs, or by ignorance.
+The scalloping loss effect has therefore been passed over in silence by either oversampling on transmit, by carefully positioning point scatterers exactly on transmit-receive beams, or by ignorance.
 In our first work where Capon beamforming was applied on a loop of images containing moving objects \cite{Asen2012, Asen} we commented on the effect, but no solution was presented other than oversampling on transmit when simulating.
@@
 In this paper we investigate the issue further.
-We show that when a moving point scatterer is imaged with the Capon beamformer with the same beam spacing sufficient for DAS, the predicted scalloping loss is observed both in simulations and \textit{in-vitro} data.
+We show that when a moving point scatterer is imaged with the Capon beamformer with the same beam spacing which is sufficient for DAS, the predicted scalloping loss is observed both in simulations and \textit{in-vitro} data.
 The question is then, what can we do in order to improve on the situation without affecting the acquisition frame rate and processing throughout.
@@
 
-The next section gives an introduction to Capon beamforming for medical ultrasound imaging together with a short introduction to lateral sampling and shift-invariance.
+The next section gives an introduction to Capon beamforming together with a short introduction to lateral sampling and shift-invariance.
 In Section \ref{sec:capon_LLSI} we study the scalloping loss yield by the Capon beamformer and its local lateral shift-invariance property.
-Different methods for reduced scalloping loss and by that improved local lateral shift-invariance are then presented in Section \ref{sec:methods}.
+Different methods for reduced scalloping loss and by that improved local lateral shift-invariance are presented in Section \ref{sec:methods}.
 Then, in Section \ref{sec:res} we present the results of applying the most promising method on \textit{in-vitro} data.
@@
 \subsection{Capon Beamforming}
-The Capon beamformer produce a weight vector, like the one in (\ref{eq:das}), based on the impinging signal and an optimization criterion. The problem is as follow \cite{Capon1969}:
+The Capon beamformer produces a weight vector, like the one in (\ref{eq:das}), based on the impinging signal and an optimization criterion. The problem is as follow \cite{Capon1969}:
 \begin{align}
@@
 where $\mat{\hat{R}}$ is a sample covariance matrix.
-Hence, the resulting weight vector will minimizes the output power while maintaining unit gain in the steering direction $\vec{a}$.
+Hence, the resulting weight vector minimizes the output power while maintaining unit gain in the steering direction $\vec{a}$.
 %The covariance matrix is usually unknown and has to be estimated from the element data.
@@
 
-The lateral sampling required in an ultrasound image for a given center frequency, $f_c$, and bandwidth, $B$, is given by the aperture size, $D$, and the wavelength of the center frequency, $\lambda_c = c/f_c$, where $c$ is the speed of sound.
+The required lateral sampling in an ultrasound image for a given center frequency, $f_c$, and bandwidth, $B$, is given by the aperture size, $D$, and the wavelength of the center frequency, $\lambda_c = c/f_c$, where $c$ is the speed of sound.
 The required Nyquist beam spacing is then \cite{Hergum2007}:
@@
 The geometric distortion is now increased to a level where the contours are not diagonal.
-The maximum scalloping loss is also larger than 1 dB.
+The maximum scalloping loss is almost 4 dB.
 Hence, the imaging system is no longer locally laterally shift-invariant.
@@
 \caption{LSV-plots of delay-and-sum beamforming.
-If the contours are diagonal and the amplitude variation among the peak points (black dots) is small, the imaging system is said to be locally laterally shift-invariant.
-Dotted vertical lines are plotted where transmit-receive beam pairs are located.
+If contours are diagonal and the amplitude variation among the peak points (black dots) is small, the imaging system is said to be locally laterally shift-invariant.
+Dotted vertical lines are plotted where transmit-receive beams are located.
 a) Sampling density according to (\ref{eq:resolution}).
 The system is locally laterally shift-invariant.
-b) Undersampling with a factor of two.
+b) Undersampling by a factor of two.
 The system is no longer locally laterally shift-invariant.}
@@
 The diagonal factor $d$ and the effective adaptive aperture size $L$ can be used to control the mainlobe width of the Capon beamformer.
-In both extreme cases, $d\gg1$ and $L=1$, the Capon beamformer becomes either a triangle or uniform-weighted DAS beamformer respectively, and a beam spacing according to (\ref{eq:resolution}) would then be sufficient.
+In the extreme cases, $d\gg1$ and $L=1$, the Capon beamformer becomes either a triangle or uniform-weighted DAS beamformer respectively, and a beam spacing according to (\ref{eq:resolution}) would then be sufficient.
 In Fig.\,\ref{fig:capon2} we show how the scalloping loss is reduced when the amount of diagonal loading is increased.
@@
 We observe that Fig.\,\ref{fig:capon2} now contains a scaled version of the pattern presented in Fig.\,\ref{fig:das_shiftvariant}.
-From Fig.\,\ref{fig:capon} it should be obvious that the Capon beamformer either requires increased lateral sampling in order to make the imaging system locally laterally shift-invariant, or the resolution has to be reduced to the same level as DAS.
+From Fig.\,\ref{fig:capon} it should be clear that Capon beamforming either requires increased lateral sampling, or the resolution has to be reduced to the same level as DAS in order to make the imaging system locally laterally shift-invariant.
 It would be unwise to do all the computation involved with Capon beamforming with a high $d$ value and end up with a resolution equal to DAS.
-However, we see that the diagonal factor can be a useful tool if resolution has to be traded for the need of oversampling.
+However, we see that diagonal loading can be a useful tool if resolution has to be traded for the need of oversampling.
 The same is true for the subarray length parameter.
@@
 a) Capon beamforming with $d=1/100$ as diagonal loading.
-The system is not locally lateral shift-invariant.
+The system is not locally laterally shift-invariant.
 b) The diagonal loading factor $d$ is increased to 1.
-The scalloping loss is reduced, but the system is still not locally lateral shift-invariant.}
+The scalloping loss is reduced, but the system is still not locally laterally shift-invariant.}
 \label{fig:capon}
@@
 \section{Oversampling Methods}\label{sec:methods}
-In this section we will discuss different methods aimed at improving the local lateral shift-invariance for the Capon beamformer.
+In this section we will discuss different methods aimed at improving the local lateral shift-invariance of the Capon beamformer.
 From the previous section it should be clear that we somehow need to increase the lateral sampling to make the Capon imaging system locally laterally shift-invariant.
@@
 A straight forward way of obtaining sufficient sampling in an active imaging system is by increasing the number of transmit beams (tx-beams) until the system is locally laterally shift-invariant.
-In Fig.\,\ref{fig:capon_oversampling_a} and Fig.\,\ref{fig:capon_oversampling_b} we have increased the lateral sampling on transmit by a $q$-factor of four and sixteen respectively.
+In Fig.\,\ref{fig:capon_oversampling_a} and Fig.\,\ref{fig:capon_oversampling_b} we have increased the tx-beam density by a $q$-factor of four and sixteen respectively.
 As expected the geometric distortion and scalloping loss decreases with increased lateral sampling.
@@
 
-From Fig.\,\ref{fig:capon_oversampling_b} we see that sixteen times oversampling reduces the scalloping loss significantly, but the maximum is still larger than 1 dB.
-A even higher level of oversampling is therefore required in order to have a completely local lateral shift-invariant Capon beamformer when a moving point scatterer is simulated without noise and with the same parameters as in Fig\,\ref{fig:capon}.
+From Fig.\,\ref{fig:capon_oversampling_b} we see that sixteen times oversampling reduces the scalloping loss significantly.
+Even though high frequency variations are reduced, the maximum is still larger than 1 dB.
+An even higher level of oversampling is therefore required in order to have a completely local lateral shift-invariant Capon beamformer when a moving point scatterer is simulated without noise and with the same parameters as in Fig\,\ref{fig:capon}.
 %$1/100$ in diagonal loading.
@@
 b) Sixteen times oversampling.
-Even though its reduced, the maximum scalloping loss is still too high for the system to be named locally laterally shift invariant.}
+The maximum scalloping loss is reduced, but it is still too high for the system to be locally laterally shift invariant.}
 \label{fig:capon_oversampling}
@@
 Parallel receive beamforming (PRB) means that multiple narrow receive beams (rx-beams) are constructed from one broad tx-beam.
-Then, by reducing the number of tx-beams and relying on multiple rx-beams in order to get sufficient sampling on receive, it is possible to achieve high frame rates in e.g.
-3D cardiac ultrasound imaging.
+Then, by reducing the number of tx-beams and relying on multiple rx-beams in order to get sufficient sampling on receive, it is possible to achieve high frame rate imaging.
 However, PRB also introduces geometrical distortion which to some degree can be compensated for by using e.g.
-synthetic transmit beamforming (STB) \cite{Hergum2007, Denarie2011, Denarie2013}.
+synthetic transmit beamforming (STB) \cite{Hergum2007, Denarie2013}.
 In \cite{Rabinovich2013} it is shown that STB works well with Capon beamforming.
@@
 The acquisition frame rate will then not be affected.
-On the other hand more rx-beams will increase the number of covariance matrices, and thereby the number of matrix-inversions with a factor equal to the number of parallel receive lines.
+On the other hand, more rx-beams will increase the number of covariance matrices and thereby the number of matrix-inversions with a factor equal to the number of parallel receive lines.
 The processing times reported in \cite{Asen} will then also increase with the same factor.
@@
 However, the steering vector can also, as in narrow band applications, be varied over a set of pre-defined vectors.
-In a narrow band setting the beam can make a sweep across the whole imaging sector by phase steering.
-This is also possible in broadband applications as long as the phase rotation is less than one pulse length (See coarse-fine beamforming in \cite{Thomenius}).
-The maximal steering angle is therefore given by
+In a narrow band setting the beam can make a sweep across the whole imaging sector through phase steering.
+This is also possible in broadband applications as long as the phase rotation is less than one pulse length (See coarse-fine beamforming in \cite{Thomenius} and its references).
+The maximal steering angle is approximately given by
 \begin{align}
@@
 For our simulations $\theta_{max} = 3\Delta$.
-It is therefore possible to make a sweep using phase rotation between a given beam and half-way to its two nearest neighbours without introducing to large errors.
-Phase rotation is actually what the Capon beamformer use internally to micro-steer away from bright point scatterers and to enhance edges.
+It is therefore possible to make a sweep based on phase rotation between a given beam and half-way to its two nearest neighbours without introducing large errors.
+Phase rotation is actually what the Capon beamformer uses internally to micro-steer away from bright point scatterers and to enhance edges.
 
@@
 
+Figure \ref{fig:capon_oversampling_pr} shows the phase rotation method applied on the same scan sequence as in Fig\,\ref{fig:capon} with four and sixteen times oversampling respectively.
+Note that the figure is zoomed compared with Fig\,\ref{fig:capon}.
+As with oversampling on transmit there is some variation when the point moves in and out of focus.
+The maximum scalloping loss is therefore large.
+The extreme local variations are however reduced significantly.
+Comparing Fig\,\ref{fig:capon_oversampling} with Fig\,\ref{fig:capon_oversampling_pr}, the errors introduced by the phase rotation method are best seen in sidelobes.
+
 Since the phase rotation method operates on the same channel data as in Fig.\,\ref{fig:capon1}, there will be no reduction in the acquisition frame rate by applying this method.
@@
 b) Sixteen times oversampling.}
-	\label{fig:benchmark_capon_bs}
+	\label{fig:capon_oversampling_pr}
 \end{figure*}
@@
 
-In Fig\,\ref{fig:image_das} one point scatterer has been selected (white box), and then used to generate LSV-plots.
+One point scatterer in Fig\,\ref{fig:image_das} has been selected [white box], and then used to generate LSV-plots.
 Figure \ref{fig:phantom_das} shows the resulting LSV-plot when using the DAS beamformer.
-The scalloping loss is less than 1 dB, and the system is therefore locally laterally shift-invariant.
-As a consequence point scatterers have a consistent appearance when subjected to motion \multimedia{Video DAS}.
+The scalloping loss is less than 1 dB, and the system is by that locally laterally shift-invariant.
+As a consequence point scatterers have a consistent appearance when subjected to motion \multimedia{Media Movie 1}.
 
@@
 In almost all the other frames the signal loss is massive (at most 10 dB).
-The result is blinking point scatterers when subjected to motion \multimedia{Video Capon}.
+The result is blinking point scatterers when subjected to motion \multimedia{Media Movie 2}.
 
@@
 In addition, due to the increase in lateral resolution, the Capon beamformer requires smaller frame-to-frame movements than DAS.
-Looking at the video of Capon beamforming with sixteen times oversampling by phase rotation \multimedia{Video Capon + 16PR} the blinking is now highly reduced.
+Looking at the video of Capon beamforming with sixteen times oversampling by phase rotation \multimedia{Media Movie 3} the blinking is now highly reduced.
 %The system is however not fully shift invariant.
@@
 \begin{tabular}{cc}
-\subfloat[]{\includegraphics[width=0.5\linewidth]{\img fig7a.eps}\label{fig:image_das}}
-\subfloat[]{\includegraphics[width=0.5\linewidth]{\img fig7b.eps}\label{fig:phantom_das}}\\
-\subfloat[]{\includegraphics[width=0.5\linewidth]{\img fig7c.eps}\label{fig:phantom_capon_LSV}}
-\subfloat[]{\includegraphics[width=0.5\linewidth]{\img fig7d.eps}\label{fig:phantom_capon_LSV_10PR}}
+\subfloat[]{\includegraphics[width=0.5\linewidth]{\img fig6a.eps}\label{fig:image_das}}
+\subfloat[]{\includegraphics[width=0.5\linewidth]{\img fig6b.eps}\label{fig:phantom_das}}\\
+\subfloat[]{\includegraphics[width=0.5\linewidth]{\img fig6c.eps}\label{fig:phantom_capon_LSV}}
+\subfloat[]{\includegraphics[width=0.5\linewidth]{\img fig6d.eps}\label{fig:phantom_capon_LSV_10PR}}
 \end{tabular}
@@
 \section{Discussion}\label{sec:dis}
-In this paper we have investigated the lateral shift-invariance property of the Capon beamformer when imaging moving objects.
-We have shown that the Capon beamformer is less sensitive to small lateral shifts if oversampling based on phase rotation is applied.
-This without reducing the acquisition frame rate and with only a minor increase in processing complexity.
-%In Section \ref{sec:capon_LLSI} it was shown that a Capon beamformer with high resolution parameters is not locally laterally shift-invariant when subjected to motion.
-The root cause was found to be too coarse lateral sampling.
-%In Section \ref{sec:methods} we presented methods for improved shift-invariance based on oversampling.
-The most promising method, with respect to improved shift-invariance, reduction in acquisition frame rate and processing complexity combined, was based on phase rotation.
-%In Section \ref{sec:res} we presented results showing that this method also works on \textit{in-vitro} data when the ultrasound probe is subjected to motion. 
-In simulations the method reduced the maximal scalloping loss by 16 dB.
-On \textit{in-vitro} data the reduction was more than 7 dB, and the imaging system is now almost locally laterally shift invariant.
-
-Note that oversampling is not only essential to keep the amplitude level of moving point scatterers.
-It should also help to improve contrast and edge definitions.
-If oversampling is not applied homogeneous speckle regions could get a too low average value and the edge delineation of a moving object will fluctuate in the same way as a moving point scatterer would have sparkled.
-
 %Note that in this paper we have investigated a single array configuration and a limited parameter range for the Capon beamformer.
@@
 In his seminal paper, Cox \cite{Cox1973} derives the required beam density for a given output SNR.
-For instance 47 dB requires a beam density for the Capon beamformer ten times the one given by (\ref{eq:resolution}).
+For instance 47 dB requires a beam density for the Capon beamformer ten times the one given by (\ref{eq:resolution}) [with $qp=1$].
 Since resolution for the Capon beamformer depends on SNR we would recommend to first use the dynamic range in the image as guidance.
@@
 Relating the presented phase rotation method to earlier work we find that it is similar to a steering vector uncertainty set.
-However, where these set-based methods typically has a criterion for selecting the best output after applying all the steering vectors contained in the set, oversampling with phase rotation outputs all.
+However, where these set-based methods typically has a criterion for selecting the best output after applying all the steering vectors contained in the set, oversampling by phase rotation outputs all.
 It can also be thought of as a narrow band approach to parallel receive beamforming where Capon's method is applied on all receive beams.
@@
 eight phase rotated beams are averaged into four.
-The geometric distortion, or stripes, seen at the bottom of the sector in the video with sixteen times oversampling \multimedia{Capon+16PR} should be possible to remove using this technique.
+The geometric distortion, or stripes, seen at the bottom of the sector in the video with sixteen times oversampling \multimedia{Media Movie 3} should be remove by such technique.
 
@@
 
+Note that oversampling is not only essential to keep the amplitude level of moving point scatterers.
+It should also help to improve contrast and edge definitions.
+If oversampling is not applied homogeneous speckle regions could get a too low average value and the edge delineation of a moving object will fluctuate in the same way as a moving point scatterer would have sparkled.
+
 Finally we want to stress the fact that this demand for oversampling does not only affect the Capon [or minimum variance beamformer].
@@
 \section{Conclusion}\label{sec:con}
-In this paper we have investigated the lateral shift-invariance properties of the Capon beamformer when imaging moving objects.
-It was shown that Capon beamforming with the same beam density as required for delay-and-sum beforming results in blinking point scatterers when subjected to motion.
-We then discussed and investigated different methods in order to reduced this effect.
-Finally we showed that the local lateral shift-invariance could be improved by oversampling based on phase rotation without affecting the acquisition frame rate and with minor changes in processing complexity.
+%In this paper we have investigated the lateral shift-invariance property of the Capon beamformer when imaging moving objects.
+We have shown that the Capon beamformer is less sensitive to small lateral shifts if oversampling based on phase rotation is applied.
+This without reducing the acquisition frame rate and with only a minor increase in processing complexity.
+%In Section \ref{sec:capon_LLSI} it was shown that a Capon beamformer with high resolution parameters is not locally laterally shift-invariant when subjected to motion.
+The root cause was found to be too coarse lateral sampling.
+%In Section \ref{sec:methods} we presented methods for improved shift-invariance based on oversampling.
+The most promising method, with respect to improved shift-invariance, reduction in acquisition frame rate and processing complexity combined, was based on phase rotation.
+%In Section \ref{sec:res} we presented results showing that this method also works on \textit{in-vitro} data when the ultrasound probe is subjected to motion.
+
+In this paper we have investigated the lateral shift-invariance property of the Capon beamformer when imaging moving objects.
+It was shown that Capon beamforming with the same beam density as required for delay-and-sum beforming results in blinking point scatterers when the system is subjected to motion.
+We then discussed and investigated different methods based on oversampling on transmit and receive in order to reduced this effect.
+Finally we showed that the local lateral shift-invariance could be improved by oversampling based on phase rotation on receive.
+This without affecting the acquisition frame rate and with a minor change in processing complexity.
+In simulations, this method reduced the maximum scalloping loss by 16 dB.
+On \textit{in-vitro} data the reduction was around 7 dB and the point scatterer blinking was highly reduced.
 
@@
 
-\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{photos/jon_petter2.eps}}]{Jon Petter \AA{}sen}
-(S12) was born in Porsgrunn, Norway in 1986.
+%\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{photos/jon_petter2.eps}}]{Jon Petter \AA{}sen}
+%(S12) was born in Porsgrunn, Norway in 1986.
 He received the B.Sc.
@@
 
-He has industry experience from GE Vingmed Ultrasound, Horten, Norway, were he completed three summer internships from 2008 to 2010.
+%He has industry experience from GE Vingmed Ultrasound, Horten, Norway, were he completed three summer internships from 2008 to 2010.
 His research interests include image and signal processing, adaptive ultrasound processing techniques and acceleration of ultrasound algorithms using Graphics Processing Units (GPUs).
-\end{IEEEbiography}
-
-\vfill{}
-
-\newpage
-
-%\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{photos/jo_inge2.eps}}]{Jo Inge Buskenes}
-%received the B.Sc.
-degree in electrical engineering from Gj\o{}vik College University, Norway, in 2007, and the M.Sc.
-degree in instrumentation for particle physics from the University of Oslo, Norway, in 2010.
-He is currently pursuing the Ph.D.
-degree in image reconstruction and technology at the University of Oslo.
-
-%His industry experience includes the European Organization for Nuclear Research (CERN), Geneva, Switzerland (2007-2008), and the Norwegian Defence Research Establishment, Kjeller, Norway (2009).
-He has lectured in digital signal processing at the Gj\o{}vik College University (2009), and at the University of Oslo (2010-2012).
-His research interests include adaptive beamforming, digital image reconstruction, high performance computing, intelligent detector design and embedded Linux platforms.
 %\end{IEEEbiography}
 
-%\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{photos/carl-inge.eps}}]{Carl-Inge Colombo Nilsen}
-%(S’06–M’10) received the M.Sc.
-and Ph.D.
-degrees in computer science from the University of Oslo, Norway, in 2005 and 2010.
-He is currently working at the University of Oslo as a postdoctoral research fellow.
-His research interests include signal and array processing for ultrasound imaging and other acoustic applications.
-%\end{IEEEbiography}
-
-% insert where needed to balance the two columns on the last page with
-% biographies
+%\vfill{}
+
 %\newpage
 
-%\vspace{-4cm}
-
-\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{photos/andreas.eps}}]{Andreas Austeng}
-was born in Oslo, Norway, in 1970.
+%\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{photos/andreas.eps}}]{Andreas Austeng}
+%was born in Oslo, Norway, in 1970.
 He received the M.S.
@@
 His research interests include signal and array processing for acoustical imaging.
-\end{IEEEbiography}
-
-%\vspace{-4cm}
-%\newpage
-
-\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{photos/sverre.eps}}]{Sverre Holm}
-(M’82-SM’02) was born in Oslo, Norway, in 1954.
+%\end{IEEEbiography}
+
+%\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{photos/sverre.eps}}]{Sverre Holm}
+%(M’82-SM’02) was born in Oslo, Norway, in 1954.
 He received the M.S.
@@
 He worked with Sonitor Technologies (2000-2005), where he developed ultrasonic technology that is the core of the indoor positioning system described in \textit{Scientific Americanin} 2008 as "a positioning system that goes where GPS can't." He has published about 130 publications internationally, plus about 10 patents.
-He has had sabbaticals at GE Global Research, NY (1998), and the Waves and
-Acoustics Laboratory of ESPCI, Paris (2008-2009).
+He has had sabbaticals at GE Global Research, NY (1998), and the Waves and Acoustics Laboratory of ESPCI, Paris (2008-2009).
 He was an associate editor of the \textit{IEEE Transactions on Ultrasonics, Ferroelectrics, and Frequency Control} from 1997 to 2002.
 His research interests are medical ultrasound imaging, modeling of tissue-ultrasound interaction, and ultrasound positioning.
-\end{IEEEbiography}
-
-\vfill
+%\end{IEEEbiography}
+
+%\vfill
+
 % You can push biographies down or up by placing

### 2014-02-18 HRUIofMovingObjects/response_to_reviewers.tex
Response to reviewers.

@@
 The paper is good but I would like it to go through a second review round based on the following comments:
+\textit{We want to thank the reviewer for an encouraging review. Below follows our response to his/her comments.}
 \\\\
@@
 I would suggest that in the introduction we do not need these details.
+\textit{(Should we remove this?)}
 \\\\
@@
 Let us say that the oversampling is used both for Capon and for reducing folding of frequencies over the bandwidth.
+\textit{We do not find this confusing, $p$ is always 2 and $q$ is 4 and 16 when oversampling on transmit is applied for the Capon beamformer.}
 \\\\
 At what depth are the LSV plots taken with respect to the focal depth? Does it matter?
+\textit{The shift-invariance caused by under-sampling will be at its maximum at the transmit focus. Simulated LSV-plots are generated from this point, as specified on page 5, first paragraph.}
 \\\\
@@
 Without this information, the work is not reproductible.
+\textit{More extensive information about the probe and scan sequence have been added. (Not yet. Need to check with GE.)}
 \\\\
 In section “IV-A. Oversampling in transmit”: please confirm (specify) that 1x1 receive beamforming is performed at this point.
+\textit{We have now specified that our starting point is a system with a one-to-one relationship between tx and rx beams.}
 \\\\
@@
 Oversampling by parallel receive beamforming”, could you explain in more details what are the differences between Capon + STB and the technique proposed by Rabinovitch (2013) – reference [27]?
+\textit{There are no differences. Rabinovitch (2013 (reference [25]) does not propose any new technique, it is just Capon + STB.}
 \\\\
 Page10 Column1 line56  write “should be removed” instead of “should be remove”
+\textit{Misspelling has been corrected.}
 \\\\
@@
 It would be interesting to see such technique implemented.
+\textit{It would, but while minimizing the length of the manuscript an actual result was left out.
+It should, however, be clear from how STB corrects errors introduced by PRB, that something similar will work for phase-rotated rx-beams.}
 \\\\
 P10 column 2 line 28-29 write “too extreme” instead of “to extreme”
+\textit{Misspelling has been corrected.}
 \\\\
@@
 6), just in case some people won’t look at the videos.
-
+\textit{We have now specified that the image is processed with delay-and-sum beamformer. Default values for $p$ and $q$ are used and are therefore not specified. (Should we add images?)}
 \end{document}

### 2014-02-18 HRUIofMovingObjects/jrnl_article.tex
Response to reviewers.

@@
 \subsection{Oversampling on Transmit}
-A straight forward way of obtaining sufficient sampling in an active imaging system is by increasing the number of transmit beams (tx-beams) until the system is locally laterally shift-invariant.
+A straight forward way of obtaining sufficient sampling in an active imaging system with an equal number of transmit (tx) and receive (rx) beams is to increasing the number of tx-beams until the system is locally laterally shift-invariant.
 In Fig.\,\ref{fig:capon_oversampling_a} and Fig.\,\ref{fig:capon_oversampling_b} we have increased the tx-beam density by a $q$-factor of four and sixteen respectively.
@@
 P-MAE is not shown due to a lack of probe movement control.
-a) First frame in the recorded loop.
+a) First frame in the recorded loop processed with delay-and-sum beamforming.
 A white box indicates the point scatterer used to construct the LSV-plots.
@@
 c) Capon beamforming without oversampling.
-d) Capon with ten times oversampling using phase rotation.}
+d) Capon with sixteen times oversampling using phase rotation.}
 \label{fig:LSV_invitro}
@@
 eight phase rotated beams are averaged into four.
-The geometric distortion, or stripes, seen at the bottom of the sector in the video with sixteen times oversampling \multimedia{Media Movie 3} should be remove by such technique.
-
-When oversampling is applied it becomes evident that the high resolution settings used in this paper might be to extreme, both with respect to the level of required oversampling and the point scatterer's appearance.
+The geometric distortion, or stripes, seen at the bottom of the sector in the video with sixteen times oversampling \multimedia{Media Movie 3} should be removed by such technique.
+
+When oversampling is applied it becomes evident that the high resolution settings used in this paper might be too extreme, both with respect to the level of required oversampling and the point scatterer's appearance.
 Traditionally in medical ultrasound imaging the point spread function is always wider laterally than in range.

### 2014-02-19 ThesisJP/Background/background.tex
Started writing compare cpu and gpu.

@@
 The last three generations of CPUs from Intel have all had on-chip integrated graphics.
-Even if an on-chip GPU is less powerful than a discrete GPU, valuable time is saved by not having to send data across the PCI Express bus, which connects the different parts of a computer to the motherboard.
+Even if an on-chip GPU is less powerful than a high-performance s discrete GPU, valuable time is saved by not having to send data across the PCI Express bus, which connects the different parts of a computer to the motherboard.
 This is the reason why the first rules of GPGPU (with discrete GPUs) is to minimize CPU-to-GPU transfers.
@@
 Together, multi-core, heterogeneous cores, and cloud cores forms the "jungle" that today's software engineers have to tackle (Fig.\,\ref{fig:jungle}).
-This thesis focus on heterogeneous computing with one multi-core CPU and a discrete high performance GPU, a cheap and high performance system which is likely to be the specification of current and future ultrasound imaging system.
+In this thesis we focus on heterogeneous computing with one multi-core CPU and a discrete high performance GPU, a cheap and high performance system which is likely to be the specification of current and future ultrasound imaging system.
 
@@
 Even though they started out as very different processors, modern CPUs are including more and more GPU-like features, and GPUs are adding CPU features.
-In this section we will see why there is a big difference in peak performance for a CPU and GPU.
+In this section we will see what these similarities are and why there is still a big difference in peak performance between a CPU and GPU.
+
+In Section \ref{sec:gpgpu} we introduced a special type of instructions, namely SIMD instructions, which are designed to execute a single instruction across multiple data element in one clock cycle.
+At the end of the 20th century, Intel introduced a SIMD instruction set for their Pentium III processor called the Streaming SIMD Extensions (\nom{SSE}{Streaming SIMD Extensions}).
+This extension has seen several updates since.
+The first SEE instructions set had 128 bit long registers which meant that four 32 bit floating point values could be processed per clock cycle\footnote{For simplicity we will assume that the whole register is processed in one clock cycle even if the actual implementation might distributes one instruction for the register across multiple cycles.}.
+Updates of the SSE instructions sets have added new instructions (SSE2, SSE3, and SSE4) and longer registers (\nom{AVX}{Advanced Vector Extensions}, AVX2, and AVX-512).
+AVX and AVX2 adds 256 bit long registers which means that eight 32 bit floating point values can be processed per clock cycle.
+This extension is found in the latest CPUs from Intel.
+AVX-512 extends the registers to 512 bit and is found in the Intel accelerator board  Xeon Phi and might show up in the next generation of CPUs.
+The latest quad-core CPU from Intel, the Haswell architecture, can therefore process 32 32 bit floating point values per clock cycle.
+With a clock frequency of 3.9 GHz this sums up to a theoretical peak single precision performance of 124.8 giga single precision floating point operations per second (GFLOPS).
+In addition, the most powerful integrated GPU found on Haswell CPUs have 40 execution units capable of processing 16 32 bit floating point values at 1.1 GHz.
+The total theoretical peak single precision FLOPS for a high-end Haswell CPU is therfore $124.8 + (40*16*1.1) = 828.8$ GFLOPS as visualized in Fig\,\ref{fig:cpu_vs_gpu}.
+
+Modern GPUs can also be interpreted as a unit consisting of multiple SIMD processors.
+The Fermi GPU architecture by Nvidia, which is used the most in this theses, have up to 16 SIMD units capable of performing 32 32 bit floats per clock cycle at 772 MHz.
+The theoretical peak single precision FLOPS for this architecture is by that found to be $16*32*772 = $ (needs to add FMA and SP-units, What about FMA on the CPU?)
 
@@
 Single instruction, multiple data (SIMD).
+
+In this section we have focused on CPUs from Intel and GPUs from Nvidia, note that GPUs and CPUs from AMD have similar features.
 
@@
 }
-\caption{Depicts how GPU threads are grouped into blocks and arranged in a grid. One thread runs a copy of a kernel function, in this example an element-wise matrix square operation.}
+\caption{Depicts how GPU threads are grouped into blocks and arranged in a grid.
+One thread runs a copy of a kernel function.
+In this example the kernel function performs an element-wise matrix square.}
 \label{fig:gpu_grid}

### 2014-02-20 HRUIofMovingObjects/response_to_reviewers.tex
Ready for 2nd submit.

@@
 The paper is good but I would like it to go through a second review round based on the following comments:
-\textit{We want to thank the reviewer for an encouraging review. Below follows our response to his/her comments.}
+\textit{We want to thank the reviewer for an encouraging review. Below follows our response to the comments given.}
 \\\\
@@
 I would suggest that in the introduction we do not need these details.
-\textit{(Should we remove this?)}
+\textit{These details have now been removed from the introduction.}
 \\\\
@@
 Let us say that the oversampling is used both for Capon and for reducing folding of frequencies over the bandwidth.
-\textit{We do not find this confusing, $p$ is always 2 and $q$ is 4 and 16 when oversampling on transmit is applied for the Capon beamformer.}
+\textit{We do not find this confusing, $p$ is always 2 and $q$ is 4 and 16 when oversampling on transmit is applied for the Capon beamformer (as mention in the first paragraph of Section IV-A).
+In all the other examples $q=1$.
+The $p$ and $q$ is there to clearly separate oversampling from the Nyquist requirement.}
 \\\\
 At what depth are the LSV plots taken with respect to the focal depth? Does it matter?
-\textit{The shift-invariance caused by under-sampling will be at its maximum at the transmit focus. Simulated LSV-plots are generated from this point, as specified on page 5, first paragraph.}
+\textit{The shift-invariance caused by under-sampling will be at its maximum at the transmit focus.
+Simulated LSV-plots are generated from this point, as specified on page 5, first paragraph.
+In the in-vitro data, the point scatterer closest to the transmit focus is selected.}
 \\\\
@@
 Without this information, the work is not reproductible.
-\textit{More extensive information about the probe and scan sequence have been added. (Not yet. Need to check with GE.)}
+\textit{More extensive information about the probe and scan sequence have been added.}
 \\\\
@@
 Oversampling by parallel receive beamforming”, could you explain in more details what are the differences between Capon + STB and the technique proposed by Rabinovitch (2013) – reference [27]?
-\textit{There are no differences. Rabinovitch (2013 (reference [25]) does not propose any new technique, it is just Capon + STB.}
+\textit{There are no differences. Rabinovitch (2013), reference [25], does not propose any new technique, it is just Capon + STB.}
 \\\\
@@
 6), just in case some people won’t look at the videos.
-\textit{We have now specified that the image is processed with delay-and-sum beamformer. Default values for $p$ and $q$ are used and are therefore not specified. (Should we add images?)}
+\textit{We have now specified that the image is processed with delay-and-sum beamformer.
+Default values for $p$ and $q$ are used and are therefore not specified.
+Images are not added in order to minimize the length of the paper.}
 \end{document}

### 2014-02-20 HRUIofMovingObjects/jrnl_article.tex
Ready for 2nd submit.

@@
 \newcommand{\R}{$\mat{\hat{R}}$ }
+\newcommand{\degree}{\ensuremath{^\circ}}
 
@@
 \IEEEPARstart{I}{n} the last decade several authors have applied the Capon beamformer (also known as the minimum variance beamformer) for medical ultrasound imaging \cite{Synnevag2007, Vignon2008, Viola}.
-The method has been shown to improve lateral resolution and edge sharpness \cite{Synnevag2007, Synnevag2009, Chen2011} in ultrasound images.
+The method has been shown to improve lateral resolution and edge sharpness in ultrasound images \cite{Synnevag2007, Synnevag2009, Chen2011}.
 This is accomplished by adaptively adjusting the array weights based on element-to-element covariance in order to minimize interference.
@@
 Hence, the interference-noise matrix ($\sigma_0^2\mat{Q}$ in \cite{Cox1973}) is the identity.
-The effect of typical subarray averaging and diagonal loading ($L=32$ and $d=1/100$) have been added to the Capon beamformer output.
+The effect of typical subarray averaging and diagonal loading have been added to the Capon beamformer output.
 We see that a DAS beamformer, using the required beam spacing, could yield a $1.9$ dB scalloping loss \cite{Harris1978} if the object is moving parallel to the array.
@@
 }
-\caption{Estimated drop in output power mid-way between receive beams for delay-and-sum and Capon beamforming ($M=64$, $L=32$, $d=0.01$) applied in a farfield-narrowband setting.
+\caption{Estimated drop in output power mid-way between receive beams for delay-and-sum and Capon beamforming (with typical robustification parameters) applied in a farfield-narrowband setting.
 %The scanned output from both beamformers is plotted for two different time instances.
@@
 Dotted vertical lines indicate position of transmit-receive beam pairs.
-The point scatterer response has been simulated using Field II \cite{Jensen1992, Jensen1996a} with a probe setup equal to the M5S probe by GE Vingmed Ultrasound, Horten, Norway.
+The point scatterer response has been simulated using Field II \cite{Jensen1992, Jensen1996a} with a probe setup equal to the M5S-D phased array by GE Vingmed Ultrasound, Horten, Norway.
 The probe was excited with a 2.5 MHz and 1.5 cycles long pulse.
@@
 
-Figure \ref{fig:capon_oversampling_pr} shows the phase rotation method applied on the same scan sequence as in Fig\,\ref{fig:capon} with four and sixteen times oversampling respectively.
+Figure \ref{fig:capon_oversampling_pr} shows the phase rotation method applied on the same scan sequence as in Fig\,\ref{fig:capon} with four and sixteen times oversampling on receive respectively ($q$-factor).
 Note that the figure is zoomed compared with Fig\,\ref{fig:capon}.
@@
 In Fig.\,\ref{fig:LSV_invitro} we present the result of applying the Capon beamformer on \textit{in-vitro} data with and without oversampling by phase rotation.
-The data was acquired using a modified Vivid E9 high-end cardiac ultrasound scanner by GE Vingmed Ultrasound, Horten, Norway, equipped with a M5S transducer operating in harmonic mode with a receive frequency of 3.7 MHz.
-The transmit beam density was set to match (\ref{eq:resolution}) based on this frequency.
+The data was acquired using a modified Vivid E9 high-end cardiac ultrasound scanner by GE Vingmed Ultrasound, Horten, Norway, equipped with a M5S-D phased transducer operating in harmonic mode at 3.34 MHz.
+The transmit beam density was set to match (\ref{eq:resolution}) based on this frequency, and on a 65\degree sector we therefore distributed 111 tx-beams.
+Transmit focus is located at 8 cm range.
 Note that in harmonic imaging the generated transmit beam at the harmonic frequency will be wider and have lower sidelobes than the beam resulting from transmitting at the harmonic frequency \cite{Fedewa2004}.
@@
 A tissue mimicking phantom containing wire targets and anechoic cysts was imaged, and the data was acquired while the probe was moved across the phantom manually.
-Fig.\,\ref{fig:image_das} shows the first frame out of 14 in the acquired loop.
+Fig.\,\ref{fig:image_das} shows the first frame out of 14 in the acquired loop processed with DAS beamforming.
 %Movies of the whole loop processed with DAS and Capon beamforming with and without ten times oversampling by phase rotation can be found in the supplementary material \multimedia{Videos}.
@@
 One point scatterer in Fig\,\ref{fig:image_das} has been selected [white box], and then used to generate LSV-plots.
+The point scatterer is located close to the transmit focus.
 Figure \ref{fig:phantom_das} shows the resulting LSV-plot when using the DAS beamformer.

### 2014-02-21 ThesisJP/Background/background.tex
Stil writing...

@@
 \centering
-\includegraphics[width=0.9\textwidth]{img/free_lunsh.png}
+\includegraphics[width=0.8\textwidth]{img/free_lunsh.png}
 \caption{Illustration of transitions between different eras in the computer industry.
@@
 When geometry is projected and rasterized the same instruction is typically needed for a lot of data at the same time.
-This makes a special kind of instruction, Single Instruction Multiple Data (\nom{SIMD}{Single instruction multiple data (instruction type)}) instruction \cite{Flynn1966}, especially suited for this kind of problem.
+This makes a special kind of instruction, Single Instruction Multiple Data (\nom{SIMD}{Single instruction multiple data (instruction)}) instruction \cite{Flynn1966}, especially suited for this kind of problem.
 The differences between how SIMD instructions are utilized in modern CPUs and GPUs are discussed in Section \ref{sec:cpu_vs_gpu}.
@@
 
+\begin{figure}
+\centering
+\includegraphics[width=\textwidth]{img/CPU-GPU.png}
+\caption{Overview of CPU and GPU architecture. (make my own)}
+\label{fig:cpu_gpu}
+\end{figure}
+
 In Section \ref{sec:gpgpu} we introduced a special type of instructions, namely SIMD instructions, which are designed to execute a single instruction across multiple data element in one clock cycle.
-At the end of the 20th century, Intel introduced a SIMD instruction set for their Pentium III processor called the Streaming SIMD Extensions (\nom{SSE}{Streaming SIMD Extensions}).
+At the end of the 20th century, Intel introduced a SIMD instruction set for their Pentium III processor called the Streaming SIMD Extensions (\nom{SSE}{Streaming SIMD extensions (instruction)}).
 This extension has seen several updates since.
 The first SEE instructions set had 128 bit long registers which meant that four 32 bit floating point values could be processed per clock cycle\footnote{For simplicity we will assume that the whole register is processed in one clock cycle even if the actual implementation might distributes one instruction for the register across multiple cycles.}.
-Updates of the SSE instructions sets have added new instructions (SSE2, SSE3, and SSE4) and longer registers (\nom{AVX}{Advanced Vector Extensions}, AVX2, and AVX-512).
-AVX and AVX2 adds 256 bit long registers which means that eight 32 bit floating point values can be processed per clock cycle.
-This extension is found in the latest CPUs from Intel.
-AVX-512 extends the registers to 512 bit and is found in the Intel accelerator board  Xeon Phi and might show up in the next generation of CPUs.
-The latest quad-core CPU from Intel, the Haswell architecture, can therefore process 32 32 bit floating point values per clock cycle.
-With a clock frequency of 3.9 GHz this sums up to a theoretical peak single precision performance of 124.8 giga single precision floating point operations per second (GFLOPS).
-In addition, the most powerful integrated GPU found on Haswell CPUs have 40 execution units capable of processing 16 32 bit floating point values at 1.1 GHz.
-The total theoretical peak single precision FLOPS for a high-end Haswell CPU is therfore $124.8 + (40*16*1.1) = 828.8$ GFLOPS as visualized in Fig\,\ref{fig:cpu_vs_gpu}.
+Updates of the SSE instructions sets have added new instructions (SSE2, SSE3, and SSE4) and longer registers (\nom{AVX}{Advanced vector extensions (instruction)}, AVX2, and AVX-512).
+AVX and AVX2 adds 256 bit long registers and fused multiply-accumulate instructions (\nom{FMA}{Fused multiply-accumulate (instruction)})  which means that sixteen 32 bit floating point values can be processed per clock cycle per core.
+A new extension, AVX-512, extends the registers to 512 bit and is found in the Intel accelerator board  Xeon Phi and might show up in the next generation of CPUs.
+The latest CPU architecture from Intel, the Haswell architecture, has two FMA ports and is therefore capable of processing 32, 32 bit, floating point values per clock cycle per core (Fig.\,\ref{fig:cpu_gpu}).
+For an eight core\footnote{An 8-core Haswell CPU is scheduled to launch during 2014.} CPU with a clock frequency of 3.8 GHz this sums to
+\begin{equation}
+32*8*3.8 = 972.8\,\text{GFLOPS},
+\end{equation}
+hence a theoretical peak single precision performance of close to one tera single precision floating point operations per second (TFLOPS).
+This is the same number as visualized in Fig\,\ref{fig:cpu_vs_gpu} for the Haswell architecture.
+In addition, the most powerful integrated GPU found on Haswell CPUs have 40 execution units capable of processing sixteen, 32 bit, floating points at 1.1 GHz,
+\begin{equation}
+16*40*1.1 = 704\,\text{GFLOPS}.
+\end{equation}
 
-Modern GPUs can also be interpreted as a unit consisting of multiple SIMD processors.
-The Fermi GPU architecture by Nvidia, which is used the most in this theses, have up to 16 SIMD units capable of performing 32 32 bit floats per clock cycle at 772 MHz.
-The theoretical peak single precision FLOPS for this architecture is by that found to be $16*32*772 = $ (needs to add FMA and SP-units, What about FMA on the CPU?)
+Modern GPUs can also be interpreted as a unit consisting of multiple SIMD processors (Fig.\,\ref{fig:cpu_gpu}).
+The latest GPU architecture by Nvidia have up to 15 SIMD units capable of processing 192, 32 bit, floats per clock cycle at 745 MHz.
+This GPU is also capable of performing FMA instructions, doubling the number of theoretical FLOPS.
+The theoretical peak single precision FLOPS for this architecture is by that found to be
+\begin{equation}
+15*192*2*745 = 4291\,\text{GFLOPS}.
+\end{equation} 
+Each SIMD core also comprise 16 special function units (\nom{SFU}{Special function unit (cos, exp, ect.)}) which add close to 1 TFLOPS in additional power.
+We see how the wide SIMD register found on GPUs results in five times the performance of a high-end CPU (without the integrated GPU).
 
-It is common to say that a CPU has few but advance generic cores, and the GPU has many but simple cores.
-
-The SIMD registers on a GPU is longer than on a CPU and the GPU have more of them.
-
- the most powerful integrated intel GPU (HD 5200) have a peak performance of nearly one tera single precision floating point operations per second (FLOPs).
-
-It it is also much less costly to switch between a pixel-thread for the GPU then a program thread for the CPU.
-
-About theoretical FLOPS: Note that these are theoretical numbers and the actual throughput typically is algorithm dependent.
-
-Compare CPU and GPU ALUs.
-SSE, AVX ...
-The GPU has super-wide registers.
-The SM in Fermi have 32 cores or 1024 bit registers.
-The SMX in Kepler have 192 cores.
-6144 bit registers.
-SSE (Streaming SIMD Extensions) instructions have 128 bit registers.
-AVX (Advanced Vector Extensions) have 256 bit registers.
-AVX-512 has 512 bit registers.
-Single instruction, multiple data (SIMD).
-
-In this section we have focused on CPUs from Intel and GPUs from Nvidia, note that GPUs and CPUs from AMD have similar features.
+Note that the derived numbers are strictly theoretical and that actual throughput typically is algorithm dependent.
+In this section we have focused on CPUs from Intel and GPUs from Nvidia, however, GPUs and CPUs from AMD have similar specifications.
 
@@
 
-\begin{figure}
-\centering
-\includegraphics[width=\textwidth]{img/CPU-GPU.png}
-\caption{Overview of CPU and GPU architecture.}
-\label{fig:cpu_gpu}
-\end{figure}
-
 \subsection{Programming a GPU}
@@
 Architecture and how to program. See Paper II.
+
+It it is also much less costly to switch between a pixel-thread for the GPU then a program thread for the CPU.
 

### 2014-02-21 ThesisJP/Introduction/introduction.tex
Stil writing...

@@
 J.\:I.\:Buskenes, \textbf{J.\:P.\:\AA{}sen}, C.-I.\:C.\:Nilsen and A.\:Austeng\\
-{\it IEEE Transactions on Oceanic Engineering, submitted.}\\\\
+{\it IEEE Transactions on Oceanic Engineering, accepted for publication.}\\\\
 The first paper describes in details how the Capon beamformer was mapped to GPU architecture.
@@
 \textbf{J.\:P.\:\AA{}sen}, A.\:Austeng and S.\:Holm\\
-{\it IEEE Transactions on Ultrasonics, Ferroelectrics, and Frequency Control, submitted.}\\\\
+{\it IEEE Transactions on Ultrasonics, Ferroelectrics, and Frequency Control, accepted for publication.}\\\\
 In the third paper we study the shift-invariant property of an imaging system based on the Capon beamformer.

### 2014-02-21 ThesisJP/Thesis.tex
Continued writing background information.
writing and added figure.
Stil writing...

@@
 \newcommand*{\nom}[2]{#1\nomenclature{#1}{#2}}
+
+% custom commands
+\newcommand{\mat}[1]{\mathbf{#1}}
+\renewcommand{\vec}[1]{\mathbf{#1}}
+\newcommand{\degree}{\ensuremath{^\circ}}
+\newcommand{\R}{$\mat{\hat{R}}$ }
 
@@
 \renewcommand\labelenumi{\bfseries\theenumi}
-	\item J.\:I.\:Buskenes, \textbf{J.\:P.\:\AA{}sen}, C.-I.\:C.\:Nilsen and A.\:Austeng, "An optimised GPU implementation of the MVDR beamformer for active sonar imaging", {\it  IEEE Transactions on Oceanic Engineering, submitted}.
+	\item J.\:I.\:Buskenes, \textbf{J.\:P.\:\AA{}sen}, C.-I.\:C.\:Nilsen and A.\:Austeng, "An optimised GPU implementation of the MVDR beamformer for active sonar imaging", {\it  IEEE Transactions on Oceanic Engineering, accepted for publication}.
 	\item \textbf{J.\:P.\:\AA{}sen}, J.\:I.\:Buskenes, C.-I.\:C.\:Nilsen, A.\:Austeng and S.\:Holm, "Implementing Capon beamforming on a GPU for real-time cardiac ultrasound imaging", {\it IEEE Transactions on Ultrasonics, Ferroelectrics, and Frequency Control, vol.
@@
 76-85, Jan 2014}.
- 	\item \textbf{J.\:P.\:\AA{}sen}, A.\:Austeng and S.\:Holm, "Capon beamforming and moving objects - An analysis of lateral shift-invariance", {\it IEEE Transactions on Ultrasonics, Ferroelectrics, and Frequency Control, submitted}.
+ 	\item \textbf{J.\:P.\:\AA{}sen}, A.\:Austeng and S.\:Holm, "Capon beamforming and moving objects - An analysis of lateral shift-invariance", {\it IEEE Transactions on Ultrasonics, Ferroelectrics, and Frequency Control, accepted for publication}.
 	\item \textbf{J.\:P.\:\AA{}sen}, E.\:Steen, G.\:Kiss, A.\:Thorstensen and S.\:I.\:Rabben, "Adaptive volume rendering of cardiac 3D ultrasound images - utilizing blood pool statistics", {\it Proc.
@@
 8320, pp.
-832008.}.
+832008}.
 	\item \textbf{J.\:P.\:\AA{}sen} and S.\:Holm, "Huygens on speed: Interactive simulation of ultrasound pressure fields", {\it Proc. IEEE Ultrasonics Symposium 2012, pp. 1643-1646}.
@@
 }
-\noindent \textit{IEEE Transactions on Oceanic Engineering, submitted}.
+\noindent \textit{IEEE Transactions on Oceanic Engineering, accepted for publication}.
 %\newpage\input{./CaponSonarGPU/CaponSonarGPU}
@@
 }
-\noindent \textit{IEEE Transactions on Ultrasonics, Ferroelectrics, and Frequency Control, submitted}.
+\noindent \textit{IEEE Transactions on Ultrasonics, Ferroelectrics, and Frequency Control, accepted for publication}.
 %\input{./CaponMovingObjects/CaponMovingObjects}

### 2014-02-22 ThesisJP/Background/background.tex
Some background writing.

@@
 The essay bears the name "Welcome to the Jungle" and is a sequel to the essay "The Free Lunch is Over" from 2004 \cite{HerbSuttera}.
-The two essays spins around the fact that chip manufactures hit a frequency-wall in the beginning of the 21st century.
-Until then, all mainstream computer programs where typically running in a single thread on a single core, and a programmer could expect the software to annually gain performance without touching the code.
+The two essays spins around the fact that CPU manufactures hit a frequency-wall in the beginning of the 21st century.
+Until then, all mainstream computer programs where typically running in a single thread on a single CPU \textit{processing core}, and a programmer could expect the software to annually gain performance without touching the code.
 This was the "free-lunch"-era as depicted in Fig.\,\ref{fig:jungle}.
@@
 This insured continuation of Moore's law, since the number of transistors per chip could continue to grow.
-For programmers, multi-core CPUs meant that computer programs now had to be multi-threaded in order to continue to gain performance.
+For programmers, multi-core CPUs meant that computer programs now had to be multi-threaded in order to annually gain performance.
 It was a concurrency revolution, as Herb Sutter named it, and  the "free lunch" was over.
@@
 In most cases it is important to balance the load equally between the CPU and GPU to obtain their combined computationally power.
-Balancing typically means that highly parallel tasks should be offloaded to the GPU and serialized work should stay on the CPU.
+Balancing typically means that highly parallel and computationally intense tasks should be offloaded to the GPU and serialized and memory intensive work should stay on the CPU.
 This paradigm of utilizing specialized cores for solving specific problems is today known as heterogeneous computing.
@@
 Even if an on-chip GPU is less powerful than a high-performance s discrete GPU, valuable time is saved by not having to send data across the PCI Express bus, which connects the different parts of a computer to the motherboard.
-This is the reason why the first rules of GPGPU (with discrete GPUs) is to minimize CPU-to-GPU transfers.
+This is the reason why one of the first rules of GPGPU (with discrete GPUs) is to minimize CPU-to-GPU memory transfers.
 
@@
 Even though they started out as very different processors, modern CPUs are including more and more GPU-like features, and GPUs are adding CPU features.
-In this section we will see what these similarities are and why there is still a big difference in peak performance between a CPU and GPU.
+In this section we will see what these similarities are and why there is still a big difference in peak performance between a CPU and a GPU.
 
@@
 \centering
-\includegraphics[width=\textwidth]{img/CPU-GPU.png}
-\caption{Overview of CPU and GPU architecture. (make my own)}
+\includegraphics[width=0.8\textwidth]{img/CPU-GPU.pdf}
+\caption{Schematic overview of CPU (Intel Haswell) and GPU (Nvidia Kepler) architecture.
+Note how more space is dedicated for compute cores on a GPU, and how more space is dedicated to control and caching on a CPU.}
 \label{fig:cpu_gpu}
@@
 The first SEE instructions set had 128 bit long registers which meant that four 32 bit floating point values could be processed per clock cycle\footnote{For simplicity we will assume that the whole register is processed in one clock cycle even if the actual implementation might distributes one instruction for the register across multiple cycles.}.
-Updates of the SSE instructions sets have added new instructions (SSE2, SSE3, and SSE4) and longer registers (\nom{AVX}{Advanced vector extensions (instruction)}, AVX2, and AVX-512).
+Updates of the SSE instruction set have added new instructions (SSE2, SSE3, and SSE4) and longer registers (\nom{AVX}{Advanced vector extensions (instruction)}, AVX2, and AVX-512).
 AVX and AVX2 adds 256 bit long registers and fused multiply-accumulate instructions (\nom{FMA}{Fused multiply-accumulate (instruction)})  which means that sixteen 32 bit floating point values can be processed per clock cycle per core.
 A new extension, AVX-512, extends the registers to 512 bit and is found in the Intel accelerator board  Xeon Phi and might show up in the next generation of CPUs.
-The latest CPU architecture from Intel, the Haswell architecture, has two FMA ports and is therefore capable of processing 32, 32 bit, floating point values per clock cycle per core (Fig.\,\ref{fig:cpu_gpu}).
+The latest CPU architecture from Intel, the Haswell architecture, has two processing ports supporting FMA-AVX2 instructions, and is therefore capable of processing 32, 32 bit, floating point values per clock cycle per core (Fig.\,\ref{fig:cpu_gpu}).
 For an eight core\footnote{An 8-core Haswell CPU is scheduled to launch during 2014.} CPU with a clock frequency of 3.8 GHz this sums to
@@
 \end{equation}
-hence a theoretical peak single precision performance of close to one tera single precision floating point operations per second (TFLOPS).
+hence a theoretical peak performance of close to one tera single precision floating point operations per second (TFLOPS).
 This is the same number as visualized in Fig\,\ref{fig:cpu_vs_gpu} for the Haswell architecture.
-In addition, the most powerful integrated GPU found on Haswell CPUs have 40 execution units capable of processing sixteen, 32 bit, floating points at 1.1 GHz,
+In addition, the most powerful integrated GPU found on Haswell CPUs have 40 execution units capable of processing sixteen, 32 bit, floating points at 1.1 GHz.
+This adds
 \begin{equation}
-16*40*1.1 = 704\,\text{GFLOPS}.
+16*40*1.1 = 704\,\text{GFLOPS}
 \end{equation}
+in extra processing power to the CPU chip.
 
@@
 The latest GPU architecture by Nvidia have up to 15 SIMD units capable of processing 192, 32 bit, floats per clock cycle at 745 MHz.
-This GPU is also capable of performing FMA instructions, doubling the number of theoretical FLOPS.
+These GPUs are also capable of performing FMA instructions, effectively doubling the number of theoretical FLOPS.
 The theoretical peak single precision FLOPS for this architecture is by that found to be
@@
 \end{equation} 
-Each SIMD core also comprise 16 special function units (\nom{SFU}{Special function unit (cos, exp, ect.)}) which add close to 1 TFLOPS in additional power.
-We see how the wide SIMD register found on GPUs results in five times the performance of a high-end CPU (without the integrated GPU).
+Each SIMD core also comprise 16 special function units (\nom{SFU}{Special function unit (sin, exp, sqrt, ect.)}) which add close to 1 TFLOPS in additional computational power.
+This adds up to five TFLOPS in computationally power as shown in Fig.\,\ref{fig:cpu_vs_gpu} for the GeForce GTX 780 TI GPU.
+From Fig.\,\ref{fig:cpu_gpu} we see how the high number of SIMD cores found on GPUs and their larger width results in five times the computational performance of a high-end CPU (without the integrated GPU).
 
-Note that the derived numbers are strictly theoretical and that actual throughput typically is algorithm dependent.
-In this section we have focused on CPUs from Intel and GPUs from Nvidia, however, GPUs and CPUs from AMD have similar specifications.
+Note that the derived numbers are strictly theoretical and that actual throughput is highly algorithm dependent.
+Nevertheless, in theory, the benefit of GPU computing over CPU computing is currently a factor of ten\footnote{A factor of five when a eight core Haswell CPU is launched later this year.} in raw computationally power.
+A claimed speed-up larger than this is typically caused by comparing a non-optimized CPU implementation with an optimized GPU implementation or by comparing uneven hardware \cite{Lee2010, Kothapalli2013}.
+Also note that in this section we have focused on CPUs from Intel and GPUs from Nvidia, however, CPUs and GPUs from AMD have similar specifications.
 
@@
 \subsection{Programming a GPU}
-Early gpgpu programming required expert knowledge about graphics and the remaining fixed funtions in the pipeline. New programming languages where needed. OpenCL and CUDA. C++ AMP.
+Early GPGPU programming required expert knowledge about computer graphics and the remaining fixed functions in the pipeline.
+When GPGPU gained interest it quickly became evident that new programing languages where needed.
+One of the first GPGPU oriented languages for the GPU was the Brook language by Buck \textit{et al.} \cite{Buck2004}.
+Buck later joined Nvidia where he led the work on \nom{CUDA}{Compute unified device architecture (GPGPU language)}, a GPGPU language by Nvidia for Nvidia GPUs only.
+The first version of CUDA was released in 2007.
+The following year, Apple launched their own GPGPU language called OpenCL, a standard which are administrated by the Krohnos Group.
+
+New programming languages where needed. OpenCL and CUDA. C++ AMP.
 

### 2014-02-22 ThesisJP/Background/background.tex
Writing about GPU programming.

@@
 Note that the derived numbers are strictly theoretical and that actual throughput is highly algorithm dependent.
-Nevertheless, in theory, the benefit of GPU computing over CPU computing is currently a factor of ten\footnote{A factor of five when a eight core Haswell CPU is launched later this year.} in raw computationally power.
+Nevertheless, in theory, the benefit of GPU computing over CPU computing is currently a factor of five and ten\footnote{Until an eight core Haswell CPU is launched later this year.} in raw computationally power.
 A claimed speed-up larger than this is typically caused by comparing a non-optimized CPU implementation with an optimized GPU implementation or by comparing uneven hardware \cite{Lee2010, Kothapalli2013}.
@@
 \includegraphics[width=\textwidth]{img/cpu_vs_gpu.pdf}
-\caption{Development in theoretical peak throughput, single precision (SP) and double precision (DP), the last decade for Intel CPUs and Nvidia GPUs.
-Image courtesy of Michael Galloy (michaelgalloy.com).}
+\caption{Development in theoretical peak throughput, single precision (SP) and double precision (DP), the last decade for Intel CPUs and Nvidia GPUs.}
 \label{fig:cpu_vs_gpu}
@@
 One of the first GPGPU oriented languages for the GPU was the Brook language by Buck \textit{et al.} \cite{Buck2004}.
-Buck later joined Nvidia where he led the work on \nom{CUDA}{Compute unified device architecture (GPGPU language)}, a GPGPU language by Nvidia for Nvidia GPUs only.
+Buck later joined Nvidia where he led the work on \nom{CUDA}{Compute unified device architecture (GPGPU language)}, a GPGPU language and framework by Nvidia for Nvidia GPUs only.
 The first version of CUDA was released in 2007.
-The following year, Apple launched their own GPGPU language called OpenCL, a standard which are administrated by the Krohnos Group.
+The following year, Apple launched a GPGPU framework called \nom{OpenCL}{Open Compute Language (GPGPU language)}, a standard which are maintained by the Khronos Group.
+These two are still today the main programming frameworks for GPGPU.
+CUDA only runs on Nvidia GPUs while OpenCL can now run on GPUs from both Nvidia and AMD, CPUs from both Intel and AMD, and on FPGAs from Altera.
 
-New programming languages where needed. OpenCL and CUDA. C++ AMP.
+\begin{figure}
+\centering
+\subfigure[Kernel]{
+	\includegraphics[width=0.5\textwidth]{img/kernel.pdf}\label{fig:gpu_kernel}
+}
+\subfigure[Threads, blocks and grid]{
+	\includegraphics[width=0.4\textwidth]{img/cuda_threads.png}\label{fig:gpu_grid}
+}
+\caption{The figure depicts how GPU threads are grouped into blocks and arranged in a grid.
+One thread runs a copy of a kernel function.
+In this example the kernel function performs an element-wise matrix square.}
+\label{fig:gpu_grid_kernel}
+\end{figure}
+
+Both CUDA and OpenCL is centered around a kernel function which are executed be a large group of threads.
+The kernel function, for both CUDA and OpenCL, has a C-like syntax with some additional specifiers.
+A CUDA kernel is depicted in Fig.\,\ref{fig:gpu_kernel}.
+The \texttt{\_\_global\_\_} specifier tells the CUDA compiler that this function is a kernel function, and, in CUDA, this kernel function is launched by using a CUDA specific syntax:
+\begin{align*}
+\texttt{square<<<grid, block>>>(a, N)}
+\end{align*}
+where \texttt{grid} and \texttt{block} is a structure containing the grid and block size as depicted in Fig.\,\ref{fig:gpu_grid}, and \texttt{a} is a pointer to a memory buffer located on the GPU.
+When launched, a copy of the kernel is executed for each thread in each block.
+Each kernel get as input its location in the compute grid (\texttt{blockIdx} and \texttt{threadIdx}).
+
+C++ AMP.
 
@@
 Add sentence about siemens and supersonic imagine scanners. They use GPUs? What does this mean? FPGAs and ASICs.
-
-\begin{figure}
-\centering
-\subfigure[Threads, blocks and grid]{
-	\includegraphics[width=0.4\textwidth]{img/cuda_threads.png}
-}
-\subfigure[Kernel]{
-	\includegraphics[width=0.5\textwidth]{img/kernel.pdf}
-}
-\caption{Depicts how GPU threads are grouped into blocks and arranged in a grid.
-One thread runs a copy of a kernel function.
-In this example the kernel function performs an element-wise matrix square.}
-\label{fig:gpu_grid}
-\end{figure}
 

### 2014-02-23 ThesisJP/Background/background.tex
Writing background...

@@
 In this chapter the background information needed in order to understand the topics of this thesis and its papers is given.
-First, in Section \ref{sec:gpgpu}, general-purpose programming on graphics processing units (\nom{GPGPU}{General-purpose programming on graphics processing units}) is reviewed, followed by an introduction to medical ultrasound imaging in Section \ref{sec:ultrasound}.
+First, in Section \ref{sec:gpgpu}, general-purpose computing on graphics processing units (\nom{GPGPU}{General-purpose computing on graphics processing units}) is reviewed, followed by an introduction to medical ultrasound imaging in Section \ref{sec:ultrasound}.
 Finally, the concepts of volume rendering and ultrasound field simulations are presented in Section \ref{sec:volren} and \ref{sec:field}.
@@
 The essay bears the name "Welcome to the Jungle" and is a sequel to the essay "The Free Lunch is Over" from 2004 \cite{HerbSuttera}.
-The two essays spins around the fact that CPU manufactures hit a frequency-wall in the beginning of the 21st century.
+The two essays spins around the fact that manufactures of central processing units (CPU) hit a frequency-wall in the beginning of the 21st century.
 Until then, all mainstream computer programs where typically running in a single thread on a single CPU \textit{processing core}, and a programmer could expect the software to annually gain performance without touching the code.
@@
 From that point, chip manufactures had to pursue a different direction  than just increasing the clock frequency.
-The immediate solution was to embed several cores in one CPU chip.
+The immediate solution was to embed several cores in one CPU.
 This insured continuation of Moore's law, since the number of transistors per chip could continue to grow.
@@
 When geometry is projected and rasterized the same instruction is typically needed for a lot of data at the same time.
-This makes a special kind of instruction, Single Instruction Multiple Data (\nom{SIMD}{Single instruction multiple data (instruction)}) instruction \cite{Flynn1966}, especially suited for this kind of problem.
+This makes a special kind of instruction, the Single Instruction Multiple Data (\nom{SIMD}{Single instruction multiple data (instruction)}) instruction \cite{Flynn1966}, especially suited for this kind of problem.
 The differences between how SIMD instructions are utilized in modern CPUs and GPUs are discussed in Section \ref{sec:cpu_vs_gpu}.
 
-For the first adopters of GPU programming, or General-purpose programming on graphics processing units (GPGPU) as it was named, it soon became evident that offloading all types of computations to the GPU was not always a good idea.
+For the first adopters general-purpose computing on GPUs (GPGPU) it soon became evident that offloading all types of computations to the GPU was not always a good idea.
 In most cases it is important to balance the load equally between the CPU and GPU to obtain their combined computationally power.
@@
 The last three generations of CPUs from Intel have all had on-chip integrated graphics.
-Even if an on-chip GPU is less powerful than a high-performance s discrete GPU, valuable time is saved by not having to send data across the PCI Express bus, which connects the different parts of a computer to the motherboard.
+Even if an on-chip GPU is less powerful than a high-performance discrete GPU, valuable time is saved by not having to send data across the PCI Express bus, which connects the different parts of a computer to the motherboard.
 This is the reason why one of the first rules of GPGPU (with discrete GPUs) is to minimize CPU-to-GPU memory transfers.
@@
 \caption{Schematic overview of CPU (Intel Haswell) and GPU (Nvidia Kepler) architecture.
-Note how more space is dedicated for compute cores on a GPU, and how more space is dedicated to control and caching on a CPU.}
+Note how more space is dedicated to compute cores on a GPU, and how more space is dedicated to control and caching on a CPU.}
 \label{fig:cpu_gpu}
@@
 This extension has seen several updates since.
-The first SEE instructions set had 128 bit long registers which meant that four 32 bit floating point values could be processed per clock cycle\footnote{For simplicity we will assume that the whole register is processed in one clock cycle even if the actual implementation might distributes one instruction for the register across multiple cycles.}.
+The first SEE instructions set had 128 bit long registers which meant that four, 32 bit, floating point values could be processed per clock cycle\footnote{For simplicity we will assume that the whole register is processed in one clock cycle even if the actual implementation might distributes one instruction for the register across multiple cycles.}.
 Updates of the SSE instruction set have added new instructions (SSE2, SSE3, and SSE4) and longer registers (\nom{AVX}{Advanced vector extensions (instruction)}, AVX2, and AVX-512).
@@
 \end{equation}
-in extra processing power to the CPU chip.
+in extra processing power to the APU.
 
 Modern GPUs can also be interpreted as a unit consisting of multiple SIMD processors (Fig.\,\ref{fig:cpu_gpu}).
-The latest GPU architecture by Nvidia have up to 15 SIMD units capable of processing 192, 32 bit, floats per clock cycle at 745 MHz.
+The Kepler architecture by Nvidia have up to 15 SIMD units capable of processing 192, 32 bit, floats per clock cycle at 745 MHz.
 These GPUs are also capable of performing FMA instructions, effectively doubling the number of theoretical FLOPS.
@@
 Each SIMD core also comprise 16 special function units (\nom{SFU}{Special function unit (sin, exp, sqrt, ect.)}) which add close to 1 TFLOPS in additional computational power.
-This adds up to five TFLOPS in computationally power as shown in Fig.\,\ref{fig:cpu_vs_gpu} for the GeForce GTX 780 TI GPU.
+This adds up to a total of five TFLOPS in computationally power as shown in Fig.\,\ref{fig:cpu_vs_gpu} for the GeForce GTX 780 TI GPU.
 From Fig.\,\ref{fig:cpu_gpu} we see how the high number of SIMD cores found on GPUs and their larger width results in five times the computational performance of a high-end CPU (without the integrated GPU).
@@
 Note that the derived numbers are strictly theoretical and that actual throughput is highly algorithm dependent.
-Nevertheless, in theory, the benefit of GPU computing over CPU computing is currently a factor of five and ten\footnote{Until an eight core Haswell CPU is launched later this year.} in raw computationally power.
+Nevertheless, in theory, the benefit of GPU computing over CPU computing is currently a factor of five or ten\footnote{Until an eight core Haswell CPU is launched later this year.} in raw computationally power.
 A claimed speed-up larger than this is typically caused by comparing a non-optimized CPU implementation with an optimized GPU implementation or by comparing uneven hardware \cite{Lee2010, Kothapalli2013}.
@@
 The following year, Apple launched a GPGPU framework called \nom{OpenCL}{Open Compute Language (GPGPU language)}, a standard which are maintained by the Khronos Group.
-These two are still today the main programming frameworks for GPGPU.
-CUDA only runs on Nvidia GPUs while OpenCL can now run on GPUs from both Nvidia and AMD, CPUs from both Intel and AMD, and on FPGAs from Altera.
+These two frameworks are still today the main programming frameworks for GPGPU.
+Where CUDA only runs on Nvidia GPUs, OpenCL can now run on GPUs from both Nvidia and AMD, CPUs from both Intel and AMD, FPGAs from Altera, and mobile processors from \nom{ARM}{Designer of mobile chip architectures}.
 
@@
 
-Both CUDA and OpenCL is centered around a kernel function which are executed be a large group of threads.
+Both CUDA and OpenCL is centered around a kernel function which are executed by a large group of threads.
 The kernel function, for both CUDA and OpenCL, has a C-like syntax with some additional specifiers.
 A CUDA kernel is depicted in Fig.\,\ref{fig:gpu_kernel}.
-The \texttt{\_\_global\_\_} specifier tells the CUDA compiler that this function is a kernel function, and, in CUDA, this kernel function is launched by using a CUDA specific syntax:
+The \texttt{\_\_global\_\_} specifier tells the CUDA compiler that this function is a kernel function, and in CUDA, this kernel function is launched by using a CUDA specific syntax:
 \begin{align*}
-\texttt{square<<<grid, block>>>(a, N)}
+\texttt{square<<<grid, block>>>(a, N)},
 \end{align*}
-where \texttt{grid} and \texttt{block} is a structure containing the grid and block size as depicted in Fig.\,\ref{fig:gpu_grid}, and \texttt{a} is a pointer to a memory buffer located on the GPU.
-When launched, a copy of the kernel is executed for each thread in each block.
-Each kernel get as input its location in the compute grid (\texttt{blockIdx} and \texttt{threadIdx}).
+where \texttt{grid} and \texttt{block} is 3-component vectors specifying grid and block size in "xyz" as depicted in Fig.\,\ref{fig:gpu_grid}, \texttt{a} is a pointer to a memory buffer located on the GPU, and \texttt{N}$*$\texttt{N} is the buffer size.
+When launched, a copy of the kernel is scheduled  for each thread in each block, and each kernel get as input its location in the compute grid (\texttt{blockIdx} and \texttt{threadIdx}) together with the grid size (\texttt{gridDim} and \texttt{blockDim}).
+This information is then typically used to calculate which memory elements to gather from the global memory buffer \texttt{a}.
+In order to formulate a given problem as a kernel function the global memory input to each thread must not rely on the output from other threads, hence it must be possible for the GPU to scheduled threads in any order.
 
-C++ AMP.
+This is of course a minimal example of GPU programing.
+A full example will involve several lines of code transferring data to and from the GPU, and the kernel could include manual administration of the L1 cache (shared memory), and intra-block thread barriers.
+This is some of the reasons why writing a high performance GPU kernel is too often found to be a complex and time consuming task.
 
-Architecture and how to program. See Paper II.
-
-It it is also much less costly to switch between a pixel-thread for the GPU then a program thread for the CPU.
-
-Hopefully, in the future, GPU programming will be the job of e.g.
-c++ compiler, and direct fiddling with GPU kernels should only be needed if maximum throughput is absolutely needed.
-(In the same way the compiler is able to translate c++ code into SSE and AVX instructions.)
-
-\subsection{Accelerators in medical ultrasound imaging }
-Add sentence about siemens and supersonic imagine scanners. They use GPUs? What does this mean? FPGAs and ASICs.
+No doubt, it is still a major overhead involved when writing GPU code, and in the future we will probably see work aimed at reducing this overhead.
+A recent development is the introduction of C++ AMP which enables developers to write accelerated code (e.g.
+by a GPU) in plain C++.
+Hopefully, in the future, GPU programming will be a job for compilers, and direct fiddling with GPU kernels should only be needed if maximum throughput is absolutely needed.
+This is similar to how SIMD instructions is handled in CPU code today.
+Most of the time developers will rely on the CPU compilers to analyze the code and auto-apply SIMD instruction whenever it is possible.
+However, when maximum throughput is needed, developers with deep knowledge of the problem at hand are often able to make faster code by manually writing SIMD instruction code.
+The relevance of auto-applied GPU instructions by compilers will also increasing due to the recent embedding of GPUs on the same chip as the CPU.
+Soon, when writing code, we will probably talk about coding a heterogeneous core where some operations are automatically offloaded to special cores (e.g.
+a GPU) by the compiler.
 
@@
 
+Delay and sum. Apodization. (See intro to Paper II or III).
+Time v.s. phase delays.
+
 Key selling points of ultrasound imaging: cost, safety and real-time user interaction.
@@
 One sentence about speckle tracking.
-
-\subsection{Cardiac anatomy}
-Small section about heart anatomy.
 
@@
 
-\subsection{Basic beamforming}
+Small section about heart anatomy.
 
-Delay and sum. Apodization. (See intro to Paper II or III).
-Time v.s. phase delays.
+\subsection{Accelerators in medical ultrasound imaging}
+Ultrasound imaging systems generates an incredible amount of data.
+A modern system capable of 3D imaging will generate several tera bytes of data in the front end of the system during a clinical scan.
+The data rate from a 2D scan is "only" in order giga bytes per second.
+To handle these amount of data, ultrasound imaging systems have relied heavily on application-specific integrated circuits (\nom{ASIC}{Application-specific integrated circuit}) for all steps in the processing pipeline.
+However, as the computationally power of CPUs and GPUs have increased, ASICs, e.g.
+for scan conversion, have gradually been converted into software \cite{Guracar2013}.
+One of  the last remaining ASICs is the beamformer, whose job is to steer and focus the ultrasound beam in space.
+This operation involves a reduction of the data rate with a factor equal to the number of elements in the ultrasound probe.
+The input data rate is therefor very high.
+Nevertheless, real-time beamforming has recently been achieved on a GPU \cite{Song2012}, and a French company (SuperSonic Imagine) are utilizing GPUs in their plane wave ultrasound beamforming algorithm \cite{Tanter2014}.
+Beamforming is therefore finally moving into software to.
+The largest players in the medical ultrasound business, General Electric and Philips, are also well aware of where the trend is going \cite{Thomenius2012} \cite{Metz2011}.
+Software beamforming opens up new possibilities for advanced processing.
+One example is Capon beamforming which is explored in this thesis (\textbf{Paper\,I} and \textbf{II}), another is advanced 3D visualization as presented in \cite{solteszova2010multidirectional}, \textbf{Paper\,IV}, and \textbf{IX}.
 

### 2014-02-23 HRUIofMovingObjects/jrnl_article.tex
Changed symbol for resolution.

@@
 \begin{align}
-\Delta < \frac{\lambda_c}{qpD}, \label{eq:resolution}
+\delta\theta < \frac{\lambda_c}{qpD}, \label{eq:resolution}
 \end{align}
@@
 \begin{align}
-\theta_{max} = \alpha p \Delta,
+\theta_{max} = \alpha p \delta\theta,
 \end{align}
-where $\alpha$ is the pulse length in $\lambda$'s, and $p$ and $\Delta$ is the oversampling factor and beam spacing respectively as defined in (\ref{eq:resolution}).
-For our simulations $\theta_{max} = 3\Delta$.
+where $\alpha$ is the pulse length in $\lambda$'s, and $p$ and $\delta\theta$ is the oversampling factor and beam spacing respectively as defined in (\ref{eq:resolution}).
+For our simulations $\theta_{max} = 3\delta\theta$.
 It is therefore possible to make a sweep based on phase rotation between a given beam and half-way to its two nearest neighbours without introducing large errors.
@@
 \end{align}
-where $x_i$ is the element position and $\theta$ is swept from $-\Delta/2$ to $\Delta/2$.
+where $x_i$ is the element position and $\theta$ is swept from $-\delta\theta/2$ to $\delta\theta/2$.
 Note that the steering vector must be calculated for an $L$ element symmetric subarray, even though the full array has $M$ elements.
 This subarray should have the same element pitch, $x_i - x_{i-1}$, as the full array.
-%The steering error introduced by this steering vector for a frequency band $B$ is less than $B/2 \times \Delta/2$.
+%The steering error introduced by this steering vector for a frequency band $B$ is less than $B/2 \times \delta\theta/2$.
 

### 2014-02-23 ThesisJP/Background/background.tex
Changed symbol for resolution.

@@
 The graphics processing pipeline had before this been a fixed function pipeline, where e.g.
-geometry and textures where fed to the graphics processing unit (GPU) in one end and a rasterized rendered image was outputted in the other.
+geometry and textures where fed to the graphics processing unit (GPU) in one end, and where shaded a rasterized geometry was outputted in the other.
 What programmable shaders  introduced was the ability to replace some steps in the fixed-function pipeline with custom computations.
@@
 Another reason is the inherent parallel nature of the rendering problem.
-When geometry is projected and rasterized the same instruction is typically needed for a lot of data at the same time.
-This makes a special kind of instruction, the Single Instruction Multiple Data (\nom{SIMD}{Single instruction multiple data (instruction)}) instruction \cite{Flynn1966}, especially suited for this kind of problem.
+When geometry is shaded, projected and rasterized the same instruction is typically needed for a lot of data at the same time.
+This makes a special kind of instruction, known as Single Instruction Multiple Data (\nom{SIMD}{Single instruction multiple data (instruction)}) \cite{Flynn1966}, especially suited for this kind of problem.
 The differences between how SIMD instructions are utilized in modern CPUs and GPUs are discussed in Section \ref{sec:cpu_vs_gpu}.
@@
 AVX and AVX2 adds 256 bit long registers and fused multiply-accumulate instructions (\nom{FMA}{Fused multiply-accumulate (instruction)})  which means that sixteen 32 bit floating point values can be processed per clock cycle per core.
-A new extension, AVX-512, extends the registers to 512 bit and is found in the Intel accelerator board  Xeon Phi and might show up in the next generation of CPUs.
+A new extension, AVX-512, extends the registers to 512 bit and is found in the Intel accelerator board  Xeon Phi and might show up in the next generation of CPUs from Intel.
 The latest CPU architecture from Intel, the Haswell architecture, has two processing ports supporting FMA-AVX2 instructions, and is therefore capable of processing 32, 32 bit, floating point values per clock cycle per core (Fig.\,\ref{fig:cpu_gpu}).
@@
 This is the same number as visualized in Fig\,\ref{fig:cpu_vs_gpu} for the Haswell architecture.
-In addition, the most powerful integrated GPU found on Haswell CPUs have 40 execution units capable of processing sixteen, 32 bit, floating points at 1.1 GHz.
+In addition, the most powerful integrated GPU found on Haswell CPUs (HD 5200) have 40 execution units capable of processing sixteen, 32 bit, floating points at 1.3 GHz.
 This adds
 \begin{equation}
-16*40*1.1 = 704\,\text{GFLOPS}
+16*40*1.3 = 832\,\text{GFLOPS}
 \end{equation}
@@
 This adds up to a total of five TFLOPS in computationally power as shown in Fig.\,\ref{fig:cpu_vs_gpu} for the GeForce GTX 780 TI GPU.
-From Fig.\,\ref{fig:cpu_gpu} we see how the high number of SIMD cores found on GPUs and their larger width results in five times the computational performance of a high-end CPU (without the integrated GPU).
+From Fig.\,\ref{fig:cpu_gpu} we see how the high number of SIMD cores found on GPUs and their larger width, compared with SIMD cores on the CPU, results in five times the computational performance of a high-end CPU (without the integrated GPU).
 
 Note that the derived numbers are strictly theoretical and that actual throughput is highly algorithm dependent.
-Nevertheless, in theory, the benefit of GPU computing over CPU computing is currently a factor of five or ten\footnote{Until an eight core Haswell CPU is launched later this year.} in raw computationally power.
-A claimed speed-up larger than this is typically caused by comparing a non-optimized CPU implementation with an optimized GPU implementation or by comparing uneven hardware \cite{Lee2010, Kothapalli2013}.
+Nevertheless, in theory, the benefit of GPU computing over CPU computing is currently a factor of five in raw computationally power.
+A claimed speed-up of several order of magnitude is typically caused by comparing a non-optimized CPU implementation with an optimized GPU implementation or by comparing uneven hardware \cite{Lee2010, Kothapalli2013}.
 Also note that in this section we have focused on CPUs from Intel and GPUs from Nvidia, however, CPUs and GPUs from AMD have similar specifications.
@@
 The first version of CUDA was released in 2007.
-The following year, Apple launched a GPGPU framework called \nom{OpenCL}{Open Compute Language (GPGPU language)}, a standard which are maintained by the Khronos Group.
+The following year, Apple launched a GPGPU framework known as \nom{OpenCL}{Open Compute Language (GPGPU language)}, a standard which are maintained by the Khronos Group.
 These two frameworks are still today the main programming frameworks for GPGPU.
@@
 
-No doubt, it is still a major overhead involved when writing GPU code, and in the future we will probably see work aimed at reducing this overhead.
+No doubt, it is still a major overhead involved with writing GPU code, and in the future we will probably see work aimed at reducing this overhead.
 A recent development is the introduction of C++ AMP which enables developers to write accelerated code (e.g.
@@
 The relevance of auto-applied GPU instructions by compilers will also increasing due to the recent embedding of GPUs on the same chip as the CPU.
-Soon, when writing code, we will probably talk about coding a heterogeneous core where some operations are automatically offloaded to special cores (e.g.
+Soon, when writing code, we will probably talk about coding for a collection of heterogeneous cores where operations are automatically offloaded to special cores (e.g.
 a GPU) by the compiler.
@@
 This is the reason why 2-4 MHz is used for cardiac imaging and a frequency around 7 MHz is used for vascular imaging.
+A piezoelectric ceramic is used to generate the ultrasonic frequencies, and is special in that way that ultrasound vibrations are generated when voltage is applied and \textit{vice versa}.
+It can therefore be used to both send and receive ultrasound signals, like a combined microphone and speaker.
+Piezoelectric elements organized in an array, is referred to as an ultrasound \textit{probe}, and a special probe is typically design for each medical imaging modality.
+For cardiac ultrasound imaging the probe footprint has to be smaller than the intercostal opening in order to avoid shadowing by ribs.
+
+If the width of a piezoelectric element is less than the wavelength of the transmitted signal the element is said to be \textit{omnidirectional}, hence energy is distributed equally in all directions.
+When multiple omnidirectional elements are arranged in an array the transmitted signal will get shaped into a beam of ultrasound by diffraction.
+The 3 dB angular width of this beam, and also the angular resolution of the array, is given by \cite{AngelUltrasound}:
+\begin{align}
+\Delta = \frac{\lambda}{D},
+\end{align}
+\nomi{$\Delta$}{Angular resolution} where \nom{$\lambda$}{Wavelength} is the ultrasound wavelength and \nom{$D$}{Aperture width} is the aperture width.
+With a speed-of-sound in human tissue of $1540\frac{\text{m}}{\text{s}}$ the angular resolution of a 2 cm cardiac array at 3 MHz is around 1.5\degree, which at 8 cm gives a lateral resolution of 2 mm.
+The axial resolution is inverse proportional to the pulse length, hence the frequency bandwidth of the probe.
+
+If time delays are applied on the signal sent to the different elements it is possible to control the steering and focusing of the beam.
+This technique is know as \textit{beamforming}, and for ultrasound imaging it is applied on both transmit and receive.
+
+The standard form of time domain array beamforming is the delay-and-sum beamformer where the output $z$ is calculated as:
+\begin{align}\label{eq:das}
+z[n] = \sum_{m = 0}^{M-1}\vec{w}_m^*\vec{x}_m[n - \Delta_m[n]] = \vec{w}^H\vec{x}[n],
+\end{align}
+where $M$ is the number of elements or channels in the array, $\Delta_m[n]$ is the per-element focusing and steering delay, and $\vec{w}$ is a weight vector or window function.
+The weight vector is typically selected in order to trade resolution for a lower side lobe level.
+Hence, the weights are typically real and their K-space response is then symmetric.
 
@@
 Sampling rates, and data size.
-%Due to the high bandwidth requirement for medical ultrasound imaging, as we will see in Section \ref{sec:ultrasound}, a cloud computing ultrasound imaging system will not feasibly for a long time.
-Because of the high data rate and processing requirements ultrasound systems where first analog systems up to the mid 90s. The first digital systems where made with custom boards and ASICs.
 

### 2014-02-23 ThesisJP/Introduction/introduction.tex
Changed symbol for resolution.

@@
 A rapid development in computer game technology and accompanying programming languages have recently provided researchers with small personal super computers, comprised in a single graphics card (\nom{GPU}{Graphics processing unit}).
-The latest graphics cards\footnote{As of January 2014.} from both \nom{Nvidia}{Graphics card producer} and \nom{AMD}{Graphics card producer} provide close to 6 T\nom{FLOPS}{Floating point operations per second} single precision and 1.5 TFLOPS double precision computations.
+The latest graphics cards\footnote{As of January 2014.} from both \nom{Nvidia}{Graphics card producer} and \nom{AMD}{Graphics card producer} provide more than 5 T\nom{FLOPS}{Floating point operations per second} single precision and 1.5 TFLOPS double precision computations.
 For double precision this is more than the worlds most powerful supercomputer could provide in 1998.

### 2014-02-23 ThesisJP/Background/background.tex
Removed vector notation.

@@
 Piezoelectric elements organized in an array, is referred to as an ultrasound \textit{probe}, and a special probe is typically design for each medical imaging modality.
-For cardiac ultrasound imaging the probe footprint has to be smaller than the intercostal opening in order to avoid shadowing by ribs.
+For cardiac ultrasound imaging the probe footprint has to be smaller than the intercostal window in order to avoid shadowing by ribs.
+A line in an ultrasound image is generated by firing a ultrasound pulse into the body and sample the received echo.
+This amplitude trace is known as an A-mode image.
 
 If the width of a piezoelectric element is less than the wavelength of the transmitted signal the element is said to be \textit{omnidirectional}, hence energy is distributed equally in all directions.
-When multiple omnidirectional elements are arranged in an array the transmitted signal will get shaped into a beam of ultrasound by diffraction.
+When multiple omnidirectional elements are arranged in an array the transmitted pulse will get shaped into a beam\footnote{A beam refers to the pulse integrated in time.} of ultrasound by diffraction.
 The 3 dB angular width of this beam, and also the angular resolution of the array, is given by \cite{AngelUltrasound}:
 \begin{align}
-\Delta = \frac{\lambda}{D},
+\delta\theta = \frac{\lambda}{D},
 \end{align}
-\nomi{$\Delta$}{Angular resolution} where \nom{$\lambda$}{Wavelength} is the ultrasound wavelength and \nom{$D$}{Aperture width} is the aperture width.
+\nomi{$\delta\theta$}{Angular resolution} where \nom{$\lambda$}{Wavelength} is the ultrasound wavelength, and \nom{$D$}{Aperture width} is the aperture width.
 With a speed-of-sound in human tissue of $1540\frac{\text{m}}{\text{s}}$ the angular resolution of a 2 cm cardiac array at 3 MHz is around 1.5\degree, which at 8 cm gives a lateral resolution of 2 mm.
-The axial resolution is inverse proportional to the pulse length, hence the frequency bandwidth of the probe.
+The axial resolution is inverse proportional to the pulse length, and is typically better than the lateral resolution.
 
-If time delays are applied on the signal sent to the different elements it is possible to control the steering and focusing of the beam.
-This technique is know as \textit{beamforming}, and for ultrasound imaging it is applied on both transmit and receive.
-
-The standard form of time domain array beamforming is the delay-and-sum beamformer where the output $z$ is calculated as:
+If time delays are applied on the signal sent to the different elements it is possible to steer and focus the ultrasound beam in space.
+This technique is know as \textit{beamforming}, and for ultrasound imaging it is applied both when transmitting and receiving ultrasound pulses.
+For medical ultrasound imaging there are two main types of ultrasound probes, linear and phased arrays \cite{AngelUltrasound}.
+A cardiac ultrasound array is a phased array, which means that the image is constructed from beams, or A-mode images, which are gradually steered from $-\theta/2$ to $\theta/2$, resulting in sector which are $\theta$ wide.
+This kind of image is known as a B-mode image (B for brightness).
+If \nom{$M$}{Number of array elements} is the number of elements in the array, and \nom{$\Delta_m[n]$}{Per-element focusing and steering delays} is the per-element focusing and steering delay, the beamforming technique known as \textit{delay-and-sum} is given as:
 \begin{align}\label{eq:das}
@@
 \end{align}
-where $M$ is the number of elements or channels in the array, $\Delta_m[n]$ is the per-element focusing and steering delay, and $\vec{w}$ is a weight vector or window function.
-The weight vector is typically selected in order to trade resolution for a lower side lobe level.
-Hence, the weights are typically real and their K-space response is then symmetric.
+where \nom{$\vec{w}$}{Array weight vector (apodization)} is know as the array weight vector or apodization, and $z[n]$ is an A-mode image.
+If the phase center of our array is the origin and $p$ is the point where we want to steer and focus the ultrasound beam we have to use delays given as:
+\begin{align}
+\Delta_m[n] =\frac{|x_m+p| - |p|}{c},
+\end{align}
+where $c$ is the speed of sound.
+
+%The weight vector is typically selected in order to trade resolution for a lower side lobe level. Hence, the weights are typically real and their K-space response is then symmetric.
 

### 2014-02-23 ThesisJP/Background/background.tex
Background.

@@
 This kind of image is known as a B-mode image (B for brightness).
-If \nom{$M$}{Number of array elements} is the number of elements in the array, and \nom{$\Delta_m[n]$}{Per-element focusing and steering delays} is the per-element focusing and steering delay, the beamforming technique known as \textit{delay-and-sum} is given as:
+If \nom{$M$}{Number of array elements} is the number of elements in the array, and \nom{$\Delta_m[n]$}{Per-element focusing and steering delays} is the per-element focusing and steering delay, the standard form of beamforming used on receive, known as \textit{delay-and-sum}, is given as:
 \begin{align}\label{eq:das}
-z[n] = \sum_{m = 0}^{M-1}\vec{w}_m^*\vec{x}_m[n - \Delta_m[n]] = \vec{w}^H\vec{x}[n],
+z[n] = \sum_{m = 0}^{M-1}w_m^*x_m[n - \Delta_m[n]] = \vec{w}^H\vec{x}[n],
 \end{align}
-where \nom{$\vec{w}$}{Array weight vector (apodization)} is know as the array weight vector or apodization, and $z[n]$ is an A-mode image.
-If the phase center of our array is the origin and $p$ is the point where we want to steer and focus the ultrasound beam we have to use delays given as:
+where \nom{$\vec{w}$}{Array weight vector (apodization)} is know as the array weight vector or apodization, and $z[n]$ is the output A-mode image.
+If $x_o$ is the phase center of our array and $p$ is the point where we want to steer and focus the ultrasound beam, we have to use the following delays:
 \begin{align}
-\Delta_m[n] =\frac{|x_m+p| - |p|}{c},
+\Delta_m[n] =\frac{|x_m+p| - |x_o+p|}{c},
 \end{align}
-where $c$ is the speed of sound.
+where \nom{$c$}{Speed of sound} is the speed of sound in human tissue.
+On transmit the inverse set of delays have to be used.
+Also note that on transmit we have to decide where to steer and focus prior to transmitting a pulse, transmit delays are therefore not time varying.
 
-%The weight vector is typically selected in order to trade resolution for a lower side lobe level. Hence, the weights are typically real and their K-space response is then symmetric.
-
-Typical probes, scan sequences, resolution and sampling.
+Resolution and sampling.
 
@@
 
-Delay and sum. Apodization. (See intro to Paper II or III).
 Time v.s. phase delays.

### 2014-02-24 ThesisJP/Background/background.tex
Minar change in background.

@@
 Ultrasound imaging encompasses technology which generates images based on sound whose frequencies we can not hear.
-For medical ultrasound imaging frequencies between 2 MHz and 10 MHz are typically used.
-Frequency dependent attenuation force us to select a lower frequency if the object we want to image is located deep into the body.
+For medical ultrasound imaging frequencies between 2 MHz and 10 MHz are typically used, where frequency dependent attenuation force us to select a lower frequency if the object we want to image is located deep into the body.
 This is the reason why 2-4 MHz is used for cardiac imaging and a frequency around 7 MHz is used for vascular imaging.
@@
 Piezoelectric elements organized in an array, is referred to as an ultrasound \textit{probe}, and a special probe is typically design for each medical imaging modality.
-For cardiac ultrasound imaging the probe footprint has to be smaller than the intercostal window in order to avoid shadowing by ribs.
+For cardiac ultrasound imaging the probe must not only be designed for a certain frequency range, its footprint also needs to be smaller than the intercostal window in order to avoid shadowing by ribs.
 A line in an ultrasound image is generated by firing a ultrasound pulse into the body and sample the received echo.
@@
 where \nom{$\vec{w}$}{Array weight vector (apodization)} is know as the array weight vector or apodization, and $z[n]$ is the output A-mode image.
-If $x_o$ is the phase center of our array and $p$ is the point where we want to steer and focus the ultrasound beam, we have to use the following delays:
+If $p_o$ is the phase center of our array and $p$ is the point where we want to steer and focus the ultrasound beam, we have to use the following delays:
 \begin{align}
-\Delta_m[n] =\frac{|x_m+p| - |x_o+p|}{c},
+\Delta_m[n] =\frac{|p_m+p| - |p_o+p|}{c},
 \end{align}
-where \nom{$c$}{Speed of sound} is the speed of sound in human tissue.
+where \nom{$c$}{Speed of sound} is the speed of sound in human tissue, and $p_m$ is the position of the m'th element.
 On transmit the inverse set of delays have to be used.

### 2014-02-26 HRUIofMovingObjects/response_to_reviewers.tex
Finished review.

@@
 I would suggest that in the introduction we do not need these details.
-\textit{These details have now been removed from the introduction.}
+\textit{We believe that this is crucial information in order to reproduce Fig 1.. L, M and d have now been replaced by their full name.}
 \\\\
@@
 Let us say that the oversampling is used both for Capon and for reducing folding of frequencies over the bandwidth.
-\textit{We do not find this confusing, $p$ is always 2 and $q$ is 4 and 16 when oversampling on transmit is applied for the Capon beamformer (as mention in the first paragraph of Section IV-A).
+\textit{We do not find this confusing, $p$ is always 2 and $q$ is 4 and 16 when oversampling is applied for the Capon beamformer (as mention in the first paragraph of Section IV-A).
 In all the other examples $q=1$.
-The $p$ and $q$ is there to clearly separate oversampling from the Nyquist requirement.}
+The $p$ and $q$ is there to clearly separate oversampling from the Nyquist requirement, and the different reasons for why to oversample from each other.}
 \\\\

### 2014-02-26 HRUIofMovingObjects/jrnl_article.tex
Removed vector notation.
Finished review.

@@
 Hence, the interference-noise matrix ($\sigma_0^2\mat{Q}$ in \cite{Cox1973}) is the identity.
-The effect of typical subarray averaging and diagonal loading have been added to the Capon beamformer output.
+The effect of typical subarray averaging and diagonal loading have been added to the Capon beamformer output (subarray length of 32 and a diagonal loading factor of of 0.01).
 We see that a DAS beamformer, using the required beam spacing, could yield a $1.9$ dB scalloping loss \cite{Harris1978} if the object is moving parallel to the array.
@@
 }
-\caption{Estimated drop in output power mid-way between receive beams for delay-and-sum and Capon beamforming (with typical robustification parameters) applied in a farfield-narrowband setting.
+\caption{Estimated drop in output power mid-way between receive beams for delay-and-sum and Capon beamforming (with subarray length of 32 and a diagonal loading factor of of 0.01) applied in a farfield-narrowband setting.
 %The scanned output from both beamformers is plotted for two different time instances.
@@
 \begin{align}\label{eq:das}
-z[n] = \sum_{m = 0}^{M-1}\vec{w}_m^*\vec{x}_m[n - \Delta_m[n]] = \vec{w}^H\vec{x}[n],
+z[n] = \sum_{m = 0}^{M-1}w_m^*x_m[n - \Delta_m[n]] = \vec{w}^H\vec{x}[n],
 \end{align}
@@
 Dotted vertical lines indicate position of transmit-receive beam pairs.
-The point scatterer response has been simulated using Field II \cite{Jensen1992, Jensen1996a} with a probe setup equal to the M5S-D phased array by GE Vingmed Ultrasound, Horten, Norway.
+The point scatterer response has been simulated using Field II \cite{Jensen1992, Jensen1996a} using a 96 element phased array with a 22 mm azimuth aperture.
 The probe was excited with a 2.5 MHz and 1.5 cycles long pulse.
 The transmit focus in azimuth and elevation were configured to be equal and the point scatterer was moved through this point.
+The beam density was calculated  based on (\ref{eq:resolution}) with the speed of sound equal to 1540.
 We see that the contours in Fig.\,\ref{fig:das_shiftinvariant} are approximately diagonal (low P-MAE) and that the amplitude variation among the maximum points is minimal (low A-MAE).
@@
 The data was acquired using a modified Vivid E9 high-end cardiac ultrasound scanner by GE Vingmed Ultrasound, Horten, Norway, equipped with a M5S-D phased transducer operating in harmonic mode at 3.34 MHz.
-The transmit beam density was set to match (\ref{eq:resolution}) based on this frequency, and on a 65\degree sector we therefore distributed 111 tx-beams.
+The transmit beam density was set to match (\ref{eq:resolution}) based on this frequency, and 111 tx-beams where therefore distributed on a 65\degree sector.
 Transmit focus is located at 8 cm range.

### 2014-03-03 HRUIofMovingObjects/response_to_reviewers.tex
Worked on comments from Sverre.

--- 2014-02-20 HRUIofMovingObjects/response_to_reviewers.tex
@@
 I would suggest that in the introduction we do not need these details.
-\textit{These details have now been removed from the introduction.}
+\textit{In order to reproduce Fig 1, L, M and d have to be known. We have now replaced abbreviations with full names (L in to subarray length etc.) in order to minimize the need for prior knowledge.}
 \\\\
@@
 Let us say that the oversampling is used both for Capon and for reducing folding of frequencies over the bandwidth.
-\textit{We do not find this confusing, $p$ is always 2 and $q$ is 4 and 16 when oversampling on transmit is applied for the Capon beamformer (as mention in the first paragraph of Section IV-A).
-In all the other examples $q=1$.
-The $p$ and $q$ is there to clearly separate oversampling from the Nyquist requirement.}
+\textit{We understand that this may be confusing and we have amended the text to make it more clear. The factor $q$ has been removed.}
 \\\\

### 2014-03-03 HRUIofMovingObjects/jrnl_article.tex
Worked on comments from Sverre.

@@
 Hence, the interference-noise matrix ($\sigma_0^2\mat{Q}$ in \cite{Cox1973}) is the identity.
-The effect of typical subarray averaging and diagonal loading have been added to the Capon beamformer output (subarray length of 32 and a diagonal loading factor of of 0.01).
+The effect of typical subarray averaging and diagonal loading have been added to the Capon beamformer output (subarray length of 32 and a diagonal loading factor of 0.01).
 We see that a DAS beamformer, using the required beam spacing, could yield a $1.9$ dB scalloping loss \cite{Harris1978} if the object is moving parallel to the array.
@@
 }
-\caption{Estimated drop in output power mid-way between receive beams for delay-and-sum and Capon beamforming (with subarray length of 32 and a diagonal loading factor of of 0.01) applied in a farfield-narrowband setting.
+\caption{Estimated drop in output power mid-way between receive beams for delay-and-sum and Capon beamforming (with subarray length of 32 and a diagonal loading factor of 0.01) applied in a farfield-narrowband setting.
 %The scanned output from both beamformers is plotted for two different time instances.
@@
 The required lateral sampling in an ultrasound image for a given center frequency, $f_c$, and bandwidth, $B$, is given by the aperture size, $D$, and the wavelength of the center frequency, $\lambda_c = c/f_c$, where $c$ is the speed of sound.
-The required Nyquist beam spacing is then \cite{Hergum2007}:
+The required Nyquist beam spacing is then:
 \begin{align}
-\delta\theta < \frac{\lambda_c}{qpD}, \label{eq:resolution}
+\delta\theta < \frac{\lambda_c}{2pD}, \label{eq:resolution}
 \end{align}
-where the oversampling factor $p$ is applied to reduce folding of frequencies between $f_c$ and $f_c + B/2$.
-A $p$-factor of two has been used throughout this paper.
-The oversampling factor $q$ will be used to oversample prior to Capon beamforming and is one if not stated otherwise.
+where the factor two is the effect of a two-way acquisition \cite{Hergum2009}, and where $p$ is an oversampling factor which can be applied to reduce folding of frequencies between $f_c$ and $f_c + B/2$.
+In this paper we will use the $p$-factor to perform oversampling prior to Capon beamforming.
 
@@
 A straight forward way of obtaining sufficient sampling in an active imaging system with an equal number of transmit (tx) and receive (rx) beams is to increasing the number of tx-beams until the system is locally laterally shift-invariant.
-In Fig.\,\ref{fig:capon_oversampling_a} and Fig.\,\ref{fig:capon_oversampling_b} we have increased the tx-beam density by a $q$-factor of four and sixteen respectively.
+In Fig.\,\ref{fig:capon_oversampling_a} and Fig.\,\ref{fig:capon_oversampling_b} we have increased the tx-beam density by a $p$-factor of four and sixteen respectively.
 As expected the geometric distortion and scalloping loss decreases with increased lateral sampling.
@@
 
-Figure \ref{fig:capon_oversampling_pr} shows the phase rotation method applied on the same scan sequence as in Fig\,\ref{fig:capon} with four and sixteen times oversampling on receive respectively ($q$-factor).
+Figure \ref{fig:capon_oversampling_pr} shows the phase rotation method applied on the same scan sequence as in Fig\,\ref{fig:capon} with four and sixteen times oversampling on receive respectively ($p$-factor).
 Note that the figure is zoomed compared with Fig\,\ref{fig:capon}.

### 2014-03-03 ThesisJP/Background/background.tex
Small changes in thesis background.

@@
 
-For the first adopters general-purpose computing on GPUs (GPGPU) it soon became evident that offloading all types of computations to the GPU was not always a good idea.
+For the first adopters of general-purpose computing on GPUs (GPGPU) it soon became evident that offloading all types of computations to the GPU was not always a good idea.
 In most cases it is important to balance the load equally between the CPU and GPU to obtain their combined computationally power.
@@
 
-The final transition currently taking place in the computer industry is the introduction of cloud computing, where compute power is provided as a web-service and automatically scales on demand.
+The final transition currently taking place in the computer industry is the introduction of cloud computing, where compute power is provided as a web service and automatically scales on demand.
 Together, multi-core, heterogeneous cores, and cloud cores forms the "jungle" that today's software engineers have to tackle (Fig.\,\ref{fig:jungle}).
@@
 \subsection{Comparing CPU and GPU performance}\label{sec:cpu_vs_gpu}
-The CPU is design to be the main control unit of a computer.
-It is therefore good at general processing.
+The CPU is design to be the main control unit of a computer, it is therefore good at general processing.
 The GPU, on the other hand, has a history of processing graphics which involves heavy vector arithmetic.
@@
 The first SEE instructions set had 128 bit long registers which meant that four, 32 bit, floating point values could be processed per clock cycle\footnote{For simplicity we will assume that the whole register is processed in one clock cycle even if the actual implementation might distributes one instruction for the register across multiple cycles.}.
-Updates of the SSE instruction set have added new instructions (SSE2, SSE3, and SSE4) and longer registers (\nom{AVX}{Advanced vector extensions (instruction)}, AVX2, and AVX-512).
-AVX and AVX2 adds 256 bit long registers and fused multiply-accumulate instructions (\nom{FMA}{Fused multiply-accumulate (instruction)})  which means that sixteen 32 bit floating point values can be processed per clock cycle per core.
+Updates of the SSE instruction set have added new instructions and longer registers (SSE2, SSE3, SSE4, \nom{AVX}{Advanced vector extensions (instruction)}, AVX2, and AVX-512).
+The newly introduced extensions, AVX and AVX2, adds 256 bit long registers and fused multiply-accumulate instructions (\nom{FMA}{Fused multiply-accumulate (instruction)})  which means that sixteen 32 bit floating point values can be processed per clock cycle per core.
 A new extension, AVX-512, extends the registers to 512 bit and is found in the Intel accelerator board  Xeon Phi and might show up in the next generation of CPUs from Intel.
@@
 This is the same number as visualized in Fig\,\ref{fig:cpu_vs_gpu} for the Haswell architecture.
-In addition, the most powerful integrated GPU found on Haswell CPUs (HD 5200) have 40 execution units capable of processing sixteen, 32 bit, floating points at 1.3 GHz.
+In addition, the most powerful integrated GPU found on Haswell CPUs, the HD 5200, have 40 execution units capable of processing sixteen, 32 bit, floating points at 1.3 GHz.
 This adds
@@
 Each SIMD core also comprise 16 special function units (\nom{SFU}{Special function unit (sin, exp, sqrt, ect.)}) which add close to 1 TFLOPS in additional computational power.
-This adds up to a total of five TFLOPS in computationally power as shown in Fig.\,\ref{fig:cpu_vs_gpu} for the GeForce GTX 780 TI GPU.
+This adds up to a total of five TFLOPS as shown in Fig.\,\ref{fig:cpu_vs_gpu} for the GeForce GTX 780 TI GPU.
 From Fig.\,\ref{fig:cpu_gpu} we see how the high number of SIMD cores found on GPUs and their larger width, compared with SIMD cores on the CPU, results in five times the computational performance of a high-end CPU (without the integrated GPU).
@@
 where \texttt{grid} and \texttt{block} is 3-component vectors specifying grid and block size in "xyz" as depicted in Fig.\,\ref{fig:gpu_grid}, \texttt{a} is a pointer to a memory buffer located on the GPU, and \texttt{N}$*$\texttt{N} is the buffer size.
-When launched, a copy of the kernel is scheduled  for each thread in each block, and each kernel get as input its location in the compute grid (\texttt{blockIdx} and \texttt{threadIdx}) together with the grid size (\texttt{gridDim} and \texttt{blockDim}).
-This information is then typically used to calculate which memory elements to gather from the global memory buffer \texttt{a}.
+When launched, a copy of the kernel is scheduled  for each thread in each block, and each kernel gets as input its location in the compute grid (\texttt{blockIdx} and \texttt{threadIdx}) together with the grid size (\texttt{gridDim} and \texttt{blockDim}).
+This information is then typically used to calculate which memory elements to gather from global memory, in this case the elements in \texttt{a}.
 In order to formulate a given problem as a kernel function the global memory input to each thread must not rely on the output from other threads, hence it must be possible for the GPU to scheduled threads in any order.
@@
 However, when maximum throughput is needed, developers with deep knowledge of the problem at hand are often able to make faster code by manually writing SIMD instruction code.
-The relevance of auto-applied GPU instructions by compilers will also increasing due to the recent embedding of GPUs on the same chip as the CPU.
+The relevance of auto-applied GPU instructions by compilers will also increase due to the recent embedding of GPUs on the same chip as the CPU.
 Soon, when writing code, we will probably talk about coding for a collection of heterogeneous cores where operations are automatically offloaded to special cores (e.g.
@@
 Ultrasound imaging encompasses technology which generates images based on sound whose frequencies we can not hear.
-For medical ultrasound imaging frequencies between 2 MHz and 10 MHz are typically used, where frequency dependent attenuation force us to select a lower frequency if the object we want to image is located deep into the body.
+For medical ultrasound imaging frequencies between 2 MHz and 10 MHz are typically used, where frequency dependent attenuation in human tissue force us to select a lower frequency if the object we want to image is located deep into the body.
 This is the reason why 2-4 MHz is used for cardiac imaging and a frequency around 7 MHz is used for vascular imaging.
@@
 It can therefore be used to both send and receive ultrasound signals, like a combined microphone and speaker.
-Piezoelectric elements organized in an array, is referred to as an ultrasound \textit{probe}, and a special probe is typically design for each medical imaging modality.
+Piezoelectric elements organized in an array, is referred to as an ultrasound \textit{probe}, and a special probe is typically designed for each medical imaging modality.
 For cardiac ultrasound imaging the probe must not only be designed for a certain frequency range, its footprint also needs to be smaller than the intercostal window in order to avoid shadowing by ribs.
@@
 \nomi{$\delta\theta$}{Angular resolution} where \nom{$\lambda$}{Wavelength} is the ultrasound wavelength, and \nom{$D$}{Aperture width} is the aperture width.
-With a speed-of-sound in human tissue of $1540\frac{\text{m}}{\text{s}}$ the angular resolution of a 2 cm cardiac array at 3 MHz is around 1.5\degree, which at 8 cm gives a lateral resolution of 2 mm.
+With a speed-of-sound in human tissue of $1540\frac{\text{m}}{\text{s}}$ the angular resolution of a 2 cm cardiac array at 3 MHz is around 1.5\degree, which at 8 cm gives a lateral resolution of around 2 mm.
 The axial resolution is inverse proportional to the pulse length, and is typically better than the lateral resolution.
@@
 For medical ultrasound imaging there are two main types of ultrasound probes, linear and phased arrays \cite{AngelUltrasound}.
-A cardiac ultrasound array is a phased array, which means that the image is constructed from beams, or A-mode images, which are gradually steered from $-\theta/2$ to $\theta/2$, resulting in sector which are $\theta$ wide.
+A cardiac ultrasound array is a phased array, which means that the image is constructed from beams, or A-mode images, which are gradually steered from $-\theta/2$ to $\theta/2$, resulting in an imaging sector which are $\theta$ wide.
 This kind of image is known as a B-mode image (B for brightness).
@@
 On transmit the inverse set of delays have to be used.
-Also note that on transmit we have to decide where to steer and focus prior to transmitting a pulse, transmit delays are therefore not time varying.
+Also note that on transmit we have to decide where to steer and focus prior to transmitting a pulse, transmit delays are therefore not time varying for an A-mode image.
 

### 2014-03-04 HRUIofMovingObjects/response_to_reviewers.tex
Followed up comments from reviewers. Made answers more friendly.

--- 2014-02-26 HRUIofMovingObjects/response_to_reviewers.tex
@@
 2. In section IIB, the opening sentence, there is an 's' missing in 'produce'.
-
+\\\\
 \textit{We want to thank the reviewer for an encouraging review. Typos have been corrected.}
@@
 The paper is good but I would like it to go through a second review round based on the following comments:
+\\\\
 \textit{We want to thank the reviewer for an encouraging review. Below follows our response to the comments given.}
@@
 I would suggest that in the introduction we do not need these details.
-\textit{We believe that this is crucial information in order to reproduce Fig 1.. L, M and d have now been replaced by their full name.}
+\\\\
+\textit{In order to reproduce Fig 1, L, M and d have to be known. We have now replaced abbreviations with full names (L in to subarray length etc.) in order to minimize the need for prior knowledge.}
 \\\\
@@
 Let us say that the oversampling is used both for Capon and for reducing folding of frequencies over the bandwidth.
-\textit{We do not find this confusing, $p$ is always 2 and $q$ is 4 and 16 when oversampling is applied for the Capon beamformer (as mention in the first paragraph of Section IV-A).
-In all the other examples $q=1$.
-The $p$ and $q$ is there to clearly separate oversampling from the Nyquist requirement, and the different reasons for why to oversample from each other.}
+\\\\
+\textit{We understand that this may be confusing and we have amended the text to make it more clear.
+The factor $p$ has been removed, and $q$ is now the oversampling factor.
+An additional 2-factor has been added to Eq.
+(8) with a new reference to sampling of two-way imaging systems ([21]).
+This factor of two replaces the previous use of $p$.}
 \\\\
 At what depth are the LSV plots taken with respect to the focal depth? Does it matter?
-\textit{The shift-invariance caused by under-sampling will be at its maximum at the transmit focus.
+\\\\
+\textit{Yes.
+Shift invariance caused by under sampling will be at its maximum at the transmit focus (See [22]) (the two-way beam is at its narrowest at this location).
 Simulated LSV-plots are generated from this point, as specified on page 5, first paragraph.
-In the in-vitro data, the point scatterer closest to the transmit focus is selected.}
+In the in-vitro data, the point scatterer closest to the transmit focus is selected.
+We have amended the text to clearly state why the transmit focus have been selected as the point of observation.}
 \\\\
@@
 Without this information, the work is not reproductible.
-\textit{More extensive information about the probe and scan sequence have been added.}
+\\\\
+\textit{More extensive information about the probe and scan sequence have been added to the text. Simulated results should now be possible to reproduce.}
 \\\\
 In section “IV-A. Oversampling in transmit”: please confirm (specify) that 1x1 receive beamforming is performed at this point.
-\textit{We have now specified that our starting point is a system with a one-to-one relationship between tx and rx beams.}
+\\\\
+\textit{We have now specified in the text that our starting point in Section IV-A is a system with an one-to-one relationship between tx and rx beams.}
 \\\\
@@
 Oversampling by parallel receive beamforming”, could you explain in more details what are the differences between Capon + STB and the technique proposed by Rabinovitch (2013) – reference [27]?
-\textit{There are no differences. Rabinovitch (2013), reference [25], does not propose any new technique, it is just Capon + STB.}
+\\\\
+\textit{Thank you for noticing that this point is hard to understand.
+As we see it there are actually no differences.
+Rabinovitch (2013), reference [25] in the first submission ([26] in the updated version), does not propose any new technique, it is just Capon + STB.
+The same combination of processing was performed in [18].
+The text has been amended to make this observation simpler and more clear.}
 \\\\
 Page10 Column1 line56  write “should be removed” instead of “should be remove”
+\\\\
 \textit{Misspelling has been corrected.}
@@
 It would be interesting to see such technique implemented.
-\textit{It would, but while minimizing the length of the manuscript an actual result was left out.
-It should, however, be clear from how STB corrects errors introduced by PRB, that something similar will work for phase-rotated rx-beams.}
+\\\\
+\textit{It would, but while minimizing the length of the manuscript an actual result was left out for future work. The text is amended to make this more clear.}
 \\\\
 P10 column 2 line 28-29 write “too extreme” instead of “to extreme”
+\\\\
 \textit{Misspelling has been corrected.}
@@
 6), just in case some people won’t look at the videos.
-\textit{We have now specified that the image is processed with delay-and-sum beamformer.
-Default values for $p$ and $q$ are used and are therefore not specified.
-Images are not added in order to minimize the length of the paper.}
+\\\\
+\textit{Thank you for noticing this lack of information.
+We have now specified that the image is processed with delay-and-sum beamforming.
+Images of Capon with and without oversampling have also been added to the manuscript.}
 \end{document}

### 2014-03-04 HRUIofMovingObjects/jrnl_article.tex
Followed up comments from reviewers. Made answers more friendly.

@@
 \begin{align}
-\delta\theta < \frac{\lambda_c}{2pD}, \label{eq:resolution}
+\delta\theta < \frac{\lambda_c}{2qD}, \label{eq:resolution}
 \end{align}
-where the factor two is the effect of a two-way acquisition \cite{Hergum2009}, and where $p$ is an oversampling factor which can be applied to reduce folding of frequencies between $f_c$ and $f_c + B/2$.
-In this paper we will use the $p$-factor to perform oversampling prior to Capon beamforming.
+where the factor two is the effect of a two-way acquisition \cite{Hergum2009}, and where $q$ is an oversampling factor which can be applied to reduce folding of frequencies between $f_c$ and $f_c + B/2$.
+In this paper we will use the $q$-factor to perform oversampling prior to Capon beamforming.
 
@@
 The transmit focus in azimuth and elevation were configured to be equal and the point scatterer was moved through this point.
+The transmit focus is where the beam is at its narrowest, hence shift variance effects will appear at this location first.
 The beam density was calculated  based on (\ref{eq:resolution}) with the speed of sound equal to 1540.
@@
 A straight forward way of obtaining sufficient sampling in an active imaging system with an equal number of transmit (tx) and receive (rx) beams is to increasing the number of tx-beams until the system is locally laterally shift-invariant.
-In Fig.\,\ref{fig:capon_oversampling_a} and Fig.\,\ref{fig:capon_oversampling_b} we have increased the tx-beam density by a $p$-factor of four and sixteen respectively.
+In Fig.\,\ref{fig:capon_oversampling_a} and Fig.\,\ref{fig:capon_oversampling_b} we have increased the tx-beam density by a $q$-factor of four and sixteen respectively.
 As expected the geometric distortion and scalloping loss decreases with increased lateral sampling.
@@
 The figure is zoomed compared with Fig\,\ref{fig:capon} to better see the beam-to-beam variation.
-a) Four times oversampling.
-b) Sixteen times oversampling.
+a) Four times oversampling ($q=4$).
+b) Sixteen times oversampling ($q=16$).
 The maximum scalloping loss is reduced, but it is still too high for the system to be locally laterally shift invariant.}
@@
 synthetic transmit beamforming (STB) \cite{Hergum2007, Denarie2013}.
-In \cite{Rabinovich2013} it is shown that STB works well with Capon beamforming.
-In \cite{Asen} we also used STB in order to increase the beam density of a cardiac ultrasound recording.
+In \cite{Rabinovich2013} it is shown that STB works well with Capon beamforming, something which is also verified in \cite{Asen}.
 In the following discussion we consider corrected PRB as being equal to oversampling on transmit when it comes to image quality.
@@
 \begin{align}
-\theta_{max} = \alpha p \delta\theta,
+\theta_{max} = \alpha q \delta\theta,
 \end{align}
-where $\alpha$ is the pulse length in $\lambda$'s, and $p$ and $\delta\theta$ is the oversampling factor and beam spacing respectively as defined in (\ref{eq:resolution}).
+where $\alpha$ is the pulse length in $\lambda$'s, and $q$ and $\delta\theta$ is the oversampling factor and beam spacing respectively as defined in (\ref{eq:resolution}).
 For our simulations $\theta_{max} = 3\delta\theta$.
@@
 
-Figure \ref{fig:capon_oversampling_pr} shows the phase rotation method applied on the same scan sequence as in Fig\,\ref{fig:capon} with four and sixteen times oversampling on receive respectively ($p$-factor).
+Figure \ref{fig:capon_oversampling_pr} shows the phase rotation method applied on the same scan sequence as in Fig\,\ref{fig:capon} with four and sixteen times oversampling on receive respectively ($q$-factor).
 Note that the figure is zoomed compared with Fig\,\ref{fig:capon}.
@@
 The figure is zoomed compared with Fig\,\ref{fig:capon} to better see the beam-to-beam variation.
-a) Four times oversampling.
-b) Sixteen times oversampling.}
+a) Four times oversampling ($q=4$).
+b) Sixteen times oversampling ($q=16$).}
 	\label{fig:capon_oversampling_pr}
@@
 Oversampling by phase rotation can therefore introduce similar block artifacts as parallel receive beamforming (PRB).
-These artifacts should be possible to remove in the same way as for PRB by using the synthetic transmit beamforming (STB) technique.
-If we distribute our steering angles in an overlapping scheme [like rx-beams are spread out for STB] we get a phase rotation version of STB where e.g.
+To remove these artifacts we could distribute our steering angles in an overlapping scheme [like rx-beams are spread out for STB].
+We then get a phase rotation version of STB where e.g.
 eight phase rotated beams are averaged into four.
-The geometric distortion, or stripes, seen at the bottom of the sector in the video with sixteen times oversampling \multimedia{Media Movie 3} should be removed by such technique.
+If the geometric distortion, or stripes, seen at the bottom of the sector in the video with sixteen times oversampling \multimedia{Media Movie 3} is removed by such technique is left for future work.
 

### 2014-03-04 HRUIofMovingObjects/jrnl_article.tex
Reade for second submit.

@@
 %
-\documentclass[draftcls]{IEEEtran}
-%\documentclass[journal]{IEEEtran}
+%\documentclass[draftcls]{IEEEtran}
+\documentclass[journal]{IEEEtran}
 %
@@
 
-\IEEEPARstart{I}{n} the last decade several authors have applied the Capon beamformer (also known as the minimum variance beamformer) for medical ultrasound imaging \cite{Synnevag2007, Vignon2008, Viola}.
+\IEEEPARstart{I}{n} the last decade several authors have applied the Capon beamformer [also known as the minimum variance beamformer] for medical ultrasound imaging \cite{Synnevag2007, Vignon2008, Viola}.
 The method has been shown to improve lateral resolution and edge sharpness in ultrasound images \cite{Synnevag2007, Synnevag2009, Chen2011}.
@@
 The maximum for each beam profile is also marked with a black dot, and the amplitude variation among these points is presented in a subplot next to the LSV-plot.
-The LSV-plots based on simulations has been up-interpolated to 128 samples in each direction using cubic interpolation if the number of receive beams is less than 128.
+The LSV-plots based on simulations have been up-interpolated to 128 samples in each direction using cubic interpolation if the number of receive beams is less than 128.
 
@@
 Dotted vertical lines indicate position of transmit-receive beam pairs.
-The point scatterer response has been simulated using Field II \cite{Jensen1992, Jensen1996a} using a 96 element phased array with a 22 mm azimuth aperture.
-The probe was excited with a 2.5 MHz and 1.5 cycles long pulse.
-The transmit focus in azimuth and elevation were configured to be equal and the point scatterer was moved through this point.
-The transmit focus is where the beam is at its narrowest, hence shift variance effects will appear at this location first.
+The point scatterer response has been simulated using Field II \cite{Jensen1992, Jensen1996a} using a 96 element phased array with a 22 mm azimuth aperture, which was excited with a 2.5 MHz and 1.5 cycles long pulse.
+Transmit focus in azimuth and elevation were configured to be equal and the point scatterer was moved through this point.
+The transmit focus is where the ultrasound beam is at its narrowest, hence shift variance effects will appear at this location first.
 The beam density was calculated  based on (\ref{eq:resolution}) with the speed of sound equal to 1540.
@@
 We see how the signal cancellation caused by the difference between the signal propagation vector and the assumed steering vector increases as the point moves away from a given transmit-receive beam.
-As a consequence the scalloping loss is tremendous, and not far from the predicted value in Fig\,\ref{fig:das_capon_beams}.
+As a consequence the scalloping loss is tremendous, and not far from the predicted value in Fig.\,\ref{fig:das_capon_beams}.
 %The difference between the predicted and the simulated loss might be caused by several factors.
@@
 Even though high frequency variations are reduced, the maximum is still larger than 1 dB.
-An even higher level of oversampling is therefore required in order to have a completely local lateral shift-invariant Capon beamformer when a moving point scatterer is simulated without noise and with the same parameters as in Fig\,\ref{fig:capon}.
+An even higher level of oversampling is therefore required in order to have a completely local lateral shift-invariant Capon beamformer when a moving point scatterer is simulated without noise and with the same parameters as in Fig.\,\ref{fig:capon}.
 %$1/100$ in diagonal loading.
@@
 }
-\caption{Capon beamforming as described in Fig\,\ref{fig:capon} with oversampling on transmit.
-The figure is zoomed compared with Fig\,\ref{fig:capon} to better see the beam-to-beam variation.
+\caption{Capon beamforming as described in Fig.\,\ref{fig:capon} with oversampling on transmit.
+The figure is zoomed compared with Fig.\,\ref{fig:capon} to better see the beam-to-beam variation.
 a) Four times oversampling ($q=4$).
@@
 \begin{align}
-\theta_{max} = \alpha q \delta\theta,
+\theta_{max} = 2\alpha \delta\theta,
 \end{align}
-where $\alpha$ is the pulse length in $\lambda$'s, and $q$ and $\delta\theta$ is the oversampling factor and beam spacing respectively as defined in (\ref{eq:resolution}).
+where $\alpha$ is the pulse length in $\lambda$'s, and $2$ and $\delta\theta$ is the oversampling factor and beam spacing from (\ref{eq:resolution}).
 For our simulations $\theta_{max} = 3\delta\theta$.
@@
 Note that the steering vector must be calculated for an $L$ element symmetric subarray, even though the full array has $M$ elements.
-This subarray should have the same element pitch, $x_i - x_{i-1}$, as the full array.
+This subarray should have the same element pitch, $|x_i - x_{i-1}|$, as the full array.
 %The steering error introduced by this steering vector for a frequency band $B$ is less than $B/2 \times \delta\theta/2$.
@@
 
-Figure \ref{fig:capon_oversampling_pr} shows the phase rotation method applied on the same scan sequence as in Fig\,\ref{fig:capon} with four and sixteen times oversampling on receive respectively ($q$-factor).
-Note that the figure is zoomed compared with Fig\,\ref{fig:capon}.
+Figure \ref{fig:capon_oversampling_pr} shows the phase rotation method applied on the same scan sequence as in Fig.\,\ref{fig:capon} with four and sixteen times oversampling on receive respectively ($q$-factor).
+Note that the figure is zoomed compared with Fig.\,\ref{fig:capon}.
 As with oversampling on transmit there is some variation when the point moves in and out of focus.
@@
 The extreme local variations are however reduced significantly.
-Comparing Fig\,\ref{fig:capon_oversampling} with Fig\,\ref{fig:capon_oversampling_pr}, the errors introduced by the phase rotation method are best seen in sidelobes.
+Comparing Fig.\,\ref{fig:capon_oversampling} with Fig.\,\ref{fig:capon_oversampling_pr}, the errors introduced by the phase rotation method are best seen in sidelobes.
 
@@
 	}
-	\caption{Capon beamforming as described in Fig\,\ref{fig:capon} with oversampling based on phase rotation.
-The figure is zoomed compared with Fig\,\ref{fig:capon} to better see the beam-to-beam variation.
+	\caption{Capon beamforming as described in Fig.\,\ref{fig:capon} with oversampling based on phase rotation.
+The figure is zoomed compared with Fig.\,\ref{fig:capon} to better see the beam-to-beam variation.
 a) Four times oversampling ($q=4$).
@@
 A tissue mimicking phantom containing wire targets and anechoic cysts was imaged, and the data was acquired while the probe was moved across the phantom manually.
-Fig.\,\ref{fig:image_das} shows the first frame out of 14 in the acquired loop processed with DAS beamforming.
+Fig.\,\ref{fig:phantom_das_image} shows the first frame out of 14 in the acquired loop processed with DAS beamforming.
 %Movies of the whole loop processed with DAS and Capon beamforming with and without ten times oversampling by phase rotation can be found in the supplementary material \multimedia{Videos}.
 
-One point scatterer in Fig\,\ref{fig:image_das} has been selected [white box], and then used to generate LSV-plots.
+One point scatterer in Fig.\,\ref{fig:phantom_das_image} has been selected [white box], and then used to generate LSV-plots.
 The point scatterer is located close to the transmit focus.
-Figure \ref{fig:phantom_das} shows the resulting LSV-plot when using the DAS beamformer.
+Figure \ref{fig:phantom_das_LSV} shows the resulting LSV-plot when using the DAS beamformer.
 The scalloping loss is less than 1 dB, and the system is by that locally laterally shift-invariant.
@@
 
-Figure \ref{fig:phantom_capon_LSV_10PR} presents the result of applying sixteen times oversampling by phase rotation.
+Figure \ref{fig:phantom_capon_LSV_16PR} presents the result of applying sixteen times oversampling by phase rotation.
 We observe how the amplitude of the point scatterer is better preserved in all frames, and the scalloping loss is not far from our 1 dB threshold.
@@
 Looking at the video of Capon beamforming with sixteen times oversampling by phase rotation \multimedia{Media Movie 3} the blinking is now highly reduced.
-%The system is however not fully shift invariant.
+The same can be seen from Fig.\,\ref{fig:img_invitro} where point scatterers appear brighter in the image with oversampling by phase rotation.%The system is however not fully shift invariant.
 To achieve this a higher level of oversampling, %We therefore conclude that the system is now locally laterally shift invariant.
@@
 \begin{tabular}{cc}
-\subfloat[]{\includegraphics[width=0.5\linewidth]{\img fig6a.eps}\label{fig:image_das}}
-\subfloat[]{\includegraphics[width=0.5\linewidth]{\img fig6b.eps}\label{fig:phantom_das}}\\
-\subfloat[]{\includegraphics[width=0.5\linewidth]{\img fig6c.eps}\label{fig:phantom_capon_LSV}}
-\subfloat[]{\includegraphics[width=0.5\linewidth]{\img fig6d.eps}\label{fig:phantom_capon_LSV_10PR}}
+\subfloat[]{\includegraphics[width=0.45\linewidth]{\img fig6a.png}\label{fig:phantom_das_image}}
+\subfloat[]{\includegraphics[width=0.45\linewidth]{\img fig6b.eps}\label{fig:phantom_das_LSV}}\\
+\subfloat[]{\includegraphics[width=0.45\linewidth]{\img fig6c.eps}\label{fig:phantom_capon_LSV}}
+\subfloat[]{\includegraphics[width=0.45\linewidth]{\img fig6d.eps}\label{fig:phantom_capon_LSV_16PR}}
 \end{tabular}
@@
 \caption{LSV-plots generated from \textit{in-vitro} phantom data where the probe has been manually dragged across the phantom.
-See the supplementary material for videos of the whole loop processed with the different beamformers \multimedia{Videos}.
+See the supplementary material for videos of the whole loop processed with the different beamformers.
 Capon beamforming has been performed with the same settings as in Fig.\,\ref{fig:capon1}.
 P-MAE is not shown due to a lack of probe movement control.
-a) First frame in the recorded loop processed with delay-and-sum beamforming.
+a) First frame in the recorded loop processed with delay-and-sum beamforming  \multimedia{Media Movie 1}.
 A white box indicates the point scatterer used to construct the LSV-plots.
-b) Delay-and-sum beamforming.
-c) Capon beamforming without oversampling.
-d) Capon with sixteen times oversampling using phase rotation.}
+b) LSV-plot for delay-and-sum beamforming.
+c) LSV-plot for Capon beamforming without oversampling.
+d) LSV-plot for Capon beamforming with sixteen times oversampling using phase rotation.}
 \label{fig:LSV_invitro}
+\end{figure*}
+
+\begin{figure*}[!t]
+\centerline{
+\subfloat[]{\includegraphics[width=0.48\linewidth]{\img fig7a.png}\label{fig:phantom_capon_image}}
+\subfloat[]{\includegraphics[width=0.48\linewidth]{\img fig7b.png}\label{fig:phantom_capon_16PR_image}}
+}
+\caption{First frame from Fig.\,\ref{fig:phantom_das_image} processed with Capon beamforming.
+a) Without oversampling \multimedia{Media Movie 2}.
+b) With sixteen times oversampling using phase rotation \multimedia{Media Movie 3}.
+Notice how the point scatterers appear brighter in the image with oversampling.}
+\label{fig:img_invitro}
 \end{figure*}
@@
 In his seminal paper, Cox \cite{Cox1973} derives the required beam density for a given output SNR.
-For instance 47 dB requires a beam density for the Capon beamformer ten times the one given by (\ref{eq:resolution}) [with $qp=1$].
+For instance 47 dB requires a beam density for the Capon beamformer ten times the one given by (\ref{eq:resolution}) [with $q=1/2$].
 Since resolution for the Capon beamformer depends on SNR we would recommend to first use the dynamic range in the image as guidance.
 The amount of diagonal loading will also dictate the maximal achievable resolution since $10\log{d}$ can be interpreted as the maximal input SNR seen by the Capon beamformer.
-Fig\,\ref{fig:das_capon_beams} shows that 25 times oversampling is required for a $d$-value of $1/100$.
+Fig.\,\ref{fig:das_capon_beams} shows that 25 times oversampling is required for a $d$-value of $1/100$.
 When \textit{in-vitro} data contains additional noise this oversampling requirement is reduced.%Cox measures resolution as the 3 dB saddle point between two impinging monochromatic plane waves.
@@
 Oversampling by phase rotation can therefore introduce similar block artifacts as parallel receive beamforming (PRB).
-To remove these artifacts we could distribute our steering angles in an overlapping scheme [like rx-beams are spread out for STB].
+To remove these artifacts we could distribute our steering vectors in an overlapping scheme [like rx-beams are spread out for STB].
 We then get a phase rotation version of STB where e.g.
@@
 
-%\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{photos/jon_petter2.eps}}]{Jon Petter \AA{}sen}
-%(S12) was born in Porsgrunn, Norway in 1986.
+\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{photos/jon_petter2.eps}}]{Jon Petter \AA{}sen}
+(S12) was born in Porsgrunn, Norway in 1986.
 He received the B.Sc.
@@
 
-%He has industry experience from GE Vingmed Ultrasound, Horten, Norway, were he completed three summer internships from 2008 to 2010.
+He has industry experience from GE Vingmed Ultrasound, Horten, Norway, were he completed three summer internships from 2008 to 2010.
 His research interests include image and signal processing, adaptive ultrasound processing techniques and acceleration of ultrasound algorithms using Graphics Processing Units (GPUs).
-%\end{IEEEbiography}
-
-%\vfill{}
+\end{IEEEbiography}
+
+\vfill{}
 
@@
 
-%\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{photos/andreas.eps}}]{Andreas Austeng}
-%was born in Oslo, Norway, in 1970.
+\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{photos/andreas.eps}}]{Andreas Austeng}
+was born in Oslo, Norway, in 1970.
 He received the M.S.
@@
 His research interests include signal and array processing for acoustical imaging.
-%\end{IEEEbiography}
-
-%\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{photos/sverre.eps}}]{Sverre Holm}
-%(M’82-SM’02) was born in Oslo, Norway, in 1954.
+\end{IEEEbiography}
+
+\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{photos/sverre.eps}}]{Sverre Holm}
+(M’82-SM’02) was born in Oslo, Norway, in 1954.
 He received the M.S.
@@
 His research interests are medical ultrasound imaging, modeling of tissue-ultrasound interaction, and ultrasound positioning.
-%\end{IEEEbiography}
-
-%\vfill
+\end{IEEEbiography}
+
+\vfill
 

### 2014-03-04 HRUIofMovingObjects/jrnl_article.tex
Changed back to draft style.

@@
 %
-%\documentclass[draftcls]{IEEEtran}
-\documentclass[journal]{IEEEtran}
+\documentclass[draftcls]{IEEEtran}
+%\documentclass[journal]{IEEEtran}
 %
@@
 \begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{photos/sverre.eps}}]{Sverre Holm}
-(M’82-SM’02) was born in Oslo, Norway, in 1954.
+(M82-SM02) was born in Oslo, Norway, in 1954.
 He received the M.S.

### 2014-03-05 HRUIofMovingObjects/jrnl_article.tex
Small final adjustments.

@@
 Hence, the interference-noise matrix ($\sigma_0^2\mat{Q}$ in \cite{Cox1973}) is the identity.
-The effect of typical subarray averaging and diagonal loading have been added to the Capon beamformer output (subarray length of 32 and a diagonal loading factor of 0.01).
+The effect of typical subarray averaging and diagonal loading have been added to the Capon beamformer output [subarray length of 32 and a diagonal loading factor of 0.01].
 We see that a DAS beamformer, using the required beam spacing, could yield a $1.9$ dB scalloping loss \cite{Harris1978} if the object is moving parallel to the array.
@@
 }
-\caption{Estimated drop in output power mid-way between receive beams for delay-and-sum and Capon beamforming (with subarray length of 32 and a diagonal loading factor of 0.01) applied in a farfield-narrowband setting.
+\caption{Estimated drop in output power mid-way between receive beams for delay-and-sum and Capon beamforming [with subarray length of 32 and a diagonal loading factor of 0.01] applied in a farfield-narrowband setting.
 %The scanned output from both beamformers is plotted for two different time instances.
@@
 The transmit focus is where the ultrasound beam is at its narrowest, hence shift variance effects will appear at this location first.
-The beam density was calculated  based on (\ref{eq:resolution}) with the speed of sound equal to 1540.
+The beam density was calculated  based on (\ref{eq:resolution}) with a speed of sound equal to 1540.
 We see that the contours in Fig.\,\ref{fig:das_shiftinvariant} are approximately diagonal (low P-MAE) and that the amplitude variation among the maximum points is minimal (low A-MAE).
@@
 
-Figure \ref{fig:capon_oversampling_pr} shows the phase rotation method applied on the same scan sequence as in Fig.\,\ref{fig:capon} with four and sixteen times oversampling on receive respectively ($q$-factor).
+Figure \ref{fig:capon_oversampling_pr} shows the phase rotation method applied on the same scan sequence as in Fig.\,\ref{fig:capon} with four and sixteen times oversampling ($q$-factor) on receive respectively.
 Note that the figure is zoomed compared with Fig.\,\ref{fig:capon}.
@@
 The figure is zoomed compared with Fig.\,\ref{fig:capon} to better see the beam-to-beam variation.
-a) Four times oversampling ($q=4$).
-b) Sixteen times oversampling ($q=16$).}
+a) Four times oversampling on receive ($q=4$).
+b) Sixteen times oversampling on receive ($q=16$).}
 	\label{fig:capon_oversampling_pr}

### 2014-03-08 ThesisJP/Background/background.tex
Background. GPU programming.

@@
 In this chapter the background information needed in order to understand the topics of this thesis and its papers is given.
-First, in Section \ref{sec:gpgpu}, general-purpose computing on graphics processing units (\nom{GPGPU}{General-purpose computing on graphics processing units}) is reviewed, followed by an introduction to medical ultrasound imaging in Section \ref{sec:ultrasound}.
+First, in Section \ref{sec:gpgpu}, the history and principle of general-purpose computing on graphics processing units (\nom{GPGPU}{General-purpose computing on graphics processing units}) is reviewed, followed by an introduction to medical ultrasound imaging and adaptive beamforming in Section \ref{sec:ultrasound}.
 Finally, the concepts of volume rendering and ultrasound field simulations are presented in Section \ref{sec:volren} and \ref{sec:field}.
@@
 The essay bears the name "Welcome to the Jungle" and is a sequel to the essay "The Free Lunch is Over" from 2004 \cite{HerbSuttera}.
-The two essays spins around the fact that manufactures of central processing units (CPU) hit a frequency-wall in the beginning of the 21st century.
+The two essays spins around the fact that manufactures of central processing units (CPUs) hit a frequency-wall in the beginning of the 21st century.
 Until then, all mainstream computer programs where typically running in a single thread on a single CPU \textit{processing core}, and a programmer could expect the software to annually gain performance without touching the code.
@@
 Actually, at the time when CPU manufacturers hit the frequency wall the GPU had already gone multi-core, first of all driven by an ever-increasing demand for more realistic computer games.
-Second, the rendering pipeline exhibits minimal execution path branching, and data are typically used once.
+Secondly, the rendering pipeline exhibits minimal execution path branching, and data are typically used once.
 Designers of GPUs could therefore skip advance feature, found in most CPUs, like branch prediction and different levels of data caching.
@@
 Balancing typically means that highly parallel and computationally intense tasks should be offloaded to the GPU and serialized and memory intensive work should stay on the CPU.
-This paradigm of utilizing specialized cores for solving specific problems is today known as heterogeneous computing.
+This paradigm of utilizing specialized cores for solving specific problems is today known as \textit{heterogeneous computing}.
 The specialized core can be a GPU or e.g.
@@
 This heterogeneous processors is often referred to as an accelerated processing unit (\nom{APU}{Accelerated processing unit}).
-The last three generations of CPUs from Intel have all had on-chip integrated graphics.
-Even if an on-chip GPU is less powerful than a high-performance discrete GPU, valuable time is saved by not having to send data across the PCI Express bus, which connects the different parts of a computer to the motherboard.
+The last three generations of CPUs from Intel have all had on-chip integrated graphics, hence they can be classified as APUs.
+An on-chip GPU is less powerful than a high-performance discrete GPU, but valuable time is saved by not having to send data across the PCI Express bus, which connects the different parts of a computer to the motherboard.
 This is the reason why one of the first rules of GPGPU (with discrete GPUs) is to minimize CPU-to-GPU memory transfers.
@@
 In Section \ref{sec:gpgpu} we introduced a special type of instructions, namely SIMD instructions, which are designed to execute a single instruction across multiple data element in one clock cycle.
-At the end of the 20th century, Intel introduced a SIMD instruction set for their Pentium III processor called the Streaming SIMD Extensions (\nom{SSE}{Streaming SIMD extensions (instruction)}).
-This extension has seen several updates since.
+At the end of the 20th century, Intel introduced a SIMD instruction set for their Pentium III processor called the Streaming SIMD Extensions (\nom{SSE}{Streaming SIMD extensions (instruction)}), which has seen several updates since.
 The first SEE instructions set had 128 bit long registers which meant that four, 32 bit, floating point values could be processed per clock cycle\footnote{For simplicity we will assume that the whole register is processed in one clock cycle even if the actual implementation might distributes one instruction for the register across multiple cycles.}.
 Updates of the SSE instruction set have added new instructions and longer registers (SSE2, SSE3, SSE4, \nom{AVX}{Advanced vector extensions (instruction)}, AVX2, and AVX-512).
-The newly introduced extensions, AVX and AVX2, adds 256 bit long registers and fused multiply-accumulate instructions (\nom{FMA}{Fused multiply-accumulate (instruction)})  which means that sixteen 32 bit floating point values can be processed per clock cycle per core.
+The newly introduced extensions, AVX and AVX2, adds 256 bit long registers and fused multiply-accumulate instructions (\nom{FMA}{Fused multiply-accumulate (instruction)})  which means that sixteen 32 bit floating point operations can be performed per clock cycle per core.
 A new extension, AVX-512, extends the registers to 512 bit and is found in the Intel accelerator board  Xeon Phi and might show up in the next generation of CPUs from Intel.
-The latest CPU architecture from Intel, the Haswell architecture, has two processing ports supporting FMA-AVX2 instructions, and is therefore capable of processing 32, 32 bit, floating point values per clock cycle per core (Fig.\,\ref{fig:cpu_gpu}).
+The latest CPU architecture from Intel, the Haswell architecture, has two processing ports per core supporting FMA-AVX2 instructions, and is therefore capable of performing 32, 32 bit, floating point operations per clock cycle per core (Fig.\,\ref{fig:cpu_gpu}).
 For an eight core\footnote{An 8-core Haswell CPU is scheduled to launch during 2014.} CPU with a clock frequency of 3.8 GHz this sums to
@@
 hence a theoretical peak performance of close to one tera single precision floating point operations per second (TFLOPS).
-This is the same number as visualized in Fig\,\ref{fig:cpu_vs_gpu} for the Haswell architecture.
-In addition, the most powerful integrated GPU found on Haswell CPUs, the HD 5200, have 40 execution units capable of processing sixteen, 32 bit, floating points at 1.3 GHz.
+This is the same number as visualized in Fig\,\ref{fig:cpu_vs_gpu} for the Haswell architecture (Single precision).
+In addition, the most powerful integrated GPU found on Haswell CPUs, the HD 5200, has 40 execution units capable of processing sixteen, 32 bit, floating points at 1.3 GHz.
 This adds
@@
 This adds up to a total of five TFLOPS as shown in Fig.\,\ref{fig:cpu_vs_gpu} for the GeForce GTX 780 TI GPU.
-From Fig.\,\ref{fig:cpu_gpu} we see how the high number of SIMD cores found on GPUs and their larger width, compared with SIMD cores on the CPU, results in five times the computational performance of a high-end CPU (without the integrated GPU).
+From Fig.\,\ref{fig:cpu_gpu} we see how the high number of SIMD cores found on GPUs and their larger width, compared with the SIMD cores on CPUs, results in five times the computational performance of a high-end CPU (without the integrated GPU).
 
@@
 Nevertheless, in theory, the benefit of GPU computing over CPU computing is currently a factor of five in raw computationally power.
-A claimed speed-up of several order of magnitude is typically caused by comparing a non-optimized CPU implementation with an optimized GPU implementation or by comparing uneven hardware \cite{Lee2010, Kothapalli2013}.
-Also note that in this section we have focused on CPUs from Intel and GPUs from Nvidia, however, CPUs and GPUs from AMD have similar specifications.
+A claimed speed-up of several order of magnitude is typically a result of comparing a non-optimized CPU implementation with an optimized GPU implementation or by comparing uneven hardware \cite{Lee2010, Kothapalli2013}.
+In this section we have focused on CPUs from Intel and GPUs from Nvidia, however, CPUs and GPUs from AMD have similar specifications and differences in theoretical peak performance.
 
@@
 \subsection{Programming a GPU}
-Early GPGPU programming required expert knowledge about computer graphics and the remaining fixed functions in the pipeline.
+Early GPGPU programming required expert knowledge about computer graphics and the remaining fixed functions in the graphics pipeline.
 When GPGPU gained interest it quickly became evident that new programing languages where needed.
@@
 One thread runs a copy of a kernel function.
-In this example the kernel function performs an element-wise matrix square.}
+In this example the kernel function performs an in-place element-wise matrix square.}
 \label{fig:gpu_grid_kernel}
@@
 This information is then typically used to calculate which memory elements to gather from global memory, in this case the elements in \texttt{a}.
-In order to formulate a given problem as a kernel function the global memory input to each thread must not rely on the output from other threads, hence it must be possible for the GPU to scheduled threads in any order.
+%Behind the scene, thread blocks are divided into smaller sets of threads called a warp (or wavefront in OpenCL).
+When a kernel is executed, multiple thread blocks are scheduled to each SIMD core where they stay until all instructions in the kernel is finished for all threads in the block.
+Thus,  in order to formulate a given problem as a kernel function the global memory input to each thread must not rely on the output from other blocks, hence it must be possible for the GPU to scheduled blocks in any order.
+As we see, a thread has a different conceptual meaning on a GPU than on a CPU.
+Where a CPU is said to execute threads which can perform SIMD instructions, a thread on a GPU can be interpreted as stream capable of processing one data element in the SIMD instruction.
 
-This is of course a minimal example of GPU programing.
+The code in Fig.\,\ref{fig:gpu_grid} is of course a minimal example of GPU programing.
 A full example will involve several lines of code transferring data to and from the GPU, and the kernel could include manual administration of the L1 cache (shared memory), and intra-block thread barriers.

### 2014-03-08 ThesisJP/Background/background.tex
Background. Ultrasound.

@@
 Where a CPU is said to execute threads which can perform SIMD instructions, a thread on a GPU can be interpreted as stream capable of processing one data element in the SIMD instruction.
+This is why GPU threads are said to be light weighted.
 
@@
 A full example will involve several lines of code transferring data to and from the GPU, and the kernel could include manual administration of the L1 cache (shared memory), and intra-block thread barriers.
-This is some of the reasons why writing a high performance GPU kernel is too often found to be a complex and time consuming task.
+This is some of the reasons why writing a high performance GPU kernel is often found to be a complex and time consuming task.
 
-No doubt, it is still a major overhead involved with writing GPU code, and in the future we will probably see work aimed at reducing this overhead.
+Because of the major overhead involved with writing GPU code we will, in the future, probably see a lot of work aimed at reducing it.
 A recent development is the introduction of C++ AMP which enables developers to write accelerated code (e.g.
@@
 Hopefully, in the future, GPU programming will be a job for compilers, and direct fiddling with GPU kernels should only be needed if maximum throughput is absolutely needed.
-This is similar to how SIMD instructions is handled in CPU code today.
+This is similar to how SIMD instructions are handled in CPU-code today.
 Most of the time developers will rely on the CPU compilers to analyze the code and auto-apply SIMD instruction whenever it is possible.
-However, when maximum throughput is needed, developers with deep knowledge of the problem at hand are often able to make faster code by manually writing SIMD instruction code.
+However, when maximum throughput is needed, developers with deep knowledge of the algorithm are often able to make faster code by manually writing SIMD instruction code.
 The relevance of auto-applied GPU instructions by compilers will also increase due to the recent embedding of GPUs on the same chip as the CPU.
-Soon, when writing code, we will probably talk about coding for a collection of heterogeneous cores where operations are automatically offloaded to special cores (e.g.
+Soon, when writing code, we will probably talk about coding for a collection of heterogeneous cores where operations are automatically offloaded to the right core (e.g.
 a GPU) by the compiler.
@@
 Ultrasound imaging encompasses technology which generates images based on sound whose frequencies we can not hear.
-For medical ultrasound imaging frequencies between 2 MHz and 10 MHz are typically used, where frequency dependent attenuation in human tissue force us to select a lower frequency if the object we want to image is located deep into the body.
-This is the reason why 2-4 MHz is used for cardiac imaging and a frequency around 7 MHz is used for vascular imaging.
-A piezoelectric ceramic is used to generate the ultrasonic frequencies, and is special in that way that ultrasound vibrations are generated when voltage is applied and \textit{vice versa}.
+For medical ultrasound imaging frequencies between 2 MHz and 10 MHz are typically used.
+Which frequency to select depends on how deep into the body we want to image, this because human tissue exhibits frequency dependent attenuation.
+This is the reason why a frequency around 3 MHz is used for cardiac imaging and a frequency around 7 MHz is used for vascular imaging.
+To generate the ultrasonic frequencies a piezoelectric ceramic is used.
+A piezoelectric material is special in that way that ultrasound vibrations are generated when voltage is applied and \textit{vice versa}.
 It can therefore be used to both send and receive ultrasound signals, like a combined microphone and speaker.
@@
 If the width of a piezoelectric element is less than the wavelength of the transmitted signal the element is said to be \textit{omnidirectional}, hence energy is distributed equally in all directions.
-When multiple omnidirectional elements are arranged in an array the transmitted pulse will get shaped into a beam\footnote{A beam refers to the pulse integrated in time.} of ultrasound by diffraction.
+When multiple omnidirectional elements are arranged in an array the transmitted pulse will get shaped into a beam\footnote{A beam refers to the pulse integrated in time.} of ultrasound by diffraction as the size (aperture) of the array grows.
 The 3 dB angular width of this beam, and also the angular resolution of the array, is given by \cite{AngelUltrasound}:

### 2014-03-08 ThesisJP/Background/background.tex
s

@@
 
-Modern GPUs can also be interpreted as a unit consisting of multiple SIMD processors (Fig.\,\ref{fig:cpu_gpu}).
-The Kepler architecture by Nvidia have up to 15 SIMD units capable of processing 192, 32 bit, floats per clock cycle at 745 MHz.
+Modern GPUs can also be interpreted as a unit consisting of multiple SIMD cores (Fig.\,\ref{fig:cpu_gpu}).
+The recent Kepler architecture by Nvidia have up to 15 SIMD units capable of processing 192, 32 bit, floating point values per clock cycle at 745 MHz\footnote{Nvidia typically refers to a core as a unit capable of performing a FMA instruction on one single precision floating point value.
+Hence, in Nvidia's world, Kepler has $15*192=2880$ cores.}.
 These GPUs are also capable of performing FMA instructions, effectively doubling the number of theoretical FLOPS.
@@
 The kernel function, for both CUDA and OpenCL, has a C-like syntax with some additional specifiers.
-A CUDA kernel is depicted in Fig.\,\ref{fig:gpu_kernel}.
+An example of a CUDA kernel is depicted in Fig.\,\ref{fig:gpu_kernel}.
 The \texttt{\_\_global\_\_} specifier tells the CUDA compiler that this function is a kernel function, and in CUDA, this kernel function is launched by using a CUDA specific syntax:
@@
 Thus,  in order to formulate a given problem as a kernel function the global memory input to each thread must not rely on the output from other blocks, hence it must be possible for the GPU to scheduled blocks in any order.
+
 As we see, a thread has a different conceptual meaning on a GPU than on a CPU.
@@
 This is why GPU threads are said to be light weighted.
+This conceptual difference also means that there is an inherent focus on breaking algorithms down and into SIMD instructions just by thinking of threads.
+
+As the width of SIMD instructions increase the only algorithms that will benefit is those who can be divided into enough concurrent parts.
+Also, to achieve peak performance, the implementation must perform one FMA operation on each core per clock cycle and all memory latency must be hidden by instructions.
+This is something which is impossible to achieve in real life applications.
 
@@
 This is some of the reasons why writing a high performance GPU kernel is often found to be a complex and time consuming task.
-
 Because of the major overhead involved with writing GPU code we will, in the future, probably see a lot of work aimed at reducing it.
@@
 The relevance of auto-applied GPU instructions by compilers will also increase due to the recent embedding of GPUs on the same chip as the CPU.
-Soon, when writing code, we will probably talk about coding for a collection of heterogeneous cores where operations are automatically offloaded to the right core (e.g.
+Soon, when writing code, we will probably talk about coding a heterogeneous collection of cores where operations are automatically offloaded to the right core (e.g.
 a GPU) by the compiler.
@@
 If the width of a piezoelectric element is less than the wavelength of the transmitted signal the element is said to be \textit{omnidirectional}, hence energy is distributed equally in all directions.
-When multiple omnidirectional elements are arranged in an array the transmitted pulse will get shaped into a beam\footnote{A beam refers to the pulse integrated in time.} of ultrasound by diffraction as the size (aperture) of the array grows.
+When multiple omnidirectional elements are arranged in an array the transmitted pulse will get shaped into a beam\footnote{A beam refers to the pulse in space integrated across time.} of ultrasound by diffraction as the size of the array  (the \textit{aperture}) grows.
 The 3 dB angular width of this beam, and also the angular resolution of the array, is given by \cite{AngelUltrasound}:
@@
 \end{align}
-\nomi{$\delta\theta$}{Angular resolution} where \nom{$\lambda$}{Wavelength} is the ultrasound wavelength, and \nom{$D$}{Aperture width} is the aperture width.
-With a speed-of-sound in human tissue of $1540\frac{\text{m}}{\text{s}}$ the angular resolution of a 2 cm cardiac array at 3 MHz is around 1.5\degree, which at 8 cm gives a lateral resolution of around 2 mm.
+\nomi{$\delta\theta$}{Angular resolution} where \nom{$\lambda$}{Wavelength} is the ultrasound wavelength, and \nom{$D$}{Aperture size} is the aperture width.
+With an average speed-of-sound in human tissue of $1540\frac{\text{m}}{\text{s}}$ the angular resolution or opening angle of a 2 cm cardiac probe operating at 3 MHz is around 1.5\degree, which at 8 cm gives a lateral resolution of around 2 mm.
 The axial resolution is inverse proportional to the pulse length, and is typically better than the lateral resolution.
@@
 This technique is know as \textit{beamforming}, and for ultrasound imaging it is applied both when transmitting and receiving ultrasound pulses.
-For medical ultrasound imaging there are two main types of ultrasound probes, linear and phased arrays \cite{AngelUltrasound}.
-A cardiac ultrasound array is a phased array, which means that the image is constructed from beams, or A-mode images, which are gradually steered from $-\theta/2$ to $\theta/2$, resulting in an imaging sector which are $\theta$ wide.
-This kind of image is known as a B-mode image (B for brightness).
+For medical ultrasound imaging there are two main types of ultrasound probes, linear and phased probes \cite{AngelUltrasound}.
+A cardiac probe is a phased array, which means that the image is constructed from beams, or A-mode images, which are gradually steered from $-\theta/2$ to $\theta/2$ radians, resulting in an imaging sector which are $\theta$ radians wide.
+This kind of image, where A-mode images are acquired from multiple spatial locations in the xz-plane, is known as a B-mode image (B for brightness).
+If A-mode images are steered in the yz-plane as well we get what is known as a volumetric or 3D ultrasound image.
+
 If \nom{$M$}{Number of array elements} is the number of elements in the array, and \nom{$\Delta_m[n]$}{Per-element focusing and steering delays} is the per-element focusing and steering delay, the standard form of beamforming used on receive, known as \textit{delay-and-sum}, is given as:
@@
 \end{align}
-where \nom{$\vec{w}$}{Array weight vector (apodization)} is know as the array weight vector or apodization, and $z[n]$ is the output A-mode image.
+where \nom{$\vec{w}$}{Array weight vector (apodization)} is know as the array \textit{weight vector} or \textit{apodization}, and $z[n]$ is the output A-mode image in the direction dictated by $\Delta_m[n]$.
 If $p_o$ is the phase center of our array and $p$ is the point where we want to steer and focus the ultrasound beam, we have to use the following delays:

### 2014-03-09 ThesisJP/Background/background.tex
Background. ULtrasound imaging.

@@
 When geometry is shaded, projected and rasterized the same instruction is typically needed for a lot of data at the same time.
-This makes a special kind of instruction, known as Single Instruction Multiple Data (\nom{SIMD}{Single instruction multiple data (instruction)}) \cite{Flynn1966}, especially suited for this kind of problem.
+This makes a special kind of instruction, known as Single Instruction Multiple Data (\nom{SIMD}{Single instruction multiple data (instruction type)}) \cite{Flynn1966}, especially suited for this kind of problem.
 The differences between how SIMD instructions are utilized in modern CPUs and GPUs are discussed in Section \ref{sec:cpu_vs_gpu}.
@@
 In Section \ref{sec:gpgpu} we introduced a special type of instructions, namely SIMD instructions, which are designed to execute a single instruction across multiple data element in one clock cycle.
-At the end of the 20th century, Intel introduced a SIMD instruction set for their Pentium III processor called the Streaming SIMD Extensions (\nom{SSE}{Streaming SIMD extensions (instruction)}), which has seen several updates since.
+At the end of the 20th century, Intel introduced a SIMD instruction set for their Pentium III processor called the Streaming SIMD Extensions (\nom{SSE}{Streaming SIMD extensions (instruction set)}), which has seen several updates since.
 The first SEE instructions set had 128 bit long registers which meant that four, 32 bit, floating point values could be processed per clock cycle\footnote{For simplicity we will assume that the whole register is processed in one clock cycle even if the actual implementation might distributes one instruction for the register across multiple cycles.}.
@@
 Early GPGPU programming required expert knowledge about computer graphics and the remaining fixed functions in the graphics pipeline.
-When GPGPU gained interest it quickly became evident that new programing languages where needed.
+When GPGPU gained interest and the GPU was used to compute problems that did not involve graphics, it quickly became evident that new programing languages where needed.
 One of the first GPGPU oriented languages for the GPU was the Brook language by Buck \textit{et al.} \cite{Buck2004}.
-Buck later joined Nvidia where he led the work on \nom{CUDA}{Compute unified device architecture (GPGPU language)}, a GPGPU language and framework by Nvidia for Nvidia GPUs only.
+Buck later joined Nvidia where he led the work on \nom{CUDA}{Compute unified device architecture (GPGPU framework)}, a GPGPU language and framework by Nvidia for Nvidia GPUs only.
 The first version of CUDA was released in 2007.
-The following year, Apple launched a GPGPU framework known as \nom{OpenCL}{Open Compute Language (GPGPU language)}, a standard which are maintained by the Khronos Group.
+The following year, Apple launched a GPGPU framework known as \nom{OpenCL}{Open Compute Language (GPGPU framework)}, a standard which are maintained by the Khronos Group.
 These two frameworks are still today the main programming frameworks for GPGPU.
@@
 \end{align*}
-where \texttt{grid} and \texttt{block} is 3-component vectors specifying grid and block size in "xyz" as depicted in Fig.\,\ref{fig:gpu_grid}, \texttt{a} is a pointer to a memory buffer located on the GPU, and \texttt{N}$*$\texttt{N} is the buffer size.
-When launched, a copy of the kernel is scheduled  for each thread in each block, and each kernel gets as input its location in the compute grid (\texttt{blockIdx} and \texttt{threadIdx}) together with the grid size (\texttt{gridDim} and \texttt{blockDim}).
+where \texttt{grid} and \texttt{block} is 3-component vectors specifying grid and block size in "xyz" as depicted in Fig.\,\ref{fig:gpu_grid}, \texttt{a} is a pointer to a memory buffer located on the GPU, and \texttt{N}$*$\texttt{N} is the size of the buffer \texttt{a}.
+
+When the kernel is launched, a copy of it is scheduled for each thread in each block, and each kernel instance gets as input its location in the compute grid (\texttt{blockIdx} and \texttt{threadIdx}) together with the grid size (\texttt{gridDim} and \texttt{blockDim}).
 This information is then typically used to calculate which memory elements to gather from global memory, in this case the elements in \texttt{a}.
@@
 When a kernel is executed, multiple thread blocks are scheduled to each SIMD core where they stay until all instructions in the kernel is finished for all threads in the block.
-Thus,  in order to formulate a given problem as a kernel function the global memory input to each thread must not rely on the output from other blocks, hence it must be possible for the GPU to scheduled blocks in any order.
+Thus,  in order to formulate a given problem as a kernel function the global memory input to each thread must not depend on the output from other blocks, hence, it must be possible for the GPU to schedule blocks in any order.
 
 As we see, a thread has a different conceptual meaning on a GPU than on a CPU.
-Where a CPU is said to execute threads which can perform SIMD instructions, a thread on a GPU can be interpreted as stream capable of processing one data element in the SIMD instruction.
+Where a CPU is said to execute threads which can perform SIMD instructions, a thread on a GPU can be interpreted as unit capable of processing one data element in the SIMD instruction.
 This is why GPU threads are said to be light weighted.
-This conceptual difference also means that there is an inherent focus on breaking algorithms down and into SIMD instructions just by thinking of threads.
+This conceptual difference also means that there is an inherent focus on breaking algorithms down and into SIMD instructions when programing for the GPU.
+This often leads to a highly optimized parallel implementation on the GPU and not on the CPU.
+Threads, or SIMD instructions, is just how one program for the GPU, there is no other way of doing it.
 
-As the width of SIMD instructions increase the only algorithms that will benefit is those who can be divided into enough concurrent parts.
-Also, to achieve peak performance, the implementation must perform one FMA operation on each core per clock cycle and all memory latency must be hidden by instructions.
-This is something which is impossible to achieve in real life applications.
+It is important to understand that not all algorithms will benefit from wider SIMD registers.
+As the width increase the only algorithms that will benefit is those who can be divided into enough concurrent fine granular parts.
+Also, to achieve peak performance, the implementation must perform one FMA operation on each core per clock cycle and all memory latency must be hidden by issuing instructions.
+All of this is something which is impossible to achieve in real life applications.
+One will always end up with a throughput which is a fraction of the peak throughput.
 
 The code in Fig.\,\ref{fig:gpu_grid} is of course a minimal example of GPU programing.
-A full example will involve several lines of code transferring data to and from the GPU, and the kernel could include manual administration of the L1 cache (shared memory), and intra-block thread barriers.
+A full example will involve several lines of C++ code transferring data to and from the GPU, and the kernel could include manual administration of the L1 cache (shared memory), and intra-block thread barriers.
 This is some of the reasons why writing a high performance GPU kernel is often found to be a complex and time consuming task.
@@
 by a GPU) in plain C++.
-Hopefully, in the future, GPU programming will be a job for compilers, and direct fiddling with GPU kernels should only be needed if maximum throughput is absolutely needed.
+Hopefully, in near future, GPU programming will be a job for compilers, and direct fiddling with GPU kernels should only be needed if maximum throughput is absolutely needed.
 This is similar to how SIMD instructions are handled in CPU-code today.
@@
 Which frequency to select depends on how deep into the body we want to image, this because human tissue exhibits frequency dependent attenuation.
-This is the reason why a frequency around 3 MHz is used for cardiac imaging and a frequency around 7 MHz is used for vascular imaging.
+Frequency dependent attenuation is the reason why a frequency around 3 MHz is used for cardiac imaging and a frequency around 7 MHz is used for more shallow imaging (e.g.
+vascular imaging).
 To generate the ultrasonic frequencies a piezoelectric ceramic is used.
@@
 
-If the width of a piezoelectric element is less than the wavelength of the transmitted signal the element is said to be \textit{omnidirectional}, hence energy is distributed equally in all directions.
+If the width of a piezoelectric element is less than the wavelength of the transmitted signal the element is said to be \textit{omnidirectional}, hence energy is distributed or sensed equally in all directions.
 When multiple omnidirectional elements are arranged in an array the transmitted pulse will get shaped into a beam\footnote{A beam refers to the pulse in space integrated across time.} of ultrasound by diffraction as the size of the array  (the \textit{aperture}) grows.
 The 3 dB angular width of this beam, and also the angular resolution of the array, is given by \cite{AngelUltrasound}:
-\begin{align}
-\delta\theta = \frac{\lambda}{D},
+\begin{align}\label{eq:res}
+\delta\theta_a = \frac{\lambda}{D},
 \end{align}
-\nomi{$\delta\theta$}{Angular resolution} where \nom{$\lambda$}{Wavelength} is the ultrasound wavelength, and \nom{$D$}{Aperture size} is the aperture width.
+\nomi{$\delta\theta_a$}{Angular resolution of an aperture} where \nom{$\lambda$}{Wavelength} is the ultrasound wavelength, and \nom{$D$}{Aperture size} is the aperture width.
 With an average speed-of-sound in human tissue of $1540\frac{\text{m}}{\text{s}}$ the angular resolution or opening angle of a 2 cm cardiac probe operating at 3 MHz is around 1.5\degree, which at 8 cm gives a lateral resolution of around 2 mm.
 The axial resolution is inverse proportional to the pulse length, and is typically better than the lateral resolution.
+
+\begin{figure}
+\centering
+\includegraphics[width=0.8\textwidth]{img/beamforming_das.eps}
+\caption{Delay-and-sum array beamforming.}
+\label{fig:das}
+\end{figure}
 
@@
 A cardiac probe is a phased array, which means that the image is constructed from beams, or A-mode images, which are gradually steered from $-\theta/2$ to $\theta/2$ radians, resulting in an imaging sector which are $\theta$ radians wide.
-This kind of image, where A-mode images are acquired from multiple spatial locations in the xz-plane, is known as a B-mode image (B for brightness).
-If A-mode images are steered in the yz-plane as well we get what is known as a volumetric or 3D ultrasound image.
+This kind of image, where A-mode images are acquired from multiple spatial locations in the $xz$-plane, is known as a B-mode image (B for brightness).
+If A-mode images are steered in the $yz$-plane as well we get what is known as a volumetric or 3D ultrasound image.
 
@@
 \end{align}
-where \nom{$\vec{w}$}{Array weight vector (apodization)} is know as the array \textit{weight vector} or \textit{apodization}, and $z[n]$ is the output A-mode image in the direction dictated by $\Delta_m[n]$.
+where \nom{$\vec{w}$}{Array weight vector (apodization)} is know as the array \textit{weight vector} or \textit{apodization}, and $z[n]$ is the output A-mode image in the direction given by $\Delta_m[n]$.
+Delay-and-sum array beamforming is depicted in Fig.\,\ref{fig:das}.
 If $p_o$ is the phase center of our array and $p$ is the point where we want to steer and focus the ultrasound beam, we have to use the following delays:
 \begin{align}
-\Delta_m[n] =\frac{|p_m+p| - |p_o+p|}{c},
+\Delta_m[n] =\frac{|p_o+p| - |p_m+p|}{c},
 \end{align}
@@
 On transmit the inverse set of delays have to be used.
-Also note that on transmit we have to decide where to steer and focus prior to transmitting a pulse, transmit delays are therefore not time varying for an A-mode image.
+Also note that on transmit we have to decide where to steer and focus prior to transmitting a pulse, transmit delays therefore do not vary within an A-mode image.
 
-Resolution and sampling.
+When generating a B-mode image we have to insure that the spatial sampling is sufficient.
+We have already seen that the lateral resolution of an ultrasound imaging probe is given by (\ref{eq:res}).
+Hergum \textit{et al.} \cite{Hergum2009} have further shown that the lateral frequency support is governed by the sum of the transmit (\nom{tx}{Transmit}) and receive (\nom{rx}{Receive}) aperture.
+The lateral resolution of an active imaging system is then found to be
+\begin{align}
+\delta\theta = \frac{\lambda}{D_{tx} + D_{rx}}.
+\end{align}
+\nomi{$\delta\theta$}{Angular resolution of a two-way imaging system}\nomi{$D_{tx}$}{Aperture size on transmit}\nomi{$D_{rx}$}{Aperture size on receive}In order to properly sample an imaging sector of $\theta$ radians we have to acquire
+\begin{align}
+N_\theta = \frac{\theta}{\delta\theta}
+\end{align}
+\nomi{$N_\theta$}{Number of lines required for proper sampling of an imaging sector}A-mode images. If \nom{$r$}{Imaging range.} is the maximum range we want to image we get an imaging frame rate of
+\begin{align}
+FR = \frac{c}{2rN_\theta}
+\end{align} 
+frames per second (\nom{FPS}{Frames per second}). With our 2 cm cardiac probe from the earlier example we then get a frame rate of $55$ FPS when imaging a 70\degree{} sector with $r=15$ cm.
 
-Sampling rates, and data size.
+The required axial sampling is dictated by the transmit frequency and the frequency bandwidth of the probe.
+From Fourier theory we know that a short impulse consist of many frequency components.
+A short ultrasound pulse, which would yield high axial resolution, therefore requires a large frequency bandwidth.
+A ultrasound probe has normally a relative bandwidth between 50 and 100 \%.
+This corresponds to transmitting a pulse which are between 2 and 1 oscillation long.
+In the front-end of an ultrasound imaging system a high sampling frequency is used (e.g.
+40 MHz).
+However, the ultrasound signal is band limited, and the data rate can therefore be reduced in a process known as \nom{IQ}{In-phase quadrature (ultrasound data format).} demodulation.
+In this process the axial sampling rate is reduced to the bandwidth of the imaging system in the following three steps: demodulation with a complex exponential, low-pass filtering, and decimation.
+Number of samples in one A-mode image is then found to be:
+\begin{align}
+N = \frac{2r}{c}B_wf_c.
+\end{align}
+\nomi{$N$}{Number of samples in range}where \nom{$B_w$}{Relative bandwidth} is the relative bandwidth and \nom{$f_c$}{Center frequency} is the center frequency of the transmitted pulse.
+The total number of samples in an B-mode image is then
+\begin{align}
+N_{\text{B-mode}} = N_\theta N.
+\end{align}
+For our cardiac probe we have to process close to $50000$ $M$-long vectors of element data every $1/55$ second.
 
@@
 Key selling points of ultrasound imaging: cost, safety and real-time user interaction.
-
-One sentence about speckle tracking.
 
@@
 
-Small section about heart anatomy.
+Small section about heart anatomy (Fig.\,\ref{fig:human_heart}).
 
@@
 
+One sentence about speckle tracking.
+
 \section{Volume rendering}\label{sec:volren}

### 2014-03-09 ThesisJP/Background/background.tex
ULtrasound imaging.

@@
 The first SEE instructions set had 128 bit long registers which meant that four, 32 bit, floating point values could be processed per clock cycle\footnote{For simplicity we will assume that the whole register is processed in one clock cycle even if the actual implementation might distributes one instruction for the register across multiple cycles.}.
-Updates of the SSE instruction set have added new instructions and longer registers (SSE2, SSE3, SSE4, \nom{AVX}{Advanced vector extensions (instruction)}, AVX2, and AVX-512).
+Updates of the SSE instruction set have added new instructions and longer registers (SSE2, SSE3, SSE4, \nom{AVX}{Advanced vector extensions (instruction set)}, AVX2, and AVX-512).
 The newly introduced extensions, AVX and AVX2, adds 256 bit long registers and fused multiply-accumulate instructions (\nom{FMA}{Fused multiply-accumulate (instruction)})  which means that sixteen 32 bit floating point operations can be performed per clock cycle per core.
@@
 \begin{align}\label{eq:res}
-\delta\theta_a = \frac{\lambda}{D},
+\delta\theta_a = \frac{\lambda}{D} = \frac{c}{fD},
 \end{align}
-\nomi{$\delta\theta_a$}{Angular resolution of an aperture} where \nom{$\lambda$}{Wavelength} is the ultrasound wavelength, and \nom{$D$}{Aperture size} is the aperture width.
-With an average speed-of-sound in human tissue of $1540\frac{\text{m}}{\text{s}}$ the angular resolution or opening angle of a 2 cm cardiac probe operating at 3 MHz is around 1.5\degree, which at 8 cm gives a lateral resolution of around 2 mm.
+\nomi{$\delta\theta_a$}{Angular resolution of an aperture}where \nom{$\lambda$}{Wavelength} and \nom{$f$}{Frequency} is the ultrasound wavelength and frequency respectively, $c$ is the speed of sound, and \nom{$D$}{Aperture size} is the aperture width.
+With an average speed-of-sound in human tissue of $1540\frac{\text{m}}{\text{s}}$, the angular resolution or opening angle of a 2 cm cardiac probe operating at 3 MHz is around 1.5\degree, which at 8 cm gives a lateral resolution of around 2 mm.
 The axial resolution is inverse proportional to the pulse length, and is typically better than the lateral resolution.
@@
 
-If time delays are applied on the signal sent to the different elements it is possible to steer and focus the ultrasound beam in space.
-This technique is know as \textit{beamforming}, and for ultrasound imaging it is applied both when transmitting and receiving ultrasound pulses.
+If time delays are applied on the signal sent to the different elements it is possible to steer and focus the ultrasound beam in space (Fig.\,\ref{fig:das}).
+This technique is known as \textit{beamforming}, and for ultrasound imaging it is applied both when transmitting and receiving ultrasound pulses.
 For medical ultrasound imaging there are two main types of ultrasound probes, linear and phased probes \cite{AngelUltrasound}.
@@
 This kind of image, where A-mode images are acquired from multiple spatial locations in the $xz$-plane, is known as a B-mode image (B for brightness).
-If A-mode images are steered in the $yz$-plane as well we get what is known as a volumetric or 3D ultrasound image.
+If A-mode images are steered in the $yz$-plane as well we get what is known as a volumetric B-mode or 3D ultrasound image.
 
-If \nom{$M$}{Number of array elements} is the number of elements in the array, and \nom{$\Delta_m[n]$}{Per-element focusing and steering delays} is the per-element focusing and steering delay, the standard form of beamforming used on receive, known as \textit{delay-and-sum}, is given as:
+If \nom{$M$}{Number of array elements} is the number of elements in the array, and \nom{$\Delta_m$}{Per-element focusing and steering delays} is the per-element focusing and steering delay, the standard form of beamforming used on receive, known as \textit{delay-and-sum}, is given as:
 \begin{align}\label{eq:das}
@@
 where \nom{$c$}{Speed of sound} is the speed of sound in human tissue, and $p_m$ is the position of the m'th element.
-On transmit the inverse set of delays have to be used.
+On transmit the inverse set of delays has to be used.
 Also note that on transmit we have to decide where to steer and focus prior to transmitting a pulse, transmit delays therefore do not vary within an A-mode image.
 
-When generating a B-mode image we have to insure that the spatial sampling is sufficient.
+When generating a B-mode image we have to insure that the spatial and temporal sampling is sufficient.
 We have already seen that the lateral resolution of an ultrasound imaging probe is given by (\ref{eq:res}).
@@
 \end{align}
-\nomi{$\delta\theta$}{Angular resolution of a two-way imaging system}\nomi{$D_{tx}$}{Aperture size on transmit}\nomi{$D_{rx}$}{Aperture size on receive}In order to properly sample an imaging sector of $\theta$ radians we have to acquire
+\nomi{$\delta\theta$}{Angular resolution of a two-way imaging system}\nomi{$D_{tx}$}{Aperture size on transmit}\nomi{$D_{rx}$}{Aperture size on receive}Then, in order to properly sample an imaging sector of $\theta$ radians we have to acquire
 \begin{align}
@@
 \end{align}
-\nomi{$N_\theta$}{Number of lines required for proper sampling of an imaging sector}A-mode images. If \nom{$r$}{Imaging range.} is the maximum range we want to image we get an imaging frame rate of
+\nomi{$N_\theta$}{Number of lines required for proper sampling of an imaging sector}A-mode images uniformly distributed across the sector.
+If \nom{$r$}{Imaging range.} is the maximum range we want to image we get an imaging frame rate of
 \begin{align}
@@
 \end{align} 
-frames per second (\nom{FPS}{Frames per second}). With our 2 cm cardiac probe from the earlier example we then get a frame rate of $55$ FPS when imaging a 70\degree{} sector with $r=15$ cm.
+frames per second (\nom{FPS}{Frames per second}). With our 2 cm cardiac probe from the earlier example we get a frame rate of around $55$ FPS when imaging a 70\degree{} sector with $r=15$ cm.
 
@@
 This corresponds to transmitting a pulse which are between 2 and 1 oscillation long.
-In the front-end of an ultrasound imaging system a high sampling frequency is used (e.g.
+In the front-end of an ultrasound imaging system a high sampling frequency is typically used (e.g.
 40 MHz).
-However, the ultrasound signal is band limited, and the data rate can therefore be reduced in a process known as \nom{IQ}{In-phase quadrature (ultrasound data format).} demodulation.
+However, the ultrasound signal is band limited, and the data rate can therefore be reduced in a process known as \textit{\nom{IQ}{In-phase quadrature (ultrasound data format).} demodulation}.
 In this process the axial sampling rate is reduced to the bandwidth of the imaging system in the following three steps: demodulation with a complex exponential, low-pass filtering, and decimation.
@@
 \begin{align}
-N = \frac{2r}{c}B_wf_c.
+N = \frac{2r}{c}B.
 \end{align}
-\nomi{$N$}{Number of samples in range}where \nom{$B_w$}{Relative bandwidth} is the relative bandwidth and \nom{$f_c$}{Center frequency} is the center frequency of the transmitted pulse.
+\nomi{$N$}{Number of samples in range}where \nom{$B$}{Frequency bandwidth} is the bandwidth, $B = B_wf_c$ (where \nom{$B_w$}{Relative bandwidth} is the relative bandwidth and \nom{$f_c$}{Center frequency of ultrasound pulse} is the center frequency of the transmitted pulse).
 The total number of samples in an B-mode image is then
@@
 \end{align}
-For our cardiac probe we have to process close to $50000$ $M$-long vectors of element data every $1/55$ second.
+For our cardiac probe has a relative bandwidth of $80$ \% we have to process almost $50000$ $M$-long vectors of element data every $1/55$ second.
 
-Time v.s. phase delays.
-
-Key selling points of ultrasound imaging: cost, safety and real-time user interaction.
+%Time v.s. phase delays. It is well described in the paper using phase delays.
 
@@
 
-Small section about heart anatomy (Fig.\,\ref{fig:human_heart}).
+Medical ultrasound imaging is considered to be a low-risk tool for non-invasive investigation.
+There are no gamma rays or high intensity magnetic fields involved, just sound pressure below well controlled limits.
+It is also far less costly than other imaging modalities and offers real-time inspection of the human body.
+Ultrasound imaging is well know for its use in fetal imaging during pregnancy, however, in this thesis we will focus on different modality, namely cardiac ultrasound imaging.
+An illustration of the human heart is therefore included in Fig.\,\ref{fig:human_heart}.
+Throughout the thesis we will refer to some of the parts shown in Fig.\,\ref{fig:human_heart}.
+Especially note the four chambers of the heart, the left and right atrium, and the left and right ventricle.
+The tissue separating the left and right ventricle is know as the inter-ventricular septum.
+The left side is the strongest and hardest working part of the heart, and it is responsible of pumping blood out to the whole body through the aortic valve.
+When blood flows out the aorta it is important that the mitral valve, that separates the left atrium and ventricle, is closed and does not leak.
+The left ventricle and the mitral valve therefore receives special attention during cardiac ultrasound examinations.
+When visualizing 3D ultrasound images we will have a strong focus on visibility of cardiac tissue, and note for later that the endocardium refers to the innermost layer of the cardiac muscle.
 
-\subsection{Accelerators in medical ultrasound imaging}
+\subsection{GPUs in medical ultrasound imaging}
 Ultrasound imaging systems generates an incredible amount of data.
@@
 However, as the computationally power of CPUs and GPUs have increased, ASICs, e.g.
-for scan conversion, have gradually been converted into software \cite{Guracar2013}.
+for converting data from a polar to a Cartesian grid (known as scan conversion), have gradually been converted into software \cite{Guracar2013}.
 One of  the last remaining ASICs is the beamformer, whose job is to steer and focus the ultrasound beam in space.
@@
 The input data rate is therefor very high.
-Nevertheless, real-time beamforming has recently been achieved on a GPU \cite{Song2012}, and a French company (SuperSonic Imagine) are utilizing GPUs in their plane wave ultrasound beamforming algorithm \cite{Tanter2014}.
-Beamforming is therefore finally moving into software to.
+Nevertheless, real-time beamforming has recently been achieved on a GPU \cite{Song2012}, and a French company (SuperSonic Imagine) are utilizing GPUs in their plane wave beamforming algorithm \cite{Tanter2014}.
+Beamforming is therefore finally moving into software too.
 The largest players in the medical ultrasound business, General Electric and Philips, are also well aware of where the trend is going \cite{Thomenius2012} \cite{Metz2011}.
-Software beamforming opens up new possibilities for advanced processing.
-One example is Capon beamforming which is explored in this thesis (\textbf{Paper\,I} and \textbf{II}), another is advanced 3D visualization as presented in \cite{solteszova2010multidirectional}, \textbf{Paper\,IV}, and \textbf{IX}.
+Software beamforming opens up new possibilities for advanced processing of ultrasound data.
+One example is Capon beamforming which is explored in this thesis (\textbf{Paper\,I} and \textbf{II}).
+The increased computational power introduced by GPUs makes it also possible to include advanced 3D visualization as presented in \cite{solteszova2010multidirectional}, \textbf{Paper\,IV}, and \textbf{IX}.
 

### 2014-03-10 ThesisJP/Background/background.tex
Medical ultrasound imaging.

@@
 Which frequency to select depends on how deep into the body we want to image, this because human tissue exhibits frequency dependent attenuation.
-Frequency dependent attenuation is the reason why a frequency around 3 MHz is used for cardiac imaging and a frequency around 7 MHz is used for more shallow imaging (e.g.
+Frequency dependent attenuation is the reason why a frequency around 3 MHz is used for fetal and cardiac imaging and a frequency around 7 MHz is used for more shallow imaging (e.g.
 vascular imaging).
-To generate the ultrasonic frequencies a piezoelectric ceramic is used.
+
+Medical ultrasound imaging is considered to be a low-risk tool for non-invasive investigation.
+There are no gamma rays or high intensity magnetic fields involved, just sound pressure below well controlled limits.
+It is also far less costly than other imaging modalities and offers real-time inspection of the human body.
+Ultrasound imaging is well know for its use in fetal imaging during pregnancy, however, in this thesis we will focus on a different modality, namely cardiac ultrasound imaging.
+An illustration of the human heart is therefore included in Fig.\,\ref{fig:human_heart}.
+Throughout the thesis we will refer to some of the parts shown in Fig.\,\ref{fig:human_heart}.
+Especially note the four chambers of the heart, the left and right atrium, and the left and right ventricle.
+The tissue separating the left and right ventricle is know as the inter-ventricular septum.
+The left side is the strongest and hardest working part of the heart, and it is responsible of pumping blood out to the whole body through the aortic valve.
+When blood flows out the aorta it is important that the mitral valve, that separates the left atrium and ventricle, is closed and does not leak.
+The left ventricle and the mitral valve therefore receives special attention during cardiac ultrasound examinations.
+When visualizing 3D ultrasound images in \textbf{Paper IV} we will have a strong focus on visibility of cardiac tissue, thus note for later that the endocardium refers to the innermost layer of the cardiac muscle.
+
+\begin{figure}
+\centering
+\subfigure{
+	\includegraphics[width=0.47\textwidth]{img/Diagram_of_the_human_heart.png}
+}
+\subfigure{
+	\includegraphics[width=0.47\textwidth]{img/HeartWall.png}
+}
+\caption{Overview of the human heart. Illustrations from wikipedia.org.}
+\label{fig:human_heart}
+\end{figure}
+
+To generate the ultrasonic frequencies used for imaging a piezoelectric ceramic is used.
 A piezoelectric material is special in that way that ultrasound vibrations are generated when voltage is applied and \textit{vice versa}.
@@
 \nomi{$\delta\theta_a$}{Angular resolution of an aperture}where \nom{$\lambda$}{Wavelength} and \nom{$f$}{Frequency} is the ultrasound wavelength and frequency respectively, $c$ is the speed of sound, and \nom{$D$}{Aperture size} is the aperture width.
-With an average speed-of-sound in human tissue of $1540\frac{\text{m}}{\text{s}}$, the angular resolution or opening angle of a 2 cm cardiac probe operating at 3 MHz is around 1.5\degree, which at 8 cm gives a lateral resolution of around 2 mm.
+With an average speed-of-sound in human tissue of $1540\frac{\text{m}}{\text{s}}$, the angular resolution or opening angle of a 2 cm cardiac probe operating at 3 MHz is around 1.5\degree, which at 8 cm range gives a lateral resolution of around 2 mm.
 The axial resolution is inverse proportional to the pulse length, and is typically better than the lateral resolution.
@@
 If time delays are applied on the signal sent to the different elements it is possible to steer and focus the ultrasound beam in space (Fig.\,\ref{fig:das}).
-This technique is known as \textit{beamforming}, and for ultrasound imaging it is applied both when transmitting and receiving ultrasound pulses.
+This technique is known as \textit{beamforming}, and for ultrasound imaging it is traditionally applied both when transmitting and receiving ultrasound pulses.
 For medical ultrasound imaging there are two main types of ultrasound probes, linear and phased probes \cite{AngelUltrasound}.
-A cardiac probe is a phased array, which means that the image is constructed from beams, or A-mode images, which are gradually steered from $-\theta/2$ to $\theta/2$ radians, resulting in an imaging sector which are $\theta$ radians wide.
+A cardiac probe is a phased probe, which means that the image is constructed from beams, or A-mode images, which are gradually steered from $-\theta/2$ to $\theta/2$ radians, resulting in an imaging sector which are $\theta$ radians wide.
 This kind of image, where A-mode images are acquired from multiple spatial locations in the $xz$-plane, is known as a B-mode image (B for brightness).
@@
 where \nom{$\vec{w}$}{Array weight vector (apodization)} is know as the array \textit{weight vector} or \textit{apodization}, and $z[n]$ is the output A-mode image in the direction given by $\Delta_m[n]$.
-Delay-and-sum array beamforming is depicted in Fig.\,\ref{fig:das}.
+A schematic overview of delay-and-sum array beamforming is depicted in Fig.\,\ref{fig:das}.
 If $p_o$ is the phase center of our array and $p$ is the point where we want to steer and focus the ultrasound beam, we have to use the following delays:
@@
 
-When generating a B-mode image we have to insure that the spatial and temporal sampling is sufficient.
-We have already seen that the lateral resolution of an ultrasound imaging probe is given by (\ref{eq:res}).
-Hergum \textit{et al.} \cite{Hergum2009} have further shown that the lateral frequency support is governed by the sum of the transmit (\nom{tx}{Transmit}) and receive (\nom{rx}{Receive}) aperture.
-The lateral resolution of an active imaging system is then found to be
+When generating a B-mode image we have to insure that the spatial and temporal sampling are sufficient.
+We have already seen that the lateral resolution of an ultrasound probe is given by (\ref{eq:res}).
+Hergum \textit{et al.} \cite{Hergum2009} have further shown that the lateral frequency support for a system which both emit and receive energy is governed by the sum of the transmit (\nom{tx}{Transmit}) and receive (\nom{rx}{Receive}) aperture.
+The lateral resolution of what is known as an active imaging system is found to be
 \begin{align}
@@
 \begin{align}
-FR = \frac{c}{2rN_\theta}
+f_{\text{FPS}} = \frac{c}{2rN_\theta}
 \end{align} 
@@
 In this process the axial sampling rate is reduced to the bandwidth of the imaging system in the following three steps: demodulation with a complex exponential, low-pass filtering, and decimation.
-Number of samples in one A-mode image is then found to be:
+The number of samples in one A-mode image is after this process given by:
 \begin{align}
-N = \frac{2r}{c}B.
+N = \frac{2r}{c}B,
 \end{align}
@@
 \end{align}
-For our cardiac probe has a relative bandwidth of $80$ \% we have to process almost $50000$ $M$-long vectors of element data every $1/55$ second.
+If our cardiac probe has a relative bandwidth of $80$ \% we have to process almost $50000$ $M$-long vectors of element data every $1/55$ second.
+The formula for number of FLOPS needed for DAS beamforming is then given as:
+\begin{align}
+\text{FLOPS}_{\text{DAS}} = MN_{\text{B-mode}}f_{\text{FPS}}
+\end{align}
+If our cardiac probe has $M=128$ elements, DAS beamforming requires a processing throughput of around $350$ MFLOPS. Hence, far below the peak processing throughput of modern GPUs and CPUs.
 
 %Time v.s. phase delays. It is well described in the paper using phase delays.
-
-\begin{figure}
-\centering
-\subfigure{
-	\includegraphics[width=0.47\textwidth]{img/Diagram_of_the_human_heart.png}
-}
-\subfigure{
-	\includegraphics[width=0.47\textwidth]{img/HeartWall.png}
-}
-\caption{Overview of the human heart. Illustrations from wikipedia.org.}
-\label{fig:human_heart}
-\end{figure}
-
-Medical ultrasound imaging is considered to be a low-risk tool for non-invasive investigation.
-There are no gamma rays or high intensity magnetic fields involved, just sound pressure below well controlled limits.
-It is also far less costly than other imaging modalities and offers real-time inspection of the human body.
-Ultrasound imaging is well know for its use in fetal imaging during pregnancy, however, in this thesis we will focus on different modality, namely cardiac ultrasound imaging.
-An illustration of the human heart is therefore included in Fig.\,\ref{fig:human_heart}.
-Throughout the thesis we will refer to some of the parts shown in Fig.\,\ref{fig:human_heart}.
-Especially note the four chambers of the heart, the left and right atrium, and the left and right ventricle.
-The tissue separating the left and right ventricle is know as the inter-ventricular septum.
-The left side is the strongest and hardest working part of the heart, and it is responsible of pumping blood out to the whole body through the aortic valve.
-When blood flows out the aorta it is important that the mitral valve, that separates the left atrium and ventricle, is closed and does not leak.
-The left ventricle and the mitral valve therefore receives special attention during cardiac ultrasound examinations.
-When visualizing 3D ultrasound images we will have a strong focus on visibility of cardiac tissue, and note for later that the endocardium refers to the innermost layer of the cardiac muscle.
 

### 2014-03-14 HowtoHowtoCapon/submission/submit1/answer/answer.tex
Adding answers to reviewers

% !TEX TS-program = pdflatex
% !TEX encoding = UTF-8 Unicode

% This is a simple template for a LaTeX document using the "article" class.
% See "book", "report", "letter" for other types of document.

\documentclass[11pt]{article} % use larger type; default would be 10pt

\usepackage[utf8]{inputenc} % set input encoding (not needed with XeLaTeX)

%%% Examples of Article customizations
% These packages are optional, depending whether you want the features they provide.
% See the LaTeX Companion or other references for full information.

%%% PAGE DIMENSIONS
\usepackage{geometry} % to change the page dimensions
\geometry{a4paper} % or letterpaper (US) or a5paper or....
% \geometry{margin=2in} % for example, change the margins to 2 inches all round
% \geometry{landscape} % set up the page for landscape
%   read geometry.pdf for detailed page layout information

\usepackage{graphicx} % support the \includegraphics command and options
\usepackage{xcolor}
% \usepackage[parfill]{parskip} % Activate to begin paragraphs with an empty line rather than an indent

%%% PACKAGES
\usepackage{booktabs} % for much better looking tables
\usepackage{array} % for better arrays (eg matrices) in maths
\usepackage{paralist} % very flexible & customisable lists (eg. enumerate/itemize, etc.)
\usepackage{verbatim} % adds environment for commenting out blocks of text & for better verbatim
\usepackage{subfig} % make it possible to include more than one captioned figure/table in a single float
% These packages are all incorporated in the memoir class to one degree or another...
\usepackage{amsmath}

%%% HEADERS & FOOTERS
\usepackage{fancyhdr} % This should be set AFTER setting up the page geometry
\pagestyle{fancy} % options: empty , plain , fancy
\renewcommand{\headrulewidth}{0pt} % customise the layout...
\lhead{}\chead{}\rhead{}
\lfoot{}\cfoot{\thepage}\rfoot{}

%%% SECTION TITLE APPEARANCE
\usepackage{sectsty}
\allsectionsfont{\sffamily\mdseries\upshape} % (See the fntguide.pdf for font help)
% (This matches ConTeXt defaults)

%%% ToC (table of contents) APPEARANCE
\usepackage[nottoc,notlof,notlot]{tocbibind} % Put the bibliography in the ToC
\usepackage[titles,subfigure]{tocloft} % Alter the style of the Table of Contents
\renewcommand{\cftsecfont}{\rmfamily\mdseries\upshape}
\renewcommand{\cftsecpagefont}{\rmfamily\mdseries\upshape} % No bold!

\newcommand\T{^{\scriptscriptstyle T}}
\renewcommand\H{^{\scriptscriptstyle H}}

\renewcommand\vec[1]{\boldsymbol{#1}}
\newcommand\mat[1]{\boldsymbol{#1}}

\newcommand\Om{O_\text{m}}
\newcommand\Oa{O_\text{a}}
\newcommand\Nl{N_\text{l}}
\newcommand\Nk{N_\text{k}}
\newcommand\1{\vec 1}
\newcommand\I{\mat I}
\renewcommand*\a{\vec a}
\renewcommand*\i{\vec i}
\renewcommand*\k{\vec k}
\newcommand*\n{\vec n}
\newcommand*\p{\vec p}
\newcommand*\s{\vec s}
\newcommand*\w{\vec w}
\newcommand*\x{\vec x}
\newcommand*\y{\vec y}

\newcommand*\A{\mat A}
\newcommand*\B{\mat B}
\newcommand*\C{\mat C}
\newcommand*\E{\mat E}
% \renewcommand*\H{\mat H}
\renewcommand*\P{\mat P}
\newcommand*\eP{\mat{\hat P}}
\newcommand*\R{\mat R}
\newcommand*\Ri{\R^{-1}}
\newcommand*\eR{\mat{\hat R}}
\newcommand*\eRi{\hat{\mat R}\;\!^{-1}}
\newcommand*\Navg{N_\text{avg}}
\newcommand*\W{\mat W}
\newcommand*\X{\mat X}
\newcommand*\Xd{\X_{\!\Delta}}
\newcommand*\Y{\mat Y}

\renewcommand*\L{\mat \Lambda}
\newcommand*\U{\mat U}
% \renewcommand*\t{\mathtiny{^T}}
% \newcommand*\h{\mathtiny{^H}}
\renewcommand*\t{^T}
\newcommand*\h{^H}

\newcommand\D{\vec\nabla} %Del: Vector differential operator - nabla
\newcommand\Dx{\vec\nabla\times}
\newcommand\Dd{\vec\nabla\cdot}

\newcommand\q[1]{\textit{#1}}
\newcommand\qi[1]{\item\q{#1}}
\newcommand\hi[1]{\item[\textbf{#1}]}
\newcommand\ans[1]{#1}
\newcommand\ai[1]{\\[.5\baselineskip]\ans{#1}}

\begin{document}
% \maketitle

\begin{center}\Large\bf
Author's response to the reviewer's comments
\end{center}

\begin{center}
\begin{tabular}{l p{.6\linewidth}}\centering
Manuscript number: & 2013JOE001295 \\
Title: & An Optimized GPU Implementation of the MVDR Beamformer for Active Sonar Imaging \\
Authors: & Buskenes,~J.I., \AA{}sen,~J.P., Nilsen,~C.-I.C, Austeng,~A.
\end{tabular}
\end{center}

We would like to thank the associate editor for his efforts and for his speedy replies to queries from the authors, and reviewers for their thorough and highly relevant comments. A summary of our answers are provided below, and subsequent sections will target each reviewer's comments in detail. We quote in \textbf{bold face} statements from the reviewers, and provide our replies in ordinary print,

\section{Summary of response to reviewer's comments}

\newpage
\section{Answer to reviewer 1}

\begin{enumerate}
% \emph{General}
\hi{General}
\qi{Mpx/s is not a widely known abbreviation. This needs to be defined upon first use.}
\ai{This is a good point. The Oxford dictionary lists "MP" as the correct abbreviation for "megapixel", so we substituted "Mpx/s" with "MP/s" and included a proper definition.}
\qi{"Matlab" should be all caps: "MATLAB" throughout the paper.}
\ai{Thank you, we were not aware of this. Corrections applied.}
%
\hi{Abstract}
\qi{A key attribute of the MVDR studied in this paper is the use of a subarray approximation. This should be stated up front in the abstract, i.e. "subarray MVDR" instead of MVDR.}
\ai{You are right that subarray averaging is important in our study. But we believe that "subarray MVDR" would be misleading, as our use of subarrays is strictly due to the application of MVDR in an active system to decorrelate signal and interference. TODO: Citation }
%
\qi{This paper advertises an impressive ("more than two orders of magnitude) improvement over a C implementation of MVDR, but this is really misleading. In reality, an improvement of up to an order of magnitude is achieved over an optimized C implementation, as later stated in the caption for figure 8. A comparison of an optimized GPU implementation to an unoptimized C implementation is not very meaningful — both should be optimized. The abstract should be revised to indicate up to an order of magnitude over an optimized C implementation.}
\ai{Thank you for this remark. We agree fully, and will state this better up front in the abstract. TODO}
%
\hi{Section I}
\qi{Typo in second paragraph: "litterature" should be "literature".}
\ai{Thank you. Fixed.}
%
\qi{Subarray processing is a common approximation and should be included in this list (reference D J Chapman, "Partial Adaptivity for the Large Array," IEEE Trans. Antennas \& Prop., Vol. Ap-24, No. 5, September 1976, pp. 685-696).}
\ai{Thank you providing us with this interesting reference. However, we can not see that the technique described by Chapman is relevant for us. He is not using overlapping subarrays like we do, and obtains a sparse array with associated grating lobe problems. We use of subarrays to (1) deal with the coherency problem in active sonar, and (2) to provide robustness to our covariance estimate. The latter is particularly important in our near single snapshot application. We attempt to emphasize this further in the report, and suggest [9] Kailath and Shan, "Adaptive beamforming for coherent signals and interference", as well as R.T.Hoctor and S.A.Kassam, "The unifying role of the coarray in aperture synthesis for coherent and incoherent imaging" as good reads on the subject. }
%
\hi{Section II}
\qi{Should reference the literature (from above comment 6) on the subarray approximation in paragraph 3.}
\ai{Certainly. Reference added.}
\hi{Section III (A-C)}
\qi{Typo in first paragraph after figure: "got" should be "get".}
\ai{Fixed.}
\qi{Move "a" in "It is a Gauss Jordan based".}
\ai{TODO}
\qi{"which are a mere scaled version" should be "which are merely scaled versions".}
\ai{Thank you. Corrected.}
%
\hi{Section IV}
\qi{The sub-plots on the top break the processing out into 3 pieces, yet the lower sub-plots reflect the entire processing string for MATLAB and C. It would be useful to show the entire processing time for the GPU for better comparison. Perhaps breaking the two rows of sub-plots into two figures would work better.}
\ai{Yes, the initial plots had the total GPU processing time included. But since there are three versions of the covariance building step, there TODO }
%
\qi{Comment about real processing improvement over optimized C should be related in the text instead of a caption. This is an important point about the comparisons made in the paper.}
\ai{It is indeed. We dedicated the second paragraph in the discussion (IV) to this topic, but will reword the relevant result paragraph slightly to hopefully cause less confusion.}
%
\qi{(caption of figure 10) "cruel" should be "crude"}
\ai{Thank you, crude grammar correction applied.}
%
\qi{The point made in the first paragraph about the data transfer from GPU to CPU being ignored further muddies the water in comparing the two approaches. This point should also be stated up front. I would suggest making a very clear itemization of the assumptions and caveats about the two approaches.}
\ai{TODO}
%
\hi{Section V}
\qi{The reason the C implementation is linear in L is because it is dominated by the calculation of the covariance matrix and data movement. The inversion is negligible here. On the other hand, the GPU optimizes the covariance calculation so much that now the inversion is significant, hence the non-linear computation time as a function of L. This is really the key point to be made in the comparison between the two approaches, in my opinion.}
\ai{Yes! This is likely the reason. Thank you for pointing this out. This will be added to the conclusions.}
\qi{Bottom of page, "hold- ups" has an extra space after the hyphen.}
\ai{Fixed.}
\hi{Section V}
\qi{Two spaces before "1 Mpx/s" in 4th paragraph. }
\ai{Thank you, corrected.}
\end{enumerate}

\newpage
\section{Answer to reviewer 2}

\q{The authors describe their work on implementing an MVDR adaptive beamformer for a high-frequency imaging sonar on a particular graphical processing unit (GPU), the prevous-generation NVIDIA Quadro 6000, using NVIDIA's CUDA C environment. The authors report 1-2 orders of magnitude faster processing with their GPU implementation, compared to their MATLAB and single-thread, unoptimized CPU implementations, without taking the input/output of data in and out of the GPU.}

\q{Because the paper is a valuable result to imaging sonar users, who will (to some extent) know that they can speed up their processing by 1-2 orders of magnitude if they use a GPU, I think the paper should be published.}

\q{However, I think the authors could greatly improve the appeal of the paper to a wider audience by adding helpful context and background on their implementation. I will suggest some questions that I think the authors could easily address toward this end. This is a report from the "front lines" of migrating scientific computing to GPUs, a topic which seems to be of great interest to everybody in our field that does numerical computing. In my review, I will comment on the information it provides to a reader interested in starting to use GPUs (and the authors for some selective elaboration).}

\subsection*{Suggestions for improving the exposition of the processing}

\q{I suspect many readers of the paper will not have the same assumptions as the authors, if they are not processing high-frequency narrowband signals with undersampled arrays. I see this disconnect between low-frequency passive processors and mid-frequency active processors, but the gap to imaging sonars is even greater. I venture that the authors can greatly broaden the appeal of their paper by addressing the questions I pose below.}

\begin{enumerate}
\qi{Here are my calculations of how the 32 elements are spaced in the HISAS 1030 system:}
\begin{align}
fc &= 100\;\text{kHz} \\
c  &= 1500\;\text{m/s} \\
\lambda &= c/f_c = 15\;\text{mm} \\
\textcolor{red}{ d } &= \textcolor{red}{\lambda/2 } \\
D &= 1200\;\text{mm} \\
\textcolor{red}{M} &= D/d = \textcolor{red}{160}
\end{align}
\ai{Thank you, we realize that this was not clear enough in the article. The 32 elements are evenly spaced along the 1.2m array, so:
\begin{align}
M &= 32 \\
d &= D / M = 3.75\;\text{cm} = 2.5\lambda
\end{align}
Further details are provided in 2.}
%
\qi{What is impact of using only 32 elements? How are they spaced? It would seem that spatial aliasing would be an issue. Authors should explain what their assumptions are.}
\ai{Thank you for pointing this out, we agree that this should be clarified. The array configuration produce the first grating lobes at $\theta = \text{arcsin}(\lambda M/D) = 23.5^\circ$, but this is not a problem since the element opening angle is $25^\circ$ ($\theta\in[12.5^\circ,12.5^\circ]$) at $f_c=100$\;kHz. In the paper we addition we perform sidescan imaging where only very narrow beams need to be beamformed. These details are now included in the paper.}
%
\qi{What is bandwidth of transmitted waveform? At 100 kHz, perhaps the authors are comfortably in a narrowband regime, and they do not need to worry about array response changing over their bandwidth (which would suggest that MVDR would need to be performed at different frequencies independently, unless some sort of pre-steering was used, which may be what the authors are doing anyway). Authors should explain what their assumptions are.}
\ai{Thank you. We mention in the beginning of section IIA that we pre-steer to every angle and range bin in the image, but will try to express this with greater clarity and emphasize that we are operating in the near field.}
%
\qi{Authors show solution of MVDR optimization as $\w = \frac{\Ri\1}{\1\T\Ri\1}$ and state this is the broadside solution. Are the authors pre-steering the array for every angle? Or are they using $w = \frac{\Ri\s}{\s\t\Ri\s}$ at off-broadside angles, with $\s$ the conventional steering vector (which would be a frequency-domain phase-shift vector, to be applied to a narrowband signal)? Pre-steering would enable disparate frequencies over a wide band to be coherently summed, to provide coherent gain in the look direction (over the band) by providing additional sample support for the sample covariance estimation.}
\ai{Yes, we are pre-steering the array for every angle. We will attempt to clarify this further, and add a remark saying that while not necessary here we could have phase steered the array by introducing the steering vector $\vec s$.}
%
\qi{Authors time average $\x\x\H$ over $K\in\{0, 1, 2\}$ time samples. Without doing some sort of reduced-rank processing, why is as little as 2 samples enough to estimate the covariance? Perhaps this is due to high SNR? At high SNR, though, why are the authors doing MVDR at all? What interference is being canceled, and what is its spatial distribution and stationarity? }
\ai{Again, thank you for a helpful remark, we will clarify this as well. After pulse compressing a chirp using the full transceiver bandwidth (30~kHz) range resolution is roughly $\Delta r \approx \frac{1}{B} = 2.5$~cm, which corresponds to a sample shift of $\Delta n = \frac{\Delta r f_s}{c} = 0.6$. Hence we cover the entire pulse compressed signal with only a few samples. $K=2$ corresponds to a window length of $2K+1=5$ samples, as is shown by (4) in the article.} 
%
\qi{I read the first paragraph of Section II, "..principle of adjusting the array's focus in range and bearing is commonly referred to as beamforming", and was confused - were the authors doing wavefront curvature beamforming, where every combination of range bin and beam angle required a distinct set of steering weights? After reading further, I would venture that the authors are steering only in angle, and that the index $n$ is over time of arrival, at the resolution of the width in time of the matched filter output (after pulse compression... the authors do not say anything about the waveform, so I have no idea whether there WAS any pulse compression). What is the resolution in time or range? }
\ai{Thank you, we understand how the confusion might arise. We operate under near field conditions and are indeed performing wavefront curvature beamforming. Hence, the index $n$ just an index over time with an undefined offset, at a resolution given by the sampling frequency. We will clarify this in the paper.}
\end{enumerate}

\subsection{Suggestions for additional GPU details so that paper can also serve as "field report" on CPU work in general}

\begin{enumerate}
\qi{It is unfortunate that so little effort was made to speed up the CPU versions of the code. For example, there are public-domain versions of optimized LAPACK libraries that provide the same sort of linear algebra codes as the authors used in their CPU implementation (for example, GoToBLAS, or Intel's MKL libraries).}
\ai{Thank you. We share these concerns but decided to leave CPU optimizations out of the paper. One reason is that we have not been able to find any other batch solvers out there and lacked the time to make our own. In our tests our custom inline Cholesky solver beat alternatives from e.g. MKL as our matrices are too small to justify the extra function call overhead. Another reason is that to make the comparison fair we feel an equal effort would have to put into writing a covariance builder for the CPU, and to minimize data movement in each step and in between them. Considering your remarks we will include these reasons in the paper.}
%
\qi{The extra input/output of data to the GPU is a burden only on the GPU implementation, so would penalize the GPU implementation only, unless the authors were to "hide" the i/o by setting it up to run in parallel with the CPU processing, which is easier and easier with the latest GPU architectures. This should perhaps be pointed out.}
\ai{Yes this is a good point. This is now included.}
%
\qi{The authors do not provide much information about the GPU hardware they used, the NVIDIA Quadro 6000. How does this compare to the desktop video cards that reside in most desktop, workstation, server and even high-end laptop systems? What is the vintage of the architecture of this GPU? How has the architecture evolved? For example, the authors tell us the Quadro 6000 (\$1900, 6GB VRAM, 448 cores) is a Fermi architecture - this is the previous architecture, having been superceded by the Kepler architecture, with a new Quadro K6000 board (\$3600, 12 GB VRAM, 2880 cores). The Quadro boards are high-end boards with video output targeting the CADAM market - there are many cheaper alternatives tar-geted at gangers and an even more expensive compute-only board called TESLA targeted at high performance computing. It's difficult to say how much of this irrelevant. }
\ai{TODO}
%
\qi{The authors state that OpenCL is an equally attractive alternative to CUDA. This would be the case if the authors had been able to find something in OpenCL akin to the recently released batch-optimized linear solver code for CUDA. Unfortunately, this is a pervasive story - the support for scientific programming in CUDA is years ahead of OpenCL. The benefit of OpenCL is that it is supposed to be "cross-platform", which is true in theory, but it is hard to imagine being true in practice considering the wide range of heterogeneous computing elements OpenCL is supposed to cover; from phones to DSP chips to FPGAs. }
\ai{These are important remarks, thank you. We have written OpenCL programs too, and run them on Intel CPUs and GPUs from AMD and Nvidia. It was no walk in the park, but with a satisfactory end result. But like you say, these are just a subset of devices that OpenCL support, and we have now changed the wording to be more specific here. Your remark on the lack a batched linear solver is spot in, the sentence is now rewritten.}
\end{enumerate}

\subsection{Not clear how R is being calculated in CPU}

\q{
Page 5, last paragraph in 1st column is puzzling. Authors say that "when a thread has finished up a diagonal, we have them wrap around to compute one of the diagonals in the lower triangular of R." The diagonals are all of different lengths, and, presumably the code to do a copy and conjugation from the upper triangular part to the lower triangular part of R requires different instructions than the moving average being calculated along a diagonal in the upper triangular part. The cores MUST execute the exact same code; or "diverge" (with one set executing one set of code while the complementary set sitting idle, then vice versa, as you would get with two if-then-else code sections), since this is a SIMD architecture. What is going on here?}

\q{This requires a better explanation.}

\ans{You are absolutely right. The copy will be serialized due to divergent branching. It is unfortunate that no Cholesky based batch linear equation solver exist, which would have eliminated the need for the lower triangular... The paragraph will be improved to explain this better.}

\subsection{Opportunities for additional GPU tutorial content?}

The authors describe the methodology they used to get things to run fast on the GPU. The analysis of the workload was informative and I think it is very informative for readers to see how the analysis progresses from less hardware specific (operations counts), to algorithm restructuring to reduce operations; to more hardware specific concerns (relative throughput of operations and memory transfers) .

Let me suggest some concepts that could be discussed a little bit more to give readers more of a sense of what programming GPUs is like.

\begin{enumerate}
\qi{What does somebody need to program in CUDA? Authors can mention that CUDA C is just C with a few extra language features to specify how to distribute the work across the cores. Authors can mention that there is a nvcc compiler, a profiler, a debugger, math libraries (CUFFT, CUBLAS), and a large number of examples of varying complexity, all from NVIDIA. Authors can mention that CUDA can be used on Windows, MacOSX, or Linux, with an NVIDIA graphics card. Authors can mention that there are also third party CUDA interfaces in Fortran and Python. When discussing the batch-optimized linear solver code from NVIDIA, the authors should cite the NVIDIA site, and also links to CULA and MAGMA, other CUDA versions of LAPACK or subsets thereof.}
\qi{Problems need to have high "arithmetic intensity" (the authors use "computational intensity") so that the cores stay busy. If there is not enough arithmetic, the cores sit idle waiting for data to be moved around. The authors do a nice job describing this situation on page 4 and the start of page 5.}
\ai{}
\qi{The authors use the term "arithmetic reduction". Unfortunately, reduction is an important word/topic in GPU work. Perhaps "minimizing arithmetic operations" would be better. Reduction is any N-to-1 mapping, which poses a conundrum for N cores having to combine their individual results at a single location all at the same time... you don't want the cores to all line up and wait their turn.}
\qi{On page 4, the authors talk about avoiding global memory, without get-ting into "coalesced" memory access patterns. This might merit a mention - for example, is at best only able to move one float for every 30 floats processed, and potentially much worse if care is not taken to use "coa-lesced memory access patterns (that avoid memory bank conflicts)". On page 5, the authors talk about "a collective access pattern that maximizes global memory bandwidth". Here too, the authors should perhaps use the "coalesced" term, since this is the term used in all of the GPGPU documentation.}
\end{enumerate}

\subsection{Localized comments}

\begin{enumerate}
\qi{Equation (2) using a "broadside steering vector" - this is confusing... are the authors only steering at broadside? clearly not... do the authors pre-steer all of the elements prior to forming R so that can "focus" in a wideband sense?}
\qi{Paragraph after equation (2) mentions "spatial averaging to avoid signal cancellation" - this problem might be better identified for readers by cit-ing the Kailath reference and using "signal cancellation due to coherent multipath".}
\qi{Equation 4 has second summation using index n', but this index does not appear in any of the terms inside the summation.}
\qi{Page 2 - R is averaged over K time samples... at what rate is beamformer output produced? is R averaged using a moving window process, so that R[n] is just a rank-1 update to R[n — 1] and there is one beamformer output for every input time sample?}
\qi{Page 4, in list of three optimizations - item 3 is not self-evident, but if I understand what the authors are doing, it is described by the 1985 reference by Kailath.}
\qi{Page 4, last paragraph of 1) Arithmetic reduction - authors should iden-tify that the "Minimized memory" version skips step 2) from list of three optimization, and that the "Minimized instructions" version uses all three steps... I had a hard time understanding how the three "arithmetic reduc-tion steps" were related to the plots in Figure 5, and was puzzled about where the "minimized memory" and "minimized instruction" variants of the processing had come from. Perhaps the authors can introduce these variants more clearly and more prominently.}
\qi{On page 6, authors say "The bottleneck in the final design is now the in-version step, which is typically 5 times slower than the build step". Surely, this merits an explanation. What is the final design being compared with? What was the expectation? What changed? Figure 3 shows FLOPs before optimization - this shows build and inversion steps of roughly the same order of magnitude in the right half of the figure. Does Figure 3 set the expectation? Figure 5 shows "relative complexity reduction" of entire pro-cess, so there is no information on relative speed of build and inversion steps. Figure 10 shows estimated FLOPs for the two steps, but since both steps have different theoretical complexities (operations counts), and the two steps do not optimize the same way, this does not provide an expec-tation of the relative times.}
\qi{On page 6, authors say their implementation was improved by factor of 1.5 - 2 when run on a K20. Authors could say a few words about what a K20 is, how it differs from the Quadro 6000, that the Kepler architecture actually has features that are dramatically better for scientific computing if the code is adapted to use them (kernels can call kernels, more registers, more shared memory, ability to eliminate some host-to-GPU i/o).}
\qi{Section I says 1-2 orders of magnitude. Section IV, page 6, says 2-3 orders of magnitude.}
\qi{Reference 9 has two authors, Shan and Kailath. 11. Reference 17 seems to be the reviewed manuscript? }
\end{enumerate}

\begin{enumerate}
\item Fig. 3 - Slightly increased spacing between subplots.
\item Fig. 8 - Abbreviating megapixels as "MP" in stead of "MPx".
\item Fig. 9 - Abbreviating megapixels as "MP" in stead of "MPx".
\item Fig. 11 - Abbreviating megapixels as "MP" in stead of "MPx".
\end{enumerate}

% 
% 
% 
% Describe how models should be loaded. Framework and formate etc.
% 
% \section{Calculation by Rasterization}
% \subsection{From 3D Models to Rasterized Points}\label{sec:modelsToPoints}
% The 3D models and sea floor has to be rasterized on to a view plane in order to calculate output intensity using CUDA.
% 
% We plan to use openGL to construct depth and reflection coefficient buffers at each along-track position. For this we need to set up a camera in a scene that mimic the sonar array imaging objects on a sea floor. The output from this step is a depth and a reflection coefficient buffer for each along-track position. Investigation has to be done to see if information is redundant along the track. We might also need to calculate per-pixel normals for use in the next step.
% 
% \subsection{Calculating Intensity}\label{sec:calcIntensity}
% For each pixel, load normal, range, and reflection coefficient. Calculating the contribution from this location is then done by calculating the reciprocal doubled range, and weight it with the reflection coefficient and the normal component towards the along-track position. Saving these calculations to a buffer, we can now find the slant-range pixel by summing all values closer than half the sampling distance for the given slant-range value.
% 
% \subsection{Parallel Design Pattern for the GPU}
% It is natural that we loop over along-track positions. At each position we generate the listed buffers in Section \ref{sec:modelsToPoints} and holds them in GPU memory. The buffers is then fed to a custom CUDA kernel that calculates the reflected intensity per buffer location. All buffers should be placed in texture memory so that resolution can be adjusted at any step. If we render each buffer listed in Section \ref{sec:modelsToPoints} at resolution $M \times N$, the step in Section \ref{sec:calcIntensity} can easily be rendered at $2M \times 2N$ by interpolation. Finally we arrive to what is believed to be the less parallel part of this approach, the sum operation. A kernel has to be launched that for each slant-range accumulates the total contribution in the given intensity buffer. We might want to use a library kernel for this if it exist. If we are going to code this step, one idea could be as follow: Assign a region to a block of treads, find min and max, and let one thread search for one ore more 
% slant-range values\footnote{Need to do some more thinking here}.
% 
% \subsection{Concluding Remarks}
% Uses a lot of memory, many calculations. Flexible. Will support point-based objects and sea floor out of the box. 
% 
% \section{Geometric Calculation of Output Intensity}
% Inspired by the Field II ultrasound simulator, where array responses are found by considering geometry, the same concept can be used for reflection off object geometry. Geometry in our case can be viewed as arrays, where the emitted energy is reflected energy from a point source at the along-track position.
% 
% Given a list of triangle patches, were the three corners of each triangle has a corresponding reflection coefficient, normal and position/depth, the reflection intensity can be found by simple geometric considerations. First we should sort triangles in depth, so that GPU-threads in a block, process vertices that are close to each other in range. This to maximize throughput of writing to global memory. One thread loads a triangle and iterates from min to max depth using the slant-range sampling distance as step length. For each step the thread calculates the intersection between the triangle and a sphere with radius given by the current depth. The intersection length is weighted with both the reciprocal depth, the reflection coefficient and the triangle normal \footnote{We should only use triangles with equal normals in each corner. Hence they are all equal to the average triangle normal. Might not be compliant.}.
% 
% \subsection{Concluding Remarks}
% Uses little memory, few calculations. Will not support point-based objects and sea floor out of the box.

\end{document}

### 2014-03-14 HowtoHowtoCapon/submission/submit1/answer/answer.tex
la2pdf commit - 2014:03:14	01:57:07	(Fri14.Mar)
la2pdf commit - 2014:03:14	15:02:53	(Fri14.Mar)

@@
 The authors do a nice job describing this situation on page 4 and the start of page 5.}
-\ai{}
+\ai{Thank you for those kind words. In the mentioned context we agree that "arithmetic" is likely a better alternative to "computational", and will make the relevant edits.}
+%
 \qi{The authors use the term "arithmetic reduction".
@@
 you don't want the cores to all line up and wait their turn.}
-\qi{On page 4, the authors talk about avoiding global memory, without get-ting into "coalesced" memory access patterns.
-This might merit a mention - for example, is at best only able to move one float for every 30 floats processed, and potentially much worse if care is not taken to use "coa-lesced memory access patterns (that avoid memory bank conflicts)".
+\ai{Yet again, very true! Your suggestion is much better, granted your permission we will use it.}
+%
+\qi{On page 4, the authors talk about avoiding global memory, without getting into "coalesced" memory access patterns.
+This might merit a mention - for example, is at best only able to move one float for every 30 floats processed, and potentially much worse if care is not taken to use "coalesced memory access patterns (that avoid memory bank conflicts)".
 On page 5, the authors talk about "a collective access pattern that maximizes global memory bandwidth".
 Here too, the authors should perhaps use the "coalesced" term, since this is the term used in all of the GPGPU documentation.}
+\ai{You are right. Initially we sought to avoid as much GPU jargon as we could, but in its current form it probably just adds confusion to leave it out. Your suggestion sounded well.}
 \end{enumerate}
@@
 \end{enumerate}
+
+\newpage
+\section{Changes to figures}
 

### 2014-03-14 HowtoHowtoCapon/submission/submit1/answer/answer.tex
la2pdf commit - 2014:03:14	15:08:45	(Fri14.Mar)
la2pdf commit - 2014:03:14	17:35:48	(Fri14.Mar)

@@
 do the authors pre-steer all of the elements prior to forming R so that can "focus" in a wideband sense?}
+\ai{We indeed presteer all the elements. TODO}
+%
 \qi{Paragraph after equation (2) mentions "spatial averaging to avoid signal cancellation" - this problem might be better identified for readers by cit-ing the Kailath reference and using "signal cancellation due to coherent multipath".}
-\qi{Equation 4 has second summation using index n', but this index does not appear in any of the terms inside the summation.}
-\qi{Page 2 - R is averaged over K time samples...
+\ai{Thank you, both suggestions are now included.}
+%
+\qi{Equation 4 has second summation using index $n'$, but this index does not appear in any of the terms inside the summation.}
+\ai{Good catch! Fixed.}
+%
+\qi{Page 2 - $\R$ is averaged over K time samples...
 at what rate is beamformer output produced?
-is R averaged using a moving window process, so that R[n] is just a rank-1 update to R[n — 1] and there is one beamformer output for every input time sample?}
+is $\R$ averaged using a moving window process, so that $\R[n]$ is just a rank-1 update to $\R$[n — 1] and there is one beamformer output for every input time sample?}
+\ai{Actually, $\R$ is averaged over $2K+1$ time samples, this can be seen from (4). TODO}
+%
 \qi{Page 4, in list of three optimizations - item 3 is not self-evident, but if I understand what the authors are doing, it is described by the 1985 reference by Kailath.}

### 2014-03-15 HowtoHowtoCapon/submission/submit1/answer/answer.tex
la2pdf commit - 2014:03:15	15:17:09	(Sat15.Mar)
la2pdf commit - 2014:03:15	15:48:34	(Sat15.Mar)

@@
 \qi{Mpx/s is not a widely known abbreviation. This needs to be defined upon first use.}
-\ai{This is a good point. The Oxford dictionary lists "MP" as the correct abbreviation for "megapixel", so we substituted "Mpx/s" with "MP/s" and included a proper definition.}
+\ai{This is a good point. "Mpx" is now corrected to "MP", in accordance with the second reviewer's suggestion and the Oxford dictionary.}
 \qi{"Matlab" should be all caps: "MATLAB" throughout the paper.}
-\ai{Thank you, we were not aware of this. Corrections applied.}
+\ai{Thank you, we agree. Corrections applied.}
 %
@@
 \ai{You are right that subarray averaging is important in our study.
-But we believe that "subarray MVDR" would be misleading, as our use of subarrays is strictly due to the application of MVDR in an active system to decorrelate signal and interference.
-TODO: Citation }
+But we believe that "subarray MVDR" would be misleading, as we use subarray averaging mainly to deal with the coherent echoes in active sonar.
+We have now stated this better up front in the article.}
 %
@@
 \ai{Certainly. Reference added.}
+%
 \hi{Section III (A-C)}
 \qi{Typo in first paragraph after figure: "got" should be "get".}
-\ai{Fixed.}
 \qi{Move "a" in "It is a Gauss Jordan based".}
-\ai{TODO}
 \qi{"which are a mere scaled version" should be "which are merely scaled versions".}
-\ai{Thank you. Corrected.}
+\ai{We thank the reviewer for these corrections (8.-10.). Corrections applied.}
 %

### 2014-03-15 ThesisJP/Background/background.tex
Added comment on IO.
Writing section on adaptive beamforming.

@@
 If the width of a piezoelectric element is less than the wavelength of the transmitted signal the element is said to be \textit{omnidirectional}, hence energy is distributed or sensed equally in all directions.
-When multiple omnidirectional elements are arranged in an array the transmitted pulse will get shaped into a beam\footnote{A beam refers to the pulse in space integrated across time.} of ultrasound by diffraction as the size of the array  (the \textit{aperture}) grows.
+When multiple omnidirectional elements are arranged in a linear array the transmitted pulse will get shaped into a beam\footnote{A beam refers to the pulse in space integrated across time.} of ultrasound by diffraction as the size of the array  (the \textit{aperture}) grows.
 The 3 dB angular width of this beam, and also the angular resolution of the array, is given by \cite{AngelUltrasound}:
@@
 This kind of image, where A-mode images are acquired from multiple spatial locations in the $xz$-plane, is known as a B-mode image (B for brightness).
-If A-mode images are steered in the $yz$-plane as well we get what is known as a volumetric B-mode or 3D ultrasound image.
+If A-mode images are acquired from the $yz$-plane as well we get what is known as a volumetric B-mode or 3D ultrasound image.
+A probe for 3D imaging has several thousand elements arranged as a matrix.
+In this thesis we will focus on linear arrays for 2D imaging when the topic is adaptive beamforming, and 3D imaging when the topic is adaptive visualization.
 
@@
 If our cardiac probe has a relative bandwidth of $80$ \% we have to process almost $50000$ $M$-long vectors of element data every $1/55$ second.
-The formula for number of FLOPS needed for DAS beamforming is then given as:
+The formula for the number of FLOPS required for real-time DAS beamforming of cardiac ultrasound is then given as:
 \begin{align}
-\text{FLOPS}_{\text{DAS}} = MN_{\text{B-mode}}f_{\text{FPS}}
+\text{FLOPS}_{\text{DAS}} = MN_{\text{B-mode}}f_{\text{FPS}} = MB_wf_c
 \end{align}
-If our cardiac probe has $M=128$ elements, DAS beamforming requires a processing throughput of around $350$ MFLOPS. Hence, far below the peak processing throughput of modern GPUs and CPUs.
+If our cardiac probe has $M=128$ elements, DAS beamforming requires a processing throughput of around $300$ MFLOPS.
+Hence, far below the peak processing throughput of modern GPUs and CPUs.
+The main challange with DAS beamforming is the amount of data that has to be moved.
+If one element sample is two bytes long, for our example, around 600 MB has to be moved on to the GPU every second.
+This is however less than the maximum bandwidth of PCI Express 2.0 (8 GB/s), even if the relative bandwidth, transmit frequency and number of array elements are increased with a total factor of 10.
 
@@
 
-\subsection{Adaptive Beamforming}\label{sec:adaptbf}
+\subsection{Adaptive beamforming}\label{sec:adaptbf}
+Adaptive beamforming has several meanings in the medical ultrasound literature.
+It can for instance refer to aberration correction which is the task of applying time delays to compensate for the distortion introduced by fatty layers in the human body.
+However, in this thesis adaptive beamforming will refer to data-dependent apodization (or weighting) derived from a given optimization criterion.
+The optimization criterion which we focus on in this thesis is the minimum variance distortionless response criterion, which is also known as Capon's method \cite{Capon1969}.
+A beamformer which uses weights derived using Capon's method will in this thesis mainly be referred to as the \textit{Capon beamformer}.
+However, it is also known as the minimum variance  (\nom{MV}{Minimum variance beamformer}) beamformer or the minimum variance distortionless respons beamformer (\nom{MVDR}{Minimum variance distortionless respons beamformer (Capon)}).
 
 \begin{figure}[t!]
-\subfigure[Delay-and-sum]{
-	\includegraphics[width=0.47\textwidth]{img/scenario_das_resp2.png}
+\subfigure[Uniform weights]{
+	\includegraphics[width=0.47\textwidth]{img/scenario_das_resp2.png}\label{fig:das_weights}
 }
-\subfigure[Capon]{
-	\includegraphics[width=0.47\textwidth]{img/scenario_mv_resp2.png}
+\subfigure[Capon weights]{
+	\includegraphics[width=0.47\textwidth]{img/scenario_mv_resp2.png}\label{fig:capon_weights}
 }
-\caption{Array beam pattern with uniform and Capon weights.
+\caption{Examples of array beampatterns where the array weights are either uniform (a) or derived using Capon's method (b).
 An interfering source is located at 20\degree.
-Notice the reduced sidelobe level for the Capon beam pattern in the direction of the interferer.}
+Notice the reduced sidelobe level for the Capon beampattern in the direction of the interfering source.
+Image courtesy of Jo Inge Buskenes.}
+\label{fig:weights}
 \end{figure}
 
-Capon beamforming\footnote{The name ''Capon beamformer'' is due to work by J. Capon \todo{add citation} on seismic arrays \todo{Move to background.}} or minimum variance beamforming.
+The Capon beamformer produces a weight vector, as introduced in (\ref{eq:das}), based on the impinging signal and an optimization criterion. The problem is as follow \cite{Capon1969}:
+\begin{align}
+&\min_{\vec{w}} E\{|z[n]|^2\} \rightarrow \min_{\vec{w}} \vec{w}^H \mat{\hat{R}} \vec{w} \label{eq:capon_optimization_criteria} \\
+&\text{subjected to } \vec{w}^H\vec{a} = 1,
+\end{align}
+where \nom{$\mat{\hat{R}}$}{Sample covariance matrix (with regularization)} is a sample covariance matrix.
+Hence, the resulting weight vector minimizes the output power (the criterion) while maintaining unit gain in the steering direction $\vec{a}$ (distortionless response).
+The benefit of adaptive beamforming is clearly seen from Fig\,\ref{fig:weights} which shows a sensitivity map (an \textit{array beampattern}) for DAS beamforming with uniform weights and the Capon beamformer.
+In this simulated case an interfering source is located at 20\degree{} which for the DAS beamformer with uniform weights (\ref{fig:das_weights}) will leak into the beamformer output with its power attenuated with around $13$ dB.
+Since the Capon weights are data-dependent the method "knows" where the interfering source is located and is able to position a zero in the direction of it.
+Hence, the interfering source will be highly attenuated independent of its direction.
+Another interesting case is when the interfering source is located within the mainlobe of the array beampattern.
+In theory a signal will be regarded by the Capon beamformer as an interfering source as long as the signal propagation vector does not match the steering vector ($\vec{a}$) exactly.
+If it is not possible to move a zero to the direction of the interferer the mainlobe will get shifted if this results in minimum output power.
+Adaptive positioning of zeros and mainlobe shifting gives rice to the super-resolution capabilities of the Capon beamformer \cite{Synnevag2007}, which is the main reason why Capon beamforming has been highly studied in the field of medical ultrasound imaging the last decade.
 
-Intro to adaptive beamforming. List other variants (LCA, beamspace, Eigen space etc.).
+To "know" the direction of interfering sources the Capon beamformer has to estimate the statistics of the impinging wave field.
+This is done by estimating a sample covariance matrix, which for an active broadband system can be estimated in the following way \cite{Synnevag2009}:
+\begin{align}
+\mat{\breve{R}}[n] = \frac{1}{N_LN_K}\sum_{n'=n-K}^{n+K} \sum_{l=0}^{N_L-1} \vec{x}_l[n']\vec{x}_l[n']^H,
+\end{align}
+where $K = M-L+1$\nomi{$K$}{Temporal averaging parameter (Capon)} typically is proportional to the pulse length, $N_K = 2K + 1$ \nomi{$N_K$}{Number of samples used for temporal averaging (Capon)}, $N_L = M-L+1$,\nomi{$N_L$}{Number of subarrays used for spatial averaging (Capon)}\nomi{$L$}{Subarray length (Capon)} and $\vec{x}_l$ is a $L$ long subarray $[x_l[n], \dotso x_{l+L}[n]]$.
+To ensure robustness against model errors and numerical stability the matrix is loaded with a diagonal factor,
+\begin{align}
+\epsilon &= d*\text{trace}\{\mat{\breve{R}}\}/L\\
+\mat{\hat{R}} &= \mat{\breve{R}} + \epsilon\mat{I}.
+\end{align} 
+Thus, a white noise component proportional to the in-coherent signal power is added to the estimate.
+
+The solution to the minimization problem in (\ref{eq:capon_optimization_criteria}) is, by the method of Lagrange multipliers \cite{VanTrees2003}, found to be
+\begin{align}\label{eq:capon_weights}
+\vec{w}[n] = \frac{\mat{\hat{R}}[n]^{-1}\vec{a}}{\vec{a}^H\mat{\hat{R}}[n]^{-1}\vec{a}} \in \mathbb{C}^L.
+\end{align}
+Since $\vec{w}$ is time (or range) dependent one matrix has to be constructed and inverted for each data vector $\vec{x}[n]$ received by the system.
+This is indeed computationally demanding, something which becomes clear when we take a closer look at each step of the beamformer.
+Construction of the covariance matrix has a computationally complexity of $O(L^2N_LN_K)$\nomi{O(.)}{Big O notation (algorithm complexity)}, inverting it has a computationally complexity of $O(L^3)$, and the final normalization and calculation of $\vec{w}$ is $O(L)$.
+For $L=M/2$, a subarray length often used in the literature, the first two steps both have a complexity of $O(L^3)$.
+These steps should therefore receive the same amount of attention when a fast implementation of Capon beamforming is attempted.
+A rough estimate of the number of FLOPS required by a Capon beamformer for real-time cardiac ultrasound imaging is then given by:
+\begin{align}
+\text{FLOPS}_{\text{Capon}} = L^3N_{\text{B-mode}}f_{\text{FPS}} = L^3B_wf_c.
+\end{align}
+Which for our example, with $L=M/2=64$, sums up to a required throughput of more than $0.5$ TFLOPS.
+Even though this is just 10\% of the peak throughput of modern GPUs remember that the latter where only theoretical numbers, and that the actual throughput is highly algorithm dependent.
+The required throughput for Capon beamforming is, for the moment, hard to achieve in practice \cite{So2011}.
+As for DAS beamforming this number also scales with higher relative bandwidth, pulse frequency, and more array elements since subarray size typically is proportional to the number of array elements.
+In the literature on medical ultrasound imaging we find few attempts on implementing a real-time Capon beamformer for medical ultrasound imaging.
+In addition to the papers in this thesis only Chen \textit{et al.} \cite{Chen2011, Chen, Chen2011a} are known to have explored GPU and FPGA implementations of the Capon beamformer for medical ultrasound imaging, however with some shortcomings  which are discussed in both \textbf{Paper I} and \textbf{II}.
+
+To reduce the processing requirement for Capon beamforming several simplifications have been proposed \cite{Asl2012, Kim}.
+The most interesting proposals include the low complexity beamformer by Synnev\aa{}g \textit{et al.} \cite{Synnevag2011} and the beamspace adaptive beamformer by Nilsen and Hafizovic \cite{Nilsen2009}.
 
@@
 
-Not phase aberration correction.
+Capon beamforming and focused systems.
 
-Add section about the computationally complexity. How many flops are required per rx-beam etc...
+The Capon beamformer is derived for narrowband, far-field applications and has typically been applied on medical ultrasound imaging "as is" even tough medical ultrasound imaging is a broadband, near-field imaging application.
+Later in this section we will explain why a broadband version of the Capon beamformer will have little effect.
+First we will take a closer look at the details of Capon beamforming.
+
+Intro to adaptive beamforming. List other variants (LCA, beamspace, Eigen space etc.).
 

### 2014-03-15 ThesisJP/Background/background.tex
Writing section on adaptive beamforming.

@@
 A ultrasound probe has normally a relative bandwidth between 50 and 100 \%.
-This corresponds to transmitting a pulse which are between 2 and 1 oscillation long.
+This corresponds to transmitting a pulse which is between 2 and 1 oscillations when entering tissue.
 In the front-end of an ultrasound imaging system a high sampling frequency is typically used (e.g.
@@
 \nomi{$N$}{Number of samples in range}where \nom{$B$}{Frequency bandwidth} is the bandwidth, $B = B_wf_c$ (where \nom{$B_w$}{Relative bandwidth} is the relative bandwidth and \nom{$f_c$}{Center frequency of ultrasound pulse} is the center frequency of the transmitted pulse).
+Note that after this process the ultrasound data will be complex values.
 The total number of samples in an B-mode image is then
@@
 If one element sample is two bytes long, for our example, around 600 MB has to be moved on to the GPU every second.
-This is however less than the maximum bandwidth of PCI Express 2.0 (8 GB/s), even if the relative bandwidth, transmit frequency and number of array elements are increased with a total factor of 10.
+This is however less than the maximum bandwidth of PCI Express 2.0 (8 GB/s), even after the relative bandwidth, transmit frequency and number of array elements are increased with a total factor of 10.
+It is also exist newer versions of PCI Express which supports even higher transfer rates.
 
@@
 A modern system capable of 3D imaging will generate several tera bytes of data in the front end of the system during a clinical scan.
-The data rate from a 2D scan is "only" in order giga bytes per second.
+The data rate from a 2D scan is "only" in order of giga bytes per second prior to IQ-demodulation.
 To handle these amount of data, ultrasound imaging systems have relied heavily on application-specific integrated circuits (\nom{ASIC}{Application-specific integrated circuit}) for all steps in the processing pipeline.
@@
 for converting data from a polar to a Cartesian grid (known as scan conversion), have gradually been converted into software \cite{Guracar2013}.
-One of  the last remaining ASICs is the beamformer, whose job is to steer and focus the ultrasound beam in space.
+One of  the last remaining ASICs in ultrasound scanners is the beamformer, whose job is to steer and focus the ultrasound beam in space.
 This operation involves a reduction of the data rate with a factor equal to the number of elements in the ultrasound probe.
 The input data rate is therefor very high.
-Nevertheless, real-time beamforming has recently been achieved on a GPU \cite{Song2012}, and a French company (SuperSonic Imagine) are utilizing GPUs in their plane wave beamforming algorithm \cite{Tanter2014}.
-Beamforming is therefore finally moving into software too.
+Nevertheless, real-time beamforming has recently been achieved on a GPU \cite{Song2012}, and a French company (SuperSonic Imagine) are for instance utilizing GPUs in their plane wave beamforming algorithm \cite{Tanter2014}.
+This is the same technique as reported implemented on a GPU by Yiu \textit{et al.} \cite{Yiu2011}.
+We therfore can conclude that beamforming is finally moving into software too.
 The largest players in the medical ultrasound business, General Electric and Philips, are also well aware of where the trend is going \cite{Thomenius2012} \cite{Metz2011}.
 Software beamforming opens up new possibilities for advanced processing of ultrasound data.
-One example is Capon beamforming which is explored in this thesis (\textbf{Paper\,I} and \textbf{II}).
+One example is Capon beamforming which is explored in this thesis (\textbf{Paper\,I} and \textbf{II}), and which we will introduce in the next section.
 The increased computational power introduced by GPUs makes it also possible to include advanced 3D visualization as presented in \cite{solteszova2010multidirectional}, \textbf{Paper\,IV}, and \textbf{IX}.
@@
 \subsection{Adaptive beamforming}\label{sec:adaptbf}
-Adaptive beamforming has several meanings in the medical ultrasound literature.
+Adaptive beamforming has several meanings in the literature on medical ultrasound imaging.
 It can for instance refer to aberration correction which is the task of applying time delays to compensate for the distortion introduced by fatty layers in the human body.
@@
 
-The Capon beamformer produces a weight vector, as introduced in (\ref{eq:das}), based on the impinging signal and an optimization criterion. The problem is as follow \cite{Capon1969}:
+The Capon beamformer produces a weight vector, as introduced in (\ref{eq:das}), based on the impinging signal and a constrained optimization problem. The problem is as follow \cite{Capon1969}:
 \begin{align}
@@
 Hence, the resulting weight vector minimizes the output power (the criterion) while maintaining unit gain in the steering direction $\vec{a}$ (distortionless response).
-The benefit of adaptive beamforming is clearly seen from Fig\,\ref{fig:weights} which shows a sensitivity map (an \textit{array beampattern}) for DAS beamforming with uniform weights and the Capon beamformer.
-In this simulated case an interfering source is located at 20\degree{} which for the DAS beamformer with uniform weights (\ref{fig:das_weights}) will leak into the beamformer output with its power attenuated with around $13$ dB.
-Since the Capon weights are data-dependent the method "knows" where the interfering source is located and is able to position a zero in the direction of it.
+The benefit of Capon beamforming is clearly seen from Fig\,\ref{fig:weights} which shows a sensitivity map (an \textit{array beampattern}) for DAS beamforming with uniform weights and the Capon beamformer.
+In this simulated case an interfering source is located at 20\degree{} which for the DAS beamformer with uniform weights (\ref{fig:das_weights}) will leak into the beamformer output with its power attenuated with around $13$ dB (first sidelobe).
+Since the Capon weights are data-dependent the method "knows" where the interfering source is located and is able to position a zero in the direction of it (Fig.\,\ref{fig:capon_weights}).
 Hence, the interfering source will be highly attenuated independent of its direction.
@@
 In theory a signal will be regarded by the Capon beamformer as an interfering source as long as the signal propagation vector does not match the steering vector ($\vec{a}$) exactly.
-If it is not possible to move a zero to the direction of the interferer the mainlobe will get shifted if this results in minimum output power.
+If it is not possible to move a zero to the direction of the interferer the mainlobe will get shifted and scaled if this results in minimum output power.
 Adaptive positioning of zeros and mainlobe shifting gives rice to the super-resolution capabilities of the Capon beamformer \cite{Synnevag2007}, which is the main reason why Capon beamforming has been highly studied in the field of medical ultrasound imaging the last decade.
@@
 
-The solution to the minimization problem in (\ref{eq:capon_optimization_criteria}) is, by the method of Lagrange multipliers \cite{VanTrees2003}, found to be
+The solution to the constrained minimization problem in (\ref{eq:capon_optimization_criteria}) is, by the method of Lagrange multipliers \cite{VanTrees2003}, found to be
 \begin{align}\label{eq:capon_weights}
@@
 Since $\vec{w}$ is time (or range) dependent one matrix has to be constructed and inverted for each data vector $\vec{x}[n]$ received by the system.
-This is indeed computationally demanding, something which becomes clear when we take a closer look at each step of the beamformer.
-Construction of the covariance matrix has a computationally complexity of $O(L^2N_LN_K)$\nomi{O(.)}{Big O notation (algorithm complexity)}, inverting it has a computationally complexity of $O(L^3)$, and the final normalization and calculation of $\vec{w}$ is $O(L)$.
+This is indeed computationally demanding, something which becomes clear when each step of the beamformer is closer examined.
+It is then found that construction of the covariance matrix has a computationally complexity of $O(L^2N_LN_K)$\nomi{O(.)}{Big O notation (algorithm complexity)}, inverting it has a computationally complexity of $O(L^3)$, and the final normalization and calculation of $\vec{w}$ is $O(L)$.
 For $L=M/2$, a subarray length often used in the literature, the first two steps both have a complexity of $O(L^3)$.
@@
 \end{align}
-Which for our example, with $L=M/2=64$, sums up to a required throughput of more than $0.5$ TFLOPS.
+Which for our example, with $L=M/2=64$, sums up to a required throughput of more than $1/2$ TFLOPS.
 Even though this is just 10\% of the peak throughput of modern GPUs remember that the latter where only theoretical numbers, and that the actual throughput is highly algorithm dependent.
-The required throughput for Capon beamforming is, for the moment, hard to achieve in practice \cite{So2011}.
-As for DAS beamforming this number also scales with higher relative bandwidth, pulse frequency, and more array elements since subarray size typically is proportional to the number of array elements.
+The required throughput for Capon beamforming might, for the moment, be hard to achieve in practice \cite{So2011}.
+As for DAS beamforming this number also scales with higher relative bandwidth, pulse frequency, and more array elements since the subarray size typically is proportional to the number of array elements.
 In the literature on medical ultrasound imaging we find few attempts on implementing a real-time Capon beamformer for medical ultrasound imaging.
@@
 
-To reduce the processing requirement for Capon beamforming several simplifications have been proposed \cite{Asl2012, Kim}.
-The most interesting proposals include the low complexity beamformer by Synnev\aa{}g \textit{et al.} \cite{Synnevag2011} and the beamspace adaptive beamformer by Nilsen and Hafizovic \cite{Nilsen2009}.
-
-Beamspace data is typically refers to the polar grid that cardiac ultrasound data is located in prior to scan conversion.
-In combination with the Capon beamformer, beamspace refers to the K-space representation of the impinging signals (hence the \nom{FFT}{Fast fourier transform} of the channel data).
+To reduce the processing requirement of the Capon beamformer for medical ultrasound imaging several simplified version have been proposed \cite{Asl2012, Kim} in the literature.
+Among the most interesting proposals we find the low complexity data-dependent beamformer by Synnev\aa{}g \textit{et al.} \cite{Synnevag2011} and the beamspace (BS) Capon beamformer by Nilsen and Hafizovic \cite{Nilsen2009}.
+In medical ultrasound imaging beamspace typically refers to the polar grid that cardiac ultrasound data is located in prior to scan conversion.
+In combination with the Capon beamformer, beamspace refers to the wavenumber-space representation of the recorded signals, hence, the spatial Fourier transform of element data.
+A GPU implementation of the BS-Capon beamformer for real-time cardiac ultrasound imaging is presented in \textbf{Paper II}.
+The complexity of BS-Capon beamforming is dominated by the transformation of data into beamspace which has a computational complexity of $O(N_LLN_b)$, where \nom{$N_b$}{Number of selected beams in beamspace} is the number of beams selected in beamspace.
+The initial complexity of BS-Capon beamforming before any tricks are performed is then found to be:
+\begin{align}
+\text{FLOPS}_{\text{BS-Capon}} = L^2N_bB_wf_c
+\end{align}
+For our example BS-Capon beamforming, with $L=M/2=64$ and $N_b=3$, will require a throughput of around $30$ GFLOPS.
+This is a level of processing which both modern CPUs and GPUs should be able to cope with.
+The rationale behind using the GPU is that a highly parallel task like the Capon beamformer should be offloaded to the processor made for parallel processing.
+The CPU is then left to do managing and all the task which by tradition occupies it.
 
@@
 First we will take a closer look at the details of Capon beamforming.
+For the low complexity data-dependent beamformer all the shifted windows are shifted less than the mainlobe width.
 

### 2014-03-16 HowtoHowtoCapon/submission/submit1/answer/answer.tex
la2pdf commit - 2014:03:16	19:56:37	(Sun16.Mar)

@@
 \ai{You are right that subarray averaging is important in our study.
-But we believe that "subarray MVDR" would be misleading, as we use subarray averaging mainly to deal with the coherent echoes in active sonar.
-We have now stated this better up front in the article.}
+But we believe that "subarray MVDR" would be misleading, as to the best of our knowledge this term is usually used to describe the use of subarrays to reduce complexity, while we average over subarrays to deal with the coherent echoes present in \emph{active} systems.
+We still apply the weights to the full array.
+To avoid confusion we have now stated this better up front in the article.}
 %
@@
 The abstract should be revised to indicate up to an order of magnitude over an optimized C implementation.}
-\ai{Thank you for this remark. We agree fully, and will state this better up front in the abstract. TODO}
+\ai{We agree with the reviewer that this should be improved.
+We still state the two order of magnitude improvement over our reference implementation as this is the total improvement we obtained, but also state that this figure would likely be 5-10 had a similar effort been carried out on the CPU side as well.}
 %

### 2014-03-16 HowtoHowtoCapon/submission/submit1/answer/answer.tex
la2pdf commit - 2014:03:16	20:50:01	(Sun16.Mar)

@@
 \ai{We agree with the reviewer that this should be improved.
-We still state the two order of magnitude improvement over our reference implementation as this is the total improvement we obtained, but also state that this figure would likely be 5-10 had a similar effort been carried out on the CPU side as well.}
+We still state the two order of magnitude improvement over our reference implementation as this ``total payoff'' of our efforts, but also state that this figure would likely be 5-10 had the CPU side received similar treatment.}
 %
@@
 685-696).}
-\ai{Thank you providing us with this interesting reference.
-However, we can not see that the technique described by Chapman is relevant for us.
-He is not using overlapping subarrays like we do, and obtains a sparse array with associated grating lobe problems.
-We use of subarrays to (1) deal with the coherency problem in active sonar, and (2) to provide robustness to our covariance estimate.
-The latter is particularly important in our near single snapshot application.
-We attempt to emphasize this further in the report, and suggest [9] Kailath and Shan, "Adaptive beamforming for coherent signals and interference", as well as R.T.Hoctor and S.A.Kassam, "The unifying role of the coarray in aperture synthesis for coherent and incoherent imaging" as good reads on the subject.
-}
+\ai{We thank the reviewer for providing this reference.
+However, unlike the subarray technique described by Chapman, which seems focused on reducing array complexity but in a way that causes grating lobe trouble, we use overlapping subarrays to (1) deal with the coherency problem in active sonar, and (2) to improve the rank of covariance matrix.
+The latter is particularly important in our case since we have hardly any temporal sample support.
+The Chapman reference,  distinction and a reference to [9] Kailath and Shan, "Adaptive beamforming for coherent signals and interference" is now included in the introduction to hopefully avoid future confusion.}
 %
@@
 \qi{Should reference the literature (from above comment 6) on the subarray approximation in paragraph 3.}
-\ai{Certainly. Reference added.}
+\ai{We agree that that this paragraph lacks a reference, but feel that [9] Kailath and Shan "Adaptive beamforming for coherent signals and interference" better describes this method in the context of active sonar.
+The Chapman reference is added to the introduction instead.}
 %
@@
 Figure 5 shows "relative complexity reduction" of entire pro-cess, so there is no information on relative speed of build and inversion steps.
-Figure 10 shows estimated FLOPs for the two steps, but since both steps have different theoretical complexities (operations counts), and the two steps do not optimize the same way, this does not provide an expec-tation of the relative times.}
+Figure 10 shows estimated FLOPs for the two steps, but since both steps have different theoretical complexities (operations counts), and the two steps do not optimize the same way, this does not provide an expectation of the relative times.}
 \qi{On page 6, authors say their implementation was improved by factor of 1.5 - 2 when run on a K20.
@@
 \qi{Reference 9 has two authors, Shan and Kailath. 11. Reference 17 seems to be the reviewed manuscript? }
+\ai{Reference 9 has been corrected. }
 \end{enumerate}
@@
 \item Fig. 3 - Slightly increased spacing between subplots.
-\item Fig. 8 - Abbreviating megapixels as "MP" in stead of "MPx".
-\item Fig. 9 - Abbreviating megapixels as "MP" in stead of "MPx".
-\item Fig. 11 - Abbreviating megapixels as "MP" in stead of "MPx".
+\item Fig. 8 - Abbreviating megapixels as "MP" instead of "MPx".
+\item Fig. 9 - Abbreviating megapixels as "MP" instead of "MPx".
+\item Fig. 11 - Abbreviating megapixels as "MP" instead of "MPx".
 \end{enumerate}

### 2014-03-16 ThesisJP/Background/background.tex
Now, only VR and Feild simulations sections remains. And some iterations with correcting language.

@@
 The specialized core can be a GPU or e.g.
-a field-programmable gate array (\nom{FPGA}{Field-programmable gate array}).
+a field-programmable gate array (\nom{FPGA}{Field-programmable gate array (processor)}).
 A recent trend is that specialized accelerator cores are embedded onto the CPU.
@@
 Software beamforming opens up new possibilities for advanced processing of ultrasound data.
-One example is Capon beamforming which is explored in this thesis (\textbf{Paper\,I} and \textbf{II}), and which we will introduce in the next section.
-The increased computational power introduced by GPUs makes it also possible to include advanced 3D visualization as presented in \cite{solteszova2010multidirectional}, \textbf{Paper\,IV}, and \textbf{IX}.
+One example is Capon beamforming which is explored in this thesis (\textbf{Paper\,I} and \textbf{II}), and which is introduce in the next section.
+The increased computational power introduced by GPUs makes it also possible to include more advanced 3D visualization as presented in \cite{solteszova2010multidirectional}, \textbf{Paper\,IV}, and \textbf{IX}.
 
@@
 The optimization criterion which we focus on in this thesis is the minimum variance distortionless response criterion, which is also known as Capon's method \cite{Capon1969}.
-A beamformer which uses weights derived using Capon's method will in this thesis mainly be referred to as the \textit{Capon beamformer}.
-However, it is also known as the minimum variance  (\nom{MV}{Minimum variance beamformer}) beamformer or the minimum variance distortionless respons beamformer (\nom{MVDR}{Minimum variance distortionless respons beamformer (Capon)}).
+A beamformer which uses weights derived using Capon's method will in this thesis mainly be referred to as the \textit{Capon beamformer} \cite{Vignon2008}.
+However, it is also known as the minimum variance  (\nom{MV}{Minimum variance beamformer (Synonym for Capon)}) beamformer \cite{Synnevag2007}, the minimum variance distortionless respons beamformer (\nom{MVDR}{Minimum variance distortionless respons beamformer (Synonym for Capon)}) (\textbf{Paper I}), or the constrained adaptive beamformer \cite{Mann2002}.
 
@@
 \end{align}
-where \nom{$\mat{\hat{R}}$}{Sample covariance matrix (with regularization)} is a sample covariance matrix.
+where \nom{$\mat{\hat{R}}$}{Sample covariance matrix (with regularization)} is a sample covariance matrix, and \nom{$(.)^H$}{Complex conjugate transpose} is the complex conjugate transpose.
 Hence, the resulting weight vector minimizes the output power (the criterion) while maintaining unit gain in the steering direction $\vec{a}$ (distortionless response).
@@
 In theory a signal will be regarded by the Capon beamformer as an interfering source as long as the signal propagation vector does not match the steering vector ($\vec{a}$) exactly.
-If it is not possible to move a zero to the direction of the interferer the mainlobe will get shifted and scaled if this results in minimum output power.
-Adaptive positioning of zeros and mainlobe shifting gives rice to the super-resolution capabilities of the Capon beamformer \cite{Synnevag2007}, which is the main reason why Capon beamforming has been highly studied in the field of medical ultrasound imaging the last decade.
+If it is not possible to position a zero in the direction of the interferer the mainlobe will get shifted and scaled if this is whats needed to achieve minimum output power.
+Adaptive positioning of zeros and mainlobe shifting gives rice to the super-resolution capabilities of the Capon beamformer \cite{Synnevag2007}, and the positive implications of this effect \cite{Synnevag2009} is the main reason why this beamformer has been highly studied in the field of medical ultrasound imaging the last decade.
 
@@
 \end{align}
-where $K = M-L+1$\nomi{$K$}{Temporal averaging parameter (Capon)} typically is proportional to the pulse length, $N_K = 2K + 1$ \nomi{$N_K$}{Number of samples used for temporal averaging (Capon)}, $N_L = M-L+1$,\nomi{$N_L$}{Number of subarrays used for spatial averaging (Capon)}\nomi{$L$}{Subarray length (Capon)} and $\vec{x}_l$ is a $L$ long subarray $[x_l[n], \dotso x_{l+L}[n]]$.
-To ensure robustness against model errors and numerical stability the matrix is loaded with a diagonal factor,
+where\nomi{$K$}{Temporal averaging parameter (Capon)} $N_K = 2K + 1$ \nomi{$N_K$}{Number of samples used for temporal averaging (Capon)} is the number of samples used for temporal smoothing,\nomi{$L$}{Subarray length (Capon)} $\vec{x}_l$ is a $L$ long subarray $[x_l[n], \dotso, x_{l+L}[n]]$, and $N_L = M-L+1$\nomi{$N_L$}{Number of subarrays used for spatial averaging (Capon)} is the number of subarrays used for spatial averaging.
+Temporal averaging is used to obtain DAS-like speckle statistics in combination with large subarrays \cite{Synnevag2007a}, and subarray averaging is applied to reduce signal cancellation caused by coherent signals \cite{Reddy1987}.
+The number of temporal samples is usually selected relative to the pulse length.
+To ensure robustness against model errors and numerical stability the matrix is further loaded with a diagonal factor,
 \begin{align}
@@
 \end{align} 
-Thus, a white noise component proportional to the in-coherent signal power is added to the estimate.
+Hence, a white noise component proportional to the in-coherent signal power is added to the estimate \cite{Featherstone1997b}.
 
@@
 Since $\vec{w}$ is time (or range) dependent one matrix has to be constructed and inverted for each data vector $\vec{x}[n]$ received by the system.
-This is indeed computationally demanding, something which becomes clear when each step of the beamformer is closer examined.
-It is then found that construction of the covariance matrix has a computationally complexity of $O(L^2N_LN_K)$\nomi{O(.)}{Big O notation (algorithm complexity)}, inverting it has a computationally complexity of $O(L^3)$, and the final normalization and calculation of $\vec{w}$ is $O(L)$.
-For $L=M/2$, a subarray length often used in the literature, the first two steps both have a complexity of $O(L^3)$.
-These steps should therefore receive the same amount of attention when a fast implementation of Capon beamforming is attempted.
+This is indeed computationally demanding, something which becomes clear when each step of the beamformer is examined closer.
+It is then found that constructing the covariance matrix has a computationally complexity of $O(L^2N_LN_K)$\nomi{O(.)}{Big O notation (algorithm complexity)}, inverting it has a computationally complexity of $O(L^3)$, and the final normalization and calculation of $\vec{w}$ is $O(L)$.
+For $L=M/2$, a subarray size often used in the literature, the first two steps both have a complexity of $O(L^3)$.
+These steps should therefore receive the same amount of attention when a real-time implementation of Capon beamforming is attempted.
+
 A rough estimate of the number of FLOPS required by a Capon beamformer for real-time cardiac ultrasound imaging is then given by:
 \begin{align}
-\text{FLOPS}_{\text{Capon}} = L^3N_{\text{B-mode}}f_{\text{FPS}} = L^3B_wf_c.
+\text{FLOPS}_{\text{Capon}} = L^3B_wf_c.
 \end{align}
 Which for our example, with $L=M/2=64$, sums up to a required throughput of more than $1/2$ TFLOPS.
-Even though this is just 10\% of the peak throughput of modern GPUs remember that the latter where only theoretical numbers, and that the actual throughput is highly algorithm dependent.
-The required throughput for Capon beamforming might, for the moment, be hard to achieve in practice \cite{So2011}.
+Even though this is just 10\% of the peak throughput of modern GPUs remember that the latter are only theoretical numbers, and that the actual throughput is highly algorithm dependent.
+The required throughput for Capon beamforming might therefore still be hard to achieve in practice \cite[\textbf{Paper II}]{So2011}.
 As for DAS beamforming this number also scales with higher relative bandwidth, pulse frequency, and more array elements since the subarray size typically is proportional to the number of array elements.
+
 In the literature on medical ultrasound imaging we find few attempts on implementing a real-time Capon beamformer for medical ultrasound imaging.
-In addition to the papers in this thesis only Chen \textit{et al.} \cite{Chen2011, Chen, Chen2011a} are known to have explored GPU and FPGA implementations of the Capon beamformer for medical ultrasound imaging, however with some shortcomings  which are discussed in both \textbf{Paper I} and \textbf{II}.
+In addition to the papers in this thesis only Chen \textit{et al.} \cite{Chen2011, Chen, Chen2011a} are known to have explored GPU and FPGA implementations of the Capon beamformer for medical ultrasound imaging.
+The differences between our work and Chen \textit{et al.} is discussed in \textbf{Paper I}, \textbf{II}, and \textbf{VI}.
 
-To reduce the processing requirement of the Capon beamformer for medical ultrasound imaging several simplified version have been proposed \cite{Asl2012, Kim} in the literature.
-Among the most interesting proposals we find the low complexity data-dependent beamformer by Synnev\aa{}g \textit{et al.} \cite{Synnevag2011} and the beamspace (BS) Capon beamformer by Nilsen and Hafizovic \cite{Nilsen2009}.
+To reduce the processing requirement of the Capon beamformer for medical ultrasound imaging several simplified version have been proposed in the literature \cite{Asl2012, Kim}.
+Among the most interesting proposals we find the low complexity adaptive (LCA) beamformer by Synnev\aa{}g \textit{et al.} \cite{Synnevag2011} and the beamspace (BS) Capon beamformer by Nilsen and Hafizovic \cite{Nilsen2009}.
 In medical ultrasound imaging beamspace typically refers to the polar grid that cardiac ultrasound data is located in prior to scan conversion.
-In combination with the Capon beamformer, beamspace refers to the wavenumber-space representation of the recorded signals, hence, the spatial Fourier transform of element data.
-A GPU implementation of the BS-Capon beamformer for real-time cardiac ultrasound imaging is presented in \textbf{Paper II}.
-The complexity of BS-Capon beamforming is dominated by the transformation of data into beamspace which has a computational complexity of $O(N_LLN_b)$, where \nom{$N_b$}{Number of selected beams in beamspace} is the number of beams selected in beamspace.
-The initial complexity of BS-Capon beamforming before any tricks are performed is then found to be:
+In combination with the Capon beamformer, beamspace refers to the wavenumber-space representation of the recorded signals, hence, the spatial Fourier transform of the element data:
+\begin{align}
+&\mat{B}[x_l[n],\,\dotso,\,x_{l+L}[n]]^T\\
+&[\mat{B}]_{p,q} = \frac{1}{\sqrt{L}}e^{-j 2 \pi p q / L}
+\end{align}
+The idea behind BS-Capon beamforming is that spatial beams with little information can be removed from $\mat{\hat{R}}$ by removing rows in $B$.
+The complexity of BS-Capon beamforming is dominated by the transformation of data into beamspace which has a computational complexity of $O(N_LLN_b)$, where \nom{$N_b$}{Number of selected beams in beamspace} is the number of selected beams.
+
+The number of FLOPS required for BS-Capon beamforming of real-time cardiac ultrasound is then found to be:
 \begin{align}
@@
 \end{align}
-For our example BS-Capon beamforming, with $L=M/2=64$ and $N_b=3$, will require a throughput of around $30$ GFLOPS.
-This is a level of processing which both modern CPUs and GPUs should be able to cope with.
-The rationale behind using the GPU is that a highly parallel task like the Capon beamformer should be offloaded to the processor made for parallel processing.
-The CPU is then left to do managing and all the task which by tradition occupies it.
+For our example, BS-Capon beamforming with $L=M/2=64$ and $N_b=3$, will require a throughput of around $30$ GFLOPS.
+This is a level of processing which both modern CPUs and GPUs should be able to handle.
+A GPU implementation of the BS-Capon beamformer for real-time cardiac ultrasound imaging is presented in \textbf{Paper II}.
+The rationale behind using the GPU is that a highly parallel task like the Capon and BS-Capon beamformer should be offloaded to the processor made for parallel processing (heterogeneous computing).
+The CPU is then left to do managing and all the task which by tradition already occupies it.
+It should also be possible to achieve higher processing throughput on a GPU than on a CPU for this kind of problem.
 
-Capon beamforming and focused systems.
+Both the LCA and BS-Capon beamformer is based upon a detailed investigation of how Capon beamfoming behaves when applied to medical ultrasound images.
+For the LCA beamformer, the windows selected by the Capon beamformer where investigated and a special set of predefined windows where defined.
+If one look closer at the windows selected for the LCA beamformer we see that half the set performs micro-steering of the mainlobe and the other half provides different mainlobe widths.
+The results from applying the LCA beamformer shows that the main benefit of applying Capon beamforming to medical ultrasound imaging is micro-steering of the mainlobe, and not attenuation of far off-axis backscatter.
 
-The Capon beamformer is derived for narrowband, far-field applications and has typically been applied on medical ultrasound imaging "as is" even tough medical ultrasound imaging is a broadband, near-field imaging application.
-Later in this section we will explain why a broadband version of the Capon beamformer will have little effect.
-First we will take a closer look at the details of Capon beamforming.
-For the low complexity data-dependent beamformer all the shifted windows are shifted less than the mainlobe width.
+The Capon beamformer is derived for narrowband applications and has traditionally been applied on medical ultrasound imaging "as is".
+Medical ultrasound imaging is, however, a broadband imaging modality and one could think that a broadband Capon beamformer therefore where needed \cite{Holfort2009}.
+The investigation performed in \cite{Nilsen2009} shows that with focused beams, energy will be centered on-axis and as little as three near-axis beams is therefore adequate in order to achieve similar performance for BS-Capon and Capon beamforming.
+The two-way beamprofile will be wider with un-focused (plane wave) transmit beams \cite{Holfort2009, Austeng2011}, but focusing on receive will still make the main amount of energy arrive on axis.
+As it is shown in \textbf{Paper III} it is possible to micro-steer with phase delays instead of time delays without introducing large errors as long as the phase shift is smaller than the pulse length.
+Since Capon beamforming mainly performs main-lobe shifts smaller than the beam width, broadband Capon beamforming will only provide a slight improvement over narrowband Capon beamforming.
 
-Intro to adaptive beamforming. List other variants (LCA, beamspace, Eigen space etc.).
+%A final important point when working with Capon beamforming is how we present our results with respect to normalization.
+Since the Capon beamforming highly attenuate signals within the mainlobe  we can get a false impression of improvements in contrast if we normalize  \cite[Fig.\,6]{Synnevag2007}
+%\todo{Need figures here...}
 
-Add section about how to present data (max v.s mean etc.)
+%Examples of other adaptive apodization methods found in the literature on medical ultrasound imaging includes the APES beamformer, the MUSIC method, and the eigenvalue method. \todo{Need references}.
 
-\subsection{Shift invariance}
+%\subsection{Shift invariance}
 
-One sentence about speckle tracking.
+%One sentence about speckle tracking.
 
@@
 
-Get section from master theses. Ray casting and opacity functions.
+Get section from master theses. Ray casting and opacity functions. (Fig.\,\ref{fig:vr})
 

### 2014-03-16 ThesisJP/Introduction/introduction.tex
Now, only VR and Feild simulations sections remains. And some iterations with correcting language.

@@
 The latest graphics cards\footnote{As of January 2014.} from both \nom{Nvidia}{Graphics card producer} and \nom{AMD}{Graphics card producer} provide more than 5 T\nom{FLOPS}{Floating point operations per second} single precision and 1.5 TFLOPS double precision computations.
-For double precision this is more than the worlds most powerful supercomputer could provide in 1998.
+For double precision this is more than the worlds most powerful supercomputer could provide in 1998, and that was a room full of computers.
 
@@
 No wonder, to achieve real-time processing for e.g.
-cardiac ultrasound imaging, an effective processing rate of around 100 MFLOPS is required.
-This number, as explained in Section \ref{sec:adaptbf}, also grows with larger subarrays, an increasing number of samples, and higher frame rates.
+cardiac ultrasound imaging, an effective processing rate of hundreds of GFLOPS is required.
+This number, as explained in Section \ref{sec:adaptbf}, also grows with larger subarrays and an increasing number of samples per second.
 Nevertheless, with modern GPUs, these levels of processing are finally available within a single card.
@@
 The improvements seen are minimal, especially with the number of calculations in mind.
-The beamformers are clearly able to decrees the lateral width of lateral point scatterers and thin structures \textit{in vivo}, and to sharpen step edges.
+The beamformers are clearly able to decrees the lateral width of point scatterers and thin structures \textit{in vivo}, and to sharpen step edges.
 On the other hand, the overall contrast is visually not improved and sometimes it is worse (Fig.\,6b of \textbf{Paper\,II}).
@@
 If only minor improvements are obtained, it would be easier to ignore this fact if the algorithm is less computationally costly.
-Finally, it would also be interesting to see how the Eigenspace-based Capon beamformer performs on cardiac images, and how its absence of a distortionless criterion will influence the image spatially and temporally.
-The work in \textbf{Paper\,II} made it clear that interesting findings are done when users are able to watch the temporal behavior of an algorithm.
-This led us to the work of \textbf{Paper\,III}.
+%Finally, it would also be interesting to see how the Eigenspace-based Capon beamformer performs on cardiac images, and how its absence of a distortionless criterion will influence the image spatially and temporally.
+The work in \textbf{Paper\,II} made it clear that interesting findings are done when users are able to watch the temporal behavior of an algorithm. This led us to the work of \textbf{Paper\,III}. 
 \\\\
@@
 It also turns out that more matrix inversions are not needed either.
-Oversampling on receive will, however,  increase the penalty of applying post-processing, like filtering,  on scan-grid data.
+Oversampling on receive will, however,  increase the penalty of applying post-processing, like filtering,  on polar-grid data.
 %The time taken to convert data from a polar scan grid to a Cartesian display grid will also increase.
@@
 For example to increase the precision of speckle and edge tracking.
-In the case where the image is generated for post processing it does not need to be visually pleasant anymore, as long as signal cancellation is avoided.
+In the case where the image is generated for post processing it does not need to be visually pleasant anymore, as long as signal cancellation is avoided and the shift-invariance property is maintained.
 Good looking speckle, or DAS-like speckle, has been the driving force behind many of the robustification techniques and the default parameters seen in the literature.
@@
 The paper does also propose an interesting method for automatic tuning of rendering parameters.
-Figure 3d in \textbf{Paper\,IV} actually shows that an optimal threshold could be found based on global and not only local statistics.
-Even though the visualization based on global statistics has more errors, with respect to endocardium visibility, the tuning of the opacity function is still fully automatic if an iterative search for the minima in Figure 3d is performed.
+Fig.\,3d in \textbf{Paper\,IV} actually shows that an optimal threshold could be found based on global and not only local statistics.
+Even though the visualization based on global statistics has more errors, with respect to endocardium visibility, the tuning of the opacity function is still fully automatic if an iterative search for the minima in Fig.\,3d in \textbf{Paper\,IV} is performed.
 A global opacity transfer function will assure a rendering which are realistic, but it will still be dependent on a successful tracking for the auto-adjustment to work.
@@
 A local opacity transfer function is just to extreme to be clinically robust.
-It would also be interesting to see if statistics based on the window selected by the Capon or the LCA beamformer could be utilized to adjust the opacity.
+In addition, it would also be interesting to see if statistics based on the window selected by the Capon or the LCA beamformer could be utilized to adjust the opacity.
 It should also be possible to use this statistic as input to image filtering.
 An adaptively selected window will be symmetric in homogeneous regions and left-right shifted close to tissue interfaces.
-In that way we can apply less filtering and higher opacity on edges.
-If this will be different than just looking at neighboring pixels in a smoothing filter is to be seen.
+In that way we can apply less smoothing and higher opacity at tissue interfaces.
+If this will be different than just looking at neighboring pixels in an edge-preserving image filter is to be seen.
 \\\\
 The fifth paper stands out from the other papers by not dealing with an adaptive processing technique.
-It is still, however, centered around the topic of implementing algorithms for ultrasound imaging on GPUs.
+It is still, however, centered around the topic of implementing algorithms for ultrasound imaging on the GPU.
 Ultrasound simulations are typically used for research and sometimes also to pre-calculate configurations of a given scan sequence, for instance to achieve a certain pressure in a given area.
 The continuously growth in computational power will, in the future, make it possible to evaluate a scan sequence on the fly, removing the need for lookup tables.
-Estimation of the mechanical and thermal index could also be improved.
+Estimation of the mechanical and thermal index could then also be improved.
 Achieving higher processing speed for an excising algorithm by implementing it on a new architecture is not academically relevant in its own.

### 2014-03-16 HowtoHowtoCapon/submission/submit1/answer/answer.tex
la2pdf commit - 2014:03:16	23:46:31	(Sun16.Mar)

@@
 \ai{We agree with the reviewer that this should be improved.
-We still state the two order of magnitude improvement over our reference implementation as this ``total payoff'' of our efforts, but also state that this figure would likely be 5-10 had the CPU side received similar treatment.}
+We still state the two order of magnitude improvement over our reference implementation as a "total payoff" for our efforts, but now also state that this figure would likely be 5-10 had the CPU side received similar treatment.}
 %
@@
 \qi{"which are a mere scaled version" should be "which are merely scaled versions".}
-\ai{We thank the reviewer for these corrections (8.-10.). Corrections applied.}
+\ai{We thank the reviewer for these corrections (8.-10.). They are all applied.}
 %

### 2014-03-16 HowtoHowtoCapon/submission/submit1/answer/answer.tex
Calling it a night

@@
 We still apply the weights to the full array.
-To avoid confusion we have now stated this better up front in the article.}
+To avoid confusion we have now stated this better up front in the article (section I paragraph 5).}
 %
@@
 Perhaps breaking the two rows of sub-plots into two figures would work better.}
-\ai{Yes, the initial plots had the total GPU processing time included. But since there are three versions of the covariance building step, there TODO }
+\ai{The reviewer has a point.
+But since there are 3 occurrences of the covariance processing time, including the total time GPU processing time would introduce three new lines.
+We feel this would clutter the plot unnecessarily as it is not too hard to derive the total time by looking at the current plots.
+Putting the total processing time in the MATLAB/C total time figure is an option, but then a logarithmic y-scale would necessary and detail would be lost.}
 %

### 2014-03-17 HowtoHowtoCapon/submission/submit1/answer/answer.tex
la2pdf commit - 2014:03:17	14:16:32	(Mon17.Mar)
la2pdf commit - 2014:03:17	15:04:37	(Mon17.Mar)

@@
 But since there are 3 occurrences of the covariance processing time, including the total time GPU processing time would introduce three new lines.
-We feel this would clutter the plot unnecessarily as it is not too hard to derive the total time by looking at the current plots.
+We feel this would clutter the plot unnecessarily as it is fairly easy derive the total time by looking at the current plots.
 Putting the total processing time in the MATLAB/C total time figure is an option, but then a logarithmic y-scale would necessary and detail would be lost.}
@@
 I would suggest making a very clear itemization of the assumptions and caveats about the two approaches.}
-\ai{TODO}
+\ai{We agree that this should hould be clarified, and have We disagree that this }
 %
@@
 \ai{Yes! This is likely the reason. Thank you for pointing this out. This will be added to the conclusions.}
-\qi{Bottom of page, "hold- ups" has an extra space after the hyphen.}
+\qi{Bottom of page, "hold-ups" has an extra space after the hyphen.}
 \ai{Fixed.}
@@
 Perhaps the authors can introduce these variants more clearly and more prominently.}
+%
 \qi{On page 6, authors say "The bottleneck in the final design is now the in-version step, which is typically 5 times slower than the build step".

### 2014-03-17 HowtoHowtoCapon/submission/submit1/answer/answer.tex
la2pdf commit - 2014:03:17	15:57:50	(Mon17.Mar)

@@
 clearly not...
-do the authors pre-steer all of the elements prior to forming R so that can "focus" in a wideband sense?}
-\ai{We indeed presteer all the elements. TODO}
+do the authors pre-steer all of the elements prior to forming $\R$ so that can "focus" in a wideband sense?}
+\ai{We indeed presteer all the elements.
+To make this clearer we have now stated it again after equation (2), and explained that a generalization is straight forward by substituting $\1$ with a steering vector.}
 %

### 2014-03-17 HowtoHowtoCapon/submission/submit1/answer/answer.tex
la2pdf commit - 2014:03:17	16:42:50	(Mon17.Mar)

@@
 %
-\qi{Paragraph after equation (2) mentions "spatial averaging to avoid signal cancellation" - this problem might be better identified for readers by cit-ing the Kailath reference and using "signal cancellation due to coherent multipath".}
+\qi{Paragraph after equation (2) mentions "spatial averaging to avoid signal cancellation" - this problem might be better identified for readers by citing the Kailath reference and using "signal cancellation due to coherent multipath".}
 \ai{Thank you, both suggestions are now included.}
@@
 is $\R$ averaged using a moving window process, so that $\R[n]$ is just a rank-1 update to $\R$[n — 1] and there is one beamformer output for every input time sample?}
-\ai{Actually, $\R$ is averaged over $2K+1$ time samples, this can be seen from (4). TODO}
+\ai{Actually, $\R$ is averaged over $2K+1$ time samples, this can be seen from (4) it in the text that follows. TODO}
 %
 \qi{Page 4, in list of three optimizations - item 3 is not self-evident, but if I understand what the authors are doing, it is described by the 1985 reference by Kailath.}
-\qi{Page 4, last paragraph of 1) Arithmetic reduction - authors should iden-tify that the "Minimized memory" version skips step 2) from list of three optimization, and that the "Minimized instructions" version uses all three steps...
-I had a hard time understanding how the three "arithmetic reduc-tion steps" were related to the plots in Figure 5, and was puzzled about where the "minimized memory" and "minimized instruction" variants of the processing had come from.
+\ai{We must have conveyed this in a flawed way.
+We are simply implementing a text-book sliding window summation, where a start sum is computed and then this is updated with a new term for each element in $\eR$.
+Item 3 has now been rewritten to hopefully express this better.}
+%
+\qi{Page 4, last paragraph of 1) Arithmetic reduction - authors should identify that the "Minimized memory" version skips step 2) from list of three optimization, and that the "Minimized instructions" version uses all three steps...
+I had a hard time understanding how the three "arithmetic reduction steps" were related to the plots in Figure 5, and was puzzled about where the "minimized memory" and "minimized instruction" variants of the processing had come from.
 Perhaps the authors can introduce these variants more clearly and more prominently.}
+\ai{The reviewer is right. This is now rewritten to be more explicit.}
 %

### 2014-03-17 HowtoHowtoCapon/submission/submit1/answer/answer.tex
la2pdf commit - 2014:03:17	17:34:18	(Mon17.Mar)
la2pdf commit - 2014:03:17	17:35:16	(Mon17.Mar)

@@
 %
-\qi{On page 6, authors say "The bottleneck in the final design is now the in-version step, which is typically 5 times slower than the build step".
+\qi{On page 6, authors say "The bottleneck in the final design is now the inversion step, which is typically 5 times slower than the build step".
 Surely, this merits an explanation.
@@
 \qi{Section I says 1-2 orders of magnitude. Section IV, page 6, says 2-3 orders of magnitude.}
-\qi{Reference 9 has two authors, Shan and Kailath. 11. Reference 17 seems to be the reviewed manuscript? }
-\ai{Reference 9 has been corrected. }
+\ai{Well seen. This is now corrected.}
+\qi{Reference 9 has two authors, Shan and Kailath. 11. Reference 17 seems to be the reviewed manuscript?}
+\ai{Reference 9 has been corrected, and reference 17 (now 18) has been updated to point to the correct article. }
 \end{enumerate}

### 2014-03-17 HowtoHowtoCapon/submission/submit1/answer/answer.tex
End of day(s)

@@
 Here too, the authors should perhaps use the "coalesced" term, since this is the term used in all of the GPGPU documentation.}
-\ai{You are right. Initially we sought to avoid as much GPU jargon as we could, but in its current form it probably just adds confusion to leave it out. Your suggestion sounded well.}
+\ai{You are right. It is better to use the "official" terms. We have applied your suggestions.}
 \end{enumerate}
@@
 is $\R$ averaged using a moving window process, so that $\R[n]$ is just a rank-1 update to $\R$[n — 1] and there is one beamformer output for every input time sample?}
-\ai{Actually, $\R$ is averaged over $2K+1$ time samples, this can be seen from (4) it in the text that follows. TODO}
+\ai{Actually, $\R$ is averaged over $2K+1$ time samples, this can be seen from (4) and in the text that follows.
+There is indeed one beamformer output for every input time sample.
+We are not sure where the article is indicating otherwise...}
 %

### 2014-03-18 HowtoHowtoCapon/submission/submit1/answer/answer.tex
la2pdf commit - 2014:03:18	09:10:45	(Tue18.Mar)
la2pdf commit - 2014:03:18	09:36:13	(Tue18.Mar)

@@
 This is really the key point to be made in the comparison between the two approaches, in my opinion.}
-\ai{Yes! This is likely the reason. Thank you for pointing this out. This will be added to the conclusions.}
+\ai{Yes! This is likely the reason. Thank you for pointing this out. This is now added to the discussion.}
 \qi{Bottom of page, "hold-ups" has an extra space after the hyphen.}
@@
 I see this disconnect between low-frequency passive processors and mid-frequency active processors, but the gap to imaging sonars is even greater.
-I venture that the authors can greatly broaden the appeal of their paper by addressing the questions I pose below.}
+I venture that the authors can greatly broaden the appeal of their paper by addressing the questions I pose below.}\\[-.5\baselineskip]
+
+\ans{It is clear we had to state this better upfront. In the 5th paragraph of the introduction we now explain that our system is a well sampled wideband sonar.}
 

### 2014-03-18 HowtoHowtoCapon/submission/submit1/answer/answer.tex
la2pdf commit - 2014:03:18	09:44:51	(Tue18.Mar)

@@
 \end{align}
-Further details are provided in 2.}
+Further details are provided in 2. A sentence elaborating on the element spacing has been added in section IV first paragraph; where the HISAS1030 is introduced.}
 %
@@
 In the paper we addition we perform sidescan imaging where only very narrow beams need to be beamformed.
-These details are now included in the paper.}
+These details are now included in the paper, some in the 5th paragraph of the introduction and some in the first paragraph of section IV.}
 %

### 2014-03-18 HowtoHowtoCapon/submission/submit1/answer/answer.tex
...

@@
 \ai{Thank you.
-We mention in the beginning of section IIA that we pre-steer to every angle and range bin in the image, but will try to express this with greater clarity and emphasize that we are operating in the near field.}
+We mention in the beginning of section IIA that we pre-steer to every angle and range bin in the image.
+but will try to express this with greater clarity and emphasize that we are operating in the near field.}
 %

### 2014-03-18 HowtoHowtoCapon/submission/submit1/answer/answer.tex
la2pdf commit - 2014:03:18	11:24:19	(Tue18.Mar)

@@
 
-\ans{It is clear we had to state this better upfront. In the 5th paragraph of the introduction we now explain that our system is a well sampled wideband sonar.}
+\ans{Thank you, it seems we should state this better upfront. In the 5th paragraph of the introduction we now explain that our system is a well sampled wideband sonar.}
 
@@
 \ai{Thank you.
-We mention in the beginning of section IIA that we pre-steer to every angle and range bin in the image.
-but will try to express this with greater clarity and emphasize that we are operating in the near field.}
+We mention in the beginning of section IIA that we pre-steer to every angle and range bin in the image, but now repeat this in IIA paragraph 3, and state that we operate in wideband conditions in the introduction.}
 %
@@
 Are the authors pre-steering the array for every angle?
-Or are they using $w = \frac{\Ri\s}{\s\t\Ri\s}$ at off-broadside angles, with $\s$ the conventional steering vector (which would be a frequency-domain phase-shift vector, to be applied to a narrowband signal)?
+Or are they using $\w = \frac{\Ri\s}{\s\t\Ri\s}$ at off-broadside angles, with $\s$ the conventional steering vector (which would be a frequency-domain phase-shift vector, to be applied to a narrowband signal)?
 Pre-steering would enable disparate frequencies over a wide band to be coherently summed, to provide coherent gain in the look direction (over the band) by providing additional sample support for the sample covariance estimation.}

### 2014-03-18 HowtoHowtoCapon/submission/submit1/answer/answer.tex
la2pdf commit - 2014:03:18	11:33:08	(Tue18.Mar)

@@
 \ai{Yes, we are pre-steering the array for every angle.
-We will attempt to clarify this further, and add a remark saying that while not necessary here we could have phase steered the array by introducing the steering vector $\vec s$.}
+We have clarified this further (see 3.), and added a remark saying that while not necessary here we could have phase steered the array by introducing the steering vector $\vec s$.}
 %
@@
 \ai{Again, thank you for a helpful remark, we will clarify this as well.
-After pulse compressing a chirp using the full transceiver bandwidth (30~kHz) range resolution is roughly $\Delta r \approx \frac{1}{B} = 2.5$~cm, which corresponds to a sample shift of $\Delta n = \frac{\Delta r f_s}{c} = 0.6$.
+After pulse compressing a chirp using the full transceiver bandwidth (30~kHz) range resolution is roughly $\Delta r \approx \frac{1}{B} = 2.5$~cm, which corresponds to a sample shift of just 1.2\footnote{$\Delta n = \frac{\Delta r f_s}{c} = 1.2$}.
 Hence we cover the entire pulse compressed signal with only a few samples.

### 2014-03-18 HowtoHowtoCapon/submission/submit1/answer/answer.tex
la2pdf commit - 2014:03:18	12:25:52	(Tue18.Mar)

@@
 Hence we cover the entire pulse compressed signal with only a few samples.
-$K=2$ corresponds to a window length of $2K+1=5$ samples, as is shown by (4) in the article.}
+$K=2$ corresponds to a window length of $2K+1=5$ samples, as is shown by (4) in the article.
+Some edits have been made to make this clearer.}
 %
@@
 We operate under near field conditions and are indeed performing wavefront curvature beamforming.
-Hence, the index $n$ just an index over time with an undefined offset, at a resolution given by the sampling frequency.
-We will clarify this in the paper.}
+Hence, the index $n$ just an index over time at a resolution given by the sampling frequency.
+This has now been clarified.}
 \end{enumerate}
@@
 \ai{Thank you.
-We share these concerns but decided to leave CPU optimizations out of the paper.
-One reason is that we have not been able to find any other batch solvers out there and lacked the time to make our own.
+We have now included our reasons for not comparing with an optimized CPU implementaiton.
+One reason is that we have not been able to find any CPU batch solvers out there and lacked the time to make our own.
 In our tests our custom inline Cholesky solver beat alternatives from e.g.
@@
 Another reason is that to make the comparison fair we feel an equal effort would have to put into writing a covariance builder for the CPU, and to minimize data movement in each step and in between them.
-Considering your remarks we will include these reasons in the paper.}
+}
 %

### 2014-03-18 HowtoHowtoCapon/submission/submit1/answer/answer.tex
la2pdf commit - 2014:03:18	13:21:37	(Tue18.Mar)

@@
 \ai{Thank you.
-We have now included our reasons for not comparing with an optimized CPU implementaiton.
+We have now included our reasons for not comparing with an optimized CPU implementation.
 One reason is that we have not been able to find any CPU batch solvers out there and lacked the time to make our own.
@@
 Another reason is that to make the comparison fair we feel an equal effort would have to put into writing a covariance builder for the CPU, and to minimize data movement in each step and in between them.
-}
+The details may be found in section V paragraph 2.}
 %
@@
 This should perhaps be pointed out.}
-\ai{Yes this is a good point. This is now included.}
+\ai{Yes this is a good point. Included in last paragraph of section IV.}
 %
@@
 For example, the authors tell us the Quadro 6000 (\$1900, 6GB VRAM, 448 cores) is a Fermi architecture - this is the previous architecture, having been superceded by the Kepler architecture, with a new Quadro K6000 board (\$3600, 12 GB VRAM, 2880 cores).
-The Quadro boards are high-end boards with video output targeting the CADAM market - there are many cheaper alternatives tar-geted at gangers and an even more expensive compute-only board called TESLA targeted at high performance computing.
-It's difficult to say how much of this irrelevant.
-}
-\ai{TODO}
+The Quadro boards are high-end boards with video output targeting the CADAM market - there are many cheaper alternatives targeted at gamers and an even more expensive compute-only board called TESLA targeted at high performance computing.
+It's difficult to say how much of this irrelevant.}
+\ai{There is certainly a potential for comparing with many other cards here, but we decided to be thorough with one architecture than to compare with everything.
+To the paragraph introducing the Quadro 6000 we have explained what its equivalent consumer grade Geforce card would be.
+We have also included remarks that we expect Kepler to perform better had the design been optimized for it.}
 %

### 2014-03-18 HowtoHowtoCapon/submission/submit1/answer/answer.tex
la2pdf commit - 2014:03:18	13:53:03	(Tue18.Mar)

@@
 It is unfortunate that no Cholesky based batch linear equation solver exist, which would have eliminated the need for the lower triangular...
-The paragraph will be improved to explain this better.}
+The paragraph has been edited to explain this better.}
 
@@
 
-The authors describe the methodology they used to get things to run fast on the GPU.
-The analysis of the workload was informative and I think it is very informative for readers to see how the analysis progresses from less hardware specific (operations counts), to algorithm restructuring to reduce operations; to more hardware specific concerns (relative throughput of operations and memory transfers) .
+\q{The authors describe the methodology they used to get things to run fast on the GPU.
+The analysis of the workload was informative and I think it is very informative for readers to see how the analysis progresses from less hardware specific (operations counts), to algorithm restructuring to reduce operations; to more hardware specific concerns (relative throughput of operations and memory transfers).}\\[.5\baselineskip]
+\ans{We thank the reviewer for the credit.}\\[-.5\baselineskip]
 
-Let me suggest some concepts that could be discussed a little bit more to give readers more of a sense of what programming GPUs is like.
+\q{Let me suggest some concepts that could be discussed a little bit more to give readers more of a sense of what programming GPUs is like.}
 
@@
 When discussing the batch-optimized linear solver code from NVIDIA, the authors should cite the NVIDIA site, and also links to CULA and MAGMA, other CUDA versions of LAPACK or subsets thereof.}
+\ai{These are good suggestions.
+We have added some of these suggestions to the paragraph where we introduce CUDA, but regrettably had to leave some of it out due to space considerations.
+Citation to the "Nvidia Registered Developers Program" is added where we speak of the batch solver.}
+%
 \qi{Problems need to have high "arithmetic intensity" (the authors use "computational intensity") so that the cores stay busy.
@@
 The authors do a nice job describing this situation on page 4 and the start of page 5.}
-\ai{Thank you for those kind words. In the mentioned context we agree that "arithmetic" is likely a better alternative to "computational", and will make the relevant edits.}
+\ai{We thank the reviewer for those kind words. In the mentioned context we agree that "arithmetic" is likely a better alternative to "computational", and will make the relevant edits.}
 %
@@
 you don't want the cores to all line up and wait their turn.}
-\ai{Yet again, very true! Your suggestion is much better, granted your permission we will use it.}
+\ai{Yet again, very true! The reviewer's suggestion is much better, we have used it.}
 %

### 2014-03-18 HowtoHowtoCapon/submission/submit1/answer/answer.tex
la2pdf commit - 2014:03:18	14:02:26	(Tue18.Mar)
...

@@
 \qi{A key attribute of the MVDR studied in this paper is the use of a subarray approximation. This should be stated up front in the abstract, i.e. "subarray MVDR" instead of MVDR.}
-\ai{You are right that subarray averaging is important in our study.
+\ai{The reviewer is right that subarray averaging is important in our study.
 But we believe that "subarray MVDR" would be misleading, as to the best of our knowledge this term is usually used to describe the use of subarrays to reduce complexity, while we average over subarrays to deal with the coherent echoes present in \emph{active} systems.
@@
 
-\ans{Thank you, it seems we should state this better upfront. In the 5th paragraph of the introduction we now explain that our system is a well sampled wideband sonar.}
+\ans{We thank the reviewer for these remarks, it seems we should state this better upfront. In the 5th paragraph of the introduction we now explain that our system is a well sampled wideband sonar.}
 
@@
 \end{align}
-\ai{Thank you, we realize that this was not clear enough in the article. The 32 elements are evenly spaced along the 1.2m array, so:
+\ai{The reviewer made us realize that this was not clear enough in the article. The 32 elements are evenly spaced along the 1.2m array, so:
 \begin{align}
@@
 \qi{What is impact of using only 32 elements? How are they spaced? It would seem that spatial aliasing would be an issue. Authors should explain what their assumptions are.}
-\ai{Thank you for pointing this out, we agree that this should be clarified.
+\ai{Appreciations for pointing this out.
+We agree that this should be clarified.
 The array configuration produce the first grating lobes at $\theta = \text{arcsin}(\lambda M/D) = 23.5^\circ$, but this is not a problem since the element opening angle is $25^\circ$ ($\theta\in[12.5^\circ,12.5^\circ]$) at $f_c=100$\;kHz.
@@
 Authors should explain what their assumptions are.}
-\ai{Thank you.
-We mention in the beginning of section IIA that we pre-steer to every angle and range bin in the image, but now repeat this in IIA paragraph 3, and state that we operate in wideband conditions in the introduction.}
+\ai{We mention in the beginning of section IIA that we pre-steer to every angle and range bin in the image, but now repeat this in IIA paragraph 3, and state that we operate in wideband conditions in the introduction.}
 %
@@
 }
-\ai{Again, thank you for a helpful remark, we will clarify this as well.
+\ai{Again, a helpful remark, we will clarify this as well.
 After pulse compressing a chirp using the full transceiver bandwidth (30~kHz) range resolution is roughly $\Delta r \approx \frac{1}{B} = 2.5$~cm, which corresponds to a sample shift of just 1.2\footnote{$\Delta n = \frac{\Delta r f_s}{c} = 1.2$}.
@@
 }
-\ai{Thank you, we understand how the confusion might arise.
+\ai{We understand how the confusion might arise.
 We operate under near field conditions and are indeed performing wavefront curvature beamforming.
@@
 For example, there are public-domain versions of optimized LAPACK libraries that provide the same sort of linear algebra codes as the authors used in their CPU implementation (for example, GoToBLAS, or Intel's MKL libraries).}
-\ai{Thank you.
-We have now included our reasons for not comparing with an optimized CPU implementation.
+\ai{We agree, and have now included our reasons for not comparing with an optimized CPU implementation.
 One reason is that we have not been able to find any CPU batch solvers out there and lacked the time to make our own.
@@
 
-\ans{You are absolutely right.
+\ans{The reviewer is absolutely right.
 The copy will be serialized due to divergent branching.
@@
 Here too, the authors should perhaps use the "coalesced" term, since this is the term used in all of the GPGPU documentation.}
-\ai{You are right. It is better to use the "official" terms. We have applied your suggestions.}
+\ai{The reviewer is right. It is better to use the "official" terms. We have applied the suggestions.}
 \end{enumerate}
@@
 \qi{Paragraph after equation (2) mentions "spatial averaging to avoid signal cancellation" - this problem might be better identified for readers by citing the Kailath reference and using "signal cancellation due to coherent multipath".}
-\ai{Thank you, both suggestions are now included.}
+\ai{Good point, both suggestions are now included.}
 %
@@
 Figure 10 shows estimated FLOPs for the two steps, but since both steps have different theoretical complexities (operations counts), and the two steps do not optimize the same way, this does not provide an expectation of the relative times.}
+\ai{TODO}
 \qi{On page 6, authors say their implementation was improved by factor of 1.5 - 2 when run on a K20.
 Authors could say a few words about what a K20 is, how it differs from the Quadro 6000, that the Kepler architecture actually has features that are dramatically better for scientific computing if the code is adapted to use them (kernels can call kernels, more registers, more shared memory, ability to eliminate some host-to-GPU i/o).}
+\ai{Good suggestions. They are now added to the last paragraph of section V.}
 \qi{Section I says 1-2 orders of magnitude. Section IV, page 6, says 2-3 orders of magnitude.}

### 2014-03-18 HowtoHowtoCapon/submission/submit1/answer/answer.tex
la2pdf commit - 2014:03:18	14:50:32	(Tue18.Mar)

@@
 We would like to thank the associate editor for his efforts and for his speedy replies to queries from the authors, and reviewers for their thorough and highly relevant comments.
-A summary of our answers are provided below, and subsequent sections will target each reviewer's comments in detail.
-We quote in \textbf{bold face} statements from the reviewers, and provide our replies in ordinary print,
+Subsequent sections will target each reviewer's comments in detail.
+We quote in \textit{italic} statements from the reviewers, and provide our replies in ordinary print,
 
-\section{Summary of response to reviewer's comments}
-
-\newpage
 \section{Answer to reviewer 1}

### 2014-03-18 HowtoHowtoCapon/submission/submit1/answer/answer.tex
la2pdf commit - 2014:03:18	14:58:18	(Tue18.Mar)

@@
 \ai{The reviewer has a point.
-But since there are 3 occurrences of the covariance processing time, including the total time GPU processing time would introduce three new lines.
-We feel this would clutter the plot unnecessarily as it is fairly easy derive the total time by looking at the current plots.
-Putting the total processing time in the MATLAB/C total time figure is an option, but then a logarithmic y-scale would necessary and detail would be lost.}
+But since there are 3 occurrences of the covariance processing time, including the total GPU processing time would introduce three new lines.
+We feel this would clutter the plot unnecessarily as it is fairly easy to derive the total time by looking at the current plots.
+Putting the total processing time in the MATLAB/C total time figure is an option, but then a logarithmic y-scale would necessary and detail would be lost.
+In stead we have included a remark in the caption saying the total GPU time is the sum of all the lines.}
 %

### 2014-03-18 HowtoHowtoCapon/submission/submit1/answer/answer.tex
la2pdf commit - 2014:03:18	15:55:51	(Tue18.Mar)
la2pdf commit - 2014:03:18	15:58:41	(Tue18.Mar)

@@
 \qi{Comment about real processing improvement over optimized C should be related in the text instead of a caption. This is an important point about the comparisons made in the paper.}
-\ai{It is indeed. We dedicated the second paragraph in the discussion (IV) to this topic, but will reword the relevant result paragraph slightly to hopefully cause less confusion.}
+\ai{It is indeed. We dedicated the second paragraph in the discussion (IV) to this topic, but have revised the paragraph to emphasize this point better.}
 %
@@
 I would suggest making a very clear itemization of the assumptions and caveats about the two approaches.}
-\ai{We agree that this should hould be clarified, and have We disagree that this }
+\ai{We agree that this should should be clarified, and }
 %

### 2014-03-18 HowtoHowtoCapon/submission/submit1/answer/answer.tex
la2pdf commit - 2014:03:18	18:42:05	(Tue18.Mar)

@@
 \qi{(caption of figure 10) "cruel" should be "crude"}
-\ai{Thank you, crude grammar correction applied.}
+\ai{Crude grammar correction applied.}
 %
@@
 I would suggest making a very clear itemization of the assumptions and caveats about the two approaches.}
-\ai{We agree that this should should be clarified, and }
+\ai{We agree that this should should be clarified, but since the transfer can happen in parallel without adding to the computation time we do not believe this point belongs in the introduction.
+An elaboration on the topic is included in the last paragraph in section IV.
+}
 %
@@
 
-\newpage
 \section{Answer to reviewer 2}
@@
 
-\ans{We thank the reviewer for these remarks, it seems we should state this better upfront. In the 5th paragraph of the introduction we now explain that our system is a well sampled wideband sonar.}
+\ans{We thank the reviewer for these remarks, it seems we should state this better upfront.
+In the 5th paragraph of the introduction we now explain that our system is a well sampled wideband sonar and that we operate in near field conditions.}
 
@@
 \end{align}
-\ai{The reviewer made us realize that this was not clear enough in the article. The 32 elements are evenly spaced along the 1.2m array, so:
+\ai{The reviewer made us realize that we should clarify this. The 32 elements are evenly spaced along the 1.2m array, so:
 \begin{align}
@@
 The array configuration produce the first grating lobes at $\theta = \text{arcsin}(\lambda M/D) = 23.5^\circ$, but this is not a problem since the element opening angle is $25^\circ$ ($\theta\in[12.5^\circ,12.5^\circ]$) at $f_c=100$\;kHz.
-In the paper we addition we perform sidescan imaging where only very narrow beams need to be beamformed.
+In the paper we perform sidescan imaging where only very narrow beams need to be beamformed.
 These details are now included in the paper, some in the 5th paragraph of the introduction and some in the first paragraph of section IV.}
@@
 \ai{Yes, we are pre-steering the array for every angle.
-We have clarified this further (see 3.), and added a remark saying that while not necessary here we could have phase steered the array by introducing the steering vector $\vec s$.}
+We have clarified this further (see 3.), and added a remark saying that while we can not use it here (due to processing a wideband signal) a phase steered version of MVDR can be obtained by introducing a steering vector $\vec s$.}
 %
@@
 It's difficult to say how much of this irrelevant.}
-\ai{There is certainly a potential for comparing with many other cards here, but we decided to be thorough with one architecture than to compare with everything.
+\ai{There is certainly a potential for comparing with many other cards here, but we decided to be thorough with one architecture rather than to compare with everything.
 To the paragraph introducing the Quadro 6000 we have explained what its equivalent consumer grade Geforce card would be.
@@
 The copy will be serialized due to divergent branching.
-It is unfortunate that no Cholesky based batch linear equation solver exist, which would have eliminated the need for the lower triangular...
+It is unfortunate that to the best of our knowledge no Cholesky based batch linear equation solver exist, which would have eliminated the need for the lower triangular...
 The paragraph has been edited to explain this better.}

### 2014-03-18 HowtoHowtoCapon/submission/submit1/answer/answer.tex
la2pdf commit - 2014:03:18	20:59:26	(Tue18.Mar)

@@
 The copy will be serialized due to divergent branching.
-It is unfortunate that to the best of our knowledge no Cholesky based batch linear equation solver exist, which would have eliminated the need for the lower triangular...
+It is unfortunate that to the best of our knowledge no Cholesky based batch linear equation solver exist, which would have eliminated the need for the lower triangular.
 The paragraph has been edited to explain this better.}
@@
 do the authors pre-steer all of the elements prior to forming $\R$ so that can "focus" in a wideband sense?}
-\ai{We indeed presteer all the elements.
-To make this clearer we have now stated it again after equation (2), and explained that a generalization is straight forward by substituting $\1$ with a steering vector.}
+\ai{We indeed presteer all the elements to deal with with our wideband signal.
+To make this clearer we have now stated it again after equation (2), and explained that in a narrowband case a generalization is straight forward by substituting $\1$ with a steering vector.}
 %

### 2014-03-18 HowtoHowtoCapon/submission/submit1/answer/answer.tex
la2pdf commit - 2014:03:19	00:32:47	(Wed19.Mar)
la2pdf commit - 2014:03:19	00:40:40	(Wed19.Mar)

@@
 Figure 10 shows estimated FLOPs for the two steps, but since both steps have different theoretical complexities (operations counts), and the two steps do not optimize the same way, this does not provide an expectation of the relative times.}
-\ai{TODO}
+\ai{The explanation is provided in the section V, the discussion.
+We have now rewritten the relevant paragraphs to include answers to these and other comments.
+Regarding the discussion of the FLOPS these were never intended to be an accurate representation of the run-times but rather to provide an indication of how efficiently we are able to harness the GPU's resources.}
+%
 \qi{On page 6, authors say their implementation was improved by factor of 1.5 - 2 when run on a K20.

### 2014-03-19 HowtoHowtoCapon/gfx/prepare.sh
la2pdf commit - 2014:03:19	02:18:35	(Wed19.Mar)

@@ -7,7 +7,7 @@ cp mvdr_build_R.eps buske4.eps
 cp buildR-breakdown.eps buske5.eps
 cp mvdr_implementation.eps buske6.eps
 cp plot_holmengraa_L16_Navg1.eps buske7.eps
-cp benchmark.eps buske8.eps
-cp code_assessment.eps buske9.eps
+cp code_assessment.eps buske8.eps
+cp benchmark.eps buske9.eps
 cp code_assess_flops.eps buske10.eps
 cp benchmark_boston.eps buske11.eps

### 2014-03-20 ThesisJP/Background/background.tex
BS-Capon complexity.
Writing background.

@@
 
-If \nom{$M$}{Number of array elements} is the number of elements in the array, and \nom{$\Delta_m$}{Per-element focusing and steering delays} is the per-element focusing and steering delay, the standard form of beamforming used on receive, known as \textit{delay-and-sum}, is given as:
+If \nom{$M$}{Number of array elements} is the number of elements in the array, $x_m$ is the per-element data, and \nom{$\Delta_m$}{Per-element focusing and steering delays} is the per-element focusing and steering delay, the standard form of beamforming used on receive in ultrasound imaging is:
 \begin{align}\label{eq:das}
-z[n] = \sum_{m = 0}^{M-1}w_m^*x_m[n - \Delta_m[n]] = \vec{w}^H\vec{x}[n],
+z[n] = \sum_{m = 0}^{M-1}w_m^*x_m[n - \Delta_m[n]] = \vec{w}^H\vec{x}[n].
 \end{align}
-where \nom{$\vec{w}$}{Array weight vector (apodization)} is know as the array \textit{weight vector} or \textit{apodization}, and $z[n]$ is the output A-mode image in the direction given by $\Delta_m[n]$.
+This is known as \textit{delay-and-sum} (DAS) beamforming.
+Here, \nom{$\vec{w}$}{Array weight vector (apodization)} is the array \textit{weight vector} or \textit{apodization}, $\vec{x}$ is a pre-delayed vector of element data, and $z[n]$ is the output or the A-mode image in the direction given by $\Delta_m[n]$.
+From now on $x_m[n]$ will refer to its delayed version $x_m[n-\Delta_m[n]]$.
 A schematic overview of delay-and-sum array beamforming is depicted in Fig.\,\ref{fig:das}.
@@
 \end{align} 
-frames per second (\nom{FPS}{Frames per second}). With our 2 cm cardiac probe from the earlier example we get a frame rate of around $55$ FPS when imaging a 70\degree{} sector with $r=15$ cm.
+frames per second (\nom{FPS}{Frames per second}). With our 2 cm cardiac probe we get a frame rate of around $55$ FPS when imaging a 70\degree{} sector with $r=15$ cm.
 
@@
 A short ultrasound pulse, which would yield high axial resolution, therefore requires a large frequency bandwidth.
-A ultrasound probe has normally a relative bandwidth between 50 and 100 \%.
-This corresponds to transmitting a pulse which is between 2 and 1 oscillations when entering tissue.
+An ultrasound system has normally a relative bandwidth between 50 \% and 100 \%.
+This corresponds to transmitting a pulse which is between one or two oscillations when entering tissue.
 In the front-end of an ultrasound imaging system a high sampling frequency is typically used (e.g.
@@
 Note that after this process the ultrasound data will be complex values.
-The total number of samples in an B-mode image is then
+The total number of samples in an B-mode image from a phased array is then
 \begin{align}
@@
 This is however less than the maximum bandwidth of PCI Express 2.0 (8 GB/s), even after the relative bandwidth, transmit frequency and number of array elements are increased with a total factor of 10.
-It is also exist newer versions of PCI Express which supports even higher transfer rates.
+Newer versions of PCI Express which supports even higher transfer rates have also been developed.
 
@@
 The data rate from a 2D scan is "only" in order of giga bytes per second prior to IQ-demodulation.
-To handle these amount of data, ultrasound imaging systems have relied heavily on application-specific integrated circuits (\nom{ASIC}{Application-specific integrated circuit}) for all steps in the processing pipeline.
-However, as the computationally power of CPUs and GPUs have increased, ASICs, e.g.
+To handle these amount of data ultrasound imaging systems have, for the last twenty years, relied heavily on application-specific integrated circuits (\nom{ASIC}{Application-specific integrated circuit}) for all steps in the processing pipeline \cite{Thomenius}.
+However, as the computationally power of CPUs have increased, ASICs, e.g.
 for converting data from a polar to a Cartesian grid (known as scan conversion), have gradually been converted into software \cite{Guracar2013}.
-One of  the last remaining ASICs in ultrasound scanners is the beamformer, whose job is to steer and focus the ultrasound beam in space.
+One of  the last remaining ASICs in ultrasound scanners is the beamformer \cite{Thomenius}, whose job is to steer and focus the ultrasound beam in space.
 This operation involves a reduction of the data rate with a factor equal to the number of elements in the ultrasound probe.
 The input data rate is therefor very high.
-Nevertheless, real-time beamforming has recently been achieved on a GPU \cite{Song2012}, and a French company (SuperSonic Imagine) are for instance utilizing GPUs in their plane wave beamforming algorithm \cite{Tanter2014}.
+Nevertheless, real-time beamforming has recently been achieved on a GPU \cite{Song2012}, and a French company (SuperSonic Imagine) are utilizing GPUs in their plane wave beamforming algorithm \cite{Tanter2014}.
 This is the same technique as reported implemented on a GPU by Yiu \textit{et al.} \cite{Yiu2011}.
@@
 Adaptive beamforming has several meanings in the literature on medical ultrasound imaging.
-It can for instance refer to aberration correction which is the task of applying time delays to compensate for the distortion introduced by fatty layers in the human body.
+It can for instance refer to aberration correction \cite{cole1996method} which is the task of applying time delays to compensate for pulse distortion introduced by fatty layers in the human body.
 However, in this thesis adaptive beamforming will refer to data-dependent apodization (or weighting) derived from a given optimization criterion.
@@
 Hence, the resulting weight vector minimizes the output power (the criterion) while maintaining unit gain in the steering direction $\vec{a}$ (distortionless response).
-The benefit of Capon beamforming is clearly seen from Fig\,\ref{fig:weights} which shows a sensitivity map (an \textit{array beampattern}) for DAS beamforming with uniform weights and the Capon beamformer.
-In this simulated case an interfering source is located at 20\degree{} which for the DAS beamformer with uniform weights (\ref{fig:das_weights}) will leak into the beamformer output with its power attenuated with around $13$ dB (first sidelobe).
+The benefit of Capon beamforming is clearly seen from Fig\,\ref{fig:weights} which present a sensitivity map (an \textit{array beampattern}) for DAS beamforming with uniform weights and the Capon beamformer.
+In this simulated case an interfering source is located at 20\degree{} which for the DAS beamformer with uniform weights (\ref{fig:das_weights}) will leak into the beamformer output with its power attenuated with around $13$ dB.
 Since the Capon weights are data-dependent the method "knows" where the interfering source is located and is able to position a zero in the direction of it (Fig.\,\ref{fig:capon_weights}).
@@
 In theory a signal will be regarded by the Capon beamformer as an interfering source as long as the signal propagation vector does not match the steering vector ($\vec{a}$) exactly.
-If it is not possible to position a zero in the direction of the interferer the mainlobe will get shifted and scaled if this is whats needed to achieve minimum output power.
+If it is not possible to position a zero in the direction of the interferer the mainlobe will get shifted and scaled if this is whats needed in order to achieve minimum output power.
 Adaptive positioning of zeros and mainlobe shifting gives rice to the super-resolution capabilities of the Capon beamformer \cite{Synnevag2007}, and the positive implications of this effect \cite{Synnevag2009} is the main reason why this beamformer has been highly studied in the field of medical ultrasound imaging the last decade.
@@
 \begin{align}
-\mat{\breve{R}}[n] = \frac{1}{N_LN_K}\sum_{n'=n-K}^{n+K} \sum_{l=0}^{N_L-1} \vec{x}_l[n']\vec{x}_l[n']^H,
+\mat{\breve{R}}[n] &= \frac{1}{N_LN_K}\sum_{n'=n-K}^{n+K} \sum_{l=0}^{N_L-1} \vec{x}_l[n']\vec{x}_l[n']^H\\
+\vec{x}_l[n'] &= [x_l[n], \dotso, x_{l+L}[n]]
 \end{align}
-where\nomi{$K$}{Temporal averaging parameter (Capon)} $N_K = 2K + 1$ \nomi{$N_K$}{Number of samples used for temporal averaging (Capon)} is the number of samples used for temporal smoothing,\nomi{$L$}{Subarray length (Capon)} $\vec{x}_l$ is a $L$ long subarray $[x_l[n], \dotso, x_{l+L}[n]]$, and $N_L = M-L+1$\nomi{$N_L$}{Number of subarrays used for spatial averaging (Capon)} is the number of subarrays used for spatial averaging.
+where\nomi{$K$}{Temporal averaging parameter (Capon)} $N_K = 2K + 1$ \nomi{$N_K$}{Number of samples used for temporal averaging (Capon)} is the number of samples used for temporal smoothing,\nomi{$L$}{Subarray length (Capon)} $\vec{x}_l$ is a $L$ long subarray, and $N_L = M-L+1$\nomi{$N_L$}{Number of subarrays used for spatial averaging (Capon)} is the number of subarrays used for spatial averaging.
 Temporal averaging is used to obtain DAS-like speckle statistics in combination with large subarrays \cite{Synnevag2007a}, and subarray averaging is applied to reduce signal cancellation caused by coherent signals \cite{Reddy1987}.
@@
 \begin{align}
-&\mat{B}[x_l[n],\,\dotso,\,x_{l+L}[n]]^T\\
-&[\mat{B}]_{p,q} = \frac{1}{\sqrt{L}}e^{-j 2 \pi p q / L}
+\vec{x}_{\text{BS},l} &= \mat{B}[x_l[n],\,\dotso,\,x_{l+L}[n]]^T\\
+[\mat{B}]_{p,q} &= \frac{1}{\sqrt{L}}e^{-j 2 \pi p q / L}.
 \end{align}
 The idea behind BS-Capon beamforming is that spatial beams with little information can be removed from $\mat{\hat{R}}$ by removing rows in $B$.
-The complexity of BS-Capon beamforming is dominated by the transformation of data into beamspace which has a computational complexity of $O(N_LLN_b)$, where \nom{$N_b$}{Number of selected beams in beamspace} is the number of selected beams.
+When investigating complexity of BS-Capon beamformer \textbf{Paper II} we find that is dominated by the transformation of data into beamspace, $O(N_LLN_b)$, where \nom{$N_b$}{Number of selected beams in beamspace} is the number of selected beams.
 
@@
 \begin{align}
-\text{FLOPS}_{\text{BS-Capon}} = L^2N_bB_wf_c
+\text{FLOPS}_{\text{BS-Capon}} = N_LLN_bB_wf_c
 \end{align}
-For our example, BS-Capon beamforming with $L=M/2=64$ and $N_b=3$, will require a throughput of around $30$ GFLOPS.
+For our example, BS-Capon beamforming with $L=M/2=64$ and $N_b=3$ will require a throughput of around $30$ GFLOPS.
 This is a level of processing which both modern CPUs and GPUs should be able to handle.
@@
 The Capon beamformer is derived for narrowband applications and has traditionally been applied on medical ultrasound imaging "as is".
-Medical ultrasound imaging is, however, a broadband imaging modality and one could think that a broadband Capon beamformer therefore where needed \cite{Holfort2009}.
-The investigation performed in \cite{Nilsen2009} shows that with focused beams, energy will be centered on-axis and as little as three near-axis beams is therefore adequate in order to achieve similar performance for BS-Capon and Capon beamforming.
-The two-way beamprofile will be wider with un-focused (plane wave) transmit beams \cite{Holfort2009, Austeng2011}, but focusing on receive will still make the main amount of energy arrive on axis.
-As it is shown in \textbf{Paper III} it is possible to micro-steer with phase delays instead of time delays without introducing large errors as long as the phase shift is smaller than the pulse length.
-Since Capon beamforming mainly performs main-lobe shifts smaller than the beam width, broadband Capon beamforming will only provide a slight improvement over narrowband Capon beamforming.
+Medical ultrasound imaging is, however, a broadband imaging modality and one could think that a broadband Capon beamformer therefore where strictly needed \cite{Holfort2009}.
+The investigation performed in \cite{Nilsen2009} showed that with pre-delayed beams, energy will be centered on-axis and as little as three near-axis beams is therefore adequate in order to achieve similar performance for BS-Capon and Capon beamforming.
+The two-way beam will be wider for unfocused (plane wave) transmit beams \cite{Holfort2009, Austeng2011}, but steering and focusing on receive will still make the main amount of energy arrive on axis.
+As shown in \textbf{Paper III} it is also possible to micro-steer with phase delays (narrowband) instead of time delays (broadband) without introducing large errors.
+Since Capon beamforming mainly performs main-lobe shifts smaller than the beam width, the difference between a broadband and narrowband Capon beamformer will be minimal \cite{Diamantis2014}.
+Broadband Capon beamforming is also significantly more computationally demanding than narrowband Capon beamforming.
 

### 2014-03-22 ThesisJP/Background/background.tex
Finished section on volume rendering. Next, adaptive VR.

@@
 \section{Volume rendering}\label{sec:volren}
-
-Get section from master theses. Ray casting and opacity functions. (Fig.\,\ref{fig:vr})
-
 \begin{figure}
@@
 \subfigure[Ray-casting]{
-	\includegraphics[width=0.4\textwidth]{img/Volumeraycasting.png}
+	\includegraphics[width=0.4\textwidth]{img/Volumeraycasting.png}\label{fig:vr_rc}
 }
 \subfigure[Opacity transfer function]{
-	\includegraphics[width=0.5\textwidth]{img/otf.png}
+	\includegraphics[width=0.5\textwidth]{img/otf.png}\label{fig:vr_otf}
 }
-\caption{.}
+\caption{The principle behind volume ray casting.
+a) A ray is casted out from every pixel in the output image.
+Along the ray, samples are gathered from the volume.
+The samples are assigned a color and an opacity before they are combined into one pixel value.
+b) The opacity transfer function used to determine how opaque a sample should be.}
 \label{fig:vr}
 \end{figure}
+A 3D ultrasound dataset is a cloud of scalar values that somehow needs to be presented to the user.
+One approach used for cardiac ultrasound  is to extract slices corresponding to different long-axis and short-axis views of the cardiac chambers.
+Another way is to project the whole volume into a two dimensional image.
+The latter technique is known as volume rendering, and it present for the user the three dimensional structure of the heart in one two-dimensional image.
+Many different methods for volume rendering exists \cite{brodlie2001}, and in this thesis we will focus on ray casting \cite{Levoy1988}.
+
+The principle behind ray casting is depicted in Fig.\,\ref{fig:vr}.
+The projection of volumetric data is with this technique performed by casting a ray out from every pixel in the output image (Fig.\,\ref{fig:vr_rc}).
+Along each ray, samples are gathered from the volume and assigned a color and opacity.
+For ultrasound visualization the color intensity is dictated by the ultrasound backscatter intensity, and the actual color is often encoded with depth information in order to improve depth perception.
+An example of a volume rendering of cardiac ultrasound data is presented in \textbf{Paper IV}.
+The opacity, however, is governed by a user controlled \textit{opacity transfer function} (\nom{OTF}{Opacity transfer function (volume rendering)}) as shown in Fig.\,\ref{fig:vr_otf}.
+For ultrasound visualization a monotonically increasing function is typically used, which means that below a certain threshold all samples will be rendered transparent (opacity = 0).
+We will refer to this as the \textit{opacity threshold}, and for the example in Fig.\,\ref{fig:vr_otf} this threshold is 25.
+
+The iterative composition scheme for front-to-back ray casting of one ray, from $s_0$ to $s_{N_s-1}$ where $s$ is the sample position and $N_s$ is the number of samples along the ray, is
+\begin{align}
+I(s_0, s_{i+1}) &= I(s_0, s_{i}) + (1 - \alpha_{\text{acc},i}) c_{i+1}\\
+\alpha_{\text{acc},i+1} &= \alpha_{\text{acc},i} + (1 - \alpha_{\text{acc},i}) \alpha(c_i+1).
+\label{eq:fronToBack}
+\end{align}
+Here, $\alpha(v)$ is the opacity transfer function, $c_{k}$ is the color intensity of sample $s_{k}$, $\alpha_{\text{acc}}$ is the accumulated opacity, and $I(s_0, s_{N_s-1})$ is the intensity value of all the projected samples combined.
+If $\alpha(v)$ is close to one for one or two consecutive samples, $\alpha_{\text{acc}}$ will quickly approach a value of one.
+An $\alpha_{acc,k}$-value close to one means that the contribution from $c_{k+1}, \cdots, c_{N_s-1}$ to the final value, $I(s_0, s_{N_s-1})$, will be minimal.
+Hence, the iteration can stop at $i=k$ without loosing visible samples.
+This optimization technique is known as \textit{early-ray-termination}, and will significantly speed up our rendering algorithm if an opaque material is rendered.
+
+For visualization of cardiac ultrasound we want $\alpha_{\text{acc}}$ to be close to one only when a tissue interface has been encountered.
+If a sample of clutter noise is assigned a high opacity value, cardiac tissue further behind will not be visible.
+It is said to be occluded by clutter noise.
+The general goal in cardiac ultrasound visualization is therefore to assign high opacity to cardiac tissue and low opacity to noise samples.
+This is a big challenge for cardiac ultrasound since the backscattered signal level from tissue varies a lot throughout the volume.
+The per-volume sample value distributions for tissue and noise therefore always overlap to some degree.
+
+Volume rendering of a single volume has a complexity of $O(N_xN_yN_s)$, where $N_xN_y$ is the number of pixels in the output image.
+With $N_x = N_y = N_s = 512$ and a required frame rate between ten and twenty, we find that the required number of FLOPS for real-time volume rendering of 3D cardiac ultrasound is in the order of several GFLOPS.
+When calculation of gradients for shading and more advanced processing is added this number continues to grow with a constant factor.
 

### 2014-03-22 ThesisJP/Background/background.tex
Finished section on adaptive volume rendering.

@@
 
-The iterative composition scheme for front-to-back ray casting of one ray, from $s_0$ to $s_{N_s-1}$ where $s$ is the sample position and $N_s$ is the number of samples along the ray, is
+The iterative composition scheme for front-to-back ray casting of one ray, from $s_0$ to $s_{N_s-1}$ where $s$ is the sample position and $N_s$ is the number of samples along the ray, is \cite{rtvg2006}
 \begin{align}
@@
 Volume rendering of a single volume has a complexity of $O(N_xN_yN_s)$, where $N_xN_y$ is the number of pixels in the output image.
-With $N_x = N_y = N_s = 512$ and a required frame rate between ten and twenty, we find that the required number of FLOPS for real-time volume rendering of 3D cardiac ultrasound is in the order of several GFLOPS.
-When calculation of gradients for shading and more advanced processing is added this number continues to grow with a constant factor.
+With $N_x = N_y = N_s = 512$ and a required frame rate for 3D cardiac ultrasound between ten and twenty, we find that the required number of FLOPS for real-time volume rendering of 3D cardiac ultrasound is in the order of several GFLOPS.
+When calculation of gradients for shading and more advanced processing is added per sample this number continues to grow with a constant factor.
 
 \subsection{Adaptive volume rendering}
+As for adaptive beamforming, adaptive volume rendering means to derive and use information about the volume being visualized from the volume it self.
+In the literature on volume rendering this is known as feature enhancement or visibility driven visualization \cite{viola2005, correa2010visibility, marchesin2010}.
+The common feature of these techniques is how the sample opacity is modulated in order to achieve visibility of relevant structures.
+If we somehow are able to classify samples as cardiac tissue we can use this information as input to an adaptive volume rendering algorithm, and by that achieve improved visibility of cardiac tissue.
 
-Visibility driven visualization.
+In \textbf{Paper IV} we extend the work by Honigmann \textit{et al.} \cite{Honigmann2003} on a global adaptive OTF by formulating an regional adaptive OTF for visualization of cardiac ultrasound volumes.
+This regional adaptive OTF is based on estimated statistics of the clutter noise inside the left ventricle (the blood pool).
+The added complexity of this technique is the statistical estimation which is basically a second volume rendering (mean and variance projection) that has to be performed before the rendered image is produced.
 

### 2014-03-22 ThesisJP/Background/background.tex
Started on section about ultrasound field simulations.

@@
 
+\subsection{Beamforming}
 If the width of a piezoelectric element is less than the wavelength of the transmitted signal the element is said to be \textit{omnidirectional}, hence energy is distributed or sensed equally in all directions.
@@
 
+\subsection{Sampling and processing complexity}
 When generating a B-mode image we have to insure that the spatial and temporal sampling are sufficient.
@@
 
-\section{Field simulations}\label{sec:field}
+\section{Ultrasound field simulations}\label{sec:field}
+Simulation of ultrasound probes are heavily used by researchers in order to test out new algorithms and probe designs.
+A  typical simulator for ultrasound probes works by approximating the probe surface with discrete points or patches.
+In software it is then possible to transmit a pulse from a given probe and measure its response in space and time.
+The most popular simulator for ultrasound field simulations is Field II \cite{Jensen1992}.
+All simulations in both \textbf{Paper II} and \textbf{III} are produced with this simulator.
 
-Small chapter about different simulation tools (See hos paper).
+Field II. Spatial impulse response.
+
+Ultrasim. Fusk. Propose. Explain the principle behind all.
+
+What is the computational complexity of Filed II and Ultrasim.
+
+Simulators implemented on the GPU.
 

### 2014-03-23 ThesisJP/Background/background.tex
Started writing about field simulations. Added hos-field figure.

@@
 \section{Volume rendering}\label{sec:volren}
-\begin{figure}
+\begin{figure}[t!]
 \centering
@@
 \subsection{Adaptive volume rendering}
-As for adaptive beamforming, adaptive volume rendering means to derive and use information about the volume being visualized from the volume it self.
+As for adaptive beamforming, adaptive volume rendering means to derive and use information about the data being visualized from the data itself.
 In the literature on volume rendering this is known as feature enhancement or visibility driven visualization \cite{viola2005, correa2010visibility, marchesin2010}.
@@
 Simulation of ultrasound probes are heavily used by researchers in order to test out new algorithms and probe designs.
-A  typical simulator for ultrasound probes works by approximating the probe surface with discrete points or patches.
-In software it is then possible to transmit a pulse from a given probe and measure its response in space and time.
-The most popular simulator for ultrasound field simulations is Field II \cite{Jensen1992}.
+A simulator for ultrasound probes works by approximating the probe surface with discrete points or patches.
+In software it is then possible to transmit a pulse from a given probe and measure its response in space and time, the \textit{spatial impulse response}.
+Figure \ref{fig:hos} shows a snapshot of a simulated pressure field from an unfocused linear array emitting a continuous wave.
+The image was simulated with the simulator from \textbf{Paper V}.
+The most popular simulator for medical ultrasound field simulations is Field II \cite{Jensen1992}.
 All simulations in both \textbf{Paper II} and \textbf{III} are produced with this simulator.
+The Field II simulator works by dividing each element into mathematical elements, and all field points needs to be in the far-field of these small elements.
+Then it is possible to approximate the spatial impulse response by summing a trapezoid impulse response function from each mathematical element.
+\todo{Bergen flops for Field II.
+From this we see that calculation of ultrasound simulation is complex.}
+
+\begin{figure}[t!]
+\centering
+\includegraphics[width=\textwidth]{img/hos.eps}
+\caption{An example of a simulated ultrasound pressure field from a linear array without focusing. The field was simulated using the simulator in \textbf{Paper V}.}
+\label{fig:hos}
+\end{figure}
 

### 2014-03-23 ThesisJP/Background/background.tex
Siste murstein på plass.

@@
 The per-volume sample value distributions for tissue and noise therefore always overlap to some degree.
+With a global OTF it is therefore impossible to find a threshold that separates tissue from noise.
 
@@
 This regional adaptive OTF is based on estimated statistics of the clutter noise inside the left ventricle (the blood pool).
-The added complexity of this technique is the statistical estimation which is basically a second volume rendering (mean and variance projection) that has to be performed before the rendered image is produced.
+The added complexity of this technique caused by the statistical estimation which is basically a second volume rendering (mean and variance projection) that has to be performed before the rendered image is produced.
 
@@
 Simulation of ultrasound probes are heavily used by researchers in order to test out new algorithms and probe designs.
-A simulator for ultrasound probes works by approximating the probe surface with discrete points or patches.
-In software it is then possible to transmit a pulse from a given probe and measure its response in space and time, the \textit{spatial impulse response}.
+A simulator for ultrasound probes typically works by approximating the probe surface with discrete points or patches.
+In software it is then possible to transmit a pulse from a given probe and measure its response in space and time, know as the \textit{spatial impulse response}.
 Figure \ref{fig:hos} shows a snapshot of a simulated pressure field from an unfocused linear array emitting a continuous wave.
 The image was simulated with the simulator from \textbf{Paper V}.
-The most popular simulator for medical ultrasound field simulations is Field II \cite{Jensen1992}.
-All simulations in both \textbf{Paper II} and \textbf{III} are produced with this simulator.
-The Field II simulator works by dividing each element into mathematical elements, and all field points needs to be in the far-field of these small elements.
-Then it is possible to approximate the spatial impulse response by summing a trapezoid impulse response function from each mathematical element.
-\todo{Bergen flops for Field II.
-From this we see that calculation of ultrasound simulation is complex.}
 
@@
 
-Field II. Spatial impulse response.
+The most popular simulator for medical ultrasound field simulations is Field II \cite{Jensen1992}.
+All simulations in both \textbf{Paper II} and \textbf{III} are produced with this simulator.
+The Field II simulator works by dividing each element into mathematical elements, and all field points needs to be in the far-field of these small elements\footnote{$l\gg\frac{w^2}{4\lambda}$, where $w$ is the mathematical element width, $l$ is the distance to the field point, and $\lambda$ is the wavelength of the transmitted pulse.
+For $w=\lambda/2$ the approximation is valid for $l>\lambda/16$.}.
+It is then possible to approximate the spatial impulse response by summing a trapezoid impulse response function originating from each mathematical element \cite{Jensen1992} in all field points.
+A rough estimate of the number of required FLOPS for calculating the spatial impulse respons as it is done in Field II is:
+\begin{align}
+\text{FLOPS}_\text{Field II} = f_s\frac{w}{c}N_{\text{elements}}N_{\text{field points}},
+\end{align}
+where $f_s$ is the sampling frequency, $w$ is the width of the mathematical elements, $f_s\frac{w}{c}$ is the maximal number of samples in each trapezoid function, $N_{\text{elements}}$ is the number of mathematical elements, and $N_{\text{field points}}$ is the number of field points.
+The required number of FLOPS for a simulation at $100$ MHz with $500$, $\lambda/2$ wide, mathematical elements across a field of $0.5$ million points is $6$ GFLOPS in order for it to complete after one second.
+If we want to simulate an ultrasound image the spatial impulse response on both transmit and recieve have to be calculated and convolved in time.
+The respons will also be different for each A-mode image.
+Hence, more than 800 GFLOPS is required in order to complete a Field II simulation with the given example values of an image consisting of 70 A-mode images in one second.
+However, Jensen and Nikolov have shown ways of reducing the computational load of Field II simulations \cite{Jensen2000}.
+Other simulators have also been introduced that are faster but less exact than Field II \cite{Hergum2009}.
 
-Ultrasim. Fusk. Propose. Explain the principle behind all.
+For the simulator presented in \textbf{Paper V} the required number of FLOPS in order to achieve a certain frame rate $f_{\text{FPS}}$ is:
+\begin{align}
+\text{FLOPS}_{\text{HOS}} = N_{\text{source points}}N_{\text{field points}}f_{\text{FPS}}.
+\end{align}
+Therefore, to simulate a field of one million points with several thousand source points at 50 FPS, more than 100 GFLOPS is required.
+Note that this simulator evaluates a cosine and sine function for each calculation, something which makes it even harder to achieve the required throughput, since especially GPUs have less silicon devoted to such tasks.
 
-What is the computational complexity of Filed II and Ultrasim.
-
-Simulators implemented on the GPU.
+The above examples shows that realistic and dense ultrasound simulations are computationally demanding.
+On the other hand we find that the levels of required processing is within the theoretical limits of modern CPUs and GPUs.
+Since ultrasound field simulations are quite easy to accelerate in parallel it should be possible, even if a theorethical throughput is almoste impossible to achieve in practise, to add increased interactivity to ultrasound simulations tools by utilizing parallel programing and SIMD instructions.
+In addition to \textbf{Paper V} threre are several simulators which have been presented the last years that does acactly this by laveraging the powers of GPUs \cite{Gjerald2012, Reichl2009, Hlawitschka2011}.
 

### 2014-03-29 ThesisJP/Background/background.tex
Started on comments from Ines.

@@
 
-In this chapter the background information needed in order to understand the topics of this thesis and its papers is given.
-First, in Section \ref{sec:gpgpu}, the history and principle of general-purpose computing on graphics processing units (\nom{GPGPU}{General-purpose computing on graphics processing units}) is reviewed, followed by an introduction to medical ultrasound imaging and adaptive beamforming in Section \ref{sec:ultrasound}.
+This chapter gives the background information needed in order to understand the topics of this thesis and its papers.
+First, in Section \ref{sec:gpgpu}, the history and the principle of general-purpose computing on graphics processing units (\nom{GPGPU}{General-purpose computing on graphics processing units}) are reviewed, followed by an introduction to medical ultrasound imaging and adaptive beamforming in Section \ref{sec:ultrasound}.
 Finally, the concepts of volume rendering and ultrasound field simulations are presented in Section \ref{sec:volren} and \ref{sec:field}.
@@
 The essay bears the name "Welcome to the Jungle" and is a sequel to the essay "The Free Lunch is Over" from 2004 \cite{HerbSuttera}.
-The two essays spins around the fact that manufactures of central processing units (CPUs) hit a frequency-wall in the beginning of the 21st century.
+The two essays spin around the fact that manufactures of central processing units (CPUs) hit a frequency-wall in the beginning of the 21st century.
 Until then, all mainstream computer programs where typically running in a single thread on a single CPU \textit{processing core}, and a programmer could expect the software to annually gain performance without touching the code.

### 2014-03-30 ThesisJP/Background/background.tex
Revised GPU text.

@@
 The essay bears the name "Welcome to the Jungle" and is a sequel to the essay "The Free Lunch is Over" from 2004 \cite{HerbSuttera}.
-The two essays spin around the fact that manufactures of central processing units (CPUs) hit a frequency-wall in the beginning of the 21st century.
+The two essays spin around the fact that manufacturers of central processing units (CPUs) hit a frequency wall in the beginning of the 21st century.
 Until then, all mainstream computer programs where typically running in a single thread on a single CPU \textit{processing core}, and a programmer could expect the software to annually gain performance without touching the code.
@@
 The increase in processing power came mostly from an ever-growing clock frequency.
-However, power consumption and heat generation where also growing at the same rate, until the level of cooling required finally became to much in 2004.
+However, power consumption and heat generation where also growing at the same rate until the required cooling finally became to much in 2004.
 From that point, chip manufactures had to pursue a different direction  than just increasing the clock frequency.
@@
 
-Around the same time as CPU manufactures hit the frequency wall, people started experimenting with programmable shaders which recently had been introduced in the field of graphics programming.
+Around the same time as CPU manufacturers hit the frequency wall, people had started experimenting with programmable shaders which recently had been introduced in the field of graphics programming.
 The graphics processing pipeline had before this been a fixed function pipeline, where e.g.
-geometry and textures where fed to the graphics processing unit (GPU) in one end, and where shaded a rasterized geometry was outputted in the other.
+geometry and colors where fed to the graphics processing unit (GPU) in one end, and where shaded and rasterized sceneries were outputted in the other.
 What programmable shaders  introduced was the ability to replace some steps in the fixed-function pipeline with custom computations.
@@
 The rationale behind this exploration was that GPUs were already highly available in consumer PCs, and hidden behind its graphics interface were processing powers an order of magnitude larger than what the CPU could provide (Fig.\,\ref{fig:cpu_vs_gpu}).
-Actually, at the time when CPU manufacturers hit the frequency wall the GPU had already gone multi-core, first of all driven by an ever-increasing demand for more realistic computer games.
-Secondly, the rendering pipeline exhibits minimal execution path branching, and data are typically used once.
-Designers of GPUs could therefore skip advance feature, found in most CPUs, like branch prediction and different levels of data caching.
-These measures reduced the silicon footprint of each core and made it possible to add multiple cores to one GPU before CPU designers where able to do the same.
-Today, even if more advanced caching has been added to GPUs, this is still the main reason why GPUs have higher peak performance than CPUs (Fig.\,\ref{fig:cpu_vs_gpu}).
-Another reason is the inherent parallel nature of the rendering problem.
-When geometry is shaded, projected and rasterized the same instruction is typically needed for a lot of data at the same time.
-This makes a special kind of instruction, known as Single Instruction Multiple Data (\nom{SIMD}{Single instruction multiple data (instruction type)}) \cite{Flynn1966}, especially suited for this kind of problem.
-The differences between how SIMD instructions are utilized in modern CPUs and GPUs are discussed in Section \ref{sec:cpu_vs_gpu}.
+Actually, at the time when CPU manufacturers hit the frequency wall, the GPU had already gone multi-core, first of all driven by an ever-increasing demand for more realistic computer games.
+
+The graphics pipeline exhibits minimal execution path branching, and data are typically used once.
+Designers of GPUs could therefore skip advanced feature found in most CPUs, like branch prediction and different levels of data caching.
+These measures reduced the silicon footprint of each core and made it possible to add multiple cores to the GPU before CPU designers where able to do the same.
+Today, even if more advanced caching has been added to GPUs, this is still the main reason why GPUs have higher peak performance than CPUs (Fig.\,\ref{fig:cpu_gpu} and \ref{fig:cpu_vs_gpu}).
+Another reason is the inherent parallel nature of the problem of rendering graphics.
+When geometry is shaded, projected, and rasterized, the same instruction is needed for a lot of data at the same time.
+This makes a special kind of instruction, known as Single Instruction Multiple Data (\nom{SIMD}{Single instruction multiple data (instruction type)}) \cite{Flynn1966}, especially suited for this problem.
+How SIMD instructions are utilized in modern CPUs and GPUs is discussed in Section \ref{sec:cpu_vs_gpu}.
 
 For the first adopters of general-purpose computing on GPUs (GPGPU) it soon became evident that offloading all types of computations to the GPU was not always a good idea.
-In most cases it is important to balance the load equally between the CPU and GPU to obtain their combined computationally power.
-Balancing typically means that highly parallel and computationally intense tasks should be offloaded to the GPU and serialized and memory intensive work should stay on the CPU.
+In most cases it is important to balance the load equally between the CPU and GPU to unify the computational power of both processors.
+Balancing typically means that highly parallel and computationally intense tasks should be offloaded to the GPU while serialized and memory intensive work should stay on the CPU.
 This paradigm of utilizing specialized cores for solving specific problems is today known as \textit{heterogeneous computing}.
@@
 A recent trend is that specialized accelerator cores are embedded onto the CPU.
-This heterogeneous processors is often referred to as an accelerated processing unit (\nom{APU}{Accelerated processing unit}).
-The last three generations of CPUs from Intel have all had on-chip integrated graphics, hence they can be classified as APUs.
-An on-chip GPU is less powerful than a high-performance discrete GPU, but valuable time is saved by not having to send data across the PCI Express bus, which connects the different parts of a computer to the motherboard.
+This heterogeneous processor is often referred to as an accelerated processing unit (\nom{APU}{Accelerated processing unit}).
+The last three generations of CPUs from Intel have all had on-chip integrated graphics, and all of them are therefore APUs.
+An on-chip GPU is less powerful than a high-performance discrete GPU, but valuable time is saved by not having to send data across the PCI Express bus, which connects the different parts of a computer to the CPU.
 This is the reason why one of the first rules of GPGPU (with discrete GPUs) is to minimize CPU-to-GPU memory transfers.
+An integrated GPU can therefore provide increased performance over a discrete GPU if the problem requires a high level of CPU-GPU data synchronization.
 
 The final transition currently taking place in the computer industry is the introduction of cloud computing, where compute power is provided as a web service and automatically scales on demand.
-Together, multi-core, heterogeneous cores, and cloud cores forms the "jungle" that today's software engineers have to tackle (Fig.\,\ref{fig:jungle}).
-In this thesis we focus on heterogeneous computing with one multi-core CPU and a discrete high performance GPU, a cheap and high performance system which is likely to be the specification of current and future ultrasound imaging system.
+Together, multi-cores, heterogeneous cores, and cloud cores form the "jungle" that today's software engineers have to handle (Fig.\,\ref{fig:jungle}).
+In this thesis we focus on heterogeneous computing with one multi-core CPU and a discrete high performance GPU.
+This is a cheap and high performance system which is likely to be the specification of current and future ultrasound imaging system.
 
 \subsection{Comparing CPU and GPU performance}\label{sec:cpu_vs_gpu}
-The CPU is design to be the main control unit of a computer, it is therefore good at general processing.
+The CPU is designed to be the main control unit of a personal computer, and is therefore good at general processing.
 The GPU, on the other hand, has a history of processing graphics which involves heavy vector arithmetic.
-Even though they started out as very different processors, modern CPUs are including more and more GPU-like features, and GPUs are adding CPU features.
-In this section we will see what these similarities are and why there is still a big difference in peak performance between a CPU and a GPU.
+Although they started out as very different processors, modern CPUs are now including more and more GPU-like features, and GPUs are adding CPU features.
+In this section we will see what these features are and why there is still a big difference in peak performance between a CPU and a GPU.
 
@@
 
-In Section \ref{sec:gpgpu} we introduced a special type of instructions, namely SIMD instructions, which are designed to execute a single instruction across multiple data element in one clock cycle.
-At the end of the 20th century, Intel introduced a SIMD instruction set for their Pentium III processor called the Streaming SIMD Extensions (\nom{SSE}{Streaming SIMD extensions (instruction set)}), which has seen several updates since.
-The first SEE instructions set had 128 bit long registers which meant that four, 32 bit, floating point values could be processed per clock cycle\footnote{For simplicity we will assume that the whole register is processed in one clock cycle even if the actual implementation might distributes one instruction for the register across multiple cycles.}.
+In Section \ref{sec:gpgpu} a special type of instructions was mentioned, namely SIMD instructions.
+These are instructions which are designed to execute a single instruction across multiple data element in one clock cycle.
+At the end of the 20th century, Intel introduced a SIMD instruction set for their Pentium III processor called the Streaming SIMD Extensions (\nom{SSE}{Streaming SIMD extensions (instruction set)}).
+The first SSE instruction set had registers of 128 bit which meant that four single-precision (32 bit) floating point values could be processed per clock cycle.
 Updates of the SSE instruction set have added new instructions and longer registers (SSE2, SSE3, SSE4, \nom{AVX}{Advanced vector extensions (instruction set)}, AVX2, and AVX-512).
-The newly introduced extensions, AVX and AVX2, adds 256 bit long registers and fused multiply-accumulate instructions (\nom{FMA}{Fused multiply-accumulate (instruction)})  which means that sixteen 32 bit floating point operations can be performed per clock cycle per core.
+The newly introduced extensions, AVX and AVX2, add registers of 256 bit and fused multiply accumulate (\nom{FMA}{Fused multiply-accumulate (instruction)}) instructions  which means that sixteen single-precision floating point operations can be performed per clock cycle per core.
 A new extension, AVX-512, extends the registers to 512 bit and is found in the Intel accelerator board  Xeon Phi and might show up in the next generation of CPUs from Intel.
-The latest CPU architecture from Intel, the Haswell architecture, has two processing ports per core supporting FMA-AVX2 instructions, and is therefore capable of performing 32, 32 bit, floating point operations per clock cycle per core (Fig.\,\ref{fig:cpu_gpu}).
-For an eight core\footnote{An 8-core Haswell CPU is scheduled to launch during 2014.} CPU with a clock frequency of 3.8 GHz this sums to
+The latest CPU architecture from Intel, the Haswell architecture, has two processing ports per core supporting FMA-AVX2 instructions, and is therefore capable of performing 32 single-precision floating point operations per clock cycle per core (Fig.\,\ref{fig:cpu_gpu}).
+For an eight-core\footnote{An 8-core Haswell CPU is scheduled to launch during 2014.} CPU with a clock frequency of 3.8 GHz this sums to
 \begin{equation}
-32*8*3.8 = 972.8\,\text{GFLOPS},
+32*8*3.8 = 972.8\,\text{GFLOPS}.
 \end{equation}
-hence a theoretical peak performance of close to one tera single precision floating point operations per second (TFLOPS).
+Hence, a theoretical peak performance of close to one tera single precision floating point operations per second (TFLOPS).
 This is the same number as visualized in Fig\,\ref{fig:cpu_vs_gpu} for the Haswell architecture (Single precision).
-In addition, the most powerful integrated GPU found on Haswell CPUs, the HD 5200, has 40 execution units capable of processing sixteen, 32 bit, floating points at 1.3 GHz.
+In addition, the most powerful integrated GPU found on Haswell CPUs, the HD 5200, has 40 execution units capable of processing sixteen single-precision floating point values at 1.3 GHz.
 This adds
@@
 Modern GPUs can also be interpreted as a unit consisting of multiple SIMD cores (Fig.\,\ref{fig:cpu_gpu}).
-The recent Kepler architecture by Nvidia have up to 15 SIMD units capable of processing 192, 32 bit, floating point values per clock cycle at 745 MHz\footnote{Nvidia typically refers to a core as a unit capable of performing a FMA instruction on one single precision floating point value.
-Hence, in Nvidia's world, Kepler has $15*192=2880$ cores.}.
+The recent Kepler architecture by Nvidia can provide up to 15 SIMD cores\footnote{Nvidia likes to refer to the unit capable of processing one single precision floating point value as a core.
+In Nvidia's world, Kepler therefore has $15*192=2880$ cores.} capable of processing 192 single-precision floating point values per clock cycle.
 These GPUs are also capable of performing FMA instructions, effectively doubling the number of theoretical FLOPS.
-The theoretical peak single precision FLOPS for this architecture is by that found to be
+The maximum theoretical peak single precision FLOPS for this architecture with a core frequency of 745 MHz is then found to be
 \begin{equation}
@@
 \end{equation} 
-Each SIMD core also comprise 16 special function units (\nom{SFU}{Special function unit (sin, exp, sqrt, ect.)}) which add close to 1 TFLOPS in additional computational power.
-This adds up to a total of five TFLOPS as shown in Fig.\,\ref{fig:cpu_vs_gpu} for the GeForce GTX 780 TI GPU.
+Each SIMD core also includes 16 special function units (\nom{SFU}{Special function unit (sin, exp, sqrt, ect.)}) which add close to 1 TFLOPS in additional computational power.
+This adds up to a total of 5 TFLOPS as shown in Fig.\,\ref{fig:cpu_vs_gpu} for the GeForce GTX 780 TI GPU.
 From Fig.\,\ref{fig:cpu_gpu} we see how the high number of SIMD cores found on GPUs and their larger width, compared with the SIMD cores on CPUs, results in five times the computational performance of a high-end CPU (without the integrated GPU).
 
-Note that the derived numbers are strictly theoretical and that actual throughput is highly algorithm dependent.
-Nevertheless, in theory, the benefit of GPU computing over CPU computing is currently a factor of five in raw computationally power.
-A claimed speed-up of several order of magnitude is typically a result of comparing a non-optimized CPU implementation with an optimized GPU implementation or by comparing uneven hardware \cite{Lee2010, Kothapalli2013}.
+Note that the derived numbers are strictly theoretical and that the actual throughput is highly algorithm dependent.
+Nevertheless, in theory, the computational power provided by modern GPUs is five times higher than that provided by CPUs.
+A claimed speed-up of several orders of magnitude is therefore typically a result of comparing a non-optimized CPU implementation with an optimized GPU implementation or by comparing uneven hardware \cite{Lee2010, Kothapalli2013}.
+However, in the next section we will see why it is often easier to obtain an optimized implementation for the GPU than for the CPU.
+
 In this section we have focused on CPUs from Intel and GPUs from Nvidia, however, CPUs and GPUs from AMD have similar specifications and differences in theoretical peak performance.
@@
 \subsection{Programming a GPU}
-Early GPGPU programming required expert knowledge about computer graphics and the remaining fixed functions in the graphics pipeline.
-When GPGPU gained interest and the GPU was used to compute problems that did not involve graphics, it quickly became evident that new programing languages where needed.
-One of the first GPGPU oriented languages for the GPU was the Brook language by Buck \textit{et al.} \cite{Buck2004}.
+Early GPGPU programming required expert knowledge about computer graphics and the graphics pipeline.
+When GPGPU gained a broader interest and the GPU started to be used to compute problems that did not involve graphics, it quickly became evident that new programing languages where needed.
+One of the first GPGPU oriented languages was the Brook language by Buck \textit{et al.} \cite{Buck2004}.
 Buck later joined Nvidia where he led the work on \nom{CUDA}{Compute unified device architecture (GPGPU framework)}, a GPGPU language and framework by Nvidia for Nvidia GPUs only.
 The first version of CUDA was released in 2007.
-The following year, Apple launched a GPGPU framework known as \nom{OpenCL}{Open Compute Language (GPGPU framework)}, a standard which are maintained by the Khronos Group.
+The following year, Apple launched a GPGPU framework known as \nom{OpenCL}{Open Compute Language (GPGPU framework)}, a standard which is maintained by the Khronos Group.
 These two frameworks are still today the main programming frameworks for GPGPU.
-Where CUDA only runs on Nvidia GPUs, OpenCL can now run on GPUs from both Nvidia and AMD, CPUs from both Intel and AMD, FPGAs from Altera, and mobile processors from \nom{ARM}{Designer of mobile chip architectures}.
+While CUDA only runs on Nvidia GPUs, OpenCL can now run on GPUs from both Nvidia and AMD, CPUs from both Intel and AMD, FPGAs from Altera, and mobile processors from \nom{ARM}{Designer of mobile chip architectures}.
 
@@
 
-Both CUDA and OpenCL is centered around a kernel function which are executed by a large group of threads.
+Both CUDA and OpenCL are centered around a kernel function which is executed by a large group of threads.
 The kernel function, for both CUDA and OpenCL, has a C-like syntax with some additional specifiers.
 An example of a CUDA kernel is depicted in Fig.\,\ref{fig:gpu_kernel}.
-The \texttt{\_\_global\_\_} specifier tells the CUDA compiler that this function is a kernel function, and in CUDA, this kernel function is launched by using a CUDA specific syntax:
+The \texttt{\_\_global\_\_} specifier tells the CUDA compiler that this function is a kernel function, and in CUDA this kernel function is launched by using a CUDA-specific syntax:
 \begin{align*}
@@
 \end{align*}
-where \texttt{grid} and \texttt{block} is 3-component vectors specifying grid and block size in "xyz" as depicted in Fig.\,\ref{fig:gpu_grid}, \texttt{a} is a pointer to a memory buffer located on the GPU, and \texttt{N}$*$\texttt{N} is the size of the buffer \texttt{a}.
+where \texttt{grid} and \texttt{block} are 3-component vectors specifying grid and block size in "xyz" as depicted in Fig.\,\ref{fig:gpu_grid}, \texttt{a} is a pointer to a memory buffer located on the GPU, and \texttt{N}$*$\texttt{N} is the size of the buffer \texttt{a}.
 
-When the kernel is launched, a copy of it is scheduled for each thread in each block, and each kernel instance gets as input its location in the compute grid (\texttt{blockIdx} and \texttt{threadIdx}) together with the grid size (\texttt{gridDim} and \texttt{blockDim}).
-This information is then typically used to calculate which memory elements to gather from global memory, in this case the elements in \texttt{a}.
+When the kernel is launched, a copy of it is scheduled for each thread in each block, and each kernel instance gets as input its location in the grid (\texttt{blockIdx} and \texttt{threadIdx}) together with the grid size (\texttt{gridDim} and \texttt{blockDim}).
+This information is then typically used to calculate which values to gather from the global memory, in this case an element of \texttt{a}.
 %Behind the scene, thread blocks are divided into smaller sets of threads called a warp (or wavefront in OpenCL).
-When a kernel is executed, multiple thread blocks are scheduled to each SIMD core where they stay until all instructions in the kernel is finished for all threads in the block.
-Thus,  in order to formulate a given problem as a kernel function the global memory input to each thread must not depend on the output from other blocks, hence, it must be possible for the GPU to schedule blocks in any order.
+When a kernel is executed, multiple thread blocks are scheduled to each SIMD core where they stay until all instructions in the kernel are finished for all threads in the block.
+Thus,  in order to formulate a given problem as a kernel function, the global memory input to each thread must not depend on the output from other blocks.
+Hence, it must be possible for the GPU to schedule blocks in any order.
 
 As we see, a thread has a different conceptual meaning on a GPU than on a CPU.
-Where a CPU is said to execute threads which can perform SIMD instructions, a thread on a GPU can be interpreted as unit capable of processing one data element in the SIMD instruction.
-This is why GPU threads are said to be light weighted.
-This conceptual difference also means that there is an inherent focus on breaking algorithms down and into SIMD instructions when programing for the GPU.
+While a CPU is said to execute threads which can perform SIMD instructions, a thread on a GPU can be interpreted as a unit capable of processing one data element in the SIMD instruction.
+The GPU is because of this sometimes referred to as a single instruction multiple threads (\nom{SIMT}{Single instruction multiple threads}) processor instead of a SIMD processor.
+This also explain why GPU threads are said to be light-weighted.
+Such a conceptual difference means that there is an inherent focus on breaking algorithms down into SIMD instructions when programing for the GPU.
 This often leads to a highly optimized parallel implementation on the GPU and not on the CPU.
-Threads, or SIMD instructions, is just how one program for the GPU, there is no other way of doing it.
+Manually typed SIMT, or SIMD, instructions is the only way of programming the GPU, while a single thread with automatic vectorization of for-loops is the standard way of programming the CPU.
 
 It is important to understand that not all algorithms will benefit from wider SIMD registers.
-As the width increase the only algorithms that will benefit is those who can be divided into enough concurrent fine granular parts.
-Also, to achieve peak performance, the implementation must perform one FMA operation on each core per clock cycle and all memory latency must be hidden by issuing instructions.
-All of this is something which is impossible to achieve in real life applications.
-One will always end up with a throughput which is a fraction of the peak throughput.
+As the width increase, the only algorithms that will benefit are those which can be divided into enough independent calculations.
+Also, to achieve the peak performance, the implementation must perform one FMA operation on each core per clock cycle, and it must be possible for the GPU to hide all memory latency by scheduling threads which are not stalled by a memory fetch.
+All of this is something which is impossible to achieve in practical applications, and therefore one will always end up with a throughput which is a fraction of the peak throughput.
 
-The code in Fig.\,\ref{fig:gpu_grid} is of course a minimal example of GPU programing.
+The code in Fig.\,\ref{fig:gpu_grid} is an incomplete and minimal example of a GPU program.
 A full example will involve several lines of C++ code transferring data to and from the GPU, and the kernel could include manual administration of the L1 cache (shared memory), and intra-block thread barriers.
-This is some of the reasons why writing a high performance GPU kernel is often found to be a complex and time consuming task.
-Because of the major overhead involved with writing GPU code we will, in the future, probably see a lot of work aimed at reducing it.
+This is part of the reasons why writing a high performance GPU kernel is often found to be a complex and time consuming task.
+Therefore, we will probably see a lot of work aimed at reducing this overhead in the future.
 A recent development is the introduction of C++ AMP which enables developers to write accelerated code (e.g.
 by a GPU) in plain C++.
-Hopefully, in near future, GPU programming will be a job for compilers, and direct fiddling with GPU kernels should only be needed if maximum throughput is absolutely needed.
+Hopefully, in near future, programming the GPU will be a job for compilers, and direct fiddling with GPU kernels should only be needed if maximum throughput is absolutely needed.
 This is similar to how SIMD instructions are handled in CPU-code today.
 Most of the time developers will rely on the CPU compilers to analyze the code and auto-apply SIMD instruction whenever it is possible.
-However, when maximum throughput is needed, developers with deep knowledge of the algorithm are often able to make faster code by manually writing SIMD instruction code.
-The relevance of auto-applied GPU instructions by compilers will also increase due to the recent embedding of GPUs on the same chip as the CPU.
-Soon, when writing code, we will probably talk about coding a heterogeneous collection of cores where operations are automatically offloaded to the right core (e.g.
+However, when maximum throughput is needed, developers with deep knowledge of the algorithm are often able to make faster code by manually writing SIMD instruction.
+The relevance of auto-applied GPU instructions will also increase due to the recent embedding of GPUs on the same chip as the CPU.
+Soon, GPU programming will probably be a matter of coding a heterogeneous collection of cores where operations are automatically offloaded to the right core (e.g.
 a GPU) by the compiler.
@@
 The idea behind BS-Capon beamforming is that spatial beams with little information can be removed from $\mat{\hat{R}}$ by removing rows in $B$.
-When investigating complexity of BS-Capon beamformer \textbf{Paper II} we find that is dominated by the transformation of data into beamspace, $O(N_LLN_b)$, where \nom{$N_b$}{Number of selected beams in beamspace} is the number of selected beams.
+When investigating the complexity of the BS-Capon beamformer (\textbf{Paper II}) we find that is dominated by the transformation of data into beamspace, $O(N_LLN_b)$, where \nom{$N_b$}{Number of selected beams in beamspace} is the number of selected beams.
 
@@
 Therefore, to simulate a field of one million points with several thousand source points at 50 FPS, more than 100 GFLOPS is required.
-Note that this simulator evaluates a cosine and sine function for each calculation, something which makes it even harder to achieve the required throughput, since especially GPUs have less silicon devoted to such tasks.
+Note that this simulator evaluates a cosine and sine function for each calculation, something which makes it even harder to achieve the required throughput, since the GPU has less silicon devoted to such tasks (SFUs).
 

### 2014-03-30 ThesisJP/Background/background.tex
Added new figure to the medical ulttrasound imaging sector.

@@
 \section {Medical ultrasound imaging}\label{sec:ultrasound}
-Ultrasound imaging encompasses technology which generates images based on sound whose frequencies we can not hear.
-For medical ultrasound imaging frequencies between 2 MHz and 10 MHz are typically used.
-Which frequency to select depends on how deep into the body we want to image, this because human tissue exhibits frequency dependent attenuation.
-Frequency dependent attenuation is the reason why a frequency around 3 MHz is used for fetal and cardiac imaging and a frequency around 7 MHz is used for more shallow imaging (e.g.
-vascular imaging).
-
 Medical ultrasound imaging is considered to be a low-risk tool for non-invasive investigation.
-There are no gamma rays or high intensity magnetic fields involved, just sound pressure below well controlled limits.
-It is also far less costly than other imaging modalities and offers real-time inspection of the human body.
-Ultrasound imaging is well know for its use in fetal imaging during pregnancy, however, in this thesis we will focus on a different modality, namely cardiac ultrasound imaging.
+Neither gamma rays nor high intensity magnetic fields are involved, just sound pressure below well-controlled limits.
+It is also far less costly than other medical imaging modalities and offers real-time inspection of the human body.
+Ultrasound imaging is well known for its use in fetal imaging during pregnancy, however, in this thesis we will focus on a different modality, namely cardiac ultrasound imaging.
 An illustration of the human heart is therefore included in Fig.\,\ref{fig:human_heart}.
@@
 The tissue separating the left and right ventricle is know as the inter-ventricular septum.
-The left side is the strongest and hardest working part of the heart, and it is responsible of pumping blood out to the whole body through the aortic valve.
-When blood flows out the aorta it is important that the mitral valve, that separates the left atrium and ventricle, is closed and does not leak.
-The left ventricle and the mitral valve therefore receives special attention during cardiac ultrasound examinations.
+The left side is responsible of pumping blood out to the whole body through the aortic valve.
+It is therefore the strongest and hardest working part of the heart.
+When blood flows out of the aorta it is important that the mitral valve, which separates the left atrium from the left ventricle, is closed and does not leak.
+The left side, and especially the left ventricle and the mitral valve, therefore receives the most attention during cardiac ultrasound examinations.
 When visualizing 3D ultrasound images in \textbf{Paper IV} we will have a strong focus on visibility of cardiac tissue, thus note for later that the endocardium refers to the innermost layer of the cardiac muscle.
@@
 \centering
-\subfigure{
-	\includegraphics[width=0.47\textwidth]{img/Diagram_of_the_human_heart.png}
+\subfigure[]{
+	\includegraphics[width=0.47\textwidth]{img/Diagram_of_the_human_heart.png}\label{fig:human_heart}
 }
-\subfigure{
-	\includegraphics[width=0.47\textwidth]{img/HeartWall.png}
+\subfigure[]{
+	\includegraphics[width=0.47\textwidth]{img/ultrasound_grid.eps}\label{fig:us_grid}
 }
-\caption{Overview of the human heart. Illustrations from wikipedia.org.}
-\label{fig:human_heart}
+\caption{a) Overview of the human heart. Illustration from wikipedia.org. b) Coordinate system for phased array imaging.}
 \end{figure}
+
+%Ultrasound imaging encompasses technology which generates images based on sound whose frequencies we can not hear.
+For medical ultrasound imaging frequencies between 2 MHz and 10 MHz are typically used.
+The choice of frequency depends on the depth of the body layers we want to image, because human tissue exhibits frequency dependent attenuation.
+This is why a frequency around 3 MHz is used for fetal and cardiac imaging and a frequency around 7 MHz is used for more shallow imaging (e.g.
+vascular imaging).
 
@@
 Piezoelectric elements organized in an array, is referred to as an ultrasound \textit{probe}, and a special probe is typically designed for each medical imaging modality.
-For cardiac ultrasound imaging the probe must not only be designed for a certain frequency range, its footprint also needs to be smaller than the intercostal window in order to avoid shadowing by ribs.
-A line in an ultrasound image is generated by firing a ultrasound pulse into the body and sample the received echo.
-This amplitude trace is known as an A-mode image.
+%For cardiac ultrasound imaging the probe must not only be designed for a certain frequency range, its footprint also needs to be smaller than the intercostal window in order to avoid shadowing by ribs.
+A line in an ultrasound image is generated by firing a ultrasound pulse into the body and sample the received echo. This amplitude trace is known as an A-mode image.
 

### 2014-03-30 ThesisJP/Background/background.tex
Added several images. Text has been updated.

@@
 Especially note the four chambers of the heart, the left and right atrium, and the left and right ventricle.
-The tissue separating the left and right ventricle is know as the inter-ventricular septum.
+The tissue separating the left and right ventricle is known as the inter-ventricular septum.
 The left side is responsible of pumping blood out to the whole body through the aortic valve.
@@
 \subfigure[]{
-	\includegraphics[width=0.47\textwidth]{img/ultrasound_grid.eps}\label{fig:us_grid}
+	\includegraphics[width=0.47\textwidth]{img/ultrasound_grid.pdf}\label{fig:us_grid}
 }
-\caption{a) Overview of the human heart. Illustration from wikipedia.org. b) Coordinate system for phased array imaging.}
+\caption{a) Overview of the human heart. Illustration from wikipedia.org. b) Illustration of probe geometry and layout for phased array imaging.}
 \end{figure}
@@
 
-To generate the ultrasonic frequencies used for imaging a piezoelectric ceramic is used.
+To generate the ultrasonic frequencies used for imaging, a piezoelectric ceramic is used.
 A piezoelectric material is special in that way that ultrasound vibrations are generated when voltage is applied and \textit{vice versa}.
@@
 Piezoelectric elements organized in an array, is referred to as an ultrasound \textit{probe}, and a special probe is typically designed for each medical imaging modality.
+An example of a probe for medical ultrasound imaging is shown in Fig.\,\ref{fig:us_grid}.
 %For cardiac ultrasound imaging the probe must not only be designed for a certain frequency range, its footprint also needs to be smaller than the intercostal window in order to avoid shadowing by ribs.
-A line in an ultrasound image is generated by firing a ultrasound pulse into the body and sample the received echo. This amplitude trace is known as an A-mode image.
 
-\subsection{Beamforming}
-If the width of a piezoelectric element is less than the wavelength of the transmitted signal the element is said to be \textit{omnidirectional}, hence energy is distributed or sensed equally in all directions.
-When multiple omnidirectional elements are arranged in a linear array the transmitted pulse will get shaped into a beam\footnote{A beam refers to the pulse in space integrated across time.} of ultrasound by diffraction as the size of the array  (the \textit{aperture}) grows.
-The 3 dB angular width of this beam, and also the angular resolution of the array, is given by \cite{AngelUltrasound}:
+If the width of a piezoelectric element is less than the wavelength of the transmitted signal the element is said to be \textit{omnidirectional}, hence, energy will be distributed or sensed equally in all directions.
+When multiple omnidirectional elements are arranged in an array the transmitted pulse will get shaped into a beam\footnote{A beam refers to the pulse in space integrated across time.} of ultrasound by diffraction as the size (the \textit{aperture}) of the array grows.
+The angular width of this beam, and also the angular resolution of the array, is approximately given by \cite{AngelUltrasound}:
 \begin{align}\label{eq:res}
@@
 \end{align}
-\nomi{$\delta\theta_a$}{Angular resolution of an aperture}where \nom{$\lambda$}{Wavelength} and \nom{$f$}{Frequency} is the ultrasound wavelength and frequency respectively, $c$ is the speed of sound, and \nom{$D$}{Aperture size} is the aperture width.
-With an average speed-of-sound in human tissue of $1540\frac{\text{m}}{\text{s}}$, the angular resolution or opening angle of a 2 cm cardiac probe operating at 3 MHz is around 1.5\degree, which at 8 cm range gives a lateral resolution of around 2 mm.
+\nomi{$\delta\theta_a$}{Angular resolution of an aperture}where \nom{$\lambda$}{Wavelength} and \nom{$f$}{Frequency} is the ultrasound wavelength and frequency respectively, $c$ is the speed of sound in human tissue, and \nom{$D$}{Aperture size} is the aperture width.
+With an average speed-of-sound in human tissue of $1540\frac{\text{m}}{\text{s}}$, the angular resolution or opening angle of a 2 cm cardiac probe operating at 3 MHz is around 1.5\degree.
 The axial resolution is inverse proportional to the pulse length, and is typically better than the lateral resolution.
+The geometry of cardiac ultrasound imaging is depicted in Fig.\,\ref{fig:us_grid}.
+A line in an ultrasound image is generated by firing a ultrasound pulse or beam into the body and sample the received echo.
+We will refer to this amplitude trace as an \textit{image line}.
+
+\subsection{Beamforming}
+If time delays are applied on the signal emitted by the elements it is possible to steer and focus the ultrasound beam in space (Fig.\,\ref{fig:das}).
+This technique is known as \textit{beamforming}, and for ultrasound imaging it is applied both when transmitting and receiving ultrasound pulses.
+A cardiac probe is a phased probe, which means that the image is constructed from image lines which are gradually steered from $-\theta/2$ to $\theta/2$ radians.
+As depicted in Fig.\,\ref{fig:us_grid}, this results in an imaging sector which is $\theta$ radians wide.
+This is known as a 2D B-mode image (B for brightness).
+If image lines are acquired for multiple B-mode images which are tilted in the $yz$-plane we get what is known as a volumetric B-mode or a 3D ultrasound image.
+A probe for 3D imaging has several thousand elements arranged as a matrix in order to facilitate steering in both the $xz$ and $yz$-plane.
+In this thesis we will focus on 2D imaging when the topic is adaptive beamforming, and 3D imaging when the topic is adaptive visualization.
 
@@
 
-If time delays are applied on the signal sent to the different elements it is possible to steer and focus the ultrasound beam in space (Fig.\,\ref{fig:das}).
-This technique is known as \textit{beamforming}, and for ultrasound imaging it is traditionally applied both when transmitting and receiving ultrasound pulses.
-For medical ultrasound imaging there are two main types of ultrasound probes, linear and phased probes \cite{AngelUltrasound}.
-A cardiac probe is a phased probe, which means that the image is constructed from beams, or A-mode images, which are gradually steered from $-\theta/2$ to $\theta/2$ radians, resulting in an imaging sector which are $\theta$ radians wide.
-This kind of image, where A-mode images are acquired from multiple spatial locations in the $xz$-plane, is known as a B-mode image (B for brightness).
-If A-mode images are acquired from the $yz$-plane as well we get what is known as a volumetric B-mode or 3D ultrasound image.
-A probe for 3D imaging has several thousand elements arranged as a matrix.
-In this thesis we will focus on linear arrays for 2D imaging when the topic is adaptive beamforming, and 3D imaging when the topic is adaptive visualization.
-
-If \nom{$M$}{Number of array elements} is the number of elements in the array, $x_m$ is the per-element data, and \nom{$\Delta_m$}{Per-element focusing and steering delays} is the per-element focusing and steering delay, the standard form of beamforming used on receive in ultrasound imaging is:
+The standard form of beamforming used on receive in ultrasound imaging is known as \textit{delay-and-sum} (DAS) beamforming: 
 \begin{align}\label{eq:das}
@@
 \end{align}
-This is known as \textit{delay-and-sum} (DAS) beamforming.
-Here, \nom{$\vec{w}$}{Array weight vector (apodization)} is the array \textit{weight vector} or \textit{apodization}, $\vec{x}$ is a pre-delayed vector of element data, and $z[n]$ is the output or the A-mode image in the direction given by $\Delta_m[n]$.
-From now on $x_m[n]$ will refer to its delayed version $x_m[n-\Delta_m[n]]$.
+Here, \nom{$M$}{Number of array elements} is the number of elements in the array, $x_m$ is the per-element data, \nom{$\Delta_m$}{Per-element focusing and steering delays} is the per-element focusing and steering delay, \nom{$\vec{w}$}{Array weight vector (apodization)} is the array \textit{weight vector} or \textit{apodization}, $\vec{x}$ is a pre-delayed vector of element data, and $z[n]$ is the output image line in the direction given by $\Delta_m[n]$.
+From now on we will omit the dependency on $\Delta_m[n]$ and $x_m$ will refer to delayed element data.
 A schematic overview of delay-and-sum array beamforming is depicted in Fig.\,\ref{fig:das}.
-If $p_o$ is the phase center of our array and $p$ is the point where we want to steer and focus the ultrasound beam, we have to use the following delays:
+If $p_o$ is the phase center of our array and $p$ is the point to which we want to steer and focus the ultrasound beam, we have to use the following delays:
 \begin{align}
@@
 On transmit the inverse set of delays has to be used.
-Also note that on transmit we have to decide where to steer and focus prior to transmitting a pulse, transmit delays therefore do not vary within an A-mode image.
+Also note that on transmit we have to decide where to steer and focus prior to transmitting a pulse.
+Transmit delays therefore do not vary within an image line.
 
@@
 We have already seen that the lateral resolution of an ultrasound probe is given by (\ref{eq:res}).
-Hergum \textit{et al.} \cite{Hergum2009} have further shown that the lateral frequency support for a system which both emit and receive energy is governed by the sum of the transmit (\nom{tx}{Transmit}) and receive (\nom{rx}{Receive}) aperture.
+Hergum \textit{et al.} \cite{Hergum2009} have further shown that the lateral frequency support for a system which both emit and receive energy is governed by the sum of the transmit ($D_{tx}$\nomi{tx}{Transmit}) and receive ($D_{rx}$\nomi{rx}{Receive}) aperture.
 The lateral resolution of what is known as an active imaging system is found to be
@@
 \end{align}
-\nomi{$N_\theta$}{Number of lines required for proper sampling of an imaging sector}A-mode images uniformly distributed across the sector.
-If \nom{$r$}{Imaging range.} is the maximum range we want to image we get an imaging frame rate of
+\nomi{$N_\theta$}{Number of lines required for proper sampling of an imaging sector}image lines distributed across the imaging sector.
+If \nom{$r$}{Imaging range.} is the maximum range in our image, as shown in Fig.\,\ref{fig:us_grid}, we can acquire
 \begin{align}
@@
 \end{align} 
-frames per second (\nom{FPS}{Frames per second}). With our 2 cm cardiac probe we get a frame rate of around $55$ FPS when imaging a 70\degree{} sector with $r=15$ cm.
+images or frames per second (\nom{FPS}{Frames per second}). With our 2 cm cardiac probe we get a frame rate of around $55$ FPS when imaging a 70\degree{} sector with $r=15$ cm.
 

### 2014-03-30 ThesisJP/Background/background.tex
Finished fixing text based on comments from Ines.

@@
 The geometry of cardiac ultrasound imaging is depicted in Fig.\,\ref{fig:us_grid}.
-A line in an ultrasound image is generated by firing a ultrasound pulse or beam into the body and sample the received echo.
+A line in an ultrasound image is generated by firing an ultrasound pulse or beam into the body and sample the received echo.
 We will refer to this amplitude trace as an \textit{image line}.
@@
 The required axial sampling is dictated by the transmit frequency and the frequency bandwidth of the probe.
-From Fourier theory we know that a short impulse consist of many frequency components.
-A short ultrasound pulse, which would yield high axial resolution, therefore requires a large frequency bandwidth.
-An ultrasound system has normally a relative bandwidth between 50 \% and 100 \%.
-This corresponds to transmitting a pulse which is between one or two oscillations when entering tissue.
-In the front-end of an ultrasound imaging system a high sampling frequency is typically used (e.g.
+From Fourier theory we know that a short pulse consists of many frequency components.
+A short ultrasound pulse, which would yield a high axial resolution, requires therefore a large frequency bandwidth.
+Medical ultrasound systems typically has a bandwidth between 33 \% and 100 \% of the pulse frequency.
+This corresponds to transmitting a pulse which is between 1 and 3 oscillations when entering tissue.
+In the analog-to-digital converter of an ultrasound imaging system a high sampling frequency is typically used (e.g.
 40 MHz).
-However, the ultrasound signal is band limited, and the data rate can therefore be reduced in a process known as \textit{\nom{IQ}{In-phase quadrature (ultrasound data format).} demodulation}.
+However, the ultrasound signal is band limited, and the data rate can therefore be reduced in a process known as \textit{in-phase quadrature (\nom{IQ}{In-phase quadrature (ultrasound data format).}) demodulation}.
 In this process the axial sampling rate is reduced to the bandwidth of the imaging system in the following three steps: demodulation with a complex exponential, low-pass filtering, and decimation.
-The number of samples in one A-mode image is after this process given by:
+The number of samples in one image line is after this process given by:
 \begin{align}
-N = \frac{2r}{c}B,
+N = \frac{2r}{c}B_wf_c,
 \end{align}
-\nomi{$N$}{Number of samples in range}where \nom{$B$}{Frequency bandwidth} is the bandwidth, $B = B_wf_c$ (where \nom{$B_w$}{Relative bandwidth} is the relative bandwidth and \nom{$f_c$}{Center frequency of ultrasound pulse} is the center frequency of the transmitted pulse).
+\nomi{$N$}{Number of samples in range}where \nom{$B_w$}{Relative bandwidth} is the relative bandwidth in percent of \nom{$f_c$}{Center frequency of ultrasound pulse}, where $f_c$ is the center frequency of the transmitted pulse.
 Note that after this process the ultrasound data will be complex values.
@@
 \end{align}
-If our cardiac probe has a relative bandwidth of $80$ \% we have to process almost $50000$ $M$-long vectors of element data every $1/55$ second.
-The formula for the number of FLOPS required for real-time DAS beamforming of cardiac ultrasound is then given as:
+If our cardiac probe, which is used to image a $70$\degree{} sector, has a relative bandwidth of $80$ \% we have to process almost $50000$ $M$-long vectors of complex element data every $1/55$ second.
+The formula for the number of FLOPS required for real-time DAS beamforming of cardiac ultrasound is then roughly given as:
 \begin{align}
-\text{FLOPS}_{\text{DAS}} = MN_{\text{B-mode}}f_{\text{FPS}} = MB_wf_c
+\text{FLOPS}_{\text{DAS}} = MN_{\text{B-mode}}f_{\text{FPS}} = MB_wf_c.
 \end{align}
@@
 The main challange with DAS beamforming is the amount of data that has to be moved.
-If one element sample is two bytes long, for our example, around 600 MB has to be moved on to the GPU every second.
-This is however less than the maximum bandwidth of PCI Express 2.0 (8 GB/s), even after the relative bandwidth, transmit frequency and number of array elements are increased with a total factor of 10.
-Newer versions of PCI Express which supports even higher transfer rates have also been developed.
+In our example, for a sample size of four bytes, around 1.2 GB has to be moved to the GPU every second.
+This is however less than the maximum bandwidth of PCI Express 2.0 (8 GB/s), even after the relative bandwidth, transmit frequency and number of array elements are increased with a total factor of 5.
+Newer versions of PCI Express which supports even higher transfer rates have also recently been introduced.
 
@@
 \subsection{GPUs in medical ultrasound imaging}
-Ultrasound imaging systems generates an incredible amount of data.
-A modern system capable of 3D imaging will generate several tera bytes of data in the front end of the system during a clinical scan.
-The data rate from a 2D scan is "only" in order of giga bytes per second prior to IQ-demodulation.
-To handle these amount of data ultrasound imaging systems have, for the last twenty years, relied heavily on application-specific integrated circuits (\nom{ASIC}{Application-specific integrated circuit}) for all steps in the processing pipeline \cite{Thomenius}.
-However, as the computationally power of CPUs have increased, ASICs, e.g.
-for converting data from a polar to a Cartesian grid (known as scan conversion), have gradually been converted into software \cite{Guracar2013}.
-One of  the last remaining ASICs in ultrasound scanners is the beamformer \cite{Thomenius}, whose job is to steer and focus the ultrasound beam in space.
-This operation involves a reduction of the data rate with a factor equal to the number of elements in the ultrasound probe.
-The input data rate is therefor very high.
+Ultrasound scanners generate an incredible amount of data.
+A modern system capable of 3D imaging will generate several tera bytes of data inside the probe handle during a clinical scan.
+Fortunately the data rate is reduced to the level of 2D imaging before the data reaches the scanner.
+The data rate of a 2D scan is "only" in order of giga bytes per second.
+To handle this amount of data, ultrasound imaging systems have, for the last twenty years, relied heavily on application-specific integrated circuits (\nom{ASIC}{Application-specific integrated circuit}) for all steps in the processing pipeline \cite{Thomenius}.
+However, as the computational power of generic processors have increased, ASICs have gradually been replaced by software \cite{Guracar2013}.
+One of  the last remaining ASICs in ultrasound scanners is used for beamforming \cite{Thomenius}.
+As seen from (\ref{eq:das}), beamforming involves a reduction of the data rate with a factor equal to the number of elements in the ultrasound probe.
+The input data rate is therefore very high.
 Nevertheless, real-time beamforming has recently been achieved on a GPU \cite{Song2012}, and a French company (SuperSonic Imagine) are utilizing GPUs in their plane wave beamforming algorithm \cite{Tanter2014}.
@@
 The largest players in the medical ultrasound business, General Electric and Philips, are also well aware of where the trend is going \cite{Thomenius2012} \cite{Metz2011}.
-Software beamforming opens up new possibilities for advanced processing of ultrasound data.
+Software beamforming opens up for new possibilities for advanced processing of ultrasound data.
 One example is Capon beamforming which is explored in this thesis (\textbf{Paper\,I} and \textbf{II}), and which is introduce in the next section.

### 2014-04-05 ThesisJP/Paper4/spie.tex
Added Paper 4.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
\begin{abstract}
In this paper we introduce and investigate an adaptive direct volume rendering (DVR) method for real-time visualization of cardiac 3D ultrasound.  DVR is commonly used in cardiac ultrasound to visualize interfaces between tissue and blood. However, this is particularly challenging with ultrasound images due to variability of the signal within tissue as well as variability of noise signal within the blood pool. Standard DVR involves a global mapping of sample values to opacity by an opacity transfer function (OTF). While a global OTF may represent the interface correctly in one part of the image, it may result in tissue dropouts, or even artificial interfaces within the blood pool in other parts of the image. In order to increase correctness of the rendered image, the presented method utilizes blood pool statistics to do regional adjustments of the OTF. The regional adaptive OTF was compared with a global OTF in a dataset of apical recordings from 18 subjects. For each recording, three renderings from standard views (apical 4-chamber (A4C), inverted A4C (IA4C) and mitral valve (MV)) were generated for both methods, and each rendering was tuned to the best visual appearance by a physician echocardiographer. For each rendering we measured the mean absolute error (MAE) between the rendering depth buffer and a validated left ventricular segmentation. The difference $d$ in MAE between the global and regional method was calculated and t-test results are reported with significant improvements for the regional adaptive method ($\bar{d}_{A4C} = 1.5 \pm 0.3$ mm, $\bar{d}_{IA4C} = 2.5 \pm 0.4$ mm, $\bar{d}_{MV} = 1.7 \pm 0.2$  mm, d.f. $= 17$, all $p<0.001$). This improvement by the regional adaptive method was confirmed through qualitative visual assessment by an experienced physician echocardiographer who concluded that the regional adaptive method produced rendered images with fewer tissue dropouts and less spurious structures inside the blood pool in the vast majority of the renderings. The algorithm has been implemented on a GPU,  running an average of 16 fps with a resolution of 512x512x100 samples (Nvidia GTX460).
\end{abstract}

\graphicspath{{./Paper4/img/}}

\section{Introduction}
Direct volume rendering (DVR)\cite{Levoy1988} is commonly used in cardiac ultrasound to visualize interfaces between tissue and blood\cite{steen1994, Rabben2010, kiss2010}. DVR has the advantage over other methods like 2D-slicing that it presents a large part of the volumetric data in a single view. Figure \ref{fig:dvr}a presents such a volume rendering of a high quality cardiac ultrasound volume. Anatomical structures of special interest are highlighted; the left ventricle (LV), the mitral valve which separates LV from the left atrium (LA), and the interventricular septum which separates the LV from the right ventricle (RV). As can be seen, interpreting cardiac ultrasound data is difficult and requires a lot of training. Clear tissue-blood boundaries are therefore of great importance. However, this is challenging to achieve with ultrasound images due to variability of the signal within tissue as well as variability of noise signal within the blood pool.
\begin{figure}  
\begin{center}
\begin{minipage}[b]{0.5\linewidth}\centering
\includegraphics[height=1\textwidth]{vr_cardiac_high_quality.eps}\\
(a)
\end{minipage}%
\hspace{-5em}
\begin{minipage}[b]{0.6\linewidth}\centering
\includegraphics[height=0.4\textwidth]{otf2.eps}\\
(b)
\end{minipage} 
\end{center}
\caption[example] { \label{fig:dvr}  a). An example of direct volume rendering (DVR) of cardiac ultrasound. Highlighted anatomical structures are: The left ventricle (LV), the right atrium (RA), mitral valve and the interventricular septum.  b). Shows two opacity transfer functions (OTF) with a fuzzy transition between voxels excluded from (values below $\tau$) or included in the rendering (values above $\tau$). Adjusting $\tau$ (blue arrow) will either include or exclude more data.}
\end{figure} 
%DVR is based on a physical model of light moving through a gaseous medium, and for simplicity, only absorption and emission along each ray are considered in its standard formulation.

At the core of DVR, optical properties (colour and opacity) are derived from a transfer function and integrated along rays of light. One ray for each pixel in the final image. Applying the transfer function is also known as classification since features are visually separated through assignment of different colours and opacities. On the other hand, selecting a transfer function with low classification error is considered a challenge and will always be related to the data at hand \cite{pfister2002transfer}. Figure \ref{fig:dvr}b shows an example of a monotonically increasing opacity transfer function (OTF) for cardiac ultrasound DVR. Unlike other medical imaging modalities like CT and MR, cardiac ultrasound data are assumed to consist of only two classes; tissue and the blood pool signal. Tissue is expected to have higher intensity than blood, but in practice the two class-distributions overlap. The OTF in Figure \ref{fig:dvr}b is therefore designed to give a fuzzy threshold between samples excluded from and samples included in the rendering. Selecting the right threshold, as for transfer functions design in general, is therefore a challenge in cardiac ultrasound visualization.

There are several reasons for the overlap between tissue and blood pool signal in ultrasound images. Sidelobes of the ultrasound beam results in signal from off-beam structures (clutter noise). Speckle pattern, caused by interference between point scatterers results in high signal variance in homogeneous regions. Multiple reflections between tissue structures (reverberations) causes false echoes. Distortion of the ultrasound wave front caused by varying speed of sound in inhomogeneous tissue (aberrations) reduce the signal amplitude. Interfaces of high reflectivity (lungs, ribs and calcifications) and how tissue layers are aligned with the ultrasound beam result in drop-outs and varying tissue signal. Attenuation of the ultrasound signal with depth gives high variation in tissue intensity across the volume. While an optimal global OTF may represent the tissue-blood interface correctly in one part of the image, the high variation in tissue intensities may result in tissue dropouts, or even artificial interfaces within the blood pool in other parts of the image. The latter artefact is visible in the upper part of LV in Figure \ref{fig:dvr}a. Figure \ref{fig:dvr}a also shows a large dropout in the lateral wall. These problems are to some extent solved by clipping, but standard clipping also has its limitations due to planar geometry. The curved shape of cardiac chambers calls either for other clipping geometries or a way to better assign high opacity to voxels associated with tissue and low opacity otherwise\cite{patent}.

A lot of work has been done in the visualization community to improve the results given by standard DVR. Special attention has been devoted to the problem of selecting the right transfer function\cite{Kindlmann1998, sereda2006visualization, wesarg2d, Woodring2009, Haidacher2010, correa2010visibility, Wang2011} and exploration of illustrative or non-photorealistic techniques to bring important information into focus\cite{viola2005, bruckner2006illustrative, Rezk-Salama2006, Malik2007, Bruckner2009}. Recently, Marchesin et al. \cite{marchesin2010} introduced a relevance function for feature enhancement in volume rendering. The paper differs from earlier contributions by also suggesting how to derive a relevance map without having any prior knowledge of voxel priorities. However, the use of local gradients as relevance indicators together with the per-ray constant opacity, yielding transparent renderings, makes it unsuitable for cardiac ultrasound. In our approach we adapt the ideas given by Marchesin et al. on how to do local opacity modulations, but with a relevance map better suited for cardiac ultrasound. For medical applications many adaptive and semi-adaptive DVR-like techniques have been proposed for enhancing anatomical structures\cite{borland2006volumetric, lindholm2010, 10.1109/TVCG.2006.100, Zhang2008}. For ultrasound, H\"{o}nigmann et al.\cite{Honigmann2003} proposed how to design a global adaptive OTF used to visualize volumetric fetal images. They also discussed the possibility of calculating one OFT for different regions in the rendering. Such a regional OTF can be derived from e.g. local edge detection or histograms. The method proposed in this paper regionally adapts the OTF by utilizing blood pool statistics derived from a real-time LV endocardial tracking algorithm\cite{orderud2006}. The purpose is to increase the correctness of the rendered image by reducing artificial tissue dropouts and spurious structures inside the blood pool.
%Our approach is highly influenced by this discussion, and in this paper we explore both regional and per-ray adaptation of the OTF.

%For a visualization method to be applicable to cardiac ultrasound it needs be both fast and robust. Cardiac ultrasound is a real-time modality, so users expect immediate feedback on the display when moving the probe. In addition, the low SNR makes e.g. gradient calculation a challenge. Many of the proposed methods in the literature are based on semi-automatic transfer function design, depending on time-consuming user interaction. For automatic methods, boundaries are often expected to be related to the local gradient or histogram.

%limitations of the ultrasound imaging system causing a low signal-to-noise ratio (SNR)

\section{Methods}
\subsection{Blood pool based regional adaptive opacity transfer function}
Selecting the correct global OTF for cardiac ultrasound visualization can be viewed as a global thresholding problem. Therefore we define the opacity threshold $\tau$ for a monotonically increasing OTF $\alpha$ as the highest sample value $v$ where all samples with lower value than $v$ will get zero opacity, $\tau = \max_{i}(v_i), \; \alpha(v_j)=0 \;\; \forall \; j<i$, hence they will be rendered transparent. 
As OTF we use a truncated polynomial:
\begin{equation}
\alpha(v) = \left\{
	\begin{array}{rl}
		0, &  v < \tau,\\
		\textrm{min}(a(v - \tau)^{\gamma}, k), & v \ge \tau.
	\end{array} \right.
\label{eq:otf}
\end{equation}
Global transparency is controlled by $a\in[0,\infty\rangle$, and in this paper $a=80$ has been selected to give a steep OTF. The truncation and gamma value are set to $k=0.5$ and $\gamma=2$ respectively. A second order polynomial has been selected instead of a standard linear OTF to increase the transparency of values near the opacity threshold. These values have a higher probability for being noise, and we let the opacity reflect this.

\begin{figure}  
\begin{center}
\begin{tabular}{cc}
\includegraphics[height=0.35\textwidth]{lot.eps} & 
\includegraphics[height=0.35\textwidth]{lotf_mesh2.eps}\\ % TODO Remove mesh and LOTF lable from this image
(a) & (b)
\end{tabular}
\end{center}
\caption[example] { \label{fig:lotf}  a) Shows ideal bimodal histograms of two density evolutions along ray $r_i$ and $r_j$. If we apply a global OTF (orange) on the given case we must select one ray were we want to separate tissue from blood. However, with a regional adapted OTF (green) we can find individual thresholds $\tau_r$ that minimizes the amount of misclassified samples along both rays. b) Estimated values of mean $\mu_r$ and standard deviation $\sigma_r$ inside the blood pool are used to regionally adjust the OTF. The regional opacity threshold $\tau_r = \mu_r + c\sigma_r$ is calculated for each ray $r$ intersecting a given left ventricular delineation.}
\end{figure} 
Since the global distributions of tissue and blood pool voxels usually overlap, adjusting $\tau$ will always involve a tradeoff between 
the number of true-positive and false-positive tissue voxels. If the opacity threshold is set too high, tissue surfaces will start to fall apart. However, if set too low, low level noise will get high opacity and the rendering will end before the actual tissue border is reached. Figure \ref{fig:lotf}a depicts the problem of having a global OTF (in orange). If rays $r_i$ and $r_j$ have noise ($n_r$) and tissue ($s_r$) levels given by $n_i$, $s_i$, $n_j$, $s_j$, with $n_i < s_i < n_j < s_j$, there are two obvious choices for $\tau$. The global opacity threshold $\tau$ can be selected as $n_i < \tau < s_i$ or $n_j < \tau < s_j$. The first choice makes $n_j$ occlude $s_j$, however the second choice will render $s_i$ transparent. This will result in an artificial hole in the final rendering along $r_i$. With our regional adaptive OTF, the idea is to let the opacity threshold $\tau$ vary across the framebuffer, making us able to define individual opacity thresholds for $r_i$ and $r_j$ (shown in green in Figure \ref{fig:lotf}a).

In optical character recognition there has been a long tradition to utilize adaptive thresholding schemes to deal with variable background levels in 2D images\cite{Niblack1985}, thus maximizing the amount of foreground output. One simple adaptive method is based on estimates of regional statistics. The local image threshold $\tau_{x,y}$ is calculated as
\begin{equation}
\tau_{x,y} = \mu_{x,y} + c\sigma_{x,y},
\label{eq:niblack}
\end{equation}
where $c$ is a constant, and $\mu_{x,y}$ and $\sigma_{x,y}$ are mean and standard deviation estimates calculated in a window $w$ around (x,y). This yields what is known as a thresholding surface, where the method assumes that the foreground has higher expected value than the background inside $w$. Hence, foreground pixels are located above the thresholding surface.

To address our problem with high variability in tissue intensity in ultrasound images, we have designed a regional adaptive OTF by estimating the regional means and standard deviations of Equation \ref{eq:niblack} from the blood pool of the LV. The blood pool was delineated by a state-estimation algorithm capable of tracking the LV endocardium in real-time\cite{orderud2006}. The temporal state of the LV is returned by the framework as a time series of polygon meshes. How we utilize this information to do regional opacity adjustments is depicted in Figure \ref{fig:lotf}b. For a given frame we estimate the signal mean and standard deviation, $\mu_r$ and $\sigma_r$, along rendering ray $r$. Note that only the blood pool, the intersection between $r$ and the given LV delineation, is used to estimate $\mu_r$ and $\sigma_r$. The estimates are then used to calculate a regional opacity threshold $\tau_r$ given as
\begin{equation}
\tau_r=\mu_r+c\sigma_r, 
\label{eq:niblack_ray}
\end{equation}
where $c$ is a global constant adjusting $\tau_r$ around $\mu_r$ with a certain number of standard deviations. The local threshold $\tau_r$ together with Equation \ref{eq:otf} form our blood pool based regional adaptive OTF. Rays that do not intersect the given LV delineation are rendered using a global OTF.

\subsection{Parameter space}
The proposed method in Equation \ref{eq:niblack_ray} introduces a parameter $c$, which controls how many standard deviations we want $\tau_r$ to deviate from the per-ray blood pool mean $\mu_r$. This parameter is clearly a substitution for the global opacity threshold, and adjusting $c$ will in the same way control the level of data visible in the rendered image. Setting $c = 0$ will render approximately $50\%$ of the blood pool samples transparent. Since we have a robust estimate of blood pool statistics, tissue signals are expected to have $\mu_{tissue} > \mu_r$ and $c > 0$ should be used.

The shape of the OTF is of great importance for the final appearance of a volume rendering. An OTF with a gentle slope will for instance result in a transparent look. In the proposed method we are more interested in the opacity threshold than the actual shape of the OTF, therefore we use the simple shifted and truncated second degree polynomial from Equation \ref{eq:otf} as a basis for the local OTF. Because it is a simple closed form expression, it can be used to quickly calculate on-the-fly opacity values.

For thresholding 2D images, the adaptive method in Equation \ref{eq:niblack} requires a window size that is directly related to the size of foreground objects. In our method, this window size depends on the ray-blood-pool intersection, and ideally no foreground (tissue) is included in the estimates. The regional opacity threshold $\tau_r$ is used along the full length of $r$, not just the intersection. Intersections with too few samples to give reliable statistics are rendered using a global OTF.

\subsection{Spatial and temporal regularization}
Introducing adaptive visualization has some negative side effects. We trade spatial and temporal coherence in the presentation of anatomical structures for a better presentation of endocardial tissue. This does not pose any problems as long as the structure of interest along a given ray is associated with a value larger than the blood pool mean. Otherwise we risk rendering this structure transparent.
%The challenging part of estimating per-voxel relevance indicators in ultrasound data is to maintain spatial and temporal coherence of anatomical structures. 
%Since relevance in our method is proportional to intensity, structures of interest can be suppressed if they are not associated with a value larger than the blood pool estimate. 
The latter situation will often arise if e.g. the endocardium, mitral valve or papillary muscles are included in the estimated statistics. Given the properties of the real-time tracking framework, the two last structures, will to some extent, always be included. To exclude the endocardium we highly depend on a successful LV endocardial tracking. All segmentation algorithms will  be prone to errors and two regularization schemes have therefore been tested. First, we do spatial smoothing of estimates across neighbouring rays. This will constrain the regional adaptivity. Second, we do temporal smoothing over successive frames to reduce the per-frame adaptivity and potential flickering\cite{Peterscha2005}. The smoothing has been implemented as averaging filters on the framebuffer containing regional statistics. All regional adaptive renderings and statistics have been produced without any regularization in this paper. However the regularization schemes have been tested, and we will comment on how they perform.

%For each recording, a polygon model tracking the left ventricle has been provided by the AutoLVQ tool from GE Vingmed Ultrasound. With this information available, estimation of $\mu_r$ and $\sigma_r$ can be restricted to blood pool samples only. For rays intersecting the model, the opacity threshold $\tau_r$ is therefore a weighted sum between first and second order blood pool statistics along the ray. Other parts of the volume can be rendered using a global OTF.

\subsection{Rendering setup}
All renderings have been produced with Phong lighting and the data have been pre-smoothed with anisotropic diffusion\cite{perona1990scale, Steen}. %As common for 3D echocardiography, volume renderings are depth encoded for enhanced depth perception. 
For ray-casting it is normal to stop integrating when no significant transparency is left (known as early-ray-termination). We have chosen to terminate each ray when only $5 \%$ transparency is left. The rendering resolution has been set to 512x512x100.

\subsection{Measuring rendering accuracy}
As a measure of rendering accuracy, we have calculated the mean absolute error (MAE) between the rendering depth buffer (given by early-ray-termination) and a validated left ventrical segmentation. The reference segmentation was performed by an experienced physician echocardiographer using the AutoLVQ tool from GE Vingmed Ultrasound. The MAE measure will increase if the rendering contains tissue dropouts, or if spurious structures occlude endocardial tissue. A minimal MAE therefore exists when a majority of rays end their traversal near the tissue-blood border depicted by the reference segmentation. A view-aligned clipping plane has been applied to produce standard long-axis and mitral valve views, clipping away half of both the ventricle and the reference segmentation. Clipping of the volume from behind has also been applied to reduce the impact of non-terminated rays versus rays terminated by clutter noise.

\section{Results}\label{sec:res}

\subsection{Quantitative evaluation}
% table showing statistics
\begin{table}
\begin{center}
\begin{tabular}{c|ccc}
& Apical 4-chamber & Inverted apical 4-chamber & Mitral valve \\ \hline \\[-1.5ex]
$\bar{d}$ & $1.5 \pm 0.3$ mm & $2.5 \pm 0.4$ mm & $1.7 \pm 0.2$ mm\\ 
\end{tabular}
\end{center}
\caption[]{\label{tab:stat1} Statistics showing a significant improvement ($p < 0.001$, $\textrm{d.f.} = 17$) in mean absolute error (MAE) for the proposed regional adaptive method in three views. The variable $\bar{d}$ is the average difference plus-minus standard error between $\textrm{MAE}_{global}$ and $\textrm{MAE}_{regional}$ in 18 subjects.
%The null hypothesis that $\bar{d}$ is less than or equal to zero, i.e. $\textrm{MAE}_{global} \le \textrm{MAE}_{regional}$, has probability $p < 0.001$ for all views using t-statistics with $\textrm{d.f.} = 17$. 
All renderings are from end-diastole and were tuned to the best visual appearance by a physician echocardiographer.}
\end{table}
The global and regional adaptive methods were compared using a dataset of apical recordings from 18 subjects. The dataset consisted of 8 healthy volunteers and 10 subjects with recent myocardial infarctions and were acquired using the GE Vivid 7 Dimensions system (GE Vingmed, Horten, Norway) and a 2.5MHz matrix array transducer (GE Vingmed, Horten, Norway), with the subjects in the left lateral decubitus position. Scans were taken from the apical imaging
window, in harmonic mode, from 4 up to 6 QRS triggered sub-volumes, during an end-expiratory breath-hold. The depth and angle of the ultrasound sector were adjusted such that the entire LV was covered.

For each recording, three renderings from standard views (apical 4-chamber (A4C), inverted A4C and mitral valve (MV)) were generated for both the global and regional adaptive methods. Each rendering was then tuned to the best visual appearance by a physician echocardiographer. For each rendering we measured the MAE between the rendering depth buffer and a validated LV segmentation. The results of these measurements are reported in Table \ref{tab:stat1}. The null hypothesis that $\bar{d}$ is less than or equal to zero, i.e. $\textrm{MAE}_{global} \le \textrm{MAE}_{regional}$, can be rejected ($p < 0.001$) for all views using t-statistics with $\textrm{d.f.} = 17$.

The algorithm has been implemented on a GPU using CUDA. It runs at an average speed of 16 fps with an Nvidia GTX 460 graphics card. Note that this card has half the texture performance in CUDA compared to Direct3D. To estimate statistics we need to add a second pass through the volume to the rendering pipeline, reducing performance from 26 fps for a global OTF to 16 fps. As mentioned the rendering resolution is 512x512x100.

\subsection{Qualitative evaluation}
\begin{table}
\begin{center}
\begin{tabular}{c|ccc}
& Apical 4-chamber & Inverted apical 4-chamber & Mitral valve \\ \hline \\[-1.5ex]
Better & 17 & 17 &  12 \\ 
Equal  & 1 & 1 & 4\\
Worse & 0 & 0 & 2\\
\end{tabular}
\end{center}
\caption[]{\label{tab:stat2} Ratings given  by an experienced physician echocardiographer when comparing the blood pool based regional adaptive OTF with a global OTF in three views.}
\end{table}
For each pair of renderings, an experienced physician echocardiographer was asked to answer the following questions regarding image quality: \emph{Does the blood pool based regional adaptive OTF produce a rendered image with less tissue dropouts and spurious structures inside the blood pool compared with the global OTF?}. This qualitative evaluation is summarized in Table \ref{tab:stat2}. As seen, he concluded that the regional adaptive method performed better in 46 of the 56 renderings, and only worse in two.

\subsection{Examples}
% one figure showing some renderings
\begin{figure}  
\begin{center}
\begin{tabular}{cc}
\includegraphics[height=0.4\textwidth]{global_otf_cn02_crown.eps} & 
\includegraphics[height=0.4\textwidth]{global_otf_crown.pdf}\\
(a) & (b)\\
\includegraphics[height=0.4\textwidth]{regional_otf_c05_crown.pdf} & 
\includegraphics[height=0.4\textwidth]{residualGraphVsC.eps}\\
(c) & (d)
\end{tabular}
\end{center}
\caption[example] { \label{fig:vis} a) Global OTF with a low opacity threshold $\tau$. A lot of noise is visible inside the LV (circle). b) Global OTF with $\tau$ corresponding to the lowest MAE in Figure d. The image contains less noise, but also a large tissue dropout (circle). c) Blood pool based regional adaptive OTF with c=0.5 which is close to the minimum in d. The papillary muscle is now more visible (arrow) and the tissue dropout (circle) is smaller without pruning the mitral valve (sqare).  d) Parameter exploration in the view shown in a, b and c. MAE is plotted versus the $c$-value. For comparison, the global OTF has been converted into a function of $c$ using the global mean and standard deviation.}
\end{figure}

Figures \ref{fig:vis}a, b and c show three renderings of the same apical 4-chamber view of one subject. With a low opacity threshold (Figure \ref{fig:vis}a), a lot of clutter noise is picked up in the apex of the LV (circle). With a threshold corresponding to the minimum MAE in Figure \ref{fig:vis}d, the clutter noise is removed (Figure \ref{fig:vis}b), but a tissue dropout (circle) has now increased in size. With the blood-pool based method (Figure \ref{fig:vis}c) clutter noise inside the LV is suppressed, giving a clear view of the endocardial tissue and a papillary muscle (arrow) without reducing the thickness of the mitral valve (square). The extent of the mentioned dropout is also reduced (circle). In Figure \ref{fig:vis}d we present a parameter exploration of the $c$-value introduced with the regional adaptive OTF. For comparison the global opacity threshold is substituted with a function of $c$, global mean and standard deviation. The figure shows that for a given view the MAE measure has a local minimum for both the global and regional methods, and that the regional OTF has a smaller MAE than the global OTF for all $c$.

% show spatial regularization

\section{Discussion}
% summarising contribution
A regional adaptive OTF based on blood-pool statistics has been proposed. The purpose was to better handle the high variation of signal within tissue as well as variability of noise signal inside the blood pool. Both quantitative and qualitative results reported in Section \ref{sec:res} shows that this has been accomplished.

% Discuss quantitative results
We believe that the significant reduction in MAE for the regional adaptive method in Table \ref{tab:stat1} is caused by a reduced amount of rendering outliers. By outliers we mean rendering rays that stop far away from the endocardial boundary depicted by the reference segmentation, thus having a high absolute error. We observe that the difference in MAE in millimetre is small, between 1.5 and 2.5 mm for the three views in Table \ref{tab:stat1}, compared to normal average end-diastole myocardial thickness (around 10mm). However, regional improvements in order of centimetres can be observed in Figure \ref{fig:vis}c (circle). These large improvements are smoothed out by the MAE measure, but remains as a significant statistical improvement ($p<0.001$) in MAE-difference between the regional and global adaptive method in all three views (Table \ref{tab:stat1}).

% Discuss qualitative results
Table \ref{tab:stat2} shows that in 46 pair of renderings the regional adaptive OTF based on blood pool statistics provided renderings with less tissue dropouts and spurious structures inside the blood pool. However, Table \ref{tab:stat2} also shows that the regional adaptive method was ranked worse in the MV view for two subjects, and in four MV-views the two methods were equal. On the other hand, the MV statistics in Table \ref{tab:stat1} are the most significant among the three views. One explanation to these diverging results is: When doing the qualitative assessment the physician echocardiographer watched the whole cardiac cycle, where the statistics in Table \ref{tab:stat1} are calculated from a single end-diastolic frame. In end-diastole the mitral valve is closed, which means that the MV is outside our tracking model and therefore not included in the blood pool statistics. When the MV is inside the tracking model, thus included in the estimates, visual artefacts can occur since the MV will be treated as being a part of the blood pool and might get rendered transparent. %It is important to prevent this, given the high clinical value that the MV represent. 
Another explanation is that all the recordings are from an apical imaging view. From this view the MV is normal to the ultrasound beam. The MV has because of this better image quality than e.g. the LV endocardium and thus less variation is observed in the MV-view statistics than for the long-axis views (Table \ref{tab:stat1}). When the image quality is good it is hard to make any regional improvement compared with a global OTF. Opposite, the endocardial boundary has usually lower image quality than the MV from the apical imaging window. For the two 4-chamber views the statistics therefore has higher variation and for the inverted 4-chamber, that had large acoustic dropouts in many of the recordings, the improvement is much larger.

% Discuss example renderings
The images in Figure \ref{fig:vis} demonstrate that our regional adaptation of the OTF provides a better tradoff between artificial tissue dropouts and spurious structures compared with a global OTF. In Figure \ref{fig:vis}d this improvement is quantified as $\textrm{MAE}_{global}(c) > \textrm{MAE}_{regional}(c), \forall c\in[-1.5, 1.5]$. This illustrate our findings that the global OTF did not produce a smaller MAE than the regional OTF in any pair of renderings if both methods where tuned to the lowest MAE. The gentle slope of $\textrm{MAE}_{regional}(c)$ in Figure \ref{fig:vis}d is another important observation. This indicates that the proposed method is less sensitive to changes in c, and therefore less tuning is often required. This observation was also done by the physician echocardiographer. 
The rendering in Figure \ref{fig:vis}c has a c-value corresponding to the minimum of $\textrm{MAE}_{regional}(c)$ in Figure \ref{fig:vis}d. We have often found the c-value at the local minimum of the MAE measure to be approximately the same c-value as both we and the physician echocradiographer would have selected in a given view. A fully automatic method could therefore iterate towards the local minimum by steepest descent, removing the need for manual tuning of $c$. It should be kept in mind that the success of this approach highly depends on a correct segmentation of the LV. This approach will also not be real-time, since an accurate segmentation of the LV usually involves manual adjustments and processing times in the order of seconds. One option is to use the real-time acquired LV delineation, used to estimate blood pool statistics, to measure rendering accuracy as well. A different endocardial detection scheme should then be used to estimate blood-pool statistics, to avoid a too close coupling between the LV delineation used for estimation and validation.

%Discuss spatial and temporal smoothing
As mentioned, no spatial regularization was applied to the results in Section \ref{sec:res}. However, spatial smoothing will decrease the variation in $\tau_r$ across the image, and with increased filter kernels the results will become equivalent to applying an OTF with a global adaptive opacity threshold $\tau_g$ based on the whole blood pool. With increased spatial smoothing less visual artefacts have been observed at the cost of a higher MAE. As no flickering has been observed, temporal smoothing had no positive effect. E.g. the rapid movement of the mitral valve caused artificial dropouts in all views when temporal smoothing was applied.

\subsection{Limitations}
The choice of using a deformable model to indicate the blood pool location imposes some limitations for the proposed method. As mentioned, some important cardiac structures are included in the depicted blood pool and might result in visual artefacts. However, how the LV endocardial boundary is derived is not relevant for the result in this paper. Local edge detectors combined with regularization may work equally well for limiting the regional estimation of means and standard deviations to the blood pool.

%Caused by the decision of using a deformable model to indicate the blood-pool location.

\section{Conclusion}
%Many adaptive algorithms for DVR have been proposed in the literature, however they are usually tested on data with considerably higher SNR than ultrasound. We have proposed a method for fast adaptive visualization of low-SNR cardiac 3D ultrasound. Compared with a global OTF we have shown that the regional adaptive method, based on blood pool statistics, is capable of providing a better tradeoff between tissue dropouts and the number of spurious structures inside the blood pool. The need for manual tuning is reduced by adapting the OTF to the data at hand, potentially leading to a fully automatic, data-dependent OTF.
In this paper we have introduced and investigated an adaptive DVR method for real-time visualization of cardiac 3D ultrasound. Compared to a global OTF we have shown, both with quantitative and qualitative results, that the regional adaptive method, based on blood pool statistics, is capable of reducing tissue dropouts and spurious structures inside the blood pool. The need for manual tuning is reduced by adapting the OTF to the data at hand, potentially leading to a fully automatic and data-dependent OTF.

### 2014-04-05 ThesisJP/Paper2/cos.tex
Added Paper2 and Paper3.

--- 2013-09-06 HRUIonGPU/jrnl_article.tex
@@
-%% bare_jrnl.tex
-%% V1.3
-%% 2007/01/11
-%% by Michael Shell
-%% see http://www.michaelshell.org/
-%% for current contact information.
-%%
-%% This is a skeleton file demonstrating the use of IEEEtran.cls
-%% (requires IEEEtran.cls version 1.7 or later) with an IEEE journal paper.
-%%
-%% Support sites:
-%% http://www.michaelshell.org/tex/ieeetran/
-%% http://www.ctan.org/tex-archive/macros/latex/contrib/IEEEtran/
-%% and
-%% http://www.ieee.org/
-
-% *** Authors should verify (and, if needed, correct) their LaTeX system  ***
-% *** with the testflow diagnostic prior to trusting their LaTeX platform ***
-% *** with production work. IEEE's font choices can trigger bugs that do  ***
-% *** not appear when using other class files.                            ***
-% The testflow support page is at:
-% http://www.michaelshell.org/tex/testflow/
-
-%%*************************************************************************
-%% Legal Notice:
-%% This code is offered as-is without any warranty either expressed or
-%% implied; without even the implied warranty of MERCHANTABILITY or
-%% FITNESS FOR A PARTICULAR PURPOSE! 
-%% User assumes all risk.
-%% In no event shall IEEE or any contributor to this code be liable for
-%% any damages or losses, including, but not limited to, incidental,
-%% consequential, or any other damages, resulting from the use or misuse
-%% of any information contained here.
-%%
-%% All comments are the opinions of their respective authors and are not
-%% necessarily endorsed by the IEEE.
-%%
-%% This work is distributed under the LaTeX Project Public License (LPPL)
-%% ( http://www.latex-project.org/ ) version 1.3, and may be freely used,
-%% distributed and modified. A copy of the LPPL, version 1.3, is included
-%% in the base LaTeX documentation of all distributions of LaTeX released
-%% 2003/12/01 or later.
-%% Retain all contribution notices and credits.
-%% ** Modified files should be clearly indicated as such, including  **
-%% ** renaming them and changing author support contact information. **
-%%
-%% File list of work: IEEEtran.cls, IEEEtran_HOWTO.pdf, bare_adv.tex,
-%%                    bare_conf.tex, bare_jrnl.tex, bare_jrnl_compsoc.tex
-%%*************************************************************************
-
-% Note that the a4paper option is mainly intended so that authors in
-% countries using A4 can easily print to A4 and see how their papers will
-% look in print - the typesetting of the document will not typically be
-% affected with changes in paper size (but the bottom and side margins will).
-% Use the testflow package mentioned above to verify correct handling of
-% both paper sizes by the user's LaTeX system.
-%
-% Also note that the "draftcls" or "draftclsnofoot", not "draft", option
-% should be used if it is desired that the figures are to be displayed in
-% draft mode.
-%
-
-\documentclass[journal]{IEEEtran}
-%\documentclass[11pt, draftcls, onecolumn]{IEEEtran}
-
-% Single column review mode:
-%\documentclass[12pt,journal,onecolumn]{IEEEtran}
-%\newcommand\CLASSINPUTbaselinestretch{2}
-%\RequirePackage{calc}
-%\RequirePackage{fp} 
-% Left margin = 1 inch + hoffset + oddsidemargin (or evensidemargin)
-% Adding the 2cm gutter width to the odd/even side margins:
-%\setlength\hoffset{0pt}
-%\setlength\oddsidemargin{4cm}       % 4cm margins on the left side
-%\addtolength\oddsidemargin{-1in}    % Subtract the initial 1 inch
-%\setlength\textwidth{21cm-8cm}    % A4 width (21cm) minus margins on either side
-
-%
-% If IEEEtran.cls has not been installed into the LaTeX system files,
-% manually specify the path to it like:
-% \documentclass[journal]{../sty/IEEEtran}
-
-% Some very useful LaTeX packages include:
-% (uncomment the ones you want to load)
-
-% *** MISC UTILITY PACKAGES ***
-%
-%\usepackage{ifpdf}
-% Heiko Oberdiek's ifpdf.sty is very useful if you need conditional
-% compilation based on whether the output is pdf or dvi.
-% usage:
-% \ifpdf
-%   % pdf code
-% \else
-%   % dvi code
-% \fi
-% The latest version of ifpdf.sty can be obtained from:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/oberdiek/
-% Also, note that IEEEtran.cls V1.7 and later provides a builtin
-% \ifCLASSINFOpdf conditional that works the same way.
-% When switching from latex to pdflatex and vice-versa, the compiler may
-% have to be run twice to clear warning/error messages.
-
-% *** CITATION PACKAGES ***
-%
-\usepackage{cite}
-% cite.sty was written by Donald Arseneau
-% V1.6 and later of IEEEtran pre-defines the format of the cite.sty package
-% \cite{} output to follow that of IEEE. Loading the cite package will
-% result in citation numbers being automatically sorted and properly
-% "compressed/ranged". e.g., [1], [9], [2], [7], [5], [6] without using
-% cite.sty will become [1], [2], [5]--[7], [9] using cite.sty. cite.sty's
-% \cite will automatically add leading space, if needed. Use cite.sty's
-% noadjust option (cite.sty V3.8 and later) if you want to turn this off.
-% cite.sty is already installed on most LaTeX systems. Be sure and use
-% version 4.0 (2003-05-27) and later if using hyperref.sty. cite.sty does
-% not currently provide for hyperlinked citations.
-% The latest version can be obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/cite/
-% The documentation is contained in the cite.sty file itself.
-
-% *** GRAPHICS RELATED PACKAGES ***
-%
-\ifCLASSINFOpdf
-  % \usepackage[pdftex]{graphicx}
-  % declare the path(s) where your graphic files are
-  % \graphicspath{{../pdf/}{../jpeg/}}
-  % and their extensions so you won't have to specify these with
-  % every instance of \includegraphics
-  % \DeclareGraphicsExtensions{.pdf,.jpeg,.png}
-\else
-  % or other class option (dvipsone, dvipdf, if not using dvips). graphicx
-  % will default to the driver specified in the system graphics.cfg if no
-  % driver is specified.
-  % \usepackage[dvips]{graphicx}
-  % declare the path(s) where your graphic files are
-  % \graphicspath{{../eps/}}
-  % and their extensions so you won't have to specify these with
-  % every instance of \includegraphics
-  % \DeclareGraphicsExtensions{.eps}
-\fi
-% graphicx was written by David Carlisle and Sebastian Rahtz. It is
-% required if you want graphics, photos, etc. graphicx.sty is already
-% installed on most LaTeX systems. The latest version and documentation can
-% be obtained at: 
-% http://www.ctan.org/tex-archive/macros/latex/required/graphics/
-% Another good source of documentation is "Using Imported Graphics in
-% LaTeX2e" by Keith Reckdahl which can be found as epslatex.ps or
-% epslatex.pdf at: http://www.ctan.org/tex-archive/info/
-%
-% latex, and pdflatex in dvi mode, support graphics in encapsulated
-% postscript (.eps) format. pdflatex in pdf mode supports graphics
-% in .pdf, .jpeg, .png and .mps (metapost) formats. Users should ensure
-% that all non-photo figures use a vector format (.eps, .pdf, .mps) and
-% not a bitmapped formats (.jpeg, .png). IEEE frowns on bitmapped formats
-% which can result in "jaggedy"/blurry rendering of lines and letters as
-% well as large increases in file sizes.
-%
-% You can find documentation about the pdfTeX application at:
-% http://www.tug.org/applications/pdftex
-
-% *** MATH PACKAGES ***
-%
-%\usepackage[cmex10]{amsmath}
-% A popular package from the American Mathematical Society that provides
-% many useful and powerful commands for dealing with mathematics. If using
-% it, be sure to load this package with the cmex10 option to ensure that
-% only type 1 fonts will utilized at all point sizes. Without this option,
-% it is possible that some math symbols, particularly those within
-% footnotes, will be rendered in bitmap form which will result in a
-% document that can not be IEEE Xplore compliant!
-%
-% Also, note that the amsmath package sets \interdisplaylinepenalty to 10000
-% thus preventing page breaks from occurring within multiline equations. Use:
-%\interdisplaylinepenalty=2500
-% after loading amsmath to restore such page breaks as IEEEtran.cls normally
-% does. amsmath.sty is already installed on most LaTeX systems. The latest
-% version and documentation can be obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/required/amslatex/math/
-
-% *** SPECIALIZED LIST PACKAGES ***
-%
-%\usepackage{algorithmic}
-% algorithmic.sty was written by Peter Williams and Rogerio Brito.
-% This package provides an algorithmic environment fo describing algorithms.
-% You can use the algorithmic environment in-text or within a figure
-% environment to provide for a floating algorithm. Do NOT use the algorithm
-% floating environment provided by algorithm.sty (by the same authors) or
-% algorithm2e.sty (by Christophe Fiorio) as IEEE does not use dedicated
-% algorithm float types and packages that provide these will not provide
-% correct IEEE style captions. The latest version and documentation of
-% algorithmic.sty can be obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/algorithms/
-% There is also a support site at:
-% http://algorithms.berlios.de/index.html
-% Also of interest may be the (relatively newer and more customizable)
-% algorithmicx.sty package by Szasz Janos:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/algorithmicx/
-
-% *** ALIGNMENT PACKAGES ***
-%
-%\usepackage{array}
-% Frank Mittelbach's and David Carlisle's array.sty patches and improves
-% the standard LaTeX2e array and tabular environments to provide better
-% appearance and additional user controls. As the default LaTeX2e table
-% generation code is lacking to the point of almost being broken with
-% respect to the quality of the end results, all users are strongly
-% advised to use an enhanced (at the very least that provided by array.sty)
-% set of table tools. array.sty is already installed on most systems. The
-% latest version and documentation can be obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/required/tools/
-
-%\usepackage{mdwmath}
-%\usepackage{mdwtab}
-% Also highly recommended is Mark Wooding's extremely powerful MDW tools,
-% especially mdwmath.sty and mdwtab.sty which are used to format equations
-% and tables, respectively. The MDWtools set is already installed on most
-% LaTeX systems. The lastest version and documentation is available at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/mdwtools/
-
-% IEEEtran contains the IEEEeqnarray family of commands that can be used to
-% generate multiline equations as well as matrices, tables, etc., of high
-% quality.
-
-%\usepackage{eqparbox}
-% Also of notable interest is Scott Pakin's eqparbox package for creating
-% (automatically sized) equal width boxes - aka "natural width parboxes".
-% Available at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/eqparbox/
-
-% *** SUBFIGURE PACKAGES ***
-%\usepackage[tight,footnotesize]{subfigure}
-% subfigure.sty was written by Steven Douglas Cochran. This package makes it
-% easy to put subfigures in your figures. e.g., "Figure 1a and 1b". For IEEE
-% work, it is a good idea to load it with the tight package option to reduce
-% the amount of white space around the subfigures. subfigure.sty is already
-% installed on most LaTeX systems. The latest version and documentation can
-% be obtained at:
-% http://www.ctan.org/tex-archive/obsolete/macros/latex/contrib/subfigure/
-% subfigure.sty has been superceeded by subfig.sty.
-
-%\usepackage[caption=false]{caption}
-%\usepackage[font=footnotesize]{subfig}
-% subfig.sty, also written by Steven Douglas Cochran, is the modern
-% replacement for subfigure.sty. However, subfig.sty requires and
-% automatically loads Axel Sommerfeldt's caption.sty which will override
-% IEEEtran.cls handling of captions and this will result in nonIEEE style
-% figure/table captions. To prevent this problem, be sure and preload
-% caption.sty with its "caption=false" package option. This is will preserve
-% IEEEtran.cls handing of captions. Version 1.3 (2005/06/28) and later 
-% (recommended due to many improvements over 1.2) of subfig.sty supports
-% the caption=false option directly:
-\usepackage[caption=false,font=footnotesize]{subfig}
-%
-% The latest version and documentation can be obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/subfig/
-% The latest version and documentation of caption.sty can be obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/caption/
-
-% *** FLOAT PACKAGES ***
-%
-%\usepackage{fixltx2e}
-% fixltx2e, the successor to the earlier fix2col.sty, was written by
-% Frank Mittelbach and David Carlisle. This package corrects a few problems
-% in the LaTeX2e kernel, the most notable of which is that in current
-% LaTeX2e releases, the ordering of single and double column floats is not
-% guaranteed to be preserved. Thus, an unpatched LaTeX2e can allow a
-% single column figure to be placed prior to an earlier double column
-% figure. The latest version and documentation can be found at:
-% http://www.ctan.org/tex-archive/macros/latex/base/
-
-%\usepackage{stfloats}
-% stfloats.sty was written by Sigitas Tolusis. This package gives LaTeX2e
-% the ability to do double column floats at the bottom of the page as well
-% as the top. (e.g., "\begin{figure*}[!b]" is not normally possible in
-% LaTeX2e). It also provides a command:
-%\fnbelowfloat
-% to enable the placement of footnotes below bottom floats (the standard
-% LaTeX2e kernel puts them above bottom floats). This is an invasive package
-% which rewrites many portions of the LaTeX2e float routines. It may not work
-% with other packages that modify the LaTeX2e float routines. The latest
-% version and documentation can be obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/sttools/
-% Documentation is contained in the stfloats.sty comments as well as in the
-% presfull.pdf file. Do not use the stfloats baselinefloat ability as IEEE
-% does not allow \baselineskip to stretch. Authors submitting work to the
-% IEEE should note that IEEE rarely uses double column equations and
-% that authors should try to avoid such use. Do not be tempted to use the
-% cuted.sty or midfloat.sty packages (also by Sigitas Tolusis) as IEEE does
-% not format its papers in such ways.
-
-%\ifCLASSOPTIONcaptionsoff
-%  \usepackage[nomarkers]{endfloat}
-% \let\MYoriglatexcaption\caption
-% \renewcommand{\caption}[2][\relax]{\MYoriglatexcaption[#2]{#2}}
-%\fi
-% endfloat.sty was written by James Darrell McCauley and Jeff Goldberg.
-% This package may be useful when used in conjunction with IEEEtran.cls'
-% captionsoff option. Some IEEE journals/societies require that submissions
-% have lists of figures/tables at the end of the paper and that
-% figures/tables without any captions are placed on a page by themselves at
-% the end of the document. If needed, the draftcls IEEEtran class option or
-% \CLASSINPUTbaselinestretch interface can be used to increase the line
-% spacing as well. Be sure and use the nomarkers option of endfloat to
-% prevent endfloat from "marking" where the figures would have been placed
-% in the text. The two hack lines of code above are a slight modification of
-% that suggested by in the endfloat docs (section 8.3.1) to ensure that
-% the full captions always appear in the list of figures/tables - even if
-% the user used the short optional argument of \caption[]{}.
-% IEEE papers do not typically make use of \caption[]'s optional argument,
-% so this should not be an issue. A similar trick can be used to disable
-% captions of packages such as subfig.sty that lack options to turn off
-% the subcaptions:
-% For subfig.sty:
-%\let\MYorigsubfloat\subfloat
-%\renewcommand{\subfloat}[2][\relax]{\MYorigsubfloat[]{#2}}
-% For subfigure.sty:
-% \let\MYorigsubfigure\subfigure
-% \renewcommand{\subfigure}[2][\relax]{\MYorigsubfigure[]{#2}}
-% However, the above trick will not work if both optional arguments of
-% the \subfloat/subfig command are used. Furthermore, there needs to be a
-% description of each subfigure *somewhere* and endfloat does not add
-% subfigure captions to its list of figures. Thus, the best approach is to
-% avoid the use of subfigure captions (many IEEE journals avoid them anyway)
-% and instead reference/explain all the subfigures within the main caption.
-% The latest version of endfloat.sty and its documentation can obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/endfloat/
-%
-% The IEEEtran \ifCLASSOPTIONcaptionsoff conditional can also be used
-% later in the document, say, to conditionally put the References on a 
-% page by themselves.
-
-% *** PDF, URL AND HYPERLINK PACKAGES ***
-%
-\usepackage{url}
-% url.sty was written by Donald Arseneau. It provides better support for
-% handling and breaking URLs. url.sty is already installed on most LaTeX
-% systems. The latest version can be obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/misc/
-% Read the url.sty source comments for usage information. Basically,
-% \url{my_url_here}.
-
-% *** Do not adjust lengths that control margins, column widths, etc. ***
-% *** Do not use packages that alter fonts (such as pslatex).         ***
-% There should be no need to do such things with IEEEtran.cls V1.6 and later.
-% (Unless specifically asked to do so by the journal or conference you plan
-% to submit to, of course. )
-
-% correct bad hyphenation here
-\hyphenation{op-tical net-works semi-conduc-tor}
-
-%%% My packages
-\usepackage{amsmath,epsfig}
-\usepackage{amsfonts}
-\usepackage{color}
-
 %%% My commands
-\newcommand{\mat}[1]{\mathbf{#1}}
+%\newcommand{\mat}[1]{\mathbf{#1}}
 \renewcommand{\vec}[1]{\mathbf{#1}}
-\newcommand{\degree}{\ensuremath{^\circ}}
-\newcommand{\R}{$\mat{\hat{R}}$ }
+%\newcommand{\degree}{\ensuremath{^\circ}}
+%\newcommand{\R}{$\mat{\hat{R}}$ }
 
@@
 
-\begin{document}
-%
-% paper title
-% can use linebreaks \\ within to get better formatting as desired
-\title{Implementing Capon Beamforming on a GPU for Real-Time Cardiac Ultrasound Imaging}
-%
-%
-% author names and IEEE memberships
-% note positions of commas and nonbreaking spaces ( ~ ) LaTeX will not break
-% a structure at a ~ so this keeps an author's name from being broken across
-% two lines.
-% use \thanks{} to gain access to the first footnote area
-% a separate \thanks must be used for each paragraph as LaTeX2e's \thanks
-% was not built to handle multiple paragraphs
-%
-
-\author{Jon Petter \AA{}sen, \IEEEmembership{Student Member,~IEEE,} Jo Inge Buskenes, Carl-Inge Colombo Nilsen, \IEEEmembership{Member,~IEEE,} Andreas Austeng, \IEEEmembership{Member,~IEEE,} and Sverre Holm, \IEEEmembership{Senior Member,~IEEE,} %<-this % stops a space
-\thanks{Manuscript received ...; accepted ...}
-\thanks{J. P. \AA{}sen and S. Holm is with the Medical Imaging Lab (MI-Lab) at the Norwegian University of Science and Technology, Trondheim, Norway.}
-\thanks{J. I. Buskenes, C.-I. C. Nilsen, A. Austeng and S. Holm is with the Department of Informatics, University of Oslo, Oslo, Norway. (e-mail: jon.p.asen@ntnu.no)} 
-}
-
-% note the % following the last \IEEEmembership and also \thanks - 
-% these prevent an unwanted space from occurring between the last author name
-% and the end of the author line. i.e., if you had this:
-% 
-% \author{....lastname \thanks{...} \thanks{...} }
-%                     ^------------^------------^----Do not want these spaces!
-%
-% a space would be appended to the last name and could cause every name on that
-% line to be shifted left slightly. This is one of those "LaTeX things". For
-% instance, "\textbf{A} \textbf{B}" will typeset as "A B" not "AB". To get
-% "AB" then you have to do: "\textbf{A}\textbf{B}"
-% \thanks is no different in this regard, so shield the last } of each \thanks
-% that ends a line with a % and do not let a space in before the next \thanks.
-% Spaces after \IEEEmembership other than the last one are OK (and needed) as
-% you are supposed to have spaces between the names. For what it is worth,
-% this is a minor point as most people would not even notice if the said evil
-% space somehow managed to creep in.
-
-% The paper headers
-\markboth{Transactions on Ultrasonics and Ferroelectrics and Frequency Control,~Vol.~x, No.~y, Z~201\ae{}}%
-{\AA{}sen \MakeLowercase{\textit{et al.}}: Capon Adaptive Beamforming on the GPU}
-% The only time the second header will appear is for the odd numbered pages
-% after the title page when using the twoside option.
-% 
-% *** Note that you probably will NOT want to include the author's ***
-% *** name in the headers of peer review papers.                   ***
-% You can use \ifCLASSOPTIONpeerreview for conditional compilation here if
-% you desire.
-
-% If you want to put a publisher's ID mark on the page you can do it like
-% this:
-%\IEEEpubid{0000--0000/00\$00.00~\copyright~2007 IEEE}
-% Remember, if you use this you must call \IEEEpubidadjcol in the second
-% column for its text to clear the IEEEpubid mark.
-
-% use for special paper notices
-%\IEEEspecialpapernotice{(Invited Paper)}
-
-% make the title area
-\maketitle
+\graphicspath{{./Paper2/}}
 
@@
 \end{abstract}
-% IEEEtran.cls defaults to using nonbold math in the Abstract.
-% This preserves the distinction between vectors and scalars. However,
-% if the journal you are submitting to favors bold math in the abstract,
-% then you can use LaTeX's standard command \boldmath at the very start
-% of the abstract to achieve this. Many IEEE journals frown on math
-% in the abstract anyway.
-
-% Note that keywords are not normally used for peerreview papers.
-\begin{IEEEkeywords}
-Adaptive beamforming, Capon beamforming, Covariance matrix, GPGPU, Graphics processing unit, Minimum variance beamforming, Parallel programming, Ultrasound imaging.
-\end{IEEEkeywords}
-
-% For peer review papers, you can put extra information on the cover
-% page as needed:
-% \ifCLASSOPTIONpeerreview
-% \begin{center} \bfseries EDICS Category: 3-BBND \end{center}
-% \fi
-%
-% For peerreview papers, this IEEEtran command inserts a page break and
-% creates the second title. It will be ignored for other modes.
-\IEEEpeerreviewmaketitle
 
 \section{Introduction}
-% The very first letter is a 2 line initial drop letter followed
-% by the rest of the first word in caps.
-% 
-% form to use if the first word consists of a single letter:
-% \IEEEPARstart{A}{demo} file is ....
-% 
-% form to use if you need the single drop letter followed by
-% normal text (unknown if ever used by IEEE):
-% \IEEEPARstart{A}{}demo file is ....
-% 
-% Some journals put the first two words in caps:
-% \IEEEPARstart{T}{his demo} file is ....
-% 
-% Here we have the typical use of a "T" for an initial drop letter
-% and "HIS" in caps to complete the first word.
-
-%\IEEEPARstart{I}{ntroduction} to Capon adaptive beamforming, ultrasound imaging and GPU/CUDA.
-\IEEEPARstart{C}{apon} beamforming \cite{Capon1969} has recently been applied to medical ultrasound imaging with promising results in several studies \cite{Synnevag2007, Austeng2008, Vignon2008, Viola, Mehdizadeh2012}.
+
+Capon beamforming \cite{Capon1969} has recently been applied to medical ultrasound imaging with promising results in several studies \cite{Synnevag2007, Austeng2008, Vignon2008, Viola, Mehdizadeh2012}.
 Most of this work focuses on the improved lateral resolution and contrast obtained with the Capon beamformer.
@@
 \begin{figure}
-\centerline{\includegraphics[width=3in]{gfx/beamforming_mv.eps}}
+\centerline{\includegraphics[width=0.8\textwidth]{gfx/beamforming_mv.eps}}
 \caption{Capon beamforming.
@@
 \begin{figure}
-\centerline{\includegraphics[width=3.5in]{gfx/gpu_layout_vertical_2.eps}}
+\centerline{\includegraphics[width=0.8\textwidth]{gfx/gpu_layout_vertical_2.eps}}
 \caption{Calculation of Capon adaptive weights mapped to GPU architecture.
@@
 \begin{figure}
-\centerline{\includegraphics[width=3.2in]{gfx/gpu_layout_bs_2.eps}}
+\centerline{\includegraphics[width=0.8\textwidth]{gfx/gpu_layout_bs_2.eps}}
 \caption{Calculation of the sliding beamspace transform.
@@
 \centerline{
-\subfloat[]{\includegraphics[width=3.4in]{gfx/benchmark_bar_64_80_540.png} \label{fig:benchUS64}}
+\subfloat[]{\includegraphics[width=0.5\textwidth]{gfx/benchmark_bar_64_80_540.png} \label{fig:benchUS64}}\\
 \hfil
-\subfloat[]{\includegraphics[width=3.4in]{gfx/benchmark_bar_96_120_540.png}%
+\subfloat[]{\includegraphics[width=0.5\textwidth]{gfx/benchmark_bar_96_120_540.png}%
 \label{fig:benchUS96}}}
@@
 \begin{figure*}[!t]
-\centerline{\subfloat[]{\includegraphics[width=3.4in]{gfx/benchmark_bar_bs_M=64_Nx=80_Ny=540_Nb=3.png} \label{fig:benchBS64}}
+\centerline{\subfloat[]{\includegraphics[width=0.5\textwidth]{gfx/benchmark_bar_bs_M=64_Nx=80_Ny=540_Nb=3.png} \label{fig:benchBS64}}
 \hfil
-\subfloat[]{\includegraphics[width=3.4in]{gfx/benchmark_bar_bs_M=96_Nx=120_Ny=540_Nb=3.png}%
+\subfloat[]{\includegraphics[width=0.5\textwidth]{gfx/benchmark_bar_bs_M=96_Nx=120_Ny=540_Nb=3.png}%
 \label{fig:benchBS96}}}
@@
 \centerline{
-\subfloat[]{\includegraphics[width=2.2in]{gfx/das_invivo.png} \label{fig:invivoDAS}}
+\subfloat[]{\includegraphics[width=0.3\textwidth]{gfx/das_invivo.png} \label{fig:invivoDAS}}
 \hfil
-\subfloat[]{\includegraphics[width=2.2in]{gfx/capon_invivo.png} \label{fig:invivoMV}}
+\subfloat[]{\includegraphics[width=0.3\textwidth]{gfx/capon_invivo.png} \label{fig:invivoMV}}
 \hfil
 \subfloat[]{
-\includegraphics[width=2.2in]{gfx/capon_bs_invivo.png} \label{fig:invivoBS}}
+\includegraphics[width=0.3\textwidth]{gfx/capon_bs_invivo.png} \label{fig:invivoBS}}
 \hfil
-\includegraphics[width=0.4in]{gfx/colorBarInvivo2.png}
+\includegraphics[height=0.3\textwidth]{gfx/colorBarInvivo2.png}
 }
@@
 \centering
-\includegraphics[width=3.3in]{gfx/speed_res_trade.png}
+\includegraphics[width=0.8\textwidth]{gfx/speed_res_trade.png}
 \caption{Trade-off between resolution and speed by reducing subarray size or reducing beamspace dimension.
@@
 
-%\section{Future work}
-%Study shift invariance and minimum sampling...
-
-%MLA's has, to some extent been used to increase later sampling.
-When using a high resolution beamformer, Nyquist tells us that the sampling frequency must be increase.
-Future work will concentrate on how much over sampling is required when using the Capon beamformer whit i curtain set of parameters...
-
-%Plane waves...
-
-% An example of a floating figure using the graphicx package.
-% Note that \label must occur AFTER (or within) \caption.
-% For figures, \caption should occur after the \includegraphics.
-% Note that IEEEtran v1.7 and later has special internal code that
-% is designed to preserve the operation of \label within \caption
-% even when the captionsoff option is in effect. However, because
-% of issues like this, it may be the safest practice to put all your
-% \label just after \caption rather than within \caption{}.
-%
-% Reminder: the "draftcls" or "draftclsnofoot", not "draft", class
-% option should be used if it is desired that the figures are to be
-% displayed while in draft mode.
-%
-%\begin{figure}[!t]
-%\centering
-%\includegraphics[width=2.5in]{myfigure}
-% where an .eps filename suffix will be assumed under latex, 
-% and a .pdf suffix will be assumed for pdflatex; or what has been declared
-% via \DeclareGraphicsExtensions.
-%\caption{Simulation Results}
-%\label{fig_sim}
-%\end{figure}
-
-% Note that IEEE typically puts floats only at the top, even when this
-% results in a large percentage of a column being occupied by floats.
-
-% An example of a double column floating figure using two subfigures.
-% (The subfig.sty package must be loaded for this to work.)
-% The subfigure \label commands are set within each subfloat command, the
-% \label for the overall figure must come after \caption.
-% \hfil must be used as a separator to get equal spacing.
-% The subfigure.sty package works much the same way, except \subfigure is
-% used instead of \subfloat.
-%
-%\begin{figure*}[!t]
-%\centerline{\subfloat[Case I]\includegraphics[width=2.5in]{subfigcase1}%
-%\label{fig_first_case}}
-%\hfil
-%\subfloat[Case II]{\includegraphics[width=2.5in]{subfigcase2}%
-%\label{fig_second_case}}}
-%\caption{Simulation results}
-%\label{fig_sim}
-%\end{figure*}
-%
-% Note that often IEEE papers with subfigures do not employ subfigure
-% captions (using the optional argument to \subfloat), but instead will
-% reference/describe all of them (a), (b), etc., within the main caption.
-
-% An example of a floating table. Note that, for IEEE style tables, the 
-% \caption command should come BEFORE the table. Table text will default to
-% \footnotesize as IEEE normally uses this smaller font for tables.
-% The \label must come after \caption as always.
-%
-%\begin{table}[!t]
-%% increase table row spacing, adjust to taste
-%\renewcommand{\arraystretch}{1.3}
-% if using array.sty, it might be a good idea to tweak the value of
-% \extrarowheight as needed to properly center the text within the cells
-%\caption{An Example of a Table}
-%\label{table_example}
-%\centering
-%% Some packages, such as MDW tools, offer better commands for making tables
-%% than the plain LaTeX2e tabular which is used here.
-%\begin{tabular}{|c||c|}
-%\hline
-%One & Two\\
-%\hline
-%Three & Four\\
-%\hline
-%\end{tabular}
-%\end{table}
-
-% Note that IEEE does not put floats in the very first column - or typically
-% anywhere on the first page for that matter. Also, in-text middle ("here")
-% positioning is not used. Most IEEE journals use top floats exclusively.
-% Note that, LaTeX2e, unlike IEEE journals, places footnotes above bottom
-% floats. This can be corrected via the \fnbelowfloat command of the
-% stfloats package.
-
-%\section{Conclusion}
-%The conclusion goes here.
-
-% if have a single appendix:
-%\appendix[Proof of the Zonklar Equations]
-% or
-%\appendix  % for no appendix heading
-% do not use \section anymore after \appendix, only \section*
-% is possibly needed
-
-% use appendices with more than one appendix
-% then use \section to start each appendix
-% you must declare a \section before using any
-% \subsection or using \label (\appendices by itself
-% starts a section numbered zero.)
-%
-
-%\appendices
-%\section{Proof of the First Zonklar Equation}
-%Appendix one text goes here.
-
-% you can choose not to have a title for an appendix
-% if you want by leaving the argument blank
-%\section{}
-%Appendix two text goes here.
-
-% use section* for acknowledgement
 \section*{Acknowledgment}
 The authors would like to thank Prof. K. Kristoffersen, GE Vingmed Ultrasound and Norwegian University of Science and Technology, for valuable comments.
-
-% Can use something like this to put references on a page
-% by themselves when using endfloat and the captionsoff option.
-\ifCLASSOPTIONcaptionsoff
-  \newpage
-\fi
-
-% trigger a \newpage just before the given reference
-% number - used to balance the columns on the last page
-% adjust value as needed - may need to be readjusted if
-% the document is modified later
-%\IEEEtriggeratref{8}
-% The "triggered" command can be changed if desired:
-%\IEEEtriggercmd{\enlargethispage{-5in}}
-
-% references section
-
-% can use a bibliography generated by BibTeX as a .bbl file
-% BibTeX documentation can be easily obtained at:
-% http://www.ctan.org/tex-archive/biblio/bibtex/contrib/doc/
-% The IEEEtran BibTeX style support page is at:
-% http://www.michaelshell.org/tex/ieeetran/bibtex/
-\bibliographystyle{IEEEtran}
-% argument is your BibTeX string definitions and bibliography database(s)
-%\bibliography{IEEEabrv,../bib/paper}
-\bibliography{mybib} % link to mendeley organized bibtex-file
-%\bibliography{IEEEabrv,mybib} 
-% <OR> manually copy in the resultant .bbl file
-% set second argument of \begin to the number of references
-% (used to reserve space for the reference number labels box)
-%\begin{thebibliography}{1}
-
-%\bibitem{IEEEhowto:kopka}
-%H.~Kopka and P.~W. Daly, \emph{A Guide to \LaTeX}, 3rd~ed.\hskip 1em plus
-%  0.5em minus 0.4em\relax Harlow, England: Addison-Wesley, 1999.
-
-%\end{thebibliography}
-
-% biography section
-% 
-% If you have an EPS/PDF photo (graphicx package needed) extra braces are
-% needed around the contents of the optional argument to biography to prevent
-% the LaTeX parser from getting confused when it sees the complicated
-% \includegraphics command within an optional argument. (You could create
-% your own custom macro containing the \includegraphics command to make things
-% simpler here.)
-%\begin{biography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{mshell}}]{Michael Shell}
-% or if you just want to reserve a space for a photo:
-
-% insert where needed to balance the two columns on the last page with
-% biographies
-%\newpage
-
-\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{photos/jon_petter2.eps}}]{Jon Petter \AA{}sen}
-(S12) was born in Porsgrunn, Norway in 1986.
-He received the B.Sc.
-and M.Sc.
-degree in computer science from the University of Oslo, Norway, in 2008 and 2010.
-He is currently pursuing his Ph.D.
-degree in medical ultrasound technology at the Norwegian University of Science and Technology (NTNU) Medical Imaging Lab (MI-Lab), Trondheim, Norway.
-
-He has industry experience from GE Vingmed Ultrasound, Horten, Norway, were he completed three summer internships from 2008 to 2010.
-His research interests include image and signal processing, adaptive ultrasound processing techniques and acceleration of ultrasound algorithms using Graphics Processing Units (GPUs).
-\end{IEEEbiography}
-
-\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{photos/jo_inge2.eps}}]{Jo Inge Buskenes}
-received the B.Sc.
-degree in electrical engineering from Gj\o{}vik College University, Norway, in 2007, and the M.Sc.
-degree in instrumentation for particle physics from the University of Oslo, Norway, in 2010.
-He is currently pursuing the Ph.D.
-degree in image reconstruction and technology at the University of Oslo.
-
-His industry experience includes the European Organization for Nuclear Research (CERN), Geneva, Switzerland (2007-2008), and the Norwegian Defence Research Establishment, Kjeller, Norway (2009).
-He has lectured in digital signal processing at the Gj\o{}vik College University (2009), and at the University of Oslo (2010-2012).
-His research interests include adaptive beamforming, digital image reconstruction, high performance computing, intelligent detector design and embedded Linux platforms.
-\end{IEEEbiography}
-
-\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{photos/carl-inge.eps}}]{Carl-Inge Colombo Nilsen}
-(S’06–M’10) received the M.Sc.
-and Ph.D.
-degrees in computer science from the University of Oslo, Norway, in 2005 and 2010.
-He is currently working at the University of Oslo as a postdoctoral research fellow.
-His research interests include signal and array processing for ultrasound imaging and other acoustic applications.
-\end{IEEEbiography}
-
-% insert where needed to balance the two columns on the last page with
-% biographies
-%\newpage
-
-\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{photos/andreas.eps}}]{Andreas Austeng}
-was born in Oslo, Norway, in 1970.
-He received the M.S.
-degree in physics in 1996 and the Ph.D.
-degree in computer science in 2001, both from the University of Oslo.
-Since 2001, he has been working at the Department of Informatics, University of Oslo, first as a postdoctoral research fellow and currently as an associate professor.
-His research interests include signal and array processing for acoustical imaging.
-\end{IEEEbiography}
-
-%\newpage
-
-\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{photos/sverre.eps}}]{Sverre Holm}
-(M’82-SM’02) was born in Oslo, Norway, in 1954.
-He received the M.S.
-and Ph.D.
-degrees in electrical engineering from the Norwegian Institute of Technology, Trondheim, in 1978 and 1982, respectively.
-His academic experience includes Yarmouk University in Jordan (1984-1986) and the Norwegian Institute of Technology and later the University of Oslo (1989-1994) as an adjunct professor.
-Since 1995, he has been a professor of signal processing in the Department of Informatics at the University of Oslo.
-In 2002, he was elected a member of the Norwegian Academy of Technological Sciences, and in 2009, he became an adjunct professor in the Norwegian University of Science and Technology (NTNU) Medical Imaging Lab, Trondheim.
-His industry experience includes GE Vingmed Ultrasound, Horten, Norway (1990-1994).
-He worked with Sonitor Technologies (2000-2005), where he developed ultrasonic technology that is the core of the indoor positioning system described in \textit{Scientific Americanin} 2008 as "a positioning system that goes where GPS can't." He has published about 130 publications internationally, plus about 10 patents.
-He has had sabbaticals at GE Global Research, NY (1998), and the Waves and
-Acoustics Laboratory of ESPCI, Paris (2008-2009).
-He was an associate editor of the \textit{IEEE Transactions on Ultrasonics, Ferroelectrics, and Frequency Control} from 1997 to 2002.
-His research interests are medical ultrasound imaging, modeling of tissue-ultrasound interaction, and ultrasound positioning.
-\end{IEEEbiography}
-
-% You can push biographies down or up by placing
-% a \vfill before or after them. The appropriate
-% use of \vfill depends on what kind of text is
-% on the last page and whether or not the columns
-% are being equalized.
-
-%\vfill
-
-% Can be used to pull up biographies so that the bottom of the last one
-% is flush with the other column.
-%\enlargethispage{-5in}
-
-% that's all folks
-\end{document}

### 2014-04-05 ThesisJP/Paper3/moving_obj.tex
Added Paper2 and Paper3.

--- 2014-03-05 HRUIofMovingObjects/jrnl_article.tex
@@
-%% bare_jrnl.tex
-%% V1.3
-%% 2007/01/11
-%% by Michael Shell
-%% see http://www.michaelshell.org/
-%% for current contact information.
-%%
-%% This is a skeleton file demonstrating the use of IEEEtran.cls
-%% (requires IEEEtran.cls version 1.7 or later) with an IEEE journal paper.
-%%
-%% Support sites:
-%% http://www.michaelshell.org/tex/ieeetran/
-%% http://www.ctan.org/tex-archive/macros/latex/contrib/IEEEtran/
-%% and
-%% http://www.ieee.org/
+%%% My commands
+%\renewcommand{\vec}[1]{\mathbf{#1}}
 
-% *** Authors should verify (and, if needed, correct) their LaTeX system  ***
-% *** with the testflow diagnostic prior to trusting their LaTeX platform ***
-% *** with production work. IEEE's font choices can trigger bugs that do  ***
-% *** not appear when using other class files.                            ***
-% The testflow support page is at:
-% http://www.michaelshell.org/tex/testflow/
-
-%%*************************************************************************
-%% Legal Notice:
-%% This code is offered as-is without any warranty either expressed or
-%% implied; without even the implied warranty of MERCHANTABILITY or
-%% FITNESS FOR A PARTICULAR PURPOSE! 
-%% User assumes all risk.
-%% In no event shall IEEE or any contributor to this code be liable for
-%% any damages or losses, including, but not limited to, incidental,
-%% consequential, or any other damages, resulting from the use or misuse
-%% of any information contained here.
-%%
-%% All comments are the opinions of their respective authors and are not
-%% necessarily endorsed by the IEEE.
-%%
-%% This work is distributed under the LaTeX Project Public License (LPPL)
-%% ( http://www.latex-project.org/ ) version 1.3, and may be freely used,
-%% distributed and modified. A copy of the LPPL, version 1.3, is included
-%% in the base LaTeX documentation of all distributions of LaTeX released
-%% 2003/12/01 or later.
-%% Retain all contribution notices and credits.
-%% ** Modified files should be clearly indicated as such, including  **
-%% ** renaming them and changing author support contact information. **
-%%
-%% File list of work: IEEEtran.cls, IEEEtran_HOWTO.pdf, bare_adv.tex,
-%%                    bare_conf.tex, bare_jrnl.tex, bare_jrnl_compsoc.tex
-%%*************************************************************************
-
-% Note that the a4paper option is mainly intended so that authors in
-% countries using A4 can easily print to A4 and see how their papers will
-% look in print - the typesetting of the document will not typically be
-% affected with changes in paper size (but the bottom and side margins will).
-% Use the testflow package mentioned above to verify correct handling of
-% both paper sizes by the user's LaTeX system.
-%
-% Also note that the "draftcls" or "draftclsnofoot", not "draft", option
-% should be used if it is desired that the figures are to be displayed in
-% draft mode.
-%
-\documentclass[draftcls]{IEEEtran}
-%\documentclass[journal]{IEEEtran}
-%
-% If IEEEtran.cls has not been installed into the LaTeX system files,
-% manually specify the path to it like:
-% \documentclass[journal]{../sty/IEEEtran}
-
-% Some very useful LaTeX packages include:
-% (uncomment the ones you want to load)
-
-% *** MISC UTILITY PACKAGES ***
-%
-%\usepackage{ifpdf}
-% Heiko Oberdiek's ifpdf.sty is very useful if you need conditional
-% compilation based on whether the output is pdf or dvi.
-% usage:
-% \ifpdf
-%   % pdf code
-% \else
-%   % dvi code
-% \fi
-% The latest version of ifpdf.sty can be obtained from:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/oberdiek/
-% Also, note that IEEEtran.cls V1.7 and later provides a builtin
-% \ifCLASSINFOpdf conditional that works the same way.
-% When switching from latex to pdflatex and vice-versa, the compiler may
-% have to be run twice to clear warning/error messages.
-
-% *** CITATION PACKAGES ***
-%
-\usepackage{cite}
-% cite.sty was written by Donald Arseneau
-% V1.6 and later of IEEEtran pre-defines the format of the cite.sty package
-% \cite{} output to follow that of IEEE. Loading the cite package will
-% result in citation numbers being automatically sorted and properly
-% "compressed/ranged". e.g., [1], [9], [2], [7], [5], [6] without using
-% cite.sty will become [1], [2], [5]--[7], [9] using cite.sty. cite.sty's
-% \cite will automatically add leading space, if needed. Use cite.sty's
-% noadjust option (cite.sty V3.8 and later) if you want to turn this off.
-% cite.sty is already installed on most LaTeX systems. Be sure and use
-% version 4.0 (2003-05-27) and later if using hyperref.sty. cite.sty does
-% not currently provide for hyperlinked citations.
-% The latest version can be obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/cite/
-% The documentation is contained in the cite.sty file itself.
-
-% *** GRAPHICS RELATED PACKAGES ***
-%
-\ifCLASSINFOpdf
-  % \usepackage[pdftex]{graphicx}
-  % declare the path(s) where your graphic files are
-  % \graphicspath{{../pdf/}{../jpeg/}}
-  % and their extensions so you won't have to specify these with
-  % every instance of \includegraphics
-  % \DeclareGraphicsExtensions{.pdf,.jpeg,.png}
-\else
-  % or other class option (dvipsone, dvipdf, if not using dvips). graphicx
-  % will default to the driver specified in the system graphics.cfg if no
-  % driver is specified.
-  % \usepackage[dvips]{graphicx}
-  % declare the path(s) where your graphic files are
-  % \graphicspath{{../eps/}}
-  % and their extensions so you won't have to specify these with
-  % every instance of \includegraphics
-  % \DeclareGraphicsExtensions{.eps}
-\fi
-% graphicx was written by David Carlisle and Sebastian Rahtz. It is
-% required if you want graphics, photos, etc. graphicx.sty is already
-% installed on most LaTeX systems. The latest version and documentation can
-% be obtained at: 
-% http://www.ctan.org/tex-archive/macros/latex/required/graphics/
-% Another good source of documentation is "Using Imported Graphics in
-% LaTeX2e" by Keith Reckdahl which can be found as epslatex.ps or
-% epslatex.pdf at: http://www.ctan.org/tex-archive/info/
-%
-% latex, and pdflatex in dvi mode, support graphics in encapsulated
-% postscript (.eps) format. pdflatex in pdf mode supports graphics
-% in .pdf, .jpeg, .png and .mps (metapost) formats. Users should ensure
-% that all non-photo figures use a vector format (.eps, .pdf, .mps) and
-% not a bitmapped formats (.jpeg, .png). IEEE frowns on bitmapped formats
-% which can result in "jaggedy"/blurry rendering of lines and letters as
-% well as large increases in file sizes.
-%
-% You can find documentation about the pdfTeX application at:
-% http://www.tug.org/applications/pdftex
-
-% *** MATH PACKAGES ***
-%
-%\usepackage[cmex10]{amsmath}
-% A popular package from the American Mathematical Society that provides
-% many useful and powerful commands for dealing with mathematics. If using
-% it, be sure to load this package with the cmex10 option to ensure that
-% only type 1 fonts will utilized at all point sizes. Without this option,
-% it is possible that some math symbols, particularly those within
-% footnotes, will be rendered in bitmap form which will result in a
-% document that can not be IEEE Xplore compliant!
-%
-% Also, note that the amsmath package sets \interdisplaylinepenalty to 10000
-% thus preventing page breaks from occurring within multiline equations. Use:
-%\interdisplaylinepenalty=2500
-% after loading amsmath to restore such page breaks as IEEEtran.cls normally
-% does. amsmath.sty is already installed on most LaTeX systems. The latest
-% version and documentation can be obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/required/amslatex/math/
-
-% *** SPECIALIZED LIST PACKAGES ***
-%
-%\usepackage{algorithmic}
-% algorithmic.sty was written by Peter Williams and Rogerio Brito.
-% This package provides an algorithmic environment fo describing algorithms.
-% You can use the algorithmic environment in-text or within a figure
-% environment to provide for a floating algorithm. Do NOT use the algorithm
-% floating environment provided by algorithm.sty (by the same authors) or
-% algorithm2e.sty (by Christophe Fiorio) as IEEE does not use dedicated
-% algorithm float types and packages that provide these will not provide
-% correct IEEE style captions. The latest version and documentation of
-% algorithmic.sty can be obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/algorithms/
-% There is also a support site at:
-% http://algorithms.berlios.de/index.html
-% Also of interest may be the (relatively newer and more customizable)
-% algorithmicx.sty package by Szasz Janos:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/algorithmicx/
-
-% *** ALIGNMENT PACKAGES ***
-%
-%\usepackage{array}
-% Frank Mittelbach's and David Carlisle's array.sty patches and improves
-% the standard LaTeX2e array and tabular environments to provide better
-% appearance and additional user controls. As the default LaTeX2e table
-% generation code is lacking to the point of almost being broken with
-% respect to the quality of the end results, all users are strongly
-% advised to use an enhanced (at the very least that provided by array.sty)
-% set of table tools. array.sty is already installed on most systems. The
-% latest version and documentation can be obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/required/tools/
-
-%\usepackage{mdwmath}
-%\usepackage{mdwtab}
-% Also highly recommended is Mark Wooding's extremely powerful MDW tools,
-% especially mdwmath.sty and mdwtab.sty which are used to format equations
-% and tables, respectively. The MDWtools set is already installed on most
-% LaTeX systems. The lastest version and documentation is available at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/mdwtools/
-
-% IEEEtran contains the IEEEeqnarray family of commands that can be used to
-% generate multiline equations as well as matrices, tables, etc., of high
-% quality.
-
-%\usepackage{eqparbox}
-% Also of notable interest is Scott Pakin's eqparbox package for creating
-% (automatically sized) equal width boxes - aka "natural width parboxes".
-% Available at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/eqparbox/
-
-% *** SUBFIGURE PACKAGES ***
-%\usepackage[tight,footnotesize]{subfigure}
-% subfigure.sty was written by Steven Douglas Cochran. This package makes it
-% easy to put subfigures in your figures. e.g., "Figure 1a and 1b". For IEEE
-% work, it is a good idea to load it with the tight package option to reduce
-% the amount of white space around the subfigures. subfigure.sty is already
-% installed on most LaTeX systems. The latest version and documentation can
-% be obtained at:
-% http://www.ctan.org/tex-archive/obsolete/macros/latex/contrib/subfigure/
-% subfigure.sty has been superceeded by subfig.sty.
-
-%\usepackage[caption=false]{caption}
-%\usepackage[font=footnotesize]{subfig}
-% subfig.sty, also written by Steven Douglas Cochran, is the modern
-% replacement for subfigure.sty. However, subfig.sty requires and
-% automatically loads Axel Sommerfeldt's caption.sty which will override
-% IEEEtran.cls handling of captions and this will result in nonIEEE style
-% figure/table captions. To prevent this problem, be sure and preload
-% caption.sty with its "caption=false" package option. This is will preserve
-% IEEEtran.cls handing of captions. Version 1.3 (2005/06/28) and later 
-% (recommended due to many improvements over 1.2) of subfig.sty supports
-% the caption=false option directly:
-\usepackage[caption=false,font=footnotesize]{subfig}
-%
-% The latest version and documentation can be obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/subfig/
-% The latest version and documentation of caption.sty can be obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/caption/
-
-% *** FLOAT PACKAGES ***
-%
-%\usepackage{fixltx2e}
-% fixltx2e, the successor to the earlier fix2col.sty, was written by
-% Frank Mittelbach and David Carlisle. This package corrects a few problems
-% in the LaTeX2e kernel, the most notable of which is that in current
-% LaTeX2e releases, the ordering of single and double column floats is not
-% guaranteed to be preserved. Thus, an unpatched LaTeX2e can allow a
-% single column figure to be placed prior to an earlier double column
-% figure. The latest version and documentation can be found at:
-% http://www.ctan.org/tex-archive/macros/latex/base/
-
-%\usepackage{stfloats}
-% stfloats.sty was written by Sigitas Tolusis. This package gives LaTeX2e
-% the ability to do double column floats at the bottom of the page as well
-% as the top. (e.g., "\begin{figure*}[!b]" is not normally possible in
-% LaTeX2e). It also provides a command:
-%\fnbelowfloat
-% to enable the placement of footnotes below bottom floats (the standard
-% LaTeX2e kernel puts them above bottom floats). This is an invasive package
-% which rewrites many portions of the LaTeX2e float routines. It may not work
-% with other packages that modify the LaTeX2e float routines. The latest
-% version and documentation can be obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/sttools/
-% Documentation is contained in the stfloats.sty comments as well as in the
-% presfull.pdf file. Do not use the stfloats baselinefloat ability as IEEE
-% does not allow \baselineskip to stretch. Authors submitting work to the
-% IEEE should note that IEEE rarely uses double column equations and
-% that authors should try to avoid such use. Do not be tempted to use the
-% cuted.sty or midfloat.sty packages (also by Sigitas Tolusis) as IEEE does
-% not format its papers in such ways.
-
-%\ifCLASSOPTIONcaptionsoff
-%  \usepackage[nomarkers]{endfloat}
-% \let\MYoriglatexcaption\caption
-% \renewcommand{\caption}[2][\relax]{\MYoriglatexcaption[#2]{#2}}
-%\fi
-% endfloat.sty was written by James Darrell McCauley and Jeff Goldberg.
-% This package may be useful when used in conjunction with IEEEtran.cls'
-% captionsoff option. Some IEEE journals/societies require that submissions
-% have lists of figures/tables at the end of the paper and that
-% figures/tables without any captions are placed on a page by themselves at
-% the end of the document. If needed, the draftcls IEEEtran class option or
-% \CLASSINPUTbaselinestretch interface can be used to increase the line
-% spacing as well. Be sure and use the nomarkers option of endfloat to
-% prevent endfloat from "marking" where the figures would have been placed
-% in the text. The two hack lines of code above are a slight modification of
-% that suggested by in the endfloat docs (section 8.3.1) to ensure that
-% the full captions always appear in the list of figures/tables - even if
-% the user used the short optional argument of \caption[]{}.
-% IEEE papers do not typically make use of \caption[]'s optional argument,
-% so this should not be an issue. A similar trick can be used to disable
-% captions of packages such as subfig.sty that lack options to turn off
-% the subcaptions:
-% For subfig.sty:
-% \let\MYorigsubfloat\subfloat
-% \renewcommand{\subfloat}[2][\relax]{\MYorigsubfloat[]{#2}}
-% For subfigure.sty:
-% \let\MYorigsubfigure\subfigure
-% \renewcommand{\subfigure}[2][\relax]{\MYorigsubfigure[]{#2}}
-% However, the above trick will not work if both optional arguments of
-% the \subfloat/subfig command are used. Furthermore, there needs to be a
-% description of each subfigure *somewhere* and endfloat does not add
-% subfigure captions to its list of figures. Thus, the best approach is to
-% avoid the use of subfigure captions (many IEEE journals avoid them anyway)
-% and instead reference/explain all the subfigures within the main caption.
-% The latest version of endfloat.sty and its documentation can obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/endfloat/
-%
-% The IEEEtran \ifCLASSOPTIONcaptionsoff conditional can also be used
-% later in the document, say, to conditionally put the References on a 
-% page by themselves.
-
-% *** PDF, URL AND HYPERLINK PACKAGES ***
-%
-%\usepackage{url}
-% url.sty was written by Donald Arseneau. It provides better support for
-% handling and breaking URLs. url.sty is already installed on most LaTeX
-% systems. The latest version can be obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/misc/
-% Read the url.sty source comments for usage information. Basically,
-% \url{my_url_here}.
-
-% *** Do not adjust lengths that control margins, column widths, etc. ***
-% *** Do not use packages that alter fonts (such as pslatex).         ***
-% There should be no need to do such things with IEEEtran.cls V1.6 and later.
-% (Unless specifically asked to do so by the journal or conference you plan
-% to submit to, of course. )
-
-% correct bad hyphenation here
-\hyphenation{op-tical net-works semi-conduc-tor}
-
-%%% My packages
-\usepackage{amsmath,epsfig}
-\usepackage{amsfonts}
-\usepackage{color}
-\usepackage{epstopdf}
-
-%%% My commands
-\newcommand{\mat}[1]{\mathbf{#1}}
-\renewcommand{\vec}[1]{\mathbf{#1}}
-\newcommand{\R}{$\mat{\hat{R}}$ }
-\newcommand{\degree}{\ensuremath{^\circ}}
-
-\newcommand{\img}{img/}
-\epstopdfsetup{outdir=\img}
+\newcommand{\img}{Paper3/img/}
+\epstopdfsetup{outdir=Paper3/img}
 
 \newcommand\multimedia[1]{\textbf{{\color{red}[#1]}}}
-\newcommand\comment[1]{\textit{{\color{red}(#1)}}}
-
-\begin{document}
-%
-% paper title
-% can use linebreaks \\ within to get better formatting as desired
-\title{Capon Beamforming and Moving Objects - An Analysis of Lateral Shift-Invariance}
-%
-%
-% author names and IEEE memberships
-% note positions of commas and nonbreaking spaces ( ~ ) LaTeX will not break
-% a structure at a ~ so this keeps an author's name from being broken across
-% two lines.
-% use \thanks{} to gain access to the first footnote area
-% a separate \thanks must be used for each paragraph as LaTeX2e's \thanks
-% was not built to handle multiple paragraphs
-%
-
-\author{
-   Jon Petter \AA{}sen, \IEEEmembership{Student Member,~IEEE,} Andreas Austeng, \IEEEmembership{Member,~IEEE,} and Sverre Holm, \IEEEmembership{Senior Member,~IEEE}% <-this % stops a space
-
-\thanks{Manuscript received ...; accepted ...}
-\thanks{J. P. \AA{}sen and S. Holm is with the Medical Imaging Lab (MI-Lab) at the Norwegian University of Science and Technology, Trondheim, Norway.}
-\thanks{A. Austeng and S. Holm is with the Department of Informatics, University of Oslo, Oslo, Norway. (e-mail: jon.p.asen@ntnu.no)} 
-%J. I. Buskenes, C.-I. C. Nilsen, 
-}
-%M. Shell is with the Department
-%of Electrical and Computer Engineering, Georgia Institute of Technology, Atlanta,
-%GA, 30332 USA e-mail: (see http://www.michaelshell.org/contact.html).}% <-this % stops a space
-%\thanks{J. Doe and J. Doe are with Anonymous University.}% <-this % stops a space
-%\thanks{Manuscript received April 19, 2005; revised January 11, 2007.}}
-
-% note the % following the last \IEEEmembership and also \thanks - 
-% these prevent an unwanted space from occurring between the last author name
-% and the end of the author line. i.e., if you had this:
-% 
-% \author{....lastname \thanks{...} \thanks{...} }
-%                     ^------------^------------^----Do not want these spaces!
-%
-% a space would be appended to the last name and could cause every name on that
-% line to be shifted left slightly. This is one of those "LaTeX things". For
-% instance, "\textbf{A} \textbf{B}" will typeset as "A B" not "AB". To get
-% "AB" then you have to do: "\textbf{A}\textbf{B}"
-% \thanks is no different in this regard, so shield the last } of each \thanks
-% that ends a line with a % and do not let a space in before the next \thanks.
-% Spaces after \IEEEmembership other than the last one are OK (and needed) as
-% you are supposed to have spaces between the names. For what it is worth,
-% this is a minor point as most people would not even notice if the said evil
-% space somehow managed to creep in.
-
-% The paper headers
-\markboth{Transactions on Ultrasonics and Ferroelectrics and Frequency Control,~Vol.~x, No.~y, January~201\AA{}}%
-{\AA{}sen \MakeLowercase{\textit{et al.}}: The Capon Adaptive Beamformer Applied on Moving Objects }
-% The only time the second header will appear is for the odd numbered pages
-% after the title page when using the twoside option.
-% 
-% *** Note that you probably will NOT want to include the author's ***
-% *** name in the headers of peer review papers.                   ***
-% You can use \ifCLASSOPTIONpeerreview for conditional compilation here if
-% you desire.
-
-% If you want to put a publisher's ID mark on the page you can do it like
-% this:
-%\IEEEpubid{0000--0000/00\$00.00~\copyright~2007 IEEE}
-% Remember, if you use this you must call \IEEEpubidadjcol in the second
-% column for its text to clear the IEEEpubid mark.
-
-% use for special paper notices
-%\IEEEspecialpapernotice{(Invited Paper)}
-
-% make the title area
-\maketitle
+%\newcommand\comment[1]{\textit{{\color{red}(#1)}}}
 
@@
 \end{abstract}
-% IEEEtran.cls defaults to using nonbold math in the Abstract.
-% This preserves the distinction between vectors and scalars. However,
-% if the journal you are submitting to favors bold math in the abstract,
-% then you can use LaTeX's standard command \boldmath at the very start
-% of the abstract to achieve this. Many IEEE journals frown on math
-% in the abstract anyway.
-
-% Note that keywords are not normally used for peerreview papers.
-\begin{IEEEkeywords}
-Adaptive beamforming, Capon beamformer, Shift Invariance, Ultrasound Imaging.
-\end{IEEEkeywords}
-
-% For peer review papers, you can put extra information on the cover
-% page as needed:
-% \ifCLASSOPTIONpeerreview
-% \begin{center} \bfseries EDICS Category: 3-BBND \end{center}
-% \fi
-%
-% For peerreview papers, this IEEEtran command inserts a page break and
-% creates the second title. It will be ignored for other modes.
-\IEEEpeerreviewmaketitle
 
 \section{Introduction}
-% The very first letter is a 2 line initial drop letter followed
-% by the rest of the first word in caps.
-% 
-% form to use if the first word consists of a single letter:
-% \IEEEPARstart{A}{demo} file is ....
-% 
-% form to use if you need the single drop letter followed by
-% normal text (unknown if ever used by IEEE):
-% \IEEEPARstart{A}{}demo file is ....
-% 
-% Some journals put the first two words in caps:
-% \IEEEPARstart{T}{his demo} file is ....
-% 
-% Here we have the typical use of a "T" for an initial drop letter
-% and "HIS" in caps to complete the first word.
 
-\IEEEPARstart{I}{n} the last decade several authors have applied the Capon beamformer [also known as the minimum variance beamformer] for medical ultrasound imaging \cite{Synnevag2007, Vignon2008, Viola}.
+In the last decade several authors have applied the Capon beamformer [also known as the minimum variance beamformer] for medical ultrasound imaging \cite{Synnevag2007, Vignon2008, Viola}.
 The method has been shown to improve lateral resolution and edge sharpness in ultrasound images \cite{Synnevag2007, Synnevag2009, Chen2011}.
@@
 \centerline{
-\includegraphics[width=\linewidth]{\img fig1_2way.eps}
+\includegraphics[width=0.8\linewidth]{\img fig1_2way.eps}
 }
@@
 In this paper we investigate the issue further.
-We show that when a moving point scatterer is imaged with the Capon beamformer with the same beam spacing which is sufficient for DAS, the predicted scalloping loss is observed both in simulations and \textit{in-vitro} data.
+We show that when a moving point scatterer is imaged with the Capon beamformer with the same beam spacing which is sufficient for DAS, the predicted scalloping loss is observed both in simulations and \textit{in vitro} data.
 The question is then, what can we do in order to improve on the situation without affecting the acquisition frame rate and processing throughout.
@@
 Different methods for reduced scalloping loss and by that improved local lateral shift-invariance are presented in Section \ref{sec:methods}.
-Then, in Section \ref{sec:res} we present the results of applying the most promising method on \textit{in-vitro} data.
+Then, in Section \ref{sec:res} we present the results of applying the most promising method on \textit{in vitro} data.
 Finally we discuss our findings in Section \ref{sec:dis} and give our conclusions in Section \ref{sec:con}.
@@
 \end{align}
-where $K = M-L+1$ typically is proportional to the pulse length, $N_K = 2K + 1$, $N_L = M-L+1$, and $\vec{x}_l$ is a $L$ long subarray $[x_l[n], \dotso x_{l+L}[n]]$.
+where $K$ is selected proportional to the pulse length, $N_K = 2K + 1$, $N_L = M-L+1$, and $\vec{x}_l$ is a $L$ long subarray $[x_l[n], \dotso x_{l+L}[n]]$.
 Finally the matrix is loaded with a diagonal factor for numerical stability and increased robustness,
@@
 
-\section{In-vitro phantom data}\label{sec:res}
-In Fig.\,\ref{fig:LSV_invitro} we present the result of applying the Capon beamformer on \textit{in-vitro} data with and without oversampling by phase rotation.
+\section{In Vitro Phantom Data}\label{sec:res}
+In Fig.\,\ref{fig:LSV_invitro} we present the result of applying the Capon beamformer on \textit{in vitro} data with and without oversampling by phase rotation.
 The data was acquired using a modified Vivid E9 high-end cardiac ultrasound scanner by GE Vingmed Ultrasound, Horten, Norway, equipped with a M5S-D phased transducer operating in harmonic mode at 3.34 MHz.
@@
 
-%\subsection{In-vitro Images}
+%\subsection{In vitro Images}
 %Acquire channel data from a phantom containing point wire targets and cysts while moving the probe laterally. Confirm the improved shift-invariance when using phase rotation.
@@
 \begin{tabular}{cc}
-\subfloat[]{\includegraphics[width=0.45\linewidth]{\img fig6a.png}\label{fig:phantom_das_image}}
-\subfloat[]{\includegraphics[width=0.45\linewidth]{\img fig6b.eps}\label{fig:phantom_das_LSV}}\\
-\subfloat[]{\includegraphics[width=0.45\linewidth]{\img fig6c.eps}\label{fig:phantom_capon_LSV}}
-\subfloat[]{\includegraphics[width=0.45\linewidth]{\img fig6d.eps}\label{fig:phantom_capon_LSV_16PR}}
+\subfloat[]{\includegraphics[width=0.5\linewidth]{\img fig6a.png}\label{fig:phantom_das_image}}
+\subfloat[]{\includegraphics[width=0.5\linewidth]{\img fig6b.eps}\label{fig:phantom_das_LSV}}\\
+\subfloat[]{\includegraphics[width=0.5\linewidth]{\img fig6c.eps}\label{fig:phantom_capon_LSV}}
+\subfloat[]{\includegraphics[width=0.5\linewidth]{\img fig6d.eps}\label{fig:phantom_capon_LSV_16PR}}
 \end{tabular}
 }
-\caption{LSV-plots generated from \textit{in-vitro} phantom data where the probe has been manually dragged across the phantom.
+\caption{LSV-plots generated from \textit{in vitro} phantom data where the probe has been manually dragged across the phantom.
 See the supplementary material for videos of the whole loop processed with the different beamformers.
@@
 \centerline{
-\subfloat[]{\includegraphics[width=0.48\linewidth]{\img fig7a.png}\label{fig:phantom_capon_image}}
-\subfloat[]{\includegraphics[width=0.48\linewidth]{\img fig7b.png}\label{fig:phantom_capon_16PR_image}}
+\subfloat[]{\includegraphics[width=0.5\linewidth]{\img fig7a.png}\label{fig:phantom_capon_image}}
+\subfloat[]{\includegraphics[width=0.5\linewidth]{\img fig7b.png}\label{fig:phantom_capon_16PR_image}}
 }
@@
 Fig.\,\ref{fig:das_capon_beams} shows that 25 times oversampling is required for a $d$-value of $1/100$.
-When \textit{in-vitro} data contains additional noise this oversampling requirement is reduced.%Cox measures resolution as the 3 dB saddle point between two impinging monochromatic plane waves.
+When \textit{in vitro} data contains additional noise this oversampling requirement is reduced.%Cox measures resolution as the 3 dB saddle point between two impinging monochromatic plane waves.
 For ultrasound data only a 50 \% increase in resolution was found in \cite[Fig.~7]{Asen} for a 30 dB output SNR when measured in a similar way as Cox [6 dB saddle point].
 However, sixteen times oversampling was required to make the imaging system shift invariant in the same paper.
-In this paper more than ten times oversampling was found to be required for a 40 dB output SNR (after array gain) in simulations, however for the \textit{in-vitro} data ten times oversampling was enough for the system to be locally laterally shift-invariant.
+In this paper more than ten times oversampling was found to be required for a 40 dB output SNR (after array gain) in simulations, however for the \textit{in vitro} data ten times oversampling was enough for the system to be locally laterally shift-invariant.
 This might be a result of improved initial sampling (as discussed in Section \ref{sec:res}) and a higher initial noise level before diagonal loading was applied.
@@
 The most promising method, with respect to improved shift-invariance, reduction in acquisition frame rate and processing complexity combined, was based on phase rotation.
-%In Section \ref{sec:res} we presented results showing that this method also works on \textit{in-vitro} data when the ultrasound probe is subjected to motion.
+%In Section \ref{sec:res} we presented results showing that this method also works on \textit{in vitro} data when the ultrasound probe is subjected to motion.
 
@@
 In simulations, this method reduced the maximum scalloping loss by 16 dB.
-On \textit{in-vitro} data the reduction was around 7 dB and the point scatterer blinking was highly reduced.
+On \textit{in vitro} data the reduction was around 7 dB and the point scatterer blinking was highly reduced.
 
-% An example of a floating figure using the graphicx package.
-% Note that \label must occur AFTER (or within) \caption.
-% For figures, \caption should occur after the \includegraphics.
-% Note that IEEEtran v1.7 and later has special internal code that
-% is designed to preserve the operation of \label within \caption
-% even when the captionsoff option is in effect. However, because
-% of issues like this, it may be the safest practice to put all your
-% \label just after \caption rather than within \caption{}.
-%
-% Reminder: the "draftcls" or "draftclsnofoot", not "draft", class
-% option should be used if it is desired that the figures are to be
-% displayed while in draft mode.
-%
-%\begin{figure}[!t]
-%\centering
-%\includegraphics[width=2.5in]{myfigure}
-% where an .eps filename suffix will be assumed under latex, 
-% and a .pdf suffix will be assumed for pdflatex; or what has been declared
-% via \DeclareGraphicsExtensions.
-%\caption{Simulation Results}
-%\label{fig_sim}
-%\end{figure}
-
-% Note that IEEE typically puts floats only at the top, even when this
-% results in a large percentage of a column being occupied by floats.
-
-% An example of a double column floating figure using two subfigures.
-% (The subfig.sty package must be loaded for this to work.)
-% The subfigure \label commands are set within each subfloat command, the
-% \label for the overall figure must come after \caption.
-% \hfil must be used as a separator to get equal spacing.
-% The subfigure.sty package works much the same way, except \subfigure is
-% used instead of \subfloat.
-%
-%\begin{figure*}[!t]
-%\centerline{\subfloat[Case I]\includegraphics[width=2.5in]{subfigcase1}%
-%\label{fig_first_case}}
-%\hfil
-%\subfloat[Case II]{\includegraphics[width=2.5in]{subfigcase2}%
-%\label{fig_second_case}}}
-%\caption{Simulation results}
-%\label{fig_sim}
-%\end{figure*}
-%
-% Note that often IEEE papers with subfigures do not employ subfigure
-% captions (using the optional argument to \subfloat), but instead will
-% reference/describe all of them (a), (b), etc., within the main caption.
-
-% An example of a floating table. Note that, for IEEE style tables, the 
-% \caption command should come BEFORE the table. Table text will default to
-% \footnotesize as IEEE normally uses this smaller font for tables.
-% The \label must come after \caption as always.
-%
-%\begin{table}[!t]
-%% increase table row spacing, adjust to taste
-%\renewcommand{\arraystretch}{1.3}
-% if using array.sty, it might be a good idea to tweak the value of
-% \extrarowheight as needed to properly center the text within the cells
-%\caption{An Example of a Table}
-%\label{table_example}
-%\centering
-%% Some packages, such as MDW tools, offer better commands for making tables
-%% than the plain LaTeX2e tabular which is used here.
-%\begin{tabular}{|c||c|}
-%\hline
-%One & Two\\
-%\hline
-%Three & Four\\
-%\hline
-%\end{tabular}
-%\end{table}
-
-% Note that IEEE does not put floats in the very first column - or typically
-% anywhere on the first page for that matter. Also, in-text middle ("here")
-% positioning is not used. Most IEEE journals use top floats exclusively.
-% Note that, LaTeX2e, unlike IEEE journals, places footnotes above bottom
-% floats. This can be corrected via the \fnbelowfloat command of the
-% stfloats package.
-
-%\section{Conclusion}
-%The conclusion goes here.
-
-% if have a single appendix:
-%\appendix[Proof of the Zonklar Equations]
-% or
-%\appendix  % for no appendix heading
-% do not use \section anymore after \appendix, only \section*
-% is possibly needed
-
-% use appendices with more than one appendix
-% then use \section to start each appendix
-% you must declare a \section before using any
-% \subsection or using \label (\appendices by itself
-% starts a section numbered zero.)
-%
-
-%\appendices
-%\section{Proof of the First Zonklar Equation}
-%Appendix one text goes here.
-
-% you can choose not to have a title for an appendix
-% if you want by leaving the argument blank
-%\section{}
-%Appendix two text goes here.
-
-% use section* for acknowledgement
 \section*{Acknowledgment}
 
-The authors would like to thank Anders S\o{}rnes at GE Vingmed Ultrasound, Horten, Norway for the help with acquiring \textit{in-vitro} data.
-
-% Can use something like this to put references on a page
-% by themselves when using endfloat and the captionsoff option.
-\ifCLASSOPTIONcaptionsoff
-  \newpage
-\fi
-
-% trigger a \newpage just before the given reference
-% number - used to balance the columns on the last page
-% adjust value as needed - may need to be readjusted if
-% the document is modified later
-%\IEEEtriggeratref{8}
-% The "triggered" command can be changed if desired:
-%\IEEEtriggercmd{\enlargethispage{-5in}}
-
-% references section
-
-% can use a bibliography generated by BibTeX as a .bbl file
-% BibTeX documentation can be easily obtained at:
-% http://www.ctan.org/tex-archive/biblio/bibtex/contrib/doc/
-% The IEEEtran BibTeX style support page is at:
-% http://www.michaelshell.org/tex/ieeetran/bibtex/
-\bibliographystyle{IEEEtran}
-% argument is your BibTeX string definitions and bibliography database(s)
-%\bibliography{IEEEabrv,../bib/paper}
-\bibliography{mybib}
-%
-% <OR> manually copy in the resultant .bbl file
-% set second argument of \begin to the number of references
-% (used to reserve space for the reference number labels box)
-%\begin{thebibliography}{1}
-
-%\bibitem{IEEEhowto:kopka}
-%H.~Kopka and P.~W. Daly, \emph{A Guide to \LaTeX}, 3rd~ed.\hskip 1em plus
-%  0.5em minus 0.4em\relax Harlow, England: Addison-Wesley, 1999.
-
-%\end{thebibliography}
-
-% biography section
-% 
-% If you have an EPS/PDF photo (graphicx package needed) extra braces are
-% needed around the contents of the optional argument to biography to prevent
-% the LaTeX parser from getting confused when it sees the complicated
-% \includegraphics command within an optional argument. (You could create
-% your own custom macro containing the \includegraphics command to make things
-% simpler here.)
-%\begin{biography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{mshell}}]{Michael Shell}
-% or if you just want to reserve a space for a photo:
-
-% insert where needed to balance the two columns on the last page with
-% biographies
-
-%\newpage
-
-\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{photos/jon_petter2.eps}}]{Jon Petter \AA{}sen}
-(S12) was born in Porsgrunn, Norway in 1986.
-He received the B.Sc.
-and M.Sc.
-degree in computer science from the University of Oslo, Norway, in 2008 and 2010.
-He is currently pursuing his Ph.D.
-degree in medical ultrasound technology at the Norwegian University of Science and Technology (NTNU) Medical Imaging Lab (MI-Lab), Trondheim, Norway.
-
-He has industry experience from GE Vingmed Ultrasound, Horten, Norway, were he completed three summer internships from 2008 to 2010.
-His research interests include image and signal processing, adaptive ultrasound processing techniques and acceleration of ultrasound algorithms using Graphics Processing Units (GPUs).
-\end{IEEEbiography}
-
-\vfill{}
-
-%\newpage
-
-\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{photos/andreas.eps}}]{Andreas Austeng}
-was born in Oslo, Norway, in 1970.
-He received the M.S.
-degree in physics in 1996 and the Ph.D.
-degree in computer science in 2001, both from the University of Oslo.
-Since 2001, he has been working at the Department of Informatics, University of Oslo, first as a postdoctoral research fellow and currently as an associate professor.
-His research interests include signal and array processing for acoustical imaging.
-\end{IEEEbiography}
-
-\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{photos/sverre.eps}}]{Sverre Holm}
-(M82-SM02) was born in Oslo, Norway, in 1954.
-He received the M.S.
-and Ph.D.
-degrees in electrical engineering from the Norwegian Institute of Technology, Trondheim, in 1978 and 1982, respectively.
-His academic experience includes Yarmouk University in Jordan (1984-1986) and the Norwegian Institute of Technology and later the University of Oslo (1989-1994) as an adjunct professor.
-Since 1995, he has been a professor of signal processing in the Department of Informatics at the University of Oslo.
-In 2002, he was elected a member of the Norwegian Academy of Technological Sciences, and in 2009, he became an adjunct professor in the Norwegian University of Science and Technology (NTNU) Medical Imaging Lab, Trondheim.
-His industry experience includes GE Vingmed Ultrasound, Horten, Norway (1990-1994).
-He worked with Sonitor Technologies (2000-2005), where he developed ultrasonic technology that is the core of the indoor positioning system described in \textit{Scientific Americanin} 2008 as "a positioning system that goes where GPS can't." He has published about 130 publications internationally, plus about 10 patents.
-He has had sabbaticals at GE Global Research, NY (1998), and the Waves and Acoustics Laboratory of ESPCI, Paris (2008-2009).
-He was an associate editor of the \textit{IEEE Transactions on Ultrasonics, Ferroelectrics, and Frequency Control} from 1997 to 2002.
-His research interests are medical ultrasound imaging, modeling of tissue-ultrasound interaction, and ultrasound positioning.
-\end{IEEEbiography}
-
-\vfill
-
-% You can push biographies down or up by placing
-% a \vfill before or after them. The appropriate
-% use of \vfill depends on what kind of text is
-% on the last page and whether or not the columns
-% are being equalized.
-
-%\vfill
-
-% Can be used to pull up biographies so that the bottom of the last one
-% is flush with the other column.
-%\enlargethispage{-5in}
-
-% that's all folks
-\end{document}
+The authors would like to thank Anders S\o{}rnes at GE Vingmed Ultrasound, Horten, Norway for the help with acquiring \textit{in vitro} data.

### 2014-04-05 ThesisJP/Thesis.tex
Writing background...
Changed symbol for resolution.
Writing background.
Added Paper 4.
Added paper 5.
Small cosmetic changes.
Added Paper2 and Paper3.

@@
 % TITLE PAGE
-\author{Jon Petter \AA{}sen}
+\author{Jon Petter Helgesen \AA{}sen}
 \title{%Integrated Ultrasound Visualization and Formation
@@
 %Accelerating adaptive ultrasound algorithms with graphics processing units 
-Accelerating adaptive ultrasound algorithms by means of general-purpose programming on graphics processing units 
+Accelerating adaptive ultrasound imaging algorithms by means of general-purpose computing on graphics processing units 
 %Accelerating adaptive ultrasound algorithms by general purpose programming on graphics processing units 
@@
 
+% from Paper 3
+\usepackage{epstopdf}
+
+% from spie.tex (Paper 4)
+\usepackage{array}
+
+% from hos (Paper 5)
+\usepackage{url}
+%\usepackage{epsfig}
+
 % symbol list
@@
 \newcommand*{\nom}[2]{#1\nomenclature{#1}{#2}}
+\newcommand*{\nomi}[2]{\nomenclature{#1}{#2}}
 
@@
 \vspace{0.3cm}
-\noindent Jon Petter \AA{}sen\\
+\noindent Jon Petter Helgesen \AA{}sen\\
 Institutt for sirkulasjon og bildediagnostikk, NTNU\\
@@
 \todo{Write acknowledgments ...}
+Takkeliste: Beate.
+Farmor.
+Familien.
+Andreas Austeng.
+Jo Inge Buskenes.
+Carl-Inge Colombo Nilsen.
+Ines Hafizovic for feedback on the introduction and background.
+Sverre Holm.
+Miljøet i Trondheim.
+PhD studentene i Trondheim spesielt.
+Miljøet i Oslo.
+Miljøet hos Vingmed.
+SquareHead.
 
@@
 	\item \textbf{J.\:P.\:\AA{}sen}, J.\:I.\:Buskenes, C.-I.\:C.\:Nilsen, A.\:Austeng and S.\:Holm, "Implementing Capon Beamforming on the GPU for Real-Time Cardiac Ultrasound Imaging", {\it Proc.
-IEEE Ultrasonics Symposium 2012}, pp.
-2133-2136.
+IEEE Ultrasonics Symposium 2012, pp.
+2133-2136}.
 
 	\item J.\:I.\:Buskenes, \textbf{J.\:P.\:\AA{}sen}, C.-I.\:C.\:Nilsen and A.\:Austeng, "Adapting the Minimum variance beamformer to a graphics processing unit for active sonar imaging systems", {\it Proc.
-Meetings on Acoustics 2013} vol.
-13.
+Meetings on Acoustics 2013, vol.
+13}.
+\textit{Invited talk}.
 
 	\item G.\:Kiss, E.\:Steen, \textbf{J.\:P.\:\AA{}sen} and H.\:Torp, "GPU volume rendering in 3D echocardiography: real-time pre-processing and ray-casting", {\it Proc.
-IEEE Ultrasonics Symposium 2010}, pp.
-193-196.
+IEEE Ultrasonics Symposium 2010, pp.
+193-196}.
 
@@
 	\item E.\:S.\:Br\o{}nstad, \textbf{J.\:P.\:\AA{}sen}, H.\:Torp and G.\:Kiss, "Visibility driven visualization of 3D cardiac ultrasound data on the GPU", {\it Proc.
-IEEE Ultrasonics Symposium 2012}, pp.
-2651-2654.
+IEEE Ultrasonics Symposium 2012, pp.
+2651-2654}.
 \end{enumerate}
@@
 %%%% CHAPTER 2 %%%%
-\begin{bibunit}[ieeetr]
+\begin{bibunit}[IEEEtran]
 \CustChapter{Background}
@@
 %%%% Paper 1 %%%%
-\begin{bibunit}[ieeetr]
+\begin{bibunit}[IEEEtran]
 \CustChapterPaper{An Optimised GPU Implementation of the MVDR Beamformer for Active Sonar Imaging}
@@
 %%%% Paper 2 %%%%
-\begin{bibunit}[ieeetr]
+\begin{bibunit}[IEEEtran]
 \CustChapterPaper{Implementing Capon Beamforming on a GPU for Real-Time Cardiac Ultrasound Imaging}
@@
 \noindent \textit{IEEE Transactions on Ultrasonics, Ferroelectrics, and Frequency Control, vol. 61, pp. 76-85, Jan 2014}.
-%\input{./CaponCardiacGPU/CaponCardiacGPU}
-%\newpage\addcontentsline{toc}{section}{References}
-%\putbib[./CaponCardiacGPU/bibCaponCardiacGPU]
+\newpage
+{\input{./Paper2/cos}}
+\newpage
+\addcontentsline{toc}{section}{References}
+\putbib[./Paper2/cosbib]
 \end{bibunit}
@@
 %%%% Paper 3 %%%%
-\begin{bibunit}[ieeetr]
+\begin{bibunit}[IEEEtran]
 \CustChapterPaper{Capon Beamforming and Moving Objects - An Analysis of Lateral Shift-Invariance}
@@
 \noindent \textit{IEEE Transactions on Ultrasonics, Ferroelectrics, and Frequency Control, accepted for publication}.
-%\input{./CaponMovingObjects/CaponMovingObjects}
-%\newpage\addcontentsline{toc}{section}{References}
-%\putbib[./CaponMovingObjects/bibCaponMovingObjects]
+\newpage\input{./Paper3/moving_obj}
+\newpage\addcontentsline{toc}{section}{References}
+\putbib[./Paper3/moving_objbib]
 \end{bibunit}
@@
 %%%% Paper 4 %%%%
-\begin{bibunit}[ieeetr]
+\begin{bibunit}[IEEEtran]
 \CustChapterPaper{Adaptive Volume Rendering of Cardiac 3D Ultrasound Images - Utilizing Blood Pool Statistics}
@@
 \noindent \textit{Proc. SPIE Medical Imaging 2012, vol. 8320, pp. 832008.}
-%\input{./AdaptiveVolumeRendering/AdaptiveVolumeRendering}
-%\newpage\addcontentsline{toc}{section}{References}
-%\putbib[./AdaptiveVolumeRendering/bibAdaptiveVolumeRendering]
+\newpage\input{./Paper4/spie}
+\newpage\addcontentsline{toc}{section}{References}
+\putbib[./Paper4/spiebib]
 \end{bibunit}
@@
 %%%% Paper 5 %%%%
-\begin{bibunit}[ieeetr]
+\begin{bibunit}[IEEEtran]
 \CustChapterPaper{Huygens on Speed: Interactive Simulation of Ultrasound Pressure Fields}
@@
 \noindent \textit{Proc. IEEE Ultrasonics Symposium 2012, pp. 1643-1646}.
-%\input{./Hos/Hos}
-%\newpage\addcontentsline{toc}{section}{References}
-%\putbib[./Hos/bibHos]
+\newpage\input{./Paper5/hos}
+\newpage\addcontentsline{toc}{section}{References}
+\putbib[./Paper5/hosbib]
 \end{bibunit}

### 2014-04-06 ThesisJP/Background/background.tex
Changed from subfigure to subfloat in Backgroun chapter.
Small update. Started on concluding remarks section in Background chapter.

@@
 \centering
-\subfigure[Kernel]{
+\subfloat[Kernel]{
 	\includegraphics[width=0.5\textwidth]{img/kernel.pdf}\label{fig:gpu_kernel}
 }
-\subfigure[Threads, blocks and grid]{
+\subfloat[Threads, blocks and grid]{
 	\includegraphics[width=0.4\textwidth]{img/cuda_threads.png}\label{fig:gpu_grid}
@@
 \centering
-\subfigure[]{
+\subfloat[]{
 	\includegraphics[width=0.47\textwidth]{img/Diagram_of_the_human_heart.png}\label{fig:human_heart}
 }
-\subfigure[]{
+\subfloat[]{
 	\includegraphics[width=0.47\textwidth]{img/ultrasound_grid.pdf}\label{fig:us_grid}
@@
 \subsection{Beamforming}
-If time delays are applied on the signal emitted by the elements it is possible to steer and focus the ultrasound beam in space (Fig.\,\ref{fig:das}).
+If time delays are applied on the signal emitted by the elements it is possible to steer and focus the ultrasound beam in space (Fig.\,\ref{fig:das_background}).
 This technique is known as \textit{beamforming}, and for ultrasound imaging it is applied both when transmitting and receiving ultrasound pulses.
@@
 \caption{Delay-and-sum array beamforming.}
-\label{fig:das}
+\label{fig:das_background}
 \end{figure}
@@
 The standard form of beamforming used on receive in ultrasound imaging is known as \textit{delay-and-sum} (DAS) beamforming: 
-\begin{align}\label{eq:das}
+\begin{align}\label{eq:das_background}
 z[n] = \sum_{m = 0}^{M-1}w_m^*x_m[n - \Delta_m[n]] = \vec{w}^H\vec{x}[n].
@@
 From now on we will omit the dependency on $\Delta_m[n]$ and $x_m$ will refer to delayed element data.
-A schematic overview of delay-and-sum array beamforming is depicted in Fig.\,\ref{fig:das}.
+A schematic overview of delay-and-sum array beamforming is depicted in Fig.\,\ref{fig:das_background}.
 If $p_o$ is the phase center of our array and $p$ is the point to which we want to steer and focus the ultrasound beam, we have to use the following delays:
@@
 One of  the last remaining ASICs in ultrasound scanners is used for beamforming \cite{Thomenius}.
-As seen from (\ref{eq:das}), beamforming involves a reduction of the data rate with a factor equal to the number of elements in the ultrasound probe.
+As seen from (\ref{eq:das_background}), beamforming involves a reduction of the data rate with a factor equal to the number of elements in the ultrasound probe.
 The input data rate is therefore very high.
@@
 \begin{figure}[t!]
-\subfigure[Uniform weights]{
+\subfloat[Uniform weights]{
 	\includegraphics[width=0.47\textwidth]{img/scenario_das_resp2.png}\label{fig:das_weights}
 }
-\subfigure[Capon weights]{
+\subfloat[Capon weights]{
 	\includegraphics[width=0.47\textwidth]{img/scenario_mv_resp2.png}\label{fig:capon_weights}
@@
 
-The Capon beamformer produces a weight vector, as introduced in (\ref{eq:das}), based on the impinging signal and a constrained optimization problem. The problem is as follow \cite{Capon1969}:
+The Capon beamformer produces a weight vector, as introduced in (\ref{eq:das_background}), based on the impinging signal and a constrained optimization problem.
+The problem is as follow \cite{Capon1969}:
 \begin{align}
-&\min_{\vec{w}} E\{|z[n]|^2\} \rightarrow \min_{\vec{w}} \vec{w}^H \mat{\hat{R}} \vec{w} \label{eq:capon_optimization_criteria} \\
+&\min_{\vec{w}} E\{|z[n]|^2\} \rightarrow \min_{\vec{w}} \vec{w}^H \mat{\hat{R}} \vec{w} \label{eq:capon_optimization_criteria_background} \\
 &\text{subjected to } \vec{w}^H\vec{a} = 1,
@@
 \mat{\breve{R}}[n] &= \frac{1}{N_LN_K}\sum_{n'=n-K}^{n+K} \sum_{l=0}^{N_L-1} \vec{x}_l[n']\vec{x}_l[n']^H\\
-\vec{x}_l[n'] &= [x_l[n], \dotso, x_{l+L}[n]]
+\vec{x}_l[n'] &= [x_l[n'], \dotso, x_{l+L}[n']]
 \end{align}
@@
 
-The solution to the constrained minimization problem in (\ref{eq:capon_optimization_criteria}) is, by the method of Lagrange multipliers \cite{VanTrees2003}, found to be
+The solution to the constrained minimization problem in (\ref{eq:capon_optimization_criteria_background}) is, by the method of Lagrange multipliers \cite{VanTrees2003}, found to be
 \begin{align}\label{eq:capon_weights}
@@
 The required throughput for Capon beamforming might therefore still be hard to achieve in practice \cite[\textbf{Paper II}]{So2011}.
-As for DAS beamforming this number also scales with higher relative bandwidth, pulse frequency, and more array elements since the subarray size typically is proportional to the number of array elements.
+As for DAS beamforming this number also scales with higher relative bandwidth, pulse frequency, and more array elements since the subarray size is selected proportional to the number of array elements.
 
@@
 The idea behind BS-Capon beamforming is that spatial beams with little information can be removed from $\mat{\hat{R}}$ by removing rows in $B$.
-When investigating the complexity of the BS-Capon beamformer (\textbf{Paper II}) we find that is dominated by the transformation of data into beamspace, $O(N_LLN_b)$, where \nom{$N_b$}{Number of selected beams in beamspace} is the number of selected beams.
+When investigating the complexity of the BS-Capon beamformer (\textbf{Paper II}) we find that it is dominated by the transformation of data into beamspace, $O(N_LLN_b)$, where \nom{$N_b$}{Number of selected beams in beamspace} is the number of selected beams.
 
@@
 
+\subsection{Narrowband vs Broadband Capon beamforming}
 Both the LCA and BS-Capon beamformer is based upon a detailed investigation of how Capon beamfoming behaves when applied to medical ultrasound images.
@@
 
-The Capon beamformer is derived for narrowband applications and has traditionally been applied on medical ultrasound imaging "as is".
+The Capon beamformer is derived for narrowband applications and has traditionally been applied to medical ultrasound imaging "as is".
 Medical ultrasound imaging is, however, a broadband imaging modality and one could think that a broadband Capon beamformer therefore where strictly needed \cite{Holfort2009}.
-The investigation performed in \cite{Nilsen2009} showed that with pre-delayed beams, energy will be centered on-axis and as little as three near-axis beams is therefore adequate in order to achieve similar performance for BS-Capon and Capon beamforming.
+The investigation performed in \cite{Nilsen2009} showed that with pre-delayed data, energy will be centered on-axis and as little as three near-axis beams is therefore adequate in order to achieve similar performance for BS-Capon and Capon beamforming.
 The two-way beam will be wider for unfocused (plane wave) transmit beams \cite{Holfort2009, Austeng2011}, but steering and focusing on receive will still make the main amount of energy arrive on axis.
@@
 \centering
-\subfigure[Ray-casting]{
+\subfloat[Ray-casting]{
 	\includegraphics[width=0.4\textwidth]{img/Volumeraycasting.png}\label{fig:vr_rc}
 }
-\subfigure[Opacity transfer function]{
+\subfloat[Opacity transfer function]{
 	\includegraphics[width=0.5\textwidth]{img/otf.png}\label{fig:vr_otf}
@@
 
+\section{Concluding remarks}
+All the presented problems in this chapter has in common that they require a processing throughput from ten to several hundred GFLOPS.
+Comparing this to the typical clock frequency of modern CPUs and GPUs (1.5 - 3 GHz) we see that GFLOPS problems requires parallel execution in order to achieve real-time performance.
+
 \endinput

### 2014-04-09 ThesisJP/Background/background.tex
Started working on comments from CI.

@@
 However, power consumption and heat generation where also growing at the same rate until the required cooling finally became to much in 2004.
-From that point, chip manufactures had to pursue a different direction  than just increasing the clock frequency.
+From that point, CPU manufactures had to pursue a different direction  than just increasing the clock frequency.
 The immediate solution was to embed several cores in one CPU.
@@
 Although they started out as very different processors, modern CPUs are now including more and more GPU-like features, and GPUs are adding CPU features.
-In this section we will see what these features are and why there is still a big difference in peak performance between a CPU and a GPU.
+In this section we will see what these features are and why there is still a big difference in theoretical peak performance between a CPU and a GPU.
 
@@
 \section {Medical ultrasound imaging}\label{sec:ultrasound}
-Medical ultrasound imaging is considered to be a low-risk tool for non-invasive investigation.
+Medical ultrasound imaging is considered to be a low-risk tool for non-invasive investigation of the human body.
 Neither gamma rays nor high intensity magnetic fields are involved, just sound pressure below well-controlled limits.
-It is also far less costly than other medical imaging modalities and offers real-time inspection of the human body.
+It is also far less costly than other medical imaging modalities and offers real-time imaging.
 Ultrasound imaging is well known for its use in fetal imaging during pregnancy, however, in this thesis we will focus on a different modality, namely cardiac ultrasound imaging.
@@
 It is therefore the strongest and hardest working part of the heart.
-When blood flows out of the aorta it is important that the mitral valve, which separates the left atrium from the left ventricle, is closed and does not leak.
+When blood flows out of the aorta, it is important that the mitral valve, which separates the left atrium from the left ventricle, is closed and does not leak.
 The left side, and especially the left ventricle and the mitral valve, therefore receives the most attention during cardiac ultrasound examinations.
@@
 To generate the ultrasonic frequencies used for imaging, a piezoelectric ceramic is used.
-A piezoelectric material is special in that way that ultrasound vibrations are generated when voltage is applied and \textit{vice versa}.
+A piezoelectric material has a special property where mechanical vibrations are generated when voltage is applied and \textit{vice versa}.
 It can therefore be used to both send and receive ultrasound signals, like a combined microphone and speaker.
-Piezoelectric elements organized in an array, is referred to as an ultrasound \textit{probe}, and a special probe is typically designed for each medical imaging modality.
+A set of piezoelectric elements organized in an array is referred to as an ultrasound \textit{probe}, and special probes are typically designed for different medical imaging modalities.
 An example of a probe for medical ultrasound imaging is shown in Fig.\,\ref{fig:us_grid}.
@@
 If the width of a piezoelectric element is less than the wavelength of the transmitted signal the element is said to be \textit{omnidirectional}, hence, energy will be distributed or sensed equally in all directions.
-When multiple omnidirectional elements are arranged in an array the transmitted pulse will get shaped into a beam\footnote{A beam refers to the pulse in space integrated across time.} of ultrasound by diffraction as the size (the \textit{aperture}) of the array grows.
-The angular width of this beam, and also the angular resolution of the array, is approximately given by \cite{AngelUltrasound}:
+When multiple omnidirectional elements are arranged in an array, the transmitted pulse will get shaped into a beam\footnote{A beam refers to the accumulated magnitude across time of a pulse traveling in space.} of ultrasound by diffraction as the width (the \textit{aperture}) of the array grows.
+The angular width of this beam, and also the angular resolution in radians of the array, is approximately given by \cite{AngelUltrasound}:
 \begin{align}\label{eq:res}
@@
 \end{align}
-\nomi{$\delta\theta_a$}{Angular resolution of an aperture}where \nom{$\lambda$}{Wavelength} and \nom{$f$}{Frequency} is the ultrasound wavelength and frequency respectively, $c$ is the speed of sound in human tissue, and \nom{$D$}{Aperture size} is the aperture width.
-With an average speed-of-sound in human tissue of $1540\frac{\text{m}}{\text{s}}$, the angular resolution or opening angle of a 2 cm cardiac probe operating at 3 MHz is around 1.5\degree.
-The axial resolution is inverse proportional to the pulse length, and is typically better than the lateral resolution.
+\nomi{$\delta\theta_a$}{Angular resolution of an aperture}where \nom{$\lambda$}{Wavelength} and \nom{$f$}{Frequency} are the ultrasound wavelength and frequency respectively, $c$ is the speed of sound in human tissue, and \nom{$D$}{Aperture size} is the aperture width.
+With an average speed of sound in human tissue of $1540\frac{\text{m}}{\text{s}}$, the angular resolution or opening angle of a 2 cm cardiac probe operating at 3 MHz is around 1.5\degree.
+The axial resolution is inversely proportional to the pulse length, and is typically better than the lateral resolution.
 The geometry of cardiac ultrasound imaging is depicted in Fig.\,\ref{fig:us_grid}.
@@
 \subsection{Beamforming}
-If time delays are applied on the signal emitted by the elements it is possible to steer and focus the ultrasound beam in space (Fig.\,\ref{fig:das_background}).
+If time delays are applied to the signal emitted by the elements it is possible to steer and focus the ultrasound beam in space (Fig.\,\ref{fig:das_background}).
 This technique is known as \textit{beamforming}, and for ultrasound imaging it is applied both when transmitting and receiving ultrasound pulses.
@@
 This is known as a 2D B-mode image (B for brightness).
-If image lines are acquired for multiple B-mode images which are tilted in the $yz$-plane we get what is known as a volumetric B-mode or a 3D ultrasound image.
-A probe for 3D imaging has several thousand elements arranged as a matrix in order to facilitate steering in both the $xz$ and $yz$-plane.
+If image lines are acquired for multiple B-mode images that are tilted in the $yz$-plane we get what is known as a volumetric B-mode or a 3D ultrasound image.
+A probe for 3D imaging has several thousand elements arranged as a matrix in order to facilitate steering in both the $xz$ and $yz$ planes.
 In this thesis we will focus on 2D imaging when the topic is adaptive beamforming, and 3D imaging when the topic is adaptive visualization.
@@
 
-The standard form of beamforming used on receive in ultrasound imaging is known as \textit{delay-and-sum} (DAS) beamforming: 
+The standard form of beamforming used in ultrasound imaging is known as \textit{delay-and-sum} (DAS): 
 \begin{align}\label{eq:das_background}
@@
 where \nom{$c$}{Speed of sound} is the speed of sound in human tissue, and $p_m$ is the position of the m'th element.
-On transmit the inverse set of delays has to be used.
+When transmitting the pipeline depicted in Fig.\,\ref{fig:das_background}, without the sum operator, is reversed and the negated set of delays has to be used.
 Also note that on transmit we have to decide where to steer and focus prior to transmitting a pulse.
@@
 \subsection{Sampling and processing complexity}
-When generating a B-mode image we have to insure that the spatial and temporal sampling are sufficient.
+When generating a B-mode image, we have to insure that the spatial and temporal sampling are sufficient.
 We have already seen that the lateral resolution of an ultrasound probe is given by (\ref{eq:res}).
-Hergum \textit{et al.} \cite{Hergum2009} have further shown that the lateral frequency support for a system which both emit and receive energy is governed by the sum of the transmit ($D_{tx}$\nomi{tx}{Transmit}) and receive ($D_{rx}$\nomi{rx}{Receive}) aperture.
+Hergum \textit{et al.} \cite{Hergum2009} have further shown that the lateral frequency support for a system that both emits and receives energy is governed by the sum of the transmit ($D_{tx}$\nomi{tx}{Transmit}) and receive ($D_{rx}$\nomi{rx}{Receive}) aperture.
 The lateral resolution of what is known as an active imaging system is found to be
@@
 \end{align}
-\nomi{$N$}{Number of samples in range}where \nom{$B_w$}{Relative bandwidth} is the relative bandwidth in percent of \nom{$f_c$}{Center frequency of ultrasound pulse}, where $f_c$ is the center frequency of the transmitted pulse.
+\nomi{$N$}{Number of samples in range (or time)}where \nom{$B_w$}{Relative bandwidth} is the relative bandwidth in percent of \nom{$f_c$}{Center frequency of ultrasound pulse}, where $f_c$ is the center frequency of the transmitted pulse.
 Note that after this process the ultrasound data will be complex values.
@@
 \end{align}
-If our cardiac probe, which is used to image a $70$\degree{} sector, has a relative bandwidth of $80$ \% we have to process almost $50000$ $M$-long vectors of complex element data every $1/55$ second.
+If our cardiac probe, which is used to image a $70$\degree{} sector, has a relative bandwidth of $80$ \% we have to process almost $50000$ complex data vectors of length $M$ every $1/55$ second.
 The formula for the number of FLOPS required for real-time DAS beamforming of cardiac ultrasound is then roughly given as:
@@
 If our cardiac probe has $M=128$ elements, DAS beamforming requires a processing throughput of around $300$ MFLOPS.
-Hence, far below the peak processing throughput of modern GPUs and CPUs.
+This is far below the peak processing throughput of modern GPUs and CPUs.
 The main challange with DAS beamforming is the amount of data that has to be moved.
@@
 This is however less than the maximum bandwidth of PCI Express 2.0 (8 GB/s), even after the relative bandwidth, transmit frequency and number of array elements are increased with a total factor of 5.
-Newer versions of PCI Express which supports even higher transfer rates have also recently been introduced.
+Newer versions of PCI Express, which supports even higher transfer rates, have also recently been introduced.
 
@@
 Ultrasound scanners generate an incredible amount of data.
-A modern system capable of 3D imaging will generate several tera bytes of data inside the probe handle during a clinical scan.
+A modern system capable of 3D imaging will generate several tera-bytes of data inside the probe handle during a clinical scan.
 Fortunately the data rate is reduced to the level of 2D imaging before the data reaches the scanner.
-The data rate of a 2D scan is "only" in order of giga bytes per second.
+The data rate of a 2D scan is "only" in order of giga-bytes per second.
 To handle this amount of data, ultrasound imaging systems have, for the last twenty years, relied heavily on application-specific integrated circuits (\nom{ASIC}{Application-specific integrated circuit}) for all steps in the processing pipeline \cite{Thomenius}.
-However, as the computational power of generic processors have increased, ASICs have gradually been replaced by software \cite{Guracar2013}.
+However, as the computational power of generic processors have increased, the expensive ASICs have gradually been replaced by software \cite{Guracar2013}.
 One of  the last remaining ASICs in ultrasound scanners is used for beamforming \cite{Thomenius}.
@@
 This is the same technique as reported implemented on a GPU by Yiu \textit{et al.} \cite{Yiu2011}.
-We therfore can conclude that beamforming is finally moving into software too.
+We can therefore conclude that beamforming is finally moving into software too.
 The largest players in the medical ultrasound business, General Electric and Philips, are also well aware of where the trend is going \cite{Thomenius2012} \cite{Metz2011}.
-Software beamforming opens up for new possibilities for advanced processing of ultrasound data.
-One example is Capon beamforming which is explored in this thesis (\textbf{Paper\,I} and \textbf{II}), and which is introduce in the next section.
+Software beamforming opens up for cost-efficient and rapid prototyping of advanced processing of ultrasound data.
+One example is Capon beamforming which is explored in this thesis (\textbf{Paper\,I} and \textbf{II}) and introduced in the next section.
 The increased computational power introduced by GPUs makes it also possible to include more advanced 3D visualization as presented in \cite{solteszova2010multidirectional}, \textbf{Paper\,IV}, and \textbf{IX}.
@@
 The idea behind BS-Capon beamforming is that spatial beams with little information can be removed from $\mat{\hat{R}}$ by removing rows in $B$.
-When investigating the complexity of the BS-Capon beamformer (\textbf{Paper II}) we find that it is dominated by the transformation of data into beamspace, $O(N_LLN_b)$, where \nom{$N_b$}{Number of selected beams in beamspace} is the number of selected beams.
+When investigating the complexity of the BS-Capon beamformer (\textbf{Paper II}) we find that it is dominated by the transformation of data into beamspace, $O(N_LLN_b)$, where \nom{$N_b$}{Number of selected beams in beamspace (BS-Capon)} is the number of selected beams.
 

### 2014-04-10 ThesisJP/Background/background.tex
Finished first set of comments from CI.

@@
 \subsection{Adaptive beamforming}\label{sec:adaptbf}
-Adaptive beamforming has several meanings in the literature on medical ultrasound imaging.
+Adaptive beamforming has several meanings in the medical ultrasound imaging literature.
 It can for instance refer to aberration correction \cite{cole1996method} which is the task of applying time delays to compensate for pulse distortion introduced by fatty layers in the human body.
-However, in this thesis adaptive beamforming will refer to data-dependent apodization (or weighting) derived from a given optimization criterion.
+However, in this thesis adaptive beamforming will refer to data-dependent weighting derived from a given optimization criterion.
 The optimization criterion which we focus on in this thesis is the minimum variance distortionless response criterion, which is also known as Capon's method \cite{Capon1969}.
-A beamformer which uses weights derived using Capon's method will in this thesis mainly be referred to as the \textit{Capon beamformer} \cite{Vignon2008}.
-However, it is also known as the minimum variance  (\nom{MV}{Minimum variance beamformer (Synonym for Capon)}) beamformer \cite{Synnevag2007}, the minimum variance distortionless respons beamformer (\nom{MVDR}{Minimum variance distortionless respons beamformer (Synonym for Capon)}) (\textbf{Paper I}), or the constrained adaptive beamformer \cite{Mann2002}.
+A beamformer that uses weights derived using Capon's method will in this thesis mainly be referred to as the \textit{Capon beamformer} \cite{Vignon2008}.
+However, it is also known as the minimum variance  (\nom{MV}{Minimum variance beamformer (Synonym for Capon)}) beamformer \cite{Synnevag2007}, the minimum variance distortionless response beamformer (\nom{MVDR}{Minimum variance distortionless response beamformer (Synonym for Capon)}) (\textbf{Paper I}), or the constrained adaptive beamformer \cite{Mann2002}.
 
@@
 The Capon beamformer produces a weight vector, as introduced in (\ref{eq:das_background}), based on the impinging signal and a constrained optimization problem.
-The problem is as follow \cite{Capon1969}:
+The problem is as follows \cite{Capon1969}:
 \begin{align}
-&\min_{\vec{w}} E\{|z[n]|^2\} \rightarrow \min_{\vec{w}} \vec{w}^H \mat{\hat{R}} \vec{w} \label{eq:capon_optimization_criteria_background} \\
-&\text{subjected to } \vec{w}^H\vec{a} = 1,
+&\min_{\vec{w}} E\{|z[n]|^2\} \approx \min_{\vec{w}} \vec{w}^H \mat{\hat{R}} \vec{w} \label{eq:capon_optimization_criteria_background} \\
+&\text{subject to } \vec{w}^H\vec{a} = 1,\notag
 \end{align}
@@
 Hence, the resulting weight vector minimizes the output power (the criterion) while maintaining unit gain in the steering direction $\vec{a}$ (distortionless response).
-The benefit of Capon beamforming is clearly seen from Fig\,\ref{fig:weights} which present a sensitivity map (an \textit{array beampattern}) for DAS beamforming with uniform weights and the Capon beamformer.
-In this simulated case an interfering source is located at 20\degree{} which for the DAS beamformer with uniform weights (\ref{fig:das_weights}) will leak into the beamformer output with its power attenuated with around $13$ dB.
-Since the Capon weights are data-dependent the method "knows" where the interfering source is located and is able to position a zero in the direction of it (Fig.\,\ref{fig:capon_weights}).
-Hence, the interfering source will be highly attenuated independent of its direction.
+The benefit of Capon beamforming is clearly seen from Fig\,\ref{fig:weights}, which presents a sensitivity map (an \textit{array beampattern}) for DAS beamforming with uniform weights and the Capon beamformer.
+In this simulated case, an interfering source is located at 20\degree{}, which for the DAS beamformer with uniform weights (\ref{fig:das_weights}) will leak into the beamformer output with its power attenuated by around $13$ dB.
+Since the Capon weights are data-dependent, the method "knows" where the interfering source is located and is able to position a zero in its direction (Fig.\,\ref{fig:capon_weights}).
+Hence, the interfering source will be highly attenuated independent of the direction from which it arrives.
 Another interesting case is when the interfering source is located within the mainlobe of the array beampattern.
 In theory a signal will be regarded by the Capon beamformer as an interfering source as long as the signal propagation vector does not match the steering vector ($\vec{a}$) exactly.
-If it is not possible to position a zero in the direction of the interferer the mainlobe will get shifted and scaled if this is whats needed in order to achieve minimum output power.
-Adaptive positioning of zeros and mainlobe shifting gives rice to the super-resolution capabilities of the Capon beamformer \cite{Synnevag2007}, and the positive implications of this effect \cite{Synnevag2009} is the main reason why this beamformer has been highly studied in the field of medical ultrasound imaging the last decade.
+If it is not possible to position a zero in the direction of the interferer the mainlobe will get shifted and scaled if this is what is needed in order to achieve minimum output power.
+Adaptive positioning of zeros and mainlobe shifting gives rise to the super-resolution capabilities of the Capon beamformer \cite{Synnevag2007}, and the positive implications of this effect \cite{Synnevag2009} is the main reason why this beamformer has been highly studied in the field of medical ultrasound imaging over the last decade.
 
 To "know" the direction of interfering sources the Capon beamformer has to estimate the statistics of the impinging wave field.
-This is done by estimating a sample covariance matrix, which for an active broadband system can be estimated in the following way \cite{Synnevag2009}:
+This is done by estimating a sample covariance matrix, which for an active broadband system can be done in the following way \cite{Synnevag2009}:
 \begin{align}
@@
 \end{align}
-where\nomi{$K$}{Temporal averaging parameter (Capon)} $N_K = 2K + 1$ \nomi{$N_K$}{Number of samples used for temporal averaging (Capon)} is the number of samples used for temporal smoothing,\nomi{$L$}{Subarray length (Capon)} $\vec{x}_l$ is a $L$ long subarray, and $N_L = M-L+1$\nomi{$N_L$}{Number of subarrays used for spatial averaging (Capon)} is the number of subarrays used for spatial averaging.
+where\nomi{$K$}{Temporal averaging parameter (Capon)} $N_K = 2K + 1$ \nomi{$N_K$}{Number of samples used for temporal averaging (Capon)} is the number of samples used for temporal smoothing,\nomi{$L$}{Subarray length (Capon)} $\vec{x}_l$ is a subarray of length $L$, and $N_L = M-L+1$\nomi{$N_L$}{Number of subarrays used for spatial averaging (Capon)} is the number of subarrays used for spatial averaging.
 Temporal averaging is used to obtain DAS-like speckle statistics in combination with large subarrays \cite{Synnevag2007a}, and subarray averaging is applied to reduce signal cancellation caused by coherent signals \cite{Reddy1987}.
@@
 \begin{align}
-\epsilon &= d*\text{trace}\{\mat{\breve{R}}\}/L\\
+\epsilon &= d\times\text{trace}\{\mat{\breve{R}}\}/L\\
 \mat{\hat{R}} &= \mat{\breve{R}} + \epsilon\mat{I}.
@@
 \end{align}
-Since $\vec{w}$ is time (or range) dependent one matrix has to be constructed and inverted for each data vector $\vec{x}[n]$ received by the system.
+Since $\vec{w}$ is time (or range) dependent, a matrix has to be constructed and inverted for each data vector $\vec{x}[n]$ received by the system.
 This is indeed computationally demanding, something which becomes clear when each step of the beamformer is examined closer.
-It is then found that constructing the covariance matrix has a computationally complexity of $O(L^2N_LN_K)$\nomi{O(.)}{Big O notation (algorithm complexity)}, inverting it has a computationally complexity of $O(L^3)$, and the final normalization and calculation of $\vec{w}$ is $O(L)$.
+It is then found that constructing the covariance matrix has a computational complexity of $O(L^2N_LN_K)$\nomi{O(.)}{Big O notation (algorithm complexity)}, inverting it has a computational complexity of $O(L^3)$, and the final normalization and calculation of $\vec{w}$ is $O(L)$.
 For $L=M/2$, a subarray size often used in the literature, the first two steps both have a complexity of $O(L^3)$.
@@
 Which for our example, with $L=M/2=64$, sums up to a required throughput of more than $1/2$ TFLOPS.
-Even though this is just 10\% of the peak throughput of modern GPUs remember that the latter are only theoretical numbers, and that the actual throughput is highly algorithm dependent.
+Even though this is just 10\% of the peak throughput of modern GPUs, remember that the latter are only theoretical numbers, and that the actual throughput is highly algorithm dependent.
 The required throughput for Capon beamforming might therefore still be hard to achieve in practice \cite[\textbf{Paper II}]{So2011}.
-As for DAS beamforming this number also scales with higher relative bandwidth, pulse frequency, and more array elements since the subarray size is selected proportional to the number of array elements.
+As for DAS beamforming, this number also scales with higher relative bandwidth, pulse frequency, and more array elements since the subarray size is typically proportional to the number of array elements.
 
-In the literature on medical ultrasound imaging we find few attempts on implementing a real-time Capon beamformer for medical ultrasound imaging.
+In the medical ultrasound imaging literature we find few attempts of implementing a Capon beamformer for real-time imaging.
 In addition to the papers in this thesis only Chen \textit{et al.} \cite{Chen2011, Chen, Chen2011a} are known to have explored GPU and FPGA implementations of the Capon beamformer for medical ultrasound imaging.
-The differences between our work and Chen \textit{et al.} is discussed in \textbf{Paper I}, \textbf{II}, and \textbf{VI}.
+The differences between our work and Chen \textit{et al.} is discussed in \textbf{Papers I}, \textbf{II}, and \textbf{VI}.
 
-To reduce the processing requirement of the Capon beamformer for medical ultrasound imaging several simplified version have been proposed in the literature \cite{Asl2012, Kim}.
+To reduce the processing requirement of the Capon beamformer for medical ultrasound imaging, several simplified version have been proposed in the literature \cite{Asl2012, Kim}.
 Among the most interesting proposals we find the low complexity adaptive (LCA) beamformer by Synnev\aa{}g \textit{et al.} \cite{Synnevag2011} and the beamspace (BS) Capon beamformer by Nilsen and Hafizovic \cite{Nilsen2009}.
-In medical ultrasound imaging beamspace typically refers to the polar grid that cardiac ultrasound data is located in prior to scan conversion.
-In combination with the Capon beamformer, beamspace refers to the wavenumber-space representation of the recorded signals, hence, the spatial Fourier transform of the element data:
+In medical ultrasound imaging beamspace typically refers to the polar grid that cardiac ultrasound data are located in prior to scan conversion.
+In combination with the Capon beamformer, beamspace refers to the wavenumber-space representation of the recorded signals, i.e., the spatial Fourier transform of the element data:
 \begin{align}
@@
 The idea behind BS-Capon beamforming is that spatial beams with little information can be removed from $\mat{\hat{R}}$ by removing rows in $B$.
-When investigating the complexity of the BS-Capon beamformer (\textbf{Paper II}) we find that it is dominated by the transformation of data into beamspace, $O(N_LLN_b)$, where \nom{$N_b$}{Number of selected beams in beamspace (BS-Capon)} is the number of selected beams.
+When investigating the complexity of the BS-Capon beamformer [\textbf{Paper II}] we find that it is dominated by the transformation of data into beamspace, $O(N_LLN_b)$, where \nom{$N_b$}{Number of selected beams in beamspace (BS-Capon)} is the number of selected beams.
 
@@
 A GPU implementation of the BS-Capon beamformer for real-time cardiac ultrasound imaging is presented in \textbf{Paper II}.
-The rationale behind using the GPU is that a highly parallel task like the Capon and BS-Capon beamformer should be offloaded to the processor made for parallel processing (heterogeneous computing).
-The CPU is then left to do managing and all the task which by tradition already occupies it.
+The rationale behind using the GPU is that in the heterogeneous computing paradigm a highly parallel task like the Capon and BS-Capon beamformer should be offloaded to the processor made for parallel processing.
+The CPU is then left to do managing and all the tasks it traditionally performs.
 It should also be possible to achieve higher processing throughput on a GPU than on a CPU for this kind of problem.
@@
 \subsection{Narrowband vs Broadband Capon beamforming}
-Both the LCA and BS-Capon beamformer is based upon a detailed investigation of how Capon beamfoming behaves when applied to medical ultrasound images.
-For the LCA beamformer, the windows selected by the Capon beamformer where investigated and a special set of predefined windows where defined.
-If one look closer at the windows selected for the LCA beamformer we see that half the set performs micro-steering of the mainlobe and the other half provides different mainlobe widths.
-The results from applying the LCA beamformer shows that the main benefit of applying Capon beamforming to medical ultrasound imaging is micro-steering of the mainlobe, and not attenuation of far off-axis backscatter.
+Both the LCA and BS-Capon beamformers are based upon a detailed investigation of how Capon beamfoming behaves when applied to medical ultrasound images.
+For the LCA beamformer, the windows selected by the Capon beamformer were investigated and a special set of predefined windows were defined.
+If one looks closer at the windows selected for the LCA beamformer we see that half the set performs micro-steering of the mainlobe and the other half provides different mainlobe widths.
+The results from applying the LCA beamformer show that the main benefit of applying Capon beamforming to medical ultrasound imaging is micro-steering of the mainlobe, and not attenuation of far off-axis backscatter.
 
 The Capon beamformer is derived for narrowband applications and has traditionally been applied to medical ultrasound imaging "as is".
-Medical ultrasound imaging is, however, a broadband imaging modality and one could think that a broadband Capon beamformer therefore where strictly needed \cite{Holfort2009}.
-The investigation performed in \cite{Nilsen2009} showed that with pre-delayed data, energy will be centered on-axis and as little as three near-axis beams is therefore adequate in order to achieve similar performance for BS-Capon and Capon beamforming.
+Medical ultrasound imaging is, however, a broadband imaging modality and one could think that a broadband Capon beamformer would therefore be needed \cite{Holfort2009}.
+The investigation performed in \cite{Nilsen2009} showed, that with pre-delayed data, energy will be centered on-axis and as little as three near-axis beams is therefore adequate in order to achieve similar performance for BS-Capon and Capon beamforming.
 The two-way beam will be wider for unfocused (plane wave) transmit beams \cite{Holfort2009, Austeng2011}, but steering and focusing on receive will still make the main amount of energy arrive on axis.
@@
 One approach used for cardiac ultrasound  is to extract slices corresponding to different long-axis and short-axis views of the cardiac chambers.
-Another way is to project the whole volume into a two dimensional image.
-The latter technique is known as volume rendering, and it present for the user the three dimensional structure of the heart in one two-dimensional image.
-Many different methods for volume rendering exists \cite{brodlie2001}, and in this thesis we will focus on ray casting \cite{Levoy1988}.
+Another way is to project the whole volume into a two-dimensional image.
+The latter technique is known as volume rendering, and it presents to the user the three dimensional structure of the heart in one two-dimensional image.
+Many different methods for volume rendering exist \cite{brodlie2001}, and in this thesis we will focus on ray casting \cite{Levoy1988}.
 
 The principle behind ray casting is depicted in Fig.\,\ref{fig:vr}.
-The projection of volumetric data is with this technique performed by casting a ray out from every pixel in the output image (Fig.\,\ref{fig:vr_rc}).
-Along each ray, samples are gathered from the volume and assigned a color and opacity.
-For ultrasound visualization the color intensity is dictated by the ultrasound backscatter intensity, and the actual color is often encoded with depth information in order to improve depth perception.
+The projection of volumetric data is performed by casting a ray out from every pixel in the output image (Fig.\,\ref{fig:vr_rc}).
+Along each ray, samples are gathered from the volume and assigned color and opacity.
+For ultrasound visualization, the color intensity is dictated by the ultrasound backscatter intensity, and the actual color is often encoded with depth information in order to improve depth perception.
 An example of a volume rendering of cardiac ultrasound data is presented in \textbf{Paper IV}.
-The opacity, however, is governed by a user controlled \textit{opacity transfer function} (\nom{OTF}{Opacity transfer function (volume rendering)}) as shown in Fig.\,\ref{fig:vr_otf}.
+The opacity, however, is governed by a user-controlled \textit{opacity transfer function} (\nom{OTF}{Opacity transfer function (volume rendering)}) as shown in Fig.\,\ref{fig:vr_otf}.
 For ultrasound visualization a monotonically increasing function is typically used, which means that below a certain threshold all samples will be rendered transparent (opacity = 0).

### 2014-04-12 ThesisJP/Background/background.tex
Finished review from CI.

@@
 
-This chapter gives the background information needed in order to understand the topics of this thesis and its papers.
-First, in Section \ref{sec:gpgpu}, the history and the principle of general-purpose computing on graphics processing units (\nom{GPGPU}{General-purpose computing on graphics processing units}) are reviewed, followed by an introduction to medical ultrasound imaging and adaptive beamforming in Section \ref{sec:ultrasound}.
-Finally, the concepts of volume rendering and ultrasound field simulations are presented in Section \ref{sec:volren} and \ref{sec:field}.
+This chapter gives the background information needed in order to understand the topics of this thesis and the associated papers.
+First, in Section \ref{sec:gpgpu}, the history and the principles of general-purpose computing on graphics processing units (\nom{GPGPU}{General-purpose computing on graphics processing units}) are reviewed, followed by an introduction to medical ultrasound imaging and adaptive beamforming in Section \ref{sec:ultrasound}.
+Finally, the concepts of volume rendering and ultrasound field simulations are presented in Sections \ref{sec:volren} and \ref{sec:field}.
 
 \section{General-purpose computing on graphics processing units}\label{sec:gpgpu}
-In a recent essay, Herb Sutter, a leading authority on software development, reviews the current state of the software and computer industry \cite{HerbSutter}.
+Herb Sutter, a leading authority on software development, reviews the current state of the software and computer industry in a recent essay \cite{HerbSutter}.
 The essay bears the name "Welcome to the Jungle" and is a sequel to the essay "The Free Lunch is Over" from 2004 \cite{HerbSuttera}.
@@
 Until then, all mainstream computer programs where typically running in a single thread on a single CPU \textit{processing core}, and a programmer could expect the software to annually gain performance without touching the code.
-This was the "free-lunch"-era as depicted in Fig.\,\ref{fig:jungle}.
+This was the "free-lunch" era as depicted in Fig.\,\ref{fig:jungle}.
 The increase in processing power came mostly from an ever-growing clock frequency.
-However, power consumption and heat generation where also growing at the same rate until the required cooling finally became to much in 2004.
-From that point, CPU manufactures had to pursue a different direction  than just increasing the clock frequency.
+However, power consumption and heat generation were also growing at the same rate until the required cooling finally became too much in 2004.
+From that point, CPU manufactures had to pursue different directions than just increasing the clock frequency.
 The immediate solution was to embed several cores in one CPU.
@@
 
-Around the same time as CPU manufacturers hit the frequency wall, people had started experimenting with programmable shaders which recently had been introduced in the field of graphics programming.
+Around the same time as CPU manufacturers hit the frequency wall, people had started experimenting with programmable shaders, which recently had been introduced in the field of graphics programming.
 The graphics processing pipeline had before this been a fixed function pipeline, where e.g.
-geometry and colors where fed to the graphics processing unit (GPU) in one end, and where shaded and rasterized sceneries were outputted in the other.
-What programmable shaders  introduced was the ability to replace some steps in the fixed-function pipeline with custom computations.
+geometry and colors were fed to the graphics processing unit (GPU) in one end, and where shaded and rasterized sceneries were outputted in the other.
+Programmable shaders introduced the ability to replace some steps in the fixed-function pipeline with custom computations.
 This made it possible to utilize the GPU for other tasks than just rendering graphics \cite{Seland2007}.
-The rationale behind this exploration was that GPUs were already highly available in consumer PCs, and hidden behind its graphics interface were processing powers an order of magnitude larger than what the CPU could provide (Fig.\,\ref{fig:cpu_vs_gpu}).
+The rationale behind this development was that GPUs were already highly available in consumer PCs, and hidden behind its graphics interface were processing powers an order of magnitude larger than what the CPU could provide (Fig.\,\ref{fig:cpu_vs_gpu}).
 Actually, at the time when CPU manufacturers hit the frequency wall, the GPU had already gone multi-core, first of all driven by an ever-increasing demand for more realistic computer games.
 
-The graphics pipeline exhibits minimal execution path branching, and data are typically used once.
-Designers of GPUs could therefore skip advanced feature found in most CPUs, like branch prediction and different levels of data caching.
+The original graphics pipeline involved no conditional branching, and data were typically used once \cite{Purcell2002}.
+Designers of GPUs could therefore skip advanced features found in most CPUs, like branch prediction and different levels of data caching.
 These measures reduced the silicon footprint of each core and made it possible to add multiple cores to the GPU before CPU designers where able to do the same.
@@
 For the first adopters of general-purpose computing on GPUs (GPGPU) it soon became evident that offloading all types of computations to the GPU was not always a good idea.
-In most cases it is important to balance the load equally between the CPU and GPU to unify the computational power of both processors.
-Balancing typically means that highly parallel and computationally intense tasks should be offloaded to the GPU while serialized and memory intensive work should stay on the CPU.
+In most cases it is important to balance the load equally between the CPU and GPU to get the most out of each processors.
+Balancing means that highly parallel and computationally intense tasks should be offloaded to the GPU while serialized and memory intensive work should stay on the CPU.
 This paradigm of utilizing specialized cores for solving specific problems is today known as \textit{heterogeneous computing}.
@@
 Together, multi-cores, heterogeneous cores, and cloud cores form the "jungle" that today's software engineers have to handle (Fig.\,\ref{fig:jungle}).
-In this thesis we focus on heterogeneous computing with one multi-core CPU and a discrete high performance GPU.
-This is a cheap and high performance system which is likely to be the specification of current and future ultrasound imaging system.
+In this thesis we focus on heterogeneous computing with a multi-core CPU and a discrete high performance GPU.
+This is a cheap, high-performance system which is likely to be the specification of current and future ultrasound imaging system.
 
@@
 \includegraphics[width=0.8\textwidth]{img/CPU-GPU.pdf}
-\caption{Schematic overview of CPU (Intel Haswell) and GPU (Nvidia Kepler) architecture.
+\caption{Schematic overview of CPU (Intel Haswell) and GPU (Nvidia Kepler) architectures.
 Note how more space is dedicated to compute cores on a GPU, and how more space is dedicated to control and caching on a CPU.}
@@
 The first SSE instruction set had registers of 128 bit which meant that four single-precision (32 bit) floating point values could be processed per clock cycle.
-Updates of the SSE instruction set have added new instructions and longer registers (SSE2, SSE3, SSE4, \nom{AVX}{Advanced vector extensions (instruction set)}, AVX2, and AVX-512).
+Updates of the SSE instruction set (SSE2, SSE3, SSE4, \nom{AVX}{Advanced vector extensions (instruction set)}, AVX2, and AVX-512) have added new instructions and longer registers.
 The newly introduced extensions, AVX and AVX2, add registers of 256 bit and fused multiply accumulate (\nom{FMA}{Fused multiply-accumulate (instruction)}) instructions  which means that sixteen single-precision floating point operations can be performed per clock cycle per core.
@@
 The latest CPU architecture from Intel, the Haswell architecture, has two processing ports per core supporting FMA-AVX2 instructions, and is therefore capable of performing 32 single-precision floating point operations per clock cycle per core (Fig.\,\ref{fig:cpu_gpu}).
-For an eight-core\footnote{An 8-core Haswell CPU is scheduled to launch during 2014.} CPU with a clock frequency of 3.8 GHz this sums to
+For an eight-core\footnote{An 8-core Haswell CPU is scheduled to launch during 2014.
+The numbers in Fig.\,\ref{fig:cpu_vs_gpu} for the last three generations of CPUs are all based on 8-core CPUs.} CPU with a clock frequency of 3.8 GHz this sums to
 \begin{equation}
-32*8*3.8 = 972.8\,\text{GFLOPS}.
+32 \times 8 \times 3.8 = 972.8\,\text{GFLOPS}.
 \end{equation}
@@
 \begin{equation}
-16*40*1.3 = 832\,\text{GFLOPS}
+16 \times 40 \times 1.3 = 832\,\text{GFLOPS}
 \end{equation}
@@
 
-Modern GPUs can also be interpreted as a unit consisting of multiple SIMD cores (Fig.\,\ref{fig:cpu_gpu}).
-The recent Kepler architecture by Nvidia can provide up to 15 SIMD cores\footnote{Nvidia likes to refer to the unit capable of processing one single precision floating point value as a core.
-In Nvidia's world, Kepler therefore has $15*192=2880$ cores.} capable of processing 192 single-precision floating point values per clock cycle.
+Modern GPUs can also be interpreted as units consisting of multiple SIMD cores (Fig.\,\ref{fig:cpu_gpu}).
+The recent Kepler architecture from Nvidia can provide up to 15 SIMD cores\footnote{Nvidia likes to refer to the unit capable of processing one single precision floating point value as a core.
+In Nvidia's world, Kepler therefore has $15\times192=2880$ cores.} capable of processing 192 single-precision floating point values per clock cycle per core.
 These GPUs are also capable of performing FMA instructions, effectively doubling the number of theoretical FLOPS.
@@
 \begin{equation}
-15*192*2*745 = 4291\,\text{GFLOPS}.
+15 \times 192 \times 2 \times 745 = 4291\,\text{GFLOPS}.
 \end{equation} 
@@
 
-\begin{figure}
+\begin{figure}[t!]
 \centering
@@
 \subsection{Programming a GPU}
-Early GPGPU programming required expert knowledge about computer graphics and the graphics pipeline.
+Early GPGPU programming required expert knowledge of computer graphics and the graphics pipeline.
 When GPGPU gained a broader interest and the GPU started to be used to compute problems that did not involve graphics, it quickly became evident that new programing languages where needed.
@@
 %Behind the scene, thread blocks are divided into smaller sets of threads called a warp (or wavefront in OpenCL).
-When a kernel is executed, multiple thread blocks are scheduled to each SIMD core where they stay until all instructions in the kernel are finished for all threads in the block.
-Thus,  in order to formulate a given problem as a kernel function, the global memory input to each thread must not depend on the output from other blocks.
-Hence, it must be possible for the GPU to schedule blocks in any order.
+When a kernel is executed, multiple arbitrary thread blocks are scheduled to each SIMD core where they stay until all instructions in the kernel are finished for all threads in all the blocks.
+This step is then repeated until all blocks have been processed.
+Thus,  in order to formulate a given problem as a kernel function, it must be possible for the GPU to schedule the resulting thread blocks in any order.
+As an example, the global memory input to each thread can therefore not depend on the output from other blocks.
 
@@
 It is important to understand that not all algorithms will benefit from wider SIMD registers.
-As the width increase, the only algorithms that will benefit are those which can be divided into enough independent calculations.
-Also, to achieve the peak performance, the implementation must perform one FMA operation on each core per clock cycle, and it must be possible for the GPU to hide all memory latency by scheduling threads which are not stalled by a memory fetch.
+As the width increase, the only algorithms that will benefit are those which can be divided into a sufficiently large number of independent calculations.
+Also, to achieve peak performance, the implementation must perform one FMA operation on each core per clock cycle, and it must be possible for the GPU to hide all memory latency by scheduling threads which are not stalled by a memory fetch.
 All of this is something which is impossible to achieve in practical applications, and therefore one will always end up with a throughput which is a fraction of the peak throughput.
@@
 The axial resolution is inversely proportional to the pulse length, and is typically better than the lateral resolution.
-The geometry of cardiac ultrasound imaging is depicted in Fig.\,\ref{fig:us_grid}.
-A line in an ultrasound image is generated by firing an ultrasound pulse or beam into the body and sample the received echo.
+An overall view of the cardiac ultrasound imaging geometry is depicted in Fig.\,\ref{fig:us_grid}.
+A line in the depicted ultrasound image is generated by firing an ultrasound pulse or beam into the body and sample the received echo.
 We will refer to this amplitude trace as an \textit{image line}.
@@
 \end{align}
-Here, \nom{$M$}{Number of array elements} is the number of elements in the array, $x_m$ is the per-element data, \nom{$\Delta_m$}{Per-element focusing and steering delays} is the per-element focusing and steering delay, \nom{$\vec{w}$}{Array weight vector (apodization)} is the array \textit{weight vector} or \textit{apodization}, $\vec{x}$ is a pre-delayed vector of element data, and $z[n]$ is the output image line in the direction given by $\Delta_m[n]$.
-From now on we will omit the dependency on $\Delta_m[n]$ and $x_m$ will refer to delayed element data.
+Here, \nom{$M$}{Number of array elements} is the number of elements in the array, $x_m$ is the per-element data, \nom{$\Delta_m$}{Per-element focusing and steering delays} is the per-element focusing and steering delay, \nom{$\vec{w}$}{Array weight vector (apodization)} is the array \textit{weight vector} or \textit{apodization}, $\vec{x}$ is a pre-delayed vector of element data, and $z[n]$ is the output image line in the direction given by $\{\Delta_m[n]\}$.
+From now on we will omit the dependency on $\{\Delta_m[n]\}$ and $x_m$ will refer to delayed element data.
 A schematic overview of delay-and-sum array beamforming is depicted in Fig.\,\ref{fig:das_background}.
@@
 \end{align}
-\nomi{$N$}{Number of samples in range (or time)}where \nom{$B_w$}{Relative bandwidth} is the relative bandwidth in percent of \nom{$f_c$}{Center frequency of ultrasound pulse}, where $f_c$ is the center frequency of the transmitted pulse.
+\nomi{$N$}{Number of samples in range (or time)}where $f_c$ is the center frequency of the transmitted pulse, and where \nom{$B_w$}{Relative bandwidth} is the relative bandwidth in percent of \nom{$f_c$}{Center frequency of ultrasound pulse}.
 Note that after this process the ultrasound data will be complex values.
@@
 The largest players in the medical ultrasound business, General Electric and Philips, are also well aware of where the trend is going \cite{Thomenius2012} \cite{Metz2011}.
+
 Software beamforming opens up for cost-efficient and rapid prototyping of advanced processing of ultrasound data.
@@
 However, in this thesis adaptive beamforming will refer to data-dependent weighting derived from a given optimization criterion.
-The optimization criterion which we focus on in this thesis is the minimum variance distortionless response criterion, which is also known as Capon's method \cite{Capon1969}.
-A beamformer that uses weights derived using Capon's method will in this thesis mainly be referred to as the \textit{Capon beamformer} \cite{Vignon2008}.
-However, it is also known as the minimum variance  (\nom{MV}{Minimum variance beamformer (Synonym for Capon)}) beamformer \cite{Synnevag2007}, the minimum variance distortionless response beamformer (\nom{MVDR}{Minimum variance distortionless response beamformer (Synonym for Capon)}) (\textbf{Paper I}), or the constrained adaptive beamformer \cite{Mann2002}.
+The optimization criterion that we focus on in this thesis is the minimum variance distortionless response criterion, which is also known as Capon's method \cite{Capon1969}.
+A beamformer that uses weights derived with Capon's method will in this thesis mainly be referred to as the \textit{Capon beamformer} \cite{Vignon2008}.
+However, it is also known as the minimum variance  (\nom{MV}{Minimum variance beamformer (Synonym for Capon)}) beamformer \cite{Synnevag2007}, the minimum variance distortionless response beamformer (\nom{MVDR}{Minimum variance distortionless response beamformer (Synonym for Capon)}) [\textbf{Paper I}], or the constrained adaptive beamformer \cite{Mann2002}.
 
@@
 In theory a signal will be regarded by the Capon beamformer as an interfering source as long as the signal propagation vector does not match the steering vector ($\vec{a}$) exactly.
-If it is not possible to position a zero in the direction of the interferer the mainlobe will get shifted and scaled if this is what is needed in order to achieve minimum output power.
-Adaptive positioning of zeros and mainlobe shifting gives rise to the super-resolution capabilities of the Capon beamformer \cite{Synnevag2007}, and the positive implications of this effect \cite{Synnevag2009} is the main reason why this beamformer has been highly studied in the field of medical ultrasound imaging over the last decade.
+Positioning of a zero in the direction of an interferer within the mainlobe will lead to a shifted and scaled mainlobe if this is what is needed in order to achieve minimum output power.
+Adaptive positioning of zeros and mainlobe shifting are the source to the super-resolution capabilities of the Capon beamformer \cite{Synnevag2007}, and the positive implications of this effect \cite{Synnevag2009} is the main reason why this beamformer has been highly studied in the field of medical ultrasound imaging over the last decade.
 
@@
 Temporal averaging is used to obtain DAS-like speckle statistics in combination with large subarrays \cite{Synnevag2007a}, and subarray averaging is applied to reduce signal cancellation caused by coherent signals \cite{Reddy1987}.
-The number of temporal samples is usually selected relative to the pulse length.
+The number of temporal samples is selected relative to the pulse length.
 To ensure robustness against model errors and numerical stability the matrix is further loaded with a diagonal factor,
@@
 
-The number of FLOPS required for BS-Capon beamforming of real-time cardiac ultrasound is then found to be:
+The number of FLOPS required for BS-Capon beamforming of real-time cardiac ultrasound is then found to be around:
 \begin{align}
@@
 This is a big challenge for cardiac ultrasound since the backscattered signal level from tissue varies a lot throughout the volume.
-The per-volume sample value distributions for tissue and noise therefore always overlap to some degree.
+The sample value distributions for tissue and noise therefore always overlap to some degree.
 With a global OTF it is therefore impossible to find a threshold that separates tissue from noise.
@@
 This regional adaptive OTF is based on estimated statistics of the clutter noise inside the left ventricle (the blood pool).
-The added complexity of this technique caused by the statistical estimation which is basically a second volume rendering (mean and variance projection) that has to be performed before the rendered image is produced.
+The added complexity of this technique is caused by the statistical estimation, which is basically a second volume rendering (mean and variance projection) that has to be performed before the rendered image is produced.
 
@@
 A simulator for ultrasound probes typically works by approximating the probe surface with discrete points or patches.
-In software it is then possible to transmit a pulse from a given probe and measure its response in space and time, know as the \textit{spatial impulse response}.
+In software it is then possible to transmit a pulse from a given probe and measure its response in space and time.
 Figure \ref{fig:hos} shows a snapshot of a simulated pressure field from an unfocused linear array emitting a continuous wave.
@@
 All simulations in both \textbf{Paper II} and \textbf{III} are produced with this simulator.
-The Field II simulator works by dividing each element into mathematical elements, and all field points needs to be in the far-field of these small elements\footnote{$l\gg\frac{w^2}{4\lambda}$, where $w$ is the mathematical element width, $l$ is the distance to the field point, and $\lambda$ is the wavelength of the transmitted pulse.
+The Field II simulator applies a far-field approximation where each element is divided into mathematical elements.
+All field points needs to be in the far-field of these small elements\footnote{$l\gg\frac{w^2}{4\lambda}$, where $w$ is the mathematical element width, $l$ is the distance to the field point, and $\lambda$ is the wavelength of the transmitted pulse.
 For $w=\lambda/2$ the approximation is valid for $l>\lambda/16$.}.
-It is then possible to approximate the spatial impulse response by summing a trapezoid impulse response function originating from each mathematical element \cite{Jensen1992} in all field points.
+It is then possible to approximate the spatial impulse response for a given probe by summing a trapezoid impulse response function originating from each mathematical element in all field points \cite{Jensen1992}.
 A rough estimate of the number of required FLOPS for calculating the spatial impulse respons as it is done in Field II is:
@@
 \end{align}
-where $f_s$ is the sampling frequency, $w$ is the width of the mathematical elements, $f_s\frac{w}{c}$ is the maximal number of samples in each trapezoid function, $N_{\text{elements}}$ is the number of mathematical elements, and $N_{\text{field points}}$ is the number of field points.
+where $f_s$ is the sampling frequency, $w$ is the width of the mathematical elements, $f_s\frac{w}{c}$ is the maximal number of samples in each trapezoid function, $N_{\text{elements}}$ is the total number of mathematical elements, and $N_{\text{field points}}$ is the number of field points.
 The required number of FLOPS for a simulation at $100$ MHz with $500$, $\lambda/2$ wide, mathematical elements across a field of $0.5$ million points is $6$ GFLOPS in order for it to complete after one second.
 If we want to simulate an ultrasound image the spatial impulse response on both transmit and recieve have to be calculated and convolved in time.
-The respons will also be different for each A-mode image.
-Hence, more than 800 GFLOPS is required in order to complete a Field II simulation with the given example values of an image consisting of 70 A-mode images in one second.
-However, Jensen and Nikolov have shown ways of reducing the computational load of Field II simulations \cite{Jensen2000}.
-Other simulators have also been introduced that are faster but less exact than Field II \cite{Hergum2009}.
+The respons will also be different for each image line.
+Hence, more than 800 GFLOPS is required in order to complete a Field II simulation with the given example values of an image consisting of 70 image lines in one second.
+%However, Jensen and Nikolov have shown ways of reducing the computational load of Field II simulations \cite{Jensen2000}.
+Other simulators have also been introduced which are faster but less exact than Field II \cite{Hergum2009}.
 
@@
 \end{align}
-Therefore, to simulate a field of one million points with several thousand source points at 50 FPS, more than 100 GFLOPS is required.
+To simulate a field of one million points with several thousand source points at 50 FPS, more than 100 GFLOPS is required.
 Note that this simulator evaluates a cosine and sine function for each calculation, something which makes it even harder to achieve the required throughput, since the GPU has less silicon devoted to such tasks (SFUs).
 
-The above examples shows that realistic and dense ultrasound simulations are computationally demanding.
-On the other hand we find that the levels of required processing is within the theoretical limits of modern CPUs and GPUs.
-Since ultrasound field simulations are quite easy to accelerate in parallel it should be possible, even if a theorethical throughput is almoste impossible to achieve in practise, to add increased interactivity to ultrasound simulations tools by utilizing parallel programing and SIMD instructions.
-In addition to \textbf{Paper V} threre are several simulators which have been presented the last years that does acactly this by laveraging the powers of GPUs \cite{Gjerald2012, Reichl2009, Hlawitschka2011}.
+The above examples show that realistic and dense ultrasound simulations are computationally demanding.
+On the other hand we find that the required amount of processing is within the theoretical limits of modern CPUs and GPUs.
+Since ultrasound field simulations are quite easy to accelerate in parallel it should be possible to add increased interactivity to ultrasound simulations tools by utilizing parallel programing and SIMD instructions.
+In addition to \textbf{Paper V} there are several simulators that have been presented the last years, which does exactly this by leveraging the powers of GPUs \cite{Gjerald2012, Reichl2009, Hlawitschka2011}.
 
 \section{Concluding remarks}
-All the presented problems in this chapter has in common that they require a processing throughput from ten to several hundred GFLOPS.
-Comparing this to the typical clock frequency of modern CPUs and GPUs (1.5 - 3 GHz) we see that GFLOPS problems requires parallel execution in order to achieve real-time performance.
+All the presented problems in this chapter have in common that they require a processing throughput from ten to several hundred GFLOPS.
+Comparing this to the typical clock frequency of modern CPUs and GPUs  at 3 and 1.5 GHz we see that GFLOPS problems require parallel execution in order to achieve real-time performance.
+With CPU threads it is only possible to achieve 24 GFLOPS in peak performance on a eight-core CPU.
+To reach a peak performance above 100 GFLOPS one has to use SIMD instructions on a CPU or SIMT instructions on a GPU.
+It also must be possible to divide the problem into a large number of independent calculations.
 

### 2014-04-12 ThesisJP/Background/background.tex
Started on comments from Ines on the introduction section.

@@
 The latter technique is known as volume rendering, and it presents to the user the three dimensional structure of the heart in one two-dimensional image.
-Many different methods for volume rendering exist \cite{brodlie2001}, and in this thesis we will focus on ray casting \cite{Levoy1988}.
+Many different methods for volume rendering exist \cite{brodlie2001}, but in this thesis we will focus on ray casting \cite{Levoy1988}.
 
@@
 Hence, the iteration can stop at $i=k$ without loosing visible samples.
-This optimization technique is known as \textit{early-ray-termination}, and will significantly speed up our rendering algorithm if an opaque material is rendered.
+This optimization technique is known as \textit{early-ray-termination}, and will significantly speed up the rendering algorithm if an opaque material is rendered.
 
@@
 The Field II simulator applies a far-field approximation where each element is divided into mathematical elements.
-All field points needs to be in the far-field of these small elements\footnote{$l\gg\frac{w^2}{4\lambda}$, where $w$ is the mathematical element width, $l$ is the distance to the field point, and $\lambda$ is the wavelength of the transmitted pulse.
+All field points then needs to be in the far-field of these small elements\footnote{$l\gg\frac{w^2}{4\lambda}$, where $w$ is the mathematical element width, $l$ is the distance to the field point, and $\lambda$ is the wavelength of the transmitted pulse.
 For $w=\lambda/2$ the approximation is valid for $l>\lambda/16$.}.
 It is then possible to approximate the spatial impulse response for a given probe by summing a trapezoid impulse response function originating from each mathematical element in all field points \cite{Jensen1992}.
-A rough estimate of the number of required FLOPS for calculating the spatial impulse respons as it is done in Field II is:
+A rough estimate of the number of required FLOPS for calculating the spatial impulse response as it is done in Field II is:
 \begin{align}
@@
 The required number of FLOPS for a simulation at $100$ MHz with $500$, $\lambda/2$ wide, mathematical elements across a field of $0.5$ million points is $6$ GFLOPS in order for it to complete after one second.
-If we want to simulate an ultrasound image the spatial impulse response on both transmit and recieve have to be calculated and convolved in time.
-The respons will also be different for each image line.
-Hence, more than 800 GFLOPS is required in order to complete a Field II simulation with the given example values of an image consisting of 70 image lines in one second.
+If we want to simulate an ultrasound image the spatial impulse response on both transmit and receive have to be calculated and convolved in time.
+The response will also be different for each image line.
+Hence, more than 800 GFLOPS is required in order to complete a Field II simulation in one second with the given example values and an image consisting of 70 image lines.
 %However, Jensen and Nikolov have shown ways of reducing the computational load of Field II simulations \cite{Jensen2000}.
@@
 Since ultrasound field simulations are quite easy to accelerate in parallel it should be possible to add increased interactivity to ultrasound simulations tools by utilizing parallel programing and SIMD instructions.
-In addition to \textbf{Paper V} there are several simulators that have been presented the last years, which does exactly this by leveraging the powers of GPUs \cite{Gjerald2012, Reichl2009, Hlawitschka2011}.
+In addition to \textbf{Paper V}, several simulators have recently accomplished this by leveraging the powers of GPUs \cite{Gjerald2012, Reichl2009, Hlawitschka2011}.
 
 \section{Concluding remarks}
-All the presented problems in this chapter have in common that they require a processing throughput from ten to several hundred GFLOPS.
-Comparing this to the typical clock frequency of modern CPUs and GPUs  at 3 and 1.5 GHz we see that GFLOPS problems require parallel execution in order to achieve real-time performance.
-With CPU threads it is only possible to achieve 24 GFLOPS in peak performance on a eight-core CPU.
-To reach a peak performance above 100 GFLOPS one has to use SIMD instructions on a CPU or SIMT instructions on a GPU.
+All the presented problems in this chapter, except DAS beamforming, have in common that they require a processing throughput of more than ten GFLOPS.
+Comparing this to the typical clock frequency of modern CPUs at 3 GHz we see that GFLOPS problems will require parallel execution in order to achieve real-time performance.
+With CPU threads the theoretical peak performance on an eight-core CPU increases to 24 GFLOPS.
+However, to reach a peak performance above 100 GFLOPS, SIMD instructions on a CPU or SIMT instructions on a GPU have to be used.
 It also must be possible to divide the problem into a large number of independent calculations.

### 2014-04-12 ThesisJP/Introduction/introduction.tex
Started on comments from Ines on the introduction section.

@@
 The latest graphics cards\footnote{As of January 2014.} from both \nom{Nvidia}{Graphics card producer} and \nom{AMD}{Graphics card producer} provide more than 5 T\nom{FLOPS}{Floating point operations per second} single precision and 1.5 TFLOPS double precision computations.
-For double precision this is more than the worlds most powerful supercomputer could provide in 1998, and that was a room full of computers.
+For double precision, this is more than the world's most powerful supercomputer could provide in 1998, and that was a room full of computers.
 
@@
 Algorithms which previously had to be implemented in hardware for performance reasons can now be implemented in software.
-And algorithms, either too complicated or costly to implement in hardware, and where a software implementation was thought to be too computationally heavy for real-time use, becomes realizable.
+%Algorithms, either too complicated or costly to implement in hardware, and where a software implementation was thought to be too computationally heavy for real-time use, become realizable. 
 It is clear that high-performance programmable processors, like the GPU, already have and will continue to make future ultrasound imaging systems more flexible, cheaper to produce, and equipped with even more cutting-edge processing.
@@
 When researchers are exploring new algorithms for ultrasound imaging it is therefore important to keep the architecture of parallel accelerators in mind.
-If a new complex algorithm is suppose to run in real time it needs to fit the programmable and parallel pipeline of modern ultrasound scanners.
+If a new complex algorithm is supposed to run in real time it needs to fit the programmable and parallel pipeline of modern ultrasound scanners.
 
@@
 
-Main focus have been the Capon beamformer, and the problem of making this computationally intense algorithm available for real-time ultrasound imaging (\textbf{Paper\,I} and \textbf{II}).
-However, two additional methods were also explored.
-The first is adaptive visualization of cardiac ultrasound volumes (\textbf{Paper\,IV}) and the latter is simulation of dense ultrasound pressure fields (\textbf{Paper\,V}).
-Both share the property of being computationally complex, but on the other hand they consist of many independent computations which makes them perfectly suited for parallel GPU acceleration.
+The main focus has been the Capon beamformer, and the problem of making this computationally intense algorithm available for real-time ultrasound imaging [\textbf{Paper\,I} and \textbf{II}].
+However, two additional methods have also been explored.
+The first one being adaptive visualization of cardiac ultrasound volumes [\textbf{Paper\,IV}], and the latter being simulation of dense ultrasound pressure fields [\textbf{Paper\,V}].
+Both share the property of being computationally complex, but on the other hand they consist of many independent computations, which makes them perfectly suited for parallel GPU acceleration.
 
@@
 {\it IEEE Transactions on Oceanic Engineering, accepted for publication.}\\\\
-The first paper describes in details how the Capon beamformer was mapped to GPU architecture.
+The first paper describes in detail how the Capon beamformer is mapped to the GPU architecture.
 Even though the paper is written within the field of active \nom{sonar}{SOund Navigation And Ranging (usually under water)} imaging, the depicted implementation is applicable to a vast range of active imaging systems.
@@
 
-The estimation of the spatial covariance matrix receives special attention in this study.
+The estimation of the spatial covariance matrix receives a special attention in this study.
 In previous literature on Capon beamforming the matrix inversion has always been regarded as the most computationally complex step.
-In this paper we show that estimation of the sample covariance matrix is actually the most complex part when spatial and temporal smoothing with common parameters are included.
-It is then shown how the arithmetic complexity of this estimation process can be reduced from cubic to square.
+In this paper we show that estimation of the sample covariance matrix is the most complex part when spatial and temporal smoothing of the estimate are included.
+It is then shown how the arithmetic complexity of this estimation procedure can be reduced from cubic to square.
 Finally, an in-depth analysis of the arithmetic throughput on multiple platforms is given.
@@
 Achieving this will facilitate further study of the method's \textit{in vivo} performance.
-This is something which has been stated as further work in many publications on Capon beamforming for medical ultrasound imaging the past decade.
+This is something which has been suggested as further work in many publications on Capon beamforming for medical ultrasound imaging the past decade.
 In \textbf{Paper\,I} arrays with no more than 32 elements were investigated.
-A linear array for cardiac ultrasound imaging typically has 64 or more elements.
+A linear array for cardiac ultrasound imaging has typically 64 or more elements.
 In \textbf{Paper\,VI}, which is summarized in \textbf{Paper\,II}, it is shown that our implementation from  \textbf{Paper\,I} does not reach the level of throughput required for real-time Capon beamforming of cardiac ultrasound imaging.
@@
 
-In this paper, to reduce the matrix size, it is taken advantage of the beamspace version of the Capon beamformer (\nom{BS-Capon}{Beamspace capon beamforming}), which is implemented on the GPU.
+In this paper, to reduce the matrix size, we take the advantage of the beamspace version of the Capon beamformer (\nom{BS-Capon}{Beamspace capon beamforming}), which is implemented on the GPU.
 For parameters previously derived to give similar performance for BS-Capon and element-space Capon (\nom{ES-Capon}{Element space capon beamforming}), we show that real-time BS-Capon beamforming is feasible for cardiac ultrasound imaging.
 The reported processing throughput is able to keep up with the acquisition frame rate in a typical cardiac ultrasound imaging system equipped with 64 and 96 element linear arrays.
-\textbf{Paper\,II} and \textbf{VI} also presents, for the first time, videos where the ES-Capon and BS-Capon beamformer have been applied on loops of simulated and \textit{in vivo} medical ultrasound images.
+\textbf{Paper\,II} and \textbf{VI} also present for the first time videos where the ES-Capon and BS-Capon beamformer have been applied on loops of simulated and \textit{in vivo} medical ultrasound images.
 
@@
 As mentioned, shift-invariant imaging is essential if the method is ever to be adopted.
-The paper was written based on observations of aliasing artifacts when imaging bright point scatterers \textit{in vitro} and in simulations.
-The point scatterers were observed to twinkle like stars in the sky.
+The paper is based on an observation of aliasing artifacts when the Capon beamformer was used to image bright point scatterers \textit{in vitro} and in simulations.
+The intensity of the point scatterers were observed to oscillate as they or the probe moved.
 In earlier work on Capon beamforming for medical ultrasound imaging the real-time nature of ultrasound imaging has often been ignored, and a large degree of oversampling on transmit might have been applied without being clearly stated.
-Oversampling on transmit is not an option if the imaging modality is used to image rapidly varying objects.
+Oversampling on transmit will reduce the acquisition frame rate, and is therefore not an option if the imaging system is used to image rapidly varying objects.
 Smooth presentation of probe motion is equally important since real-time user interaction is one of the key selling points of ultrasound imaging.
-The frame rate should therefore be higher than 10 frames per second.
 
-Further it has been investigated how the Capon beamforming achieves its super resolution, and how this makes it highly sensitive to differences between the assumed steering vector and the signal propagation vector.
-Earlier work, focusing on single images, has typically referred to this effect as speed-of-sound errors, however, it also comes naturally into play when objects moves from frame to frame and the beam density is insufficient.
-With a narrowband and farfield model it is shown that the Capon beamformer suffers from beam-to-beam gain variations as large as 27 dB for parameters previously used to obtain high resolution medical ultrasound images.
+Further, it has been investigated how the Capon beamforming achieves its super-resolution, and how this makes it highly sensitive to differences between the assumed steering vector and the signal propagation vector.
+Earlier work, focusing on single images, has typically referred to this effect as speed-of-sound errors, however, it also comes naturally into play when objects moves across an imaging sector which has insufficient lateral sampling.
+With a narrowband and farfield model it is shown that the Capon beamformer suffers from lateral gain variations as large as 27 dB for parameters previously used to obtain high resolution medical ultrasound images.
 To obtain the same lateral gain variations as with \nom{DAS}{Delay-and-sum beamforming} beamforming, 25 times oversampling is shown to be required.
@@
 {\it Proc. SPIE Medical Imaging 2012, vol. 8320, pp. 832008.}\\\\
-The fourth paper propose an adaptive volume rendering method based on statistics derived from the blood pool of the left ventricle.
+The fourth paper proposes an adaptive volume rendering method based on statistics derived from the blood pool of the left ventricle.
 Noise located in the blood pool tends to occlude cardiac tissue when rendered in 3D, and is often impossible to remove by adjusting global rendering parameters.
-This because ultrasound signals have high variability within blood as well as high variability within tissue, even when heavily smoothed.
+This because ultrasound signals have high variability within blood, as well as high variability within tissue, even when heavily smoothed.
 Delineation of the blood pool is done by a state estimation algorithm, capable of tracking the endocardium in real time.
@@
 {\it Proc. IEEE Ultrasonics Symposium 2012, pp. 1643-1646.}\\\\
-The fifth paper presents an interactive simulations tool capable of simulating dense pressure fields from ultrasound arrays in real time.
+The fifth paper presents an interactive simulation tool capable of simulating dense pressure fields from ultrasound arrays in real time.
 Simulation tools are heavily used by researchers in order to test out new algorithms and to deduce the performance of new array designs prior to manufacturing.
@@
 The simulator is linked with both a Paint-like interface for interactive drawing of arrays, and a Matlab interface for precise scripting of array configurations.
-The main contributions in this paper is the increased performance of the GPU implementation compared with a CPU and Matlab\footnote{The Ultrasim toolbox \todo{add citation}} version.
-Second, the paint-like interface provides a neat way of demonstrating array beamforming principle in real time.
+The main contribution in this paper is the increased performance of the GPU implementation compared with a CPU and Matlab\footnote{The Ultrasim toolbox.} version.
+Second, the Paint-like interface provides a neat way of demonstrating array beamforming principle in real time.
 We believe this to be of great value both for array designers and teachers.
@@
 \item a GPU implementation of the beamspace Capon beamformer.
-\item that real-time beamspace Capon beamforming is achieved for cardiac ultrasound imaging.
-\item the first investigation of Capon beamforming applied on multiple frames in medical ultrasound imaging (simulated, \textit{in vitro}, and \textit{in vivo}).
+\item a real-time beamspace Capon beamforming is achieved for cardiac ultrasound imaging.
+\item a first-time investigation of Capon beamforming applied on multiple frames in medical ultrasound imaging (simulated, \textit{in vitro}, and \textit{in vivo}).
 \item an investigation of shift-invariance when the Capon beamformer is applied on consecutive frames.
@@
 This result shows how hard it is to reach the theoretical level of throughput, and that how close one can get is highly algorithm depended.
-It might be ways to improve on these numbers, but both the solver and covariance estimation step consist of several points where fine granular synchronization and serialized instructions are needed.
+There might be ways to improve on these numbers, but both the solver and covariance estimation step consist of several points where fine granular synchronization and serialized instructions are needed.
 This will restrict even the most fine tuned implementation from reaching the maximum throughput of the GPU.

### 2014-04-12 ThesisJP/Introduction/introduction.tex
Started on Introduction 2.5.

@@
 \section{Discussion and future work}
-Even with all our effort in \textbf{Paper\,I} we were only able to achieve a fraction of the maximum throughput of the target GPU.
-This result shows how hard it is to reach the theoretical level of throughput, and that how close one can get is highly algorithm depended.
-There might be ways to improve on these numbers, but both the solver and covariance estimation step consist of several points where fine granular synchronization and serialized instructions are needed.
-This will restrict even the most fine tuned implementation from reaching the maximum throughput of the GPU.
-An interesting observation can be found in Fig.\,10 of \textbf{Paper\,I}.
-Here we see how the achieved throughput of the covariance estimation kernel is more than three times higher than the theoretical throughput of a modern, high-end, \nom{CPU}{Central processing unit}.
+\subsection{Paper\,I}
+Even with all our effort in \textbf{Paper\,I} we were only able to achieve a fraction of the peak theoretic throughput of the target GPU.
+This shows how hard it is to reach the theoretical level of throughput, and that how close one can get is highly algorithm depended.
+There might be ways to improve on these numbers, but both the matrix equation solver and the covariance matrix estimation step consist of several points where fine granular synchronization and serialized instructions are needed.
+These points will restrict even the most fine tuned implementation from reaching the maximum throughput of the GPU.
+The work in \textbf{Paper I} also showed that the covariance estimation kernel was memory bound, hence there is not enough instructions to hide all the memory latency.
+Further improvements through improved memory management could therefore be possible.
+
+An interesting observation in \textbf{Paper\,I} can be found in Fig.\,10 \todo{Direct ref}.
+Here we see how the achieved throughput of the covariance estimation kernel is more than three times higher than the theoretical throughput of a high-end \nom{CPU}{Central processing unit} (The target CPU is from 2010).
+Note that the target GPU, the Nvidia Quadro 6000, is also from 2010.
 This clearly shows the benefit of modern GPU computing.
-Solving was performed by a batched Gauss Jordan solver implemented by Nvidia.
-In our work we focused on the covariance estimation step, and did not try to beat Nvidia at home.
-Nevertheless, with the measured throughput in mind, a natural next step would be to analyze the current solver, and figure out if any further optimization is possible.
-Since the sample covariance matrix is hermitian, a solver based on Cholesky decomposition might also help to improve the solver's throughput.
-\\\\
+
+The matrix equation found in the Capon beamformer is solved using a batched Gauss Jordan solver implemented by Nvidia.
+In our work we focused on the covariance matrix estimation step, and did not try to beat Nvidia at home.
+Nevertheless, \textbf{Paper\,I} showed that the solver had significantly lower throughput than the covariance matrix estimation kernel.
+A natural next step would therefore be to analyze the current solver, and figure out if any further optimization is possible.
+Since the sample covariance matrix is hermitian, a solver based on Cholesky decomposition should also help to improve the throughput.
+
+\subsection{Paper\,II}
 In \textbf{Paper\,II} we focused on achieving real time Capon processing for cardiac ultrasound by implementing the BS-Capon beamformer on a GPU.
@@
 The work in \textbf{Paper\,II} made it clear that interesting findings are done when users are able to watch the temporal behavior of an algorithm. This led us to the work of \textbf{Paper\,III}.
-\\\\
+
+\subsection{Paper\,III}
 In \textbf{Paper\,III} we investigate the inherent self nulling involved with Capon beamforming, and how this effect comes naturally into play when the beamformer is subjected to linear steering vector errors caused by lateral undersampling.
@@
 Hence, more aggressive parameters might be used if the image produced is intended as input to another algorithm and not for the human eye.
-\\\\
+
+\subsection{Paper\,IV}
 %Objections to \textbf{Paper\,IV} includes its inherent chicken-or-egg dilemma.
@@
 If this will be different than just looking at neighboring pixels in an edge-preserving image filter is to be seen.
-\\\\
+
+\subsection{Paper\,V}
 The fifth paper stands out from the other papers by not dealing with an adaptive processing technique.
@@
 It also has close to a "whopping" 1000 views on youtube.
-\\\\
+\\\\\\
 Finally it is worth mentioning that the code for both the Capon beamformer\footnote{github.com/jpaasen/cos} and the ultrasound field simulator\footnote{github.com/jpaasen/hos} have been made open source and is available on github.

### 2014-04-13 ThesisJP/Background/background.tex
Finished acknowledgments.

@@
 
-\subsection{Narrowband vs Broadband Capon beamforming}
+\subsection{Narrowband or broadband Capon beamforming}
 Both the LCA and BS-Capon beamformers are based upon a detailed investigation of how Capon beamfoming behaves when applied to medical ultrasound images.

### 2014-04-13 ThesisJP/Introduction/introduction.tex
Finished acknowledgments.

@@
 Since the sample covariance matrix is hermitian, a solver based on Cholesky decomposition should also help to improve the throughput.
+However, a triangular matrix can make memory management and finding the best thread layout even more difficult.
 
 \subsection{Paper\,II}
-In \textbf{Paper\,II} we focused on achieving real time Capon processing for cardiac ultrasound by implementing the BS-Capon beamformer on a GPU.
+In \textbf{Paper\,II} we focused on achieving real time Capon beamforming for cardiac ultrasound imaging by implementing the BS-Capon beamformer on a GPU.
 However, we also present the first medical ultrasound loops processed with ES- and BS-Capon beamforming.
@@
 Yet, initially it has been proven hard to transfer all results obtained in simulations to \textit{in vivo} images.
-The improvements seen are minimal, especially with the number of calculations in mind.
-The beamformers are clearly able to decrees the lateral width of point scatterers and thin structures \textit{in vivo}, and to sharpen step edges.
-On the other hand, the overall contrast is visually not improved and sometimes it is worse (Fig.\,6b of \textbf{Paper\,II}).
-Since the Capon beamformer only affect directional noise, it is clear that contrast will get worse in low \nom{SNR}{Signal-to-noise ratio} situations when signal cancellation occurs.
-The oversampling approach presented in \textbf{Paper\,III} might improve on this, but it will only reduce signal cancellation if the phase shifts across the aperture are close to linear and small.
-Another issue is that bright points with a wide lateral profile exhibits better visual contrast when surrounded by speckle noise than points with narrow lateral profiles.
-In future work a detailed investigation of why some results, especially the contrast, obtained in simulations are not transferable to \textit{in vivo} needs to be performed.
-It will also be crucial to show larger improvements \textit{in vivio} than what is presented in \textbf{Paper\,II}.
-Both to justify all our computations, and to show an image which has both highly improved resolution and contrast.
+The observed improvements are minimal, especially with the number of calculations in mind.
+The beamformers are clearly able to decrease the lateral width of point scatterers and thin structures \textit{in vivo}, and to sharpen step edges.
+On the other hand, the observed contrast is not improved and sometimes it is worse (Fig.\,6b of \textbf{Paper\,II}).
+Since the Capon beamformer only affects directional noise, it is clear that contrast will get worse in low \nom{SNR}{Signal-to-noise ratio} situations when signal cancellation occurs.
+The oversampling approach presented in \textbf{Paper\,III} might improve on this, but it will only reduce signal cancellation if the phase shifts across the aperture are close to linear and are small.
+Another issue is that bright points with a wide lateral profile exhibit better visual contrast when surrounded by speckle noise than points with narrow lateral profiles.
+Future work should address a detailed investigation on why some results, especially the contrast obtained in simulations is so hard to achieved \textit{in vivo}.
+It will also be crucial to show larger improvements \textit{in vivo} than what is presented in \textbf{Paper\,II}.
+Both to justify all our computations, and to show an image with both highly improved resolution and contrast.
+
 As pointed out at the end of \textbf{Paper\,II}, cardiac ultrasound imaging might not be the best modality for Capon beamforming.
-Cardiac ultrasound imaging was mainly selected because it is the medical ultrasound modality which requires most calculations per second.
-An application where the focus is detection of closely spaced point targets, might be better suited.
-A better suited adaptive beamformer for cardiac applications might be the low-complexity adaptive beamformer (\nom{LCA}{Low complexity adaptive (beamformer)} beamformer), where the Capon optimization problem is applied on a set of predefined windows.
-This method has linear complexity and shares many properties with the Capon beamformer.
+Cardiac ultrasound imaging was mainly selected because it is the medical ultrasound modality which requires the most calculations per second.
+A better suited adaptive beamformer for cardiac ultrasound imaging might be the low-complexity adaptive beamformer (\nom{LCA}{Low complexity adaptive (beamformer)} beamformer), where the Capon optimization problem is applied on a set of predefined windows.
+This method has linear computational complexity and has similar super-resolution properties as the Capon beamformer.
 If only minor improvements are obtained, it would be easier to ignore this fact if the algorithm is less computationally costly.
 %Finally, it would also be interesting to see how the Eigenspace-based Capon beamformer performs on cardiac images, and how its absence of a distortionless criterion will influence the image spatially and temporally.
-The work in \textbf{Paper\,II} made it clear that interesting findings are done when users are able to watch the temporal behavior of an algorithm. This led us to the work of \textbf{Paper\,III}.
+
+The work in \textbf{Paper\,II} made it clear that interesting findings are done when a real-time implementation makes it possible to watch the temporal behavior of an algorithm.
+This encourage for the work of \textbf{Paper\,III}.
 
@@
 In \textbf{Paper\,III} we investigate the inherent self nulling involved with Capon beamforming, and how this effect comes naturally into play when the beamformer is subjected to linear steering vector errors caused by lateral undersampling.
-These errors are naturally reduced when the lateral sampling is increased.
-This simple observation is something which has been missing in the literature on Capon beamforming for medical ultrasound imaging.
-The reason could be ignorance or its possible negative impact on imaging frame rate.
-A lack of studies where Capon beamforming is performed on consecutive frames might also explain the lack of comments.
-Fortunately, as shown in \textbf{Paper\,III}, it is possible to maintain the same frame rate as with DAS beamforming by applying oversampling on receive using a set of steering vectors (phase delays).
+This observation is something which has been missing in the literature on Capon beamforming for medical ultrasound imaging.
+The reason could be that researchers have focused more on the theoretical aspects of Capon beamforming than on the practical aspects of ultrasound imaging.
+A lack of studies where Capon beamforming is performed on consecutive frames might also explain the lack of discussions on this theme.
+Fortunately, as shown in \textbf{Paper\,III}, it is possible to maintain the same acquisition frame rate as for DAS beamforming by applying oversampling on receive using a set of steering vectors (phase delays).
 It also turns out that more matrix inversions are not needed either.
@@
 The Cartesian grid should also have high resolution in order to preserve the un-sharpening introduced by Capon beamforming.
-As just mentioned in the discussion of \textbf{Paper\,II}, reduced signal cancellation should prevent contrast from degrade to less than the contrast obtain with DAS.
-It will, however,  not correct for signal cancellation caused by phase aberrations.
+
+As just mentioned in the discussion of \textbf{Paper\,II}, reduced signal cancellation should prevent contrast from degrading to less than the contrast obtained with DAS beamforming.
+%It will, however,  not correct for signal cancellation caused by phase aberrations.
 At best it will correct for the linear term introduced by the aberrator if the term is small.
@@
 However, the method could still be used to clean up still images of standard cardiac views.
+
 The paper does also propose an interesting method for automatic tuning of rendering parameters.
-Fig.\,3d in \textbf{Paper\,IV} actually shows that an optimal threshold could be found based on global and not only local statistics.
+Figure 3d \todo{Direct ref} in \textbf{Paper\,IV} actually shows that an optimal threshold could be found based on global and not only local statistics.
 Even though the visualization based on global statistics has more errors, with respect to endocardium visibility, the tuning of the opacity function is still fully automatic if an iterative search for the minima in Fig.\,3d in \textbf{Paper\,IV} is performed.
-A global opacity transfer function will assure a rendering which are realistic, but it will still be dependent on a successful tracking for the auto-adjustment to work.
-An automatically tuned opacity function based on global statistics combined with an improved smoothing scheme is likely to be better suited for cardiac ultrasound imaging than a local opacity transfer function.
-A local opacity transfer function is just to extreme to be clinically robust.
-In addition, it would also be interesting to see if statistics based on the window selected by the Capon or the LCA beamformer could be utilized to adjust the opacity.
+A global opacity transfer function will insure a rendering which is realistic, but it will still be dependent on a successful tracking for the auto-adjustment to work.
+
+An automatically tuned opacity transfer function based on global statistics combined with an improved smoothing scheme is likely to be better suited for cardiac ultrasound imaging than a local opacity transfer function.
+A local opacity transfer function is just too extreme to be clinically robust.
+
+For future work, it would also be interesting to see if statistics based on the weight vector selected by the Capon or the LCA beamformer could be utilized to control local opacity modulations.
 It should also be possible to use this statistic as input to image filtering.
-An adaptively selected window will be symmetric in homogeneous regions and left-right shifted close to tissue interfaces.
-In that way we can apply less smoothing and higher opacity at tissue interfaces.
+An adaptively selected weight vector will be symmetric in homogeneous regions, and left-right shifted close to tissue interfaces.
+In that way we can, with improved confident, apply less smoothing and higher opacity values at tissue interfaces.
 If this will be different than just looking at neighboring pixels in an edge-preserving image filter is to be seen.
@@
 The fifth paper stands out from the other papers by not dealing with an adaptive processing technique.
-It is still, however, centered around the topic of implementing algorithms for ultrasound imaging on the GPU.
+However, it is still centered around the topic of implementing algorithms for ultrasound imaging on the GPU.
 Ultrasound simulations are typically used for research and sometimes also to pre-calculate configurations of a given scan sequence, for instance to achieve a certain pressure in a given area.
-The continuously growth in computational power will, in the future, make it possible to evaluate a scan sequence on the fly, removing the need for lookup tables.
-Estimation of the mechanical and thermal index could then also be improved.
+The continuous growth in computational power will in the future make it possible to evaluate a scan sequence on the fly.
+For example to improve the estimation of the mechanical and thermal index.
+It will also add improved interactivity to ultrasound simulation tools.
+
 Achieving higher processing speed for an excising algorithm by implementing it on a new architecture is not academically relevant in its own.
-It has to be combined with either a good analysis of what special adjustments that had to be made to the algorithm in order for the increase to happen or a suggestion of how to utilize the speed gained.
+A good analysis has to be given of how the algorithm was modified in order for the increase to happen, or how to utilize the gained speed.
 These are points we have tried to answer in both \textbf{Paper\,I}, \textbf{II},  \textbf{IV}, and \textbf{V}.
+
 In \textbf{Paper\,V} the increased speed made an interactive ''Paint-program'' feasible.
 As of January 2014 the simulator has been download 200 times and counting.
-It also has close to a "whopping" 1000 views on youtube.
-\\\\\\
-Finally it is worth mentioning that the code for both the Capon beamformer\footnote{github.com/jpaasen/cos} and the ultrasound field simulator\footnote{github.com/jpaasen/hos} have been made open source and is available on github.
+It also has close to a 1000 views on Youtube.
+
+\section{Software}
+Finally it is worth mentioning that the code for both the Capon beamformer\footnote{github.com/jpaasen/cos} and the ultrasound field simulator\footnote{github.com/jpaasen/hos} have been released under an open source license.
 \endinput

### 2014-04-13 ThesisJP/Thesis.tex
Finished acknowledgments.

@@
 %Accelerating ultrasound algorithms using graphics processing units 
+%Pinner og romber fra det yttre rom
 \vspace{0.5cm}}
@@
 The research was funded by \textit{Medical Imaging Laboratory} (MI-Lab), and was carried out at the Department of Circulation and Medical Imaging.
-The main supervisor has been Professor Sverre Holm  in the Department of Informatics, University of Oslo, and Professor II at MI-Lab, NTNU.
-Co-supervisors have been PhD Stein Inge Rabben, PhD Erik Steen and Professor Hans Torp.
+The main supervisor has been Professor Sverre Holm in the Department of Informatics, University of Oslo (UiO), and Professor II at MI-Lab, NTNU.
+Co-supervisors have been Ph.D.
+Stein Inge Rabben, Ph.D.
+Erik Steen, both at GE Vingmed Ultrasound AS, and Professor Hans Torp in the Department of Circulation and Medical Imaging, NTNU.
 
 \subsubsection{Acknowledgments}
-\todo{Write acknowledgments ...}
-Takkeliste: Beate.
-Farmor.
-Familien.
-Andreas Austeng.
-Jo Inge Buskenes.
-Carl-Inge Colombo Nilsen.
-Ines Hafizovic for feedback on the introduction and background.
-Sverre Holm.
-Miljøet i Trondheim.
-PhD studentene i Trondheim spesielt.
-Miljøet i Oslo.
-Miljøet hos Vingmed.
-SquareHead.
+First of all I would like to thank my main supervisor Sverre Holm and the rest of my team of supervisors for all help and guidance through the Ph.D project.
+Andreas Austeng also deserves big thank you.
+In the second part of my Ph.D.
+project you have been my supervisor more than anyone.
+Thank you to all my co-authors for their contributions.
+I also want to thank the Faculty of Medicine, MI-Lab and the Department of Circulation and Medical Imaging, NTNU for believing in me and making it possible to conduct this work.
 
+During these years I have had the great pleasure of working with and getting to know several research communities and companies.
+I therefore would like to thank all colleagues at the Department of Circulation and Medical Imaging, NTNU, the Department of Informatics, UiO, GE Vingmed Ultrasound AS, and Squarehead Technology AS for providing great working environments.
+A special thanks goes to all my fellow Ph.D.
+students at MI-Lab.
+I will never forget our conference trips.
+A special thanks also goes to Jo Inge Buskenes at "office" 4470 for making every working hour so much fun.
+
+Thank you to Ines Hafizovic and Carl-Inge Colombo Nilsen for extensive help with reviewing this thesis.
+
+I would like to thank my family for all support. But most of all, I want to thank Beate for all her love, patient and support. Without it I would never have manage to complete this theses.
+\null
+\vfill
+\hfill Dedicated to Sofie Mathilde \AA{}sen
 \CustChapter*{List of publications}

### 2014-04-13 ThesisJP/Introduction/introduction.tex
Added paper 1.

@@
 \subsection{Paper\,I}
-\textbf{An Optimised GPU Implementation of the MVDR Beamformer for Active Sonar Imaging}\\
+\textbf{An Optimized GPU Implementation of the MVDR Beamformer for Active Sonar Imaging}\\
 J.\:I.\:Buskenes, \textbf{J.\:P.\:\AA{}sen}, C.-I.\:C.\:Nilsen and A.\:Austeng\\

### 2014-04-13 ThesisJP/Paper1/howto_howto_capon.tex
Added paper 1.

\newif\ifPeerReview\PeerReviewfalse             % Whether to create the PeerReview version or
                                                % Journal version
\newif\ifFlatArchive\FlatArchivefalse           % Whether archive is flat (messy) or contain 
                                                % subfolders for graphics etc.
\newif\ifFloatAtEnd\FloatAtEndfalse             % Available in PeerReview mode:

\graphicspath{{./Paper1/}}

\definecolor{tabBlue}{HTML}{AACCFF}

%%%%%%%%%%                                      ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
% MACROS %       
%%%%%%%%%%                                      ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

\newcommand\Fig[1]{Fig.~\ref{#1}}

\newcommand\nn{\nonumber\\}

\newcommand\nmat[1]{\begin{matrix}#1\end{matrix}}
\newcommand\bmat[1]{\begin{bmatrix}#1\end{bmatrix}}
\newcommand\case[1]{\begin{cases}#1\end{cases}}
\newcommand\textbox[2]{\footnotesize\text{\parbox{#1}{\centering\emph{#2}}}}

\newcommand\rand{\text{rand}}
\newcommand\randn{\text{randn}}
\newcommand\rect{\text{rect}}
\newcommand\sinc{\text{sinc}}
\newcommand\tr{\text{tr}}
\newcommand\adj{\text{adj}}

% \newcommand\max{\text{max}}
\newcommand\argmin[1]{\text{arg}\;\underset{#1}{\text{min}}}

\newcommand\qqquad{\quad\qquad}
\newcommand\qqqquad{\qquad\qquad}

% \renewcommand\l[1]{\left#1}
% \renewcommand\r[1]{\right#1}

% {\text{\parbox{1.5cm}{\centering volume hyper- sphere}}}

%Keyword colouring:
\newcommand\kw[1]{#1}
\newcommand\parm[1]{#1}%\color{Black}#1\color{Black}}

\newcommand\of[1]{\scriptstyle(\parm{#1})\displaystyle}
\newcommand\df[1]{\scriptstyle[\parm{#1}]\displaystyle}
\newcommand\var[3]{#1_\text{#2}\of{#3}}

\newcommand\diag{\text{diag}}

% \raisebox{lift}[extend-above-baseline][extend-below-baseline]{text}
\newcommand\mt[1]{\text{\emph{#1}}} %mt = mathtext
\newcommand\mathnorm{\textstyle}
\newcommand\mathbig[1]{\displaystyle#1\mathnorm}
\newcommand\mathsmall[1]{\scriptstyle#1\mathnorm}
\newcommand\mathtiny[1]{\scriptscriptstyle#1\mathnorm}
\newcommand\sfrac[2]{\scriptstyle\raisebox{0.25pt}[0pt][0pt]{$\frac{#1}{#2}$}\mathnorm}
\newcommand\nfrac[2]{\textstyle\frac{#1}{#2}\displaystyle}

\newcommand\sumu[1]{\sum\limits^{#1}\;}
\newcommand\suml[1]{\sum\limits_{#1}\;}
\newcommand\sumb[2]{\sum\limits_{#1}^{#2}\;}

\newcommand\produ[1]{\prod\limits^{#1}\;}
\newcommand\prodl[1]{\prod\limits_{#1}\;}
\newcommand\prodb[2]{\prod\limits_{#1}^{#2}\;}

\newcommand\defeq{\overset{\underset{\mathrm{def}}{}}{=}}

%Math macros:
\newcommand\diff[2]{\frac{\kw{d}\;\textstyle #1\scriptstyle}{\kw{d\parm{#2}}}\displaystyle}
\newcommand\ddiff[2]{\frac{\kw{d^2}\;\displaystyle #1\scriptstyle}{\kw{d\parm{#2}}^2}\displaystyle}

\renewcommand\d[1]{\scriptstyle\kw{\;d\parm{#1}}\displaystyle}

% These commands are mutually exclusive. Remember to "renew" in v2.
\newcommand\intb[4]{\int\limits_{#3}^{#4} #1 \d{#2}} % \int{exp}{var}{from}{to}
\newcommand\intl[3]{\int\limits_{#3} #1 \d{#2}} % \int{exp}{var}{for all}
\newcommand\intu[2]{\int #1 \d{#2}} % \int{exp}{var}{for all}

\newcommand\T{^{\scriptscriptstyle T}}
\renewcommand\H{^{\scriptscriptstyle H}}

\renewcommand\vec[1]{\boldsymbol{#1}}
\renewcommand\mat[1]{\boldsymbol{#1}}

\newcommand\Om{O_\text{m}}
\newcommand\Oa{O_\text{a}}
\newcommand\Nl{N_\text{l}}
\newcommand\Nk{N_\text{k}}
\newcommand\1{\vec 1}
\newcommand\I{\mat I}
\renewcommand*\a{\vec a}
\renewcommand*\i{\vec i}
\renewcommand*\k{\vec k}
\newcommand*\n{\vec n}
\newcommand*\p{\vec p}
\newcommand*\s{\vec s}
\newcommand*\w{\vec w}
\newcommand*\x{\vec x}
\newcommand*\y{\vec y}

\newcommand*\A{\mat A}
\newcommand*\B{\mat B}
\newcommand*\C{\mat C}
\newcommand*\E{\mat E}
% \renewcommand*\H{\mat H}
\renewcommand*\P{\mat P}
\newcommand*\eP{\mat{\hat P}}
\renewcommand*\R{\mat R}
\newcommand*\Ri{\R^{-1}}
\newcommand*\eR{\mat{\hat R}}
\newcommand*\eRi{\hat{\mat R}\;\!^{-1}}
\newcommand*\Navg{N_\text{avg}}
\newcommand*\W{\mat W}
\newcommand*\X{\mat X}
\newcommand*\Xd{\X_{\!\Delta}}
\newcommand*\Y{\mat Y}

\renewcommand*\L{\mat \Lambda}
\newcommand*\U{\mat U}
% \renewcommand*\t{\mathtiny{^T}}
% \newcommand*\h{\mathtiny{^H}}
\renewcommand*\t{^T}
\newcommand*\h{^H}

\newcommand\D{\vec\nabla} %Del: Vector differential operator - nabla
\newcommand\Dx{\vec\nabla\times}
\newcommand\Dd{\vec\nabla\cdot}

\newenvironment{outline}
{\begin{itemize}}
{\end{itemize}}

\begin{abstract}
% Two limiting factors of modern phased array sonar imaging systems are the angular resolution and contrast. Adaptive beamformers target these limitatitons by weighting sensors dynamically to fit the impinging wavefield.

The Minimum Variance Distortionless Response (MVDR) beamformer has recently been proposed as an attractive alternative to conventional beamformers in active sonar imaging. Unfortunately, it is very computationally complex because a spatial covariance matrix must be estimated and inverted for each image pixel. This may discourage its use unnecessarily in sonar systems which are continuously being pushed to ever higher imaging ranges and resolutions.

In this study we show that for active sonar systems up to 32 channels, the computation time can be significantly reduced by performing arithmetic optimizations, and by implementing the MVDR on a Graphics Processing Unit (GPU). We point out important hardware limitations for these devices, and assess the design in terms of how efficiently it is able to use the GPU's resources. On a quad-core Intel Xeon system with a high-end Nvidia GPU, our GPU implementation renders more than a million pixels per second (1\;MP/s). Compared to our initial CPU implementation the optimizations described herein led to a speedup of more than 2 orders of magnitude, or an expected 5-10 times improvement had the CPU received similar optimization effort. This throughput enables real-time processing of sonar data, and makes the MVDR a viable alternative to conventional methods in practical systems.
% If the HISAS1030 was attached to a platform traveling at 1.8m/s, creating full-coverage sectorscan images in real-time would be possible with this performance.

% We benchmarked optimized implementations of the MV beamformer in MATLAB and C, running on a quad-core Intel Xeon E5620, and in CUDA running on a high-end nVidia card. The beamformers operated on data from the 32x4 element Kongsberg Martime HISAS1030 sonar, using a conservative set of robustification parameters. 
%The CPU would then be free to take on the rest of the beamforming as well.
\end{abstract}

\section{Introduction}

% Images from a phased array active sonar system are formed in the post-processing by focusing on one pixel at a time. This is achieved by delaying and weighting each of the array's channels, providing coarse and fine adjustments to the region of focus, respectively. This is commonly referred to as beamforming. Assuming far-field conditions, the delays are usually chosen to steer the pixel of interest to broadside, and the weights chosen to reject the noise impinging on the array from non-broadside directions.
% 
% The algorithm responsible for carrying out this task is referred to as a beamformer. Depending on whether data is evaluated when choosing a suitable set of delays and weights, beamformers can be branded either conventional or adaptive. A good example of the former is the The Delay-and-Sum (DAS) beamformer, which - as the name implies - simply delay the sensor channels appropriately prior to summing them up. Applying various weight sets, or windows, to the data prior to summation allows the lateral \gls{SNR} to be improved, but at the inevitable expense of a deteriorated lateral resolution. \todo{aka. the curse of physics.}
% 
% The Minimum Variance Distortionless Response (MVDR) (or \gls{MVDR}) beamformer, have the potential to overcome this limitation. While ensuring unit gain in the look direction, it computes the set of weights that minimizes the energy accumulated by the array from other directions. In other words, whenever the distribution of noise and interference is primarily focused around specific angles, an array response with a high degree of suppression at these angles are chosen, resulting in an image with a lateral \gls{SNR} and resolution superior to anything the \gls{DAS} can come up with.

% 
% In active sonar imaging the constrast and resolution can be improved by the use of adaptive beamforming. 
% 
% Adaptive beamforming 
% A critical part in the sonar image formation is the 
% The image quality is mainly limited by beampattern of the imaging arrays.  resolution and noise suppression
% 
% The quality of sonar images 
% 
% An essential part of a sonar imaging system is the beamformer. 
% 
% 
% A sonar image is typically formed by using receive beamforming to focus on one pixel at a time. The beamformer's ability to suppress interference   to focus on one pixel at the time. How well a pixel value is estimated, and thus how 
% 
% A sonar image is built one pixel at the time the 
% 
% The image quality of a sonar system will depend on the imaging arrays ability to focus in 
% 
% 
% 
% Forming images done by combining data from multiple receivers in a way that to obtain focus on one pixel at the time, and assigning the output value to the intensity of that pixel. The 
% 
% Most modern active sonar systems employ phased arrays to 
% The basic way of a sonar image was formed 
% 
% A sonar system's ability to 
% 
% Adaptive beamforming has 
% 
% The image quality 
% 
% To improve the image quality of 
% 
% In high resolution sonar imaging the 
% 
% The image 
% 
% 
% In a typical sonar imaging scenario an encoded signal is transmitted to highlight a surface of interest, and the reflected wavefield sampled using a receiver array. A image is then formed in the post-processing by coherently combining the receiver outputs to focus at one pixel at the time. The principle of adjusting the arrays focus is commonly referred to as beamforming, and involves assigning suitable delays and weights to the array's channels.
%

% A sonar image is typically formed by using receive beamforming to focus on one pixel at a time.   The image resolution and contrast The image quality The quality of active sonar images depend - amoung other things - on the beamformer's ability to suppress interference  resolution and constrast
% 
% A sonar image is typically formed by using receive beamforming to focus on one pixel at a time. The quality of the final image will depend on the beamformer's ability to suppress interference and noise from off-pixel.    to focus on one pixel at the time. How well a pixel value is estimated, and thus how
% 
% 
% In conventional active sonar imaging systems 
% 
% Various methods have been applied to active sonar imaging. One of these methods, 
% 
%  eamforming techniques have been applied in several fields to improve image. methods can be applied to 
% 
% Most active sonar imaging systems are based on conventional beamforming methods. These are robust and simple, but does not take the incoming wavefield into account. offer  in the image formation process. While being inherently robust, these methods 
% 
% 
% With the ever increasing processing power found in modern computers 
% 
% As the processing power of todays modern computer systems keeps improving the feasibility of 
% 
% As the processing power of modern computer systems keep improving, the use of data adaptive methods in various imaging fields.

Data driven methods have been introduced in various imaging fields over the years in attempts to improve image quality. One such method is the Minimum Variance Distortionless Response (MVDR) beamformer. When compared to conventional methods, MVDR is often able to improve image contrast and resolution, as shown in e.g. radar~\cite{Benitz1997}, ultrasound~\cite{Synnevag2007}, and active sonar~\cite{Blomberg2012a,Blomberg2011,Dursun2009,Lo2004}.

Despite its inherent potential, the MVDR beamformer has yet to see widespread adoption in the active sonar field. There may be several reasons for this. For one, the method is not inherently robust, and may suffer from a phenomenon called signal cancellation in active systems~\cite{Widrow1982}. Another reason is that in its original form, the computational complexity is cubic with the number of channels, O($M^3$), while conventional beamformers are at O($M$). This is because a spatial covariance matrix is estimated and inverted for each image pixel.

% Robustification:
% R estimate accuracy vs. dynamic trackability: (1) Trade observation time for bandwidth (2) Partial adaptiveness (less degrees of freedom than channels) (3) Quadratic constraint on sensor weights to limit the norm to some preset value.
% -  
To ensure MVDR robustness, the literature suggests combining means such as temporal and spatial averaging, and regularization~\cite{VanTrees2002,Kailath1985}. The complexity issue, on the other hand, can be handled by introducing well-founded approximations. For instance, some studies assume data stationarity which allows the formation of a Toeplitz-structured covariance matrix that is simpler to invert~\cite{Asl2012a,Jakobsson2000}. Other studies perform MVDR beamforming in beamspace (the spatial frequency domain), which can be considered sparse due to the limited angular extent of the received beam~\cite{VanTrees2002,Nilsen2009}. Exploiting this can lead to considerable performance improvements, especially in high channel count systems.

% Some low complexity approximations to the MVDR beamformer is also available, such as the Low Complexity Adaptive (LCA) beamformer,

% Such devices can often this by combining up to thousands of computing cores that each performs modestly, but combined performs impressively.

% We will show that by exploiting symmetry and avoiding unnecessary computations, the spatial averaging will also aid in reducing the overall arithmetic complexity\todo{make sure this is shown clearly somewhere}. for Unlike other work, which focus modifying or transforming\todo{get low channel count in, as it makes a case against the other solutions.} We make no approximations, but rather stay with a textbook example of the method that handle complex basebanded data\todo{misplaced?}.

% Other work (keywords: mvdr, capon, computation, efficient, complexity):
% - Partially adaptive beamforming. DAS on non-overlapping subarrays, then MVDR \cite{Swingler1999a}.
% - Beamspace / frequency based methods.
% - Low complexity. PCA + LC. 
% - Systolic arrays. Radar \cite{Ward1984} 
% - Toeplitz. \cite{Yapanel} Spectrum estimation - speech recognition. 
% 
% Us:
% - Aim for the GPU. Simple is often better.
% - Therefore: Optimize the standard MVDR chain.
% 1. Reduce arithmetic instructions
% 2. Implement for full utilisation of the arithmetic throughput on the GPU.
% 
% (Made partially adaptive by employing subarray averaging to reduce degrees of freedom to L.)

In this work, we show that such approximations may not be necessary for sonar systems up to 32 channels, since here the computation time can be reduced by 1-2 orders of magnitude by implementing MVDR on a Graphics Processing Unit (GPU). Similar work may be found in e.g. medical ultrasound imaging, where Chen \emph{et al.} have demonstrated a GPU implementation of the MVDR that operates on real valued data~\cite{Chen2011,Chen2011a}. While this implementation is fast, the real valued data constraint does not allow the creation of asymmetric and shifted responses, and hence prevents the MVDR's potential to be fully reached. Our implementation is not restricted in this manner.

The implementation is adapted to a well sampled 100\;kHz wideband active sonar system operating in the near field. The system has a range resolution comparable to the sample period so our MVDR implementation runs near single snapshot mode. To deal with this and avoid signal cancellation we compute and average the covariance estimates from overlapping subarrays~\cite{Kailath1985}. This may resemble techniques such as "subarray MVDR"~\cite{Chapman1976}, but unlike that technique we are not using subarrays to reduce complexity but to make MVDR in an active system possible. In related work, we investigated using the same MVDR GPU implementation in cardiac ultrasound imaging~\cite{Asen2012,Asen2013}, and in active sonar imaging we have compared the performance of this implementation to comparable adaptive methods~\cite{Buskenes2013}.

This article is outlined as follows: In section \ref{methods} we introduce the concept of adaptive beamforming and provide details on the MVDR method. Then in \ref{maptogpu} we investigate the complexity issue, discuss means for reducing arithmetic complexity, and detail an implementation that makes efficient use of the GPU's parallel resources. The final design is assessed in \ref{images_and_benchmarks}-\ref{discussion}, where we provide benchmarks, comparisons with similar CPU implementations, and measures of how efficiently our implementation makes use of the GPU's resources.

%  Furthermore, by implementing it on a modern \gls{GPU} we have gained a speedup of around 2 orders of magnitude compared to an optimized single-thread implementation in C.

% These devices have in recent years emerged as a flexible and highly potent parallel technology well suited for many DSP tasks, and are available off-the-shelf at reasonable prices\todo{skip this?}.

% These devices have in recent years emerged as a flexible and highly potent parallel technology well suited for many DSP tasks, and are available off-the-shelf at reasonable prices.

% Yay or nay? $\Rightarrow$ GPU ultrasound (\cite{So2011,Chen2011}). Complex data, all robustification techniques.

% - Adaptive beamformers offers image quality edge, but are usually computationally complex.
% - Who's tried before?
% - We: Active sonar MVDR, GPU, high performance.

\section{Methods}\label{methods}

Consider a sonar imaging scenario where an encoded signal is transmitted to highlight a surface of interest, and assume that the backscattered wavefield is properly sampled using a uniform linear array of receivers. An image can then be formed by coherently combining the receiver outputs to focus at one angle at the time. The principle of adjusting the array's focus in range and direction is commonly referred to as beamforming, and involves assigning suitable delays and weights to the array's channels.

\subsection{Beamforming}

\ifPeerReview
\begin{figure}[H]\centering
\includegraphics[width=0.8\linewidth]{gfx/buske1.eps}
\else
\begin{figure}[!t]\centering
\includegraphics[width=0.8\linewidth]{gfx/beamforming.eps}
\fi%
\caption{Beamforming principle. Signal signature is first removed by matched filtering. Then - prior to summation - a suitable set of delays, $\Delta$, and weights, $w$, are applied to focus on a pixel of interest at angle and range ($\theta,n$).}\label{beamforming}
\end{figure}
Let the receiver be an $M$ element uniform linear array, and assume that the signature of the transmitted signal has been removed by a matched filter (Fig. \ref{beamforming}). Further assume that the array channels are digitally delayed to focus at a pixel with azimuth angle $\theta$ and range sample $n$, such that the delayed data from the $m$th channel can be expressed as $x_m[\theta,n]$. To simplify notation, we make the dependence on $\theta$ implicit from now on.

By definition the beamformer output $z[n]$ can now be expressed as the weighted sum of all the delayed data samples:
\begin{align}
z[n] = \w\H[n]\x[n] = \bmat{w_0[n]\\w_1[n]\\\vdots\\w_{M-1}[n]}^H \bmat{x_0[n]\\x_1[n]\\\vdots\\x_{M-1}[n]},\label{z}
\end{align}
where $w_m$ is the weight factor assigned to channel $m$. With static weights this would be referred to as the conventional delay-and-sum (DAS) beamformer. A large variety of weighting functions exists here for trading lateral resolution for improved noise suppression (contrast), but one always ends up with a compromise between the two~\cite{Harris1978}.

Various adaptive beamformers target this limitation by allowing the weights to change for each pixel to better fit the dynamic nature of the incoming wavefield. In other words, they attempt to use the \emph{a priori} information present in the data to improve image quality. The MVDR beamformer is one such method. It finds the set of complex weights that minimizes the beamformer's expected output power, while ensuring unity gain in the look direction~\cite{Capon1969}. This is a convex optimization problem that can be solved using Lagrange multipliers to yield the solution:
\begin{gather}
\vec w[n] = \frac{\Ri[n]\1}{\1\T\Ri[n]\1},\label{weights}
\end{gather}
where $\1$ is a row vector of ones that represents broadside steering, and $\R=E\{\x[n]\x\H[n]\} \in\mathbb{C}^{M,M}$ is the spatial covariance matrix for the full array. Since we pre-steer our data to every pixel in the image we will always look to broadside, but (\ref{weights}) can be easily generalized by substituting $\1$ with a suitable steering vector. Since $\R$ is unknown, we estimate it by computing a sample covariance matrix $\eR$. In this computation we perform some degree of:
\begin{itemize}
\item \emph{spatial averaging} to avoid signal cancellation by decorrelating coherent echoes~\cite{Kailath1985},
\item \emph{temporal averaging} over an interval comparable to the pulse length (1-5 samples) to maintain true speckle statistics~\cite{Synnevag2009a}, and
\item \emph{diagonal loading} to improve robustness to parameter errors~\cite{Cox1987,Maksym1979}.
\end{itemize}%
%
% will perform some degree of \emph{spatial averaging} to avoid signal cancellation caused by coherent echoes~\cite{Kailath1985}, \emph{temporal averaging} to maintain true speckle statistics~\cite{Synnevag2009a}, and \emph{diagonal loading} to improve robustness to parameter errors ~\cite{Cox1987,Maksym1979}. 
%
Combined, these steps will also ensure a numerically well conditioned $\eR$.

To form $\eR$ we will first compute an intermediate sample covariance matrix $\breve{\R}$, which we form using temporal and spatial averaging. To do this we need to segment our array into subarrays. If we let $x_l[n]$ represent the data vector from subarray $l$,
\begin{gather}
\x_l[n] = \bmat{x_l[n] & x_{l+1}[n] & \dots & x_{l+L-1}[n]}\T,
\end{gather}
then $\breve{\R}$ can be calculated as:
\begin{gather}
\breve{\R}[n] =  \frac{1}{N_K N_L} \sumb{l=0}{M-L}\sumb{n'=n-K}{n+K} \x_l[n']\x_l\H[n'] \in\mathbb{C}^{L,L},\label{spatialR}
\end{gather}
where $N_K = 2K+1$ is the number of temporal samples to perform averaging over, and $N_L = M-L+1$ is the number of subarrays.

The final estimate $\eR$ is found by adding a fraction $d$ of the total power of $\breve{\R}[n]$ to its diagonal~\cite{Synnevag2007}:
\begin{align}
\eR[n] = \breve{\R}[n] + \I \frac{d}{L} \tr\{\breve{\R}[n]\},\label{finalR}
\end{align}
where $\I$ is an identity matrix, $\tr\{\cdot\}$ represents the matrix trace operation, and $\tr\{\breve{\R}[n]\}$ is an estimate of the energy received from this pixel.

Note how subarray averaging led to a size reduction of $\eR$ from $\mathbb{C}^{M,M}$ to $\mathbb{C}^{L,L}$, and hence will produce an $L$-element weight set when substituted into (\ref{z}). This weight set is applied to all the subarrays, prior to computing the beamformer output as in (\ref{z}). Or, equivalently, we may apply the weight set to the sum of all the subarrays:
\begin{align}
z[n] = \w\H[n] \sumb{l=0}{M-L} \x_l[n].\label{finalZ}
\end{align}
\ifPeerReview
\begin{figure}[!t]\centering
\includegraphics[width=.8\linewidth]{gfx/buske2.eps}
\else
\begin{figure}[!t]\centering
\includegraphics[width=.9\linewidth]{gfx/implementation.eps}
\fi%
\caption{MVDR beamforming. For each pixel in range and azimuth,\newline
1. an $L\times{}L$ sample covariance matrix $\eR$ is computed, \newline
2. the term $\eR^{-1}\1$ is found using a linear equation solver,\newline
3. and the beamformer output $z$ is computed from (\ref{finalZ}), where $\w$ is found by substituting $\eR^{-1}\1$ into (\ref{weights}). } \label{mvdr_beamforming}
\end{figure}
% \begin{figure}[!t]\centering
% \includegraphics[width=\linewidth]{gfx/algorithm_structure.eps}
% \caption{MVDR beamforming. First a spatial covariance matrix is estimated from the delayed data (\ref{spatialR}-\ref{finalR}), then the weights are computed (\ref{weights}) and finally applied to the delayed channel data (\ref{z}).}
% \label{implementation}
% \end{figure}
As summarized in Fig. \ref{mvdr_beamforming}, the MVDR method is applied to each pixel independently, by:
\begin{enumerate}
\item computing the sample covariance matrix $\eR$ in (\ref{finalR}),
\item computing $\eRi\1$ in (\ref{weights}), and
\item computing the beamformer output $z$ in (\ref{finalZ}).
\end{enumerate}
% This is illustrated in .
% \begin{figure}[!t]\centering
% \includegraphics[width=\linewidth]{gfx/buildR-complex_data-average.eps}
% \caption{.}\label{complexity}
% \end{figure}
Next we will evaluate these steps in terms of arithmetic complexity, and then discuss their mappability to parallel hardware.

\subsection{Computational Complexity}

% Computing and inverting $\eR$ is by far the most computationally intensive tasks in MVDR beamforming. This

If we neglect spatial and temporal averaging, then the computation of $\eR$ is reduced to a single outer product with complexity of O($M^2$), and the inversion is then of O($M^3$). This might lead us to believe that the inversion step is by far the most complex. But if we implement spatial and temporal averaging as in (\ref{spatialR}), then computing $\eR$ is of O($N_K N_L L^2$) and the inversion of O($L^3$). Computing $\eR$ is now the most complex operation whenever $N_LN_K>L$.
\ifPeerReview
\begin{figure}[!t]\centering
\includegraphics[width=.8\linewidth]{gfx/buske3.eps}
\else
\begin{figure}[!t]\centering
\includegraphics[width=.9\linewidth]{gfx/mvdr_complexity.eps}
\fi%
\caption{Per-pixel computational complexity of the steps in MVDR beamforming (prior to any optimizations). To avoid signal cancellation in an active sonar system we usually set $L<\frac{M}{2}$, in which region the computation of $\eR$ dominates in terms of arithmetic complexity, especially when performing temporal averaging.}\label{mvdr_complexity}
\end{figure}
To visualize these relations, and include the effects of using complex numbers, we set up complexity formulas that account for the total number of arithmetic operations for each step in the MVDR process (see appendix \ref{mvdr_formulas}). We only excluded diagonal loading performed in the computation of $\eR$, and partial pivoting used in the inversion step, since these operations contribute marginally to the end result. The entire range of possible subarray sizes from $L\in[1,M]$ was finally evaluated, with temporal averaging set to $K\in\{0,1,2\}$, and the number of channels set to $M\in\{8,16,32\}$. The results are shown in \Fig{mvdr_complexity}.

We note how the computation of $\eR$ completely dominates at smaller subarray sizes, that solving $\eRi\1$ only plays a notable role for larger array and subarray sizes, and that the computation of $z$ has a negligible impact on computation time. Also notice how temporal averaging comes with a high computational penalty. This is because building $\eR$ is heavy on complex multiplications, which require 3 times as many arithmetic instructions as a complex addition, and a lot of these are repeated unnecessarily.

% Not corrected for fused multiply accumulate

% Also, the need for partial pivoting has been ignored, as we ensure a numerically well conditioned $\eR$ by applying diagonal loading. Since this is the most common scenario, we will have a thorough look at how it can be optimized.

% and does not take into that does not  feasibility parallel decomposition  memory consumption.

\section{Mapping the MVDR to a GPU}\label{maptogpu}

% \begin{itemize}
% \item Keep delay step on CPU, as this is a common step in all beamforming
% \item Increasing matrix size reduces the number of threads per block, since the entire matrix must be in shared memory. 1 matrix 72x72 complex (8 bytes) max per block - 72 threads. same as 207 5x5 matrixes, 5*207=1036 threads. max threads in block 512.
% \end{itemize}

% An important feature of MVDR beamforming, and with beamforming in general, is that each pixel can be computed independently and in an identical fashion. Considering that sonar images often end up having millions of pixels, this makes a compelling case for investigating technologies that support massive data level parallelism. The technology that does this best is perhaps Graphics Processing Units (GPUs). These chips contain up to thousands of small computation cores, which combined deliver superior peak floating point performance compared multi-core systems such as CPUs.

An important feature of MVDR beamforming, as with beamforming in general, is that each pixel can be computed independently and in an identical fashion. Furthermore, a typical sonar image may contain millions of pixels. This represents a level of data parallelism that appears very well suited for a massively parallel architecture such as the GPU.

We decided to investigate the feasibility of running MVDR on a GPU by mapping it to a Nvidia GeForce Quadro 6000, which performance-wise is or par with or perhaps slightly inferior to the Geforce 480. This is a high-end Compute Unified Device Architecture (CUDA) enabled GPU based on Nvidia's Fermi architecture, which as now been superseded by the Kepler architecture. The code was written in Nvidia's ``C for CUDA'' framework, which adds a few extra features to the C language to specify how to distribute work across a large number of threads. It would have been possible to use GPUs from AMD and the cross-platform OpenCL framework from the Khronos group instead, but CUDA has a batch linear equation solver~\cite{Nvidia} that no OpenCL library provides. As will be discussed later, this is a key component in our design.

To use Nvidia's own terminology, the Quadro 6000 is comprised of 14 streaming multiprocessors (SMs), each having 32 CUDA cores that execute a common program called a kernel. Combined these cores deliver a peak performance of more than 1\;Tflop/s (appendix \ref{throughput}). In practice this performance is hard to obtain, but one can get fairly close by balancing the load evenly on all cores, and by trying to avoid that some cores are forced to idle due to a pending data transfer or thread synchronization. As the steps in the MVDR method require different strategies for achieving this, we have designed a different kernel and configuration for each of them. We will pay particular attention to building $\eR$, since the potential for gaining overall speedups are greatest here.

%  this assumes that the implementation is solely bound by arithmetic throughput.  r bound by memory bandwidth, arithmetic latency or arithmetic bandwidth.  or  is solely bound by processing power.   These cores need to be kept busy if we are to That make for a total of 448 CUDA cores which we need to keep busy if we want to tap the full potential.

\ifPeerReview
\begin{figure}[t]\centering
\includegraphics[width=.8\linewidth]{gfx/buske4.eps}
\else
\begin{figure}[t]\centering
\includegraphics[width=.9\linewidth]{gfx/mvdr_build_R.eps}
\fi%
\caption{Step 1: Building $\eR$. This is a visualization of how $\eR$ could be built in a case with $M=5$ sensors, with subarray size $L=3$ and temporal averaging set to $K=1$. Here $\R'_{l}$ is the sample covariance matrix for the $l$th subarray, and $\breve{\R}$ is the average of $N_K$ of these. Note that instead of performing the temporal sum last as here, one could take more temporal samples into consideration in the computation of each $r_{ij}$.}\label{mvdr_build_R}
\end{figure}

% \todo{Another shitty GPU transition}
% Before proceeding we should make a quick note on the GPU memory. Note that to obtain the best possible performance, we need to make sure that CUDA threads find the data they need in cache or registers most of the time. Only these two memory types have a high enough bandwidth to promote sustained maximum utilisation of all CUDA cores. Unfortunately, they are a sparse resource.

% This makes it apparent that computing one pixel per thread is not a feasible solution. \todo{...}

% When optimizing an algorithm for speed, it is if of importance to know whether an algorithm is bound by computations or memory. One way to get an approximate idea of this is to compute the ratio of peak arithmetic throughput to peak memory throughput of the in the target platform, and compare this ratio to the ratio of arithmetical operations to memory operations in the algorithm.

% follow collective access patterns to memory.

% At first glance, it might seem like a good idea to compute one pixel per CUDA core. Each pixel is computed using the same instructions, and without any exchange of data. However, this approach is unfeasible due to the limited cache and register memory available to each thread. One could resolve to using the slower GDDR5 memory, but this would largely mitigate the performance gain and defeat the purpose of using a GPU in the first place.

%Bandwidth becomes crucial, as getting data in to the cores and out again must happen i a frantic pace. %the feeding all these cores with data and harvesting the result

% To effectively harness the computing potential of a GPU we need to make sure all computing cores stays busy most of the time.

% Memory latency can be hidden in two ways. One is to keep a lot of threads queued up on each SM at all times, such that when some threads perform memory operations other threads can be scheduled to run instead. This is referred to as data level parallelism (DLP). Alternatively, or preferrably in addition, the designer may promote instruction level parallelism (ILP) by letting subsequent instructions in a thread be independent. While ILP can be very effective when done correctly, DLP is generally simpler to design for.
% 
% In the CUDA framework, the threads on an SM is managed through a software abstraction call a compute block. The block is a 3 dimensional grid where each element correspond to a thread that is scheduled to be executed on the SM at some point.

% load balancing
% - distribute load - optimize symmetry
% - avoid branching
% 
% memory latency hiding
% - make threads use high bandwidth memory
% - hide latency with dlp, ilp, or both
% 
% thread sync hiding
% - queue blocks up on an sm

%This is referred to as a single-program, multiple-data (SPMD) architecture\todo{malplaced}.

 %An SM is a single-program multiple-data (SPMD) processor, meaning that all the CUDA cores on an instruction unit, L1 cache and registers is shared by all the CUDA cores on that SM.  %
% GPUs, on the other hand, does not attempt to do ILP, but use all available transistors to support massive DLP. This leads to designs such as the nNvidia GeForce GTX 580. 
% each with 48kB of L1 cache (shared memory) and 128kB of register memory.

% -bank conflicts when L > 32. not treated.

\subsection{Computing the Spatial Covariance Matrix, $\hat{\boldsymbol{R}}$}\label{computing_eR}

% As Fig. \ref{mvdr_complexity} illustrates the computation of $\eR$

As observed in Fig. \ref{mvdr_complexity}, a direct computation of $\eR$ by implementing (\ref{spatialR}) and (\ref{finalR}) is the greatest computational burden. We target this issue by first \emph{minimizing arithmetic operations}, then aim to get as close to the peak \emph{arithmetic throughput} on the GPU as possible.

\subsubsection{Minimizing Arithmetic Operations}

In Fig. \ref{mvdr_build_R} we illustrate how to reduce the arithmetic complexity of building $\eR$ in a system with $M=5$ sensors, a subarray size of $L=3$ and temporal averaging is set to $K=1$. Here the spatial sum is carried out before the temporal sum. To reduce arithmetic operations we may:
%
\begin{enumerate}
\item exploit the fact that $\eR$ is Hermitian positive semi-definite to compute only one half of it,
\item avoid redundant multiplications, and
\item avoid redundant summations by first computing the upper row of $\eR$ and then add/subtract to these iteratively to find the diagonals of $\eR$.
\end{enumerate}
%
\ifPeerReview
\begin{figure}[!t]\centering
\includegraphics[width=.8\linewidth]{gfx/buske5.eps}
\else
\begin{figure}[!t]\centering
\includegraphics[width=.9\linewidth]{gfx/buildR-breakdown.eps}
\fi%
\caption{Arithmetic optimization of computing $\eR$: Relative reduction in arithmetic complexity compared to the initial implementation shown in Fig. \ref{mvdr_complexity} (higher is better) . Note how the arithmetic count is reduced by a factor 4-10 in the memory optimized case, and by a factor 6-22 in the instruction optimized case.}\label{mvdr_complexity_optimized}
\end{figure}

To study the effect of these optimizations we altered the complexity formulas to take them into account. Exploiting symmetry and performing iterative summing (step 1 and 3) is always desirable, but avoiding redundant multiplications (step 2) comes at the cost of increased memory consumption. When all optimizations are applied we call it a "minimum instructions" solution, and when step 2 is left out we call it a "minimum instructions" solution. Fig. \ref{mvdr_complexity_optimized} compares the two solutions, and formulas may be found in appendix \ref{mvdr_formulas}. Here the complexity curves are relative to the reference implementation in Fig. \ref{mvdr_complexity}. Both solutions reduce the complexity considerably, and recomputing the multiplications does not make that much of a difference. We selected the memory minimized version since our algorithm is severely memory bound, as we will see next.

% \begin{table}[b]\centering%\normalsize
% \begin{tabular}[c]{l r@{:}  l}\hline
% \rowcolor{tabBlue}\bf Memory type & \multicolumn{2}{>{\columncolor{tabBlue}}c}{\bf BW$_\text{memory}$/BW$_\text{arithmetic}$} \\\hline
% Global memory &\hspace{30pt} 1 & 30 \\
% Shared memory & 1 & 4 \\
% Registers & $>$3 & 2~\cite{Vasilyy}
% \end{tabular}
% \caption{Nvidia Quadro 6000 bandwidth ratios.}\label{tabbandwidth}
% \end{table}

\begin{table}[b]\centering%\normalsize
\begin{tabular}[c]{l r r r@{}  l}\hline
\rowcolor{tabBlue} & \multicolumn{1}{>{\columncolor{tabBlue}}c}{\bf B$_\text{arith}$} & \multicolumn{1}{>{\columncolor{tabBlue}}c}{\bf B$_\text{mem}$} & \multicolumn{2}{>{\columncolor{tabBlue}}c}{\bf B$_\text{mem}$/B$_\text{arith}$} \\\hline
Arithmetic & 1.03 Tflop/s & & &\\
Global memory & & 36 Gfloats/s & \hspace{30pt} 1 &:30 \\
Shared memory & & 257 Gfloats/s & 1 &:4 \\
Registers & & $>$1.5 Tfloats/s & $>$3 &:2~\cite{Vasilyy}
\end{tabular}
\caption{Nvidia Quadro 6000: Memory throughput, $B_{\lowercase{\text{mem}}}$, compared to arithmetic throughput, $B_\text{arith}$.}\label{throughputs}
\end{table}

\ifPeerReview
\begin{figure}[!t]\centering
\includegraphics[width=.8\linewidth]{gfx/buske6.eps}
\else
\begin{figure}[!t]\centering
\includegraphics[width=.9\linewidth]{gfx/mvdr_implementation.eps}
\fi%
\caption{MVDR implementated on a GPU. We do this in 3 steps, where each step process the full image before moving on to the next step. \emph{Step~1}: The sample covariance matrix $\eR$ is formed by threads running along its diagonals. This allows spatial averaging to be implemented in a computationally efficient manner and minimizes inter-thread communication. \emph{Step~2}: $\eRi\1$ is computed using the heavily optimized batched linear equation solver from Nvidia~\cite{Nvidia}. \emph{Step~3}: The beamformer output $z$ is computed in a straight forward fashion using $L$ threads that first sum the subarrays and then apply the MVDR weighting function. A single thread finally sum all the channels up.}\label{mvdr_implementation}
\end{figure}

\subsubsection{Arithmetic Throughput}

When we compute $\eR$, we very rarely perform more than 1-3 floating point operations for every float read or written to memory. Unfortunately, the GPU prefers kernels to be more arithmetically intensive than this. This can be inferred from Table \ref{throughputs}, where the peak bandwidth of the three types of GPU memory are compared to the peak arithmetic throughput (see appendix \ref{throughput} for derivations). Global memory (RAM), in which all data must reside at some point, is at best only able to move one float for every 30 floats processed by the CUDA cores, and potentially much worse if care is not taken to use "coalesced" memory access patterns (that avoid memory bank conflicts). This is why the usage of global memory should be minimized. Shared memory, on the other hand, is a very fast level 1 cache that is shared by all threads on an SM. At peak performance it is almost able to keep up with the computing units, but to obtain this performance we must make sure that:
\begin{enumerate}
\item the data needed to compute a pixel can fit in shared memory, while
\item the access patterns we use promote maximum bandwidth.
\end{enumerate}
 %Further alleviation is also possible by exposing instruction level parallelism within a thread, as this promotes the use of registers with a bandwidth at least 6 times higher than that of shared memory~\cite{Vasilyy}\todo{so why don't we? rephrase}.

These challenges are very closely linked. Since the Quadro 6000 architecture is of compute capability 2.0, it has 48\;kB shared memory (L1 cache) and 128\;kB registers per streaming multiprocessor (SM)~\cite{Nvidia2012}. This memory is shared by all active threads on that SM, a number that should be no less than 768~\cite{Nvidia2012a}. This is to expose a sufficient level of data parallelism to ensure that memory latency is completely hidden (in accordance with Little's law). If we divide the shared memory evenly on all 768 threads we find that each thread should store no more than 8 single precision complex numbers in shared memory, and 24 stored in registers. This should make it apparent why computing a single pixel per thread is a bad idea, and why we need to keep each thread as light on memory consumption as possible.

% What about the GPU implementation? This one was tricky. Balance load, avoid inter-thread dependencies when performing iterative summations, keep thread memory consumption at a minimum, and promote collective access patterns (coalesced reads/writes) to global memory.

In Fig. \ref{mvdr_implementation} we illustrate how we get around these challenges. First we make each SM compute entire range lines of pixels, then load the subset of $\x$ that all these pixels depend upon into shared memory. This lets us perform temporal averaging without having to read additional channel data.

\ifPeerReview\else
\begin{figure}[!t]\centering
\includegraphics[width=.9\linewidth]{gfx/plot_holmengraa_L16_Navg1.eps}
\caption{HISAS sidescan sonar (SSS) image of the shipwreck Holmengraa that lies on a slanted seabed at 77\;m depth outside of Horten, Norway. The image was processed with $M=32$, $L=16$, $K=1$ and $d=1\%$. Prior to display the image was linearly upinterpolated by a factor 2 in azimuth making its total size 1.46 megapixels (MP). Note how MVDR improves edge definition and reduces noise in shadow regions.}\label{holmengraa}
\end{figure}
\fi%

Second, we assign $L$ threads per pixel to traverse the diagonals of $\eR$. On the top row a full computation is carried out, then that row is saved back to global memory following a coalesced access pattern that maximizes global memory bandwidth. For subsequent rows the threads move along the diagonals while performing iterative summations; the result from the previous element on the diagonal is updated by adding and removing the correlation coefficients that enters and exits the sum, respectively. To minimize memory consumption, we compute the coefficients again every time we need them. When a thread has finished up a diagonal, we have them wrap around to compute one of the diagonals in the lower triangular of $\eR$. Since $\eR$ is conjugate symmetric, the values in the leftmost column is obtained by a complex conjugate copy of the relevant value in the first row. This is slightly non-optimal as it causes divergent branching, but is only needed because the solver is Gauss Jordan based and thus require a 
full $\eR$. Combined these steps balance the load evenly on all threads, is almost completely free of arithmetic redundancy, and consumes less memory.

% Hence, whenever $N_\text{avg}>1$ the sliding window approach will demand less instructions at the expense of a slightly higher memory consumption.

% However, an advantage with this approach is that the summations can be carried out in any order, and hence this form exhibits great deal of parallelism. Can we optimize the algorithm without sacrificing parallelism?

% - Order of summations - beamspace capable

% 
% In the upcoming results, we have beamformed pre-delayed data. This is a large dataset, and the latency experienced when copying it from the \gls{CPU} side to the \gls{GPU} was significant. However, in a practical system only the channel data should be transferred as the delaying of data can effortlessly be performed by the \gls{GPU}. The remaining latency can be hidden as long as the \gls{GPU} is kept busy, hence the data transfer times will not be reported in the upcoming results.
%

% \begin{itemize}
% \item Arrange order of summations to reduce data dimensions as early as possible.
% \item Sliding window in time/space to reduce instructions at the cost of extra memory consumption. Only pays off to do this over subarrays.
% \end{itemize}
% 
% \item Between GPU and CPU: Several concurrent blocks (contexts) per SM.
% \item Job queue (how does this work)
% \item Register pressure and shared memory. Thread context must fit in local memories.
% \item Warp computational intensity should be high to make up for the memory cost.

% increasing performance hit several walls:
% \begin{itemize}
% \item memory couldnt keep track - remedy cache.
% \item instruction parallelism - complex to identity
% \item power consumption cubic with frequency
% GPUs:
% \begin{itemize}
% \item lower frequency, less power
% \item no instruction parallelism
% \item massive amounts of lightweight cores
% \end{itemize}\end{itemize}

% \cite{Owens2007}\todo{Find a good spot for these}

\subsection{Solving $\hat{\boldsymbol{R}}$$\;\!^{-1}$$\mathbf{1}$}

As intuition might suggest the matrix product $\eRi\1$ can be carried out by first inverting the matrix $\eR$, then performing the inner product with $\1$. However, one can also solve the linear equation $\hat{\boldsymbol{R}}\boldsymbol{b} = \1$ for $\boldsymbol{b}$, where $\boldsymbol{b} = \eRi\1$. This is the preferred approach since one vector In terms of arithmetic count the latter is preferred , and when comparing direct implementations, the latter appears to be the preferred option.

An important thing to note, however, is that unlike the problems most GPU libraries tries to solve we do not attempt so solve large linear equations or invert large matrices; our matrices are small, but we have a very large number of them. Fortunately Nvidia recently released a highly optimized batched linear equation solver tailored for this particular task. It is Gauss Jordan based and supports partial pivoting and complex numbers. The only downside to using it in our application is that it does not exploit the Hermitian property of our sample covariance matrix. A better choice might be a solver based on Cholesky decomposition. These are designed for Hermitian positive semi-definite matrices, and require roughly half the arithmetic operations compared to a Gauss Jordan solver.

% \emph{Computing} $\eRi[n]\1$ (from (\ref{weights})). While intuition may suggest that this step is carried out by inverting $\eR$, it is better to solve the linear equation $\R\beta = \1$ for $\beta$ instead, which gives us $\beta = \eRi\1$ directly. We have tested various solvers for this task, both in-house and proprietary implementations, and achieved the best performance by using an unofficial batch solver from nVidia. Inverting $\eR$ is by far the most computationally intensive task, and a key area of focus for further improvements.

\ifPeerReview\else
\begin{figure}[!t]\centering
\includegraphics[width=.9\linewidth]{gfx/code_assessment.eps}
\caption{Execution time of an arithmetic-only and a memory-only version of the MVDR code. A dataset from an $M=32$ array was processed for all $L$ using $K=1$, and the mean execution time for a 1\;MP image was used here. From this plot we can infer that the kernel building $\eR$ is memory bound, as the time the kernel spends performing memory transactions is higher than the corresponding time it spends carrying out arithmetical operations. Furthermore, when the total runtime is larger than the restricted kernels this can largely be attributed to latency, which we can see that building $\eR$ suffers from with the chosen parameters.}\label{code_assessment}
\end{figure}
\fi%

\subsection{Computing $z$}

The beamformer output $z$ is computed on a per-pixel basis by first computing the MVDR weights $\w$, which are merely scaled versions of $\eRi\1$ (\ref{weights}). Then the sum in (\ref{finalZ}) is found by assigning a group of $L$ threads to respective elements of $x_l$, which then proceed to accumulate these for all $N_L$ subarrays. The resulting data vector is finally multiplied with the weight vector using the same threads, and then a single thread is used to sum these products to obtain $z$.

\subsection{Summary of GPU implementation}

The GPU implementation can be summarized as follows:
\begin{itemize}
\item Pixels are processed in batches.
\item Multiple threads per pixel.
\item $\eR$ is computed directly from raw data to minimize memory footprint.
\item Inversion step is performed by Nvidias batch linear equation solver.
\item Design makes efficient use of cache and data transfers are kept to a minimum.
\end{itemize}

Compared to our CPU implementations, where:
\begin{itemize}
\item Pixels are processed one by one.
\item Single thread per pixel.
\item Computation of $\eR$ is optimized for minumum number of instructions, as in \ref{mvdr_build_R}. Only upper half of $\eR$ is computed (see next point).
\item Inversion step carried out by a custom inline Cholesky based solver that only requires the upper triangular of $eR$.
\item Design has generally not been optimized for efficient use of memory.
\end{itemize}

%  The subarray data can, as here, be reduced to coincide in length with the subarray weights, or the weights can be extended to $M$ in size and applied directly on $\x[n]$ and summed.

% \item \emph{Computing the spatial covariance matrix} $\eR[n]$ (\ref{spatialR}-\ref{finalR}). A group of threads are created that slide along the diagonals of $\eR$. In this way we exploited the fact that entries on the diagonals overlap across subarrays and time, and keep the numbers of both data reads and writes at a minimum.

% There are essentially two ways to accelerate an algorithm on parallel hardware. One is to identify independent instructions for each thread, and run these in in parallel, and the other is to support the parallel execution of as many threads as possible. This is referred to as instruction level parallelism (ILP) and data level parallelism (DLP), respectively. Most general purpose processors, such as a CPUs, support both DLP by featuring multiple cores with single-data multiple-data (SIMD) capabilities, and ILP by through means such as branch prediction and out-of-order execution.

% GPUs, on the other hand, does not attempt to do ILP, but use all available transistors to support massive DLP. This leads to designs such as the nNvidia GeForce GTX 580. Using their own terminology, it is comprised of 16 ``streaming'' multiprocessors (SMs), each having having 32 CUDA\todo{make sure CUDA is introduced} cores that execute a common program called a kernel. %An SM is a single-program multiple-data (SPMD) processor, meaning that all the CUDA cores on an instruction unit, L1 cache and registers is shared by all the CUDA cores on that SM.  %
% each with 48kB of L1 cache (shared memory) and 128kB of register memory.

% In short, designing for a GPU involves finding the optimal balance between hiding memory latency (high occupancy) and resource utilization.
% 
% blocks in grid - high enough so that all SMs have atleast one block to execute. Optimally in the thousands to scale for future devices, and to allow an SM to switch between blocks to keep hardware busy when running synchronizing threads.
% 
% threads in block - min 64. 128-256 better starting point.

% Instead of having a few complex core that can boost massive single thread performance, designers chose to create hundreds of lightweight cores that each is much powerful than a single CPU core, but combined boasts a massive artithmetical performance.
% 
% \begin{itemize}
% \item Lots of cores, little memory for each.
% \item Few data transfers, lots of computations.
% \item Data transfers are moved around by hardware that can operate asynchronously from the compute units.
% \item Memory latency hidden as long as it takes longer to compute than to move data around.
% \item Tasks can be queued on SM (blocks) and on CPU to hide internal and external memory latencies.
% \end{itemize}
% 
% Vast amount of processing power, which are easily accessible but hard to fully harness. This is because in the pursuit of an ever increasing amount of arithmetical computing cores, the 
% 
% Unlike a CPU, a GPU 
% 
% If we compare a GPU with a CPU, 
% 
% When implementing an algorithm on a GPU, it is important to be aware
% 
% People friendly:
% \begin{itemize}
% \item More and more compute units, but less memory for each.
% \item Need to keep each compute unit busy.
% \item Few data transfers, lots of computations.
% \item Data transfers are moved around by hardware that can operate asynchronously from the compute units.
% \item Memory latency hidden as long as it takes longer to compute than to move data around.
% \item Tasks can be queued on SM (blocks) and on CPU to hide internal and external memory latencies.
% \end{itemize}
% 
% The ``fast'' memory of the GPU is the registers and cache. All threads executed 
% 
% 
% \begin{itemize}
% \item Hiding memory access.
% \begin{itemize}
% \item Between GPU and CPU: Several concurrent blocks (contexts) per SM.
% \item Job queue (how does this work)
% \end{itemize}
% \item Register pressure and shared memory. Thread context must fit in local memories.
% \item Warp computational intensity should be high to make up for the memory cost.
% \item Optimized for small L, high M, hardly any time averaging. A typical sonar scenario.
% \end{itemize}
% - Split program into equally sized portions - load balancing
% - Some portions depend on other potions already being carried outer - sequential dependencies
% - Effort timed in order to finish at the same time - synchronization
% 
% - task level parallelism - compute images independent of each other
% - data level parallelism - compute pixels independent of each other

% - Compute bound vs. bandwidth bound

\section{Images and Benchmarks}\label{images_and_benchmarks}

\ifPeerReview
\begin{figure}[!t]\centering
\includegraphics[width=.8\linewidth]{gfx/buske7.eps}
\caption{HISAS sidescan sonar (SSS) image of the shipwreck Holmengraa that lies on a slanted seabed at 77\;m depth outside of Horten, Norway. The image was processed with $M=32$, $L=16$, $K=1$ and $d=1\%$. Prior to display the image was linearly upinterpolated by a factor 2 in azimuth making its total size 1.46\;MP. Note how MVDR improves edge definition and reduces noise in shadow regions.}\label{holmengraa}
\end{figure}
\fi%

\ifPeerReview
\begin{figure}[!t]\centering
\includegraphics[width=.8\linewidth]{gfx/buske8.eps}
\else
\begin{figure}[!t]\centering
\includegraphics[width=.9\linewidth]{gfx/benchmark.eps}
\fi%
\caption{\protect MVDR benchmarks. A 1\;MP image from a $M=32$ channel array was processed for all $L$, and for $K\in\{0,1,2\}$.\newline \emph{Top:} The time the GPU spent on building $\eR$, solving $\eRi\1$, and computing $z$. Note the major speedup of building $\eR$ when compared to the complexity plot in Fig. \ref{mvdr_complexity}.\newline
\emph{Bottom:} Compared to a reference MATLAB and single thread C implementation running on a CPU the GPU offered a speedup of 2-3 orders of magnitude, but these numbers are somewhat misleading. If the C implementation was properly optimized we expect the GPU to be no more than a factor 5-10 faster, even if its theoretical peak performance is ~20 times higher than that of the CPU.}\label{benchmarks}
\end{figure}

\ifPeerReview
\begin{figure}[!t]\centering
\includegraphics[width=.8\linewidth]{gfx/buske9.eps}
\caption{Execution time of an arithmetic-only and a memory-only version of the MVDR code. A dataset from an $M=32$ array was processed for all $L$ using $K=1$, and the mean execution time for a 1 megapixel (MP) image was used here. From this plot we can infer that the kernel building $\eR$ is memory bound, as the time the kernel spends performing memory transactions is higher than the corresponding time it spends carrying out arithmetical operations. Furthermore, when the total runtime is larger than the restricted kernels this can largely be attributed to latency, which we can see that building $\eR$ suffers from with the chosen parameters.}\label{code_assessment}
\end{figure}
\fi%

\ifPeerReview
\begin{figure}[!t]\centering
\includegraphics[width=.8\linewidth]{gfx/buske10.eps}
\else
\begin{figure}[!t]\centering
\includegraphics[width=.9\linewidth]{gfx/code_assess_flops.eps}
\fi
\caption{\protect Code efficiency. An estimate of the number of floating point operations per second (Flop/s), found by dividing the theoretical complexity curves by actual run-times. This is a crude measure as it does not include any other instructions than the actual arithmetic operations in the MVDR computation.}\label{code_assess_flops}
\end{figure}

\ifPeerReview
\begin{figure}[!t]\centering
\includegraphics[width=.8\linewidth]{gfx/buske11.eps}
\else
\begin{figure}[!t]\centering
\includegraphics[width=.9\linewidth]{gfx/benchmark_boston.eps}
\fi
\caption{\protect MVDR benchmarks from Boston HPC centre with the new high-end Nvidia K20 Kepler GPU. The exact same scenario and code as in Fig. \ref{benchmarks} was used here. With no code alterations the performance was only improved marginally compared to running on the Quadro 6000.}\label{benchmarks_boston}
\end{figure}

% In a practical system it should be moved to the GPU as well. Therefore, in the upcoming benchmarks, CPU computation time and data transfer to the GPU is neglected.

To demonstrate the imaging capability of the MVDR beamformer, we have processed experimental datasets from the $M=32$ element Kongsberg Maritime HISAS1030 sonar mounted on a HUGIN AUV~\cite{Hansen2009}. HISAS1030 is a high resolution synthetic aperture sonar with an array length of 1.2\;m, operating frequency of 100\;kHz, and bandwidth of 30\;kHz. The element size and spacing is 2.5$\lambda$ and the opening angle is 25$^\circ$. To produce the image shown in \Fig{holmengraa} the sonar was operated in sidescan mode. The studied object is the 1500 dwt oil tanker wreck Holmengraa. It is 68\;m long and 9\;m wide, and lies at a slanted seabed at 77\;m depth outside of Horten, Norway. The 1\;MP MVDR image were here processed with parameters $L=16$, $K=1$, and $d=1\%$.

The computational performance of our implementation was first assessed on a test system with a quad-core Intel Xeon E5620, 64\;GB of RAM and an Nvidia Quadro 6000 card. The results were obtained by processing a 1\;MP image from the data from a 32 channel array, for all subarray sizes $L$, and for $K\in\{0,1,2\}$. Run-time measurements are shown in \Fig{benchmarks}, the run-times of memory-only and arithmetic-only GPU kernels are depicted in \Fig{code_assessment}, and an estimate of computation efficiency is presented in \Fig{code_assess_flops}. We were also granted a test drive at the Boston HPC centre on a machine with an Intel Xeon E5-2670, 32\;GB RAM and an Nvidia K20. The results from this run are shown in \Fig{benchmarks_boston}.

As seen in the run-time comparisons presented in \Fig{benchmarks}, the GPU method is now 2-3 orders of magnitude faster than the C implemtation we started out with. The bottleneck in the final design is now the inversion step, which is typically 5 times slower than the build step. In most cases the processing speed of the Quadro 6000 is above 1\;MP/s, and this was only improved by a factor 1.5-2 when run on the new K20 GPU. The benchmarks of our memory-only and arithmetic-only kernels show that the kernels spend roughly the same time on both these tasks, so optimising only one of these further will have marginal effect. All GPU kernels were compiled with \texttt{nvcc} at optimization level \texttt{O2}. Excluded from these benchmarks is the data transfer time from CPU to GPU, which account for 2-20\% of the total processing time. We believe it makes little sense to include them since it keeps getting easier perform these data transfers in parallel by offloading the task to the Direct Memory Access (DMA) 
controllers present on modern GPUs. Furthermore, the data rates in active sonar are relatively low compared to the bandwidth available for these transfers. 
% 
% We have tested our \gls{GPU} implementation of the \gls{MVDR} beamformer experimental datasets from the 32 element Kongsberg Maritime HISAS1030 sonar\todo{shouldn't we state that it's SAS?} being attached to the HUGIN \gls{AUV}~\cite{Hansen2009}. To obtain the studied object, the 1500 dwt oil tanker wreck Holmengraa, the sonar was operated in sidescan mode to image (\Fig{hlmengraa}). Holmengraa is roughly 68\;m long and 9\;m wide, and lies on a slanted seabed at 77\;m depth. The image was processed with \gls{MVDR} parameters $L=16$, $N_\text{avg}=3$, and $d=1\%$, which proved to be a reasonable selection for this scenario.
%

\section{Discussion}\label{discussion}

In accordance with previous studies, \Fig{holmengraa} demonstrates the MVDR beamformer's ability to produce images with suppressed interference and improved detail resolution. Compared to the DAS beamformer, the ship's edges appear sharper and the shadows less noisy. In this scenario the MVDR's performance was not particularly sensitive to parameter adjustments. Similar performance was obtained with arbitrary combinations of $L\in\{12,16,24\}$, $K\in\{1,2,3\}$ and $d\in[0.01,0.05]$, and no adjustments had to be made when processing other parts of the scene.
%But while \gls{DAS} processed this 1.3\;Mpixel image in milliseconds, our MATLAB implementation\todo{never mentioned matlab vefore} of MVDR beamformer needed minutes.

As observed in \Fig{benchmarks} the combination of minimizing arithmetic operations and implementing the MVDR on a GPU lead to a significant improvement in processing speed. Compared to a MATLAB and single thread C implementation an order 2 and 3 speedup was achieved, respectively. Note that we do not consider this comparison fair. In fact, the theoretical peak throughput of this GPU is roughly 20 times higher than that the CPU in question, meaning the potential of the CPU was far from reached in our initial implementations. The key reason for this is that the GPU design process pixels in batches, which allows us to minimize data transfers and maximize the use of fast memory cache. In other aspects the designs are similar, both CPU implementations compute $\eR$ efficiently, and they make use of either an optimized custom Cholesky based solver or one from the Intel Math Kernel Library (MKL). Unfortunately we had little time to write a custom batch based solver and covariance builder for the CPU, but we expect 
the speed difference would be 5-10 times if we had compared two such designs.

Note from \Fig{benchmarks} how the building of $\eR$ now only takes up a fraction of the total processing time, and recall that it was the other way around when we presented the theoretical complexity curves in \Fig{mvdr_complexity}. Also, the benchmark curves for building $\eR$ now seem linear. The main reason for this is that the optimization step negated much of the extra complexity introduced by averaging, i.e. reduced the complexity from O($N_KN_LL^2$) to less than O($L^2$). Another reason for the run-time to appear more like O($L$) is likely that our design makes better use of the GPU's resources when there is more processing involved per pixel. The inversion step, on the other hand, gains less from being implemented on a GPU. This is because its nature is less data parallel. In particular, the back substitution step involved in its computation is mostly sequential. This difference be can observed in \Fig{mvdr_complexity}. Alternative solvers, such as one based on Cholesky decomposition that exploits 
the Hermitian property of $\eR$, can in theory reduce complexity by a factor two, but we question whether this result can be obtained in practice using a GPU. The CPU implementations perform Cholesky based inversion, but this does not explain why the C and MATLAB implementation have a total run-time that is linearly dependent on $L$. We believe this is because these implementations are dominated by the calculation of the covariance matrix and data movement, and not by the inversion step.

It is important to know whether a GPU design is limited by memory bandwidth or arithmetical throughput. To measure this  we made two versions of the MVDR kernel, one that only performs arithmetic operations, and one that only performs memory operations. Then we benchmarked these kernels and compared their run-times to the total run-time (\Fig{code_assessment}). A first thing to note is how these kernels seem equally occupied performing memory and arithmetic operations. This is a good sign, but since we consistently minimized memory consumption at the expense of some extra computations when building $\eR$ , we know it to be bound by memory bandwidth. It also has a problem with latency, which can be inferred from the total run-time being significantly higher than that of the two single-function kernels. Since the GPU hardware can carry out memory and arithmetic operations concurrently and independently, this gap indicates that the GPU sometimes does ``nothing''. In our case this is due to synchronization hold-
ups when performing temporal averaging, which is carried out in a sequential manner.

Even with all our efforts we were only able to obtain processing rates of 40\;MFlop/s on average in the desirable range of subarray sizes (\Fig{mvdr_complexity}). This is mainly due to the inversion step, since we for the most part obtained more than 100\;MFlop/s when building $\eR$. While these numbers are likely slightly underestimated, they are still far away from the Quadro 6000's peak of 1\;Flop/s. Again, we believe this is due to memory bandwidth. This belief is further supported by the test results from running the code on the new K20 Kepler card, where we saw a modest factor 1.5-2 speedup. Although the Kepler card peaks at 3.5-4\;Tflop/s, its shared memory bandwidth is only approximately twice that of the Quadro 6000. However, it is likely that the Kepler card would perform better had we optimized our design for it. For scientific computing the Kepler boards are considered more attractive, with the introduction of features such as kernels calling kernels, more registers, more shared memory, and 
better mechanisms for hiding or removing data transfers.

% Interesting to note is also that solving $\eRi\1$ remains the main bottleneck in the GPU implementation, and it gets even worse for larger $L$'s. However, this is a general Gauss Jordan solver which does not exploit symmetry properties of $\eR$, so creating, e.g. a batch based Cholesky solver should improve the runtimes further. Both the C and MATLAB implementation in this benchmark take this symmetry into account, which partially explains why the complexity curves of these implementations are different from that of the GPU implementation.

% If the HISAS1030 was attached to a platform moving at 1.8m/s, for which the maximum range is 250\;m~\cite{Hansen2010}, the pulse repetition frequency could be set to 3\;Hz. We could further assume a complex sampling frequency of 40\;kHz to accomodate the  HISAS1030's bandwidth of 30\;kHz, and beamform $M=32$ lateral beams. The throughput required for real-time processing of these images would then be 1\;MP/s. As we have seen, this is something our GPU implementation can handle if $L$ is not too large.\todo{droppe denne kanskje}%, leaving the \gls{CPU} free to take on other assignments.

% What may this performance be used for? Let us assume that the 32 element, 1.2\;m long HISAS1030 array is mounted on a platform moving at 1.8\;m/s. Then the pulse repetition frequency (PRF) could be set to 3\;Hz to ensure critical along-track sampling, 
% To illustrate what this performance might be used for, consider a moving platform travelling at 1.8\;m/s. 
% If the sonar system is Maximum range 
% There is a limit to the along-track sampli
% To avoid along track undersampling, the

% HUGIN: 1.8m/s - range 250m \\
% fs = 100kHz*30\%*$\frac{4}{3}$ = 40kHz \\
% $N_{range px} = \frac{2\;250m}{1500m/s}40kHz = 13.3kpx$ \\
% $\times 32$ beams = 426kpx \\\\
% 
% PRF = $\frac{1500m/s}{2 250m} = 3Hz$ \\
% TP = 215px 3Hz = 1.28MP/s \\
% 4 arrays: 1.28MP/s * 4 = 5.12MP/s
% 
% 
% \begin{itemize}
% \item Need for speed: HUGIN 4 banks of 32 elements, can be processed faster than the ping repetition rate, with margins to spare.
% \end{itemize}

\section{Conclusion}\label{conclusion}

The MVDR beamformer is an algorithm capable of producing images with improved detail resolution and contrast compared to conventional DAS beamforming. The downside is its inherent need for robustification and the high computational complexity associated with estimating and inverting the spatial covariance matrix.

We have shown that for systems containing up to 32 channels the problem can be largely mitigated by building the covariance matrix in a clever way, and by making use of the massive computational power available in modern GPUs. We were able to improve upon the run-time of a single thread C-implementation by roughly two orders of magnitude. For most choices of parameters the GPU was able to create images at $\sim$1\;MP/s at an average data processing rate of ~40\;GFlop/s. This is less than 5\% of the peak performance of the GPU, but we believe it to be near optimal given the constraints of memory bandwidth and the sequential nature of some parts of the MVDR implementation.

%  This performance is sufficient for computing properly sampled full-coverage sectorscan images from the HISAS1030 sonar in real-time.

All in all, the MVDR maps well to the GPU since the computations involved are independent on the pixel level, and partially also within each pixel. The GPU allows MVDR to be used in real-time processing of sonar data, and makes the MVDR a viable alternative to conventional methods in practical systems.

% \begin{figure}[t]
% \centering
% \includegraphics[width=\linewidth]{gfx/benchmark_tagged.eps}
% \caption{Realtime requirement for sectorscan imaging.}\label{real-time_sectorscan}
% \end{figure}

%%%%%%%%%%%%%%%%%%                              ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
% DOCUMENT APPENDICES %
%%%%%%%%%%%%%%%%%%                              ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

% \titleformat{<command>}[<shape>]{<format}{<label>}{<sep>}{<before>}[<after>]

% \titleformat{\section}[hang]{\bf}
% {\thesection.\enspace}%\thesubsubsection}%  {\footnotesize \enspace \emph{Sec.}  }
%    {0pt}{\MakeUppercase}[]
% \titlespacing*{\section}{0pt}{2\lineheight}{\lineheight}

% {-2ex plus -.5ex minus -.2ex}{1.0ex plus .2ex minus .2ex}

% \renewcommand\section[1]{{\Large\bf\MakeUppercase #1}}
% \renewcommand\section*[1]{{\Large\bf\MakeUppercase #1}}

% 
% 
% 
% 
% 
% 
% \section{Introduction}
% 
% \todopar{Case:
% \begin{itemize}
% \item UFFC(?)
% \item Ultrasound. Arrays up to 128 channels. Processing power vital. What's the best way to implement 
% \item Other articles evaluates Capon image quality, and some have shown that speedups can be achieved by reducing the complexity of the traditional Capon algorithm or by implementing specific versions on a GPU.
% \end{itemize}
% }
% 
% \begin{itemize}
% \item As briefly as possible, introduce beamforming.
% \item Adaptive beamformer's potential lies in its ability to suppress interference power
% \item Why adaptive beamformers struggle in active sonar systems. Correlated noise, robustification kills the adaptive potential. Quite computationally intensive. Constraints must be applied in one way or another - parameters must be tuned.
% \item The three main ways to implement Capon \cite{Capon1969} is \todo{not entirely true?}
% \begin{itemize}
% \item Traditional way: Build $\eR$ and solve $\w = \frac{\Ri\a}{\a\H\Ri\a}$. 
% \item Iterative methods: Woodbury, Conjugated gradients
% \item Beamspace
% \item LCA (or other methods, as reference)
% \end{itemize}
% fs = 100kHz*30\%*$\frac{4}{3}$ = 40kHz \\
% $N_{range px} = \frac{2\;250m}{1500m/s}40kHz = 13.3kpx$ \\
% $\times 32$ beams = 426kpx \\\\
% 
% PRF = $\frac{1500m/s}{2 250m} = 3Hz$ \\
% TP = 215px 3Hz = 1.28MP/s \\
% 4 arrays: 1.28MP/s * 4 = 5.12MP/s
% 
% 
% 
% 1Tflops (ops/cycle?)
% 144GB/s global
% 1TB/s shared
% x6 regs
% 
% latency - arithmetic (20 cycles) and memory (400+ cycles)
% 
% Little's law - needed parallelism = latency * throughput
% arithmetic: 23 cycles * 32ops/SM/cycle = 576 ops/SM parallelism
% memory: <800 cycles * <144GB/s=125B/cycle = <100kB in the flight
% 
% On GF104: must have some ILP to get >0.66 of peak:
% - 48cores/SM, one instruction is broadcast across 16 cores
% - need 3 instructions per cycle
% - but only have 2 warp schedulers
% - instead, can issue 2 instructions per warp in the same cycle
%

\section{Appendices}

\subsection{MVDR Complexity Formulas}\label{mvdr_formulas}

To estimate the number of floating point operations needed to MVDR beamform a single pixel, we formed expressions that accumulate the number of complex arithmetic operations found in the MVDR process. From observing the generated assembly code we inferred that each complex addition and multiplication would require $\Oa = 2$ and $\Om = 6$ floating point operations, respectively. The formulas are listed for each step below for reference.

\emph{Building $\eR$}. The initial complexity of this step can be inferred directly from (\ref{spatialR}) and (\ref{finalR}):
\begin{align}
O_{\underset{\text{initial}}{\text{Build $\eR$}}} &= \underbrace{\Om\Nk\Nl L^2}_\text{Multiplications} + \Oa(\Nk+\Nl-2)L^2 + O_\text{dload},
\end{align}
where $O_\text{dload} = (2L-1)\Oa + \Om$ is the minor cost of performing diagonal loading. If we apply the optimization strategies discussed in section \ref{computing_eR} we can arrive at the following instead:
\begin{align}
O_{\underset{\text{min arith}}{\text{Build $\eR$}}} &= \underbrace{\Om\frac{M+\Nl}{2}L}_\text{Multiplications} + \underbrace{\Oa(\Nk-1)(\Nl-1)L}_\text{First row additions} \nonumber\\
&+ \underbrace{\frac{(L-1)(L-2)}{2} \bigg[ 2\Oa + 2(\Nk-1)\Oa \Big]}_\text{Iteration additions} \nonumber\\
&+ O_\text{dload}
\end{align}
Of the solutions discussed, this is the least expensive in terms of arithmetic instructions. However, if memory bandwidth is a limiting factor a better solution is to recompute multiplications where they are needed:
\begin{align}
O_{\underset{\text{min mem}}{\text{Build $\eR$}}} &= \underbrace{\Om\Nk\Nl{}L}_\text{First row multiplications} + \underbrace{O_a(\Nk-1)(\Nl-1)L}_\text{First row additions} \nonumber\\
&+ \underbrace{\frac{(L-1)(L-2)}{2}\bigg[ 2\Oa + 2(\Nk-1)\Oa + 2\Nk\Om \bigg]}_\text{Iteration multiplications and additions} \nonumber\\
&+ O_\text{dload}.
\end{align}

\emph{Solving $\eRi\1$} is achieved by using a batched Gauss Jordan solver with support for complex numbers and partial pivoting. Its complexity - with partial pivoting excluded - can be expressed as: 
\begin{align}
O_{\text{Solve $\eRi\1$}} = \sumb{r=0}{L} \bigg[&\underbrace{\big(L-r\big)\big((L-r+2)\Oa + (L-r+3)\Om\big)}_\text{Reduction}\nonumber\\
+ &\underbrace{(r-1)\Oa + r\Om}_\text{Backsubstitution}\bigg],
\end{align}
where $r$ is a running variable $r$ that indexes rows in the augmented matrix $\bmat{\eR|\1}$.

\emph{Computing $z$} is very simple once the covariance matrix $\eR$ is built and inverted, and has an near negligible impact on performance: 
\begin{align}
O_\text{Compute $z$} = \Oa(2L-2)2 + \Om 3 L.
\end{align}

\subsection{GPU Throughput}\label{throughput}

In the context of determining whether an implementation is computationally bound or memory bound, one should first compare the target platform's sustained computational throughput to sustained memory throughput. Let us start with the former.

The Quadro 6000 has 32 CUDA cores per SM, each operating at a rate of 1148\;MHz and being able to perform 2 floating point operations (flop) per clock cycle when multiply-add instructions are used. Theoretical peak arithmetic throughput is then given as:
\begin{align}
\text{B}_\text{arith} &= 2\cdot1148\;\text{flop/s/core}\cdot 32\;\text{cores/SM}\cdot 14\;\text{SMs}\nn
&= 1.03\;\text{Tflop/s}.\label{flops}
\end{align}
Now let us compare this to the memory throughput. The ``global'' GDDR5 memory bus is 384\;bit wide, and operates at 3\;Ghz where 2 bits are sent every cycle. Its peak bandwidth is then:
\begin{align}
\text{B}_\text{gmem} &= \frac{2\cdot 3\;\text{Gbit/s}\cdot384\;\text{bit}}{8\;\text{bit/byte}}\nn
&= 144\;\text{GB/s}\ (36\;\text{Gfloats/s}).\label{bwglobal}
\end{align}
The shared memory, on the other hand, is organized into 32 banks per SM, each being 32\;bit wide and operating at 1148\;MHz where 1 bit is sent per cycle. Its peak aggregated bandwidth is then:
\begin{align}
\text{B}_\text{smem} &= \frac{\frac{1148}{2}\;\text{Mbit/s}\cdot32\;\text{bit/bank}\cdot 32\;\text{banks/SM}\cdot 14\;\text{SMs}}{8\;\text{bit/byte}}\nn
&\approx 1.03\;\text{TB/s}\ (257\;\text{Gfloats/s}).\label{bwshared}
\end{align}
The bandwidths are compared in Tab. \ref{throughputs}. Note that even when using shared memory at least 4 floating point operations must be carried out per float transferred to the CUDA cores, otherwise the algorithm will be memory bound and the peak arithmetic throughput can not be reached. %In such cases instruction level parallelism can be exploited to move data from shared memory into the even faster register memory~\cite{Vasilyy}, but this is outside the scope of this article.

%%%%%%%%%%%%%%%%%%                              ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
% DOCUMENT APPENDICES %
%%%%%%%%%%%%%%%%%%                              ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

\appendices

\section*{Acknowledgment}
The authors would like to thank Kongsberg Maritime and the Norwegian Defence Research Establishment (FFI) for providing experimental data, and thank Nvidia for providing support for running the batched linear equation solver as well as for granting us a testdrive at the Boston HPC center.

### 2014-04-13 ThesisJP/Thesis.tex
Added paper 1.

@@
 \newcommand\todo[1]{{\textit{\color{red}(#1)}}}
+
+% from Paper 1
+\usepackage{colortbl}
+\usepackage{appendix}
 
@@
 
-I would like to thank my family for all support. But most of all, I want to thank Beate for all her love, patient and support. Without it I would never have manage to complete this theses.
+I would like to thank my family and all my friends for all support.
+But most of all, I want to thank Beate for all her love, patient and support.
+Without it I would never have manage to complete this theses.
 \null
 \vfill
-\hfill Dedicated to Sofie Mathilde \AA{}sen
+\hfill To: Sofie Mathilde \AA{}sen
 \CustChapter*{List of publications}
@@
 \renewcommand\labelenumi{\bfseries\theenumi}
-	\item J.\:I.\:Buskenes, \textbf{J.\:P.\:\AA{}sen}, C.-I.\:C.\:Nilsen and A.\:Austeng, "An optimised GPU implementation of the MVDR beamformer for active sonar imaging", {\it  IEEE Transactions on Oceanic Engineering, accepted for publication}.
+	\item J.\:I.\:Buskenes, \textbf{J.\:P.\:\AA{}sen}, C.-I.\:C.\:Nilsen and A.\:Austeng, "An optimized GPU implementation of the MVDR beamformer for active sonar imaging", {\it  IEEE Transactions on Oceanic Engineering, accepted for publication}.
 	\item \textbf{J.\:P.\:\AA{}sen}, J.\:I.\:Buskenes, C.-I.\:C.\:Nilsen, A.\:Austeng and S.\:Holm, "Implementing Capon beamforming on a GPU for real-time cardiac ultrasound imaging", {\it IEEE Transactions on Ultrasonics, Ferroelectrics, and Frequency Control, vol.
@@
 \begin{bibunit}[IEEEtran]
-\CustChapterPaper{An Optimised GPU Implementation of the MVDR Beamformer for Active Sonar Imaging}
+\CustChapterPaper{An Optimized GPU Implementation of the MVDR Beamformer for Active Sonar Imaging}
 \runningtitle{GPU MVDR Beamformer for Active Sonar Imaging}
@@
 \noindent \textit{IEEE Transactions on Oceanic Engineering, accepted for publication}.
-%\newpage\input{./CaponSonarGPU/CaponSonarGPU}
-%\newpage\addcontentsline{toc}{section}{References}
-%\putbib[./CaponSonarGPU/bibCaponSonarGPU]
+\newpage
+{
+\input{./Paper1/howto_howto_capon}
+}
+\newpage\addcontentsline{toc}{section}{References}
+\putbib[./Paper1/library]
 \end{bibunit}

### 2014-04-13 ThesisJP/Paper2/cos.tex
Fixed figure layout. Checked all labels.

@@
 Benchmark results and resulting images for both implementations are presented in Section \ref{sec:bench} and \ref{sec:images}, together with a discussion of trading image quality for speed in Section \ref{sec:trade}.
-Finally we discuss the results in Section \ref{sec:dis}, and draw conclusions in Section \ref{sec:con}.
+Finally we discuss the results in Section \ref{sec:dis_paper2}, and draw conclusions in Section \ref{sec:con_paper2}.
 
@@
 \centerline{
-\subfloat[]{\includegraphics[width=0.5\textwidth]{gfx/benchmark_bar_64_80_540.png} \label{fig:benchUS64}}\\
-\hfil
-\subfloat[]{\includegraphics[width=0.5\textwidth]{gfx/benchmark_bar_96_120_540.png}%
-\label{fig:benchUS96}}}
-%\centerline{\includegraphics[width=3.5in]{gfx/benchmark_bar_64_80_540.png}}
+\begin{tabular}{c}
+\subfloat[]{\includegraphics[width=0.7\textwidth]{gfx/benchmark_bar_64_80_540.png} \label{fig:benchUS64}}\\
+\subfloat[]{\includegraphics[width=0.7\textwidth]{gfx/benchmark_bar_96_120_540.png}%
+\label{fig:benchUS96}}
+\end{tabular}
+}
 \caption{Benchmarks of GPU Capon beamforming in element space for a cardiac image covering a $70\degree$ wide $15.4$ cm deep sector.
@@
 \begin{figure*}[!t]
-\centerline{\subfloat[]{\includegraphics[width=0.5\textwidth]{gfx/benchmark_bar_bs_M=64_Nx=80_Ny=540_Nb=3.png} \label{fig:benchBS64}}
-\hfil
-\subfloat[]{\includegraphics[width=0.5\textwidth]{gfx/benchmark_bar_bs_M=96_Nx=120_Ny=540_Nb=3.png}%
-\label{fig:benchBS96}}}
+\centerline{
+\begin{tabular}{c}
+\subfloat[]{\includegraphics[width=0.7\textwidth]{gfx/benchmark_bar_bs_M=64_Nx=80_Ny=540_Nb=3.png} \label{fig:benchBS64}}\\
+\subfloat[]{\includegraphics[width=0.7\textwidth]{gfx/benchmark_bar_bs_M=96_Nx=120_Ny=540_Nb=3.png}%
+\label{fig:benchBS96}}
+\end{tabular}
+}
 \caption{Benchmarks of GPU beamspace Capon beamforming for a cardiac image covering a $70 \degree$ and $15.4$ cm deep sector.
@@
 \centerline{
-\subfloat[]{\includegraphics[width=0.3\textwidth]{gfx/das_invivo.png} \label{fig:invivoDAS}}
-\hfil
-\subfloat[]{\includegraphics[width=0.3\textwidth]{gfx/capon_invivo.png} \label{fig:invivoMV}}
-\hfil
-\subfloat[]{
-\includegraphics[width=0.3\textwidth]{gfx/capon_bs_invivo.png} \label{fig:invivoBS}}
-\hfil
-\includegraphics[height=0.3\textwidth]{gfx/colorBarInvivo2.png}
+\begin{tabular}{cc}
+\subfloat[]{\includegraphics[width=0.45\textwidth]{gfx/das_invivo.png} \label{fig:invivoDAS}} &
+\includegraphics[height=0.45\textwidth]{gfx/colorBarInvivo2.png}\\
+\subfloat[]{\includegraphics[width=0.45\textwidth]{gfx/capon_invivo.png} \label{fig:invivoMV}} &
+\subfloat[]{\includegraphics[width=0.45\textwidth]{gfx/capon_bs_invivo.png} \label{fig:invivoBS}}
+\end{tabular}
 }
@@
 
-\section{Discussion}\label{sec:dis}
+\section{Discussion}\label{sec:dis_paper2}
 In this paper we have presented a parallel implementation of the BS-Capon beamformer for high-resolution cardiac ultrasound imaging.
@@
 
-\section{Conclusion}\label{sec:con}
+\section{Conclusion}\label{sec:con_paper2}
 In this paper a parallel formulation and implementation of the beamspace Capon beamformer have been presented, and

### 2014-04-22 ThesisJP/Background/background.tex
Changes summary section in Background chap.

@@
 \section{Concluding remarks}
-All the presented problems in this chapter, except DAS beamforming, have in common that they require a processing throughput of more than ten GFLOPS.
+All the presented problems in this chapter, except DAS beamforming, have in common that they require a processing throughput in order of GFLOPS.
 Comparing this to the typical clock frequency of modern CPUs at 3 GHz we see that GFLOPS problems will require parallel execution in order to achieve real-time performance.

### 2014-04-22 HowtoHowtoCapon/gfx/prepare.sh
la2pdf commit - 2014:04:22	11:43:51	(Tue22.Apr)

@@ -1,13 +1,66 @@
 #!/bin/bash
 
-cp beamforming.eps buske1.eps
-cp implementation.eps buske2.eps
-cp mvdr_complexity.eps buske3.eps
-cp mvdr_build_R.eps buske4.eps
-cp buildR-breakdown.eps buske5.eps
-cp mvdr_implementation.eps buske6.eps
-cp plot_holmengraa_L16_Navg1.eps buske7.eps
-cp benchmark.eps buske8.eps
-cp code_assessment.eps buske9.eps
-cp code_assess_flops.eps buske10.eps
-cp benchmark_boston.eps buske11.eps
+FILES=" \
+   beamforming.svg \
+   implementation.svg \
+   mvdr_complexity.svg \
+   mvdr_build_R.svg \
+   buildR-breakdown.svg \
+   mvdr_implementation.svg \
+   plot_holmengraa_L16_Navg1.eps \
+   benchmark.eps \
+   code_assessment.eps \
+   code_assess_flops.eps \
+   benchmark_boston.eps \
+   "
+
+cp beamforming.eps buske1_online.eps
+cp implementation.eps buske2_online.eps
+cp mvdr_complexity.eps buske3_online.eps
+cp mvdr_build_R.eps buske4_online.eps
+cp buildR-breakdown.eps buske5_online.eps
+cp mvdr_implementation.eps buske6_online.eps
+cp plot_holmengraa_L16_Navg1.eps buske7_online.eps
+cp benchmark.eps buske8_online.eps
+cp code_assessment.eps buske9_online.eps
+cp code_assess_flops.eps buske10_online.eps
+cp benchmark_boston.eps buske11_online.eps
+
+# Coverts eps file to black-white using ghostscript
+function makeEpsBW {
+   gs -sDEVICE=psgray -dNOPAUSE -dBATCH -dSAFER -sOutputFile=tmp.eps $1
+   ps2eps tmp.eps
+   mv tmp.eps.eps $2
+}
+
+inkscape --select=svg2 --verb org.inkscape.color.grayscale --export-eps=buske1_bw.eps beamforming.svg
+inkscape --select=svg2 --verb org.inkscape.color.grayscale --export-eps=buske2_bw.eps implementation.svg
+inkscape --select=svg2 --verb org.inkscape.color.grayscale --export-eps=buske4_bw.eps mvdr_build_R.svg
[…]

### 2014-04-22 HowtoHowtoCapon/gfx/prepare.sh
la2pdf commit - 2014:04:22	13:23:05	(Tue22.Apr)

@@ -17,9 +17,11 @@ FILES=" \
 cp beamforming.eps buske1_online.eps
 cp implementation.eps buske2_online.eps
 cp mvdr_complexity.eps buske3_online.eps
+cp mvdr_complexity_bw.eps buske3_bw.eps
 cp mvdr_build_R.eps buske4_online.eps
 cp buildR-breakdown.eps buske5_online.eps
 cp mvdr_implementation.eps buske6_online.eps
+cp plot_holmengraa_L16_Navg1.eps buske7_bw.eps
 cp plot_holmengraa_L16_Navg1.eps buske7_online.eps
 cp benchmark.eps buske8_online.eps
 cp code_assessment.eps buske9_online.eps
@@ -33,18 +35,17 @@ function makeEpsBW {
    mv tmp.eps.eps $2
 }
 
-inkscape --select=svg2 --verb org.inkscape.color.grayscale --export-eps=buske1_bw.eps beamforming.svg
-inkscape --select=svg2 --verb org.inkscape.color.grayscale --export-eps=buske2_bw.eps implementation.svg
-inkscape --select=svg2 --verb org.inkscape.color.grayscale --export-eps=buske4_bw.eps mvdr_build_R.svg
-inkscape --select=svg2 --verb org.inkscape.color.grayscale --export-eps=buske6_bw.eps mvdr_implementation.svg
-cp plot_holmengraa_L16_Navg1.eps buske7_bw.eps
+# inkscape --select=svg2 --verb org.inkscape.color.grayscale --export-eps=buske1_bw.eps beamforming.svg
+# inkscape --select=svg2 --verb org.inkscape.color.grayscale --export-eps=buske2_bw.eps implementation.svg
+# inkscape --select=svg2 --verb org.inkscape.color.grayscale --export-eps=buske4_bw.eps mvdr_build_R.svg
+# inkscape --select=svg2 --verb org.inkscape.color.grayscale --export-eps=buske6_bw.eps mvdr_implementation.svg
 
-makeEpsBW buske3_online.eps buske3_bw.eps
-makeEpsBW buske5_online.eps buske5_bw.eps
-makeEpsBW buske8_online.eps buske8_bw.eps
-makeEpsBW buske9_online.eps buske9_bw.eps
-makeEpsBW buske10_online.eps buske10_bw.eps
-makeEpsBW buske11_online.eps buske11_bw.eps
[…]

### 2014-04-22 HowtoHowtoCapon/gfx/prepare.sh
la2pdf commit - 2014:04:22	15:38:42	(Tue22.Apr)

@@ -16,10 +16,15 @@ FILES=" \
 
 cp beamforming.eps buske1_online.eps
 cp implementation.eps buske2_online.eps
-cp mvdr_complexity.eps buske3_online.eps
+
 cp mvdr_complexity_bw.eps buske3_bw.eps
+cp mvdr_complexity.eps buske3_online.eps
+
+cp mvdr_build_R_bw.eps buske4_bw.eps
 cp mvdr_build_R.eps buske4_online.eps
+
 cp buildR-breakdown.eps buske5_online.eps
+
 cp mvdr_implementation.eps buske6_online.eps
 cp plot_holmengraa_L16_Navg1.eps buske7_bw.eps
 cp plot_holmengraa_L16_Navg1.eps buske7_online.eps

### 2014-04-22 HowtoHowtoCapon/gfx/prepare.sh
la2pdf commit - 2014:04:22	16:11:09	(Tue22.Apr)

@@ -23,12 +23,18 @@ cp mvdr_complexity.eps buske3_online.eps
 cp mvdr_build_R_bw.eps buske4_bw.eps
 cp mvdr_build_R.eps buske4_online.eps
 
+cp buildR-breakdown_bw.eps buske5_bw.eps
 cp buildR-breakdown.eps buske5_online.eps
 
+cp mvdr_implementation_bw.eps buske6_bw.eps
 cp mvdr_implementation.eps buske6_online.eps
+
 cp plot_holmengraa_L16_Navg1.eps buske7_bw.eps
 cp plot_holmengraa_L16_Navg1.eps buske7_online.eps
+
+cp benchmark_bw.eps buske8_bw.eps
 cp benchmark.eps buske8_online.eps
+
 cp code_assessment.eps buske9_online.eps
 cp code_assess_flops.eps buske10_online.eps
 cp benchmark_boston.eps buske11_online.eps

### 2014-04-22 HowtoHowtoCapon/gfx/prepare.sh
la2pdf commit - 2014:04:22	16:49:37	(Tue22.Apr)

@@ -35,8 +35,12 @@ cp plot_holmengraa_L16_Navg1.eps buske7_online.eps
 cp benchmark_bw.eps buske8_bw.eps
 cp benchmark.eps buske8_online.eps
 
+cp code_assessment_bw.eps buske9_bw.eps
 cp code_assessment.eps buske9_online.eps
+
+cp code_assess_flops_bw.eps buske10_bw.eps
 cp code_assess_flops.eps buske10_online.eps
+
 cp benchmark_boston.eps buske11_online.eps
 
 # Coverts eps file to black-white using ghostscript

### 2014-04-22 HowtoHowtoCapon/gfx/prepare.sh
la2pdf commit - 2014:04:22	21:58:48	(Tue22.Apr)

@@ -41,6 +41,7 @@ cp code_assessment.eps buske9_online.eps
 cp code_assess_flops_bw.eps buske10_bw.eps
 cp code_assess_flops.eps buske10_online.eps
 
+cp benchmark_boston_bw.eps buske11_bw.eps
 cp benchmark_boston.eps buske11_online.eps
 
 # Coverts eps file to black-white using ghostscript

### 2014-04-23 HowtoHowtoCapon/bio/andreas_austeng.tex
JOE Final submit

--- 2014-03-04 HRUIofMovingObjects/jrnl_article.tex
@@
-%% bare_jrnl.tex
-%% V1.3
-%% 2007/01/11
-%% by Michael Shell
-%% see http://www.michaelshell.org/
-%% for current contact information.
-%%
-%% This is a skeleton file demonstrating the use of IEEEtran.cls
-%% (requires IEEEtran.cls version 1.7 or later) with an IEEE journal paper.
-%%
-%% Support sites:
-%% http://www.michaelshell.org/tex/ieeetran/
-%% http://www.ctan.org/tex-archive/macros/latex/contrib/IEEEtran/
-%% and
-%% http://www.ieee.org/
-
-% *** Authors should verify (and, if needed, correct) their LaTeX system  ***
-% *** with the testflow diagnostic prior to trusting their LaTeX platform ***
-% *** with production work. IEEE's font choices can trigger bugs that do  ***
-% *** not appear when using other class files.                            ***
-% The testflow support page is at:
-% http://www.michaelshell.org/tex/testflow/
-
-%%*************************************************************************
-%% Legal Notice:
-%% This code is offered as-is without any warranty either expressed or
-%% implied; without even the implied warranty of MERCHANTABILITY or
-%% FITNESS FOR A PARTICULAR PURPOSE! 
-%% User assumes all risk.
-%% In no event shall IEEE or any contributor to this code be liable for
-%% any damages or losses, including, but not limited to, incidental,
-%% consequential, or any other damages, resulting from the use or misuse
-%% of any information contained here.
-%%
-%% All comments are the opinions of their respective authors and are not
-%% necessarily endorsed by the IEEE.
-%%
-%% This work is distributed under the LaTeX Project Public License (LPPL)
-%% ( http://www.latex-project.org/ ) version 1.3, and may be freely used,
-%% distributed and modified. A copy of the LPPL, version 1.3, is included
-%% in the base LaTeX documentation of all distributions of LaTeX released
-%% 2003/12/01 or later.
-%% Retain all contribution notices and credits.
-%% ** Modified files should be clearly indicated as such, including  **
-%% ** renaming them and changing author support contact information. **
-%%
-%% File list of work: IEEEtran.cls, IEEEtran_HOWTO.pdf, bare_adv.tex,
-%%                    bare_conf.tex, bare_jrnl.tex, bare_jrnl_compsoc.tex
-%%*************************************************************************
-
-% Note that the a4paper option is mainly intended so that authors in
-% countries using A4 can easily print to A4 and see how their papers will
-% look in print - the typesetting of the document will not typically be
-% affected with changes in paper size (but the bottom and side margins will).
-% Use the testflow package mentioned above to verify correct handling of
-% both paper sizes by the user's LaTeX system.
-%
-% Also note that the "draftcls" or "draftclsnofoot", not "draft", option
-% should be used if it is desired that the figures are to be displayed in
-% draft mode.
-%
-\documentclass[draftcls]{IEEEtran}
-%\documentclass[journal]{IEEEtran}
-%
-% If IEEEtran.cls has not been installed into the LaTeX system files,
-% manually specify the path to it like:
-% \documentclass[journal]{../sty/IEEEtran}
-
-% Some very useful LaTeX packages include:
-% (uncomment the ones you want to load)
-
-% *** MISC UTILITY PACKAGES ***
-%
-%\usepackage{ifpdf}
-% Heiko Oberdiek's ifpdf.sty is very useful if you need conditional
-% compilation based on whether the output is pdf or dvi.
-% usage:
-% \ifpdf
-%   % pdf code
-% \else
-%   % dvi code
-% \fi
-% The latest version of ifpdf.sty can be obtained from:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/oberdiek/
-% Also, note that IEEEtran.cls V1.7 and later provides a builtin
-% \ifCLASSINFOpdf conditional that works the same way.
-% When switching from latex to pdflatex and vice-versa, the compiler may
-% have to be run twice to clear warning/error messages.
-
-% *** CITATION PACKAGES ***
-%
-\usepackage{cite}
-% cite.sty was written by Donald Arseneau
-% V1.6 and later of IEEEtran pre-defines the format of the cite.sty package
-% \cite{} output to follow that of IEEE. Loading the cite package will
-% result in citation numbers being automatically sorted and properly
-% "compressed/ranged". e.g., [1], [9], [2], [7], [5], [6] without using
-% cite.sty will become [1], [2], [5]--[7], [9] using cite.sty. cite.sty's
-% \cite will automatically add leading space, if needed. Use cite.sty's
-% noadjust option (cite.sty V3.8 and later) if you want to turn this off.
-% cite.sty is already installed on most LaTeX systems. Be sure and use
-% version 4.0 (2003-05-27) and later if using hyperref.sty. cite.sty does
-% not currently provide for hyperlinked citations.
-% The latest version can be obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/cite/
-% The documentation is contained in the cite.sty file itself.
-
-% *** GRAPHICS RELATED PACKAGES ***
-%
-\ifCLASSINFOpdf
-  % \usepackage[pdftex]{graphicx}
-  % declare the path(s) where your graphic files are
-  % \graphicspath{{../pdf/}{../jpeg/}}
-  % and their extensions so you won't have to specify these with
-  % every instance of \includegraphics
-  % \DeclareGraphicsExtensions{.pdf,.jpeg,.png}
-\else
-  % or other class option (dvipsone, dvipdf, if not using dvips). graphicx
-  % will default to the driver specified in the system graphics.cfg if no
-  % driver is specified.
-  % \usepackage[dvips]{graphicx}
-  % declare the path(s) where your graphic files are
-  % \graphicspath{{../eps/}}
-  % and their extensions so you won't have to specify these with
-  % every instance of \includegraphics
-  % \DeclareGraphicsExtensions{.eps}
-\fi
-% graphicx was written by David Carlisle and Sebastian Rahtz. It is
-% required if you want graphics, photos, etc. graphicx.sty is already
-% installed on most LaTeX systems. The latest version and documentation can
-% be obtained at: 
-% http://www.ctan.org/tex-archive/macros/latex/required/graphics/
-% Another good source of documentation is "Using Imported Graphics in
-% LaTeX2e" by Keith Reckdahl which can be found as epslatex.ps or
-% epslatex.pdf at: http://www.ctan.org/tex-archive/info/
-%
-% latex, and pdflatex in dvi mode, support graphics in encapsulated
-% postscript (.eps) format. pdflatex in pdf mode supports graphics
-% in .pdf, .jpeg, .png and .mps (metapost) formats. Users should ensure
-% that all non-photo figures use a vector format (.eps, .pdf, .mps) and
-% not a bitmapped formats (.jpeg, .png). IEEE frowns on bitmapped formats
-% which can result in "jaggedy"/blurry rendering of lines and letters as
-% well as large increases in file sizes.
-%
-% You can find documentation about the pdfTeX application at:
-% http://www.tug.org/applications/pdftex
-
-% *** MATH PACKAGES ***
-%
-%\usepackage[cmex10]{amsmath}
-% A popular package from the American Mathematical Society that provides
-% many useful and powerful commands for dealing with mathematics. If using
-% it, be sure to load this package with the cmex10 option to ensure that
-% only type 1 fonts will utilized at all point sizes. Without this option,
-% it is possible that some math symbols, particularly those within
-% footnotes, will be rendered in bitmap form which will result in a
-% document that can not be IEEE Xplore compliant!
-%
-% Also, note that the amsmath package sets \interdisplaylinepenalty to 10000
-% thus preventing page breaks from occurring within multiline equations. Use:
-%\interdisplaylinepenalty=2500
-% after loading amsmath to restore such page breaks as IEEEtran.cls normally
-% does. amsmath.sty is already installed on most LaTeX systems. The latest
-% version and documentation can be obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/required/amslatex/math/
-
-% *** SPECIALIZED LIST PACKAGES ***
-%
-%\usepackage{algorithmic}
-% algorithmic.sty was written by Peter Williams and Rogerio Brito.
-% This package provides an algorithmic environment fo describing algorithms.
-% You can use the algorithmic environment in-text or within a figure
-% environment to provide for a floating algorithm. Do NOT use the algorithm
-% floating environment provided by algorithm.sty (by the same authors) or
-% algorithm2e.sty (by Christophe Fiorio) as IEEE does not use dedicated
-% algorithm float types and packages that provide these will not provide
-% correct IEEE style captions. The latest version and documentation of
-% algorithmic.sty can be obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/algorithms/
-% There is also a support site at:
-% http://algorithms.berlios.de/index.html
-% Also of interest may be the (relatively newer and more customizable)
-% algorithmicx.sty package by Szasz Janos:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/algorithmicx/
-
-% *** ALIGNMENT PACKAGES ***
-%
-%\usepackage{array}
-% Frank Mittelbach's and David Carlisle's array.sty patches and improves
-% the standard LaTeX2e array and tabular environments to provide better
-% appearance and additional user controls. As the default LaTeX2e table
-% generation code is lacking to the point of almost being broken with
-% respect to the quality of the end results, all users are strongly
-% advised to use an enhanced (at the very least that provided by array.sty)
-% set of table tools. array.sty is already installed on most systems. The
-% latest version and documentation can be obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/required/tools/
-
-%\usepackage{mdwmath}
-%\usepackage{mdwtab}
-% Also highly recommended is Mark Wooding's extremely powerful MDW tools,
-% especially mdwmath.sty and mdwtab.sty which are used to format equations
-% and tables, respectively. The MDWtools set is already installed on most
-% LaTeX systems. The lastest version and documentation is available at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/mdwtools/
-
-% IEEEtran contains the IEEEeqnarray family of commands that can be used to
-% generate multiline equations as well as matrices, tables, etc., of high
-% quality.
-
-%\usepackage{eqparbox}
-% Also of notable interest is Scott Pakin's eqparbox package for creating
-% (automatically sized) equal width boxes - aka "natural width parboxes".
-% Available at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/eqparbox/
-
-% *** SUBFIGURE PACKAGES ***
-%\usepackage[tight,footnotesize]{subfigure}
-% subfigure.sty was written by Steven Douglas Cochran. This package makes it
-% easy to put subfigures in your figures. e.g., "Figure 1a and 1b". For IEEE
-% work, it is a good idea to load it with the tight package option to reduce
-% the amount of white space around the subfigures. subfigure.sty is already
-% installed on most LaTeX systems. The latest version and documentation can
-% be obtained at:
-% http://www.ctan.org/tex-archive/obsolete/macros/latex/contrib/subfigure/
-% subfigure.sty has been superceeded by subfig.sty.
-
-%\usepackage[caption=false]{caption}
-%\usepackage[font=footnotesize]{subfig}
-% subfig.sty, also written by Steven Douglas Cochran, is the modern
-% replacement for subfigure.sty. However, subfig.sty requires and
-% automatically loads Axel Sommerfeldt's caption.sty which will override
-% IEEEtran.cls handling of captions and this will result in nonIEEE style
-% figure/table captions. To prevent this problem, be sure and preload
-% caption.sty with its "caption=false" package option. This is will preserve
-% IEEEtran.cls handing of captions. Version 1.3 (2005/06/28) and later 
-% (recommended due to many improvements over 1.2) of subfig.sty supports
-% the caption=false option directly:
-\usepackage[caption=false,font=footnotesize]{subfig}
-%
-% The latest version and documentation can be obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/subfig/
-% The latest version and documentation of caption.sty can be obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/caption/
-
-% *** FLOAT PACKAGES ***
-%
-%\usepackage{fixltx2e}
-% fixltx2e, the successor to the earlier fix2col.sty, was written by
-% Frank Mittelbach and David Carlisle. This package corrects a few problems
-% in the LaTeX2e kernel, the most notable of which is that in current
-% LaTeX2e releases, the ordering of single and double column floats is not
-% guaranteed to be preserved. Thus, an unpatched LaTeX2e can allow a
-% single column figure to be placed prior to an earlier double column
-% figure. The latest version and documentation can be found at:
-% http://www.ctan.org/tex-archive/macros/latex/base/
-
-%\usepackage{stfloats}
-% stfloats.sty was written by Sigitas Tolusis. This package gives LaTeX2e
-% the ability to do double column floats at the bottom of the page as well
-% as the top. (e.g., "\begin{figure*}[!b]" is not normally possible in
-% LaTeX2e). It also provides a command:
-%\fnbelowfloat
-% to enable the placement of footnotes below bottom floats (the standard
-% LaTeX2e kernel puts them above bottom floats). This is an invasive package
-% which rewrites many portions of the LaTeX2e float routines. It may not work
-% with other packages that modify the LaTeX2e float routines. The latest
-% version and documentation can be obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/sttools/
-% Documentation is contained in the stfloats.sty comments as well as in the
-% presfull.pdf file. Do not use the stfloats baselinefloat ability as IEEE
-% does not allow \baselineskip to stretch. Authors submitting work to the
-% IEEE should note that IEEE rarely uses double column equations and
-% that authors should try to avoid such use. Do not be tempted to use the
-% cuted.sty or midfloat.sty packages (also by Sigitas Tolusis) as IEEE does
-% not format its papers in such ways.
-
-%\ifCLASSOPTIONcaptionsoff
-%  \usepackage[nomarkers]{endfloat}
-% \let\MYoriglatexcaption\caption
-% \renewcommand{\caption}[2][\relax]{\MYoriglatexcaption[#2]{#2}}
-%\fi
-% endfloat.sty was written by James Darrell McCauley and Jeff Goldberg.
-% This package may be useful when used in conjunction with IEEEtran.cls'
-% captionsoff option. Some IEEE journals/societies require that submissions
-% have lists of figures/tables at the end of the paper and that
-% figures/tables without any captions are placed on a page by themselves at
-% the end of the document. If needed, the draftcls IEEEtran class option or
-% \CLASSINPUTbaselinestretch interface can be used to increase the line
-% spacing as well. Be sure and use the nomarkers option of endfloat to
-% prevent endfloat from "marking" where the figures would have been placed
-% in the text. The two hack lines of code above are a slight modification of
-% that suggested by in the endfloat docs (section 8.3.1) to ensure that
-% the full captions always appear in the list of figures/tables - even if
-% the user used the short optional argument of \caption[]{}.
-% IEEE papers do not typically make use of \caption[]'s optional argument,
-% so this should not be an issue. A similar trick can be used to disable
-% captions of packages such as subfig.sty that lack options to turn off
-% the subcaptions:
-% For subfig.sty:
-% \let\MYorigsubfloat\subfloat
-% \renewcommand{\subfloat}[2][\relax]{\MYorigsubfloat[]{#2}}
-% For subfigure.sty:
-% \let\MYorigsubfigure\subfigure
-% \renewcommand{\subfigure}[2][\relax]{\MYorigsubfigure[]{#2}}
-% However, the above trick will not work if both optional arguments of
-% the \subfloat/subfig command are used. Furthermore, there needs to be a
-% description of each subfigure *somewhere* and endfloat does not add
-% subfigure captions to its list of figures. Thus, the best approach is to
-% avoid the use of subfigure captions (many IEEE journals avoid them anyway)
-% and instead reference/explain all the subfigures within the main caption.
-% The latest version of endfloat.sty and its documentation can obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/endfloat/
-%
-% The IEEEtran \ifCLASSOPTIONcaptionsoff conditional can also be used
-% later in the document, say, to conditionally put the References on a 
-% page by themselves.
-
-% *** PDF, URL AND HYPERLINK PACKAGES ***
-%
-%\usepackage{url}
-% url.sty was written by Donald Arseneau. It provides better support for
-% handling and breaking URLs. url.sty is already installed on most LaTeX
-% systems. The latest version can be obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/misc/
-% Read the url.sty source comments for usage information. Basically,
-% \url{my_url_here}.
-
-% *** Do not adjust lengths that control margins, column widths, etc. ***
-% *** Do not use packages that alter fonts (such as pslatex).         ***
-% There should be no need to do such things with IEEEtran.cls V1.6 and later.
-% (Unless specifically asked to do so by the journal or conference you plan
-% to submit to, of course. )
-
-% correct bad hyphenation here
-\hyphenation{op-tical net-works semi-conduc-tor}
-
-%%% My packages
-\usepackage{amsmath,epsfig}
-\usepackage{amsfonts}
-\usepackage{color}
-\usepackage{epstopdf}
-
-%%% My commands
-\newcommand{\mat}[1]{\mathbf{#1}}
-\renewcommand{\vec}[1]{\mathbf{#1}}
-\newcommand{\R}{$\mat{\hat{R}}$ }
-\newcommand{\degree}{\ensuremath{^\circ}}
-
-\newcommand{\img}{img/}
-\epstopdfsetup{outdir=\img}
-
-\newcommand\multimedia[1]{\textbf{{\color{red}[#1]}}}
-\newcommand\comment[1]{\textit{{\color{red}(#1)}}}
-
-\begin{document}
-%
-% paper title
-% can use linebreaks \\ within to get better formatting as desired
-\title{Capon Beamforming and Moving Objects - An Analysis of Lateral Shift-Invariance}
-%
-%
-% author names and IEEE memberships
-% note positions of commas and nonbreaking spaces ( ~ ) LaTeX will not break
-% a structure at a ~ so this keeps an author's name from being broken across
-% two lines.
-% use \thanks{} to gain access to the first footnote area
-% a separate \thanks must be used for each paragraph as LaTeX2e's \thanks
-% was not built to handle multiple paragraphs
-%
-
-\author{
-   Jon Petter \AA{}sen, \IEEEmembership{Student Member,~IEEE,} Andreas Austeng, \IEEEmembership{Member,~IEEE,} and Sverre Holm, \IEEEmembership{Senior Member,~IEEE}% <-this % stops a space
-
-\thanks{Manuscript received ...; accepted ...}
-\thanks{J. P. \AA{}sen and S. Holm is with the Medical Imaging Lab (MI-Lab) at the Norwegian University of Science and Technology, Trondheim, Norway.}
-\thanks{A. Austeng and S. Holm is with the Department of Informatics, University of Oslo, Oslo, Norway. (e-mail: jon.p.asen@ntnu.no)} 
-%J. I. Buskenes, C.-I. C. Nilsen, 
-}
-%M. Shell is with the Department
-%of Electrical and Computer Engineering, Georgia Institute of Technology, Atlanta,
-%GA, 30332 USA e-mail: (see http://www.michaelshell.org/contact.html).}% <-this % stops a space
-%\thanks{J. Doe and J. Doe are with Anonymous University.}% <-this % stops a space
-%\thanks{Manuscript received April 19, 2005; revised January 11, 2007.}}
-
-% note the % following the last \IEEEmembership and also \thanks - 
-% these prevent an unwanted space from occurring between the last author name
-% and the end of the author line. i.e., if you had this:
-% 
-% \author{....lastname \thanks{...} \thanks{...} }
-%                     ^------------^------------^----Do not want these spaces!
-%
-% a space would be appended to the last name and could cause every name on that
-% line to be shifted left slightly. This is one of those "LaTeX things". For
-% instance, "\textbf{A} \textbf{B}" will typeset as "A B" not "AB". To get
-% "AB" then you have to do: "\textbf{A}\textbf{B}"
-% \thanks is no different in this regard, so shield the last } of each \thanks
-% that ends a line with a % and do not let a space in before the next \thanks.
-% Spaces after \IEEEmembership other than the last one are OK (and needed) as
-% you are supposed to have spaces between the names. For what it is worth,
-% this is a minor point as most people would not even notice if the said evil
-% space somehow managed to creep in.
-
-% The paper headers
-\markboth{Transactions on Ultrasonics and Ferroelectrics and Frequency Control,~Vol.~x, No.~y, January~201\AA{}}%
-{\AA{}sen \MakeLowercase{\textit{et al.}}: The Capon Adaptive Beamformer Applied on Moving Objects }
-% The only time the second header will appear is for the odd numbered pages
-% after the title page when using the twoside option.
-% 
-% *** Note that you probably will NOT want to include the author's ***
-% *** name in the headers of peer review papers.                   ***
-% You can use \ifCLASSOPTIONpeerreview for conditional compilation here if
-% you desire.
-
-% If you want to put a publisher's ID mark on the page you can do it like
-% this:
-%\IEEEpubid{0000--0000/00\$00.00~\copyright~2007 IEEE}
-% Remember, if you use this you must call \IEEEpubidadjcol in the second
-% column for its text to clear the IEEEpubid mark.
-
-% use for special paper notices
-%\IEEEspecialpapernotice{(Invited Paper)}
-
-% make the title area
-\maketitle
-
-\begin{abstract}
-%\boldmath
-If an ultrasound imaging system provides a presentation of a moving object which is sensitive to small spatial shifts, the system is said to be locally spatially shift-variant.
-This can for instance happen if the axial or lateral sampling is insufficient.
-The Capon beamformer has been shown to provide increased lateral resolution in ultrasound images.
-Increased lateral resolution should demand denser lateral sampling.
-However, in previous literature on Capon beamforming for medical ultrasound imaging only single frame scenarios have been simulated.
-Temporal behaviour and effects caused by the increased resolution and lack of oversampling have therefore been neglected.
-
-In this paper we analyse the local lateral shift-invariance of the Capon beamformer when imaging moving objects.
-We show that insufficient lateral sampling makes an imaging system based on the Capon beamformer laterally shift-variant.
-Different methods for oversampling on transmit and receive are then discussed and investigated in order to improve on it.
-It is shown that lateral shift-invariance can be improved by oversampling based on phase rotation on receive.
-This without affecting the acquisition frame rate and with a minor change in processing complexity.
-\end{abstract}
-% IEEEtran.cls defaults to using nonbold math in the Abstract.
-% This preserves the distinction between vectors and scalars. However,
-% if the journal you are submitting to favors bold math in the abstract,
-% then you can use LaTeX's standard command \boldmath at the very start
-% of the abstract to achieve this. Many IEEE journals frown on math
-% in the abstract anyway.
-
-% Note that keywords are not normally used for peerreview papers.
-\begin{IEEEkeywords}
-Adaptive beamforming, Capon beamformer, Shift Invariance, Ultrasound Imaging.
-\end{IEEEkeywords}
-
-% For peer review papers, you can put extra information on the cover
-% page as needed:
-% \ifCLASSOPTIONpeerreview
-% \begin{center} \bfseries EDICS Category: 3-BBND \end{center}
-% \fi
-%
-% For peerreview papers, this IEEEtran command inserts a page break and
-% creates the second title. It will be ignored for other modes.
-\IEEEpeerreviewmaketitle
-
-\section{Introduction}
-% The very first letter is a 2 line initial drop letter followed
-% by the rest of the first word in caps.
-% 
-% form to use if the first word consists of a single letter:
-% \IEEEPARstart{A}{demo} file is ....
-% 
-% form to use if you need the single drop letter followed by
-% normal text (unknown if ever used by IEEE):
-% \IEEEPARstart{A}{}demo file is ....
-% 
-% Some journals put the first two words in caps:
-% \IEEEPARstart{T}{his demo} file is ....
-% 
-% Here we have the typical use of a "T" for an initial drop letter
-% and "HIS" in caps to complete the first word.
-
-\IEEEPARstart{I}{n} the last decade several authors have applied the Capon beamformer [also known as the minimum variance beamformer] for medical ultrasound imaging \cite{Synnevag2007, Vignon2008, Viola}.
-The method has been shown to improve lateral resolution and edge sharpness in ultrasound images \cite{Synnevag2007, Synnevag2009, Chen2011}.
-This is accomplished by adaptively adjusting the array weights based on element-to-element covariance in order to minimize interference.
-The method is, however, also highly sensitive to model errors \cite{Mestre2006, Widrow1982, Wax1996, Wax1996a}.
-This has led to the development of several robust versions of the Capon beamformer with e.g.
-spatial smoothing \cite{Shan1985} or diagonal loading \cite{JianLi2003} of the covariance estimate, or by applying steering vector uncertainty sets \cite{Lorenz2005, Rubsamen2013}.
-Spatial smoothing and diagonal loading, together with temporal smoothing of the covariance estimate, have been applied in earlier work on Capon beamforming for medical ultrasound imaging \cite{Synnevag2009}.
-The first two trade resolution for robustness and temporal smoothing helps to preserve speckle statistics for high resolution parameters \cite{Synnevag2007a}.
-Steering vector uncertainty sets, however, have not been applied for medical ultrasound imaging.
-
-% but at an increased computational expense \cite{So2011}.
-
-%In recent work we introduced an implementation of the Capon beamformer capable of processing ultrasound images at interactive frame rates \cite{Asen}.
-In the same paper we discovered that oversampling had to be applied to make the imaging system locally spatially shift-invariant.
-%It is therefore now interesting how a moving object appears after Capon beamforming has been applied.
-%Earlier work on Capon beamforming for medical ultrasound imaging has investigated the method's capabilities in single frames only, and temporal behaviour and effects have therefore been neglected.
-%
-
-A steering vector uncertainty set reduces signal-nulling which appears if the signal of interest is present during the covariance estimation process and if this estimate is later used with a steering vector that does not match the signal propagation vector exactly.
-This self-nulling is actually how the Capon beamformer obtains its super resolution capabilities, but the signal can also get lost in the process.
-In \cite{JianLi2003} it was shown that most methods based on steering vector uncertainty sets are equivalent to diagonal loading where the loading factor depends on the steering vector uncertainty.
-In that way they make up for a given uncertainty by making the Capon beam wider.
-Hence, the maximal achievable resolution is decreased if the initial steering vector spacing was too coarse.
-In medical ultrasound imaging this initial set of steering vectors is governed by the number of receive beams.
-In this paper we will make use of an additional set of steering vectors to reduce the self-nulling effect by means of oversampling.
-The maximal achievable resolution is then maintained without the presence of severe self-nulling.
-
-In \cite{Cox1973}, Cox presents in great detail how the amount of mismatch between the selected steering vector and the signal propagation vector impacts the beamformer output in simple situations.
-He also states the following important fact about the Capon beamformer [the $\vec{k}_3$ processor]:
-\begin{quote}
-"\textit{...
-a $\vec{k}_3$ processor requires more closely spaced beams than a ...
-$\vec{k}_1$-processor in order to avoid serious signal suppression effects being introduced on signals arriving from directions between the beams}",
-\end{quote}
-where $\vec{k}_1$ is the delay-and-sum (DAS) beamformer.
-In Fig.\,\ref{fig:das_capon_beams} we have plotted the two-way response for DAS and the Capon beamformer using (26) and (32) in \cite{Cox1973}.
-A 64-element linear array with $\lambda/2$ element spacing, where $\lambda$ is the wavelength, is scanned over a sector equal to four times the array resolution using narrow band excitation.
-Located in the farfield of the sector is an object moving parallel to the array surface.
-The plot shows the high precision scanned output when the object is located at plus and minus half the required beam spacing for DAS [See (\ref{eq:resolution})].
-%No spatially white noise is present.
-Hence, the interference-noise matrix ($\sigma_0^2\mat{Q}$ in \cite{Cox1973}) is the identity.
-The effect of typical subarray averaging and diagonal loading have been added to the Capon beamformer output (subarray length of 32 and a diagonal loading factor of 0.01).
-We see that a DAS beamformer, using the required beam spacing, could yield a $1.9$ dB scalloping loss \cite{Harris1978} if the object is moving parallel to the array.
-However, if the Capon beamformer is used with the same beam density the loss could be as large as $27$ dB.
-This imaging system is then highly sensitive to small spatial shifts, either by the object or the array itself.
-A $1.9$ dB scalloping loss for the Capon beamforming is obtained for this specific example when the beam spacing or object distance is decreased with a factor of 25.
-
-\begin{figure}[!t]
-\centerline{
-\includegraphics[width=\linewidth]{\img fig1_2way.eps}
-}
-\caption{Estimated drop in output power mid-way between receive beams for delay-and-sum and Capon beamforming (with subarray length of 32 and a diagonal loading factor of 0.01) applied in a farfield-narrowband setting.
-%The scanned output from both beamformers is plotted for two different time instances.
-For the first instances an object is located at an angle equal to half the required beam spacing for delay-and-sum, and for the latter the object is moved an angular distance corresponding to the system resolution.
-If the beam density was reduced to one beam per system resolution the maximum scalloping loss is found at zero degrees.
-}
-\label{fig:das_capon_beams}
-\end{figure}
-
-In previous literature on Capon beamforming for medical ultrasound imaging only single frame scenarios have been simulated.
-The scalloping loss effect has therefore been passed over in silence by either oversampling on transmit, by carefully positioning point scatterers exactly on transmit-receive beams, or by ignorance.
-In our first work where Capon beamforming was applied on a loop of images containing moving objects \cite{Asen2012, Asen} we commented on the effect, but no solution was presented other than oversampling on transmit when simulating.
-The same was done by Jensen and Austeng \cite{Jensen2012}.
-
-In this paper we investigate the issue further.
-We show that when a moving point scatterer is imaged with the Capon beamformer with the same beam spacing which is sufficient for DAS, the predicted scalloping loss is observed both in simulations and \textit{in-vitro} data.
-The question is then, what can we do in order to improve on the situation without affecting the acquisition frame rate and processing throughout.
-%We try two answer this by studying the scalloping loss of the Capon beamforming and investigating different methods aimed at reducing it.
-We also discuss how the different methods will affect the acquisition frame rate and the processing complexity.
-
-The next section gives an introduction to Capon beamforming together with a short introduction to lateral sampling and shift-invariance.
-In Section \ref{sec:capon_LLSI} we study the scalloping loss yield by the Capon beamformer and its local lateral shift-invariance property.
-Different methods for reduced scalloping loss and by that improved local lateral shift-invariance are presented in Section \ref{sec:methods}.
-Then, in Section \ref{sec:res} we present the results of applying the most promising method on \textit{in-vitro} data.
-Finally we discuss our findings in Section \ref{sec:dis} and give our conclusions in Section \ref{sec:con}.
-
-%\comment{Next sections summary.}
-
-% needed in second column of first page if using \IEEEpubid
-%\IEEEpubidadjcol
-
-\section{Background}
-\subsection{Standard Beamforming}
-The standard form of time domain array beamforming is the DAS beamformer where the output $z$ is calculated as:
-\begin{align}\label{eq:das}
-z[n] = \sum_{m = 0}^{M-1}w_m^*x_m[n - \Delta_m[n]] = \vec{w}^H\vec{x}[n],
-\end{align}
-where $M$ is the number of elements or channels in the array, $\Delta_m[n]$ is the per-element focusing and steering delay, and $\vec{w}$ is a weight vector or window function.
-The weight vector is typically selected in order to trade resolution for a lower side lobe level.
-Hence, the weights are typically real and their K-space response is then symmetric.
-
-\subsection{Capon Beamforming}
-The Capon beamformer produces a weight vector, like the one in (\ref{eq:das}), based on the impinging signal and an optimization criterion. The problem is as follow \cite{Capon1969}:
-\begin{align}
-&\min_{\vec{w}} E\{|z[n]|^2\} \rightarrow \min_{\vec{w}} \vec{w}^H \mat{\hat{R}} \vec{w} \label{eq:capon_optimization_criteria} \\
-&\text{subjected to } \vec{w}^H\vec{a} = 1,
-\end{align}
-where $\mat{\hat{R}}$ is a sample covariance matrix.
-Hence, the resulting weight vector minimizes the output power while maintaining unit gain in the steering direction $\vec{a}$.
-%The covariance matrix is usually unknown and has to be estimated from the element data.
-The covariance matrix in (\ref{eq:capon_optimization_criteria}) is therefore substituted with the sample covariance matrix.
-
-The sample covariance matrix can be estimated for an active broadband system in the following way \cite{Synnevag2009}:
-\begin{align}
-\mat{\breve{R}}[n] = \frac{1}{N_LN_K}\sum_{n'=n-K}^{n+K} \sum_{l=0}^{N_L-1} \vec{x}_l[n']\vec{x}_l[n']^H,
-\end{align}
-where $K = M-L+1$ typically is proportional to the pulse length, $N_K = 2K + 1$, $N_L = M-L+1$, and $\vec{x}_l$ is a $L$ long subarray $[x_l[n], \dotso x_{l+L}[n]]$.
-Finally the matrix is loaded with a diagonal factor for numerical stability and increased robustness,
-\begin{align}
-\epsilon &= d*\text{trace}\{\mat{\breve{R}}\}/L\\
-\mat{\hat{R}} &= \mat{\breve{R}} + \epsilon\mat{I}.
-\end{align}
-
-The solution to the minimization problem in (\ref{eq:capon_optimization_criteria}) is
-\begin{align}\label{eq:capon_weights}
-\vec{w}[n] = \frac{\mat{\hat{R}}[n]^{-1}\vec{a}}{\vec{a}^H\mat{\hat{R}}[n]^{-1}\vec{a}} \in \mathbb{C}^L.
-\end{align}
-Note that one matrix has to be constructed and inverted for each data vector $\vec{x}[n]$ received by the system.
-%This is indeed computational demanding, but following the innovation in GPU computing, and by applying a novel method in order to reduce the size of the matrix inversion problem \cite{Nilsen2009}, real time processing is now possible \cite{Asen}.
-From (\ref{eq:capon_weights}) it is also clear that complex data [in-phase and quadrature data] can lead to complex weights.
-The weights can therefore have an asymmetric K-space response with a shifted main lobe and zeros at arbitrary positions.
-%The beam can therefore be micro-steered by phase rotation.
-
-%Nilsen et.
-al has proposed to apply beamspace processing to further reduce the computations required.
-With beamspace we mean transforming the channel data from element space to a fan of beams covering the imaging sector, $\vec{x}_{BS} = \mat{B}x$.
-The steering vectors in $\mat{B}$, the so-called Butler matrix, determines each beams direction in space.
-We can therefore easily remove beams where no interference is present by removing rows in $\mat{B}$.
-For a focused system like medical ultrasound, where the received signal is concentrated in a narrow band around broadside, there is a small percentage of the beams that contains almost all of the energy.
-Nilsen et.
-al has shown that as little as three beams still produced results comparable with applying the full Capon estimation in element space.
-
-\subsection{Lateral Sampling and Shift-Invariance}
-
-The required lateral sampling in an ultrasound image for a given center frequency, $f_c$, and bandwidth, $B$, is given by the aperture size, $D$, and the wavelength of the center frequency, $\lambda_c = c/f_c$, where $c$ is the speed of sound.
-The required Nyquist beam spacing is then:
-\begin{align}
-\delta\theta < \frac{\lambda_c}{2qD}, \label{eq:resolution}
-\end{align}
-where the factor two is the effect of a two-way acquisition \cite{Hergum2009}, and where $q$ is an oversampling factor which can be applied to reduce folding of frequencies between $f_c$ and $f_c + B/2$.
-In this paper we will use the $q$-factor to perform oversampling prior to Capon beamforming.
-
-Equation (\ref{eq:resolution}) was used to calculate the object positions in Fig.\,\ref{fig:das_capon_beams}.
-Therefore, if (\ref{eq:resolution}) is not fulfilled, the resulting scalloping loss for the DAS beamformer could become visible in the image.
-The system output will then be sensitive to small spatial shifts.
-We say that the system is locally spatially shift-variant.
-Since Capon beamforming improves the lateral resolution we will focus on sensitivity to small lateral shifts, or in other words the local lateral shift-invariance property of the imaging system.%This invariance to local movements is what makes ultrasound imaging feasible in general, but it is specially important for quantitative techniques, like speckle tracking of tissue and blood, that correlate time-varying features.
-%In this paper we analyze the lateral shift-invariance property of the Capon beamformer by studying the beamformer output when applied on simulated and \textit{in vitro} images of moving objects.
-
-In an ultrasound image with 50 dB dynamic range mapped to 256 gray levels, a 1 dB loss corresponds to 5 gray levels.
-This is approximately equal to the visibility threshold [Weber fraction of 2\%] for gray scale images.
-A loss larger than 1 dB could therefore end up being visible to the observer.
-%However, what eventually matters is whether the loss is visible in the image or not.%In our further analysis a system would be regarded as locally laterally shift-invariant if the scalloping loss is less than 1 dB.
-%However, depending on viewing light conditions and e.g.
-the selected gray level mapping, a larger loss might be acceptable.
-
-In order to visualize the local lateral shift-invariance of a given imaging system, we will make use of lateral shift-variance plots (LSV-plots) \cite{Hergum2007}.
-A LSV-plot is constructed for a given imaging system by imaging a point scatterer moving laterally at constant speed.
-For each lateral position the root-mean-square (RMS) of the image data is calculated in range, and the resulting beam profiles are stacked in to an image.
-%For a given scan sequence there are twice as many point positions than there are receive lines.
-The LSV-plots in this paper are displayed as a contour plot with contours on -1, -2, -3, -6, -12 and -24 dB.
-The maximum for each beam profile is also marked with a black dot, and the amplitude variation among these points is presented in a subplot next to the LSV-plot.
-The LSV-plots based on simulations have been up-interpolated to 128 samples in each direction using cubic interpolation if the number of receive beams is less than 128.
-
-To further quantify the shift-invariance we have calculated the mean absolute error (MAE) between the maximum point position in each beam profile and a straight line ($f(x)=x$).
-We will refer to this measure as P-MAE.
-The MAE of the maximum points' amplitude versus the data maximum is also measured.
-We refer to this measure as A-MAE.
-The first measure quantifies the amount of geometric distortion \cite{Hergum2007}, and the latter quantifies the peak gain variation or in other words the scalloping loss.
-The maximum amplitude deviation, or maximum scalloping loss, along this point trace is also calculated.
-We present all these numbers together with the LSV-plot.
-
-A given imaging system will then be locally laterally shift-invariant if the contours are diagonal and if the amplitude variation among the maximum points is less than 1 dB.
-Then there is minimal geometric distortion (low P-MAE) and the scalloping loss will not be visible for the observer (low A-MAE).
-%Since both effects can occur at the same time it is important to have a measure not effected by geometric distortion.
-
-Figure \ref{fig:das_shiftinvariant} presents such a LSV-plot for an imaging system using the DAS beamformer with a beam spacing equal to (\ref{eq:resolution}).
-Dotted vertical lines indicate position of transmit-receive beam pairs.
-The point scatterer response has been simulated using Field II \cite{Jensen1992, Jensen1996a} using a 96 element phased array with a 22 mm azimuth aperture, which was excited with a 2.5 MHz and 1.5 cycles long pulse.
-Transmit focus in azimuth and elevation were configured to be equal and the point scatterer was moved through this point.
-The transmit focus is where the ultrasound beam is at its narrowest, hence shift variance effects will appear at this location first.
-The beam density was calculated  based on (\ref{eq:resolution}) with the speed of sound equal to 1540.
-We see that the contours in Fig.\,\ref{fig:das_shiftinvariant} are approximately diagonal (low P-MAE) and that the amplitude variation among the maximum points is minimal (low A-MAE).
-We therefore conclude that the system is locally laterally shift-invariant.
-%In this, and the following LSV-plots based on simulations, the scan sequence always has a beam at zero degrees and the point scatterer is located in the center of this beam at one time instance.
-
-%The scalloping loss in Fig.\,\ref{fig:das_shiftinvariant} is smaller than what was predicted by Fig.\,\ref{fig:das_capon_beams}.
-To understand why note that Fig.\,\ref{fig:das_capon_beams} shows the scanned response for a narrowband system while the imaging system in Fig.\,\ref{fig:das_shiftinvariant} uses a broadband pulse.
-The beam in Fig.\,\ref{fig:das_shiftinvariant} will therefore be more similar to a squared gaussian function than the squared $sinc$ functions depicted in Fig.\,\ref{fig:das_capon_beams}.
-The imaging system in Fig.\,\ref{fig:das_shiftinvariant} also uses an expanding cosine window on receive.
-
-%The scalloping loss in Fig.\,\ref{fig:das_shiftinvariant} is smaller than what was predicted by Fig.\,\ref{fig:das_capon_beams}.
-To understand why note that Fig.\,\ref{fig:das_capon_beams} shows the one-way response of a passive array, and the system resolution has therefore been calculated based on the receive aperture only.
-Whereas the beamspacing in (\ref{eq:resolution}) is derived from the fact that the two-way lateral response of an active rectangular aperture is a $D_{tx}+D_{rx}$ wide triangle in K-space \cite{Hergum2009}.
-Even though the lateral bandwidth is increased on receive, the distance to the first zero (The Rayleigh criteria) remains the same when multiplying two sinc functions.
-Sampling according to (\ref{eq:resolution}) is therefore more than what is strictly needed in practise.
-The loss in Fig.\,\ref{fig:das_shiftinvariant} is because of this smaller than what was predicted.
-
-In Fig.\,\ref{fig:das_shiftvariant} the beam density is reduced by a factor of two compared to Fig.\,\ref{fig:das_shiftinvariant}.
-The geometric distortion is now increased to a level where the contours are not diagonal.
-The maximum scalloping loss is almost 4 dB.
-Hence, the imaging system is no longer locally laterally shift-invariant.
-We observe that the maximum peak is both geometrically distorted and attenuated when the point is located between two beams and the distance between the beams is too high.
-
-\begin{figure*}[!t]
-\centerline{
-\subfloat[]{\includegraphics[width=0.5\linewidth]{\img fig2a.eps}\label{fig:das_shiftinvariant}}
-\hfill{}
-\subfloat[]{\includegraphics[width=0.5\linewidth]{\img fig2b.eps}\label{fig:das_shiftvariant}}
-}
-\caption{LSV-plots of delay-and-sum beamforming.
-If contours are diagonal and the amplitude variation among the peak points (black dots) is small, the imaging system is said to be locally laterally shift-invariant.
-Dotted vertical lines are plotted where transmit-receive beams are located.
-a) Sampling density according to (\ref{eq:resolution}).
-The system is locally laterally shift-invariant.
-b) Undersampling by a factor of two.
-The system is no longer locally laterally shift-invariant.}
-\label{fig:das}
-\end{figure*}
-
-\section{Local Lateral Shift-Invariance of the Capon Beamformer}\label{sec:capon_LLSI}
-We will now investigate the local lateral shift-invariance of the Capon beamformer with high resolution parameters \cite{Synnevag2009, Asen}.
-In Fig.\,\ref{fig:capon1} the same point scatterer response as presented in Fig.\,\ref{fig:das_shiftinvariant} has been weighted with Capon weights calculated using (\ref{eq:capon_weights}) with $L = M/2 = 48$, $K=1$ and $d=1/100$.
-We see how the signal cancellation caused by the difference between the signal propagation vector and the assumed steering vector increases as the point moves away from a given transmit-receive beam.
-As a consequence the scalloping loss is tremendous, and not far from the predicted value in Fig.\,\ref{fig:das_capon_beams}.
-%The difference between the predicted and the simulated loss might be caused by several factors.
-The most important one being that Fig.\,\ref{fig:das_capon_beams} shows a single frequency whereas the RMS operator used in Fig.\,\ref{fig:capon1} average a broadband signal in range.
-For a slice through the center of the pulse the scalloping loss is below -30dB.
-
-The diagonal factor $d$ and the effective adaptive aperture size $L$ can be used to control the mainlobe width of the Capon beamformer.
-In the extreme cases, $d\gg1$ and $L=1$, the Capon beamformer becomes either a triangle or uniform-weighted DAS beamformer respectively, and a beam spacing according to (\ref{eq:resolution}) would then be sufficient.
-In Fig.\,\ref{fig:capon2} we show how the scalloping loss is reduced when the amount of diagonal loading is increased.
-
-We observe that Fig.\,\ref{fig:capon2} now contains a scaled version of the pattern presented in Fig.\,\ref{fig:das_shiftvariant}.
-From Fig.\,\ref{fig:capon} it should be clear that Capon beamforming either requires increased lateral sampling, or the resolution has to be reduced to the same level as DAS in order to make the imaging system locally laterally shift-invariant.
-It would be unwise to do all the computation involved with Capon beamforming with a high $d$ value and end up with a resolution equal to DAS.
-However, we see that diagonal loading can be a useful tool if resolution has to be traded for the need of oversampling.
-The same is true for the subarray length parameter.
-
-\begin{figure*}[!t]
-\centerline{
-\subfloat[]{
-\includegraphics[width=0.5\linewidth]{\img fig3a.eps}\label{fig:capon1}
-}
-\subfloat[]{
-\includegraphics[width=0.5\linewidth]{\img fig3b.eps}\label{fig:capon2}
-}
-}
-\caption{LSV-plots of Capon beamforming with $L=M/2=48$ and $K=1$, and with sampling corresponding to (\ref{eq:resolution}).
-a) Capon beamforming with $d=1/100$ as diagonal loading.
-The system is not locally laterally shift-invariant.
-b) The diagonal loading factor $d$ is increased to 1.
-The scalloping loss is reduced, but the system is still not locally laterally shift-invariant.}
-\label{fig:capon}
-\end{figure*}
-
-\section{Oversampling Methods}\label{sec:methods}
-In this section we will discuss different methods aimed at improving the local lateral shift-invariance of the Capon beamformer.
-From the previous section it should be clear that we somehow need to increase the lateral sampling to make the Capon imaging system locally laterally shift-invariant.
-In this section we will investigate how this oversampling should be conducted, how large it needs to be and how the proposed methods will affect the acquisition frame rate and the processing complexity.
-
-\subsection{Oversampling on Transmit}
-A straight forward way of obtaining sufficient sampling in an active imaging system with an equal number of transmit (tx) and receive (rx) beams is to increasing the number of tx-beams until the system is locally laterally shift-invariant.
-In Fig.\,\ref{fig:capon_oversampling_a} and Fig.\,\ref{fig:capon_oversampling_b} we have increased the tx-beam density by a $q$-factor of four and sixteen respectively.
-As expected the geometric distortion and scalloping loss decreases with increased lateral sampling.
-Note that Fig.\,\ref{fig:capon_oversampling_b} is zoomed compared to Fig.\,\ref{fig:capon} in order to better see the beam-to-beam variation.
-
-From Fig.\,\ref{fig:capon_oversampling_b} we see that sixteen times oversampling reduces the scalloping loss significantly.
-Even though high frequency variations are reduced, the maximum is still larger than 1 dB.
-An even higher level of oversampling is therefore required in order to have a completely local lateral shift-invariant Capon beamformer when a moving point scatterer is simulated without noise and with the same parameters as in Fig.\,\ref{fig:capon}.
-%$1/100$ in diagonal loading.
-Note that a diagonal loading factor of $1/100$ will introduce a white noise component in the sample covariance matrix $20$ dB lower than the signal present.
-Hence, the covariance has been estimated from data which appeared to have an SNR of 20 dB.
-
-Even though increasing the number of transmits improves the lateral shift-invariance it will also reduce the acquisition frame rate with a factor equal to the oversampling factor.
-The processing complexity will also increase equally.
-For these two reasons, oversampling on transmit is something which is desirable to avoid in real time ultrasound applications, and therefore we will not discuss this method further.
-
-\begin{figure*}[!t]
-\centerline{
-\subfloat[]{
-\includegraphics[width=0.5\linewidth]{\img fig4a.eps}\label{fig:capon_oversampling_a}
-}
-\subfloat[]{
-\includegraphics[width=0.5\linewidth]{\img fig4b.eps}\label{fig:capon_oversampling_b}
-}
-}
-\caption{Capon beamforming as described in Fig.\,\ref{fig:capon} with oversampling on transmit.
-The figure is zoomed compared with Fig.\,\ref{fig:capon} to better see the beam-to-beam variation.
-a) Four times oversampling ($q=4$).
-b) Sixteen times oversampling ($q=16$).
-The maximum scalloping loss is reduced, but it is still too high for the system to be locally laterally shift invariant.}
-\label{fig:capon_oversampling}
-\end{figure*}
-
-\subsection{Oversampling by Parallel Receive Beamforming}
-Parallel receive beamforming (PRB) means that multiple narrow receive beams (rx-beams) are constructed from one broad tx-beam.
-Then, by reducing the number of tx-beams and relying on multiple rx-beams in order to get sufficient sampling on receive, it is possible to achieve high frame rate imaging.
-However, PRB also introduces geometrical distortion which to some degree can be compensated for by using e.g.
-synthetic transmit beamforming (STB) \cite{Hergum2007, Denarie2013}.
-In \cite{Rabinovich2013} it is shown that STB works well with Capon beamforming, something which is also verified in \cite{Asen}.
-In the following discussion we consider corrected PRB as being equal to oversampling on transmit when it comes to image quality.
-The LSV-plots for PRB are therefore not included.
-
-In cardiac ultrasound imaging, PRB is typically used in order to keep the number of rx-beams high while the number of tx-beams are reduced.
-When using PRB for oversampling we would like to maintain the same number of tx-beams while the number of rx-beams are increased.
-The acquisition frame rate will then not be affected.
-On the other hand, more rx-beams will increase the number of covariance matrices and thereby the number of matrix-inversions with a factor equal to the number of parallel receive lines.
-The processing times reported in \cite{Asen} will then also increase with the same factor.
-Since more than sixteen times oversampling is required in order to achieve lateral-shift-invariant and high-resolution imaging of a moving point scatterer, none of the implementations described in \cite{Asen} would be real time any more.
-We therefore need a different way of oversampling our imaging sector.
-
-\subsection{Oversampling by Phase Rotation}
-The Capon beamformer formula in (\ref{eq:capon_weights}) gives a direct opportunity for oversampling using phase rotation.
-The steering vector $\vec{a}$ has in previous work been set to $\vec{1}$ since ultrasound data have to be pre-delayed.
-However, the steering vector can also, as in narrow band applications, be varied over a set of pre-defined vectors.
-In a narrow band setting the beam can make a sweep across the whole imaging sector through phase steering.
-This is also possible in broadband applications as long as the phase rotation is less than one pulse length (See coarse-fine beamforming in \cite{Thomenius} and its references).
-The maximal steering angle is approximately given by
-\begin{align}
-\theta_{max} = 2\alpha \delta\theta,
-\end{align}
-where $\alpha$ is the pulse length in $\lambda$'s, and $2$ and $\delta\theta$ is the oversampling factor and beam spacing from (\ref{eq:resolution}).
-For our simulations $\theta_{max} = 3\delta\theta$.
-It is therefore possible to make a sweep based on phase rotation between a given beam and half-way to its two nearest neighbours without introducing large errors.
-Phase rotation is actually what the Capon beamformer uses internally to micro-steer away from bright point scatterers and to enhance edges.
-
-For a uniform linear array located along the x-axis the steering vector $\vec{a}_\theta$ should be calculated as 
-\begin{align}
-\vec{a}_\theta = 
-\begin{bmatrix}
-e^{-j\frac{2\pi}{\lambda_c}x_0\sin(\theta)} \\
-e^{-j\frac{2\pi}{\lambda_c}x_1\sin(\theta)} \\
-\vdots \\ 
-e^{-j\frac{2\pi}{\lambda_c}x_{L-1}\sin(\theta)}
-\end{bmatrix}
-\end{align}
-where $x_i$ is the element position and $\theta$ is swept from $-\delta\theta/2$ to $\delta\theta/2$.
-Note that the steering vector must be calculated for an $L$ element symmetric subarray, even though the full array has $M$ elements.
-This subarray should have the same element pitch, $|x_i - x_{i-1}|$, as the full array.
-%The steering error introduced by this steering vector for a frequency band $B$ is less than $B/2 \times \delta\theta/2$.
-
-The steering angles can be spread out in many different ways.
-If the number of steering angles is $N_\theta$ we have selected a uniform distribution of $\theta$  with $N_\theta/2$ angles on each side of the original steering direction when $N_\theta$ is even.%, and $(N_\theta-1)/2$ [including the original steering direction] if $N_\theta$ is odd.
-
-Figure \ref{fig:capon_oversampling_pr} shows the phase rotation method applied on the same scan sequence as in Fig.\,\ref{fig:capon} with four and sixteen times oversampling on receive respectively ($q$-factor).
-Note that the figure is zoomed compared with Fig.\,\ref{fig:capon}.
-As with oversampling on transmit there is some variation when the point moves in and out of focus.
-The maximum scalloping loss is therefore large.
-The extreme local variations are however reduced significantly.
-Comparing Fig.\,\ref{fig:capon_oversampling} with Fig.\,\ref{fig:capon_oversampling_pr}, the errors introduced by the phase rotation method are best seen in sidelobes.
-
-Since the phase rotation method operates on the same channel data as in Fig.\,\ref{fig:capon1}, there will be no reduction in the acquisition frame rate by applying this method.
-In addition, since the same number of covariance matrices are constructed and inverted, there is no extra cost involved here as well.
-The only additional cost will be the multiplication of the extra steering vectors with the inverted covariance matrix.
-If the equation $\mat{\hat{R}}\vec{a} = \vec{b}$, as in \cite{Asen}, is solved using the Gauss-Jordan algorithm, we can solve the problem for all steering vectors at once by row reducing the matrix $[\mat{\hat{R}},  \vec{a}_{\theta_1}, \vec{a}_{\theta_2}, \cdots \vec{a}_{\theta_{N_\theta}}]$ to the following form $[\mat{I},  \vec{b}_{\theta_1}, \vec{b}_{\theta_2}, \cdots \vec{b}_{\theta_{N_\theta}}]$.
-This is an $O(L^3)$ operation as long as $N_\theta$ is not much greater than $L$.
-
-\begin{figure*}[!t]
-	\centerline{
-		\subfloat[]{
-			\includegraphics[width=0.5\linewidth]{\img fig5a.eps}
-		}
-		\subfloat[]{
-			\includegraphics[width=0.5\linewidth]{\img fig5b.eps}
-		}
-	}
-	\caption{Capon beamforming as described in Fig.\,\ref{fig:capon} with oversampling based on phase rotation.
-The figure is zoomed compared with Fig.\,\ref{fig:capon} to better see the beam-to-beam variation.
-a) Four times oversampling ($q=4$).
-b) Sixteen times oversampling ($q=16$).}
-	\label{fig:capon_oversampling_pr}
-\end{figure*}
-
-\section{In-vitro phantom data}\label{sec:res}
-In Fig.\,\ref{fig:LSV_invitro} we present the result of applying the Capon beamformer on \textit{in-vitro} data with and without oversampling by phase rotation.
-The data was acquired using a modified Vivid E9 high-end cardiac ultrasound scanner by GE Vingmed Ultrasound, Horten, Norway, equipped with a M5S-D phased transducer operating in harmonic mode at 3.34 MHz.
-The transmit beam density was set to match (\ref{eq:resolution}) based on this frequency, and 111 tx-beams where therefore distributed on a 65\degree sector.
-Transmit focus is located at 8 cm range.
-Note that in harmonic imaging the generated transmit beam at the harmonic frequency will be wider and have lower sidelobes than the beam resulting from transmitting at the harmonic frequency \cite{Fedewa2004}.
-The lateral sampling is therefore increased compared with our simulations.
-A tissue mimicking phantom containing wire targets and anechoic cysts was imaged, and the data was acquired while the probe was moved across the phantom manually.
-Fig.\,\ref{fig:phantom_das_image} shows the first frame out of 14 in the acquired loop processed with DAS beamforming.
-%Movies of the whole loop processed with DAS and Capon beamforming with and without ten times oversampling by phase rotation can be found in the supplementary material \multimedia{Videos}.
-
-One point scatterer in Fig.\,\ref{fig:phantom_das_image} has been selected [white box], and then used to generate LSV-plots.
-The point scatterer is located close to the transmit focus.
-Figure \ref{fig:phantom_das_LSV} shows the resulting LSV-plot when using the DAS beamformer.
-The scalloping loss is less than 1 dB, and the system is by that locally laterally shift-invariant.
-As a consequence point scatterers have a consistent appearance when subjected to motion \multimedia{Media Movie 1}.
-
-Figure \ref{fig:phantom_capon_LSV} shows the same point scatterer response processed with the Capon beamformer without oversampling.
-For the frame where the point scatterer is at -6 degrees the receive beam direction is very close to the signal propagation vector.
-Hence, we see a high amplitude value at this location.
-In almost all the other frames the signal loss is massive (at most 10 dB).
-The result is blinking point scatterers when subjected to motion \multimedia{Media Movie 2}.
-
-Figure \ref{fig:phantom_capon_LSV_16PR} presents the result of applying sixteen times oversampling by phase rotation.
-We observe how the amplitude of the point scatterer is better preserved in all frames, and the scalloping loss is not far from our 1 dB threshold.
-Note that the oscillation along the diagonal does not indicate shift-variance.
-It is a result of lack of control when the probe was manually dragged across the phantom.
-In order to get a diagonal without peaks the probe has to be dragged across the phantom with a speed which matches the acquisition frame rate.
-In addition, due to the increase in lateral resolution, the Capon beamformer requires smaller frame-to-frame movements than DAS.
-Looking at the video of Capon beamforming with sixteen times oversampling by phase rotation \multimedia{Media Movie 3} the blinking is now highly reduced.
-The same can be seen from Fig.\,\ref{fig:img_invitro} where point scatterers appear brighter in the image with oversampling by phase rotation.%The system is however not fully shift invariant.
-To achieve this a higher level of oversampling, %We therefore conclude that the system is now locally laterally shift invariant.
-
-%Acquire channel data from a phantom containing point wire targets and cysts while moving the probe laterally.
-Confirm the improved shift-invariance when using phase rotation.
-See if improved sampling also increase cyst contrast.
-%Make plot of contrast between two areas in a simulation for different values of $L$.
-Make the same plot for different number of rx beams.
-Do the same for an in-vivo image.
-The rational being that contrast is also improved by denser rx-sampling not just shift-invariance.
-
-%\begin{figure}\caption{Plot of contrast vs oversampling factor (phase rotation) in in-vivo data.}
-%\end{figure}
-
-%\subsection{In-vitro Images}
-%Acquire channel data from a phantom containing point wire targets and cysts while moving the probe laterally. Confirm the improved shift-invariance when using phase rotation.
-
-\begin{figure*}[!t]
-\centerline{
-\begin{tabular}{cc}
-\subfloat[]{\includegraphics[width=0.45\linewidth]{\img fig6a.png}\label{fig:phantom_das_image}}
-\subfloat[]{\includegraphics[width=0.45\linewidth]{\img fig6b.eps}\label{fig:phantom_das_LSV}}\\
-\subfloat[]{\includegraphics[width=0.45\linewidth]{\img fig6c.eps}\label{fig:phantom_capon_LSV}}
-\subfloat[]{\includegraphics[width=0.45\linewidth]{\img fig6d.eps}\label{fig:phantom_capon_LSV_16PR}}
-\end{tabular}
-}
-\caption{LSV-plots generated from \textit{in-vitro} phantom data where the probe has been manually dragged across the phantom.
-See the supplementary material for videos of the whole loop processed with the different beamformers.
-Capon beamforming has been performed with the same settings as in Fig.\,\ref{fig:capon1}.
-P-MAE is not shown due to a lack of probe movement control.
-a) First frame in the recorded loop processed with delay-and-sum beamforming  \multimedia{Media Movie 1}.
-A white box indicates the point scatterer used to construct the LSV-plots.
-b) LSV-plot for delay-and-sum beamforming.
-c) LSV-plot for Capon beamforming without oversampling.
-d) LSV-plot for Capon beamforming with sixteen times oversampling using phase rotation.}
-\label{fig:LSV_invitro}
-\end{figure*}
-
-\begin{figure*}[!t]
-\centerline{
-\subfloat[]{\includegraphics[width=0.48\linewidth]{\img fig7a.png}\label{fig:phantom_capon_image}}
-\subfloat[]{\includegraphics[width=0.48\linewidth]{\img fig7b.png}\label{fig:phantom_capon_16PR_image}}
-}
-\caption{First frame from Fig.\,\ref{fig:phantom_das_image} processed with Capon beamforming.
-a) Without oversampling \multimedia{Media Movie 2}.
-b) With sixteen times oversampling using phase rotation \multimedia{Media Movie 3}.
-Notice how the point scatterers appear brighter in the image with oversampling.}
-\label{fig:img_invitro}
-\end{figure*}
-
-%\subsection{In-Vivo Images}
-%Investigate if phase rotation improves contrast in in-vivo images of the heart or liver.
-
-%\begin{figure*}[!t]
-%\centerline{
-%\subfloat[]{\includegraphics[width=0.5\linewidth]{\img phantom1_capon_processing_K=0_L=24_full.png}\label{fig:in-vivo_capon}}
-%\hfill{}
-%\subfloat[]{\includegraphics[width=0.5\linewidth]{\img phantom1_capon_processing_K=0_L=24_8pr_full.png}\label{fig:in-vivo_capon_8pr}}
-%\hfill{}
-%\subfloat[]{\includegraphics[width=0.3\linewidth]{\img das_undersampled2_2x.eps}\label{fig:das_shiftvariant}}
-%}
-%\caption{\comment{For the moment phantom data. Need to acquire some high quality in-vivo data.}. In-vivo data. a) Capon image. b) Capon image with oversampling using 8PR.}
-%\label{fig:das}
-%\end{figure*}
-
-\section{Discussion}\label{sec:dis}
-%Note that in this paper we have investigated a single array configuration and a limited parameter range for the Capon beamformer.
-The goal was not to make a reference for which oversampling factor to use in different scenarios, but rather do a detailed investigation of the scalloping loss effect.
-We hope with this investigation that future readers will be able to find out them self what oversampling factor to use for their system.
-
-Faced with an imaging scene containing bright point scatterers or sharp edges, one could wonder how large the oversampling factor needs to be in order to avoid visible scalloping losses.
-In his seminal paper, Cox \cite{Cox1973} derives the required beam density for a given output SNR.
-For instance 47 dB requires a beam density for the Capon beamformer ten times the one given by (\ref{eq:resolution}) [with $q=1/2$].
-Since resolution for the Capon beamformer depends on SNR we would recommend to first use the dynamic range in the image as guidance.
-The amount of diagonal loading will also dictate the maximal achievable resolution since $10\log{d}$ can be interpreted as the maximal input SNR seen by the Capon beamformer.
-Fig.\,\ref{fig:das_capon_beams} shows that 25 times oversampling is required for a $d$-value of $1/100$.
-When \textit{in-vitro} data contains additional noise this oversampling requirement is reduced.%Cox measures resolution as the 3 dB saddle point between two impinging monochromatic plane waves.
-For ultrasound data only a 50 \% increase in resolution was found in \cite[Fig.~7]{Asen} for a 30 dB output SNR when measured in a similar way as Cox [6 dB saddle point].
-However, sixteen times oversampling was required to make the imaging system shift invariant in the same paper.
-In this paper more than ten times oversampling was found to be required for a 40 dB output SNR (after array gain) in simulations, however for the \textit{in-vitro} data ten times oversampling was enough for the system to be locally laterally shift-invariant.
-This might be a result of improved initial sampling (as discussed in Section \ref{sec:res}) and a higher initial noise level before diagonal loading was applied.
-
-Relating the presented phase rotation method to earlier work we find that it is similar to a steering vector uncertainty set.
-However, where these set-based methods typically has a criterion for selecting the best output after applying all the steering vectors contained in the set, oversampling by phase rotation outputs all.
-It can also be thought of as a narrow band approach to parallel receive beamforming where Capon's method is applied on all receive beams.
-
-Oversampling by phase rotation can therefore introduce similar block artifacts as parallel receive beamforming (PRB).
-To remove these artifacts we could distribute our steering vectors in an overlapping scheme [like rx-beams are spread out for STB].
-We then get a phase rotation version of STB where e.g.
-eight phase rotated beams are averaged into four.
-If the geometric distortion, or stripes, seen at the bottom of the sector in the video with sixteen times oversampling \multimedia{Media Movie 3} is removed by such technique is left for future work.
-
-When oversampling is applied it becomes evident that the high resolution settings used in this paper might be too extreme, both with respect to the level of required oversampling and the point scatterer's appearance.
-Traditionally in medical ultrasound imaging the point spread function is always wider laterally than in range.
-With the Capon beamformer this is now changed.
-A more symmetric point spread function while preserving speckle statistics is obtained with the parameters $L=M/4$ and $K=0$ \cite{Synnevag2007a}.
-The level of required oversampling and the computational complexity are also reduced.
-%Another option is to perform pulse inversion if a spiky symmetric point spread function is desirable.
-
-Note that oversampling is not only essential to keep the amplitude level of moving point scatterers.
-It should also help to improve contrast and edge definitions.
-If oversampling is not applied homogeneous speckle regions could get a too low average value and the edge delineation of a moving object will fluctuate in the same way as a moving point scatterer would have sparkled.
-
-Finally we want to stress the fact that this demand for oversampling does not only affect the Capon [or minimum variance beamformer].
-Other methods claiming super resolution, like the Eigenspace method, low complexity beamformer etc.
-\cite{Synnevag2011, Nilsen2009, Mehdizadeh2012a, Kim}, would also require oversampling when improved lateral resolution is obtained.
-The same applies for high resolution spectral estimation \cite{Ekroll2012}.
-
-%When signal detection (absolute value and square root) is performed before an ultrasound image is displayed the lateral bandwidth is increased.
-In order to not introduce aliasing in the image, IQ-interpolation is typically used on ultrasound systems to double the lateral beam density before detection.
-For the same reason we need to increase the lateral sampling when Capon beamforming is applied, but now on a channel data level.
-The Capon beamformer improves the lateral resolution, hence the lateral bandwidth is increased, and higher lateral sampling is required.
-
-%One way of reducing the number of matrix inversions in PRB is to estimate the covariance matrix from multiple beams \cite{Jensen2012}, generating one covariance matrix per range.
-Tailoring the method of Jensen and Austeng to cardiac ultrasound imaging it makes more sense, due to motion, to construct one covariance matrix per tx-beam using multiple parallel rx-beams.
-With this solution the number of matrix inversions will not changes, but the time taken to construct the matrices will increase.
-We will not investigate this method further in this paper.
-
-%\subsection{Impact on Real Time Performance}
-%Discuss, and show with benchmarks how the different methods for making Capon beamforming shift-invariant impact the total execution time.
-
-\section{Conclusion}\label{sec:con}
-%In this paper we have investigated the lateral shift-invariance property of the Capon beamformer when imaging moving objects.
-We have shown that the Capon beamformer is less sensitive to small lateral shifts if oversampling based on phase rotation is applied.
-This without reducing the acquisition frame rate and with only a minor increase in processing complexity.
-%In Section \ref{sec:capon_LLSI} it was shown that a Capon beamformer with high resolution parameters is not locally laterally shift-invariant when subjected to motion.
-The root cause was found to be too coarse lateral sampling.
-%In Section \ref{sec:methods} we presented methods for improved shift-invariance based on oversampling.
-The most promising method, with respect to improved shift-invariance, reduction in acquisition frame rate and processing complexity combined, was based on phase rotation.
-%In Section \ref{sec:res} we presented results showing that this method also works on \textit{in-vitro} data when the ultrasound probe is subjected to motion.
-
-In this paper we have investigated the lateral shift-invariance property of the Capon beamformer when imaging moving objects.
-It was shown that Capon beamforming with the same beam density as required for delay-and-sum beforming results in blinking point scatterers when the system is subjected to motion.
-We then discussed and investigated different methods based on oversampling on transmit and receive in order to reduced this effect.
-Finally we showed that the local lateral shift-invariance could be improved by oversampling based on phase rotation on receive.
-This without affecting the acquisition frame rate and with a minor change in processing complexity.
-In simulations, this method reduced the maximum scalloping loss by 16 dB.
-On \textit{in-vitro} data the reduction was around 7 dB and the point scatterer blinking was highly reduced.
-
-% An example of a floating figure using the graphicx package.
-% Note that \label must occur AFTER (or within) \caption.
-% For figures, \caption should occur after the \includegraphics.
-% Note that IEEEtran v1.7 and later has special internal code that
-% is designed to preserve the operation of \label within \caption
-% even when the captionsoff option is in effect. However, because
-% of issues like this, it may be the safest practice to put all your
-% \label just after \caption rather than within \caption{}.
-%
-% Reminder: the "draftcls" or "draftclsnofoot", not "draft", class
-% option should be used if it is desired that the figures are to be
-% displayed while in draft mode.
-%
-%\begin{figure}[!t]
-%\centering
-%\includegraphics[width=2.5in]{myfigure}
-% where an .eps filename suffix will be assumed under latex, 
-% and a .pdf suffix will be assumed for pdflatex; or what has been declared
-% via \DeclareGraphicsExtensions.
-%\caption{Simulation Results}
-%\label{fig_sim}
-%\end{figure}
-
-% Note that IEEE typically puts floats only at the top, even when this
-% results in a large percentage of a column being occupied by floats.
-
-% An example of a double column floating figure using two subfigures.
-% (The subfig.sty package must be loaded for this to work.)
-% The subfigure \label commands are set within each subfloat command, the
-% \label for the overall figure must come after \caption.
-% \hfil must be used as a separator to get equal spacing.
-% The subfigure.sty package works much the same way, except \subfigure is
-% used instead of \subfloat.
-%
-%\begin{figure*}[!t]
-%\centerline{\subfloat[Case I]\includegraphics[width=2.5in]{subfigcase1}%
-%\label{fig_first_case}}
-%\hfil
-%\subfloat[Case II]{\includegraphics[width=2.5in]{subfigcase2}%
-%\label{fig_second_case}}}
-%\caption{Simulation results}
-%\label{fig_sim}
-%\end{figure*}
-%
-% Note that often IEEE papers with subfigures do not employ subfigure
-% captions (using the optional argument to \subfloat), but instead will
-% reference/describe all of them (a), (b), etc., within the main caption.
-
-% An example of a floating table. Note that, for IEEE style tables, the 
-% \caption command should come BEFORE the table. Table text will default to
-% \footnotesize as IEEE normally uses this smaller font for tables.
-% The \label must come after \caption as always.
-%
-%\begin{table}[!t]
-%% increase table row spacing, adjust to taste
-%\renewcommand{\arraystretch}{1.3}
-% if using array.sty, it might be a good idea to tweak the value of
-% \extrarowheight as needed to properly center the text within the cells
-%\caption{An Example of a Table}
-%\label{table_example}
-%\centering
-%% Some packages, such as MDW tools, offer better commands for making tables
-%% than the plain LaTeX2e tabular which is used here.
-%\begin{tabular}{|c||c|}
-%\hline
-%One & Two\\
-%\hline
-%Three & Four\\
-%\hline
-%\end{tabular}
-%\end{table}
-
-% Note that IEEE does not put floats in the very first column - or typically
-% anywhere on the first page for that matter. Also, in-text middle ("here")
-% positioning is not used. Most IEEE journals use top floats exclusively.
-% Note that, LaTeX2e, unlike IEEE journals, places footnotes above bottom
-% floats. This can be corrected via the \fnbelowfloat command of the
-% stfloats package.
-
-%\section{Conclusion}
-%The conclusion goes here.
-
-% if have a single appendix:
-%\appendix[Proof of the Zonklar Equations]
-% or
-%\appendix  % for no appendix heading
-% do not use \section anymore after \appendix, only \section*
-% is possibly needed
-
-% use appendices with more than one appendix
-% then use \section to start each appendix
-% you must declare a \section before using any
-% \subsection or using \label (\appendices by itself
-% starts a section numbered zero.)
-%
-
-%\appendices
-%\section{Proof of the First Zonklar Equation}
-%Appendix one text goes here.
-
-% you can choose not to have a title for an appendix
-% if you want by leaving the argument blank
-%\section{}
-%Appendix two text goes here.
-
-% use section* for acknowledgement
-\section*{Acknowledgment}
-
-The authors would like to thank Anders S\o{}rnes at GE Vingmed Ultrasound, Horten, Norway for the help with acquiring \textit{in-vitro} data.
-
-% Can use something like this to put references on a page
-% by themselves when using endfloat and the captionsoff option.
-\ifCLASSOPTIONcaptionsoff
-  \newpage
-\fi
-
-% trigger a \newpage just before the given reference
-% number - used to balance the columns on the last page
-% adjust value as needed - may need to be readjusted if
-% the document is modified later
-%\IEEEtriggeratref{8}
-% The "triggered" command can be changed if desired:
-%\IEEEtriggercmd{\enlargethispage{-5in}}
-
-% references section
-
-% can use a bibliography generated by BibTeX as a .bbl file
-% BibTeX documentation can be easily obtained at:
-% http://www.ctan.org/tex-archive/biblio/bibtex/contrib/doc/
-% The IEEEtran BibTeX style support page is at:
-% http://www.michaelshell.org/tex/ieeetran/bibtex/
-\bibliographystyle{IEEEtran}
-% argument is your BibTeX string definitions and bibliography database(s)
-%\bibliography{IEEEabrv,../bib/paper}
-\bibliography{mybib}
-%
-% <OR> manually copy in the resultant .bbl file
-% set second argument of \begin to the number of references
-% (used to reserve space for the reference number labels box)
-%\begin{thebibliography}{1}
-
-%\bibitem{IEEEhowto:kopka}
-%H.~Kopka and P.~W. Daly, \emph{A Guide to \LaTeX}, 3rd~ed.\hskip 1em plus
-%  0.5em minus 0.4em\relax Harlow, England: Addison-Wesley, 1999.
-
-%\end{thebibliography}
-
-% biography section
-% 
-% If you have an EPS/PDF photo (graphicx package needed) extra braces are
-% needed around the contents of the optional argument to biography to prevent
-% the LaTeX parser from getting confused when it sees the complicated
-% \includegraphics command within an optional argument. (You could create
-% your own custom macro containing the \includegraphics command to make things
-% simpler here.)
-%\begin{biography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{mshell}}]{Michael Shell}
-% or if you just want to reserve a space for a photo:
-
-% insert where needed to balance the two columns on the last page with
-% biographies
-
-%\newpage
-
-\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{photos/jon_petter2.eps}}]{Jon Petter \AA{}sen}
-(S12) was born in Porsgrunn, Norway in 1986.
-He received the B.Sc.
-and M.Sc.
-degree in computer science from the University of Oslo, Norway, in 2008 and 2010.
-He is currently pursuing his Ph.D.
-degree in medical ultrasound technology at the Norwegian University of Science and Technology (NTNU) Medical Imaging Lab (MI-Lab), Trondheim, Norway.
-
-He has industry experience from GE Vingmed Ultrasound, Horten, Norway, were he completed three summer internships from 2008 to 2010.
-His research interests include image and signal processing, adaptive ultrasound processing techniques and acceleration of ultrasound algorithms using Graphics Processing Units (GPUs).
-\end{IEEEbiography}
-
-\vfill{}
-
-%\newpage
-
-\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{photos/andreas.eps}}]{Andreas Austeng}
+\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{bio/andreas.eps}}]{Andreas Austeng}
 was born in Oslo, Norway, in 1970.
-He received the M.S.
+He received the M.Sc.
 degree in physics in 1996 and the Ph.D.
@@
 \end{IEEEbiography}
-
-\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{photos/sverre.eps}}]{Sverre Holm}
-(M82-SM02) was born in Oslo, Norway, in 1954.
-He received the M.S.
-and Ph.D.
-degrees in electrical engineering from the Norwegian Institute of Technology, Trondheim, in 1978 and 1982, respectively.
-His academic experience includes Yarmouk University in Jordan (1984-1986) and the Norwegian Institute of Technology and later the University of Oslo (1989-1994) as an adjunct professor.
-Since 1995, he has been a professor of signal processing in the Department of Informatics at the University of Oslo.
-In 2002, he was elected a member of the Norwegian Academy of Technological Sciences, and in 2009, he became an adjunct professor in the Norwegian University of Science and Technology (NTNU) Medical Imaging Lab, Trondheim.
-His industry experience includes GE Vingmed Ultrasound, Horten, Norway (1990-1994).
-He worked with Sonitor Technologies (2000-2005), where he developed ultrasonic technology that is the core of the indoor positioning system described in \textit{Scientific Americanin} 2008 as "a positioning system that goes where GPS can't." He has published about 130 publications internationally, plus about 10 patents.
-He has had sabbaticals at GE Global Research, NY (1998), and the Waves and Acoustics Laboratory of ESPCI, Paris (2008-2009).
-He was an associate editor of the \textit{IEEE Transactions on Ultrasonics, Ferroelectrics, and Frequency Control} from 1997 to 2002.
-His research interests are medical ultrasound imaging, modeling of tissue-ultrasound interaction, and ultrasound positioning.
-\end{IEEEbiography}
-
-\vfill
-
-% You can push biographies down or up by placing
-% a \vfill before or after them. The appropriate
-% use of \vfill depends on what kind of text is
-% on the last page and whether or not the columns
-% are being equalized.
-
-%\vfill
-
-% Can be used to pull up biographies so that the bottom of the last one
-% is flush with the other column.
-%\enlargethispage{-5in}
-
-% that's all folks
-\end{document}

### 2014-04-23 HowtoHowtoCapon/bio/jo_inge_buskenes.tex
JOE Final submit

--- 2013-04-03 HRUIonGPU/jrnl_article.tex
@@
-%% bare_jrnl.tex
-%% V1.3
-%% 2007/01/11
-%% by Michael Shell
-%% see http://www.michaelshell.org/
-%% for current contact information.
-%%
-%% This is a skeleton file demonstrating the use of IEEEtran.cls
-%% (requires IEEEtran.cls version 1.7 or later) with an IEEE journal paper.
-%%
-%% Support sites:
-%% http://www.michaelshell.org/tex/ieeetran/
-%% http://www.ctan.org/tex-archive/macros/latex/contrib/IEEEtran/
-%% and
-%% http://www.ieee.org/
-
-% *** Authors should verify (and, if needed, correct) their LaTeX system  ***
-% *** with the testflow diagnostic prior to trusting their LaTeX platform ***
-% *** with production work. IEEE's font choices can trigger bugs that do  ***
-% *** not appear when using other class files.                            ***
-% The testflow support page is at:
-% http://www.michaelshell.org/tex/testflow/
-
-%%*************************************************************************
-%% Legal Notice:
-%% This code is offered as-is without any warranty either expressed or
-%% implied; without even the implied warranty of MERCHANTABILITY or
-%% FITNESS FOR A PARTICULAR PURPOSE! 
-%% User assumes all risk.
-%% In no event shall IEEE or any contributor to this code be liable for
-%% any damages or losses, including, but not limited to, incidental,
-%% consequential, or any other damages, resulting from the use or misuse
-%% of any information contained here.
-%%
-%% All comments are the opinions of their respective authors and are not
-%% necessarily endorsed by the IEEE.
-%%
-%% This work is distributed under the LaTeX Project Public License (LPPL)
-%% ( http://www.latex-project.org/ ) version 1.3, and may be freely used,
-%% distributed and modified. A copy of the LPPL, version 1.3, is included
-%% in the base LaTeX documentation of all distributions of LaTeX released
-%% 2003/12/01 or later.
-%% Retain all contribution notices and credits.
-%% ** Modified files should be clearly indicated as such, including  **
-%% ** renaming them and changing author support contact information. **
-%%
-%% File list of work: IEEEtran.cls, IEEEtran_HOWTO.pdf, bare_adv.tex,
-%%                    bare_conf.tex, bare_jrnl.tex, bare_jrnl_compsoc.tex
-%%*************************************************************************
-
-% Note that the a4paper option is mainly intended so that authors in
-% countries using A4 can easily print to A4 and see how their papers will
-% look in print - the typesetting of the document will not typically be
-% affected with changes in paper size (but the bottom and side margins will).
-% Use the testflow package mentioned above to verify correct handling of
-% both paper sizes by the user's LaTeX system.
-%
-% Also note that the "draftcls" or "draftclsnofoot", not "draft", option
-% should be used if it is desired that the figures are to be displayed in
-% draft mode.
-%
-
-%\documentclass[journal]{IEEEtran}
-\documentclass[11pt, draftcls, onecolumn]{IEEEtran}
-
-% Single column review mode:
-%\documentclass[12pt,journal,onecolumn]{IEEEtran}
-%\newcommand\CLASSINPUTbaselinestretch{2}
-%\RequirePackage{calc}
-%\RequirePackage{fp} 
-% Left margin = 1 inch + hoffset + oddsidemargin (or evensidemargin)
-% Adding the 2cm gutter width to the odd/even side margins:
-%\setlength\hoffset{0pt}
-%\setlength\oddsidemargin{4cm}       % 4cm margins on the left side
-%\addtolength\oddsidemargin{-1in}    % Subtract the initial 1 inch
-%\setlength\textwidth{21cm-8cm}    % A4 width (21cm) minus margins on either side
-
-%
-% If IEEEtran.cls has not been installed into the LaTeX system files,
-% manually specify the path to it like:
-% \documentclass[journal]{../sty/IEEEtran}
-
-% Some very useful LaTeX packages include:
-% (uncomment the ones you want to load)
-
-% *** MISC UTILITY PACKAGES ***
-%
-%\usepackage{ifpdf}
-% Heiko Oberdiek's ifpdf.sty is very useful if you need conditional
-% compilation based on whether the output is pdf or dvi.
-% usage:
-% \ifpdf
-%   % pdf code
-% \else
-%   % dvi code
-% \fi
-% The latest version of ifpdf.sty can be obtained from:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/oberdiek/
-% Also, note that IEEEtran.cls V1.7 and later provides a builtin
-% \ifCLASSINFOpdf conditional that works the same way.
-% When switching from latex to pdflatex and vice-versa, the compiler may
-% have to be run twice to clear warning/error messages.
-
-% *** CITATION PACKAGES ***
-%
-\usepackage{cite}
-% cite.sty was written by Donald Arseneau
-% V1.6 and later of IEEEtran pre-defines the format of the cite.sty package
-% \cite{} output to follow that of IEEE. Loading the cite package will
-% result in citation numbers being automatically sorted and properly
-% "compressed/ranged". e.g., [1], [9], [2], [7], [5], [6] without using
-% cite.sty will become [1], [2], [5]--[7], [9] using cite.sty. cite.sty's
-% \cite will automatically add leading space, if needed. Use cite.sty's
-% noadjust option (cite.sty V3.8 and later) if you want to turn this off.
-% cite.sty is already installed on most LaTeX systems. Be sure and use
-% version 4.0 (2003-05-27) and later if using hyperref.sty. cite.sty does
-% not currently provide for hyperlinked citations.
-% The latest version can be obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/cite/
-% The documentation is contained in the cite.sty file itself.
-
-% *** GRAPHICS RELATED PACKAGES ***
-%
-\ifCLASSINFOpdf
-  % \usepackage[pdftex]{graphicx}
-  % declare the path(s) where your graphic files are
-  % \graphicspath{{../pdf/}{../jpeg/}}
-  % and their extensions so you won't have to specify these with
-  % every instance of \includegraphics
-  % \DeclareGraphicsExtensions{.pdf,.jpeg,.png}
-\else
-  % or other class option (dvipsone, dvipdf, if not using dvips). graphicx
-  % will default to the driver specified in the system graphics.cfg if no
-  % driver is specified.
-  % \usepackage[dvips]{graphicx}
-  % declare the path(s) where your graphic files are
-  % \graphicspath{{../eps/}}
-  % and their extensions so you won't have to specify these with
-  % every instance of \includegraphics
-  % \DeclareGraphicsExtensions{.eps}
-\fi
-% graphicx was written by David Carlisle and Sebastian Rahtz. It is
-% required if you want graphics, photos, etc. graphicx.sty is already
-% installed on most LaTeX systems. The latest version and documentation can
-% be obtained at: 
-% http://www.ctan.org/tex-archive/macros/latex/required/graphics/
-% Another good source of documentation is "Using Imported Graphics in
-% LaTeX2e" by Keith Reckdahl which can be found as epslatex.ps or
-% epslatex.pdf at: http://www.ctan.org/tex-archive/info/
-%
-% latex, and pdflatex in dvi mode, support graphics in encapsulated
-% postscript (.eps) format. pdflatex in pdf mode supports graphics
-% in .pdf, .jpeg, .png and .mps (metapost) formats. Users should ensure
-% that all non-photo figures use a vector format (.eps, .pdf, .mps) and
-% not a bitmapped formats (.jpeg, .png). IEEE frowns on bitmapped formats
-% which can result in "jaggedy"/blurry rendering of lines and letters as
-% well as large increases in file sizes.
-%
-% You can find documentation about the pdfTeX application at:
-% http://www.tug.org/applications/pdftex
-
-% *** MATH PACKAGES ***
-%
-%\usepackage[cmex10]{amsmath}
-% A popular package from the American Mathematical Society that provides
-% many useful and powerful commands for dealing with mathematics. If using
-% it, be sure to load this package with the cmex10 option to ensure that
-% only type 1 fonts will utilized at all point sizes. Without this option,
-% it is possible that some math symbols, particularly those within
-% footnotes, will be rendered in bitmap form which will result in a
-% document that can not be IEEE Xplore compliant!
-%
-% Also, note that the amsmath package sets \interdisplaylinepenalty to 10000
-% thus preventing page breaks from occurring within multiline equations. Use:
-%\interdisplaylinepenalty=2500
-% after loading amsmath to restore such page breaks as IEEEtran.cls normally
-% does. amsmath.sty is already installed on most LaTeX systems. The latest
-% version and documentation can be obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/required/amslatex/math/
-
-% *** SPECIALIZED LIST PACKAGES ***
-%
-%\usepackage{algorithmic}
-% algorithmic.sty was written by Peter Williams and Rogerio Brito.
-% This package provides an algorithmic environment fo describing algorithms.
-% You can use the algorithmic environment in-text or within a figure
-% environment to provide for a floating algorithm. Do NOT use the algorithm
-% floating environment provided by algorithm.sty (by the same authors) or
-% algorithm2e.sty (by Christophe Fiorio) as IEEE does not use dedicated
-% algorithm float types and packages that provide these will not provide
-% correct IEEE style captions. The latest version and documentation of
-% algorithmic.sty can be obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/algorithms/
-% There is also a support site at:
-% http://algorithms.berlios.de/index.html
-% Also of interest may be the (relatively newer and more customizable)
-% algorithmicx.sty package by Szasz Janos:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/algorithmicx/
-
-% *** ALIGNMENT PACKAGES ***
-%
-%\usepackage{array}
-% Frank Mittelbach's and David Carlisle's array.sty patches and improves
-% the standard LaTeX2e array and tabular environments to provide better
-% appearance and additional user controls. As the default LaTeX2e table
-% generation code is lacking to the point of almost being broken with
-% respect to the quality of the end results, all users are strongly
-% advised to use an enhanced (at the very least that provided by array.sty)
-% set of table tools. array.sty is already installed on most systems. The
-% latest version and documentation can be obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/required/tools/
-
-%\usepackage{mdwmath}
-%\usepackage{mdwtab}
-% Also highly recommended is Mark Wooding's extremely powerful MDW tools,
-% especially mdwmath.sty and mdwtab.sty which are used to format equations
-% and tables, respectively. The MDWtools set is already installed on most
-% LaTeX systems. The lastest version and documentation is available at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/mdwtools/
-
-% IEEEtran contains the IEEEeqnarray family of commands that can be used to
-% generate multiline equations as well as matrices, tables, etc., of high
-% quality.
-
-%\usepackage{eqparbox}
-% Also of notable interest is Scott Pakin's eqparbox package for creating
-% (automatically sized) equal width boxes - aka "natural width parboxes".
-% Available at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/eqparbox/
-
-% *** SUBFIGURE PACKAGES ***
-%\usepackage[tight,footnotesize]{subfigure}
-% subfigure.sty was written by Steven Douglas Cochran. This package makes it
-% easy to put subfigures in your figures. e.g., "Figure 1a and 1b". For IEEE
-% work, it is a good idea to load it with the tight package option to reduce
-% the amount of white space around the subfigures. subfigure.sty is already
-% installed on most LaTeX systems. The latest version and documentation can
-% be obtained at:
-% http://www.ctan.org/tex-archive/obsolete/macros/latex/contrib/subfigure/
-% subfigure.sty has been superceeded by subfig.sty.
-
-%\usepackage[caption=false]{caption}
-%\usepackage[font=footnotesize]{subfig}
-% subfig.sty, also written by Steven Douglas Cochran, is the modern
-% replacement for subfigure.sty. However, subfig.sty requires and
-% automatically loads Axel Sommerfeldt's caption.sty which will override
-% IEEEtran.cls handling of captions and this will result in nonIEEE style
-% figure/table captions. To prevent this problem, be sure and preload
-% caption.sty with its "caption=false" package option. This is will preserve
-% IEEEtran.cls handing of captions. Version 1.3 (2005/06/28) and later 
-% (recommended due to many improvements over 1.2) of subfig.sty supports
-% the caption=false option directly:
-\usepackage[caption=false,font=footnotesize]{subfig}
-%
-% The latest version and documentation can be obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/subfig/
-% The latest version and documentation of caption.sty can be obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/caption/
-
-% *** FLOAT PACKAGES ***
-%
-%\usepackage{fixltx2e}
-% fixltx2e, the successor to the earlier fix2col.sty, was written by
-% Frank Mittelbach and David Carlisle. This package corrects a few problems
-% in the LaTeX2e kernel, the most notable of which is that in current
-% LaTeX2e releases, the ordering of single and double column floats is not
-% guaranteed to be preserved. Thus, an unpatched LaTeX2e can allow a
-% single column figure to be placed prior to an earlier double column
-% figure. The latest version and documentation can be found at:
-% http://www.ctan.org/tex-archive/macros/latex/base/
-
-%\usepackage{stfloats}
-% stfloats.sty was written by Sigitas Tolusis. This package gives LaTeX2e
-% the ability to do double column floats at the bottom of the page as well
-% as the top. (e.g., "\begin{figure*}[!b]" is not normally possible in
-% LaTeX2e). It also provides a command:
-%\fnbelowfloat
-% to enable the placement of footnotes below bottom floats (the standard
-% LaTeX2e kernel puts them above bottom floats). This is an invasive package
-% which rewrites many portions of the LaTeX2e float routines. It may not work
-% with other packages that modify the LaTeX2e float routines. The latest
-% version and documentation can be obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/sttools/
-% Documentation is contained in the stfloats.sty comments as well as in the
-% presfull.pdf file. Do not use the stfloats baselinefloat ability as IEEE
-% does not allow \baselineskip to stretch. Authors submitting work to the
-% IEEE should note that IEEE rarely uses double column equations and
-% that authors should try to avoid such use. Do not be tempted to use the
-% cuted.sty or midfloat.sty packages (also by Sigitas Tolusis) as IEEE does
-% not format its papers in such ways.
-
-%\ifCLASSOPTIONcaptionsoff
-%  \usepackage[nomarkers]{endfloat}
-% \let\MYoriglatexcaption\caption
-% \renewcommand{\caption}[2][\relax]{\MYoriglatexcaption[#2]{#2}}
-%\fi
-% endfloat.sty was written by James Darrell McCauley and Jeff Goldberg.
-% This package may be useful when used in conjunction with IEEEtran.cls'
-% captionsoff option. Some IEEE journals/societies require that submissions
-% have lists of figures/tables at the end of the paper and that
-% figures/tables without any captions are placed on a page by themselves at
-% the end of the document. If needed, the draftcls IEEEtran class option or
-% \CLASSINPUTbaselinestretch interface can be used to increase the line
-% spacing as well. Be sure and use the nomarkers option of endfloat to
-% prevent endfloat from "marking" where the figures would have been placed
-% in the text. The two hack lines of code above are a slight modification of
-% that suggested by in the endfloat docs (section 8.3.1) to ensure that
-% the full captions always appear in the list of figures/tables - even if
-% the user used the short optional argument of \caption[]{}.
-% IEEE papers do not typically make use of \caption[]'s optional argument,
-% so this should not be an issue. A similar trick can be used to disable
-% captions of packages such as subfig.sty that lack options to turn off
-% the subcaptions:
-% For subfig.sty:
-%\let\MYorigsubfloat\subfloat
-%\renewcommand{\subfloat}[2][\relax]{\MYorigsubfloat[]{#2}}
-% For subfigure.sty:
-% \let\MYorigsubfigure\subfigure
-% \renewcommand{\subfigure}[2][\relax]{\MYorigsubfigure[]{#2}}
-% However, the above trick will not work if both optional arguments of
-% the \subfloat/subfig command are used. Furthermore, there needs to be a
-% description of each subfigure *somewhere* and endfloat does not add
-% subfigure captions to its list of figures. Thus, the best approach is to
-% avoid the use of subfigure captions (many IEEE journals avoid them anyway)
-% and instead reference/explain all the subfigures within the main caption.
-% The latest version of endfloat.sty and its documentation can obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/endfloat/
-%
-% The IEEEtran \ifCLASSOPTIONcaptionsoff conditional can also be used
-% later in the document, say, to conditionally put the References on a 
-% page by themselves.
-
-% *** PDF, URL AND HYPERLINK PACKAGES ***
-%
-\usepackage{url}
-% url.sty was written by Donald Arseneau. It provides better support for
-% handling and breaking URLs. url.sty is already installed on most LaTeX
-% systems. The latest version can be obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/misc/
-% Read the url.sty source comments for usage information. Basically,
-% \url{my_url_here}.
-
-% *** Do not adjust lengths that control margins, column widths, etc. ***
-% *** Do not use packages that alter fonts (such as pslatex).         ***
-% There should be no need to do such things with IEEEtran.cls V1.6 and later.
-% (Unless specifically asked to do so by the journal or conference you plan
-% to submit to, of course. )
-
-% correct bad hyphenation here
-\hyphenation{op-tical net-works semi-conduc-tor}
-
-%%% My packages
-\usepackage{amsmath,epsfig}
-\usepackage{amsfonts}
-\usepackage{color}
-
-%%% My commands
-\newcommand{\mat}[1]{\mathbf{#1}}
-\renewcommand{\vec}[1]{\mathbf{#1}}
-\newcommand{\degree}{\ensuremath{^\circ}}
-\newcommand{\R}{$\mat{\hat{R}}$ }
-
-%\newcommand{\Red}[1]{{\color{red}#1}}
-\newcommand\multimedia[1]{\textbf{{\color{red}#1}}}
-
-\begin{document}
-%
-% paper title
-% can use linebreaks \\ within to get better formatting as desired
-\title{Implementing Capon Beamforming on a GPU for Real-Time Cardiac Ultrasound Imaging}
-%
-%
-% author names and IEEE memberships
-% note positions of commas and nonbreaking spaces ( ~ ) LaTeX will not break
-% a structure at a ~ so this keeps an author's name from being broken across
-% two lines.
-% use \thanks{} to gain access to the first footnote area
-% a separate \thanks must be used for each paragraph as LaTeX2e's \thanks
-% was not built to handle multiple paragraphs
-%
-
-\author{Jon Petter \AA{}sen, \IEEEmembership{Student Member,~IEEE,} Jo Inge Buskenes, Carl-Inge Colombo Nilsen, \IEEEmembership{Member,~IEEE,} Andreas Austeng, \IEEEmembership{Member,~IEEE,} and Sverre Holm, \IEEEmembership{Senior Member,~IEEE,} %<-this % stops a space
-\thanks{Manuscript received ...; accepted ...}
-\thanks{J. P. \AA{}sen and S. Holm is with the Medical Imaging Lab (MI-Lab) at the Norwegian University of Science and Technology, Trondheim, Norway.}
-\thanks{J. I. Buskenes, C.-I. C. Nilsen, A. Austeng and S. Holm is with the Department of Informatics, University of Oslo, Oslo, Norway. (e-mail: jon.p.asen@ntnu.no)} 
-}
-
-% note the % following the last \IEEEmembership and also \thanks - 
-% these prevent an unwanted space from occurring between the last author name
-% and the end of the author line. i.e., if you had this:
-% 
-% \author{....lastname \thanks{...} \thanks{...} }
-%                     ^------------^------------^----Do not want these spaces!
-%
-% a space would be appended to the last name and could cause every name on that
-% line to be shifted left slightly. This is one of those "LaTeX things". For
-% instance, "\textbf{A} \textbf{B}" will typeset as "A B" not "AB". To get
-% "AB" then you have to do: "\textbf{A}\textbf{B}"
-% \thanks is no different in this regard, so shield the last } of each \thanks
-% that ends a line with a % and do not let a space in before the next \thanks.
-% Spaces after \IEEEmembership other than the last one are OK (and needed) as
-% you are supposed to have spaces between the names. For what it is worth,
-% this is a minor point as most people would not even notice if the said evil
-% space somehow managed to creep in.
-
-% The paper headers
-\markboth{Transactions on Ultrasonics and Ferroelectrics and Frequency Control,~Vol.~x, No.~y, Z~201\ae{}}%
-{\AA{}sen \MakeLowercase{\textit{et al.}}: Capon Adaptive Beamforming on the GPU}
-% The only time the second header will appear is for the odd numbered pages
-% after the title page when using the twoside option.
-% 
-% *** Note that you probably will NOT want to include the author's ***
-% *** name in the headers of peer review papers.                   ***
-% You can use \ifCLASSOPTIONpeerreview for conditional compilation here if
-% you desire.
-
-% If you want to put a publisher's ID mark on the page you can do it like
-% this:
-%\IEEEpubid{0000--0000/00\$00.00~\copyright~2007 IEEE}
-% Remember, if you use this you must call \IEEEpubidadjcol in the second
-% column for its text to clear the IEEEpubid mark.
-
-% use for special paper notices
-%\IEEEspecialpapernotice{(Invited Paper)}
-
-% make the title area
-\maketitle
-
-\begin{abstract}
-%\boldmath
-Capon beamforming is associated with a high computational complexity, which limits its use as a real-time method in many applications.
-In this paper we present an implementation of the Capon beamformer that exhibits real-time performance when applied in a typical cardiac ultrasound imaging setting.
-To achieve this performance we make use of the parallel processing power found in modern Graphics Processing Units (GPUs), combined with beamspace processing to reduce the computational complexity as the number of array elements increases.
-
-For a three dimensional beamspace we show that processing rates supporting real-time cardiac ultrasound imaging are possible, meaning that images can be processed faster than the image acquisition rate for a wide range of parameters.
-Image quality is investigated in videos of both a simulated phantom, and in an \textit{in-vivo} cardiac dataset.
-These results show that Capon beamforming is feasible for cardiac ultrasound imaging, providing images with improved lateral resolution and no local shift-variance effects, both in element-space and beamspace.
-\end{abstract}
-% IEEEtran.cls defaults to using nonbold math in the Abstract.
-% This preserves the distinction between vectors and scalars. However,
-% if the journal you are submitting to favors bold math in the abstract,
-% then you can use LaTeX's standard command \boldmath at the very start
-% of the abstract to achieve this. Many IEEE journals frown on math
-% in the abstract anyway.
-
-% Note that keywords are not normally used for peerreview papers.
-\begin{IEEEkeywords}
-Adaptive beamforming, Capon beamforming, Covariance matrix, GPGPU, Graphics processing unit, Minimum variance beamforming, Parallel programming, Ultrasound imaging.
-\end{IEEEkeywords}
-
-% For peer review papers, you can put extra information on the cover
-% page as needed:
-% \ifCLASSOPTIONpeerreview
-% \begin{center} \bfseries EDICS Category: 3-BBND \end{center}
-% \fi
-%
-% For peerreview papers, this IEEEtran command inserts a page break and
-% creates the second title. It will be ignored for other modes.
-\IEEEpeerreviewmaketitle
-
-\section{Introduction}
-% The very first letter is a 2 line initial drop letter followed
-% by the rest of the first word in caps.
-% 
-% form to use if the first word consists of a single letter:
-% \IEEEPARstart{A}{demo} file is ....
-% 
-% form to use if you need the single drop letter followed by
-% normal text (unknown if ever used by IEEE):
-% \IEEEPARstart{A}{}demo file is ....
-% 
-% Some journals put the first two words in caps:
-% \IEEEPARstart{T}{his demo} file is ....
-% 
-% Here we have the typical use of a "T" for an initial drop letter
-% and "HIS" in caps to complete the first word.
-
-%\IEEEPARstart{I}{ntroduction} to Capon adaptive beamforming, ultrasound imaging and GPU/CUDA.
-\IEEEPARstart{C}{apon} beamforming \cite{Capon1969} has recently been applied to medical ultrasound imaging with promising results in several studies \cite{Synnevag2007, Austeng2008, Vignon2008, Viola, Mehdizadeh2012}.
-Most of this work focuses on the improved lateral resolution and contrast obtained with the Capon beamformer.
-However, some interesting trade-offs have also been introduced.
-In \cite{Synnevag2009}, Synnev\aa{}g \textit{et al.} explains how image resolution can be maintained at higher framerates, with smaller probes, or for deeper penetration.
-Despite these benefits, widespread adoption of the Capon beamformer has not been seen due to its high computational complexity \cite{So2011}.
-In this paper we present an implementation of the Capon beamformer exhibiting real-time performance when applied in a typical cardiac ultrasound imaging setting.
-To achieve this performance we make use of the parallel processing power found in modern Graphics Processing Units (GPUs) (found in almost all PCs), combined with novel methods to reduce the computational complexity as the number of array elements increases.
-
-Previous work on parallel implementations of the Capon beamformer have been focused on radar and passive sonar systems using custom made systolic array processors \cite{McWhirter1989, Moonen1993, Sinha2002}.
-For medical ultrasound, Chen \textit{et al.} \cite{Chen2011a, Chen2011, Chen} implemented a Capon beamformer on the GPU that produces real apodization weights based on real data.
-However, the Capon beamformer is meant to process complex data.
-Under the constraint of using real data, the Capon beamformer is not allowed to micro-steer the main lobe, and therefore its full potential is not reached.
-Micro-steering of the main lobe is an important property of the Capon beamformer in order for it to improve structural details and edge definitions in an ultrasound image \cite[Fig~9.]{Synnevag2009}.
-Complex arithmetic leads to more operations per sample, but also allows the sampling frequency to be reduced.
-The number of samples needed per image can therefore be minimized.
-All results in this paper are reported with the use of baseband (or I/Q) data, where the sampling frequency is near the system bandwidth.
-
-%Using real data will results in symmetric windows, and even if the number of samples they managed to process per second was high, both the number of floating point operations per sample and the imaging region is reduced using real data compared with use of complex data.
-It is crucial to use complex data within the Capon beamformer framework, both for reducing the amount of data per image, and to allow for asymmetric windows.
-All results in this paper is reported with use of critically sampled complex data, both laterally and in range.
-(ikke helt kritisk samplet, ikke si det, bare at det er i baseband).
-
-% denne må utelates i IUS-artikkelen
-In recent work \cite{asen2012, Buskenes, Buskenes2013} we have applied a GPU-based Capon beamformer for both active sonar imaging and for cardiac ultrasound imaging.
-Sonar has a less strict real-time requirement than medical ultrasound imaging, which means that most array configurations can be handled in real time using the GPU.
-Capon beamforming in medical ultrasound imaging, and in particular cardiac ultrasound imaging, demands frame rates that not even a high-end GPU can provide.
-Since the execution time is cubic with respect to the number of array elements, we need methods that reduce the computational complexity for large arrays.
-Various approaches to low-complexity Capon beamforming for medical ultrasound imaging have therefore been proposed in the literature \cite{Synnevag2011, Asl2012, Jensen2012, Kim}.
-In this paper we will make use of a beamspace transformation described by Nilsen and Hafizovic \cite{Nilsen2009}, in order to maintain high framerates as the number of elements increases.
-%The beamspace transformation matches well with the focused transmit beams used in cardiac ultrasound imaging. 
-%(Referer til JOE-artikkelen, få med beskrivelse av hva som blir anerledes for ultralyd.)
-
-In the next section we present background information on Capon beamforming in both element-space and beamspace and the GPU computing model.
-In Section \ref{sec:meth} we recap our parallel implementation of the Capon beamformer, before the GPU implementation of the beamspace Capon beamformer is presented in Section \ref{sec:bs}.
-Benchmark results and resulting images are presented in Section \ref{sec:bench} and \ref{sec:images}, together with a discussion of trading image quality for speed in Section \ref{sec:trade}.
-Finally we discuss the results in Section \ref{sec:dis}, and draw conclusions in Section \ref{sec:con}.
-
-\section{Background}\label{background}
-
-\subsection{Capon Beamforming}\label{sec:es-capon}
-
-%\textbf{Bruk riktig terminologi. $N_{avg}$ er K. Antall sub-arrayer må hete noe annet. F.eks $N_L$}
-
-%\textbf{Presis på hvilke formler vi skal implementere.}
-
-%\textbf{De skal forstå MV etter denne seksjonen.}
-
-%\textbf{Ton ned introduksjonen, og sett opp akkurat de ligningene vi implementerer.}
-
-%\textbf{(Skriv om den introduserende setningen, start rett på linje to, Kombinere data fra mange sensorer, Leseren kan Ultralyd ikke gpu)}
-
-The conventional form of array beamforming in medical ultrasound imaging is delay-and-sum (DAS).
-Here, the signal recorded at element number $m \in [0,M-1]$ at time $n$ is delayed with an appropriate delay $\Delta_m[n]$ to focus and steer the ultrasound beam.
-This delayed data is here denoted $x_m[n]$.
-The beamformer output $z[n]$ is then the sum of all $M$ elements weighted with a factor $w_m$,
-\begin{align}
-z[n] = \sum_{m = 0}^{M-1}w_m^*x_m[n] = \vec{w}^H\vec{x}[n], \label{eq:z}
-\end{align}
-where $(\cdot)^H$ is the Hermitian transpose.
-The weight vector, or window, $\vec{w}$ is usually real in DAS beamforming, and is applied to trade resolution for a better signal-to-noise ratio (SNR).
-Although different windows can be used in different areas of the images, $\vec{w}$ is usually predefined.
-
-As shown in Fig.
-\ref{fig:mvbf}, the basic idea of an adaptive beamformer is to form a complex weight vector based on the received data.
-The Capon beamformer produces weights that minimize the beamformer output power while maintaining unity gain in the focus direction.
-The result is that interference impinging from other directions is suppressed \cite{Synnevag2007}.
-The complex weights cause a micro-steering which is minute compared to the initial steering and focusing step, therefore the beamformer also works well for broadband ultrasound signals.
-
-%Formally stated, the adaptive weights are found by solving the following minimization problem:
-%\begin{align}
-%&\min_{\vec{w}} E\{|z[n]|^2\} = \min_{\vec{w}}\vec{w}^H\mat{R}\vec{w}\label{eq:minz}\\
-%&\text{subjected to } \vec{w}^H\vec{a} = 1,\label{eq:constraint}
-%\end{align}
-%where $\mat{R} = E\{\vec{x}[n]\vec{x}[n]^H\}$ is the spatial covariance matrix, and $\vec{a} = \left[ \begin{array}{cccc} e^{-jk_{\tilde{x}}\tilde{x}_0}& e^{-jk_{\tilde{x}}\tilde{x}_1}& ...& e^{-jk_{\tilde{x}}\tilde{x}_{M-1}}\end{array}\right] ^T$ is a far-field steering vector towards angle $\theta = \arcsin(\frac{\lambda k_{\tilde{x}}}{2\pi})$.
-%The constraint in (\ref{eq:constraint}) is there to avoid the trivial solution $\vec{w} = 0$, and a distorted respons in the direction of $\vec{a}$. 
-%The solution to the minimization problem in (\ref{eq:minz}) and (\ref{eq:constraint}) is
-
-\begin{figure}
-\centerline{\includegraphics[width=3in]{gfx/beamforming_mv.eps}}
-\caption{Capon beamforming.
-Impinging signal is demodulated, and aligned in phase by applying steering and focusing delays $\Delta_m[n]$.
-Adaptive weights, $\vec{w}[n]$, are then calculated based on in-phase data, and finally the output is formed by a Capon-weighted coherent sum.}
-\label{fig:mvbf}
-\end{figure}
-
-The Capon beamformer, following the approach in \cite{Synnevag2009}, can be divided into three main steps: estimation of a sample covariance matrix, solving a system of linear equations, and calculation of the beamformer output.
-The sample covariance matrix is estimated as
-\begin{align}
-\mat{\breve{R}}[n] = \frac{1}{N_LN_K}\sum_{n'=n-K}^{n+K} \sum_{l=0}^{N_L-1} \vec{x}_l[n']\vec{x}_l[n']^H,\label{eq:R}
-\end{align}
-where  $N_K = 2K + 1$ is the number of temporal samples included, $\vec{x}_l = [x_l[n], \dotso, x_{l+L-1}[n]]$ is the $l\text{th}$ subarray of length $L$, and $N_L = M-L+1$ is the number of subarrays.
-Note that the covariance matrix dimension is $L \times L$.
-The covariance matrix is loaded with a diagonal factor $\epsilon$ to ensure numerical stability and increased robustness:
-\begin{align}\label{eq:diag}
-\mat{\hat{R}}[n] = \mat{\breve{R}}[n] + \epsilon[n]\mat{I} = f(\vec{x}[n]).
-\end{align}
-The operator $f(\cdot)$ represents the process of estimating $\mat{\hat{R}}$ from the input $\vec{x}$.
-A factor proportional to the output power is often used to reduce the need for parameter adjustments, and the trace of $\mat{\breve{R}}$, 
-\begin{align}\label{eq:diag_adapt}
-\epsilon[n] &= d \times \frac{\text{trace}\{\mat{\breve{R}}[n]\}}{L},
-\end{align}
-has been applied in much of the recent literature on Capon beamforming for medical ultrasound imaging \cite{Synnevag2007, Nilsen2009, Wang2009, Mehdizadeh2012}.
-It has also been applied to HF and VHF antenna processing \cite{Featherstone1997b}.
-
-Next, a linear system of equations has to be solved,
-\begin{align}\label{eq:b}
-\vec{b}[n] = \mat{\hat{R}}[n]^{-1}\vec{a} \in \mathbb{C}^L,
-\end{align}
-where the steering vector $\vec{a} = \vec{1}_L = [1_0, 1_1, ..., 1_{L-1}]^T$ since $\vec{x}$ is pre-delayed.
-
-Finally the Capon weight vector is formed from (\ref{eq:b}) as
-\begin{align}\label{eq:w}
-\vec{w}[n] = \frac{\vec{b}[n]}{\vec{1}^H\vec{b}[n]} \in \mathbb{C}^L,
-\end{align}
-and the beamformer output is calculated by combining all $N_L$ subarrays weighted with the same set of adaptive weights. This is formally stated as
-\begin{align}
-z[n] &= \frac{1}{N_L}\vec{w}[n]^H \sum_{l=0}^{N_L-1} \vec{x}_l[n] \label{eq:z_mv}\\
-&= 1/N_L \, (\vec{w}[n] * \vec{1}_{N_L})^H\vec{x}[n] \label{eq:z_mv2},
-\end{align}
-%where (this matrix is wrong)
-%\begin{align}
-%\mat{C} = \left(
-%\begin{matrix}
-%\vec{1}_1 & \vec{1}_2 & \cdots & \vec{1}_{N_L} & \vec{0}_1 & \cdots & \vec{0}_{N_L-1} & \vec{0}_{N_L}\\
-%\vec{0}_{N_L-1} & \vec{0}_{N_L-2} & \cdots & \vec{0}_{0} & \vec{1}_{N_L-1} & \cdots & \vec{1}_{2} & \vec{1}_{1}\\
-%\end{matrix}
-%\right)^T
-%\end{align}
-where $*$ is a non-truncated convolution across elements.
-Note how we can either choose to sum all subarrays (\ref{eq:z_mv}), or accumulate a weighting per element (\ref{eq:z_mv2}).
-
-In passive systems, assuming stationary statistics, the sample covariance matrix in (\ref{eq:R}) is typically averaged over a long sequence of samples\cite{Krima}.
-Since medical ultrasound imaging is an active, short-pulsed system, there is a limited amount of data from which $\mat{\hat{R}}$ can be estimated.
-The estimated matrix is therefore often found to be both inaccurate and poorly conditioned.
-Another challenge with active systems is the high correlation between signal and interference, resulting in signal cancellation \cite{Reddy1987}.
-Thus, the beamformer output can be lower than expected, even when there is signal present and the distortionless-response criterion is applied.
-
-In order to get a well-conditioned $\mat{\hat{R}}$, avoid signal cancellation, and to get DAS-like speckle, $\mat{\hat{R}}$ has to be averaged over $L\le M/2$ long subarrays and $N_K \sim T_p/T_s$ time samples \cite{Synnevag2007a}, where, $T_p$ is the pulse length in seconds and $T_s$ is the sampling period.
-For our baseband data, one pulse length ($1.5\lambda$) is around 3 samples ($K=1$).
-Typical values for $d$ is between $1/10$ and $1/100$.
-
-% Få med at vi processerer hele bildet på en gang. Forklar forskjellen med å prosessere en beam av gangen. Hva med MLA?
-% Illustrer tråder og tider for hvordan processeringen går.
-% Vi snakker om mottak-stråleformeren.
-
-%\begin{figure}
-%\includegraphics[width=3in]{gfx/beamforming_mv_lowres.png}
-%\caption{}
-%\label{fig:mv}
-%\end{figure}
-
-%Building and calculating the inverse of $\hat{R}$ is computational demanding, limiting the methods accessibility. Following the innovation in GPU computing, this has now started to change. 
-%In this paper we introduce the first GPU implementation of the Capon beamformer capable of processing a 70 degree sector cardiac image from a 64 element phased array at interactive frame rates using both spatial and temporal smoothing.
-Both types of smoothing are important for maintaining delay-and-sum-like speckle and at the same time keep the high resolution property.
-This is achieved using complex base band data, which reduces the size of each range line to a minimum.
-In addition, complex data results in complex weights; hence this is a correct implementation of Capon beamforming with all degrees of freedom preserved.
-For high channel count systems we propose two ways of reducing computational complexity while maintaining
-
-\subsection{Beamspace Processing}\label{bs-capon}
-Nilsen and Hafizovic \cite{Nilsen2009} have proposed to apply beamspace processing to reduce the required computations of the Capon beamformer when applied to medical ultrasound imaging.
-Beamspace is an alternative representation of the per-element data, transformed into a fan of beams covering the sector illuminated by transmit beams.
-The transformation is formally stated as
-\begin{align}
-\vec{x}_\text{BS}[n] &= \mat{B}\vec{x}[n], \ \ \ \ \ \ \ \ \ \ \ \ \ \, \mat{B} \in \mathbb{C}^{N,M} \label{eq:butler} \\ 
-[\mat{B}]_{p,q} &= \frac{1}{\sqrt{M}}e^{-j 2 \pi p q / M} \ \ \ N \le M.
-\end{align} 
-The steering vectors in $\mat{B}$, the so-called Butler or normalized discrete fourier transform (DFT) matrix, determine each beam's direction in space.
-Beams where no interference is present can therefore be removed by simply removing rows from $\mat{B}$.
-
-For a focused system like cardiac ultrasound imaging, where the received signal is concentrated in a narrow sector around the transmit direction, a small percentage of the beams contains almost all the received energy.
-Nilsen and Hafizovic concluded that as little as three beams around (and including) the transmit direction were adequate to produce results comparable to applying Capon beamforming in element space.
-Hence, for equally pitched arrays the inversion steps is found to be constant with respect to the number of array elements.
-
-The beamspace version of the Capon beamformer is derived by replacing the covariance matrix in (\ref{eq:w}) with the beamspace covariance matrix 
-\begin{align}
-\mat{\hat{R}}_\text{BS} = f(\vec{x}_\text{BS}) = f(\mat{B}\vec{x}), 
-\end{align}
-where $\mat{B} \in \mathbb{C}^{N_b,L}$, $N_b$ is the number of selected beams, and $f(\cdot)$ represent the covariance estimation process as defined in (\ref{eq:diag}).
-
-In the remaining sections, the methods described in this section and Section \ref{sec:es-capon} will be referred to as beamspace Capon (BS-Capon) and element-space Capon (ES-Capon) respectively.
-The name Capon will refer to both methods.
-
-\subsection{GPU Compute Model}
-
-%\textbf{(Ta med mer banale ting, dette er GPU for ultralydfolk)}
-
-%\textbf{(Etter denne så må de forstå GPU og hvorfor MV passer.)}
-
-In this section we give a brief discussion of GPU computing, laying the basis for our discussions on parallel implementation strategies.
-The Capon beamformers have been implemented using the CUDA GPU framework by Nvidia \cite{Nvidia2013}, however the challenges and proposed methods described in this paper are valid for other GPU framworks as well, like e.g.
-OpenCL.
-Independent of framework, the key point is that GPUs are designed to execute large sets of small concurrent tasks, like beamforming a pixel in a large image, much faster than a CPU.
-While the CPU has few cores, but logic to process complex threads, the GPU has many cores, but each core needs many (preferably simple) threads to keep it occupied and by that achieve maximum throughput.
-
-In CUDA programming a kernel function is executed over a large set of threads, where the kernel function describes the work to be done in each thread.
-As depicted in Fig.
-\ref{fig:gpulayout}, threads are organized in a grid of thread-blocks, with maximal size of 1024 threads per block (e.g.
-$32\times32$).
-Threads inside a block support low-cost synchronization if needed.
-From a CUDA perspective the GPU consist of one or more streaming multi-processors (SM), where each SM is capable of 32 or more concurrent multiplications or additions, or both, depending on the architecture.
-A group of blocks is scheduled to each SM until all blocks have been processed.
-
-An SM has a limited amount of shared memory (or near-core cache) and registers.
-Therefore, to achieve maximum GPU occupancy (defined as having the maximum number of threads residing on each SM), a conservative use of registers and shared memory per thread is required.
-Each thread should perform enough instructions per transferred byte in order to hide memory latency, and the problem should be divided into enough threads in order to hide instruction latency.
-Global GPU memory transactions, in addition to CPU-GPU transfers, are slow and must be minimized.
-All of these constraints combined pose a real challenge as the target problem has to map well to the GPU framework on both a grid, block, thread, and memory consumption level in order to benefit from parallel acceleration.
-In the next two sections, we will describe how this mapping can be done for the Capon beamformers.
-
-\begin{figure}
-\centerline{\includegraphics[width=3.5in]{gfx/gpu_layout_vertical_2.eps}}
-\caption{Calculation of Capon adaptive weights mapped to GPU architecture.
-A recorded echo across the array for a given range and angle (selected cell) is assigned to a group of $L$ threads that can run independently of all other groups.
-Note that several samples can be processed per thread-block if $L$ is small.
-The task of these $L$ threads is to calculate the Capon beamformer output in the three depicted steps.}
-\label{fig:gpulayout}
-\end{figure}
-
-%When implementing algorithms on the GPU it is important to keep data as close to the core as possible. Utilizing available registers and shared memory to its maximum is therefore important.
-
-%More important, the CUDA 2.x compute capability (CC) has available 48 KB of near-core shared memory per SM that can be utilized by resident blocks.
-The maximum number of resident blocks per SM is 8, but in addition the maximum number of resident warps per SM is restricted to 48.
-Hence, with a block size of 512 only $48/(512/32) = 3$ blocks can occupy the SM at once.
-In addition, if one thread where to use 64B of shared memory the number of allowed resident blocks with a block size of 512 would be reduced to 1.
-Up to a curtain point it is beneficial to have as many warps as possible occupying the SM at once to hide latency from accessing memory.
-The SM has for CC 2.x 32K 32-bit long registers.
-When implementing algorithms on the GPU it is important to keep data as close to the core as possible.
-Utilizing available registers and shared memory to its maximum is therefore important.
-
-%\subsection{Optimizing kernel launch}
-%Set the size of grid adaptively based on $L$, $M$ and $Yavg$.
-%It is better to slide in space than time. $K$ is typically much larger than $2*Yavg + 1$.
-
-% needed in second column of first page if using \IEEEpubid
-%\IEEEpubidadjcol
-
-\section{Parallel ES-Capon}\label{sec:meth}
-In this section we will recap our parallel implementation of the ES-Capon beamformer as presented in \cite{asen2012, Buskenes}.
-This implementation was our starting point when the BS-Capon beamformer was implemented.
-
-If the beamformer in (\ref{eq:z_mv}) is analyzed in combination with the matrix equation in (\ref{eq:b}) and the covariance estimation process in (\ref{eq:diag}), we see that the calculation of a set of Capon weights is independent across the outputs $z[n]$.
-As shown in Fig.
-\ref{fig:gpulayout}, a first level of granularity is therefore to divide the image into a grid, where the output in each cell can be calculated independently.
-Further, based on memory dependencies and to hold memory consumption per thread at a minimum, a group of threads should process one output in the final image.
-Continuing with the calculation of one weight vector, the algorithm can be broken down into three main steps that, to some extent, can be divided further into a set of parallel processes.
-That is, for each data vector $\vec{x}$ in the image:
-\begin{align*}
-%\begin{array}{l}
-\text{1) } &\text{Calculate the sample covariance matrix } \mat{\hat{R}} \text{ as in } \text{(\ref{eq:diag})}.\\
-\text{2) } &\text{Solve the linear system of equations as in (\ref{eq:b}}).\\ %\vec{b} =\mat{\hat{R}}^{-1}\vec{1}_L\\
-\text{3) } &\text{Calculate } \vec{w} = \vec{b}/\vec{1}_{L}^H\vec{b} \text{ and the amplitude as in (\ref{eq:z_mv})}.
-%\end{array}
-\end{align*}
-%\begin{enumerate}
-%\item Calculate the sample covariance matrix $\mat{\hat{R}}$ as in (\ref{eq:R})
-%\end{enumerate}
-%\begin{enumerate}
-%\item[2)] Solve the linear system of equations $\mat{\hat{R}}\vec{b} =\vec{ 1}$
-%\end{enumerate}
-%\begin{enumerate}
-%\item[3)] Calculate $\vec{w} = \vec{b}/\vec{1}^H\vec{b}$ and the amplitude $\vec{w}^H\vec{x}$
-%\end{enumerate}
-The computational complexities of these three steps are $O(N_KN_LL^2)$, $O(L^3)$, and $O(L)$, respectively.
-As $M$ (and thereby $L$) increases, the first two steps quickly become unmanageable, and real-time performance in cardiac ultrasound imaging can only be achieved by exploring methods with reduced complexity.
-
-%(The three steps have, for modularity, been implemented using three different kernels. Joining all kernels into one could possible lead to an increase in speed.)
-
-%\begin{figure}
-%\includegraphics[width=3in]{gfx/gpu_layout_lowres.png}%
-%\caption{}
-%\label{fig:gpulayout}
-%\end{figure}
-
-\subsection{Calculation of Multiple Sample Covariance Matrices}\label{sec:calcR}
-
-%The straightforward sum in (\ref{eq:R}) has complexity $O(L^2N_LN_K) = O(L^3)$ since $N_L$, the number of subarrays, is a function of $L$.
-This is actually the same as for inverting an $L \times L$ matrix.
-However little attention has been devoted to this step in the literature before.
-Note that for a typical array configuration we do more spacial smoothing than temporal smoothing.
-Hence, $N_K$ is usually small compared to $N_L$.
-The construction of $\mat{\hat{R}}$ is dominated by the number of elements $M$ ($N_L = M - L + 1$) for short subarrays, but since $L$ is selected in most cases to be between $M/2$ and $M/4$ a large $M$ implies a large $L$.
-
-Calculation of one element of \R is from (\ref{eq:R}) found to be independent of calculation of the other elements.
-A natural level of granularity would then be to assign one thread to each element of $\mat{\hat{R}}$ \cite{Chen2011}.
-This will minimize global reads and writes needed per thread, but since we want to avoid block-to-block synchronization, the maximum supported subarray length is limited to 32 by this approach.
-Since most medical ultrasound arrays have 64 elements or more, an implementation not limited by $L_{max}=M/2=32$ is preferable.
-
-In \cite{asen2012, Buskenes} we demonstrated how the cubic complexity with $L$ and linear complexity with $N_K$ can be reduced by realizing that calculations in (\ref{eq:R}) overlap across subarrays.
-Element $(i,j)$ of $\mat{\breve{R}}[n]$ can be calculated from element $(i-1, j-1), \text{for } i,j \in [1, L-1]$ in a sliding window maner as
-\begin{align}
-\mat{\breve{R}}_{i,j}[n] &=  \mat{\breve{R}}_{i-1,j-1}[n]  \nonumber \\
-&+ \sum_{n'=n-K}^{n+K} (x_{i+N_L-1}[n']x_{j+N_L-1}[n']^* \nonumber \\
-&- x_{i-1}[n']x_{j-1}[n']^H). \label{eq:sliding}
-\end{align}
-%Along the time dimension we also find that
-%\begin{align}
-%\mat{\breve{R}}[n] &= \mat{\breve{R}}[n-1] + \sum_{l=0}^{N_L-1} (\vec{x}_l[n+K]\vec{x}_l[n+K]^H \nonumber \\
-%&- \vec{x}[n-K-1]\vec{x}[n-K-1]^H).
-%\end{align}
-%- \mat{\breve{R}}[n-K-1] + \mat{\breve{R}}[n-K].
-Sliding windows across subarrays and in time can be realized simultaneously if $\mat{\hat{R}}$, averaged over subarrays, is calculated from an $N_K$-average in time of the outer products $\vec{x}[n]\vec{x}[n]^H$.
-However, since $N_K$ typically is small, and the GPU has a limited amount of shared memory available per thread, we focus on subarrays only.
-In that case, the complexity of calculating $\mat{\hat{R}}$ is reduced to $O(N_K(L(N_L-2)+2L^2))$.
-For $L=M/2$ this reduces to $O(N_KM^2)$.
-
-The sliding-window approach will cause dependence among the iterative calculations.
-The level of task granularity is therefore increased.
-Calculations are still independent across time, but not along diagonals of $\mat{\hat{R}}$.
-One thread is therefore assigned to each diagonal in the upper triangle of $\mat{\hat{R}}$.
-For small values of $L$, several covariance matrices are calculated per thread block (Fig.
-\ref{fig:gpulayout}).
-Maximum supported $L$ is now increased to the maximum number of threads per block (currently 1024).
-However, for small-$L$-high-$M$ configurations, memory consumption per thread will be too high to achieve full GPU occupancy.
-
-Taking symmetry into account, only the upper half of $\mat{\hat{R}}$ needs to be calculated.
-However, the GPU computes $\mat{\hat{R}}$ in such a way that the lower triangle can be found with little overhead \cite{Buskenes}.
-This also means that no special solver is needed in order to solve $\mat{\hat{R}}\vec{b} = \vec{1}_L$, even if a symmetric solver could have saved some instructions.
-
-For the adaptive diagonal loading factor in (\ref{eq:diag}), the trace of $\mat{\hat{R}}$ has to be known before diagonal elements are written back to global memory.
-Since one row for all covariance matrices is written to global memory per kernel iteration, the diagonal elements need to be held in shared memory until the trace is accumulated.
-When that is done, scattered writes of the scaled diagonal elements to global memory has to be performed.
-
-%For subarray averaging, the first row and column of $\mat{\hat{R}}$ has to be fully calculated, and for time averaging $\mat{\hat{R}}[0]$ has to be fully calculated.
-
-%Since $L$ usually is larger than $N$, we have
-
-%Building $\hat{\mat{R}}$ using a sliding window approach across $K$ and $N_{avg}$.
-We have a limited amount of fast, near-core memory on the GPU.
-On NVIDIA architecture this memory is known as shared memory, and is restricted to 48KB per compute block.
-The maximum block size of 32x32, further restrict how large arrays we can handle inside a single compute block.
-It is therefore natural to restrict $L$ to a maxima of $32$ elements.
-Since $L \le M/2$, $M \le 64$.
-We can not afford to hold the full $\hat{\mat{R}}$ ($M = L$) when $\hat{\mat{R}} > 32$, because of limited amount of memory.
-...
-Explain in detail how \texttt{buildR} is implemented in a kernel.
-...
-Give more details on how much shared memory we have, and how it can be divided between compute blocks occupying one stream multiprocessor (SM).
-
-%Discuss the bottleneck adaptive diagonal loading causes. Possible solutions includes, use trace from previous image, use pre-calculated array power, constant weight.
-
-%Time averaging takes a lot of resources.
-It has been shown that time averaging can maintain both resolution and delay-and-sum-like speckle.
-The same speckle statistics can be obtain with small subarrays, but then we loose resolution.
-Small sub-arrays benefits from being less computational demanding.
-
-%\subsubsection{CUDA Compute Model}
-%The CUDA compute model is based on execution of a kernel across a grid of compute threads.
-Each position in the grid holds a block of threads.
-Each block is further divided into groups of 32 threads, known as a warp.
-The maximum number of threads inside a block is restricted to 512, hence 16 warps.
-From a CUDA perspective the GPU consist of one or more stream multi processors (SM), %where each SM is capable of 32 or more, depending on the architecture, concurrent multiplications and/or additions.
-The CUDA 2.x compute capability (CC) has available 48 KB of near-core shared memory per SM that can be utilized by resident blocks.
-The maximum number of resident blocks per SM is 8, but in addition the maximum number of resident warps per SM is restricted to 48.
-Hence, %with a block size of 512 only $48/(512/32) = 3$ blocks can occupy the SM at once.
-In addition, if one thread where to use 64B of shared memory the number of allowed resident blocks with a block size of 512 would be reduced to 1.
-Up to a curtain point
-%it is beneficial to have as many warps as possible occupying the SM at once to hide latency from accessing memory and executing time consuming mathematical functions.
-The SM has for CC 2.x 32K 32-bit long registers.
-When implementing algorithms on the GPU it is important to keep data as close to the core as possible.
-Utilizing available registers and shared memory to its maximum is %therefore important.
-
-\subsection{Solving Multiple Small Linear Systems}
-The literature has mostly focused on large systems when solving systems of linear equations on the GPU.
-GPU libraries therefore often lack a comprehensive collection of batched solvers for large sets of small systems.
-The reason for this is two-fold.
-First the GPU needs to solve thousands of small systems in order to beat the low memory latency of the CPU.
-Second there is a range of system dimensions that does not fit well with the GPU architecture, where only a small amount of shared memory and registers are available.
-Because of this, it has been proven hard to optimize a manageable set of solvers that provides speedups compared to the CPU in every case.
-It is therefore not given that solving on the GPU is faster than solving on the CPU in general.
-However, there are two arguments for solving on the GPU.
-First the CPU, which might already run a lot of ultrasound-related algorithms, is offloaded.
-Second, the transfer of several thousand covariance matrices over the PCI-Express bus, from the CPU to the GPU, is avoided.
-
-For the results presented in this paper, we have used a GPU implementation of batched Gauss-Jordan elimination made by Nvidia to solve $\mat{\hat{R}}\vec{b} = \vec{1}_L$, where $\vec{b} = \mat{\hat{R}}^{-1}\vec{1}_L$, for all pixels in an image in one kernel launch.
-The solver is available online for registered CUDA developers \cite{Nvidiaa}.%For smaller systems, $L < 10$, a direct solver using one thread per system is preferable.
-For this purpose we have implemented Cramer's rule for $2 \times 2$, $3 \times 3$ and $4 \times 4$ matrices.
-All solvers have been benchmarked against the latest Intel MKL libraries.
-
-%However we will go through an implementation of an $\mat{U}^H\mat{D}\mat{U}$-decomposition-based solver to point out challenges when implementing batched solving of small linear systems on the GPU.
-...
-(Unsure if this should be included).
-
-%Can include details on GPU implementation of $\mat{U}^H\mat{D}\mat{U}$, but this has not proved to be faster than NVIDIA's GJ implementation.
-However, the complexity for solving with $\mat{U}^H\mat{D}\mat{U}$ decomposition is supposed to be $1/2$ of GJ.
-
-%Skriv om utfordringene om å lage en solver på GPU. Ikke dra inn UHDU.
-
-%In the final journal article we could discuss Conjugated gradient and Woodbury and the benefits they bring.
-%Other solvers that provides different benefits includes Conjugated gradient and the use of the Woodbury matrix identity.
-
-\subsection{Compute Beamformer Output}
-The calculation of the beamformer output in (\ref{eq:z_mv}) has been implemented using one thread per element weight ($L$ threads per weight vector).
-After the solution vector $\vec{b}$ has been placed in shared memory, one of the $L$ threads finds the sum of $\vec{b}$.
-%Another option is to let several threads cooperate to form the sum using a binary reduction sum. However, in the following benchmarks the first approach has been used. 
-Each of the $L$ threads then reads one element of $\vec{b}$, and scales it with the inverse sum to get the weight vector $\vec{w}[n]$.
-
-The final output can be formed by either reducing the data down to length $L$ as in (\ref{eq:z_mv}), or increasing the weight vector to length $M$ as in (\ref{eq:z_mv2}).
-Since the design is $L$ threads per output, the data is reduced to length $L$ by letting thread $r$ calculate
-\begin{align}\label{eq:subsum}
-\breve{x}_{r} = \sum_{k=r}^{r+N_L-1}x_{k} \text{ for } r \in [0, L-1].
-\end{align}
-Each thread now computes $z_r = w_r^*x_r$, and finally one thread computes and outputs $z = \frac{1}{N_L}\sum_{r=0}^{L-1} z_r$ to global memory.
-%The last sum can also be performed using a binary reduction sum across several threads for increased speed.
-
-\begin{figure}
-\centerline{\includegraphics[width=3.2in]{gfx/gpu_layout_bs_2.eps}}
-\caption{Calculation of the sliding beamspace transform.
-A group of $N_b$ threads is assigned a set of subarrays which they transform to beamspace in a sliding window fashion.
-In this example $N_b=3$, $l$ is the current subarray and $k$ is the current beam.}
-\label{fig:gpulayoutbs}
-\end{figure}
-
-\section{Parallel BS-Capon}\label{sec:bs}
-In this section we will derive a parallel implementation of the BS-Capon beamformer, and comment on how it deviates from the implementation in Section \ref{sec:meth}.
-The three main steps described in Section \ref{sec:meth} are present here as well, but in addition we need to transform the per-element data to beamspace.
-This can be done using a FFT, but for few beams ($N_b < \log{L}$) a matrix multiply with the truncated Butler matrix is less costly.
-
-In order to reduce the complexity of the matrix inversion problem the transformation must be performed before or after the covariance matrix has been estimated.
-If the transformation is applied before, one transformation is needed per subarray.
-This results in subarrays that no longer overlap.
-The amount of input data are therefore changed from $M$ to $N_LN_b$ elements per sample, but all remaining steps should get lower computational complexity as long as $N_b < L$.
-
-The other option is to transform the element space covariance matrix to beamspace:
-\begin{align}
-\mat{\hat{R}}_\text{BS} &= \mat{B}\mat{\hat{R}}\mat{B}^H.
-\end{align}
-The advantage of this approach is that minor changes have to be made to the pipeline described in Section \ref{sec:meth}, but it will only speed up the solving step if the solution $\vec{b}_{BS}$ is transformed back to element space.
-In this paper we will focus on the first approach, and how it has been implemented on the GPU.
-%By this approach we go from solving $\mat{\hat{R}}\vec{b} = \vec{1}_L$, which is an $L \times L$ system, to solving the $N_b \times N_b$ system $\mat{\hat{R}}_{BS}\vec{b}_{BS} = (\sqrt{L}\vec{e}_1)$.
-Now, one can either choose to transform the subarray sums, as in (\ref{eq:subsum}), to beamspace or transform the result, $\vec{b}_{BS}$, back to element space.
-In both cases solving has to be performed against a beamspace steering vector
-\begin{align}
-\vec{a}_\text{BS} &= \mat{B}\vec{1}_L = \sqrt{L}\vec{e}_1,
-\end{align}
-where $\vec{e}_i$ has the value one in the $i$th position and zeroes elsewhere.
-
-As mentioned, transforming subarrays into beamspace yields subarrays that no longer overlap.
-This means that the spatial overlap previously exploited when calculating the sample covariance matrix in element space is no longer available.
-On the other hand, since $N_b$ should be small ($N_b < 32$), one thread assigned to each element in $\mat{\hat{R}}_\text{BS}$ is now an option.
-A kernel similar to the one described in \cite{Chen2011} has therefore been implemented.
-In addition, for small $N_b$ values, several covariance matrices are calculated per thread block to achieve high GPU occupancy.
-The computational complexity of this step is now $O(N_KN_LN_b^2)$, which for $L=M/2$ is less than the complexity of building $\mat{\hat{R}}$ in element space as long as $N_b < \sqrt{M}$.
-
-%The implementation calculating covariance matrices therefore needs to be adjusted to handle non-overlapping subarrays.
-The complexity of constructing the covariance matrix is changed to $O(N_b^2L)$, so constructing the covariance matrices will be faster in beamspace as long as $N_b < \sqrt{L}$.
-%Derivation of sliding beamspace transformation. 
-We will now show how to minimize the amount of operations required for the beamspace transformation by utilizing the overlap between subarrays previously exploited when estimating the covariance matrix in element space.
-Given a $M$ long data vector divided into $N_L$ subarrays, the beamspace transformation of the $l$th subarray for the $k$th beam is from (\ref{eq:butler}) found to be:
-\begin{align}\label{eq:bs_formula}
-x_{\text{BS},l,k} &= \mat{B}_k\vec{x}_l \,\,\,\,\,\,\,\,\, k \in [0, N_b-1] \\
-&= \frac{1}{\sqrt{L}}\sum_{p=i}^{l+L-1}x_p e^{-j2\pi k(p-l)/L},
-\end{align}
-where $\mat{B}_k$ is the $k$th row of $\mat{B}$. Manipulating this expression we find that for $l \in [1, L-1]$
-\begin{align}\label{eq:sliding_bs}
-x_{\text{BS},l,k} &= \frac{1}{\sqrt{L}}((x_{\text{BS},l-1,k} - x_{l-1})e^{-j2\pi k/L} \nonumber \\ &+ x_{i+L-1}e^{-j2\pi k(L-1)/L}) \nonumber \\
-&= \frac{1}{\sqrt{L}}(x_{\text{BS},l-1,k} - x_{l-1} + x_{i+L-1})e^{-j2\pi k/L}.
-\end{align}
-We denote this the sliding-window beamspace transformation (SBS), and it is equal to a normalized sliding DFT \cite{Lyons2003}.
-For $l=0$ a full computation has to be performed using (\ref{eq:bs_formula}).
-
-Equation (\ref{eq:sliding_bs}) shows that instructions are saved if we utilize the overlap between subarrays, but with an increase in task level granularity (as discussed in Section \ref{sec:calcR}).
-In this situation, it might be tempting to use one thread per beam (index $k$) that calculates this beam for all subarrays.
-However, since the number of selected beams usually is small ($N_b \ll M$) we will end up with a large memory-per-thread ratio ($M/N_b$).
-Using one thread per beam is therefore not feasible for small-$N_b$-large-$M$ configurations.
-One solution, as depicted in Fig.
-\ref{fig:gpulayoutbs}, is to combine SBS with full computations.
-A number of sub thread groups of size $N_b$ is created that targets different sets of subarrays which they transform using the SBS algorithm.
-In that way, the number of threads per sample is increased and thus the memory-per-thread ratio is decreased.
-
-After solving $\vec{b}_\text{BS} = \sqrt{L}\mat{\hat{R}}_\text{BS}^{-1}\vec{e}_1$, the beamspace weight vector is calculated as 
-\begin{align}
-\vec{w}_\text{BS} = \frac{\vec{b}_\text{BS}}{\sqrt{L}\vec{e}_1\vec{b}_\text{BS}} \in \mathbb{C}^{N_b}
-\end{align}  
-using one thread per beam, and applied to the beamspace data summed across subarrays,
-\begin{align}
-z[n] = \vec{w}_\text{BS}^H\sum_{l=0}^{L-1}\vec{x}_{\text{BS},l}. 
-\end{align}
-Several pixels are calculated per thread block for small values of $N_b$ in order to achieve high GPU occupancy. The beamspace data summed across subarrays are calculated as part of the SBS kernel.
-
-%How do we calculate the number of possible subarrays $N_L$ with overlap $L-s$ in a data vector of length $M$.
-%\begin{align}
-%N_L = floor(\frac{M}{s}) - floor(\frac{L-1}{s})
-%\end{align}
-
-As presented in (\ref{eq:diag}), adaptive diagonal loading is performed by loading the sample covariance matrix with the average energy per element, that is $\text{trace}\{\mat{\breve{R}}\}/L$.
-When data are transformed into beamspace, a high percentage of the total energy is mapped to a small subset of beams.
-Therefore, if a set of high energy beams is selected, the average energy per beam will be higher than the average energy per element ($\text{trace}\{\mat{\breve{R}}\}/L < \text{trace}\{\mat{\breve{R}}_{\text{BS}}\}/N_b$).
-This means that less diagonal loading (smaller $d$) has to be applied when the covariance is estimated among beams and not elements.
-
-To summarize, by transforming data to beamspace the complexity of all subsequent steps is potentially reduced.
-Calculation of the covariance matrix is now $O(N_KN_LN_b^2)$, the inversion step is $O(N_b^3)$, and calculation of weights and the beamformer output are both $O(N_b)$.
-The added complexity of SBS is $O(N_bL)$.
-To maximize performance, large adjustments to the processing pipeline described in Section \ref{sec:meth} is required.
-
-%(Go in detail about the implementation using sliding DFT).
-
-%(Go in detail about what kind of modifications we need to do to each of the three steps)
-
-%Note that $\mat{B}\mat{B}^H = \mat{I}$ and $\mat{B}^H\mat{B} \ne \mat{I}$ when $N_b < L$.
-
-%Another way of computing the beamspace transformation is to transform the steering vector $\vec{a}$ and covariance vector $\mat{\hat{R}}$ beforing solving the following system of linear equations: $(\mat{B}\mat{\hat{R}}\mat{B}^H)^{-1}(\mat{B}a) = b_{BS}$.
-
-%\subsection{Iterative Capon}
-%The idea behind iterative Capon is to perform Capon using small subarrays in several iterations until there is only two subarrays to sum.
-In this way we fix the size of L, hence the complexity of solving $\mat{R}\vec{b} = \vec{1}$.
-However, for each iteration we re-obtain L-1 degrees of freedom until all are retrieved after $K = M-L+1$ iterations.
-Since subarray averaging has been applied to decorrelated the signal, we might loose this effect when reestablishing all degrees of freedom.
-The number of iterations should therefore be restricted to $K/2$ to equal the normal choice of $L = M/2$ when it comes to degrees of freedom.
-
-\section{Benchmarks}\label{sec:bench}
-
-%Describe the computer system (CPU, GPU...). Describe the data that are processed (field II, Channel data from GE Vingmed).
-% stolpediagram, med de ulike metodenes kjøretid oppehverandre.
-
-\begin{figure}
-%\centerline{\subfloat[]{\includegraphics[width=3.5in]{gfx/benchmark_bar_32_40_400.png} \label{fig:benchUS32}}
-%\hfil
-%\subfloat[]{\includegraphics[width=3.5in]{gfx/benchmark_bar_64_80_606.png}%
-%\label{fig:benchUS64}}}
-\centerline{\includegraphics[width=3.5in]{gfx/benchmark_bar_64_80_540.png}}
-\caption{Benchmark of GPU Capon beamforming in element space for a cardiac image covering a $70\degree$ wide $15$ cm deep sector.
-Execution time is plotted as a function of subarray length $L$.
-The plot shows execution times for the three steps listed in Section \ref{sec:meth}: Calculating covariance matrices with three different choices of temporal averaging, solver, and the final beamformer sum.}
-\label{fig:bench}
-\end{figure}
-
-\begin{figure*}[!t]
-\centerline{\subfloat[]{\includegraphics[width=3.45in]{gfx/benchmark_bar_bs_M=64_Nx=80_Ny=540_Nb=3.png} \label{fig:benchBS64}}
-\hfil
-\subfloat[]{\includegraphics[width=3.45in]{gfx/benchmark_bar_bs_M=96_Nx=120_Ny=540_Nb=3.png}%
-\label{fig:benchBS96}}}
-\caption{Benchmarks of GPU beamspace Capon beamforming for a cardiac image covering a $70 \degree$ and $15$ cm deep sector.
-Execution time is plotted as a function of subarray length $L$.
-If $L$ had been kept constant on $M/2$ and execution time was plotted against $N_b$, the plot would be similar to the one in Fig.
-\ref{fig:bench}.
-The variation in execution time is caused by adaptive kernel launch configurations.}
-\label{fig:benchBS}
-\end{figure*}
-In this section we present benchmarks of the implementations presented in Sections \ref{sec:meth} and \ref{sec:bs} performed on a Nvidia Quadro 6000 graphics card (14 SMs, 1 TFlops theoretical single precision performance, 6 GB global memory, and 144 GB/s memory bandwidth).
-Problem sizes are calculated based on 64 and 96 element phased arrays with a 3.4 MHz center frequency.
-Assuming a system bandwidth of 80\% (relative to the center frequency) we need a complex sampling frequency of 2.7 MHz, and when imaging down to a depth of 15.4 cm we have approximately 540 samples per range line.
-To sample a $70\degree$ sector with 64- and 96-element arrays ($d=\lambda/2$ pitch) at twice the Rayleigh criterion ($\lambda/(2Md)$), we need approximately 80 and 120 receive beams respectively.
-The presented execution times do not include transfer of data from CPU- to the GPU-side, since data are assumed to be pre-delayed and available in global GPU memory.
-What is included are all calculations, memory transfers from global to shared GPU memory, and shared memory access.
-
-A benchmark of the ES-Capon is presented in Fig.
-\ref{fig:bench}.
-The benchmark is based on a 64-element phased array, using 80 receive beams.
-For each subarray size, the total execution time is split into one bar for each of the three steps described in Section \ref{sec:meth}, and the calculation of $\mat{\hat{R}}$ is presented for three different values of $K$.
-For $M=64$, $L=32$ and $K=2$, a typical setup for high-quality cardiac imaging, processing throughput is 7 frames per second.
-
-In Fig.
-\ref{fig:benchBS} we present similar benchmarks for the BS-Capon beamformer.
-Benchmarks are based on 64- and 96-element phased arrays, using 80 and 120 receive beams.
-The figure only presents results with use of temporal averaging ($K=2$).
-For $M=64$, $L=32$, $K=2$ and $N_b=3$, processing throughput is 125 frames per second (8 ms).
-For $M=96$, $L=48$, $K=2$ and $N_b=3$, processing throughput is 77 frames per second (13 ms).
-In all four benchmarks there is a line indicating the time it takes to acquire the underlying image using one receive line per transmit.
-Hence, below this line we are capable of performing real-time processing.
-
-\section{Simulated phantom and \textit{in-vivo} cardiac images}\label{sec:images}
-
-\begin{figure*}[!t]
-\centerline{
-\subfloat[]{ \includegraphics[width=2.2in]{gfx/das.png} \label{fig:phantomDAS} }
-\hfil
-\subfloat[]{ \includegraphics[width=2.2in]{gfx/capon.png} \label{fig:phantomMV} }
-\hfil
-\subfloat[]{ 
-\includegraphics[width=2.2in]{gfx/capon_bs.png} \label{fig:phantomBS} 
-\hfil
-\includegraphics[width=0.35in]{gfx/colorBarPhantom2.png}
-}
-}
-\caption{Simulated phantom containing one bright (20 dB) and one dark (-20 dB) circular area and two point targets (30 dB) in speckle (0 dB).
-The simulation was done using Field II with a 64-element, 3.4 MHz, and $\lambda/2$ pitch phased array.
-Dynamic range is -20 dB to 20 dB with mean value represented as 0 dB.
-The image is the 15th frame in the attached videos.
-a) Delay-and-sum with uniform weights.
-b) Element-space Capon ($L=32, K=2, d=0.1$) \multimedia{Media-Movie 1}.
-c) Beamspace Capon ($L=32, K=2, N_b=3, d=0.01$) \multimedia{Media-Movie 2}.}
-\label{fig:phantom}
-\end{figure*}
-
-\begin{figure*}[!t]
-\centerline{
-\subfloat[]{\includegraphics[width=2.2in]{gfx/das_invivo.png} \label{fig:invivoDAS}}
-\hfil
-\subfloat[]{\includegraphics[width=2.2in]{gfx/capon_invivo.png} \label{fig:invivoMV}}
-\hfil
-\subfloat[]{
-\includegraphics[width=2.2in]{gfx/capon_bs_invivo.png} \label{fig:invivoBS}}
-\hfil
-\includegraphics[width=0.4in]{gfx/colorBarInvivo2.png}
-}
-\caption{\textit{In-vivo} harmonic cardiac ultrasound image acquired using a 3.5 MHz, 64 elements phased array.
-Number of receive beams for the $70\degree$ sector is 98, and number of samples in range is 610.
-Dynamic range is -20 dB to 20 dB with mean value represented as 0 dB.
-The image is the first frame in the attached videos.
-a) Delay-and-sum with uniform weights \multimedia{Media-Movie 3}.
-b) Element-space Capon ($L=32, K=2, d=0.1$) \multimedia{Media-Movie 4}.
-c) Beamspace Capon ($L=32, K=2, N_b=3, d=0.01$) \multimedia{Media-Movie 5}.}
-\label{fig:invivo}
-\end{figure*}
-
-\begin{figure*}[!t]
-\centerline{
-\subfloat[]{\includegraphics[width=3in]{gfx/simulation_slice.png} \label{fig:simulationSlice}}
-\hfil
-\subfloat[]{\includegraphics[width=3in]{gfx/invivo_slice.png} \label{fig:invivoSlice}}
-}
-\caption{Lateral slices taken from the images in Fig.
-\ref{fig:phantom} and Fig.
-\ref{fig:invivo}.
-a) From Fig.
-\ref{fig:phantom} at 93 mm constant depth, through the two bright point scatterers.
-b) From Fig.
-\ref{fig:invivo} at 96 mm range, through the tip of the mitral valve.}
-\label{fig:slices}
-\end{figure*}
-
-To investigate the image quality of both the ES and BS-Capon beamformer, a phantom was simulated using Field II \cite{Jensen1992, Jensen1996a} (Fig.
-\ref{fig:phantom}).
-The simulation was performed using a 64-element, $\lambda/2$ pitch, 10 $\mu$m kerf, 80\% bandwidth, and 10 mm high phased array.
-The transducer was excited with a 3.4 MHz and $1.5\lambda$ long pulse with transmit focus at 90 mm.
-The phantom contains two circular areas, one bright at 20 dB and one dark at -20 dB, and two point targets at 30 dB, located at 90 mm depth.
-Speckle level is at 0 dB, covering an area from $\pm90\degree$ in azimuth.
-The resulting image has been processed both using DAS with uniform weights, ES- and BS-Capon.
-The presented image is the first frame in a series of images where motion is imposed on both circles, point targets and the probe.
-The entire simulation is presented in two attached videos \multimedia{Media-Movie 1 and 2}.
-
-Fig.
-\ref{fig:invivo} presents results of applying ES- and BS-Capon on an \textit{in-vivo} image of the left ventricle.
-The image was acquired using a 64-element and $0.65\lambda$ pitched phased array in harmonic mode (1.7 MHz at transmit and 3.4 MHz at receive).
-The image is extracted from a cardiac cycle, that can be viewed in its full length in the attached videos \multimedia{Media-Movie 3, 4 and 5}.
-It is included to show that both ES-Capon and BS-Capon beamforming are feasible for \textit{in-vivo} cardiac imaging, not just in simulations.
-
-To get a better impression of the improved lateral resolution obtained with Capon beamforming, two slices taken from the images in Fig.
-\ref{fig:phantom} and \ref{fig:invivo} are included in Fig.
-\ref{fig:slices}.
-For the phantom, the cut goes through the two point targets at 90 mm constant radius, and for the cardiac image it goes through the tip of the mitral valve at 96 mm constant radius.
-
-Both data sets have been processed with the following settings for the ES-Capon beamformer: $L = 32$, $K = 2$, and $d=0.1$.
-These settings have previously been shown to provide improved lateral resolution while maintaining DAS-like speckle \cite{Synnevag2007a}.
-For BS-Capon the following parameters have been used: $L=32$, $K=2$, $d=0.01$, and $N_b=3$.
-For the simulated phantom, 16x lateral oversampling on transmit has been performed in order to avoid under sampling of the two bright point scatterers as they move.
-The cardiac recording of a 70\degree{} sector has a minimal lateral sampling of 98 receive beams ($\lambda/2Md$).
-
-\section{Trading resolution for speed}\label{sec:trade}
-As seen from Fig.
-\ref{fig:bench}, execution time is reduced when $L$ is reduced.
-In Fig.
-\ref{fig:benchBS} we see that execution time is also reduced when moving to beamspace.
-So to gain an  increase in processing throughput we can reduce $L$ in element space or $N_b$ in beamspace.
-In this section we investigate how adjusting $L$ for ES-Capon and $N_b$ for BS-Capon impacts the lateral resolution.
-In Fig.
-\ref{fig:speed_res_trade}, the minimum distance between the two point targets in Fig.
-\ref{fig:phantom} (while preserving a 6 dB saddle point in between the two) is plotted against $L$ (when processed with ES-Capon) and $N_b$ (when processed with BS-Capon).
-To produce the result in Fig.
-\ref{fig:speed_res_trade}, $d=0.01$ has been used on both methods in order to obtain the same resolution for $L=N_b=M/2$.
-We considered this to be a small value for ES-Capon and a normal value for BS-Capon (cf.
-Section \ref{sec:bs}), hence the resolution for ES-Capon is here better than what is presented in Fig.
-\ref{fig:phantom}.
-The precision of the presented measurements is 0.12 mm, dictated by the simulated frame rate, 25 frames per second (FPS), and the relative speed of the two point scatterers, 3 mm/s.
-
-For high values of $L$ and $N_b$ the two methods are equal, but for lower values, BS-Capon provides improved resolution compared to using small subarrays.
-We see that the resolution for BS-Capon gradually decreases due to an automatic increase in diagonal loading (cf.
-end of Section \ref{sec:bs}).
-If we compensate by reducing $d$, BS-Capon will provide resolution equal to ES-Capon (with $L=M/2$) for all values of $N_b$ except one \cite{Nilsen2009}.
-For $N_b=1$ the resolution is worse than for DAS because of a triangular apodization caused by subarray averaging.
-
-In element space, resolution decays rapidly since the effective adaptive aperture is proportional to $L$.
-For $L = N_b=3$, BS-Capon provides almost 1 mm improved resolution over DAS and ES-Capon.
-However, note that the resolution of the Capon beamformers strongly depends on image SNR and the choice of parameters.
-Thus, different SNR scenarios and choices of parameters will lead to different results.
-It is also worth noting that ES-Capon beamforming in this example performs worse than DAS for small values of $L$.
-For these small subarrays, the method is not able to steer zeros at each point target or accurately micro-steer the main lobe.
-Subarray averaging further impose a triangular or trapezoid apodization function, which leads to worse resolution if the introduced adaptivity can not compensate for the increase in main lobe width.
-For $L=1$ the ES-Capon is equal to DAS.
-
-\begin{figure}
-\centering
-\includegraphics[width=3in]{gfx/speed_res_trade_d=0-01.png}
-\caption{Trade-off between resolution and speed by reducing subarray size or reducing beamspace dimension (based on Fig.
-\ref{fig:phantom}).
-As $L$ or $N_b$ gets smaller, the execution time is reduced (Fig.
-\ref{fig:bench} and \ref{fig:benchBS}).
-When $N_b$ is varied, $L$ is fixed on 32.}
-\label{fig:speed_res_trade}
-\end{figure}
-
-\section{Discussion}\label{sec:dis}
-In this paper we have presented a parallel implementation of the Capon beamformer in both element- and beamspace for high-resolution cardiac ultrasound imaging.
-The results show that it is now possible to apply BS-Capon in real-time in cardiac ultrasound imaging by utilizing the power of GPUs.
-According to Fig.
-\ref{fig:benchBS} it is also possible to process two parallel receive lines per transmit beam (for both $M=64$ and $M=96$.
-$L=M/2$).
-These extra lines might be needed for increased sampling instead of increased frame rate when using the Capon beamformers in high SNR situations (Fig \ref{fig:phantom}).
-If parallel receive lines are used to obtain denser sampling, acquisition time remains the same but processing times are increased with a factor equal to the amount of oversampling.
-If parallel receive lines are used for increased frame rate, the acquisition time is lowered, and the processing time remains constant.
-Hence, four parallel receive lines are not possible to process in real-time using BS-Capon (with $N_b=3$) in combination with our single GPU setup.
-
-For ES-Capon, processing rates for high quality settings are still too high to support real-time cardiac ultrasound imaging.
-From Fig.
-\ref{fig:bench} we see that for $L = M/2 = 32$ and $K=1$, the total execution time is 140 ms (7 FPS).
-Taking into account that we use complex data, but less samples per image, these results are similar to the performance reported by Chen et al.
-\cite{Chen2011}.
-
-A major part of the total execution time for ES-Capon comes from solving the system of linear equations.
-This limitation is targeted by reducing the sample covariance matrix size in beamspace, and as shown in Fig.
-\ref{fig:benchBS}, inverting the $3\times3$ beamspace covariance matrix is then barely visible.
-The time it takes to construct all covariance matrices is also reduced, but for few beams this is now the major contributor to the total execution time.
-If real-time is defined as 10 FPS \cite{Chen2011} almost all execution times in both Fig.
-\ref{fig:bench} and \ref{fig:benchBS} are real-time.
-More natural for cardiac ultrasound imaging is to define real-time as the time it takes to acquire the underlying image.
-With this requirement the images formed using a 64 and 96 element array have a real-time requirement of 62.5 and 41.7 FPS respectively using single line acquisition.
-Each benchmark plot has a line indicating this real-time requirement.
-With beamspace processing, all processing times for large subarrays are below this line.
-
-This paper further shows how to obtain a parallel formulation of the ES-Capon and BS-Capon beamformer, and how to implement them in a GPU framework.
-The per-pixel granularity suggests that the beamformer is easily dividable across multiple GPUs.
-Real-time execution for all cases presented in this paper is therefore possible if several GPUs are used simultaneously.
-However, there is usually a space-cost-energy limitation for how many GPUs that can be fitted in an ultrasound scanner.
-
-Turning to image quality we see from Fig.
-\ref{fig:phantom}, and the attached video of the simulation, that Capon beamforming provides improved lateral resolution of moving point targets both in element- and in a reduced beam-space ($N_b=3$).
-Our preliminary finding is that the Capon beamformers are local shift invariant as long as the lateral sampling is increased to a curtain level \multimedia{Media-Movie 1 and 2}.
-How large this oversampling factor needs to be and how the oversampling should be conducted is outside the scope of this paper.
-No major improvement in contrast is visually observed in the presented images and videos, other than slightly rounder circles due to mainlobe micro steering.
-In the cardiac recording (Fig.
-\ref{fig:invivo}), structures are thinner and better resolved when the Capon beamformers are applied.
-However we observe a loss in contrast, especially in the septum (Fig.
-\ref{fig:slices}).
-This loss can be partially explained by the presence of highly correlated signals and mismatch between the assumed steering vector and signal.
-Future work should focus on solving this issue.
-
-Nilsen and Hafizovic suggested to do automatic selection of beams in beamspace, hence picking the beams containing most energy.
-In this paper the center beams are always picked.
-This seems to work fine for focused systems, but an adaptive selection of beams should be implemented if interference at high angles is shown to contaminate the image.
-
-%We expect that a large speedup had the Capon beamformer been implemented on the CPU using threads and SIMD (single instruction multiple data) instructions.
-
-%For a typical cardiac ultrasound image of 80*400 pixels (70 degree sector, 15cm range) acquired using a 2.5MHz, $M=64$ element phased array, the result is 10fps (subarray length $L=M/2$, temporal smoothing over 3 samples).
-If we do a 2-element pre-beamforming to get the channel count down to 32, the rate increases to 44fps.
-For a 32 element phased array we need less beams to cover the sector ($40 \times 400$ pixels), hence with the same parameters the frame rate increases to 87fps.
-The bottleneck of the Capon algorithm is to find the inverse covariance matrix for all pixels.
-This is an $O(L^3)$ operation, and results show that when L is larger than 16, real time performance is not possible for cardiac imaging with current high-end hardware.
-The target GPU has been Nvidia’s Quadro 6000, capable of 1Tflops.
-
-\section{Conclusion}\label{sec:con}
-In this paper a parallel formulation and implementation of the element- and beamspace Capon beamformer have been presented, and
-both beamformers have successfully been implemented on the GPU.
-For a three dimensional beamspace, processing rates supporting real-time cardiac ultrasound imaging was shown to be possible, meaning that processing rates higher than the image acquisition rate are obtained for a wide range of parameters.
-Image quality was investigated both in a simulated phantom and an \textit{in-vivo} cardiac dataset.
-These results showed that Capon beamforming is feasible for cardiac ultrasound imaging, providing images with improved lateral resolution and no local shift-variance effects, both in element- and beamspace.
-
-%\section{Future work}
-%Study shift invariance and minimum sampling...
-
-%MLA's has, to some extent been used to increase later sampling.
-When using a high resolution beamformer, Nyquist tells us that the sampling frequency must be increase.
-Future work will concentrate on how much over sampling is required when using the Capon beamformer whit i curtain set of parameters...
-
-%Plane waves...
-
-% An example of a floating figure using the graphicx package.
-% Note that \label must occur AFTER (or within) \caption.
-% For figures, \caption should occur after the \includegraphics.
-% Note that IEEEtran v1.7 and later has special internal code that
-% is designed to preserve the operation of \label within \caption
-% even when the captionsoff option is in effect. However, because
-% of issues like this, it may be the safest practice to put all your
-% \label just after \caption rather than within \caption{}.
-%
-% Reminder: the "draftcls" or "draftclsnofoot", not "draft", class
-% option should be used if it is desired that the figures are to be
-% displayed while in draft mode.
-%
-%\begin{figure}[!t]
-%\centering
-%\includegraphics[width=2.5in]{myfigure}
-% where an .eps filename suffix will be assumed under latex, 
-% and a .pdf suffix will be assumed for pdflatex; or what has been declared
-% via \DeclareGraphicsExtensions.
-%\caption{Simulation Results}
-%\label{fig_sim}
-%\end{figure}
-
-% Note that IEEE typically puts floats only at the top, even when this
-% results in a large percentage of a column being occupied by floats.
-
-% An example of a double column floating figure using two subfigures.
-% (The subfig.sty package must be loaded for this to work.)
-% The subfigure \label commands are set within each subfloat command, the
-% \label for the overall figure must come after \caption.
-% \hfil must be used as a separator to get equal spacing.
-% The subfigure.sty package works much the same way, except \subfigure is
-% used instead of \subfloat.
-%
-%\begin{figure*}[!t]
-%\centerline{\subfloat[Case I]\includegraphics[width=2.5in]{subfigcase1}%
-%\label{fig_first_case}}
-%\hfil
-%\subfloat[Case II]{\includegraphics[width=2.5in]{subfigcase2}%
-%\label{fig_second_case}}}
-%\caption{Simulation results}
-%\label{fig_sim}
-%\end{figure*}
-%
-% Note that often IEEE papers with subfigures do not employ subfigure
-% captions (using the optional argument to \subfloat), but instead will
-% reference/describe all of them (a), (b), etc., within the main caption.
-
-% An example of a floating table. Note that, for IEEE style tables, the 
-% \caption command should come BEFORE the table. Table text will default to
-% \footnotesize as IEEE normally uses this smaller font for tables.
-% The \label must come after \caption as always.
-%
-%\begin{table}[!t]
-%% increase table row spacing, adjust to taste
-%\renewcommand{\arraystretch}{1.3}
-% if using array.sty, it might be a good idea to tweak the value of
-% \extrarowheight as needed to properly center the text within the cells
-%\caption{An Example of a Table}
-%\label{table_example}
-%\centering
-%% Some packages, such as MDW tools, offer better commands for making tables
-%% than the plain LaTeX2e tabular which is used here.
-%\begin{tabular}{|c||c|}
-%\hline
-%One & Two\\
-%\hline
-%Three & Four\\
-%\hline
-%\end{tabular}
-%\end{table}
-
-% Note that IEEE does not put floats in the very first column - or typically
-% anywhere on the first page for that matter. Also, in-text middle ("here")
-% positioning is not used. Most IEEE journals use top floats exclusively.
-% Note that, LaTeX2e, unlike IEEE journals, places footnotes above bottom
-% floats. This can be corrected via the \fnbelowfloat command of the
-% stfloats package.
-
-%\section{Conclusion}
-%The conclusion goes here.
-
-% if have a single appendix:
-%\appendix[Proof of the Zonklar Equations]
-% or
-%\appendix  % for no appendix heading
-% do not use \section anymore after \appendix, only \section*
-% is possibly needed
-
-% use appendices with more than one appendix
-% then use \section to start each appendix
-% you must declare a \section before using any
-% \subsection or using \label (\appendices by itself
-% starts a section numbered zero.)
-%
-
-%\appendices
-%\section{Proof of the First Zonklar Equation}
-%Appendix one text goes here.
-
-% you can choose not to have a title for an appendix
-% if you want by leaving the argument blank
-%\section{}
-%Appendix two text goes here.
-
-% use section* for acknowledgement
-\section*{Acknowledgment}
-The authors would like to thank Dr. K. Kristoffersen, GE Vingmed Ultrasound, for valuable comments.
-
-% Can use something like this to put references on a page
-% by themselves when using endfloat and the captionsoff option.
-\ifCLASSOPTIONcaptionsoff
-  \newpage
-\fi
-
-% trigger a \newpage just before the given reference
-% number - used to balance the columns on the last page
-% adjust value as needed - may need to be readjusted if
-% the document is modified later
-%\IEEEtriggeratref{8}
-% The "triggered" command can be changed if desired:
-%\IEEEtriggercmd{\enlargethispage{-5in}}
-
-% references section
-
-% can use a bibliography generated by BibTeX as a .bbl file
-% BibTeX documentation can be easily obtained at:
-% http://www.ctan.org/tex-archive/biblio/bibtex/contrib/doc/
-% The IEEEtran BibTeX style support page is at:
-% http://www.michaelshell.org/tex/ieeetran/bibtex/
-\bibliographystyle{IEEEtran}
-% argument is your BibTeX string definitions and bibliography database(s)
-%\bibliography{IEEEabrv,../bib/paper}
-\bibliography{mybib} % link to mendeley organized bibtex-file
-%\bibliography{IEEEabrv,mybib} 
-% <OR> manually copy in the resultant .bbl file
-% set second argument of \begin to the number of references
-% (used to reserve space for the reference number labels box)
-%\begin{thebibliography}{1}
-
-%\bibitem{IEEEhowto:kopka}
-%H.~Kopka and P.~W. Daly, \emph{A Guide to \LaTeX}, 3rd~ed.\hskip 1em plus
-%  0.5em minus 0.4em\relax Harlow, England: Addison-Wesley, 1999.
-
-%\end{thebibliography}
-
-% biography section
-% 
-% If you have an EPS/PDF photo (graphicx package needed) extra braces are
-% needed around the contents of the optional argument to biography to prevent
-% the LaTeX parser from getting confused when it sees the complicated
-% \includegraphics command within an optional argument. (You could create
-% your own custom macro containing the \includegraphics command to make things
-% simpler here.)
-%\begin{biography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{mshell}}]{Michael Shell}
-% or if you just want to reserve a space for a photo:
-
-% insert where needed to balance the two columns on the last page with
-% biographies
-%\newpage
-
-\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{photos/jon_petter2.eps}}]{Jon Petter \AA{}sen}
-(S12) was born in Porsgrunn, Norway in 1986.
-He received the B.Sc.
-and M.Sc.
-degree in computer science from the University of Oslo, Norway, in 2008 and 2010.
-He is currently pursuing his Ph.D.
-degree in medical ultrasound technology at the Norwegian University of Science and Technology (NTNU) Medical Imaging Lab (MI-Lab), Trondheim, Norway.
-
-He has industry experience from GE Vingmed Ultrasound, Horten, Norway, were he completed three summer internships from 2008 to 2010.
-His research interests include adaptive ultrasound processing techniques and acceleration of ultrasound algorithms using Graphics Processing Units (GPUs).
-\end{IEEEbiography}
-
-\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{photos/jo_inge2.eps}}]{Jo Inge Buskenes}
+\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{bio/jo_inge.eps}}]{Jo Inge Buskenes}
 received the B.Sc.
@@
 He has lectured in digital signal processing at the Gj\o{}vik College University (2009), and at the University of Oslo (2010-2012).
-His research interests include adaptive beamforming, digital image reconstruction, high performance computing, intelligent detector design and embedded Linux platforms.
+His research interests include adaptive beamforming, digital image reconstruction, high performance computing, intelligent detector design and open source software.
 \end{IEEEbiography}
-
-\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{photos/carl-inge.eps}}]{Carl-Inge Colombo Nilsen}
-(S’06–M’10) received the M.Sc.
-and Ph.D.
-degrees in computer science from the University of Oslo, Norway, in 2005 and 2010.
-He is currently working at the University of Oslo as a postdoctoral research fellow.
-His research interests include signal and array processing for ultrasound imaging and other acoustic applications.
-\end{IEEEbiography}
-
-% insert where needed to balance the two columns on the last page with
-% biographies
-%\newpage
-
-\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{photos/andreas.eps}}]{Andreas Austeng}
-was born in Oslo, Norway, in 1970.
-He received the M.S.
-degree in physics in 1996 and the Ph.D.
-degree in computer science in 2001, both from the University of Oslo.
-Since 2001, he has been working at the Department of Informatics, University of Oslo, first as a postdoctoral research fellow and currently as an associate professor.
-His research interests include signal and array processing for acoustical imaging.
-\end{IEEEbiography}
-
-%\newpage
-
-\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{photos/sverre.eps}}]{Sverre Holm}
-(M’82-SM’02) was born in Oslo, Norway, in 1954.
-He received the M.S.
-and Ph.D.
-degrees in electrical engineering from the Norwegian Institute of Technology, Trondheim, in 1978 and 1982, respectively.
-His academic experience includes Yarmouk University in Jordan (1984-1986) and the Norwegian Institute of Technology and later the University of Oslo (1989-1994) as an adjunct professor.
-Since 1995, he has been a professor of signal processing in the Department of Informatics at the University of Oslo.
-In 2002, he was elected a member of the Norwegian Academy of Technological Sciences, and in 2009, he became an adjunct professor in the Norwegian University of Science and Technology (NTNU) Medical Imaging Lab, Trondheim.
-His industry experience includes GE Vingmed Ultrasound, Horten, Norway (1990-1994).
-He worked with Sonitor Technologies (2000-2005), where he developed ultrasonic technology that is the core of the indoor positioning system described in \textit{Scientific Americanin} 2008 as "a positioning system that goes where GPS can't." He has published about 130 publications internationally, plus about 10 patents.
-He has had sabbaticals at GE Global Research, NY (1998), and the Waves and
-Acoustics Laboratory of ESPCI, Paris (2008-2009).
-He was an associate editor of the \textit{IEEE Transactions on Ultrasonics, Ferroelectrics, and Frequency Control} from 1997 to 2002.
-His research interests are medical ultrasound imaging, modeling of tissue-ultrasound interaction, and ultrasound positioning.
-\end{IEEEbiography}
-
-% You can push biographies down or up by placing
-% a \vfill before or after them. The appropriate
-% use of \vfill depends on what kind of text is
-% on the last page and whether or not the columns
-% are being equalized.
-
-%\vfill
-
-% Can be used to pull up biographies so that the bottom of the last one
-% is flush with the other column.
-%\enlargethispage{-5in}
-
-% that's all folks
-\end{document}

### 2014-04-23 HowtoHowtoCapon/submission/final/gfx/convert.sh

@@ -1,0 +1,25 @@
+#!/bin/bash
+
+for i in $(ls ../../../gfx/buske*eps); do
+   FN=${i##*/}
+   echo "cp $i ${FN%*.eps}_online.eps"
+   cp $i ${FN%*.eps}_online.eps
+   echo "${FN%*.eps}_online.eps > make black and white > $FN"
+   gs -sDEVICE=psgray -dNOPAUSE -dBATCH -dSAFER -sOutputFile=tmp.eps ${FN%*.eps}_online.eps
+   ps2eps tmp.eps
+   mv tmp.eps.eps $FN
+#    echo "../bw_convert ${FN%*.eps}_online.eps $FN"
+#    ../bw_convert ${FN%*.eps}_online.eps $FN
+done
+# ../bw_convert buske
+# cp beamforming.eps buske1.eps
+# cp implementation.eps buske2.eps
+# cp mvdr_complexity.eps buske3.eps
+# cp mvdr_build_R.eps buske4.eps
+# # # # cp buildR-breakdown.eps buske5.eps
+# cp mvdr_implementation.eps buske6.eps
+# cp plot_holmengraa_L16_Navg1.eps buske7.eps
+# cp benchmark.eps buske8.eps
+# cp code_assessment.eps buske9.eps
+# cp code_assess_flops.eps buske10.eps
+# cp benchmark_boston.eps buske11.eps

### 2014-04-23 HRUIofMovingObjects/jrnl_article_1st.tex

--- 2014-03-04 HRUIofMovingObjects/jrnl_article.tex
@@
 \newcommand{\R}{$\mat{\hat{R}}$ }
-\newcommand{\degree}{\ensuremath{^\circ}}
 
@@
 
-\IEEEPARstart{I}{n} the last decade several authors have applied the Capon beamformer [also known as the minimum variance beamformer] for medical ultrasound imaging \cite{Synnevag2007, Vignon2008, Viola}.
-The method has been shown to improve lateral resolution and edge sharpness in ultrasound images \cite{Synnevag2007, Synnevag2009, Chen2011}.
+\IEEEPARstart{T}{he} last decade several authors have applied the Capon beamformer (also known as the minimum variance beamformer) for medical ultrasound imaging \cite{Synnevag2007, Vignon2008, Viola}.
+The method has been shown to improve lateral resolution and edge sharpness \cite{Synnevag2007, Synnevag2009, Chen2011} in ultrasound images.
 This is accomplished by adaptively adjusting the array weights based on element-to-element covariance in order to minimize interference.
@@
 Hence, the interference-noise matrix ($\sigma_0^2\mat{Q}$ in \cite{Cox1973}) is the identity.
-The effect of typical subarray averaging and diagonal loading have been added to the Capon beamformer output (subarray length of 32 and a diagonal loading factor of 0.01).
+The effect of typical subarray averaging and diagonal loading ($L=32$ and $d=1/100$) have been added to the Capon beamformer output.
 We see that a DAS beamformer, using the required beam spacing, could yield a $1.9$ dB scalloping loss \cite{Harris1978} if the object is moving parallel to the array.
@@
 }
-\caption{Estimated drop in output power mid-way between receive beams for delay-and-sum and Capon beamforming (with subarray length of 32 and a diagonal loading factor of 0.01) applied in a farfield-narrowband setting.
+\caption{Estimated drop in output power mid-way between receive beams for delay-and-sum and Capon beamforming ($M=64$, $L=32$, $d=0.01$) applied in a farfield-narrowband setting.
 %The scanned output from both beamformers is plotted for two different time instances.
@@
 \begin{align}\label{eq:das}
-z[n] = \sum_{m = 0}^{M-1}w_m^*x_m[n - \Delta_m[n]] = \vec{w}^H\vec{x}[n],
+z[n] = \sum_{m = 0}^{M-1}\vec{w}_m^*\vec{x}_m[n - \Delta_m[n]] = \vec{w}^H\vec{x}[n],
 \end{align}
@@
 \subsection{Capon Beamforming}
-The Capon beamformer produces a weight vector, like the one in (\ref{eq:das}), based on the impinging signal and an optimization criterion. The problem is as follow \cite{Capon1969}:
+The Capon beamformer produce a weight vector, like the one in (\ref{eq:das}), based on the impinging signal and an optimization criterion. The problem is as follow \cite{Capon1969}:
 \begin{align}
@@
 The required lateral sampling in an ultrasound image for a given center frequency, $f_c$, and bandwidth, $B$, is given by the aperture size, $D$, and the wavelength of the center frequency, $\lambda_c = c/f_c$, where $c$ is the speed of sound.
-The required Nyquist beam spacing is then:
+The required Nyquist beam spacing is then \cite{Hergum2007}:
 \begin{align}
-\delta\theta < \frac{\lambda_c}{2qD}, \label{eq:resolution}
+\Delta < \frac{\lambda_c}{qpD}, \label{eq:resolution}
 \end{align}
-where the factor two is the effect of a two-way acquisition \cite{Hergum2009}, and where $q$ is an oversampling factor which can be applied to reduce folding of frequencies between $f_c$ and $f_c + B/2$.
-In this paper we will use the $q$-factor to perform oversampling prior to Capon beamforming.
+where the oversampling factor $p$ is applied to reduce folding of frequencies between $f_c$ and $f_c + B/2$.
+A $p$-factor of two has been used throughout this paper.
+The oversampling factor $q$ will be used to oversample prior to Capon beamforming and is one if not stated otherwise.
 
@@
 The maximum for each beam profile is also marked with a black dot, and the amplitude variation among these points is presented in a subplot next to the LSV-plot.
-The LSV-plots based on simulations have been up-interpolated to 128 samples in each direction using cubic interpolation if the number of receive beams is less than 128.
+The LSV-plots based on simulations has been up-interpolated to 128 samples in each direction using cubic interpolation if the number of receive beams is less than 128.
 
@@
 Dotted vertical lines indicate position of transmit-receive beam pairs.
-The point scatterer response has been simulated using Field II \cite{Jensen1992, Jensen1996a} using a 96 element phased array with a 22 mm azimuth aperture, which was excited with a 2.5 MHz and 1.5 cycles long pulse.
-Transmit focus in azimuth and elevation were configured to be equal and the point scatterer was moved through this point.
-The transmit focus is where the ultrasound beam is at its narrowest, hence shift variance effects will appear at this location first.
-The beam density was calculated  based on (\ref{eq:resolution}) with the speed of sound equal to 1540.
+The point scatterer response has been simulated using Field II \cite{Jensen1992, Jensen1996a} with a probe setup equal to the M5S probe by GE Vingmed Ultrasound, Horten, Norway.
+The probe was excited with a 2.5 MHz and 1.5 cycles long pulse.
+The transmit focus in azimuth and elevation were configured to be equal and the point scatterer was moved through this point.
 We see that the contours in Fig.\,\ref{fig:das_shiftinvariant} are approximately diagonal (low P-MAE) and that the amplitude variation among the maximum points is minimal (low A-MAE).
@@
 We see how the signal cancellation caused by the difference between the signal propagation vector and the assumed steering vector increases as the point moves away from a given transmit-receive beam.
-As a consequence the scalloping loss is tremendous, and not far from the predicted value in Fig.\,\ref{fig:das_capon_beams}.
+As a consequence the scalloping loss is tremendous, and not far from the predicted value in Fig\,\ref{fig:das_capon_beams}.
 %The difference between the predicted and the simulated loss might be caused by several factors.
@@
 \subsection{Oversampling on Transmit}
-A straight forward way of obtaining sufficient sampling in an active imaging system with an equal number of transmit (tx) and receive (rx) beams is to increasing the number of tx-beams until the system is locally laterally shift-invariant.
+A straight forward way of obtaining sufficient sampling in an active imaging system is by increasing the number of transmit beams (tx-beams) until the system is locally laterally shift-invariant.
 In Fig.\,\ref{fig:capon_oversampling_a} and Fig.\,\ref{fig:capon_oversampling_b} we have increased the tx-beam density by a $q$-factor of four and sixteen respectively.
@@
 Even though high frequency variations are reduced, the maximum is still larger than 1 dB.
-An even higher level of oversampling is therefore required in order to have a completely local lateral shift-invariant Capon beamformer when a moving point scatterer is simulated without noise and with the same parameters as in Fig.\,\ref{fig:capon}.
+An even higher level of oversampling is therefore required in order to have a completely local lateral shift-invariant Capon beamformer when a moving point scatterer is simulated without noise and with the same parameters as in Fig\,\ref{fig:capon}.
 %$1/100$ in diagonal loading.
@@
 }
-\caption{Capon beamforming as described in Fig.\,\ref{fig:capon} with oversampling on transmit.
-The figure is zoomed compared with Fig.\,\ref{fig:capon} to better see the beam-to-beam variation.
-a) Four times oversampling ($q=4$).
-b) Sixteen times oversampling ($q=16$).
+\caption{Capon beamforming as described in Fig\,\ref{fig:capon} with oversampling on transmit.
+The figure is zoomed compared with Fig\,\ref{fig:capon} to better see the beam-to-beam variation.
+a) Four times oversampling.
+b) Sixteen times oversampling.
 The maximum scalloping loss is reduced, but it is still too high for the system to be locally laterally shift invariant.}
@@
 synthetic transmit beamforming (STB) \cite{Hergum2007, Denarie2013}.
-In \cite{Rabinovich2013} it is shown that STB works well with Capon beamforming, something which is also verified in \cite{Asen}.
+In \cite{Rabinovich2013} it is shown that STB works well with Capon beamforming.
+In \cite{Asen} we also used STB in order to increase the beam density of a cardiac ultrasound recording.
 In the following discussion we consider corrected PRB as being equal to oversampling on transmit when it comes to image quality.
@@
 \begin{align}
-\theta_{max} = 2\alpha \delta\theta,
+\theta_{max} = \alpha p \Delta,
 \end{align}
-where $\alpha$ is the pulse length in $\lambda$'s, and $2$ and $\delta\theta$ is the oversampling factor and beam spacing from (\ref{eq:resolution}).
-For our simulations $\theta_{max} = 3\delta\theta$.
+where $\alpha$ is the pulse length in $\lambda$'s, and $p$ and $\Delta$ is the oversampling factor and beam spacing respectively as defined in (\ref{eq:resolution}).
+For our simulations $\theta_{max} = 3\Delta$.
 It is therefore possible to make a sweep based on phase rotation between a given beam and half-way to its two nearest neighbours without introducing large errors.
@@
 \end{align}
-where $x_i$ is the element position and $\theta$ is swept from $-\delta\theta/2$ to $\delta\theta/2$.
+where $x_i$ is the element position and $\theta$ is swept from $-\Delta/2$ to $\Delta/2$.
 Note that the steering vector must be calculated for an $L$ element symmetric subarray, even though the full array has $M$ elements.
-This subarray should have the same element pitch, $|x_i - x_{i-1}|$, as the full array.
-%The steering error introduced by this steering vector for a frequency band $B$ is less than $B/2 \times \delta\theta/2$.
+This subarray should have the same element pitch, $x_i - x_{i-1}$, as the full array.
+%The steering error introduced by this steering vector for a frequency band $B$ is less than $B/2 \times \Delta/2$.
 
@@
 
-Figure \ref{fig:capon_oversampling_pr} shows the phase rotation method applied on the same scan sequence as in Fig.\,\ref{fig:capon} with four and sixteen times oversampling on receive respectively ($q$-factor).
-Note that the figure is zoomed compared with Fig.\,\ref{fig:capon}.
+Figure \ref{fig:capon_oversampling_pr} shows the phase rotation method applied on the same scan sequence as in Fig\,\ref{fig:capon} with four and sixteen times oversampling respectively.
+Note that the figure is zoomed compared with Fig\,\ref{fig:capon}.
 As with oversampling on transmit there is some variation when the point moves in and out of focus.
@@
 The extreme local variations are however reduced significantly.
-Comparing Fig.\,\ref{fig:capon_oversampling} with Fig.\,\ref{fig:capon_oversampling_pr}, the errors introduced by the phase rotation method are best seen in sidelobes.
+Comparing Fig\,\ref{fig:capon_oversampling} with Fig\,\ref{fig:capon_oversampling_pr}, the errors introduced by the phase rotation method are best seen in sidelobes.
 
@@
 	}
-	\caption{Capon beamforming as described in Fig.\,\ref{fig:capon} with oversampling based on phase rotation.
-The figure is zoomed compared with Fig.\,\ref{fig:capon} to better see the beam-to-beam variation.
-a) Four times oversampling ($q=4$).
-b) Sixteen times oversampling ($q=16$).}
+	\caption{Capon beamforming as described in Fig\,\ref{fig:capon} with oversampling based on phase rotation.
+The figure is zoomed compared with Fig\,\ref{fig:capon} to better see the beam-to-beam variation.
+a) Four times oversampling.
+b) Sixteen times oversampling.}
 	\label{fig:capon_oversampling_pr}
@@
 In Fig.\,\ref{fig:LSV_invitro} we present the result of applying the Capon beamformer on \textit{in-vitro} data with and without oversampling by phase rotation.
-The data was acquired using a modified Vivid E9 high-end cardiac ultrasound scanner by GE Vingmed Ultrasound, Horten, Norway, equipped with a M5S-D phased transducer operating in harmonic mode at 3.34 MHz.
-The transmit beam density was set to match (\ref{eq:resolution}) based on this frequency, and 111 tx-beams where therefore distributed on a 65\degree sector.
-Transmit focus is located at 8 cm range.
+The data was acquired using a modified Vivid E9 high-end cardiac ultrasound scanner by GE Vingmed Ultrasound, Horten, Norway, equipped with a M5S transducer operating in harmonic mode with a receive frequency of 3.7 MHz.
+The transmit beam density was set to match (\ref{eq:resolution}) based on this frequency.
 Note that in harmonic imaging the generated transmit beam at the harmonic frequency will be wider and have lower sidelobes than the beam resulting from transmitting at the harmonic frequency \cite{Fedewa2004}.
@@
 A tissue mimicking phantom containing wire targets and anechoic cysts was imaged, and the data was acquired while the probe was moved across the phantom manually.
-Fig.\,\ref{fig:phantom_das_image} shows the first frame out of 14 in the acquired loop processed with DAS beamforming.
+Fig.\,\ref{fig:image_das} shows the first frame out of 14 in the acquired loop.
 %Movies of the whole loop processed with DAS and Capon beamforming with and without ten times oversampling by phase rotation can be found in the supplementary material \multimedia{Videos}.
 
-One point scatterer in Fig.\,\ref{fig:phantom_das_image} has been selected [white box], and then used to generate LSV-plots.
-The point scatterer is located close to the transmit focus.
-Figure \ref{fig:phantom_das_LSV} shows the resulting LSV-plot when using the DAS beamformer.
+One point scatterer in Fig\,\ref{fig:image_das} has been selected [white box], and then used to generate LSV-plots.
+Figure \ref{fig:phantom_das} shows the resulting LSV-plot when using the DAS beamformer.
 The scalloping loss is less than 1 dB, and the system is by that locally laterally shift-invariant.
@@
 
-Figure \ref{fig:phantom_capon_LSV_16PR} presents the result of applying sixteen times oversampling by phase rotation.
+Figure \ref{fig:phantom_capon_LSV_10PR} presents the result of applying sixteen times oversampling by phase rotation.
 We observe how the amplitude of the point scatterer is better preserved in all frames, and the scalloping loss is not far from our 1 dB threshold.
@@
 Looking at the video of Capon beamforming with sixteen times oversampling by phase rotation \multimedia{Media Movie 3} the blinking is now highly reduced.
-The same can be seen from Fig.\,\ref{fig:img_invitro} where point scatterers appear brighter in the image with oversampling by phase rotation.%The system is however not fully shift invariant.
+%The system is however not fully shift invariant.
 To achieve this a higher level of oversampling, %We therefore conclude that the system is now locally laterally shift invariant.
@@
 \begin{tabular}{cc}
-\subfloat[]{\includegraphics[width=0.45\linewidth]{\img fig6a.png}\label{fig:phantom_das_image}}
-\subfloat[]{\includegraphics[width=0.45\linewidth]{\img fig6b.eps}\label{fig:phantom_das_LSV}}\\
-\subfloat[]{\includegraphics[width=0.45\linewidth]{\img fig6c.eps}\label{fig:phantom_capon_LSV}}
-\subfloat[]{\includegraphics[width=0.45\linewidth]{\img fig6d.eps}\label{fig:phantom_capon_LSV_16PR}}
+\subfloat[]{\includegraphics[width=0.5\linewidth]{\img fig6a.eps}\label{fig:image_das}}
+\subfloat[]{\includegraphics[width=0.5\linewidth]{\img fig6b.eps}\label{fig:phantom_das}}\\
+\subfloat[]{\includegraphics[width=0.5\linewidth]{\img fig6c.eps}\label{fig:phantom_capon_LSV}}
+\subfloat[]{\includegraphics[width=0.5\linewidth]{\img fig6d.eps}\label{fig:phantom_capon_LSV_10PR}}
 \end{tabular}
@@
 \caption{LSV-plots generated from \textit{in-vitro} phantom data where the probe has been manually dragged across the phantom.
-See the supplementary material for videos of the whole loop processed with the different beamformers.
+See the supplementary material for videos of the whole loop processed with the different beamformers \multimedia{Videos}.
 Capon beamforming has been performed with the same settings as in Fig.\,\ref{fig:capon1}.
 P-MAE is not shown due to a lack of probe movement control.
-a) First frame in the recorded loop processed with delay-and-sum beamforming  \multimedia{Media Movie 1}.
+a) First frame in the recorded loop.
 A white box indicates the point scatterer used to construct the LSV-plots.
-b) LSV-plot for delay-and-sum beamforming.
-c) LSV-plot for Capon beamforming without oversampling.
-d) LSV-plot for Capon beamforming with sixteen times oversampling using phase rotation.}
+b) Delay-and-sum beamforming.
+c) Capon beamforming without oversampling.
+d) Capon with ten times oversampling using phase rotation.}
 \label{fig:LSV_invitro}
-\end{figure*}
-
-\begin{figure*}[!t]
-\centerline{
-\subfloat[]{\includegraphics[width=0.48\linewidth]{\img fig7a.png}\label{fig:phantom_capon_image}}
-\subfloat[]{\includegraphics[width=0.48\linewidth]{\img fig7b.png}\label{fig:phantom_capon_16PR_image}}
-}
-\caption{First frame from Fig.\,\ref{fig:phantom_das_image} processed with Capon beamforming.
-a) Without oversampling \multimedia{Media Movie 2}.
-b) With sixteen times oversampling using phase rotation \multimedia{Media Movie 3}.
-Notice how the point scatterers appear brighter in the image with oversampling.}
-\label{fig:img_invitro}
 \end{figure*}
@@
 In his seminal paper, Cox \cite{Cox1973} derives the required beam density for a given output SNR.
-For instance 47 dB requires a beam density for the Capon beamformer ten times the one given by (\ref{eq:resolution}) [with $q=1/2$].
+For instance 47 dB requires a beam density for the Capon beamformer ten times the one given by (\ref{eq:resolution}) [with $qp=1$].
 Since resolution for the Capon beamformer depends on SNR we would recommend to first use the dynamic range in the image as guidance.
 The amount of diagonal loading will also dictate the maximal achievable resolution since $10\log{d}$ can be interpreted as the maximal input SNR seen by the Capon beamformer.
-Fig.\,\ref{fig:das_capon_beams} shows that 25 times oversampling is required for a $d$-value of $1/100$.
+Fig\,\ref{fig:das_capon_beams} shows that 25 times oversampling is required for a $d$-value of $1/100$.
 When \textit{in-vitro} data contains additional noise this oversampling requirement is reduced.%Cox measures resolution as the 3 dB saddle point between two impinging monochromatic plane waves.
@@
 Oversampling by phase rotation can therefore introduce similar block artifacts as parallel receive beamforming (PRB).
-To remove these artifacts we could distribute our steering vectors in an overlapping scheme [like rx-beams are spread out for STB].
-We then get a phase rotation version of STB where e.g.
+These artifacts should be possible to remove in the same way as for PRB by using the synthetic transmit beamforming (STB) technique.
+If we distribute our steering angles in an overlapping scheme [like rx-beams are spread out for STB] we get a phase rotation version of STB where e.g.
 eight phase rotated beams are averaged into four.
-If the geometric distortion, or stripes, seen at the bottom of the sector in the video with sixteen times oversampling \multimedia{Media Movie 3} is removed by such technique is left for future work.
-
-When oversampling is applied it becomes evident that the high resolution settings used in this paper might be too extreme, both with respect to the level of required oversampling and the point scatterer's appearance.
+The geometric distortion, or stripes, seen at the bottom of the sector in the video with sixteen times oversampling \multimedia{Media Movie 3} should be remove by such technique.
+
+When oversampling is applied it becomes evident that the high resolution settings used in this paper might be to extreme, both with respect to the level of required oversampling and the point scatterer's appearance.
 Traditionally in medical ultrasound imaging the point spread function is always wider laterally than in range.
@@
 
-He has industry experience from GE Vingmed Ultrasound, Horten, Norway, were he completed three summer internships from 2008 to 2010.
+%He has industry experience from GE Vingmed Ultrasound, Horten, Norway, were he completed three summer internships from 2008 to 2010.
 His research interests include image and signal processing, adaptive ultrasound processing techniques and acceleration of ultrasound algorithms using Graphics Processing Units (GPUs).
-\end{IEEEbiography}
-
-\vfill{}
+%\end{IEEEbiography}
+
+%\vfill{}
 
@@
 
-\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{photos/andreas.eps}}]{Andreas Austeng}
-was born in Oslo, Norway, in 1970.
+%\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{photos/andreas.eps}}]{Andreas Austeng}
+%was born in Oslo, Norway, in 1970.
 He received the M.S.
@@
 His research interests include signal and array processing for acoustical imaging.
-\end{IEEEbiography}
-
-\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{photos/sverre.eps}}]{Sverre Holm}
-(M82-SM02) was born in Oslo, Norway, in 1954.
+%\end{IEEEbiography}
+
+%\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{photos/sverre.eps}}]{Sverre Holm}
+%(M’82-SM’02) was born in Oslo, Norway, in 1954.
 He received the M.S.
@@
 His research interests are medical ultrasound imaging, modeling of tissue-ultrasound interaction, and ultrasound positioning.
-\end{IEEEbiography}
-
-\vfill
+%\end{IEEEbiography}
+
+%\vfill
 

### 2014-04-23 ThesisJP/Introduction/introduction.tex

@@
 
-An interesting observation in \textbf{Paper\,I} can be found in Fig.\,10 \todo{Direct ref}.
+An interesting observation in \textbf{Paper\,I} can be found in Fig.\,\ref{code_assess_flops}.
 Here we see how the achieved throughput of the covariance estimation kernel is more than three times higher than the theoretical throughput of a high-end \nom{CPU}{Central processing unit} (The target CPU is from 2010).
@@
 The paper does also propose an interesting method for automatic tuning of rendering parameters.
-Figure 3d \todo{Direct ref} in \textbf{Paper\,IV} actually shows that an optimal threshold could be found based on global and not only local statistics.
-Even though the visualization based on global statistics has more errors, with respect to endocardium visibility, the tuning of the opacity function is still fully automatic if an iterative search for the minima in Fig.\,3d in \textbf{Paper\,IV} is performed.
+Figure \ref{fig:vis}d in \textbf{Paper\,IV} actually shows that an optimal threshold could be found based on global and not only local statistics.
+Even though the visualization based on global statistics has more errors, with respect to endocardium visibility, the tuning of the opacity function is still fully automatic if an iterative search for the minima in Fig.\,\ref{fig:vis}d is performed.
 A global opacity transfer function will insure a rendering which is realistic, but it will still be dependent on a successful tracking for the auto-adjustment to work.

### 2014-04-23 ThesisJP/Paper3/moving_obj.tex

@@
 Different methods for reduced scalloping loss and by that improved local lateral shift-invariance are presented in Section \ref{sec:methods}.
-Then, in Section \ref{sec:res} we present the results of applying the most promising method on \textit{in vitro} data.
-Finally we discuss our findings in Section \ref{sec:dis} and give our conclusions in Section \ref{sec:con}.
+Then, in Section \ref{sec:res_paper3} we present the results of applying the most promising method on \textit{in vitro} data.
+Finally we discuss our findings in Section \ref{sec:dis_paper3} and give our conclusions in Section \ref{sec:con_paper3}.
 
@@
 \centerline{
-\subfloat[]{\includegraphics[width=0.5\linewidth]{\img fig2a.eps}\label{fig:das_shiftinvariant}}
-\hfill{}
-\subfloat[]{\includegraphics[width=0.5\linewidth]{\img fig2b.eps}\label{fig:das_shiftvariant}}
+\begin{tabular}{c}
+\subfloat[]{\includegraphics[width=0.7\linewidth]{\img fig2a.eps}\label{fig:das_shiftinvariant}}\\
+\subfloat[]{\includegraphics[width=0.7\linewidth]{\img fig2b.eps}\label{fig:das_shiftvariant}}
+\end{tabular}
 }
@@
 \centerline{
+\begin{tabular}{c}
 \subfloat[]{
-\includegraphics[width=0.5\linewidth]{\img fig3a.eps}\label{fig:capon1}
+\includegraphics[width=0.7\linewidth]{\img fig3a.eps}\label{fig:capon1}
+}\\
+\subfloat[]{
+\includegraphics[width=0.7\linewidth]{\img fig3b.eps}\label{fig:capon2}
 }
-\subfloat[]{
-\includegraphics[width=0.5\linewidth]{\img fig3b.eps}\label{fig:capon2}
-}
+\end{tabular}
 }
@@
 \centerline{
+\begin{tabular}{c}
 \subfloat[]{
-\includegraphics[width=0.5\linewidth]{\img fig4a.eps}\label{fig:capon_oversampling_a}
+\includegraphics[width=0.7\linewidth]{\img fig4a.eps}\label{fig:capon_oversampling_a}
+}\\
+\subfloat[]{
+\includegraphics[width=0.7\linewidth]{\img fig4b.eps}\label{fig:capon_oversampling_b}
 }
-\subfloat[]{
-\includegraphics[width=0.5\linewidth]{\img fig4b.eps}\label{fig:capon_oversampling_b}
-}
+\end{tabular}
 }
@@
 	\centerline{
+		\begin{tabular}{c}
 		\subfloat[]{
-			\includegraphics[width=0.5\linewidth]{\img fig5a.eps}
+			\includegraphics[width=0.7\linewidth]{\img fig5a.eps}
+		}\\
+		\subfloat[]{
+			\includegraphics[width=0.7\linewidth]{\img fig5b.eps}
 		}
-		\subfloat[]{
-			\includegraphics[width=0.5\linewidth]{\img fig5b.eps}
-		}
+		\end{tabular}
 	}
@@
 
-\section{In Vitro Phantom Data}\label{sec:res}
+\section{In Vitro Phantom Data}\label{sec:res_paper3}
 In Fig.\,\ref{fig:LSV_invitro} we present the result of applying the Capon beamformer on \textit{in vitro} data with and without oversampling by phase rotation.
@@
 \centerline{
-\subfloat[]{\includegraphics[width=0.5\linewidth]{\img fig7a.png}\label{fig:phantom_capon_image}}
-\subfloat[]{\includegraphics[width=0.5\linewidth]{\img fig7b.png}\label{fig:phantom_capon_16PR_image}}
+\begin{tabular}{c}
+\subfloat[]{\includegraphics[width=0.7\linewidth]{\img fig7a.png}\label{fig:phantom_capon_image}}\\
+\subfloat[]{\includegraphics[width=0.7\linewidth]{\img fig7b.png}\label{fig:phantom_capon_16PR_image}}
+\end{tabular}
 }
@@
 
-\section{Discussion}\label{sec:dis}
+\section{Discussion}\label{sec:dis_paper3}
 %Note that in this paper we have investigated a single array configuration and a limited parameter range for the Capon beamformer.
@@
 
-\section{Conclusion}\label{sec:con}
+\section{Conclusion}\label{sec:con_paper3}
 %In this paper we have investigated the lateral shift-invariance property of the Capon beamformer when imaging moving objects.

### 2014-04-24 HowtoHowtoCapon/gfx/prepare.sh
...

@@ -1,13 +1,83 @@
 #!/bin/bash
 
-cp beamforming.eps buske1.eps
-cp implementation.eps buske2.eps
-cp mvdr_complexity.eps buske3.eps
-cp mvdr_build_R.eps buske4.eps
-cp buildR-breakdown.eps buske5.eps
-cp mvdr_implementation.eps buske6.eps
-cp plot_holmengraa_L16_Navg1.eps buske7.eps
-cp code_assessment.eps buske8.eps
-cp benchmark.eps buske9.eps
-cp code_assess_flops.eps buske10.eps
-cp benchmark_boston.eps buske11.eps
+FILES=" \
+   beamforming.svg \
+   implementation.svg \
+   mvdr_complexity.svg \
+   mvdr_build_R.svg \
+   buildR-breakdown.svg \
+   mvdr_implementation.svg \
+   plot_holmengraa_L16_Navg1.eps \
+   benchmark.eps \
+   code_assessment.eps \
+   code_assess_flops.eps \
+   benchmark_boston.eps \
+   "
+
+cp beamforming.eps buske1_online.eps
+cp implementation.eps buske2_online.eps
+
+cp mvdr_complexity_bw.eps buske3_bw.eps
+cp mvdr_complexity.eps buske3_online.eps
+
+cp mvdr_build_R_bw.eps buske4_bw.eps
+cp mvdr_build_R.eps buske4_online.eps
+
+cp buildR-breakdown_bw.eps buske5_bw.eps
+cp buildR-breakdown.eps buske5_online.eps
+
+cp mvdr_implementation_bw.eps buske6_bw.eps
+cp mvdr_implementation.eps buske6_online.eps
+
+cp plot_holmengraa_L16_Navg1.eps buske7_bw.eps
+cp plot_holmengraa_L16_Navg1.eps buske7_online.eps
+
+cp benchmark_bw.eps buske8_bw.eps
+cp benchmark.eps buske8_online.eps
+
+cp code_assessment_bw.eps buske9_bw.eps
+cp code_assessment.eps buske9_online.eps
+
+cp code_assess_flops_bw.eps buske10_bw.eps
+cp code_assess_flops.eps buske10_online.eps
+
+cp benchmark_boston_bw.eps buske11_bw.eps
+cp benchmark_boston.eps buske11_online.eps
+
+# Coverts eps file to black-white using ghostscript
+function makeEpsBW {
[…]

### 2014-04-24 HRUIofMovingObjects/response_to_reviewers.tex

@@
 \\\\
-\textit{We understand that this may be confusing and we have amended the text to make it more clear.
+\textit{We understand that this may be confusing and we have amended the text to make it more clear ($q$ is e.g.
+4 and 16 when oversampling is applied for Capon beamforming).
 The factor $p$ has been removed, and $q$ is now the oversampling factor.
@@
 In the in-vitro data, the point scatterer closest to the transmit focus is selected.
-We have amended the text to clearly state why the transmit focus have been selected as the point of observation.}
+We have amended the text to clearly state why the transmit focus has been selected as the point of observation.}
 \\\\

### 2014-04-24 HRUIofMovingObjects/jrnl_article.tex
Background. GPU programming.

@@
 
-\section{In-vitro phantom data}\label{sec:res}
-In Fig.\,\ref{fig:LSV_invitro} we present the result of applying the Capon beamformer on \textit{in-vitro} data with and without oversampling by phase rotation.
+\section{In vitro phantom data}\label{sec:res}
+In Fig.\,\ref{fig:LSV_invitro} we present the result of applying the Capon beamformer on \textit{in vitro} data with and without oversampling by phase rotation.
 The data was acquired using a modified Vivid E9 high-end cardiac ultrasound scanner by GE Vingmed Ultrasound, Horten, Norway, equipped with a M5S-D phased transducer operating in harmonic mode at 3.34 MHz.
@@
 }
-\caption{LSV-plots generated from \textit{in-vitro} phantom data where the probe has been manually dragged across the phantom.
+\caption{LSV-plots generated from \textit{in vitro} phantom data where the probe has been manually dragged across the phantom.
 See the supplementary material for videos of the whole loop processed with the different beamformers.
@@
 In simulations, this method reduced the maximum scalloping loss by 16 dB.
-On \textit{in-vitro} data the reduction was around 7 dB and the point scatterer blinking was highly reduced.
+On \textit{in vitro} data the reduction was around 7 dB and the point scatterer blinking was highly reduced.
 
@@
 
-The authors would like to thank Anders S\o{}rnes at GE Vingmed Ultrasound, Horten, Norway for the help with acquiring \textit{in-vitro} data.
+The authors would like to thank Anders S\o{}rnes at GE Vingmed Ultrasound, Horten, Norway for the help with acquiring \textit{in vitro} data.
 

### 2014-04-24 HowtoHowtoCapon/submission/final/article.tex

--- 2014-04-13 ThesisJP/Paper1/howto_howto_capon.tex
@@
-\newif\ifPeerReview\PeerReviewfalse             % Whether to create the PeerReview version or
+%%%%%%%%%%%%%%%%                                ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
+% CONDITIONALS %
+%%%%%%%%%%%%%%%%                                ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
+
+\newif\ifPeerReview\PeerReviewtrue              % Whether to create the PeerReview version or
                                                 % Journal version
+\newif\ifOnlineColor\OnlineColortrue            % Compile online color version?
+
 \newif\ifFlatArchive\FlatArchivefalse           % Whether archive is flat (messy) or contain 
@@
 \newif\ifFloatAtEnd\FloatAtEndfalse             % Available in PeerReview mode:
-
-\graphicspath{{./Paper1/}}
-
+                                                % Place floats at end of document?
+\newif\ifTODO\TODOtrue                        % Use todo notes?
+
+\ifOnlineColor
+   \newcommand\figPostfix{_online}              % Colored images online
+\else
+   \newcommand\figPostfix{_bw}                  % Black and white in journal
+\fi
+
+%%%%%%%%%%%%                                    ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
+% IEEEtran %
+%%%%%%%%%%%%                                    ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
+
+\ifPeerReview
+\documentclass[12pt,journal,draftclsnofoot,onecolumn]{IEEEtran}
+% \newcommand\CLASSINPUTbaselinestretch{1.66}     % http://theoval.cmp.uea.ac.uk/~nlct/latex/thesis/node17.html
+\else
+\documentclass[journal]{IEEEtran}
+\fi
+
+% \RequirePackage[latin1]{inputenc}%              % Set input encoding (optionally latin1)
+% \RequirePackage[T1]{fontenc}%                 % Set font encoding
+% \usepackage[norsk]{babel}
+
+% Single column review mode:
+%\documentclass[12pt,journal,onecolumn]{IEEEtran}
+%\newcommand\CLASSINPUTbaselinestretch{2}
+%\RequirePackage{calc}
+%\RequirePackage{fp} 
+% Left margin = 1 inch + hoffset + oddsidemargin (or evensidemargin)
+% Adding the 2cm gutter width to the odd/even side margins:
+%\setlength\hoffset{0pt}
+%\setlength\oddsidemargin{4cm}       % 4cm margins on the left side
+%\addtolength\oddsidemargin{-1in}    % Subtract the initial 1 inch
+%\setlength\textwidth{21cm-8cm}    % A4 width (21cm) minus margins on either side
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%                  ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
+% IEEE ''APPROVED'' PACKAGES %
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%                  ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
+
+\usepackage{cite}
+
+\ifCLASSINFOpdf
+   \usepackage[dvips]{graphicx}                 % Might not work. Use 'latex' instead of 
+   \ifFlatArchive\else                          % 'pdflatex'
+      \graphicspath{./gfx/}
+   \fi
+\else
+   \usepackage[dvips]{graphicx}
+   \ifFlatArchive\else
+      \graphicspath{./gfx/}
+   \fi
+\fi
+
+\RequirePackage[table,dvipsnames,svgnames]{xcolor}
+
+\usepackage[cmex10]{amsmath}                    % cmex10 option to be IEEE explore compliant
+\interdisplaylinepenalty=2500                   % Allows multiline equations to be broken
+
+% \RequirePackage{amssymb}
+
+\RequirePackage{array}
+
+\ifCLASSOPTIONcompsoc
+   \usepackage[caption=false,font=normalsize,labelfont=sf,textfont=sf]{subfig}
+\else
+   \usepackage[caption=false,font=footnotesize]{subfig}
+\fi
+\ifCLASSOPTIONcaptionsoff                       % IEEE promoted hack to turn off captions from the 
+   \let\MYorigsubfloat\subfloat                 % subfloat package should the captionsoff option
+   \renewcommand{\subfloat}[2][\relax]{\MYorigsubfloat[]{#2}} % be specified.
+\fi
+
+\ifFloatAtEnd
+\ifCLASSOPTIONcaptionsoff                       % Places float at the end of the document when the
+  \usepackage[nomarkers]{endfloat}              % captionsoff options is specified to IEEEtrans.cls
+  \let\MYoriglatexcaption\caption               % (PeerReview mode)
+  \renewcommand{\caption}[2][\relax]{\MYoriglatexcaption[#2]{#2}}
+\fi
+\fi
+
+\usepackage{fixltx2e}                           % Fix some twocolumn float problems
+
+%\usepackage{stfloats}                          % Allows: \begin{figure*}[!b]
+                                                % (double column figures on top/bottom)
+
+\usepackage{url}                                % Support for handling and breaking URLs
+
+% NOTE: PDF hyperlink and bookmark features are not required in IEEE
+%       papers and their use requires extra complexity and work.
+\newcommand\MYhyperrefoptions{bookmarks=true,bookmarksnumbered=true,
+pdfpagemode={UseOutlines},plainpages=false,pdfpagelabels=true,
+colorlinks=true,linkcolor={black},citecolor={black},urlcolor={black},
+pdftitle={An Optimized GPU Implementation of the MVDR Beamformer for Active Sonar Imaging},
+pdfsubject={},
+pdfauthor={Jo Inge Buskenes},
+pdfkeywords={adaptive beamforming, beamforming, complexity, sonar, active}}%
+\ifCLASSINFOpdf
+   \usepackage[\MYhyperrefoptions,pdftex]{hyperref}
+\else
+   \usepackage[\MYhyperrefoptions,breaklinks=true,dvips]{hyperref}
+   \usepackage{breakurl}                        % Allows 'dvips' driver to break links
+\fi
+
+%%%%%%%%%%%%%%%%%%%%%%%                         ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
+% ADDITIONAL PACKAGES %       
+%%%%%%%%%%%%%%%%%%%%%%%                         ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
+
+% \ifPeerReview
+% \let\OldIncludegraphics{\includegraphics}
+% \usepackage{letltxmacro}
+% \LetLtxMacro{\OldIncludegrsaphics}{\includegraphics}
+% \renewcommand{\includegraphics}[2][]{\OldIncludegraphics[width=\linewidth, #1]{#2}}
+% \fi
+
+% % \makeatletter
+% % \def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth
+% % \else\Gin@nat@width\fi}
+% % \makeatother
+% % \let\Oldincludegraphics\includegraphics
+% % \renewcommand{\includegraphics}[1]{\Oldincludegraphics[width=\maxwidth]{#1}}
+
+\usepackage[maxfloats=40]{morefloats}
+\newcounter{todoidx}
+% \setcounter{todoidx}
+
+\ifTODO
+   \definecolor{todobackground}{rgb}{0.95,0.95,0.95}
+   \setlength\marginparsep{1pt}
+   \setlength\marginparwidth{35pt}
+   \newlength\marginparwidthsmall
+   \setlength\marginparwidthsmall{\marginparwidth}
+   \addtolength\marginparwidthsmall{-7pt}
+   \newcommand\todo[1]{%
+      \addtocounter{todoidx}{1}%
+      {\color{Red}\bf(\thetodoidx{})}%%\fbox{\bf\thetodoidx{}}}%
+      \marginpar{%
+         {\vspace*{-10pt}\color{Red}\fbox{\bf\thetodoidx{}}}\\%
+         \fcolorbox{red}{todobackground}{\parbox{\marginparwidthsmall}{\scriptsize #1}}}}
+
+   \newcommand\todopar[1]{\fcolorbox{red}{white}{\parbox{0.97\linewidth}{#1}}}
+\else
+%    \usepackage[disable]{./todonotes} 
+   \newcommand\todo[1]{}
+\fi
+
+\newenvironment{narrow}[2]{%
+\begin{list}{}{%
+\setlength{\topsep}{0pt}%
+\setlength{\leftmargin}{#1}%
+\setlength{\rightmargin}{#2}%
+\setlength{\listparindent}{\parindent}%
+\setlength{\itemindent}{\parindent}%
+\setlength{\parsep}{\parskip}}%
+\item[]}{\end{list}}
+
+\usepackage{float}
+
+\ifOnlineColor
    \definecolor{tabBlue}{HTML}{AACCFF}
+\else
+   \definecolor{tabBlue}{HTML}{CCCCCC}
+\fi
 
@@
 
+% \DeclareMathOperator*{\argmin}{\text{arg}\;\text{min}}
+
 \newcommand\Fig[1]{Fig.~\ref{#1}}
+
+\newcommand\Grey[1]{{\color{Grey}#1}}
+\newcommand\Red[1]{{\color{Red}#1}}
+\newcommand\Blue[1]{{\color{Blue}#1}}
+\newcommand\DarkBlue[1]{{\color{DarkBlue}#1}}
+\newcommand\LightBlue[1]{{\color{LightBlue}#1}}
+\newcommand\Brown[1]{{\color{Brown}#1}}
+\newcommand\Green[1]{{\color{Green}#1}}
+\newcommand\SeaGreen[1]{{\color{SeaGreen}#1}}
+\newcommand\Yellow[1]{{\color{yellow}#1}}
+\newcommand\Orange[1]{{\color{orange}#1}}
 
@@
 %Math macros:
-\newcommand\diff[2]{\frac{\kw{d}\;\textstyle #1\scriptstyle}{\kw{d\parm{#2}}}\displaystyle}
-\newcommand\ddiff[2]{\frac{\kw{d^2}\;\displaystyle #1\scriptstyle}{\kw{d\parm{#2}}^2}\displaystyle}
-
-\renewcommand\d[1]{\scriptstyle\kw{\;d\parm{#1}}\displaystyle}
-
-% These commands are mutually exclusive. Remember to "renew" in v2.
-\newcommand\intb[4]{\int\limits_{#3}^{#4} #1 \d{#2}} % \int{exp}{var}{from}{to}
-\newcommand\intl[3]{\int\limits_{#3} #1 \d{#2}} % \int{exp}{var}{for all}
-\newcommand\intu[2]{\int #1 \d{#2}} % \int{exp}{var}{for all}
-
 \newcommand\T{^{\scriptscriptstyle T}}
@@
 \renewcommand\vec[1]{\boldsymbol{#1}}
-\renewcommand\mat[1]{\boldsymbol{#1}}
+\newcommand\mat[1]{\boldsymbol{#1}}
 
@@
 \newcommand*\E{\mat E}
-% \renewcommand*\H{\mat H}
 \renewcommand*\P{\mat P}
 \newcommand*\eP{\mat{\hat P}}
-\renewcommand*\R{\mat R}
+\newcommand*\R{\mat R}
 \newcommand*\Ri{\R^{-1}}
@@
 
-\newcommand\D{\vec\nabla} %Del: Vector differential operator - nabla
-\newcommand\Dx{\vec\nabla\times}
-\newcommand\Dd{\vec\nabla\cdot}
+\usepackage{tikz}
+\usetikzlibrary{shapes,snakes}
+\usepackage{amsmath,amssymb}
+% \usepackage{datatool}
+% \usepackage{glossaries}
 
@@
 
+%    \definecolor{todobackground}{rgb}{0.95,0.95,0.95}
+%    \setlength\marginparsep{3pt}
+%    \setlength\marginparwidth{42pt}
+%    \newlength\marginparwidthsmall
+%    \setlength\marginparwidthsmall{\marginparwidth}
+%    \addtolength\marginparwidthsmall{-7pt}
+%    \newcommand\todo[1]{%
+%       \addtocounter{todoidx}{1}%
+%       {\color{Red}\fbox{\bf\thetodoidx{}}}%
+%       \marginpar{%
+%          {\vspace*{-10pt}\color{Red}\fbox{\bf\thetodoidx{}}}\\%
+%          \fcolorbox{red}{todobackground}{\parbox{\marginparwidthsmall}{#1}}}}
+%
+
+% correct bad hyphenation here
+% \hyphenation{op-tical net-works semi-conduc-tor}
+
+%%%%%%%%%%%%                                    ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
+% GLOSSARY %
+%%%%%%%%%%%%                                    ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
+
+% \makeglossaries
+% \input{../../Library/glossary}
+
+%%%%%%%%%%%%%%%%%%                              ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
+% DOCUMENT START %
+%%%%%%%%%%%%%%%%%%                              ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
+
+% \usepackage{yfonts}
+
+\graphicspath{submission/final/gfx/}
+
+\begin{document}
+
+\title{An Optimized GPU Implementation of the MVDR Beamformer for Active Sonar Imaging}
+
+\author{Jo~Inge~Buskenes,~\IEEEmembership{Student Member,~IEEE,} %
+        Jon~Petter~\AA{}sen,~\IEEEmembership{Student Member,~IEEE,} %
+        Carl-Inge~Colombo~Nilsen,~\IEEEmembership{Member,~IEEE,} %
+        Andreas~Austeng,~\IEEEmembership{Member,~IEEE}%
+\thanks{J. I. Buskenes, C.-I. C. Nilsen and A. Austeng are with the Department of Informatics, University of Oslo, Norway.}%
+\thanks{J. P. \AA{}sen is with MI-Lab, Norwegian University of Science and Technology, Trondheim, Norway.}% <-this % stops a space
+% \thanks{Manuscript received April 19, 2005; revised January 11, 2007.}
+}
+
+% The paper headers
+\markboth{IEEE Journal of Oceanic Engineering}%
+{A Robust and Optimized GPU Implementation of the MVDR beamformer}
+
+% Publishers ID mark:
+%\IEEEpubid{0000--0000/00\$00.00~\copyright~2007 IEEE}
+
+% use for special paper notices
+%\IEEEspecialpapernotice{(Invited Paper)}
+
+% for Computer Society papers, we must declare the abstract and index terms
+% PRIOR to the title within the \IEEEcompsoctitleabstractindextext IEEEtran
+% command as these need to go into the title area created by \maketitle.
+\IEEEcompsoctitleabstractindextext{%
 \begin{abstract}
@@
 
+% Keywords (normally not used for peer reviews)
+\ifPeerReview\else
+\begin{IEEEkeywords}
+Adaptive beamforming, MVDR, Capon, sonar, active sonar, GPU, graphics processing unit, complexity.
+\end{IEEEkeywords}
+\fi}
+% \fi
+
+% make the title area
+\maketitle
+
+% This command fixes abstract positioning for compsoc articles:
+\IEEEdisplaynotcompsoctitleabstractindextext
+
+% (Optional) Add some extra info on cover page of peer review papers:
+% \ifCLASSOPTIONpeerreview
+% \begin{center} \bfseries EDICS Category: 3-BBND \end{center}
+% \fi
+
+% Insert page break and insert second title (peer review mode)
+\IEEEpeerreviewmaketitle
+
+\maketitle
+
 \section{Introduction}
@@
 
-Data driven methods have been introduced in various imaging fields over the years in attempts to improve image quality.
+\IEEEPARstart{D}{ata} driven methods have been introduced in various imaging fields over the years in attempts to improve image quality.
 One such method is the Minimum Variance Distortionless Response (MVDR) beamformer.
@@
 
-The implementation is adapted to a well sampled 100\;kHz wideband active sonar system operating in the near field.
-The system has a range resolution comparable to the sample period so our MVDR implementation runs near single snapshot mode.
-To deal with this and avoid signal cancellation we compute and average the covariance estimates from overlapping subarrays~\cite{Kailath1985}.
-This may resemble techniques such as "subarray MVDR"~\cite{Chapman1976}, but unlike that technique we are not using subarrays to reduce complexity but to make MVDR in an active system possible.
+Our implementation is adapted to a well sampled 100\;kHz wideband active sonar system operating in the near field.
+The system has a range resolution comparable to the sample period so our MVDR implementation runs in near single snapshot mode.
+To deal with this and to avoid signal cancellation we compute and average the covariance estimates from overlapping subarrays~\cite{Kailath1985}.
+This may resemble techniques such as "subarray MVDR"~\cite{Chapman1976}, but unlike that technique we are not using subarrays to reduce complexity but to make it feasible to use MVDR in an active system.
 In related work, we investigated using the same MVDR GPU implementation in cardiac ultrasound imaging~\cite{Asen2012,Asen2013}, and in active sonar imaging we have compared the performance of this implementation to comparable adaptive methods~\cite{Buskenes2013}.
@@
 
-%  Furthermore, by implementing it on a modern \gls{GPU} we have gained a speedup of around 2 orders of magnitude compared to an optimized single-thread implementation in C.
-
-% These devices have in recent years emerged as a flexible and highly potent parallel technology well suited for many DSP tasks, and are available off-the-shelf at reasonable prices\todo{skip this?}.
-
-% These devices have in recent years emerged as a flexible and highly potent parallel technology well suited for many DSP tasks, and are available off-the-shelf at reasonable prices.
-
 % Yay or nay? $\Rightarrow$ GPU ultrasound (\cite{So2011,Chen2011}). Complex data, all robustification techniques.
-
-% - Adaptive beamformers offers image quality edge, but are usually computationally complex.
-% - Who's tried before?
-% - We: Active sonar MVDR, GPU, high performance.
 
@@
 \begin{figure}[H]\centering
-\includegraphics[width=0.8\linewidth]{gfx/buske1.eps}
-\else
-\begin{figure}[!t]\centering
-\includegraphics[width=0.8\linewidth]{gfx/beamforming.eps}
+\includegraphics[width=0.8\linewidth]{gfx/buske1\figPostfix.eps}
+\else
+\begin{figure}[!t]\centering
+\includegraphics[width=\linewidth]{gfx/beamforming.eps}
 \fi%
@@
 Signal signature is first removed by matched filtering.
-Then - prior to summation - a suitable set of delays, $\Delta$, and weights, $w$, are applied to focus on a pixel of interest at angle and range ($\theta,n$).}\label{beamforming}
+Then - before summation - a suitable set of delays, $\Delta$, and weights, $w$, are applied to focus on a pixel of interest at angle and range ($\theta,n$).}\label{beamforming}
 \end{figure}
@@
 \begin{gather}
-\vec w[n] = \frac{\Ri[n]\1}{\1\T\Ri[n]\1},\label{weights}
+\vec w[n] = \frac{\Ri[n]\a}{\a\T\Ri[n]\a},\label{weights}
 \end{gather}
-where $\1$ is a row vector of ones that represents broadside steering, and $\R=E\{\x[n]\x\H[n]\} \in\mathbb{C}^{M,M}$ is the spatial covariance matrix for the full array.
-Since we pre-steer our data to every pixel in the image we will always look to broadside, but (\ref{weights}) can be easily generalized by substituting $\1$ with a suitable steering vector.
-Since $\R$ is unknown, we estimate it by computing a sample covariance matrix $\eR$.
+where $\a$ is a steering vector and $\R=E\{\x[n]\x\H[n]\} \in\mathbb{C}^{M,M}$ is the spatial covariance matrix for the full array.
+Since we pre-steer our data to every pixel in the image we simplify (\ref{weights}) by substituting $\a$ with a row vector $\1$ that represents broadside phase-steering.
+To estimate $\R$ we compute a sample covariance matrix $\eR$.
 In this computation we perform some degree of:
@@
 
-To form $\eR$ we will first compute an intermediate sample covariance matrix $\breve{\R}$, which we form using temporal and spatial averaging.
+We perform \emph{temporal} and \emph{spatial averaging} first and put the result in an intermediate sample covariance matrix $\breve{\R}$.
 To do this we need to segment our array into subarrays.
@@
 Note how subarray averaging led to a size reduction of $\eR$ from $\mathbb{C}^{M,M}$ to $\mathbb{C}^{L,L}$, and hence will produce an $L$-element weight set when substituted into (\ref{z}).
-This weight set is applied to all the subarrays, prior to computing the beamformer output as in (\ref{z}).
+This weight set is applied to all the subarrays, before computing the beamformer output as in (\ref{z}).
 Or, equivalently, we may apply the weight set to the sum of all the subarrays:
@@
 \begin{figure}[!t]\centering
-\includegraphics[width=.8\linewidth]{gfx/buske2.eps}
-\else
-\begin{figure}[!t]\centering
-\includegraphics[width=.9\linewidth]{gfx/implementation.eps}
+\includegraphics[width=.8\linewidth]{gfx/buske2\figPostfix.eps}
+\else
+\begin{figure}[!t]\centering
+\includegraphics[width=\linewidth]{gfx/implementation.eps}
 \fi%
@@
 \end{figure}
-% \begin{figure}[!t]\centering
-% \includegraphics[width=\linewidth]{gfx/algorithm_structure.eps}
-% \caption{MVDR beamforming.
-First a spatial covariance matrix is estimated from the delayed data (\ref{spatialR}-\ref{finalR}), then the weights are computed (\ref{weights}) and finally applied to the delayed channel data (\ref{z}).}
-% \label{implementation}
-% \end{figure}
 As summarized in Fig. \ref{mvdr_beamforming}, the MVDR method is applied to each pixel independently, by:
@@
 \end{enumerate}
-% This is illustrated in .
-% \begin{figure}[!t]\centering
-% \includegraphics[width=\linewidth]{gfx/buildR-complex_data-average.eps}
-% \caption{.}\label{complexity}
-% \end{figure}
 Next we will evaluate these steps in terms of arithmetic complexity, and then discuss their mappability to parallel hardware.
@@
 \begin{figure}[!t]\centering
-\includegraphics[width=.8\linewidth]{gfx/buske3.eps}
-\else
-\begin{figure}[!t]\centering
-\includegraphics[width=.9\linewidth]{gfx/mvdr_complexity.eps}
+\includegraphics[width=.8\linewidth]{gfx/buske3\figPostfix.eps}
+\else
+\begin{figure}[!t]\centering
+\includegraphics[width=\linewidth]{gfx/mvdr_complexity.eps}
 \fi%
-\caption{Per-pixel computational complexity of the steps in MVDR beamforming (prior to any optimizations).
+\caption{Per-pixel computational complexity of the steps in MVDR beamforming (before any optimizations).
 To avoid signal cancellation in an active sonar system we usually set $L<\frac{M}{2}$, in which region the computation of $\eR$ dominates in terms of arithmetic complexity, especially when performing temporal averaging.}\label{mvdr_complexity}
@@
 To visualize these relations, and include the effects of using complex numbers, we set up complexity formulas that account for the total number of arithmetic operations for each step in the MVDR process (see appendix \ref{mvdr_formulas}).
-We only excluded diagonal loading performed in the computation of $\eR$, and partial pivoting used in the inversion step, since these operations contribute marginally to the end result.
+The only excluded operation was partial pivoting used in the inversion step, but this should contribute marginally to the end result.
 The entire range of possible subarray sizes from $L\in[1,M]$ was finally evaluated, with temporal averaging set to $K\in\{0,1,2\}$, and the number of channels set to $M\in\{8,16,32\}$.
@@
 
-We note how the computation of $\eR$ completely dominates at smaller subarray sizes, that solving $\eRi\1$ only plays a notable role for larger array and subarray sizes, and that the computation of $z$ has a negligible impact on computation time.
+Note how the computation of $\eR$ completely dominates at smaller subarray sizes, that solving $\eRi\1$ only plays a notable role for larger array and subarray sizes, and that the computation of $z$ has a negligible impact on computation time.
 Also notice how temporal averaging comes with a high computational penalty.
@@
 
-% Not corrected for fused multiply accumulate
-
-% Also, the need for partial pivoting has been ignored, as we ensure a numerically well conditioned $\eR$ by applying diagonal loading.
-Since this is the most common scenario, we will have a thorough look at how it can be optimized.
-
-% and does not take into that does not  feasibility parallel decomposition  memory consumption.
-
 \section{Mapping the MVDR to a GPU}\label{maptogpu}
-
-% \begin{itemize}
-% \item Keep delay step on CPU, as this is a common step in all beamforming
-% \item Increasing matrix size reduces the number of threads per block, since the entire matrix must be in shared memory.
-1 matrix 72x72 complex (8 bytes) max per block - 72 threads.
-same as 207 5x5 matrixes, 5*207=1036 threads.
-max threads in block 512.
-% \end{itemize}
-
-% An important feature of MVDR beamforming, and with beamforming in general, is that each pixel can be computed independently and in an identical fashion.
-Considering that sonar images often end up having millions of pixels, this makes a compelling case for investigating technologies that support massive data level parallelism.
-The technology that does this best is perhaps Graphics Processing Units (GPUs).
-These chips contain up to thousands of small computation cores, which combined deliver superior peak floating point performance compared multi-core systems such as CPUs.
 
@@
 
-We decided to investigate the feasibility of running MVDR on a GPU by mapping it to a Nvidia GeForce Quadro 6000, which performance-wise is or par with or perhaps slightly inferior to the Geforce 480.
+We decided to investigate the feasibility of running MVDR on a GPU by mapping it to a Nvidia GeForce Quadro 6000, which performs similarly to the consumer grade Geforce GTX 480.
 This is a high-end Compute Unified Device Architecture (CUDA) enabled GPU based on Nvidia's Fermi architecture, which as now been superseded by the Kepler architecture.
@@
 \begin{figure}[t]\centering
-\includegraphics[width=.8\linewidth]{gfx/buske4.eps}
+\includegraphics[width=.8\linewidth]{gfx/buske4\figPostfix.eps}
 \else
 \begin{figure}[t]\centering
-\includegraphics[width=.9\linewidth]{gfx/mvdr_build_R.eps}
+\includegraphics[width=\linewidth]{gfx/mvdr_build_R.eps}
 \fi%
@@
 
-% \todo{Another shitty GPU transition}
-% Before proceeding we should make a quick note on the GPU memory.
-Note that to obtain the best possible performance, we need to make sure that CUDA threads find the data they need in cache or registers most of the time.
-Only these two memory types have a high enough bandwidth to promote sustained maximum utilisation of all CUDA cores.
-Unfortunately, they are a sparse resource.
-
-% This makes it apparent that computing one pixel per thread is not a feasible solution. \todo{...}
-
-% When optimizing an algorithm for speed, it is if of importance to know whether an algorithm is bound by computations or memory.
-One way to get an approximate idea of this is to compute the ratio of peak arithmetic throughput to peak memory throughput of the in the target platform, and compare this ratio to the ratio of arithmetical operations to memory operations in the algorithm.
-
-% follow collective access patterns to memory.
-
-% At first glance, it might seem like a good idea to compute one pixel per CUDA core.
-Each pixel is computed using the same instructions, and without any exchange of data.
-However, this approach is unfeasible due to the limited cache and register memory available to each thread.
-One could resolve to using the slower GDDR5 memory, but this would largely mitigate the performance gain and defeat the purpose of using a GPU in the first place.
-
-%Bandwidth becomes crucial, as getting data in to the cores and out again must happen i a frantic pace. %the feeding all these cores with data and harvesting the result
-
-% To effectively harness the computing potential of a GPU we need to make sure all computing cores stays busy most of the time.
-
 % Memory latency can be hidden in two ways.
@@
 While ILP can be very effective when done correctly, DLP is generally simpler to design for.
-% 
-% In the CUDA framework, the threads on an SM is managed through a software abstraction call a compute block.
-The block is a 3 dimensional grid where each element correspond to a thread that is scheduled to be executed on the SM at some point.
 
@@
 \begin{figure}[!t]\centering
-\includegraphics[width=.8\linewidth]{gfx/buske5.eps}
-\else
-\begin{figure}[!t]\centering
-\includegraphics[width=.9\linewidth]{gfx/buildR-breakdown.eps}
+\includegraphics[width=.8\linewidth]{gfx/buske5\figPostfix.eps}
+\else
+\begin{figure}[!t]\centering
+\includegraphics[width=\linewidth]{gfx/buildR-breakdown.eps}
 \fi%
@@
 Exploiting symmetry and performing iterative summing (step 1 and 3) is always desirable, but avoiding redundant multiplications (step 2) comes at the cost of increased memory consumption.
-When all optimizations are applied we call it a "minimum instructions" solution, and when step 2 is left out we call it a "minimum instructions" solution.
+When all optimizations are applied we call it a "minimum instructions" solution, and when step 2 is left out we call it a "minimum memory" solution.
 Fig.
@@
 
-% \begin{table}[b]\centering%\normalsize
-% \begin{tabular}[c]{l r@{:}  l}\hline
-% \rowcolor{tabBlue}\bf Memory type & \multicolumn{2}{>{\columncolor{tabBlue}}c}{\bf BW$_\text{memory}$/BW$_\text{arithmetic}$} \\\hline
-% Global memory &\hspace{30pt} 1 & 30 \\
-% Shared memory & 1 & 4 \\
-% Registers & $>$3 & 2~\cite{Vasilyy}
-% \end{tabular}
-% \caption{Nvidia Quadro 6000 bandwidth ratios.}\label{tabbandwidth}
-% \end{table}
-
-\begin{table}[b]\centering%\normalsize
-\begin{tabular}[c]{l r r r@{}  l}\hline
-\rowcolor{tabBlue} & \multicolumn{1}{>{\columncolor{tabBlue}}c}{\bf B$_\text{arith}$} & \multicolumn{1}{>{\columncolor{tabBlue}}c}{\bf B$_\text{mem}$} & \multicolumn{2}{>{\columncolor{tabBlue}}c}{\bf B$_\text{mem}$/B$_\text{arith}$} \\\hline
-Arithmetic & 1.03 Tflop/s & & &\\
-Global memory & & 36 Gfloats/s & \hspace{30pt} 1 &:30 \\
-Shared memory & & 257 Gfloats/s & 1 &:4 \\
-Registers & & $>$1.5 Tfloats/s & $>$3 &:2~\cite{Vasilyy}
-\end{tabular}
-\caption{Nvidia Quadro 6000: Memory throughput, $B_{\lowercase{\text{mem}}}$, compared to arithmetic throughput, $B_\text{arith}$.}\label{throughputs}
-\end{table}
-
 \ifPeerReview
 \begin{figure}[!t]\centering
-\includegraphics[width=.8\linewidth]{gfx/buske6.eps}
-\else
-\begin{figure}[!t]\centering
-\includegraphics[width=.9\linewidth]{gfx/mvdr_implementation.eps}
+\includegraphics[width=.8\linewidth]{gfx/buske6\figPostfix.eps}
+\else
+\begin{figure}[!t]\centering
+\includegraphics[width=\linewidth]{gfx/mvdr_implementation.eps}
 \fi%
@@
 Unfortunately, the GPU prefers kernels to be more arithmetically intensive than this.
-This can be inferred from Table \ref{throughputs}, where the peak bandwidth of the three types of GPU memory are compared to the peak arithmetic throughput (see appendix \ref{throughput} for derivations).
+This can be inferred from Table \ref{throughputs}, %
+%
+\begin{table}[!b]\centering%\normalsize
+\begin{tabular}[c]{l r r r@{}  l}\hline
+\rowcolor{tabBlue} & \multicolumn{1}{>{\columncolor{tabBlue}}c}{\bf B$_\text{arith}$} & \multicolumn{1}{>{\columncolor{tabBlue}}c}{\bf B$_\text{mem}$} & \multicolumn{2}{>{\columncolor{tabBlue}}c}{\bf B$_\text{mem}$/B$_\text{arith}$} \\\hline
+Arithmetic & 1.03 Tflop/s & & &\\
+Global memory & & 36 Gfloats/s & \hspace{30pt} 1 &:30 \\
+Shared memory & & 257 Gfloats/s & 1 &:4 \\
+Registers & & $>$1.5 Tfloats/s & $>$3 &:2~\cite{Vasilyy}
+\end{tabular}
+\caption{Nvidia Quadro 6000: Memory throughput, $B_{\lowercase{\text{mem}}}$, compared to arithmetic throughput, $B_\text{arith}$.}\label{throughputs}
+\end{table}%
+%
+where the peak bandwidth of the three types of GPU memory are compared to the peak arithmetic throughput (see appendix \ref{throughput} for derivations).
 Global memory (RAM), in which all data must reside at some point, is at best only able to move one float for every 30 floats processed by the CUDA cores, and potentially much worse if care is not taken to use "coalesced" memory access patterns (that avoid memory bank conflicts).
@@
 
-% What about the GPU implementation?
-This one was tricky.
-Balance load, avoid inter-thread dependencies when performing iterative summations, keep thread memory consumption at a minimum, and promote collective access patterns (coalesced reads/writes) to global memory.
-
 In Fig.
@@
 This lets us perform temporal averaging without having to read additional channel data.
-
-\ifPeerReview\else
-\begin{figure}[!t]\centering
-\includegraphics[width=.9\linewidth]{gfx/plot_holmengraa_L16_Navg1.eps}
-\caption{HISAS sidescan sonar (SSS) image of the shipwreck Holmengraa that lies on a slanted seabed at 77\;m depth outside of Horten, Norway.
-The image was processed with $M=32$, $L=16$, $K=1$ and $d=1\%$.
-Prior to display the image was linearly upinterpolated by a factor 2 in azimuth making its total size 1.46 megapixels (MP).
-Note how MVDR improves edge definition and reduces noise in shadow regions.}\label{holmengraa}
-\end{figure}
-\fi%
 
@@
 Since $\eR$ is conjugate symmetric, the values in the leftmost column is obtained by a complex conjugate copy of the relevant value in the first row.
-This is slightly non-optimal as it causes divergent branching, but is only needed because the solver is Gauss Jordan based and thus require a
-full $\eR$. Combined these steps balance the load evenly on all threads, is almost completely free of arithmetic redundancy, and consumes less memory.
-
-% Hence, whenever $N_\text{avg}>1$ the sliding window approach will demand less instructions at the expense of a slightly higher memory consumption.
-
-% However, an advantage with this approach is that the summations can be carried out in any order, and hence this form exhibits great deal of parallelism.
-Can we optimize the algorithm without sacrificing parallelism?
-
-% - Order of summations - beamspace capable
-
-% 
-% In the upcoming results, we have beamformed pre-delayed data.
-This is a large dataset, and the latency experienced when copying it from the \gls{CPU} side to the \gls{GPU} was significant.
-However, in a practical system only the channel data should be transferred as the delaying of data can effortlessly be performed by the \gls{GPU}.
-The remaining latency can be hidden as long as the \gls{GPU} is kept busy, hence the data transfer times will not be reported in the upcoming results.
-%
-
-% \begin{itemize}
-% \item Arrange order of summations to reduce data dimensions as early as possible.
-% \item Sliding window in time/space to reduce instructions at the cost of extra memory consumption. Only pays off to do this over subarrays.
-% \end{itemize}
-% 
-% \item Between GPU and CPU: Several concurrent blocks (contexts) per SM.
-% \item Job queue (how does this work)
-% \item Register pressure and shared memory. Thread context must fit in local memories.
-% \item Warp computational intensity should be high to make up for the memory cost.
-
-% increasing performance hit several walls:
-% \begin{itemize}
-% \item memory couldnt keep track - remedy cache.
-% \item instruction parallelism - complex to identity
-% \item power consumption cubic with frequency
-% GPUs:
-% \begin{itemize}
-% \item lower frequency, less power
-% \item no instruction parallelism
-% \item massive amounts of lightweight cores
-% \end{itemize}\end{itemize}
+This is slightly non-optimal as it causes divergent branching, but is needed because the solver is Gauss Jordan based and thus require a full $\eR$.
+Combined these steps balance the load evenly on all threads, is almost completely free of arithmetic redundancy, and consumes less memory.
 
@@
 However, one can also solve the linear equation $\hat{\boldsymbol{R}}\boldsymbol{b} = \1$ for $\boldsymbol{b}$, where $\boldsymbol{b} = \eRi\1$.
-This is the preferred approach since one vector In terms of arithmetic count the latter is preferred , and when comparing direct implementations, the latter appears to be the preferred option.
+This is the preferred approach since the solution is obtained directly and without the added computational cost of the final inner product.
 
@@
 The only downside to using it in our application is that it does not exploit the Hermitian property of our sample covariance matrix.
-A better choice might be a solver based on Cholesky decomposition.
-These are designed for Hermitian positive semi-definite matrices, and require roughly half the arithmetic operations compared to a Gauss Jordan solver.
-
-% \emph{Computing} $\eRi[n]\1$ (from (\ref{weights})).
-While intuition may suggest that this step is carried out by inverting $\eR$, it is better to solve the linear equation $\R\beta = \1$ for $\beta$ instead, which gives us $\beta = \eRi\1$ directly.
-We have tested various solvers for this task, both in-house and proprietary implementations, and achieved the best performance by using an unofficial batch solver from nVidia.
-Inverting $\eR$ is by far the most computationally intensive task, and a key area of focus for further improvements.
-
-\ifPeerReview\else
-\begin{figure}[!t]\centering
-\includegraphics[width=.9\linewidth]{gfx/code_assessment.eps}
-\caption{Execution time of an arithmetic-only and a memory-only version of the MVDR code.
-A dataset from an $M=32$ array was processed for all $L$ using $K=1$, and the mean execution time for a 1\;MP image was used here.
-From this plot we can infer that the kernel building $\eR$ is memory bound, as the time the kernel spends performing memory transactions is higher than the corresponding time it spends carrying out arithmetical operations.
-Furthermore, when the total runtime is larger than the restricted kernels this can largely be attributed to latency, which we can see that building $\eR$ suffers from with the chosen parameters.}\label{code_assessment}
-\end{figure}
-\fi%
+A better choice would be a solver based on Cholesky decomposition.
+These are designed for Hermitian positive semi-definite matrices such as our covariance matrix, and require only half the arithmetic operations of a Gauss Jordan solver.
 
@@
 
-\subsection{Summary of GPU implementation}
-
-The GPU implementation can be summarized as follows:
-\begin{itemize}
-\item Pixels are processed in batches.
-\item Multiple threads per pixel.
-\item $\eR$ is computed directly from raw data to minimize memory footprint.
-\item Inversion step is performed by Nvidias batch linear equation solver.
-\item Design makes efficient use of cache and data transfers are kept to a minimum.
-\end{itemize}
-
-Compared to our CPU implementations, where:
-\begin{itemize}
-\item Pixels are processed one by one.
-\item Single thread per pixel.
-\item Computation of $\eR$ is optimized for minumum number of instructions, as in \ref{mvdr_build_R}. Only upper half of $\eR$ is computed (see next point).
-\item Inversion step carried out by a custom inline Cholesky based solver that only requires the upper triangular of $eR$.
-\item Design has generally not been optimized for efficient use of memory.
-\end{itemize}
-
-%  The subarray data can, as here, be reduced to coincide in length with the subarray weights, or the weights can be extended to $M$ in size and applied directly on $\x[n]$ and summed.
-
-% \item \emph{Computing the spatial covariance matrix} $\eR[n]$ (\ref{spatialR}-\ref{finalR}).
-A group of threads are created that slide along the diagonals of $\eR$.
-In this way we exploited the fact that entries on the diagonals overlap across subarrays and time, and keep the numbers of both data reads and writes at a minimum.
+\subsection{Implementation summary}
+
+% The GPU implementation can be summarized as follows:
+% \begin{itemize}
+% \item Pixels are processed in batches.
+% \item Multiple threads per pixel.
+% \item $\eR$ is computed directly from raw data to minimize memory footprint.
+% \item Inversion step is performed by Nvidias batch linear equation solver.
+% \item Design makes efficient use of cache and data transfers are kept to a minimum.
+% \end{itemize}
+% 
+% We compare this to our CPU implementations, where:
+% \begin{itemize}
+% \item Pixels are processed one by one.
+% \item Single thread per pixel.
+% \item Computation of $\eR$ is optimized for minumum number of instructions, as in \ref{mvdr_build_R}. Only upper half of $\eR$ is computed (see next point).
+% \item Inversion step carried out by a custom inline Cholesky based solver that only requires the upper triangular of $\eR$.
+% \item Design has generally not been optimized for efficient use of memory.
+% \end{itemize}
+
+The GPU design is summarized and compared to our CPU designs in table \ref{implementation_summary}.
+\begin{table}[!h]\centering%\normalsize
+% \begin{tabular}[c]{l r r r@{}  l}\hline
+\begin{tabular}[l]{l >{\raggedright}p{.27\linewidth} p{.28\linewidth}}\hline
+\rowcolor{tabBlue} & \bf GPU & \bf CPU \\\hline
+Pixel processing & In batches & One by one \\
+Threads per pixel & Multiple & One \\
+1. Building $\eR$:  & & \\
+\ \ \ Values computed & All & Only upper triangular \\
+\ \ \ Optimization    & Minimized memory consumption & Minimized number of instructions \\
+2. Computing $\eRi\1$: & & \\
+\ \ \ Method & Nvidia's batch\newline Gauss Jordan solver & Custom inline\newline Cholesky solver \\
+Cache utilization & Efficient & Not efficient \\
+Data transfers & Minimum & Moderate
+\end{tabular}
+\caption{Implementation summary.}\label{implementation_summary}
+\end{table}
 
@@
 
-% In short, designing for a GPU involves finding the optimal balance between hiding memory latency (high occupancy) and resource utilization.
-% 
-% blocks in grid - high enough so that all SMs have atleast one block to execute.
-Optimally in the thousands to scale for future devices, and to allow an SM to switch between blocks to keep hardware busy when running synchronizing threads.
-% 
-% threads in block - min 64. 128-256 better starting point.
-
-% Instead of having a few complex core that can boost massive single thread performance, designers chose to create hundreds of lightweight cores that each is much powerful than a single CPU core, but combined boasts a massive artithmetical performance.
-% 
-% \begin{itemize}
-% \item Lots of cores, little memory for each.
-% \item Few data transfers, lots of computations.
-% \item Data transfers are moved around by hardware that can operate asynchronously from the compute units.
-% \item Memory latency hidden as long as it takes longer to compute than to move data around.
-% \item Tasks can be queued on SM (blocks) and on CPU to hide internal and external memory latencies.
-% \end{itemize}
-% 
-% Vast amount of processing power, which are easily accessible but hard to fully harness. This is because in the pursuit of an ever increasing amount of arithmetical computing cores, the 
-% 
-% Unlike a CPU, a GPU 
-% 
-% If we compare a GPU with a CPU, 
-% 
-% When implementing an algorithm on a GPU, it is important to be aware
-% 
-% People friendly:
-% \begin{itemize}
-% \item More and more compute units, but less memory for each.
-% \item Need to keep each compute unit busy.
-% \item Few data transfers, lots of computations.
-% \item Data transfers are moved around by hardware that can operate asynchronously from the compute units.
-% \item Memory latency hidden as long as it takes longer to compute than to move data around.
-% \item Tasks can be queued on SM (blocks) and on CPU to hide internal and external memory latencies.
-% \end{itemize}
-% 
-% The ``fast'' memory of the GPU is the registers and cache. All threads executed 
-% 
-% 
-% \begin{itemize}
-% \item Hiding memory access.
-% \begin{itemize}
-% \item Between GPU and CPU: Several concurrent blocks (contexts) per SM.
-% \item Job queue (how does this work)
-% \end{itemize}
-% \item Register pressure and shared memory. Thread context must fit in local memories.
-% \item Warp computational intensity should be high to make up for the memory cost.
-% \item Optimized for small L, high M, hardly any time averaging. A typical sonar scenario.
-% \end{itemize}
-% - Split program into equally sized portions - load balancing
-% - Some portions depend on other potions already being carried outer - sequential dependencies
-% - Effort timed in order to finish at the same time - synchronization
-% 
-% - task level parallelism - compute images independent of each other
-% - data level parallelism - compute pixels independent of each other
-
-% - Compute bound vs. bandwidth bound
-
 \section{Images and Benchmarks}\label{images_and_benchmarks}
 
+To demonstrate the imaging capability of the MVDR beamformer, we have processed experimental datasets from the $M=32$ element Kongsberg Maritime HISAS1030 sonar mounted on a HUGIN AUV~\cite{Hansen2009}.
+HISAS1030 is a high resolution synthetic aperture sonar with an array length of 1.2\;m, operating frequency of 100\;kHz, and bandwidth of 30\;kHz.
+The element size and spacing is 2.5$\lambda$ and the opening angle is 25$^\circ$.
+To produce the image shown in \Fig{holmengraa} the sonar was operated in sidescan mode.
+The studied object is the 1500 dwt oil tanker wreck Holmengraa.
+It is 68\;m long and 9\;m wide, and lies at a slanted seabed at 77\;m depth outside of Horten, Norway.
+The 1 megapixel (MP) MVDR image were here processed with parameters $L=16$, $K=1$, and $d=1\%$.
+
 \ifPeerReview
 \begin{figure}[!t]\centering
-\includegraphics[width=.8\linewidth]{gfx/buske7.eps}
+\includegraphics[width=.8\linewidth]{gfx/buske7\figPostfix.eps}
+\else
+\begin{figure}[!t]\centering
+\includegraphics[width=\linewidth]{gfx/plot_holmengraa_L16_Navg1.eps}
+\fi%
 \caption{HISAS sidescan sonar (SSS) image of the shipwreck Holmengraa that lies on a slanted seabed at 77\;m depth outside of Horten, Norway.
 The image was processed with $M=32$, $L=16$, $K=1$ and $d=1\%$.
-Prior to display the image was linearly upinterpolated by a factor 2 in azimuth making its total size 1.46\;MP.
+The image was linearly upinterpolated by a factor 2 in azimuth making its total size 1.46 megapixels (MP).
 Note how MVDR improves edge definition and reduces noise in shadow regions.}\label{holmengraa}
 \end{figure}
-\fi%
+
+The computational performance of our implementation was first assessed on a test system with a quad-core Intel Xeon E5620, 64\;GB of RAM and an Nvidia Quadro 6000 card.
+The results were obtained by processing a 1\;MP image from the data from a 32 channel array, for all subarray sizes $L$, and for $K\in\{0,1,2\}$.
+Run-time measurements are shown in \Fig{benchmarks}, the run-times of memory-only and arithmetic-only GPU kernels are depicted in \Fig{code_assessment}, and an estimate of computation efficiency is presented in \Fig{code_assess_flops}.
+We were also granted a test drive at the Boston HPC centre on a machine with an Intel Xeon E5-2670, 32\;GB RAM and an Nvidia K20.
+The results from this run are shown in \Fig{benchmarks_boston}.
 
@@
 \begin{figure}[!t]\centering
-\includegraphics[width=.8\linewidth]{gfx/buske8.eps}
-\else
-\begin{figure}[!t]\centering
-\includegraphics[width=.9\linewidth]{gfx/benchmark.eps}
+\includegraphics[width=.8\linewidth]{gfx/buske8\figPostfix.eps}
+\else
+\begin{figure}[!t]\centering
+\includegraphics[width=\linewidth]{gfx/benchmark.eps}
 \fi%
@@
 \begin{figure}[!t]\centering
-\includegraphics[width=.8\linewidth]{gfx/buske9.eps}
+\includegraphics[width=.8\linewidth]{gfx/buske9\figPostfix.eps}
+\else
+\begin{figure}[!t]\centering
+\includegraphics[width=\linewidth]{gfx/code_assessment.eps}
+\fi%
 \caption{Execution time of an arithmetic-only and a memory-only version of the MVDR code.
@@
 \end{figure}
-\fi%
 
@@
 \begin{figure}[!t]\centering
-\includegraphics[width=.8\linewidth]{gfx/buske10.eps}
-\else
-\begin{figure}[!t]\centering
-\includegraphics[width=.9\linewidth]{gfx/code_assess_flops.eps}
+\includegraphics[width=.8\linewidth]{gfx/buske10\figPostfix.eps}
+\else
+\begin{figure}[!t]\centering
+\includegraphics[width=\linewidth]{gfx/code_assess_flops.eps}
 \fi
@@
 \begin{figure}[!t]\centering
-\includegraphics[width=.8\linewidth]{gfx/buske11.eps}
-\else
-\begin{figure}[!t]\centering
-\includegraphics[width=.9\linewidth]{gfx/benchmark_boston.eps}
+\includegraphics[width=.8\linewidth]{gfx/buske11\figPostfix.eps}
+\else
+\begin{figure}[!t]\centering
+\includegraphics[width=\linewidth]{gfx/benchmark_boston.eps}
 \fi
@@
 
-% In a practical system it should be moved to the GPU as well. Therefore, in the upcoming benchmarks, CPU computation time and data transfer to the GPU is neglected.
-
-To demonstrate the imaging capability of the MVDR beamformer, we have processed experimental datasets from the $M=32$ element Kongsberg Maritime HISAS1030 sonar mounted on a HUGIN AUV~\cite{Hansen2009}.
-HISAS1030 is a high resolution synthetic aperture sonar with an array length of 1.2\;m, operating frequency of 100\;kHz, and bandwidth of 30\;kHz.
-The element size and spacing is 2.5$\lambda$ and the opening angle is 25$^\circ$.
-To produce the image shown in \Fig{holmengraa} the sonar was operated in sidescan mode.
-The studied object is the 1500 dwt oil tanker wreck Holmengraa.
-It is 68\;m long and 9\;m wide, and lies at a slanted seabed at 77\;m depth outside of Horten, Norway.
-The 1\;MP MVDR image were here processed with parameters $L=16$, $K=1$, and $d=1\%$.
-
-The computational performance of our implementation was first assessed on a test system with a quad-core Intel Xeon E5620, 64\;GB of RAM and an Nvidia Quadro 6000 card.
-The results were obtained by processing a 1\;MP image from the data from a 32 channel array, for all subarray sizes $L$, and for $K\in\{0,1,2\}$.
-Run-time measurements are shown in \Fig{benchmarks}, the run-times of memory-only and arithmetic-only GPU kernels are depicted in \Fig{code_assessment}, and an estimate of computation efficiency is presented in \Fig{code_assess_flops}.
-We were also granted a test drive at the Boston HPC centre on a machine with an Intel Xeon E5-2670, 32\;GB RAM and an Nvidia K20.
-The results from this run are shown in \Fig{benchmarks_boston}.
-
-As seen in the run-time comparisons presented in \Fig{benchmarks}, the GPU method is now 2-3 orders of magnitude faster than the C implemtation we started out with.
-The bottleneck in the final design is now the inversion step, which is typically 5 times slower than the build step.
+As seen in the run-time comparisons presented in \Fig{benchmarks}, the GPU method became 2-3 orders of magnitude faster than the C implemtation we started out with.
+The remaining bottleneck in the final design is the inversion step, which is typically 5 times slower than the build step.
 In most cases the processing speed of the Quadro 6000 is above 1\;MP/s, and this was only improved by a factor 1.5-2 when run on the new K20 GPU.
-The benchmarks of our memory-only and arithmetic-only kernels show that the kernels spend roughly the same time on both these tasks, so optimising only one of these further will have marginal effect.
+The benchmarks of our memory-only and arithmetic-only kernels show that the kernels spend roughly the same time on both these tasks, so optimising only one of these further will have marginal impact.
 All GPU kernels were compiled with \texttt{nvcc} at optimization level \texttt{O2}.
 Excluded from these benchmarks is the data transfer time from CPU to GPU, which account for 2-20\% of the total processing time.
-We believe it makes little sense to include them since it keeps getting easier perform these data transfers in parallel by offloading the task to the Direct Memory Access (DMA)
-controllers present on modern GPUs. Furthermore, the data rates in active sonar are relatively low compared to the bandwidth available for these transfers. 
-% 
-% We have tested our \gls{GPU} implementation of the \gls{MVDR} beamformer experimental datasets from the 32 element Kongsberg Maritime HISAS1030 sonar\todo{shouldn't we state that it's SAS?} being attached to the HUGIN \gls{AUV}~\cite{Hansen2009}.
-To obtain the studied object, the 1500 dwt oil tanker wreck Holmengraa, the sonar was operated in sidescan mode to image (\Fig{hlmengraa}).
-Holmengraa is roughly 68\;m long and 9\;m wide, and lies on a slanted seabed at 77\;m depth.
-The image was processed with \gls{MVDR} parameters $L=16$, $N_\text{avg}=3$, and $d=1\%$, which proved to be a reasonable selection for this scenario.
-%
+We believe it makes little sense to include them since it keeps getting easier to perform these data transfers in parallel by offloading the task to the Direct Memory Access (DMA) controllers present on modern GPUs.
+Furthermore, the data rates in active sonar are relatively low compared to the bandwidth available for these transfers.
 
@@
 Similar performance was obtained with arbitrary combinations of $L\in\{12,16,24\}$, $K\in\{1,2,3\}$ and $d\in[0.01,0.05]$, and no adjustments had to be made when processing other parts of the scene.
-%But while \gls{DAS} processed this 1.3\;Mpixel image in milliseconds, our MATLAB implementation\todo{never mentioned matlab vefore} of MVDR beamformer needed minutes.
 
@@
 In other aspects the designs are similar, both CPU implementations compute $\eR$ efficiently, and they make use of either an optimized custom Cholesky based solver or one from the Intel Math Kernel Library (MKL).
-Unfortunately we had little time to write a custom batch based solver and covariance builder for the CPU, but we expect
-the speed difference would be 5-10 times if we had compared two such designs.
+Unfortunately we had little time to write a custom batch based solver and covariance builder for the CPU, but we expect the speed difference to be 5-10 times if we had compared two such designs.
 
@@
 In particular, the back substitution step involved in its computation is mostly sequential.
-This difference be can observed in \Fig{mvdr_complexity}.
-Alternative solvers, such as one based on Cholesky decomposition that exploits
-the Hermitian property of $\eR$, can in theory reduce complexity by a factor two, but we question whether this result can be obtained in practice using a GPU.
+This difference can be observed in \Fig{mvdr_complexity}.
+Alternative solvers, such as one based on Cholesky decomposition that exploits the Hermitian property of $\eR$, can in theory reduce complexity by a factor two, but we question whether this result can be obtained in practice using a GPU.
 The CPU implementations perform Cholesky based inversion, but this does not explain why the C and MATLAB implementation have a total run-time that is linearly dependent on $L$.
-We believe this is because these implementations are dominated by the calculation of the covariance matrix and data movement, and not by the inversion step.
-
-It is important to know whether a GPU design is limited by memory bandwidth or arithmetical throughput.
+We believe this happens because these implementations are dominated by the calculation of the covariance matrix and data movement, and not by the inversion step.
+
+When optimizing a GPU design it is important to know whether it is limited by memory bandwidth or arithmetical throughput.
 To measure this we made two versions of the MVDR kernel, one that only performs arithmetic operations, and one that only performs memory operations.
@@
 Since the GPU hardware can carry out memory and arithmetic operations concurrently and independently, this gap indicates that the GPU sometimes does ``nothing''.
-In our case this is due to synchronization hold-
-ups when performing temporal averaging, which is carried out in a sequential manner.
-
-Even with all our efforts we were only able to obtain processing rates of 40\;MFlop/s on average in the desirable range of subarray sizes (\Fig{mvdr_complexity}).
-This is mainly due to the inversion step, since we for the most part obtained more than 100\;MFlop/s when building $\eR$.
-While these numbers are likely slightly underestimated, they are still far away from the Quadro 6000's peak of 1\;Flop/s.
-Again, we believe this is due to memory bandwidth.
+In our case this is due to synchronization hold-ups when performing temporal averaging, which is carried out in a sequential manner.
+
+Even with all our efforts we were only able to obtain processing rates of 40\;GFlop/s on average in the desirable range of subarray sizes (\Fig{mvdr_complexity}).
+This is mainly due to the inversion step, since we for the most part obtained more than 100\;GFlop/s when building $\eR$.
+While these numbers are slightly underestimated, they are regardless far away from the Quadro 6000's peak of 1\;TFlop/s.
+Again, we believe this is due to the design being bound by memory bandwidth.
 This belief is further supported by the test results from running the code on the new K20 Kepler card, where we saw a modest factor 1.5-2 speedup.
-Although the Kepler card peaks at 3.5-4\;Tflop/s, its shared memory bandwidth is only approximately twice that of the Quadro 6000.
+Although the Kepler card peaks at 3.5-4\;Tflop/s, its shared memory bandwidth is only approximately twice that of the Quadro 6000, hence matches the observed speed-up.
 However, it is likely that the Kepler card would perform better had we optimized our design for it.
-For scientific computing the Kepler boards are considered more attractive, with the introduction of features such as kernels calling kernels, more registers, more shared memory, and
-better mechanisms for hiding or removing data transfers.
-
-% Interesting to note is also that solving $\eRi\1$ remains the main bottleneck in the GPU implementation, and it gets even worse for larger $L$'s.
-However, this is a general Gauss Jordan solver which does not exploit symmetry properties of $\eR$, so creating, e.g.
-a batch based Cholesky solver should improve the runtimes further.
-Both the C and MATLAB implementation in this benchmark take this symmetry into account, which partially explains why the complexity curves of these implementations are different from that of the GPU implementation.
+For scientific computing the Kepler boards are considered more attractive, with the introduction of features such as kernels calling kernels, more registers, more shared memory, and better mechanisms for hiding or removing data transfers.
 
@@
 
-We have shown that for systems containing up to 32 channels the problem can be largely mitigated by building the covariance matrix in a clever way, and by making use of the massive computational power available in modern GPUs.
+We have shown that for systems consisting of up to 32 channels the problem can be largely mitigated by building the covariance matrix in a clever way, and by making use of the massive computational power available in modern GPUs.
 We were able to improve upon the run-time of a single thread C-implementation by roughly two orders of magnitude.
-For most choices of parameters the GPU was able to create images at $\sim$1\;MP/s at an average data processing rate of ~40\;GFlop/s.
+For most choices of parameters the GPU was able to create images at $\sim$1\;MP/s at an average data processing rate of $\sim$40\;GFlop/s.
 This is less than 5\% of the peak performance of the GPU, but we believe it to be near optimal given the constraints of memory bandwidth and the sequential nature of some parts of the MVDR implementation.
@@
 The GPU allows MVDR to be used in real-time processing of sonar data, and makes the MVDR a viable alternative to conventional methods in practical systems.
-
-% \begin{figure}[t]
-% \centering
-% \includegraphics[width=\linewidth]{gfx/benchmark_tagged.eps}
-% \caption{Realtime requirement for sectorscan imaging.}\label{real-time_sectorscan}
-% \end{figure}
 
@@
 % 
-% 
-% \section{Introduction}
-% 
-% \todopar{Case:
-% \begin{itemize}
-% \item UFFC(?)
-% \item Ultrasound. Arrays up to 128 channels. Processing power vital. What's the best way to implement 
-% \item Other articles evaluates Capon image quality, and some have shown that speedups can be achieved by reducing the complexity of the traditional Capon algorithm or by implementing specific versions on a GPU.
-% \end{itemize}
-% }
-% 
-% \begin{itemize}
-% \item As briefly as possible, introduce beamforming.
-% \item Adaptive beamformer's potential lies in its ability to suppress interference power
-% \item Why adaptive beamformers struggle in active sonar systems.
-Correlated noise, robustification kills the adaptive potential.
-Quite computationally intensive.
-Constraints must be applied in one way or another - parameters must be tuned.
-% \item The three main ways to implement Capon \cite{Capon1969} is \todo{not entirely true?}
-% \begin{itemize}
-% \item Traditional way: Build $\eR$ and solve $\w = \frac{\Ri\a}{\a\H\Ri\a}$. 
-% \item Iterative methods: Woodbury, Conjugated gradients
-% \item Beamspace
-% \item LCA (or other methods, as reference)
-% \end{itemize}
-% fs = 100kHz*30\%*$\frac{4}{3}$ = 40kHz \\
-% $N_{range px} = \frac{2\;250m}{1500m/s}40kHz = 13.3kpx$ \\
-% $\times 32$ beams = 426kpx \\\\
 % 
@@
 
-\section{Appendices}
-
-\subsection{MVDR Complexity Formulas}\label{mvdr_formulas}
+\appendices
+
+\section{MVDR Complexity Formulas}\label{mvdr_formulas}
 
@@
 \begin{align}
-O_{\underset{\text{initial}}{\text{Build $\eR$}}} &= \underbrace{\Om\Nk\Nl L^2}_\text{Multiplications} + \Oa(\Nk+\Nl-2)L^2 + O_\text{dload},
+O_{\underset{\text{initial}}{\text{Build $\eR$}}} &= \underbrace{\Om\Nk\Nl L^2}_\text{Multiplications} + \underbrace{\Oa(\Nk+\Nl-2)L^2}_\text{Additions} + O_\text{dload},
 \end{align}
@@
 
-\subsection{GPU Throughput}\label{throughput}
+\section{GPU Throughput}\label{throughput}
 
@@
 The Quadro 6000 has 32 CUDA cores per SM, each operating at a rate of 1148\;MHz and being able to perform 2 floating point operations (flop) per clock cycle when multiply-add instructions are used.
-Theoretical peak arithmetic throughput is then given as:
+The theoretical peak arithmetic throughput is then given as:
 \begin{align}
-\text{B}_\text{arith} &= 2\cdot1148\;\text{flop/s/core}\cdot 32\;\text{cores/SM}\cdot 14\;\text{SMs}\nn
+\text{B}_\text{arith} &= 2\cdot1148\;\text{Mflop/s/core}\cdot 32\;\text{cores/SM}\cdot 14\;\text{SMs}\nn
 &= 1.03\;\text{Tflop/s}.\label{flops}
@@
 Note that even when using shared memory at least 4 floating point operations must be carried out per float transferred to the CUDA cores, otherwise the algorithm will be memory bound and the peak arithmetic throughput can not be reached.
-%In such cases instruction level parallelism can be exploited to move data from shared memory into the even faster register memory~\cite{Vasilyy}, but this is outside the scope of this article.
 
@@
 
+% use section* for acknowledgement
+\ifCLASSOPTIONcompsoc% % This command fixes abstract positioning for compsoc articles:
+% \IEEEdisplaynotcompsoctitleabstractindextext
+% 
+% % (Optional) Add some extra info on cover page of peer review papers:
+% % \ifCLASSOPTIONpeerreview
+% % \begin{center} \bfseries EDICS Category: 3-BBND \end{center}
+% % \fi
+% 
+% % Insert page break and insert second title (peer review mode)
+% \IEEEpeerreviewmaketitle
+% 
+% 
+%
+
+  \section*{Acknowledgments}
+\else
   \section*{Acknowledgment}
-The authors would like to thank Kongsberg Maritime and the Norwegian Defence Research Establishment (FFI) for providing experimental data, and thank Nvidia for providing support for running the batched linear equation solver as well as for granting us a testdrive at the Boston HPC center.
+\fi
+
+The authors would like to thank Kongsberg Maritime and the Norwegian Defence Research Establishment (FFI) for providing experimental data, and thank Nvidia for providing support on running the batched linear equation solver and for granting us a testdrive at the Boston HPC center.
+
+% Can use something like this to put references on a page
+% by themselves when using endfloat and the captionsoff option.
+\ifCLASSOPTIONcaptionsoff
+  \newpage
+\fi
+
+% \bibliographystyle{IEEEtran}
+% \bibliography{library.bib}
+
+% Generated by IEEEtran.bst, version: 1.13 (2008/09/30)
+\begin{thebibliography}{10}
+\def\url#1{}
+\csname url@samestyle\endcsname
+\providecommand{\newblock}{\relax}
+\providecommand{\bibinfo}[2]{#2}
+\providecommand{\BIBentrySTDinterwordspacing}{\spaceskip=0pt\relax}
+\providecommand{\BIBentryALTinterwordstretchfactor}{4}
+\providecommand{\BIBentryALTinterwordspacing}{\spaceskip=\fontdimen2\font plus
+\BIBentryALTinterwordstretchfactor\fontdimen3\font minus
+  \fontdimen4\font\relax}
+\providecommand{\BIBforeignlanguage}[2]{{%
+\expandafter\ifx\csname l@#1\endcsname\relax
+\typeout{** WARNING: IEEEtran.bst: No hyphenation pattern has been}%
+\typeout{** loaded for the language `#1'. Using the pattern for}%
+\typeout{** the default language instead.}%
+\else
+\language=\csname l@#1\endcsname
+\fi
+#2}}
+\providecommand{\BIBdecl}{\relax}
+\BIBdecl
+
+\bibitem{Benitz1997}
+\BIBentryALTinterwordspacing
+G.~R. Benitz, ``{High-Definition Vector Imaging},'' \emph{Lincoln Laboratory
+  Journal}, vol.~10, no.~2, pp. 147--170, 1997.
+  \url{http://www.ll.mit.edu/publications/journal/journalarchives10-2.html}
+\BIBentrySTDinterwordspacing
+
+\bibitem{Synnevag2007}
+\BIBentryALTinterwordspacing
+J.-F. Synnev\aa{}g, A.~Austeng, and S.~Holm, ``{Adaptive beamforming applied to
+  medical ultrasound imaging.}'' \emph{IEEE transactions on ultrasonics,
+  ferroelectrics, and frequency control}, vol.~54, no.~8, pp. 1606--13, Aug.
+  2007.  \url{http://www.ncbi.nlm.nih.gov/pubmed/19811995}
+\BIBentrySTDinterwordspacing
+
+\bibitem{Blomberg2012a}
+\BIBentryALTinterwordspacing
+A.~E.~A. Blomberg, A.~Austeng, and R.~E. Hansen, ``{Adaptive Beamforming
+  Applied to a Cylindrical Sonar Array Using an Interpolated Array
+  Transformation},'' \emph{IEEE Journal of Oceanic Engineering}, vol.~37,
+  no.~1, pp. 25--34, Jan. 2012.
+  \url{http://ieeexplore.ieee.org/lpdocs/epic03/wrapper.htm?arnumber=6112183}
+\BIBentrySTDinterwordspacing
+
+\bibitem{Blomberg2011}
+A.~E.~A. Blomberg, R.~E. Hansen, S.~A.~V. Synnes, and A.~Austeng, ``{Improved
+  interferometric sonar performance in shallow water using adaptive
+  beamforming},'' in \emph{Proceedings of the International Conference \&
+  Exhibition on Underwater Acoustic Measurements (UAM), Kos, Greece, June},
+  2011.
+
+\bibitem{Dursun2009}
+S.~Dursun, A.~Austeng, R.~E. Hansen, and S.~Holm, ``{Minimum variance
+  beamforming in active sonar imaging},'' in \emph{Proceedings of the 3rd
+  International Conference \& Exhibition on Underwater Acoustic Measurements:
+  Technologies and Results}, B.~e. {John S. Papadakis \& Leif}, Ed., 2009, pp.
+  1373--1378.
+
+\bibitem{Lo2004}
+\BIBentryALTinterwordspacing
+K.~Lo, ``{Adaptive Array Processing for Wide-Band Active Sonars},'' \emph{IEEE
+  Journal of Oceanic Engineering}, vol.~29, no.~3, pp. 837--846, Jul. 2004.
+  \url{http://ieeexplore.ieee.org/lpdocs/epic03/wrapper.htm?arnumber=1353435}
+\BIBentrySTDinterwordspacing
+
+\bibitem{Widrow1982}
+\BIBentryALTinterwordspacing
+B.~Widrow, K.~Duvall, R.~Gooch, and W.~Newman, ``{Signal cancellation phenomena
+  in adaptive antennas: Causes and cures},'' \emph{IEEE Transactions on
+  Antennas and Propagation}, vol.~30, no.~3, pp. 469--478, May 1982.
+  \url{http://ieeexplore.ieee.org/lpdocs/epic03/wrapper.htm?arnumber=1142804}
+\BIBentrySTDinterwordspacing
+
+\bibitem{VanTrees2002}
+\BIBentryALTinterwordspacing
+H.~L. {Van Trees}, \emph{{Optimum Array Processing}}.\hskip 1em plus 0.5em
+  minus 0.4em\relax New York, USA: John Wiley \& Sons, Inc., Mar. 2002.
+  \url{http://doi.wiley.com/10.1002/0471221104}
+\BIBentrySTDinterwordspacing
+
+\bibitem{Kailath1985}
+\BIBentryALTinterwordspacing
+T.~Kailath and T.-J. Shan, ``{Adaptive beamforming for coherent signals and
+  interference},'' \emph{IEEE Transactions on Acoustics, Speech, and Signal
+  Processing}, vol.~33, no.~3, pp. 527--536, Jun. 1985.
+  \url{http://ieeexplore.ieee.org/lpdocs/epic03/wrapper.htm?arnumber=1164583}
+\BIBentrySTDinterwordspacing
+
+\bibitem{Asl2012a}
+\BIBentryALTinterwordspacing
+B.~M. Asl and A.~Mahloojifar, ``{A low-complexity adaptive beamformer for
+  ultrasound imaging using structured covariance matrix.}'' \emph{IEEE
+  transactions on ultrasonics, ferroelectrics, and frequency control}, vol.~59,
+  no.~4, pp. 660--7, Apr. 2012.
+  \url{http://www.ncbi.nlm.nih.gov/pubmed/22547277}
+\BIBentrySTDinterwordspacing
+
+\bibitem{Jakobsson2000}
+\BIBentryALTinterwordspacing
+A.~Jakobsson, S.~Marple, and P.~Stoica, ``{Computationally efficient
+  two-dimensional Capon spectrum analysis},'' \emph{IEEE Transactions on Signal
+  Processing}, vol.~48, no.~9, pp. 2651--2661, 2000.
+  \url{http://ieeexplore.ieee.org/lpdocs/epic03/wrapper.htm?arnumber=863072}
+\BIBentrySTDinterwordspacing
+
+\bibitem{Nilsen2009}
+\BIBentryALTinterwordspacing
+C.-I.~C. Nilsen and I.~Hafizovic, ``{Digital beamforming using a GPU},'' in
+  \emph{2009 IEEE International Conference on Acoustics, Speech and Signal
+  Processing}.\hskip 1em plus 0.5em minus 0.4em\relax IEEE, Apr. 2009, pp.
+  609--612.
+  \url{http://ieeexplore.ieee.org/lpdocs/epic03/wrapper.htm?arnumber=4959657}
+\BIBentrySTDinterwordspacing
+
+\bibitem{Chen2011}
+\BIBentryALTinterwordspacing
+J.~Chen, Y.~Yiu, and H.~So, ``{Real-time GPU-based adaptive beamformer for high
+  quality ultrasound imaging},'' \emph{IEEE Ultrasonics Symposium}, vol.~1,
+  no.~1, pp. 1--4, 2011.  \url{http://hub.hku.hk/handle/10722/140228}
+\BIBentrySTDinterwordspacing
+
+\bibitem{Chen2011a}
+\BIBentryALTinterwordspacing
+J.~Chen, B.~Y. Yiu, B.~K. Hamilton, A.~C. Yu, and H.~K.-H. So, ``{Design space
+  exploration of adaptive beamforming acceleration for bedside and portable
+  medical ultrasound imaging},'' \emph{ACM SIGARCH Computer Architecture News},
+  vol.~39, no.~4, p.~20, Dec. 2011.
+  \url{http://dl.acm.org/citation.cfm?doid=2082156.2082162}
+\BIBentrySTDinterwordspacing
+
+\bibitem{Chapman1976}
+\BIBentryALTinterwordspacing
+D.~Chapman, ``{Partial adaptivity for the large array},'' \emph{IEEE
+  Transactions on Antennas and Propagation}, vol.~24, no.~5, pp. 685--696, Sep.
+  1976.
+  \url{http://ieeexplore.ieee.org/lpdocs/epic03/wrapper.htm?arnumber=1141408}
+\BIBentrySTDinterwordspacing
+
+\bibitem{Asen2012}
+J.~P. \AA{}sen, J.~I. Buskenes, C.-I.~C. Nilsen, A.~Austeng, and S.~Holm,
+  ``{Implementing Capon Beamforming on the GPU for Real Time Cardiac Ultrasound
+  Imaging},'' in \emph{Proceedings IEEE Ultrasonics Symposium}, 2012.
+
+\bibitem{Asen2013}
+\BIBentryALTinterwordspacing
+J.~P. \AA{}sen, J.~I. Buskenes, C.-I. {Colombo Nilsen}, A.~Austeng, and
+  S.~Holm, ``{Implementing capon beamforming on a GPU for real-time cardiac
+  ultrasound imaging.}'' \emph{IEEE transactions on ultrasonics,
+  ferroelectrics, and frequency control}, vol.~61, no.~1, pp. 76--85, Jan.
+  2014.  \url{http://www.ncbi.nlm.nih.gov/pubmed/24402897}
+\BIBentrySTDinterwordspacing
+
+\bibitem{Buskenes2013}
+\BIBentryALTinterwordspacing
+J.~I. Buskenes, J.~P. \AA{}sen, C.-I.~C. Nilsen, and A.~Austeng, ``{Adapting
+  the minimum variance beamformer to a graphics processing unit for active
+  sonar imaging systems},'' \emph{The Journal of the Acoustical Society of
+  America}, vol. 133, no.~5, p. 3613, 2013.
+  \url{http://link.aip.org/link/JASMAN/v133/i5/p3613/s2\&Agg=doi}
+\BIBentrySTDinterwordspacing
+
+\bibitem{Harris1978}
+\BIBentryALTinterwordspacing
+F.~Harris, ``{On the use of windows for harmonic analysis with the discrete
+  Fourier transform},'' \emph{Proceedings of the IEEE}, vol.~66, no.~1, pp.
+  51--83, 1978.
+  \url{http://ieeexplore.ieee.org/lpdocs/epic03/wrapper.htm?arnumber=1455106}
+\BIBentrySTDinterwordspacing
+
+\bibitem{Capon1969}
+\BIBentryALTinterwordspacing
+J.~Capon, ``{High-resolution frequency-wavenumber spectrum analysis},''
+  \emph{Proceedings of the IEEE}, vol.~57, no.~8, pp. 1408--1418, 1969.
+  \url{http://ieeexplore.ieee.org/lpdocs/epic03/wrapper.htm?arnumber=1449208}
+\BIBentrySTDinterwordspacing
+
+\bibitem{Synnevag2009a}
+\BIBentryALTinterwordspacing
+J.-F. Synnev\aa{}g, A.~Austeng, and S.~Holm, ``{Benefits of minimum-variance
+  beamforming in medical ultrasound imaging},'' \emph{IEEE Transactions on
+  Ultrasonics, Ferroelectrics and Frequency Control}, vol.~56, no.~9, pp.
+  1868--1879, Sep. 2009.
+  \url{http://ieeexplore.ieee.org/lpdocs/epic03/wrapper.htm?arnumber=5278437}
+\BIBentrySTDinterwordspacing
+
+\bibitem{Cox1987}
+\BIBentryALTinterwordspacing
+H.~Cox, R.~Zeskind, and M.~Owen, ``{Robust adaptive beamforming},'' \emph{IEEE
+  Transactions on Acoustics, Speech, and Signal Processing}, vol.~35, no.~10,
+  pp. 1365--1376, Oct. 1987.
+  \url{http://ieeexplore.ieee.org/lpdocs/epic03/wrapper.htm?arnumber=1165054}
+\BIBentrySTDinterwordspacing
+
+\bibitem{Maksym1979}
+\BIBentryALTinterwordspacing
+J.~N. Maksym, ``{A robust formulation of an optimum cross-spectral beamformer
+  for line arrays},'' \emph{The Journal of the Acoustical Society of America},
+  vol.~65, no.~4, p. 971, 1979.
+  \url{http://link.aip.org/link/JASMAN/v65/i4/p971/s1\&Agg=doi}
+\BIBentrySTDinterwordspacing
+
+\bibitem{Nvidia}
+\BIBentryALTinterwordspacing
+Nvidia, ``{Nvidia registered developers program}.''
+  \url{https://developer.nvidia.com/registered-developer-programs}
+\BIBentrySTDinterwordspacing
+
+\bibitem{Vasilyy}
+\BIBentryALTinterwordspacing
+V.~Volkov, ``{GTC: Better Performance at Lower Occupancy},'' 2010.
+  \url{http://www.cs.berkeley.edu/~volkov/volkov10-GTC.pdf}
+\BIBentrySTDinterwordspacing
+
+\bibitem{Nvidia2012}
+\BIBentryALTinterwordspacing
+Nvidia, \emph{{CUDA C Programming Guide v4.2}}, 2012.
+  \url{http://developer.nvidia.com/cuda/nvidia-gpu-computing-documentation}
+\BIBentrySTDinterwordspacing
+
+\bibitem{Nvidia2012a}
+\BIBentryALTinterwordspacing
+Nvidia, \emph{{CUDA C Best Practices Guide v4.1}}, 2012.
+  \url{http://developer.nvidia.com/cuda/nvidia-gpu-computing-documentation}
+\BIBentrySTDinterwordspacing
+
+\bibitem{Hansen2009}
+R.~E. Hansen, H.~J. Callow, T.~O. S\ae{}b\o, S.~A. Synnes, P.~E. Hagen,
+  G.~Fossum, and B.~Langli, ``{Synthetic aperture sonar in challenging
+  environments: Results from the HISAS 1030.}'' in \emph{Proceedings of the 3rd
+  International Conference \& Exhibition on Underwater Acoustic Measurements:
+  Technologies and Results}, 2009.
+
+\end{thebibliography}
+
+\input{bio/jo_inge_buskenes}
+\input{bio/jon_petter_aasen}
+\input{bio/carl-inge_nilsen}
+\input{bio/andreas_austeng}
+
+% 
+% IMAGE METRICS
+%
+% - Point spread function (res via main lobe width, SAS gain via PDF height)
+% - Constrast measures
+%
+% ACR = cL/2
+% C = (s+n)/n ratio
+% - 
+% WHO's needing processing power
+% - Centre for Maritime Research and Experimentation
+
+\vfill
+
+\end{document}

### 2014-04-24 ThesisJP/Paper5/hos.tex

\graphicspath{{./Paper5/}}

\begin{abstract}
The introduction of graphics processing unit (GPU) computing has made it possible to speed up computationally demanding algorithms. One of these algorithms is the calculation of pressure fields from acoustic transducers. Here, the execution time often limits the number of elements and field points we can select in order to get results back in finite time.

In this paper we present a simple GPU-based simulator capable of simulating high resolution pressure fields at interactive frame rates. The simulator is based on the same principle as the Ultrasim toolbox, where responses from several point sources are accumulated in a set of observation points, hence solving the Rayleigh-Sommerfeld integral. The cumulative sum for each observation point is independent of all other observation points, making the problem perfect for GPU processing. For the simulator we provide both a Paint-like interface for interactive drawing of uniform linear arrays and free-hand shapes, and a Matlab interface for precise scripting of element positions and observation points.

The presented GPU simulator was compared both with a multi threaded C-version, Ultrasim, and Field II. Compared with Ultrasim we report a 400 times speedup when simulating a varying number of source points on a 150 K points observation grid. The test system consisted of a low-end GPU (Nvidia Quadro 600) and an Intel i7-870 2.93 GHz quad-core CPU.

\end{abstract}

\section{Introduction}
Software for simulation of pressure fields from ultrasound transducers is an important tool for gaining knowledge on how a transducer performs. This brings both great insight to researchers and helps lowering production costs. A downside is that calculating large fields is often found to be time consuming, and the number of field points and the level of transducer approximation have to be adjusted low enough to have the result back in finite time. The introduction of graphics processing unit (GPU) computing has made it possible to speed up computationally demanding algorithms. In this paper we present a simple GPU-based simulator capable of simulating high resolution pressure fields at interactive frame rates.

Today there exist a vast collection of ultrasound simulators. There is Field II by Jensen\cite{Jensen1992}, considered the gold standard of ultrasound simulation. However, there are also simulators that provides different functionality and perform different trade-offs than Field II. One example is the Ultrasim toolbox\cite{Holm2001}. As for Field II, this is also a Matlab plugin. However, where Field II is mostly used to simulate ultrasound images and beams, Ultrasim can only simulate snapshots of transducer transmit fields. Another difference, is the way these toolboxes are organized. Field II is a collection of functions that makes it easy to run large batched simulations where different acquisition setups are to be investigated. Ultrasim, on the other hand, is an interactive tool with a graphical user interface controlling all simulation parameters. Ultrasim is therefore more accessible if one only need to simulate transducer transmit pressure. Field II and Ultrasim are based on two different concepts, the spatial impulse response and Huygens' principle respectively. In addition there are also simulators based on the angular spectrum method \cite{Prieur2012} capable of simulating 3D harmonic fields. And lately, simulators, capable of simulating 3D ultrasound images in orders of seconds have also been introduced \cite{Hergum2009}.

\begin{figure}[!t]
\centering
\includegraphics[width=3in]{img/huygens_principle.eps}
\caption{Huygens' principle showing the superposition of spherical waves from many point sources.}
%Huygens' principle. A wave front, here represented by the diffraction pattern formed by a narrow slit, can be represented as the superposition of spherical waves originating from an infinite set of point sources placed inside the opening. If we limit the calculation to a finite set of source points  $\{S_i\}_{i=0}^{N-1}$, field pressure, $p$,  at a given location $\vec{x}$, can be found as the coherent sum of the $N$ contributing sources.}
\label{fig:huygens_principle}
\end{figure}

Like Ultrasim, the presented simulator, Huygens on Speed (HOS), is based on Huygens' principle \cite{huygens} as depicted in Fig. \ref{fig:huygens_principle}. Hence, solving the Rayleight-Sommerfeld integral\cite{Holm2001}:
\begin{equation}
\phi = \frac{1}{2\pi} \int_S \frac{u_n(r_0, t-r/c)}{r}dS.
\end{equation}
where $u_n$, the normal velocity, is integrated over the transducer surface $S$ weighted with the reciprocal of distance from the field point to the transducer.

We easily see that each calculation at a given field point is independent of the result at all other field points. The problem is therefore perfect for GPU computing, since one thread on the GPU can process one field point independent of all other threads. The only inputs needed are the field point and source point properties. Examples of other simulators implemented using GPU computing includes \cite{Hlawitschka2010} and \cite{Gjerald2012}. The first simulates 3D pressure fields at high speeds using a fast near field method, and the latter is capable of simulating ultrasound images in real time using convolution and an assumption of a shift-invariant imaging system.

The simulator presented in this paper is designed to be an interactive tool for drawing and exploring different array geometries.  It is our belief that the simulator can be used both in research and academic teaching, as a neat way of demonstrating array beamforming principles.

%Methods for calculating pressure from rectangular pistons \cite{Mast2007, McGough2004}.

\section{Method}\label{sec:method}

The presented simulator works by discretizing the surface $S$ found in the Rayleigh-Sommerfeld integral into a set of point sources, $\{S_i\}_{i=0}^{N-1}$. As shown in Fig \ref{fig:huygens_principle}, the total field pressure, $p$, at position $\vec{x}$ can then be found as the coherent sum of $N$ monochromatic spherical waves:
\begin{equation}\label{eq:sum}
p(\vec{x}) = \sum_{i=0}^{N-1} \frac{A_i}{r_i}e^{2 \pi j f_i t_i}.
\end{equation}
Here, $A_i$ is the source amplitude or apodization, $f_i$ is the source frequency, $r_i$ is the distance from $\vec{x}$ to $S_i$, and $t_i$ is the wave travelling time given by $t_i = r_i/c - t + t_{S_i} + \Delta_i$, where $c$ is the speed of sound, $t$ is the snap shot time stamp, $t_{S_i}$ is the source time stamp, and $\Delta_i$ is the steering and focusing delay. Hence, a simple simulator based on (\ref{eq:sum}) works by rendering the contribution from a set of point sources with their corresponding properties in a set of observation points. How the source and observation points are positioned and what they represent is left for the user to decide.

\section{Design}
\begin{figure*}[!t]
\centering
\includegraphics[width=\textwidth]{img/huygens_gpu_layout.eps}
\caption{Implementation of Huygens' principle on the GPU. One thread is launched per observation point, here presented as an azimuth plane. Each thread then reads the corresponding observation point, and collaborates on loading source point properties to near-core memory. From near-core memory, source properties are broadcast to registers when the kernel iterates over the set of sources. Note the coalesced memory layout of $\vec{x}$ and $S$ in global memory. The figure reflects the notation used in Section \ref{sec:method}.}
\label{fig:huygens_gpu}
\end{figure*}

From (\ref{eq:sum}) we see that field energy at a given observation point, $\vec{x}$, can be found independently from all other locations in space, and just as important, the result for each observation point is written to different bits of memory. Such a problem is perfect considering the GPU's ability of processing hundreds of pixels/threads in parallel. The implementation of (\ref{eq:sum}) used in the presented simulator is depicted in Fig \ref{fig:huygens_gpu}. One thread is launched per observation point. The reason for using this level of granularity is that the number of observation points is usually large compare to the number of source points. We also end up with enough instructions per thread to hide memory and instruction latency. On the other hand, if we have more source points than observation points, we should parallelize across source points instead. However, the result from each source point has to be written to equal positions in memory, making barriers that would significantly lower the method's throughput. Continuing the discussion of Fig. \ref{fig:huygens_gpu}, the threads inside a given block work together to load all source points and their corresponding properties to near-core memory. Since all threads inside a block work on the same source simultaneously, we make use of broadcasting to feed one source to all threads in a single read per property. Note also how observation and source points are organized in global memory for coalesced reads, and by that maximizing global memory throughput.

As front-end to the presented simulator we provide both a Matlab scripting environment and a Paint-like user interface for interactive drawing of arrays and freehand shapes. The Matlab interface is minimal but highly flexible. The functional interface takes as arguments the values listed in global memory in Fig. \ref{fig:huygens_gpu}, and returns the calculated pressure for each observation point. Constructing and discretizing transducer arrays into points and calculating delays etc. is left for the user. However, an example script of how this could be done is provided.

After increased speed, the main contributions in this work is how the user interface has been composed to make an interactive and interesting experience out of exploring acoustical arrays and their properties. The user interface works like a paint program, where point sources are added based on mouse interaction. There are several supported modes of interaction, and extensions to more advanced modes are obviously possible. However, when selecting modes of drawing, precision in source point positions have not been our focus. For that, one have the Matlab interface.  Supported modes include click-and-drag of $n\lambda$-spaced arrays and freehand drawing. For both these modes there exist both pulsed and continuous wave modes.

In click-and-drag mode, equidistant point sources are created along a straight line from where the left mouse button was clicked to where it is released using the specified $\lambda$-spacing. The result is an array with one point source per element. Hence, there is no element directivity. In freehand drawing, a new point source is added for each mouse-moved event that occurs when the left mouse button is down, just like freehand drawing in Paint.

After point sources have been added, they can all be focused in a common field point by right-click-and-drag. This can be used e.g. to illustrate how ultrasound arrays are able to sweep the beam in space, and how the beam can be focused at different depths. In pulsed-wave mode this can be used to illustrate how a pulse evolves when moving through the focus. Other concepts that can be interactively visualized both in pulsed and continuous wave mode include among other; frequency dependent resolution and depth of field, the effect of apodization, grating lobes and side lobes, near- and far-field transitions, and multiple focus zones.

\begin{figure}[!t]
\centering
\includegraphics[width=0.8\textwidth]{img/hos_vs_ultrasim_plot.eps}
\caption{Benchmark of HOS and Ultrasim. Test setup consisted of 150 K observation points.Note that execution times for the presented simulator are in order of milliseconds whereas execution times for Ultrasim are reported in seconds. The target platform was a low-end Nvidia GPU (Quadro 600, 96 cores GPU) and a high-end Intel quad-core CPU (i7-870, 2.93 GHz).}
\label{fig:huygens_vs_ultrasim}
\end{figure}

\section{Results and Discussion}

\begin{figure*}[!t]
\centerline{
\subfloat[Lateral pressure]{\includegraphics[width=0.5\textwidth]{img/hos_ultrasim_field_lateral.eps}\label{fig:lateral_pressure}}
\newline
\subfloat[Axial pressure]{\includegraphics[width=0.5\textwidth]{img/compare_hos_field_usim2.eps}\label{fig:radial_pressure}}
}
\caption{Profiles of pressure generated by HOS, Ultrasim and Field II. The simulated array has 64 elements, a center frequency of 2.5 MHz, $\lambda/2$ pitch, and $\lambda/2$ wide and $10\lambda$ high elements. Uniform apodization is used and the array is focused at 8 cm depth. 5 source points (mathematical elements in Field II) were used in azimuth and 20 in elevation.}
\label{fig:huygens_vs_ultrasim_fieldII}
\end{figure*}

\begin{figure}[!t]
\centerline{
\subfloat[HOS]{\includegraphics[height=0.8\textwidth]{img/hos_pressure3.eps}\label{fig:hos_field}}
\subfloat[Ultrasim]{\includegraphics[height=0.8\textwidth]{img/ultrasim_pressure3.eps}\label{fig:ultrasim_field}}
}
\caption{Azimuth pressure field in dB generated by the presented simulator and Ultrasim for the same array simulated in Figure \ref{fig:huygens_vs_ultrasim_fieldII}. The resolution is 5.6 observation points per mm for both (250K observation points in total).}
\label{fig:huygens_vs_ultrasim_pressure_field}
\end{figure}

\begin{figure*}[!t]
\centering
\includegraphics[width=\textwidth]{img/freeHandDrawing2.png}
\caption{Free hand drawing of source points using the interactive user interface.}
\label{fig:huygens_free_hand}
\end{figure*}

Implementing Huygens' principle on the GPU results in a significant speedup, that again allows for interactive simulations of dense pressure fields from acoustical arrays. We show this in Fig. \ref{fig:huygens_vs_ultrasim} by presenting a benchmark made between HOS and Ultrasim. Note that execution times are reported in seconds for Ultrasim and milliseconds for HOS. Thus the average speedup is 400x in favour of HOS across the selected range of source points.

Fig. \ref{fig:lateral_pressure} shows the in-focus (at 80mm) lateral pressure from a 2.5MHz, 64 element ($32\lambda$ wide and $10\lambda$ high, $5 \text{ azimuth} \times 20 \text{ elevation}$ points/subelements per aperture) uniform linear array calculated with HOS, Field II and Ultrasim. The beam profiles are strikingly similar. In Fig. \ref{fig:radial_pressure} we also see that all three simulators yield the same result along an axial cut through focus. The continuous pressure field calculated with Field II was found by taking the fourier transform of the total transmit beam energy at the center frequency. 
%Better element directivity might cause the focus offset using Field II compared with the other simulators.

Fig. \ref{fig:huygens_vs_ultrasim_pressure_field} shows the azimuth pressure field calculated using HOS (a) and Ultrasim (b). The resolution is 5.6 observation points per mm for both, hence 250K observation points in total. HOS and Ultrasim used 2 seconds and 3 minutes 40 seconds respectively in order to generate the image, hence a 110x speedup. The hardware used in this comparison was an Nvidia GT520, 48 cores GPU, and a i7-870 Intel quad-core CPU. A multi threaded C-version of the presented simulator used 1 minute and 45 seconds to generate the same image. Field II used 2 minutes 51 seconds. Note that Field II has to calculate the spatial impulse response for all observation points, and is therefore not directly comparable to Ultrasim's snap-shot-mode. The spatial impulse response can be calculated using HOS by iterating in time. Using the available low-end GPU and high-end CPU, the presented simulator and Field II are equally fast on this operation (without using streams on the GPU). However, for massive parallel calculation of the spatial impulse response one might want to apply a different parallelization scheme and level of granularity than what is described in this paper.

Fig. \ref{fig:huygens_free_hand} shows how arrays of any shape can be interactively drawn using the provided user interface. A video of how the interactive user interface works can be found at \cite{hos}.

% An example of a floating figure using the graphicx package.
% Note that \label must occur AFTER (or within) \caption.
% For figures, \caption should occur after the \includegraphics.
% Note that IEEEtran v1.7 and later has special internal code that
% is designed to preserve the operation of \label within \caption
% even when the captionsoff option is in effect. However, because
% of issues like this, it may be the safest practice to put all your
% \label just after \caption rather than within \caption{}.
%
% Reminder: the "draftcls" or "draftclsnofoot", not "draft", class
% option should be used if it is desired that the figures are to be
% displayed while in draft mode.
%
%\begin{figure}[!t]
%\centering
%\includegraphics[width=2.5in]{myfigure}
% where an .eps filename suffix will be assumed under latex, 
% and a .pdf suffix will be assumed for pdflatex; or what has been declared
% via \DeclareGraphicsExtensions.
%\caption{Simulation Results}
%\label{fig_sim}
%\end{figure}

% Note that IEEE typically puts floats only at the top, even when this
% results in a large percentage of a column being occupied by floats.

% An example of a double column floating figure using two subfigures.
% (The subfig.sty package must be loaded for this to work.)
% The subfigure \label commands are set within each subfloat command, the
% \label for the overall figure must come after \caption.
% \hfil must be used as a separator to get equal spacing.
% The subfigure.sty package works much the same way, except \subfigure is
% used instead of \subfloat.
%
%\begin{figure*}[!t]
%\centerline{\subfloat[Case I]\includegraphics[width=2.5in]{subfigcase1}%
%\label{fig_first_case}}
%\hfil
%\subfloat[Case II]{\includegraphics[width=2.5in]{subfigcase2}%
%\label{fig_second_case}}}
%\caption{Simulation results}
%\label{fig_sim}
%\end{figure*}
%
% Note that often IEEE papers with subfigures do not employ subfigure
% captions (using the optional argument to \subfloat), but instead will
% reference/describe all of them (a), (b), etc., within the main caption.

% An example of a floating table. Note that, for IEEE style tables, the 
% \caption command should come BEFORE the table. Table text will default to
% \footnotesize as IEEE normally uses this smaller font for tables.
% The \label must come after \caption as always.
%
%\begin{table}[!t]
%% increase table row spacing, adjust to taste
%\renewcommand{\arraystretch}{1.3}
% if using array.sty, it might be a good idea to tweak the value of
% \extrarowheight as needed to properly center the text within the cells
%\caption{An Example of a Table}
%\label{table_example}
%\centering
%% Some packages, such as MDW tools, offer better commands for making tables
%% than the plain LaTeX2e tabular which is used here.
%\begin{tabular}{|c||c|}
%\hline
%One & Two\\
%\hline
%Three & Four\\
%\hline
%\end{tabular}
%\end{table}

% Note that IEEE does not put floats in the very first column - or typically
% anywhere on the first page for that matter. Also, in-text middle ("here")
% positioning is not used. Most IEEE journals/conferences use top floats
% exclusively. Note that, LaTeX2e, unlike IEEE journals/conferences, places
% footnotes above bottom floats. This can be corrected via the \fnbelowfloat
% command of the stfloats package.

%\section{Conclusion}
%The conclusion goes here.

% conference papers do not normally have an appendix

The simulator can be downloaded from the same site \cite{hos}. The source code has been made available at github \cite{hosGit}. Contributions are welcome.

% use section* for acknowledgement
\section*{Acknowledgment}
The authors would like to thank Tore Bj\aa{}stad, Torbj\o{}rn Hergum and Hans Torp for the initial idea of creating the HOS simulator.

### 2014-04-24 ThesisJP/Paper3/moving_obj.tex
Minor update in moving objects article.

@@
 \begin{align}
-&\min_{\vec{w}} E\{|z[n]|^2\} \rightarrow \min_{\vec{w}} \vec{w}^H \mat{\hat{R}} \vec{w} \label{eq:capon_optimization_criteria} \\
+&\min_{\vec{w}} E\{|z[n]|^2\} \approx \min_{\vec{w}} \vec{w}^H \mat{\hat{R}} \vec{w} \label{eq:capon_optimization_criteria} \\
 &\text{subjected to } \vec{w}^H\vec{a} = 1,

### 2014-04-24 HowtoHowtoCapon/bio/jon_petter_aasen.tex

\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{bio/jon_petter.eps}}]{Jon Petter \AA{}sen}
(S'12) was born in Porsgrunn, Norway in 1986. He received the B.Sc. and M.Sc. degree in computer science from the University of Oslo, Norway, in 2010. He is currently pursuing his Ph.D. degree in medical ultrasound technology at the Norwegian University of Science and Technology (NTNU) Medical Imaging Lab (MI-Lab), Trondheim, Norway. His research interests include adaptive ultrasound processing techniques and acceleration of ultrasound algorithms using Graphics Processing Units (GPUs). 
\end{IEEEbiography}

### 2014-04-24 HowtoHowtoCapon/submission/submit1/answer/answer.tex

--- 2014-03-18 HowtoHowtoCapon/submission/submit1/answer/answer.tex
@@
 Does Figure 3 set the expectation?
-Figure 5 shows "relative complexity reduction" of entire pro-cess, so there is no information on relative speed of build and inversion steps.
+Figure 5 shows "relative complexity reduction" of entire process, so there is no information on relative speed of build and inversion steps.
 Figure 10 shows estimated FLOPs for the two steps, but since both steps have different theoretical complexities (operations counts), and the two steps do not optimize the same way, this does not provide an expectation of the relative times.}
-\ai{TODO}
+\ai{The explanation is provided in the section V, the discussion.
+We have now rewritten the relevant paragraphs to include answers to these and other comments.
+Regarding the discussion of the FLOPS these were never intended to be an accurate representation of the run-times but rather to provide an indication of how efficiently we are able to harness the GPU's resources.
+We have reworded the paragraph slightly to emphasize this.}
+%
 \qi{On page 6, authors say their implementation was improved by factor of 1.5 - 2 when run on a K20.

### 2014-04-24 ThesisJP/Background/background.tex
Corrected thank-you section.

@@
 \includegraphics[width=\textwidth]{img/cpu_vs_gpu.pdf}
-\caption{Development in theoretical peak throughput, single precision (SP) and double precision (DP), the last decade for Intel CPUs and Nvidia GPUs.}
+\caption{Development in theoretical peak throughput, single precision (SP) and double precision (DP), the last decade for Intel CPUs and Nvidia GPUs (github.com/mgalloy/cpu-vs-gpu).}
 \label{fig:cpu_vs_gpu}

### 2014-04-24 ThesisJP/Thesis.tex
Fixed figure layout. Checked all labels.
Some minor corrections.
Corrected thank-you section.

@@
 \subsubsection{Acknowledgments}
-First of all I would like to thank my main supervisor Sverre Holm and the rest of my team of supervisors for all help and guidance through the Ph.D project.
-Andreas Austeng also deserves big thank you.
-In the second part of my Ph.D.
-project you have been my supervisor more than anyone.
+First of all I would like to thank my main supervisor Sverre Holm and the rest of my team of supervisors for all the help and guidance through the Ph.D.
+project.
+Andreas Austeng also deserves a big thank you for all valuable help and for showing so much interest in my work.
 Thank you to all my co-authors for their contributions.
@@
 
-Thank you to Ines Hafizovic and Carl-Inge Colombo Nilsen for extensive help with reviewing this thesis.
+Ines Hafizovic and Carl-Inge Colombo Nilsen have used their spare time to review this thesis, for which I am very grateful.
 
-I would like to thank my family and all my friends for all support.
-But most of all, I want to thank Beate for all her love, patient and support.
-Without it I would never have manage to complete this theses.
+Finally I would like to thank my family and all my friends for their support, and for providing a world away from work.
+But most of all, I want to thank Beate for her love and patient.
+Without your support I would never have managed to complete this thesis.
 \null
 \vfill
-\hfill To: Sofie Mathilde \AA{}sen
+\hfill Dedicated to Sofie Mathilde \AA{}sen
 \CustChapter*{List of publications}
@@
 
+\setcounter{chapter}{1}
+
 %%%% Paper 2 %%%%

### 2014-04-27 UAM2014/Article/article.tex
UAM2014 Latex files added.

%===============%                               %~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~%
% DOCUMENTCLASS %                                See full option description in "mytemplate.cls"
%===============%                               %~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~%

\documentclass[
   UAM                                          % Document type (non-standard)
 , 12pt                                         % Text size
%  , draft
%  , final                                        % Quality
%  , xelatex                                      % Use the XeLaTeX compiler
%  , biblatex                                     % Use 'biblatex' for references (best by far)
 , bibtex                                       % Use 'bibtex' for references (oldschool :)
%  , movie
%  , notodos                                      % Disable todos
 , layout
%  , defaultformat                                % Attempt to use default formatting options
%  , glossary                                     % Use a glossary
]{common/mytemplate}
% \bibliography{references}
\hypersetup{
   bookmarksopen=true
 , bookmarksopenlevel=2
}

\newlength\oldparindent
\setlength\oldparindent{\parindent}

\newlength\figwidth
\setlength\figwidth{0.7\linewidth}

% \setlength\floatsep{\baselineskip}
\setcounter{topnumber}{1}
% \setlength\baselineskip{12pt}
% \selectfont

\renewcommand\vec[1]{\boldsymbol{#1}}
\newcommand\mat[1]{\boldsymbol{#1}}

\renewcommand*\P{\mat P}
\newcommand*\V{\mat V}
\newcommand*\M{\mat M}

\usepackage{subcaption}

\begin{document}
% Load the references.bib file
% \bibliography{../../../library.bib}

\pagestyle{plain}
% \layout
% \printinunitsof{cm}
% \prntlen{\baselineskip}
% \the\textwidth
% \the\linewidth
% \prntlen{\textwidth}\\
% \prntlen\linewidth\\
% \prntlen\oddsidemargin\\
% \prntlen\oddsidemargin
% \newpage

\title{A GPU Sonar Simulator for Automatic Target Recognition}%
%
\author{J.I.Buskenes\firstAddress, C.-I.C.Nilsen\secondAddress, A.Austeng\secondAddress}%
%
\begin{contact}
  \firstAddress Dept. of Informatics, P.O.Box 1080 Blindern, N-0316 OSLO
  \secondAddress Norwegian Defence Research Establishment (FFI) P.O.Box 25, NO-2027 Kjeller
\end{contact}%
%
\begin{contact}
Contact author: Jo Inge Buskenes\\
Dept. of Informatics, Univ. of Oslo, P.Box 1080 Blindern, N-0316 OSLO\\
\href{mailto:joibu@ifi.uio.no}{joibu@ifi.uio.no}
\end{contact}%
%
\begin{abstract}
Template matching is a common technique used when classifying objects in synthetic aperture sonar (SAS) images. The principle is to isolate an image segment containing an object of interest, correlate it with a set of template images, and assign it to the class of the template yielding the highest correlation coefficient. The challenge is to come up with a suitable set of template images considering that no seabed or object is truly alike.

We target this challenge with a sonar simulator that first take as input a seabed model derived from the real sonar image. Then it places an object model on the seabed, renders the scene, and adds the resulting image to the template set. For any object position, alignment, type and material, the procedure is repeated, and a correlation coefficient computed. The faster we are able to perform these simulations, the better we can expect the classification result to be. Therefore the simulator is written in OpenGL and OpenCL and run on graphics processing units (GPUs).

The result is a fast performing, mobile and portable on-the-fly template generator which can adapt its behavior to the nature of the current scene. We believe this can prove a powerful tool for mobile sonar imaging platforms such as autonomous underwater vehicles (AUVs).
\end{abstract}%
%
\keywords{Adaptive beamforming, minimum variance, MVDR, LCA, complexity, sonar}

\section{Introduction}

A synthetic aperture sonar (SAS) utilizes a wider along-track beam to ensonify the same part of the seafloor with several sucessive pings. Later this data is combined coherently to allow the synthetization of a much larger receiving array, which drastically improves the along-track resolution.

\begin{figure}[t]\centering%
\graphicsAI[drawing,width=\linewidth]{gfx/specs.svg}%
\caption{Simulator specifications.}\label{specs}%
\end{figure}

OpenGL is a framework that

\section{Methods}

\subsection{Setting up the scene}

Before we can do anything with OpenGL we first have to set up the scene. This involves loading the models and then placing them where we want them with the proper orientation.

Our models are stored in regular 3D model files. The simulator loads these with the Open Asset Import library (\texttt{assimp}). This is a portable open source library that supports loading a wide range of 3D model formats in a uniform manner. It handles the intricacies of the various model formats and outputs the model as a node tree. Each node contains data such as vertices, facets and textures, and since they are formatted in an OpenGL friendly way we simply iterate through the nodes and load it.

The next step is to scale, orient and position the objects in the scene, and finally project this scene onto an image plane. This is achieved by applying a set of transformations to each vertex in the model. In OpenGL, the set of transformations are usually formalized like this:
\begin{align*}
\bmat{x\\y\\0\\1}_\text{image} = \P \cdot \V \cdot \M \cdot \bmat{x\\y\\z\\1}_\text{model},
\end{align*}
where is the loaded model, $\bmat{\cdot}_\text{image}$ is the output image, and $\M$, $\V$ and $\P$ are transformation matrices. The first transform we apply is the model matrix $\M$, which scales and rotates the model in its local coordinate system. Then we map this model into world coordinates using the view matrix $\V$. This involves translating the model to its respectful place in the world, and then moving it into the view of the camera. Finally the relevant part of the scene is projected onto the image plane with a projection matrix $\P$.

\subsection{OpenGL Rendering}

When our models loaded into the OpenGL scene we are ready to render.

\begin{itemize}
\item Lambertian surface scattering model. Rough, isotropic, reflects energy equally in all directions. Intensity depends only on the angle between the look direction and surface normal.
\item 
\end{itemize}

It is also possible to compute the sonar image using the OpenGL ``blend'' feature.

\section{Results}

\begin{figure}[t]\centering%
\begin{subfigure}[b]{5cm}
\graphicsAI[drawing,width=\linewidth]{gfx/data_cylinder_submerged.svg}%
\caption{SAS image of a cylinder with 2.6\;m length and 0.53\;m radius.}\label{cutmean}%
\end{subfigure}%
\begin{subfigure}[b]{5cm}
\graphicsAI[drawing,width=\linewidth]{gfx/data_cylinder_submerged.svg}%
\caption{SAS image of a cylinder with 2.6\;m length and 0.53\;m radius.}\label{cutmean}%
\end{subfigure} 
\end{figure}

% Ligger nesten med kortsiden til.
% Bilde og templat segmenteres hver for seg.
% Ikke adaptiv - dårlig. Feil lengde, skyggen for kort (synket). Orientering feil. Feil spacing. 
% Adaptiv - målt objekt. Passer bra. 
% 
% Videre:
% - Mer testing
% - Robustifisering
% - Tuning

% 
% \newpage
% \bibliography{references}
% 
% % If biblatex (file is loaded in preamble):
% % \printreferences

\end{document}

### 2014-04-27 ThesisJP/Paper3/moving_obj.tex

@@
 If an ultrasound imaging system provides a presentation of a moving object which is sensitive to small spatial shifts, the system is said to be locally spatially shift-variant.
-This can for instance happen if the axial or lateral sampling is insufficient.
+This can, for instance, happen if the axial or lateral sampling is insufficient.
 The Capon beamformer has been shown to provide increased lateral resolution in ultrasound images.
 Increased lateral resolution should demand denser lateral sampling.
-However, in previous literature on Capon beamforming for medical ultrasound imaging only single frame scenarios have been simulated.
+However, in previous literature on Capon beamforming for medical ultrasound imaging only single-frame scenarios have been simulated.
 Temporal behaviour and effects caused by the increased resolution and lack of oversampling have therefore been neglected.
 
-In this paper we analyse the local lateral shift-invariance of the Capon beamformer when imaging moving objects.
+In this paper, we analyse the local lateral shift-invariance of the Capon beamformer when imaging moving objects.
 We show that insufficient lateral sampling makes an imaging system based on the Capon beamformer laterally shift-variant.
-Different methods for oversampling on transmit and receive are then discussed and investigated in order to improve on it.
+Different methods for oversampling on transmit and receive are then discussed and investigated to improve on the Capon beamformer.
 It is shown that lateral shift-invariance can be improved by oversampling based on phase rotation on receive.
@@
 The method has been shown to improve lateral resolution and edge sharpness in ultrasound images \cite{Synnevag2007, Synnevag2009, Chen2011}.
-This is accomplished by adaptively adjusting the array weights based on element-to-element covariance in order to minimize interference.
+This is accomplished by adaptively adjusting the array weights based on element-to-element covariance to minimize interference.
 The method is, however, also highly sensitive to model errors \cite{Mestre2006, Widrow1982, Wax1996, Wax1996a}.
-This has led to the development of several robust versions of the Capon beamformer with e.g.
-spatial smoothing \cite{Shan1985} or diagonal loading \cite{JianLi2003} of the covariance estimate, or by applying steering vector uncertainty sets \cite{Lorenz2005, Rubsamen2013}.
+This has led to the development of several robust versions of the Capon beamformer with e.g., spatial smoothing \cite{Shan1985} or diagonal loading \cite{JianLi2003} of the covariance estimate, or application of steering vector uncertainty sets \cite{Lorenz2005, Rubsamen2013}.
 Spatial smoothing and diagonal loading, together with temporal smoothing of the covariance estimate, have been applied in earlier work on Capon beamforming for medical ultrasound imaging \cite{Synnevag2009}.
@@
 This self-nulling is actually how the Capon beamformer obtains its super resolution capabilities, but the signal can also get lost in the process.
-In \cite{JianLi2003} it was shown that most methods based on steering vector uncertainty sets are equivalent to diagonal loading where the loading factor depends on the steering vector uncertainty.
-In that way they make up for a given uncertainty by making the Capon beam wider.
+In \cite{JianLi2003}, it was shown that most methods based on steering vector uncertainty sets are equivalent to diagonal loading where the loading factor depends on the steering vector uncertainty.
+In that way, they make up for a given uncertainty by making the Capon beam wider.
 Hence, the maximal achievable resolution is decreased if the initial steering vector spacing was too coarse.
-In medical ultrasound imaging this initial set of steering vectors is governed by the number of receive beams.
-In this paper we will make use of an additional set of steering vectors to reduce the self-nulling effect by means of oversampling.
+In medical ultrasound imaging, this initial set of steering vectors is governed by the number of receive beams.
+In this paper, we will make use of an additional set of steering vectors to reduce the self-nulling effect by means of oversampling.
 The maximal achievable resolution is then maintained without the presence of severe self-nulling.
@@
 In \cite{Cox1973}, Cox presents in great detail how the amount of mismatch between the selected steering vector and the signal propagation vector impacts the beamformer output in simple situations.
-He also states the following important fact about the Capon beamformer [the $\vec{k}_3$ processor]:
+He also states the following important fact about the Capon beamformer (the $\vec{k}_3$ processor):
 \begin{quote}
@@
 where $\vec{k}_1$ is the delay-and-sum (DAS) beamformer.
-In Fig.\,\ref{fig:das_capon_beams} we have plotted the two-way response for DAS and the Capon beamformer using (26) and (32) in \cite{Cox1973}.
+In Fig.\,\ref{fig:das_capon_beams} we have plotted the two-way response for DAS and the Capon beamformer using \cite[Eqs.
+(26) and (32)]{Cox1973}.
 A 64-element linear array with $\lambda/2$ element spacing, where $\lambda$ is the wavelength, is scanned over a sector equal to four times the array resolution using narrow band excitation.
@@
 We see that a DAS beamformer, using the required beam spacing, could yield a $1.9$ dB scalloping loss \cite{Harris1978} if the object is moving parallel to the array.
-However, if the Capon beamformer is used with the same beam density the loss could be as large as $27$ dB.
+However, if the Capon beamformer is used with the same beam density, the loss could be as large as $27$ dB.
 This imaging system is then highly sensitive to small spatial shifts, either by the object or the array itself.
@@
 }
-\caption{Estimated drop in output power mid-way between receive beams for delay-and-sum and Capon beamforming [with subarray length of 32 and a diagonal loading factor of 0.01] applied in a farfield-narrowband setting.
+\caption{Estimated drop in output power mid-way between receive beams for delay-and-sum and Capon beamforming (with subarray length of 32 and a diagonal loading factor of 0.01) applied in a far-field-narrowband setting.
 %The scanned output from both beamformers is plotted for two different time instances.
@@
 
-In previous literature on Capon beamforming for medical ultrasound imaging only single frame scenarios have been simulated.
+In previous literature on Capon beamforming for medical ultrasound imaging, only single frame scenarios have been simulated.
 The scalloping loss effect has therefore been passed over in silence by either oversampling on transmit, by carefully positioning point scatterers exactly on transmit-receive beams, or by ignorance.
-In our first work where Capon beamforming was applied on a loop of images containing moving objects \cite{Asen2012, Asen} we commented on the effect, but no solution was presented other than oversampling on transmit when simulating.
+In our first work in which Capon beamforming was applied on a loop of images containing moving objects \cite{Asen2012, Asen}, we commented on the effect, but no solution was presented other than oversampling on transmit when simulating.
 The same was done by Jensen and Austeng \cite{Jensen2012}.
 
-In this paper we investigate the issue further.
+In this paper, we investigate the issue further.
 We show that when a moving point scatterer is imaged with the Capon beamformer with the same beam spacing which is sufficient for DAS, the predicted scalloping loss is observed both in simulations and \textit{in vitro} data.
-The question is then, what can we do in order to improve on the situation without affecting the acquisition frame rate and processing throughout.
+The problem is then to find what can be done to improve on the situation without affecting the acquisition frame rate and processing throughout.
 %We try two answer this by studying the scalloping loss of the Capon beamforming and investigating different methods aimed at reducing it.
@@
 Different methods for reduced scalloping loss and by that improved local lateral shift-invariance are presented in Section \ref{sec:methods}.
-Then, in Section \ref{sec:res_paper3} we present the results of applying the most promising method on \textit{in vitro} data.
-Finally we discuss our findings in Section \ref{sec:dis_paper3} and give our conclusions in Section \ref{sec:con_paper3}.
+Then, in Section \ref{sec:res_paper3}, we present the results of applying the most promising method on \textit{in vitro} data.
+Finally, we discuss our findings in Section \ref{sec:dis_paper3} and give our conclusions in Section \ref{sec:con_paper3}.
 
@@
 \subsection{Standard Beamforming}
-The standard form of time domain array beamforming is the DAS beamformer where the output $z$ is calculated as:
+The standard form of time-domain array beamforming is the DAS beamformer, for which the output $z$ is calculated as
 \begin{align}\label{eq:das}
@@
 where $M$ is the number of elements or channels in the array, $\Delta_m[n]$ is the per-element focusing and steering delay, and $\vec{w}$ is a weight vector or window function.
-The weight vector is typically selected in order to trade resolution for a lower side lobe level.
+The weight vector is typically selected to trade resolution for a lower side-lobe level.
 Hence, the weights are typically real and their K-space response is then symmetric.
@@
 \end{align}
-where $K$ is selected proportional to the pulse length, $N_K = 2K + 1$, $N_L = M-L+1$, and $\vec{x}_l$ is a $L$ long subarray $[x_l[n], \dotso x_{l+L}[n]]$.
+where $K$ is selected proportional to the pulse length, $N_K = 2K + 1$, $N_L = M-L+1$, and $\vec{x}_l$ is a $L$-long subarray $[x_l[n], \dotso x_{l+L}[n]]$.
 Finally the matrix is loaded with a diagonal factor for numerical stability and increased robustness,
@@
 where the factor two is the effect of a two-way acquisition \cite{Hergum2009}, and where $q$ is an oversampling factor which can be applied to reduce folding of frequencies between $f_c$ and $f_c + B/2$.
-In this paper we will use the $q$-factor to perform oversampling prior to Capon beamforming.
+In this paper, we will use the $q$-factor to perform oversampling before Capon beamforming.
 
@@
 We say that the system is locally spatially shift-variant.
-Since Capon beamforming improves the lateral resolution we will focus on sensitivity to small lateral shifts, or in other words the local lateral shift-invariance property of the imaging system.%This invariance to local movements is what makes ultrasound imaging feasible in general, but it is specially important for quantitative techniques, like speckle tracking of tissue and blood, that correlate time-varying features.
+Since Capon beamforming improves the lateral resolution we will focus on sensitivity to small lateral shifts, or in other words, the local lateral shift-invariance property of the imaging system.%This invariance to local movements is what makes ultrasound imaging feasible in general, but it is specially important for quantitative techniques, like speckle tracking of tissue and blood, that correlate time-varying features.
 %In this paper we analyze the lateral shift-invariance property of the Capon beamformer by studying the beamformer output when applied on simulated and \textit{in vitro} images of moving objects.
@@
 In an ultrasound image with 50 dB dynamic range mapped to 256 gray levels, a 1 dB loss corresponds to 5 gray levels.
-This is approximately equal to the visibility threshold [Weber fraction of 2\%] for gray scale images.
+This is approximately equal to the visibility threshold [Weber fraction of 2\%] for grayscale images.
 A loss larger than 1 dB could therefore end up being visible to the observer.
@@
 
-In order to visualize the local lateral shift-invariance of a given imaging system, we will make use of lateral shift-variance plots (LSV-plots) \cite{Hergum2007}.
-A LSV-plot is constructed for a given imaging system by imaging a point scatterer moving laterally at constant speed.
-For each lateral position the root-mean-square (RMS) of the image data is calculated in range, and the resulting beam profiles are stacked in to an image.
+To visualize the local lateral shift-invariance of a given imaging system, we will make use of lateral shift-variance plots (LSV-plots) \cite{Hergum2007}.
+An LSV-plot is constructed for a given imaging system by imaging a point scatterer moving laterally at constant speed.
+For each lateral position, the rms of the image data is calculated in range, and the resulting beam profiles are stacked in to an image.
 %For a given scan sequence there are twice as many point positions than there are receive lines.
@@
 
-To further quantify the shift-invariance we have calculated the mean absolute error (MAE) between the maximum point position in each beam profile and a straight line ($f(x)=x$).
+To further quantify the shift-invariance we have calculated the mean absolute error (MAE) between the maximum point position in each beam profile and a straight line [$f(x)=x$].
 We will refer to this measure as P-MAE.
@@
 We refer to this measure as A-MAE.
-The first measure quantifies the amount of geometric distortion \cite{Hergum2007}, and the latter quantifies the peak gain variation or in other words the scalloping loss.
+The first measure quantifies the amount of geometric distortion \cite{Hergum2007}; and the latter quantifies the peak gain variation, or in other words, the scalloping loss.
 The maximum amplitude deviation, or maximum scalloping loss, along this point trace is also calculated.
@@
 Transmit focus in azimuth and elevation were configured to be equal and the point scatterer was moved through this point.
-The transmit focus is where the ultrasound beam is at its narrowest, hence shift variance effects will appear at this location first.
+The transmit focus is where the ultrasound beam is at its narrowest; hence shift variance effects will appear at this location first.
 The beam density was calculated  based on (\ref{eq:resolution}) with a speed of sound equal to 1540.
@@
 We observe that Fig.\,\ref{fig:capon2} now contains a scaled version of the pattern presented in Fig.\,\ref{fig:das_shiftvariant}.
-From Fig.\,\ref{fig:capon} it should be clear that Capon beamforming either requires increased lateral sampling, or the resolution has to be reduced to the same level as DAS in order to make the imaging system locally laterally shift-invariant.
+From Fig.\,\ref{fig:capon} it should be clear that Capon beamforming either requires increased lateral sampling, or the resolution has to be reduced to the same level as DAS to make the imaging system locally laterally shift-invariant.
 It would be unwise to do all the computation involved with Capon beamforming with a high $d$ value and end up with a resolution equal to DAS.
@@
 \section{Oversampling Methods}\label{sec:methods}
-In this section we will discuss different methods aimed at improving the local lateral shift-invariance of the Capon beamformer.
-From the previous section it should be clear that we somehow need to increase the lateral sampling to make the Capon imaging system locally laterally shift-invariant.
-In this section we will investigate how this oversampling should be conducted, how large it needs to be and how the proposed methods will affect the acquisition frame rate and the processing complexity.
+In this section, we will discuss different methods aimed at improving the local lateral shift-invariance of the Capon beamformer.
+From the previous section, it should be clear that we somehow need to increase the lateral sampling to make the Capon imaging system locally laterally shift-invariant.
+In this section, we will investigate how this oversampling should be conducted, how large it must be, and how the proposed methods will affect the acquisition frame rate and the processing complexity.
 
 \subsection{Oversampling on Transmit}
-A straight forward way of obtaining sufficient sampling in an active imaging system with an equal number of transmit (tx) and receive (rx) beams is to increasing the number of tx-beams until the system is locally laterally shift-invariant.
-In Fig.\,\ref{fig:capon_oversampling_a} and Fig.\,\ref{fig:capon_oversampling_b} we have increased the tx-beam density by a $q$-factor of four and sixteen respectively.
-As expected the geometric distortion and scalloping loss decreases with increased lateral sampling.
-Note that Fig.\,\ref{fig:capon_oversampling_b} is zoomed compared to Fig.\,\ref{fig:capon} in order to better see the beam-to-beam variation.
+A straightforward way of obtaining sufficient sampling in an active imaging system with an equal number of transmit (tx) and receive (rx) beams is to increasing the number of tx-beams until the system is locally laterally shift-invariant.
+In Fig.\,\ref{fig:capon_oversampling_a} and Fig.\,\ref{fig:capon_oversampling_b} we have increased the tx-beam density by $q$-factors of 4 and 16 respectively.
+As expected, the geometric distortion and scalloping loss decrease with increased lateral sampling.
+Note that Fig.\,\ref{fig:capon_oversampling_b} is zoomed-in compared to Fig.\,\ref{fig:capon} to better see the beam-to-beam variation.
 
-From Fig.\,\ref{fig:capon_oversampling_b} we see that sixteen times oversampling reduces the scalloping loss significantly.
-Even though high frequency variations are reduced, the maximum is still larger than 1 dB.
-An even higher level of oversampling is therefore required in order to have a completely local lateral shift-invariant Capon beamformer when a moving point scatterer is simulated without noise and with the same parameters as in Fig.\,\ref{fig:capon}.
+From Fig.\,\ref{fig:capon_oversampling_b}, we see that 16 times oversampling reduces the scalloping loss significantly.
+Even though high-frequency variations are reduced, the maximum is still larger than 1 dB.
+An even higher level of oversampling is therefore required to have a completely local lateral shift-invariant Capon beamformer when a moving point scatterer is simulated without noise and with the same parameters as in Fig.\,\ref{fig:capon}.
 %$1/100$ in diagonal loading.
@@
 
-Even though increasing the number of transmits improves the lateral shift-invariance it will also reduce the acquisition frame rate with a factor equal to the oversampling factor.
+Even though increasing the number of transmits improves the lateral shift-invariance, it will also reduce the acquisition frame rate with a factor equal to the oversampling factor.
 The processing complexity will also increase equally.
@@
 \caption{Capon beamforming as described in Fig.\,\ref{fig:capon} with oversampling on transmit.
-The figure is zoomed compared with Fig.\,\ref{fig:capon} to better see the beam-to-beam variation.
-a) Four times oversampling ($q=4$).
-b) Sixteen times oversampling ($q=16$).
+The figure is zoomed-in compared with Fig.\,\ref{fig:capon} to better see the beam-to-beam variation.
+a) 4 times oversampling ($q=4$).
+b) 16 times oversampling ($q=16$).
 The maximum scalloping loss is reduced, but it is still too high for the system to be locally laterally shift invariant.}
@@
 Parallel receive beamforming (PRB) means that multiple narrow receive beams (rx-beams) are constructed from one broad tx-beam.
-Then, by reducing the number of tx-beams and relying on multiple rx-beams in order to get sufficient sampling on receive, it is possible to achieve high frame rate imaging.
-However, PRB also introduces geometrical distortion which to some degree can be compensated for by using e.g.
-synthetic transmit beamforming (STB) \cite{Hergum2007, Denarie2013}.
-In \cite{Rabinovich2013} it is shown that STB works well with Capon beamforming, something which is also verified in \cite{Asen}.
-In the following discussion we consider corrected PRB as being equal to oversampling on transmit when it comes to image quality.
+Then, by reducing the number of tx-beams and relying on multiple rx-beams to get sufficient sampling on receive, it is possible to achieve high frame rate imaging.
+However, PRB also introduces geometrical distortion which, to some degree, can be compensated for by using, e.g., synthetic transmit beamforming (STB) \cite{Hergum2007, Denarie2013}.
+In \cite{Rabinovich2013}, it is shown that STB works well with Capon beamforming, something which is also verified in \cite{Asen}.
+In the following discussion, we consider corrected PRB as being equal to oversampling on transmit when it comes to image quality.
 The LSV-plots for PRB are therefore not included.
 
-In cardiac ultrasound imaging, PRB is typically used in order to keep the number of rx-beams high while the number of tx-beams are reduced.
-When using PRB for oversampling we would like to maintain the same number of tx-beams while the number of rx-beams are increased.
+In cardiac ultrasound imaging, PRB is typically used to keep the number of rx-beams high while the number of tx-beams are reduced.
+When using PRB for oversampling, we would like to maintain the same number of tx-beams while the number of rx-beams are increased.
 The acquisition frame rate will then not be affected.
@@
 The processing times reported in \cite{Asen} will then also increase with the same factor.
-Since more than sixteen times oversampling is required in order to achieve lateral-shift-invariant and high-resolution imaging of a moving point scatterer, none of the implementations described in \cite{Asen} would be real time any more.
+Because more than 16 times oversampling is required to achieve lateral-shift-invariant and high-resolution imaging of a moving point scatterer, none of the implementations described in \cite{Asen} would be possible in real time.
 We therefore need a different way of oversampling our imaging sector.
@@
 The Capon beamformer formula in (\ref{eq:capon_weights}) gives a direct opportunity for oversampling using phase rotation.
-The steering vector $\vec{a}$ has in previous work been set to $\vec{1}$ since ultrasound data have to be pre-delayed.
-However, the steering vector can also, as in narrow band applications, be varied over a set of pre-defined vectors.
-In a narrow band setting the beam can make a sweep across the whole imaging sector through phase steering.
+The steering vector $\vec{a}$ has in previous work been set to $\vec{1}$ because ultrasound data must be pre-delayed.
+However, the steering vector can also, as in narrow band applications, be varied over a set of predefined vectors.
+In a narrow band setting, the beam can make a sweep across the whole imaging sector through phase steering.
 This is also possible in broadband applications as long as the phase rotation is less than one pulse length (See coarse-fine beamforming in \cite{Thomenius} and its references).
@@
 \end{align}
-where $\alpha$ is the pulse length in $\lambda$'s, and $2$ and $\delta\theta$ is the oversampling factor and beam spacing from (\ref{eq:resolution}).
+where $\alpha$ is the pulse length in terms of $\lambda$, and $2$ and $\delta\theta$ is the oversampling factor and beam spacing from (\ref{eq:resolution}).
 For our simulations $\theta_{max} = 3\delta\theta$.
@@
 
-Figure \ref{fig:capon_oversampling_pr} shows the phase rotation method applied on the same scan sequence as in Fig.\,\ref{fig:capon} with four and sixteen times oversampling ($q$-factor) on receive respectively.
-Note that the figure is zoomed compared with Fig.\,\ref{fig:capon}.
-As with oversampling on transmit there is some variation when the point moves in and out of focus.
+Figure \ref{fig:capon_oversampling_pr} shows the phase rotation method applied on the same scan sequence as in Fig.\,\ref{fig:capon} with 4 and 16 times oversampling ($q$-factor) on receive respectively.
+Note that the figure is zoomed-in compared with Fig.\,\ref{fig:capon}.
+As with oversampling on transmit, there is some variation when the point moves in and out of focus.
 The maximum scalloping loss is therefore large.
-The extreme local variations are however reduced significantly.
+The extreme local variations are, however, reduced significantly.
 Comparing Fig.\,\ref{fig:capon_oversampling} with Fig.\,\ref{fig:capon_oversampling_pr}, the errors introduced by the phase rotation method are best seen in sidelobes.
@@
 Since the phase rotation method operates on the same channel data as in Fig.\,\ref{fig:capon1}, there will be no reduction in the acquisition frame rate by applying this method.
-In addition, since the same number of covariance matrices are constructed and inverted, there is no extra cost involved here as well.
+In addition, since the same number of covariance matrices are constructed and inverted, there also is no extra cost involved.
 The only additional cost will be the multiplication of the extra steering vectors with the inverted covariance matrix.
-If the equation $\mat{\hat{R}}\vec{a} = \vec{b}$, as in \cite{Asen}, is solved using the Gauss-Jordan algorithm, we can solve the problem for all steering vectors at once by row reducing the matrix $[\mat{\hat{R}},  \vec{a}_{\theta_1}, \vec{a}_{\theta_2}, \cdots \vec{a}_{\theta_{N_\theta}}]$ to the following form $[\mat{I},  \vec{b}_{\theta_1}, \vec{b}_{\theta_2}, \cdots \vec{b}_{\theta_{N_\theta}}]$.
+If the equation $\mat{\hat{R}}\vec{a} = \vec{b}$, as in \cite{Asen}, is solved using the Gauss-Jordan algorithm, we can solve the problem for all steering vectors at once by row reducing the matrix $[\mat{\hat{R}},  \vec{a}_{\theta_1}, \vec{a}_{\theta_2}, \cdots \vec{a}_{\theta_{N_\theta}}]$ to the form $[\mat{I},  \vec{b}_{\theta_1}, \vec{b}_{\theta_2}, \cdots \vec{b}_{\theta_{N_\theta}}]$.
 This is an $O(L^3)$ operation as long as $N_\theta$ is not much greater than $L$.
@@
 	\caption{Capon beamforming as described in Fig.\,\ref{fig:capon} with oversampling based on phase rotation.
-The figure is zoomed compared with Fig.\,\ref{fig:capon} to better see the beam-to-beam variation.
-a) Four times oversampling on receive ($q=4$).
-b) Sixteen times oversampling on receive ($q=16$).}
+The figure is zoomed-in compared with Fig.\,\ref{fig:capon} to better see the beam-to-beam variation.
+a) 4 times oversampling on receive ($q=4$).
+b) 16 times oversampling on receive ($q=16$).}
 	\label{fig:capon_oversampling_pr}
@@
 In Fig.\,\ref{fig:LSV_invitro} we present the result of applying the Capon beamformer on \textit{in vitro} data with and without oversampling by phase rotation.
-The data was acquired using a modified Vivid E9 high-end cardiac ultrasound scanner by GE Vingmed Ultrasound, Horten, Norway, equipped with a M5S-D phased transducer operating in harmonic mode at 3.34 MHz.
+The data was acquired using a modified Vivid E9 high-end cardiac ultrasound scanner (GE Vingmed Ultrasound, Horten, Norway) equipped with a M5S-D phased transducer operating in harmonic mode at 3.34 MHz.
 The transmit beam density was set to match (\ref{eq:resolution}) based on this frequency, and 111 tx-beams where therefore distributed on a 65\degree sector.
-Transmit focus is located at 8 cm range.
-Note that in harmonic imaging the generated transmit beam at the harmonic frequency will be wider and have lower sidelobes than the beam resulting from transmitting at the harmonic frequency \cite{Fedewa2004}.
+Transmit focus was located at 8 cm range.
+Note that in harmonic imaging, the generated transmit beam at the harmonic frequency will be wider and have lower sidelobes than the beam resulting from transmitting at the harmonic frequency \cite{Fedewa2004}.
 The lateral sampling is therefore increased compared with our simulations.
@@
 It is a result of lack of control when the probe was manually dragged across the phantom.
-In order to get a diagonal without peaks the probe has to be dragged across the phantom with a speed which matches the acquisition frame rate.
-In addition, due to the increase in lateral resolution, the Capon beamformer requires smaller frame-to-frame movements than DAS.
+To get a diagonal without peaks the probe has to be dragged across the phantom with a speed which matches the acquisition frame rate.
+In addition, because of the increase in lateral resolution, the Capon beamformer requires smaller frame-to-frame movements than DAS.
 Looking at the video of Capon beamforming with sixteen times oversampling by phase rotation \multimedia{Media Movie 3} the blinking is now highly reduced.
@@
 
-Faced with an imaging scene containing bright point scatterers or sharp edges, one could wonder how large the oversampling factor needs to be in order to avoid visible scalloping losses.
+Faced with an imaging scene containing bright point scatterers or sharp edges, one could wonder how large the oversampling factor needs to be to avoid visible scalloping losses.
 In his seminal paper, Cox \cite{Cox1973} derives the required beam density for a given output SNR.
@@
 Relating the presented phase rotation method to earlier work we find that it is similar to a steering vector uncertainty set.
-However, where these set-based methods typically has a criterion for selecting the best output after applying all the steering vectors contained in the set, oversampling by phase rotation outputs all.
-It can also be thought of as a narrow band approach to parallel receive beamforming where Capon's method is applied on all receive beams.
+However, where these set-based methods typically have a criterion for selecting the best output after applying all the steering vectors contained in the set, oversampling by phase rotation outputs all.
+It can also be thought of as a narrowband approach to parallel receive beamforming where Capon's method is applied on all receive beams.
 
@@
 eight phase rotated beams are averaged into four.
-If the geometric distortion, or stripes, seen at the bottom of the sector in the video with sixteen times oversampling \multimedia{Media Movie 3} is removed by such technique is left for future work.
+Whether the geometric distortion, or stripes, seen at the bottom of the sector in the video with 16 times oversampling \multimedia{Media Movie 3} is removed by such technique is left for future work.
 
-When oversampling is applied it becomes evident that the high resolution settings used in this paper might be too extreme, both with respect to the level of required oversampling and the point scatterer's appearance.
-Traditionally in medical ultrasound imaging the point spread function is always wider laterally than in range.
+When oversampling is applied it becomes evident that the high-resolution settings used in this paper might be too extreme, both with respect to the level of required oversampling and the point scatterer's appearance.
+Traditionally in medical ultrasound imaging, the point spread function is always wider laterally than in range.
 With the Capon beamformer this is now changed.
@@
 %Another option is to perform pulse inversion if a spiky symmetric point spread function is desirable.
-
 Note that oversampling is not only essential to keep the amplitude level of moving point scatterers.
 It should also help to improve contrast and edge definitions.
-If oversampling is not applied homogeneous speckle regions could get a too low average value and the edge delineation of a moving object will fluctuate in the same way as a moving point scatterer would have sparkled.
+If oversampling is not applied, homogeneous speckle regions could get a too-low average value and the edge delineation of a moving object will fluctuate in the same way as a moving point scatterer would have sparkled.
 
-Finally we want to stress the fact that this demand for oversampling does not only affect the Capon [or minimum variance beamformer].
+Finally, we want to stress the fact that this demand for oversampling does not only affect the Capon [or minimum variance beamformer].
 Other methods claiming super resolution, like the Eigenspace method, low complexity beamformer etc.
@@
 It was shown that Capon beamforming with the same beam density as required for delay-and-sum beforming results in blinking point scatterers when the system is subjected to motion.
-We then discussed and investigated different methods based on oversampling on transmit and receive in order to reduced this effect.
-Finally we showed that the local lateral shift-invariance could be improved by oversampling based on phase rotation on receive.
+We then discussed and investigated different methods based on oversampling on transmit and receive to reduced this effect.
+Finally, we showed that the local lateral shift-invariance could be improved by oversampling based on phase rotation on receive.
 This without affecting the acquisition frame rate and with a minor change in processing complexity.
 In simulations, this method reduced the maximum scalloping loss by 16 dB.
-On \textit{in vitro} data the reduction was around 7 dB and the point scatterer blinking was highly reduced.
+On \textit{in vitro} data, the reduction was around 7 dB and the point scatterer blinking was highly reduced.
 
@@
 
-The authors would like to thank Anders S\o{}rnes at GE Vingmed Ultrasound, Horten, Norway for the help with acquiring \textit{in vitro} data.
+The authors thanks A. S\o{}rnes at GE Vingmed Ultrasound, Horten, Norway for help in acquiring the \textit{in vitro} data.

### 2014-04-27 ThesisJP/Background/background.tex
Finished comments from CI and my own from easter.

@@
 \includegraphics[width=0.8\textwidth]{img/free_lunsh.png}
-\caption{Illustration of transitions between different eras in the computer industry.
-Currently there are three ongoing transitions; multi-core, heterogeneous-core and cloud-core.
-Image courtesy of Herb Sutter (herbsutter.com).}
+\caption{Illustration of transitions between different eras in the computer industry. Currently there are three ongoing transitions; multi-core, heterogeneous-core and cloud-core (herbsutter.com).}
 \label{fig:jungle}
@@
 Hence, a theoretical peak performance of close to one tera single precision floating point operations per second (TFLOPS).
-This is the same number as visualized in Fig\,\ref{fig:cpu_vs_gpu} for the Haswell architecture (Single precision).
+This is the same number as visualized in Fig.\,\ref{fig:cpu_vs_gpu} for the Haswell architecture (Single precision).
 In addition, the most powerful integrated GPU found on Haswell CPUs, the HD 5200, has 40 execution units capable of processing sixteen single-precision floating point values at 1.3 GHz.
@@
 The recent Kepler architecture from Nvidia can provide up to 15 SIMD cores\footnote{Nvidia likes to refer to the unit capable of processing one single precision floating point value as a core.
-In Nvidia's world, Kepler therefore has $15\times192=2880$ cores.} capable of processing 192 single-precision floating point values per clock cycle per core.
+In Nvidia's world, Kepler therefore has $15\times192=2880$ cores.
+A SIMD core on a Nvidia GPU is called a Streaming Multiprocessor (\nom{SM}{Streaming multiprocessor (SIMD core on Nvidia GPUs)}).} capable of processing 192 single-precision floating point values per clock cycle per core.
 These GPUs are also capable of performing FMA instructions, effectively doubling the number of theoretical FLOPS.
@@
 Also, to achieve peak performance, the implementation must perform one FMA operation on each core per clock cycle, and it must be possible for the GPU to hide all memory latency by scheduling threads which are not stalled by a memory fetch.
-All of this is something which is impossible to achieve in practical applications, and therefore one will always end up with a throughput which is a fraction of the peak throughput.
+All of this is something that is not likely to happen in practical applications, and therefore one will always end up with a throughput which is a fraction of the peak throughput.
 
@@
 Among the most interesting proposals we find the low complexity adaptive (LCA) beamformer by Synnev\aa{}g \textit{et al.} \cite{Synnevag2011} and the beamspace (BS) Capon beamformer by Nilsen and Hafizovic \cite{Nilsen2009}.
-In medical ultrasound imaging beamspace typically refers to the polar grid that cardiac ultrasound data are located in prior to scan conversion.
+In medical ultrasound imaging beamspace typically refers to the polar grid that cardiac ultrasound data are located in before it is scan converted into a Cartesian grid.
 In combination with the Capon beamformer, beamspace refers to the wavenumber-space representation of the recorded signals, i.e., the spatial Fourier transform of the element data:
@@
 The projection of volumetric data is performed by casting a ray out from every pixel in the output image (Fig.\,\ref{fig:vr_rc}).
-Along each ray, samples are gathered from the volume and assigned color and opacity.
-For ultrasound visualization, the color intensity is dictated by the ultrasound backscatter intensity, and the actual color is often encoded with depth information in order to improve depth perception.
+Along each ray, samples are gathered from the volume and assigned a color and opacity.
+For ultrasound visualization, the color intensity is dictated by the ultrasound backscatter intensity, and the colorfulness is often encoded with depth information in order to improve depth perception.
 An example of a volume rendering of cardiac ultrasound data is presented in \textbf{Paper IV}.
@@
 \end{align}
-Here, $\alpha(v)$ is the opacity transfer function, $c_{k}$ is the color intensity of sample $s_{k}$, $\alpha_{\text{acc}}$ is the accumulated opacity, and $I(s_0, s_{N_s-1})$ is the intensity value of all the projected samples combined.
+Here, $\alpha(v)$ is the opacity transfer function, $c_{k}$ is the color of sample $s_{k}$, $\alpha_{\text{acc}}$ is the accumulated opacity, and $I(s_0, s_{N_s-1})$ is the intensity value of all the projected samples combined.
 If $\alpha(v)$ is close to one for one or two consecutive samples, $\alpha_{\text{acc}}$ will quickly approach a value of one.
@@
 \section{Concluding remarks}
-All the presented problems in this chapter, except DAS beamforming, have in common that they require a processing throughput in order of GFLOPS.
-Comparing this to the typical clock frequency of modern CPUs at 3 GHz we see that GFLOPS problems will require parallel execution in order to achieve real-time performance.
+All the presented problems in this chapter, except DAS beamforming, have in common that they require a processing throughput of several GFLOPS.
+Comparing this to the typical clock frequency of modern CPUs at 3 GHz we see that GFLOPS problems quickly require parallel execution in order to achieve real-time performance.
 With CPU threads the theoretical peak performance on an eight-core CPU increases to 24 GFLOPS.
 However, to reach a peak performance above 100 GFLOPS, SIMD instructions on a CPU or SIMT instructions on a GPU have to be used.
-It also must be possible to divide the problem into a large number of independent calculations.
+However, to apply threads and SIMD instructions, it must be possible to divide the problem into a large number of independent calculations.
 

### 2014-04-27 ThesisJP/Introduction/introduction.tex
Finished comments from CI and my own from easter.

@@
 This immense rise in computational power and improved programmability are currently changing how ultrasound imaging systems are designed.
-Algorithms which previously had to be implemented in hardware for performance reasons can now be implemented in software.
+Algorithms that previously had to be implemented in hardware, for performance reasons, can now be implemented in software.
 %Algorithms, either too complicated or costly to implement in hardware, and where a software implementation was thought to be too computationally heavy for real-time use, become realizable. 
@@
 
-The Capon beamformer  is a good example of an algorithm which has been highly studied for the last decade for the application of medical ultrasound imaging, but where the number of calculations was thought to be too high for real-time processing to happen any time soon.
-No wonder, to achieve real-time processing for e.g.
+The Capon beamformer in ultrasound imaging is a good example of an algorithm that has been widely studied over the last decade where the number of calculations was thought to be too high for real-time processing.
+To achieve real-time processing for e.g.
 cardiac ultrasound imaging, an effective processing rate of hundreds of GFLOPS is required.
 This number, as explained in Section \ref{sec:adaptbf}, also grows with larger subarrays and an increasing number of samples per second.
-Nevertheless, with modern GPUs, these levels of processing are finally available within a single card.
-When researchers are exploring new algorithms for ultrasound imaging it is therefore important to keep the architecture of parallel accelerators in mind.
-If a new complex algorithm is supposed to run in real time it needs to fit the programmable and parallel pipeline of modern ultrasound scanners.
+Nevertheless, with modern GPUs, this amount of processing is finally available within a single card.
+When researchers are exploring new algorithms for ultrasound imaging, it is important to keep the architecture of parallel accelerators in mind.
+If a new complex algorithm is supposed to run in real time, it needs to fit the programmable and parallel pipeline of modern ultrasound scanners.
 
@@
 However, two additional methods have also been explored.
-The first one being adaptive visualization of cardiac ultrasound volumes [\textbf{Paper\,IV}], and the latter being simulation of dense ultrasound pressure fields [\textbf{Paper\,V}].
+The first one being adaptive visualization of cardiac ultrasound volumes [\textbf{Paper\,IV}], and the second one being the simulation of dense ultrasound pressure fields [\textbf{Paper\,V}].
 Both share the property of being computationally complex, but on the other hand they consist of many independent computations, which makes them perfectly suited for parallel GPU acceleration.
 
-Well on my way into the project, when a real-time Capon beamformer was realized, and loops of images for the first time were processed at real-time frame rates, new issues where discovered and had to be solved.
+Well into the project, when a real-time Capon beamformer was realized, and loops of images for the first time were processed at real-time frame rates, new issues where discovered and had to be solved.
 This led to a more theoretical study of the Capon beamformer in \textbf{Paper\,III}, with special attention on how to obtain high lateral resolution while preserving the important shift-invariant property of ultrasound imaging.
@@
 The first paper describes in detail how the Capon beamformer is mapped to the GPU architecture.
-Even though the paper is written within the field of active \nom{sonar}{SOund Navigation And Ranging (usually under water)} imaging, the depicted implementation is applicable to a vast range of active imaging systems.
+Even though the paper is written within the field of active \nom{sonar}{SOund Navigation And Ranging (usually under water)} imaging, the presented implementation is applicable to a range of active imaging systems.
 A similar discussion for cardiac ultrasound imaging can be found in \textbf{Paper\,VI}.
-Active sonar imaging typically differs from medical ultrasound imaging by a lower real-time requirement and fewer array elements.
+Active sonar imaging typically differs from medical ultrasound imaging by a less strict real-time requirement and fewer array elements.
 This makes it somewhat easier to reach the goal of real-time processing.
 
-The estimation of the spatial covariance matrix receives a special attention in this study.
+The estimation of the spatial covariance matrix receives special attention in this study.
 In previous literature on Capon beamforming the matrix inversion has always been regarded as the most computationally complex step.
-In this paper we show that estimation of the sample covariance matrix is the most complex part when spatial and temporal smoothing of the estimate are included.
-It is then shown how the arithmetic complexity of this estimation procedure can be reduced from cubic to square.
+In this paper we show that estimation of the sample covariance matrix is the most complex part when spatial and temporal smoothing is included.
+We then show how the computational complexity of this estimation procedure can be reduced from cubic to square.
 Finally, an in-depth analysis of the arithmetic throughput on multiple platforms is given.
-Despite our effort, the number of effective FLOPS is only 4 \% and 14 \% of the theoretical throughput of the target GPU for the matrix equation solver and covariance estimator respectively.
-Still, the reported throughput of 1 Mpx/s on a high-end GPU is enough to provide real-time processing for common sonar scan sequences.
+Despite our efforts, the number of effective FLOPS is only 4 \% and 14 \% of the theoretical throughput of the target GPU for the matrix equation solver and covariance matrix estimator respectively.
+Still, the reported throughput of 1 Mpx/s on a high-end GPU is enough to provide real-time processing for sonar imaging.
 
@@
 Achieving this will facilitate further study of the method's \textit{in vivo} performance.
-This is something which has been suggested as further work in many publications on Capon beamforming for medical ultrasound imaging the past decade.
-In \textbf{Paper\,I} arrays with no more than 32 elements were investigated.
-A linear array for cardiac ultrasound imaging has typically 64 or more elements.
-In \textbf{Paper\,VI}, which is summarized in \textbf{Paper\,II}, it is shown that our implementation from  \textbf{Paper\,I} does not reach the level of throughput required for real-time Capon beamforming of cardiac ultrasound imaging.
+This is something that has been suggested as further work in many publications on Capon beamforming for medical ultrasound imaging over the past decade.
+In \textbf{Paper\,I}, arrays with no more than 32 elements were investigated.
+A linear array for cardiac ultrasound imaging typically has 64 or more elements.
+In \textbf{Paper\,VI}, which is summarized in \textbf{Paper\,II}, it is shown that our implementation from  \textbf{Paper\,I} does not reach the level of throughput required for real-time Capon beamforming in cardiac ultrasound imaging.
 The matrices that have to be inverted become to large, and the frames that need to be processed per second are too many.
 
-In this paper, to reduce the matrix size, we take the advantage of the beamspace version of the Capon beamformer (\nom{BS-Capon}{Beamspace capon beamforming}), which is implemented on the GPU.
-For parameters previously derived to give similar performance for BS-Capon and element-space Capon (\nom{ES-Capon}{Element space capon beamforming}), we show that real-time BS-Capon beamforming is feasible for cardiac ultrasound imaging.
-The reported processing throughput is able to keep up with the acquisition frame rate in a typical cardiac ultrasound imaging system equipped with 64 and 96 element linear arrays.
-\textbf{Paper\,II} and \textbf{VI} also present for the first time videos where the ES-Capon and BS-Capon beamformer have been applied on loops of simulated and \textit{in vivo} medical ultrasound images.
+In this paper, we implement the beamspace version of the Capon beamformer (\nom{BS-Capon}{Beamspace capon beamforming}) on the GPU  to reduce the covariance matrix size.
+For parameters that give similar performance for BS-Capon and element-space Capon (\nom{ES-Capon}{Element space capon beamforming}), we show that real-time BS-Capon beamforming is feasible for cardiac ultrasound imaging.
+The reported processing throughput is able to keep up with the acquisition frame rate in a typical cardiac ultrasound imaging system equipped with 64- and 96-element linear arrays.
+\textbf{Paper\,II} and \textbf{VI} also present, for the first time, videos where the ES-Capon and BS-Capon beamformer have been applied on loops of simulated and \textit{in vivo} medical ultrasound images.
 
@@
 In the third paper we study the shift-invariant property of an imaging system based on the Capon beamformer.
-As mentioned, shift-invariant imaging is essential if the method is ever to be adopted.
+As mentioned, shift-invariant imaging is essential if the method is ever to be used in practice.
 The paper is based on an observation of aliasing artifacts when the Capon beamformer was used to image bright point scatterers \textit{in vitro} and in simulations.
-The intensity of the point scatterers were observed to oscillate as they or the probe moved.
-In earlier work on Capon beamforming for medical ultrasound imaging the real-time nature of ultrasound imaging has often been ignored, and a large degree of oversampling on transmit might have been applied without being clearly stated.
+The intensity of the point scatterers were observed to oscillate as either they or the probe moved.
+In earlier work on Capon beamforming for medical ultrasound imaging the real-time nature of ultrasound imaging has often been ignored, and a large degree of oversampling on transmit might have been applied without this being clearly stated.
 Oversampling on transmit will reduce the acquisition frame rate, and is therefore not an option if the imaging system is used to image rapidly varying objects.
@@
 With a narrowband and farfield model it is shown that the Capon beamformer suffers from lateral gain variations as large as 27 dB for parameters previously used to obtain high resolution medical ultrasound images.
-To obtain the same lateral gain variations as with \nom{DAS}{Delay-and-sum beamforming} beamforming, 25 times oversampling is shown to be required.
-For broadband nearfield simulations the same gain variations are observed.
-Finally we show that the gain variation can be reduced, hence improved shift-invariance, by oversampling on receive using phase rotation steering.
-This is achieved without affecting the acquisition frame rate and with a minor increase in computational complexity.
+To obtain the same lateral gain variations as for \nom{DAS}{Delay-and-sum beamforming} beamforming, 25 times oversampling is shown to be required.
+The same gain variations are observed for broadband nearfield simulations.
+Finally, we show that the gain variation can be reduced by oversampling on receive using phase rotation steering, hence improving shift-invariance.
+This is achieved without affecting the acquisition frame rate and with only a minor increase in computational complexity.
 The method is successfully applied on simulations and \textit{in vitro} phantom data.
@@
 The fifth paper presents an interactive simulation tool capable of simulating dense pressure fields from ultrasound arrays in real time.
-Simulation tools are heavily used by researchers in order to test out new algorithms and to deduce the performance of new array designs prior to manufacturing.
-However, dense simulations involve extensive calculation and computational time.
+Simulation tools are heavily used by researchers to test new algorithms and the performance of new array designs prior to manufacturing.
+However, dense simulations involve extensive calculations and computational time.
 With the presented simulator we aim at reducing the time needed for calculating pressure fields by means of GPU acceleration.
-The tool is based on Huygens' principle which describes diffraction caused by a slit as the superposition of several point sources located inside the opening.
+The tool is based on Huygens' principle, which describes diffraction caused by a slit as the superposition of several point sources located inside the opening.
 Thus, the simulator works by accumulating the contribution from a collection of point sources in a set of observation points.
@@
 Point sources positioned along a line will for instance simulate an ultrasound array.
-The simulator is linked with both a Paint-like interface for interactive drawing of arrays, and a Matlab interface for precise scripting of array configurations.
-The main contribution in this paper is the increased performance of the GPU implementation compared with a CPU and Matlab\footnote{The Ultrasim toolbox.} version.
-Second, the Paint-like interface provides a neat way of demonstrating array beamforming principle in real time.
-We believe this to be of great value both for array designers and teachers.
+The simulator is linked with both a painting interface for interactive drawing of arrays, and a Matlab interface for precise scripting of array configurations.
+The main contribution in this paper is the increased performance of the GPU implementation compared with CPU and Matlab\footnote{The Ultrasim toolbox.} versions.
+Second, the painting interface provides a neat way of demonstrating array beamforming principles in real time.
+We believe this to be of great value both for array designers and when teaching array processing.
 
@@
 \item a GPU implementation of the beamspace Capon beamformer.
-\item a real-time beamspace Capon beamforming is achieved for cardiac ultrasound imaging.
+\item achieving real-time beamspace Capon beamforming for cardiac ultrasound imaging.
 \item a first-time investigation of Capon beamforming applied on multiple frames in medical ultrasound imaging (simulated, \textit{in vitro}, and \textit{in vivo}).
-\item an investigation of shift-invariance when the Capon beamformer is applied on consecutive frames.
-\item a method for improved shift-invariance of the Capon beamformer when applied to medical ultrasound imaging.
-\item a method for reduced blood-pool noise in volume renderings of cardiac ultrasound volumes.
-\item a GPU implementation of this adaptive volume rendering method.
-\item a GPU implementation of simple Paint-like simulation tool for rapid visualization ultrasound pressure fields.
+\item an analysis of shift-invariance of the Capon beamformer when applied to medical ultrasound imaging.
+\item a method for real-time reduction of blood-pool noise in volume renderings of cardiac ultrasound volumes.
+\item a GPU implementation of a painting tool for rapid simulation and visualization of ultrasound pressure fields.
 \end{enumerate}
-In addition, the thesis provides several discussions on how to utilize the GPU to accelerate advanced ultrasound imaging algorithms.
+In addition, the thesis provides several more general discussions on how to utilize the GPU to accelerate advanced ultrasound imaging algorithms.
 
@@
 \subsection{Paper\,I}
-Even with all our effort in \textbf{Paper\,I} we were only able to achieve a fraction of the peak theoretic throughput of the target GPU.
-This shows how hard it is to reach the theoretical level of throughput, and that how close one can get is highly algorithm depended.
+As mentioned, we were only able to achieve a fraction of the peak theoretic throughput of the target GPU in \textbf{Paper\,I}.
+This shows how hard it is to reach the theoretical level of throughput, and that how close one can get is highly algorithm dependent.
 There might be ways to improve on these numbers, but both the matrix equation solver and the covariance matrix estimation step consist of several points where fine granular synchronization and serialized instructions are needed.
-These points will restrict even the most fine tuned implementation from reaching the maximum throughput of the GPU.
-The work in \textbf{Paper I} also showed that the covariance estimation kernel was memory bound, hence there is not enough instructions to hide all the memory latency.
+These facts will restrict even the most finely tuned implementation from reaching the maximum throughput of the GPU.
+The work in \textbf{Paper I} also showed that the covariance estimation kernel was memory bound, hence there are not enough instructions to hide all the memory latency.
 Further improvements through improved memory management could therefore be possible.
@@
 Here we see how the achieved throughput of the covariance estimation kernel is more than three times higher than the theoretical throughput of a high-end \nom{CPU}{Central processing unit} (The target CPU is from 2010).
-Note that the target GPU, the Nvidia Quadro 6000, is also from 2010.
+Note that the target high-end GPU, the Nvidia Quadro 6000, is also from 2010.
 This clearly shows the benefit of modern GPU computing.
@@
 The matrix equation found in the Capon beamformer is solved using a batched Gauss Jordan solver implemented by Nvidia.
-In our work we focused on the covariance matrix estimation step, and did not try to beat Nvidia at home.
+In our work we focused on the covariance matrix estimation step.
 Nevertheless, \textbf{Paper\,I} showed that the solver had significantly lower throughput than the covariance matrix estimation kernel.
@@
 Yet, initially it has been proven hard to transfer all results obtained in simulations to \textit{in vivo} images.
-The observed improvements are minimal, especially with the number of calculations in mind.
-The beamformers are clearly able to decrease the lateral width of point scatterers and thin structures \textit{in vivo}, and to sharpen step edges.
-On the other hand, the observed contrast is not improved and sometimes it is worse (Fig.\,6b of \textbf{Paper\,II}).
+Improvements are observed, but they are not large enough to justify the large number of calculations involved with Capon Beamforming.
+The beamformers are clearly able to decrease the lateral width of point scatterers and thin structures \textit{in vivo}, and to sharpen edges.
+On the other hand, the observed contrast is not improved, and sometimes it is even worse (Fig.\,6b of \textbf{Paper\,II}).
 Since the Capon beamformer only affects directional noise, it is clear that contrast will get worse in low \nom{SNR}{Signal-to-noise ratio} situations when signal cancellation occurs.
-The oversampling approach presented in \textbf{Paper\,III} might improve on this, but it will only reduce signal cancellation if the phase shifts across the aperture are close to linear and are small.
+The oversampling approach presented in \textbf{Paper\,III} might improve on this, but it will only reduce signal cancellation if the phase shifts across the aperture are close to linear and small.
 Another issue is that bright points with a wide lateral profile exhibit better visual contrast when surrounded by speckle noise than points with narrow lateral profiles.
-Future work should address a detailed investigation on why some results, especially the contrast obtained in simulations is so hard to achieved \textit{in vivo}.
-It will also be crucial to show larger improvements \textit{in vivo} than what is presented in \textbf{Paper\,II}.
-Both to justify all our computations, and to show an image with both highly improved resolution and contrast.
+Future work should address a detailed investigation of why some results, especially the contrast obtained in simulations, are so hard to achieve \textit{in vivo}.
+It will also be crucial to show larger improvements \textit{in vivo} than what is presented in \textbf{Paper\,II}, both to justify all our computations, and to show an image with both highly improved resolution and contrast.
 
 As pointed out at the end of \textbf{Paper\,II}, cardiac ultrasound imaging might not be the best modality for Capon beamforming.
-Cardiac ultrasound imaging was mainly selected because it is the medical ultrasound modality which requires the most calculations per second.
+Cardiac ultrasound imaging was mainly selected because it is the medical ultrasound modality that requires the most calculations per second.
 A better suited adaptive beamformer for cardiac ultrasound imaging might be the low-complexity adaptive beamformer (\nom{LCA}{Low complexity adaptive (beamformer)} beamformer), where the Capon optimization problem is applied on a set of predefined windows.
@@
 The work in \textbf{Paper\,II} made it clear that interesting findings are done when a real-time implementation makes it possible to watch the temporal behavior of an algorithm.
-This encourage for the work of \textbf{Paper\,III}.
+An observation of temporal artifacts encourage for the work of \textbf{Paper\,III}.
 
 \subsection{Paper\,III}
-In \textbf{Paper\,III} we investigate the inherent self nulling involved with Capon beamforming, and how this effect comes naturally into play when the beamformer is subjected to linear steering vector errors caused by lateral undersampling.
+In \textbf{Paper\,III} we investigate the inherent self nulling of Capon beamforming, and how this effect comes naturally into play when the beamformer is subjected to linear steering vector errors caused by lateral undersampling.
 This observation is something which has been missing in the literature on Capon beamforming for medical ultrasound imaging.
@@
 At best it will correct for the linear term introduced by the aberrator if the term is small.
-With improved shift-invariance it will also be interesting to see if high resolution beamforming could act as a post-processing step for other algorithms.
-For example to increase the precision of speckle and edge tracking.
+With improved shift invariance it will also be interesting to see if high-resolution beamforming could act as a post-processing step for other algorithms, for example to increase the precision of speckle and edge tracking.
 In the case where the image is generated for post processing it does not need to be visually pleasant anymore, as long as signal cancellation is avoided and the shift-invariance property is maintained.
-Good looking speckle, or DAS-like speckle, has been the driving force behind many of the robustification techniques and the default parameters seen in the literature.
+DAS-like speckle has been the driving force behind many of the robustification techniques and the default parameters seen in the literature.
 Hence, more aggressive parameters might be used if the image produced is intended as input to another algorithm and not for the human eye.
@@
 One limitation of the algorithm described in \textbf{Paper\,IV} is its dependence on a successful tracking.
-If the tracking fails the proposed method will fail.
+If the tracking fails, then the proposed method will fail.
 Another issue is the view-dependent visibility, where an object could become visible or disappear while rotating or if the local statistics change a lot during the cardiac cycle.
@@
 Figure \ref{fig:vis}d in \textbf{Paper\,IV} actually shows that an optimal threshold could be found based on global and not only local statistics.
-Even though the visualization based on global statistics has more errors, with respect to endocardium visibility, the tuning of the opacity function is still fully automatic if an iterative search for the minima in Fig.\,\ref{fig:vis}d is performed.
+Even though the visualization based on global statistics has more errors with respect to endocardium visibility, the tuning of the opacity function is still automatic if an iterative search for the minima in Fig.\,\ref{fig:vis}d is performed.
 A global opacity transfer function will insure a rendering which is realistic, but it will still be dependent on a successful tracking for the auto-adjustment to work.
@@
 
-In \textbf{Paper\,V} the increased speed made an interactive ''Paint-program'' feasible.
-As of January 2014 the simulator has been download 200 times and counting.
-It also has close to a 1000 views on Youtube.
+In \textbf{Paper\,V} the increased speed made an interactive painting simulation program feasible.
+As of January 2014 the simulator has been download 200 times and counting\footnote{folk.uio.no/jpaasen/huygens}.
+It also has close to a 1000 views on Youtube\footnote{youtube.com/watch?v=rLgsfskliJM}.
+
+\section{Multimedia content}
+In \textbf{Paper II} and \textbf{III} there will be references to several videos that can be accessed online\footnote{folk.uio.no/jpaasen/thesis}.
 
@@
 Finally it is worth mentioning that the code for both the Capon beamformer\footnote{github.com/jpaasen/cos} and the ultrasound field simulator\footnote{github.com/jpaasen/hos} have been released under an open source license.
+
 \endinput

### 2014-04-27 ThesisJP/Paper1/howto_howto_capon.tex

@@
 \fi%
-\caption{Per-pixel computational complexity of the steps in MVDR beamforming (prior to any optimizations).
+\caption{Per-pixel computational complexity of the steps in MVDR beamforming (before any optimizations).
 To avoid signal cancellation in an active sonar system we usually set $L<\frac{M}{2}$, in which region the computation of $\eR$ dominates in terms of arithmetic complexity, especially when performing temporal averaging.}\label{mvdr_complexity}
@@
 We selected the memory minimized version since our algorithm is severely memory bound, as we will see next.
-
-% \begin{table}[b]\centering%\normalsize
-% \begin{tabular}[c]{l r@{:}  l}\hline
-% \rowcolor{tabBlue}\bf Memory type & \multicolumn{2}{>{\columncolor{tabBlue}}c}{\bf BW$_\text{memory}$/BW$_\text{arithmetic}$} \\\hline
-% Global memory &\hspace{30pt} 1 & 30 \\
-% Shared memory & 1 & 4 \\
-% Registers & $>$3 & 2~\cite{Vasilyy}
-% \end{tabular}
-% \caption{Nvidia Quadro 6000 bandwidth ratios.}\label{tabbandwidth}
-% \end{table}
-
-\begin{table}[b]\centering%\normalsize
-\begin{tabular}[c]{l r r r@{}  l}\hline
-\rowcolor{tabBlue} & \multicolumn{1}{>{\columncolor{tabBlue}}c}{\bf B$_\text{arith}$} & \multicolumn{1}{>{\columncolor{tabBlue}}c}{\bf B$_\text{mem}$} & \multicolumn{2}{>{\columncolor{tabBlue}}c}{\bf B$_\text{mem}$/B$_\text{arith}$} \\\hline
-Arithmetic & 1.03 Tflop/s & & &\\
-Global memory & & 36 Gfloats/s & \hspace{30pt} 1 &:30 \\
-Shared memory & & 257 Gfloats/s & 1 &:4 \\
-Registers & & $>$1.5 Tfloats/s & $>$3 &:2~\cite{Vasilyy}
-\end{tabular}
-\caption{Nvidia Quadro 6000: Memory throughput, $B_{\lowercase{\text{mem}}}$, compared to arithmetic throughput, $B_\text{arith}$.}\label{throughputs}
-\end{table}
 
@@
 
-% In short, designing for a GPU involves finding the optimal balance between hiding memory latency (high occupancy) and resource utilization.
-% 
-% blocks in grid - high enough so that all SMs have atleast one block to execute.
-Optimally in the thousands to scale for future devices, and to allow an SM to switch between blocks to keep hardware busy when running synchronizing threads.
-% 
-% threads in block - min 64. 128-256 better starting point.
-
-% Instead of having a few complex core that can boost massive single thread performance, designers chose to create hundreds of lightweight cores that each is much powerful than a single CPU core, but combined boasts a massive artithmetical performance.
-% 
-% \begin{itemize}
-% \item Lots of cores, little memory for each.
-% \item Few data transfers, lots of computations.
-% \item Data transfers are moved around by hardware that can operate asynchronously from the compute units.
-% \item Memory latency hidden as long as it takes longer to compute than to move data around.
-% \item Tasks can be queued on SM (blocks) and on CPU to hide internal and external memory latencies.
-% \end{itemize}
-% 
-% Vast amount of processing power, which are easily accessible but hard to fully harness. This is because in the pursuit of an ever increasing amount of arithmetical computing cores, the 
-% 
-% Unlike a CPU, a GPU 
-% 
-% If we compare a GPU with a CPU, 
-% 
-% When implementing an algorithm on a GPU, it is important to be aware
-% 
-% People friendly:
-% \begin{itemize}
-% \item More and more compute units, but less memory for each.
-% \item Need to keep each compute unit busy.
-% \item Few data transfers, lots of computations.
-% \item Data transfers are moved around by hardware that can operate asynchronously from the compute units.
-% \item Memory latency hidden as long as it takes longer to compute than to move data around.
-% \item Tasks can be queued on SM (blocks) and on CPU to hide internal and external memory latencies.
-% \end{itemize}
-% 
-% The ``fast'' memory of the GPU is the registers and cache. All threads executed 
-% 
-% 
-% \begin{itemize}
-% \item Hiding memory access.
-% \begin{itemize}
-% \item Between GPU and CPU: Several concurrent blocks (contexts) per SM.
-% \item Job queue (how does this work)
-% \end{itemize}
-% \item Register pressure and shared memory. Thread context must fit in local memories.
-% \item Warp computational intensity should be high to make up for the memory cost.
-% \item Optimized for small L, high M, hardly any time averaging. A typical sonar scenario.
-% \end{itemize}
-% - Split program into equally sized portions - load balancing
-% - Some portions depend on other potions already being carried outer - sequential dependencies
-% - Effort timed in order to finish at the same time - synchronization
-% 
-% - task level parallelism - compute images independent of each other
-% - data level parallelism - compute pixels independent of each other
-
-% - Compute bound vs. bandwidth bound
-
 \section{Images and Benchmarks}\label{images_and_benchmarks}
 
-\ifPeerReview
-\begin{figure}[!t]\centering
-\includegraphics[width=.8\linewidth]{gfx/buske7.eps}
+To demonstrate the imaging capability of the MVDR beamformer, we have processed experimental datasets from the $M=32$ element Kongsberg Maritime HISAS1030 sonar mounted on a HUGIN AUV~\cite{Hansen2009}.
+HISAS1030 is a high resolution synthetic aperture sonar with an array length of 1.2\;m, operating frequency of 100\;kHz, and bandwidth of 30\;kHz.
+The element size and spacing is 2.5$\lambda$ and the opening angle is 25$^\circ$.
+To produce the image shown in \Fig{holmengraa} the sonar was operated in sidescan mode.
+The studied object is the 1500 dwt oil tanker wreck Holmengraa.
+It is 68\;m long and 9\;m wide, and lies at a slanted seabed at 77\;m depth outside of Horten, Norway.
+The 1 megapixel (MP) MVDR image were here processed with parameters $L=16$, $K=1$, and $d=1\%$.
+
+\ifPeerReview
+\begin{figure}[!t]\centering
+\includegraphics[width=.8\linewidth]{gfx/buske7\figPostfix.eps}
+\else
+\begin{figure}[!t]\centering
+\includegraphics[width=\linewidth]{gfx/plot_holmengraa_L16_Navg1.eps}
+\fi%
 \caption{HISAS sidescan sonar (SSS) image of the shipwreck Holmengraa that lies on a slanted seabed at 77\;m depth outside of Horten, Norway.
 The image was processed with $M=32$, $L=16$, $K=1$ and $d=1\%$.
-Prior to display the image was linearly upinterpolated by a factor 2 in azimuth making its total size 1.46\;MP.
+The image was linearly upinterpolated by a factor 2 in azimuth making its total size 1.46 megapixels (MP).
 Note how MVDR improves edge definition and reduces noise in shadow regions.}\label{holmengraa}
 \end{figure}
-\fi%
-
-\ifPeerReview
-\begin{figure}[!t]\centering
-\includegraphics[width=.8\linewidth]{gfx/buske8.eps}
+
+The computational performance of our implementation was first assessed on a test system with a quad-core Intel Xeon E5620, 64\;GB of RAM and an Nvidia Quadro 6000 card.
+The results were obtained by processing a 1\;MP image from the data from a 32 channel array, for all subarray sizes $L$, and for $K\in\{0,1,2\}$.
+Run-time measurements are shown in \Fig{benchmarks}, the run-times of memory-only and arithmetic-only GPU kernels are depicted in \Fig{code_assessment}, and an estimate of computation efficiency is presented in \Fig{code_assess_flops}.
+We were also granted a test drive at the Boston HPC centre on a machine with an Intel Xeon E5-2670, 32\;GB RAM and an Nvidia K20.
+The results from this run are shown in \Fig{benchmarks_boston}.
+
+\ifPeerReview
+\begin{figure}[!t]\centering
+\includegraphics[width=.8\linewidth]{gfx/buske8\figPostfix.eps}
 \else
@@
 The Quadro 6000 has 32 CUDA cores per SM, each operating at a rate of 1148\;MHz and being able to perform 2 floating point operations (flop) per clock cycle when multiply-add instructions are used.
-Theoretical peak arithmetic throughput is then given as:
-\begin{align}
-\text{B}_\text{arith} &= 2\cdot1148\;\text{flop/s/core}\cdot 32\;\text{cores/SM}\cdot 14\;\text{SMs}\nn
+The theoretical peak arithmetic throughput is then given as:
+\begin{align}
+\text{B}_\text{arith} &= 2\cdot1148\;\text{Mflop/s/core}\cdot 32\;\text{cores/SM}\cdot 14\;\text{SMs}\nn
 &= 1.03\;\text{Tflop/s}.\label{flops}

### 2014-04-27 UAM2014/Article/article.tex
...
la2pdf commit - 2014:04:27	21:55:03	(Sun27.Apr)
la2pdf commit - 2014:04:27	22:56:08	(Sun27.Apr)

@@
 
+% \usepackage{caption}
 \usepackage{subcaption}
@@
 
+Citations
+\begin{itemize}
+\item Midelfart/Midtgaard: Adaptive templates \cite{Midelfart2010}
+\item Simulation of sidescan images (Bell thesis \cite{Bell1997}, pre-work by V.Blake \cite{Blake1993})
+\item NURC SIGMAS+ (Coiras GPU article at OCEANS 2009 \cite{Coiras2009a}, Coiras on SIGMAS maths \cite{Coiras2009b})
+\end{itemize}
+
 \section{Methods}
@@
 
+To produce the sonar templates we assume rough, isotropic surface that reflects energy equally in all directions.
+This well with a Lambertian scattering model where the backscatter intensity depends only on the angle between the incident angle and surface normal.
+
 When our models loaded into the OpenGL scene we are ready to render.
@@
 \begin{figure}[t]\centering%
-\begin{subfigure}[b]{5cm}
+\begin{subfigure}[b]{.5\linewidth-0.5cm}
 \graphicsAI[drawing,width=\linewidth]{gfx/data_cylinder_submerged.svg}%
 \caption{SAS image of a cylinder with 2.6\;m length and 0.53\;m radius.}\label{cutmean}%
-\end{subfigure}%
-\begin{subfigure}[b]{5cm}
+\end{subfigure}\hspace{1cm}%
+\begin{subfigure}[b]{.5\linewidth-0.5cm}
 \graphicsAI[drawing,width=\linewidth]{gfx/data_cylinder_submerged.svg}%
@@
 % Bilde og templat segmenteres hver for seg.
-% Ikke adaptiv - dårlig. Feil lengde, skyggen for kort (synket). Orientering feil. Feil spacing. 
-% Adaptiv - målt objekt. Passer bra. 
+% Ikke adaptiv - darlig. Feil lengde, skyggen for kort (synket). Orientering feil. Feil spacing. 
+% Adaptiv - malt objekt. Passer bra. 
 % 
@@
 % 
-% \newpage
-% \bibliography{references}
+\newpage
+\bibliography{references}
+% \bibliography{../../../library.bib}
 % 

### 2014-04-27 UAM2014/Article/article.tex
la2pdf commit - 2014:04:27	23:31:47	(Sun27.Apr)

@@
 
-To produce the sonar templates we assume rough, isotropic surface that reflects energy equally in all directions.
-This well with a Lambertian scattering model where the backscatter intensity depends only on the angle between the incident angle and surface normal.
+To produce the sonar templates we assume a rough, isotropic surface that reflects energy equally in all directions.
+This permits us to use a Lambertian scattering model where the backscatter intensity depends only on the incidence angle~\cite{Zhang1999}.
+It does not consider observation angle or sound frequency\todo{is this unproblematic?}.
 
 When our models loaded into the OpenGL scene we are ready to render.
-
-\begin{itemize}
-\item Lambertian surface scattering model. Rough, isotropic, reflects energy equally in all directions. Intensity depends only on the angle between the look direction and surface normal.
-\item 
-\end{itemize}
 
@@
 \begin{subfigure}[b]{.5\linewidth-0.5cm}
-\graphicsAI[drawing,width=\linewidth]{gfx/data_cylinder_submerged.svg}%
+\graphicsAI[drawing,width=\linewidth]{gfx/data_cylinder_submerged_bw.svg}%
 \caption{SAS image of a cylinder with 2.6\;m length and 0.53\;m radius.}\label{cutmean}%

### 2014-04-27 UAM2014/Article/article.tex
la2pdf commit - 2014:04:28	00:14:37	(Mon28.Apr)

@@
 \usepackage{subcaption}
+
+% Tune floats
+\renewcommand{\topfraction}{0.9}  % max fraction of floats at top
+\renewcommand{\bottomfraction}{0.8} % max fraction of floats at bottom
+%   Parameters for TEXT pages (not float pages):
+\setcounter{topnumber}{2}
+\setcounter{bottomnumber}{2}
+\setcounter{totalnumber}{4}     % 2 may work better
+\setcounter{dbltopnumber}{2}    % for 2-column pages
+\renewcommand{\dbltopfraction}{0.9} % fit big float above 2-col. text
+\renewcommand{\textfraction}{0.07}  % allow minimal text w. figs
+%   Parameters for FLOAT pages (not text pages):
+\renewcommand{\floatpagefraction}{0.85}  % require fuller float pages
+% N.B.: floatpagefraction MUST be less than topfraction !!
+\renewcommand{\dblfloatpagefraction}{0.85} % require fuller float pages
 
@@
 
-\begin{figure}[t]\centering%
-\begin{subfigure}[b]{.5\linewidth-0.5cm}
+\newlength\imgspacing\setlength\imgspacing{.5cm}
+
+\begin{figure}[tp]\centering%
+\begin{subfigure}[b]{.5\linewidth-\imgspacing/2}
 \graphicsAI[drawing,width=\linewidth]{gfx/data_cylinder_submerged.svg}%
-\caption{SAS image of a cylinder with 2.6\;m length and 0.53\;m radius.}\label{cutmean}%
-\end{subfigure}\hspace{1cm}%
-\begin{subfigure}[b]{.5\linewidth-0.5cm}
+\caption{SAS image of a cylinder with 2.6\;m length and 0.53\;m radius.}\label{data_sylinder}%
+\end{subfigure}\hspace{\imgspacing}%
+\begin{subfigure}[b]{.5\linewidth-\imgspacing/2}
 \graphicsAI[drawing,width=\linewidth]{gfx/data_cylinder_submerged_bw.svg}%
-\caption{SAS image of a cylinder with 2.6\;m length and 0.53\;m radius.}\label{cutmean}%
+\caption{SAS image of a cylinder with 2.6\;m length and 0.53\;m radius.}\label{data_cylinder_bw}%
 \end{subfigure} 
+\end{figure}
+
+\begin{figure}[tp]\centering%
+\begin{subfigure}[b]{.5\linewidth-\imgspacing/2}
+\graphicsAI[drawing,width=\linewidth]{gfx/sim_cylinder_submerged.svg}%
+\caption{Simulated cylinder.}\label{sim_cylinder}%
+\end{subfigure}\hspace{\imgspacing}%
+\begin{subfigure}[b]{.5\linewidth-\imgspacing/2}
+\graphicsAI[drawing,width=\linewidth]{gfx/sim_cylinder_submerged_bw.svg}%
+\caption{Simulated cylinder.}\label{sim_cylinder_bw}%
+\end{subfigure} 
+\end{figure}
+
+\begin{figure}[tp]\centering%
+\begin{subfigure}[b]{.5\linewidth-\imgspacing/2}
+\graphicsAI[drawing,width=\linewidth]{gfx/overlay_cylinder_submerged.svg}%
+\caption{Overlay.}\label{overlay_cylinder}%
+\end{subfigure}\hspace{\imgspacing}%
+\begin{subfigure}[b]{.5\linewidth-\imgspacing/2}
+\graphicsAI[width=\linewidth]{gfx/overlay_cylinder_submerged_adaptive.svg}%
+\caption{Overlay (adaptive).}\label{overlay_cylinder_adaptive}%
+\end{subfigure} 
+\end{figure}
+
+\begin{figure}[tp]\centering%
+\graphicsAI[drawing,width=\linewidth]{gfx/overlay_colors.svg}%
+\caption{Overlay.}\label{overlay_cylinder}%
 \end{figure}

### 2014-04-27 ThesisJP/Background/background.tex
Feedback from Jo Inge finished.

@@
 The increase in processing power came mostly from an ever-growing clock frequency.
-However, power consumption and heat generation were also growing at the same rate until the required cooling finally became too much in 2004.
-From that point, CPU manufactures had to pursue different directions than just increasing the clock frequency.
-The immediate solution was to embed several cores in one CPU.
-This insured continuation of Moore's law, since the number of transistors per chip could continue to grow.
+However, power consumption and heat generation were also growing until the required cooling finally became too much in 2004.
+From that point, CPU manufactures had to look at different strategies than just increasing the clock frequency.
+One strategy was to embed several cores in one CPU.
+This ensured continuation of Moore's law, since the number of transistors per chip (CPU) could continue to grow.
 For programmers, multi-core CPUs meant that computer programs now had to be multi-threaded in order to annually gain performance.
@@
 
-Around the same time as CPU manufacturers hit the frequency wall, people had started experimenting with programmable shaders, which recently had been introduced in the field of graphics programming.
-The graphics processing pipeline had before this been a fixed function pipeline, where e.g.
-geometry and colors were fed to the graphics processing unit (GPU) in one end, and where shaded and rasterized sceneries were outputted in the other.
+Around the same time as CPU manufacturers hit the frequency wall, people had started experimenting with programmable shaders in the field of graphics programming.
+The graphics processing pipeline had before programmable shaders been a fixed function pipeline, where geometry and colors were converted into shaded and rasterized sceneries by the graphics processing unit (GPU).
 Programmable shaders introduced the ability to replace some steps in the fixed-function pipeline with custom computations.
 This made it possible to utilize the GPU for other tasks than just rendering graphics \cite{Seland2007}.
-The rationale behind this development was that GPUs were already highly available in consumer PCs, and hidden behind its graphics interface were processing powers an order of magnitude larger than what the CPU could provide (Fig.\,\ref{fig:cpu_vs_gpu}).
+The rationale behind this development was that GPUs were already available in consumer PCs, and hidden behind its graphics interface were processing capabilities an order of magnitude larger than what the CPU could provide (Fig.\,\ref{fig:cpu_vs_gpu}).
 Actually, at the time when CPU manufacturers hit the frequency wall, the GPU had already gone multi-core, first of all driven by an ever-increasing demand for more realistic computer games.
 
-The original graphics pipeline involved no conditional branching, and data were typically used once \cite{Purcell2002}.
+The original graphics pipeline involved no conditional branching (caused by if-else control flow statements), and data were typically used once \cite{Purcell2002}.
 Designers of GPUs could therefore skip advanced features found in most CPUs, like branch prediction and different levels of data caching.
@@
 In most cases it is important to balance the load equally between the CPU and GPU to get the most out of each processors.
-Balancing means that highly parallel and computationally intense tasks should be offloaded to the GPU while serialized and memory intensive work should stay on the CPU.
+Balancing means that highly parallel and computationally intensive tasks should be offloaded to the GPU while serialized and memory intensive work should stay on the CPU.
 This paradigm of utilizing specialized cores for solving specific problems is today known as \textit{heterogeneous computing}.
@@
 The last three generations of CPUs from Intel have all had on-chip integrated graphics, and all of them are therefore APUs.
-An on-chip GPU is less powerful than a high-performance discrete GPU, but valuable time is saved by not having to send data across the PCI Express bus, which connects the different parts of a computer to the CPU.
-This is the reason why one of the first rules of GPGPU (with discrete GPUs) is to minimize CPU-to-GPU memory transfers.
-An integrated GPU can therefore provide increased performance over a discrete GPU if the problem requires a high level of CPU-GPU data synchronization.
+An on-chip GPU is less powerful than a high-performance discrete GPU, but valuable time can be saved by not having to send data across the PCI Express bus, which connects the different parts of a computer to the CPU.
+This is the reason why one of the first rules of GPGPU (with discrete GPUs) is to minimize (or hide) CPU-to-GPU memory transfers.
+An integrated GPU can therefore provide increased performance over a discrete GPU if the problem requires a lot of CPU-GPU data transfers.
 
@@
 In this thesis we focus on heterogeneous computing with a multi-core CPU and a discrete high performance GPU.
-This is a cheap, high-performance system which is likely to be the specification of current and future ultrasound imaging system.
+This is a cheap, high-performance system which is likely to be the specification of current and future ultrasound imaging systems.
 
@@
 Although they started out as very different processors, modern CPUs are now including more and more GPU-like features, and GPUs are adding CPU features.
-In this section we will see what these features are and why there is still a big difference in theoretical peak performance between a CPU and a GPU.
+In this section we will see what these features are and why there is still a big difference in theoretical peak arithmetic performance between a CPU and a GPU.
 
@@
 
-In Section \ref{sec:gpgpu} a special type of instructions was mentioned, namely SIMD instructions.
-These are instructions which are designed to execute a single instruction across multiple data element in one clock cycle.
-At the end of the 20th century, Intel introduced a SIMD instruction set for their Pentium III processor called the Streaming SIMD Extensions (\nom{SSE}{Streaming SIMD extensions (instruction set)}).
+A special kind of instructions called SIMD was mentioned in Section \ref{sec:gpgpu}.
+These are instructions that are designed to execute a single instruction on multiple data elements in one clock cycle.
+In the late 1990s, Intel introduced a SIMD instruction set for their Pentium III processor called the Streaming SIMD Extensions (\nom{SSE}{Streaming SIMD extensions (instruction set)}).
 The first SSE instruction set had registers of 128 bit which meant that four single-precision (32 bit) floating point values could be processed per clock cycle.
 Updates of the SSE instruction set (SSE2, SSE3, SSE4, \nom{AVX}{Advanced vector extensions (instruction set)}, AVX2, and AVX-512) have added new instructions and longer registers.
-The newly introduced extensions, AVX and AVX2, add registers of 256 bit and fused multiply accumulate (\nom{FMA}{Fused multiply-accumulate (instruction)}) instructions  which means that sixteen single-precision floating point operations can be performed per clock cycle per core.
-A new extension, AVX-512, extends the registers to 512 bit and is found in the Intel accelerator board  Xeon Phi and might show up in the next generation of CPUs from Intel.
+The recently introduced extensions, AVX and AVX2, add registers of 256 bit and fused multiply-add (\nom{FMA}{Fused multiply-add (instruction)}) instructions  which means that sixteen single-precision floating point operations can be performed per clock cycle per core.
+The new extension AVX-512, that extends the registers to 512 bit, is found in the Intel accelerator board Xeon Phi, and might show up in the next generation of CPUs from Intel.
 The latest CPU architecture from Intel, the Haswell architecture, has two processing ports per core supporting FMA-AVX2 instructions, and is therefore capable of performing 32 single-precision floating point operations per clock cycle per core (Fig.\,\ref{fig:cpu_gpu}).
 For an eight-core\footnote{An 8-core Haswell CPU is scheduled to launch during 2014.
-The numbers in Fig.\,\ref{fig:cpu_vs_gpu} for the last three generations of CPUs are all based on 8-core CPUs.} CPU with a clock frequency of 3.8 GHz this sums to
+The numbers in Fig.\,\ref{fig:cpu_vs_gpu} for the last three generations of CPUs are all based on 8-core CPUs.} CPU with a clock frequency of 3.8 GHz this sums to a theoretical peak arithmetic performance of
 \begin{equation}
-32 \times 8 \times 3.8 = 972.8\,\text{GFLOPS}.
+32 \times 8 \times 3.8 = 972.8\,\text{GFLOPS},
 \end{equation}
-Hence, a theoretical peak performance of close to one tera single precision floating point operations per second (TFLOPS).
-This is the same number as visualized in Fig.\,\ref{fig:cpu_vs_gpu} for the Haswell architecture (Single precision).
+where FLOPS is short for; floating point operations per second.
+This is the same number as visualized in Fig.\,\ref{fig:cpu_vs_gpu} for the Haswell architecture (single precision).
 In addition, the most powerful integrated GPU found on Haswell CPUs, the HD 5200, has 40 execution units capable of processing sixteen single-precision floating point values at 1.3 GHz.
@@
 \end{equation}
-in extra processing power to the APU.
+in extra processing capabilities to the APU.
 
 Modern GPUs can also be interpreted as units consisting of multiple SIMD cores (Fig.\,\ref{fig:cpu_gpu}).
-The recent Kepler architecture from Nvidia can provide up to 15 SIMD cores\footnote{Nvidia likes to refer to the unit capable of processing one single precision floating point value as a core.
+The recent Kepler architecture from Nvidia provides up to 15 SIMD cores\footnote{Nvidia likes to refer to the unit capable of processing one single precision floating point value as a core.
 In Nvidia's world, Kepler therefore has $15\times192=2880$ cores.
-A SIMD core on a Nvidia GPU is called a Streaming Multiprocessor (\nom{SM}{Streaming multiprocessor (SIMD core on Nvidia GPUs)}).} capable of processing 192 single-precision floating point values per clock cycle per core.
+A SIMD core on a Nvidia GPU is called a Streaming Multiprocessor (\nom{SM}{Streaming multiprocessor (SIMD core on Nvidia GPUs)} or SMX).} capable of processing 192 single-precision floating point values per clock cycle per core.
 These GPUs are also capable of performing FMA instructions, effectively doubling the number of theoretical FLOPS.
@@
 \end{equation} 
-Each SIMD core also includes 16 special function units (\nom{SFU}{Special function unit (sin, exp, sqrt, ect.)}) which add close to 1 TFLOPS in additional computational power.
-This adds up to a total of 5 TFLOPS as shown in Fig.\,\ref{fig:cpu_vs_gpu} for the GeForce GTX 780 TI GPU.
-From Fig.\,\ref{fig:cpu_gpu} we see how the high number of SIMD cores found on GPUs and their larger width, compared with the SIMD cores on CPUs, results in five times the computational performance of a high-end CPU (without the integrated GPU).
+Each SIMD core also includes 16 special function units (\nom{SFU}{Special function unit (sin, exp, sqrt, ect.)}) with close to 1 TFLOPS in computational capabilities.
+One Kepler GPU therefore provides more than 5 TFLOPS as shown in Fig.\,\ref{fig:cpu_vs_gpu} for the GeForce GTX 780 TI GPU.
+From Fig.\,\ref{fig:cpu_gpu} we see how the high number of SIMD cores found on GPUs and their larger width, compared to the SIMD cores on CPUs, result in five times the computational performance of a high-end CPU (without the integrated GPU).
 
@@
 Nevertheless, in theory, the computational power provided by modern GPUs is five times higher than that provided by CPUs.
-A claimed speed-up of several orders of magnitude is therefore typically a result of comparing a non-optimized CPU implementation with an optimized GPU implementation or by comparing uneven hardware \cite{Lee2010, Kothapalli2013}.
+A claimed speed-up of several orders of magnitude is therefore typically a result of comparing a non-optimized CPU implementation with an optimized GPU implementation or by comparing incomparable hardware \cite{Lee2010, Kothapalli2013}.
 However, in the next section we will see why it is often easier to obtain an optimized implementation for the GPU than for the CPU.
@@
 }
-\caption{The figure depicts how GPU threads are grouped into blocks and arranged in a grid.
+\caption{GPU threads are grouped into blocks and arranged in a grid.
 One thread runs a copy of a kernel function.
-In this example the kernel function performs an in-place element-wise matrix square.}
+In this example the kernel function performs an in-place element-wise matrix square ((b) is from http://docs.nvidia.com/cuda).}
 \label{fig:gpu_grid_kernel}
@@
 While a CPU is said to execute threads which can perform SIMD instructions, a thread on a GPU can be interpreted as a unit capable of processing one data element in the SIMD instruction.
-The GPU is because of this sometimes referred to as a single instruction multiple threads (\nom{SIMT}{Single instruction multiple threads}) processor instead of a SIMD processor.
+Because of this the GPU is sometimes referred to as a single instruction multiple threads (\nom{SIMT}{Single instruction multiple threads}) processor instead of a SIMD processor.
 This also explain why GPU threads are said to be light-weighted.
@@
 Also, to achieve peak performance, the implementation must perform one FMA operation on each core per clock cycle, and it must be possible for the GPU to hide all memory latency by scheduling threads which are not stalled by a memory fetch.
-All of this is something that is not likely to happen in practical applications, and therefore one will always end up with a throughput which is a fraction of the peak throughput.
+All of this is something that is not likely to happen in practical applications, and therefore one will usually end up with a throughput which is a fraction of the peak throughput.
 
@@
 The relevance of auto-applied GPU instructions will also increase due to the recent embedding of GPUs on the same chip as the CPU.
-Soon, GPU programming will probably be a matter of coding a heterogeneous collection of cores where operations are automatically offloaded to the right core (e.g.
+Soon, GPU programming will hopefully be a matter of coding a heterogeneous collection of cores where operations are automatically offloaded to the right core (e.g.
 a GPU) by the compiler.
@@
 This technique is known as \textit{beamforming}, and for ultrasound imaging it is applied both when transmitting and receiving ultrasound pulses.
-A cardiac probe is a phased probe, which means that the image is constructed from image lines which are gradually steered from $-\theta/2$ to $\theta/2$ radians.
-As depicted in Fig.\,\ref{fig:us_grid}, this results in an imaging sector which is $\theta$ radians wide.
+A cardiac probe is a phased probe, which means that the image is constructed from image lines which are gradually steered from $-\theta/2$ to $\theta/2$.
+As depicted in Fig.\,\ref{fig:us_grid}, this results in an imaging sector which is $\theta$ wide.
 This is known as a 2D B-mode image (B for brightness).
@@
 \includegraphics[width=0.8\textwidth]{img/beamforming_das.eps}
-\caption{Delay-and-sum array beamforming.}
+\caption{Delay-and-sum array beamforming. Image courtesy of Jo Inge Buskenes.}
 \label{fig:das_background}
@@
 \subsection{Sampling and processing complexity}
-When generating a B-mode image, we have to insure that the spatial and temporal sampling are sufficient.
+When generating a B-mode image, we have to ensure that the spatial and temporal sampling are sufficient.
 We have already seen that the lateral resolution of an ultrasound probe is given by (\ref{eq:res}).
@@
 \end{align} 
-Hence, a white noise component proportional to the in-coherent signal power is added to the estimate \cite{Featherstone1997b}.
+This corresponds to a white noise component proportional to the in-coherent signal power is added to the estimate \cite{Featherstone1997b}.
 
@@
 Along the ray, samples are gathered from the volume.
-The samples are assigned a color and an opacity before they are combined into one pixel value.
+The samples are assigned a color and an opacity before they are combined into one pixel value (wikipedia.org/wiki/Volume\_ray\_casting).
 b) The opacity transfer function used to determine how opaque a sample should be.}

### 2014-04-27 ThesisJP/Introduction/introduction.tex
Feedback from Jo Inge finished.

@@
 A rapid development in computer game technology and accompanying programming languages have recently provided researchers with small personal super computers, comprised in a single graphics card (\nom{GPU}{Graphics processing unit}).
-The latest graphics cards\footnote{As of January 2014.} from both \nom{Nvidia}{Graphics card producer} and \nom{AMD}{Graphics card producer} provide more than 5 T\nom{FLOPS}{Floating point operations per second} single precision and 1.5 TFLOPS double precision computations.
+The latest graphics cards\footnote{As of January 2014.} from both \nom{Nvidia}{Graphics card vendor} and \nom{AMD}{Graphics card vendor} provide more than 5 T\nom{FLOPS}{Floating point operations per second} single precision and 1.5 TFLOPS double precision computations.
 For double precision, this is more than the world's most powerful supercomputer could provide in 1998, and that was a room full of computers.
@@
 In previous literature on Capon beamforming the matrix inversion has always been regarded as the most computationally complex step.
-In this paper we show that estimation of the sample covariance matrix is the most complex part when spatial and temporal smoothing is included.
+In this paper we show that estimation of the sample covariance matrix is the most complex part when spatial and temporal smoothing is performed with common parameters.
 We then show how the computational complexity of this estimation procedure can be reduced from cubic to square.

### 2014-04-27 UAM2014/Article/article.tex
la2pdf commit - 2014:04:28	00:51:19	(Mon28.Apr)

@@
 \usepackage{subcaption}
+\usepackage{color}
 
@@
 
+\begin{figure}[t]\centering%
+\graphicsAI[drawing,width=\linewidth]{gfx/simulator.svg}%
+\caption{Simulator.}\label{buildup}%
+\end{figure}
+
 OpenGL is a framework that
@@
 \begin{subfigure}[b]{.5\linewidth-\imgspacing/2}
+\graphicsAI[drawing,width=\linewidth]{gfx/sim_cylinder_submerged_adaptive.svg}%
+\caption{Simulated cylinder (adaptive).}\label{sim_cylinder}%
+\end{subfigure}\hspace{\imgspacing}%
+\begin{subfigure}[b]{.5\linewidth-\imgspacing/2}
+\graphicsAI[drawing,width=\linewidth]{gfx/sim_cylinder_submerged_adaptive_bw.svg}%
+\caption{Simulated cylinder (adaptive).}\label{sim_cylinder_bw}%
+\end{subfigure} 
+\end{figure}
+
+\newcommand\cdesc[2]{{\raggedright\setlength\fboxsep{0pt}
+\fbox{\colorbox[HTML]{#1}{\vrule height8.5pt depth3.5pt width0pt\hspace{.5cm}}}\ \ #2\\}}
+\begin{figure}[tp]\centering%
+\begin{subfigure}[b]{.5\linewidth-\imgspacing/2}
 \graphicsAI[drawing,width=\linewidth]{gfx/overlay_cylinder_submerged.svg}%
@@
 \end{subfigure}
-\end{figure}
-
-\begin{figure}[tp]\centering%
-\graphicsAI[drawing,width=\linewidth]{gfx/overlay_colors.svg}%
-\caption{Overlay.}\label{overlay_cylinder}%
+\hfill\parbox{\linewidth-30pt}{
+\cdesc{800000}{Pixels in both template and image highlight segments}
+\cdesc{FF1000}{Pixels only in template highlight segment}
+\cdesc{FFEB00}{Pixels only in image highlight segment}
+\cdesc{83FF7C}{Background pixels}
+\cdesc{00EFFF}{Pixels only in the image shadow segment}
+\cdesc{00A7FF}{Pixels in the image segment, but in template shadow}
+\cdesc{0014FF}{Pixels only in the image template highlight segment}
+\cdesc{000083}{Pixels in both the template and image shadow segments}}
 \end{figure}

### 2014-04-28 UAM2014/Article/article.tex
la2pdf commit - 2014:04:28	01:35:48	(Mon28.Apr)
la2pdf commit - 2014:04:28	11:03:27	(Mon28.Apr)

@@
 
+\definecolor{tabBlue}{HTML}{AACCFF}
 \newlength\oldparindent
@@
 %
-\author{J.I.Buskenes\firstAddress, C.-I.C.Nilsen\secondAddress, A.Austeng\secondAddress}%
+\author{J.I.Buskenes\firstAddress, \O{}ivind Midtgaard\secondAddress, Herman Midelfart\secondAddress}%
 %
@@
 
-A synthetic aperture sonar (SAS) utilizes a wider along-track beam to ensonify the same part of the seafloor with several sucessive pings.
+% A synthetic aperture sonar (SAS) utilizes a wider along-track beam to ensonify the same part of the seafloor with several sucessive pings.
 Later this data is combined coherently to allow the synthetization of a much larger receiving array, which drastically improves the along-track resolution.
@@
 \end{figure}
-
-OpenGL is a framework that
 
@@
 \graphicsAI[drawing,width=\linewidth]{gfx/sim_cylinder_submerged.svg}%
-\caption{Simulated cylinder.}\label{sim_cylinder}%
+\caption{Cylinder simulation with ``typical'' parameters: Length 2.6\;m, burial depth 0.1\;m, aspect angle 100$^\circ$. }\label{sim_cylinder}%
 \end{subfigure}\hspace{\imgspacing}%
@@
 \graphicsAI[drawing,width=\linewidth]{gfx/sim_cylinder_submerged_bw.svg}%
-\caption{Simulated cylinder.}\label{sim_cylinder_bw}%
+\caption{Cylinder simulation with ``typical'' parameters: Length 2.6\;m, burial depth 0.1\;m, aspect angle 100$^\circ$.}\label{sim_cylinder_bw}%
 \end{subfigure} 
@@
 \graphicsAI[drawing,width=\linewidth]{gfx/sim_cylinder_submerged_adaptive.svg}%
-\caption{Simulated cylinder (adaptive).}\label{sim_cylinder}%
+\caption{Adaptive cylinder simulation with parameters estimated from SAS image: Length 2.6\;m, burial depth 0.263\;m, aspect angle 105$^\circ$.}\label{sim_cylinder}%
 \end{subfigure}\hspace{\imgspacing}%
@@
 \graphicsAI[drawing,width=\linewidth]{gfx/sim_cylinder_submerged_adaptive_bw.svg}%
-\caption{Simulated cylinder (adaptive).}\label{sim_cylinder_bw}%
+\caption{Adaptive cylinder simulation with parameters estimated from SAS image: Length 2.6\;m, burial depth 0.263\;m, aspect angle 105$^\circ$.}\label{sim_cylinder_bw}%
 \end{subfigure} 
@@
 
+\begin{table}[bh]
+TODO: Probably not necessary to have this table...\\
+\begin{tabular}[c]{l l l l l}\hline
+\rowcolor{tabBlue} & cor   & ecor  & scor  & essum \\\hline
+Standard template & 0.461 & 0.698 & 0.528 & 0.613 \\
+Adaptive template & 0.673 & 0.818 & 0.807 & 0.813
+\end{tabular}
+\caption{Correlation scores. Description:\newline
+cor - correlation between image and template calculated over all pixels in the image\newline
+ecor - correlation calculated over the union of the echo pixels in the image\newline
+scor - correlation calculated over the union of the echo pixels in image and template\newline
+essum - 0.5*(ecor+scor). Average of the echo and shadows}
+\end{table}
+
+Reference correlation scores~\cite{Midelfart2010}
+
 % Ligger nesten med kortsiden til.

### 2014-04-28 UAM2014/Article/article.tex
la2pdf commit - 2014:04:28	12:12:50	(Mon28.Apr)
Writing UA conference paper

@@
 %
-\author{J.I.Buskenes\firstAddress, \O{}ivind Midtgaard\secondAddress, Herman Midelfart\secondAddress}%
+\author{Jo Inge Buskenes\firstAddress, \O{}ivind Midtgaard\secondAddress, Herman Midelfart\secondAddress}%
 %
 \begin{contact}
-  \firstAddress Dept. of Informatics, P.O.Box 1080 Blindern, N-0316 OSLO
+  \firstAddress Dept. of Informatics, P.O.Box 1080 Blindern, N-0316 OSLO\\
   \secondAddress Norwegian Defence Research Establishment (FFI) P.O.Box 25, NO-2027 Kjeller
@@
 %
-\keywords{Adaptive beamforming, minimum variance, MVDR, LCA, complexity, sonar}
+\keywords{Sonar, SAS, simulator, template, OpenGL}
 
@@
 \begin{figure}[tp]\centering%
+\begin{subfigure}[t]{.5\linewidth-\imgspacing/2}
+\graphicsAI[drawing,width=\linewidth]{gfx/data_cylinder_submerged.svg}%
+\caption{SAS image of a cylinder with 2.6\;m length and 0.53\;m radius.}\label{data_sylinder}%
+\end{subfigure}
+\end{figure}
+
+\begin{figure}[tp]\centering%
+\begin{subfigure}[t]{.5\linewidth-\imgspacing/2}
+\graphicsAI[drawing,width=\linewidth]{gfx/sim_cylinder_submerged.svg}%
+\caption{``Typical'' parameters: Length 2.6\;m, burial depth 0.1\;m, aspect angle 100$^\circ$. }%
+\end{subfigure}\hspace{\imgspacing}%
+\begin{subfigure}[t]{.5\linewidth-\imgspacing/2}
+\graphicsAI[drawing,width=\linewidth]{gfx/sim_cylinder_submerged_adaptive.svg}%
+\caption{Parameters adapted to fit SAS image: Length 2.6\;m, burial depth 0.263\;m, aspect angle 105$^\circ$.}%
+\end{subfigure}
+\caption{Cylinder simulations}\label{sim_cylinder}
+\end{figure}
+
+\newcommand\cdesc[2]{{\raggedright\setlength\fboxsep{0pt}
+\fbox{\colorbox[HTML]{#1}{\vrule height8.5pt depth3.5pt width0pt\hspace{.5cm}}}\ \ #2\\}}
+\begin{figure}[tp]\centering%
+\begin{subfigure}[b]{.5\linewidth-\imgspacing/2}
+\graphicsAI[drawing,width=\linewidth]{gfx/overlay_cylinder_submerged.svg}%
+\caption{Overlay.}\label{overlay_cylinder}%
+\end{subfigure}\hspace{\imgspacing}%
+\begin{subfigure}[b]{.5\linewidth-\imgspacing/2}
+\graphicsAI[width=\linewidth]{gfx/overlay_cylinder_submerged_adaptive.svg}%
+\caption{Overlay (adaptive).}\label{overlay_cylinder_adaptive}%
+\end{subfigure}
+\hfill\parbox{\linewidth-30pt}{
+\cdesc{800000}{Pixels in both template and image highlight segments}
+\cdesc{FF1000}{Pixels only in template highlight segment}
+\cdesc{FFEB00}{Pixels only in image highlight segment}
+\cdesc{83FF7C}{Background pixels}
+\cdesc{00EFFF}{Pixels only in the image shadow segment}
+\cdesc{00A7FF}{Pixels in the image segment, but in template shadow}
+\cdesc{0014FF}{Pixels only in the image template highlight segment}
+\cdesc{000083}{Pixels in both the template and image shadow segments}}
+\end{figure}
+
+\begin{figure}[tp]\centering%
 \begin{subfigure}[b]{.5\linewidth-\imgspacing/2}
@@
 \end{subfigure} 
-\end{figure}
-
-\newcommand\cdesc[2]{{\raggedright\setlength\fboxsep{0pt}
-\fbox{\colorbox[HTML]{#1}{\vrule height8.5pt depth3.5pt width0pt\hspace{.5cm}}}\ \ #2\\}}
-\begin{figure}[tp]\centering%
-\begin{subfigure}[b]{.5\linewidth-\imgspacing/2}
-\graphicsAI[drawing,width=\linewidth]{gfx/overlay_cylinder_submerged.svg}%
-\caption{Overlay.}\label{overlay_cylinder}%
-\end{subfigure}\hspace{\imgspacing}%
-\begin{subfigure}[b]{.5\linewidth-\imgspacing/2}
-\graphicsAI[width=\linewidth]{gfx/overlay_cylinder_submerged_adaptive.svg}%
-\caption{Overlay (adaptive).}\label{overlay_cylinder_adaptive}%
-\end{subfigure}
-\hfill\parbox{\linewidth-30pt}{
-\cdesc{800000}{Pixels in both template and image highlight segments}
-\cdesc{FF1000}{Pixels only in template highlight segment}
-\cdesc{FFEB00}{Pixels only in image highlight segment}
-\cdesc{83FF7C}{Background pixels}
-\cdesc{00EFFF}{Pixels only in the image shadow segment}
-\cdesc{00A7FF}{Pixels in the image segment, but in template shadow}
-\cdesc{0014FF}{Pixels only in the image template highlight segment}
-\cdesc{000083}{Pixels in both the template and image shadow segments}}
 \end{figure}

### 2014-04-29 UAM2014/Article/article.tex
la2pdf commit - 2014:04:28	14:42:31	(Mon28.Apr)
la2pdf commit - 2014:04:28	15:44:10	(Mon28.Apr)
...
la2pdf commit - 2014:04:29	10:02:13	(Tue29.Apr)

@@
 %   Parameters for TEXT pages (not float pages):
-\setcounter{topnumber}{2}
+\setcounter{topnumber}{1}
 \setcounter{bottomnumber}{2}
@@
 
+We have developed fast a synthetic aperture sonar (SAS) simulator that run on graphics processing units (GPUs)  using a combination of OpenGL for graphics processing and OpenCL for post-processing.
+This implementation resembles the SIGMAS+ simulator developed at the NATO Undersea Research Centre (NURC).
+They both take advantage of graphics processing to render a 2D image from 3D models and assume a Lambertian scattering model.
+
 \begin{figure}[t]\centering%
@@
 \section{Methods}
+
+The basic idea behind using a graphics
+
+OpenGL is a multi-platform application programming interface (API) for rendering 2D and 3D vector graphics.
 
@@
 
+\newpage
+
 \section{Results}
@@
 \newlength\imgspacing\setlength\imgspacing{.5cm}
+
+The simulator performance were tested on an image from the HISAS 1030 synthetic aperture sonar mounted on the HUGIN AUV.
+Out of the objects in the scene we selected a 2.6\;m cylinder that was slightly buried in the sea sediments (Fig.
+\ref{data_sylinder}).
 
@@
 \end{subfigure}
+\begin{subfigure}[t]{.5\linewidth-\imgspacing/2}
+\graphicsAI[drawing,width=\linewidth]{gfx/sim_cylinder_submerged_adaptive.svg}%
+\caption{Cylinder simulation with parameters adapted to fit the SAS image: Length 2.6\;m, burial depth 0.263\;m, aspect angle 105$^\circ$.}%
+\end{subfigure}
+\caption{}
 \end{figure}
+
+\setcounter{topnumber}{2}
 

### 2014-04-29 UAM2014/Article/article.tex
la2pdf commit - 2014:04:29	11:10:06	(Tue29.Apr)

--- 2014-04-28 UAM2014/Article/article.tex
@@
 
+% To obtain optimal classification performance using template matching techniques a large set of templates must be availab
+
+We have developed a simulator for synthetic aperture sonar (SAS) that creates object templates very quickly with the aid of the massive computing power in graphics processing units (GPUs).
+It uses OpenGL to render a optical 2D image and depth map from the 3D model and seafloor representation, and OpenCL compute a sonar template based on this.
+
+By coincidence this implementation is not so different from the SIGMAS+ simulator developed at the NATO Undersea Research Centre (NURC).
+They both take advantage of graphics processing to render a 2D image from 3D models, are designed for side-looking sonar, and assume a Lambertian scattering model.
+However, SIGMAS+ is not designed for creating object templates, and adds e.g.
+noise to the images for them to more closely
+
 \begin{figure}[t]\centering%
@@
 \section{Methods}
+
+OpenGL is a popular, well-matured, multi-platform application programming interface (API) for rendering 2D and 3D vector graphics.
 

### 2014-04-29 UAM2014/Article/article.tex
la2pdf commit - 2014:04:29	11:52:05	(Tue29.Apr)

@@
 We have developed a simulator for synthetic aperture sonar (SAS) that creates object templates very quickly with the aid of the massive computing power in graphics processing units (GPUs).
-It uses OpenGL to render a optical 2D image and depth map from the 3D model and seafloor representation, and OpenCL compute a sonar template based on this.
+It uses OpenGL to render a optical 2D image and depth map of scene with a custom 3D model of the object and seafloor, and OpenCL to post-process this data to form sonar templates.
 
-By coincidence this implementation is not so different from the SIGMAS+ simulator developed at the NATO Undersea Research Centre (NURC).
-They both take advantage of graphics processing to render a 2D image from 3D models, are designed for side-looking sonar, and assume a Lambertian scattering model.
-However, SIGMAS+ is not designed for creating object templates, and adds e.g.
-noise to the images for them to more closely
+By coincidence our implementation is not so different from the SIGMAS+ simulator developed at the NATO Undersea Research Centre (NURC)~\cite{Coiras2009a, Coiras2009b}.
+They both take advantage of OpenGL to render a 2D image from 3D models, are designed for side-looking sonar, and assume a Lambertian scattering model.
+However, while SIGMAS+ is geared more towards realistically looking sonar images and includes effects such as noise, we do not.
+Also, SIGMAS+ sum images obtained from multiple OpenGL rendering passes to create the sonar image, while we render once and tailor the sonar image with the flexibility of OpenCL.
+
+does not use OpenCL but a OpenGL a multi-pass technique, and includes ting  W
+is not designed for creating object templates, and adds e.g. noise to the images for them to more closely
 
@@
 \item Simulation of sidescan images (Bell thesis \cite{Bell1997}, pre-work by V.Blake \cite{Blake1993})
-\item NURC SIGMAS+ (Coiras GPU article at OCEANS 2009 \cite{Coiras2009a}, Coiras on SIGMAS maths \cite{Coiras2009b})
 \end{itemize}

### 2014-04-29 UAM2014/Article/article.tex
la2pdf commit - 2014:04:29	12:41:09	(Tue29.Apr)

@@
 We have developed a simulator for synthetic aperture sonar (SAS) that creates object templates very quickly with the aid of the massive computing power in graphics processing units (GPUs).
-It uses OpenGL to render a optical 2D image and depth map of scene with a custom 3D model of the object and seafloor, and OpenCL to post-process this data to form sonar templates.
+It loads a 3D model of the sea floor and an object of interest into OpenGL, where emitted sound waves are modeled with a light source placed at the sonar transmit location.
+The backscattered sound intensity is assumed to follow a Lambertian model, i.e.
+only being dependent on incidence angle of the transmitted signal onto the model surface.
+After rendering OpenGL is set up to produce an optical 2D intensity image and a map of the distance from the propagation axis to each pixel in the image.
+This data is finally combined with OpenCL to produce the image templates.
 

### 2014-04-29 UAM2014/Article/article.tex
la2pdf commit - 2014:04:29	14:02:09	(Tue29.Apr)

--- 2014-04-29 UAM2014/Article/article.tex
@@
 We have developed a simulator for synthetic aperture sonar (SAS) that creates object templates very quickly with the aid of the massive computing power in graphics processing units (GPUs).
-It uses OpenGL to render a optical 2D image and depth map of scene with a custom 3D model of the object and seafloor, and OpenCL to post-process this data to form sonar templates.
+It loads a 3D model of the sea floor and an object of interest into OpenGL, where emitted sound waves are modeled with a light source placed at the sonar transmit location.
+We assume rough, isotropic surfaces that reflects sound energy equally in all directions.
+This can be modeled with a Lambertian scattering model~\cite{Bell1997} where the intensity of the backscattered sound is said to depend only on the on incidence angle of the transmitted signal onto the model surface.
+When rendering OpenGL is set up to produce an optical 2D intensity image and depth map that reveals the distance from each pixels to the propagation axis.
+For maximum flexibility this data is finally combined with OpenCL to produce the image templates.
 
@@
 They both take advantage of OpenGL to render a 2D image from 3D models, are designed for side-looking sonar, and assume a Lambertian scattering model.
-However, while SIGMAS+ is geared more towards realistically looking sonar images and includes effects such as noise, we do not.
+However, while SIGMAS+ is geared more towards realistically looking sonar images by including effects such as noise, we do not.
 Also, SIGMAS+ sum images obtained from multiple OpenGL rendering passes to create the sonar image, while we render once and tailor the sonar image with the flexibility of OpenCL.
-
-does not use OpenCL but a OpenGL a multi-pass technique, and includes ting  W
-is not designed for creating object templates, and adds e.g. noise to the images for them to more closely
 
@@
 
-Citations
 \begin{itemize}
@@
 
-OpenGL is a popular, well-matured, multi-platform application programming interface (API) for rendering 2D and 3D vector graphics.
+Creating a 2D view of an arbitrary complex 3D scene is a non-trivial matter.
+This is why we decided to use the core OpenGL pipeline to do this for us.
+OpenGL is a popular, well-matured, and multi-platform application programming interface (API) for rendering 2D and 3D vector graphics.
+It relieves us from the intricacies of handling vertices, faces and textures, as well as from projecting this onto a suitable 2D image plane.
+This section explains how we set up OpenGL to do this for us, and then proceeds to describe how we post process the OpenGL images with OpenCL to form the final sonar image.
 
@@
 It does not consider observation angle or sound frequency\todo{is this unproblematic?}.
+
+If we place the camera at the sonar and look in the direction of the image OpenGL will generate a image of all the pixels that are ``seen'' from the sonar.
+However, this can only provide us with optically looking images, while was not designed to does not produce camera where the sonar is and render the scene   handling computing the projection of a  and rather post-process its result with OpenCL to obtain
 

### 2014-04-29 UAM2014/Article/article.tex
More on UAC2014

@@
 They both take advantage of OpenGL to render a 2D image from 3D models, are designed for side-looking sonar, and assume a Lambertian scattering model.
-However, while SIGMAS+ is geared more towards realistically looking sonar images by including effects such as noise, we do not.
-Also, SIGMAS+ sum images obtained from multiple OpenGL rendering passes to create the sonar image, while we render once and tailor the sonar image with the flexibility of OpenCL.
+However, while SIGMAS+ is geared more towards realistically looking sonar images, and include effects such as noise, we do not.
+Also, their simulator sums images obtained from multiple OpenGL rendering passes to create the sonar image, while we render once and post-process this result with OpenCL to tailor the sonar image full flexibility.
 
@@
 
-Before we can do anything with OpenGL we first have to set up the scene. This involves loading the models and then placing them where we want them with the proper orientation.
+Before we can render anything OpenGL we first have to set up the scene. This involves loading the models and then placing them where we want them with the proper orientation.
 
@@
 \end{subfigure}
-\caption{}
+\caption{A SAS image of a cylinder and a template simulation adapted to it.}
 \end{figure}
@@
 \setcounter{topnumber}{2}
-
-\begin{figure}[tp]\centering%
-\begin{subfigure}[t]{.5\linewidth-\imgspacing/2}
-\graphicsAI[drawing,width=\linewidth]{gfx/sim_cylinder_submerged.svg}%
-\caption{``Typical'' parameters: Length 2.6\;m, burial depth 0.1\;m, aspect angle 100$^\circ$. }%
-\end{subfigure}\hspace{\imgspacing}%
-\begin{subfigure}[t]{.5\linewidth-\imgspacing/2}
-\graphicsAI[drawing,width=\linewidth]{gfx/sim_cylinder_submerged_adaptive.svg}%
-\caption{Parameters adapted to fit SAS image: Length 2.6\;m, burial depth 0.263\;m, aspect angle 105$^\circ$.}%
-\end{subfigure}
-\caption{Cylinder simulations}\label{sim_cylinder}
-\end{figure}
 
@@
 \cdesc{000083}{Pixels in both the template and image shadow segments}}
-\end{figure}
-
-\begin{figure}[tp]\centering%
-\begin{subfigure}[b]{.5\linewidth-\imgspacing/2}
-\graphicsAI[drawing,width=\linewidth]{gfx/data_cylinder_submerged.svg}%
-\caption{SAS image of a cylinder with 2.6\;m length and 0.53\;m radius.}\label{data_sylinder}%
-\end{subfigure}\hspace{\imgspacing}%
-\begin{subfigure}[b]{.5\linewidth-\imgspacing/2}
-\graphicsAI[drawing,width=\linewidth]{gfx/data_cylinder_submerged_bw.svg}%
-\caption{SAS image of a cylinder with 2.6\;m length and 0.53\;m radius.}\label{data_cylinder_bw}%
-\end{subfigure} 
-\end{figure}
-
-\begin{figure}[tp]\centering%
-\begin{subfigure}[b]{.5\linewidth-\imgspacing/2}
-\graphicsAI[drawing,width=\linewidth]{gfx/sim_cylinder_submerged.svg}%
-\caption{Cylinder simulation with ``typical'' parameters: Length 2.6\;m, burial depth 0.1\;m, aspect angle 100$^\circ$. }\label{sim_cylinder}%
-\end{subfigure}\hspace{\imgspacing}%
-\begin{subfigure}[b]{.5\linewidth-\imgspacing/2}
-\graphicsAI[drawing,width=\linewidth]{gfx/sim_cylinder_submerged_bw.svg}%
-\caption{Cylinder simulation with ``typical'' parameters: Length 2.6\;m, burial depth 0.1\;m, aspect angle 100$^\circ$.}\label{sim_cylinder_bw}%
-\end{subfigure} 
-\end{figure}
-
-\begin{figure}[tp]\centering%
-\begin{subfigure}[b]{.5\linewidth-\imgspacing/2}
-\graphicsAI[drawing,width=\linewidth]{gfx/sim_cylinder_submerged_adaptive.svg}%
-\caption{Adaptive cylinder simulation with parameters estimated from SAS image: Length 2.6\;m, burial depth 0.263\;m, aspect angle 105$^\circ$.}\label{sim_cylinder}%
-\end{subfigure}\hspace{\imgspacing}%
-\begin{subfigure}[b]{.5\linewidth-\imgspacing/2}
-\graphicsAI[drawing,width=\linewidth]{gfx/sim_cylinder_submerged_adaptive_bw.svg}%
-\caption{Adaptive cylinder simulation with parameters estimated from SAS image: Length 2.6\;m, burial depth 0.263\;m, aspect angle 105$^\circ$.}\label{sim_cylinder_bw}%
-\end{subfigure} 
 \end{figure}

### 2014-04-30 UAM2014/Article/article.tex
la2pdf commit - 2014:04:30	11:03:01	(Wed30.Apr)

@@
 
-\section{Results}
+\section{Results \& Discussion}
 
@@
 
-The simulator performance were tested on an image from the HISAS 1030 synthetic aperture sonar mounted on the HUGIN AUV.
-Out of the objects in the scene we selected a 2.6\;m cylinder that was slightly buried in the sea sediments (Fig.
-\ref{data_sylinder}).
+The main advantage of a GPU-based simulator over a standard simulator is its ability to create templates incredibly fast.
+This makes it possible to tailor the templates more closely to an image as the parameters needed by the simulator can be directly estimated from the image.
+In a standard approach, on the other hand, a template library must be created beforehand from a limited set of parameter values.
+Hence, it may happen that no template in the library is a close fit of the object in the image even though the object actually is a target of interest.
+
+This performance was demonstrated on an image from the HISAS 1030 synthetic aperture sonar mounted on the HUGIN AUV.
+Out of the objects in the scene we selected a 2.6\;m long cylinder that was slightly buried in the sea sediments (Fig.
+\ref{data_cylinder}).
+The length, burial depth, and aspect angle of this cylinder were estimated from the SAS image using our  adaptive template matching approach~\cite{Midelfart2010}.
+Then these parameters were used by the simulator to create an adaptive template (Fig.
+\ref{sim_cylinder}), which is an almost a perfect match of the cylinder.
+We also created standard templates for a regular template matching approach.
+In this case, we assumed that the cylinder mine had a length of 2\;m (like a MP80 or a Murena mine) and was proud on the seafloor.
+Moreover, templates were created for very 10$^\text{th}$ degree of the aspect angle.
+We believe the assumptions are quite typical for a template library for cylinder mines.
+
+The standard and adaptive templates were then matched to the HISAS image.
+This is illustrated in Fig.
+4a for a standard template and in Fig.
+4b for an adaptive template.
+These images were created by segmenting the image and the templates into highlight, shadow, and background regions.
+The regions from a template were then laid on top of the regions of the image to illustrate how well these regions matched.
+From these figures, it was quite clear the adaptive template obtained a much closer fit to image than the standard template.
+This was also reflected in the correlation scores (which were created with the method described in \cite{Midelfart2010}) that were 0.613 for the standard template and 0.813 for the adaptive template.
+Hence, the standard approach was less likely to classify the cylinder in the image correctly as the standard templates were not created specifically for the target.
 
@@
 \graphicsAI[drawing,width=\linewidth]{gfx/data_cylinder_submerged.svg}%
-\caption{SAS image of a cylinder with 2.6\;m length and 0.53\;m radius.}\label{data_sylinder}%
+\caption{SAS image of a cylinder with 2.6\;m length and 0.53\;m radius.}\label{data_cylinder}%
 \end{subfigure}
@@
 \graphicsAI[drawing,width=\linewidth]{gfx/sim_cylinder_submerged_adaptive.svg}%
-\caption{Cylinder simulation with parameters adapted to fit the SAS image: Length 2.6\;m, burial depth 0.263\;m, aspect angle 105$^\circ$.}%
+\caption{Cylinder simulation with parameters adapted to fit the SAS image: Length 2.6\;m, burial depth 0.263\;m, aspect angle 105$^\circ$.}\label{sim_cylinder}%
 \end{subfigure}

### 2014-04-30 UAM2014/Article/article.tex
la2pdf commit - 2014:04:30	11:34:56	(Wed30.Apr)

@@
 \begin{abstract}
-Template matching is a common technique used when classifying objects in synthetic aperture sonar (SAS) images.
-The principle is to isolate an image segment containing an object of interest, correlate it with a set of template images, and assign it to the class of the template yielding the highest correlation coefficient.
-The challenge is to come up with a suitable set of template images considering that no seabed or object is truly alike.
+Template matching is a common technique for automatic classification of objects in synthetic aperture sonar (SAS) images.
+The principle is to isolate an image segment containing the object of interest, correlate it with a set of template images, and assign it to the class of the template yielding the highest correlation coefficient.
+The challenge is to come up with a representative set of template images covering the actual configurations of object and seabed.
 
-We target this challenge with a sonar simulator that first take as input a seabed model derived from the real sonar image.
-Then it places an object model on the seabed, renders the scene, and adds the resulting image to the template set.
-For any object position, alignment, type and material, the procedure is repeated, and a correlation coefficient computed.
-The faster we are able to perform these simulations, the better we can expect the classification result to be.
-Therefore the simulator is written in OpenGL and OpenCL and run on graphics processing units (GPUs).
-
-The result is a fast performing, mobile and portable on-the-fly template generator which can adapt its behavior to the nature of the current scene.
-We believe this can prove a powerful tool for mobile sonar imaging platforms such as autonomous underwater vehicles (AUVs).
+We target this challenge with a sonar simulator that first takes as input a seabed model derived from the real sonar image.
+Then it places a 3D object model on the seabed, renders the scene, and adds the resulting image to the template set.
+For any object type, position, alignment and material, the procedure is repeated, and a correlation coefficient computed.
+Simulation speed is fundamental when templates are created as part of the classification process.
+The simulator is therefore written in OpenGL and OpenCL and runs on graphics processing units (GPUs).
+The result is a fast performing and portable on-the-fly template generator which can adapt to the characteristics of the current scene.
 \end{abstract}%
@@
 
-% A synthetic aperture sonar (SAS) utilizes a wider along-track beam to ensonify the same part of the seafloor with several sucessive pings.
-Later this data is combined coherently to allow the synthetization of a much larger receiving array, which drastically improves the along-track resolution.
+Automatic target recognition (ATR) is a fundamental component to optimize the use of autonomous underwater vehicles (AUVs), as it allows the vehicle to adapt its mission plan to e.g.
+revisit detected objects for closer examination.
+The ATR algorithms must, however, be sufficiently fast to process the sensor data onboard the AUV in delayed real-time.
 
-% To obtain optimal classification performance using template matching techniques a large set of templates must be availab
+Template matching is a common technique for classification of objects in synthetic aperture sonar (SAS) images.
+The principle is to isolate an image segment containing the object of interest, correlate it with a set of template images of the relevant object classes, and assign it to the class of the template yielding the highest correlation coefficient.
+The challenge is to come up with a representative set of template images covering  the actual configurations of object and seabed.
+The use of a static, preset template library may be impractical and/or yield inaccurate results [Midelfart 4].
+
+We have developed a sonar simulator that runs sufficiently fast to create templates adapted to the actual scene as part of the classification process.
+This is achieved with the aid of the massive computing power in graphics processing units (GPUs) and optimized software libraries for scene rendering.
+The simulator loads a 3D model of the seafloor and an object class model into OpenGL, where emitted sound waves are modeled with a light source placed at the sonar transmit location.
+We assume rough, isotropic surfaces reflecting sound energy equally in all directions.
+This can be modeled with a Lambertian scattering model [1], where the intensity of the backscattered sound only depends on the incidence angle of the transmitted signal onto the model surface.
+When rendering OpenGL is set up to produce an optical 2D intensity image and depth map that reveals the distance from each pixel to the propagation axis.
+For maximum flexibility this data is finally combined with OpenCL to produce the image templates.
+
+Various simulators for high frequency, side-looking sonar imagery have been published, e.g.
+[Bell][Sammelmann].
+Most implementations have however prioritized accurate acoustic modeling at the expense of execution speed.
+One exception is the SIGMAS+ simulator in [2,3] which, similar to our approach, takes advantage of parallel processing at GPUs and uses OpenGL to render a 2D image from a 3D model, under the assumption of Lambertian scattering.
+However, SIGMAS+ is still aimed more towards realistically looking images, including effects like ambient noise, which is irrelevant for our purpose of template generation.
+Also, their simulator sums images obtained from multiple OpenGL rendering passes to create the sonar image, while we render once and post-process this result with OpenCL to tailor the sonar image full flexibility.
 
@@
 
-\newpage
-
 \section{Results \& Discussion}
@@
 
-\begin{figure}[tp]\centering%
+\setcounter{topnumber}{2}
+
+\begin{figure}[p]\centering%
 \begin{subfigure}[t]{.5\linewidth-\imgspacing/2}
@@
 
-\setcounter{topnumber}{2}
-
 \newcommand\cdesc[2]{{\raggedright\setlength\fboxsep{0pt}
 \fbox{\colorbox[HTML]{#1}{\vrule height8.5pt depth3.5pt width0pt\hspace{.5cm}}}\ \ #2\\}}
-\begin{figure}[tp]\centering%
-\begin{subfigure}[b]{.5\linewidth-\imgspacing/2}
+\begin{figure}[p]\centering%
+\begin{subfigure}[t]{.5\linewidth-\imgspacing/2}
 \graphicsAI[drawing,width=\linewidth]{gfx/overlay_cylinder_submerged.svg}%
-\caption{Overlay.}\label{overlay_cylinder}%
+\caption{Standard template. Predefine template's cylinder length, alignment and burial depth do no match. Shadow and highlight regions to not correspond well.}\label{overlay_cylinder}%
 \end{subfigure}\hspace{\imgspacing}%
-\begin{subfigure}[b]{.5\linewidth-\imgspacing/2}
+\begin{subfigure}[t]{.5\linewidth-\imgspacing/2}
 \graphicsAI[width=\linewidth]{gfx/overlay_cylinder_submerged_adaptive.svg}%
-\caption{Overlay (adaptive).}\label{overlay_cylinder_adaptive}%
+\caption{Adapted template. Template simulated with adapted parameters. Good fit in both highlight and shadow regions.}\label{overlay_cylinder_adaptive}%
 \end{subfigure}
-\hfill\parbox{\linewidth-30pt}{
-\cdesc{800000}{Pixels in both template and image highlight segments}
-\cdesc{FF1000}{Pixels only in template highlight segment}
-\cdesc{FFEB00}{Pixels only in image highlight segment}
+\hfill\parbox{\linewidth-0pt}{\small\centering
+Segmentation overlay color description:\\[.5\baselineskip]
+\begin{minipage}{.49\linewidth}
+\cdesc{800000}{Template highlight and image highlight}
+\cdesc{FF1000}{Template highlight only}
+\cdesc{FFEB00}{Image highlight only}
 \cdesc{83FF7C}{Background pixels}
-\cdesc{00EFFF}{Pixels only in the image shadow segment}
-\cdesc{00A7FF}{Pixels in the image segment, but in template shadow}
-\cdesc{0014FF}{Pixels only in the image template highlight segment}
-\cdesc{000083}{Pixels in both the template and image shadow segments}}
+\end{minipage}%
+\begin{minipage}{.49\linewidth}
+\cdesc{000083}{Template shadow and image shadow}
+\cdesc{0014FF}{Template shadow only}
+\cdesc{00EFFF}{Image shadow only}
+\cdesc{00A7FF}{Template shadow and image highlight}
+\end{minipage}}
+\vspace{\baselineskip}
+\caption{Segmented simulation image overlaid the segmented SAS image.}
 \end{figure}
@@
 % 
-\newpage
 \bibliography{references}

### 2014-04-30 UAM2014/Article/article.tex
la2pdf commit - 2014:04:30	12:09:07	(Wed30.Apr)

@@
 %
-\author{Jo Inge Buskenes\firstAddress, \O{}ivind Midtgaard\secondAddress, Herman Midelfart\secondAddress}%
+\author{Jo Inge Buskenes\firstAddress, Herman Midelfart\secondAddress, \O{}ivind Midtgaard\secondAddress}%
 %
@@
 The principle is to isolate an image segment containing the object of interest, correlate it with a set of template images of the relevant object classes, and assign it to the class of the template yielding the highest correlation coefficient.
-The challenge is to come up with a representative set of template images covering  the actual configurations of object and seabed.
-The use of a static, preset template library may be impractical and/or yield inaccurate results [Midelfart 4].
+The challenge is to come up with a representative set of template images covering the actual configurations of object and seabed\todo{objects?
+seabeds?}.
+The use of a static, predetermined template library may be impractical, yield inaccurate results, or both~\cite{Midelfart2010}.
 
@@
 We assume rough, isotropic surfaces reflecting sound energy equally in all directions.
-This can be modeled with a Lambertian scattering model [1], where the intensity of the backscattered sound only depends on the incidence angle of the transmitted signal onto the model surface.
+This can be modeled with a Lambertian scattering model~\cite{Blake1993,Bell1995}, where the intensity of the backscattered sound only depends on the incidence angle of the transmitted signal onto the model surface.
 When rendering OpenGL is set up to produce an optical 2D intensity image and depth map that reveals the distance from each pixel to the propagation axis.
@@
 
-Various simulators for high frequency, side-looking sonar imagery have been published, e.g.
-[Bell][Sammelmann].
+Various simulators for high frequency, side-looking sonar imagery have been published, e.g.~\cite{Bell1997,Sammelmann2004}.
 Most implementations have however prioritized accurate acoustic modeling at the expense of execution speed.
-One exception is the SIGMAS+ simulator in [2,3] which, similar to our approach, takes advantage of parallel processing at GPUs and uses OpenGL to render a 2D image from a 3D model, under the assumption of Lambertian scattering.
+One exception is the SIGMAS+ simulator in~\cite{Coiras2009a, Coiras2009b} which, similar to our approach, takes advantage of parallel processing at GPUs and uses OpenGL to render a 2D image from a 3D model, under the assumption of Lambertian scattering.
 However, SIGMAS+ is still aimed more towards realistically looking images, including effects like ambient noise, which is irrelevant for our purpose of template generation.
-Also, their simulator sums images obtained from multiple OpenGL rendering passes to create the sonar image, while we render once and post-process this result with OpenCL to tailor the sonar image full flexibility.
+Also, their simulator sums images obtained from multiple OpenGL rendering passes to create the sonar image, while we render once and post-process this result with OpenCL to tailor the sonar image with more flexibility.
 
-We have developed a simulator for synthetic aperture sonar (SAS) that creates object templates very quickly with the aid of the massive computing power in graphics processing units (GPUs).
+% We have developed a simulator for synthetic aperture sonar (SAS) that creates object templates very quickly with the aid of the massive computing power in graphics processing units (GPUs).
 It loads a 3D model of the sea floor and an object of interest into OpenGL, where emitted sound waves are modeled with a light source placed at the sonar transmit location.
@@
 For maximum flexibility this data is finally combined with OpenCL to produce the image templates.
-
-By coincidence our implementation is not so different from the SIGMAS+ simulator developed at the NATO Undersea Research Centre (NURC)~\cite{Coiras2009a, Coiras2009b}.
+% 
+% By coincidence our implementation is not so different from the SIGMAS+ simulator developed at the NATO Undersea Research Centre (NURC)~\cite{Coiras2009a, Coiras2009b}.
 They both take advantage of OpenGL to render a 2D image from 3D models, are designed for side-looking sonar, and assume a Lambertian scattering model.
@@
 
-\begin{itemize}
-\item Midelfart/Midtgaard: Adaptive templates \cite{Midelfart2010}
-\item Simulation of sidescan images (Bell thesis \cite{Bell1997}, pre-work by V.Blake \cite{Blake1993})
-\end{itemize}
-
 \section{Methods}
@@
 OpenGL is a popular, well-matured, and multi-platform application programming interface (API) for rendering 2D and 3D vector graphics.
-It relieves us from the intricacies of handling vertices, faces and textures, as well as from projecting this onto a suitable 2D image plane.
+It relieves us from the intricacies of projecting vertices, faces and textures defined in a 3D space onto a suitable 2D image plane.
 This section explains how we set up OpenGL to do this for us, and then proceeds to describe how we post process the OpenGL images with OpenCL to form the final sonar image.
@@
 This is a portable open source library that supports loading a wide range of 3D model formats in a uniform manner.
-It handles the intricacies of the various model formats and outputs the model as a node tree.
-Each node contains data such as vertices, facets and textures, and since they are formatted in an OpenGL friendly way we simply iterate through the nodes and load it.
+Given a model file it outputs the model as a node tree where node contains data such as vertices, facets and textures, all formatted in an OpenGL friendly way.
 
@@
 This permits us to use a Lambertian scattering model where the backscatter intensity depends only on the incidence angle~\cite{Zhang1999}.
-It does not consider observation angle or sound frequency\todo{is this unproblematic?}.
+It does not consider observation angle or sound frequency, but the purpose of creating templates this is needed.
 
@@
 
-\begin{table}[bh]
-TODO: Probably not necessary to have this table...\\
-\begin{tabular}[c]{l l l l l}\hline
-\rowcolor{tabBlue} & cor   & ecor  & scor  & essum \\\hline
-Standard template & 0.461 & 0.698 & 0.528 & 0.613 \\
-Adaptive template & 0.673 & 0.818 & 0.807 & 0.813
-\end{tabular}
-\caption{Correlation scores. Description:\newline
-cor - correlation between image and template calculated over all pixels in the image\newline
-ecor - correlation calculated over the union of the echo pixels in the image\newline
-scor - correlation calculated over the union of the echo pixels in image and template\newline
-essum - 0.5*(ecor+scor). Average of the echo and shadows}
-\end{table}
-
-Reference correlation scores~\cite{Midelfart2010}
-
 % Ligger nesten med kortsiden til.

### 2014-04-30 UAM2014/Article/article.tex
la2pdf commit - 2014:04:30	12:50:38	(Wed30.Apr)

@@
 This permits us to use a Lambertian scattering model where the backscatter intensity depends only on the incidence angle~\cite{Zhang1999}.
-It does not consider observation angle or sound frequency, but the purpose of creating templates this is needed.
+It does not consider observation angle or sound frequency, but for the purpose of creating templates this is needed.
 
@@
 
-The main advantage of a GPU-based simulator over a standard simulator is its ability to create templates incredibly fast.
+Whenever conventional computer graphics processing can be used to 
+GPUs are designed to provide the best possible graphics processing performance, and there large ecosystem of tools and libraries  available 
+Using a GPU to drive a simulator is advantageous since
+
+Using a GPU-based simulator is attractive since computer is an attractive solution as there are a  graphics traits.
+For one over a CPU-based standard simulator is two-fold: First is its ability to create templates incredibly fast.
 This makes it possible to tailor the templates more closely to an image as the parameters needed by the simulator can be directly estimated from the image.
@@
 This is illustrated in Fig.
-4a for a standard template and in Fig.
-4b for an adaptive template.
+\ref{overlay_cylinder} for a standard template and in Fig.
+\ref{overlay_cylinder_adaptive} for an adaptive template.
 These images were created by segmenting the image and the templates into highlight, shadow, and background regions.
@@
 
+\section{Conclusion}
+
+Modern GPUs are customized to perform computer graphics.
+Their theoretical peak performance is typically and order of magnitude higher than a CPU in a comparable price range.
+This performance can not always be achieved in practice, unless the algorithm can be adapted to use a well matured and optimized graphics processing API such as OpenGL.
+Our simulator uses OpenGL for most of the , then it makes little
+
+Whether this performance can be achieved in practice 
+Whenever conventional computer graphics processing can be used to 
+GPUs are designed to provide the best possible graphics processing performance, and there large ecosystem of tools and libraries  available 
+Using a GPU to drive a simulator is advantageous since
+
 % Ligger nesten med kortsiden til.

### 2014-04-30 UAM2014/Article/article.tex
la2pdf commit - 2014:04:30	13:37:52	(Wed30.Apr)

@@
 
+One way to automatically classify objects in SAS images is compare the imaged objects with a predefined set of templates.
+These are typically simple simulations of common objects with various sizes and shapes.
+With luck one might be a close match, but
+
+Template matching is a common technique for automatic classification of objects in synthetic aperture sonar (SAS) images.
+
+We have implemented a SAS simulator that creates object templates in delayed real-time based on parameters estimated from the current scene.
+Compared to the conventional method  having a predefined set of object templates
+
+, and not just a  on-the-fly based on adapted parameters, and not just a predefined set. 
 Modern GPUs are customized to perform computer graphics.
 Their theoretical peak performance is typically and order of magnitude higher than a CPU in a comparable price range.
-This performance can not always be achieved in practice, unless the algorithm can be adapted to use a well matured and optimized graphics processing API such as OpenGL.
-Our simulator uses OpenGL for most of the , then it makes little
-
-Whether this performance can be achieved in practice 
-Whenever conventional computer graphics processing can be used to 
-GPUs are designed to provide the best possible graphics processing performance, and there large ecosystem of tools and libraries  available 
-Using a GPU to drive a simulator is advantageous since
+While it is hard to obtain this performance in practice, one can get very close by using a well matured and highly optimized graphics processing API such as OpenGL.
+Because of this we use OpenGL for simulating a scene with 3D models of the seabed and objects.
 

### 2014-04-30 UAM2014/Article/article.tex
la2pdf commit - 2014:04:30	14:15:28	(Wed30.Apr)

--- 2014-04-30 UAM2014/Article/article.tex
@@
 
-\subsection{OpenGL Rendering}
+\subsection{OpenGL rendering}
 
@@
 
-If we place the camera at the sonar and look in the direction of the image OpenGL will generate a image of all the pixels that are ``seen'' from the sonar.
-However, this can only provide us with optically looking images, while was not designed to does not produce camera where the sonar is and render the scene   handling computing the projection of a  and rather post-process its result with OpenCL to obtain
+The rendered image will appear as if we placed a window at the sonar and looked through it in the direction of the image.
+This window is resized to make sure that the template image perfectly fills it.
 
-When our models loaded into the OpenGL scene we are ready to render.
+\subsection{OpenCL post-processing}
 
-It is also possible to compute the sonar image using the OpenGL ``blend'' feature.
+The image rendered with OpenGL is not in $(x,y)$ coordinates as we want it to be.
+It does, however, tell us what parts of the scene that is visible from the sonar.
+In addition, OpenGL can produce a depth map that reveals the distance to each template pixel.
+This information can be converted to a ranged sonar-like image by simply adding up all the intensity values that share the same depth for each range line.
+We perform this computation in OpenCL, which allow general purpose programming on GPUs and can interoperate with OpenGL quite nicely.
+This way we keep all the calculations on the GPU.
+
+% It is also possible to compute the sonar image using the OpenGL ``blend'' feature.
 
@@
 
-Modern GPUs are customized to perform computer graphics.
-Their theoretical peak performance is typically and order of magnitude higher than a CPU in a comparable price range.
-This performance can not always be achieved in practice, unless the algorithm can be adapted to use a well matured and optimized graphics processing API such as OpenGL.
-Our simulator uses OpenGL for most of the , then it makes little
+One way to automatically classify objects in SAS images is compare the imaged objects with a predefined set of templates.
+However, this is suboptimal as it is infeasible to create templates for all configurations of seabeds and objects.
+Instead we have implemented a SAS simulator that creates the templates in delayed real-time based on parameters estimated from the current scene.
+In our studies this improves the match between the SAS objects and the template considerably.
 
-Whether this performance can be achieved in practice 
-Whenever conventional computer graphics processing can be used to 
-GPUs are designed to provide the best possible graphics processing performance, and there large ecosystem of tools and libraries  available 
-Using a GPU to drive a simulator is advantageous since
+To obtain the delayed real-time performance we wanted to use modern GPUs.
+Their theoretical peak performance is typically an order of magnitude higher than a CPU in a comparable price range.
+While it is hard to obtain this performance in practice, we got very close by using a well matured and highly optimized graphics processing API such as OpenGL.
+Because of this we use OpenGL for rendering a 2D projection of a scene with 3D models of the seabed and objects.
+This is finally converted using the OpenCL language, which can interoperate with OpenGLTo convert this into a sonar template image we use Some additional prost-processing is required as well to form the actual template, but
 

### 2014-04-30 UAM2014/Article/article.tex
la2pdf commit - 2014:04:30	23:33:59	(Wed30.Apr)

@@
 
-Various simulators for high frequency, side-looking sonar imagery have been published, e.g.~\cite{Bell1997,Sammelmann2004}.
+Various simulators for high frequency, side-looking sonar imagery have been published, e.g.~\cite{Bell1997,Sammelm2003}.
 Most implementations have however prioritized accurate acoustic modeling at the expense of execution speed.
@@
 \newlength\imgspacing\setlength\imgspacing{.5cm}
-
-Whenever conventional computer graphics processing can be used to 
-GPUs are designed to provide the best possible graphics processing performance, and there large ecosystem of tools and libraries  available 
-Using a GPU to drive a simulator is advantageous since
-
-Using a GPU-based simulator is attractive since computer is an attractive solution as there are a  graphics traits.
-For one over a CPU-based standard simulator is two-fold: First is its ability to create templates incredibly fast.
-This makes it possible to tailor the templates more closely to an image as the parameters needed by the simulator can be directly estimated from the image.
-In a standard approach, on the other hand, a template library must be created beforehand from a limited set of parameter values.
-Hence, it may happen that no template in the library is a close fit of the object in the image even though the object actually is a target of interest.
-
-This performance was demonstrated on an image from the HISAS 1030 synthetic aperture sonar mounted on the HUGIN AUV.
-Out of the objects in the scene we selected a 2.6\;m long cylinder that was slightly buried in the sea sediments (Fig.
-\ref{data_cylinder}).
-The length, burial depth, and aspect angle of this cylinder were estimated from the SAS image using our  adaptive template matching approach~\cite{Midelfart2010}.
-Then these parameters were used by the simulator to create an adaptive template (Fig.
-\ref{sim_cylinder}), which is an almost a perfect match of the cylinder.
-We also created standard templates for a regular template matching approach.
-In this case, we assumed that the cylinder mine had a length of 2\;m (like a MP80 or a Murena mine) and was proud on the seafloor.
-Moreover, templates were created for very 10$^\text{th}$ degree of the aspect angle.
-We believe the assumptions are quite typical for a template library for cylinder mines.
-
-The standard and adaptive templates were then matched to the HISAS image.
-This is illustrated in Fig.
-\ref{overlay_cylinder} for a standard template and in Fig.
-\ref{overlay_cylinder_adaptive} for an adaptive template.
-These images were created by segmenting the image and the templates into highlight, shadow, and background regions.
-The regions from a template were then laid on top of the regions of the image to illustrate how well these regions matched.
-From these figures, it was quite clear the adaptive template obtained a much closer fit to image than the standard template.
-This was also reflected in the correlation scores (which were created with the method described in \cite{Midelfart2010}) that were 0.613 for the standard template and 0.813 for the adaptive template.
-Hence, the standard approach was less likely to classify the cylinder in the image correctly as the standard templates were not created specifically for the target.
-
 \setcounter{topnumber}{2}
@@
 
+% Whenever conventional computer graphics processing can be used to 
+% GPUs are designed to provide the best possible graphics processing performance, and there large ecosystem of tools and libraries  available 
+% Using a GPU to drive a simulator is advantageous since
+% 
+% 
+% Creating a GPU-based simulator is attractive since computer is an attractive solution as there are a  graphics traits. For one over a CPU-based standard simulator is two-fold:
+
+Our GPU-based simulator is very fast.
+On a computer with a six-core Intel Core i7-3930K and a Radeon HD7970 it computes almost 1000 templates per second with a size of 1000x1000 pixels (1 megapixel).
+This makes it possible to tailor the templates more closely to an image as the parameters needed by the simulator can be directly estimated from the image.
+In a standard approach, on the other hand, a template library must be created beforehand from a limited set of parameter values.
+Hence, it may happen that no template in the library is a close fit of the object in the image even though the object actually is a target of interest.
+
+This performance was demonstrated on an image from the HISAS 1030 synthetic aperture sonar mounted on the HUGIN AUV.
+Out of the objects in the scene we selected a 2.6\;m long cylinder that was slightly buried in the sea sediments (Fig.
+\ref{data_cylinder}).
+The length, burial depth, and aspect angle of this cylinder were estimated from the SAS image using our  adaptive template matching approach~\cite{Midelfart2010}.
+Then these parameters were used by the simulator to create an adaptive template (Fig.
+\ref{sim_cylinder}), which is an almost a perfect match of the cylinder.
+We also created standard templates for a regular template matching approach.
+In this case, we assumed that the cylinder mine had a length of 2\;m (like a MP80 or a Murena mine) and was proud on the seafloor.
+Moreover, templates were created for very 10$^\text{th}$ degree of the aspect angle.
+We believe the assumptions are quite typical for a template library for cylinder mines.
+
+The standard and adaptive templates were then matched to the HISAS image.
+This is illustrated in Fig.
+\ref{overlay_cylinder} for a standard template and in Fig.
+\ref{overlay_cylinder_adaptive} for an adaptive template.
+These images were created by segmenting the image and the templates into highlight, shadow, and background regions.
+The regions from a template were then laid on top of the regions of the image to illustrate how well these regions matched.
+From these figures, it was quite clear the adaptive template obtained a much closer fit to image than the standard template.
+This was also reflected in the correlation scores (which were created with the method described in \cite{Midelfart2010}) that were 0.613 for the standard template and 0.813 for the adaptive template.
+Hence, the standard approach was less likely to classify the cylinder in the image correctly as the standard templates were not created specifically for the target.
+
 \section{Conclusion}
@@
 Instead we have implemented a SAS simulator that creates the templates in delayed real-time based on parameters estimated from the current scene.
-In our studies this improves the match between the SAS objects and the template considerably.
+In our studies this improves the match between the SAS objects and the template significantly in most cases.
 
-To obtain the delayed real-time performance we wanted to use modern GPUs.
+To obtain the delayed real-time performance we implemented the simulator on a GPU.
 Their theoretical peak performance is typically an order of magnitude higher than a CPU in a comparable price range.
-While it is hard to obtain this performance in practice, we got very close by using a well matured and highly optimized graphics processing API such as OpenGL.
-Because of this we use OpenGL for rendering a 2D projection of a scene with 3D models of the seabed and objects.
-This is finally converted using the OpenCL language, which can interoperate with OpenGLTo convert this into a sonar template image we use Some additional prost-processing is required as well to form the actual template, but
+This potential is effectively utilized with a well matured and highly optimized graphics processing API such as OpenGL.
+Our simulator use OpenGL for most for most of the scene processing.
+A final post-processing step is performed in OpenCL
+
+\begin{itemize}
+\item Mer testing
+\item Robustifisering
+\item Tuning
+\end{itemize}
 

### 2014-04-30 UAM2014/Article/article.tex
la2pdf commit - 2014:05:01	00:04:53	(Thu 1.May)

@@
 For any object type, position, alignment and material, the procedure is repeated, and a correlation coefficient computed.
-Simulation speed is fundamental when templates are created as part of the classification process.
+Simulation speed is critical when templates are created as part of the classification process.
 The simulator is therefore written in OpenGL and OpenCL and runs on graphics processing units (GPUs).
@@
 %
-\keywords{Sonar, SAS, simulator, template, OpenGL}
+\keywords{Sonar, SAS, simulator, template matching, OpenGL, OpenCL}
 
@@
 The principle is to isolate an image segment containing the object of interest, correlate it with a set of template images of the relevant object classes, and assign it to the class of the template yielding the highest correlation coefficient.
-The challenge is to come up with a representative set of template images covering the actual configurations of object and seabed\todo{objects?
-seabeds?}.
-The use of a static, predetermined template library may be impractical, yield inaccurate results, or both~\cite{Midelfart2010}.
+The challenge is to come up with a representative set of template images covering the actual configurations of object and seabed.
+The use of a static, predetermined template library is often impractical and yield inaccurate results~\cite{Midelfart2010}.
 
@@
 This is why we decided to use the core OpenGL pipeline to do this for us.
-OpenGL is a popular, well-matured, and multi-platform application programming interface (API) for rendering 2D and 3D vector graphics.
+OpenGL is a popular, well-matured and multi-platform application programming interface (API) for rendering 2D and 3D vector graphics.
 It relieves us from the intricacies of projecting vertices, faces and textures defined in a 3D space onto a suitable 2D image plane.
-This section explains how we set up OpenGL to do this for us, and then proceeds to describe how we post process the OpenGL images with OpenCL to form the final sonar image.
+This section explains how we set up OpenGL for this task, and then proceeds to describe how we post process the OpenGL images with OpenCL to form the final sonar image.
 
@@
 
-Before we can render anything OpenGL we first have to set up the scene. This involves loading the models and then placing them where we want them with the proper orientation.
+Before we can render anything with OpenGL we first have to set up the scene. This involves loading the models and then placing them where we want them with the proper orientation.
 
@@
 This is achieved by applying a set of transformations to each vertex in the model.
-In OpenGL, the set of transformations are usually formalized like this:
+In OpenGL, the set of transformations is usually formalized like this:
 \begin{align*}
@@
 This permits us to use a Lambertian scattering model where the backscatter intensity depends only on the incidence angle~\cite{Zhang1999}.
-It does not consider observation angle or sound frequency, but for the purpose of creating templates this is needed.
+It does not consider observation angle or sound frequency, but for the purpose of creating templates this is not needed.
 
@@
 
-The image rendered with OpenGL is not in $(x,y)$ coordinates as we want it to be.
-It does, however, tell us what parts of the scene that is visible from the sonar.
+The image rendered with OpenGL is not in along-track and cross-track coordinates as we want it to be.
+It does, however, tell us what parts of the scene that are visible from the sonar.
 In addition, OpenGL can produce a depth map that reveals the distance to each template pixel.
@@
 \graphicsAI[drawing,width=\linewidth]{gfx/sim_cylinder_submerged_adaptive.svg}%
-\caption{Cylinder simulation with parameters adapted to fit the SAS image: Length 2.6\;m, burial depth 0.263\;m, aspect angle 105$^\circ$.}\label{sim_cylinder}%
+\caption{Cylinder simulation with parameters estimated from SAS image: Length 2.6\;m, burial depth 0.263\;m, aspect angle 105$^\circ$.}\label{sim_cylinder}%
 \end{subfigure}
@@
 \graphicsAI[drawing,width=\linewidth]{gfx/overlay_cylinder_submerged.svg}%
-\caption{Standard template. Predefine template's cylinder length, alignment and burial depth do no match. Shadow and highlight regions to not correspond well.}\label{overlay_cylinder}%
+\caption{Best match of a set of predefined templates. Inaccurate cylinder length, alignment and immersion depth. The segmented shadow and highlight regions are misaligned.}\label{overlay_cylinder}%
 \end{subfigure}\hspace{\imgspacing}%
@@
 \graphicsAI[width=\linewidth]{gfx/overlay_cylinder_submerged_adaptive.svg}%
-\caption{Adapted template. Template simulated with adapted parameters. Good fit in both highlight and shadow regions.}\label{overlay_cylinder_adaptive}%
+\caption{Adapted template. Template simulated with adapted parameters. Good fit in both highlight and shadow segments.}\label{overlay_cylinder_adaptive}%
 \end{subfigure}
@@
 This performance was demonstrated on an image from the HISAS 1030 synthetic aperture sonar mounted on the HUGIN AUV.
-Out of the objects in the scene we selected a 2.6\;m long cylinder that was slightly buried in the sea sediments (Fig.
+Out of the objects in the scene we selected a 2.6\;m long cylinder that was partly buried in the sea sediments (Fig.
 \ref{data_cylinder}).
-The length, burial depth, and aspect angle of this cylinder were estimated from the SAS image using our  adaptive template matching approach~\cite{Midelfart2010}.
+The length, immersion depth and aspect angle of this cylinder were estimated from the SAS image using our  adaptive template matching approach~\cite{Midelfart2010}.
 Then these parameters were used by the simulator to create an adaptive template (Fig.
-\ref{sim_cylinder}), which is an almost a perfect match of the cylinder.
+\ref{sim_cylinder}), which is an almost perfect match of the cylinder.
 We also created standard templates for a regular template matching approach.
-In this case, we assumed that the cylinder mine had a length of 2\;m (like a MP80 or a Murena mine) and was proud on the seafloor.
-Moreover, templates were created for very 10$^\text{th}$ degree of the aspect angle.
-We believe the assumptions are quite typical for a template library for cylinder mines.
+In this case, we assumed that the cylinder mine had a generic length of 2\;m (like a MP80 or a Murena mine) and was proud on the seafloor.
+Moreover, templates were created for every 10$^\text{th}$ degree of the aspect angle.
+We believe these assumptions to be typical for a template library for cylinder mines.
 
@@
 
-One way to automatically classify objects in SAS images is compare the imaged objects with a predefined set of templates.
+One way to automatically classify objects in SAS images is to compare the imaged objects with a predefined set of templates.
 However, this is suboptimal as it is infeasible to create templates for all configurations of seabeds and objects.
@@
 This potential is effectively utilized with a well matured and highly optimized graphics processing API such as OpenGL.
-Our simulator use OpenGL for most for most of the scene processing.
-A final post-processing step is performed in OpenCL
+Our simulator use OpenGL for most of the scene processing.
+A final post-processing step is performed in OpenCL.
 

### 2014-05-01 ThesisJP/Background/background.tex
Finished english abstract.

@@
 Note that the derived numbers are strictly theoretical and that the actual throughput is highly algorithm dependent.
-Nevertheless, in theory, the computational power provided by modern GPUs is five times higher than that provided by CPUs.
+Nevertheless, in theory, the computational throughput provided by modern GPUs is five times higher than that provided by CPUs.
 A claimed speed-up of several orders of magnitude is therefore typically a result of comparing a non-optimized CPU implementation with an optimized GPU implementation or by comparing incomparable hardware \cite{Lee2010, Kothapalli2013}.
@@
 where \nom{$c$}{Speed of sound} is the speed of sound in human tissue, and $p_m$ is the position of the m'th element.
-When transmitting the pipeline depicted in Fig.\,\ref{fig:das_background}, without the sum operator, is reversed and the negated set of delays has to be used.
+When transmitting, the pipeline depicted in Fig.\,\ref{fig:das_background}, without the sum operator, is reversed and the negated set of delays has to be used.
 Also note that on transmit we have to decide where to steer and focus prior to transmitting a pulse.
@@
 Medical ultrasound systems typically has a bandwidth between 33 \% and 100 \% of the pulse frequency.
-This corresponds to transmitting a pulse which is between 1 and 3 oscillations when entering tissue.
+This corresponds to transmitting a pulse which is between 1 and 3 oscillations long.
 In the analog-to-digital converter of an ultrasound imaging system a high sampling frequency is typically used (e.g.
@@
 For ultrasound visualization, the color intensity is dictated by the ultrasound backscatter intensity, and the colorfulness is often encoded with depth information in order to improve depth perception.
-An example of a volume rendering of cardiac ultrasound data is presented in \textbf{Paper IV}.
+An example of a volume rendering of cardiac ultrasound data is presented in Fig.\,\ref{fig:dvr}.
 The opacity, however, is governed by a user-controlled \textit{opacity transfer function} (\nom{OTF}{Opacity transfer function (volume rendering)}) as shown in Fig.\,\ref{fig:vr_otf}.

### 2014-05-01 ThesisJP/Introduction/introduction.tex
Finished english abstract.

@@
 
-This immense rise in computational power and improved programmability are currently changing how ultrasound imaging systems are designed.
+This immense rise in computational capabilities and improved programmability are currently changing how ultrasound imaging systems are designed.
 Algorithms that previously had to be implemented in hardware, for performance reasons, can now be implemented in software.
@@
 The proposed method is implemented on a GPU with real-time frame rates as the result.
-This results shows that with modern GPUs it is possible to add more advanced visualization to an ultrasound imaging system and still have real-time performance.
+This results shows that with modern GPUs it is possible to add more advanced visualization methods to an ultrasound imaging system and still have real-time performance.
 
@@
 Achieving higher processing speed for an excising algorithm by implementing it on a new architecture is not academically relevant in its own.
-A good analysis has to be given of how the algorithm was modified in order for the increase to happen, or how to utilize the gained speed.
+A good analysis has to be given of how the algorithm was modified for the architecture, or how to utilize the gained speed.
 These are points we have tried to answer in both \textbf{Paper\,I}, \textbf{II},  \textbf{IV}, and \textbf{V}.
 
-In \textbf{Paper\,V} the increased speed made an interactive painting simulation program feasible.
+In \textbf{Paper\,V} the increased speed made an interactive painting ultrasound simulation program feasible.
 As of January 2014 the simulator has been download 200 times and counting\footnote{folk.uio.no/jpaasen/huygens}.

### 2014-05-01 ThesisJP/Thesis.tex
Finished english abstract.

@@
 
+%\setcounter{page}{1}
 \renewcommand{\thepage}{\roman{page}}% Roman numerals for page counter
@@
 \\
-Doctoral theses at NTNU, 20014:XX\\
+Doctoral theses at NTNU, 2014:XX\\
 \\
@@
 \noindent\large 
-Sammendrag p\aa{} norsk.
+\textbf{Akselerering av adaptive algoritmer for ultralydavbildning ved hjelp a programmerbare grafikkprosessorer}
 %\vspace{0.2cm}\\%
@@
 \normalsize
-\todo{Skriv sammendrag p\aa{} norsk}
+En rivende utvikling innen spillteknologi og tilh\o{}rende programmeringsspr\aa{}k ...
 
@@
 \addcontentsline{toc}{chapter}{Abstract}
-%\chapter*{Abstract}
-%\thispagestyle{empty}%
-\todo{Write abstract ...}
+A rapid development in computer game technology and accompanying programming languages have recently provided researchers with small personal super computers, comprised in a single graphics card (GPU).
+This immense rise in computational capabilities and improved programmability are currently changing how ultrasound imaging systems are designed.
+When researchers are exploring new algorithms for ultrasound imaging, it is therefore important to keep the architecture of parallel accelerators in mind.
+If a new complex algorithm is supposed to run in real time, it needs to fit the programmable and parallel pipeline of modern ultrasound scanners.
+
+The aim of this study has been to investigate the possibility of utilizing GPUs for advanced processing in an ultrasound imaging system.
+Among the investigated problems are both adaptive beamforming, adaptive visualization of ultrasound volumes, and ultrasound simulations.
+The presented problems have in common that they require parallel programming in order to reach real-time processing.
+
+In the first part of the thesis, the Capon adaptive beamformer is investigated and implemented on a GPU for the application of real-time sonar (\textbf{Paper\,I}) and medical ultrasound imaging (\textbf{Paper\,II}).
+Real-time frame rates are achieved for both modalities.
+\textbf{Paper\,II} also present, for the first time, videos where the Capon beamformer has been applied on loops of simulated and \textit{in vivo} medical ultrasound images.
+An investigation of shift-invariance for real-time Capon beamforming is performed in \textbf{Paper\,III}.
+Shift-invariant imaging is essential if the method is ever to be used in practice.
+
+In paper \textbf{Paper IV} we propose an adaptive method for visualization of volumetric cardiac ultrasound images.
+The method is capable of removing noise that by traditionally methods would have occlude cardiac tissue.
+This work also shows that with modern GPUs it is possible to add advanced visualization methods to an ultrasound imaging system and still have real-time performance.
+
+Finally, we investigate how GPUs can be utilized to accelerate ultrasound simulations (\textbf{Paper V}).
+The result of this work is a simulation program where array geometries can be interactively drawn and where the resulting pressure field is simulated and visualized in real time.
 

### 2014-05-01 ThesisJP/Thesis.tex
Finished norwegian abstract.

@@
 \normalsize
-En rivende utvikling innen spillteknologi og tilh\o{}rende programmeringsspr\aa{}k ...
+En rivende utvikling innen spillteknologi og tilh\o{}rende programmeringsspr\aa{}k har nylig gjort super-datamaskiner tilgjengelige for forskere i form av et lite grafikkort (GPU).
+Denne utrolige \o{}kningen i beregningskraft og forbedret programmerbarhet er for \o{}yeblikket med p\aa{} \aa{} endre hvordan ultralydsystemer designes.
+N\aa{}r forskere utforsker nye algoritmer for ultralydavbildning er det derfor viktig at arkitekturen til parallelle akseleratorer tas h\o{}yde for.
+Hvis en ny, kompleks algoritme er tenkt til \aa{} kj\o{}res i sanntid, er det viktig at den kan tilpasses den programmerbare og parallelle prosesseringskjeden i moderne ultralydskannere.
+
+M\aa{}let med denne studien har v\ae{}rt \aa{} utforske hvordan GPUer kan utnyttes til avansert prosessering i et avbildende ultralydsystem.
+Blant de problemene vi har sett n\ae{}rmere p\aa{} finner man b\aa{}de adaptiv str\aa{}leforming, adaptiv visualisering av volumetrisk ultralyd og ultralydsimuleringer.
+Felles for disse problemene er at de alle krever parallelprogrammering for at sanntidsprosessering skal oppn\aa{}s.
+
+I f\o{}rste del av avhandlingen utforskes Capon's adaptive str\aa{}leformings-metode.
+Metoden implementeres p\aa{} en GPU  for \aa{} muliggj\o{}re sanntidsprosessering for avbildende sonar (\textbf{Artikkel\,I}) og medisinsk ultralydavbildning (\textbf{Artikkel\,II}).
+Sanntidsprosessering oppn\aa{}s for begge modalitetene.
+\textbf{Artikkel\,II} presenterer ogs\aa{}, for f\o{}rste gang, videoer hvor b\aa{}de simulerte og \textit{in vivo} medisinsk ultralyddata er prosessert med Capon-str\aa{}leforming.
+I \textbf{Artikkel\,III} viser vi at Capon-str\aa{}leformeren ikke har skift-invariant oppf\o{}rsel n\aa{}r den anvendes til sanntidsavbildning.
+Videre foresl\aa{}s det en metode som gir mer shift-invariant avbildning.
+Skift-invariant avbildning er avgj\o{}rende for at metoden skal kunne tas ibruk i praktisk.
+
+I \textbf{Artikkel\,IV} foresl\aa{}r vi en adaptiv metode for visualisering av volumetriske ultralydbilder av hjertet.
+Metoden undertrykker st\o{}y som med konvensjonelle metoder hadde skygget for hjertevev.
+Dette arbeidet viser ogs\aa{} at det med moderne GPUer er mulig \aa{} utf\o{}re avansert visualisering i et ultralydsystem og fortsatt ha sanntidsytelse.
+
+I siste del av avhandlingen utforsker vi hvordan GPUer kan utnyttes til \aa{} akselerere ultralydsimuleringer.
+Resultatet av dette arbeidet ble et simuleringsprogram hvor geometrien til en gitt ultralydprobe kan interaktivt tegnes, og hvor det resulterende trykkfeltet simuleres i sanntid.
 
@@
 \vspace{0.2cm}
-\noindent \emph{Ovennevnte avhandling er funnet verdig til \aa{} forsvares offentlig for graden Philosophiae Doctor (PhD) i medisinsk teknologi. Disputas finner sted ... .}
+\noindent \emph{Ovennevnte avhandling er funnet verdig til \aa{} forsvares offentlig for graden Philosophiae Doctor (Ph.D.) i medisinsk teknologi. Disputas finner sted ... .}
 
@@
 \textbf{Paper\,II} also present, for the first time, videos where the Capon beamformer has been applied on loops of simulated and \textit{in vivo} medical ultrasound images.
-An investigation of shift-invariance for real-time Capon beamforming is performed in \textbf{Paper\,III}.
+In \textbf{Paper\,III} we show that Capon beamforming does not provide shift-invariant imaging in a real-time imaging setting.
+A method is then proposed that improves the shift-invariant property.
 Shift-invariant imaging is essential if the method is ever to be used in practice.
 
-In paper \textbf{Paper IV} we propose an adaptive method for visualization of volumetric cardiac ultrasound images.
-The method is capable of removing noise that by traditionally methods would have occlude cardiac tissue.
+In paper \textbf{Paper\,IV} we propose an adaptive method for visualization of volumetric cardiac ultrasound images.
+The method is capable of removing noise that by conventional methods would have occlude cardiac tissue.
 This work also shows that with modern GPUs it is possible to add advanced visualization methods to an ultrasound imaging system and still have real-time performance.
 
-Finally, we investigate how GPUs can be utilized to accelerate ultrasound simulations (\textbf{Paper V}).
-The result of this work is a simulation program where array geometries can be interactively drawn and where the resulting pressure field is simulated and visualized in real time.
+Finally, we investigate how GPUs can be utilized to accelerate ultrasound simulations (\textbf{Paper\,V}).
+The result of this work was a simulation program where ultrasound probe geometries can be interactively drawn and where the resulting pressure field is simulated and visualized in real time.
 

### 2014-05-01 ThesisJP/Thesis.tex
Some small final corrections.

@@
 \normalsize
-En rivende utvikling innen spillteknologi og tilh\o{}rende programmeringsspr\aa{}k har nylig gjort super-datamaskiner tilgjengelige for forskere i form av et lite grafikkort (GPU).
+En rivende utvikling innen spillteknologi og tilh\o{}rende programmeringsspr\aa{}k har nylig gjort super-datamaskiner i form av et lite grafikkort (GPU) tilgjengelige for forskere.
 Denne utrolige \o{}kningen i beregningskraft og forbedret programmerbarhet er for \o{}yeblikket med p\aa{} \aa{} endre hvordan ultralydsystemer designes.
@@
 Metoden implementeres p\aa{} en GPU  for \aa{} muliggj\o{}re sanntidsprosessering for avbildende sonar (\textbf{Artikkel\,I}) og medisinsk ultralydavbildning (\textbf{Artikkel\,II}).
-Sanntidsprosessering oppn\aa{}s for begge modalitetene.
+Sanntidsprosessering oppn\aa{}s for begge modaliteter.
 \textbf{Artikkel\,II} presenterer ogs\aa{}, for f\o{}rste gang, videoer hvor b\aa{}de simulerte og \textit{in vivo} medisinsk ultralyddata er prosessert med Capon-str\aa{}leforming.
@@
 \textbf{Paper\,II} also present, for the first time, videos where the Capon beamformer has been applied on loops of simulated and \textit{in vivo} medical ultrasound images.
-In \textbf{Paper\,III} we show that Capon beamforming does not provide shift-invariant imaging in a real-time imaging setting.
+In \textbf{Paper\,III}, we show that Capon beamforming does not provide shift-invariant imaging in a real-time imaging setting.
 A method is then proposed that improves the shift-invariant property.

### 2014-05-02 UAM2014/Article/article.tex
la2pdf commit - 2014:05:02	15:49:08	(Fri 2.May)

@@
 Then it places a 3D object model on the seabed, renders the scene, and adds the resulting image to the template set.
-For any object type, position, alignment and material, the procedure is repeated, and a correlation coefficient computed.
-Simulation speed is critical when templates are created as part of the classification process.
+For every object type, position, alignment and material, the procedure is repeated, and a correlation coefficient computed.
+The best performance is obtained when these parameters are estimated from the sonar image as part of the classification process.
 The simulator is therefore written in OpenGL and OpenCL and runs on graphics processing units (GPUs).
@@
 \graphicsAI[drawing,width=\linewidth]{gfx/overlay_cylinder_submerged.svg}%
-\caption{Best match of a set of predefined templates. Inaccurate cylinder length, alignment and immersion depth. The segmented shadow and highlight regions are misaligned.}\label{overlay_cylinder}%
+\caption{Best template out of a predefined set. Inaccurate cylinder length, alignment and immersion depth. The segmented shadow and highlight regions are misaligned.}\label{overlay_cylinder}%
 \end{subfigure}\hspace{\imgspacing}%
@@
 Instead we have implemented a SAS simulator that creates the templates in delayed real-time based on parameters estimated from the current scene.
-In our studies this improves the match between the SAS objects and the template significantly in most cases.
+In our studies this improves the match between the SAS objects and their corresponding template significantly in most cases.
 
@@
 Our simulator use OpenGL for most of the scene processing.
-A final post-processing step is performed in OpenCL.
 
-\begin{itemize}
-\item Mer testing
-\item Robustifisering
-\item Tuning
-\end{itemize}
+A final post-processing step is performed in OpenCL, which allow general purpose GPU programming.
+It interoperates well with OpenGL and adds a lot of flexibility to the simulation process.
+This is valuable as the simulator is still being actively developed and will likely need new features that can not easily be obtained using OpenGL alone.
+
+% 
+% 
+% \begin{itemize}
+% \item Mer testing
+% \item Robustifisering
+% \item Tuning
+% \end{itemize}
+%
 

### 2014-05-02 UAM2014/Article/article.tex
la2pdf commit - 2014:05:02	22:54:47	(Fri 2.May)

@@
 
-Automatic target recognition (ATR) is a fundamental component to optimize the use of autonomous underwater vehicles (AUVs), as it allows the vehicle to adapt its mission plan to e.g.
+Automatic target recognition (ATR) is an important component in autonomous underwater vehicles (AUVs), as it allows the vehicle to adapt its mission plan to e.g.
 revisit detected objects for closer examination.
-The ATR algorithms must, however, be sufficiently fast to process the sensor data onboard the AUV in delayed real-time.
-
-Template matching is a common technique for classification of objects in synthetic aperture sonar (SAS) images.
+One way to classify these objects is by using template matching.
 The principle is to isolate an image segment containing the object of interest, correlate it with a set of template images of the relevant object classes, and assign it to the class of the template yielding the highest correlation coefficient.
@@
 Also, their simulator sums images obtained from multiple OpenGL rendering passes to create the sonar image, while we render once and post-process this result with OpenCL to tailor the sonar image with more flexibility.
+
+% The ATR algorithms must, however, be sufficiently fast to process the sensor data onboard the AUV in delayed real-time.
 

### 2014-05-02 UAM2014/Article/article.tex
la2pdf commit - 2014:05:02	23:51:52	(Fri 2.May)

@@
 One way to classify these objects is by using template matching.
-The principle is to isolate an image segment containing the object of interest, correlate it with a set of template images of the relevant object classes, and assign it to the class of the template yielding the highest correlation coefficient.
+The principle involved  is to isolate an image segment containing the object of interest, correlate it with a set of template images of the relevant object classes, and assign it to the class of the template yielding the highest correlation coefficient.
 The challenge is to come up with a representative set of template images covering the actual configurations of object and seabed.
-The use of a static, predetermined template library is often impractical and yield inaccurate results~\cite{Midelfart2010}.
+The alternative of using of a static, predetermined template library has a computational complexity that is exponential with the number of parameters.
+This limits its use to only a few parameters being coarsely sampled, ultimately yielding inaccurate results~\cite{Midelfart2010}.
 

### 2014-05-02 UAM2014/Article/article.tex
la2pdf commit - 2014:05:03	00:23:05	(Sat 3.May)

@@
 
-We have developed a sonar simulator that runs sufficiently fast to create templates adapted to the actual scene as part of the classification process.
+We have developed a synthetic aperture sonar (SAS) simulator that avoids this problem by running sufficiently fast to create templates adapted to the actual scene as part of the classification process.
 This is achieved with the aid of the massive computing power in graphics processing units (GPUs) and optimized software libraries for scene rendering.
@@
 \end{align*}
-where is the loaded model, $\bmat{\cdot}_\text{image}$ is the output image, and $\M$, $\V$ and $\P$ are transformation matrices.
+where $\bmat{\cdot}_\text{model}$ is the loaded model, $\bmat{\cdot}_\text{image}$ is the output image, and $\M$, $\V$ and $\P$ are transformation matrices.
 The first transform we apply is the model matrix $\M$, which scales and rotates the model in its local coordinate system.
@@
 
-The image rendered with OpenGL is not in along-track and cross-track coordinates as we want it to be.
+The ``camera image'' rendered with OpenGL is not in along-track and cross-track coordinates as we want it to be.
 It does, however, tell us what parts of the scene that are visible from the sonar.
-In addition, OpenGL can produce a depth map that reveals the distance to each template pixel.
+OpenGL can also produce a depth map that reveals the distance to each of image pixels.
 This information can be converted to a ranged sonar-like image by simply adding up all the intensity values that share the same depth for each range line.
@@
 Our GPU-based simulator is very fast.
-On a computer with a six-core Intel Core i7-3930K and a Radeon HD7970 it computes almost 1000 templates per second with a size of 1000x1000 pixels (1 megapixel).
-This makes it possible to tailor the templates more closely to an image as the parameters needed by the simulator can be directly estimated from the image.
-In a standard approach, on the other hand, a template library must be created beforehand from a limited set of parameter values.
+On a computer with a six-core Intel Core i7-3930K and a Radeon HD7970 it computes almost 1000 templates per second, each being 1 megapixel large.
+This makes it possible to tailor the templates very closely imaged objects.
+In contrast, in the standard approach a template library must be created beforehand from a limited set of parameter values.
 Hence, it may happen that no template in the library is a close fit of the object in the image even though the object actually is a target of interest.
@@
 The regions from a template were then laid on top of the regions of the image to illustrate how well these regions matched.
-From these figures, it was quite clear the adaptive template obtained a much closer fit to image than the standard template.
+Note how the adaptive template obtained a much closer fit to image than the standard template.
 This was also reflected in the correlation scores (which were created with the method described in \cite{Midelfart2010}) that were 0.613 for the standard template and 0.813 for the adaptive template.
@@
 One way to automatically classify objects in SAS images is to compare the imaged objects with a predefined set of templates.
-However, this is suboptimal as it is infeasible to create templates for all configurations of seabeds and objects.
-Instead we have implemented a SAS simulator that creates the templates in delayed real-time based on parameters estimated from the current scene.
+However, this is suboptimal as it is infeasible to create accurate templates for all the relevant configurations of seabeds and objects.
+To solve this we have implemented a SAS simulator that creates the templates in delayed real-time based on parameters estimated from the current scene.
 In our studies this improves the match between the SAS objects and their corresponding template significantly in most cases.
@@
 To obtain the delayed real-time performance we implemented the simulator on a GPU.
-Their theoretical peak performance is typically an order of magnitude higher than a CPU in a comparable price range.
-This potential is effectively utilized with a well matured and highly optimized graphics processing API such as OpenGL.
-Our simulator use OpenGL for most of the scene processing.
+These devices have a theoretical peak performance that is typically an order of magnitude higher than CPUs in a comparable price range.
+We experienced that this potential was effectively utilized with a well matured and highly optimized graphics processing API such as OpenGL.
+Our simulator use this framework for most of the scene processing.
 

### 2014-05-02 UAM2014/Article/article.tex
la2pdf commit - 2014:05:03	01:11:18	(Sat 3.May)

@@
 These devices have a theoretical peak performance that is typically an order of magnitude higher than CPUs in a comparable price range.
-We experienced that this potential was effectively utilized with a well matured and highly optimized graphics processing API such as OpenGL.
+We found this potential to be effectively utilized with the well matured and highly optimized OpenGL graphics processing API.
 Our simulator use this framework for most of the scene processing.
@@
 It interoperates well with OpenGL and adds a lot of flexibility to the simulation process.
-This is valuable as the simulator is still being actively developed and will likely need new features that can not easily be obtained using OpenGL alone.
+This is valuable as the simulator is still being actively developed and will likely see new features that can not be easily implemented in OpenGL alone.
 

### 2014-05-03 Simulator/article.tex
Started on journal version of simulator

--- 2014-05-02 UAM2014/Article/article.tex
@@
-%===============%                               %~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~%
-% DOCUMENTCLASS %                                See full option description in "mytemplate.cls"
-%===============%                               %~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~%
+%%%%%%%%%%%%%%%%                                ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
+% CONDITIONALS %
+%%%%%%%%%%%%%%%%                                ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
 
-\documentclass[
-   UAM                                          % Document type (non-standard)
- , 12pt                                         % Text size
-%  , draft
-%  , final                                        % Quality
-%  , xelatex                                      % Use the XeLaTeX compiler
-%  , biblatex                                     % Use 'biblatex' for references (best by far)
- , bibtex                                       % Use 'bibtex' for references (oldschool :)
-%  , movie
-%  , notodos                                      % Disable todos
- , layout
-%  , defaultformat                                % Attempt to use default formatting options
-%  , glossary                                     % Use a glossary
-]{common/mytemplate}
-% \bibliography{references}
-\hypersetup{
-   bookmarksopen=true
- , bookmarksopenlevel=2
-}
+\newif\ifPeerReview\PeerReviewfalse              % Whether to create the PeerReview version or
+                                                % Journal version
+\newif\ifOnlineColor\OnlineColortrue            % Compile online color version?
 
+\newif\ifFlatArchive\FlatArchivefalse           % Whether archive is flat (messy) or contain 
+                                                % subfolders for graphics etc.
+\newif\ifFloatAtEnd\FloatAtEndfalse             % Available in PeerReview mode:
+                                                % Place floats at end of document?
+\newif\ifTODO\TODOtrue                        % Use todo notes?
+
+\ifOnlineColor
+   \newcommand\figPostfix{_online}              % Colored images online
+\else
+   \newcommand\figPostfix{_bw}                  % Black and white in journal
+\fi
+
+%%%%%%%%%%%%                                    ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
+% IEEEtran %
+%%%%%%%%%%%%                                    ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
+
+\ifPeerReview
+\documentclass[12pt,journal,draftclsnofoot,onecolumn]{IEEEtran}
+% \newcommand\CLASSINPUTbaselinestretch{1.66}     % http://theoval.cmp.uea.ac.uk/~nlct/latex/thesis/node17.html
+\else
+\documentclass[journal]{IEEEtran}
+\fi
+
+% \RequirePackage[latin1]{inputenc}%              % Set input encoding (optionally latin1)
+% \RequirePackage[T1]{fontenc}%                 % Set font encoding
+% \usepackage[norsk]{babel}
+
+% Single column review mode:
+%\documentclass[12pt,journal,onecolumn]{IEEEtran}
+%\newcommand\CLASSINPUTbaselinestretch{2}
+%\RequirePackage{calc}
+%\RequirePackage{fp} 
+% Left margin = 1 inch + hoffset + oddsidemargin (or evensidemargin)
+% Adding the 2cm gutter width to the odd/even side margins:
+%\setlength\hoffset{0pt}
+%\setlength\oddsidemargin{4cm}       % 4cm margins on the left side
+%\addtolength\oddsidemargin{-1in}    % Subtract the initial 1 inch
+%\setlength\textwidth{21cm-8cm}    % A4 width (21cm) minus margins on either side
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%                  ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
+% IEEE ''APPROVED'' PACKAGES %
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%                  ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
+
+\usepackage{cite}
+
+\ifCLASSINFOpdf
+   \usepackage[dvips]{graphicx}                 % Might not work. Use 'latex' instead of 
+   \ifFlatArchive\else                          % 'pdflatex'
+      \graphicspath{./gfx/}
+   \fi
+\else
+   \usepackage[dvips]{graphicx}
+   \ifFlatArchive\else
+      \graphicspath{./gfx/}
+   \fi
+\fi
+
+\RequirePackage[table,dvipsnames,svgnames]{xcolor}
+
+\usepackage[cmex10]{amsmath}                    % cmex10 option to be IEEE explore compliant
+\interdisplaylinepenalty=2500                   % Allows multiline equations to be broken
+
+% \RequirePackage{amssymb}
+
+\RequirePackage{array}
+% 
+% \ifCLASSOPTIONcompsoc
+%    \usepackage[caption=false,font=normalsize,labelfont=sf,textfont=sf]{subfig}
+% \else
+%    \usepackage[caption=false,font=footnotesize]{subfig}
+% \fi
+
+% \usepackage{caption}
+\usepackage{subcaption}
+\usepackage{color}
+\usepackage{calc}
+\usepackage{fp}
+
+% 
+% \ifCLASSOPTIONcaptionsoff                       % IEEE promoted hack to turn off captions from the 
+%    \let\MYorigsubfloat\subfloat                 % subfloat package should the captionsoff option
+%    \renewcommand{\subfloat}[2][\relax]{\MYorigsubfloat[]{#2}} % be specified.
+% \fi
+
+\ifFloatAtEnd
+\ifCLASSOPTIONcaptionsoff                       % Places float at the end of the document when the
+  \usepackage[nomarkers]{endfloat}              % captionsoff options is specified to IEEEtrans.cls
+  \let\MYoriglatexcaption\caption               % (PeerReview mode)
+  \renewcommand{\caption}[2][\relax]{\MYoriglatexcaption[#2]{#2}}
+\fi
+\fi
+
+\usepackage{fixltx2e}                           % Fix some twocolumn float problems
+
+%\usepackage{stfloats}                          % Allows: \begin{figure*}[!b]
+                                                % (double column figures on top/bottom)
+
+\usepackage{url}                                % Support for handling and breaking URLs
+
+% NOTE: PDF hyperlink and bookmark features are not required in IEEE
+%       papers and their use requires extra complexity and work.
+\newcommand\MYhyperrefoptions{bookmarks=true,bookmarksnumbered=true,
+pdfpagemode={UseOutlines},plainpages=false,pdfpagelabels=true,
+colorlinks=true,linkcolor={black},citecolor={black},urlcolor={black},
+pdftitle={An Optimized GPU Implementation of the MVDR Beamformer for Active Sonar Imaging},
+pdfsubject={},
+pdfauthor={Jo Inge Buskenes},
+pdfkeywords={adaptive beamforming, beamforming, complexity, sonar, active}}%
+\ifCLASSINFOpdf
+   \usepackage[\MYhyperrefoptions,pdftex]{hyperref}
+\else
+   \usepackage[\MYhyperrefoptions,breaklinks=true,dvips]{hyperref}
+   \usepackage{breakurl}                        % Allows 'dvips' driver to break links
+\fi
+
+%%%%%%%%%%%%%%%%%%%%%%%                         ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
+% ADDITIONAL PACKAGES %       
+%%%%%%%%%%%%%%%%%%%%%%%                         ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
+
+% \ifPeerReview
+% \let\OldIncludegraphics{\includegraphics}
+% \usepackage{letltxmacro}
+% \LetLtxMacro{\OldIncludegrsaphics}{\includegraphics}
+% \renewcommand{\includegraphics}[2][]{\OldIncludegraphics[width=\linewidth, #1]{#2}}
+% \fi
+
+% % \makeatletter
+% % \def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth
+% % \else\Gin@nat@width\fi}
+% % \makeatother
+% % \let\Oldincludegraphics\includegraphics
+% % \renewcommand{\includegraphics}[1]{\Oldincludegraphics[width=\maxwidth]{#1}}
+
+\usepackage[maxfloats=40]{morefloats}
+\newcounter{todoidx}
+% \setcounter{todoidx}
+
+\ifTODO
+   \definecolor{todobackground}{rgb}{0.95,0.95,0.95}
+   \setlength\marginparsep{1pt}
+   \setlength\marginparwidth{35pt}
+   \newlength\marginparwidthsmall
+   \setlength\marginparwidthsmall{\marginparwidth}
+   \addtolength\marginparwidthsmall{-7pt}
+   \newcommand\todo[1]{%
+      \addtocounter{todoidx}{1}%
+      {\color{Red}\bf(\thetodoidx{})}%%\fbox{\bf\thetodoidx{}}}%
+      \marginpar{%
+         {\vspace*{-10pt}\color{Red}\fbox{\bf\thetodoidx{}}}\\%
+         \fcolorbox{red}{todobackground}{\parbox{\marginparwidthsmall}{\scriptsize #1}}}}
+
+   \newcommand\todopar[1]{\fcolorbox{red}{white}{\parbox{0.97\linewidth}{#1}}}
+\else
+%    \usepackage[disable]{./todonotes} 
+   \newcommand\todo[1]{}
+\fi
+
+\newenvironment{narrow}[2]{%
+\begin{list}{}{%
+\setlength{\topsep}{0pt}%
+\setlength{\leftmargin}{#1}%
+\setlength{\rightmargin}{#2}%
+\setlength{\listparindent}{\parindent}%
+\setlength{\itemindent}{\parindent}%
+\setlength{\parsep}{\parskip}}%
+\item[]}{\end{list}}
+
+\usepackage{float}
+
+\ifOnlineColor
    \definecolor{tabBlue}{HTML}{AACCFF}
-\newlength\oldparindent
-\setlength\oldparindent{\parindent}
+\else
+   \definecolor{tabBlue}{HTML}{CCCCCC}
+\fi
 
-\newlength\figwidth
-\setlength\figwidth{0.7\linewidth}
+%%%%%%%%%%                                      ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
+% MACROS %       
+%%%%%%%%%%                                      ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
 
-% \setlength\floatsep{\baselineskip}
-\setcounter{topnumber}{1}
-% \setlength\baselineskip{12pt}
-% \selectfont
+\newcommand\graphicsAI[2][]{%
+  \immediate\write18{./bin/laFigure #2 #1}%
+  \input{result}}%
+
+% \DeclareMathOperator*{\argmin}{\text{arg}\;\text{min}}
+
+\newcommand\Fig[1]{Fig.~\ref{#1}}
+
+\newcommand\Grey[1]{{\color{Grey}#1}}
+\newcommand\Red[1]{{\color{Red}#1}}
+\newcommand\Blue[1]{{\color{Blue}#1}}
+\newcommand\DarkBlue[1]{{\color{DarkBlue}#1}}
+\newcommand\LightBlue[1]{{\color{LightBlue}#1}}
+\newcommand\Brown[1]{{\color{Brown}#1}}
+\newcommand\Green[1]{{\color{Green}#1}}
+\newcommand\SeaGreen[1]{{\color{SeaGreen}#1}}
+\newcommand\Yellow[1]{{\color{yellow}#1}}
+\newcommand\Orange[1]{{\color{orange}#1}}
+
+\newcommand\nn{\nonumber\\}
+
+\newcommand\nmat[1]{\begin{matrix}#1\end{matrix}}
+\newcommand\bmat[1]{\begin{bmatrix}#1\end{bmatrix}}
+\newcommand\case[1]{\begin{cases}#1\end{cases}}
+\newcommand\textbox[2]{\footnotesize\text{\parbox{#1}{\centering\emph{#2}}}}
+
+\newcommand\rand{\text{rand}}
+\newcommand\randn{\text{randn}}
+\newcommand\rect{\text{rect}}
+\newcommand\sinc{\text{sinc}}
+\newcommand\tr{\text{tr}}
+\newcommand\adj{\text{adj}}
+
+% \newcommand\max{\text{max}}
+\newcommand\argmin[1]{\text{arg}\;\underset{#1}{\text{min}}}
+
+\newcommand\qqquad{\quad\qquad}
+\newcommand\qqqquad{\qquad\qquad}
+
+% \renewcommand\l[1]{\left#1}
+% \renewcommand\r[1]{\right#1}
+
+% {\text{\parbox{1.5cm}{\centering volume hyper- sphere}}}
+
+%Keyword colouring:
+\newcommand\kw[1]{#1}
+\newcommand\parm[1]{#1}%\color{Black}#1\color{Black}}
+
+\newcommand\of[1]{\scriptstyle(\parm{#1})\displaystyle}
+\newcommand\df[1]{\scriptstyle[\parm{#1}]\displaystyle}
+\newcommand\var[3]{#1_\text{#2}\of{#3}}
+
+\newcommand\diag{\text{diag}}
+
+% \raisebox{lift}[extend-above-baseline][extend-below-baseline]{text}
+\newcommand\mt[1]{\text{\emph{#1}}} %mt = mathtext
+\newcommand\mathnorm{\textstyle}
+\newcommand\mathbig[1]{\displaystyle#1\mathnorm}
+\newcommand\mathsmall[1]{\scriptstyle#1\mathnorm}
+\newcommand\mathtiny[1]{\scriptscriptstyle#1\mathnorm}
+\newcommand\sfrac[2]{\scriptstyle\raisebox{0.25pt}[0pt][0pt]{$\frac{#1}{#2}$}\mathnorm}
+\newcommand\nfrac[2]{\textstyle\frac{#1}{#2}\displaystyle}
+
+\newcommand\sumu[1]{\sum\limits^{#1}\;}
+\newcommand\suml[1]{\sum\limits_{#1}\;}
+\newcommand\sumb[2]{\sum\limits_{#1}^{#2}\;}
+
+\newcommand\produ[1]{\prod\limits^{#1}\;}
+\newcommand\prodl[1]{\prod\limits_{#1}\;}
+\newcommand\prodb[2]{\prod\limits_{#1}^{#2}\;}
+
+\newcommand\defeq{\overset{\underset{\mathrm{def}}{}}{=}}
+
+%Math macros:
+\newcommand\T{^{\scriptscriptstyle T}}
+\renewcommand\H{^{\scriptscriptstyle H}}
 
@@
 \newcommand\mat[1]{\boldsymbol{#1}}
+
+\newcommand\Om{O_\text{m}}
+\newcommand\Oa{O_\text{a}}
+\newcommand\Nl{N_\text{l}}
+\newcommand\Nk{N_\text{k}}
+\newcommand\1{\vec 1}
+\newcommand\I{\mat I}
+\renewcommand*\a{\vec a}
+\renewcommand*\i{\vec i}
+\renewcommand*\k{\vec k}
+\newcommand*\n{\vec n}
+\newcommand*\p{\vec p}
+\newcommand*\s{\vec s}
+\newcommand*\w{\vec w}
+\newcommand*\x{\vec x}
+\newcommand*\y{\vec y}
+
+\newcommand*\A{\mat A}
+\newcommand*\B{\mat B}
+\newcommand*\C{\mat C}
+\newcommand*\E{\mat E}
+\renewcommand*\P{\mat P}
+\newcommand*\eP{\mat{\hat P}}
+\newcommand*\R{\mat R}
+\newcommand*\Ri{\R^{-1}}
+\newcommand*\eR{\mat{\hat R}}
+\newcommand*\eRi{\hat{\mat R}\;\!^{-1}}
+\newcommand*\Navg{N_\text{avg}}
+\newcommand*\W{\mat W}
+\newcommand*\X{\mat X}
+\newcommand*\Xd{\X_{\!\Delta}}
+\newcommand*\Y{\mat Y}
 
@@
 
-% \usepackage{caption}
-\usepackage{subcaption}
-\usepackage{color}
+\renewcommand*\L{\mat \Lambda}
+\newcommand*\U{\mat U}
+% \renewcommand*\t{\mathtiny{^T}}
+% \newcommand*\h{\mathtiny{^H}}
+\renewcommand*\t{^T}
+\newcommand*\h{^H}
 
-% Tune floats
-\renewcommand{\topfraction}{0.9}  % max fraction of floats at top
-\renewcommand{\bottomfraction}{0.8} % max fraction of floats at bottom
-%   Parameters for TEXT pages (not float pages):
-\setcounter{topnumber}{1}
-\setcounter{bottomnumber}{2}
-\setcounter{totalnumber}{4}     % 2 may work better
-\setcounter{dbltopnumber}{2}    % for 2-column pages
-\renewcommand{\dbltopfraction}{0.9} % fit big float above 2-col. text
-\renewcommand{\textfraction}{0.07}  % allow minimal text w. figs
-%   Parameters for FLOAT pages (not text pages):
-\renewcommand{\floatpagefraction}{0.85}  % require fuller float pages
-% N.B.: floatpagefraction MUST be less than topfraction !!
-\renewcommand{\dblfloatpagefraction}{0.85} % require fuller float pages
+\usepackage{tikz}
+\usetikzlibrary{shapes,snakes}
+\usepackage{amsmath,amssymb}
+% \usepackage{datatool}
+% \usepackage{glossaries}
+
+\newenvironment{outline}
+{\begin{itemize}}
+{\end{itemize}}
+
+% correct bad hyphenation here
+% \hyphenation{op-tical net-works semi-conduc-tor}
+
+%%%%%%%%%%%%                                    ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
+% GLOSSARY %
+%%%%%%%%%%%%                                    ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
+
+% \makeglossaries
+% \input{../../Library/glossary}
+
+%%%%%%%%%%%%%%%%%%                              ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
+% DOCUMENT START %
+%%%%%%%%%%%%%%%%%%                              ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
+
+% \usepackage{yfonts}
+
+% \graphicspath{submission/final/gfx/}
 
 \begin{document}
-% Load the references.bib file
-% \bibliography{../../../library.bib}
 
-\pagestyle{plain}
-% \layout
-% \printinunitsof{cm}
-% \prntlen{\baselineskip}
-% \the\textwidth
-% \the\linewidth
-% \prntlen{\textwidth}\\
-% \prntlen\linewidth\\
-% \prntlen\oddsidemargin\\
-% \prntlen\oddsidemargin
-% \newpage
+\title{A GPU Sonar Simulator for Automatic Target Recognition}
 
-\title{A GPU Sonar Simulator for Automatic Target Recognition}%
-%
-\author{Jo Inge Buskenes\firstAddress, Herman Midelfart\secondAddress, \O{}ivind Midtgaard\secondAddress}%
-%
-\begin{contact}
-  \firstAddress Dept. of Informatics, P.O.Box 1080 Blindern, N-0316 OSLO\\
-  \secondAddress Norwegian Defence Research Establishment (FFI) P.O.Box 25, NO-2027 Kjeller
-\end{contact}%
-%
-\begin{contact}
-Contact author: Jo Inge Buskenes\\
-Dept. of Informatics, Univ. of Oslo, P.Box 1080 Blindern, N-0316 OSLO\\
-\href{mailto:joibu@ifi.uio.no}{joibu@ifi.uio.no}
-\end{contact}%
-%
+\author{{Jo~Inge~Buskenes,~\IEEEmembership{Student Member,~IEEE,} %
+        Herman~Midelfart, %
+        \O{}ivind~Midtgaard} %
+\thanks{J. I. Buskenes is with the Department of Informatics, University of Oslo, Norway.}%
+\thanks{H. Midelfart and \O. Midtgaard are with The Norwegian Defence Research Establishment (FFI), Norway.}% <-this % stops a space
+% \thanks{Manuscript received April 19, 2005; revised January 11, 2007.}
+}
+
+% The paper headers
+\markboth{IEEE Journal of Oceanic Engineering}%
+{A GPU Sonar Simulator for Automatic Target Recognition}
+
+% Publishers ID mark:
+%\IEEEpubid{0000--0000/00\$00.00~\copyright~2007 IEEE}
+
+% use for special paper notices
+%\IEEEspecialpapernotice{(Invited Paper)}
+
+% for Computer Society papers, we must declare the abstract and index terms
+% PRIOR to the title within the \IEEEcompsoctitleabstractindextext IEEEtran
+% command as these need to go into the title area created by \maketitle.
+\IEEEcompsoctitleabstractindextext{%
 \begin{abstract}
@@
 The principle is to isolate an image segment containing the object of interest, correlate it with a set of template images, and assign it to the class of the template yielding the highest correlation coefficient.
-The challenge is to come up with a representative set of template images covering the actual configurations of object and seabed.
+The challenge is to come up with a representative set of template images covering the relevant configurations of object and seabed.
 
@@
 The result is a fast performing and portable on-the-fly template generator which can adapt to the characteristics of the current scene.
-\end{abstract}%
-%
-\keywords{Sonar, SAS, simulator, template matching, OpenGL, OpenCL}
+\end{abstract}
+
+% Keywords (normally not used for peer reviews)
+\ifPeerReview\else
+\begin{IEEEkeywords}
+Sonar, SAS, simulator, template matching, OpenGL, OpenCL.
+\end{IEEEkeywords}
+\fi}
+% \fi
+
+% make the title area
+\maketitle
+
+% This command fixes abstract positioning for compsoc articles:
+\IEEEdisplaynotcompsoctitleabstractindextext
+
+% (Optional) Add some extra info on cover page of peer review papers:
+% \ifCLASSOPTIONpeerreview
+% \begin{center} \bfseries EDICS Category: 3-BBND \end{center}
+% \fi
+
+% Insert page break and insert second title (peer review mode)
+\IEEEpeerreviewmaketitle
+
+\maketitle
 
@@
 \graphicsAI[drawing,width=\linewidth]{gfx/simulator.svg}%
-\caption{Simulator.}\label{buildup}%
+\caption{Simulator concept.}\label{buildup}%
 \end{figure}
@@
 
-\section{Conclusion}
+\section{Conclusion}\label{conclusion}
 
@@
 
+% \appendices
+% 
+% \section{MVDR Complexity Formulas}\label{mvdr_formulas}
+% 
+% \section{GPU Throughput}\label{throughput}
+% 
+% bs bs
+
+% use section* for acknowledgement
+\ifCLASSOPTIONcompsoc% % This command fixes abstract positioning for compsoc articles:
+% \IEEEdisplaynotcompsoctitleabstractindextext
+% 
+% % (Optional) Add some extra info on cover page of peer review papers:
+% % \ifCLASSOPTIONpeerreview
+% % \begin{center} \bfseries EDICS Category: 3-BBND \end{center}
+% % \fi
+% 
+% % Insert page break and insert second title (peer review mode)
+% \IEEEpeerreviewmaketitle
 % 
 % 
-% \begin{itemize}
-% \item Mer testing
-% \item Robustifisering
-% \item Tuning
-% \end{itemize}
 %
 
-% Ligger nesten med kortsiden til.
-% Bilde og templat segmenteres hver for seg.
-% Ikke adaptiv - darlig. Feil lengde, skyggen for kort (synket). Orientering feil. Feil spacing. 
-% Adaptiv - malt objekt. Passer bra. 
-% 
-% Videre:
-% - Mer testing
-% - Robustifisering
-% - Tuning
+  \section*{Acknowledgments}
+\else
+  \section*{Acknowledgment}
+\fi
+
+The authors would like to thank the Norwegian Defence Research Establishment (FFI) for funding to develop the simulator\todo{more?}.
+
+% Can use something like this to put references on a page
+% by themselves when using endfloat and the captionsoff option.
+\ifCLASSOPTIONcaptionsoff
+  \newpage
+\fi
+
+\bibliographystyle{IEEEtran}
+\bibliography{references}
+
+% Generated by IEEEtran.bst, version: 1.13 (2008/09/30)
+% \begin{thebibliography}{10}
+% stuff here
+% \end{thebibliography}
+
+\input{bio/jo_inge_buskenes}
+% \input{bio/jon_petter_aasen}
+% \input{bio/carl-inge_nilsen}
+% \input{bio/andreas_austeng}
 
 % 
-\bibliography{references}
-% \bibliography{../../../library.bib}
+% IMAGE METRICS
 %
-% % If biblatex (file is loaded in preamble):
-% % \printreferences
+% - Point spread function (res via main lobe width, SAS gain via PDF height)
+% - Constrast measures
+%
+% ACR = cL/2
+% C = (s+n)/n ratio
+% - 
+% WHO's needing processing power
+% - Centre for Maritime Research and Experimentation
+
+\vfill
 

### 2014-05-09 Simulator/article.tex
la2pdf commit - 2014:05:08	12:03:57	(Thu 8.May)
la2pdf commit - 2014:05:08	18:06:57	(Thu 8.May)
la2pdf commit - 2014:05:09	10:43:41	(Fri 9.May)

@@
 
+\begin{align*}
+\left(\begin{matrix}
+x_c \\ y_c \\ z_c \\ w_c
+\end{matrix}\right)
+= \left(\begin{matrix}
+\frac{2}{r-l}  &  0              &  0              &  -\frac{r+l}{r-l} \\
+0              &  \frac{2}{t-b}  &  0              &  -\frac{t+b}{t-b} \\
+0              &  0              &  \frac{-2}{f-n} &  -\frac{f+n}{f-n} \\
+0              &  0              &  0              &  1
+\end{matrix}\right)
+\left(\begin{matrix}
+x_e \\ y_e \\ z_e \\ w_e
+\end{matrix}\right)
+\end{align*}
+where
+\begin{align*}
+l &=   &  r &= \\
+b &=   &  t &= \\
+f &=   &  n &= 
+\end{align*}
+
 \subsection{OpenGL rendering}
@@
 
-% \appendices
-% 
+\appendix
+
+We implement rotations in a 3D euclidean space by combining matrices that rotate around each of the three respective dimensions..
+For rotation by an angle $\phi_x$ clockwise around the $x$-axis we have:
+\begin{align}
+\R_x(\phi_x) &= \bmat{
+1  &  0           &  0           &  0 \\
+0  &  \cos\phi_x  &  -\sin\phi_x &  0 \\
+0  &  \sin\phi_x  &  \cos\phi_x  &  0 \\
+0  &  0           &  0           &  1 \\
+},
+\end{align}
+and for the other two euclidean dimensions we have:
+\begin{align}
+\R_y(\phi_y) &= \bmat{
+\cos\phi_y  &  0  &  \sin\phi_y  &  0 \\
+0           &  1  &  0           &  0 \\
+-\sin\phi_y &  0  &  \cos\phi_y  &  0 \\
+0           &  0  &  0           &  1 \\
+},
+\end{align}
+and
+\begin{align}
+\R_z(\phi_z) &= \bmat{
+\cos\phi_z  &  -\sin\phi_z &  0  &  0 \\
+\sin\phi_z  &  \cos\phi_z  &  0  &  0 \\
+0           &  0           &  1  &  0 \\
+0           &  0           &  0  &  1 \\
+}.
+\end{align}
+The order in which these are applied matters. We rotate around the $x$-axis first, then around $y$ and finally around $z$:
+\begin{align}
+\R(\phi_x, \phi_y, \phi_z) &= \R_z(\phi_z)\;\R_y(\phi_y)\;\R_x(\phi_x).
+\end{align}
+
+\begin{align}
+\boldsymbol{T} &= \bmat{
+1  &  0  &  0  &  dx \\
+0  &  1  &  0  &  dy \\
+0  &  0  &  1  &  dz \\
+0  &  0  &  0  &  1 \\
+}
+\end{align}
+
+\begin{align}
+\boldsymbol{S} &= \bmat{
+s_x  &  0    &  0    &  1 \\
+0    &  s_y  &  0    &  1 \\
+0    &  0    &  s_z  &  1 \\
+0    &  0    &  0    &  1 \\
+}
+\end{align}
+
 % \section{MVDR Complexity Formulas}\label{mvdr_formulas}

### 2014-05-09 Simulator/article.tex
la2pdf commit - 2014:05:09	11:14:09	(Fri 9.May)

@@
 
+We place, resize and orient objects in an euclidean 3D space using suitable transformation matrices that are applied to each of the object's vertices:
+\begin{align}
+\bmat{x_t \\ y_t \\ z_t \\ w} &= \boldsymbol{T}\R\boldsymbol{S} \bmat{x \\ y \\ z \\ w}_\text{model}.
+\end{align}
+Here a model vertex at position $(x,y,z)$ is first scaled with $\boldsymbol{S}$, then rotated with $\R$
+There is nothing special to these matrices, but we supply them here for completeness.
+
 We implement rotations in a 3D euclidean space by combining matrices that rotate around each of the three respective dimensions..
@@
 \begin{align}
-\boldsymbol{T} &= \bmat{
+\boldsymbol{T}(dx,dy,dz) &= \bmat{
 1  &  0  &  0  &  dx \\
@@
 \begin{align}
-\boldsymbol{S} &= \bmat{
-s_x  &  0    &  0    &  1 \\
-0    &  s_y  &  0    &  1 \\
-0    &  0    &  s_z  &  1 \\
+\boldsymbol{S}(s_x,s_y,s_z) &= \bmat{
+s_x  &  0    &  0    &  0 \\
+0    &  s_y  &  0    &  0 \\
+0    &  0    &  s_z  &  0 \\
 0    &  0    &  0    &  1 \\

### 2014-05-12 Simulator/article.tex
la2pdf commit - 2014:05:09	12:00:25	(Fri 9.May)
la2pdf commit - 2014:05:09	12:46:36	(Fri 9.May)
la2pdf commit - 2014:05:12	13:12:15	(Mon12.May)
la2pdf commit - 2014:05:12	14:23:15	(Mon12.May)

@@
 
-%\usepackage{stfloats}                          % Allows: \begin{figure*}[!b]
+\usepackage{stfloats}                          % Allows: \begin{figure*}[!b]
                                                 % (double column figures on top/bottom)
@@
 
-\begin{figure}[t]\centering%
+\begin{figure*}[t]\centering%
 \graphicsAI[drawing,width=\linewidth]{gfx/simulator.svg}%
 \caption{Simulator concept.}\label{buildup}%
-\end{figure}
+\end{figure*}
 
@@
 where $\bmat{\cdot}_\text{model}$ is the loaded model, $\bmat{\cdot}_\text{image}$ is the output image, and $\M$, $\V$ and $\P$ are transformation matrices.
-The first transform we apply is the model matrix $\M$, which scales and rotates the model in its local coordinate system.
+The first transform we apply is the model matrix $\M$ which operates on the model's local coordinate system.
+It is defimed as
+\begin{align}
+\bmat{x_t \\ y_t \\ z_t \\ w}_\text{\parbox{1cm}{\setlength\baselineskip{0.3cm}transformed\\model}} &= \boldsymbol{T}\R\boldsymbol{S} \bmat{x \\ y \\ z \\ w}_\text{model}.
+\end{align}
+where $\boldsymbol{S}, \R and \boldsymbol{T}$ are matrices for scaling, rotation and translation, respectively. These are described in Appendix \ref{transformation_matrices}.
+
+which scales and rotates the model in its local coordinate system.
 Then we map this model into world coordinates using the view matrix $\V$.
@@
 
-\begin{align*}
-\left(\begin{matrix}
-x_c \\ y_c \\ z_c \\ w_c
-\end{matrix}\right)
-= \left(\begin{matrix}
+\begin{align}
+\P = \left(\begin{matrix}
 \frac{2}{r-l}  &  0              &  0              &  -\frac{r+l}{r-l} \\
@@
 \end{matrix}\right)
-\left(\begin{matrix}
-x_e \\ y_e \\ z_e \\ w_e
-\end{matrix}\right)
-\end{align*}
+\end{align}
 where
 \begin{align*}
-l &=   &  r &= \\
+l &= x_\text{min}  &  r &= x_\text{max}\\
 b &=   &  t &= \\
@@
 
+\section{Transformation matrices}\label{transformation_matrices} 
 We place, resize and orient objects in an euclidean 3D space using suitable transformation matrices that are applied to each of the object's vertices:
 \begin{align}
-\bmat{x_t \\ y_t \\ z_t \\ w} &= \boldsymbol{T}\R\boldsymbol{S} \bmat{x \\ y \\ z \\ w}_\text{model}.
+\bmat{x_t \\ y_t \\ z_t \\ w}_\text{\parbox{1cm}{\setlength\baselineskip{0.3cm}transformed\\model}} &= \boldsymbol{T}\R\boldsymbol{S} \bmat{x \\ y \\ z \\ w}_\text{model}.
 \end{align}

### 2014-05-12 Simulator/article.tex
la2pdf commit - 2014:05:12	15:21:52	(Mon12.May)

@@
 where $\bmat{\cdot}_\text{model}$ is the loaded model, $\bmat{\cdot}_\text{image}$ is the output image, and $\M$, $\V$ and $\P$ are transformation matrices.
-The first transform we apply is the model matrix $\M$ which operates on the model's local coordinate system.
-It is defimed as
+
+The first transform we apply is the model matrix $\M$ which operates on the model's local coordinate system. It is defimed as
 \begin{align}
@@
 \end{align}
-where $\boldsymbol{S}, \R and \boldsymbol{T}$ are matrices for scaling, rotation and translation, respectively. These are described in Appendix \ref{transformation_matrices}.
+where $\boldsymbol{S}, \R$ and $\boldsymbol{T}$ are matrices for scaling, rotation and translation, respectively. These are described in Appendix \ref{transformation_matrices}.
 
-which scales and rotates the model in its local coordinate system.
-Then we map this model into world coordinates using the view matrix $\V$.
+Next we map the model into world coordinates using the view matrix $\V$.
 This involves translating the model to its respectful place in the world, and then moving it into the view of the camera.
+It is defined as
+
+\begin{align}
+a
+\end{align}
+
 Finally the relevant part of the scene is projected onto the image plane with a projection matrix $\P$.

### 2014-05-12 Simulator/article.tex
la2pdf commit - 2014:05:13	00:29:57	(Tue13.May)
la2pdf commit - 2014:05:13	01:05:48	(Tue13.May)

@@
 Finally the relevant part of the scene is projected onto the image plane with a projection matrix $\P$.
-
+There are two main types of projection, perspective and orthographic.
+Perspective projection scales all $x$ and $y$ coordinates inversely with the distance to the objects to create a sense of perspective in the image.
+Orthographic projection, on the other hand, maps all the scene's vertices to a squared cube where each dimension has the range [-1,1].
+This will render all objects equally large whether they are in the distance or up close.
+Its projection matrix is defined as
 \begin{align}
@@
 \end{align}
-where
+where $l,r,b,t,n$ and $f$ are left, right, bottom, top, near and far boundaries of the projection plane, respectively.
 \begin{align*}
@@
 \end{align*}
+
+\begin{figure}[t]\centering%
+\graphicsAI[drawing,width=\linewidth]{gfx/opengl_part.svg}%
+\caption{Setting camera top and bottom boundaries.}\label{camera_top_bottom}%
+\end{figure}
 

### 2014-05-12 Simulator/article.tex
la2pdf commit - 2014:05:13	01:45:17	(Tue13.May)

@@
 \end{align}
-where $l,r,b,t,n$ and $f$ are left, right, bottom, top, near and far boundaries of the projection plane, respectively.
-\begin{align*}
-l &= x_\text{min}  &  r &= x_\text{max}\\
+where $l,r,b,t,n$ and $f$ are the left, right, bottom, top, near and far boundaries of the projection plane, respectively.
+They define the window through which we observe our scene.
+We want this adapted to perfectly fit the specified image size.
+This is straightforward for left and right since it matches the $x$-axis
+\begin{align}
+l &= x_\text{min}  &  r &= x_\text{max}
+\end{align}
+For the near and far limits one simply needs to make sure the scene will fit in it
+\begin{align}
 b &=   &  t &= \\
-f &=   &  n &= 
-\end{align*}
+f &= 0.1   &  n &= 200 
+\end{align}
+How to set the bottom and top bounaries can be inferred from \Fig{camera_top_bottom}. First we define an origin vector that points to the center of the image
+\begin{align}
+\vec o &= \left[\frac{x_\text{min}+x_\text{max}}{2}, \frac{y_\text{min}+y_\text{max}}{2}, \frac{z_\text{min}+z_\text{max}}{2} \right]\T,
+\end{align}
+and the distance from the image origin to its $y$-boundaries is given as
+\begin{align}
+\Delta y &= \frac{y_\text{max} - y_\text{min}}{2}.
+\end{align}
+To find the nearest and farthest point in the image that will be part of the projected scene, $\alpha$ and $\beta$, we must consider the seafloor rotation angle around the $x$-axis, $\Phi_x$ 
+\begin{align}
+\alpha &= \vec 0 - \left[0, \Delta y, \Delta y \times \arctan\Phi_\text{x} \right]
+\end{align}
 

### 2014-05-14 ThesisJP/Background/background.tex
Some small final corrections.
Some small corrections.

@@
 \end{align} 
-This corresponds to a white noise component proportional to the in-coherent signal power is added to the estimate \cite{Featherstone1997b}.
+This adds a white noise component proportional to the in-coherent signal power to the estimate \cite{Featherstone1997b}.
 
@@
 For ultrasound visualization a monotonically increasing function is typically used, which means that below a certain threshold all samples will be rendered transparent (opacity = 0).
-We will refer to this as the \textit{opacity threshold}, and for the example in Fig.\,\ref{fig:vr_otf} this threshold is 25.
+We will refer to this as the \textit{opacity threshold} (\nom{$\tau$}{Opacity threshold}), and for the example in Fig.\,\ref{fig:vr_otf} this threshold is 25.
 
@@
 
-In \textbf{Paper IV} we extend the work by Honigmann \textit{et al.} \cite{Honigmann2003} on a global adaptive OTF by formulating an regional adaptive OTF for visualization of cardiac ultrasound volumes.
+In \textbf{Paper IV} we extend the work by Honigmann \textit{et al.} \cite{Honigmann2003} on a global adaptive OTF by formulating a regional adaptive OTF for visualization of cardiac ultrasound volumes.
 This regional adaptive OTF is based on estimated statistics of the clutter noise inside the left ventricle (the blood pool).

### 2014-05-19 ThesisJP/Thesis.tex

@@
 \noindent\large 
-\textbf{Akselerering av adaptive algoritmer for ultralydavbildning ved hjelp a programmerbare grafikkprosessorer}
+\textbf{Akselerering av adaptive algoritmer for ultralydavbildning ved hjelp av programmerbare grafikkprosessorer}
 %\vspace{0.2cm}\\%
@@
 \normalsize
-En rivende utvikling innen spillteknologi og tilh\o{}rende programmeringsspr\aa{}k har nylig gjort super-datamaskiner i form av et lite grafikkort (GPU) tilgjengelige for forskere.
+En rivende utvikling innen spillteknologi og tilh\o{}rende programmeringsspr\aa{}k har nylig gjort superdatamaskiner i form av et lite programmerbart grafikkort (GPU) tilgjengelige for forskere.
 Denne utrolige \o{}kningen i beregningskraft og forbedret programmerbarhet er for \o{}yeblikket med p\aa{} \aa{} endre hvordan ultralydsystemer designes.
-N\aa{}r forskere utforsker nye algoritmer for ultralydavbildning er det derfor viktig at arkitekturen til parallelle akseleratorer tas h\o{}yde for.
+N\aa{}r forskere utforsker nye algoritmer for ultralydavbildning er det derfor viktig at arkitekturen til parallelle akseleratorer, som en GPU er et eksempel p\aa{}, tas h\o{}yde for.
 Hvis en ny, kompleks algoritme er tenkt til \aa{} kj\o{}res i sanntid, er det viktig at den kan tilpasses den programmerbare og parallelle prosesseringskjeden i moderne ultralydskannere.
@@
 I siste del av avhandlingen utforsker vi hvordan GPUer kan utnyttes til \aa{} akselerere ultralydsimuleringer.
-Resultatet av dette arbeidet ble et simuleringsprogram hvor geometrien til en gitt ultralydprobe kan interaktivt tegnes, og hvor det resulterende trykkfeltet simuleres i sanntid.
+Resultatet av dette arbeidet ble et simuleringsprogram hvor geometrien til et gitt ultralydarray kan interaktivt tegnes, og hvor det resulterende trykkfeltet simuleres i sanntid.
 
@@
 \addcontentsline{toc}{chapter}{Abstract}
-A rapid development in computer game technology and accompanying programming languages have recently provided researchers with small personal super computers, comprised in a single graphics card (GPU).
+A rapid development in computer game technology and accompanying programming languages have recently provided researchers with small personal supercomputers, comprised in a single graphics card (GPU).
 This immense rise in computational capabilities and improved programmability are currently changing how ultrasound imaging systems are designed.
-When researchers are exploring new algorithms for ultrasound imaging, it is therefore important to keep the architecture of parallel accelerators in mind.
+When researchers are exploring new algorithms for ultrasound imaging, it is therefore important to keep the architecture of parallel accelerators like the GPU in mind.
 If a new complex algorithm is supposed to run in real time, it needs to fit the programmable and parallel pipeline of modern ultrasound scanners.
@@
 Finally, we investigate how GPUs can be utilized to accelerate ultrasound simulations (\textbf{Paper\,V}).
-The result of this work was a simulation program where ultrasound probe geometries can be interactively drawn and where the resulting pressure field is simulated and visualized in real time.
+The result of this work was a simulation program where ultrasound array geometries can be interactively drawn and where the resulting pressure field is simulated and visualized in real time.
 
@@
 %\thispagestyle{empty}%
-The present thesis is submitted in partial fulfillment of the requirements for the degree of \textit{Philosophiae Doctor} (Ph.D.) at the Faculty of Medicine at the Norwegian University of Science and Technology (NTNU).
+This thesis is submitted in partial fulfillment of the requirements for the degree of \textit{Philosophiae Doctor} (Ph.D.) at the Faculty of Medicine at the Norwegian University of Science and Technology (NTNU).
 The research was funded by \textit{Medical Imaging Laboratory} (MI-Lab), and was carried out at the Department of Circulation and Medical Imaging.

### 2014-05-21 ThesisJP/Introduction/introduction.tex

@@
 \section{Motivation}
-A rapid development in computer game technology and accompanying programming languages have recently provided researchers with small personal super computers, comprised in a single graphics card (\nom{GPU}{Graphics processing unit}).
+A rapid development in computer game technology and accompanying programming languages have recently provided researchers with small personal supercomputers, comprised in a single graphics card (\nom{GPU}{Graphics processing unit}).
 The latest graphics cards\footnote{As of January 2014.} from both \nom{Nvidia}{Graphics card vendor} and \nom{AMD}{Graphics card vendor} provide more than 5 T\nom{FLOPS}{Floating point operations per second} single precision and 1.5 TFLOPS double precision computations.
@@
 
-The Capon beamformer in ultrasound imaging is a good example of an algorithm that has been widely studied over the last decade where the number of calculations was thought to be too high for real-time processing.
+The Capon beamformer in ultrasound imaging is a good example of an algorithm that has been widely studied over the last decade where the number of calculations was thought to be too high for real-time processing to happen in near future.
 To achieve real-time processing for e.g.

### 2014-05-23 Simulator/article.tex
More on simulator article

@@
 Various simulators for high frequency, side-looking sonar imagery have been published, e.g.~\cite{Bell1997,Sammelm2003}.
-Most implementations have however prioritized accurate acoustic modeling at the expense of execution speed.
+However, most implementations have prioritized accurate acoustic modeling at the expense of execution speed.
 One exception is the SIGMAS+ simulator in~\cite{Coiras2009a, Coiras2009b} which, similar to our approach, takes advantage of parallel processing at GPUs and uses OpenGL to render a 2D image from a 3D model, under the assumption of Lambertian scattering.
@@
 Also, their simulator sums images obtained from multiple OpenGL rendering passes to create the sonar image, while we render once and post-process this result with OpenCL to tailor the sonar image with more flexibility.
+
+\begin{itemize}
+\item OpenGL does not support scattered writes. OpenCL does.
+\end{itemize}
 
@@
 This section explains how we set up OpenGL for this task, and then proceeds to describe how we post process the OpenGL images with OpenCL to form the final sonar image.
+
+\begin{itemize}
+\item Load arbitrary complex models
+\item Simulate sonar with OpenGL
+\item Speed considerations
+\end{itemize}
 
@@
 \end{align}
-where $l,r,b,t,n$ and $f$ are the left, right, bottom, top, near and far boundaries of the projection plane, respectively.
-They define the window through which we observe our scene.
+where $l,r,b$ and $t$ are the left, right, bottom and top boundaries of the projection plane, respectively, and where $n$ and $f$ is the near and far limits of the cube that encompass the scene.
+We need these parameters set They define the window through which we observe our scene.
 We want this adapted to perfectly fit the specified image size.
@@
 \begin{align}
-b &=   &  t &= \\
-f &= 0.1   &  n &= 200 
+n &= 0\;\text{m}   &  f &= 300\;\text{m} 
 \end{align}
@@
 The ``camera image'' rendered with OpenGL is not in along-track and cross-track coordinates as we want it to be.
-It does, however, tell us what parts of the scene that are visible from the sonar.
+It does, however, show us what parts of the scene that are visible from the sonar.
 OpenGL can also produce a depth map that reveals the distance to each of image pixels.
 This information can be converted to a ranged sonar-like image by simply adding up all the intensity values that share the same depth for each range line.
-We perform this computation in OpenCL, which allow general purpose programming on GPUs and can interoperate with OpenGL quite nicely.
+We perform this computation in OpenCL.
+It allows general purpose programming on GPUs and can interoperate with OpenGL quite nicely.
 This way we keep all the calculations on the GPU.
 
+\begin{itemize}
+\item Kernel tuning.
+\item Sharing. PBO. Asynchronous.
+\end{itemize}
+
 % It is also possible to compute the sonar image using the OpenGL ``blend'' feature.
+
+\subsection{Speed considerations}
+
+Using OpenGL and OpenGL to generate sonar templates on the GPU allow hundreds of sonar templates to be formed per second.
+While this is sufficient for our needs.
+However, initializing the OpenGL and OpenCL context and loading 3D models from file into it is a slow process, and typically takes several seconds for every invocation.
+Therefore, to alleviate the simulation performance we either had to supply it with a list of parameters or ensure that it did not have to be shut down for every invocation.
+
+The rendering loop that uses OpenGL and OpenCL to form
+
+\begin{itemize}
+\item Write to file
+\end{itemize}
 

### 2014-05-23 Simulator/article.tex
la2pdf commit - 2014:05:23	15:23:07	(Fri23.May)
la2pdf commit - 2014:05:23	15:53:28	(Fri23.May)

@@
 This is a portable open source library that supports loading a wide range of 3D model formats in a uniform manner.
-Given a model file it outputs the model as a node tree where node contains data such as vertices, facets and textures, all formatted in an OpenGL friendly way.
+It converts the model into a node tree where each node contains data such as vertices, facets and textures, all formatted in an OpenGL friendly way.
 
@@
 This is achieved by applying a set of transformations to each vertex in the model.
-In OpenGL, the set of transformations is usually formalized like this:
+In OpenGL, the set of transformations are usually formalized like this:
 \begin{align*}
@@
 \end{align*}
-where $\bmat{\cdot}_\text{model}$ is the loaded model, $\bmat{\cdot}_\text{image}$ is the output image, and $\M$, $\V$ and $\P$ are transformation matrices.
+where $\bmat{\cdot}_\text{model}$ is the loaded model, $\bmat{\cdot}_\text{image}$ is the output image, and $\M$, $\V$ and $\P$ are transformation matrices. These will be described next.
 
-The first transform we apply is the model matrix $\M$ which operates on the model's local coordinate system. It is defimed as
+The first transform we apply is the model matrix $\M$ which operates in the model's local coordinate system. It is defined as
 \begin{align}
-\bmat{x_t \\ y_t \\ z_t \\ w}_\text{\parbox{1cm}{\setlength\baselineskip{0.3cm}transformed\\model}} &= \boldsymbol{T}\R\boldsymbol{S} \bmat{x \\ y \\ z \\ w}_\text{model}.
+\bmat{x_t \\ y_t \\ z_t \\ w}_\text{\parbox{1cm}{\setlength\baselineskip{0.3cm}transformed\\model}} &= \boldsymbol{T} \cdot \R \cdot \boldsymbol{S} \bmat{x \\ y \\ z \\ w}_\text{model}.
 \end{align}
@@
 Next we map the model into world coordinates using the view matrix $\V$.
-This involves translating the model to its respectful place in the world, and then moving it into the view of the camera.
-It is defined as
-
+This involves positioning and orienting the model relative to the camera.
+In our case we let the camera be the phase center of the sonar, and define it as the world origin.
+The transformation take the form of
 \begin{align}
-a
+\V &= \mat{T}(0, dy, dz) \cdot \R_\text{x}(\phi_x, 0, 0) \nonumber\\
+&= 
+\bmat{
+1  &  0  &  0  &  0 \\
+0  &  1  &  0  &  dy \\
+0  &  0  &  1  &  dz \\
+0  &  0  &  0  &  1 \\
+} \cdot
+\bmat{
+1  &  0           &  0           &  0 \\
+0  &  \cos\phi_x  &  -\sin\phi_x &  0 \\
+0  &  \sin\phi_x  &  \cos\phi_x  &  0 \\
+0  &  0           &  0           &  1 \\
+},
 \end{align}
+where
+\begin{align*}
+dy &= \\
+dz &= \\
+\phi_x &=
+\end{align*}
 

### 2014-05-23 Simulator/article.tex
la2pdf commit - 2014:05:23	16:36:55	(Fri23.May)
la2pdf commit - 2014:05:23	18:00:53	(Fri23.May)

@@
 We assume rough, isotropic surfaces reflecting sound energy equally in all directions.
-This can be modeled with a Lambertian scattering model~\cite{Blake1993,Bell1995}, where the intensity of the backscattered sound only depends on the incidence angle of the transmitted signal onto the model surface.
+This can be modeled with a Lambertian scattering model~\cite{Blake1993,Bell1995}, where the intensity of the backscattered sound solely depends on the incidence angle of the transmitted signal onto the model surface.
 When rendering OpenGL is set up to produce an optical 2D intensity image and depth map that reveals the distance from each pixel to the propagation axis.
@@
 This is achieved by applying a set of transformations to each vertex in the model.
-In OpenGL, the set of transformations are usually formalized like this:
+In OpenGL, the set of transformations are usually formalized as
 \begin{align*}
@@
 
-The first transform we apply is the model matrix $\M$ which operates in the model's local coordinate system. It is defined as
+The first transform we apply is the model matrix $\M$, which operates in the model's local coordinate system. It is defined as
 \begin{align}
-\bmat{x_t \\ y_t \\ z_t \\ w}_\text{\parbox{1cm}{\setlength\baselineskip{0.3cm}transformed\\model}} &= \boldsymbol{T} \cdot \R \cdot \boldsymbol{S} \bmat{x \\ y \\ z \\ w}_\text{model}.
+\bmat{x_t \\ y_t \\ z_t \\ w}_\text{\parbox{1cm}{\setlength\baselineskip{0.3cm}transformed\\model}} &= \boldsymbol{T} \cdot \R \cdot \boldsymbol{S} \bmat{x \\ y \\ z \\ w}_\text{model},
 \end{align}
@@
 dz &= \\
-\phi_x &=
+\phi_x &= \arctan\left(\frac{y_\text{mean}}{x_\text{mean}}\right)
 \end{align*}
@@
 We implement rotations in a 3D euclidean space by combining matrices that rotate around each of the three respective dimensions..
-For rotation by an angle $\phi_x$ clockwise around the $x$-axis we have:
+For rotation by an angle $\phi_x$ clockwise around the $x$-axis we have
 \begin{align}
@@
 \end{align}
-and for the other two euclidean dimensions we have:
+and similarly, for the other two euclidean dimensions we have
 \begin{align}

### 2014-05-24 Simulator/article.tex
Added a few stray figures

@@
 colorlinks=true,linkcolor={black},citecolor={black},urlcolor={black},
-pdftitle={An Optimized GPU Implementation of the MVDR Beamformer for Active Sonar Imaging},
+pdftitle={A Fast GPU Sonar Simulator for Automatic Target Recognition},
 pdfsubject={},

### 2014-05-28 ThesisJP/Background/background.tex
Started integrated comments from sverre.

@@
 The two essays spin around the fact that manufacturers of central processing units (CPUs) hit a frequency wall in the beginning of the 21st century.
-Until then, all mainstream computer programs where typically running in a single thread on a single CPU \textit{processing core}, and a programmer could expect the software to annually gain performance without touching the code.
+Until then, all mainstream computer programs were typically running in a single thread on a single CPU \textit{processing core}, and a programmer could expect the software to annually gain performance without touching the code.
 This was the "free-lunch" era as depicted in Fig.\,\ref{fig:jungle}.

### 2014-05-28 ThesisJP/Thesis.tex
Started integrated comments from sverre.

@@
 \addcontentsline{toc}{chapter}{Abstract}
-A rapid development in computer game technology and accompanying programming languages have recently provided researchers with small personal supercomputers, comprised in a single graphics card (GPU).
+A rapid development in computer game technology and accompanying programming languages have recently provided researchers with small personal supercomputers, comprised in a single graphics processing unit (GPU).
 This immense rise in computational capabilities and improved programmability are currently changing how ultrasound imaging systems are designed.
@@
 Real-time frame rates are achieved for both modalities.
-\textbf{Paper\,II} also present, for the first time, videos where the Capon beamformer has been applied on loops of simulated and \textit{in vivo} medical ultrasound images.
+\textbf{Paper\,II} also presents, for the first time, videos where the Capon beamformer has been applied on loops of simulated and \textit{in vivo} medical ultrasound images.
 In \textbf{Paper\,III}, we show that Capon beamforming does not provide shift-invariant imaging in a real-time imaging setting.
@@
 In paper \textbf{Paper\,IV} we propose an adaptive method for visualization of volumetric cardiac ultrasound images.
-The method is capable of removing noise that by conventional methods would have occlude cardiac tissue.
+The method is capable of removing noise that by conventional methods would have occluded cardiac tissue.
 This work also shows that with modern GPUs it is possible to add advanced visualization methods to an ultrasound imaging system and still have real-time performance.
@@
 The research was funded by \textit{Medical Imaging Laboratory} (MI-Lab), and was carried out at the Department of Circulation and Medical Imaging.
-The main supervisor has been Professor Sverre Holm in the Department of Informatics, University of Oslo (UiO), and Professor II at MI-Lab, NTNU.
+The main supervisor has been Professor Sverre Holm in the Department of Informatics (IFI), University of Oslo (UiO), and Professor II at MI-Lab, NTNU.
 Co-supervisors have been Ph.D.
@@
 I will never forget our conference trips.
-A special thanks also goes to Jo Inge Buskenes at "office" 4470 for making every working hour so much fun.
+A special thanks also goes to Jo Inge Buskenes in "office" 4470 at IFI, UiO for making every working hour so much fun.
 
@@
 
-Finally I would like to thank my family and all my friends for their support, and for providing a world away from work.
-But most of all, I want to thank Beate for her love and patient.
+Finally I would like to thank my family and all my friends for being there, and for providing a world away from work.
+But most of all, I want to thank Beate for her love and patience.
 Without your support I would never have managed to complete this thesis.

### 2014-05-28 Simulator/article.tex
More simulator

@@
 \ifPeerReview
-\documentclass[12pt,journal,draftclsnofoot,onecolumn]{IEEEtran}
+\documentclass[10pt,journal,draftclsnofoot,onecolumn]{IEEEtran}
 % \newcommand\CLASSINPUTbaselinestretch{1.66}     % http://theoval.cmp.uea.ac.uk/~nlct/latex/thesis/node17.html
@@
 \RequirePackage{array}
-% 
-% \ifCLASSOPTIONcompsoc
-%    \usepackage[caption=false,font=normalsize,labelfont=sf,textfont=sf]{subfig}
-% \else
-%    \usepackage[caption=false,font=footnotesize]{subfig}
-% \fi
+
+\ifCLASSOPTIONcompsoc
+   \usepackage[caption=false,font=normalsize,labelfont=sf,textfont=sf]{subfig}
+\else
+   \usepackage[caption=false,font=footnotesize]{subfig}
+\fi
 
@@
 However, most implementations have prioritized accurate acoustic modeling at the expense of execution speed.
-One exception is the SIGMAS+ simulator in~\cite{Coiras2009a, Coiras2009b} which, similar to our approach, takes advantage of parallel processing at GPUs and uses OpenGL to render a 2D image from a 3D model, under the assumption of Lambertian scattering.
+One exception is the SIGMAS+ simulator in~\cite{Coiras2009a, Coiras2009b}.
+Similarly to our approach it takes advantage of parallel processing on GPUs, and it uses OpenGL to render a 2D image from a 3D model, under the assumption of Lambertian scattering.
 However, SIGMAS+ is still aimed more towards realistically looking images, including effects like ambient noise, which is irrelevant for our purpose of template generation.
@@
 This is achieved by applying a set of transformations to each vertex in the model.
-In OpenGL, the set of transformations are usually formalized as
+In OpenGL, the set of transformations can be formalized as
 \begin{align*}
@@
 \end{align}
-where $\boldsymbol{S}, \R$ and $\boldsymbol{T}$ are matrices for scaling, rotation and translation, respectively. These are described in Appendix \ref{transformation_matrices}.
+where $\boldsymbol{S}, \R$ and $\boldsymbol{T}$ are matrices\todo{be explicit on dependencies?} for scaling, rotation and translation, respectively.
+These are provided for reference in Appendix \ref{transformation_matrices}.
 
 Next we map the model into world coordinates using the view matrix $\V$.
-This involves positioning and orienting the model relative to the camera.
-In our case we let the camera be the phase center of the sonar, and define it as the world origin.
+This involves aligning and positioning the model into the view of the camera.
+In our case the camera represents the phase center of the sonar, and it is placed in the world origin.
 The transformation take the form of
 \begin{align}
-\V &= \mat{T}(0, dy, dz) \cdot \R_\text{x}(\phi_x, 0, 0) \nonumber\\
+\V &= \mat{T}(x_\text{c}, y_\text{c}, z_\text{c}) \cdot \R_\text{x}(\phi_x, 0, 0) \nonumber\\
 &= 
 \bmat{
-1  &  0  &  0  &  0 \\
-0  &  1  &  0  &  dy \\
-0  &  0  &  1  &  dz \\
+1  &  0  &  0  &  x_\text{c} \\
+0  &  1  &  0  &  y_\text{c} \\
+0  &  0  &  1  &  z_\text{c} \\
 0  &  0  &  0  &  1 \\
@@
 \end{align}
-where
+where $(x_\text{c}, y_\text{c}, z_\text{c})$ is the image center coordinates for along track, cross track, and depth, respectively, and $\Phi_x$ is the elevation angle\todo{is this correct?} of the sonar.
+It is given as
 \begin{align*}
-dy &= \\
-dz &= \\
-\phi_x &= \arctan\left(\frac{y_\text{mean}}{x_\text{mean}}\right)
-\end{align*}
+\phi_x &= \arctan\left(\frac{y_\text{mean}}{-z_\text{mean}}\right).
+\end{align*}\todo{use absolute sign?}
 
@@
 This will render all objects equally large whether they are in the distance or up close.
-Its projection matrix is defined as
+The matrix that gives orthographic projection is defined as
 \begin{align}
@@
 where $l,r,b$ and $t$ are the left, right, bottom and top boundaries of the projection plane, respectively, and where $n$ and $f$ is the near and far limits of the cube that encompass the scene.
-We need these parameters set They define the window through which we observe our scene.
-We want this adapted to perfectly fit the specified image size.
-This is straightforward for left and right since it matches the $x$-axis
+These parameters need to be adapted to provide a 2D image that matches the image size, which we specify in along track and cross track coordinates ($x,y$).
+This is straightforward for left and right since matches the $x$-axis
+% They define the window through which we observe our scene.
 \begin{align}
@@
 \begin{figure}[p]\centering%
-\begin{subfigure}[t]{.5\linewidth-\imgspacing/2}
+\subfloat[SAS image of a cylinder with 2.6\;m length and 0.53\;m radius.]{%
 \graphicsAI[drawing,width=\linewidth]{gfx/data_cylinder_submerged.svg}%
-\caption{SAS image of a cylinder with 2.6\;m length and 0.53\;m radius.}\label{data_cylinder}%
-\end{subfigure}
-\begin{subfigure}[t]{.5\linewidth-\imgspacing/2}
+\label{data_cylinder}}%
+\subfloat[Cylinder simulation with parameters estimated from SAS image: Length 2.6\;m, burial depth 0.263\;m, aspect angle 105$^\circ$.]{
 \graphicsAI[drawing,width=\linewidth]{gfx/sim_cylinder_submerged_adaptive.svg}%
-\caption{Cylinder simulation with parameters estimated from SAS image: Length 2.6\;m, burial depth 0.263\;m, aspect angle 105$^\circ$.}\label{sim_cylinder}%
-\end{subfigure}
+\label{sim_cylinder}}%
 \caption{A SAS image of a cylinder and a template simulation adapted to it.}
@@
 \fbox{\colorbox[HTML]{#1}{\vrule height8.5pt depth3.5pt width0pt\hspace{.5cm}}}\ \ #2\\}}
+
 \begin{figure}[p]\centering%
-\begin{subfigure}[t]{.5\linewidth-\imgspacing/2}
+\subfloat[Best template out of a predefined set. Inaccurate cylinder length, alignment and immersion depth. The segmented shadow and highlight regions are misaligned.]{
 \graphicsAI[drawing,width=\linewidth]{gfx/overlay_cylinder_submerged.svg}%
-\caption{Best template out of a predefined set. Inaccurate cylinder length, alignment and immersion depth. The segmented shadow and highlight regions are misaligned.}\label{overlay_cylinder}%
-\end{subfigure}\hspace{\imgspacing}%
-\begin{subfigure}[t]{.5\linewidth-\imgspacing/2}
+\label{overlay_cylinder}}%
+% \hfil%
+\subfloat[Adapted template. Template simulated with adapted parameters. Good fit in both highlight and shadow segments.]{%
 \graphicsAI[width=\linewidth]{gfx/overlay_cylinder_submerged_adaptive.svg}%
-\caption{Adapted template. Template simulated with adapted parameters. Good fit in both highlight and shadow segments.}\label{overlay_cylinder_adaptive}%
-\end{subfigure}
-\hfill\parbox{\linewidth-0pt}{\small\centering
+\label{overlay_cylinder_adaptive}}%
+\hfill\parbox{\linewidth}{\small\centering
 Segmentation overlay color description:\\[.5\baselineskip]
-\begin{minipage}{.49\linewidth}
+\begin{minipage}{\linewidth}
 \cdesc{800000}{Template highlight and image highlight}
@@
 \end{minipage}%
-\begin{minipage}{.49\linewidth}
+\begin{minipage}{\linewidth}
 \cdesc{000083}{Template shadow and image shadow}
@@
 Hence, the standard approach was less likely to classify the cylinder in the image correctly as the standard templates were not created specifically for the target.
+
+\begin{itemize}
+\item GPUs low power. Good for integration in AUVs.
+\item Good accuracy even with a real-time constraint in e.g. a AUV.
+\item Improved inertial navigation? 
+\end{itemize}
 

### 2014-05-31 ThesisJP/Background/background.tex
Finished all comments from Sverre on the english text.

@@
 These measures reduced the silicon footprint of each core and made it possible to add multiple cores to the GPU before CPU designers where able to do the same.
-Today, even if more advanced caching has been added to GPUs, this is still the main reason why GPUs have higher peak performance than CPUs (Fig.\,\ref{fig:cpu_gpu} and \ref{fig:cpu_vs_gpu}).
+Today, even if more advanced caching has been added to GPUs, this is still the main reason why GPUs have higher peak performance than CPUs (Fig.\,\ref{fig:cpu_vs_gpu}).
 Another reason is the inherent parallel nature of the problem of rendering graphics.
@@
 How SIMD instructions are utilized in modern CPUs and GPUs is discussed in Section \ref{sec:cpu_vs_gpu}.
+
+\begin{figure}[t!]
+\centering
+\includegraphics[width=\textwidth]{img/cpu_vs_gpu.pdf}
+\caption{Development in theoretical peak throughput, single precision (SP) and double precision (DP), the last decade for Intel CPUs and Nvidia GPUs (github.com/mgalloy/cpu-vs-gpu).}
+\label{fig:cpu_vs_gpu}
+\end{figure}
 
@@
 The final transition currently taking place in the computer industry is the introduction of cloud computing, where compute power is provided as a web service and automatically scales on demand.
-Together, multi-cores, heterogeneous cores, and cloud cores form the "jungle" that today's software engineers have to handle (Fig.\,\ref{fig:jungle}).
+Together, multi-cores, heterogeneous cores, and cloud cores form the "jungle" that today's software engineers have to handle.
 In this thesis we focus on heterogeneous computing with a multi-core CPU and a discrete high performance GPU.
@@
 \end{equation}
-where FLOPS is short for; floating point operations per second.
+where FLOPS stands for floating point operations per second.
 This is the same number as visualized in Fig.\,\ref{fig:cpu_vs_gpu} for the Haswell architecture (single precision).
@@
 One Kepler GPU therefore provides more than 5 TFLOPS as shown in Fig.\,\ref{fig:cpu_vs_gpu} for the GeForce GTX 780 TI GPU.
-From Fig.\,\ref{fig:cpu_gpu} we see how the high number of SIMD cores found on GPUs and their larger width, compared to the SIMD cores on CPUs, result in five times the computational performance of a high-end CPU (without the integrated GPU).
+From Fig.\,\ref{fig:cpu_gpu} we see how the high number of SIMD cores found on GPUs and their larger width, compared to the SIMD cores on CPUs, results in five times the computational performance of a high-end CPU (without the integrated GPU).
 
@@
 
-\begin{figure}[t!]
-\centering
-\includegraphics[width=\textwidth]{img/cpu_vs_gpu.pdf}
-\caption{Development in theoretical peak throughput, single precision (SP) and double precision (DP), the last decade for Intel CPUs and Nvidia GPUs (github.com/mgalloy/cpu-vs-gpu).}
-\label{fig:cpu_vs_gpu}
-\end{figure}
-
 \subsection{Programming a GPU}
 Early GPGPU programming required expert knowledge of computer graphics and the graphics pipeline.
-When GPGPU gained a broader interest and the GPU started to be used to compute problems that did not involve graphics, it quickly became evident that new programing languages where needed.
+When GPGPU gained a broader interest and the GPU started to be used to compute problems that did not involve graphics, it quickly became evident that new programing languages were needed.
 One of the first GPGPU oriented languages was the Brook language by Buck \textit{et al.} \cite{Buck2004}.
-Buck later joined Nvidia where he led the work on \nom{CUDA}{Compute unified device architecture (GPGPU framework)}, a GPGPU language and framework by Nvidia for Nvidia GPUs only.
+Buck later joined Nvidia where he led the work on \nom{CUDA}{Compute unified device architecture (GPGPU framework)} (Compute unified device architecture), a GPGPU language and framework by Nvidia for Nvidia GPUs only.
 The first version of CUDA was released in 2007.
@@
 While a CPU is said to execute threads which can perform SIMD instructions, a thread on a GPU can be interpreted as a unit capable of processing one data element in the SIMD instruction.
-Because of this the GPU is sometimes referred to as a single instruction multiple threads (\nom{SIMT}{Single instruction multiple threads}) processor instead of a SIMD processor.
+Because of this the GPU is sometimes referred to as a single instruction multiple threads (\nom{SIMT}{Single instruction multiple threads}) processor instead of a single instruction multiple data (SIMD) processor.
 This also explain why GPU threads are said to be light-weighted.
@@
 It is important to understand that not all algorithms will benefit from wider SIMD registers.
-As the width increase, the only algorithms that will benefit are those which can be divided into a sufficiently large number of independent calculations.
+As the width increases, the only algorithms that will benefit are those which can be divided into a sufficiently large number of independent calculations.
 Also, to achieve peak performance, the implementation must perform one FMA operation on each core per clock cycle, and it must be possible for the GPU to hide all memory latency by scheduling threads which are not stalled by a memory fetch.
@@
 Therefore, we will probably see a lot of work aimed at reducing this overhead in the future.
-A recent development is the introduction of C++ AMP which enables developers to write accelerated code (e.g.
+A recent development is the introduction of \nom{C++ AMP}{C++ Accelerated Massive Parallelism (programming model)} which enables developers to write accelerated code (e.g.
 by a GPU) in plain C++.
@@
 Throughout the thesis we will refer to some of the parts shown in Fig.\,\ref{fig:human_heart}.
-Especially note the four chambers of the heart, the left and right atrium, and the left and right ventricle.
+Especially note the four chambers of the heart, the left and right atria, and the left and right ventricles.
 The tissue separating the left and right ventricle is known as the inter-ventricular septum.
-The left side is responsible of pumping blood out to the whole body through the aortic valve.
+The left side is responsible for pumping blood out to the whole body through the aortic valve.
 It is therefore the strongest and hardest working part of the heart.
@@
 A set of piezoelectric elements organized in an array is referred to as an ultrasound \textit{probe}, and special probes are typically designed for different medical imaging modalities.
-An example of a probe for medical ultrasound imaging is shown in Fig.\,\ref{fig:us_grid}.
+Examples of probes for imaging of the carotid artery in the neck and for cardiac imaging are shown in Fig.\,\ref{fig:us_grid}, to the left and right respectively.
 %For cardiac ultrasound imaging the probe must not only be designed for a certain frequency range, its footprint also needs to be smaller than the intercostal window in order to avoid shadowing by ribs.
 
-If the width of a piezoelectric element is less than the wavelength of the transmitted signal the element is said to be \textit{omnidirectional}, hence, energy will be distributed or sensed equally in all directions.
+If the width of a piezoelectric element is less than the wavelength of the transmitted signal the element approaches \textit{omnidirectional} characteristics, hence, energy will be distributed or sensed equally in all directions.
 When multiple omnidirectional elements are arranged in an array, the transmitted pulse will get shaped into a beam\footnote{A beam refers to the accumulated magnitude across time of a pulse traveling in space.} of ultrasound by diffraction as the width (the \textit{aperture}) of the array grows.
@@
 \subsection{Sampling and processing complexity}
-When generating a B-mode image, we have to ensure that the spatial and temporal sampling are sufficient.
+When generating a B-mode image, we have to ensure that the spatial and temporal sampling is sufficient.
 We have already seen that the lateral resolution of an ultrasound probe is given by (\ref{eq:res}).
@@
 A short ultrasound pulse, which would yield a high axial resolution, requires therefore a large frequency bandwidth.
-Medical ultrasound systems typically has a bandwidth between 33 \% and 100 \% of the pulse frequency.
+Medical ultrasound systems typically have a bandwidth between 33 \% and 100 \% of the pulse frequency.
 This corresponds to transmitting a pulse which is between 1 and 3 oscillations long.
@@
 \nomi{$N$}{Number of samples in range (or time)}where $f_c$ is the center frequency of the transmitted pulse, and where \nom{$B_w$}{Relative bandwidth} is the relative bandwidth in percent of \nom{$f_c$}{Center frequency of ultrasound pulse}.
-Note that after this process the ultrasound data will be complex values.
-The total number of samples in an B-mode image from a phased array is then
+Note that after this process the ultrasound data will be complex valued.
+The total number of samples in a B-mode image from a phased array is then
 \begin{align}
@@
 This is far below the peak processing throughput of modern GPUs and CPUs.
-The main challange with DAS beamforming is the amount of data that has to be moved.
+The main challenge with DAS beamforming is the amount of data that has to be moved.
 In our example, for a sample size of four bytes, around 1.2 GB has to be moved to the GPU every second.
 This is however less than the maximum bandwidth of PCI Express 2.0 (8 GB/s), even after the relative bandwidth, transmit frequency and number of array elements are increased with a total factor of 5.
-Newer versions of PCI Express, which supports even higher transfer rates, have also recently been introduced.
+Newer versions of PCI Express, which support even higher transfer rates, have also recently been introduced.
 
@@
 Software beamforming opens up for cost-efficient and rapid prototyping of advanced processing of ultrasound data.
-One example is Capon beamforming which is explored in this thesis (\textbf{Paper\,I} and \textbf{II}) and introduced in the next section.
-The increased computational power introduced by GPUs makes it also possible to include more advanced 3D visualization as presented in \cite{solteszova2010multidirectional}, \textbf{Paper\,IV}, and \textbf{IX}.
+One example is Capon beamforming which is explored in this thesis (\textbf{Papers\,I} and \textbf{II}) and introduced in the next section.
+The increased computational power introduced by GPUs makes it also possible to include more advanced 3D visualization as presented in \cite{solteszova2010multidirectional}, \textbf{Papers\,IV} and \textbf{IX}.
 
@@
 However, in this thesis adaptive beamforming will refer to data-dependent weighting derived from a given optimization criterion.
-The optimization criterion that we focus on in this thesis is the minimum variance distortionless response criterion, which is also known as Capon's method \cite{Capon1969}.
+The optimization criterion that we focus on is the minimum variance distortionless response criterion, which is also known as Capon's method \cite{Capon1969}.
 A beamformer that uses weights derived with Capon's method will in this thesis mainly be referred to as the \textit{Capon beamformer} \cite{Vignon2008}.
@@
 \caption{Examples of array beampatterns where the array weights are either uniform (a) or derived using Capon's method (b).
-An interfering source is located at 20\degree.
+An interfering echo is coming from 20\degree.
 Notice the reduced sidelobe level for the Capon beampattern in the direction of the interfering source.
@@
 Hence, the resulting weight vector minimizes the output power (the criterion) while maintaining unit gain in the steering direction $\vec{a}$ (distortionless response).
-The benefit of Capon beamforming is clearly seen from Fig\,\ref{fig:weights}, which presents a sensitivity map (an \textit{array beampattern}) for DAS beamforming with uniform weights and the Capon beamformer.
-In this simulated case, an interfering source is located at 20\degree{}, which for the DAS beamformer with uniform weights (\ref{fig:das_weights}) will leak into the beamformer output with its power attenuated by around $13$ dB.
-Since the Capon weights are data-dependent, the method "knows" where the interfering source is located and is able to position a zero in its direction (Fig.\,\ref{fig:capon_weights}).
-Hence, the interfering source will be highly attenuated independent of the direction from which it arrives.
-Another interesting case is when the interfering source is located within the mainlobe of the array beampattern.
-In theory a signal will be regarded by the Capon beamformer as an interfering source as long as the signal propagation vector does not match the steering vector ($\vec{a}$) exactly.
+The benefit of Capon beamforming is seen from Fig\,\ref{fig:weights}, which presents a sensitivity map (an \textit{array beampattern}) for DAS beamforming with uniform weights and the Capon beamformer.
+In this simulated case, an interfering echo is coming from 20\degree{}, which for the DAS beamformer with uniform weights (\ref{fig:das_weights}) will leak into the beamformer output with its power attenuated by around $13$ dB.
+Since the Capon weights are data-dependent, the method "knows" where the interfering echo is coming from and is able to position a zero in its direction (Fig.\,\ref{fig:capon_weights}).
+Hence, the interfering echo will be highly attenuated independent of the direction from which it arrives.
+Another interesting case is when the interfering echo has a direction-of-arrival within the mainlobe of the array beampattern.
+In theory, an echo will be regarded by the Capon beamformer as an interferer as long as the signal propagation vector of the echo does not match the steering vector ($\vec{a}$) exactly.
 Positioning of a zero in the direction of an interferer within the mainlobe will lead to a shifted and scaled mainlobe if this is what is needed in order to achieve minimum output power.
-Adaptive positioning of zeros and mainlobe shifting are the source to the super-resolution capabilities of the Capon beamformer \cite{Synnevag2007}, and the positive implications of this effect \cite{Synnevag2009} is the main reason why this beamformer has been highly studied in the field of medical ultrasound imaging over the last decade.
+Adaptive positioning of zeros and mainlobe shifting are the sources of the super-resolution capabilities of the Capon beamformer \cite{Synnevag2007}, and the positive implications of this effect \cite{Synnevag2009} is the main reason why this beamformer has been highly studied in the field of medical ultrasound imaging over the last decade.
 
-To "know" the direction of interfering sources the Capon beamformer has to estimate the statistics of the impinging wave field.
+To "know" the direction of interfering echoes the Capon beamformer has to estimate the statistics of the impinging wave field.
 This is done by estimating a sample covariance matrix, which for an active broadband system can be done in the following way \cite{Synnevag2009}:
@@
 \end{align}
-Which for our example, with $L=M/2=64$, sums up to a required throughput of more than $1/2$ TFLOPS.
-Even though this is just 10\% of the peak throughput of modern GPUs, remember that the latter are only theoretical numbers, and that the actual throughput is highly algorithm dependent.
+For our example, with $L=M/2=64$, this sums up to a required throughput of more than $1/2$ TFLOPS.
+This is more than thousand times that needed for real-time DAS beamforming.
+Even though $1/2$ TFLOPS is just 10\% of the peak throughput of modern GPUs, remember that the latter is only a theoretical number, and that the actual throughput is highly algorithm dependent.
 The required throughput for Capon beamforming might therefore still be hard to achieve in practice \cite[\textbf{Paper II}]{So2011}.
@@
 In addition to the papers in this thesis only Chen \textit{et al.} \cite{Chen2011, Chen, Chen2011a} are known to have explored GPU and FPGA implementations of the Capon beamformer for medical ultrasound imaging.
-The differences between our work and Chen \textit{et al.} is discussed in \textbf{Papers I}, \textbf{II}, and \textbf{VI}.
+The differences between our work and Chen \textit{et al.} are discussed in \textbf{Papers I}, \textbf{II}, and \textbf{VI}.
 
-To reduce the processing requirement of the Capon beamformer for medical ultrasound imaging, several simplified version have been proposed in the literature \cite{Asl2012, Kim}.
-Among the most interesting proposals we find the low complexity adaptive (LCA) beamformer by Synnev\aa{}g \textit{et al.} \cite{Synnevag2011} and the beamspace (BS) Capon beamformer by Nilsen and Hafizovic \cite{Nilsen2009}.
-In medical ultrasound imaging beamspace typically refers to the polar grid that cardiac ultrasound data are located in before it is scan converted into a Cartesian grid.
+To reduce the processing requirement of the Capon beamformer for medical ultrasound imaging, several simplified versions have been proposed in the literature \cite{Asl2012, Kim}.
+Among the most interesting proposals we find the low complexity adaptive (LCA) beamformer by Synnev\aa{}g \textit{et al.} \cite{Synnevag2011},  and the beamspace (BS) Capon beamformer by Nilsen and Hafizovic \cite{Nilsen2009}.
+The LCA beamformer uses a predefined set of weights, and the BS-Capon beamformer takes advantage of the narrow transmit and receive beams used in medical ultrasound imaging.
+Beamspace typically refers to the polar grid that cardiac ultrasound data are located in before it is scan converted into a Cartesian grid.
 In combination with the Capon beamformer, beamspace refers to the wavenumber-space representation of the recorded signals, i.e., the spatial Fourier transform of the element data:
@@
 If one looks closer at the windows selected for the LCA beamformer we see that half the set performs micro-steering of the mainlobe and the other half provides different mainlobe widths.
-The results from applying the LCA beamformer show that the main benefit of applying Capon beamforming to medical ultrasound imaging is micro-steering of the mainlobe, and not attenuation of far off-axis backscatter.
+The results from applying the LCA beamformer \cite[Fig.\,9]{Synnevag2011} and the success of the BS-Capon beamformer shows that the main benefit of applying Capon beamforming to medical ultrasound imaging is micro-steering of the mainlobe, and not attenuation of far off-axis backscatter.
 
@@
 Medical ultrasound imaging is, however, a broadband imaging modality and one could think that a broadband Capon beamformer would therefore be needed \cite{Holfort2009}.
-The investigation performed in \cite{Nilsen2009} showed, that with pre-delayed data, energy will be centered on-axis and as little as three near-axis beams is therefore adequate in order to achieve similar performance for BS-Capon and Capon beamforming.
+The investigation performed in \cite{Nilsen2009} showed, that with pre-delayed data, energy will be centered on-axis and as little as three near-axis beams in beamspace is therefore adequate in order to achieve similar performance for BS-Capon and Capon beamforming.
 The two-way beam will be wider for unfocused (plane wave) transmit beams \cite{Holfort2009, Austeng2011}, but steering and focusing on receive will still make the main amount of energy arrive on axis.
 As shown in \textbf{Paper III} it is also possible to micro-steer with phase delays (narrowband) instead of time delays (broadband) without introducing large errors.
-Since Capon beamforming mainly performs main-lobe shifts smaller than the beam width, the difference between a broadband and narrowband Capon beamformer will be minimal \cite{Diamantis2014}.
+Since Capon beamforming mainly performs main-lobe shifts smaller than the beam width \cite{Synnevag2011, Nilsen2009}, the difference between a broadband and narrowband Capon beamformer will be minimal \cite{Diamantis2014}.
 Broadband Capon beamforming is also significantly more computationally demanding than narrowband Capon beamforming.
@@
 \caption{The principle behind volume ray casting.
-a) A ray is casted out from every pixel in the output image.
+a) A ray is casted out from every pixel in a 2D image.
 Along the ray, samples are gathered from the volume.
-The samples are assigned a color and an opacity before they are combined into one pixel value (wikipedia.org/wiki/Volume\_ray\_casting).
+The samples are assigned a color and an opacity before they are combined into a single pixel value (wikipedia.org/wiki/Volume\_ray\_casting).
 b) The opacity transfer function used to determine how opaque a sample should be.}
@@
 Another way is to project the whole volume into a two-dimensional image.
-The latter technique is known as volume rendering, and it presents to the user the three dimensional structure of the heart in one two-dimensional image.
+The latter technique is known as volume rendering, and it presents to the user the three dimensional structure of the heart in a single two-dimensional image.
 Many different methods for volume rendering exist \cite{brodlie2001}, but in this thesis we will focus on ray casting \cite{Levoy1988}.
@@
 The principle behind ray casting is depicted in Fig.\,\ref{fig:vr}.
-The projection of volumetric data is performed by casting a ray out from every pixel in the output image (Fig.\,\ref{fig:vr_rc}).
+The projection of volumetric data is performed by casting a ray out from every pixel in the 2D image (Fig.\,\ref{fig:vr_rc}).
 Along each ray, samples are gathered from the volume and assigned a color and opacity.

### 2014-05-31 ThesisJP/Introduction/introduction.tex
Finished all comments from Sverre on the english text.

@@
 \section{Motivation}
-A rapid development in computer game technology and accompanying programming languages have recently provided researchers with small personal supercomputers, comprised in a single graphics card (\nom{GPU}{Graphics processing unit}).
-The latest graphics cards\footnote{As of January 2014.} from both \nom{Nvidia}{Graphics card vendor} and \nom{AMD}{Graphics card vendor} provide more than 5 T\nom{FLOPS}{Floating point operations per second} single precision and 1.5 TFLOPS double precision computations.
+The rapid development of computer game technology and accompanying programming languages, as shown in Chapter \ref{chap1}, have recently provided researchers with small personal supercomputers, comprised in a single graphics processing unit (\nom{GPU}{Graphics processing unit}).
+%The latest graphics cards\footnote{As of January 2014.} from both \nom{Nvidia}{Graphics card vendor} and \nom{AMD}{Graphics card vendor} provide more than 5 T\nom{FLOPS}{Floating point operations per second} single precision and 1.5 TFLOPS double precision computations.
 For double precision, this is more than the world's most powerful supercomputer could provide in 1998, and that was a room full of computers.
-
 This immense rise in computational capabilities and improved programmability are currently changing how ultrasound imaging systems are designed.
@@
 The Capon beamformer in ultrasound imaging is a good example of an algorithm that has been widely studied over the last decade where the number of calculations was thought to be too high for real-time processing to happen in near future.
-To achieve real-time processing for e.g.
-cardiac ultrasound imaging, an effective processing rate of hundreds of GFLOPS is required.
+As shown in Section \ref{sec:adaptbf}, an effective processing rate of hundreds of GFLOPS is required to achieve real-time processing for cardiac ultrasound imaging.
 This number, as explained in Section \ref{sec:adaptbf}, also grows with larger subarrays and an increasing number of samples per second.
-Nevertheless, with modern GPUs, this amount of processing is finally available within a single card.
+Nevertheless, with modern GPUs, this amount of processing is finally available within a single GPU card.
 When researchers are exploring new algorithms for ultrasound imaging, it is important to keep the architecture of parallel accelerators in mind.
@@
 
-The main focus has been the Capon beamformer, and the problem of making this computationally intense algorithm available for real-time ultrasound imaging [\textbf{Paper\,I} and \textbf{II}].
+The main focus has been the Capon beamformer, and the problem of making this computationally intense algorithm available for real-time ultrasound imaging [\textbf{Papers\,I} and \textbf{II}].
 However, two additional methods have also been explored.
@@
 This makes it somewhat easier to reach the goal of real-time processing.
+The transmit beams used for sonar imaging are also much wider than the once used for ultrasound imaging.
+This makes techniques like the BS-Capon beamformer less attractive for sonar imaging.
+However, as shown in this paper, low-complexity methods are not needed to reach real time Capon beamforming for sonar imaging.
 
@@
 The second paper aims at implementing real-time Capon beamforming for cardiac ultrasound imaging.
-Achieving this will facilitate further study of the method's \textit{in vivo} performance.
+Achieving this will facilitate further study of the method's \textit{in vivo} performance in the future.
 This is something that has been suggested as further work in many publications on Capon beamforming for medical ultrasound imaging over the past decade.
@@
 The reported processing throughput is able to keep up with the acquisition frame rate in a typical cardiac ultrasound imaging system equipped with 64- and 96-element linear arrays.
-\textbf{Paper\,II} and \textbf{VI} also present, for the first time, videos where the ES-Capon and BS-Capon beamformer have been applied on loops of simulated and \textit{in vivo} medical ultrasound images.
+\textbf{Papers\,II} and \textbf{VI} also present, for the first time, videos where the ES-Capon and BS-Capon beamformer have been applied on loops of simulated and \textit{in vivo} medical ultrasound images.
 
@@
 Further, it has been investigated how the Capon beamforming achieves its super-resolution, and how this makes it highly sensitive to differences between the assumed steering vector and the signal propagation vector.
-Earlier work, focusing on single images, has typically referred to this effect as speed-of-sound errors, however, it also comes naturally into play when objects moves across an imaging sector which has insufficient lateral sampling.
+Earlier work, focusing on single images, has typically referred to this effect as speed-of-sound errors, however, it also comes naturally into play when objects move across an imaging sector which has insufficient lateral sampling.
 With a narrowband and farfield model it is shown that the Capon beamformer suffers from lateral gain variations as large as 27 dB for parameters previously used to obtain high resolution medical ultrasound images.
@@
 Noise located in the blood pool tends to occlude cardiac tissue when rendered in 3D, and is often impossible to remove by adjusting global rendering parameters.
-This because ultrasound signals have high variability within blood, as well as high variability within tissue, even when heavily smoothed.
+This is because ultrasound signals have high variability within blood, as well as high variability within tissue, even when heavily smoothed.
 Delineation of the blood pool is done by a state estimation algorithm, capable of tracking the endocardium in real time.
@@
 The work in \textbf{Paper\,II} made it clear that interesting findings are done when a real-time implementation makes it possible to watch the temporal behavior of an algorithm.
-An observation of temporal artifacts encourage for the work of \textbf{Paper\,III}.
+An observation of temporal artifacts was the fact that stimulated for the work of \textbf{Paper\,III}.
 

### 2014-05-31 ThesisJP/Thesis.tex
Finished all comments from Sverre on the english text.
Finished norwegian abstract.

@@
 \newpage
+
+%\begin{otherlanguage}{norwegian}
+
 %\thispagestyle{empty}%
-~\\%
-~\\%
-~\\
+%~\\%
+%~\\%
+%~\\
 \noindent\large 
-\textbf{Akselerering av adaptive algoritmer for ultralydavbildning ved hjelp av programmerbare grafikkprosessorer}
+\textbf{Akselerering av adaptive algoritmer for ultralydavbildning ved hjelp av programmerbare skjermkort}
 %\vspace{0.2cm}\\%
@@
 \normalsize
-En rivende utvikling innen spillteknologi og tilh\o{}rende programmeringsspr\aa{}k har nylig gjort superdatamaskiner i form av et lite programmerbart grafikkort (GPU) tilgjengelige for forskere.
-Denne utrolige \o{}kningen i beregningskraft og forbedret programmerbarhet er for \o{}yeblikket med p\aa{} \aa{} endre hvordan ultralydsystemer designes.
-N\aa{}r forskere utforsker nye algoritmer for ultralydavbildning er det derfor viktig at arkitekturen til parallelle akseleratorer, som en GPU er et eksempel p\aa{}, tas h\o{}yde for.
-Hvis en ny, kompleks algoritme er tenkt til \aa{} kj\o{}res i sanntid, er det viktig at den kan tilpasses den programmerbare og parallelle prosesseringskjeden i moderne ultralydskannere.
+De siste årene har det vært en rivende utvikling innen bruken av skjermkort (GPU-er), som egentlig er laget for å kjøre dataspill, til å gjøre numeriske beregninger for naturvitenskapelige anvendelser.
+Et sistegenerasjons skjermkort har nå like mye beregningskraft som verdens kraftigste datamaskin hadde rundt år 2000.
+Utfordringen er at problemet som skal løses må bestå av mange uavhengige deler som kan beregnes i parallell.
+For ultralydmaskiner har denne utviklingen gjort at algoritmer som tideligere ble trykket i hardware, nå kan programmeres i software og kjøres i sanntid på et rimelig skjermkort.
+Et ultralydsystem som bruker skjermkort til å gjøre beregninger vil være mye rimeligere, mer fleksibelt og gjøre det lettere å implementere avansert prosessering enn et system hvor alle algoritmer er og må "hugges i stein".
 
-M\aa{}let med denne studien har v\ae{}rt \aa{} utforske hvordan GPUer kan utnyttes til avansert prosessering i et avbildende ultralydsystem.
-Blant de problemene vi har sett n\ae{}rmere p\aa{} finner man b\aa{}de adaptiv str\aa{}leforming, adaptiv visualisering av volumetrisk ultralyd og ultralydsimuleringer.
-Felles for disse problemene er at de alle krever parallelprogrammering for at sanntidsprosessering skal oppn\aa{}s.
+Målet med denne studien har vært å utforske hvordan programmerbare skjermkort kan utnyttes til avansert prosessering i et avbildende ultralydsystem.
+Blant de problemene vi har sett nærmere på finner man både adaptiv bildedanelse (stråleforming), adaptiv visualisering av volumetrisk ultralyd og ultralydsimuleringer.
+Felles for disse problemene er at de krever parallellprogrammering for at sanntidsprosessering skal oppnås.
 
-I f\o{}rste del av avhandlingen utforskes Capon's adaptive str\aa{}leformings-metode.
-Metoden implementeres p\aa{} en GPU  for \aa{} muliggj\o{}re sanntidsprosessering for avbildende sonar (\textbf{Artikkel\,I}) og medisinsk ultralydavbildning (\textbf{Artikkel\,II}).
-Sanntidsprosessering oppn\aa{}s for begge modaliteter.
-\textbf{Artikkel\,II} presenterer ogs\aa{}, for f\o{}rste gang, videoer hvor b\aa{}de simulerte og \textit{in vivo} medisinsk ultralyddata er prosessert med Capon-str\aa{}leforming.
-I \textbf{Artikkel\,III} viser vi at Capon-str\aa{}leformeren ikke har skift-invariant oppf\o{}rsel n\aa{}r den anvendes til sanntidsavbildning.
-Videre foresl\aa{}s det en metode som gir mer shift-invariant avbildning.
-Skift-invariant avbildning er avgj\o{}rende for at metoden skal kunne tas ibruk i praktisk.
+I første del av avhandlingen utforskes Capons adaptive stråleformingsmetode.
+Vi starter med å implementere metoden på et skjermkort for sanntidsprosessering av sonar (\textbf{Artikkel\,I}) og medisinsk ultralyd (\textbf{Artikkel\,II}).
+Sanntidsprosessering oppnås for begge modaliteter.
+\textbf{Artikkel\,II} presenterer også, for første gang, videoer hvor både simulerte og \textit{in vivo} medisinsk ultralyddata er prosessert med Capon-stråleforming.
+I \textbf{Artikkel\,III} viser vi at Capon-stråleformeren er følsom for små skift, enten ved at ultralydproben eller objektet man avbilder beveger seg.
+Videre foreslås det en metode som gjør at stråleformeren ikke lenger er like følsom for slike skift.
+Dette er viktig å få på plass for at Capon-stråleformeren skal kunne tas i bruk i praksis.
 
-I \textbf{Artikkel\,IV} foresl\aa{}r vi en adaptiv metode for visualisering av volumetriske ultralydbilder av hjertet.
-Metoden undertrykker st\o{}y som med konvensjonelle metoder hadde skygget for hjertevev.
-Dette arbeidet viser ogs\aa{} at det med moderne GPUer er mulig \aa{} utf\o{}re avansert visualisering i et ultralydsystem og fortsatt ha sanntidsytelse.
+I \textbf{Artikkel\,IV} foreslår vi en adaptiv metode for visualisering av volumetriske ultralydbilder av hjertet.
+Metoden undertrykker støy som med konvensjonelle metoder hadde skygget for hjertevev.
+Dette arbeidet viser at det med moderne skjermkort er mulig å utføre avansert visualisering i et ultralydsystem og fortsatt ha sanntidsytelse.
 
-I siste del av avhandlingen utforsker vi hvordan GPUer kan utnyttes til \aa{} akselerere ultralydsimuleringer.
-Resultatet av dette arbeidet ble et simuleringsprogram hvor geometrien til et gitt ultralydarray kan interaktivt tegnes, og hvor det resulterende trykkfeltet simuleres i sanntid.
+I siste del av avhandlingen utforsker vi hvordan skjermkort kan utnyttes til å akselerere ultralydsimuleringer.
+Resultatet av dette arbeidet ble et simuleringsprogram hvor geometrien til et gitt ultralydarray kan tegnes interaktivt, og hvor det resulterende trykkfeltet simuleres og visualiseres i sanntid.
 
@@
 \noindent \emph{Ovennevnte avhandling er funnet verdig til \aa{} forsvares offentlig for graden Philosophiae Doctor (Ph.D.) i medisinsk teknologi. Disputas finner sted ... .}
+
+%\end{otherlanguage}
 
@@
 \begin{bibunit}[IEEEtran]
-\CustChapter{Background}
+\CustChapter{Background}\label{chap1}
 \runningtitle{Background}
 \input{./Background/background}
-\newpage\addcontentsline{toc}{section}{References}
+
+\CustChapter{Introduction to Papers} 
+\runningtitle{Introduction to Papers}
+\input{./Introduction/introduction}
+
+\newpage\addcontentsline{toc}{chapter}{References} 
 \putbib[./Background/background]
@@
 %%%% CHAPTER 1 %%%%
-\CustChapter{Introduction} 
-\runningtitle{Introduction}
-\input{./Introduction/introduction}
 

### 2014-06-05 ThesisJP/Background/background.tex
Final?

@@
 This is the same technique as reported implemented on a GPU by Yiu \textit{et al.} \cite{Yiu2011}.
+Software beamforming is also available in the Vantage research ultrasound system by Verasonics\footnote{http://www.verasonics.com/25}.
 We can therefore conclude that beamforming is finally moving into software too.
@@
 
-\subsection{Narrowband or broadband Capon beamforming}
+\subsection{Capon beamforming of focused broadband beams}
 Both the LCA and BS-Capon beamformers are based upon a detailed investigation of how Capon beamfoming behaves when applied to medical ultrasound images.
@@
 The results from applying the LCA beamformer \cite[Fig.\,9]{Synnevag2011} and the success of the BS-Capon beamformer shows that the main benefit of applying Capon beamforming to medical ultrasound imaging is micro-steering of the mainlobe, and not attenuation of far off-axis backscatter.
+This is mainly due to the focused beams used in medical ultrasound which highly limits the amount of energy arriving off axis.
+This focusing of energy also explain why Capon beamforming works on broadband ultrasound data.
 
@@
 Medical ultrasound imaging is, however, a broadband imaging modality and one could think that a broadband Capon beamformer would therefore be needed \cite{Holfort2009}.
-The investigation performed in \cite{Nilsen2009} showed, that with pre-delayed data, energy will be centered on-axis and as little as three near-axis beams in beamspace is therefore adequate in order to achieve similar performance for BS-Capon and Capon beamforming.
+The investigation performed for the BS-Capon beamformer \cite{Nilsen2009} showed, that with pre-delayed data, energy will be centered on-axis and as little as three near-axis beams in beamspace is therefore adequate in order to achieve similar performance for BS-Capon and Capon beamforming.
 The two-way beam will be wider for unfocused (plane wave) transmit beams \cite{Holfort2009, Austeng2011}, but steering and focusing on receive will still make the main amount of energy arrive on axis.

### 2014-06-06 ThesisJP/Thesis.tex
Corrected norwegian abstract based on comments from Sverre.

@@
 \noindent\large 
-\textbf{Akselerering av adaptive algoritmer for ultralydavbildning ved hjelp av programmerbare skjermkort}
+\textbf{Hurtig eksekvering av adaptive algoritmer for ultralydavbildning ved hjelp av programmerbare skjermkort}
 %\vspace{0.2cm}\\%
@@
 Utfordringen er at problemet som skal løses må bestå av mange uavhengige deler som kan beregnes i parallell.
-For ultralydmaskiner har denne utviklingen gjort at algoritmer som tideligere ble trykket i hardware, nå kan programmeres i software og kjøres i sanntid på et rimelig skjermkort.
+For ultralydmaskiner har denne utviklingen gjort at algoritmer som tideligere krevde spesialhardware, nå kan programmeres i software og kjøres i sanntid på et rimelig skjermkort.
 Et ultralydsystem som bruker skjermkort til å gjøre beregninger vil være mye rimeligere, mer fleksibelt og gjøre det lettere å implementere avansert prosessering enn et system hvor alle algoritmer er og må "hugges i stein".
@@
 Målet med denne studien har vært å utforske hvordan programmerbare skjermkort kan utnyttes til avansert prosessering i et avbildende ultralydsystem.
-Blant de problemene vi har sett nærmere på finner man både adaptiv bildedanelse (stråleforming), adaptiv visualisering av volumetrisk ultralyd og ultralydsimuleringer.
-Felles for disse problemene er at de krever parallellprogrammering for at sanntidsprosessering skal oppnås.
+Blant de problemene vi har sett nærmere på finner man både adaptiv bildedannelse (stråleforming), adaptiv visualisering av 3D ultralyd og ultralydsimuleringer.
+Felles for disse problemene er at de krever parallellprogrammering for å gå i sann tid.
 
 I første del av avhandlingen utforskes Capons adaptive stråleformingsmetode.
-Vi starter med å implementere metoden på et skjermkort for sanntidsprosessering av sonar (\textbf{Artikkel\,I}) og medisinsk ultralyd (\textbf{Artikkel\,II}).
+Vi starter med å implementere metoden på et skjermkort for sanntidsprosessering av sonardata (\textbf{Artikkel\,I}) og av data fra medisinsk ultralyd (\textbf{Artikkel\,II}).
 Sanntidsprosessering oppnås for begge modaliteter.
@@
 I \textbf{Artikkel\,III} viser vi at Capon-stråleformeren er følsom for små skift, enten ved at ultralydproben eller objektet man avbilder beveger seg.
-Videre foreslås det en metode som gjør at stråleformeren ikke lenger er like følsom for slike skift.
+Videre foreslås det en metode som gjør at stråleformeren ikke lenger er like følsom for slike forflytninger.
 Dette er viktig å få på plass for at Capon-stråleformeren skal kunne tas i bruk i praksis.
 
-I \textbf{Artikkel\,IV} foreslår vi en adaptiv metode for visualisering av volumetriske ultralydbilder av hjertet.
+I \textbf{Artikkel\,IV} foreslår vi en adaptiv metode for visualisering av 3D ultralydbilder av hjertet.
 Metoden undertrykker støy som med konvensjonelle metoder hadde skygget for hjertevev.

### 2014-06-19 ThesisJP/Introduction/introduction.tex
Final changes.

@@
 J.\:I.\:Buskenes, \textbf{J.\:P.\:\AA{}sen}, C.-I.\:C.\:Nilsen and A.\:Austeng\\
-{\it IEEE Transactions on Oceanic Engineering, accepted for publication.}\\\\
+{\it IEEE Transactions on Oceanic Engineering, Jul 2014.}\\\\
 The first paper describes in detail how the Capon beamformer is mapped to the GPU architecture.
@@
 \textbf{J.\:P.\:\AA{}sen}, A.\:Austeng and S.\:Holm\\
-{\it IEEE Transactions on Ultrasonics, Ferroelectrics, and Frequency Control, accepted for publication.}\\\\
+{\it IEEE Transactions on Ultrasonics, Ferroelectrics, and Frequency Control, , vol. 61, no. 7, pp. 1152-1160, Jul 2014.}\\\\
 In the third paper we study the shift-invariant property of an imaging system based on the Capon beamformer.
@@
 
-Achieving higher processing speed for an excising algorithm by implementing it on a new architecture is not academically relevant in its own.
+Achieving higher processing speed for an existing algorithm by implementing it on a new architecture is not academically relevant in its own.
 A good analysis has to be given of how the algorithm was modified for the architecture, or how to utilize the gained speed.
@@
 In \textbf{Paper II} and \textbf{III} there will be references to several videos that can be accessed online\footnote{folk.uio.no/jpaasen/thesis}.
+There is one folder for each paper having multimedia content.
 

### 2014-07-04 Simulator/article.tex
Added UAC2014 talk

@@
 \begin{itemize}
+\item SWAT (Shallow Water Acoustic Toolset): 
+\item Bell's thesis:
+\item SIGMA:
 \item OpenGL does not support scattered writes. OpenCL does.
+\item 2D template approximation:
+\item Geometric distortions in SSS \cite{Cobra1992}
+\item Sidescan sonar and synthetic aperture sonar. In seafloor coordinates. Maps to orthogonal rendering.
+\item Layover causes ambiguity, different seafloor phenomena can cause the same output image.
+Lots of cornercases when approximating the model in 2D.
+Instead use a 3D model and project it to a 2D one.
+\item Viewer can be used to train an operator for finding non-trivial objects on the seafloor. Input for automated target recognition standard.
+\item OpenGL rendering can have infinite resolution, only limited by the pixel density in the image.
+\end{itemize}
+
+What's new?
+\begin{itemize}
+\item Tuned for speed using GPUs, combining OpenGL and OpenCL to take advantage of existing functionality for computer graphics with the added flexibility of general purpose programming on the GPUs.
+\item Performance to spare: Game-like viewer for the visualization and verification of the rendered scene and corresponding sonar image.
+\item First simulator of its kinds aimed towards improved ATR and ultimately better AUVs.
 \end{itemize}
@@
 
+\begin{itemize}
+\item Motion compensation
+\item 
+\end{itemize}
+
 \appendix

### 2014-07-04 ThesisJP/Background/background.tex
Corrected norwegian abstract based on comments from Sverre.

@@
 \subsection{Beamforming}
-If time delays are applied to the signal emitted by the elements it is possible to steer and focus the ultrasound beam in space (Fig.\,\ref{fig:das_background}).
+If time delays are applied to the signal emitted by the array elements it is possible to steer and focus the ultrasound beam in space (Fig.\,\ref{fig:das_background}).
 This technique is known as \textit{beamforming}, and for ultrasound imaging it is applied both when transmitting and receiving ultrasound pulses.
@@
 \end{align}
-Here, \nom{$M$}{Number of array elements} is the number of elements in the array, $x_m$ is the per-element data, \nom{$\Delta_m$}{Per-element focusing and steering delays} is the per-element focusing and steering delay, \nom{$\vec{w}$}{Array weight vector (apodization)} is the array \textit{weight vector} or \textit{apodization}, $\vec{x}$ is a pre-delayed vector of element data, and $z[n]$ is the output image line in the direction given by $\{\Delta_m[n]\}$.
+Here, \nom{$M$}{Number of array elements} is the number of elements in the array, $x_m$ is the per-element data, \nom{$\Delta_m$}{Per-element focusing and steering delays} is the per-element focusing and steering delay, \nom{$\vec{w}$}{Array weight vector (apodization)} is the array \textit{weight vector} or \textit{apodization}, $\vec{x}$ is a vector of pre-delayed element data, and $z[n]$ is the output image line in the direction given by $\{\Delta_m[n]\}$.
 From now on we will omit the dependency on $\{\Delta_m[n]\}$ and $x_m$ will refer to delayed element data.
@@
 \end{align}
-\nomi{$N$}{Number of samples in range (or time)}where $f_c$ is the center frequency of the transmitted pulse, and where \nom{$B_w$}{Relative bandwidth} is the relative bandwidth in percent of \nom{$f_c$}{Center frequency of ultrasound pulse}.
+\nomi{$N$}{Number of samples in range (or time)}where \nom{$f_c$}{Center frequency of ultrasound pulse} is the center frequency of the transmitted pulse, and where \nom{$B_w$}{Relative bandwidth} is the relative bandwidth in percent.
 Note that after this process the ultrasound data will be complex valued.
@@
 An interfering echo is coming from 20\degree.
-Notice the reduced sidelobe level for the Capon beampattern in the direction of the interfering source.
+Notice the reduced sidelobe level for the Capon beampattern in the direction of the interferer.
 Image courtesy of Jo Inge Buskenes.}
@@
 \end{align}
-For our example, with $L=M/2=64$, this sums up to a required throughput of more than $1/2$ TFLOPS.
-This is more than thousand times that needed for real-time DAS beamforming.
-Even though $1/2$ TFLOPS is just 10\% of the peak throughput of modern GPUs, remember that the latter is only a theoretical number, and that the actual throughput is highly algorithm dependent.
+For our example, with $L=M/2=64$, this sums up to a required throughput of more than $1/2$ TFLOPS, which is more than thousand times that needed for real-time DAS beamforming.
+On the other hand, $1/2$ TFLOPS is just 10\% of the peak throughput of modern GPUs, but remember that the latter is only a theoretical number, and that the actual throughput is highly algorithm dependent.
 The required throughput for Capon beamforming might therefore still be hard to achieve in practice \cite[\textbf{Paper II}]{So2011}.
@@
 This is mainly due to the focused beams used in medical ultrasound which highly limits the amount of energy arriving off axis.
-This focusing of energy also explain why Capon beamforming works on broadband ultrasound data.
+This focusing of energy also explains why Capon beamforming works on broadband ultrasound data.
 

### 2014-07-04 ThesisJP/Thesis.tex
Corrected norwegian title.

@@
 \noindent\large 
-\textbf{Hurtig eksekvering av adaptive algoritmer for ultralydavbildning ved hjelp av programmerbare skjermkort}
+\textbf{Hurtig beregning av adaptive algoritmer for ultralydavbildning ved hjelp av programmerbare skjermkort}
 %\vspace{0.2cm}\\%
@@
 \renewcommand\labelenumi{\bfseries\theenumi}
-	\item J.\:I.\:Buskenes, \textbf{J.\:P.\:\AA{}sen}, C.-I.\:C.\:Nilsen and A.\:Austeng, "An optimized GPU implementation of the MVDR beamformer for active sonar imaging", {\it  IEEE Transactions on Oceanic Engineering, accepted for publication}.
+	\item J.\:I.\:Buskenes, \textbf{J.\:P.\:\AA{}sen}, C.-I.\:C.\:Nilsen and A.\:Austeng, "An optimized GPU implementation of the MVDR beamformer for active sonar imaging", {\it  IEEE Transactions on Oceanic Engineering, Jul 2014}.
 	\item \textbf{J.\:P.\:\AA{}sen}, J.\:I.\:Buskenes, C.-I.\:C.\:Nilsen, A.\:Austeng and S.\:Holm, "Implementing Capon beamforming on a GPU for real-time cardiac ultrasound imaging", {\it IEEE Transactions on Ultrasonics, Ferroelectrics, and Frequency Control, vol.
@@
 76-85, Jan 2014}.
- 	\item \textbf{J.\:P.\:\AA{}sen}, A.\:Austeng and S.\:Holm, "Capon beamforming and moving objects - An analysis of lateral shift-invariance", {\it IEEE Transactions on Ultrasonics, Ferroelectrics, and Frequency Control, accepted for publication}.
+ 	\item \textbf{J.\:P.\:\AA{}sen}, A.\:Austeng and S.\:Holm, "Capon beamforming and moving objects - An analysis of lateral shift-invariance", {\it IEEE Transactions on Ultrasonics, Ferroelectrics, and Frequency Control, vol.
+61, no.
+7, pp.
+1152-1160, Jul 2014}.
 	\item \textbf{J.\:P.\:\AA{}sen}, E.\:Steen, G.\:Kiss, A.\:Thorstensen and S.\:I.\:Rabben, "Adaptive volume rendering of cardiac 3D ultrasound images - utilizing blood pool statistics", {\it Proc.
@@
 }
-\noindent \textit{IEEE Transactions on Oceanic Engineering, accepted for publication}.
+\noindent \textit{IEEE Transactions on Oceanic Engineering, Jun 2014}.
 \newpage
@@
 }
-\noindent \textit{IEEE Transactions on Ultrasonics, Ferroelectrics, and Frequency Control, accepted for publication}.
+\noindent \textit{IEEE Transactions on Ultrasonics, Ferroelectrics, and Frequency Control, , vol. 61, no. 7, pp. 1152-1160, Jul 2014}.
 \newpage\input{./Paper3/moving_obj}

### 2014-07-18 Simulator/article.tex
Added UA2014 ship thumbnail

@@
 The principle involved is to isolate an image segment containing the object of interest, correlate it with a set of template images of the relevant object classes, and assign it to the class of the template yielding the highest correlation coefficient.
-The challenge is to come up with a representative set of template images covering the actual configurations of object and seabed.
-The alternative of using of a static, predetermined template library has a computational complexity that is exponential with the number of parameters.
+The challenge is to come up with a representative set of template images covering the relevant configurations of object and seabed.
+The alternative of using a static, predetermined template library has a computational complexity that is exponential with the number of parameters.
 This limits its use to only a few parameters being coarsely sampled, ultimately yielding inaccurate results~\cite{Midelfart2010}.
 
-We have developed a synthetic aperture sonar (SAS) simulator that avoids this problem by running sufficiently fast to create templates adapted to the actual scene as part of the classification process.
-This is achieved with the aid of the massive computing power in graphics processing units (GPUs) and optimized software libraries for scene rendering.
-The simulator loads a 3D model of the seafloor and an object class model into OpenGL, where emitted sound waves are modeled with a light source placed at the sonar transmit location.
+To avoid this problem we have developed a very fast side-looking sonar simulator that generates templates adapted to the actual scene.
+
+synthetic aperture sonar (SAS)
+
+ as part of the classification process. This is achieved with the aid of the massive computing power in graphics processing units (GPUs) and optimized software libraries for scene rendering.
+
+ The simulator loads a 3D model of the seafloor and an object class model into OpenGL, where emitted sound waves are modeled with a cylindrical light source placed at the sonar transmit location.
+
 We assume rough, isotropic surfaces reflecting sound energy equally in all directions.
@@
 However, SIGMAS+ is still aimed more towards realistically looking images, including effects like ambient noise, which is irrelevant for our purpose of template generation.
-Also, their simulator sums images obtained from multiple OpenGL rendering passes to create the sonar image, while we render once and post-process this result with OpenCL to tailor the sonar image with more flexibility.
+Also, their simulator superimpose images obtained from multiple OpenGL rendering passes to create the sonar image, while we render once and post-process this result with OpenCL to tailor the sonar image with greater flexibility.
 
@@
 \item 2D template approximation:
-\item Geometric distortions in SSS \cite{Cobra1992}
+\item Geometric distortions in SSS \cite{Cobra1992} Bell
 \item Sidescan sonar and synthetic aperture sonar. In seafloor coordinates. Maps to orthogonal rendering.
@@
 Instead use a 3D model and project it to a 2D one.
+Easier to operate, models are more general purpose.
+Less parameters to adjust.
 \item Viewer can be used to train an operator for finding non-trivial objects on the seafloor. Input for automated target recognition standard.
+\item Improved image quality of selected object, reduced data collection cost (price estimate?).
 \item OpenGL rendering can have infinite resolution, only limited by the pixel density in the image.
+\item Accuracy in terms of classification result. Need to embody static effects in a deterministic way. Incoherent synthesis of the image.
+\item Sound speed assumed constant.
+\item High frequency lead to little (define) penetration, lower range and higher resolution. Often modeled accurately with ray tracing techniques.
+\end{itemize}
+
+Model
+\begin{itemize}
+\item Ray solution: Separate amplitude and phase components of the Helmholtz eq.
+by applying geometric optics approximation.
+Ok as long as amplitude varies more slowly with position than the phase.
+Implications: High frequency, because the curvature of a ray over a wavelength must be small, and the fractional change in the sound speed must be small over a wavelength.
+Further simplification available by Snell's Law.
+Fast to compute.
+Not able to handle diffraction, focal points and caustics, and will cause (too) abrupt changes on distinct changes in sea environment.
+Can be improved upon with fuzzy beam modeling / Gaussian beam tracing.
+\begin{itemize}
+\item Pro: Easy to compute. Visual result. Handles complex bathymetry. Array and model response can be accounted for.
+\item Con: Only works for high frequency and small sound variations. No diffraction, caustics and focus. Infinite energy caustics/sharp variations. Incoherent(?)
+\end{itemize}
+\item Normal modes: Separate amplitude and phase into a range dependent term, and a depth dependent term.
+Assumes minimal range dependence in sound speed profile and depth.
+Limited near field accuracy.
+Acoustic field does not need to be calculated at intermediate ranges.
+Penetration can be modeled.
+can be calculated
+\begin{itemize}
+\item Pro: Ok to compute. Handles penetration. Low frequency, shallow water.
+\item Con: Horizontal stratification. Far field only. 
+\end{itemize}
+\item Parabolic equation: One way propagation.
+Marching solution.
+Initial starting solution must be known, often computed with normal mode.
+Only valid for narrow beams $\pm$20.
+Able to handle range dependent environments.
+Computing full field is easy.
+Computation time is of O(f$^2$), hence mostly useful for low-frequency scenarios.
+Unable to deal with horizontal refraction (shadow zones from e.g.
+conical mounts)
+\item Fast field programs / Green's function solutions.
+Exact full wave solution for acoustic fields in horizontally stratified media.
+Can handle both compressional and shear waves, and hence both fluid and solid media.
+Far field assumption.
+Similar to normal mode techniques.
+\item Finite element: Scene segmented into triangles or rectangles with a size of one tenth of a wavelength, or less.
+At each node the wavefield can be determined from the wave equation.
+Computationally limited to low frequencies.
+\item 2D model: Acoustic field in range and depth.
+Assumes that a wave transmitted in a vertical plane remains in it.
+Reasonable when the sound speed can be assumed constant in the horizontal plane, and when the seafloor normals are in the vertical range-depth plane (avoid horizontal refraction, or actually reflection).
 \end{itemize}
@@
 \item First simulator of its kinds aimed towards improved ATR and ultimately better AUVs.
+\item 3D models are still novel. 
 \end{itemize}
@@
 \item Improved inertial navigation?
+\item Aid in understanding the key parts of the SAS image and in discovering potential flaws in the SAS image. Anomalies can be flagged to make SAS more robust?
 \end{itemize}

### 2014-08-22 Simulator/article.tex
...

--- 2014-07-04 Simulator/article.tex
@@
 One way to classify these objects is by using template matching.
-The principle involved  is to isolate an image segment containing the object of interest, correlate it with a set of template images of the relevant object classes, and assign it to the class of the template yielding the highest correlation coefficient.
-The challenge is to come up with a representative set of template images covering the actual configurations of object and seabed.
+The principle involved  is to isolate an image segment containing the object of interest, compare it with a set of template images of the relevant object classes, and assign it to the class of the template with the the best fit.
+
+The challenge is to come up with accurate template images covering the actual configurations of object and seabed.
+
+Estimate 2D template directly
+  ->
+
 The alternative of using of a static, predetermined template library has a computational complexity that is exponential with the number of parameters.

### 2014-08-22 Simulator/article.tex

--- 2014-07-18 Simulator/article.tex
@@
 One way to classify these objects is by using template matching.
-The principle involved is to isolate an image segment containing the object of interest, correlate it with a set of template images of the relevant object classes, and assign it to the class of the template yielding the highest correlation coefficient.
-The challenge is to come up with a representative set of template images covering the relevant configurations of object and seabed.
-The alternative of using a static, predetermined template library has a computational complexity that is exponential with the number of parameters.
+The principle involved  is to isolate an image segment containing the object of interest, compare it with a set of template images of the relevant object classes, and assign it to the class of the template with the the best fit.
+
+The challenge is to come up with accurate template images covering the actual configurations of object and seabed.
+
+Estimate 2D template directly
+  ->
+
+The alternative of using of a static, predetermined template library has a computational complexity that is exponential with the number of parameters.
 This limits its use to only a few parameters being coarsely sampled, ultimately yielding inaccurate results~\cite{Midelfart2010}.

### 2014-09-10 ThesisJP/Thesis.tex
Added final numbers before printing (ISBN etc.).

@@
 \\
-\copyright\,Jon Petter \AA{}sen\\
+\copyright\,Jon Petter Helgesen \AA{}sen\\
 \\
-ISBN XXX-XX-XXX-XXXX-X (printed version)\\
-ISBN XXX-XX-XXX-XXXX-X (electronic version)\\
-ISSN XX-XX\\
+ISBN 978-82-326-0448-7 (printed version)\\
+ISBN 978-82-326-0449-4 (electronic version)\\
+ISSN 1503-8181\\
 \\
-Doctoral theses at NTNU, 2014:XX\\
+Doctoral theses at NTNU, 2014:268\\
 \\
@@
 \vspace{0.2cm}
-\noindent \emph{Ovennevnte avhandling er funnet verdig til \aa{} forsvares offentlig for graden Philosophiae Doctor (Ph.D.) i medisinsk teknologi. Disputas finner sted ... .}
+\noindent \emph{Ovennevnte avhandling er funnet verdig til \aa{} forsvares offentlig for graden Philosophiae Doctor (Ph.D.) i Medisinsk Teknologi.
+Disputas finner sted i Auditoriet, Medisinsk teknisk forskningssenter, torsdag 9.
+oktober 2014 kl.
+12.15.}
 

### 2014-10-16 Simulator/article.tex
la2pdf commit - 2014:10:16	18:12:32	(Thu16.Oct)

--- 2014-08-22 Simulator/article.tex
@@
 The simulator is therefore written in OpenGL and OpenCL and runs on graphics processing units (GPUs).
-The result is a fast performing and portable on-the-fly template generator which can adapt to the characteristics of the current scene.
+The result is a fast performing and portable template generator which can adapt to the characteristics of the current scene on-the-fly.
 \end{abstract}
@@
 
-Automatic target recognition (ATR) is an important component in autonomous underwater vehicles (AUVs), as it allows the vehicle to adapt its mission plan to e.g.
-revisit detected objects for closer examination.
-One way to classify these objects is by using template matching.
-The principle involved  is to isolate an image segment containing the object of interest, compare it with a set of template images of the relevant object classes, and assign it to the class of the template with the the best fit.
+One of the major shortcomings of autonomous underwater vehicles (AUVs) is their lack of curiosity.
+When encountering an interesting object, an AUV with a mounted sonar imaging system should be able to detect it, and perhaps move in for closer examination.
+Besides the apparent challenge of trying to detect something with a long list of unknown parameters, this must happen in real-time and be very power efficient.
 
-The challenge is to come up with accurate template images covering the actual configurations of object and seabed.
-
-Estimate 2D template directly
-  ->
-
-The alternative of using of a static, predetermined template library has a computational complexity that is exponential with the number of parameters.
+The detection process is well described in the field of automatic target recognition (ATR).
+The classical approach involves isolating an object of interest, extracting parameters from it and feeding those into a classifier.
+However, it is hard to find parameters that describe the objects uniquely, in particular due to the ambiguities arising from interpreting a 3D world from its 2D projection.
+Another approach is to use template matching, where the image segment is compared to a set of template images covering the relevant object classes, and assigned to the class of the template yielding the the best fit.
+The challenge here is to come up with accurate template images covering the actual configurations of object and seabed, especially if a static template library is used, which will lead to a computational complexity that is exponential with the number of model parameters.
 This limits its use to only a few parameters being coarsely sampled, ultimately yielding inaccurate results~\cite{Midelfart2010}.
@@
 This is achieved with the aid of the massive computing power in graphics processing units (GPUs) and optimized software libraries for scene rendering.
-The simulator loads a 3D model of the seafloor and an object class model into OpenGL, where emitted sound waves are modeled with a light source placed at the sonar transmit location.
+The simulator loads a 3D model of the seafloor and an object class model into OpenGL, where emitted sound waves are modeled with a light tube extending all the way down the propagation path of the AUV.
 We assume rough, isotropic surfaces reflecting sound energy equally in all directions.
@@
 
-Various simulators for high frequency, side-looking sonar imagery have been published, e.g.~\cite{Bell1997,Sammelm2003}.
+Various simulators for high frequency, side-looking sonar imagery have been published.
+e.g.~\cite{Bell1997,Sammelm2003}.
 However, most implementations have prioritized accurate acoustic modeling at the expense of execution speed.

### 2014-10-20 ThesisJo/derivations.tex
Checked in thesis template

\newpage

\section{Roughness}

\begin{figure}[th]
\graphicsAI[drawing]{gfx/roughness.svg}
\end{figure}

\begin{align}
p(\xi) = \frac{1}{\sqrt{2\pi}\sigma_{\!\xi}}e^{-\frac{\xi^2}{2\sigma_{\!\xi}^2}}
\end{align}

\begin{align}
p(\xi) = \frac{1}{\sqrt{2\pi}\sigma_{\!\xi}}e^{-\frac{\xi^2}{2\sigma_{\!\xi}^2}}
\end{align}

### 2014-10-20 ThesisJo/introduction.tex

Wonders happen every day. What our senses reveal our brain make real. This is the beauty of signal processing.

- The wonder of how the brain process images. Great information carrier.

Electromagnetic radiation - poor range due to attenuation and scattering. Use sound instead. Inferior information carrier. Lower frequency and propagation speed.

Sonar.

- Animals use sound both for imaging and communication underwater.

- Quality. Depends on e.g. sonar frequency, pulse length and directivity. Range resolution given (possibly pulse compressed) signal pulse width. Less data in along-track. resolution given on matched filtered pulse

- Low PRF - long range. High PRF - short range, high along-track resolution. Propagation time large compared to  SAS, low PRF long antenna. Both long range and high along-track resolution.

- Can use models for sidescan. Main difference improved along-track resolution. Existing models accurate under certain assumptions, but can not be used directly to form side-looking image. Model should include physical processes such as acoustic propagation, seabed reverberation and transducer characteristics.

SAS: Range and frequency independent resolution.

Performance:
- AUV revisit. Better image quality, lower data collection price.

### 2014-10-20 ThesisJo/derivations.tex
la2pdf commit - 2014:10:20	17:09:58	(Mon20.Oct)

@@
 
+For a plane monochromatic wave the Rayleigh roughness parameter is quantified by:
+\begin{align}
+\Delta\varphi = 2k\xi\cos\theta.
+\end{align}
+Here $\theta$ is the incidence angle, $k = \frac{2\pi}{\lambda}$ is the wavenumber of the incoming wave, $\xi$ is the distance from the scattering point to the mean surface plane, and $\Delta\varphi$ is the phase difference.
+
 \begin{figure}[th]
-\graphicsAI[drawing]{gfx/roughness.svg}
+\graphicsAI[drawing,width=\linewidth]{gfx/roughness.svg}
 \end{figure}
+
+To see how roughness affects the specular reflection we compute the mean reflected wave as
+\begin{align}
+<p> &= \int\limits_{-\infty}^{\infty} e^{-j\Delta\varphi(\xi)} p(\xi) d\xi,
+\end{align}
+where $p(\xi)$ is the probability density function of $\xi$. If we assume normal distribution, we get:
+\begin{align}
+<p> &= \int\limits_{-\infty}^{\infty} e^{-j2k\xi\cos\theta} \frac{1}{\sqrt{2\pi}\sigma_{\!\xi}}e^{-\frac{\xi^2}{2\sigma_{\!\xi}^2}} d\xi \nn
+&= \frac{1}{\sqrt{2\pi}\sigma_{\!\xi}} \int\limits_{-\infty}^{\infty} e^{-j2k\xi\cos\theta} e^{-\frac{\xi^2}{2\sigma_{\!\xi}^2}} d\xi \nn
+\end{align}
 

### 2014-10-22 ThesisJo/introduction.tex
End of day.

@@
 - AUV revisit. Better image quality, lower data collection price.
+
+\section{Scattering models}
+
+Defining parameters:
+- Roughness
+- Incident angle
+- Frequency
+
+Reflection and scattering from surfaces:
+
+Roughness >> wavelength:
+- Roughness > wavelength: Helmholtz-Kirchhoff (physical optics approach) (Hovem). Leads to Huygins priciple.
+- Roughness < wavelength: Rayleigh-Rice (perturbation technique) (Hovem). More accurate at slightly rough surfaces.
+
+Rayleigh. chi = 2 k sigma sin(theta): Rayleigh criterion: Surface is acoustically smooth when chi << 1.

### 2014-10-24 ThesisJo/derivations.tex
la2pdf commit - 2014:10:20	18:25:51	(Mon20.Oct)
la2pdf commit - 2014:10:20	21:07:17	(Mon20.Oct)
la2pdf commit - 2014:10:20	21:37:26	(Mon20.Oct)
Added derivation of Rayleigh specular formula
la2pdf commit - 2014:10:24	17:34:47	(Fri24.Oct)

@@
 \newpage
+
+\section{Wave Equation}
+
+\begin{table}[!b]\centering%\normalsize
+\begin{tabular}[c]{l r r r@{}  l}\hline
+\rowcolor{tabBlue} & \multicolumn{1}{>{\columncolor{tabBlue}}c}{\bf B$_\text{arith}$} & \multicolumn{1}{>{\columncolor{tabBlue}}c}{\bf B$_\text{mem}$} & \multicolumn{2}{>{\columncolor{tabBlue}}c}{\bf B$_\text{mem}$/B$_\text{arith}$} \\\hline
+Arithmetic & 1.03 Tflop/s & & &\\
+Global memory & & 36 Gfloats/s & \hspace{30pt} 1 &:30 \\
+Shared memory & & 257 Gfloats/s & 1 &:4 \\
+Registers & & $>$1.5 Tfloats/s & $>$3 &:2~\cite{Vasilyy}
+\end{tabular}
+\caption{Nvidia Quadro 6000: Memory throughput, $B_{\lowercase{\text{mem}}}$, compared to arithmetic throughput, $B_\text{arith}$.}\label{throughputs}
+\end{table}%
 
@@
 \begin{align}
-<p> &= \int\limits_{-\infty}^{\infty} e^{-j\Delta\varphi(\xi)} p(\xi) d\xi,
+<p> &= V\int\limits_{-\infty}^{\infty} e^{-j\Delta\varphi(\xi)} p(\xi) d\xi,
 \end{align}
@@
 \begin{align}
-<p> &= \int\limits_{-\infty}^{\infty} e^{-j2k\xi\cos\theta} \frac{1}{\sqrt{2\pi}\sigma_{\!\xi}}e^{-\frac{\xi^2}{2\sigma_{\!\xi}^2}} d\xi \nn
-&= \frac{1}{\sqrt{2\pi}\sigma_{\!\xi}} \int\limits_{-\infty}^{\infty} e^{-j2k\xi\cos\theta} e^{-\frac{\xi^2}{2\sigma_{\!\xi}^2}} d\xi \nn
+<p> &= V\int\limits_{-\infty}^{\infty} e^{-j2k\xi\cos\theta} \frac{1}{\sqrt{2\pi}\sigma}e^{-\frac{\xi^2}{2\sigma^2}} d\xi \nn
+&= \frac{V}{\sqrt{2\pi}\sigma} \int\limits_{-\infty}^{\infty} e^{-\frac{\xi^2}{2\sigma^2}-j2k\xi\cos\theta} d\xi &\Big|{\begin{matrix}\scriptscriptstyle\hspace{-0.64cm} a = \frac{1}{2\sigma^2}\\
+\scriptscriptstyle b = -j2k\cos\theta\end{matrix}}\nn
+&= A \int\limits_{-\infty}^{\infty} e^{-a\xi^2 + b\xi} d\xi \nn
+&= A \int\limits_{-\infty}^{\infty} e^{-a(\xi - \frac{b}{2a})^2 - \frac{b^2}{4a}} d\xi \nn
+&= A e^{\frac{-b^2}{4a}} \int\limits_{-\infty}^{\infty} e^{-a(\xi - \frac{b}{2a})^2} d\xi & \Big|_{y = \xi - \frac{b}{2a}}\nn
+&= A e^{\frac{-b^2}{4a}} \int\limits_{-\infty}^{\infty} e^{-ay^2} dy & \Big|_{z = \sqrt{a} y}\nn
+&= A e^{\frac{-b^2}{4a}}\frac{1}{\sqrt{a}} \int\limits_{-\infty}^{\infty} e^{-z^2} dz \nn
+&= A e^{\frac{-b^2}{4a}}\sqrt{\frac{\pi}{a}} \nn
+&= \frac{V}{\sqrt{2\pi}\sigma} \sqrt{\frac{\pi}{\frac{1}{2\sigma^2}}} e^{\frac{-(-j2k\cos\theta)^2}{4\frac{1}{2\sigma^2}}} \nn
+&= V e^{-2k^2\sigma^2\cos^2\theta}
 \end{align}
-
-\begin{align}
-p(\xi) = \frac{1}{\sqrt{2\pi}\sigma_{\!\xi}}e^{-\frac{\xi^2}{2\sigma_{\!\xi}^2}}
-\end{align}
-
-\begin{align}
-p(\xi) = \frac{1}{\sqrt{2\pi}\sigma_{\!\xi}}e^{-\frac{\xi^2}{2\sigma_{\!\xi}^2}}
-\end{align}
+This reflection coefficient is valid when the Rayleigh parameter is small (i.e.
+small frequency and relief amplitude), and at grazing incidence.
+Conventional limit of validity is $\frac{\pi}{2}$, which corresponds to $\sigma = \frac{\lambda}{8\cos\theta}$ and a coherent loss of 10.7dB.

### 2014-10-27 ThesisJo/derivations.tex
...

@@
 
-\begin{table}[!b]\centering%\normalsize
-\begin{tabular}[c]{l r r r@{}  l}\hline
-\rowcolor{tabBlue} & \multicolumn{1}{>{\columncolor{tabBlue}}c}{\bf B$_\text{arith}$} & \multicolumn{1}{>{\columncolor{tabBlue}}c}{\bf B$_\text{mem}$} & \multicolumn{2}{>{\columncolor{tabBlue}}c}{\bf B$_\text{mem}$/B$_\text{arith}$} \\\hline
-Arithmetic & 1.03 Tflop/s & & &\\
-Global memory & & 36 Gfloats/s & \hspace{30pt} 1 &:30 \\
-Shared memory & & 257 Gfloats/s & 1 &:4 \\
-Registers & & $>$1.5 Tfloats/s & $>$3 &:2~\cite{Vasilyy}
+The wave equation describes how a wave move as a function of both time and space.
+Its importance and application is great, so we will spend some time to derive it here.
+We will treat acoustical waves since they are most relevant to the topic in this thesis, but we could have derived it from electromagnetic waves using Maxwells equations with some boundary conditions.
+
+Define a small volumetric element $dV = dx dy dz$. 
+\begin{wraptable}{r}{8.9cm}%\normalsize
+\begin{tabular}[c]{c c l}\hline
+\rowcolor{tabBlue}\bf Symbol & \bf Unit & \bf Description \\\hline
+$p$ & Pa = N/m$^2$ & Pressure \\
+$\rho$ & kg/m$^3$ & Density \\
+$S$ &  & Entropy \\
+$\vec v$ & m/s & Mass velocity \\
+$\dot m$ & kg/s & Mass flow
 \end{tabular}
-\caption{Nvidia Quadro 6000: Memory throughput, $B_{\lowercase{\text{mem}}}$, compared to arithmetic throughput, $B_\text{arith}$.}\label{throughputs}
-\end{table}%
+\end{wraptable}%
 

### 2014-10-27 ThesisJo/introduction.tex
...

@@
 
+\section{The Wave Equation}
+
+% \begin{align*}
+% \nabla^2 g(\p,t) - \frac{1}{c^2(\p)}
+% \end{align*}
+
+% where $\nabla^2$ is the Laplacian operator
+
+\section{Synethic Aperture Sonar Simulation}
+
+For an autonomous underwater vehicle to reach its full
+
+Most autonomous devices designed today only perform very basic tasks.
+
+One of the areas where very fast where the development of faster and faster processing hardware might play a
+One of the technologies in which graphics processing units might make a welcome addition is in autonomous underwater vehicles.
+
+This thesis will describe a simulator for synthetic aperture sonar that
+
 \section{Scattering models}
@@
 Rayleigh. chi = 2 k sigma sin(theta): Rayleigh criterion: Surface is acoustically smooth when chi << 1.
+
+\begin{itemize}
+\item SWAT (Shallow Water Acoustic Toolset): 
+\item Bell's thesis:
+\item SIGMA:
+\item OpenGL does not support scattered writes. OpenCL does.
+\item 2D template approximation:
+\item Geometric distortions in SSS \cite{Cobra1992} Bell
+\item Sidescan sonar and synthetic aperture sonar. In seafloor coordinates. Maps to orthogonal rendering.
+\item Layover causes ambiguity, different seafloor phenomena can cause the same output image.
+Lots of cornercases when approximating the model in 2D.
+Instead use a 3D model and project it to a 2D one.
+Easier to operate, models are more general purpose.
+Less parameters to adjust.
+\item Viewer can be used to train an operator for finding non-trivial objects on the seafloor. Input for automated target recognition standard.
+\item Improved image quality of selected object, reduced data collection cost (price estimate?).
+\item OpenGL rendering can have infinite resolution, only limited by the pixel density in the image.
+\item Accuracy in terms of classification result. Need to embody static effects in a deterministic way. Incoherent synthesis of the image.
+\item Sound speed assumed constant.
+\item High frequency lead to little (define) penetration, lower range and higher resolution. Often modeled accurately with ray tracing techniques.
+\end{itemize}
+
+Model
+\begin{itemize}
+\item Ray solution: Separate amplitude and phase components of the Helmholtz eq.
+by applying geometric optics approximation.
+Ok as long as amplitude varies more slowly with position than the phase.
+Implications: High frequency, because the curvature of a ray over a wavelength must be small, and the fractional change in the sound speed must be small over a wavelength.
+Further simplification available by Snell's Law.
+Fast to compute.
+Not able to handle diffraction, focal points and caustics, and will cause (too) abrupt changes on distinct changes in sea environment.
+Can be improved upon with fuzzy beam modeling / Gaussian beam tracing.
+\begin{itemize}
+\item Pro: Easy to compute. Visual result. Handles complex bathymetry. Array and model response can be accounted for.
+\item Con: Only works for high frequency and small sound variations. No diffraction, caustics and focus. Infinite energy caustics/sharp variations. Incoherent(?)
+\end{itemize}
+\item Normal modes: Separate amplitude and phase into a range dependent term, and a depth dependent term.
+Assumes minimal range dependence in sound speed profile and depth.
+Limited near field accuracy.
+Acoustic field does not need to be calculated at intermediate ranges.
+Penetration can be modeled.
+can be calculated
+\begin{itemize}
+\item Pro: Ok to compute. Handles penetration. Low frequency, shallow water.
+\item Con: Horizontal stratification. Far field only. 
+\end{itemize}
+\item Parabolic equation: One way propagation.
+Marching solution.
+Initial starting solution must be known, often computed with normal mode.
+Only valid for narrow beams $\pm$20.
+Able to handle range dependent environments.
+Computing full field is easy.
+Computation time is of O(f$^2$), hence mostly useful for low-frequency scenarios.
+Unable to deal with horizontal refraction (shadow zones from e.g.
+conical mounts)
+\item Fast field programs / Green's function solutions.
+Exact full wave solution for acoustic fields in horizontally stratified media.
+Can handle both compressional and shear waves, and hence both fluid and solid media.
+Far field assumption.
+Similar to normal mode techniques.
+\item Finite element: Scene segmented into triangles or rectangles with a size of one tenth of a wavelength, or less.
+At each node the wavefield can be determined from the wave equation.
+Computationally limited to low frequencies.
+\item 2D model: Acoustic field in range and depth.
+Assumes that a wave transmitted in a vertical plane remains in it.
+Reasonable when the sound speed can be assumed constant in the horizontal plane, and when the seafloor normals are in the vertical range-depth plane (avoid horizontal refraction, or actually reflection).
+\end{itemize}
+
+What's new?
+\begin{itemize}
+\item Tuned for speed using GPUs, combining OpenGL and OpenCL to take advantage of existing functionality for computer graphics with the added flexibility of general purpose programming on the GPUs.
+\item Performance to spare: Game-like viewer for the visualization and verification of the rendered scene and corresponding sonar image.
+\item First simulator of its kinds aimed towards improved ATR and ultimately better AUVs.
+\item 3D models are still novel. 
+\end{itemize}

### 2014-10-27 ThesisJo/derivations.tex
la2pdf commit - 2014:10:27	16:25:27	(Mon27.Oct)
la2pdf commit - 2014:10:27	17:00:34	(Mon27.Oct)
la2pdf commit - 2014:10:27	17:39:25	(Mon27.Oct)
More on wave equation

@@
 The wave equation describes how a wave move as a function of both time and space.
-Its importance and application is great, so we will spend some time to derive it here.
+It forms the basis for all mathematical modeling of waves, so we will spend some time to derive it here.
 We will treat acoustical waves since they are most relevant to the topic in this thesis, but we could have derived it from electromagnetic waves using Maxwells equations with some boundary conditions.
 
-Define a small volumetric element $dV = dx dy dz$. 
-\begin{wraptable}{r}{8.9cm}%\normalsize
+\begin{figure}[ht]
+\begin{floatrow}
+
+\floatbox{figure}[\linewidth][\FBheight][t]{%
+\caption{Mass flow}}{%
+\graphicsAI[drawing,width=.77\linewidth]{gfx/wave_equation_mass_flow.svg}}%
+
+\floatbox{table}[\linewidth][\FBheight][t]{%
+\caption{Symbol description}}{%
 \begin{tabular}[c]{c c l}\hline
@@
 $\dot m$ & kg/s & Mass flow
-\end{tabular}
-\end{wraptable}%
+\end{tabular}}
+
+\end{floatrow}
+\end{figure}
+
+We begin by defining the mass flow in and out of a volumetric element $dV = dx dy dz$:
+
+\begin{align}
+\dot m_\text{in} = \rho\,\v\,dV \qquad\qquad \text{and} \qquad\qquad \dot m_\text{out} = \rho\,\v\,dV + 
+\end{align}
 

### 2014-10-28 ThesisJo/derivations.tex
la2pdf commit - 2014:10:28	14:33:53	(Tue28.Oct)

@@
 $\rho$ & kg/m$^3$ & Density \\
-$S$ &  & Entropy \\
+$S$ & Nm/K & Entropy \\
 $\vec v$ & m/s & Mass velocity \\
@@
 
-We begin by defining the mass flow in and out of a volumetric element $dV = dx dy dz$:
+We begin by defining the net mass flow in and out of a volumetric element $dV = dx dy dz$:
 
 \begin{align}
-\dot m_\text{in} = \rho\,\v\,dV \qquad\qquad \text{and} \qquad\qquad \dot m_\text{out} = \rho\,\v\,dV + 
+\dot m_\text{net} = \Big(\frac{\partial \rho\,\v}{\partial x} + \frac{\partial \rho\,\v}{\partial y} + \frac{\partial \rho\,\v}{\partial z} \Big) dV = \nabla \cdot (\rho\v)\, dV
 \end{align}

### 2014-10-28 ThesisJo/derivations.tex
la2pdf commit - 2014:10:28	15:04:01	(Tue28.Oct)

--- 2014-10-27 ThesisJo/derivations.tex
@@
 $\rho$ & kg/m$^3$ & Density \\
-$S$ &  & Entropy \\
+$S$ & Nm/K & Entropy \\
 $\vec v$ & m/s & Mass velocity \\
-$\dot m$ & kg/s & Mass flow
+$\dot{\boldsymbol{m}}$ & kg/s & Mass flow
 \end{tabular}}
@@
 
-We begin by defining the mass flow in and out of a volumetric element $dV = dx dy dz$:
-
+We begin by defining the net \textbf{mass flow} through a volumetric element $dV = dx\,dy\,dz$:
+%
 \begin{align}
-\dot m_\text{in} = \rho\,\v\,dV \qquad\qquad \text{and} \qquad\qquad \dot m_\text{out} = \rho\,\v\,dV + 
+\frac{\dot m_\text{net}}{dV} = \frac{\partial \rho\,\v}{\partial x} \ux + \frac{\partial \rho\,\v}{\partial y} \uy + \frac{\partial \rho\,\v}{\partial z} \uz = \nabla \cdot (\rho\v):
 \end{align}
+%
+The space and time dependence of the mass density $\rho = \rho(\p,t)$ and particle velocity $\v = \v(\p,t)$ was omitted here for brevity.
+If the volume of the element is fixed a change in mass must equal a corresponding change in density.
+This is the law of \textbf{mass conservation}:
+%
+\begin{align}
+\frac{\partial\rho}{\partial t} = -\nabla \cdot (\rho\v)
+\end{align}
+%
 

### 2014-10-28 ThesisJo/derivations.tex
la2pdf commit - 2014:10:28	15:36:59	(Tue28.Oct)

@@
 \begin{align}
-\frac{\dot m_\text{net}}{dV} = \frac{\partial \rho\,\v}{\partial x} \ux + \frac{\partial \rho\,\v}{\partial y} \uy + \frac{\partial \rho\,\v}{\partial z} \uz = \nabla \cdot (\rho\v):
+\frac{\dot{\boldsymbol{m}}_\text{net}}{dV} = \ux\frac{\partial \rho\,\v}{\partial x} + \uy\frac{\partial \rho\,\v}{\partial y} + \uz\frac{\partial \rho\,\v}{\partial z} = \nabla \cdot (\rho\v).
 \end{align}
 %
-The space and time dependence of the mass density $\rho = \rho(\p,t)$ and particle velocity $\v = \v(\p,t)$ was omitted here for brevity.
+The space and time dependence of the mass density $\rho = \rho(\p,t)$ and particle velocity $\v = \v(\p,t)$ is omitted here and in the following text for brevity.
 If the volume of the element is fixed a change in mass must equal a corresponding change in density.
@@
 \begin{align}
-\frac{\partial\rho}{\partial t} = -\nabla \cdot (\rho\v)
+\frac{\partial\rho}{\partial t} = -\nabla \cdot (\rho\v).
+\end{align}
+%
+An additional constraint we may surely apply is the Euler's equation, a modification of Newton's second law:
+%
+\begin{align}
+\frac{F}{dV} = \frac{m}{dV} a = \rho\Big(\frac{\partial\v}{\partial t} + \nabla\v\Big) = -\nabla p.
 \end{align}

### 2014-10-28 ThesisJo/derivations.tex
la2pdf commit - 2014:10:28	16:07:33	(Tue28.Oct)

@@
 It forms the basis for all mathematical modeling of waves, so we will spend some time to derive it here.
-We will treat acoustical waves since they are most relevant to the topic in this thesis, but we could have derived it from electromagnetic waves using Maxwells equations with some boundary conditions.
+We will treat acoustical waves since they are most relevant to the topic in this thesis, but we could have derived it from electromagnetic waves using Maxwell's equations with some boundary conditions.
 
@@
 %
-An additional constraint we may surely apply is the Euler's equation, a modification of Newton's second law:
+An additional constraint we may surely apply is the Euler's equation, a modification of Newton's second law the the total acceleration $a$ is is a function of both time and space:
 %
 \begin{align}
-\frac{F}{dV} = \frac{m}{dV} a = \rho\Big(\frac{\partial\v}{\partial t} + \nabla\v\Big) = -\nabla p.
+\frac{F}{dV} = \frac{m}{dV} a = \rho\Big(\frac{\partial\v}{\partial t} + \v\cdot\nabla\v\Big) = -\nabla p.
 \end{align}
 %
+Next, an \textbf{equation of state} is needed to relate a change in density to a change in pressure.
+We assume our volume element to be isentropic, i.e.
+that the entropy is constant, or equivalently that total energy of the system is proportional to its temperature.
+This is true for any system that is reversible and adiabatic, i.e.
+that when no transfer of matter or heat occurs between the system and its surroundings.
+In other words, we assume the system to be \emph{loss-less} system.
+In reality a passing sound wave will heat the fluid and add to the entropy, ultimately leading to additional attenuation of the wave, but we will not deal with this topic in this thesis.
+
+The pressure of the system then just depends on the mass density:
+%
+\begin{align}
+p = p(\rho)
+\end{align}
 

### 2014-10-28 ThesisJo/derivations.tex
la2pdf commit - 2014:10:28	16:37:42	(Tue28.Oct)

@@
 \begin{align}
-\frac{\partial\rho}{\partial t} = -\nabla \cdot (\rho\v).
+\frac{\partial\rho}{\partial t} = -\nabla \cdot (\rho\v). \label{wave_eq_mass_conservation}
 \end{align}
@@
 \begin{align}
-\frac{F}{dV} = \frac{m}{dV} a = \rho\Big(\frac{\partial\v}{\partial t} + \v\cdot\nabla\v\Big) = -\nabla p.
+\frac{F}{dV} = \frac{m}{dV} a = \rho\Big(\frac{\partial\v}{\partial t} + \v\cdot\nabla\v\Big) = -\nabla p. \label{wave_eq_euler}
 \end{align}
@@
 that when no transfer of matter or heat occurs between the system and its surroundings.
-In other words, we assume the system to be \emph{loss-less} system.
-In reality a passing sound wave will heat the fluid and add to the entropy, ultimately leading to additional attenuation of the wave, but we will not deal with this topic in this thesis.
-
 The pressure of the system then just depends on the mass density:
@@
 \begin{align}
-p = p(\rho)
+p = p(\rho) \label{wave_eq_sound_pressure}
+\end{align}
+
+With this we have assumed the system to be \emph{loss-less}.
+In reality a passing sound wave will heat the fluid and add to the entropy, causing additional attenuation of the wave, but we will not be concerned with attenuation in this thesis.
+
+Equation (\ref{wave_eq_mass_conservation}), (\ref{wave_eq_euler}) and (\ref{wave_eq_sound_pressure}) are non-linear, but can be linearized by assuming each quantity to be a steady-state, time-independent value plus a fluctuation term:
+%
+\begin{align}
+p = p_0 + p'  \qquad\qquad \rho = \rho_0 + \rho' \qquad\qquad \v = \v_0 + \v'
+\end{align}
+%
+Taylor expanding (\ref{wave_eq_sound_pressure}) around zero yields:
+%
+\begin{align}
+p(\rho) = p_0 + \frac{\partial p}{\partial\rho}(\rho-\rho_0) + \frac{1}{2}\frac{\partial^2 p}{\partial\rho^2}(\rho-\rho_0)^2 + \dots\label{wave_eq_sound_taylor_full}
+\end{align}
+%
+and if we can assume the first order to be sufficient we get:
+%
+\begin{align}
+p(\rho) = p_0 + \frac{\partial p}{\partial\rho}(\rho-\rho_0)\label{wave_eq_sound_taylor_first}
 \end{align}

### 2014-10-28 ThesisJo/derivations.tex
la2pdf commit - 2014:10:28	17:07:58	(Tue28.Oct)

@@
 
+\subsection{Linearization}
+
 Equation (\ref{wave_eq_mass_conservation}), (\ref{wave_eq_euler}) and (\ref{wave_eq_sound_pressure}) are non-linear, but can be linearized by assuming each quantity to be a steady-state, time-independent value plus a fluctuation term:
@@
 %
-and if we can assume the first order to be sufficient we get:
+Assuming a linear relationship between changes in pressure and volume the pressure perturbation term is simply:
 %
 \begin{align}
-p(\rho) = p_0 + \frac{\partial p}{\partial\rho}(\rho-\rho_0)\label{wave_eq_sound_taylor_first}
+p' = \frac{\partial p}{\partial\rho}(\rho-\rho_0) = \frac{\partial p}{\partial\rho}\rho' = c^2\rho'. \label{wave_eq_sound_taylor_first}
+\end{align}
+%
+When treating the Euler equation the same way we obtain:
+%
+\begin{align}
+-\nabla (p_0 + p') &= (\rho_0 + \rho') \Big(\frac{\partial(\v_0 + \v')}{\partial t} + (\v_0 + \v')\cdot\nabla(\v_0 + \v')\Big) \nn
+-\nabla p' &= (\rho_0 + \rho') \Big(\frac{\partial\v'}{\partial t} + \v'\cdot\nabla\v'\Big) \nn
+\end{align}
+%
+When acceleration mainly depends on time, and the pressure fluctuations are relatively small,
+%
+\begin{align}
+\frac{\partial\v'}{\partial t} >> \v'\cdot\nabla\v' \qquad\qquad\text{and}\qquad\qquad \rho_0 >> \rho',
+\end{align}
+%
+then
+%
+\begin{align}
+-\nabla p' &= \rho_0 \frac{\partial\v'}{\partial t}.
+\end{align}
+%
+Finally, for the mass conservation we have:
+%
+\begin{align}
+\frac{\partial(\rho_0 + \rho')}{\partial t} = -\nabla \cdot ((\rho_0 + \rho')(\v_0 + \v')). \label{wave_eq_mass_conservation}
 \end{align}

### 2014-11-11 ThesisJo/derivations.tex
Wave equation derived
...
la2pdf commit - 2014:11:11	18:09:08	(Tue11.Nov)

@@
 \rowcolor{tabBlue}\bf Symbol & \bf Unit & \bf Description \\\hline
-$p$ & Pa = N/m$^2$ & Pressure \\
+$s$ & Pa = N/m$^2$ & Pressure \\
 $\rho$ & kg/m$^3$ & Density \\
@@
 \begin{align}
-\frac{F}{dV} = \frac{m}{dV} a = \rho\Big(\frac{\partial\v}{\partial t} + \v\cdot\nabla\v\Big) = -\nabla p. \label{wave_eq_euler}
+\frac{F}{dV} = \frac{m}{dV} a = \rho\Big(\frac{\partial\v}{\partial t} + \v\cdot\nabla\v\Big) = -\nabla s. \label{wave_eq_euler}
 \end{align}
@@
 \begin{align}
-p = p(\rho) \label{wave_eq_sound_pressure}
+s = s(\rho) \label{wave_eq_sound_pressure}
 \end{align}
@@
 \begin{align}
-p = p_0 + p'  \qquad\qquad \rho = \rho_0 + \rho' \qquad\qquad \v = \v_0 + \v'
+s = s_0 + s'  \qquad\qquad \rho = \rho_0 + \rho' \qquad\qquad \v = \v_0 + \v'
 \end{align}
@@
 \begin{align}
-p(\rho) = p_0 + \frac{\partial p}{\partial\rho}(\rho-\rho_0) + \frac{1}{2}\frac{\partial^2 p}{\partial\rho^2}(\rho-\rho_0)^2 + \dots\label{wave_eq_sound_taylor_full}
+s(\rho) = s_0 + \frac{\partial s}{\partial\rho}(\rho-\rho_0) + \frac{1}{2}\frac{\partial^2 s}{\partial\rho^2}(\rho-\rho_0)^2 + \dots\label{wave_eq_sound_taylor_full}
 \end{align}
@@
 \begin{align}
-p' = \frac{\partial p}{\partial\rho}(\rho-\rho_0) = \frac{\partial p}{\partial\rho}\rho' = c^2\rho'. \label{wave_eq_sound_taylor_first}
+s' = \frac{\partial s}{\partial\rho}(\rho-\rho_0) = \frac{\partial s}{\partial\rho}\rho' = c^2\rho'. \label{wave_eq_sound_taylor_first}
 \end{align}
@@
 \begin{align}
--\nabla (p_0 + p') &= (\rho_0 + \rho') \Big(\frac{\partial(\v_0 + \v')}{\partial t} + (\v_0 + \v')\cdot\nabla(\v_0 + \v')\Big) \nn
--\nabla p' &= (\rho_0 + \rho') \Big(\frac{\partial\v'}{\partial t} + \v'\cdot\nabla\v'\Big) \nn
+-\nabla (s_0 + s') &= (\rho_0 + \rho') \Big(\frac{\partial(\v_0 + \v')}{\partial t} + (\v_0 + \v')\cdot\nabla(\v_0 + \v')\Big) \nn
+-\nabla s' &= (\rho_0 + \rho') \Big(\frac{\partial\v'}{\partial t} + \v'\cdot\nabla\v'\Big) \nn
 \end{align}
@@
 \begin{align}
--\nabla p' &= \rho_0 \frac{\partial\v'}{\partial t}.
+-\nabla s' &= \rho_0 \frac{\partial\v'}{\partial t}.
 \end{align}
@@
 \begin{align}
-\frac{\partial(\rho_0 + \rho')}{\partial t} = -\nabla \cdot ((\rho_0 + \rho')(\v_0 + \v')). \label{wave_eq_mass_conservation}
+\frac{\partial(\rho_0 + \rho')}{\partial t} &= -\nabla \cdot \big((\rho_0 + \rho')(\v_0 + \v')\big)\nn
+\frac{\partial\rho'}{\partial t} &= -\rho_0\nabla \cdot \v'. \label{wave_eq_mass_conservation2}
 \end{align}
+%
+This assumes that the density is constant inside the volume element.
+
+\begin{align}
+\nabla\cdot(-\nabla s') &= \nabla\cdot\Big(\rho_0 \frac{\partial\v'}{\partial t}\Big) &
+\frac{\partial}{\partial t}\frac{\partial\rho'}{\partial t} &= \frac{\partial}{\partial t}\Big(-\rho_0\nabla \cdot \v'\Big)\nn
+-\nabla^2 s' &= \rho_0 \nabla\cdot\frac{\partial\v'}{\partial t} &
+\frac{\partial^2\rho'}{\partial t^2} &= \frac{1}{c^2}\frac{\partial^2 s}{\partial t^2} = -\rho_0\nabla\cdot\frac{\partial \v'}{\partial t}
+\end{align}
+%
+and we can now see that
+%
+\begin{align}
+\nabla^2 s' - \frac{1}{c^2}\frac{\partial^2 s'}{\partial t^2} = 0.
+\end{align}
+%
+This is called the homogeneous lossless wave equation, because there is no source term. The inhomogeneous equation can be obtained in a similar way, by altering the linearized continuity equation to:
+%
+\begin{align}
+\frac{\partial(\rho_0 + \rho')}{\partial t} &= -\nabla \cdot \big((\rho_0 + \rho')(\v_0 + \v')\big)\nn
+\frac{\partial\rho'}{\partial t} &= -\rho_0\nabla \cdot \v'. \label{wave_eq_mass_conservation2}
+\end{align}
+%
 

### 2014-11-12 ThesisJo/derivations.tex
la2pdf commit - 2014:11:12	12:38:23	(Wed12.Nov)

@@
 
-The wave equation describes how a wave move as a function of both time and space.
+The wave equation describes how a wave moves as a function of both time and space.
 It forms the basis for all mathematical modeling of waves, so we will spend some time to derive it here.
@@
 \begin{align}
-\frac{\partial(\rho_0 + \rho')}{\partial t} &= -\nabla \cdot \big((\rho_0 + \rho')(\v_0 + \v')\big)\nn
-\frac{\partial\rho'}{\partial t} &= -\rho_0\nabla \cdot \v'. \label{wave_eq_mass_conservation2}
+\frac{\partial\rho'}{\partial t} &= -\rho_0\nabla \cdot \v' - \rho_0f(\p,t). \label{wave_eq_mass_conservation_source}
+\end{align}
+%
+where $\rho_0f(\p,t)$ is the rate of mass change (in $\left[\frac{\text{kg}}{\text{m}^3\text{s}}\right]$) caused by the source. Following the exact same procedure as before we obtain:
+%
+\begin{align}
+\nabla^2 s' - \frac{1}{c^2}\frac{\partial^2 s'}{\partial t^2} = \rho_0\frac{\partial f(\p,t)}{\partial t}.
 \end{align}

### 2014-11-12 ThesisJo/derivations.tex
la2pdf commit - 2014:11:12	19:03:21	(Wed12.Nov)

@@
 
+\section{Solution}
+
+Instead of solving the wave equation directly, we will make a qualified guess that the solution takes the form of a complex exponential:
+%
+\begin{align}
+s(\p,t) &= A\,e^{j(\omega t - \k\dot\p)}
+\end{align}
+%
+
+\section{Absorption}
+
+Our derivation of the wave equation assumed no loss of energy.
+In practice the wave's energy is abosorbed through viscous losses, heat conduction losses and intermolecular losses.
+These are strongly dependent on frequency.
+The wave equation can be adapted to account for absorption by assuming a complex sound speed, which will lead to a solution to the wave equation with a real exponential term.
+In this thesis we will not concern ourselves
+
 \section{Roughness}

### 2014-11-12 ThesisJo/derivations.tex
la2pdf commit - 2014:11:12	19:47:04	(Wed12.Nov)

@@
 
-Instead of solving the wave equation directly, we will make a qualified guess that the solution takes the form of a complex exponential:
+Instead of solving the wave equation directly, we will make a qualified guess that the solution takes the form of a complex exponential in both time and space:
 %
 \begin{align}
-s(\p,t) &= A\,e^{j(\omega t - \k\dot\p)}
+s(\p,t) &= A\,e^{j(\omega t - \k\cdot\p)}
 \end{align}
 %
+Inserting this into the wave equation yields:
+%
+\begin{align}
+\nabla^2 s = \frac{\partial^2 s}{\partial x^2} + \frac{\partial^2 s}{\partial y^2} + \frac{\partial^2 s}{\partial z^2} &= \frac{1}{c^2}\frac{\partial^2 s}{\partial t^2} \nn
+\frac{\partial^2 s}{\partial x^2} + \frac{\partial^2 s}{\partial y^2} + \frac{\partial^2 s}{\partial z^2} &= \frac{1}{c^2}\frac{\partial^2 s}{\partial t^2} \nn
+\end{align}
 

### 2014-11-17 ThesisJo/derivations.tex
la2pdf commit - 2014:11:17	14:31:50	(Mon17.Nov)

--- 2014-11-12 ThesisJo/derivations.tex
@@
 
-Instead of solving the wave equation directly, we will make a qualified guess that the solution takes the form of a complex exponential:
+Instead of solving the wave equation directly, we will make a qualified guess that the solution takes the form of a complex exponential in both time and space.
+For rectangular coordinates this can be written as:
 %
 \begin{align}
-s(\p,t) &= A\,e^{j(\omega t - \k\dot\p)}
+s(\p,t) &= A\,e^{j(\omega t - \k\cdot\p)}
 \end{align}
 %
+Inserting this into the wave equation yields:
+%
+\begin{align}
+\nabla^2 s = \frac{\partial^2 s}{\partial x^2} + \frac{\partial^2 s}{\partial y^2} + \frac{\partial^2 s}{\partial z^2} &= \frac{1}{c^2}\frac{\partial^2 s}{\partial t^2} \nn
+(-jk_x)^2 s + (-jk_y)^2 s + (-jk_z)^2 s &= \frac{(-j\omega)^2}{c^2} s \nn
+k_x^2 + k_y^2 + k_z^2 &= |\k|^2 = \frac{\omega^2}{c^2} \\
+|\k| &= \frac{\omega}{c}
+\end{align}
+%
+Hence, whenever we can assume the spatial and temporal frequency to be related directly by the sound speed, $|\k| = \frac{\omega}{c}$, then this assuming a complex exponential should be fair.
 

### 2014-11-18 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2012:03:09	02:25:52	(Fri 9.Mar)
la2pdf commit - 2012:03:09	02:56:52	(Fri 9.Mar)
la2pdf commit - 2012:03:09	05:24:15	(Fri 9.Mar)
LCA article on the way
la2pdf commit - 2012:03:09	18:08:10	(Fri 9.Mar)
la2pdf commit - 2012:03:12	17:51:05	(Mon12.Mar)
la2pdf commit - 2012:03:13	10:48:10	(Tue13.Mar)
la2pdf commit - 2012:03:13	16:52:18	(Tue13.Mar)
la2pdf commit - 2012:03:14	13:58:41	(Wed14.Mar)
la2pdf commit - 2012:03:15	13:29:07	(Thu15.Mar)
la2pdf commit - 2014:11:17	14:31:50	(Mon17.Nov)
Patching together LCA article

@@
                                                 % Journal version
+\newif\ifOnlineColor\OnlineColortrue            % Compile online color version?
+
 \newif\ifFlatArchive\FlatArchivefalse           % Whether archive is flat (messy) or contain 
@@
 
+\ifOnlineColor
+   \newcommand\figPostfix{_online}              % Colored images online
+\else
+   \newcommand\figPostfix{_bw}                  % Black and white in journal
+\fi
+
 %%%%%%%%%%%%                                    ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
@@
 \ifPeerReview
-\documentclass[12pt,journal,captionsoff,onecolumn]{IEEEtran}
-\newcommand\CLASSINPUTbaselinestretch{1.66}     % http://theoval.cmp.uea.ac.uk/~nlct/latex/thesis/node17.html
+\documentclass[10pt,journal,draftclsnofoot,onecolumn]{IEEEtran}
+% \newcommand\CLASSINPUTbaselinestretch{1.66}     % http://theoval.cmp.uea.ac.uk/~nlct/latex/thesis/node17.html
 \else
@@
 
+% \RequirePackage[latin1]{inputenc}%              % Set input encoding (optionally latin1)
+% \RequirePackage[T1]{fontenc}%                 % Set font encoding
+% \usepackage[norsk]{babel}
+
+% Single column review mode:
+%\documentclass[12pt,journal,onecolumn]{IEEEtran}
+%\newcommand\CLASSINPUTbaselinestretch{2}
+%\RequirePackage{calc}
+%\RequirePackage{fp} 
+% Left margin = 1 inch + hoffset + oddsidemargin (or evensidemargin)
+% Adding the 2cm gutter width to the odd/even side margins:
+%\setlength\hoffset{0pt}
+%\setlength\oddsidemargin{4cm}       % 4cm margins on the left side
+%\addtolength\oddsidemargin{-1in}    % Subtract the initial 1 inch
+%\setlength\textwidth{21cm-8cm}    % A4 width (21cm) minus margins on either side
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%                  ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%                  ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
+
+\usepackage{cite}
 
@@
 \fi
-\ifCLASSOPTIONcaptionsoff                       % IEEE promoted hack to turn off captions from the 
-   \let\MYorigsubfloat\subfloat                 % subfloat package should the captionsoff option
-   \renewcommand{\subfloat}[2][\relax]{\MYorigsubfloat[]{#2}} % be specified.
-\fi
+
+% \usepackage{caption}
+\usepackage{subcaption}
+\usepackage{color}
+\usepackage{calc}
+\usepackage{fp}
+
+% 
+% \ifCLASSOPTIONcaptionsoff                       % IEEE promoted hack to turn off captions from the 
+%    \let\MYorigsubfloat\subfloat                 % subfloat package should the captionsoff option
+%    \renewcommand{\subfloat}[2][\relax]{\MYorigsubfloat[]{#2}} % be specified.
+% \fi
 
@@
 colorlinks=true,linkcolor={black},citecolor={black},urlcolor={black},
-pdftitle={Low Complexity Adaptive Beamformer for Active Sonar Imaging},
+pdftitle={A Fast GPU Sonar Simulator for Automatic Target Recognition},
 pdfsubject={},
@@
 
+% \ifPeerReview
+% \let\OldIncludegraphics{\includegraphics}
+% \usepackage{letltxmacro}
+% \LetLtxMacro{\OldIncludegrsaphics}{\includegraphics}
+% \renewcommand{\includegraphics}[2][]{\OldIncludegraphics[width=\linewidth, #1]{#2}}
+% \fi
+
+% % \makeatletter
+% % \def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth
+% % \else\Gin@nat@width\fi}
+% % \makeatother
+% % \let\Oldincludegraphics\includegraphics
+% % \renewcommand{\includegraphics}[1]{\Oldincludegraphics[width=\maxwidth]{#1}}
+
+\usepackage[maxfloats=40]{morefloats}
 \newcounter{todoidx}
@@
       \addtocounter{todoidx}{1}%
-      {\color{Red}\fbox{\bf\thetodoidx{}}}%
+      {\color{Red}\bf(\thetodoidx{})}%%\fbox{\bf\thetodoidx{}}}%
       \marginpar{%
          {\vspace*{-10pt}\color{Red}\fbox{\bf\thetodoidx{}}}\\%
-         \fcolorbox{red}{todobackground}{\parbox{\marginparwidthsmall}{#1}}}}
+         \fcolorbox{red}{todobackground}{\parbox{\marginparwidthsmall}{\scriptsize #1}}}}
 
@@
 
+\newenvironment{narrow}[2]{%
+\begin{list}{}{%
+\setlength{\topsep}{0pt}%
+\setlength{\leftmargin}{#1}%
+\setlength{\rightmargin}{#2}%
+\setlength{\listparindent}{\parindent}%
+\setlength{\itemindent}{\parindent}%
+\setlength{\parsep}{\parskip}}%
+\item[]}{\end{list}}
+
+\usepackage{float}
+
+\ifOnlineColor
+   \definecolor{tabBlue}{HTML}{AACCFF}
+\else
+   \definecolor{tabBlue}{HTML}{CCCCCC}
+\fi
+
 %%%%%%%%%%                                      ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
@@
 %%%%%%%%%%                                      ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
+
+\newcommand\graphicsAI[2][]{%
+  \immediate\write18{./bin/laFigure #2 #1}%
+  \input{result}}%
+
+% \DeclareMathOperator*{\argmin}{\text{arg}\;\text{min}}
+
+\newcommand\Fig[1]{Fig.~\ref{#1}}
 
@@
 % \newcommand\max{\text{max}}
-\newcommand\argmin{\text{argmin}}
+\newcommand\argmin[1]{\text{arg}\;\underset{#1}{\text{min}}}
 
@@
 
-\renewcommand\l[1]{\left#1}
-\renewcommand\r[1]{\right#1}
+% \renewcommand\l[1]{\left#1}
+% \renewcommand\r[1]{\right#1}
 
@@
 
-\newcommand\sumu[1]{\sum\limits^{#1}\,}
-\newcommand\suml[1]{\sum\limits_{#1}\,}
-\newcommand\sumb[2]{\sum\limits_{#1}^{#2}\,}
+\newcommand\sumu[1]{\sum\limits^{#1}\;}
+\newcommand\suml[1]{\sum\limits_{#1}\;}
+\newcommand\sumb[2]{\sum\limits_{#1}^{#2}\;}
 
-\newcommand\produ[1]{\prod\limits^{#1}\,}
-\newcommand\prodl[1]{\prod\limits_{#1}\,}
-\newcommand\prodb[2]{\prod\limits_{#1}^{#2}\,}
+\newcommand\produ[1]{\prod\limits^{#1}\;}
+\newcommand\prodl[1]{\prod\limits_{#1}\;}
+\newcommand\prodb[2]{\prod\limits_{#1}^{#2}\;}
+
+\newcommand\defeq{\overset{\underset{\mathrm{def}}{}}{=}}
 
 %Math macros:
-\newcommand\diff[2]{\frac{\kw{d}\,\textstyle #1\scriptstyle}{\kw{d\parm{#2}}}\displaystyle}
-\newcommand\ddiff[2]{\frac{\kw{d^2}\,\displaystyle #1\scriptstyle}{\kw{d\parm{#2}}^2}\displaystyle}
-
-\renewcommand\d[1]{\scriptstyle\kw{\,d\parm{#1}}\displaystyle}
-
-% These commands are mutually exclusive. Remember to "renew" in v2.
-\newcommand\intb[4]{\int\limits_{#3}^{#4} #1 \d{#2}} % \int{exp}{var}{from}{to}
-\newcommand\intl[3]{\int\limits_{#3} #1 \d{#2}} % \int{exp}{var}{for all}
-\newcommand\intu[2]{\int #1 \d{#2}} % \int{exp}{var}{for all}
-
 \newcommand\T{^{\scriptscriptstyle T}}
@@
 
+\newcommand\Om{O_\text{m}}
+\newcommand\Oa{O_\text{a}}
+\newcommand\Nl{N_\text{l}}
+\newcommand\Nk{N_\text{k}}
+\newcommand\1{\vec 1}
+\newcommand\I{\mat I}
 \renewcommand*\a{\vec a}
@@
 \newcommand*\E{\mat E}
-% \renewcommand*\H{\mat H}
 \renewcommand*\P{\mat P}
@@
 \newcommand*\eR{\mat{\hat R}}
+\newcommand*\eRi{\hat{\mat R}\;\!^{-1}}
+\newcommand*\Navg{N_\text{avg}}
 \newcommand*\W{\mat W}
@@
 \newcommand*\Y{\mat Y}
+
+\renewcommand\argmin{\text{argmin}}
+
+\renewcommand*\P{\mat P}
+\newcommand*\V{\mat V}
+\newcommand*\M{\mat M}
 
@@
 
-\newcommand\D{\vec\nabla} %Del: Vector differential operator - nabla
-\newcommand\Dx{\vec\nabla\times}
-\newcommand\Dd{\vec\nabla\cdot}
-
 \usepackage{tikz}
@@
 \usepackage{amsmath,amssymb}
+% \usepackage{datatool}
+% \usepackage{glossaries}
 
@@
 {\end{itemize}}
+
+% correct bad hyphenation here
+% \hyphenation{op-tical net-works semi-conduc-tor}
+
+%%%%%%%%%%%%                                    ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
+% GLOSSARY %
+%%%%%%%%%%%%                                    ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
+
+% \makeglossaries
+% \input{../../Library/glossary}
+
+%%%%%%%%%%%%%%%%%%                              ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
+% DOCUMENT START %
+%%%%%%%%%%%%%%%%%%                              ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
+
+% \usepackage{yfonts}
+
+% \graphicspath{submission/final/gfx/}
+
+% 
+% %%%%%%%%%%                                      ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
+% % MACROS %       
+% %%%%%%%%%%                                      ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
+% 
+% \newcommand\Fig[1]{Fig.~\ref{#1}}
+% 
+% \newcommand\Grey[1]{{\color{Grey}#1}}
+% \newcommand\Red[1]{{\color{Red}#1}}
+% \newcommand\Blue[1]{{\color{Blue}#1}}
+% \newcommand\DarkBlue[1]{{\color{DarkBlue}#1}}
+% \newcommand\LightBlue[1]{{\color{LightBlue}#1}}
+% \newcommand\Brown[1]{{\color{Brown}#1}}
+% \newcommand\Green[1]{{\color{Green}#1}}
+% \newcommand\SeaGreen[1]{{\color{SeaGreen}#1}}
+% \newcommand\Yellow[1]{{\color{yellow}#1}}
+% \newcommand\Orange[1]{{\color{orange}#1}}
+% 
+% \newcommand\nn{\nonumber\\}
+% 
+% \newcommand\nmat[1]{\begin{matrix}#1\end{matrix}}
+% \newcommand\bmat[1]{\begin{bmatrix}#1\end{bmatrix}}
+% \newcommand\case[1]{\begin{cases}#1\end{cases}}
+% \newcommand\textbox[2]{\footnotesize\text{\parbox{#1}{\centering\emph{#2}}}}
+% 
+% \newcommand\rand{\text{rand}}
+% \newcommand\randn{\text{randn}}
+% \newcommand\rect{\text{rect}}
+% \newcommand\sinc{\text{sinc}}
+% \newcommand\tr{\text{tr}}
+% \newcommand\adj{\text{adj}}
+% 
+% % \newcommand\max{\text{max}}
+% \newcommand\argmin{\text{argmin}}
+% 
+% \newcommand\qqquad{\quad\qquad}
+% \newcommand\qqqquad{\qquad\qquad}
+% 
+% \renewcommand\l[1]{\left#1}
+% \renewcommand\r[1]{\right#1}
+% 
+% % {\text{\parbox{1.5cm}{\centering volume hyper- sphere}}}
+% 
+% %Keyword colouring:
+% \newcommand\kw[1]{#1}
+% \newcommand\parm[1]{#1}%\color{Black}#1\color{Black}}
+% 
+% \newcommand\of[1]{\scriptstyle(\parm{#1})\displaystyle}
+% \newcommand\df[1]{\scriptstyle[\parm{#1}]\displaystyle}
+% \newcommand\var[3]{#1_\text{#2}\of{#3}}
+% 
+% \newcommand\diag{\text{diag}}
+% 
+% % \raisebox{lift}[extend-above-baseline][extend-below-baseline]{text}
+% \newcommand\mt[1]{\text{\emph{#1}}} %mt = mathtext
+% \newcommand\mathnorm{\textstyle}
+% \newcommand\mathbig[1]{\displaystyle#1\mathnorm}
+% \newcommand\mathsmall[1]{\scriptstyle#1\mathnorm}
+% \newcommand\mathtiny[1]{\scriptscriptstyle#1\mathnorm}
+% \newcommand\sfrac[2]{\scriptstyle\raisebox{0.25pt}[0pt][0pt]{$\frac{#1}{#2}$}\mathnorm}
+% \newcommand\nfrac[2]{\textstyle\frac{#1}{#2}\displaystyle}
+% 
+% \newcommand\sumu[1]{\sum\limits^{#1}\,}
+% \newcommand\suml[1]{\sum\limits_{#1}\,}
+% \newcommand\sumb[2]{\sum\limits_{#1}^{#2}\,}
+% 
+% \newcommand\produ[1]{\prod\limits^{#1}\,}
+% \newcommand\prodl[1]{\prod\limits_{#1}\,}
+% \newcommand\prodb[2]{\prod\limits_{#1}^{#2}\,}
+% 
+% %Math macros:
+% \newcommand\diff[2]{\frac{\kw{d}\,\textstyle #1\scriptstyle}{\kw{d\parm{#2}}}\displaystyle}
+% \newcommand\ddiff[2]{\frac{\kw{d^2}\,\displaystyle #1\scriptstyle}{\kw{d\parm{#2}}^2}\displaystyle}
+% 
+% \renewcommand\d[1]{\scriptstyle\kw{\,d\parm{#1}}\displaystyle}
+% 
+% % These commands are mutually exclusive. Remember to "renew" in v2.
+% \newcommand\intb[4]{\int\limits_{#3}^{#4} #1 \d{#2}} % \int{exp}{var}{from}{to}
+% \newcommand\intl[3]{\int\limits_{#3} #1 \d{#2}} % \int{exp}{var}{for all}
+% \newcommand\intu[2]{\int #1 \d{#2}} % \int{exp}{var}{for all}
+% 
+% \newcommand\T{^{\scriptscriptstyle T}}
+% \renewcommand\H{^{\scriptscriptstyle H}}
+% 
+% \renewcommand\vec[1]{\boldsymbol{#1}}
+% \newcommand\mat[1]{\boldsymbol{#1}}
+% 
+% 
+% \renewcommand*\a{\vec a}
+% \renewcommand*\i{\vec i}
+% \renewcommand*\k{\vec k}
+% \newcommand*\n{\vec n}
+% \newcommand*\p{\vec p}
+% \newcommand*\s{\vec s}
+% \newcommand*\w{\vec w}
+% \newcommand*\x{\vec x}
+% \newcommand*\y{\vec y}
+% 
+% \newcommand*\A{\mat A}
+% \newcommand*\B{\mat B}
+% \newcommand*\C{\mat C}
+% \newcommand*\E{\mat E}
+% % \renewcommand*\H{\mat H}
+% \renewcommand*\P{\mat P}
+% \newcommand*\eP{\mat{\hat P}}
+% \newcommand*\R{\mat R}
+% \newcommand*\Ri{\R^{-1}}
+% \newcommand*\eR{\mat{\hat R}}
+% \newcommand*\W{\mat W}
+% \newcommand*\X{\mat X}
+% \newcommand*\Xd{\X_{\!\Delta}}
+% \newcommand*\Y{\mat Y}
+% 
+% \renewcommand*\L{\mat \Lambda}
+% \newcommand*\U{\mat U}
+% % \renewcommand*\t{\mathtiny{^T}}
+% % \newcommand*\h{\mathtiny{^H}}
+% \renewcommand*\t{^T}
+% \newcommand*\h{^H}
+% 
+% \newcommand\D{\vec\nabla} %Del: Vector differential operator - nabla
+% \newcommand\Dx{\vec\nabla\times}
+% \newcommand\Dd{\vec\nabla\cdot}
+% 
+% \usepackage{tikz}
+% \usetikzlibrary{shapes,snakes}
+% \usepackage{amsmath,amssymb}
+% 
+% \newenvironment{outline}
+% {\begin{itemize}}
+% {\end{itemize}}
 
@@
 
+The minimum variance distortionless response (NVDR) beamformer is a popular 
+Acoustic imaging systems can often benefit from adapting their behaviour situation at hand. When the data is received with a phased array.
+
+Most modern acoustic imaging methods rely on the data from a phased array
+
 \begin{itemize}
@@
 \IEEEpeerreviewmaketitle
+
+\begin{abstract}
+The angular resolution and contrast in active sonar images depend on the beamformer's ability to receive signals from directions of interest, while suppressing noise and interference emanating from other directions.
+For sonar arrays, this is achieved by applying weights to the array channels.
+
+While classical beamformers use predefined windows, adaptive beamformers estimate the optimal window by analytical evaluation of the data.
+The minimum variance (MV) beamformer, for instance, calculates the set of weights that minimises the variance of the beamformer's output.
+
+We have implemented a Low Complexity Adaptive (LCA) beamformer, which adaptively selects a window from a predefined set.
+The set is comprised of windows that are typical solutions found by the Minimum Variance method.
+The LCA beamformer was tested using simulated and experimental data from the Kongsberg Maritime HISAS 1030 sonar.
+On a simulated scene with speckle, highlight and shadow, the beamformer offered better lateral edge definition compared to the MV beamformer, and speckle intensity and shape comparable to DAS and MV beamformers.
+These results were verified by the experimental data.
+
+An attactive trait of the LCA is its low computational complexity; while the MV beamformer is of O($M^3$), the LCA method is of O($MW$), with $M$ being the number of channels and $W$ the number of windows.
+We made the LCA perform like the MV method using a well designed set of 30 windows.
+Hence, unless the array is very small, the proposed method will perform like the MV beamformer or better, and at a fraction of the computational cost.
+\end{abstract}%
+%
+\keywords{Beamforming, Capon, Minimum variance, Low Complexity, Sonar}
+
+% \titlepage
+
+\newpage
+% \the\lineheight
+\section{Introduction}
+
+In active sonar array imaging, an acoustic wave is transmitted and the received echoes are usually recorded using an array of sensors.
+Each of these may be independently delayed and weighted such that signals emanating from directions of interest are added constructively, while noise and interference from other directions add destructively.
+Additionally, a window may be applied to the sensor channels to further adjust the arrays spatial response.
+This process is known as beamforming.% The process of combining the signals from each sensor is known as beamforming.
+
+Traditional beamformers such as Delay-and-Sum (DAS) weight the sensors using a predefined window.
+However, since the optimal spatial response is likely to change with time, it is often preferred to use beamformers that continously adapt the weights to the dynamic data.
+This usually involves solving some optimization problem.
+
+The Minimum Variance (MV) beamformer selects the window that minimizes the beamformer's output power~\cite{cap69}.
+It has recently been investigated for active sonar imaging in e.g.~\cite{saf09}, and in ultrasound imaging in e.g.~\cite{syn07}.
+We suggest using a Low Complexity Adaptive (LCA) beamformer, which for each pixel applies a set of predefined windows and selects one of them based on the MV criterion.
+Using the LCA beamformer results in increased robustness and improved edge definition compared to the MV and DAS beamformers.
+
+% When the wavefield is sampled using an array of sensors, the data from each sensor may be 
+% 
+% Traditional beamformers such as Delay-and-Sum (DAS) apply predefined windows to all incoming data.
+However, due to the non-stationary nature of sonar data, the optimal window for any given time instant will generally differ from the next.
+This is where adaptive beamformers thrive, because they compute the optimal window coefficients for the data at each time instant.
+The choice of optimisation criteria is what mainly differentiates the various adaptive beamformers.
+The Minimum Variance (MV) beamformer, for instance, selects the set of weights that minimizes the beamformer output power for any given time instant~\cite{cap69}.
+
+\section{Background}
+
+% Adaptive beamformers usually compute the window coefficients by estimating and inverting a spatial covariance matrix.
+There are two inherent problems to this procedure.
+First, for improved estimation accuracy the data is either averaged in space or time, and the significance of the covariance matrix can be reduced using a regularisation parameter~\cite{Carl Inge}.
+These are all attempts to constrain the beamformer to ensure the window coefficience are not over-adapted to the data.
+Second, inverting the covariance matrix has a computational complexity of O($M^3$)~\cite{Carl Inge}.
+For larger arrays the computational burden becomes significant, and the question arises whether this processing power can be better utilised.
+
+The MV beamformer computes the window coefficients by estimating and inverting a spatial covariance matrix.
+For improved estimation accuracy the data is averaged in space or time, or both.
+In addition, a certain amount of diagonal loading is often added to the covariance matrix before inversion.
+These actions, while leading to statistical robustness, tend to deteriorate the beamformer's performance.
+Also, the inversion of the covariance matrix has a computational complexity of O($M^3$), where $M$ is the number of elements.
+The computational burden makes the MV beamformer for large arrays less attractive, and sometimes even infeasible.
+
+Based on the proposed method by Synnevåg~\cite{syn11}, we have implemented the LCA beamformer that keeps the minimum variance optimisation criterion but reduces the solution space to a discrete set of predefined windows.
+This reduces the computational complexity to O($MW$), where $W$ is the number of windows in the set.
+We use the Kaiser window function because it allows us to design a wide range of windows with different mainlobe widths and sidelobe suppression by adjusting the tradeoff parameter $\beta$~\cite{kai66}.
+In addition we apply steering to each of the windows and constrain the window design to ensure unit gain in the look direction.
+
+The following will show that the proposed method performs similarly to or better than the MV method.
+
+% \begin{figure}[p]\centering%
+% \subfloat[Image of a single speckle realisation.]{%
+% \includegraphics[width=\linewidth]{gfx/img_das_lca_mv.pdf}%
+% \label{speckleA}}%
+% % \hfil%
+% \subfloat[Mean image of 100 speckle realisations.]{%
+% \includegraphics[width=\linewidth]{gfx/img_das_lca_mv_mean100.pdf}%
+% \label{speckleB}}%
+% \caption{A simulated circular object buried in speckle is imaged from above.}\label{speckle}
+% \end{figure}
+
+% 
+% 
+% \begin{figure}[tp]\centering%
+% \subfigure{%
+% \graphicsAI[width=\linewidth]{gfx/img_das_lca_mv.pdf}}
+% \subfigure{%
+% \graphicsAI[width=\linewidth]{gfx/img_das_lca_mv_mean100.pdf}}%
+% \caption{A simulated circular object buried in speckle is imaged from above.\newline\parbox[t]{1.50cm}{Top:\newline Bottom:}\parbox[t]{6.8cm}{Image of a single speckle realisation.\newline Mean image of 100 speckle realisations.}}\label{speckle}%%
+% \end{figure}%
+% 
+% \begin{figure}[t]\centering%
+% \graphicsAI[width=\linewidth]{gfx/img_mean_cut.pdf}%
+% \caption{Lateral cut in the mean image at range 41.1\,m (highlight) and 44.5\,m (shadow).}\label{cutmean}%
+% \end{figure}
+% % \begin{figure}[t]\centering
+% % % \graphicsAI[drawing,width=\linewidth]{gfx/the_box.svg}\\
+% % \includegraphics[width=\linewidth]{gfx/selected_windows.pdf}
+% % \caption{Selected windows}\label{selected_windows}
+% % \end{figure}
+% % \begin{figure}[t]\centering
+% % \includegraphics[width=\linewidth]{gfx/selected_windows_angle.pdf}
+% % \caption{Selected windows - Colored by angle}\label{selected_windows_angle}
+% % \includegraphics[width=\linewidth]{gfx/selected_windows_sym_angle.pdf}
+% % \caption{Selected windows - Colored symmetrically by angle}\label{selected_windows_sym}
+% % \end{figure}
+% \begin{figure}[t]\centering
+% \graphicsAI[width=\linewidth]{gfx/img_holmengraa.pdf}%
+% \caption{HISAS sidescan sonar (SSS) image of the shipwreck Holmengraa.}\label{holmengraa}%
+% \end{figure}
+
+% Adaptive beamformers usually compute the window coefficients by estimating and inverting a spatial covariance matrix.
+There are two inherent problems to this procedure.
+First, for improved estimation accuracy the data is either averaged in space or time, and the significance of the covariance matrix can be reduced using a regularisation parameter~\cite{Carl Inge}.
+These are all attempts to constrain the beamformer to ensure the window coefficience are not over-adapted to the data.
+Second, inverting the covariance matrix has a computational complexity of O($M^3$)~\cite{Carl Inge}.
+For larger arrays the computational burden becomes significant, and the question arises whether this processing power can be better utilised.
+% 
+% Based on the proposed method by Synnevåg~\cite{syn11}, we have implemented a Low Complexity Adaptive (LCA) beamformer that keeps the minimum variance optimisation criteria but reduces the solution space to a discrete set of predefined windows.
+We use the Kaiser window function because it allows us to design a wide range of windows with different mainlobe widths and sidelobe suppression by adjusting the tradeoff parameter $\beta$.
+In addition we apply steering $\varphi$ to each of the windows and and constrain the window design to ensure unit gain in the look direction.
+
+\section{Experimental Setup}
+
+For the LCA to perform well it is important to let it have a diverse selection of windows to choose from.
+This selection is based on observations of the solutions found by the MV beamformer.
+We obtained good results in our experiments by letting the LCA beamformer select among 5 Kaiser windows with uniformly distributed $\beta$'s in the range $[0.05, 0.5]$ and a rectangular window.
+Each window was steered in 5 different directions, uniformly distributed within 80\% of the mainlobe width of the rectangular window, which is the most narrow.
+This adds to a total of 30 windows.
+
+To make the choice of window less susceptible to pixel value uncertainties, the LCA beamformer was set to apply the most frequently selected window in an 11 pixel range region to the center pixel in that region.
+In physical terms, this is a region of approximately 20\,cm.
+The MV beamformer estimated the covariance matrix as described in~\cite{syn07} by averaging over 16 subarrays and 11 range pixels, and applying 3\% diagonal loading.
+
+% Furthermore, the algorithm maps well to single-instruction-multiple-data (SIMD) hardware such as graphics processing units (GPUs).
+We show that even a modest amount of optimisation work has resulted in a factor 10 speed-up of this algorithm when implemented on a GPU as opposed to on a CPU.
+
+% \newpage
+\section{Results and Discussion}
+
+We have tested the beamformers using data aquired by the Kongsberg Maritime HISAS1030 sonar~\cite{han09}, and on simulations of the same sonar with 100 realisations of speckle.
+
+The simulations contained a few highlighted cylindric regions with 2\,m diameter and a constant intensity 15.4\,dB over the average speckle level.
+We focused on an object centered at 41\,m range and -3 degrees azimuth angle.
+The images produced by the LCA, MV and unweighted DAS for a single speckle realisation and a mean image of all 100 realisations are shown in \Fig{speckle}.
+Each image was normalised by their respective speckle level, and the dynamic range was clamped at \{-30,~15\}\,dB.
+The physical extent of the of the simulated object was superimposed for reference.
+%\ref{saf09}
+
+In \Fig{speckle} we observe that both the adaptive beamformers produce images with a more clearly defined shadow.
+The edges are sharper and the shadow more distinct.
+This is because the adaptive beamformers have better sidelobe suppression, and thus allow less signal energy to leak from the speckle region into the shadow region.
+The same effects can also be observed laterally around the highlight region of the mean image; DAS is more prone to smear energy into the speckle region than the adaptive beamformers are.
+
+\Fig{cutmean} displays two lateral cuts in the mean image, one through the highlight at 41.1\,m range and one through the shadow at 44.5\,m range.
+Each cut was computed as the mean of a 40\,cm range band centered at the respective ranges.
+The transition region between highlight and shadow is shortest for the LCA, which is advantageous since it results in a more accurate representation of the object size.
+The slightly inferior performance of the MV beamformer is due to the subarray averaging, which makes the effective array smaller~\cite{syn07}.
+
+By studying the LCA performance on the 100 speckle realisations, we found the choice of window to be highly dependent on the phenomena being imaged.
+In speckle regions LCA favored narrow spatial responses with minor steering, while wide and fully steered responses were often selected in shadow regions.
+There was also a clear distinction between the responses selected in sidelobe regions and in speckle regions.
+
+% \Fig{hist} illustrates how often a particular window is chosen in general.
+There are 10 $\phi$'s for each of the 10 $\beta$'s.
+The first 10 values are rectangular windows.
+Value 5 is a non-steered rectangular window, and we see that this is most frequently selected.
+At higher steering angles the windows are rarely selected, and thus we might get better results by tightening our steering boundaries.
+At value 15 we find inverted Kaiser windows with $\beta = 0.05$, and then $\beta$ is increased in increments of 0.505 for every 10th value.
+At value 65 the $\beta$ value is 2.025, and these and the remaining windows are hardly selected.
+
+A sidescan image of the 1500 dwt oil tanker wreck Holmengraa is shown in \Fig{holmengraa}.
+It is about 68\,m long and 9\,m wide, and lies on a slanted seabed at 77\,m depth~\cite{holmengraa}.
+The sidescan image was created using data from the HISAS 1030 sonar, which is rather unsuited for this purpose because of its large opening angle.
+This, and the fact that the wreck was imaged at a range of about 105\,m makes the image quality poor, but sufficient to compare our beamformers.
+
+In the Holmengraa image we note again that the LCA produces a cleaner shadow and better edge definition. The MV method performed almost identically to LCA in this case, and was omitted.
+
+\section{Conclusion}
+
+The LCA method performs similar to the MV method, but at a fraction of the computational cost.
+The key to achieve this success lies in the design of the window set.
+Windows that yield vagely steered narrow responses are preferred in speckle regions, while wide and steered responses are typically preferred in highlight and shadow regions.
+A window set of 6 different responses each steered in 5 different directions proved sufficient in our experiment to match the performance of the MV beamformer.
+
+The tendency of the LCA to perform similarly to the MV beamformer seems to indicate that a full solution space is rarely needed in real world scenarios.
 
@@
 
-\input{gfx/gfx_input}
+See end of document...
 % \begin{figure*}[!t]
@@
 
-% trigger a \newpage just before the given reference
-% number - used to balance the columns on the last page
-% adjust value as needed - may need to be readjusted if
-% the document is modified later
-%\IEEEtriggeratref{8}
-% The "triggered" command can be changed if desired:
-%\IEEEtriggercmd{\enlargethispage{-5in}}
-
-% references section
-
-\bibliographystyle{IEEEtran}
-\bibliography{../../Library/library}
-
-% End up doing this:
-% \begin{thebibliography}{1}
+% \bibliographystyle{IEEEtran}
+% \bibliography{references}
 % 
-% \bibitem{IEEEhowto:kopka}
-% H.~Kopka and P.~W. Daly, \emph{A Guide to {\LaTeX}}, 3rd~ed.\hskip 1em plus
-%   0.5em minus 0.4em\relax Harlow, England: Addison-Wesley, 1999.
 % 
-% \end{thebibliography}
-
-% biography section
-% 
-% If you have an EPS/PDF photo (graphicx package needed) extra braces are
-% needed around the contents of the optional argument to biography to prevent
-% the LaTeX parser from getting confused when it sees the complicated
-% \includegraphics command within an optional argument. (You could create
-% your own custom macro containing the \includegraphics command to make things
-% simpler here.)
-%\begin{biography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{mshell}}]{Michael Shell}
-% or if you just want to reserve a space for a photo:
+% \input{bio/jo_inge_buskenes}
+% % \input{bio/jon_petter_aasen}
+% % \input{bio/carl-inge_nilsen}
+% \input{bio/andreas_austeng}
 
@@
 
-\begin{IEEEbiography}{Noname dude}
-He doesn't exist.
-\end{IEEEbiography}
-
-% if you will not have a photo at all:
-\begin{IEEEbiographynophoto}{Jo Inge Buskenes}
-Not in at the moment.
-\end{IEEEbiographynophoto}
-
-%\newpage
-
-% \begin{IEEEbiographynophoto}{Jo Inge Buskenes}
-% Not in at the moment.
-% \end{IEEEbiographynophoto}
-
-%\vfill
-
-% Can be used to pull up biographies so that the bottom of the last one
-% is flush with the other column.
-%\enlargethispage{-5in}
-
 \end{document}

### 2014-11-26 ThesisJo/derivations.tex
la2pdf commit - 2014:11:24	14:53:56	(Mon24.Nov)
la2pdf commit - 2014:11:24	17:42:43	(Mon24.Nov)
la2pdf commit - 2014:11:24	18:32:43	(Mon24.Nov)
la2pdf commit - 2014:11:24	19:08:58	(Mon24.Nov)
la2pdf commit - 2014:11:26	19:01:05	(on.26.nov.)

@@
 Conventional limit of validity is $\frac{\pi}{2}$, which corresponds to $\sigma = \frac{\lambda}{8\cos\theta}$ and a coherent loss of 10.7dB.
+
+\section{LCA tricks}
+
+Let $\x$ be the data received by an array, pre-steered to some pixel in the image, and $\w$ be the window applied to these spatial data.
+Then we defined the beamformer outout $z$ as the weighted sum of the data:
+%
+\begin{align}
+z(\w,\x) = \w\H\x = \sumb{m=0}{M-1} w_m x_m^*
+\end{align}
+%
+Now let us choose a trigonometric window function:
+%
+\begin{align}
+\w = \frac{e^{j\varphi}}{2\pi\alpha}\big(\alpha + (1-\alpha)\cos\left(\frac{2\pi n}{N-1}\right)\big)
+\end{align}
+%
+Inserting this yields:
+%
+\begin{align}
+\w\H\x &= \sumb{m=0}{M-1} e^{j\varphi}\Big(\alpha + (1-\alpha)\cos\theta\Big) x_m^* \nn
+&= \sumb{m=0}{M-1} e^{j\varphi}\Big(\alpha + (1-\alpha)\cos\theta\Big) x_m^*
+\end{align}
+%
+where $\theta = \frac{2\pi n}{N-1}$.
+
+LCA can then be defined as
+%
+\begin{align}
+\argmin{\w}\ |z|^2 = \argmin{\w}\ z\,z^* = %\argmin{\w}\ \w\H\x = \sumb{m=0}{M-1} w_m^* x_m
+\end{align}
+
+\newpage
+\section{MVDR}
+
+By definition, the beamformer output $z[n]$ can now be expressed as the weighted sum of all the delayed data samples:
+\begin{align}
+z[n] = \w\H[n]\x[n] = \bmat{w_0[n]\\w_1[n]\\\vdots\\w_{M-1}[n]}^H \bmat{x_0[n]\\x_1[n]\\\vdots\\x_{M-1}[n]},\label{z}
+\end{align}
+where $w_m$ is the weight factor assigned to channel $m$.
+With static weights this would be referred to as the conventional delay-and-sum (DAS) beamformer.
+A large variety of weighting functions exists here for trading lateral resolution for improved noise suppression (contrast), but one always ends up with a compromise between the two~\cite{Harris1978}.
+
+Various adaptive beamformers target this limitation by allowing the weights to change for each pixel to better fit the dynamic nature of the incoming wavefield.
+In other words, they attempt to use the \emph{a priori} information present in the data to improve image quality.
+The MVDR beamformer is one such method.
+It finds the set of complex weights that minimizes the beamformer's expected output power, while ensuring unity gain in the look direction~\cite{Capon1969}.
+This is a convex optimization problem that can be solved using Lagrange multipliers to yield the solution
+\begin{gather}
+\vec w[n] = \frac{\Ri[n]\a}{\a\T\Ri[n]\a},\label{weights}
+\end{gather}
+where $\a$ is a steering vector and $\R=E\{\x[n]\x\H[n]\}\in\mathbb{C}^{M,M}$ is the spatial covariance matrix for the full array.
+Since we pre-steer our data to every pixel in the image we simplify (\ref{weights}) by substituting $\a$ with a row vector $\1$ that represents broadside phase-steering.
+To estimate $\R$ we compute a sample covariance matrix $\eR$.
+In this computation we perform some degree of:
+\begin{itemize}
+\item \emph{spatial averaging} to avoid signal cancellation by decorrelating coherent echoes~\cite{Kailath1985};
+\item \emph{temporal averaging} over an interval comparable to the pulse length (one to five samples) to maintain true speckle statistics~\cite{Synnevag2009a};
+\item \emph{diagonal loading} to improve robustness to parameter errors~\cite{Cox1987,Maksym1979}.
+\end{itemize}%

### 2014-12-09 ThesisJo/derivations.tex
la2pdf commit - 2014:11:26	19:44:01	(on.26.nov.)
More derivations
la2pdf commit - 2014:12:09	10:55:51	(Tue 9.Dec)

@@
 This is a convex optimization problem that can be solved using Lagrange multipliers to yield the solution
+%
 \begin{gather}
@@
 \end{gather}
+%
 where $\a$ is a steering vector and $\R=E\{\x[n]\x\H[n]\}\in\mathbb{C}^{M,M}$ is the spatial covariance matrix for the full array.
@@
 In this computation we perform some degree of:
+%
 \begin{itemize}
@@
 \end{itemize}%
+
+\subsection{Diagonal loading}
+
+Assume the data to be a combination of signal and noise, $\x = \s + \n = e^{j\theta[n]}$. Then the beamformer output is
+%
+\begin{align}
+z = \w\H\x = \w\H(\s+\n),
+\end{align}
+%
+and the output power becomes
+%
+\begin{align}
+|z|^2 = \w\H\x\x\H\w &= \w\H(\s+\n)(\s+\n)\H\w \nn
+&= \begin{cases}
+\w\H\s\s\H\w + \w\H\s\n\H\w + \w\H\n\s\H\w + \w\H\n\n\H\w &\text{when correlated}\\
+\w\H\s\s\H\w + \w\H\n\n\H\w &\text{when uncorrelated}
+\end{cases}
+\end{align}
+%
+Assuming that the data is predelayed broadside, and that the window is rectangular, we may set $\a = \s$ and get
+%
+\begin{align}
+|z|^2 = \w\H\s\s\H\w + \w\H\n\n\H\w
+\end{align}
+
+\section{Beamforming}
+
+\begin{align}
+z[n] = \w\H[n]\x[n]
+\end{align}

### 2014-12-19 ThesisJo/derivations.tex
la2pdf commit - 2014:12:10	12:08:06	(Wed10.Dec)
la2pdf commit - 2014:12:19	10:39:54	(Fri19.Dec)

@@
 \begin{align}
-s(\p,t) &= A\,e^{j(\omega t - \k\cdot\p)}
+s(\p,t) &= A\,e^{j(\omega t - \k\cdot\p)},
 \end{align}
@@
 %
-Hence, whenever we can assume the spatial and temporal frequency to be related directly by the sound speed, $|\k| = \frac{\omega}{c}$, then this assuming a complex exponential should be fair.
+This means that when we can assume the spatial and temporal frequency to be related directly by the sound speed, $|\k| = \frac{\omega}{c}$, then assuming a complex exponential to be a valid solution.
+Another solution of the wave equation is that of a spherical wave:
+%
+\begin{align}
+s(r,t) &= \frac{A}{r}\,e^{j(\omega t - kr)}.
+\end{align}
 
@@
 
+\newpage
 \section{Roughness}
@@
 
+\newpage
 \section{Beamforming}

### 2014-12-30 ThesisJo/introduction.tex
la2pdf commit - 2014:10:27	16:25:27	(Mon27.Oct)
Some KM presentation files added

@@
+The best processing unit we know of is the human mind.
+Its several billion neurons process information at a staggering rate, and concurrently.
+As described by this year's nobel price winners, memories of visual objects gets stored in a similar spacial structure in the brain.
+
+Similar to how the process visual input from our eyes and turn them into the images we perceive, we want turn echo from sonar systems into images of the physical objects the sound was scattered by.
+
+Want to compare acoustic array to the eyes.
+Similar, this images from sound should be possible.
+Brain parallel, and superior.
+Thus, processing power must be essential to the continued improvement of current sonar systems.
+But *how* should be additional processing power be used?
+
+Signal processing is the art of combining physics, mathematics and computational engineering
+
 Wonders happen every day. What our senses reveal our brain make real. This is the beauty of signal processing.
@@
 
+\chapter{Introduction}
+
 \section{The Wave Equation}
 
-% \begin{align*}
-% \nabla^2 g(\p,t) - \frac{1}{c^2(\p)}
-% \end{align*}
+\begin{align*}
+\nabla^2 f(\p,t) - \frac{1}{c^2(\p)}\frac{\partial^2 f(\p,t)}{\partial t^2} = -s(t) \delta(\p-\p_0)
+\end{align*}
 
-% where $\nabla^2$ is the Laplacian operator
+where $\nabla^2$ is the Laplacian operator
 

### 2015-02-02 ThesisJo/derivations.tex
la2pdf commit - 2015:02:02	14:08:17	(Mon 2.Feb)
la2pdf commit - 2015:02:02	15:11:14	(Mon 2.Feb)

@@
 
-\section{LCA tricks}
+\section{LCA}
 
-Let $\x$ be the data received by an array, pre-steered to some pixel in the image, and $\w$ be the window applied to these spatial data.
-Then we defined the beamformer outout $z$ as the weighted sum of the data:
+Let $\X[n]$ be the data received by an array, pre-steered to all pixels in the image, and $\w$ be the window applied to these spatial data.
+Then we defined the beamformer output $z$ as the weighted sum of the data:
 %
 \begin{align}
-z(\w,\x) = \w\H\x = \sumb{m=0}{M-1} w_m x_m^*
+z[\w,n] = \w\H\X[n] = \sumb{m=0}{M-1} w_m x_m^*[n]
 \end{align}
 %
+From a set of $P$ windows the LCA beamformer selects the window $p$ that yield the least output power:
+%
+\begin{align}
+\argmin{p}{E\big\{|z|^2\big\}} = \argmin{p}{E\big\{|\w_p\H\X[n]|^2\big\}} \qquad\text{subject to}\qquad \w\H\a = 1
+\end{align}
+
+\subsection{Tricks}
+
 Now let us choose a trigonometric window function:

### 2015-02-26 Simulator/article.tex
Changed simulator title
la2pdf commit - 2015:02:26	18:05:17	(Thu26.Feb)

--- 2014-08-22 Simulator/article.tex
@@
 % \usepackage{caption}
-\usepackage{subcaption}
+% \usepackage{subcaption}
 \usepackage{color}
@@
 
-\title{A GPU Sonar Simulator for Automatic Target Recognition}
+\title{A Real-Time Sonar Simulator for Automatic Target Recognition}
 
@@
 
-Automatic target recognition (ATR) is an important component in autonomous underwater vehicles (AUVs), as it allows the vehicle to adapt its mission plan to e.g.
+One of the major shortcomings of autonomous underwater vehicles (AUVs) is their lack of curiosity.
+When encountering an interesting object, an AUV with a mounted sonar imaging system should be able to detect it, and perhaps move in for closer examination.
+Besides the apparent challenge of trying to detect something with a long list of unknown parameters, this must happen in real-time and be very power efficient.
+
+% Automatic target recognition (ATR) is an important component in autonomous underwater vehicles (AUVs), as it allows the vehicle to adapt its mission plan to e.g.
 revisit detected objects for closer examination.
@@
 
-The challenge is to come up with accurate template images covering the actual configurations of object and seabed.
+The detection process is well described in the field of automatic target recognition (ATR).
+The classical approach involves isolating an object of interest, extracting parameters from it and feeding those into a classifier.
+However, it is hard to find parameters that describe the objects uniquely, in particular due to the ambiguities arising from interpreting a 3D world from its 2D projection.
+Another approach is to use template matching, where the image segment is compared to a set of template images covering the relevant object classes, and assigned to the class of the template yielding the the best fit.
+The challenge here is to come up with accurate template images covering the actual configurations of object and seabed, especially if a static template library is used, which will lead to a computational complexity that is exponential with the number of model parameters.
+This limits its use to only a few parameters being coarsely sampled, ultimately yielding inaccurate results~\cite{Midelfart2010}.
 
-Estimate 2D template directly
-  ->
+We have developed a synthetic aperture sonar (SAS) simulator that avoids this problem by running sufficiently fast to create templates adapted to the actual scene as part of the classification process.
+This is achieved with the aid of the massive computing power in graphics processing units (GPUs) and optimized software libraries for scene rendering.
+The simulator loads a 3D model of the seafloor and an object class model into OpenGL, where emitted sound waves are modeled with a light tube extending all the way down the propagation path of the AUV.
+We assume rough, isotropic surfaces reflecting sound energy equally in all directions.
+This can be modeled with a Lambertian scattering model~\cite{Blake1993,Bell1995}, where the intensity of the backscattered sound solely depends on the incidence angle of the transmitted signal onto the model surface.
+When rendering OpenGL is set up to produce an optical 2D intensity image and depth map that reveals the distance from each pixel to the propagation axis.
+For maximum flexibility this data is finally combined with OpenCL to produce the image templates.
 

### 2015-02-26 ThesisJo/introduction.tex
la2pdf commit - 2015:02:26	17:29:41	(Thu26.Feb)
la2pdf commit - 2015:02:26	18:05:17	(Thu26.Feb)
la2pdf commit - 2015:02:26	19:38:45	(Thu26.Feb)
la2pdf commit - 2015:02:26	20:09:33	(Thu26.Feb)

@@
 
-\chapter{Introduction}
+\chapter{Background}
+
+\section{Coordinate System}
+
+\begin{figure}[th]
+\graphicsAI[width=\linewidth]{gfx/geometry.svg}
+\caption{The coordinate system used throughout the thesis.}\label{geometry}
+\end{figure}
+
+In this thesis we will relate all geometrical interpretations to the euclidean, spherical coordinate system shown in \Fig{geometry}.
+When assessing the imaging performance of a sonar its origin will be in the center of the array, but when assessing scenes its origin will be in the center of the image.
+A unit length vector $\vec u$ can be written as:
+%
+\begin{align*}
+\vec u =
+\begin{bmatrix}
+\sin\theta \cos\phi \\
+\sin\theta \sin\phi \\
+\cos\theta
+\end{bmatrix}
+\end{align*}
+%
 
 \section{The Wave Equation}
+
+The wave equation is the single most important equation governing a wave's propagation through a medium. When the medium is loss-less it can be expressed as:
 

### 2015-03-02 ThesisJo/introduction.tex
la2pdf commit - 2015:03:02	13:41:35	(Mon 2.Mar)
la2pdf commit - 2015:03:02	14:36:38	(Mon 2.Mar)

@@
 
+When a rock is thrown into a pound of water it abruptly displaces water at the location of impact.
+This is an unstable state where various physical forces compete for order.
+Gravity wants the water back where it wants, but due to inertia the mass will resist any change in its motion.
+The wrestling of these forces result in a spherical wave propagating from the displacement centre.
+
 The wave equation is the single most important equation governing a wave's propagation through a medium. When the medium is loss-less it can be expressed as:
@@
 
-where $\nabla^2$ is the Laplacian operator
+where $\nabla^2$ is the Laplacian operator. The wave equation constrains the magnitude of the wavenumber,
+%
+\begin{align*}
+\vec k = \frac{\omega}{c} = \frac{2\pi}{\lambda},
+\end{align*}
+%
+so only the direction of $\k$ varies.
 

### 2015-03-02 ThesisJo/introduction.tex
la2pdf commit - 2015:03:02	16:50:06	(Mon 2.Mar)

@@
 When a rock is thrown into a pound of water it abruptly displaces water at the location of impact.
-This is an unstable state where various physical forces compete for order.
-Gravity wants the water back where it wants, but due to inertia the mass will resist any change in its motion.
-The wrestling of these forces result in a spherical wave propagating from the displacement centre.
+%This is an unstable state where various physical phenomena wrestle for control. Gravity wants the water back where it was, inertia cause the mass to resist any change in its state of motion.
+This results in spherical waves propagating from the displacement center.
+On the surface is a wave that has particles moving in circles relative to the propagation direction, while in the depths particles will move parallel to the propagation direction, alternately being compressed and spread apart.
+Sound is pressure waves.
+
+Unlike electromagnetic waves, these
 

### 2015-03-03 ThesisJo/introduction.tex
la2pdf commit - 2015:03:03	11:57:24	(Tue 3.Mar)

--- 2015-02-26 ThesisJo/introduction.tex
@@
 
+Light fast, sound slow.
+Sound propagate far, light attenuated.
+
 \section{Coordinate System}
@@
 
+When a rock is thrown into a pound of water it abruptly displaces water away from the center of impact. 
+%This is an unstable state where various physical phenomena wrestle for control. Gravity wants the water back where it was, inertia cause the mass to resist any change in its state of motion.
+This results in spherical waves propagating from the displacement center, transporting the kinetic energy of the rock.
+These waves are described by how their particles move.
+At the surface particles will move down up and down, and back and forth, in a circular motion.
+In the depths particles move longitudinally, transporting energy by means of alternating pressure.
+
 The wave equation is the single most important equation governing a wave's propagation through a medium. When the medium is loss-less it can be expressed as:
@@
 
-where $\nabla^2$ is the Laplacian operator
+where $\nabla^2$ is the Laplacian operator.
+
+The wave equation constrains the magnitude of the wavenumber,
+%
+\begin{align*}
+\vec k = \frac{\omega}{c} = \frac{2\pi}{\lambda},
+\end{align*}
+%
+so only the direction of $\k$ varies.
 

### 2015-03-03 ThesisJo/background.tex
la2pdf commit - 2015:03:03	14:50:18	(Tue 3.Mar)

--- 2015-03-03 ThesisJo/introduction.tex
@@
-The best processing unit we know of is the human mind.
-Its several billion neurons process information at a staggering rate, and concurrently.
-As described by this year's nobel price winners, memories of visual objects gets stored in a similar spacial structure in the brain.
-
-Similar to how the process visual input from our eyes and turn them into the images we perceive, we want turn echo from sonar systems into images of the physical objects the sound was scattered by.
-
-Want to compare acoustic array to the eyes.
-Similar, this images from sound should be possible.
-Brain parallel, and superior.
-Thus, processing power must be essential to the continued improvement of current sonar systems.
-But *how* should be additional processing power be used?
-
-Signal processing is the art of combining physics, mathematics and computational engineering
-
-Wonders happen every day. What our senses reveal our brain make real. This is the beauty of signal processing.
-
-- The wonder of how the brain process images. Great information carrier.
-
-Electromagnetic radiation - poor range due to attenuation and scattering. Use sound instead. Inferior information carrier. Lower frequency and propagation speed.
-
-Sonar.
-
-- Animals use sound both for imaging and communication underwater.
-
-- Quality.
-Depends on e.g.
-sonar frequency, pulse length and directivity.
-Range resolution given (possibly pulse compressed) signal pulse width.
-Less data in along-track.
-resolution given on matched filtered pulse
-
-- Low PRF - long range. High PRF - short range, high along-track resolution. Propagation time large compared to  SAS, low PRF long antenna. Both long range and high along-track resolution.
-
-- Can use models for sidescan.
-Main difference improved along-track resolution.
-Existing models accurate under certain assumptions, but can not be used directly to form side-looking image.
-Model should include physical processes such as acoustic propagation, seabed reverberation and transducer characteristics.
-
-SAS: Range and frequency independent resolution.
-
-Performance:
-- AUV revisit. Better image quality, lower data collection price.
-
 \chapter{Background}
@@
 Sound propagate far, light attenuated.
+
+The aim of this chapter is to review the basic background material on which this thesis is founded.
 
@@
 
-\section{The Wave Equation}
+\section{Wave Propagation}
 

### 2015-03-04 ThesisJo/background.tex
la2pdf commit - 2015:03:04	10:09:38	(Wed 4.Mar)
la2pdf commit - 2015:03:04	10:53:03	(Wed 4.Mar)
la2pdf commit - 2015:03:04	11:24:34	(Wed 4.Mar)

@@
 
+MVDR/LCA:
+- Beamforming techniques.
+
+Simulator:
+- Scattering theory
+- Coordinate systems
+
+Common denominator:
+- Image progressing
+- High processing computing
+- GPUs
+
 Light fast, sound slow.
-Sound propagate far, light attenuated.
+Sound propagates far, light attenuated.
 
 The aim of this chapter is to review the basic background material on which this thesis is founded.
-
-\section{Coordinate System}
-
-\begin{figure}[th]
-\graphicsAI[width=\linewidth]{gfx/geometry.svg}
-\caption{The coordinate system used throughout the thesis.}\label{geometry}
-\end{figure}
-
-In this thesis we will relate all geometrical interpretations to the euclidean, spherical coordinate system shown in \Fig{geometry}.
-When assessing the imaging performance of a sonar its origin will be in the center of the array, but when assessing scenes its origin will be in the center of the image.
-A unit length vector $\vec u$ can be written as:
-%
-\begin{align*}
-\vec u =
-\begin{bmatrix}
-\sin\theta \cos\phi \\
-\sin\theta \sin\phi \\
-\cos\theta
-\end{bmatrix}
-\end{align*}
-%
 
@@
 
-\section{Synethic Aperture Sonar Simulation}
+\section{Array Processing}
+
+\begin{figure}[th]
+\graphicsAI[width=\linewidth]{gfx/geometry.svg}
+\caption{The coordinate system used throughout the thesis.}\label{geometry}
+\end{figure}
+
+Consider a monochromatic plane wave impinging on an $M$-element array (\Fig{geometry}). Let the position of sensor $m$ be $\p_m$, th  of sensors at locations 
+In this thesis we will relate all geometrical interpretations to the euclidean, spherical coordinate system shown in \Fig{geometry}.
+When assessing the imaging performance of a sonar its origin will be in the center of the array, but when assessing scenes its origin will be in the center of the image.
+A unit length vector $\vec u$ can be written as:
+%
+\begin{align*}
+\vec u =
+\begin{bmatrix}
+\sin\theta \cos\phi \\
+\sin\theta \sin\phi \\
+\cos\theta
+\end{bmatrix}
+\end{align*}
+%
+
+Signal model. Monochromatic, plane wave:
+%
+\begin{align}
+\s[n] = \begin{bmatrix}
+         s_0[n] \\ s_1[n] \\ \vdots \\ s_{M-1}[n]
+        \end{bmatrix}
+      = \begin{bmatrix}
+         Ae^{j(\Omega n - \k\T\p_0)} \\
+         Ae^{j(\Omega n - \k\T\p_1)} \\
+         \vdots \\
+         Ae^{j(\Omega n - \k\T\p_{M-1})}
+        \end{bmatrix}
+      = Ae^{j\Omega n}\a(\k)
+\end{align}
+%
+where we call $\a(\k)$ the signal's \emph{steering vector}.
+
+\section{Synethic Aperture Sonar}
 

### 2015-03-04 ThesisJo/derivations.tex
la2pdf commit - 2015:02:03	12:29:14	(Tue 3.Feb)
la2pdf commit - 2015:02:03	14:24:40	(Tue 3.Feb)
la2pdf commit - 2015:02:26	14:27:18	(Thu26.Feb)
la2pdf commit - 2015:03:04	14:13:58	(Wed 4.Mar)

@@
 %
-This is called the homogeneous lossless wave equation, because there is no source term. The inhomogeneous equation can be obtained in a similar way, by altering the linearized continuity equation to:
+This is called the homogeneous lossless wave equation, because there is no source term.
+When our signal $s$ signifies pressure (N/m$^2$), this links time and space by means of mass flow rate (kg/(m$^3$s).
+The inhomogeneous equation can be obtained in a similar way, by altering the linearized continuity equation to include a source:
 %
@@
 \end{align}
+%
+We can estimate $E\big\{|\w_p\H\X[n]|^2\big\}$ as
+%
+\begin{align*}
+\sigma^2_{z_p}[n] = \frac{1}{N_k} \sumb{n'=n-K}{n+K} | z_p[n'] |^2
+\end{align*}
+%
+To add some extra flexibility, we can also estimate the variance by a weighted combination of local pixels
+%
+\begin{align*}
+\sigma^2_{z_p}[x,n] = \frac{1}{N_x N_k} \sumb{x'=x-X}{x+X} \sumb{n=n-K}{n+K} w_\text{prf}[x',n']\big| z_p[x',n'] \big|^2
+\end{align*}
+%
+where $w_\text{prf}[x',n']$ is a normalized 2 dimensional weight function.
+
+Kaiser:
+%
+\begin{align*}
+w[n] = \frac{I_0\big( \pi\alpha\sqrt{1-(\frac{2n}{N-1}-1)^2}\big)}{I_0(\pi\alpha)}
+\end{align*}
+%
+where $I_0$ is a zeroth order modified Bessel function of the first kind.
 

### 2015-03-05 ThesisJo/derivations.tex
la2pdf commit - 2015:03:05	14:50:56	(Thu 5.Mar)

@@
 When our signal $s$ signifies pressure (N/m$^2$), this links time and space by means of mass flow rate (kg/(m$^3$s).
-The inhomogeneous equation can be obtained in a similar way, by altering the linearized continuity equation to include a source:
+
+The inhomogeneous equation can be obtained in a similar way by including a source term.
+If we model the source as an injection of a volume of fluid, we can include it in the linearized continuity equation:
 %
@@
 %
-where $\rho_0f(\p,t)$ is the rate of mass change (in $\left[\frac{\text{kg}}{\text{m}^3\text{s}}\right]$) caused by the source. Following the exact same procedure as before we obtain:
+where $\rho_0f(\p,t)$ is the rate of mass change (in $\left[\frac{\text{kg}}{\text{m}^3\text{s}}\right]$) caused by the mass injection. Following the exact same procedure as before we obtain:
 %
@@
 \begin{align}
-\w = \frac{e^{j\varphi}}{2\pi\alpha}\big(\alpha + (1-\alpha)\cos\left(\frac{2\pi n}{N-1}\right)\big)
+\w = \frac{e^{j\varphi}}{2\pi\alpha}\big(\alpha + (1-\alpha)\cos\left(\frac{2\pi m}{M-1}\right)\big)
 \end{align}
@@
 \begin{align}
-\w\H\x &= \sumb{m=0}{M-1} e^{j\varphi}\Big(\alpha + (1-\alpha)\cos\theta\Big) x_m^* \nn
-&= \sumb{m=0}{M-1} e^{j\varphi}\Big(\alpha + (1-\alpha)\cos\theta\Big) x_m^*
+\w\H\x &= \sumb{m=0}{M-1} e^{j\varphi}\Big(\alpha + (1-\alpha)\cos\theta[n]\Big) x_m^* \nn
+&= \sumb{m=0}{M-1} e^{j\varphi}\Big(\alpha + (1-\alpha)\cos\theta[n]\Big) x_m^*
 \end{align}

### 2015-03-06 ThesisJo/derivations.tex
la2pdf commit - 2015:03:05	16:22:39	(Thu 5.Mar)
la2pdf commit - 2015:03:05	21:14:23	(Thu 5.Mar)
la2pdf commit - 2015:03:05	23:18:15	(Thu 5.Mar)
la2pdf commit - 2015:03:06	09:49:54	(Fri 6.Mar)
la2pdf commit - 2015:03:06	10:20:16	(Fri 6.Mar)
la2pdf commit - 2015:03:06	10:50:23	(Fri 6.Mar)
la2pdf commit - 2015:03:06	12:17:54	(Fri 6.Mar)
la2pdf commit - 2015:03:06	16:37:17	(Fri 6.Mar)

@@
 \begin{align}
-\w = \frac{e^{j\varphi}}{2\pi\alpha}\big(\alpha + (1-\alpha)\cos\left(\frac{2\pi m}{M-1}\right)\big)
+\w = e^{j\phi_\theta[m]}\frac{1}{2\pi\alpha?}\big(\alpha + (1-\alpha)\sin\left(\frac{2\pi m}{M-1}\right)\big)
 \end{align}
 %
-Inserting this yields:
+where $\phi_\theta[m] = \frac{2\pi}{\lambda}dm\sin\theta$. Inserting this yields:
 %
 \begin{align}
-\w\H\x &= \sumb{m=0}{M-1} e^{j\varphi}\Big(\alpha + (1-\alpha)\cos\theta[n]\Big) x_m^* \nn
-&= \sumb{m=0}{M-1} e^{j\varphi}\Big(\alpha + (1-\alpha)\cos\theta[n]\Big) x_m^*
+\w\H\x &= \sumb{m=0}{M-1} e^{j\phi_\theta[m]}\Big(\alpha + (1-\alpha)\sin\left(\frac{2\pi m}{M-1}\right)\Big) x_m^* \nn
+&= \alpha\sumb{m=0}{M-1} e^{j\phi_\theta[m]} x_m^* + (1-\alpha)\sumb{m=0}{M-1} e^{j\varphi[m]}\sin\left(\frac{2\pi m}{M-1}\right) x_m^* \\
+&= \alpha \a + (1-\alpha)\b
 \end{align}
 %
-where $\theta = \frac{2\pi n}{N-1}$.
+where:
+%
+\begin{align}
+a = \sumb{m=0}{M-1} e^{j\phi_\theta[m]}x_m^* \qquad \text{and} \qquad b_m = \sumb{m=0}{M-1} e^{j\phi_\theta[m]}\sin\left(\frac{2\pi m}{M-1}\right)x_m^*
+\end{align}
+%
+where
+%
+\begin{align}
+|\w\H\x|^2 &= \Big| \alpha a + (1-\alpha)b \Big|^2 \nn
+&= \Big[\alpha a + (1-\alpha)b\Big] \Big[\alpha a^* + (1-\alpha)b^*\Big] \nn
+&= \alpha^2 aa^* + \alpha a(1-\alpha)b^* + \alpha a^*(1-\alpha)b + (1-\alpha)^2bb^* \nn
+&= \alpha^2 aa^* + \alpha(1-\alpha)(ab^* + a^*b) + (1-\alpha)^2bb^* \nn
+\end{align}
+%
 

### 2015-03-06 ThesisJo/derivations.tex
la2pdf commit - 2015:03:06	20:01:38	(Fri 6.Mar)

@@
 &= \alpha^2 aa^* + \alpha(1-\alpha)(ab^* + a^*b) + (1-\alpha)^2bb^* \nn
+&= \alpha^2 (aa^* - (ab^* + a^*b) - bb^*) + \alpha(ab^* + a^*b) + bb^* \nn
 \end{align}

### 2015-03-13 ThesisJo/derivations.tex
la2pdf commit - 2015:03:10	15:06:05	(Tue10.Mar)
la2pdf commit - 2015:03:12	14:05:15	(Thu12.Mar)
la2pdf commit - 2015:03:12	17:16:15	(Thu12.Mar)
la2pdf commit - 2015:03:13	13:56:37	(Fri13.Mar)
la2pdf commit - 2015:03:13	14:29:03	(Fri13.Mar)

@@
 \begin{align}
-\w = e^{j\phi_\theta[m]}\frac{1}{2\pi\alpha?}\big(\alpha + (1-\alpha)\sin\left(\frac{2\pi m}{M-1}\right)\big)
+\w = e^{j\phi_\theta[m]}\big(\alpha - (1-\alpha)\cos\left(\frac{2\pi m}{M-1}\right)\big)
 \end{align}
@@
 \begin{align}
-\w\H\x &= \sumb{m=0}{M-1} e^{j\phi_\theta[m]}\Big(\alpha + (1-\alpha)\sin\left(\frac{2\pi m}{M-1}\right)\Big) x_m^* \nn
+\w\H\x &= \sumb{m=0}{M-1} e^{j\phi_\theta[m]}\Big(\alpha - (1-\alpha)\cos\left(\frac{2\pi m}{M-1}\right)\Big) x_m^* \nn
 &= \alpha\sumb{m=0}{M-1} e^{j\phi_\theta[m]} x_m^* + (1-\alpha)\sumb{m=0}{M-1} e^{j\varphi[m]}\sin\left(\frac{2\pi m}{M-1}\right) x_m^* \\
-&= \alpha \a + (1-\alpha)\b
+&= \alpha \a\T\x + (1-\alpha)\b\T\x
 \end{align}
@@
 \begin{align}
-a = \sumb{m=0}{M-1} e^{j\phi_\theta[m]}x_m^* \qquad \text{and} \qquad b_m = \sumb{m=0}{M-1} e^{j\phi_\theta[m]}\sin\left(\frac{2\pi m}{M-1}\right)x_m^*
+\a = \begin{bmatrix}
+     1 \\
+     e^{j\frac{2\pi}{\lambda}d\sin\theta} \\
+     e^{j\frac{2\pi}{\lambda}2d\sin\theta} \\
+     \vdots \\
+     e^{j\frac{2\pi}{\lambda}d(M-1)\sin\theta}
+     \end{bmatrix} \qquad \text{and} \qquad
+\b = \begin{bmatrix}
+     0 \\
+     e^{j\frac{2\pi}{\lambda}d\sin\theta}\sin\left(\frac{2\pi m}{M-1}\right) \\
+     e^{j\frac{2\pi}{\lambda}2d\sin\theta}\sin\left(\frac{2\pi m}{M-1}\right) \\
+     \vdots \\
+     0
+     \end{bmatrix}
 \end{align}
 %
+To attain unity gain in the look direction the weights must sum to one. 
+%
+\begin{align}
+\w\T\1 &= \big(\alpha\a - (1-\alpha)\b\big)\T\1 = 1
+\end{align}
+%
+This is true for any value of $\alpha$ if $\a\T\1 = 1$ and $\b\T\1 = 1$.
+
 where
@@
 %
+\begin{align}
+\frac{\partial}{\partial\alpha} |\w\H\x|^2 
+&= 2\alpha (aa^* - (ab^* + a^*b) - bb^*) + ab^* + a^*b = 0
+\end{align}
+%
+\begin{align}
+\alpha &= \frac{ab^* + a^*b}{2(aa^* - (ab^* + a^*b) - bb^*)}
+\end{align}
 
@@
 \end{align}
+
+\subsection{Element}
+
+The frequency response of an element is is given by the Fourier transform of the element's window function.
+%
+\begin{align}
+W(\k) = \int\limits_{V} w(\p)\,e^{-j\k\dot\p} dV
+\end{align}
+%
+where $\p = \bmat{x & y & z}\T$ and $dV = dx\,dy\,dz$. For a rectangular 2D element spanning the xy-plane, having sides with length $d_x$ and $d_y$, the frequency response becomes:
+%
+\begin{align}
+W_\text{rect}(k_x, k_y) = \frac{\sin\frac{k_x\,d_x}{2}}{\frac{k_x\,d_x}{2}}
+\end{align}
+%
+If we assume a rectangular

### 2015-03-13 ThesisJo/derivations.tex
la2pdf commit - 2015:03:13	14:59:07	(Fri13.Mar)
la2pdf commit - 2015:03:13	15:29:32	(Fri13.Mar)

@@
 \w\H\x &= \sumb{m=0}{M-1} e^{j\phi_\theta[m]}\Big(\alpha - (1-\alpha)\cos\left(\frac{2\pi m}{M-1}\right)\Big) x_m^* \nn
-&= \alpha\sumb{m=0}{M-1} e^{j\phi_\theta[m]} x_m^* + (1-\alpha)\sumb{m=0}{M-1} e^{j\varphi[m]}\sin\left(\frac{2\pi m}{M-1}\right) x_m^* \\
+&= \alpha\sumb{m=0}{M-1} e^{j\phi_\theta[m]} x_m^* - (1-\alpha)\sumb{m=0}{M-1} e^{j\varphi[m]}\cos\left(\frac{2\pi m}{M-1}\right) x_m^* \\
 &= \alpha \a\T\x + (1-\alpha)\b\T\x
@@
 \begin{align}
-W_\text{rect}(k_x, k_y) = \frac{\sin\frac{k_x\,d_x}{2}}{\frac{k_x\,d_x}{2}}
+W_\text{rect}(k_x, k_y) = \frac{\sin\frac{k_x\,d_x}{2}}{\frac{k_x\,d_x}{2}}\,\frac{\sin\frac{k_y\,d_y}{2}}{\frac{k_y\,d_y}{2}} = \sinc\frac{k_x\,d_x}{2}\,\sinc\frac{k_y\,d_y}{2}
+\end{align}
+
+\subsubsection{Opening angle and element size}
+
+What is the element size of the HISAS1030 sonar? This can be inferred from the element frequency response:
+%
+\begin{align}
+W_\text{rect}(k_x) = \frac{\sin\frac{k_x\,d_x}{2}}{\frac{k_x\,d_x}{2}} = \frac{1}{\sqrt{2}}
 \end{align}
 %
-If we assume a rectangular
+This can be solved for $\frac{k_x\,d_x}{2}$ either by series expansion of $\sin(\cdot)$, or numerically to find that $\frac{k_x\,d_x}{2} \approx 1.39$.
+For HISAS1030, with an opening angle of 25$^\circ$ at 100kHz in both directions, this means the element size must be quadratic with sides:
+%
+\begin{align}
+d = \frac{2*1.39}{k_x} = \frac{2*1.39}{2\pi \frac{f_c}{c}\sin\theta} = \frac{1.39*1500\text{ms}^{-1}}{\pi\, 100\text{kHz}\,\sin{12.5^\circ}} = 3.07\text{cm}
+\end{align}
+%
+This leaves a little space between each sensor since the element spacing is around 3.75cm.

### 2015-03-18 ThesisJo/derivations.tex
la2pdf commit - 2015:03:18	15:51:13	(Wed18.Mar)

@@
 
+\subsection{Oversampling and steering}
+
+Steering wildly is not a good idea.
+
+Too much steering, beamformer becomes too agressive. Uncertainty in source amplitude grows.
+
+\begin{itemize}
+\item More steering -> more aggressive suppression of interference at the cost of boosted sidelobes / noise.
+\item More steering -> Larger variations in amplitude of sources
+\end{itemize}
+
+Idea to steer within a fraction of the rectangular window's mainlobe.
+
+\begin{align*}
+d = \frac{\pi\,\theta_\text{open}\,\lambda}{M\,d\,O_s}
+\end{align*}
+
+HISAS1030:
+\begin{itemize}
+\item 25$^\circ$ opening angle.
+\item M$=32$ elements.
+\item 
+\end{itemize}
+
+- Retain source amplitude
+- Retain speckle statistics
+- Improve resolution
+- Easy to use
+
+Skifte til 1/4 av bredden
+
 \subsection{Tricks}

### 2015-04-14 LCAforActiveSonarImaging/lca.tex
More on LCA

@@
 
+Most modern active sonar imaging systems are usually far better at resolving details in range than in the angular direction.
+This is because it is relatively cheap and easy to collect a long and well sampled temporal representation of the incoming wavefield, but hard to do so spatially.
+It would require a very large array with a lot of sensors, which is costly, complex and impractical.
+Alternatively, a larger array can be synthesized from several successive pings recorded by a smaller physical array using a technique called synthetic aperture sonar (SAS).
+Unfortunately, not many of these systems exist due to the difficulty of navigating with sufficiently high precision under water.
+
+It would be more convenient to avoid 
+Hence, if the angular resolution can be improved by  the angular resoliutio
+Since A solution that is more convenient, cost-effective and simple solution would be to
+
+than itof to collect information from an extended part of the incoming wave in time it is easier to sample the incoming wavefield both the wavefield is better the angular resolution is usually far inferior to the The quality of active sonar imaging systems are
+
+Improving the angular resulution of active sonar systems are 
+In general the 
+In modern active sonar imaging systems it is generally harder to obtain sufficient is to transmit and receive wideband acustic pulses using multi-sensor arrays.
+and process their record the backscattered wavefield using an array of hydrophones.
+Through a coherent combination of the sensor data these
+
 In active sonar array imaging, an acoustic wave is transmitted and the received echoes are usually recorded using an array of sensors.
 Each of these may be independently delayed and weighted such that signals emanating from directions of interest are added constructively, while noise and interference from other directions add destructively.
-Additionally, a window may be applied to the sensor channels to further adjust the arrays spatial response.
+Additionally, a window may be applied to the sensor channels to further adjust the array's spatial response.
 This process is known as beamforming.% The process of combining the signals from each sensor is known as beamforming.

### 2015-04-14 ThesisJo/derivations.tex
More on LCA

@@
 
+Tommelfingerregel for styring
+- 1/2 av rektangulær for punkter
+- 1/2 Hamming/Kaiser for skarpe transisjoner
+
+- Omtrent samme beta i transisjoner
+- Samme beta i spekkel/highlight
+- Annen beta i skygge
+
+Bildeparametre:
+- Oppløsning
+- Kontrast.
+- Dynamisk område.
+- I sonar er spekkel tekstur! Vi kan tillate en undertrykking av den, men ønsker forholdsvis lik spekkelstatistikk.
+- Temporal and spatial shift variance. Et punkt burde avbildes med omtrent samme amplitude uansett hvor det er i bildet.
+- Sensitivitet
+- Steered response.
+- Coherence
+- Qualitative assessment
+
+Hvor mange trenger vi.
+1. Brute force. Tester alle kombinasjoner av
+   - Oversampling
+   - Beta
+   - Styring
+2. Hva det koker ned til.
+   - Oversampling. Virker som om 2*M er en god regel, trenger litt mer om vi tillater mye styring.
+   - Beta. Holder med 3-6. Kan regnes anaAnalytisk
+   - Styring. Desto mer styring, desto mer agressiv beamforming. Burde ikke styre mer enn av vi får litt variasjon av nullpunkt, men med liten skift-invarians i amplitide.
+
+Sammenlignet med MVDR
+- Hvorfor trenger vi ikke subarray / tidsmidling?
+- MVDR er ``for'' adaptiv.
+Selv med 1-gain rett frem kan signalet derfra kansellers.
+Subarray midling hjelper mot dette, men tar ned oppløsningen.
+LCA har ikke frihet til å velge amplitude/phase kombinasjoner som tillater signalkansellering, og det viser seg at vi da ikke trenger å midle for å oppnå ønsket resultat.
+
 \subsection{Tricks}

### 2015-04-14 Simulator/article.tex
More on LCA

@@
 % \usepackage{caption}
-% \usepackage{subcaption}
+\usepackage{subcaption}
 \usepackage{color}
@@
 However, it is hard to find parameters that describe the objects uniquely, in particular due to the ambiguities arising from interpreting a 3D world from its 2D projection.
-Another approach is to use template matching, where the image segment is compared to a set of template images covering the relevant object classes, and assigned to the class of the template yielding the the best fit.
+An alternative approach is to use template matching, where the image segment is compared to a set of template images covering the relevant object classes, and assigned to the class of the template yielding the the best fit.
 The challenge here is to come up with accurate template images covering the actual configurations of object and seabed, especially if a static template library is used, which will lead to a computational complexity that is exponential with the number of model parameters.
@@
 This can be modeled with a Lambertian scattering model~\cite{Blake1993,Bell1995}, where the intensity of the backscattered sound solely depends on the incidence angle of the transmitted signal onto the model surface.
-When rendering OpenGL is set up to produce an optical 2D intensity image and depth map that reveals the distance from each pixel to the propagation axis.
+The OpenGL rendering produces an optical 2D intensity image and depth map that reveals the distance from each pixel to the propagation axis.
 For maximum flexibility this data is finally combined with OpenCL to produce the image templates.

### 2015-04-14 ThesisJo/derivations.tex

--- 2015-03-13 ThesisJo/derivations.tex
@@
 \begin{align}
-\w = e^{j\phi_\theta[m]}\big(\alpha - (1-\alpha)\cos\left(\frac{2\pi m}{M-1}\right)\big)
+\w = \frac{e^{j\varphi}}{2\pi\alpha}\big(\alpha + (1-\alpha)\cos\left(\frac{2\pi n}{N-1}\right)\big)
 \end{align}
 %
-where $\phi_\theta[m] = \frac{2\pi}{\lambda}dm\sin\theta$. Inserting this yields:
+Inserting this yields:
 %
 \begin{align}
-\w\H\x &= \sumb{m=0}{M-1} e^{j\phi_\theta[m]}\Big(\alpha - (1-\alpha)\cos\left(\frac{2\pi m}{M-1}\right)\Big) x_m^* \nn
-&= \alpha\sumb{m=0}{M-1} e^{j\phi_\theta[m]} x_m^* - (1-\alpha)\sumb{m=0}{M-1} e^{j\varphi[m]}\cos\left(\frac{2\pi m}{M-1}\right) x_m^* \\
-&= \alpha \a\T\x + (1-\alpha)\b\T\x
+\w\H\x &= \sumb{m=0}{M-1} e^{j\varphi}\Big(\alpha + (1-\alpha)\cos\theta\Big) x_m^* \nn
+&= \sumb{m=0}{M-1} e^{j\varphi}\Big(\alpha + (1-\alpha)\cos\theta\Big) x_m^*
 \end{align}
 %
-where:
-%
-\begin{align}
-\a = \begin{bmatrix}
-     1 \\
-     e^{j\frac{2\pi}{\lambda}d\sin\theta} \\
-     e^{j\frac{2\pi}{\lambda}2d\sin\theta} \\
-     \vdots \\
-     e^{j\frac{2\pi}{\lambda}d(M-1)\sin\theta}
-     \end{bmatrix} \qquad \text{and} \qquad
-\b = \begin{bmatrix}
-     0 \\
-     e^{j\frac{2\pi}{\lambda}d\sin\theta}\sin\left(\frac{2\pi m}{M-1}\right) \\
-     e^{j\frac{2\pi}{\lambda}2d\sin\theta}\sin\left(\frac{2\pi m}{M-1}\right) \\
-     \vdots \\
-     0
-     \end{bmatrix}
-\end{align}
-%
-To attain unity gain in the look direction the weights must sum to one. 
-%
-\begin{align}
-\w\T\1 &= \big(\alpha\a - (1-\alpha)\b\big)\T\1 = 1
-\end{align}
-%
-This is true for any value of $\alpha$ if $\a\T\1 = 1$ and $\b\T\1 = 1$.
-
-where
-%
-\begin{align}
-|\w\H\x|^2 &= \Big| \alpha a + (1-\alpha)b \Big|^2 \nn
-&= \Big[\alpha a + (1-\alpha)b\Big] \Big[\alpha a^* + (1-\alpha)b^*\Big] \nn
-&= \alpha^2 aa^* + \alpha a(1-\alpha)b^* + \alpha a^*(1-\alpha)b + (1-\alpha)^2bb^* \nn
-&= \alpha^2 aa^* + \alpha(1-\alpha)(ab^* + a^*b) + (1-\alpha)^2bb^* \nn
-&= \alpha^2 (aa^* - (ab^* + a^*b) - bb^*) + \alpha(ab^* + a^*b) + bb^* \nn
-\end{align}
-%
-\begin{align}
-\frac{\partial}{\partial\alpha} |\w\H\x|^2 
-&= 2\alpha (aa^* - (ab^* + a^*b) - bb^*) + ab^* + a^*b = 0
-\end{align}
-%
-\begin{align}
-\alpha &= \frac{ab^* + a^*b}{2(aa^* - (ab^* + a^*b) - bb^*)}
-\end{align}
+where $\theta = \frac{2\pi n}{N-1}$.
 
@@
 \end{align}
-
-\subsection{Element}
-
-The frequency response of an element is is given by the Fourier transform of the element's window function.
-%
-\begin{align}
-W(\k) = \int\limits_{V} w(\p)\,e^{-j\k\dot\p} dV
-\end{align}
-%
-where $\p = \bmat{x & y & z}\T$ and $dV = dx\,dy\,dz$. For a rectangular 2D element spanning the xy-plane, having sides with length $d_x$ and $d_y$, the frequency response becomes:
-%
-\begin{align}
-W_\text{rect}(k_x, k_y) = \frac{\sin\frac{k_x\,d_x}{2}}{\frac{k_x\,d_x}{2}}\,\frac{\sin\frac{k_y\,d_y}{2}}{\frac{k_y\,d_y}{2}} = \sinc\frac{k_x\,d_x}{2}\,\sinc\frac{k_y\,d_y}{2}
-\end{align}
-
-\subsubsection{Opening angle and element size}
-
-What is the element size of the HISAS1030 sonar? This can be inferred from the element frequency response:
-%
-\begin{align}
-W_\text{rect}(k_x) = \frac{\sin\frac{k_x\,d_x}{2}}{\frac{k_x\,d_x}{2}} = \frac{1}{\sqrt{2}}
-\end{align}
-%
-This can be solved for $\frac{k_x\,d_x}{2}$ either by series expansion of $\sin(\cdot)$, or numerically to find that $\frac{k_x\,d_x}{2} \approx 1.39$.
-For HISAS1030, with an opening angle of 25$^\circ$ at 100kHz in both directions, this means the element size must be quadratic with sides:
-%
-\begin{align}
-d = \frac{2*1.39}{k_x} = \frac{2*1.39}{2\pi \frac{f_c}{c}\sin\theta} = \frac{1.39*1500\text{ms}^{-1}}{\pi\, 100\text{kHz}\,\sin{12.5^\circ}} = 3.07\text{cm}
-\end{align}
-%
-This leaves a little space between each sensor since the element spacing is around 3.75cm.

### 2015-04-14 ThesisJo/introduction.tex

--- 2015-03-02 ThesisJo/introduction.tex
@@
+\chapter{Introduction}
+
 The best processing unit we know of is the human mind.
@@
 
+\section{Aims of this thesis}
+
+\section{Motivation and scope}
+
+\section{Thesis outline}
+
 Signal processing is the art of combining physics, mathematics and computational engineering
@@
 - The wonder of how the brain process images. Great information carrier.
+
+\gls{ASIC}
 
@@
 - AUV revisit. Better image quality, lower data collection price.
-
-\chapter{Background}
-
-\section{Coordinate System}
-
-\begin{figure}[th]
-\graphicsAI[width=\linewidth]{gfx/geometry.svg}
-\caption{The coordinate system used throughout the thesis.}\label{geometry}
-\end{figure}
-
-In this thesis we will relate all geometrical interpretations to the euclidean, spherical coordinate system shown in \Fig{geometry}.
-When assessing the imaging performance of a sonar its origin will be in the center of the array, but when assessing scenes its origin will be in the center of the image.
-A unit length vector $\vec u$ can be written as:
-%
-\begin{align*}
-\vec u =
-\begin{bmatrix}
-\sin\theta \cos\phi \\
-\sin\theta \sin\phi \\
-\cos\theta
-\end{bmatrix}
-\end{align*}
-%
-
-\section{The Wave Equation}
-
-When a rock is thrown into a pound of water it abruptly displaces water at the location of impact.
-This is an unstable state where various physical forces compete for order.
-Gravity wants the water back where it wants, but due to inertia the mass will resist any change in its motion.
-The wrestling of these forces result in a spherical wave propagating from the displacement centre.
-
-The wave equation is the single most important equation governing a wave's propagation through a medium. When the medium is loss-less it can be expressed as:
-
-\begin{align*}
-\nabla^2 f(\p,t) - \frac{1}{c^2(\p)}\frac{\partial^2 f(\p,t)}{\partial t^2} = -s(t) \delta(\p-\p_0)
-\end{align*}
-
-where $\nabla^2$ is the Laplacian operator. The wave equation constrains the magnitude of the wavenumber,
-%
-\begin{align*}
-\vec k = \frac{\omega}{c} = \frac{2\pi}{\lambda},
-\end{align*}
-%
-so only the direction of $\k$ varies.
-
-\section{Synethic Aperture Sonar Simulation}
-
-For an autonomous underwater vehicle to reach its full
-
-Most autonomous devices designed today only perform very basic tasks.
-
-One of the areas where very fast where the development of faster and faster processing hardware might play a
-One of the technologies in which graphics processing units might make a welcome addition is in autonomous underwater vehicles.
-
-This thesis will describe a simulator for synthetic aperture sonar that
-
-\section{Scattering models}
-
-Defining parameters:
-- Roughness
-- Incident angle
-- Frequency
-
-Reflection and scattering from surfaces:
-
-Roughness >> wavelength:
-- Roughness > wavelength: Helmholtz-Kirchhoff (physical optics approach) (Hovem). Leads to Huygins priciple.
-- Roughness < wavelength: Rayleigh-Rice (perturbation technique) (Hovem). More accurate at slightly rough surfaces.
-
-Rayleigh. chi = 2 k sigma sin(theta): Rayleigh criterion: Surface is acoustically smooth when chi << 1.
-
-\begin{itemize}
-\item SWAT (Shallow Water Acoustic Toolset): 
-\item Bell's thesis:
-\item SIGMA:
-\item OpenGL does not support scattered writes. OpenCL does.
-\item 2D template approximation:
-\item Geometric distortions in SSS \cite{Cobra1992} Bell
-\item Sidescan sonar and synthetic aperture sonar. In seafloor coordinates. Maps to orthogonal rendering.
-\item Layover causes ambiguity, different seafloor phenomena can cause the same output image.
-Lots of cornercases when approximating the model in 2D.
-Instead use a 3D model and project it to a 2D one.
-Easier to operate, models are more general purpose.
-Less parameters to adjust.
-\item Viewer can be used to train an operator for finding non-trivial objects on the seafloor. Input for automated target recognition standard.
-\item Improved image quality of selected object, reduced data collection cost (price estimate?).
-\item OpenGL rendering can have infinite resolution, only limited by the pixel density in the image.
-\item Accuracy in terms of classification result. Need to embody static effects in a deterministic way. Incoherent synthesis of the image.
-\item Sound speed assumed constant.
-\item High frequency lead to little (define) penetration, lower range and higher resolution. Often modeled accurately with ray tracing techniques.
-\end{itemize}
-
-Model
-\begin{itemize}
-\item Ray solution: Separate amplitude and phase components of the Helmholtz eq.
-by applying geometric optics approximation.
-Ok as long as amplitude varies more slowly with position than the phase.
-Implications: High frequency, because the curvature of a ray over a wavelength must be small, and the fractional change in the sound speed must be small over a wavelength.
-Further simplification available by Snell's Law.
-Fast to compute.
-Not able to handle diffraction, focal points and caustics, and will cause (too) abrupt changes on distinct changes in sea environment.
-Can be improved upon with fuzzy beam modeling / Gaussian beam tracing.
-\begin{itemize}
-\item Pro: Easy to compute. Visual result. Handles complex bathymetry. Array and model response can be accounted for.
-\item Con: Only works for high frequency and small sound variations. No diffraction, caustics and focus. Infinite energy caustics/sharp variations. Incoherent(?)
-\end{itemize}
-\item Normal modes: Separate amplitude and phase into a range dependent term, and a depth dependent term.
-Assumes minimal range dependence in sound speed profile and depth.
-Limited near field accuracy.
-Acoustic field does not need to be calculated at intermediate ranges.
-Penetration can be modeled.
-can be calculated
-\begin{itemize}
-\item Pro: Ok to compute. Handles penetration. Low frequency, shallow water.
-\item Con: Horizontal stratification. Far field only. 
-\end{itemize}
-\item Parabolic equation: One way propagation.
-Marching solution.
-Initial starting solution must be known, often computed with normal mode.
-Only valid for narrow beams $\pm$20.
-Able to handle range dependent environments.
-Computing full field is easy.
-Computation time is of O(f$^2$), hence mostly useful for low-frequency scenarios.
-Unable to deal with horizontal refraction (shadow zones from e.g.
-conical mounts)
-\item Fast field programs / Green's function solutions.
-Exact full wave solution for acoustic fields in horizontally stratified media.
-Can handle both compressional and shear waves, and hence both fluid and solid media.
-Far field assumption.
-Similar to normal mode techniques.
-\item Finite element: Scene segmented into triangles or rectangles with a size of one tenth of a wavelength, or less.
-At each node the wavefield can be determined from the wave equation.
-Computationally limited to low frequencies.
-\item 2D model: Acoustic field in range and depth.
-Assumes that a wave transmitted in a vertical plane remains in it.
-Reasonable when the sound speed can be assumed constant in the horizontal plane, and when the seafloor normals are in the vertical range-depth plane (avoid horizontal refraction, or actually reflection).
-\end{itemize}
-
-What's new?
-\begin{itemize}
-\item Tuned for speed using GPUs, combining OpenGL and OpenCL to take advantage of existing functionality for computer graphics with the added flexibility of general purpose programming on the GPUs.
-\item Performance to spare: Game-like viewer for the visualization and verification of the rendered scene and corresponding sonar image.
-\item First simulator of its kinds aimed towards improved ATR and ultimately better AUVs.
-\item 3D models are still novel. 
-\end{itemize}

### 2015-04-17 LCAforActiveSonarImaging/lca.tex
More on LCA
Another day, no progres

@@
 Most modern active sonar imaging systems are usually far better at resolving details in range than in the angular direction.
-This is because it is relatively cheap and easy to collect a long and well sampled temporal representation of the incoming wavefield, but hard to do so spatially.
-It would require a very large array with a lot of sensors, which is costly, complex and impractical.
-Alternatively, a larger array can be synthesized from several successive pings recorded by a smaller physical array using a technique called synthetic aperture sonar (SAS).
-Unfortunately, not many of these systems exist due to the difficulty of navigating with sufficiently high precision under water.
-
-It would be more convenient to avoid 
+This is because it is relatively cheap and easy to collect a long and well sampled temporal representation of the incoming wavefield, but hard to get the spatial equivalent.
+It would require a very large array with a lot of sensors, which is costly, complicated and impractical.
+An alternative is to synthesize a large array from a small array by coherently combining data from successive pings.
+This technique is called synthetic aperture sonar (SAS).
+Unfortunately, it is very challenging to navigate with sufficient precision under water this is a very challenging task due to the difficulty of navigating with sufficiently high precision under water.
+
+Due to the challenges of solving this issue with hardware, a lot of research is put into improving the image reconstruction algorithms.
+At the heart of these is a technique called beamforming, which seeks to estimate scattering strength for each pixel location.
+It does this by applying a suitable set of the estimate the involves applying a suitable set of delays and weights to the sensors to coherently combine the sensors' data using a seeks to where focus is obtained in  the where delays and weights are used on the array's channels to form a coherent combination ly combine
+
+It would be more convenient 
 Hence, if the angular resolution can be improved by  the angular resoliutio
 Since A solution that is more convenient, cost-effective and simple solution would be to
+
+Reminder: Always hit you equal in war, +/- 2 ranks.
+Example: If your position on the war map is #20, then pick a target in the range #18-#22.
+Only exception is for our top 10-15 players, which may use second attack for cleanup.
 
@@
 \section{Results and Discussion}
+
+% \begin{table}[!b]\centering%\normalsize
+% \begin{tabular}[c]{l r r r@{}  l}\hline
+% \rowcolor{tabBlue} & \multicolumn{1}{>{\columncolor{tabBlue}}c}{\bf B$_\text{arith}$} & \multicolumn{1}{>{\columncolor{tabBlue}}c}{\bf B$_\text{mem}$} & \multicolumn{2}{>{\columncolor{tabBlue}}c}{\bf B$_\text{mem}$/B$_\text{arith}$} \\\hline
+% Arithmetic & 1.03 Tflop/s & & &\\
+% Global memory & & 36 Gfloats/s & \hspace{30pt} 1 &:30 \\
+% Shared memory & & 257 Gfloats/s & 1 &:4 \\
+% Registers & & $>$1.5 Tfloats/s & $>$3 &:2~\cite{Vasilyy}
+% \end{tabular}
+% \caption{Nvidia Quadro 6000: Memory throughput, $B_{\lowercase{\text{mem}}}$, compared to arithmetic throughput, $B_\text{arith}$.}\label{throughputs}
+% \end{table}%
+
+\newcommand\cc[1]{\multicolumn{1}{>{\columncolor{tabBlue}}c}{\bf #1}}
+\begin{table}[!b]\centering%\normalsize
+\begin{tabular}[c]{l r r r@{}  r r r}\hline
+\rowcolor{tabBlue} & \cc{Resolution} & \cc{S} & \cc{$\beta$} & \cc{$\varphi^*$} &  \cc{N$_\beta$} & \cc{N$_\varphi$} \\\hline
+& 50\%  & 28 & [0,3] & \hspace{.15cm} 50\% & 2 & 3 \\
+& 50\%  & 28 & [0,6] & \hspace{.15cm} 50\% & 2 & 3 \\
+& 50\%  & 28 & [0,9] & \hspace{.15cm} 50\% & 2 & 3 \\
+& 100\% & 32 &  &:30 \\
+& 150\% & 32 & 1 &:4\\
+\end{tabular}\\
+* Steering is relative to the -3dB width of the window being applied.
+\caption{Qualitative study}\label{parameter_study}
+\end{table}%
 

### 2015-04-23 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:04:23	19:23:56	(Thu23.Apr)

@@
 Most modern active sonar imaging systems are usually far better at resolving details in range than in the angular direction.
-This is because it is relatively cheap and easy to collect a long and well sampled temporal representation of the incoming wavefield, but hard to get the spatial equivalent.
+This is because it is relatively cheap and easy to collect a long and well sampled temporal representation of the incoming wave field, but hard to get the spatial equivalent.
 It would require a very large array with a lot of sensors, which is costly, complicated and impractical.
-An alternative is to synthesize a large array from a small array by coherently combining data from successive pings.
-This technique is called synthetic aperture sonar (SAS).
-Unfortunately, it is very challenging to navigate with sufficient precision under water this is a very challenging task due to the difficulty of navigating with sufficiently high precision under water.
-
-Due to the challenges of solving this issue with hardware, a lot of research is put into improving the image reconstruction algorithms.
-At the heart of these is a technique called beamforming, which seeks to estimate scattering strength for each pixel location.
-It does this by applying a suitable set of the estimate the involves applying a suitable set of delays and weights to the sensors to coherently combine the sensors' data using a seeks to where focus is obtained in  the where delays and weights are used on the array's channels to form a coherent combination ly combine
-
-It would be more convenient 
-Hence, if the angular resolution can be improved by  the angular resoliutio
-Since A solution that is more convenient, cost-effective and simple solution would be to
-
-Reminder: Always hit you equal in war, +/- 2 ranks.
-Example: If your position on the war map is #20, then pick a target in the range #18-#22.
-Only exception is for our top 10-15 players, which may use second attack for cleanup.
+An alternative is to synthesize a large array from a small array through a coherent combination of data from successive pings, a technique known as synthetic aperture sonar (SAS).
+Unfortunately, the difficult nature of navigating with high precision under water limits this technique to systems with advanced navigational equipment.
+
+Due to the limitations imposed on hardware, a lot of research is put into improving the part of the digital image formation process.
+A core component here is the beamformer, a spatial band-pass filter used on the wave field to estimate each pixel value.
+To allow the use of wide-band signals a coarse steering is first applied by delaying the sensor channels, followed by an individual weighting of them.
+Using complex weights, this implies fine adjustment of both the amplitude and the phase of the spatial filter.
+While conventional beamformers apply static weight functions, adaptive methods tailor the weights to the wave field.
+
+One of the best studied adaptive methods is the minimum variance distortion-less response (MVDR) beamformer.
+It solves an optimization problem to come up with a weight set that yields unity gain in the desired direction, while minimizing the effect of noise and interference from other directions.
+The result is often good looking non-linear image with improved contrast and detail resolution.
+However, MVDR is inherently neither statistically nor numerically robust, and typically needs to average data in space and time, and regularization.
+This constrains the solution space of the weights.
+
+We will show that when applied to experimental data, the weights typically chosen by a robustified MVDR yield responses that are typically symmetric and steered within a fraction of the -3dB resolution of the respective window.
+Then we make Kaiser windows of different types, and with varying degree of steering, that combined discretizes the MVDR solution space.
+
+The LCA method was first introduced in clinical ultrasound imaging, where it was shown to improve image quality compared to conventional static beamforming methods \cite{Synnevag2007}.
+
+MVDR is a fully adaptive method, meaning an optimal set of weights are produced given its optimization problem. Its only constraint is that the weights must sum to one.
+
+From our work with MVDR it
+
+ tailor the weights to the incoming wavefield.
+
+Traditionally beamformers were static, but as processors get more powerful so with the emergence of more processing power
+
+\section{Methods}
+
+At heart of the image formation process is a technique called beamforming, which creates a focus point at the pixel of interest.
+It does this by applying a suitable set of delays and weights to the array elements, such that signals emanating from the pixel location are emphasized constructively, while noise and interference from other directions sum destructively.
 
@@
 
-% \bibliographystyle{IEEEtran}
-% \bibliography{references}
-% 
+\bibliographystyle{IEEEtran}
+\bibliography{references}
+
 % 

### 2015-05-28 ThesisJo/background.tex
la2pdf commit - 2015:05:28	11:31:07	(Thu28.May)

@@
 When assessing the imaging performance of a sonar its origin will be in the center of the array, but when assessing scenes its origin will be in the center of the image.
-A unit length vector $\vec u$ can be written as:
-%
-\begin{align*}
-\vec u =
-\begin{bmatrix}
-\sin\theta \cos\phi \\
-\sin\theta \sin\phi \\
-\cos\theta
-\end{bmatrix}
-\end{align*}
-%
 
@@
 %
-where we call $\a(\k)$ the signal's \emph{steering vector}.
+where $A$ is the signal amplitude and $\Omega$ is normalized angular frequency. $\a(\k)$ is what we call signal's \emph{steering vector}.
+
+A unit length vector $\vec u$ can be written as:
+%
+\begin{align*}
+\vec u =
+\begin{bmatrix}
+\sin\theta \cos\phi \\
+\sin\theta \sin\phi \\
+\cos\theta
+\end{bmatrix}
+\end{align*}
+%
+
+\section{Active Sonar}
+
+Most modern active sonar imaging systems are usually far better at resolving details in range than in the angular direction.
+This is because it is relatively cheap and easy to collect a long and well sampled temporal representation of the incoming wave field, but hard to get the spatial equivalent.
+It would require a very large array with a lot of sensors, which is costly, complicated and impractical.
+An alternative is to synthesize a large array from a small array through a coherent combination of data from successive pings, a technique known as synthetic aperture sonar (SAS).
+Unfortunately, the difficult nature of navigating with high precision under water limits this technique to systems with advanced navigational equipment.
+
+Due to the limitations imposed on hardware, a lot of research is put into improving the part of the digital image formation process.
+A core component here is the beamformer, a spatial band-pass filter used on the wave field to estimate each pixel value.
+A way to do this while allowing wide-band signals is to steer coarsely by applying delays to the sensor channels, followed by an individual weighting of them.
+Using complex weights, this implies fine adjustment of both the amplitude and the phase of the spatial filter.
+While conventional beamformers apply static weight functions, adaptive methods dynamically tailor the weights to the wave field.
 

### 2015-07-07 ThesisJo/introduction.tex
la2pdf commit - 2015:07:07	12:34:34	(Tue 7.Jul)

@@
 But *how* should be additional processing power be used?
+
+One way to improve the performance of imaging devices is to make them more intelligent. 
+To improve the performance of modern imaging devices is becoming increasingly intelligent, they adapt their behaviour to best fit the situation they are in.
+A regular compact camera for instance, adjusts its shutter opening and speed to
+
+The image quality of any imaging device is mostly its by its hardware components, and .
 

### 2015-07-22 ThesisJo/introduction.tex
la2pdf commit - 2015:07:22	13:35:27	(Wed22.Jul)

@@
 \section{Thesis outline}
+
+C-I:
+
+Introduction
+   Beamforming
+   Wave fields and Array Processing
+   Conventional Beamforming
+   Adaptive Beamforming
+      MVDR
+      Capon/MPDR
+      APES
+      Wiener
+   Medical Ultrasound Imaging
+   Medical Ultrasound Imaging
+   Digital Beamforming
+      Compact Beamformers & Delta Sigma
+      Adaptive Beamforming in Medical Ultrasound Imaging
+         MVDR-based Beamformers and Related Methods
+         The Coherence Factor
+         Other Adaptive Methods
+Summary of publications
+   Paper I-VII
+Discussion and future work
+   Summary and discussion
+   Future Work
+
+J-F:
+
+Propagating Waves
+Beamforming
+  Beampattern
+Minimum Variance Beamforming
+  Signal Cancellation
+  Robust Minimum Variance Beamforming
+  Other Adaptive Methods
+  Other High Resolution Methods
+Medical Ultrasound Imaging
+  Broadband Near Field Minimum Variance Beamforming
+  Estimation of the Spatial Covarance Matrix
+Adaptive Beamforming in Medical Ultrasound Imaging: State of the Art
+Blind Source Separation
+  Independent Component Analysis
+  Relation to Adaptive Beamforming
+Summary  of Papers
+Discussion
+Conclusion and further research
+
+Ann:
+
+Introduction
+   Aims of the thesis
+   Motivation and scope
+   Outline
+Background
+   Sonar
+      Historical Perspective
+      Sonar Imaging
+      Scattering and Reflection
+      Key challenges
+   Propagating sound waves
+   Array processing and beamforming
+      Signal model
+      Conventional beamforming
+      The beampattern
+   Adaptive Beamforming
+      Optimal MPDR and MVDR
+      Adaptive MVDR
+      Adaptive APES
+      Low Complexity LCA
+      Adaptive methods based on aperture coherence
+      Robustness of adaptive beamformers
+   Performance metrics
+Summary of publications
+Summary and Discussion
+Future work
+
+Wei:
+
+Introduction
+   Overview of fractional calculus
+   Challenges of fractional calculus
+   Definition of fractional calculus
+   Fractional calculus and anomalous phenomena
+      Non-Fourier heat conduction
+      Modeling arbitrary power law attentuation
+   Solving fractional differential equations
+      Analytical methods
+      Numerical models
+   Optimization and parallelization
+      Locality and data reuse
+      Vectorization
+      Parallelization
+   Performance modeling
+      Serial Performance model
+      Parallel Performance model
+Summary of papers
+Future work
+
+JP
+
+Background
+   General-purpose computing on GPUs
+      Comparing CPU and GPy performance
+      Programming a GPU
+   Medical Ultrasound Imaging
+      Beamforming
+      Sampling and processing complexity
+      GPUs in medical ultrasound imaging
+      Adaptive beamforming
+      Capon beamforming of focused broadband beams
+   Volume rendering
+      Adaptive volume rendering
+   Ultrasound field simulations
+   Concluding remarks
+Introduction to papers
+   Motivation
+   Aims of study
+   Summary of papers
+   Main contributions
+   Discussion and future work
+   Multimedia content
+   Software
+References
+Papers
 

### 2015-07-22 Simulator/article.tex
la2pdf commit - 2015:07:22	14:08:05	(Wed22.Jul)

@@
 
-\begin{figure}[p]\centering%
+What's going on?
+
+\begin{figure*}[h]\centering%
 \subfloat[SAS image of a cylinder with 2.6\;m length and 0.53\;m radius.]{%
-\graphicsAI[drawing,width=\linewidth]{gfx/data_cylinder_submerged.svg}%
+\graphicsAI[drawing,width=0.5\linewidth]{gfx/data_cylinder_submerged.svg}%
 \label{data_cylinder}}%
 \subfloat[Cylinder simulation with parameters estimated from SAS image: Length 2.6\;m, burial depth 0.263\;m, aspect angle 105$^\circ$.]{
-\graphicsAI[drawing,width=\linewidth]{gfx/sim_cylinder_submerged_adaptive.svg}%
+\graphicsAI[drawing,width=0.5\linewidth]{gfx/sim_cylinder_submerged_adaptive.svg}%
 \label{sim_cylinder}}%
 \caption{A SAS image of a cylinder and a template simulation adapted to it.}
-\end{figure}
+\end{figure*}
 
@@
 
-\begin{figure}[p]\centering%
-\subfloat[Best template out of a predefined set. Inaccurate cylinder length, alignment and immersion depth. The segmented shadow and highlight regions are misaligned.]{
-\graphicsAI[drawing,width=\linewidth]{gfx/overlay_cylinder_submerged.svg}%
-\label{overlay_cylinder}}%
-% \hfil%
-\subfloat[Adapted template. Template simulated with adapted parameters. Good fit in both highlight and shadow segments.]{%
-\graphicsAI[width=\linewidth]{gfx/overlay_cylinder_submerged_adaptive.svg}%
-\label{overlay_cylinder_adaptive}}%
-\hfill\parbox{\linewidth}{\small\centering
-Segmentation overlay color description:\\[.5\baselineskip]
-\begin{minipage}{\linewidth}
-\cdesc{800000}{Template highlight and image highlight}
-\cdesc{FF1000}{Template highlight only}
-\cdesc{FFEB00}{Image highlight only}
-\cdesc{83FF7C}{Background pixels}
-\end{minipage}%
-\begin{minipage}{\linewidth}
-\cdesc{000083}{Template shadow and image shadow}
-\cdesc{0014FF}{Template shadow only}
-\cdesc{00EFFF}{Image shadow only}
-\cdesc{00A7FF}{Template shadow and image highlight}
-\end{minipage}}
-\vspace{\baselineskip}
-\caption{Segmented simulation image overlaid the segmented SAS image.}
-\end{figure}
+BS!!!
+
+\begin{figure*}[!tb]\centering%
+
+\subfloat[Best template out of a predefined set. Inaccurate cylinder length, alignment and immersion depth. The segmented shadow and highlight regions are misaligned.\label{overlay_cylinder}]{
+\graphicsAI[drawing,width=.5\linewidth]{gfx/overlay_cylinder_submerged.svg}%
+}%
+% % \hfil%
+\subfloat[Adapted template. Template simulated with adapted parameters. Good fit in both highlight and shadow segments.\label{overlay_cylinder_adaptive}]{%
+\graphicsAI[width=.5\linewidth]{gfx/overlay_cylinder_submerged_adaptive.svg}%
+}%
+\subfloat[][...]{hmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmm}
+% \parbox{1cm}{HMMMMMMMMMMMMMMMMMMMMMMMMMMMMMMMMMMMMMMMMMMMMMMmmmmmmmmmmmmmmmmmmmm}
+% \hfill
+% 
+% % \newline
+% \raggedright
+% \mbox{\parbox{1cm}{\small\centering
+% HMmmmmmmmmmmmmmmmmmm mmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmm
+% % Segmentation overlay color description:\\[.5\baselineskip]
+% % \begin{minipage}{.49\linewidth}
+% % \cdesc{800000}{Template highlight and image highlight}
+% % \cdesc{FF1000}{Template highlight only}
+% % \cdesc{FFEB00}{Image highlight only}
+% % \cdesc{83FF7C}{Background pixels}
+% % \end{minipage}%
+% % \begin{minipage}{.49\linewidth}
+% % \cdesc{000083}{Template shadow and image shadow}
+% % \cdesc{0014FF}{Template shadow only}
+% % \cdesc{00EFFF}{Image shadow only}
+% % \cdesc{00A7FF}{Template shadow and image highlight}
+% % \end{minipage}
+% }
+% \vspace{\baselineskip}
+% \caption{Segmented simulation image overlaid the segmented SAS image.}
+\end{figure*}
 

### 2015-08-08 ThesisJo/introduction.tex
la2pdf commit - 2015:08:08	09:50:29	(Sat 8.Aug)

@@
    Digital Beamforming
-      Compact Beamformers & Delta Sigma
+      Compact Beamformers \& Delta Sigma
       Adaptive Beamforming in Medical Ultrasound Imaging

### 2015-08-08 ThesisJo/derivations.tex
la2pdf commit - 2015:08:08	09:50:29	(Sat 8.Aug)
la2pdf commit - 2015:08:08	10:20:34	(Sat 8.Aug)
la2pdf commit - 2015:08:08	10:56:32	(Sat 8.Aug)
la2pdf commit - 2015:08:08	11:26:45	(Sat 8.Aug)
la2pdf commit - 2015:08:08	14:11:37	(Sat 8.Aug)

--- 2015-04-14 ThesisJo/derivations.tex
@@
 This leaves a little space between each sensor since the element spacing is around 3.75cm.
+
+\newpage
+\section{Window Parameters}
+
+Signal model, monochromatic place wave from broadside with gaussian noise:
+\begin{align}
+x_m[n] = s_m[n] + q[n] = Ae^{j\Omega_0mn} + q[n]
+\end{align}
+
+Beamformer output
+\begin{align}
+z[n] = \sumb{n}{} w^*[n]\;x[n]
+\end{align}
+
+Frequency response of the window
+
+\begin{align}
+W(k_x) = \sumb{m}{} w_m\;e^{jk_x \p_x} =
+\begin{cases}
+\frac{\sin\left(\frac{M}{2}k_x d\right)}{\sin\left(\frac{1}{2}k_x d\right)} & \text{rectangular}                                         
+\end{cases}
+\end{align}
+
+Applying a window reduces the signal power because it gets attenuated at both ends. We quantify this effect with the coherent power gain:
+
+\begin{align}
+W(\k) &= \sumb{m}{} w_m\;e^{j\k\cdot\p_m}
+\end{align}
+
+\begin{align}
+Z(\k) &= \sumb{m}{} w_m\;(s_m+q_m)\;e^{-j\k\cdot\p_m} \nn
+&= \sumb{m}{} w_m\;A\;e^{j\k\cdot\p_m}\;e^{-j\k\cdot\p_m} + \sumb{m}{} w_m\;q_m\;e^{j\k\cdot\p_m} \nn
+&= A \underbrace{\sumb{m}{} w_m}_\text{signal gain} + \sumb{m}{} w_m\;q_m\;e^{j\k\cdot\p_m}
+\end{align}
+
+Worst Case Processing Loss = Prosessing Loss + Scallopping Loss
+
+=> a measure of the worst case reduction in SNR which results from the combination of the window function and the worst case frequency location.

### 2015-08-08 ThesisJo/derivations.tex
la2pdf commit - 2015:08:08	15:03:27	(Sat 8.Aug)
la2pdf commit - 2015:08:08	15:34:50	(Sat 8.Aug)

@@
 
+\begin{align}
+\text{Coherent Gain} &= W(0) = \sumb{m}{}w_m
+\end{align}
+
+\begin{align}
+\text{Coherent Power Gain} &= \big|W(0)\big|^2 = \big|\sumb{m}{}w_m \big|^2
+\end{align}
+
+\begin{align}
+\text{Incoherent Power Gain} &= \sumb{m}{}w_m^2
+\end{align}
+
+\begin{align}
+\text{Equivalent Noise Bandwidth} &= \big|W(0)\big|^2 = \big|\sumb{m}{}w_m \big|^2
+\end{align}
+
+\begin{align}
+\text{Equivalent Noise Bandwidth} &= \frac{\text{Coherent Power Gain}}{\text{Equivalent Noise Bandwidth}} = \frac{\big|\sumb{m}{}w_m \big|^2}{\big|\sumb{m}{}w_m \big|^2}
+\end{align}
+
+Scallopping loss:
+\begin{align}
+\text{SL} &= \frac{\left|W\left(\frac{\pi}{2OMd}\right)\right|}{W(0)} = \frac{\left|\sumb{m}{}w_m\;e^{-j\k\cdot\p_m}\right|}{\sumb{m}{}w_m}
+\end{align}
+
 Worst Case Processing Loss = Prosessing Loss + Scallopping Loss
@@
 => a measure of the worst case reduction in SNR which results from the combination of the window function and the worst case frequency location.
+
+Narrow ones:
+- Best coherent gain, lowest equivalent noise bandwidth, and most sensitive to steering meaning not needing to steer as much for some effect.

### 2015-08-08 ThesisJo/derivations.tex
la2pdf commit - 2015:08:08	16:10:58	(Sat 8.Aug)

--- 2015-08-08 ThesisJo/derivations.tex
@@
 
+\begin{align}
+\text{Coherent Gain} &= W(0) = \sumb{m}{}w_m
+\end{align}
+
+\begin{align}
+\text{Coherent Power Gain} &= \big|W(0)\big|^2 = \big|\sumb{m}{}w_m \big|^2
+\end{align}
+
+\begin{align}
+\text{Incoherent Power Gain} &= \sumb{m}{}w_m^2
+\end{align}
+
+\begin{align}
+\text{Equivalent Noise Bandwidth} &= \frac{\text{Noise Power}}{\text{Coherent Power Gain}} = \frac{N_0\sumb{m}{}w_m^2}{\big|\sumb{m}{}w_m\big|^2}
+\end{align}
+
+\begin{align}
+\text{Processing Loss} &= \frac{\text{Coherent Power Gain}}{\text{Equivalent Noise Bandwidth}} = \frac{\big|\sumb{m}{}w_m \big|^2}{\big|\sumb{m}{}w_m \big|^2}
+\end{align}
+
+\begin{align}
+\text{Scallopping loss} &= \frac{\left|W\left(\frac{\pi}{2OMd}\right)\right|}{W(0)} = \frac{\left|\sumb{m}{}w_m\;e^{-j\k\cdot\p_m}\right|}{\sumb{m}{}w_m}
+\end{align}
+
 Worst Case Processing Loss = Prosessing Loss + Scallopping Loss

### 2015-08-08 ThesisJo/derivations.tex
la2pdf commit - 2015:08:08	23:04:38	(Sat 8.Aug)

@@
 \begin{align}
-\text{Equivalent Noise Bandwidth} &= \frac{\text{Noise Power}}{\text{Coherent Power Gain}} = \frac{N_0\sumb{m}{}w_m^2}{\big|\sumb{m}{}w_m\big|^2}
+\text{Equivalent Noise Bandwidth} &= \frac{\text{Noise Power}}{\text{Coherent Power Gain}} = \frac{\frac{N_0}{{\color{red}d}}\sumb{m}{}w_m^2}{\big|\sumb{m}{}w_m\big|^2}
 \end{align}
@@
 \begin{align}
-\text{Scallopping loss} &= \frac{\left|W\left(\frac{\pi}{2OMd}\right)\right|}{W(0)} = \frac{\left|\sumb{m}{}w_m\;e^{-j\k\cdot\p_m}\right|}{\sumb{m}{}w_m}
+\text{Scallopping loss} &= \frac{\left|W\left(\frac{\pi}{2{\color{red}O}Md}\right)\right|}{W(0)} = \frac{\left|\sumb{m}{}w_m\;e^{-j\k\cdot\p_m}\right|}{\sumb{m}{}w_m}
 \end{align}
 
+\begin{align}
+\text{Scallopping loss} &= \text{Prosessing Loss}_\text{dB} + \text{Scallopping Loss}_\text{dB} 
+= 
+\end{align}
 Worst Case Processing Loss = Prosessing Loss + Scallopping Loss

### 2015-08-09 ThesisJo/derivations.tex
la2pdf commit - 2015:08:09	15:19:42	(Sun 9.Aug)

@@
 \begin{align}
+\text{Processing Gain} &= \frac{S_o/N_o}{S_i/N_i} = \frac{A^2\big|\sumb{m}{}w_m\big|^2 \big/ \sigma_q^2\sumb{m}{}w_m^2}{A^2 / \sigma_q^2} = \frac{\big|\sumb{m}{}w_m\big|^2}{\sumb{m}{}w_m^2}
+\end{align}
+
+\begin{align}
 \text{Processing Loss} &= \frac{\text{Coherent Power Gain}}{\text{Equivalent Noise Bandwidth}} = \frac{\big|\sumb{m}{}w_m \big|^2}{\big|\sumb{m}{}w_m \big|^2}
@@
 - Best coherent gain, lowest equivalent noise bandwidth, and most sensitive to steering meaning not needing to steer as much for some effect.
+
+Beamforming:
+- Maximize resolution
+- Minimize noise/error, both coherent and incoherent
+- Minimize signal cancellation
+
+Gauss: Miminized time-bandwidth product
+  -
+
+Rectangular
+  - 
+Dolph-Cheb: Minimize mainlobe width given sidelobe suppression
+Kaiser: For a restricted energy and time duration, maximize energy in the band of frequencies W.
+
+Adaptive windows perform best in detection of closely separated angles of significantly amplitudes.

### 2015-08-10 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:08:10	15:36:48	(Mon10.Aug)

--- 2015-04-17 LCAforActiveSonarImaging/lca.tex
@@
 
-\ifCLASSOPTIONcompsoc
-   \usepackage[caption=false,font=normalsize,labelfont=sf,textfont=sf]{subfig}
-\else
-   \usepackage[caption=false,font=footnotesize]{subfig}
-\fi
+% \ifCLASSOPTIONcompsoc
+%    \usepackage[caption=false,font=normalsize,labelfont=sf,textfont=sf]{subfig}
+% \else
+%    \usepackage[caption=false,font=footnotesize]{subfig}
+% \fi
 
@@
 
-\usepackage[maxfloats=40]{morefloats}
+% \usepackage[maxfloats=40]{morefloats}
 \newcounter{todoidx}
@@
 \renewcommand*\a{\vec a}
+\newcommand*\f{\vec f}
 \renewcommand*\i{\vec i}
@@
 \begin{abstract}
-
-The minimum variance distortionless response (NVDR) beamformer is a popular 
-Acoustic imaging systems can often benefit from adapting their behaviour situation at hand. When the data is received with a phased array.
-
-Most modern acoustic imaging methods rely on the data from a phased array
-
-\begin{itemize}
-\item Dynamic weights $\Rightarrow$ greater contrast/resolution potential
-\item But:
-\begin{itemize}
-\item Active system $\Rightarrow$ noise and signal correlated $\Rightarrow$ adaptive beamformers no longer robust.
-\item Subarray averaging decorrelate the noise, but in the process we sacrifice resolution.
-\item Computationally complex O($M^3$) due to inversion of a spatial covariance matrix.
+\color{gray}
+
+% 
+% \begin{itemize}
+% \item Dynamic weights $\Rightarrow$ greater contrast/resolution potential
+% \item But:
+% \begin{itemize}
+% \item Active system $\Rightarrow$ noise and signal correlated $\Rightarrow$ adaptive beamformers no longer robust.
+% \item Subarray averaging decorrelate the noise, but in the process we sacrifice resolution.
+% \item Computationally complex O($M^3$) due to inversion of a spatial covariance matrix.
 LCA is only O($MW$).
 Even if Capon subspace/beamspace versions exist, they remain more complex than the LCA, and is not as ideally suited for implementation of SIMD hardware (GPUs, DSPs, ...).
-\end{itemize}
-\item LCA is inherently robust, is ``parameter free'' once setup with a well designed window set, and can often be implemented with minor modifications to already existing DAS beamformers.
-\end{itemize}
-
-\end{abstract}
-
-% Keywords (normally not used for peer reviews)
-\ifPeerReview\else
-\begin{IEEEkeywords}
-Beamforming, adaptive beamforming, sonar, active, complexity.
-\end{IEEEkeywords}
-\fi}
-% \fi
-
-% make the title area
-\maketitle
-
-% This command fixes abstract positioning for compsoc articles:
-\IEEEdisplaynotcompsoctitleabstractindextext
-
-% (Optional) Add some extra info on cover page of peer review papers:
-% \ifCLASSOPTIONpeerreview
-% \begin{center} \bfseries EDICS Category: 3-BBND \end{center}
-% \fi
-
-% Insert page break and insert second title (peer review mode)
-\IEEEpeerreviewmaketitle
-
-\begin{abstract}
-The angular resolution and contrast in active sonar images depend on the beamformer's ability to receive signals from directions of interest, while suppressing noise and interference emanating from other directions.
+% \end{itemize}
+% \item LCA is inherently robust, is ``parameter free'' once setup with a well designed window set, and can often be implemented with minor modifications to already existing DAS beamformers.
+% 
+% The angular resolution and contrast in active sonar images depend on the beamformer's ability to receive signals from directions of interest, while suppressing noise and interference emanating from other directions.
 For sonar arrays, this is achieved by applying weights to the array channels.
-
-While classical beamformers use predefined windows, adaptive beamformers estimate the optimal window by analytical evaluation of the data.
+% 
+% While classical beamformers use predefined windows, adaptive beamformers estimate the optimal window by analytical evaluation of the data.
 The minimum variance (MV) beamformer, for instance, calculates the set of weights that minimises the variance of the beamformer's output.
-
+% 
 We have implemented a Low Complexity Adaptive (LCA) beamformer, which adaptively selects a window from a predefined set.
@@
 These results were verified by the experimental data.
-
-An attactive trait of the LCA is its low computational complexity; while the MV beamformer is of O($M^3$), the LCA method is of O($MW$), with $M$ being the number of channels and $W$ the number of windows.
+% 
+% An attactive trait of the LCA is its low computational complexity; while the MV beamformer is of O($M^3$), the LCA method is of O($MW$), with $M$ being the number of channels and $W$ the number of windows.
 We made the LCA perform like the MV method using a well designed set of 30 windows.
 Hence, unless the array is very small, the proposed method will perform like the MV beamformer or better, and at a fraction of the computational cost.
-\end{abstract}%
-%
-\keywords{Beamforming, Capon, Minimum variance, Low Complexity, Sonar}
-
-% \titlepage
-
-\newpage
-% \the\lineheight
+
+% \end{itemize}
+
+\end{abstract}
+
+% Keywords (normally not used for peer reviews)
+\ifPeerReview\else
+\begin{IEEEkeywords}\color{gray}
+Beamforming, adaptive beamforming, MVDR, LCA, sonar, active, complexity.
+\end{IEEEkeywords}
+\fi}
+% \fi
+
+% make the title area
+\maketitle
+
+% This command fixes abstract positioning for compsoc articles:
+\IEEEdisplaynotcompsoctitleabstractindextext
+
+% (Optional) Add some extra info on cover page of peer review papers:
+% \ifCLASSOPTIONpeerreview
+% \begin{center} \bfseries EDICS Category: 3-BBND \end{center}
+% \fi
+
+% Insert page break and insert second title (peer review mode)
+\IEEEpeerreviewmaketitle
+
 \section{Introduction}
 
-Most modern active sonar imaging systems are usually far better at resolving details in range than in the angular direction.
-This is because it is relatively cheap and easy to collect a long and well sampled temporal representation of the incoming wavefield, but hard to get the spatial equivalent.
-It would require a very large array with a lot of sensors, which is costly, complicated and impractical.
-An alternative is to synthesize a large array from a small array by coherently combining data from successive pings.
-This technique is called synthetic aperture sonar (SAS).
-Unfortunately, it is very challenging to navigate with sufficient precision under water this is a very challenging task due to the difficulty of navigating with sufficiently high precision under water.
-
-Due to the challenges of solving this issue with hardware, a lot of research is put into improving the image reconstruction algorithms.
-At the heart of these is a technique called beamforming, which seeks to estimate scattering strength for each pixel location.
-It does this by applying a suitable set of the estimate the involves applying a suitable set of delays and weights to the sensors to coherently combine the sensors' data using a seeks to where focus is obtained in  the where delays and weights are used on the array's channels to form a coherent combination ly combine
-
-It would be more convenient 
-Hence, if the angular resolution can be improved by  the angular resoliutio
-Since A solution that is more convenient, cost-effective and simple solution would be to
-
-Reminder: Always hit you equal in war, +/- 2 ranks.
-Example: If your position on the war map is #20, then pick a target in the range #18-#22.
-Only exception is for our top 10-15 players, which may use second attack for cleanup.
-
-than itof to collect information from an extended part of the incoming wave in time it is easier to sample the incoming wavefield both the wavefield is better the angular resolution is usually far inferior to the The quality of active sonar imaging systems are
-
-Improving the angular resulution of active sonar systems are 
-In general the 
-In modern active sonar imaging systems it is generally harder to obtain sufficient is to transmit and receive wideband acustic pulses using multi-sensor arrays.
-and process their record the backscattered wavefield using an array of hydrophones.
-Through a coherent combination of the sensor data these
+- Why?
+
+\IEEEPARstart{A}{daptive} methods detector design is a term methods have been introduced in various imaging fields over the years in attempts to improve image quality.
+One such method is the minimum variance distortionless response (MVDR) beamformer.
+Compared to conventional methods, MVDR is often able to improve image contrast and resolution, as shown in~\cite{Blomberg2013,Blomberg2012a,Dursun2009,Lo2004}.
+
+Despite its inherent potential, the MVDR beamformer has yet to see widespread adoption in the active sonar field.
+There may be several reasons for this.
+For one, the method is not inherently robust, and may suffer from a phenomenon called signal cancellation in active systems~\cite{Widrow1982}.
+Another reason is that in its original form, the computational complexity is cubic with the number of channels, O($M^3$), while conventional beamformers are at O($M$).
+This is because a spatial covariance matrix is estimated and inverted for each image pixel.
+
+One of the best studied adaptive methods is the minimum variance distortion-less response (MVDR) beamformer~\cite{Capon1969}.
+It computes a weight set that minimizes the power of noise and interference solves an optimization problem to come up with a weight set that yields unity gain in the desired direction, while minimizing the effect of noise and interference from other directions.
+The result is often good looking non-linear image with improved contrast and detail resolution.
+However, MVDR is neither statistically nor numerically robust by default, and typically needs to average data in space and time, and regularization.
+This constrains the solution space of the weights.
+
+Traditional beamformers such as Delay-and-Sum (DAS) weight the sensors using a predefined window.
+However, since the optimal spatial response is likely to change with time, it is often preferred to use beamformers that continously adapt the weights to the dynamic data.
+This usually involves solving some optimization problem.
+
+The Minimum Variance (MV) beamformer selects the window that minimizes the beamformer's output power~\cite{cap69}.
+It has recently been investigated for active sonar imaging in e.g.~\cite{saf09}, and in ultrasound imaging in e.g.~\cite{syn07}.
+We suggest using a Low Complexity Adaptive (LCA) beamformer, which for each pixel applies a set of predefined windows and selects one of them based on the MV criterion.
+Using the LCA beamformer results in increased robustness and improved edge definition compared to the MV and DAS beamformers.
+
+We will show that when applied to experimental data, the weights chosen by a robustified MVDR yield responses that tend to be symmetric and steered within a fraction of the -3dB resolution of the respective window.
+Then we create Kaiser windows of different types, and with varying degree of steering, that combined form a discrete version of the MVDR solution space.
+
+The LCA method was first introduced by Synnevåg et.
+al.
+in clinical ultrasound imaging, who demonstrated its ability to improve image quality compared to conventional static beamforming methods \cite{Synnevag2007}.
+
+This article is outlined as follows:
+
+% This article is outlined as follows: In Section \ref{methods}, we introduce the concept of adaptive beamforming and provide details on the MVDR and LCA method.
+Then, in \ref{maptogpu}, we investigate the complexity issue, discuss means for reducing arithmetic complexity, and detail an implementation that makes efficient use of the GPU's parallel resources.
+The final design is assessed in Sections \ref{images_and_benchmarks} and \ref{discussion}, where we provide benchmarks, comparisons with similar CPU implementations, and measures of how efficiently our implementation makes use of the GPU's resources.
+
+% - How much oversampling
+% - How many parameters, how great a span.
+% 
+% - Experiemental data, Holmengraa/cross
+% 
+% - Note on efficiency.
+% 
+% 
+% From our work with MVDR it 
+% 
+% 
+% Traditionally beamformers were static, but as processors get more powerful so with the emergence of more processing power
+
+\section{Methods}
+
+At heart of the image formation process is a technique called beamforming, which creates a focus point at the pixel of interest.
+It does this by applying a suitable set of delays and weights to the array elements, such that signals emanating from the pixel location are emphasized constructively, while noise and interference from other directions sum destructively.
 
@@
 
-Traditional beamformers such as Delay-and-Sum (DAS) weight the sensors using a predefined window.
-However, since the optimal spatial response is likely to change with time, it is often preferred to use beamformers that continously adapt the weights to the dynamic data.
-This usually involves solving some optimization problem.
-
-The Minimum Variance (MV) beamformer selects the window that minimizes the beamformer's output power~\cite{cap69}.
-It has recently been investigated for active sonar imaging in e.g.~\cite{saf09}, and in ultrasound imaging in e.g.~\cite{syn07}.
-We suggest using a Low Complexity Adaptive (LCA) beamformer, which for each pixel applies a set of predefined windows and selects one of them based on the MV criterion.
-Using the LCA beamformer results in increased robustness and improved edge definition compared to the MV and DAS beamformers.
-
-% When the wavefield is sampled using an array of sensors, the data from each sensor may be 
-% 
+3rd order Maclaurin / Tayler approximation
+\begin{align}
+\phi_\text{-3dB bounds} \approx -0.002894\beta^3 + 0.02451\beta^2 + -0.01308\beta + 0.2980
+\end{align}
+
 % Traditional beamformers such as Delay-and-Sum (DAS) apply predefined windows to all incoming data.
@@
 
-% \begin{figure}[p]\centering%
-% \subfloat[Image of a single speckle realisation.]{%
-% \includegraphics[width=\linewidth]{gfx/img_das_lca_mv.pdf}%
-% \label{speckleA}}%
-% % \hfil%
-% \subfloat[Mean image of 100 speckle realisations.]{%
-% \includegraphics[width=\linewidth]{gfx/img_das_lca_mv_mean100.pdf}%
-% \label{speckleB}}%
-% \caption{A simulated circular object buried in speckle is imaged from above.}\label{speckle}
+\begin{figure}[tbhp]%\centering%
+\includegraphics[width=\linewidth]{gfx/calc_kaiser_3dB.pdf}%
+\caption{Steering within -3dB for each window}\label{windows_steering}
+\end{figure}
+
+\subsection{Beamforming}
+
+Let the receiver be an $M$ element uniform linear array, and assume that the signature of the transmitted signal has been removed by a matched filter.
+Further assume that the array channels are digitally delayed to focus at a pixel with azimuth angle $\theta$ and range sample $n$, such that the delayed data from the $m$th channel can be expressed as $x_m[\theta,n]$.
+To simplify notation, we make the dependence on $\theta$ implicit from now on.
+
+By definition, the beamformer output $z[n]$ can be expressed as the weighted sum of all the delayed data samples:
+\begin{align}
+z[n] = \w\H[n]\x[n] = \bmat{w_0[n]\\w_1[n]\\\vdots\\w_{M-1}[n]}^H \bmat{x_0[n]\\x_1[n]\\\vdots\\x_{M-1}[n]},\label{z}
+\end{align}
+where $w_m$ is the weight factor assigned to channel $m$.
+With static weights this would be referred to as the conventional delay-and-sum (DAS) beamformer.
+A large variety of weighting functions exists here for trading lateral resolution for improved noise suppression (contrast), but one always ends up with a compromise between the two~\cite{Harris1978}.
+
+Various adaptive beamformers target this limitation by allowing the weights to change for each pixel to better fit the dynamic nature of the incoming wavefield.
+In other words, they attempt to use the \emph{a priori} information present in the data to improve image quality.
+The MVDR beamformer is one such method.
+It finds the set of complex weights that minimizes the beamformer's expected output power, while ensuring unity gain in the look direction~\cite{Capon1969}:
+
+\begin{align}
+\underset{\w[n]}{\argmin}\, E\{|z[n]|^2\} &= \underset{\w[n]}{\argmin}\, \w[n]\x\x\H\w\H[n]\\
+\text{subject to } \w[n]\1 &= 1
+\end{align}
+
+This convex optimization problem has the solution:
+
+\begin{gather}
+\vec w[n] = \frac{\Ri[n]\a}{\a\T\Ri[n]\a},\label{weights}
+\end{gather}
+where $\a$ is a steering vector and $\R=E\{\x[n]\x\H[n]\}\in\mathbb{C}^{M,M}$ is the spatial covariance matrix for the full array.
+Since we pre-steer our data to every pixel in the image we simplify (\ref{weights}) by substituting $\a$ with a row vector $\1$ that represents broadside phase-steering.
+To estimate $\R$ we compute a sample covariance matrix $\eR$.
+In this computation we perform some degree of:
+\begin{itemize}
+\item \emph{spatial averaging} to avoid signal cancellation by decorrelating coherent echoes~\cite{Kailath1985};
+\item \emph{temporal averaging} over an interval comparable to the pulse length (one to five samples) to maintain true speckle statistics~\cite{Synnevag2009a};
+\item \emph{diagonal loading} to improve robustness to parameter errors~\cite{Cox1987,Maksym1979}.
+\end{itemize}%
+%
+% will perform some degree of \emph{spatial averaging} to avoid signal cancellation caused by coherent echoes~\cite{Kailath1985}, \emph{temporal averaging} to maintain true speckle statistics~\cite{Synnevag2009a}, and \emph{diagonal loading} to improve robustness to parameter errors ~\cite{Cox1987,Maksym1979}.
+%
+Combined, these steps will also ensure a numerically well conditioned $\eR$.
+
+We perform \emph{temporal} and \emph{spatial averaging} first and put the result in an intermediate sample covariance matrix $\breve{\R}$.
+To do this we need to segment our array into subarrays.
+If we let $x_l[n]$ represent the data vector from subarray $l$ by
+\begin{gather}
+\x_l[n] = \bmat{x_l[n] & x_{l+1}[n] & \dots & x_{l+L-1}[n]}\T,
+\end{gather}
+then $\breve{\R}$ can be calculated as
+\begin{gather}
+\breve{\R}[n] =  \frac{1}{N_K N_L} \sumb{l=0}{M-L}\sumb{n'=n-K}{n+K} \x_l[n']\x_l\H[n'] \in\mathbb{C}^{L,L},\label{spatialR}
+\end{gather}
+where $N_K = 2K+1$ is the number of temporal samples to perform averaging over, and $N_L = M-L+1$ is the number of subarrays.
+
+The final estimate $\eR$ is found by adding a fraction $d$ of the total power of $\breve{\R}[n]$ to its diagonal~\cite{Synnevag2007}:
+\begin{align}
+\eR[n] = \breve{\R}[n] + \I \frac{d}{L} \tr\{\breve{\R}[n]\},\label{finalR}
+\end{align}
+where $\I$ is an identity matrix, $\tr\{\cdot\}$ represents the matrix trace operation, and $\tr\{\breve{\R}[n]\}$ is an estimate of the energy received from this pixel.
+
+Note how subarray averaging led to a size reduction of $\eR$ from $\mathbb{C}^{M,M}$ to $\mathbb{C}^{L,L}$, and hence will produce an $L$-element weight set when substituted into (\ref{z}).
+This weight set is applied to all the subarrays, before computing the beamformer output as in (\ref{z}).
+Or, equivalently, we may apply the weight set to the sum of all the subarrays by
+\begin{align}
+z[n] = \w\H[n] \sumb{l=0}{M-L} \x_l[n].\label{finalZ}
+\end{align}
+% \ifPeerReview
+% \begin{figure}[!t]\centering
+% \includegraphics[width=.8\linewidth]{gfx/buske2\figPostfix.eps}
+% \else
+% \begin{figure}[!t]\centering
+% \includegraphics[width=\linewidth]{gfx/implementation.eps}
+% \fi%
+% \caption{MVDR beamforming. For each pixel in range and azimuth:\newline
+% 1. an $L\times{}L$ sample covariance matrix $\eR$ is computed; \newline
+% 2. the term $\eR^{-1}\1$ is found using a linear equation solver;\newline
+% 3. and the beamformer output $z$ is computed from (\ref{finalZ}), where $\w$ is found by substituting $\eR^{-1}\1$ into (\ref{weights}). } \label{mvdr_beamforming}
 % \end{figure}
-
-% 
-% 
-% \begin{figure}[tp]\centering%
-% \subfigure{%
-% \graphicsAI[width=\linewidth]{gfx/img_das_lca_mv.pdf}}
-% \subfigure{%
-% \graphicsAI[width=\linewidth]{gfx/img_das_lca_mv_mean100.pdf}}%
-% \caption{A simulated circular object buried in speckle is imaged from above.\newline\parbox[t]{1.50cm}{Top:\newline Bottom:}\parbox[t]{6.8cm}{Image of a single speckle realisation.\newline Mean image of 100 speckle realisations.}}\label{speckle}%%
-% \end{figure}%
-% 
-% \begin{figure}[t]\centering%
-% \graphicsAI[width=\linewidth]{gfx/img_mean_cut.pdf}%
-% \caption{Lateral cut in the mean image at range 41.1\,m (highlight) and 44.5\,m (shadow).}\label{cutmean}%
-% \end{figure}
-% % \begin{figure}[t]\centering
-% % % \graphicsAI[drawing,width=\linewidth]{gfx/the_box.svg}\\
-% % \includegraphics[width=\linewidth]{gfx/selected_windows.pdf}
-% % \caption{Selected windows}\label{selected_windows}
-% % \end{figure}
-% % \begin{figure}[t]\centering
-% % \includegraphics[width=\linewidth]{gfx/selected_windows_angle.pdf}
-% % \caption{Selected windows - Colored by angle}\label{selected_windows_angle}
-% % \includegraphics[width=\linewidth]{gfx/selected_windows_sym_angle.pdf}
-% % \caption{Selected windows - Colored symmetrically by angle}\label{selected_windows_sym}
-% % \end{figure}
-% \begin{figure}[t]\centering
-% \graphicsAI[width=\linewidth]{gfx/img_holmengraa.pdf}%
-% \caption{HISAS sidescan sonar (SSS) image of the shipwreck Holmengraa.}\label{holmengraa}%
-% \end{figure}
+As summarized in Fig. \ref{mvdr_beamforming}, the MVDR method is applied to each pixel independently, by:
+\begin{enumerate}
+\item computing the sample covariance matrix $\eR$ in (\ref{finalR});
+\item computing $\eRi\1$ in (\ref{weights});
+\item computing the beamformer output $z$ in (\ref{finalZ}).
+\end{enumerate}
+Next we will evaluate these steps in terms of arithmetic complexity, and then discuss their mappability to parallel hardware.
+
+\begin{itemize}
+\item \emph{Window function}
+\item \emph{Parameter boundaries}
+\item \emph{Number of windows needed}
+\item \emph{Oversampling need}
+\item \emph{Image quality}
+\end{itemize}
+
+\section{Kaiser windows}
+
+As will be demonstrated in upcoming sections, the LCA beamformer works very well with windows generated from the Kaiser-Bessel function. It is defined as:
+%
+\begin{align}
+f_m = \begin{cases}
+\frac{I_0\left(\pi\beta\sqrt{1-\left(\frac{2m}{M-1}-1\right)^2}\right)}{I_0(\pi\beta)}, & 0\le m\le M-1 \\
+0 & \text{otherwise},
+\end{cases}
+\end{align}
+%
+where $I_0$ is the zeroth order Modified Bessel function of the first kind:
+%
+\begin{align}
+I_0(x) = \sumb{a=0}{\infty} \left[ \frac{\left(\frac{x}{2}\right)^a}{a!} \right]^2.
+\end{align}
+%
+The Kaiser-Bessel window is near optimal in the sense of having its peak energy concentration around $\theta=0^\circ$, for a given time-bandwidth product specified with the parameter $\beta$:
+%
+\begin{align}
+\beta = \frac{TB}{2},
+\end{align}
+%
+where $T$ is the extent of the window in time and $B$ is its bandwidth.
+Adjusting $\beta$ changes the trade-off between mainlobe width and sidelobe level.
+At small values the windows approach the response of a rectangular window, and at large values ($>5$) the window converges to a Gaussian both in time and frequency.
+This class of windows is generally considered well suited for separating closely spaced sources with amplitudes of a high dynamic range~\cite{Harris1978}, they are easy to make, and they are optimal for any value of $\beta$.
+
+Before we apply the window to the sensor data, we normalize it to ensure unit gain in the direction of interest:
+%
+\begin{align}
+\w = \frac{\vec f}{\f\T\1} \quad\quad\text{where}\quad\quad \f =
+\bmat{
+f_0 \\ \vdots \\ f_{M-1}                                                           
+}
+\end{align}
+%
+Note that the normalization factor $\f\T\1$ is the reciprocal of the window's coherent gain.
+Since the signal-to-noise gain is constant for a given window, this normalization also proportionally increases the  incoherent noise gain.
+
+\begin{align}
+\w = \frac{\a\f}{\f\T\a} = 
+\end{align}
+
+\begin{align}
+\a = \bmat{
+1 \\
+e^{-j\frac{md}{\lambda}\sin(\phi)} \\
+\vdots\\
+e^{-j\frac{(M-1)d}{\lambda}\sin(\phi)}
+}
+\end{align}
+
+% \begin{align}
+% \z = \w\H\a 
+% \end{align}
+
+% 
+% Narrow ones:
+% - Best coherent gain, lowest equivalent noise bandwidth, and most sensitive to steering meaning not needing to steer as much for some effect.
+% 
+% 
+% Beamforming:
+% - Maximize resolution
+% - Minimize noise/error, both coherent and incoherent
+% - Minimize signal cancellation
+% 
+% Gauss: Miminized time-bandwidth product
+%   - 
+%   
+% Rectangular
+%   - 
+% Dolph-Cheb: Minimize mainlobe width given sidelobe suppression
+% Kaiser: For a restricted energy and time duration, maximize energy in the band of frequencies W.
+% 
+% Adaptive windows perform best in detection of closely separated angles of significantly amplitudes.
 
@@
 % 
-% Based on the proposed method by Synnevåg~\cite{syn11}, we have implemented a Low Complexity Adaptive (LCA) beamformer that keeps the minimum variance optimisation criteria but reduces the solution space to a discrete set of predefined windows.
-We use the Kaiser window function because it allows us to design a wide range of windows with different mainlobe widths and sidelobe suppression by adjusting the tradeoff parameter $\beta$.
-In addition we apply steering $\varphi$ to each of the windows and and constrain the window design to ensure unit gain in the look direction.
-
-\section{Experimental Setup}
-
-For the LCA to perform well it is important to let it have a diverse selection of windows to choose from.
-This selection is based on observations of the solutions found by the MV beamformer.
+%
+
+% \section{Experimental Setup}
+% 
+% For the LCA to perform well it is important to let it have a diverse selection of windows to choose from.
+The question is how this selection should be With our ovarall goal being a reduction of computational complexity, it The choice of window function does not seem to matter.
+% 
+% 
+% 
+% Using Kaiser windows as their extremal values areLimit our study to Kaiser windows
+% 
+% 
+% 
+% This selection is based on observations of the solutions found by the MV beamformer.
 We obtained good results in our experiments by letting the LCA beamformer select among 5 Kaiser windows with uniformly distributed $\beta$'s in the range $[0.05, 0.5]$ and a rectangular window.
@@
 This adds to a total of 30 windows.
-
-To make the choice of window less susceptible to pixel value uncertainties, the LCA beamformer was set to apply the most frequently selected window in an 11 pixel range region to the center pixel in that region.
+% 
+% To make the choice of window less susceptible to pixel value uncertainties, the LCA beamformer was set to apply the most frequently selected window in an 11 pixel range region to the center pixel in that region.
 In physical terms, this is a region of approximately 20\,cm.
@@
 
-% \newpage
+\begin{figure*}[tbhp!]\centering%
+\includegraphics[width=\linewidth]{gfx/mvdr_selected_windows_holmengraa.pdf}%
+\caption{Typical frequency responses for windows computed by a robust MVDR.
+The image is of the oiltanker Holmengraa, data collected from the 32 element HISAS1030 sonar with , operated in ssectorscan mode with an oversampling factor of ~10.
+The window responses was computed from 40x40 pixel groups in the shadow, highlight and speckle region.
+The dashed red lines at -13dB and -43dB marks the peak sidelobe level of a rectangular and Hamming window, respectively.
+Note how the responses are more or less symmetric, with very little steering in the shadow, moderate steering in speckle and steering within roughly 3dB in highlight.
+The phase varies mostly in highlight.
+Full opening angle.
+Parameters.
+}\label{mvdr_selected_windows}
+\end{figure*}
+
+\newpage
+
 \section{Results and Discussion}
+
+To test the performance of the MVDR and LCA beamformers, we have processed data aquired by the Kongsberg Maritime HUGIN AUV carrying their 32 element HISAS1030 sonar~\cite{Hansen2009}.
+It is a high resolution synthetic aperture sonar with array length 1.2\;m, 100\;kHz operating frequency, a bandwidth of 40\;kHz and transmits at a 25$^\circ$ opening angle.
+
+Two different scenes will be presented.
+One of the 1500 dwt oiltanker wreck Holmengraa lying at a slanted seabed at 77\;m depth outside of Horten, Norway~\cite{holmengraa}.
+It measures 68\;m by 9\;m and fills most of the highlit sector.
+The other scene is of an iron cross connected to a an anchor (fig ???).
+
+A quick summary of the results and the order they will be presented is as follows:
+
+\begin{itemize}
+\item \emph{Window function}.
+Frequency responses of typical windows chosen by MVDR in shadow, speckle and highlight regions in an image is shown in \Fig{mvdr_selected_windows}.
+The responses tend to be symmetric, minimally steered and have slightly varying mainlobe width and sidelobe suppression.
+We use Kaiser windows to mimic these responses.
+\item \emph{Window parameters}.
+The Kaiser window parameter $\beta$ adjusts the compromise between resolution and sidelobe suppression.
+At $\beta=0$ it yields a rectangular window, and at and $\phi$ controls the amount of steering..
+$\beta=0$ ) and a gaussian curve (
+\end{itemize}
+
+The following sections with expand and discuss these results further.
+
+\subsection{Window function}
+
+Assuming that a robust MVDR computes the optimal windows in the minimum variance sense, it makes sense to study them.
+For this we selected a scene of the oil tanker wreck Holmengraa, which lies at a slanted seabed at nearly 100m\;depth.
+A sectorscan image of a single ping of data collected by the HISAS1030 syntetic aperture sonar is shown in \Fig{mvdr_selected_windows}, along with the window responses MVDR typically chooses in shadow, speckle and highlight regions in this image.
+A fairly agressive yet stable set of parameters for MVDR where used; subarray size $L=16$, temporal averaging of $K=1$, and $d=1\%$ of diagonal loading.
+
+While MVDR is free to choose non-symmetric windows, we can observe from \Fig{mvdr_selected_windows} that it prefers symmetric ones.
+In shadow and speckle the windows are fully symmetric, while in highlight the windows are approximately symmetric within the opening angle of the sonar.
+
+\subsection{Window parameters}
+
+The Kaiser window parameter $\beta$ adjusts the compromise between resolution and sidelobe suppression, and $\phi$ controls the amount of steering.
+At $\beta=0$ it becomes a rectangular window, and at larger values it approaches a gaussian curve.
+Typical values are in the range $\beta\in[0,10]$.
+
+$\phi$ controls the amount of steering.
+
+The Kaiser window parameter $\beta$ adjusts the compromise between resolution and sidelobe suppression.
+At $\beta=0$ it yields a rectangular window, and at and $\phi$ controls the amount of steering..
+$\beta=0$ ) and a gaussian curve (
+
+\begin{figure*}[tbhp!]%\centering%
+\includegraphics[width=\textwidth]{gfx/oversampling_mosaic_bounds_lca_windows_beta.pdf}%
+\caption{Window parameter boundaries. Oversampled image, 19x20 windows.}\label{oversampling_mosaic_bounds}
+\end{figure*}
+
+\newpage
+\begin{figure*}[tbhp!]%\centering%
+\includegraphics[width=\textwidth]{gfx/oversampling_mosaic_bounds.pdf}%
+\caption{Window parameter boundaries.
+LCA images using an oversampled image and a relatively large window database with $N_\beta=20$ window types, each steered $N_\phi=19$.
+}\label{oversampling_mosaic_bounds_beta}
+\end{figure*}
+
+Variables of interest:
+
+\begin{itemize}
+\item Window function. Mostly symmetrical, in particular in shadow and for the most part in speckle. In highlight regions more varying.
+\item Mainlobe width. Wide in shadows, narrow in detailed regions. Adding windows in between these extremal values.
+\item Steering. No steering in shadows, slight steering in speckle and no steering in highlight. Steering more than the -3dB width of the window function has no effect.
+\item Oversampling. From our observations one should have at least twice the number of beams as channels. Image is more visually pleasing and appears more detailed up to 8M.
+\end{itemize}
+\begin{table}[!b]\centering%\normalsize
+\begin{tabular}[c]{l c c c}\hline
+\rowcolor{tabBlue} & \bf Highlight & \bf Shadow & \bf Speckle  \\\hline
+Symmetry             & Moderate  & High   & High \\
+Mainlobe width       & Narrow    & Wide   & Moderate \\
+Sidelobe suppression & Low       & High   & Moderate \\
+Steering             & Low       & Up to -3dB width & Moderate \\
+Oversampling         & 2x-8x     & 2x-8x  & 2x-8x
+\end{tabular}
+\caption{Summary}\label{tab:summary}
+\end{table}%
+
+\newpage
+
+% the performance of the LCA beamformer is surprisingly similar regardless of what windows it is allowed to choose from.
+
+\newpage
+
+What's wrong here?
+
+\begin{figure*}[tbhp!]%\centering%
+\includegraphics[width=\textwidth]{gfx/oversampling_mosaic.pdf}%
+\caption{Window parameter boundaries. Oversampled image, 19x20 windows.}\label{oversampling_mosaic}
+\end{figure*}
+
+\begin{itemize}
+\item Oversampling tests. Both Holmengraa and cross.
+\item Mosaic indicating the number of windows needed.
+\item 
+\end{itemize}
 
@@
 
-We have tested the beamformers using data aquired by the Kongsberg Maritime HISAS1030 sonar~\cite{han09}, and on simulations of the same sonar with 100 realisations of speckle.
-
-The simulations contained a few highlighted cylindric regions with 2\,m diameter and a constant intensity 15.4\,dB over the average speckle level.
+% The simulations contained a few highlighted cylindric regions with 2\,m diameter and a constant intensity 15.4\,dB over the average speckle level.
 We focused on an object centered at 41\,m range and -3 degrees azimuth angle.
@@
 %\ref{saf09}
-
-In \Fig{speckle} we observe that both the adaptive beamformers produce images with a more clearly defined shadow.
+% 
+% In \Fig{speckle} we observe that both the adaptive beamformers produce images with a more clearly defined shadow.
 The edges are sharper and the shadow more distinct.
@@
 The same effects can also be observed laterally around the highlight region of the mean image; DAS is more prone to smear energy into the speckle region than the adaptive beamformers are.
-
-\Fig{cutmean} displays two lateral cuts in the mean image, one through the highlight at 41.1\,m range and one through the shadow at 44.5\,m range.
+% 
+% \Fig{cutmean} displays two lateral cuts in the mean image, one through the highlight at 41.1\,m range and one through the shadow at 44.5\,m range.
 Each cut was computed as the mean of a 40\,cm range band centered at the respective ranges.
@@
 The slightly inferior performance of the MV beamformer is due to the subarray averaging, which makes the effective array smaller~\cite{syn07}.
-
-By studying the LCA performance on the 100 speckle realisations, we found the choice of window to be highly dependent on the phenomena being imaged.
+% 
+% By studying the LCA performance on the 100 speckle realisations, we found the choice of window to be highly dependent on the phenomena being imaged.
 In speckle regions LCA favored narrow spatial responses with minor steering, while wide and fully steered responses were often selected in shadow regions.
 There was also a clear distinction between the responses selected in sidelobe regions and in speckle regions.
-
-% \Fig{hist} illustrates how often a particular window is chosen in general.
+% 
+% % \Fig{hist} illustrates how often a particular window is chosen in general.
 There are 10 $\phi$'s for each of the 10 $\beta$'s.
@@
 At value 65 the $\beta$ value is 2.025, and these and the remaining windows are hardly selected.
-
-A sidescan image of the 1500 dwt oil tanker wreck Holmengraa is shown in \Fig{holmengraa}.
+% 
+% A sidescan image of the 1500 dwt oil tanker wreck Holmengraa is shown in \Fig{holmengraa}.
 It is about 68\,m long and 9\,m wide, and lies on a slanted seabed at 77\,m depth~\cite{holmengraa}.
@@
 This, and the fact that the wreck was imaged at a range of about 105\,m makes the image quality poor, but sufficient to compare our beamformers.
-
-In the Holmengraa image we note again that the LCA produces a cleaner shadow and better edge definition. The MV method performed almost identically to LCA in this case, and was omitted.
+% 
+% In the Holmengraa image we note again that the LCA produces a cleaner shadow and better edge definition. The MV method performed almost identically to LCA in this case, and was omitted.
 
@@
 
+\subsection{Summary and order of results}
+
+In short, our findings regarding the optimal window types for LCA is as follows:
+
+\begin{itemize}
+\item \emph{Window function} should be symmetric, . Studying typical MVDR responses for shadow, speckle and highlight regions, we believe it to be insignificant
+\item \emph{Window parameters}. 
+\end{itemize}
+
+The following sections with expand and discuss these results further.
+
 \section{Introduction}
 
-\begin{itemize}
-\item Sell it!
-\item As briefly as possible, introduce beamforming.
-\item Adaptive beamformer's potential lies in its ability to suppress interference power
-\item Why adaptive beamformers struggle in active sonar systems.
+% \begin{itemize}
+% \item Sell it!
+% \item As briefly as possible, introduce beamforming.
+% \item Adaptive beamformer's potential lies in its ability to suppress interference power
+% \item Why adaptive beamformers struggle in active sonar systems.
 Correlated noise, robustification kills the adaptive potential.
@@
 Constraints must be applied in one way or another - parameters must be tuned.
-\begin{itemize}
-\item Noise and signal is correlated. Spatial averaging required. 
-\item Increases variance of speckle (not only in active systems?), spatial compunting \cite{Vignon2009} (US) or time averaging?.
+% \begin{itemize}
+% \item Noise and signal is correlated. Spatial averaging required. 
+% \item Increases variance of speckle (not only in active systems?), spatial compunting \cite{Vignon2009} (US) or time averaging?.
 Subarray averaging applied here \cite{Synnevag2007} (US) \todo{Is it bad to mix ultrasound/ sonar refs?}
-\end{itemize}
-\item LCA is like DAS, only we test several windows and choose the one that yield the lowest beamformer output power.
+% \end{itemize}
+% \item LCA is like DAS, only we test several windows and choose the one that yield the lowest beamformer output power.
 Ultrasound \cite{Synnevag2008}, active sonar \cite{Buskenes2011}\cite{Blomberg2011}\cite{Blomberg2012}\todo{who was that guy in the 80ties?}
-\item Outline: 
-\begin{itemize}
-\item Methods.
-\item See what happens to Capon's response %\cite{Synnevag2007}  when means for robustification are added.
-\item Setup the window set for LCA to match this.
-\item Notice that the LCA performs like Capon, even with a small window database.
-\item Start with Kaiser, see that other window functions also work well.
-\end{itemize}
-\end{itemize}
-\todopar{Out of the blue idea: Weigh CF with LCA? This guy did it with MVDR \cite{Wang2009}. Probably you know already... Okay, back on track.}
+% \item Outline: 
+% \begin{itemize}
+% \item Methods.
+% \item See what happens to Capon's response %\cite{Synnevag2007}  when means for robustification are added.
+% \item Setup the window set for LCA to match this.
+% \item Notice that the LCA performs like Capon, even with a small window database.
+% \item Start with Kaiser, see that other window functions also work well.
+% \end{itemize}
+% \end{itemize}
+% \todopar{Out of the blue idea: Weigh CF with LCA? This guy did it with MVDR \cite{Wang2009}. Probably you know already... Okay, back on track.}
 
@@
 
-% \bibliographystyle{IEEEtran}
-% \bibliography{references}
+\bibliographystyle{IEEEtran}
+\bibliography{references}
 % 

### 2015-08-10 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:08:10	15:44:37	(Mon10.Aug)
la2pdf commit - 2015:08:10	17:08:15	(Mon10.Aug)

@@
 
-- Why?
+% - Nature and scope
+% - Past work
+% - Methods
+% - Brief summary results
+% - Brief conclusions
+
+\IEEEPARstart{P}{erhaps} greatest untapped potential of active sonar systems lies in making them adaptive.
+State-of-the-art here is found in nature, for instance in bats who use a high frequency sonar to detect, identify and track prey in caves in a chaos of other bats.
+They are masters at adapting their sonars to the scenario at hand, and act accordingly.
 
@@
 % 
-% - Experiemental data, Holmengraa/cross
+% - Experiementl data, Holmengraa/cross
 % 

### 2015-08-11 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:08:11	10:31:06	(Tue11.Aug)

--- 2015-08-10 LCAforActiveSonarImaging/lca.tex
@@
 
-- Why?
+% - Nature and scope
+% - Past work
+% - Methods
+% - Brief summary results
+% - Brief conclusions
+
+\IEEEPARstart{T}{he} best adaptive sonars are found in nature.
+Bats, for instance, use a high frequency sonar to detect, identify and track prey with amazing precision in caves amongst a myriad of other bats.
+In comparison, most human made active sonars inhibit a static behaviour when they collect data, but
 

### 2015-08-11 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:08:11	11:37:35	(Tue11.Aug)

@@
 Bats, for instance, use a high frequency sonar to detect, identify and track prey with amazing precision in caves amongst a myriad of other bats.
-In comparison, most human made active sonars inhibit a static behaviour when they collect data, but
-
-\IEEEPARstart{A}{daptive} methods detector design is a term methods have been introduced in various imaging fields over the years in attempts to improve image quality.
-One such method is the minimum variance distortionless response (MVDR) beamformer.
+In comparison, human made sonars inhibit minimal adaptability in the data collection process, but try to make up for it in the post processing of the data.
+Perhaps the most studied adaptive method for reconstructing a sonar image is the minimum variance distortionless response (MVDR) beamformer.
+
 Compared to conventional methods, MVDR is often able to improve image contrast and resolution, as shown in~\cite{Blomberg2013,Blomberg2012a,Dursun2009,Lo2004}.

### 2015-08-12 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:08:12	17:45:00	(Wed12.Aug)

@@
 
-Compared to conventional methods, MVDR is often able to improve image contrast and resolution, as shown in~\cite{Blomberg2013,Blomberg2012a,Dursun2009,Lo2004}.
+MVDR is often able to improve image contrast and resolution compared to conventional static methods~\cite{Blomberg2013,Blomberg2012a,Dursun2009,Lo2004}.
+However, it is computationally intensive, needs robustification to work well, and should be used on images that are oversampled 10 times in azimuth.
+
+In prior studies we have investigated ways to reduce the complexity of MVDR and accelerate it using graphics computing units (GPUs)~\cite{Buskenes2014,Aasen2013}.
 

### 2015-08-12 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:08:12	18:15:16	(Wed12.Aug)

@@
 In comparison, human made sonars inhibit minimal adaptability in the data collection process, but try to make up for it in the post processing of the data.
-Perhaps the most studied adaptive method for reconstructing a sonar image is the minimum variance distortionless response (MVDR) beamformer.
+Perhaps the most studied adaptive method for reconstructing a sonar image is the minimum variance distortionless response (MVDR) beamformer~\cite{Capon1969}.
+It computes the window that is optimal in the sense of suppressing as much energy from the noise and interference as possible.
 
 MVDR is often able to improve image contrast and resolution compared to conventional static methods~\cite{Blomberg2013,Blomberg2012a,Dursun2009,Lo2004}.
-However, it is computationally intensive, needs robustification to work well, and should be used on images that are oversampled 10 times in azimuth.
-
-In prior studies we have investigated ways to reduce the complexity of MVDR and accelerate it using graphics computing units (GPUs)~\cite{Buskenes2014,Aasen2013}.
+However, this comes at a cost of a computational complexity that is cubic with the number of channels, O($M^3$), while conventional beamformers are at O($M$).
+We have shown in previous studies that this complexity can be reduced, and the method accelerated significantly using graphics computing units (GPUs)~\cite{Buskenes2014,Asen2013}.
+However, in this work we propose
+
+We suggest using a Low Complexity Adaptive (LCA) beamformer, which for each pixel applies a set of predefined windows and selects one of them based on the MV criterion.
+Using the LCA beamformer results in increased robustness and improved edge definition compared to the MV and DAS beamformers.
 
@@
 
-One of the best studied adaptive methods is the minimum variance distortion-less response (MVDR) beamformer~\cite{Capon1969}.
-It computes a weight set that minimizes the power of noise and interference solves an optimization problem to come up with a weight set that yields unity gain in the desired direction, while minimizing the effect of noise and interference from other directions.
-The result is often good looking non-linear image with improved contrast and detail resolution.
-However, MVDR is neither statistically nor numerically robust by default, and typically needs to average data in space and time, and regularization.
-This constrains the solution space of the weights.
+However, MVDR is neither statistically nor numerically robust by default, and typically needs to average data in space and time, and regularization. This constrains the solution space of the weights.
 
@@
 This usually involves solving some optimization problem.
-
-The Minimum Variance (MV) beamformer selects the window that minimizes the beamformer's output power~\cite{cap69}.
-It has recently been investigated for active sonar imaging in e.g.~\cite{saf09}, and in ultrasound imaging in e.g.~\cite{syn07}.
-We suggest using a Low Complexity Adaptive (LCA) beamformer, which for each pixel applies a set of predefined windows and selects one of them based on the MV criterion.
-Using the LCA beamformer results in increased robustness and improved edge definition compared to the MV and DAS beamformers.
 
@@
 \end{align}
+
+oversampled 10 times in azimuth~\cite{Asen2014}.
 

### 2015-08-13 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:08:13	11:09:44	(Thu13.Aug)

@@
 Bats, for instance, use a high frequency sonar to detect, identify and track prey with amazing precision in caves amongst a myriad of other bats.
-In comparison, human made sonars inhibit minimal adaptability in the data collection process, but try to make up for it in the post processing of the data.
+Human made sonars have little to no adaptibility when data is collected, but some methods have been investigated in the data post processing.
 Perhaps the most studied adaptive method for reconstructing a sonar image is the minimum variance distortionless response (MVDR) beamformer~\cite{Capon1969}.
-It computes the window that is optimal in the sense of suppressing as much energy from the noise and interference as possible.
+It computes the window that is optimal in the sense of suppressing noise as much as possible while under the restriction of  preserving the signal of interest.
 
 MVDR is often able to improve image contrast and resolution compared to conventional static methods~\cite{Blomberg2013,Blomberg2012a,Dursun2009,Lo2004}.
-However, this comes at a cost of a computational complexity that is cubic with the number of channels, O($M^3$), while conventional beamformers are at O($M$).
-We have shown in previous studies that this complexity can be reduced, and the method accelerated significantly using graphics computing units (GPUs)~\cite{Buskenes2014,Asen2013}.
-However, in this work we propose
-
-We suggest using a Low Complexity Adaptive (LCA) beamformer, which for each pixel applies a set of predefined windows and selects one of them based on the MV criterion.
+However, while the computational complexity of conventional beamformers are linear with the number of channels, O($M$), MVDR is at O($M^3$).
+This may discourage its use and in many cases make it infeasible.
+We have shown in previous studies that MVDR's complexity can be reduced, as well as accelerated significantly using graphics computing units (GPUs)~\cite{Buskenes2014,Asen2013}, but there may be a way to avoid it altogether.
+
+We propose a low complexity version of MVDR called the Low Complexity Adaptive (LCA) beamformer.
+This method applies a set of predefined windows and selects the one that is best able to suppress the noise.
 Using the LCA beamformer results in increased robustness and improved edge definition compared to the MV and DAS beamformers.
 
-Despite its inherent potential, the MVDR beamformer has yet to see widespread adoption in the active sonar field.
 There may be several reasons for this.

### 2015-08-13 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:08:13	13:21:00	(Thu13.Aug)

@@
 
-MVDR is often able to improve image contrast and resolution compared to conventional static methods~\cite{Blomberg2013,Blomberg2012a,Dursun2009,Lo2004}.
-However, while the computational complexity of conventional beamformers are linear with the number of channels, O($M$), MVDR is at O($M^3$).
+In many cases MVDR can improve image contrast and resolution compared to conventional static methods~\cite{Blomberg2013,Blomberg2012a,Dursun2009,Lo2004}.
+However, the computational complexity of conventional beamformers are linear with the number of channels, O($M$), while MVDR is at O($M^3$).
 This may discourage its use and in many cases make it infeasible.
-We have shown in previous studies that MVDR's complexity can be reduced, as well as accelerated significantly using graphics computing units (GPUs)~\cite{Buskenes2014,Asen2013}, but there may be a way to avoid it altogether.
-
-We propose a low complexity version of MVDR called the Low Complexity Adaptive (LCA) beamformer.
-This method applies a set of predefined windows and selects the one that is best able to suppress the noise.
+We have shown in previous studies that MVDR's complexity can be reduced, as well as accelerated significantly using graphics computing units (GPUs)~\cite{Buskenes2014,Asen2013}, but there is a way to avoid it altogether while achieving similar performance.
+
+This method is called Low Complexity Adaptive (LCA) beamformer and was first proposed by Synnevåg et.
+al.
+in clinical medical imaging~\cite{Synnevag2008}, who demonstrated its ability to obtain very similar results to MVDR.
+It does this by applying a set of predefined windows and selects the one that is best able to suppress noise.
+Since it uses the same optimisation criterion as MVDR, it may be considered a version of MVDR with a reduced solution space.
+As we will demonstrate, this works well because MVDR needs
+
+The LCA method was first introduced by Synnevåg et.
+al.
+in clinical ultrasound imaging, who demonstrated its ability to improve image quality compared to conventional static beamforming methods \cite{Synnevag2007}.
+
 Using the LCA beamformer results in increased robustness and improved edge definition compared to the MV and DAS beamformers.
@@
 Then we create Kaiser windows of different types, and with varying degree of steering, that combined form a discrete version of the MVDR solution space.
-
-The LCA method was first introduced by Synnevåg et.
-al.
-in clinical ultrasound imaging, who demonstrated its ability to improve image quality compared to conventional static beamforming methods \cite{Synnevag2007}.
 

### 2015-08-13 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:08:13	14:03:11	(Thu13.Aug)

@@
 However, the computational complexity of conventional beamformers are linear with the number of channels, O($M$), while MVDR is at O($M^3$).
-This may discourage its use and in many cases make it infeasible.
+This is impracticable and sometimes infeasible.
 We have shown in previous studies that MVDR's complexity can be reduced, as well as accelerated significantly using graphics computing units (GPUs)~\cite{Buskenes2014,Asen2013}, but there is a way to avoid it altogether while achieving similar performance.
 
-This method is called Low Complexity Adaptive (LCA) beamformer and was first proposed by Synnevåg et.
+This method is called Low Complexity Adaptive (LCA) beamformer and was first proposed by Synnev\aa{}g et.
 al.
@@
 It does this by applying a set of predefined windows and selects the one that is best able to suppress noise.
-Since it uses the same optimisation criterion as MVDR, it may be considered a version of MVDR with a reduced solution space.
-As we will demonstrate, this works well because MVDR needs
-
-The LCA method was first introduced by Synnevåg et.
-al.
-in clinical ultrasound imaging, who demonstrated its ability to improve image quality compared to conventional static beamforming methods \cite{Synnevag2007}.
+Since it uses the same optimisation criterion as MVDR, it may be considered a version of MVDR with a reduced window solution space.
+As we will demonstrate, this works well because a robust MVDR design tend to create windows that resembles slightly varied versions of conventional ones.
+
+This article investigates a few questions that remain unanswered.
+One is what window functions LCA should be allowed to choose from.
+This hardly seems to matter,   Our studies show minimal difference between the traditional windows, so the common and reasonable Kaiser-Harris window where the compromize between What windows should the LCA be allowed to choose from?
+How many is needed?
+by Synnev\aa{}g demonstrated the potential of LCA,
 

### 2015-08-13 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:08:13	14:58:38	(Thu13.Aug)

@@
 Since it uses the same optimisation criterion as MVDR, it may be considered a version of MVDR with a reduced window solution space.
-As we will demonstrate, this works well because a robust MVDR design tend to create windows that resembles slightly varied versions of conventional ones.
-
-This article investigates a few questions that remain unanswered.
-One is what window functions LCA should be allowed to choose from.
+As we will demonstrate, this works well because a robust MVDR design tends to create windows that resembles slightly varied versions of conventional ones.
+
+In this work we investigate a few questions related to the LCA that remain unanswered.
+One is what window function we can use, and how many windows we will need.
+We limit our study to the use of variations of the Kaiser-Harris window, since we observed neglible performance difference to using this function or any of the common trigonometric ones.
+The Kaiser windows are easy to compute, is fairly insensitive to coefficient errors, and can create windows with shapes spanning from rectangular to Gaussian.
+windows are needed, and what variations of the windows are necessary?
+
+Another question is what it takes to achieve double the resolution.
+
+ne is what window functions LCA should be allowed to choose from.
 This hardly seems to matter,   Our studies show minimal difference between the traditional windows, so the common and reasonable Kaiser-Harris window where the compromize between What windows should the LCA be allowed to choose from?
@@
 Using the LCA beamformer results in increased robustness and improved edge definition compared to the MV and DAS beamformers.
-
-There may be several reasons for this.
-For one, the method is not inherently robust, and may suffer from a phenomenon called signal cancellation in active systems~\cite{Widrow1982}.
-Another reason is that in its original form, the computational complexity is cubic with the number of channels, O($M^3$), while conventional beamformers are at O($M$).
-This is because a spatial covariance matrix is estimated and inverted for each image pixel.
-
-However, MVDR is neither statistically nor numerically robust by default, and typically needs to average data in space and time, and regularization. This constrains the solution space of the weights.
-
-Traditional beamformers such as Delay-and-Sum (DAS) weight the sensors using a predefined window.
-However, since the optimal spatial response is likely to change with time, it is often preferred to use beamformers that continously adapt the weights to the dynamic data.
-This usually involves solving some optimization problem.
 

### 2015-08-13 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:08:13	16:47:15	(Thu13.Aug)

@@
 
-In this work we investigate a few questions related to the LCA that remain unanswered.
-One is what window function we can use, and how many windows we will need.
-We limit our study to the use of variations of the Kaiser-Harris window, since we observed neglible performance difference to using this function or any of the common trigonometric ones.
-The Kaiser windows are easy to compute, is fairly insensitive to coefficient errors, and can create windows with shapes spanning from rectangular to Gaussian.
-windows are needed, and what variations of the windows are necessary?
+Synnev\aa{}g used a window set comprised of rectangular, Kaiser and inverted Kaiser functions.
+Of these only a specific Kaiser window were steered We limit our study to the use of variations of the parametric Kaiser-Harris window, since we observed neglible performance difference to using this function or any of the common trigonometric ones.
+The Kaiser windows are easy to compute, is fairly insensitive to coefficient inaccuracies, and can create windows with shapes spanning from rectangular to Gaussian.
+
+In this work we investigate a few questions related to the LCA that remain unanswered. One is what window function we can use, and how many windows we will need.
 
 Another question is what it takes to achieve double the resolution.
-
-ne is what window functions LCA should be allowed to choose from.
-This hardly seems to matter,   Our studies show minimal difference between the traditional windows, so the common and reasonable Kaiser-Harris window where the compromize between What windows should the LCA be allowed to choose from?
-How many is needed?
-by Synnev\aa{}g demonstrated the potential of LCA,
 

### 2015-08-14 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:08:14	17:04:42	(Fri14.Aug)

@@
 Synnev\aa{}g used a window set comprised of rectangular, Kaiser and inverted Kaiser functions.
-Of these only a specific Kaiser window were steered We limit our study to the use of variations of the parametric Kaiser-Harris window, since we observed neglible performance difference to using this function or any of the common trigonometric ones.
+Of these a specific Kaiser window were micro-steered.
+We limit our study to the use of variations of the parametric Kaiser-Harris window, since we observed neglible performance difference to using this function or any of the common trigonometric ones.
 The Kaiser windows are easy to compute, is fairly insensitive to coefficient inaccuracies, and can create windows with shapes spanning from rectangular to Gaussian.
@@
 This process is known as beamforming.% The process of combining the signals from each sensor is known as beamforming.
-
-3rd order Maclaurin / Tayler approximation
-\begin{align}
-\phi_\text{-3dB bounds} \approx -0.002894\beta^3 + 0.02451\beta^2 + -0.01308\beta + 0.2980
-\end{align}
 
@@
 oversampled 10 times in azimuth~\cite{Asen2014}.
+
+\subsection{Relative steering}
+
+3rd order Maclaurin / Tayler approximation
+\begin{align}
+\phi_\text{-3dB bounds} \approx -0.002894\beta^3 + 0.02451\beta^2 + -0.01308\beta + 0.2980
+\end{align}
 

### 2015-08-14 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:08:14	19:54:32	(Fri14.Aug)

@@
 \includegraphics[width=\linewidth]{gfx/calc_kaiser_3dB.pdf}%
-\caption{Steering within -3dB for each window}\label{windows_steering}
+\caption{Each Kaiser window is steered in the interval $\phi=[0, \phi_{\text{-3dB}}(\beta)]$.
+The angle $\phi_{\text{-3dB}}(\beta)$ is the amount of steering needed for the steered response to have a -3dB crossing that is exactly half that of the unsteered window.
+With each window steered this way we expect the resolution gain to remain predictable and independent of $\beta$, and we also effectively constrain the white noise gain of the beamformer.}\label{windows_steering}
 \end{figure}
@@
 \end{align}
+
+There is an upper bound to how much a window should be steered.
+Since each window's response is constrained to a unit gain in the look direction, each steered window's response is gained as shown in \Fig{window_steering}.
+moreUnder the constraint of a unit gain in the look direction, the effect of
+When steering the windows it is desirable to limit the amount of steering
+
+Each Kaiser window is steered in the interval $\phi=[0, \phi_{\text{-3dB}}(\beta)]$.
+The angle $\phi_{\text{-3dB}}(\beta)$ is the amount of steering needed for the steered response to have a -3dB crossing that is exactly half that of the unsteered window.
+With each window steered this way we expect the resolution gain to remain predictable and independent of $\beta$, and we also effectively constrain the white noise gain of the beamformer.
 

### 2015-08-14 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:08:14	20:25:22	(Fri14.Aug)

@@
 
-\begin{figure}[tbhp]%\centering%
+\begin{figure}[tbp]%
 \includegraphics[width=\linewidth]{gfx/calc_kaiser_3dB.pdf}%
@@
 
-oversampled 10 times in azimuth~\cite{Asen2014}.
+The steering is subject to an upper steering bound specified, which will be discussed next.
 
@@
 
-3rd order Maclaurin / Tayler approximation
-\begin{align}
-\phi_\text{-3dB bounds} \approx -0.002894\beta^3 + 0.02451\beta^2 + -0.01308\beta + 0.2980
-\end{align}
-
-There is an upper bound to how much a window should be steered.
-Since each window's response is constrained to a unit gain in the look direction, each steered window's response is gained as shown in \Fig{window_steering}.
-moreUnder the constraint of a unit gain in the look direction, the effect of
-When steering the windows it is desirable to limit the amount of steering
-
-Each Kaiser window is steered in the interval $\phi=[0, \phi_{\text{-3dB}}(\beta)]$.
-The angle $\phi_{\text{-3dB}}(\beta)$ is the amount of steering needed for the steered response to have a -3dB crossing that is exactly half that of the unsteered window.
+There is an upper bound to the amount of steering we should apply to a window.
+Since each window's response is constrained to a unit gain in the look direction, each steered window's response is gained as shown in \Fig{windows_steering}.
+This increases the white noise gain and deteriorates the signal to noise ratio.
+To limit this we placing an upper bound to the steering, taking into account that a wide window can be allowed more steering than a narrow one.
+
+We call the upper bound for steering $\phi_{\text{-3dB}}(\beta)$, and define it as the steering angle needed for the steered response to have a -3dB crossing that is exactly half that of the unsteered window.
+We numerically searched for the steering angle $\phi$ that fulfilled this criterion for a wide range of $\beta$-values.
+An explicit expression for was finally obtained with a third degree polynomial fit:
+%
+\begin{align}
+\phi_\text{-3dB}(\beta) \approx -0.002894\beta^3 + 0.02451\beta^2 + -0.01308\beta + 0.2980
+\end{align}
+%
+Any steering will be specified relative to this upper bound.
 With each window steered this way we expect the resolution gain to remain predictable and independent of $\beta$, and we also effectively constrain the white noise gain of the beamformer.
 
-% \begin{align}
-% \z = \w\H\a 
-% \end{align}
-
-% 
 % Narrow ones:
@@
 \end{table}%
+
+oversampled 10 times in azimuth~\cite{Asen2014}.
 

### 2015-08-14 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:08:14	23:16:40	(Fri14.Aug)

@@
 
+A beamformer is a spatial filter we apply to backscattered wavefield data to estimate the scatter strength from a particular point or direction.
+A coarse adjustment is first obtained by applying delays to the sensor channels to negate the difference in travel time from the point of interest, thereby aligning the phase of the signals coming from that source.
+Next a set of weights, or window, is applied to the sensors to Finally a set of weights  , which in effect achieve this by applying delays and weigths to the sensor channels to coherently combine signals from the direction of interest.
+
+Let the array be an $M$ element uniform linear array, 
 By definition, the beamformer output $z[n]$ can be expressed as the weighted sum of all the delayed data samples:

### 2015-08-14 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:08:14	23:48:15	(Fri14.Aug)

@@
 
-A beamformer is a spatial filter we apply to backscattered wavefield data to estimate the scatter strength from a particular point or direction.
-A coarse adjustment is first obtained by applying delays to the sensor channels to negate the difference in travel time from the point of interest, thereby aligning the phase of the signals coming from that source.
+A beamformer is a spatial filter we apply to backscattered wavefield data to perform e.g. source localization and amplitude estimation. In sonar image reconstruction we desire
+
+ A coarse adjustment is usually achieved by applying delays to the sensor channels to negate the difference in travel time from the point of interest, and cause the  thereby aligning the phase of the signals coming from that source.
 Next a set of weights, or window, is applied to the sensors to Finally a set of weights  , which in effect achieve this by applying delays and weigths to the sensor channels to coherently combine signals from the direction of interest.

### 2015-08-17 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:08:17	10:54:49	(Mon17.Aug)

@@
 
-\subsection{Beamforming}
+\subsection{Receive Beamforming}
 
@@
 
-A beamformer is a spatial filter we apply to backscattered wavefield data to perform e.g. source localization and amplitude estimation. In sonar image reconstruction we desire
+To form a sonar image we need to estimate source locations and amplitudes.
+For this purpose we apply a spatial bandpass filter to the backscattered wavefield data, commonly called a receive beamformer.
 

### 2015-08-17 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:08:17	11:59:13	(Mon17.Aug)

@@
 
-Let the receiver be an $M$ element uniform linear array, and assume that the signature of the transmitted signal has been removed by a matched filter.
-Further assume that the array channels are digitally delayed to focus at a pixel with azimuth angle $\theta$ and range sample $n$, such that the delayed data from the $m$th channel can be expressed as $x_m[\theta,n]$.
+To form a sonar image we need to estimate source locations and amplitudes.
+For this purpose we apply a spatial bandpass filter to the backscattered wavefield data.
+The filter is called an array processor, or receive beamformer.
+The basic principle is to apply a delay and weights to the sensor channels before summing them up, chosen such that signals from the location of interest are summed coherently, while other sources sum incoherently.
+Assume that the signature of the transmitted signal has been removed by a matched filter.
+
+Let the receiver be an $M$ element uniform linear array.
+The Assume that the array channels are digitally delayed to focus at a pixel with azimuth angle $\theta$ and range sample $n$, such that the delayed data from the $m$th channel can be expressed as $x_m[\theta,n]$.
 To simplify notation, we make the dependence on $\theta$ implicit from now on.
 
-To form a sonar image we need to estimate source locations and amplitudes.
-For this purpose we apply a spatial bandpass filter to the backscattered wavefield data, commonly called a receive beamformer.
-
-A coarse adjustment is usually achieved by applying delays to the sensor channels to negate the difference in travel time from the point of interest, and cause the  thereby aligning the phase of the signals coming from that source.
-Next a set of weights, or window, is applied to the sensors to Finally a set of weights  , which in effect achieve this by applying delays and weigths to the sensor channels to coherently combine signals from the direction of interest.
-
-Let the array be an $M$ element uniform linear array, 
 By definition, the beamformer output $z[n]$ can be expressed as the weighted sum of all the delayed data samples:
@@
 A large variety of weighting functions exists here for trading lateral resolution for improved noise suppression (contrast), but one always ends up with a compromise between the two~\cite{Harris1978}.
+
+ coarse delay adjustment is usually achieved by applying delays to the sensor channels to negate the difference in travel time from the point of interest, and cause the  thereby aligning the phase of the signals coming from that source.
+Next a set of weights, or window, is applied to the sensors to Finally a set of weights  , which in effect achieve this by applying delays and weigths to the sensor channels to coherently combine signals from the direction of interest.
+
+Let the array be an $M$ element uniform linear array,
 

### 2015-08-17 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:08:17	13:05:27	(Mon17.Aug)

@@
 Let the receiver be an $M$ element uniform linear array.
-The Assume that the array channels are digitally delayed to focus at a pixel with azimuth angle $\theta$ and range sample $n$, such that the delayed data from the $m$th channel can be expressed as $x_m[\theta,n]$.
+Call the the data from the $m$th channel $x_m[\theta,n]$, where the azimuth angle $\theta$ and range sample $n$ is the focus point.
 To simplify notation, we make the dependence on $\theta$ implicit from now on.

### 2015-08-17 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:08:17	14:06:04	(Mon17.Aug)
la2pdf commit - 2015:08:17	14:36:33	(Mon17.Aug)

@@
 
-% correct bad hyphenation here
-% \hyphenation{op-tical net-works semi-conduc-tor}
-
-%%%%%%%%%%%%                                    ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
-% GLOSSARY %
-%%%%%%%%%%%%                                    ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
-
-% \makeglossaries
-% \input{../../Library/glossary}
-
-%%%%%%%%%%%%%%%%%%                              ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
-% DOCUMENT START %
-%%%%%%%%%%%%%%%%%%                              ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
-
-% \usepackage{yfonts}
-
-% \graphicspath{submission/final/gfx/}
-
-% 
-% %%%%%%%%%%                                      ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
-% % MACROS %       
-% %%%%%%%%%%                                      ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
-% 
-% \newcommand\Fig[1]{Fig.~\ref{#1}}
-% 
-% \newcommand\Grey[1]{{\color{Grey}#1}}
-% \newcommand\Red[1]{{\color{Red}#1}}
-% \newcommand\Blue[1]{{\color{Blue}#1}}
-% \newcommand\DarkBlue[1]{{\color{DarkBlue}#1}}
-% \newcommand\LightBlue[1]{{\color{LightBlue}#1}}
-% \newcommand\Brown[1]{{\color{Brown}#1}}
-% \newcommand\Green[1]{{\color{Green}#1}}
-% \newcommand\SeaGreen[1]{{\color{SeaGreen}#1}}
-% \newcommand\Yellow[1]{{\color{yellow}#1}}
-% \newcommand\Orange[1]{{\color{orange}#1}}
-% 
-% \newcommand\nn{\nonumber\\}
-% 
-% \newcommand\nmat[1]{\begin{matrix}#1\end{matrix}}
-% \newcommand\bmat[1]{\begin{bmatrix}#1\end{bmatrix}}
-% \newcommand\case[1]{\begin{cases}#1\end{cases}}
-% \newcommand\textbox[2]{\footnotesize\text{\parbox{#1}{\centering\emph{#2}}}}
-% 
-% \newcommand\rand{\text{rand}}
-% \newcommand\randn{\text{randn}}
-% \newcommand\rect{\text{rect}}
-% \newcommand\sinc{\text{sinc}}
-% \newcommand\tr{\text{tr}}
-% \newcommand\adj{\text{adj}}
-% 
-% % \newcommand\max{\text{max}}
-% \newcommand\argmin{\text{argmin}}
-% 
-% \newcommand\qqquad{\quad\qquad}
-% \newcommand\qqqquad{\qquad\qquad}
-% 
-% \renewcommand\l[1]{\left#1}
-% \renewcommand\r[1]{\right#1}
-% 
-% % {\text{\parbox{1.5cm}{\centering volume hyper- sphere}}}
-% 
-% %Keyword colouring:
-% \newcommand\kw[1]{#1}
-% \newcommand\parm[1]{#1}%\color{Black}#1\color{Black}}
-% 
-% \newcommand\of[1]{\scriptstyle(\parm{#1})\displaystyle}
-% \newcommand\df[1]{\scriptstyle[\parm{#1}]\displaystyle}
-% \newcommand\var[3]{#1_\text{#2}\of{#3}}
-% 
-% \newcommand\diag{\text{diag}}
-% 
-% % \raisebox{lift}[extend-above-baseline][extend-below-baseline]{text}
-% \newcommand\mt[1]{\text{\emph{#1}}} %mt = mathtext
-% \newcommand\mathnorm{\textstyle}
-% \newcommand\mathbig[1]{\displaystyle#1\mathnorm}
-% \newcommand\mathsmall[1]{\scriptstyle#1\mathnorm}
-% \newcommand\mathtiny[1]{\scriptscriptstyle#1\mathnorm}
-% \newcommand\sfrac[2]{\scriptstyle\raisebox{0.25pt}[0pt][0pt]{$\frac{#1}{#2}$}\mathnorm}
-% \newcommand\nfrac[2]{\textstyle\frac{#1}{#2}\displaystyle}
-% 
-% \newcommand\sumu[1]{\sum\limits^{#1}\,}
-% \newcommand\suml[1]{\sum\limits_{#1}\,}
-% \newcommand\sumb[2]{\sum\limits_{#1}^{#2}\,}
-% 
-% \newcommand\produ[1]{\prod\limits^{#1}\,}
-% \newcommand\prodl[1]{\prod\limits_{#1}\,}
-% \newcommand\prodb[2]{\prod\limits_{#1}^{#2}\,}
-% 
-% %Math macros:
-% \newcommand\diff[2]{\frac{\kw{d}\,\textstyle #1\scriptstyle}{\kw{d\parm{#2}}}\displaystyle}
-% \newcommand\ddiff[2]{\frac{\kw{d^2}\,\displaystyle #1\scriptstyle}{\kw{d\parm{#2}}^2}\displaystyle}
-% 
-% \renewcommand\d[1]{\scriptstyle\kw{\,d\parm{#1}}\displaystyle}
-% 
-% % These commands are mutually exclusive. Remember to "renew" in v2.
-% \newcommand\intb[4]{\int\limits_{#3}^{#4} #1 \d{#2}} % \int{exp}{var}{from}{to}
-% \newcommand\intl[3]{\int\limits_{#3} #1 \d{#2}} % \int{exp}{var}{for all}
-% \newcommand\intu[2]{\int #1 \d{#2}} % \int{exp}{var}{for all}
-% 
-% \newcommand\T{^{\scriptscriptstyle T}}
-% \renewcommand\H{^{\scriptscriptstyle H}}
-% 
-% \renewcommand\vec[1]{\boldsymbol{#1}}
-% \newcommand\mat[1]{\boldsymbol{#1}}
-% 
-% 
-% \renewcommand*\a{\vec a}
-% \renewcommand*\i{\vec i}
-% \renewcommand*\k{\vec k}
-% \newcommand*\n{\vec n}
-% \newcommand*\p{\vec p}
-% \newcommand*\s{\vec s}
-% \newcommand*\w{\vec w}
-% \newcommand*\x{\vec x}
-% \newcommand*\y{\vec y}
-% 
-% \newcommand*\A{\mat A}
-% \newcommand*\B{\mat B}
-% \newcommand*\C{\mat C}
-% \newcommand*\E{\mat E}
-% % \renewcommand*\H{\mat H}
-% \renewcommand*\P{\mat P}
-% \newcommand*\eP{\mat{\hat P}}
-% \newcommand*\R{\mat R}
-% \newcommand*\Ri{\R^{-1}}
-% \newcommand*\eR{\mat{\hat R}}
-% \newcommand*\W{\mat W}
-% \newcommand*\X{\mat X}
-% \newcommand*\Xd{\X_{\!\Delta}}
-% \newcommand*\Y{\mat Y}
-% 
-% \renewcommand*\L{\mat \Lambda}
-% \newcommand*\U{\mat U}
-% % \renewcommand*\t{\mathtiny{^T}}
-% % \newcommand*\h{\mathtiny{^H}}
-% \renewcommand*\t{^T}
-% \newcommand*\h{^H}
-% 
-% \newcommand\D{\vec\nabla} %Del: Vector differential operator - nabla
-% \newcommand\Dx{\vec\nabla\times}
-% \newcommand\Dd{\vec\nabla\cdot}
-% 
-% \usepackage{tikz}
-% \usetikzlibrary{shapes,snakes}
-% \usepackage{amsmath,amssymb}
-% 
-% \newenvironment{outline}
-% {\begin{itemize}}
-% {\end{itemize}}
-
-%    \definecolor{todobackground}{rgb}{0.95,0.95,0.95}
-%    \setlength\marginparsep{3pt}
-%    \setlength\marginparwidth{42pt}
-%    \newlength\marginparwidthsmall
-%    \setlength\marginparwidthsmall{\marginparwidth}
-%    \addtolength\marginparwidthsmall{-7pt}
-%    \newcommand\todo[1]{%
-%       \addtocounter{todoidx}{1}%
-%       {\color{Red}\fbox{\bf\thetodoidx{}}}%
-%       \marginpar{%
-%          {\vspace*{-10pt}\color{Red}\fbox{\bf\thetodoidx{}}}\\%
-%          \fcolorbox{red}{todobackground}{\parbox{\marginparwidthsmall}{#1}}}}
-%
-
-% correct bad hyphenation here
-% \hyphenation{op-tical net-works semi-conduc-tor}
-
-%%%%%%%%%%%%%%%%%%                              ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
-% DOCUMENT START %
-%%%%%%%%%%%%%%%%%%                              ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
-
-% \RequirePackage[utf8]{inputenc}%              % Set input encoding (optionally latin1)
-% \RequirePackage[T1]{fontenc}%                 % Set font encoding
-
 \begin{document}
@@
 
-% Adaptive beamformers usually compute the window coefficients by estimating and inverting a spatial covariance matrix.
-There are two inherent problems to this procedure.
-First, for improved estimation accuracy the data is either averaged in space or time, and the significance of the covariance matrix can be reduced using a regularisation parameter~\cite{Carl Inge}.
-These are all attempts to constrain the beamformer to ensure the window coefficience are not over-adapted to the data.
-Second, inverting the covariance matrix has a computational complexity of O($M^3$)~\cite{Carl Inge}.
-For larger arrays the computational burden becomes significant, and the question arises whether this processing power can be better utilised.
-
 The MV beamformer computes the window coefficients by estimating and inverting a spatial covariance matrix.
@@
 To simplify notation, we make the dependence on $\theta$ implicit from now on.
-
 By definition, the beamformer output $z[n]$ can be expressed as the weighted sum of all the delayed data samples:
@@
 where $w_m$ is the weight factor assigned to channel $m$.
+
 With static weights this would be referred to as the conventional delay-and-sum (DAS) beamformer.
 A large variety of weighting functions exists here for trading lateral resolution for improved noise suppression (contrast), but one always ends up with a compromise between the two~\cite{Harris1978}.
-
- coarse delay adjustment is usually achieved by applying delays to the sensor channels to negate the difference in travel time from the point of interest, and cause the  thereby aligning the phase of the signals coming from that source.
-Next a set of weights, or window, is applied to the sensors to Finally a set of weights  , which in effect achieve this by applying delays and weigths to the sensor channels to coherently combine signals from the direction of interest.
-
-Let the array be an $M$ element uniform linear array,
 
@@
 
+We limit our study to the use of variations of the parametric Kaiser-Harris window, since we observed neglible performance difference to using this function or any of the common trigonometric ones.
+The Kaiser windows are easy to compute, is fairly insensitive to coefficient inaccuracies, and can create windows with shapes spanning from rectangular to Gaussian.
+
 % if have a single appendix:

### 2015-08-17 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:08:17	21:24:04	(Mon17.Aug)

@@
 However, the computational complexity of conventional beamformers are linear with the number of channels, O($M$), while MVDR is at O($M^3$).
-This is impracticable and sometimes infeasible.
+This makes MVDR impracticable and sometimes infeasible.
 We have shown in previous studies that MVDR's complexity can be reduced, as well as accelerated significantly using graphics computing units (GPUs)~\cite{Buskenes2014,Asen2013}, but there is a way to avoid it altogether while achieving similar performance.
 
-This method is called Low Complexity Adaptive (LCA) beamformer and was first proposed by Synnev\aa{}g et.
+This method is called the Low Complexity Adaptive (LCA) beamformer.
+It was first proposed by Synnev\aa{}g et.
 al.
@@
 
-\section{Methods}
-
-At heart of the image formation process is a technique called beamforming, which creates a focus point at the pixel of interest.
+% 
+% \section{Methods}
+% 
+% At heart of the image formation process is a technique called beamforming, which creates a focus point at the pixel of interest.
 It does this by applying a suitable set of delays and weights to the array elements, such that signals emanating from the pixel location are emphasized constructively, while noise and interference from other directions sum destructively.
-
-In active sonar array imaging, an acoustic wave is transmitted and the received echoes are usually recorded using an array of sensors.
+% 
+% In active sonar array imaging, an acoustic wave is transmitted and the received echoes are usually recorded using an array of sensors.
 Each of these may be independently delayed and weighted such that signals emanating from directions of interest are added constructively, while noise and interference from other directions add destructively.
@@
 This process is known as beamforming.% The process of combining the signals from each sensor is known as beamforming.
-
-% Traditional beamformers such as Delay-and-Sum (DAS) apply predefined windows to all incoming data.
+% 
+% 
+% 
+% 
+% % Traditional beamformers such as Delay-and-Sum (DAS) apply predefined windows to all incoming data.
 However, due to the non-stationary nature of sonar data, the optimal window for any given time instant will generally differ from the next.
@@
 The Minimum Variance (MV) beamformer, for instance, selects the set of weights that minimizes the beamformer output power for any given time instant~\cite{cap69}.
-
-\section{Background}
-
-The MV beamformer computes the window coefficients by estimating and inverting a spatial covariance matrix.
+% 
+% 
+% \section{Background}
+% 
+% The MV beamformer computes the window coefficients by estimating and inverting a spatial covariance matrix.
 For improved estimation accuracy the data is averaged in space or time, or both.
@@
 The computational burden makes the MV beamformer for large arrays less attractive, and sometimes even infeasible.
-
-Based on the proposed method by Synnevåg~\cite{syn11}, we have implemented the LCA beamformer that keeps the minimum variance optimisation criterion but reduces the solution space to a discrete set of predefined windows.
+% 
+% Based on the proposed method by Synnevåg~\cite{syn11}, we have implemented the LCA beamformer that keeps the minimum variance optimisation criterion but reduces the solution space to a discrete set of predefined windows.
 This reduces the computational complexity to O($MW$), where $W$ is the number of windows in the set.
@@
 In addition we apply steering to each of the windows and constrain the window design to ensure unit gain in the look direction.
-
-The following will show that the proposed method performs similarly to or better than the MV method.
+% 
+% The following will show that the proposed method performs similarly to or better than the MV method.
 

### 2015-08-18 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:08:18	13:57:32	(Tue18.Aug)

@@
 
+\newpage
 \subsection{Receive Beamforming}
@@
 
-Let the receiver be an $M$ element uniform linear array.
-Call the the data from the $m$th channel $x_m[\theta,n]$, where the azimuth angle $\theta$ and range sample $n$ is the focus point.
+Assume that the wavefield has been sampled by an $M$ element uniform linear array, and that the signal signature has been removed by a matched filter.
+The data from the $m$th channel $x_m[\theta,n]$, where the azimuth angle $\theta$ and range sample $n$ is the focus point.
 To simplify notation, we make the dependence on $\theta$ implicit from now on.

### 2015-08-18 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:08:18	14:28:11	(Tue18.Aug)

@@
 Assume that the wavefield has been sampled by an $M$ element uniform linear array, and that the signal signature has been removed by a matched filter.
-The data from the $m$th channel $x_m[\theta,n]$, where the azimuth angle $\theta$ and range sample $n$ is the focus point.
-To simplify notation, we make the dependence on $\theta$ implicit from now on.
-By definition, the beamformer output $z[n]$ can be expressed as the weighted sum of all the delayed data samples:
+Let $x_m[\theta,n]$ be the delayed data from the $m$th channel, where the $\theta$ and $n$ is the azimuth angle and range sample of the focus point, respectively.
+Each angle $\theta$ can be processed independently, so to simplify notation we will assume the dependence on $\theta$ to be implicit from now on.
+
+The beamformer output $z[n]$ can now be expressed as the weighted sum of all the delayed data samples:
 \begin{align}
@@
 where $w_m$ is the weight factor assigned to channel $m$.
+We will refer to a set of weights as a window, but it is also commonly called a taper function.
+Conventional beamformers all have static weights, and by far the most common of these is the delay-and-sum (DAS) beamformer, which does exactly what the name implies.
+
+ See section \ref{kaiser_windows}.
 
@@
 \end{align}
-% \ifPeerReview
-% \begin{figure}[!t]\centering
-% \includegraphics[width=.8\linewidth]{gfx/buske2\figPostfix.eps}
-% \else
-% \begin{figure}[!t]\centering
-% \includegraphics[width=\linewidth]{gfx/implementation.eps}
-% \fi%
-% \caption{MVDR beamforming. For each pixel in range and azimuth:\newline
-% 1. an $L\times{}L$ sample covariance matrix $\eR$ is computed; \newline
-% 2. the term $\eR^{-1}\1$ is found using a linear equation solver;\newline
-% 3. and the beamformer output $z$ is computed from (\ref{finalZ}), where $\w$ is found by substituting $\eR^{-1}\1$ into (\ref{weights}). } \label{mvdr_beamforming}
-% \end{figure}
-As summarized in Fig. \ref{mvdr_beamforming}, the MVDR method is applied to each pixel independently, by:
-\begin{enumerate}
-\item computing the sample covariance matrix $\eR$ in (\ref{finalR});
-\item computing $\eRi\1$ in (\ref{weights});
-\item computing the beamformer output $z$ in (\ref{finalZ}).
-\end{enumerate}
+
 Next we will evaluate these steps in terms of arithmetic complexity, and then discuss their mappability to parallel hardware.
@@
 
-\section{Kaiser windows}
+\section{Kaiser windows}\label{kaiser_windows}
 

### 2015-08-18 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:08:18	19:01:20	(Tue18.Aug)

@@
 We will refer to a set of weights as a window, but it is also commonly called a taper function.
-Conventional beamformers all have static weights, and by far the most common of these is the delay-and-sum (DAS) beamformer, which does exactly what the name implies.
-
- See section \ref{kaiser_windows}.
-
-With static weights this would be referred to as the conventional delay-and-sum (DAS) beamformer.
-A large variety of weighting functions exists here for trading lateral resolution for improved noise suppression (contrast), but one always ends up with a compromise between the two~\cite{Harris1978}.
+Conventional beamformers all have static weights, and the reference here is the delay-and-sum (DAS) beamformer, or the backprojection algortithm.
+As the name implies it delays each pixel into focus, then applies a suitable window, and finally sums the data.
 
@@
 It finds the set of complex weights that minimizes the beamformer's expected output power, while ensuring unity gain in the look direction~\cite{Capon1969}:
+
+See section \ref{kaiser_windows}.
+
+With static weights this would be referred to as the conventional delay-and-sum (DAS) beamformer.
+A large variety of weighting functions exists here for trading lateral resolution for improved noise suppression (contrast), but one always ends up with a compromise between the two~\cite{Harris1978}.
 

### 2015-08-18 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:08:18	19:37:06	(Tue18.Aug)

--- 2015-08-18 LCAforActiveSonarImaging/lca.tex
@@
 The basic principle is to apply a delay and weights to the sensor channels before summing them up, chosen such that signals from the location of interest are summed coherently, while other sources sum incoherently.
-Assume that the signature of the transmitted signal has been removed by a matched filter.
 
@@
 We will refer to a set of weights as a window, but it is also commonly called a taper function.
-Conventional beamformers all have static weights, and by far the most common of these is the delay-and-sum (DAS) beamformer, which does exactly what the name implies.
-
- See section \ref{kaiser_windows}.
-
-With static weights this would be referred to as the conventional delay-and-sum (DAS) beamformer.
-A large variety of weighting functions exists here for trading lateral resolution for improved noise suppression (contrast), but one always ends up with a compromise between the two~\cite{Harris1978}.
-
-Various adaptive beamformers target this limitation by allowing the weights to change for each pixel to better fit the dynamic nature of the incoming wavefield.
-In other words, they attempt to use the \emph{a priori} information present in the data to improve image quality.
-The MVDR beamformer is one such method.
-It finds the set of complex weights that minimizes the beamformer's expected output power, while ensuring unity gain in the look direction~\cite{Capon1969}:
-
+Applying a window that trails off towards the edges improves noise suppression at the cost of reduced resolution.
+Real windows lead to symmetric responses, while complex weights allow asymmetric responses.
+
+Conventional beamformers all have static weights.
+The reference here is the delay-and-sum (DAS) beamformer, also know as the backprojection algortithm.
+It delays each pixel into focus, then applies a suitable window, and finally sums the data.
+The virtue of DAS is its simplicity, robustness to parameter errors, linear processing of the image and the ease of which it can be implemented on parallel hardware.
+
+Adaptive beamformers are generally not so simple.
+They allow the weights/delays to change to adjust the array response to better fit the incoming wavefield.
+Of these methods the minimum variance distortionless response (MVDR) beamformer is perhaps the best studied and well known.
+It seeks to minimize the power of the beamformer, under the constraint of a unity gain in the desired direction $\a$~\cite{Capon1969}:
+%
 \begin{align}
 \underset{\w[n]}{\argmin}\, E\{|z[n]|^2\} &= \underset{\w[n]}{\argmin}\, \w[n]\x\x\H\w\H[n]\\
-\text{subject to } \w[n]\1 &= 1
-\end{align}
-
-This convex optimization problem has the solution:
-
+\text{subject to } \w[n]\a &= 1
+\end{align}
+%
+where $\a$ is a steering vector. This is a convex optimization problem with the solution:
+%
 \begin{gather}
@@
 \end{gather}
-where $\a$ is a steering vector and $\R=E\{\x[n]\x\H[n]\}\in\mathbb{C}^{M,M}$ is the spatial covariance matrix for the full array.
-Since we pre-steer our data to every pixel in the image we simplify (\ref{weights}) by substituting $\a$ with a row vector $\1$ that represents broadside phase-steering.
-To estimate $\R$ we compute a sample covariance matrix $\eR$.
-In this computation we perform some degree of:
-\begin{itemize}
-\item \emph{spatial averaging} to avoid signal cancellation by decorrelating coherent echoes~\cite{Kailath1985};
-\item \emph{temporal averaging} over an interval comparable to the pulse length (one to five samples) to maintain true speckle statistics~\cite{Synnevag2009a};
-\item \emph{diagonal loading} to improve robustness to parameter errors~\cite{Cox1987,Maksym1979}.
-\end{itemize}%
+where $\R=E\{\x[n]\x\H[n]\}\in\mathbb{C}^{M,M}$ is the spatial covariance matrix for the full array.
+Estimating and inverting this spatial covariance matrix is a non trivial matter.
+Spatial averaging is needed to avoid signal cancellation~\cite{Kailath1985}, temporal averaging is needed to obtain true speckle statistics~\cite{Synnevag2009a}, and diagonal loading is needed to improve robustness to parameter errors~\cite{Cox1987,Maksym1979}.
+
+See section \ref{kaiser_windows}.
+
+.
+A large variety of weighting functions exists here for trading lateral resolution for improved noise suppression (contrast), but one always ends up with a compromise between the two~\cite{Harris1978}.
+
 %

### 2015-08-18 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:08:18	21:57:03	(Tue18.Aug)

@@
 Of these methods the minimum variance distortionless response (MVDR) beamformer is perhaps the best studied and well known.
-It seeks to minimize the power of the beamformer, under the constraint of a unity gain in the desired direction $\a$~\cite{Capon1969}:
+
+\section{MVDR}
+
+MVDR seeks to minimize the power of the beamformer, under the constraint of a unity gain in the desired direction $\a$~\cite{Capon1969}:
 %
@@
 where $\R=E\{\x[n]\x\H[n]\}\in\mathbb{C}^{M,M}$ is the spatial covariance matrix for the full array.
-Estimating and inverting this spatial covariance matrix is a non trivial matter.
+The problem lies in estimating and inverting this spatial covariance matrix.
 Spatial averaging is needed to avoid signal cancellation~\cite{Kailath1985}, temporal averaging is needed to obtain true speckle statistics~\cite{Synnevag2009a}, and diagonal loading is needed to improve robustness to parameter errors~\cite{Cox1987,Maksym1979}.
+The steps are also needed to ensure that the covariance matrix is numerically well conditioned and hence invertible.
 
@@
 
-%
-% will perform some degree of \emph{spatial averaging} to avoid signal cancellation caused by coherent echoes~\cite{Kailath1985}, \emph{temporal averaging} to maintain true speckle statistics~\cite{Synnevag2009a}, and \emph{diagonal loading} to improve robustness to parameter errors ~\cite{Cox1987,Maksym1979}.
-%
-Combined, these steps will also ensure a numerically well conditioned $\eR$.
-
-We perform \emph{temporal} and \emph{spatial averaging} first and put the result in an intermediate sample covariance matrix $\breve{\R}$.
-To do this we need to segment our array into subarrays.
-If we let $x_l[n]$ represent the data vector from subarray $l$ by
-\begin{gather}
-\x_l[n] = \bmat{x_l[n] & x_{l+1}[n] & \dots & x_{l+L-1}[n]}\T,
-\end{gather}
-then $\breve{\R}$ can be calculated as
-\begin{gather}
-\breve{\R}[n] =  \frac{1}{N_K N_L} \sumb{l=0}{M-L}\sumb{n'=n-K}{n+K} \x_l[n']\x_l\H[n'] \in\mathbb{C}^{L,L},\label{spatialR}
-\end{gather}
-where $N_K = 2K+1$ is the number of temporal samples to perform averaging over, and $N_L = M-L+1$ is the number of subarrays.
-
-The final estimate $\eR$ is found by adding a fraction $d$ of the total power of $\breve{\R}[n]$ to its diagonal~\cite{Synnevag2007}:
-\begin{align}
-\eR[n] = \breve{\R}[n] + \I \frac{d}{L} \tr\{\breve{\R}[n]\},\label{finalR}
-\end{align}
-where $\I$ is an identity matrix, $\tr\{\cdot\}$ represents the matrix trace operation, and $\tr\{\breve{\R}[n]\}$ is an estimate of the energy received from this pixel.
-
-Note how subarray averaging led to a size reduction of $\eR$ from $\mathbb{C}^{M,M}$ to $\mathbb{C}^{L,L}$, and hence will produce an $L$-element weight set when substituted into (\ref{z}).
-This weight set is applied to all the subarrays, before computing the beamformer output as in (\ref{z}).
-Or, equivalently, we may apply the weight set to the sum of all the subarrays by
-\begin{align}
-z[n] = \w\H[n] \sumb{l=0}{M-L} \x_l[n].\label{finalZ}
-\end{align}
-
-Next we will evaluate these steps in terms of arithmetic complexity, and then discuss their mappability to parallel hardware.
+\section{LCA}
 
@@
 \end{itemize}
+
+Let $\X[n]$ be the data received by an array, pre-steered to all pixels in the image, and $\w$ be the window applied to these spatial data.
+Then we defined the beamformer output $z$ as the weighted sum of the data:
+%
+\begin{align}
+z[\w,n] = \w\H\X[n] = \sumb{m=0}{M-1} w_m x_m^*[n]
+\end{align}
+%
+From a set of $P$ windows the LCA beamformer selects the window $p$ that yield the least output power:
+%
+\begin{align}
+\argmin{p}{E\big\{|z|^2\big\}} = \argmin{p}{E\big\{|\w_p\H\X[n]|^2\big\}} \qquad\text{subject to}\qquad \w\H\a = 1
+\end{align}
+%
+We can estimate $E\big\{|\w_p\H\X[n]|^2\big\}$ as
+%
+\begin{align*}
+\sigma^2_{z_p}[n] = \frac{1}{N_k} \sumb{n'=n-K}{n+K} | z_p[n'] |^2
+\end{align*}
+%
+To add some extra flexibility, we can also estimate the variance by a weighted combination of local pixels
+%
+\begin{align*}
+\sigma^2_{z_p}[x,n] = \frac{1}{N_x N_k} \sumb{x'=x-X}{x+X} \sumb{n=n-K}{n+K} w_\text{prf}[x',n']\big| z_p[x',n'] \big|^2
+\end{align*}
+%
+where $w_\text{prf}[x',n']$ is a normalized 2 dimensional weight function.
+
+Kaiser:
+%
+\begin{align*}
+w[n] = \frac{I_0\big( \pi\alpha\sqrt{1-(\frac{2n}{N-1}-1)^2}\big)}{I_0(\pi\alpha)}
+\end{align*}
+%
+where $I_0$ is a zeroth order modified Bessel function of the first kind.
+
+\subsection{Oversampling and steering}
+
+Steering wildly is not a good idea.
+
+Too much steering, beamformer becomes too agressive. Uncertainty in source amplitude grows.
+
+\begin{itemize}
+\item More steering -> more aggressive suppression of interference at the cost of boosted sidelobes / noise.
+\item More steering -> Larger variations in amplitude of sources
+\end{itemize}
+
+Idea to steer within a fraction of the rectangular window's mainlobe.
+
+\begin{align*}
+d = \frac{\pi\,\theta_\text{open}\,\lambda}{M\,d\,O_s}
+\end{align*}
 
@@
 \end{align}
-If $\eR$ is the estimate of $\R$, the solution to the MVDR beamformer is
-\begin{gather*}
-\vec w[n] = \frac{\hat{\mat R}\,\!^{-1}[n]\vec a}{\vec a\H\hat{\mat R}\,\!^{-1}[n]\vec a} = \frac{\text{\raisebox{1.9pt}{$\vec\chi$}}[n]}{\vec a\H\text{\raisebox{1.9pt}{$\vec\chi$}}[n]}
-\end{gather*}
-A robust estimate of $\eR$ is found by
-\begin{itemize}
-\item Spatial averaging to decorrelate noise from signal. \emph{Subarray averaging} used here.
-\item Temporal averaging to ensure valid speckle statistics.
-\item \emph{Diagonal loading} to ensure  numerical stability prior to inversion.
-\item Choice of robustification parameters \cite{Synnevag2007}
-\end{itemize}
-
-Here $\vec a$ is a \emph{steering vector} that if set to $\vec 1$ steers towards broadside, and $\hat{\mat R}$ is an estimate of the spatial covariance matrix, often computed as
-\begin{gather*}
-r_{ij}[n] = \frac{1}{K(2N_{\text{avg}}+1)}\sumb{k=0}{K-1} \sumb{n'=n-N_{\text{avg}}}{n+N_{\text{avg}}} x_{i+k}[n']\,x_{j+k}[n'],
-\end{gather*}
-where $r_{ij}$ is the $(i,j)$'th element of the covariance matrix, $K$ is the number of subarrays, $2N_{\text{avg}}+1$ is the number of samples to perform time averaging over, and $\vec x_i[n']$ is the data recorded by the $i$'th sensor at sample $n'$.
-
-\emph{Diagonal loading} chosen as \cite{Synnevag2007}
-\begin{align}
-\varepsilon = \Delta \cdot \tr\{\eR[n]\}
-\end{align}
 

### 2015-08-19 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:08:19	10:32:22	(Wed19.Aug)

@@
 We will refer to a set of weights as a window, but it is also commonly called a taper function.
-Applying a window that trails off towards the edges improves noise suppression at the cost of reduced resolution.
+Applying a window that trails off towards the edges improves noise suppression at the cost of reduced resolution~\cite{Harris1978}.
 Real windows lead to symmetric responses, while complex weights allow asymmetric responses.
@@
 \begin{align}
-\underset{\w[n]}{\argmin}\, E\{|z[n]|^2\} &= \underset{\w[n]}{\argmin}\, \w[n]\x\x\H\w\H[n]\\
+\underset{\w[n]}{\argmin}\, E\{|z[n]|^2\} &= \underset{\w[n]}{\argmin}\, \w[n]\R[n]\w\H[n]\\
 \text{subject to } \w[n]\a &= 1
@@
 %
-where $\a$ is a steering vector. This is a convex optimization problem with the solution:
+where $\a$ is a steering vector and $\R[n] = E\{\x\x\H\}$ is the spatial covariance matrix. This is a convex optimization problem with the solution:
 %
@@
 
-See section \ref{kaiser_windows}.
-
-.
-A large variety of weighting functions exists here for trading lateral resolution for improved noise suppression (contrast), but one always ends up with a compromise between the two~\cite{Harris1978}.
-
 \section{LCA}
+
+From a set of $P$ windows the LCA beamformer selects the window $p$ that yields the least output power:
+%
+\begin{align}
+\underset{p}{\argmin}\, E\{|z[n]|^2\} = \underset{p}{\argmin}\, E\big\{|\w_p\H\x[n]|^2\big\}\\
+\qquad\text{subject to}\qquad \w\H\a = 1
+\end{align}
+%
+We can estimate $E\big\{|\w_p\H\x[n]|^2\big\}$ as
+%
+\begin{align*}
+\sigma^2_{z_p}[n] = \frac{1}{N_k} \sumb{n'=n-K}{n+K} | z_p[n'] |^2
+\end{align*}
+%
+To add some extra flexibility, we can also estimate the variance by a weighted combination of local pixels
+%
+\begin{align*}
+\sigma^2_{z_p}[x,n] = \frac{1}{N_x N_k} \sumb{x'=x-X}{x+X} \sumb{n=n-K}{n+K} w_\text{prf}[x',n']\big| z_p[x',n'] \big|^2
+\end{align*}
+%
+where $w_\text{prf}[x',n']$ is a normalized 2 dimensional weight function.
 
@@
 
-Let $\X[n]$ be the data received by an array, pre-steered to all pixels in the image, and $\w$ be the window applied to these spatial data.
-Then we defined the beamformer output $z$ as the weighted sum of the data:
-%
-\begin{align}
-z[\w,n] = \w\H\X[n] = \sumb{m=0}{M-1} w_m x_m^*[n]
-\end{align}
-%
-From a set of $P$ windows the LCA beamformer selects the window $p$ that yield the least output power:
-%
-\begin{align}
-\argmin{p}{E\big\{|z|^2\big\}} = \argmin{p}{E\big\{|\w_p\H\X[n]|^2\big\}} \qquad\text{subject to}\qquad \w\H\a = 1
-\end{align}
-%
-We can estimate $E\big\{|\w_p\H\X[n]|^2\big\}$ as
-%
-\begin{align*}
-\sigma^2_{z_p}[n] = \frac{1}{N_k} \sumb{n'=n-K}{n+K} | z_p[n'] |^2
-\end{align*}
-%
-To add some extra flexibility, we can also estimate the variance by a weighted combination of local pixels
-%
-\begin{align*}
-\sigma^2_{z_p}[x,n] = \frac{1}{N_x N_k} \sumb{x'=x-X}{x+X} \sumb{n=n-K}{n+K} w_\text{prf}[x',n']\big| z_p[x',n'] \big|^2
-\end{align*}
-%
-where $w_\text{prf}[x',n']$ is a normalized 2 dimensional weight function.
-
-Kaiser:
-%
-\begin{align*}
-w[n] = \frac{I_0\big( \pi\alpha\sqrt{1-(\frac{2n}{N-1}-1)^2}\big)}{I_0(\pi\alpha)}
-\end{align*}
-%
-where $I_0$ is a zeroth order modified Bessel function of the first kind.
-
-\subsection{Oversampling and steering}
-
-Steering wildly is not a good idea.
-
-Too much steering, beamformer becomes too agressive. Uncertainty in source amplitude grows.
-
-\begin{itemize}
-\item More steering -> more aggressive suppression of interference at the cost of boosted sidelobes / noise.
-\item More steering -> Larger variations in amplitude of sources
-\end{itemize}
-
-Idea to steer within a fraction of the rectangular window's mainlobe.
-
-\begin{align*}
-d = \frac{\pi\,\theta_\text{open}\,\lambda}{M\,d\,O_s}
-\end{align*}
-
 \section{Kaiser windows}\label{kaiser_windows}
@@
 With each window steered this way we expect the resolution gain to remain predictable and independent of $\beta$, and we also effectively constrain the white noise gain of the beamformer.
+
+\subsection{Oversampling and steering}
+
+Steering wildly is not a good idea.
+
+Too much steering, beamformer becomes too agressive. Uncertainty in source amplitude grows.
+
+\begin{itemize}
+\item More steering -> more aggressive suppression of interference at the cost of boosted sidelobes / noise.
+\item More steering -> Larger variations in amplitude of sources
+\end{itemize}
+
+Idea to steer within a fraction of the rectangular window's mainlobe.
+
+\begin{align*}
+d = \frac{\pi\,\theta_\text{open}\,\lambda}{M\,d\,O_s}
+\end{align*}
+
+\begin{align}
+\w[\beta, \phi] = \text{kaiser}(\beta) = \bmat{
+1 \\
+e^{-j\frac{md}{\lambda}\sin(\phi)} \\
+\vdots\\
+e^{-j\frac{(M-1)d}{\lambda}\sin(\phi)}
+}
+\end{align}
+
+See section \ref{kaiser_windows}.
 

### 2015-08-19 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:08:19	13:19:10	(Wed19.Aug)
la2pdf commit - 2015:08:19	13:49:31	(Wed19.Aug)
la2pdf commit - 2015:08:19	14:37:03	(Wed19.Aug)

@@
 
-\begin{figure}[tbp]%
-\includegraphics[width=\linewidth]{gfx/calc_kaiser_3dB.pdf}%
-\caption{Each Kaiser window is steered in the interval $\phi=[0, \phi_{\text{-3dB}}(\beta)]$.
-The angle $\phi_{\text{-3dB}}(\beta)$ is the amount of steering needed for the steered response to have a -3dB crossing that is exactly half that of the unsteered window.
-With each window steered this way we expect the resolution gain to remain predictable and independent of $\beta$, and we also effectively constrain the white noise gain of the beamformer.}\label{windows_steering}
-\end{figure}
-
 \newpage
@@
 \underset{\w[n]}{\argmin}\, E\{|z[n]|^2\} &= \underset{\w[n]}{\argmin}\, \w[n]\R[n]\w\H[n]\\
-\text{subject to } \w[n]\a &= 1
-\end{align}
-%
-where $\a$ is a steering vector and $\R[n] = E\{\x\x\H\}$ is the spatial covariance matrix. This is a convex optimization problem with the solution:
+\text{subject to } \w[n]\a_\phi &= 1
+\end{align}
+%
+where $\a_\phi$ is a steering vector and $\R[n] = E\{\x\x\H\}$ is the spatial covariance matrix. This is a convex optimization problem with the solution:
 %
@@
 
-From a set of $P$ windows the LCA beamformer selects the window $p$ that yields the least output power:
-%
-\begin{align}
-\underset{p}{\argmin}\, E\{|z[n]|^2\} = \underset{p}{\argmin}\, E\big\{|\w_p\H\x[n]|^2\big\}\\
+The LCA beamformer interates through a set of $P$ windows and $p$ that yields the least output power:
+%
+\begin{align}
+\underset{p}{\argmin}\, E\{|z[n]|^2\} = \underset{p}{\argmin}\, E\big\{|\w_{\beta,\phi}\H\x[n]|^2\big\}\\
 \qquad\text{subject to}\qquad \w\H\a = 1
-\end{align}
+\end{align}\label{lca_criterion}
 %
@@
 \begin{align}
-f_m = \begin{cases}
-\frac{I_0\left(\pi\beta\sqrt{1-\left(\frac{2m}{M-1}-1\right)^2}\right)}{I_0(\pi\beta)}, & 0\le m\le M-1 \\
-0 & \text{otherwise},
-\end{cases}
-\end{align}
-%
-where $I_0$ is the zeroth order Modified Bessel function of the first kind:
+\f[\beta] = \bmat{
+f_0[\beta] \\
+\vdots\\
+f_{M-1}[\beta]
+},
+\end{align}
+%
+where
+%
+\begin{align}
+f_m = \frac{I_0\left(\pi\beta\sqrt{1-\left(\frac{2m}{M-1}-1\right)^2}\right)}{I_0(\pi\beta)}
+\end{align}
+%
+and $I_0$ is the zeroth order modified Bessel function of the first kind:
 %
@@
 
-Before we apply the window to the sensor data, we normalize it to ensure unit gain in the direction of interest:
-%
-\begin{align}
-\w = \frac{\vec f}{\f\T\1} \quad\quad\text{where}\quad\quad \f =
-\bmat{
-f_0 \\ \vdots \\ f_{M-1}                                                           
-}
-\end{align}
-%
-Note that the normalization factor $\f\T\1$ is the reciprocal of the window's coherent gain.
-Since the signal-to-noise gain is constant for a given window, this normalization also proportionally increases the  incoherent noise gain.
-
-\begin{align}
-\w = \frac{\a\f}{\f\T\a} = 
-\end{align}
-
-\begin{align}
-\a = \bmat{
+\begin{figure}[tbp]%
+\includegraphics[width=\linewidth]{gfx/calc_kaiser_3dB.pdf}%
+\caption{Each Kaiser window is steered in the interval $\phi=[0, \phi_{\text{-3dB}}(\beta)]$.
+The angle $\phi_{\text{-3dB}}(\beta)$ is the amount of steering needed for the steered response to have a -3dB crossing that is exactly half that of the unsteered window.
+With each window steered this way we expect the resolution gain to remain predictable and independent of $\beta$, and we also effectively constrain the white noise gain of the beamformer.}\label{windows_steering}
+\end{figure}
+
+\subsection{Steering}
+
+\begin{align}
+\w_{\beta,\phi} = \f_\beta\H\diag{(\a_\phi)}\\
+\end{align}
+%
+where
+%
+\begin{align}
+\a_\phi = \bmat{
 1 \\
@@
 
+Before we apply the window to the sensor data, we normalize it to ensure unit gain in the direction of interest:
+%
+\begin{align}
+\w = \frac{\vec f}{\f\T\1} \quad\quad\text{where}\quad\quad \f =
+\bmat{
+f_0 \\ \vdots \\ f_{M-1}                                                           
+}
+\end{align}
+%
+Note that the normalization factor $\f\T\1$ is the reciprocal of the window's coherent gain.
+Since the signal-to-noise gain is constant for a given window, this normalization also proportionally increases the  incoherent noise gain.
+
+\begin{align}
+\w = \frac{\a\f}{\f\T\a} = 
+\end{align}
+
 The steering is subject to an upper steering bound specified, which will be discussed next.
 
-\subsection{Relative steering}
+\subsection{Steering bounds}
 
@@
 \subsection{Oversampling and steering}
-
-Steering wildly is not a good idea.
 

### 2015-08-20 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:08:19	16:08:56	(Wed19.Aug)
la2pdf commit - 2015:08:20	15:01:26	(Thu20.Aug)

@@
 \underset{p}{\argmin}\, E\{|z[n]|^2\} = \underset{p}{\argmin}\, E\big\{|\w_{\beta,\phi}\H\x[n]|^2\big\}\\
-\qquad\text{subject to}\qquad \w\H\a = 1
+\qquad\text{subject to}\qquad \w\H\a = 1\label{eq_lca_constraint}
 \end{align}\label{lca_criterion}
@@
 
-\section{Kaiser windows}\label{kaiser_windows}
+\subsection{Window function: Kaiser}\label{kaiser_windows}
 
@@
 
-\begin{align}
-\w_{\beta,\phi} = \f_\beta\H\diag{(\a_\phi)}\\
+To allow LCA some added flexibility we also create variations of each window that are steered a little to each side.
+%
+\begin{align}
+\w_{\beta,\phi} = \frac{\f_\beta\H\diag{(\a_\phi)}}{\f_\beta\H\a_\phi}\\
 \end{align}
@@
 e^{-j\frac{(M-1)d}{\lambda}\sin(\phi)}
-}
-\end{align}
-
-Before we apply the window to the sensor data, we normalize it to ensure unit gain in the direction of interest:
-%
-\begin{align}
-\w = \frac{\vec f}{\f\T\1} \quad\quad\text{where}\quad\quad \f =
-\bmat{
-f_0 \\ \vdots \\ f_{M-1}                                                           
-}
-\end{align}
-%
-Note that the normalization factor $\f\T\1$ is the reciprocal of the window's coherent gain.
+}.
+\end{align}
+%
+Here $d$ is the element spacing and $\lambda$ is the wavelength.
+The normalization factor $\f_\beta\H\a_\phi$ is the reciprocal of the window's coherent gain.
+It is needed to ensures unit gain in the direction of interest as required by (\ref{eq_lca_constraint}).
 Since the signal-to-noise gain is constant for a given window, this normalization also proportionally increases the  incoherent noise gain.
-
-\begin{align}
-\w = \frac{\a\f}{\f\T\a} = 
-\end{align}
 

### 2015-08-20 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:08:20	15:33:07	(Thu20.Aug)

@@
 %
-\begin{align}
-\f[\beta] = \bmat{
-f_0[\beta] \\
+\begin{align*}
+\f_\beta = \bmat{
+f_0(\beta) \\
 \vdots\\
-f_{M-1}[\beta]
-},
-\end{align}
+f_{M-1}(\beta)
+}
+\end{align*}
 %
@@
 \begin{align}
-f_m = \frac{I_0\left(\pi\beta\sqrt{1-\left(\frac{2m}{M-1}-1\right)^2}\right)}{I_0(\pi\beta)}
+f_m(\beta) = \frac{I_0\left(\pi\beta\sqrt{1-\left(\frac{2m}{M-1}-1\right)^2}\right)}{I_0(\pi\beta)}
 \end{align}
@@
 The normalization factor $\f_\beta\H\a_\phi$ is the reciprocal of the window's coherent gain.
-It is needed to ensures unit gain in the direction of interest as required by (\ref{eq_lca_constraint}).
-Since the signal-to-noise gain is constant for a given window, this normalization also proportionally increases the  incoherent noise gain.
-
-The steering is subject to an upper steering bound specified, which will be discussed next.
+It ensures unit gain in the direction of interest as required by (\ref{eq_lca_constraint}).
+Since the signal-to-noise ratio is constant for a specific window (same $\beta$-value), this normalization also proportionally increases the incoherent noise gain.
 
@@
 We call the upper bound for steering $\phi_{\text{-3dB}}(\beta)$, and define it as the steering angle needed for the steered response to have a -3dB crossing that is exactly half that of the unsteered window.
-We numerically searched for the steering angle $\phi$ that fulfilled this criterion for a wide range of $\beta$-values.
-An explicit expression for was finally obtained with a third degree polynomial fit:
+We performed a numeric search for the steering angle $\phi$ that fulfilled this criterion for a wide range of $\beta$-values.
+A third degree polynomial fit to these points is:
 %
@@
 \includegraphics[width=\linewidth]{gfx/mvdr_selected_windows_holmengraa.pdf}%
-\caption{Typical frequency responses for windows computed by a robust MVDR.
-The image is of the oiltanker Holmengraa, data collected from the 32 element HISAS1030 sonar with , operated in ssectorscan mode with an oversampling factor of ~10.
-The window responses was computed from 40x40 pixel groups in the shadow, highlight and speckle region.
+\caption{Typical frequency responses for windows computed by a robust MVDR.\newline
+Left:\hfill
+\parbox[t]{.95\linewidth}{A sectorscan image of the oiltanker Holmengraa, using data collected from the 32 element HISAS1030 sonar attached to the HUGIN AUV.
+Image reconstructed with MVDR with $K=1$ temporal oversampling, $L=16$ subarray size and $d=1\%$ This is a sectorscan image with full opening angle 8 times lateral oversampling.}\newline
+Right:\hfill
+\parbox[t]{.95\linewidth}{The window responses was computed from 40x40 pixel groups in the shadow, highlight and speckle region.
 The dashed red lines at -13dB and -43dB marks the peak sidelobe level of a rectangular and Hamming window, respectively.
@@
 Full opening angle.
-Parameters.
-}\label{mvdr_selected_windows}
+Parameters.} }\label{mvdr_selected_windows}
 \end{figure*}

### 2015-08-20 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:08:20	22:20:19	(Thu20.Aug)

@@
 \caption{Typical frequency responses for windows computed by a robust MVDR.\newline
-Left:\hfill
-\parbox[t]{.95\linewidth}{A sectorscan image of the oiltanker Holmengraa, using data collected from the 32 element HISAS1030 sonar attached to the HUGIN AUV.
-Image reconstructed with MVDR with $K=1$ temporal oversampling, $L=16$ subarray size and $d=1\%$ This is a sectorscan image with full opening angle 8 times lateral oversampling.}\newline
-Right:\hfill
-\parbox[t]{.95\linewidth}{The window responses was computed from 40x40 pixel groups in the shadow, highlight and speckle region.
-The dashed red lines at -13dB and -43dB marks the peak sidelobe level of a rectangular and Hamming window, respectively.
+\emph{Left:}\hfill
+\parbox[t]{.95\linewidth}{MVDR sectorscan image of the oiltanker Holmengraa, using data collected from the 32 element HISAS1030 sonar attached to the HUGIN AUV.
+MVDR was run with $K=1$ temporal oversampling, $L=16$ subarray size, $d=1\%$ diagonal loading and 8 times lateral oversampling.}\newline
+\emph{Right:}\hfill
+\parbox[t]{.95\linewidth}{Window responses computed from 40x40 pixel groups in the shadow, highlight and speckle region of the Holmengraa image.
+The dashed red lines at -13dB and -43dB marks the peak sidelobe levels of a rectangular and Hamming window, respectively.
 Note how the responses are more or less symmetric, with very little steering in the shadow, moderate steering in speckle and steering within roughly 3dB in highlight.
-The phase varies mostly in highlight.
-Full opening angle.
-Parameters.} }\label{mvdr_selected_windows}
+The phase varies mostly in highlight.} }\label{mvdr_selected_windows}
 \end{figure*}
@@
 To test the performance of the MVDR and LCA beamformers, we have processed data aquired by the Kongsberg Maritime HUGIN AUV carrying their 32 element HISAS1030 sonar~\cite{Hansen2009}.
-It is a high resolution synthetic aperture sonar with array length 1.2\;m, 100\;kHz operating frequency, a bandwidth of 40\;kHz and transmits at a 25$^\circ$ opening angle.
+It is a high resolution synthetic aperture sonar with array length 1.2\;m, 100\;kHz operating frequency, a bandwidth of 40\;kHz and 25$^\circ$ opening angle.
 

### 2015-08-21 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:08:21	15:21:10	(Fri21.Aug)

@@
 
+Adaptive beamformers process each pixel in a different way in an attempt to surpass the conventional limit of resolution and noise suppression.
+This non-linear processing of the image increases its bandwidth and introduces a need for oversampling.
+For MVDR a lateral oversampling factor of up to 10 times is often needed to ensure spatial shift-invariance~\cite{Asen2014}.
+
 Too much steering, beamformer becomes too agressive. Uncertainty in source amplitude grows.
@@
 \parbox[t]{.95\linewidth}{Window responses computed from 40x40 pixel groups in the shadow, highlight and speckle region of the Holmengraa image.
+The responses are overlayed each other.
 The dashed red lines at -13dB and -43dB marks the peak sidelobe levels of a rectangular and Hamming window, respectively.
@@
 
-\newpage
+\begin{figure*}[tbhp!]%\centering%
+\includegraphics[width=\textwidth]{gfx/oversampling_mosaic_bounds_lca_windows_beta.pdf}%
+\caption{Window parameter boundaries. Oversampled image, 19x20 windows.}\label{oversampling_mosaic_bounds}
+\end{figure*}
 
@@
 $\beta=0$ ) and a gaussian curve (
-
-\begin{figure*}[tbhp!]%\centering%
-\includegraphics[width=\textwidth]{gfx/oversampling_mosaic_bounds_lca_windows_beta.pdf}%
-\caption{Window parameter boundaries. Oversampled image, 19x20 windows.}\label{oversampling_mosaic_bounds}
-\end{figure*}
 
@@
 \end{table}%
-
-oversampled 10 times in azimuth~\cite{Asen2014}.
 

### 2015-08-21 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:08:21	15:51:54	(Fri21.Aug)

@@
 \begin{figure}[tbp]%
-\includegraphics[width=\linewidth]{gfx/calc_kaiser_3dB.pdf}%
+\includegraphics[width=\linewidth]{gfx/calc_kaiser_3dB.eps}%
 \caption{Each Kaiser window is steered in the interval $\phi=[0, \phi_{\text{-3dB}}(\beta)]$.
@@
 
-\begin{figure*}[tbhp!]\centering%
-\includegraphics[width=\linewidth]{gfx/mvdr_selected_windows_holmengraa.pdf}%
+\begin{figure*}[t]\centering%
+\includegraphics[width=\linewidth]{gfx/mvdr_selected_windows_holmengraa.eps}%
 \caption{Typical frequency responses for windows computed by a robust MVDR.\newline
@@
 
-\begin{figure*}[tbhp!]%\centering%
-\includegraphics[width=\textwidth]{gfx/oversampling_mosaic_bounds_lca_windows_beta.pdf}%
-\caption{Window parameter boundaries. Oversampled image, 19x20 windows.}\label{oversampling_mosaic_bounds}
+\begin{figure*}[t]%\centering%
+\includegraphics[width=\textwidth]{gfx/oversampling_mosaic_bounds_lca_windows_beta.eps}%
+\caption{Window parameter boundaries. Oversampled image, 19x20 windows.}\label{oversampling_mosaic_bounds_beta}
 \end{figure*}
 
+\begin{figure*}[t]%\centering%
+\includegraphics[width=\textwidth]{gfx/oversampling_mosaic_bounds.eps}%
+\caption{Window parameter boundaries.
+LCA images using an oversampled image and a relatively large window database with $N_\beta=20$ window types, each steered $N_\phi=19$.
+}\label{oversampling_mosaic_bounds}
+\end{figure*}
+
+\begin{figure*}[t]%\centering%
+\includegraphics[width=\textwidth]{gfx/oversampling_mosaic.eps}%
+\caption{Window parameter boundaries. Oversampled image, 19x20 windows.}\label{oversampling_mosaic}
+\end{figure*}
+
+\newpage
 \section{Results and Discussion}
@@
 $\beta=0$ ) and a gaussian curve (
-
-\newpage
-\begin{figure*}[tbhp!]%\centering%
-\includegraphics[width=\textwidth]{gfx/oversampling_mosaic_bounds.pdf}%
-\caption{Window parameter boundaries.
-LCA images using an oversampled image and a relatively large window database with $N_\beta=20$ window types, each steered $N_\phi=19$.
-}\label{oversampling_mosaic_bounds_beta}
-\end{figure*}
 
@@
 What's wrong here?
-
-\begin{figure*}[tbhp!]%\centering%
-\includegraphics[width=\textwidth]{gfx/oversampling_mosaic.pdf}%
-\caption{Window parameter boundaries. Oversampled image, 19x20 windows.}\label{oversampling_mosaic}
-\end{figure*}
 
@@
 
-\todopar{Threw some formulas in to get a feel of them}
+% \todopar{Threw some formulas in to get a feel of them}
 

### 2015-08-21 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:08:21	17:50:27	(Fri21.Aug)

@@
 %       papers and their use requires extra complexity and work.
-\newcommand\MYhyperrefoptions{bookmarks=true,bookmarksnumbered=true,
-pdfpagemode={UseOutlines},plainpages=false,pdfpagelabels=true,
-colorlinks=true,linkcolor={black},citecolor={black},urlcolor={black},
-pdftitle={A Fast GPU Sonar Simulator for Automatic Target Recognition},
-pdfsubject={},
-pdfauthor={Jo Inge Buskenes},
-pdfkeywords={adaptive beamforming, beamforming, complexity, sonar, active}}%
-\ifCLASSINFOpdf
-   \usepackage[\MYhyperrefoptions,pdftex]{hyperref}
-\else
-   \usepackage[\MYhyperrefoptions,breaklinks=true,dvips]{hyperref}
-   \usepackage{breakurl}                        % Allows 'dvips' driver to break links
-\fi
+% \newcommand\MYhyperrefoptions{bookmarks=true,bookmarksnumbered=true,
+% pdfpagemode={UseOutlines},plainpages=false,pdfpagelabels=true,
+% colorlinks=true,linkcolor={black},citecolor={black},urlcolor={black},
+% pdftitle={A Fast GPU Sonar Simulator for Automatic Target Recognition},
+% pdfsubject={},
+% pdfauthor={Jo Inge Buskenes},
+% pdfkeywords={adaptive beamforming, beamforming, complexity, sonar, active}}%
+% \ifCLASSINFOpdf
+%    \usepackage[\MYhyperrefoptions,pdftex]{hyperref}
+% \else
+%    \usepackage[\MYhyperrefoptions,breaklinks=true,dvips]{hyperref}
+%    \usepackage{breakurl}                        % Allows 'dvips' driver to break links
+% \fi
 
@@
 
-\usepackage{float}
+% \usepackage{float}
 
@@
 
-\newenvironment{outline}
-{\begin{itemize}}
-{\end{itemize}}
+% \newenvironment{outline}
+% {\begin{itemize}}
+% {\end{itemize}}
+%
 
@@
 \begin{figure}[tbp]%
-\includegraphics[width=\linewidth]{gfx/calc_kaiser_3dB.eps}%
+\includegraphics[width=.9\linewidth]{gfx/calc_kaiser_3dB.eps}%
 \caption{Each Kaiser window is steered in the interval $\phi=[0, \phi_{\text{-3dB}}(\beta)]$.

### 2015-08-23 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:08:23	10:45:30	(Sun23.Aug)

@@
 
-\usepackage{stfloats}                          % Allows: \begin{figure*}[!b]
+% \usepackage{stfloats}                          % Allows: \begin{figure*}[!b]
                                                 % (double column figures on top/bottom)
@@
 
-\section{LCA}
-
-The LCA beamformer interates through a set of $P$ windows and $p$ that yields the least output power:
-%
-\begin{align}
-\underset{p}{\argmin}\, E\{|z[n]|^2\} = \underset{p}{\argmin}\, E\big\{|\w_{\beta,\phi}\H\x[n]|^2\big\}\\
-\qquad\text{subject to}\qquad \w\H\a = 1\label{eq_lca_constraint}
-\end{align}\label{lca_criterion}
-%
-We can estimate $E\big\{|\w_p\H\x[n]|^2\big\}$ as
-%
-\begin{align*}
-\sigma^2_{z_p}[n] = \frac{1}{N_k} \sumb{n'=n-K}{n+K} | z_p[n'] |^2
-\end{align*}
-%
-To add some extra flexibility, we can also estimate the variance by a weighted combination of local pixels
-%
-\begin{align*}
-\sigma^2_{z_p}[x,n] = \frac{1}{N_x N_k} \sumb{x'=x-X}{x+X} \sumb{n=n-K}{n+K} w_\text{prf}[x',n']\big| z_p[x',n'] \big|^2
-\end{align*}
-%
-where $w_\text{prf}[x',n']$ is a normalized 2 dimensional weight function.
-
-\begin{itemize}
-\item \emph{Window function}
-\item \emph{Parameter boundaries}
-\item \emph{Number of windows needed}
-\item \emph{Oversampling need}
-\item \emph{Image quality}
-\end{itemize}
-
-\subsection{Window function: Kaiser}\label{kaiser_windows}
-
-As will be demonstrated in upcoming sections, the LCA beamformer works very well with windows generated from the Kaiser-Bessel function. It is defined as:
-%
-\begin{align*}
-\f_\beta = \bmat{
-f_0(\beta) \\
-\vdots\\
-f_{M-1}(\beta)
-}
-\end{align*}
-%
-where
-%
-\begin{align}
-f_m(\beta) = \frac{I_0\left(\pi\beta\sqrt{1-\left(\frac{2m}{M-1}-1\right)^2}\right)}{I_0(\pi\beta)}
-\end{align}
-%
-and $I_0$ is the zeroth order modified Bessel function of the first kind:
-%
-\begin{align}
-I_0(x) = \sumb{a=0}{\infty} \left[ \frac{\left(\frac{x}{2}\right)^a}{a!} \right]^2.
-\end{align}
-%
-The Kaiser-Bessel window is near optimal in the sense of having its peak energy concentration around $\theta=0^\circ$, for a given time-bandwidth product specified with the parameter $\beta$:
-%
-\begin{align}
-\beta = \frac{TB}{2},
-\end{align}
-%
-where $T$ is the extent of the window in time and $B$ is its bandwidth.
-Adjusting $\beta$ changes the trade-off between mainlobe width and sidelobe level.
-At small values the windows approach the response of a rectangular window, and at large values ($>5$) the window converges to a Gaussian both in time and frequency.
-This class of windows is generally considered well suited for separating closely spaced sources with amplitudes of a high dynamic range~\cite{Harris1978}, they are easy to make, and they are optimal for any value of $\beta$.
-
 \begin{figure}[tbp]%
@@
 
-\subsection{Steering}
-
-To allow LCA some added flexibility we also create variations of each window that are steered a little to each side.
-%
-\begin{align}
-\w_{\beta,\phi} = \frac{\f_\beta\H\diag{(\a_\phi)}}{\f_\beta\H\a_\phi}\\
-\end{align}
-%
-where
-%
-\begin{align}
-\a_\phi = \bmat{
-1 \\
-e^{-j\frac{md}{\lambda}\sin(\phi)} \\
-\vdots\\
-e^{-j\frac{(M-1)d}{\lambda}\sin(\phi)}
-}.
-\end{align}
-%
-Here $d$ is the element spacing and $\lambda$ is the wavelength.
-The normalization factor $\f_\beta\H\a_\phi$ is the reciprocal of the window's coherent gain.
-It ensures unit gain in the direction of interest as required by (\ref{eq_lca_constraint}).
-Since the signal-to-noise ratio is constant for a specific window (same $\beta$-value), this normalization also proportionally increases the incoherent noise gain.
-
-\subsection{Steering bounds}
-
-There is an upper bound to the amount of steering we should apply to a window.
-Since each window's response is constrained to a unit gain in the look direction, each steered window's response is gained as shown in \Fig{windows_steering}.
-This increases the white noise gain and deteriorates the signal to noise ratio.
-To limit this we placing an upper bound to the steering, taking into account that a wide window can be allowed more steering than a narrow one.
-
-We call the upper bound for steering $\phi_{\text{-3dB}}(\beta)$, and define it as the steering angle needed for the steered response to have a -3dB crossing that is exactly half that of the unsteered window.
-We performed a numeric search for the steering angle $\phi$ that fulfilled this criterion for a wide range of $\beta$-values.
-A third degree polynomial fit to these points is:
-%
-\begin{align}
-\phi_\text{-3dB}(\beta) \approx -0.002894\beta^3 + 0.02451\beta^2 + -0.01308\beta + 0.2980
-\end{align}
-%
-Any steering will be specified relative to this upper bound.
-With each window steered this way we expect the resolution gain to remain predictable and independent of $\beta$, and we also effectively constrain the white noise gain of the beamformer.
-
-\subsection{Oversampling and steering}
-
-Adaptive beamformers process each pixel in a different way in an attempt to surpass the conventional limit of resolution and noise suppression.
-This non-linear processing of the image increases its bandwidth and introduces a need for oversampling.
-For MVDR a lateral oversampling factor of up to 10 times is often needed to ensure spatial shift-invariance~\cite{Asen2014}.
-
-Too much steering, beamformer becomes too agressive. Uncertainty in source amplitude grows.
-
-\begin{itemize}
-\item More steering -> more aggressive suppression of interference at the cost of boosted sidelobes / noise.
-\item More steering -> Larger variations in amplitude of sources
-\end{itemize}
-
-Idea to steer within a fraction of the rectangular window's mainlobe.
-
-\begin{align*}
-d = \frac{\pi\,\theta_\text{open}\,\lambda}{M\,d\,O_s}
-\end{align*}
-
-\begin{align}
-\w[\beta, \phi] = \text{kaiser}(\beta) = \bmat{
-1 \\
-e^{-j\frac{md}{\lambda}\sin(\phi)} \\
-\vdots\\
-e^{-j\frac{(M-1)d}{\lambda}\sin(\phi)}
-}
-\end{align}
-
-See section \ref{kaiser_windows}.
-
-% Narrow ones:
-% - Best coherent gain, lowest equivalent noise bandwidth, and most sensitive to steering meaning not needing to steer as much for some effect.
-% 
-% 
-% Beamforming:
-% - Maximize resolution
-% - Minimize noise/error, both coherent and incoherent
-% - Minimize signal cancellation
-% 
-% Gauss: Miminized time-bandwidth product
-%   - 
-%   
-% Rectangular
-%   - 
-% Dolph-Cheb: Minimize mainlobe width given sidelobe suppression
-% Kaiser: For a restricted energy and time duration, maximize energy in the band of frequencies W.
-% 
-% Adaptive windows perform best in detection of closely separated angles of significantly amplitudes.
-
-% Adaptive beamformers usually compute the window coefficients by estimating and inverting a spatial covariance matrix.
-There are two inherent problems to this procedure.
-First, for improved estimation accuracy the data is either averaged in space or time, and the significance of the covariance matrix can be reduced using a regularisation parameter~\cite{Carl Inge}.
-These are all attempts to constrain the beamformer to ensure the window coefficience are not over-adapted to the data.
-Second, inverting the covariance matrix has a computational complexity of O($M^3$)~\cite{Carl Inge}.
-For larger arrays the computational burden becomes significant, and the question arises whether this processing power can be better utilised.
-% 
-%
-
-% \section{Experimental Setup}
-% 
-% For the LCA to perform well it is important to let it have a diverse selection of windows to choose from.
-The question is how this selection should be With our ovarall goal being a reduction of computational complexity, it The choice of window function does not seem to matter.
-% 
-% 
-% 
-% Using Kaiser windows as their extremal values areLimit our study to Kaiser windows
-% 
-% 
-% 
-% This selection is based on observations of the solutions found by the MV beamformer.
-We obtained good results in our experiments by letting the LCA beamformer select among 5 Kaiser windows with uniformly distributed $\beta$'s in the range $[0.05, 0.5]$ and a rectangular window.
-Each window was steered in 5 different directions, uniformly distributed within 80\% of the mainlobe width of the rectangular window, which is the most narrow.
-This adds to a total of 30 windows.
-% 
-% To make the choice of window less susceptible to pixel value uncertainties, the LCA beamformer was set to apply the most frequently selected window in an 11 pixel range region to the center pixel in that region.
-In physical terms, this is a region of approximately 20\,cm.
-The MV beamformer estimated the covariance matrix as described in~\cite{syn07} by averaging over 16 subarrays and 11 range pixels, and applying 3\% diagonal loading.
-
-% Furthermore, the algorithm maps well to single-instruction-multiple-data (SIMD) hardware such as graphics processing units (GPUs).
-We show that even a modest amount of optimisation work has resulted in a factor 10 speed-up of this algorithm when implemented on a GPU as opposed to on a CPU.
+\setcounter{topnumber}{1}
+\setcounter{dbltopnumber}{1}
 
 \begin{figure*}[t]\centering%
-\includegraphics[width=\linewidth]{gfx/mvdr_selected_windows_holmengraa.eps}%
+\includegraphics[width=\linewidth]{gfx/mvdr_selected_windows_holmengraa.pdf}%
 \caption{Typical frequency responses for windows computed by a robust MVDR.\newline
@@
 \begin{figure*}[t]%\centering%
-\includegraphics[width=\textwidth]{gfx/oversampling_mosaic_bounds_lca_windows_beta.eps}%
+\includegraphics[width=\textwidth]{gfx/oversampling_mosaic_bounds_lca_windows_beta.pdf}%
 \caption{Window parameter boundaries. Oversampled image, 19x20 windows.}\label{oversampling_mosaic_bounds_beta}
@@
 \begin{figure*}[t]%\centering%
-\includegraphics[width=\textwidth]{gfx/oversampling_mosaic_bounds.eps}%
+\includegraphics[width=\textwidth]{gfx/oversampling_mosaic_bounds.pdf}%
 \caption{Window parameter boundaries.
@@
 \begin{figure*}[t]%\centering%
-\includegraphics[width=\textwidth]{gfx/oversampling_mosaic.eps}%
+\includegraphics[width=\textwidth]{gfx/oversampling_mosaic.pdf}%
 \caption{Window parameter boundaries. Oversampled image, 19x20 windows.}\label{oversampling_mosaic}
 \end{figure*}
+
+\section{LCA}
+
+The LCA beamformer interates through a set of $P$ windows and $p$ that yields the least output power:
+%
+\begin{align}
+\underset{p}{\argmin}\, E\{|z[n]|^2\} = \underset{p}{\argmin}\, E\big\{|\w_{\beta,\phi}\H\x[n]|^2\big\}\\
+\qquad\text{subject to}\qquad \w\H\a = 1\label{eq_lca_constraint}
+\end{align}\label{lca_criterion}
+%
+We can estimate $E\big\{|\w_p\H\x[n]|^2\big\}$ as
+%
+\begin{align*}
+\sigma^2_{z_p}[n] = \frac{1}{N_k} \sumb{n'=n-K}{n+K} | z_p[n'] |^2
+\end{align*}
+%
+To add some extra flexibility, we can also estimate the variance by a weighted combination of local pixels
+%
+\begin{align*}
+\sigma^2_{z_p}[x,n] = \frac{1}{N_x N_k} \sumb{x'=x-X}{x+X} \sumb{n=n-K}{n+K} w_\text{prf}[x',n']\big| z_p[x',n'] \big|^2
+\end{align*}
+%
+where $w_\text{prf}[x',n']$ is a normalized 2 dimensional weight function.
+
+\begin{itemize}
+\item \emph{Window function}
+\item \emph{Parameter boundaries}
+\item \emph{Number of windows needed}
+\item \emph{Oversampling need}
+\item \emph{Image quality}
+\end{itemize}
+
+\subsection{Window function: Kaiser}\label{kaiser_windows}
+
+As will be demonstrated in upcoming sections, the LCA beamformer works very well with windows generated from the Kaiser-Bessel function. It is defined as:
+%
+\begin{align*}
+\f_\beta = \bmat{
+f_0(\beta) \\
+\vdots\\
+f_{M-1}(\beta)
+}
+\end{align*}
+%
+where
+%
+\begin{align}
+f_m(\beta) = \frac{I_0\left(\pi\beta\sqrt{1-\left(\frac{2m}{M-1}-1\right)^2}\right)}{I_0(\pi\beta)}
+\end{align}
+%
+and $I_0$ is the zeroth order modified Bessel function of the first kind:
+%
+\begin{align}
+I_0(x) = \sumb{a=0}{\infty} \left[ \frac{\left(\frac{x}{2}\right)^a}{a!} \right]^2.
+\end{align}
+%
+The Kaiser-Bessel window is near optimal in the sense of having its peak energy concentration around $\theta=0^\circ$, for a given time-bandwidth product specified with the parameter $\beta$:
+%
+\begin{align}
+\beta = \frac{TB}{2},
+\end{align}
+%
+where $T$ is the extent of the window in time and $B$ is its bandwidth.
+Adjusting $\beta$ changes the trade-off between mainlobe width and sidelobe level.
+At small values the windows approach the response of a rectangular window, and at large values ($>5$) the window converges to a Gaussian both in time and frequency.
+This class of windows is generally considered well suited for separating closely spaced sources with amplitudes of a high dynamic range~\cite{Harris1978}, they are easy to make, and they are optimal for any value of $\beta$.
+
+\subsection{Steering}
+
+To allow LCA some added flexibility we also create variations of each window that are steered a little to each side.
+%
+\begin{align}
+\w_{\beta,\phi} = \frac{\f_\beta\H\diag{(\a_\phi)}}{\f_\beta\H\a_\phi}\\
+\end{align}
+%
+where
+%
+\begin{align}
+\a_\phi = \bmat{
+1 \\
+e^{-j\frac{md}{\lambda}\sin(\phi)} \\
+\vdots\\
+e^{-j\frac{(M-1)d}{\lambda}\sin(\phi)}
+}.
+\end{align}
+%
+Here $d$ is the element spacing and $\lambda$ is the wavelength.
+The normalization factor $\f_\beta\H\a_\phi$ is the reciprocal of the window's coherent gain.
+It ensures unit gain in the direction of interest as required by (\ref{eq_lca_constraint}).
+Since the signal-to-noise ratio is constant for a specific window (same $\beta$-value), this normalization also proportionally increases the incoherent noise gain.
+
+\subsection{Steering bounds}
+
+There is an upper bound to the amount of steering we should apply to a window.
+Since each window's response is constrained to a unit gain in the look direction, each steered window's response is gained as shown in \Fig{windows_steering}.
+This increases the white noise gain and deteriorates the signal to noise ratio.
+To limit this we placing an upper bound to the steering, taking into account that a wide window can be allowed more steering than a narrow one.
+
+We call the upper bound for steering $\phi_{\text{-3dB}}(\beta)$, and define it as the steering angle needed for the steered response to have a -3dB crossing that is exactly half that of the unsteered window.
+We performed a numeric search for the steering angle $\phi$ that fulfilled this criterion for a wide range of $\beta$-values.
+A third degree polynomial fit to these points is:
+%
+\begin{align}
+\phi_\text{-3dB}(\beta) \approx -0.002894\beta^3 + 0.02451\beta^2 + -0.01308\beta + 0.2980
+\end{align}
+%
+Any steering will be specified relative to this upper bound.
+With each window steered this way we expect the resolution gain to remain predictable and independent of $\beta$, and we also effectively constrain the white noise gain of the beamformer.
+
+\subsection{Oversampling and steering}
+
+Adaptive beamformers process each pixel in a different way in an attempt to surpass the conventional limit of resolution and noise suppression.
+This non-linear processing of the image increases its bandwidth and introduces a need for oversampling.
+For MVDR a lateral oversampling factor of up to 10 times is often needed to ensure spatial shift-invariance~\cite{Asen2014}.
+
+Too much steering, beamformer becomes too agressive. Uncertainty in source amplitude grows.
+
+\begin{itemize}
+\item More steering -> more aggressive suppression of interference at the cost of boosted sidelobes / noise.
+\item More steering -> Larger variations in amplitude of sources
+\end{itemize}
+
+Idea to steer within a fraction of the rectangular window's mainlobe.
+
+\begin{align*}
+d = \frac{\pi\,\theta_\text{open}\,\lambda}{M\,d\,O_s}
+\end{align*}
+
+\begin{align}
+\w[\beta, \phi] = \text{kaiser}(\beta) = \bmat{
+1 \\
+e^{-j\frac{md}{\lambda}\sin(\phi)} \\
+\vdots\\
+e^{-j\frac{(M-1)d}{\lambda}\sin(\phi)}
+}
+\end{align}
+
+See section \ref{kaiser_windows}.
+
+% Narrow ones:
+% - Best coherent gain, lowest equivalent noise bandwidth, and most sensitive to steering meaning not needing to steer as much for some effect.
+% 
+% 
+% Beamforming:
+% - Maximize resolution
+% - Minimize noise/error, both coherent and incoherent
+% - Minimize signal cancellation
+% 
+% Gauss: Miminized time-bandwidth product
+%   - 
+%   
+% Rectangular
+%   - 
+% Dolph-Cheb: Minimize mainlobe width given sidelobe suppression
+% Kaiser: For a restricted energy and time duration, maximize energy in the band of frequencies W.
+% 
+% Adaptive windows perform best in detection of closely separated angles of significantly amplitudes.
+
+% Adaptive beamformers usually compute the window coefficients by estimating and inverting a spatial covariance matrix.
+There are two inherent problems to this procedure.
+First, for improved estimation accuracy the data is either averaged in space or time, and the significance of the covariance matrix can be reduced using a regularisation parameter~\cite{Carl Inge}.
+These are all attempts to constrain the beamformer to ensure the window coefficience are not over-adapted to the data.
+Second, inverting the covariance matrix has a computational complexity of O($M^3$)~\cite{Carl Inge}.
+For larger arrays the computational burden becomes significant, and the question arises whether this processing power can be better utilised.
+% 
+%
+
+% \section{Experimental Setup}
+% 
+% For the LCA to perform well it is important to let it have a diverse selection of windows to choose from.
+The question is how this selection should be With our ovarall goal being a reduction of computational complexity, it The choice of window function does not seem to matter.
+% 
+% 
+% 
+% Using Kaiser windows as their extremal values areLimit our study to Kaiser windows
+% 
+% 
+% 
+% This selection is based on observations of the solutions found by the MV beamformer.
+We obtained good results in our experiments by letting the LCA beamformer select among 5 Kaiser windows with uniformly distributed $\beta$'s in the range $[0.05, 0.5]$ and a rectangular window.
+Each window was steered in 5 different directions, uniformly distributed within 80\% of the mainlobe width of the rectangular window, which is the most narrow.
+This adds to a total of 30 windows.
+% 
+% To make the choice of window less susceptible to pixel value uncertainties, the LCA beamformer was set to apply the most frequently selected window in an 11 pixel range region to the center pixel in that region.
+In physical terms, this is a region of approximately 20\,cm.
+The MV beamformer estimated the covariance matrix as described in~\cite{syn07} by averaging over 16 subarrays and 11 range pixels, and applying 3\% diagonal loading.
+
+% Furthermore, the algorithm maps well to single-instruction-multiple-data (SIMD) hardware such as graphics processing units (GPUs).
+We show that even a modest amount of optimisation work has resulted in a factor 10 speed-up of this algorithm when implemented on a GPU as opposed to on a CPU.
 

### 2015-08-23 LCAforActiveSonarImaging/lca.tex
Adding old files.

@@
 
-\subsection{Oversampling and steering}
-
-Adaptive beamformers process each pixel in a different way in an attempt to surpass the conventional limit of resolution and noise suppression.
-This non-linear processing of the image increases its bandwidth and introduces a need for oversampling.
-For MVDR a lateral oversampling factor of up to 10 times is often needed to ensure spatial shift-invariance~\cite{Asen2014}.
-
-Too much steering, beamformer becomes too agressive. Uncertainty in source amplitude grows.
-
 \begin{itemize}
@@
 
+\subsection{Oversampling}
+
+Adaptive beamformers process each pixel in a different way in an attempt to surpass the conventional limit of resolution and noise suppression.
+This non-linear processing of the image increases its bandwidth and introduces a need for oversampling.
+For MVDR a lateral oversampling factor of up to 10 times is often needed to ensure spatial shift-invariance in source amplitude~\cite{Asen2014}.
+
 Idea to steer within a fraction of the rectangular window's mainlobe.
 
-\begin{align*}
+\begin{align}
 d = \frac{\pi\,\theta_\text{open}\,\lambda}{M\,d\,O_s}
-\end{align*}
-
-\begin{align}
-\w[\beta, \phi] = \text{kaiser}(\beta) = \bmat{
-1 \\
-e^{-j\frac{md}{\lambda}\sin(\phi)} \\
-\vdots\\
-e^{-j\frac{(M-1)d}{\lambda}\sin(\phi)}
-}
-\end{align}
-
-See section \ref{kaiser_windows}.
+\end{align}
+
+% See section \ref{kaiser_windows}.
 

### 2015-08-27 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:08:27	16:38:57	(Thu27.Aug)

@@
 
-As will be demonstrated in upcoming sections, the LCA beamformer works very well with windows generated from the Kaiser-Bessel function. It is defined as:
+As will be demonstrated in upcoming sections, the LCA beamformer works very well with windows generated from the Kaiser-Bessel function. Let it be expressed on vector form as:
 %
@@
 Adjusting $\beta$ changes the trade-off between mainlobe width and sidelobe level.
-At small values the windows approach the response of a rectangular window, and at large values ($>5$) the window converges to a Gaussian both in time and frequency.
+When $\beta=0$ the window becomes a rectangular, while at large values ($\beta>5$) the window converges to a Gaussian both in time and frequency.
 This class of windows is generally considered well suited for separating closely spaced sources with amplitudes of a high dynamic range~\cite{Harris1978}, they are easy to make, and they are optimal for any value of $\beta$.
@@
 %
-Here $d$ is the element spacing and $\lambda$ is the wavelength.
+Here $d$ is the element spacing, $\lambda$ is the wavelength and $\phi$ is the steering amount.
 The normalization factor $\f_\beta\H\a_\phi$ is the reciprocal of the window's coherent gain.
@@
 Since each window's response is constrained to a unit gain in the look direction, each steered window's response is gained as shown in \Fig{windows_steering}.
-This increases the white noise gain and deteriorates the signal to noise ratio.
+This increases the white noise gain and deteriorates the signal-to-noise ratio.
 To limit this we placing an upper bound to the steering, taking into account that a wide window can be allowed more steering than a narrow one.
@@
 We call the upper bound for steering $\phi_{\text{-3dB}}(\beta)$, and define it as the steering angle needed for the steered response to have a -3dB crossing that is exactly half that of the unsteered window.
-We performed a numeric search for the steering angle $\phi$ that fulfilled this criterion for a wide range of $\beta$-values.
-A third degree polynomial fit to these points is:
+Using a numerical search for the steering angle $\phi$ that fulfilled this criterion for a wide range of $\beta$-values, we obtained a third degree polynomial fit to these points:
 %

### 2015-08-28 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:08:28	22:01:27	(Fri28.Aug)

@@
 
+\newcommand\multimedia[1]{\textbf{{\color{red}[#1]}}}
+
 \begin{document}
@@
 \includegraphics[width=\textwidth]{gfx/oversampling_mosaic.pdf}%
-\caption{Window parameter boundaries. Oversampled image, 19x20 windows.}\label{oversampling_mosaic}
+\caption{LCA images using window sets with varying parameter boundaries. Oversampled image, 19x20 windows.}\label{oversampling_mosaic}
 \end{figure*}
@@
 In shadow and speckle the windows are fully symmetric, while in highlight the windows are approximately symmetric within the opening angle of the sonar.
+
+Media file \multimedia{Media Movie 1} shows how \Fig{mvdr_selected_windows} changes as a function of the MVDR subarray length $L$.
+It demonstrates that at $L$=1 the window responses are rectangular in all areas in the image.
+At $L$=2 we can observe that pixels in highlight are processed with windows with slightly varying amplitude responses, but with zero locations on similar locations and highlight regions being
 

### 2015-08-28 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:08:28	22:42:14	(Fri28.Aug)

@@
 % colorlinks=true,linkcolor={black},citecolor={black},urlcolor={black},
-% pdftitle={A Fast GPU Sonar Simulator for Automatic Target Recognition},
+% pdftitle={Optimal Window Design for the Low Complexity Adaptive Beamformer in Active Sonar Imaging},
 % pdfsubject={},
@@
 % pdfkeywords={adaptive beamforming, beamforming, complexity, sonar, active}}%
-% \ifCLASSINFOpdf
+\ifCLASSINFOpdf
 %    \usepackage[\MYhyperrefoptions,pdftex]{hyperref}
-% \else
-%    \usepackage[\MYhyperrefoptions,breaklinks=true,dvips]{hyperref}
-%    \usepackage{breakurl}                        % Allows 'dvips' driver to break links
-% \fi
+   \usepackage[pdftex,
+    final,                %Keep hyperlink stuff even when in draft mode
+    breaklinks=true,      %Break links when necessary
+    linktocpage=true,     %Enable link to page?
+    linkcolor=black,  %Colour of links to labels within document
+    citecolor=black,      %Colour of links to the biliography
+    filecolor=red,        %Colour of links to local files
+    pagecolor=black,        %Colour of links to other pages withing document
+    urlcolor=Blue,        %Colour of the links to external URLs
+    colorlinks=true,      %??
+    plainpages=false,     %Store roman/arabic numbering differently to avoid
+    bookmarksnumbered ]{hyperref}
+\else
+   \usepackage[\MYhyperrefoptions,breaklinks=true,dvips]{hyperref}
+   \usepackage{breakurl}                        % Allows 'dvips' driver to break links
+\fi
 
@@
 
-\newcommand\multimedia[1]{\textbf{{\color{red}[#1]}}}
+% \newcommand\multimedia[1]{\textbf{{\color{red}[#1]}}}
+\newcommand\multimedia[2]{\href{#1}{#2}}
+
+\newcommand\mediaPath{gfx/media}
+\newcommand\mediaI{\multimedia{\mediaPath/media1.mp4}{Media Movie 1}}
+\newcommand\mediaII{\multimedia{\mediaPath/media2.mp4}{Media Movie 2}}
 
@@
 
-Adaptive beamformers are generally not so simple.
-They allow the weights/delays to change to adjust the array response to better fit the incoming wavefield.
-Of these methods the minimum variance distortionless response (MVDR) beamformer is perhaps the best studied and well known.
+Adaptive beamformers allow the weights and delays to change to adjust the array response to better fit the incoming wavefield.
+One of the most extensively studied methods here is the minimum variance distortionless response (MVDR) beamformer.
 
@@
 
-Media file \multimedia{Media Movie 1} shows how \Fig{mvdr_selected_windows} changes as a function of the MVDR subarray length $L$.
+\mediaI{} shows how \Fig{mvdr_selected_windows} changes as a function of the MVDR subarray length $L$.
 It demonstrates that at $L$=1 the window responses are rectangular in all areas in the image.

### 2015-08-29 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:08:29	13:22:53	(Sat29.Aug)

@@
 
-MVDR seeks to minimize the power of the beamformer, under the constraint of a unity gain in the desired direction $\a$~\cite{Capon1969}:
+MVDR seeks to minimize the power of the beamformer, under the constraint of a unity gain in the desired direction $\phi$~\cite{Capon1969}:
 %
@@
 \underset{\w[n]}{\argmin}\, E\{|z[n]|^2\} &= \underset{\w[n]}{\argmin}\, \w[n]\R[n]\w\H[n]\\
-\text{subject to } \w[n]\a_\phi &= 1
-\end{align}
-%
-where $\a_\phi$ is a steering vector and $\R[n] = E\{\x\x\H\}$ is the spatial covariance matrix. This is a convex optimization problem with the solution:
+\text{subject to } \w[n]\a_\phi &= 1.
+\end{align}
+%
+Here $\a_\phi$ is a steering vector, $E\{\cdot\}$ is the expectation operator, and $\R[n] = E\{\x\x\H\}$ is the spatial covariance matrix. This is a convex optimization problem with the solution:
 %
 \begin{gather}
-\vec w[n] = \frac{\Ri[n]\a}{\a\T\Ri[n]\a},\label{weights}
+\vec w[n] = \frac{\Ri[n]\a_\phi}{\a_\phi\T\Ri[n]\a_\phi},\label{weights}
 \end{gather}
@@
 
-The LCA beamformer interates through a set of $P$ windows and $p$ that yields the least output power:
-%
-\begin{align}
-\underset{p}{\argmin}\, E\{|z[n]|^2\} = \underset{p}{\argmin}\, E\big\{|\w_{\beta,\phi}\H\x[n]|^2\big\}\\
-\qquad\text{subject to}\qquad \w\H\a = 1\label{eq_lca_constraint}
+A much less complex alternative to MVDR is the low complexity adaptive (LCA) beamformer.
+It iterates through a set of $P$ windows and selects the window $p$ that best fulfills the minimum variance criteron:
+%
+\begin{align}
+\underset{p}{\argmin}\, E\{|z[n]|^2\} = \underset{p}{\argmin}\, E\big\{|\w_{p}\H\x[n]|^2\big\}\\
+\qquad\text{subject to}\qquad \w\H\a_\phi = 1\label{eq_lca_constraint}
 \end{align}\label{lca_criterion}
@@
 %
-To add some extra flexibility, we can also estimate the variance by a weighted combination of local pixels
+which is the sample covariance averaged over $N_k$=$2*K+1$ temporal samples. To add some extra flexibility, we can also estimate the variance by a weighted combination of local pixels
 %

### 2015-08-29 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:08:29	17:44:21	(Sat29.Aug)

@@
 \newcommand\Fig[1]{Fig.~\ref{#1}}
+\newcommand\eq[1]{(\ref{#1})}
 
@@
 \begin{align}
-\underset{\w[n]}{\argmin}\, E\{|z[n]|^2\} &= \underset{\w[n]}{\argmin}\, \w[n]\R[n]\w\H[n]\\
-\text{subject to } \w[n]\a_\phi &= 1.
+\underset{\w[n]}{\argmin}\, E\{|z[n]|^2\} &= \underset{\w[n]}{\argmin}\, \w[n]\R[n]\w\H[n]\nn
+\text{subject to } \w[n]\a_\phi &= 1.\label{eq_mvdr_definition}
 \end{align}
@@
 \begin{align}
-\underset{p}{\argmin}\, E\{|z[n]|^2\} = \underset{p}{\argmin}\, E\big\{|\w_{p}\H\x[n]|^2\big\}\\
-\qquad\text{subject to}\qquad \w\H\a_\phi = 1\label{eq_lca_constraint}
+\underset{p}{\argmin}\, E\{|z[n]|^2\} = \underset{p}{\argmin}\, E\big\{|\w_{p}\H\x[n]|^2\big\}\nn
+\qquad\text{subject to}\qquad \w\H\a_\phi = 1.\label{eq_lca_definition}
 \end{align}\label{lca_criterion}
 %
+Note how closely LCA is related to the MVDR definition in \eq{eq_mvdr_definition}.
+The optimization criterion and constraint is the same, but LCA has a finite and discrete solutionspace for the weights.
+As will be demonstrated in section \ref{lca_window_function}, this can be done because a robustified MVDR implementation seems to be constrained to window functions similar to the conventional ones.
+
 We can estimate $E\big\{|\w_p\H\x[n]|^2\big\}$ as
@@
 %
-which is the sample covariance averaged over $N_k$=$2*K+1$ temporal samples. To add some extra flexibility, we can also estimate the variance by a weighted combination of local pixels
+which is the sample covariance averaged over $N_k=2K+1$ temporal samples. To add some extra flexibility, we can also estimate the variance by a weighted combination of local pixels
 %
@@
 
-\subsection{Window function}
+\subsection{Window function}\label{lca_window_function}
 

### 2015-08-29 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:08:29	19:21:57	(Sat29.Aug)

@@
 The optimization criterion and constraint is the same, but LCA has a finite and discrete solutionspace for the weights.
-As will be demonstrated in section \ref{lca_window_function}, this can be done because a robustified MVDR implementation seems to be constrained to window functions similar to the conventional ones.
+As will be demonstrated in section \ref{lca_window_function}, LCA will perform similarly because a robustified MVDR implementation seems to be constrained to window functions similar to the ones we let LCA choose from.
 

### 2015-08-29 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:08:29	19:52:25	(Sat29.Aug)

--- 2015-08-29 LCAforActiveSonarImaging/lca.tex
@@
 The problem lies in estimating and inverting this spatial covariance matrix.
-Spatial averaging is needed to avoid signal cancellation~\cite{Kailath1985}, temporal averaging is needed to obtain true speckle statistics~\cite{Synnevag2009a}, and diagonal loading is needed to improve robustness to parameter errors~\cite{Cox1987,Maksym1979}.
+We have described the exact steps in~\cite{Buskenes2014}.
+To avoid signal cancellation we apply spatial averaging by computing a mean covariance matrix from a set of subarrays with length $L$~\cite{Kailath1985}.
+For true speckle statistics we perform temporal averaging over $N_k = 2K+1$ temporal samples~\cite{Synnevag2009a}.
+Finally, to improve robustness to parameter errors we apply $d$ percent diagonal loading~\cite{Cox1987,Maksym1979}.
 The steps are also needed to ensure that the covariance matrix is numerically well conditioned and hence invertible.
@@
 The optimization criterion and constraint is the same, but LCA has a finite and discrete solutionspace for the weights.
-As will be demonstrated in section \ref{lca_window_function}, this can be done because a robustified MVDR implementation seems to be constrained to window functions similar to the conventional ones.
-
-We can estimate $E\big\{|\w_p\H\x[n]|^2\big\}$ as
+As will be demonstrated in section \ref{lca_window_function}, LCA will perform similarly to MVDR because a robustified MVDR implementation seems to be constrained to window functions similar to the ones we let LCA choose from.
+
+In practice, we estimate the beamformer output power as
 %
 \begin{align*}
-\sigma^2_{z_p}[n] = \frac{1}{N_k} \sumb{n'=n-K}{n+K} | z_p[n'] |^2
+E\big\{|z[n]|^2\big\} = \frac{1}{N_k} \sumb{n'=n-K}{n+K} | z_p[n'] |^2
 \end{align*}
 %
-which is the sample covariance averaged over $N_k=2K+1$ temporal samples. To add some extra flexibility, we can also estimate the variance by a weighted combination of local pixels
+which computes the sample power average over $N_k=2K+1$ temporal samples.
+This is the same as we did for MVDR.
+Our signal is only a few samples long after matched filtering so we have set $K=1$ for both MVDR and LCA.
+
+If extra flexibility is desired, we can also estimate the variance by a weighted combination of local pixels
 %
@@
 %
-where $w_\text{prf}[x',n']$ is a normalized 2 dimensional weight function.
+where $N_x = 2X+1$ is the number of azimuth lines to average over, and $w_\text{prf}[x',n']$ is a normalized 2 dimensional weight function.
 
@@
 
-As will be demonstrated in upcoming sections, the LCA beamformer works very well with windows generated from the Kaiser-Bessel function. Let it be expressed on vector form as:
+As will be demonstrated in upcoming sections, the LCA beamformer works very well with windows generated from the Kaiser-Bessel function. We will express it in vector form as:
 %
@@
 The normalization factor $\f_\beta\H\a_\phi$ is the reciprocal of the window's coherent gain.
-It ensures unit gain in the direction of interest as required by (\ref{eq_lca_constraint}).
+It ensures unit gain in the direction of interest as required by (\ref{eq_lca_definition}).
 Since the signal-to-noise ratio is constant for a specific window (same $\beta$-value), this normalization also proportionally increases the incoherent noise gain.
@@
 There is an upper bound to the amount of steering we should apply to a window.
-Since each window's response is constrained to a unit gain in the look direction, each steered window's response is gained as shown in \Fig{windows_steering}.
-This increases the white noise gain and deteriorates the signal-to-noise ratio.
+Since each window's response is constrained to a unit gain in the look direction, the steered window will have increased white noise gain and deteriorated signal-to-noise ratio as shown in \Fig{windows_steering}.
 To limit this we placing an upper bound to the steering, taking into account that a wide window can be allowed more steering than a narrow one.

### 2015-08-29 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:08:29	21:26:08	(Sat29.Aug)

@@
 \includegraphics[width=\textwidth]{gfx/oversampling_mosaic_bounds_lca_windows_beta.pdf}%
-\caption{Window parameter boundaries. Oversampled image, 19x20 windows.}\label{oversampling_mosaic_bounds_beta}
+\caption{$\beta$-values of the windows LCA selects for each image pixel.
+The underlying image is the same cross as shown in \Fig{oversampling_mosaic_bounds}.
+Oversampled image, 19x20 windows.}\label{oversampling_mosaic_bounds_beta}
 \end{figure*}
@@
 \begin{align}
-\underset{p}{\argmin}\, E\{|z[n]|^2\} = \underset{p}{\argmin}\, E\big\{|\w_{p}\H\x[n]|^2\big\}\nn
+\underset{p}{\argmin}\, E\{|z_p[n]|^2\} = \underset{p}{\argmin}\, E\big\{|\w_{p}\H\x[n]|^2\big\}\nn
 \qquad\text{subject to}\qquad \w\H\a_\phi = 1.\label{eq_lca_definition}
@@
 \begin{align*}
-E\big\{|z[n]|^2\big\} = \frac{1}{N_k} \sumb{n'=n-K}{n+K} | z_p[n'] |^2
+E\big\{|z_p[n]|^2\big\} = \frac{1}{N_k} \sumb{n'=n-K}{n+K} | z_p[n'] |^2
 \end{align*}
@@
 %
-\begin{align*}
-\sigma^2_{z_p}[x,n] = \frac{1}{N_x N_k} \sumb{x'=x-X}{x+X} \sumb{n=n-K}{n+K} w_\text{prf}[x',n']\big| z_p[x',n'] \big|^2
-\end{align*}
+\begin{align}
+E\big\{|z_p[x,n]|^2\big\} = \frac{1}{N_x N_k} \sumb{x'=x-X}{x+X} \sumb{n=n-K}{n+K} w_\text{prf}[x',n']\big| z_p[x',n'] \big|^2
+\end{align}
 %
@@
 Since each window's response is constrained to a unit gain in the look direction, the steered window will have increased white noise gain and deteriorated signal-to-noise ratio as shown in \Fig{windows_steering}.
-To limit this we placing an upper bound to the steering, taking into account that a wide window can be allowed more steering than a narrow one.
+To limit this we placing an upper bound to the steering, realizing that  a wide window will be allowed more steering than a narrow one.
 
@@
 %
-Any steering will be specified relative to this upper bound.
+Henceforth any steering will be specified relative to this upper bound.
 With each window steered this way we expect the resolution gain to remain predictable and independent of $\beta$, and we also effectively constrain the white noise gain of the beamformer.
@@
 This non-linear processing of the image increases its bandwidth and introduces a need for oversampling.
-For MVDR a lateral oversampling factor of up to 10 times is often needed to ensure spatial shift-invariance in source amplitude~\cite{Asen2014}.
+For MVDR often a factor 10 lateral oversampling is often needed to ensure spatial shift-invariance in source amplitude~\cite{Asen2014}.
+To ensure that no
 

### 2015-08-29 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:08:29	22:12:39	(Sat29.Aug)

@@
 
-\begin{figure*}[t]%\centering%
-\includegraphics[width=\textwidth]{gfx/oversampling_mosaic_bounds_lca_windows_beta.pdf}%
-\caption{$\beta$-values of the windows LCA selects for each image pixel.
-The underlying image is the same cross as shown in \Fig{oversampling_mosaic_bounds}.
-Oversampled image, 19x20 windows.}\label{oversampling_mosaic_bounds_beta}
-\end{figure*}
+\setcounter{dbltopnumber}{2}
 
@@
 \end{figure*}
+
+\begin{figure*}[t]%\centering%
+\includegraphics[width=\textwidth]{gfx/oversampling_mosaic_bounds_lca_windows_beta.pdf}%
+\caption{$\beta$-values of the windows LCA selects for each image pixel.
+The underlying image is the one shown in \Fig{oversampling_mosaic_bounds}, and the location of the anchor and cut line is overlaid.
+}\label{oversampling_mosaic_bounds_beta}
+\end{figure*}
+
+% \setcounter{dbltopnumber}{1}
 
@@
 For MVDR often a factor 10 lateral oversampling is often needed to ensure spatial shift-invariance in source amplitude~\cite{Asen2014}.
-To ensure that no
+To compare apples to apples, we will use the same factor for both LCA and MVDR.
+
+Need $M$ in the image sector. Why???
+
+- Abs -> adds a factor 2
+- log -> infinity?
+- near field -> range and azimuth is coupled
 
@@
 \item \emph{Window function}.
-Frequency responses of typical windows chosen by MVDR in shadow, speckle and highlight regions in an image is shown in \Fig{mvdr_selected_windows}.
+\Fig{mvdr_selected_windows} show frequency responses of typical windows chosen by MVDR in shadow, speckle and highlight regions in an image.
 The responses tend to be symmetric, minimally steered and have slightly varying mainlobe width and sidelobe suppression.

### 2015-08-29 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:08:29	22:46:48	(Sat29.Aug)

@@
 \begin{itemize}
-\item \emph{Window function}.
+\item \emph{Window type}.
 \Fig{mvdr_selected_windows} show frequency responses of typical windows chosen by MVDR in shadow, speckle and highlight regions in an image.
@@
 We use Kaiser windows to mimic these responses.
-\item \emph{Window parameters}.
+\item \emph{Window parameter boundaries}.
 The Kaiser window parameter $\beta$ adjusts the compromise between resolution and sidelobe suppression.
@@
 $\beta=0$ ) and a gaussian curve (
+\item \emph{Window database size}.
+\item \emph{Performance}.
+\item \emph{Oversampling need}.
 \end{itemize}
@@
 Assuming that a robust MVDR computes the optimal windows in the minimum variance sense, it makes sense to study them.
-For this we selected a scene of the oil tanker wreck Holmengraa, which lies at a slanted seabed at nearly 100m\;depth.
+For this we selected a scene of the oil tanker wreck Holmengraa, which lies at a slanted seabed at nearly 100\;m depth.
 A sectorscan image of a single ping of data collected by the HISAS1030 syntetic aperture sonar is shown in \Fig{mvdr_selected_windows}, along with the window responses MVDR typically chooses in shadow, speckle and highlight regions in this image.
@@
 In shadow and speckle the windows are fully symmetric, while in highlight the windows are approximately symmetric within the opening angle of the sonar.
+
+Kaiser
+Unlike the method described by Synne\aa{}g~\cite{Synnevag2008} we do not let LCA choose from  inverted Kaiser windows.
+These have a spatial response with mainlobe width narrower than that of a rectangular window, but at the expense of drastically increased white noise gain.
+We have found that these windows hardly ever get used on experimental data, in particular steered versions which ever worsened white noise gain.
+We can infer the same from \Fig{mvdr_selected_windows} by noting that the maximum sidelobe level rarely exceeds the rectangular window level of -13\;dB.
 

### 2015-08-30 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:08:30	12:26:53	(Sun30.Aug)

@@
 
-% \ifCLASSOPTIONcompsoc
-%    \usepackage[caption=false,font=normalsize,labelfont=sf,textfont=sf]{subfig}
-% \else
-%    \usepackage[caption=false,font=footnotesize]{subfig}
-% \fi
+\ifCLASSOPTIONcompsoc
+   \usepackage[caption=false,font=normalsize,labelfont=sf,textfont=sf]{subfig}
+\else
+   \usepackage[caption=false,font=footnotesize]{subfig}
+\fi
 
@@
 \includegraphics[width=\linewidth]{gfx/mvdr_selected_windows_holmengraa.pdf}%
-\caption{Typical frequency responses for windows computed by a robust MVDR.\newline
+\caption{\emph{Determining LCA window type}. Typical frequency responses for windows computed by a robust MVDR.\newline
+\mbox{}\hfill\parbox[t]{.95\linewidth}{\emph{Goal: Determining LCA window type.}}
+\newline
 \emph{Left:}\hfill
 \parbox[t]{.95\linewidth}{MVDR sectorscan image of the oiltanker Holmengraa, using data collected from the 32 element HISAS1030 sonar attached to the HUGIN AUV.
-MVDR was run with $K=1$ temporal oversampling, $L=16$ subarray size, $d=1\%$ diagonal loading and 8 times lateral oversampling.}\newline
+MVDR was run with $K=1$ temporal oversampling, $L=16$ subarray size, $d=1\%$ diagonal loading and 8 times lateral oversampling.}\protect\\\hspace{\textwidth}
 \emph{Right:}\hfill
-\parbox[t]{.95\linewidth}{Window responses computed from 40x40 pixel groups in the shadow, highlight and speckle region of the Holmengraa image.
+\parbox[t]{.95\linewidth}{MVDR window responses computed from 40x40 pixel groups in the shadow, highlight and speckle region of the Holmengraa image.
 The responses are overlayed each other.
@@
 where $N_x = 2X+1$ is the number of azimuth lines to average over, and $w_\text{prf}[x',n']$ is a normalized 2 dimensional weight function.
+\todo{Might want to be careful with this when we steer the windows, or drop it altogether}
 
@@
 Kaiser
+
 Unlike the method described by Synne\aa{}g~\cite{Synnevag2008} we do not let LCA choose from  inverted Kaiser windows.

### 2015-08-30 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:08:30	13:37:45	(Sun30.Aug)

@@
 
-% \usepackage{stfloats}                          % Allows: \begin{figure*}[!b]
+\usepackage{stfloats}                          % Allows: \begin{figure*}[!b]
                                                 % (double column figures on top/bottom)
@@
 \caption{\emph{Determining LCA window type}. Typical frequency responses for windows computed by a robust MVDR.\newline
-\mbox{}\hfill\parbox[t]{.95\linewidth}{\emph{Goal: Determining LCA window type.}}
-\newline
 \emph{Left:}\hfill
@@
 The responses are overlayed each other.
-The dashed red lines at -13dB and -43dB marks the peak sidelobe levels of a rectangular and Hamming window, respectively.
+The dashed red lines at -13dB and -43dB marks the peak sidelobe levels of an unsteered rectangular and Hamming window, respectively.
 Note how the responses are more or less symmetric, with very little steering in the shadow, moderate steering in speckle and steering within roughly 3dB in highlight.
@@
 \begin{align}
-E\big\{|z_p[x,n]|^2\big\} = \frac{1}{N_x N_k} \sumb{x'=x-X}{x+X} \sumb{n=n-K}{n+K} w_\text{prf}[x',n']\big| z_p[x',n'] \big|^2
-\end{align}
+E\big\{|z_p[x,n]|^2\big\} = \frac{1}{N_x N_k} \sumb{\nmat{\scriptstyle x'=\\[-2mm]\scriptstyle x-X}}{x+X} \sumb{\nmat{\scriptstyle n'=\\[-2mm]\scriptstyle n-K}}{n+K} w_\text{prf}[x',n']\big| z_p[x',n'] \big|^2\end{align}
 %
@@
 Adjusting $\beta$ changes the trade-off between mainlobe width and sidelobe level.
-When $\beta=0$ the window becomes a rectangular, while at large values ($\beta>5$) the window converges to a Gaussian both in time and frequency.
+When $\beta=0$ the window becomes rectangular, while at large values ($\beta>5$) the window converges to a Gaussian both in time and frequency.
 This class of windows is generally considered well suited for separating closely spaced sources with amplitudes of a high dynamic range~\cite{Harris1978}, they are easy to make, and they are optimal for any value of $\beta$.
@@
 %
-where
+where we define the steering vector $\a_\phi$ as:
 %

### 2015-08-30 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:08:30	14:11:20	(Sun30.Aug)
la2pdf commit - 2015:08:30	14:41:21	(Sun30.Aug)

@@
          {\vspace*{-10pt}\color{Red}\fbox{\bf\thetodoidx{}}}\\%
-         \fcolorbox{red}{todobackground}{\parbox{\marginparwidthsmall}{\scriptsize #1}}}}
+         \fcolorbox{red}{todobackground}{\parbox{\marginparwidthsmall}{\raggedright\scriptsize #1}}}}
 
@@
 
+\newcommand\minus{\scalebox{0.75}[1.0]{$-$}}
+
 % \usepackage{tikz}
@@
 \begin{align}
-\underset{\w[n]}{\argmin}\, E\{|z[n]|^2\} &= \underset{\w[n]}{\argmin}\, \w[n]\R[n]\w\H[n]\nn
+\underset{\w[n]}{\argmin}\, E\{\big|z[n]\big|^2\} &= \underset{\w[n]}{\argmin}\, \w[n]\R[n]\w\H[n]\nn
 \text{subject to } \w[n]\a_\phi &= 1.\label{eq_mvdr_definition}
@@
 \includegraphics[width=\linewidth]{gfx/mvdr_selected_windows_holmengraa.pdf}%
-\caption{\emph{Determining LCA window type}. Typical frequency responses for windows computed by a robust MVDR.\newline
+\caption{\emph{Determining LCA window type}: Typical frequency responses for windows computed by a robust MVDR.\newline
 \emph{Left:}\hfill
@@
 \includegraphics[width=\textwidth]{gfx/oversampling_mosaic_bounds.pdf}%
-\caption{Window parameter boundaries.
-LCA images using an oversampled image and a relatively large window database with $N_\beta=20$ window types, each steered $N_\phi=19$.
+\caption{\emph{Determining Kaiser parameter boundaries}: LCA images created with a large and equal size window database, but where the upper bounds for $\beta$ and $\phi$ is varied.
+The upper left image is equal to DAS with rectangular window.
+While \Fig{mvdr_selected_windows} suggests that an upper bound
+\newline
+\emph{Upper bound $\beta$:}\hfill
+\parbox[t]{.9\linewidth}{.}\newline
+\emph{Right:}\hfill
+\parbox[t]{.95\linewidth}{.}
 }\label{oversampling_mosaic_bounds}
@@
 \begin{align}
-\underset{p}{\argmin}\, E\{|z_p[n]|^2\} = \underset{p}{\argmin}\, E\big\{|\w_{p}\H\x[n]|^2\big\}\nn
+\underset{p}{\argmin}\, E\{\big|z_p[n]\big|^2\} = \underset{p}{\argmin}\, E\big\{\big|\w_{p}\H\x[n]\big|^2\big\}\nn
 \qquad\text{subject to}\qquad \w\H\a_\phi = 1.\label{eq_lca_definition}
@@
 \begin{align*}
-E\big\{|z_p[n]|^2\big\} = \frac{1}{N_k} \sumb{n'=n-K}{n+K} | z_p[n'] |^2
+E\big\{\big|z[n]\big|^2\big\} = \frac{1}{N_k} \sumb{n'=n-K}{n+K} \big| z[n'] \big|^2
 \end{align*}
@@
 \begin{align}
-E\big\{|z_p[x,n]|^2\big\} = \frac{1}{N_x N_k} \sumb{\nmat{\scriptstyle x'=\\[-2mm]\scriptstyle x-X}}{x+X} \sumb{\nmat{\scriptstyle n'=\\[-2mm]\scriptstyle n-K}}{n+K} w_\text{prf}[x',n']\big| z_p[x',n'] \big|^2\end{align}
+E\big\{\big|z[x,n]\big|^2\big\} = \frac{1}{N_x N_k} \sumb{\nmat{\scriptstyle x'=\\[-2mm]\scriptstyle x-X}}{x+X} \sumb{\nmat{\scriptstyle n'=\\[-2mm]\scriptstyle n-K}}{n+K} w[x',n']\big| z[x',n'] \big|^2
+\end{align}
 %
@@
 \todo{Might want to be careful with this when we steer the windows, or drop it altogether}
-
-\begin{itemize}
-\item \emph{Window function}
-\item \emph{Parameter boundaries}
-\item \emph{Number of windows needed}
-\item \emph{Oversampling need}
-\item \emph{Image quality}
-\end{itemize}
+% 
+% \begin{itemize}
+% \item \emph{Window function}
+% \item \emph{Parameter boundaries}
+% \item \emph{Number of windows needed}
+% \item \emph{Oversampling need}
+% \item \emph{Image quality}
+% \end{itemize}
 
@@
 %
-where we define the steering vector $\a_\phi$ as:
+where diag($\a_\phi$) is a diagonal matrix constructed from the steering vector $\a_\phi$:
 %
@@
 \begin{align}
-\phi_\text{-3dB}(\beta) \approx -0.002894\beta^3 + 0.02451\beta^2 + -0.01308\beta + 0.2980
+\phi_\text{3dB}(\beta) \approx \minus{}0.002894\beta^3 + 0.02451\beta^2 - 0.01308\beta + 0.2980
 \end{align}
@@
 In shadow and speckle the windows are fully symmetric, while in highlight the windows are approximately symmetric within the opening angle of the sonar.
+The peak sidelobe levels are between that of the rectangular and Hamming window.
+
+This correspond to Kaiser windows with $\beta$-values of 0 and 2, respectively.
+\Fig{oversampling_mosaic_bounds_beta} illustrates which $\beta$-value LCA prefers for different regions in the cross image.
 
@@
 Typical values are in the range $\beta\in[0,10]$.
-
-$\phi$ controls the amount of steering.
-
-The Kaiser window parameter $\beta$ adjusts the compromise between resolution and sidelobe suppression.
-At $\beta=0$ it yields a rectangular window, and at and $\phi$ controls the amount of steering..
-$\beta=0$ ) and a gaussian curve (
 

### 2015-08-30 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:08:30	23:09:09	(Sun30.Aug)

@@
 \includegraphics[width=\linewidth]{gfx/mvdr_selected_windows_holmengraa.pdf}%
-\caption{\emph{Determining LCA window type}: Typical frequency responses for windows computed by a robust MVDR.\newline
+\caption{\emph{Determining LCA window type}: MVDR images with typical frequency responses chosen for various pixel regions.\newline
 \emph{Left:}\hfill
@@
 \caption{\emph{Determining Kaiser parameter boundaries}: LCA images created with a large and equal size window database, but where the upper bounds for $\beta$ and $\phi$ is varied.
-The upper left image is equal to DAS with rectangular window.
-While \Fig{mvdr_selected_windows} suggests that an upper bound
+The upper left image is equal to DAS with a rectangular window.
+For all images we measure the distance from the maximum to the -3dB point of the anchor
 \newline
 \emph{Upper bound $\beta$:}\hfill
-\parbox[t]{.9\linewidth}{.}\newline
-\emph{Right:}\hfill
-\parbox[t]{.95\linewidth}{.}
+\parbox[t]{.89\linewidth}{Observe that LCA gets better at suppressing sidelobes in the image as we increase the upper bounds for $\beta$.
+\Fig{mvdr_selected_windows} suggests MVDR hardly choose windows with maximum sidelobe level lower than that of Kaiser with $\beta=2$, but here we observe further improvement going to $\beta=5$.}\newline
+\emph{Upper bound $\phi$:}\hfill
+\parbox[t]{.89\linewidth}{As we increase the upper bound of the steering $\phi$, we also increase the lateral image resolution. }
 }\label{oversampling_mosaic_bounds}

### 2015-08-31 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:08:31	20:48:57	(Mon31.Aug)

@@
 Note how the responses are more or less symmetric, with very little steering in the shadow, moderate steering in speckle and steering within roughly 3dB in highlight.
-The phase varies mostly in highlight.} }\label{mvdr_selected_windows}
+The phase varies mostly in highlight to allow the steering there.} }\label{mvdr_selected_windows}
 \end{figure*}
@@
 
-To test the performance of the MVDR and LCA beamformers, we have processed data aquired by the Kongsberg Maritime HUGIN AUV carrying their 32 element HISAS1030 sonar~\cite{Hansen2009}.
+To test the performance of the MVDR and LCA beamformers, we have processed data acquired by the Kongsberg Maritime HUGIN AUV carrying their 32 element HISAS1030 sonar~\cite{Hansen2009}.
 It is a high resolution synthetic aperture sonar with array length 1.2\;m, 100\;kHz operating frequency, a bandwidth of 40\;kHz and 25$^\circ$ opening angle.
@@
 Two different scenes will be presented.
-One of the 1500 dwt oiltanker wreck Holmengraa lying at a slanted seabed at 77\;m depth outside of Horten, Norway~\cite{holmengraa}.
-It measures 68\;m by 9\;m and fills most of the highlit sector.
+One of the 1500 DWT oiltanker wreck Holmengraa lying at a slanted seabed at 77\;m depth outside of Horten, Norway~\cite{holmengraa}.
+It measures 68\;m by 9\;m and fills most of the highlighted sector.
 The other scene is of an iron cross connected to a an anchor (fig ???).
@@
 A sectorscan image of a single ping of data collected by the HISAS1030 syntetic aperture sonar is shown in \Fig{mvdr_selected_windows}, along with the window responses MVDR typically chooses in shadow, speckle and highlight regions in this image.
-A fairly agressive yet stable set of parameters for MVDR where used; subarray size $L=16$, temporal averaging of $K=1$, and $d=1\%$ of diagonal loading.
+A fairly aggressive yet stable set of parameters for MVDR where used; subarray size $L=16$, temporal averaging of $K=1$, and $d=1\%$ of diagonal loading.
 
@@
 The peak sidelobe levels are between that of the rectangular and Hamming window.
-
 This correspond to Kaiser windows with $\beta$-values of 0 and 2, respectively.
-\Fig{oversampling_mosaic_bounds_beta} illustrates which $\beta$-value LCA prefers for different regions in the cross image.
-
-Kaiser
+
+\Fig{oversampling_mosaic_bounds_beta} illustrates which $\beta$-value LCA prefers for different regions in the image of the cross.
 

### 2015-08-31 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:08:31	21:58:12	(Mon31.Aug)

@@
 The peak sidelobe levels are between that of the rectangular and Hamming window.
-This correspond to Kaiser windows with $\beta$-values of 0 and 2, respectively.
+This corresponds to Kaiser windows with $\beta$-values of 0 and 2, respectively.
 
@@
 
-Unlike the method described by Synne\aa{}g~\cite{Synnevag2008} we do not let LCA choose from  inverted Kaiser windows.
+Unlike the method described by Synnev\aa{}g~\cite{Synnevag2008} we do not let LCA choose from  inverted Kaiser windows.
 These have a spatial response with mainlobe width narrower than that of a rectangular window, but at the expense of drastically increased white noise gain.
@@
 
-\newpage
-
 % the performance of the LCA beamformer is surprisingly similar regardless of what windows it is allowed to choose from.
-
-\newpage
 

### 2015-08-31 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:08:31	22:28:39	(Mon31.Aug)

@@
 Bats, for instance, use a high frequency sonar to detect, identify and track prey with amazing precision in caves amongst a myriad of other bats.
-Human made sonars have little to no adaptibility when data is collected, but some methods have been investigated in the data post processing.
-Perhaps the most studied adaptive method for reconstructing a sonar image is the minimum variance distortionless response (MVDR) beamformer~\cite{Capon1969}.
+Human made sonars have little to no adaptability when data is collected, but some methods have been investigated in the data post processing.
+Perhaps the most studied adaptive method for reconstructing a sonar image is the minimum variance distortion-less response (MVDR) beamformer~\cite{Capon1969}.
 It computes the window that is optimal in the sense of suppressing noise as much as possible while under the restriction of  preserving the signal of interest.
@@
 It does this by applying a set of predefined windows and selects the one that is best able to suppress noise.
-Since it uses the same optimisation criterion as MVDR, it may be considered a version of MVDR with a reduced window solution space.
+Since it uses the same optimization criterion as MVDR, it may be considered a version of MVDR with a reduced window solution space.
 As we will demonstrate, this works well because a robust MVDR design tends to create windows that resembles slightly varied versions of conventional ones.
@@
 Of these a specific Kaiser window were micro-steered.
-We limit our study to the use of variations of the parametric Kaiser-Harris window, since we observed neglible performance difference to using this function or any of the common trigonometric ones.
+We limit our study to the use of variations of the parametric Kaiser-Harris window, since we observed negligible performance difference to using this function or any of the common trigonometric ones.
 The Kaiser windows are easy to compute, is fairly insensitive to coefficient inaccuracies, and can create windows with shapes spanning from rectangular to Gaussian.
@@
 Conventional beamformers all have static weights.
-The reference here is the delay-and-sum (DAS) beamformer, also know as the backprojection algortithm.
+The reference here is the delay-and-sum (DAS) beamformer, also know as the backprojection algorithm.
 It delays each pixel into focus, then applies a suitable window, and finally sums the data.
@@
 Adaptive beamformers allow the weights and delays to change to adjust the array response to better fit the incoming wavefield.
-One of the most extensively studied methods here is the minimum variance distortionless response (MVDR) beamformer.
+One of the most extensively studied methods here is the minimum variance distortion-less response (MVDR) beamformer.
 
@@
 A much less complex alternative to MVDR is the low complexity adaptive (LCA) beamformer.
-It iterates through a set of $P$ windows and selects the window $p$ that best fulfills the minimum variance criteron:
+It iterates through a set of $P$ windows and selects the window $p$ that best fulfills the minimum variance criterion:
 %
@@
 Note how closely LCA is related to the MVDR definition in \eq{eq_mvdr_definition}.
-The optimization criterion and constraint is the same, but LCA has a finite and discrete solutionspace for the weights.
+The optimization criterion and constraint is the same, but LCA has a finite and discrete solution space for the weights.
 As will be demonstrated in section \ref{lca_window_function}, LCA will perform similarly to MVDR because a robustified MVDR implementation seems to be constrained to window functions similar to the ones we let LCA choose from.
@@
 Two different scenes will be presented.
-One of the 1500 DWT oiltanker wreck Holmengraa lying at a slanted seabed at 77\;m depth outside of Horten, Norway~\cite{holmengraa}.
+One of the 1500 DWT oil tanker wreck Holmengraa lying at a slanted seabed at 77\;m depth outside of Horten, Norway~\cite{holmengraa}.
 It measures 68\;m by 9\;m and fills most of the highlighted sector.
@@
 \item \emph{Window parameter boundaries}.
-The Kaiser window parameter $\beta$ adjusts the compromise between resolution and sidelobe suppression.
-At $\beta=0$ it yields a rectangular window, and at and $\phi$ controls the amount of steering..
-$\beta=0$ ) and a gaussian curve (
+\Fig{oversampling_mosaic_bounds} show LCA images made with the same sized window database, but containing windows with varied upper boundaries for $\beta$ and $\phi$.
+We suggest an upper bound for $\beta$ between 2 and 5, where 5 is slightly better at suppressing sidelobes.
+For $\phi$ we suggest to steer either to 50\% or 100\% of the -3dB width, which leads to a resolution of 73\% or 54\% of the rectangular window, respectively.
+As lower boundaries we choose $\beta=0$ to always include a rectangular window, and $\phi=0$ to always include the centered window.
 \item \emph{Window database size}.
@@
 For this we selected a scene of the oil tanker wreck Holmengraa, which lies at a slanted seabed at nearly 100\;m depth.
-A sectorscan image of a single ping of data collected by the HISAS1030 syntetic aperture sonar is shown in \Fig{mvdr_selected_windows}, along with the window responses MVDR typically chooses in shadow, speckle and highlight regions in this image.
-A fairly aggressive yet stable set of parameters for MVDR where used; subarray size $L=16$, temporal averaging of $K=1$, and $d=1\%$ of diagonal loading.
+A sectorscan image of a single ping of data collected by the HISAS1030 synthetic aperture sonar is shown in \Fig{mvdr_selected_windows}, along with the window responses MVDR typically chooses in shadow, speckle and highlight regions in this image.
+A fairly aggressive yet stable set of parameters for MVDR where used; subarray size $L=16$, temporal averaging of $K=1$, and $d=1\%$ diagonal loading.
 
 While MVDR is free to choose non-symmetric windows, we can observe from \Fig{mvdr_selected_windows} that it prefers symmetric ones.
-In shadow and speckle the windows are fully symmetric, while in highlight the windows are approximately symmetric within the opening angle of the sonar.
+In shadow and speckle the windows are fully symmetric, while in highlight the windows are approximately symmetric within the highlighted sector of the sonar.
 The peak sidelobe levels are between that of the rectangular and Hamming window.
@@
 
-\subsection{Window parameters}
+\subsection{Window parameter boundaries}
 
 The Kaiser window parameter $\beta$ adjusts the compromise between resolution and sidelobe suppression, and $\phi$ controls the amount of steering.
-At $\beta=0$ it becomes a rectangular window, and at larger values it approaches a gaussian curve.
+At $\beta=0$ it becomes a rectangular window, and at larger values it approaches a Gaussian curve.
 Typical values are in the range $\beta\in[0,10]$.
 
+LCA is largely insensitive to the choice of windows and steering. 
+LCA works well for all the boundaries
 Variables of interest:
@@
 \item Steering. No steering in shadows, slight steering in speckle and no steering in highlight. Steering more than the -3dB width of the window function has no effect.
-\item Oversampling. From our observations one should have at least twice the number of beams as channels. Image is more visually pleasing and appears more detailed up to 8M.
+\item Oversampling. From our observations one should have at least twice the number of beams as channels. Image is more visually pleasing and appears more detailed up to $8M$.
 \end{itemize}

### 2015-09-01 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:09:01	11:23:41	(Tue 1.Sep)

@@
 \item \emph{Window parameter boundaries}.
-\Fig{oversampling_mosaic_bounds} show LCA images made with the same sized window database, but containing windows with varied upper boundaries for $\beta$ and $\phi$.
+\Fig{mvdr_selected_windows} show LCA images made with the same sized window database, but containing windows with varied upper boundaries for $\beta$ and $\phi$.
 We suggest an upper bound for $\beta$ between 2 and 5, where 5 is slightly better at suppressing sidelobes.

### 2015-09-01 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:09:01	11:55:38	(Tue 1.Sep)

@@
 Let $x_m[\theta,n]$ be the delayed data from the $m$th channel, where the $\theta$ and $n$ is the azimuth angle and range sample of the focus point, respectively.
-Each angle $\theta$ can be processed independently, so to simplify notation we will assume the dependence on $\theta$ to be implicit from now on.
-
-The beamformer output $z[n]$ can now be expressed as the weighted sum of all the delayed data samples:
+Each angle $\theta$ will be processed independently, so to simplify notation we will assume the dependence on $\theta$ to be implicit from now on.
+
+The output $z[n]$ of a beamformer is defined as the weighted sum of all the delayed data samples:
+%
 \begin{align}
@@
 \end{align}
+%
 where $w_m$ is the weight factor assigned to channel $m$.
-We will refer to a set of weights as a window, but it is also commonly called a taper function.
+A weight set is commonly called a window or a taper function.
 Applying a window that trails off towards the edges improves noise suppression at the cost of reduced resolution~\cite{Harris1978}.
@@
 Conventional beamformers all have static weights.
-The reference here is the delay-and-sum (DAS) beamformer, also know as the backprojection algorithm.
+The reference method is the delay-and-sum (DAS) beamformer, also know as the backprojection algorithm.
 It delays each pixel into focus, then applies a suitable window, and finally sums the data.
@@
 \begin{figure}[tbp]%
-\includegraphics[width=.9\linewidth]{gfx/calc_kaiser_3dB.eps}%
+\includegraphics[width=\linewidth]{gfx/calc_kaiser_3dB.eps}%
 \caption{Each Kaiser window is steered in the interval $\phi=[0, \phi_{\text{-3dB}}(\beta)]$.
@@
 With each window steered this way we expect the resolution gain to remain predictable and independent of $\beta$, and we also effectively constrain the white noise gain of the beamformer.}\label{windows_steering}
+\end{figure}
+
+\begin{figure}[tbp]%
+\includegraphics[width=\linewidth]{gfx/cross.jpg}%
+\caption{The cross and accompanying anchor used to test the beamformers.}\label{windows_steering}
 \end{figure}
@@
 \parbox[t]{.89\linewidth}{Observe that LCA gets better at suppressing sidelobes in the image as we increase the upper bounds for $\beta$.
-\Fig{mvdr_selected_windows} suggests MVDR hardly choose windows with maximum sidelobe level lower than that of Kaiser with $\beta=2$, but here we observe further improvement going to $\beta=5$.}\newline
+\Fig{mvdr_selected_windows} suggests that MVDR tend to choose windows with levels lower than that of Kaiser with $\beta=2$, but here we observe further improvement going to $\beta=5$.}\newline
 \emph{Upper bound $\phi$:}\hfill
@@
 
-To allow LCA some added flexibility we also create variations of each window that are steered a little to each side.
-%
-\begin{align}
-\w_{\beta,\phi} = \frac{\f_\beta\H\diag{(\a_\phi)}}{\f_\beta\H\a_\phi}\\
+Adding slightly steered versions of each window to the window database gives LCA greater flexibility in searching for optimal window. A Kaiser window $\f_\beta$ steered to $\phi$ can be expressed as 
+%
+\begin{align}
+\w_{\beta,\phi} = \frac{\f_\beta\H\diag{(\a_\phi)}}{\f_\beta\H\a_\phi}
 \end{align}

### 2015-09-01 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:09:01	13:38:35	(Tue 1.Sep)

@@
 
-Adaptive beamformers allow the weights and delays to change to adjust the array response to better fit the incoming wavefield.
-One of the most extensively studied methods here is the minimum variance distortion-less response (MVDR) beamformer.
+Adaptive beamformers are dynamic and seek to adjust the array response to better fit the incoming wavefield.
+This may be achieved by allowing either the weights or delays to change, or both.
+One of the most extensively studied of these methods is the minimum variance distortion-less response (MVDR) beamformer.
 
@@
 
-MVDR seeks to minimize the power of the beamformer, under the constraint of a unity gain in the desired direction $\phi$~\cite{Capon1969}:
+MVDR seeks to minimize the power of the beamformer, under the constraint of a unity gain in some desired direction $\phi$~\cite{Capon1969}:
 %
@@
 Finally, to improve robustness to parameter errors we apply $d$ percent diagonal loading~\cite{Cox1987,Maksym1979}.
-The steps are also needed to ensure that the covariance matrix is numerically well conditioned and hence invertible.
+These steps are also needed to ensure that the covariance matrix is numerically well conditioned and hence invertible.
 
@@
 which computes the sample power average over $N_k=2K+1$ temporal samples.
-This is the same as we did for MVDR.
-Our signal is only a few samples long after matched filtering so we have set $K=1$ for both MVDR and LCA.
+This is the same temporal averaging method we use for MVDR.
+In our case the matched filtered signal is a just few samples long, hence $K=1$ will be used for both MVDR and LCA throughout this work.
 

### 2015-09-01 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:09:01	14:09:51	(Tue 1.Sep)

@@
 \includegraphics[width=\linewidth]{gfx/mvdr_selected_windows_holmengraa.pdf}%
-\caption{\emph{Determining LCA window type}: MVDR images with typical frequency responses chosen for various pixel regions.\newline
+\caption{\emph{Determining LCA window type}: MVDR images with typical frequency responses for windows used in various pixel regions.\newline
 \emph{Left:}\hfill
@@
 \caption{$\beta$-values of the windows LCA selects for each image pixel.
-The underlying image is the one shown in \Fig{oversampling_mosaic_bounds}, and the location of the anchor and cut line is overlaid.
+The underlying image is the one shown in \Fig{oversampling_mosaic_bounds}, with the same parameters.
+The location of the anchor and cut line marked in red.
+Note how LCA prefers a rectangular window ($\beta=0$) on the anchor.
 }\label{oversampling_mosaic_bounds_beta}

### 2015-09-03 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:09:02	21:34:49	(Wed 2.Sep)
la2pdf commit - 2015:09:03	08:37:55	(Thu 3.Sep)
la2pdf commit - 2015:09:03	11:31:05	(Thu 3.Sep)

@@
 \includegraphics[width=\linewidth]{gfx/cross.jpg}%
-\caption{The cross and accompanying anchor used to test the beamformers.}\label{windows_steering}
+\caption{The cross and accompanying anchor used to test the beamformers.}\label{cross}
 \end{figure}
@@
 
+The window database we will construct will contain $N_\beta$ Kaiser windows with unique $\beta$-values, each being steered in $N_\phi$ different directions.
+This gives us $N_w = N_\beta N_\phi$ unique windows.
+
 \subsection{Steering bounds}
@@
 It measures 68\;m by 9\;m and fills most of the highlighted sector.
-The other scene is of an iron cross connected to a an anchor (fig ???).
+The other scene is of an iron cross connected to a an anchor (\Fig{cross}).
 

### 2015-09-03 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:09:03	12:01:55	(Thu 3.Sep)

@@
 \item \emph{Window parameter boundaries}.
-\Fig{mvdr_selected_windows} show LCA images made with the same sized window database, but containing windows with varied upper boundaries for $\beta$ and $\phi$.
+\Fig{oversampling_mosaic_bounds} show LCA images made with the same sized window database, but containing windows with varied upper boundaries for $\beta$ and $\phi$.
 We suggest an upper bound for $\beta$ between 2 and 5, where 5 is slightly better at suppressing sidelobes.
@@
 As lower boundaries we choose $\beta=0$ to always include a rectangular window, and $\phi=0$ to always include the centered window.
-\item \emph{Window database size}.
+\item \emph{Window database size}. \Fig{oversampling_mosaic}
 \item \emph{Performance}.
@@
 
-Assuming that a robust MVDR computes the optimal windows in the minimum variance sense, it makes sense to study them.
-For this we selected a scene of the oil tanker wreck Holmengraa, which lies at a slanted seabed at nearly 100\;m depth.
-A sectorscan image of a single ping of data collected by the HISAS1030 synthetic aperture sonar is shown in \Fig{mvdr_selected_windows}, along with the window responses MVDR typically chooses in shadow, speckle and highlight regions in this image.
+Assuming that a robust MVDR computes the optimal windows in the minimum variance sense, it seems sensible to create similar ones for LCA.
+We study this in \Fig{mvdr_selected_windows}, where an MVDR sectorscan image of the Holmengraa wreck is presented, along with the window responses MVDR chose in the shadow, speckle and highlight regions of this image.
 A fairly aggressive yet stable set of parameters for MVDR where used; subarray size $L=16$, temporal averaging of $K=1$, and $d=1\%$ diagonal loading.
@@
 While MVDR is free to choose non-symmetric windows, we can observe from \Fig{mvdr_selected_windows} that it prefers symmetric ones.
-In shadow and speckle the windows are fully symmetric, while in highlight the windows are approximately symmetric within the highlighted sector of the sonar.
+In shadow and speckle the windows are fully symmetric, while in highlight the windows are approximately symmetric within the illuminated\todo{valid term?} sector of the seafloor.
 The peak sidelobe levels are between that of the rectangular and Hamming window.
@@
 \Fig{oversampling_mosaic_bounds_beta} illustrates which $\beta$-value LCA prefers for different regions in the image of the cross.
+Red marker lines are used to pinpoint the cut line and location of the anchor.
+Observe that in the anchor region only rectangular windows are used, which is the reason for the anchor resolution being measure to the same value regardless of the upper bound of $\beta$ in \Fig{oversampling_mosaic_bounds}.
 
@@
 
-\mediaI{} shows how \Fig{mvdr_selected_windows} changes as a function of the MVDR subarray length $L$.
-It demonstrates that at $L$=1 the window responses are rectangular in all areas in the image.
-At $L$=2 we can observe that pixels in highlight are processed with windows with slightly varying amplitude responses, but with zero locations on similar locations and highlight regions being
+\mediaI{} shows how \Fig{mvdr_selected_windows} changes as a function of the MVDR subarray length $L$ and temporal averaging $K$.
+It demonstrates that at $L=1$ the window responses are rectangular in all areas in the image.
+At $L=2$ we can observe that pixels in highlight are processed with windows with slightly varying amplitude responses, but with the phase being 0$^\circ$ in the illuminated seafloor sector.
+Yet MVDR is able to greatly suppress noise, which seems to emphasize the importance varying the zero locations slightly to suppress noise.
+The media file also shows MVDR can be run with subarray sizes $L\in[M/2,5M/8]$, but only with temporal averaging $K=1$ or above.
 

### 2015-09-03 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:09:03	13:26:52	(Thu 3.Sep)

@@
 
-\setcounter{dbltopnumber}{2}
+% \setcounter{dbltopnumber}{2}
 
@@
 
-The Kaiser window parameter $\beta$ adjusts the compromise between resolution and sidelobe suppression, and $\phi$ controls the amount of steering.
-At $\beta=0$ it becomes a rectangular window, and at larger values it approaches a Gaussian curve.
-Typical values are in the range $\beta\in[0,10]$.
+\Fig{oversampling_mosaic_bounds} show LCA images made with the same sized window database, but containing windows with varied upper boundaries for $\beta$ and $\phi$.
+We suggest an upper bound for $\beta$ between 2 and 5, where 5 is slightly better at suppressing sidelobes.
+For $\phi$ we suggest to steer either to 50\% or 100\% of the -3dB width, which leads to a resolution of 73\% or 54\% of the rectangular window, respectively.
+As lower boundaries we choose $\beta=0$ to always include a rectangular window, and $\phi=0$ to always include the centered window.
+
+\subsection{Window database size}
+
+\subsection{Oversampling}
+
+From our observations one should have at least twice the number of beams as channels. Image is more visually pleasing and appears more detailed up to $8M$.
 
@@
 \begin{itemize}
-\item Window function. Mostly symmetrical, in particular in shadow and for the most part in speckle. In highlight regions more varying.
 \item Mainlobe width. Wide in shadows, narrow in detailed regions. Adding windows in between these extremal values.
@@
 \end{itemize}
-\begin{table}[!b]\centering%\normalsize
-\begin{tabular}[c]{l c c c}\hline
-\rowcolor{tabBlue} & \bf Highlight & \bf Shadow & \bf Speckle  \\\hline
-Symmetry             & Moderate  & High   & High \\
-Mainlobe width       & Narrow    & Wide   & Moderate \\
-Sidelobe suppression & Low       & High   & Moderate \\
-Steering             & Low       & Up to -3dB width & Moderate \\
-Oversampling         & 2x-8x     & 2x-8x  & 2x-8x
-\end{tabular}
-\caption{Summary}\label{tab:summary}
-\end{table}%
 
@@
 The tendency of the LCA to perform similarly to the MV beamformer seems to indicate that a full solution space is rarely needed in real world scenarios.
+
+\begin{table}[!b]\centering%\normalsize
+\begin{tabular}[c]{l c c c}\hline
+\rowcolor{tabBlue} & \bf Highlight & \bf Shadow & \bf Speckle  \\\hline
+Symmetry             & Moderate  & High   & High \\
+Mainlobe width       & Narrow    & Wide   & Moderate \\
+Sidelobe suppression & Low       & High   & Moderate \\
+Steering             & Low       & Up to -3dB width & Moderate \\
+Oversampling         & 2x-8x     & 2x-8x  & 2x-8x
+\end{tabular}
+\caption{Summary}\label{tab:summary}
+\end{table}%
 

### 2015-09-04 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:09:03	15:17:02	(Thu 3.Sep)
la2pdf commit - 2015:09:04	12:58:22	(Fri 4.Sep)

@@
 \includegraphics[width=\linewidth]{gfx/mvdr_selected_windows_holmengraa.pdf}%
-\caption{\emph{Determining LCA window type}: MVDR images with typical frequency responses for windows used in various pixel regions.\newline
+\caption{\emph{Determining LCA window type}: MVDR image with typical frequency responses for windows used in various pixel regions.\newline
 \emph{Left:}\hfill
@@
 The upper left image is equal to DAS with a rectangular window.
-For all images we measure the distance from the maximum to the -3dB point of the anchor
+For all images we measure the lateral distance $\Delta x$ between the two -3dB points of the anchor.
+These are specified relatively to the reference distance  $\Delta x_\text{ref}$ from the upper left DAS image.
 \newline
@@
 
-Adding slightly steered versions of each window to the window database gives LCA greater flexibility in searching for optimal window. A Kaiser window $\f_\beta$ steered to $\phi$ can be expressed as 
+Adding slightly steered versions of each window to the window database gives LCA greater flexibility in searching for optimal window.
+A Kaiser window $\f_\beta$ steered to the angle $\phi$ can be expressed as
 %
@@
 
-We call the upper bound for steering $\phi_{\text{-3dB}}(\beta)$, and define it as the steering angle needed for the steered response to have a -3dB crossing that is exactly half that of the unsteered window.
+We call the upper bound for steering $\phi_{\text{3dB}}(\beta)$, and define it as the steering angle needed for the steered response to have a -3dB crossing that is exactly half that of the unsteered window.
 Using a numerical search for the steering angle $\phi$ that fulfilled this criterion for a wide range of $\beta$-values, we obtained a third degree polynomial fit to these points:
@@
 
-Need $M$ in the image sector. Why???
-
-- Abs -> adds a factor 2
-- log -> infinity?
-- near field -> range and azimuth is coupled
+\todo{Need $M$ in the image sector. Why???}
+
+\todo{
+- Abs -> adds a factor 2\\
+- log -> infinity?\\
+- near field -> range and azimuth is coupled}
 
@@
 The peak sidelobe levels are between that of the rectangular and Hamming window.
-This corresponds to Kaiser windows with $\beta$-values of 0 and 2, respectively.
+
+All the responses in \Fig{mvdr_selected_windows} can be approximated with the Kaiser-Bessel function.
+By adjusting the window parameter $\beta$ we can adjust the trade-off between mainlobe width and sidelobe level.
+To generate windows with different mainlobe width and sidelobe level,
+
+A parametric window like Kaiser-Bessel allow us to create approximately these and the responses in between.
+
+The Kaiser-Bessel windows s with $\beta$-values of 0 and 2 allows us to approximately trade between these two boundaries. \todo{add the advantages here?}
+
+In our studies
 
@@
 
-\Fig{oversampling_mosaic_bounds} show LCA images made with the same sized window database, but containing windows with varied upper boundaries for $\beta$ and $\phi$.
+Letting LCA choose from Kaiser windows seems like a good choice due to their similarity to the 
+Which Kaiser windows should we include in the windows database? The parameter $\beta$ controls the window shape, and $\phi$ the steering.
+
+Having chosen to fill the window database with steered Kaiser windows, we should  function While having decided to use steered Kaiser windows, we still  To decide what the ideal boundaries for $\beta$ and $\phi$ should be, we constructed large and equally sized window databases containing windows with different parameter boundaries.
+The images produced are shown in \Fig{oversampling_mosaic_bounds}.
+To include and compare with a rectangular weighted DAS, we set the lower boundary to 0 for both parameters.
+
+varied upper boundaries for $\beta$ and $\phi$.
+
+ The upper left image has $\beta=0$ and no steering, and is thus equal to a rectangularly weighted DAS.
+
 We suggest an upper bound for $\beta$ between 2 and 5, where 5 is slightly better at suppressing sidelobes.
@@
 As lower boundaries we choose $\beta=0$ to always include a rectangular window, and $\phi=0$ to always include the centered window.
+
+LCA images created with a large and equal size window database, but where the upper bounds for $\beta$ and $\phi$ is varied.
+The upper left image is equal to DAS with a rectangular window.
+For all images we measure the lateral distance $\Delta x$ between the two -3dB points of the anchor.
+These are specified relatively to the reference distance  $\Delta x_\text{ref}$ from the upper left DAS image.
 

### 2015-09-05 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:09:05	15:27:56	(Sat 5.Sep)

@@
 As we will demonstrate, this works well because a robust MVDR design tends to create windows that resembles slightly varied versions of conventional ones.
-
-Synnev\aa{}g used a window set comprised of rectangular, Kaiser and inverted Kaiser functions.
+But which windows should LCA choose from?
+
+Synnev\aa{}g et.
+al.
+used a window set comprised of rectangular, Kaiser and inverted Kaiser functions\footnote{May also be called Kaiser-Harris windows.}.
 Of these a specific Kaiser window were micro-steered.
-We limit our study to the use of variations of the parametric Kaiser-Harris window, since we observed negligible performance difference to using this function or any of the common trigonometric ones.
-The Kaiser windows are easy to compute, is fairly insensitive to coefficient inaccuracies, and can create windows with shapes spanning from rectangular to Gaussian.
-
-In this work we investigate a few questions related to the LCA that remain unanswered. One is what window function we can use, and how many windows we will need.
-
-Another question is what it takes to achieve double the resolution.
-
-Using the LCA beamformer results in increased robustness and improved edge definition compared to the MV and DAS beamformers.
-
-We will show that when applied to experimental data, the weights chosen by a robustified MVDR yield responses that tend to be symmetric and steered within a fraction of the -3dB resolution of the respective window.
-Then we create Kaiser windows of different types, and with varying degree of steering, that combined form a discrete version of the MVDR solution space.
+His database had 12 windows in it, which proved to work well.
+But just how well did this work compared to MVDR?
+Can LCA perform even better with more windows, and could it work just as well with even fewer?
+
+In this work we aim to provide some intuitive understanding of how the window database may be designed.
+Like Synnev\aa{}g et.
+al.
+we limit our study to the use of variations of the parametric Kaiser window, since we observed negligible performance difference to using this function or any of the common trigonometric ones.
+However, we steer all our windows and not just some, and provide reason why the Kaiser window is a suitable one, what variations of Kaiser we should include, and how much it should be steered.
+
+In short, our findings show that when applied to experimental data, the weights chosen by a robustified MVDR yield responses that tend to be symmetric and steered within a fraction of the -3dB resolution of the respective window.
+When mimicking these responses with Kaiser windows, we obtain images with both improved noise suppression and resolution.
+We resolution gain can be made predictable and depending on the amount of steering we apply to the windows.
+Overall, the image quality produced by LCA is comparable to MVDR, and sometimes even better.
+LCA is also inherently robust, easy to implement and fairly easy to understand.
 
@@
 \begin{figure}[tbp]%
-\includegraphics[width=\linewidth]{gfx/cross.jpg}%
+\includegraphics[width=\linewidth]{gfx/cross_tagged.png}%
 \caption{The cross and accompanying anchor used to test the beamformers.}\label{cross}
@@
 In our studies
+
+The Kaiser windows are easy to compute, is fairly insensitive to coefficient inaccuracies, and can create windows with shapes spanning from rectangular to Gaussian.
 
@@
 
+Using the LCA beamformer results in increased robustness and improved edge definition compared to the MV and DAS beamformers.
+
 % if have a single appendix:

### 2015-09-06 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:09:05	19:04:33	(Sat 5.Sep)
la2pdf commit - 2015:09:06	13:33:56	(Sun 6.Sep)

@@
 
-In short, our findings show that when applied to experimental data, the weights chosen by a robustified MVDR yield responses that tend to be symmetric and steered within a fraction of the -3dB resolution of the respective window.
-When mimicking these responses with Kaiser windows, we obtain images with both improved noise suppression and resolution.
-We resolution gain can be made predictable and depending on the amount of steering we apply to the windows.
+Our findings show that when applied to experimental data, the weights chosen by a robustified MVDR yield responses that tend to be symmetric and steered within a fraction of the -3dB resolution of the respective window.
+When letting LCA choose from Kaiser windows with similar responses, we obtain images with both improved noise suppression and resolution.
+The resolution gain can be made predictable and depending on the amount of steering we apply to the windows.
 Overall, the image quality produced by LCA is comparable to MVDR, and sometimes even better.
@@
 \includegraphics[width=\linewidth]{gfx/cross_tagged.png}%
-\caption{The cross and accompanying anchor used to test the beamformers.}\label{cross}
+\caption{One of the objects used to test the beamformers is a 1\,m long cross attached to an anchor with a diameter of 13\,cm.}\label{cross}
 \end{figure}
@@
 The underlying image is the one shown in \Fig{oversampling_mosaic_bounds}, with the same parameters.
-The location of the anchor and cut line marked in red.
-Note how LCA prefers a rectangular window ($\beta=0$) on the anchor.
-}\label{oversampling_mosaic_bounds_beta}
+The location of the anchor and cut line is marked in red.
+Note how LCA prefers a rectangular window ($\beta=0$) on the anchor.}\label{oversampling_mosaic_bounds_beta}
 \end{figure*}
@@
 \includegraphics[width=\textwidth]{gfx/oversampling_mosaic.pdf}%
-\caption{LCA images using window sets with varying parameter boundaries. Oversampled image, 19x20 windows.}\label{oversampling_mosaic}
+\caption{LCA images using different window database sizes. The window boundaries are the same.}\label{oversampling_mosaic}
 \end{figure*}

### 2015-09-06 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:09:06	14:13:20	(Sun 6.Sep)

@@
 \includegraphics[width=\textwidth]{gfx/oversampling_mosaic.pdf}%
-\caption{LCA images using different window database sizes. The window boundaries are the same.}\label{oversampling_mosaic}
+\caption{LCA images using different window database sizes.
+The window boundaries are the same.
+The upper left image is identical to a rectangularly weighted DAS.
+Observe that sampling the $\beta$-range more finely increases the sidelobe suppression but leaves resolution unchanged.
+Adding more steering-variations improves both sidelobe suppression and resolution.
+However, using more than  $N_\beta=2$ window types and $N_\beta=3$ steering angles makes minimal difference.}\label{oversampling_mosaic}
 \end{figure*}
@@
 
-\newpage
 \section{Results and Discussion}
@@
 
-A quick summary of the results and the order they will be presented is as follows:
+\subsection{Summary}
+
+The results and the order they will be presented is as follows:
 
@@
 \subsection{Window database size}
+
+So far we have determined it reasonable to use Kaiser windows in the range $\beta\in[0,5]$, steered within the range $|\phi|\in[0,\phi_\mathrm{3dB}]$. But many windows do we need in these ranges?
+
+In \Fig{oversampling_mosaic} we have compared LCA images made from windows databases of different sizes, but with the same parameter boundaries.
+We included the results from using a single window ($N_\beta=1$) and steering angle ($N_\phi=1$), which corresponds to a rectangularly weighted DAS.
+
+Observe that sampling the $\beta$-range more finely increases the sidelobe suppression but leaves resolution unchanged.
+Adding more steering-variations improves both sidelobe suppression and resolution.
+
+However, using more than  $N_\beta=2$ window types and $N_\beta=3$ steering angles makes minimal
+
+\ref{oversampling_mosaic}
 

### 2015-09-06 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:09:06	22:05:20	(Sun 6.Sep)

@@
 It measures 68\;m by 9\;m and fills most of the highlighted sector.
-The other scene is of an iron cross connected to a an anchor (\Fig{cross}).
+The other scene is of an iron cross of 1\,m length in both directions, connected to a an anchor with a 13\,cm diameter (\Fig{cross}).
+
+MVDR was run with a fairly aggressive yet stable set of parameters; subarray size $L=16$, temporal averaging of $K=1$, and $d=1\%$ diagonal loading. Temporal averaging of $K=1$ was also used with LCA.
 
@@
 We study this in \Fig{mvdr_selected_windows}, where an MVDR sectorscan image of the Holmengraa wreck is presented, along with the window responses MVDR chose in the shadow, speckle and highlight regions of this image.
-A fairly aggressive yet stable set of parameters for MVDR where used; subarray size $L=16$, temporal averaging of $K=1$, and $d=1\%$ diagonal loading.
-
-While MVDR is free to choose non-symmetric windows, we can observe from \Fig{mvdr_selected_windows} that it prefers symmetric ones.
-In shadow and speckle the windows are fully symmetric, while in highlight the windows are approximately symmetric within the illuminated\todo{valid term?} sector of the seafloor.
-The peak sidelobe levels are between that of the rectangular and Hamming window.
-
+
+Observe from \Fig{mvdr_selected_windows} that MVDR seems prefers symmetric window responses, even if it is free to choose non-symmetric ones.
+The symmetry is most predominant in shadow and speckle regions, while in highlight the windows are approximately symmetric within the illuminated\todo{valid term?} sector of the seafloor.
+The symmetry is a side-effect of robustifying MVDR to handle coherent interference and the low sample support of covariance statistics.
+
+If we look at the peak sidelobe levels of the MVDR windows they are for the most part between -13\,dB and -43\,dB.
+This corresponds to that of the rectangular and Hamming window, or a Kaiser window with $\beta=0$ and $\beta=2$.
+Choosing another value for $\beta$ will  This suggests This and symmetry suggests
+
+Since the responses as similar to those 
 All the responses in \Fig{mvdr_selected_windows} can be approximated with the Kaiser-Bessel function.
@@
 
-\subsection{Window parameter boundaries}
-
-Letting LCA choose from Kaiser windows seems like a good choice due to their similarity to the 
-Which Kaiser windows should we include in the windows database? The parameter $\beta$ controls the window shape, and $\phi$ the steering.
-
-Having chosen to fill the window database with steered Kaiser windows, we should  function While having decided to use steered Kaiser windows, we still  To decide what the ideal boundaries for $\beta$ and $\phi$ should be, we constructed large and equally sized window databases containing windows with different parameter boundaries.
+\subsection{Kaiser parameter boundaries}
+
+What is the best set of Kaiser windows to use with LCA, and how much should they be steered?
+
+To investigate this we constructed large and equally sized window databases containing windows with varied upper boundaries for $\beta$ and $\phi$.
+As lower boundaries we chose $\beta=0$ and $\phi=\pm0^\circ$, which includes the rectangularly weighted DAS image.
 The images produced are shown in \Fig{oversampling_mosaic_bounds}.
-To include and compare with a rectangular weighted DAS, we set the lower boundary to 0 for both parameters.
-
-varied upper boundaries for $\beta$ and $\phi$.
-
- The upper left image has $\beta=0$ and no steering, and is thus equal to a rectangularly weighted DAS.
 
@@
 
-The LCA method performs similar to the MV method, but at a fraction of the computational cost.
-The key to achieve this success lies in the design of the window set.
+LCA can be viewed as either a simplification of MVDR, or as an extension to DAS.
+It creates images similar to MVDR, but at a fraction of the computational cost.
+It is also inherently robust and is largely insensitive to the exact set of windows it is run with, as long as they are sensible.
+
+Using experimental data we have demonstrated that when MVDR is made robust we constrain the solution space of the window coefficients it can choose from into mostly symmetric responses that varies between the rectangular and Hamming.
+We construct a window database using the parametric Kaiser function, which easily allows us to create responses similar to these and those in between.
+The Kaiser window is optimal in the sense of maximum power near the direction of interest.
+
+% The images produced by LCA appears sharper than those of MVDR, but 
+% 
+% The key to achieve this success lies in the design of the window set.
 Windows that yield vagely steered narrow responses are preferred in speckle regions, while wide and steered responses are typically preferred in highlight and shadow regions.
-A window set of 6 different responses each steered in 5 different directions proved sufficient in our experiment to match the performance of the MV beamformer.
+% 
+% A window set of 6 different responses each steered in 5 different directions proved sufficient in our experiment to match the performance of the MV beamformer.
 

### 2015-09-06 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:09:06	22:37:29	(Sun 6.Sep)

@@
 \author{Jo~Inge~Buskenes, %
-        Andreas~Austeng, %
-        Carl-Inge~Columbo~Nilsen%
+        Roy~Edgar~Hansen,%
+        Andreas~Austeng%
 \IEEEcompsocitemizethanks{\IEEEcompsocthanksitem All authors are with the Department
@@
 
-\begin{itemize}
-\item More steering -> more aggressive suppression of interference at the cost of boosted sidelobes / noise.
-\item More steering -> Larger variations in amplitude of sources
-\end{itemize}
-
 \subsection{Oversampling}
 
-Adaptive beamformers process each pixel in a different way in an attempt to surpass the conventional limit of resolution and noise suppression.
-This non-linear processing of the image increases its bandwidth and introduces a need for oversampling.
-For MVDR often a factor 10 lateral oversampling is often needed to ensure spatial shift-invariance in source amplitude~\cite{Asen2014}.
-To compare apples to apples, we will use the same factor for both LCA and MVDR.
+The ultimate goal of adaptive beamformers is to take the information available in the wavefield and use it to either improve image resolution, noise suppression, or both.
+This non-linear processing increases the image bandwidth and hence introduces a to oversample it to capture all the details.
+For MVDR a factor 10 oversampling laterally compared to the Nyquist rate is often needed to ensure  spatially local shift-invariance in source amplitude~\cite{Asen2014}.
+Since LCA operates with the same optimization criterion, we will be using approximately a factor 8 lateral oversampling for both beamformers.
 
@@
 
-Idea to steer within a fraction of the rectangular window's mainlobe.
-
-\begin{align}
-d = \frac{\pi\,\theta_\text{open}\,\lambda}{M\,d\,O_s}
-\end{align}
+% Idea to steer within a fraction of the rectangular window's mainlobe.
+
+% \begin{align}
+% d = \frac{\pi\,\theta_\text{open}\,\lambda}{M\,d\,O_s}
+% \end{align}
 
@@
 
-Assuming that a robust MVDR computes the optimal windows in the minimum variance sense, it seems sensible to create similar ones for LCA.
+Assuming that a robust MVDR is the reference method we want LCA to perform similarly to, it seems sensible to create window database for LCA with responses that resembles that of MVDR.
 We study this in \Fig{mvdr_selected_windows}, where an MVDR sectorscan image of the Holmengraa wreck is presented, along with the window responses MVDR chose in the shadow, speckle and highlight regions of this image.
@@
 If we look at the peak sidelobe levels of the MVDR windows they are for the most part between -13\,dB and -43\,dB.
-This corresponds to that of the rectangular and Hamming window, or a Kaiser window with $\beta=0$ and $\beta=2$.
-Choosing another value for $\beta$ will  This suggests This and symmetry suggests
-
-Since the responses as similar to those 
-All the responses in \Fig{mvdr_selected_windows} can be approximated with the Kaiser-Bessel function.
-By adjusting the window parameter $\beta$ we can adjust the trade-off between mainlobe width and sidelobe level.
-To generate windows with different mainlobe width and sidelobe level,
-
-A parametric window like Kaiser-Bessel allow us to create approximately these and the responses in between.
-
-The Kaiser-Bessel windows s with $\beta$-values of 0 and 2 allows us to approximately trade between these two boundaries. \todo{add the advantages here?}
-
-In our studies
-
+This corresponds to that of the rectangular and Hamming window, or a Kaiser window with $\beta=0$ and $\beta=2$, respectively.
+Using the Kaiser window in this case makes sense since it is optimal in the sense of having its peak power concentration near the steering angle, and because we can also create any response in between.
 The Kaiser windows are easy to compute, is fairly insensitive to coefficient inaccuracies, and can create windows with shapes spanning from rectangular to Gaussian.
@@
 
-\subsection{Kaiser parameter boundaries}
+\subsection{Kaiser parameter $\beta$ and steering $\phi$}
 
@@
 These are specified relatively to the reference distance  $\Delta x_\text{ref}$ from the upper left DAS image.
+
+\begin{itemize}
+\item More steering -> more aggressive suppression of interference at the cost of boosted sidelobes / noise.
+\item More steering -> Larger variations in amplitude of sources
+\end{itemize}
 
@@
 
-\newcommand\cc[1]{\multicolumn{1}{>{\columncolor{tabBlue}}c}{\bf #1}}
-\begin{table}[!b]\centering%\normalsize
-\begin{tabular}[c]{l r r r@{}  r r r}\hline
-\rowcolor{tabBlue} & \cc{Resolution} & \cc{S} & \cc{$\beta$} & \cc{$\varphi^*$} &  \cc{N$_\beta$} & \cc{N$_\varphi$} \\\hline
-& 50\%  & 28 & [0,3] & \hspace{.15cm} 50\% & 2 & 3 \\
-& 50\%  & 28 & [0,6] & \hspace{.15cm} 50\% & 2 & 3 \\
-& 50\%  & 28 & [0,9] & \hspace{.15cm} 50\% & 2 & 3 \\
-& 100\% & 32 &  &:30 \\
-& 150\% & 32 & 1 &:4\\
-\end{tabular}\\
-* Steering is relative to the -3dB width of the window being applied.
-\caption{Qualitative study}\label{parameter_study}
-\end{table}%
+% \newcommand\cc[1]{\multicolumn{1}{>{\columncolor{tabBlue}}c}{\bf #1}}
+% \begin{table}[!b]\centering%\normalsize
+% \begin{tabular}[c]{l r r r@{}  r r r}\hline
+% \rowcolor{tabBlue} & \cc{Resolution} & \cc{S} & \cc{$\beta$} & \cc{$\varphi^*$} &  \cc{N$_\beta$} & \cc{N$_\varphi$} \\\hline
+% & 50\%  & 28 & [0,3] & \hspace{.15cm} 50\% & 2 & 3 \\
+% & 50\%  & 28 & [0,6] & \hspace{.15cm} 50\% & 2 & 3 \\
+% & 50\%  & 28 & [0,9] & \hspace{.15cm} 50\% & 2 & 3 \\
+% & 100\% & 32 &  &:30 \\
+% & 150\% & 32 & 1 &:4\\
+% \end{tabular}\\
+% * Steering is relative to the -3dB width of the window being applied.
+% \caption{Qualitative study}\label{parameter_study}
+% \end{table}%
 
@@
 The Kaiser window is optimal in the sense of maximum power near the direction of interest.
+
+% We limit our study to the use of variations of the parametric Kaiser-Harris window, since we observed neglible performance difference to using this function or any of the common trigonometric ones.
+The Kaiser windows are easy to compute, is fairly insensitive to coefficient inaccuracies, and can create windows with shapes spanning from rectangular to Gaussian.
+% 
+% 
+% Using the LCA beamformer results in increased robustness and improved edge definition compared to the MV and DAS beamformers.
 
@@
 
-\subsection{Summary and order of results}
-
-In short, our findings regarding the optimal window types for LCA is as follows:
-
-\begin{itemize}
-\item \emph{Window function} should be symmetric, . Studying typical MVDR responses for shadow, speckle and highlight regions, we believe it to be insignificant
-\item \emph{Window parameters}. 
-\end{itemize}
-
-The following sections with expand and discuss these results further.
-
-\section{Introduction}
-
 % \begin{itemize}
-% \item Sell it!
-% \item As briefly as possible, introduce beamforming.
 % \item Adaptive beamformer's potential lies in its ability to suppress interference power
@@
 % \end{itemize}
-% \item LCA is like DAS, only we test several windows and choose the one that yield the lowest beamformer output power.
-Ultrasound \cite{Synnevag2008}, active sonar \cite{Buskenes2011}\cite{Blomberg2011}\cite{Blomberg2012}\todo{who was that guy in the 80ties?}
 % \item Outline: 
 % \begin{itemize}
-% \item Methods.
-% \item See what happens to Capon's response %\cite{Synnevag2007}  when means for robustification are added.
 % \item Setup the window set for LCA to match this.
@@
 
-\section{Methods}
-
-% \todopar{Threw some formulas in to get a feel of them}
-
-\begin{align}
-z[n] = \sumb{m=0}{M-1} w_m[n]^*x_m[n-\Delta_m] = \w\H[n]\Xd[n]
-\end{align}
-where
-\begin{align}
-\w[n] = \bmat{w_0[n]\\w_1[n]\\\vdots\\w_{M-1}[n]} \quad \text{and} \quad\Xd[n] = \bmat{x_0[n]\\x_1[n]\\\vdots\\x_{M-1}[n]}.
-\end{align}
-
-Minimum variance distortionless resonse \cite{Capon1969}
-\begin{align}
-\underset{\w[n]}{\argmin}\, E\{ |z[n]|^2 \} = \underset{\w[n]}{\argmin}\, \w\H[n]\R[n]\w[n], 
-\end{align}
-subject to
-\begin{align}
-\w\H[n]\a = 1,
-\end{align}
-where
-\begin{align}
-\R[n] = E\{ \x\x\H \}.
-\end{align}
-
-\section{Results}
-
-% \begin{figure}[!t]\centering
-% % \includegraphics[width=2.5in]{}
-% \caption{Simulation Results}
-% \label{fig_sim}
-% \end{figure}
-
-See end of document...
-% \begin{figure*}[!t]
-% \centerline{\subfloat[Case I]\includegraphics[width=2.5in]{subfigcase1}%
-% \label{fig_first_case}}
-% \hfil
-% \subfloat[Case II]{\includegraphics[width=2.5in]{subfigcase2}%
-% \label{fig_second_case}}}
-% \caption{Simulation results}
-% \label{fig_sim}
-% \end{figure*}
-
-\section{Conclusion}
-
-\begin{itemize}
-\item Something similar to the abstract.
-\end{itemize}
-
-% \ \\
 % \IEEEPARstart{T}{o} form images from a modern phased array sonar system the received wavefield is usually recorded, and then postprocessed by a digital beamformer.
@@
 % \begin{itemize}
-% \item Capon's resonse when applying robustification
 % \item Choice of window functions makes little difference.
@@
 %
-% 
-% \begin{align}
-% z[n] = \sumb{m=0}{M-1} w_m[n]^*x_m[n-\Delta_m] = \w\H[n]\x[n-\Delta_m]
-% \end{align}
-% 
-% 
-% \section{Methods}
-% 
-% Basically, we are working with a practical implementation of the Capon beamformer that computes a set of weights $\vec w$ for every single sample $n$ by solving:
-% \begin{gather*}
-% \vec w[n] = \frac{\hat{\mat R}\,\!^{-1}[n]\vec a}{\vec a\H\hat{\mat R}\,\!^{-1}[n]\vec a} = \frac{\text{\raisebox{1.9pt}{$\vec\chi$}}[n]}{\vec a\H\text{\raisebox{1.9pt}{$\vec\chi$}}[n]}
-% \end{gather*}%
-% where
-% % \newcommand\X{\text{\raisebox{2pt}{$\vec\chi$}}}
-% \begin{gather*}
-% \text{\raisebox{1.9pt}{$\vec\chi$}} = \hat{\mat R}\,\!^{-1}\vec a \qquad\qquad\text{is the solution to}\qquad\qquad \hat{\mat R}\text{\raisebox{1.9pt}{$\vec\chi$}} = \vec a.
-% \end{gather*}
-
-% 
+
 % , and maximum suppression of while ensuring that the beamformer digitally  before each of the pixels are estimated one at a time.
 The resolution and contrast of such a system will depend on the systems spatial response, which ideally should be narrow  be very sharp in the desired direction its ability to achieve  fundamental principle of forming a sonar image is to record the received wavefield,
-% 
-% image quality of a phased array sonar imaging systems depend on  the choice of weights to apply to each of the sensors are crucial. 
-% 
-% A modern phased array imaging system may be thought of as a spatial filter. To achieve the best possible performance, the 
-% 
-% resolution and contrast 
-% 
-% Adaptive beamformers have only recently been introduced in active sonar imaging.
-For a while they were considered unsuited for this purpose because the backscattered signal is largely correlated with the
-% 
-% 
-%
-
-%\begin{figure}[!t]
-%\centering
-%\includegraphics[width=2.5in]{myfigure}
-% where an .eps filename suffix will be assumed under latex, 
-% and a .pdf suffix will be assumed for pdflatex; or what has been declared
-% via \DeclareGraphicsExtensions.
-%\caption{Simulation Results}
-%\label{fig_sim}
-%\end{figure}
-
-% An example of a double column floating figure using two subfigures.
-% (The subfig.sty package must be loaded for this to work.)
-% The subfigure \label commands are set within each subfloat command, the
-% \label for the overall figure must come after \caption.
-% \hfil must be used as a separator to get equal spacing.
-% The subfigure.sty package works much the same way, except \subfigure is
-% used instead of \subfloat.
-%
-% \begin{figure*}[!t]
-% \centerline{\subfloat[Case I]\includegraphics[width=2.5in]{subfigcase1}%
-% \label{fig_first_case}}
-% \hfil
-% \subfloat[Case II]{\includegraphics[width=2.5in]{subfigcase2}%
-% \label{fig_second_case}}}
-% \caption{Simulation results}
-% \label{fig_sim}
-% \end{figure*}
-%
-% Note that often IEEE papers with subfigures do not employ subfigure
-% captions (using the optional argument to \subfloat), but instead will
-% reference/describe all of them (a), (b), etc., within the main caption.
-
-% An example of a floating table. Note that, for IEEE style tables, the 
-% \caption command should come BEFORE the table. Table text will default to
-% \footnotesize as IEEE normally uses this smaller font for tables.
-% The \label must come after \caption as always.
-%
-%\begin{table}[!t]
-%% increase table row spacing, adjust to taste
-%\renewcommand{\arraystretch}{1.3}
-% if using array.sty, it might be a good idea to tweak the value of
-% \extrarowheight as needed to properly center the text within the cells
-%\caption{An Example of a Table}
-%\label{table_example}
-%\centering
-%% Some packages, such as MDW tools, offer better commands for making tables
-%% than the plain LaTeX2e tabular which is used here.
-%\begin{tabular}{|c||c|}
-%\hline
-%One & Two\\
-%\hline
-%Three & Four\\
-%\hline
-%\end{tabular}
-%\end{table}
-
-% Note that IEEE does not put floats in the very first column - or typically
-% anywhere on the first page for that matter. Also, in-text middle ("here")
-% positioning is not used. Most IEEE journals use top floats exclusively.
-% However, Computer Society journals sometimes do use bottom floats - bear
-% this in mind when choosing appropriate optional arguments for the
-% figure/table environments.
-% Note that, LaTeX2e, unlike IEEE journals, places footnotes above bottom
-% floats. This can be corrected via the \fnbelowfloat command of the
-% stfloats package.
-
-\section{Conclusion}
-The conclusion goes here.
-
-We limit our study to the use of variations of the parametric Kaiser-Harris window, since we observed neglible performance difference to using this function or any of the common trigonometric ones.
-The Kaiser windows are easy to compute, is fairly insensitive to coefficient inaccuracies, and can create windows with shapes spanning from rectangular to Gaussian.
-
-Using the LCA beamformer results in increased robustness and improved edge definition compared to the MV and DAS beamformers.
-
-% if have a single appendix:
-%\appendix[Proof of the Zonklar Equations]
-% or
-%\appendix  % for no appendix heading
-% do not use \section anymore after \appendix, only \section*
-% is possibly needed
-
-% use appendices with more than one appendix
-% then use \section to start each appendix
-% you must declare a \section before using any
-% \subsection or using \label (\appendices by itself
-% starts a section numbered zero.)
-%
 
@@
 % 
-% \input{bio/jo_inge_buskenes}
-% % \input{bio/jon_petter_aasen}
-% % \input{bio/carl-inge_nilsen}
-% \input{bio/andreas_austeng}
+\input{bio/jo_inge_buskenes}
+\input{bio/andreas_austeng}
 

### 2015-09-07 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:09:07	11:08:35	(Mon 7.Sep)

@@
 
+The results and the order they will be presented is as follows:
+
 \subsection{Summary}
-
-The results and the order they will be presented is as follows:
 
@@
 Red marker lines are used to pinpoint the cut line and location of the anchor.
-Observe that in the anchor region only rectangular windows are used, which is the reason for the anchor resolution being measure to the same value regardless of the upper bound of $\beta$ in \Fig{oversampling_mosaic_bounds}.
+Observe that in the anchor region only rectangular windows are used, which is the reason for the anchor resolution being measured to the same value regardless of the upper bound of $\beta$ in \Fig{oversampling_mosaic_bounds}.
 
@@
 What is the best set of Kaiser windows to use with LCA, and how much should they be steered?
-
 To investigate this we constructed large and equally sized window databases containing windows with varied upper boundaries for $\beta$ and $\phi$.

### 2015-09-07 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:09:07	12:31:14	(Mon 7.Sep)

@@
 The ultimate goal of adaptive beamformers is to take the information available in the wavefield and use it to either improve image resolution, noise suppression, or both.
-This non-linear processing increases the image bandwidth and hence introduces a to oversample it to capture all the details.
-For MVDR a factor 10 oversampling laterally compared to the Nyquist rate is often needed to ensure  spatially local shift-invariance in source amplitude~\cite{Asen2014}.
+This non-linear processing increases the image bandwidth introduces a need for a sampling rate higher than the Nyquist rate.
+For MVDR a factor 10 oversampling laterally compared to the Nyquist rate is often needed to ensure  spatialhift-invariance in source amplitude~\cite{Asen2014}.
 Since LCA operates with the same optimization criterion, we will be using approximately a factor 8 lateral oversampling for both beamformers.
@@
 
-% \section{Experimental Setup}
-% 
-% For the LCA to perform well it is important to let it have a diverse selection of windows to choose from.
-The question is how this selection should be With our ovarall goal being a reduction of computational complexity, it The choice of window function does not seem to matter.
-% 
-% 
-% 
-% Using Kaiser windows as their extremal values areLimit our study to Kaiser windows
-% 
-% 
 % 
@@
 
-% Furthermore, the algorithm maps well to single-instruction-multiple-data (SIMD) hardware such as graphics processing units (GPUs).
-We show that even a modest amount of optimisation work has resulted in a factor 10 speed-up of this algorithm when implemented on a GPU as opposed to on a CPU.
-
 \section{Results and Discussion}
@@
 
-MVDR was run with a fairly aggressive yet stable set of parameters; subarray size $L=16$, temporal averaging of $K=1$, and $d=1\%$ diagonal loading. Temporal averaging of $K=1$ was also used with LCA.
-
-The results and the order they will be presented is as follows:
+MVDR was run with a fairly aggressive yet stable set of parameters; subarray size $L=16$ and $d=1\%$ diagonal loading. Both MVDR and LCA was run with $K=1$ temporal averaging.
+
+The results will be presented in the following order: In section \ref{lca_window_function} we discuss typical window responses computed by MVDR (\Fig{mvdr_selected_windows}), and hypothesize  that these can be mimicked by Kaiser windows.
+In section \ref{lca_kaiser_parameters} we investigate what sensible boundaries for the Kaiser parameter $\beta$ and window steering $\phi$ should be (\ref{oversampling_mosaic_bounds}).
+
+\subsection{Window function}\label{lca_window_function}
+
+Assuming that a robust MVDR is the reference method we want LCA to perform similarly to, it seems sensible to create window database for LCA with responses that resembles that of MVDR.
+The typical responses MVDR creates for shadow, speckle and highlight regions of a sectorscan Holmengraa image is shown in \Fig{mvdr_selected_windows}.
+
+Observe from \Fig{mvdr_selected_windows} that MVDR seems prefers symmetric window responses, even if it is free to choose non-symmetric ones.
+The symmetry is most predominant in shadow and speckle regions, while in highlight the windows are approximately symmetric within the illuminated\todo{valid term?} sector of the seafloor.
+The symmetry is a side-effect of robustifying MVDR to handle coherent interference and the low sample support of covariance statistics.
+
+If we look at the peak sidelobe levels of the MVDR windows they are for the most part between -13\,dB and -43\,dB.
+This corresponds to that of the rectangular and Hamming window, or a Kaiser window with $\beta=0$ and $\beta=2$, respectively.
+Using the Kaiser window in this case makes sense since it is optimal in the sense of having its peak power concentration near the steering angle, and because we can also create any response in between.
+The Kaiser windows are easy to compute, is fairly insensitive to coefficient inaccuracies, and can create windows with shapes spanning from rectangular to Gaussian.
+
+\Fig{oversampling_mosaic_bounds_beta} illustrates which $\beta$-value LCA prefers for different regions in the image of the cross.
+Red marker lines are used to pinpoint the cut line and location of the anchor.
+Observe that in the anchor region only rectangular windows are used, which is the reason for the anchor resolution being measured to the same value regardless of the upper bound of $\beta$ in \Fig{oversampling_mosaic_bounds}.
+
+Unlike the method described by Synnev\aa{}g~\cite{Synnevag2008} we do not let LCA choose from  inverted Kaiser windows.
+These have a spatial response with mainlobe width narrower than that of a rectangular window, but at the expense of drastically increased white noise gain.
+We have found that these windows hardly ever get used on experimental data, in particular steered versions which ever worsened white noise gain.
+We can infer the same from \Fig{mvdr_selected_windows} by noting that the maximum sidelobe level rarely exceeds the rectangular window level of -13\;dB.
+
+\mediaI{} shows how \Fig{mvdr_selected_windows} changes as a function of the MVDR subarray length $L$ and temporal averaging $K$.
+It demonstrates that at $L=1$ the window responses are rectangular in all areas in the image.
+At $L=2$ we can observe that pixels in highlight are processed with windows with slightly varying amplitude responses, but with the phase being 0$^\circ$ in the illuminated seafloor sector.
+Yet MVDR is able to greatly suppress noise, which seems to emphasize the importance varying the zero locations slightly to suppress noise.
+The media file also shows MVDR can be run with subarray sizes $L\in[M/2,5M/8]$, but only with temporal averaging $K=1$ or above.
+
+\subsection{Kaiser parameter $\beta$ and steering $\phi$}\label{lca_kaiser_parameters}
+
+What is the best set of Kaiser windows to use with LCA, and how much should they be steered?
+To investigate this we constructed large and equally sized window databases containing windows with varied upper boundaries for $\beta$ and $\phi$.
+As lower boundaries we chose $\beta=0$ and $\phi=\pm0^\circ$, which includes the rectangularly weighted DAS image.
+The images produced are shown in \Fig{oversampling_mosaic_bounds}.
+
+We suggest an upper bound for $\beta$ between 2 and 5, where 5 is slightly better at suppressing sidelobes.
+For $\phi$ we suggest to steer either to 50\% or 100\% of the -3dB width, which leads to a resolution of 73\% or 54\% of the rectangular window, respectively.
+As lower boundaries we choose $\beta=0$ to always include a rectangular window, and $\phi=0$ to always include the centered window.
+
+LCA images created with a large and equal size window database, but where the upper bounds for $\beta$ and $\phi$ is varied.
+The upper left image is equal to DAS with a rectangular window.
+For all images we measure the lateral distance $\Delta x$ between the two -3dB points of the anchor.
+These are specified relatively to the reference distance  $\Delta x_\text{ref}$ from the upper left DAS image.
+
+\begin{itemize}
+\item More steering -> more aggressive suppression of interference at the cost of boosted sidelobes / noise.
+\item More steering -> Larger variations in amplitude of sources
+\end{itemize}
+
+\subsection{Window database size}
+
+So far we have determined it reasonable to use Kaiser windows in the range $\beta\in[0,5]$, steered within the range $|\phi|\in[0,\phi_\mathrm{3dB}]$. But many windows do we need in these ranges?
+
+In \Fig{oversampling_mosaic} we have compared LCA images made from windows databases of different sizes, but with the same parameter boundaries.
+We included the results from using a single window ($N_\beta=1$) and steering angle ($N_\phi=1$), which corresponds to a rectangularly weighted DAS.
+
+Observe that sampling the $\beta$-range more finely increases the sidelobe suppression but leaves resolution unchanged.
+Adding more steering-variations improves both sidelobe suppression and resolution.
+
+However, using more than  $N_\beta=2$ window types and $N_\beta=3$ steering angles makes minimal
+
+\ref{oversampling_mosaic}
+
+\subsection{Oversampling}
+
+From our observations one should have at least twice the number of beams as channels. Image is more visually pleasing and appears more detailed up to $8M$.
+
+LCA is largely insensitive to the choice of windows and steering. 
+LCA works well for all the boundaries
+Variables of interest:
+
+\begin{itemize}
+\item Mainlobe width. Wide in shadows, narrow in detailed regions. Adding windows in between these extremal values.
+\item Steering. No steering in shadows, slight steering in speckle and no steering in highlight. Steering more than the -3dB width of the window function has no effect.
+\item Oversampling. From our observations one should have at least twice the number of beams as channels. Image is more visually pleasing and appears more detailed up to $8M$.
+\end{itemize}
+
+\subsection{Computational complexity}
+
+% Furthermore, the algorithm maps well to single-instruction-multiple-data (SIMD) hardware such as graphics processing units (GPUs).
+We show that even a modest amount of optimisation work has resulted in a factor 10 speed-up of this algorithm when implemented on a GPU as opposed to on a CPU.
+
+% the performance of the LCA beamformer is surprisingly similar regardless of what windows it is allowed to choose from.
 
@@
 \end{itemize}
-
-The following sections with expand and discuss these results further.
-
-\subsection{Window function}\label{lca_window_function}
-
-Assuming that a robust MVDR is the reference method we want LCA to perform similarly to, it seems sensible to create window database for LCA with responses that resembles that of MVDR.
-We study this in \Fig{mvdr_selected_windows}, where an MVDR sectorscan image of the Holmengraa wreck is presented, along with the window responses MVDR chose in the shadow, speckle and highlight regions of this image.
-
-Observe from \Fig{mvdr_selected_windows} that MVDR seems prefers symmetric window responses, even if it is free to choose non-symmetric ones.
-The symmetry is most predominant in shadow and speckle regions, while in highlight the windows are approximately symmetric within the illuminated\todo{valid term?} sector of the seafloor.
-The symmetry is a side-effect of robustifying MVDR to handle coherent interference and the low sample support of covariance statistics.
-
-If we look at the peak sidelobe levels of the MVDR windows they are for the most part between -13\,dB and -43\,dB.
-This corresponds to that of the rectangular and Hamming window, or a Kaiser window with $\beta=0$ and $\beta=2$, respectively.
-Using the Kaiser window in this case makes sense since it is optimal in the sense of having its peak power concentration near the steering angle, and because we can also create any response in between.
-The Kaiser windows are easy to compute, is fairly insensitive to coefficient inaccuracies, and can create windows with shapes spanning from rectangular to Gaussian.
-
-\Fig{oversampling_mosaic_bounds_beta} illustrates which $\beta$-value LCA prefers for different regions in the image of the cross.
-Red marker lines are used to pinpoint the cut line and location of the anchor.
-Observe that in the anchor region only rectangular windows are used, which is the reason for the anchor resolution being measured to the same value regardless of the upper bound of $\beta$ in \Fig{oversampling_mosaic_bounds}.
-
-Unlike the method described by Synnev\aa{}g~\cite{Synnevag2008} we do not let LCA choose from  inverted Kaiser windows.
-These have a spatial response with mainlobe width narrower than that of a rectangular window, but at the expense of drastically increased white noise gain.
-We have found that these windows hardly ever get used on experimental data, in particular steered versions which ever worsened white noise gain.
-We can infer the same from \Fig{mvdr_selected_windows} by noting that the maximum sidelobe level rarely exceeds the rectangular window level of -13\;dB.
-
-\mediaI{} shows how \Fig{mvdr_selected_windows} changes as a function of the MVDR subarray length $L$ and temporal averaging $K$.
-It demonstrates that at $L=1$ the window responses are rectangular in all areas in the image.
-At $L=2$ we can observe that pixels in highlight are processed with windows with slightly varying amplitude responses, but with the phase being 0$^\circ$ in the illuminated seafloor sector.
-Yet MVDR is able to greatly suppress noise, which seems to emphasize the importance varying the zero locations slightly to suppress noise.
-The media file also shows MVDR can be run with subarray sizes $L\in[M/2,5M/8]$, but only with temporal averaging $K=1$ or above.
-
-\subsection{Kaiser parameter $\beta$ and steering $\phi$}
-
-What is the best set of Kaiser windows to use with LCA, and how much should they be steered?
-To investigate this we constructed large and equally sized window databases containing windows with varied upper boundaries for $\beta$ and $\phi$.
-As lower boundaries we chose $\beta=0$ and $\phi=\pm0^\circ$, which includes the rectangularly weighted DAS image.
-The images produced are shown in \Fig{oversampling_mosaic_bounds}.
-
-We suggest an upper bound for $\beta$ between 2 and 5, where 5 is slightly better at suppressing sidelobes.
-For $\phi$ we suggest to steer either to 50\% or 100\% of the -3dB width, which leads to a resolution of 73\% or 54\% of the rectangular window, respectively.
-As lower boundaries we choose $\beta=0$ to always include a rectangular window, and $\phi=0$ to always include the centered window.
-
-LCA images created with a large and equal size window database, but where the upper bounds for $\beta$ and $\phi$ is varied.
-The upper left image is equal to DAS with a rectangular window.
-For all images we measure the lateral distance $\Delta x$ between the two -3dB points of the anchor.
-These are specified relatively to the reference distance  $\Delta x_\text{ref}$ from the upper left DAS image.
-
-\begin{itemize}
-\item More steering -> more aggressive suppression of interference at the cost of boosted sidelobes / noise.
-\item More steering -> Larger variations in amplitude of sources
-\end{itemize}
-
-\subsection{Window database size}
-
-So far we have determined it reasonable to use Kaiser windows in the range $\beta\in[0,5]$, steered within the range $|\phi|\in[0,\phi_\mathrm{3dB}]$. But many windows do we need in these ranges?
-
-In \Fig{oversampling_mosaic} we have compared LCA images made from windows databases of different sizes, but with the same parameter boundaries.
-We included the results from using a single window ($N_\beta=1$) and steering angle ($N_\phi=1$), which corresponds to a rectangularly weighted DAS.
-
-Observe that sampling the $\beta$-range more finely increases the sidelobe suppression but leaves resolution unchanged.
-Adding more steering-variations improves both sidelobe suppression and resolution.
-
-However, using more than  $N_\beta=2$ window types and $N_\beta=3$ steering angles makes minimal
-
-\ref{oversampling_mosaic}
-
-\subsection{Oversampling}
-
-From our observations one should have at least twice the number of beams as channels. Image is more visually pleasing and appears more detailed up to $8M$.
-
-LCA is largely insensitive to the choice of windows and steering. 
-LCA works well for all the boundaries
-Variables of interest:
-
-\begin{itemize}
-\item Mainlobe width. Wide in shadows, narrow in detailed regions. Adding windows in between these extremal values.
-\item Steering. No steering in shadows, slight steering in speckle and no steering in highlight. Steering more than the -3dB width of the window function has no effect.
-\item Oversampling. From our observations one should have at least twice the number of beams as channels. Image is more visually pleasing and appears more detailed up to $8M$.
-\end{itemize}
-
-% the performance of the LCA beamformer is surprisingly similar regardless of what windows it is allowed to choose from.
-
-What's wrong here?
-
-\begin{itemize}
-\item Oversampling tests. Both Holmengraa and cross.
-\item Mosaic indicating the number of windows needed.
-\item 
-\end{itemize}
-
-% \begin{table}[!b]\centering%\normalsize
-% \begin{tabular}[c]{l r r r@{}  l}\hline
-% \rowcolor{tabBlue} & \multicolumn{1}{>{\columncolor{tabBlue}}c}{\bf B$_\text{arith}$} & \multicolumn{1}{>{\columncolor{tabBlue}}c}{\bf B$_\text{mem}$} & \multicolumn{2}{>{\columncolor{tabBlue}}c}{\bf B$_\text{mem}$/B$_\text{arith}$} \\\hline
-% Arithmetic & 1.03 Tflop/s & & &\\
-% Global memory & & 36 Gfloats/s & \hspace{30pt} 1 &:30 \\
-% Shared memory & & 257 Gfloats/s & 1 &:4 \\
-% Registers & & $>$1.5 Tfloats/s & $>$3 &:2~\cite{Vasilyy}
-% \end{tabular}
-% \caption{Nvidia Quadro 6000: Memory throughput, $B_{\lowercase{\text{mem}}}$, compared to arithmetic throughput, $B_\text{arith}$.}\label{throughputs}
-% \end{table}%
 

### 2015-09-07 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:09:07	14:29:15	(Mon 7.Sep)

@@
 The results will be presented in the following order: In section \ref{lca_window_function} we discuss typical window responses computed by MVDR (\Fig{mvdr_selected_windows}), and hypothesize  that these can be mimicked by Kaiser windows.
-In section \ref{lca_kaiser_parameters} we investigate what sensible boundaries for the Kaiser parameter $\beta$ and window steering $\phi$ should be (\ref{oversampling_mosaic_bounds}).
+In section \ref{lca_kaiser_parameters} we investigate what sensible boundaries for the Kaiser parameter $\beta$ and window steering $\phi$ should be (\Fig{oversampling_mosaic_bounds}).
+In section \ref{lca_database_size} we discuss how many variations of windows between the just determined boundaries should be (\Fig{oversampling_mosaic_bounds_beta}), and determine max window database size.
 
@@
 
-\subsection{Window database size}
+\subsection{Window database size}\label{lca_database_size}
 

### 2015-09-08 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:09:08	11:12:04	(Tue 8.Sep)

@@
 The other scene is of an iron cross of 1\,m length in both directions, connected to a an anchor with a 13\,cm diameter (\Fig{cross}).
-
-MVDR was run with a fairly aggressive yet stable set of parameters; subarray size $L=16$ and $d=1\%$ diagonal loading. Both MVDR and LCA was run with $K=1$ temporal averaging.
+Both scenes is imaged in sectorscan mode.
+
+In the image reconstruction we have run MVDR with a fairly aggressive yet stable set of parameters; subarray size $L=16$ and $d=1\%$ diagonal loading.
+Both MVDR and LCA was run with $K=1$ temporal averaging.
 
@@
 In section \ref{lca_kaiser_parameters} we investigate what sensible boundaries for the Kaiser parameter $\beta$ and window steering $\phi$ should be (\Fig{oversampling_mosaic_bounds}).
-In section \ref{lca_database_size} we discuss how many variations of windows between the just determined boundaries should be (\Fig{oversampling_mosaic_bounds_beta}), and determine max window database size.
-
-\subsection{Window function}\label{lca_window_function}
-
-Assuming that a robust MVDR is the reference method we want LCA to perform similarly to, it seems sensible to create window database for LCA with responses that resembles that of MVDR.
-The typical responses MVDR creates for shadow, speckle and highlight regions of a sectorscan Holmengraa image is shown in \Fig{mvdr_selected_windows}.
+In section \ref{lca_database_size} we discuss how many window variations is needed for LCA to perform well (\Fig{oversampling_mosaic}).
+
+\subsection{LCA window function}\label{lca_window_function}
+
+Assuming that a robust MVDR is the reference method we want LCA to perform similarly to, it seems sensible to create window database for LCA with responses that resembles the ones MVDR computes.
+We study this in \Fig{mvdr_selected_windows}, where the typical responses MVDR creates for shadow, speckle and highlight regions in the Holmengraa scene is presented.
 
@@
 The symmetry is most predominant in shadow and speckle regions, while in highlight the windows are approximately symmetric within the illuminated\todo{valid term?} sector of the seafloor.
-The symmetry is a side-effect of robustifying MVDR to handle coherent interference and the low sample support of covariance statistics.
-
+The symmetry is a side-effect of robustifying MVDR to operate in near snapshot mode in an active system.
 If we look at the peak sidelobe levels of the MVDR windows they are for the most part between -13\,dB and -43\,dB.
 This corresponds to that of the rectangular and Hamming window, or a Kaiser window with $\beta=0$ and $\beta=2$, respectively.
-Using the Kaiser window in this case makes sense since it is optimal in the sense of having its peak power concentration near the steering angle, and because we can also create any response in between.
-The Kaiser windows are easy to compute, is fairly insensitive to coefficient inaccuracies, and can create windows with shapes spanning from rectangular to Gaussian.
+
+We hypothesize that LCA should work even if it only selects from Kaiser windows.
+If they span a decent range of $\beta$-values, and some steering variations is applied to each, we should have responses that resembles those in \Fig{mvdr_selected_windows}.
+The Kaiser window is easy to compute, is fairly insensitive to coefficient inaccuracies, and can create windows with shapes spanning from rectangular to Gaussian.
+
+it is optimal in the sense of having its peak power concentration near the steering angle
 

### 2015-09-08 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:09:08	14:54:44	(Tue 8.Sep)

@@
 
-Assuming that a robust MVDR is the reference method we want LCA to perform similarly to, it seems sensible to create window database for LCA with responses that resembles the ones MVDR computes.
+Assuming that a robust MVDR is the reference method we want LCA to perform similarly to, it seems sensible to create window database for LCA with with responses similar to the ones MVDR computes.
 We study this in \Fig{mvdr_selected_windows}, where the typical responses MVDR creates for shadow, speckle and highlight regions in the Holmengraa scene is presented.
@@
 The symmetry is most predominant in shadow and speckle regions, while in highlight the windows are approximately symmetric within the illuminated\todo{valid term?} sector of the seafloor.
-The symmetry is a side-effect of robustifying MVDR to operate in near snapshot mode in an active system.
-If we look at the peak sidelobe levels of the MVDR windows they are for the most part between -13\,dB and -43\,dB.
+We have observed this symmetry to be a side-effect of running MVDR in near snapshot mode in an active system, due to the steps needed to build the sample covariance matrix.If we look at the peak sidelobe levels of the MVDR windows they are for the most part between -13\,dB and -43\,dB.
 This corresponds to that of the rectangular and Hamming window, or a Kaiser window with $\beta=0$ and $\beta=2$, respectively.
@@
 The Kaiser window is easy to compute, is fairly insensitive to coefficient inaccuracies, and can create windows with shapes spanning from rectangular to Gaussian.
-
-it is optimal in the sense of having its peak power concentration near the steering angle
+It is also optimal in the sense of having its peak power concentration near the steering angle.
 

### 2015-09-08 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:09:08	16:47:18	(Tue 8.Sep)
la2pdf commit - 2015:09:08	17:46:31	(Tue 8.Sep)

@@
 
-% \setcounter{dbltopnumber}{2}
-
 \begin{figure*}[t]%\centering%
@@
 
-\begin{figure*}[t]%\centering%
-\includegraphics[width=\textwidth]{gfx/oversampling_mosaic_bounds_lca_windows_beta.pdf}%
-\caption{$\beta$-values of the windows LCA selects for each image pixel.
+% \setcounter{dbltopnumber}{2}
+
+\begin{figure*}[t]\centering
+\subfloat[$\beta$-values of the Kaiser window chosen for each image pixel.
+Note how LCA prefers a rectangular window ($\beta=0$) on the anchor.]{\includegraphics[width=\textwidth]{gfx/oversampling_mosaic_bounds_lca_windows_beta.pdf}%%
+\label{oversampling_mosaic_bounds_beta}}
+\hfil
+\subfloat[$\beta$-values of the Kaiser window chosen for each image pixel. ]{\includegraphics[width=\textwidth]{gfx/oversampling_mosaic_bounds_lca_windows_phi.pdf}%
+\label{oversampling_mosaic_bounds_phi}}
+\caption{Kaiser windows chosen for each pixel in the image.
+The underlying image is the one shown in \Fig{oversampling_mosaic_bounds}, with the same parameters.
+The location of the anchor, cut line and main scatter locations of the cross is marked in red.}
+\label{fig_sim}
+\end{figure*}
+
+% \begin{figure*}[t]%\centering%
+% \includegraphics[width=\textwidth]{gfx/oversampling_mosaic_bounds_lca_windows_beta.pdf}%
+% \caption{$\beta$-values of the windows LCA selects for each image pixel.
 The underlying image is the one shown in \Fig{oversampling_mosaic_bounds}, with the same parameters.
@@
 Note how LCA prefers a rectangular window ($\beta=0$) on the anchor.}\label{oversampling_mosaic_bounds_beta}
-\end{figure*}
-
-% \setcounter{dbltopnumber}{1}
+% \end{figure*} 
+% 
+% \begin{figure*}[t]%\centering%
+% \includegraphics[width=\textwidth]{gfx/oversampling_mosaic_bounds_lca_windows_phi.pdf}%
+% \caption{$\beta$-values of the windows LCA selects for each image pixel.
+The underlying image is the one shown in \Fig{oversampling_mosaic_bounds}, with the same parameters.
+The location of the anchor and cut line is marked in red.
+Note how LCA prefers a rectangular window ($\beta=0$) on the anchor.}\label{oversampling_mosaic_bounds_beta}
+% \end{figure*}
+
+% % \setcounter{dbltopnumber}{1}
 

### 2015-09-08 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:09:08	18:24:38	(Tue 8.Sep)

@@
 \subfloat[$\beta$-values of the Kaiser window chosen for each image pixel.
-Note how LCA prefers a rectangular window ($\beta=0$) on the anchor.]{\includegraphics[width=\textwidth]{gfx/oversampling_mosaic_bounds_lca_windows_beta.pdf}%%
+Note how LCA prefers a a narrow response ($\beta=0$) on the anchor, and on other strong sources with little lateral interference.
+On the sources where lateral interference is present it chooses the widest response with the best sidelobe suppression.]{\includegraphics[width=\textwidth]{gfx/oversampling_mosaic_bounds_lca_windows_beta.pdf}%%
 \label{oversampling_mosaic_bounds_beta}}
 \hfil
-\subfloat[$\beta$-values of the Kaiser window chosen for each image pixel. ]{\includegraphics[width=\textwidth]{gfx/oversampling_mosaic_bounds_lca_windows_phi.pdf}%
+\subfloat[$\phi$-values of the Kaiser window chosen for each image pixel.
+On strong sources LCA prefers selects windows that are steered away from the source.
+This is what improves the FWHM measurement in \Fig{oversampling_mosaic_bounds}.]{\includegraphics[width=\textwidth]{gfx/oversampling_mosaic_bounds_lca_windows_phi.pdf}%
 \label{oversampling_mosaic_bounds_phi}}
@@
 This non-linear processing increases the image bandwidth introduces a need for a sampling rate higher than the Nyquist rate.
-For MVDR a factor 10 oversampling laterally compared to the Nyquist rate is often needed to ensure  spatialhift-invariance in source amplitude~\cite{Asen2014}.
+For MVDR a factor 10 oversampling laterally compared to the Nyquist rate is often needed to ensure  spatial shift-invariance in source amplitude~\cite{Asen2014}.
 Since LCA operates with the same optimization criterion, we will be using approximately a factor 8 lateral oversampling for both beamformers.
 
-\todo{Need $M$ in the image sector. Why???}
+% \todo{Need $M$ in the image sector. Why???}
 
@@
 To test the performance of the MVDR and LCA beamformers, we have processed data acquired by the Kongsberg Maritime HUGIN AUV carrying their 32 element HISAS1030 sonar~\cite{Hansen2009}.
-It is a high resolution synthetic aperture sonar with array length 1.2\;m, 100\;kHz operating frequency, a bandwidth of 40\;kHz and 25$^\circ$ opening angle.
+It is a high resolution synthetic aperture sonar with 1.2\;m array length, 100\;kHz operating frequency, 40\;kHz bandwidth and 25$^\circ$ opening angle on transmission.
 
@@
 
-In the image reconstruction we have run MVDR with a fairly aggressive yet stable set of parameters; subarray size $L=16$ and $d=1\%$ diagonal loading.
+In the image reconstruction we have run MVDR with subarray length $L=16$ and $d=1\%$ diagonal loading.
 Both MVDR and LCA was run with $K=1$ temporal averaging.
+This is a fairly aggressive yet stable set of parameters.
 
@@
 The symmetry is most predominant in shadow and speckle regions, while in highlight the windows are approximately symmetric within the illuminated\todo{valid term?} sector of the seafloor.
-We have observed this symmetry to be a side-effect of running MVDR in near snapshot mode in an active system, due to the steps needed to build the sample covariance matrix.If we look at the peak sidelobe levels of the MVDR windows they are for the most part between -13\,dB and -43\,dB.
+We have observed this symmetry to be a side-effect of the averaging steps needed to build the sample covariance matrix, which is key to allow  MVDR to run in near snapshot mode in an active system.
+The peak sidelobe levels of the MVDR windows they are mostly between -13\,dB and -43\,dB.
 This corresponds to that of the rectangular and Hamming window, or a Kaiser window with $\beta=0$ and $\beta=2$, respectively.
 
-We hypothesize that LCA should work even if it only selects from Kaiser windows.
+From this we hypothesize that LCA should work even if it only selects from various Kaiser windows.
 If they span a decent range of $\beta$-values, and some steering variations is applied to each, we should have responses that resembles those in \Fig{mvdr_selected_windows}.
@@
 To investigate this we constructed large and equally sized window databases containing windows with varied upper boundaries for $\beta$ and $\phi$.
-As lower boundaries we chose $\beta=0$ and $\phi=\pm0^\circ$, which includes the rectangularly weighted DAS image.
+As lower boundaries we chose $\beta=0$ and $\phi=0^\circ$, which includes the rectangularly weighted DAS image.
 The images produced are shown in \Fig{oversampling_mosaic_bounds}.
-
-We suggest an upper bound for $\beta$ between 2 and 5, where 5 is slightly better at suppressing sidelobes.
-For $\phi$ we suggest to steer either to 50\% or 100\% of the -3dB width, which leads to a resolution of 73\% or 54\% of the rectangular window, respectively.
+The lateral distance $\Delta x$ between the -3dB points of the anchor is measured for each image, with the reference $\Delta x_\text{ref}$ being the rectangularly weighted DAS image.
+This measure is also called the full width half maximum (FWHM).
+In the case of a imaging a point scatter, it is closely related to the resolution of of the system.
+For a rectangular window, for instance, the resolution is 0.89$\cdot$FWHM.
+
+The images show much improved sidelobe suppression with a an upper for bound for $\beta$ of 2, but it improved further as this bound was adjusted up to 5.
+After 5 no noticeable difference was observed.
+The images also demonstrate that increasing the steering $\phi$ improves the resolution of the strong scatterers.
+For this sonar the Rayleigh resolution 0.72$^\circ$, or roughly 24\,cm at the imaging range of 19\,m.
+we suggest to steer either to 50\% or 100\% of the -3dB width, which leads to a resolution of 73\% or 54\% of the rectangular window, respectively.
 As lower boundaries we choose $\beta=0$ to always include a rectangular window, and $\phi=0$ to always include the centered window.

### 2015-09-08 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:09:08	19:05:54	(Tue 8.Sep)

@@
 
-\Fig{oversampling_mosaic_bounds_beta} illustrates which $\beta$-value LCA prefers for different regions in the image of the cross.
-Red marker lines are used to pinpoint the cut line and location of the anchor.
-Observe that in the anchor region only rectangular windows are used, which is the reason for the anchor resolution being measured to the same value regardless of the upper bound of $\beta$ in \Fig{oversampling_mosaic_bounds}.
+\Fig{oversampling_mosaic_bounds_beta} and \Fig{oversampling_mosaic_bounds_phi} illustrates which $\beta$ and $\phi$ values LCA prefers for different regions in the image of the cross.
+Red marker lines are used to pinpoint the cut line, location of the anchor and strong scatter points on the cross.
+Observe that in the anchor region only narrow responses ($\beta=0$) are used, which is the reason for the anchor FWHM being measured to the same value regardless of the upper bound of $\beta$ in \Fig{oversampling_mosaic_bounds}.
+On other scatterers with strong nearby lateral interference LCA prefers wider windows to suppress the interference.
+When allowed LCA always choose windows steered away from the sources, and should not be allowed to select windows that are steered more than the 3\,dB width of the window.
+We observed image oscillations when exceeding this bound.
 
@@
 These have a spatial response with mainlobe width narrower than that of a rectangular window, but at the expense of drastically increased white noise gain.
-We have found that these windows hardly ever get used on experimental data, in particular steered versions which ever worsened white noise gain.
+We have found that these windows hardly ever get used on experimental data, in particular steered versions with an even worse white noise gain.
 We can infer the same from \Fig{mvdr_selected_windows} by noting that the maximum sidelobe level rarely exceeds the rectangular window level of -13\;dB.
@@
 To investigate this we constructed large and equally sized window databases containing windows with varied upper boundaries for $\beta$ and $\phi$.
-As lower boundaries we chose $\beta=0$ and $\phi=0^\circ$, which includes the rectangularly weighted DAS image.
+As lower boundaries we chose $\beta=0$ and $\phi=0^\circ$, which includes the symmetric rectangularly weighted DAS image.
 The images produced are shown in \Fig{oversampling_mosaic_bounds}.
-The lateral distance $\Delta x$ between the -3dB points of the anchor is measured for each image, with the reference $\Delta x_\text{ref}$ being the rectangularly weighted DAS image.
+In each image we computed the lateral distance $\Delta x$ between the -3dB points of the anchor, with the reference $\Delta x_\text{ref}$ being the rectangularly weighted DAS image.
+A forth order polynomial fit was used in this computation.
 This measure is also called the full width half maximum (FWHM).
 In the case of a imaging a point scatter, it is closely related to the resolution of of the system.
-For a rectangular window, for instance, the resolution is 0.89$\cdot$FWHM.
-
-The images show much improved sidelobe suppression with a an upper for bound for $\beta$ of 2, but it improved further as this bound was adjusted up to 5.
-After 5 no noticeable difference was observed.
+For a rectangular window, for instance, the system resolution is $\Delta = \text{FWHM}/0.89$~\cite{Harris1978}.
+
+The images in \Fig{oversampling_mosaic_bounds} much improved sidelobe suppression with a an upper for bound for $\beta$ of 2, but it improved further as this bound was adjusted up to 5.
+No noticeable difference were observed for higher values than 5.
 The images also demonstrate that increasing the steering $\phi$ improves the resolution of the strong scatterers.
-For this sonar the Rayleigh resolution 0.72$^\circ$, or roughly 24\,cm at the imaging range of 19\,m.
-we suggest to steer either to 50\% or 100\% of the -3dB width, which leads to a resolution of 73\% or 54\% of the rectangular window, respectively.
-As lower boundaries we choose $\beta=0$ to always include a rectangular window, and $\phi=0$ to always include the centered window.
-
-LCA images created with a large and equal size window database, but where the upper bounds for $\beta$ and $\phi$ is varied.
-The upper left image is equal to DAS with a rectangular window.
-For all images we measure the lateral distance $\Delta x$ between the two -3dB points of the anchor.
-These are specified relatively to the reference distance  $\Delta x_\text{ref}$ from the upper left DAS image.
+For this sonar the Rayleigh resolution is $\Delta=0.0125$ radians (0.72$^\circ$), thus a point scatterer imaged at 19\,m with a rectangular DAS would have a lateral FWHM of $0.125\cdot19\,\text{m}\cdot0.89=21\,\text{cm}$.
+This is close to the FWHM of the anchor in the DAS image in \Fig{oversampling_mosaic_bounds}.
+As we increase steering to either 50\% or 100\% of the -3dB width, the FWHM drops to 73\% or 53\% to that of the rectangular window, respectively.
 
@@
 \subsection{Computational complexity}
+
+TODO
 

### 2015-09-08 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:09:08	22:03:25	(Tue 8.Sep)

--- 2015-09-08 LCAforActiveSonarImaging/lca.tex
@@
 For MVDR a factor 10 oversampling laterally compared to the Nyquist rate is often needed to ensure  spatial shift-invariance in source amplitude~\cite{Asen2014}.
+
 Since LCA operates with the same optimization criterion, we will be using approximately a factor 8 lateral oversampling for both beamformers.
-
+The images appear more visually pleasing and detailed up to 8, but we consider an absolute minimum to be 2.
+In \mediaII{} we demonstrate this with LCA images changing as a function of a lateral oversampling factor.
+The displayed images are all upinterpolated to the same size, but the detail level varies.
+% 
 % \todo{Need $M$ in the image sector. Why???}
@@
 
-\section{Results and Discussion}
-
-To test the performance of the MVDR and LCA beamformers, we have processed data acquired by the Kongsberg Maritime HUGIN AUV carrying their 32 element HISAS1030 sonar~\cite{Hansen2009}.
+% 
+% 
+% \section{Results and Discussion}\label{results_discussion}
+% 
+% 
+% To test the performance of the MVDR and LCA beamformers, we have processed data acquired by the Kongsberg Maritime HUGIN AUV carrying their 32 element HISAS1030 sonar~\cite{Hansen2009}.
 It is a high resolution synthetic aperture sonar with 1.2\;m array length, 100\;kHz operating frequency, 40\;kHz bandwidth and 25$^\circ$ opening angle on transmission.
-
-Two different scenes will be presented.
+% 
+% Two different scenes will be presented.
 One of the 1500 DWT oil tanker wreck Holmengraa lying at a slanted seabed at 77\;m depth outside of Horten, Norway~\cite{holmengraa}.
@@
 Both scenes is imaged in sectorscan mode.
-
-In the image reconstruction we have run MVDR with subarray length $L=16$ and $d=1\%$ diagonal loading.
+% 
+% In the image reconstruction we have run MVDR with subarray length $L=16$ and $d=1\%$ diagonal loading.
 Both MVDR and LCA was run with $K=1$ temporal averaging.
 This is a fairly aggressive yet stable set of parameters.
-
-The results will be presented in the following order: In section \ref{lca_window_function} we discuss typical window responses computed by MVDR (\Fig{mvdr_selected_windows}), and hypothesize  that these can be mimicked by Kaiser windows.
+% 
+% The results will be presented in the following order: In section \ref{lca_window_function} we discuss typical window responses computed by MVDR (\Fig{mvdr_selected_windows}), and hypothesize  that these can be mimicked by Kaiser windows.
 In section \ref{lca_kaiser_parameters} we investigate what sensible boundaries for the Kaiser parameter $\beta$ and window steering $\phi$ should be (\Fig{oversampling_mosaic_bounds}).
 In section \ref{lca_database_size} we discuss how many window variations is needed for LCA to perform well (\Fig{oversampling_mosaic}).
-
-\subsection{LCA window function}\label{lca_window_function}
-
-Assuming that a robust MVDR is the reference method we want LCA to perform similarly to, it seems sensible to create window database for LCA with with responses similar to the ones MVDR computes.
+% 
+% \subsection{LCA window function}\label{lca_window_function}
+% 
+% Assuming that a robust MVDR is the reference method we want LCA to perform similarly to, it seems sensible to create window database for LCA with with responses similar to the ones MVDR computes.
 We study this in \Fig{mvdr_selected_windows}, where the typical responses MVDR creates for shadow, speckle and highlight regions in the Holmengraa scene is presented.
-
-Observe from \Fig{mvdr_selected_windows} that MVDR seems prefers symmetric window responses, even if it is free to choose non-symmetric ones.
+% 
+% Observe from \Fig{mvdr_selected_windows} that MVDR seems prefers symmetric window responses, even if it is free to choose non-symmetric ones.
 The symmetry is most predominant in shadow and speckle regions, while in highlight the windows are approximately symmetric within the illuminated\todo{valid term?} sector of the seafloor.
@@
 This corresponds to that of the rectangular and Hamming window, or a Kaiser window with $\beta=0$ and $\beta=2$, respectively.
-
-From this we hypothesize that LCA should work even if it only selects from various Kaiser windows.
+% 
+% From this we hypothesize that LCA should work even if it only selects from various Kaiser windows.
 If they span a decent range of $\beta$-values, and some steering variations is applied to each, we should have responses that resembles those in \Fig{mvdr_selected_windows}.
@@
 It is also optimal in the sense of having its peak power concentration near the steering angle.
-
-\Fig{oversampling_mosaic_bounds_beta} illustrates which $\beta$-value LCA prefers for different regions in the image of the cross.
-Red marker lines are used to pinpoint the cut line and location of the anchor.
-Observe that in the anchor region only rectangular windows are used, which is the reason for the anchor resolution being measured to the same value regardless of the upper bound of $\beta$ in \Fig{oversampling_mosaic_bounds}.
-
-Unlike the method described by Synnev\aa{}g~\cite{Synnevag2008} we do not let LCA choose from  inverted Kaiser windows.
+% 
+% Unlike the method described by Synnev\aa{}g~\cite{Synnevag2008} we do not let LCA choose from  inverted Kaiser windows.
 These have a spatial response with mainlobe width narrower than that of a rectangular window, but at the expense of drastically increased white noise gain.
-We have found that these windows hardly ever get used on experimental data, in particular steered versions which ever worsened white noise gain.
+We have found that these windows hardly ever get used on experimental data, in particular steered versions with an even worse white noise gain.
 We can infer the same from \Fig{mvdr_selected_windows} by noting that the maximum sidelobe level rarely exceeds the rectangular window level of -13\;dB.
-
-\mediaI{} shows how \Fig{mvdr_selected_windows} changes as a function of the MVDR subarray length $L$ and temporal averaging $K$.
+% 
+% \mediaI{} shows how \Fig{mvdr_selected_windows} changes as a function of the MVDR subarray length $L$ and temporal averaging $K$.
 It demonstrates that at $L=1$ the window responses are rectangular in all areas in the image.
@@
 The media file also shows MVDR can be run with subarray sizes $L\in[M/2,5M/8]$, but only with temporal averaging $K=1$ or above.
-
-\subsection{Kaiser parameter $\beta$ and steering $\phi$}\label{lca_kaiser_parameters}
-
-What is the best set of Kaiser windows to use with LCA, and how much should they be steered?
+% 
+% 
+% 
+% \subsection{Kaiser parameter $\beta$ and steering $\phi$}\label{lca_kaiser_parameters}
+% 
+% What is the best set of Kaiser windows to use with LCA, and how much should they be steered?
 To investigate this we constructed large and equally sized window databases containing windows with varied upper boundaries for $\beta$ and $\phi$.
-As lower boundaries we chose $\beta=0$ and $\phi=0^\circ$, which includes the rectangularly weighted DAS image.
+As lower boundaries we chose $\beta=0$ and $\phi=0^\circ$, which includes the symmetric rectangularly weighted DAS image.
 The images produced are shown in \Fig{oversampling_mosaic_bounds}.
-The lateral distance $\Delta x$ between the -3dB points of the anchor is measured for each image, with the reference $\Delta x_\text{ref}$ being the rectangularly weighted DAS image.
+In each image we computed the lateral distance $\Delta x$ between the -3dB points of the anchor, with the reference $\Delta x_\text{ref}$ being the rectangularly weighted DAS image.
+A forth order polynomial fit was used in this computation.
 This measure is also called the full width half maximum (FWHM).
 In the case of a imaging a point scatter, it is closely related to the resolution of of the system.
-For a rectangular window, for instance, the resolution is 0.89$\cdot$FWHM.
-
-The images show much improved sidelobe suppression with a an upper for bound for $\beta$ of 2, but it improved further as this bound was adjusted up to 5.
-After 5 no noticeable difference was observed.
+For a rectangular window, for instance, the system resolution is $\Delta = \text{FWHM}/0.89$~\cite{Harris1978}.
+% 
+% The images in \Fig{oversampling_mosaic_bounds} much improved sidelobe suppression with a an upper for bound for $\beta$ of 2, but it improved further as this bound was adjusted up to 5.
+No noticeable difference were observed for higher values than 5.
 The images also demonstrate that increasing the steering $\phi$ improves the resolution of the strong scatterers.
-For this sonar the Rayleigh resolution 0.72$^\circ$, or roughly 24\,cm at the imaging range of 19\,m.
-we suggest to steer either to 50\% or 100\% of the -3dB width, which leads to a resolution of 73\% or 54\% of the rectangular window, respectively.
-As lower boundaries we choose $\beta=0$ to always include a rectangular window, and $\phi=0$ to always include the centered window.
-
-LCA images created with a large and equal size window database, but where the upper bounds for $\beta$ and $\phi$ is varied.
-The upper left image is equal to DAS with a rectangular window.
-For all images we measure the lateral distance $\Delta x$ between the two -3dB points of the anchor.
-These are specified relatively to the reference distance  $\Delta x_\text{ref}$ from the upper left DAS image.
-
-\begin{itemize}
-\item More steering -> more aggressive suppression of interference at the cost of boosted sidelobes / noise.
-\item More steering -> Larger variations in amplitude of sources
-\end{itemize}
-
-\subsection{Window database size}\label{lca_database_size}
-
-So far we have determined it reasonable to use Kaiser windows in the range $\beta\in[0,5]$, steered within the range $|\phi|\in[0,\phi_\mathrm{3dB}]$. But many windows do we need in these ranges?
-
-In \Fig{oversampling_mosaic} we have compared LCA images made from windows databases of different sizes, but with the same parameter boundaries.
+For this sonar the Rayleigh resolution is $\Delta=0.0125$ radians (0.72$^\circ$), thus a point scatterer imaged at 19\,m with a rectangular DAS would have a lateral FWHM of $0.125\cdot19\,\text{m}\cdot0.89=21\,\text{cm}$.
+This is close to the FWHM of the anchor in the DAS image in \Fig{oversampling_mosaic_bounds}.
+As we increase steering to either 50\% or 100\% of the -3dB width, the FWHM drops to 73\% or 53\% to that of the rectangular window, respectively.
+% 
+% \Fig{oversampling_mosaic_bounds_beta} and \Fig{oversampling_mosaic_bounds_phi} illustrates which $\beta$ and $\phi$ values LCA prefers for different regions in the image of the cross.
+Red marker lines are used to pinpoint the cut line, location of the anchor and strong scatter points on the cross.
+Observe that in the anchor region only narrow responses ($\beta=0$) are used, which is the reason for the anchor FWHM being measured to the same value regardless of the upper bound of $\beta$ in \Fig{oversampling_mosaic_bounds}.
+On other scatterers with strong nearby lateral interference LCA prefers wider windows to suppress the interference.
+When allowed LCA always choose windows steered away from the sources, and should not be allowed to select windows that are steered more than the 3\,dB width of the window.
+We observed image oscillations when exceeding this bound.
+% 
+% 
+% \subsection{Window database size}\label{lca_database_size}
+% 
+% So far we have determined it reasonable to use Kaiser windows in the range $\beta\in[0,5]$, steered within the range $|\phi|\in[0,\phi_\mathrm{3dB}]$.
+But how many windows do we need in these ranges?
+% 
+% In \Fig{oversampling_mosaic} we have compared LCA images made from windows databases of different sizes, but with the same parameter boundaries.
 We included the results from using a single window ($N_\beta=1$) and steering angle ($N_\phi=1$), which corresponds to a rectangularly weighted DAS.
-
-Observe that sampling the $\beta$-range more finely increases the sidelobe suppression but leaves resolution unchanged.
+% 
+% Observe that sampling the $\beta$-range more finely increases the sidelobe suppression but leaves resolution unchanged.
 Adding more steering-variations improves both sidelobe suppression and resolution.
-
-However, using more than  $N_\beta=2$ window types and $N_\beta=3$ steering angles makes minimal
-
-\ref{oversampling_mosaic}
-
-\subsection{Oversampling}
-
-From our observations one should have at least twice the number of beams as channels. Image is more visually pleasing and appears more detailed up to $8M$.
-
-LCA is largely insensitive to the choice of windows and steering. 
-LCA works well for all the boundaries
-Variables of interest:
-
-\begin{itemize}
-\item Mainlobe width. Wide in shadows, narrow in detailed regions. Adding windows in between these extremal values.
-\item Steering. No steering in shadows, slight steering in speckle and no steering in highlight. Steering more than the -3dB width of the window function has no effect.
-\item Oversampling. From our observations one should have at least twice the number of beams as channels. Image is more visually pleasing and appears more detailed up to $8M$.
-\end{itemize}
-
-\subsection{Computational complexity}
-
-% Furthermore, the algorithm maps well to single-instruction-multiple-data (SIMD) hardware such as graphics processing units (GPUs).
+We can observe a major improvement going to $N_\beta=2$ and $N_\phi=3$, but minimal improvement by adding yet more windows.
+% 
+% 
+% 
+% LCA is largely insensitive to the choice of windows and steering. 
+% LCA works well for all the boundaries
+% Variables of interest:
+% 
+% \begin{itemize}
+% \item Mainlobe width. Wide in shadows, narrow in detailed regions. Adding windows in between these extremal values.
+% \item Steering. No steering in shadows, slight steering in speckle and no steering in highlight. Steering more than the -3dB width of the window function has no effect.
+% \item Oversampling. From our observations one should have at least twice the number of beams as channels. Image is more visually pleasing and appears more detailed up to $8M$.
+% \end{itemize}
+% 
+% \subsection{Computational complexity}
+% 
+% The computational complexity of MVDR is generally considered to be of O($M^3$).
+However, the implementation we are using is optimized so that building the spatial covariance matrix is of O($N_kN_L,L^2$), and inverting it is of O($L^3$).
+We describe this implementation in \cite{Buskenes2014}, and a beamspace version in \cite{Asen2013}, where we optimize if for a graphics processing unit (GPU) for nearly two orders of magnitude speed increase.
+However, MVDR is not particularly well suited for this due to the complex data dependencies of the covariance computation, and at best we only managed to utilize 10\% of the GPUs potential.
+% 
+% In comparison, LCA is of O($M N_w$), and extremely well suited for GPUs.
+% 
+% % Furthermore, the algorithm maps well to single-instruction-multiple-data (SIMD) hardware such as graphics processing units (GPUs).
 We show that even a modest amount of optimisation work has resulted in a factor 10 speed-up of this algorithm when implemented on a GPU as opposed to on a CPU.
-
-% the performance of the LCA beamformer is surprisingly similar regardless of what windows it is allowed to choose from.
-
-\subsection{Summary}
-
-\begin{itemize}
-\item \emph{Window type}.
+% 
+% 
+% % the performance of the LCA beamformer is surprisingly similar regardless of what windows it is allowed to choose from.
+% 
+% 
+% \subsection{Summary}
+% 
+% \begin{itemize}
+% \item \emph{Window type}.
 \Fig{mvdr_selected_windows} show frequency responses of typical windows chosen by MVDR in shadow, speckle and highlight regions in an image.
@@
 We use Kaiser windows to mimic these responses.
-\item \emph{Window parameter boundaries}.
+% \item \emph{Window parameter boundaries}.
 \Fig{oversampling_mosaic_bounds} show LCA images made with the same sized window database, but containing windows with varied upper boundaries for $\beta$ and $\phi$.
@@
 As lower boundaries we choose $\beta=0$ to always include a rectangular window, and $\phi=0$ to always include the centered window.
-\item \emph{Window database size}. \Fig{oversampling_mosaic}
-\item \emph{Performance}.
-\item \emph{Oversampling need}.
-\end{itemize}
-
-% \newcommand\cc[1]{\multicolumn{1}{>{\columncolor{tabBlue}}c}{\bf #1}}
-% \begin{table}[!b]\centering%\normalsize
-% \begin{tabular}[c]{l r r r@{}  r r r}\hline
-% \rowcolor{tabBlue} & \cc{Resolution} & \cc{S} & \cc{$\beta$} & \cc{$\varphi^*$} &  \cc{N$_\beta$} & \cc{N$_\varphi$} \\\hline
-% & 50\%  & 28 & [0,3] & \hspace{.15cm} 50\% & 2 & 3 \\
-% & 50\%  & 28 & [0,6] & \hspace{.15cm} 50\% & 2 & 3 \\
-% & 50\%  & 28 & [0,9] & \hspace{.15cm} 50\% & 2 & 3 \\
-% & 100\% & 32 &  &:30 \\
-% & 150\% & 32 & 1 &:4\\
-% \end{tabular}\\
-% * Steering is relative to the -3dB width of the window being applied.
-% \caption{Qualitative study}\label{parameter_study}
-% \end{table}%
-
-% The simulations contained a few highlighted cylindric regions with 2\,m diameter and a constant intensity 15.4\,dB over the average speckle level.
+% \item \emph{Window database size}. \Fig{oversampling_mosaic}
+% \item \emph{Performance}.
+% \item \emph{Oversampling need}.
+% 
+% 
+% % \end{itemize}
+% 
+% 
+% % \newcommand\cc[1]{\multicolumn{1}{>{\columncolor{tabBlue}}c}{\bf #1}}
+% % \begin{table}[!b]\centering%\normalsize
+% % \begin{tabular}[c]{l r r r@{}  r r r}\hline
+% % \rowcolor{tabBlue} & \cc{Resolution} & \cc{S} & \cc{$\beta$} & \cc{$\varphi^*$} &  \cc{N$_\beta$} & \cc{N$_\varphi$} \\\hline
+% % & 50\%  & 28 & [0,3] & \hspace{.15cm} 50\% & 2 & 3 \\
+% % & 50\%  & 28 & [0,6] & \hspace{.15cm} 50\% & 2 & 3 \\
+% % & 50\%  & 28 & [0,9] & \hspace{.15cm} 50\% & 2 & 3 \\
+% % & 100\% & 32 &  &:30 \\
+% % & 150\% & 32 & 1 &:4\\
+% % \end{tabular}\\
+% % * Steering is relative to the -3dB width of the window being applied.
+% % \caption{Qualitative study}\label{parameter_study}
+% % \end{table}%
+% 
+% 
+% % The simulations contained a few highlighted cylindric regions with 2\,m diameter and a constant intensity 15.4\,dB over the average speckle level.
 We focused on an object centered at 41\,m range and -3 degrees azimuth angle.
@@
 %\ref{saf09}
-% 
-% In \Fig{speckle} we observe that both the adaptive beamformers produce images with a more clearly defined shadow.
+% % 
+% % In \Fig{speckle} we observe that both the adaptive beamformers produce images with a more clearly defined shadow.
 The edges are sharper and the shadow more distinct.
@@
 The same effects can also be observed laterally around the highlight region of the mean image; DAS is more prone to smear energy into the speckle region than the adaptive beamformers are.
-% 
-% \Fig{cutmean} displays two lateral cuts in the mean image, one through the highlight at 41.1\,m range and one through the shadow at 44.5\,m range.
+% % 
+% % \Fig{cutmean} displays two lateral cuts in the mean image, one through the highlight at 41.1\,m range and one through the shadow at 44.5\,m range.
 Each cut was computed as the mean of a 40\,cm range band centered at the respective ranges.
@@
 The slightly inferior performance of the MV beamformer is due to the subarray averaging, which makes the effective array smaller~\cite{syn07}.
-% 
-% By studying the LCA performance on the 100 speckle realisations, we found the choice of window to be highly dependent on the phenomena being imaged.
+% % 
+% % By studying the LCA performance on the 100 speckle realisations, we found the choice of window to be highly dependent on the phenomena being imaged.
 In speckle regions LCA favored narrow spatial responses with minor steering, while wide and fully steered responses were often selected in shadow regions.
 There was also a clear distinction between the responses selected in sidelobe regions and in speckle regions.
-% 
-% % \Fig{hist} illustrates how often a particular window is chosen in general.
+% % 
+% % % \Fig{hist} illustrates how often a particular window is chosen in general.
 There are 10 $\phi$'s for each of the 10 $\beta$'s.
@@
 At value 65 the $\beta$ value is 2.025, and these and the remaining windows are hardly selected.
-% 
-% A sidescan image of the 1500 dwt oil tanker wreck Holmengraa is shown in \Fig{holmengraa}.
+% % 
+% % A sidescan image of the 1500 dwt oil tanker wreck Holmengraa is shown in \Fig{holmengraa}.
 It is about 68\,m long and 9\,m wide, and lies on a slanted seabed at 77\,m depth~\cite{holmengraa}.
@@
 This, and the fact that the wreck was imaged at a range of about 105\,m makes the image quality poor, but sufficient to compare our beamformers.
-% 
-% In the Holmengraa image we note again that the LCA produces a cleaner shadow and better edge definition. The MV method performed almost identically to LCA in this case, and was omitted.
-
-\section{Conclusion}
-
-LCA can be viewed as either a simplification of MVDR, or as an extension to DAS.
+% % 
+% % In the Holmengraa image we note again that the LCA produces a cleaner shadow and better edge definition. The MV method performed almost identically to LCA in this case, and was omitted.
+% 
+% 
+% \section{Conclusion}
+% 
+% LCA can be viewed as either a simplification of MVDR, or as an extension to DAS.
 It creates images similar to MVDR, but at a fraction of the computational cost.
 It is also inherently robust and is largely insensitive to the exact set of windows it is run with, as long as they are sensible.
-
-Using experimental data we have demonstrated that when MVDR is made robust we constrain the solution space of the window coefficients it can choose from into mostly symmetric responses that varies between the rectangular and Hamming.
+% 
+% Using experimental data we have demonstrated that when MVDR is made robust we constrain the solution space of the window coefficients it can choose from into mostly symmetric responses that varies between the rectangular and Hamming.
 We construct a window database using the parametric Kaiser function, which easily allows us to create responses similar to these and those in between.
 The Kaiser window is optimal in the sense of maximum power near the direction of interest.
-
-% We limit our study to the use of variations of the parametric Kaiser-Harris window, since we observed neglible performance difference to using this function or any of the common trigonometric ones.
+% 
+% % We limit our study to the use of variations of the parametric Kaiser-Harris window, since we observed neglible performance difference to using this function or any of the common trigonometric ones.
 The Kaiser windows are easy to compute, is fairly insensitive to coefficient inaccuracies, and can create windows with shapes spanning from rectangular to Gaussian.
-% 
-% 
-% Using the LCA beamformer results in increased robustness and improved edge definition compared to the MV and DAS beamformers.
-
-% The images produced by LCA appears sharper than those of MVDR, but 
-% 
-% The key to achieve this success lies in the design of the window set.
+% % 
+% % 
+% % Using the LCA beamformer results in increased robustness and improved edge definition compared to the MV and DAS beamformers.
+% 
+% 
+% % The images produced by LCA appears sharper than those of MVDR, but 
+% % 
+% % The key to achieve this success lies in the design of the window set.
 Windows that yield vagely steered narrow responses are preferred in speckle regions, while wide and steered responses are typically preferred in highlight and shadow regions.
-% 
-% A window set of 6 different responses each steered in 5 different directions proved sufficient in our experiment to match the performance of the MV beamformer.
-
-The tendency of the LCA to perform similarly to the MV beamformer seems to indicate that a full solution space is rarely needed in real world scenarios.
-
-\begin{table}[!b]\centering%\normalsize
-\begin{tabular}[c]{l c c c}\hline
-\rowcolor{tabBlue} & \bf Highlight & \bf Shadow & \bf Speckle  \\\hline
-Symmetry             & Moderate  & High   & High \\
-Mainlobe width       & Narrow    & Wide   & Moderate \\
-Sidelobe suppression & Low       & High   & Moderate \\
-Steering             & Low       & Up to -3dB width & Moderate \\
-Oversampling         & 2x-8x     & 2x-8x  & 2x-8x
-\end{tabular}
-\caption{Summary}\label{tab:summary}
-\end{table}%
+% % 
+% % A window set of 6 different responses each steered in 5 different directions proved sufficient in our experiment to match the performance of the MV beamformer.
+% 
+% The tendency of the LCA to perform similarly to the MV beamformer seems to indicate that a full solution space is rarely needed in real world scenarios. 
+% 
+% \begin{table}[!b]\centering%\normalsize
+% \begin{tabular}[c]{l c c c}\hline
+% \rowcolor{tabBlue} & \bf Highlight & \bf Shadow & \bf Speckle  \\\hline
+% Symmetry             & Moderate  & High   & High \\
+% Mainlobe width       & Narrow    & Wide   & Moderate \\
+% Sidelobe suppression & Low       & High   & Moderate \\
+% Steering             & Low       & Up to -3dB width & Moderate \\
+% Oversampling         & 2x-8x     & 2x-8x  & 2x-8x
+% \end{tabular}
+% \caption{Summary}\label{tab:summary}
+% \end{table}%
 

### 2015-09-08 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:09:08	22:34:55	(Tue 8.Sep)

--- 2015-09-08 LCAforActiveSonarImaging/lca.tex
@@
 1 \\
-e^{-j\frac{md}{\lambda}\sin(\phi)} \\
+e^{-j\frac{2\pi d}{\lambda}\sin(\phi)} \\
 \vdots\\
-e^{-j\frac{(M-1)d}{\lambda}\sin(\phi)}
-}.
+e^{-j\frac{2\pi (M-1)d}{\lambda}\sin(\phi)}
+}.\label{steering_vector}
 \end{align}
@@
 
-\section{Results and Discussion}
+\section{Results and Discussion}\label{results_discussion}
 
@@
 It is also optimal in the sense of having its peak power concentration near the steering angle.
-
-\Fig{oversampling_mosaic_bounds_beta} and \Fig{oversampling_mosaic_bounds_phi} illustrates which $\beta$ and $\phi$ values LCA prefers for different regions in the image of the cross.
-Red marker lines are used to pinpoint the cut line, location of the anchor and strong scatter points on the cross.
-Observe that in the anchor region only narrow responses ($\beta=0$) are used, which is the reason for the anchor FWHM being measured to the same value regardless of the upper bound of $\beta$ in \Fig{oversampling_mosaic_bounds}.
-On other scatterers with strong nearby lateral interference LCA prefers wider windows to suppress the interference.
-When allowed LCA always choose windows steered away from the sources, and should not be allowed to select windows that are steered more than the 3\,dB width of the window.
-We observed image oscillations when exceeding this bound.
 
@@
 
-\begin{itemize}
-\item More steering -> more aggressive suppression of interference at the cost of boosted sidelobes / noise.
-\item More steering -> Larger variations in amplitude of sources
-\end{itemize}
+\Fig{oversampling_mosaic_bounds_beta} and \Fig{oversampling_mosaic_bounds_phi} illustrates which $\beta$ and $\phi$ values LCA prefers for different regions in the image of the cross.
+Red marker lines are used to pinpoint the cut line, location of the anchor and strong scatter points on the cross.
+Observe that in the anchor region only narrow responses ($\beta=0$) are used, which is the reason for the anchor FWHM being measured to the same value regardless of the upper bound of $\beta$ in \Fig{oversampling_mosaic_bounds}.
+On other scatterers with strong nearby lateral interference LCA prefers wider windows to suppress the interference.
+When allowed LCA always choose windows steered away from the sources, and should not be allowed to select windows that are steered more than the 3\,dB width of the window.
+We observed image oscillations when exceeding this bound.
 
@@
 
-So far we have determined it reasonable to use Kaiser windows in the range $\beta\in[0,5]$, steered within the range $|\phi|\in[0,\phi_\mathrm{3dB}]$. But many windows do we need in these ranges?
+So far we have determined it reasonable to use Kaiser windows in the range $\beta\in[0,5]$, steered within the range $|\phi|\in[0,\phi_\mathrm{3dB}]$. But how many windows do we need in these ranges?
 
@@
 Adding more steering-variations improves both sidelobe suppression and resolution.
-
-However, using more than  $N_\beta=2$ window types and $N_\beta=3$ steering angles makes minimal
-
-\ref{oversampling_mosaic}
-
-\subsection{Oversampling}
-
-From our observations one should have at least twice the number of beams as channels. Image is more visually pleasing and appears more detailed up to $8M$.
+We can observe a major improvement going to $N_\beta=2$ and $N_\phi=3$, but minimal improvement by adding yet more windows.
 
@@
 
-TODO
+The computational complexity of MVDR is generally considered to be of O($M^3$).
+However, the implementation we are using is optimized so that building the spatial covariance matrix is of O($N_kN_L,L^2$), and inverting it is of O($L^3$).
+We describe this implementation in \cite{Buskenes2014}, and a beamspace version in \cite{Asen2013}, where we optimize if for a graphics processing unit (GPU) for nearly two orders of magnitude speed increase compared to a straightforward C implementation.
+However, MVDR is not ideal for GPUs due to the complex data dependencies present in the covariance computation, and at best we only managed to utilize 10\% of the GPUs potential.
+
+In comparison, LCA is of O($M N_w$), and extremely well suited for GPUs.
+Each pixel depends on only a small subset of data, and the windows can be precomputed and stored in GPU cache for near immediate access.
+Furthermore, since we have observed LCA to perform similarly using other parametric window variants, the complexity can be reduced to near DAS.
 
@@
 
+\section{Reducing LCA complexity}
+
+\renewcommand\b{\vec b}
+
+Now let us choose a trigonometric window function:
+%
+\begin{align}
+\w = e^{j\phi_\theta[m]}\big(\alpha - (1-\alpha)\cos\left(\frac{2\pi m}{M-1}\right)\big)
+\end{align}
+%
+where $\phi_\theta[m] = \frac{2\pi}{\lambda}dm\sin\theta$. Inserting this yields:
+%
+\begin{align}
+\w\H\x &= \sumb{m=0}{M-1} e^{j\phi_\theta[m]}\Big(\alpha - (1-\alpha)\cos\left(\frac{2\pi m}{M-1}\right)\Big) x_m^* \nn
+&= \alpha\sumb{m=0}{M-1} e^{j\phi_\theta[m]} x_m^* - (1-\alpha)\sumb{m=0}{M-1} e^{j\varphi[m]}\cos\left(\frac{2\pi m}{M-1}\right) x_m^* \nn
+&= \alpha \a\T\x + (1-\alpha)\b\T\x
+\end{align}
+%
+where $\a_\phi$ is the steering vector defined in (\ref{steering_vector}), and 
+%
+\begin{align}
+\a = \begin{bmatrix}
+     1 \\
+     e^{j\frac{2\pi}{\lambda}d\sin\theta} \\
+     e^{j\frac{2\pi}{\lambda}2d\sin\theta} \\
+     \vdots \\
+     e^{j\frac{2\pi}{\lambda}d(M-1)\sin\theta}
+     \end{bmatrix} \qquad \text{and} \qquad
+\b = \begin{bmatrix}
+     0 \\
+     e^{j\frac{2\pi}{\lambda}d\sin\theta}\sin\left(\frac{2\pi m}{M-1}\right) \\
+     e^{j\frac{2\pi}{\lambda}2d\sin\theta}\sin\left(\frac{2\pi m}{M-1}\right) \\
+     \vdots \\
+     0
+     \end{bmatrix}\a
+\end{align}
+%
+To attain unity gain in the look direction the weights must sum to one. 
+%
+\begin{align}
+\w\T\1 &= \big(\alpha\a - (1-\alpha)\b\big)\T\1 = 1
+\end{align}
+%
+This is true for any value of $\alpha$ if $\a\T\1 = 1$ and $\b\T\1 = 1$.
+
+where
+%
+\begin{align}
+|\w\H\x|^2 &= \Big| \alpha a + (1-\alpha)b \Big|^2 \nn
+&= \Big[\alpha a + (1-\alpha)b\Big] \Big[\alpha a^* + (1-\alpha)b^*\Big] \nn
+&= \alpha^2 aa^* + \alpha a(1-\alpha)b^* + \alpha a^*(1-\alpha)b + (1-\alpha)^2bb^* \nn
+&= \alpha^2 aa^* + \alpha(1-\alpha)(ab^* + a^*b) + (1-\alpha)^2bb^* \nn
+&= \alpha^2 (aa^* - (ab^* + a^*b) - bb^*) + \alpha(ab^* + a^*b) + bb^* \nn
+\end{align}
+%
+\begin{align}
+\frac{\partial}{\partial\alpha} |\w\H\x|^2 
+&= 2\alpha (aa^* - (ab^* + a^*b) - bb^*) + ab^* + a^*b = 0
+\end{align}
+%
+\begin{align}
+\alpha &= \frac{ab^* + a^*b}{2(aa^* - (ab^* + a^*b) - bb^*)}
+\end{align}
+
+LCA can then be defined as
+%
+\begin{align}
+\argmin{\w}\ |z|^2 = \argmin{\w}\ z\,z^* = %\argmin{\w}\ \w\H\x = \sumb{m=0}{M-1} w_m^* x_m
+\end{align}
+
 % use section* for acknowledgement

### 2015-09-14 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:09:14	14:11:08	(Mon14.Sep)

@@
 
-LCA is largely insensitive to the choice of windows and steering. 
-LCA works well for all the boundaries
-Variables of interest:
-
-\begin{itemize}
-\item Mainlobe width. Wide in shadows, narrow in detailed regions. Adding windows in between these extremal values.
-\item Steering. No steering in shadows, slight steering in speckle and no steering in highlight. Steering more than the -3dB width of the window function has no effect.
-\item Oversampling. From our observations one should have at least twice the number of beams as channels. Image is more visually pleasing and appears more detailed up to $8M$.
-\end{itemize}
+\subsection{Image quality}
+
+TODO. Add image of DAS vs. MVDR vs. LCA Kaiser vs. LCA Trig
 
@@
 The Kaiser window is optimal in the sense of maximum power near the direction of interest.
+
+% 
+% LCA is largely insensitive to the choice of windows and steering. 
+% LCA works well for all the boundaries
+% Variables of interest:
 
@@
 %
-where $\phi_\theta[m] = \frac{2\pi}{\lambda}dm\sin\theta$. Inserting this yields:
+where $\phi_\theta[m] = \frac{2\pi}{\lambda}dm\sin\theta$ is the steering angle. Inserting this into the beamformer equation yields:
 %
@@
 %
-where $\a_\phi$ is the steering vector defined in (\ref{steering_vector}), and 
-%
-\begin{align}
-\a = \begin{bmatrix}
-     1 \\
-     e^{j\frac{2\pi}{\lambda}d\sin\theta} \\
-     e^{j\frac{2\pi}{\lambda}2d\sin\theta} \\
-     \vdots \\
-     e^{j\frac{2\pi}{\lambda}d(M-1)\sin\theta}
-     \end{bmatrix} \qquad \text{and} \qquad
-\b = \begin{bmatrix}
-     0 \\
-     e^{j\frac{2\pi}{\lambda}d\sin\theta}\sin\left(\frac{2\pi m}{M-1}\right) \\
-     e^{j\frac{2\pi}{\lambda}2d\sin\theta}\sin\left(\frac{2\pi m}{M-1}\right) \\
-     \vdots \\
-     0
-     \end{bmatrix}\a
+where $\a$ is the steering vector defined in (\ref{steering_vector}), and 
+%
+\begin{align}
+\b = \text{diag}(\a)\begin{bmatrix}
+     1 &
+     \cos\left(\frac{2\pi}{M-1}\right) &
+     \cos\left(\frac{2\pi 2}{M-1}\right) &
+     \hdots &
+     1
+     \end{bmatrix}\T.
 \end{align}
@@
 %
-This is true for any value of $\alpha$ if $\a\T\1 = 1$ and $\b\T\1 = 1$.
-
-where
+This is true for any value of $\alpha$ if $\a\T\1 = 1$ and $\b\T\1 = 1$. Hence, as long as we normalize $\a$ and $\b$ the unity gain constraint is preserved.
+
+Now let $a=\a\T x$ and $b=\b\T x$. The beamformer output can then be written as:
 %
@@
 %
+Differentiating this with respect to $\alpha$ yields:
+%
 \begin{align}
@@
 %
+which has the solution:
+%
 \begin{align}
@@
 \end{align}
-
-LCA can then be defined as
-%
-\begin{align}
-\argmin{\w}\ |z|^2 = \argmin{\w}\ z\,z^* = %\argmin{\w}\ \w\H\x = \sumb{m=0}{M-1} w_m^* x_m
-\end{align}
+%
+In this computation there are only 4 unique additions and multiplications.
+If we used this to analytically solve for $\alpha$, but perform the search for $\phi$, the LCA's computational complexity would be of O($8MN_\phi)$ instead of O($MN_\alpha N_\phi)$.
+The value for $\alpha$ would also be the optimal.
 

### 2015-09-16 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:09:16	22:56:55	(Wed16.Sep)

@@
 \author{Jo~Inge~Buskenes, %
-        Roy~Edgar~Hansen,%
+        Roy~Edgar~Hansen, %
         Andreas~Austeng%
@@
 This makes MVDR impracticable and sometimes infeasible.
-We have shown in previous studies that MVDR's complexity can be reduced, as well as accelerated significantly using graphics computing units (GPUs)~\cite{Buskenes2014,Asen2013}, but there is a way to avoid it altogether while achieving similar performance.
+We have shown in previous studies that MVDR's complexity can be reduced, as well as accelerated significantly using graphics computing units (GPUs)~\cite{Buskenes2014,Asen2013}, but there is a way to avoid using MVDR altogether while achieving similar performance.
 
@@
 
-In this work we aim to provide some intuitive understanding of how the window database may be designed.
+In this work apply the LCA method to active sonar imaging, and aim to provide some intuitive understanding of how the window database may be designed.
 Like Synnev\aa{}g et.
@@
 \subfloat[$\phi$-values of the Kaiser window chosen for each image pixel.
-On strong sources LCA prefers selects windows that are steered away from the source.
+On strong sources LCA selects windows that are steered away from the source.
 This is what improves the FWHM measurement in \Fig{oversampling_mosaic_bounds}.]{\includegraphics[width=\textwidth]{gfx/oversampling_mosaic_bounds_lca_windows_phi.pdf}%
@@
 \label{fig_sim}
+\end{figure*}
+
+\begin{figure*}[t]\centering
+\subfloat[]{\includegraphics[width=\textwidth]{gfx/beamformer_comparison_holmengraa.pdf}%%
+\label{oversampling_mosaic_bounds_beta}}
+\hfil
+\subfloat[]{\includegraphics[width=\textwidth]{gfx/beamformer_comparison_cross.pdf}%
+\label{oversampling_mosaic_bounds_phi}}
+\caption{}\label{fig_sim}
 \end{figure*}
@@
 A forth order polynomial fit was used in this computation.
-This measure is also called the full width half maximum (FWHM).
+This measure is also commonly called the full width half maximum (FWHM).
 In the case of a imaging a point scatter, it is closely related to the resolution of of the system.
-For a rectangular window, for instance, the system resolution is $\Delta = \text{FWHM}/0.89$~\cite{Harris1978}.
+For a rectangular window, for instance, the system resolution is approximately $\Delta \approx \text{FWHM}/0.89$~\cite{Harris1978}.
 

### 2015-09-16 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:09:16	23:29:32	(Wed16.Sep)

@@
 The ultimate goal of adaptive beamformers is to take the information available in the wavefield and use it to either improve image resolution, noise suppression, or both.
-This non-linear processing increases the image bandwidth introduces a need for a sampling rate higher than the Nyquist rate.
-For MVDR a factor 10 oversampling laterally compared to the Nyquist rate is often needed to ensure  spatial shift-invariance in source amplitude~\cite{Asen2014}.
+This non-linear processing increases the image bandwidth and introduces a need for a sampling rate higher than the Nyquist rate.
+For MVDR a lateral oversampling factor of 10 compared to the Nyquist rate is often needed to ensure minimal spatial shift-variance in source amplitude~\cite{Asen2014}.
 
@@
 
-Now let us choose a trigonometric window function:
-%
-\begin{align}
-\w = e^{j\phi_\theta[m]}\big(\alpha - (1-\alpha)\cos\left(\frac{2\pi m}{M-1}\right)\big)
-\end{align}
-%
-where $\phi_\theta[m] = \frac{2\pi}{\lambda}dm\sin\theta$ is the steering angle. Inserting this into the beamformer equation yields:
+Throughout this article we have used LCA with variations of the Kaiser window function. However, we could instead have used a trigonometric window function:
+%
+\begin{align}
+\w = e^{j\phi_\theta[m]}\big(\alpha - (1-\alpha)\cos\left(\frac{2\pi m}{M-1}\right)\big)\label{trig_window_function}
+\end{align}
+%
+where $\phi_\theta[m] = \frac{2\pi}{\lambda}dm\sin\theta$ is the steering angle.
+For this window there is an analytic solution for the optimal value of $\alpha$ in the minimum variance sense.
+To derive it, we start by inserting the window function (\ref{trig_window_function}) into the beamformer equation (\ref{z}):
 %
@@
 \w\H\x &= \sumb{m=0}{M-1} e^{j\phi_\theta[m]}\Big(\alpha - (1-\alpha)\cos\left(\frac{2\pi m}{M-1}\right)\Big) x_m^* \nn
-&= \alpha\sumb{m=0}{M-1} e^{j\phi_\theta[m]} x_m^* - (1-\alpha)\sumb{m=0}{M-1} e^{j\varphi[m]}\cos\left(\frac{2\pi m}{M-1}\right) x_m^* \nn
 &= \alpha \a\T\x + (1-\alpha)\b\T\x
@@
 \begin{align}
-\b = \text{diag}(\a)\begin{bmatrix}
+\b = \text{diag}(\a)\cdot\begin{bmatrix}
      1 &
@@
 %
-To attain unity gain in the look direction the weights must sum to one. 
+To attain unity gain in the look direction the weights must sum to one:
 %
@@
 %
-This is true for any value of $\alpha$ if $\a\T\1 = 1$ and $\b\T\1 = 1$. Hence, as long as we normalize $\a$ and $\b$ the unity gain constraint is preserved.
+This is true for any value of $\alpha$ if $\a\T\1 = 1$ and $\b\T\1 = 1$. Hence, we can preserve the unity gain constraint as long as we normalize $\a$ and $\b$.
 
@@
 %
-Differentiating this with respect to $\alpha$ yields:
+This is a convex function with a single minimum, which we obtain by differentiating with respect to $\alpha$ and setting equal to 0:
 %
@@
 In this computation there are only 4 unique additions and multiplications.
-If we used this to analytically solve for $\alpha$, but perform the search for $\phi$, the LCA's computational complexity would be of O($8MN_\phi)$ instead of O($MN_\alpha N_\phi)$.
-The value for $\alpha$ would also be the optimal.
+If we used this to analytically solve for $\alpha$, but perform the search for $\phi$, the computational complexity of LCA would be of O($8MN_\phi)$ instead of O($MN_\alpha N_\phi)$.
+The value for $\alpha$ would also be the optimal in the minimum variance sense.
 

### 2015-09-24 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:09:24	13:46:50	(Thu24.Sep)

@@
 \includegraphics[width=\linewidth]{gfx/cross_tagged.png}%
-\caption{One of the objects used to test the beamformers is a 1\,m long cross attached to an anchor with a diameter of 13\,cm.}\label{cross}
+\caption{One of the objects used to test the beamformers is a 1\,m long cross attached to an anchor with a diameter of $\approx$13\,cm.}\label{cross}
 \end{figure}
@@
 Since LCA operates with the same optimization criterion, we will be using approximately a factor 8 lateral oversampling for both beamformers.
-The images appear more visually pleasing and detailed up to 8, but we consider an absolute minimum to be 2.
-In \mediaII{} we demonstrate this with LCA images changing as a function of a lateral oversampling factor.
+The images appear more visually pleasing and detailed up to 8.
+We consider an absolute minimum to be 2, due to the non-linear nature of delaying pixels that isn't strictly in the far field, and of displaying the absolute value of the pixels on a decibel scale.
+In \mediaII{} we visualize the effect of changing the lateral oversampling factor on a set of LCA images.
 The displayed images are all upinterpolated to the same size, but the detail level varies.
@@
 % \todo{Need $M$ in the image sector. Why???}
-
-\todo{
-- Abs -> adds a factor 2\\
-- log -> infinity?\\
-- near field -> range and azimuth is coupled}
 
@@
 Observe from \Fig{mvdr_selected_windows} that MVDR seems prefers symmetric window responses, even if it is free to choose non-symmetric ones.
-The symmetry is most predominant in shadow and speckle regions, while in highlight the windows are approximately symmetric within the illuminated\todo{valid term?} sector of the seafloor.
+The symmetry is most predominant in shadow and speckle regions, while in highlight the windows are approximately symmetric within the illuminated sector of the seafloor.
 We have observed this symmetry to be a side-effect of the averaging steps needed to build the sample covariance matrix, which is key to allow  MVDR to run in near snapshot mode in an active system.
@@
 For this sonar the Rayleigh resolution is $\Delta=0.0125$ radians (0.72$^\circ$), thus a point scatterer imaged at 19\,m with a rectangular DAS would have a lateral FWHM of $0.125\cdot19\,\text{m}\cdot0.89=21\,\text{cm}$.
-This is close to the FWHM of the anchor in the DAS image in \Fig{oversampling_mosaic_bounds}.
+This is close to the FWHM of the anchor in the DAS image in \Fig{oversampling_mosaic_bounds}, which means that the acoustic fingerprint of the anchor is similar to a point source.
+This suggests the anchor reflections to be specular and that we view its rounded side.
 As we increase steering to either 50\% or 100\% of the -3dB width, the FWHM drops to 73\% or 53\% to that of the rectangular window, respectively.

### 2015-09-25 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:09:25	14:02:29	(Fri25.Sep)

@@
 \includegraphics[width=\linewidth]{gfx/cross_tagged.png}%
-\caption{One of the objects used to test the beamformers is a 1\,m long cross attached to an anchor with a diameter of $\approx$13\,cm.}\label{cross}
+\caption{One of the objects used to test the beamformers is a 1\,m long cross attached to an anchor with a diameter of approximately 13\,cm.}\label{cross}
 \end{figure}

### 2015-09-29 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:09:28	16:32:50	(Mon28.Sep)
la2pdf commit - 2015:09:28	18:41:30	(Mon28.Sep)
la2pdf commit - 2015:09:29	16:07:54	(Tue29.Sep)

@@
 
-\begin{figure*}[t]%\centering%
+\begin{figure*}[t]%
 \includegraphics[width=\textwidth]{gfx/oversampling_mosaic_bounds.pdf}%
@@
 \begin{figure*}[t]\centering
-\subfloat[]{\includegraphics[width=\textwidth]{gfx/beamformer_comparison_holmengraa.pdf}%%
-\label{oversampling_mosaic_bounds_beta}}
+\subfloat[LCA using Kaiser windows, LCA using trigonometric windows, and MVDR.
+Two lateral image cuts are shown in the leftmost figures, along with the response of a rectangularly weighted DAS.
+Note how the FWHM of LCA sits in between that of DAS and MVDR.
+Also note how LCA is insensitive to the window type; its performance is near identical whether it uses Kasier or trigonometric windows.]{\includegraphics[width=\textwidth]{gfx/beamformer_comparison_cross.pdf}%
+\label{beamformer_comparison_cross}}
 \hfil
-\subfloat[]{\includegraphics[width=\textwidth]{gfx/beamformer_comparison_cross.pdf}%
-\label{oversampling_mosaic_bounds_phi}}
-\caption{}\label{fig_sim}
+\subfloat[Image of Holmengraa.
+All the adaptive beamformers offer improvement compared to DAS.
+The shadows are deeper, the edges better refined and the detail level is higher.]{\includegraphics[width=\textwidth]{gfx/beamformer_comparison_holmengraa.pdf}%%
+\label{beamformer_comparison_holmengraa}}
+\caption{Image quality: Comparing DAS with MVDR and LCA with Kaiser and trigonometric windows.}\label{image_quality}
 \end{figure*}
@@
 
+% \begin{figure*}[t]%\centering%
+% \includegraphics[width=\textwidth]{gfx/beamformer_comparison_cross.pdf}%
+% \caption{}\label{beamformer_comparison_cross}
+% \end{figure*}
+% 
+% \begin{figure*}[t]%\centering%
+% \includegraphics[width=\textwidth]{gfx/beamformer_comparison_holmengraa.pdf}%
+% \caption{}\label{beamformer_comparison_holmengraa}
+% \end{figure*}
+
 \section{LCA}
@@
 
+\subsection{Using trigonometric windows instead of Kaiser}
+
+\begin{align}
+\phi_\text{3dB}(\beta) \approx \minus{}2.214\beta^3 + 6.084\beta^2 - 5.710\beta + 2.136
+\end{align}
+
 \subsection{Oversampling}
@@
 Assuming that a robust MVDR is the reference method we want LCA to perform similarly to, it seems sensible to create window database for LCA with with responses similar to the ones MVDR computes.
-We study this in \Fig{mvdr_selected_windows}, where the typical responses MVDR creates for shadow, speckle and highlight regions in the Holmengraa scene is presented.
-
-Observe from \Fig{mvdr_selected_windows} that MVDR seems prefers symmetric window responses, even if it is free to choose non-symmetric ones.
+We study this in \Fig{mvdr_selected_windows}, where we present the typical responses that MVDR creates for shadow, speckle and highlight regions of the Holmengraa scene.
+
+Observe from \Fig{mvdr_selected_windows} that MVDR seems to prefer symmetric window responses, even if it is free to choose non-symmetric ones.
 The symmetry is most predominant in shadow and speckle regions, while in highlight the windows are approximately symmetric within the illuminated sector of the seafloor.
-We have observed this symmetry to be a side-effect of the averaging steps needed to build the sample covariance matrix, which is key to allow  MVDR to run in near snapshot mode in an active system.
-The peak sidelobe levels of the MVDR windows they are mostly between -13\,dB and -43\,dB.
+We have observed this symmetry to be a side-effect of the averaging steps needed to build the sample covariance matrix, which is required to allow MVDR to run in near snapshot mode in an active system.
+The peak sidelobe levels of the MVDR windows are mostly between -13\,dB and -43\,dB.
 This corresponds to that of the rectangular and Hamming window, or a Kaiser window with $\beta=0$ and $\beta=2$, respectively.
@@
 
-TODO. Add image of DAS vs. MVDR vs. LCA Kaiser vs. LCA Trig
+We compare LCA image quality to that of DAS and MVDR in \Fig{image_quality}.
+\Fig{beamformer_comparison_cross} are of the cross, were we display the images computed by LCA using Kaiser windows, LCA using trigonometric windows, and MVDR.
+
+ \Fig{beamformer_comparison_holmengraa}
 

### 2015-09-30 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:09:30	15:06:29	(Wed30.Sep)

@@
 \begin{align}
-\w_{\beta,\phi} = \frac{\f_\beta\H\diag{(\a_\phi)}}{\f_\beta\H\a_\phi}
+\w_{\beta,\phi} = \frac{\f_\beta\H\diag{(\a_\phi)}}{\f_\beta\H\a_\phi}\label{steered_window}
 \end{align}
@@
 
-\subsection{Steering bounds}
+\subsection{Steering bounds}\label{steering_bounds}
 
@@
 \begin{align}
-\phi_\text{3dB}(\beta) \approx \minus{}0.002894\beta^3 + 0.02451\beta^2 - 0.01308\beta + 0.2980
+\phi_\text{3dB}(\beta) \approx \minus{}0.002894\beta^3 + 0.02451\beta^2 - 0.01308\beta + 0.2980\label{steering_kaiser_3dB}
 \end{align}
@@
 
-\subsection{Using trigonometric windows instead of Kaiser}
-
-\begin{align}
-\phi_\text{3dB}(\beta) \approx \minus{}2.214\beta^3 + 6.084\beta^2 - 5.710\beta + 2.136
+\subsection{Trigonometric windows}
+
+While we focus on using LCA with Kaiser windows, we will for completeness compare the performance with that of using trigonometric windows. This is defined as:
+%
+\begin{align}
+f_m(\alpha) = \alpha - (1-\alpha)\cos\Big(\frac{2\pi m}{M-1}\Big)\label{trig_window_function}
+\end{align}
+%
+where we apply the contraint $\alpha\in[0.5,1]$ to avoid windows with negative coefficients.
+Steering is applied as in (\ref{steered_window}), and the $\alpha$-value that halves the 3\,dB distance is found using the method described in section \ref{steering_bounds}:
+%
+\begin{align}
+\phi_\text{3dB}(\alpha) \approx \minus{}2.214\beta^3 + 6.084\beta^2 - 5.710\beta + 2.136.
 \end{align}
@@
 Throughout this article we have used LCA with variations of the Kaiser window function. However, we could instead have used a trigonometric window function:
-%
-\begin{align}
-\w = e^{j\phi_\theta[m]}\big(\alpha - (1-\alpha)\cos\left(\frac{2\pi m}{M-1}\right)\big)\label{trig_window_function}
+
+\ref{trig_window_function} \ref{steering_bounds}
+
+%
+\begin{align}
+f_m(\alpha) = e^{j\phi_\theta[m]}\Big[\alpha - (1-\alpha)\cos\Big(\frac{2\pi m}{M-1}\Big)\Big]\label{trig_window_function}
 \end{align}

### 2015-09-30 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:09:30	16:10:27	(Wed30.Sep)

@@
 %
-\begin{align*}
+\begin{align}
 \f_\beta = \bmat{
@@
 f_{M-1}(\beta)
-}
-\end{align*}
+}\label{kaiser_window_vector}
+\end{align}
 %
@@
 \begin{align}
-f_m(\beta) = \frac{I_0\left(\pi\beta\sqrt{1-\left(\frac{2m}{M-1}-1\right)^2}\right)}{I_0(\pi\beta)}
+f_m(\beta) = \frac{I_0\left(\pi\beta\sqrt{1-\left(\frac{2m}{M-1}-1\right)^2}\right)}{I_0(\pi\beta)}\label{kaiser_window_element}
 \end{align}
@@
 \begin{align}
-I_0(x) = \sumb{a=0}{\infty} \left[ \frac{\left(\frac{x}{2}\right)^a}{a!} \right]^2.
+I_0(x) = \sumb{a=0}{\infty} \left[ \frac{\left(\frac{x}{2}\right)^a}{a!} \right]^2.\label{modified_bessel_first_kind}
 \end{align}
@@
 \end{table}%
+
+% use section* for acknowledgement
+\ifCLASSOPTIONcompsoc
+  \section*{Acknowledgments}
+\else
+  \section*{Acknowledgment}
+\fi
+
+The authors would like to thank...
 
@@
 
-Throughout this article we have used LCA with variations of the Kaiser window function. However, we could instead have used a trigonometric window function:
-
-\ref{trig_window_function} \ref{steering_bounds}
+Throughout this article we have used LCA with variations of the Kaiser window function defined in (\ref{kaiser_window_element}), steered in .
+However, as shown in \Fig{image_quality} we can obtain similar performance using the trigonometric window function:
+
+\ref{trig_window_function} \ref{steering_window}
 
@@
 %
+(\ref{steering_vector})
 where $\phi_\theta[m] = \frac{2\pi}{\lambda}dm\sin\theta$ is the steering angle.
@@
 %
-This is a convex function with a single minimum, which we obtain by differentiating with respect to $\alpha$ and setting equal to 0:
+This is a convex function with a single minimum, which we find by differentiating with respect to $\alpha$ and setting equal to 0:
 %
@@
 \frac{\partial}{\partial\alpha} |\w\H\x|^2 
-&= 2\alpha (aa^* - (ab^* + a^*b) - bb^*) + ab^* + a^*b = 0
+&= 2\alpha (aa^* - (ab^* + a^*b) - bb^*) + ab^* + a^*b = 0,
 \end{align}
@@
 \begin{align}
-\alpha &= \frac{ab^* + a^*b}{2(aa^* - (ab^* + a^*b) - bb^*)}
+\alpha &= \frac{ab^* + a^*b}{2(aa^* - (ab^* + a^*b) - bb^*)}.
 \end{align}
@@
 The value for $\alpha$ would also be the optimal in the minimum variance sense.
-
-% use section* for acknowledgement
-\ifCLASSOPTIONcompsoc
-  \section*{Acknowledgments}
-\else
-  \section*{Acknowledgment}
-\fi
-
-The authors would like to thank...
 

### 2015-09-30 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:09:30	20:44:00	(Wed30.Sep)

@@
 
-Throughout this article we have used LCA with variations of the Kaiser window function defined in (\ref{kaiser_window_element}), steered in .
-However, as shown in \Fig{image_quality} we can obtain similar performance using the trigonometric window function:
-
-\ref{trig_window_function} \ref{steering_window}
-
-%
-\begin{align}
-f_m(\alpha) = e^{j\phi_\theta[m]}\Big[\alpha - (1-\alpha)\cos\Big(\frac{2\pi m}{M-1}\Big)\Big]\label{trig_window_function}
-\end{align}
-%
-(\ref{steering_vector})
-where $\phi_\theta[m] = \frac{2\pi}{\lambda}dm\sin\theta$ is the steering angle.
+Throughout this article we have used LCA with variations of the Kaiser window function defined in (\ref{kaiser_window_element}), steered as described in (\ref{steered_window}).
+However, as shown in \Fig{image_quality} we can obtain similar performance using the trigonometric window function defined in (\ref{trig_window_function} ):
+%
+\begin{align}
+f_m(\alpha) = e^{j\theta_\phi[m]}\Big[\alpha - (1-\alpha)\cos\Big(\frac{2\pi m}{M-1}\Big)\Big]\label{trig_window_function}
+\end{align}
+%
+where $\theta_\phi[m] = \frac{2\pi dm}{\lambda}\sin\phi$ is the $m$th component of the steering vector defined in (\ref{steering_vector}).
 For this window there is an analytic solution for the optimal value of $\alpha$ in the minimum variance sense.
@@
 \begin{align}
-\w\H\x &= \sumb{m=0}{M-1} e^{j\phi_\theta[m]}\Big(\alpha - (1-\alpha)\cos\left(\frac{2\pi m}{M-1}\right)\Big) x_m^* \nn
-&= \alpha \a\T\x + (1-\alpha)\b\T\x
+\w\H\x &= \sumb{m=0}{M-1} e^{j\theta_\phi[m]}\Big(\alpha - (1-\alpha)\cos\left(\frac{2\pi m}{M-1}\right)\Big) x_m^* \nn
+&= \alpha \a\T\x - (1-\alpha)\b\T\x
 \end{align}

### 2015-09-30 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:09:30	21:24:30	(Wed30.Sep)

@@
 \w\H\x &= \sumb{m=0}{M-1} e^{j\theta_\phi[m]}\Big(\alpha - (1-\alpha)\cos\left(\frac{2\pi m}{M-1}\right)\Big) x_m^* \nn
-&= \alpha \a\T\x - (1-\alpha)\b\T\x
-\end{align}
-%
-where $\a$ is the steering vector defined in (\ref{steering_vector}), and 
-%
-\begin{align}
-\b = \text{diag}(\a)\cdot\begin{bmatrix}
+&= \alpha \a_\phi\T\x - (1-\alpha)\b_\phi\T\x
+\end{align}
+%
+where $\a_\phi$ is the steering vector defined in (\ref{steering_vector}), and 
+%
+\begin{align}
+\b_\phi = \text{diag}(\a_\phi)\cdot\begin{bmatrix}
      1 &
@@
 %
-To attain unity gain in the look direction the weights must sum to one:
-%
-\begin{align}
-\w\T\1 &= \big(\alpha\a - (1-\alpha)\b\big)\T\1 = 1
+Unity gain in the look direction is ensured as long as the weights sum to one:
+%
+\begin{align}
+\w\T\1 &= \big(\alpha\a_\phi - (1-\alpha)\b_\phi\big)\T\1 = 1.
 \end{align}
@@
 
-Now let $a=\a\T x$ and $b=\b\T x$. The beamformer output can then be written as:
-%
-\begin{align}
-|\w\H\x|^2 &= \Big| \alpha a + (1-\alpha)b \Big|^2 \nn
-&= \Big[\alpha a + (1-\alpha)b\Big] \Big[\alpha a^* + (1-\alpha)b^*\Big] \nn
-&= \alpha^2 aa^* + \alpha a(1-\alpha)b^* + \alpha a^*(1-\alpha)b + (1-\alpha)^2bb^* \nn
-&= \alpha^2 aa^* + \alpha(1-\alpha)(ab^* + a^*b) + (1-\alpha)^2bb^* \nn
-&= \alpha^2 (aa^* - (ab^* + a^*b) - bb^*) + \alpha(ab^* + a^*b) + bb^* \nn
+Now let $a=\a\T\x$ and $b=\b\T\x$. The beamformer output can then be written as:
+%
+\begin{align}
+|\w\H\x|^2 &= \Big| \alpha a - (1-\alpha)b \Big|^2 \nn
+&= \alpha^2 (aa^* - (ab^* + a^*b) - bb^*) + \alpha(ab^* + a^*b) + bb^*
 \end{align}

### 2015-09-30 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:09:30	22:11:22	(Wed30.Sep)

@@
 %
-This is true for any value of $\alpha$ if $\a\T\1 = 1$ and $\b\T\1 = 1$. Hence, we can preserve the unity gain constraint as long as we normalize $\a$ and $\b$.
+This is true for any value of $\alpha$ if $\a_\phi\T\1 = 1$ and $\b_\phi\T\1 = 1$. Hence, we can preserve the unity gain constraint as long as we normalize $\a$ and $\b$.
 
@@
 |\w\H\x|^2 &= \Big| \alpha a - (1-\alpha)b \Big|^2 \nn
-&= \alpha^2 (aa^* - (ab^* + a^*b) - bb^*) + \alpha(ab^* + a^*b) + bb^*
+&= \alpha^2 (aa^* + ab^* + a^*b - bb^*) - \alpha(ab^* + a^*b + 2bb^*) + bb^*
 \end{align}

### 2015-09-30 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:09:30	23:12:27	(Wed30.Sep)

@@
 %
-This is true for any value of $\alpha$ if $\a_\phi\T\1 = 1$ and $\b_\phi\T\1 = 1$. Hence, we can preserve the unity gain constraint as long as we normalize $\a$ and $\b$.
-
-Now let $a=\a\T\x$ and $b=\b\T\x$. The beamformer output can then be written as:
+This is true for any value of $\alpha$ if $\a_\phi\T\1 = 1$ and $\b_\phi\T\1 = 1$. Hence, we can preserve the unity gain constraint as long as we normalize $\a_\phi$ and $\b_\phi$.
+
+Now let $a=\a_\phi\T\x$ and $b=\b_\phi\T\x$. The beamformer output can then be written as:
 %
@@
 |\w\H\x|^2 &= \Big| \alpha a - (1-\alpha)b \Big|^2 \nn
-&= \alpha^2 (aa^* + ab^* + a^*b - bb^*) - \alpha(ab^* + a^*b + 2bb^*) + bb^*
+&= \alpha^2 (aa^* + ab^* + a^*b + bb^*) \nn
+&- \alpha(ab^* + a^*b + 2bb^*) + bb^*
 \end{align}
@@
 \frac{\partial}{\partial\alpha} |\w\H\x|^2 
-&= 2\alpha (aa^* - (ab^* + a^*b) - bb^*) + ab^* + a^*b = 0,
+&= 2\alpha (aa^* + ab^* + a^*b + bb^*) - (ab^* + a^*b + 2bb^*) = 0,
 \end{align}

### 2015-10-01 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:09:30	23:45:36	(Wed30.Sep)
la2pdf commit - 2015:10:01	12:01:44	(Thu 1.Oct)

@@
 
-Throughout this article we have used LCA with variations of the Kaiser window function defined in (\ref{kaiser_window_element}), steered as described in (\ref{steered_window}).
-However, as shown in \Fig{image_quality} we can obtain similar performance using the trigonometric window function defined in (\ref{trig_window_function} ):
-%
-\begin{align}
-f_m(\alpha) = e^{j\theta_\phi[m]}\Big[\alpha - (1-\alpha)\cos\Big(\frac{2\pi m}{M-1}\Big)\Big]\label{trig_window_function}
-\end{align}
-%
-where $\theta_\phi[m] = \frac{2\pi dm}{\lambda}\sin\phi$ is the $m$th component of the steering vector defined in (\ref{steering_vector}).
-For this window there is an analytic solution for the optimal value of $\alpha$ in the minimum variance sense.
-To derive it, we start by inserting the window function (\ref{trig_window_function}) into the beamformer equation (\ref{z}):
+Throughout this article we have used LCA with the Kaiser window function.
+However, as shown in \Fig{image_quality} we can obtain similar performance using the trigonometric window function.
+For this window we can obtain an analytic solution for the optimal value of $\alpha$ in the minimum variance sense.
+To derive it, we start by inserting the trigonometric function (\ref{trig_window_function}), with applied steering (\ref{steering_vector}), into the beamformer equation (\ref{z}):
 %
@@
 %
-where $\a_\phi$ is the steering vector defined in (\ref{steering_vector}), and 
+where $\theta_\phi[m] = \frac{2\pi dm}{\lambda}\sin\phi$ is the $m$th component of the steering vector $\a_\phi$ defined in (\ref{steering_vector}), and 
 %
@@
 \frac{\partial}{\partial\alpha} |\w\H\x|^2 
-&= 2\alpha (aa^* + ab^* + a^*b + bb^*) - (ab^* + a^*b + 2bb^*) = 0,
+&= 2\alpha (aa^* + ab^* + a^*b + bb^*) \nn
+&- (ab^* + a^*b + 2bb^*) = 0,
 \end{align}
@@
 \begin{align}
-\alpha &= \frac{ab^* + a^*b}{2(aa^* - (ab^* + a^*b) - bb^*)}.
+\alpha &= \frac{ab^* + a^*b + 2bb^*}{2(aa^* + ab^* + a^*b + bb^*)}.
 \end{align}

### 2015-10-01 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:10:01	14:23:47	(Thu 1.Oct)

@@
 \begin{align}
-\w\H\x &= \sumb{m=0}{M-1} e^{j\theta_\phi[m]}\Big(\alpha - (1-\alpha)\cos\left(\frac{2\pi m}{M-1}\right)\Big) x_m^* \nn
+\w\H\x &= \sumb{m=0}{M-1} [\a_\phi]_m\Big(\alpha - (1-\alpha)\cos\left(\frac{2\pi m}{M-1}\right)\Big) x_m^* \nn
 &= \alpha \a_\phi\T\x - (1-\alpha)\b_\phi\T\x
@@
 %
-where $\theta_\phi[m] = \frac{2\pi dm}{\lambda}\sin\phi$ is the $m$th component of the steering vector $\a_\phi$ defined in (\ref{steering_vector}), and 
+where $[\a_\phi]_m$ is the $m$th component of the steering vector $\a_\phi$ defined in (\ref{steering_vector}), and 
 %
@@
 %
-This is true for any value of $\alpha$ if $\a_\phi\T\1 = 1$ and $\b_\phi\T\1 = 1$. Hence, we can preserve the unity gain constraint as long as we normalize $\a_\phi$ and $\b_\phi$.
+This is true for any value of $\alpha$ if $\a_\phi\T\1 = 1$ and $\b_\phi\T\1 = 1$. Hence, we  preserve the unity gain constraint as long as we normalize $\a_\phi$ and $\b_\phi$.
 
@@
 %
-In this computation there are only 4 unique additions and multiplications.
-If we used this to analytically solve for $\alpha$, but perform the search for $\phi$, the computational complexity of LCA would be of O($8MN_\phi)$ instead of O($MN_\alpha N_\phi)$.
+In this computation there are only 4 and 7 unique complex additions and multiplications, respectively.
+If we used this to analytically solve for $\alpha$, but perform the search for $\phi$, the computational complexity of LCA would be of O($11MN_\phi)$ instead of O($MN_\alpha N_\phi)$.
 The value for $\alpha$ would also be the optimal in the minimum variance sense.

### 2015-10-01 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:10:01	15:44:40	(Thu 1.Oct)

@@
     bookmarksnumbered ]{hyperref}
+
 \else
@@
 In this computation there are only 4 and 7 unique complex additions and multiplications, respectively.
-If we used this to analytically solve for $\alpha$, but perform the search for $\phi$, the computational complexity of LCA would be of O($11MN_\phi)$ instead of O($MN_\alpha N_\phi)$.
-The value for $\alpha$ would also be the optimal in the minimum variance sense.
+If we used this to analytically solve for $\alpha$, but perform the search for $\phi$, the computational complexity of LCA would be of O($MN_\phi)$ instead of O($MN_\alpha N_\phi)$.
+The solution for $\alpha$ would also yield the optimal beamformer output in the minimum variance sense.
 
@@
 \input{bio/jo_inge_buskenes}
+\input{bio/roy_edgar_hansen}
 \input{bio/andreas_austeng}

### 2015-10-01 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:10:01	16:14:50	(Thu 1.Oct)

@@
 
+\begin{figure*}[t]%\centering%
+\includegraphics[width=\textwidth]{gfx/oversampling_mosaic.pdf}%
+\caption{LCA images using different window database sizes.
+The window boundaries are the same.
+The upper left image is identical to a rectangularly weighted DAS.
+Observe that sampling the $\beta$-range more finely increases the sidelobe suppression but leaves resolution unchanged.
+Adding more steering-variations improves both sidelobe suppression and resolution.
+However, using more than  $N_\beta=2$ window types and $N_\beta=3$ steering angles makes minimal difference.}\label{oversampling_mosaic}
+\end{figure*}
+
 \begin{figure*}[t]\centering
@@
 All the adaptive beamformers offer improvement compared to DAS.
-The shadows are deeper, the edges better refined and the detail level is higher.]{\includegraphics[width=\textwidth]{gfx/beamformer_comparison_holmengraa.pdf}%%
+The shadows are deeper, the edges better defined and the detail level is higher.]{\includegraphics[width=\textwidth]{gfx/beamformer_comparison_holmengraa.pdf}%%
 \label{beamformer_comparison_holmengraa}}
@@
 
-% % \setcounter{dbltopnumber}{1}
-
-\begin{figure*}[t]%\centering%
-\includegraphics[width=\textwidth]{gfx/oversampling_mosaic.pdf}%
-\caption{LCA images using different window database sizes.
-The window boundaries are the same.
-The upper left image is identical to a rectangularly weighted DAS.
-Observe that sampling the $\beta$-range more finely increases the sidelobe suppression but leaves resolution unchanged.
-Adding more steering-variations improves both sidelobe suppression and resolution.
-However, using more than  $N_\beta=2$ window types and $N_\beta=3$ steering angles makes minimal difference.}\label{oversampling_mosaic}
-\end{figure*}
+% % \setcounter{dbltopnumber}{1
 
@@
 We compare LCA image quality to that of DAS and MVDR in \Fig{image_quality}.
-\Fig{beamformer_comparison_cross} are of the cross, were we display the images computed by LCA using Kaiser windows, LCA using trigonometric windows, and MVDR.
-
- \Fig{beamformer_comparison_holmengraa}
+In \Fig{beamformer_comparison_cross} we display the images computed by LCA using Kaiser windows, LCA using trigonometric windows, and MVDR.
+Two selected lateral cuts are presented in the left plot.
+The images and these cuts show that there is minimal difference to using Kaiser windows compared to trigonometric windows.
+Compared to DAS the LCA produces targets with an FWHM that lies in between that of DAS and MVDR.
+
+In \Fig{beamformer_comparison_holmengraa} we show full sector images of the Holmengraa wreck.
+We compare the images produced by a Hamming weidted DAS, MVDR and LCA with Kaiser or trigonometric windows.
+All the adaptive beamformers produce a sharper and less noisy image than DAS, but the difference between LCA and MVDR is minimal.
+Also here it is nearly impossible to observe any difference to using Kaiser or trigonometric windows with LCA.
 
@@
 
-In comparison, LCA is of O($M N_w$), and extremely well suited for GPUs.
+In comparison, LCA is of O($M N_\beta N_\phi$), and extremely well suited for GPUs.
 Each pixel depends on only a small subset of data, and the windows can be precomputed and stored in GPU cache for near immediate access.
-Furthermore, since we have observed LCA to perform similarly using other parametric window variants, the complexity can be reduced to near DAS.
+If we use a trigonometric window function we solve for the window parameter analytically as described in appendix \ref{reducing_lca_complexity}.
+This reduces the complexity to O($M N_\phi$), and the only remaining LCA parameter is the steering amount.
 
@@
 
-\section{Reducing LCA complexity}
+\section{Reducing LCA complexity}\label{reducing_lca_complexity}
 

### 2015-10-01 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:10:01	18:09:20	(Thu 1.Oct)

@@
 \includegraphics[width=\linewidth]{gfx/cross_tagged.png}%
-\caption{One of the objects used to test the beamformers is a 1\,m long cross attached to an anchor with a diameter of approximately 13\,cm.}\label{cross}
+\caption{One of the objects used to test the beamformers is a 1\,m long cross attached to an anchor with a diameter of approximately 13\,cm.
+Source image curtesy of Bundeswehr Technical Center for Ships and Naval Weapons Bundeswehr, Maritime Technology and Research (WTD 71).}\label{cross}
 \end{figure}
@@
 
-Adding slightly steered versions of each window to the window database gives LCA greater flexibility in searching for optimal window.
+Adding slightly steered versions of each window to the window database gives LCA greater flexibility in searching for an optimal window.
 A Kaiser window $\f_\beta$ steered to the angle $\phi$ can be expressed as
@@
 It measures 68\;m by 9\;m and fills most of the highlighted sector.
+This data was collected by the Norwegian Defense Research Establishment and Kongsberg Maritime.
 The other scene is of an iron cross of 1\,m length in both directions, connected to a an anchor with a 13\,cm diameter (\Fig{cross}).
+This data was collected by the Bundeswehr Technical Center for Ships and Naval Weapons Bundeswehr, Maritime Technology and Research (WTD 71).
 Both scenes is imaged in sectorscan mode.
@@
 % the performance of the LCA beamformer is surprisingly similar regardless of what windows it is allowed to choose from.
+
+% \newcommand\cc[1]{\multicolumn{1}{>{\columncolor{tabBlue}}c}{\bf #1}}
+% \begin{table}[!b]\centering%\normalsize
+% \begin{tabular}[c]{l r r r@{}  r r r}\hline
+% \rowcolor{tabBlue} & \cc{Resolution} & \cc{S} & \cc{$\beta$} & \cc{$\varphi^*$} &  \cc{N$_\beta$} & \cc{N$_\varphi$} \\\hline
+% & 50\%  & 28 & [0,3] & \hspace{.15cm} 50\% & 2 & 3 \\
+% & 50\%  & 28 & [0,6] & \hspace{.15cm} 50\% & 2 & 3 \\
+% & 50\%  & 28 & [0,9] & \hspace{.15cm} 50\% & 2 & 3 \\
+% & 100\% & 32 &  &:30 \\
+% & 150\% & 32 & 1 &:4\\
+% \end{tabular}\\
+% * Steering is relative to the -3dB width of the window being applied.
+% \caption{Qualitative study}\label{parameter_study}
+% \end{table}%
+
+% The simulations contained a few highlighted cylindric regions with 2\,m diameter and a constant intensity 15.4\,dB over the average speckle level.
+We focused on an object centered at 41\,m range and -3 degrees azimuth angle.
+The images produced by the LCA, MV and unweighted DAS for a single speckle realisation and a mean image of all 100 realisations are shown in \Fig{speckle}.
+Each image was normalised by their respective speckle level, and the dynamic range was clamped at \{-30,~15\}\,dB.
+The physical extent of the of the simulated object was superimposed for reference.
+%\ref{saf09}
+% 
+% In \Fig{speckle} we observe that both the adaptive beamformers produce images with a more clearly defined shadow.
+The edges are sharper and the shadow more distinct.
+This is because the adaptive beamformers have better sidelobe suppression, and thus allow less signal energy to leak from the speckle region into the shadow region.
+The same effects can also be observed laterally around the highlight region of the mean image; DAS is more prone to smear energy into the speckle region than the adaptive beamformers are.
+% 
+% \Fig{cutmean} displays two lateral cuts in the mean image, one through the highlight at 41.1\,m range and one through the shadow at 44.5\,m range.
+Each cut was computed as the mean of a 40\,cm range band centered at the respective ranges.
+The transition region between highlight and shadow is shortest for the LCA, which is advantageous since it results in a more accurate representation of the object size.
+The slightly inferior performance of the MV beamformer is due to the subarray averaging, which makes the effective array smaller~\cite{syn07}.
+% 
+% By studying the LCA performance on the 100 speckle realisations, we found the choice of window to be highly dependent on the phenomena being imaged.
+In speckle regions LCA favored narrow spatial responses with minor steering, while wide and fully steered responses were often selected in shadow regions.
+There was also a clear distinction between the responses selected in sidelobe regions and in speckle regions.
+% 
+% % \Fig{hist} illustrates how often a particular window is chosen in general.
+There are 10 $\phi$'s for each of the 10 $\beta$'s.
+The first 10 values are rectangular windows.
+Value 5 is a non-steered rectangular window, and we see that this is most frequently selected.
+At higher steering angles the windows are rarely selected, and thus we might get better results by tightening our steering boundaries.
+At value 15 we find inverted Kaiser windows with $\beta = 0.05$, and then $\beta$ is increased in increments of 0.505 for every 10th value.
+At value 65 the $\beta$ value is 2.025, and these and the remaining windows are hardly selected.
+% 
+% A sidescan image of the 1500 dwt oil tanker wreck Holmengraa is shown in \Fig{holmengraa}.
+It is about 68\,m long and 9\,m wide, and lies on a slanted seabed at 77\,m depth~\cite{holmengraa}.
+The sidescan image was created using data from the HISAS 1030 sonar, which is rather unsuited for this purpose because of its large opening angle.
+This, and the fact that the wreck was imaged at a range of about 105\,m makes the image quality poor, but sufficient to compare our beamformers.
+% 
+% In the Holmengraa image we note again that the LCA produces a cleaner shadow and better edge definition. The MV method performed almost identically to LCA in this case, and was omitted.
+
+\section{Conclusion}
+
+\begin{table}[!b]\centering%\normalsize
+\begin{tabular}[c]{l c c c}\hline
+\rowcolor{tabBlue} & \bf Highlight & \bf Shadow & \bf Speckle  \\\hline
+Symmetry             & Moderate  & High   & High \\
+Mainlobe width       & Narrow    & Wide   & Moderate \\
+Sidelobe suppression & Low       & High   & Moderate \\
+Steering             & Low       & Up to -3dB width & Moderate \\
+Oversampling         & 2x-8x     & 2x-8x  & 2x-8x
+\end{tabular}
+\caption{Summary}\label{tab:summary}
+\end{table}%
+
+LCA iterates through a set of windows in search of the one that minimizes beamformer output power.
+It can be viewed as either a reduced window solution space version of  MVDR, or as an extension to DAS.
+It creates images similar to MVDR, but at a fraction of the computational cost.
+It is also inherently robust and insensitive to the particular set of windows it selects from, supported by the observation that LCA can be used with either Kaiser or trigonometric windows and produce near identical images.
+
+Using experimental data we have demonstrated that when MVDR is made robust we constrain the solution space of the window coefficients it can choose from into mostly symmetric responses that varies between the rectangular and Hamming.
+We constructed a window database using the parametric Kaiser function, which easily allows us to create responses similar to these and those in between.
 
@@
 
-% \newcommand\cc[1]{\multicolumn{1}{>{\columncolor{tabBlue}}c}{\bf #1}}
-% \begin{table}[!b]\centering%\normalsize
-% \begin{tabular}[c]{l r r r@{}  r r r}\hline
-% \rowcolor{tabBlue} & \cc{Resolution} & \cc{S} & \cc{$\beta$} & \cc{$\varphi^*$} &  \cc{N$_\beta$} & \cc{N$_\varphi$} \\\hline
-% & 50\%  & 28 & [0,3] & \hspace{.15cm} 50\% & 2 & 3 \\
-% & 50\%  & 28 & [0,6] & \hspace{.15cm} 50\% & 2 & 3 \\
-% & 50\%  & 28 & [0,9] & \hspace{.15cm} 50\% & 2 & 3 \\
-% & 100\% & 32 &  &:30 \\
-% & 150\% & 32 & 1 &:4\\
-% \end{tabular}\\
-% * Steering is relative to the -3dB width of the window being applied.
-% \caption{Qualitative study}\label{parameter_study}
-% \end{table}%
-
-% The simulations contained a few highlighted cylindric regions with 2\,m diameter and a constant intensity 15.4\,dB over the average speckle level.
-We focused on an object centered at 41\,m range and -3 degrees azimuth angle.
-The images produced by the LCA, MV and unweighted DAS for a single speckle realisation and a mean image of all 100 realisations are shown in \Fig{speckle}.
-Each image was normalised by their respective speckle level, and the dynamic range was clamped at \{-30,~15\}\,dB.
-The physical extent of the of the simulated object was superimposed for reference.
-%\ref{saf09}
-% 
-% In \Fig{speckle} we observe that both the adaptive beamformers produce images with a more clearly defined shadow.
-The edges are sharper and the shadow more distinct.
-This is because the adaptive beamformers have better sidelobe suppression, and thus allow less signal energy to leak from the speckle region into the shadow region.
-The same effects can also be observed laterally around the highlight region of the mean image; DAS is more prone to smear energy into the speckle region than the adaptive beamformers are.
-% 
-% \Fig{cutmean} displays two lateral cuts in the mean image, one through the highlight at 41.1\,m range and one through the shadow at 44.5\,m range.
-Each cut was computed as the mean of a 40\,cm range band centered at the respective ranges.
-The transition region between highlight and shadow is shortest for the LCA, which is advantageous since it results in a more accurate representation of the object size.
-The slightly inferior performance of the MV beamformer is due to the subarray averaging, which makes the effective array smaller~\cite{syn07}.
-% 
-% By studying the LCA performance on the 100 speckle realisations, we found the choice of window to be highly dependent on the phenomena being imaged.
-In speckle regions LCA favored narrow spatial responses with minor steering, while wide and fully steered responses were often selected in shadow regions.
-There was also a clear distinction between the responses selected in sidelobe regions and in speckle regions.
-% 
-% % \Fig{hist} illustrates how often a particular window is chosen in general.
-There are 10 $\phi$'s for each of the 10 $\beta$'s.
-The first 10 values are rectangular windows.
-Value 5 is a non-steered rectangular window, and we see that this is most frequently selected.
-At higher steering angles the windows are rarely selected, and thus we might get better results by tightening our steering boundaries.
-At value 15 we find inverted Kaiser windows with $\beta = 0.05$, and then $\beta$ is increased in increments of 0.505 for every 10th value.
-At value 65 the $\beta$ value is 2.025, and these and the remaining windows are hardly selected.
-% 
-% A sidescan image of the 1500 dwt oil tanker wreck Holmengraa is shown in \Fig{holmengraa}.
-It is about 68\,m long and 9\,m wide, and lies on a slanted seabed at 77\,m depth~\cite{holmengraa}.
-The sidescan image was created using data from the HISAS 1030 sonar, which is rather unsuited for this purpose because of its large opening angle.
-This, and the fact that the wreck was imaged at a range of about 105\,m makes the image quality poor, but sufficient to compare our beamformers.
-% 
-% In the Holmengraa image we note again that the LCA produces a cleaner shadow and better edge definition. The MV method performed almost identically to LCA in this case, and was omitted.
-
-\section{Conclusion}
-
-LCA can be viewed as either a simplification of MVDR, or as an extension to DAS.
-It creates images similar to MVDR, but at a fraction of the computational cost.
-It is also inherently robust and is largely insensitive to the exact set of windows it is run with, as long as they are sensible.
-
-Using experimental data we have demonstrated that when MVDR is made robust we constrain the solution space of the window coefficients it can choose from into mostly symmetric responses that varies between the rectangular and Hamming.
-We construct a window database using the parametric Kaiser function, which easily allows us to create responses similar to these and those in between.
-The Kaiser window is optimal in the sense of maximum power near the direction of interest.
-
 % 
@@
 The tendency of the LCA to perform similarly to the MV beamformer seems to indicate that a full solution space is rarely needed in real world scenarios.
-
-\begin{table}[!b]\centering%\normalsize
-\begin{tabular}[c]{l c c c}\hline
-\rowcolor{tabBlue} & \bf Highlight & \bf Shadow & \bf Speckle  \\\hline
-Symmetry             & Moderate  & High   & High \\
-Mainlobe width       & Narrow    & Wide   & Moderate \\
-Sidelobe suppression & Low       & High   & Moderate \\
-Steering             & Low       & Up to -3dB width & Moderate \\
-Oversampling         & 2x-8x     & 2x-8x  & 2x-8x
-\end{tabular}
-\caption{Summary}\label{tab:summary}
-\end{table}%
-
-% use section* for acknowledgement
-\ifCLASSOPTIONcompsoc
-  \section*{Acknowledgments}
-\else
-  \section*{Acknowledgment}
-\fi
-
-The authors would like to thank...
 
@@
 
+% use section* for acknowledgement
+\ifCLASSOPTIONcompsoc
+  \section*{Acknowledgments}
+\else
+  \section*{Acknowledgment}
+\fi
+
+The authors would like to thank Kongsberg Maritime, the Norwegian Defence Research Establishment (FFI), and Bundeswehr Technical Center for Ships and Naval Weapons Bundeswehr (WTD 71) for the equipment and effort needed to acquire data of the cross and Holmengraa, as well as for allowing us to use it.
+
 % Can use something like this to put references on a page

### 2015-10-01 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:10:01	21:39:48	(Thu 1.Oct)

@@
 \emph{Left:}\hfill
-\parbox[t]{.95\linewidth}{MVDR sectorscan image of the oiltanker Holmengraa, using data collected from the 32 element HISAS1030 sonar attached to the HUGIN AUV.
-MVDR was run with $K=1$ temporal oversampling, $L=16$ subarray size, $d=1\%$ diagonal loading and 8 times lateral oversampling.}\protect\\\hspace{\textwidth}
+\parbox[t]{.95\linewidth}{MVDR sectorscan image of the oiltanker Holmengraa, using data collected by the 32 element HISAS1030 sonar attached to the HUGIN AUV.}\protect\\\hspace{\textwidth}
 \emph{Right:}\hfill
-\parbox[t]{.95\linewidth}{MVDR window responses computed from 40x40 pixel groups in the shadow, highlight and speckle region of the Holmengraa image.
-The responses are overlayed each other.
+\parbox[t]{.95\linewidth}{MVDR window responses computed from 40x40 pixel groups in the shadow, highlight and speckle region of the image.
+The responses are overlayed each other and the amount of overlap is colored with a logarithmic scale.
 The dashed red lines at -13dB and -43dB marks the peak sidelobe levels of an unsteered rectangular and Hamming window, respectively.
@@
 The peak sidelobe levels of the MVDR windows are mostly between -13\,dB and -43\,dB.
-This corresponds to that of the rectangular and Hamming window, or a Kaiser window with $\beta=0$ and $\beta=2$, respectively.
+This corresponds to that of an unsteered rectangular and Hamming window, or an unsteered Kaiser window with $\beta=0$ and $\beta=2$, respectively.
 
@@
 
-The images in \Fig{oversampling_mosaic_bounds} much improved sidelobe suppression with a an upper for bound for $\beta$ of 2, but it improved further as this bound was adjusted up to 5.
+The LCA images in \Fig{oversampling_mosaic_bounds} demonstrate the effect of adjusting the upper bound of $\beta$ and $\phi$.
+As suggested by \Fig{mvdr_selected_windows} we attempted first to use $\beta\in[0,2]$.
+While this resulted in much improved sidelobe suppression, there was also a slight improvement up to $\beta\in[0,5]$.
+This is likely needed to compensate for the increased sidelobe levels caused by steering.
 No noticeable difference were observed for higher values than 5.
@@
 
-LCA iterates through a set of windows in search of the one that minimizes beamformer output power.
-It can be viewed as either a reduced window solution space version of  MVDR, or as an extension to DAS.
+LCA iterates through a set of windows in search of the one that minimizes the noise and interference power in each image pixel.
+It can be viewed as either a reduced solution space version of MVDR, or as an multi-taper extension to DAS.
 It creates images similar to MVDR, but at a fraction of the computational cost.
 It is also inherently robust and insensitive to the particular set of windows it selects from, supported by the observation that LCA can be used with either Kaiser or trigonometric windows and produce near identical images.
+
+We have observed that when MVDR is constrained to work in an active system it tends to create near symmetric responses with maximum sidelobe levels between that of the rectangular and Hamming window, or Kaiser windows with $\beta=0$ and $\beta=2$, respectively.
 

### 2015-10-01 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:10:01	22:52:00	(Thu 1.Oct)

@@
 Perhaps the most studied adaptive method for reconstructing a sonar image is the minimum variance distortion-less response (MVDR) beamformer~\cite{Capon1969}.
-It computes the window that is optimal in the sense of suppressing noise as much as possible while under the restriction of  preserving the signal of interest.
-
-In many cases MVDR can improve image contrast and resolution compared to conventional static methods~\cite{Blomberg2013,Blomberg2012a,Dursun2009,Lo2004}.
+It adapts the sonar array's response to minimize the influence of noise and interference in the final image, while preserving the signal of interest.
+
+In many cases MVDR can improve the contrast and resolution of active sonar images compared to conventional static methods~\cite{Blomberg2013,Blomberg2012a,Dursun2009,Lo2004}.
 However, the computational complexity of conventional beamformers are linear with the number of channels, O($M$), while MVDR is at O($M^3$).
 This makes MVDR impracticable and sometimes infeasible.
-We have shown in previous studies that MVDR's complexity can be reduced, as well as accelerated significantly using graphics computing units (GPUs)~\cite{Buskenes2014,Asen2013}, but there is a way to avoid using MVDR altogether while achieving similar performance.
+We have shown in previous studies that MVDR's complexity can be reduced, as well as accelerated significantly using graphics computing units (GPUs)~\cite{Buskenes2014,Asen2013}, but we also a way to avoid using MVDR altogether while achieving similar performance.
 
@@
 \caption{One of the objects used to test the beamformers is a 1\,m long cross attached to an anchor with a diameter of approximately 13\,cm.
-Source image curtesy of Bundeswehr Technical Center for Ships and Naval Weapons Bundeswehr, Maritime Technology and Research (WTD 71).}\label{cross}
+Source image curtesy of Bundeswehr Technical Center for Ships and Naval Weapons, Maritime Technology and Research (WTD 71).}\label{cross}
 \end{figure}
@@
 
-So far we have determined it reasonable to use Kaiser windows in the range $\beta\in[0,5]$, steered within the range $|\phi|\in[0,\phi_\mathrm{3dB}]$. But how many windows do we need in these ranges?
+So far we have determined it reasonable to use Kaiser windows in the range $\beta\in[0,5]$, each steered within the range $|\phi|\in[0,\phi_\mathrm{3dB}(\beta)]$.
+But how many windows do we need in these ranges?
 
@@
 In \Fig{beamformer_comparison_cross} we display the images computed by LCA using Kaiser windows, LCA using trigonometric windows, and MVDR.
-Two selected lateral cuts are presented in the left plot.
+Two lateral image cuts through all images are presented in the left plot.
 The images and these cuts show that there is minimal difference to using Kaiser windows compared to trigonometric windows.
@@
 In \Fig{beamformer_comparison_holmengraa} we show full sector images of the Holmengraa wreck.
-We compare the images produced by a Hamming weidted DAS, MVDR and LCA with Kaiser or trigonometric windows.
+We compare the images produced by a Hamming weighted DAS, by MVDR and by LCA with Kaiser or trigonometric windows.
 All the adaptive beamformers produce a sharper and less noisy image than DAS, but the difference between LCA and MVDR is minimal.
-Also here it is nearly impossible to observe any difference to using Kaiser or trigonometric windows with LCA.
+The LCA image produced with Kaiser windows appear identical to that from trigonometric windows.
 

### 2015-10-02 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:10:02	16:35:00	(Fri 2.Oct)

@@
         Andreas~Austeng%
-\IEEEcompsocitemizethanks{\IEEEcompsocthanksitem All authors are with the Department
-of Informatics, University of Oslo, Norway.}% <-this % stops a space
+\IEEEcompsocitemizethanks{\IEEEcompsocthanksitem All authors are with the Department of Informatics, University of Oslo, Norway.}%
+\IEEEcompsocitemizethanks{\IEEEcompsocthanksitem R.E.Hansen is also with The Norwegian Defense Research Establishment (FFI), Norway.}% <-this % stops a space
 
@@
 
-This article is outlined as follows:
+This article is outlined as follows: In section \ref{beamforming} we introduce beamforming, \ref{mvdr} MVDR \ref{lca} LCA \ref{} \ref{} \ref{}
 
@@
 \newpage
-\subsection{Receive Beamforming}
+\subsection{Receive Beamforming}\label{beamforming}
 
@@
 
-\section{MVDR}
+\section{MVDR}\label{mvdr}
 
@@
 
-\section{LCA}
+\section{LCA}\label{lca}
 
@@
 
-The authors would like to thank Kongsberg Maritime, the Norwegian Defence Research Establishment (FFI), and Bundeswehr Technical Center for Ships and Naval Weapons Bundeswehr (WTD 71) for the equipment and effort needed to acquire data of the cross and Holmengraa, as well as for allowing us to use it.
+The authors would like to thank Kongsberg Maritime, the Norwegian Defence Research Establishment (FFI), and Bundeswehr Technical Center for Ships and Naval Weapons, Maritime Technology and Research (WTD 71), for the equipment and effort needed to acquire data of the cross and Holmengraa, as well as for allowing us to use it.
 

### 2015-10-02 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:10:02	17:09:02	(Fri 2.Oct)

@@
 \IEEEcompsocitemizethanks{\IEEEcompsocthanksitem All authors are with the Department of Informatics, University of Oslo, Norway.}%
-\IEEEcompsocitemizethanks{\IEEEcompsocthanksitem R.E.Hansen is also with The Norwegian Defense Research Establishment (FFI), Norway.}% <-this % stops a space
+\IEEEcompsocitemizethanks{\IEEEcompsocthanksitem R.~E.~Hansen is also with the Norwegian Defense Research Establishment (FFI), Norway.}% <-this % stops a space
 
@@
 
-This article is outlined as follows: In section \ref{beamforming} we introduce beamforming, \ref{mvdr} MVDR \ref{lca} LCA \ref{} \ref{} \ref{}
+This article is outlined as follows: In Section \ref{sec:beamforming} we introduce beamforming, followed by background on the MVDR beamformer in Section \ref{sec:mvdr}.
+In \ref{sec:lca} we provide complete background on the LCA method, by presenting the Kaiser window function in \ref{sec:lca_kaiser_windows}, the method for steering it in \ref{sec:lca_steering} as well as a note on maximum steering amount in \ref{sec:lca_steering_bounds}, and a mention of the alternative trigonometric window function in \ref{sec:lca_trigonometric} and sampling considerations in \ref{sec:lca_oversampling}.
 
@@
 \newpage
-\subsection{Receive Beamforming}\label{beamforming}
+\section{Receive Beamforming}\label{sec:beamforming}
 
@@
 \begin{align}
-z[n] = \w\H[n]\x[n] = \bmat{w_0[n]\\w_1[n]\\\vdots\\w_{M-1}[n]}^H \bmat{x_0[n]\\x_1[n]\\\vdots\\x_{M-1}[n]},\label{z}
+z[n] = \w\H[n]\x[n] = \bmat{w_0[n]\\w_1[n]\\\vdots\\w_{M-1}[n]}^H \bmat{x_0[n]\\x_1[n]\\\vdots\\x_{M-1}[n]},\label{eq:beamformer_output}
 \end{align}
@@
 
-\section{MVDR}\label{mvdr}
+\section{MVDR}\label{sec:mvdr}
 
@@
 \underset{\w[n]}{\argmin}\, E\{\big|z[n]\big|^2\} &= \underset{\w[n]}{\argmin}\, \w[n]\R[n]\w\H[n]\nn
-\text{subject to } \w[n]\a_\phi &= 1.\label{eq_mvdr_definition}
+\text{subject to } \w[n]\a_\phi &= 1.\label{eq:mvdr_definition}
 \end{align}
@@
 \begin{gather}
-\vec w[n] = \frac{\Ri[n]\a_\phi}{\a_\phi\T\Ri[n]\a_\phi},\label{weights}
+\vec w[n] = \frac{\Ri[n]\a_\phi}{\a_\phi\T\Ri[n]\a_\phi},\label{eq:mvdr_weights}
 \end{gather}
@@
 
-\section{LCA}\label{lca}
+\section{LCA}\label{sec:lca}
 
@@
 \underset{p}{\argmin}\, E\{\big|z_p[n]\big|^2\} = \underset{p}{\argmin}\, E\big\{\big|\w_{p}\H\x[n]\big|^2\big\}\nn
-\qquad\text{subject to}\qquad \w\H\a_\phi = 1.\label{eq_lca_definition}
+\qquad\text{subject to}\qquad \w\H\a_\phi = 1.\label{eq:lca_definition}
 \end{align}\label{lca_criterion}
 %
-Note how closely LCA is related to the MVDR definition in \eq{eq_mvdr_definition}.
+Note how closely LCA is related to the MVDR definition in \eq{eq:mvdr_definition}.
 The optimization criterion and constraint is the same, but LCA has a finite and discrete solution space for the weights.
-As will be demonstrated in section \ref{lca_window_function}, LCA will perform similarly to MVDR because a robustified MVDR implementation seems to be constrained to window functions similar to the ones we let LCA choose from.
+As will be demonstrated in section \ref{sec:results_lca_window_function}, LCA will perform similarly to MVDR because a robustified MVDR implementation seems to be constrained to window functions similar to the ones we let LCA choose from.
 
@@
 
-\subsection{Window function: Kaiser}\label{kaiser_windows}
+\subsection{Window function: Kaiser}\label{sec:lca_kaiser_windows}
 
@@
 f_{M-1}(\beta)
-}\label{kaiser_window_vector}
+}\label{eq:kaiser_window_vector}
 \end{align}
@@
 \begin{align}
-f_m(\beta) = \frac{I_0\left(\pi\beta\sqrt{1-\left(\frac{2m}{M-1}-1\right)^2}\right)}{I_0(\pi\beta)}\label{kaiser_window_element}
+f_m(\beta) = \frac{I_0\left(\pi\beta\sqrt{1-\left(\frac{2m}{M-1}-1\right)^2}\right)}{I_0(\pi\beta)}\label{eq:kaiser_window_element}
 \end{align}
@@
 \begin{align}
-I_0(x) = \sumb{a=0}{\infty} \left[ \frac{\left(\frac{x}{2}\right)^a}{a!} \right]^2.\label{modified_bessel_first_kind}
+I_0(x) = \sumb{a=0}{\infty} \left[ \frac{\left(\frac{x}{2}\right)^a}{a!} \right]^2.\label{eq:modified_bessel_first_kind}
 \end{align}
@@
 
-\subsection{Steering}
+\subsection{Steering}\label{sec:lca_steering}
 
@@
 \begin{align}
-\w_{\beta,\phi} = \frac{\f_\beta\H\diag{(\a_\phi)}}{\f_\beta\H\a_\phi}\label{steered_window}
+\w_{\beta,\phi} = \frac{\f_\beta\H\diag{(\a_\phi)}}{\f_\beta\H\a_\phi}\label{eq:steered_window}
 \end{align}
@@
 e^{-j\frac{2\pi (M-1)d}{\lambda}\sin(\phi)}
-}.\label{steering_vector}
+}.\label{eq:steering_vector}
 \end{align}
@@
 The normalization factor $\f_\beta\H\a_\phi$ is the reciprocal of the window's coherent gain.
-It ensures unit gain in the direction of interest as required by (\ref{eq_lca_definition}).
+It ensures unit gain in the direction of interest as required by (\ref{eq:lca_definition}).
 Since the signal-to-noise ratio is constant for a specific window (same $\beta$-value), this normalization also proportionally increases the incoherent noise gain.
@@
 
-\subsection{Steering bounds}\label{steering_bounds}
+\subsection{Steering bounds}\label{sec:lca_steering_bounds}
 
@@
 \begin{align}
-\phi_\text{3dB}(\beta) \approx \minus{}0.002894\beta^3 + 0.02451\beta^2 - 0.01308\beta + 0.2980\label{steering_kaiser_3dB}
+\phi_\text{3dB}(\beta) \approx \minus{}0.002894\beta^3 + 0.02451\beta^2 - 0.01308\beta + 0.2980\label{eq:steering_kaiser_3dB}
 \end{align}
@@
 
-\subsection{Trigonometric windows}
+\subsection{Trigonometric windows}\label{sec:lca_trigonometric}
 
@@
 \begin{align}
-f_m(\alpha) = \alpha - (1-\alpha)\cos\Big(\frac{2\pi m}{M-1}\Big)\label{trig_window_function}
+f_m(\alpha) = \alpha - (1-\alpha)\cos\Big(\frac{2\pi m}{M-1}\Big)\label{eq:trig_window_function}
 \end{align}
@@
 where we apply the contraint $\alpha\in[0.5,1]$ to avoid windows with negative coefficients.
-Steering is applied as in (\ref{steered_window}), and the $\alpha$-value that halves the 3\,dB distance is found using the method described in section \ref{steering_bounds}:
+Steering is applied as in (\ref{eq:steered_window}), and the $\alpha$-value that halves the 3\,dB distance is found using the method described in section \ref{sec:lca_steering_bounds}:
 %
@@
 
-\subsection{Oversampling}
+\subsection{Oversampling}\label{sec:lca_oversampling}
 
@@
 
-\section{Results and Discussion}\label{results_discussion}
+\section{Results and Discussion}\label{sec:results_discussion}
 
@@
 
-The results will be presented in the following order: In section \ref{lca_window_function} we discuss typical window responses computed by MVDR (\Fig{mvdr_selected_windows}), and hypothesize  that these can be mimicked by Kaiser windows.
-In section \ref{lca_kaiser_parameters} we investigate what sensible boundaries for the Kaiser parameter $\beta$ and window steering $\phi$ should be (\Fig{oversampling_mosaic_bounds}).
-In section \ref{lca_database_size} we discuss how many window variations is needed for LCA to perform well (\Fig{oversampling_mosaic}).
-
-\subsection{LCA window function}\label{lca_window_function}
+The results will be presented in the following order: In section \ref{sec:results_lca_window_function} we discuss typical window responses computed by MVDR (\Fig{mvdr_selected_windows}), and hypothesize  that these can be mimicked by Kaiser windows.
+In section \ref{sec:results_window_parameters} we investigate what sensible boundaries for the Kaiser parameter $\beta$ and window steering $\phi$ should be (\Fig{oversampling_mosaic_bounds}).
+In section \ref{sec:results_database_size} we discuss how many window variations is needed for LCA to perform well (\Fig{oversampling_mosaic}).
+
+\subsection{LCA window function}\label{sec:results_lca_window_function}
 
@@
 
-\subsection{Kaiser parameter $\beta$ and steering $\phi$}\label{lca_kaiser_parameters}
+\subsection{Kaiser parameter $\beta$ and steering $\phi$}\label{sec:results_window_parameters}
 
@@
 
-\subsection{Window database size}\label{lca_database_size}
+\subsection{Window database size}\label{sec:results_database_size}
 
@@
 
-\subsection{Image quality}
+\subsection{Image quality}\label{sec:results_image_quality}
 
@@
 
-\subsection{Computational complexity}
+\subsection{Computational complexity}\label{sec:results_complexity}
 
@@
 Each pixel depends on only a small subset of data, and the windows can be precomputed and stored in GPU cache for near immediate access.
-If we use a trigonometric window function we solve for the window parameter analytically as described in appendix \ref{reducing_lca_complexity}.
+If we use a trigonometric window function we solve for the window parameter analytically as described in appendix \ref{sec:reducing_lca_complexity}.
 This reduces the complexity to O($M N_\phi$), and the only remaining LCA parameter is the steering amount.
@@
 
-\section{Conclusion}
+\section{Conclusion}\label{sec:conclusion}
 
@@
 
-\subsection{Summary}
+\subsection{Summary}\label{sec:summary}
 
@@
 
-\section{Reducing LCA complexity}\label{reducing_lca_complexity}
+\section{Reducing LCA complexity}\label{sec:reducing_lca_complexity}
 
@@
 For this window we can obtain an analytic solution for the optimal value of $\alpha$ in the minimum variance sense.
-To derive it, we start by inserting the trigonometric function (\ref{trig_window_function}), with applied steering (\ref{steering_vector}), into the beamformer equation (\ref{z}):
+To derive it, we start by inserting the trigonometric function (\ref{eq:trig_window_function}), with applied steering (\ref{eq:steering_vector}), into the beamformer equation (\ref{eq:beamformer_output}):
 %
@@
 %
-where $[\a_\phi]_m$ is the $m$th component of the steering vector $\a_\phi$ defined in (\ref{steering_vector}), and 
+where $[\a_\phi]_m$ is the $m$th component of the steering vector $\a_\phi$ defined in (\ref{eq:steering_vector}), and 
 %

### 2015-10-02 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:10:02	19:28:27	(Fri 2.Oct)

@@
 In \ref{sec:lca} we provide complete background on the LCA method, by presenting the Kaiser window function in \ref{sec:lca_kaiser_windows}, the method for steering it in \ref{sec:lca_steering} as well as a note on maximum steering amount in \ref{sec:lca_steering_bounds}, and a mention of the alternative trigonometric window function in \ref{sec:lca_trigonometric} and sampling considerations in \ref{sec:lca_oversampling}.
+Results and discussion is provided in Section \ref{sec:results_discussion}, where we study MVDR to find suitable LCA windows in \ref{sec:results_lca_window_function}, assess how big the range of window types and steering should be in \ref{sec:results_window_parameters}, study image quality to determine the number of windows needed in \ref{sec:results_database_size}, and
 

### 2015-10-04 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:10:04	20:58:58	(Sun 4.Oct)

@@
 
-This article is outlined as follows: In Section \ref{sec:beamforming} we introduce beamforming, followed by background on the MVDR beamformer in Section \ref{sec:mvdr}.
-In \ref{sec:lca} we provide complete background on the LCA method, by presenting the Kaiser window function in \ref{sec:lca_kaiser_windows}, the method for steering it in \ref{sec:lca_steering} as well as a note on maximum steering amount in \ref{sec:lca_steering_bounds}, and a mention of the alternative trigonometric window function in \ref{sec:lca_trigonometric} and sampling considerations in \ref{sec:lca_oversampling}.
+This article is outlined as follows: In Section \ref{sec:beamforming} we offer a gentle introduction to beamforming.
+Then we move on to adaptive beamforming and describe the MVDR and LCA methods \ref{sec:mvdr} and \ref{sec:lca}, respectively.
+We use LCA with the Kaiser window which we describe in \ref{sec:lca_kaiser_windows}, and explain how we steer it in \ref{sec:lca_steering}, and by how much in \ref{sec:lca_steering_bounds}.
+Finally we add some remarks on using the trigonometric window instead of Kaiser in \ref{sec:lca_trigonometric}, and of of sampling considerations in \ref{sec:lca_oversampling}.
 Results and discussion is provided in Section \ref{sec:results_discussion}, where we study MVDR to find suitable LCA windows in \ref{sec:results_lca_window_function}, assess how big the range of window types and steering should be in \ref{sec:results_window_parameters}, study image quality to determine the number of windows needed in \ref{sec:results_database_size}, and

### 2015-10-04 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:10:04	21:33:05	(Sun 4.Oct)

@@
 Perhaps the most studied adaptive method for reconstructing a sonar image is the minimum variance distortion-less response (MVDR) beamformer~\cite{Capon1969}.
-It adapts the sonar array's response to minimize the influence of noise and interference in the final image, while preserving the signal of interest.
+It adapts the sonar array's response to minimize the influence of noise and interference in the final image.
 
 In many cases MVDR can improve the contrast and resolution of active sonar images compared to conventional static methods~\cite{Blomberg2013,Blomberg2012a,Dursun2009,Lo2004}.
-However, the computational complexity of conventional beamformers are linear with the number of channels, O($M$), while MVDR is at O($M^3$).
+However, while the computational complexity of conventional beamformers are linear with the number of channels, O($M$), MVDR is at O($M^3$).
 This makes MVDR impracticable and sometimes infeasible.
-We have shown in previous studies that MVDR's complexity can be reduced, as well as accelerated significantly using graphics computing units (GPUs)~\cite{Buskenes2014,Asen2013}, but we also a way to avoid using MVDR altogether while achieving similar performance.
+We have shown in previous studies that MVDR's complexity can be reduced, as well as accelerated significantly using graphics computing units (GPUs)~\cite{Buskenes2014,Asen2013}.
+However, while MVDR relies on estimating and inverting a spatial covariance matrix, there exist a method that avoid this altogether.
 
@@
 Finally we add some remarks on using the trigonometric window instead of Kaiser in \ref{sec:lca_trigonometric}, and of of sampling considerations in \ref{sec:lca_oversampling}.
-Results and discussion is provided in Section \ref{sec:results_discussion}, where we study MVDR to find suitable LCA windows in \ref{sec:results_lca_window_function}, assess how big the range of window types and steering should be in \ref{sec:results_window_parameters}, study image quality to determine the number of windows needed in \ref{sec:results_database_size}, and
+Results and discussion a provided in Section \ref{sec:results_discussion}, where we study MVDR's windows to find suitable LCA ones in \ref{sec:results_lca_window_function}, assess the range of window types and steering in \ref{sec:results_window_parameters}, determine the number of windows needed in \ref{sec:results_database_size}, and add a note on computational complexity in \ref{sec:results_complexity}.
+A conclusion is provided in \ref{sec:conclusion}.
 
@@
 \includegraphics[width=\textwidth]{gfx/oversampling_mosaic.pdf}%
-\caption{LCA images using different window database sizes.
+\caption{\emph{LCA images using different window database sizes}.
 The window boundaries are the same.
@@
 \begin{figure*}[t]\centering
-\subfloat[LCA using Kaiser windows, LCA using trigonometric windows, and MVDR.
+\subfloat[\emph{Image quality of approximate point scatterers found in the cross scene}.
 Two lateral image cuts are shown in the leftmost figures, along with the response of a rectangularly weighted DAS.
@@
 \hfil
-\subfloat[Image of Holmengraa.
-All the adaptive beamformers offer improvement compared to DAS.
-The shadows are deeper, the edges better defined and the detail level is higher.]{\includegraphics[width=\textwidth]{gfx/beamformer_comparison_holmengraa.pdf}%%
+\subfloat[\emph{Image quality of the full sector Holmengraa scene}.
+Compared to DAS the adaptive beamformers produce deeper shadows, sharper edges and a higher detail level.
+The adaptive methods produce nearly identical images.]{\includegraphics[width=\textwidth]{gfx/beamformer_comparison_holmengraa.pdf}%%
 \label{beamformer_comparison_holmengraa}}
-\caption{Image quality: Comparing DAS with MVDR and LCA with Kaiser and trigonometric windows.}\label{image_quality}
+\caption{Comparing image quality of DAS, MVDR and LCA with Kaiser or trigonometric windows.}\label{image_quality}
 \end{figure*}

### 2015-10-04 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:10:04	22:10:59	(Sun 4.Oct)

@@
 % 
-We have implemented a Low Complexity Adaptive (LCA) beamformer, which adaptively selects a window from a predefined set.
-The set is comprised of windows that are typical solutions found by the Minimum Variance method.
+We have implemented and studied the low complexity adaptive (LCA) beamformer in active sonar imaging.
+LCA is an attractive alternative to the minimum variance distortionless response (MVDR) method for minimizing noise and interference in sonar images.
+While MVDR estimates and inverts a spatial covariance matrix to compute an optimal window for the sonar array, LCA selects the optimal window from a predefined set using the same criterion as MVDR.
+
+We show that a robust MVDR implementation typically chooses approximately symmetric windows with maximum sidelobe levels between that of the rectangular and Hamming window.
+We then set up LCA to use Kaiser windows with similar responses to those of MVDR, and apply various amounts of steering.
+We find that when Steering to -3\,dB width of the window.
+This lead to a predictable resolution gain of .
+design LCA's window database with windows that resembles  constrained to work in active sonar imaging the typical window responses LCA is allowed to choose from windows that are typical solutions found by the minimum variance distortionless response (MVDR) method.
+
 The LCA beamformer was tested using simulated and experimental data from the Kongsberg Maritime HISAS 1030 sonar.
@@
 This method is called the Low Complexity Adaptive (LCA) beamformer.
-It was first proposed by Synnev\aa{}g et.
-al.
-in clinical medical imaging~\cite{Synnevag2008}, who demonstrated its ability to obtain very similar results to MVDR.
-It does this by applying a set of predefined windows and selects the one that is best able to suppress noise.
-Since it uses the same optimization criterion as MVDR, it may be considered a version of MVDR with a reduced window solution space.
+It was first proposed by Synnev\aa{}g in clinical medical imaging~\cite{Synnevag2008}, who demonstrated its ability to obtain very similar results to MVDR.
+It does this by applying a set of predefined windows and selects the one provide the best noise suppression.
+It uses the same optimization criterion as MVDR, so it may be considered a version of MVDR with a reduced window solution space.
 As we will demonstrate, this works well because a robust MVDR design tends to create windows that resembles slightly varied versions of conventional ones.
@@
 
-Synnev\aa{}g et.
-al.
-used a window set comprised of rectangular, Kaiser and inverted Kaiser functions\footnote{May also be called Kaiser-Harris windows.}.
+Synnev\aa{}g used a window set comprised of rectangular, Kaiser and inverted Kaiser functions\footnote{May also be called Kaiser-Harris windows.}.
 Of these a specific Kaiser window were micro-steered.
@@
 Our findings show that when applied to experimental data, the weights chosen by a robustified MVDR yield responses that tend to be symmetric and steered within a fraction of the -3dB resolution of the respective window.
-When letting LCA choose from Kaiser windows with similar responses, we obtain images with both improved noise suppression and resolution.
-The resolution gain can be made predictable and depending on the amount of steering we apply to the windows.
-Overall, the image quality produced by LCA is comparable to MVDR, and sometimes even better.
+When letting LCA choose from Kaiser windows with similar responses, we obtain images with improved noise suppression and resolution.
+The resolution gain is made predictable by limiting the maximum steering to a fraction of the window's -3\,dB width.
+Overall, LCA produces images that are similar to MVDR, but with a resolution in between that of DAS and MVDR.
 LCA is also inherently robust, easy to implement and fairly easy to understand.

### 2015-10-05 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:10:05	14:02:50	(Mon 5.Oct)
la2pdf commit - 2015:10:05	14:33:37	(Mon 5.Oct)

@@
 
-We show that a robust MVDR implementation typically chooses approximately symmetric windows with maximum sidelobe levels between that of the rectangular and Hamming window.
-We then set up LCA to use Kaiser windows with similar responses to those of MVDR, and apply various amounts of steering.
+We show that a robust MVDR implementation typically chooses symmetric windows with responses in between that of the rectangular and Hamming window.
+When allowing LCA to choose from similar use Kaiser windows with similar responses to those of MVDR, and apply various amounts of steering.
 We find that when Steering to -3\,dB width of the window.
@@
 
-\newpage
 \section{Receive Beamforming}\label{sec:beamforming}

### 2015-10-05 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:10:05	15:44:44	(Mon 5.Oct)

@@
 LCA is an attractive alternative to the minimum variance distortionless response (MVDR) method for minimizing noise and interference in sonar images.
-While MVDR estimates and inverts a spatial covariance matrix to compute an optimal window for the sonar array, LCA selects the optimal window from a predefined set using the same criterion as MVDR.
-
-We show that a robust MVDR implementation typically chooses symmetric windows with responses in between that of the rectangular and Hamming window.
-When allowing LCA to choose from similar use Kaiser windows with similar responses to those of MVDR, and apply various amounts of steering.
+While MVDR estimates and inverts a spatial covariance matrix to achieve this goal, LCA iterates through a predefined set of sensor weights selecting the one that best fulfills the minimum variance criterion.
+
+We show that a robust MVDR implementation typically chooses windows with responses ranging between that of the rectangular and Hamming, or Kaiser windows with $\beta\in[0,2]$, respectively.
+We let LCA choose from these Kaiser windows with similar responses to those of MVDR, and apply some variations of steering to each, , .
 We find that when Steering to -3\,dB width of the window.

### 2015-10-05 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:10:05	20:02:19	(Mon 5.Oct)

@@
 We have implemented and studied the low complexity adaptive (LCA) beamformer in active sonar imaging.
-LCA is an attractive alternative to the minimum variance distortionless response (MVDR) method for minimizing noise and interference in sonar images.
+LCA can be viewed as either a simplification of the minimum variance distortionless response (MVDR) beamformer, or as an adaptive extension to the delay and sum (DAS) beamformer.
+Both LCA and MVDR seek to minmimize noise and interference in the image, but while MVDR achieves this goal by estimating and inverting a spatial covariance matrix, LCA applies the optimal window iterates through a predefined set of sensor weights selecting the one that best fulfills the minimum variance criterion
+
+ LCA is an attractive alternative to the  for minimizing noise and interference in sonar images.
 While MVDR estimates and inverts a spatial covariance matrix to achieve this goal, LCA iterates through a predefined set of sensor weights selecting the one that best fulfills the minimum variance criterion.
+using the same criterion as MVDR.
 

### 2015-10-05 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:10:05	20:40:15	(Mon 5.Oct)

--- 2015-10-05 LCAforActiveSonarImaging/lca.tex
@@
 We have implemented and studied the low complexity adaptive (LCA) beamformer in active sonar imaging.
-LCA is an attractive alternative to the minimum variance distortionless response (MVDR) method for minimizing noise and interference in sonar images.
-While MVDR estimates and inverts a spatial covariance matrix to compute an optimal window for the sonar array, LCA selects the optimal window from a predefined set using the same criterion as MVDR.
-
-We show that a robust MVDR implementation typically chooses symmetric windows with responses in between that of the rectangular and Hamming window.
-When allowing LCA to choose from similar use Kaiser windows with similar responses to those of MVDR, and apply various amounts of steering.
+LCA can be viewed as either a simplification of the minimum variance distortionless response (MVDR) beamformer, or as an adaptive extension to the delay and sum (DAS) beamformer.
+Both LCA and MVDR attempt to minimize the power of noise and interference in the image, but while MVDR tries to find the optimal solution by estimating and inverting a spatial covariance matrix, LCA iterates through a predefined set of sensor weights (or window) and selects the closest to optimal one.
+
+We demonstrate that a robust MVDR implementation typically creates responses ranging between that of the rectangular and Hamming, or Kaiser windows with $\beta\in[0,2]$, respectively.
+We design the LCA window database with variations of Kaiser windows in this range, and add some steered versions of each.
+
  We find that when Steering to -3\,dB width of the window.

### 2015-10-06 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:10:06	10:49:10	(Tue 6.Oct)

@@
 \begin{abstract}
-\color{gray}
-
 % 
@@
 LCA can be viewed as either a simplification of the minimum variance distortionless response (MVDR) beamformer, or as an adaptive extension to the delay and sum (DAS) beamformer.
-Both LCA and MVDR attempt to minimize the power of noise and interference in the image, but while MVDR tries to find the optimal solution by estimating and inverting a spatial covariance matrix, LCA iterates through a predefined set of sensor weights (or window) and selects the closest to optimal one.
-
-We demonstrate that a robust MVDR implementation typically creates responses ranging between that of the rectangular and Hamming, or Kaiser windows with $\beta\in[0,2]$, respectively.
-We design the LCA window database with variations of Kaiser windows in this range, and add some steered versions of each.
+While both LCA and MVDR attempt to minimize the power of noise and interference in the image, MVDR tries to find the optimal solution by estimating and inverting a spatial covariance matrix, while LCA iterates through a predefined set of sensor weights (or window) and selects the closest to optimal one.
+
+We show that a robust MVDR implementation typically creates responses ranging between that of the rectangular and Hamming, or Kaiser windows with $\beta\in[0,2]$, respectively.
+We let LCA select from a set of Kaiser windows in this range, and add some variations of steering to each window.
+When up to their -3dB width.
+
+each with a little steering added.
 

### 2015-10-06 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:10:06	11:46:24	(Tue 6.Oct)

@@
 
-We show that a robust MVDR implementation typically creates responses ranging between that of the rectangular and Hamming, or Kaiser windows with $\beta\in[0,2]$, respectively.
-We let LCA select from a set of Kaiser windows in this range, and add some variations of steering to each window.
-When up to their -3dB width.
-
-each with a little steering added.
-
- We find that when Steering to -3\,dB width of the window.
-This lead to a predictable resolution gain of .
-design LCA's window database with windows that resembles  constrained to work in active sonar imaging the typical window responses LCA is allowed to choose from windows that are typical solutions found by the minimum variance distortionless response (MVDR) method.
+We show that a robust MVDR implementation typically creates responses ranging between that of the rectangular and Hamming.
+We let LCA select from a set of Kaiser windows with responses in this range, and add some steered variations of each.
+The amount of steering is limited to roughly half the response's -3\,dB width.
+
+Using experimental data we find that LCA and MVDR produces nearly identical images of large scenes.
+On point targets LCA is able to double the resolution compared to DAS, of provide half that of MVDR.
+This performance is achieved with a total of 6 windows; the unsteered rectangular and Kaiser window with $\beta=5$, and left and right steered versions out to the steering limit.
+Adding a few extra windows makes the image slightly smoother.
 
@@
 \ifPeerReview\else
-\begin{IEEEkeywords}\color{gray}
+\begin{IEEEkeywords}
 Beamforming, adaptive beamforming, MVDR, LCA, sonar, active, complexity.

### 2015-10-06 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:10:06	13:40:18	(Tue 6.Oct)

@@
 % 
-We have implemented and studied the low complexity adaptive (LCA) beamformer in active sonar imaging.
+We have studied the low complexity adaptive (LCA) beamformer in active sonar imaging.
 LCA can be viewed as either a simplification of the minimum variance distortionless response (MVDR) beamformer, or as an adaptive extension to the delay and sum (DAS) beamformer.
@@
 The amount of steering is limited to roughly half the response's -3\,dB width.
-
-Using experimental data we find that LCA and MVDR produces nearly identical images of large scenes.
-On point targets LCA is able to double the resolution compared to DAS, of provide half that of MVDR.
+Using experimental data from the Kongsberg Maritime HISAS 1030 sonar we find that LCA and MVDR produce nearly identical images of large scenes, both being superior to DAS.
+On point targets LCA is able to double the resolution compared to DAS, or provide half that of MVDR.
 This performance is achieved with a total of 6 windows; the unsteered rectangular and Kaiser window with $\beta=5$, and left and right steered versions out to the steering limit.
-Adding a few extra windows makes the image slightly smoother.
-
-The LCA beamformer was tested using simulated and experimental data from the Kongsberg Maritime HISAS 1030 sonar.
-On a simulated scene with speckle, highlight and shadow, the beamformer offered better lateral edge definition compared to the MV beamformer, and speckle intensity and shape comparable to DAS and MV beamformers.
-These results were verified by the experimental data.
-% 
+Going to 15 windows yield a slightly smoother image, but past this the difference is minimal.
+We show that LCA creates near identical images using Kaiser or trigonometric windows.
+
+All our observations point to LCA being very easy to manage.
+It simply works, and is surprisingly insensitive to the exact type of window function, steering amount, or number of windows.
+It can also be very efficiently implemented and run on any scene without the need for parameter adjustments.
+
 % An attactive trait of the LCA is its low computational complexity; while the MV beamformer is of O($M^3$), the LCA method is of O($MW$), with $M$ being the number of channels and $W$ the number of windows.
@@
 We have shown in previous studies that MVDR's complexity can be reduced, as well as accelerated significantly using graphics computing units (GPUs)~\cite{Buskenes2014,Asen2013}.
-However, while MVDR relies on estimating and inverting a spatial covariance matrix, there exist a method that avoid this altogether.
+However, while MVDR relies on estimating and inverting a spatial covariance matrix, there exist a method that avoid this altogether.\todo{samme avsnitt?}
 
@@
 
-We have observed that when MVDR is constrained to work in an active system it tends to create near symmetric responses with maximum sidelobe levels between that of the rectangular and Hamming window, or Kaiser windows with $\beta=0$ and $\beta=2$, respectively.
-
-Using experimental data we have demonstrated that when MVDR is made robust we constrain the solution space of the window coefficients it can choose from into mostly symmetric responses that varies between the rectangular and Hamming.
-We constructed a window database using the parametric Kaiser function, which easily allows us to create responses similar to these and those in between.
-
-\subsection{Summary}\label{sec:summary}
-
-\begin{itemize}
-\item \emph{Window type}.
-\Fig{mvdr_selected_windows} show frequency responses of typical windows chosen by MVDR in shadow, speckle and highlight regions in an image.
-The responses tend to be symmetric, minimally steered and have slightly varying mainlobe width and sidelobe suppression.
-We use Kaiser windows to mimic these responses.
-\item \emph{Window parameter boundaries}.
-\Fig{oversampling_mosaic_bounds} show LCA images made with the same sized window database, but containing windows with varied upper boundaries for $\beta$ and $\phi$.
+When using LCA with Kaiser windows we made their responses similar to those typically chosen by a robust MVDR in key areas in the image.
 We suggest an upper bound for $\beta$ between 2 and 5, where 5 is slightly better at suppressing sidelobes.
 For $\phi$ we suggest to steer either to 50\% or 100\% of the -3dB width, which leads to a resolution of 73\% or 54\% of the rectangular window, respectively.
-As lower boundaries we choose $\beta=0$ to always include a rectangular window, and $\phi=0$ to always include the centered window.
-\item \emph{Window database size}. \Fig{oversampling_mosaic}
-\item \emph{Performance}.
-\item \emph{Oversampling need}.
-\end{itemize}
+As lower boundaries we set $\beta=0$ to always include a rectangular window, and $\phi=0$ to always include the centered window.
+
+The tendency of the LCA to perform similarly to the MV beamformer seems to indicate that a full solution space is rarely needed in real world scenarios.
 
@@
 % A window set of 6 different responses each steered in 5 different directions proved sufficient in our experiment to match the performance of the MV beamformer.
-
-The tendency of the LCA to perform similarly to the MV beamformer seems to indicate that a full solution space is rarely needed in real world scenarios.
 

### 2015-10-06 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:10:06	14:13:02	(Tue 6.Oct)

@@
 This method is called the Low Complexity Adaptive (LCA) beamformer.
-It was first proposed by Synnev\aa{}g in clinical medical imaging~\cite{Synnevag2008}, who demonstrated its ability to obtain very similar results to MVDR.
-It does this by applying a set of predefined windows and selects the one provide the best noise suppression.
-It uses the same optimization criterion as MVDR, so it may be considered a version of MVDR with a reduced window solution space.
-As we will demonstrate, this works well because a robust MVDR design tends to create windows that resembles slightly varied versions of conventional ones.
-But which windows should LCA choose from?
+It was first proposed by Synnev\aa{}g in clinical medical imaging~\cite{Synnevag2008}, who demonstrated its ability to obtain very similar results to MVDR in a system with focused transmit beam.
+LCA applies a set of predefined windows and selects the one offering the best noise suppression.
+It relies on the same optimization criterion as MVDR, and may be considered a version of MVDR with a reduced window solution space.
+Alternatively, it may be viewed as a delay-and-sum (DAS) beamformer with an adaptive extension.
+As we will demonstrate, LCA performs surprisingly well because a robust MVDR design tends to create windows that resembles slightly varied versions of conventional ones.
 
 Synnev\aa{}g used a window set comprised of rectangular, Kaiser and inverted Kaiser functions\footnote{May also be called Kaiser-Harris windows.}.
-Of these a specific Kaiser window were micro-steered.
-His database had 12 windows in it, which proved to work well.
-But just how well did this work compared to MVDR?
-Can LCA perform even better with more windows, and could it work just as well with even fewer?
-
-In this work apply the LCA method to active sonar imaging, and aim to provide some intuitive understanding of how the window database may be designed.
-Like Synnev\aa{}g et.
-al.
-we limit our study to the use of variations of the parametric Kaiser window, since we observed negligible performance difference to using this function or any of the common trigonometric ones.
+Of these one specific Kaiser window was micro-steered.
+One of the Kaiser windows were micro-steered in 6 different directions.
+In total he let LCA choose from 12 unique windows, which worked well.
+But no comparison with MVDR was made, and it was not made clear how LCA performs with more or less windows, and if how steering affects it all.
+
+In this work we apply the LCA method to active sonar imaging.
+Our overall goal is to provide some intuitive understanding\todo{kanskje ikke helt riktig ord her} of how the window database may be designed.
+Like Synnev\aa{}g we limit our study to the use of variations of the parametric Kaiser window, since we observed negligible performance difference to using this function or any of the common trigonometric ones.
 However, we steer all our windows and not just some, and provide reason why the Kaiser window is a suitable one, what variations of Kaiser we should include, and how much it should be steered.
-
-Our findings show that when applied to experimental data, the weights chosen by a robustified MVDR yield responses that tend to be symmetric and steered within a fraction of the -3dB resolution of the respective window.
+Furthermore, unlike Synnev\aa{}g we testing it on data with unfocused transmit beam.
+
+We will show that when a robust MVDR is applied to experimental data, the weights it computes has responses that tend to be symmetric and steered within a fraction of the -3dB resolution of the respective window.
 When letting LCA choose from Kaiser windows with similar responses, we obtain images with improved noise suppression and resolution.
@@
 \emph{Right:}\hfill
-\parbox[t]{.95\linewidth}{MVDR window responses computed from 40x40 pixel groups in the shadow, highlight and speckle region of the image.
-The responses are overlayed each other and the amount of overlap is colored with a logarithmic scale.
-The dashed red lines at -13dB and -43dB marks the peak sidelobe levels of an unsteered rectangular and Hamming window, respectively.
-Note how the responses are more or less symmetric, with very little steering in the shadow, moderate steering in speckle and steering within roughly 3dB in highlight.
-The phase varies mostly in highlight to allow the steering there.} }\label{mvdr_selected_windows}
+\parbox[t]{.95\linewidth}{MVDR window amplitude and phase responses computed from 40x40 pixel groups in the shadow, highlight and speckle region of the image.
+The responses are overlayed each other and the amount of overlap is colored using a logarithmic scale.
+The dashed red lines at -13\,dB and -43\,dB marks the peak sidelobe levels of an unsteered rectangular and Hamming window, respectively.
+Note how the responses are more or less symmetric, with very little steering in the shadow, moderate steering in speckle and steering within roughly 3\,dB in highlight.
+The phase varies most in the highlight region where we see the highest contrast.} }\label{mvdr_selected_windows}
 \end{figure*}
@@
 Both MVDR and LCA was run with $K=1$ temporal averaging.
-This is a fairly aggressive yet stable set of parameters.
-
-The results will be presented in the following order: In section \ref{sec:results_lca_window_function} we discuss typical window responses computed by MVDR (\Fig{mvdr_selected_windows}), and hypothesize  that these can be mimicked by Kaiser windows.
-In section \ref{sec:results_window_parameters} we investigate what sensible boundaries for the Kaiser parameter $\beta$ and window steering $\phi$ should be (\Fig{oversampling_mosaic_bounds}).
-In section \ref{sec:results_database_size} we discuss how many window variations is needed for LCA to perform well (\Fig{oversampling_mosaic}).
+This is a fairly aggressive yet stable set of parameters.\todo{suitable ref?}
+
+The results will be presented in the following order: In section \ref{sec:results_lca_window_function} we discuss typical window responses computed by MVDR, and hypothesize  that these can be mimicked by Kaiser windows.
+In section \ref{sec:results_window_parameters} we investigate what sensible boundaries for the Kaiser parameter $\beta$ and window steering $\phi$ should be .
+In section \ref{sec:results_database_size} we discuss how many window variations is needed for LCA to perform well.
 

### 2015-10-06 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:10:06	16:04:07	(Tue 6.Oct)

@@
 LCA can be viewed as either a simplification of the minimum variance distortionless response (MVDR) beamformer, or as an adaptive extension to the delay and sum (DAS) beamformer.
-While both LCA and MVDR attempt to minimize the power of noise and interference in the image, MVDR tries to find the optimal solution by estimating and inverting a spatial covariance matrix, while LCA iterates through a predefined set of sensor weights (or window) and selects the closest to optimal one.
+While both LCA and MVDR attempt to minimize the power of noise and interference in the image, MVDR tries find an optimal way of doing it by estimating and inverting the array's spatial covariance matrix, while LCA selects the best of a  predefined set of sensor window functions.
 
@@
 We consider an absolute minimum to be 2, due to the non-linear nature of delaying pixels that isn't strictly in the far field, and of displaying the absolute value of the pixels on a decibel scale.
-In \mediaII{} we visualize the effect of changing the lateral oversampling factor on a set of LCA images.
+In \mediaI{} we visualize the effect of changing the lateral oversampling factor on a set of LCA images.
 The displayed images are all upinterpolated to the same size, but the detail level varies.
@@
 
-\mediaI{} shows how \Fig{mvdr_selected_windows} changes as a function of the MVDR subarray length $L$ and temporal averaging $K$.
+\mediaII{} shows how \Fig{mvdr_selected_windows} changes as a function of the MVDR subarray length $L$ and temporal averaging $K$.
 It demonstrates that at $L=1$ the window responses are rectangular in all areas in the image.
@@
 The images also demonstrate that increasing the steering $\phi$ improves the resolution of the strong scatterers.
-For this sonar the Rayleigh resolution is $\Delta=0.0125$ radians (0.72$^\circ$), thus a point scatterer imaged at 19\,m with a rectangular DAS would have a lateral FWHM of $0.125\cdot19\,\text{m}\cdot0.89=21\,\text{cm}$.
+For this sonar the Rayleigh resolution is $\Delta=0.0125$ radians (0.72$^\circ$), thus a point scatterer imaged at 19\,m with a rectangular DAS would have a lateral $\text{FWHM}\approx0.125\cdot19\,\text{m}\cdot0.89=21\,\text{cm}$.
 This is close to the FWHM of the anchor in the DAS image in \Fig{oversampling_mosaic_bounds}, which means that the acoustic fingerprint of the anchor is similar to a point source.

### 2015-10-06 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:10:06	18:09:38	(Tue 6.Oct)

@@
 LCA can be viewed as either a simplification of the minimum variance distortionless response (MVDR) beamformer, or as an adaptive extension to the delay and sum (DAS) beamformer.
-While both LCA and MVDR attempt to minimize the power of noise and interference in the image, MVDR tries find an optimal way of doing it by estimating and inverting the array's spatial covariance matrix, while LCA selects the best of a  predefined set of sensor window functions.
+While both LCA and MVDR attempt to minimize the power of noise and interference in the image, MVDR seeks the optimal solution by estimating and inverting the array's spatial covariance matrix, while LCA selects the best performer from a predefined set of sensor windows.
 
@@
 We let LCA select from a set of Kaiser windows with responses in this range, and add some steered variations of each.
-The amount of steering is limited to roughly half the response's -3\,dB width.
+We limit the steering to roughly half the -3\,dB width of the window's amplitude response.
 Using experimental data from the Kongsberg Maritime HISAS 1030 sonar we find that LCA and MVDR produce nearly identical images of large scenes, both being superior to DAS.
 On point targets LCA is able to double the resolution compared to DAS, or provide half that of MVDR.
-This performance is achieved with a total of 6 windows; the unsteered rectangular and Kaiser window with $\beta=5$, and left and right steered versions out to the steering limit.
-Going to 15 windows yield a slightly smoother image, but past this the difference is minimal.
-We show that LCA creates near identical images using Kaiser or trigonometric windows.
-
-All our observations point to LCA being very easy to manage.
+This performance is achieved with a total of 6 windows; the rectangular and Kaiser window with $\beta=5$, in unsteered version, and left and right steered to the steering limit.
+Slightly smoother images are produced if the window count is increased to 15, but past this we observe minimal difference.
+Finally we show that LCA work just as well if Kaiser windows are substituted with trigonometric ones.
+
+All our observations and experiences point to LCA being very easy to manage.
 It simply works, and is surprisingly insensitive to the exact type of window function, steering amount, or number of windows.
-It can also be very efficiently implemented and run on any scene without the need for parameter adjustments.
+It can be efficiently implemented on parallel hardware, and handles any scene without the need for parameter adjustments.
 
@@
 Bats, for instance, use a high frequency sonar to detect, identify and track prey with amazing precision in caves amongst a myriad of other bats.
-Human made sonars have little to no adaptability when data is collected, but some methods have been investigated in the data post processing.
-Perhaps the most studied adaptive method for reconstructing a sonar image is the minimum variance distortion-less response (MVDR) beamformer~\cite{Capon1969}.
+Active human made sonars have little to no adaptability on transmit, but several adaptive methods have been investigated post process the received data.
+Perhaps the most studied method for reconstructing a sonar image is the minimum variance distortion-less response (MVDR) beamformer~\cite{Capon1969}.
 It adapts the sonar array's response to minimize the influence of noise and interference in the final image.

### 2015-10-06 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:10:06	18:49:45	(Tue 6.Oct)

@@
 Active human made sonars have little to no adaptability on transmit, but several adaptive methods have been investigated post process the received data.
-Perhaps the most studied method for reconstructing a sonar image is the minimum variance distortion-less response (MVDR) beamformer~\cite{Capon1969}.
+Perhaps the best studied method for reconstructing a sonar image is the minimum variance distortion-less response (MVDR) beamformer~\cite{Capon1969}.
 It adapts the sonar array's response to minimize the influence of noise and interference in the final image.
@@
 This makes MVDR impracticable and sometimes infeasible.
-We have shown in previous studies that MVDR's complexity can be reduced, as well as accelerated significantly using graphics computing units (GPUs)~\cite{Buskenes2014,Asen2013}.
+We have shown in previous studies that MVDR's complexity can be reduced, and its implementation accelerated significantly using graphics computing units (GPUs)~\cite{Buskenes2014,Asen2013}.
 However, while MVDR relies on estimating and inverting a spatial covariance matrix, there exist a method that avoid this altogether.\todo{samme avsnitt?}
@@
 This method is called the Low Complexity Adaptive (LCA) beamformer.
-It was first proposed by Synnev\aa{}g in clinical medical imaging~\cite{Synnevag2008}, who demonstrated its ability to obtain very similar results to MVDR in a system with focused transmit beam.
+It was first proposed by Synnev\aa{}g in clinical medical imaging~\cite{Synnevag2008}, who demonstrated its ability to obtain very similar results to MVDR in a system with focused transmit beams.
 LCA applies a set of predefined windows and selects the one offering the best noise suppression.
@@
 Alternatively, it may be viewed as a delay-and-sum (DAS) beamformer with an adaptive extension.
-As we will demonstrate, LCA performs surprisingly well because a robust MVDR design tends to create windows that resembles slightly varied versions of conventional ones.
 
 Synnev\aa{}g used a window set comprised of rectangular, Kaiser and inverted Kaiser functions\footnote{May also be called Kaiser-Harris windows.}.
-Of these one specific Kaiser window was micro-steered.
-One of the Kaiser windows were micro-steered in 6 different directions.
-In total he let LCA choose from 12 unique windows, which worked well.
-But no comparison with MVDR was made, and it was not made clear how LCA performs with more or less windows, and if how steering affects it all.
-
-In this work we apply the LCA method to active sonar imaging.
-Our overall goal is to provide some intuitive understanding\todo{kanskje ikke helt riktig ord her} of how the window database may be designed.
-Like Synnev\aa{}g we limit our study to the use of variations of the parametric Kaiser window, since we observed negligible performance difference to using this function or any of the common trigonometric ones.
-However, we steer all our windows and not just some, and provide reason why the Kaiser window is a suitable one, what variations of Kaiser we should include, and how much it should be steered.
-Furthermore, unlike Synnev\aa{}g we testing it on data with unfocused transmit beam.
+He made LCA choose from 12 unique windows, 6 of which were micro-steered variants of a fairly wide Kaiser window.
+While achieving good results, it was not made clear how LCA performs with more or fewer windows, and how steering affects it all.
+
+In this work we apply the LCA method to data from the Kongsberg Maritime HISAS 1030 synthetic aperture sonar (SAS).
+It has 32 sensors, operates at 100\,kHz and an opening angle of 25$^\circ$.
+We focus on using LCA with variations of the parametric Kaiser window, but a comparison to using trigonometric windows will also be provided.
+Unlike Synnev\aa{}g we steer all our windows and not just some, and elaborate on the type of Kaiser windows to include and the method of steering.
 

### 2015-10-06 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:10:06	19:32:19	(Tue 6.Oct)

@@
 
-In this work we apply the LCA method to data from the Kongsberg Maritime HISAS 1030 synthetic aperture sonar (SAS).
-It has 32 sensors, operates at 100\,kHz and an opening angle of 25$^\circ$.
-We focus on using LCA with variations of the parametric Kaiser window, but a comparison to using trigonometric windows will also be provided.
-Unlike Synnev\aa{}g we steer all our windows and not just some, and elaborate on the type of Kaiser windows to include and the method of steering.
-
-We will show that when a robust MVDR is applied to experimental data, the weights it computes has responses that tend to be symmetric and steered within a fraction of the -3dB resolution of the respective window.
+We use LCA with variations of the Kaiser window, but a comparison to using trigonometric windows will also be provided.
+We extend Synnev\aa{}g's work by steering all our windows and not just some, and elaborate on the type of Kaiser windows to include and the method of steering.
+Then the method is applied to experimental data from the 32 element Kongsberg Maritime HISAS 1030 synthetic aperture sonar (SAS), with an operating frequency of 100\,kHz and transmit angle of 25$^\circ$.
+
+Our results show that MVDR computes weights with responses that tend to be symmetric and steered within a fraction of the window's -\,3dB resolution.
 When letting LCA choose from Kaiser windows with similar responses, we obtain images with improved noise suppression and resolution.
@@
 Then we move on to adaptive beamforming and describe the MVDR and LCA methods \ref{sec:mvdr} and \ref{sec:lca}, respectively.
-We use LCA with the Kaiser window which we describe in \ref{sec:lca_kaiser_windows}, and explain how we steer it in \ref{sec:lca_steering}, and by how much in \ref{sec:lca_steering_bounds}.
+We use LCA with the Kaiser window which we describe in \ref{sec:lca_kaiser_windows}, explain how we steer it in \ref{sec:lca_steering}, and by how much in \ref{sec:lca_steering_bounds}.
 Finally we add some remarks on using the trigonometric window instead of Kaiser in \ref{sec:lca_trigonometric}, and of of sampling considerations in \ref{sec:lca_oversampling}.
 Results and discussion a provided in Section \ref{sec:results_discussion}, where we study MVDR's windows to find suitable LCA ones in \ref{sec:results_lca_window_function}, assess the range of window types and steering in \ref{sec:results_window_parameters}, determine the number of windows needed in \ref{sec:results_database_size}, and add a note on computational complexity in \ref{sec:results_complexity}.
-A conclusion is provided in \ref{sec:conclusion}.
+Finally we conclude in \ref{sec:conclusion}.
 

### 2015-10-06 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:10:06	22:07:36	(Tue 6.Oct)

@@
 The filter is called an array processor, or receive beamformer.
-The basic principle is to apply a delay and weights to the sensor channels before summing them up, chosen such that signals from the location of interest are summed coherently, while other sources sum incoherently.
+The basic principle is to apply delay and weights to the sensor channels before summing them up, chosen such that signals from the location of interest are summed coherently, while other sources sum incoherently.
 
@@
 
-MVDR seeks to minimize the power of the beamformer, under the constraint of a unity gain in some desired direction $\phi$~\cite{Capon1969}:
+MVDR seeks to minimize the output power of the beamformer, under the constraint of a unity gain in some desired direction $\phi$~\cite{Capon1969}:
 %
@@
 The problem lies in estimating and inverting this spatial covariance matrix.
-We have described the exact steps in~\cite{Buskenes2014}.
-To avoid signal cancellation we apply spatial averaging by computing a mean covariance matrix from a set of subarrays with length $L$~\cite{Kailath1985}.
+We have described the exact steps in~\cite{Buskenes2014}: To avoid signal cancellation we apply spatial averaging by computing a mean covariance matrix from a set of subarrays with length $L$~\cite{Kailath1985}.
 For true speckle statistics we perform temporal averaging over $N_k = 2K+1$ temporal samples~\cite{Synnevag2009a}.

### 2015-10-06 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:10:06	22:43:04	(Tue 6.Oct)

@@
 The problem lies in estimating and inverting this spatial covariance matrix.
-We have described the exact steps in~\cite{Buskenes2014}: To avoid signal cancellation we apply spatial averaging by computing a mean covariance matrix from a set of subarrays with length $L$~\cite{Kailath1985}.
-For true speckle statistics we perform temporal averaging over $N_k = 2K+1$ temporal samples~\cite{Synnevag2009a}.
-Finally, to improve robustness to parameter errors we apply $d$ percent diagonal loading~\cite{Cox1987,Maksym1979}.
+We have described the exact steps in~\cite{Buskenes2014}: To avoid signal cancellation we apply spatial averaging by computing a mean covariance matrix from a set of subarrays with length $L$~\cite{Kailath1985}, for true speckle statistics we perform temporal averaging over $N_k = 2K+1$ temporal samples~\cite{Synnevag2009a}, and to improve robustness to parameter errors we add $d$ percent diagonal loading~\cite{Cox1987,Maksym1979}.
 These steps are also needed to ensure that the covariance matrix is numerically well conditioned and hence invertible.
@@
 \includegraphics[width=\linewidth]{gfx/cross_tagged.png}%
-\caption{One of the objects used to test the beamformers is a 1\,m long cross attached to an anchor with a diameter of approximately 13\,cm.
+\caption{One of the objects used to test the beamformers is a 1\,m by 1\,m cross attached to an anchor with a diameter of approximately 13\,cm.
 Source image curtesy of Bundeswehr Technical Center for Ships and Naval Weapons, Maritime Technology and Research (WTD 71).}\label{cross}
@@
 \emph{Left:}\hfill
-\parbox[t]{.95\linewidth}{MVDR sectorscan image of the oiltanker Holmengraa, using data collected by the 32 element HISAS1030 sonar attached to the HUGIN AUV.}\protect\\\hspace{\textwidth}
+\parbox[t]{.95\linewidth}{MVDR sectorscan image of the oiltanker Holmengraa, with the pixel groups are indicated with red boxes.}\protect\\\hspace{\textwidth}
 \emph{Right:}\hfill
@@
 
-If extra flexibility is desired, we can also estimate the variance by a weighted combination of local pixels
+\todo{Drop this paragraph altogether?}If extra flexibility is desired, we can also estimate the variance by a weighted combination of local pixels
 %
@@
 where $N_x = 2X+1$ is the number of azimuth lines to average over, and $w_\text{prf}[x',n']$ is a normalized 2 dimensional weight function. 
-\todo{Might want to be careful with this when we steer the windows, or drop it altogether}
 % 
@@
 %
-where $T$ is the extent of the window in time and $B$ is its bandwidth.
+where $T$ is the extent of the window in time\todo{Should perhaps say space-bandwidth product?} and $B$ is its bandwidth.
 Adjusting $\beta$ changes the trade-off between mainlobe width and sidelobe level.
 When $\beta=0$ the window becomes rectangular, while at large values ($\beta>5$) the window converges to a Gaussian both in time and frequency.
-This class of windows is generally considered well suited for separating closely spaced sources with amplitudes of a high dynamic range~\cite{Harris1978}, they are easy to make, and they are optimal for any value of $\beta$.
+This class of windows is generally considered well suited for separating closely spaced\todo{better word?} sources with amplitudes of a high dynamic range~\cite{Harris1978}, they are easy to make, and they are optimal for any value of $\beta$.
 
@@
 %
-Here $d$ is the element spacing, $\lambda$ is the wavelength and $\phi$ is the steering amount.
+Here $d$\todo{$d$ is reused as element spacing, find an alternative} is the element spacing, $\lambda$ is the wavelength and $\phi$ is the steering amount.
 The normalization factor $\f_\beta\H\a_\phi$ is the reciprocal of the window's coherent gain.
@@
 
-There is an upper bound to the amount of steering we should apply to a window.
-Since each window's response is constrained to a unit gain in the look direction, the steered window will have increased white noise gain and deteriorated signal-to-noise ratio as shown in \Fig{windows_steering}.
-To limit this we placing an upper bound to the steering, realizing that  a wide window will be allowed more steering than a narrow one.
-
-We call the upper bound for steering $\phi_{\text{3dB}}(\beta)$, and define it as the steering angle needed for the steered response to have a -3dB crossing that is exactly half that of the unsteered window.
+Each window's response is constrained to unit gain in the look direction.
+Hence, when it is steered the white noise gain must increase and the the signal-to-noise ratio decrease, as shown in \Fig{windows_steering}.
+To limit this we place an upper bound to the steering, chosen such that wide windows is allowed to be steered more than narrow ones.
+We call this upper bound for steering $\phi_{\text{3dB}}(\beta)$, and define it as the steering angle needed for the steered response to have a \minus{}3\,dB crossing that is exactly half that of the unsteered window.
 Using a numerical search for the steering angle $\phi$ that fulfilled this criterion for a wide range of $\beta$-values, we obtained a third degree polynomial fit to these points:
@@
 
-Since LCA operates with the same optimization criterion, we will be using approximately a factor 8 lateral oversampling for both beamformers.
+Since LCA operates with the same optimization criterion as MVDR, we will be using approximately a factor 8 lateral oversampling for both beamformers.
 The images appear more visually pleasing and detailed up to 8.
@@
 In \mediaI{} we visualize the effect of changing the lateral oversampling factor on a set of LCA images.
-The displayed images are all upinterpolated to the same size, but the detail level varies.
+Upon display the images are all bilinearly upinterpolated to the same size.
 % 
@@
 
-Two different scenes will be presented.
+Two different scenes will be studied.
 One of the 1500 DWT oil tanker wreck Holmengraa lying at a slanted seabed at 77\;m depth outside of Horten, Norway~\cite{holmengraa}.
@@
 This data was collected by the Norwegian Defense Research Establishment and Kongsberg Maritime.
-The other scene is of an iron cross of 1\,m length in both directions, connected to a an anchor with a 13\,cm diameter (\Fig{cross}).
-This data was collected by the Bundeswehr Technical Center for Ships and Naval Weapons Bundeswehr, Maritime Technology and Research (WTD 71).
-Both scenes is imaged in sectorscan mode.
-
-In the image reconstruction we have run MVDR with subarray length $L=16$ and $d=1\%$ diagonal loading.
+The other scene is of an 1\,m by 1\,m iron cross, connected to a an anchor with 13\,cm diameter (\Fig{cross}).
+This data was collected by the Bundeswehr Technical Center for Ships and Naval Weapons, Maritime Technology and Research (WTD 71).
+Both scenes are imaged in sectorscan mode.
+
+In the image reconstruction we have run MVDR with subarray length $L=16$ and $d=1\%$ diagonal loading\todo{also used for element distance}.
 Both MVDR and LCA was run with $K=1$ temporal averaging.
@@
 The results will be presented in the following order: In section \ref{sec:results_lca_window_function} we discuss typical window responses computed by MVDR, and hypothesize that these can be mimicked by Kaiser windows.
-In section \ref{sec:results_window_parameters} we investigate what sensible boundaries for the Kaiser parameter $\beta$ and window steering $\phi$ should be .
-In section \ref{sec:results_database_size} we discuss how many window variations is needed for LCA to perform well.
+In section \ref{sec:results_window_parameters} we determine sensible boundaries for the Kaiser parameter $\beta$ and window steering $\phi$.
+In section \ref{sec:results_database_size} we discuss how many window variations are needed for LCA to perform well.
 
@@
 
-Assuming that a robust MVDR is the reference method we want LCA to perform similarly to, it seems sensible to create window database for LCA with with responses similar to the ones MVDR computes.
-We study this in \Fig{mvdr_selected_windows}, where we present the typical responses that MVDR creates for shadow, speckle and highlight regions of the Holmengraa scene.
+Assuming that a robust MVDR is the reference method we want LCA to perform similarly to, it seems sensible to create a window database for LCA with with responses similar to the ones that MVDR computes.
+We study this in \Fig{mvdr_selected_windows}, where we present the typical amplitude and phase responses that MVDR creates for shadow, speckle and highlight regions of the Holmengraa scene.
 
@@
 The symmetry is most predominant in shadow and speckle regions, while in highlight the windows are approximately symmetric within the illuminated sector of the seafloor.
-We have observed this symmetry to be a side-effect of the averaging steps needed to build the sample covariance matrix, which is required to allow MVDR to run in near snapshot mode in an active system.
-The peak sidelobe levels of the MVDR windows are mostly between -13\,dB and -43\,dB.
+We have observed this symmetry to be a side-effect of the averaging steps needed to build the sample covariance matrix, which are required for MVDR to operate in near snapshot mode in an active system.
+The peak sidelobe levels of the MVDR windows are mostly between \minus{}13\,dB and \minus{}43\,dB.
 This corresponds to that of an unsteered rectangular and Hamming window, or an unsteered Kaiser window with $\beta=0$ and $\beta=2$, respectively.

### 2015-10-06 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:10:06	23:37:16	(Tue 6.Oct)

@@
 
-From this we hypothesize that LCA should work even if it only selects from various Kaiser windows.
-If they span a decent range of $\beta$-values, and some steering variations is applied to each, we should have responses that resembles those in \Fig{mvdr_selected_windows}.
+From this we hypothesize that a good window database LCA can be made using a varied  set of Kaiser windows.
+If they span a suitable range of $\beta$-values, and some steering variations are applied to each, we should have responses that resemble those in \Fig{mvdr_selected_windows}.
 The Kaiser window is easy to compute, is fairly insensitive to coefficient inaccuracies, and can create windows with shapes spanning from rectangular to Gaussian.
@@
 Unlike the method described by Synnev\aa{}g~\cite{Synnevag2008} we do not let LCA choose from inverted Kaiser windows.
-These have a spatial response with mainlobe width narrower than that of a rectangular window, but at the expense of drastically increased white noise gain.
-We have found that these windows hardly ever get used on experimental data, in particular steered versions with an even worse white noise gain.
-We can infer the same from \Fig{mvdr_selected_windows} by noting that the maximum sidelobe level rarely exceeds the rectangular window level of -13\;dB.
-
-\mediaII{} shows how \Fig{mvdr_selected_windows} changes as a function of the MVDR subarray length $L$ and temporal averaging $K$.
-It demonstrates that at $L=1$ the window responses are rectangular in all areas in the image.
-At $L=2$ we can observe that pixels in highlight are processed with windows with slightly varying amplitude responses, but with the phase being 0$^\circ$ in the illuminated seafloor sector.
-Yet MVDR is able to greatly suppress noise, which seems to emphasize the importance varying the zero locations slightly to suppress noise.
-The media file also shows MVDR can be run with subarray sizes $L\in[M/2,5M/8]$, but only with temporal averaging $K=1$ or above.
+These have a mainlobe width narrower than that of the rectangular window, but at the expense of a very poor white noise gain.
+We have found that these windows hardly ever get used on experimental data, in particular steered versions of them.
+We can infer the same from \Fig{mvdr_selected_windows} by noting that the maximum sidelobe level rarely exceeds the rectangular window level of \minus{}13\,dB.
+
+\mediaII{} animates how \Fig{mvdr_selected_windows} changes as a function of the MVDR subarray length $L$ and temporal averaging $K$.
+At $L=1$ the window responses are rectangular in all areas in the image.
+At $L=2$ we observe window responses with slight amplitude variations in the highlight region, but with the phase being 0$^\circ$ in the illuminated seafloor sector.
+Already at $L=2$ the MVDR is able to greatly suppress noise.
+This is a common observation; adding a little flexibility to adapt to the scene has a dramatic effect, but further allowing full flexibility is much less significant.
+The media file also shows that MVDR can be run with subarray sizes $L\in[M/2,5M/8]$, but only with temporal averaging $K=1$ or above.
 
@@
 
-What is the best set of Kaiser windows to use with LCA, and how much should they be steered?
-To investigate this we constructed large and equally sized window databases containing windows with varied upper boundaries for $\beta$ and $\phi$.
-As lower boundaries we chose $\beta=0$ and $\phi=0^\circ$, which includes the symmetric rectangularly weighted DAS image.
-The images produced are shown in \Fig{oversampling_mosaic_bounds}.
-In each image we computed the lateral distance $\Delta x$ between the -3dB points of the anchor, with the reference $\Delta x_\text{ref}$ being the rectangularly weighted DAS image.
+To determine a suitable range for the Kaiser parameter $\beta$ and steering $\phi$, we constructed large and equally sized window databases containing Kaiser windows with varied upper boundaries for $\beta$ and $\phi$.
+The lower boundaries were chosen as $\beta=0$ and $\phi=0^\circ$, which includes the rectangularly weighted DAS image.
+The resulting images are shown in \Fig{oversampling_mosaic_bounds}.
+In each image we computed the lateral distance $\Delta x$ between the \minus{}3dB points of the anchor, with the reference $\Delta x_\text{ref}$ being the rectangularly weighted DAS image.
 A forth order polynomial fit was used in this computation.

### 2015-10-07 Simulator/article.tex
la2pdf commit - 2015:07:23	20:45:57	(Thu23.Jul)
la2pdf commit - 2015:10:07	09:27:27	(Wed 7.Oct)

@@
 
-\title{A Real-Time Sonar Simulator for Automatic Target Recognition}
-
-\author{{Jo~Inge~Buskenes,~\IEEEmembership{Student Member,~IEEE,} %
-        Herman~Midelfart, %
-        \O{}ivind~Midtgaard} %
-\thanks{J. I. Buskenes is with the Department of Informatics, University of Oslo, Norway.}%
-\thanks{H. Midelfart and \O. Midtgaard are with The Norwegian Defence Research Establishment (FFI), Norway.}% <-this % stops a space
-% \thanks{Manuscript received April 19, 2005; revised January 11, 2007.}
-}
-
-% The paper headers
-\markboth{IEEE Journal of Oceanic Engineering}%
-{A GPU Sonar Simulator for Automatic Target Recognition}
-
-% Publishers ID mark:
-%\IEEEpubid{0000--0000/00\$00.00~\copyright~2007 IEEE}
-
-% use for special paper notices
-%\IEEEspecialpapernotice{(Invited Paper)}
-
-% for Computer Society papers, we must declare the abstract and index terms
-% PRIOR to the title within the \IEEEcompsoctitleabstractindextext IEEEtran
-% command as these need to go into the title area created by \maketitle.
-\IEEEcompsoctitleabstractindextext{%
-\begin{abstract}
-Template matching is a common technique for automatic classification of objects in synthetic aperture sonar (SAS) images.
+% 
+% \title{A Real-Time Sonar Simulator for Automatic Target Recognition}
+% 
+% \author{{Jo~Inge~Buskenes,~\IEEEmembership{Student Member,~IEEE,} %
+%         Herman~Midelfart, %
+%         \O{}ivind~Midtgaard} %
+% \thanks{J. I. Buskenes is with the Department of Informatics, University of Oslo, Norway.}%
+% \thanks{H. Midelfart and \O. Midtgaard are with The Norwegian Defence Research Establishment (FFI), Norway.}% <-this % stops a space
+% % \thanks{Manuscript received April 19, 2005; revised January 11, 2007.}
+% }
+% 
+% % The paper headers
+% \markboth{IEEE Journal of Oceanic Engineering}%
+% {A GPU Sonar Simulator for Automatic Target Recognition}
+% 
+% % Publishers ID mark:
+% %\IEEEpubid{0000--0000/00\$00.00~\copyright~2007 IEEE}
+% 
+% % use for special paper notices
+% %\IEEEspecialpapernotice{(Invited Paper)}
+% 
+% % for Computer Society papers, we must declare the abstract and index terms
+% % PRIOR to the title within the \IEEEcompsoctitleabstractindextext IEEEtran
+% % command as these need to go into the title area created by \maketitle.
+% \IEEEcompsoctitleabstractindextext{%
+% \begin{abstract}
+% Template matching is a common technique for automatic classification of objects in synthetic aperture sonar (SAS) images.
 The principle is to isolate an image segment containing the object of interest, correlate it with a set of template images, and assign it to the class of the template yielding the highest correlation coefficient.
 The challenge is to come up with a representative set of template images covering the relevant configurations of object and seabed.
-
-We target this challenge with a sonar simulator that first takes as input a seabed model derived from the real sonar image.
+% 
+% We target this challenge with a sonar simulator that first takes as input a seabed model derived from the real sonar image.
 Then it places a 3D object model on the seabed, renders the scene, and adds the resulting image to the template set.
@@
 The result is a fast performing and portable template generator which can adapt to the characteristics of the current scene on-the-fly.
-\end{abstract}
-
-% Keywords (normally not used for peer reviews)
-\ifPeerReview\else
-\begin{IEEEkeywords}
-Sonar, SAS, simulator, template matching, OpenGL, OpenCL.
-\end{IEEEkeywords}
-\fi}
-% \fi
-
-% make the title area
-\maketitle
-
-% This command fixes abstract positioning for compsoc articles:
-\IEEEdisplaynotcompsoctitleabstractindextext
-
-% (Optional) Add some extra info on cover page of peer review papers:
-% \ifCLASSOPTIONpeerreview
-% \begin{center} \bfseries EDICS Category: 3-BBND \end{center}
-% \fi
-
-% Insert page break and insert second title (peer review mode)
-\IEEEpeerreviewmaketitle
-
-\maketitle
-
-\section{Introduction}
-
-One of the major shortcomings of autonomous underwater vehicles (AUVs) is their lack of curiosity.
+% \end{abstract}
+% 
+% % Keywords (normally not used for peer reviews)
+% \ifPeerReview\else
+% \begin{IEEEkeywords}
+% Sonar, SAS, simulator, template matching, OpenGL, OpenCL.
+% \end{IEEEkeywords}
+% \fi}
+% % \fi
+% 
+% % make the title area
+% \maketitle
+% 
+% % This command fixes abstract positioning for compsoc articles:
+% \IEEEdisplaynotcompsoctitleabstractindextext
+% 
+% % (Optional) Add some extra info on cover page of peer review papers:
+% % \ifCLASSOPTIONpeerreview
+% % \begin{center} \bfseries EDICS Category: 3-BBND \end{center}
+% % \fi
+% 
+% % Insert page break and insert second title (peer review mode)
+% \IEEEpeerreviewmaketitle
+% 
+% \maketitle
+%
+% \section{Introduction}
+% 
+% 
+% One of the major shortcomings of autonomous underwater vehicles (AUVs) is their lack of curiosity.
 When encountering an interesting object, an AUV with a mounted sonar imaging system should be able to detect it, and perhaps move in for closer examination.
 Besides the apparent challenge of trying to detect something with a long list of unknown parameters, this must happen in real-time and be very power efficient.
-
-% Automatic target recognition (ATR) is an important component in autonomous underwater vehicles (AUVs), as it allows the vehicle to adapt its mission plan to e.g.
+% 
+% 
+% % Automatic target recognition (ATR) is an important component in autonomous underwater vehicles (AUVs), as it allows the vehicle to adapt its mission plan to e.g.
 revisit detected objects for closer examination.
@@
 The principle involved  is to isolate an image segment containing the object of interest, compare it with a set of template images of the relevant object classes, and assign it to the class of the template with the the best fit.
-
-The detection process is well described in the field of automatic target recognition (ATR).
+% 
+% The detection process is well described in the field of automatic target recognition (ATR).
 The classical approach involves isolating an object of interest, extracting parameters from it and feeding those into a classifier.
@@
 This limits its use to only a few parameters being coarsely sampled, ultimately yielding inaccurate results~\cite{Midelfart2010}.
-
-We have developed a synthetic aperture sonar (SAS) simulator that avoids this problem by running sufficiently fast to create templates adapted to the actual scene as part of the classification process.
+% 
+% We have developed a synthetic aperture sonar (SAS) simulator that avoids this problem by running sufficiently fast to create templates adapted to the actual scene as part of the classification process.
 This is achieved with the aid of the massive computing power in graphics processing units (GPUs) and optimized software libraries for scene rendering.
@@
 For maximum flexibility this data is finally combined with OpenCL to produce the image templates.
-
-The alternative of using of a static, predetermined template library has a computational complexity that is exponential with the number of parameters.
+% 
+% The alternative of using of a static, predetermined template library has a computational complexity that is exponential with the number of parameters.
 This limits its use to only a few parameters being coarsely sampled, ultimately yielding inaccurate results~\cite{Midelfart2010}.
-
-To avoid this problem we have developed a very fast side-looking sonar simulator that generates templates adapted to the actual scene.
-
-synthetic aperture sonar (SAS)
-
- as part of the classification process. This is achieved with the aid of the massive computing power in graphics processing units (GPUs) and optimized software libraries for scene rendering.
-
- The simulator loads a 3D model of the seafloor and an object class model into OpenGL, where emitted sound waves are modeled with a cylindrical light source placed at the sonar transmit location.
-
-We assume rough, isotropic surfaces reflecting sound energy equally in all directions.
+% 
+% To avoid this problem we have developed a very fast side-looking sonar simulator that generates templates adapted to the actual scene. 
+% 
+% 
+% synthetic aperture sonar (SAS)
+% 
+% 
+%  as part of the classification process. This is achieved with the aid of the massive computing power in graphics processing units (GPUs) and optimized software libraries for scene rendering.
+% 
+% 
+% 
+%  The simulator loads a 3D model of the seafloor and an object class model into OpenGL, where emitted sound waves are modeled with a cylindrical light source placed at the sonar transmit location. 
+% 
+% We assume rough, isotropic surfaces reflecting sound energy equally in all directions.
 This can be modeled with a Lambertian scattering model~\cite{Blake1993,Bell1995}, where the intensity of the backscattered sound solely depends on the incidence angle of the transmitted signal onto the model surface.
@@
 For maximum flexibility this data is finally combined with OpenCL to produce the image templates.
-
-Various simulators for high frequency, side-looking sonar imagery have been published, e.g.~\cite{Bell1997,Sammelm2003}.
+% 
+% Various simulators for high frequency, side-looking sonar imagery have been published, e.g.~\cite{Bell1997,Sammelm2003}.
 However, most implementations have prioritized accurate acoustic modeling at the expense of execution speed.
@@
 Also, their simulator superimpose images obtained from multiple OpenGL rendering passes to create the sonar image, while we render once and post-process this result with OpenCL to tailor the sonar image with greater flexibility.
-
-\begin{itemize}
-\item SWAT (Shallow Water Acoustic Toolset): 
-\item Bell's thesis:
-\item SIGMA:
-\item OpenGL does not support scattered writes. OpenCL does.
-\item 2D template approximation:
-\item Geometric distortions in SSS \cite{Cobra1992} Bell
-\item Sidescan sonar and synthetic aperture sonar. In seafloor coordinates. Maps to orthogonal rendering.
-\item Layover causes ambiguity, different seafloor phenomena can cause the same output image.
+% 
+% \begin{itemize}
+% \item SWAT (Shallow Water Acoustic Toolset): 
+% \item Bell's thesis:
+% \item SIGMA:
+% \item OpenGL does not support scattered writes. OpenCL does.
+% \item 2D template approximation:
+% \item Geometric distortions in SSS \cite{Cobra1992} Bell
+% \item Sidescan sonar and synthetic aperture sonar. In seafloor coordinates. Maps to orthogonal rendering.
+% \item Layover causes ambiguity, different seafloor phenomena can cause the same output image.
 Lots of cornercases when approximating the model in 2D.
@@
 Less parameters to adjust.
-\item Viewer can be used to train an operator for finding non-trivial objects on the seafloor. Input for automated target recognition standard.
-\item Improved image quality of selected object, reduced data collection cost (price estimate?).
-\item OpenGL rendering can have infinite resolution, only limited by the pixel density in the image.
-\item Accuracy in terms of classification result. Need to embody static effects in a deterministic way. Incoherent synthesis of the image.
-\item Sound speed assumed constant.
-\item High frequency lead to little (define) penetration, lower range and higher resolution. Often modeled accurately with ray tracing techniques.
-\end{itemize}
-
-Model
-\begin{itemize}
-\item Ray solution: Separate amplitude and phase components of the Helmholtz eq.
+% \item Viewer can be used to train an operator for finding non-trivial objects on the seafloor. Input for automated target recognition standard.
+% \item Improved image quality of selected object, reduced data collection cost (price estimate?).
+% \item OpenGL rendering can have infinite resolution, only limited by the pixel density in the image.
+% \item Accuracy in terms of classification result. Need to embody static effects in a deterministic way. Incoherent synthesis of the image.
+% \item Sound speed assumed constant.
+% \item High frequency lead to little (define) penetration, lower range and higher resolution. Often modeled accurately with ray tracing techniques.
+% \end{itemize}
+% 
+% Model
+% \begin{itemize}
+% \item Ray solution: Separate amplitude and phase components of the Helmholtz eq.
 by applying geometric optics approximation.
@@
 Can be improved upon with fuzzy beam modeling / Gaussian beam tracing.
-\begin{itemize}
-\item Pro: Easy to compute. Visual result. Handles complex bathymetry. Array and model response can be accounted for.
-\item Con: Only works for high frequency and small sound variations. No diffraction, caustics and focus. Infinite energy caustics/sharp variations. Incoherent(?)
-\end{itemize}
-\item Normal modes: Separate amplitude and phase into a range dependent term, and a depth dependent term.
+% \begin{itemize}
+% \item Pro: Easy to compute. Visual result. Handles complex bathymetry. Array and model response can be accounted for.
+% \item Con: Only works for high frequency and small sound variations. No diffraction, caustics and focus. Infinite energy caustics/sharp variations. Incoherent(?)
+% \end{itemize}
+% \item Normal modes: Separate amplitude and phase into a range dependent term, and a depth dependent term.
 Assumes minimal range dependence in sound speed profile and depth.
@@
 can be calculated
-\begin{itemize}
-\item Pro: Ok to compute. Handles penetration. Low frequency, shallow water.
-\item Con: Horizontal stratification. Far field only. 
-\end{itemize}
-\item Parabolic equation: One way propagation.
+% \begin{itemize}
+% \item Pro: Ok to compute. Handles penetration. Low frequency, shallow water.
+% \item Con: Horizontal stratification. Far field only. 
+% \end{itemize}
+% \item Parabolic equation: One way propagation.
 Marching solution.
@@
 conical mounts)
-\item Fast field programs / Green's function solutions.
+% \item Fast field programs / Green's function solutions.
 Exact full wave solution for acoustic fields in horizontally stratified media.
@@
 Similar to normal mode techniques.
-\item Finite element: Scene segmented into triangles or rectangles with a size of one tenth of a wavelength, or less.
+% \item Finite element: Scene segmented into triangles or rectangles with a size of one tenth of a wavelength, or less.
 At each node the wavefield can be determined from the wave equation.
 Computationally limited to low frequencies.
-\item 2D model: Acoustic field in range and depth.
+% \item 2D model: Acoustic field in range and depth.
 Assumes that a wave transmitted in a vertical plane remains in it.
 Reasonable when the sound speed can be assumed constant in the horizontal plane, and when the seafloor normals are in the vertical range-depth plane (avoid horizontal refraction, or actually reflection).
-\end{itemize}
-
-What's new?
-\begin{itemize}
-\item Tuned for speed using GPUs, combining OpenGL and OpenCL to take advantage of existing functionality for computer graphics with the added flexibility of general purpose programming on the GPUs.
-\item Performance to spare: Game-like viewer for the visualization and verification of the rendered scene and corresponding sonar image.
-\item First simulator of its kinds aimed towards improved ATR and ultimately better AUVs.
-\item 3D models are still novel. 
-\end{itemize}
-
-% The ATR algorithms must, however, be sufficiently fast to process the sensor data onboard the AUV in delayed real-time.
-
-% We have developed a simulator for synthetic aperture sonar (SAS) that creates object templates very quickly with the aid of the massive computing power in graphics processing units (GPUs).
+% \end{itemize}
+% 
+% 
+% What's new?
+% \begin{itemize}
+% \item Tuned for speed using GPUs, combining OpenGL and OpenCL to take advantage of existing functionality for computer graphics with the added flexibility of general purpose programming on the GPUs.
+% \item Performance to spare: Game-like viewer for the visualization and verification of the rendered scene and corresponding sonar image.
+% \item First simulator of its kinds aimed towards improved ATR and ultimately better AUVs.
+% \item 3D models are still novel. 
+% \end{itemize}
+% 
+% 
+% 
+% 
+% % The ATR algorithms must, however, be sufficiently fast to process the sensor data onboard the AUV in delayed real-time.
+% 
+% % We have developed a simulator for synthetic aperture sonar (SAS) that creates object templates very quickly with the aid of the massive computing power in graphics processing units (GPUs).
 It loads a 3D model of the sea floor and an object of interest into OpenGL, where emitted sound waves are modeled with a light source placed at the sonar transmit location.
@@
 For maximum flexibility this data is finally combined with OpenCL to produce the image templates.
-% 
-% By coincidence our implementation is not so different from the SIGMAS+ simulator developed at the NATO Undersea Research Centre (NURC)~\cite{Coiras2009a, Coiras2009b}.
+% % 
+% % By coincidence our implementation is not so different from the SIGMAS+ simulator developed at the NATO Undersea Research Centre (NURC)~\cite{Coiras2009a, Coiras2009b}.
 They both take advantage of OpenGL to render a 2D image from 3D models, are designed for side-looking sonar, and assume a Lambertian scattering model.
@@
 Also, their simulator sums images obtained from multiple OpenGL rendering passes to create the sonar image, while we render once and post-process this result with OpenCL to tailor the sonar image full flexibility.
-
-\begin{figure}[t]\centering%
-\graphicsAI[drawing,width=\linewidth]{gfx/specs.svg}%
-\caption{Simulator specifications.}\label{specs}%
-\end{figure}
-
-\begin{figure*}[t]\centering%
-\graphicsAI[drawing,width=\linewidth]{gfx/simulator.svg}%
-\caption{Simulator concept.}\label{buildup}%
-\end{figure*}
-
-\section{Methods}
-
-Creating a 2D view of an arbitrary complex 3D scene is a non-trivial matter.
+% 
+% 
+% \begin{figure}[t]\centering%
+% \graphicsAI[drawing,width=\linewidth]{gfx/specs.svg}%
+% \caption{Simulator specifications.}\label{specs}%
+% \end{figure}
+% 
+% \begin{figure*}[t]\centering%
+% \graphicsAI[drawing,width=\linewidth]{gfx/simulator.svg}%
+% \caption{Simulator concept.}\label{buildup}%
+% \end{figure*}
+% 
+% \section{Methods}
+% 
+% Creating a 2D view of an arbitrary complex 3D scene is a non-trivial matter.
 This is why we decided to use the core OpenGL pipeline to do this for us.
@@
 This section explains how we set up OpenGL for this task, and then proceeds to describe how we post process the OpenGL images with OpenCL to form the final sonar image.
-
-\begin{itemize}
-\item Load arbitrary complex models
-\item Simulate sonar with OpenGL
-\item Speed considerations
-\end{itemize}
-
-\subsection{Setting up the scene}
-
-Before we can render anything with OpenGL we first have to set up the scene. This involves loading the models and then placing them where we want them with the proper orientation.
-
-Our models are stored in regular 3D model files.
+% 
+% \begin{itemize}
+% \item Load arbitrary complex models
+% \item Simulate sonar with OpenGL
+% \item Speed considerations
+% \end{itemize}
+% 
+% 
+% 
+% \subsection{Setting up the scene}
+% 
+% Before we can render anything with OpenGL we first have to set up the scene. This involves loading the models and then placing them where we want them with the proper orientation. 
+% 
+% Our models are stored in regular 3D model files.
 The simulator loads these with the Open Asset Import library (\texttt{assimp}).
@@
 It converts the model into a node tree where each node contains data such as vertices, facets and textures, all formatted in an OpenGL friendly way.
-
-The next step is to scale, orient and position the objects in the scene, and finally project this scene onto an image plane.
+% 
+% The next step is to scale, orient and position the objects in the scene, and finally project this scene onto an image plane.
 This is achieved by applying a set of transformations to each vertex in the model.
 In OpenGL, the set of transformations can be formalized as
-\begin{align*}
-\bmat{x\\y\\0\\1}_\text{image} = \P \cdot \V \cdot \M \cdot \bmat{x\\y\\z\\1}_\text{model},
-\end{align*}
-where $\bmat{\cdot}_\text{model}$ is the loaded model, $\bmat{\cdot}_\text{image}$ is the output image, and $\M$, $\V$ and $\P$ are transformation matrices. These will be described next.
-
-The first transform we apply is the model matrix $\M$, which operates in the model's local coordinate system. It is defined as
-\begin{align}
-\bmat{x_t \\ y_t \\ z_t \\ w}_\text{\parbox{1cm}{\setlength\baselineskip{0.3cm}transformed\\model}} &= \boldsymbol{T} \cdot \R \cdot \boldsymbol{S} \bmat{x \\ y \\ z \\ w}_\text{model},
-\end{align}
-where $\boldsymbol{S}, \R$ and $\boldsymbol{T}$ are matrices\todo{be explicit on dependencies?} for scaling, rotation and translation, respectively.
+% \begin{align*}
+% \bmat{x\\y\\0\\1}_\text{image} = \P \cdot \V \cdot \M \cdot \bmat{x\\y\\z\\1}_\text{model},
+% \end{align*}
+% where $\bmat{\cdot}_\text{model}$ is the loaded model, $\bmat{\cdot}_\text{image}$ is the output image, and $\M$, $\V$ and $\P$ are transformation matrices. These will be described next.
+% 
+% The first transform we apply is the model matrix $\M$, which operates in the model's local coordinate system. It is defined as
+% \begin{align}
+% \bmat{x_t \\ y_t \\ z_t \\ w}_\text{\parbox{1cm}{\setlength\baselineskip{0.3cm}transformed\\model}} &= \boldsymbol{T} \cdot \R \cdot \boldsymbol{S} \bmat{x \\ y \\ z \\ w}_\text{model},
+% \end{align}
+% where $\boldsymbol{S}, \R$ and $\boldsymbol{T}$ are matrices\todo{be explicit on dependencies?} for scaling, rotation and translation, respectively.
 These are provided for reference in Appendix \ref{transformation_matrices}.
-
-Next we map the model into world coordinates using the view matrix $\V$.
+% 
+% Next we map the model into world coordinates using the view matrix $\V$.
 This involves aligning and positioning the model into the view of the camera.
@@
 The transformation take the form of
-\begin{align}
-\V &= \mat{T}(x_\text{c}, y_\text{c}, z_\text{c}) \cdot \R_\text{x}(\phi_x, 0, 0) \nonumber\\
-&= 
-\bmat{
-1  &  0  &  0  &  x_\text{c} \\
-0  &  1  &  0  &  y_\text{c} \\
-0  &  0  &  1  &  z_\text{c} \\
-0  &  0  &  0  &  1 \\
-} \cdot
-\bmat{
-1  &  0           &  0           &  0 \\
-0  &  \cos\phi_x  &  -\sin\phi_x &  0 \\
-0  &  \sin\phi_x  &  \cos\phi_x  &  0 \\
-0  &  0           &  0           &  1 \\
-},
-\end{align}
-where $(x_\text{c}, y_\text{c}, z_\text{c})$ is the image center coordinates for along track, cross track, and depth, respectively, and $\Phi_x$ is the elevation angle\todo{is this correct?} of the sonar.
+% \begin{align}
+% \V &= \mat{T}(x_\text{c}, y_\text{c}, z_\text{c}) \cdot \R_\text{x}(\phi_x, 0, 0) \nonumber\\
+% &= 
+% \bmat{
+% 1  &  0  &  0  &  x_\text{c} \\
+% 0  &  1  &  0  &  y_\text{c} \\
+% 0  &  0  &  1  &  z_\text{c} \\
+% 0  &  0  &  0  &  1 \\
+% } \cdot
+% \bmat{
+% 1  &  0           &  0           &  0 \\
+% 0  &  \cos\phi_x  &  -\sin\phi_x &  0 \\
+% 0  &  \sin\phi_x  &  \cos\phi_x  &  0 \\
+% 0  &  0           &  0           &  1 \\
+% },
+% \end{align}
+% where $(x_\text{c}, y_\text{c}, z_\text{c})$ is the image center coordinates for along track, cross track, and depth, respectively, and $\Phi_x$ is the elevation angle\todo{is this correct?} of the sonar.
 It is given as
-\begin{align*}
-\phi_x &= \arctan\left(\frac{y_\text{mean}}{-z_\text{mean}}\right).
-\end{align*}\todo{use absolute sign?}
-
-Finally the relevant part of the scene is projected onto the image plane with a projection matrix $\P$.
+% \begin{align*}
+% \phi_x &= \arctan\left(\frac{y_\text{mean}}{-z_\text{mean}}\right).
+% \end{align*}\todo{use absolute sign?}
+% 
+% Finally the relevant part of the scene is projected onto the image plane with a projection matrix $\P$.
 There are two main types of projection, perspective and orthographic.
@@
 The matrix that gives orthographic projection is defined as
-\begin{align}
-\P = \left(\begin{matrix}
-\frac{2}{r-l}  &  0              &  0              &  -\frac{r+l}{r-l} \\
-0              &  \frac{2}{t-b}  &  0              &  -\frac{t+b}{t-b} \\
-0              &  0              &  \frac{-2}{f-n} &  -\frac{f+n}{f-n} \\
-0              &  0              &  0              &  1
-\end{matrix}\right)
-\end{align}
-where $l,r,b$ and $t$ are the left, right, bottom and top boundaries of the projection plane, respectively, and where $n$ and $f$ is the near and far limits of the cube that encompass the scene.
+% \begin{align}
+% \P = \left(\begin{matrix}
+% \frac{2}{r-l}  &  0              &  0              &  -\frac{r+l}{r-l} \\
+% 0              &  \frac{2}{t-b}  &  0              &  -\frac{t+b}{t-b} \\
+% 0              &  0              &  \frac{-2}{f-n} &  -\frac{f+n}{f-n} \\
+% 0              &  0              &  0              &  1
+% \end{matrix}\right)
+% \end{align}
+% where $l,r,b$ and $t$ are the left, right, bottom and top boundaries of the projection plane, respectively, and where $n$ and $f$ is the near and far limits of the cube that encompass the scene.
 These parameters need to be adapted to provide a 2D image that matches the image size, which we specify in along track and cross track coordinates ($x,y$).
 This is straightforward for left and right since matches the $x$-axis
-% They define the window through which we observe our scene.
-\begin{align}
-l &= x_\text{min}  &  r &= x_\text{max}
-\end{align}
-For the near and far limits one simply needs to make sure the scene will fit in it
-\begin{align}
-n &= 0\;\text{m}   &  f &= 300\;\text{m} 
-\end{align}
-How to set the bottom and top bounaries can be inferred from \Fig{camera_top_bottom}. First we define an origin vector that points to the center of the image
-\begin{align}
-\vec o &= \left[\frac{x_\text{min}+x_\text{max}}{2}, \frac{y_\text{min}+y_\text{max}}{2}, \frac{z_\text{min}+z_\text{max}}{2} \right]\T,
-\end{align}
-and the distance from the image origin to its $y$-boundaries is given as
-\begin{align}
-\Delta y &= \frac{y_\text{max} - y_\text{min}}{2}.
-\end{align}
-To find the nearest and farthest point in the image that will be part of the projected scene, $\alpha$ and $\beta$, we must consider the seafloor rotation angle around the $x$-axis, $\Phi_x$ 
-\begin{align}
-\alpha &= \vec 0 - \left[0, \Delta y, \Delta y \times \arctan\Phi_\text{x} \right]
-\end{align}
-
-\begin{figure}[t]\centering%
-\graphicsAI[drawing,width=\linewidth]{gfx/opengl_part.svg}%
-\caption{Setting camera top and bottom boundaries.}\label{camera_top_bottom}%
-\end{figure}
-
-\subsection{OpenGL rendering}
-
-To produce the sonar templates we assume a rough, isotropic surface that reflects energy equally in all directions.
+% % They define the window through which we observe our scene.
+% \begin{align}
+% l &= x_\text{min}  &  r &= x_\text{max}
+% \end{align}
+% For the near and far limits one simply needs to make sure the scene will fit in it
+% \begin{align}
+% n &= 0\;\text{m}   &  f &= 300\;\text{m} 
+% \end{align}
+% How to set the bottom and top bounaries can be inferred from \Fig{camera_top_bottom}. First we define an origin vector that points to the center of the image
+% \begin{align}
+% \vec o &= \left[\frac{x_\text{min}+x_\text{max}}{2}, \frac{y_\text{min}+y_\text{max}}{2}, \frac{z_\text{min}+z_\text{max}}{2} \right]\T,
+% \end{align}
+% and the distance from the image origin to its $y$-boundaries is given as
+% \begin{align}
+% \Delta y &= \frac{y_\text{max} - y_\text{min}}{2}.
+% \end{align}
+% To find the nearest and farthest point in the image that will be part of the projected scene, $\alpha$ and $\beta$, we must consider the seafloor rotation angle around the $x$-axis, $\Phi_x$ 
+% \begin{align}
+% \alpha &= \vec 0 - \left[0, \Delta y, \Delta y \times \arctan\Phi_\text{x} \right]
+% \end{align}
+% 
+% \begin{figure}[t]\centering%
+% \graphicsAI[drawing,width=\linewidth]{gfx/opengl_part.svg}%
+% \caption{Setting camera top and bottom boundaries.}\label{camera_top_bottom}%
+% \end{figure}
+% 
+% 
+% 
+% \subsection{OpenGL rendering}
+% 
+% To produce the sonar templates we assume a rough, isotropic surface that reflects energy equally in all directions.
 This permits us to use a Lambertian scattering model where the backscatter intensity depends only on the incidence angle~\cite{Zhang1999}.
 It does not consider observation angle or sound frequency, but for the purpose of creating templates this is not needed.
-
-The rendered image will appear as if we placed a window at the sonar and looked through it in the direction of the image.
+% 
+% The rendered image will appear as if we placed a window at the sonar and looked through it in the direction of the image.
 This window is resized to make sure that the template image perfectly fills it.
-
-\subsection{OpenCL post-processing}
-
-The ``camera image'' rendered with OpenGL is not in along-track and cross-track coordinates as we want it to be.
+% 
+% \subsection{OpenCL post-processing}
+% 
+% The ``camera image'' rendered with OpenGL is not in along-track and cross-track coordinates as we want it to be.
 It does, however, show us what parts of the scene that are visible from the sonar.
@@
 This way we keep all the calculations on the GPU.
-
-\begin{itemize}
-\item Kernel tuning.
-\item Sharing. PBO. Asynchronous.
-\end{itemize}
-
-% It is also possible to compute the sonar image using the OpenGL ``blend'' feature.
-
-\subsection{Speed considerations}
-
-Using OpenGL and OpenGL to generate sonar templates on the GPU allow hundreds of sonar templates to be formed per second.
+% 
+% \begin{itemize}
+% \item Kernel tuning.
+% \item Sharing. PBO. Asynchronous.
+% \end{itemize}
+% 
+% 
+% % It is also possible to compute the sonar image using the OpenGL ``blend'' feature.
+% 
+% \subsection{Speed considerations}
+% 
+% Using OpenGL and OpenGL to generate sonar templates on the GPU allow hundreds of sonar templates to be formed per second.
 While this is sufficient for our needs.
@@
 Therefore, to alleviate the simulation performance we either had to supply it with a list of parameters or ensure that it did not have to be shut down for every invocation.
-
-The rendering loop that uses OpenGL and OpenCL to form
-
-\begin{itemize}
-\item Write to file
-\end{itemize}
-
-\section{Results \& Discussion}
-
-\newlength\imgspacing\setlength\imgspacing{.5cm}
-\setcounter{topnumber}{2}
-
-What's going on?
-
-\begin{figure*}[h]\centering%
-\subfloat[SAS image of a cylinder with 2.6\;m length and 0.53\;m radius.]{%
-\graphicsAI[drawing,width=0.5\linewidth]{gfx/data_cylinder_submerged.svg}%
-\label{data_cylinder}}%
-\subfloat[Cylinder simulation with parameters estimated from SAS image: Length 2.6\;m, burial depth 0.263\;m, aspect angle 105$^\circ$.]{
-\graphicsAI[drawing,width=0.5\linewidth]{gfx/sim_cylinder_submerged_adaptive.svg}%
-\label{sim_cylinder}}%
-\caption{A SAS image of a cylinder and a template simulation adapted to it.}
-\end{figure*}
-
-\newcommand\cdesc[2]{{\raggedright\setlength\fboxsep{0pt}
-\fbox{\colorbox[HTML]{#1}{\vrule height8.5pt depth3.5pt width0pt\hspace{.5cm}}}\ \ #2\\}}
-
-BS!!!
-
-\begin{figure*}[!tb]\centering%
-
-\subfloat[Best template out of a predefined set. Inaccurate cylinder length, alignment and immersion depth. The segmented shadow and highlight regions are misaligned.\label{overlay_cylinder}]{
-\graphicsAI[drawing,width=.5\linewidth]{gfx/overlay_cylinder_submerged.svg}%
-}%
-% % \hfil%
-\subfloat[Adapted template. Template simulated with adapted parameters. Good fit in both highlight and shadow segments.\label{overlay_cylinder_adaptive}]{%
-\graphicsAI[width=.5\linewidth]{gfx/overlay_cylinder_submerged_adaptive.svg}%
-}%
-\subfloat[][...]{hmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmm}
-% \parbox{1cm}{HMMMMMMMMMMMMMMMMMMMMMMMMMMMMMMMMMMMMMMMMMMMMMMmmmmmmmmmmmmmmmmmmmm}
-% \hfill
 % 
-% % \newline
-% \raggedright
-% \mbox{\parbox{1cm}{\small\centering
-% HMmmmmmmmmmmmmmmmmmm mmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmmm
-% % Segmentation overlay color description:\\[.5\baselineskip]
-% % \begin{minipage}{.49\linewidth}
+% The rendering loop that uses OpenGL and OpenCL to form 
+% 
+% \begin{itemize}
+% \item Write to file
+% \end{itemize}
+% 
+% 
+% 
+% \section{Results \& Discussion}
+% 
+% \newlength\imgspacing\setlength\imgspacing{.5cm}
+% \setcounter{topnumber}{2}
+% 
+% % What's going on?
+% % 
+% % \begin{figure*}[h]\centering%
+% % \subfloat[SAS image of a cylinder with 2.6\;m length and 0.53\;m radius.]{%
+% % \graphicsAI[drawing,width=0.5\linewidth]{gfx/data_cylinder_submerged.svg}%
+% % \label{data_cylinder}}%
+% % \subfloat[Cylinder simulation with parameters estimated from SAS image: Length 2.6\;m, burial depth 0.263\;m, aspect angle 105$^\circ$.]{
+% % \graphicsAI[drawing,width=0.5\linewidth]{gfx/sim_cylinder_submerged_adaptive.svg}%
+% % \label{sim_cylinder}}%
+% % \caption{A SAS image of a cylinder and a template simulation adapted to it.}
+% % \end{figure*}
+% % 
+% % \newcommand\cdesc[2]{{\raggedright\setlength\fboxsep{0pt}
+% % \fbox{\colorbox[HTML]{#1}{\vrule height8.5pt depth3.5pt width0pt\hspace{.5cm}}}\ \ #2\\}}
+% 
+% % BS!!!
+% 
+% % \begin{figure*}[!tb]\centering%
+% % 
+% % \subfloat[Best template out of a predefined set. Inaccurate cylinder length, alignment and immersion depth. The segmented shadow and highlight regions are misaligned.\label{overlay_cylinder}]{
+% % \graphicsAI[drawing,width=.5\linewidth]{gfx/overlay_cylinder_submerged.svg}%
+% % }%
+% % % % \hfil%
+% % \subfloat[Adapted template. Template simulated with adapted parameters. Good fit in both highlight and shadow segments.\label{overlay_cylinder_adaptive}]{%
+% % \graphicsAI[width=.5\linewidth]{gfx/overlay_cylinder_submerged_adaptive.svg}%
+% % }
+% % \caption{Segmented simulation image overlaid the segmented SAS image. Segmentation overlay color description:
+% % % \hfill
+% % % \parbox{1cm}{\small\centering
+% % \newline\centering\newline %\\[.5\baselineskip]
+% % \begin{minipage}{.35\linewidth}
 % % \cdesc{800000}{Template highlight and image highlight}
@@
 % % \end{minipage}%
-% % \begin{minipage}{.49\linewidth}
+% % \begin{minipage}{.35\linewidth}
 % % \cdesc{000083}{Template shadow and image shadow}
@@
 % % \end{minipage}
-% }
-% \vspace{\baselineskip}
-% \caption{Segmented simulation image overlaid the segmented SAS image.}
-\end{figure*}
-
-% Whenever conventional computer graphics processing can be used to 
-% GPUs are designed to provide the best possible graphics processing performance, and there large ecosystem of tools and libraries  available 
-% Using a GPU to drive a simulator is advantageous since
+% % }
+% % % \vspace{\baselineskip}
+% % % \caption{Segmented simulation image overlaid the segmented SAS image.}
+% % \end{figure*}
 % 
+% % Whenever conventional computer graphics processing can be used to 
+% % GPUs are designed to provide the best possible graphics processing performance, and there large ecosystem of tools and libraries  available 
+% % Using a GPU to drive a simulator is advantageous since
+% % 
+% % 
+% % Creating a GPU-based simulator is attractive since computer is an attractive solution as there are a  graphics traits. For one over a CPU-based standard simulator is two-fold: 
 % 
-% Creating a GPU-based simulator is attractive since computer is an attractive solution as there are a  graphics traits. For one over a CPU-based standard simulator is two-fold:
-
-Our GPU-based simulator is very fast.
+% Our GPU-based simulator is very fast.
 On a computer with a six-core Intel Core i7-3930K and a Radeon HD7970 it computes almost 1000 templates per second, each being 1 megapixel large.
@@
 Hence, it may happen that no template in the library is a close fit of the object in the image even though the object actually is a target of interest.
-
-This performance was demonstrated on an image from the HISAS 1030 synthetic aperture sonar mounted on the HUGIN AUV.
+% 
+% This performance was demonstrated on an image from the HISAS 1030 synthetic aperture sonar mounted on the HUGIN AUV.
 Out of the objects in the scene we selected a 2.6\;m long cylinder that was partly buried in the sea sediments (Fig.
@@
 We believe these assumptions to be typical for a template library for cylinder mines.
-
-The standard and adaptive templates were then matched to the HISAS image.
+% 
+% The standard and adaptive templates were then matched to the HISAS image.
 This is illustrated in Fig.
@@
 Hence, the standard approach was less likely to classify the cylinder in the image correctly as the standard templates were not created specifically for the target.
-
-\begin{itemize}
-\item GPUs low power. Good for integration in AUVs.
-\item Good accuracy even with a real-time constraint in e.g. a AUV.
-\item Improved inertial navigation?
-\item Aid in understanding the key parts of the SAS image and in discovering potential flaws in the SAS image. Anomalies can be flagged to make SAS more robust?
-\end{itemize}
-
-\section{Conclusion}\label{conclusion}
-
-One way to automatically classify objects in SAS images is to compare the imaged objects with a predefined set of templates.
+% 
+% % \begin{itemize}
+% % \item GPUs low power. Good for integration in AUVs.
+% % \item Good accuracy even with a real-time constraint in e.g. a AUV.
+% % \item Improved inertial navigation?
+% % \item Aid in understanding the key parts of the SAS image and in discovering potential flaws in the SAS image. Anomalies can be flagged to make SAS more robust?
+% % \end{itemize}
+% 
+% 
+% \section{Conclusion}\label{conclusion}
+% 
+% One way to automatically classify objects in SAS images is to compare the imaged objects with a predefined set of templates.
 However, this is suboptimal as it is infeasible to create accurate templates for all the relevant configurations of seabeds and objects.
@@
 In our studies this improves the match between the SAS objects and their corresponding template significantly in most cases.
-
-To obtain the delayed real-time performance we implemented the simulator on a GPU.
+% 
+% To obtain the delayed real-time performance we implemented the simulator on a GPU.
 These devices have a theoretical peak performance that is typically an order of magnitude higher than CPUs in a comparable price range.
@@
 Our simulator use this framework for most of the scene processing.
-
-A final post-processing step is performed in OpenCL, which allow general purpose GPU programming.
+% 
+% A final post-processing step is performed in OpenCL, which allow general purpose GPU programming.
 It interoperates well with OpenGL and adds a lot of flexibility to the simulation process.
 This is valuable as the simulator is still being actively developed and will likely see new features that can not be easily implemented in OpenGL alone.
-
-\begin{itemize}
-\item Motion compensation
-\item 
-\end{itemize}
-
-\appendix
-
-\section{Transformation matrices}\label{transformation_matrices} 
-We place, resize and orient objects in an euclidean 3D space using suitable transformation matrices that are applied to each of the object's vertices:
-\begin{align}
-\bmat{x_t \\ y_t \\ z_t \\ w}_\text{\parbox{1cm}{\setlength\baselineskip{0.3cm}transformed\\model}} &= \boldsymbol{T}\R\boldsymbol{S} \bmat{x \\ y \\ z \\ w}_\text{model}.
-\end{align}
-Here a model vertex at position $(x,y,z)$ is first scaled with $\boldsymbol{S}$, then rotated with $\R$
-There is nothing special to these matrices, but we supply them here for completeness.
-
-We implement rotations in a 3D euclidean space by combining matrices that rotate around each of the three respective dimensions..
+% 
+% \begin{itemize}
+% \item Motion compensation
+% \item 
+% \end{itemize}
+% 
+% 
+% \appendix
+% 
+% \section{Transformation matrices}\label{transformation_matrices} 
+% We place, resize and orient objects in an euclidean 3D space using suitable transformation matrices that are applied to each of the object's vertices:
+% \begin{align}
+% \bmat{x_t \\ y_t \\ z_t \\ w}_\text{\parbox{1cm}{\setlength\baselineskip{0.3cm}transformed\\model}} &= \boldsymbol{T}\R\boldsymbol{S} \bmat{x \\ y \\ z \\ w}_\text{model}.
+% \end{align}
+% Here a model vertex at position $(x,y,z)$ is first scaled with $\boldsymbol{S}$, then rotated with $\R$
+% There is nothing special to these matrices, but we supply them here for completeness.
+% 
+% We implement rotations in a 3D euclidean space by combining matrices that rotate around each of the three respective dimensions..
 For rotation by an angle $\phi_x$ clockwise around the $x$-axis we have
-\begin{align}
-\R_x(\phi_x) &= \bmat{
-1  &  0           &  0           &  0 \\
-0  &  \cos\phi_x  &  -\sin\phi_x &  0 \\
-0  &  \sin\phi_x  &  \cos\phi_x  &  0 \\
-0  &  0           &  0           &  1 \\
-},
-\end{align}
-and similarly, for the other two euclidean dimensions we have
-\begin{align}
-\R_y(\phi_y) &= \bmat{
-\cos\phi_y  &  0  &  \sin\phi_y  &  0 \\
-0           &  1  &  0           &  0 \\
--\sin\phi_y &  0  &  \cos\phi_y  &  0 \\
-0           &  0  &  0           &  1 \\
-},
-\end{align}
-and
-\begin{align}
-\R_z(\phi_z) &= \bmat{
-\cos\phi_z  &  -\sin\phi_z &  0  &  0 \\
-\sin\phi_z  &  \cos\phi_z  &  0  &  0 \\
-0           &  0           &  1  &  0 \\
-0           &  0           &  0  &  1 \\
-}.
-\end{align}
-The order in which these are applied matters. We rotate around the $x$-axis first, then around $y$ and finally around $z$:
-\begin{align}
-\R(\phi_x, \phi_y, \phi_z) &= \R_z(\phi_z)\;\R_y(\phi_y)\;\R_x(\phi_x).
-\end{align}
-
-\begin{align}
-\boldsymbol{T}(dx,dy,dz) &= \bmat{
-1  &  0  &  0  &  dx \\
-0  &  1  &  0  &  dy \\
-0  &  0  &  1  &  dz \\
-0  &  0  &  0  &  1 \\
-}
-\end{align}
-
-\begin{align}
-\boldsymbol{S}(s_x,s_y,s_z) &= \bmat{
-s_x  &  0    &  0    &  0 \\
-0    &  s_y  &  0    &  0 \\
-0    &  0    &  s_z  &  0 \\
-0    &  0    &  0    &  1 \\
-}
-\end{align}
+% \begin{align}
+% \R_x(\phi_x) &= \bmat{
+% 1  &  0           &  0           &  0 \\
+% 0  &  \cos\phi_x  &  -\sin\phi_x &  0 \\
+% 0  &  \sin\phi_x  &  \cos\phi_x  &  0 \\
+% 0  &  0           &  0           &  1 \\
+% },
+% \end{align}
+% and similarly, for the other two euclidean dimensions we have
+% \begin{align}
+% \R_y(\phi_y) &= \bmat{
+% \cos\phi_y  &  0  &  \sin\phi_y  &  0 \\
+% 0           &  1  &  0           &  0 \\
+% -\sin\phi_y &  0  &  \cos\phi_y  &  0 \\
+% 0           &  0  &  0           &  1 \\
+% },
+% \end{align}
+% and
+% \begin{align}
+% \R_z(\phi_z) &= \bmat{
+% \cos\phi_z  &  -\sin\phi_z &  0  &  0 \\
+% \sin\phi_z  &  \cos\phi_z  &  0  &  0 \\
+% 0           &  0           &  1  &  0 \\
+% 0           &  0           &  0  &  1 \\
+% }.
+% \end{align}
+% The order in which these are applied matters. We rotate around the $x$-axis first, then around $y$ and finally around $z$:
+% \begin{align}
+% \R(\phi_x, \phi_y, \phi_z) &= \R_z(\phi_z)\;\R_y(\phi_y)\;\R_x(\phi_x).
+% \end{align}
+% 
+% \begin{align}
+% \boldsymbol{T}(dx,dy,dz) &= \bmat{
+% 1  &  0  &  0  &  dx \\
+% 0  &  1  &  0  &  dy \\
+% 0  &  0  &  1  &  dz \\
+% 0  &  0  &  0  &  1 \\
+% }
+% \end{align}
+% 
+% \begin{align}
+% \boldsymbol{S}(s_x,s_y,s_z) &= \bmat{
+% s_x  &  0    &  0    &  0 \\
+% 0    &  s_y  &  0    &  0 \\
+% 0    &  0    &  s_z  &  0 \\
+% 0    &  0    &  0    &  1 \\
+% }
+% \end{align}
 

### 2015-10-07 Simulator/article.tex
la2pdf commit - 2015:10:07	09:57:31	(Wed 7.Oct)

--- 2015-04-14 Simulator/article.tex
@@
 %       papers and their use requires extra complexity and work.
-\newcommand\MYhyperrefoptions{bookmarks=true,bookmarksnumbered=true,
-pdfpagemode={UseOutlines},plainpages=false,pdfpagelabels=true,
-colorlinks=true,linkcolor={black},citecolor={black},urlcolor={black},
-pdftitle={A Fast GPU Sonar Simulator for Automatic Target Recognition},
-pdfsubject={},
-pdfauthor={Jo Inge Buskenes},
-pdfkeywords={adaptive beamforming, beamforming, complexity, sonar, active}}%
+% \newcommand\MYhyperrefoptions{bookmarks=true,bookmarksnumbered=true,
+% pdfpagemode={UseOutlines},plainpages=false,pdfpagelabels=true,
+% colorlinks=true,linkcolor={black},citecolor={black},urlcolor={black},
+% pdftitle={Optimal Window Design for the Low Complexity Adaptive Beamformer in Active Sonar Imaging},
+% pdfsubject={},
+% pdfauthor={Jo Inge Buskenes},
+% pdfkeywords={adaptive beamforming, beamforming, complexity, sonar, active}}%
 \ifCLASSINFOpdf
-   \usepackage[\MYhyperrefoptions,pdftex]{hyperref}
+%    \usepackage[\MYhyperrefoptions,pdftex]{hyperref}
+   \usepackage[pdftex,
+    final,                %Keep hyperlink stuff even when in draft mode
+    breaklinks=true,      %Break links when necessary
+    linktocpage=true,     %Enable link to page?
+    linkcolor=black,  %Colour of links to labels within document
+    citecolor=black,      %Colour of links to the biliography
+    filecolor=red,        %Colour of links to local files
+    pagecolor=black,        %Colour of links to other pages withing document
+    urlcolor=Blue,        %Colour of the links to external URLs
+    colorlinks=true,      %??
+    plainpages=false,     %Store roman/arabic numbering differently to avoid
+    bookmarksnumbered ]{hyperref}
+
 \else
@@
 
+\renewcommand\argmin{\text{argmin}}
+
 \renewcommand*\P{\mat P}
@@
 
-\usepackage{tikz}
-\usetikzlibrary{shapes,snakes}
+\newcommand\minus{\scalebox{0.75}[1.0]{$-$}}
+
+% \usepackage{tikz}
+% \usetikzlibrary{shapes,snakes}
 \usepackage{amsmath,amssymb}
-% \usepackage{datatool}
-% \usepackage{glossaries}
-
-\newenvironment{outline}
-{\begin{itemize}}
-{\end{itemize}}
 
@@
 
-\begin{figure}[t]\centering%
+\begin{figure}[tbp]\centering%
 \graphicsAI[drawing,width=\linewidth]{gfx/specs.svg}%
@@
 
-\begin{figure}[p]\centering%
+What's going on?
+
+\begin{figure*}[h]\centering%
 \subfloat[SAS image of a cylinder with 2.6\;m length and 0.53\;m radius.]{%
-\graphicsAI[drawing,width=\linewidth]{gfx/data_cylinder_submerged.svg}%
+\graphicsAI[drawing,width=0.5\linewidth]{gfx/data_cylinder_submerged.svg}%
 \label{data_cylinder}}%
 \subfloat[Cylinder simulation with parameters estimated from SAS image: Length 2.6\;m, burial depth 0.263\;m, aspect angle 105$^\circ$.]{
-\graphicsAI[drawing,width=\linewidth]{gfx/sim_cylinder_submerged_adaptive.svg}%
+\graphicsAI[drawing,width=0.5\linewidth]{gfx/sim_cylinder_submerged_adaptive.svg}%
 \label{sim_cylinder}}%
 \caption{A SAS image of a cylinder and a template simulation adapted to it.}
-\end{figure}
+\end{figure*}
 
-\newcommand\cdesc[2]{{\raggedright\setlength\fboxsep{0pt}
-\fbox{\colorbox[HTML]{#1}{\vrule height8.5pt depth3.5pt width0pt\hspace{.5cm}}}\ \ #2\\}}
+% \newcommand\cdesc[2]{{\raggedright\setlength\fboxsep{0pt}
+% \fbox{\colorbox[HTML]{#1}{\vrule height8.5pt depth3.5pt width0pt\hspace{.5cm}}}\ \ #2\\}}
 
-\begin{figure}[p]\centering%
+\newcommand\cdesc[2]{#2}
+
+BS!!!
+
+\begin{figure*}[!tb]\centering%
+
 \subfloat[Best template out of a predefined set. Inaccurate cylinder length, alignment and immersion depth. The segmented shadow and highlight regions are misaligned.]{
-\graphicsAI[drawing,width=\linewidth]{gfx/overlay_cylinder_submerged.svg}%
-\label{overlay_cylinder}}%
-% \hfil%
+\graphicsAI[drawing,width=.5\linewidth]{gfx/overlay_cylinder_submerged.svg}\label{overlay_cylinder}%
+}%
+\hfil%
 \subfloat[Adapted template. Template simulated with adapted parameters. Good fit in both highlight and shadow segments.]{%
-\graphicsAI[width=\linewidth]{gfx/overlay_cylinder_submerged_adaptive.svg}%
-\label{overlay_cylinder_adaptive}}%
-\hfill\parbox{\linewidth}{\small\centering
-Segmentation overlay color description:\\[.5\baselineskip]
-\begin{minipage}{\linewidth}
-\cdesc{800000}{Template highlight and image highlight}
-\cdesc{FF1000}{Template highlight only}
-\cdesc{FFEB00}{Image highlight only}
-\cdesc{83FF7C}{Background pixels}
-\end{minipage}%
-\begin{minipage}{\linewidth}
-\cdesc{000083}{Template shadow and image shadow}
-\cdesc{0014FF}{Template shadow only}
-\cdesc{00EFFF}{Image shadow only}
-\cdesc{00A7FF}{Template shadow and image highlight}
-\end{minipage}}
-\vspace{\baselineskip}
+\graphicsAI[width=.5\linewidth]{gfx/overlay_cylinder_submerged_adaptive.svg}%
+}\caption{Segmented simulation image overlaid the segmented SAS image. Segmentation overlay color description:\newline
+% \hfill
+% \parbox{1cm}{\small\centering
+% \newline\centering\newline %\\[.5\baselineskip]
+% \begin{minipage}{.35\linewidth}
+% \cdesc{800000}{Template highlight and image highlight}
+% \cdesc{FF1000}{Template highlight only}
+% \cdesc{FFEB00}{Image highlight only}
+% \cdesc{83FF7C}{Background pixels}
+% \end{minipage}%
+% \begin{minipage}{.35\linewidth}
+% \cdesc{000083}{Template shadow and image shadow}
+% \cdesc{0014FF}{Template shadow only}
+% \cdesc{00EFFF}{Image shadow only}
+% \cdesc{00A7FF}{Template shadow and image highlight}
+% \end{minipage}
+}\label{overlay_cylinder_adaptive}%
+% % \vspace{\baselineskip}
 \caption{Segmented simulation image overlaid the segmented SAS image.}
-\end{figure}
+\end{figure*}
 

### 2015-10-07 Simulator/article.tex
la2pdf commit - 2015:10:07	11:23:21	(Wed 7.Oct)

@@
 % pdfkeywords={adaptive beamforming, beamforming, complexity, sonar, active}}%
-\ifCLASSINFOpdf
-%    \usepackage[\MYhyperrefoptions,pdftex]{hyperref}
-   \usepackage[pdftex,
-    final,                %Keep hyperlink stuff even when in draft mode
-    breaklinks=true,      %Break links when necessary
-    linktocpage=true,     %Enable link to page?
-    linkcolor=black,  %Colour of links to labels within document
-    citecolor=black,      %Colour of links to the biliography
-    filecolor=red,        %Colour of links to local files
-    pagecolor=black,        %Colour of links to other pages withing document
-    urlcolor=Blue,        %Colour of the links to external URLs
-    colorlinks=true,      %??
-    plainpages=false,     %Store roman/arabic numbering differently to avoid
-    bookmarksnumbered ]{hyperref}
-
-\else
-   \usepackage[\MYhyperrefoptions,breaklinks=true,dvips]{hyperref}
-   \usepackage{breakurl}                        % Allows 'dvips' driver to break links
-\fi
+% \ifCLASSINFOpdf
+% %    \usepackage[\MYhyperrefoptions,pdftex]{hyperref}
+%    \usepackage[pdftex,
+%     final,                %Keep hyperlink stuff even when in draft mode
+%     breaklinks=true,      %Break links when necessary
+%     linktocpage=true,     %Enable link to page?
+%     linkcolor=black,  %Colour of links to labels within document
+%     citecolor=black,      %Colour of links to the biliography
+%     filecolor=red,        %Colour of links to local files
+%     pagecolor=black,        %Colour of links to other pages withing document
+%     urlcolor=Blue,        %Colour of the links to external URLs
+%     colorlinks=true,      %??
+%     plainpages=false,     %Store roman/arabic numbering differently to avoid
+%     bookmarksnumbered ]{hyperref}
+% 
+% \else
+%    \usepackage[\MYhyperrefoptions,breaklinks=true,dvips]{hyperref}
+%    \usepackage{breakurl}                        % Allows 'dvips' driver to break links
+% \fi
 

### 2015-10-07 Simulator/article.tex
la2pdf commit - 2015:10:07	11:53:29	(Wed 7.Oct)

--- 2015-10-07 Simulator/article.tex
@@
 %       papers and their use requires extra complexity and work.
-% \newcommand\MYhyperrefoptions{bookmarks=true,bookmarksnumbered=true,
-% pdfpagemode={UseOutlines},plainpages=false,pdfpagelabels=true,
-% colorlinks=true,linkcolor={black},citecolor={black},urlcolor={black},
-% pdftitle={Optimal Window Design for the Low Complexity Adaptive Beamformer in Active Sonar Imaging},
-% pdfsubject={},
-% pdfauthor={Jo Inge Buskenes},
-% pdfkeywords={adaptive beamforming, beamforming, complexity, sonar, active}}%
+\newcommand\MYhyperrefoptions{bookmarks=true,bookmarksnumbered=true,
+pdfpagemode={UseOutlines},plainpages=false,pdfpagelabels=true,
+colorlinks=true,linkcolor={black},citecolor={black},urlcolor={black},
+pdftitle={A Fast GPU Sonar Simulator for Automatic Target Recognition},
+pdfsubject={},
+pdfauthor={Jo Inge Buskenes},
+pdfkeywords={adaptive beamforming, beamforming, complexity, sonar, active}}%
 \ifCLASSINFOpdf
-%    \usepackage[\MYhyperrefoptions,pdftex]{hyperref}
-   \usepackage[pdftex,
-    final,                %Keep hyperlink stuff even when in draft mode
-    breaklinks=true,      %Break links when necessary
-    linktocpage=true,     %Enable link to page?
-    linkcolor=black,  %Colour of links to labels within document
-    citecolor=black,      %Colour of links to the biliography
-    filecolor=red,        %Colour of links to local files
-    pagecolor=black,        %Colour of links to other pages withing document
-    urlcolor=Blue,        %Colour of the links to external URLs
-    colorlinks=true,      %??
-    plainpages=false,     %Store roman/arabic numbering differently to avoid
-    bookmarksnumbered ]{hyperref}
-
+   \usepackage[\MYhyperrefoptions,pdftex]{hyperref}
 \else
@@
 
-% \usepackage[maxfloats=40]{morefloats}
+\usepackage[maxfloats=40]{morefloats}
 \newcounter{todoidx}
@@
          {\vspace*{-10pt}\color{Red}\fbox{\bf\thetodoidx{}}}\\%
-         \fcolorbox{red}{todobackground}{\parbox{\marginparwidthsmall}{\raggedright\scriptsize #1}}}}
+         \fcolorbox{red}{todobackground}{\parbox{\marginparwidthsmall}{\scriptsize #1}}}}
 
@@
 
-% \usepackage{float}
+\usepackage{float}
 
@@
 \newcommand\Fig[1]{Fig.~\ref{#1}}
-\newcommand\eq[1]{(\ref{#1})}
 
@@
 \renewcommand*\a{\vec a}
-\newcommand*\f{\vec f}
 \renewcommand*\i{\vec i}
@@
 
-\renewcommand\argmin{\text{argmin}}
-
 \renewcommand*\P{\mat P}
@@
 
-\newcommand\minus{\scalebox{0.75}[1.0]{$-$}}
+\usepackage{tikz}
+\usetikzlibrary{shapes,snakes}
+\usepackage{amsmath,amssymb}
+% \usepackage{datatool}
+% \usepackage{glossaries}
 
-% \usepackage{tikz}
-% \usetikzlibrary{shapes,snakes}
-\usepackage{amsmath,amssymb}
+\newenvironment{outline}
+{\begin{itemize}}
+{\end{itemize}}
 
@@
 
-\begin{figure}[tbp]\centering%
+\begin{figure}[t]\centering%
 \graphicsAI[drawing,width=\linewidth]{gfx/specs.svg}%
@@
 
-\begin{figure*}[h]\centering%
+\begin{figure*}[t]\centering%
 \subfloat[SAS image of a cylinder with 2.6\;m length and 0.53\;m radius.]{%
@@
 
-\begin{figure*}[!tb]\centering%
+\begin{figure*}[t]\centering%
 
@@
 \graphicsAI[width=.5\linewidth]{gfx/overlay_cylinder_submerged_adaptive.svg}%
-}\caption{Segmented simulation image overlaid the segmented SAS image. Segmentation overlay color description:\newline
-% \hfill
+}\caption[LoF entry]{Segmented simulation image overlaid the segmented SAS image. Segmentation overlay color description:\parbox{1cm}{\small\centering 1}}%
+% % \hfill
 % \parbox{1cm}{\small\centering
 % \newline\centering\newline %\\[.5\baselineskip]
-% \begin{minipage}{.35\linewidth}
+\begin{minipage}{.35\linewidth}
 % \cdesc{800000}{Template highlight and image highlight}
@@
 % \cdesc{83FF7C}{Background pixels}
-% \end{minipage}%
+\end{minipage}%
 % \begin{minipage}{.35\linewidth}
@@
 % \end{minipage}
-}\label{overlay_cylinder_adaptive}%
+% }\label{overlay_cylinder_adaptive}%
 % % \vspace{\baselineskip}

### 2015-10-08 Simulator/article.tex
la2pdf commit - 2015:10:08	10:53:43	(Thu 8.Oct)

@@
 
-\begin{figure*}[t]\centering%
+\begin{figure*}[!tbp]\centering%
 \subfloat[SAS image of a cylinder with 2.6\;m length and 0.53\;m radius.]{%
@@
 
-% \newcommand\cdesc[2]{{\raggedright\setlength\fboxsep{0pt}
-% \fbox{\colorbox[HTML]{#1}{\vrule height8.5pt depth3.5pt width0pt\hspace{.5cm}}}\ \ #2\\}}
+\newcommand\cdesc[2]{{\raggedright\setlength\fboxsep{0pt}
+\fbox{\colorbox[HTML]{#1}{\vrule height8.5pt depth3.5pt width0pt\hspace{.5cm}}}\ \ #2\\}}
 
-\newcommand\cdesc[2]{#2}
+% \newcommand\cdesc[2]{#2}
 
-BS!!!
+\begin{figure*}[!tbp]\centering%
+\subfloat[Best template out of a predefined set. Inaccurate cylinder length, alignment and immersion depth. The segmented shadow and highlight regions are misaligned.]{
+\graphicsAI[drawing,width=.48\linewidth]{gfx/overlay_cylinder_submerged.svg}\label{overlay_cylinder}%
+}%
+\hfill%
+\subfloat[Adapted template. Template simulated with adapted parameters. Good fit in both highlight and shadow segments.]{%
+\graphicsAI[width=.48\linewidth]{gfx/overlay_cylinder_submerged_adaptive.svg}%
+}\caption[LoF entry]{\raggedright Simulation image overlaid the segmented SAS image. Segmentation overlay color description:
 
-\begin{figure*}[t]\centering%
-
-\subfloat[Best template out of a predefined set. Inaccurate cylinder length, alignment and immersion depth. The segmented shadow and highlight regions are misaligned.]{
-\graphicsAI[drawing,width=.5\linewidth]{gfx/overlay_cylinder_submerged.svg}\label{overlay_cylinder}%
-}%
-\hfil%
-\subfloat[Adapted template. Template simulated with adapted parameters. Good fit in both highlight and shadow segments.]{%
-\graphicsAI[width=.5\linewidth]{gfx/overlay_cylinder_submerged_adaptive.svg}%
-}\caption[LoF entry]{Segmented simulation image overlaid the segmented SAS image. Segmentation overlay color description:\parbox{1cm}{\small\centering 1}}%
-% % \hfill
-% \parbox{1cm}{\small\centering
+\parbox{\linewidth}{\small\centering\vspace{.5\baselineskip}
 % \newline\centering\newline %\\[.5\baselineskip]
-\begin{minipage}{.35\linewidth}
-% \cdesc{800000}{Template highlight and image highlight}
-% \cdesc{FF1000}{Template highlight only}
-% \cdesc{FFEB00}{Image highlight only}
-% \cdesc{83FF7C}{Background pixels}
+\begin{minipage}{.4\linewidth}
+\cdesc{800000}{Template highlight and image highlight}
+\cdesc{FF1000}{Template highlight only}
+\cdesc{FFEB00}{Image highlight only}
+\cdesc{83FF7C}{Background pixels}
 \end{minipage}%
-% \begin{minipage}{.35\linewidth}
-% \cdesc{000083}{Template shadow and image shadow}
-% \cdesc{0014FF}{Template shadow only}
-% \cdesc{00EFFF}{Image shadow only}
-% \cdesc{00A7FF}{Template shadow and image highlight}
-% \end{minipage}
+\begin{minipage}{.4\linewidth}
+\cdesc{000083}{Template shadow and image shadow}
+\cdesc{0014FF}{Template shadow only}
+\cdesc{00EFFF}{Image shadow only}
+\cdesc{00A7FF}{Template shadow and image highlight}
+\end{minipage}
+}}
 % }\label{overlay_cylinder_adaptive}%
 % % \vspace{\baselineskip}
-\caption{Segmented simulation image overlaid the segmented SAS image.}
+% \caption{Segmented simulation image overlaid the segmented SAS image.}
 \end{figure*}

### 2015-10-09 LCAforActiveSonarImaging/lca.tex
Hmm

@@
 The LCA images in \Fig{oversampling_mosaic_bounds} demonstrate the effect of adjusting the upper bound of $\beta$ and $\phi$.
-As suggested by \Fig{mvdr_selected_windows} we attempted first to use $\beta\in[0,2]$.
-While this resulted in much improved sidelobe suppression, there was also a slight improvement up to $\beta\in[0,5]$.
-This is likely needed to compensate for the increased sidelobe levels caused by steering.
-No noticeable difference were observed for higher values than 5.
-The images also demonstrate that increasing the steering $\phi$ improves the resolution of the strong scatterers.
-For this sonar the Rayleigh resolution is $\Delta=0.0125$ radians (0.72$^\circ$), thus a point scatterer imaged at 19\,m with a rectangular DAS would have a lateral $\text{FWHM}\approx0.125\cdot19\,\text{m}\cdot0.89=21\,\text{cm}$.
+As suggested by \Fig{mvdr_selected_windows} we first attempted to use $\beta\in[0,2]$.
+While this significantly improved sidelobe suppression, a slight further improvement was observed up to $\beta\in[0,5]$.
+This can likely be explained by the need to compensate for the increased sidelobe levels caused by steering.
+No noticeable difference were observed for higher values than $\beta=5$.
+The images also demonstrate that increasing the upper bound for steering $\phi$ improves the resolution of the strong scatterers.
+For this sonar the Rayleigh resolution is $\Delta\approx0.0125$ radians (0.72$^\circ$), so a point scatterer imaged at 19\,m range with a rectangular DAS would have a lateral $\text{FWHM}\approx0.125\cdot19\,\text{m}\cdot0.89=21\,\text{cm}$.
 This is close to the FWHM of the anchor in the DAS image in \Fig{oversampling_mosaic_bounds}, which means that the acoustic fingerprint of the anchor is similar to a point source.
@@
 \Fig{oversampling_mosaic_bounds_beta} and \Fig{oversampling_mosaic_bounds_phi} illustrates which $\beta$ and $\phi$ values LCA prefers for different regions in the image of the cross.
-Red marker lines are used to pinpoint the cut line, location of the anchor and strong scatter points on the cross.
+Red marker lines are used to pinpoint the cut line, location of the anchor and strong scatters on the cross.
 Observe that in the anchor region only narrow responses ($\beta=0$) are used, which is the reason for the anchor FWHM being measured to the same value regardless of the upper bound of $\beta$ in \Fig{oversampling_mosaic_bounds}.
 On other scatterers with strong nearby lateral interference LCA prefers wider windows to suppress the interference.
-When allowed LCA always choose windows steered away from the sources, and should not be allowed to select windows that are steered more than the 3\,dB width of the window.
-We observed image oscillations when exceeding this bound.
+When allowed LCA prefers to steer windows away from the sources, but never exceed the $\phi_\text{3dB}(\beta)$ limit.
+When exceeding this bound we observed oscillation artifacts in the image.
 
@@
 
-So far we have determined it reasonable to use Kaiser windows in the range $\beta\in[0,5]$, each steered within the range $|\phi|\in[0,\phi_\mathrm{3dB}(\beta)]$.
-But how many windows do we need in these ranges?
-
-In \Fig{oversampling_mosaic} we have compared LCA images made from windows databases of different sizes, but with the same parameter boundaries.
-We included the results from using a single window ($N_\beta=1$) and steering angle ($N_\phi=1$), which corresponds to a rectangularly weighted DAS.
+So far we have found it reasonable to use Kaiser windows in the range $\beta\in[0,5]$, each steered within the range $|\phi|\in[0,\phi_\mathrm{3dB}(\beta)]$.
+What remains is to determine the number of windows needed.
+We study this in \Fig{oversampling_mosaic}, where we compared LCA images made from windows databases of different sizes, but with the same parameter boundaries.
+We included the results from using a single window ($N_\beta=1$) and steering angle ($N_\phi=1$), which corresponds to an unsteered rectangularly weighted DAS.
 
@@
 Two lateral image cuts through all images are presented in the left plot.
-The images and these cuts show that there is minimal difference to using Kaiser windows compared to trigonometric windows.
+The images and their corresponding cuts are near identical for the LCA version with Kaiser windows and the one with trigonometric windows.
 Compared to DAS the LCA produces targets with an FWHM that lies in between that of DAS and MVDR.
@@
 
-\begin{table}[!b]\centering%\normalsize
-\begin{tabular}[c]{l c c c}\hline
-\rowcolor{tabBlue} & \bf Highlight & \bf Shadow & \bf Speckle  \\\hline
-Symmetry             & Moderate  & High   & High \\
-Mainlobe width       & Narrow    & Wide   & Moderate \\
-Sidelobe suppression & Low       & High   & Moderate \\
-Steering             & Low       & Up to -3dB width & Moderate \\
-Oversampling         & 2x-8x     & 2x-8x  & 2x-8x
-\end{tabular}
-\caption{Summary}\label{tab:summary}
-\end{table}%
+% \begin{table}[!b]\centering%\normalsize
+% \begin{tabular}[c]{l c c c}\hline
+% \rowcolor{tabBlue} & \bf Highlight & \bf Shadow & \bf Speckle  \\\hline
+% Symmetry             & Moderate  & High   & High \\
+% Mainlobe width       & Narrow    & Wide   & Moderate \\
+% Sidelobe suppression & Low       & High   & Moderate \\
+% Steering             & Low       & Up to -3dB width & Moderate \\
+% Oversampling         & 2x-8x     & 2x-8x  & 2x-8x
+% \end{tabular}
+% \caption{Summary}\label{tab:summary}
+% \end{table}%
 
@@
 We suggest an upper bound for $\beta$ between 2 and 5, where 5 is slightly better at suppressing sidelobes.
-For $\phi$ we suggest to steer either to 50\% or 100\% of the -3dB width, which leads to a resolution of 73\% or 54\% of the rectangular window, respectively.
+For $\phi$ we suggest to steer either to 50\% or 100\% of the \minus{}3\,dB width, which leads to a resolution of 73\% or 54\% of the rectangular window, respectively.
 As lower boundaries we set $\beta=0$ to always include a rectangular window, and $\phi=0$ to always include the centered window.

### 2015-10-09 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:10:09	15:17:23	(Fri 9.Oct)

@@
+\synctex=1
 %%%%%%%%%%%%%%%%                                ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
@@
 The problem lies in estimating and inverting this spatial covariance matrix.
-We have described the exact steps in~\cite{Buskenes2014}: To avoid signal cancellation we apply spatial averaging by computing a mean covariance matrix from a set of subarrays with length $L$~\cite{Kailath1985}, for true speckle statistics we perform temporal averaging over $N_k = 2K+1$ temporal samples~\cite{Synnevag2009a}, and to improve robustness to parameter errors we add $d$ percent diagonal loading~\cite{Cox1987,Maksym1979}.
+We have described the exact steps in~\cite{Buskenes2014}: To avoid signal cancellation we apply spatial averaging by computing a mean covariance matrix from a set of subarrays with length $L$~\cite{Kailath1985}, for true speckle statistics we perform temporal averaging over $N_k = 2K+1$ temporal samples~\cite{Synnevag2009a}, and to improve robustness to parameter errors we add {\large$\epsilon$} percent diagonal loading~\cite{Cox1987,Maksym1979}.
 These steps are also needed to ensure that the covariance matrix is numerically well conditioned and hence invertible.
@@
 %
-Here $d$\todo{$d$ is reused as element spacing, find an alternative} is the element spacing, $\lambda$ is the wavelength and $\phi$ is the steering amount.
+Here $d$ is the element spacing, $\lambda$ is the wavelength and $\phi$ is the steering amount.
 The normalization factor $\f_\beta\H\a_\phi$ is the reciprocal of the window's coherent gain.
@@
 
-In the image reconstruction we have run MVDR with subarray length $L=16$ and $d=1\%$ diagonal loading\todo{also used for element distance}.
+In the image reconstruction we have run MVDR with subarray length $L=16$ and $\text{\large$\epsilon$}=1\%$ diagonal loading.
 Both MVDR and LCA was run with $K=1$ temporal averaging.

### 2015-10-09 ThesisJo/derivations.tex
la2pdf commit - 2015:08:17	14:36:33	(Mon17.Aug)
la2pdf commit - 2015:10:09	15:48:03	(Fri 9.Oct)
la2pdf commit - 2015:10:09	20:08:04	(Fri 9.Oct)

@@
 
+%  coarse delay adjustment is usually achieved by applying delays to the sensor channels to negate the difference in travel time from the point of interest, and cause the  thereby aligning the phase of the signals coming from that source.
+Next a set of weights, or window, is applied to the sensors to Finally a set of weights  , which in effect achieve this by applying delays and weigths to the sensor channels to coherently combine signals from the direction of interest.
+%
+
 \subsection{Tricks}
@@
 Adaptive windows perform best in detection of closely separated angles of significantly amplitudes.
+
+\section{Mathematical Proofs}
+
+\subsection[Proof d/dw\^{}H w\^{}HRw = 2Rw]{Proof: $\frac{\partial}{\partial\w\H} \w\H\R\w = 2\R\w$}
+
+\begin{align}
+\w\H\R\w &= \sumb{i=0}{N}w_i^* \sumb{j=0}{N}  R_{ij} w_j \nn
+&= \sumb{\substack{i=0\\i\ne k}}{N} w_i^* \sumb{j=0}{N} R_{ij} w_j + w_k^*\sumb{j=0}{N} R_{kj} w_j \nn
+&= \sumb{\substack{i=0\\i\ne k}}{N} w_i^* \sumb{\substack{j=0\\j\ne k}}{N} R_{ij} w_j + w_k^*\sumb{j=0}{N} R_{kj} w_j + w_k\sumb{i=0}{N}w_i^* R_{ik} \nn
+&= \sumb{\substack{i=0\\i\ne k}}{N} w_i^* \sumb{\substack{j=0\\j\ne k}}{N} R_{ij} w_j + w_k^*\sumb{j=0}{N} R_{kj} w_j + w_k^*\sumb{i=0}{N}w_i R_{ik}^* \nn
+&= \sumb{\substack{i=0\\i\ne k}}{N} w_i^* \sumb{\substack{j=0\\j\ne k}}{N} R_{ij} w_j + 2w_k^*\sumb{i=0}{N} R_{ki} w_i
+\end{align}
+%
+Partial differentiation with respect to $\w\H$ then yields:
+%
+\begin{align}
+\frac{\partial}{\partial w_k^*}\;\w\H\R\w
+&= \sumb{\substack{i=0\\i\ne k}}{N} w_i^* \sumb{j=0}{N} R_{ij} w_j + w_k^*\sumb{j=0}{N} R_{kj} w_j \nn
+&= \sumb{\substack{i=0\\i\ne k}}{N} w_i^* \sumb{j=0}{N} R_{ij} w_j + w_k^*\sumb{j=0}{N} R_{kj} w_j
+\end{align}

### 2015-10-09 ThesisJo/derivations.tex
la2pdf commit - 2015:10:09	21:07:06	(Fri 9.Oct)

--- 2015-08-09 ThesisJo/derivations.tex
@@
 Adaptive windows perform best in detection of closely separated angles of significantly amplitudes.
+
+\section{Mathematical Proofs}
+
+\subsection[Proof d/dw\^{}H w\^{}HRw = 2Rw]{Proof: $\frac{\partial}{\partial\w\H} \w\H\R\w = 2\R\w$}
+
+To prove this relation we will write the matrix operations out on index form, and 
+%
+\begin{align}
+\w\H\R\w &= \sumb{i=1}{N}w_i^* \sumb{j=1}{N}  R_{ij} w_j \nn
+&= \sumb{\substack{i=1\\i\ne k}}{N} w_i^* \sumb{j=1}{N} R_{ij} w_j + w_k^*\sumb{j=1}{N} R_{kj} w_j \nn
+&= \sumb{\substack{i=1\\i\ne k}}{N} w_i^* \sumb{\substack{j=1\\j\ne k}}{N} R_{ij} w_j + w_k^*\sumb{j=1}{N} R_{kj} w_j + w_k\sumb{i=1}{N}w_i^* R_{ik} \nn
+&= \sumb{\substack{i=1\\i\ne k}}{N} w_i^* \sumb{\substack{j=1\\j\ne k}}{N} R_{ij} w_j + w_k^*\sumb{j=1}{N} R_{kj} w_j + w_k^*\sumb{i=1}{N}w_i R_{ik}^* \nn
+&\overset{R_{ij}=R_{ji}^*}{=} \sumb{\substack{i=1\\i\ne k}}{N} w_i^* \sumb{\substack{j=1\\j\ne k}}{N} R_{ij} w_j + 2w_k^*\sumb{i=1}{N} R_{ki} w_i
+\end{align}
+%
+Partial differentiation with respect to $w_k^*$ then gives:
+%
+\begin{align}
+\frac{\partial}{\partial w_k^*}\;\w\H\R\w
+&= 2\sumb{i=1}{N} R_{kj} w_i
+\end{align}
+%
+Hence, when we apply the partial differentiation as a row operation we get:
+%
+\begin{align}
+\frac{\partial}{\partial \w^H}\;\w\H\R\w
+&= \bmat{
+\frac{\partial}{\partial w_1^*}\;\w\H\R\w \\
+\frac{\partial}{\partial w_2^*}\;\w\H\R\w \\ \vdots \\
+\frac{\partial}{\partial w_M^*}\;\w\H\R\w}
+= \bmat{
+2\sum_{i=1}^{N} R_{1j} w_j\\
+2\sum_{i=1}^{N} R_{2j} w_j\\ \vdots \\
+2\sum_{i=1}^{N} R_{Nj} w_j}
+= 2\R\w
+\end{align}

### 2015-10-09 ThesisJo/background.tex
la2pdf commit - 2015:10:09	23:43:54	(Fri 9.Oct)

@@
 %
+
+\section{Adaptive Beamforming}
+
+\subsection{MVDR}
+
+By definition, the beamformer output $z[n]$ can now be expressed as the weighted sum of all the delayed data samples:
+\begin{align}
+z[n] = \w\H[n]\x[n] = \bmat{w_0[n]\\w_1[n]\\\vdots\\w_{M-1}[n]}^H \bmat{x_0[n]\\x_1[n]\\\vdots\\x_{M-1}[n]},\label{z}
+\end{align}
+where $w_m$ is the weight factor assigned to channel $m$.
+With static weights this would be referred to as the conventional delay-and-sum (DAS) beamformer.
+A large variety of weighting functions exists here for trading lateral resolution for improved noise suppression (contrast), but one always ends up with a compromise between the two~\cite{Harris1978}.
+
+Various adaptive beamformers target this limitation by allowing the weights to change for each pixel to better fit the dynamic nature of the incoming wavefield.
+In other words, they attempt to use the \emph{a priori} information present in the data to improve image quality.
+The MVDR beamformer is one such method.
+It finds the set of complex weights that minimizes the beamformer's expected output power, while ensuring unity gain in the look direction~\cite{Capon1969}.
+This is a convex optimization problem that can be solved using Lagrange multipliers to yield the solution
+%
+\begin{gather}
+\vec w[n] = \frac{\Ri[n]\a}{\a\T\Ri[n]\a},\label{weights}
+\end{gather}
+%
+where $\a$ is a steering vector and $\R=E\{\x[n]\x\H[n]\}\in\mathbb{C}^{M,M}$ is the spatial covariance matrix for the full array.
+Since we pre-steer our data to every pixel in the image we simplify (\ref{weights}) by substituting $\a$ with a row vector $\1$ that represents broadside phase-steering.
+To estimate $\R$ we compute a sample covariance matrix $\eR$.
+In this computation we perform some degree of:
+%
+\begin{itemize}
+\item \emph{spatial averaging} to avoid signal cancellation by decorrelating coherent echoes~\cite{Kailath1985};
+\item \emph{temporal averaging} over an interval comparable to the pulse length (one to five samples) to maintain true speckle statistics~\cite{Synnevag2009a};
+\item \emph{diagonal loading} to improve robustness to parameter errors~\cite{Cox1987,Maksym1979}.
+\end{itemize}%
 

### 2015-10-11 ThesisJo/derivations.tex
la2pdf commit - 2015:10:09	23:43:54	(Fri 9.Oct)
la2pdf commit - 2015:10:11	19:15:42	(Sun11.Oct)
la2pdf commit - 2015:10:11	19:49:50	(Sun11.Oct)
la2pdf commit - 2015:10:11	19:59:19	(Sun11.Oct)

@@
 \newpage
+
+\chapter{Derivations}
 
@@
 \newpage
+
 \section{MVDR}
@@
 \begin{align}
-z[n] = \w\H[n]\x[n] = \bmat{w_0[n]\\w_1[n]\\\vdots\\w_{M-1}[n]}^H \bmat{x_0[n]\\x_1[n]\\\vdots\\x_{M-1}[n]},\label{z}
+\argmin{\w}{E\big\{||\w\H\x||^2\big\}} \qquad\text{subject to }\w\H\a = c
 \end{align}
-where $w_m$ is the weight factor assigned to channel $m$.
-With static weights this would be referred to as the conventional delay-and-sum (DAS) beamformer.
-A large variety of weighting functions exists here for trading lateral resolution for improved noise suppression (contrast), but one always ends up with a compromise between the two~\cite{Harris1978}.
-
-Various adaptive beamformers target this limitation by allowing the weights to change for each pixel to better fit the dynamic nature of the incoming wavefield.
-In other words, they attempt to use the \emph{a priori} information present in the data to improve image quality.
-The MVDR beamformer is one such method.
-It finds the set of complex weights that minimizes the beamformer's expected output power, while ensuring unity gain in the look direction~\cite{Capon1969}.
-This is a convex optimization problem that can be solved using Lagrange multipliers to yield the solution
 %
-\begin{gather}
-\vec w[n] = \frac{\Ri[n]\a}{\a\T\Ri[n]\a},\label{weights}
-\end{gather}
+\begin{align}
+E\big\{||\w\H\x||^2\big\}
+&= E\big\{\big(\w\H\x\big)\big(\w\H\x\big)^*\big\} \nn
+&= \w\H E\big\{\x\x\H\big\}\w \nn
+&= \w\H\R\w
+\end{align}
 %
-where $\a$ is a steering vector and $\R=E\{\x[n]\x\H[n]\}\in\mathbb{C}^{M,M}$ is the spatial covariance matrix for the full array.
-Since we pre-steer our data to every pixel in the image we simplify (\ref{weights}) by substituting $\a$ with a row vector $\1$ that represents broadside phase-steering.
-To estimate $\R$ we compute a sample covariance matrix $\eR$.
-In this computation we perform some degree of:
+When solving these convex optimization problems we can make use Lagrange multipliers.
 %
-\begin{itemize}
-\item \emph{spatial averaging} to avoid signal cancellation by decorrelating coherent echoes~\cite{Kailath1985};
-\item \emph{temporal averaging} over an interval comparable to the pulse length (one to five samples) to maintain true speckle statistics~\cite{Synnevag2009a};
-\item \emph{diagonal loading} to improve robustness to parameter errors~\cite{Cox1987,Maksym1979}.
-\end{itemize}%
+\begin{align}
+\frac{\partial}{\partial\w} \Big( \w\H\R\w + \lambda(\w\H\a - c)\big) = \R\w + \lambda\a = 0
+\end{align}
+%
+Solving for $\w$ gives
+%
+\begin{align}
+\w = -\lambda\Ri\a 
+\end{align}
+%
+Substituting this into the constraint equation gives
+%
+\begin{align}
+(-\lambda\Ri\a)\H\a = -\lambda\a\H\Ri\a = c
+\end{align}
+%
+\begin{align}
+\lambda = -\frac{c}{\a\H\Ri\a}
+\end{align}
+%
+Finally, insering this into the weight equation we have
+%
+\begin{align}
+\w = c\frac{\Ri\a}{\a\H\Ri\a}
+\end{align}
+%
+With $c=1$ we have the minimum variance distortionless response (MVDR) beamformer.
 

### 2015-10-12 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:10:12	11:42:46	(Mon12.Oct)
la2pdf commit - 2015:10:12	15:34:32	(Mon12.Oct)
la2pdf commit - 2015:10:12	18:31:47	(Mon12.Oct)

@@
 
-We show that a robust MVDR implementation typically creates responses ranging between that of the rectangular and Hamming.
-We let LCA select from a set of Kaiser windows with responses in this range, and add some steered variations of each.
+We show that a robust MVDR implementation typically creates windows spanning between the rectangular and Hamming.
+We let LCA select from a set of Kaiser windows with responses in this span, and add some steered variations of each.
 We limit the steering to roughly half the -3\,dB width of the window's amplitude response.
@@
 \end{IEEEkeywords}
-\fi}
+\fi
+}
 % \fi
@@
 We have shown in previous studies that MVDR's complexity can be reduced, and its implementation accelerated significantly using graphics computing units (GPUs)~\cite{Buskenes2014,Asen2013}.
-However, while MVDR relies on estimating and inverting a spatial covariance matrix, there exist a method that avoid this altogether.\todo{samme avsnitt?}
-
-This method is called the Low Complexity Adaptive (LCA) beamformer.
+However, this implementation is complex and still leaves something to be desired in speed.
+
+An alternative to MVDR is the Low Complexity Adaptive (LCA) beamformer.
 It was first proposed by Synnev\aa{}g in clinical medical imaging~\cite{Synnevag2008}, who demonstrated its ability to obtain very similar results to MVDR in a system with focused transmit beams.
-LCA applies a set of predefined windows and selects the one offering the best noise suppression.
+LCA applies a set of predefined windows, or tapers, and selects the one offering the best noise suppression.
 It relies on the same optimization criterion as MVDR, and may be considered a version of MVDR with a reduced window solution space.
-Alternatively, it may be viewed as a delay-and-sum (DAS) beamformer with an adaptive extension.
+Alternatively, it may be viewed as a delay-and-sum (DAS) beamformer with an adaptive multi-taper extension.
+\cite{Stankwitz1995}
 

### 2015-10-13 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:10:13	15:38:04	(Tue13.Oct)

@@
 We have shown in previous studies that MVDR's complexity can be reduced, and its implementation accelerated significantly using graphics computing units (GPUs)~\cite{Buskenes2014,Asen2013}.
-However, this implementation is complex and still leaves something to be desired in speed.
-
-An alternative to MVDR is the Low Complexity Adaptive (LCA) beamformer.
+However, the implementation is still complex, relatively slow and not intutive.
+
+A light weight alternative to MVDR is the Low Complexity Adaptive (LCA) beamformer.
 It was first proposed by Synnev\aa{}g in clinical medical imaging~\cite{Synnevag2008}, who demonstrated its ability to obtain very similar results to MVDR in a system with focused transmit beams.

### 2015-10-14 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:10:13	17:29:25	(Tue13.Oct)
la2pdf commit - 2015:10:14	09:47:32	(Wed14.Oct)

--- 2015-10-12 LCAforActiveSonarImaging/lca.tex
@@
 We have shown in previous studies that MVDR's complexity can be reduced, and its implementation accelerated significantly using graphics computing units (GPUs)~\cite{Buskenes2014,Asen2013}.
-However, this implementation is complex and still leaves something to be desired in speed.
-
-An alternative to MVDR is the Low Complexity Adaptive (LCA) beamformer.
-It was first proposed by Synnev\aa{}g in clinical medical imaging~\cite{Synnevag2008}, who demonstrated its ability to obtain very similar results to MVDR in a system with focused transmit beams.
+However, even with a near optimal implementation it is still slow compared to DAS.
+
+We propose using a light weight alternative to MVDR called the Low Complexity Adaptive (LCA) beamformer.
+It was first introduced by Synnev\aa{}g in clinical medical imaging~\cite{Synnevag2008}, who demonstrated its ability to obtain very similar results to MVDR in a system with focused transmit beams.
 LCA applies a set of predefined windows, or tapers, and selects the one offering the best noise suppression.
-It relies on the same optimization criterion as MVDR, and may be considered a version of MVDR with a reduced window solution space.
+It relies on the same optimization criterion as MVDR, thus may be considered a version of MVDR with a reduced window solution space.
 Alternatively, it may be viewed as a delay-and-sum (DAS) beamformer with an adaptive multi-taper extension.
-\cite{Stankwitz1995}
+Such extensions have been studied in e.g.
+radar imaging up to \cite{Stankwitz1995}, but unlike these
 

### 2015-10-14 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:10:14	12:25:36	(Wed14.Oct)

@@
 
-We propose using a light weight alternative to MVDR called the Low Complexity Adaptive (LCA) beamformer.
-It was first introduced by Synnev\aa{}g in clinical medical imaging~\cite{Synnevag2008}, who demonstrated its ability to obtain very similar results to MVDR in a system with focused transmit beams.
+A very fast and simple alternative to MVDR is the Low Complexity Adaptive (LCA) beamformer.
+It was first introduced by Synnev\aa{}g in clinical medical imaging~\cite{Synnevag2008}, who demonstrated its ability to obtain very similar results to MVDR in a system with focused transmit beams\todo{ha med at i US \o{}nsker man ulike vindus-typer p\aa{} forskjellig range?}.
 LCA applies a set of predefined windows, or tapers, and selects the one offering the best noise suppression.
@@
 Alternatively, it may be viewed as a delay-and-sum (DAS) beamformer with an adaptive multi-taper extension.
-Such extensions have been studied in e.g.
-radar imaging up to \cite{Stankwitz1995}, but unlike these
-
-Synnev\aa{}g used a window set comprised of rectangular, Kaiser and inverted Kaiser functions\footnote{May also be called Kaiser-Harris windows.}.
+Such extensions up to 3 windows by e.g.
+Stankwitz in radar imaging~\cite{Stankwitz1995}, but unlike these LCA more windows, and add micro-steered variations of them.
+
+Synnev\aa{}g used a window set comprised of rectangular, Kaiser (or Kaiser-Harris) and inverted Kaiser functions.
 He made LCA choose from 12 unique windows, 6 of which were micro-steered variants of a fairly wide Kaiser window.
 While achieving good results, it was not made clear how LCA performs with more or fewer windows, and how steering affects it all.
+His system
 

### 2015-10-14 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:10:14	13:18:07	(Wed14.Oct)

@@
 
-\newcommand\mediaPath{gfx/media}
+% \newcommand\mediaPath{gfx/media}
+\newcommand\mediaPath{http://folk.uio.no/joibu/media/2015_JOE_LCA}
+
 \newcommand\mediaI{\multimedia{\mediaPath/media1.mp4}{Media Movie 1}}
@@
 We have shown in previous studies that MVDR's complexity can be reduced, and its implementation accelerated significantly using graphics computing units (GPUs)~\cite{Buskenes2014,Asen2013}.
-However, even with a near optimal implementation it is still slow compared to DAS.
-
-A very fast and simple alternative to MVDR is the Low Complexity Adaptive (LCA) beamformer.
-It was first introduced by Synnev\aa{}g in clinical medical imaging~\cite{Synnevag2008}, who demonstrated its ability to obtain very similar results to MVDR in a system with focused transmit beams\todo{ha med at i US \o{}nsker man ulike vindus-typer p\aa{} forskjellig range?}.
+However, even this near optimal implementation is slow compared to DAS.
+
+A much faster and simpler alternative to MVDR is the Low Complexity Adaptive (LCA) beamformer.
+It was first introduced by Synnev\aa{}g in clinical medical imaging~\cite{Synnevag2008}, who demonstrated its ability to obtain very similar image quality to MVDR in a system with focused transmit beams\todo{ha med at i US \o{}nsker man ulike vindus-typer p\aa{} forskjellig range?}.
 LCA applies a set of predefined windows, or tapers, and selects the one offering the best noise suppression.

### 2015-10-14 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:10:14	14:39:08	(Wed14.Oct)

@@
 A much faster and simpler alternative to MVDR is the Low Complexity Adaptive (LCA) beamformer.
-It was first introduced by Synnev\aa{}g in clinical medical imaging~\cite{Synnevag2008}, who demonstrated its ability to obtain very similar image quality to MVDR in a system with focused transmit beams\todo{ha med at i US \o{}nsker man ulike vindus-typer p\aa{} forskjellig range?}.
-LCA applies a set of predefined windows, or tapers, and selects the one offering the best noise suppression.
+Based on an idea by Vignon~\cite{Vignon2008}, it was first introduced by Synnev\aa{}g in clinical medical imaging~\cite{Synnevag2008}, who demonstrated its ability to obtain very similar image quality to MVDR in a system with focused transmit beams\todo{ha med at i US \o{}nsker man ulike vindus-typer p\aa{} forskjellig range?}.
+LCA applies a set of predefined windows, or tapers, and selects the one offering the best sidelobe suppression.
 It relies on the same optimization criterion as MVDR, thus may be considered a version of MVDR with a reduced window solution space.

### 2015-10-14 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:10:14	21:20:17	(Wed14.Oct)

@@
 However, while the computational complexity of conventional beamformers are linear with the number of channels, O($M$), MVDR is at O($M^3$).
-This makes MVDR impracticable and sometimes infeasible.
-We have shown in previous studies that MVDR's complexity can be reduced, and its implementation accelerated significantly using graphics computing units (GPUs)~\cite{Buskenes2014,Asen2013}.
+This is because MVDR relies on estimating and inverting a covariance matrix.
+When the number of channels are less than 32, we have shown that the estimation step is the bottleneck and how this can be how complexity can be reduced and its implementation accelerated significantly using graphics computing units (GPUs)~\cite{Buskenes2014}.
+In clinical ultrasound where the systems tend to be larger and the beams are focused we have ported the same algorithm to beamspace for  performance~\cite{Asen2013}.
+Asl  a space reduction technique , which is a technique to .
 However, even this near optimal implementation is slow compared to DAS.

### 2015-10-14 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:10:14	22:02:02	(Wed14.Oct)

--- 2015-10-14 LCAforActiveSonarImaging/lca.tex
@@
 However, while the computational complexity of conventional beamformers are linear with the number of channels, O($M$), MVDR is at O($M^3$).
-This makes MVDR impracticable and sometimes infeasible.
-We have shown in previous studies that MVDR's complexity can be reduced, and its implementation accelerated significantly using graphics computing units (GPUs)~\cite{Buskenes2014,Asen2013}.
+This is because MVDR relies on estimating and inverting a covariance matrix.
+We have shown that in systems with less than $\approx$32 channels the complexity can be reduced and the implementation accelerated significantly using graphics computing units (GPUs)~\cite{Buskenes2014}.
+In larger systems the inversion step is the bottleneck and space reduction techniques more viable.
+In clinical ultrasound, where the transmit beams are focused and the spatial sparse, we obtained similar results running our GPU implementation of MVDR in beamspace~\cite{Asen2013}.
+Closely related techniques include assuming spatial stationarity in the narrow beam to form a Toepliz structued covariance matrix that is easier to invert~\cite{Asl2012} a space reduction technique , which is a technique to .
 However, even this near optimal implementation is slow compared to DAS.

### 2015-10-14 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:10:14	22:33:26	(Wed14.Oct)

@@
 In larger systems the inversion step is the bottleneck and space reduction techniques more viable.
-In clinical ultrasound, where the transmit beams are focused and the spatial sparse, we obtained similar results running our GPU implementation of MVDR in beamspace~\cite{Asen2013}.
-Closely related techniques include assuming spatial stationarity in the narrow beam to form a Toepliz structued covariance matrix that is easier to invert~\cite{Asl2012} a space reduction technique , which is a technique to .
-However, even this near optimal implementation is slow compared to DAS.
+In clinical ultrasound, where the transmit beams are focused and the spatial domain is sparse, we reduced  running our GPU implementation of MVDR in beamspace~\cite{Asen2013}.
+Two closely related techniques to recude is to form a simpler use principal component analysis (PCA)~\cite{Kim2014}, and to assume spatial stationarity to form more easily invertible Toepliz matrices~\cite{Asl2012}.
+At best these methods reduce the complexity to O($M^2$).
 

### 2015-10-14 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:10:14	23:05:15	(Wed14.Oct)

@@
 We have shown that in systems with less than $\approx$32 channels the complexity can be reduced and the implementation accelerated significantly using graphics computing units (GPUs)~\cite{Buskenes2014}.
-In larger systems the inversion step is the bottleneck and space reduction techniques more viable.
-In clinical ultrasound, where the transmit beams are focused and the spatial domain is sparse, we reduced  running our GPU implementation of MVDR in beamspace~\cite{Asen2013}.
-Two closely related techniques to recude is to form a simpler use principal component analysis (PCA)~\cite{Kim2014}, and to assume spatial stationarity to form more easily invertible Toepliz matrices~\cite{Asl2012}.
-At best these methods reduce the complexity to O($M^2$).
+For larger systems the inversion step is the bottleneck and space reduction techniques are desirable.
+In clinical ultrasound, where the transmit beams are focused, we processed our data in beamspace where the spatial sparsity allows much smaller matrices to be inverted~\cite{Asen2013}.
+Two closely related dimensionality reducing techniques used with MVDR includes using principal component analysis (PCA)~\cite{Kim2014}, and to assume spatial stationarity to form more easily invertible Toepliz matrices~\cite{Asl2012}.
+At best the mentioned methods are able to reduce the complexity to O($M^2$).
 
 A much faster and simpler alternative to MVDR is the Low Complexity Adaptive (LCA) beamformer.
-Based on an idea by Vignon~\cite{Vignon2008}, it was first introduced by Synnev\aa{}g in clinical medical imaging~\cite{Synnevag2008}, who demonstrated its ability to obtain very similar image quality to MVDR in a system with focused transmit beams\todo{ha med at i US \o{}nsker man ulike vindus-typer p\aa{} forskjellig range?}.
+Based on an idea by Vignon~\cite{Vignon2008}, it was first introduced by Synnev\aa{}g in clinical medical imaging~\cite{Synnevag2008}, who demonstrated its ability to obtain very similar image quality to MVDR in systems with focused transmit beams\todo{ha med at i US \o{}nsker man ulike vindus-typer p\aa{} forskjellig range?}.
 LCA applies a set of predefined windows, or tapers, and selects the one offering the best sidelobe suppression.
@@
 Alternatively, it may be viewed as a delay-and-sum (DAS) beamformer with an adaptive multi-taper extension.
-Such extensions up to 3 windows by e.g.
-Stankwitz in radar imaging~\cite{Stankwitz1995}, but unlike these LCA more windows, and add micro-steered variations of them.
+Multi-taper methods that select from up to 3 windows has been investigated by e.g.
+Stankwitz in radar imaging~\cite{Stankwitz1995}.
+LCA selects from even more windows, and also from micro-steered variations of them.
 

### 2015-10-14 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:10:14	23:37:52	(Wed14.Oct)

@@
 For larger systems the inversion step is the bottleneck and space reduction techniques are desirable.
-In clinical ultrasound, where the transmit beams are focused, we processed our data in beamspace where the spatial sparsity allows much smaller matrices to be inverted~\cite{Asen2013}.
+In clinical ultrasound, where the transmit beams are focused, we adapted our GPU implementation to operate in beamspace where the spatial sparsity allows much smaller matrices to be inverted~\cite{Asen2013}.
 Two closely related dimensionality reducing techniques used with MVDR includes using principal component analysis (PCA)~\cite{Kim2014}, and to assume spatial stationarity to form more easily invertible Toepliz matrices~\cite{Asl2012}.
@@
 It relies on the same optimization criterion as MVDR, thus may be considered a version of MVDR with a reduced window solution space.
-Alternatively, it may be viewed as a delay-and-sum (DAS) beamformer with an adaptive multi-taper extension.
-Multi-taper methods that select from up to 3 windows has been investigated by e.g.
-Stankwitz in radar imaging~\cite{Stankwitz1995}.
-LCA selects from even more windows, and also from micro-steered variations of them.
+Alternatively, it may be viewed as a an multi-taper technique.
+One such method is described by Stankwitz in radar imaging~\cite{Stankwitz1995}.
+Here the best of 3 windows are selected, but no micro-steered variations are used like LCA does.
 
 Synnev\aa{}g used a window set comprised of rectangular, Kaiser (or Kaiser-Harris) and inverted Kaiser functions.
-He made LCA choose from 12 unique windows, 6 of which were micro-steered variants of a fairly wide Kaiser window.
-While achieving good results, it was not made clear how LCA performs with more or fewer windows, and how steering affects it all.
-His system
+In total he made LCA choose from 12 unique windows, 6 of which were micro-steered variants of a fairly wide Kaiser window.
+While achieving good results, it was not made clear how LCA performs with more or fewer windows, and how they should be steered.
+Also, it is not clear how LCA performs in systems with unfocused transmit beams.
 
@@
 Finally we add some remarks on using the trigonometric window instead of Kaiser in \ref{sec:lca_trigonometric}, and of of sampling considerations in \ref{sec:lca_oversampling}.
-Results and discussion a provided in Section \ref{sec:results_discussion}, where we study MVDR's windows to find suitable LCA ones in \ref{sec:results_lca_window_function}, assess the range of window types and steering in \ref{sec:results_window_parameters}, determine the number of windows needed in \ref{sec:results_database_size}, and add a note on computational complexity in \ref{sec:results_complexity}.
+Results and discussion are provided in Section \ref{sec:results_discussion}, where we study MVDR's windows to find suitable LCA ones in \ref{sec:results_lca_window_function}, assess the range of window types and steering in \ref{sec:results_window_parameters}, determine the number of windows needed in \ref{sec:results_database_size}, and add a note on computational complexity in \ref{sec:results_complexity}.
 Finally we conclude in \ref{sec:conclusion}.
@@
 Assume that the wavefield has been sampled by an $M$ element uniform linear array, and that the signal signature has been removed by a matched filter.
-Let $x_m[\theta,n]$ be the delayed data from the $m$th channel, where the $\theta$ and $n$ is the azimuth angle and range sample of the focus point, respectively.
+Let $x_m[\theta,n]$ be the delayed data from the $m$th channel, where the $\theta$ and $n$ are the azimuth angle and range sample of the focus point, respectively.
 Each angle $\theta$ will be processed independently, so to simplify notation we will assume the dependence on $\theta$ to be implicit from now on.
@@
 \underset{\w[n]}{\argmin}\, E\{\big|z[n]\big|^2\} &= \underset{\w[n]}{\argmin}\, \w[n]\R[n]\w\H[n]\nn
-\text{subject to } \w[n]\a_\phi &= 1.\label{eq:mvdr_definition}
-\end{align}
-%
-Here $\a_\phi$ is a steering vector, $E\{\cdot\}$ is the expectation operator, and $\R[n] = E\{\x\x\H\}$ is the spatial covariance matrix. This is a convex optimization problem with the solution:
+\text{subject to } \w^H[n]\a_\phi &= 1.\label{eq:mvdr_definition}
+\end{align}
+%
+Here $\a_\phi$ is a steering vector, $E\{\cdot\}$ is the expectation operator, and $\R[n] = E\{\x\x\H\}$ is the spatial covariance matrix for the full array.
+This is a convex optimization problem with the solution:
 %
 \begin{gather}
-\vec w[n] = \frac{\Ri[n]\a_\phi}{\a_\phi\T\Ri[n]\a_\phi},\label{eq:mvdr_weights}
+\vec w[n] = \frac{\Ri[n]\a_\phi}{\a_\phi\T\Ri[n]\a_\phi}.\label{eq:mvdr_weights}
 \end{gather}
-where $\R=E\{\x[n]\x\H[n]\}\in\mathbb{C}^{M,M}$ is the spatial covariance matrix for the full array.
+.
 The problem lies in estimating and inverting this spatial covariance matrix.

### 2015-10-15 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:10:15	10:14:21	(Thu15.Oct)

@@
 
-MVDR seeks to minimize the output power of the beamformer, under the constraint of a unity gain in some desired direction $\phi$~\cite{Capon1969}:
+MVDR seeks to minimize the output variance of the beamformer, under the constraint of a unity gain in some desired direction $\phi$~\cite{Capon1969}.
+With the assumption of zero mean data this can be formulated as:
 %
@@
 %
-Here $\a_\phi$ is a steering vector, $E\{\cdot\}$ is the expectation operator, and $\R[n] = E\{\x\x\H\}$ is the spatial covariance matrix for the full array.
+Here $\a_\phi$ is a steering vector, $E\{\cdot\}$ is the expectation operator, and $\R[n] = E\{\x[n]\x\H[n]\}$ is the spatial covariance matrix for the full array.
 This is a convex optimization problem with the solution:
@@
 \end{gather}
-.
 The problem lies in estimating and inverting this spatial covariance matrix.
@@
 This is the same temporal averaging method we use for MVDR.
-In our case the matched filtered signal is a just few samples long, hence $K=1$ will be used for both MVDR and LCA throughout this work.
-
-\todo{Drop this paragraph altogether?}If extra flexibility is desired, we can also estimate the variance by a weighted combination of local pixels
+In our case the bandwidth of our matched filtered signal is a just few samples long, hence $K=1$ will be used for both MVDR and LCA throughout this work.
+
+Since pixels are correlated laterally, we can include a weighted combination of these as well to improve the variance estimation:
 %
@@
 where $N_x = 2X+1$ is the number of azimuth lines to average over, and $w_\text{prf}[x',n']$ is a normalized 2 dimensional weight function.
+However, we observed no visual improvement doing this, and the effect of micro-steered windows is hard to predict.
+Henceforth no lateral averaging will be applied.
 % 
@@
 %
-The Kaiser-Bessel window is near optimal in the sense of having its peak energy concentration around $\theta=0^\circ$, for a given time-bandwidth product specified with the parameter $\beta$:
+The Kaiser-Bessel window is near optimal in the sense of having its peak energy concentration around $\theta=0^\circ$, for a given space-bandwidth product specified with the parameter $\beta$:
 %
@@
 %
-where $T$ is the extent of the window in time\todo{Should perhaps say space-bandwidth product?} and $B$ is its bandwidth.
+where $T$ in our case is the spatial extent of the window and $B$ is its bandwidth.
 Adjusting $\beta$ changes the trade-off between mainlobe width and sidelobe level.
 When $\beta=0$ the window becomes rectangular, while at large values ($\beta>5$) the window converges to a Gaussian both in time and frequency.
-This class of windows is generally considered well suited for separating closely spaced\todo{better word?} sources with amplitudes of a high dynamic range~\cite{Harris1978}, they are easy to make, and they are optimal for any value of $\beta$.
+This class of windows is generally considered well suited for separating closely spaced sources with amplitudes of a high dynamic range~\cite{Harris1978}, they are easy to make, and they are optimal for any value of $\beta$.
 

### 2015-10-15 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:10:15	16:37:19	(Thu15.Oct)

@@
 This is because MVDR relies on estimating and inverting a covariance matrix.
-We have shown that in systems with less than $\approx$32 channels the complexity can be reduced and the implementation accelerated significantly using graphics computing units (GPUs)~\cite{Buskenes2014}.
-For larger systems the inversion step is the bottleneck and space reduction techniques are desirable.
-In clinical ultrasound, where the transmit beams are focused, we adapted our GPU implementation to operate in beamspace where the spatial sparsity allows much smaller matrices to be inverted~\cite{Asen2013}.
-Two closely related dimensionality reducing techniques used with MVDR includes using principal component analysis (PCA)~\cite{Kim2014}, and to assume spatial stationarity to form more easily invertible Toepliz matrices~\cite{Asl2012}.
+In sonar systems with less than 32 channels the complexity can be reduced and the implementation accelerated significantly using graphics computing units (GPUs)~\cite{Buskenes2014}.
+For larger systems the inversion step dominates the computation, the use of approximations through space reduction techniques become attractive.
+Some examples of such techniques include beamspace proessing on a GPU~\cite{Asen2013}, the closely related principal component analysis (PCA) method~\cite{Kim2014}, or to assume spatial stationarity to form more easily invertible Toepliz matrices~\cite{Asl2012}.
 At best the mentioned methods are able to reduce the complexity to O($M^2$).
@@
 A much faster and simpler alternative to MVDR is the Low Complexity Adaptive (LCA) beamformer.
-Based on an idea by Vignon~\cite{Vignon2008}, it was first introduced by Synnev\aa{}g in clinical medical imaging~\cite{Synnevag2008}, who demonstrated its ability to obtain very similar image quality to MVDR in systems with focused transmit beams\todo{ha med at i US \o{}nsker man ulike vindus-typer p\aa{} forskjellig range?}.
+Based on an idea by Vignon~\cite{Vignon2008}, it was first introduced by Synnev\aa{}g \emph{et al}.
+in clinical medical imaging~\cite{Synnevag2008}, who demonstrated its ability to obtain very similar image quality to MVDR in systems with focused transmit beams\todo{ha med at i US \o{}nsker man ulike vindus-typer p\aa{} forskjellig range?}.
 LCA applies a set of predefined windows, or tapers, and selects the one offering the best sidelobe suppression.
@@
 Alternatively, it may be viewed as a an multi-taper technique.
-One such method is described by Stankwitz in radar imaging~\cite{Stankwitz1995}.
-Here the best of 3 windows are selected, but no micro-steered variations are used like LCA does.
-
-Synnev\aa{}g used a window set comprised of rectangular, Kaiser (or Kaiser-Harris) and inverted Kaiser functions.
+One such method is described by Stankwitz \emph{et al}.
+in radar imaging~\cite{Stankwitz1995}.
+Here the best of 3 windows are selected, but unlike with LCA no micro-steered variations are used.
+
+Synnev\aa{}g \emph{et al}.
+suggested using a window set comprised of rectangular, Kaiser (or Kaiser-Harris) and inverted Kaiser functions.
 In total he made LCA choose from 12 unique windows, 6 of which were micro-steered variants of a fairly wide Kaiser window.
@@
 Both MVDR and LCA was run with $K=1$ temporal averaging.
-This is a fairly aggressive yet stable set of parameters.\todo{suitable ref?}
+This is a fairly aggressive yet stable set of parameters~\cite{Synnevag2009a}.
 
@@
 We suggest an upper bound for $\beta$ between 2 and 5, where 5 is slightly better at suppressing sidelobes.
-For $\phi$ we suggest to steer either to 50\% or 100\% of the \minus{}3\,dB width, which leads to a resolution of 73\% or 54\% of the rectangular window, respectively.
+For $\phi$ we suggest to steer either to 50\% or 100\% of the \minus{}3\,dB width, which leads to a resolution of 73\% or 53\% of the rectangular window, respectively.
 As lower boundaries we set $\beta=0$ to always include a rectangular window, and $\phi=0$ to always include the centered window.

### 2015-10-16 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:10:15	17:09:07	(Thu15.Oct)
la2pdf commit - 2015:10:16	07:51:03	(Fri16.Oct)

@@
 in radar imaging~\cite{Stankwitz1995}.
-Here the best of 3 windows are selected, but unlike with LCA no micro-steered variations are used.
+Here the best of 2 or 3 windows are selected, but unlike with LCA no micro-steered variations are used.
 
@@
 We extend Synnev\aa{}g's work by steering all our windows and not just some, and elaborate on the type of Kaiser windows to include and the method of steering.
-Then the method is applied to experimental data from the 32 element Kongsberg Maritime HISAS 1030 synthetic aperture sonar (SAS), with an operating frequency of 100\,kHz and transmit angle of 25$^\circ$.
+Then the method is applied to experimental data from the 32 element Kongsberg Maritime HISAS 1030 synthetic aperture sonar (SAS), with a 23$^\circ$ transmit angle operating frequency of 100\,kHz and .
 

### 2015-10-16 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:10:16	08:26:02	(Fri16.Oct)

@@
 In sonar systems with less than 32 channels the complexity can be reduced and the implementation accelerated significantly using graphics computing units (GPUs)~\cite{Buskenes2014}.
-For larger systems the inversion step dominates the computation, the use of approximations through space reduction techniques become attractive.
-Some examples of such techniques include beamspace proessing on a GPU~\cite{Asen2013}, the closely related principal component analysis (PCA) method~\cite{Kim2014}, or to assume spatial stationarity to form more easily invertible Toepliz matrices~\cite{Asl2012}.
-At best the mentioned methods are able to reduce the complexity to O($M^2$).
+For larger systems the inversion step dominates the computation.
+This can be dealt with by approximating the full covariance matrix with a small one using space reduction techniques such as beamspace processing~\cite{Asen2013}, or the closely related principal component analysis (PCA) method~\cite{Kim2014}.
+Another method is to assume spatial stationarity to form more easily invertible Toepliz matrices~\cite{Asl2012}.
+However, at best the mentioned methods are able to reduce the complexity to O($M^2$).
 
@@
 Based on an idea by Vignon~\cite{Vignon2008}, it was first introduced by Synnev\aa{}g \emph{et al}.
-in clinical medical imaging~\cite{Synnevag2008}, who demonstrated its ability to obtain very similar image quality to MVDR in systems with focused transmit beams\todo{ha med at i US \o{}nsker man ulike vindus-typer p\aa{} forskjellig range?}.
+in clinical medical imaging~\cite{Synnevag2008}, who demonstrated its ability to obtain very similar image quality to MVDR in systems with focused transmit beams.
 LCA applies a set of predefined windows, or tapers, and selects the one offering the best sidelobe suppression.
 It relies on the same optimization criterion as MVDR, thus may be considered a version of MVDR with a reduced window solution space.
-Alternatively, it may be viewed as a an multi-taper technique.
+Alternatively, it may be viewed as a multi-taper technique.
 One such method is described by Stankwitz \emph{et al}.
-in radar imaging~\cite{Stankwitz1995}.
-Here the best of 2 or 3 windows are selected, but unlike with LCA no micro-steered variations are used.
+in radar imaging~\cite{Stankwitz1995}, where the best out of 2 or 3 windows are selected.
+However, LCA is different in that it typically selects from more windows and allow micro-steered variations of them.
+This gives LCA an adaptive freedom that is somewhere in between that of traditional multi-taper techniques and MVDR.
 
@@
 suggested using a window set comprised of rectangular, Kaiser (or Kaiser-Harris) and inverted Kaiser functions.
-In total he made LCA choose from 12 unique windows, 6 of which were micro-steered variants of a fairly wide Kaiser window.
-While achieving good results, it was not made clear how LCA performs with more or fewer windows, and how they should be steered.
-Also, it is not clear how LCA performs in systems with unfocused transmit beams.
+In total he made LCA choose from 12 unique windows, 6 of which were micro-steered variants of a fairly wide Kaiser window, and with the rectangular window included.
+This was motivated by the desire to use wide windows with maximum sidelobe suppression near the receiver, and narrow windows at greater depths for maximum sensitivity and depth penetration.
+While achieving good results, it was not made clear how how many windows LCA needs, or how they ideally should be steered.
+Also, it is not apparent how LCA performs in sonar systems where the transmit beam is unfocused.
 

### 2015-10-16 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:10:16	10:03:13	(Fri16.Oct)

@@
 
-We use LCA with variations of the Kaiser window, but a comparison to using trigonometric windows will also be provided.
-We extend Synnev\aa{}g's work by steering all our windows and not just some, and elaborate on the type of Kaiser windows to include and the method of steering.
-Then the method is applied to experimental data from the 32 element Kongsberg Maritime HISAS 1030 synthetic aperture sonar (SAS), with a 23$^\circ$ transmit angle operating frequency of 100\,kHz and .
+In this study apply LCA to active sonar imaging.
+We use variations of the parametric Kaiser window, but also compare it to the trigonometric window function.
+We extend Synnev\aa{}g's work by steering all our windows and not just some, and demonstrate how LCA's performance is affected by type and number of windows included.
+Our results are obtained using experimental data from the 32 element Kongsberg Maritime HISAS 1030 synthetic aperture sonar (SAS), which operates at 100\,kHz with with a transmit angle of 23$^\circ$.
 

### 2015-10-16 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:10:16	10:34:17	(Fri16.Oct)

@@
 
-\newcommand\mediaI{\multimedia{\mediaPath/media1.mp4}{Media Movie 1}}
-\newcommand\mediaII{\multimedia{\mediaPath/media2.mp4}{Media Movie 2}}
+\newcommand\mediaI{\multimedia{\mediaPath/media1.mp4}{Media Movie~1}}
+\newcommand\mediaII{\multimedia{\mediaPath/media2.mp4}{Media Movie~2}}
 
@@
 We extend Synnev\aa{}g's work by steering all our windows and not just some, and demonstrate how LCA's performance is affected by type and number of windows included.
-Our results are obtained using experimental data from the 32 element Kongsberg Maritime HISAS 1030 synthetic aperture sonar (SAS), which operates at 100\,kHz with with a transmit angle of 23$^\circ$.
+Our results are obtained using experimental data from the 32 element Kongsberg Maritime HISAS 1030 synthetic aperture sonar (SAS), which operates at 100\,kHz with an element opening angle of 23$^\circ$.
 
@@
 
-To form a sonar image we need to estimate source locations and amplitudes.
-For this purpose we apply a spatial bandpass filter to the backscattered wavefield data.
+A sonar image is formed by estimate source locations and amplitudes.
+For this purpose a spatial bandpass filter is applied to the backscattered wavefield data.
 The filter is called an array processor, or receive beamformer.
@@
 \includegraphics[width=\textwidth]{gfx/oversampling_mosaic_bounds.pdf}%
-\caption{\emph{Determining Kaiser parameter boundaries}: LCA images created with a large and equal size window database, but where the upper bounds for $\beta$ and $\phi$ is varied.
+\caption{\emph{Determining Kaiser parameter boundaries}: LCA images of the cross and anchor created with a large and equal size window database, but where the upper bounds for $\beta$ and $\phi$ is varied.
 The upper left image is equal to DAS with a rectangular window.
-For all images we measure the lateral distance $\Delta x$ between the two -3dB points of the anchor.
+For all images we measure the lateral distance $\Delta x$ between the two -3dB points of the anchor (red object).
 These are specified relatively to the reference distance  $\Delta x_\text{ref}$ from the upper left DAS image.
@@
 
-The ultimate goal of adaptive beamformers is to take the information available in the wavefield and use it to either improve image resolution, noise suppression, or both.
+The ultimate goal of adaptive beamformers is to take the information available in the wavefield and use it to either improve image resolution, sidelobe suppression, or preferably both.
 This non-linear processing increases the image bandwidth and introduces a need for a sampling rate higher than the Nyquist rate.
@@
 
-To test the performance of the MVDR and LCA beamformers, we have processed data acquired by the Kongsberg Maritime HUGIN AUV carrying their 32 element HISAS1030 sonar~\cite{Hansen2009}.
-It is a high resolution synthetic aperture sonar with 1.2\;m array length, 100\;kHz operating frequency, 40\;kHz bandwidth and 25$^\circ$ opening angle on transmission.
+To test the performance of the MVDR and LCA beamformers, we have processed data acquired by the Kongsberg Maritime HUGIN AUV carrying their 32 element HISAS1030 sonar~\cite{Hansen2011}.
+It is a high resolution synthetic aperture sonar with 1.2\;m array length, 100\;kHz operating frequency, 30\;kHz bandwidth and 23$^\circ$ receiver element beamwidth.
 
@@
 
-The authors would like to thank Kongsberg Maritime, the Norwegian Defence Research Establishment (FFI), and Bundeswehr Technical Center for Ships and Naval Weapons, Maritime Technology and Research (WTD 71), for the equipment and effort needed to acquire data of the cross and Holmengraa, as well as for allowing us to use it.
+The authors would like to thank Kongsberg Maritime, the Norwegian Defence Research Establishment (FFI), and Bundeswehr Technical Center for Ships and Naval Weapons, Maritime Technology and Research (WTD 71), Germany, for the equipment and effort needed to acquire data of the resolution test cross and Holmengraa, as well as for allowing us to use it.
 

### 2015-10-16 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:10:16	11:08:35	(Fri16.Oct)

@@
 
-In many cases MVDR can improve the contrast and resolution of active sonar images compared to conventional static methods~\cite{Blomberg2013,Blomberg2012a,Dursun2009,Lo2004}.
+In many cases MVDR can improve the contrast and resolution of active sonar images compared to conventional static methods~\cite{Blomberg2013,Blomberg2012a,Lo2004}.
 However, while the computational complexity of conventional beamformers are linear with the number of channels, O($M$), MVDR is at O($M^3$).
@@
 We extend Synnev\aa{}g's work by steering all our windows and not just some, and demonstrate how LCA's performance is affected by type and number of windows included.
-Our results are obtained using experimental data from the 32 element Kongsberg Maritime HISAS 1030 synthetic aperture sonar (SAS), which operates at 100\,kHz with an element opening angle of 23$^\circ$.
+Our results are obtained using experimental data from the 32 element Kongsberg Maritime HISAS 1030 synthetic aperture sonar (SAS), which operates at 100\,kHz with an element beamwidth of 23$^\circ$.
 

### 2015-10-21 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:10:21	13:19:57	(Wed21.Oct)

@@
 
-\newif\ifPeerReview\PeerReviewfalse              % Whether to create the PeerReview version or
+\newif\ifPeerReview\PeerReviewtrue              % Whether to create the PeerReview version or
                                                 % Journal version
@@
 \newif\ifTODO\TODOtrue                        % Use todo notes?
+
+\newif\ifBuildBibliography\BuildBibliographytrue  % Whether to include the references.bib file or use inline references.
 
@@
 
-\title{Low Complexity Adaptive Beamformer\\ for Active Sonar Imaging}
+\title{Low Complexity Adaptive Sonar Imaging}
 
@@
 Slightly smoother images are produced if the window count is increased to 15, but past this we observe minimal difference.
-Finally we show that LCA work just as well if Kaiser windows are substituted with trigonometric ones.
-
-All our observations and experiences point to LCA being very easy to manage.
+Finally we show that LCA works just as well if Kaiser windows are substituted with trigonometric ones.
+
+All our observations and experiences point to LCA being very easy to understand and manage.
 It simply works, and is surprisingly insensitive to the exact type of window function, steering amount, or number of windows.
@@
 Bats, for instance, use a high frequency sonar to detect, identify and track prey with amazing precision in caves amongst a myriad of other bats.
-Active human made sonars have little to no adaptability on transmit, but several adaptive methods have been investigated post process the received data.
+Active human made sonars have little to no adaptability on transmit, but several adaptive methods have been investigated to post process the received data.
 Perhaps the best studied method for reconstructing a sonar image is the minimum variance distortion-less response (MVDR) beamformer~\cite{Capon1969}.
-It adapts the sonar array's response to minimize the influence of noise and interference in the final image.
+It adapts the sonar array's spatial response to minimize the influence of noise and interference in the final image.
 
@@
 This can be dealt with by approximating the full covariance matrix with a small one using space reduction techniques such as beamspace processing~\cite{Asen2013}, or the closely related principal component analysis (PCA) method~\cite{Kim2014}.
-Another method is to assume spatial stationarity to form more easily invertible Toepliz matrices~\cite{Asl2012}.
-However, at best the mentioned methods are able to reduce the complexity to O($M^2$).
+Another alternative is to assume spatial stationarity to form more easily invertible Toepliz matrices~\cite{Asl2012}.
+However, the  mentioned methods are still relatively slow compared to DAS.
 
@@
 in clinical medical imaging~\cite{Synnevag2008}, who demonstrated its ability to obtain very similar image quality to MVDR in systems with focused transmit beams.
-LCA applies a set of predefined windows, or tapers, and selects the one offering the best sidelobe suppression.
+LCA applies a set of predefined windows, or apodization weights, and selects the one offering the best sidelobe suppression.
 It relies on the same optimization criterion as MVDR, thus may be considered a version of MVDR with a reduced window solution space.
-Alternatively, it may be viewed as a multi-taper technique.
+Alternatively, it may be viewed as a multi-apodization technique.
 One such method is described by Stankwitz \emph{et al}.
 in radar imaging~\cite{Stankwitz1995}, where the best out of 2 or 3 windows are selected.
-However, LCA is different in that it typically selects from more windows and allow micro-steered variations of them.
-This gives LCA an adaptive freedom that is somewhere in between that of traditional multi-taper techniques and MVDR.
+However, LCA is different in that it typically selects from a larger pool of windows and allows micro-steered variations of them.
+This gives LCA an adaptive freedom that is somewhere in between that of traditional multi-apodization techniques and MVDR.
 
@@
 
-In this study apply LCA to active sonar imaging.
+In this study we apply LCA to active sonar imaging.
 We use variations of the parametric Kaiser window, but also compare it to the trigonometric window function.
@@
 
-Our results show that MVDR computes weights with responses that tend to be symmetric and steered within a fraction of the window's -\,3dB resolution.
+Our results show that MVDR, when adapted to work in active sonar imaging, computes weights with responses that tend to be symmetric and steered within a fraction of the window's -\,3dB resolution.
 When letting LCA choose from Kaiser windows with similar responses, we obtain images with improved noise suppression and resolution.
@@
 Overall, LCA produces images that are similar to MVDR, but with a resolution in between that of DAS and MVDR.
-LCA is also inherently robust, easy to implement and fairly easy to understand.
+LCA is also inherently robust, easy to implement, and fairly easy to understand.
 
@@
 Results and discussion are provided in Section \ref{sec:results_discussion}, where we study MVDR's windows to find suitable LCA ones in \ref{sec:results_lca_window_function}, assess the range of window types and steering in \ref{sec:results_window_parameters}, determine the number of windows needed in \ref{sec:results_database_size}, and add a note on computational complexity in \ref{sec:results_complexity}.
-Finally we conclude in \ref{sec:conclusion}.
+A conclusion is provided in \ref{sec:conclusion}.
 
@@
 
-A sonar image is formed by estimate source locations and amplitudes.
+A sonar image is formed by estimating source locations and amplitudes.
 For this purpose a spatial bandpass filter is applied to the backscattered wavefield data.
-The filter is called an array processor, or receive beamformer.
+The filter is called an array processor or receive beamformer.
 The basic principle is to apply delay and weights to the sensor channels before summing them up, chosen such that signals from the location of interest are summed coherently, while other sources sum incoherently.
@@
 where $w_m$ is the weight factor assigned to channel $m$.
-A weight set is commonly called a window or a taper function.
+A weight set is commonly called a window, apodization or taper function.
 Applying a window that trails off towards the edges improves noise suppression at the cost of reduced resolution~\cite{Harris1978}.
@@
 
-\begin{figure}[tbp]%
+\begin{figure}[tbp]\centering%
+\ifPeerReview%
+\includegraphics[width=.5\linewidth]{gfx/calc_kaiser_3dB.eps}%
+\else%
 \includegraphics[width=\linewidth]{gfx/calc_kaiser_3dB.eps}%
+\fi%
 \caption{Each Kaiser window is steered in the interval $\phi=[0, \phi_{\text{-3dB}}(\beta)]$.
@@
 
-\begin{figure}[tbp]%
+\begin{figure}[tbp]\centering%
+\ifPeerReview%
+\includegraphics[width=.5\linewidth]{gfx/cross_tagged.png}%
+\else%
 \includegraphics[width=\linewidth]{gfx/cross_tagged.png}%
-\caption{One of the objects used to test the beamformers is a 1\,m by 1\,m cross attached to an anchor with a diameter of approximately 13\,cm.
+\fi%
+\caption{One of the objects used to test the beamformers is a 1\,m by 1\,m resolution test cross attached to an anchor with a diameter of approximately 13\,cm.
 Source image curtesy of Bundeswehr Technical Center for Ships and Naval Weapons, Maritime Technology and Research (WTD 71).}\label{cross}
@@
 \includegraphics[width=\textwidth]{gfx/oversampling_mosaic_bounds.pdf}%
-\caption{\emph{Determining Kaiser parameter boundaries}: LCA images of the cross and anchor created with a large and equal size window database, but where the upper bounds for $\beta$ and $\phi$ is varied.
+\caption{\emph{Determining Kaiser parameter boundaries}: LCA images of the resolution test cross and anchor created with a large and equal size window database, but where the upper bounds for $\beta$ and $\phi$ is varied.
 The upper left image is equal to DAS with a rectangular window.
@@
 Adding more steering-variations improves both sidelobe suppression and resolution.
-However, using more than  $N_\beta=2$ window types and $N_\beta=3$ steering angles makes minimal difference.}\label{oversampling_mosaic}
+However, using more than  $N_\beta=2$ window types and $N_\phi=3$ steering angles makes minimal difference.}\label{oversampling_mosaic}
 \end{figure*}
@@
 \begin{figure*}[t]\centering
-\subfloat[\emph{Image quality of approximate point scatterers found in the cross scene}.
+\subfloat[\emph{Image quality of approximate point scatterers found in the resolution test cross scene}.
 Two lateral image cuts are shown in the leftmost figures, along with the response of a rectangularly weighted DAS.
@@
 
-In practice, we estimate the beamformer output power as
+In practice, we estimate the beamformer output power by computing a sample power average $s_z^2$:
 %
 \begin{align*}
-E\big\{\big|z[n]\big|^2\big\} = \frac{1}{N_k} \sumb{n'=n-K}{n+K} \big| z[n'] \big|^2
+E\big\{\big|z[n]\big|^2\big\} \approx s_z^2 = \frac{1}{N_k} \sumb{n'=n-K}{n+K} \big| z[n'] \big|^2
 \end{align*}
@@
 \begin{align}
-E\big\{\big|z[x,n]\big|^2\big\} = \frac{1}{N_x N_k} \sumb{\nmat{\scriptstyle x'=\\[-2mm]\scriptstyle x-X}}{x+X} \sumb{\nmat{\scriptstyle n'=\\[-2mm]\scriptstyle n-K}}{n+K} w[x',n']\big| z[x',n'] \big|^2
-\end{align}
-%
-where $N_x = 2X+1$ is the number of azimuth lines to average over, and $w_\text{prf}[x',n']$ is a normalized 2 dimensional weight function.
+s_z^2 = \frac{1}{N_x N_k} \sumb{\nmat{\scriptstyle x'=\\[-2mm]\scriptstyle x-X}}{x+X} \sumb{\nmat{\scriptstyle n'=\\[-2mm]\scriptstyle n-K}}{n+K} \omega[x',n']\big| z[x',n'] \big|^2
+\end{align}
+%
+where $N_x = 2X+1$ is the number of azimuth lines to average over, and $\omega[x',n']$ is a normalized 2 dimensional weight function.
 However, we observed no visual improvement doing this, and the effect of micro-steered windows is hard to predict.
@@
 %
-The Kaiser-Bessel window is near optimal in the sense of having its peak energy concentration around $\theta=0^\circ$, for a given space-bandwidth product specified with the parameter $\beta$:
+The Kaiser-Bessel window is near optimal in the sense of having its peak energy concentration around $\theta=0^\circ$, for a given space-bandwidth product related to the Kaiser parameter $\beta$ as:
 %
@@
 This data was collected by the Norwegian Defense Research Establishment and Kongsberg Maritime.
-The other scene is of an 1\,m by 1\,m iron cross, connected to a an anchor with 13\,cm diameter (\Fig{cross}).
+The other scene is of an 1\,m by 1\,m iron resolution test cross, connected to a an anchor with 13\,cm diameter (\Fig{cross}).
 This data was collected by the Bundeswehr Technical Center for Ships and Naval Weapons, Maritime Technology and Research (WTD 71).
@@
 
-Assuming that a robust MVDR is the reference method we want LCA to perform similarly to, it seems sensible to create a window database for LCA with with responses similar to the ones that MVDR computes.
+Assuming that a robust MVDR is the reference method we want LCA to perform similarly to, it seems sensible to create a window database for LCA with spatial responses similar to the ones that MVDR computes.
 We study this in \Fig{mvdr_selected_windows}, where we present the typical amplitude and phase responses that MVDR creates for shadow, speckle and highlight regions of the Holmengraa scene.
@@
 In each image we computed the lateral distance $\Delta x$ between the \minus{}3dB points of the anchor, with the reference $\Delta x_\text{ref}$ being the rectangularly weighted DAS image.
-A forth order polynomial fit was used in this computation.
+A fourth order polynomial fit was used in this computation.
 This measure is also commonly called the full width half maximum (FWHM).
@@
 
-\Fig{oversampling_mosaic_bounds_beta} and \Fig{oversampling_mosaic_bounds_phi} illustrates which $\beta$ and $\phi$ values LCA prefers for different regions in the image of the cross.
+\Fig{oversampling_mosaic_bounds_beta} and \Fig{oversampling_mosaic_bounds_phi} illustrates which $\beta$ and $\phi$ values LCA prefers for different regions in the image of the resolution test cross.
 Red marker lines are used to pinpoint the cut line, location of the anchor and strong scatters on the cross.
@@
 Two lateral image cuts through all images are presented in the left plot.
-The images and their corresponding cuts are near identical for the LCA version with Kaiser windows and the one with trigonometric windows.
+The images and their corresponding cuts are nearly identical for the LCA version with Kaiser windows and the one with trigonometric windows.
 Compared to DAS the LCA produces targets with an FWHM that lies in between that of DAS and MVDR.
@@
 
-LCA iterates through a set of windows in search of the one that minimizes the noise and interference power in each image pixel.
-It can be viewed as either a reduced solution space version of MVDR, or as an multi-taper extension to DAS.
-It creates images similar to MVDR, but at a fraction of the computational cost.
-It is also inherently robust and insensitive to the particular set of windows it selects from, supported by the observation that LCA can be used with either Kaiser or trigonometric windows and produce near identical images.
-
-When using LCA with Kaiser windows we made their responses similar to those typically chosen by a robust MVDR in key areas in the image.
-We suggest an upper bound for $\beta$ between 2 and 5, where 5 is slightly better at suppressing sidelobes.
-For $\phi$ we suggest to steer either to 50\% or 100\% of the \minus{}3\,dB width, which leads to a resolution of 73\% or 53\% of the rectangular window, respectively.
-As lower boundaries we set $\beta=0$ to always include a rectangular window, and $\phi=0$ to always include the centered window.
-
-The tendency of the LCA to perform similarly to the MV beamformer seems to indicate that a full solution space is rarely needed in real world scenarios.
-
+LCA seeks to improve image resolution and contrast by minimizing the power of noise and interference in each image pixel.
+This is the same optimization criterion as MVDR uses, but instead of computing an optimal array window like MVDR does, it selects the best window out of a predefined set.
+Hence, it can be viewed as either a reduced and discrete window space version of MVDR, or as an advanced multi-apodization technique.
+What we have studied in this paper is which windows we should ideally let LCA choose from.
+
+We have found the LCA to be surprisingly insensitive to the exact window function it is used with, supported by the observation that near identical images were produced whether we used Kaiser or trigonometric windows.
+However, it needs to be able to select between wide and narrow window responses, since the wide ones offer the best sidelobe suppression and the narrow ones offer the best sensitivity.
+We achieved good results using 2-3 Kaiser windows in the range $\beta\in[0,5]$, and observed minimal difference when adding more windows or adjusting the upper bound of $\beta$.
+For the steering angle we suggest a value in the range $\phi\in[0,\pm\delta\phi_\text{3dB}(\beta)]$, where $\phi_\text{3dB}(\beta)$ is the bandwidth of the respective window, and $\delta$ is a scaling parameters.
+Setting $\delta$ to either 50\% or 100\% reduces the lateral point target size to 73\% or 53\% compared to the rectangular window, respectively.
+Hence, it can be used to control aggressiveness and resolution gain of LCA.
+
+In summary, LCA is an attractive alternative to other adaptive beamformers due to being very fast, simple to understand, practically parameter-free, inherently robust, and able produce images similar to that of MVDR.
+Contrary to MVDR, it has a low computational complexity and can be easily and efficiently accelerated using GPUs.
+Since LCA performs just as well if we used trigonometric windows instead of Kaiser, we can even compute the optimal window parameter analytically for any given steering angle, and avoid the search over the window parameter entirely.
+
+% The tendency of LCA to produce images of similar quality as MVDR seems to indicate that a full solution space is rarely needed in real world scenarios.
+
+%  It initially very is also very fast, and is very well suited for acceleration using  very fast, easy to implement even on parallel hardware,
+
+% Its computational complexity is only of O($N_wM$), and its data parallel nature makes it very well suited for acceleration using e.g. GPUs. compared to MVDR's O($M^3$).
 % 
@@
 
+\ifBuildBibliography
    \bibliographystyle{IEEEtran}
    \bibliography{references}
-% 
-% 
+
+\else
+
+% Generated by IEEEtran.bst, version: 1.13 (2008/09/30)
+\begin{thebibliography}{10}
+\providecommand{\url}[1]{#1}
+\csname url@samestyle\endcsname
+\providecommand{\newblock}{\relax}
+\providecommand{\bibinfo}[2]{#2}
+\providecommand{\BIBentrySTDinterwordspacing}{\spaceskip=0pt\relax}
+\providecommand{\BIBentryALTinterwordstretchfactor}{4}
+\providecommand{\BIBentryALTinterwordspacing}{\spaceskip=\fontdimen2\font plus
+\BIBentryALTinterwordstretchfactor\fontdimen3\font minus
+  \fontdimen4\font\relax}
+\providecommand{\BIBforeignlanguage}[2]{{%
+\expandafter\ifx\csname l@#1\endcsname\relax
+\typeout{** WARNING: IEEEtran.bst: No hyphenation pattern has been}%
+\typeout{** loaded for the language `#1'. Using the pattern for}%
+\typeout{** the default language instead.}%
+\else
+\language=\csname l@#1\endcsname
+\fi
+#2}}
+\providecommand{\BIBdecl}{\relax}
+\BIBdecl
+
+\bibitem{Capon1969}
+J.~Capon, ``{High-resolution frequency-wavenumber spectrum analysis},''
+  \emph{Proceedings of the IEEE}, vol.~57, no.~8, pp. 1408--1418, 1969.
+
+\bibitem{Blomberg2013}
+A.~E.~A. Blomberg, A.~Austeng, R.~E. Hansen, and S.~A.~V. Synnes, ``{Improving
+  Sonar Performance in Shallow Water Using Adaptive Beamforming},'' \emph{IEEE
+  Journal of Oceanic Engineering}, vol.~38, no.~2, pp. 297--307, Apr. 2013.
+
+\bibitem{Blomberg2012a}
+A.~E.~A. Blomberg, A.~Austeng, and R.~E. Hansen, ``{Adaptive Beamforming
+  Applied to a Cylindrical Sonar Array Using an Interpolated Array
+  Transformation},'' \emph{IEEE Journal of Oceanic Engineering}, vol.~37,
+  no.~1, pp. 25--34, Jan. 2012.
+
+\bibitem{Lo2004}
+K.~Lo, ``{Adaptive Array Processing for Wide-Band Active Sonars},'' \emph{IEEE
+  Journal of Oceanic Engineering}, vol.~29, no.~3, pp. 837--846, Jul. 2004.
+
+\bibitem{Buskenes2014}
+J.~I. Buskenes, J.~P. Asen, C.-I.~C. Nilsen, and A.~Austeng, ``{An Optimized
+  GPU Implementation of the MVDR Beamformer for Active Sonar Imaging},''
+  \emph{IEEE Journal of Oceanic Engineering}, pp. 1--13, 2014.
+
+\bibitem{Asen2013}
+J.~P. \AA{}sen, J.~I. Buskenes, C.-I. {Colombo Nilsen}, A.~Austeng, and
+  S.~Holm, ``{Implementing capon beamforming on a GPU for real-time cardiac
+  ultrasound imaging.}'' \emph{IEEE transactions on ultrasonics,
+  ferroelectrics, and frequency control}, vol.~61, no.~1, pp. 76--85, Jan.
+  2014.
+
+\bibitem{Kim2014}
+K.~Kim, S.~Park, J.~Kim, S.-B. Park, and M.~Bae, ``{A fast minimum variance
+  beamforming method using principal component analysis},'' \emph{IEEE
+  Transactions on Ultrasonics, Ferroelectrics, and Frequency Control}, vol.~61,
+  no.~6, pp. 930--945, Jun. 2014.
+
+\bibitem{Asl2012}
+B.~M. Asl and A.~Mahloojifar, ``{A low-complexity adaptive beamformer for
+  ultrasound imaging using structured covariance matrix},'' \emph{IEEE
+  Transactions on Ultrasonics, Ferroelectrics and Frequency Control}, vol.~59,
+  no.~4, pp. 660--667, Apr. 2012.
+
+\bibitem{Vignon2008}
+F.~Vignon and M.~R. Burcher, ``{Capon beamforming in medical ultrasound imaging
+  with focused beams.}'' \emph{IEEE transactions on ultrasonics,
+  ferroelectrics, and frequency control}, vol.~55, no.~3, pp. 619--28, Mar.
+  2008.
+
+\bibitem{Synnevag2008}
+J.-F. Synnev\aa{}g, S.~Holm, and A.~Austeng, ``{A low complexity data-dependent
+  beamformer},'' in \emph{2008 IEEE Ultrasonics Symposium}, no.~2.\hskip 1em
+  plus 0.5em minus 0.4em\relax IEEE, Nov. 2008, pp. 1084--1087.
+
+\bibitem{Stankwitz1995}
+H.~Stankwitz, R.~Dallaire, and J.~Fienup, ``{Nonlinear apodization for sidelobe
+  control in SAR imagery},'' \emph{IEEE Transactions on Aerospace and
+  Electronic Systems}, vol.~31, no.~1, pp. 267--279, Jan. 1995.
+
+\bibitem{Harris1978}
+F.~Harris, ``{On the use of windows for harmonic analysis with the discrete
+  Fourier transform},'' \emph{Proceedings of the IEEE}, vol.~66, no.~1, pp.
+  51--83, 1978.
+
+\bibitem{Kailath1985}
+T.~Kailath and T.-J. Shan, ``{Adaptive beamforming for coherent signals and
+  interference},'' \emph{IEEE Transactions on Acoustics, Speech, and Signal
+  Processing}, vol.~33, no.~3, pp. 527--536, Jun. 1985.
+
+\bibitem{Synnevag2009a}
+J.-F. Synnev\aa{}g, A.~Austeng, and S.~Holm, ``{Benefits of minimum-variance
+  beamforming in medical ultrasound imaging},'' \emph{IEEE Transactions on
+  Ultrasonics, Ferroelectrics and Frequency Control}, vol.~56, no.~9, pp.
+  1868--1879, Sep. 2009.
+
+\bibitem{Cox1987}
+H.~Cox, R.~Zeskind, and M.~Owen, ``{Robust adaptive beamforming},'' \emph{IEEE
+  Transactions on Acoustics, Speech, and Signal Processing}, vol.~35, no.~10,
+  pp. 1365--1376, Oct. 1987.
+
+\bibitem{Maksym1979}
+J.~N. Maksym, ``{A robust formulation of an optimum cross-spectral beamformer
+  for line arrays},'' \emph{The Journal of the Acoustical Society of America},
+  vol.~65, no.~4, p. 971, 1979.
+
+\bibitem{Asen2014}
+J.~P. \AA{}sen, A.~Austeng, and S.~Holm, ``{Capon beamforming and moving
+  objects--an analysis of lateral shift-invariance.}'' \emph{IEEE transactions
+  on ultrasonics, ferroelectrics, and frequency control}, vol.~61, no.~7, pp.
+  1152--60, Jul. 2014.
+
+\bibitem{Hansen2011}
+R.~E. Hansen, H.~J. Callow, T.~O. Sabo, and S.~A.~V. Synnes, ``{Challenges in
+  Seafloor Imaging and Mapping With Synthetic Aperture Sonar},'' \emph{IEEE
+  Transactions on Geoscience and Remote Sensing}, vol.~49, no.~10, pp.
+  3677--3687, Oct. 2011.
+
+\bibitem{holmengraa}
+{Article of Holmengraa on dykkepedia.com}.
+
+\end{thebibliography}
+
+\fi
+
 \input{bio/jo_inge_buskenes}

### 2015-10-21 Simulator/article.tex
la2pdf commit - 2015:10:21	13:19:57	(Wed21.Oct)

@@
 
-One of the major shortcomings of autonomous underwater vehicles (AUVs) is their lack of curiosity.
-When encountering an interesting object, an AUV with a mounted sonar imaging system should be able to detect it, and perhaps move in for closer examination.
-Besides the apparent challenge of trying to detect something with a long list of unknown parameters, this must happen in real-time and be very power efficient.
+One of the major shortcomings of autonomous underwater vehicles (AUVs) is their lack of intelligence.
+When encountering an interesting object, an AUV mounted with a sonar imaging system should be able to tell that this object might be worth closer examination, and act accordingly.
+This is challenging as what we deem as interesting is often something that stand out form the background, i.e.
+it relies on previous knowledge.
+It would also need to process this information is near real-time to act in an efficient manner.
 

### 2015-10-21 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:10:21	14:23:06	(Wed21.Oct)

@@
 
-\newif\ifPeerReview\PeerReviewtrue              % Whether to create the PeerReview version or
+\newif\ifPeerReview\PeerReviewfalse              % Whether to create the PeerReview version or
                                                 % Journal version
@@
 Bats, for instance, use a high frequency sonar to detect, identify and track prey with amazing precision in caves amongst a myriad of other bats.
-Active human made sonars have little to no adaptability on transmit, but several adaptive methods have been investigated to post process the received data.
+Active human made sonars have little to no adaptability on transmit, but several methods have been investigated to adaptively post process the received data.
 Perhaps the best studied method for reconstructing a sonar image is the minimum variance distortion-less response (MVDR) beamformer~\cite{Capon1969}.
@@
 
-In this study we apply LCA to active sonar imaging.
+In this study we expand our previous work of applying LCA to active sonar imaging~\cite{Buskenes2011}.
 We use variations of the parametric Kaiser window, but also compare it to the trigonometric window function.
-We extend Synnev\aa{}g's work by steering all our windows and not just some, and demonstrate how LCA's performance is affected by type and number of windows included.
+We extend Synnev\aa{}g's work by steering all our windows and not just some, and demonstrate how LCA's performance is affected by the type and number of windows included.
 Our results are obtained using experimental data from the 32 element Kongsberg Maritime HISAS 1030 synthetic aperture sonar (SAS), which operates at 100\,kHz with an element beamwidth of 23$^\circ$.
 
-Our results show that MVDR, when adapted to work in active sonar imaging, computes weights with responses that tend to be symmetric and steered within a fraction of the window's -\,3dB resolution.
+Our results show that when MVDR is modified to work in active sonar imaging, it computes weights with spatial responses that tend to be symmetric and steered within a fraction of the window's -\,3dB resolution.
 When letting LCA choose from Kaiser windows with similar responses, we obtain images with improved noise suppression and resolution.
@@
 This article is outlined as follows: In Section \ref{sec:beamforming} we offer a gentle introduction to beamforming.
-Then we move on to adaptive beamforming and describe the MVDR and LCA methods \ref{sec:mvdr} and \ref{sec:lca}, respectively.
+Then we move on to adaptive beamforming and describe the MVDR and LCA methods in \ref{sec:mvdr} and \ref{sec:lca}, respectively.
 We use LCA with the Kaiser window which we describe in \ref{sec:lca_kaiser_windows}, explain how we steer it in \ref{sec:lca_steering}, and by how much in \ref{sec:lca_steering_bounds}.
@@
 Results and discussion are provided in Section \ref{sec:results_discussion}, where we study MVDR's windows to find suitable LCA ones in \ref{sec:results_lca_window_function}, assess the range of window types and steering in \ref{sec:results_window_parameters}, determine the number of windows needed in \ref{sec:results_database_size}, and add a note on computational complexity in \ref{sec:results_complexity}.
-A conclusion is provided in \ref{sec:conclusion}.
+Section \ref{sec:conclusion} provides a conclusion.
 
@@
 We have found the LCA to be surprisingly insensitive to the exact window function it is used with, supported by the observation that near identical images were produced whether we used Kaiser or trigonometric windows.
-However, it needs to be able to select between wide and narrow window responses, since the wide ones offer the best sidelobe suppression and the narrow ones offer the best sensitivity.
+However, it needs to be able to choose between wide and narrow window responses, since the wide ones offer the best sidelobe suppression and the narrow ones offer the best sensitivity.
 We achieved good results using 2-3 Kaiser windows in the range $\beta\in[0,5]$, and observed minimal difference when adding more windows or adjusting the upper bound of $\beta$.
@@
 Setting $\delta$ to either 50\% or 100\% reduces the lateral point target size to 73\% or 53\% compared to the rectangular window, respectively.
-Hence, it can be used to control aggressiveness and resolution gain of LCA.
+Hence, this parameter can be used to control the aggressiveness and resolution gain of LCA.
 
 In summary, LCA is an attractive alternative to other adaptive beamformers due to being very fast, simple to understand, practically parameter-free, inherently robust, and able produce images similar to that of MVDR.
-Contrary to MVDR, it has a low computational complexity and can be easily and efficiently accelerated using GPUs.
+Contrary to MVDR, it has a low computational complexity and can be easily and efficiently accelerated using e.g.
+GPUs.
 Since LCA performs just as well if we used trigonometric windows instead of Kaiser, we can even compute the optimal window parameter analytically for any given steering angle, and avoid the search over the window parameter entirely.
@@
 For this window we can obtain an analytic solution for the optimal value of $\alpha$ in the minimum variance sense.
-To derive it, we start by inserting the trigonometric function (\ref{eq:trig_window_function}), with applied steering (\ref{eq:steering_vector}), into the beamformer equation (\ref{eq:beamformer_output}):
+The derivation generalizes the one described for a non-steered window in~\cite{Stoica2005} to also apply for the steered (complex) window.
+We start by inserting the trigonometric function (\ref{eq:trig_window_function}), with applied steering (\ref{eq:steering_vector}), into the beamformer equation (\ref{eq:beamformer_output}):
 %

### 2015-10-27 Simulator/article.tex
la2pdf commit - 2015:10:27	11:49:36	(Tue27.Oct)

@@
 
-\begin{itemize}
-\item SWAT (Shallow Water Acoustic Toolset): 
-\item Bell's thesis:
-\item SIGMA:
-\item OpenGL does not support scattered writes. OpenCL does.
-\item 2D template approximation:
-\item Geometric distortions in SSS \cite{Cobra1992} Bell
-\item Sidescan sonar and synthetic aperture sonar. In seafloor coordinates. Maps to orthogonal rendering.
-\item Layover causes ambiguity, different seafloor phenomena can cause the same output image.
-Lots of cornercases when approximating the model in 2D.
-Instead use a 3D model and project it to a 2D one.
-Easier to operate, models are more general purpose.
-Less parameters to adjust.
-\item Viewer can be used to train an operator for finding non-trivial objects on the seafloor. Input for automated target recognition standard.
-\item Improved image quality of selected object, reduced data collection cost (price estimate?).
-\item OpenGL rendering can have infinite resolution, only limited by the pixel density in the image.
-\item Accuracy in terms of classification result. Need to embody static effects in a deterministic way. Incoherent synthesis of the image.
-\item Sound speed assumed constant.
-\item High frequency lead to little (define) penetration, lower range and higher resolution. Often modeled accurately with ray tracing techniques.
-\end{itemize}
-
-Model
-\begin{itemize}
-\item Ray solution: Separate amplitude and phase components of the Helmholtz eq.
-by applying geometric optics approximation.
-Ok as long as amplitude varies more slowly with position than the phase.
-Implications: High frequency, because the curvature of a ray over a wavelength must be small, and the fractional change in the sound speed must be small over a wavelength.
-Further simplification available by Snell's Law.
-Fast to compute.
-Not able to handle diffraction, focal points and caustics, and will cause (too) abrupt changes on distinct changes in sea environment.
-Can be improved upon with fuzzy beam modeling / Gaussian beam tracing.
-\begin{itemize}
-\item Pro: Easy to compute. Visual result. Handles complex bathymetry. Array and model response can be accounted for.
-\item Con: Only works for high frequency and small sound variations. No diffraction, caustics and focus. Infinite energy caustics/sharp variations. Incoherent(?)
-\end{itemize}
-\item Normal modes: Separate amplitude and phase into a range dependent term, and a depth dependent term.
-Assumes minimal range dependence in sound speed profile and depth.
-Limited near field accuracy.
-Acoustic field does not need to be calculated at intermediate ranges.
-Penetration can be modeled.
-can be calculated
-\begin{itemize}
-\item Pro: Ok to compute. Handles penetration. Low frequency, shallow water.
-\item Con: Horizontal stratification. Far field only. 
-\end{itemize}
-\item Parabolic equation: One way propagation.
-Marching solution.
-Initial starting solution must be known, often computed with normal mode.
-Only valid for narrow beams $\pm$20.
-Able to handle range dependent environments.
-Computing full field is easy.
-Computation time is of O(f$^2$), hence mostly useful for low-frequency scenarios.
-Unable to deal with horizontal refraction (shadow zones from e.g.
-conical mounts)
-\item Fast field programs / Green's function solutions.
-Exact full wave solution for acoustic fields in horizontally stratified media.
-Can handle both compressional and shear waves, and hence both fluid and solid media.
-Far field assumption.
-Similar to normal mode techniques.
-\item Finite element: Scene segmented into triangles or rectangles with a size of one tenth of a wavelength, or less.
-At each node the wavefield can be determined from the wave equation.
-Computationally limited to low frequencies.
-\item 2D model: Acoustic field in range and depth.
-Assumes that a wave transmitted in a vertical plane remains in it.
-Reasonable when the sound speed can be assumed constant in the horizontal plane, and when the seafloor normals are in the vertical range-depth plane (avoid horizontal refraction, or actually reflection).
-\end{itemize}
-
-What's new?
-\begin{itemize}
-\item Tuned for speed using GPUs, combining OpenGL and OpenCL to take advantage of existing functionality for computer graphics with the added flexibility of general purpose programming on the GPUs.
-\item Performance to spare: Game-like viewer for the visualization and verification of the rendered scene and corresponding sonar image.
-\item First simulator of its kinds aimed towards improved ATR and ultimately better AUVs.
-\item 3D models are still novel. 
-\end{itemize}
-
-% The ATR algorithms must, however, be sufficiently fast to process the sensor data onboard the AUV in delayed real-time.
-
-% We have developed a simulator for synthetic aperture sonar (SAS) that creates object templates very quickly with the aid of the massive computing power in graphics processing units (GPUs).
-It loads a 3D model of the sea floor and an object of interest into OpenGL, where emitted sound waves are modeled with a light source placed at the sonar transmit location.
-We assume rough, isotropic surfaces that reflects sound energy equally in all directions.
-This can be modeled with a Lambertian scattering model~\cite{Bell1997} where the intensity of the backscattered sound is said to depend only on the on incidence angle of the transmitted signal onto the model surface.
-When rendering OpenGL is set up to produce an optical 2D intensity image and depth map that reveals the distance from each pixels to the propagation axis.
-For maximum flexibility this data is finally combined with OpenCL to produce the image templates.
-% 
-% By coincidence our implementation is not so different from the SIGMAS+ simulator developed at the NATO Undersea Research Centre (NURC)~\cite{Coiras2009a, Coiras2009b}.
-They both take advantage of OpenGL to render a 2D image from 3D models, are designed for side-looking sonar, and assume a Lambertian scattering model.
-However, while SIGMAS+ is geared more towards realistically looking sonar images, and include effects such as noise, we do not.
-Also, their simulator sums images obtained from multiple OpenGL rendering passes to create the sonar image, while we render once and post-process this result with OpenCL to tailor the sonar image full flexibility.
+\begin{figure}[t]\centering%
+\graphicsAI[width=\linewidth]{gfx/scene_photograph.svg}%
+\caption{Simulator principle.}\label{simulator_principe}%
+\end{figure}
 

### 2015-10-28 Simulator/article.tex
la2pdf commit - 2015:10:27	15:59:37	(Tue27.Oct)
la2pdf commit - 2015:10:28	11:41:14	(Wed28.Oct)
la2pdf commit - 2015:10:28	19:17:10	(Wed28.Oct)
la2pdf commit - 2015:10:29	00:28:02	(Thu29.Oct)

@@
 \newif\ifTODO\TODOtrue                        % Use todo notes?
+
+\newif\ifBuildBibliography\BuildBibliographyfalse  % Whether to include the references.bib file or use inline references.
 
@@
 
-\usepackage{tikz}
-\usetikzlibrary{shapes,snakes}
+\newcommand\minus{\scalebox{0.75}[1.0]{$-$}}
+
+% \usepackage{tikz}
+% \usetikzlibrary{shapes,snakes}
 \usepackage{amsmath,amssymb}
@@
 
-\newenvironment{outline}
-{\begin{itemize}}
-{\end{itemize}}
-
-% correct bad hyphenation here
-% \hyphenation{op-tical net-works semi-conduc-tor}
-
-%%%%%%%%%%%%                                    ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
-% GLOSSARY %
-%%%%%%%%%%%%                                    ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
-
-% \makeglossaries
-% \input{../../Library/glossary}
-
-%%%%%%%%%%%%%%%%%%                              ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
-% DOCUMENT START %
-%%%%%%%%%%%%%%%%%%                              ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
-
-% \usepackage{yfonts}
-
-% \graphicspath{submission/final/gfx/}
+% \newenvironment{outline}
+% {\begin{itemize}}
+% {\end{itemize}}
 
@@
 
-\maketitle
+\section{Introduction}
 
-\section{Introduction}
+\setcounter{totalnumber}{1}
+\setcounter{topnumber}{1}
+\setcounter{dbltopnumber}{1}
+
+\begin{figure*}[t]\centering%
+\graphicsAI[width=\linewidth]{gfx/scene_coordinate_system.svg}%
+\caption{Simulator principle.}\label{simulator_coordinate_system}%
+\end{figure*}
+
+% \begin{figure}[t!]\centering%
+% \graphicsAI[drawing,width=\linewidth]{gfx/specs.svg}%
+% \caption{Simulator specifications.}\label{specs}%
+% \end{figure}
+
+\begin{figure*}[t]\centering%
+\graphicsAI[drawing,width=\linewidth]{gfx/simulator.svg}%
+\caption{Simulator concept.}\label{buildup}%
+\end{figure*}
 
@@
 Also, their simulator superimpose images obtained from multiple OpenGL rendering passes to create the sonar image, while we render once and post-process this result with OpenCL to tailor the sonar image with greater flexibility.
-
-\begin{figure}[t]\centering%
-\graphicsAI[width=\linewidth]{gfx/scene_photograph.svg}%
-\caption{Simulator principle.}\label{simulator_principe}%
-\end{figure}
-
-\begin{figure}[t]\centering%
-\graphicsAI[drawing,width=\linewidth]{gfx/specs.svg}%
-\caption{Simulator specifications.}\label{specs}%
-\end{figure}
-
-\begin{figure*}[t]\centering%
-\graphicsAI[drawing,width=\linewidth]{gfx/simulator.svg}%
-\caption{Simulator concept.}\label{buildup}%
-\end{figure*}
 
@@
 
-We implement rotations in a 3D euclidean space by combining matrices that rotate around each of the three respective dimensions..
-For rotation by an angle $\phi_x$ clockwise around the $x$-axis we have
+We implement rotations in a 3D euclidean space by combining matrices that rotate around each of the three respective dimensions. For rotation by an angle $\phi_x$ clockwise around the $x$-axis we have
 \begin{align}

### 2015-10-29 Simulator/article.tex
la2pdf commit - 2015:10:29	11:01:10	(Thu29.Oct)

@@
 One of the major shortcomings of autonomous underwater vehicles (AUVs) is their lack of intelligence.
-When encountering an interesting object, an AUV mounted with a sonar imaging system should be able to tell that this object might be worth closer examination, and act accordingly.
+When an AUV equipped with a sonar encounter an interesting object, it would be very beneficial if adapted its behavior to examine the object further.
 This is challenging as what we deem as interesting is often something that stand out form the background, i.e.

### 2015-10-29 Simulator/article.tex
la2pdf commit - 2015:10:29	11:48:08	(Thu29.Oct)

@@
 
-One of the major shortcomings of autonomous underwater vehicles (AUVs) is their lack of intelligence.
+\IEEEPARstart{O}{ne} of the greatest visions of science is to create intelligent machines.
+This is This is especially true for devices designed to sense and act without aid, such as autonomous underwater vehicles (AUVs).
+The potential here seems only limited by our imagination.
 When an AUV equipped with a sonar encounter an interesting object, it would be very beneficial if adapted its behavior to examine the object further.

### 2015-10-31 Simulator/article.tex
la2pdf commit - 2015:10:31	09:31:54	(Sat31.Oct)

--- 2015-10-29 Simulator/article.tex
@@
 
-One of the major shortcomings of autonomous underwater vehicles (AUVs) is their lack of intelligence.
+\IEEEPARstart{P}{erhaps} of the greatest visions of science is to create intelligent machines.
+The potential seems immense and perhaps only limited by our imagination.
+If we , and is perhaps This is especially true for devices designed to sense and act without aid, such as autonomous underwater vehicles (AUVs).
+The potential here seems only limited by our imagination.
 When an AUV equipped with a sonar encounter an interesting object, it would be very beneficial if adapted its behavior to examine the object further.
-This is challenging as what we deem as interesting is often something that stand out form the background, i.e.
+This is challenging as it would need to learn in what ways the interesting objects differ from the background, i.e.
 it relies on previous knowledge.

### 2015-12-17 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:12:17	09:57:16	(to.17.des.)

@@
 
-\newif\ifBuildBibliography\BuildBibliographytrue  % Whether to include the references.bib file or use inline references.
+\newif\ifBuildBibliography\BuildBibliographyfalse  % Whether to include the references.bib file or use inline references.
 
@@
 Hence, it can be viewed as either a reduced and discrete window space version of MVDR, or as an advanced multi-apodization technique.
-What we have studied in this paper is which windows we should ideally let LCA choose from.
-
-We have found the LCA to be surprisingly insensitive to the exact window function it is used with, supported by the observation that near identical images were produced whether we used Kaiser or trigonometric windows.
-However, it needs to be able to choose between wide and narrow window responses, since the wide ones offer the best sidelobe suppression and the narrow ones offer the best sensitivity.
-We achieved good results using 2-3 Kaiser windows in the range $\beta\in[0,5]$, and observed minimal difference when adding more windows or adjusting the upper bound of $\beta$.
-For the steering angle we suggest a value in the range $\phi\in[0,\pm\delta\phi_\text{3dB}(\beta)]$, where $\phi_\text{3dB}(\beta)$ is the bandwidth of the respective window, and $\delta$ is a scaling parameters.
+In this paper we have studied which windows we should ideally let LCA choose from.
+
+We found the LCA to be surprisingly insensitive to the exact window function it is used with, supported by the observation that near identical images were produced whether we used Kaiser windows or trigonometric ones.
+However, LCA must be able to choose between wide and narrow window responses, since the wide ones offer the best sidelobe suppression and the narrow ones offer the best sensitivity.
+We achieved good results with 2-3 Kaiser windows in the range $\beta\in[0,5]$, and observed minimal difference when adding more windows or adjusting the upper bound of $\beta$.
+For the steering angle we suggest a value in the range $\phi\in[\pm\delta\phi_\text{3dB}(\beta)]$, where $\phi_\text{3dB}(\beta)$ is the bandwidth of the respective window, and $\delta$ is a scaling parameter.
 Setting $\delta$ to either 50\% or 100\% reduces the lateral point target size to 73\% or 53\% compared to the rectangular window, respectively.
@@
 GPUs.
-Since LCA performs just as well if we used trigonometric windows instead of Kaiser, we can even compute the optimal window parameter analytically for any given steering angle, and avoid the search over the window parameter entirely.
+Since LCA performs just as well if we use trigonometric windows instead of Kaiser ones, we can also compute the optimal window parameter analytically for any given steering angle, and avoid the search over the window parameter entirely.
 
@@
 For this window we can obtain an analytic solution for the optimal value of $\alpha$ in the minimum variance sense.
-The derivation generalizes the one described for a non-steered window in~\cite{Stoica2005} to also apply for the steered (complex) window.
+The following derivation generalizes that of the non-steered (real) window described in e.g.~\cite{Stoica2005} to also apply for the steered (complex) window.
 We start by inserting the trigonometric function (\ref{eq:trig_window_function}), with applied steering (\ref{eq:steering_vector}), into the beamformer equation (\ref{eq:beamformer_output}):
@@
 
+\bibitem{Buskenes2011}
+J.~I. Buskenes, C.-I.~C. Nilsen, and A.~Austeng, ``{A Low Complexity Adaptive
+  Beamformer for Active Sonar Imaging},'' in \emph{Proceedings of the
+  Underwater Acoustic Measurements: Technologies \& Results}.\hskip 1em plus
+  0.5em minus 0.4em\relax Underwater Acoustic Measurements: Technologies \&
+  Results, 2011.
+
 \bibitem{Harris1978}
@@
 
+\bibitem{Stoica2005}
+P.~Stoica and R.~L. Moses, ``{Nonparametric Methods},'' in \emph{Spectral
+  Analysis of Signals}.\hskip 1em plus 0.5em minus 0.4em\relax Prentice Hall,
+  2005, ch.~2, pp. 23--89.
+
 \end{thebibliography}

### 2015-12-17 Simulator/article.tex
la2pdf commit - 2015:12:17	09:57:16	(to.17.des.)
la2pdf commit - 2015:12:17	11:01:02	(to.17.des.)

--- 2015-10-29 Simulator/article.tex
@@
 \graphicsAI[width=\linewidth]{gfx/scene_coordinate_system.svg}%
-\caption{Simulator principle.}\label{simulator_coordinate_system}%
+\caption{Simulator principle.
+A model of the seafloor and object is loaded into OpenGL.
+All points in the imaged seafloor region that are visible from the sonar are orthogonally projected onto a view plane.
+}\label{simulator_coordinate_system}%
 \end{figure*}
@@
 
-One of the major shortcomings of autonomous underwater vehicles (AUVs) is their lack of intelligence.
-When an AUV equipped with a sonar encounter an interesting object, it would be very beneficial if adapted its behavior to examine the object further.
-This is challenging as what we deem as interesting is often something that stand out form the background, i.e.
+\IEEEPARstart{O}{ne} of the greatest visions in science is that of artificial intelligence.
+If we could make machines not only do what we say, but also able to adapt their behaviour to handle the unexpected, the possibilities are perhaps only limited by our imagination.
+An apparent area of use is devices designed to sense and act without aid, such as autonomous underwater vehicles (AUVs).
+When an AUV with a sonar imaging system encounter an interesting object, it would be very beneficial if adapted its behavior to examine the object further.
+This is challenging as it would need to learn in what ways the interesting objects differ from the background, i.e.
 it relies on previous knowledge.

### 2016-01-07 Simulator/article.tex
la2pdf commit - 2016:01:07	09:19:40	(to. 7.jan.)

@@
 
+% - Adapt simulation to its behaviour, see if they still match.
+
 % Automatic target recognition (ATR) is an important component in autonomous underwater vehicles (AUVs), as it allows the vehicle to adapt its mission plan to e.g.
@@
 This is achieved with the aid of the massive computing power in graphics processing units (GPUs) and optimized software libraries for scene rendering.
-The simulator loads a 3D model of the seafloor and an object class model into OpenGL, where emitted sound waves are modeled with a light tube extending all the way down the propagation path of the AUV.
+The simulator loads a 3D model of the seafloor and object class into OpenGL and positions the camera at the sonar location.
+The scene is highlighted using a light tube that extends to infinity along the AUV's axis of probagation.
 We assume rough, isotropic surfaces reflecting sound energy equally in all directions.
@@
 This limits its use to only a few parameters being coarsely sampled, ultimately yielding inaccurate results~\cite{Midelfart2010}.
-
-To avoid this problem we have developed a very fast side-looking sonar simulator that generates templates adapted to the actual scene.
-
-synthetic aperture sonar (SAS)
 

### 2016-01-15 Simulator/article.tex
la2pdf commit - 2016:01:08	14:04:57	(fr. 8.jan.)
Adding files from LCA submit 1.

@@
 
-What's going on?
+We have tested the template matching techniqiues on experimental data from the HISAS1030 synthetic aperture sonar (SAS) attached to the HUGIN autonomous underwater vehicle (AUV), both developed by Kongsberg Maritime.
+Three objects from Jesusbukta (Jesus Bay) in Norway will be presented; two roughly 3\,m long cylinders imaged a couple days apart in Janyary 2009, and a roughly 5\m long torpedo imaged in June 2012.
+
+The resulting classification performance of the template match ing techniques
+
+Data from Jesusbukta
+
+Cylinder 1: 2.9m (short side towards sonar)
+Cylinder 2: 3.05m (diagonal)
+Torpedo: 5.06m (looks good)
 

### 2016-01-15 LCAforActiveSonarImaging/submission/submit1/answer/answer.tex
la2pdf commit - 2016:01:15	21:14:58	(fr.15.jan.)

--- 2014-03-18 HowtoHowtoCapon/submission/submit1/answer/answer.tex
@@
 \begin{tabular}{l p{.6\linewidth}}\centering
-Manuscript number: & 2013JOE001295 \\
-Title: & An Optimized GPU Implementation of the MVDR Beamformer for Active Sonar Imaging \\
-Authors: & Buskenes,~J.I., \AA{}sen,~J.P., Nilsen,~C.-I.C, Austeng,~A.
+Manuscript number: & 2015JOE001805 \\
+Title: & Low Complexity Adaptive Sonar Imaging \\
+Authors: & Buskenes,~J.I., Hansen,~R.E., Austeng,~A.
 \end{tabular}
@@
 
+\subsection*{General}
+
+\q{The paper is well written and addresses an actual problem. The solution however, is barely innovative. Still the results are interesting enough for publication. Some minors details.}
+
+\subsection*{Specifics}
+
 \begin{enumerate}
 % \emph{General}
-\hi{General}
-\qi{Mpx/s is not a widely known abbreviation. This needs to be defined upon first use.}
-\ai{This is a good point. "Mpx" is now corrected to "MP", in accordance with the second reviewer's suggestion and the Oxford dictionary.}
-\qi{"Matlab" should be all caps: "MATLAB" throughout the paper.}
-\ai{Thank you, we agree. Corrections applied.}
+\hi{Introduction}
+\qi{The authors show no evidence that bats use adaptive processing, remove this example.}
+\ai{}
 %
-\hi{Abstract}
-\qi{A key attribute of the MVDR studied in this paper is the use of a subarray approximation. This should be stated up front in the abstract, i.e. "subarray MVDR" instead of MVDR.}
-\ai{The reviewer is right that subarray averaging is important in our study.
-But we believe that "subarray MVDR" would be misleading, as to the best of our knowledge this term is usually used to describe the use of subarrays to reduce complexity, while we average over subarrays to deal with the coherent echoes present in \emph{active} systems.
-We still apply the weights to the full array.
-To avoid confusion we have now stated this better up front in the article (section I paragraph 5).}
+\qi{It is not clear whether LCA is really an adaptive method that adapts weights on the fly or a method that a posteriori decides which window shows best results.}
+\ai{}
 %
-\qi{This paper advertises an impressive ("more than two orders of magnitude) improvement over a C implementation of MVDR, but this is really misleading.
-In reality, an improvement of up to an order of magnitude is achieved over an optimized C implementation, as later stated in the caption for figure 8.
-A comparison of an optimized GPU implementation to an unoptimized C implementation is not very meaningful — both should be optimized.
-The abstract should be revised to indicate up to an order of magnitude over an optimized C implementation.}
-\ai{We agree with the reviewer that this should be improved.
-We still state the two order of magnitude improvement over our reference implementation as a "total payoff" for our efforts, but now also state that this figure would likely be 5-10 had the CPU side received similar treatment.}
+\qi{The use of non-SI units like cm is discouraged.}
+\ai{}
 %
-\hi{Section I}
-\qi{Typo in second paragraph: "litterature" should be "literature".}
-\ai{Thank you. Fixed.}
-%
-\qi{Subarray processing is a common approximation and should be included in this list (reference D J Chapman, "Partial Adaptivity for the Large Array," IEEE Trans.
-Antennas \& Prop., Vol.
-Ap-24, No.
-5, September 1976, pp.
-685-696).}
-\ai{We thank the reviewer for providing this reference.
-However, unlike the subarray technique described by Chapman, which seems focused on reducing array complexity but in a way that causes grating lobe trouble, we use overlapping subarrays to (1) deal with the coherency problem in active sonar, and (2) to improve the rank of covariance matrix.
-The latter is particularly important in our case since we have hardly any temporal sample support.
-The Chapman reference,  distinction and a reference to [9] Kailath and Shan, "Adaptive beamforming for coherent signals and interference" is now included in the introduction to hopefully avoid future confusion.}
-%
-\hi{Section II}
-\qi{Should reference the literature (from above comment 6) on the subarray approximation in paragraph 3.}
-\ai{We agree that that this paragraph lacks a reference, but feel that [9] Kailath and Shan "Adaptive beamforming for coherent signals and interference" better describes this method in the context of active sonar.
-The Chapman reference is added to the introduction instead.}
-%
-\hi{Section III (A-C)}
-\qi{Typo in first paragraph after figure: "got" should be "get".}
-\qi{Move "a" in "It is a Gauss Jordan based".}
-\qi{"which are a mere scaled version" should be "which are merely scaled versions".}
-\ai{We thank the reviewer for these corrections (8.-10.). They are all applied.}
-%
-\hi{Section IV}
-\qi{The sub-plots on the top break the processing out into 3 pieces, yet the lower sub-plots reflect the entire processing string for MATLAB and C.
-It would be useful to show the entire processing time for the GPU for better comparison.
-Perhaps breaking the two rows of sub-plots into two figures would work better.}
-\ai{The reviewer has a point.
-But since there are 3 occurrences of the covariance processing time, including the total GPU processing time would introduce three new lines.
-We feel this would clutter the plot unnecessarily as it is fairly easy to derive the total time by looking at the current plots.
-Putting the total processing time in the MATLAB/C total time figure is an option, but then a logarithmic y-scale would necessary and detail would be lost.
-In stead we have included a remark in the caption saying the total GPU time is the sum of all the lines.}
-%
-\qi{Comment about real processing improvement over optimized C should be related in the text instead of a caption. This is an important point about the comparisons made in the paper.}
-\ai{It is indeed. We dedicated the second paragraph in the discussion (IV) to this topic, but have revised the paragraph to emphasize this point better.}
-%
-\qi{(caption of figure 10) "cruel" should be "crude"}
-\ai{Crude grammar correction applied.}
-%
-\qi{The point made in the first paragraph about the data transfer from GPU to CPU being ignored further muddies the water in comparing the two approaches.
-This point should also be stated up front.
-I would suggest making a very clear itemization of the assumptions and caveats about the two approaches.}
-\ai{We agree that this should should be clarified, but since the transfer can happen in parallel without adding to the computation time we do not believe this point belongs in the introduction.
-An elaboration on the topic is included in the last paragraph in section IV.
-}
-%
-\hi{Section V}
-\qi{The reason the C implementation is linear in L is because it is dominated by the calculation of the covariance matrix and data movement.
-The inversion is negligible here.
-On the other hand, the GPU optimizes the covariance calculation so much that now the inversion is significant, hence the non-linear computation time as a function of L.
-This is really the key point to be made in the comparison between the two approaches, in my opinion.}
-\ai{Yes! This is likely the reason. Thank you for pointing this out. This is now added to the discussion.}
-\qi{Bottom of page, "hold-ups" has an extra space after the hyphen.}
-\ai{Fixed.}
-\hi{Section V}
-\qi{Two spaces before "1 Mpx/s" in 4th paragraph. }
-\ai{Thank you, corrected.}
 \end{enumerate}
 
-\section{Answer to reviewer 2}
+\section{Answer to reviewer 3}
 
-\q{The authors describe their work on implementing an MVDR adaptive beamformer for a high-frequency imaging sonar on a particular graphical processing unit (GPU), the prevous-generation NVIDIA Quadro 6000, using NVIDIA's CUDA C environment.
-The authors report 1-2 orders of magnitude faster processing with their GPU implementation, compared to their MATLAB and single-thread, unoptimized CPU implementations, without taking the input/output of data in and out of the GPU.}
+\subsection*{General}
 
-\q{Because the paper is a valuable result to imaging sonar users, who will (to some extent) know that they can speed up their processing by 1-2 orders of magnitude if they use a GPU, I think the paper should be published.}
+Although the topic is potentially interesting for all readers, due to many specific terms used in the text, in the present form the article would probably only be highly appreciated by expert readers.
+I would recommend some minor changes in order to widen the frame of potential readers.
 
-\q{However, I think the authors could greatly improve the appeal of the paper to a wider audience by adding helpful context and background on their implementation.
-I will suggest some questions that I think the authors could easily address toward this end.
-This is a report from the "front lines" of migrating scientific computing to GPUs, a topic which seems to be of great interest to everybody in our field that does numerical computing.
-In my review, I will comment on the information it provides to a reader interested in starting to use GPUs (and the authors for some selective elaboration).}
+\subsection*{Specific points}
+
+\begin{enumerate}
+% \emph{General}
+\hi{Abstract}
+\qi{"the array's spatial covariance matrix" might be a bit shocking for the non-expert Reader --> The formulation in the Introduction might fit better "the incluence of noise ..."}
+\ai{}
+%
+\hi{Page 2}
+\qi{"6 of which were micro-steered" --> a short explanation or reference might help to understand what is meant by "micro-steered"}
+\ai{}
+%
+\qi{"it was not clear how how many" --> delete 1 x "how"}
+\ai{}
+%
+\qi{"with an element beamwidth" --> what is meant by element beamwidth? Maybe an additional figure could be used to explain this.}
+\ai{}
+%
+\hi{Page 3, after (1)}
+\qi{"Applying a window that trails off ....
+Minimum variance distortion-less response (MVDR) beamformer."--> real and complex Windows?
+The introduction should be more formal: x ist introduced as delayed data, weights or delays to change, or both is explained later.
+--> The difference between noise and sidelobes should be pointed out.}
+\ai{}
+%
+\hi{Page 5, equation (5)}
+\qi{The lateral correlation and the reason why equation (5) is a good choice should be better explained.}
+\ai{}
+%
+\hi{Page 6, last row}
+\qi{"they are optimal for any value" --> optimal in which sense?}
+\ai{}
+%
+\hi{Fig 3}
+\qi{why is a 40x40 pixel group a good choice?
+could the associated broadside direction be made visible in the figure, too?}
+\ai{}
+%
+\hi{Page 8}
+\qi{reference to Fig.1 --> could the sequence of figures be changes to have Fig. 1 closer to this paragraph}
+\ai{}
+%
+\hi{Page 11}
+\qi{where is the synthetic aperture needed/applied in this work, or is it just the name of the array? how does the work on LCA/MVDR Change the Quality of synthetic aperture?
+what exactly is the sectorscan mode?}
+\ai{}
+%
+\end{enumerate}
 
 \subsection*{Suggestions for improving the exposition of the processing}
-
-\q{I suspect many readers of the paper will not have the same assumptions as the authors, if they are not processing high-frequency narrowband signals with undersampled arrays.
-I see this disconnect between low-frequency passive processors and mid-frequency active processors, but the gap to imaging sonars is even greater.
-I venture that the authors can greatly broaden the appeal of their paper by addressing the questions I pose below.}\\[-.5\baselineskip]
-
-\ans{We thank the reviewer for these remarks, it seems we should state this better upfront.
-In the 5th paragraph of the introduction we now explain that our system is a well sampled wideband sonar and that we operate in near field conditions.}
-
-\begin{enumerate}
-\qi{Here are my calculations of how the 32 elements are spaced in the HISAS 1030 system:}
-\begin{align}
-fc &= 100\;\text{kHz} \\
-c  &= 1500\;\text{m/s} \\
-\lambda &= c/f_c = 15\;\text{mm} \\
-\textcolor{red}{ d } &= \textcolor{red}{\lambda/2 } \\
-D &= 1200\;\text{mm} \\
-\textcolor{red}{M} &= D/d = \textcolor{red}{160}
-\end{align}
-\ai{The reviewer made us realize that we should clarify this. The 32 elements are evenly spaced along the 1.2m array, so:
-\begin{align}
-M &= 32 \\
-d &= D / M = 3.75\;\text{cm} = 2.5\lambda
-\end{align}
-Further details are provided in 2. A sentence elaborating on the element spacing has been added in section IV first paragraph; where the HISAS1030 is introduced.}
-%
-\qi{What is impact of using only 32 elements? How are they spaced? It would seem that spatial aliasing would be an issue. Authors should explain what their assumptions are.}
-\ai{Appreciations for pointing this out.
-We agree that this should be clarified.
-The array configuration produce the first grating lobes at $\theta = \text{arcsin}(\lambda M/D) = 23.5^\circ$, but this is not a problem since the element opening angle is $25^\circ$ ($\theta\in[12.5^\circ,12.5^\circ]$) at $f_c=100$\;kHz.
-In the paper we perform sidescan imaging where only very narrow beams need to be beamformed.
-These details are now included in the paper, some in the 5th paragraph of the introduction and some in the first paragraph of section IV.}
-%
-\qi{What is bandwidth of transmitted waveform?
-At 100 kHz, perhaps the authors are comfortably in a narrowband regime, and they do not need to worry about array response changing over their bandwidth (which would suggest that MVDR would need to be performed at different frequencies independently, unless some sort of pre-steering was used, which may be what the authors are doing anyway).
-Authors should explain what their assumptions are.}
-\ai{We mention in the beginning of section IIA that we pre-steer to every angle and range bin in the image, but now repeat this in IIA paragraph 3, and state that we operate in wideband conditions in the introduction.}
-%
-\qi{Authors show solution of MVDR optimization as $\w = \frac{\Ri\1}{\1\T\Ri\1}$ and state this is the broadside solution.
-Are the authors pre-steering the array for every angle?
-Or are they using $\w = \frac{\Ri\s}{\s\t\Ri\s}$ at off-broadside angles, with $\s$ the conventional steering vector (which would be a frequency-domain phase-shift vector, to be applied to a narrowband signal)?
-Pre-steering would enable disparate frequencies over a wide band to be coherently summed, to provide coherent gain in the look direction (over the band) by providing additional sample support for the sample covariance estimation.}
-\ai{Yes, we are pre-steering the array for every angle.
-We have clarified this further (see 3.), and added a remark saying that while we can not use it here (due to processing a wideband signal) a phase steered version of MVDR can be obtained by introducing a steering vector $\vec s$.}
-%
-\qi{Authors time average $\x\x\H$ over $K\in\{0, 1, 2\}$ time samples.
-Without doing some sort of reduced-rank processing, why is as little as 2 samples enough to estimate the covariance?
-Perhaps this is due to high SNR?
-At high SNR, though, why are the authors doing MVDR at all?
-What interference is being canceled, and what is its spatial distribution and stationarity?
-}
-\ai{Again, a helpful remark, we will clarify this as well.
-After pulse compressing a chirp using the full transceiver bandwidth (30~kHz) range resolution is roughly $\Delta r \approx \frac{1}{B} = 2.5$~cm, which corresponds to a sample shift of just 1.2\footnote{$\Delta n = \frac{\Delta r f_s}{c} = 1.2$}.
-Hence we cover the entire pulse compressed signal with only a few samples.
-$K=2$ corresponds to a window length of $2K+1=5$ samples, as is shown by (4) in the article.
-Some edits have been made to make this clearer.}
-%
-\qi{I read the first paragraph of Section II, "..principle of adjusting the array's focus in range and bearing is commonly referred to as beamforming", and was confused - were the authors doing wavefront curvature beamforming, where every combination of range bin and beam angle required a distinct set of steering weights?
-After reading further, I would venture that the authors are steering only in angle, and that the index $n$ is over time of arrival, at the resolution of the width in time of the matched filter output (after pulse compression...
-the authors do not say anything about the waveform, so I have no idea whether there WAS any pulse compression).
-What is the resolution in time or range?
-}
-\ai{We understand how the confusion might arise.
-We operate under near field conditions and are indeed performing wavefront curvature beamforming.
-Hence, the index $n$ just an index over time at a resolution given by the sampling frequency.
-This has now been clarified.}
-\end{enumerate}
-
-\subsection{Suggestions for additional GPU details so that paper can also serve as "field report" on CPU work in general}
-
-\begin{enumerate}
-\qi{It is unfortunate that so little effort was made to speed up the CPU versions of the code.
-For example, there are public-domain versions of optimized LAPACK libraries that provide the same sort of linear algebra codes as the authors used in their CPU implementation (for example, GoToBLAS, or Intel's MKL libraries).}
-\ai{We agree, and have now included our reasons for not comparing with an optimized CPU implementation.
-One reason is that we have not been able to find any CPU batch solvers out there and lacked the time to make our own.
-In our tests our custom inline Cholesky solver beat alternatives from e.g.
-MKL as our matrices are too small to justify the extra function call overhead.
-Another reason is that to make the comparison fair we feel an equal effort would have to put into writing a covariance builder for the CPU, and to minimize data movement in each step and in between them.
-The details may be found in section V paragraph 2.}
-%
-\qi{The extra input/output of data to the GPU is a burden only on the GPU implementation, so would penalize the GPU implementation only, unless the authors were to "hide" the i/o by setting it up to run in parallel with the CPU processing, which is easier and easier with the latest GPU architectures.
-This should perhaps be pointed out.}
-\ai{Yes this is a good point. Included in last paragraph of section IV.}
-%
-\qi{The authors do not provide much information about the GPU hardware they used, the NVIDIA Quadro 6000.
-How does this compare to the desktop video cards that reside in most desktop, workstation, server and even high-end laptop systems?
-What is the vintage of the architecture of this GPU?
-How has the architecture evolved?
-For example, the authors tell us the Quadro 6000 (\$1900, 6GB VRAM, 448 cores) is a Fermi architecture - this is the previous architecture, having been superceded by the Kepler architecture, with a new Quadro K6000 board (\$3600, 12 GB VRAM, 2880 cores).
-The Quadro boards are high-end boards with video output targeting the CADAM market - there are many cheaper alternatives targeted at gamers and an even more expensive compute-only board called TESLA targeted at high performance computing.
-It's difficult to say how much of this irrelevant.}
-\ai{There is certainly a potential for comparing with many other cards here, but we decided to be thorough with one architecture rather than to compare with everything.
-To the paragraph introducing the Quadro 6000 we have explained what its equivalent consumer grade Geforce card would be.
-We have also included remarks that we expect Kepler to perform better had the design been optimized for it.}
-%
-\qi{The authors state that OpenCL is an equally attractive alternative to CUDA.
-This would be the case if the authors had been able to find something in OpenCL akin to the recently released batch-optimized linear solver code for CUDA.
-Unfortunately, this is a pervasive story - the support for scientific programming in CUDA is years ahead of OpenCL.
-The benefit of OpenCL is that it is supposed to be "cross-platform", which is true in theory, but it is hard to imagine being true in practice considering the wide range of heterogeneous computing elements OpenCL is supposed to cover; from phones to DSP chips to FPGAs.
-}
-\ai{These are important remarks, thank you.
-We have written OpenCL programs too, and run them on Intel CPUs and GPUs from AMD and Nvidia.
-It was no walk in the park, but with a satisfactory end result.
-But like you say, these are just a subset of devices that OpenCL support, and we have now changed the wording to be more specific here.
-Your remark on the lack a batched linear solver is spot in, the sentence is now rewritten.}
-\end{enumerate}
-
-\subsection{Not clear how R is being calculated in CPU}
-
-\q{
-Page 5, last paragraph in 1st column is puzzling.
-Authors say that "when a thread has finished up a diagonal, we have them wrap around to compute one of the diagonals in the lower triangular of R." The diagonals are all of different lengths, and, presumably the code to do a copy and conjugation from the upper triangular part to the lower triangular part of R requires different instructions than the moving average being calculated along a diagonal in the upper triangular part.
-The cores MUST execute the exact same code; or "diverge" (with one set executing one set of code while the complementary set sitting idle, then vice versa, as you would get with two if-then-else code sections), since this is a SIMD architecture.
-What is going on here?}
-
-\q{This requires a better explanation.}
-
-\ans{The reviewer is absolutely right.
-The copy will be serialized due to divergent branching.
-It is unfortunate that to the best of our knowledge no Cholesky based batch linear equation solver exist, which would have eliminated the need for the lower triangular.
-The paragraph has been edited to explain this better.}
-
-\subsection{Opportunities for additional GPU tutorial content?}
-
-\q{The authors describe the methodology they used to get things to run fast on the GPU.
-The analysis of the workload was informative and I think it is very informative for readers to see how the analysis progresses from less hardware specific (operations counts), to algorithm restructuring to reduce operations; to more hardware specific concerns (relative throughput of operations and memory transfers).}\\[.5\baselineskip]
-\ans{We thank the reviewer for the credit.}\\[-.5\baselineskip]
-
-\q{Let me suggest some concepts that could be discussed a little bit more to give readers more of a sense of what programming GPUs is like.}
-
-\begin{enumerate}
-\qi{What does somebody need to program in CUDA?
-Authors can mention that CUDA C is just C with a few extra language features to specify how to distribute the work across the cores.
-Authors can mention that there is a nvcc compiler, a profiler, a debugger, math libraries (CUFFT, CUBLAS), and a large number of examples of varying complexity, all from NVIDIA.
-Authors can mention that CUDA can be used on Windows, MacOSX, or Linux, with an NVIDIA graphics card.
-Authors can mention that there are also third party CUDA interfaces in Fortran and Python.
-When discussing the batch-optimized linear solver code from NVIDIA, the authors should cite the NVIDIA site, and also links to CULA and MAGMA, other CUDA versions of LAPACK or subsets thereof.}
-\ai{These are good suggestions.
-We have added some of these suggestions to the paragraph where we introduce CUDA, but regrettably had to leave some of it out due to space considerations.
-Citation to the "Nvidia Registered Developers Program" is added where we speak of the batch solver.}
-%
-\qi{Problems need to have high "arithmetic intensity" (the authors use "computational intensity") so that the cores stay busy.
-If there is not enough arithmetic, the cores sit idle waiting for data to be moved around.
-The authors do a nice job describing this situation on page 4 and the start of page 5.}
-\ai{We thank the reviewer for those kind words. In the mentioned context we agree that "arithmetic" is likely a better alternative to "computational", and will make the relevant edits.}
-%
-\qi{The authors use the term "arithmetic reduction".
-Unfortunately, reduction is an important word/topic in GPU work.
-Perhaps "minimizing arithmetic operations" would be better.
-Reduction is any N-to-1 mapping, which poses a conundrum for N cores having to combine their individual results at a single location all at the same time...
-you don't want the cores to all line up and wait their turn.}
-\ai{Yet again, very true! The reviewer's suggestion is much better, we have used it.}
-%
-\qi{On page 4, the authors talk about avoiding global memory, without getting into "coalesced" memory access patterns.
-This might merit a mention - for example, is at best only able to move one float for every 30 floats processed, and potentially much worse if care is not taken to use "coalesced memory access patterns (that avoid memory bank conflicts)".
-On page 5, the authors talk about "a collective access pattern that maximizes global memory bandwidth".
-Here too, the authors should perhaps use the "coalesced" term, since this is the term used in all of the GPGPU documentation.}
-\ai{The reviewer is right. It is better to use the "official" terms. We have applied the suggestions.}
-\end{enumerate}
-
-\subsection{Localized comments}
-
-\begin{enumerate}
-\qi{Equation (2) using a "broadside steering vector" - this is confusing...
-are the authors only steering at broadside?
-clearly not...
-do the authors pre-steer all of the elements prior to forming $\R$ so that can "focus" in a wideband sense?}
-\ai{We indeed presteer all the elements to deal with with our wideband signal.
-To make this clearer we have now stated it again after equation (2), and explained that in a narrowband case a generalization is straight forward by substituting $\1$ with a steering vector.}
-%
-\qi{Paragraph after equation (2) mentions "spatial averaging to avoid signal cancellation" - this problem might be better identified for readers by citing the Kailath reference and using "signal cancellation due to coherent multipath".}
-\ai{Good point, both suggestions are now included.}
-%
-\qi{Equation 4 has second summation using index $n'$, but this index does not appear in any of the terms inside the summation.}
-\ai{Good catch! Fixed.}
-%
-\qi{Page 2 - $\R$ is averaged over K time samples...
-at what rate is beamformer output produced?
-is $\R$ averaged using a moving window process, so that $\R[n]$ is just a rank-1 update to $\R$[n — 1] and there is one beamformer output for every input time sample?}
-\ai{Actually, $\R$ is averaged over $2K+1$ time samples, this can be seen from (4) and in the text that follows.
-There is indeed one beamformer output for every input time sample.
-We are not sure where the article is indicating otherwise...}
-%
-\qi{Page 4, in list of three optimizations - item 3 is not self-evident, but if I understand what the authors are doing, it is described by the 1985 reference by Kailath.}
-\ai{We must have conveyed this in a flawed way.
-We are simply implementing a text-book sliding window summation, where a start sum is computed and then this is updated with a new term for each element in $\eR$.
-Item 3 has now been rewritten to hopefully express this better.}
-%
-\qi{Page 4, last paragraph of 1) Arithmetic reduction - authors should identify that the "Minimized memory" version skips step 2) from list of three optimization, and that the "Minimized instructions" version uses all three steps...
-I had a hard time understanding how the three "arithmetic reduction steps" were related to the plots in Figure 5, and was puzzled about where the "minimized memory" and "minimized instruction" variants of the processing had come from.
-Perhaps the authors can introduce these variants more clearly and more prominently.}
-\ai{The reviewer is right. This is now rewritten to be more explicit.}
-%
-\qi{On page 6, authors say "The bottleneck in the final design is now the inversion step, which is typically 5 times slower than the build step".
-Surely, this merits an explanation.
-What is the final design being compared with?
-What was the expectation?
-What changed?
-Figure 3 shows FLOPs before optimization - this shows build and inversion steps of roughly the same order of magnitude in the right half of the figure.
-Does Figure 3 set the expectation?
-Figure 5 shows "relative complexity reduction" of entire pro-cess, so there is no information on relative speed of build and inversion steps.
-Figure 10 shows estimated FLOPs for the two steps, but since both steps have different theoretical complexities (operations counts), and the two steps do not optimize the same way, this does not provide an expectation of the relative times.}
-\ai{TODO}
-\qi{On page 6, authors say their implementation was improved by factor of 1.5 - 2 when run on a K20.
-Authors could say a few words about what a K20 is, how it differs from the Quadro 6000, that the Kepler architecture actually has features that are dramatically better for scientific computing if the code is adapted to use them (kernels can call kernels, more registers, more shared memory, ability to eliminate some host-to-GPU i/o).}
-\ai{Good suggestions. They are now added to the last paragraph of section V.}
-\qi{Section I says 1-2 orders of magnitude. Section IV, page 6, says 2-3 orders of magnitude.}
-\ai{Well seen. This is now corrected.}
-\qi{Reference 9 has two authors, Shan and Kailath. 11. Reference 17 seems to be the reviewed manuscript?}
-\ai{Reference 9 has been corrected, and reference 17 (now 18) has been updated to point to the correct article. }
-\end{enumerate}
 
@@
 \begin{enumerate}
-\item Fig. 3 - Slightly increased spacing between subplots.
-\item Fig. 8 - Abbreviating megapixels as "MP" instead of "MPx".
-\item Fig. 9 - Abbreviating megapixels as "MP" instead of "MPx".
-\item Fig. 11 - Abbreviating megapixels as "MP" instead of "MPx".
+\item Fig. X. Changed bla bla bla
 \end{enumerate}

### 2016-01-16 LCAforActiveSonarImaging/submission/submit1/answer/answer.tex
la2pdf commit - 2016:01:16	15:40:15	(lø.16.jan.)

@@
 
-\q{The paper is well written and addresses an actual problem. The solution however, is barely innovative. Still the results are interesting enough for publication. Some minors details.}
+\q{The paper is well written and addresses an actual problem. The solution however, is barely innovative. Still the results are interesting enough for publication. Some minors details.} \\\\
+\ans{Barely innvative? Can we argue better for why it is?}
 
@@
 
-Although the topic is potentially interesting for all readers, due to many specific terms used in the text, in the present form the article would probably only be highly appreciated by expert readers.
-I would recommend some minor changes in order to widen the frame of potential readers.
+\q{Although the topic is potentially interesting for all readers, due to many specific terms used in the text, in the present form the article would probably only be highly appreciated by expert readers.
+I would recommend some minor changes in order to widen the frame of potential readers.}
+
+\ans{Thank you for the thorough review and very constructive remarks. We believe in this philosophy and will make changes accordingly.}
 
@@
 \qi{"the array's spatial covariance matrix" might be a bit shocking for the non-expert Reader --> The formulation in the Introduction might fit better "the incluence of noise ..."}
-\ai{}
+\ai{Thank you, this is a good point. We have reformulated the sentence to speak of computing the spatial statistics instead.}
 %
@@
 \qi{"it was not clear how how many" --> delete 1 x "how"}
-\ai{}
+\ai{Thank you. Corrected.}
 %
@@
 \qi{"they are optimal for any value" --> optimal in which sense?}
-\ai{}
+\ai{Thank you. This was mentioned earlier in the same paragraph, just after (8).}
 %
 \hi{Fig 3}
-\qi{why is a 40x40 pixel group a good choice?
-could the associated broadside direction be made visible in the figure, too?}
-\ai{}
+\qi{why is a 40x40 pixel group a good choice? could the associated broadside direction be made visible in the figure, too?}
+\ai{This is an excellent question. This should perhaps be related to the detail level in the image, which in turn depends on the array's imaging performance.}
 %
@@
 \qi{reference to Fig.1 --> could the sequence of figures be changes to have Fig. 1 closer to this paragraph}
-\ai{}
+\ai{Valid point. In our two column print the figure is on the page before its first reference. }
 %
@@
 what exactly is the sectorscan mode?}
-\ai{}
+\ai{Thank you, we see that our formulation should be clarified.
+We are creating an image of the sector highlighted by a single ping of data (sectorscan), but are using data recorded by a synthetic aperture sonar (SAS).
+We mention it since SAS systems have a wide transmit beam intended to highlight multiple pings and thus does not produce the best sectorscan images.}
 %

### 2016-02-16 LCAforActiveSonarImaging/lca_formula.tex
Working on the review

\newpage
%
The angular frequency response of a window steered to $\phi'$ ($\lambda$ disappears if we express element distance $d$ relative to the wavelength):
%
\begin{align*}
W_{M,d,\beta,\phi'}(\phi) &= \sum\limits_{m=0}^{M-1} w_{m,M,d,\beta,\phi'}\, e^{j\frac{2\pi}{\lambda}md\sin\phi} \\
&= \sum\limits_{m=0}^{M-1} e^{-j\frac{2\pi}{\lambda}md\sin\phi'}\,f_{m,M}(\beta)\,e^{j\frac{2\pi}{\lambda}md\sin\phi} \\
&= \sum\limits_{m=0}^{M-1} f_{m,M}(\beta)\,e^{j\frac{2\pi}{\lambda}md\big[\sin\phi'-\sin\phi\big]}
\end{align*}
%
Or, expressed as a function of wavenumber $k_x = \|\vec k\|\sin\phi = \frac{2\pi}{\lambda}\sin\phi$:
%
\begin{align*}
W_{M,d,\beta,k_x'}(k_x) &= \sum\limits_{m=0}^{M-1} w_{m,M,d,\beta,\phi'}\, e^{jk_xmd} \\
&= \sum\limits_{m=0}^{M-1} e^{-jk_x'(\phi) md}\,f_{m,M}(\beta)\,e^{jk_xmd} \\
&= \sum\limits_{m=0}^{M-1} f_{m,M}(\beta)\,e^{j\big[k_x'(\phi)-k_x\big]md}
\end{align*}
%
What I want to know is what steering $k_x'$ that causes the amplitude to be -3dB at some predetermined location $k_x^0$, under the constraint that weights always sum to 1 (for unit gain at $k_x=0$):
%
\begin{align*}
\Big|W_{M,d,\beta,k'}(k_x^0)\Big| &= \frac{1}{\sqrt{2}} \quad\quad\text{subject to:}\quad\quad \sum\limits_{m=0}^{M-1} w_{m,M,d,\beta,\phi'} = 1.
\end{align*}
%
Is this possible to solve somehow? I don't mind if it needs to be a numeric solution or an approximation. I need it in my article and some kind of formula would be better than the algorithm I have now that performs a search to get the answer. I was wondering if the equation can be separated into a term/factor that independently depends on $M$, $d$ and $\beta$ somehow, then numerically solve each and create a compounded equation/approximation of the results, but my math skills are struggling :P

\newpage

### 2016-02-16 LCAforActiveSonarImaging/lca_rev1.tex
Working on the review

--- 2015-10-21 LCAforActiveSonarImaging/lca.tex
@@
-\synctex=1
 %%%%%%%%%%%%%%%%                                ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
@@
 \newif\ifTODO\TODOtrue                        % Use todo notes?
-
-\newif\ifBuildBibliography\BuildBibliographytrue  % Whether to include the references.bib file or use inline references.
 
@@
 
-% \newcommand\mediaPath{gfx/media}
-\newcommand\mediaPath{http://folk.uio.no/joibu/media/2015_JOE_LCA}
-
-\newcommand\mediaI{\multimedia{\mediaPath/media1.mp4}{Media Movie~1}}
-\newcommand\mediaII{\multimedia{\mediaPath/media2.mp4}{Media Movie~2}}
+\newcommand\mediaPath{gfx/media}
+\newcommand\mediaI{\multimedia{\mediaPath/media1.mp4}{Media Movie 1}}
+\newcommand\mediaII{\multimedia{\mediaPath/media2.mp4}{Media Movie 2}}
 
@@
 
-\title{Low Complexity Adaptive Sonar Imaging}
+\title{Low Complexity Adaptive Beamformer\\ for Active Sonar Imaging}
 
@@
 
-We show that a robust MVDR implementation typically creates windows spanning between the rectangular and Hamming.
-We let LCA select from a set of Kaiser windows with responses in this span, and add some steered variations of each.
+We show that a robust MVDR implementation typically creates responses ranging between that of the rectangular and Hamming.
+We let LCA select from a set of Kaiser windows with responses in this range, and add some steered variations of each.
 We limit the steering to roughly half the -3\,dB width of the window's amplitude response.
@@
 Slightly smoother images are produced if the window count is increased to 15, but past this we observe minimal difference.
-Finally we show that LCA works just as well if Kaiser windows are substituted with trigonometric ones.
-
-All our observations and experiences point to LCA being very easy to understand and manage.
+Finally we show that LCA work just as well if Kaiser windows are substituted with trigonometric ones.
+
+All our observations and experiences point to LCA being very easy to manage.
 It simply works, and is surprisingly insensitive to the exact type of window function, steering amount, or number of windows.
@@
 \end{IEEEkeywords}
-\fi
-}
+\fi}
 % \fi
@@
 Bats, for instance, use a high frequency sonar to detect, identify and track prey with amazing precision in caves amongst a myriad of other bats.
-Active human made sonars have little to no adaptability on transmit, but several adaptive methods have been investigated to post process the received data.
+Active human made sonars have little to no adaptability on transmit, but several adaptive methods have been investigated post process the received data.
 Perhaps the best studied method for reconstructing a sonar image is the minimum variance distortion-less response (MVDR) beamformer~\cite{Capon1969}.
-It adapts the sonar array's spatial response to minimize the influence of noise and interference in the final image.
-
-In many cases MVDR can improve the contrast and resolution of active sonar images compared to conventional static methods~\cite{Blomberg2013,Blomberg2012a,Lo2004}.
+It adapts the sonar array's response to minimize the influence of noise and interference in the final image.
+
+In many cases MVDR can improve the contrast and resolution of active sonar images compared to conventional static methods~\cite{Blomberg2013,Blomberg2012a,Dursun2009,Lo2004}.
 However, while the computational complexity of conventional beamformers are linear with the number of channels, O($M$), MVDR is at O($M^3$).
-This is because MVDR relies on estimating and inverting a covariance matrix.
-In sonar systems with less than 32 channels the complexity can be reduced and the implementation accelerated significantly using graphics computing units (GPUs)~\cite{Buskenes2014}.
-For larger systems the inversion step dominates the computation.
-This can be dealt with by approximating the full covariance matrix with a small one using space reduction techniques such as beamspace processing~\cite{Asen2013}, or the closely related principal component analysis (PCA) method~\cite{Kim2014}.
-Another alternative is to assume spatial stationarity to form more easily invertible Toepliz matrices~\cite{Asl2012}.
-However, the  mentioned methods are still relatively slow compared to DAS.
-
-A much faster and simpler alternative to MVDR is the Low Complexity Adaptive (LCA) beamformer.
-Based on an idea by Vignon~\cite{Vignon2008}, it was first introduced by Synnev\aa{}g \emph{et al}.
-in clinical medical imaging~\cite{Synnevag2008}, who demonstrated its ability to obtain very similar image quality to MVDR in systems with focused transmit beams.
-LCA applies a set of predefined windows, or apodization weights, and selects the one offering the best sidelobe suppression.
-It relies on the same optimization criterion as MVDR, thus may be considered a version of MVDR with a reduced window solution space.
-Alternatively, it may be viewed as a multi-apodization technique.
-One such method is described by Stankwitz \emph{et al}.
-in radar imaging~\cite{Stankwitz1995}, where the best out of 2 or 3 windows are selected.
-However, LCA is different in that it typically selects from a larger pool of windows and allows micro-steered variations of them.
-This gives LCA an adaptive freedom that is somewhere in between that of traditional multi-apodization techniques and MVDR.
-
-Synnev\aa{}g \emph{et al}.
-suggested using a window set comprised of rectangular, Kaiser (or Kaiser-Harris) and inverted Kaiser functions.
-In total he made LCA choose from 12 unique windows, 6 of which were micro-steered variants of a fairly wide Kaiser window, and with the rectangular window included.
-This was motivated by the desire to use wide windows with maximum sidelobe suppression near the receiver, and narrow windows at greater depths for maximum sensitivity and depth penetration.
-While achieving good results, it was not made clear how how many windows LCA needs, or how they ideally should be steered.
-Also, it is not apparent how LCA performs in sonar systems where the transmit beam is unfocused.
-
-In this study we apply LCA to active sonar imaging.
-We use variations of the parametric Kaiser window, but also compare it to the trigonometric window function.
-We extend Synnev\aa{}g's work by steering all our windows and not just some, and demonstrate how LCA's performance is affected by type and number of windows included.
-Our results are obtained using experimental data from the 32 element Kongsberg Maritime HISAS 1030 synthetic aperture sonar (SAS), which operates at 100\,kHz with an element beamwidth of 23$^\circ$.
-
-Our results show that MVDR, when adapted to work in active sonar imaging, computes weights with responses that tend to be symmetric and steered within a fraction of the window's -\,3dB resolution.
+This makes MVDR impracticable and sometimes infeasible.
+We have shown in previous studies that MVDR's complexity can be reduced, and its implementation accelerated significantly using graphics computing units (GPUs)~\cite{Buskenes2014,Asen2013}.
+However, while MVDR relies on estimating and inverting a spatial covariance matrix, there exist a method that avoid this altogether.\todo{samme avsnitt?}
+
+This method is called the Low Complexity Adaptive (LCA) beamformer.
+It was first proposed by Synnev\aa{}g in clinical medical imaging~\cite{Synnevag2008}, who demonstrated its ability to obtain very similar results to MVDR in a system with focused transmit beams.
+LCA applies a set of predefined windows and selects the one offering the best noise suppression.
+It relies on the same optimization criterion as MVDR, and may be considered a version of MVDR with a reduced window solution space.
+Alternatively, it may be viewed as a delay-and-sum (DAS) beamformer with an adaptive extension.
+
+Synnev\aa{}g used a window set comprised of rectangular, Kaiser and inverted Kaiser functions\footnote{May also be called Kaiser-Harris windows.}.
+He made LCA choose from 12 unique windows, 6 of which were micro-steered variants of a fairly wide Kaiser window.
+While achieving good results, it was not made clear how LCA performs with more or fewer windows, and how steering affects it all.
+
+We use LCA with variations of the Kaiser window, but a comparison to using trigonometric windows will also be provided.
+We extend Synnev\aa{}g's work by steering all our windows and not just some, and elaborate on the type of Kaiser windows to include and the method of steering.
+Then the method is applied to experimental data from the 32 element Kongsberg Maritime HISAS 1030 synthetic aperture sonar (SAS), with an operating frequency of 100\,kHz and transmit angle of 25$^\circ$.
+
+Our results show that MVDR computes weights with responses that tend to be symmetric and steered within a fraction of the window's -\,3dB resolution.
 When letting LCA choose from Kaiser windows with similar responses, we obtain images with improved noise suppression and resolution.
@@
 Overall, LCA produces images that are similar to MVDR, but with a resolution in between that of DAS and MVDR.
-LCA is also inherently robust, easy to implement, and fairly easy to understand.
+LCA is also inherently robust, easy to implement and fairly easy to understand.
 
@@
 Finally we add some remarks on using the trigonometric window instead of Kaiser in \ref{sec:lca_trigonometric}, and of of sampling considerations in \ref{sec:lca_oversampling}.
-Results and discussion are provided in Section \ref{sec:results_discussion}, where we study MVDR's windows to find suitable LCA ones in \ref{sec:results_lca_window_function}, assess the range of window types and steering in \ref{sec:results_window_parameters}, determine the number of windows needed in \ref{sec:results_database_size}, and add a note on computational complexity in \ref{sec:results_complexity}.
-A conclusion is provided in \ref{sec:conclusion}.
+Results and discussion a provided in Section \ref{sec:results_discussion}, where we study MVDR's windows to find suitable LCA ones in \ref{sec:results_lca_window_function}, assess the range of window types and steering in \ref{sec:results_window_parameters}, determine the number of windows needed in \ref{sec:results_database_size}, and add a note on computational complexity in \ref{sec:results_complexity}.
+Finally we conclude in \ref{sec:conclusion}.
 
@@
 
-A sonar image is formed by estimating source locations and amplitudes.
-For this purpose a spatial bandpass filter is applied to the backscattered wavefield data.
-The filter is called an array processor or receive beamformer.
+To form a sonar image we need to estimate source locations and amplitudes.
+For this purpose we apply a spatial bandpass filter to the backscattered wavefield data.
+The filter is called an array processor, or receive beamformer.
 The basic principle is to apply delay and weights to the sensor channels before summing them up, chosen such that signals from the location of interest are summed coherently, while other sources sum incoherently.
@@
 Assume that the wavefield has been sampled by an $M$ element uniform linear array, and that the signal signature has been removed by a matched filter.
-Let $x_m[\theta,n]$ be the delayed data from the $m$th channel, where the $\theta$ and $n$ are the azimuth angle and range sample of the focus point, respectively.
+Let $x_m[\theta,n]$ be the delayed data from the $m$th channel, where the $\theta$ and $n$ is the azimuth angle and range sample of the focus point, respectively.
 Each angle $\theta$ will be processed independently, so to simplify notation we will assume the dependence on $\theta$ to be implicit from now on.
@@
 where $w_m$ is the weight factor assigned to channel $m$.
-A weight set is commonly called a window, apodization or taper function.
+A weight set is commonly called a window or a taper function.
 Applying a window that trails off towards the edges improves noise suppression at the cost of reduced resolution~\cite{Harris1978}.
@@
 
-MVDR seeks to minimize the output variance of the beamformer, under the constraint of a unity gain in some desired direction $\phi$~\cite{Capon1969}.
-With the assumption of zero mean data this can be formulated as:
+MVDR seeks to minimize the output power of the beamformer, under the constraint of a unity gain in some desired direction $\phi$~\cite{Capon1969}:
 %
@@
 \underset{\w[n]}{\argmin}\, E\{\big|z[n]\big|^2\} &= \underset{\w[n]}{\argmin}\, \w[n]\R[n]\w\H[n]\nn
-\text{subject to } \w^H[n]\a_\phi &= 1.\label{eq:mvdr_definition}
-\end{align}
-%
-Here $\a_\phi$ is a steering vector, $E\{\cdot\}$ is the expectation operator, and $\R[n] = E\{\x[n]\x\H[n]\}$ is the spatial covariance matrix for the full array.
-This is a convex optimization problem with the solution:
+\text{subject to } \w[n]\a_\phi &= 1.\label{eq:mvdr_definition}
+\end{align}
+%
+Here $\a_\phi$ is a steering vector, $E\{\cdot\}$ is the expectation operator, and $\R[n] = E\{\x\x\H\}$ is the spatial covariance matrix. This is a convex optimization problem with the solution:
 %
 \begin{gather}
-\vec w[n] = \frac{\Ri[n]\a_\phi}{\a_\phi\T\Ri[n]\a_\phi}.\label{eq:mvdr_weights}
+\vec w[n] = \frac{\Ri[n]\a_\phi}{\a_\phi\T\Ri[n]\a_\phi},\label{eq:mvdr_weights}
 \end{gather}
+where $\R=E\{\x[n]\x\H[n]\}\in\mathbb{C}^{M,M}$ is the spatial covariance matrix for the full array.
 The problem lies in estimating and inverting this spatial covariance matrix.
-We have described the exact steps in~\cite{Buskenes2014}: To avoid signal cancellation we apply spatial averaging by computing a mean covariance matrix from a set of subarrays with length $L$~\cite{Kailath1985}, for true speckle statistics we perform temporal averaging over $N_k = 2K+1$ temporal samples~\cite{Synnevag2009a}, and to improve robustness to parameter errors we add {\large$\epsilon$} percent diagonal loading~\cite{Cox1987,Maksym1979}.
+We have described the exact steps in~\cite{Buskenes2014}: To avoid signal cancellation we apply spatial averaging by computing a mean covariance matrix from a set of subarrays with length $L$~\cite{Kailath1985}, for true speckle statistics we perform temporal averaging over $N_k = 2K+1$ temporal samples~\cite{Synnevag2009a}, and to improve robustness to parameter errors we add $d$ percent diagonal loading~\cite{Cox1987,Maksym1979}.
 These steps are also needed to ensure that the covariance matrix is numerically well conditioned and hence invertible.
 
-\begin{figure}[tbp]\centering%
-\ifPeerReview%
-\includegraphics[width=.5\linewidth]{gfx/calc_kaiser_3dB.eps}%
-\else%
+\begin{figure}[tbp]%
 \includegraphics[width=\linewidth]{gfx/calc_kaiser_3dB.eps}%
-\fi%
 \caption{Each Kaiser window is steered in the interval $\phi=[0, \phi_{\text{-3dB}}(\beta)]$.
@@
 
-\begin{figure}[tbp]\centering%
-\ifPeerReview%
-\includegraphics[width=.5\linewidth]{gfx/cross_tagged.png}%
-\else%
+\begin{figure}[tbp]%
 \includegraphics[width=\linewidth]{gfx/cross_tagged.png}%
-\fi%
-\caption{One of the objects used to test the beamformers is a 1\,m by 1\,m resolution test cross attached to an anchor with a diameter of approximately 13\,cm.
+\caption{One of the objects used to test the beamformers is a 1\,m by 1\,m cross attached to an anchor with a diameter of approximately 13\,cm.
 Source image curtesy of Bundeswehr Technical Center for Ships and Naval Weapons, Maritime Technology and Research (WTD 71).}\label{cross}
@@
 \includegraphics[width=\textwidth]{gfx/oversampling_mosaic_bounds.pdf}%
-\caption{\emph{Determining Kaiser parameter boundaries}: LCA images of the resolution test cross and anchor created with a large and equal size window database, but where the upper bounds for $\beta$ and $\phi$ is varied.
+\caption{\emph{Determining Kaiser parameter boundaries}: LCA images created with a large and equal size window database, but where the upper bounds for $\beta$ and $\phi$ is varied.
 The upper left image is equal to DAS with a rectangular window.
-For all images we measure the lateral distance $\Delta x$ between the two -3dB points of the anchor (red object).
+For all images we measure the lateral distance $\Delta x$ between the two -3dB points of the anchor.
 These are specified relatively to the reference distance  $\Delta x_\text{ref}$ from the upper left DAS image.
@@
 Adding more steering-variations improves both sidelobe suppression and resolution.
-However, using more than  $N_\beta=2$ window types and $N_\phi=3$ steering angles makes minimal difference.}\label{oversampling_mosaic}
+However, using more than  $N_\beta=2$ window types and $N_\beta=3$ steering angles makes minimal difference.}\label{oversampling_mosaic}
 \end{figure*}
@@
 \begin{figure*}[t]\centering
-\subfloat[\emph{Image quality of approximate point scatterers found in the resolution test cross scene}.
+\subfloat[\emph{Image quality of approximate point scatterers found in the cross scene}.
 Two lateral image cuts are shown in the leftmost figures, along with the response of a rectangularly weighted DAS.
@@
 
-In practice, we estimate the beamformer output power by computing a sample power average $s_z^2$:
+In practice, we estimate the beamformer output power as
 %
 \begin{align*}
-E\big\{\big|z[n]\big|^2\big\} \approx s_z^2 = \frac{1}{N_k} \sumb{n'=n-K}{n+K} \big| z[n'] \big|^2
+E\big\{\big|z[n]\big|^2\big\} = \frac{1}{N_k} \sumb{n'=n-K}{n+K} \big| z[n'] \big|^2
 \end{align*}
@@
 This is the same temporal averaging method we use for MVDR.
-In our case the bandwidth of our matched filtered signal is a just few samples long, hence $K=1$ will be used for both MVDR and LCA throughout this work.
-
-Since pixels are correlated laterally, we can include a weighted combination of these as well to improve the variance estimation:
-%
-\begin{align}
-s_z^2 = \frac{1}{N_x N_k} \sumb{\nmat{\scriptstyle x'=\\[-2mm]\scriptstyle x-X}}{x+X} \sumb{\nmat{\scriptstyle n'=\\[-2mm]\scriptstyle n-K}}{n+K} \omega[x',n']\big| z[x',n'] \big|^2
-\end{align}
-%
-where $N_x = 2X+1$ is the number of azimuth lines to average over, and $\omega[x',n']$ is a normalized 2 dimensional weight function.
-However, we observed no visual improvement doing this, and the effect of micro-steered windows is hard to predict.
-Henceforth no lateral averaging will be applied.
+In our case the matched filtered signal is a just few samples long, hence $K=1$ will be used for both MVDR and LCA throughout this work.
+
+\todo{Drop this paragraph altogether?}If extra flexibility is desired, we can also estimate the variance by a weighted combination of local pixels
+%
+\begin{align}
+E\big\{\big|z[x,n]\big|^2\big\} = \frac{1}{N_x N_k} \sumb{\nmat{\scriptstyle x'=\\[-2mm]\scriptstyle x-X}}{x+X} \sumb{\nmat{\scriptstyle n'=\\[-2mm]\scriptstyle n-K}}{n+K} w[x',n']\big| z[x',n'] \big|^2
+\end{align}
+%
+where $N_x = 2X+1$ is the number of azimuth lines to average over, and $w_\text{prf}[x',n']$ is a normalized 2 dimensional weight function. 
 % 
@@
 %
-The Kaiser-Bessel window is near optimal in the sense of having its peak energy concentration around $\theta=0^\circ$, for a given space-bandwidth product related to the Kaiser parameter $\beta$ as:
+The Kaiser-Bessel window is near optimal in the sense of having its peak energy concentration around $\theta=0^\circ$, for a given time-bandwidth product specified with the parameter $\beta$:
 %
@@
 %
-where $T$ in our case is the spatial extent of the window and $B$ is its bandwidth.
+where $T$ is the extent of the window in time\todo{Should perhaps say space-bandwidth product?} and $B$ is its bandwidth.
 Adjusting $\beta$ changes the trade-off between mainlobe width and sidelobe level.
 When $\beta=0$ the window becomes rectangular, while at large values ($\beta>5$) the window converges to a Gaussian both in time and frequency.
-This class of windows is generally considered well suited for separating closely spaced sources with amplitudes of a high dynamic range~\cite{Harris1978}, they are easy to make, and they are optimal for any value of $\beta$.
+This class of windows is generally considered well suited for separating closely spaced\todo{better word?} sources with amplitudes of a high dynamic range~\cite{Harris1978}, they are easy to make, and they are optimal for any value of $\beta$.
 
@@
 
-The ultimate goal of adaptive beamformers is to take the information available in the wavefield and use it to either improve image resolution, sidelobe suppression, or preferably both.
+The ultimate goal of adaptive beamformers is to take the information available in the wavefield and use it to either improve image resolution, noise suppression, or both.
 This non-linear processing increases the image bandwidth and introduces a need for a sampling rate higher than the Nyquist rate.
@@
 
-To test the performance of the MVDR and LCA beamformers, we have processed data acquired by the Kongsberg Maritime HUGIN AUV carrying their 32 element HISAS1030 sonar~\cite{Hansen2011}.
-It is a high resolution synthetic aperture sonar with 1.2\;m array length, 100\;kHz operating frequency, 30\;kHz bandwidth and 23$^\circ$ receiver element beamwidth.
+To test the performance of the MVDR and LCA beamformers, we have processed data acquired by the Kongsberg Maritime HUGIN AUV carrying their 32 element HISAS1030 sonar~\cite{Hansen2009}.
+It is a high resolution synthetic aperture sonar with 1.2\;m array length, 100\;kHz operating frequency, 40\;kHz bandwidth and 25$^\circ$ opening angle on transmission.
 
@@
 This data was collected by the Norwegian Defense Research Establishment and Kongsberg Maritime.
-The other scene is of an 1\,m by 1\,m iron resolution test cross, connected to a an anchor with 13\,cm diameter (\Fig{cross}).
+The other scene is of an 1\,m by 1\,m iron cross, connected to a an anchor with 13\,cm diameter (\Fig{cross}).
 This data was collected by the Bundeswehr Technical Center for Ships and Naval Weapons, Maritime Technology and Research (WTD 71).
@@
 
-In the image reconstruction we have run MVDR with subarray length $L=16$ and $\text{\large$\epsilon$}=1\%$ diagonal loading.
+In the image reconstruction we have run MVDR with subarray length $L=16$ and $d=1\%$ diagonal loading\todo{also used for element distance}.
 Both MVDR and LCA was run with $K=1$ temporal averaging.
-This is a fairly aggressive yet stable set of parameters~\cite{Synnevag2009a}.
+This is a fairly aggressive yet stable set of parameters.\todo{suitable ref?}
 
@@
 
-Assuming that a robust MVDR is the reference method we want LCA to perform similarly to, it seems sensible to create a window database for LCA with spatial responses similar to the ones that MVDR computes.
+Assuming that a robust MVDR is the reference method we want LCA to perform similarly to, it seems sensible to create a window database for LCA with with responses similar to the ones that MVDR computes.
 We study this in \Fig{mvdr_selected_windows}, where we present the typical amplitude and phase responses that MVDR creates for shadow, speckle and highlight regions of the Holmengraa scene.
@@
 In each image we computed the lateral distance $\Delta x$ between the \minus{}3dB points of the anchor, with the reference $\Delta x_\text{ref}$ being the rectangularly weighted DAS image.
-A fourth order polynomial fit was used in this computation.
+A forth order polynomial fit was used in this computation.
 This measure is also commonly called the full width half maximum (FWHM).
@@
 
-\Fig{oversampling_mosaic_bounds_beta} and \Fig{oversampling_mosaic_bounds_phi} illustrates which $\beta$ and $\phi$ values LCA prefers for different regions in the image of the resolution test cross.
+\Fig{oversampling_mosaic_bounds_beta} and \Fig{oversampling_mosaic_bounds_phi} illustrates which $\beta$ and $\phi$ values LCA prefers for different regions in the image of the cross.
 Red marker lines are used to pinpoint the cut line, location of the anchor and strong scatters on the cross.
@@
 Two lateral image cuts through all images are presented in the left plot.
-The images and their corresponding cuts are nearly identical for the LCA version with Kaiser windows and the one with trigonometric windows.
+The images and their corresponding cuts are near identical for the LCA version with Kaiser windows and the one with trigonometric windows.
 Compared to DAS the LCA produces targets with an FWHM that lies in between that of DAS and MVDR.
@@
 
-LCA seeks to improve image resolution and contrast by minimizing the power of noise and interference in each image pixel.
-This is the same optimization criterion as MVDR uses, but instead of computing an optimal array window like MVDR does, it selects the best window out of a predefined set.
-Hence, it can be viewed as either a reduced and discrete window space version of MVDR, or as an advanced multi-apodization technique.
-What we have studied in this paper is which windows we should ideally let LCA choose from.
-
-We have found the LCA to be surprisingly insensitive to the exact window function it is used with, supported by the observation that near identical images were produced whether we used Kaiser or trigonometric windows.
-However, it needs to be able to select between wide and narrow window responses, since the wide ones offer the best sidelobe suppression and the narrow ones offer the best sensitivity.
-We achieved good results using 2-3 Kaiser windows in the range $\beta\in[0,5]$, and observed minimal difference when adding more windows or adjusting the upper bound of $\beta$.
-For the steering angle we suggest a value in the range $\phi\in[0,\pm\delta\phi_\text{3dB}(\beta)]$, where $\phi_\text{3dB}(\beta)$ is the bandwidth of the respective window, and $\delta$ is a scaling parameters.
-Setting $\delta$ to either 50\% or 100\% reduces the lateral point target size to 73\% or 53\% compared to the rectangular window, respectively.
-Hence, it can be used to control aggressiveness and resolution gain of LCA.
-
-In summary, LCA is an attractive alternative to other adaptive beamformers due to being very fast, simple to understand, practically parameter-free, inherently robust, and able produce images similar to that of MVDR.
-Contrary to MVDR, it has a low computational complexity and can be easily and efficiently accelerated using GPUs.
-Since LCA performs just as well if we used trigonometric windows instead of Kaiser, we can even compute the optimal window parameter analytically for any given steering angle, and avoid the search over the window parameter entirely.
-
-% The tendency of LCA to produce images of similar quality as MVDR seems to indicate that a full solution space is rarely needed in real world scenarios.
-
-%  It initially very is also very fast, and is very well suited for acceleration using  very fast, easy to implement even on parallel hardware,
-
-% Its computational complexity is only of O($N_wM$), and its data parallel nature makes it very well suited for acceleration using e.g. GPUs. compared to MVDR's O($M^3$).
+LCA iterates through a set of windows in search of the one that minimizes the noise and interference power in each image pixel.
+It can be viewed as either a reduced solution space version of MVDR, or as an multi-taper extension to DAS.
+It creates images similar to MVDR, but at a fraction of the computational cost.
+It is also inherently robust and insensitive to the particular set of windows it selects from, supported by the observation that LCA can be used with either Kaiser or trigonometric windows and produce near identical images.
+
+When using LCA with Kaiser windows we made their responses similar to those typically chosen by a robust MVDR in key areas in the image.
+We suggest an upper bound for $\beta$ between 2 and 5, where 5 is slightly better at suppressing sidelobes.
+For $\phi$ we suggest to steer either to 50\% or 100\% of the \minus{}3\,dB width, which leads to a resolution of 73\% or 54\% of the rectangular window, respectively.
+As lower boundaries we set $\beta=0$ to always include a rectangular window, and $\phi=0$ to always include the centered window.
+
+The tendency of the LCA to perform similarly to the MV beamformer seems to indicate that a full solution space is rarely needed in real world scenarios.
+
 % 
@@
 
-The authors would like to thank Kongsberg Maritime, the Norwegian Defence Research Establishment (FFI), and Bundeswehr Technical Center for Ships and Naval Weapons, Maritime Technology and Research (WTD 71), Germany, for the equipment and effort needed to acquire data of the resolution test cross and Holmengraa, as well as for allowing us to use it.
+The authors would like to thank Kongsberg Maritime, the Norwegian Defence Research Establishment (FFI), and Bundeswehr Technical Center for Ships and Naval Weapons, Maritime Technology and Research (WTD 71), for the equipment and effort needed to acquire data of the cross and Holmengraa, as well as for allowing us to use it.
 
@@
 
-\ifBuildBibliography
 \bibliographystyle{IEEEtran}
 \bibliography{references}
-
-\else
-
-% Generated by IEEEtran.bst, version: 1.13 (2008/09/30)
-\begin{thebibliography}{10}
-\providecommand{\url}[1]{#1}
-\csname url@samestyle\endcsname
-\providecommand{\newblock}{\relax}
-\providecommand{\bibinfo}[2]{#2}
-\providecommand{\BIBentrySTDinterwordspacing}{\spaceskip=0pt\relax}
-\providecommand{\BIBentryALTinterwordstretchfactor}{4}
-\providecommand{\BIBentryALTinterwordspacing}{\spaceskip=\fontdimen2\font plus
-\BIBentryALTinterwordstretchfactor\fontdimen3\font minus
-  \fontdimen4\font\relax}
-\providecommand{\BIBforeignlanguage}[2]{{%
-\expandafter\ifx\csname l@#1\endcsname\relax
-\typeout{** WARNING: IEEEtran.bst: No hyphenation pattern has been}%
-\typeout{** loaded for the language `#1'. Using the pattern for}%
-\typeout{** the default language instead.}%
-\else
-\language=\csname l@#1\endcsname
-\fi
-#2}}
-\providecommand{\BIBdecl}{\relax}
-\BIBdecl
-
-\bibitem{Capon1969}
-J.~Capon, ``{High-resolution frequency-wavenumber spectrum analysis},''
-  \emph{Proceedings of the IEEE}, vol.~57, no.~8, pp. 1408--1418, 1969.
-
-\bibitem{Blomberg2013}
-A.~E.~A. Blomberg, A.~Austeng, R.~E. Hansen, and S.~A.~V. Synnes, ``{Improving
-  Sonar Performance in Shallow Water Using Adaptive Beamforming},'' \emph{IEEE
-  Journal of Oceanic Engineering}, vol.~38, no.~2, pp. 297--307, Apr. 2013.
-
-\bibitem{Blomberg2012a}
-A.~E.~A. Blomberg, A.~Austeng, and R.~E. Hansen, ``{Adaptive Beamforming
-  Applied to a Cylindrical Sonar Array Using an Interpolated Array
-  Transformation},'' \emph{IEEE Journal of Oceanic Engineering}, vol.~37,
-  no.~1, pp. 25--34, Jan. 2012.
-
-\bibitem{Lo2004}
-K.~Lo, ``{Adaptive Array Processing for Wide-Band Active Sonars},'' \emph{IEEE
-  Journal of Oceanic Engineering}, vol.~29, no.~3, pp. 837--846, Jul. 2004.
-
-\bibitem{Buskenes2014}
-J.~I. Buskenes, J.~P. Asen, C.-I.~C. Nilsen, and A.~Austeng, ``{An Optimized
-  GPU Implementation of the MVDR Beamformer for Active Sonar Imaging},''
-  \emph{IEEE Journal of Oceanic Engineering}, pp. 1--13, 2014.
-
-\bibitem{Asen2013}
-J.~P. \AA{}sen, J.~I. Buskenes, C.-I. {Colombo Nilsen}, A.~Austeng, and
-  S.~Holm, ``{Implementing capon beamforming on a GPU for real-time cardiac
-  ultrasound imaging.}'' \emph{IEEE transactions on ultrasonics,
-  ferroelectrics, and frequency control}, vol.~61, no.~1, pp. 76--85, Jan.
-  2014.
-
-\bibitem{Kim2014}
-K.~Kim, S.~Park, J.~Kim, S.-B. Park, and M.~Bae, ``{A fast minimum variance
-  beamforming method using principal component analysis},'' \emph{IEEE
-  Transactions on Ultrasonics, Ferroelectrics, and Frequency Control}, vol.~61,
-  no.~6, pp. 930--945, Jun. 2014.
-
-\bibitem{Asl2012}
-B.~M. Asl and A.~Mahloojifar, ``{A low-complexity adaptive beamformer for
-  ultrasound imaging using structured covariance matrix},'' \emph{IEEE
-  Transactions on Ultrasonics, Ferroelectrics and Frequency Control}, vol.~59,
-  no.~4, pp. 660--667, Apr. 2012.
-
-\bibitem{Vignon2008}
-F.~Vignon and M.~R. Burcher, ``{Capon beamforming in medical ultrasound imaging
-  with focused beams.}'' \emph{IEEE transactions on ultrasonics,
-  ferroelectrics, and frequency control}, vol.~55, no.~3, pp. 619--28, Mar.
-  2008.
-
-\bibitem{Synnevag2008}
-J.-F. Synnev\aa{}g, S.~Holm, and A.~Austeng, ``{A low complexity data-dependent
-  beamformer},'' in \emph{2008 IEEE Ultrasonics Symposium}, no.~2.\hskip 1em
-  plus 0.5em minus 0.4em\relax IEEE, Nov. 2008, pp. 1084--1087.
-
-\bibitem{Stankwitz1995}
-H.~Stankwitz, R.~Dallaire, and J.~Fienup, ``{Nonlinear apodization for sidelobe
-  control in SAR imagery},'' \emph{IEEE Transactions on Aerospace and
-  Electronic Systems}, vol.~31, no.~1, pp. 267--279, Jan. 1995.
-
-\bibitem{Harris1978}
-F.~Harris, ``{On the use of windows for harmonic analysis with the discrete
-  Fourier transform},'' \emph{Proceedings of the IEEE}, vol.~66, no.~1, pp.
-  51--83, 1978.
-
-\bibitem{Kailath1985}
-T.~Kailath and T.-J. Shan, ``{Adaptive beamforming for coherent signals and
-  interference},'' \emph{IEEE Transactions on Acoustics, Speech, and Signal
-  Processing}, vol.~33, no.~3, pp. 527--536, Jun. 1985.
-
-\bibitem{Synnevag2009a}
-J.-F. Synnev\aa{}g, A.~Austeng, and S.~Holm, ``{Benefits of minimum-variance
-  beamforming in medical ultrasound imaging},'' \emph{IEEE Transactions on
-  Ultrasonics, Ferroelectrics and Frequency Control}, vol.~56, no.~9, pp.
-  1868--1879, Sep. 2009.
-
-\bibitem{Cox1987}
-H.~Cox, R.~Zeskind, and M.~Owen, ``{Robust adaptive beamforming},'' \emph{IEEE
-  Transactions on Acoustics, Speech, and Signal Processing}, vol.~35, no.~10,
-  pp. 1365--1376, Oct. 1987.
-
-\bibitem{Maksym1979}
-J.~N. Maksym, ``{A robust formulation of an optimum cross-spectral beamformer
-  for line arrays},'' \emph{The Journal of the Acoustical Society of America},
-  vol.~65, no.~4, p. 971, 1979.
-
-\bibitem{Asen2014}
-J.~P. \AA{}sen, A.~Austeng, and S.~Holm, ``{Capon beamforming and moving
-  objects--an analysis of lateral shift-invariance.}'' \emph{IEEE transactions
-  on ultrasonics, ferroelectrics, and frequency control}, vol.~61, no.~7, pp.
-  1152--60, Jul. 2014.
-
-\bibitem{Hansen2011}
-R.~E. Hansen, H.~J. Callow, T.~O. Sabo, and S.~A.~V. Synnes, ``{Challenges in
-  Seafloor Imaging and Mapping With Synthetic Aperture Sonar},'' \emph{IEEE
-  Transactions on Geoscience and Remote Sensing}, vol.~49, no.~10, pp.
-  3677--3687, Oct. 2011.
-
-\bibitem{holmengraa}
-{Article of Holmengraa on dykkepedia.com}.
-
-\end{thebibliography}
-
-\fi
-
+% 
+% 
 \input{bio/jo_inge_buskenes}

### 2016-02-17 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2016:02:17	13:35:53	(on.17.feb.)

@@
    \ifFlatArchive\else                          % 'pdflatex'
-      \graphicspath{./gfx/}
    \fi
@@
    \ifFlatArchive\else
-      \graphicspath{./gfx/}
    \fi
@@
 
+\ifPeerReview
+\graphicspath{./submission/submit1}
+\fi
+
 % \newcommand\multimedia[1]{\textbf{{\color{red}[#1]}}}
@@
 LCA can be viewed as either a simplification of the minimum variance distortionless response (MVDR) beamformer, or as an adaptive extension to the delay and sum (DAS) beamformer.
-While both LCA and MVDR attempt to minimize the power of noise and interference in the image, MVDR seeks the optimal solution by estimating and inverting the array's spatial covariance matrix, while LCA selects the best performer from a predefined set of sensor windows.
+While both LCA and MVDR attempt to minimize the power of noise and interference in the image by adapting the array's weights, MVDR seeks an optimal solution by computing the spatial statistics of the wavefield, while LCA simply selects the best window out of a predefined set.
 
@@
 This was motivated by the desire to use wide windows with maximum sidelobe suppression near the receiver, and narrow windows at greater depths for maximum sensitivity and depth penetration.
-While achieving good results, it was not made clear how how many windows LCA needs, or how they ideally should be steered.
+While achieving good results, it was not made clear how many windows LCA needs, or how they ideally should be steered.
 Also, it is not apparent how LCA performs in sonar systems where the transmit beam is unfocused.
@@
 \ifPeerReview%
-\includegraphics[width=.5\linewidth]{gfx/calc_kaiser_3dB.eps}%
+\includegraphics[width=.5\linewidth]{gfx/buske1_online.eps}%
 \else%
-\includegraphics[width=\linewidth]{gfx/calc_kaiser_3dB.eps}%
+\includegraphics[width=\linewidth]{gfx/buske1_online.eps}%
 \fi%
@@
 \ifPeerReview%
-\includegraphics[width=.5\linewidth]{gfx/cross_tagged.png}%
+\includegraphics[width=.5\linewidth]{gfx/buske2_online.png}%
 \else%
-\includegraphics[width=\linewidth]{gfx/cross_tagged.png}%
+\includegraphics[width=\linewidth]{gfx/buske2_online.png}%
 \fi%
@@
 \begin{figure*}[t]\centering%
-\includegraphics[width=\linewidth]{gfx/mvdr_selected_windows_holmengraa.pdf}%
+\includegraphics[width=\linewidth]{gfx/buske3_online.pdf}%
 \caption{\emph{Determining LCA window type}: MVDR image with typical frequency responses for windows used in various pixel regions.\newline
@@
 \begin{figure*}[t]%
-\includegraphics[width=\textwidth]{gfx/oversampling_mosaic_bounds.pdf}%
+\includegraphics[width=\textwidth]{gfx/buske4_online.pdf}%
 \caption{\emph{Determining Kaiser parameter boundaries}: LCA images of the resolution test cross and anchor created with a large and equal size window database, but where the upper bounds for $\beta$ and $\phi$ is varied.
@@
 Note how LCA prefers a a narrow response ($\beta=0$) on the anchor, and on other strong sources with little lateral interference.
-On the sources where lateral interference is present it chooses the widest response with the best sidelobe suppression.]{\includegraphics[width=\textwidth]{gfx/oversampling_mosaic_bounds_lca_windows_beta.pdf}%%
+On the sources where lateral interference is present it chooses the widest response with the best sidelobe suppression.]{\includegraphics[width=\textwidth]{gfx/buske5a_online.pdf}%%
 \label{oversampling_mosaic_bounds_beta}}
@@
 On strong sources LCA selects windows that are steered away from the source.
-This is what improves the FWHM measurement in \Fig{oversampling_mosaic_bounds}.]{\includegraphics[width=\textwidth]{gfx/oversampling_mosaic_bounds_lca_windows_phi.pdf}%
+This is what improves the FWHM measurement in \Fig{oversampling_mosaic_bounds}.]{\includegraphics[width=\textwidth]{gfx/buske5b_online.pdf}%
 \label{oversampling_mosaic_bounds_phi}}
@@
 \begin{figure*}[t]%\centering%
-\includegraphics[width=\textwidth]{gfx/oversampling_mosaic.pdf}%
+\includegraphics[width=\textwidth]{gfx/buske6_online.pdf}%
 \caption{\emph{LCA images using different window database sizes}.
@@
 Note how the FWHM of LCA sits in between that of DAS and MVDR.
-Also note how LCA is insensitive to the window type; its performance is near identical whether it uses Kasier or trigonometric windows.]{\includegraphics[width=\textwidth]{gfx/beamformer_comparison_cross.pdf}%
+Also note how LCA is insensitive to the window type; its performance is near identical whether it uses Kasier or trigonometric windows.]{\includegraphics[width=\textwidth]{gfx/buske7a_online.pdf}%
 \label{beamformer_comparison_cross}}
@@
 Compared to DAS the adaptive beamformers produce deeper shadows, sharper edges and a higher detail level.
-The adaptive methods produce nearly identical images.]{\includegraphics[width=\textwidth]{gfx/beamformer_comparison_holmengraa.pdf}%%
+The adaptive methods produce nearly identical images.]{\includegraphics[width=\textwidth]{gfx/buske7b_online.pdf}%%
 \label{beamformer_comparison_holmengraa}}
@@
 Hence, when it is steered the white noise gain must increase and the the signal-to-noise ratio decrease, as shown in \Fig{windows_steering}.
-To limit this we place an upper bound to the steering, chosen such that wide windows is allowed to be steered more than narrow ones.
+To limit this we place an upper bound to the steering, chosen such that wide windows are allowed to be steered more than narrow ones.
 We call this upper bound for steering $\phi_{\text{3dB}}(\beta)$, and define it as the steering angle needed for the steered response to have a \minus{}3\,dB crossing that is exactly half that of the unsteered window.

### 2016-02-17 LCAforActiveSonarImaging/submission/submit1/answer/answer.tex
la2pdf commit - 2016:02:17	13:35:53	(on.17.feb.)

@@
 \lfoot{}\cfoot{\thepage}\rfoot{}
+
+\usepackage{url}
 
@@
 
+\newpage
 \section{Answer to reviewer 1}
@@
 \q{The paper is well written and addresses an actual problem. The solution however, is barely innovative. Still the results are interesting enough for publication. Some minors details.} \\\\
-\ans{Barely innvative? Can we argue better for why it is?}
+\ans{We appreciate the reviewer's }
 
@@
 \qi{The authors show no evidence that bats use adaptive processing, remove this example.}
-\ai{}
+\ai{The reviewer makes a good point, this may not have been obvious. We have provided a reference for the adaptive processing used by bats, but would prefer to keep the example if possible.}
 %
 \qi{It is not clear whether LCA is really an adaptive method that adapts weights on the fly or a method that a posteriori decides which window shows best results.}
-\ai{}
+\ai{We appreciate having this pointed out.
+We are using a rather loose definition of adaptivity that includes any kind of dynamic and data dependent behaviour, hence feel that both beamformers fall in that category.
+However, as the reviewer pointed out, the method .
+}
 %
 \qi{The use of non-SI units like cm is discouraged.}
-\ai{}
+\ai{We agree that non-SI units should be avoided, however had the belief that cm was a valid unit? In the IEEE Styles Manual this unit is used several times as an example: 
+\url{http://www.ieee.org/publications\_standards/publications/journmag/online\_style\_manual-10292015.pdf}} %(on page 24 paragraph ``The En, Em, or Two-Em Dash'', and several times on page 25 paragraph ``Math'' bullet point 16.
+Should we be adhering to a different style guide?}
 %
@@
 
+\newpage
 \section{Answer to reviewer 3}

### 2016-02-17 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2016:02:17	16:44:20	(on.17.feb.)

@@
 LCA can be viewed as either a simplification of the minimum variance distortionless response (MVDR) beamformer, or as an adaptive extension to the delay and sum (DAS) beamformer.
-While both LCA and MVDR attempt to minimize the power of noise and interference in the image by adapting the array's weights, MVDR seeks an optimal solution by computing the spatial statistics of the wavefield, while LCA simply selects the best window out of a predefined set.
-
-We show that a robust MVDR implementation typically creates windows spanning between the rectangular and Hamming.
+While both LCA and MVDR attempt to minimize the power of noise and interference in the image, MVDR achieves this by computing optimal array weights from the spatial statistics of the wavefield, while LCA selects the best performing weights out of a predefined set.
+
+To build confidence in the LCA method we show that a robust MVDR implementation typically creates weights spanning between the rectangular and Hamming window function.
 We let LCA select from a set of Kaiser windows with responses in this span, and add some steered variations of each.
@@
 
-\IEEEPARstart{T}{he} best adaptive sonars are found in nature.
-Bats, for instance, use a high frequency sonar to detect, identify and track prey with amazing precision in caves amongst a myriad of other bats.
-Active human made sonars have little to no adaptability on transmit, but several methods have been investigated to adaptively post process the received data.
+% The promise is already demonstrated in nature.
+
+The sonar systems of the future
+
+Future sonar systems
+
+Great promise in adaptivity
+
+The greatest untapped
+
+The sonars of the future might benefit greatly from adaptivity.
+
+Future sonars will likely be
+
+\IEEEPARstart{A}{daptive} sonars holds great promise.
+This promise is allured to in nature by for instance bats, which use a high frequency sonar to detect, identify and track prey with amazing precision in caves amongst a myriad of other bats.
+No man made sonar can currenly match this feat, because it is hard to determine what the correct action is for an arbitrary situation, and too slow to process all the information required to typically slow to process all the information needed to act correctly the situation in time, and non trivial to determine exactly how it should adapt.
+In current active sonars there is little to no adaptability on transmit, several methods have been investigated to adaptively post process the received data.
 Perhaps the best studied method for reconstructing a sonar image is the minimum variance distortion-less response (MVDR) beamformer~\cite{Capon1969}.

### 2016-02-17 LCAforActiveSonarImaging/submission/submit1/answer/answer.tex
la2pdf commit - 2016:02:17	16:44:20	(on.17.feb.)

@@
 \q{The paper is well written and addresses an actual problem. The solution however, is barely innovative. Still the results are interesting enough for publication. Some minors details.} \\\\
-\ans{We appreciate the reviewer's }
+\ans{We appreciate the reviewer's view and would like to thank him for this time and patience reviewing this article.}
 
@@
 \qi{The use of non-SI units like cm is discouraged.}
-\ai{We agree that non-SI units should be avoided, however had the belief that cm was a valid unit? In the IEEE Styles Manual this unit is used several times as an example: 
-\url{http://www.ieee.org/publications\_standards/publications/journmag/online\_style\_manual-10292015.pdf}} %(on page 24 paragraph ``The En, Em, or Two-Em Dash'', and several times on page 25 paragraph ``Math'' bullet point 16.
-Should we be adhering to a different style guide?}
+\ai{We agree that non-SI units should be avoided, however had the belief that cm being a prefixed SI unit was valid? In the IEEE Styles Manual this unit is used several times as an example: 
+\url{http://www.ieee.org/publications_standards/publications/journmag/online_style_manual-10292015.pdf} (on page 24 paragraph ``The En, Em, or Two-Em Dash'', and several times on page 25 paragraph ``Math'' bullet point 16.
+Will leave this as is for now pending an answer for this.}
 %
@@
 
-\ans{Thank you for the thorough review and very constructive remarks. We believe in this philosophy and will make changes accordingly.}
+\ans{We thank the reviewer for the thorough review and very constructive remarks. We believe in this philosophy and will make changes accordingly.}
 

### 2016-02-22 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2016:02:22	14:54:15	(ma.22.feb.)

@@
 % The promise is already demonstrated in nature.
-
-The sonar systems of the future
-
-Future sonar systems
-
-Great promise in adaptivity
-
-The greatest untapped
-
-The sonars of the future might benefit greatly from adaptivity.
-
-Future sonars will likely be
-
-\IEEEPARstart{A}{daptive} sonars holds great promise.
-This promise is allured to in nature by for instance bats, which use a high frequency sonar to detect, identify and track prey with amazing precision in caves amongst a myriad of other bats.
-No man made sonar can currenly match this feat, because it is hard to determine what the correct action is for an arbitrary situation, and too slow to process all the information required to typically slow to process all the information needed to act correctly the situation in time, and non trivial to determine exactly how it should adapt.
-In current active sonars there is little to no adaptability on transmit, several methods have been investigated to adaptively post process the received data.
+% 
+% The sonar systems of the future 
+% 
+% Future sonar systems 
+% 
+% Great promise in adaptivity
+% 
+% The greatest untapped 
+% 
+% The sonars of the future might benefit greatly from adaptivity.
+% 
+% 
+% Future sonars will likely be
+
+% To make a sonar truly perform better, we need to make it smarter. 
+% 
+% \IEEEPARstart{F}{uture} sonar systems will merit a much more sophisticated level of adaptivity. 
+% 
+% \IEEEPARstart{T}{he} next generation sonar systems will be adaptive.
+% 
+% Making sonars adaptive is the way to take them into the future. 
+% 
+% The virtue of future sonar systems will be sophisticated adaptivity. 
+% 
+% Intelligent
+% 
+% Reactive
+% 
+% Observe, process, adapt, act. 
+% 
+% Adaptivity - sense, process, respond
+% 
+% Faster, smarter
+% 
+% The best sonars in nature are adaptive. 
+% Nature allures us to be this bold because
+
+\IEEEPARstart{T}{he} best sonars in existence is perhaps found in nature.
+Bats, for instance, use a high frequency sonar to detect, identify and track prey with amazing precision in caves amongst a myriad of other bats.
+No man made sonar can currenly match this feat, much due to the challenge of processing and adapting to a complex environment fast enough.
+While human made sonars have little to no adaptivity on transmit, several methods have been investigated to adaptively post process the received data.
 Perhaps the best studied method for reconstructing a sonar image is the minimum variance distortion-less response (MVDR) beamformer~\cite{Capon1969}.
@@
 This is because MVDR relies on estimating and inverting a covariance matrix.
-In sonar systems with less than 32 channels the complexity can be reduced and the implementation accelerated significantly using graphics computing units (GPUs)~\cite{Buskenes2014}.
-For larger systems the inversion step dominates the computation.
+The estimation step dominates the computation in sonar systems with less than 32 channels, but here computational complexity can be reduced and the implementation accelerated significantly using graphics computing units (GPUs)~\cite{Buskenes2014}.
+For larger systems the inversion step dominates.
 This can be dealt with by approximating the full covariance matrix with a small one using space reduction techniques such as beamspace processing~\cite{Asen2013}, or the closely related principal component analysis (PCA) method~\cite{Kim2014}.
@@
 In total he made LCA choose from 12 unique windows, 6 of which were micro-steered variants of a fairly wide Kaiser window, and with the rectangular window included.
+Micro-steering is means to phase-shift the array's elements to finetune the azimuth angle the array is steered towards.
 This was motivated by the desire to use wide windows with maximum sidelobe suppression near the receiver, and narrow windows at greater depths for maximum sensitivity and depth penetration.

### 2016-02-22 LCAforActiveSonarImaging/submission/submit1/answer/answer.tex
la2pdf commit - 2016:02:22	14:54:15	(ma.22.feb.)

@@
 \qi{It is not clear whether LCA is really an adaptive method that adapts weights on the fly or a method that a posteriori decides which window shows best results.}
-\ai{We appreciate having this pointed out.
-We are using a rather loose definition of adaptivity that includes any kind of dynamic and data dependent behaviour, hence feel that both beamformers fall in that category.
-However, as the reviewer pointed out, the method .
-}
+\ai{We appreciate the remark.
+We are using a system level definition of adaptivity that includes any kind of dynamic and data dependent behaviour, hence feel that both beamformers fall in that category.
+Both methods process the pixels in a data dependent manner, however, perhaps it should be made even clearer that the LCA solution space is a sampled version of that of MVDR.
+We will attempt to express this better.}
 %
@@
 \ai{We agree that non-SI units should be avoided, however had the belief that cm being a prefixed SI unit was valid? In the IEEE Styles Manual this unit is used several times as an example: 
-\url{http://www.ieee.org/publications_standards/publications/journmag/online_style_manual-10292015.pdf} (on page 24 paragraph ``The En, Em, or Two-Em Dash'', and several times on page 25 paragraph ``Math'' bullet point 16.
+\url{http://www.ieee.org/publications_standards/publications/journmag/online_style_manual-10292015.pdf} (e.g.
+on page 24 paragraph ``The En, Em, or Two-Em Dash'', and several times on page 25 paragraph ``Math'' bullet point 16).
 Will leave this as is for now pending an answer for this.}
@@
 \qi{"the array's spatial covariance matrix" might be a bit shocking for the non-expert Reader --> The formulation in the Introduction might fit better "the incluence of noise ..."}
-\ai{Thank you, this is a good point. We have reformulated the sentence to speak of computing the spatial statistics instead.}
+\ai{This is a good point. We have reformulated the sentence to speak of computing the spatial statistics instead.}
 %
@@
 \qi{"6 of which were micro-steered" --> a short explanation or reference might help to understand what is meant by "micro-steered"}
-\ai{}
+\ai{Nice catch. A short sentence elaborating on this expression added.}
 %

### 2016-02-22 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2016:02:22	15:58:01	(ma.22.feb.)

@@
 We extend Synnev\aa{}g's work by steering all our windows and not just some, and demonstrate how LCA's performance is affected by the type and number of windows included.
-Our results are obtained using experimental data from the 32 element Kongsberg Maritime HISAS 1030 synthetic aperture sonar (SAS), which operates at 100\,kHz with an element beamwidth of 23$^\circ$.
+Our results are obtained using experimental data from the 32 element Kongsberg Maritime HISAS 1030 synthetic aperture sonar (SAS), which operates at 100\,kHz with an -3\,dB element opening angle of 23$^\circ$.
 
@@
 To test the performance of the MVDR and LCA beamformers, we have processed data acquired by the Kongsberg Maritime HUGIN AUV carrying their 32 element HISAS1030 sonar~\cite{Hansen2011}.
-It is a high resolution synthetic aperture sonar with 1.2\;m array length, 100\;kHz operating frequency, 30\;kHz bandwidth and 23$^\circ$ receiver element beamwidth.
+It is a high resolution synthetic aperture sonar with 1.2\;m array length, 100\;kHz operating frequency, 30\;kHz bandwidth and 23$^\circ$ receiver element -3\,dB opening angle.
 

### 2016-02-22 LCAforActiveSonarImaging/submission/submit1/answer/answer.tex
la2pdf commit - 2016:02:22	15:58:01	(ma.22.feb.)

@@
 We are using a system level definition of adaptivity that includes any kind of dynamic and data dependent behaviour, hence feel that both beamformers fall in that category.
-Both methods process the pixels in a data dependent manner, however, perhaps it should be made even clearer that the LCA solution space is a sampled version of that of MVDR.
+Both methods process the pixels in a data dependent manner, and result in non linear image.
+However, perhaps it should be made even clearer that the LCA solution space is a sampled version of that of MVDR.
 We will attempt to express this better.}
@@
 \qi{"6 of which were micro-steered" --> a short explanation or reference might help to understand what is meant by "micro-steered"}
-\ai{Nice catch. A short sentence elaborating on this expression added.}
+\ai{Nice catch. A short sentence elaborating on this expression is added.}
 %
@@
 \qi{"with an element beamwidth" --> what is meant by element beamwidth? Maybe an additional figure could be used to explain this.}
-\ai{}
+\ai{Good remark. We meant the -3dB opening angle of the elements on transmit. This has now been reworded.}
 %
@@
 \qi{The lateral correlation and the reason why equation (5) is a good choice should be better explained.}
-\ai{}
+\ai{Indeed it should, this paragraph is now rewritten.}
 %

### 2016-03-01 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2016:03:01	13:37:50	(ti. 1.mars)

@@
 This is because MVDR relies on estimating and inverting a covariance matrix.
-The estimation step dominates the computation in sonar systems with less than 32 channels, but here computational complexity can be reduced and the implementation accelerated significantly using graphics computing units (GPUs)~\cite{Buskenes2014}.
+The estimation step dominates the computation in sonar systems with less than 32 channels, but here the computational complexity can be reduced and the implementation accelerated significantly using graphics computing units (GPUs)~\cite{Buskenes2014}.
 For larger systems the inversion step dominates.
@@
 suggested using a window set comprised of rectangular, Kaiser (or Kaiser-Harris) and inverted Kaiser functions.
-In total he made LCA choose from 12 unique windows, 6 of which were micro-steered variants of a fairly wide Kaiser window, and with the rectangular window included.
-Micro-steering is means to phase-shift the array's elements to finetune the azimuth angle the array is steered towards.
-This was motivated by the desire to use wide windows with maximum sidelobe suppression near the receiver, and narrow windows at greater depths for maximum sensitivity and depth penetration.
+In total he made LCA choose from 12 unique windows, 6 of which were phase shifted versions of a fairly wide Kaiser window, and one rectangular.
+Phase shifting the windows has the effect of slighly adjusting the angle the array is steered towards.
+His composition of windows was motivated by the desire to use wide windows with maximum sidelobe suppression near the receiver, and narrow windows at greater depths for maximum sensitivity and depth penetration.
 While achieving good results, it was not made clear how many windows LCA needs, or how they ideally should be steered.

### 2016-03-01 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2016:03:01	20:09:13	(ti. 1.mars)

@@
 % \newcommand\mediaPath{gfx/media}
-\newcommand\mediaPath{http://folk.uio.no/joibu/media/2015_JOE_LCA}
-
-\newcommand\mediaI{\multimedia{\mediaPath/media1.mp4}{Media Movie~1}}
-\newcommand\mediaII{\multimedia{\mediaPath/media2.mp4}{Media Movie~2}}
+\newcommand\articlePath{http://folk.uio.no/joibu/articles/2015_JOE_LCA}
+
+\newcommand\mediaI{\multimedia{\articlePath/media1.mp4}{Media Movie~1}}
+\newcommand\mediaII{\multimedia{\articlePath/media2.mp4}{Media Movie~2}}
 
@@
 We consider an absolute minimum to be 2, due to the non-linear nature of delaying pixels that isn't strictly in the far field, and of displaying the absolute value of the pixels on a decibel scale.
-In \mediaI{} we visualize the effect of changing the lateral oversampling factor on a set of LCA images.
+In \mediaI{}\footnote{For media location refer to Appendix \ref{source_code}.} we visualize the effect of changing the lateral oversampling factor on a set of LCA images.
 Upon display the images are all bilinearly upinterpolated to the same size.
@@
 
-\mediaII{} animates how \Fig{mvdr_selected_windows} changes as a function of the MVDR subarray length $L$ and temporal averaging $K$.
+\mediaII{}\footnote{For media location refer to Appendix \ref{source_code}.} animates how \Fig{mvdr_selected_windows} changes as a function of the MVDR subarray length $L$ and temporal averaging $K$.
 At $L=1$ the window responses are rectangular in all areas in the image.
@@
 
+% \section{Steering angle}
+
+\begin{figure}[t]%\centering%
+\includegraphics[width=\linewidth]{gfx/buske8_online.eps}%
+\caption{\emph{Steering angle needed to cut the -3\,dB distance of the Kaiser window's spatial response in half}.
+The needed amount of steering depends on the number of channels $M$, the element spacing relative to the wavelength $\frac{d}{\lambda}$, and the Kaiser parameter $\beta$.
+}\label{steering_angle_3dB}
+\end{figure}
+%
+One of the goals of this article was to characterize the windows used by the MVDR method, and then identify a subset of these that were suitable for use with the LCA method.
+We determined that the Kaiser function could fit the role of producing the relevant windows.
+The spatial response of a steered Kaiser window depends on the number of channels $M$, the element spacing relative to the wavelength $\frac{d}{\lambda}$, the Kaiser parameter $\beta$, and the amount of steering $\phi$.
+
+To control the resolution gain of LCA, we wanted to determine the extent that the windows need to be steered to cut the angle of the -3\,dB point of the window's spatial amplitude response by a fixed amount, say, by a factor 2 as reference.
+We computed this steering angle for some common configurations of the mentioned parameters.
+Refer to \Fig{steering_angle_3dB} for the result.
+The boundary values of the Kaiser window is shown, i.e.
+the rectangular window at $\beta=0$ and the near gaussian window at $\beta=5$, as well as one in between at $\beta=2.5$.
+Observe that for a given $M$ and $\beta$, the angle $\phi_\mathrm{3dB}$ scales near proportionally with $\frac{d}{\lambda}$.
+We found this to be true also for $\frac{d}{\lambda}\in\{0.5,4\}$.
+Hence, an approximate figure for a rather wide range of system parameters can be derived off this figure.
+
+For the sake of completeness we also supply the source code, see Appendix \ref{source_code}.
+
 \section{Reducing LCA complexity}\label{sec:reducing_lca_complexity}
@@
 The solution for $\alpha$ would also yield the optimal beamformer output in the minimum variance sense.
+
+% \section{Multimedia Files and Source Code}\label{source_code}
+
+This article is accompanied by some multimedia files and source code for determining the steering angles for LCA's windows. It is located at:
+
+\url{\articlePath}
 

### 2016-03-01 LCAforActiveSonarImaging/submission/submit1/answer/answer.tex
la2pdf commit - 2016:03:01	13:37:50	(ti. 1.mars)
la2pdf commit - 2016:03:01	20:09:13	(ti. 1.mars)
la2pdf commit - 2016:03:01	20:27:55	(ti. 1.mars)

@@
 on page 24 paragraph ``The En, Em, or Two-Em Dash'', and several times on page 25 paragraph ``Math'' bullet point 16).
-Will leave this as is for now pending an answer for this.}
+Will leave this as is for now pending any objections?}
 %
@@
 \qi{"6 of which were micro-steered" --> a short explanation or reference might help to understand what is meant by "micro-steered"}
-\ai{Nice catch. A short sentence elaborating on this expression is added.}
+\ai{Another good point. This wording is now changed and a sentence elaborating on its meaning is added.}
 %
@@
 \qi{"with an element beamwidth" --> what is meant by element beamwidth? Maybe an additional figure could be used to explain this.}
-\ai{Good remark. We meant the -3dB opening angle of the elements on transmit. This has now been reworded.}
+\ai{Good remark. We meant the -3dB opening angle of the elements on transmit. Changes made to reflect this.}
 %
@@
 \newpage
-\section{Changes to figures}
+
+\section{Figure changes}
 
 \begin{enumerate}
-\item Fig. X. Changed bla bla bla
+\item Fig 8 is new.
+It was added to provide details on how the steering bound $\phi_\mathrm{3dB}$ is affected by the number of elements $M$ and the element spacing relative to the wavelength $\frac{d}{\lambda}$.
+\end{enumerate}
+
+\section{Other changes}
+
+The steering bound $\phi_\mathrm{3dB}$ depends on not just $\beta$, but also on the number of elements $M$ and the element spacing relative to the wavelength $\frac{d}{\lambda}$.
+\begin{enumerate}
+\item We have supplied additional info on how the steering is affected by these parameters in appendix C.
+\item Equations (12) and (14) only gave the correct results for our system configuration, and were not very intuitive.
+Appendix C now replace and extends these equations, and we removed them in hope this will make the article easier to read.
 \end{enumerate}

### 2016-03-02 LCAforActiveSonarImaging/submission/submit1/answer/answer.tex
la2pdf commit - 2016:03:02	08:53:57	(on. 2.mars)

@@
 We are using a system level definition of adaptivity that includes any kind of dynamic and data dependent behaviour, hence feel that both beamformers fall in that category.
-Both methods process the pixels in a data dependent manner, and result in non linear image.
-However, perhaps it should be made even clearer that the LCA solution space is a sampled version of that of MVDR.
-We will attempt to express this better.}
+Both methods process the pixels in a data dependent manner, resulting in non linear image.
+We have tried to express it clearer that the LCA solution space is merely a sampled version of that of MVDR, hence can not be considered fully adaptive such as MVDR (although this, too, is heavily constrained).}
 %
@@
 
-\ans{We thank the reviewer for the thorough review and very constructive remarks. We believe in this philosophy and will make changes accordingly.}
+\ans{We thank the reviewer for the thorough review and very constructive remarks. We believe in this philosophy and attempt to make changes accordingly.}
 

### 2016-03-02 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2016:03:01	20:27:55	(ti. 1.mars)
la2pdf commit - 2016:03:02	09:22:10	(on. 2.mars)

@@
 Bats, for instance, use a high frequency sonar to detect, identify and track prey with amazing precision in caves amongst a myriad of other bats.
-No man made sonar can currenly match this feat, much due to the challenge of processing and adapting to a complex environment fast enough.
-While human made sonars have little to no adaptivity on transmit, several methods have been investigated to adaptively post process the received data.
-Perhaps the best studied method for reconstructing a sonar image is the minimum variance distortion-less response (MVDR) beamformer~\cite{Capon1969}.
+No man made sonar can currenly match this feat, much due to the challenge of processing and adapting to such a complex environment fast enough.
+This is why human made sonars have little to no adaptivity on transmit.
+On the receiving side, however, several methods have been investigated to adaptively post process the received data.
+Perhaps the most well known method for reconstructing a sonar image out of the data is the minimum variance distortion-less response (MVDR) beamformer~\cite{Capon1969}.
 It adapts the sonar array's spatial response to minimize the influence of noise and interference in the final image.
@@
 
-% \section{Steering angle}
+\section{Steering angle}
 
@@
 
-% \section{Multimedia Files and Source Code}\label{source_code}
+\section{Multimedia Files and Source Code}\label{source_code}
 

### 2016-03-02 LCAforActiveSonarImaging/submission/submit1/answer/answer.tex
la2pdf commit - 2016:03:02	09:22:10	(on. 2.mars)

@@
 \qi{The authors show no evidence that bats use adaptive processing, remove this example.}
-\ai{The reviewer makes a good point, this may not have been obvious. We have provided a reference for the adaptive processing used by bats, but would prefer to keep the example if possible.}
+\ai{Very true.
+We have reworded the example to say that the bats have a sophisticated sonar instead.
+The term adaptive sonar is now only used on human made systems.
+We hope this change is satisfactory.
+}
 %
@@
 \ai{We appreciate the remark.
-We are using a system level definition of adaptivity that includes any kind of dynamic and data dependent behaviour, hence feel that both beamformers fall in that category.
+We are using a system level definition of adaptivity that includes any kind of dynamic and data dependent behavior, hence feel that both beamformers fall in that category.
 Both methods process the pixels in a data dependent manner, resulting in non linear image.
@@
 \hi{Abstract}
-\qi{"the array's spatial covariance matrix" might be a bit shocking for the non-expert Reader --> The formulation in the Introduction might fit better "the incluence of noise ..."}
+\qi{"the array's spatial covariance matrix" might be a bit shocking for the non-expert Reader --> The formulation in the Introduction might fit better "the influence of noise ..."}
 \ai{This is a good point. We have reformulated the sentence to speak of computing the spatial statistics instead.}

### 2016-03-02 LCAforActiveSonarImaging/submission/submit1/answer/answer.tex
la2pdf commit - 2016:03:02	11:26:25	(on. 2.mars)

@@
 Both methods process the pixels in a data dependent manner, resulting in non linear image.
-We have tried to express it clearer that the LCA solution space is merely a sampled version of that of MVDR, hence can not be considered fully adaptive such as MVDR (although this, too, is heavily constrained).}
+We have tried to emphasize with improved clarity that the LCA solution space is merely a sampled version of that of MVDR, hence can not be considered fully adaptive such as MVDR (although this, too, is heavily constrained).}
 %
@@
 on page 24 paragraph ``The En, Em, or Two-Em Dash'', and several times on page 25 paragraph ``Math'' bullet point 16).
-Will leave this as is for now pending any objections?}
+Will leave it as is for now pending clarification.}
 %

### 2016-03-02 LCAforActiveSonarImaging/submission/submit1/answer/answer.tex
la2pdf commit - 2016:03:02	11:31:57	(on. 2.mars)

@@
 Both methods process the pixels in a data dependent manner, resulting in non linear image.
-We have tried to emphasize with improved clarity that the LCA solution space is merely a sampled version of that of MVDR, hence can not be considered fully adaptive such as MVDR (although this, too, is heavily constrained).}
+We have tried to emphasize with improved clarity that the LCA solution space is merely a sampled version of that of MVDR, hence can not be considered a fully adaptive (yet heavily constrained) method such as MVDR.}
 %

### 2016-03-02 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2016:03:02	11:31:57	(on. 2.mars)
la2pdf commit - 2016:03:02	12:45:10	(on. 2.mars)

@@
 This is why human made sonars have little to no adaptivity on transmit.
-On the receiving side, however, several methods have been investigated to adaptively post process the received data.
-Perhaps the most well known method for reconstructing a sonar image out of the data is the minimum variance distortion-less response (MVDR) beamformer~\cite{Capon1969}.
+However, several methods have been investigated to adaptively post process the received data.
+Perhaps the best studied method for reconstructing a sonar image out of the data is the minimum variance distortion-less response (MVDR) beamformer~\cite{Capon1969}.
 It adapts the sonar array's spatial response to minimize the influence of noise and interference in the final image.

### 2016-03-02 LCAforActiveSonarImaging/submission/submit1/answer/answer.tex
la2pdf commit - 2016:03:02	12:45:10	(on. 2.mars)

@@
 \hi{Abstract}
-\qi{"the array's spatial covariance matrix" might be a bit shocking for the non-expert Reader --> The formulation in the Introduction might fit better "the influence of noise ..."}
+\qi{"the array's spatial covariance matrix" might be a bit shocking for the non-expert Reader $\rightarrow$ The formulation in the Introduction might fit better "the influence of noise ..."}
 \ai{This is a good point. We have reformulated the sentence to speak of computing the spatial statistics instead.}
@@
 \qi{"with an element beamwidth" --> what is meant by element beamwidth? Maybe an additional figure could be used to explain this.}
-\ai{Good remark. We meant the -3dB opening angle of the elements on transmit. Changes made to reflect this.}
+\ai{Good remark. We meant the -3dB opening angle of the elements. Changes are now made to reflect this.}
 %

### 2016-03-02 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2016:03:02	12:55:42	(on. 2.mars)

@@
 LCA applies a set of predefined windows, or apodization weights, and selects the one offering the best sidelobe suppression.
-It relies on the same optimization criterion as MVDR, thus may be considered a version of MVDR with a reduced window solution space.
+It relies on the same optimization criterion as MVDR, thus may be considered a version of MVDR with a discrete and static window solution space.
 Alternatively, it may be viewed as a multi-apodization technique.
@@
 This gives LCA an adaptive freedom that is somewhere in between that of traditional multi-apodization techniques and MVDR.
+The bigger the space the weights span, the greater the adaptive freedom, but only the solutions that are valid in the constrained solution space can make a difference.
 
@@
 To test the performance of the MVDR and LCA beamformers, we have processed data acquired by the Kongsberg Maritime HUGIN AUV carrying their 32 element HISAS1030 sonar~\cite{Hansen2011}.
-It is a high resolution synthetic aperture sonar with 1.2\;m array length, 100\;kHz operating frequency, 30\;kHz bandwidth and 23$^\circ$ receiver element -3\,dB opening angle.
+It is a high resolution synthetic aperture sonar with 1.2\;m array length, 100\;kHz operating frequency, 30\;kHz bandwidth and 23$^\circ$ element -3\,dB opening angle.
 

### 2016-03-02 LCAforActiveSonarImaging/submission/submit1/answer/answer.tex
la2pdf commit - 2016:03:02	12:55:42	(on. 2.mars)

@@
 \hi{Page 2}
-\qi{"6 of which were micro-steered" --> a short explanation or reference might help to understand what is meant by "micro-steered"}
+\qi{"6 of which were micro-steered" $\rightarrow$ a short explanation or reference might help to understand what is meant by "micro-steered"}
 \ai{Another good point. This wording is now changed and a sentence elaborating on its meaning is added.}
 %
-\qi{"it was not clear how how many" --> delete 1 x "how"}
+\qi{"it was not clear how how many" $\rightarrow$ delete 1 x "how"}
 \ai{Thank you. Corrected.}
 %
-\qi{"with an element beamwidth" --> what is meant by element beamwidth? Maybe an additional figure could be used to explain this.}
-\ai{Good remark. We meant the -3dB opening angle of the elements. Changes are now made to reflect this.}
+\qi{"with an element beamwidth" $\rightarrow$ what is meant by element beamwidth? Maybe an additional figure could be used to explain this.}
+\ai{Good remark. We meant the -3dB opening angle of the elements. We have reworded this the two places it is mentioned.}
 %
@@
 \qi{"Applying a window that trails off ....
-Minimum variance distortion-less response (MVDR) beamformer."--> real and complex Windows?
+Minimum variance distortion-less response (MVDR) beamformer." $\rightarrow$ real and complex Windows?
 The introduction should be more formal: x ist introduced as delayed data, weights or delays to change, or both is explained later.
---> The difference between noise and sidelobes should be pointed out.}
+$\rightarrow$ The difference between noise and sidelobes should be pointed out.}
 \ai{}
@@
 \hi{Page 6, last row}
-\qi{"they are optimal for any value" --> optimal in which sense?}
+\qi{"they are optimal for any value" $\rightarrow$ optimal in which sense?}
 \ai{Thank you. This was mentioned earlier in the same paragraph, just after (8).}
@@
 \hi{Page 8}
-\qi{reference to Fig.1 --> could the sequence of figures be changes to have Fig. 1 closer to this paragraph}
+\qi{reference to Fig.1 $\rightarrow$ could the sequence of figures be changes to have Fig. 1 closer to this paragraph}
 \ai{Valid point. In our two column print the figure is on the page before its first reference. }

### 2016-03-02 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2016:03:02	13:35:59	(on. 2.mars)

@@
 We limit the steering to roughly half the -3\,dB width of the window's amplitude response.
-Using experimental data from the Kongsberg Maritime HISAS 1030 sonar we find that LCA and MVDR produce nearly identical images of large scenes, both being superior to DAS.
+Using experimental data from the Kongsberg Maritime HISAS1030 sonar we find that LCA and MVDR produce nearly identical images of large scenes, both being superior to DAS.
 On point targets LCA is able to double the resolution compared to DAS, or provide half that of MVDR.
@@
 Based on an idea by Vignon~\cite{Vignon2008}, it was first introduced by Synnev\aa{}g \emph{et al}.
-in clinical medical imaging~\cite{Synnevag2008}, who demonstrated its ability to obtain very similar image quality to MVDR in systems with focused transmit beams.
-LCA applies a set of predefined windows, or apodization weights, and selects the one offering the best sidelobe suppression.
+in clinical medical ultrasound imaging~\cite{Synnevag2008}, who demonstrated its ability to obtain very similar image quality to MVDR in systems with focused transmit beams.
+LCA applies a set of predefined and static windows, or apodization weights, and selects the one offering the best noise suppression.
 It relies on the same optimization criterion as MVDR, thus may be considered a version of MVDR with a discrete and static window solution space.
@@
 in radar imaging~\cite{Stankwitz1995}, where the best out of 2 or 3 windows are selected.
-However, LCA is different in that it typically selects from a larger pool of windows and allows micro-steered variations of them.
-This gives LCA an adaptive freedom that is somewhere in between that of traditional multi-apodization techniques and MVDR.
-The bigger the space the weights span, the greater the adaptive freedom, but only the solutions that are valid in the constrained solution space can make a difference.
+However, LCA is different in that it typically selects from a larger pool of windows and allows phase-steered variations of them.
+This gives LCA an adaptive freedom that lies somewhere in between that of traditional multi-apodization techniques and MVDR.
+How similarly LCA performs compared to MVDR depends on how well the predefined windows represent the constrained solution space of MVDR.
+
+% While using weights that span a bigger space gives LCA greater freedom, not The bigger the space the weights span, the greater the adaptive freedom, but only the solutions that are valid in the constrained solution space can make a difference.
+
+% Instead of adaptively computing the weights, it adaptively selects them.
 
@@
 suggested using a window set comprised of rectangular, Kaiser (or Kaiser-Harris) and inverted Kaiser functions.
-In total he made LCA choose from 12 unique windows, 6 of which were phase shifted versions of a fairly wide Kaiser window, and one rectangular.
-Phase shifting the windows has the effect of slighly adjusting the angle the array is steered towards.
-His composition of windows was motivated by the desire to use wide windows with maximum sidelobe suppression near the receiver, and narrow windows at greater depths for maximum sensitivity and depth penetration.
+In total he made LCA choose from 12 unique windows, 6 of which were phase-shifted versions of a fairly wide Kaiser window, and one were rectangular.
+Phase-shifting the windows has the effect of slighly adjusting the angle the array is steered towards.
+Synnev\aa{}g's composition of windows was motivated by the desire to use wide windows with maximum sidelobe suppression near the receiver, and narrow windows at greater depths for maximum sensitivity and depth penetration.
 While achieving good results, it was not made clear how many windows LCA needs, or how they ideally should be steered.
@@
 We extend Synnev\aa{}g's work by steering all our windows and not just some, and demonstrate how LCA's performance is affected by the type and number of windows included.
-Our results are obtained using experimental data from the 32 element Kongsberg Maritime HISAS 1030 synthetic aperture sonar (SAS), which operates at 100\,kHz with an -3\,dB element opening angle of 23$^\circ$.
+Our results are obtained using experimental data from the 32 element Kongsberg Maritime HISAS1030, which operates at 100\,kHz with an -3\,dB element opening angle of 23$^\circ$.
+Although this system was designed as synthetic aperture sonar (SAS) where an image is synthesized from several pings of data, we will only create images out of single pings (sectorscan imaging) in this study.
 
@@
 where $N_x = 2X+1$ is the number of azimuth lines to average over, and $\omega[x',n']$ is a normalized 2 dimensional weight function.
-However, we observed no visual improvement doing this, and the effect of micro-steered windows is hard to predict.
+However, we observed no visual improvement doing this, and the effect of phase-steered windows is hard to predict.
 Henceforth no lateral averaging will be applied.
@@
 To test the performance of the MVDR and LCA beamformers, we have processed data acquired by the Kongsberg Maritime HUGIN AUV carrying their 32 element HISAS1030 sonar~\cite{Hansen2011}.
-It is a high resolution synthetic aperture sonar with 1.2\;m array length, 100\;kHz operating frequency, 30\;kHz bandwidth and 23$^\circ$ element -3\,dB opening angle.
+It is a high resolution sonar with 1.2\;m array length, 100\;kHz operating frequency, 30\;kHz bandwidth and 23$^\circ$ element -3\,dB opening angle.
 

### 2016-03-02 LCAforActiveSonarImaging/submission/submit1/answer/answer.tex
la2pdf commit - 2016:03:02	13:35:59	(on. 2.mars)

@@
 \hi{Page 11}
-\qi{where is the synthetic aperture needed/applied in this work, or is it just the name of the array? how does the work on LCA/MVDR Change the Quality of synthetic aperture?
-what exactly is the sectorscan mode?}
-\ai{Thank you, we see that our formulation should be clarified.
-We are creating an image of the sector highlighted by a single ping of data (sectorscan), but are using data recorded by a synthetic aperture sonar (SAS).
-We mention it since SAS systems have a wide transmit beam intended to highlight multiple pings and thus does not produce the best sectorscan images.}
+\qi{where is the synthetic aperture needed/applied in this work, or is it just the name of the array?
+how does the work on LCA/MVDR Change the Quality of synthetic aperture?
+What exactly is the sectorscan mode?}
+\ai{The reviewer is quite right, it is a system designed for SAS imaging, but we only create image sectors for its data.
+That is, the images are from single ping, rather than synthesized from several pings.
+We agree this was worded in an ambiguous way and have now made relevant changes, as well as added a remark of what it meant by sectorscan.}
 %

### 2016-03-02 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2016:03:02	13:56:02	(on. 2.mars)

@@
 We extend Synnev\aa{}g's work by steering all our windows and not just some, and demonstrate how LCA's performance is affected by the type and number of windows included.
-Our results are obtained using experimental data from the 32 element Kongsberg Maritime HISAS1030, which operates at 100\,kHz with an -3\,dB element opening angle of 23$^\circ$.
+Our results are obtained using experimental data from the 32 element Kongsberg Maritime HISAS1030 sonar, which operates at 100\,kHz with an -3\,dB element opening angle of 23$^\circ$.
 Although this system was designed as synthetic aperture sonar (SAS) where an image is synthesized from several pings of data, we will only create images out of single pings (sectorscan imaging) in this study.
@@
 A weight set is commonly called a window, apodization or taper function.
-Applying a window that trails off towards the edges improves noise suppression at the cost of reduced resolution~\cite{Harris1978}.
-Real windows lead to symmetric responses, while complex weights allow asymmetric responses.
-
-Conventional beamformers all have static weights.
+If the window real its response is symmetric.
+Applying a real window that trails off towards the edges creates a response with lower sidelobe levels and wider mainlobe, which translates into improved noise suppression at the cost of reduced resolution, respectively~\cite{Harris1978}.
+
+Conventional beamformers all have static and usually real weights.
 The reference method is the delay-and-sum (DAS) beamformer, also know as the backprojection algorithm.
 It delays each pixel into focus, then applies a suitable window, and finally sums the data.
-The virtue of DAS is its simplicity, robustness to parameter errors, linear processing of the image and the ease of which it can be implemented on parallel hardware.
+The virtue of DAS is its simplicity, robustness to parameter errors, linear processing of the image and the ease of which it can be implemented in parallel hardware.
 
@@
 This may be achieved by allowing either the weights or delays to change, or both.
-One of the most extensively studied of these methods is the minimum variance distortion-less response (MVDR) beamformer.
+Here the weights are usually complex, which allows asymmetric responses.
+One of the most extensively studied adaptive methods is the minimum variance distortion-less response (MVDR) beamformer.
 

### 2016-03-02 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2016:03:02	14:55:20	(on. 2.mars)

@@
 A weight set is commonly called a window, apodization or taper function.
-If the window real its response is symmetric.
+When the window is real its response is symmetric.
 Applying a real window that trails off towards the edges creates a response with lower sidelobe levels and wider mainlobe, which translates into improved noise suppression at the cost of reduced resolution, respectively~\cite{Harris1978}.
@@
 
-Since pixels are correlated laterally, we can include a weighted combination of these as well to improve the variance estimation:
+If the pixels are correlated laterally, we can include a weighted combination of these as well to improve the variance estimation:
 %
@@
 where $N_x = 2X+1$ is the number of azimuth lines to average over, and $\omega[x',n']$ is a normalized 2 dimensional weight function.
-However, we observed no visual improvement doing this, and the effect of phase-steered windows is hard to predict.
+Since we oversample slightly laterally when we delay each pixel (see Section \ref{sec:lca_oversampling}), we thought the lateral correlation to be sufficient to benefit from this.
+However, we observed no visual improvement applying this averaging in our case.
+Also, since the effect of phase-steering windows is hard to predict.
 Henceforth no lateral averaging will be applied.

### 2016-03-02 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2016:03:02	15:25:47	(on. 2.mars)

@@
 The responses are overlayed each other and the amount of overlap is colored using a logarithmic scale.
+Since each pixel is pre-delayed into focus the unsteered responses all have their center at broadside.
 The dashed red lines at -13\,dB and -43\,dB marks the peak sidelobe levels of an unsteered rectangular and Hamming window, respectively.
@@
 Since we oversample slightly laterally when we delay each pixel (see Section \ref{sec:lca_oversampling}), we thought the lateral correlation to be sufficient to benefit from this.
-However, we observed no visual improvement applying this averaging in our case.
-Also, since the effect of phase-steering windows is hard to predict.
-Henceforth no lateral averaging will be applied.
+However, we observed no visual improvement from applying this technique, and decided not to use it.
 % 

### 2016-03-02 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2016:03:02	16:12:30	(on. 2.mars)

@@
 We let LCA select from a set of Kaiser windows with responses in this span, and add some steered variations of each.
-We limit the steering to roughly half the -3\,dB width of the window's amplitude response.
+We limit the steering to roughly half the \minus{}3\,dB width of the window's amplitude response.
 Using experimental data from the Kongsberg Maritime HISAS1030 sonar we find that LCA and MVDR produce nearly identical images of large scenes, both being superior to DAS.
@@
 We extend Synnev\aa{}g's work by steering all our windows and not just some, and demonstrate how LCA's performance is affected by the type and number of windows included.
-Our results are obtained using experimental data from the 32 element Kongsberg Maritime HISAS1030 sonar, which operates at 100\,kHz with an -3\,dB element opening angle of 23$^\circ$.
+Our results are obtained using experimental data from the 32 element Kongsberg Maritime HISAS1030 sonar, which operates at 100\,kHz with an \minus{}3\,dB element opening angle of 23$^\circ$.
 Although this system was designed as synthetic aperture sonar (SAS) where an image is synthesized from several pings of data, we will only create images out of single pings (sectorscan imaging) in this study.
 
-Our results show that when MVDR is modified to work in active sonar imaging, it computes weights with spatial responses that tend to be symmetric and steered within a fraction of the window's -\,3dB resolution.
+Our results show that when MVDR is modified to work in active sonar imaging, it computes weights with spatial responses that tend to be symmetric and steered within a fraction of the window's \minus{}3\,dB resolution.
 When letting LCA choose from Kaiser windows with similar responses, we obtain images with improved noise suppression and resolution.
-The resolution gain is made predictable by limiting the maximum steering to a fraction of the window's -3\,dB width.
+The resolution gain is made predictable by limiting the maximum steering to a fraction of the window's \minus{}3\,dB width.
 Overall, LCA produces images that are similar to MVDR, but with a resolution in between that of DAS and MVDR.
@@
 \fi%
-\caption{Each Kaiser window is steered in the interval $\phi=[0, \phi_{\text{-3dB}}(\beta)]$.
-The angle $\phi_{\text{-3dB}}(\beta)$ is the amount of steering needed for the steered response to have a -3dB crossing that is exactly half that of the unsteered window.
+\caption{Each Kaiser window is steered in the interval $\phi=[0, \phi_{\mathrm{3\,dB}}(\beta)]$.
+The angle $\phi_{\mathrm{3\,dB}}(\beta)$ is the amount of steering needed for the steered response to have a \protect\minus{}3\,dB crossing that is exactly half that of the unsteered window.
 With each window steered this way we expect the resolution gain to remain predictable and independent of $\beta$, and we also effectively constrain the white noise gain of the beamformer.}\label{windows_steering}
@@
 \emph{Right:}\hfill
-\parbox[t]{.95\linewidth}{MVDR window amplitude and phase responses computed from 40x40 pixel groups in the shadow, highlight and speckle region of the image.
+\parbox[t]{.95\linewidth}{MVDR window amplitude and phase responses computed from 40x40 pixel groups to provide sufficient statistics for the shadow, highlight and speckle region of the image.
 The responses are overlayed each other and the amount of overlap is colored using a logarithmic scale.
@@
 The upper left image is equal to DAS with a rectangular window.
-For all images we measure the lateral distance $\Delta x$ between the two -3dB points of the anchor (red object).
+For all images we measure the lateral distance $\Delta x$ between the two \protect\minus{}3\,dB points of the anchor (red object).
 These are specified relatively to the reference distance  $\Delta x_\text{ref}$ from the upper left DAS image.
@@
 To limit this we place an upper bound to the steering, chosen such that wide windows are allowed to be steered more than narrow ones.
-We call this upper bound for steering $\phi_{\text{3dB}}(\beta)$, and define it as the steering angle needed for the steered response to have a \minus{}3\,dB crossing that is exactly half that of the unsteered window.
-Using a numerical search for the steering angle $\phi$ that fulfilled this criterion for a wide range of $\beta$-values, we obtained a third degree polynomial fit to these points:
-%
-\begin{align}
-\phi_\text{3dB}(\beta) \approx \minus{}0.002894\beta^3 + 0.02451\beta^2 - 0.01308\beta + 0.2980\label{eq:steering_kaiser_3dB}
-\end{align}
-%
+We call this upper bound for steering $\phi_{\mathrm{3\,dB}}(\beta)$, and define it as the steering angle needed for the steered response to have a \minus{}3\,dB crossing that is exactly half that of the unsteered window.
+The method we devised for finding it is described in Appendix \ref{appendix:steering_angles}.
+
 Henceforth any steering will be specified relative to this upper bound.
@@
 where we apply the contraint $\alpha\in[0.5,1]$ to avoid windows with negative coefficients.
-Steering is applied as in (\ref{eq:steered_window}), and the $\alpha$-value that halves the 3\,dB distance is found using the method described in section \ref{sec:lca_steering_bounds}:
-%
-\begin{align}
-\phi_\text{3dB}(\alpha) \approx \minus{}2.214\beta^3 + 6.084\beta^2 - 5.710\beta + 2.136.
-\end{align}
+Steering is applied as in (\ref{eq:steered_window}), and the $\alpha$-value that halves the \minus{}3\,dB distance is found using the method described in Appendix \ref{appendix:steering_angles}.
 
@@
 We consider an absolute minimum to be 2, due to the non-linear nature of delaying pixels that isn't strictly in the far field, and of displaying the absolute value of the pixels on a decibel scale.
-In \mediaI{}\footnote{For media location refer to Appendix \ref{source_code}.} we visualize the effect of changing the lateral oversampling factor on a set of LCA images.
+In \mediaI{}\footnote{For media location refer to Appendix \ref{app:source_code}.} we visualize the effect of changing the lateral oversampling factor on a set of LCA images.
 Upon display the images are all bilinearly upinterpolated to the same size.
@@
 To test the performance of the MVDR and LCA beamformers, we have processed data acquired by the Kongsberg Maritime HUGIN AUV carrying their 32 element HISAS1030 sonar~\cite{Hansen2011}.
-It is a high resolution sonar with 1.2\;m array length, 100\;kHz operating frequency, 30\;kHz bandwidth and 23$^\circ$ element -3\,dB opening angle.
+It is a high resolution sonar with 1.2\;m array length, 100\;kHz operating frequency, 30\;kHz bandwidth and 23$^\circ$ element \minus{}3\,dB opening angle.
 
@@
 This data was collected by the Bundeswehr Technical Center for Ships and Naval Weapons, Maritime Technology and Research (WTD 71).
-Both scenes are imaged in sectorscan mode.
 
@@
 
-\mediaII{}\footnote{For media location refer to Appendix \ref{source_code}.} animates how \Fig{mvdr_selected_windows} changes as a function of the MVDR subarray length $L$ and temporal averaging $K$.
+\mediaII{}\footnote{For media location refer to Appendix \ref{app:source_code}.} animates how \Fig{mvdr_selected_windows} changes as a function of the MVDR subarray length $L$ and temporal averaging $K$.
 At $L=1$ the window responses are rectangular in all areas in the image.
@@
 The resulting images are shown in \Fig{oversampling_mosaic_bounds}.
-In each image we computed the lateral distance $\Delta x$ between the \minus{}3dB points of the anchor, with the reference $\Delta x_\text{ref}$ being the rectangularly weighted DAS image.
+In each image we computed the lateral distance $\Delta x$ between the \minus{}3\,dB points of the anchor, with the reference $\Delta x_\text{ref}$ being the rectangularly weighted DAS image.
 A fourth order polynomial fit was used in this computation.
@@
 This suggests the anchor reflections to be specular and that we view its rounded side.
-As we increase steering to either 50\% or 100\% of the -3dB width, the FWHM drops to 73\% or 53\% to that of the rectangular window, respectively.
+As we increase steering to either 50\% or 100\% of the \minus{}3\,dB width, the FWHM drops to 73\% or 53\% to that of the rectangular window, respectively.
 
@@
 On other scatterers with strong nearby lateral interference LCA prefers wider windows to suppress the interference.
-When allowed LCA prefers to steer windows away from the sources, but never exceed the $\phi_\text{3dB}(\beta)$ limit.
+When allowed LCA prefers to steer windows away from the sources, but never exceed the $\phi_\mathrm{3\,dB}(\beta)$ limit.
 When exceeding this bound we observed oscillation artifacts in the image.
@@
 
-So far we have found it reasonable to use Kaiser windows in the range $\beta\in[0,5]$, each steered within the range $|\phi|\in[0,\phi_\mathrm{3dB}(\beta)]$.
+So far we have found it reasonable to use Kaiser windows in the range $\beta\in[0,5]$, each steered within the range $|\phi|\in[0,\phi_\mathrm{3\,dB}(\beta)]$.
 What remains is to determine the number of windows needed.
@@
 % \end{tabular}\\
-% * Steering is relative to the -3dB width of the window being applied.
+% * Steering is relative to the \minus{}3\,dB width of the window being applied.
 % \caption{Qualitative study}\label{parameter_study}
@@
 % Sidelobe suppression & Low       & High   & Moderate \\
-% Steering             & Low       & Up to -3dB width & Moderate \\
+% Steering             & Low       & Up to \minus{}3\,dB width & Moderate \\
 % Oversampling         & 2x-8x     & 2x-8x  & 2x-8x
@@
 We achieved good results with 2-3 Kaiser windows in the range $\beta\in[0,5]$, and observed minimal difference when adding more windows or adjusting the upper bound of $\beta$.
-For the steering angle we suggest a value in the range $\phi\in[\pm\delta\phi_\text{3dB}(\beta)]$, where $\phi_\text{3dB}(\beta)$ is the bandwidth of the respective window, and $\delta$ is a scaling parameter.
+For the steering angle we suggest a value in the range $\phi\in[\pm\delta\phi_\mathrm{3\,dB}(\beta)]$, where $\phi_\mathrm{3\,dB}(\beta)$ is the bandwidth of the respective window, and $\delta$ is a scaling parameter.
 Setting $\delta$ to either 50\% or 100\% reduces the lateral point target size to 73\% or 53\% compared to the rectangular window, respectively.
@@
 
-\section{Steering angle}
+\section{Steering angle}\label{appendix:steering_angles}
 
@@
 \includegraphics[width=\linewidth]{gfx/buske8_online.eps}%
-\caption{\emph{Steering angle needed to cut the -3\,dB distance of the Kaiser window's spatial response in half}.
+\caption{\emph{Steering angle needed to cut the \protect\minus{}3\,dB distance of the Kaiser window's spatial response in half}.
 The needed amount of steering depends on the number of channels $M$, the element spacing relative to the wavelength $\frac{d}{\lambda}$, and the Kaiser parameter $\beta$.
@@
 
-To control the resolution gain of LCA, we wanted to determine the extent that the windows need to be steered to cut the angle of the -3\,dB point of the window's spatial amplitude response by a fixed amount, say, by a factor 2 as reference.
+To control the resolution gain of LCA, we wanted to determine the extent that the windows need to be steered to cut the angle of the \minus{}3\,dB point of the window's spatial amplitude response by a fixed amount, say, by a factor 2 as reference.
 We computed this steering angle for some common configurations of the mentioned parameters.
@@
 the rectangular window at $\beta=0$ and the near gaussian window at $\beta=5$, as well as one in between at $\beta=2.5$.
-Observe that for a given $M$ and $\beta$, the angle $\phi_\mathrm{3dB}$ scales near proportionally with $\frac{d}{\lambda}$.
+Observe that for a given $M$ and $\beta$, the angle $\phi_\mathrm{3\,dB}$ scales near proportionally with $\frac{d}{\lambda}$.
 We found this to be true also for $\frac{d}{\lambda}\in\{0.5,4\}$.
@@
 
-For the sake of completeness we also supply the source code, see Appendix \ref{source_code}.
+For the sake of completeness we also supply the source code, see Appendix \ref{app:source_code}.
 
@@
 
-\section{Multimedia Files and Source Code}\label{source_code}
+\section{Multimedia Files and Source Code}\label{app:source_code}
 

### 2016-03-02 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2016:03:02	18:04:38	(on. 2.mars)

@@
 
-\newif\ifBuildBibliography\BuildBibliographyfalse  % Whether to include the references.bib file or use inline references.
+\newif\ifBuildBibliography\BuildBibliographytrue  % Whether to include the references.bib file or use inline references.
 
@@
 On point targets LCA is able to double the resolution compared to DAS, or provide half that of MVDR.
-This performance is achieved with a total of 6 windows; the rectangular and Kaiser window with $\beta=5$, in unsteered version, and left and right steered to the steering limit.
+This performance is achieved with a total of 6 windows; the rectangular window and the Kaiser window with $\beta=5$, in an unsteered version, and versions that are left and right steered to the steering limit.
 Slightly smoother images are produced if the window count is increased to 15, but past this we observe minimal difference.
@@
 Based on an idea by Vignon~\cite{Vignon2008}, it was first introduced by Synnev\aa{}g \emph{et al}.
-in clinical medical ultrasound imaging~\cite{Synnevag2008}, who demonstrated its ability to obtain very similar image quality to MVDR in systems with focused transmit beams.
+in clinical medical ultrasound imaging~\cite{Synnevag2011}, who demonstrated its ability to obtain very similar image quality to MVDR in systems with focused transmit beams.
 LCA applies a set of predefined and static windows, or apodization weights, and selects the one offering the best noise suppression.
@@
 \emph{Right:}\hfill
-\parbox[t]{.95\linewidth}{MVDR window amplitude and phase responses computed from 40x40 pixel groups to provide sufficient statistics for the shadow, highlight and speckle region of the image.
+\parbox[t]{.95\linewidth}{MVDR window amplitude and phase responses computed from 40x40 pixel groups for the shadow, highlight and speckle region of the image.
 The responses are overlayed each other and the amount of overlap is colored using a logarithmic scale.
+The size of the pixel groups were chosen ad-hoc for the histograms to be near invariant to a shift in position within the same region.
 Since each pixel is pre-delayed into focus the unsteered responses all have their center at broadside.
@@
 
-Unlike the method described by Synnev\aa{}g~\cite{Synnevag2008} we do not let LCA choose from inverted Kaiser windows.
+Unlike the method described by Synnev\aa{}g~\cite{Synnevag2011} we do not let LCA choose from inverted Kaiser windows.
 These have a mainlobe width narrower than that of the rectangular window, but at the expense of a very poor white noise gain.

### 2016-03-02 LCAforActiveSonarImaging/submission/submit1/answer/answer.tex
la2pdf commit - 2016:03:02	13:56:02	(on. 2.mars)
la2pdf commit - 2016:03:02	16:12:30	(on. 2.mars)
la2pdf commit - 2016:03:02	18:04:38	(on. 2.mars)

@@
 \qi{"with an element beamwidth" $\rightarrow$ what is meant by element beamwidth? Maybe an additional figure could be used to explain this.}
-\ai{Good remark. We meant the -3dB opening angle of the elements. We have reworded this the two places it is mentioned.}
+\ai{Good remark. We meant the -3dB opening angle of the elements. We have reworded this the two places it was mentioned.}
 %
@@
 $\rightarrow$ The difference between noise and sidelobes should be pointed out.}
-\ai{}
+\ai{Good observations. We realize this was ambiguous and have now rewritten this paragraph in hopes it will be clearerr.}
 %
@@
 \qi{The lateral correlation and the reason why equation (5) is a good choice should be better explained.}
-\ai{Indeed it should, this paragraph is now rewritten.}
+\ai{Indeed. We have rewritten the paragraph slightly, but the technique is mentioned mostly for completeness. It wasn't used.}
 %
@@
 \qi{"they are optimal for any value" $\rightarrow$ optimal in which sense?}
-\ai{Thank you. This was mentioned earlier in the same paragraph, just after (8).}
+\ai{Thank you. This was mentioned earlier in the same paragraph, just after (8). Is this insufficient?}
 %
@@
 \qi{why is a 40x40 pixel group a good choice? could the associated broadside direction be made visible in the figure, too?}
-\ai{This is an excellent question. This should perhaps be related to the detail level in the image, which in turn depends on the array's imaging performance.}
+\ai{This is an excellent question.
+This should perhaps be related to the detail level in the image, which in turn depends on the array's imaging performance.
+We took an ad-hoc approach to this and selected 40x40 pixels to get sufficient statistics for a visually good looking histogram.
+We have now written this into the caption of Fig.
+3.}
 %
@@
 \qi{reference to Fig.1 $\rightarrow$ could the sequence of figures be changes to have Fig. 1 closer to this paragraph}
-\ai{Valid point. In our two column print the figure is on the page before its first reference. }
+\ai{We agree this would be better. In our two column print the figure is on the page before its first reference. We assume this will be placed properly when processed for publishing.}
 %
@@
 
-\subsection*{Suggestions for improving the exposition of the processing}
-
 \newpage
@@
 \begin{enumerate}
+\item Fig 3.
+Caption changed to indicate that the size 40x40 pixels was chosen for sufficient statistics to make the histogram look well.
+Also added a sentence explaining that the steering is relative to broadside for all pixels, since each of them are pre-delayed into focus.
 \item Fig 8 is new.
@@
 
-The steering bound $\phi_\mathrm{3dB}$ depends on not just $\beta$, but also on the number of elements $M$ and the element spacing relative to the wavelength $\frac{d}{\lambda}$.
+\begin{enumerate}
+\item The steering bound $\phi_\mathrm{3dB}$ depends on not just $\beta$, but also on the number of elements $M$ and the element spacing relative to the wavelength $\frac{d}{\lambda}$. 
 \begin{enumerate}
@@
 Appendix C now replace and extends these equations, and we removed them in hope this will make the article easier to read.
+\end{enumerate}
+\item Various spelling flaws attempted corrected.
+\item Appendix C added. Contains information of where source code and media files can be located. Before the article goes to publishing we will make sure that this link is permanent.
 \end{enumerate}

### 2016-03-02 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2016:03:02	18:11:20	(on. 2.mars)

@@
 
+\bibitem{Synnevag2011}
+J.-F. Synnev\aa{}g, A.~Austeng, and S.~Holm, ``{A low-complexity data-dependent
+  beamformer},'' \emph{IEEE Transactions on Ultrasonics, Ferroelectrics and
+  Frequency Control}, vol.~58, no.~2, pp. 281--289, feb 2011.
+
 \bibitem{Synnevag2008}

### 2016-03-03 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2016:03:03	09:43:51	(to. 3.mars)

@@
 
-\begin{figure}[t]%\centering%
+\begin{figure}[t]
+\ifPeerReview\centering%
+\includegraphics[width=.5\linewidth]{gfx/buske8_online.eps}%
+\else%
 \includegraphics[width=\linewidth]{gfx/buske8_online.eps}%
+\fi%
 \caption{\emph{Steering angle needed to cut the \protect\minus{}3\,dB distance of the Kaiser window's spatial response in half}.
 The needed amount of steering depends on the number of channels $M$, the element spacing relative to the wavelength $\frac{d}{\lambda}$, and the Kaiser parameter $\beta$.
+Note how the angle scales near proportionally with $\frac{d}{\lambda}$.
 }\label{steering_angle_3dB}
@@
 
-This article is accompanied by some multimedia files and source code for determining the steering angles for LCA's windows. It is located at:
+This article is accompanied by some multimedia files and source code for determining the steering angles for LCA's windows. They are located at:
 
@@
 
-\bibitem{Synnevag2008}
-J.-F. Synnev\aa{}g, S.~Holm, and A.~Austeng, ``{A low complexity data-dependent
-  beamformer},'' in \emph{2008 IEEE Ultrasonics Symposium}, no.~2.\hskip 1em
-  plus 0.5em minus 0.4em\relax IEEE, Nov. 2008, pp. 1084--1087.
-
 \bibitem{Stankwitz1995}

### 2016-03-03 LCAforActiveSonarImaging/submission/submit1/answer/answer.tex
la2pdf commit - 2016:03:02	18:11:20	(on. 2.mars)
la2pdf commit - 2016:03:03	09:43:51	(to. 3.mars)

@@
 We would like to thank the associate editor for his efforts and for his speedy replies to queries from the authors, and reviewers for their thorough and highly relevant comments.
-Subsequent sections will target each reviewer's comments in detail.
-We quote in \textit{italic} statements from the reviewers, and provide our replies in ordinary print,
+The following sections mentions some general changes and the reviewer's comments are answered in detail in the next pages.
+We quote in \textit{italic} statements from the reviewers, and provide our replies in ordinary print.
+
+\newpage
+
+\section{Figure changes}
+
+\begin{enumerate}
+\item Fig 3.
+Caption changed to indicate that the size 40x40 pixels was chosen for sufficient statistics to make the histogram look well.
+Also added a sentence explaining that the steering is relative to broadside for all pixels, since each of them are pre-delayed into focus.
+\item Fig 8 is new.
+It was added to provide details on how the steering bound $\phi_\mathrm{3dB}$ is affected by the number of elements $M$ and the element spacing relative to the wavelength $\frac{d}{\lambda}$.
+\end{enumerate}
+
+\section{Other changes}
+
+\begin{enumerate}
+\item The steering bound $\phi_\mathrm{3dB}$ depends on not just $\beta$, but also on the number of elements $M$ and the element spacing relative to the wavelength $\frac{d}{\lambda}$. 
+\begin{enumerate}
+\item We have supplied additional info on how the steering is affected by these parameters in appendix C.
+\item Equations (12) and (14) only gave the correct results for our system configuration, and were not very intuitive.
+Appendix C now replace and extends these equations, and we removed them in hope this will make the article easier to read.
+\end{enumerate}
+\item Various spelling flaws attempted corrected.
+\item Appendix C added. Contains information of where source code and media files can be located. Before the article goes to publishing we will make sure that this link is permanent.
+\item Reference to Synnevåg's proceedings article from 2008, ``A low-complexity data-dependent beamformer'' replaced with his 2011 journal article with the same name.
+We never intended to reference the proceedings article since this work extends his journal version.
+\end{enumerate}
 
@@
 
-\newpage
-
-\section{Figure changes}
-
-\begin{enumerate}
-\item Fig 3.
-Caption changed to indicate that the size 40x40 pixels was chosen for sufficient statistics to make the histogram look well.
-Also added a sentence explaining that the steering is relative to broadside for all pixels, since each of them are pre-delayed into focus.
-\item Fig 8 is new.
-It was added to provide details on how the steering bound $\phi_\mathrm{3dB}$ is affected by the number of elements $M$ and the element spacing relative to the wavelength $\frac{d}{\lambda}$.
-\end{enumerate}
-
-\section{Other changes}
-
-\begin{enumerate}
-\item The steering bound $\phi_\mathrm{3dB}$ depends on not just $\beta$, but also on the number of elements $M$ and the element spacing relative to the wavelength $\frac{d}{\lambda}$. 
-\begin{enumerate}
-\item We have supplied additional info on how the steering is affected by these parameters in appendix C.
-\item Equations (12) and (14) only gave the correct results for our system configuration, and were not very intuitive.
-Appendix C now replace and extends these equations, and we removed them in hope this will make the article easier to read.
-\end{enumerate}
-\item Various spelling flaws attempted corrected.
-\item Appendix C added. Contains information of where source code and media files can be located. Before the article goes to publishing we will make sure that this link is permanent.
-\end{enumerate}
-
 % 

### 2016-05-01 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2016:05:01	20:45:42	(sø. 1.mai)

@@
 
-To build confidence in the LCA method we show that a robust MVDR implementation typically creates weights spanning between the rectangular and Hamming window function.
+To build confidence in the LCA method we show that a robust MVDR implementation typically creates weight sets with shapes spanning between a rectangular and Hamming window function.
 We let LCA select from a set of Kaiser windows with responses in this span, and add some steered variations of each.
@@
 This is why human made sonars have little to no adaptivity on transmit.
-However, several methods have been investigated to adaptively post process the received data.
-Perhaps the best studied method for reconstructing a sonar image out of the data is the minimum variance distortion-less response (MVDR) beamformer~\cite{Capon1969}.
+However, several methods have been investigated to adaptively process the received acustic data.
+Perhaps the best studied method for reconstructing a sonar image out of such data is the minimum variance distortion-less response (MVDR) beamformer~\cite{Capon1969}.
 It adapts the sonar array's spatial response to minimize the influence of noise and interference in the final image.
@@
 For larger systems the inversion step dominates.
-This can be dealt with by approximating the full covariance matrix with a small one using space reduction techniques such as beamspace processing~\cite{Asen2013}, or the closely related principal component analysis (PCA) method~\cite{Kim2014}.
+This can be dealt with by approximating the full covariance matrix with a small one using space reduction techniques such as beamspace processing~\cite{Asen2013}, or the closely related principal component analysis (PCA)~\cite{Kim2014}.
 Another alternative is to assume spatial stationarity to form more easily invertible Toepliz matrices~\cite{Asl2012}.
@@
 One such method is described by Stankwitz \emph{et al}.
-in radar imaging~\cite{Stankwitz1995}, where the best out of 2 or 3 windows are selected.
-However, LCA is different in that it typically selects from a larger pool of windows and allows phase-steered variations of them.
+in radar imaging~\cite{Stankwitz1995}, where the best out of 2 or 3 windows is selected.
+However, LCA differs in that it typically selects from a larger pool of windows and allows phase-steered variations of them.
+Phase-shifting the windows has the effect of slighly adjusting the angle the array is steered towards.
 This gives LCA an adaptive freedom that lies somewhere in between that of traditional multi-apodization techniques and MVDR.
@@
 suggested using a window set comprised of rectangular, Kaiser (or Kaiser-Harris) and inverted Kaiser functions.
-In total he made LCA choose from 12 unique windows, 6 of which were phase-shifted versions of a fairly wide Kaiser window, and one were rectangular.
-Phase-shifting the windows has the effect of slighly adjusting the angle the array is steered towards.
+In total he let LCA choose from 12 unique windows, 6 of which were phase-shifted versions of a fairly wide Kaiser window, and one were rectangular.
 Synnev\aa{}g's composition of windows was motivated by the desire to use wide windows with maximum sidelobe suppression near the receiver, and narrow windows at greater depths for maximum sensitivity and depth penetration.
-While achieving good results, it was not made clear how many windows LCA needs, or how they ideally should be steered.
+While he achieved good results, it was not made clear how many windows LCA needs, or how they ideally should be steered.
 Also, it is not apparent how LCA performs in sonar systems where the transmit beam is unfocused.
@@
 Our results are obtained using experimental data from the 32 element Kongsberg Maritime HISAS1030 sonar, which operates at 100\,kHz with an \minus{}3\,dB element opening angle of 23$^\circ$.
-Although this system was designed as synthetic aperture sonar (SAS) where an image is synthesized from several pings of data, we will only create images out of single pings (sectorscan imaging) in this study.
-
-Our results show that when MVDR is modified to work in active sonar imaging, it computes weights with spatial responses that tend to be symmetric and steered within a fraction of the window's \minus{}3\,dB resolution.
+Although this system was designed as a synthetic aperture sonar (SAS) where an image is synthesized from several pings of data, we will only create images out of single pings (sectorscan imaging) in this study.
+
+Our results show that when MVDR is robustified to work in active sonar imaging, it computes weights with spatial amplitude responses that tend to be symmetric and steered within a fraction of their \minus{}3\,dB widths.
 When letting LCA choose from Kaiser windows with similar responses, we obtain images with improved noise suppression and resolution.
-The resolution gain is made predictable by limiting the maximum steering to a fraction of the window's \minus{}3\,dB width.
+The resolution gain is made predictable by limiting the maximum steering to a fraction of the \minus{}3\,dB width of each window's spatial amplitude response.
 Overall, LCA produces images that are similar to MVDR, but with a resolution in between that of DAS and MVDR.
@@
 Results and discussion are provided in Section \ref{sec:results_discussion}, where we study MVDR's windows to find suitable LCA ones in \ref{sec:results_lca_window_function}, assess the range of window types and steering in \ref{sec:results_window_parameters}, determine the number of windows needed in \ref{sec:results_database_size}, and add a note on computational complexity in \ref{sec:results_complexity}.
-Section \ref{sec:conclusion} provides a conclusion.
+Finally, Section \ref{sec:conclusion} provides a conclusion.
 
@@
 
-MVDR seeks to minimize the output variance of the beamformer, under the constraint of a unity gain in some desired direction $\phi$~\cite{Capon1969}.
-With the assumption of zero mean data this can be formulated as:
+MVDR seeks to minimize the power of noise and interference in the output of the beamformer, under the constraint of unity gain in some desired direction $\phi$~\cite{Capon1969}.
+With the assumption of zero mean data this takes the form of minimizing output variance:
 %
@@
 %
-Here $\a_\phi$ is a steering vector, $E\{\cdot\}$ is the expectation operator, and $\R[n] = E\{\x[n]\x\H[n]\}$ is the spatial covariance matrix for the full array.
-This is a convex optimization problem with the solution:
+Here $\a_\phi$ is a steering vector for an azimuthal steering angle $\phi$, $E\{\cdot\}$ is the expectation operator, and $\R[n] = E\{\x[n]\x\H[n]\}$ is the spatial covariance matrix for the full array.
+This is a convex optimization problem with the unique solution:
 %
@@
 \end{gather}
-The problem lies in estimating and inverting this spatial covariance matrix.
-We have described the exact steps in~\cite{Buskenes2014}: To avoid signal cancellation we apply spatial averaging by computing a mean covariance matrix from a set of subarrays with length $L$~\cite{Kailath1985}, for true speckle statistics we perform temporal averaging over $N_k = 2K+1$ temporal samples~\cite{Synnevag2009a}, and to improve robustness to parameter errors we add {\large$\epsilon$} percent diagonal loading~\cite{Cox1987,Maksym1979}.
+The problem lies in estimating and inverting the spatial covariance matrix.
+We have described the exact steps in~\cite{Buskenes2014}: To avoid signal cancellation we apply spatial averaging by computing a mean covariance matrix from a set of subarrays with length $L$~\cite{Kailath1985}, for true speckle statistics we perform temporal averaging over $N_k = 2K+1$ temporal samples~\cite{Synnevag2009a}, and to improve robustness to parameter errors we add {\large$\epsilon$} percent of the total output power to the matrix's diagonal~\cite{Cox1987,Maksym1979}.
 These steps are also needed to ensure that the covariance matrix is numerically well conditioned and hence invertible.
@@
 \fi%
-\caption{Each Kaiser window is steered in the interval $\phi=[0, \phi_{\mathrm{3\,dB}}(\beta)]$.
-The angle $\phi_{\mathrm{3\,dB}}(\beta)$ is the amount of steering needed for the steered response to have a \protect\minus{}3\,dB crossing that is exactly half that of the unsteered window.
+\caption{Each Kaiser window is steered in the interval $\phi\in[0, \phi_{\mathrm{3\,dB}}(\beta)]$.
+The angle $\phi_{\mathrm{3\,dB}}(\beta)$ is the amount of steering needed for the steered amplitude response to have a \protect\minus{}3\,dB crossing that is exactly half that of the unsteered window.
 With each window steered this way we expect the resolution gain to remain predictable and independent of $\beta$, and we also effectively constrain the white noise gain of the beamformer.}\label{windows_steering}
@@
 \fi%
-\caption{One of the objects used to test the beamformers is a 1\,m by 1\,m resolution test cross attached to an anchor with a diameter of approximately 13\,cm.
+\caption{The submerged object used to test beamforming resolution is a 1\,m by 1\,m test cross attached to an anchor with a diameter of approximately 13\,cm.
 Source image curtesy of Bundeswehr Technical Center for Ships and Naval Weapons, Maritime Technology and Research (WTD 71).}\label{cross}
@@
 \emph{Left:}\hfill
-\parbox[t]{.95\linewidth}{MVDR sectorscan image of the oiltanker Holmengraa, with the pixel groups are indicated with red boxes.}\protect\\\hspace{\textwidth}
+\parbox[t]{.95\linewidth}{MVDR sectorscan image of the oiltanker Holmengraa, with 40x40 pixel groups for the shadow, highlight and speckle region of the image indicated with red boxes.}\protect\\\hspace{\textwidth}
 \emph{Right:}\hfill
-\parbox[t]{.95\linewidth}{MVDR window amplitude and phase responses computed from 40x40 pixel groups for the shadow, highlight and speckle region of the image.
+\parbox[t]{.95\linewidth}{MVDR window amplitude and phase responses computed from the 40x40 pixel groups.
 The responses are overlayed each other and the amount of overlap is colored using a logarithmic scale.
-The size of the pixel groups were chosen ad-hoc for the histograms to be near invariant to a shift in position within the same region.
+The size of the pixel groups were chosen ad-hoc for the histograms to be visually invariant to a shift in position within the same region.
 Since each pixel is pre-delayed into focus the unsteered responses all have their center at broadside.
@@
 \includegraphics[width=\textwidth]{gfx/buske4_online.pdf}%
-\caption{\emph{Determining Kaiser parameter boundaries}: LCA images of the resolution test cross and anchor created with a large and equal size window database, but where the upper bounds for $\beta$ and $\phi$ is varied.
+\caption{\emph{Determining Kaiser parameter boundaries}: LCA images of the resolution test cross and anchor created with a large window database, but where the upper bounds for $\beta$ and $\phi$ are varied.
 The upper left image is equal to DAS with a rectangular window.
@@
 \parbox[t]{.89\linewidth}{Observe that LCA gets better at suppressing sidelobes in the image as we increase the upper bounds for $\beta$.
-\Fig{mvdr_selected_windows} suggests that MVDR tend to choose windows with levels lower than that of Kaiser with $\beta=2$, but here we observe further improvement going to $\beta=5$.}\newline
+\Fig{mvdr_selected_windows} suggests that MVDR prefers windows with sidelobe levels lower than that of Kaiser with $\beta=2$, but here we observe further improvement going to $\beta=5$.}\newline
 \emph{Upper bound $\phi$:}\hfill
@@
 \subfloat[$\beta$-values of the Kaiser window chosen for each image pixel.
-Note how LCA prefers a a narrow response ($\beta=0$) on the anchor, and on other strong sources with little lateral interference.
+Note how LCA prefers a narrow response ($\beta=0$) on the anchor, and on other strong sources with little lateral interference.
 On the sources where lateral interference is present it chooses the widest response with the best sidelobe suppression.]{\includegraphics[width=\textwidth]{gfx/buske5a_online.pdf}%%
@@
 \label{oversampling_mosaic_bounds_phi}}
-\caption{Kaiser windows chosen for each pixel in the image.
+\caption{Kaiser windows chosen for each pixel in the image of the resolution test cross.
 The underlying image is the one shown in \Fig{oversampling_mosaic_bounds}, with the same parameters.
@@
 The upper left image is identical to a rectangularly weighted DAS.
-Observe that sampling the $\beta$-range more finely increases the sidelobe suppression but leaves resolution unchanged.
+Observe that finer sampling of the $\beta$-range improves sidelobe suppression but leaves resolution unchanged.
 Adding more steering-variations improves both sidelobe suppression and resolution.
@@
 The optimization criterion and constraint is the same, but LCA has a finite and discrete solution space for the weights.
-As will be demonstrated in section \ref{sec:results_lca_window_function}, LCA will perform similarly to MVDR because a robustified MVDR implementation seems to be constrained to window functions similar to the ones we let LCA choose from.
+As will be demonstrated in Section \ref{sec:results_lca_window_function}, LCA performs similarly to MVDR because a robustified MVDR implementation seems to be constrained to window functions similar to the ones we let LCA choose from.
 
@@
 
-If the pixels are correlated laterally, we can include a weighted combination of these as well to improve the variance estimation:
+If the pixels are correlated laterally, we can include a weighted combination of these to improve the variance estimation:
 %
@@
 Adding slightly steered versions of each window to the window database gives LCA greater flexibility in searching for an optimal window.
-A Kaiser window $\f_\beta$ steered to the angle $\phi$ can be expressed as
+A Kaiser window $\f_\beta$ steered to the azimuth angle $\phi$ can be expressed as
 %
@@
 
-Each window's response is constrained to unit gain in the look direction.
+Each window's spatial response is constrained to unit gain in the look direction.
 Hence, when it is steered the white noise gain must increase and the the signal-to-noise ratio decrease, as shown in \Fig{windows_steering}.
 To limit this we place an upper bound to the steering, chosen such that wide windows are allowed to be steered more than narrow ones.
-We call this upper bound for steering $\phi_{\mathrm{3\,dB}}(\beta)$, and define it as the steering angle needed for the steered response to have a \minus{}3\,dB crossing that is exactly half that of the unsteered window.
+We call this upper bound for steering $\phi_{\mathrm{3\,dB}}(\beta)$, and define it as the steering angle needed for the steered amplitude response to have a \minus{}3\,dB crossing that is exactly half that of the unsteered window.
 The method we devised for finding it is described in Appendix \ref{appendix:steering_angles}.
@@
 
-While we focus on using LCA with Kaiser windows, we will for completeness compare the performance with that of using trigonometric windows. This is defined as:
-%
-\begin{align}
-f_m(\alpha) = \alpha - (1-\alpha)\cos\Big(\frac{2\pi m}{M-1}\Big)\label{eq:trig_window_function}
+While we focus on using LCA with the Kaiser window function, we will for comparison also test it with the trigonometric window function. Its definition is:
+%
+\begin{align}
+f_m(\alpha) = \alpha - (1-\alpha)\cos\Big(\frac{2\pi m}{M-1}\Big),\label{eq:trig_window_function}
 \end{align}
 %
-where we apply the contraint $\alpha\in[0.5,1]$ to avoid windows with negative coefficients.
+where we apply the constraint $\alpha\in[0.5,1]$ to avoid windows with negative coefficients.
 Steering is applied as in (\ref{eq:steered_window}), and the $\alpha$-value that halves the \minus{}3\,dB distance is found using the method described in Appendix \ref{appendix:steering_angles}.
@@
 In \mediaI{}\footnote{For media location refer to Appendix \ref{app:source_code}.} we visualize the effect of changing the lateral oversampling factor on a set of LCA images.
-Upon display the images are all bilinearly upinterpolated to the same size.
+Upon display the images were all bilinearly upinterpolated to the same size.
 % 
@@
 
-The results will be presented in the following order: In section \ref{sec:results_lca_window_function} we discuss typical window responses computed by MVDR, and hypothesize that these can be mimicked by Kaiser windows.
-In section \ref{sec:results_window_parameters} we determine sensible boundaries for the Kaiser parameter $\beta$ and window steering $\phi$.
-In section \ref{sec:results_database_size} we discuss how many window variations are needed for LCA to perform well.
+The results will be presented in the following order: In Section \ref{sec:results_lca_window_function} we discuss typical window responses computed by MVDR, and hypothesize that these can be mimicked by Kaiser windows.
+In Section \ref{sec:results_window_parameters} we determine sensible boundaries for the Kaiser parameter $\beta$ and window steering $\phi$.
+In Section \ref{sec:results_database_size} we discuss how many window variations are needed for LCA to perform well.
 
@@
 Assuming that a robust MVDR is the reference method we want LCA to perform similarly to, it seems sensible to create a window database for LCA with spatial responses similar to the ones that MVDR computes.
-We study this in \Fig{mvdr_selected_windows}, where we present the typical amplitude and phase responses that MVDR creates for shadow, speckle and highlight regions of the Holmengraa scene.
+We study this in \Fig{mvdr_selected_windows}, where we present the typical amplitude and phase responses for the windows that MVDR creates in shadow, speckle and highlight regions of the Holmengraa scene.
 
@@
 The symmetry is most predominant in shadow and speckle regions, while in highlight the windows are approximately symmetric within the illuminated sector of the seafloor.
-We have observed this symmetry to be a side-effect of the averaging steps needed to build the sample covariance matrix, which are required for MVDR to operate in near snapshot mode in an active system.
+We have observed this symmetry to be a side-effect of the averaging steps needed to build the sample covariance matrix, which are required for MVDR to operate with very little to no temporal sample support in an active system.
 The peak sidelobe levels of the MVDR windows are mostly between \minus{}13\,dB and \minus{}43\,dB.
@@
 If they span a suitable range of $\beta$-values, and some steering variations are applied to each, we should have responses that resemble those in \Fig{mvdr_selected_windows}.
-The Kaiser window is easy to compute, is fairly insensitive to coefficient inaccuracies, and can create windows with shapes spanning from rectangular to Gaussian.
+The Kaiser window is easy to compute, is fairly insensitive to coefficient inaccuracies, and span shapes from rectangular to Gaussian.
 It is also optimal in the sense of having its peak power concentration near the steering angle.
@@
 Unlike the method described by Synnev\aa{}g~\cite{Synnevag2011} we do not let LCA choose from inverted Kaiser windows.
-These have a mainlobe width narrower than that of the rectangular window, but at the expense of a very poor white noise gain.
+These have a mainlobe width narrower than that of the rectangular window, but at the expense of very poor white noise gain.
 We have found that these windows hardly ever get used on experimental data, in particular steered versions of them.
@@
 Already at $L=2$ the MVDR is able to greatly suppress noise.
-This is a common observation; adding a little flexibility to adapt to the scene has a dramatic effect, but further allowing full flexibility is much less significant.
+This is a common observation; adding a little flexibility to adapt to the scene has a dramatic effect, but allowing full flexibility is much less significant.
 The media file also shows that MVDR can be run with subarray sizes $L\in[M/2,5M/8]$, but only with temporal averaging $K=1$ or above.
@@
 
-To determine a suitable range for the Kaiser parameter $\beta$ and steering $\phi$, we constructed large and equally sized window databases containing Kaiser windows with varied upper boundaries for $\beta$ and $\phi$.
+To determine a suitable range for the Kaiser parameter $\beta$ and phase steering $\phi$, we constructed large and equally sized window databases containing Kaiser windows with varied upper boundaries for $\beta$ and $\phi$.
 The lower boundaries were chosen as $\beta=0$ and $\phi=0^\circ$, which includes the rectangularly weighted DAS image.
@@
 A fourth order polynomial fit was used in this computation.
-This measure is also commonly called the full width half maximum (FWHM).
-In the case of a imaging a point scatter, it is closely related to the resolution of of the system.
-For a rectangular window, for instance, the system resolution is approximately $\Delta \approx \text{FWHM}/0.89$~\cite{Harris1978}.
+The measure $\Delta x$ is also commonly called the full width half maximum (FWHM).
+In the case of a imaging a point scatter, it is closely related to the resolution of the system.
+For a rectangular window, for instance, the system resolution is approximately $\delta x \approx \Delta x/0.89$~\cite{Harris1978}.
 
@@
 The images also demonstrate that increasing the upper bound for steering $\phi$ improves the resolution of the strong scatterers.
-For this sonar the Rayleigh resolution is $\Delta\approx0.0125$ radians (0.72$^\circ$), so a point scatterer imaged at 19\,m range with a rectangular DAS would have a lateral $\text{FWHM}\approx0.125\cdot19\,\text{m}\cdot0.89=21\,\text{cm}$.
+For this sonar the Rayleigh resolution is $\delta x\approx0.0125$ radians (0.72$^\circ$), so a point scatterer imaged at 19\,m range with a rectangular DAS would have a lateral $\Delta x\approx0.0125\cdot19\,\text{m}\cdot0.89=21\,\text{cm}$.
 This is close to the FWHM of the anchor in the DAS image in \Fig{oversampling_mosaic_bounds}, which means that the acoustic fingerprint of the anchor is similar to a point source.
-This suggests the anchor reflections to be specular and that we view its rounded side.
+This further implies that the anchor reflections are specular and that we view its rounded side.
 As we increase steering to either 50\% or 100\% of the \minus{}3\,dB width, the FWHM drops to 73\% or 53\% to that of the rectangular window, respectively.
@@
 On other scatterers with strong nearby lateral interference LCA prefers wider windows to suppress the interference.
-When allowed LCA prefers to steer windows away from the sources, but never exceed the $\phi_\mathrm{3\,dB}(\beta)$ limit.
+When allowed LCA prefers to steer windows away from the sources, but we never let it exceed the $\phi_\mathrm{3\,dB}(\beta)$ limit.
 When exceeding this bound we observed oscillation artifacts in the image.
@@
 What remains is to determine the number of windows needed.
-We study this in \Fig{oversampling_mosaic}, where we compared LCA images made from windows databases of different sizes, but with the same parameter boundaries.
-We included the results from using a single window ($N_\beta=1$) and steering angle ($N_\phi=1$), which corresponds to an unsteered rectangularly weighted DAS.
-
-Observe that sampling the $\beta$-range more finely increases the sidelobe suppression but leaves resolution unchanged.
+We study this in \Fig{oversampling_mosaic}, where we compare LCA images made from windows databases of different sizes, but with the same parameter boundaries.
+Included are the results from using a single window ($N_\beta=1$) and steering angle ($N_\phi=1$), which corresponds to an unsteered rectangularly weighted DAS.
+
+Observe that a finer sampling of the $\beta$-range improves sidelobe suppression but leaves resolution unchanged.
 Adding more steering-variations improves both sidelobe suppression and resolution.
-We can observe a major improvement going to $N_\beta=2$ and $N_\phi=3$, but minimal improvement by adding yet more windows.
+We observe major improvement going to $N_\beta=2$ and $N_\phi=3$, but minimal improvement by adding yet more windows.
 
@@
 The images and their corresponding cuts are nearly identical for the LCA version with Kaiser windows and the one with trigonometric windows.
-Compared to DAS the LCA produces targets with an FWHM that lies in between that of DAS and MVDR.
+Compared to DAS the LCA produces images with an FWHM that lies in between that of DAS and MVDR.
 
@@
 However, the implementation we are using is optimized so that building the spatial covariance matrix is of O($N_kN_L,L^2$), and inverting it is of O($L^3$).
-We describe this implementation in \cite{Buskenes2014}, and a beamspace version in \cite{Asen2013}, where we optimize if for a graphics processing unit (GPU) for nearly two orders of magnitude speed increase compared to a straightforward C implementation.
-However, MVDR is not ideal for GPUs due to the complex data dependencies present in the covariance computation, and at best we only managed to utilize 10\% of the GPUs potential.
+We describe this implementation in \cite{Buskenes2014}, and a beamspace version in \cite{Asen2013}, where we implement it on a graphics processing unit (GPU) for nearly two orders of magnitude speed increase compared to a straightforward C implementation.
+However, MVDR is not ideal for GPUs due to the complex data dependencies present in the covariance computation, and at best we only managed to utilize 10\% of the GPUs maximum theoretical potential.
 
@@
 Each pixel depends on only a small subset of data, and the windows can be precomputed and stored in GPU cache for near immediate access.
-If we use a trigonometric window function we solve for the window parameter analytically as described in appendix \ref{sec:reducing_lca_complexity}.
-This reduces the complexity to O($M N_\phi$), and the only remaining LCA parameter is the steering amount.
+Furthermore, if we use the trigonometric window function we can solve for the window parameter analytically as described in Appendix \ref{sec:reducing_lca_complexity}.
+This reduces the complexity to O($M N_\phi$), because the only remaining LCA parameter to create windows for will be the steering angle.
 
@@
 
-We found the LCA to be surprisingly insensitive to the exact window function it is used with, supported by the observation that near identical images were produced whether we used Kaiser windows or trigonometric ones.
+We found LCA to be surprisingly insensitive to the exact window function it is used with, supported by the observation that near identical images were produced whether we used Kaiser windows or trigonometric ones.
 However, LCA must be able to choose between wide and narrow window responses, since the wide ones offer the best sidelobe suppression and the narrow ones offer the best sensitivity.
@@
 
-In summary, LCA is an attractive alternative to other adaptive beamformers due to being very fast, simple to understand, practically parameter-free, inherently robust, and able produce images similar to that of MVDR.
+In summary, LCA is an attractive alternative to other adaptive beamformers due to being very fast, simple to understand, practically parameter-free, inherently robust, and able to produce images similar to that of MVDR.
 Contrary to MVDR, it has a low computational complexity and can be easily and efficiently accelerated using e.g.
 GPUs.
-Since LCA performs just as well if we use trigonometric windows instead of Kaiser ones, we can also compute the optimal window parameter analytically for any given steering angle, and avoid the search over the window parameter entirely.
+Also, since LCA performs just as well if we use trigonometric windows instead of Kaiser ones, we can compute the optimal window parameter analytically for any given steering angle and avoid the search over the window parameter entirely.
 
@@
 To control the resolution gain of LCA, we wanted to determine the extent that the windows need to be steered to cut the angle of the \minus{}3\,dB point of the window's spatial amplitude response by a fixed amount, say, by a factor 2 as reference.
-We computed this steering angle for some common configurations of the mentioned parameters.
-Refer to \Fig{steering_angle_3dB} for the result.
+We computed this steering angle for some common configurations of the mentioned parameters, see the result in \Fig{steering_angle_3dB}.
 The boundary values of the Kaiser window is shown, i.e.
@@
 
-This article is accompanied by some multimedia files and source code for determining the steering angles for LCA's windows. They are located at:
+This article is accompanied by some multimedia files and source code for determining steering boundaries for LCA's windows. The files are located at:
 
@@
 
-\input{bio/jo_inge_buskenes}
-\input{bio/roy_edgar_hansen}
+\input{bio/jo_inge_buskenes_new}
+\input{bio/roy_edgar_hansen_new}
 \input{bio/andreas_austeng}

### 2016-05-01 ThesisJo/introduction.tex
Working on the review
la2pdf commit - 2016:05:01	20:45:42	(sø. 1.mai)

@@
+Faster, smarter, better.
+
+Preface?
+
+Perhaps the most advanced trait of human intelligence is the ability of ask why. We are masters at both understanding what a problem is, as well as finding solutions to it.
+
+What if we could create intelligent machines?
+
+When we face a problem we do not understand, To understand what the problem is
+
+This thesis will eventially narrow down to adaptive processing sonar images, and technicalities of the implementation process.
+
 \chapter{Introduction}
@@
 The image quality of any imaging device is mostly its by its hardware components, and .
+
+There is no best algorithm/system for all tasks. Best would be to use them all as they complement each other.
+
+Hard to determine the most vital information in an image. The better the image the more information is contains, and the more processing power and memory bandwidth is needed to handle it.
 

### 2016-07-28 Simulator/article.tex
la2pdf commit - 2016:07:28	12:02:03	(to.28.juli)

@@
 We have tested the template matching techniqiues on experimental data from the HISAS1030 synthetic aperture sonar (SAS) attached to the HUGIN autonomous underwater vehicle (AUV), both developed by Kongsberg Maritime.
-Three objects from Jesusbukta (Jesus Bay) in Norway will be presented; two roughly 3\,m long cylinders imaged a couple days apart in Janyary 2009, and a roughly 5\m long torpedo imaged in June 2012.
+Three objects from Jesusbukta (Jesus Bay) in Norway will be presented; two roughly 3\,m long cylinders imaged a couple days apart in Janyary 2009, and a roughly 5\,m long torpedo imaged in June 2012.
 

### 2016-08-30 Simulator/article.tex
la2pdf commit - 2016:08:04	20:15:01	(to. 4.aug.)
la2pdf commit - 2016:08:30	10:50:10	(ti.30.aug.)

@@
 \begin{figure*}[!tbp]\centering%
+\graphicsAI[width=\linewidth]{gfx/fig_images_sonar_simulation.pdf}%
+\label{fig_sonar_simulation}%
+\caption{SAS images of 2 cylinders and a torpedo located in the Jesus Bay, Norway. These detections are just 3 objects out of 1000+ detected in this run.}
+\end{figure*}
+
+\begin{figure*}[!tbp]\centering%
+\graphicsAI[width=\linewidth]{gfx/fig_rocs.pdf}%
+\label{fig_sonar_simulation}%
+\caption{Receiver operating characteristics curves.}
+\end{figure*}
+
+\begin{figure*}[!tbp]\centering%
+\subfloat[SAS images of 2 cylinders and a torpedo.]{%
+\graphicsAI[width=0.5\linewidth]{gfx/fig_images_sonar_simulation.pdf}%
+\label{fig_sonar_simulation}}%
+\subfloat[Cylinder simulation with parameters estimated from SAS image: Length 2.6\;m, burial depth 0.263\;m, aspect angle 105$^\circ$.]{
+\graphicsAI[drawing,width=0.5\linewidth]{gfx/sim_cylinder_submerged_adaptive.svg}%
+\label{sim_cylinder}}%
+\caption{A SAS image of a cylinder and a template simulation adapted to it.}
+\end{figure*}
+
+\begin{figure*}[!tbp]\centering%
 \subfloat[SAS image of a cylinder with 2.6\;m length and 0.53\;m radius.]{%

### 2017-04-19 Simulator/article.tex
la2pdf commit - 2017:04:19	21:11:29	(on.19.april)

--- 2016-01-15 Simulator/article.tex
@@
 
-\usepackage{fixltx2e}                           % Fix some twocolumn float problems
+% \usepackage{fixltx2e}                           % Fix some twocolumn float problems
 
@@
     bookmarksnumbered ]{hyperref}
-
-\else
-   \usepackage[\MYhyperrefoptions,breaklinks=true,dvips]{hyperref}
-   \usepackage{breakurl}                        % Allows 'dvips' driver to break links
+% 
+% \else
+%    \usepackage[\MYhyperrefoptions,breaklinks=true,dvips]{hyperref}%
 \fi
+%    \usepackage{breakurl}                        % Allows 'dvips' driver to break links
 
@@
 \newcommand*\B{\mat B}
-\newcommand*\C{\mat C}
+% \newcommand*\C{\mat C}
 \newcommand*\E{\mat E}
@@
 \renewcommand*\L{\mat \Lambda}
-\newcommand*\U{\mat U}
+% \newcommand*\U{\mat U}
 % \renewcommand*\t{\mathtiny{^T}}
@@
 \caption{Simulator principle.
-A model of the seafloor and object is loaded into OpenGL.
-All points in the imaged seafloor region that are visible from the sonar are orthogonally projected onto a view plane.
-}\label{simulator_coordinate_system}%
+A model of the seafloor and object is first loaded into OpenGL, and an orthogonal camera view set up to match the desired image region on the seafloor.
+Then the scene is rendered and an intensity image and a depth map is created.
+The intensity image along with the depth information is finally combined to form a sonar like image.
+This is our template.}\label{simulator_coordinate_system}%
 \end{figure*}
-
-% \begin{figure}[t!]\centering%
-% \graphicsAI[drawing,width=\linewidth]{gfx/specs.svg}%
-% \caption{Simulator specifications.}\label{specs}%
-% \end{figure}
 
@@
 
+For lack of a better name, the simulator will be referred to as "FFISim" throughout this article.
+
+Show:
+\begin{itemize}
+\item Adaptive template matching to avoid large static template libraries,
+\item Improved classification performance by feeding simulator with parameters derived off SAS image,
+\item Further improved classification by performing a fine-search within a narrow search space around the SAS derived parameters. GPU makes this more powerful.
+\end{itemize}
+
 \section{Methods}
@@
 
-Before we can render anything with OpenGL we first have to set up the scene. This involves loading the models and then placing them where we want them with the proper orientation.
+Our models are stored in regular 3D model files.
+The simulator begins by loading these into a tree of nodes, each representing a vertex in the model with corresponding facet and texture information.
+Then the vertices are placed into a virtual scene in OpenGL, and a camera set up to capture the part of the scene corresponding to the image.
+When OpenGL renders the image, it first computes which vertex contributes to which pixel in the image, then computes the intensity and color of the pixel based on the node's facet, texture and scattering model.
+The code that operates on vertices are called a vertex shader, and the code that operates on pixels are called a fragment shader.
 
-Our models are stored in regular 3D model files.
-The simulator loads these with the Open Asset Import library (\texttt{assimp}).
-This is a portable open source library that supports loading a wide range of 3D model formats in a uniform manner.
-It converts the model into a node tree where each node contains data such as vertices, facets and textures, all formatted in an OpenGL friendly way.
-
-The next step is to scale, orient and position the objects in the scene, and finally project this scene onto an image plane.
-This is achieved by applying a set of transformations to each vertex in the model.
-In OpenGL, the set of transformations can be formalized as
+The mapping of the vertices from object coordinates to image coordinates happen through a set of transformations:
+%
 \begin{align*}
-\bmat{x\\y\\0\\1}_\text{image} = \P \cdot \V \cdot \M \cdot \bmat{x\\y\\z\\1}_\text{model},
+\bmat{x\\y\\0\\1}_\text{image} = \P \cdot \V \cdot \M \cdot \bmat{x\\y\\z\\1}_\text{object},
 \end{align*}
+%
 where $\bmat{\cdot}_\text{model}$ is the loaded model, $\bmat{\cdot}_\text{image}$ is the output image, and $\M$, $\V$ and $\P$ are transformation matrices. These will be described next.
@@
 The first transform we apply is the model matrix $\M$, which operates in the model's local coordinate system. It is defined as
+%
 \begin{align}
-\bmat{x_t \\ y_t \\ z_t \\ w}_\text{\parbox{1cm}{\setlength\baselineskip{0.3cm}transformed\\model}} &= \boldsymbol{T} \cdot \R \cdot \boldsymbol{S} \bmat{x \\ y \\ z \\ w}_\text{model},
+\M = \bmat{x_t \\ y_t \\ z_t \\ w}_\text{\parbox{1cm}{\setlength\baselineskip{0.3cm}transformed\\model}} &= \boldsymbol{T} \cdot \R \cdot \boldsymbol{S} \bmat{x \\ y \\ z \\ w}_\text{model},
 \end{align}
+%
 where $\boldsymbol{S}, \R$ and $\boldsymbol{T}$ are matrices\todo{be explicit on dependencies?} for scaling, rotation and translation, respectively.
@@
 Next we map the model into world coordinates using the view matrix $\V$.
-This involves aligning and positioning the model into the view of the camera.
+This involves aligning the camera to precisely fit the image region.
 In our case the camera represents the phase center of the sonar, and it is placed in the world origin.
-The transformation take the form of
+The transformation then take the form of
+%
 \begin{align}
@@
 \end{align}
-where $(x_\text{c}, y_\text{c}, z_\text{c})$ is the image center coordinates for along track, cross track, and depth, respectively, and $\Phi_x$ is the elevation angle\todo{is this correct?} of the sonar.
+%
+where $(x_\text{c}, y_\text{c}, z_\text{c})$ is the image center coordinates for along track, cross track, and depth, respectively, and $\phi_x$ is the elevation angle\todo{is this correct?} of the sonar.
 It is given as
+%
 \begin{align*}
@@
 The matrix that gives orthographic projection is defined as
+%
 \begin{align}
@@
 \end{align}
+%
 where $l,r,b$ and $t$ are the left, right, bottom and top boundaries of the projection plane, respectively, and where $n$ and $f$ is the near and far limits of the cube that encompass the scene.
@@
 \end{align}
-For the near and far limits one simply needs to make sure the scene will fit in it
+%
+For the near and far limits one simply needs to make sure the scene will fit in it, say
+%
 \begin{align}
-n &= 0\;\text{m}   &  f &= 300\;\text{m} 
+n &= 0\;\text{m}   &  f &= 300\;\text{m}.
 \end{align}
-How to set the bottom and top bounaries can be inferred from \Fig{camera_top_bottom}. First we define an origin vector that points to the center of the image
+%
+The bottom and top boundaries can be inferred from \Fig{camera_top_bottom}. First we define an origin vector that points to the center of the image
+%
 \begin{align}
@@
 \end{align}
+%
 and the distance from the image origin to its $y$-boundaries is given as
+%
 \begin{align}
-\Delta y &= \frac{y_\text{max} - y_\text{min}}{2}.
+\Delta y &= \frac{|y_\text{max} - y_\text{min}|}{2}.
 \end{align}
+%
 To find the nearest and farthest point in the image that will be part of the projected scene, $\alpha$ and $\beta$, we must consider the seafloor rotation angle around the $x$-axis, $\Phi_x$ 
+%
 \begin{align}
-\alpha &= \vec 0 - \left[0, \Delta y, \Delta y \times \arctan\Phi_\text{x} \right]
+\vec\alpha &= \vec 0 - \left[0, \Delta y, \Delta y \cdot \tan\Phi_\text{x} \right] \nn
+\vec\beta &= \vec 0 + \left[0, \Delta y, \Delta y \cdot \tan\Phi_\text{x} \right]
 \end{align}
@@
 
-\begin{itemize}
-\item Kernel tuning.
-\item Sharing. PBO. Asynchronous.
-\end{itemize}
+\todo{
+- Kernel tuning.
+- Sharing. PBO. Asynchronous.
+}
+
+\subsection{Template matching}
+
+To narrow down the search space of possible templates we derive some reliable statistics of the seafloor and object from the SAS image.
+While FFISim accepts an arbitrarily complex seabed, we find that modeling it as a plane wave with the right tilt works well.
+The object is more difficult.
+For instance, we estimate its length and height from the shape of its shadow, but this fails if the object is notably immersed into the seabed at one end.
+
+After the a template is created we decompose it into its highlight and shadow component, and compute a correlation score between these and the corresponding areas in the SAS image.
+Then the two scores are summed and normalized to form the final correlation score of the template.
 
@@
 
-\begin{itemize}
-\item Write to file
-\end{itemize}
+\todo{
+Write to file
+}
+
+%    """
+%    
+%    stm: Standard template matching, made with SIGMAS.
+%    atm: Adaptive template matching, made with FFISim.
+%    atm_bury: Adaptive template matching where a search of bury depths is used.
+%    atm_bury_restr: Restricted?
+%    
+%    Adaptiveness made by letting the template matcher use 5 different rotation
+%    angles in the ground-range axis. 
+%    
+%    
+%    Data from Jesusbukta
+%    
+%    Figures are from run 2012.06.16
+%    det210 - Cylinder 1: est 2.92m (short side towards sonar)
+%    det233 - Cylinder 2: est 3.05m (diagonal)
+%    det379 - Torpedo: est 5.06m (looks good)
+%    
+%    ROC curves from 3 different runs, first two just 2 days apart in 2009, last one in 2012.
+%    
+%    run_20120616 - last one. Much better with adaptive templates.
+%    
+%    ROC curves made by average correlation score of shadow/highlight from template matching.
+%    
+%    
+%    fig4b(2015.11.16)
+%       Bug that skipped masking of port side (negative y) detections past 180m range fixed
+%       
+%    fig4 (2015.11.16)
+%       Bug leading to incorrect estimation of cylinder orientation fixed.
+%          This affects run090601_1. Should be better after this.
+%          
+%    fig3 (2015.11.09)
+%       Detections outside 180m range have been masked (but bug for port side fixed in fig4b).
+%       Fixed an incorrectly annotated detection.
+%       
+%    fig2 (2015.11.02)
+%       Height of object estimated from shadow length
+%       Let adaptive method try 5 different rotations in ground-range (-4, -2, 0, 2, 4)
+%          Improved results, but incorrect classification of:
+%          5 detections in first run, and
+%          3 detections in second run
+%          => For FPR=0.2 the results are slightly worse than with normal adaptive (particularly first run).
+%          Images are problematic in these cases:
+%             2 detections get wrong orientation because a line (anchor) cross the image
+%             Other misclassifications in image regions with poor echo or shadow strength
+%          => Not a big issue as not enough time to detect more than ~20% of the detections in a run
+%             In a MCM operation these detections would never be used.
+%             The runs have 3000 detections each, photographing 20% of these is too much.
+%          Performance FPR>0.2 not interesting.
+%          Performance FPR<0.2 much better with adaptive template matching.
+%          
+%    fig (2015.10.23)
+%       atm - cylinder orientation and length estimated from SAS image and fed into simulator.
+%       
+%       ROC curves from last run. Computing the mean correlation score of the shadow and highlight/echo.
+%       Early results. Can only check vs. cylinder-like objects.
+%       Standard template library created based on estimation of physical size of known cylinder mines.
+%       Not possible for adaptive method to perform better when hitting such a target.
+%       
+%  
+%    """
 
@@
 
-\newlength\imgspacing\setlength\imgspacing{.5cm}
-\setcounter{topnumber}{2}
-
-We have tested the template matching techniqiues on experimental data from the HISAS1030 synthetic aperture sonar (SAS) attached to the HUGIN autonomous underwater vehicle (AUV), both developed by Kongsberg Maritime.
-Three objects from Jesusbukta (Jesus Bay) in Norway will be presented; two roughly 3\,m long cylinders imaged a couple days apart in Janyary 2009, and a roughly 5\m long torpedo imaged in June 2012.
-
-The resulting classification performance of the template match ing techniques
-
-Data from Jesusbukta
-
-Cylinder 1: 2.9m (short side towards sonar)
-Cylinder 2: 3.05m (diagonal)
-Torpedo: 5.06m (looks good)
+\begin{figure*}[!tbp]\centering%
+\graphicsAI[width=\linewidth]{gfx/fig_images_sonar_tagged.svg}\label{fig_image_sonar}%
+\caption{SAS images of 2 cylinders and a torpedo located in the Jesus Bay, Norway. Most of the 1000+ detected objects this run are similar to these.}
+\end{figure*}
 
 \begin{figure*}[!tbp]\centering%
-\subfloat[SAS image of a cylinder with 2.6\;m length and 0.53\;m radius.]{%
-\graphicsAI[drawing,width=0.5\linewidth]{gfx/data_cylinder_submerged.svg}%
-\label{data_cylinder}}%
-\subfloat[Cylinder simulation with parameters estimated from SAS image: Length 2.6\;m, burial depth 0.263\;m, aspect angle 105$^\circ$.]{
-\graphicsAI[drawing,width=0.5\linewidth]{gfx/sim_cylinder_submerged_adaptive.svg}%
-\label{sim_cylinder}}%
-\caption{A SAS image of a cylinder and a template simulation adapted to it.}
+\graphicsAI[width=\linewidth]{gfx/fig_images_simulator.pdf}\label{fig_image_simulation}%
+\caption{SAS images of 2 cylinders and a torpedo located in the Jesus Bay, Norway. Most of the 1000+ detected objects this run are similar to these.}
 \end{figure*}
 
+\begin{figure*}[!tbp]\centering%
+\graphicsAI[width=\linewidth]{gfx/fig_rocs.pdf}\label{fig_roc_curves}%
+\caption{Receiver operating characteristics (ROC) curves.}
+\end{figure*}
+
+\begin{figure*}[!tbp]\centering%
+\graphicsAI[width=\linewidth]{gfx/fig_overlay.pdf}%
+% \parbox{\linewidth}{\small\centering\vspace{.5\baselineskip}
+% \newline\centering\newline %\\[.5\baselineskip]
 \newcommand\cdesc[2]{{\raggedright\setlength\fboxsep{0pt}
 \fbox{\colorbox[HTML]{#1}{\vrule height8.5pt depth3.5pt width0pt\hspace{.5cm}}}\ \ #2\\}}
-
-% \newcommand\cdesc[2]{#2}
-
-\begin{figure*}[!tbp]\centering%
-\subfloat[Best template out of a predefined set. Inaccurate cylinder length, alignment and immersion depth. The segmented shadow and highlight regions are misaligned.]{
-\graphicsAI[drawing,width=.48\linewidth]{gfx/overlay_cylinder_submerged.svg}\label{overlay_cylinder}%
-}%
-\hfill%
-\subfloat[Adapted template. Template simulated with adapted parameters. Good fit in both highlight and shadow segments.]{%
-\graphicsAI[width=.48\linewidth]{gfx/overlay_cylinder_submerged_adaptive.svg}%
-}\caption[LoF entry]{\raggedright Simulation image overlaid the segmented SAS image. Segmentation overlay color description:
-
-\parbox{\linewidth}{\small\centering\vspace{.5\baselineskip}
-% \newline\centering\newline %\\[.5\baselineskip]
 \begin{minipage}{.4\linewidth}
@@
 \cdesc{00A7FF}{Template shadow and image highlight}
-\end{minipage}
-}}
-% }\label{overlay_cylinder_adaptive}%
-% % \vspace{\baselineskip}
-% \caption{Segmented simulation image overlaid the segmented SAS image.}
+\end{minipage}%}
+\caption{Comparison of classification performance for the standard and adaptive template technique.
+The three objects were more or less arbitrary picked from the scene.
+For the roughly 3\,m long cylinders both methods perform similarly, the standard method seemingly with a slight edge.
+This is no surprise given that the 3\,m long cylinder model that make up the static template database fit the actual object well here.
+However, for the longer torpedo-like object the adaptive technique is clearly better.
+This is its merit; for objects and geometries that are not included in the static template database we can expect the adaptive techniques to perform better.}\label{fig_image_simulation}%
 \end{figure*}
+
+\newlength\imgspacing\setlength\imgspacing{.5cm}
+\setcounter{topnumber}{2}
+
+The performance of our simulator and template matching techniques where measured using experimental data from an interferometric HISAS1030 synthetic aperture sonar (SAS) attached to a HUGIN autonomous underwater vehicle (AUV), both developed by Kongsberg Maritime, Norway.
+The HISAS1030 is a fully digital phased array with 32 hydrophones, 100\,kHz center frequency and 30\,kHz bandwidth.
+It is 1.2\,m long, has a half-power beamwidth (HPBW) of 23$\circ$, and is capable of generating synthetic images with a theoretical resolution of 3-4\,cm in both dimensions.
+The version of HUGIN used also carries an camera that is used for optical inspection of interesting objects.
+
+We present data from three different HUGIN runs from the Jesus Bay in Norway.
+The first two are taken a couple days apart in January 2009, the last in June 2012.
+There are roughly 3000 detections in each, most representing by cylinder shaped objects such as barrels, pipe mines and torpedoes.
+Therefore, the standard template library were created with SIGMAS+ using a basic  cylinder model with a set of geometrical properties commonly found in the scene\todo{Should be more precise.
+How many templates?
+How finely were the parameters sampled?}.
+For a fair comparison, FFISim was restricted to the same cylinder model as well.
+
+In \Fig{fig_image_sonar} we display SAS images of three typical objects in the 2012 run.
+All are cylindric; the length two first being estimated to roughly 3\,m and the last 5\,m, likely representing cylinder mines and a torpedo, respectively.
+
+Rotation around ground range axis: -4,-2,0,2,4 degrees.
+
+Like the majority of objects in the Jesus Bay they are all cylinders with a length of a few meters.
+The same objects are simulated in \Fig{fig_image_simulation}, one with SIGMAS and the two other with FFISim.
+Each of these simulations were based on
+
+The classification performance is quantified with receiver operating characteristics (ROC) curves from all runs in \Fig{fig_roc_curves}.
+As expected, the 2009 runs proves a greater challenge than that from 2012, with 5 and 3 false positive detections at FPR=20\% in the January 6.
+and 8.
+run, respectively.
+This can partly be explained by incorrect estimation of orientation for a few objects in close proximity to fish trawler tracks, and partly by poor quality in highlight, shadow, or both.
+However, for FPR<20\% the adaptive techniques do much better than the static template library.
+For e.g.
+a mine counter measures (MCM) operation, this is the only region of interest because only a few objects out of the 3000 detections will be visually inspected up close.
+
+- Stm stimulated with length matching test cylinder. Perfect score against this.
+- Atm - not given to improve score. Shadow and highlight both need to match, and overfitting could lead to reduced performance.
+
+Good statistics drawn from the image -> adaptive better. Poor statistics -> hit and miss.
+
+1. Comparison with standard template library (ROC enough?)
+   ROC: Adaptive better at FPR<0.2 in all cases.
+2. Comparison with SIGMAS
+   Fairly similar images, performance is similar. 
+   Used to create the standard template library.
+3. Comparison with search space
+   ROC: Additional search always better FPR<0.2.
+   Explain search criterion.
+
+SAS images - look good. Random pick objects. Lengths derived from image.
+Simulated images - look good. Adaptive simulation more accurate. Otherwise similar. Different background level. Adaptive technique excels for less normal objects/geometries.
+Overlap - Provide quantification measure.
+
+1. Adaptive better? Compare with standard. Doing this in ROC
+2. Improved classification? ROCs
+3. Further improvement by doing fine-search? ROC
+
+An issue the adaptive technique must deal with when estimating object length
+
+As the ROC curves indicate, the adaptive template matching techniques both  ROC curves demonstrate the
+
+The majority of the objects in the Jesus Bay are cylinder shaped. A SAS image of three such objects are displayed in \Fig{fig_image_sonar}
+
+% \begin{figure*}[!tbp]\centering%
+% \subfloat[SAS images of 2 cylinders and a torpedo.]{%
+% \graphicsAI[width=0.5\linewidth]{gfx/fig_images_sonar_simulation.pdf}%
+% \label{fig_sonar_simulation}}%
+% \subfloat[Cylinder simulation with parameters estimated from SAS image: Length 2.6\;m, burial depth 0.263\;m, aspect angle 105$^\circ$.]{
+% \graphicsAI[drawing,width=0.5\linewidth]{gfx/sim_cylinder_submerged_adaptive.svg}%
+% \label{sim_cylinder}}%
+% \caption{A SAS image of a cylinder and a template simulation adapted to it.}
+% \end{figure*}
+% 
+% 
+% \begin{figure*}[!tbp]\centering%
+% \subfloat[SAS image of a cylinder with 2.6\;m length and 0.53\;m radius.]{%
+% \graphicsAI[drawing,width=0.5\linewidth]{gfx/data_cylinder_submerged.svg}%
+% \label{data_cylinder}}%
+% \subfloat[Cylinder simulation with parameters estimated from SAS image: Length 2.6\;m, burial depth 0.263\;m, aspect angle 105$^\circ$.]{
+% \graphicsAI[drawing,width=0.5\linewidth]{gfx/sim_cylinder_submerged_adaptive.svg}%
+% \label{sim_cylinder}}%
+% \caption{A SAS image of a cylinder and a template simulation adapted to it.}
+% \end{figure*}
+
+% \newcommand\cdesc[2]{#2}
 

### 2017-04-19 ThesisJo/introduction.tex
la2pdf commit - 2017:04:19	21:11:29	(on.19.april)

@@
 
+big data. needed to evolve ai intelligence. e.g. deep learning.  need processing speed to handle it. especially near sensors
+
+data now enough? we have all of evolution.
+
+processimg loop small. learn to adapt behaviour then and there.
+
+arrays allow focus and width.but requires processing.
+
+human information 2000 titanics of thumbdrives. ai used to map genes 
 Preface?
@@
 \chapter{Introduction}
+
+One of mankind's greatest dreams is that of creating intelligent machines.
+Imagine things you can strike a conversation with, that aides us in our daily lives, that never grow old, impatient or boring.
+Yesterday science fiction, tomorrow reality.
+It is not a question whether the intelligent machines are coming, but when and how.
+Around the corner are self driving cars, virtual reality, ...
+
+To understand how we can evolve the intelligence of machines, perhaps it is best to look at how nature has done it.
+Our 5 basic senses are sight, smell, hearing, taste and touch.
+The information here is converted from the respective physical phenomena to electrical impulses through various cells.
+Then these electrical signals travel to a vast network of neurons that make up our brain.
+Since each neuron is linked to several others, the data dependency is many to many and very hard to understand.
+
+Intelligence is by its essence a feedback loop. 
+This requires them to sense their surroundings and To do so, we need equip it with sensors we grant them sensory input by converting energy in various physical forms into electrical signals, process these using analog and digital electronics, and imbue them with  Most of the components for doing so
+
+Two big topics in recent years are big data and articifical intelligence.
 

### 2017-04-19 HRUIonGPU/gfx/all_plots.py

@@ -1,279 +1,0 @@
-'''
-Created on Mar 22, 2011
-
-@author: me
-'''
-
-#import framework.calc as calc
-#from framework.calc import db
-#from framework.mynumpy import size, arange, log10, pi, linspace, mean, cos, sin, arcsin, unique, zeros
-from numpy import size, arange, log10, pi, linspace, mean, cos, sin, arcsin, unique, zeros
-#from framework.mynumpy import mode_
-#import framework.mynumpy as np
-import numpy as np
-from scipy import io
-#from framework.
-from pylab import legend,grid, plot, imshow, ion, ioff, figure, savefig, cm, xlim, ylim, clim, xticks, subplot, subplots_adjust, get_current_fig_manager, setp, colorbar, xlabel, ylabel,show, gca
-#import mypylab as pl
-#from mpl_toolkits.axes_grid1 import ImageGrid
-#from framework.gfuncs import error
-
-GFX_PATH='/home/me/Work/Phd/Code/gfx'
-
-
-
-def plotBenchmarkGPU(filename, M, Nx, Ny, fromL, toL):
-
-   print "Reading ", filename
-   
-   timings = np.loadtxt(filename)
-
-   print 'Timings ', timings.shape
-   
-   min_L = fromL
-   max_L = toL
-
-   yavg1 = timings[0::3,:]
-   yavg2 = timings[1::3,:]
-   yavg3 = timings[2::3,:]
-   yavg1_sum = yavg1.sum(1)
-   yavg2_sum = yavg2.sum(1)
-   yavg3_sum = yavg3.sum(1)
-   
-   print 'Yavg1 ', yavg1.shape
-   print 'Yavg2 ', yavg2.shape
-   print 'Yavg3 ', yavg3.shape
-   
-   fig = figure()
-   ax = fig.add_subplot(1,1,1)
-   cax = ax.plot(range(min_L,max_L+1), yavg1[:,0], 'r')
-   cax = ax.plot(range(min_L,max_L+1), yavg1[:,1], 'g')
-   cax = ax.plot(range(min_L,max_L+1), yavg1[:,2], 'b')
-   cax = ax.plot(range(min_L,max_L+1), yavg1_sum, 'k')
-   
-   cax = ax.plot([0,0.001],[0,0], 'k')
[…]

### 2017-04-19 HRUIonGPU/gfx/make_plots.py

@@ -1,22 +1,0 @@
-from all_plots import plotBenchmarkGPU as plot
-from all_plots import barBenchmarkGPU as bar
-from all_plots import barBeamspaceBenchmarkGPU as barBS
-from all_plots import barBeamspaceBenchmarkGPU2 as barBS2
-
-#plot('timer_values_64_80_400.txt', 64, 80, 400, 2, 32)
-#plot('timer_values_32_40_400.txt', 32, 40, 400, 2, 32)
-
-#bar('timer_values_64_80_400.txt', 64, 80, 400, 2, 32)
-#bar('timer_values_32_40_400.txt', 32, 40, 400, 2, 16)
-
-#barBS('timer_values_M=64_Nx=80_Ny=400_L=2-32_Nb=3_slidingBS.txt', 64, 3, 80, 400, 3, 32)
-#barBS('timer_values_M=96_Nx=120_Ny=400_L=3-48_Nb=3_slidingBS.txt', 96, 3, 120, 400, 3, 48)
-
-bar('timer_values_M=64_Nx=80_Ny=540_L=2-32.txt', 64, 80, 540, 2, 32)
-bar('timer_values_M=96_Nx=120_Ny=540_L=2-48.txt', 96, 120, 540, 2, 48, 2)
-
-#barBS('timer_values_M=64_Nx=80_Ny=540_L=16-48_Nb=3.txt', 64, 3, 80, 540, 16, 47, False)
-#barBS('timer_values_M=96_Nx=120_Ny=540_L=24-72_Nb=3.txt', 96, 3, 120, 540, 24, 71)
-
-#barBS2('timer_values_M=64_Nx=80_Ny=606_L=32_Nb=1-5.txt', 64, 32, 80, 606, 1, 5, False)
-#barBS2('timer_values_M=96_Nx=120_Ny=606_L=48_Nb=1-5.txt', 96, 48, 120, 606, 1, 5)

### 2017-05-11 Simulator/article.tex
la2pdf commit - 2017:05:11	12:34:36	(to.11.mai)

@@
 
-\IEEEPARstart{O}{ne} of the greatest visions in science is that of artificial intelligence.
-If we could make machines not only do what we say, but also able to adapt their behaviour to handle the unexpected, the possibilities are perhaps only limited by our imagination.
-An apparent area of use is devices designed to sense and act without aid, such as autonomous underwater vehicles (AUVs).
+% \IEEEPARstart{W}{hen} 
+% Autonomous underwater vehicles (AUVs).
 When an AUV with a sonar imaging system encounter an interesting object, it would be very beneficial if adapted its behavior to examine the object further.
@@
 
+\IEEEPARstart{A}{utonomous} underwater vehicles (AUVs) are becoming increasingly important in human underwater endeavors.
+When equipped with a synthetic aperture sonar (SAS), they can e.g.
+be used to map the seafloor, survey underwater biology, help localize wrecks, inspect oil pipes, fields and electricity lines, or hunt for mines and biohazard waste.
+Such independent systems have much to gain from being able to adapt to the scenario at hand.
+For instance, with a well designed automatic target recognition (ATR) system on board they can adapt their mission plan to collect more information on particularly interesting objects, or fine-tune the sonar and navigation system to improve imaging performance.
+
 % Automatic target recognition (ATR) is an important component in autonomous underwater vehicles (AUVs), as it allows the vehicle to adapt its mission plan to e.g.
@@
 
-The detection process is well described in the field of automatic target recognition (ATR).
-The classical approach involves isolating an object of interest, extracting parameters from it and feeding those into a classifier.
-However, it is hard to find parameters that describe the objects uniquely, in particular due to the ambiguities arising from interpreting a 3D world from its 2D projection.
-An alternative approach is to use template matching, where the image segment is compared to a set of template images covering the relevant object classes, and assigned to the class of the template yielding the the best fit.
-The challenge here is to come up with accurate template images covering the actual configurations of object and seabed, especially if a static template library is used, which will lead to a computational complexity that is exponential with the number of model parameters.
+In conventional ATR an object of interest is first isolated, its properties attempted estimated, and a classifier used to determine what object class these properties belong to.
+The challenge is finding unique and intrinsic object properties, such as its type, shape and size, and to do so independently of extrinsic properties such as its alignment, burial depth, degeneration level or view angle.
+While SAS provide images with state of the art resolution and dynamic range, they still only represent an ambiguous acoustical 2D projection of the 3D object, littered with a highly coherent noise floor.
+
+An alternative approach is to use template matching, where the isolated image segment is compared to a set of template images covering the relevant object classes, and assigned to the class of the template with the best fit.
+Since the templates are computed from 3D models, this reduces the issue of image ambiguity.
+However, a challenge is finding a set of templates that represent the actual configurations of object and seafloor, Another issue is that the size of precomputed library of templates grows exponentially with the number of parameters needed.
 This limits its use to only a few parameters being coarsely sampled, ultimately yielding inaccurate results~\cite{Midelfart2010}.
 
-We have developed a synthetic aperture sonar (SAS) simulator that avoids this problem by running sufficiently fast to create templates adapted to the actual scene as part of the classification process.
+% We avoid the problems of conventional template matching by computing the templates on-the-fly as part of the classification process, thereby not needing a precomputed set of templates at all. 
+%  that avoids this problem by running sufficiently fast to create templates adapted to the actual scene as part of the classification process.
+% Classification process
+
+We tackle the problems of conventional template matching by performing:
+%
+\begin{itemize}
+\item \emph{Adaptive template matching}; where a template is selected based on object and seafloor parameters estimated from the SAS image~\cite{Midelfart2010}, and
+\item \emph{Real-time template simulation}; where a template is created tailored to the estimated parameters, improving its accuracy and avoiding the need of a precomputed template library.
+\end{itemize}
+%
+
+The methods presented herein will focus on the template simulator, but some details will also be given on adaptive template matching for completeness.
+More details on this technique can be found in~\cite{Midelfart2010}.
+
  This is achieved with the aid of the massive computing power in graphics processing units (GPUs) and optimized software libraries for scene rendering.
@@
 
-The alternative of using of a static, predetermined template library has a computational complexity that is exponential with the number of parameters.
-This limits its use to only a few parameters being coarsely sampled, ultimately yielding inaccurate results~\cite{Midelfart2010}.
-
 as part of the classification process. This is achieved with the aid of the massive computing power in graphics processing units (GPUs) and optimized software libraries for scene rendering.
@@
 
-\subsection{Setting up the scene}
+Models are not ROCCAN (???)
+
+\subsection{Loading the scene}
 
@@
 
+1. Segmentation
+2. Parameter estimation
+   Available: Range, altitude, resolution => Input to both standard and adaptive. In standard the best fit out of a predefined template database is selected, then resampled to correct resolution.
+   Additional for adaptive, parameters that modify object model: Target aspect angle, seafloor slope, object length and height.
+
+   Target aspect angle computed with Radon Transform, seafloor under object modeled as plane with a slope estimated from bathymetric data, length estimated from longest part of highlight region, and height estimated from the shape of the shadow.
+The burial depth is n
+
+3.
+
+   Details: \ref{Midelfart2010}.
+
+Let $S(x,y)$ be the isolated part of the
+
+Matching the isolated image segment with the template take the form of
+
+\begin{align}
+c_{h,i} = \underset{x',y'}{\text{argmax}} \sum\limits_{\forall x}\sum\limits_{\forall y} \big| H(x,y) - T(x-x',y-y') \big|
+\end{align}
+
+\begin{align}
+c_i = \frac{c_{h,i} + c_{s,i}}{2}
+\end{align}
+
+And the final template used is the one with the greatest correlation coefficient
+
+\begin{align}
+\underset{T_i(x,y)}{\text{argmax}} C_i\sum\limits_{\forall x}\sum\limits_{\forall y}  S(x,y) T(x-x',y-y')
+\end{align}
+
+Then assign object to class $i$.
+
 To narrow down the search space of possible templates we derive some reliable statistics of the seafloor and object from the SAS image.
@@
 Then the two scores are summed and normalized to form the final correlation score of the template.
+
+We tag the template matchers as either semi- or fully adaptive.
+The former refers to using the best matching SIGMAS+ generated template out of a precomputed set of templates, while the latter refers to the proposed method of using FFISim to generate templates on the-the-fly.
+\todo{Should be more precise.
+How many templates?
+Parameter span?
+How finely were the parameters sampled?}
 
@@
 \begin{figure*}[!tbp]\centering%
-\graphicsAI[width=\linewidth]{gfx/fig_images_sonar_tagged.svg}\label{fig_image_sonar}%
-\caption{SAS images of 2 cylinders and a torpedo located in the Jesus Bay, Norway. Most of the 1000+ detected objects this run are similar to these.}
-\end{figure*}
-
-\begin{figure*}[!tbp]\centering%
-\graphicsAI[width=\linewidth]{gfx/fig_images_simulator.pdf}\label{fig_image_simulation}%
-\caption{SAS images of 2 cylinders and a torpedo located in the Jesus Bay, Norway. Most of the 1000+ detected objects this run are similar to these.}
+\graphicsAI[width=\linewidth]{gfx/fig_images_sonar_simulator_tagged.svg}\label{fig_images_sonar_simulation}%
+\caption{Selected objects.
+Top row show SAS images of 2 cylinders and a torpedo located in the Jesus Bay, Norway.
+Center row show the best match from a conventional template matcher, and bottom row show the adaptive template matcher.
+The contour of the SAS object is indicated with red lines.
+Note that the adaptive technique produce templates with a better fit to the highlight, but that this does not necessarily lead to a better match against the shadow.}
 \end{figure*}
@@
 We present data from three different HUGIN runs from the Jesus Bay in Norway.
-The first two are taken a couple days apart in January 2009, the last in June 2012.
-There are roughly 3000 detections in each, most representing by cylinder shaped objects such as barrels, pipe mines and torpedoes.
-Therefore, the standard template library were created with SIGMAS+ using a basic  cylinder model with a set of geometrical properties commonly found in the scene\todo{Should be more precise.
-How many templates?
-How finely were the parameters sampled?}.
-For a fair comparison, FFISim was restricted to the same cylinder model as well.
-
-In \Fig{fig_image_sonar} we display SAS images of three typical objects in the 2012 run.
-All are cylindric; the length two first being estimated to roughly 3\,m and the last 5\,m, likely representing cylinder mines and a torpedo, respectively.
+The first two are taken a couple days apart in January 2009, containing roughly 2700 and 3200 detections, respectively.
+The last is from June 2012 and contains roughly 650 detections.
+The detections represent mostly cylinder shaped objects such as barrels, pipe mines and torpedoes.
+Therefore, all templates were based on a basic cylinder model.
+
+The two template matchers can be visually inspected in \Fig{fig_images_sonar_simulator_tagged}.
+
+The classification performance is quantified with receiver operating characteristics (ROC) curves from all runs in \Fig{fig_roc_curves}.
+The 2009 runs proves a greater challenge than that from 2012, with 5 and 3 false positive detections at FPR=20\% in the January 6.
+and 8.
+run, respectively.
+This came as no surprise due to the 2009 runs containing fish trawler tracks and having poorer highlight and shadow quality than the run from 2012.
+However, for FPR<20\% the adaptive techniques outperform the static template library.
+For e.g.
+a mine counter measures (MCM) operation, this is the only region of interest because only a few objects out of the 3000 detections will be visually inspected up close.
+
+- Stm stimulated with length matching test cylinder. Perfect score against this.
+- Atm - not given to improve score. Shadow and highlight both need to match, and overfitting could lead to reduced performance.
+
+In \Fig{fig_images_sonar_simulator_tagged} we compare the SAS image to the templates generated by the conventional and adaptive template matcher.
+Three typical objects from the 2012 were chosen, all being cylindric.
+The length two first being estimated to roughly 3\,m and the last 5\,m, likely representing cylinder mines and a torpedo, respectively.
+
+Selected objects.
+Top row show SAS images of 2 cylinders and a torpedo located in the Jesus Bay, Norway.
+Center row show the best match from a conventional template matcher, and bottom row show the adaptive template matcher.
+The contour of the SAS object is indicated with red lines.
+Note that the adaptive technique produce templates with a better fit to the highlight, but that this does not necessarily lead to a better match against the shadow.
 
@@
 Each of these simulations were based on
-
-The classification performance is quantified with receiver operating characteristics (ROC) curves from all runs in \Fig{fig_roc_curves}.
-As expected, the 2009 runs proves a greater challenge than that from 2012, with 5 and 3 false positive detections at FPR=20\% in the January 6.
-and 8.
-run, respectively.
-This can partly be explained by incorrect estimation of orientation for a few objects in close proximity to fish trawler tracks, and partly by poor quality in highlight, shadow, or both.
-However, for FPR<20\% the adaptive techniques do much better than the static template library.
-For e.g.
-a mine counter measures (MCM) operation, this is the only region of interest because only a few objects out of the 3000 detections will be visually inspected up close.
-
-- Stm stimulated with length matching test cylinder. Perfect score against this.
-- Atm - not given to improve score. Shadow and highlight both need to match, and overfitting could lead to reduced performance.
 

### 2017-05-11 ThesisJo/introduction.tex
la2pdf commit - 2017:05:11	12:34:36	(to.11.mai)

@@
 \chapter{Introduction}
+
+\IEEEPARstart{O}{ne} of the greatest visions in science is that of artificial intelligence.
+If we could make machines not only do what we say, but also able to adapt their behaviour to handle the unexpected, the possibilities are perhaps only limited by our imagination.
 
@@
 \section{Thesis outline}
+
+JIB
+Smarter, better, faster - On the future of machine intelligence in active sonar systems
+
+Preface
+  Way too complex a topic to be covered in fully detail. Perhaps impossible?
+  Interesting enough to be covered, even if just partially.
+  Complexity issue - picture from essay
+
+Introduction
+   A marvel of science, the world, and beyond
+      Intelligence / conscience. What is it?
+      HPC bridging learning and big data
+   Machine intelligence / adaptivity / Deep learning
+      Means for intelligence?
+      Adapts to data
+      Proven powerful
+      Hard to understand, hard to control
+      Data hungry
+      Fascinating perspectives:
+      - Local dependencies work on e.g. images. Due to lack of data?
+      - First layer similar to template matching (convolitional network)
+      - If we can't understand it, perhaps we can parent it? Society's norms a giant assemble of prior knowledge.
+      - Deep neural nets interconnected, learning from each other?
+      - The brain adapts its wiring to solve tasks, does this carry over?
+      - Neuman networks - recursive
+   Big data
+      Fundament of intelligence?
+      Amount of data recorded and stored immense. Sensors everywhere.
+      Phones, internet searches, smart watches, cars, surveillance cams,   
+   High performance computing
+      Engine of intelligence?
+      Combining deep learning with big data
+      Improving at a rapid pace
+
+Background
+   Acoustic waves
+      Wave equation
+   Image reconstruction
+      Conventional beamforming
+      Adaptive beamforming
+         LCA
+         MVDR
+   Image modeling
+      Images based on prior knowledge
+      Can be used as input to deep learning
+
+Summary of publications
+   - Red line: HPC, adaptivity,
+   Journal:
+      2017 - JOE - Simulator
+
+      2017 - JOE - Low-Complexity Adaptive Sonar Imaging
+      2015 - JOE - An optimized GPU implementation of the MVDR beamformer for active sonar imaging
+      2014 - TUFFC - Implementing capon beamforming on a GPU for real-time cardiac ultrasound imaging
+
+   Related work:
+      2014 - UAM2014 - A GPU sonar simulator for automatic target recognition
+      2013 - ICA2013 - Adapting the minimum variance beamformer to a graphics processing unit for active sonar imaging systems
+      2012 - ECUA2012 - GPU-Based Adaptive Beamforming for Active Sonar Imaging
+      2012 - IEEE-US - Implementing Capon Beamforming on the GPU for Real Time Cardiac Ultrasound Imaging
+      2011 - UAM2011 - A low complexity adaptive beamformer for active sonar imaging
+
+Discussion and future work
+   Summary and discussion
+      Simulator:
+      - The way to train and constrain deep learning algorithms?
+      MVDR:
+      - Analytical optimization - full adaptivity.
+      - Heavy on computations
+      LCA:
+      - Optimization by trial and error - discrete solution space.
+      - Works almost as well as MVDR in most cases
+      - Fast!
+      - Well suited for hardware
+
+      Lessions from deep learning:
+      -
+
+   Future work
 

### 2018-01-04 Simulator/article.tex
la2pdf commit - 2018:01:04	11:27:54	(to. 4.jan.)

@@
 
-\newcommand\graphicsAI[2][]{%
+\let\oldincludegraphics\includegraphics
+\renewcommand\includegraphics[2][]{%
   \immediate\write18{./bin/laFigure #2 #1}%
@@
 \begin{figure*}[t]\centering%
-\graphicsAI[width=\linewidth]{gfx/scene_coordinate_system.svg}%
+\includegraphics[width=\linewidth]{gfx/scene_coordinate_system.svg}%
 \caption{Simulator principle.
@@
 \begin{figure*}[t]\centering%
-\graphicsAI[drawing,width=\linewidth]{gfx/simulator.svg}%
+\includegraphics[drawing,width=\linewidth]{gfx/simulator.svg}%
 \caption{Simulator concept.}\label{buildup}%
@@
 \begin{align*}
-\bmat{x\\y\\0\\1}_\text{image} = \P \cdot \V \cdot \M \cdot \bmat{x\\y\\z\\1}_\text{object},
+\bmat{x\\y\\0\\1}_\text{image} = \boldsymbol{P} \cdot \boldsymbol{V} \cdot \boldsymbol{M} \cdot \bmat{x\\y\\z\\1}_\text{object},
 \end{align*}
@@
 \begin{align}
-\M = \bmat{x_t \\ y_t \\ z_t \\ w}_\text{\parbox{1cm}{\setlength\baselineskip{0.3cm}transformed\\model}} &= \boldsymbol{T} \cdot \R \cdot \boldsymbol{S} \bmat{x \\ y \\ z \\ w}_\text{model},
-\end{align}
-%
-where $\boldsymbol{S}, \R$ and $\boldsymbol{T}$ are matrices\todo{be explicit on dependencies?} for scaling, rotation and translation, respectively.
+\M = \bmat{x_t \\ y_t \\ z_t \\ w}_\text{\parbox{1cm}{\setlength\baselineskip{0.3cm}transformed\\model}} &= \boldsymbol{T} \cdot \boldsymbol{R} \cdot \boldsymbol{S} \bmat{x \\ y \\ z \\ w}_\text{model},
+\end{align}
+%
+where $\boldsymbol{S}$, $\boldsymbol{R}$ and $\boldsymbol{T}$ are matrices\todo{be explicit on dependencies?} for scaling, rotation and translation, respectively.
 These are provided for reference in Appendix \ref{transformation_matrices}.
@@
 \begin{figure}[t]\centering%
-\graphicsAI[drawing,width=\linewidth]{gfx/opengl_part.svg}%
+\includegraphics[drawing,width=\linewidth]{gfx/opengl_part.svg}%
 \caption{Setting camera top and bottom boundaries.}\label{camera_top_bottom}%
@@
 \begin{figure*}[!tbp]\centering%
-\graphicsAI[width=\linewidth]{gfx/fig_images_sonar_simulator_tagged.svg}\label{fig_images_sonar_simulation}%
+\includegraphics[width=\linewidth]{gfx/fig_images_sonar_simulator_tagged.svg}\label{fig_images_sonar_simulation}%
 \caption{Selected objects.
@@
 \begin{figure*}[!tbp]\centering%
-\graphicsAI[width=\linewidth]{gfx/fig_rocs.pdf}\label{fig_roc_curves}%
+\includegraphics[width=\linewidth]{gfx/fig_rocs.pdf}\label{fig_roc_curves}%
 \caption{Receiver operating characteristics (ROC) curves.}
@@
 \begin{figure*}[!tbp]\centering%
-\graphicsAI[width=\linewidth]{gfx/fig_overlay.pdf}%
+\includegraphics[width=\linewidth]{gfx/fig_overlay.pdf}%
 % \parbox{\linewidth}{\small\centering\vspace{.5\baselineskip}
@@
 % \subfloat[SAS images of 2 cylinders and a torpedo.]{%
-% \graphicsAI[width=0.5\linewidth]{gfx/fig_images_sonar_simulation.pdf}%
+% \includegraphics[width=0.5\linewidth]{gfx/fig_images_sonar_simulation.pdf}%
 % \label{fig_sonar_simulation}}%
 % \subfloat[Cylinder simulation with parameters estimated from SAS image: Length 2.6\;m, burial depth 0.263\;m, aspect angle 105$^\circ$.]{
-% \graphicsAI[drawing,width=0.5\linewidth]{gfx/sim_cylinder_submerged_adaptive.svg}%
+% \includegraphics[drawing,width=0.5\linewidth]{gfx/sim_cylinder_submerged_adaptive.svg}%
 % \label{sim_cylinder}}%
@@
 % \subfloat[SAS image of a cylinder with 2.6\;m length and 0.53\;m radius.]{%
-% \graphicsAI[drawing,width=0.5\linewidth]{gfx/data_cylinder_submerged.svg}%
+% \includegraphics[drawing,width=0.5\linewidth]{gfx/data_cylinder_submerged.svg}%
 % \label{data_cylinder}}%
 % \subfloat[Cylinder simulation with parameters estimated from SAS image: Length 2.6\;m, burial depth 0.263\;m, aspect angle 105$^\circ$.]{
-% \graphicsAI[drawing,width=0.5\linewidth]{gfx/sim_cylinder_submerged_adaptive.svg}%
+% \includegraphics[drawing,width=0.5\linewidth]{gfx/sim_cylinder_submerged_adaptive.svg}%
 % \label{sim_cylinder}}%

### 2018-01-04 ThesisJo/introduction.tex
la2pdf commit - 2018:01:04	11:27:54	(to. 4.jan.)

@@
   Interesting enough to be covered, even if just partially.
-  Complexity issue - picture from essay
+  Complexity issue - picture from essay.
+     - We observe the world from one location. The internet lets machine see everything.
+     - Mental constraint. Can machines break through it?
 
@@
       HPC bridging learning and big data
+      Deep Patient, Google Brain, Nvidia self driving cars, Bloomberg money machine, image and speech analysis, ...
+   Information processing
+      Instrinsic and extrinsic information
+      Data decomposition
+      Data volume
+      Amount of information, rank, dimensions
+
    Machine intelligence / adaptivity / Deep learning
@@
       - Neuman networks - recursive
+      - Energy efficient. But the body does it MUCH better
+      - Only make the choices that are relevant. Adaptivity vs. computational/data requirements
+        - The body has a lot of information stored in the genome. It doesn't start from scratch
+      - On images - convolutional networks
+        - Layers offer image feature representation of different scales (intuition: create extrinsic shift invariance)
    Big data
@@
       Improving at a rapid pace
+      Quantum computers seem very adapt at deep learning, with node-like structure
+      FPGA might fill a gap too.
+Hardware generated adaptively.
+Efficient information exchange often more effective than processing power.
+Current architectures limited by bandwidth for most applications.
 

### 2018-01-04 Simulator/article.tex
la2pdf commit - 2018:01:04	12:28:36	(to. 4.jan.)

@@
 \section{Transformation matrices}\label{transformation_matrices} 
+%
 We place, resize and orient objects in an euclidean 3D space using suitable transformation matrices that are applied to each of the object's vertices:

### 2018-02-06 ASA2013/presentation/presentation.tex
Major commit before messing with PhD thesis

\documentclass[
    beamer                                       % Document type (non-standard)
%  , handout
%   , xelatex                                      % Use the XeLaTeX compiler
%  , movie
  ,table,dvipsnames,svgnames
]{common/mytemplate}

\pdfpageattr {/Group << /S /Transparency /I true /CS /DeviceRGB>>}

% \mode<handout>{\setbeamercolor{background canvas}{bg=black!5}}

% \RequirePackage[table,dvipsnames,svgnames]{xcolor}
\definecolor{tabBlue}{HTML}{AACCFF}
\usepackage{template/beamerthemeUiO}
% \usepackage{common/oxygentheme}
% \setbeameroption{show only notes}
% \setbeameroption{notes on second screen=left}

% \newcommand\remotesource{http://www.joinge.net/compet}

\title{\newline\newline Adapting the MVDR beamformer to a GPU for differently sized active sonar imaging systems}
\subtitle{}
\author[]{\textbf{Jo Inge Buskenes}$^{\text{a}}$, Jon Petter Åsen$^{\text{b}}$, Carl-Inge~Colombo~Nilsen$^{\text{a}}$, Andreas~Austeng$^{\text{a}}$}
\date[Jo Inge Buskenes at ASA/ICA, Montreal, June\ 2013]{Montreal, Canada, 2013}
\institute[Dept.\ of Informatics, University of Oslo]{\bf$^{\text{a}}$Department of Informatics, University of Oslo, Norway\\
                                                      \bf$^{\text{b}}$MI-Lab, Norwegian University of Science and Technology}
% \date{Sist revidert: 27.01.2009}

\definecolor{LightBlue}{rgb}{0.3,0.3,1}
\definecolor{DarkBlue}{rgb}{0.2,0.2,0.7}
\setbeamercolor{block title}{bg=DarkBlue,fg=white}%bg=background, fg= foreground
\setbeamercolor{block body}{bg=gray!20,fg=black}%bg=background, fg= foreground
\setbeamertemplate{blocks}[rounded]

\RequirePackage[latin1]{inputenc}%              % Set input encoding (optionally latin1) utf8
\RequirePackage[T1]{fontenc}%                 % Set font encoding

\begin{document}

\begin{frame}
\vspace*{2\baselineskip}
  \titlepage
\end{frame}
\note{\null{}}

% \AtBeginSection[]
% {
%   \frame<handout:0>
%   {
%     \frametitle{Agenda}
%     \tableofcontents %[currentsection,hideallsubsections]
%   }
% }

% \AtBeginSubsection[]
% {
%   \frame<handout:0>
%   {
%     \frametitle{Agenda}
%     \tableofcontents[sectionstyle=show/hide,subsectionstyle=show/shaded/hide]
%   }
% }

\newcommand<>{\highlighton}[1]{%
  \alt#2{\structure{#1}}{{#1}}
}

\newcommand{\icon}[1]{\pgfimage[height=1em]{#1}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%% Content starts here %%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% {

% \section{Case study}
% \subsection{Realtime sectorcan imaging}
% 
% {
% \setbeamertemplate{background}{\graphicsAI[width=\paperwidth]{gfx/use_case_sectorscan.svg}}%
% \begin{frame}
% \vspace{-80pt}\hspace{-25pt}
% \frametitle{\color{white}Realtime requirements?}
% \framesubtitle{\vspace{-10pt}\color{white}Sectorscan imaging}
% \vspace{100pt}\ 
% % \begin{figure}[H]
% % \begin{narrow}{-1.6cm}{-1.6cm}
% % \graphicsAI<1>[drawing,width=\linewidth]{gfx/use_case_sectorscan3.svg}
% % \end{narrow}
% % \end{figure}   
% \end{frame}
% }

% \renewcommand{\frametitle}[2]{{\vspace*{10pt}\bf\Large #1\par}}
% \begin{frame}
% \frametitle{Context}
% \framesubtitle{Active Sonar}
% \vspace{-25pt}
% \begin{figure}[H]
% \begin{narrow}{-0.5cm}{0pt}
% \hspace{-10pt}\graphicsAI<1>[drawing,width=1.05\linewidth]{gfx/SonarPrinciple.svg}
% \end{narrow}
% \end{figure}
% % \begin{itemize}
% % \item Each pixel value is estimated by focusing the receiver on that point.
% % \end{itemize}
% \end{frame}

% \begin{frame}
% \frametitle{Context}
% \framesubtitle{Imaging techniques}
% \vspace{-10pt}
% \begin{figure}[H]
% \begin{narrow}{-0.5cm}{0pt}
% \hspace{-10pt}\graphicsAI<1>[drawing,width=1.05\linewidth]{gfx/imaging_concepts.png}
% \end{narrow}
% \end{figure}
% A phased array can be used in several imaging modes:
% \begin{itemize}
% \item \emph{Sector scan}: Image computed from a single ping
% \item \emph{Sidescan}: Image created by stacking images from several subsequent pings
% \item \emph{Synthetic aperture sonar}: Image computed from overlapping pings
% \end{itemize}
% % \begin{block}{Proposed method: Generic, can be used in all modes}
% % \vspace{-25pt}
% % \end{block}
% \end{frame}
% }

% \section{}
% \begin{frame}
%   \frametitle{Agenda}
%   \tableofcontents%[%section=1,
%                    %hidesubsections]
% \end{frame}

\renewcommand{\frametitle}[2]{{\vspace*{10pt}\bf\Large #1\par}}

% \begin{frame}
% \frametitle{Acknowledgements}
% \framesubtitle{}
% \vspace{-3pt}
% \begin{itemize}
% \item Kongsberg Maritime.
% \item FFI guys.
% \end{itemize}
% \end{frame}

% General system, may be applied to several different datasets

% \section{Introduction}
% 
% 
% \begin{frame}
% \frametitle{What we do}
% \framesubtitle{}
% \vspace{-10pt}
% \begin{figure}[H]
% \begin{narrow}{-0.5cm}{0pt}
% \hspace{-10pt}\graphicsAI<1>[drawing,width=1.05\linewidth]{gfx/applications.svg}
% \end{narrow}
% \end{figure}
% % A phased array can be used in several imaging modes:
% % \begin{itemize}
% % \item \emph{Sector scan}: Image computed from a single ping
% % \item \emph{Sidescan}: Image created by stacking images from several subsequent pings
% % \item \emph{Synthetic aperture sonar}: Image computed from overlapping pings
% % \end{itemize}
% % % \begin{block}{Proposed method: Generic, can be used in all modes}
% % % \vspace{-25pt}
% % % \end{block}
% \end{frame}
%

\begin{frame}
\frametitle{Image Reconstruction}
\framesubtitle{Medical Ultrasound Imaging}
% \vspace{-25pt}
\begin{figure}[H]
\begin{narrow}{-0.5cm}{0pt}
\graphicsAI<1>[drawing,width=1.05\linewidth]{gfx/applications.svg}
\end{narrow}
\end{figure}
\vspace{-10pt}
\begin{itemize}
\item Videos and 3D processing $\rightarrow$ High computational requirements.
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{\hspace{-.5cm}Image Reconstruction}
\framesubtitle{\hspace{-.5cm}Synthetic\\\hspace{-.5cm}Aperture\\\hspace{-.5cm}Sonar}
\framesubtitle{Active Sonar}
\vspace{-45pt}
\begin{figure}[H]
\begin{narrow}{.1cm}{-0.6cm}
\flushright\hspace{-10pt}\graphicsAI<1>[drawing,width=.92\linewidth]{gfx/rock_formation.svg}
\end{narrow}
\end{figure}
\vspace{-10pt}
\begin{narrow}{-0.5cm}{-0.5cm}
\begin{itemize}\small
\item SAS image of a 25 x 25m wide by 5m high rock formation at 70m depth.
\item Theoretical resolution of 3-4cm throughout the image.
\item Data collected by HUGIN AUV carrying the HISAS interferometric SAS. Courtesy of the Norwegian Defence Research Establishment.
\end{itemize}
\end{narrow}
\end{frame}

\begin{frame}
\frametitle{Beamforming}
\framesubtitle{The Delay and Sum beamformer (conventional method)}
\begin{figure}[H]
\begin{narrow}{-0.9cm}{-0.9cm}
\graphicsAI<1>[drawing,width=\linewidth]{gfx/beamforming_das.svg} %{../Docs/GeiloSrc/Figs/das.pdf}
\end{narrow}
\end{figure}                                                                                                                                                                                                                                                                                                   
\end{frame}

{
\setbeamertemplate{background}{\graphicsAI[width=\paperwidth]{gfx/scenario_interference.svg}}%
\begin{frame}
\vspace{-151pt}
\frametitle{\hspace{-.5cm}Simulated Responses}
\framesubtitle{\hspace{-.5cm}What determines a\\\hspace{-.5cm}beamformer's ability to\\\hspace{-.5cm}suppress interference?}
\end{frame}
}

{
\setbeamertemplate{background}{\graphicsAI[width=\paperwidth]{gfx/scenario_interference_das.svg}}%
\begin{frame}
\vspace{-160pt}
\frametitle{\hspace{-.5cm}Simulated Responses}
\framesubtitle{\hspace{-.5cm}The Delay-and-Sum\\\hspace{-.5cm}(DAS) beamformer}
\end{frame}
}

{
\setbeamertemplate{background}{\graphicsAI[width=\paperwidth]{gfx/scenario_interference_mvdr.svg}}%
\begin{frame}
\vspace{-142pt}
\frametitle{\hspace{-.5cm}Simulated Responses}
\framesubtitle{\hspace{-.5cm}The Minimum Variance\\\hspace{-.5cm}Distortionless Response\\\hspace{-.5cm}(MVDR)\\\hspace{-.5cm}beamformer}
\end{frame}
}

% \begin{frame}
% \frametitle{Beamforming}
% \framesubtitle{Minimum Variance Distortionless Response (MVDR) beamformer}
% \begin{figure}[H]
% \begin{narrow}{-0.9cm}{-0.9cm}
% \graphicsAI<1>[drawing,width=\linewidth]{gfx/beamforming_mv.svg} %{../Docs/GeiloSrc/Figs/das.pdf}
% \end{narrow}
% \end{figure}                                                                                                                                                                                                                                                                                                   
% \end{frame}

\begin{frame}
\frametitle{MVDR beamformer}
\framesubtitle{Computational steps and complexity}\\
% \vspace{-15pt}
\begin{figure}[H]
\begin{narrow}{-.5cm}{-.5cm}
\graphicsAI<1>[drawing,width=\linewidth]{gfx/beamforming_mv_light.svg} %{../Docs/GeiloSrc/Figs/das.pdf}
\end{narrow}
\end{figure}
\vspace{-5pt}\textbf{Complexity increases fast with system size, but:}
\begin{itemize}
\item Each pixel in the image can be processed independently.
\item Use Graphics Processing Units (GPUs)?
\end{itemize}
\end{frame}
% 
% 
% % \begin{frame}
% % \frametitle{MVDR}
% % \framesubtitle{Estimating $\R$}
% % \vspace{15pt}
% % % An estimate $\eR$ of $\R$ can be computed as
% % % \begin{align*}
% % % \R = E\{\x[n]\x\H[n]\} = %\frac{1}{(2N_{\text{avg}}}\sumb{n'=n-Navg}{Navg}\sumb{k=0}{M-L+1} \x_k[n]\x_k\H[n]
% % % \end{align*}
% % \begin{itemize}
% % \item Spatial (subarray) averaging to improve estimate
% % \item Temporal averaging to improve speckle statistics
% % \item Diagonal loading improve numerical conditioning
% % \end{itemize}
% % \end{frame}
% 
% \subsection{Graphics processing Units (GPUs)}
% 
\begin{frame}
\makeatletter%
% \special{pdf: put @thispage <</Group << /S /Transparency /I true /CS /DeviceRGB>> >>}%
\makeatother%
\frametitle{GPUs for image reconstruction}
\framesubtitle{}
\vspace{-3pt}
\begin{figure}[H]
\begin{narrow}{0cm}{0cm}
\graphicsAI<1>[drawing,width=\linewidth]{gfx/gpus.svg}%{../Docs/GeiloSrc/Figs/das.pdf}
\end{narrow}
\end{figure}
\begin{itemize}
\item CPUs: Optimized for running \emph{a few} potentially complex threads, that may have \emph{complex} data dependencies.
\item GPUs: Optimized for running \emph{hundreds} of light-weight threads, with \emph{simple} data dependencies.
\end{itemize}
\end{frame}

% 
% \section{Implementation on a GPU}
% 
\begin{frame}
\frametitle{Implementation}
\framesubtitle{MVDR on GPU}
\vspace{-15pt}
\begin{figure}[H]
\begin{narrow}{0cm}{0cm}
\graphicsAI<1>[drawing,width=\linewidth]{gfx/implementation.svg}%{../Docs/GeiloSrc/Figs/das.pdf}
\end{narrow}
\end{figure}
\end{frame}

\begin{frame}
\frametitle{Implementation}
\framesubtitle{Beamspace MVDR on GPU (frequency domain version)}
\vspace{-13pt}
\begin{figure}[H]
\begin{narrow}{0cm}{0cm}
\graphicsAI<1>[drawing,width=\linewidth]{gfx/implementation_beamspace.svg}%{../Docs/GeiloSrc/Figs/das.pdf}
\end{narrow}
\end{figure}
\end{frame}

\begin{frame}
\frametitle{Bandwidth limited design!!!}
\framesubtitle{}
\begin{center}
\begin{minipage}{4cm}
\begin{block}{}
\centering $A = B\;c + D$
\end{block}
\end{minipage}
\end{center}
\begin{itemize}
\item 4 floats over I/O (3 reads, 1 write) 
\item 2 floating point operations.
\end{itemize}
\begin{table}[H]\centering%\normalsize
\begin{tabular}[c]{l r r r@{}  l}\hline
\rowcolor{tabBlue} &
\multicolumn{1}{>{\columncolor{tabBlue}}c}{\bf B$_\text{arith}$} & \multicolumn{1}{>{\columncolor{tabBlue}}c}{\bf B$_\text{mem}$} & \multicolumn{2}{>{\columncolor{tabBlue}}c}{\bf B$_\text{mem}$/B$_\text{arith}$} \\\hline
Arithmetic & 1.03 Tflop/s & & &\\
Global memory & & 36 Gfloats/s & \hspace{30pt} 1 &:30 \\
Shared memory & & 257 Gfloats/s & 1 &:4 \\
Registers & & $>$1.5 Tfloats/s & $>$3 &:2
\end{tabular}
\caption{Nvidia Quadro 6000: Memory throughput, $B_{\lowercase{\text{mem}}}$, compared to arithmetic throughput, $B_\text{arith}$.}
\end{table}
\end{frame}

% \begin{frame}
% \frametitle{Experimental Setup}
% \framesubtitle{HISAS 1030}
% \begin{figure}[H]
% \graphicsAI<1>[drawing,width=0.6\linewidth]{gfx/hisas_on_hugin.jpg} %{../Docs/GeiloSrc/Figs/das.pdf}
% \end{figure}
% \begin{itemize}
% \item High resolution interferometric SAS
% \item 2x32 element phased array transmitter/receiver
% \item Array length: 120cm
% \item Sampling frequency: 100kHz
% \item Opening angle: 15deg (TX), 23deg (RX)
% \end{itemize}
% \end{frame}

% \section{Results}
% 
% \subsection{Image Quality}

\begin{frame}
\frametitle{Results (benchmarks)}
\framesubtitle{}
\vspace{-3pt}
\begin{itemize}
\item Test system: Intel Xeon quad core CPU, and a Nvidia Quadro 6000 GPU (equivalent to Geforce GTX480).
\end{itemize}
\begin{narrow}{-.5cm}{-.5cm}
\begin{figure}[H]\centering
\graphicsAI<1>[drawing,width=1\linewidth]{gfx/benchmark1_tagged.svg}
\end{figure}
\end{narrow}
\vspace{-10pt}
\begin{itemize}
\item GPU version $\approx$ 2 orders of magnitude faster than (plain) C.
\end{itemize}
\end{frame}

\section{Case study}
% \subsection{Realtime sectorcan imaging}

{
\setbeamertemplate{background}{\graphicsAI[width=\paperwidth]{gfx/use_case_sectorscan3.svg}}%
\begin{frame}
\vspace{-150pt}\hspace{-25pt}
\frametitle{\color{white}Realtime requirements?}
\framesubtitle{\vspace{-10pt}\color{white}Sectorscan imaging}
% \vspace{100pt}\ 
% \begin{figure}[H]
% \begin{narrow}{-1.6cm}{-1.6cm}
% \graphicsAI<1>[drawing,width=\linewidth]{gfx/use_case_sectorscan3.svg}
% \end{narrow}
% \end{figure}   
\end{frame}
}

{
\setbeamertemplate{background}{\graphicsAI[width=\paperwidth]{gfx/scenario_interference_lca.svg}}%
\begin{frame}
\vspace{-142pt}
\frametitle{\hspace{-.5cm}Alternative for very large systems}
\framesubtitle{\hspace{-.5cm}The Low Complexity\\\hspace{-.5cm}Adaptive\\\hspace{-.5cm}(LCA)\\\hspace{-.5cm}beamformer}
\end{frame}
}

\begin{frame}
\frametitle{Image Quality}
\framesubtitle{\scriptsize Shipwreck Holmengraa imaged with HISAS1030 from Kongsberg Maritime}
\vspace{-5pt}
\begin{figure}[H]
\begin{narrow}{-2cm}{-2cm}\centering
\graphicsAI<1>[width=0.85\linewidth]{gfx/plot_holmengraa_L16_Navg1.pdf}
\end{narrow}
\end{figure}
\vspace{-10pt}
\begin{itemize}
\item All adaptive beamformers improve edge definition and suppress noise compared to DAS.
\end{itemize}
\end{frame}

% \subsection{Benchmarks}

{
\setbeamertemplate{background}{\graphicsAI[width=\paperwidth]{gfx/AUVenv.png}}%
\begin{frame}
\vspace{-25pt}\frametitle{\color{white}Conclusion}
\framesubtitle{}
\vspace{-5pt}\color{white}
\begin{itemize}
\item \color{white}GPUs are well suited to accelerate algorithms such as the MVDR beamformer
\item \color{white}Our GPU implementation of the MVDR beamformer
\begin{itemize}
\item \color{white}fully supports active sonar systems,
\item \color{white}easily outperforms CPU implementations, and
\item \color{white}can run in beamspace mode to maintain imaging speeds for large systems.
\end{itemize}
\item \color{white}LCA: A lightweight and easily implementable alternative for very large systems.
\end{itemize}
\vfill
\textbf{\color{black}Acknowledgements}:
\begin{itemize}
\item Thanks to the Norwegian Defence Research Establishment (FFI) and Kongsberg Maritime for providing us with data.
\end{itemize}
\end{frame}
}

% 
% {
% \setbeamertemplate{background}{\graphicsAI[width=\paperwidth]{gfx/AUVenv.png}}%
% \begin{frame}
% \vspace{-15pt}\frametitle{\color{white}Conclusion}
% \framesubtitle{}
% \vspace{-5pt}\color{white}
% \begin{itemize}
% \item \color{white}GPUs are well suited to accelerate algorithms such as the MVDR beamformer
% \item \color{white}Our GPU implementation of the MVDR beamformer
% \begin{itemize}
% \item \color{white}fully supports active sonar systems,
% \item \color{white}easily outperforms CPU implementations, and
% \item \color{white}runs fast enough to support realtime sectorscan imaging.
% \end{itemize}
% \end{itemize}
% \begin{figure}[H]%[width=0.5\linewidth]
% \graphicsAI<1>[width=0.3\linewidth]{gfx/happy_fish.png}\\
% \color{black}\centering\LARGE\textbf{Questions?}
% \end{figure}
% % \vspace{-10pt}
% \end{frame}
% }
% 
% % \begin{frame}
% % \frametitle{}
% % \framesubtitle{}
% % 
% % \begin{figure}[H]%[width=0.5\linewidth]
% % \graphicsAI<1>[width=0.3\linewidth]{gfx/happy_fish.png}
% % \end{figure}
% % \vspace{-10pt}
% % \centering\LARGE\textbf{Questions?}
% % 
% % \end{frame}
%

\end{document}

% \documentclass[
%   ucs,
%   utf8
%   12pt,                   %extsize
%   draft,                  %for empty decorations
%     hyperref={
%       breaklinks=true,      %Break links when necessary
%       linktocpage=true,     %Enable link to page?
%       linkcolor=PineGreen,  %Colour of links to labels within document
%       citecolor=Brown,      %Colour of links to the biliography
%       filecolor=Red,        %Colour of links to local files
%       pagecolor=Red,        %Colour of links to other pages withing document
%       urlcolor=Blue,        %Colour of the links to external URLs
%       colorlinks=true,      %??
%       plainpages=false,     %Store roman/arabic numbering differently to avoid
%       bookmarksnumbered
%     },                      %``duplicate'' warning.
% %   usepdftitle={},         %
%     xcolor={
%       table,                %For colour in tabulars (will pull in colortbl)
%       dvipsnames,
%       svgnames
%     }
%   c,                      %centered
%   t,                      %top aligned
%   compress,               %
%   trans,                  %
%   noamsthm,               %
%   notheorems,             %
%   envcountsec,            %
%   ignoreonframetext,      %
%   handout,                %
%   notes={}                %
% ]{beamer}

### 2018-02-06 ECUA2012/template/ECUA_2012_template.tex
Major commit before messing with PhD thesis

\documentclass[10pt,a4paper]{article}

% formatting requires ecua.sty style file 
\usepackage{ecua}
% load some other useful packages
\usepackage{cite} % improved formatting for citation lists
\usepackage{graphicx}

\title{\textrm{\LaTeX} template for the 11th European Conference on Underwater Acoustics, ECUA 2012 [max. 3 lines for title, left justified]}

% please edit the following tabular environment
% for \author block to match conference formatting
% use & to separate author and institution
% end each author line with \\
\author{%
\begin{tabular}{p{30mm}p{125mm}}
AN Author & Institute of Acoustics, St Albans, Herts., UK \\
AN Other &  Dept. of Electrical \& Electronic Engineering, Loughborough University of Technology, Loughborough, UK\\
D Bird & Cricket Acoustics Ltd., Dolphinston, Roxburghshire, UK \\
\end{tabular}
}

\begin{document}

\maketitle

\section{Introduction}
This document provides formatting suitable for paper submissions to ECUA 2012. It uses package ecua.sty available by separate download from the conference website. In case of difficulty in obtaining the style file or to report bugs or deficiencies please contact c.capus@hw.ac.uk. Some good references for help in using \textrm{\LaTeX} are listed below\cite{kopka:latex, goosens:latex, pakin:latexSymbols}.

\section{Formatting [Section Headers are set in Uppercase `L\MakeLowercase{arge}'  type, 14pt size]}
This exemplifies the section heading style. For specific cases where lowercase is needed for understanding, as above, it can be forced using the command \textbackslash\texttt{MakeLowercase}\{\texttt{text to be set in lowercase}\} --- this can also be applied in the paper title and subsection headers if required.

\subsection{[Subsection Headers are set in Uppercase `\MakeLowercase{large}'  type, 12pt size]}
This exemplifies the subsection heading style.

\subsubsection{[Subsubsection Headers are Set in `\MakeLowercase{large}' Type, 12pt Size]}
This exemplifies the subsubsection heading style, which should be set in titlecase --- that is, capitalise only the first letter of each word. Further heading levels are strongly discouraged.

\subsection{Main body text}
Main body text is typeset in 10pt Helvetica sans serif font with single spacing. Page margins are set at: left 27mm; right 27mm; bottom 30mm; top 40mm on title page and 30mm on subsequent pages. Text should be fully justified.

Do not include any page numbers. These will be added by the typesetters once all of the papers have been collated.

\subsection{Headers and Footers}
Every page shall carry the header: Proceedings of the 12th European Conference on Underwater Acoustics. Every page shall carry the footer: Vol. XX. Pt. Y 2009. The volume and part numbers will be provided by the conference organisers.  All headers and footers are left justified and set in bold type.

\section{Tables and Figures}

Tables and Figures are set in the floating table and figure environments. Table and Figure positioning is suppressed at the top of the first page.

\subsection{Figures}

The graphics and graphicx packages are recommended for Figure placement, an example is given in Figure~\ref{fig1:egFig}. The caption is included below the Figure.

\begin{figure}
\begin{center}
\includegraphics[width=10cm]{./egFigure.png}
\caption{\label{fig1:egFig} Example graph -- please avoid unnecessary use of colour and ensure line and marker styles allow discrimination when printed or copied in black and white}
\end{center}
\end{figure}

\subsection{Tables}
Tables are included using the tabular environment within a table float. An example is given in Table~\ref{tab1:egTable}. At the authors' discretion, entries in the titles line may be set in bold type, as in this example. The caption is included below the Table.

\begin{table}[h]
%\renewcommand{\tabcolsep}{1em}
\begin{center}
\begin{tabular}{|r|rrr|}
\hline
\textbf{X value} & \textbf{Y1 value} & \textbf{Y2 value} & \textbf{Y3 value} \\
\textbf{(units)} & \textbf{(units)} & \textbf{(units)} & \textbf{(units)} \\
\hline
1 & 2 & 1.5 & 1  \\
2 & 4 & 3.0 & 4  \\
3 & 6 & 4.5 & 9 \\
4 & 8 & 6.0 & 16 \\
5 & 10 & 7.5 & 25 \\
6 & 12 & 9.0 & 36 \\
7 & 14 & 10.5 & 49 \\
8 & 16 & 12.0 & 64 \\
9 & 18 & 13.5 & 81 \\
10 & 20 & 15.0 & 100 \\
\hline
\end{tabular}
\caption{\label{tab1:egTable} Example table}
\end{center}
\end{table}

\section{Other Formatting}

\subsection{Equations and Equation Numbering}

Equations are included with the equation and eqnarray environments. Numbers should appear in brackets to the right of each numbered equation line. Equation references in the text are of the form equation~(\ref{eqn1}).

\begin{equation}
\Re = \sqrt{\frac{\alpha}{\beta}}^n
\label{eqn1}
\end{equation}

\subsection{Footnotes}
Footnotes are discouraged due to the reduction in font sizes in the reprinted materials. If deemed necessary they can be included using the \textbackslash\texttt{footnote}\{\} command.\footnote{which will set in microscopically small text at the bottom of the current page if space is available.}

\subsection{Alternative Font Styles}

If required for particular emphasis, a limited number of alternative font styles can be selected as follows:

\begin{tabular}{ll}
COMMAND & RESULT \\
\textbackslash\texttt{texttt}\{\texttt{Teletype font}\} & \texttt{Teletype font} \\
\textbackslash\texttt{textrm}\{\texttt{Roman font}\} & \textrm{Roman font} \\
\textbackslash\texttt{textsc}\{\texttt{Small Caps font}\} & \textsc{Small Caps font} \\
\textbackslash\texttt{textbf}\{\texttt{Bold font}\} & \textbf{Bold font} \\
\textbackslash\texttt{textit}\{\texttt{Italic font}\}  & \textit{Italic font} \\
\textbackslash\texttt{emph}\{\texttt{Italic font}\}  & \emph{Italic font} \\
\end{tabular}

\section{Citations and References}
An ECUA \textrm{\BibTeX} style file, ecua.bst, is available from the conference website. This is based on the standard \texttt{plain} style file. Citations are made in the usual way using the \textbackslash\texttt{cite}\{\} command. Examples of various entry types can be found in the References section below: a journal article~\cite{capus:biosonar}; an article published in conference proceedings~\cite{thode:marRecord}; a book~\cite{au:dolphins}; a chapter in an edited book~\cite{abramowitz:handbook}; a fully attributed section from an edited volume~\cite{green:mcmc}; a PhD thesis~\cite{muller:bats}; a technical report~\cite{APL-UW:handbook}.

\section*{Acknowledgements}
Any acknowledgements section should immediately precede the References with no section number. This is achieved using the starred \textrm{\LaTeX}\,tag \textbackslash\texttt{section*}\{\texttt{Acknowledgements}\}.

\bibliographystyle{./ecua}
\bibliography{ecua}

\end{document}

### 2018-02-06 LCA/article.tex
Major commit before messing with PhD thesis

--- 2017-05-11 Simulator/article.tex
@@
 \newif\ifTODO\TODOtrue                        % Use todo notes?
-
-\newif\ifBuildBibliography\BuildBibliographyfalse  % Whether to include the references.bib file or use inline references.
 
@@
 
-% \usepackage{fixltx2e}                           % Fix some twocolumn float problems
-
-% \usepackage{stfloats}                          % Allows: \begin{figure*}[!b]
+\usepackage{fixltx2e}                           % Fix some twocolumn float problems
+
+\usepackage{stfloats}                          % Allows: \begin{figure*}[!b]
                                                 % (double column figures on top/bottom)
@@
 %       papers and their use requires extra complexity and work.
-% \newcommand\MYhyperrefoptions{bookmarks=true,bookmarksnumbered=true,
-% pdfpagemode={UseOutlines},plainpages=false,pdfpagelabels=true,
-% colorlinks=true,linkcolor={black},citecolor={black},urlcolor={black},
-% pdftitle={Optimal Window Design for the Low Complexity Adaptive Beamformer in Active Sonar Imaging},
-% pdfsubject={},
-% pdfauthor={Jo Inge Buskenes},
-% pdfkeywords={adaptive beamforming, beamforming, complexity, sonar, active}}%
+\newcommand\MYhyperrefoptions{bookmarks=true,bookmarksnumbered=true,
+pdfpagemode={UseOutlines},plainpages=false,pdfpagelabels=true,
+colorlinks=true,linkcolor={black},citecolor={black},urlcolor={black},
+pdftitle={A Fast GPU Sonar Simulator for Automatic Target Recognition},
+pdfsubject={},
+pdfauthor={Jo Inge Buskenes},
+pdfkeywords={adaptive beamforming, beamforming, complexity, sonar, active}}%
 \ifCLASSINFOpdf
-%    \usepackage[\MYhyperrefoptions,pdftex]{hyperref}
-   \usepackage[pdftex,
-    final,                %Keep hyperlink stuff even when in draft mode
-    breaklinks=true,      %Break links when necessary
-    linktocpage=true,     %Enable link to page?
-    linkcolor=black,  %Colour of links to labels within document
-    citecolor=black,      %Colour of links to the biliography
-    filecolor=red,        %Colour of links to local files
-    pagecolor=black,        %Colour of links to other pages withing document
-    urlcolor=Blue,        %Colour of the links to external URLs
-    colorlinks=true,      %??
-    plainpages=false,     %Store roman/arabic numbering differently to avoid
-    bookmarksnumbered ]{hyperref}
-% 
-% \else
-%    \usepackage[\MYhyperrefoptions,breaklinks=true,dvips]{hyperref}%
-\fi
-%    \usepackage{breakurl}                        % Allows 'dvips' driver to break links
+   \usepackage[\MYhyperrefoptions,pdftex]{hyperref}
+\else
+   \usepackage[\MYhyperrefoptions,breaklinks=true,dvips]{hyperref}
+   \usepackage{breakurl}                        % Allows 'dvips' driver to break links
+\fi
 
@@
 \newcommand*\B{\mat B}
-% \newcommand*\C{\mat C}
+\newcommand*\C{\mat C}
 \newcommand*\E{\mat E}
@@
 \renewcommand*\L{\mat \Lambda}
-% \newcommand*\U{\mat U}
+\newcommand*\U{\mat U}
 % \renewcommand*\t{\mathtiny{^T}}
@@
 
-\newcommand\minus{\scalebox{0.75}[1.0]{$-$}}
-
-% \usepackage{tikz}
-% \usetikzlibrary{shapes,snakes}
+\usepackage{tikz}
+\usetikzlibrary{shapes,snakes}
 \usepackage{amsmath,amssymb}
@@
 
-% \newenvironment{outline}
-% {\begin{itemize}}
-% {\end{itemize}}
+\newenvironment{outline}
+{\begin{itemize}}
+{\end{itemize}}
+
+% correct bad hyphenation here
+% \hyphenation{op-tical net-works semi-conduc-tor}
+
+%%%%%%%%%%%%                                    ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
+% GLOSSARY %
+%%%%%%%%%%%%                                    ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
+
+% \makeglossaries
+% \input{../../Library/glossary}
+
+%%%%%%%%%%%%%%%%%%                              ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
+% DOCUMENT START %
+%%%%%%%%%%%%%%%%%%                              ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
+
+% \usepackage{yfonts}
+
+% \graphicspath{submission/final/gfx/}
 
@@
 
+\maketitle
+
 \section{Introduction}
 
-\setcounter{totalnumber}{1}
-\setcounter{topnumber}{1}
-\setcounter{dbltopnumber}{1}
-
-\begin{figure*}[t]\centering%
-\graphicsAI[width=\linewidth]{gfx/scene_coordinate_system.svg}%
-\caption{Simulator principle.
-A model of the seafloor and object is first loaded into OpenGL, and an orthogonal camera view set up to match the desired image region on the seafloor.
-Then the scene is rendered and an intensity image and a depth map is created.
-The intensity image along with the depth information is finally combined to form a sonar like image.
-This is our template.}\label{simulator_coordinate_system}%
-\end{figure*}
-
-\begin{figure*}[t]\centering%
-\graphicsAI[drawing,width=\linewidth]{gfx/simulator.svg}%
-\caption{Simulator concept.}\label{buildup}%
-\end{figure*}
-
-% \IEEEPARstart{W}{hen} 
-% Autonomous underwater vehicles (AUVs).
-When an AUV with a sonar imaging system encounter an interesting object, it would be very beneficial if adapted its behavior to examine the object further.
-This is challenging as it would need to learn in what ways the interesting objects differ from the background, i.e.
-it relies on previous knowledge.
-It would also need to process this information is near real-time to act in an efficient manner.
-
-% - Adapt simulation to its behaviour, see if they still match.
-
-\IEEEPARstart{A}{utonomous} underwater vehicles (AUVs) are becoming increasingly important in human underwater endeavors.
-When equipped with a synthetic aperture sonar (SAS), they can e.g.
-be used to map the seafloor, survey underwater biology, help localize wrecks, inspect oil pipes, fields and electricity lines, or hunt for mines and biohazard waste.
-Such independent systems have much to gain from being able to adapt to the scenario at hand.
-For instance, with a well designed automatic target recognition (ATR) system on board they can adapt their mission plan to collect more information on particularly interesting objects, or fine-tune the sonar and navigation system to improve imaging performance.
+One of the major shortcomings of autonomous underwater vehicles (AUVs) is their lack of curiosity.
+When encountering an interesting object, an AUV with a mounted sonar imaging system should be able to detect it, and perhaps move in for closer examination.
+Besides the apparent challenge of trying to detect something with a long list of unknown parameters, this must happen in real-time and be very power efficient.
 
@@
 
-In conventional ATR an object of interest is first isolated, its properties attempted estimated, and a classifier used to determine what object class these properties belong to.
-The challenge is finding unique and intrinsic object properties, such as its type, shape and size, and to do so independently of extrinsic properties such as its alignment, burial depth, degeneration level or view angle.
-While SAS provide images with state of the art resolution and dynamic range, they still only represent an ambiguous acoustical 2D projection of the 3D object, littered with a highly coherent noise floor.
-
-An alternative approach is to use template matching, where the isolated image segment is compared to a set of template images covering the relevant object classes, and assigned to the class of the template with the best fit.
-Since the templates are computed from 3D models, this reduces the issue of image ambiguity.
-However, a challenge is finding a set of templates that represent the actual configurations of object and seafloor, Another issue is that the size of precomputed library of templates grows exponentially with the number of parameters needed.
+The detection process is well described in the field of automatic target recognition (ATR).
+The classical approach involves isolating an object of interest, extracting parameters from it and feeding those into a classifier.
+However, it is hard to find parameters that describe the objects uniquely, in particular due to the ambiguities arising from interpreting a 3D world from its 2D projection.
+Another approach is to use template matching, where the image segment is compared to a set of template images covering the relevant object classes, and assigned to the class of the template yielding the the best fit.
+The challenge here is to come up with accurate template images covering the actual configurations of object and seabed, especially if a static template library is used, which will lead to a computational complexity that is exponential with the number of model parameters.
 This limits its use to only a few parameters being coarsely sampled, ultimately yielding inaccurate results~\cite{Midelfart2010}.
 
-% We avoid the problems of conventional template matching by computing the templates on-the-fly as part of the classification process, thereby not needing a precomputed set of templates at all. 
-%  that avoids this problem by running sufficiently fast to create templates adapted to the actual scene as part of the classification process.
-% Classification process
-
-We tackle the problems of conventional template matching by performing:
-%
-\begin{itemize}
-\item \emph{Adaptive template matching}; where a template is selected based on object and seafloor parameters estimated from the SAS image~\cite{Midelfart2010}, and
-\item \emph{Real-time template simulation}; where a template is created tailored to the estimated parameters, improving its accuracy and avoiding the need of a precomputed template library.
-\end{itemize}
-%
-
-The methods presented herein will focus on the template simulator, but some details will also be given on adaptive template matching for completeness.
-More details on this technique can be found in~\cite{Midelfart2010}.
-
+We have developed a synthetic aperture sonar (SAS) simulator that avoids this problem by running sufficiently fast to create templates adapted to the actual scene as part of the classification process.
 This is achieved with the aid of the massive computing power in graphics processing units (GPUs) and optimized software libraries for scene rendering.
-The simulator loads a 3D model of the seafloor and object class into OpenGL and positions the camera at the sonar location.
-The scene is highlighted using a light tube that extends to infinity along the AUV's axis of probagation.
+The simulator loads a 3D model of the seafloor and an object class model into OpenGL, where emitted sound waves are modeled with a light tube extending all the way down the propagation path of the AUV.
 We assume rough, isotropic surfaces reflecting sound energy equally in all directions.
 This can be modeled with a Lambertian scattering model~\cite{Blake1993,Bell1995}, where the intensity of the backscattered sound solely depends on the incidence angle of the transmitted signal onto the model surface.
-The OpenGL rendering produces an optical 2D intensity image and depth map that reveals the distance from each pixel to the propagation axis.
+When rendering OpenGL is set up to produce an optical 2D intensity image and depth map that reveals the distance from each pixel to the propagation axis.
 For maximum flexibility this data is finally combined with OpenCL to produce the image templates.
+
+The alternative of using of a static, predetermined template library has a computational complexity that is exponential with the number of parameters.
+This limits its use to only a few parameters being coarsely sampled, ultimately yielding inaccurate results~\cite{Midelfart2010}.
+
+To avoid this problem we have developed a very fast side-looking sonar simulator that generates templates adapted to the actual scene.
+
+synthetic aperture sonar (SAS)
 
@@
 
-For lack of a better name, the simulator will be referred to as "FFISim" throughout this article.
-
-Show:
 \begin{itemize}
-\item Adaptive template matching to avoid large static template libraries,
-\item Improved classification performance by feeding simulator with parameters derived off SAS image,
-\item Further improved classification by performing a fine-search within a narrow search space around the SAS derived parameters. GPU makes this more powerful.
+\item SWAT (Shallow Water Acoustic Toolset): 
+\item Bell's thesis:
+\item SIGMA:
+\item OpenGL does not support scattered writes. OpenCL does.
+\item 2D template approximation:
+\item Geometric distortions in SSS \cite{Cobra1992} Bell
+\item Sidescan sonar and synthetic aperture sonar. In seafloor coordinates. Maps to orthogonal rendering.
+\item Layover causes ambiguity, different seafloor phenomena can cause the same output image.
+Lots of cornercases when approximating the model in 2D.
+Instead use a 3D model and project it to a 2D one.
+Easier to operate, models are more general purpose.
+Less parameters to adjust.
+\item Viewer can be used to train an operator for finding non-trivial objects on the seafloor. Input for automated target recognition standard.
+\item Improved image quality of selected object, reduced data collection cost (price estimate?).
+\item OpenGL rendering can have infinite resolution, only limited by the pixel density in the image.
+\item Accuracy in terms of classification result. Need to embody static effects in a deterministic way. Incoherent synthesis of the image.
+\item Sound speed assumed constant.
+\item High frequency lead to little (define) penetration, lower range and higher resolution. Often modeled accurately with ray tracing techniques.
 \end{itemize}
+
+Model
+\begin{itemize}
+\item Ray solution: Separate amplitude and phase components of the Helmholtz eq.
+by applying geometric optics approximation.
+Ok as long as amplitude varies more slowly with position than the phase.
+Implications: High frequency, because the curvature of a ray over a wavelength must be small, and the fractional change in the sound speed must be small over a wavelength.
+Further simplification available by Snell's Law.
+Fast to compute.
+Not able to handle diffraction, focal points and caustics, and will cause (too) abrupt changes on distinct changes in sea environment.
+Can be improved upon with fuzzy beam modeling / Gaussian beam tracing.
+\begin{itemize}
+\item Pro: Easy to compute. Visual result. Handles complex bathymetry. Array and model response can be accounted for.
+\item Con: Only works for high frequency and small sound variations. No diffraction, caustics and focus. Infinite energy caustics/sharp variations. Incoherent(?)
+\end{itemize}
+\item Normal modes: Separate amplitude and phase into a range dependent term, and a depth dependent term.
+Assumes minimal range dependence in sound speed profile and depth.
+Limited near field accuracy.
+Acoustic field does not need to be calculated at intermediate ranges.
+Penetration can be modeled.
+can be calculated
+\begin{itemize}
+\item Pro: Ok to compute. Handles penetration. Low frequency, shallow water.
+\item Con: Horizontal stratification. Far field only. 
+\end{itemize}
+\item Parabolic equation: One way propagation.
+Marching solution.
+Initial starting solution must be known, often computed with normal mode.
+Only valid for narrow beams $\pm$20.
+Able to handle range dependent environments.
+Computing full field is easy.
+Computation time is of O(f$^2$), hence mostly useful for low-frequency scenarios.
+Unable to deal with horizontal refraction (shadow zones from e.g.
+conical mounts)
+\item Fast field programs / Green's function solutions.
+Exact full wave solution for acoustic fields in horizontally stratified media.
+Can handle both compressional and shear waves, and hence both fluid and solid media.
+Far field assumption.
+Similar to normal mode techniques.
+\item Finite element: Scene segmented into triangles or rectangles with a size of one tenth of a wavelength, or less.
+At each node the wavefield can be determined from the wave equation.
+Computationally limited to low frequencies.
+\item 2D model: Acoustic field in range and depth.
+Assumes that a wave transmitted in a vertical plane remains in it.
+Reasonable when the sound speed can be assumed constant in the horizontal plane, and when the seafloor normals are in the vertical range-depth plane (avoid horizontal refraction, or actually reflection).
+\end{itemize}
+
+What's new?
+\begin{itemize}
+\item Tuned for speed using GPUs, combining OpenGL and OpenCL to take advantage of existing functionality for computer graphics with the added flexibility of general purpose programming on the GPUs.
+\item Performance to spare: Game-like viewer for the visualization and verification of the rendered scene and corresponding sonar image.
+\item First simulator of its kinds aimed towards improved ATR and ultimately better AUVs.
+\item 3D models are still novel. 
+\end{itemize}
+
+% The ATR algorithms must, however, be sufficiently fast to process the sensor data onboard the AUV in delayed real-time.
+
+% We have developed a simulator for synthetic aperture sonar (SAS) that creates object templates very quickly with the aid of the massive computing power in graphics processing units (GPUs).
+It loads a 3D model of the sea floor and an object of interest into OpenGL, where emitted sound waves are modeled with a light source placed at the sonar transmit location.
+We assume rough, isotropic surfaces that reflects sound energy equally in all directions.
+This can be modeled with a Lambertian scattering model~\cite{Bell1997} where the intensity of the backscattered sound is said to depend only on the on incidence angle of the transmitted signal onto the model surface.
+When rendering OpenGL is set up to produce an optical 2D intensity image and depth map that reveals the distance from each pixels to the propagation axis.
+For maximum flexibility this data is finally combined with OpenCL to produce the image templates.
+% 
+% By coincidence our implementation is not so different from the SIGMAS+ simulator developed at the NATO Undersea Research Centre (NURC)~\cite{Coiras2009a, Coiras2009b}.
+They both take advantage of OpenGL to render a 2D image from 3D models, are designed for side-looking sonar, and assume a Lambertian scattering model.
+However, while SIGMAS+ is geared more towards realistically looking sonar images, and include effects such as noise, we do not.
+Also, their simulator sums images obtained from multiple OpenGL rendering passes to create the sonar image, while we render once and post-process this result with OpenCL to tailor the sonar image full flexibility.
+
+\begin{figure}[t]\centering%
+\graphicsAI[drawing,width=\linewidth]{gfx/specs.svg}%
+\caption{Simulator specifications.}\label{specs}%
+\end{figure}
+
+\begin{figure*}[t]\centering%
+\graphicsAI[drawing,width=\linewidth]{gfx/simulator.svg}%
+\caption{Simulator concept.}\label{buildup}%
+\end{figure*}
 
@@
 
-Models are not ROCCAN (???)
-
-\subsection{Loading the scene}
+\subsection{Setting up the scene}
+
+Before we can render anything with OpenGL we first have to set up the scene. This involves loading the models and then placing them where we want them with the proper orientation.
 
 Our models are stored in regular 3D model files.
-The simulator begins by loading these into a tree of nodes, each representing a vertex in the model with corresponding facet and texture information.
-Then the vertices are placed into a virtual scene in OpenGL, and a camera set up to capture the part of the scene corresponding to the image.
-When OpenGL renders the image, it first computes which vertex contributes to which pixel in the image, then computes the intensity and color of the pixel based on the node's facet, texture and scattering model.
-The code that operates on vertices are called a vertex shader, and the code that operates on pixels are called a fragment shader.
-
-The mapping of the vertices from object coordinates to image coordinates happen through a set of transformations:
-%
+The simulator loads these with the Open Asset Import library (\texttt{assimp}).
+This is a portable open source library that supports loading a wide range of 3D model formats in a uniform manner.
+It converts the model into a node tree where each node contains data such as vertices, facets and textures, all formatted in an OpenGL friendly way.
+
+The next step is to scale, orient and position the objects in the scene, and finally project this scene onto an image plane.
+This is achieved by applying a set of transformations to each vertex in the model.
+In OpenGL, the set of transformations can be formalized as
 \begin{align*}
-\bmat{x\\y\\0\\1}_\text{image} = \P \cdot \V \cdot \M \cdot \bmat{x\\y\\z\\1}_\text{object},
+\bmat{x\\y\\0\\1}_\text{image} = \P \cdot \V \cdot \M \cdot \bmat{x\\y\\z\\1}_\text{model},
 \end{align*}
-%
 where $\bmat{\cdot}_\text{model}$ is the loaded model, $\bmat{\cdot}_\text{image}$ is the output image, and $\M$, $\V$ and $\P$ are transformation matrices. These will be described next.
@@
 The first transform we apply is the model matrix $\M$, which operates in the model's local coordinate system. It is defined as
-%
-\begin{align}
-\M = \bmat{x_t \\ y_t \\ z_t \\ w}_\text{\parbox{1cm}{\setlength\baselineskip{0.3cm}transformed\\model}} &= \boldsymbol{T} \cdot \R \cdot \boldsymbol{S} \bmat{x \\ y \\ z \\ w}_\text{model},
-\end{align}
-%
+\begin{align}
+\bmat{x_t \\ y_t \\ z_t \\ w}_\text{\parbox{1cm}{\setlength\baselineskip{0.3cm}transformed\\model}} &= \boldsymbol{T} \cdot \R \cdot \boldsymbol{S} \bmat{x \\ y \\ z \\ w}_\text{model},
+\end{align}
 where $\boldsymbol{S}, \R$ and $\boldsymbol{T}$ are matrices\todo{be explicit on dependencies?} for scaling, rotation and translation, respectively.
@@
 Next we map the model into world coordinates using the view matrix $\V$.
-This involves aligning the camera to precisely fit the image region.
+This involves aligning and positioning the model into the view of the camera.
 In our case the camera represents the phase center of the sonar, and it is placed in the world origin.
-The transformation then take the form of
-%
+The transformation take the form of
 \begin{align}
@@
 \end{align}
-%
-where $(x_\text{c}, y_\text{c}, z_\text{c})$ is the image center coordinates for along track, cross track, and depth, respectively, and $\phi_x$ is the elevation angle\todo{is this correct?} of the sonar.
+where $(x_\text{c}, y_\text{c}, z_\text{c})$ is the image center coordinates for along track, cross track, and depth, respectively, and $\Phi_x$ is the elevation angle\todo{is this correct?} of the sonar.
 It is given as
-%
 \begin{align*}
@@
 The matrix that gives orthographic projection is defined as
-%
 \begin{align}
@@
 \end{align}
-%
 where $l,r,b$ and $t$ are the left, right, bottom and top boundaries of the projection plane, respectively, and where $n$ and $f$ is the near and far limits of the cube that encompass the scene.
@@
 \end{align}
-%
-For the near and far limits one simply needs to make sure the scene will fit in it, say
-%
-\begin{align}
-n &= 0\;\text{m}   &  f &= 300\;\text{m}.
-\end{align}
-%
-The bottom and top boundaries can be inferred from \Fig{camera_top_bottom}. First we define an origin vector that points to the center of the image
-%
+For the near and far limits one simply needs to make sure the scene will fit in it
+\begin{align}
+n &= 0\;\text{m}   &  f &= 300\;\text{m} 
+\end{align}
+How to set the bottom and top bounaries can be inferred from \Fig{camera_top_bottom}. First we define an origin vector that points to the center of the image
 \begin{align}
@@
 \end{align}
-%
 and the distance from the image origin to its $y$-boundaries is given as
-%
-\begin{align}
-\Delta y &= \frac{|y_\text{max} - y_\text{min}|}{2}.
-\end{align}
-%
+\begin{align}
+\Delta y &= \frac{y_\text{max} - y_\text{min}}{2}.
+\end{align}
 To find the nearest and farthest point in the image that will be part of the projected scene, $\alpha$ and $\beta$, we must consider the seafloor rotation angle around the $x$-axis, $\Phi_x$ 
-%
-\begin{align}
-\vec\alpha &= \vec 0 - \left[0, \Delta y, \Delta y \cdot \tan\Phi_\text{x} \right] \nn
-\vec\beta &= \vec 0 + \left[0, \Delta y, \Delta y \cdot \tan\Phi_\text{x} \right]
+\begin{align}
+\alpha &= \vec 0 - \left[0, \Delta y, \Delta y \times \arctan\Phi_\text{x} \right]
 \end{align}
@@
 
-\todo{
-- Kernel tuning.
-- Sharing. PBO. Asynchronous.
-}
-
-\subsection{Template matching}
-
-1. Segmentation
-2. Parameter estimation
-   Available: Range, altitude, resolution => Input to both standard and adaptive. In standard the best fit out of a predefined template database is selected, then resampled to correct resolution.
-   Additional for adaptive, parameters that modify object model: Target aspect angle, seafloor slope, object length and height.
-
-   Target aspect angle computed with Radon Transform, seafloor under object modeled as plane with a slope estimated from bathymetric data, length estimated from longest part of highlight region, and height estimated from the shape of the shadow.
-The burial depth is n
-
-3.
-
-   Details: \ref{Midelfart2010}.
-
-Let $S(x,y)$ be the isolated part of the
-
-Matching the isolated image segment with the template take the form of
-
-\begin{align}
-c_{h,i} = \underset{x',y'}{\text{argmax}} \sum\limits_{\forall x}\sum\limits_{\forall y} \big| H(x,y) - T(x-x',y-y') \big|
-\end{align}
-
-\begin{align}
-c_i = \frac{c_{h,i} + c_{s,i}}{2}
-\end{align}
-
-And the final template used is the one with the greatest correlation coefficient
-
-\begin{align}
-\underset{T_i(x,y)}{\text{argmax}} C_i\sum\limits_{\forall x}\sum\limits_{\forall y}  S(x,y) T(x-x',y-y')
-\end{align}
-
-Then assign object to class $i$.
-
-To narrow down the search space of possible templates we derive some reliable statistics of the seafloor and object from the SAS image.
-While FFISim accepts an arbitrarily complex seabed, we find that modeling it as a plane wave with the right tilt works well.
-The object is more difficult.
-For instance, we estimate its length and height from the shape of its shadow, but this fails if the object is notably immersed into the seabed at one end.
-
-After the a template is created we decompose it into its highlight and shadow component, and compute a correlation score between these and the corresponding areas in the SAS image.
-Then the two scores are summed and normalized to form the final correlation score of the template.
-
-We tag the template matchers as either semi- or fully adaptive.
-The former refers to using the best matching SIGMAS+ generated template out of a precomputed set of templates, while the latter refers to the proposed method of using FFISim to generate templates on the-the-fly.
-\todo{Should be more precise.
-How many templates?
-Parameter span?
-How finely were the parameters sampled?}
+\begin{itemize}
+\item Kernel tuning.
+\item Sharing. PBO. Asynchronous.
+\end{itemize}
 
@@
 
-\todo{
-Write to file
-}
-
-%    """
-%    
-%    stm: Standard template matching, made with SIGMAS.
-%    atm: Adaptive template matching, made with FFISim.
-%    atm_bury: Adaptive template matching where a search of bury depths is used.
-%    atm_bury_restr: Restricted?
-%    
-%    Adaptiveness made by letting the template matcher use 5 different rotation
-%    angles in the ground-range axis. 
-%    
-%    
-%    Data from Jesusbukta
-%    
-%    Figures are from run 2012.06.16
-%    det210 - Cylinder 1: est 2.92m (short side towards sonar)
-%    det233 - Cylinder 2: est 3.05m (diagonal)
-%    det379 - Torpedo: est 5.06m (looks good)
-%    
-%    ROC curves from 3 different runs, first two just 2 days apart in 2009, last one in 2012.
-%    
-%    run_20120616 - last one. Much better with adaptive templates.
-%    
-%    ROC curves made by average correlation score of shadow/highlight from template matching.
-%    
-%    
-%    fig4b(2015.11.16)
-%       Bug that skipped masking of port side (negative y) detections past 180m range fixed
-%       
-%    fig4 (2015.11.16)
-%       Bug leading to incorrect estimation of cylinder orientation fixed.
-%          This affects run090601_1. Should be better after this.
-%          
-%    fig3 (2015.11.09)
-%       Detections outside 180m range have been masked (but bug for port side fixed in fig4b).
-%       Fixed an incorrectly annotated detection.
-%       
-%    fig2 (2015.11.02)
-%       Height of object estimated from shadow length
-%       Let adaptive method try 5 different rotations in ground-range (-4, -2, 0, 2, 4)
-%          Improved results, but incorrect classification of:
-%          5 detections in first run, and
-%          3 detections in second run
-%          => For FPR=0.2 the results are slightly worse than with normal adaptive (particularly first run).
-%          Images are problematic in these cases:
-%             2 detections get wrong orientation because a line (anchor) cross the image
-%             Other misclassifications in image regions with poor echo or shadow strength
-%          => Not a big issue as not enough time to detect more than ~20% of the detections in a run
-%             In a MCM operation these detections would never be used.
-%             The runs have 3000 detections each, photographing 20% of these is too much.
-%          Performance FPR>0.2 not interesting.
-%          Performance FPR<0.2 much better with adaptive template matching.
-%          
-%    fig (2015.10.23)
-%       atm - cylinder orientation and length estimated from SAS image and fed into simulator.
-%       
-%       ROC curves from last run. Computing the mean correlation score of the shadow and highlight/echo.
-%       Early results. Can only check vs. cylinder-like objects.
-%       Standard template library created based on estimation of physical size of known cylinder mines.
-%       Not possible for adaptive method to perform better when hitting such a target.
-%       
-%  
-%    """
+\begin{itemize}
+\item Write to file
+\end{itemize}
 
@@
 
-\begin{figure*}[!tbp]\centering%
-\graphicsAI[width=\linewidth]{gfx/fig_images_sonar_simulator_tagged.svg}\label{fig_images_sonar_simulation}%
-\caption{Selected objects.
-Top row show SAS images of 2 cylinders and a torpedo located in the Jesus Bay, Norway.
-Center row show the best match from a conventional template matcher, and bottom row show the adaptive template matcher.
-The contour of the SAS object is indicated with red lines.
-Note that the adaptive technique produce templates with a better fit to the highlight, but that this does not necessarily lead to a better match against the shadow.}
-\end{figure*}
-
-\begin{figure*}[!tbp]\centering%
-\graphicsAI[width=\linewidth]{gfx/fig_rocs.pdf}\label{fig_roc_curves}%
-\caption{Receiver operating characteristics (ROC) curves.}
-\end{figure*}
-
-\begin{figure*}[!tbp]\centering%
-\graphicsAI[width=\linewidth]{gfx/fig_overlay.pdf}%
-% \parbox{\linewidth}{\small\centering\vspace{.5\baselineskip}
-% \newline\centering\newline %\\[.5\baselineskip]
+\newlength\imgspacing\setlength\imgspacing{.5cm}
+\setcounter{topnumber}{2}
+
+\begin{figure}[p]\centering%
+\subfloat[SAS image of a cylinder with 2.6\;m length and 0.53\;m radius.]{%
+\graphicsAI[drawing,width=\linewidth]{gfx/data_cylinder_submerged.svg}%
+\label{data_cylinder}}%
+\subfloat[Cylinder simulation with parameters estimated from SAS image: Length 2.6\;m, burial depth 0.263\;m, aspect angle 105$^\circ$.]{
+\graphicsAI[drawing,width=\linewidth]{gfx/sim_cylinder_submerged_adaptive.svg}%
+\label{sim_cylinder}}%
+\caption{A SAS image of a cylinder and a template simulation adapted to it.}
+\end{figure}
+
 \newcommand\cdesc[2]{{\raggedright\setlength\fboxsep{0pt}
 \fbox{\colorbox[HTML]{#1}{\vrule height8.5pt depth3.5pt width0pt\hspace{.5cm}}}\ \ #2\\}}
-\begin{minipage}{.4\linewidth}
+
+\begin{figure}[p]\centering%
+\subfloat[Best template out of a predefined set. Inaccurate cylinder length, alignment and immersion depth. The segmented shadow and highlight regions are misaligned.]{
+\graphicsAI[drawing,width=\linewidth]{gfx/overlay_cylinder_submerged.svg}%
+\label{overlay_cylinder}}%
+% \hfil%
+\subfloat[Adapted template. Template simulated with adapted parameters. Good fit in both highlight and shadow segments.]{%
+\graphicsAI[width=\linewidth]{gfx/overlay_cylinder_submerged_adaptive.svg}%
+\label{overlay_cylinder_adaptive}}%
+\hfill\parbox{\linewidth}{\small\centering
+Segmentation overlay color description:\\[.5\baselineskip]
+\begin{minipage}{\linewidth}
 \cdesc{800000}{Template highlight and image highlight}
@@
 \end{minipage}%
-\begin{minipage}{.4\linewidth}
+\begin{minipage}{\linewidth}
 \cdesc{000083}{Template shadow and image shadow}
@@
 \cdesc{00A7FF}{Template shadow and image highlight}
-\end{minipage}%}
-\caption{Comparison of classification performance for the standard and adaptive template technique.
-The three objects were more or less arbitrary picked from the scene.
-For the roughly 3\,m long cylinders both methods perform similarly, the standard method seemingly with a slight edge.
-This is no surprise given that the 3\,m long cylinder model that make up the static template database fit the actual object well here.
-However, for the longer torpedo-like object the adaptive technique is clearly better.
-This is its merit; for objects and geometries that are not included in the static template database we can expect the adaptive techniques to perform better.}\label{fig_image_simulation}%
-\end{figure*}
-
-\newlength\imgspacing\setlength\imgspacing{.5cm}
-\setcounter{topnumber}{2}
-
-The performance of our simulator and template matching techniques where measured using experimental data from an interferometric HISAS1030 synthetic aperture sonar (SAS) attached to a HUGIN autonomous underwater vehicle (AUV), both developed by Kongsberg Maritime, Norway.
-The HISAS1030 is a fully digital phased array with 32 hydrophones, 100\,kHz center frequency and 30\,kHz bandwidth.
-It is 1.2\,m long, has a half-power beamwidth (HPBW) of 23$\circ$, and is capable of generating synthetic images with a theoretical resolution of 3-4\,cm in both dimensions.
-The version of HUGIN used also carries an camera that is used for optical inspection of interesting objects.
-
-We present data from three different HUGIN runs from the Jesus Bay in Norway.
-The first two are taken a couple days apart in January 2009, containing roughly 2700 and 3200 detections, respectively.
-The last is from June 2012 and contains roughly 650 detections.
-The detections represent mostly cylinder shaped objects such as barrels, pipe mines and torpedoes.
-Therefore, all templates were based on a basic cylinder model.
-
-The two template matchers can be visually inspected in \Fig{fig_images_sonar_simulator_tagged}.
-
-The classification performance is quantified with receiver operating characteristics (ROC) curves from all runs in \Fig{fig_roc_curves}.
-The 2009 runs proves a greater challenge than that from 2012, with 5 and 3 false positive detections at FPR=20\% in the January 6.
-and 8.
-run, respectively.
-This came as no surprise due to the 2009 runs containing fish trawler tracks and having poorer highlight and shadow quality than the run from 2012.
-However, for FPR<20\% the adaptive techniques outperform the static template library.
-For e.g.
-a mine counter measures (MCM) operation, this is the only region of interest because only a few objects out of the 3000 detections will be visually inspected up close.
-
-- Stm stimulated with length matching test cylinder. Perfect score against this.
-- Atm - not given to improve score. Shadow and highlight both need to match, and overfitting could lead to reduced performance.
-
-In \Fig{fig_images_sonar_simulator_tagged} we compare the SAS image to the templates generated by the conventional and adaptive template matcher.
-Three typical objects from the 2012 were chosen, all being cylindric.
-The length two first being estimated to roughly 3\,m and the last 5\,m, likely representing cylinder mines and a torpedo, respectively.
-
-Selected objects.
-Top row show SAS images of 2 cylinders and a torpedo located in the Jesus Bay, Norway.
-Center row show the best match from a conventional template matcher, and bottom row show the adaptive template matcher.
-The contour of the SAS object is indicated with red lines.
-Note that the adaptive technique produce templates with a better fit to the highlight, but that this does not necessarily lead to a better match against the shadow.
-
-Rotation around ground range axis: -4,-2,0,2,4 degrees.
-
-Like the majority of objects in the Jesus Bay they are all cylinders with a length of a few meters.
-The same objects are simulated in \Fig{fig_image_simulation}, one with SIGMAS and the two other with FFISim.
-Each of these simulations were based on
-
-Good statistics drawn from the image -> adaptive better. Poor statistics -> hit and miss.
-
-1. Comparison with standard template library (ROC enough?)
-   ROC: Adaptive better at FPR<0.2 in all cases.
-2. Comparison with SIGMAS
-   Fairly similar images, performance is similar. 
-   Used to create the standard template library.
-3. Comparison with search space
-   ROC: Additional search always better FPR<0.2.
-   Explain search criterion.
-
-SAS images - look good. Random pick objects. Lengths derived from image.
-Simulated images - look good. Adaptive simulation more accurate. Otherwise similar. Different background level. Adaptive technique excels for less normal objects/geometries.
-Overlap - Provide quantification measure.
-
-1. Adaptive better? Compare with standard. Doing this in ROC
-2. Improved classification? ROCs
-3. Further improvement by doing fine-search? ROC
-
-An issue the adaptive technique must deal with when estimating object length
-
-As the ROC curves indicate, the adaptive template matching techniques both  ROC curves demonstrate the
-
-The majority of the objects in the Jesus Bay are cylinder shaped. A SAS image of three such objects are displayed in \Fig{fig_image_sonar}
-
-% \begin{figure*}[!tbp]\centering%
-% \subfloat[SAS images of 2 cylinders and a torpedo.]{%
-% \graphicsAI[width=0.5\linewidth]{gfx/fig_images_sonar_simulation.pdf}%
-% \label{fig_sonar_simulation}}%
-% \subfloat[Cylinder simulation with parameters estimated from SAS image: Length 2.6\;m, burial depth 0.263\;m, aspect angle 105$^\circ$.]{
-% \graphicsAI[drawing,width=0.5\linewidth]{gfx/sim_cylinder_submerged_adaptive.svg}%
-% \label{sim_cylinder}}%
-% \caption{A SAS image of a cylinder and a template simulation adapted to it.}
-% \end{figure*}
-% 
-% 
-% \begin{figure*}[!tbp]\centering%
-% \subfloat[SAS image of a cylinder with 2.6\;m length and 0.53\;m radius.]{%
-% \graphicsAI[drawing,width=0.5\linewidth]{gfx/data_cylinder_submerged.svg}%
-% \label{data_cylinder}}%
-% \subfloat[Cylinder simulation with parameters estimated from SAS image: Length 2.6\;m, burial depth 0.263\;m, aspect angle 105$^\circ$.]{
-% \graphicsAI[drawing,width=0.5\linewidth]{gfx/sim_cylinder_submerged_adaptive.svg}%
-% \label{sim_cylinder}}%
-% \caption{A SAS image of a cylinder and a template simulation adapted to it.}
-% \end{figure*}
-
-% \newcommand\cdesc[2]{#2}
+\end{minipage}}
+\vspace{\baselineskip}
+\caption{Segmented simulation image overlaid the segmented SAS image.}
+\end{figure}
 
@@
 
-We implement rotations in a 3D euclidean space by combining matrices that rotate around each of the three respective dimensions. For rotation by an angle $\phi_x$ clockwise around the $x$-axis we have
+We implement rotations in a 3D euclidean space by combining matrices that rotate around each of the three respective dimensions..
+For rotation by an angle $\phi_x$ clockwise around the $x$-axis we have
 \begin{align}

### 2018-02-06 LCAforActiveSonarImaging/bio/jo_inge_buskenes_new.tex
Major commit before messing with PhD thesis

--- 2014-04-23 HowtoHowtoCapon/bio/jo_inge_buskenes.tex
@@
 He is currently pursuing the Ph.D.
-degree in image reconstruction and technology at the University of Oslo.
+degree in acoustic image reconstruction and high performance computing at the University of Oslo.
 
-His industry experience includes the European Organization for Nuclear Research (CERN), Geneva, Switzerland (2007-2008), and the Norwegian Defence Research Establishment, Kjeller, Norway (2009).
-He has lectured in digital signal processing at the Gj\o{}vik College University (2009), and at the University of Oslo (2010-2012).
-His research interests include adaptive beamforming, digital image reconstruction, high performance computing, intelligent detector design and open source software.
+His industry experience includes development of digital electronics at the European Organization for Nuclear Research (CERN), Geneva, Switzerland (2007-2008).
+He has lectured in digital signal processing at the Gj\o{}vik College University (2009), and at the University of Oslo (2010-2013).
+Current affiliation is with The Norwegian Defence Research Establishment, Kjeller, Norway, for which he is developing radar systems (2015-), and formerly sonar systems (2009, 2013).
+
+His research interests include radar and sonar technology, aptive image reconstruction, high performance computing, intelligent detector design and open source software.
 \end{IEEEbiography}

### 2018-02-06 LCAforActiveSonarImaging/diagram/overview.tex
Major commit before messing with PhD thesis

\documentclass[tikz]{standalone}

\usepackage{pgf}
% % \usepackage{pgf-blur}
\usepackage{tikz}
\usetikzlibrary{arrows,automata,graphs,
		fadings,mindmap,positioning,
		shadows,shadows.blur,
	        shapes} 
% \usegdlibrary{layered}

\pagestyle{empty}

\tikzstyle{decision} = [diamond, draw, fill=blue!20, 
    text width=4.5em, text badly centered, node distance=3cm, inner sep=0pt]

\tikzstyle{section} = [
   draw,
   ellipse,
   fill=gray!50!gray,
%    drop shadow,
   blur shadow={shadow blur steps=8},
   node distance=3cm,
   minimum height=2em,
   shade, shading=axis, top color=black!10!white, bottom color=black!25!white %shading angle=45
   ]

\tikzstyle{goal} = [
%    draw,
%    rounded rectangle,
%    fill=white,
%    drop shadow,
%    blur shadow={shadow blur steps=8},
   node distance=0.5cm,
   minimum height=0.2cm,
   minimum height=0.2cm,
   ]

\tikzstyle{structure} = [
   draw, ellipse,
   fill=white!20,
   drop shadow,
   node distance=3cm,
   minimum height=2em]

\tikzset{state/.style={
	rectangle,
	rounded corners,
	draw=black, very thick,
	minimum height=2em,
	inner sep=2pt,
	text centered,
	},
}

\begin{document}
\begin{tikzpicture}[->,>=stealth']

\node[section](abstract){ \bf Abstract };

\node[section, below of=abstract](introduction){ \bf Introduction };

\node[section, below of=introduction](methods){ \bf Methods };

\node[section, below of=methods](results){ \bf Results };

\node[section, below of=results](discussion){ \bf Discussion };

\node[section, below of=discussion](conclusion){ \bf Conclusion };

\node[goal, left of=introduction](intro_goals){ \bf Goals };
\node[goal, left of=intro_goals](intro_goals_why){ \bf Why };
\node[goal, below of=intro_goals_why](intro_goals_what){ \bf What };

\path
(abstract)        edge  (introduction)
(introduction)    edge  (methods)
   (introduction) edge  (goals)
(methods)         edge  (results)
(results)         edge  (discussion)
(discussion)      edge  (conclusion);

\end{tikzpicture}
\end{document}

### 2018-02-06 LCAforActiveSonarImaging/gfx/generated/gfx_input.tex
Major commit before messing with PhD thesis

\newpage
\begin{figure*}[!t]
\begin{narrow}{-1.2cm}{-1.2cm}\centering\vspace{-1.0cm}
\textbf{1. LCA with trigonometric and Kaiser windows - Capon shining}\\
\begin{tabular}[c]{l l l l}
\bf General & M = 32                            & $\Delta r = \frac{c}{2B}$ = 2.5 cm & $\frac{640\,\text{pixels}] / 12\,\text{m}}{\Delta r} = \frac{4}{3}$ \\
\bf LCA     & $\beta \in [0,10]$ (9 values) & $\phi \in [-1.07,1.07]$ deg (9 values) & Navg = 3 \\
\bf Capon   & $\Delta$ = 0.01                 & L = 16                           & Navg = 3 \\
\end{tabular}
\subfloat[LCA Window Response]{\includegraphics[width=0.49\linewidth]{gfx/1_window_response.png}}\hfill
\subfloat[Mean images]{\includegraphics[width=0.49\linewidth]{gfx/1_mean_imgs.png}}\\
\subfloat[Windows ($\beta$)]{\includegraphics[width=0.49\linewidth]{gfx/1_windows_beta.png}}\hfill
\subfloat[Windows ($\phi$)]{\includegraphics[width=0.48\linewidth]{gfx/1_windows_phi.png}}\\
\subfloat[Capon win. resp. through shadow]{\includegraphics[width=0.49\linewidth]{gfx/1_win_resp_cut_shadow.png}}\hfill
\subfloat[Capon win. resp. through highlight]{\includegraphics[width=0.49\linewidth]{gfx/1_win_resp_cut_highlight.png}}\\
\end{narrow}
\end{figure*}
\newpage
\begin{figure*}[!t]
\begin{narrow}{-1.2cm}{-1.2cm}\centering\vspace{-1.0cm}
\textbf{2. Capon: Tuning regularisation.}\\
\begin{tabular}[c]{l l l l}
\bf General & M = 32                            & $\Delta r = \frac{c}{2B}$ = 2.5 cm & $\frac{640\,\text{pixels}] / 12\,\text{m}}{\Delta r} = \frac{4}{3}$ \\
\bf LCA     & $\beta \in [0,10]$ (9 values) & $\phi \in [-1.07,1.07]$ deg (9 values) & Navg = 3 \\
\bf Capon   & $\Delta$ = 0.05                 & L = 16                           & Navg = 3 \\
\end{tabular}
\subfloat[LCA Window Response]{\includegraphics[width=0.49\linewidth]{gfx/2_window_response.png}}\hfill
\subfloat[Mean images]{\includegraphics[width=0.49\linewidth]{gfx/2_mean_imgs.png}}\\
\subfloat[Windows ($\beta$)]{\includegraphics[width=0.49\linewidth]{gfx/2_windows_beta.png}}\hfill
\subfloat[Windows ($\phi$)]{\includegraphics[width=0.48\linewidth]{gfx/2_windows_phi.png}}\\
\subfloat[Capon win. resp. through shadow]{\includegraphics[width=0.49\linewidth]{gfx/2_win_resp_cut_shadow.png}}\hfill
\subfloat[Capon win. resp. through highlight]{\includegraphics[width=0.49\linewidth]{gfx/2_win_resp_cut_highlight.png}}\\
\end{narrow}
\end{figure*}
\newpage
\begin{figure*}[!t]
\begin{narrow}{-1.2cm}{-1.2cm}\centering\vspace{-1.0cm}
\textbf{3. Capon: Tuning regularisation.}\\
\begin{tabular}[c]{l l l l}
\bf General & M = 32                            & $\Delta r = \frac{c}{2B}$ = 2.5 cm & $\frac{640\,\text{pixels}] / 12\,\text{m}}{\Delta r} = \frac{4}{3}$ \\
\bf LCA     & $\beta \in [0,10]$ (9 values) & $\phi \in [-1.07,1.07]$ deg (9 values) & Navg = 3 \\
\bf Capon   & $\Delta$ = 0.20                 & L = 16                           & Navg = 3 \\
\end{tabular}
\subfloat[LCA Window Response]{\includegraphics[width=0.49\linewidth]{gfx/3_window_response.png}}\hfill
\subfloat[Mean images]{\includegraphics[width=0.49\linewidth]{gfx/3_mean_imgs.png}}\\
\subfloat[Windows ($\beta$)]{\includegraphics[width=0.49\linewidth]{gfx/3_windows_beta.png}}\hfill
\subfloat[Windows ($\phi$)]{\includegraphics[width=0.48\linewidth]{gfx/3_windows_phi.png}}\\
\subfloat[Capon win. resp. through shadow]{\includegraphics[width=0.49\linewidth]{gfx/3_win_resp_cut_shadow.png}}\hfill
\subfloat[Capon win. resp. through highlight]{\includegraphics[width=0.49\linewidth]{gfx/3_win_resp_cut_highlight.png}}\\
\end{narrow}
\end{figure*}
\newpage
\begin{figure*}[!t]
\begin{narrow}{-1.2cm}{-1.2cm}\centering\vspace{-1.0cm}
\textbf{4. Capon: Tuning subarray.}\\
\begin{tabular}[c]{l l l l}
\bf General & M = 32                            & $\Delta r = \frac{c}{2B}$ = 2.5 cm & $\frac{640\,\text{pixels}] / 12\,\text{m}}{\Delta r} = \frac{4}{3}$ \\
\bf LCA     & $\beta \in [0,10]$ (9 values) & $\phi \in [-1.07,1.07]$ deg (9 values) & Navg = 3 \\
\bf Capon   & $\Delta$ = 0.01                 & L = 20                           & Navg = 3 \\
\end{tabular}
\subfloat[LCA Window Response]{\includegraphics[width=0.49\linewidth]{gfx/4_window_response.png}}\hfill
\subfloat[Mean images]{\includegraphics[width=0.49\linewidth]{gfx/4_mean_imgs.png}}\\
\subfloat[Windows ($\beta$)]{\includegraphics[width=0.49\linewidth]{gfx/4_windows_beta.png}}\hfill
\subfloat[Windows ($\phi$)]{\includegraphics[width=0.48\linewidth]{gfx/4_windows_phi.png}}\\
\subfloat[Capon win. resp. through shadow]{\includegraphics[width=0.49\linewidth]{gfx/4_win_resp_cut_shadow.png}}\hfill
\subfloat[Capon win. resp. through highlight]{\includegraphics[width=0.49\linewidth]{gfx/4_win_resp_cut_highlight.png}}\\
\end{narrow}
\end{figure*}
\newpage
\begin{figure*}[!t]
\begin{narrow}{-1.2cm}{-1.2cm}\centering\vspace{-1.0cm}
\textbf{5. Capon: Tuning subarray.}\\
\begin{tabular}[c]{l l l l}
\bf General & M = 32                            & $\Delta r = \frac{c}{2B}$ = 2.5 cm & $\frac{640\,\text{pixels}] / 12\,\text{m}}{\Delta r} = \frac{4}{3}$ \\
\bf LCA     & $\beta \in [0,10]$ (9 values) & $\phi \in [-1.07,1.07]$ deg (9 values) & Navg = 3 \\
\bf Capon   & $\Delta$ = 0.01                 & L = 16                           & Navg = 3 \\
\end{tabular}
\subfloat[LCA Window Response]{\includegraphics[width=0.49\linewidth]{gfx/5_window_response.png}}\hfill
\subfloat[Mean images]{\includegraphics[width=0.49\linewidth]{gfx/5_mean_imgs.png}}\\
\subfloat[Windows ($\beta$)]{\includegraphics[width=0.49\linewidth]{gfx/5_windows_beta.png}}\hfill
\subfloat[Windows ($\phi$)]{\includegraphics[width=0.48\linewidth]{gfx/5_windows_phi.png}}\\
\subfloat[Capon win. resp. through shadow]{\includegraphics[width=0.49\linewidth]{gfx/5_win_resp_cut_shadow.png}}\hfill
\subfloat[Capon win. resp. through highlight]{\includegraphics[width=0.49\linewidth]{gfx/5_win_resp_cut_highlight.png}}\\
\end{narrow}
\end{figure*}
\newpage
\begin{figure*}[!t]
\begin{narrow}{-1.2cm}{-1.2cm}\centering\vspace{-1.0cm}
\textbf{6. Capon: Tuning subarray.}\\
\begin{tabular}[c]{l l l l}
\bf General & M = 32                            & $\Delta r = \frac{c}{2B}$ = 2.5 cm & $\frac{640\,\text{pixels}] / 12\,\text{m}}{\Delta r} = \frac{4}{3}$ \\
\bf LCA     & $\beta \in [0,10]$ (9 values) & $\phi \in [-1.07,1.07]$ deg (9 values) & Navg = 3 \\
\bf Capon   & $\Delta$ = 0.01                 & L = 12                           & Navg = 3 \\
\end{tabular}
\subfloat[LCA Window Response]{\includegraphics[width=0.49\linewidth]{gfx/6_window_response.png}}\hfill
\subfloat[Mean images]{\includegraphics[width=0.49\linewidth]{gfx/6_mean_imgs.png}}\\
\subfloat[Windows ($\beta$)]{\includegraphics[width=0.49\linewidth]{gfx/6_windows_beta.png}}\hfill
\subfloat[Windows ($\phi$)]{\includegraphics[width=0.48\linewidth]{gfx/6_windows_phi.png}}\\
\subfloat[Capon win. resp. through shadow]{\includegraphics[width=0.49\linewidth]{gfx/6_win_resp_cut_shadow.png}}\hfill
\subfloat[Capon win. resp. through highlight]{\includegraphics[width=0.49\linewidth]{gfx/6_win_resp_cut_highlight.png}}\\
\end{narrow}
\end{figure*}
\newpage
\begin{figure*}[!t]
\begin{narrow}{-1.2cm}{-1.2cm}\centering\vspace{-1.0cm}
\textbf{7. Capon: Tuning time averaging.}\\
\begin{tabular}[c]{l l l l}
\bf General & M = 32                            & $\Delta r = \frac{c}{2B}$ = 2.5 cm & $\frac{640\,\text{pixels}] / 12\,\text{m}}{\Delta r} = \frac{4}{3}$ \\
\bf LCA     & $\beta \in [0,10]$ (9 values) & $\phi \in [-1.07,1.07]$ deg (9 values) & Navg = 1 \\
\bf Capon   & $\Delta$ = 0.01                 & L = 16                           & Navg = 1 \\
\end{tabular}
\subfloat[LCA Window Response]{\includegraphics[width=0.49\linewidth]{gfx/7_window_response.png}}\hfill
\subfloat[Mean images]{\includegraphics[width=0.49\linewidth]{gfx/7_mean_imgs.png}}\\
\subfloat[Windows ($\beta$)]{\includegraphics[width=0.49\linewidth]{gfx/7_windows_beta.png}}\hfill
\subfloat[Windows ($\phi$)]{\includegraphics[width=0.48\linewidth]{gfx/7_windows_phi.png}}\\
\subfloat[Capon win. resp. through shadow]{\includegraphics[width=0.49\linewidth]{gfx/7_win_resp_cut_shadow.png}}\hfill
\subfloat[Capon win. resp. through highlight]{\includegraphics[width=0.49\linewidth]{gfx/7_win_resp_cut_highlight.png}}\\
\end{narrow}
\end{figure*}
\newpage
\begin{figure*}[!t]
\begin{narrow}{-1.2cm}{-1.2cm}\centering\vspace{-1.0cm}
\textbf{8. Capon: Tuning time averaging.}\\
\begin{tabular}[c]{l l l l}
\bf General & M = 32                            & $\Delta r = \frac{c}{2B}$ = 2.5 cm & $\frac{640\,\text{pixels}] / 12\,\text{m}}{\Delta r} = \frac{4}{3}$ \\
\bf LCA     & $\beta \in [0,10]$ (9 values) & $\phi \in [-1.07,1.07]$ deg (9 values) & Navg = 3 \\
\bf Capon   & $\Delta$ = 0.01                 & L = 16                           & Navg = 3 \\
\end{tabular}
\subfloat[LCA Window Response]{\includegraphics[width=0.49\linewidth]{gfx/8_window_response.png}}\hfill
\subfloat[Mean images]{\includegraphics[width=0.49\linewidth]{gfx/8_mean_imgs.png}}\\
\subfloat[Windows ($\beta$)]{\includegraphics[width=0.49\linewidth]{gfx/8_windows_beta.png}}\hfill
\subfloat[Windows ($\phi$)]{\includegraphics[width=0.48\linewidth]{gfx/8_windows_phi.png}}\\
\subfloat[Capon win. resp. through shadow]{\includegraphics[width=0.49\linewidth]{gfx/8_win_resp_cut_shadow.png}}\hfill
\subfloat[Capon win. resp. through highlight]{\includegraphics[width=0.49\linewidth]{gfx/8_win_resp_cut_highlight.png}}\\
\end{narrow}
\end{figure*}
\newpage
\begin{figure*}[!t]
\begin{narrow}{-1.2cm}{-1.2cm}\centering\vspace{-1.0cm}
\textbf{9. Capon: Tuning time averaging.}\\
\begin{tabular}[c]{l l l l}
\bf General & M = 32                            & $\Delta r = \frac{c}{2B}$ = 2.5 cm & $\frac{640\,\text{pixels}] / 12\,\text{m}}{\Delta r} = \frac{4}{3}$ \\
\bf LCA     & $\beta \in [0,10]$ (9 values) & $\phi \in [-1.07,1.07]$ deg (9 values) & Navg = 5 \\
\bf Capon   & $\Delta$ = 0.01                 & L = 16                           & Navg = 5 \\
\end{tabular}
\subfloat[LCA Window Response]{\includegraphics[width=0.49\linewidth]{gfx/9_window_response.png}}\hfill
\subfloat[Mean images]{\includegraphics[width=0.49\linewidth]{gfx/9_mean_imgs.png}}\\
\subfloat[Windows ($\beta$)]{\includegraphics[width=0.49\linewidth]{gfx/9_windows_beta.png}}\hfill
\subfloat[Windows ($\phi$)]{\includegraphics[width=0.48\linewidth]{gfx/9_windows_phi.png}}\\
\subfloat[Capon win. resp. through shadow]{\includegraphics[width=0.49\linewidth]{gfx/9_win_resp_cut_shadow.png}}\hfill
\subfloat[Capon win. resp. through highlight]{\includegraphics[width=0.49\linewidth]{gfx/9_win_resp_cut_highlight.png}}\\
\end{narrow}
\end{figure*}
\newpage
\begin{figure*}[!t]
\begin{narrow}{-1.2cm}{-1.2cm}\centering\vspace{-1.0cm}
\textbf{10. Capon: Tuning time averaging.}\\
\begin{tabular}[c]{l l l l}
\bf General & M = 32                            & $\Delta r = \frac{c}{2B}$ = 2.5 cm & $\frac{640\,\text{pixels}] / 12\,\text{m}}{\Delta r} = \frac{4}{3}$ \\
\bf LCA     & $\beta \in [0,10]$ (9 values) & $\phi \in [-1.07,1.07]$ deg (9 values) & Navg = 7 \\
\bf Capon   & $\Delta$ = 0.01                 & L = 16                           & Navg = 7 \\
\end{tabular}
\subfloat[LCA Window Response]{\includegraphics[width=0.49\linewidth]{gfx/10_window_response.png}}\hfill
\subfloat[Mean images]{\includegraphics[width=0.49\linewidth]{gfx/10_mean_imgs.png}}\\
\subfloat[Windows ($\beta$)]{\includegraphics[width=0.49\linewidth]{gfx/10_windows_beta.png}}\hfill
\subfloat[Windows ($\phi$)]{\includegraphics[width=0.48\linewidth]{gfx/10_windows_phi.png}}\\
\subfloat[Capon win. resp. through shadow]{\includegraphics[width=0.49\linewidth]{gfx/10_win_resp_cut_shadow.png}}\hfill
\subfloat[Capon win. resp. through highlight]{\includegraphics[width=0.49\linewidth]{gfx/10_win_resp_cut_highlight.png}}\\
\end{narrow}
\end{figure*}
\newpage
\begin{figure*}[!t]
\begin{narrow}{-1.2cm}{-1.2cm}\centering\vspace{-1.0cm}
\textbf{11. LCA: Reasonable settings.}\\
\begin{tabular}[c]{l l l l}
\bf General & M = 32                            & $\Delta r = \frac{c}{2B}$ = 2.5 cm & $\frac{640\,\text{pixels}] / 12\,\text{m}}{\Delta r} = \frac{4}{3}$ \\
\bf LCA     & $\beta \in [0,10]$ (9 values) & $\phi \in [-0.57,0.57]$ deg (9 values) & Navg = 7 \\
\bf Capon   & $\Delta$ = 0.01                 & L = 16                           & Navg = 7 \\
\end{tabular}
\subfloat[LCA Window Response]{\includegraphics[width=0.49\linewidth]{gfx/11_window_response.png}}\hfill
\subfloat[Mean images]{\includegraphics[width=0.49\linewidth]{gfx/11_mean_imgs.png}}\\
\subfloat[Windows ($\beta$)]{\includegraphics[width=0.49\linewidth]{gfx/11_windows_beta.png}}\hfill
\subfloat[Windows ($\phi$)]{\includegraphics[width=0.48\linewidth]{gfx/11_windows_phi.png}}\\
\subfloat[Capon win. resp. through shadow]{\includegraphics[width=0.49\linewidth]{gfx/11_win_resp_cut_shadow.png}}\hfill
\subfloat[Capon win. resp. through highlight]{\includegraphics[width=0.49\linewidth]{gfx/11_win_resp_cut_highlight.png}}\\
\end{narrow}
\end{figure*}
\newpage
\begin{figure*}[!t]
\begin{narrow}{-1.2cm}{-1.2cm}\centering\vspace{-1.0cm}
\textbf{12. LCA: Lots of Kaiser windows - wide span.}\\
\begin{tabular}[c]{l l l l}
\bf General & M = 32                            & $\Delta r = \frac{c}{2B}$ = 2.5 cm & $\frac{640\,\text{pixels}] / 12\,\text{m}}{\Delta r} = \frac{4}{3}$ \\
\bf LCA     & $\beta \in [0,20]$ (19 values) & $\phi \in [-1.07,1.07]$ deg (19 values) & Navg = 7 \\
\bf Capon   & $\Delta$ = 0.01                 & L = 16                           & Navg = 7 \\
\end{tabular}
\subfloat[LCA Window Response]{\includegraphics[width=0.49\linewidth]{gfx/12_window_response.png}}\hfill
\subfloat[Mean images]{\includegraphics[width=0.49\linewidth]{gfx/12_mean_imgs.png}}\\
\subfloat[Windows ($\beta$)]{\includegraphics[width=0.49\linewidth]{gfx/12_windows_beta.png}}\hfill
\subfloat[Windows ($\phi$)]{\includegraphics[width=0.48\linewidth]{gfx/12_windows_phi.png}}\\
\subfloat[Capon win. resp. through shadow]{\includegraphics[width=0.49\linewidth]{gfx/12_win_resp_cut_shadow.png}}\hfill
\subfloat[Capon win. resp. through highlight]{\includegraphics[width=0.49\linewidth]{gfx/12_win_resp_cut_highlight.png}}\\
\end{narrow}
\end{figure*}
\newpage
\begin{figure*}[!t]
\begin{narrow}{-1.2cm}{-1.2cm}\centering\vspace{-1.0cm}
\textbf{13. LCA: Lots of trigonometric windows - wide span.}\\
\begin{tabular}[c]{l l l l}
\bf General & M = 32                            & $\Delta r = \frac{c}{2B}$ = 2.5 cm & $\frac{640\,\text{pixels}] / 12\,\text{m}}{\Delta r} = \frac{4}{3}$ \\
\bf LCA     & $\beta \in [0,20]$ (19 values) & $\phi \in [-1.07,1.07]$ deg (19 values) & Navg = 7 \\
\bf Capon   & $\Delta$ = 0.01                 & L = 16                           & Navg = 7 \\
\end{tabular}
\subfloat[LCA Window Response]{\includegraphics[width=0.49\linewidth]{gfx/13_window_response.png}}\hfill
\subfloat[Mean images]{\includegraphics[width=0.49\linewidth]{gfx/13_mean_imgs.png}}\\
\subfloat[Windows ($\beta$)]{\includegraphics[width=0.49\linewidth]{gfx/13_windows_beta.png}}\hfill
\subfloat[Windows ($\phi$)]{\includegraphics[width=0.48\linewidth]{gfx/13_windows_phi.png}}\\
\subfloat[Capon win. resp. through shadow]{\includegraphics[width=0.49\linewidth]{gfx/13_win_resp_cut_shadow.png}}\hfill
\subfloat[Capon win. resp. through highlight]{\includegraphics[width=0.49\linewidth]{gfx/13_win_resp_cut_highlight.png}}\\
\end{narrow}
\end{figure*}
\newpage
\begin{figure*}[!t]
\begin{narrow}{-1.2cm}{-1.2cm}\centering\vspace{-1.0cm}
\textbf{14. LCA: Back on Kaiser - reducing steering span.}\\
\begin{tabular}[c]{l l l l}
\bf General & M = 32                            & $\Delta r = \frac{c}{2B}$ = 2.5 cm & $\frac{640\,\text{pixels}] / 12\,\text{m}}{\Delta r} = \frac{4}{3}$ \\
\bf LCA     & $\beta \in [0,20]$ (19 values) & $\phi \in [-0.72,0.72]$ deg (19 values) & Navg = 7 \\
\bf Capon   & $\Delta$ = 0.01                 & L = 16                           & Navg = 7 \\
\end{tabular}
\subfloat[LCA Window Response]{\includegraphics[width=0.49\linewidth]{gfx/14_window_response.png}}\hfill
\subfloat[Mean images]{\includegraphics[width=0.49\linewidth]{gfx/14_mean_imgs.png}}\\
\subfloat[Windows ($\beta$)]{\includegraphics[width=0.49\linewidth]{gfx/14_windows_beta.png}}\hfill
\subfloat[Windows ($\phi$)]{\includegraphics[width=0.48\linewidth]{gfx/14_windows_phi.png}}\\
\subfloat[Capon win. resp. through shadow]{\includegraphics[width=0.49\linewidth]{gfx/14_win_resp_cut_shadow.png}}\hfill
\subfloat[Capon win. resp. through highlight]{\includegraphics[width=0.49\linewidth]{gfx/14_win_resp_cut_highlight.png}}\\
\end{narrow}
\end{figure*}
\newpage
\begin{figure*}[!t]
\begin{narrow}{-1.2cm}{-1.2cm}\centering\vspace{-1.0cm}
\textbf{15. LCA: Reducing steering span.}\\
\begin{tabular}[c]{l l l l}
\bf General & M = 32                            & $\Delta r = \frac{c}{2B}$ = 2.5 cm & $\frac{640\,\text{pixels}] / 12\,\text{m}}{\Delta r} = \frac{4}{3}$ \\
\bf LCA     & $\beta \in [0,20]$ (19 values) & $\phi \in [-0.36,0.36]$ deg (19 values) & Navg = 7 \\
\bf Capon   & $\Delta$ = 0.01                 & L = 16                           & Navg = 7 \\
\end{tabular}
\subfloat[LCA Window Response]{\includegraphics[width=0.49\linewidth]{gfx/15_window_response.png}}\hfill
\subfloat[Mean images]{\includegraphics[width=0.49\linewidth]{gfx/15_mean_imgs.png}}\\
\subfloat[Windows ($\beta$)]{\includegraphics[width=0.49\linewidth]{gfx/15_windows_beta.png}}\hfill
\subfloat[Windows ($\phi$)]{\includegraphics[width=0.48\linewidth]{gfx/15_windows_phi.png}}\\
\subfloat[Capon win. resp. through shadow]{\includegraphics[width=0.49\linewidth]{gfx/15_win_resp_cut_shadow.png}}\hfill
\subfloat[Capon win. resp. through highlight]{\includegraphics[width=0.49\linewidth]{gfx/15_win_resp_cut_highlight.png}}\\
\end{narrow}
\end{figure*}
\newpage
\begin{figure*}[!t]
\begin{narrow}{-1.2cm}{-1.2cm}\centering\vspace{-1.0cm}
\textbf{16. LCA: Adjusting resolution/SNR.}\\
\begin{tabular}[c]{l l l l}
\bf General & M = 32                            & $\Delta r = \frac{c}{2B}$ = 2.5 cm & $\frac{640\,\text{pixels}] / 12\,\text{m}}{\Delta r} = \frac{4}{3}$ \\
\bf LCA     & $\beta \in [0,5]$ (19 values) & $\phi \in [-0.72,0.72]$ deg (19 values) & Navg = 7 \\
\bf Capon   & $\Delta$ = 0.01                 & L = 16                           & Navg = 7 \\
\end{tabular}
\subfloat[LCA Window Response]{\includegraphics[width=0.49\linewidth]{gfx/16_window_response.png}}\hfill
\subfloat[Mean images]{\includegraphics[width=0.49\linewidth]{gfx/16_mean_imgs.png}}\\
\subfloat[Windows ($\beta$)]{\includegraphics[width=0.49\linewidth]{gfx/16_windows_beta.png}}\hfill
\subfloat[Windows ($\phi$)]{\includegraphics[width=0.48\linewidth]{gfx/16_windows_phi.png}}\\
\subfloat[Capon win. resp. through shadow]{\includegraphics[width=0.49\linewidth]{gfx/16_win_resp_cut_shadow.png}}\hfill
\subfloat[Capon win. resp. through highlight]{\includegraphics[width=0.49\linewidth]{gfx/16_win_resp_cut_highlight.png}}\\
\end{narrow}
\end{figure*}
\newpage
\begin{figure*}[!t]
\begin{narrow}{-1.2cm}{-1.2cm}\centering\vspace{-1.0cm}
\textbf{17. LCA: Adjusting resolution/SNR.}\\
\begin{tabular}[c]{l l l l}
\bf General & M = 32                            & $\Delta r = \frac{c}{2B}$ = 2.5 cm & $\frac{640\,\text{pixels}] / 12\,\text{m}}{\Delta r} = \frac{4}{3}$ \\
\bf LCA     & $\beta \in [5,20]$ (19 values) & $\phi \in [-0.72,0.72]$ deg (19 values) & Navg = 7 \\
\bf Capon   & $\Delta$ = 0.01                 & L = 16                           & Navg = 7 \\
\end{tabular}
\subfloat[LCA Window Response]{\includegraphics[width=0.49\linewidth]{gfx/17_window_response.png}}\hfill
\subfloat[Mean images]{\includegraphics[width=0.49\linewidth]{gfx/17_mean_imgs.png}}\\
\subfloat[Windows ($\beta$)]{\includegraphics[width=0.49\linewidth]{gfx/17_windows_beta.png}}\hfill
\subfloat[Windows ($\phi$)]{\includegraphics[width=0.48\linewidth]{gfx/17_windows_phi.png}}\\
\subfloat[Capon win. resp. through shadow]{\includegraphics[width=0.49\linewidth]{gfx/17_win_resp_cut_shadow.png}}\hfill
\subfloat[Capon win. resp. through highlight]{\includegraphics[width=0.49\linewidth]{gfx/17_win_resp_cut_highlight.png}}\\
\end{narrow}
\end{figure*}
\newpage
\begin{figure*}[!t]
\begin{narrow}{-1.2cm}{-1.2cm}\centering\vspace{-1.0cm}
\textbf{18. LCA: Fewer windows.}\\
\begin{tabular}[c]{l l l l}
\bf General & M = 32                            & $\Delta r = \frac{c}{2B}$ = 2.5 cm & $\frac{640\,\text{pixels}] / 12\,\text{m}}{\Delta r} = \frac{4}{3}$ \\
\bf LCA     & $\beta \in [0,10]$ (9 values) & $\phi \in [-0.72,0.72]$ deg (19 values) & Navg = 7 \\
\bf Capon   & $\Delta$ = 0.01                 & L = 16                           & Navg = 7 \\
\end{tabular}
\subfloat[LCA Window Response]{\includegraphics[width=0.49\linewidth]{gfx/18_window_response.png}}\hfill
\subfloat[Mean images]{\includegraphics[width=0.49\linewidth]{gfx/18_mean_imgs.png}}\\
\subfloat[Windows ($\beta$)]{\includegraphics[width=0.49\linewidth]{gfx/18_windows_beta.png}}\hfill
\subfloat[Windows ($\phi$)]{\includegraphics[width=0.48\linewidth]{gfx/18_windows_phi.png}}\\
\subfloat[Capon win. resp. through shadow]{\includegraphics[width=0.49\linewidth]{gfx/18_win_resp_cut_shadow.png}}\hfill
\subfloat[Capon win. resp. through highlight]{\includegraphics[width=0.49\linewidth]{gfx/18_win_resp_cut_highlight.png}}\\
\end{narrow}
\end{figure*}
\newpage
\begin{figure*}[!t]
\begin{narrow}{-1.2cm}{-1.2cm}\centering\vspace{-1.0cm}
\textbf{19. LCA: Fewer steering angles.}\\
\begin{tabular}[c]{l l l l}
\bf General & M = 32                            & $\Delta r = \frac{c}{2B}$ = 2.5 cm & $\frac{640\,\text{pixels}] / 12\,\text{m}}{\Delta r} = \frac{4}{3}$ \\
\bf LCA     & $\beta \in [0,10]$ (19 values) & $\phi \in [-0.72,0.72]$ deg (9 values) & Navg = 7 \\
\bf Capon   & $\Delta$ = 0.01                 & L = 16                           & Navg = 7 \\
\end{tabular}
\subfloat[LCA Window Response]{\includegraphics[width=0.49\linewidth]{gfx/19_window_response.png}}\hfill
\subfloat[Mean images]{\includegraphics[width=0.49\linewidth]{gfx/19_mean_imgs.png}}\\
\subfloat[Windows ($\beta$)]{\includegraphics[width=0.49\linewidth]{gfx/19_windows_beta.png}}\hfill
\subfloat[Windows ($\phi$)]{\includegraphics[width=0.48\linewidth]{gfx/19_windows_phi.png}}\\
\subfloat[Capon win. resp. through shadow]{\includegraphics[width=0.49\linewidth]{gfx/19_win_resp_cut_shadow.png}}\hfill
\subfloat[Capon win. resp. through highlight]{\includegraphics[width=0.49\linewidth]{gfx/19_win_resp_cut_highlight.png}}\\
\end{narrow}
\end{figure*}
\newpage
\begin{figure*}[!t]
\begin{narrow}{-1.2cm}{-1.2cm}\centering\vspace{-1.0cm}
\textbf{20. LCA: Fewer both.}\\
\begin{tabular}[c]{l l l l}
\bf General & M = 32                            & $\Delta r = \frac{c}{2B}$ = 2.5 cm & $\frac{640\,\text{pixels}] / 12\,\text{m}}{\Delta r} = \frac{4}{3}$ \\
\bf LCA     & $\beta \in [0,10]$ (9 values) & $\phi \in [-0.72,0.72]$ deg (9 values) & Navg = 7 \\
\bf Capon   & $\Delta$ = 0.01                 & L = 16                           & Navg = 7 \\
\end{tabular}
\subfloat[LCA Window Response]{\includegraphics[width=0.49\linewidth]{gfx/20_window_response.png}}\hfill
\subfloat[Mean images]{\includegraphics[width=0.49\linewidth]{gfx/20_mean_imgs.png}}\\
\subfloat[Windows ($\beta$)]{\includegraphics[width=0.49\linewidth]{gfx/20_windows_beta.png}}\hfill
\subfloat[Windows ($\phi$)]{\includegraphics[width=0.48\linewidth]{gfx/20_windows_phi.png}}\\
\subfloat[Capon win. resp. through shadow]{\includegraphics[width=0.49\linewidth]{gfx/20_win_resp_cut_shadow.png}}\hfill
\subfloat[Capon win. resp. through highlight]{\includegraphics[width=0.49\linewidth]{gfx/20_win_resp_cut_highlight.png}}\\
\end{narrow}
\end{figure*}
\newpage
\begin{figure*}[!t]
\begin{narrow}{-1.2cm}{-1.2cm}\centering\vspace{-1.0cm}
\textbf{21. LCA: Even less.}\\
\begin{tabular}[c]{l l l l}
\bf General & M = 32                            & $\Delta r = \frac{c}{2B}$ = 2.5 cm & $\frac{640\,\text{pixels}] / 12\,\text{m}}{\Delta r} = \frac{4}{3}$ \\
\bf LCA     & $\beta \in [0,5]$ (9 values) & $\phi \in [-0.72,0.72]$ deg (5 values) & Navg = 7 \\
\bf Capon   & $\Delta$ = 0.01                 & L = 16                           & Navg = 7 \\
\end{tabular}
\subfloat[LCA Window Response]{\includegraphics[width=0.49\linewidth]{gfx/21_window_response.png}}\hfill
\subfloat[Mean images]{\includegraphics[width=0.49\linewidth]{gfx/21_mean_imgs.png}}\\
\subfloat[Windows ($\beta$)]{\includegraphics[width=0.49\linewidth]{gfx/21_windows_beta.png}}\hfill
\subfloat[Windows ($\phi$)]{\includegraphics[width=0.48\linewidth]{gfx/21_windows_phi.png}}\\
\subfloat[Capon win. resp. through shadow]{\includegraphics[width=0.49\linewidth]{gfx/21_win_resp_cut_shadow.png}}\hfill
\subfloat[Capon win. resp. through highlight]{\includegraphics[width=0.49\linewidth]{gfx/21_win_resp_cut_highlight.png}}\\
\end{narrow}
\end{figure*}
\newpage
\begin{figure*}[!t]
\begin{narrow}{-1.2cm}{-1.2cm}\centering\vspace{-1.0cm}
\textbf{22. LCA: Hardly any.}\\
\begin{tabular}[c]{l l l l}
\bf General & M = 32                            & $\Delta r = \frac{c}{2B}$ = 2.5 cm & $\frac{640\,\text{pixels}] / 12\,\text{m}}{\Delta r} = \frac{4}{3}$ \\
\bf LCA     & $\beta \in [0,5]$ (5 values) & $\phi \in [-0.72,0.72]$ deg (5 values) & Navg = 7 \\
\bf Capon   & $\Delta$ = 0.01                 & L = 16                           & Navg = 7 \\
\end{tabular}
\subfloat[LCA Window Response]{\includegraphics[width=0.49\linewidth]{gfx/22_window_response.png}}\hfill
\subfloat[Mean images]{\includegraphics[width=0.49\linewidth]{gfx/22_mean_imgs.png}}\\
\subfloat[Windows ($\beta$)]{\includegraphics[width=0.49\linewidth]{gfx/22_windows_beta.png}}\hfill
\subfloat[Windows ($\phi$)]{\includegraphics[width=0.48\linewidth]{gfx/22_windows_phi.png}}\\
\subfloat[Capon win. resp. through shadow]{\includegraphics[width=0.49\linewidth]{gfx/22_win_resp_cut_shadow.png}}\hfill
\subfloat[Capon win. resp. through highlight]{\includegraphics[width=0.49\linewidth]{gfx/22_win_resp_cut_highlight.png}}\\
\end{narrow}
\end{figure*}

### 2018-02-06 LCAforActiveSonarImaging/lca_submit1.tex
Major commit before messing with PhD thesis

--- 2015-10-21 LCAforActiveSonarImaging/lca.tex
@@
 
+\section{Steering angle}
+
 \section{Reducing LCA complexity}\label{sec:reducing_lca_complexity}
@@
 The solution for $\alpha$ would also yield the optimal beamformer output in the minimum variance sense.
+
+\section{Multimedia Files and Source Code}\label{source_code}
 

### 2018-02-06 LCAforActiveSonarImaging/lca_submit2.tex
Major commit before messing with PhD thesis

--- 2015-12-17 LCAforActiveSonarImaging/lca.tex
@@
 LCA can be viewed as either a simplification of the minimum variance distortionless response (MVDR) beamformer, or as an adaptive extension to the delay and sum (DAS) beamformer.
-While both LCA and MVDR attempt to minimize the power of noise and interference in the image, MVDR seeks the optimal solution by estimating and inverting the array's spatial covariance matrix, while LCA selects the best performer from a predefined set of sensor windows.
-
-We show that a robust MVDR implementation typically creates windows spanning between the rectangular and Hamming.
+While both LCA and MVDR attempt to minimize the power of noise and interference in the image, MVDR achieves this by computing optimal array weights from the spatial statistics of the wavefield, while LCA selects the best performing weights out of a predefined set.
+
+To build confidence in the LCA method we show that a robust MVDR implementation typically creates weights spanning between the rectangular and Hamming window function.
 We let LCA select from a set of Kaiser windows with responses in this span, and add some steered variations of each.
-We limit the steering to roughly half the -3\,dB width of the window's amplitude response.
-Using experimental data from the Kongsberg Maritime HISAS 1030 sonar we find that LCA and MVDR produce nearly identical images of large scenes, both being superior to DAS.
+We limit the steering to roughly half the \minus{}3\,dB width of the window's amplitude response.
+Using experimental data from the Kongsberg Maritime HISAS1030 sonar we find that LCA and MVDR produce nearly identical images of large scenes, both being superior to DAS.
 On point targets LCA is able to double the resolution compared to DAS, or provide half that of MVDR.
-This performance is achieved with a total of 6 windows; the rectangular and Kaiser window with $\beta=5$, in unsteered version, and left and right steered to the steering limit.
+This performance is achieved with a total of 6 windows; the rectangular window and the Kaiser window with $\beta=5$, in an unsteered version, and versions that are left and right steered to the steering limit.
 Slightly smoother images are produced if the window count is increased to 15, but past this we observe minimal difference.
@@
 
-\IEEEPARstart{T}{he} best adaptive sonars are found in nature.
+% The promise is already demonstrated in nature.
+% 
+% The sonar systems of the future 
+% 
+% Future sonar systems 
+% 
+% Great promise in adaptivity
+% 
+% The greatest untapped 
+% 
+% The sonars of the future might benefit greatly from adaptivity.
+% 
+% 
+% Future sonars will likely be
+
+% To make a sonar truly perform better, we need to make it smarter. 
+% 
+% \IEEEPARstart{F}{uture} sonar systems will merit a much more sophisticated level of adaptivity. 
+% 
+% \IEEEPARstart{T}{he} next generation sonar systems will be adaptive.
+% 
+% Making sonars adaptive is the way to take them into the future. 
+% 
+% The virtue of future sonar systems will be sophisticated adaptivity. 
+% 
+% Intelligent
+% 
+% Reactive
+% 
+% Observe, process, adapt, act. 
+% 
+% Adaptivity - sense, process, respond
+% 
+% Faster, smarter
+% 
+% The best sonars in nature are adaptive. 
+% Nature allures us to be this bold because
+
+\IEEEPARstart{T}{he} best sonars in existence is perhaps found in nature.
 Bats, for instance, use a high frequency sonar to detect, identify and track prey with amazing precision in caves amongst a myriad of other bats.
-Active human made sonars have little to no adaptability on transmit, but several methods have been investigated to adaptively post process the received data.
-Perhaps the best studied method for reconstructing a sonar image is the minimum variance distortion-less response (MVDR) beamformer~\cite{Capon1969}.
+No man made sonar can currenly match this feat, much due to the challenge of processing and adapting to such a complex environment fast enough.
+This is why human made sonars have little to no adaptivity on transmit.
+However, several methods have been investigated to adaptively post process the received data.
+Perhaps the best studied method for reconstructing a sonar image out of the data is the minimum variance distortion-less response (MVDR) beamformer~\cite{Capon1969}.
 It adapts the sonar array's spatial response to minimize the influence of noise and interference in the final image.
@@
 This is because MVDR relies on estimating and inverting a covariance matrix.
-In sonar systems with less than 32 channels the complexity can be reduced and the implementation accelerated significantly using graphics computing units (GPUs)~\cite{Buskenes2014}.
-For larger systems the inversion step dominates the computation.
+The estimation step dominates the computation in sonar systems with less than 32 channels, but here the computational complexity can be reduced and the implementation accelerated significantly using graphics computing units (GPUs)~\cite{Buskenes2014}.
+For larger systems the inversion step dominates.
 This can be dealt with by approximating the full covariance matrix with a small one using space reduction techniques such as beamspace processing~\cite{Asen2013}, or the closely related principal component analysis (PCA) method~\cite{Kim2014}.
@@
 Based on an idea by Vignon~\cite{Vignon2008}, it was first introduced by Synnev\aa{}g \emph{et al}.
-in clinical medical imaging~\cite{Synnevag2008}, who demonstrated its ability to obtain very similar image quality to MVDR in systems with focused transmit beams.
-LCA applies a set of predefined windows, or apodization weights, and selects the one offering the best sidelobe suppression.
-It relies on the same optimization criterion as MVDR, thus may be considered a version of MVDR with a reduced window solution space.
+in clinical medical ultrasound imaging~\cite{Synnevag2011}, who demonstrated its ability to obtain very similar image quality to MVDR in systems with focused transmit beams.
+LCA applies a set of predefined and static windows, or apodization weights, and selects the one offering the best noise suppression.
+It relies on the same optimization criterion as MVDR, thus may be considered a version of MVDR with a discrete and static window solution space.
 Alternatively, it may be viewed as a multi-apodization technique.
@@
 in radar imaging~\cite{Stankwitz1995}, where the best out of 2 or 3 windows are selected.
-However, LCA is different in that it typically selects from a larger pool of windows and allows micro-steered variations of them.
-This gives LCA an adaptive freedom that is somewhere in between that of traditional multi-apodization techniques and MVDR.
+However, LCA is different in that it typically selects from a larger pool of windows and allows phase-steered variations of them.
+This gives LCA an adaptive freedom that lies somewhere in between that of traditional multi-apodization techniques and MVDR.
+How similarly LCA performs compared to MVDR depends on how well the predefined windows represent the constrained solution space of MVDR.
+
+% While using weights that span a bigger space gives LCA greater freedom, not The bigger the space the weights span, the greater the adaptive freedom, but only the solutions that are valid in the constrained solution space can make a difference.
+
+% Instead of adaptively computing the weights, it adaptively selects them.
 
@@
 suggested using a window set comprised of rectangular, Kaiser (or Kaiser-Harris) and inverted Kaiser functions.
-In total he made LCA choose from 12 unique windows, 6 of which were micro-steered variants of a fairly wide Kaiser window, and with the rectangular window included.
-This was motivated by the desire to use wide windows with maximum sidelobe suppression near the receiver, and narrow windows at greater depths for maximum sensitivity and depth penetration.
-While achieving good results, it was not made clear how how many windows LCA needs, or how they ideally should be steered.
+In total he made LCA choose from 12 unique windows, 6 of which were phase-shifted versions of a fairly wide Kaiser window, and one were rectangular.
+Phase-shifting the windows has the effect of slighly adjusting the angle the array is steered towards.
+Synnev\aa{}g's composition of windows was motivated by the desire to use wide windows with maximum sidelobe suppression near the receiver, and narrow windows at greater depths for maximum sensitivity and depth penetration.
+While achieving good results, it was not made clear how many windows LCA needs, or how they ideally should be steered.
 Also, it is not apparent how LCA performs in sonar systems where the transmit beam is unfocused.
@@
 We extend Synnev\aa{}g's work by steering all our windows and not just some, and demonstrate how LCA's performance is affected by the type and number of windows included.
-Our results are obtained using experimental data from the 32 element Kongsberg Maritime HISAS 1030 synthetic aperture sonar (SAS), which operates at 100\,kHz with an element beamwidth of 23$^\circ$.
-
-Our results show that when MVDR is modified to work in active sonar imaging, it computes weights with spatial responses that tend to be symmetric and steered within a fraction of the window's -\,3dB resolution.
+Our results are obtained using experimental data from the 32 element Kongsberg Maritime HISAS1030 sonar, which operates at 100\,kHz with an \minus{}3\,dB element opening angle of 23$^\circ$.
+Although this system was designed as synthetic aperture sonar (SAS) where an image is synthesized from several pings of data, we will only create images out of single pings (sectorscan imaging) in this study.
+
+Our results show that when MVDR is modified to work in active sonar imaging, it computes weights with spatial responses that tend to be symmetric and steered within a fraction of the window's \minus{}3\,dB resolution.
 When letting LCA choose from Kaiser windows with similar responses, we obtain images with improved noise suppression and resolution.
-The resolution gain is made predictable by limiting the maximum steering to a fraction of the window's -3\,dB width.
+The resolution gain is made predictable by limiting the maximum steering to a fraction of the window's \minus{}3\,dB width.
 Overall, LCA produces images that are similar to MVDR, but with a resolution in between that of DAS and MVDR.
@@
 A weight set is commonly called a window, apodization or taper function.
-Applying a window that trails off towards the edges improves noise suppression at the cost of reduced resolution~\cite{Harris1978}.
-Real windows lead to symmetric responses, while complex weights allow asymmetric responses.
-
-Conventional beamformers all have static weights.
+When the window is real its response is symmetric.
+Applying a real window that trails off towards the edges creates a response with lower sidelobe levels and wider mainlobe, which translates into improved noise suppression at the cost of reduced resolution, respectively~\cite{Harris1978}.
+
+Conventional beamformers all have static and usually real weights.
 The reference method is the delay-and-sum (DAS) beamformer, also know as the backprojection algorithm.
 It delays each pixel into focus, then applies a suitable window, and finally sums the data.
-The virtue of DAS is its simplicity, robustness to parameter errors, linear processing of the image and the ease of which it can be implemented on parallel hardware.
+The virtue of DAS is its simplicity, robustness to parameter errors, linear processing of the image and the ease of which it can be implemented in parallel hardware.
 
@@
 \ifPeerReview%
-\includegraphics[width=.5\linewidth]{gfx/calc_kaiser_3dB.eps}%
+\includegraphics[width=.5\linewidth]{gfx/buske1_online.eps}%
 \else%
-\includegraphics[width=\linewidth]{gfx/calc_kaiser_3dB.eps}%
+\includegraphics[width=\linewidth]{gfx/buske1_online.eps}%
 \fi%
-\caption{Each Kaiser window is steered in the interval $\phi=[0, \phi_{\text{-3dB}}(\beta)]$.
-The angle $\phi_{\text{-3dB}}(\beta)$ is the amount of steering needed for the steered response to have a -3dB crossing that is exactly half that of the unsteered window.
+\caption{Each Kaiser window is steered in the interval $\phi=[0, \phi_{\mathrm{3\,dB}}(\beta)]$.
+The angle $\phi_{\mathrm{3\,dB}}(\beta)$ is the amount of steering needed for the steered response to have a \protect\minus{}3\,dB crossing that is exactly half that of the unsteered window.
 With each window steered this way we expect the resolution gain to remain predictable and independent of $\beta$, and we also effectively constrain the white noise gain of the beamformer.}\label{windows_steering}
@@
 \emph{Right:}\hfill
-\parbox[t]{.95\linewidth}{MVDR window amplitude and phase responses computed from 40x40 pixel groups in the shadow, highlight and speckle region of the image.
+\parbox[t]{.95\linewidth}{MVDR window amplitude and phase responses computed from 40x40 pixel groups for the shadow, highlight and speckle region of the image.
 The responses are overlayed each other and the amount of overlap is colored using a logarithmic scale.
+The size of the pixel groups were chosen ad-hoc for the histograms to be near invariant to a shift in position within the same region.
+Since each pixel is pre-delayed into focus the unsteered responses all have their center at broadside.
 The dashed red lines at -13\,dB and -43\,dB marks the peak sidelobe levels of an unsteered rectangular and Hamming window, respectively.
@@
 \begin{figure*}[t]%
-\includegraphics[width=\textwidth]{gfx/oversampling_mosaic_bounds.pdf}%
+\includegraphics[width=\textwidth]{gfx/buske4_online.pdf}%
 \caption{\emph{Determining Kaiser parameter boundaries}: LCA images of the resolution test cross and anchor created with a large and equal size window database, but where the upper bounds for $\beta$ and $\phi$ is varied.
 The upper left image is equal to DAS with a rectangular window.
-For all images we measure the lateral distance $\Delta x$ between the two -3dB points of the anchor (red object).
+For all images we measure the lateral distance $\Delta x$ between the two \protect\minus{}3\,dB points of the anchor (red object).
 These are specified relatively to the reference distance  $\Delta x_\text{ref}$ from the upper left DAS image.
@@
 where $N_x = 2X+1$ is the number of azimuth lines to average over, and $\omega[x',n']$ is a normalized 2 dimensional weight function.
-However, we observed no visual improvement doing this, and the effect of micro-steered windows is hard to predict.
-Henceforth no lateral averaging will be applied.
+Since we oversample slightly laterally when we delay each pixel (see Section \ref{sec:lca_oversampling}), we thought the lateral correlation to be sufficient to benefit from this.
+However, we observed no visual improvement from applying this technique, and decided not to use it.
 % 
@@
 Hence, when it is steered the white noise gain must increase and the the signal-to-noise ratio decrease, as shown in \Fig{windows_steering}.
-To limit this we place an upper bound to the steering, chosen such that wide windows is allowed to be steered more than narrow ones.
-We call this upper bound for steering $\phi_{\text{3dB}}(\beta)$, and define it as the steering angle needed for the steered response to have a \minus{}3\,dB crossing that is exactly half that of the unsteered window.
-Using a numerical search for the steering angle $\phi$ that fulfilled this criterion for a wide range of $\beta$-values, we obtained a third degree polynomial fit to these points:
-%
-\begin{align}
-\phi_\text{3dB}(\beta) \approx \minus{}0.002894\beta^3 + 0.02451\beta^2 - 0.01308\beta + 0.2980\label{eq:steering_kaiser_3dB}
-\end{align}
-%
+To limit this we place an upper bound to the steering, chosen such that wide windows are allowed to be steered more than narrow ones.
+We call this upper bound for steering $\phi_{\mathrm{3\,dB}}(\beta)$, and define it as the steering angle needed for the steered response to have a \minus{}3\,dB crossing that is exactly half that of the unsteered window.
+The method we devised for finding it is described in Appendix \ref{appendix:steering_angles}.
+
 Henceforth any steering will be specified relative to this upper bound.
@@
 We consider an absolute minimum to be 2, due to the non-linear nature of delaying pixels that isn't strictly in the far field, and of displaying the absolute value of the pixels on a decibel scale.
-In \mediaI{} we visualize the effect of changing the lateral oversampling factor on a set of LCA images.
+In \mediaI{}\footnote{For media location refer to Appendix \ref{app:source_code}.} we visualize the effect of changing the lateral oversampling factor on a set of LCA images.
 Upon display the images are all bilinearly upinterpolated to the same size.
@@
 To test the performance of the MVDR and LCA beamformers, we have processed data acquired by the Kongsberg Maritime HUGIN AUV carrying their 32 element HISAS1030 sonar~\cite{Hansen2011}.
-It is a high resolution synthetic aperture sonar with 1.2\;m array length, 100\;kHz operating frequency, 30\;kHz bandwidth and 23$^\circ$ receiver element beamwidth.
+It is a high resolution sonar with 1.2\;m array length, 100\;kHz operating frequency, 30\;kHz bandwidth and 23$^\circ$ element \minus{}3\,dB opening angle.
 
@@
 
-\mediaII{} animates how \Fig{mvdr_selected_windows} changes as a function of the MVDR subarray length $L$ and temporal averaging $K$.
+\mediaII{}\footnote{For media location refer to Appendix \ref{app:source_code}.} animates how \Fig{mvdr_selected_windows} changes as a function of the MVDR subarray length $L$ and temporal averaging $K$.
 At $L=1$ the window responses are rectangular in all areas in the image.
@@
 
+\section{Steering angle}\label{appendix:steering_angles}
+
+\begin{figure}[t]
+\ifPeerReview\centering%
+\includegraphics[width=.5\linewidth]{gfx/buske8_online.eps}%
+\else%
+\includegraphics[width=\linewidth]{gfx/buske8_online.eps}%
+\fi%
+\caption{\emph{Steering angle needed to cut the \protect\minus{}3\,dB distance of the Kaiser window's spatial response in half}.
+The needed amount of steering depends on the number of channels $M$, the element spacing relative to the wavelength $\frac{d}{\lambda}$, and the Kaiser parameter $\beta$.
+Note how the angle scales near proportionally with $\frac{d}{\lambda}$.
+}\label{steering_angle_3dB}
+\end{figure}
+%
+One of the goals of this article was to characterize the windows used by the MVDR method, and then identify a subset of these that were suitable for use with the LCA method.
+We determined that the Kaiser function could fit the role of producing the relevant windows.
+The spatial response of a steered Kaiser window depends on the number of channels $M$, the element spacing relative to the wavelength $\frac{d}{\lambda}$, the Kaiser parameter $\beta$, and the amount of steering $\phi$.
+
+To control the resolution gain of LCA, we wanted to determine the extent that the windows need to be steered to cut the angle of the \minus{}3\,dB point of the window's spatial amplitude response by a fixed amount, say, by a factor 2 as reference.
+We computed this steering angle for some common configurations of the mentioned parameters.
+Refer to \Fig{steering_angle_3dB} for the result.
+The boundary values of the Kaiser window is shown, i.e.
+the rectangular window at $\beta=0$ and the near gaussian window at $\beta=5$, as well as one in between at $\beta=2.5$.
+Observe that for a given $M$ and $\beta$, the angle $\phi_\mathrm{3\,dB}$ scales near proportionally with $\frac{d}{\lambda}$.
+We found this to be true also for $\frac{d}{\lambda}\in\{0.5,4\}$.
+Hence, an approximate figure for a rather wide range of system parameters can be derived off this figure.
+
+For the sake of completeness we also supply the source code, see Appendix \ref{app:source_code}.
+
 \section{Reducing LCA complexity}\label{sec:reducing_lca_complexity}
@@
 The solution for $\alpha$ would also yield the optimal beamformer output in the minimum variance sense.
+
+\section{Multimedia Files and Source Code}\label{app:source_code}
+
+This article is accompanied by some multimedia files and source code for determining the steering angles for LCA's windows. They are located at:
+
+\url{\articlePath}
 
@@
 
-\bibitem{Synnevag2008}
-J.-F. Synnev\aa{}g, S.~Holm, and A.~Austeng, ``{A low complexity data-dependent
-  beamformer},'' in \emph{2008 IEEE Ultrasonics Symposium}, no.~2.\hskip 1em
-  plus 0.5em minus 0.4em\relax IEEE, Nov. 2008, pp. 1084--1087.
+\bibitem{Synnevag2011}
+J.-F. Synnev\aa{}g, A.~Austeng, and S.~Holm, ``{A low-complexity data-dependent
+  beamformer},'' \emph{IEEE Transactions on Ultrasonics, Ferroelectrics and
+  Frequency Control}, vol.~58, no.~2, pp. 281--289, feb 2011.
 

### 2018-02-06 LCAforActiveSonarImaging/submission/submit2/answer/answer.tex
Major commit before messing with PhD thesis

--- 2016-03-03 LCAforActiveSonarImaging/submission/submit1/answer/answer.tex
@@
 
-\newpage
-
 \section{Figure changes}
@@
 \begin{enumerate}
-\item We have supplied additional info on how the steering is affected by these parameters in appendix C.
+\item We have supplied additional info on how the steering is affected by these parameters in appendix A.
 \item Equations (12) and (14) only gave the correct results for our system configuration, and were not very intuitive.
-Appendix C now replace and extends these equations, and we removed them in hope this will make the article easier to read.
+Appendix C now replace and extends these equations, so removed them in hope this will make the article easier to read.
 \end{enumerate}
-\item Various spelling flaws attempted corrected.
+\item Various spelling flaws corrected.
 \item Appendix C added. Contains information of where source code and media files can be located. Before the article goes to publishing we will make sure that this link is permanent.
-\item Reference to Synnevåg's proceedings article from 2008, ``A low-complexity data-dependent beamformer'' replaced with his 2011 journal article with the same name.
-We never intended to reference the proceedings article since this work extends his journal version.
+\item Reference to Synnevåg's proceedings article from 2008, ``A low-complexity data-dependent beamformer'' is replaced with his 2011 journal article with the same name.
+This was an error from our side, we never intended to reference the proceedings article since this work extends his journal version.
 \end{enumerate}
@@
 That is, the images are from single ping, rather than synthesized from several pings.
-We agree this was worded in an ambiguous way and have now made relevant changes, as well as added a remark of what it meant by sectorscan.}
+We agree this was worded in an ambiguous way and have rewritten it for improved clarify, as well as elaborated on the meaning of sectorscan imaging.}
 %

### 2018-02-06 Simulator/article.bak.tex
Major commit before messing with PhD thesis

--- 2018-01-04 Simulator/article.tex
@@
 
+  % \newcommand\includegraphics[2][]{%
+  % \immediate\write18{./bin/laFigure #2 #1}%
+  % \input{result}}%
+
 % \DeclareMathOperator*{\argmin}{\text{arg}\;\text{min}}
@@
 
-% - Adapt simulation to its behaviour, see if they still match.
-
-\IEEEPARstart{A}{utonomous} underwater vehicles (AUVs) are becoming increasingly important in human underwater endeavors.
-When equipped with a synthetic aperture sonar (SAS), they can e.g.
-be used to map the seafloor, survey underwater biology, help localize wrecks, inspect oil pipes, fields and electricity lines, or hunt for mines and biohazard waste.
-Such independent systems have much to gain from being able to adapt to the scenario at hand.
-For instance, with a well designed automatic target recognition (ATR) system on board they can adapt their mission plan to collect more information on particularly interesting objects, or fine-tune the sonar and navigation system to improve imaging performance.
-
-% Automatic target recognition (ATR) is an important component in autonomous underwater vehicles (AUVs), as it allows the vehicle to adapt its mission plan to e.g.
-revisit detected objects for closer examination.
-One way to classify these objects is by using template matching.
-The principle involved  is to isolate an image segment containing the object of interest, compare it with a set of template images of the relevant object classes, and assign it to the class of the template with the the best fit.
-
-In conventional ATR an object of interest is first isolated, its properties attempted estimated, and a classifier used to determine what object class these properties belong to.
-The challenge is finding unique and intrinsic object properties, such as its type, shape and size, and to do so independently of extrinsic properties such as its alignment, burial depth, degeneration level or view angle.
-While SAS provide images with state of the art resolution and dynamic range, they still only represent an ambiguous acoustical 2D projection of the 3D object, littered with a highly coherent noise floor.
-
-An alternative approach is to use template matching, where the isolated image segment is compared to a set of template images covering the relevant object classes, and assigned to the class of the template with the best fit.
-Since the templates are computed from 3D models, this reduces the issue of image ambiguity.
-However, a challenge is finding a set of templates that represent the actual configurations of object and seafloor, Another issue is that the size of precomputed library of templates grows exponentially with the number of parameters needed.
-This limits its use to only a few parameters being coarsely sampled, ultimately yielding inaccurate results~\cite{Midelfart2010}.
-
-% We avoid the problems of conventional template matching by computing the templates on-the-fly as part of the classification process, thereby not needing a precomputed set of templates at all. 
-%  that avoids this problem by running sufficiently fast to create templates adapted to the actual scene as part of the classification process.
-% Classification process
-
-We tackle the problems of conventional template matching by performing:
-%
-\begin{itemize}
-\item \emph{Adaptive template matching}; where a template is selected based on object and seafloor parameters estimated from the SAS image~\cite{Midelfart2010}, and
-\item \emph{Real-time template simulation}; where a template is created tailored to the estimated parameters, improving its accuracy and avoiding the need of a precomputed template library.
-\end{itemize}
-%
-
-The methods presented herein will focus on the template simulator, but some details will also be given on adaptive template matching for completeness.
-More details on this technique can be found in~\cite{Midelfart2010}.
-
- This is achieved with the aid of the massive computing power in graphics processing units (GPUs) and optimized software libraries for scene rendering.
-The simulator loads a 3D model of the seafloor and object class into OpenGL and positions the camera at the sonar location.
-The scene is highlighted using a light tube that extends to infinity along the AUV's axis of probagation.
-We assume rough, isotropic surfaces reflecting sound energy equally in all directions.
-This can be modeled with a Lambertian scattering model~\cite{Blake1993,Bell1995}, where the intensity of the backscattered sound solely depends on the incidence angle of the transmitted signal onto the model surface.
-The OpenGL rendering produces an optical 2D intensity image and depth map that reveals the distance from each pixel to the propagation axis.
-For maximum flexibility this data is finally combined with OpenCL to produce the image templates.
-
-as part of the classification process. This is achieved with the aid of the massive computing power in graphics processing units (GPUs) and optimized software libraries for scene rendering.
-
-The simulator loads a 3D model of the seafloor and an object class model into OpenGL, where emitted sound waves are modeled with a cylindrical light source placed at the sonar transmit location.
-
-We assume rough, isotropic surfaces reflecting sound energy equally in all directions.
-This can be modeled with a Lambertian scattering model~\cite{Blake1993,Bell1995}, where the intensity of the backscattered sound solely depends on the incidence angle of the transmitted signal onto the model surface.
-When rendering OpenGL is set up to produce an optical 2D intensity image and depth map that reveals the distance from each pixel to the propagation axis.
-For maximum flexibility this data is finally combined with OpenCL to produce the image templates.
-
-Various simulators for high frequency, side-looking sonar imagery have been published, e.g.~\cite{Bell1997,Sammelm2003}.
-However, most implementations have prioritized accurate acoustic modeling at the expense of execution speed.
-One exception is the SIGMAS+ simulator in~\cite{Coiras2009a, Coiras2009b}.
-Similarly to our approach it takes advantage of parallel processing on GPUs, and it uses OpenGL to render a 2D image from a 3D model, under the assumption of Lambertian scattering.
-However, SIGMAS+ is still aimed more towards realistically looking images, including effects like ambient noise, which is irrelevant for our purpose of template generation.
-Also, their simulator superimpose images obtained from multiple OpenGL rendering passes to create the sonar image, while we render once and post-process this result with OpenCL to tailor the sonar image with greater flexibility.
-
-For lack of a better name, the simulator will be referred to as "FFISim" throughout this article.
-
-Show:
-\begin{itemize}
-\item Adaptive template matching to avoid large static template libraries,
-\item Improved classification performance by feeding simulator with parameters derived off SAS image,
-\item Further improved classification by performing a fine-search within a narrow search space around the SAS derived parameters. GPU makes this more powerful.
-\end{itemize}
-
-\section{Methods}
-
-Creating a 2D view of an arbitrary complex 3D scene is a non-trivial matter.
-This is why we decided to use the core OpenGL pipeline to do this for us.
-OpenGL is a popular, well-matured and multi-platform application programming interface (API) for rendering 2D and 3D vector graphics.
-It relieves us from the intricacies of projecting vertices, faces and textures defined in a 3D space onto a suitable 2D image plane.
-This section explains how we set up OpenGL for this task, and then proceeds to describe how we post process the OpenGL images with OpenCL to form the final sonar image.
-
-\begin{itemize}
-\item Load arbitrary complex models
-\item Simulate sonar with OpenGL
-\item Speed considerations
-\end{itemize}
-
-Models are not ROCCAN (???)
-
-\subsection{Loading the scene}
-
-Our models are stored in regular 3D model files.
-The simulator begins by loading these into a tree of nodes, each representing a vertex in the model with corresponding facet and texture information.
-Then the vertices are placed into a virtual scene in OpenGL, and a camera set up to capture the part of the scene corresponding to the image.
-When OpenGL renders the image, it first computes which vertex contributes to which pixel in the image, then computes the intensity and color of the pixel based on the node's facet, texture and scattering model.
-The code that operates on vertices are called a vertex shader, and the code that operates on pixels are called a fragment shader.
-
-The mapping of the vertices from object coordinates to image coordinates happen through a set of transformations:
-%
-\begin{align*}
-\bmat{x\\y\\0\\1}_\text{image} = \boldsymbol{P} \cdot \boldsymbol{V} \cdot \boldsymbol{M} \cdot \bmat{x\\y\\z\\1}_\text{object},
-\end{align*}
-%
-where $\bmat{\cdot}_\text{model}$ is the loaded model, $\bmat{\cdot}_\text{image}$ is the output image, and $\M$, $\V$ and $\P$ are transformation matrices. These will be described next.
-
-The first transform we apply is the model matrix $\M$, which operates in the model's local coordinate system. It is defined as
-%
-\begin{align}
-\M = \bmat{x_t \\ y_t \\ z_t \\ w}_\text{\parbox{1cm}{\setlength\baselineskip{0.3cm}transformed\\model}} &= \boldsymbol{T} \cdot \boldsymbol{R} \cdot \boldsymbol{S} \bmat{x \\ y \\ z \\ w}_\text{model},
-\end{align}
-%
-where $\boldsymbol{S}$, $\boldsymbol{R}$ and $\boldsymbol{T}$ are matrices\todo{be explicit on dependencies?} for scaling, rotation and translation, respectively.
+% - Adapt simula th\baselineskip{0.3cm}transformed\\model}} &= \boldsymbol{T} \cdot \R \cdot \boldsymbol{S} \bmat{x \\ y \\ z \\ w}_\text{model},
+\end{align}
+%
+where $\boldsymbol{S}, \R$ and $\boldsymbol{T}$ are matrices\todo{be explicit on dependencies?} for scaling, rotation and translation, respectively.
 These are provided for reference in Appendix \ref{transformation_matrices}.

### 2018-02-06 Simulator/article.tex
Major commit before messing with PhD thesis

@@
 
-The mapping of the vertices from object coordinates to image coordinates happen through a set of transformations:
+Mapping of the vertices from object coordinates to image coordinates happen through a set of transformations:
 %
 \begin{align*}
-\bmat{x\\y\\0\\1}_\text{image} = \boldsymbol{P} \cdot \boldsymbol{V} \cdot \boldsymbol{M} \cdot \bmat{x\\y\\z\\1}_\text{object},
+\bmat{x\\y\\0\\1}_\text{image} = \boldsymbol{P} \cdot \boldsymbol{V} \cdot \boldsymbol{M} \cdot \bmat{x\\y\\z\\1}_\text{model},
 \end{align*}
 %
-where $\bmat{\cdot}_\text{model}$ is the loaded model, $\bmat{\cdot}_\text{image}$ is the output image, and $\M$, $\V$ and $\P$ are transformation matrices. These will be described next.
-
-The first transform we apply is the model matrix $\M$, which operates in the model's local coordinate system. It is defined as
-%
-\begin{align}
-\M = \bmat{x_t \\ y_t \\ z_t \\ w}_\text{\parbox{1cm}{\setlength\baselineskip{0.3cm}transformed\\model}} &= \boldsymbol{T} \cdot \boldsymbol{R} \cdot \boldsymbol{S} \bmat{x \\ y \\ z \\ w}_\text{model},
-\end{align}
-%
-where $\boldsymbol{S}$, $\boldsymbol{R}$ and $\boldsymbol{T}$ are matrices\todo{be explicit on dependencies?} for scaling, rotation and translation, respectively.
-These are provided for reference in Appendix \ref{transformation_matrices}.
-
-Next we map the model into world coordinates using the view matrix $\V$.
+where $\bmat{\cdot}_\text{model}$ is the loaded model, $\bmat{\cdot}_\text{image}$ is the output image, and $\M$, $\V$ and $\P$ are model, view and transformation matrices, respectively.
+These will be described next.
+
+The model matrix $\M$ is the first transform applied to the object, and is defined as:
+%
+\begin{align}
+\M = \bmat{x_t \\ y_t \\ z_t \\ w}_\text{\parbox{1cm}{\setlength\baselineskip{0.3cm}transformed\\model}} &= \boldsymbol{T} \cdot \boldsymbol{R} \cdot \boldsymbol{S} \cdot \bmat{x \\ y \\ z \\ w}_\text{model},
+\end{align}
+%
+where $\boldsymbol{S}$, $\boldsymbol{R}$ and $\boldsymbol{T}$ are transformation matrices\todo{be explicit on dependencies?} that sets the object's scale, rotation and translation, respectively.
+Combined these transforms map the object from its local coordinate system to the world's coordinate system.
+They are provided for reference in Appendix \ref{transformation_matrices}.
+
+Next the object is mapped into world coordinates with the view matrix $\V$.
 This involves aligning the camera to precisely fit the image region.
@@
 \section{Transformation matrices}\label{transformation_matrices}
-%
-We place, resize and orient objects in an euclidean 3D space using suitable transformation matrices that are applied to each of the object's vertices:
-\begin{align}
-\bmat{x_t \\ y_t \\ z_t \\ w}_\text{\parbox{1cm}{\setlength\baselineskip{0.3cm}transformed\\model}} &= \boldsymbol{T}\R\boldsymbol{S} \bmat{x \\ y \\ z \\ w}_\text{model}.
-\end{align}
-Here a model vertex at position $(x,y,z)$ is first scaled with $\boldsymbol{S}$, then rotated with $\R$
-There is nothing special to these matrices, but we supply them here for completeness.
-
-We implement rotations in a 3D euclidean space by combining matrices that rotate around each of the three respective dimensions. For rotation by an angle $\phi_x$ clockwise around the $x$-axis we have
-\begin{align}
-\R_x(\phi_x) &= \bmat{
-1  &  0           &  0           &  0 \\
-0  &  \cos\phi_x  &  -\sin\phi_x &  0 \\
-0  &  \sin\phi_x  &  \cos\phi_x  &  0 \\
+
+Each data model that we load into OpenGL is defined in its own local coordinate system.
+To map them into OpenGL world coordinates a set of common transformation matrices are applied to each model's vertices:
+%
+\begin{align}
+\bmat{x_t \\ y_t \\ z_t \\ w}_\text{\parbox{1cm}{\setlength\baselineskip{0.3cm}transformed\\model}} &= \boldsymbol{T} \cdot \boldsymbol{R} \cdot \boldsymbol{S} \cdot \bmat{x \\ y \\ z \\ w}_\text{model}.
+\end{align}
+%
+Here a model vertex at position $(x,y,z)$ is first scaled with $\boldsymbol{S}$, then rotated with $\boldsymbol{R}$ and finally translated with $\boldsymbol{T}$, resulting in the transformed coordinates $(x_t,y_t,z_t)$.
+The parameter $w$ defines whether the vertex is a position ($w=1$) or a direction ($w=0$), which need to be handled differently.
+While these matrices are well-known in the field of computer graphics, there exist several different conventions but we supply them here for completeness.
+
+The scaling and translation matrices are simple:
+%
+\begin{align}
+\boldsymbol{S}(s_x,s_y,s_z) &= \bmat{
+   s_x  &  0    &  0    &  0 \\
+   0    &  s_y  &  0    &  0 \\
+   0    &  0    &  s_z  &  0 \\
    0    &  0    &  0    &  1 \\
@@
   \end{align}
-and similarly, for the other two euclidean dimensions we have
-\begin{align}
-\R_y(\phi_y) &= \bmat{
-\cos\phi_y  &  0  &  \sin\phi_y  &  0 \\
-0           &  1  &  0           &  0 \\
--\sin\phi_y &  0  &  \cos\phi_y  &  0 \\
-0           &  0  &  0           &  1 \\
-},
-\end{align}
+%
 and
-\begin{align}
-\R_z(\phi_z) &= \bmat{
-\cos\phi_z  &  -\sin\phi_z &  0  &  0 \\
-\sin\phi_z  &  \cos\phi_z  &  0  &  0 \\
-0           &  0           &  1  &  0 \\
-0           &  0           &  0  &  1 \\
-}.
-\end{align}
-The order in which these are applied matters. We rotate around the $x$-axis first, then around $y$ and finally around $z$:
-\begin{align}
-\R(\phi_x, \phi_y, \phi_z) &= \R_z(\phi_z)\;\R_y(\phi_y)\;\R_x(\phi_x).
-\end{align}
-
+
+TZF64LJO2AUK2XEM
+%
 \begin{align}
@@
 	0  &  0  &  0  &  1 \\
-}
-\end{align}
-
-\begin{align}
-\boldsymbol{S}(s_x,s_y,s_z) &= \bmat{
-s_x  &  0    &  0    &  0 \\
-0    &  s_y  &  0    &  0 \\
-0    &  0    &  s_z  &  0 \\
+},
+\end{align}
+%
+where ($s_x$, $s_y$, $s_z$) and ($dx$, $dy$, $dz$) is scaling and translation factors.
+
+Rotations will be specified using intrinsic Euler angles, i.e.
+as yaw ($\phi$), pitch ($\theta$) and roll ($\psi$), representing clockwise rotations around the $z$-, $y$- and $x$-axis, respectively.
+The Brian-Tait $Z$-$Y'$-$X''$ convention is used for rotation order, in which roll is applied first, then pitch and finally yaw:
+%
+\begin{align}
+\R(\phi_x) = \R_z(\phi)\R_y(\theta)\R_x(\psi),
+\end{align}
+%
+where
+%
+\begin{align}
+\R_x(\psi) &= \bmat{
+1  &  0           &  0           &  0 \\
+0  &  \cos\psi  &  -\sin\psi &  0 \\
+0  &  \sin\psi  &  \cos\psi  &  0 \\
 0  &  0           &  0           &  1 \\
-}
-\end{align}
+},
+\end{align}
+%
+\begin{align}
+\R_y(\theta) & = \bmat{
+\cos\theta	& 0          & \sin\theta & 0 \\
+0					& 1                   & 0          & 0          &  \\
+-\sin\theta  & 0                   & \cos\theta & 0          &  \\
+0            & 0                   & 0          & 1          &  \\
+},           &
+\end{align}
+%
+and
+%
+\begin{align}
+\R_z(\phi) &= \bmat{
+\cos\phi  &  -\sin\phi &  0  &  0 \\
+\sin\phi  &  \cos\phi  &  0  &  0 \\
+0				&  0           &  1  &  0 \\
+0           &  0           &  0  &  1 \\
+}.
+\end{align}
+%
+Note that at $\theta=\frac{\pi}{2}$ we have $\phi=-\psi$, and at $\theta=-\frac{\pi}{2}$ we have $\phi=\psi$.
+Hence, one degree of freedom is lost at these points, an effect known as Gimbal lock.
+This can be solved by avoiding Eulean angles, e.g.
+by direct specification of orientation in terms of a rotation axis and angle.
+This can be achieved using the Euler-Rodrigues formula or with quaternions.
+The latter is the most common, due to added benefits such as reduced computational complexity, improved numerical stability and simple ways to perform spherical interpolations.
+However, in for our application this is not needed.
+%
+Translation is performed last. 
+%
 
@@
 % 
-
   \section*{Acknowledgments}
@@
 
-The authors would like to thank the Norwegian Defence Research Establishment (FFI) for funding to develop the simulator\todo{more?}.
+The authors would like to express their gratitude to the Norwegian Defence Research Establishment (FFI) for funding the development of the simulator.
 
@@
 
-\input{bio/jo_inge_buskenes}
+\input{bio/jo_inge_buskenes_new}
 % \input{bio/jon_petter_aasen}
@@
 
+\input{derivations}
+
 \end{document}

### 2018-02-06 Simulator/bio/jo_inge_buskenes_new.tex
Major commit before messing with PhD thesis

--- 2018-02-06 LCAforActiveSonarImaging/bio/jo_inge_buskenes_new.tex
@@
 
-His research interests include radar and sonar technology, aptive image reconstruction, high performance computing, intelligent detector design and open source software.
+His research interests include radar and sonar technology, adaptive image reconstruction, high performance computing, intelligent detector design and open source software.
 \end{IEEEbiography}

### 2018-02-06 Simulator/bio/roy_edgar_hansen.tex
Major commit before messing with PhD thesis

\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{bio/roy_color.eps}}]{Roy Edgar Hansen} (M'07) received the M.Sc. and Ph.D. degrees in physics from the University of Troms\o, Troms\o, Norway, in 1992 and 1999, respectively.

From 1992 to 2000, he was with the Norwegian research company TRIAD, working on multistatic sonar, multistatic radar, synthetic aperture radar (SAR), and underwater communications. Since 2000, he has been with the Norwegian Defence Research Establishment (FFI), Kjeller, Norway. He is currently Principal Scientist and team leader for the autonomous underwater vehicle development and the synthetic aperture sonar development at FFI. He is also Adjunct Associated Professor at the University of Oslo, Oslo, Norway.
\end{IEEEbiography}

### 2018-02-06 Simulator/bio/roy_edgar_hansen_new.tex
Major commit before messing with PhD thesis

--- 2018-02-06 Simulator/bio/roy_edgar_hansen.tex
@@
 degrees in physics from the University of Troms\o, Troms\o, Norway, in 1992 and 1999, respectively.
-
+The PhD thesis is named Measurements in the Mixed Layer by a Bistatic Multi-CW Doppler Sonar. 
 From 1992 to 2000, he was with the Norwegian research company TRIAD, working on multistatic sonar, multistatic radar, synthetic aperture radar (SAR), and underwater communications.
 Since 2000, he has been with the Norwegian Defence Research Establishment (FFI), Kjeller, Norway.
-He is currently Principal Scientist and team leader for the autonomous underwater vehicle development and the synthetic aperture sonar development at FFI.
-He is also Adjunct Associated Professor at the University of Oslo, Oslo, Norway.
+In the period 2009-2015, he was research manager for the autonomous underwater vehicle development at FFI, including the fields underwater navigation, battery and fuel-cell technologies, decisional autonomy, and synthetic aperture sonar.
+He is currently principal scientist at FFI. 
+He is also adjunct associated professor at Department of informatics at University in Oslo, Norway.
+His research interests include synthetic aperture sonar and radar, ultrasound imaging, and array signal processing.
 \end{IEEEbiography}

### 2018-02-06 ThesisJo/introduction.tex
Major commit before messing with PhD thesis

@@
 
-\IEEEPARstart{O}{ne} of the greatest visions in science is that of artificial intelligence.
-If we could make machines not only do what we say, but also able to adapt their behaviour to handle the unexpected, the possibilities are perhaps only limited by our imagination.
+One of the greatest visions in science is that of artificial intelligence.
+If we could make machines not only do what we say, but also able to adapt their behavior to handle the unexpected, the possibilities are perhaps only limited by our imagination.
 

### 2018-02-06 ThesisJo/preface.tex
Major commit before messing with PhD thesis

\chapter*{Preface}

This thesis has been submitted to the Faculty of Mathematics and Natural Sciences at the University of Oslo in partial fulfillment of the requirements for the degree \emph{Philosophiae Doctor} (Ph.D.). This work was carried out at the Digital Signal Processing and Image Analysis group at the Department of Informatics, University of Oslo, under the supervision of Professor Andreas Austeng and Professor Sverre Holm. The work was financed by the University of Oslo.

\section*{Acknowledgements}

### 2018-02-06 LCAforActiveSonarImaging/diagram/cmd.sh

@@ -1,0 +1,3 @@
+#latexmk -pdf -pvc -lualatex -interaction=nonstopmode overview
+latexmk -pdf -pvc -interaction=nonstopmode overview
+#latexmk -pdf -pvc -xelatex -interaction=nonstopmode overview

### 2018-02-06 LCAforActiveSonarImaging/gfx/prepare.sh

@@ -1,0 +1,68 @@
+#!/bin/bash
+
+# FILES=" \
+#    beamforming.svg \
+#    implementation.svg \
+#    mvdr_complexity.svg \
+#    mvdr_build_R.svg \
+#    buildR-breakdown.svg \
+#    mvdr_implementation.svg \
+#    plot_holmengraa_L16_Navg1.eps \
+#    benchmark.eps \
+#    code_assessment.eps \
+#    code_assess_flops.eps \
+#    benchmark_boston.eps \
+#    "
+
+cp calc_kaiser_3dB.eps                             buske1_online.eps
+cp cross_tagged.png                                buske2_online.png
+cp mvdr_selected_windows_holmengraa.pdf            buske3_online.pdf
+cp oversampling_mosaic_bounds.pdf                  buske4_online.pdf
+cp oversampling_mosaic_bounds_lca_windows_beta.pdf buske5a_online.pdf
+cp oversampling_mosaic_bounds_lca_windows_phi.pdf  buske5b_online.pdf
+cp oversampling_mosaic.pdf                         buske6_online.pdf
+cp beamformer_comparison_cross.pdf                 buske7a_online.pdf
+cp beamformer_comparison_holmengraa.pdf            buske7b_online.pdf
+cp phi_lookup.eps                                  buske8_online.eps
+# cp  buske_online.eps
+
+
+
+# cp benchmark_boston.eps buske11_online.eps
+# Coverts eps file to black-white using ghostscript
+# function makeEpsBW {
+#    gs -sDEVICE=psgray -dNOPAUSE -dBATCH -dSAFER -sOutputFile=tmp.eps $1
+#    ps2eps tmp.eps
+#    mv tmp.eps.eps $2
+# }
+
+# inkscape --select=svg2 --verb org.inkscape.color.grayscale --export-eps=buske1_bw.eps beamforming.svg
+# inkscape --select=svg2 --verb org.inkscape.color.grayscale --export-eps=buske2_bw.eps implementation.svg
+# inkscape --select=svg2 --verb org.inkscape.color.grayscale --export-eps=buske4_bw.eps mvdr_build_R.svg
+# inkscape --select=svg2 --verb org.inkscape.color.grayscale --export-eps=buske6_bw.eps mvdr_implementation.svg
+
[…]

### 2018-02-06 LCAforActiveSonarImaging/src/lca_complexity.py

@@ -1,0 +1,23 @@
+from sympy import *
+from sympy import pprint
+
+A,B,a0,b0 = symbols('A B a0 b0')
+
+#a = A*exp(a0)
+#b = B*exp(b0)
+
+a,b = symbols('a b', complex=True)
+#b = B*exp(b0)
+
+alpha = symbols('alpha', real=True)
+
+z = (alpha*a - (1-alpha)*b)*conjugate((alpha*a - (1-alpha)*b))
+
+dz = diff(z,alpha)
+
+sol = solve(dz,alpha)
+
+
+
+pprint(expand(dz))
+pprint(sol)

### 2018-02-06 LCAforActiveSonarImaging/src/num_tests.py

@@ -1,0 +1,68 @@
+from numpy import *
+from numpy import pi
+import numpy as np
+import pylab as pl
+
+M = 6.0
+N = 1.0
+alpha = linspace(0,1,1000)
+m = arange(M)
+
+lmda = 1.0
+d = lmda/2.0
+theta = 5.0*pi/180
+w = cos(2*pi*m/(M-1))
+aa = exp(1j*2*pi*d*m/lmda*sin(theta))
+bb = aa * w
+
+aa = aa / aa.sum()
+bb = bb / bb.sum()
+print aa.sum(), bb.sum()
+
+x = np.random.uniform(size=(M,N)) + 1j*np.random.uniform(size=(M,N))
+
+a = dot(aa,x)
+b = dot(bb,x)
+
+f = outer(alpha,a) - outer((1-alpha),b)
+
+z1 = np.abs(f)**2 + 0.05
+
+z2 = outer(alpha**2, a*a.conj() + a*b.conj() + a.conj()*b + b*b.conj()) \
+   - outer(alpha,    a*b.conj() + a.conj()*b + 2*b*b.conj()) \
+   + b*b.conj()
+
+sol = (a*b.conj() + a.conj()*b + 2*b*b.conj())/(2*(a*a.conj() + a*b.conj() + a.conj()*b + b*b.conj()))
+
+#z1 = outer(alpha,a) - outer((1-alpha),b)
+
+fn = pl.figure()
+ax = fn.add_subplot(111)
+#ax.plot(z2.T)
+ax.plot(alpha,z1)
+ax.plot(alpha,z2)
+ax.plot([sol]*2,[alpha.min(),alpha.max()],'k')
+
+#print sol
+#z1 = dot(outer(array(meshgrid(ones(M),alpha)[0]) - outer((1-alpha),w)),x)
+
+
+pl.show()
+
+#print a
+#print b
+
+
+#A,B,a0,b0 = symbols('A B a0 b0')
+
+##a = A*exp(a0)
+##b = B*exp(b0)
+
+#a,b = symbols('a b', complex=True)
+##b = B*exp(b0)
+
+#alpha = symbols('alpha', real=True)
+
+#hmm = expand((alpha*a - (1-alpha)*b)*conjugate((alpha*a - (1-alpha)*b)))
+
+#pprint(hmm)

### 2018-02-06 LCAforActiveSonarImaging/submission/submit2/src/calcWindow3dB.py

@@ -1,0 +1,216 @@
+"""
+@file
+@brief Code for computing LCA steering angles
+
+ Sample code for computing steering angles for the
+ Low Complexity Adaptive (LCA) beamformer.
+
+ Released as complementary source code to the article titled
+ "Low Complexity Adaptive Beamformer for Active Sonar Imaging",
+ submitted for publishing to IEEE Journal of Oceanic Engineering 2015.
+
+ No license restrictions applies. Use, alter and/or redistribute as you wish.
+     Author:      Jo Inge Buskenes
+     Affiliation: University of Oslo / The Norwegian Defense Research Establishment (FFI)
+     Date:        February 2016
+"""
+
+from mynumpy import *
+import mypylab as pl
+
+  #######################################################################################
+ ## WINDOW FUNCTIONS ##
+######################
+   
+def getTrig(M=10, alpha=0.54, phi=0, d=0.5, k=2*pi):
+   """ Generates a steered Trigonometric window
+       
+         @param[in] M         Number of array elements
+         @param[in] beta      Trigonometric window parameter
+         @param[in] phi       Vector of steering angles  [rad]
+         @param[in] d         Element spacing            [m]
+         @param[in] k         Size of wavenumber         [rad/m]
+
+         @return              Steered Trigonometric windows
+   """
+   m    = linspace(0,1,M)
+   wc   = alpha - (1-alpha)*cos(2*pi*m)         # Window coefficients
+   m    = arange(0,M)                           # Create M indices from 0 to 1
+   a    = exp(-1j*k*m*d*sin(phi))               # Steering vector
+   ws   = dot(wc, a)                            # Normalization factor
+   win  = a * wc / ws                           # Steered and normalized window   
+   
+
+def getKaiser(M=10, beta=3, phi=0, d=0.5, k=2*pi):
+   """ Generates a steered Kaiser window
+       
+         @param[in] M         Number of array elements
+         @param[in] beta      Kaiser window parameter
+         @param[in] phi       Vector of steering angles  [rad]
+         @param[in] d         Element spacing            [m]
+         @param[in] k         Size of wavenumber         [rad/m]
+
+         @return              Steered Kaiser windows
+   """
[…]

### 2018-02-06 ThesisJo/src/derivations.py

@@ -1,0 +1,28 @@
+# -*- coding: utf-8 -*-
+"""
+Created on Wed Nov 12 16:27:38 2014
+
+@author: me
+"""
+
+import sympy as sp
+
+from sympy.physics.vector import ReferenceFrame
+from sympy.physics.vector import curl
+
+R = ReferenceFrame('R')
+v1 = R[1]*R[2]*R.x + R[0]*R[2]*R.y + R[0]*R[1]*R.z 
+
+def waveEquation():
+    
+    c = sp.Symbol('c')
+    phi = sp.Symbol('phi')
+    
+    omega = sp.Symbol('omega')
+    t = sp.Symbol('t')
+    
+    kx,ky,kz = sp.Symbol('kx', 'ky', 'kz')
+    x,y,z = sp.Symbol('x', 'y', 'z')
+    
+    
+    A = 

### 2018-02-06 UAM2013/Presentation/bin/hmm.sh

@@ -1,0 +1,4 @@
+#! /bin/bash
+
+mplayer gfx/test.avi
+

### 2018-02-06 UAM2014/Article/data/herman/manta_problem/run_2014_05_05_manta3.m

@@ -1,0 +1,236 @@
+if isunix
+   [~, hostname] = system('hostname');
+   [~, user]     = system('whoami');
+   joinge = strcmp(user(1:end-1),'me') && strcmp(hostname(1:end-1), 'Desktop');       % AMD Desktop with Linux
+else
+   [~, hostname]  = system('hostname');
+   [~, user]      = system('echo %USERNAME%');
+   joinge_laptop  = strcmp(user(1:end-1),'Laptop') && strcmp(hostname(1:end-1), 'Laptop'); % Nvidia Laptop with Windows
+   joinge_desktop = strcmp(user(1:end-1),'me')     && strcmp(hostname(1:end-1), 'Desktop'); % AMD Desktop with Windows 
+   
+   joinge = joinge_laptop || joinge_desktop;
+end
+%slantrange = 100.324018;
+%altitude = 22.743674;
+slantrange = 104.678520;
+altitude = 24.909445;
+groundrange = sqrt(slantrange^2 - altitude^2);
+burydepth = 0.533-0.27;
+angle = 105;
+length = 2.6;
+
+h = 0.18; h2 = 0.47; s = h/(altitude-h)*slantrange; slope = (asin(h2/s)-asin(altitude/slantrange))* 180/pi;
+%slope = 0
+
+cam_pos = [0 -groundrange altitude];
+r_bounds = [(groundrange - 5) (groundrange + 5)];
+c = 1500;
+fc = 100e3;
+fs = 40e3;
+d  = 1.2;
+
+% Don't set image size too low or strange things might happen.
+% If you experience images with large areas of 1s or 0s then
+% this is probably the reason.
+image_Nx = 300;
+image_Ny = 400;
+x_os = 3;
+y_os = 2;
+
+screen_Nx = 1200;
+screen_Ny = 800;
+
+%objfile = 'data-models/unit_cylinder.dae';
+objfile = 'data-models/manta3.dae';
+% seafile = '../../tools/vrml-models/cylinder_plane_red.dae';
+% objfile = '../../data-models/bench.obj';
+seafile = 'data-models/plane.dae';
+%objtransform = [0 groundrange -altitude+0.533/2-burydepth  0 0 (angle-90) 0.533 length 0.533];
+objtransform = [0 groundrange -altitude  slope 0 0 1 1 1];
+seatransform = [0 groundrange -altitude  slope 0 0  2 2 2];
[…]

### 2018-02-09 HowtoHowtoCapon/document.tex
Getting the hang of including articles

--- 2014-04-24 HowtoHowtoCapon/submission/final/article.tex
@@
-%%%%%%%%%%%%%%%%                                ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
-% CONDITIONALS %
-%%%%%%%%%%%%%%%%                                ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
-
-\newif\ifPeerReview\PeerReviewtrue              % Whether to create the PeerReview version or
-                                                % Journal version
-\newif\ifOnlineColor\OnlineColortrue            % Compile online color version?
-
-\newif\ifFlatArchive\FlatArchivefalse           % Whether archive is flat (messy) or contain 
-                                                % subfolders for graphics etc.
-\newif\ifFloatAtEnd\FloatAtEndfalse             % Available in PeerReview mode:
-                                                % Place floats at end of document?
-\newif\ifTODO\TODOtrue                        % Use todo notes?
-
-\ifOnlineColor
-   \newcommand\figPostfix{_online}              % Colored images online
-\else
-   \newcommand\figPostfix{_bw}                  % Black and white in journal
-\fi
-
-%%%%%%%%%%%%                                    ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
-% IEEEtran %
-%%%%%%%%%%%%                                    ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
-
-\ifPeerReview
-\documentclass[12pt,journal,draftclsnofoot,onecolumn]{IEEEtran}
-% \newcommand\CLASSINPUTbaselinestretch{1.66}     % http://theoval.cmp.uea.ac.uk/~nlct/latex/thesis/node17.html
-\else
-\documentclass[journal]{IEEEtran}
-\fi
-
-% \RequirePackage[latin1]{inputenc}%              % Set input encoding (optionally latin1)
-% \RequirePackage[T1]{fontenc}%                 % Set font encoding
-% \usepackage[norsk]{babel}
-
-% Single column review mode:
-%\documentclass[12pt,journal,onecolumn]{IEEEtran}
-%\newcommand\CLASSINPUTbaselinestretch{2}
-%\RequirePackage{calc}
-%\RequirePackage{fp} 
-% Left margin = 1 inch + hoffset + oddsidemargin (or evensidemargin)
-% Adding the 2cm gutter width to the odd/even side margins:
-%\setlength\hoffset{0pt}
-%\setlength\oddsidemargin{4cm}       % 4cm margins on the left side
-%\addtolength\oddsidemargin{-1in}    % Subtract the initial 1 inch
-%\setlength\textwidth{21cm-8cm}    % A4 width (21cm) minus margins on either side
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%                  ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
-% IEEE ''APPROVED'' PACKAGES %
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%                  ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
-
-\usepackage{cite}
-
-\ifCLASSINFOpdf
-   \usepackage[dvips]{graphicx}                 % Might not work. Use 'latex' instead of 
-   \ifFlatArchive\else                          % 'pdflatex'
-      \graphicspath{./gfx/}
-   \fi
-\else
-   \usepackage[dvips]{graphicx}
-   \ifFlatArchive\else
-      \graphicspath{./gfx/}
-   \fi
-\fi
-
-\RequirePackage[table,dvipsnames,svgnames]{xcolor}
-
-\usepackage[cmex10]{amsmath}                    % cmex10 option to be IEEE explore compliant
-\interdisplaylinepenalty=2500                   % Allows multiline equations to be broken
-
-% \RequirePackage{amssymb}
-
-\RequirePackage{array}
-
-\ifCLASSOPTIONcompsoc
-   \usepackage[caption=false,font=normalsize,labelfont=sf,textfont=sf]{subfig}
-\else
-   \usepackage[caption=false,font=footnotesize]{subfig}
-\fi
-\ifCLASSOPTIONcaptionsoff                       % IEEE promoted hack to turn off captions from the 
-   \let\MYorigsubfloat\subfloat                 % subfloat package should the captionsoff option
-   \renewcommand{\subfloat}[2][\relax]{\MYorigsubfloat[]{#2}} % be specified.
-\fi
-
-\ifFloatAtEnd
-\ifCLASSOPTIONcaptionsoff                       % Places float at the end of the document when the
-  \usepackage[nomarkers]{endfloat}              % captionsoff options is specified to IEEEtrans.cls
-  \let\MYoriglatexcaption\caption               % (PeerReview mode)
-  \renewcommand{\caption}[2][\relax]{\MYoriglatexcaption[#2]{#2}}
-\fi
-\fi
-
-\usepackage{fixltx2e}                           % Fix some twocolumn float problems
-
-%\usepackage{stfloats}                          % Allows: \begin{figure*}[!b]
-                                                % (double column figures on top/bottom)
-
-\usepackage{url}                                % Support for handling and breaking URLs
-
-% NOTE: PDF hyperlink and bookmark features are not required in IEEE
-%       papers and their use requires extra complexity and work.
-\newcommand\MYhyperrefoptions{bookmarks=true,bookmarksnumbered=true,
-pdfpagemode={UseOutlines},plainpages=false,pdfpagelabels=true,
-colorlinks=true,linkcolor={black},citecolor={black},urlcolor={black},
-pdftitle={An Optimized GPU Implementation of the MVDR Beamformer for Active Sonar Imaging},
-pdfsubject={},
-pdfauthor={Jo Inge Buskenes},
-pdfkeywords={adaptive beamforming, beamforming, complexity, sonar, active}}%
-\ifCLASSINFOpdf
-   \usepackage[\MYhyperrefoptions,pdftex]{hyperref}
-\else
-   \usepackage[\MYhyperrefoptions,breaklinks=true,dvips]{hyperref}
-   \usepackage{breakurl}                        % Allows 'dvips' driver to break links
-\fi
-
-%%%%%%%%%%%%%%%%%%%%%%%                         ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
-% ADDITIONAL PACKAGES %       
-%%%%%%%%%%%%%%%%%%%%%%%                         ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
-
-% \ifPeerReview
-% \let\OldIncludegraphics{\includegraphics}
-% \usepackage{letltxmacro}
-% \LetLtxMacro{\OldIncludegrsaphics}{\includegraphics}
-% \renewcommand{\includegraphics}[2][]{\OldIncludegraphics[width=\linewidth, #1]{#2}}
-% \fi
-
-% % \makeatletter
-% % \def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth
-% % \else\Gin@nat@width\fi}
-% % \makeatother
-% % \let\Oldincludegraphics\includegraphics
-% % \renewcommand{\includegraphics}[1]{\Oldincludegraphics[width=\maxwidth]{#1}}
-
-\usepackage[maxfloats=40]{morefloats}
-\newcounter{todoidx}
-% \setcounter{todoidx}
-
-\ifTODO
-   \definecolor{todobackground}{rgb}{0.95,0.95,0.95}
-   \setlength\marginparsep{1pt}
-   \setlength\marginparwidth{35pt}
-   \newlength\marginparwidthsmall
-   \setlength\marginparwidthsmall{\marginparwidth}
-   \addtolength\marginparwidthsmall{-7pt}
-   \newcommand\todo[1]{%
-      \addtocounter{todoidx}{1}%
-      {\color{Red}\bf(\thetodoidx{})}%%\fbox{\bf\thetodoidx{}}}%
-      \marginpar{%
-         {\vspace*{-10pt}\color{Red}\fbox{\bf\thetodoidx{}}}\\%
-         \fcolorbox{red}{todobackground}{\parbox{\marginparwidthsmall}{\scriptsize #1}}}}
-
-   \newcommand\todopar[1]{\fcolorbox{red}{white}{\parbox{0.97\linewidth}{#1}}}
-\else
-%    \usepackage[disable]{./todonotes} 
-   \newcommand\todo[1]{}
-\fi
-
-\newenvironment{narrow}[2]{%
-\begin{list}{}{%
-\setlength{\topsep}{0pt}%
-\setlength{\leftmargin}{#1}%
-\setlength{\rightmargin}{#2}%
-\setlength{\listparindent}{\parindent}%
-\setlength{\itemindent}{\parindent}%
-\setlength{\parsep}{\parskip}}%
-\item[]}{\end{list}}
-
-\usepackage{float}
-
-\ifOnlineColor
-   \definecolor{tabBlue}{HTML}{AACCFF}
-\else
-   \definecolor{tabBlue}{HTML}{CCCCCC}
-\fi
-
-%%%%%%%%%%                                      ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
-% MACROS %       
-%%%%%%%%%%                                      ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
-
-% \DeclareMathOperator*{\argmin}{\text{arg}\;\text{min}}
-
-\newcommand\Fig[1]{Fig.~\ref{#1}}
-
-\newcommand\Grey[1]{{\color{Grey}#1}}
-\newcommand\Red[1]{{\color{Red}#1}}
-\newcommand\Blue[1]{{\color{Blue}#1}}
-\newcommand\DarkBlue[1]{{\color{DarkBlue}#1}}
-\newcommand\LightBlue[1]{{\color{LightBlue}#1}}
-\newcommand\Brown[1]{{\color{Brown}#1}}
-\newcommand\Green[1]{{\color{Green}#1}}
-\newcommand\SeaGreen[1]{{\color{SeaGreen}#1}}
-\newcommand\Yellow[1]{{\color{yellow}#1}}
-\newcommand\Orange[1]{{\color{orange}#1}}
-
-\newcommand\nn{\nonumber\\}
-
-\newcommand\nmat[1]{\begin{matrix}#1\end{matrix}}
-\newcommand\bmat[1]{\begin{bmatrix}#1\end{bmatrix}}
-\newcommand\case[1]{\begin{cases}#1\end{cases}}
-\newcommand\textbox[2]{\footnotesize\text{\parbox{#1}{\centering\emph{#2}}}}
-
-\newcommand\rand{\text{rand}}
-\newcommand\randn{\text{randn}}
-\newcommand\rect{\text{rect}}
-\newcommand\sinc{\text{sinc}}
-\newcommand\tr{\text{tr}}
-\newcommand\adj{\text{adj}}
-
-% \newcommand\max{\text{max}}
-\newcommand\argmin[1]{\text{arg}\;\underset{#1}{\text{min}}}
-
-\newcommand\qqquad{\quad\qquad}
-\newcommand\qqqquad{\qquad\qquad}
-
-% \renewcommand\l[1]{\left#1}
-% \renewcommand\r[1]{\right#1}
-
-% {\text{\parbox{1.5cm}{\centering volume hyper- sphere}}}
-
-%Keyword colouring:
-\newcommand\kw[1]{#1}
-\newcommand\parm[1]{#1}%\color{Black}#1\color{Black}}
-
-\newcommand\of[1]{\scriptstyle(\parm{#1})\displaystyle}
-\newcommand\df[1]{\scriptstyle[\parm{#1}]\displaystyle}
-\newcommand\var[3]{#1_\text{#2}\of{#3}}
-
-\newcommand\diag{\text{diag}}
-
-% \raisebox{lift}[extend-above-baseline][extend-below-baseline]{text}
-\newcommand\mt[1]{\text{\emph{#1}}} %mt = mathtext
-\newcommand\mathnorm{\textstyle}
-\newcommand\mathbig[1]{\displaystyle#1\mathnorm}
-\newcommand\mathsmall[1]{\scriptstyle#1\mathnorm}
-\newcommand\mathtiny[1]{\scriptscriptstyle#1\mathnorm}
-\newcommand\sfrac[2]{\scriptstyle\raisebox{0.25pt}[0pt][0pt]{$\frac{#1}{#2}$}\mathnorm}
-\newcommand\nfrac[2]{\textstyle\frac{#1}{#2}\displaystyle}
-
-\newcommand\sumu[1]{\sum\limits^{#1}\;}
-\newcommand\suml[1]{\sum\limits_{#1}\;}
-\newcommand\sumb[2]{\sum\limits_{#1}^{#2}\;}
-
-\newcommand\produ[1]{\prod\limits^{#1}\;}
-\newcommand\prodl[1]{\prod\limits_{#1}\;}
-\newcommand\prodb[2]{\prod\limits_{#1}^{#2}\;}
-
-\newcommand\defeq{\overset{\underset{\mathrm{def}}{}}{=}}
-
-%Math macros:
-\newcommand\T{^{\scriptscriptstyle T}}
-\renewcommand\H{^{\scriptscriptstyle H}}
-
-\renewcommand\vec[1]{\boldsymbol{#1}}
-\newcommand\mat[1]{\boldsymbol{#1}}
-
-\newcommand\Om{O_\text{m}}
-\newcommand\Oa{O_\text{a}}
-\newcommand\Nl{N_\text{l}}
-\newcommand\Nk{N_\text{k}}
-\newcommand\1{\vec 1}
-\newcommand\I{\mat I}
-\renewcommand*\a{\vec a}
-\renewcommand*\i{\vec i}
-\renewcommand*\k{\vec k}
-\newcommand*\n{\vec n}
-\newcommand*\p{\vec p}
-\newcommand*\s{\vec s}
-\newcommand*\w{\vec w}
-\newcommand*\x{\vec x}
-\newcommand*\y{\vec y}
-
-\newcommand*\A{\mat A}
-\newcommand*\B{\mat B}
-\newcommand*\C{\mat C}
-\newcommand*\E{\mat E}
-\renewcommand*\P{\mat P}
-\newcommand*\eP{\mat{\hat P}}
-\newcommand*\R{\mat R}
-\newcommand*\Ri{\R^{-1}}
-\newcommand*\eR{\mat{\hat R}}
-\newcommand*\eRi{\hat{\mat R}\;\!^{-1}}
-\newcommand*\Navg{N_\text{avg}}
-\newcommand*\W{\mat W}
-\newcommand*\X{\mat X}
-\newcommand*\Xd{\X_{\!\Delta}}
-\newcommand*\Y{\mat Y}
-
-\renewcommand*\L{\mat \Lambda}
-\newcommand*\U{\mat U}
-% \renewcommand*\t{\mathtiny{^T}}
-% \newcommand*\h{\mathtiny{^H}}
-\renewcommand*\t{^T}
-\newcommand*\h{^H}
-
-\usepackage{tikz}
-\usetikzlibrary{shapes,snakes}
-\usepackage{amsmath,amssymb}
-% \usepackage{datatool}
-% \usepackage{glossaries}
-
-\newenvironment{outline}
-{\begin{itemize}}
-{\end{itemize}}
-
-%    \definecolor{todobackground}{rgb}{0.95,0.95,0.95}
-%    \setlength\marginparsep{3pt}
-%    \setlength\marginparwidth{42pt}
-%    \newlength\marginparwidthsmall
-%    \setlength\marginparwidthsmall{\marginparwidth}
-%    \addtolength\marginparwidthsmall{-7pt}
-%    \newcommand\todo[1]{%
-%       \addtocounter{todoidx}{1}%
-%       {\color{Red}\fbox{\bf\thetodoidx{}}}%
-%       \marginpar{%
-%          {\vspace*{-10pt}\color{Red}\fbox{\bf\thetodoidx{}}}\\%
-%          \fcolorbox{red}{todobackground}{\parbox{\marginparwidthsmall}{#1}}}}
-%
-
-% correct bad hyphenation here
-% \hyphenation{op-tical net-works semi-conduc-tor}
-
-%%%%%%%%%%%%                                    ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
-% GLOSSARY %
-%%%%%%%%%%%%                                    ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
-
-% \makeglossaries
-% \input{../../Library/glossary}
-
-%%%%%%%%%%%%%%%%%%                              ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
-% DOCUMENT START %
-%%%%%%%%%%%%%%%%%%                              ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
-
-% \usepackage{yfonts}
-
-\graphicspath{submission/final/gfx/}
-
-\begin{document}
-
 \title{An Optimized GPU Implementation of the MVDR Beamformer for Active Sonar Imaging}
@@
 \author{Jo~Inge~Buskenes,~\IEEEmembership{Student Member,~IEEE,} %
-        Jon~Petter~\AA{}sen,~\IEEEmembership{Student Member,~IEEE,} %
+        Jon~Petter~\AA{}Åsen,~\IEEEmembership{Student Member,~IEEE,} %
         Carl-Inge~Colombo~Nilsen,~\IEEEmembership{Member,~IEEE,} %
         Andreas~Austeng,~\IEEEmembership{Member,~IEEE}%
-\thanks{J. I. Buskenes, C.-I. C. Nilsen and A. Austeng are with the Department of Informatics, University of Oslo, Norway.}%
-\thanks{J. P. \AA{}sen is with MI-Lab, Norwegian University of Science and Technology, Trondheim, Norway.}% <-this % stops a space
+\thanks{J. I. Buskenes, C.-I. C. Nilsen and A. Austeng are with the Department of Informatics, University of Oslo, 0316 Oslo, Norway.}%
+\thanks{J. P. \AA{}sen is with MI-Lab, Norwegian University of Science and Technology, 7491 Trondheim, Norway.}% <-this % stops a space
 % \thanks{Manuscript received April 19, 2005; revised January 11, 2007.}
@@
 
-The Minimum Variance Distortionless Response (MVDR) beamformer has recently been proposed as an attractive alternative to conventional beamformers in active sonar imaging.
+The minimum variance distortionless response (MVDR) beamformer has recently been proposed as an attractive alternative to conventional beamformers in active sonar imaging.
 Unfortunately, it is very computationally complex because a spatial covariance matrix must be estimated and inverted for each image pixel.
@@
 
-In this study we show that for active sonar systems up to 32 channels, the computation time can be significantly reduced by performing arithmetic optimizations, and by implementing the MVDR on a Graphics Processing Unit (GPU).
+In this study we show that for active sonar systems up to 32 channels, the computation time can be significantly reduced by performing arithmetic optimizations, and by implementing the MVDR beamformer on a graphics processing unit (GPU).
 We point out important hardware limitations for these devices, and assess the design in terms of how efficiently it is able to use the GPU's resources.
 On a quad-core Intel Xeon system with a high-end Nvidia GPU, our GPU implementation renders more than a million pixels per second (1\;MP/s).
-Compared to our initial CPU implementation the optimizations described herein led to a speedup of more than 2 orders of magnitude, or an expected 5-10 times improvement had the CPU received similar optimization effort.
+Compared to our initial central processing unit (CPU) implementation the optimizations described herein led to a speedup of more than two orders of magnitude, or an expected five to ten times improvement had the CPU received similar optimization effort.
 This throughput enables real-time processing of sonar data, and makes the MVDR a viable alternative to conventional methods in practical systems.
@@
 % Keywords (normally not used for peer reviews)
-\ifPeerReview\else
 \begin{IEEEkeywords}
@@
 \end{IEEEkeywords}
-\fi}
-% \fi
+}
 
@@
 
-\IEEEPARstart{D}{ata} driven methods have been introduced in various imaging fields over the years in attempts to improve image quality.
-One such method is the Minimum Variance Distortionless Response (MVDR) beamformer.
-When compared to conventional methods, MVDR is often able to improve image contrast and resolution, as shown in e.g.
-radar~\cite{Benitz1997}, ultrasound~\cite{Synnevag2007}, and active sonar~\cite{Blomberg2012a,Blomberg2011,Dursun2009,Lo2004}.
+\IEEEPARstart{D}{ata-driven} methods have been introduced in various imaging fields over the years in attempts to improve image quality.
+One such method is the minimum variance distortionless response (MVDR) beamformer.
+When compared to conventional methods, MVDR is often able to improve image contrast and resolution, as shown in, for example, radar~\cite{Benitz1997}, ultrasound~\cite{Synnevag2007}, and active sonar~\cite{Blomberg2013,Blomberg2012a,Dursun2009,Lo2004}.
 
@@
 To ensure MVDR robustness, the literature suggests combining means such as temporal and spatial averaging, and regularization~\cite{VanTrees2002,Kailath1985}.
-The complexity issue, on the other hand, can be handled by introducing well-founded approximations.
+On the other hand, the complexity issue can be handled by introducing well-founded approximations.
 For instance, some studies assume data stationarity which allows the formation of a Toeplitz-structured covariance matrix that is simpler to invert~\cite{Asl2012a,Jakobsson2000}.
@@
 
-In this work, we show that such approximations may not be necessary for sonar systems up to 32 channels, since here the computation time can be reduced by 1-2 orders of magnitude by implementing MVDR on a Graphics Processing Unit (GPU).
-Similar work may be found in e.g.
-medical ultrasound imaging, where Chen \emph{et al.} have demonstrated a GPU implementation of the MVDR that operates on real valued data~\cite{Chen2011,Chen2011a}.
+In this work, we show that such approximations may not be necessary for sonar systems up to 32 channels, since here the computation time can be reduced by one to two orders of magnitude by implementing MVDR on a graphics processing unit (GPU).
+Similar work may be found in, for example, medical ultrasound imaging, where Chen \emph{et al.} have demonstrated a GPU implementation of the MVDR that operates on real valued data~\cite{Chen2011,Chen2011a}.
 While this implementation is fast, the real valued data constraint does not allow the creation of asymmetric and shifted responses, and hence prevents the MVDR's potential to be fully reached.
@@
 Our implementation is adapted to a well sampled 100\;kHz wideband active sonar system operating in the near field.
-The system has a range resolution comparable to the sample period so our MVDR implementation runs in near single snapshot mode.
+The system has a range resolution comparable to the sampling period so our MVDR implementation runs in near single snapshot mode.
 To deal with this and to avoid signal cancellation we compute and average the covariance estimates from overlapping subarrays~\cite{Kailath1985}.
 This may resemble techniques such as "subarray MVDR"~\cite{Chapman1976}, but unlike that technique we are not using subarrays to reduce complexity but to make it feasible to use MVDR in an active system.
-In related work, we investigated using the same MVDR GPU implementation in cardiac ultrasound imaging~\cite{Asen2012,Asen2013}, and in active sonar imaging we have compared the performance of this implementation to comparable adaptive methods~\cite{Buskenes2013}.
-
-This article is outlined as follows: In section \ref{methods} we introduce the concept of adaptive beamforming and provide details on the MVDR method.
-Then in \ref{maptogpu} we investigate the complexity issue, discuss means for reducing arithmetic complexity, and detail an implementation that makes efficient use of the GPU's parallel resources.
-The final design is assessed in \ref{images_and_benchmarks}-\ref{discussion}, where we provide benchmarks, comparisons with similar CPU implementations, and measures of how efficiently our implementation makes use of the GPU's resources.
+In related work, we investigated using the same MVDR GPU implementation in cardiac ultrasound imaging~\cite{Asen2012,Asen2013}, and in active sonar imaging we have compared the performance of this implementation to comparable adaptive beamformers~\cite{Buskenes2013}.
+
+This article is outlined as follows: In Section \ref{methods}, we introduce the concept of adaptive beamforming and provide details on the MVDR method.
+Then, in \ref{maptogpu}, we investigate the complexity issue, discuss means for reducing arithmetic complexity, and detail an implementation that makes efficient use of the GPU's parallel resources.
+The final design is assessed in Sections \ref{images_and_benchmarks} and \ref{discussion}, where we provide benchmarks, comparisons with similar CPU implementations, and measures of how efficiently our implementation makes use of the GPU's resources.
 
@@
 Signal signature is first removed by matched filtering.
-Then - before summation - a suitable set of delays, $\Delta$, and weights, $w$, are applied to focus on a pixel of interest at angle and range ($\theta,n$).}\label{beamforming}
+Then, before summation, a suitable set of delays $\Delta$ and weights $w$ are applied to focus on a pixel of interest at angle and range ($\theta,n$).}\label{beamforming}
 \end{figure}
@@
 
-By definition the beamformer output $z[n]$ can now be expressed as the weighted sum of all the delayed data samples:
+By definition, the beamformer output $z[n]$ can now be expressed as the weighted sum of all the delayed data samples:
 \begin{align}
@@
 It finds the set of complex weights that minimizes the beamformer's expected output power, while ensuring unity gain in the look direction~\cite{Capon1969}.
-This is a convex optimization problem that can be solved using Lagrange multipliers to yield the solution:
+This is a convex optimization problem that can be solved using Lagrange multipliers to yield the solution
 \begin{gather}
@@
 \end{gather}
-where $\a$ is a steering vector and $\R=E\{\x[n]\x\H[n]\} \in\mathbb{C}^{M,M}$ is the spatial covariance matrix for the full array.
+where $\a$ is a steering vector and $\R=E\{\x[n]\x\H[n]\}\in\mathbb{C}^{M,M}$ is the spatial covariance matrix for the full array.
 Since we pre-steer our data to every pixel in the image we simplify (\ref{weights}) by substituting $\a$ with a row vector $\1$ that represents broadside phase-steering.
@@
 \begin{itemize}
-\item \emph{spatial averaging} to avoid signal cancellation by decorrelating coherent echoes~\cite{Kailath1985},
-\item \emph{temporal averaging} over an interval comparable to the pulse length (1-5 samples) to maintain true speckle statistics~\cite{Synnevag2009a}, and
+\item \emph{spatial averaging} to avoid signal cancellation by decorrelating coherent echoes~\cite{Kailath1985};
+\item \emph{temporal averaging} over an interval comparable to the pulse length (one to five samples) to maintain true speckle statistics~\cite{Synnevag2009a};
 \item \emph{diagonal loading} to improve robustness to parameter errors~\cite{Cox1987,Maksym1979}.
@@
 To do this we need to segment our array into subarrays.
-If we let $x_l[n]$ represent the data vector from subarray $l$,
+If we let $x_l[n]$ represent the data vector from subarray $l$ by
 \begin{gather}
@@
 \end{gather}
-then $\breve{\R}$ can be calculated as:
+then $\breve{\R}$ can be calculated as
 \begin{gather}
@@
 This weight set is applied to all the subarrays, before computing the beamformer output as in (\ref{z}).
-Or, equivalently, we may apply the weight set to the sum of all the subarrays:
+Or, equivalently, we may apply the weight set to the sum of all the subarrays by
 \begin{align}
@@
 \fi%
-\caption{MVDR beamforming. For each pixel in range and azimuth,\newline
-1. an $L\times{}L$ sample covariance matrix $\eR$ is computed, \newline
-2. the term $\eR^{-1}\1$ is found using a linear equation solver,\newline
+\caption{MVDR beamforming. For each pixel in range and azimuth:\newline
+1. an $L\times{}L$ sample covariance matrix $\eR$ is computed; \newline
+2. the term $\eR^{-1}\1$ is found using a linear equation solver;\newline
 3. and the beamformer output $z$ is computed from (\ref{finalZ}), where $\w$ is found by substituting $\eR^{-1}\1$ into (\ref{weights}). } \label{mvdr_beamforming}
@@
 \begin{enumerate}
-\item computing the sample covariance matrix $\eR$ in (\ref{finalR}),
-\item computing $\eRi\1$ in (\ref{weights}), and
+\item computing the sample covariance matrix $\eR$ in (\ref{finalR});
+\item computing $\eRi\1$ in (\ref{weights});
 \item computing the beamformer output $z$ in (\ref{finalZ}).
@@
 \subsection{Computational Complexity}
+
+% \the\linewidth
 
@@
 Also notice how temporal averaging comes with a high computational penalty.
-This is because building $\eR$ is heavy on complex multiplications, which require 3 times as many arithmetic instructions as a complex addition, and a lot of these are repeated unnecessarily.
+This is because building $\eR$ is heavy on complex multiplications, which require three times as many arithmetic instructions as a complex addition, and a lot of these are repeated unnecessarily.
 
@@
 We decided to investigate the feasibility of running MVDR on a GPU by mapping it to a Nvidia GeForce Quadro 6000, which performs similarly to the consumer grade Geforce GTX 480.
-This is a high-end Compute Unified Device Architecture (CUDA) enabled GPU based on Nvidia's Fermi architecture, which as now been superseded by the Kepler architecture.
+This is a high-end Compute Unified Device Architecture (CUDA)-enabled GPU based on Nvidia's Fermi architecture, which as now been superseded by the Kepler architecture.
 The code was written in Nvidia's ``C for CUDA'' framework, which adds a few extra features to the C language to specify how to distribute work across a large number of threads.
@@
 
-To use Nvidia's own terminology, the Quadro 6000 is comprised of 14 streaming multiprocessors (SMs), each having 32 CUDA cores that execute a common program called a kernel.
+To use Nvidia's own terminology, the Quadro 6000 comprises 14 streaming multiprocessors (SMs), each having 32 CUDA cores that execute a common program called a kernel.
 Combined these cores deliver a peak performance of more than 1\;Tflop/s (appendix \ref{throughput}).
@@
 \ref{mvdr_complexity}, a direct computation of $\eR$ by implementing (\ref{spatialR}) and (\ref{finalR}) is the greatest computational burden.
-We target this issue by first \emph{minimizing arithmetic operations}, then aim to get as close to the peak \emph{arithmetic throughput} on the GPU as possible.
+We target this issue by first minimizing arithmetic operations, then aim to get as close to the peak arithmetic throughput on the GPU as possible.
 
@@
 In Fig.
-\ref{mvdr_build_R} we illustrate how to reduce the arithmetic complexity of building $\eR$ in a system with $M=5$ sensors, a subarray size of $L=3$ and temporal averaging is set to $K=1$.
+\ref{mvdr_build_R}, we illustrate how to reduce the arithmetic complexity of building $\eR$ in a system with $M=5$ sensors, a subarray size of $L=3$, and temporal averaging is set to $K=1$.
 Here the spatial sum is carried out before the temporal sum.
@@
 \begin{enumerate}
-\item exploit the fact that $\eR$ is Hermitian positive semi-definite to compute only one half of it,
-\item avoid redundant multiplications, and
-\item avoid redundant summations by first computing the upper row of $\eR$ and then add/subtract to these iteratively to find the diagonals of $\eR$.
+\item exploit the fact that $\eR$ is Hermitian positive semidefinite to compute only one half of it;
+\item avoid redundant multiplications;
+\item avoid redundant summations by first computing the upper row of $\eR$ and then adding/subtracting to these iteratively to find the diagonals of $\eR$.
 \end{enumerate}
@@
 \ref{mvdr_complexity} (higher is better) .
-Note how the arithmetic count is reduced by a factor 4-10 in the memory optimized case, and by a factor 6-22 in the instruction optimized case.}\label{mvdr_complexity_optimized}
+Note how the arithmetic count is reduced by a factor 4-10 in the memory-optimized case, and by a factor 6-22 in the instruction-optimized case.}\label{mvdr_complexity_optimized}
 \end{figure}
@@
 Exploiting symmetry and performing iterative summing (step 1 and 3) is always desirable, but avoiding redundant multiplications (step 2) comes at the cost of increased memory consumption.
-When all optimizations are applied we call it a "minimum instructions" solution, and when step 2 is left out we call it a "minimum memory" solution.
+When all optimizations are applied, we call it a "minimum instructions" solution, and when step 2 is left out, we call it a "minimum memory" solution.
 Fig.
-\ref{mvdr_complexity_optimized} compares the two solutions, and formulas may be found in appendix \ref{mvdr_formulas}.
+\ref{mvdr_complexity_optimized} compares the two solutions, and formulas may be found in Appendix \ref{mvdr_formulas}.
 Here the complexity curves are relative to the reference implementation in Fig.
@@
 \emph{Step~1}: The sample covariance matrix $\eR$ is formed by threads running along its diagonals.
-This allows spatial averaging to be implemented in a computationally efficient manner and minimizes inter-thread communication.
+This allows spatial averaging to be implemented in a computationally efficient manner and minimizes interthread communication.
 \emph{Step~2}: $\eRi\1$ is computed using the heavily optimized batched linear equation solver from Nvidia~\cite{Nvidia}.
-\emph{Step~3}: The beamformer output $z$ is computed in a straight forward fashion using $L$ threads that first sum the subarrays and then apply the MVDR weighting function.
-A single thread finally sum all the channels up.}\label{mvdr_implementation}
+\emph{Step~3}: The beamformer output $z$ is computed in a straightforward fashion using $L$ threads that first sum the subarrays and then apply the MVDR weighting function.
+A single thread finally sums all the channels up.}\label{mvdr_implementation}
 \end{figure}
@@
 Shared memory, on the other hand, is a very fast level 1 cache that is shared by all threads on an SM.
-At peak performance it is almost able to keep up with the computing units, but to obtain this performance we must make sure that:
+At peak performance, it is almost able to keep up with the computing units, but to obtain this performance we must make sure that:
 \begin{enumerate}
@@
 These challenges are very closely linked.
-Since the Quadro 6000 architecture is of compute capability 2.0, it has 48\;kB shared memory (L1 cache) and 128\;kB registers per streaming multiprocessor (SM)~\cite{Nvidia2012}.
+Since the Quadro 6000 architecture is of compute capability 2.0, it has 48\;kB shared memory (L1 cache) and 128\;kB of register memory per streaming multiprocessor (SM)~\cite{Nvidia2012}.
 This memory is shared by all active threads on that SM, a number that should be no less than 768~\cite{Nvidia2012a}.
@@
 In Fig.
-\ref{mvdr_implementation} we illustrate how we get around these challenges.
+\ref{mvdr_implementation}, we illustrate how we get around these challenges.
 First we make each SM compute entire range lines of pixels, then load the subset of $\x$ that all these pixels depend upon into shared memory.
@@
 Second, we assign $L$ threads per pixel to traverse the diagonals of $\eR$.
-On the top row a full computation is carried out, then that row is saved back to global memory following a coalesced access pattern that maximizes global memory bandwidth.
-For subsequent rows the threads move along the diagonals while performing iterative summations; the result from the previous element on the diagonal is updated by adding and removing the correlation coefficients that enters and exits the sum, respectively.
+On the top row, a full computation is carried out, then that row is saved back to global memory following a coalesced access pattern that maximizes global memory bandwidth.
+For subsequent rows, the threads move along the diagonals while performing iterative summations; the result from the previous element on the diagonal is updated by adding and subtracting the correlation coefficients that enter and exit the sum, respectively.
 To minimize memory consumption, we compute the coefficients again every time we need them.
 When a thread has finished up a diagonal, we have them wrap around to compute one of the diagonals in the lower triangular of $\eR$.
-Since $\eR$ is conjugate symmetric, the values in the leftmost column is obtained by a complex conjugate copy of the relevant value in the first row.
-This is slightly non-optimal as it causes divergent branching, but is needed because the solver is Gauss Jordan based and thus require a full $\eR$.
-Combined these steps balance the load evenly on all threads, is almost completely free of arithmetic redundancy, and consumes less memory.
+Since $\eR$ is conjugate symmetric, the values in the leftmost column are obtained by a complex conjugate copy of the relevant value in the first row.
+This is slightly nonoptimal as it causes divergent branching, but it is needed because the solver is Gauss Jordan based and thus require a full $\eR$.
+Combined these steps balance the load evenly on all threads, is almost completely free of arithmetic redundancy, and cause minimal memory consumption.
 
@@
 
-An important thing to note, however, is that unlike the problems most GPU libraries tries to solve we do not attempt so solve large linear equations or invert large matrices; our matrices are small, but we have a very large number of them.
-Fortunately Nvidia recently released a highly optimized batched linear equation solver tailored for this particular task.
+An important thing to note here is that unlike the problems most GPU libraries try to solve, we do not attempt so solve large linear equations or invert large matrices; our matrices are small, but we have a very large number of them.
+Fortunately, Nvidia recently released a highly optimized batched linear equation solver tailored for this particular task.
 It is Gauss Jordan based and supports partial pivoting and complex numbers.
@@
 A better choice would be a solver based on Cholesky decomposition.
-These are designed for Hermitian positive semi-definite matrices such as our covariance matrix, and require only half the arithmetic operations of a Gauss Jordan solver.
+These are designed for Hermitian positive-semidefinite matrices such as our covariance matrix, and require only half the arithmetic operations of a Gauss Jordan solver.
 
@@
 
-To demonstrate the imaging capability of the MVDR beamformer, we have processed experimental datasets from the $M=32$ element Kongsberg Maritime HISAS1030 sonar mounted on a HUGIN AUV~\cite{Hansen2009}.
-HISAS1030 is a high resolution synthetic aperture sonar with an array length of 1.2\;m, operating frequency of 100\;kHz, and bandwidth of 30\;kHz.
-The element size and spacing is 2.5$\lambda$ and the opening angle is 25$^\circ$.
-To produce the image shown in \Fig{holmengraa} the sonar was operated in sidescan mode.
-The studied object is the 1500 dwt oil tanker wreck Holmengraa.
-It is 68\;m long and 9\;m wide, and lies at a slanted seabed at 77\;m depth outside of Horten, Norway.
-The 1 megapixel (MP) MVDR image were here processed with parameters $L=16$, $K=1$, and $d=1\%$.
+To demonstrate the imaging capability of the MVDR beamformer, we have processed experimental datasets from the $M=32$ element Kongsberg Maritime HISAS1030 sonar mounted on a HUGIN autonomous underwater vehicle (AUV)~\cite{Hansen2009}.
+HISAS1030 is a high resolution synthetic aperture sonar (SAS) with an array length of 1.2\;m, operating frequency of 100\;kHz, and bandwidth of 30\;kHz.
+The element size and spacing is 2.5\;$\lambda$ and the opening angle is 25$^\circ$.
+To produce the image shown in \Fig{holmengraa}, the sonar was operated in sidescan mode.
+The studied object is the 1500-dwt oil tanker wreck Holmengraa.
+It is 68\;m long and 9\;m wide, and lies at a slanted seabed at 77\;m depth outside Horten, Norway.
+The 1 megapixel (MP) MVDR image was here processed with parameters $L=16$, $K=1$, and $d=1\%$.
 
@@
 The results were obtained by processing a 1\;MP image from the data from a 32 channel array, for all subarray sizes $L$, and for $K\in\{0,1,2\}$.
-Run-time measurements are shown in \Fig{benchmarks}, the run-times of memory-only and arithmetic-only GPU kernels are depicted in \Fig{code_assessment}, and an estimate of computation efficiency is presented in \Fig{code_assess_flops}.
+Runtime measurements are shown in \Fig{benchmarks}.
+The runtimes of memory-only and arithmetic-only GPU kernels are depicted in \Fig{code_assessment}.
+An estimate of computation efficiency is presented in \Fig{code_assess_flops}.
 We were also granted a test drive at the Boston HPC centre on a machine with an Intel Xeon E5-2670, 32\;GB RAM and an Nvidia K20.
@@
 
-As seen in the run-time comparisons presented in \Fig{benchmarks}, the GPU method became 2-3 orders of magnitude faster than the C implemtation we started out with.
-The remaining bottleneck in the final design is the inversion step, which is typically 5 times slower than the build step.
+As seen in the runtime comparisons presented in \Fig{benchmarks}, the GPU method became two to three orders of magnitude faster than the C implementation we started out with.
+The remaining bottleneck in the final design is the inversion step, which is typically five times slower than the build step.
 In most cases the processing speed of the Quadro 6000 is above 1\;MP/s, and this was only improved by a factor 1.5-2 when run on the new K20 GPU.
@@
 Excluded from these benchmarks is the data transfer time from CPU to GPU, which account for 2-20\% of the total processing time.
-We believe it makes little sense to include them since it keeps getting easier to perform these data transfers in parallel by offloading the task to the Direct Memory Access (DMA) controllers present on modern GPUs.
-Furthermore, the data rates in active sonar are relatively low compared to the bandwidth available for these transfers.
+We believe it makes little sense to include them since it keeps getting easier to perform these data transfers in parallel by offloading the task to the direct memory access (DMA) controllers present on modern GPUs.
+Furthermore, the data rates in an active sonar are relatively low compared to the bandwidth available for these transfers.
 
@@
 
-As observed in \Fig{benchmarks} the combination of minimizing arithmetic operations and implementing the MVDR on a GPU lead to a significant improvement in processing speed.
-Compared to a MATLAB and single thread C implementation an order 2 and 3 speedup was achieved, respectively.
+As observed in \Fig{benchmarks}, the combination of minimizing arithmetic operations and implementing the MVDR on a GPU lead to a significant improvement in processing speed.
+Compared to a MATLAB and single thread C implementation, an order 2 and 3 speedup was achieved, respectively.
 Note that we do not consider this comparison fair.
-In fact, the theoretical peak throughput of this GPU is roughly 20 times higher than that the CPU in question, meaning the potential of the CPU was far from reached in our initial implementations.
-The key reason for this is that the GPU design process pixels in batches, which allows us to minimize data transfers and maximize the use of fast memory cache.
+In fact, the theoretical peak throughput of this GPU is roughly 20 times higher than that the CPU in question, meaning the potential of the CPU was far from being reached in our initial implementations.
+The key reason for this is that the GPU design processes pixels in batches, which allows us to minimize data transfers and maximize the use of fast memory cache.
 In other aspects the designs are similar, both CPU implementations compute $\eR$ efficiently, and they make use of either an optimized custom Cholesky based solver or one from the Intel Math Kernel Library (MKL).
-Unfortunately we had little time to write a custom batch based solver and covariance builder for the CPU, but we expect the speed difference to be 5-10 times if we had compared two such designs.
+Unfortunately we had little time to write a custom batch based solver and covariance builder for the CPU, but we expect the speed difference to be five to ten times, if we had compared two such designs.
 
@@
 Also, the benchmark curves for building $\eR$ now seem linear.
-The main reason for this is that the optimization step negated much of the extra complexity introduced by averaging, i.e.
-reduced the complexity from O($N_KN_LL^2$) to less than O($L^2$).
-Another reason for the run-time to appear more like O($L$) is likely that our design makes better use of the GPU's resources when there is more processing involved per pixel.
+The main reason for this is that the optimization step negated much of the extra complexity introduced by averaging, i.e., reduced the complexity from O($N_KN_LL^2$) to less than O($L^2$).
+Another reason for the runtime to appear more like O($L$) is likely that our design makes better use of the GPU's resources when there is more processing involved per pixel.
 The inversion step, on the other hand, gains less from being implemented on a GPU.
@@
 Alternative solvers, such as one based on Cholesky decomposition that exploits the Hermitian property of $\eR$, can in theory reduce complexity by a factor two, but we question whether this result can be obtained in practice using a GPU.
-The CPU implementations perform Cholesky based inversion, but this does not explain why the C and MATLAB implementation have a total run-time that is linearly dependent on $L$.
+The CPU implementations perform Cholesky-based inversion, but this does not explain why the C and MATLAB implementation have a total runtime that is linearly dependent on $L$.
 We believe this happens because these implementations are dominated by the calculation of the covariance matrix and data movement, and not by the inversion step.
@@
 When optimizing a GPU design it is important to know whether it is limited by memory bandwidth or arithmetical throughput.
-To measure this we made two versions of the MVDR kernel, one that only performs arithmetic operations, and one that only performs memory operations.
-Then we benchmarked these kernels and compared their run-times to the total run-time (\Fig{code_assessment}).
+To measure this we made two versions of the MVDR kernel: One that only performs arithmetic operations, and one that only performs memory operations.
+Then, we benchmarked these kernels and compared their runtimes to the total runtime (\Fig{code_assessment}).
 A first thing to note is how these kernels seem equally occupied performing memory and arithmetic operations.
 This is a good sign, but since we consistently minimized memory consumption at the expense of some extra computations when building $\eR$ , we know it to be bound by memory bandwidth.
-It also has a problem with latency, which can be inferred from the total run-time being significantly higher than that of the two single-function kernels.
+It also has a problem with latency, which can be inferred from the total runtime being significantly higher than that of the two single-function kernels.
 Since the GPU hardware can carry out memory and arithmetic operations concurrently and independently, this gap indicates that the GPU sometimes does ``nothing''.
@@
 Even with all our efforts we were only able to obtain processing rates of 40\;GFlop/s on average in the desirable range of subarray sizes (\Fig{mvdr_complexity}).
-This is mainly due to the inversion step, since we for the most part obtained more than 100\;GFlop/s when building $\eR$.
+This is mainly due to the inversion step, since for the most part we obtained more than 100\;GFlop/s when building $\eR$.
 While these numbers are slightly underestimated, they are regardless far away from the Quadro 6000's peak of 1\;TFlop/s.
@@
 This belief is further supported by the test results from running the code on the new K20 Kepler card, where we saw a modest factor 1.5-2 speedup.
-Although the Kepler card peaks at 3.5-4\;Tflop/s, its shared memory bandwidth is only approximately twice that of the Quadro 6000, hence matches the observed speed-up.
+Although the Kepler card peaks at 3.5-4\;Tflop/s, its shared memory bandwidth is only approximately twice that of the Quadro 6000, hence matches the observed speedup.
 However, it is likely that the Kepler card would perform better had we optimized our design for it.
-For scientific computing the Kepler boards are considered more attractive, with the introduction of features such as kernels calling kernels, more registers, more shared memory, and better mechanisms for hiding or removing data transfers.
+For scientific computing, the Kepler boards are considered more attractive, with the introduction of features such as kernels calling kernels, more registers, more shared memory, and better mechanisms for hiding or removing data transfers.
 
@@
 We have shown that for systems consisting of up to 32 channels the problem can be largely mitigated by building the covariance matrix in a clever way, and by making use of the massive computational power available in modern GPUs.
-We were able to improve upon the run-time of a single thread C-implementation by roughly two orders of magnitude.
+We were able to improve upon the runtime of a single thread C-implementation by roughly two orders of magnitude.
 For most choices of parameters the GPU was able to create images at $\sim$1\;MP/s at an average data processing rate of $\sim$40\;GFlop/s.
@@
 
-\emph{Solving $\eRi\1$} is achieved by using a batched Gauss Jordan solver with support for complex numbers and partial pivoting.
-Its complexity - with partial pivoting excluded - can be expressed as:
+\emph{Solving $\eRi\1$} is achieved by using a batched Gauss Jordan solver with support for complex numbers and partial pivoting. Its complexity, with partial pivoting excluded, can be expressed as: 
 \begin{align}
@@
 The Quadro 6000 has 32 CUDA cores per SM, each operating at a rate of 1148\;MHz and being able to perform 2 floating point operations (flop) per clock cycle when multiply-add instructions are used.
-The theoretical peak arithmetic throughput is then given as:
+The theoretical peak arithmetic throughput is then given as
 \begin{align}
@@
 \end{align}
-Now let us compare this to the memory throughput. The ``global'' GDDR5 memory bus is 384\;bit wide, and operates at 3\;Ghz where 2 bits are sent every cycle. Its peak bandwidth is then:
+Now let us compare this to the memory throughput. The ``global'' GDDR5 memory bus is 384\;bit wide, and operates at 3\;Ghz where 2 bits are sent every cycle. Its peak bandwidth is then
 \begin{align}
@@
 \end{align}
-The shared memory, on the other hand, is organized into 32 banks per SM, each being 32\;bit wide and operating at 1148\;MHz where 1 bit is sent per cycle. Its peak aggregated bandwidth is then:
+The shared memory, on the other hand, is organized into 32 banks per SM, each being 32\;bit wide and operating at 1148\;MHz where 1 bit is sent per cycle. Its peak aggregated bandwidth is then
 \begin{align}
@@
 
+\bibitem{Blomberg2013}
+\BIBentryALTinterwordspacing
+A.~E.~A. Blomberg, A.~Austeng, R.~E. Hansen, and S.~A.~V. Synnes, ``{Improving
+  Sonar Performance in Shallow Water Using Adaptive Beamforming},'' \emph{IEEE
+  Journal of Oceanic Engineering}, vol.~38, no.~2, pp. 297--307, Apr. 2013.
+  \url{http://ieeexplore.ieee.org/lpdocs/epic03/wrapper.htm?arnumber=6401207}
+\BIBentrySTDinterwordspacing
+
 \bibitem{Blomberg2012a}
@@
 \BIBentrySTDinterwordspacing
-
-\bibitem{Blomberg2011}
-A.~E.~A. Blomberg, R.~E. Hansen, S.~A.~V. Synnes, and A.~Austeng, ``{Improved
-  interferometric sonar performance in shallow water using adaptive
-  beamforming},'' in \emph{Proceedings of the International Conference \&
-  Exhibition on Underwater Acoustic Measurements (UAM), Kos, Greece, June},
-  2011.
 
@@
 
+\ifPhdDoc\else
 \input{bio/jo_inge_buskenes}
@@
 \input{bio/andreas_austeng}
+\fi
 
@@
 \vfill
-
-\end{document}

### 2018-02-16 ThesisJo/abstract.tex
Parallel build starting to work

\input{variables}\ifMonolithic\else\input{header}\input{subheader}\fi
%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

\chapter{Abstract}

%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
\ifMonolithic\else\input{subfooter}\fi

### 2018-02-16 ThesisJo/literature_review.tex
Parallel build starting to work

--- 2015-03-04 ThesisJo/background.tex
@@
-\chapter{Background}
+\input{variables}\ifMonolithic\else\input{header}\input{subheader}\fi
+%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
 
-MVDR/LCA:
-- Beamforming techniques.
+\chapter{Litterature Review}
 
-Simulator:
-- Scattering theory
-- Coordinate systems
-
-Common denominator:
-- Image progressing
-- High processing computing
-- GPUs
-
-Light fast, sound slow.
-Sound propagates far, light attenuated.
-
-The aim of this chapter is to review the basic background material on which this thesis is founded.
-
-\section{Wave Propagation}
-
-When a rock is thrown into a pound of water it abruptly displaces water away from the center of impact. 
-%This is an unstable state where various physical phenomena wrestle for control. Gravity wants the water back where it was, inertia cause the mass to resist any change in its state of motion.
-This results in spherical waves propagating from the displacement center, transporting the kinetic energy of the rock.
-These waves are described by how their particles move.
-At the surface particles will move down up and down, and back and forth, in a circular motion.
-In the depths particles move longitudinally, transporting energy by means of alternating pressure.
-
-The wave equation is the single most important equation governing a wave's propagation through a medium. When the medium is loss-less it can be expressed as:
-
-\begin{align*}
-\nabla^2 f(\p,t) - \frac{1}{c^2(\p)}\frac{\partial^2 f(\p,t)}{\partial t^2} = -s(t) \delta(\p-\p_0)
-\end{align*}
-
-where $\nabla^2$ is the Laplacian operator.
-
-The wave equation constrains the magnitude of the wavenumber,
-%
-\begin{align*}
-\vec k = \frac{\omega}{c} = \frac{2\pi}{\lambda},
-\end{align*}
-%
-so only the direction of $\k$ varies.
-
-\section{Array Processing}
-
-\begin{figure}[th]
-\graphicsAI[width=\linewidth]{gfx/geometry.svg}
-\caption{The coordinate system used throughout the thesis.}\label{geometry}
-\end{figure}
-
-Consider a monochromatic plane wave impinging on an $M$-element array (\Fig{geometry}). Let the position of sensor $m$ be $\p_m$, th  of sensors at locations 
-In this thesis we will relate all geometrical interpretations to the euclidean, spherical coordinate system shown in \Fig{geometry}.
-When assessing the imaging performance of a sonar its origin will be in the center of the array, but when assessing scenes its origin will be in the center of the image.
-A unit length vector $\vec u$ can be written as:
-%
-\begin{align*}
-\vec u =
-\begin{bmatrix}
-\sin\theta \cos\phi \\
-\sin\theta \sin\phi \\
-\cos\theta
-\end{bmatrix}
-\end{align*}
-%
-
-Signal model. Monochromatic, plane wave:
-%
-\begin{align}
-\s[n] = \begin{bmatrix}
-         s_0[n] \\ s_1[n] \\ \vdots \\ s_{M-1}[n]
-        \end{bmatrix}
-      = \begin{bmatrix}
-         Ae^{j(\Omega n - \k\T\p_0)} \\
-         Ae^{j(\Omega n - \k\T\p_1)} \\
-         \vdots \\
-         Ae^{j(\Omega n - \k\T\p_{M-1})}
-        \end{bmatrix}
-      = Ae^{j\Omega n}\a(\k)
-\end{align}
-%
-where we call $\a(\k)$ the signal's \emph{steering vector}.
-
-\section{Synethic Aperture Sonar}
-
-For an autonomous underwater vehicle to reach its full
-
-Most autonomous devices designed today only perform very basic tasks.
-
-One of the areas where very fast where the development of faster and faster processing hardware might play a
-One of the technologies in which graphics processing units might make a welcome addition is in autonomous underwater vehicles.
-
-This thesis will describe a simulator for synthetic aperture sonar that
-
-\section{Scattering models}
-
-Defining parameters:
-- Roughness
-- Incident angle
-- Frequency
-
-Reflection and scattering from surfaces:
-
-Roughness >> wavelength:
-- Roughness > wavelength: Helmholtz-Kirchhoff (physical optics approach) (Hovem). Leads to Huygins priciple.
-- Roughness < wavelength: Rayleigh-Rice (perturbation technique) (Hovem). More accurate at slightly rough surfaces.
-
-Rayleigh. chi = 2 k sigma sin(theta): Rayleigh criterion: Surface is acoustically smooth when chi << 1.
+\section{Simulators}
 
 \begin{itemize}
-\item SWAT (Shallow Water Acoustic Toolset): 
-\item Bell's thesis:
-\item SIGMA:
-\item OpenGL does not support scattered writes. OpenCL does.
-\item 2D template approximation:
-\item Geometric distortions in SSS \cite{Cobra1992} Bell
-\item Sidescan sonar and synthetic aperture sonar. In seafloor coordinates. Maps to orthogonal rendering.
-\item Layover causes ambiguity, different seafloor phenomena can cause the same output image.
-Lots of cornercases when approximating the model in 2D.
-Instead use a 3D model and project it to a 2D one.
-Easier to operate, models are more general purpose.
-Less parameters to adjust.
-\item Viewer can be used to train an operator for finding non-trivial objects on the seafloor. Input for automated target recognition standard.
-\item Improved image quality of selected object, reduced data collection cost (price estimate?).
-\item OpenGL rendering can have infinite resolution, only limited by the pixel density in the image.
-\item Accuracy in terms of classification result. Need to embody static effects in a deterministic way. Incoherent synthesis of the image.
-\item Sound speed assumed constant.
-\item High frequency lead to little (define) penetration, lower range and higher resolution. Often modeled accurately with ray tracing techniques.
+\item SST - Sonar Simulation Toolkit. \cite{Goddard}
 \end{itemize}
 
-Model
+General Modeling tools
 \begin{itemize}
-\item Ray solution: Separate amplitude and phase components of the Helmholtz eq.
-by applying geometric optics approximation.
-Ok as long as amplitude varies more slowly with position than the phase.
-Implications: High frequency, because the curvature of a ray over a wavelength must be small, and the fractional change in the sound speed must be small over a wavelength.
-Further simplification available by Snell's Law.
-Fast to compute.
-Not able to handle diffraction, focal points and caustics, and will cause (too) abrupt changes on distinct changes in sea environment.
-Can be improved upon with fuzzy beam modeling / Gaussian beam tracing.
-\begin{itemize}
-\item Pro: Easy to compute. Visual result. Handles complex bathymetry. Array and model response can be accounted for.
-\item Con: Only works for high frequency and small sound variations. No diffraction, caustics and focus. Infinite energy caustics/sharp variations. Incoherent(?)
-\end{itemize}
-\item Normal modes: Separate amplitude and phase into a range dependent term, and a depth dependent term.
-Assumes minimal range dependence in sound speed profile and depth.
-Limited near field accuracy.
-Acoustic field does not need to be calculated at intermediate ranges.
-Penetration can be modeled.
-can be calculated
-\begin{itemize}
-\item Pro: Ok to compute. Handles penetration. Low frequency, shallow water.
-\item Con: Horizontal stratification. Far field only. 
-\end{itemize}
-\item Parabolic equation: One way propagation.
-Marching solution.
-Initial starting solution must be known, often computed with normal mode.
-Only valid for narrow beams $\pm$20.
-Able to handle range dependent environments.
-Computing full field is easy.
-Computation time is of O(f$^2$), hence mostly useful for low-frequency scenarios.
-Unable to deal with horizontal refraction (shadow zones from e.g.
-conical mounts)
-\item Fast field programs / Green's function solutions.
-Exact full wave solution for acoustic fields in horizontally stratified media.
-Can handle both compressional and shear waves, and hence both fluid and solid media.
-Far field assumption.
-Similar to normal mode techniques.
-\item Finite element: Scene segmented into triangles or rectangles with a size of one tenth of a wavelength, or less.
-At each node the wavefield can be determined from the wave equation.
-Computationally limited to low frequencies.
-\item 2D model: Acoustic field in range and depth.
-Assumes that a wave transmitted in a vertical plane remains in it.
-Reasonable when the sound speed can be assumed constant in the horizontal plane, and when the seafloor normals are in the vertical range-depth plane (avoid horizontal refraction, or actually reflection).
+\item CASS - Comprehensive Acoustic System Simulation. \cite{Goddard}
 \end{itemize}
 
-What's new?
-\begin{itemize}
-\item Tuned for speed using GPUs, combining OpenGL and OpenCL to take advantage of existing functionality for computer graphics with the added flexibility of general purpose programming on the GPUs.
-\item Performance to spare: Game-like viewer for the visualization and verification of the rendered scene and corresponding sonar image.
-\item First simulator of its kinds aimed towards improved ATR and ultimately better AUVs.
-\item 3D models are still novel. 
-\end{itemize}
+GSM - Generic Sonar Model.
+
+FOCUS simulator.
+
+%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
+\ifMonolithic\else\input{subfooter}\fi

### 2018-02-16 ThesisJo/bin/pagecnt.sh

@@ -1,0 +1,15 @@
+#!/bin/bash -e
+
+V1=$(cat cnt.txt)
+
+perl -pe 's/(glsnumberformat\}\{)(\d+)/$1.($2 + '"$V1"')/e' ${1}.glo > ${1}_cumulative.glo
+
+V2=$(pdfinfo ${1}.pdf | grep Pages | awk '{print $2}')
+PAGES=$((${V1}+${V2}))
+echo ${PAGES} > cnt.txt
+echo "====P1: ${1}===="
+echo "====V1: ${V1}===="
+echo "====V2: ${V2}===="
+echo "====Number of pages total: ${PAGES}===="
+
+# perl -pe 's/(glsnumberformat\}\{)(\d+)/$1.($2 + '"$var"')/e' introduction.glo

### 2018-02-19 HowtoHowtoCapon/document.tex
Parallel build starting to work
2_MVDR ok. 3_LCA inlcuded

@@
-\title{An Optimized GPU Implementation of the MVDR Beamformer for Active Sonar Imaging}
-
+% 
+% \graphicspath{gfx/../}
+%
+\title{An Optimized GPU Implementation of the MVDR Beamformer for Active Sonar Imaging}%
+%
 \author{Jo~Inge~Buskenes,~\IEEEmembership{Student Member,~IEEE,} %
@@
 % \thanks{Manuscript received April 19, 2005; revised January 11, 2007.}
-}
-
+}%
+%
 % The paper headers
 \markboth{IEEE Journal of Oceanic Engineering}%
-{A Robust and Optimized GPU Implementation of the MVDR beamformer}
-
+{A Robust and Optimized GPU Implementation of the MVDR beamformer}%
+%
 % Publishers ID mark:
 %\IEEEpubid{0000--0000/00\$00.00~\copyright~2007 IEEE}
-
+%
 % use for special paper notices
 %\IEEEspecialpapernotice{(Invited Paper)}
-
+%
 % for Computer Society papers, we must declare the abstract and index terms
@@
 \IEEEcompsoctitleabstractindextext{%
-\begin{abstract}
+\begin{abstract}%
 % Two limiting factors of modern phased array sonar imaging systems are the angular resolution and contrast.
 Adaptive beamformers target these limitatitons by weighting sensors dynamically to fit the impinging wavefield.
-
+%
 The minimum variance distortionless response (MVDR) beamformer has recently been proposed as an attractive alternative to conventional beamformers in active sonar imaging.
@@
 %The CPU would then be free to take on the rest of the beamforming as well.
-\end{abstract}
-
+\end{abstract}%
+%
 % Keywords (normally not used for peer reviews)
@@
 \end{IEEEkeywords}
-}
+}%
 
@@
 \subsection{Beamforming}
-
+\ifPhdDoc
+\begin{figure}[htbp]\centering
+\includegraphics[width=\linewidth]{gfx/buske1\figPostfix.eps}
+\else
 \ifPeerReview
-\begin{figure}[H]\centering
+\begin{figure}[htbp]\centering
 \includegraphics[width=0.8\linewidth]{gfx/buske1\figPostfix.eps}
@@
 \includegraphics[width=\linewidth]{gfx/beamforming.eps}
-\fi%
+\fi\fi%
 \caption{Beamforming principle.
@@
 Or, equivalently, we may apply the weight set to the sum of all the subarrays by
+%
 \begin{align}
@@
 \end{align}
+%
+\ifPhdDoc
+\begin{figure}[tbp]\centering
+\includegraphics[width=\linewidth]{gfx/buske2\figPostfix.eps}
+\else
 \ifPeerReview
@@
 \includegraphics[width=\linewidth]{gfx/implementation.eps}
-\fi%
+\fi\fi%
 \caption{MVDR beamforming. For each pixel in range and azimuth:\newline
@@
 Computing $\eR$ is now the most complex operation whenever $N_LN_K>L$.
+%
+\ifPhdDoc
+\begin{figure}[tbp]\centering
+\includegraphics[width=\linewidth]{gfx/buske3\figPostfix.eps}
+\else
 \ifPeerReview
@@
 \includegraphics[width=\linewidth]{gfx/mvdr_complexity.eps}
-\fi%
+\fi\fi%
 \caption{Per-pixel computational complexity of the steps in MVDR beamforming (before any optimizations).
@@
 
+\ifPhdDoc
+\begin{figure}[tbp]\centering
+\includegraphics[width=\linewidth]{gfx/buske4\figPostfix.eps}
+\else
 \ifPeerReview
@@
 \includegraphics[width=\linewidth]{gfx/mvdr_build_R.eps}
-\fi%
+\fi\fi%
 \caption{Step 1: Building $\eR$.
@@
 
-\subsection{Computing the Spatial Covariance Matrix, $\hat{\boldsymbol{R}}$}\label{computing_eR}
+\subsection[Computing the Spatial Covariance Matrix]{Computing the Spatial Covariance Matrix, $\hat{\boldsymbol{R}}$}\label{computing_eR}
 
@@
 \end{enumerate}
-%
+
+\ifPhdDoc
+\begin{figure}[tp]\centering
+\includegraphics[width=\linewidth]{gfx/buske5\figPostfix.eps}
+\else
 \ifPeerReview
@@
 \includegraphics[width=\linewidth]{gfx/buildR-breakdown.eps}
-\fi%
+\fi\fi%
 \caption{Arithmetic optimization of computing $\eR$: Relative reduction in arithmetic complexity compared to the initial implementation shown in Fig.
@@
 
+\ifPhdDoc
+\begin{figure}[p]\centering
+\includegraphics[width=.997\linewidth]{gfx/buske6\figPostfix.eps}
+% Caption is the same, but using newlines for the items in it (since it is now on a page)
+\caption{MVDR implemented on a GPU.
+We do this in 3 steps, where each step process the full image before moving on to the next step.
+\\[.5\baselineskip]\textbf{Step~1}: The sample covariance matrix $\eR$ is formed by threads running along its diagonals.
+This allows spatial averaging to be implemented in a computationally efficient manner and minimizes interthread communication.
+\\[.5\baselineskip]\textbf{Step~2}: $\eRi\1$ is computed using the heavily optimized batched linear equation solver from Nvidia~\cite{Nvidia}. 
+\\[.5\baselineskip]\textbf{Step~3}: The beamformer output $z$ is computed in a straightforward fashion using $L$ threads that first sum the subarrays and then apply the MVDR weighting function.
+A single thread finally sums all the channels up.}\label{mvdr_implementation}
+\else
 \ifPeerReview
@@
 \includegraphics[width=\linewidth]{gfx/mvdr_implementation.eps}
-\fi%
-\caption{MVDR implementated on a GPU.
+\fi
+\caption{MVDR implemented on a GPU.
 We do this in 3 steps, where each step process the full image before moving on to the next step.
@@
 A single thread finally sums all the channels up.}\label{mvdr_implementation}
+\fi%
 \end{figure}
@@
 
-\subsection{Solving $\hat{\boldsymbol{R}}$$\;\!^{-1}$$\mathbf{1}$}
+\subsection[Inverting R]{Solving $\hat{\boldsymbol{R}}$$\;\!^{-1}$$\mathbf{1}$}
 
@@
 
-\subsection{Computing $z$}
+\subsection[Computing z]{Computing $z$}
 
@@
 
-The GPU design is summarized and compared to our CPU designs in table \ref{implementation_summary}.
-\begin{table}[!h]\centering%\normalsize
+The GPU design is summarized and compared to our CPU designs in Table \ref{implementation_summary}.
+%
+\begin{table}[!ht]\centering%\normalsize
 % \begin{tabular}[c]{l r r r@{}  l}\hline
@@
 
+\ifPhdDoc
+\begin{figure}[tbp]\centering
+\includegraphics[width=\linewidth]{gfx/buske7\figPostfix.eps}
+\else
 \ifPeerReview
@@
 \includegraphics[width=\linewidth]{gfx/plot_holmengraa_L16_Navg1.eps}
-\fi%
+\fi\fi%
 \caption{HISAS sidescan sonar (SSS) image of the shipwreck Holmengraa that lies on a slanted seabed at 77\;m depth outside of Horten, Norway.
@@
 
+\ifPhdDoc
+\begin{figure}[p]\centering
+\includegraphics[width=\linewidth]{gfx/buske8\figPostfix.eps}
+% Same caption but slightly different formatting since it is now on its own page
+\caption{\protect MVDR benchmarks. A 1\;MP image from a $M=32$ channel array was processed for all $L$, and for $K\in\{0,1,2\}$.
+\\[.5\baselineskip]\textbf{Top:} The time the GPU spent on building $\eR$, solving $\eRi\1$, and computing $z$.
+Note the major speedup of building $\eR$ when compared to the complexity plot in Fig.
+\ref{mvdr_complexity}.
+\\[.5\baselineskip]\textbf{Bottom:} Compared to a reference MATLAB and single thread C implementation running on a CPU the GPU offered a speedup of 2-3 orders of magnitude, but these numbers are somewhat misleading.
+If the C implementation was properly optimized we expect the GPU to be no more than a factor 5-10 faster, even if its theoretical peak performance is ~20 times higher than that of the CPU.}\label{benchmarks}
+\else
 \ifPeerReview
@@
 \includegraphics[width=\linewidth]{gfx/benchmark.eps}
-\fi%
-\caption{\protect MVDR benchmarks.
+\fi\caption{\protect MVDR benchmarks.
 A 1\;MP image from a $M=32$ channel array was processed for all $L$, and for $K\in\{0,1,2\}$.\newline \emph{Top:} The time the GPU spent on building $\eR$, solving $\eRi\1$, and computing $z$.
@@
 If the C implementation was properly optimized we expect the GPU to be no more than a factor 5-10 faster, even if its theoretical peak performance is ~20 times higher than that of the CPU.}\label{benchmarks}
+\fi%
 \end{figure}
 
+\ifPhdDoc
+\begin{figure}[p]\centering
+\includegraphics[width=\linewidth]{gfx/buske9\figPostfix.eps}
+\else
 \ifPeerReview
@@
 \includegraphics[width=\linewidth]{gfx/code_assessment.eps}
-\fi%
+\fi\fi%
 \caption{Execution time of an arithmetic-only and a memory-only version of the MVDR code.
@@
 
+\ifPhdDoc
+\begin{figure}[p]\centering
+\includegraphics[width=\linewidth]{gfx/buske10\figPostfix.eps}
+\else
 \ifPeerReview
@@
 \includegraphics[width=\linewidth]{gfx/code_assess_flops.eps}
-\fi
+\fi\fi
 \caption{\protect Code efficiency.
@@
 
+\ifPhdDoc
+\begin{figure}[tbp]\centering
+\includegraphics[width=\linewidth]{gfx/buske11\figPostfix.eps}
+\else
 \ifPeerReview
@@
 \includegraphics[width=\linewidth]{gfx/benchmark_boston.eps}
-\fi
+\fi\fi
 \caption{\protect MVDR benchmarks from Boston HPC centre with the new high-end Nvidia K20 Kepler GPU.
@@
 
-Even with all our efforts we were only able to obtain processing rates of 40\;GFlop/s on average in the desirable range of subarray sizes (\Fig{mvdr_complexity}).
+Even with all our efforts we were unable to obtain processing rates higher than 40\;GFlop/s on average in the desirable range of subarray sizes (\Fig{mvdr_complexity}).
 This is mainly due to the inversion step, since for the most part we obtained more than 100\;GFlop/s when building $\eR$.
@@
 
+\ifPhdDoc
+\clearpage
+\appendix
+% \section*{Appendix}
+\renewcommand\thesection{\Roman{section}}
+\else
 \appendices
+\fi
 
@@
 
-\appendices
+\appendix
 
@@
 
+\ifPhdDoc
+   \printbibliography[title=References,heading=subbibliography]
+%    \bibliographysty
+%    \bibliography{library.bib}
+%    \print
+\else
+
+% Originally build with bibtex:
 % \bibliographystyle{IEEEtran}
@@
 
-\ifPhdDoc\else
 \input{bio/jo_inge_buskenes}

### 2018-02-19 LCAforActiveSonarImaging/document.tex
2_MVDR ok. 3_LCA inlcuded

--- 2016-05-01 LCAforActiveSonarImaging/lca.tex
@@
-\synctex=1
-%%%%%%%%%%%%%%%%                                ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
-% CONDITIONALS %
-%%%%%%%%%%%%%%%%                                ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
-
-\newif\ifPeerReview\PeerReviewfalse              % Whether to create the PeerReview version or
-                                                % Journal version
-\newif\ifOnlineColor\OnlineColortrue            % Compile online color version?
-
-\newif\ifFlatArchive\FlatArchivefalse           % Whether archive is flat (messy) or contain 
-                                                % subfolders for graphics etc.
-\newif\ifFloatAtEnd\FloatAtEndfalse             % Available in PeerReview mode:
-                                                % Place floats at end of document?
-\newif\ifTODO\TODOtrue                        % Use todo notes?
-
-\newif\ifBuildBibliography\BuildBibliographyfalse  % Whether to include the references.bib file or use inline references.
-
-\ifOnlineColor
-   \newcommand\figPostfix{_online}              % Colored images online
-\else
-   \newcommand\figPostfix{_bw}                  % Black and white in journal
-\fi
-
-%%%%%%%%%%%%                                    ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
-% IEEEtran %
-%%%%%%%%%%%%                                    ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
-
-\ifPeerReview
-\documentclass[10pt,journal,draftclsnofoot,onecolumn]{IEEEtran}
-% \newcommand\CLASSINPUTbaselinestretch{1.66}     % http://theoval.cmp.uea.ac.uk/~nlct/latex/thesis/node17.html
-\else
-\documentclass[journal]{IEEEtran}
-\fi
-
-% \RequirePackage[latin1]{inputenc}%              % Set input encoding (optionally latin1)
-% \RequirePackage[T1]{fontenc}%                 % Set font encoding
-% \usepackage[norsk]{babel}
-
-% Single column review mode:
-%\documentclass[12pt,journal,onecolumn]{IEEEtran}
-%\newcommand\CLASSINPUTbaselinestretch{2}
-%\RequirePackage{calc}
-%\RequirePackage{fp} 
-% Left margin = 1 inch + hoffset + oddsidemargin (or evensidemargin)
-% Adding the 2cm gutter width to the odd/even side margins:
-%\setlength\hoffset{0pt}
-%\setlength\oddsidemargin{4cm}       % 4cm margins on the left side
-%\addtolength\oddsidemargin{-1in}    % Subtract the initial 1 inch
-%\setlength\textwidth{21cm-8cm}    % A4 width (21cm) minus margins on either side
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%                  ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
-% IEEE ''APPROVED'' PACKAGES %
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%                  ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
-
-\usepackage{cite}
-
-\ifCLASSINFOpdf
-   \usepackage[dvips]{graphicx}                 % Might not work. Use 'latex' instead of 
-   \ifFlatArchive\else                          % 'pdflatex'
-   \fi
-\else
-   \usepackage[dvips]{graphicx}
-   \ifFlatArchive\else
-   \fi
-\fi
-
-\RequirePackage[table,dvipsnames,svgnames]{xcolor}
-
-\usepackage[cmex10]{amsmath}                    % cmex10 option to be IEEE explore compliant
-\interdisplaylinepenalty=2500                   % Allows multiline equations to be broken
-
-% \RequirePackage{amssymb}
-
-\RequirePackage{array}
-
-\ifCLASSOPTIONcompsoc
-   \usepackage[caption=false,font=normalsize,labelfont=sf,textfont=sf]{subfig}
-\else
-   \usepackage[caption=false,font=footnotesize]{subfig}
-\fi
-
-% \usepackage{caption}
-% \usepackage{subcaption}
-\usepackage{color}
-\usepackage{calc}
-\usepackage{fp}
-
-% 
-% \ifCLASSOPTIONcaptionsoff                       % IEEE promoted hack to turn off captions from the 
-%    \let\MYorigsubfloat\subfloat                 % subfloat package should the captionsoff option
-%    \renewcommand{\subfloat}[2][\relax]{\MYorigsubfloat[]{#2}} % be specified.
-% \fi
-
-\ifFloatAtEnd
-\ifCLASSOPTIONcaptionsoff                       % Places float at the end of the document when the
-  \usepackage[nomarkers]{endfloat}              % captionsoff options is specified to IEEEtrans.cls
-  \let\MYoriglatexcaption\caption               % (PeerReview mode)
-  \renewcommand{\caption}[2][\relax]{\MYoriglatexcaption[#2]{#2}}
-\fi
-\fi
-
-\usepackage{fixltx2e}                           % Fix some twocolumn float problems
-
-% \usepackage{stfloats}                          % Allows: \begin{figure*}[!b]
-                                                % (double column figures on top/bottom)
-
-\usepackage{url}                                % Support for handling and breaking URLs
-
-% NOTE: PDF hyperlink and bookmark features are not required in IEEE
-%       papers and their use requires extra complexity and work.
-% \newcommand\MYhyperrefoptions{bookmarks=true,bookmarksnumbered=true,
-% pdfpagemode={UseOutlines},plainpages=false,pdfpagelabels=true,
-% colorlinks=true,linkcolor={black},citecolor={black},urlcolor={black},
-% pdftitle={Optimal Window Design for the Low Complexity Adaptive Beamformer in Active Sonar Imaging},
-% pdfsubject={},
-% pdfauthor={Jo Inge Buskenes},
-% pdfkeywords={adaptive beamforming, beamforming, complexity, sonar, active}}%
-\ifCLASSINFOpdf
-%    \usepackage[\MYhyperrefoptions,pdftex]{hyperref}
-   \usepackage[pdftex,
-    final,                %Keep hyperlink stuff even when in draft mode
-    breaklinks=true,      %Break links when necessary
-    linktocpage=true,     %Enable link to page?
-    linkcolor=black,  %Colour of links to labels within document
-    citecolor=black,      %Colour of links to the biliography
-    filecolor=red,        %Colour of links to local files
-    pagecolor=black,        %Colour of links to other pages withing document
-    urlcolor=Blue,        %Colour of the links to external URLs
-    colorlinks=true,      %??
-    plainpages=false,     %Store roman/arabic numbering differently to avoid
-    bookmarksnumbered ]{hyperref}
-
-\else
-   \usepackage[\MYhyperrefoptions,breaklinks=true,dvips]{hyperref}
-   \usepackage{breakurl}                        % Allows 'dvips' driver to break links
-\fi
-
-%%%%%%%%%%%%%%%%%%%%%%%                         ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
-% ADDITIONAL PACKAGES %       
-%%%%%%%%%%%%%%%%%%%%%%%                         ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
-
-% \ifPeerReview
-% \let\OldIncludegraphics{\includegraphics}
-% \usepackage{letltxmacro}
-% \LetLtxMacro{\OldIncludegrsaphics}{\includegraphics}
-% \renewcommand{\includegraphics}[2][]{\OldIncludegraphics[width=\linewidth, #1]{#2}}
-% \fi
-
-% % \makeatletter
-% % \def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth
-% % \else\Gin@nat@width\fi}
-% % \makeatother
-% % \let\Oldincludegraphics\includegraphics
-% % \renewcommand{\includegraphics}[1]{\Oldincludegraphics[width=\maxwidth]{#1}}
-
-% \usepackage[maxfloats=40]{morefloats}
-\newcounter{todoidx}
-% \setcounter{todoidx}
-
-\ifTODO
-   \definecolor{todobackground}{rgb}{0.95,0.95,0.95}
-   \setlength\marginparsep{1pt}
-   \setlength\marginparwidth{35pt}
-   \newlength\marginparwidthsmall
-   \setlength\marginparwidthsmall{\marginparwidth}
-   \addtolength\marginparwidthsmall{-7pt}
-   \newcommand\todo[1]{%
-      \addtocounter{todoidx}{1}%
-      {\color{Red}\bf(\thetodoidx{})}%%\fbox{\bf\thetodoidx{}}}%
-      \marginpar{%
-         {\vspace*{-10pt}\color{Red}\fbox{\bf\thetodoidx{}}}\\%
-         \fcolorbox{red}{todobackground}{\parbox{\marginparwidthsmall}{\raggedright\scriptsize #1}}}}
-
-   \newcommand\todopar[1]{\fcolorbox{red}{white}{\parbox{0.97\linewidth}{#1}}}
-\else
-%    \usepackage[disable]{./todonotes} 
-   \newcommand\todo[1]{}
-\fi
-
-\newenvironment{narrow}[2]{%
-\begin{list}{}{%
-\setlength{\topsep}{0pt}%
-\setlength{\leftmargin}{#1}%
-\setlength{\rightmargin}{#2}%
-\setlength{\listparindent}{\parindent}%
-\setlength{\itemindent}{\parindent}%
-\setlength{\parsep}{\parskip}}%
-\item[]}{\end{list}}
-
-% \usepackage{float}
-
-\ifOnlineColor
-   \definecolor{tabBlue}{HTML}{AACCFF}
-\else
-   \definecolor{tabBlue}{HTML}{CCCCCC}
-\fi
-
-%%%%%%%%%%                                      ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
-% MACROS %       
-%%%%%%%%%%                                      ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
-
-\newcommand\graphicsAI[2][]{%
-  \immediate\write18{./bin/laFigure #2 #1}%
-  \input{result}}%
-
-% \DeclareMathOperator*{\argmin}{\text{arg}\;\text{min}}
-
-\newcommand\Fig[1]{Fig.~\ref{#1}}
-\newcommand\eq[1]{(\ref{#1})}
-
-\newcommand\Grey[1]{{\color{Grey}#1}}
-\newcommand\Red[1]{{\color{Red}#1}}
-\newcommand\Blue[1]{{\color{Blue}#1}}
-\newcommand\DarkBlue[1]{{\color{DarkBlue}#1}}
-\newcommand\LightBlue[1]{{\color{LightBlue}#1}}
-\newcommand\Brown[1]{{\color{Brown}#1}}
-\newcommand\Green[1]{{\color{Green}#1}}
-\newcommand\SeaGreen[1]{{\color{SeaGreen}#1}}
-\newcommand\Yellow[1]{{\color{yellow}#1}}
-\newcommand\Orange[1]{{\color{orange}#1}}
-
-\newcommand\nn{\nonumber\\}
-
-\newcommand\nmat[1]{\begin{matrix}#1\end{matrix}}
-\newcommand\bmat[1]{\begin{bmatrix}#1\end{bmatrix}}
-\newcommand\case[1]{\begin{cases}#1\end{cases}}
-\newcommand\textbox[2]{\footnotesize\text{\parbox{#1}{\centering\emph{#2}}}}
-
-\newcommand\rand{\text{rand}}
-\newcommand\randn{\text{randn}}
-\newcommand\rect{\text{rect}}
-\newcommand\sinc{\text{sinc}}
-\newcommand\tr{\text{tr}}
-\newcommand\adj{\text{adj}}
-
-% \newcommand\max{\text{max}}
-\newcommand\argmin[1]{\text{arg}\;\underset{#1}{\text{min}}}
-
-\newcommand\qqquad{\quad\qquad}
-\newcommand\qqqquad{\qquad\qquad}
-
-% \renewcommand\l[1]{\left#1}
-% \renewcommand\r[1]{\right#1}
-
-% {\text{\parbox{1.5cm}{\centering volume hyper- sphere}}}
-
-%Keyword colouring:
-\newcommand\kw[1]{#1}
-\newcommand\parm[1]{#1}%\color{Black}#1\color{Black}}
-
-\newcommand\of[1]{\scriptstyle(\parm{#1})\displaystyle}
-\newcommand\df[1]{\scriptstyle[\parm{#1}]\displaystyle}
-\newcommand\var[3]{#1_\text{#2}\of{#3}}
-
-\newcommand\diag{\text{diag}}
-
-% \raisebox{lift}[extend-above-baseline][extend-below-baseline]{text}
-\newcommand\mt[1]{\text{\emph{#1}}} %mt = mathtext
-\newcommand\mathnorm{\textstyle}
-\newcommand\mathbig[1]{\displaystyle#1\mathnorm}
-\newcommand\mathsmall[1]{\scriptstyle#1\mathnorm}
-\newcommand\mathtiny[1]{\scriptscriptstyle#1\mathnorm}
-\newcommand\sfrac[2]{\scriptstyle\raisebox{0.25pt}[0pt][0pt]{$\frac{#1}{#2}$}\mathnorm}
-\newcommand\nfrac[2]{\textstyle\frac{#1}{#2}\displaystyle}
-
-\newcommand\sumu[1]{\sum\limits^{#1}\;}
-\newcommand\suml[1]{\sum\limits_{#1}\;}
-\newcommand\sumb[2]{\sum\limits_{#1}^{#2}\;}
-
-\newcommand\produ[1]{\prod\limits^{#1}\;}
-\newcommand\prodl[1]{\prod\limits_{#1}\;}
-\newcommand\prodb[2]{\prod\limits_{#1}^{#2}\;}
-
-\newcommand\defeq{\overset{\underset{\mathrm{def}}{}}{=}}
-
-%Math macros:
-\newcommand\T{^{\scriptscriptstyle T}}
-\renewcommand\H{^{\scriptscriptstyle H}}
-
-\renewcommand\vec[1]{\boldsymbol{#1}}
-\newcommand\mat[1]{\boldsymbol{#1}}
-
-\newcommand\Om{O_\text{m}}
-\newcommand\Oa{O_\text{a}}
-\newcommand\Nl{N_\text{l}}
-\newcommand\Nk{N_\text{k}}
-\newcommand\1{\vec 1}
-\newcommand\I{\mat I}
-\renewcommand*\a{\vec a}
-\newcommand*\f{\vec f}
-\renewcommand*\i{\vec i}
-\renewcommand*\k{\vec k}
-\newcommand*\n{\vec n}
-\newcommand*\p{\vec p}
-\newcommand*\s{\vec s}
-\newcommand*\w{\vec w}
-\newcommand*\x{\vec x}
-\newcommand*\y{\vec y}
-
-\newcommand*\A{\mat A}
-\newcommand*\B{\mat B}
-\newcommand*\C{\mat C}
-\newcommand*\E{\mat E}
-\renewcommand*\P{\mat P}
-\newcommand*\eP{\mat{\hat P}}
-\newcommand*\R{\mat R}
-\newcommand*\Ri{\R^{-1}}
-\newcommand*\eR{\mat{\hat R}}
-\newcommand*\eRi{\hat{\mat R}\;\!^{-1}}
-\newcommand*\Navg{N_\text{avg}}
-\newcommand*\W{\mat W}
-\newcommand*\X{\mat X}
-\newcommand*\Xd{\X_{\!\Delta}}
-\newcommand*\Y{\mat Y}
-
-\renewcommand\argmin{\text{argmin}}
-
-\renewcommand*\P{\mat P}
-\newcommand*\V{\mat V}
-\newcommand*\M{\mat M}
-
-\renewcommand*\L{\mat \Lambda}
-\newcommand*\U{\mat U}
-% \renewcommand*\t{\mathtiny{^T}}
-% \newcommand*\h{\mathtiny{^H}}
-\renewcommand*\t{^T}
-\newcommand*\h{^H}
-
-\newcommand\minus{\scalebox{0.75}[1.0]{$-$}}
-
-% \usepackage{tikz}
-% \usetikzlibrary{shapes,snakes}
-\usepackage{amsmath,amssymb}
-% \usepackage{datatool}
-% \usepackage{glossaries}
-
-% \newenvironment{outline}
-% {\begin{itemize}}
-% {\end{itemize}}
-%
-
-\ifPeerReview
-\graphicspath{./submission/submit1}
-\fi
-
-% \newcommand\multimedia[1]{\textbf{{\color{red}[#1]}}}
-\newcommand\multimedia[2]{\href{#1}{#2}}
-
-% \newcommand\mediaPath{gfx/media}
-\newcommand\articlePath{http://folk.uio.no/joibu/articles/2015_JOE_LCA}
-
-\newcommand\mediaI{\multimedia{\articlePath/media1.mp4}{Media Movie~1}}
-\newcommand\mediaII{\multimedia{\articlePath/media2.mp4}{Media Movie~2}}
-
-\begin{document}
-
 \title{Low Complexity Adaptive Sonar Imaging}
@@
 % 
-% \begin{itemize}
-% \item Dynamic weights $\Rightarrow$ greater contrast/resolution potential
-% \item But:
-% \begin{itemize}
-% \item Active system $\Rightarrow$ noise and signal correlated $\Rightarrow$ adaptive beamformers no longer robust.
-% \item Subarray averaging decorrelate the noise, but in the process we sacrifice resolution.
-% \item Computationally complex O($M^3$) due to inversion of a spatial covariance matrix.
-LCA is only O($MW$).
-Even if Capon subspace/beamspace versions exist, they remain more complex than the LCA, and is not as ideally suited for implementation of SIMD hardware (GPUs, DSPs, ...).
-% \end{itemize}
-% \item LCA is inherently robust, is ``parameter free'' once setup with a well designed window set, and can often be implemented with minor modifications to already existing DAS beamformers.
-% 
-% The angular resolution and contrast in active sonar images depend on the beamformer's ability to receive signals from directions of interest, while suppressing noise and interference emanating from other directions.
-For sonar arrays, this is achieved by applying weights to the array channels.
-% 
-% While classical beamformers use predefined windows, adaptive beamformers estimate the optimal window by analytical evaluation of the data.
-The minimum variance (MV) beamformer, for instance, calculates the set of weights that minimises the variance of the beamformer's output.
 % 
@@
 
-% An attactive trait of the LCA is its low computational complexity; while the MV beamformer is of O($M^3$), the LCA method is of O($MW$), with $M$ being the number of channels and $W$ the number of windows.
-We made the LCA perform like the MV method using a well designed set of 30 windows.
-Hence, unless the array is very small, the proposed method will perform like the MV beamformer or better, and at a fraction of the computational cost.
-
 % \end{itemize}
@@
 
-% (Optional) Add some extra info on cover page of peer review papers:
-% \ifCLASSOPTIONpeerreview
-% \begin{center} \bfseries EDICS Category: 3-BBND \end{center}
-% \fi
-
 % Insert page break and insert second title (peer review mode)
@@
 \section{Introduction}
-
-% - Nature and scope
-% - Past work
-% - Methods
-% - Brief summary results
-% - Brief conclusions
-
-% The promise is already demonstrated in nature.
-% 
-% The sonar systems of the future 
-% 
-% Future sonar systems 
-% 
-% Great promise in adaptivity
-% 
-% The greatest untapped 
-% 
-% The sonars of the future might benefit greatly from adaptivity.
-% 
-% 
-% Future sonars will likely be
-
-% To make a sonar truly perform better, we need to make it smarter. 
-% 
-% \IEEEPARstart{F}{uture} sonar systems will merit a much more sophisticated level of adaptivity. 
-% 
-% \IEEEPARstart{T}{he} next generation sonar systems will be adaptive.
-% 
-% Making sonars adaptive is the way to take them into the future. 
-% 
-% The virtue of future sonar systems will be sophisticated adaptivity. 
-% 
-% Intelligent
-% 
-% Reactive
-% 
-% Observe, process, adapt, act. 
-% 
-% Adaptivity - sense, process, respond
-% 
-% Faster, smarter
-% 
-% The best sonars in nature are adaptive. 
-% Nature allures us to be this bold because
 
@@
 
-% While using weights that span a bigger space gives LCA greater freedom, not The bigger the space the weights span, the greater the adaptive freedom, but only the solutions that are valid in the constrained solution space can make a difference.
-
-% Instead of adaptively computing the weights, it adaptively selects them.
-
 Synnev\aa{}g \emph{et al}.
@@
 
-% This article is outlined as follows: In Section \ref{methods}, we introduce the concept of adaptive beamforming and provide details on the MVDR and LCA method.
-Then, in \ref{maptogpu}, we investigate the complexity issue, discuss means for reducing arithmetic complexity, and detail an implementation that makes efficient use of the GPU's parallel resources.
-The final design is assessed in Sections \ref{images_and_benchmarks} and \ref{discussion}, where we provide benchmarks, comparisons with similar CPU implementations, and measures of how efficiently our implementation makes use of the GPU's resources.
-
-% - How much oversampling
-% - How many parameters, how great a span.
-% 
-% - Experiementl data, Holmengraa/cross
-% 
-% - Note on efficiency.
-% 
-% 
-% From our work with MVDR it 
-% 
-% 
-% Traditionally beamformers were static, but as processors get more powerful so with the emergence of more processing power
-
-% 
-% \section{Methods}
-% 
-% At heart of the image formation process is a technique called beamforming, which creates a focus point at the pixel of interest.
-It does this by applying a suitable set of delays and weights to the array elements, such that signals emanating from the pixel location are emphasized constructively, while noise and interference from other directions sum destructively.
-% 
-% In active sonar array imaging, an acoustic wave is transmitted and the received echoes are usually recorded using an array of sensors.
-Each of these may be independently delayed and weighted such that signals emanating from directions of interest are added constructively, while noise and interference from other directions add destructively.
-Additionally, a window may be applied to the sensor channels to further adjust the array's spatial response.
-This process is known as beamforming.% The process of combining the signals from each sensor is known as beamforming.
-% 
-% 
-% 
-% 
-% % Traditional beamformers such as Delay-and-Sum (DAS) apply predefined windows to all incoming data.
-However, due to the non-stationary nature of sonar data, the optimal window for any given time instant will generally differ from the next.
-This is where adaptive beamformers thrive, because they compute the optimal window coefficients for the data at each time instant.
-The choice of optimisation criteria is what mainly differentiates the various adaptive beamformers.
-The Minimum Variance (MV) beamformer, for instance, selects the set of weights that minimizes the beamformer output power for any given time instant~\cite{cap69}.
-% 
-% 
-% \section{Background}
-% 
-% The MV beamformer computes the window coefficients by estimating and inverting a spatial covariance matrix.
-For improved estimation accuracy the data is averaged in space or time, or both.
-In addition, a certain amount of diagonal loading is often added to the covariance matrix before inversion.
-These actions, while leading to statistical robustness, tend to deteriorate the beamformer's performance.
-Also, the inversion of the covariance matrix has a computational complexity of O($M^3$), where $M$ is the number of elements.
-The computational burden makes the MV beamformer for large arrays less attractive, and sometimes even infeasible.
-% 
-% Based on the proposed method by Synnevåg~\cite{syn11}, we have implemented the LCA beamformer that keeps the minimum variance optimisation criterion but reduces the solution space to a discrete set of predefined windows.
-This reduces the computational complexity to O($MW$), where $W$ is the number of windows in the set.
-We use the Kaiser window function because it allows us to design a wide range of windows with different mainlobe widths and sidelobe suppression by adjusting the tradeoff parameter $\beta$~\cite{kai66}.
-In addition we apply steering to each of the windows and constrain the window design to ensure unit gain in the look direction.
-% 
-% The following will show that the proposed method performs similarly to or better than the MV method.
-
 \section{Receive Beamforming}\label{sec:beamforming}
@@
 \begin{align}
-\underset{\w[n]}{\argmin}\, E\{\big|z[n]\big|^2\} &= \underset{\w[n]}{\argmin}\, \w[n]\R[n]\w\H[n]\nn
+\argmin{\w[n]}\, E\{\big|z[n]\big|^2\} &= \argmin{\w[n]}\, \w[n]\R[n]\w\H[n]\nn
 \text{subject to } \w^H[n]\a_\phi &= 1.\label{eq:mvdr_definition}
@@
 
+\ifPhdDoc
+   \begin{figure}[tbp]\centering%
+   \includegraphics[width=\linewidth]{gfx/fig1.eps}%
+\else
    \begin{figure}[tbp]\centering%
@@
       \includegraphics[width=\linewidth]{gfx/buske1_online.eps}%
-\fi%
+\fi\fi%
 \caption{Each Kaiser window is steered in the interval $\phi\in[0, \phi_{\mathrm{3\,dB}}(\beta)]$.
@@
 
+\ifPhdDoc
+   \begin{figure}[tbp]\centering%
+   \includegraphics[width=\linewidth]{gfx/fig2.png}%
+\else
    \begin{figure}[tbp]\centering%
@@
       \includegraphics[width=\linewidth]{gfx/buske2_online.png}%
-\fi%
+\fi\fi%
 \caption{The submerged object used to test beamforming resolution is a 1\,m by 1\,m test cross attached to an anchor with a diameter of approximately 13\,cm.
@@
 
+\ifPhdDoc
+   \begin{figure*}[tbp]\centering%
+   \includegraphics[width=\linewidth]{gfx/fig3.pdf}%
+\else
    \begin{figure*}[t]\centering%
    \includegraphics[width=\linewidth]{gfx/buske3_online.pdf}%
+\fi%
 \caption{\emph{Determining LCA window type}: MVDR image with typical frequency responses for windows used in various pixel regions.\newline
@@
 
+\ifPhdDoc
+   \begin{figure*}[tbp]\centering%
+   \includegraphics[width=\linewidth]{gfx/fig4.pdf}%
+\else
    \begin{figure*}[t]%
    \includegraphics[width=\textwidth]{gfx/buske4_online.pdf}%
+\fi%
 \caption{\emph{Determining Kaiser parameter boundaries}: LCA images of the resolution test cross and anchor created with a large window database, but where the upper bounds for $\beta$ and $\phi$ are varied.
@@
 
-% \setcounter{dbltopnumber}{2}
-
+\setcounter{dbltopnumber}{2}
+
+\ifPhdDoc
+   \begin{figure*}[t]\centering
+   \subfloat[$\beta$-values of the Kaiser window chosen for each image pixel.
+Note how LCA prefers a narrow response ($\beta=0$) on the anchor, and on other strong sources with little lateral interference.
+On the sources where lateral interference is present it chooses the widest response with the best sidelobe suppression.]{\includegraphics[width=\textwidth]{gfx/fig5a.pdf}%%
+   \label{oversampling_mosaic_bounds_beta}}
+   \hfil
+   \subfloat[$\phi$-values of the Kaiser window chosen for each image pixel.
+On strong sources LCA selects windows that are steered away from the source.
+This is what improves the FWHM measurement in \Fig{oversampling_mosaic_bounds}.]{\includegraphics[width=\textwidth]{gfx/fig5b.pdf}%
+   \label{oversampling_mosaic_bounds_phi}}
+   \caption{Kaiser windows chosen for each pixel in the image of the resolution test cross.
+The underlying image is the one shown in \Fig{oversampling_mosaic_bounds}, with the same parameters.
+The location of the anchor, cut line and main scatter locations of the cross is marked in red.}
+   \label{fig_sim}
+   \end{figure*}
+\else
    \begin{figure*}[t]\centering
@@
    \end{figure*}
-
+\fi
+
+\ifPhdDoc
+   \begin{figure*}[tbp]\centering%
+   \includegraphics[width=\linewidth]{gfx/fig6.pdf}%
+\else
    \begin{figure*}[t]%\centering%
    \includegraphics[width=\textwidth]{gfx/buske6_online.pdf}%
+\fi
 \caption{\emph{LCA images using different window database sizes}.
@@
 
+\ifPhdDoc
+   \begin{figure*}[t]\centering
+   \subfloat[\emph{Image quality of approximate point scatterers found in the resolution test cross scene}.
+Two lateral image cuts are shown in the leftmost figures, along with the response of a rectangularly weighted DAS.
+Note how the FWHM of LCA sits in between that of DAS and MVDR.
+Also note how LCA is insensitive to the window type; its performance is near identical whether it uses Kasier or trigonometric windows.]{\includegraphics[width=\textwidth]{gfx/fig7a.pdf}%
+   \label{beamformer_comparison_cross}}
+   \hfil
+   \subfloat[\emph{Image quality of the full sector Holmengraa scene}.
+Compared to DAS the adaptive beamformers produce deeper shadows, sharper edges and a higher detail level.
+The adaptive methods produce nearly identical images.]{\includegraphics[width=\textwidth]{gfx/fig7b.pdf}%%
+   \label{beamformer_comparison_holmengraa}}
+   \caption{Comparing image quality of DAS, MVDR and LCA with Kaiser or trigonometric windows.}\label{image_quality}
+   \end{figure*}
+\else
    \begin{figure*}[t]\centering
@@
    \end{figure*}
-
-% \begin{figure*}[t]%\centering%
-% \includegraphics[width=\textwidth]{gfx/oversampling_mosaic_bounds_lca_windows_beta.pdf}%
-% \caption{$\beta$-values of the windows LCA selects for each image pixel.
-The underlying image is the one shown in \Fig{oversampling_mosaic_bounds}, with the same parameters.
-The location of the anchor and cut line is marked in red.
-Note how LCA prefers a rectangular window ($\beta=0$) on the anchor.}\label{oversampling_mosaic_bounds_beta}
-% \end{figure*} 
-% 
-% \begin{figure*}[t]%\centering%
-% \includegraphics[width=\textwidth]{gfx/oversampling_mosaic_bounds_lca_windows_phi.pdf}%
-% \caption{$\beta$-values of the windows LCA selects for each image pixel.
-The underlying image is the one shown in \Fig{oversampling_mosaic_bounds}, with the same parameters.
-The location of the anchor and cut line is marked in red.
-Note how LCA prefers a rectangular window ($\beta=0$) on the anchor.}\label{oversampling_mosaic_bounds_beta}
-% \end{figure*}
-
-% % \setcounter{dbltopnumber}{1
-
-% \begin{figure*}[t]%\centering%
-% \includegraphics[width=\textwidth]{gfx/beamformer_comparison_cross.pdf}%
-% \caption{}\label{beamformer_comparison_cross}
-% \end{figure*}
-% 
-% \begin{figure*}[t]%\centering%
-% \includegraphics[width=\textwidth]{gfx/beamformer_comparison_holmengraa.pdf}%
-% \caption{}\label{beamformer_comparison_holmengraa}
-% \end{figure*}
+\fi
 
@@
 \begin{align}
-\underset{p}{\argmin}\, E\{\big|z_p[n]\big|^2\} = \underset{p}{\argmin}\, E\big\{\big|\w_{p}\H\x[n]\big|^2\big\}\nn
+\argmin{p}\, E\{\big|z_p[n]\big|^2\} = \argmin{p}\, E\big\{\big|\w_{p}\H\x[n]\big|^2\big\}\nn
 \qquad\text{subject to}\qquad \w\H\a_\phi = 1.\label{eq:lca_definition}
@@
 %
-% \begin{itemize}
-% \item \emph{Window function}
-% \item \emph{Parameter boundaries}
-% \item \emph{Number of windows needed}
-% \item \emph{Oversampling need}
-% \item \emph{Image quality}
-% \end{itemize}
 
@@
 Upon display the images were all bilinearly upinterpolated to the same size.
-% 
-% \todo{Need $M$ in the image sector. Why???}
-
-% Idea to steer within a fraction of the rectangular window's mainlobe.
-
-% \begin{align}
-% d = \frac{\pi\,\theta_\text{open}\,\lambda}{M\,d\,O_s}
-% \end{align}
-
-% See section \ref{kaiser_windows}.
-
-% Narrow ones:
-% - Best coherent gain, lowest equivalent noise bandwidth, and most sensitive to steering meaning not needing to steer as much for some effect.
-% 
-% 
-% Beamforming:
-% - Maximize resolution
-% - Minimize noise/error, both coherent and incoherent
-% - Minimize signal cancellation
-% 
-% Gauss: Miminized time-bandwidth product
-%   - 
-%   
-% Rectangular
-%   - 
-% Dolph-Cheb: Minimize mainlobe width given sidelobe suppression
-% Kaiser: For a restricted energy and time duration, maximize energy in the band of frequencies W.
-% 
-% Adaptive windows perform best in detection of closely separated angles of significantly amplitudes.
-
-% Adaptive beamformers usually compute the window coefficients by estimating and inverting a spatial covariance matrix.
-There are two inherent problems to this procedure.
-First, for improved estimation accuracy the data is either averaged in space or time, and the significance of the covariance matrix can be reduced using a regularisation parameter~\cite{Carl Inge}.
-These are all attempts to constrain the beamformer to ensure the window coefficience are not over-adapted to the data.
-Second, inverting the covariance matrix has a computational complexity of O($M^3$)~\cite{Carl Inge}.
-For larger arrays the computational burden becomes significant, and the question arises whether this processing power can be better utilised.
-% 
-%
-
-% 
-% This selection is based on observations of the solutions found by the MV beamformer.
-We obtained good results in our experiments by letting the LCA beamformer select among 5 Kaiser windows with uniformly distributed $\beta$'s in the range $[0.05, 0.5]$ and a rectangular window.
-Each window was steered in 5 different directions, uniformly distributed within 80\% of the mainlobe width of the rectangular window, which is the most narrow.
-This adds to a total of 30 windows.
-% 
-% To make the choice of window less susceptible to pixel value uncertainties, the LCA beamformer was set to apply the most frequently selected window in an 11 pixel range region to the center pixel in that region.
-In physical terms, this is a region of approximately 20\,cm.
-The MV beamformer estimated the covariance matrix as described in~\cite{syn07} by averaging over 16 subarrays and 11 range pixels, and applying 3\% diagonal loading.
 
@@
 
-\subsection{Kaiser parameter $\beta$ and steering $\phi$}\label{sec:results_window_parameters}
+\subsection[Kaiser parameter β and steering Φ]{Kaiser parameter $\beta$ and steering $\phi$}\label{sec:results_window_parameters}
 
@@
 
-% Furthermore, the algorithm maps well to single-instruction-multiple-data (SIMD) hardware such as graphics processing units (GPUs).
-We show that even a modest amount of optimisation work has resulted in a factor 10 speed-up of this algorithm when implemented on a GPU as opposed to on a CPU.
-
-% the performance of the LCA beamformer is surprisingly similar regardless of what windows it is allowed to choose from.
-
-% \newcommand\cc[1]{\multicolumn{1}{>{\columncolor{tabBlue}}c}{\bf #1}}
-% \begin{table}[!b]\centering%\normalsize
-% \begin{tabular}[c]{l r r r@{}  r r r}\hline
-% \rowcolor{tabBlue} & \cc{Resolution} & \cc{S} & \cc{$\beta$} & \cc{$\varphi^*$} &  \cc{N$_\beta$} & \cc{N$_\varphi$} \\\hline
-% & 50\%  & 28 & [0,3] & \hspace{.15cm} 50\% & 2 & 3 \\
-% & 50\%  & 28 & [0,6] & \hspace{.15cm} 50\% & 2 & 3 \\
-% & 50\%  & 28 & [0,9] & \hspace{.15cm} 50\% & 2 & 3 \\
-% & 100\% & 32 &  &:30 \\
-% & 150\% & 32 & 1 &:4\\
-% \end{tabular}\\
-% * Steering is relative to the \minus{}3\,dB width of the window being applied.
-% \caption{Qualitative study}\label{parameter_study}
-% \end{table}%
-
-% The simulations contained a few highlighted cylindric regions with 2\,m diameter and a constant intensity 15.4\,dB over the average speckle level.
-We focused on an object centered at 41\,m range and -3 degrees azimuth angle.
-The images produced by the LCA, MV and unweighted DAS for a single speckle realisation and a mean image of all 100 realisations are shown in \Fig{speckle}.
-Each image was normalised by their respective speckle level, and the dynamic range was clamped at \{-30,~15\}\,dB.
-The physical extent of the of the simulated object was superimposed for reference.
-%\ref{saf09}
-% 
-% In \Fig{speckle} we observe that both the adaptive beamformers produce images with a more clearly defined shadow.
-The edges are sharper and the shadow more distinct.
-This is because the adaptive beamformers have better sidelobe suppression, and thus allow less signal energy to leak from the speckle region into the shadow region.
-The same effects can also be observed laterally around the highlight region of the mean image; DAS is more prone to smear energy into the speckle region than the adaptive beamformers are.
-% 
-% \Fig{cutmean} displays two lateral cuts in the mean image, one through the highlight at 41.1\,m range and one through the shadow at 44.5\,m range.
-Each cut was computed as the mean of a 40\,cm range band centered at the respective ranges.
-The transition region between highlight and shadow is shortest for the LCA, which is advantageous since it results in a more accurate representation of the object size.
-The slightly inferior performance of the MV beamformer is due to the subarray averaging, which makes the effective array smaller~\cite{syn07}.
-% 
-% By studying the LCA performance on the 100 speckle realisations, we found the choice of window to be highly dependent on the phenomena being imaged.
-In speckle regions LCA favored narrow spatial responses with minor steering, while wide and fully steered responses were often selected in shadow regions.
-There was also a clear distinction between the responses selected in sidelobe regions and in speckle regions.
-% 
-% % \Fig{hist} illustrates how often a particular window is chosen in general.
-There are 10 $\phi$'s for each of the 10 $\beta$'s.
-The first 10 values are rectangular windows.
-Value 5 is a non-steered rectangular window, and we see that this is most frequently selected.
-At higher steering angles the windows are rarely selected, and thus we might get better results by tightening our steering boundaries.
-At value 15 we find inverted Kaiser windows with $\beta = 0.05$, and then $\beta$ is increased in increments of 0.505 for every 10th value.
-At value 65 the $\beta$ value is 2.025, and these and the remaining windows are hardly selected.
-% 
-% A sidescan image of the 1500 dwt oil tanker wreck Holmengraa is shown in \Fig{holmengraa}.
-It is about 68\,m long and 9\,m wide, and lies on a slanted seabed at 77\,m depth~\cite{holmengraa}.
-The sidescan image was created using data from the HISAS 1030 sonar, which is rather unsuited for this purpose because of its large opening angle.
-This, and the fact that the wreck was imaged at a range of about 105\,m makes the image quality poor, but sufficient to compare our beamformers.
-% 
-% In the Holmengraa image we note again that the LCA produces a cleaner shadow and better edge definition. The MV method performed almost identically to LCA in this case, and was omitted.
-
 \section{Conclusion}\label{sec:conclusion}
-
-% \begin{table}[!b]\centering%\normalsize
-% \begin{tabular}[c]{l c c c}\hline
-% \rowcolor{tabBlue} & \bf Highlight & \bf Shadow & \bf Speckle  \\\hline
-% Symmetry             & Moderate  & High   & High \\
-% Mainlobe width       & Narrow    & Wide   & Moderate \\
-% Sidelobe suppression & Low       & High   & Moderate \\
-% Steering             & Low       & Up to \minus{}3\,dB width & Moderate \\
-% Oversampling         & 2x-8x     & 2x-8x  & 2x-8x
-% \end{tabular}
-% \caption{Summary}\label{tab:summary}
-% \end{table}%
 
@@
 
-% The tendency of LCA to produce images of similar quality as MVDR seems to indicate that a full solution space is rarely needed in real world scenarios.
-
-%  It initially very is also very fast, and is very well suited for acceleration using  very fast, easy to implement even on parallel hardware,
-
-% Its computational complexity is only of O($N_wM$), and its data parallel nature makes it very well suited for acceleration using e.g. GPUs. compared to MVDR's O($M^3$).
-% 
-% LCA is largely insensitive to the choice of windows and steering. 
-% LCA works well for all the boundaries
-% Variables of interest:
-
-% We limit our study to the use of variations of the parametric Kaiser-Harris window, since we observed neglible performance difference to using this function or any of the common trigonometric ones.
-The Kaiser windows are easy to compute, is fairly insensitive to coefficient inaccuracies, and can create windows with shapes spanning from rectangular to Gaussian.
-% 
-% 
-% Using the LCA beamformer results in increased robustness and improved edge definition compared to the MV and DAS beamformers.
-
-% The images produced by LCA appears sharper than those of MVDR, but 
-% 
-% The key to achieve this success lies in the design of the window set.
-Windows that yield vagely steered narrow responses are preferred in speckle regions, while wide and steered responses are typically preferred in highlight and shadow regions.
-% 
-% A window set of 6 different responses each steered in 5 different directions proved sufficient in our experiment to match the performance of the MV beamformer.
-
-% \begin{itemize}
-% \item Adaptive beamformer's potential lies in its ability to suppress interference power
-% \item Why adaptive beamformers struggle in active sonar systems.
-Correlated noise, robustification kills the adaptive potential.
-Quite computationally intensive.
-Constraints must be applied in one way or another - parameters must be tuned.
-% \begin{itemize}
-% \item Noise and signal is correlated. Spatial averaging required. 
-% \item Increases variance of speckle (not only in active systems?), spatial compunting \cite{Vignon2009} (US) or time averaging?.
-Subarray averaging applied here \cite{Synnevag2007} (US) \todo{Is it bad to mix ultrasound/ sonar refs?}
-% \end{itemize}
-% \item Outline: 
-% \begin{itemize}
-% \item Setup the window set for LCA to match this.
-% \item Notice that the LCA performs like Capon, even with a small window database.
-% \item Start with Kaiser, see that other window functions also work well.
-% \end{itemize}
-% \end{itemize}
-% \todopar{Out of the blue idea: Weigh CF with LCA? This guy did it with MVDR \cite{Wang2009}. Probably you know already... Okay, back on track.}
-
-% \IEEEPARstart{T}{o} form images from a modern phased array sonar system the received wavefield is usually recorded, and then postprocessed by a digital beamformer.
-The beamformer applies delays and weights to the sensor channels, the beamformer adjusts the arrays spatial response to focus at one pixel at a time.
-such that signals emanating from regions of interest add constructively, while ensuring that noise and interference from other angles do not.
-% 
-% The imaging capabilities of a modern phased array sonar system depend on physical attributes such element response and array geometry, the transmitted signal, as well as the beamforming method being used on transmission and reception.
-Beamforming is the concept of applying delays and weights to the sensors channels to steer the arrays response to points of interest.
-
-% 
-% 
-% Outline:
-% \begin{itemize}
-% \item Choice of window functions makes little difference.
-% \item Steering and mainlobewidths have outer bounds.
-% \item Beamspace?
-% \item Chosen window plots - what may they tell us? Variance intensity values when using various windows.
-% \item Assymmetric windows?
-% \end{itemize}
-%
-
-% , and maximum suppression of while ensuring that the beamformer digitally  before each of the pixels are estimated one at a time.
-The resolution and contrast of such a system will depend on the systems spatial response, which ideally should be narrow  be very sharp in the desired direction its ability to achieve  fundamental principle of forming a sonar image is to record the received wavefield,
-
 %%%%%%%%%%%%%%%%%%                              ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
@@
 \begin{figure}[t]
+\ifPhdDoc\centering
+   \includegraphics[width=\linewidth]{gfx/fig8.eps}%
+\else
    \ifPeerReview\centering%
@@
       \includegraphics[width=\linewidth]{gfx/buske8_online.eps}%
+   \fi
 \fi%
@@
 
+\ifPhdDoc
+\typeout{heyheyhey}
+   \printbibliography[title=References,heading=subbibliography]
+%    \bibliographysty
+%    \bibliography{library.bib}
+%    \print
+\else
+\typeout{hohoho}
    \ifBuildBibliography
@@
 
-\end{document}
+\fi

### 2018-02-19 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2016:07:28	12:02:03	(to.28.juli)
2_MVDR ok. 3_LCA inlcuded

--- 2016-03-03 LCAforActiveSonarImaging/lca.tex
@@
 
-% \RequirePackage[latin1]{inputenc}%              % Set input encoding (optionally latin1)
-% \RequirePackage[T1]{fontenc}%                 % Set font encoding
-% \usepackage[norsk]{babel}
-
-% Single column review mode:
-%\documentclass[12pt,journal,onecolumn]{IEEEtran}
-%\newcommand\CLASSINPUTbaselinestretch{2}
-%\RequirePackage{calc}
-%\RequirePackage{fp} 
-% Left margin = 1 inch + hoffset + oddsidemargin (or evensidemargin)
-% Adding the 2cm gutter width to the odd/even side margins:
-%\setlength\hoffset{0pt}
-%\setlength\oddsidemargin{4cm}       % 4cm margins on the left side
-%\addtolength\oddsidemargin{-1in}    % Subtract the initial 1 inch
-%\setlength\textwidth{21cm-8cm}    % A4 width (21cm) minus margins on either side
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%                  ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
-% IEEE ''APPROVED'' PACKAGES %
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%                  ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
+%%%%%%%%%%%%%%%%%%%%%%%%%%                      ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
+% IEEE APPROVED PACKAGES %
+%%%%%%%%%%%%%%%%%%%%%%%%%%                      ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
 
@@
 \usepackage{fp}
-
-% 
-% \ifCLASSOPTIONcaptionsoff                       % IEEE promoted hack to turn off captions from the 
-%    \let\MYorigsubfloat\subfloat                 % subfloat package should the captionsoff option
-%    \renewcommand{\subfloat}[2][\relax]{\MYorigsubfloat[]{#2}} % be specified.
-% \fi
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%                         ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
-
-% \ifPeerReview
-% \let\OldIncludegraphics{\includegraphics}
-% \usepackage{letltxmacro}
-% \LetLtxMacro{\OldIncludegrsaphics}{\includegraphics}
-% \renewcommand{\includegraphics}[2][]{\OldIncludegraphics[width=\linewidth, #1]{#2}}
-% \fi
-
-% % \makeatletter
-% % \def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth
-% % \else\Gin@nat@width\fi}
-% % \makeatother
-% % \let\Oldincludegraphics\includegraphics
-% % \renewcommand{\includegraphics}[1]{\Oldincludegraphics[width=\maxwidth]{#1}}
 
@@
 
-\title{Low Complexity Adaptive Sonar Imaging}
-
-\author{Jo~Inge~Buskenes, %
-        Roy~Edgar~Hansen, %
-        Andreas~Austeng%
-\IEEEcompsocitemizethanks{\IEEEcompsocthanksitem All authors are with the Department of Informatics, University of Oslo, Norway.}%
-\IEEEcompsocitemizethanks{\IEEEcompsocthanksitem R.~E.~Hansen is also with the Norwegian Defense Research Establishment (FFI), Norway.}% <-this % stops a space
-
-% \thanks{Manuscript received April 19, 2005; revised January 11, 2007.}
-}
-
-% The paper headers
-\markboth{IEEE Journal of Oceanic Engineering}%
-{Low Complexity Adaptive Beamformer for Active Sonar Imaging}
-
-% Publishers ID mark:
-%\IEEEpubid{0000--0000/00\$00.00~\copyright~2007 IEEE}
-
-% use for special paper notices
-%\IEEEspecialpapernotice{(Invited Paper)}
-
-% for Computer Society papers, we must declare the abstract and index terms
-% PRIOR to the title within the \IEEEcompsoctitleabstractindextext IEEEtran
-% command as these need to go into the title area created by \maketitle.
-\IEEEcompsoctitleabstractindextext{%
-\begin{abstract}
-% 
-% \begin{itemize}
-% \item Dynamic weights $\Rightarrow$ greater contrast/resolution potential
-% \item But:
-% \begin{itemize}
-% \item Active system $\Rightarrow$ noise and signal correlated $\Rightarrow$ adaptive beamformers no longer robust.
-% \item Subarray averaging decorrelate the noise, but in the process we sacrifice resolution.
-% \item Computationally complex O($M^3$) due to inversion of a spatial covariance matrix.
-LCA is only O($MW$).
-Even if Capon subspace/beamspace versions exist, they remain more complex than the LCA, and is not as ideally suited for implementation of SIMD hardware (GPUs, DSPs, ...).
-% \end{itemize}
-% \item LCA is inherently robust, is ``parameter free'' once setup with a well designed window set, and can often be implemented with minor modifications to already existing DAS beamformers.
-% 
-% The angular resolution and contrast in active sonar images depend on the beamformer's ability to receive signals from directions of interest, while suppressing noise and interference emanating from other directions.
-For sonar arrays, this is achieved by applying weights to the array channels.
-% 
-% While classical beamformers use predefined windows, adaptive beamformers estimate the optimal window by analytical evaluation of the data.
-The minimum variance (MV) beamformer, for instance, calculates the set of weights that minimises the variance of the beamformer's output.
-% 
-We have studied the low complexity adaptive (LCA) beamformer in active sonar imaging.
-LCA can be viewed as either a simplification of the minimum variance distortionless response (MVDR) beamformer, or as an adaptive extension to the delay and sum (DAS) beamformer.
-While both LCA and MVDR attempt to minimize the power of noise and interference in the image, MVDR achieves this by computing optimal array weights from the spatial statistics of the wavefield, while LCA selects the best performing weights out of a predefined set.
-
-To build confidence in the LCA method we show that a robust MVDR implementation typically creates weights spanning between the rectangular and Hamming window function.
-We let LCA select from a set of Kaiser windows with responses in this span, and add some steered variations of each.
-We limit the steering to roughly half the \minus{}3\,dB width of the window's amplitude response.
-Using experimental data from the Kongsberg Maritime HISAS1030 sonar we find that LCA and MVDR produce nearly identical images of large scenes, both being superior to DAS.
-On point targets LCA is able to double the resolution compared to DAS, or provide half that of MVDR.
-This performance is achieved with a total of 6 windows; the rectangular window and the Kaiser window with $\beta=5$, in an unsteered version, and versions that are left and right steered to the steering limit.
-Slightly smoother images are produced if the window count is increased to 15, but past this we observe minimal difference.
-Finally we show that LCA works just as well if Kaiser windows are substituted with trigonometric ones.
-
-All our observations and experiences point to LCA being very easy to understand and manage.
-It simply works, and is surprisingly insensitive to the exact type of window function, steering amount, or number of windows.
-It can be efficiently implemented on parallel hardware, and handles any scene without the need for parameter adjustments.
-
-% An attactive trait of the LCA is its low computational complexity; while the MV beamformer is of O($M^3$), the LCA method is of O($MW$), with $M$ being the number of channels and $W$ the number of windows.
-We made the LCA perform like the MV method using a well designed set of 30 windows.
-Hence, unless the array is very small, the proposed method will perform like the MV beamformer or better, and at a fraction of the computational cost.
-
-% \end{itemize}
-
-\end{abstract}
-
-% Keywords (normally not used for peer reviews)
-\ifPeerReview\else
-\begin{IEEEkeywords}
-Beamforming, adaptive beamforming, MVDR, LCA, sonar, active, complexity.
-\end{IEEEkeywords}
-\fi
-}
-% \fi
-
-% make the title area
-\maketitle
-
-% This command fixes abstract positioning for compsoc articles:
-\IEEEdisplaynotcompsoctitleabstractindextext
-
-% (Optional) Add some extra info on cover page of peer review papers:
-% \ifCLASSOPTIONpeerreview
-% \begin{center} \bfseries EDICS Category: 3-BBND \end{center}
-% \fi
-
-% Insert page break and insert second title (peer review mode)
-\IEEEpeerreviewmaketitle
-
-\section{Introduction}
-
-% - Nature and scope
-% - Past work
-% - Methods
-% - Brief summary results
-% - Brief conclusions
-
-% The promise is already demonstrated in nature.
-% 
-% The sonar systems of the future 
-% 
-% Future sonar systems 
-% 
-% Great promise in adaptivity
-% 
-% The greatest untapped 
-% 
-% The sonars of the future might benefit greatly from adaptivity.
-% 
-% 
-% Future sonars will likely be
-
-% To make a sonar truly perform better, we need to make it smarter. 
-% 
-% \IEEEPARstart{F}{uture} sonar systems will merit a much more sophisticated level of adaptivity. 
-% 
-% \IEEEPARstart{T}{he} next generation sonar systems will be adaptive.
-% 
-% Making sonars adaptive is the way to take them into the future. 
-% 
-% The virtue of future sonar systems will be sophisticated adaptivity. 
-% 
-% Intelligent
-% 
-% Reactive
-% 
-% Observe, process, adapt, act. 
-% 
-% Adaptivity - sense, process, respond
-% 
-% Faster, smarter
-% 
-% The best sonars in nature are adaptive. 
-% Nature allures us to be this bold because
-
-\IEEEPARstart{T}{he} best sonars in existence is perhaps found in nature.
-Bats, for instance, use a high frequency sonar to detect, identify and track prey with amazing precision in caves amongst a myriad of other bats.
-No man made sonar can currenly match this feat, much due to the challenge of processing and adapting to such a complex environment fast enough.
-This is why human made sonars have little to no adaptivity on transmit.
-However, several methods have been investigated to adaptively post process the received data.
-Perhaps the best studied method for reconstructing a sonar image out of the data is the minimum variance distortion-less response (MVDR) beamformer~\cite{Capon1969}.
-It adapts the sonar array's spatial response to minimize the influence of noise and interference in the final image.
-
-In many cases MVDR can improve the contrast and resolution of active sonar images compared to conventional static methods~\cite{Blomberg2013,Blomberg2012a,Lo2004}.
-However, while the computational complexity of conventional beamformers are linear with the number of channels, O($M$), MVDR is at O($M^3$).
-This is because MVDR relies on estimating and inverting a covariance matrix.
-The estimation step dominates the computation in sonar systems with less than 32 channels, but here the computational complexity can be reduced and the implementation accelerated significantly using graphics computing units (GPUs)~\cite{Buskenes2014}.
-For larger systems the inversion step dominates.
-This can be dealt with by approximating the full covariance matrix with a small one using space reduction techniques such as beamspace processing~\cite{Asen2013}, or the closely related principal component analysis (PCA) method~\cite{Kim2014}.
-Another alternative is to assume spatial stationarity to form more easily invertible Toepliz matrices~\cite{Asl2012}.
-However, the mentioned methods are still relatively slow compared to DAS.
-
-A much faster and simpler alternative to MVDR is the Low Complexity Adaptive (LCA) beamformer.
-Based on an idea by Vignon~\cite{Vignon2008}, it was first introduced by Synnev\aa{}g \emph{et al}.
-in clinical medical ultrasound imaging~\cite{Synnevag2011}, who demonstrated its ability to obtain very similar image quality to MVDR in systems with focused transmit beams.
-LCA applies a set of predefined and static windows, or apodization weights, and selects the one offering the best noise suppression.
-It relies on the same optimization criterion as MVDR, thus may be considered a version of MVDR with a discrete and static window solution space.
-Alternatively, it may be viewed as a multi-apodization technique.
-One such method is described by Stankwitz \emph{et al}.
-in radar imaging~\cite{Stankwitz1995}, where the best out of 2 or 3 windows are selected.
-However, LCA is different in that it typically selects from a larger pool of windows and allows phase-steered variations of them.
-This gives LCA an adaptive freedom that lies somewhere in between that of traditional multi-apodization techniques and MVDR.
-How similarly LCA performs compared to MVDR depends on how well the predefined windows represent the constrained solution space of MVDR.
-
-% While using weights that span a bigger space gives LCA greater freedom, not The bigger the space the weights span, the greater the adaptive freedom, but only the solutions that are valid in the constrained solution space can make a difference.
-
-% Instead of adaptively computing the weights, it adaptively selects them.
-
-Synnev\aa{}g \emph{et al}.
-suggested using a window set comprised of rectangular, Kaiser (or Kaiser-Harris) and inverted Kaiser functions.
-In total he made LCA choose from 12 unique windows, 6 of which were phase-shifted versions of a fairly wide Kaiser window, and one were rectangular.
-Phase-shifting the windows has the effect of slighly adjusting the angle the array is steered towards.
-Synnev\aa{}g's composition of windows was motivated by the desire to use wide windows with maximum sidelobe suppression near the receiver, and narrow windows at greater depths for maximum sensitivity and depth penetration.
-While achieving good results, it was not made clear how many windows LCA needs, or how they ideally should be steered.
-Also, it is not apparent how LCA performs in sonar systems where the transmit beam is unfocused.
-
-In this study we expand our previous work of applying LCA to active sonar imaging~\cite{Buskenes2011}.
-We use variations of the parametric Kaiser window, but also compare it to the trigonometric window function.
-We extend Synnev\aa{}g's work by steering all our windows and not just some, and demonstrate how LCA's performance is affected by the type and number of windows included.
-Our results are obtained using experimental data from the 32 element Kongsberg Maritime HISAS1030 sonar, which operates at 100\,kHz with an \minus{}3\,dB element opening angle of 23$^\circ$.
-Although this system was designed as synthetic aperture sonar (SAS) where an image is synthesized from several pings of data, we will only create images out of single pings (sectorscan imaging) in this study.
-
-Our results show that when MVDR is modified to work in active sonar imaging, it computes weights with spatial responses that tend to be symmetric and steered within a fraction of the window's \minus{}3\,dB resolution.
-When letting LCA choose from Kaiser windows with similar responses, we obtain images with improved noise suppression and resolution.
-The resolution gain is made predictable by limiting the maximum steering to a fraction of the window's \minus{}3\,dB width.
-Overall, LCA produces images that are similar to MVDR, but with a resolution in between that of DAS and MVDR.
-LCA is also inherently robust, easy to implement, and fairly easy to understand.
-
-This article is outlined as follows: In Section \ref{sec:beamforming} we offer a gentle introduction to beamforming.
-Then we move on to adaptive beamforming and describe the MVDR and LCA methods in \ref{sec:mvdr} and \ref{sec:lca}, respectively.
-We use LCA with the Kaiser window which we describe in \ref{sec:lca_kaiser_windows}, explain how we steer it in \ref{sec:lca_steering}, and by how much in \ref{sec:lca_steering_bounds}.
-Finally we add some remarks on using the trigonometric window instead of Kaiser in \ref{sec:lca_trigonometric}, and of of sampling considerations in \ref{sec:lca_oversampling}.
-Results and discussion are provided in Section \ref{sec:results_discussion}, where we study MVDR's windows to find suitable LCA ones in \ref{sec:results_lca_window_function}, assess the range of window types and steering in \ref{sec:results_window_parameters}, determine the number of windows needed in \ref{sec:results_database_size}, and add a note on computational complexity in \ref{sec:results_complexity}.
-Section \ref{sec:conclusion} provides a conclusion.
-
-% This article is outlined as follows: In Section \ref{methods}, we introduce the concept of adaptive beamforming and provide details on the MVDR and LCA method.
-Then, in \ref{maptogpu}, we investigate the complexity issue, discuss means for reducing arithmetic complexity, and detail an implementation that makes efficient use of the GPU's parallel resources.
-The final design is assessed in Sections \ref{images_and_benchmarks} and \ref{discussion}, where we provide benchmarks, comparisons with similar CPU implementations, and measures of how efficiently our implementation makes use of the GPU's resources.
-
-% - How much oversampling
-% - How many parameters, how great a span.
-% 
-% - Experiementl data, Holmengraa/cross
-% 
-% - Note on efficiency.
-% 
-% 
-% From our work with MVDR it 
-% 
-% 
-% Traditionally beamformers were static, but as processors get more powerful so with the emergence of more processing power
-
-% 
-% \section{Methods}
-% 
-% At heart of the image formation process is a technique called beamforming, which creates a focus point at the pixel of interest.
-It does this by applying a suitable set of delays and weights to the array elements, such that signals emanating from the pixel location are emphasized constructively, while noise and interference from other directions sum destructively.
-% 
-% In active sonar array imaging, an acoustic wave is transmitted and the received echoes are usually recorded using an array of sensors.
-Each of these may be independently delayed and weighted such that signals emanating from directions of interest are added constructively, while noise and interference from other directions add destructively.
-Additionally, a window may be applied to the sensor channels to further adjust the array's spatial response.
-This process is known as beamforming.% The process of combining the signals from each sensor is known as beamforming.
-% 
-% 
-% 
-% 
-% % Traditional beamformers such as Delay-and-Sum (DAS) apply predefined windows to all incoming data.
-However, due to the non-stationary nature of sonar data, the optimal window for any given time instant will generally differ from the next.
-This is where adaptive beamformers thrive, because they compute the optimal window coefficients for the data at each time instant.
-The choice of optimisation criteria is what mainly differentiates the various adaptive beamformers.
-The Minimum Variance (MV) beamformer, for instance, selects the set of weights that minimizes the beamformer output power for any given time instant~\cite{cap69}.
-% 
-% 
-% \section{Background}
-% 
-% The MV beamformer computes the window coefficients by estimating and inverting a spatial covariance matrix.
-For improved estimation accuracy the data is averaged in space or time, or both.
-In addition, a certain amount of diagonal loading is often added to the covariance matrix before inversion.
-These actions, while leading to statistical robustness, tend to deteriorate the beamformer's performance.
-Also, the inversion of the covariance matrix has a computational complexity of O($M^3$), where $M$ is the number of elements.
-The computational burden makes the MV beamformer for large arrays less attractive, and sometimes even infeasible.
-% 
-% Based on the proposed method by Synnevåg~\cite{syn11}, we have implemented the LCA beamformer that keeps the minimum variance optimisation criterion but reduces the solution space to a discrete set of predefined windows.
-This reduces the computational complexity to O($MW$), where $W$ is the number of windows in the set.
-We use the Kaiser window function because it allows us to design a wide range of windows with different mainlobe widths and sidelobe suppression by adjusting the tradeoff parameter $\beta$~\cite{kai66}.
-In addition we apply steering to each of the windows and constrain the window design to ensure unit gain in the look direction.
-% 
-% The following will show that the proposed method performs similarly to or better than the MV method.
-
-\section{Receive Beamforming}\label{sec:beamforming}
-
-A sonar image is formed by estimating source locations and amplitudes.
-For this purpose a spatial bandpass filter is applied to the backscattered wavefield data.
-The filter is called an array processor or receive beamformer.
-The basic principle is to apply delay and weights to the sensor channels before summing them up, chosen such that signals from the location of interest are summed coherently, while other sources sum incoherently.
-
-Assume that the wavefield has been sampled by an $M$ element uniform linear array, and that the signal signature has been removed by a matched filter.
-Let $x_m[\theta,n]$ be the delayed data from the $m$th channel, where the $\theta$ and $n$ are the azimuth angle and range sample of the focus point, respectively.
-Each angle $\theta$ will be processed independently, so to simplify notation we will assume the dependence on $\theta$ to be implicit from now on.
-
-The output $z[n]$ of a beamformer is defined as the weighted sum of all the delayed data samples:
-%
-\begin{align}
-z[n] = \w\H[n]\x[n] = \bmat{w_0[n]\\w_1[n]\\\vdots\\w_{M-1}[n]}^H \bmat{x_0[n]\\x_1[n]\\\vdots\\x_{M-1}[n]},\label{eq:beamformer_output}
-\end{align}
-%
-where $w_m$ is the weight factor assigned to channel $m$.
-A weight set is commonly called a window, apodization or taper function.
-When the window is real its response is symmetric.
-Applying a real window that trails off towards the edges creates a response with lower sidelobe levels and wider mainlobe, which translates into improved noise suppression at the cost of reduced resolution, respectively~\cite{Harris1978}.
-
-Conventional beamformers all have static and usually real weights.
-The reference method is the delay-and-sum (DAS) beamformer, also know as the backprojection algorithm.
-It delays each pixel into focus, then applies a suitable window, and finally sums the data.
-The virtue of DAS is its simplicity, robustness to parameter errors, linear processing of the image and the ease of which it can be implemented in parallel hardware.
-
-Adaptive beamformers are dynamic and seek to adjust the array response to better fit the incoming wavefield.
-This may be achieved by allowing either the weights or delays to change, or both.
-Here the weights are usually complex, which allows asymmetric responses.
-One of the most extensively studied adaptive methods is the minimum variance distortion-less response (MVDR) beamformer.
-
-\section{MVDR}\label{sec:mvdr}
-
-MVDR seeks to minimize the output variance of the beamformer, under the constraint of a unity gain in some desired direction $\phi$~\cite{Capon1969}.
-With the assumption of zero mean data this can be formulated as:
-%
-\begin{align}
-\underset{\w[n]}{\argmin}\, E\{\big|z[n]\big|^2\} &= \underset{\w[n]}{\argmin}\, \w[n]\R[n]\w\H[n]\nn
-\text{subject to } \w^H[n]\a_\phi &= 1.\label{eq:mvdr_definition}
-\end{align}
-%
-Here $\a_\phi$ is a steering vector, $E\{\cdot\}$ is the expectation operator, and $\R[n] = E\{\x[n]\x\H[n]\}$ is the spatial covariance matrix for the full array.
-This is a convex optimization problem with the solution:
-%
-\begin{gather}
-\vec w[n] = \frac{\Ri[n]\a_\phi}{\a_\phi\T\Ri[n]\a_\phi}.\label{eq:mvdr_weights}
-\end{gather}
-The problem lies in estimating and inverting this spatial covariance matrix.
-We have described the exact steps in~\cite{Buskenes2014}: To avoid signal cancellation we apply spatial averaging by computing a mean covariance matrix from a set of subarrays with length $L$~\cite{Kailath1985}, for true speckle statistics we perform temporal averaging over $N_k = 2K+1$ temporal samples~\cite{Synnevag2009a}, and to improve robustness to parameter errors we add {\large$\epsilon$} percent diagonal loading~\cite{Cox1987,Maksym1979}.
-These steps are also needed to ensure that the covariance matrix is numerically well conditioned and hence invertible.
-
-\begin{figure}[tbp]\centering%
-\ifPeerReview%
-\includegraphics[width=.5\linewidth]{gfx/buske1_online.eps}%
-\else%
-\includegraphics[width=\linewidth]{gfx/buske1_online.eps}%
-\fi%
-\caption{Each Kaiser window is steered in the interval $\phi=[0, \phi_{\mathrm{3\,dB}}(\beta)]$.
-The angle $\phi_{\mathrm{3\,dB}}(\beta)$ is the amount of steering needed for the steered response to have a \protect\minus{}3\,dB crossing that is exactly half that of the unsteered window.
-With each window steered this way we expect the resolution gain to remain predictable and independent of $\beta$, and we also effectively constrain the white noise gain of the beamformer.}\label{windows_steering}
-\end{figure}
-
-\begin{figure}[tbp]\centering%
-\ifPeerReview%
-\includegraphics[width=.5\linewidth]{gfx/buske2_online.png}%
-\else%
-\includegraphics[width=\linewidth]{gfx/buske2_online.png}%
-\fi%
-\caption{One of the objects used to test the beamformers is a 1\,m by 1\,m resolution test cross attached to an anchor with a diameter of approximately 13\,cm.
-Source image curtesy of Bundeswehr Technical Center for Ships and Naval Weapons, Maritime Technology and Research (WTD 71).}\label{cross}
-\end{figure}
-
-\setcounter{topnumber}{1}
-\setcounter{dbltopnumber}{1}
-
-\begin{figure*}[t]\centering%
-\includegraphics[width=\linewidth]{gfx/buske3_online.pdf}%
-\caption{\emph{Determining LCA window type}: MVDR image with typical frequency responses for windows used in various pixel regions.\newline
-\emph{Left:}\hfill
-\parbox[t]{.95\linewidth}{MVDR sectorscan image of the oiltanker Holmengraa, with the pixel groups are indicated with red boxes.}\protect\\\hspace{\textwidth}
-\emph{Right:}\hfill
-\parbox[t]{.95\linewidth}{MVDR window amplitude and phase responses computed from 40x40 pixel groups for the shadow, highlight and speckle region of the image.
-The responses are overlayed each other and the amount of overlap is colored using a logarithmic scale.
-The size of the pixel groups were chosen ad-hoc for the histograms to be near invariant to a shift in position within the same region.
-Since each pixel is pre-delayed into focus the unsteered responses all have their center at broadside.
-The dashed red lines at -13\,dB and -43\,dB marks the peak sidelobe levels of an unsteered rectangular and Hamming window, respectively.
-Note how the responses are more or less symmetric, with very little steering in the shadow, moderate steering in speckle and steering within roughly 3\,dB in highlight.
-The phase varies most in the highlight region where we see the highest contrast.} }\label{mvdr_selected_windows}
-\end{figure*}
-
-\begin{figure*}[t]%
-\includegraphics[width=\textwidth]{gfx/buske4_online.pdf}%
-\caption{\emph{Determining Kaiser parameter boundaries}: LCA images of the resolution test cross and anchor created with a large and equal size window database, but where the upper bounds for $\beta$ and $\phi$ is varied.
-The upper left image is equal to DAS with a rectangular window.
-For all images we measure the lateral distance $\Delta x$ between the two \protect\minus{}3\,dB points of the anchor (red object).
-These are specified relatively to the reference distance  $\Delta x_\text{ref}$ from the upper left DAS image.
-\newline
-\emph{Upper bound $\beta$:}\hfill
-\parbox[t]{.89\linewidth}{Observe that LCA gets better at suppressing sidelobes in the image as we increase the upper bounds for $\beta$.
-\Fig{mvdr_selected_windows} suggests that MVDR tend to choose windows with levels lower than that of Kaiser with $\beta=2$, but here we observe further improvement going to $\beta=5$.}\newline
-\emph{Upper bound $\phi$:}\hfill
-\parbox[t]{.89\linewidth}{As we increase the upper bound of the steering $\phi$, we also increase the lateral image resolution. }
-}\label{oversampling_mosaic_bounds}
-\end{figure*}
-
-% \setcounter{dbltopnumber}{2}
-
-\begin{figure*}[t]\centering
-\subfloat[$\beta$-values of the Kaiser window chosen for each image pixel.
-Note how LCA prefers a a narrow response ($\beta=0$) on the anchor, and on other strong sources with little lateral interference.
-On the sources where lateral interference is present it chooses the widest response with the best sidelobe suppression.]{\includegraphics[width=\textwidth]{gfx/buske5a_online.pdf}%%
-\label{oversampling_mosaic_bounds_beta}}
-\hfil
-\subfloat[$\phi$-values of the Kaiser window chosen for each image pixel.
-On strong sources LCA selects windows that are steered away from the source.
-This is what improves the FWHM measurement in \Fig{oversampling_mosaic_bounds}.]{\includegraphics[width=\textwidth]{gfx/buske5b_online.pdf}%
-\label{oversampling_mosaic_bounds_phi}}
-\caption{Kaiser windows chosen for each pixel in the image.
-The underlying image is the one shown in \Fig{oversampling_mosaic_bounds}, with the same parameters.
-The location of the anchor, cut line and main scatter locations of the cross is marked in red.}
-\label{fig_sim}
-\end{figure*}
-
-\begin{figure*}[t]%\centering%
-\includegraphics[width=\textwidth]{gfx/buske6_online.pdf}%
-\caption{\emph{LCA images using different window database sizes}.
-The window boundaries are the same.
-The upper left image is identical to a rectangularly weighted DAS.
-Observe that sampling the $\beta$-range more finely increases the sidelobe suppression but leaves resolution unchanged.
-Adding more steering-variations improves both sidelobe suppression and resolution.
-However, using more than  $N_\beta=2$ window types and $N_\phi=3$ steering angles makes minimal difference.}\label{oversampling_mosaic}
-\end{figure*}
-
-\begin{figure*}[t]\centering
-\subfloat[\emph{Image quality of approximate point scatterers found in the resolution test cross scene}.
-Two lateral image cuts are shown in the leftmost figures, along with the response of a rectangularly weighted DAS.
-Note how the FWHM of LCA sits in between that of DAS and MVDR.
-Also note how LCA is insensitive to the window type; its performance is near identical whether it uses Kasier or trigonometric windows.]{\includegraphics[width=\textwidth]{gfx/buske7a_online.pdf}%
-\label{beamformer_comparison_cross}}
-\hfil
-\subfloat[\emph{Image quality of the full sector Holmengraa scene}.
-Compared to DAS the adaptive beamformers produce deeper shadows, sharper edges and a higher detail level.
-The adaptive methods produce nearly identical images.]{\includegraphics[width=\textwidth]{gfx/buske7b_online.pdf}%%
-\label{beamformer_comparison_holmengraa}}
-\caption{Comparing image quality of DAS, MVDR and LCA with Kaiser or trigonometric windows.}\label{image_quality}
-\end{figure*}
-
-% \begin{figure*}[t]%\centering%
-% \includegraphics[width=\textwidth]{gfx/oversampling_mosaic_bounds_lca_windows_beta.pdf}%
-% \caption{$\beta$-values of the windows LCA selects for each image pixel.
-The underlying image is the one shown in \Fig{oversampling_mosaic_bounds}, with the same parameters.
-The location of the anchor and cut line is marked in red.
-Note how LCA prefers a rectangular window ($\beta=0$) on the anchor.}\label{oversampling_mosaic_bounds_beta}
-% \end{figure*} 
-% 
-% \begin{figure*}[t]%\centering%
-% \includegraphics[width=\textwidth]{gfx/oversampling_mosaic_bounds_lca_windows_phi.pdf}%
-% \caption{$\beta$-values of the windows LCA selects for each image pixel.
-The underlying image is the one shown in \Fig{oversampling_mosaic_bounds}, with the same parameters.
-The location of the anchor and cut line is marked in red.
-Note how LCA prefers a rectangular window ($\beta=0$) on the anchor.}\label{oversampling_mosaic_bounds_beta}
-% \end{figure*}
-
-% % \setcounter{dbltopnumber}{1
-
-% \begin{figure*}[t]%\centering%
-% \includegraphics[width=\textwidth]{gfx/beamformer_comparison_cross.pdf}%
-% \caption{}\label{beamformer_comparison_cross}
-% \end{figure*}
-% 
-% \begin{figure*}[t]%\centering%
-% \includegraphics[width=\textwidth]{gfx/beamformer_comparison_holmengraa.pdf}%
-% \caption{}\label{beamformer_comparison_holmengraa}
-% \end{figure*}
-
-\section{LCA}\label{sec:lca}
-
-A much less complex alternative to MVDR is the low complexity adaptive (LCA) beamformer.
-It iterates through a set of $P$ windows and selects the window $p$ that best fulfills the minimum variance criterion:
-%
-\begin{align}
-\underset{p}{\argmin}\, E\{\big|z_p[n]\big|^2\} = \underset{p}{\argmin}\, E\big\{\big|\w_{p}\H\x[n]\big|^2\big\}\nn
-\qquad\text{subject to}\qquad \w\H\a_\phi = 1.\label{eq:lca_definition}
-\end{align}\label{lca_criterion}
-%
-Note how closely LCA is related to the MVDR definition in \eq{eq:mvdr_definition}.
-The optimization criterion and constraint is the same, but LCA has a finite and discrete solution space for the weights.
-As will be demonstrated in section \ref{sec:results_lca_window_function}, LCA will perform similarly to MVDR because a robustified MVDR implementation seems to be constrained to window functions similar to the ones we let LCA choose from.
-
-In practice, we estimate the beamformer output power by computing a sample power average $s_z^2$:
-%
-\begin{align*}
-E\big\{\big|z[n]\big|^2\big\} \approx s_z^2 = \frac{1}{N_k} \sumb{n'=n-K}{n+K} \big| z[n'] \big|^2
-\end{align*}
-%
-which computes the sample power average over $N_k=2K+1$ temporal samples.
-This is the same temporal averaging method we use for MVDR.
-In our case the bandwidth of our matched filtered signal is a just few samples long, hence $K=1$ will be used for both MVDR and LCA throughout this work.
-
-If the pixels are correlated laterally, we can include a weighted combination of these as well to improve the variance estimation:
-%
-\begin{align}
-s_z^2 = \frac{1}{N_x N_k} \sumb{\nmat{\scriptstyle x'=\\[-2mm]\scriptstyle x-X}}{x+X} \sumb{\nmat{\scriptstyle n'=\\[-2mm]\scriptstyle n-K}}{n+K} \omega[x',n']\big| z[x',n'] \big|^2
-\end{align}
-%
-where $N_x = 2X+1$ is the number of azimuth lines to average over, and $\omega[x',n']$ is a normalized 2 dimensional weight function.
-Since we oversample slightly laterally when we delay each pixel (see Section \ref{sec:lca_oversampling}), we thought the lateral correlation to be sufficient to benefit from this.
-However, we observed no visual improvement from applying this technique, and decided not to use it.
-% 
-% \begin{itemize}
-% \item \emph{Window function}
-% \item \emph{Parameter boundaries}
-% \item \emph{Number of windows needed}
-% \item \emph{Oversampling need}
-% \item \emph{Image quality}
-% \end{itemize}
-
-\subsection{Window function: Kaiser}\label{sec:lca_kaiser_windows}
-
-As will be demonstrated in upcoming sections, the LCA beamformer works very well with windows generated from the Kaiser-Bessel function. We will express it in vector form as:
-%
-\begin{align}
-\f_\beta = \bmat{
-f_0(\beta) \\
-\vdots\\
-f_{M-1}(\beta)
-}\label{eq:kaiser_window_vector}
-\end{align}
-%
-where
-%
-\begin{align}
-f_m(\beta) = \frac{I_0\left(\pi\beta\sqrt{1-\left(\frac{2m}{M-1}-1\right)^2}\right)}{I_0(\pi\beta)}\label{eq:kaiser_window_element}
-\end{align}
-%
-and $I_0$ is the zeroth order modified Bessel function of the first kind:
-%
-\begin{align}
-I_0(x) = \sumb{a=0}{\infty} \left[ \frac{\left(\frac{x}{2}\right)^a}{a!} \right]^2.\label{eq:modified_bessel_first_kind}
-\end{align}
-%
-The Kaiser-Bessel window is near optimal in the sense of having its peak energy concentration around $\theta=0^\circ$, for a given space-bandwidth product related to the Kaiser parameter $\beta$ as:
-%
-\begin{align}
-\beta = \frac{TB}{2},
-\end{align}
-%
-where $T$ in our case is the spatial extent of the window and $B$ is its bandwidth.
-Adjusting $\beta$ changes the trade-off between mainlobe width and sidelobe level.
-When $\beta=0$ the window becomes rectangular, while at large values ($\beta>5$) the window converges to a Gaussian both in time and frequency.
-This class of windows is generally considered well suited for separating closely spaced sources with amplitudes of a high dynamic range~\cite{Harris1978}, they are easy to make, and they are optimal for any value of $\beta$.
-
-\subsection{Steering}\label{sec:lca_steering}
-
-Adding slightly steered versions of each window to the window database gives LCA greater flexibility in searching for an optimal window.
-A Kaiser window $\f_\beta$ steered to the angle $\phi$ can be expressed as
-%
-\begin{align}
-\w_{\beta,\phi} = \frac{\f_\beta\H\diag{(\a_\phi)}}{\f_\beta\H\a_\phi}\label{eq:steered_window}
-\end{align}
-%
-where diag($\a_\phi$) is a diagonal matrix constructed from the steering vector $\a_\phi$:
-%
-\begin{align}
-\a_\phi = \bmat{
-1 \\
-e^{-j\frac{2\pi d}{\lambda}\sin(\phi)} \\
-\vdots\\
-e^{-j\frac{2\pi (M-1)d}{\lambda}\sin(\phi)}
-}.\label{eq:steering_vector}
-\end{align}
-%
-Here $d$ is the element spacing, $\lambda$ is the wavelength and $\phi$ is the steering amount.
-The normalization factor $\f_\beta\H\a_\phi$ is the reciprocal of the window's coherent gain.
-It ensures unit gain in the direction of interest as required by (\ref{eq:lca_definition}).
-Since the signal-to-noise ratio is constant for a specific window (same $\beta$-value), this normalization also proportionally increases the incoherent noise gain.
-
-The window database we will construct will contain $N_\beta$ Kaiser windows with unique $\beta$-values, each being steered in $N_\phi$ different directions.
-This gives us $N_w = N_\beta N_\phi$ unique windows.
-
-\subsection{Steering bounds}\label{sec:lca_steering_bounds}
-
-Each window's response is constrained to unit gain in the look direction.
-Hence, when it is steered the white noise gain must increase and the the signal-to-noise ratio decrease, as shown in \Fig{windows_steering}.
-To limit this we place an upper bound to the steering, chosen such that wide windows are allowed to be steered more than narrow ones.
-We call this upper bound for steering $\phi_{\mathrm{3\,dB}}(\beta)$, and define it as the steering angle needed for the steered response to have a \minus{}3\,dB crossing that is exactly half that of the unsteered window.
-The method we devised for finding it is described in Appendix \ref{appendix:steering_angles}.
-
-Henceforth any steering will be specified relative to this upper bound.
-With each window steered this way we expect the resolution gain to remain predictable and independent of $\beta$, and we also effectively constrain the white noise gain of the beamformer.
-
-\subsection{Trigonometric windows}\label{sec:lca_trigonometric}
-
-While we focus on using LCA with Kaiser windows, we will for completeness compare the performance with that of using trigonometric windows. This is defined as:
-%
-\begin{align}
-f_m(\alpha) = \alpha - (1-\alpha)\cos\Big(\frac{2\pi m}{M-1}\Big)\label{eq:trig_window_function}
-\end{align}
-%
-where we apply the contraint $\alpha\in[0.5,1]$ to avoid windows with negative coefficients.
-Steering is applied as in (\ref{eq:steered_window}), and the $\alpha$-value that halves the \minus{}3\,dB distance is found using the method described in Appendix \ref{appendix:steering_angles}.
-
-\subsection{Oversampling}\label{sec:lca_oversampling}
-
-The ultimate goal of adaptive beamformers is to take the information available in the wavefield and use it to either improve image resolution, sidelobe suppression, or preferably both.
-This non-linear processing increases the image bandwidth and introduces a need for a sampling rate higher than the Nyquist rate.
-For MVDR a lateral oversampling factor of 10 compared to the Nyquist rate is often needed to ensure minimal spatial shift-variance in source amplitude~\cite{Asen2014}.
-
-Since LCA operates with the same optimization criterion as MVDR, we will be using approximately a factor 8 lateral oversampling for both beamformers.
-The images appear more visually pleasing and detailed up to 8.
-We consider an absolute minimum to be 2, due to the non-linear nature of delaying pixels that isn't strictly in the far field, and of displaying the absolute value of the pixels on a decibel scale.
-In \mediaI{}\footnote{For media location refer to Appendix \ref{app:source_code}.} we visualize the effect of changing the lateral oversampling factor on a set of LCA images.
-Upon display the images are all bilinearly upinterpolated to the same size.
-% 
-% \todo{Need $M$ in the image sector. Why???}
-
-% Idea to steer within a fraction of the rectangular window's mainlobe.
-
-% \begin{align}
-% d = \frac{\pi\,\theta_\text{open}\,\lambda}{M\,d\,O_s}
-% \end{align}
-
-% See section \ref{kaiser_windows}.
-
-% Narrow ones:
-% - Best coherent gain, lowest equivalent noise bandwidth, and most sensitive to steering meaning not needing to steer as much for some effect.
-% 
-% 
-% Beamforming:
-% - Maximize resolution
-% - Minimize noise/error, both coherent and incoherent
-% - Minimize signal cancellation
-% 
-% Gauss: Miminized time-bandwidth product
-%   - 
-%   
-% Rectangular
-%   - 
-% Dolph-Cheb: Minimize mainlobe width given sidelobe suppression
-% Kaiser: For a restricted energy and time duration, maximize energy in the band of frequencies W.
-% 
-% Adaptive windows perform best in detection of closely separated angles of significantly amplitudes.
-
-% Adaptive beamformers usually compute the window coefficients by estimating and inverting a spatial covariance matrix.
-There are two inherent problems to this procedure.
-First, for improved estimation accuracy the data is either averaged in space or time, and the significance of the covariance matrix can be reduced using a regularisation parameter~\cite{Carl Inge}.
-These are all attempts to constrain the beamformer to ensure the window coefficience are not over-adapted to the data.
-Second, inverting the covariance matrix has a computational complexity of O($M^3$)~\cite{Carl Inge}.
-For larger arrays the computational burden becomes significant, and the question arises whether this processing power can be better utilised.
-% 
-%
-
-% 
-% This selection is based on observations of the solutions found by the MV beamformer.
-We obtained good results in our experiments by letting the LCA beamformer select among 5 Kaiser windows with uniformly distributed $\beta$'s in the range $[0.05, 0.5]$ and a rectangular window.
-Each window was steered in 5 different directions, uniformly distributed within 80\% of the mainlobe width of the rectangular window, which is the most narrow.
-This adds to a total of 30 windows.
-% 
-% To make the choice of window less susceptible to pixel value uncertainties, the LCA beamformer was set to apply the most frequently selected window in an 11 pixel range region to the center pixel in that region.
-In physical terms, this is a region of approximately 20\,cm.
-The MV beamformer estimated the covariance matrix as described in~\cite{syn07} by averaging over 16 subarrays and 11 range pixels, and applying 3\% diagonal loading.
-
-\section{Results and Discussion}\label{sec:results_discussion}
-
-To test the performance of the MVDR and LCA beamformers, we have processed data acquired by the Kongsberg Maritime HUGIN AUV carrying their 32 element HISAS1030 sonar~\cite{Hansen2011}.
-It is a high resolution sonar with 1.2\;m array length, 100\;kHz operating frequency, 30\;kHz bandwidth and 23$^\circ$ element \minus{}3\,dB opening angle.
-
-Two different scenes will be studied.
-One of the 1500 DWT oil tanker wreck Holmengraa lying at a slanted seabed at 77\;m depth outside of Horten, Norway~\cite{holmengraa}.
-It measures 68\;m by 9\;m and fills most of the highlighted sector.
-This data was collected by the Norwegian Defense Research Establishment and Kongsberg Maritime.
-The other scene is of an 1\,m by 1\,m iron resolution test cross, connected to a an anchor with 13\,cm diameter (\Fig{cross}).
-This data was collected by the Bundeswehr Technical Center for Ships and Naval Weapons, Maritime Technology and Research (WTD 71).
-
-In the image reconstruction we have run MVDR with subarray length $L=16$ and $\text{\large$\epsilon$}=1\%$ diagonal loading.
-Both MVDR and LCA was run with $K=1$ temporal averaging.
-This is a fairly aggressive yet stable set of parameters~\cite{Synnevag2009a}.
-
-The results will be presented in the following order: In section \ref{sec:results_lca_window_function} we discuss typical window responses computed by MVDR, and hypothesize that these can be mimicked by Kaiser windows.
-In section \ref{sec:results_window_parameters} we determine sensible boundaries for the Kaiser parameter $\beta$ and window steering $\phi$.
-In section \ref{sec:results_database_size} we discuss how many window variations are needed for LCA to perform well.
-
-\subsection{LCA window function}\label{sec:results_lca_window_function}
-
-Assuming that a robust MVDR is the reference method we want LCA to perform similarly to, it seems sensible to create a window database for LCA with spatial responses similar to the ones that MVDR computes.
-We study this in \Fig{mvdr_selected_windows}, where we present the typical amplitude and phase responses that MVDR creates for shadow, speckle and highlight regions of the Holmengraa scene.
-
-Observe from \Fig{mvdr_selected_windows} that MVDR seems to prefer symmetric window responses, even if it is free to choose non-symmetric ones.
-The symmetry is most predominant in shadow and speckle regions, while in highlight the windows are approximately symmetric within the illuminated sector of the seafloor.
-We have observed this symmetry to be a side-effect of the averaging steps needed to build the sample covariance matrix, which are required for MVDR to operate in near snapshot mode in an active system.
-The peak sidelobe levels of the MVDR windows are mostly between \minus{}13\,dB and \minus{}43\,dB.
-This corresponds to that of an unsteered rectangular and Hamming window, or an unsteered Kaiser window with $\beta=0$ and $\beta=2$, respectively.
-
-From this we hypothesize that a good window database LCA can be made using a varied  set of Kaiser windows.
-If they span a suitable range of $\beta$-values, and some steering variations are applied to each, we should have responses that resemble those in \Fig{mvdr_selected_windows}.
-The Kaiser window is easy to compute, is fairly insensitive to coefficient inaccuracies, and can create windows with shapes spanning from rectangular to Gaussian.
-It is also optimal in the sense of having its peak power concentration near the steering angle.
-
-Unlike the method described by Synnev\aa{}g~\cite{Synnevag2011} we do not let LCA choose from inverted Kaiser windows.
-These have a mainlobe width narrower than that of the rectangular window, but at the expense of a very poor white noise gain.
-We have found that these windows hardly ever get used on experimental data, in particular steered versions of them.
-We can infer the same from \Fig{mvdr_selected_windows} by noting that the maximum sidelobe level rarely exceeds the rectangular window level of \minus{}13\,dB.
-
-\mediaII{}\footnote{For media location refer to Appendix \ref{app:source_code}.} animates how \Fig{mvdr_selected_windows} changes as a function of the MVDR subarray length $L$ and temporal averaging $K$.
-At $L=1$ the window responses are rectangular in all areas in the image.
-At $L=2$ we observe window responses with slight amplitude variations in the highlight region, but with the phase being 0$^\circ$ in the illuminated seafloor sector.
-Already at $L=2$ the MVDR is able to greatly suppress noise.
-This is a common observation; adding a little flexibility to adapt to the scene has a dramatic effect, but further allowing full flexibility is much less significant.
-The media file also shows that MVDR can be run with subarray sizes $L\in[M/2,5M/8]$, but only with temporal averaging $K=1$ or above.
-
-\subsection{Kaiser parameter $\beta$ and steering $\phi$}\label{sec:results_window_parameters}
-
-To determine a suitable range for the Kaiser parameter $\beta$ and steering $\phi$, we constructed large and equally sized window databases containing Kaiser windows with varied upper boundaries for $\beta$ and $\phi$.
-The lower boundaries were chosen as $\beta=0$ and $\phi=0^\circ$, which includes the rectangularly weighted DAS image.
-The resulting images are shown in \Fig{oversampling_mosaic_bounds}.
-In each image we computed the lateral distance $\Delta x$ between the \minus{}3\,dB points of the anchor, with the reference $\Delta x_\text{ref}$ being the rectangularly weighted DAS image.
-A fourth order polynomial fit was used in this computation.
-This measure is also commonly called the full width half maximum (FWHM).
-In the case of a imaging a point scatter, it is closely related to the resolution of of the system.
-For a rectangular window, for instance, the system resolution is approximately $\Delta \approx \text{FWHM}/0.89$~\cite{Harris1978}.
-
-The LCA images in \Fig{oversampling_mosaic_bounds} demonstrate the effect of adjusting the upper bound of $\beta$ and $\phi$.
-As suggested by \Fig{mvdr_selected_windows} we first attempted to use $\beta\in[0,2]$.
-While this significantly improved sidelobe suppression, a slight further improvement was observed up to $\beta\in[0,5]$.
-This can likely be explained by the need to compensate for the increased sidelobe levels caused by steering.
-No noticeable difference were observed for higher values than $\beta=5$.
-The images also demonstrate that increasing the upper bound for steering $\phi$ improves the resolution of the strong scatterers.
-For this sonar the Rayleigh resolution is $\Delta\approx0.0125$ radians (0.72$^\circ$), so a point scatterer imaged at 19\,m range with a rectangular DAS would have a lateral $\text{FWHM}\approx0.125\cdot19\,\text{m}\cdot0.89=21\,\text{cm}$.
-This is close to the FWHM of the anchor in the DAS image in \Fig{oversampling_mosaic_bounds}, which means that the acoustic fingerprint of the anchor is similar to a point source.
-This suggests the anchor reflections to be specular and that we view its rounded side.
-As we increase steering to either 50\% or 100\% of the \minus{}3\,dB width, the FWHM drops to 73\% or 53\% to that of the rectangular window, respectively.
-
-\Fig{oversampling_mosaic_bounds_beta} and \Fig{oversampling_mosaic_bounds_phi} illustrates which $\beta$ and $\phi$ values LCA prefers for different regions in the image of the resolution test cross.
-Red marker lines are used to pinpoint the cut line, location of the anchor and strong scatters on the cross.
-Observe that in the anchor region only narrow responses ($\beta=0$) are used, which is the reason for the anchor FWHM being measured to the same value regardless of the upper bound of $\beta$ in \Fig{oversampling_mosaic_bounds}.
-On other scatterers with strong nearby lateral interference LCA prefers wider windows to suppress the interference.
-When allowed LCA prefers to steer windows away from the sources, but never exceed the $\phi_\mathrm{3\,dB}(\beta)$ limit.
-When exceeding this bound we observed oscillation artifacts in the image.
-
-\subsection{Window database size}\label{sec:results_database_size}
-
-So far we have found it reasonable to use Kaiser windows in the range $\beta\in[0,5]$, each steered within the range $|\phi|\in[0,\phi_\mathrm{3\,dB}(\beta)]$.
-What remains is to determine the number of windows needed.
-We study this in \Fig{oversampling_mosaic}, where we compared LCA images made from windows databases of different sizes, but with the same parameter boundaries.
-We included the results from using a single window ($N_\beta=1$) and steering angle ($N_\phi=1$), which corresponds to an unsteered rectangularly weighted DAS.
-
-Observe that sampling the $\beta$-range more finely increases the sidelobe suppression but leaves resolution unchanged.
-Adding more steering-variations improves both sidelobe suppression and resolution.
-We can observe a major improvement going to $N_\beta=2$ and $N_\phi=3$, but minimal improvement by adding yet more windows.
-
-\subsection{Image quality}\label{sec:results_image_quality}
-
-We compare LCA image quality to that of DAS and MVDR in \Fig{image_quality}.
-In \Fig{beamformer_comparison_cross} we display the images computed by LCA using Kaiser windows, LCA using trigonometric windows, and MVDR.
-Two lateral image cuts through all images are presented in the left plot.
-The images and their corresponding cuts are nearly identical for the LCA version with Kaiser windows and the one with trigonometric windows.
-Compared to DAS the LCA produces targets with an FWHM that lies in between that of DAS and MVDR.
-
-In \Fig{beamformer_comparison_holmengraa} we show full sector images of the Holmengraa wreck.
-We compare the images produced by a Hamming weighted DAS, by MVDR and by LCA with Kaiser or trigonometric windows.
-All the adaptive beamformers produce a sharper and less noisy image than DAS, but the difference between LCA and MVDR is minimal.
-The LCA image produced with Kaiser windows appear identical to that from trigonometric windows.
-
-\subsection{Computational complexity}\label{sec:results_complexity}
-
-The computational complexity of MVDR is generally considered to be of O($M^3$).
-However, the implementation we are using is optimized so that building the spatial covariance matrix is of O($N_kN_L,L^2$), and inverting it is of O($L^3$).
-We describe this implementation in \cite{Buskenes2014}, and a beamspace version in \cite{Asen2013}, where we optimize if for a graphics processing unit (GPU) for nearly two orders of magnitude speed increase compared to a straightforward C implementation.
-However, MVDR is not ideal for GPUs due to the complex data dependencies present in the covariance computation, and at best we only managed to utilize 10\% of the GPUs potential.
-
-In comparison, LCA is of O($M N_\beta N_\phi$), and extremely well suited for GPUs.
-Each pixel depends on only a small subset of data, and the windows can be precomputed and stored in GPU cache for near immediate access.
-If we use a trigonometric window function we solve for the window parameter analytically as described in appendix \ref{sec:reducing_lca_complexity}.
-This reduces the complexity to O($M N_\phi$), and the only remaining LCA parameter is the steering amount.
-
-% Furthermore, the algorithm maps well to single-instruction-multiple-data (SIMD) hardware such as graphics processing units (GPUs).
-We show that even a modest amount of optimisation work has resulted in a factor 10 speed-up of this algorithm when implemented on a GPU as opposed to on a CPU.
-
-% the performance of the LCA beamformer is surprisingly similar regardless of what windows it is allowed to choose from.
-
-% \newcommand\cc[1]{\multicolumn{1}{>{\columncolor{tabBlue}}c}{\bf #1}}
-% \begin{table}[!b]\centering%\normalsize
-% \begin{tabular}[c]{l r r r@{}  r r r}\hline
-% \rowcolor{tabBlue} & \cc{Resolution} & \cc{S} & \cc{$\beta$} & \cc{$\varphi^*$} &  \cc{N$_\beta$} & \cc{N$_\varphi$} \\\hline
-% & 50\%  & 28 & [0,3] & \hspace{.15cm} 50\% & 2 & 3 \\
-% & 50\%  & 28 & [0,6] & \hspace{.15cm} 50\% & 2 & 3 \\
-% & 50\%  & 28 & [0,9] & \hspace{.15cm} 50\% & 2 & 3 \\
-% & 100\% & 32 &  &:30 \\
-% & 150\% & 32 & 1 &:4\\
-% \end{tabular}\\
-% * Steering is relative to the \minus{}3\,dB width of the window being applied.
-% \caption{Qualitative study}\label{parameter_study}
-% \end{table}%
-
-% The simulations contained a few highlighted cylindric regions with 2\,m diameter and a constant intensity 15.4\,dB over the average speckle level.
-We focused on an object centered at 41\,m range and -3 degrees azimuth angle.
-The images produced by the LCA, MV and unweighted DAS for a single speckle realisation and a mean image of all 100 realisations are shown in \Fig{speckle}.
-Each image was normalised by their respective speckle level, and the dynamic range was clamped at \{-30,~15\}\,dB.
-The physical extent of the of the simulated object was superimposed for reference.
-%\ref{saf09}
-% 
-% In \Fig{speckle} we observe that both the adaptive beamformers produce images with a more clearly defined shadow.
-The edges are sharper and the shadow more distinct.
-This is because the adaptive beamformers have better sidelobe suppression, and thus allow less signal energy to leak from the speckle region into the shadow region.
-The same effects can also be observed laterally around the highlight region of the mean image; DAS is more prone to smear energy into the speckle region than the adaptive beamformers are.
-% 
-% \Fig{cutmean} displays two lateral cuts in the mean image, one through the highlight at 41.1\,m range and one through the shadow at 44.5\,m range.
-Each cut was computed as the mean of a 40\,cm range band centered at the respective ranges.
-The transition region between highlight and shadow is shortest for the LCA, which is advantageous since it results in a more accurate representation of the object size.
-The slightly inferior performance of the MV beamformer is due to the subarray averaging, which makes the effective array smaller~\cite{syn07}.
-% 
-% By studying the LCA performance on the 100 speckle realisations, we found the choice of window to be highly dependent on the phenomena being imaged.
-In speckle regions LCA favored narrow spatial responses with minor steering, while wide and fully steered responses were often selected in shadow regions.
-There was also a clear distinction between the responses selected in sidelobe regions and in speckle regions.
-% 
-% % \Fig{hist} illustrates how often a particular window is chosen in general.
-There are 10 $\phi$'s for each of the 10 $\beta$'s.
-The first 10 values are rectangular windows.
-Value 5 is a non-steered rectangular window, and we see that this is most frequently selected.
-At higher steering angles the windows are rarely selected, and thus we might get better results by tightening our steering boundaries.
-At value 15 we find inverted Kaiser windows with $\beta = 0.05$, and then $\beta$ is increased in increments of 0.505 for every 10th value.
-At value 65 the $\beta$ value is 2.025, and these and the remaining windows are hardly selected.
-% 
-% A sidescan image of the 1500 dwt oil tanker wreck Holmengraa is shown in \Fig{holmengraa}.
-It is about 68\,m long and 9\,m wide, and lies on a slanted seabed at 77\,m depth~\cite{holmengraa}.
-The sidescan image was created using data from the HISAS 1030 sonar, which is rather unsuited for this purpose because of its large opening angle.
-This, and the fact that the wreck was imaged at a range of about 105\,m makes the image quality poor, but sufficient to compare our beamformers.
-% 
-% In the Holmengraa image we note again that the LCA produces a cleaner shadow and better edge definition. The MV method performed almost identically to LCA in this case, and was omitted.
-
-\section{Conclusion}\label{sec:conclusion}
-
-% \begin{table}[!b]\centering%\normalsize
-% \begin{tabular}[c]{l c c c}\hline
-% \rowcolor{tabBlue} & \bf Highlight & \bf Shadow & \bf Speckle  \\\hline
-% Symmetry             & Moderate  & High   & High \\
-% Mainlobe width       & Narrow    & Wide   & Moderate \\
-% Sidelobe suppression & Low       & High   & Moderate \\
-% Steering             & Low       & Up to \minus{}3\,dB width & Moderate \\
-% Oversampling         & 2x-8x     & 2x-8x  & 2x-8x
-% \end{tabular}
-% \caption{Summary}\label{tab:summary}
-% \end{table}%
-
-LCA seeks to improve image resolution and contrast by minimizing the power of noise and interference in each image pixel.
-This is the same optimization criterion as MVDR uses, but instead of computing an optimal array window like MVDR does, it selects the best window out of a predefined set.
-Hence, it can be viewed as either a reduced and discrete window space version of MVDR, or as an advanced multi-apodization technique.
-In this paper we have studied which windows we should ideally let LCA choose from.
-
-We found the LCA to be surprisingly insensitive to the exact window function it is used with, supported by the observation that near identical images were produced whether we used Kaiser windows or trigonometric ones.
-However, LCA must be able to choose between wide and narrow window responses, since the wide ones offer the best sidelobe suppression and the narrow ones offer the best sensitivity.
-We achieved good results with 2-3 Kaiser windows in the range $\beta\in[0,5]$, and observed minimal difference when adding more windows or adjusting the upper bound of $\beta$.
-For the steering angle we suggest a value in the range $\phi\in[\pm\delta\phi_\mathrm{3\,dB}(\beta)]$, where $\phi_\mathrm{3\,dB}(\beta)$ is the bandwidth of the respective window, and $\delta$ is a scaling parameter.
-Setting $\delta$ to either 50\% or 100\% reduces the lateral point target size to 73\% or 53\% compared to the rectangular window, respectively.
-Hence, this parameter can be used to control the aggressiveness and resolution gain of LCA.
-
-In summary, LCA is an attractive alternative to other adaptive beamformers due to being very fast, simple to understand, practically parameter-free, inherently robust, and able produce images similar to that of MVDR.
-Contrary to MVDR, it has a low computational complexity and can be easily and efficiently accelerated using e.g.
-GPUs.
-Since LCA performs just as well if we use trigonometric windows instead of Kaiser ones, we can also compute the optimal window parameter analytically for any given steering angle, and avoid the search over the window parameter entirely.
-
-% The tendency of LCA to produce images of similar quality as MVDR seems to indicate that a full solution space is rarely needed in real world scenarios.
-
-%  It initially very is also very fast, and is very well suited for acceleration using  very fast, easy to implement even on parallel hardware,
-
-% Its computational complexity is only of O($N_wM$), and its data parallel nature makes it very well suited for acceleration using e.g. GPUs. compared to MVDR's O($M^3$).
-% 
-% LCA is largely insensitive to the choice of windows and steering. 
-% LCA works well for all the boundaries
-% Variables of interest:
-
-% We limit our study to the use of variations of the parametric Kaiser-Harris window, since we observed neglible performance difference to using this function or any of the common trigonometric ones.
-The Kaiser windows are easy to compute, is fairly insensitive to coefficient inaccuracies, and can create windows with shapes spanning from rectangular to Gaussian.
-% 
-% 
-% Using the LCA beamformer results in increased robustness and improved edge definition compared to the MV and DAS beamformers.
-
-% The images produced by LCA appears sharper than those of MVDR, but 
-% 
-% The key to achieve this success lies in the design of the window set.
-Windows that yield vagely steered narrow responses are preferred in speckle regions, while wide and steered responses are typically preferred in highlight and shadow regions.
-% 
-% A window set of 6 different responses each steered in 5 different directions proved sufficient in our experiment to match the performance of the MV beamformer.
-
-% \begin{itemize}
-% \item Adaptive beamformer's potential lies in its ability to suppress interference power
-% \item Why adaptive beamformers struggle in active sonar systems.
-Correlated noise, robustification kills the adaptive potential.
-Quite computationally intensive.
-Constraints must be applied in one way or another - parameters must be tuned.
-% \begin{itemize}
-% \item Noise and signal is correlated. Spatial averaging required. 
-% \item Increases variance of speckle (not only in active systems?), spatial compunting \cite{Vignon2009} (US) or time averaging?.
-Subarray averaging applied here \cite{Synnevag2007} (US) \todo{Is it bad to mix ultrasound/ sonar refs?}
-% \end{itemize}
-% \item Outline: 
-% \begin{itemize}
-% \item Setup the window set for LCA to match this.
-% \item Notice that the LCA performs like Capon, even with a small window database.
-% \item Start with Kaiser, see that other window functions also work well.
-% \end{itemize}
-% \end{itemize}
-% \todopar{Out of the blue idea: Weigh CF with LCA? This guy did it with MVDR \cite{Wang2009}. Probably you know already... Okay, back on track.}
-
-% \IEEEPARstart{T}{o} form images from a modern phased array sonar system the received wavefield is usually recorded, and then postprocessed by a digital beamformer.
-The beamformer applies delays and weights to the sensor channels, the beamformer adjusts the arrays spatial response to focus at one pixel at a time.
-such that signals emanating from regions of interest add constructively, while ensuring that noise and interference from other angles do not.
-% 
-% The imaging capabilities of a modern phased array sonar system depend on physical attributes such element response and array geometry, the transmitted signal, as well as the beamforming method being used on transmission and reception.
-Beamforming is the concept of applying delays and weights to the sensors channels to steer the arrays response to points of interest.
-
-% 
-% 
-% Outline:
-% \begin{itemize}
-% \item Choice of window functions makes little difference.
-% \item Steering and mainlobewidths have outer bounds.
-% \item Beamspace?
-% \item Chosen window plots - what may they tell us? Variance intensity values when using various windows.
-% \item Assymmetric windows?
-% \end{itemize}
-%
-
-% , and maximum suppression of while ensuring that the beamformer digitally  before each of the pixels are estimated one at a time.
-The resolution and contrast of such a system will depend on the systems spatial response, which ideally should be narrow  be very sharp in the desired direction its ability to achieve  fundamental principle of forming a sonar image is to record the received wavefield,
-
-%%%%%%%%%%%%%%%%%%                              ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
-% DOCUMENT APPENDICES %
-%%%%%%%%%%%%%%%%%%                              ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
-
-\appendices
-
-\section{Steering angle}\label{appendix:steering_angles}
-
-\begin{figure}[t]
-\ifPeerReview\centering%
-\includegraphics[width=.5\linewidth]{gfx/buske8_online.eps}%
-\else%
-\includegraphics[width=\linewidth]{gfx/buske8_online.eps}%
-\fi%
-\caption{\emph{Steering angle needed to cut the \protect\minus{}3\,dB distance of the Kaiser window's spatial response in half}.
-The needed amount of steering depends on the number of channels $M$, the element spacing relative to the wavelength $\frac{d}{\lambda}$, and the Kaiser parameter $\beta$.
-Note how the angle scales near proportionally with $\frac{d}{\lambda}$.
-}\label{steering_angle_3dB}
-\end{figure}
-%
-One of the goals of this article was to characterize the windows used by the MVDR method, and then identify a subset of these that were suitable for use with the LCA method.
-We determined that the Kaiser function could fit the role of producing the relevant windows.
-The spatial response of a steered Kaiser window depends on the number of channels $M$, the element spacing relative to the wavelength $\frac{d}{\lambda}$, the Kaiser parameter $\beta$, and the amount of steering $\phi$.
-
-To control the resolution gain of LCA, we wanted to determine the extent that the windows need to be steered to cut the angle of the \minus{}3\,dB point of the window's spatial amplitude response by a fixed amount, say, by a factor 2 as reference.
-We computed this steering angle for some common configurations of the mentioned parameters.
-Refer to \Fig{steering_angle_3dB} for the result.
-The boundary values of the Kaiser window is shown, i.e.
-the rectangular window at $\beta=0$ and the near gaussian window at $\beta=5$, as well as one in between at $\beta=2.5$.
-Observe that for a given $M$ and $\beta$, the angle $\phi_\mathrm{3\,dB}$ scales near proportionally with $\frac{d}{\lambda}$.
-We found this to be true also for $\frac{d}{\lambda}\in\{0.5,4\}$.
-Hence, an approximate figure for a rather wide range of system parameters can be derived off this figure.
-
-For the sake of completeness we also supply the source code, see Appendix \ref{app:source_code}.
-
-\section{Reducing LCA complexity}\label{sec:reducing_lca_complexity}
-
-\renewcommand\b{\vec b}
-
-Throughout this article we have used LCA with the Kaiser window function.
-However, as shown in \Fig{image_quality} we can obtain similar performance using the trigonometric window function.
-For this window we can obtain an analytic solution for the optimal value of $\alpha$ in the minimum variance sense.
-The following derivation generalizes that of the non-steered (real) window described in e.g.~\cite{Stoica2005} to also apply for the steered (complex) window.
-We start by inserting the trigonometric function (\ref{eq:trig_window_function}), with applied steering (\ref{eq:steering_vector}), into the beamformer equation (\ref{eq:beamformer_output}):
-%
-\begin{align}
-\w\H\x &= \sumb{m=0}{M-1} [\a_\phi]_m\Big(\alpha - (1-\alpha)\cos\left(\frac{2\pi m}{M-1}\right)\Big) x_m^* \nn
-&= \alpha \a_\phi\T\x - (1-\alpha)\b_\phi\T\x
-\end{align}
-%
-where $[\a_\phi]_m$ is the $m$th component of the steering vector $\a_\phi$ defined in (\ref{eq:steering_vector}), and 
-%
-\begin{align}
-\b_\phi = \text{diag}(\a_\phi)\cdot\begin{bmatrix}
-     1 &
-     \cos\left(\frac{2\pi}{M-1}\right) &
-     \cos\left(\frac{2\pi 2}{M-1}\right) &
-     \hdots &
-     1
-     \end{bmatrix}\T.
-\end{align}
-%
-Unity gain in the look direction is ensured as long as the weights sum to one:
-%
-\begin{align}
-\w\T\1 &= \big(\alpha\a_\phi - (1-\alpha)\b_\phi\big)\T\1 = 1.
-\end{align}
-%
-This is true for any value of $\alpha$ if $\a_\phi\T\1 = 1$ and $\b_\phi\T\1 = 1$. Hence, we  preserve the unity gain constraint as long as we normalize $\a_\phi$ and $\b_\phi$.
-
-Now let $a=\a_\phi\T\x$ and $b=\b_\phi\T\x$. The beamformer output can then be written as:
-%
-\begin{align}
-|\w\H\x|^2 &= \Big| \alpha a - (1-\alpha)b \Big|^2 \nn
-&= \alpha^2 (aa^* + ab^* + a^*b + bb^*) \nn
-&- \alpha(ab^* + a^*b + 2bb^*) + bb^*
-\end{align}
-%
-This is a convex function with a single minimum, which we find by differentiating with respect to $\alpha$ and setting equal to 0:
-%
-\begin{align}
-\frac{\partial}{\partial\alpha} |\w\H\x|^2 
-&= 2\alpha (aa^* + ab^* + a^*b + bb^*) \nn
-&- (ab^* + a^*b + 2bb^*) = 0,
-\end{align}
-%
-which has the solution:
-%
-\begin{align}
-\alpha &= \frac{ab^* + a^*b + 2bb^*}{2(aa^* + ab^* + a^*b + bb^*)}.
-\end{align}
-%
-In this computation there are only 4 and 7 unique complex additions and multiplications, respectively.
-If we used this to analytically solve for $\alpha$, but perform the search for $\phi$, the computational complexity of LCA would be of O($MN_\phi)$ instead of O($MN_\alpha N_\phi)$.
-The solution for $\alpha$ would also yield the optimal beamformer output in the minimum variance sense.
-
-\section{Multimedia Files and Source Code}\label{app:source_code}
-
-This article is accompanied by some multimedia files and source code for determining the steering angles for LCA's windows. They are located at:
-
-\url{\articlePath}
-
-% use section* for acknowledgement
-\ifCLASSOPTIONcompsoc
-  \section*{Acknowledgments}
-\else
-  \section*{Acknowledgment}
-\fi
-
-The authors would like to thank Kongsberg Maritime, the Norwegian Defence Research Establishment (FFI), and Bundeswehr Technical Center for Ships and Naval Weapons, Maritime Technology and Research (WTD 71), Germany, for the equipment and effort needed to acquire data of the resolution test cross and Holmengraa, as well as for allowing us to use it.
-
-% Can use something like this to put references on a page
-% by themselves when using endfloat and the captionsoff option.
-\ifCLASSOPTIONcaptionsoff
-  \newpage
-\fi
-
-\ifBuildBibliography
-   \bibliographystyle{IEEEtran}
-   \bibliography{references}
-
-\else
-
-% Generated by IEEEtran.bst, version: 1.13 (2008/09/30)
-\begin{thebibliography}{10}
-\providecommand{\url}[1]{#1}
-\csname url@samestyle\endcsname
-\providecommand{\newblock}{\relax}
-\providecommand{\bibinfo}[2]{#2}
-\providecommand{\BIBentrySTDinterwordspacing}{\spaceskip=0pt\relax}
-\providecommand{\BIBentryALTinterwordstretchfactor}{4}
-\providecommand{\BIBentryALTinterwordspacing}{\spaceskip=\fontdimen2\font plus
-\BIBentryALTinterwordstretchfactor\fontdimen3\font minus
-  \fontdimen4\font\relax}
-\providecommand{\BIBforeignlanguage}[2]{{%
-\expandafter\ifx\csname l@#1\endcsname\relax
-\typeout{** WARNING: IEEEtran.bst: No hyphenation pattern has been}%
-\typeout{** loaded for the language `#1'. Using the pattern for}%
-\typeout{** the default language instead.}%
-\else
-\language=\csname l@#1\endcsname
-\fi
-#2}}
-\providecommand{\BIBdecl}{\relax}
-\BIBdecl
-
-\bibitem{Capon1969}
-J.~Capon, ``{High-resolution frequency-wavenumber spectrum analysis},''
-  \emph{Proceedings of the IEEE}, vol.~57, no.~8, pp. 1408--1418, 1969.
-
-\bibitem{Blomberg2013}
-A.~E.~A. Blomberg, A.~Austeng, R.~E. Hansen, and S.~A.~V. Synnes, ``{Improving
-  Sonar Performance in Shallow Water Using Adaptive Beamforming},'' \emph{IEEE
-  Journal of Oceanic Engineering}, vol.~38, no.~2, pp. 297--307, Apr. 2013.
-
-\bibitem{Blomberg2012a}
-A.~E.~A. Blomberg, A.~Austeng, and R.~E. Hansen, ``{Adaptive Beamforming
-  Applied to a Cylindrical Sonar Array Using an Interpolated Array
-  Transformation},'' \emph{IEEE Journal of Oceanic Engineering}, vol.~37,
-  no.~1, pp. 25--34, Jan. 2012.
-
-\bibitem{Lo2004}
-K.~Lo, ``{Adaptive Array Processing for Wide-Band Active Sonars},'' \emph{IEEE
-  Journal of Oceanic Engineering}, vol.~29, no.~3, pp. 837--846, Jul. 2004.
-
-\bibitem{Buskenes2014}
-J.~I. Buskenes, J.~P. Asen, C.-I.~C. Nilsen, and A.~Austeng, ``{An Optimized
-  GPU Implementation of the MVDR Beamformer for Active Sonar Imaging},''
-  \emph{IEEE Journal of Oceanic Engineering}, pp. 1--13, 2014.
-
-\bibitem{Asen2013}
-J.~P. \AA{}sen, J.~I. Buskenes, C.-I. {Colombo Nilsen}, A.~Austeng, and
-  S.~Holm, ``{Implementing capon beamforming on a GPU for real-time cardiac
-  ultrasound imaging.}'' \emph{IEEE transactions on ultrasonics,
-  ferroelectrics, and frequency control}, vol.~61, no.~1, pp. 76--85, Jan.
-  2014.
-
-\bibitem{Kim2014}
-K.~Kim, S.~Park, J.~Kim, S.-B. Park, and M.~Bae, ``{A fast minimum variance
-  beamforming method using principal component analysis},'' \emph{IEEE
-  Transactions on Ultrasonics, Ferroelectrics, and Frequency Control}, vol.~61,
-  no.~6, pp. 930--945, Jun. 2014.
-
-\bibitem{Asl2012}
-B.~M. Asl and A.~Mahloojifar, ``{A low-complexity adaptive beamformer for
-  ultrasound imaging using structured covariance matrix},'' \emph{IEEE
-  Transactions on Ultrasonics, Ferroelectrics and Frequency Control}, vol.~59,
-  no.~4, pp. 660--667, Apr. 2012.
-
-\bibitem{Vignon2008}
-F.~Vignon and M.~R. Burcher, ``{Capon beamforming in medical ultrasound imaging
-  with focused beams.}'' \emph{IEEE transactions on ultrasonics,
-  ferroelectrics, and frequency control}, vol.~55, no.~3, pp. 619--28, Mar.
-  2008.
-
-\bibitem{Synnevag2011}
-J.-F. Synnev\aa{}g, A.~Austeng, and S.~Holm, ``{A low-complexity data-dependent
-  beamformer},'' \emph{IEEE Transactions on Ultrasonics, Ferroelectrics and
-  Frequency Control}, vol.~58, no.~2, pp. 281--289, feb 2011.
-
-\bibitem{Stankwitz1995}
-H.~Stankwitz, R.~Dallaire, and J.~Fienup, ``{Nonlinear apodization for sidelobe
-  control in SAR imagery},'' \emph{IEEE Transactions on Aerospace and
-  Electronic Systems}, vol.~31, no.~1, pp. 267--279, Jan. 1995.
-
-\bibitem{Buskenes2011}
-J.~I. Buskenes, C.-I.~C. Nilsen, and A.~Austeng, ``{A Low Complexity Adaptive
-  Beamformer for Active Sonar Imaging},'' in \emph{Proceedings of the
-  Underwater Acoustic Measurements: Technologies \& Results}.\hskip 1em plus
-  0.5em minus 0.4em\relax Underwater Acoustic Measurements: Technologies \&
-  Results, 2011.
-
-\bibitem{Harris1978}
-F.~Harris, ``{On the use of windows for harmonic analysis with the discrete
-  Fourier transform},'' \emph{Proceedings of the IEEE}, vol.~66, no.~1, pp.
-  51--83, 1978.
-
-\bibitem{Kailath1985}
-T.~Kailath and T.-J. Shan, ``{Adaptive beamforming for coherent signals and
-  interference},'' \emph{IEEE Transactions on Acoustics, Speech, and Signal
-  Processing}, vol.~33, no.~3, pp. 527--536, Jun. 1985.
-
-\bibitem{Synnevag2009a}
-J.-F. Synnev\aa{}g, A.~Austeng, and S.~Holm, ``{Benefits of minimum-variance
-  beamforming in medical ultrasound imaging},'' \emph{IEEE Transactions on
-  Ultrasonics, Ferroelectrics and Frequency Control}, vol.~56, no.~9, pp.
-  1868--1879, Sep. 2009.
-
-\bibitem{Cox1987}
-H.~Cox, R.~Zeskind, and M.~Owen, ``{Robust adaptive beamforming},'' \emph{IEEE
-  Transactions on Acoustics, Speech, and Signal Processing}, vol.~35, no.~10,
-  pp. 1365--1376, Oct. 1987.
-
-\bibitem{Maksym1979}
-J.~N. Maksym, ``{A robust formulation of an optimum cross-spectral beamformer
-  for line arrays},'' \emph{The Journal of the Acoustical Society of America},
-  vol.~65, no.~4, p. 971, 1979.
-
-\bibitem{Asen2014}
-J.~P. \AA{}sen, A.~Austeng, and S.~Holm, ``{Capon beamforming and moving
-  objects--an analysis of lateral shift-invariance.}'' \emph{IEEE transactions
-  on ultrasonics, ferroelectrics, and frequency control}, vol.~61, no.~7, pp.
-  1152--60, Jul. 2014.
-
-\bibitem{Hansen2011}
-R.~E. Hansen, H.~J. Callow, T.~O. Sabo, and S.~A.~V. Synnes, ``{Challenges in
-  Seafloor Imaging and Mapping With Synthetic Aperture Sonar},'' \emph{IEEE
-  Transactions on Geoscience and Remote Sensing}, vol.~49, no.~10, pp.
-  3677--3687, Oct. 2011.
-
-\bibitem{holmengraa}
-{Article of Holmengraa on dykkepedia.com}.
-
-\bibitem{Stoica2005}
-P.~Stoica and R.~L. Moses, ``{Nonparametric Methods},'' in \emph{Spectral
-  Analysis of Signals}.\hskip 1em plus 0.5em minus 0.4em\relax Prentice Hall,
-  2005, ch.~2, pp. 23--89.
-
-\end{thebibliography}
-
-\fi
-
-\input{bio/jo_inge_buskenes}
-\input{bio/roy_edgar_hansen}
-\input{bio/andreas_austeng}
-
-\vfill
+\input{document}
 

### 2018-02-24 HRUIonGPU/2nd submit/jrnl_article_bak.tex
Adding HRUIonGPU article

--- 2013-06-05 HRUIonGPU/jrnl_article.tex
@@
 In this paper we extend the work done in \cite{asen2012} and \cite{Buskenes2013} by utilizing a beamspace transformation.
-The beamspace transform has been proposed used in ultrasound imaging by Nilsen and Hafizovic \cite{Nilsen2009} in order to maintain high frame rates as the number of elements increases while maintaining almost the same lateral resolution and sidelobe reduction capabilities of full Capon beamforming.
+The beamspace transform has been proposed used in ultrasound imaging by Nilsen and Hafizovic \cite{Nilsen2009} in order to maintain high frame rates as the number of elements increases while maintaining almost the same lateral resolution and sidelobe reduction capabilities of element-space Capon applied in narrow-beam systems.
 %The beamspace transformation matches well with the focused transmit beams used in cardiac ultrasound imaging. 
@@
 Threads inside a block support low-cost synchronization if needed.
-From a CUDA perspective the GPU consist of one or more streaming multi-processors (SM), where each SM is capable of 32 or more concurrent multiplications or additions, or both, depending on the architecture.
-A group of blocks is scheduled to each SM until all blocks have been processed.
+From a CUDA perspective the GPU consist of one or more streaming multi-processors (SM), where each SM is capable of executing multiple arithmetic operations per clock cycle.
+%32 or more concurrent multiplications or additions, or both, depending on the architecture. A group of blocks is scheduled to each SM until all blocks have been processed.
 
@@
 Assuming a system bandwidth of 80\% (relative to the center frequency) we need a complex sampling frequency of 2.7 MHz, and when imaging down to a depth of 15.4 cm we have approximately 540 samples per range line.
-The angular sampling rate is given by twice the Rayleigh criterion \cite{Hergum2007}
+The angular sampling is given by the Rayleigh criterion divided by two \cite{Hergum2007},
 \begin{equation}\label{eq:rayleigh}
-\delta{}\theta = \lambda/(2Mp).
+\delta{}\theta = \lambda/(2Mp),
 \end{equation}
-To sample a $70\degree$ sector with $64$- and $96$-element arrays with pitch $p=\lambda/2$ we therefore need approximately 80 and 120 receive beams respectively.
+where $\lambda$ is the pulse wavelength and $p$ is the element pitch. 
+To sample a $70\degree$ sector with $64$- and $96$-element arrays with $p=\lambda/2$ we therefore need approximately 80 and 120 receive beams respectively.
 
@@
 \ref{fig:bench}.
-The benchmarks are based on a 64-element and 96-elements phased array respectively using 80 receive beams.
 For each subarray size, the total execution time is split into one bar for each of the three steps described in Section \ref{sec:meth}, and the calculation of $\mat{\hat{R}}$ is presented for three different values of $K$.
@@
 \ref{fig:benchBS} we present similar benchmarks for the BS-Capon beamformer.
-Benchmarks are based on 64- and 96-element phased arrays, using 80 and 120 receive beams.
 The figure only presents results with use of temporal averaging ($K=2$).
@@
 
-For the simulated phantom, modulation effects were observed when the two bright point scatterers moved across receive-beams.
-The artefact did not disappear until 16x lateral oversampling relative to (\ref{eq:rayleigh}) was used on transmit.
+When the simulated phantom was processed with the ES-Capon beamformer, modulation effects were observed when the two bright point scatterers moved across receive-beams.
+The artifact did not disappear until 16x lateral oversampling relative to (\ref{eq:rayleigh}) was used on transmit.
 The cardiac recording of a 70\degree{} sector has a minimal lateral sampling of 98 receive beams given by (\ref{eq:rayleigh}).
-Despite this, the same artefacts were not observed here.
+Despite this, the same artifacts were not observed here.
 The image was acquired using 50 transmit beams and "Synthetic Transmit Beamforming" was used on receive \cite{Hergum2007}.
@@
 \ref{fig:speed_res_trade}, $d=0.01$ has been used on both methods in order to obtain the same resolution for $L=N_b=M/2$.
-We considered this to be a small value for ES-Capon and a normal value for BS-Capon (cf.
+In Fig.
+\ref{fig:phantom}, the diagonal loading factor for BS-Capon was adjusted to make the two results equal for $L=32$ and $N_b=3$.
+The resolution for ES-Capon is therefore better here than what is presented in Fig.
+\ref{fig:phantom}.
+%We considered this to be a small value for ES-Capon and a normal value for BS-Capon (cf.
 Section \ref{sec:bs}), hence the resolution for ES-Capon is here better than what is presented in Fig.
 \ref{fig:phantom}.
-The precision of the presented measurements is 0.12 mm, dictated by the simulated frame rate, 25 FPS, and the relative speed of the two point scatterers, 3 mm/s.
+The precision of the presented measurements is 0.12 mm, dictated by the simulated frame rate of 25 FPS, and the relative speed of the two point scatterers, 3 mm/s.
 
@@
 \section{Discussion}\label{sec:dis}
-In this paper we have presented a parallel implementation of the Capon beamformer in beamspace for high-resolution cardiac ultrasound imaging.
+In this paper we have presented a parallel implementation of the BS-Capon beamformer for high-resolution cardiac ultrasound imaging.
 The results show that it is now possible to apply BS-Capon in real-time in cardiac ultrasound imaging by utilizing the power of a GPU.
@@
 $L=M/2$).
-These extra lines might be needed for increased sampling instead of increased frame rate since we had to use oversampling in our simulation.
-If parallel receive lines are used to obtain denser sampling, acquisition time remains the same but processing times are increased with a factor equal to the amount of oversampling.
-If parallel receive lines are used for increased frame rate, the acquisition time is lowered, and the processing time remains constant.
+However, since we had to use oversampling in our simulation, these extra lines might be needed for increased sampling instead of increasing the frame rate.
+If parallel receive lines are used to obtain denser sampling, the acquisition time remains the same but the processing time is increased with a factor equal to the amount of oversampling.
+If parallel receive lines are used to increase the frame rate, the acquisition time is lowered, and the processing time remains constant.
 Hence, four parallel receive lines are not possible to process in real-time using BS-Capon with $N_b=3$ in combination with our single GPU setup.
@@
 The per-pixel granularity suggests that the beamformer is easily dividable across multiple GPUs.
-Real-time execution for all cases presented in this paper is therefore possible if several GPUs are used simultaneously.
-However, there is usually a space-cost-energy limitation for how many GPUs that can be fitted in an ultrasound scanner.
+It is therefore possible to achieve higher frame rates if several GPUs are used simultaneously.
+However, there is usually a space-cost-energy limitation for how many GPUs that can be assembled in an ultrasound scanner.
 
@@
 \ref{fig:phantom}, and the attached video of the simulation, that Capon beamforming provides improved lateral resolution of moving point targets both in element- and in a reduced beam-space ($N_b=3$).
-Our preliminary finding is that the Capon beamformers are local shift invariant as long as the lateral sampling is increased to a certain level \multimedia{Media-Movie 1 and 2}.
-How large this oversampling factor needs to be and how the oversampling should be conducted is outside the scope of this paper.
+Our preliminary finding is that the Capon beamformers are local shift invariant as long as the lateral sampling is increased to a certain level in simulations \multimedia{Media-Movie 1 and 2}.
+%How large this oversampling factor needs to be and how the oversampling in the end is conducted will impact the real time requirement, and .
+Future work should focus on how large this oversampling needs to be and how it should be conducted in order to still achieve real time processing.
 No major improvement in contrast is visually observed in the presented images and videos, other than slightly rounder circles due to mainlobe micro steering.
@@
 Future work should focus on solving this issue.
-Cardiac ultrasound has not been selected as the target modality in this paper because Capon beamforming shows large improvements in image quality in particular, it has been selected because it is the most computationally demanding modality (measured in float operations per second).
-The rationale is that if BS-Capon can run in real time for cardiac ultrasound it can run in real time for all medical ultrasound modalities.
-We believe that Capon will show more improvements in other imaging modalities.
-
-Nilsen and Hafizovic suggested to do automatic selection of beams in beamspace, hence picking the beams containing most energy.
+
+In the end we want to point out that cardiac ultrasound has not been selected as the target modality in this paper because Capon beamforming shows large improvements in image quality in particular.
+It has been selected because it is the most computationally demanding modality, measured in floating points operations per second.
+The rationale being that if BS-Capon can run in real time for cardiac ultrasound it can run in real time for all medical ultrasound modalities.
+We believe that Capon beamforming may show more improvements in other imaging modalities where detection of small point targets is important.
+
+%Nilsen and Hafizovic suggested to do automatic selection of beams in beamspace, hence picking the beams containing most energy.
 In this paper the center beams are always picked.
@@
 In this paper a parallel formulation and implementation of the beamspace Capon beamformer have been presented, and
-the beamformer have successfully been implemented on the GPU.
+the beamformer has successfully been implemented on the GPU.
 For a three dimensional beamspace, processing rates supporting real-time cardiac ultrasound imaging was shown to be possible, meaning that processing rates higher than the image acquisition rate are obtained for a wide range of parameters.

### 2018-02-24 HRUIonGPU/conf_article.tex
Adding HRUIonGPU article

@@
 % can use linebreaks \\ within to get better formatting as desired
-\title{Implementing Capon Adaptive Beamforming on the GPU for Real Time Cardiac Ultrasound Imaging}
+\title{Implementing Capon Beamforming on the GPU for Real Time Cardiac Ultrasound Imaging}
 
@@
 \boldmath
-Recent results \cite{Synnevag2007, Austeng2008, Vignon2008, Viola, Mehdizadeh2012} suggest that the Capon beamformer can provide increased lateral resolution when applied in medical ultrasound imaging.
-In this paper, the high computational complexity of the Capon beamformer is targeted with the use of Graphics Processing Units (GPUs).
-We also present in-vivo results with Capon beamforming applied on a full cardiac cycle in addition to simulations.
-With GPUs we are able to process a $70\degree$ sector cardiac image from a 64 element phased array at interactive frame rates using both spatial and temporal smoothing.
+Recent work on Capon beamforming suggest that it can provide increased lateral resolution when applied in a medical ultrasound imaging setting.
+In this paper, the high computational complexity of the Capon beamformer is targeted with the use of a Graphics Processing Unit (GPU).
+\textsl{In-vivo} results with Capon beamforming applied on a full cardiac cycle in addition to simulations are presented.
+With the GPU we are able to process a $70\degree$ sector cardiac image from a 64 element phased array at interactive frame rates using both spatial and temporal smoothing.
 
 For a typical cardiac ultrasound image of $80\times400$ pixels ($70\degree$ sector, 15 cm range) acquired using a 2.5 MHz, M=64 element phased array, we obtain 10 fps (subarray length L=M/2, temporal smoothing over 3 samples).
-If we do a 2-element pre-beamforming to get the channel count down to 32, the frame rate increases to 44 fps.
+If we perform a 2-element pre-beamforming, the channel count is reduced to 32, and frame rate is increased to 44 fps.
 For a 32 element phased array we need less beams to cover the sector ($40\times400$ pixels), hence with the same parameters the frame rate increases to 87 fps.
-After optimization the bottleneck of the Capon algorithm is to find the inverse covariance matrix for all pixels.
+The target GPU was the Nvidia’s Quadro 6000, capable of 1 Tflops.
+
+%After optimization the bottleneck of the Capon algorithm is to find the inverse covariance matrix for all pixels.
 This is an $O(L^3)$ operation, and when L is larger than 16 (M = 64), real time performance is not possible for cardiac imaging with current high-end hardware.
-The target GPU has been Nvidia’s Quadro 6000, capable of 1 Tflops.
 
@@
 \section{Introduction}
-Capon adaptive beamforming \cite{Capon1969} applied to medical ultrasound imaging has recently shown promising results in several studies \cite{Synnevag2007, Austeng2008, Vignon2008, Viola, Mehdizadeh2012}.
-For instance, some interesting trade-offs in addition to just improved resolution were introduced, like maintaining image resolution in combination with higher framerates, smaller probes, or deeper penetration \cite{Synnevag2009}.
+Capon beamforming \cite{Capon1969} applied to medical ultrasound imaging has recently shown promising results in several studies \cite{Synnevag2007, Austeng2008, Vignon2008, Viola, Mehdizadeh2012}.
+Some interesting trade-offs in addition to just improved resolution has been introduced, like maintaining image resolution in combination with higher framerates, smaller probes, or deeper penetration \cite{Synnevag2009}.
 Despite these benefits, widespread use has been limited by the immense computational complexity associated with the method \cite{So2011}.
-In this paper we introduce the first implementation of the Capon beamformer exhibiting real-time performance when applied in a typical cardiac ultrasound imaging setting.
+In this paper we introduce an implementation of the Capon beamformer aiming for real-time performance when applied in a typical cardiac ultrasound imaging scenario.
 To accomplish this task we make use of the massive, and easily accessed, parallel processing power found in modern Graphics Processing Units (GPUs).
@@
 For medical ultrasound, Chen et al.
-\cite{Chen2011a, Chen2011} have implemented a Capon beamformer on the GPU that produced real apodization weights based on real data.
-Under this constraint the Capon beamformer is not allowed to micro steer the main lobe, and therefore its full potential is not reached.
-Micro steering of the main lobe is an important property of the Capon beamformer in order for it to improve edge definitions and contrast throughout an ultrasound image.
-Another aspect of using complex data is that even though the number of aritmetic operations per sample is increased we can now sample at the system bandwidth, keeping the number of samples for a full cardiac imaging sector at a minimum.
+\cite{Chen2011a, Chen2011} implemented a Capon beamformer on the GPU that produced real apodization weights based on real data.
+Under this constraint the Capon beamformer is not allowed to micro-steer the main lobe, and therefore its full potential is not reached.
+Micro-steering of the main lobe is an important property of the Capon beamformer in order for it to improve structural details and edge definitions throughout an ultrasound image.
+Another aspect of using complex data is that even though the number of aritmetic operations per sample is increased we can sample at the system bandwidth, keeping the number of samples for a full cardiac imaging sector at a minimum.
 All results in this paper are reported with the use of baseband data.
@@
 \section{Background}
-\subsection{Capon Adaptive Beamforming}
-The conventional form of array beamforming is delay-and-sum (DAS).
+\subsection{Capon Beamforming}
+The conventional form of array beamforming in medical ultrasound imaging is delay-and-sum (DAS).
 Here, the signal $x_m[n]$ recorded at element number $m \in [0,M-1]$ at time $n$ is delayed to broadside with an appropriate delay $\Delta_m[n]$.
@@
 
-As shown in Figure \ref{fig:mvbf}, the basic idea of an adaptive beamformer is to form a weight vector based on the data at hand.
-For Capon adaptive beamforming the weights are found that minimize the beamformer output power while maintaining unit gain at broadside.
+As shown in Fig.
+\ref{fig:mvbf}, the basic idea of an adaptive beamformer is to form a weight vector based on the data at hand.
+In Capon beamforming the weights are found that minimize the beamformer output power while maintaining unit gain at broadside.
 The result is that interference impinging from other directions than the point of focus is suppressed \cite{Synnevag2007}.
@@
 \begin{figure}
-\centerline{\includegraphics[width=3in]{gfx/beamforming_mv.pdf}}
-\caption{Adaptive beamforming.
+\centerline{\includegraphics[width=3.5in]{gfx/beamforming_mv.pdf}}
+\caption{Beamforming with adaptive generation of weights.
 Impinging signal is demodulated, and aligned in phase by applying steering and focusing delays $\Delta_m[n]$.
@@
 
-To calculate the Capon beamformer output we need three equations. First the sample covariance matrix has to be calculated,
+The Capon beamformer can be summarized in three main steps. First a sample covariance matrix has to be calculated,
 \begin{align}
@@
 Note that the dimension of the covariance matrix is $L \times L$.
-The covariance matrix is usually loaded with a diagonal factor $\epsilon$ to ensure numerical stability,
+The covariance matrix is then loaded with a diagonal factor $\epsilon$ to ensure numerical stability,
 \begin{align}\label{eq:diag}
-\mat{\hat{R}} = \mat{\breve{R}} + \epsilon\mat{I}.
+\mat{\hat{R}}[n] = \mat{\breve{R}}[n] + \epsilon\mat{I}.
 \end{align}
@@
 \begin{align}\label{eq:diag_adapt}
-\epsilon &= d \times tr\{\mat{\breve{R}}\}, \\
-d &= 1/L
+\epsilon[n] &= d \times \frac{tr\{\mat{\breve{R}}[n]\}}{L},
 \end{align}
-has been applied in much of the resent literature on Capon beamforming for medical ultrasound imaging \cite{Synnevag2007, Mehdizadeh2012}.
+has been applied in much of the recent literature on Capon beamforming for medical ultrasound imaging \cite{Synnevag2007, Mehdizadeh2012}.
 
-Second, we calculate the Capon weight vector as
+Second, we solve the linear system of equations 
+\begin{align}\label{eq:b}
+\vec{b}[n] = \mat{\hat{R}}[n]^{-1}\vec{a},
+\end{align}
+from which we derive the Capon weight vector as
 \begin{align}\label{eq:w}
-\vec{w}[n] = \frac{\mat{\hat{R}}[n]^{-1}\vec{a}}{\vec{a}^H\mat{\hat{R}}[n]^{-1}\vec{a}} = \frac{\vec{b}}{\vec{a}^H\vec{b}},
+\vec{w}[n] = \frac{\vec{b}[n]}{\vec{a}^H\vec{b}[n]}.
 \end{align}
-where $\vec{a} = [1_0, 1_1, ..., 1_{L-1}]^T = \vec{1}_L$ when $\vec{x}$ is pre-delayed. 
-And last, the beamformer output is calculated by combining all $N_L$ subarrays weighted with the same set of adaptive weights,
+Note that  $\vec{a} = [1_0, 1_1, ..., 1_{L-1}]^T = \vec{1}_L$ since $\vec{x}^\Delta$ is pre-delayed. 
+Finally, the beamformer output is calculated by combining all $N_L$ subarrays weighted with the same set of adaptive weights,
 \begin{align}
@@
 \subsection{GPU Compute Model}
-CUDA by Nvidia has been our choice of GPU developer environment \cite{Nvidia2011}.
-The CUDA compute model is based on execution of a kernel function across a grid of compute threads, where the kernel function describes the work to be done by each thread.
-As depicted in Figure \ref{fig:gpulayout}, each position in the grid holds a block of threads, with maximal size of 1024 threads, that supports low-cost synchronization if needed.
+CUDA by Nvidia has been our choice of GPU developing environment \cite{Nvidia2011}.
+The CUDA compute model is based on execution of a kernel function across a large set of threads, where the kernel function describes the work to be done in each thread.
+As depicted in Fig.
+\ref{fig:gpulayout}, threads are organized in a grid of thread-blocks, with maximal size of 1024 threads per block.
+Threads inside a block supports low-cost synchronization if needed.
 %Each block is further divided into groups of 32 threads, known as a warp. The maximum number of threads inside a block is restricted to 512, hence 16 warps. 
-From a CUDA perspective the GPU consists of one or more stream multi processors (SMs), where each SM performs single-instruction-multiple-data (SIMD) operations.
-A group of blocks is scheduled to each SM until all blocks in the grid have been processed.
+From a CUDA perspective the GPU consists of one or more streaming multi processors (SMs), where each SM performs single-instruction-multiple-data (SIMD) operations.
+A group of blocks is scheduled to each SM until all blocks have been processed.
 
-An SM has available a limited amount of near-core memory and registers.
-Each thread therefore has to be as light on memory as possible and has to perform enough instructions per used byte in order to hide memory latency.
+An SM has a limited amount of near-core memory and registers.
+Each thread therefore has to be as light on memory as possible and has to perform enough instructions per used byte in order to hide memory and instruction latency.
 Global reads and writes on the GPU, in addition to CPU-GPU transfers, have high penalty, and should be minimized as much as possible.
 All of these constraints combined pose a real challenge as the target problem has to map on to the parallel framework both on a grid, block and memory consumption level in order to benefits from parallel acceleration.
-In the next section, we will describe how this mapping can be done for the Capon beamformer.
+In the next section, we will describe how this mapping can be performed for the Capon beamformer.
 
 \begin{figure}
-\centerline{\includegraphics[width=3in]{gfx/gpu_layout_vertical_lr.png}}
+\centerline{\includegraphics[width=3.5in]{gfx/gpu_layout_vertical_lr.png}}
 \caption{Calculation of Capon adaptive weights mapped to GPU architecture.
-A recorded echo across the array for a given range and angle (selected cell) is assigned to a group of $L$ threads that can run independently of all other groups.
-Note that several samples can be processed per block of threads if $L$ is small.
+A recorded echo across the array for a given range and angle (selected cell) is assigned to a group of $L$ threads that can run independently of all other groups.} %Note that several samples can be processed per block of threads if $L$ is small.
 The task of these $L$ threads is to calculate the Capon beamformer output in the three depicted steps.}
@@
 If we analyze (\ref{eq:z_mv}) in combination with (\ref{eq:w}) and (\ref{eq:R}), we see that each calculation of a set of Capon weights is independent across the output $z[n]$.
-As shown in Figure \ref{fig:gpulayout}, a first level of granularity is therefore to divide the image into a grid, where the output in each cell can be calculated independently of all others.
-Further, based on memory dependencies and to hold memory consumption per thread at a minimum, one should recognise that a group of threads should process one output in the final image.
+As shown in Fig.
+\ref{fig:gpulayout}, a first level of granularity is therefore to divide the image at sample level.
+Further, based on memory dependencies and to hold memory consumption per thread at a minimum, a group of threads should process one output in the final image.
 Continuing with the calculation of one weight vector, the algorithm can be broken down into three main steps that, to some extent, can be further divided into a small set of parallel processes.
@@
 %\begin{array}{l}
-\text{1) } &\text{Calculate the sample covariance matrix } \mat{\hat{R}} \text{ as in } \text{(\ref{eq:diag})}\\
-\text{2) } &\text{Solve the linear system of equations } \vec{b} =\mat{\hat{R}}^{-1}\vec{1}_L\\
-\text{3) } &\text{Calculate } \vec{w} = \vec{b}/\vec{1}_{L}^H\vec{b} \text{ and the amplitude as in (\ref{eq:z_mv})}
+\text{1) } &\text{Calculate the sample covariance matrix } \mat{\hat{R}} \text{ as in } \text{(\ref{eq:diag})}.\\
+\text{2) } &\text{Solve the linear system of equations presented in (\ref{eq:b})}.\\
+\text{3) } &\text{Calculate } \vec{w} \text{ as in (\ref{eq:w}) and } z \text{ as in (\ref{eq:z_mv}).}
 %\end{array}
@@
 
-The computational complexity of these three steps is $O(L^3)$, $O(L^3)$, and $O(L)$ respectively.
+The computational complexity of these three steps is $O(L^2)$ (with optimization), $O(L^3)$, and $O(L)$, respectively.
 
 \subsection{Calculating Multiple Sample Covariance Matrices}
-The straightforward sum in (\ref{eq:R}) has complexity $O(L^2N_LN_K)$.
-Note that for a typical array configuration ($M \in [32, 128]$) we do more spatial than temporal smoothing.
+%The straightforward sum in (\ref{eq:R}) has complexity $O(L^2N_LN_K)$. 
+%Note that for a typical array configuration ($M \in [32, 128]$) we do more spatial than temporal smoothing.
 Hence, $N_K$ is usually small compared to the number of subarrays ($N_L$).
@@
 In this interval, calculating the sample covariance matrix is cubic with $L$.
-This is actually the same as for inverting an $L \times L$ matrix.
-However little attention has been devoted to this step in the literature.
 
@@
 A natural level of granularity would then be to assign one thread to each element of $\mat{\hat{R}}$ as in \cite{Chen2011}.
-This will minimize global reads and writes per thread, but unfortunately the maximum number of threads per block puts an upper limit on the subarray length we then can process, $L_{max}=32$, in order to avoid block-to-block synchronization.
+This will minimize global reads and writes per thread, but unfortunately the maximum number of threads per block puts an upper limit on the subarray length one then can process, $L_{max}=32$, in order to avoid block-to-block synchronization.
 
+The straightforward sum in (\ref{eq:R}) has complexity $O(L^2N_LN_K) = O(L^3)$. %O((\alpha-1)L^3+L^2)$ ($\alpha = M/L$). 
+This is the same as for inverting an $L \times L$ matrix.
+In this work we therefore focus on optimizing this step in order to reduce the total complexity.
 The cubic complexity with $L$ can be reduced by realizing that the calculation in (\ref{eq:R}) overlaps across subarrays.
@@
 The complexity of calculating $\mat{\hat{R}}$ is now reduced to $O(L^2)$.
-Important to notice is that sliding will break independence in the sliding dimension.
+It is important to notice that sliding will break independence along the sliding dimension.
 The calculation in (\ref{eq:sliding}) is still independent across time, but not along diagonals in $\mat{\hat{R}}$.
 We have therefore selected to use one thread per diagonal in the upper triangle of $\mat{\hat{R}}$.
-And for small subarrays, several covariance matrices are calculated per thread block (Figure \ref{fig:gpulayout}).
-Maximum supported $L$ is with this approach increased to 512, however for small $L$ high $M$ configurations, memory consumption per thread will be too high to achieve full occupancy of the GPU.
+And for small subarrays, several covariance matrices are calculated per thread block (Fig.
+\ref{fig:gpulayout}).
+Maximum supported $L$ is with this approach increased to 512.%, however for small $L$ high $M$ configurations, memory consumption per thread will be too high to achieve full occupancy of the GPU.
 
@@
 However, the way we have decided to organize the GPU compute threads per $\mat{\hat{R}}$, the lower triangle can be calculated with little overhead on the GPU.
-This also means that we do not rely on an upper triangular solver to solve $\mat{\hat{R}}\vec{b} = \vec{1}_L$, even if that could have saved some instructions.
+This also means that we do not rely on an upper triangular solver to solve $\mat{\hat{R}}\vec{b} = \vec{1}_L$.%, even if that could have saved some instructions.
 
@@
 GPU libraries therefore often lack a comprehensive collection of batched solvers.
-The reason for this is two-fold, first the GPU needs to solve thousands of small systems in order to beat the low memory latency of the CPU, second there is a range of system dimensions that does not fit well with the GPU architecture, where only a small amount of near-core memory and registers are available.
-Because of this, it has proven hard to optimize a manageable set of solvers that provides speedups compared with the CPU in every case.
-It is therefore not given that solving on the GPU is faster than solving on the CPU in general.
+The reason for this is two-fold.
+First the GPU needs to solve thousands of small systems in order to beat the low memory latency of the CPU, second there is a range of system dimensions that does not fit well with the GPU architecture, where only a small amount of near-core memory and registers are available.
+%Because of this, it has proven hard to optimize a manageable set of solvers that provides speedups compared with the CPU in every case. 
+It is therefore not necessarily true that solving on the GPU is faster than solving on the CPU.
 However, there are two arguments for doing solving on the GPU.
 First the CPU, which might already runs a lot of algorithms, is offloaded.
-Second the transfer of several thousand covariance matrices from CPU-side to GPU-side is spared.
+Second the transfer of several thousand covariance matrices from CPU-side to GPU-side is avoided.
 
-For the results presented in this paper, we have used an unreleased GPU implementation of batched Gauss-Jordan (GJ) elimination (by Nvidia) to solve $\mat{\hat{R}}\vec{b} = \vec{1}_L$.
+For the results presented in this paper, we have used a GPU implementation of batched Gauss-Jordan (GJ) elimination by Nvidia to solve $\mat{\hat{R}}\vec{b} = \vec{1}_L$ ($L_{max} = 76$).
 
@@
 The formula in (\ref{eq:z_mv}) has been implemented with one thread per element-weight ($L$ threads per weight vector).
-After the solution vector $\vec{b}$ has been placed in near-core memory, one of the $L$ threads finds the sum of $\vec{b}$.
-Each of the $L$ threads then reads one element of $\vec{b}$ and scales it with the inverse sum just found to get the weight vector $\vec{w}[n]$.
+After the solution vector $\vec{b}[n]$ has been placed in near-core memory, one of the $L$ threads finds the sum of $\vec{b}[n]$.
+Each of the $L$ threads then reads one element of $\vec{b}[n]$ and scales it with the inverse sum just found to get the weight vector $\vec{w}[n]$.
 
@@
 \begin{figure*}[!t]
-\centerline{\subfloat[]{\includegraphics[width=3.49in]{gfx/benchmark_bar_32_40_400.png} \label{fig:benchUS32}}
+\centerline{\subfloat[]{\includegraphics[width=3.7in]{gfx/benchmark_bar_32_40_400.png} \label{fig:benchUS32}}
 \hfil
-\subfloat[]{\includegraphics[width=3.49in]{gfx/benchmark_bar_64_80_400.png}%
+\subfloat[]{\includegraphics[width=3.7in]{gfx/benchmark_bar_64_80_400.png}%
 \label{fig:benchUS64}}}
@@
 Execution time is plotted as a function of subarray length $L \in [1, M/2]$.
-The figures show execution times for the three steps listed in Section \ref{sec:meth}: Calculation of covariance matrices in light gray, solver in gray, and the final beamformer sum in dark gray.
-For the calculation of covariance matrices, there is one bar per choice of temporal averaging, $K \in [0, 1, 2]$.
+The figures show execution times for the three steps listed in Section \ref{sec:meth}.
 (a) A 32 element array.
@@
 \section{Benchmarks and Videos}
-Figure \ref{fig:bench} shows the results from benchmarking the implementation described in Section \ref{sec:meth} on an Nvidia Quadro 6000 graphics card (14 SMs, 1 TFlops single precision, 6 GB global memory, and 144 GB/s memory bandwidth).
+Fig.
+\ref{fig:bench} shows the results from benchmarking the implementation described in Section \ref{sec:meth} on an Nvidia Quadro 6000 graphics card (14 SMs, 1 TFlops single precision, 6 GB global memory, and 144 GB/s memory bandwidth).
 The problem sizes are calculated based on 32 and 64 element arrays with center frequency 2.5 MHz.
@@
 If we image down to a depth of 15 cm we get approximately 400 samples per range line.
-To correctly sample a $70\degree$ sector with 64 and 32 element arrays ($\lambda/2$ pitch) we need at least 80 and 40 beams respectively.
+To correctly sample a $70\degree$ sector with 64 and 32 element arrays ($\lambda/2$ pitch), we need at least 80 and 40 beams, respectively.
 The presented execution times do not include transfer of data from CPU-side to the GPU.
@@
 
-At (cite) one will find videos of a cardiac loop, 
+There are two videos attached to this paper. One cardiac loop, 
 %(add ref to video)
-and a simulated phantom 
+and one simulation, both 
 %(add ref to videos) 
-processed with different types of settings given to the Capon beamformer.
+processed with the following Capon beamformer settings: $L = 32$, $K = 2$, $d=0.01$.
+These settings have previously been shown to provide high resolution while maintaining DAS-like speckle \cite{Synnevag2007a} in ultrasound images.
 Cardiac data was acquired using a 64 element and 3.5 MHz phased array.
 Simulations were done using a 64 element 2.5 MHz phased array.
+For the simulation, 16x lateral over-sampling has been used, in order to avoid under sampling of the two bright point scatterers.
+The cardiac recording only has $30\%$ lateral over sampling.
 
 \section{Discussion}
-From Figure \ref{fig:benchUS64} we see that for $L = M/2 = 32$ and $K=1$, total execution time is 100 ms (10 fps).
+From Fig.
+\ref{fig:benchUS64} we see that for $L = M/2 = 32$ and $K=1$, total execution time is 100 ms (10 fps).
 Taking into account that we use complex data, but less samples per image, this is similar to the performance reported by Chen et al.
 \cite{Chen2011}.
-For the 32 element array in Figure (\ref{fig:benchUS32}), execution time is reduced to 11.5 ms (87 fps) using the same settings ($L=M/2, K = 1$).
+For the 32 element array in Fig.
+(\ref{fig:benchUS32}), execution time is reduced to 11.5 ms (87 fps) using the same settings ($L=M/2, K = 1$).
 Doubling this number can be interpreted as pre-beamforming a 64 element array down to 32.
 The frame rate for a 64 element array is then increased from 10 to 44 fps.
+Note that lateral oversampling will increase the total execution time with a factor equal the amount of oversampling.
 
@@
 As for now, we have shown that the Capon beamformer can be implemented in a GPU framework.
-And if we as in \cite{Chen2011} define real-time as 10 fps, all execution times  in Figure (\ref{fig:bench}) are real time.
+And if we as in \cite{Chen2011} define real-time as 10 fps, all execution times  in Fig.
+(\ref{fig:bench}) are real time.
 If we on the other hand define real-time as the time it takes to acquire a given image, the image formed using a 32 and 64 element array has a real time requirement of 125 fps and 62.5 fps respectively.
@@
 However we see that real time, by the last definition, is possible for $M=64, L < 16$, and for $M = 32, L < 13$.
-(Should comment on videos? Ran out of space)
+Another option to reach real-time execution is to use several GPUs simultaneously.
+Note that multi-line acquisition will increase the frame rate requirement accordingly.
+
+In the attached video of a simulated phantom one can see that Capon beamforming highly resolves moving point targets.
+The beamformer is shift invariant as long as the lateral sampling is increased.
+No major improvement in contrast is observed.
+In the cardiac recording, structures are thinner when the Capon beamformer is applied, however we observe a loss in contrast due to signal cancellation.
+Future work should focus on solving this issue.
 
@@
 \section*{Acknowledgment}
-The authors would like to thank Tore Bj\aa{}stad at GE Vingmed Ultrasound for providing cardiac data, and Nvidia for their batched solver written in CUDA.
+The authors would like to thank Nvidia for their batched linear equation solver written in CUDA.
 

### 2018-02-24 LCAforActiveSonarImaging/document.tex
Adding HRUIonGPU article

@@
    \begin{figure}[tbp]\centering%
-   \includegraphics[width=\linewidth]{gfx/fig1.eps}%
+   \includegraphics[width=.8\linewidth]{gfx/fig1.eps}%
 \else
@@
    \begin{figure}[tbp]\centering%
-   \includegraphics[width=\linewidth]{gfx/fig2.png}%
+   \includegraphics[width=.7\linewidth]{gfx/fig2.png}%
 \else
@@
 
+\ifPhdDoc
+   \begin{figure*}[tbp]\centering%
+   \includegraphics[width=\linewidth]{gfx/fig3_phd.pdf}%
+   \caption{\emph{Determining LCA window type}: MVDR image with typical frequency responses for windows used in various pixel regions.\\[.5\baselineskip]
+   \textbf{Left:} MVDR sectorscan image of the oiltanker Holmengraa, with 40x40 pixel groups for the shadow, highlight and speckle region of the image indicated with red boxes.\\[.5\baselineskip]
+   \textbf{Right:} MVDR window amplitude and phase responses computed from the 40x40 pixel groups.
+The responses are overlayed each other and the amount of overlap is colored using a logarithmic scale.
+The size of the pixel groups were chosen ad-hoc for the histograms to be visually invariant to a shift in position within the same region.
+Since each pixel is pre-delayed into focus the unsteered responses all have their center at broadside.
+The dashed red lines at -13\,dB and -43\,dB marks the peak sidelobe levels of an unsteered rectangular and Hamming window, respectively.
+Note how the responses are more or less symmetric, with very little steering in the shadow, moderate steering in speckle and steering within roughly 3\,dB in highlight.
+The phase varies most in the highlight region where we see the highest contrast.}\label{mvdr_selected_windows}
+\else
 \setcounter{topnumber}{1}
@@
 
-\ifPhdDoc
-   \begin{figure*}[tbp]\centering%
-   \includegraphics[width=\linewidth]{gfx/fig3.pdf}%
-\else
    \begin{figure*}[t]\centering%
    \includegraphics[width=\linewidth]{gfx/buske3_online.pdf}%
-\fi%
    \caption{\emph{Determining LCA window type}: MVDR image with typical frequency responses for windows used in various pixel regions.\newline
@@
 The phase varies most in the highlight region where we see the highest contrast.} }\label{mvdr_selected_windows}
+\fi%
+
 \end{figure*}
@@
    \includegraphics[width=\linewidth]{gfx/fig4.pdf}%
+   \caption{\emph{Determining Kaiser parameter boundaries}: LCA images of the resolution test cross and anchor created with a large window database, but where the upper bounds for $\beta$ and $\phi$ are varied.
+The upper left image is equal to DAS with a rectangular window.
+For all images we measure the lateral distance $\Delta x$ between the two \protect\minus{}3\,dB points of the anchor (red object).
+These are specified relatively to the reference distance  $\Delta x_\text{ref}$ from the upper left DAS image.
+   \\[.5\baselineskip]
+   \textbf{Upper bound $\beta$:} Observe that LCA gets better at suppressing sidelobes in the image as we increase the upper bounds for $\beta$.
+\Fig{mvdr_selected_windows} suggests that MVDR prefers windows with sidelobe levels lower than that of Kaiser with $\beta=2$, but here we observe further improvement going to $\beta=5$.
+   \\[.5\baselineskip]
+   \textbf{Upper bound $\phi$:} As we increase the upper bound of the steering $\phi$, we also increase the lateral image resolution.
+   }
 \else
@@
    \includegraphics[width=\textwidth]{gfx/buske4_online.pdf}%
-\fi%
    \caption{\emph{Determining Kaiser parameter boundaries}: LCA images of the resolution test cross and anchor created with a large window database, but where the upper bounds for $\beta$ and $\phi$ are varied.
@@
    \parbox[t]{.89\linewidth}{As we increase the upper bound of the steering $\phi$, we also increase the lateral image resolution. }
-}\label{oversampling_mosaic_bounds}
+   }
+\fi%
+\label{oversampling_mosaic_bounds}
 \end{figure*}
@@
 \ifPhdDoc
-   \begin{figure*}[tbp]\centering%
+   \begin{figure*}[tb]\centering%
    \includegraphics[width=\linewidth]{gfx/fig6.pdf}%
@@
 
+\ifPhdDoc
+\clearpage
+\appendix
+\renewcommand\thesection{\Roman{section}}
+\else
 \appendices
+\fi
 

### 2018-02-24 HRUIonGPU/gfx/all_plots.py

@@ -1,0 +1,279 @@
+'''
+Created on Mar 22, 2011
+
+@author: me
+'''
+
+#import framework.calc as calc
+#from framework.calc import db
+#from framework.mynumpy import size, arange, log10, pi, linspace, mean, cos, sin, arcsin, unique, zeros
+from numpy import size, arange, log10, pi, linspace, mean, cos, sin, arcsin, unique, zeros
+#from framework.mynumpy import mode_
+#import framework.mynumpy as np
+import numpy as np
+from scipy import io
+#from framework.
+from pylab import legend,grid, plot, imshow, ion, ioff, figure, savefig, cm, xlim, ylim, clim, xticks, subplot, subplots_adjust, get_current_fig_manager, setp, colorbar, xlabel, ylabel,show, gca
+#import mypylab as pl
+#from mpl_toolkits.axes_grid1 import ImageGrid
+#from framework.gfuncs import error
+
+GFX_PATH='/home/me/Work/Phd/Code/gfx'
+
+
+
+def plotBenchmarkGPU(filename, M, Nx, Ny, fromL, toL):
+
+   print "Reading ", filename
+   
+   timings = np.loadtxt(filename)
+
+   print 'Timings ', timings.shape
+   
+   min_L = fromL
+   max_L = toL
+
+   yavg1 = timings[0::3,:]
+   yavg2 = timings[1::3,:]
+   yavg3 = timings[2::3,:]
+   yavg1_sum = yavg1.sum(1)
+   yavg2_sum = yavg2.sum(1)
+   yavg3_sum = yavg3.sum(1)
+   
+   print 'Yavg1 ', yavg1.shape
+   print 'Yavg2 ', yavg2.shape
+   print 'Yavg3 ', yavg3.shape
+   
+   fig = figure()
+   ax = fig.add_subplot(1,1,1)
+   cax = ax.plot(range(min_L,max_L+1), yavg1[:,0], 'r')
+   cax = ax.plot(range(min_L,max_L+1), yavg1[:,1], 'g')
+   cax = ax.plot(range(min_L,max_L+1), yavg1[:,2], 'b')
+   cax = ax.plot(range(min_L,max_L+1), yavg1_sum, 'k')
+   
+   cax = ax.plot([0,0.001],[0,0], 'k')
[…]

### 2018-02-24 HRUIonGPU/gfx/make_plots.py

@@ -1,0 +1,22 @@
+from all_plots import plotBenchmarkGPU as plot
+from all_plots import barBenchmarkGPU as bar
+from all_plots import barBeamspaceBenchmarkGPU as barBS
+from all_plots import barBeamspaceBenchmarkGPU2 as barBS2
+
+#plot('timer_values_64_80_400.txt', 64, 80, 400, 2, 32)
+#plot('timer_values_32_40_400.txt', 32, 40, 400, 2, 32)
+
+#bar('timer_values_64_80_400.txt', 64, 80, 400, 2, 32)
+#bar('timer_values_32_40_400.txt', 32, 40, 400, 2, 16)
+
+#barBS('timer_values_M=64_Nx=80_Ny=400_L=2-32_Nb=3_slidingBS.txt', 64, 3, 80, 400, 3, 32)
+#barBS('timer_values_M=96_Nx=120_Ny=400_L=3-48_Nb=3_slidingBS.txt', 96, 3, 120, 400, 3, 48)
+
+bar('timer_values_M=64_Nx=80_Ny=540_L=2-32.txt', 64, 80, 540, 2, 32)
+bar('timer_values_M=96_Nx=120_Ny=540_L=2-48.txt', 96, 120, 540, 2, 48, 2)
+
+#barBS('timer_values_M=64_Nx=80_Ny=540_L=16-48_Nb=3.txt', 64, 3, 80, 540, 16, 47, False)
+#barBS('timer_values_M=96_Nx=120_Ny=540_L=24-72_Nb=3.txt', 96, 3, 120, 540, 24, 71)
+
+#barBS2('timer_values_M=64_Nx=80_Ny=606_L=32_Nb=1-5.txt', 64, 32, 80, 606, 1, 5, False)
+#barBS2('timer_values_M=96_Nx=120_Ny=606_L=48_Nb=1-5.txt', 96, 48, 120, 606, 1, 5)

### 2018-03-05 HRUIonGPU/document.tex
Individual build and precompiled headers ok

--- 2013-09-06 HRUIonGPU/jrnl_article.tex
@@
-%% bare_jrnl.tex
-%% V1.3
-%% 2007/01/11
-%% by Michael Shell
-%% see http://www.michaelshell.org/
-%% for current contact information.
-%%
-%% This is a skeleton file demonstrating the use of IEEEtran.cls
-%% (requires IEEEtran.cls version 1.7 or later) with an IEEE journal paper.
-%%
-%% Support sites:
-%% http://www.michaelshell.org/tex/ieeetran/
-%% http://www.ctan.org/tex-archive/macros/latex/contrib/IEEEtran/
-%% and
-%% http://www.ieee.org/
-
-% *** Authors should verify (and, if needed, correct) their LaTeX system  ***
-% *** with the testflow diagnostic prior to trusting their LaTeX platform ***
-% *** with production work. IEEE's font choices can trigger bugs that do  ***
-% *** not appear when using other class files.                            ***
-% The testflow support page is at:
-% http://www.michaelshell.org/tex/testflow/
-
-%%*************************************************************************
-%% Legal Notice:
-%% This code is offered as-is without any warranty either expressed or
-%% implied; without even the implied warranty of MERCHANTABILITY or
-%% FITNESS FOR A PARTICULAR PURPOSE! 
-%% User assumes all risk.
-%% In no event shall IEEE or any contributor to this code be liable for
-%% any damages or losses, including, but not limited to, incidental,
-%% consequential, or any other damages, resulting from the use or misuse
-%% of any information contained here.
-%%
-%% All comments are the opinions of their respective authors and are not
-%% necessarily endorsed by the IEEE.
-%%
-%% This work is distributed under the LaTeX Project Public License (LPPL)
-%% ( http://www.latex-project.org/ ) version 1.3, and may be freely used,
-%% distributed and modified. A copy of the LPPL, version 1.3, is included
-%% in the base LaTeX documentation of all distributions of LaTeX released
-%% 2003/12/01 or later.
-%% Retain all contribution notices and credits.
-%% ** Modified files should be clearly indicated as such, including  **
-%% ** renaming them and changing author support contact information. **
-%%
-%% File list of work: IEEEtran.cls, IEEEtran_HOWTO.pdf, bare_adv.tex,
-%%                    bare_conf.tex, bare_jrnl.tex, bare_jrnl_compsoc.tex
-%%*************************************************************************
-
-% Note that the a4paper option is mainly intended so that authors in
-% countries using A4 can easily print to A4 and see how their papers will
-% look in print - the typesetting of the document will not typically be
-% affected with changes in paper size (but the bottom and side margins will).
-% Use the testflow package mentioned above to verify correct handling of
-% both paper sizes by the user's LaTeX system.
-%
-% Also note that the "draftcls" or "draftclsnofoot", not "draft", option
-% should be used if it is desired that the figures are to be displayed in
-% draft mode.
-%
-
-\documentclass[journal]{IEEEtran}
-%\documentclass[11pt, draftcls, onecolumn]{IEEEtran}
-
-% Single column review mode:
-%\documentclass[12pt,journal,onecolumn]{IEEEtran}
-%\newcommand\CLASSINPUTbaselinestretch{2}
-%\RequirePackage{calc}
-%\RequirePackage{fp} 
-% Left margin = 1 inch + hoffset + oddsidemargin (or evensidemargin)
-% Adding the 2cm gutter width to the odd/even side margins:
-%\setlength\hoffset{0pt}
-%\setlength\oddsidemargin{4cm}       % 4cm margins on the left side
-%\addtolength\oddsidemargin{-1in}    % Subtract the initial 1 inch
-%\setlength\textwidth{21cm-8cm}    % A4 width (21cm) minus margins on either side
-
-%
-% If IEEEtran.cls has not been installed into the LaTeX system files,
-% manually specify the path to it like:
-% \documentclass[journal]{../sty/IEEEtran}
-
-% Some very useful LaTeX packages include:
-% (uncomment the ones you want to load)
-
-% *** MISC UTILITY PACKAGES ***
-%
-%\usepackage{ifpdf}
-% Heiko Oberdiek's ifpdf.sty is very useful if you need conditional
-% compilation based on whether the output is pdf or dvi.
-% usage:
-% \ifpdf
-%   % pdf code
-% \else
-%   % dvi code
-% \fi
-% The latest version of ifpdf.sty can be obtained from:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/oberdiek/
-% Also, note that IEEEtran.cls V1.7 and later provides a builtin
-% \ifCLASSINFOpdf conditional that works the same way.
-% When switching from latex to pdflatex and vice-versa, the compiler may
-% have to be run twice to clear warning/error messages.
-
-% *** CITATION PACKAGES ***
-%
-\usepackage{cite}
-% cite.sty was written by Donald Arseneau
-% V1.6 and later of IEEEtran pre-defines the format of the cite.sty package
-% \cite{} output to follow that of IEEE. Loading the cite package will
-% result in citation numbers being automatically sorted and properly
-% "compressed/ranged". e.g., [1], [9], [2], [7], [5], [6] without using
-% cite.sty will become [1], [2], [5]--[7], [9] using cite.sty. cite.sty's
-% \cite will automatically add leading space, if needed. Use cite.sty's
-% noadjust option (cite.sty V3.8 and later) if you want to turn this off.
-% cite.sty is already installed on most LaTeX systems. Be sure and use
-% version 4.0 (2003-05-27) and later if using hyperref.sty. cite.sty does
-% not currently provide for hyperlinked citations.
-% The latest version can be obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/cite/
-% The documentation is contained in the cite.sty file itself.
-
-% *** GRAPHICS RELATED PACKAGES ***
-%
-\ifCLASSINFOpdf
-  % \usepackage[pdftex]{graphicx}
-  % declare the path(s) where your graphic files are
-  % \graphicspath{{../pdf/}{../jpeg/}}
-  % and their extensions so you won't have to specify these with
-  % every instance of \includegraphics
-  % \DeclareGraphicsExtensions{.pdf,.jpeg,.png}
-\else
-  % or other class option (dvipsone, dvipdf, if not using dvips). graphicx
-  % will default to the driver specified in the system graphics.cfg if no
-  % driver is specified.
-  % \usepackage[dvips]{graphicx}
-  % declare the path(s) where your graphic files are
-  % \graphicspath{{../eps/}}
-  % and their extensions so you won't have to specify these with
-  % every instance of \includegraphics
-  % \DeclareGraphicsExtensions{.eps}
-\fi
-% graphicx was written by David Carlisle and Sebastian Rahtz. It is
-% required if you want graphics, photos, etc. graphicx.sty is already
-% installed on most LaTeX systems. The latest version and documentation can
-% be obtained at: 
-% http://www.ctan.org/tex-archive/macros/latex/required/graphics/
-% Another good source of documentation is "Using Imported Graphics in
-% LaTeX2e" by Keith Reckdahl which can be found as epslatex.ps or
-% epslatex.pdf at: http://www.ctan.org/tex-archive/info/
-%
-% latex, and pdflatex in dvi mode, support graphics in encapsulated
-% postscript (.eps) format. pdflatex in pdf mode supports graphics
-% in .pdf, .jpeg, .png and .mps (metapost) formats. Users should ensure
-% that all non-photo figures use a vector format (.eps, .pdf, .mps) and
-% not a bitmapped formats (.jpeg, .png). IEEE frowns on bitmapped formats
-% which can result in "jaggedy"/blurry rendering of lines and letters as
-% well as large increases in file sizes.
-%
-% You can find documentation about the pdfTeX application at:
-% http://www.tug.org/applications/pdftex
-
-% *** MATH PACKAGES ***
-%
-%\usepackage[cmex10]{amsmath}
-% A popular package from the American Mathematical Society that provides
-% many useful and powerful commands for dealing with mathematics. If using
-% it, be sure to load this package with the cmex10 option to ensure that
-% only type 1 fonts will utilized at all point sizes. Without this option,
-% it is possible that some math symbols, particularly those within
-% footnotes, will be rendered in bitmap form which will result in a
-% document that can not be IEEE Xplore compliant!
-%
-% Also, note that the amsmath package sets \interdisplaylinepenalty to 10000
-% thus preventing page breaks from occurring within multiline equations. Use:
-%\interdisplaylinepenalty=2500
-% after loading amsmath to restore such page breaks as IEEEtran.cls normally
-% does. amsmath.sty is already installed on most LaTeX systems. The latest
-% version and documentation can be obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/required/amslatex/math/
-
-% *** SPECIALIZED LIST PACKAGES ***
-%
-%\usepackage{algorithmic}
-% algorithmic.sty was written by Peter Williams and Rogerio Brito.
-% This package provides an algorithmic environment fo describing algorithms.
-% You can use the algorithmic environment in-text or within a figure
-% environment to provide for a floating algorithm. Do NOT use the algorithm
-% floating environment provided by algorithm.sty (by the same authors) or
-% algorithm2e.sty (by Christophe Fiorio) as IEEE does not use dedicated
-% algorithm float types and packages that provide these will not provide
-% correct IEEE style captions. The latest version and documentation of
-% algorithmic.sty can be obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/algorithms/
-% There is also a support site at:
-% http://algorithms.berlios.de/index.html
-% Also of interest may be the (relatively newer and more customizable)
-% algorithmicx.sty package by Szasz Janos:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/algorithmicx/
-
-% *** ALIGNMENT PACKAGES ***
-%
-%\usepackage{array}
-% Frank Mittelbach's and David Carlisle's array.sty patches and improves
-% the standard LaTeX2e array and tabular environments to provide better
-% appearance and additional user controls. As the default LaTeX2e table
-% generation code is lacking to the point of almost being broken with
-% respect to the quality of the end results, all users are strongly
-% advised to use an enhanced (at the very least that provided by array.sty)
-% set of table tools. array.sty is already installed on most systems. The
-% latest version and documentation can be obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/required/tools/
-
-%\usepackage{mdwmath}
-%\usepackage{mdwtab}
-% Also highly recommended is Mark Wooding's extremely powerful MDW tools,
-% especially mdwmath.sty and mdwtab.sty which are used to format equations
-% and tables, respectively. The MDWtools set is already installed on most
-% LaTeX systems. The lastest version and documentation is available at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/mdwtools/
-
-% IEEEtran contains the IEEEeqnarray family of commands that can be used to
-% generate multiline equations as well as matrices, tables, etc., of high
-% quality.
-
-%\usepackage{eqparbox}
-% Also of notable interest is Scott Pakin's eqparbox package for creating
-% (automatically sized) equal width boxes - aka "natural width parboxes".
-% Available at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/eqparbox/
-
-% *** SUBFIGURE PACKAGES ***
-%\usepackage[tight,footnotesize]{subfigure}
-% subfigure.sty was written by Steven Douglas Cochran. This package makes it
-% easy to put subfigures in your figures. e.g., "Figure 1a and 1b". For IEEE
-% work, it is a good idea to load it with the tight package option to reduce
-% the amount of white space around the subfigures. subfigure.sty is already
-% installed on most LaTeX systems. The latest version and documentation can
-% be obtained at:
-% http://www.ctan.org/tex-archive/obsolete/macros/latex/contrib/subfigure/
-% subfigure.sty has been superceeded by subfig.sty.
-
-%\usepackage[caption=false]{caption}
-%\usepackage[font=footnotesize]{subfig}
-% subfig.sty, also written by Steven Douglas Cochran, is the modern
-% replacement for subfigure.sty. However, subfig.sty requires and
-% automatically loads Axel Sommerfeldt's caption.sty which will override
-% IEEEtran.cls handling of captions and this will result in nonIEEE style
-% figure/table captions. To prevent this problem, be sure and preload
-% caption.sty with its "caption=false" package option. This is will preserve
-% IEEEtran.cls handing of captions. Version 1.3 (2005/06/28) and later 
-% (recommended due to many improvements over 1.2) of subfig.sty supports
-% the caption=false option directly:
-\usepackage[caption=false,font=footnotesize]{subfig}
-%
-% The latest version and documentation can be obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/subfig/
-% The latest version and documentation of caption.sty can be obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/caption/
-
-% *** FLOAT PACKAGES ***
-%
-%\usepackage{fixltx2e}
-% fixltx2e, the successor to the earlier fix2col.sty, was written by
-% Frank Mittelbach and David Carlisle. This package corrects a few problems
-% in the LaTeX2e kernel, the most notable of which is that in current
-% LaTeX2e releases, the ordering of single and double column floats is not
-% guaranteed to be preserved. Thus, an unpatched LaTeX2e can allow a
-% single column figure to be placed prior to an earlier double column
-% figure. The latest version and documentation can be found at:
-% http://www.ctan.org/tex-archive/macros/latex/base/
-
-%\usepackage{stfloats}
-% stfloats.sty was written by Sigitas Tolusis. This package gives LaTeX2e
-% the ability to do double column floats at the bottom of the page as well
-% as the top. (e.g., "\begin{figure*}[!b]" is not normally possible in
-% LaTeX2e). It also provides a command:
-%\fnbelowfloat
-% to enable the placement of footnotes below bottom floats (the standard
-% LaTeX2e kernel puts them above bottom floats). This is an invasive package
-% which rewrites many portions of the LaTeX2e float routines. It may not work
-% with other packages that modify the LaTeX2e float routines. The latest
-% version and documentation can be obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/sttools/
-% Documentation is contained in the stfloats.sty comments as well as in the
-% presfull.pdf file. Do not use the stfloats baselinefloat ability as IEEE
-% does not allow \baselineskip to stretch. Authors submitting work to the
-% IEEE should note that IEEE rarely uses double column equations and
-% that authors should try to avoid such use. Do not be tempted to use the
-% cuted.sty or midfloat.sty packages (also by Sigitas Tolusis) as IEEE does
-% not format its papers in such ways.
-
-%\ifCLASSOPTIONcaptionsoff
-%  \usepackage[nomarkers]{endfloat}
-% \let\MYoriglatexcaption\caption
-% \renewcommand{\caption}[2][\relax]{\MYoriglatexcaption[#2]{#2}}
-%\fi
-% endfloat.sty was written by James Darrell McCauley and Jeff Goldberg.
-% This package may be useful when used in conjunction with IEEEtran.cls'
-% captionsoff option. Some IEEE journals/societies require that submissions
-% have lists of figures/tables at the end of the paper and that
-% figures/tables without any captions are placed on a page by themselves at
-% the end of the document. If needed, the draftcls IEEEtran class option or
-% \CLASSINPUTbaselinestretch interface can be used to increase the line
-% spacing as well. Be sure and use the nomarkers option of endfloat to
-% prevent endfloat from "marking" where the figures would have been placed
-% in the text. The two hack lines of code above are a slight modification of
-% that suggested by in the endfloat docs (section 8.3.1) to ensure that
-% the full captions always appear in the list of figures/tables - even if
-% the user used the short optional argument of \caption[]{}.
-% IEEE papers do not typically make use of \caption[]'s optional argument,
-% so this should not be an issue. A similar trick can be used to disable
-% captions of packages such as subfig.sty that lack options to turn off
-% the subcaptions:
-% For subfig.sty:
-%\let\MYorigsubfloat\subfloat
-%\renewcommand{\subfloat}[2][\relax]{\MYorigsubfloat[]{#2}}
-% For subfigure.sty:
-% \let\MYorigsubfigure\subfigure
-% \renewcommand{\subfigure}[2][\relax]{\MYorigsubfigure[]{#2}}
-% However, the above trick will not work if both optional arguments of
-% the \subfloat/subfig command are used. Furthermore, there needs to be a
-% description of each subfigure *somewhere* and endfloat does not add
-% subfigure captions to its list of figures. Thus, the best approach is to
-% avoid the use of subfigure captions (many IEEE journals avoid them anyway)
-% and instead reference/explain all the subfigures within the main caption.
-% The latest version of endfloat.sty and its documentation can obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/endfloat/
-%
-% The IEEEtran \ifCLASSOPTIONcaptionsoff conditional can also be used
-% later in the document, say, to conditionally put the References on a 
-% page by themselves.
-
-% *** PDF, URL AND HYPERLINK PACKAGES ***
-%
-\usepackage{url}
-% url.sty was written by Donald Arseneau. It provides better support for
-% handling and breaking URLs. url.sty is already installed on most LaTeX
-% systems. The latest version can be obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/misc/
-% Read the url.sty source comments for usage information. Basically,
-% \url{my_url_here}.
-
-% *** Do not adjust lengths that control margins, column widths, etc. ***
-% *** Do not use packages that alter fonts (such as pslatex).         ***
-% There should be no need to do such things with IEEEtran.cls V1.6 and later.
-% (Unless specifically asked to do so by the journal or conference you plan
-% to submit to, of course. )
-
-% correct bad hyphenation here
-\hyphenation{op-tical net-works semi-conduc-tor}
-
-%%% My packages
-\usepackage{amsmath,epsfig}
-\usepackage{amsfonts}
-\usepackage{color}
-
-%%% My commands
-\newcommand{\mat}[1]{\mathbf{#1}}
-\renewcommand{\vec}[1]{\mathbf{#1}}
-\newcommand{\degree}{\ensuremath{^\circ}}
-\newcommand{\R}{$\mat{\hat{R}}$ }
-
-%\newcommand{\Red}[1]{{\color{red}#1}}
-\newcommand\multimedia[1]{\textbf{{\color{red}#1}}}
-
-\begin{document}
 %
@@
 % denne må utelates i IUS-artikkelen
-In recent work we have applied a GPU-based Capon beamformer for both active sonar imaging \cite{Buskenes, Buskenes2013} and cardiac ultrasound imaging \cite{asen2012}.
+In recent work we have applied a GPU-based Capon beamformer for both active sonar imaging \cite{Buskenes, Buskenes2013} and cardiac ultrasound imaging \cite{Asen2012}.
 Sonar has a less strict real-time requirement than medical ultrasound imaging, which means that most array configurations can be handled in real time using the GPU.
-In \cite{asen2012} and in this paper we show that Capon beamforming applied in medical ultrasound imaging, and in particular cardiac ultrasound imaging, demands frame rates that not even a cluster of high-end GPUs can provide.
+In \cite{Asen2012} and in this paper we show that Capon beamforming applied in medical ultrasound imaging, and in particular cardiac ultrasound imaging, demands frame rates that not even a cluster of high-end GPUs can provide.
 Since the execution time of Capon beamforming is cubic with respect to the number of array elements, we need methods that reduce the computational complexity for large arrays.
 Various approaches to low-complexity Capon beamforming for medical ultrasound imaging have been proposed in the literature \cite{Synnevag2011, Asl2012, Jensen2012, Kim}.
-In this paper we extend the work done in \cite{asen2012} and \cite{Buskenes2013} by utilizing a beamspace transformation.
+In this paper we extend the work done in \cite{Asen2012} and \cite{Buskenes2013} by utilizing a beamspace transformation.
 The beamspace transform has been proposed used in ultrasound imaging by Nilsen and Hafizovic \cite{Nilsen2009} in order to maintain high frame rates as the number of elements increases while maintaining almost the same lateral resolution and sidelobe reduction capabilities of element-space Capon when applied in narrow-beam systems.
@@
 \begin{figure}
-\centerline{\includegraphics[width=3in]{gfx/beamforming_mv.eps}}
+\centerline{\includegraphics[width=.7\linewidth]{gfx/beamforming_mv.eps}}
 \caption{Capon beamforming.
@@
 The sample covariance matrix is estimated as
+%
 \begin{align}
@@
 \end{align}
-where  $N_K = 2K + 1$ is the number of temporal samples included, $\vec{x}_l = [x_l[n], \dotso, x_{l+L-1}[n]]$ is the $l\text{th}$ subarray of length $L$, and $N_L = M-L+1$ is the number of subarrays.
+%
+where  $N_K = 2K + 1$ is the number of temporal samples included, $N_L = M-L+1$ is the number of subarrays, and $\vec{x}_l = [x_l[n], \dotso, x_{l+L-1}[n]]$ is the $l\text{th}$ subarray of length $L$.
 Note that the covariance matrix dimension is $L \times L$.
 The covariance matrix is loaded with a diagonal factor $\epsilon$ to ensure numerical stability and increased robustness:
+%
 \begin{align}\label{eq:diag}
@@
 \end{align}
+%
 The operator $f(\cdot)$ represents the process of estimating $\mat{\hat{R}}$ from the input $\vec{x}$.
 A factor proportional to the output power is often used to reduce the need for parameter adjustments, and the trace of $\mat{\breve{R}}$, 
+%
 \begin{align}\label{eq:diag_adapt}
@@
 \end{align}
+%
 has been applied in much of the recent literature on Capon beamforming for medical ultrasound imaging \cite{Synnevag2007, Nilsen2009, Wang2009, Mehdizadeh2012}.
@@
 Next, a linear system of equations has to be solved,
+%
 \begin{align}\label{eq:b}
@@
 \end{align}
+%
 where the steering vector $\vec{a} = \vec{1}_L = [1_0, 1_1, ..., 1_{L-1}]^T$ since $\vec{x}$ is pre-delayed.
@@
 Finally the Capon weight vector is formed from (\ref{eq:b}) as
+%
 \begin{align}\label{eq:w}
@@
 \end{align}
+%
 and the beamformer output is calculated by combining all $N_L$ subarrays weighted with the same set of adaptive weights. This is formally stated as
+%
 \begin{align}
@@
 \section{Parallel ES-Capon}\label{sec:meth}
-In this section we will give a brief summary of our parallel implementation of the ES-Capon beamformer as presented in \cite{asen2012} and \cite{Buskenes}.
+
+In this section we will give a brief summary of our parallel implementation of the ES-Capon beamformer as presented in \cite{Asen2012} and \cite{Buskenes}.
 This implementation was our starting point when the BS-Capon beamformer was implemented.
@@
 
-In \cite{asen2012, Buskenes} we demonstrated how the cubic complexity with $L$ %and linear complexity with $N_K$ 
+In \cite{Asen2012, Buskenes} we demonstrated how the cubic complexity with $L$ %and linear complexity with $N_K$ 
 can be reduced by realizing that calculations in (\ref{eq:R}) overlap across subarrays.
@@
 
-In \cite{asen2012} we discussed how it is not given that solving on the GPU is faster than solving on the CPU in general.
+In \cite{Asen2012} we discussed how it is not given that solving on the GPU is faster than solving on the CPU in general.
 However, there are two arguments for solving on the GPU.
@@
 Since we selected to use $L$ threads per output, the data is reduced to length $L$ by letting thread $r$ calculate
+%
 \begin{align}\label{eq:subsum}
@@
 \end{align}
+%
 Each thread now computes $z_r = w_r^*x_r$, and finally one thread computes and outputs $z = \frac{1}{N_L}\sum_{r=0}^{L-1} z_r$ to global memory.
@@
 \section{Parallel BS-Capon}\label{sec:bs}
+
 In this section we will derive a parallel implementation of the BS-Capon beamformer, and comment on how it deviates from the implementation in Section \ref{sec:meth}.
@@
 The other option is to transform the element space covariance matrix to beamspace:
+%
 \begin{align}
@@
 \end{align}
+%
 The advantage of this approach is that minor changes have to be made to the pipeline described in Section \ref{sec:meth}, but it will only speed up the solving step if the solution $\vec{b}_{BS}$ is transformed back to element space.
@@
 After solving $\vec{b}_\text{BS} = \sqrt{L}\mat{\hat{R}}_\text{BS}^{-1}\vec{e}_1$, the beamspace weight vector is calculated as
+%
 \begin{align}
@@
 \end{align}  
+%
 using one thread per beam, and applied to the beamspace data summed across subarrays,
+%
 \begin{align}
@@
 \end{align}
-%Several pixels are calculated per thread block for small values of $N_b$ in order to achieve high GPU occupancy. The beamspace data summed across subarrays are calculated as part of the SBS kernel.
-
+%Several pixels are calculated per thread block for small values of $N_b$ in order to achieve high GPU occupancy.
+The beamspace data summed across subarrays are calculated as part of the SBS kernel.
+%
 %How do we calculate the number of possible subarrays $N_L$ with overlap $L-s$ in a data vector of length $M$.
@@
 %\end{align}
-
+%
 As presented in (\ref{eq:diag}), adaptive diagonal loading is performed by loading the sample covariance matrix with the average energy per element, that is $\text{trace}\{\mat{\breve{R}}\}/L$.
@@
 
+\ifPhdDoc
+  \begin{figure*}[!t]
+  \centerline{\subfloat[]{\includegraphics[width=3.8in]{gfx/benchmark_bar_bs_M=64_Nx=80_Ny=540_Nb=3.png} \label{fig:benchBS64}}
+%   \hfil
+  \subfloat[]{\includegraphics[width=3.8in]{gfx/benchmark_bar_bs_M=96_Nx=120_Ny=540_Nb=3.png}%
+  \label{fig:benchBS96}}}
+\else
   \begin{figure*}[!t]
@@
   \label{fig:benchBS96}}}
+\fi%
 \caption{Benchmarks of GPU beamspace Capon beamforming for a cardiac image covering a $70 \degree$ and $15.4$ cm deep sector.
@@
 \end{figure*}
+%
 In this section we present benchmarks of the implementations presented in Sections \ref{sec:meth} and \ref{sec:bs} performed on a Nvidia Quadro 6000 graphics card (14 SMs, 1 TFlops theoretical single precision performance, 6 GB global memory, and 144 GB/s memory bandwidth).
@@
 The angular sampling is given by the Rayleigh criterion divided by two \cite{Hergum2007},
+%
 \begin{equation}\label{eq:rayleigh}
@@
 \end{equation}
+%
 where $\lambda$ is the pulse wavelength and $p$ is the element pitch. 
@@
 
+\ifPhdDoc
+   \printbibliography[title=References,heading=subbibliography]
+%    \bibliographysty
+%    \bibliography{library.bib}
+%    \print
+\else
+
   % can use a bibliography generated by BibTeX as a .bbl file
@@
 
-% that's all folks
-\end{document}
+\fi

### 2018-03-05 ThesisJo/2_publications.tex
Individual build and precompiled headers ok

%&2_publications
% !TeX root = 2_publications

\input{header}
\input{subheader}
\endofdump

\input{glossary}
\makeglossaries

\begin{document}
%\documentclass[float=false, crop=false]{standalone}
%\usepackage[subpreambles=true]{standalone}
%\input{header}
%\input{subheader}

\chapter{Publications}

The upcoming papers will investigate various ways in which acoustical imaging systems can be improved by making them smarter in an operational sense. This involves granting them the ability to adapt their behavior to the information at hand, and the processing power needed to handle the information fast enough.

%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
\section{Paper I}\label{sec:paperI} %
%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

\paper{I} describes the use and implementation of the \gls{MVDR} beamformer in active sonar imaging in its most general and exact form.

Its widespread use and generic implementation makes it the reference. thus will be considered the reference method. , and this will be referred to as a fully adaptive technique. Due to its ability to analytically  to apply the minimum variance criteria in an exact manner, Due to its popularity and We view this as the reference method for We refer to it as fully adaptive since it computes analytically evaluates   yields  This is a fully adaptive method   This method is  on a \gls{GPGPU} for active sonar imaging. This is to prove that using a \gls{MVDR}

Paper II takes the \gls{MVDR} method and applies it to clinical Ultrasound Imaging

We point out important hardware limitations for these devices, and assess the design in terms of how efficiently it is able to use the GPU's resources. On a quad-core Intel Xeon system with a high-end Nvidia GPU, our GPU implementation renders more than a million pixels per second (1\;MP/s). Compared to our initial central processing unit (CPU) implementation the optimizations described herein led to a speedup of more than two orders of magnitude, or an expected five to ten times improvement had the CPU received similar optimization effort. This throu

%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
\section{Paper I}\label{sec:paperI} %
%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
\textbf{An Optimized GPU implementation of the MVDR Beamformer for Active Sonar Imaging}\\
J.I. Buskenes, J.P. Åsen, C.-I.C. Nilsen and A. Austeng\\
\textit{IEEE Journal of Oceanic Engineering}\\
Volume 40, Issue 2, April 2015

\begin{itemize}
\item Minimum variance
\item Full search space, continuous
\item Element-space
\item Sonar
\item Computationally complex
\item Common: Minimal temporal statistics (near single snapshot)
\end{itemize}

This is a popular method that analytically estimates the array's weight set that best comply with the minimum variance criterion. Its analytic nature allows it to chose from an infinite number of possible weight sets, but with a solution space limited by the constraint of unit gain in the look direction.

The first paper is concerned with applying the Minimum Variance Distortionless Response (MVDR) beamformer to active sonar imaging, with emphasis on performing arithmetic reductions and solving the practical its high computational complexity. It shows that  with emphasis on solving and with the process of implementing this method on a \gls{GPU}. The goal is to improve sonar image quality without suffering the consequences of the MVDR's inherently high computational complexity.

Combined, these steps led to a speedup of an order of magnitude or more compared to its straight-forward \gls{CPU} implementation.

In functionality this method acts as our reference method, but its practicality is limited by a high computational complexity. The paper suggests means for reducing the arithmetic complexity and provide extensive details for fitting the implementation on a \gls{GPU} for a major speedup. It also relates the performance to theoretical maximums and investigates remaining bottlenecks.

%Applying the MVDR-beamformer may not seem like fully adaptive technique 
%The Minimum Variance Distortionless Response (MVDR) beamformer is a technique predominant is a fully adaptive method that 
%
%
%When performing image processing data from phased arrays an inherent question arise 
%In the pursuit of creating images with more detail and less noise from phased-array data, a fully adaptive beamformer such as the the Minimum Variance Distortionless Response (MVDR) can be used in the receive

% The minimum variance distortionless response (MVDR) beamformer has recently been proposed as an attractive alternative to conventional beamformers in active sonar imaging. Unfortunately, it is very computationally complex because a spatial covariance matrix must be estimated and inverted for each image pixel. This may discourage its use unnecessarily in sonar systems which are continuously being pushed to ever higher imaging ranges and resolutions.
% 
% In this study we show that for active sonar systems up to 32 channels, the computation time can be significantly reduced by performing arithmetic optimizations, and by implementing the MVDR beamformer on a graphics processing unit (GPU). We point out important hardware limitations for these devices, and assess the design in terms of how efficiently it is able to use the GPU's resources. On a quad-core Intel Xeon system with a high-end Nvidia GPU, our GPU implementation renders more than a million pixels per second (1\;MP/s). Compared to our initial central processing unit (CPU) implementation the optimizations described herein led to a speedup of more than two orders of magnitude, or an expected five to ten times improvement had the CPU received similar optimization effort. This throughput enables real-time processing of sonar data, and makes the MVDR a viable alternative to conventional methods in practical systems.

%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
\section{Paper II}\label{sec:paperII} %
%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
\textbf{Implementing Capon Beamformer on a GPU for\\ Real-Time Cardiac Ultrasound Imaging}\\
J.P. Åsen, J.I. Buskenes, C.-I.C. Nilsen, A. Austeng and S. Holm\\
\textit{IEEE Transactions on Ultrasonics, Ferroelectrics, and Frequency Control}\\
Volume 61, Issue 1, January 2015

\begin{itemize}
\item Minimum variance
\item Full search space, continuous
\item Beam-space
\item Ultrasound
\end{itemize}

%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
\section{Paper III}\label{sec:paperIII} %
%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
\textbf{Low Complexity Adaptive Beamforming}~\cite{Buskenes2015}\\
J.I. Buskenes, R.E. Hansen and A. Austeng\\
\textit{IEEE Journal of Oceanic Engineering}\\
Volume 42, Issue 1, January 2017

\begin{itemize}
\item Minimum variance
\item Smaller search space, discrete
\item Element-space
\item sonar
\end{itemize}

%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
\section{Paper IV}\label{sec:paperIV} %
%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

\begin{itemize}
\item Template-generation
\item Autonomy
\item Sonar
\end{itemize}

\section{Onlook}

R dim > 10000
- Inverting such matrices better to do with gradient decent

\end{document}

### 2018-03-05 ThesisJo/3_discussion.tex
Individual build and precompiled headers ok

%&header
% !TeX root = 3_discussion
\documentclass[thesis.tex]{subfiles}
\input{subheader}
\begin{document}

\chapter{Discussion and Outlook}

\begin{itemize}
\item Improving imaging systems by making them smarter.
\item Big data - more processing power.
\item Processing speed
\begin{itemize}
	\item Choose the best tool (algorithm) for the job
	\item More complex not always better, but slower
	\item Extracting the essential information is key 
	\item Exploit hardware resources available
\end{itemize}
\item Faster IS smarter
 \begin{itemize}
 	\item Short feedback loops allow real-time adaptivity
 	\item More data contains more information
 	\item Faster learning
 \end{itemize}
\item Smarter IS better
 \begin{itemize}
	\item Better image quality
	\item Improved autonomy
\end{itemize}
\end{itemize}

\begin{itemize}
\item Improving imaging systems by making them smarter.
\item Big data - more processing power.
\item IV - Simulator
\begin{itemize}
	\item Incredibly fast template matching possible.
	\item Possibly a powerful tool to pass prior knowledge to deep learning
	\item Extracting the essential information is key 
	\item Exploit hardware resources available
\end{itemize}
\item Faster IS smarter
 \begin{itemize}
 	\item Short feedback loops allow real-time adaptivity
 	\item More data contains more information
 	\item Faster learning
 \end{itemize}
\item Smarter IS better
 \begin{itemize}
	\item Better image quality
	\item Improved autonomy
\end{itemize}
\end{itemize}

\end{document}

### 2018-03-05 ThesisJo/literature_review.tex
Individual build and precompiled headers ok

@@
 
+\section{Deep learning}
+
+\begin{itemize}
+\item AlphaGo - https://storage.googleapis.com/deepmind-media/alphago/AlphaGoNaturePaper.pdf
+\item AlphaZero - Chess
+\item Google’s Tacotron 2 text-to-speech system (autoregressive), based on WaveNet
+\end{itemize}
+
+Search spaces, learning speed,
+
+Brain a good tutor - but AI should use operational definition and not a cognitive one. (Alan Turing / Tom M. Mitchell)
+
+Autoregressive methods
+Recurrent architectures are slower, and focus is on finding alternatives
+
+ML by output classification
+- Classification - discrete number of outputs (classes). Usually supervised.
+- Regression - continuous outputs. Supervised.
+- Clustering - unsupervised.
+- Density estimation - finding the distribution of inputs in some space.
+- Dimensionality reduction - simplifies inputs by mapping them to a lower-dimensional space. Topic modeling is related, finding human language similarities
+- Learning to learn - learns its own inductive bias.
+- Developmental learning / robot learning - generates its own sequences of learning situations to learn new skills through autonomous self-exploration and social interaction.
+
+Keywords:
+- Search space
+- (deep) Reinforcement learning
+  - Standard methods
+    * Example: Deep Q-learning
+  - Evolution strategies
+    * Don't rely on gradients
+    * Good when data isn't iid, error signals are sparse, and when need for exploration
+    * Scale well with parallel hardware / machine clusters
+- Supervised learning
+
+Thinking fast and slow - Kahneman
+
+Decision-based methods
+- Supervised
+  - Statistical methods
+    - Parametric
+      - Maximum likelihood
+      - Bayesian estimation
+    - Non-parametric
+      - Window-method
+      - Nearest-neighbor
+      - Series expansion
+  - Optimization methods
+    - Linear discrimination functions
+    - Generalized l.d.f.
+- Non-supervised
+  - Parametric methods
+    - Maximum likelihood
+    - Bayesian estimation
+  - Cluster analysis
+    - Optimization methods
+    - Hierarchical methods
+
+Structural methods
+  - Syntactic methods
+    - Syntax analysis
+    - String matching
+
 %~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

### 2018-03-06 Simulator/article.tex
All compiling, but lots of work to be done

--- 2018-01-04 Simulator/article.tex
@@
 
-\let\oldincludegraphics\includegraphics
-\renewcommand\includegraphics[2][]{%
-  \immediate\write18{./bin/laFigure #2 #1}%
-  \input{result}}%
+\input{macros}
 
-% \DeclareMathOperator*{\argmin}{\text{arg}\;\text{min}}
-
-\newcommand\Fig[1]{Fig.~\ref{#1}}
-\newcommand\eq[1]{(\ref{#1})}
-
-\newcommand\Grey[1]{{\color{Grey}#1}}
-\newcommand\Red[1]{{\color{Red}#1}}
-\newcommand\Blue[1]{{\color{Blue}#1}}
-\newcommand\DarkBlue[1]{{\color{DarkBlue}#1}}
-\newcommand\LightBlue[1]{{\color{LightBlue}#1}}
-\newcommand\Brown[1]{{\color{Brown}#1}}
-\newcommand\Green[1]{{\color{Green}#1}}
-\newcommand\SeaGreen[1]{{\color{SeaGreen}#1}}
-\newcommand\Yellow[1]{{\color{yellow}#1}}
-\newcommand\Orange[1]{{\color{orange}#1}}
-
-\newcommand\nn{\nonumber\\}
-
-\newcommand\nmat[1]{\begin{matrix}#1\end{matrix}}
-\newcommand\bmat[1]{\begin{bmatrix}#1\end{bmatrix}}
-\newcommand\case[1]{\begin{cases}#1\end{cases}}
-\newcommand\textbox[2]{\footnotesize\text{\parbox{#1}{\centering\emph{#2}}}}
-
-\newcommand\rand{\text{rand}}
-\newcommand\randn{\text{randn}}
-\newcommand\rect{\text{rect}}
-\newcommand\sinc{\text{sinc}}
-\newcommand\tr{\text{tr}}
-\newcommand\adj{\text{adj}}
-
-% \newcommand\max{\text{max}}
-\newcommand\argmin[1]{\text{arg}\;\underset{#1}{\text{min}}}
-
-\newcommand\qqquad{\quad\qquad}
-\newcommand\qqqquad{\qquad\qquad}
-
-% \renewcommand\l[1]{\left#1}
-% \renewcommand\r[1]{\right#1}
-
-% {\text{\parbox{1.5cm}{\centering volume hyper- sphere}}}
-
-%Keyword colouring:
-\newcommand\kw[1]{#1}
-\newcommand\parm[1]{#1}%\color{Black}#1\color{Black}}
-
-\newcommand\of[1]{\scriptstyle(\parm{#1})\displaystyle}
-\newcommand\df[1]{\scriptstyle[\parm{#1}]\displaystyle}
-\newcommand\var[3]{#1_\text{#2}\of{#3}}
-
-\newcommand\diag{\text{diag}}
-
-% \raisebox{lift}[extend-above-baseline][extend-below-baseline]{text}
-\newcommand\mt[1]{\text{\emph{#1}}} %mt = mathtext
-\newcommand\mathnorm{\textstyle}
-\newcommand\mathbig[1]{\displaystyle#1\mathnorm}
-\newcommand\mathsmall[1]{\scriptstyle#1\mathnorm}
-\newcommand\mathtiny[1]{\scriptscriptstyle#1\mathnorm}
-\newcommand\sfrac[2]{\scriptstyle\raisebox{0.25pt}[0pt][0pt]{$\frac{#1}{#2}$}\mathnorm}
-\newcommand\nfrac[2]{\textstyle\frac{#1}{#2}\displaystyle}
-
-\newcommand\sumu[1]{\sum\limits^{#1}\;}
-\newcommand\suml[1]{\sum\limits_{#1}\;}
-\newcommand\sumb[2]{\sum\limits_{#1}^{#2}\;}
-
-\newcommand\produ[1]{\prod\limits^{#1}\;}
-\newcommand\prodl[1]{\prod\limits_{#1}\;}
-\newcommand\prodb[2]{\prod\limits_{#1}^{#2}\;}
-
-\newcommand\defeq{\overset{\underset{\mathrm{def}}{}}{=}}
-
-%Math macros:
-\newcommand\T{^{\scriptscriptstyle T}}
-\renewcommand\H{^{\scriptscriptstyle H}}
-
-\renewcommand\vec[1]{\boldsymbol{#1}}
-\newcommand\mat[1]{\boldsymbol{#1}}
-
-\newcommand\Om{O_\text{m}}
-\newcommand\Oa{O_\text{a}}
-\newcommand\Nl{N_\text{l}}
-\newcommand\Nk{N_\text{k}}
-\newcommand\1{\vec 1}
-\newcommand\I{\mat I}
-\renewcommand*\a{\vec a}
-\newcommand*\f{\vec f}
-\renewcommand*\i{\vec i}
-\renewcommand*\k{\vec k}
-\newcommand*\n{\vec n}
-\newcommand*\p{\vec p}
-\newcommand*\s{\vec s}
-\newcommand*\w{\vec w}
-\newcommand*\x{\vec x}
-\newcommand*\y{\vec y}
-
-\newcommand*\A{\mat A}
-\newcommand*\B{\mat B}
-% \newcommand*\C{\mat C}
-\newcommand*\E{\mat E}
-\renewcommand*\P{\mat P}
-\newcommand*\eP{\mat{\hat P}}
-\newcommand*\R{\mat R}
-\newcommand*\Ri{\R^{-1}}
-\newcommand*\eR{\mat{\hat R}}
-\newcommand*\eRi{\hat{\mat R}\;\!^{-1}}
-\newcommand*\Navg{N_\text{avg}}
-\newcommand*\W{\mat W}
-\newcommand*\X{\mat X}
-\newcommand*\Xd{\X_{\!\Delta}}
-\newcommand*\Y{\mat Y}
-
-\renewcommand\argmin{\text{argmin}}
-
-\renewcommand*\P{\mat P}
-\newcommand*\V{\mat V}
-\newcommand*\M{\mat M}
-
-\renewcommand*\L{\mat \Lambda}
-% \newcommand*\U{\mat U}
-% \renewcommand*\t{\mathtiny{^T}}
-% \newcommand*\h{\mathtiny{^H}}
-\renewcommand*\t{^T}
-\newcommand*\h{^H}
-
-\newcommand\minus{\scalebox{0.75}[1.0]{$-$}}
-
-% \usepackage{tikz}
-% \usetikzlibrary{shapes,snakes}
-\usepackage{amsmath,amssymb}
-% \usepackage{datatool}
-% \usepackage{glossaries}
-
-% \newenvironment{outline}
-% {\begin{itemize}}
-% {\end{itemize}}
+%\let\oldincludegraphics\includegraphics
+%\renewcommand\includegraphics[2][]{%
+%  \immediate\write18{./bin/laFigure #2 #1}%
+%  \input{result}}%
+%  
+%% \DeclareMathOperator*{\argmin}{\text{arg}\;\text{min}}
+%
+%\newcommand\Fig[1]{Fig.~\ref{#1}}
+%\newcommand\eq[1]{(\ref{#1})}
+%
+%\newcommand\Grey[1]{{\color{Grey}#1}}
+%\newcommand\Red[1]{{\color{Red}#1}}
+%\newcommand\Blue[1]{{\color{Blue}#1}}
+%\newcommand\DarkBlue[1]{{\color{DarkBlue}#1}}
+%\newcommand\LightBlue[1]{{\color{LightBlue}#1}}
+%\newcommand\Brown[1]{{\color{Brown}#1}}
+%\newcommand\Green[1]{{\color{Green}#1}}
+%\newcommand\SeaGreen[1]{{\color{SeaGreen}#1}}
+%\newcommand\Yellow[1]{{\color{yellow}#1}}
+%\newcommand\Orange[1]{{\color{orange}#1}}
+%
+%\newcommand\nn{\nonumber\\}
+%
+%\newcommand\nmat[1]{\begin{matrix}#1\end{matrix}}
+%\newcommand\bmat[1]{\begin{bmatrix}#1\end{bmatrix}}
+%\newcommand\case[1]{\begin{cases}#1\end{cases}}
+%\newcommand\textbox[2]{\footnotesize\text{\parbox{#1}{\centering\emph{#2}}}}
+%
+%\newcommand\rand{\text{rand}}
+%\newcommand\randn{\text{randn}}
+%\newcommand\rect{\text{rect}}
+%\newcommand\sinc{\text{sinc}}
+%\newcommand\tr{\text{tr}}
+%\newcommand\adj{\text{adj}}
+%
+%% \newcommand\max{\text{max}}
+%\newcommand\argmin[1]{\text{arg}\;\underset{#1}{\text{min}}}
+%
+%\newcommand\qqquad{\quad\qquad}
+%\newcommand\qqqquad{\qquad\qquad}
+%
+%% \renewcommand\l[1]{\left#1}
+%% \renewcommand\r[1]{\right#1}
+%
+%% {\text{\parbox{1.5cm}{\centering volume hyper- sphere}}}
+%
+%%Keyword colouring:
+%\newcommand\kw[1]{#1}
+%\newcommand\parm[1]{#1}%\color{Black}#1\color{Black}}
+%
+%\newcommand\of[1]{\scriptstyle(\parm{#1})\displaystyle}
+%\newcommand\df[1]{\scriptstyle[\parm{#1}]\displaystyle}
+%\newcommand\var[3]{#1_\text{#2}\of{#3}}
+%
+%\newcommand\diag{\text{diag}}
+%
+%% \raisebox{lift}[extend-above-baseline][extend-below-baseline]{text}
+%\newcommand\mt[1]{\text{\emph{#1}}} %mt = mathtext
+%\newcommand\mathnorm{\textstyle}
+%\newcommand\mathbig[1]{\displaystyle#1\mathnorm}
+%\newcommand\mathsmall[1]{\scriptstyle#1\mathnorm}
+%\newcommand\mathtiny[1]{\scriptscriptstyle#1\mathnorm}
+%\newcommand\sfrac[2]{\scriptstyle\raisebox{0.25pt}[0pt][0pt]{$\frac{#1}{#2}$}\mathnorm}
+%\newcommand\nfrac[2]{\textstyle\frac{#1}{#2}\displaystyle}
+%
+%\newcommand\sumu[1]{\sum\limits^{#1}\;}
+%\newcommand\suml[1]{\sum\limits_{#1}\;}
+%\newcommand\sumb[2]{\sum\limits_{#1}^{#2}\;}
+%
+%\newcommand\produ[1]{\prod\limits^{#1}\;}
+%\newcommand\prodl[1]{\prod\limits_{#1}\;}
+%\newcommand\prodb[2]{\prod\limits_{#1}^{#2}\;}
+%
+%\newcommand\defeq{\overset{\underset{\mathrm{def}}{}}{=}}
+%
+%%Math macros:
+%\newcommand\T{^{\scriptscriptstyle T}}
+%\renewcommand\H{^{\scriptscriptstyle H}}
+%
+%\renewcommand\vec[1]{\boldsymbol{#1}}
+%\newcommand\mat[1]{\boldsymbol{#1}}
+%
+%\newcommand\Om{O_\text{m}}
+%\newcommand\Oa{O_\text{a}}
+%\newcommand\Nl{N_\text{l}}
+%\newcommand\Nk{N_\text{k}}
+%\newcommand\1{\vec 1}
+%\newcommand\I{\mat I}
+%\renewcommand*\a{\vec a}
+%\newcommand*\f{\vec f}
+%\renewcommand*\i{\vec i}
+%\renewcommand*\k{\vec k}
+%\newcommand*\n{\vec n}
+%\newcommand*\p{\vec p}
+%\newcommand*\s{\vec s}
+%\newcommand*\w{\vec w}
+%\newcommand*\x{\vec x}
+%\newcommand*\y{\vec y}
+%
+%\newcommand*\A{\mat A}
+%\newcommand*\B{\mat B}
+%% \newcommand*\C{\mat C}
+%\newcommand*\E{\mat E}
+%\renewcommand*\P{\mat P}
+%\newcommand*\eP{\mat{\hat P}}
+%\newcommand*\R{\mat R}
+%\newcommand*\Ri{\R^{-1}}
+%\newcommand*\eR{\mat{\hat R}}
+%\newcommand*\eRi{\hat{\mat R}\;\!^{-1}}
+%\newcommand*\Navg{N_\text{avg}}
+%\newcommand*\W{\mat W}
+%\newcommand*\X{\mat X}
+%\newcommand*\Xd{\X_{\!\Delta}}
+%\newcommand*\Y{\mat Y}
+%
+%\renewcommand\argmin{\text{argmin}}
+%
+%
+%\renewcommand*\P{\mat P}
+%\newcommand*\V{\mat V}
+%\newcommand*\M{\mat M}
+%
+%\renewcommand*\L{\mat \Lambda}
+%% \newcommand*\U{\mat U}
+%% \renewcommand*\t{\mathtiny{^T}}
+%% \newcommand*\h{\mathtiny{^H}}
+%\renewcommand*\t{^T}
+%\newcommand*\h{^H}
+%
+%\newcommand\minus{\scalebox{0.75}[1.0]{$-$}}
+%
+%% \usepackage{tikz}
+%% \usetikzlibrary{shapes,snakes}
+%\usepackage{amsmath,amssymb}
+%% \usepackage{datatool}
+%% \usepackage{glossaries}
+%
+%% \newenvironment{outline}
+%% {\begin{itemize}}
+%% {\end{itemize}}
 
@@
 
-\title{A Real-Time Sonar Simulator for Automatic Target Recognition}
-
-\author{{Jo~Inge~Buskenes,~\IEEEmembership{Student Member,~IEEE,} %
-        Herman~Midelfart, %
-        \O{}ivind~Midtgaard} %
-\thanks{J. I. Buskenes is with the Department of Informatics, University of Oslo, Norway.}%
-\thanks{H. Midelfart and \O. Midtgaard are with The Norwegian Defence Research Establishment (FFI), Norway.}% <-this % stops a space
-% \thanks{Manuscript received April 19, 2005; revised January 11, 2007.}
-}
-
-% The paper headers
-\markboth{IEEE Journal of Oceanic Engineering}%
-{A GPU Sonar Simulator for Automatic Target Recognition}
-
-% Publishers ID mark:
-%\IEEEpubid{0000--0000/00\$00.00~\copyright~2007 IEEE}
-
-% use for special paper notices
-%\IEEEspecialpapernotice{(Invited Paper)}
-
-% for Computer Society papers, we must declare the abstract and index terms
-% PRIOR to the title within the \IEEEcompsoctitleabstractindextext IEEEtran
-% command as these need to go into the title area created by \maketitle.
-\IEEEcompsoctitleabstractindextext{%
-\begin{abstract}
-Template matching is a common technique for automatic classification of objects in synthetic aperture sonar (SAS) images.
-The principle is to isolate an image segment containing the object of interest, correlate it with a set of template images, and assign it to the class of the template yielding the highest correlation coefficient.
-The challenge is to come up with a representative set of template images covering the relevant configurations of object and seabed.
-
-We target this challenge with a sonar simulator that first takes as input a seabed model derived from the real sonar image.
-Then it places a 3D object model on the seabed, renders the scene, and adds the resulting image to the template set.
-For every object type, position, alignment and material, the procedure is repeated, and a correlation coefficient computed.
-The best performance is obtained when these parameters are estimated from the sonar image as part of the classification process.
-The simulator is therefore written in OpenGL and OpenCL and runs on graphics processing units (GPUs).
-The result is a fast performing and portable template generator which can adapt to the characteristics of the current scene on-the-fly.
-\end{abstract}
-
-% Keywords (normally not used for peer reviews)
-\ifPeerReview\else
-\begin{IEEEkeywords}
-Sonar, SAS, simulator, template matching, OpenGL, OpenCL.
-\end{IEEEkeywords}
-\fi}
-% \fi
-
-% make the title area
-\maketitle
-
-% This command fixes abstract positioning for compsoc articles:
-\IEEEdisplaynotcompsoctitleabstractindextext
-
-% (Optional) Add some extra info on cover page of peer review papers:
-% \ifCLASSOPTIONpeerreview
-% \begin{center} \bfseries EDICS Category: 3-BBND \end{center}
-% \fi
-
-% Insert page break and insert second title (peer review mode)
-\IEEEpeerreviewmaketitle
-
-\section{Introduction}
-
-\setcounter{totalnumber}{1}
-\setcounter{topnumber}{1}
-\setcounter{dbltopnumber}{1}
-
-\begin{figure*}[t]\centering%
-\includegraphics[width=\linewidth]{gfx/scene_coordinate_system.svg}%
-\caption{Simulator principle.
-A model of the seafloor and object is first loaded into OpenGL, and an orthogonal camera view set up to match the desired image region on the seafloor.
-Then the scene is rendered and an intensity image and a depth map is created.
-The intensity image along with the depth information is finally combined to form a sonar like image.
-This is our template.}\label{simulator_coordinate_system}%
-\end{figure*}
-
-\begin{figure*}[t]\centering%
-\includegraphics[drawing,width=\linewidth]{gfx/simulator.svg}%
-\caption{Simulator concept.}\label{buildup}%
-\end{figure*}
-
-% \IEEEPARstart{W}{hen} 
-% Autonomous underwater vehicles (AUVs).
-When an AUV with a sonar imaging system encounter an interesting object, it would be very beneficial if adapted its behavior to examine the object further.
-This is challenging as it would need to learn in what ways the interesting objects differ from the background, i.e.
-it relies on previous knowledge.
-It would also need to process this information is near real-time to act in an efficient manner.
-
-% - Adapt simulation to its behaviour, see if they still match.
-
-\IEEEPARstart{A}{utonomous} underwater vehicles (AUVs) are becoming increasingly important in human underwater endeavors.
-When equipped with a synthetic aperture sonar (SAS), they can e.g.
-be used to map the seafloor, survey underwater biology, help localize wrecks, inspect oil pipes, fields and electricity lines, or hunt for mines and biohazard waste.
-Such independent systems have much to gain from being able to adapt to the scenario at hand.
-For instance, with a well designed automatic target recognition (ATR) system on board they can adapt their mission plan to collect more information on particularly interesting objects, or fine-tune the sonar and navigation system to improve imaging performance.
-
-% Automatic target recognition (ATR) is an important component in autonomous underwater vehicles (AUVs), as it allows the vehicle to adapt its mission plan to e.g.
-revisit detected objects for closer examination.
-One way to classify these objects is by using template matching.
-The principle involved  is to isolate an image segment containing the object of interest, compare it with a set of template images of the relevant object classes, and assign it to the class of the template with the the best fit.
-
-In conventional ATR an object of interest is first isolated, its properties attempted estimated, and a classifier used to determine what object class these properties belong to.
-The challenge is finding unique and intrinsic object properties, such as its type, shape and size, and to do so independently of extrinsic properties such as its alignment, burial depth, degeneration level or view angle.
-While SAS provide images with state of the art resolution and dynamic range, they still only represent an ambiguous acoustical 2D projection of the 3D object, littered with a highly coherent noise floor.
-
-An alternative approach is to use template matching, where the isolated image segment is compared to a set of template images covering the relevant object classes, and assigned to the class of the template with the best fit.
-Since the templates are computed from 3D models, this reduces the issue of image ambiguity.
-However, a challenge is finding a set of templates that represent the actual configurations of object and seafloor, Another issue is that the size of precomputed library of templates grows exponentially with the number of parameters needed.
-This limits its use to only a few parameters being coarsely sampled, ultimately yielding inaccurate results~\cite{Midelfart2010}.
-
-% We avoid the problems of conventional template matching by computing the templates on-the-fly as part of the classification process, thereby not needing a precomputed set of templates at all. 
-%  that avoids this problem by running sufficiently fast to create templates adapted to the actual scene as part of the classification process.
-% Classification process
-
-We tackle the problems of conventional template matching by performing:
-%
-\begin{itemize}
-\item \emph{Adaptive template matching}; where a template is selected based on object and seafloor parameters estimated from the SAS image~\cite{Midelfart2010}, and
-\item \emph{Real-time template simulation}; where a template is created tailored to the estimated parameters, improving its accuracy and avoiding the need of a precomputed template library.
-\end{itemize}
-%
-
-The methods presented herein will focus on the template simulator, but some details will also be given on adaptive template matching for completeness.
-More details on this technique can be found in~\cite{Midelfart2010}.
-
- This is achieved with the aid of the massive computing power in graphics processing units (GPUs) and optimized software libraries for scene rendering.
-The simulator loads a 3D model of the seafloor and object class into OpenGL and positions the camera at the sonar location.
-The scene is highlighted using a light tube that extends to infinity along the AUV's axis of probagation.
-We assume rough, isotropic surfaces reflecting sound energy equally in all directions.
-This can be modeled with a Lambertian scattering model~\cite{Blake1993,Bell1995}, where the intensity of the backscattered sound solely depends on the incidence angle of the transmitted signal onto the model surface.
-The OpenGL rendering produces an optical 2D intensity image and depth map that reveals the distance from each pixel to the propagation axis.
-For maximum flexibility this data is finally combined with OpenCL to produce the image templates.
-
-as part of the classification process. This is achieved with the aid of the massive computing power in graphics processing units (GPUs) and optimized software libraries for scene rendering.
-
-The simulator loads a 3D model of the seafloor and an object class model into OpenGL, where emitted sound waves are modeled with a cylindrical light source placed at the sonar transmit location.
-
-We assume rough, isotropic surfaces reflecting sound energy equally in all directions.
-This can be modeled with a Lambertian scattering model~\cite{Blake1993,Bell1995}, where the intensity of the backscattered sound solely depends on the incidence angle of the transmitted signal onto the model surface.
-When rendering OpenGL is set up to produce an optical 2D intensity image and depth map that reveals the distance from each pixel to the propagation axis.
-For maximum flexibility this data is finally combined with OpenCL to produce the image templates.
-
-Various simulators for high frequency, side-looking sonar imagery have been published, e.g.~\cite{Bell1997,Sammelm2003}.
-However, most implementations have prioritized accurate acoustic modeling at the expense of execution speed.
-One exception is the SIGMAS+ simulator in~\cite{Coiras2009a, Coiras2009b}.
-Similarly to our approach it takes advantage of parallel processing on GPUs, and it uses OpenGL to render a 2D image from a 3D model, under the assumption of Lambertian scattering.
-However, SIGMAS+ is still aimed more towards realistically looking images, including effects like ambient noise, which is irrelevant for our purpose of template generation.
-Also, their simulator superimpose images obtained from multiple OpenGL rendering passes to create the sonar image, while we render once and post-process this result with OpenCL to tailor the sonar image with greater flexibility.
-
-For lack of a better name, the simulator will be referred to as "FFISim" throughout this article.
-
-Show:
-\begin{itemize}
-\item Adaptive template matching to avoid large static template libraries,
-\item Improved classification performance by feeding simulator with parameters derived off SAS image,
-\item Further improved classification by performing a fine-search within a narrow search space around the SAS derived parameters. GPU makes this more powerful.
-\end{itemize}
-
-\section{Methods}
-
-Creating a 2D view of an arbitrary complex 3D scene is a non-trivial matter.
-This is why we decided to use the core OpenGL pipeline to do this for us.
-OpenGL is a popular, well-matured and multi-platform application programming interface (API) for rendering 2D and 3D vector graphics.
-It relieves us from the intricacies of projecting vertices, faces and textures defined in a 3D space onto a suitable 2D image plane.
-This section explains how we set up OpenGL for this task, and then proceeds to describe how we post process the OpenGL images with OpenCL to form the final sonar image.
-
-\begin{itemize}
-\item Load arbitrary complex models
-\item Simulate sonar with OpenGL
-\item Speed considerations
-\end{itemize}
-
-Models are not ROCCAN (???)
-
-\subsection{Loading the scene}
-
-Our models are stored in regular 3D model files.
-The simulator begins by loading these into a tree of nodes, each representing a vertex in the model with corresponding facet and texture information.
-Then the vertices are placed into a virtual scene in OpenGL, and a camera set up to capture the part of the scene corresponding to the image.
-When OpenGL renders the image, it first computes which vertex contributes to which pixel in the image, then computes the intensity and color of the pixel based on the node's facet, texture and scattering model.
-The code that operates on vertices are called a vertex shader, and the code that operates on pixels are called a fragment shader.
-
-The mapping of the vertices from object coordinates to image coordinates happen through a set of transformations:
-%
-\begin{align*}
-\bmat{x\\y\\0\\1}_\text{image} = \boldsymbol{P} \cdot \boldsymbol{V} \cdot \boldsymbol{M} \cdot \bmat{x\\y\\z\\1}_\text{object},
-\end{align*}
-%
-where $\bmat{\cdot}_\text{model}$ is the loaded model, $\bmat{\cdot}_\text{image}$ is the output image, and $\M$, $\V$ and $\P$ are transformation matrices. These will be described next.
-
-The first transform we apply is the model matrix $\M$, which operates in the model's local coordinate system. It is defined as
-%
-\begin{align}
-\M = \bmat{x_t \\ y_t \\ z_t \\ w}_\text{\parbox{1cm}{\setlength\baselineskip{0.3cm}transformed\\model}} &= \boldsymbol{T} \cdot \boldsymbol{R} \cdot \boldsymbol{S} \bmat{x \\ y \\ z \\ w}_\text{model},
-\end{align}
-%
-where $\boldsymbol{S}$, $\boldsymbol{R}$ and $\boldsymbol{T}$ are matrices\todo{be explicit on dependencies?} for scaling, rotation and translation, respectively.
-These are provided for reference in Appendix \ref{transformation_matrices}.
-
-Next we map the model into world coordinates using the view matrix $\V$.
-This involves aligning the camera to precisely fit the image region.
-In our case the camera represents the phase center of the sonar, and it is placed in the world origin.
-The transformation then take the form of
-%
-\begin{align}
-\V &= \mat{T}(x_\text{c}, y_\text{c}, z_\text{c}) \cdot \R_\text{x}(\phi_x, 0, 0) \nonumber\\
-&= 
-\bmat{
-1  &  0  &  0  &  x_\text{c} \\
-0  &  1  &  0  &  y_\text{c} \\
-0  &  0  &  1  &  z_\text{c} \\
-0  &  0  &  0  &  1 \\
-} \cdot
-\bmat{
-1  &  0           &  0           &  0 \\
-0  &  \cos\phi_x  &  -\sin\phi_x &  0 \\
-0  &  \sin\phi_x  &  \cos\phi_x  &  0 \\
-0  &  0           &  0           &  1 \\
-},
-\end{align}
-%
-where $(x_\text{c}, y_\text{c}, z_\text{c})$ is the image center coordinates for along track, cross track, and depth, respectively, and $\phi_x$ is the elevation angle\todo{is this correct?} of the sonar.
-It is given as
-%
-\begin{align*}
-\phi_x &= \arctan\left(\frac{y_\text{mean}}{-z_\text{mean}}\right).
-\end{align*}\todo{use absolute sign?}
-
-Finally the relevant part of the scene is projected onto the image plane with a projection matrix $\P$.
-There are two main types of projection, perspective and orthographic.
-Perspective projection scales all $x$ and $y$ coordinates inversely with the distance to the objects to create a sense of perspective in the image.
-Orthographic projection, on the other hand, maps all the scene's vertices to a squared cube where each dimension has the range [-1,1].
-This will render all objects equally large whether they are in the distance or up close.
-The matrix that gives orthographic projection is defined as
-%
-\begin{align}
-\P = \left(\begin{matrix}
-\frac{2}{r-l}  &  0              &  0              &  -\frac{r+l}{r-l} \\
-0              &  \frac{2}{t-b}  &  0              &  -\frac{t+b}{t-b} \\
-0              &  0              &  \frac{-2}{f-n} &  -\frac{f+n}{f-n} \\
-0              &  0              &  0              &  1
-\end{matrix}\right)
-\end{align}
-%
-where $l,r,b$ and $t$ are the left, right, bottom and top boundaries of the projection plane, respectively, and where $n$ and $f$ is the near and far limits of the cube that encompass the scene.
-These parameters need to be adapted to provide a 2D image that matches the image size, which we specify in along track and cross track coordinates ($x,y$).
-This is straightforward for left and right since matches the $x$-axis
-% They define the window through which we observe our scene.
-\begin{align}
-l &= x_\text{min}  &  r &= x_\text{max}
-\end{align}
-%
-For the near and far limits one simply needs to make sure the scene will fit in it, say
-%
-\begin{align}
-n &= 0\;\text{m}   &  f &= 300\;\text{m}.
-\end{align}
-%
-The bottom and top boundaries can be inferred from \Fig{camera_top_bottom}. First we define an origin vector that points to the center of the image
-%
-\begin{align}
-\vec o &= \left[\frac{x_\text{min}+x_\text{max}}{2}, \frac{y_\text{min}+y_\text{max}}{2}, \frac{z_\text{min}+z_\text{max}}{2} \right]\T,
-\end{align}
-%
-and the distance from the image origin to its $y$-boundaries is given as
-%
-\begin{align}
-\Delta y &= \frac{|y_\text{max} - y_\text{min}|}{2}.
-\end{align}
-%
-To find the nearest and farthest point in the image that will be part of the projected scene, $\alpha$ and $\beta$, we must consider the seafloor rotation angle around the $x$-axis, $\Phi_x$ 
-%
-\begin{align}
-\vec\alpha &= \vec 0 - \left[0, \Delta y, \Delta y \cdot \tan\Phi_\text{x} \right] \nn
-\vec\beta &= \vec 0 + \left[0, \Delta y, \Delta y \cdot \tan\Phi_\text{x} \right]
-\end{align}
-
-\begin{figure}[t]\centering%
-\includegraphics[drawing,width=\linewidth]{gfx/opengl_part.svg}%
-\caption{Setting camera top and bottom boundaries.}\label{camera_top_bottom}%
-\end{figure}
-
-\subsection{OpenGL rendering}
-
-To produce the sonar templates we assume a rough, isotropic surface that reflects energy equally in all directions.
-This permits us to use a Lambertian scattering model where the backscatter intensity depends only on the incidence angle~\cite{Zhang1999}.
-It does not consider observation angle or sound frequency, but for the purpose of creating templates this is not needed.
-
-The rendered image will appear as if we placed a window at the sonar and looked through it in the direction of the image.
-This window is resized to make sure that the template image perfectly fills it.
-
-\subsection{OpenCL post-processing}
-
-The ``camera image'' rendered with OpenGL is not in along-track and cross-track coordinates as we want it to be.
-It does, however, show us what parts of the scene that are visible from the sonar.
-OpenGL can also produce a depth map that reveals the distance to each of image pixels.
-This information can be converted to a ranged sonar-like image by simply adding up all the intensity values that share the same depth for each range line.
-We perform this computation in OpenCL.
-It allows general purpose programming on GPUs and can interoperate with OpenGL quite nicely.
-This way we keep all the calculations on the GPU.
-
-\todo{
-- Kernel tuning.
-- Sharing. PBO. Asynchronous.
-}
-
-\subsection{Template matching}
-
-1. Segmentation
-2. Parameter estimation
-   Available: Range, altitude, resolution => Input to both standard and adaptive. In standard the best fit out of a predefined template database is selected, then resampled to correct resolution.
-   Additional for adaptive, parameters that modify object model: Target aspect angle, seafloor slope, object length and height.
-
-   Target aspect angle computed with Radon Transform, seafloor under object modeled as plane with a slope estimated from bathymetric data, length estimated from longest part of highlight region, and height estimated from the shape of the shadow.
-The burial depth is n
-
-3.
-
-   Details: \ref{Midelfart2010}.
-
-Let $S(x,y)$ be the isolated part of the
-
-Matching the isolated image segment with the template take the form of
-
-\begin{align}
-c_{h,i} = \underset{x',y'}{\text{argmax}} \sum\limits_{\forall x}\sum\limits_{\forall y} \big| H(x,y) - T(x-x',y-y') \big|
-\end{align}
-
-\begin{align}
-c_i = \frac{c_{h,i} + c_{s,i}}{2}
-\end{align}
-
-And the final template used is the one with the greatest correlation coefficient
-
-\begin{align}
-\underset{T_i(x,y)}{\text{argmax}} C_i\sum\limits_{\forall x}\sum\limits_{\forall y}  S(x,y) T(x-x',y-y')
-\end{align}
-
-Then assign object to class $i$.
-
-To narrow down the search space of possible templates we derive some reliable statistics of the seafloor and object from the SAS image.
-While FFISim accepts an arbitrarily complex seabed, we find that modeling it as a plane wave with the right tilt works well.
-The object is more difficult.
-For instance, we estimate its length and height from the shape of its shadow, but this fails if the object is notably immersed into the seabed at one end.
-
-After the a template is created we decompose it into its highlight and shadow component, and compute a correlation score between these and the corresponding areas in the SAS image.
-Then the two scores are summed and normalized to form the final correlation score of the template.
-
-We tag the template matchers as either semi- or fully adaptive.
-The former refers to using the best matching SIGMAS+ generated template out of a precomputed set of templates, while the latter refers to the proposed method of using FFISim to generate templates on the-the-fly.
-\todo{Should be more precise.
-How many templates?
-Parameter span?
-How finely were the parameters sampled?}
-
-% It is also possible to compute the sonar image using the OpenGL ``blend'' feature.
-
-\subsection{Speed considerations}
-
-Using OpenGL and OpenGL to generate sonar templates on the GPU allow hundreds of sonar templates to be formed per second.
-While this is sufficient for our needs.
-However, initializing the OpenGL and OpenCL context and loading 3D models from file into it is a slow process, and typically takes several seconds for every invocation.
-Therefore, to alleviate the simulation performance we either had to supply it with a list of parameters or ensure that it did not have to be shut down for every invocation.
-
-The rendering loop that uses OpenGL and OpenCL to form
-
-\todo{
-Write to file
-}
-
-%    """
-%    
-%    stm: Standard template matching, made with SIGMAS.
-%    atm: Adaptive template matching, made with FFISim.
-%    atm_bury: Adaptive template matching where a search of bury depths is used.
-%    atm_bury_restr: Restricted?
-%    
-%    Adaptiveness made by letting the template matcher use 5 different rotation
-%    angles in the ground-range axis. 
-%    
-%    
-%    Data from Jesusbukta
-%    
-%    Figures are from run 2012.06.16
-%    det210 - Cylinder 1: est 2.92m (short side towards sonar)
-%    det233 - Cylinder 2: est 3.05m (diagonal)
-%    det379 - Torpedo: est 5.06m (looks good)
-%    
-%    ROC curves from 3 different runs, first two just 2 days apart in 2009, last one in 2012.
-%    
-%    run_20120616 - last one. Much better with adaptive templates.
-%    
-%    ROC curves made by average correlation score of shadow/highlight from template matching.
-%    
-%    
-%    fig4b(2015.11.16)
-%       Bug that skipped masking of port side (negative y) detections past 180m range fixed
-%       
-%    fig4 (2015.11.16)
-%       Bug leading to incorrect estimation of cylinder orientation fixed.
-%          This affects run090601_1. Should be better after this.
-%          
-%    fig3 (2015.11.09)
-%       Detections outside 180m range have been masked (but bug for port side fixed in fig4b).
-%       Fixed an incorrectly annotated detection.
-%       
-%    fig2 (2015.11.02)
-%       Height of object estimated from shadow length
-%       Let adaptive method try 5 different rotations in ground-range (-4, -2, 0, 2, 4)
-%          Improved results, but incorrect classification of:
-%          5 detections in first run, and
-%          3 detections in second run
-%          => For FPR=0.2 the results are slightly worse than with normal adaptive (particularly first run).
-%          Images are problematic in these cases:
-%             2 detections get wrong orientation because a line (anchor) cross the image
-%             Other misclassifications in image regions with poor echo or shadow strength
-%          => Not a big issue as not enough time to detect more than ~20% of the detections in a run
-%             In a MCM operation these detections would never be used.
-%             The runs have 3000 detections each, photographing 20% of these is too much.
-%          Performance FPR>0.2 not interesting.
-%          Performance FPR<0.2 much better with adaptive template matching.
-%          
-%    fig (2015.10.23)
-%       atm - cylinder orientation and length estimated from SAS image and fed into simulator.
-%       
-%       ROC curves from last run. Computing the mean correlation score of the shadow and highlight/echo.
-%       Early results. Can only check vs. cylinder-like objects.
-%       Standard template library created based on estimation of physical size of known cylinder mines.
-%       Not possible for adaptive method to perform better when hitting such a target.
-%       
-%  
-%    """
-
-\section{Results \& Discussion}
-
-\begin{figure*}[!tbp]\centering%
-\includegraphics[width=\linewidth]{gfx/fig_images_sonar_simulator_tagged.svg}\label{fig_images_sonar_simulation}%
-\caption{Selected objects.
-Top row show SAS images of 2 cylinders and a torpedo located in the Jesus Bay, Norway.
-Center row show the best match from a conventional template matcher, and bottom row show the adaptive template matcher.
-The contour of the SAS object is indicated with red lines.
-Note that the adaptive technique produce templates with a better fit to the highlight, but that this does not necessarily lead to a better match against the shadow.}
-\end{figure*}
-
-\begin{figure*}[!tbp]\centering%
-\includegraphics[width=\linewidth]{gfx/fig_rocs.pdf}\label{fig_roc_curves}%
-\caption{Receiver operating characteristics (ROC) curves.}
-\end{figure*}
-
-\begin{figure*}[!tbp]\centering%
-\includegraphics[width=\linewidth]{gfx/fig_overlay.pdf}%
-% \parbox{\linewidth}{\small\centering\vspace{.5\baselineskip}
-% \newline\centering\newline %\\[.5\baselineskip]
-\newcommand\cdesc[2]{{\raggedright\setlength\fboxsep{0pt}
-\fbox{\colorbox[HTML]{#1}{\vrule height8.5pt depth3.5pt width0pt\hspace{.5cm}}}\ \ #2\\}}
-\begin{minipage}{.4\linewidth}
-\cdesc{800000}{Template highlight and image highlight}
-\cdesc{FF1000}{Template highlight only}
-\cdesc{FFEB00}{Image highlight only}
-\cdesc{83FF7C}{Background pixels}
-\end{minipage}%
-\begin{minipage}{.4\linewidth}
-\cdesc{000083}{Template shadow and image shadow}
-\cdesc{0014FF}{Template shadow only}
-\cdesc{00EFFF}{Image shadow only}
-\cdesc{00A7FF}{Template shadow and image highlight}
-\end{minipage}%}
-\caption{Comparison of classification performance for the standard and adaptive template technique.
-The three objects were more or less arbitrary picked from the scene.
-For the roughly 3\,m long cylinders both methods perform similarly, the standard method seemingly with a slight edge.
-This is no surprise given that the 3\,m long cylinder model that make up the static template database fit the actual object well here.
-However, for the longer torpedo-like object the adaptive technique is clearly better.
-This is its merit; for objects and geometries that are not included in the static template database we can expect the adaptive techniques to perform better.}\label{fig_image_simulation}%
-\end{figure*}
-
-\newlength\imgspacing\setlength\imgspacing{.5cm}
-\setcounter{topnumber}{2}
-
-The performance of our simulator and template matching techniques where measured using experimental data from an interferometric HISAS1030 synthetic aperture sonar (SAS) attached to a HUGIN autonomous underwater vehicle (AUV), both developed by Kongsberg Maritime, Norway.
-The HISAS1030 is a fully digital phased array with 32 hydrophones, 100\,kHz center frequency and 30\,kHz bandwidth.
-It is 1.2\,m long, has a half-power beamwidth (HPBW) of 23$\circ$, and is capable of generating synthetic images with a theoretical resolution of 3-4\,cm in both dimensions.
-The version of HUGIN used also carries an camera that is used for optical inspection of interesting objects.
-
-We present data from three different HUGIN runs from the Jesus Bay in Norway.
-The first two are taken a couple days apart in January 2009, containing roughly 2700 and 3200 detections, respectively.
-The last is from June 2012 and contains roughly 650 detections.
-The detections represent mostly cylinder shaped objects such as barrels, pipe mines and torpedoes.
-Therefore, all templates were based on a basic cylinder model.
-
-The two template matchers can be visually inspected in \Fig{fig_images_sonar_simulator_tagged}.
-
-The classification performance is quantified with receiver operating characteristics (ROC) curves from all runs in \Fig{fig_roc_curves}.
-The 2009 runs proves a greater challenge than that from 2012, with 5 and 3 false positive detections at FPR=20\% in the January 6.
-and 8.
-run, respectively.
-This came as no surprise due to the 2009 runs containing fish trawler tracks and having poorer highlight and shadow quality than the run from 2012.
-However, for FPR<20\% the adaptive techniques outperform the static template library.
-For e.g.
-a mine counter measures (MCM) operation, this is the only region of interest because only a few objects out of the 3000 detections will be visually inspected up close.
-
-- Stm stimulated with length matching test cylinder. Perfect score against this.
-- Atm - not given to improve score. Shadow and highlight both need to match, and overfitting could lead to reduced performance.
-
-In \Fig{fig_images_sonar_simulator_tagged} we compare the SAS image to the templates generated by the conventional and adaptive template matcher.
-Three typical objects from the 2012 were chosen, all being cylindric.
-The length two first being estimated to roughly 3\,m and the last 5\,m, likely representing cylinder mines and a torpedo, respectively.
-
-Selected objects.
-Top row show SAS images of 2 cylinders and a torpedo located in the Jesus Bay, Norway.
-Center row show the best match from a conventional template matcher, and bottom row show the adaptive template matcher.
-The contour of the SAS object is indicated with red lines.
-Note that the adaptive technique produce templates with a better fit to the highlight, but that this does not necessarily lead to a better match against the shadow.
-
-Rotation around ground range axis: -4,-2,0,2,4 degrees.
-
-Like the majority of objects in the Jesus Bay they are all cylinders with a length of a few meters.
-The same objects are simulated in \Fig{fig_image_simulation}, one with SIGMAS and the two other with FFISim.
-Each of these simulations were based on
-
-Good statistics drawn from the image -> adaptive better. Poor statistics -> hit and miss.
-
-1. Comparison with standard template library (ROC enough?)
-   ROC: Adaptive better at FPR<0.2 in all cases.
-2. Comparison with SIGMAS
-   Fairly similar images, performance is similar. 
-   Used to create the standard template library.
-3. Comparison with search space
-   ROC: Additional search always better FPR<0.2.
-   Explain search criterion.
-
-SAS images - look good. Random pick objects. Lengths derived from image.
-Simulated images - look good. Adaptive simulation more accurate. Otherwise similar. Different background level. Adaptive technique excels for less normal objects/geometries.
-Overlap - Provide quantification measure.
-
-1. Adaptive better? Compare with standard. Doing this in ROC
-2. Improved classification? ROCs
-3. Further improvement by doing fine-search? ROC
-
-An issue the adaptive technique must deal with when estimating object length
-
-As the ROC curves indicate, the adaptive template matching techniques both  ROC curves demonstrate the
-
-The majority of the objects in the Jesus Bay are cylinder shaped. A SAS image of three such objects are displayed in \Fig{fig_image_sonar}
-
-% \begin{figure*}[!tbp]\centering%
-% \subfloat[SAS images of 2 cylinders and a torpedo.]{%
-% \includegraphics[width=0.5\linewidth]{gfx/fig_images_sonar_simulation.pdf}%
-% \label{fig_sonar_simulation}}%
-% \subfloat[Cylinder simulation with parameters estimated from SAS image: Length 2.6\;m, burial depth 0.263\;m, aspect angle 105$^\circ$.]{
-% \includegraphics[drawing,width=0.5\linewidth]{gfx/sim_cylinder_submerged_adaptive.svg}%
-% \label{sim_cylinder}}%
-% \caption{A SAS image of a cylinder and a template simulation adapted to it.}
-% \end{figure*}
-% 
-% 
-% \begin{figure*}[!tbp]\centering%
-% \subfloat[SAS image of a cylinder with 2.6\;m length and 0.53\;m radius.]{%
-% \includegraphics[drawing,width=0.5\linewidth]{gfx/data_cylinder_submerged.svg}%
-% \label{data_cylinder}}%
-% \subfloat[Cylinder simulation with parameters estimated from SAS image: Length 2.6\;m, burial depth 0.263\;m, aspect angle 105$^\circ$.]{
-% \includegraphics[drawing,width=0.5\linewidth]{gfx/sim_cylinder_submerged_adaptive.svg}%
-% \label{sim_cylinder}}%
-% \caption{A SAS image of a cylinder and a template simulation adapted to it.}
-% \end{figure*}
-
-% \newcommand\cdesc[2]{#2}
-
-% Whenever conventional computer graphics processing can be used to 
-% GPUs are designed to provide the best possible graphics processing performance, and there large ecosystem of tools and libraries  available 
-% Using a GPU to drive a simulator is advantageous since
-% 
-% 
-% Creating a GPU-based simulator is attractive since computer is an attractive solution as there are a  graphics traits. For one over a CPU-based standard simulator is two-fold:
-
-Our GPU-based simulator is very fast.
-On a computer with a six-core Intel Core i7-3930K and a Radeon HD7970 it computes almost 1000 templates per second, each being 1 megapixel large.
-This makes it possible to tailor the templates very closely imaged objects.
-In contrast, in the standard approach a template library must be created beforehand from a limited set of parameter values.
-Hence, it may happen that no template in the library is a close fit of the object in the image even though the object actually is a target of interest.
-
-This performance was demonstrated on an image from the HISAS 1030 synthetic aperture sonar mounted on the HUGIN AUV.
-Out of the objects in the scene we selected a 2.6\;m long cylinder that was partly buried in the sea sediments (Fig.
-\ref{data_cylinder}).
-The length, immersion depth and aspect angle of this cylinder were estimated from the SAS image using our  adaptive template matching approach~\cite{Midelfart2010}.
-Then these parameters were used by the simulator to create an adaptive template (Fig.
-\ref{sim_cylinder}), which is an almost perfect match of the cylinder.
-We also created standard templates for a regular template matching approach.
-In this case, we assumed that the cylinder mine had a generic length of 2\;m (like a MP80 or a Murena mine) and was proud on the seafloor.
-Moreover, templates were created for every 10$^\text{th}$ degree of the aspect angle.
-We believe these assumptions to be typical for a template library for cylinder mines.
-
-The standard and adaptive templates were then matched to the HISAS image.
-This is illustrated in Fig.
-\ref{overlay_cylinder} for a standard template and in Fig.
-\ref{overlay_cylinder_adaptive} for an adaptive template.
-These images were created by segmenting the image and the templates into highlight, shadow, and background regions.
-The regions from a template were then laid on top of the regions of the image to illustrate how well these regions matched.
-Note how the adaptive template obtained a much closer fit to image than the standard template.
-This was also reflected in the correlation scores (which were created with the method described in \cite{Midelfart2010}) that were 0.613 for the standard template and 0.813 for the adaptive template.
-Hence, the standard approach was less likely to classify the cylinder in the image correctly as the standard templates were not created specifically for the target.
-
-\begin{itemize}
-\item GPUs low power. Good for integration in AUVs.
-\item Good accuracy even with a real-time constraint in e.g. a AUV.
-\item Improved inertial navigation?
-\item Aid in understanding the key parts of the SAS image and in discovering potential flaws in the SAS image. Anomalies can be flagged to make SAS more robust?
-\end{itemize}
-
-\section{Conclusion}\label{conclusion}
-
-One way to automatically classify objects in SAS images is to compare the imaged objects with a predefined set of templates.
-However, this is suboptimal as it is infeasible to create accurate templates for all the relevant configurations of seabeds and objects.
-To solve this we have implemented a SAS simulator that creates the templates in delayed real-time based on parameters estimated from the current scene.
-In our studies this improves the match between the SAS objects and their corresponding template significantly in most cases.
-
-To obtain the delayed real-time performance we implemented the simulator on a GPU.
-These devices have a theoretical peak performance that is typically an order of magnitude higher than CPUs in a comparable price range.
-We found this potential to be effectively utilized with the well matured and highly optimized OpenGL graphics processing API.
-Our simulator use this framework for most of the scene processing.
-
-A final post-processing step is performed in OpenCL, which allow general purpose GPU programming.
-It interoperates well with OpenGL and adds a lot of flexibility to the simulation process.
-This is valuable as the simulator is still being actively developed and will likely see new features that can not be easily implemented in OpenGL alone.
-
-\begin{itemize}
-\item Motion compensation
-\item 
-\end{itemize}
-
-\appendix
-
-\section{Transformation matrices}\label{transformation_matrices} 
-%
-We place, resize and orient objects in an euclidean 3D space using suitable transformation matrices that are applied to each of the object's vertices:
-\begin{align}
-\bmat{x_t \\ y_t \\ z_t \\ w}_\text{\parbox{1cm}{\setlength\baselineskip{0.3cm}transformed\\model}} &= \boldsymbol{T}\R\boldsymbol{S} \bmat{x \\ y \\ z \\ w}_\text{model}.
-\end{align}
-Here a model vertex at position $(x,y,z)$ is first scaled with $\boldsymbol{S}$, then rotated with $\R$
-There is nothing special to these matrices, but we supply them here for completeness.
-
-We implement rotations in a 3D euclidean space by combining matrices that rotate around each of the three respective dimensions. For rotation by an angle $\phi_x$ clockwise around the $x$-axis we have
-\begin{align}
-\R_x(\phi_x) &= \bmat{
-1  &  0           &  0           &  0 \\
-0  &  \cos\phi_x  &  -\sin\phi_x &  0 \\
-0  &  \sin\phi_x  &  \cos\phi_x  &  0 \\
-0  &  0           &  0           &  1 \\
-},
-\end{align}
-and similarly, for the other two euclidean dimensions we have
-\begin{align}
-\R_y(\phi_y) &= \bmat{
-\cos\phi_y  &  0  &  \sin\phi_y  &  0 \\
-0           &  1  &  0           &  0 \\
--\sin\phi_y &  0  &  \cos\phi_y  &  0 \\
-0           &  0  &  0           &  1 \\
-},
-\end{align}
-and
-\begin{align}
-\R_z(\phi_z) &= \bmat{
-\cos\phi_z  &  -\sin\phi_z &  0  &  0 \\
-\sin\phi_z  &  \cos\phi_z  &  0  &  0 \\
-0           &  0           &  1  &  0 \\
-0           &  0           &  0  &  1 \\
-}.
-\end{align}
-The order in which these are applied matters. We rotate around the $x$-axis first, then around $y$ and finally around $z$:
-\begin{align}
-\R(\phi_x, \phi_y, \phi_z) &= \R_z(\phi_z)\;\R_y(\phi_y)\;\R_x(\phi_x).
-\end{align}
-
-\begin{align}
-\boldsymbol{T}(dx,dy,dz) &= \bmat{
-1  &  0  &  0  &  dx \\
-0  &  1  &  0  &  dy \\
-0  &  0  &  1  &  dz \\
-0  &  0  &  0  &  1 \\
-}
-\end{align}
-
-\begin{align}
-\boldsymbol{S}(s_x,s_y,s_z) &= \bmat{
-s_x  &  0    &  0    &  0 \\
-0    &  s_y  &  0    &  0 \\
-0    &  0    &  s_z  &  0 \\
-0    &  0    &  0    &  1 \\
-}
-\end{align}
-
-% \section{MVDR Complexity Formulas}\label{mvdr_formulas}
-% 
-% \section{GPU Throughput}\label{throughput}
-% 
-% bs bs
-
-% use section* for acknowledgement
-\ifCLASSOPTIONcompsoc% % This command fixes abstract positioning for compsoc articles:
-% \IEEEdisplaynotcompsoctitleabstractindextext
-% 
-% % (Optional) Add some extra info on cover page of peer review papers:
-% % \ifCLASSOPTIONpeerreview
-% % \begin{center} \bfseries EDICS Category: 3-BBND \end{center}
-% % \fi
-% 
-% % Insert page break and insert second title (peer review mode)
-% \IEEEpeerreviewmaketitle
-% 
-% 
-%
-
-  \section*{Acknowledgments}
-\else
-  \section*{Acknowledgment}
-\fi
-
-The authors would like to thank the Norwegian Defence Research Establishment (FFI) for funding to develop the simulator\todo{more?}.
-
-% Can use something like this to put references on a page
-% by themselves when using endfloat and the captionsoff option.
-\ifCLASSOPTIONcaptionsoff
-  \newpage
-\fi
-
-\bibliographystyle{IEEEtran}
-\bibliography{references}
-
-% Generated by IEEEtran.bst, version: 1.13 (2008/09/30)
-% \begin{thebibliography}{10}
-% stuff here
-% \end{thebibliography}
-
-\input{bio/jo_inge_buskenes}
-% \input{bio/jon_petter_aasen}
-% \input{bio/carl-inge_nilsen}
-% \input{bio/andreas_austeng}
-
-% 
-% IMAGE METRICS
-%
-% - Point spread function (res via main lobe width, SAS gain via PDF height)
-% - Constrast measures
-%
-% ACR = cL/2
-% C = (s+n)/n ratio
-% - 
-% WHO's needing processing power
-% - Centre for Maritime Research and Experimentation
-
-\vfill
+\input{document}
 

### 2018-10-11 HRUIonGPU/document.tex
Testing change of frame letters

@@
 We also give an introduction to the GPU computing model.
-In Section \ref{sec:meth} we give a summary of our parallel implementation of the element-space Capon beamformer, before the GPU implementation of the beamspace Capon beamformer is presented in Section \ref{sec:bs}.
-Benchmark results and resulting images for both implementations are presented in Section \ref{sec:bench} and \ref{sec:images}, together with a discussion of trading image quality for speed in Section \ref{sec:trade}.
-Finally we discuss the results in Section \ref{sec:dis}, and draw conclusions in Section \ref{sec:con}.
-
-\section{Background}\label{background}
-
-\subsection{Capon Beamforming}\label{sec:es-capon}
+In Section \ref{II_sec:meth} we give a summary of our parallel implementation of the element-space Capon beamformer, before the GPU implementation of the beamspace Capon beamformer is presented in Section \ref{II_sec:bs}.
+Benchmark results and resulting images for both implementations are presented in Section \ref{II_sec:bench} and \ref{II_sec:images}, together with a discussion of trading image quality for speed in Section \ref{II_sec:trade}.
+Finally we discuss the results in Section \ref{II_sec:dis}, and draw conclusions in Section \ref{II_sec:con}.
+
+\section{Background}\label{II_background}
+
+\subsection{Capon Beamforming}\label{II_sec:es-capon}
 
@@
 \begin{align}
-z[n] = \sum_{m = 0}^{M-1}w_m^*x_m[n] = \vec{w}^H\vec{x}[n], \label{eq:z}
+z[n] = \sum_{m = 0}^{M-1}w_m^*x_m[n] = \vec{w}^H\vec{x}[n], \label{II_eq:z}
 \end{align}
@@
 
-As shown in Fig.\ \ref{fig:mvbf}, the basic idea of an adaptive beamformer is to form a complex weight vector based on the received data.
+As shown in Fig.\ \ref{II_fig:mvbf}, the basic idea of an adaptive beamformer is to form a complex weight vector based on the received data.
 The Capon beamformer produces weights that minimize the beamformer output power while maintaining unity gain in the focus direction.
@@
 %\begin{align}
-%&\min_{\vec{w}} E\{|z[n]|^2\} = \min_{\vec{w}}\vec{w}^H\mat{R}\vec{w}\label{eq:minz}\\
-%&\text{subjected to } \vec{w}^H\vec{a} = 1,\label{eq:constraint}
+%&\min_{\vec{w}} E\{|z[n]|^2\} = \min_{\vec{w}}\vec{w}^H\mat{R}\vec{w}\label{II_eq:minz}\\
+%&\text{subjected to } \vec{w}^H\vec{a} = 1,\label{II_eq:constraint}
 %\end{align}
 %where $\mat{R} = E\{\vec{x}[n]\vec{x}[n]^H\}$ is the spatial covariance matrix, and $\vec{a} = \left[ \begin{array}{cccc} e^{-jk_{\tilde{x}}\tilde{x}_0}& e^{-jk_{\tilde{x}}\tilde{x}_1}& ...& e^{-jk_{\tilde{x}}\tilde{x}_{M-1}}\end{array}\right] ^T$ is a far-field steering vector towards angle $\theta = \arcsin(\frac{\lambda k_{\tilde{x}}}{2\pi})$.
-%The constraint in (\ref{eq:constraint}) is there to avoid the trivial solution $\vec{w} = 0$, and a distorted respons in the direction of $\vec{a}$. 
-%The solution to the minimization problem in (\ref{eq:minz}) and (\ref{eq:constraint}) is
+%The constraint in (\ref{II_eq:constraint}) is there to avoid the trivial solution $\vec{w} = 0$, and a distorted respons in the direction of $\vec{a}$. 
+%The solution to the minimization problem in (\ref{II_eq:minz}) and (\ref{II_eq:constraint}) is
 
@@
 Adaptive weights, $\vec{w}[n]$, are then calculated based on in-phase data, and finally the output is formed by a Capon-weighted coherent sum.}
-\label{fig:mvbf}
+\label{II_fig:mvbf}
 \end{figure}
@@
 \begin{align}
-\mat{\breve{R}}[n] = \frac{1}{N_LN_K}\sum_{n'=n-K}^{n+K} \sum_{l=0}^{N_L-1} \vec{x}_l[n']\vec{x}_l[n']^H,\label{eq:R}
+\mat{\breve{R}}[n] = \frac{1}{N_LN_K}\sum_{n'=n-K}^{n+K} \sum_{l=0}^{N_L-1} \vec{x}_l[n']\vec{x}_l[n']^H,\label{II_eq:R}
 \end{align}
@@
 %
-\begin{align}\label{eq:diag}
+\begin{align}\label{II_eq:diag}
 \mat{\hat{R}}[n] = \mat{\breve{R}}[n] + \epsilon[n]\mat{I} = f(\vec{x}[n]).
@@
 %
-\begin{align}\label{eq:diag_adapt}
+\begin{align}\label{II_eq:diag_adapt}
 \epsilon[n] &= d \times \frac{\text{trace}\{\mat{\breve{R}}[n]\}}{L},
@@
 %
-\begin{align}\label{eq:b}
+\begin{align}\label{II_eq:b}
 \vec{b}[n] = \mat{\hat{R}}[n]^{-1}\vec{a} \in \mathbb{C}^L,
@@
 
-Finally the Capon weight vector is formed from (\ref{eq:b}) as
-%
-\begin{align}\label{eq:w}
+Finally the Capon weight vector is formed from (\ref{II_eq:b}) as
+%
+\begin{align}\label{II_eq:w}
 \vec{w}[n] = \frac{\vec{b}[n]}{\vec{1}^H\vec{b}[n]} \in \mathbb{C}^L,
@@
 \begin{align}
-z[n] &= \frac{1}{N_L}\vec{w}[n]^H \sum_{l=0}^{N_L-1} \vec{x}_l[n] \label{eq:z_mv}\\
-&= 1/N_L \, (\vec{w}[n] * \vec{1}_{N_L})^H\vec{x}[n] \label{eq:z_mv2},
+z[n] &= \frac{1}{N_L}\vec{w}[n]^H \sum_{l=0}^{N_L-1} \vec{x}_l[n] \label{II_eq:z_mv}\\
+&= 1/N_L \, (\vec{w}[n] * \vec{1}_{N_L})^H\vec{x}[n] \label{II_eq:z_mv2},
 \end{align}
@@
 where $*$ is a non-truncated convolution across elements.
-Note how we can either choose to sum all subarrays (\ref{eq:z_mv}), or accumulate a weighting per element (\ref{eq:z_mv2}).
-
-In passive systems, assuming stationary statistics, the sample covariance matrix in (\ref{eq:R}) is typically averaged over a long sequence of samples\cite{Krima}.
+Note how we can either choose to sum all subarrays (\ref{II_eq:z_mv}), or accumulate a weighting per element (\ref{II_eq:z_mv2}).
+
+In passive systems, assuming stationary statistics, the sample covariance matrix in (\ref{II_eq:R}) is typically averaged over a long sequence of samples\cite{Krima}.
 Since medical ultrasound imaging is an active, short-pulsed system, there is a limited amount of data from which $\mat{\hat{R}}$ can be estimated.
@@
 %\caption{}
-%\label{fig:mv}
+%\label{II_fig:mv}
 %\end{figure}
@@
 
-\subsection{Beamspace Processing}\label{bs-capon}
+\subsection{Beamspace Processing}\label{II_bs-capon}
 Nilsen and Hafizovic \cite{Nilsen2009} have proposed to apply beamspace processing to reduce the required computations of the Capon beamformer when applied to medical ultrasound imaging.
@@
 \begin{align}
-\vec{x}_\text{BS}[n] &= \mat{B}\vec{x}[n], \ \ \ \ \ \ \ \ \ \ \ \ \ \, \mat{B} \in \mathbb{C}^{N,M} \label{eq:butler} \\ 
+\vec{x}_\text{BS}[n] &= \mat{B}\vec{x}[n], \ \ \ \ \ \ \ \ \ \ \ \ \ \, \mat{B} \in \mathbb{C}^{N,M} \label{II_eq:butler} \\ 
 [\mat{B}]_{p,q} &= \frac{1}{\sqrt{M}}e^{-j 2 \pi p q / M} \ \ \ N \le M.
@@
 
-The beamspace version of the Capon beamformer is derived by replacing the covariance matrix in (\ref{eq:w}) with the beamspace covariance matrix 
+The beamspace version of the Capon beamformer is derived by replacing the covariance matrix in (\ref{II_eq:w}) with the beamspace covariance matrix 
 \begin{align}
@@
 \end{align}
-where $\mat{B} \in \mathbb{C}^{N_b,L}$, $N_b$ is the number of selected beams, and $f(\cdot)$ represent the covariance estimation process as defined in (\ref{eq:diag}).
-
-In the remaining sections, the methods described in this section and Section \ref{sec:es-capon} will be referred to as beamspace Capon (BS-Capon) and element-space Capon (ES-Capon) respectively.
+where $\mat{B} \in \mathbb{C}^{N_b,L}$, $N_b$ is the number of selected beams, and $f(\cdot)$ represent the covariance estimation process as defined in (\ref{II_eq:diag}).
+
+In the remaining sections, the methods described in this section and Section \ref{II_sec:es-capon} will be referred to as beamspace Capon (BS-Capon) and element-space Capon (ES-Capon) respectively.
 The name Capon will refer to both methods.
@@
 In CUDA programming a kernel function is executed over a large set of threads, where the kernel function describes the work to be done in each thread.
-As depicted in Fig.\ \ref{fig:gpulayout}, threads are organized in a grid of thread-blocks, with maximal size of 1024 threads per block (e.g.
+As depicted in Fig.\ \ref{II_fig:gpulayout}, threads are organized in a grid of thread-blocks, with maximal size of 1024 threads per block (e.g.
 $32\times32$).
@@
 The task of these $L$ threads is to calculate the Capon beamformer output in the three depicted steps.}
-\label{fig:gpulayout}
+\label{II_fig:gpulayout}
 \end{figure}
@@
 
-\section{Parallel ES-Capon}\label{sec:meth}
+\section{Parallel ES-Capon}\label{II_sec:meth}
 
@@
 
-If the beamformer in (\ref{eq:z_mv}) is analyzed in combination with the matrix equation in (\ref{eq:b}) and the covariance estimation process in (\ref{eq:diag}), we see that the calculation of a set of Capon weights is independent across the outputs $z[n]$.
-As shown in Fig.\ \ref{fig:gpulayout}, a first level of granularity is therefore to divide the image into a grid, where the output in each cell can be calculated independently.
+If the beamformer in (\ref{II_eq:z_mv}) is analyzed in combination with the matrix equation in (\ref{II_eq:b}) and the covariance estimation process in (\ref{II_eq:diag}), we see that the calculation of a set of Capon weights is independent across the outputs $z[n]$.
+As shown in Fig.\ \ref{II_fig:gpulayout}, a first level of granularity is therefore to divide the image into a grid, where the output in each cell can be calculated independently.
 Further, based on memory dependencies and to hold memory consumption per thread at a minimum, a group of threads should process one output in the final image.
@@
 %\begin{array}{l}
-\text{1) } &\text{Calculate the sample covariance matrix } \mat{\hat{R}} \text{ as in } \text{(\ref{eq:diag})}.\\
-\text{2) } &\text{Solve the linear system of equations as in (\ref{eq:b}}).\\ %\vec{b} =\mat{\hat{R}}^{-1}\vec{1}_L\\
-\text{3) } &\text{Calculate } \vec{w} = \vec{b}/\vec{1}_{L}^H\vec{b} \text{ and the amplitude as in (\ref{eq:z_mv})}.
+\text{1) } &\text{Calculate the sample covariance matrix } \mat{\hat{R}} \text{ as in } \text{(\ref{II_eq:diag})}.\\
+\text{2) } &\text{Solve the linear system of equations as in (\ref{II_eq:b}}).\\ %\vec{b} =\mat{\hat{R}}^{-1}\vec{1}_L\\
+\text{3) } &\text{Calculate } \vec{w} = \vec{b}/\vec{1}_{L}^H\vec{b} \text{ and the amplitude as in (\ref{II_eq:z_mv})}.
 %\end{array}
@@
 %\begin{enumerate}
-%\item Calculate the sample covariance matrix $\mat{\hat{R}}$ as in (\ref{eq:R})
+%\item Calculate the sample covariance matrix $\mat{\hat{R}}$ as in (\ref{II_eq:R})
 %\end{enumerate}
@@
 %\caption{}
-%\label{fig:gpulayout}
+%\label{II_fig:gpulayout}
 %\end{figure}
 
-\subsection{Calculation of Multiple Sample Covariance Matrices}\label{sec:calcR}
-
-%The straightforward sum in (\ref{eq:R}) has complexity $O(L^2N_LN_K) = O(L^3)$ since $N_L$, the number of subarrays, is a function of $L$.
+\subsection{Calculation of Multiple Sample Covariance Matrices}\label{II_sec:calcR}
+
+%The straightforward sum in (\ref{II_eq:R}) has complexity $O(L^2N_LN_K) = O(L^3)$ since $N_L$, the number of subarrays, is a function of $L$.
 This is actually the same as for inverting an $L \times L$ matrix.
@@
 
-Calculation of one element of \R is from (\ref{eq:R}) found to be independent of calculation of the other elements.
+Calculation of one element of \R is from (\ref{II_eq:R}) found to be independent of calculation of the other elements.
 A natural level of granularity would then be to assign one thread to each element of $\mat{\hat{R}}$ \cite{Chen2011}.
@@
 In \cite{Asen2012, Buskenes} we demonstrated how the cubic complexity with $L$ %and linear complexity with $N_K$ 
-can be reduced by realizing that calculations in (\ref{eq:R}) overlap across subarrays.
+can be reduced by realizing that calculations in (\ref{II_eq:R}) overlap across subarrays.
 %Element $(i,j)$ of $\mat{\breve{R}}[n]$ can be calculated from element $(i-1, j-1), \text{for } i,j \in [1, L-1]$ in a sliding window maner as
@@
 %&+ \sum_{n'=n-K}^{n+K} (x_{i+N_L-1}[n']x_{j+N_L-1}[n']^* \nonumber \\
-%&- x_{i-1}[n']x_{j-1}[n']^H). \label{eq:sliding}
+%&- x_{i-1}[n']x_{j-1}[n']^H). \label{II_eq:sliding}
 %\end{align}
@@
 For small values of $L$, several covariance matrices are calculated per thread block (Fig.
-\ref{fig:gpulayout}).
+\ref{II_fig:gpulayout}).
 Maximum supported $L$ is also increased to the maximum number of threads per block (currently 1024).
@@
 
-%For the adaptive diagonal loading factor in (\ref{eq:diag}), the trace of $\mat{\hat{R}}$ has to be known before diagonal elements are written back to global memory.
+%For the adaptive diagonal loading factor in (\ref{II_eq:diag}), the trace of $\mat{\hat{R}}$ has to be known before diagonal elements are written back to global memory.
 One row for all covariance matrices is written to global memory per kernel iteration, the diagonal elements therefore need to reside in shared memory until the trace is accumulated.
@@
 \subsection{Compute Beamformer Output}
-%The calculation of the beamformer output in (\ref{eq:z_mv}) has been implemented using one thread per element weight ($L$ threads per weight vector).
+%The calculation of the beamformer output in (\ref{II_eq:z_mv}) has been implemented using one thread per element weight ($L$ threads per weight vector).
 After the solution vector $\vec{b}$ has been placed in shared memory, one of the $L$ threads finds the sum of $\vec{b}$.
@@
 %Each of the $L$ threads then reads one element of $\vec{b}$, and scales it with the inverse sum to get the weight vector $\vec{w}[n]$. 
-The final output can be formed by either reducing the data down to length $L$ as in (\ref{eq:z_mv}), or increasing the weight vector to length $M$ as in (\ref{eq:z_mv2}).
+The final output can be formed by either reducing the data down to length $L$ as in (\ref{II_eq:z_mv}), or increasing the weight vector to length $M$ as in (\ref{II_eq:z_mv2}).
 Since we selected to use $L$ threads per output, the data is reduced to length $L$ by letting thread $r$ calculate
 %
-\begin{align}\label{eq:subsum}
+\begin{align}\label{II_eq:subsum}
 \breve{x}_{r} = \sum_{k=r}^{r+N_L-1}x_{k} \text{ for } r \in [0, L-1].
@@
 In this example $N_b=3$, $l$ is the current subarray and $k$ is the current beam.}
-\label{fig:gpulayoutbs}
+\label{II_fig:gpulayoutbs}
 \end{figure}
 
-\section{Parallel BS-Capon}\label{sec:bs}
-
-In this section we will derive a parallel implementation of the BS-Capon beamformer, and comment on how it deviates from the implementation in Section \ref{sec:meth}.
-The three main steps described in Section \ref{sec:meth} are present here as well, but in addition we need to transform the per-element data to beamspace.
+\section{Parallel BS-Capon}\label{II_sec:bs}
+
+In this section we will derive a parallel implementation of the BS-Capon beamformer, and comment on how it deviates from the implementation in Section \ref{II_sec:meth}.
+The three main steps described in Section \ref{II_sec:meth} are present here as well, but in addition we need to transform the per-element data to beamspace.
 This can be done using a FFT, but for few beams ($N_b < \log{L}$) a matrix multiply with the truncated Butler matrix is less costly.
@@
 %
-The advantage of this approach is that minor changes have to be made to the pipeline described in Section \ref{sec:meth}, but it will only speed up the solving step if the solution $\vec{b}_{BS}$ is transformed back to element space.
+The advantage of this approach is that minor changes have to be made to the pipeline described in Section \ref{II_sec:meth}, but it will only speed up the solving step if the solution $\vec{b}_{BS}$ is transformed back to element space.
 In this paper we will focus on the first approach, and how it has been implemented on the GPU.
 %By this approach we go from solving $\mat{\hat{R}}\vec{b} = \vec{1}_L$, which is an $L \times L$ system, to solving the $N_b \times N_b$ system $\mat{\hat{R}}_{BS}\vec{b}_{BS} = (\sqrt{L}\vec{e}_1)$.
-Now, one can either choose to transform the subarray sums, as in (\ref{eq:subsum}), to beamspace or transform the result, $\vec{b}_{BS}$, back to element space.
+Now, one can either choose to transform the subarray sums, as in (\ref{II_eq:subsum}), to beamspace or transform the result, $\vec{b}_{BS}$, back to element space.
 In both cases solving has to be performed against a beamspace steering vector
@@
 The overlap between subarrays that were exploited when estimating the covariance matrix in element space, can now be utilized when calculating the beamspace transform. 
-Given a $M$ long data vector divided into $N_L$ subarrays, the beamspace transformation of the $l$th subarray for the $k$th beam is from (\ref{eq:butler}) found to be:
-\begin{align}\label{eq:bs_formula}
+Given a $M$ long data vector divided into $N_L$ subarrays, the beamspace transformation of the $l$th subarray for the $k$th beam is from (\ref{II_eq:butler}) found to be:
+\begin{align}\label{II_eq:bs_formula}
 x_{\text{BS},l,k} &= \mat{B}_k\vec{x}_l \,\,\,\,\,\,\,\,\, k \in [0, N_b-1] \\
@@
 where $\mat{B}_k$ is the $k$th row of $\mat{B}$. %Manipulating this expression we find that for $l \in [1, L-1]$
-%\begin{align}\label{eq:sliding_bs}
+%\begin{align}\label{II_eq:sliding_bs}
 %x_{\text{BS},l,k} &= \frac{1}{\sqrt{L}}((x_{\text{BS},l-1,k} - x_{l-1})e^{-j2\pi k/L} \nonumber \\ &+ x_{i+L-1}e^{-j2\pi k(L-1)/L}) \nonumber \\
@@
 %We denote this the sliding-window beamspace transformation (SBS), and it is equal to a normalized sliding DFT \cite{Lyons2003}.
-For $l=0$ a full computation has to be performed using (\ref{eq:bs_formula}).
+For $l=0$ a full computation has to be performed using (\ref{II_eq:bs_formula}).
 This formula has been implemented using a normalized sliding DFT (SDFT)\cite{Lyons2003}.
-%Equation (\ref{eq:sliding_bs}) shows that 
+%Equation (\ref{II_eq:sliding_bs}) shows that 
 The SDFT allows us to still save instructions by utilizing the overlap between subarrays, but with an increase in task level granularity.
@@
 Using one thread per beam is therefore not feasible for small-$N_b$-large-$M$ configurations.
-One solution, as depicted in Fig.\ \ref{fig:gpulayoutbs}, is to combine SDFT with full computations.
+One solution, as depicted in Fig.\ \ref{II_fig:gpulayoutbs}, is to combine SDFT with full computations.
 A number of sub thread groups of size $N_b$ is created that targets different sets of subarrays which they transform using the SDFT algorithm.
@@
 %
-As presented in (\ref{eq:diag}), adaptive diagonal loading is performed by loading the sample covariance matrix with the average energy per element, that is $\text{trace}\{\mat{\breve{R}}\}/L$.
+As presented in (\ref{II_eq:diag}), adaptive diagonal loading is performed by loading the sample covariance matrix with the average energy per element, that is $\text{trace}\{\mat{\breve{R}}\}/L$.
 When data are transformed into beamspace, a high percentage of the total energy is mapped to a small subset of beams.
@@
 The added complexity of SDFT is $O(N_bL)$.
-To maximize performance, large adjustments to the processing pipeline described in Section \ref{sec:meth} is required.
+To maximize performance, large adjustments to the processing pipeline described in Section \ref{II_sec:meth} is required.
 
@@
 
-\section{Benchmarks}\label{sec:bench}
+\section{Benchmarks}\label{II_sec:bench}
 
@@
 \centerline{
-\subfloat[]{\includegraphics[width=3.4in]{gfx/benchmark_bar_64_80_540.png} \label{fig:benchUS64}}
+\subfloat[]{\includegraphics[width=3.4in]{gfx/benchmark_bar_64_80_540.png} \label{II_fig:benchUS64}}
 \hfil
 \subfloat[]{\includegraphics[width=3.4in]{gfx/benchmark_bar_96_120_540.png}%
-\label{fig:benchUS96}}}
+\label{II_fig:benchUS96}}}
 %\centerline{\includegraphics[width=3.5in]{gfx/benchmark_bar_64_80_540.png}}
@@
 Execution time is plotted as a function of subarray length $L$.
-The figures show execution times for the three steps listed in Section \ref{sec:meth}: Calculating covariance matrices with three different choices of temporal averaging, solver, and the final beamformer sum.}
-\label{fig:bench}
+The figures show execution times for the three steps listed in Section \ref{II_sec:meth}: Calculating covariance matrices with three different choices of temporal averaging, solver, and the final beamformer sum.}
+\label{II_fig:bench}
 \end{figure*}
@@
   \begin{figure*}[!t]
-  \centerline{\subfloat[]{\includegraphics[width=3.8in]{gfx/benchmark_bar_bs_M=64_Nx=80_Ny=540_Nb=3.png} \label{fig:benchBS64}}
+  \centerline{\subfloat[]{\includegraphics[width=3.8in]{gfx/benchmark_bar_bs_M=64_Nx=80_Ny=540_Nb=3.png} \label{II_fig:benchBS64}}
 %   \hfil
   \subfloat[]{\includegraphics[width=3.8in]{gfx/benchmark_bar_bs_M=96_Nx=120_Ny=540_Nb=3.png}%
-  \label{fig:benchBS96}}}
+  \label{II_fig:benchBS96}}}
 \else
   \begin{figure*}[!t]
-  \centerline{\subfloat[]{\includegraphics[width=3.4in]{gfx/benchmark_bar_bs_M=64_Nx=80_Ny=540_Nb=3.png} \label{fig:benchBS64}}
+  \centerline{\subfloat[]{\includegraphics[width=3.4in]{gfx/benchmark_bar_bs_M=64_Nx=80_Ny=540_Nb=3.png} \label{II_fig:benchBS64}}
   \hfil
   \subfloat[]{\includegraphics[width=3.4in]{gfx/benchmark_bar_bs_M=96_Nx=120_Ny=540_Nb=3.png}%
-  \label{fig:benchBS96}}}
+  \label{II_fig:benchBS96}}}
 \fi%
@@
 Execution time is plotted as a function of subarray length $L$.
-If $L$ had been fixed at $M/2$ and execution time was plotted against $N_b$, the solver would have shown the same quadratic growth as in Fig.\ \ref{fig:bench}.
+If $L$ had been fixed at $M/2$ and execution time was plotted against $N_b$, the solver would have shown the same quadratic growth as in Fig.\ \ref{II_fig:bench}.
 When $N_b$ is fixed, calculating the covariance matrix and the BS-transform is inverse proportional and proportional to $L$ respectively.
 The large jumps in execution times are caused by adaptive kernel launch configurations, which depends on $M$, $L$ and $N_b$.}
-\label{fig:benchBS}
+\label{II_fig:benchBS}
 \end{figure*}
 %
-In this section we present benchmarks of the implementations presented in Sections \ref{sec:meth} and \ref{sec:bs} performed on a Nvidia Quadro 6000 graphics card (14 SMs, 1 TFlops theoretical single precision performance, 6 GB global memory, and 144 GB/s memory bandwidth).
+In this section we present benchmarks of the implementations presented in Sections \ref{II_sec:meth} and \ref{II_sec:bs} performed on a Nvidia Quadro 6000 graphics card (14 SMs, 1 TFlops theoretical single precision performance, 6 GB global memory, and 144 GB/s memory bandwidth).
 Problem sizes are calculated based on 64 and 96 element phased arrays with a 3.4 MHz center frequency.
@@
 %
-\begin{equation}\label{eq:rayleigh}
+\begin{equation}\label{II_eq:rayleigh}
 \delta{}\theta = \lambda/(2Mp),
@@
 
-Benchmarks of the ES-Capon beamformer are presented in Fig.\ \ref{fig:bench}.
-For each subarray size, the total execution time is split into one bar for each of the three steps described in Section \ref{sec:meth}, and the calculation of $\mat{\hat{R}}$ is presented for three different values of $K$.
+Benchmarks of the ES-Capon beamformer are presented in Fig.\ \ref{II_fig:bench}.
+For each subarray size, the total execution time is split into one bar for each of the three steps described in Section \ref{II_sec:meth}, and the calculation of $\mat{\hat{R}}$ is presented for three different values of $K$.
 For $M=64$, $L=32$ and $K=2$, a typical setup for high-quality cardiac imaging, processing throughput is 8.3 frames per second (FPS).
@@
 
-In Fig.\ \ref{fig:benchBS} we present similar benchmarks for the BS-Capon beamformer.
+In Fig.\ \ref{II_fig:benchBS} we present similar benchmarks for the BS-Capon beamformer.
 The figure only presents results with use of temporal averaging ($K=2$).
@@
 
-\section{In-vivo cardiac images}\label{sec:images}
+\section{In-vivo cardiac images}\label{II_sec:images}
 
@@
 %\centerline{
-%\subfloat[]{ \includegraphics[width=2.2in]{gfx/das.png} \label{fig:phantomDAS} }
+%\subfloat[]{ \includegraphics[width=2.2in]{gfx/das.png} \label{II_fig:phantomDAS} }
 %\hfil
-%\subfloat[]{ \includegraphics[width=2.2in]{gfx/capon.png} \label{fig:phantomMV} }
+%\subfloat[]{ \includegraphics[width=2.2in]{gfx/capon.png} \label{II_fig:phantomMV} }
 %\hfil
 %\subfloat[]{ 
-%\includegraphics[width=2.2in]{gfx/capon_bs.png} \label{fig:phantomBS} 
+%\includegraphics[width=2.2in]{gfx/capon_bs.png} \label{II_fig:phantomBS} 
 %\hfil
@@
 c) Beamspace Capon ($L=32, K=2, N_b=3, d=0.01$) \multimedia{Media-Movie 2}.}
-%\label{fig:phantom}
+%\label{II_fig:phantom}
 %\end{figure*}
@@
 \centerline{
-\subfloat[]{\includegraphics[width=2.2in]{gfx/das_invivo.png} \label{fig:invivoDAS}}
+\subfloat[]{\includegraphics[width=2.2in]{gfx/das_invivo.png} \label{II_fig:invivoDAS}}
 \hfil
-\subfloat[]{\includegraphics[width=2.2in]{gfx/capon_invivo.png} \label{fig:invivoMV}}
+\subfloat[]{\includegraphics[width=2.2in]{gfx/capon_invivo.png} \label{II_fig:invivoMV}}
 \hfil
 \subfloat[]{
-\includegraphics[width=2.2in]{gfx/capon_bs_invivo.png} \label{fig:invivoBS}}
+\includegraphics[width=2.2in]{gfx/capon_bs_invivo.png} \label{II_fig:invivoBS}}
 \hfil
@@
 c) Beamspace Capon ($L=32, K=2, N_b=3, d=0.01$) \multimedia{Media-Movie 3}.}
-\label{fig:invivo}
+\label{II_fig:invivo}
 \end{figure*}
@@
 %\centerline{
-%\subfloat[]{\includegraphics[width=3in]{gfx/simulation_slice.png} \label{fig:simulationSlice}}
+%\subfloat[]{\includegraphics[width=3in]{gfx/simulation_slice.png} \label{II_fig:simulationSlice}}
 %\hfil
-%\subfloat[]{\includegraphics[width=3in]{gfx/invivo_slice.png} \label{fig:invivoSlice}}
+%\subfloat[]{\includegraphics[width=3in]{gfx/invivo_slice.png} \label{II_fig:invivoSlice}}
 %}
 %\caption{Lateral slices taken from the images in Fig.
-\ref{fig:phantom} and Fig.
-\ref{fig:invivo}.
+\ref{II_fig:phantom} and Fig.
+\ref{II_fig:invivo}.
 a) From Fig.
-\ref{fig:phantom} at 93 mm constant depth, through the two bright point scatterers.
+\ref{II_fig:phantom} at 93 mm constant depth, through the two bright point scatterers.
 b) From Fig.
-\ref{fig:invivo} at 96 mm range, through the tip of the mitral valve.}
-%\label{fig:slices}
+\ref{II_fig:invivo} at 96 mm range, through the tip of the mitral valve.}
+%\label{II_fig:slices}
 %\end{figure*}
@@
 %To investigate the image quality of both the ES and BS-Capon beamformer, a phantom was simulated using Field II \cite{Jensen1992, Jensen1996a} (Fig.
-\ref{fig:phantom}).
+\ref{II_fig:phantom}).
 The simulation was performed using a 64-element, $p=\lambda/2$ pitch, 10 $\mu$m kerf, 80\% bandwidth, and 10 mm high phased array.
@@
 
-Fig.\ \ref{fig:invivo} presents results of applying ES- and BS-Capon on an \textit{in-vivo} image of the left ventricle.
+Fig.\ \ref{II_fig:invivo} presents results of applying ES- and BS-Capon on an \textit{in-vivo} image of the left ventricle.
 The image was acquired using a 64-element and $p=0.65\lambda$ pitched phased array in harmonic mode (1.7 MHz at transmit and 3.4 MHz at receive).
 The image frame is extracted from a cardiac cycle, that can be viewed in its full length in the attached videos \multimedia{Media-Movie 1, 2 and 3}.
-The region of interest is 70\degree{} with a minimal lateral sampling of 98 receive beams given by (\ref{eq:rayleigh}).
+The region of interest is 70\degree{} with a minimal lateral sampling of 98 receive beams given by (\ref{II_eq:rayleigh}).
 The image was acquired using 50 transmit beams and "Synthetic Transmit Beamforming" was used on receive \cite{Hergum2007}.
@@
 %To get a better impression of the improved lateral resolution obtained with Capon beamforming, two slices taken from the images in Fig.
-\ref{fig:phantom} and \ref{fig:invivo} are included in Fig.
-\ref{fig:slices}.
+\ref{II_fig:phantom} and \ref{II_fig:invivo} are included in Fig.
+\ref{II_fig:slices}.
 For the phantom, the cut goes through the two point targets at 90 mm constant radius, and for the cardiac image it goes through the tip of the mitral valve at 96 mm constant radius.
@@
 %When the simulated phantom was processed with the ES-Capon beamformer, modulation effects were observed when the two bright point scatterers moved across receive-beams.
-The artifact did not disappear until 16x lateral oversampling relative to (\ref{eq:rayleigh}) was used on transmit.
-%The cardiac recording shows a 70\degree{} sector with a minimal lateral sampling of 98 receive beams given by (\ref{eq:rayleigh}). %Despite this, the same artifacts were not observed here. 
+The artifact did not disappear until 16x lateral oversampling relative to (\ref{II_eq:rayleigh}) was used on transmit.
+%The cardiac recording shows a 70\degree{} sector with a minimal lateral sampling of 98 receive beams given by (\ref{II_eq:rayleigh}). %Despite this, the same artifacts were not observed here. 
 %The image was acquired using 50 transmit beams and "Synthetic Transmit Beamforming" was used on receive \cite{Hergum2007}.
 
-\section{Trading resolution for speed}\label{sec:trade}
-As seen from Fig.\ \ref{fig:bench} the execution time, for the solver in particular, is reduced when $L$ is reduced.
+\section{Trading resolution for speed}\label{II_sec:trade}
+As seen from Fig.\ \ref{II_fig:bench} the execution time, for the solver in particular, is reduced when $L$ is reduced.
 The same is true when $N_b$ is reduced in beamspace.
@@
 
-In Fig.\ \ref{fig:speed_res_trade}, the minimum distance between two point targets (while preserving a 6 dB saddle point in between the two) is plotted against $L$ (when processed with ES-Capon) and $N_b$ (when processed with BS-Capon).
-To produce the result in Fig.\ \ref{fig:speed_res_trade} $K=2$ and $d=0.01$ has been used on both methods in order to obtain the same resolution for $L=N_b=M/2$.
-In Fig.\ \ref{fig:invivo}, the diagonal loading factor for BS-Capon was adjusted to make the two results equal for $L=32$ and $N_b=3$.
+In Fig.\ \ref{II_fig:speed_res_trade}, the minimum distance between two point targets (while preserving a 6 dB saddle point in between the two) is plotted against $L$ (when processed with ES-Capon) and $N_b$ (when processed with BS-Capon).
+To produce the result in Fig.\ \ref{II_fig:speed_res_trade} $K=2$ and $d=0.01$ has been used on both methods in order to obtain the same resolution for $L=N_b=M/2$.
+In Fig.\ \ref{II_fig:invivo}, the diagonal loading factor for BS-Capon was adjusted to make the two results equal for $L=32$ and $N_b=3$.
 %The resolution for ES-Capon is therefore better here than what is presented in Fig.
-\ref{fig:phantom}.
+\ref{II_fig:phantom}.
 %We considered this to be a small value for ES-Capon and a normal value for BS-Capon (cf.
-Section \ref{sec:bs}), hence the resolution for ES-Capon is here better than what is presented in Fig.
-\ref{fig:phantom}.
+Section \ref{II_sec:bs}), hence the resolution for ES-Capon is here better than what is presented in Fig.
+\ref{II_fig:phantom}.
 
@@
 We see that the resolution for BS-Capon gradually decreases due to an automatic increase in diagonal loading (cf.
-end of Section \ref{sec:bs}).
+end of Section \ref{II_sec:bs}).
 If we compensate by reducing $d$, BS-Capon will provide resolution equal to ES-Capon (with $L=M/2$) for all values of $N_b$ except one \cite{Nilsen2009}.
@@
 \caption{Trade-off between resolution and speed by reducing subarray size or reducing beamspace dimension.
-As $L$ or $N_b$ gets smaller, the execution time is reduced (Fig.\ \ref{fig:bench} and \ref{fig:benchBS}).
+As $L$ or $N_b$ gets smaller, the execution time is reduced (Fig.\ \ref{II_fig:bench} and \ref{II_fig:benchBS}).
 When $N_b$ is varied, $L$ is fixed on 32.}
-\label{fig:speed_res_trade}
+\label{II_fig:speed_res_trade}
 \end{figure}
 
-\section{Discussion}\label{sec:dis}
+\section{Discussion}\label{II_sec:dis}
 In this paper we have presented a parallel implementation of the BS-Capon beamformer for high-resolution cardiac ultrasound imaging.
 The results show that it is now possible to apply BS-Capon in real-time in cardiac ultrasound imaging by utilizing the power of a GPU.
-According to Fig.\ \ref{fig:benchBS} it is also possible to process two parallel receive lines per transmit beam (for both $M=64$ and $M=96$.
+According to Fig.\ \ref{II_fig:benchBS} it is also possible to process two parallel receive lines per transmit beam (for both $M=64$ and $M=96$.
 $L=M/2$).
@@
 For ES-Capon, processing rates for high quality settings are still too high to support real-time cardiac ultrasound imaging.
-From Fig.\ \ref{fig:benchUS64} we see that for $L = M/2 = 32$ and $K=1$, the total execution time is 110 ms (9 FPS).
+From Fig.\ \ref{II_fig:benchUS64} we see that for $L = M/2 = 32$ and $K=1$, the total execution time is 110 ms (9 FPS).
 Taking into account that we use complex data, but less samples per image, these results are similar to the performance reported by Chen et al.
@@
 A major part of the total execution time for ES-Capon comes from solving the system of linear equations.
-This limitation is targeted by reducing the sample covariance matrix size in beamspace, and as shown in Fig.\ \ref{fig:benchBS}, inverting the $3\times3$ beamspace covariance matrix is then barely visible.
+This limitation is targeted by reducing the sample covariance matrix size in beamspace, and as shown in Fig.\ \ref{II_fig:benchBS}, inverting the $3\times3$ beamspace covariance matrix is then barely visible.
 The time it takes to construct all covariance matrices is also reduced, but for few beams this is now the major contributor to the total execution time.
-If real-time is defined as 10 FPS \cite{Chen2011} almost all execution times in both Fig.\ \ref{fig:benchUS64} and \ref{fig:benchBS} are real-time.
+If real-time is defined as 10 FPS \cite{Chen2011} almost all execution times in both Fig.\ \ref{II_fig:benchUS64} and \ref{II_fig:benchBS} are real-time.
 More natural for cardiac ultrasound imaging is to define real-time as the time it takes to acquire the underlying image.
@@
 %Turning to image quality we see from Fig.
-\ref{fig:phantom}, and the attached video of the simulation, that Capon beamforming provides improved lateral resolution of moving point targets both in element- and in a reduced beam-space ($N_b=3$).
+\ref{II_fig:phantom}, and the attached video of the simulation, that Capon beamforming provides improved lateral resolution of moving point targets both in element- and in a reduced beam-space ($N_b=3$).
 Our preliminary finding is that the Capon beamformers are local shift invariant as long as the lateral sampling is increased to a certain level in simulations \multimedia{Media-Movie 1 and 2}.
@@
 
-Turning to image quality we see from Fig.\ \ref{fig:speed_res_trade} and the attached videos, that Capon beamforming provides improved lateral resolution of moving point targets both in element- and in a reduced beam-space ($N_b=3$).
-In the cardiac recording (Fig.\ \ref{fig:invivo}), structures are thinner and better resolved when the Capon beamformers are applied.
+Turning to image quality we see from Fig.\ \ref{II_fig:speed_res_trade} and the attached videos, that Capon beamforming provides improved lateral resolution of moving point targets both in element- and in a reduced beam-space ($N_b=3$).
+In the cardiac recording (Fig.\ \ref{II_fig:invivo}), structures are thinner and better resolved when the Capon beamformers are applied.
 One example is the mitral valve located at 10 cm depth and -1.5 cm width.
-On the other hand we observe a loss in contrast, especially in the septum (Fig.\ \ref{fig:invivo}, 6 cm depth, -2 cm width).
+On the other hand we observe a loss in contrast, especially in the septum (Fig.\ \ref{II_fig:invivo}, 6 cm depth, -2 cm width).
 This loss can be partially explained by the presence of highly correlated signals and mismatch between the assumed steering vector and signal.
-This is the same issue that made us use 16x oversampling when producing Fig.\ \ref{fig:speed_res_trade}.
+This is the same issue that made us use 16x oversampling when producing Fig.\ \ref{II_fig:speed_res_trade}.
 Future work should focus on how large this oversampling needs to be and how it should be conducted in order to still achieve real time processing.
@@
 
-\section{Conclusion}\label{sec:con}
+\section{Conclusion}\label{II_sec:con}
 In this paper a parallel formulation and implementation of the beamspace Capon beamformer have been presented, and
@@
 %\caption{Simulation Results}
-%\label{fig_sim}
+%\label{II_fig_sim}
 %\end{figure}
@@
 %\centerline{\subfloat[Case I]\includegraphics[width=2.5in]{subfigcase1}%
-%\label{fig_first_case}}
+%\label{II_fig_first_case}}
 %\hfil
 %\subfloat[Case II]{\includegraphics[width=2.5in]{subfigcase2}%
-%\label{fig_second_case}}}
+%\label{II_fig_second_case}}}
 %\caption{Simulation results}
-%\label{fig_sim}
+%\label{II_fig_sim}
 %\end{figure*}
@@
 %\caption{An Example of a Table}
-%\label{table_example}
+%\label{II_table_example}
 %\centering
@@
 % the document is modified later
-%\IEEEtriggeratref{8}
+%\IEEEtriggeratref{II_8}
 % The "triggered" command can be changed if desired:

### 2018-10-11 HowtoHowtoCapon/document.tex
Testing change of frame letters

@@
 
-This article is outlined as follows: In Section \ref{methods}, we introduce the concept of adaptive beamforming and provide details on the MVDR method.
-Then, in \ref{maptogpu}, we investigate the complexity issue, discuss means for reducing arithmetic complexity, and detail an implementation that makes efficient use of the GPU's parallel resources.
-The final design is assessed in Sections \ref{images_and_benchmarks} and \ref{discussion}, where we provide benchmarks, comparisons with similar CPU implementations, and measures of how efficiently our implementation makes use of the GPU's resources.
+This article is outlined as follows: In Section \ref{I_methods}, we introduce the concept of adaptive beamforming and provide details on the MVDR method.
+Then, in \ref{I_maptogpu}, we investigate the complexity issue, discuss means for reducing arithmetic complexity, and detail an implementation that makes efficient use of the GPU's parallel resources.
+The final design is assessed in Sections \ref{I_images_and_benchmarks} and \ref{I_discussion}, where we provide benchmarks, comparisons with similar CPU implementations, and measures of how efficiently our implementation makes use of the GPU's resources.
 
@@
 
-\section{Methods}\label{methods}
+\section{Methods}\label{I_methods}
 
@@
 Signal signature is first removed by matched filtering.
-Then, before summation, a suitable set of delays $\Delta$ and weights $w$ are applied to focus on a pixel of interest at angle and range ($\theta,n$).}\label{beamforming}
+Then, before summation, a suitable set of delays $\Delta$ and weights $w$ are applied to focus on a pixel of interest at angle and range ($\theta,n$).}\label{I_beamforming}
 \end{figure}
 Let the receiver be an $M$ element uniform linear array, and assume that the signature of the transmitted signal has been removed by a matched filter (Fig.
-\ref{beamforming}).
+\ref{I_beamforming}).
 Further assume that the array channels are digitally delayed to focus at a pixel with azimuth angle $\theta$ and range sample $n$, such that the delayed data from the $m$th channel can be expressed as $x_m[\theta,n]$.
@@
 \begin{align}
-z[n] = \w\H[n]\x[n] = \bmat{w_0[n]\\w_1[n]\\\vdots\\w_{M-1}[n]}^H \bmat{x_0[n]\\x_1[n]\\\vdots\\x_{M-1}[n]},\label{z}
+z[n] = \w\H[n]\x[n] = \bmat{w_0[n]\\w_1[n]\\\vdots\\w_{M-1}[n]}^H \bmat{x_0[n]\\x_1[n]\\\vdots\\x_{M-1}[n]},\label{I_z}
 \end{align}
@@
 \begin{gather}
-\vec w[n] = \frac{\Ri[n]\a}{\a\T\Ri[n]\a},\label{weights}
+\vec w[n] = \frac{\Ri[n]\a}{\a\T\Ri[n]\a},\label{I_weights}
 \end{gather}
 where $\a$ is a steering vector and $\R=E\{\x[n]\x\H[n]\}\in\mathbb{C}^{M,M}$ is the spatial covariance matrix for the full array.
-Since we pre-steer our data to every pixel in the image we simplify (\ref{weights}) by substituting $\a$ with a row vector $\1$ that represents broadside phase-steering.
+Since we pre-steer our data to every pixel in the image we simplify (\ref{I_weights}) by substituting $\a$ with a row vector $\1$ that represents broadside phase-steering.
 To estimate $\R$ we compute a sample covariance matrix $\eR$.
@@
 \begin{gather}
-\breve{\R}[n] =  \frac{1}{N_K N_L} \sumb{l=0}{M-L}\sumb{n'=n-K}{n+K} \x_l[n']\x_l\H[n'] \in\mathbb{C}^{L,L},\label{spatialR}
+\breve{\R}[n] =  \frac{1}{N_K N_L} \sumb{l=0}{M-L}\sumb{n'=n-K}{n+K} \x_l[n']\x_l\H[n'] \in\mathbb{C}^{L,L},\label{I_spatialR}
 \end{gather}
@@
 \begin{align}
-\eR[n] = \breve{\R}[n] + \I \frac{d}{L} \tr\{\breve{\R}[n]\},\label{finalR}
+\eR[n] = \breve{\R}[n] + \I \frac{d}{L} \tr\{\breve{\R}[n]\},\label{I_finalR}
 \end{align}
@@
 
-Note how subarray averaging led to a size reduction of $\eR$ from $\mathbb{C}^{M,M}$ to $\mathbb{C}^{L,L}$, and hence will produce an $L$-element weight set when substituted into (\ref{z}).
-This weight set is applied to all the subarrays, before computing the beamformer output as in (\ref{z}).
+Note how subarray averaging led to a size reduction of $\eR$ from $\mathbb{C}^{M,M}$ to $\mathbb{C}^{L,L}$, and hence will produce an $L$-element weight set when substituted into (\ref{I_z}).
+This weight set is applied to all the subarrays, before computing the beamformer output as in (\ref{I_z}).
 Or, equivalently, we may apply the weight set to the sum of all the subarrays by
@@
 \begin{align}
-z[n] = \w\H[n] \sumb{l=0}{M-L} \x_l[n].\label{finalZ}
+z[n] = \w\H[n] \sumb{l=0}{M-L} \x_l[n].\label{I_finalZ}
 \end{align}
@@
 2. the term $\eR^{-1}\1$ is found using a linear equation solver;\newline
-3. and the beamformer output $z$ is computed from (\ref{finalZ}), where $\w$ is found by substituting $\eR^{-1}\1$ into (\ref{weights}). } \label{mvdr_beamforming}
+3. and the beamformer output $z$ is computed from (\ref{I_finalZ}), where $\w$ is found by substituting $\eR^{-1}\1$ into (\ref{I_weights}). } \label{I_mvdr_beamforming}
 \end{figure}
-As summarized in Fig. \ref{mvdr_beamforming}, the MVDR method is applied to each pixel independently, by:
+As summarized in Fig. \ref{I_mvdr_beamforming}, the MVDR method is applied to each pixel independently, by:
 \begin{enumerate}
-\item computing the sample covariance matrix $\eR$ in (\ref{finalR});
-\item computing $\eRi\1$ in (\ref{weights});
-\item computing the beamformer output $z$ in (\ref{finalZ}).
+\item computing the sample covariance matrix $\eR$ in (\ref{I_finalR});
+\item computing $\eRi\1$ in (\ref{I_weights});
+\item computing the beamformer output $z$ in (\ref{I_finalZ}).
 \end{enumerate}
@@
 This might lead us to believe that the inversion step is by far the most complex.
-But if we implement spatial and temporal averaging as in (\ref{spatialR}), then computing $\eR$ is of O($N_K N_L L^2$) and the inversion of O($L^3$).
+But if we implement spatial and temporal averaging as in (\ref{I_spatialR}), then computing $\eR$ is of O($N_K N_L L^2$) and the inversion of O($L^3$).
 Computing $\eR$ is now the most complex operation whenever $N_LN_K>L$.
@@
 \caption{Per-pixel computational complexity of the steps in MVDR beamforming (before any optimizations).
-To avoid signal cancellation in an active sonar system we usually set $L<\frac{M}{2}$, in which region the computation of $\eR$ dominates in terms of arithmetic complexity, especially when performing temporal averaging.}\label{mvdr_complexity}
+To avoid signal cancellation in an active sonar system we usually set $L<\frac{M}{2}$, in which region the computation of $\eR$ dominates in terms of arithmetic complexity, especially when performing temporal averaging.}\label{I_mvdr_complexity}
 \end{figure}
-To visualize these relations, and include the effects of using complex numbers, we set up complexity formulas that account for the total number of arithmetic operations for each step in the MVDR process (see appendix \ref{mvdr_formulas}).
+To visualize these relations, and include the effects of using complex numbers, we set up complexity formulas that account for the total number of arithmetic operations for each step in the MVDR process (see appendix \ref{I_mvdr_formulas}).
 The only excluded operation was partial pivoting used in the inversion step, but this should contribute marginally to the end result.
 The entire range of possible subarray sizes from $L\in[1,M]$ was finally evaluated, with temporal averaging set to $K\in\{0,1,2\}$, and the number of channels set to $M\in\{8,16,32\}$.
-The results are shown in \Fig{mvdr_complexity}.
+The results are shown in \Fig{I_mvdr_complexity}.
 
@@
 
-\section{Mapping the MVDR to a GPU}\label{maptogpu}
+\section{Mapping the MVDR to a GPU}\label{I_maptogpu}
 
@@
 To use Nvidia's own terminology, the Quadro 6000 comprises 14 streaming multiprocessors (SMs), each having 32 CUDA cores that execute a common program called a kernel.
-Combined these cores deliver a peak performance of more than 1\;Tflop/s (appendix \ref{throughput}).
+Combined these cores deliver a peak performance of more than 1\;Tflop/s (appendix \ref{I_throughput}).
 In practice this performance is hard to obtain, but one can get fairly close by balancing the load evenly on all cores, and by trying to avoid that some cores are forced to idle due to a pending data transfer or thread synchronization.
@@
 Here $\R'_{l}$ is the sample covariance matrix for the $l$th subarray, and $\breve{\R}$ is the average of $N_K$ of these.
-Note that instead of performing the temporal sum last as here, one could take more temporal samples into consideration in the computation of each $r_{ij}$.}\label{mvdr_build_R}
+Note that instead of performing the temporal sum last as here, one could take more temporal samples into consideration in the computation of each $r_{ij}$.}\label{I_mvdr_build_R}
 \end{figure}
@@
 
-\subsection[Computing the Spatial Covariance Matrix]{Computing the Spatial Covariance Matrix, $\hat{\boldsymbol{R}}$}\label{computing_eR}
-
-% As Fig. \ref{mvdr_complexity} illustrates the computation of $\eR$
+\subsection[Computing the Spatial Covariance Matrix]{Computing the Spatial Covariance Matrix, $\hat{\boldsymbol{R}}$}\label{I_computing_eR}
+
+% As Fig. \ref{I_mvdr_complexity} illustrates the computation of $\eR$
 
 As observed in Fig.
-\ref{mvdr_complexity}, a direct computation of $\eR$ by implementing (\ref{spatialR}) and (\ref{finalR}) is the greatest computational burden.
+\ref{I_mvdr_complexity}, a direct computation of $\eR$ by implementing (\ref{I_spatialR}) and (\ref{I_finalR}) is the greatest computational burden.
 We target this issue by first minimizing arithmetic operations, then aim to get as close to the peak arithmetic throughput on the GPU as possible.
@@
 In Fig.
-\ref{mvdr_build_R}, we illustrate how to reduce the arithmetic complexity of building $\eR$ in a system with $M=5$ sensors, a subarray size of $L=3$, and temporal averaging is set to $K=1$.
+\ref{I_mvdr_build_R}, we illustrate how to reduce the arithmetic complexity of building $\eR$ in a system with $M=5$ sensors, a subarray size of $L=3$, and temporal averaging is set to $K=1$.
 Here the spatial sum is carried out before the temporal sum.
@@
 \caption{Arithmetic optimization of computing $\eR$: Relative reduction in arithmetic complexity compared to the initial implementation shown in Fig.
-\ref{mvdr_complexity} (higher is better) .
-Note how the arithmetic count is reduced by a factor 4-10 in the memory-optimized case, and by a factor 6-22 in the instruction-optimized case.}\label{mvdr_complexity_optimized}
+\ref{I_mvdr_complexity} (higher is better) .
+Note how the arithmetic count is reduced by a factor 4-10 in the memory-optimized case, and by a factor 6-22 in the instruction-optimized case.}\label{I_mvdr_complexity_optimized}
 \end{figure}
@@
 Fig.
-\ref{mvdr_complexity_optimized} compares the two solutions, and formulas may be found in Appendix \ref{mvdr_formulas}.
+\ref{I_mvdr_complexity_optimized} compares the two solutions, and formulas may be found in Appendix \ref{I_mvdr_formulas}.
 Here the complexity curves are relative to the reference implementation in Fig.
-\ref{mvdr_complexity}.
+\ref{I_mvdr_complexity}.
 Both solutions reduce the complexity considerably, and recomputing the multiplications does not make that much of a difference.
@@
 \\[.5\baselineskip]\textbf{Step~3}: The beamformer output $z$ is computed in a straightforward fashion using $L$ threads that first sum the subarrays and then apply the MVDR weighting function.
-A single thread finally sums all the channels up.}\label{mvdr_implementation}
+A single thread finally sums all the channels up.}\label{I_mvdr_implementation}
 \else
@@
 \emph{Step~3}: The beamformer output $z$ is computed in a straightforward fashion using $L$ threads that first sum the subarrays and then apply the MVDR weighting function.
-A single thread finally sums all the channels up.}\label{mvdr_implementation}
+A single thread finally sums all the channels up.}\label{I_mvdr_implementation}
 \fi%
@@
 Unfortunately, the GPU prefers kernels to be more arithmetically intensive than this.
-This can be inferred from Table \ref{throughputs}, %
+This can be inferred from Table \ref{I_throughputs}, %
 %
@@
 \end{tabular}
-\caption{Nvidia Quadro 6000: Memory throughput, $B_{\lowercase{\text{mem}}}$, compared to arithmetic throughput, $B_\text{arith}$.}\label{throughputs}
+\caption{Nvidia Quadro 6000: Memory throughput, $B_{\lowercase{\text{mem}}}$, compared to arithmetic throughput, $B_\text{arith}$.}\label{I_throughputs}
 \end{table}%
 %
-where the peak bandwidth of the three types of GPU memory are compared to the peak arithmetic throughput (see appendix \ref{throughput} for derivations).
+where the peak bandwidth of the three types of GPU memory are compared to the peak arithmetic throughput (see appendix \ref{I_throughput} for derivations).
 Global memory (RAM), in which all data must reside at some point, is at best only able to move one float for every 30 floats processed by the CUDA cores, and potentially much worse if care is not taken to use "coalesced" memory access patterns (that avoid memory bank conflicts).
@@
 In Fig.
-\ref{mvdr_implementation}, we illustrate how we get around these challenges.
+\ref{I_mvdr_implementation}, we illustrate how we get around these challenges.
 First we make each SM compute entire range lines of pixels, then load the subset of $\x$ that all these pixels depend upon into shared memory.
@@
 
-The beamformer output $z$ is computed on a per-pixel basis by first computing the MVDR weights $\w$, which are merely scaled versions of $\eRi\1$ (\ref{weights}).
-Then the sum in (\ref{finalZ}) is found by assigning a group of $L$ threads to respective elements of $x_l$, which then proceed to accumulate these for all $N_L$ subarrays.
+The beamformer output $z$ is computed on a per-pixel basis by first computing the MVDR weights $\w$, which are merely scaled versions of $\eRi\1$ (\ref{I_weights}).
+Then the sum in (\ref{I_finalZ}) is found by assigning a group of $L$ threads to respective elements of $x_l$, which then proceed to accumulate these for all $N_L$ subarrays.
 The resulting data vector is finally multiplied with the weight vector using the same threads, and then a single thread is used to sum these products to obtain $z$.
@@
 % \item Single thread per pixel.
-% \item Computation of $\eR$ is optimized for minumum number of instructions, as in \ref{mvdr_build_R}. Only upper half of $\eR$ is computed (see next point).
+% \item Computation of $\eR$ is optimized for minumum number of instructions, as in \ref{I_mvdr_build_R}. Only upper half of $\eR$ is computed (see next point).
 % \item Inversion step carried out by a custom inline Cholesky based solver that only requires the upper triangular of $\eR$.
@@
 
-The GPU design is summarized and compared to our CPU designs in Table \ref{implementation_summary}.
+The GPU design is summarized and compared to our CPU designs in Table \ref{I_implementation_summary}.
 %
@@
 \end{tabular}
-\caption{Implementation summary.}\label{implementation_summary}
+\caption{Implementation summary.}\label{I_implementation_summary}
 \end{table}
@@
 
-\section{Images and Benchmarks}\label{images_and_benchmarks}
+\section{Images and Benchmarks}\label{I_images_and_benchmarks}
 
@@
 The element size and spacing is 2.5\;$\lambda$ and the opening angle is 25$^\circ$.
-To produce the image shown in \Fig{holmengraa}, the sonar was operated in sidescan mode.
+To produce the image shown in \Fig{I_holmengraa}, the sonar was operated in sidescan mode.
 The studied object is the 1500-dwt oil tanker wreck Holmengraa.
@@
 The image was linearly upinterpolated by a factor 2 in azimuth making its total size 1.46 megapixels (MP).
-Note how MVDR improves edge definition and reduces noise in shadow regions.}\label{holmengraa}
+Note how MVDR improves edge definition and reduces noise in shadow regions.}\label{I_holmengraa}
 \end{figure}
@@
 The results were obtained by processing a 1\;MP image from the data from a 32 channel array, for all subarray sizes $L$, and for $K\in\{0,1,2\}$.
-Runtime measurements are shown in \Fig{benchmarks}.
-The runtimes of memory-only and arithmetic-only GPU kernels are depicted in \Fig{code_assessment}.
-An estimate of computation efficiency is presented in \Fig{code_assess_flops}.
+Runtime measurements are shown in \Fig{I_benchmarks}.
+The runtimes of memory-only and arithmetic-only GPU kernels are depicted in \Fig{I_code_assessment}.
+An estimate of computation efficiency is presented in \Fig{I_code_assess_flops}.
 We were also granted a test drive at the Boston HPC centre on a machine with an Intel Xeon E5-2670, 32\;GB RAM and an Nvidia K20.
-The results from this run are shown in \Fig{benchmarks_boston}.
+The results from this run are shown in \Fig{I_benchmarks_boston}.
 
@@
 Note the major speedup of building $\eR$ when compared to the complexity plot in Fig.
-\ref{mvdr_complexity}.
+\ref{I_mvdr_complexity}.
 \\[.5\baselineskip]\textbf{Bottom:} Compared to a reference MATLAB and single thread C implementation running on a CPU the GPU offered a speedup of 2-3 orders of magnitude, but these numbers are somewhat misleading.
-If the C implementation was properly optimized we expect the GPU to be no more than a factor 5-10 faster, even if its theoretical peak performance is ~20 times higher than that of the CPU.}\label{benchmarks}
+If the C implementation was properly optimized we expect the GPU to be no more than a factor 5-10 faster, even if its theoretical peak performance is ~20 times higher than that of the CPU.}\label{I_benchmarks}
 \else
@@
 Note the major speedup of building $\eR$ when compared to the complexity plot in Fig.
-\ref{mvdr_complexity}.\newline
+\ref{I_mvdr_complexity}.\newline
 \emph{Bottom:} Compared to a reference MATLAB and single thread C implementation running on a CPU the GPU offered a speedup of 2-3 orders of magnitude, but these numbers are somewhat misleading.
-If the C implementation was properly optimized we expect the GPU to be no more than a factor 5-10 faster, even if its theoretical peak performance is ~20 times higher than that of the CPU.}\label{benchmarks}
+If the C implementation was properly optimized we expect the GPU to be no more than a factor 5-10 faster, even if its theoretical peak performance is ~20 times higher than that of the CPU.}\label{I_benchmarks}
 \fi%
@@
 From this plot we can infer that the kernel building $\eR$ is memory bound, as the time the kernel spends performing memory transactions is higher than the corresponding time it spends carrying out arithmetical operations.
-Furthermore, when the total runtime is larger than the restricted kernels this can largely be attributed to latency, which we can see that building $\eR$ suffers from with the chosen parameters.}\label{code_assessment}
+Furthermore, when the total runtime is larger than the restricted kernels this can largely be attributed to latency, which we can see that building $\eR$ suffers from with the chosen parameters.}\label{I_code_assessment}
 \end{figure}
@@
 An estimate of the number of floating point operations per second (Flop/s), found by dividing the theoretical complexity curves by actual run-times.
-This is a crude measure as it does not include any other instructions than the actual arithmetic operations in the MVDR computation.}\label{code_assess_flops}
+This is a crude measure as it does not include any other instructions than the actual arithmetic operations in the MVDR computation.}\label{I_code_assess_flops}
 \end{figure}
@@
 The exact same scenario and code as in Fig.
-\ref{benchmarks} was used here.
-With no code alterations the performance was only improved marginally compared to running on the Quadro 6000.}\label{benchmarks_boston}
+\ref{I_benchmarks} was used here.
+With no code alterations the performance was only improved marginally compared to running on the Quadro 6000.}\label{I_benchmarks_boston}
 \end{figure}
 
-As seen in the runtime comparisons presented in \Fig{benchmarks}, the GPU method became two to three orders of magnitude faster than the C implementation we started out with.
+As seen in the runtime comparisons presented in \Fig{I_benchmarks}, the GPU method became two to three orders of magnitude faster than the C implementation we started out with.
 The remaining bottleneck in the final design is the inversion step, which is typically five times slower than the build step.
@@
 
-\section{Discussion}\label{discussion}
-
-In accordance with previous studies, \Fig{holmengraa} demonstrates the MVDR beamformer's ability to produce images with suppressed interference and improved detail resolution.
+\section{Discussion}\label{I_discussion}
+
+In accordance with previous studies, \Fig{I_holmengraa} demonstrates the MVDR beamformer's ability to produce images with suppressed interference and improved detail resolution.
 Compared to the DAS beamformer, the ship's edges appear sharper and the shadows less noisy.
@@
 
-As observed in \Fig{benchmarks}, the combination of minimizing arithmetic operations and implementing the MVDR on a GPU lead to a significant improvement in processing speed.
+As observed in \Fig{I_benchmarks}, the combination of minimizing arithmetic operations and implementing the MVDR on a GPU lead to a significant improvement in processing speed.
 Compared to a MATLAB and single thread C implementation, an order 2 and 3 speedup was achieved, respectively.
@@
 
-Note from \Fig{benchmarks} how the building of $\eR$ now only takes up a fraction of the total processing time, and recall that it was the other way around when we presented the theoretical complexity curves in \Fig{mvdr_complexity}.
+Note from \Fig{I_benchmarks} how the building of $\eR$ now only takes up a fraction of the total processing time, and recall that it was the other way around when we presented the theoretical complexity curves in \Fig{I_mvdr_complexity}.
 Also, the benchmark curves for building $\eR$ now seem linear.
@@
 In particular, the back substitution step involved in its computation is mostly sequential.
-This difference can be observed in \Fig{mvdr_complexity}.
+This difference can be observed in \Fig{I_mvdr_complexity}.
 Alternative solvers, such as one based on Cholesky decomposition that exploits the Hermitian property of $\eR$, can in theory reduce complexity by a factor two, but we question whether this result can be obtained in practice using a GPU.
@@
 To measure this we made two versions of the MVDR kernel: One that only performs arithmetic operations, and one that only performs memory operations.
-Then, we benchmarked these kernels and compared their runtimes to the total runtime (\Fig{code_assessment}).
+Then, we benchmarked these kernels and compared their runtimes to the total runtime (\Fig{I_code_assessment}).
 A first thing to note is how these kernels seem equally occupied performing memory and arithmetic operations.
@@
 
-Even with all our efforts we were unable to obtain processing rates higher than 40\;GFlop/s on average in the desirable range of subarray sizes (\Fig{mvdr_complexity}).
+Even with all our efforts we were unable to obtain processing rates higher than 40\;GFlop/s on average in the desirable range of subarray sizes (\Fig{I_mvdr_complexity}).
 This is mainly due to the inversion step, since for the most part we obtained more than 100\;GFlop/s when building $\eR$.
@@
 
-\section{Conclusion}\label{conclusion}
+\section{Conclusion}\label{I_conclusion}
 
@@
 
-\section{MVDR Complexity Formulas}\label{mvdr_formulas}
+\section{MVDR Complexity Formulas}\label{I_mvdr_formulas}
 
@@
 
-\emph{Building $\eR$}. The initial complexity of this step can be inferred directly from (\ref{spatialR}) and (\ref{finalR}):
+\emph{Building $\eR$}. The initial complexity of this step can be inferred directly from (\ref{I_spatialR}) and (\ref{I_finalR}):
 \begin{align}
@@
 where $O_\text{dload} = (2L-1)\Oa + \Om$ is the minor cost of performing diagonal loading.
-If we apply the optimization strategies discussed in section \ref{computing_eR} we can arrive at the following instead:
+If we apply the optimization strategies discussed in section \ref{I_computing_eR} we can arrive at the following instead:
 \begin{align}
@@
 
-\section{GPU Throughput}\label{throughput}
+\section{GPU Throughput}\label{I_throughput}
 
@@
 \text{B}_\text{arith} &= 2\cdot1148\;\text{Mflop/s/core}\cdot 32\;\text{cores/SM}\cdot 14\;\text{SMs}\nn
-&= 1.03\;\text{Tflop/s}.\label{flops}
+&= 1.03\;\text{Tflop/s}.\label{I_flops}
 \end{align}
@@
 \text{B}_\text{gmem} &= \frac{2\cdot 3\;\text{Gbit/s}\cdot384\;\text{bit}}{8\;\text{bit/byte}}\nn
-&= 144\;\text{GB/s}\ (36\;\text{Gfloats/s}).\label{bwglobal}
+&= 144\;\text{GB/s}\ (36\;\text{Gfloats/s}).\label{I_bwglobal}
 \end{align}
@@
 \text{B}_\text{smem} &= \frac{\frac{1148}{2}\;\text{Mbit/s}\cdot32\;\text{bit/bank}\cdot 32\;\text{banks/SM}\cdot 14\;\text{SMs}}{8\;\text{bit/byte}}\nn
-&\approx 1.03\;\text{TB/s}\ (257\;\text{Gfloats/s}).\label{bwshared}
+&\approx 1.03\;\text{TB/s}\ (257\;\text{Gfloats/s}).\label{I_bwshared}
 \end{align}
 The bandwidths are compared in Tab.
-\ref{throughputs}.
+\ref{I_throughputs}.
 Note that even when using shared memory at least 4 floating point operations must be carried out per float transferred to the CUDA cores, otherwise the algorithm will be memory bound and the peak arithmetic throughput can not be reached.

### 2018-10-11 LCAforActiveSonarImaging/document.tex
All compiling, but lots of work to be done
Testing change of frame letters

@@
+% !TeX root = III_lca
 \title{Low Complexity Adaptive Sonar Imaging}
@@
 
-This article is outlined as follows: In Section \ref{sec:beamforming} we offer a gentle introduction to beamforming.
-Then we move on to adaptive beamforming and describe the MVDR and LCA methods in \ref{sec:mvdr} and \ref{sec:lca}, respectively.
-We use LCA with the Kaiser window which we describe in \ref{sec:lca_kaiser_windows}, explain how we steer it in \ref{sec:lca_steering}, and by how much in \ref{sec:lca_steering_bounds}.
-Finally we add some remarks on using the trigonometric window instead of Kaiser in \ref{sec:lca_trigonometric}, and of of sampling considerations in \ref{sec:lca_oversampling}.
-Results and discussion are provided in Section \ref{sec:results_discussion}, where we study MVDR's windows to find suitable LCA ones in \ref{sec:results_lca_window_function}, assess the range of window types and steering in \ref{sec:results_window_parameters}, determine the number of windows needed in \ref{sec:results_database_size}, and add a note on computational complexity in \ref{sec:results_complexity}.
-Finally, Section \ref{sec:conclusion} provides a conclusion.
+This article is outlined as follows: In Section \ref{III_sec:beamforming} we offer a gentle introduction to beamforming.
+Then we move on to adaptive beamforming and describe the MVDR and LCA methods in \ref{III_sec:mvdr} and \ref{III_sec:lca}, respectively.
+We use LCA with the Kaiser window which we describe in \ref{III_sec:lca_kaiser_windows}, explain how we steer it in \ref{III_sec:lca_steering}, and by how much in \ref{III_sec:lca_steering_bounds}.
+Finally we add some remarks on using the trigonometric window instead of Kaiser in \ref{III_sec:lca_trigonometric}, and of of sampling considerations in \ref{III_sec:lca_oversampling}.
+Results and discussion are provided in Section \ref{III_sec:results_discussion}, where we study MVDR's windows to find suitable LCA ones in \ref{III_sec:results_lca_window_function}, assess the range of window types and steering in \ref{III_sec:results_window_parameters}, determine the number of windows needed in \ref{III_sec:results_database_size}, and add a note on computational complexity in \ref{III_sec:results_complexity}.
+Finally, Section \ref{III_sec:conclusion} provides a conclusion.
 
-\section{Receive Beamforming}\label{sec:beamforming}
+\section{Receive Beamforming}\label{III_sec:beamforming}
 
@@
 \begin{align}
-z[n] = \w\H[n]\x[n] = \bmat{w_0[n]\\w_1[n]\\\vdots\\w_{M-1}[n]}^H \bmat{x_0[n]\\x_1[n]\\\vdots\\x_{M-1}[n]},\label{eq:beamformer_output}
+z[n] = \w\H[n]\x[n] = \bmat{w_0[n]\\w_1[n]\\\vdots\\w_{M-1}[n]}^H \bmat{x_0[n]\\x_1[n]\\\vdots\\x_{M-1}[n]},\label{III_eq:beamformer_output}
 \end{align}
@@
 
-\section{MVDR}\label{sec:mvdr}
+\section{MVDR}\label{III_sec:mvdr}
 
@@
 \argmin{\w[n]}\, E\{\big|z[n]\big|^2\} &= \argmin{\w[n]}\, \w[n]\R[n]\w\H[n]\nn
-\text{subject to } \w^H[n]\a_\phi &= 1.\label{eq:mvdr_definition}
+\text{subject to } \w^H[n]\a_\phi &= 1.\label{III_eq:mvdr_definition}
 \end{align}
@@
 \begin{gather}
-\vec w[n] = \frac{\Ri[n]\a_\phi}{\a_\phi\T\Ri[n]\a_\phi}.\label{eq:mvdr_weights}
+\vec w[n] = \frac{\Ri[n]\a_\phi}{\a_\phi\T\Ri[n]\a_\phi}.\label{III_eq:mvdr_weights}
 \end{gather}
@@
 The angle $\phi_{\mathrm{3\,dB}}(\beta)$ is the amount of steering needed for the steered amplitude response to have a \protect\minus{}3\,dB crossing that is exactly half that of the unsteered window.
-With each window steered this way we expect the resolution gain to remain predictable and independent of $\beta$, and we also effectively constrain the white noise gain of the beamformer.}\label{windows_steering}
+With each window steered this way we expect the resolution gain to remain predictable and independent of $\beta$, and we also effectively constrain the white noise gain of the beamformer.}\label{III_windows_steering}
 \end{figure}
@@
 \caption{The submerged object used to test beamforming resolution is a 1\,m by 1\,m test cross attached to an anchor with a diameter of approximately 13\,cm.
-Source image curtesy of Bundeswehr Technical Center for Ships and Naval Weapons, Maritime Technology and Research (WTD 71).}\label{cross}
+Source image curtesy of Bundeswehr Technical Center for Ships and Naval Weapons, Maritime Technology and Research (WTD 71).}\label{III_cross}
 \end{figure}
@@
 Note how the responses are more or less symmetric, with very little steering in the shadow, moderate steering in speckle and steering within roughly 3\,dB in highlight.
-The phase varies most in the highlight region where we see the highest contrast.}\label{mvdr_selected_windows}
+The phase varies most in the highlight region where we see the highest contrast.}\label{III_mvdr_selected_windows}
 \else
@@
 Note how the responses are more or less symmetric, with very little steering in the shadow, moderate steering in speckle and steering within roughly 3\,dB in highlight.
-The phase varies most in the highlight region where we see the highest contrast.} }\label{mvdr_selected_windows}
+The phase varies most in the highlight region where we see the highest contrast.} }\label{III_mvdr_selected_windows}
 \fi%
@@
    \textbf{Upper bound $\beta$:} Observe that LCA gets better at suppressing sidelobes in the image as we increase the upper bounds for $\beta$.
-\Fig{mvdr_selected_windows} suggests that MVDR prefers windows with sidelobe levels lower than that of Kaiser with $\beta=2$, but here we observe further improvement going to $\beta=5$.
+\Fig{III_mvdr_selected_windows} suggests that MVDR prefers windows with sidelobe levels lower than that of Kaiser with $\beta=2$, but here we observe further improvement going to $\beta=5$.
    \\[.5\baselineskip]
@@
    \parbox[t]{.89\linewidth}{Observe that LCA gets better at suppressing sidelobes in the image as we increase the upper bounds for $\beta$.
-\Fig{mvdr_selected_windows} suggests that MVDR prefers windows with sidelobe levels lower than that of Kaiser with $\beta=2$, but here we observe further improvement going to $\beta=5$.}\newline
+\Fig{III_mvdr_selected_windows} suggests that MVDR prefers windows with sidelobe levels lower than that of Kaiser with $\beta=2$, but here we observe further improvement going to $\beta=5$.}\newline
    \emph{Upper bound $\phi$:}\hfill
@@
 \fi%
-\label{oversampling_mosaic_bounds}
+\label{III_oversampling_mosaic_bounds}
 \end{figure*}
@@
 On the sources where lateral interference is present it chooses the widest response with the best sidelobe suppression.]{\includegraphics[width=\textwidth]{gfx/fig5a.pdf}%%
-   \label{oversampling_mosaic_bounds_beta}}
+   \label{III_oversampling_mosaic_bounds_beta}}
    \hfil
@@
 On strong sources LCA selects windows that are steered away from the source.
-This is what improves the FWHM measurement in \Fig{oversampling_mosaic_bounds}.]{\includegraphics[width=\textwidth]{gfx/fig5b.pdf}%
-   \label{oversampling_mosaic_bounds_phi}}
+This is what improves the FWHM measurement in \Fig{III_oversampling_mosaic_bounds}.]{\includegraphics[width=\textwidth]{gfx/fig5b_phd.pdf}%
+   \label{III_oversampling_mosaic_bounds_phi}}
    \caption{Kaiser windows chosen for each pixel in the image of the resolution test cross.
-The underlying image is the one shown in \Fig{oversampling_mosaic_bounds}, with the same parameters.
+The underlying image is the one shown in \Fig{III_oversampling_mosaic_bounds}, with the same parameters.
 The location of the anchor, cut line and main scatter locations of the cross is marked in red.}
-   \label{fig_sim}
+   \label{III_fig_sim}
    \end{figure*}
@@
 On the sources where lateral interference is present it chooses the widest response with the best sidelobe suppression.]{\includegraphics[width=\textwidth]{gfx/buske5a_online.pdf}%%
-   \label{oversampling_mosaic_bounds_beta}}
+   \label{III_oversampling_mosaic_bounds_beta}}
    \hfil
@@
 On strong sources LCA selects windows that are steered away from the source.
-This is what improves the FWHM measurement in \Fig{oversampling_mosaic_bounds}.]{\includegraphics[width=\textwidth]{gfx/buske5b_online.pdf}%
-   \label{oversampling_mosaic_bounds_phi}}
+This is what improves the FWHM measurement in \Fig{III_oversampling_mosaic_bounds}.]{\includegraphics[width=\textwidth]{gfx/buske5b_online.pdf}%
+   \label{III_oversampling_mosaic_bounds_phi}}
    \caption{Kaiser windows chosen for each pixel in the image of the resolution test cross.
-The underlying image is the one shown in \Fig{oversampling_mosaic_bounds}, with the same parameters.
+The underlying image is the one shown in \Fig{III_oversampling_mosaic_bounds}, with the same parameters.
 The location of the anchor, cut line and main scatter locations of the cross is marked in red.}
-   \label{fig_sim}
+   \label{III_fig_sim}
    \end{figure*}
@@
 Adding more steering-variations improves both sidelobe suppression and resolution.
-However, using more than  $N_\beta=2$ window types and $N_\phi=3$ steering angles makes minimal difference.}\label{oversampling_mosaic}
+However, using more than  $N_\beta=2$ window types and $N_\phi=3$ steering angles makes minimal difference.}\label{III_oversampling_mosaic}
 \end{figure*}
@@
 Also note how LCA is insensitive to the window type; its performance is near identical whether it uses Kasier or trigonometric windows.]{\includegraphics[width=\textwidth]{gfx/fig7a.pdf}%
-   \label{beamformer_comparison_cross}}
+   \label{III_beamformer_comparison_cross}}
    \hfil
@@
 Compared to DAS the adaptive beamformers produce deeper shadows, sharper edges and a higher detail level.
-The adaptive methods produce nearly identical images.]{\includegraphics[width=\textwidth]{gfx/fig7b.pdf}%%
-   \label{beamformer_comparison_holmengraa}}
-   \caption{Comparing image quality of DAS, MVDR and LCA with Kaiser or trigonometric windows.}\label{image_quality}
+The adaptive methods produce nearly identical images.]{\includegraphics[width=\textwidth]{gfx/fig7b_phd.pdf}%%
+   \label{III_beamformer_comparison_holmengraa}}
+   \caption{Comparing image quality of DAS, MVDR and LCA with Kaiser or trigonometric windows.}\label{III_image_quality}
    \end{figure*}
@@
 Also note how LCA is insensitive to the window type; its performance is near identical whether it uses Kasier or trigonometric windows.]{\includegraphics[width=\textwidth]{gfx/buske7a_online.pdf}%
-   \label{beamformer_comparison_cross}}
+   \label{III_beamformer_comparison_cross}}
    \hfil
@@
 The adaptive methods produce nearly identical images.]{\includegraphics[width=\textwidth]{gfx/buske7b_online.pdf}%%
-   \label{beamformer_comparison_holmengraa}}
-   \caption{Comparing image quality of DAS, MVDR and LCA with Kaiser or trigonometric windows.}\label{image_quality}
+   \label{III_beamformer_comparison_holmengraa}}
+   \caption{Comparing image quality of DAS, MVDR and LCA with Kaiser or trigonometric windows.}\label{III_image_quality}
    \end{figure*}
@@
 
-\section{LCA}\label{sec:lca}
+\section{LCA}\label{III_sec:lca}
 
@@
 \argmin{p}\, E\{\big|z_p[n]\big|^2\} = \argmin{p}\, E\big\{\big|\w_{p}\H\x[n]\big|^2\big\}\nn
-\qquad\text{subject to}\qquad \w\H\a_\phi = 1.\label{eq:lca_definition}
-\end{align}\label{lca_criterion}
+\qquad\text{subject to}\qquad \w\H\a_\phi = 1.\label{III_eq:lca_definition}
+\end{align}\label{III_lca_criterion}
 %
-Note how closely LCA is related to the MVDR definition in \eq{eq:mvdr_definition}.
+Note how closely LCA is related to the MVDR definition in \eq{III_eq:mvdr_definition}.
 The optimization criterion and constraint is the same, but LCA has a finite and discrete solution space for the weights.
-As will be demonstrated in Section \ref{sec:results_lca_window_function}, LCA performs similarly to MVDR because a robustified MVDR implementation seems to be constrained to window functions similar to the ones we let LCA choose from.
+As will be demonstrated in Section \ref{III_sec:results_lca_window_function}, LCA performs similarly to MVDR because a robustified MVDR implementation seems to be constrained to window functions similar to the ones we let LCA choose from.
 
@@
 where $N_x = 2X+1$ is the number of azimuth lines to average over, and $\omega[x',n']$ is a normalized 2 dimensional weight function.
-Since we oversample slightly laterally when we delay each pixel (see Section \ref{sec:lca_oversampling}), we thought the lateral correlation to be sufficient to benefit from this.
+Since we oversample slightly laterally when we delay each pixel (see Section \ref{III_sec:lca_oversampling}), we thought the lateral correlation to be sufficient to benefit from this.
 However, we observed no visual improvement from applying this technique, and decided not to use it.
@@
 
-\subsection{Window function: Kaiser}\label{sec:lca_kaiser_windows}
+\subsection{Window function: Kaiser}\label{III_sec:lca_kaiser_windows}
 
@@
 f_{M-1}(\beta)
-}\label{eq:kaiser_window_vector}
+}\label{III_eq:kaiser_window_vector}
 \end{align}
@@
 \begin{align}
-f_m(\beta) = \frac{I_0\left(\pi\beta\sqrt{1-\left(\frac{2m}{M-1}-1\right)^2}\right)}{I_0(\pi\beta)}\label{eq:kaiser_window_element}
+f_m(\beta) = \frac{I_0\left(\pi\beta\sqrt{1-\left(\frac{2m}{M-1}-1\right)^2}\right)}{I_0(\pi\beta)}\label{III_eq:kaiser_window_element}
 \end{align}
@@
 \begin{align}
-I_0(x) = \sumb{a=0}{\infty} \left[ \frac{\left(\frac{x}{2}\right)^a}{a!} \right]^2.\label{eq:modified_bessel_first_kind}
+I_0(x) = \sumb{a=0}{\infty} \left[ \frac{\left(\frac{x}{2}\right)^a}{a!} \right]^2.\label{III_eq:modified_bessel_first_kind}
 \end{align}
@@
 
-\subsection{Steering}\label{sec:lca_steering}
+\subsection{Steering}\label{III_sec:lca_steering}
 
@@
 \begin{align}
-\w_{\beta,\phi} = \frac{\f_\beta\H\diag{(\a_\phi)}}{\f_\beta\H\a_\phi}\label{eq:steered_window}
+\w_{\beta,\phi} = \frac{\f_\beta\H\diag{(\a_\phi)}}{\f_\beta\H\a_\phi}\label{III_eq:steered_window}
 \end{align}
@@
 e^{-j\frac{2\pi (M-1)d}{\lambda}\sin(\phi)}
-}.\label{eq:steering_vector}
+}.\label{III_eq:steering_vector}
 \end{align}
@@
 The normalization factor $\f_\beta\H\a_\phi$ is the reciprocal of the window's coherent gain.
-It ensures unit gain in the direction of interest as required by (\ref{eq:lca_definition}).
+It ensures unit gain in the direction of interest as required by (\ref{III_eq:lca_definition}).
 Since the signal-to-noise ratio is constant for a specific window (same $\beta$-value), this normalization also proportionally increases the incoherent noise gain.
@@
 
-\subsection{Steering bounds}\label{sec:lca_steering_bounds}
+\subsection{Steering bounds}\label{III_sec:lca_steering_bounds}
 
 Each window's spatial response is constrained to unit gain in the look direction.
-Hence, when it is steered the white noise gain must increase and the the signal-to-noise ratio decrease, as shown in \Fig{windows_steering}.
+Hence, when it is steered the white noise gain must increase and the the signal-to-noise ratio decrease, as shown in \Fig{III_windows_steering}.
 To limit this we place an upper bound to the steering, chosen such that wide windows are allowed to be steered more than narrow ones.
 We call this upper bound for steering $\phi_{\mathrm{3\,dB}}(\beta)$, and define it as the steering angle needed for the steered amplitude response to have a \minus{}3\,dB crossing that is exactly half that of the unsteered window.
-The method we devised for finding it is described in Appendix \ref{appendix:steering_angles}.
+The method we devised for finding it is described in Appendix \ref{III_appendix:steering_angles}.
 
@@
 
-\subsection{Trigonometric windows}\label{sec:lca_trigonometric}
+\subsection{Trigonometric windows}\label{III_sec:lca_trigonometric}
 
@@
 \begin{align}
-f_m(\alpha) = \alpha - (1-\alpha)\cos\Big(\frac{2\pi m}{M-1}\Big),\label{eq:trig_window_function}
+f_m(\alpha) = \alpha - (1-\alpha)\cos\Big(\frac{2\pi m}{M-1}\Big),\label{III_eq:trig_window_function}
 \end{align}
@@
 where we apply the constraint $\alpha\in[0.5,1]$ to avoid windows with negative coefficients.
-Steering is applied as in (\ref{eq:steered_window}), and the $\alpha$-value that halves the \minus{}3\,dB distance is found using the method described in Appendix \ref{appendix:steering_angles}.
+Steering is applied as in (\ref{III_eq:steered_window}), and the $\alpha$-value that halves the \minus{}3\,dB distance is found using the method described in Appendix \ref{III_appendix:steering_angles}.
 
-\subsection{Oversampling}\label{sec:lca_oversampling}
+\subsection{Oversampling}\label{III_sec:lca_oversampling}
 
@@
 We consider an absolute minimum to be 2, due to the non-linear nature of delaying pixels that isn't strictly in the far field, and of displaying the absolute value of the pixels on a decibel scale.
-In \mediaI{}\footnote{For media location refer to Appendix \ref{app:source_code}.} we visualize the effect of changing the lateral oversampling factor on a set of LCA images.
+In \mediaI{}\footnote{For media location refer to Appendix \ref{III_app:source_code}.} we visualize the effect of changing the lateral oversampling factor on a set of LCA images.
 Upon display the images were all bilinearly upinterpolated to the same size.
 
-\section{Results and Discussion}\label{sec:results_discussion}
+\section{Results and Discussion}\label{III_sec:results_discussion}
 
@@
 This data was collected by the Norwegian Defense Research Establishment and Kongsberg Maritime.
-The other scene is of an 1\,m by 1\,m iron resolution test cross, connected to a an anchor with 13\,cm diameter (\Fig{cross}).
+The other scene is of an 1\,m by 1\,m iron resolution test cross, connected to a an anchor with 13\,cm diameter (\Fig{III_cross}).
 This data was collected by the Bundeswehr Technical Center for Ships and Naval Weapons, Maritime Technology and Research (WTD 71).
@@
 
-The results will be presented in the following order: In Section \ref{sec:results_lca_window_function} we discuss typical window responses computed by MVDR, and hypothesize that these can be mimicked by Kaiser windows.
-In Section \ref{sec:results_window_parameters} we determine sensible boundaries for the Kaiser parameter $\beta$ and window steering $\phi$.
-In Section \ref{sec:results_database_size} we discuss how many window variations are needed for LCA to perform well.
+The results will be presented in the following order: In Section \ref{III_sec:results_lca_window_function} we discuss typical window responses computed by MVDR, and hypothesize that these can be mimicked by Kaiser windows.
+In Section \ref{III_sec:results_window_parameters} we determine sensible boundaries for the Kaiser parameter $\beta$ and window steering $\phi$.
+In Section \ref{III_sec:results_database_size} we discuss how many window variations are needed for LCA to perform well.
 
-\subsection{LCA window function}\label{sec:results_lca_window_function}
+\subsection{LCA window function}\label{III_sec:results_lca_window_function}
 
 Assuming that a robust MVDR is the reference method we want LCA to perform similarly to, it seems sensible to create a window database for LCA with spatial responses similar to the ones that MVDR computes.
-We study this in \Fig{mvdr_selected_windows}, where we present the typical amplitude and phase responses for the windows that MVDR creates in shadow, speckle and highlight regions of the Holmengraa scene.
+We study this in \Fig{III_mvdr_selected_windows}, where we present the typical amplitude and phase responses for the windows that MVDR creates in shadow, speckle and highlight regions of the Holmengraa scene.
 
-Observe from \Fig{mvdr_selected_windows} that MVDR seems to prefer symmetric window responses, even if it is free to choose non-symmetric ones.
+Observe from \Fig{III_mvdr_selected_windows} that MVDR seems to prefer symmetric window responses, even if it is free to choose non-symmetric ones.
 The symmetry is most predominant in shadow and speckle regions, while in highlight the windows are approximately symmetric within the illuminated sector of the seafloor.
@@
 From this we hypothesize that a good window database LCA can be made using a varied  set of Kaiser windows.
-If they span a suitable range of $\beta$-values, and some steering variations are applied to each, we should have responses that resemble those in \Fig{mvdr_selected_windows}.
+If they span a suitable range of $\beta$-values, and some steering variations are applied to each, we should have responses that resemble those in \Fig{III_mvdr_selected_windows}.
 The Kaiser window is easy to compute, is fairly insensitive to coefficient inaccuracies, and span shapes from rectangular to Gaussian.
@@
 We have found that these windows hardly ever get used on experimental data, in particular steered versions of them.
-We can infer the same from \Fig{mvdr_selected_windows} by noting that the maximum sidelobe level rarely exceeds the rectangular window level of \minus{}13\,dB.
+We can infer the same from \Fig{III_mvdr_selected_windows} by noting that the maximum sidelobe level rarely exceeds the rectangular window level of \minus{}13\,dB.
 
-\mediaII{}\footnote{For media location refer to Appendix \ref{app:source_code}.} animates how \Fig{mvdr_selected_windows} changes as a function of the MVDR subarray length $L$ and temporal averaging $K$.
+\mediaII{}\footnote{For media location refer to Appendix \ref{III_app:source_code}.} animates how \Fig{III_mvdr_selected_windows} changes as a function of the MVDR subarray length $L$ and temporal averaging $K$.
 At $L=1$ the window responses are rectangular in all areas in the image.
@@
 
-\subsection[Kaiser parameter β and steering Φ]{Kaiser parameter $\beta$ and steering $\phi$}\label{sec:results_window_parameters}
+\subsection{Kaiser parameter \texorpdfstring{$\beta$}{} and steering \texorpdfstring{$\phi$}{}}\label{III_sec:results_window_parameters}
 
@@
 The lower boundaries were chosen as $\beta=0$ and $\phi=0^\circ$, which includes the rectangularly weighted DAS image.
-The resulting images are shown in \Fig{oversampling_mosaic_bounds}.
+The resulting images are shown in \Fig{III_oversampling_mosaic_bounds}.
 In each image we computed the lateral distance $\Delta x$ between the \minus{}3\,dB points of the anchor, with the reference $\Delta x_\text{ref}$ being the rectangularly weighted DAS image.
@@
 
-The LCA images in \Fig{oversampling_mosaic_bounds} demonstrate the effect of adjusting the upper bound of $\beta$ and $\phi$.
-As suggested by \Fig{mvdr_selected_windows} we first attempted to use $\beta\in[0,2]$.
+The LCA images in \Fig{III_oversampling_mosaic_bounds} demonstrate the effect of adjusting the upper bound of $\beta$ and $\phi$.
+As suggested by \Fig{III_mvdr_selected_windows} we first attempted to use $\beta\in[0,2]$.
 While this significantly improved sidelobe suppression, a slight further improvement was observed up to $\beta\in[0,5]$.
@@
 For this sonar the Rayleigh resolution is $\delta x\approx0.0125$ radians (0.72$^\circ$), so a point scatterer imaged at 19\,m range with a rectangular DAS would have a lateral $\Delta x\approx0.0125\cdot19\,\text{m}\cdot0.89=21\,\text{cm}$.
-This is close to the FWHM of the anchor in the DAS image in \Fig{oversampling_mosaic_bounds}, which means that the acoustic fingerprint of the anchor is similar to a point source.
+This is close to the FWHM of the anchor in the DAS image in \Fig{III_oversampling_mosaic_bounds}, which means that the acoustic fingerprint of the anchor is similar to a point source.
 This further implies that the anchor reflections are specular and that we view its rounded side.
@@
 
-\Fig{oversampling_mosaic_bounds_beta} and \Fig{oversampling_mosaic_bounds_phi} illustrates which $\beta$ and $\phi$ values LCA prefers for different regions in the image of the resolution test cross.
+\Fig{III_oversampling_mosaic_bounds_beta} and \Fig{III_oversampling_mosaic_bounds_phi} illustrates which $\beta$ and $\phi$ values LCA prefers for different regions in the image of the resolution test cross.
 Red marker lines are used to pinpoint the cut line, location of the anchor and strong scatters on the cross.
-Observe that in the anchor region only narrow responses ($\beta=0$) are used, which is the reason for the anchor FWHM being measured to the same value regardless of the upper bound of $\beta$ in \Fig{oversampling_mosaic_bounds}.
+Observe that in the anchor region only narrow responses ($\beta=0$) are used, which is the reason for the anchor FWHM being measured to the same value regardless of the upper bound of $\beta$ in \Fig{III_oversampling_mosaic_bounds}.
 On other scatterers with strong nearby lateral interference LCA prefers wider windows to suppress the interference.
@@
 
-\subsection{Window database size}\label{sec:results_database_size}
+\subsection{Window database size}\label{III_sec:results_database_size}
 
@@
 What remains is to determine the number of windows needed.
-We study this in \Fig{oversampling_mosaic}, where we compare LCA images made from windows databases of different sizes, but with the same parameter boundaries.
+We study this in \Fig{III_oversampling_mosaic}, where we compare LCA images made from windows databases of different sizes, but with the same parameter boundaries.
 Included are the results from using a single window ($N_\beta=1$) and steering angle ($N_\phi=1$), which corresponds to an unsteered rectangularly weighted DAS.
@@
 
-\subsection{Image quality}\label{sec:results_image_quality}
+\subsection{Image quality}\label{III_sec:results_image_quality}
 
-We compare LCA image quality to that of DAS and MVDR in \Fig{image_quality}.
-In \Fig{beamformer_comparison_cross} we display the images computed by LCA using Kaiser windows, LCA using trigonometric windows, and MVDR.
+We compare LCA image quality to that of DAS and MVDR in \Fig{III_image_quality}.
+In \Fig{III_beamformer_comparison_cross} we display the images computed by LCA using Kaiser windows, LCA using trigonometric windows, and MVDR.
 Two lateral image cuts through all images are presented in the left plot.
@@
 
-In \Fig{beamformer_comparison_holmengraa} we show full sector images of the Holmengraa wreck.
+In \Fig{III_beamformer_comparison_holmengraa} we show full sector images of the Holmengraa wreck.
 We compare the images produced by a Hamming weighted DAS, by MVDR and by LCA with Kaiser or trigonometric windows.
@@
 
-\subsection{Computational complexity}\label{sec:results_complexity}
+\subsection{Computational complexity}\label{III_sec:results_complexity}
 
@@
 Each pixel depends on only a small subset of data, and the windows can be precomputed and stored in GPU cache for near immediate access.
-Furthermore, if we use the trigonometric window function we can solve for the window parameter analytically as described in Appendix \ref{sec:reducing_lca_complexity}.
+Furthermore, if we use the trigonometric window function we can solve for the window parameter analytically as described in Appendix \ref{III_sec:reducing_lca_complexity}.
 This reduces the complexity to O($M N_\phi$), because the only remaining LCA parameter to create windows for will be the steering angle.
 
-\section{Conclusion}\label{sec:conclusion}
+\section{Conclusion}\label{III_sec:conclusion}
 
@@
 
-\section{Steering angle}\label{appendix:steering_angles}
+\section{Steering angle}\label{III_appendix:steering_angles}
 
@@
 Note how the angle scales near proportionally with $\frac{d}{\lambda}$.
-}\label{steering_angle_3dB}
+}\label{III_steering_angle_3dB}
 \end{figure}
@@
 To control the resolution gain of LCA, we wanted to determine the extent that the windows need to be steered to cut the angle of the \minus{}3\,dB point of the window's spatial amplitude response by a fixed amount, say, by a factor 2 as reference.
-We computed this steering angle for some common configurations of the mentioned parameters, see the result in \Fig{steering_angle_3dB}.
+We computed this steering angle for some common configurations of the mentioned parameters, see the result in \Fig{III_steering_angle_3dB}.
 The boundary values of the Kaiser window is shown, i.e.
@@
 
-For the sake of completeness we also supply the source code, see Appendix \ref{app:source_code}.
+For the sake of completeness we also supply the source code, see Appendix \ref{III_app:source_code}.
 
-\section{Reducing LCA complexity}\label{sec:reducing_lca_complexity}
+\section{Reducing LCA complexity}\label{III_sec:reducing_lca_complexity}
 
@@
 Throughout this article we have used LCA with the Kaiser window function.
-However, as shown in \Fig{image_quality} we can obtain similar performance using the trigonometric window function.
+However, as shown in \Fig{III_image_quality} we can obtain similar performance using the trigonometric window function.
 For this window we can obtain an analytic solution for the optimal value of $\alpha$ in the minimum variance sense.
 The following derivation generalizes that of the non-steered (real) window described in e.g.~\cite{Stoica2005} to also apply for the steered (complex) window.
-We start by inserting the trigonometric function (\ref{eq:trig_window_function}), with applied steering (\ref{eq:steering_vector}), into the beamformer equation (\ref{eq:beamformer_output}):
+We start by inserting the trigonometric function (\ref{III_eq:trig_window_function}), with applied steering (\ref{III_eq:steering_vector}), into the beamformer equation (\ref{III_eq:beamformer_output}):
 %
@@
 %
-where $[\a_\phi]_m$ is the $m$th component of the steering vector $\a_\phi$ defined in (\ref{eq:steering_vector}), and 
+where $[\a_\phi]_m$ is the $m$th component of the steering vector $\a_\phi$ defined in (\ref{III_eq:steering_vector}), and 
 %
@@
 
-\section{Multimedia Files and Source Code}\label{app:source_code}
+\section{Multimedia Files and Source Code}\label{III_app:source_code}
 

### 2018-10-11 Simulator/article.tex
Testing change of frame letters

@@
 \newif\ifTODO\TODOtrue                        % Use todo notes?
+\newif\ifOPTtodos\OPTtodosfalse
+\newif\ifPhdDoc\PhdDocfalse
 
-\newif\ifBuildBibliography\BuildBibliographyfalse  % Whether to include the references.bib file or use inline references.
+\newif\ifBuildBibliography\BuildBibliographytrue  % Whether to include the references.bib file or use inline references.
+\newif\ifOverLeaf\OverLeaftrue                  % Using OverLeaf to edit?
 
@@
 
+%\usepackage[backend=biber,bibstyle=ieee]{biblatex}
+%\addbibresource{references.bib}
+
+\RequirePackage{amsfonts}
 \usepackage{cite}
@@
 
-\usepackage[cmex10]{amsmath}                    % cmex10 option to be IEEE explore compliant
+%\usepackage[cmex10]{amsmath}                    % cmex10 option to be IEEE explore compliant
+\usepackage[cmex10]{mathtools}                    % cmex10 option to be IEEE explore compliant
 \interdisplaylinepenalty=2500                   % Allows multiline equations to be broken
-
-% \RequirePackage{amssymb}
+\RequirePackage{amssymb}
 
@@
 
+\usepackage{accents}
+
 % NOTE: PDF hyperlink and bookmark features are not required in IEEE
@@
 
+\usepackage{tikz}
+\usetikzlibrary{calc,positioning,shapes,snakes}
+%\newcommand{\tikzmark}[1]{\tikz[remember picture]\coordinate(#1){(-.35em,-.545ex)};}
+\newcommand{\tikzmark}[2]{\tikz[remember picture,baseline=(#1.base)]\node[shape=rectangle,inner sep=0pt](#1){#2};}
+\newcommand*\circled[1]{\tikz[baseline=(char.base)]{
+            \node[shape=circle,draw,inner sep=2pt] (char) {#1};}}
+
+\usepackage{multirow}
+\usepackage{tabularx}
+
+%\usepackage[active,float]{preview}
+%\PreviewEnvironment{gather} 
 % \ifPeerReview
@@
 % \usepackage[maxfloats=40]{morefloats}
-\newcounter{todoidx}
-% \setcounter{todoidx}
+%\newcounter{todoidx}
+%% \setcounter{todoidx}
+%
+%\ifTODO
+%   \definecolor{todobackground}{rgb}{0.95,0.95,0.95}
+%   \setlength\marginparsep{1pt}
+%   \setlength\marginparwidth{35pt}
+%   \newlength\marginparwidthsmall
+%   \setlength\marginparwidthsmall{\marginparwidth}
+%   \addtolength\marginparwidthsmall{-7pt}
+%   \newcommand\todo[1]{%
+%      \addtocounter{todoidx}{1}%
+%      {\color{Red}\bf(\thetodoidx{})}%%\fbox{\bf\thetodoidx{}}}%
+%      \marginpar{%
+%         {\vspace*{-10pt}\color{Red}\fbox{\bf\thetodoidx{}}}\\%
+%         \fcolorbox{red}{todobackground}{\parbox{\marginparwidthsmall}{\raggedright\scriptsize #1}}}}
+%
+%   \newcommand\todopar[1]{\fcolorbox{red}{white}{\parbox{0.97\linewidth}{#1}}}
+%\else
+%%    \usepackage[disable]{./todonotes} 
+%   \newcommand\todo[1]{}
+%\fi
 
-\ifTODO
-   \definecolor{todobackground}{rgb}{0.95,0.95,0.95}
-   \setlength\marginparsep{1pt}
-   \setlength\marginparwidth{35pt}
-   \newlength\marginparwidthsmall
-   \setlength\marginparwidthsmall{\marginparwidth}
-   \addtolength\marginparwidthsmall{-7pt}
-   \newcommand\todo[1]{%
-      \addtocounter{todoidx}{1}%
-      {\color{Red}\bf(\thetodoidx{})}%%\fbox{\bf\thetodoidx{}}}%
-      \marginpar{%
-         {\vspace*{-10pt}\color{Red}\fbox{\bf\thetodoidx{}}}\\%
-         \fcolorbox{red}{todobackground}{\parbox{\marginparwidthsmall}{\raggedright\scriptsize #1}}}}
-
-   \newcommand\todopar[1]{\fcolorbox{red}{white}{\parbox{0.97\linewidth}{#1}}}
-\else
-%    \usepackage[disable]{./todonotes} 
-   \newcommand\todo[1]{}
-\fi
+\newcommand\gls[1]{}
 

### 2018-10-11 ThesisJo/2_publications.tex
Almost perfect build of all targets with precompiled header
Almost there with all contents, but encoding issues. Trying Xelatex
Testing change of frame letters

@@
-%&2_publications
+%&header
 % !TeX root = 2_publications
-
-\input{header}
 \input{subheader}
@@
 
+\ifRootBuild\else
   \input{glossary}
   \makeglossaries
+\fi
 
 \begin{document}
-%\documentclass[float=false, crop=false]{standalone}
-%\usepackage[subpreambles=true]{standalone}
-%\input{header}
-%\input{subheader}
 
@@
 %~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
-\section{Paper I}\label{sec:paperI} %
+\section{Intro}\label{sec:intro} %
 %~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
 
-\paper{I} describes the use and implementation of the \gls{MVDR} beamformer in active sonar imaging in its most general and exact form.
-
-Its widespread use and generic implementation makes it the reference.
+%
+%\paper{I} describes the use and implementation of the \gls{MVDR} beamformer in active sonar imaging in its most general and exact form. 
+%
+%Its widespread use and generic implementation makes it the reference.
 thus will be considered the reference method.
@@
 This is to prove that using a \gls{MVDR}
-
-Paper II takes the \gls{MVDR} method and applies it to clinical Ultrasound Imaging
-
-We point out important hardware limitations for these devices, and assess the design in terms of how efficiently it is able to use the GPU's resources.
+%
+%Paper II takes the \gls{MVDR} method and applies it to clinical Ultrasound Imaging
+%
+%
+%We point out important hardware limitations for these devices, and assess the design in terms of how efficiently it is able to use the GPU's resources.
 On a quad-core Intel Xeon system with a high-end Nvidia GPU, our GPU implementation renders more than a million pixels per second (1\;MP/s).
@@
 This throu
+
+Common:
+\begin{itemize}\setlength\baselineskip{0pt}
+\item Adaptive
+\item Processing speed
+\end{itemize}
 
@@
 
-\begin{itemize}
-\item Minimum variance
-\item Full search space, continuous
-\item Element-space
+\begin{itemize}\setlength\baselineskip{0pt}
+\item Optimization: Minimum variance
+\item Solution space: Continuous, closed when made robust
+\item Time domain
 \item Sonar
-\item Computationally complex
-\item Common: Minimal temporal statistics (near single snapshot)
+\item Computationally complex, $O(N^3)$.
+\item Information significance: Minimal temporal sample support (~single snapshot)
 \end{itemize}
 
-This is a popular method that analytically estimates the array's weight set that best comply with the minimum variance criterion.
+%This is a popular method that analytically estimates the array's weight set that best comply with the minimum variance criterion.
 Its analytic nature allows it to chose from an infinite number of possible weight sets, but with a solution space limited by the constraint of unit gain in the look direction.
-
-The first paper is concerned with applying the Minimum Variance Distortionless Response (MVDR) beamformer to active sonar imaging, with emphasis on performing arithmetic reductions and solving the practical its high computational complexity.
+%
+%The first paper is concerned with applying the Minimum Variance Distortionless Response (MVDR) beamformer to active sonar imaging, with emphasis on performing arithmetic reductions and solving the practical its high computational complexity.
 It shows that  with emphasis on solving and with the process of implementing this method on a \gls{GPU}.
 The goal is to improve sonar image quality without suffering the consequences of the MVDR's inherently high computational complexity.
-
-Combined, these steps led to a speedup of an order of magnitude or more compared to its straight-forward \gls{CPU} implementation.
-
-In functionality this method acts as our reference method, but its practicality is limited by a high computational complexity.
+%
+%Combined, these steps led to a speedup of an order of magnitude or more compared to its straight-forward \gls{CPU} implementation.
+%
+%In functionality this method acts as our reference method, but its practicality is limited by a high computational complexity.
 The paper suggests means for reducing the arithmetic complexity and provide extensive details for fitting the implementation on a \gls{GPU} for a major speedup.
@@
 %~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
-\textbf{Low Complexity Adaptive Beamforming}~\cite{Buskenes2015}\\
+\textbf{Low Complexity Adaptive Beamforming}~\cite{Buskenes2014}\\
 J.I. Buskenes, R.E. Hansen and A. Austeng\\
@@
 %~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
+\textbf{A Real-Time Synthetic Aperture Sonar Simulator for Automatic Target Recognition: Notation and application}\\
+J.I. Buskenes, H. Midelfart\\
+\textit{IEEE Journal of Oceanic Engineering}\\
+Submitted DD.MM.2018%\\\\
+%Volume 42, Issue 1, January 2017
 
@@
 \item Sonar
+	\item Incredibly fast template matching possible.
+	\item Possibly a powerful tool to pass prior knowledge to deep learning
+	\item Extracting the essential information is key 
+	\item Exploit hardware resources available
 \end{itemize}
@@
 \section{Onlook}
+
+Machine cognition:
+\begin{itemize}
+\item Ethical dilemmas. Decision and actions have consequences. Who's legally responsible for a machine's doing? Itself?
+\item Safety. Machine flexibility allow more possible outcomes, and a wider range of security issues to consider.
+\end{itemize}
 

### 2018-10-11 ThesisJo/3_discussion.tex
Almost there with all contents, but encoding issues. Trying Xelatex
Testing change of frame letters

@@
 % !TeX root = 3_discussion
-\documentclass[thesis.tex]{subfiles}
 \input{subheader}
+\endofdump
+
+\ifRootBuild\else
+  \input{glossary}
+  \makeglossaries
+\fi
+
 \begin{document}
@@
 
+Constrain attention to beamformers and simulators (not ATR?).
+
+\section{Trends}
+
+The future is faster, smarter.
+
+\subsection{Hardware}
+
 \begin{itemize}
-\item Improving imaging systems by making them smarter.
-\item Big data - more processing power.
-\item Processing speed
+\item Ever increasing processing speed.
 \begin{itemize}
@@
 	\item Exploit hardware resources available
+   \item Intel entering the discrete graphics stage
    \end{itemize}
+\item Software defined transceiver/receivers
+\begin{itemize}
+\item System on Chips with integrated ADCs/DACs. Big power in small package.
+\item Compact, flexible, direct conversion
+\item Software defined signal generation, filtering, synchronization, rate-conversion and mixing.
+\item Multipurpose, hardware reuse, easier implementation of advanced signal processing, faster development and lower price.
+\item 
+\item Compact multi-band Wifi
+\item MIMO (in new wifi). Breaks array SNR boost. Not viable for long range.
+\item Phones: Wifi 2.4-2.5G, 5G (massive MIMO, heavy SDR, 600M-6G), LTE (4G, Long Term Evolution 800M), GSM (2G, Global System for Mobile communications 900M,1.8-1.9M), UMTS (3G, Universal Mobile Telecommunications System 1.9-2.1M), GPS (Global Positioning System 1.5G), Bluetooth (2.4G), NFC (Near-Field Communications 13.56M), FM (100M).
+Multi-radio in a single SDR package!
+\end{itemize}
+\item Short feedback loops allow real-time adaptivity
+\end{itemize}
+
+\subsection{Algorithms}
+
+\begin{itemize}
+\item Intelligence - Ability to process data fast, and retain key information
+\item Adaptive methods ever more attractive.
+   \begin{itemize}
+   \item Processing more data to increase significance of extracted information
+   \item Reducing reaction time: Act when the data is fresh/valid.
+   \end{itemize}
+\item Better image quality
+\item Improved autonomy
+\end{itemize}
+
+\subsection{Connectivity}
+
+\begin{itemize}
+\item Internet of Things
+\item Big data
+\item Intelligent mesh of nodes
+\end{itemize}
+
+\section{Risks \& Challenges}
+
+\begin{itemize}
+\item Improving imaging systems by making them smarter.
+\item Big data - more processing power.
 \item Faster IS smarter
@@
  \end{itemize}
-\item Smarter IS better
- \begin{itemize}
-	\item Better image quality
-	\item Improved autonomy
-\end{itemize}
 \end{itemize}
 
+%Discussion and future work
+%   Summary and discussion
+%      Simulator:
+%      - The way to train and constrain deep learning algorithms?
+%      MVDR:
+%      - Analytical optimization - full adaptivity.
+%      - Heavy on computations
+%      LCA:
+%      - Optimization by trial and error - discrete solution space.
+%      - Works almost as well as MVDR in most cases
+%      - Fast!
+%      - Well suited for hardware
+%      
+%      Lessions from deep learning:
+%      - 
+%      
+%   Future work
+
+\section{MVDR}
+
+Recent developments:
+- A low complexity minimum variance beamformer for ultrasound imaging using dominant mode rejection~\cite{Jiang2016}
+- Google deep beamforming
+
 \begin{itemize}
-\item Improving imaging systems by making them smarter.
-\item Big data - more processing power.
-\item IV - Simulator
-\begin{itemize}
-	\item Incredibly fast template matching possible.
-	\item Possibly a powerful tool to pass prior knowledge to deep learning
-	\item Extracting the essential information is key 
-	\item Exploit hardware resources available
-\end{itemize}
-\item Faster IS smarter
- \begin{itemize}
- 	\item Short feedback loops allow real-time adaptivity
- 	\item More data contains more information
- 	\item Faster learning
- \end{itemize}
-\item Smarter IS better
- \begin{itemize}
-	\item Better image quality
-	\item Improved autonomy
-\end{itemize}
+\item SDR - Entirely software defined radio, radar and sonar, bigger potential for adaptive methods
+\item Cognitive radar/sonar.
+1.
+Adaptive signal processing, learn from data.
+2.
+Feedback from receiver to transmitter.
+3.
+Preservation of the information in received data.
+All of this is present in bats.
 \end{itemize}

### 2018-10-11 ThesisJo/frontmatter.tex
Almost there with all contents, but encoding issues. Trying Xelatex
Testing change of frame letters

--- 2018-02-06 ThesisJo/preface.tex
@@
+%&header
+% !TeX root = thesis
+\input{subheader}
+\endofdump
+
+\ifRootBuild\else
+  \input{glossary}
+  \makeglossaries
+\fi
+
+\begin{document}
+
+\begin{markdown}
+Hey
+\end{markdown}
+
 \chapter*{Preface}
@@
 \section*{Acknowledgements}
+
+\newpage
+\chapter*{Abstract}
+
+\newpage
+\chapter*{List of Publications}
+
+This thesis is based on the following four journal papers, referred to by their roman literal:
+
+\newcommand\pubEntry[4]{\item #2, "#1", \textit{#3}, #4}
+{\renewcommand{\theenumi}{\textbf{\Roman{enumi}}}
+\renewcommand\cleardoublepage{\newpage}
+\begin{enumerate}
+\pubEntry{%
+An Optimized GPU implementation of the MVDR Beamformer for Active Sonar Imaging}%
+{J.I. Buskenes, J.P. Åsen, C.-I.C. Nilsen and A. Austeng}%
+{IEEE Journal of Oceanic Engineering}%
+{Volume 40, Issue 2, April 2015}
+%
+\pubEntry{%
+Implementing Capon Beamformer on a GPU for Real-Time Cardiac Ultrasound Imaging}%
+{J.P. Åsen, J.I. Buskenes, C.-I.C. Nilsen, A. Austeng and S. Holm}%
+{IEEE Transactions on Ultrasonics, Ferroelectrics, and Frequency Control}%
+{Volume 61, Issue 1, January 2015}
+%
+\pubEntry%
+{Low Complexity Adaptive Beamforming}%
+{J.I. Buskenes, R.E. Hansen and A. Austeng}%
+{IEEE Journal of Oceanic Engineering}%
+{Volume 42, Issue 1, January 2017}
+%
+\pubEntry%
+{A Real-Time Synthetic Aperture Sonar Simulator for Automatic Target Recognition: Notation and application}%
+{J.I. Buskenes, H. Midelfart}%
+{IEEE Journal of Oceanic Engineering}%
+{Submitted DD.MM.2018}
+\end{enumerate}
+%}
+
+%\newpage
+\chapter*{Related Publications}
+
+\begin{enumerate}
+\pubEntry{%
+A GPU sonar simulator for automatic target recognition}%
+{J.I. Buskenes, H. Midelfart, Ø. Midtgaard}%
+{Underwater acoustics international conference and exhibition}%
+{June 2014}
+%
+\pubEntry{%
+Adapting the minimum variance beamformer to a graphics processing unit for active sonar imaging systems}%
+{J.I. Buskenes, J.P. Åsen, C.-I.C. Nilsen, and A. Austeng}%
+{Journal of the Acoustical Society of America}%
+{Volume 133, 2013}
+%
+\pubEntry%
+{GPU-Based Adaptive Beamforming for Active Sonar Imaging}%
+{J.I. Buskenes, J.P. Åsen, C.-I.C. Nilsen, and A. Austeng}%
+{11th European Conference on Underwater Acoustics (ECUA)}%
+{ISBN: 978-1-62276-192-0 p. 860-865, 2012}
+%
+\pubEntry%
+{Implementing Capon Beamforming on the GPU for Real Time Cardiac Ultrasound Imaging}%
+{J.P. Åsen, J.I. Buskenes, C.-I.C. Nilsen, A. Austeng, and S. Holm}%
+{Proceedings - IEEE Ultrasonics Symposium}%
+{p. 2133-2136, 2012}
+%
+\pubEntry%
+{A Low Complexity Adaptive Beamformer for Active Sonar Imaging}%
+{J.I. Buskenes, C.-I.C. Nilsen, and A. Austeng}%
+{Underwater Acoustic Measurements: Technology and Results}%
+{2011}
+\end{enumerate}
+}
+
+\end{document}

### 2019-04-18 Simulator/article.tex
Long since last commit
Simulator: End of day
2019-04-18

@@
 \newif\ifBuildBibliography\BuildBibliographytrue  % Whether to include the references.bib file or use inline references.
-\newif\ifOverLeaf\OverLeaftrue                  % Using OverLeaf to edit?
+\newif\ifOverLeaf\OverLeaffalse                  % Using OverLeaf to edit?
 
@@
 \RequirePackage{amsfonts}
+% \usepackage[makeroom]{cancel}
+% \usepackage{cancel}
+% \usepackage{ulem}
 \usepackage{cite}
@@
 \usepackage{accents}
+\usepackage{moresize}
 
@@
 \usepackage{tikz}
-\usetikzlibrary{calc,positioning,shapes,snakes}
+\usetikzlibrary{arrows.meta,calc,positioning,shadows,shadows.blur,shapes,snakes}
 %\newcommand{\tikzmark}[1]{\tikz[remember picture]\coordinate(#1){(-.35em,-.545ex)};}
 \newcommand{\tikzmark}[2]{\tikz[remember picture,baseline=(#1.base)]\node[shape=rectangle,inner sep=0pt](#1){#2};}
+%\newcommand{\tikzmark}[2][]{\tikz[remember picture,overlay]\coordinate[#1](#2);}
 \newcommand*\circled[1]{\tikz[baseline=(char.base)]{
-            \node[shape=circle,draw,inner sep=2pt] (char) {#1};}}
+            \node[shape=circle,draw,inner sep=2.5pt] (char) {#1};}}
+
+\tikzset{-{to[length=.5em,width=.5em]}}
+
+\newcommand\step[1]{\tikz[baseline=(char.base)]{%
+            \node[shape=circle,draw,color=Green,shade,top color=black!5!white,bottom color=black!15!white,inner sep=.75pt] (char) {\color{Green}#1};}}
+\newcommand\figstep[2]{\step{#1}\mbox{}\hfill\parbox[t]{.97\linewidth}{#2}}
+
+\makeatletter
+\let\@oldseccntformat\@seccntformat
+\DeclareRobustCommand{\@seccntformat}[1]{%
+  \def\temp@@a{#1}%
+  \def\temp@@b{subsubsection}%
+  \ifx\temp@@a\temp@@b
+  \else
+  \@oldseccntformat{#1}
+  \fi
+}
+\renewcommand\subsubsection{\@startsection{subsubsection}{3}{\parindent\step{\normalfont\arabic{subsubsection}}\enskip}{0ex plus 0.1ex minus 0.1ex}{0ex}{\normalfont\normalsize\itshape}}%
+\makeatother
+
+%\titleformat{\subsubsection}[hang]{\itshape}{\thesubsubsection}{0.6em}{}
+
+%\makeatletter
+%\xpatchcmd{\@sect}{%
+%  \protected@edef\@svsec{\@seccntformat{#1}\relax}%
+%}{%
+%  \ifnum0=\pdfstrcmp{#1}{section}%
+%  \def\@svsec{\circlednumber{\number\value{section}}\quad\relax}%
+%  \else
+%  \protected@edef\@svsec{\@seccntformat{#1}\relax}%
+%  \fi
+%  }{}{}
+%\makeatother
+
+%\titleformat{\subsubsection}{\Large\bfseries}{}{0em}{%
+%\begin{picture}(0,0)
+%\put(12,4){\circle{28}}
+%\end{picture}\makebox[1.45em][c]{\thesubsubsection}\quad}
 
@@
 \usepackage{tabularx}
+\usepackage{collcell}
+\usepackage{booktabs}
+\usepackage{calc}
 
@@
 \fi
+
+% The below line commands ensure proper fill color along with the booktabs package
+% \specialrule[width]{abovespace}{belowspace}
+\colorlet{tableheadcolor}{tabBlue}
+\newlength\arulesep\setlength\arulesep{\aboverulesep}
+\newlength\brulesep\setlength\brulesep{\belowrulesep}
+\setlength{\aboverulesep}{0pt}
+\setlength{\belowrulesep}{0pt}  
+\newcommand\topline{\arrayrulecolor{black}\specialrule{\heavyrulewidth}{\arulesep}{0pt}%
+            \arrayrulecolor{tableheadcolor}\specialrule{\brulesep}{0pt}{0pt}%
+            \arrayrulecolor{black}}
+\newcommand\midline{\arrayrulecolor{tableheadcolor}\specialrule{\arulesep}{0pt}{0pt}%
+            \arrayrulecolor{black}\specialrule{\lightrulewidth}{0pt}{0pt}%
+            \arrayrulecolor{white}\specialrule{\brulesep}{0pt}{0pt}%
+            \arrayrulecolor{black}}
+\newcommand\cmidline[2][]{\arrayrulecolor{tableheadcolor}\cmidrule[\arulesep]{#2}%
+                        \arrayrulecolor{black}\cmidrule[\lightrulewidth]{#2}%
+                        \arrayrulecolor{tableheadcolor}\cmidrule[\brulesep]{#2}%
+%                         \arrayrulecolor{tableheadcolor}\cmidrule[\belowrulesep]{#1}%
+%                         \vspace{-\belowrulesep}
+%                         \arrayrulecolor{black}\cmidrule[\lightrulewidth]{#1}
+%                         \arrayrulecolor{tableheadcolor}\cmidrule[\belowrulesep]{#1}%
+%                        \arrayrulecolor{black}\cmidrule[\lightrulewidth]{#1}%\specialrule{\lightrulewidth}{0pt}{0pt}%
+%                        \arrayrulecolor{white}\cmidrule[0\belowrulesep]{#1}%
+                        \arrayrulecolor{black}}
+%\newcommand\cbmidline[1]{\arrayrulecolor{tableheadcolor}\cmidrule[0\aboverulesep]{#1}%
+%                        \arrayrulecolor{black}\cmidrule[\lightrulewidth]{#1}%\specialrule{\lightrulewidth}{0pt}{0pt}%
+%                        \arrayrulecolor{white}\cmidrule[0\belowrulesep]{#1}%
+%                        \arrayrulecolor{black}}
+%\newlength\arulesep{}\setlength\arulesep{\aboverulesep}\addtolength\arulesep{\lightrulewidth}
+%\newcommand\cmidline[1]{\arrayrulecolor{tableheadcolor}\specialrule{\aboverulesep}{0pt}{-\aboverulesep}%
+%            \arrayrulecolor{black}\ule{#1}
+%            \arrayrulecolor{tableheadcolor}\specialrule{\belowrulesep}{-\belowrulesep}{0pt}%
+%            \arrayrulecolor{black}}
+%\newlength\trulesep{}\setlength  \trulesep{\aboverulesep}
+%                     \addtolength\trulesep{\lightrulewidth}
+%                     \addtolength\trulesep{\belowrulesep}
+%\newlength\brulesep{}\setlength  \brulesep{\lightrulewidth}
+%                     \addtolength\brulesep{\belowrulesep}
+%\newlength\mrulesep{}\setlength  \mrulesep{\aboverulesep}
+%                     \addtolength\mrulesep{\belowrulesep}
+%\newcommand*{\cmidline}[1]{\cmidrule(rl){#1}}%
+%                      \arrayrulecolor{tableheadcolor}\specialrule{\trulesep}{0pt}{-\trulesep}\arrayrulecolor{black}
+%                      \cmidrule(rl){#1}
+%%                      \arrayrulecolor{white}\specialrule{0pt}{-\mrulesep}{0pt}%
+%                      \arrayrulecolor{black}
+%}
+
+%\newcommand*\cmidwrap[1]{\cmidrule(rl){#1}}%
+%                      \arrayrulecolor{tableheadcolor}\specialrule{\trulesep}{0pt}{-\trulesep}\arrayrulecolor{black}
+%                      \cmidrule(rl){#1}
+%%                      \arrayrulecolor{white}\specialrule{0pt}{-\mrulesep}{0pt}%
+%                      \arrayrulecolor{black}
+%}
+
+%             \arrayrulecolor{tableheadcolor} \specialrule{6pt}{0pt}{-6pt} \arrayrulecolor{black}
+%             
+%             \headcol Blue & Red & Yellow\\
+%\newcommand\cmidline[1]{\arrayrulecolor{tableheadcolor}\specialrule{\aboverulesep}{0pt}{0pt}%
+%                        \arrayrulecolor{black}\specialrule{\lightrulewidth}{0pt}{0pt}%
+%                        \arrayrulecolor{white}\specialrule{\belowrulesep}{0pt}{0pt}%
+%                        \arrayrulecolor{black}}
+%            
+%\cmidrule(lr){2-3}
+%
+%\usepackage{xparse}
+%\newcommand{\cmidline}{\noalign\bgroup\CMidRuleZaux{}}
+%
+%\ExplSyntaxOn
+%\makeatletter
+%\NewDocumentCommand{\CMidRuleZaux}{
+%    m % Material to reinsert before cmidrule.
+%%    O{0.0ex} % #1 = left adjust
+%%    O{0.0ex} % #1 = right adjust
+%    m  %   #3 = columns to span
+%}{
+%    \peek_meaning_remove_ignore_spaces:NTF \cmidline
+%      { \CMidRuleZaux { #1 \beforerule\cmidrule[\cmidrulewidth](lr){#2} } }%(l{#2}r{#3})
+%      { \egroup #1 \cmidrule[\cmidrulewidth](lr){#2} }
+%}
+%\makeatother
+%\ExplSyntaxOff
+%
+%                     
+%\newcommand*\beforerule{\arrayrulecolor{tableheadcolor}\specialrule{\trulesep}{0pt}{-\trulesep}\arrayrulecolor{black}}
+%\newcommand*\afterrule{\arrayrulecolor{white}\specialrule{0pt}{-\mrulesep}{0pt}\arrayrulecolor{black}}
+
+\newlength\trulesep{}\setlength  \trulesep{\arulesep}
+                     \addtolength\trulesep{\lightrulewidth}
+                     \addtolength\trulesep{\brulesep}
+\newcommand\cmidwrap[1]{%\arrayrulecolor{tableheadcolor}\specialrule{\trulesep}{0pt}{-\trulesep}%
+                        \arrayrulecolor{tableheadcolor}\specialrule{\arulesep}{0pt}{0pt}%
+                        \arrayrulecolor{tableheadcolor}\specialrule{\lightrulewidth}{0pt}{-\lightrulewidth}%
+%                        \arrayrulecolor{black}\cmidrule[\lightrulewidth]{#2}%
+%                        \arrayrulecolor{tableheadcolor}\cmidrule[\brulesep]{#2}%
+                        \arrayrulecolor{black}#1
+                        \arrayrulecolor{tableheadcolor}\specialrule{\brulesep}{0pt}{0pt}%
+%            \setlength{\aboverulesep}{10pt}
+%            \setlength{\belowrulesep}{10pt}
+%            #1
+            }
 

### 2019-05-05 Simulator/bio/jo_inge_buskenes_new.tex
2019-05-04

@@
 He has lectured in digital signal processing at the Gj\o{}vik College University (2009), and at the University of Oslo (2010-2013).
-Current affiliation is with The Norwegian Defence Research Establishment, Kjeller, Norway, for which he is developing radar systems (2015-), and formerly sonar systems (2009, 2013).
+Current affiliation is with The Norwegian Defence Research Establishment, Kjeller, Norway, where he develops sonar and radar systems.
 

### 2019-05-11 Simulator/article.tex
2019-04-19
2019-04-22
2019-05-04
2019-05-09
2019-05-10

@@
     plainpages=false,     %Store roman/arabic numbering differently to avoid
-    bookmarksnumbered ]{hyperref}
+    unicode=true,
+    bookmarksopen=true,
+%    bookmarksopenlevel=3,
+    bookmarksnumbered=false ]{hyperref}
 % 
@@
 %%%%%%%%%%%%%%%%%%%%%%%                         ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
+
+% Exotic orchids
+% https://visme.co/blog/wp-content/uploads/2016/01/63-1024x512.jpg
+%\definecolor{cDarkBlue}{HTML}{192e5b}
+%\definecolor{cBlue}{HTML}{1d65a6}
+%\definecolor{cLightBlue}{HTML}{73a2c0}
+%\definecolor{cGreen}{HTML}{00743f}
+%\definecolor{cYellow}{HTML}{f2a104}
+%\colorlet{cBase}{cLightBlue}
+%\colorlet{cBaseMark}{cGreen}
+%\colorlet{cBrightText}{cYellow}
+%\colorlet{cDarkText}{cDarkBlue}
+
+\ifOnlineColor
+   \definecolor{tabBlue}{HTML}{AACCFF}
+\else
+   \definecolor{tabBlue}{HTML}{CCCCCC}
+\fi
+
+%\definecolor{cBase}{HTML}{375e97}
+% \colorlet{cBase}{yellow}
+\colorlet{cBase}{tabBlue}
+%\definecolor{cBaseMark}{HTML}{3f681c}
+\colorlet{cBaseMark}{Green}
+\definecolor{cDarkText}{HTML}{ff662f}
+\definecolor{cBrightText}{HTML}{3492c2}
+% \definecolor{cBrightText}{HTML}{2492c2}
+% \definecolor{cDarkArrow}{HTML}{ff662f}
+% \definecolor{cBrightArrow}{HTML}{3492c2}
+\colorlet{cDarkArrow}{gray}
+\colorlet{cBrightArrow}{gray}
+
+\newcommand\supcarry[1]{{\color{cBrightText}#1}}
+\newcommand\subcarry[1]{{\color{cDarkText}#1}}
 
@@
 \newcommand\step[1]{\tikz[baseline=(char.base)]{%
-            \node[shape=circle,draw,color=Green,shade,top color=black!5!white,bottom color=black!15!white,inner sep=.75pt] (char) {\color{Green}#1};}}
+            \node[shape=circle,draw,color=cBaseMark,shade,top color=black!5!white,bottom color=black!15!white,inner sep=.75pt] (char) {\color{cBaseMark}#1};}}
 \newcommand\figstep[2]{\step{#1}\mbox{}\hfill\parbox[t]{.97\linewidth}{#2}}
@@
 }
+\let\oldsubsubsection\subsubsection
 \renewcommand\subsubsection{\@startsection{subsubsection}{3}{\parindent\step{\normalfont\arabic{subsubsection}}\enskip}{0ex plus 0.1ex minus 0.1ex}{0ex}{\normalfont\normalsize\itshape}}%
@@
 \usepackage{calc}
+%\usepackage{ifthen}
 
@@
 
-\ifOnlineColor
-   \definecolor{tabBlue}{HTML}{AACCFF}
-\else
-   \definecolor{tabBlue}{HTML}{CCCCCC}
-\fi
+%% Cool blues (sea)
+%\definecolor{deep_aqua}{HTML}{003b46}
+%\definecolor{ocean}{HTML}{07575b}
+%\definecolor{wave}{HTML}{66a5ad}
+%\definecolor{sea_foam}{HTML}{c4dfe6}
+%\colorlet{base}{wave}
+%\colorlet{bright}{ocean}
+%\colorlet{dark}{deep_aqua}
+%
+%% Invigorating
+%% https://visme.co/blog/wp-content/uploads/2016/01/26-1024x512.jpg
+%\definecolor{cDarkBlue}{HTML}{003b46}
+%\definecolor{cBlue}{HTML}{07575b}
+%\definecolor{cLightBlue}{HTML}{66a5ad}
+%\definecolor{cYellow}{HTML}{c4dfe6}
+%\definecolor{cRed}{HTML}{c4dfe6}
+%\colorlet{base}{wave}
+%\colorlet{bright}{ocean}
+%\colorlet{dark}{deep_aqua}
+
+%% Invigorating
+%% https://visme.co/blog/wp-content/uploads/2016/01/26-1024x512.jpg
+%\definecolor{cBlue}{HTML}{80add7}
+%\definecolor{cGreen}{HTML}{0abda0}
+%\definecolor{cWhite}{HTML}{ebf2ea}
+%\definecolor{cYellow}{HTML}{d4dca9}
+%\definecolor{cBrown}{HTML}{bf9d7a}
+%\colorlet{cBase}{cBlue}
+%\colorlet{cBaseMark}{cGreen}
+%\colorlet{cBrightText}{cYellow}
+%\colorlet{cDarkText}{cBrown}
 
@@
 % \specialrule[width]{abovespace}{belowspace}
-\colorlet{tableheadcolor}{tabBlue}
+\colorlet{tableheadcolor}{cBase}
 \newlength\arulesep\setlength\arulesep{\aboverulesep}

### 2019-06-22 Simulator/article.tex
2019-05-19
2019-06-05
2019-06-21

--- 2018-10-11 Simulator/article.tex
@@
 
-\usepackage{tikz}
-\usetikzlibrary{calc,positioning,shapes,snakes}
-%\newcommand{\tikzmark}[1]{\tikz[remember picture]\coordinate(#1){(-.35em,-.545ex)};}
-\newcommand{\tikzmark}[2]{\tikz[remember picture,baseline=(#1.base)]\node[shape=rectangle,inner sep=0pt](#1){#2};}
-\newcommand*\circled[1]{\tikz[baseline=(char.base)]{
-            \node[shape=circle,draw,inner sep=2pt] (char) {#1};}}
+% Exotic orchids
+% https://visme.co/blog/wp-content/uploads/2016/01/63-1024x512.jpg
+%\definecolor{cDarkBlue}{HTML}{192e5b}
+%\definecolor{cBlue}{HTML}{1d65a6}
+%\definecolor{cLightBlue}{HTML}{73a2c0}
+%\definecolor{cGreen}{HTML}{00743f}
+%\definecolor{cYellow}{HTML}{f2a104}
+%\colorlet{cBase}{cLightBlue}
+%\colorlet{cBaseMark}{cGreen}
+%\colorlet{cPoint}{cYellow}
+%\colorlet{cBasis}{cDarkBlue}
+
+\ifOnlineColor
+   \definecolor{tabBlue}{HTML}{AACCFF}
+\else
+   \definecolor{tabBlue}{HTML}{CCCCCC}
+\fi
+
+%\titleformat{\subsubsection}[hang]{\itshape}{\thesubsubsection}{0.6em}{}
+
+%\makeatletter
+%\xpatchcmd{\@sect}{%
+%  \protected@edef\@svsec{\@seccntformat{#1}\relax}%
+%}{%
+%  \ifnum0=\pdfstrcmp{#1}{section}%
+%  \def\@svsec{\circlednumber{\number\value{section}}\quad\relax}%
+%  \else
+%  \protected@edef\@svsec{\@seccntformat{#1}\relax}%
+%  \fi
+%  }{}{}
+%\makeatother
+
+%\titleformat{\subsubsection}{\Large\bfseries}{}{0em}{%
+%\begin{picture}(0,0)
+%\put(12,4){\circle{28}}
+%\end{picture}\makebox[1.45em][c]{\thesubsubsection}\quad}
 
@@
 \usepackage{tabularx}
+\usepackage{collcell}
+\usepackage{booktabs}
+\usepackage{calc}
+%\usepackage{ifthen}
 
@@
 
+\usepackage{listings}
+%\usepackage[outputdir=build]{minted}
+
+\lstset{
+  %These 3 parameters should be re-specified for each listing:
+   language=bash%
+ , caption=%
+ , label=%
+% , extendedchars=false
+%%  , inputencoding=utf8
+%  %
+%%  ,  fancyvrb=false % Beamer hates fancyvrb
+%%   morecomment=[1]\ ,
+%%   fvcmdparams=<\frac><2>,
+% , backgroundcolor=\color{codeBackground}%
+%  %
+ , basicstyle=\small
+% , keywordstyle=\color{DarkBlue}%
+%% , identifierstyle=\color{green}%
+% , commentstyle=\color{ForestGreen}\itshape%
+%  %morecomment=[|]{//}
+% , stringstyle=\color{Brown}%
+%  %
+%% , showspaces=false%
+%% , showstringspaces=false%
+%% , showtabs=false%
+%  %
+%% , frame=single                            %single, shadowbox, framed
+ , frame=bt
+%%  , xtopmargin=0pt%
+%%  , xbottommargin=0pt%
+%% , framesep=20pt
+%% , rulesep=20pt
+%% , lineskip=1ex
+ , aboveskip=\baselineskip % space above listing
+ , belowskip=\baselineskip % space below listing
+ , framexleftmargin=1em%
+% , framexrightmargin=0pt%
+% , linewidth=\linewidth%
+ , xleftmargin=1em%        %To vertically align the box with the rest of the text in the
+% , xrightmargin=0em%       %document, rather than having the text aligned and the box overshoot.
+% , tabsize=2%
+% , abovecaptionskip=1\baselineskip
+% , belowcaptionskip=0pt
+%  %  
+ , captionpos=b%     %Caption at bottom
+%  % 
+% , breaklines=true%  
+% , breakatwhitespace=false%
+%  %
+ , mathescape=true%
+%% , literate=*{-}{-}1
+% , columns=fullflexible
+%% , float
+%% , extendedchars=false
+%% , escapeinside={\$}{\$}
+%%   escapebegin={|},
+%%   escapeend={|},
+%% , inputencoding=utf8x
+%  %
+%% , numbers=left%
+%% , numberstyle=\tiny\color{SeaGreen}%
+%% , stepnumber=1%
+%% , numbersep=2em 
+ }
+
 \newcommand\gls[1]{}
@@
 
-\ifOnlineColor
-   \definecolor{tabBlue}{HTML}{AACCFF}
-\else
-   \definecolor{tabBlue}{HTML}{CCCCCC}
-\fi
+%% Cool blues (sea)
+%\definecolor{deep_aqua}{HTML}{003b46}
+%\definecolor{ocean}{HTML}{07575b}
+%\definecolor{wave}{HTML}{66a5ad}
+%\definecolor{sea_foam}{HTML}{c4dfe6}
+%\colorlet{base}{wave}
+%\colorlet{bright}{ocean}
+%\colorlet{dark}{deep_aqua}
+%
+%% Invigorating
+%% https://visme.co/blog/wp-content/uploads/2016/01/26-1024x512.jpg
+%\definecolor{cDarkBlue}{HTML}{003b46}
+%\definecolor{cBlue}{HTML}{07575b}
+%\definecolor{cLightBlue}{HTML}{66a5ad}
+%\definecolor{cYellow}{HTML}{c4dfe6}
+%\definecolor{cRed}{HTML}{c4dfe6}
+%\colorlet{base}{wave}
+%\colorlet{bright}{ocean}
+%\colorlet{dark}{deep_aqua}
+
+%% Invigorating
+%% https://visme.co/blog/wp-content/uploads/2016/01/26-1024x512.jpg
+%\definecolor{cBlue}{HTML}{80add7}
+%\definecolor{cGreen}{HTML}{0abda0}
+%\definecolor{cWhite}{HTML}{ebf2ea}
+%\definecolor{cYellow}{HTML}{d4dca9}
+%\definecolor{cBrown}{HTML}{bf9d7a}
+%\colorlet{cBase}{cBlue}
+%\colorlet{cBaseMark}{cGreen}
+%\colorlet{cPoint}{cYellow}
+%\colorlet{cBasis}{cBrown}
 

### 2019-06-22 ThesisJo/frontmatter.tex
2019-06-22

@@
 \begin{document}
+\pagestyle{normal}
 
@@
 \chapter*{Preface}
+% \cfoot{\thepage}
 
@@
 
-\newcommand\pubEntry[4]{\item #2, "#1", \textit{#3}, #4}
+\newcommand\pubEntry[4]{\item #2\\"#1"\\\textit{#3}\\#4}
 {\renewcommand{\theenumi}{\textbf{\Roman{enumi}}}
@@
 \pubEntry%
-{A Real-Time Synthetic Aperture Sonar Simulator for Automatic Target Recognition: Notation and application}%
+{Real-Time Synthetic Aperture Sonar Simulation for Automatic Target Recognition: Notation and Application}%
 {J.I. Buskenes, H. Midelfart}%
 {IEEE Journal of Oceanic Engineering}%
-{Submitted DD.MM.2018}
+{Submitted DD.MM.2019}
 \end{enumerate}

### 2019-06-27 LCAforActiveSonarImaging/document.tex
2019-06-26

@@
 \ifPhdDoc
-   \begin{figure*}[tbp]\centering%
-   \includegraphics[width=\linewidth]{gfx/fig3_phd.pdf}%
+   \begin{figure*}[t]\centering%
+   \makebox[\linewidth][c]{\includegraphics[width=1.25\linewidth]{gfx/fig3_phd.pdf}}
    \caption{\emph{Determining LCA window type}: MVDR image with typical frequency responses for windows used in various pixel regions.\\[.5\baselineskip]
@@
    \begin{figure*}[tbp]\centering%
-   \includegraphics[width=\linewidth]{gfx/fig4.pdf}%
+   \makebox[\linewidth][c]{\includegraphics[width=1.25\linewidth]{gfx/fig4.pdf}}
+%    \includegraphics[width=\linewidth]{gfx/fig4.pdf}%
    \caption{\emph{Determining Kaiser parameter boundaries}: LCA images of the resolution test cross and anchor created with a large window database, but where the upper bounds for $\beta$ and $\phi$ are varied.
@@
 Note how LCA prefers a narrow response ($\beta=0$) on the anchor, and on other strong sources with little lateral interference.
-On the sources where lateral interference is present it chooses the widest response with the best sidelobe suppression.]{\includegraphics[width=\textwidth]{gfx/fig5a.pdf}%%
+On the sources where lateral interference is present it chooses the widest response with the best sidelobe suppression.]{\makebox[\linewidth][c]{\includegraphics[width=1.25\linewidth]{gfx/fig5a.pdf}}%%
    \label{III_oversampling_mosaic_bounds_beta}}
@@
 On strong sources LCA selects windows that are steered away from the source.
-This is what improves the FWHM measurement in \Fig{III_oversampling_mosaic_bounds}.]{\includegraphics[width=\textwidth]{gfx/fig5b_phd.pdf}%
+This is what improves the FWHM measurement in \Fig{III_oversampling_mosaic_bounds}.]{\makebox[\linewidth][c]{\includegraphics[width=1.25\linewidth]{gfx/fig5b_phd.pdf}}%
    \label{III_oversampling_mosaic_bounds_phi}}
@@
    \begin{figure*}[tb]\centering%
-   \includegraphics[width=\linewidth]{gfx/fig6.pdf}%
+   \makebox[\linewidth][c]{\includegraphics[width=1.25\linewidth]{gfx/fig6.pdf}}%
 \else
@@
 Note how the FWHM of LCA sits in between that of DAS and MVDR.
-Also note how LCA is insensitive to the window type; its performance is near identical whether it uses Kasier or trigonometric windows.]{\includegraphics[width=\textwidth]{gfx/fig7a.pdf}%
+Also note how LCA is insensitive to the window type; its performance is near identical whether it uses Kasier or trigonometric windows.]{\makebox[\linewidth][c]{\includegraphics[width=1.25\textwidth]{gfx/fig7a.pdf}}%
    \label{III_beamformer_comparison_cross}}
@@
 Compared to DAS the adaptive beamformers produce deeper shadows, sharper edges and a higher detail level.
-The adaptive methods produce nearly identical images.]{\includegraphics[width=\textwidth]{gfx/fig7b_phd.pdf}%%
+The adaptive methods produce nearly identical images.]{\makebox[\linewidth][c]{\includegraphics[width=1.25\textwidth]{gfx/fig7b_phd.pdf}}%%
    \label{III_beamformer_comparison_holmengraa}}
@@
 \appendix
-\renewcommand\thesection{\Roman{section}}
+\renewcommand\thesection{\Alph{section}}
+\renewcommand\thesubsection{\Alph{section}.\arabic{subsection}}
 \else
@@
    \printbibliography[title=References,heading=subbibliography]
+   \addcontentsline{toc}{section}{References}
 %    \bibliographysty

### 2019-06-27 ThesisJo/2_publications.tex
2019-06-26

@@
 
-The upcoming papers will investigate various ways in which acoustical imaging systems can be improved by making them smarter in an operational sense.
+%The upcoming papers will investigate various ways in which acoustical imaging systems can be improved by making them smarter in an operational sense.
 This involves granting them the ability to adapt their behavior to the information at hand, and the processing power needed to handle the information fast enough.
 
-%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
-\section{Intro}\label{sec:intro} %
-%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
+%%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
+%\section{Intro}\label{sec:intro} %
+%%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
 
@@
 
-Common:
-\begin{itemize}\setlength\baselineskip{0pt}
-\item Adaptive
-\item Processing speed
-\end{itemize}
+% Classification accuracy
+% Logarithmic loss
+% Confusion matrix
+% Area under curve
+% F1 score
+% Mean absolute error
+% Mean squared error
+%
+%
+
+We present four journal-papers concerned with the \emph{adaptive} and \emph{hardware accelerated} formation of acoustic images: the first three reconstructs them from acoustic echoes, and the last synthesizes them through acoustic simulation.
+
+%
+%Common:
+%
+%GPUs
+%Experimental validation on HISAS1030 data.
+
+%\begin{itemize}\setlength\baselineskip{0pt}
+%\item \emph{Optimization method}, or criteria for adaptiveness and constraints;
+%\item \emph{Solution space}, or range of possible values for an optimization problem that satisfy a problem's constraints;
+%\item \emph{Domain}, or data representation, 
+%\item \emph{Application}, 
+%\item \emph{Computational complexity}, $O(N^3)$.
+%\item Hardware acceleration. 
+%\item Information significance: Minimal temporal sample support (~single snapshot)
+%\end{itemize}
+%
+%All:
+%- GPU
 
@@
 %~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
-\textbf{An Optimized GPU implementation of the MVDR Beamformer for Active Sonar Imaging}\\
+\textbf{An Optimized GPU implementation of the\\ MVDR Beamformer for Active Sonar Imaging}\\
 J.I. Buskenes, J.P. Åsen, C.-I.C. Nilsen and A. Austeng\\
@@
 
-\begin{itemize}\setlength\baselineskip{0pt}
-\item Optimization: Minimum variance
-\item Solution space: Continuous, closed when made robust
-\item Time domain
-\item Sonar
-\item Computationally complex, $O(N^3)$.
-\item Information significance: Minimal temporal sample support (~single snapshot)
-\end{itemize}
+This article evaluates time domain MVDR on GPUs for use in active sonar imaging.
+We refer to MVDR as fully adaptive, because it computes array weights analytically from spatial features of the data, yielding a continuous solution space.
+Like any data-driven algorithm, it requires statistically distinguishable signal and noise.
+However, in typical active systems, the signal is highly correlated with the noise and has poor temporal sample support.
+Therefore, in its unconstrained form, MVDR overfits the data, yields high variance and fails to generate representative images.
+
+We relax MVDR through a combination of temporal and spatial averaging, and regularization.
+This makes it robust and produces images with slightly more details and less noise than delay-and-sum.
+The resulting solution space is still continuous, but tighter bounded.
+
+We find that implementing MVDR on GPUs instead of CPUs accelerates it by two orders of magnitude, but requires considerable and delicate efforts to reduce arithmetic operations and fine-tune cache usage.
+The design is limited to 32 channels due data dependencies that are complicated and not inherently parallel, combined with the relatively limited bandwidth of this hardware architecture.
+
+%
+%%\begin{itemize}\setlength\baselineskip{0pt}
+%\item Solution space is continuous but strictly constrained due to robustification,
+%\item Time domain
+%\item Computationally complex, $O(N^3)$
+%\item Information significance: Minimal temporal sample support (~single snapshot)
+%\end{itemize}
 
@@
 %
-%The first paper is concerned with applying the Minimum Variance Distortionless Response (MVDR) beamformer to active sonar imaging, with emphasis on performing arithmetic reductions and solving the practical its high computational complexity.
-It shows that  with emphasis on solving and with the process of implementing this method on a \gls{GPU}.
-The goal is to improve sonar image quality without suffering the consequences of the MVDR's inherently high computational complexity.
-%
-%Combined, these steps led to a speedup of an order of magnitude or more compared to its straight-forward \gls{CPU} implementation.
-%
+
 %In functionality this method acts as our reference method, but its practicality is limited by a high computational complexity.
@@
 
-%Applying the MVDR-beamformer may not seem like fully adaptive technique 
-%The Minimum Variance Distortionless Response (MVDR) beamformer is a technique predominant is a fully adaptive method that 
-%
 %
@@
 
-\begin{itemize}
-\item Minimum variance
-\item Full search space, continuous
-\item Beam-space
-\item Ultrasound
-\end{itemize}
+This article evaluates frequency domain (or beamspace) MVDR on GPUs for use in cardiac ultrasound imaging.
+These systems typically have real-time requirements and more than 32 channels, so we apply a space reduction technique to reduce the computational effort.
+Since the transmitted beams are narrow, most acoustic energy falls within three to five center frequencies, and a good approximation follows from just computing those.
+
+We find that beamspace MVDR delivers real-time performance independent of channel-count and an image quality indistinguishable from time domain MVDR.
+
+%The first article implements the fully adaptive MVDR beamformer on a GPU for use in active sonar imaging.
+Like any data-driven algorithm, it requires data with statistically distinguishable signal and noise.
+However, in typical active systems, the signal is highly correlated with the noise and has weak temporal sample support.
+Therefore, MVDR in its unconstrained form overfits the data, yields high variance and fails.
+%
+%We relax MVDR through a combination of temporal and spatial averaging, and regularization. This makes it robust and produces images of slightly higher quality than delay-and-sum. 
+%
+%Implementing MVDR on a GPU instead of a CPU led to a speed-up of two orders of magnitude, but required considerable and delicate efforts to reduce arithmetic operations and fine-tune cache usage.
+The data dependencies are complicated and not inherently parallel.
 
@@
 
-\begin{itemize}
-\item Minimum variance
-\item Smaller search space, discrete
-\item Element-space
-\item sonar
-\end{itemize}
+This article describes LCA, an alternative to MVDR that iterates through precomputed sets of array weights and picks the one that best fulfills the minimum variance criterion.
+It can be viewed as either a simplification of MVDR or as an adaptive extension to DAS.
+
+We find that constrained MVDR typically generates spatial responses between that of uniform weights and Hamming weights.
+When LCA is configured with six weight sets covering this range, it matches MVDR in image quality.
+As a bonus, it is inherently robust and easy to work with, understand and accelerate with GPUs.
 
@@
 \textit{IEEE Journal of Oceanic Engineering}\\
-Submitted DD.MM.2018%\\\\
+Submitted DD.MM.2019%\\\\
 %Volume 42, Issue 1, January 2017
 
-\begin{itemize}
-\item Template-generation
-\item Autonomy
-\item Sonar
-	\item Incredibly fast template matching possible.
-	\item Possibly a powerful tool to pass prior knowledge to deep learning
-	\item Extracting the essential information is key 
-	\item Exploit hardware resources available
-\end{itemize}
+This article describes an acoustic simulator for SAS, implemented on GPUs using OpenGL and OpenCL.
+Its purpose is to enable AUVs to compare its images on-the-fly to simulations of various known objects, for more responsive, accurate and flexible automatic target recognition (ATR).
 
-\section{Onlook}
+We find that simulating objects on-the-fly instead of pre-simulating them consistently improves classification accuracy at false positive rates (FPR) below 20\%, and ties above.
 
-Machine cognition:
-\begin{itemize}
-\item Ethical dilemmas. Decision and actions have consequences. Who's legally responsible for a machine's doing? Itself?
-\item Safety. Machine flexibility allow more possible outcomes, and a wider range of security issues to consider.
-\end{itemize}
+We discuss the design in-depth using a new navigational notation system, extended with topological concepts from robotics and homogeneous coordinates.
+We find it incredibly helpful for keeping track of complex geometries, numerical representations and higher level abstractions such as manipulating rigid objects.
 
-R dim > 10000
-- Inverting such matrices better to do with gradient decent
+%Acoustic image reconstruction and synthesis are closely related, and
+
+%
+%
+%
+%\begin{itemize}
+%\item Template-generation
+%\item Autonomy
+%\item Sonar
+%	\item Incredibly fast template matching possible.
+%	\item Possibly a powerful tool to pass prior knowledge to deep learning
+%	\item Extracting the essential information is key 
+%	\item Exploit hardware resources available
+%\end{itemize}
 

### 2019-06-28 ThesisJo/2_publications.tex
2019-06-27

@@
 
-We present four journal-papers concerned with the \emph{adaptive} and \emph{hardware accelerated} formation of acoustic images: the first three reconstructs them from acoustic echoes, and the last synthesizes them through acoustic simulation.
+We present four journal published papers concerned with \emph{adaptive} and \emph{hardware accelerated} formation of acoustic images: the first three reconstruct them from acoustic echoes and the last synthesizes them through acoustic simulation.
 
@@
 This makes it robust and produces images with slightly more details and less noise than delay-and-sum.
-The resulting solution space is still continuous, but tighter bounded.
+The resulting solution space is still continuous, but more tightly bound.
 
 We find that implementing MVDR on GPUs instead of CPUs accelerates it by two orders of magnitude, but requires considerable and delicate efforts to reduce arithmetic operations and fine-tune cache usage.
-The design is limited to 32 channels due data dependencies that are complicated and not inherently parallel, combined with the relatively limited bandwidth of this hardware architecture.
+The design is limited to 32 channels due to data dependencies that are complicated and not inherently parallel, combined with the relatively limited bandwidth of this hardware architecture.
 
@@
 This article describes an acoustic simulator for SAS, implemented on GPUs using OpenGL and OpenCL.
-Its purpose is to enable AUVs to compare its images on-the-fly to simulations of various known objects, for more responsive, accurate and flexible automatic target recognition (ATR).
+It enables AUVs to compare its images on-the-fly to simulations of various known objects, for more responsive, accurate and flexible automatic target recognition (ATR).
 
@@
 
-We discuss the design in-depth using a new navigational notation system, extended with topological concepts from robotics and homogeneous coordinates.
+The design is discussed in-depth using a new navigational notation system, extended with topological concepts from robotics and homogeneous coordinates.
 We find it incredibly helpful for keeping track of complex geometries, numerical representations and higher level abstractions such as manipulating rigid objects.

### 2019-06-28 ThesisJo/frontmatter.tex
2019-06-26
2019-06-27

@@
 \endofdump
+
+\cleardoublepage
+
+\vspace*{\fill}
+\begin{center}
+{\huge\bf Til Joakim} \\[1.5cm]
+{\Large aldri gi opp}
+\end{center}
+\vspace*{\fill}
+
+\clearpage
 
@@
 
-\begin{markdown}
-Hey
-\end{markdown}
-
 \chapter*{Preface}
+\addcontentsline{toc}{chapter}{Preface}
 % \cfoot{\thepage}
@@
 
+This journey seems to have started a different life ago.
+Along the way more appreciation has piled up in the depths of my heart than I believed possible.
+I can only hope to give a little back by extending my sincerest gratitude to
+
+\emph{Professor Andreas Austeng}, for offering great advice, insights, ever-presence, persistence and unwaivering optimism---simply being the best supervisor imaginable;
+
+\emph{Jon Petter Åsen}, for being a wonderful and joyful office mate, colleague, coauthor and friend during the first years;
+
+\emph{Cherie}, my dear faraway friend, for the steady stream of encouraging thoughts, patience and help---asking nothing in return;
+
+\emph{the Department of Informatics, IFE}, for initially granting me this opportunity and later extending it;
+
+\emph{the Norwegian Defence Research Institute, FFI}, for providing data to study and being an amazing workplace;
+
+\emph{Professor Roy Hansen} (FFI/IFE), for initially sparking my interest for sonars and this PhD;
+
+\emph{Professor Sverre Holm} (FFI/IFE), for being my second supervisor, an inspiration, a bottomless pit of knowledge and an aid while I lectured signal processing;
+
+\emph{Stig Synnes} (FFI/IFE), for getting me into volleyball and the mutual friendly nudges to finish writing;
+
+\emph{Kenneth Gade} (FFI), for volleyball enthusiasm and the inspiration and help developing the notation in paper IV,
+
+\emph{Carl-Inge C. Nielsen} (IFE), for the time, knowledge and flying start you offered when this all began;
+
+\emph{Herman Midelfart (FFI)}, for getting a chance to work with you before your days were numbered;
+
+\emph{other colleagues and students} at IFE and FFI, for making work and life enjoyable, every day;
+
+\emph{Friskvernklinikken}, Asker, for showing me a path where I thought there were none;
+
+\emph{Marthe Arnesen}, for the gesture you once made yourself, for what is and for what may come;
+
+\emph{my family}, for all your support, love and presence---through good and bad;
+
+(\emph{myself}, I could not have done it without you;)
+
+\emph{Joakim}, for giving me purpose, perspective and joy. This is for you.
+
 \newpage
@@
 
+\textit{Cognition:}\\The mental action or process or acquiring knowledge and understanding through thought, experience, and the senses.
+
 \newpage
-\chapter*{List of Publications}
+\chapter*{Included publications}
+\addcontentsline{toc}{chapter}{Included publications}
 
@@
 %\newpage
-\chapter*{Related Publications}
+\chapter*{Excluded Publications}
+\addcontentsline{toc}{chapter}{Excluded publications}
 

### 2019-06-30 ThesisJo/2_publications.tex
2019-06-29

@@
 
-We present four journal published papers concerned with \emph{adaptive} and \emph{hardware accelerated} formation of acoustic images: the first three reconstruct them from acoustic echoes and the last synthesizes them through acoustic simulation.
+We present four journal published papers concerned with \emph{adaptive} and \emph{hardware accelerated} formation of acoustic images: the first three reconstruct from echoes and the last synthesizes through simulation.
 

### 2019-06-30 ThesisJo/frontmatter.tex
2019-06-29

@@
 Along the way more appreciation has piled up in the depths of my heart than I believed possible.
-I can only hope to give a little back by extending my sincerest gratitude to
+I can only hope to give a little back by extending my sincerest gratitude to the following:
 
@@
 
-\emph{the Department of Informatics, IFE}, for initially granting me this opportunity and later extending it;
+\emph{The Department of Informatics, IFE}, for initially granting me this opportunity and later extending it;
 
-\emph{the Norwegian Defence Research Institute, FFI}, for providing data to study and being an amazing workplace;
+\emph{The Norwegian Defence Research Institute, FFI}, for providing data to study and being an amazing workplace;
 
@@
 
-\emph{my family}, for all your support, love and presence---through good and bad;
+\emph{My family}, for all your support, love and presence---through good and bad;
 
-(\emph{myself}, I could not have done it without you;)
+\emph{Myself}, I could not have done it without you ;)
 
@@
 
-\textit{Cognition:}\\The mental action or process or acquiring knowledge and understanding through thought, experience, and the senses.
+\textit{Cognition:}\\The mental action or process of acquiring knowledge and understanding through thought, experience and the senses.
 

### 2019-08-19 ThesisJo/2_background.tex
2019-08-18

--- 2013-01-23 ASA2013/asa2013.tex
@@
-\newif\ifTODO\TODOfalse                        % Use todo notes?
+%&header
+% !TeX root = 1_introduction
 
-%===============%                               %~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~%
-% DOCUMENTCLASS %                                See full option description in "mytemplate.cls"
-%===============%                               %~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~%
+\input{subheader}
+\endofdump
 
-\documentclass[
-a4paper,10pt
-]{ica2013_2}
-
-% \usepackage[utf8x]{inputenc}
-\usepackage[table,dvipsnames,svgnames]{xcolor}
-\usepackage{titlesec}
-
-\newcommand\Fig[1]{Fig. \ref{#1}}
-
-\newcommand\Grey[1]{{\color{Grey}#1}}
-\newcommand\Red[1]{{\color{Red}#1}}
-\newcommand\Blue[1]{{\color{Blue}#1}}
-\newcommand\DarkBlue[1]{{\color{DarkBlue}#1}}
-\newcommand\LightBlue[1]{{\color{LightBlue}#1}}
-\newcommand\Brown[1]{{\color{Brown}#1}}
-\newcommand\Green[1]{{\color{Green}#1}}
-\newcommand\SeaGreen[1]{{\color{SeaGreen}#1}}
-\newcommand\Yellow[1]{{\color{yellow}#1}}
-\newcommand\Orange[1]{{\color{orange}#1}}
-
-\ifTODO
-   \newcounter{todoidx}
-   \definecolor{todobackground}{rgb}{0.95,0.95,0.95}
-   \setlength\marginparsep{1pt}
-   \setlength\marginparwidth{50pt}
-   \newlength\marginparwidthsmall
-   \setlength\marginparwidthsmall{\marginparwidth}
-   \addtolength\marginparwidthsmall{-7pt}
-   \newcommand\todo[1]{%
-      \addtocounter{todoidx}{1}%
-      {\color{Red}\bf(\thetodoidx{})}%%\fbox{\bf\thetodoidx{}}}%
-      \marginpar{%
-         {\vspace*{-10pt}\color{Red}\fbox{\bf\thetodoidx{}}}\\%
-         \fcolorbox{red}{todobackground}{\parbox{\marginparwidthsmall}{\scriptsize #1}}}}
-
-   \newcommand\todopar[1]{\fcolorbox{red}{white}{\parbox{0.97\linewidth}{#1}}}
-\else
-%    \usepackage[disable]{./todonotes} 
-   \newcommand\todo[1]{}
+\ifRootBuild\else
+  \input{glossary}
+  \makeglossaries
 \fi
 
-\newcommand\nn{\nonumber\\}
-
-\newcommand\nmat[1]{\begin{matrix}#1\end{matrix}}
-\newcommand\bmat[1]{\begin{bmatrix}#1\end{bmatrix}}
-\newcommand\case[1]{\begin{cases}#1\end{cases}}
-\newcommand\textbox[2]{\footnotesize\text{\parbox{#1}{\centering\emph{#2}}}}
-
-\newcommand\rand{\text{rand}}
-\newcommand\randn{\text{randn}}
-\newcommand\rect{\text{rect}}
-\newcommand\sinc{\text{sinc}}
-\newcommand\tr{\text{tr}}
-\newcommand\adj{\text{adj}}
-
-% \newcommand\max{\text{max}}
-\newcommand\argmin[1]{\text{arg}\,\underset{#1}{\text{min}}}
-
-\newcommand\qqquad{\quad\qquad}
-\newcommand\qqqquad{\qquad\qquad}
-
-\renewcommand\l[1]{\left#1}
-\renewcommand\r[1]{\right#1}
-
-% {\text{\parbox{1.5cm}{\centering volume hyper- sphere}}}
-
-%Keyword colouring:
-\newcommand\kw[1]{#1}
-\newcommand\parm[1]{#1}%\color{Black}#1\color{Black}}
-
-\newcommand\of[1]{\scriptstyle(\parm{#1})\displaystyle}
-\newcommand\df[1]{\scriptstyle[\parm{#1}]\displaystyle}
-\newcommand\var[3]{#1_\text{#2}\of{#3}}
-
-\newcommand\diag{\text{diag}}
-
-% \raisebox{lift}[extend-above-baseline][extend-below-baseline]{text}
-\newcommand\mt[1]{\text{\emph{#1}}} %mt = mathtext
-\newcommand\mathnorm{\textstyle}
-\newcommand\mathbig[1]{\displaystyle#1\mathnorm}
-\newcommand\mathsmall[1]{\scriptstyle#1\mathnorm}
-\newcommand\mathtiny[1]{\scriptscriptstyle#1\mathnorm}
-\newcommand\sfrac[2]{\scriptstyle\raisebox{0.25pt}[0pt][0pt]{$\frac{#1}{#2}$}\mathnorm}
-\newcommand\nfrac[2]{\textstyle\frac{#1}{#2}\displaystyle}
-
-\newcommand\sumu[1]{\sum\limits^{#1}\,}
-\newcommand\suml[1]{\sum\limits_{#1}\,}
-\newcommand\sumb[2]{\sum\limits_{#1}^{#2}\,}
-
-\newcommand\produ[1]{\prod\limits^{#1}\,}
-\newcommand\prodl[1]{\prod\limits_{#1}\,}
-\newcommand\prodb[2]{\prod\limits_{#1}^{#2}\,}
-
-\newcommand\defeq{\overset{\underset{\mathrm{def}}{}}{=}}
-
-%Math macros:
-\newcommand\diff[2]{\frac{\kw{d}\,\textstyle #1\scriptstyle}{\kw{d\parm{#2}}}\displaystyle}
-\newcommand\ddiff[2]{\frac{\kw{d^2}\,\displaystyle #1\scriptstyle}{\kw{d\parm{#2}}^2}\displaystyle}
-
-\renewcommand\d[1]{\scriptstyle\kw{\,d\parm{#1}}\displaystyle}
-
-% These commands are mutually exclusive. Remember to "renew" in v2.
-\newcommand\intb[4]{\int\limits_{#3}^{#4} #1 \d{#2}} % \int{exp}{var}{from}{to}
-\newcommand\intl[3]{\int\limits_{#3} #1 \d{#2}} % \int{exp}{var}{for all}
-\newcommand\intu[2]{\int #1 \d{#2}} % \int{exp}{var}{for all}
-
-\newcommand\T{^{\scriptscriptstyle T}}
-\renewcommand\H{^{\scriptscriptstyle H}}
-
-\renewcommand\vec[1]{\boldsymbol{#1}}
-\newcommand\mat[1]{\boldsymbol{#1}}
-
-\newcommand\1{\vec 1}
-\newcommand\I{\mat I}
-\renewcommand*\a{\vec a}
-\renewcommand*\i{\vec i}
-\renewcommand*\k{\vec k}
-\newcommand*\n{\vec n}
-\newcommand*\p{\vec p}
-\newcommand*\s{\vec s}
-\newcommand*\w{\vec w}
-\newcommand*\x{\vec x}
-\newcommand*\y{\vec y}
-
-\newcommand*\A{\mat A}
-\newcommand*\B{\mat B}
-\newcommand*\C{\mat C}
-\newcommand*\E{\mat E}
-% \renewcommand*\H{\mat H}
-\renewcommand*\P{\mat P}
-\newcommand*\eP{\mat{\hat P}}
-\newcommand*\R{\mat R}
-\newcommand*\Ri{\R^{-1}}
-\newcommand*\eR{\mat{\hat R}}
-\newcommand*\eRi{\hat{\mat R}\,\!^{-1}}
-\newcommand*\Navg{N_\text{avg}}
-\newcommand*\W{\mat W}
-\newcommand*\X{\mat X}
-\newcommand*\Xd{\X_{\!\Delta}}
-\newcommand*\Y{\mat Y}
-\renewcommand*\B{\mat B}
-
-\renewcommand*\L{\mat \Lambda}
-\newcommand*\U{\mat U}
-% \renewcommand*\t{\mathtiny{^T}}
-% \newcommand*\h{\mathtiny{^H}}
-\renewcommand*\t{^T}
-\newcommand*\h{^H}
-
-\newcommand\D{\vec\nabla} %Del: Vector differential operator - nabla
-\newcommand\Dx{\vec\nabla\times}
-\newcommand\Dd{\vec\nabla\cdot}
-
-\usepackage{tikz}
-\usetikzlibrary{shapes,snakes}
-\usepackage{amsmath,amssymb}
-\usepackage{glossaries}
-\usepackage{common/slashbox}
-
-\newcommand\graphicsAI[2][]{%
-  \immediate\write18{./bin/laFigure #2 #1}%
-  \input{result}}%
-
-\definecolor{tabBlue}{HTML}{AACCFF}
-
-\newenvironment{outline}
-{\begin{itemize}}
-{\end{itemize}}
+%\beforeepigraphskip
 
 \begin{document}
-% \setlength{\headrulewidth}{0.0pt}
 
-%~~ TitlePage ~~%                               %~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~%
-\thispagestyle{empty}
-\pagenumbering{arabic} %Normal numbers
+\newcommand\chapref[2]{\hyperref[#1]{#2.\ref{#1}}}
 
-{\Large\bf Adapting the MVDR beamformer to a GPU for differently sized active sonar imaging systems.}
+\chapter{Background}
 
-\begin{abstract}
-The MVDR beamformer has been shown to improve active sonar image quality compared to conventional methods.
-Unfortunately, it is also significantly more computationally expensive because a spatial covariance matrix must be estimated and inverted for each image pixel.
-We target this challenge by altering and mapping the MVDR beamformer to a GPU, and suggest three different solutions depending on the system size.
+{
 
-For systems with relatively few channels, we suggest arithmetic optimizations for the estimation step, and show how a GPU can be used to yield image creation rates of more than 1\,Mpx/s.
-For larger systems we show that frequency domain processing is preferable, as this promotes high processing rates at a negligible reduction in image quality.
-These GPU implementations consistently reduced the runtime by 2-3 orders of magnitude compared to our reference C and Matlab implementations.
+% \newcommand\mat[1]{\boldsymbol{#1}}
+\newcommand\bmat[1]{\begin{bmatrix}#1\end{bmatrix}}
+\renewcommand*\eR{\mat{\hat R}}
+\renewcommand\1{\vec 1}
 
-For even larger systems we suggest employing the LCA beamformer.
-It does not calculate a weightset, but merely computes the beamformer output for each of a predefined set of weights, and selects the one that best fulfils the MVDR criterion.
-The LCA creates images with a quality comparable to MVDR, and it is perfectly suited for a GPU.
-\end{abstract}
-\vspace{2cm}
-(Forsiden genereres av ASA.)
-% \ \\
-% Total: 195 words (200 max)
-% 
-% \ \\
-% Noen tanker jeg har selv naa:
-% \begin{itemize}
-% \item Det mangler noe om bruksomraader.
-% \item Maa passe paa aa ikke selge LCA saa bra at MVDR ser helt haaplos ut(?)
-% \end{itemize}
-% 
-% 
-% \ \\
-% Questions they wanted answered:
-% \begin{itemize}
-% \item This session will feature work on adapting computationally demanding modeling and signal processing tasks to take advantage of various parallel architectures, such as cloud computing, computing clusters, multicore CPUs, multi-core graphic processing units (GPUs), DSP chips, and FPGAs.
-% \item If algorithm is slow - What technology will help me?
-% \item What skills are needed?
-% \item What tools are available and what's the cost?
-% \item What acceleration factor are we looking at?
-% \item How can I get started?
-% \item What we did. Which tools? Was it hard? Speedup?
-% \end{itemize}
-
-\newpage
 \section{Introduction}
 
-Graphics Processing Units (GPUs) is a technology tailored to perform image processing as fast as possible.
+% Graphics Processing Units (GPUs) is a technology tailored to perform image processing as fast as possible.
 They are available off-the-shelf at reasonable prices, are very power efficient, and can be programmed to carry out a wide range of computational tasks, often displaying speed increases of an order of magnitude compared to similar CPU designs.
@@
 While this is true for simple techniques such as conventional beamformers, the more complex per-pixel computations present in adaptive techniques are harder for the GPU to handle.
-
-Through a case study of on such technique, the Minimum Variance Distortionless Response (MVDR) beamformer, we identify these challenges and explain how they were overcome.
+% 
+% Through a case study of on such technique, the Minimum Variance Distortionless Response (MVDR) beamformer, we identify these challenges and explain how they were overcome.
 The MVDR technique is known for its notable potential for improving image quality over simple conventional methods, but it is also far more computationally complex.
 This is because a spatial covariance matrix must be estimated and inverted for each image pixel, the former dominating for systems with less than 20-30 channels, and the latter otherwise.
-
-We propose three different solutions to the complexity issue depending on system size.
+% 
+% We propose three different solutions to the complexity issue depending on system size.
 For smaller systems we use the standard time-domain MVDR method, arithmetically optimize the estimation step, and show that image creation rates of more than 1\,Mpx/s can be achieved when this algorithm is run on a GPU.
@@
 
-Related work can be found e.g.
+% Related work can be found e.g.
 in ultrasound imaging, where Chen et.
@@
 
-This article is outlined as follows: First we will present the mathematical formulation and evaluate complexity aspects of the MVDR method operated in time or frequency domain, and on the LCA beamformer.
+Waves are made of endless tiny energy transactions, a dance of particles effecting each other in a search for equilibrium.
+They are caused by disturbances in the forces of nature.
+The drop of a rock into flat water or rapid displacement of still air cause mechanical waves, while synchronized oscillations of electric and magnetic fields cause electromagnetic waves.
+Regardless of whether the waves are mechanical or electromagnetic, move perpendicular or parallel to the axis of propagation, what medium they propagate through or the means by which they do so, they are---perhaps surprisingly---all governed by a partial differential equation known as the \emph{wave equation}:
+%
+\begin{align}
+\Delta f = 
+\frac{\delta^2 f}{\delta x^2} + \frac{\delta^2 f}{\delta y^2} + \frac{\delta^2 f}{\delta z^2} = \frac{1}{c}\frac{\delta^2 f}{\delta t^2},
+\end{align}
+%
+where $\Delta$ is the Laplace operator, $f$ is field strength and $c\in\mathbb{R}^+$ is wave propagation speed.
+If $f$ represents the displacement of a vibrating string, the equation states that the force pulling the string towards equilibrium (right side) is proportional to the string's curvature (left side).
+In acoustics $f$ represents sound pressure, but since the mathematics are all the same we will simply think of $f$ as a signal: a unit-less scalar function with wave characteristics.
+
+Suppose an array of sensors has a reference frame $\ubar{A} \triangleq (\udot{A},\uvec{A})$ attached to it, where $\udot{A}\in\mathbb{A}$ is its absolute position and $\uvec{A}\triangleq\big(\hat{x}_{A},\hat{y}_{A},\hat{z}_{A}\big)\in\mathbb{V}^3$ its absolute orientation (where $\hat{\cdot}$ denotes unit length).
+A signal source is located at a point $\udot{S}$.
+We can represent the field observed at $\udot{A}$ as $f(\dvec{p}^A,t)$, where position is represented by $\dvec{p}^A=[x_A,y_A,z_A]^T\in\mathbb{R}^3$ and time by $t\in\mathbb{R}$,
+
+An introduction to such notation can be found in \chapref{IV_methods}{IV}.
+
+reference frame $\ubar{A} \triangleq (\udot{A},\uvec{A})$ to it, where $\udot{A}\in\mathbb{A}$ is its absolute position and $\uvec{A}\triangleq\big(\hat{x}_{A},\hat{y}_{A},\hat{z}_{A}\big)\in\mathbb{V}^3$ its absolute orientation (where $\hat{\cdot}$ denotes unit length)
+
+%
+\begin{align}
+f_\textrm{plane}(\dvec{p}^A,t)  &= A e^{j(\omega t - \dvec{k}^T \dvec{p}^A)} \\
+f_\textrm{spheric}(\dvec{p}^A,t) &= \frac{A}{\lVert \dvec{p}^A-\dvec{p}_{AS}^A \rVert} e^{j(\omega t - \dvec{k}^T \lVert \dvec{p}^A-\dvec{p}_{AS}^A \rVert)}
+\end{align}
+%
+% \hyperref[IV_methods]{\chapterref{IV_methods}.\sectionref{IV_methods}}
+% \renewcommand{\thesection}{\thechapter.\arabic{section}}
+
+% This article is outlined as follows: First we will present the mathematical formulation and evaluate complexity aspects of the MVDR method operated in time or frequency domain, and on the LCA beamformer.
 Then we mention some of our experiences with implementing the MVDR on a GPU, before concluding with some results and final thoughts.
 
-\section{Methods}\label{methods}
+\section{Image reconstruction}\label{methods}
 
-Suppose that a backscattered wavefield is recorded by an $M$ element uniform linear array, that a suitable set of delays are applied to the channels for the pixel of interest, and let the delayed output from the $m$'th channel be given as $x_m$.
-By definition a beamformer's output $z[n]$ at time instant $n$ is a weighted sum all the delayed data outputs:
+Consider a sonar imaging scenario where an encoded signal is transmitted to highlight a surface of interest, and assume that the backscattered wavefield is properly sampled using a uniform linear array of receivers.
+After stripping the carrier frequency by heterodyning and the signal signature by  matched filtering (Fig.
+\ref{I_beamforming}), an image can then be formed by coherently combining the receiver outputs to focus at one angle (or point) at the time.
+The principle of adjusting the array's focus in range and direction is commonly referred to as beamforming, and involves assigning suitable delays and weights to the array's channels.
+
+\subsection{Beamforming}
+% \ifPhdDoc
+% \begin{figure}[htbp]\centering
+% \includegraphics[width=\linewidth]{gfx/buske1\figPostfix.eps}
+% \else
+% \ifPeerReview
+% \begin{figure}[htbp]\centering
+% \includegraphics[width=0.8\linewidth]{gfx/buske1\figPostfix.eps}
+% \else
+% \begin{figure}[!t]\centering
+% \includegraphics[width=\linewidth]{gfx/beamforming.eps}
+% \fi\fi%
+% \caption{Beamforming principle.
+Signal signature is first removed by matched filtering.
+Then, before summation, a suitable set of delays $\Delta$ and weights $w$ are applied to focus on a pixel of interest at angle and range ($\theta,n$).}\label{I_beamforming}
+% \end{figure}
+
+Suppose we record a wavefield with $M$ sensors arranged in a uniformly spaced linear array (ULA) and attach a reference frame $\ubar{A} \triangleq (\udot{A},\uvec{A})$ to it, where $\udot{A}\in\mathbb{A}$ is its absolute position and $\uvec{A}\triangleq\big(\hat{x}_{A},\hat{y}_{A},\hat{z}_{A}\big)\in\mathbb{V}^3$ its absolute orientation (where $\hat{\cdot}$ denotes unit length).
+Assume that the array channels are digitally delayed to focus at a pixel with azimuth angle $\theta$ and range sample $n$, such that the delayed data from the $m$th channel can be expressed as $x_m[\theta,n-\Delta_m]\in\mathbb{C}$.
+We simply notation by making the dependence on $\theta$ implicit from now on.
+
+The beamformer output $z[n]\in\mathbb{C}$ is by definition the weighted sum of all the delayed data samples:
+%
 \begin{align}
-z[n] = \w\H[n]\x[n] = \bmat{w_0[n]\\w_1[n]\\\vdots\\w_{M-1}[n]}^H \bmat{x_0[n]\\x_1[n]\\\vdots\\x_{M-1}[n]},\label{z}
+z[n] = \dvec{w}^H[n]\dvec{x}[n] = \bmat{w_0[n]\\w_1[n]\\\vdots\\w_{M-1}[n]}^H \bmat{x_0[n]\\x_1[n]\\\vdots\\x_{M-1}[n]},\label{2_eq_z}
 \end{align}
-where $w_m$ is the weight factor assigned to channel $m$.
+%
+where $w_m\in\mathbb{C}$ is the weight factor assigned to channel $m$.
 With static weights this would be referred to as the conventional delay-and-sum (DAS) beamformer.
-Various weighting patterns exists here for trading lateral resolution for improved noise suppression (contrast), but one always ends up with a compromise between the two\todo{don't quite like this one}~\cite{Harris1978}.
+A large variety of weighting functions exists here for trading lateral resolution for improved noise suppression (contrast), but one always ends up with a compromise between the two~\cite{Harris1978}.
 However, there exists a class of adaptive beamformers that target this limitation by adapting the weights to the impinging wavefield.
@@
 
-\subsection{MVDR beamforming in the time domain}
+\subsection{Time-domain MVDR}
 
@@
 2. the term $\eR^{-1}\1$ is found using a linear equation solver,\newline
-3. and the beamformer output is computed from $z$ from (\ref{finalZ}), where $\w$ is found by substituting $\eR^{-1}\1$ into (\ref{weights}). }\label{mvdr_beamforming}
+3. and the beamformer output is computed from $z$ from (\ref{finalZ}), where $\w$ is found by substituting $\eR^{-1}\1$ into (\ref{2_eq_weights}). }\label{mvdr_beamforming}
 \end{figure}
-The basic idea behind the MVDR method is to compute the set of weights that minimizes the noise and signal power accumulated by the array, but with the constraint of unity gain for signals coming from the pixel of interest.
+%
+The basic idea behind the MVDR method is to compute the set of weights that minimizes noise and signal power accumulated by the array, while ensuring unity gain for signals coming from the pixel of interest.
 When the data is delayed this can be formulated as follows:
+%
 \begin{gather}
-\argmin{\w}\ E\Big\{\big|\w\H[n]\x[n]\big|^2\Big\} \qquad\text{subject to}\qquad \w\H[n]\1 = 1. \label{mvdr_def}
-% \vec w[n] = \frac{\Ri[n]\1}{\1\T\Ri[n]\1},\label{weights}
+\argmin{\dvec{w}}\ E\Big\{\big|\dvec{w}^H[n]\dvec{x}[n]\big|^2\Big\} \qquad\text{subject to}\qquad \dvec{w}^H[n]\dvec{1} = 1. \label{2_eq_mvdr_def}
+% \vec w[n] = \frac{\Ri[n]\1}{\1^T\Ri[n]\1},\label{2_eq_weights}
 \end{gather}
+%
 This is a convex optimization problem with the solution:
+%
 \begin{gather}
-\vec w[n] = \frac{\Ri[n]\1}{\1\T\Ri[n]\1},\label{weights}
+\vec w[n] = \frac{\Ri[n]\dvec{1}}{\dvec{1}^T\Ri[n]\dvec{1}},\label{2_eq_weights}
 \end{gather}
-where $\1$ is a row vector of ones that represents broadside steering, and $\R=E\{\x[n]\x\H[n]\}$ is the spatial covariance matrix for the full array.
+%
+where $\1$ is a row vector of ones that represents broadside steering, and $\R=E\{\x[n]\x^H[n]\}$ is the spatial covariance matrix for the full array.
 Since $\R$ is unknown, we estimate it by by computing a sample covariance matrix $\eR$.
@@
 If we let $x_l[n]$ represent the datavector from subarray $l$,
+%
 \begin{gather}
-\x_l[n] = \bmat{x_l[n] & x_{l+1}[n] & \dots & x_{l+L-1}[n]}\T.
+\dvec{x}_l[n] = \bmat{x_l[n] & x_{l+1}[n] & \dots & x_{l+L-1}[n]}^T.
 \end{gather}
+%
 then the sample covariance matrix formed with temporal and spatial averaging, $\eR$, is given as:
+%
 \begin{gather}
-\eR[n] =  \frac{1}{N_K N_L} \sumb{l=0}{N_L-1}\sumb{n'=n-K}{n+K} \x_l[n]\x_l\H[n] \in\mathbb{C}^{L,L},\label{spatialR}
+\eR[n] =  \frac{1}{N_K N_L} \sumb{l=0}{N_L-1}\sumb{n'=n-K}{n+K} \x_l[n]\x_l^H[n] \in\mathbb{C}^{L,L},\label{2_eq_spatialR}
 \end{gather}
+%
 where $N_K = 2K+1$ is the number of temporal samples to perform averaging over, and $N_L = M-L+1$ is the number of subarrays.
@@
 
-To summarize as in Figure \ref{mvdr_beamforming}: For every pixel in the image we estimate $\eR$ (\ref{spatialR}), solve the linear equation $\eRi\1$ (\ref{weights}), the finally substitute some to get the beamformer output (\ref{mvdr_def}-\ref{weights}).
+To summarize as in Figure \ref{mvdr_beamforming}: For every pixel in the image we estimate $\eR$ (\ref{2_eq_spatialR}), solve the linear equation $\eRi\1$ (\ref{2_eq_weights}), the finally substitute some to get the beamformer output (\ref{2_eq_mvdr_def}-\ref{2_eq_weights}).
 
@@
 
-If (\ref{spatialR}) were to be implemented directly, we note that the estimation step would have a complexity of O($N_K N_L L^2$).
+If (\ref{2_eq_spatialR}) were to be implemented directly, we note that the estimation step would have a complexity of O($N_K N_L L^2$).
 By comparison, performing a Gauss Jordan inversion is generally of O($L^3$), which implies that whenever $N_K\,N_L = N_K(M-L+1) \gg L$ the estimation step will dominate.
@@
 
-A few properties let us improve these matters notably. First, we may infer from (\ref{spatialR}) that:
+A few properties let us improve these matters notably. First, we may infer from (\ref{2_eq_spatialR}) that:
+%
 \begin{itemize}
-\item The subarray covariance matrices $R_l[n] = \x_l[n]\x_l\H[n]$ are Hermitian semi-definite, which means that we can get away with computing only one of their triagonals.
+\item The subarray covariance matrices $R_l[n] = \x_l[n]\x_l^H[n]$ are Hermitian semi-definite, which means that we can get away with computing only one of their triagonals.
 \item When summing $R_l[n]$ most of the multiplications need only be computed only once, and most sums can be updated iteratively.
 \end{itemize}
+%
 Second, each of the diagonals of $R_l[n]\ \forall\ l$ can be computed independently, which is good news for our GPU design.
 
-\subsection{MVDR beamforming in the frequency domain}
+\subsection{Frequency-domain MVDR}
 
@@
 The beamspace MVDR formulation very closely resembles that of the time domain MVDR. The main difference is the initial need to Fourier transform the data:
+%
 \begin{gather}
-\x_{\text{BS},l}[n] = \B\x_l[n], \qquad \B\in \mathbb{C}^{N_b,L}, \qquad \text{where} \qquad [\B]_{p,q} = \frac{1}{\sqrt{L}} e^{-j2\pi\,p\,q/L}.\label{beamspace}
+\x_{\text{BS},l}[n] = \mat{B}\x_l[n], \qquad \mat{B}\in \mathbb{C}^{N_b,L}, \qquad \text{where} \qquad [\mat{B}]_{p,q} = \frac{1}{\sqrt{L}} e^{-j2\pi\,p\,q/L}.\label{2_eq_beamspace}
 \end{gather}
-Here $\B$ is referred to as the Butler matrix, $l$ is the subarray index, and $N_b$ is the number of beams that should be computed.
-Note has once $\x_{\text{BS},l}[n]$ is computed, $\eR_\text{BS}$ can be found as in (\ref{spatialR}), $\w_\text{BS}$ as in (\ref{weights}), and $z$ as in (\ref{z}).
+%
+Here $\mat{B}$ is referred to as the Butler matrix, $l$ is the subarray index, and $N_b$ is the number of beams that should be computed.
+Note has once $\x_{\text{BS},l}[n]$ is computed, $\eR_\text{BS}$ can be found as in (\ref{2_eq_spatialR}), $\w_\text{BS}$ as in (\ref{2_eq_weights}), and $z$ as in (\ref{z}).
 
@@
 
-The Low Complexity Adaptive (LCA) beamformer is another take on the optimization problem in (\ref{mvdr_def}).
-Instead of calculating the weights like in (\ref{weights}), it simply computes the beamformer output for each window in a larger set of windows, and finally applies the one that best fulfils the given optimization criterion.
+The Low Complexity Adaptive (LCA) beamformer is another take on the optimization problem in (\ref{2_eq_mvdr_def}).
+Instead of calculating the weights like in (\ref{2_eq_weights}), it simply computes the beamformer output for each window in a larger set of windows, and finally applies the one that best fulfils the given optimization criterion.
 If we let $\w_p$ be the $p$'th of a total of $P$ windows, then this can be formulated as~\cite{Synnevag2008}:
 \begin{gather}
-\argmin{p}\ E\Big\{\big|\w_p\H\x[n]\big|^2\Big\} \qquad\text{subject to}\qquad \w_p\H\1 = 1. \label{lca_def}
+\argmin{p}\ E\Big\{\big|\w_p^H\x[n]\big|^2\Big\} \qquad\text{subject to}\qquad \w_p^H\1 = 1. \label{2_eq_lca_def}
 \end{gather}
-Note the similarity between (\ref{mvdr_def}) and (\ref{lca_def}).
+Note the similarity between (\ref{2_eq_mvdr_def}) and (\ref{2_eq_lca_def}).
 The LCA and MVDR uses essentially the same optimization criterion, but whereas MVDR searches a continuous solution space for the best set of weights to apply, the LCA beamformer only have $P$ solutions to select from.
@@
 \graphicsAI[drawing,width=\linewidth]{gfx/mvdr_implementation.svg}%
-\caption{MVDR implementation.}\label{mvdr_implementation}
+\caption{MVDR implementation.}\label{2_eq_mvdr_implementation}
 \end{figure}
 
-Our GPU implementation consists of three main kernels, one for each of the computing steps illustrated in Figure \ref{mvdr_implementation}.
+Our GPU implementation consists of three main kernels, one for each of the computing steps illustrated in Figure \ref{2_eq_mvdr_implementation}.
 In the first step we estimate $\eR$, in the second we solve the linear equation $\eRi\1$, and finally we perform some substitutions to obtain the beamformer output, $z$.
@@
 It is 68\,m long and 9\,m wide, and lies at a slanted seabed at 77\,m depth outside of Horten, Norway.
-The results are shown in Figure \ref{holmengraa}.
+The results are shown in Figure \ref{2_fig_holmengraa}.
 The MVDR images were processed with parameters $L$=16, $K$=3, and 1\% diagonal loading.
@@
 
-A few benchmarks of these methods are presented in Figure \ref{benchmark}.
+A few benchmarks of these methods are presented in Figure \ref{2_fig_benchmark}.
 On our test system with a quad-core Intel Xeon E5620, 64Gb of RAM and an nVidia Quadro 6000 card, the MVDR methods managed to create images at rates just over 1\,Mpx/s.
@@
 For beamspace MVDR we set $N_b=L/2$.
-LCA was set up with a window database consisting of 5 Kaiser windows with the window-parameter set in the range $\beta\in[0,10]$, and then each of these were steed in 5 distinct directions leaving a total of 25 windows for the LCA to select from..}\label{holmengraa}
+LCA was set up with a window database consisting of 5 Kaiser windows with the window-parameter set in the range $\beta\in[0,10]$, and then each of these were steed in 5 distinct directions leaving a total of 25 windows for the LCA to select from..}\label{2_fig_holmengraa}
 \end{figure}
@@
 \caption{Benchmarks for the our time-domain and frequency-domain MVDR method running on a GPU.
-The grey line marks the realtime requirement for the HUGIN AUV to perform full-coverage sectorscan images with a ping rate of 3 pings/s.}\label{benchmark}
+The grey line marks the realtime requirement for the HUGIN AUV to perform full-coverage sectorscan images with a ping rate of 3 pings/s.}\label{2_fig_benchmark}
 \end{figure}
@@
 The MVDR beamformer is a technique that is capable of improving image quality over conventional methods.
-This can be seen in Figure \ref{holmengraa}, where the MVDR produces images with cleaner shadows and slightly improved edge definitions.
+This can be seen in Figure \ref{2_fig_holmengraa}, where the MVDR produces images with cleaner shadows and slightly improved edge definitions.
 Contrary to common belief, it is also possible to create these images relatively fast.
@@
 It is based on the same optimisation criterion as the MVDR method, but avoids the estimation and inversion step altogether by limiting the weight's solution space to only a few predefined windows.
-As seen the Figure \ref{holmengraa}, it produces images with a quality comparable to the MVDR method, but computing the images takes only a fraction of the time.
+As seen the Figure \ref{2_fig_holmengraa}, it produces images with a quality comparable to the MVDR method, but computing the images takes only a fraction of the time.
 
-% Finally, throughout our work with porting the MVDR beamformer to a GPU, one challenge has been particularly difficult to solve: The GPU will only perform efficiently when each computing thread is very computationally intense.
-This should be apparent by studying Table \ref{tabbandwidth}.
-It compares theoretical peak values for the memory bandwidth and arithmetic throughput for an Nvidia Quadro 6000 GPU.
-Looking at global memory, we note that it is only able to move 1 float for every 30 floating point operation that the GPU is able to execute.
-Now, consider the simple equation we the simple equation $C=A+B$, where the ratio is 3 floats read/written for every 1 floating point operation, we realise that global memory should be avoided whenever possible.
-
-% 
-% $\underset{p}{\argmin}$
-% 
-% \begin{gather}
-% \vec w[n] = \frac{\Ri[n]\1}{\1\T\Ri[n]\1},\label{weights}
-% \end{gather}
-
-% \begin{figure}[!t]\centering
-% \includegraphics[width=\linewidth]{gfx/algorithm_structure.eps}
-% \caption{MVDR beamforming.
-First a spatial covariance matrix is estimated from the delayed data (\ref{spatialR}-\ref{finalR}), then the weights are computed (\ref{weights}) and finally applied to the delayed channel data (\ref{z}).}
-% \label{implementation}
-% \end{figure}
-% As summarized in Fig. \ref{mvdr_beamforming}, the MVDR method is applied to each pixel independently, by
-% \begin{enumerate}
-% \item computing the sample covariance matrix $\eR$ in (\ref{finalR}),
-% \item computing $\eRi\1$ in (\ref{weights}), and
-% \item computing the beamformer output $z$ in (\ref{finalZ}).
-% \end{enumerate}
-
-% 
-% \section{Results}
-% 
-% \begin{figure}[t]\centering
-% \graphicsAI[width=\linewidth]{gfx/implementation.svg}
-% \caption{MVDR beamforming. For one of the total number of pixels in range and azimuth, $N_y$ and $N_x$,\newline
-% 1. an $L\times{}L$ sample covariance matrix $\eR$ is computed, \newline
-% 2. the term $\eR^{-1}\1$ is found using a linear equation solver,\newline
-% 3. and the beamformer output is computed from $z$ from (\ref{finalZ}), where $\w$ is found by substituting $\eR^{-1}\1$ into (\ref{weights}). } \label{mvdr_beamforming}
-% \end{figure}
-
-% 
-% \ \\
-% Total: 192 words (200 max)
-% 
-% \ \\
-% Noen tanker jeg har selv naa:
-% \begin{itemize}
-% \item Det mangler noe om bruksomraader.
-% \item Maa passe paa aa ikke selge LCA saa bra at MVDR ser helt haapløs ut(?)
-% \item 200 ord er ingenting. :)
-% \end{itemize}
-% 
-% 
-% \ \\
-% Questions they wanted answered:\cite{Kailath1985}
-% \begin{itemize}
-% \item This session will feature work on adapting computationally demanding modeling and signal processing tasks to take advantage of various parallel architectures, such as cloud computing, computing clusters, multicore CPUs, multi-core graphic processing units (GPUs), DSP chips, and FPGAs.
-% \item If algorithm is slow - What technology will help me?
-% \item What skills are needed?
-% \item What tools are available and what's the cost?
-% \item What acceleration factor are we looking at?
-% \item How can I get started?
-% \item What we did. Which tools? Was it hard? Speedup?
-% \end{itemize}
-% \end{abstract}
-
-\bibliographystyle{jasanum}
-% \bibliography{../../Library/library}
-\bibliography{library}
-
-% \section{Introduction}
+}
 

### 2019-08-19 ThesisJo/3_publications.tex
2019-08-18

--- 2019-06-30 ThesisJo/2_publications.tex
@@
 This article evaluates time domain MVDR on GPUs for use in active sonar imaging.
-We refer to MVDR as fully adaptive, because it computes array weights analytically from spatial features of the data, yielding a continuous solution space.
+We refer to MVDR as fully adaptive because it computes array weights analytically from spatial features of the data, yielding a continuous solution space.
 Like any data-driven algorithm, it requires statistically distinguishable signal and noise.
@@
 This article evaluates frequency domain (or beamspace) MVDR on GPUs for use in cardiac ultrasound imaging.
-These systems typically have real-time requirements and more than 32 channels, so we apply a space reduction technique to reduce the computational effort.
+These systems typically have real-time requirements and exceed 32 channels, so we apply a space reduction technique to reduce computational effort.
 Since the transmitted beams are narrow, most acoustic energy falls within three to five center frequencies, and a good approximation follows from just computing those.
 
-We find that beamspace MVDR delivers real-time performance independent of channel-count and an image quality indistinguishable from time domain MVDR.
+We find that beamspace MVDR delivers real-time performance independent of channel-count and image quality indistinguishable from time domain MVDR.
 

### 2019-08-20 ThesisJo/2_background.tex
2019-08-19

@@
 
-Waves are made of endless tiny energy transactions, a dance of particles effecting each other in a search for equilibrium.
+Waves are made of endless tiny energy transactions, a dance of particles effecting each other in search of equilibrium.
 They are caused by disturbances in the forces of nature.
-The drop of a rock into flat water or rapid displacement of still air cause mechanical waves, while synchronized oscillations of electric and magnetic fields cause electromagnetic waves.
+The drop of a rock into calm water or rapid displacement of still air cause mechanical waves, while synchronized oscillations of electric and magnetic fields cause electromagnetic waves.
 Regardless of whether the waves are mechanical or electromagnetic, move perpendicular or parallel to the axis of propagation, what medium they propagate through or the means by which they do so, they are---perhaps surprisingly---all governed by a partial differential equation known as the \emph{wave equation}:
@@
 
-Suppose an array of sensors has a reference frame $\ubar{A} \triangleq (\udot{A},\uvec{A})$ attached to it, where $\udot{A}\in\mathbb{A}$ is its absolute position and $\uvec{A}\triangleq\big(\hat{x}_{A},\hat{y}_{A},\hat{z}_{A}\big)\in\mathbb{V}^3$ its absolute orientation (where $\hat{\cdot}$ denotes unit length).
-A signal source is located at a point $\udot{S}$.
+A few solutions to the wave equation is of particular interest, that of a plane and spherical wave
+%
+\begin{align}
+f_\textrm{plane}(\dvec{p}^A,t)  &= A e^{j(\omega t - \vec{k} \cdot \vec{p}_{SO})} \\
+f_\textrm{spheric}(\dvec{p}^A,t) &= \frac{A}{\lVert \dvec{p}_{SO}^S \rVert} e^{j(\omega t - \dvec{k}^T \lVert \dvec{p}_{SO}^A \rVert)}
+\end{align}
+%
+
+One is that of plane wave and the other a  A signal source with the position $\udot{S}$ .
 We can represent the field observed at $\udot{A}$ as $f(\dvec{p}^A,t)$, where position is represented by $\dvec{p}^A=[x_A,y_A,z_A]^T\in\mathbb{R}^3$ and time by $t\in\mathbb{R}$,
@@
 %
-\begin{align}
-f_\textrm{plane}(\dvec{p}^A,t)  &= A e^{j(\omega t - \dvec{k}^T \dvec{p}^A)} \\
-f_\textrm{spheric}(\dvec{p}^A,t) &= \frac{A}{\lVert \dvec{p}^A-\dvec{p}_{AS}^A \rVert} e^{j(\omega t - \dvec{k}^T \lVert \dvec{p}^A-\dvec{p}_{AS}^A \rVert)}
-\end{align}
+
 %
@@
 % \renewcommand{\thesection}{\thechapter.\arabic{section}}
+
+% Suppose an array of sensors has a reference frame $\ubar{A} \triangleq (\udot{A},\uvec{A})$ attached to it, where $\udot{A}\in\mathbb{A}$ is its absolute position and $\uvec{A}\triangleq\big(\hat{x}_{A},\hat{y}_{A},\hat{z}_{A}\big)\in\mathbb{V}^3$ its absolute orientation (where $\hat{\cdot}$ denotes unit length).
 
@@
 After stripping the carrier frequency by heterodyning and the signal signature by  matched filtering (Fig.
-\ref{I_beamforming}), an image can then be formed by coherently combining the receiver outputs to focus at one angle (or point) at the time.
+\ref{I_beamforming}), an image is formed by coherently combining the receiver outputs to focus at one angle (or point) at a time.
 The principle of adjusting the array's focus in range and direction is commonly referred to as beamforming, and involves assigning suitable delays and weights to the array's channels.
@@
 
-Suppose we record a wavefield with $M$ sensors arranged in a uniformly spaced linear array (ULA) and attach a reference frame $\ubar{A} \triangleq (\udot{A},\uvec{A})$ to it, where $\udot{A}\in\mathbb{A}$ is its absolute position and $\uvec{A}\triangleq\big(\hat{x}_{A},\hat{y}_{A},\hat{z}_{A}\big)\in\mathbb{V}^3$ its absolute orientation (where $\hat{\cdot}$ denotes unit length).
+Imagine we record a wavefield with $M$ sensors arranged in a uniformly spaced linear array (ULA) and attach a reference frame $\ubar{A} \triangleq (\udot{A},\uvec{A})$ to it, where $\udot{A}\in\mathbb{A}$ is its absolute position and $\uvec{A}\triangleq\big(\hat{x}_{A},\hat{y}_{A},\hat{z}_{A}\big)\in\mathbb{V}^3$ its absolute orientation (where $\hat{\cdot}$ denotes unit length).
 Assume that the array channels are digitally delayed to focus at a pixel with azimuth angle $\theta$ and range sample $n$, such that the delayed data from the $m$th channel can be expressed as $x_m[\theta,n-\Delta_m]\in\mathbb{C}$.
-We simply notation by making the dependence on $\theta$ implicit from now on.
+We simplify notation by making the dependence on $\theta$ implicit from now on.
 
@@
 With static weights this would be referred to as the conventional delay-and-sum (DAS) beamformer.
-A large variety of weighting functions exists here for trading lateral resolution for improved noise suppression (contrast), but one always ends up with a compromise between the two~\cite{Harris1978}.
+A large variety of weighting functions exist for trading lateral resolution for improved noise suppression (contrast), but a compromise between the two is unavoidable~\cite{Harris1978}.
 However, there exists a class of adaptive beamformers that target this limitation by adapting the weights to the impinging wavefield.
@@
 
-\subsubsection{Complexity and implementation}
-
-If (\ref{2_eq_spatialR}) were to be implemented directly, we note that the estimation step would have a complexity of O($N_K N_L L^2$).
-By comparison, performing a Gauss Jordan inversion is generally of O($L^3$), which implies that whenever $N_K\,N_L = N_K(M-L+1) \gg L$ the estimation step will dominate.
-This tends to be true in active systems because we often want to set $L<M/2$ and $N_K\ge>3$ for the MVDR to be robust enough to yield the expected results.
-
-A few properties let us improve these matters notably. First, we may infer from (\ref{2_eq_spatialR}) that:
-%
-\begin{itemize}
-\item The subarray covariance matrices $R_l[n] = \x_l[n]\x_l^H[n]$ are Hermitian semi-definite, which means that we can get away with computing only one of their triagonals.
-\item When summing $R_l[n]$ most of the multiplications need only be computed only once, and most sums can be updated iteratively.
-\end{itemize}
-%
-Second, each of the diagonals of $R_l[n]\ \forall\ l$ can be computed independently, which is good news for our GPU design.
-
 \subsection{Frequency-domain MVDR}
@@
 In its basic form its computational complexity only of O($M\,P$).
-
-\section{Implementation}\label{implementation}
-
-\begin{figure}[t]\centering%
-\graphicsAI[drawing,width=\linewidth]{gfx/mvdr_implementation.svg}%
-\caption{MVDR implementation.}\label{2_eq_mvdr_implementation}
-\end{figure}
-
-Our GPU implementation consists of three main kernels, one for each of the computing steps illustrated in Figure \ref{2_eq_mvdr_implementation}.
-In the first step we estimate $\eR$, in the second we solve the linear equation $\eRi\1$, and finally we perform some substitutions to obtain the beamformer output, $z$.
-All pixels are processed by one step before we move on to the next step.
-Our efforts here were mainly focused on estimating $\eR$, because a highly optimised batch linear equation solver is already available from Nvidia, and computing the beamformer output is a rather simple task.
-
-When computing $\eR$ we we assign $L$ threads per pixel to traverse the diagonals of $\eR$.
-On the top row a full computation is carried out, then that entire row is saved back to global memory following a collective access pattern that maximizes global memory bandwidth.
-For subsequent rows the threads move along the diagonals while performing iterative summations; the result from the previous element on the diagonal is updated by adding and removing the correlation coefficients that enters and exits the sum, respectively.
-To minimize memory consumption, we compute the coefficients again every time we need them.
-When a thread has finished up a diagonal, we have them wrap around to compute one of the diagonals in the lower triangular of $\eR$.
-Since $\eR$ is conjugate symmetric, the values in the leftmost column is obtained by a complex conjugate copy of the relevant value in the first row.
-This procedure balances the load evenly on all threads, is almost completely free of arithmetic redundancy, and consumes very little memory.
-
-The implementation of beamspace MVDR is very similar, but the data access pattern is different.
-In the time domain subsequent subarrays overlap on all but one element in memory, but the Fourier transformed data does not overlap at all.
-Such a seemingly innocent detail can make all the difference in a GPU design, because designing to make efficient use of memory is key to obtaining good performance.
 

### 2019-08-22 ThesisJo/2_background.tex
2019-08-21

@@
 \begin{align}
-\Delta f = 
-\frac{\delta^2 f}{\delta x^2} + \frac{\delta^2 f}{\delta y^2} + \frac{\delta^2 f}{\delta z^2} = \frac{1}{c}\frac{\delta^2 f}{\delta t^2},
+\nabla^2 f 
+ = \frac{1}{c}\frac{\delta^2 f}{\delta t^2},
 \end{align}
+%\frac{\delta^2 f}{\delta x^2} + \frac{\delta^2 f}{\delta y^2} + \frac{\delta^2 f}{\delta z^2}
 %
-where $\Delta$ is the Laplace operator, $f$ is field strength and $c\in\mathbb{R}^+$ is wave propagation speed.
+where $\nabla^2$ is the Laplace operator, $f$ is field strength and $c$ is wave propagation speed.
 If $f$ represents the displacement of a vibrating string, the equation states that the force pulling the string towards equilibrium (right side) is proportional to the string's curvature (left side).
@@
 
-A few solutions to the wave equation is of particular interest, that of a plane and spherical wave
+A few solutions to the wave equation is of interest, in particular that of a plane and spherical wave,
 %
 \begin{align}
-f_\textrm{plane}(\dvec{p}^A,t)  &= A e^{j(\omega t - \vec{k} \cdot \vec{p}_{SO})} \\
-f_\textrm{spheric}(\dvec{p}^A,t) &= \frac{A}{\lVert \dvec{p}_{SO}^S \rVert} e^{j(\omega t - \dvec{k}^T \lVert \dvec{p}_{SO}^A \rVert)}
+f_\textrm{plane}(\vec{p},t)  &= A e^{j(\omega t - \vec{k} \cdot \vec{p})} \\
+f_\textrm{spheric}(\vec{p},t) &= \frac{A}{\lVert \vec{p} \rVert} e^{j(\omega t - \lVert\vec{k}\rVert \lVert \vec{p} \rVert)},
 \end{align}
 % 
+where $A$ is amplitude, $\omega$ is temporal wave frequency, $t$ is time, $\lVert\vec{k}\rVert=\frac{\omega}{c}=\frac{2\pi}{\lambda}$ is spatial wave frequency (wavenumber) and $\vec{p}$ is position of the source relative to the observer.
 
-One is that of plane wave and the other a  A signal source with the position $\udot{S}$ .
+% One is that of plane wave and the other a  A signal source with the position $\udot{S}$ .
 We can represent the field observed at $\udot{A}$ as $f(\dvec{p}^A,t)$, where position is represented by $\dvec{p}^A=[x_A,y_A,z_A]^T\in\mathbb{R}^3$ and time by $t\in\mathbb{R}$,
-
-An introduction to such notation can be found in \chapref{IV_methods}{IV}.
-
-reference frame $\ubar{A} \triangleq (\udot{A},\uvec{A})$ to it, where $\udot{A}\in\mathbb{A}$ is its absolute position and $\uvec{A}\triangleq\big(\hat{x}_{A},\hat{y}_{A},\hat{z}_{A}\big)\in\mathbb{V}^3$ its absolute orientation (where $\hat{\cdot}$ denotes unit length)
+% 
+% An introduction to such notation can be found in \chapref{IV_methods}{IV}.
+% 
+% 
+% reference frame $\ubar{A} \triangleq (\udot{A},\uvec{A})$ to it, where $\udot{A}\in\mathbb{A}$ is its absolute position and $\uvec{A}\triangleq\big(\hat{x}_{A},\hat{y}_{A},\hat{z}_{A}\big)\in\mathbb{V}^3$ its absolute orientation (where $\hat{\cdot}$ denotes unit length)
 
@@
 Assume that the array channels are digitally delayed to focus at a pixel with azimuth angle $\theta$ and range sample $n$, such that the delayed data from the $m$th channel can be expressed as $x_m[\theta,n-\Delta_m]\in\mathbb{C}$.
-We simplify notation by making the dependence on $\theta$ implicit from now on.
+By making the dependence on $\theta$ implicit and assuming the data to be pre-delayed to the pixel of interest, we can simply the notation to $x_m[\theta,n-\Delta_m] = x_m[n]$.
 

### 2019-08-22 ThesisJo/4_discussion.tex
2019-08-18
2019-08-21

--- 2018-10-11 ThesisJo/3_discussion.tex
@@
 
-Constrain attention to beamformers and simulators (not ATR?).
+The contributions of this thesis are:
+
+Paper I - Time-domain MVDR beamforming on GPUs in active sonar imaging:
+\begin{enumerate}
+\item Accelerates the MVDR beamformer through algorithmic reductions and implementation on GPU.
+\item Demonstrates speeds exceeding 1\,Mpx/s on systems with no more than 32 channels, two orders of magnitude faster than a semi-optimized C-implementation on a CPU.
+\item Offers uncommon insights into hardware limitations, design implications and computation efficiency.
+\end{enumerate}
+
+Paper II - Frequency-domain MVDR beamforming on GPUs in clinical ultrasound imaging:
+\begin{enumerate}
+\item Alters the implementation in Paper I to operate in a frequency domain approximated by its three most significant components.
+\item Demonstrates speeds exceeding 1\,Mpx/s on systems with 64 and 96 channels.
+\item Yields similar image quality to time-domain MVDR.
+\end{enumerate}
+
+Paper III - LCA beamforming compared to MVDR in active sonar imaging:
+\begin{enumerate}
+\item Contrasts LCA, that selects the best performing weights from a predefined set, to MVDR, that computes the weights from an estimated covariance matrix formed from the input data.
+\item 
+\item Yields similar image quality to MVDR.
+\end{enumerate}
+
+Paper IV - Real-time sonar image synthesis:
+\begin{enumerate}
+\item Contrasts LCA, that selects the best performing weights from a predefined set, to MVDR, that computes the weights from an estimated covariance matrix formed from the input data.
+\item Extends a navigational notation system with topological ideas from robotics and homogeneous coordinates, promoting geometrical clarity and validity while aiding the designer in focusing on higher level concepts, such as manipulating rigid objects.
+\item Yields similar image quality to MVDR.
+\end{enumerate}
 
@@
 
+\section{Onlook}
+
+Machine cognition:
+\begin{itemize}
+\item Ethical dilemmas. Decision and actions have consequences. Who's legally responsible for a machine's doing? Itself?
+\item Safety. Machine flexibility allow more possible outcomes, and a wider range of security issues to consider.
+\end{itemize}
+
+R dim > 10000
+- Inverting such matrices better to do with gradient decent
+
+Deep learning. Simulator used to trained deep neural networks. Lots of data needed.
+
+\section{}
+
 \end{document}

### 2019-09-02 ThesisJo/2_background.tex
2019-09-01

@@
 
-Waves are made of endless tiny energy transactions, a dance of particles effecting each other in search of equilibrium.
-They are caused by disturbances in the forces of nature.
-The drop of a rock into calm water or rapid displacement of still air cause mechanical waves, while synchronized oscillations of electric and magnetic fields cause electromagnetic waves.
-Regardless of whether the waves are mechanical or electromagnetic, move perpendicular or parallel to the axis of propagation, what medium they propagate through or the means by which they do so, they are---perhaps surprisingly---all governed by a partial differential equation known as the \emph{wave equation}:
+Any wave, regardless of its nature, are governed by a partial differential equation known as the \emph{wave equation}:
 %

### 2019-09-02 ThesisJo/4_discussion.tex
2019-09-01

--- 2018-03-05 ThesisJo/3_discussion.tex
@@
 
-\begin{itemize}
-\item Improving imaging systems by making them smarter.
-\item Big data - more processing power.
-\item Processing speed
-\begin{itemize}
-	\item Choose the best tool (algorithm) for the job
-	\item More complex not always better, but slower
-	\item Extracting the essential information is key 
-	\item Exploit hardware resources available
-\end{itemize}
-\item Faster IS smarter
- \begin{itemize}
- 	\item Short feedback loops allow real-time adaptivity
- 	\item More data contains more information
- 	\item Faster learning
- \end{itemize}
-\item Smarter IS better
- \begin{itemize}
-	\item Better image quality
-	\item Improved autonomy
-\end{itemize}
-\end{itemize}
+The main contributions of this thesis can be summarized as follows:
 
-\begin{itemize}
-\item Improving imaging systems by making them smarter.
-\item Big data - more processing power.
-\item IV - Simulator
-\begin{itemize}
-	\item Incredibly fast template matching possible.
-	\item Possibly a powerful tool to pass prior knowledge to deep learning
-	\item Extracting the essential information is key 
-	\item Exploit hardware resources available
-\end{itemize}
-\item Faster IS smarter
- \begin{itemize}
- 	\item Short feedback loops allow real-time adaptivity
- 	\item More data contains more information
- 	\item Faster learning
- \end{itemize}
-\item Smarter IS better
- \begin{itemize}
-	\item Better image quality
-	\item Improved autonomy
-\end{itemize}
-\end{itemize}
+\textbf{Paper I} - \textit{Time-domain MVDR beamforming on GPUs in active sonar imaging}:
+\begin{enumerate}
+\item Accelerates the MVDR beamformer through algorithmic reductions and fine-tuning to a GPU, yielding a real-time solution.
+\item Demonstrates speeds exceeding 1\,Mpixel/s on systems with no more than 32 channels, two orders of magnitude faster than a semi-optimized C-implementation on a CPU.
+\item Offers rarely documented insights into hardware limitations, design implications and computational efficiency.
+\end{enumerate}
+
+Time-domain MVDR is the reference method for creating sonar images using the minimum variance criterion.
+It computes array weights analytically, yielding optimal weights in the minimum variance sense.
+Initially, this may seem ideal, but its use is cumbersome: parameters for temporal averaging, spatial averaging and regularization must often be fine-tuned ad-hoc to obtain improved image quality.
+GPU acceleration mitigates the high computational complexity resulting from estimating and inverting a spatial covariance matrix, yielding real-time performance.
+
+\textbf{Paper II} - \textit{Frequency-domain MVDR beamforming on GPUs in clinical ultrasound imaging}:
+\begin{enumerate}
+\item Alters the implementation in Paper I to operate in beamspace (frequency domain), reduced to its three center beams (the most significant ones).
+\item Demonstrates speeds exceeding 1\,Mpixel/s on systems with 64 and 96 channels.
+\item Yields similar image quality to time-domain MVDR, in accordance with former studies in active sonar.
+\end{enumerate}
+
+\textbf{Paper III} - \textit{LCA beamforming versus MVDR in active sonar imaging}:
+\begin{enumerate}
+\item Shows that LCA beamformer performs similarly to MVDR with only 6 weight sets to choose from: The rectangular and Kaiser function with $\beta=6$, steered to $\phi\in[0,\pm\frac{1}{2}\delta{\phi_{-3dB}}]$, where $\delta{\phi_{-3dB}}$ is the $-3$\,dB bandwidth of the amplitude response to the weight set.
+%Contrasts the LCA and MVDR beamformer, two different ways to solve the minimum variance criterion.
+The former selects the best performing weights from a predefined set, the latter computes them from a sample covariance matrix estimated from the input data.
+\item Finds no further improvement above 15 suitable weight sets.
+\item Provides insights into typical amplitude responses of MVDR constrained for active sonar imaging and pinpoints their resemblance to the chosen LCA weight sets. 
+\item Suggests swapping the Kaiser function trigonometric one parametrized by $\alpha$, yielding similar image quality while allowing an analytic solution for $\alpha$ to be found.
+This reduces the algorithm to simply iterating through steering angles.
+\end{enumerate}
+
+\textbf{Paper IV} - \textit{Real-time sonar image synthesis on GPUs for automatic target recognition}:
+\begin{enumerate}
+\item Implements a SAS simulator on GPU that allows AUVs to compare its images on-the-fly to vast numbers of objects and configurations, yielding superior classification results and real-time responses.
+\item Extends a navigational notation system with topological ideas from robotics and homogeneous coordinates, promoting geometrical clarity and validity while aiding the designer to focus on higher level concepts, such as manipulating rigid objects.
+\item Demonstrates superior classification accuracy on experimental data of a few thousand objects from Jesus Bay, Norway, at false positive rates (FRP) below 20, similar %, and tied with it above. 
+\end{enumerate}
+
+% We grant them perception by performing automatic target recognition (ATR) with an
+% acoustic simulator for synthetic aperture sonar (SAS). With it, an AUV can compare
+% its images on-the-fly to vast numbers of objects and configurations, yielding superior
+% classification results and real-time responses. This is validated on experimental data of
+% a few thousand objects from Jesus Bay, Norway, recorded by the Kongsberg Maritime
+% HISAS1030 sonar. Our simulator consistently outperformed a conventional method using
+% predefined images at false positive rates (FRP) below 20 %, and tied with it above.
+
+In clinical ultrasound imaging
+
+\section{Onlook}
+
+Machine intelligence requires high fidelity sensory data, sophisticated adaptive signal processing and massive computing power.
+If history is any indication, the future promises more of everything in these categories.
+
+The future holds a promise of more of everything: more sensors (data), more signal processing and more processing power.
+
+\emph{Big data} and \emph{Internet of things} (IoT) and \emph{big data} will continue to empower us with ever more information.
+
+\emph{Machine learning} has made leaps with the combination of deep neural networks and big data.
+
+\subsection{Sensors}
+
+\emph{Phased arrays}
+
+\section{Signal processing}
+
+\emph{Software defined sonar}
+
+\subsection{Hardware}
+
+\emph{System on Chips} (Soc) are becoming increasingly popular as a heterogeneous computing platform.
+One of these chips, the Ultrascale+ Radio Frequency SoC (RFSoC) from Xilinx, contains a quad-core CPU, a dual-core real-time processor, a GPU, Field Programmable Gate Array (FPGA) resources and up to 16 Analog-to-Digital and Digital-to-Analog Converters (ADCs and DACs)---all in one CMOS chip.
+It targets the need for multiple-input-multiple-output (MIMO) can easily replace most analog front-end electronics in a sonar or ultrasound system with a software-defined solution, yielding better flexibility, reduced design complexity, improved bandwidth and lowered latency, all in a tiny physical footprint.
+Although the GPU
+
+\emph{Phased arrays} samples and records wavefields for digital processing. It is a prerequisite for many of the image reconstruction algorithms
+
+\subsection{Software}
+
+% % 
+% % \item System on Chips with integrated ADCs/DACs. Big power in small package.
+% % \item Compact, flexible, direct conversion
+% % \item Software defined signal generation, filtering, synchronization, rate-conversion and mixing.
+% % \item Multipurpose, hardware reuse, easier implementation of advanced signal processing, faster development and lower price.
+% % \item 
+% % \item Compact multi-band Wifi
+% % \item MIMO (in new wifi). Breaks array SNR boost. Not viable for long range.
+% % \item Phones: Wifi 2.4-2.5G, 5G (massive MIMO, heavy SDR, 600M-6G), LTE (4G, Long Term Evolution 800M), GSM (2G, Global System for Mobile communications 900M,1.8-1.9M), UMTS (3G, Universal Mobile Telecommunications System 1.9-2.1M), GPS (Global Positioning System 1.5G), Bluetooth (2.4G), NFC (Near-Field Communications 13.56M), FM (100M).
+Multi-radio in a single SDR package!
+% 
+% 
+% 
+% 
+% 
+% 
+% 
+% 
+% 
+% 
+% The potential for intelligence improves with the amount of high quality data, the sophistication of the signal processing algorithms and the hardware processing that glues it all together. 
+% 
+% The acquisition of ever more data 
+% Instruments acquire ever 
+% 
+% The intelligence of autonomous instruments require boils down to a tight and fast integration of data collection, processing relies on high quality data, processing and to three key ingredients: high quality sensory data, fast data processing and adaptive signal processing.
+% 
+% More high quality data
+% The processing power to digest it fast
+% Adaptive 
+% 
+% \section{Trends}
+% 
+% The future is faster, smarter.
+% 
+% \subsection{Hardware}
+% 
+% \begin{itemize}
+% \item Ever increasing processing speed.
+% \begin{itemize}
+% 	\item Choose the best tool (algorithm) for the job
+% 	\item More complex not always better, but slower
+% 	\item Extracting the essential information is key 
+% 	\item Exploit hardware resources available
+%    \item Intel entering the discrete graphics stage
+%    \end{itemize}
+% \item Software defined transceiver/receivers
+% \begin{itemize}
+% \item System on Chips with integrated ADCs/DACs. Big power in small package.
+% \item Compact, flexible, direct conversion
+% \item Software defined signal generation, filtering, synchronization, rate-conversion and mixing.
+% \item Multipurpose, hardware reuse, easier implementation of advanced signal processing, faster development and lower price.
+% \item 
+% \item Compact multi-band Wifi
+% \item MIMO (in new wifi). Breaks array SNR boost. Not viable for long range.
+% \item Phones: Wifi 2.4-2.5G, 5G (massive MIMO, heavy SDR, 600M-6G), LTE (4G, Long Term Evolution 800M), GSM (2G, Global System for Mobile communications 900M,1.8-1.9M), UMTS (3G, Universal Mobile Telecommunications System 1.9-2.1M), GPS (Global Positioning System 1.5G), Bluetooth (2.4G), NFC (Near-Field Communications 13.56M), FM (100M).
+Multi-radio in a single SDR package!
+% \end{itemize}
+% \item Short feedback loops allow real-time adaptivity
+% \end{itemize}
+% 
+% 
+% \subsection{Algorithms}
+% 
+% \begin{itemize}
+% \item Intelligence - Ability to process data fast, and retain key information
+% \item Adaptive methods ever more attractive.
+%    \begin{itemize}
+%    \item Processing more data to increase significance of extracted information
+%    \item Reducing reaction time: Act when the data is fresh/valid.
+%    \end{itemize}
+% \item Better image quality
+% \item Improved autonomy
+% \end{itemize}
+% 
+% 
+% \subsection{Connectivity}
+% 
+% \begin{itemize}
+% \item Internet of Things
+% \item Big data
+% \item Intelligent mesh of nodes
+% \end{itemize}
+% 
+% \section{Risks \& Challenges}
+% 
+% 
+% 
+% \begin{itemize}
+% \item Improving imaging systems by making them smarter.
+% \item Big data - more processing power.
+% \item Faster IS smarter
+%  \begin{itemize}
+%  	\item Short feedback loops allow real-time adaptivity
+%  	\item More data contains more information
+%  	\item Faster learning
+%  \end{itemize}
+% \end{itemize}
+% 
+% 
+% 
+% %Discussion and future work
+% %   Summary and discussion
+% %      Simulator:
+% %      - The way to train and constrain deep learning algorithms?
+% %      MVDR:
+% %      - Analytical optimization - full adaptivity.
+% %      - Heavy on computations
+% %      LCA:
+% %      - Optimization by trial and error - discrete solution space.
+% %      - Works almost as well as MVDR in most cases
+% %      - Fast!
+% %      - Well suited for hardware
+% %      
+% %      Lessions from deep learning:
+% %      - 
+% %      
+% %   Future work
+% 
+% 
+% \section{MVDR}
+% 
+% Recent developments:
+% - A low complexity minimum variance beamformer for ultrasound imaging using dominant mode rejection~\cite{Jiang2016}
+% - Google deep beamforming
+% 
+% \begin{itemize}
+% \item SDR - Entirely software defined radio, radar and sonar, bigger potential for adaptive methods
+% \item Cognitive radar/sonar.
+1.
+Adaptive signal processing, learn from data.
+2.
+Feedback from receiver to transmitter.
+3.
+Preservation of the information in received data.
+All of this is present in bats.
+% \end{itemize}
+% 
+% 
+% \section{Onlook}
+% 
+% Machine cognition:
+% \begin{itemize}
+% \item Ethical dilemmas. Decision and actions have consequences. Who's legally responsible for a machine's doing? Itself?
+% \item Safety. Machine flexibility allow more possible outcomes, and a wider range of security issues to consider.
+% \end{itemize}
+% 
+% R dim > 10000
+% - Inverting such matrices better to do with gradient decent
+% 
+% 
+% Deep learning. Simulator used to trained deep neural networks. Lots of data needed.
+% 
+% 
+% 
+% 
+% \section{}
 

### 2019-09-09 ThesisJo/3_publications.tex
2019-09-08

@@
 
-We present four journal published papers concerned with \emph{adaptive} and \emph{hardware accelerated} formation of acoustic images: the first three reconstruct from echoes and the last synthesizes through simulation.
+We present four journal published papers concerned with \emph{adaptive} and \emph{hardware accelerated} formation of acoustic images: the first three through \emph{reconstruct} from echoes and the last \emph{synthesizes} through simulation.
 
@@
 %~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
-\textbf{An Optimized GPU implementation of the\\ MVDR Beamformer for Active Sonar Imaging}\\
+\textbf{An Optimized GPU implementation of the\\
+MVDR Beamformer for Active Sonar Imaging}\\
 J.I. Buskenes, J.P. Åsen, C.-I.C. Nilsen and A. Austeng\\
@@
 
-This article evaluates time domain MVDR on GPUs for use in active sonar imaging.
-We refer to MVDR as fully adaptive because it computes array weights analytically from spatial features of the data, yielding a continuous solution space.
+This article accelerates time-domain MVDR with GPUs for use in active sonar imaging.
+Time-domain MVDR is the reference method for creating sonar images using the minimum variance optimization, aimed at reducing noise and interference while preserving signal integrity while.
+We refer to it as fully adaptive because it computes array weights analytically from spatial features of the data, yielding array weights from a continuous solution space that are optimal in the minimum variance sense.
 Like any data-driven algorithm, it requires statistically distinguishable signal and noise.
@@
 The design is limited to 32 channels due to data dependencies that are complicated and not inherently parallel, combined with the relatively limited bandwidth of this hardware architecture.
+
+%Time-domain MVDR is the reference method for creating sonar images using the minimum variance criterion.
+It computes array weights analytically, yielding optimal weights in the minimum variance sense.
+Initially, this may seem ideal, but its use is cumbersome: parameters for temporal averaging, spatial averaging and regularization must often be fine-tuned ad-hoc to obtain improved image quality.
+GPU acceleration mitigates the high computational complexity resulting from estimating and inverting a spatial covariance matrix, yielding real-time performance.
 

### 2019-09-26 ThesisJo/2_background.tex
2019-09-17

@@
 
-Consider a sonar imaging scenario where an encoded signal is transmitted to highlight a surface of interest, and assume that the backscattered wavefield is properly sampled using a uniform linear array of receivers.
+Consider an acoustic signal transmitted to highlight a surface of interest.
+Assume the backscattered wavefield is properly sampled using a uniform linear array (ULA) of receivers.
 After stripping the carrier frequency by heterodyning and the signal signature by matched filtering (Fig.
@@
 
-Imagine we record a wavefield with $M$ sensors arranged in a uniformly spaced linear array (ULA) and attach a reference frame $\ubar{A} \triangleq (\udot{A},\uvec{A})$ to it, where $\udot{A}\in\mathbb{A}$ is its absolute position and $\uvec{A}\triangleq\big(\hat{x}_{A},\hat{y}_{A},\hat{z}_{A}\big)\in\mathbb{V}^3$ its absolute orientation (where $\hat{\cdot}$ denotes unit length).
+Imagine an array with $M$ sensors with a reference frame $\ubar{A} \triangleq (\udot{A},\uvec{A})$ attached to it, where $\udot{A}\in\mathbb{A}$ is its absolute position and $\uvec{A}\triangleq\big(\hat{x}_{A},\hat{y}_{A},\hat{z}_{A}\big)\in\mathbb{V}^3$ its absolute orientation (where $\hat{\cdot}$ denotes unit length).
 Assume that the array channels are digitally delayed to focus at a pixel with azimuth angle $\theta$ and range sample $n$, such that the delayed data from the $m$th channel can be expressed as $x_m[\theta,n-\Delta_m]\in\mathbb{C}$.

### 2019-09-26 ThesisJo/3_publications.tex
2019-09-17

@@
 
+\begin{figure}[tp]
+\makebox[\linewidth][c]{%
+\graphicsAI[drawing,width=1.3\linewidth]{gfx/article_relation.svg}}
+\caption{Article relations.
+The first three articles reconstructs images from acoustic echoes.
+The last article synthesizes image templates from 3D models and object parameters estimated from the reconstructed image.
+The two key traits are: accelerated adaptivity.
+}
+\end{figure}
+
 %
@@
 %- GPU
+% MVDR: Needs acceleration
+% LCA: No particular need of acceleration
+% Sim: Needs acceleration
 

### 2019-09-30 ThesisJo/2_background.tex
2019-09-30

@@
 
-A few solutions to the wave equation is of interest, in particular that of a plane and spherical wave,
+A few solutions to the wave equation are of interest, in particular that of a plane and spherical wave:
 %
@@
 % 
-where $A$ is amplitude, $\omega$ is temporal wave frequency, $t$ is time, $\lVert\vec{k}\rVert=\frac{\omega}{c}=\frac{2\pi}{\lambda}$ is spatial wave frequency (wavenumber) and $\vec{p}$ is position of the source relative to the observer.
+Here $A$ is amplitude, $\omega$ is temporal wave frequency, $t$ is time, $\lVert\vec{k}\rVert=\frac{\omega}{c}=\frac{2\pi}{\lambda}$ is spatial wave frequency (wavenumber) and $\vec{p}$ is position of the source relative to the observer.
 
@@
 \begin{align}
-z[n] = \dvec{w}^H[n]\dvec{x}[n] = \bmat{w_0[n]\\w_1[n]\\\vdots\\w_{M-1}[n]}^H \bmat{x_0[n]\\x_1[n]\\\vdots\\x_{M-1}[n]},\label{2_eq_z}
+z[n] = \dvec{w}^H[n]\dvec{x}[n] = \bmat{w_0[n]\\w_1[n]\\\vdots\\w_{M-1}[n]}^H \bmat{x_0[n]\\x_1[n]\\\vdots\\x_{M-1}[n]} \in\mathbb{C},\label{2_eq_z}
 \end{align}
@@
 where $w_m\in\mathbb{C}$ is the weight factor assigned to channel $m$.
-With static weights this would be referred to as the conventional delay-and-sum (DAS) beamformer.
-A large variety of weighting functions exist for trading lateral resolution for improved noise suppression (contrast), but a compromise between the two is unavoidable~\cite{Harris1978}.
-However, there exists a class of adaptive beamformers that target this limitation by adapting the weights to the impinging wavefield.
-The MVDR method is one such technique.
+With static weights this is referred to as the conventional delay-and-sum (DAS) beamformer.
+A variety of weighting functions exist for trading lateral resolution for improved noise suppression (contrast), but a compromise between the two is unavoidable~\cite{Harris1978}.
+A class of adaptive beamformers target this limitation by adapting the weights to the impinging wavefield.
+One such technique is the minimum variace distortionless response (MVDR) method.
+
+\begin{figure}[tp]\label{II_fig_article_relation}
+\makebox[\linewidth][c]{%
+\graphicsAI[drawing,width=1.3\linewidth]{gfx/article_relation.svg}}
+\caption{Method relations.
+This thesis explores three beamformers reconstructing images from acoustic echoes: time-domain MVDR (reference), frequency-domain MVDR (close approximation) and time-domain LCA (educated trial-and-error).
+Background on synthesis of image templates from the reconstructed images will be presented in article IV.
+}
+\end{figure}
+
+% Accelerated adaptivity is key: it enables rapid dynamic responses to mission events. }
 
@@
 \begin{figure}[t]\centering%
-\graphicsAI[drawing,width=0.7\linewidth]{gfx/implementation.svg}%
+\graphicsAI[drawing,width=1\linewidth]{gfx/implementation.svg}%
 \caption{MVDR beamforming. For one of the total number of pixels in range and azimuth, $N_y$ and $N_x$,\newline
@@
 %
-The basic idea behind the MVDR method is to compute the set of weights that minimizes noise and signal power accumulated by the array, while ensuring unity gain for signals coming from the pixel of interest.
-When the data is delayed this can be formulated as follows:
+The basic idea behind MVDR is to compute the set of weights that minimizes noise and signal power accumulated by the array, while seeking unit gain for signals from the pixel of interest.
+With pre-delayed data this can be formulated as follows:
 %
@@
 %
-This is a convex optimization problem with the solution:
+This optimization problem has the solution:
 %
@@
 %
-where $\1$ is a row vector of ones that represents broadside steering, and $\R=E\{\x[n]\x^H[n]\}$ is the spatial covariance matrix for the full array.
-Since $\R$ is unknown, we estimate it by by computing a sample covariance matrix $\eR$.
-In this computation we perform some degree of \emph{spatial averaging} to avoid signal cancellation, \emph{temporal averaging} to maintain true speckle statistics, and \emph{diagonal loading} to improve robustness to parameter errors ~\cite{Synnevag2009a}.
+where $\dvec1$ is a row vector of ones representing broadside steering and $\R=E\{\x[n]\x^H[n]\}$ is the spatial covariance matrix for the full array.
+We estimate $\R$ by computing a sample covariance matrix $\eR$.
+A good estimate requires some degree of \emph{spatial averaging} to avoid signal cancellation, \emph{temporal averaging} to maintain true speckle statistics, and \emph{diagonal loading} to improve robustness to parameter errors ~\cite{Synnevag2009a}.
 Combined, these techniques will also ensure a numerically well conditioned $\eR$.

### 2019-09-30 ThesisJo/abstract.tex
2019-09-30

@@
-\input{variables}\ifMonolithic\else\input{header}\input{subheader}\fi
+% \input{variables}\ifMonolithic\else\input{header}\input{subheader}\fi
 %~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
 
-\chapter{Abstract}
+The world is highly dynamic, constantly changing from moment to moment.
+An autonomous device aspiring to navigate it must be situationally perceptive and reactive.
+This requires numerous sensors gathering high quality data at high throughput, adaptive signal processing algorithms and powerful parallel hardware capable of near-realtime responses.
+
+This thesis explores ways to accelerate adaptive acoustic image reconstruction and synthesis.
+The reconstruction process relies heavily on beamforming: the art of delaying and weighting sensor arrays to achieve desired spatial responses.
+A beamformer can be viewed as a spatial filter or model, and works best---as any model---when it is neither too simple nor too complex.
+
+All models have a sweet spot in complexity, a balance point where its estimates offer an ideal compromise between bias and variance.
+
+Such a spatial filter, or model, . model complexity should be just right. not too simple, not too complex. bias vs variance.
+
+sensory faculty
+
+as adaptive as possible with the least possible computational load.
+
+gather information
+dynamically analyze it
+and react to it
+Får
+
+adapt
+persist
+prevail
+
+Adaptive algorithm and the computation power to drive it.
+
+Necessitates an algorithm that adapts to changing data.
+
+Cognition is the mental action or process of acquiring knowledge and understanding through thought, experience and the senses.
+
+The intelligence and autonomy of acoustical imaging devices require more sensory data, algorithmic adaptivity and processing power. All are vital and equally important.
+
+improves with more high quality sensory data, sophisticated adaptive signal processing and the hardware resources to deal with it all.
+
+more sensory data, algorithmic adaptivity and processing power.
+
+adaptive processing and processing power.
+
+benefits from more sensors, sensory data,   detectors rely on
+
+autonomy improves with data amount, adaptive processing, processing power.
+
+fast dynamic information gathering and processing.
+
+sensory input, sophisticated information and the processing power to drive it all.
+
+rely on fast adaptive data processing. have an intrinsic need for adaptive data processing and hardware acceleration.
+
+Accelerated adaptive image reconstruction and synthesis.
+
+information gathering and processing.
+
+One use case autonomy
+
+Key challenges that are addressed include active sonar and
+
+- Active sonar and clinical ultrasound
+- Different ways to achieve adaptivity. Analytical assessment of the data, computing weights. Approximating it. Trial-and-error. Different means to the same adaptivity
+- Computational complexity
+- Hardware acceleration
+-
+
+Spatial sensitivity is like a balloon; squeezing it on one side inflates it on the other.
+
+World is highly dynamic, constantly changing from moment to moment. Necessitates an algorithm that adapts to changing data.
 
 %~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
-\ifMonolithic\else\input{subfooter}\fi
+% \ifMonolithic\else\input{subfooter}\fi

### 2019-10-01 ThesisJo/2_background.tex
2019-09-31

@@
 
-Any wave, regardless of its nature, are governed by a partial differential equation known as the \emph{wave equation}:
+Any wave, regardless of its nature, is governed by a partial differential equation known as the \emph{wave equation}:
 %
@@
 If $f$ represents the displacement of a vibrating string, the equation states that the force pulling the string towards equilibrium (right side) is proportional to the string's curvature (left side).
-In acoustics $f$ represents sound pressure, but since the mathematics are all the same we will simply think of $f$ as a signal: a unit-less scalar function with wave characteristics.
+In acoustics $f$ represents sound pressure, but since the mathematics are independent of the physics involved we will simply think of $f$ as a signal: a unit-less scalar function with wave characteristics.
 
-A few solutions to the wave equation are of interest, in particular that of a plane and spherical wave:
+The wave equation has a few solutions of interest, in particular that of a plane and spherical wave:
 %
@@
 
+\begin{figure}[tp]\label{II_fig_article_relation}
+\makebox[\linewidth][c]{%
+\graphicsAI[drawing,width=1.3\linewidth]{gfx/article_relation.svg}}
+\caption{Method relations.
+This thesis explores three beamformers reconstructing images from acoustic echoes: time-domain MVDR (reference), frequency-domain MVDR (close approximation) and time-domain LCA (educated trial-and-error).
+Background on synthesis of image templates from the reconstructed images will be presented in article IV.
+}
+\end{figure}
+
 \subsection{Beamforming}
@@
 
-\begin{figure}[tp]\label{II_fig_article_relation}
-\makebox[\linewidth][c]{%
-\graphicsAI[drawing,width=1.3\linewidth]{gfx/article_relation.svg}}
-\caption{Method relations.
-This thesis explores three beamformers reconstructing images from acoustic echoes: time-domain MVDR (reference), frequency-domain MVDR (close approximation) and time-domain LCA (educated trial-and-error).
-Background on synthesis of image templates from the reconstructed images will be presented in article IV.
-}
-\end{figure}
-
 % Accelerated adaptivity is key: it enables rapid dynamic responses to mission events. }
@@
 %
-The basic idea behind MVDR is to compute the set of weights that minimizes noise and signal power accumulated by the array, while seeking unit gain for signals from the pixel of interest.
+The basic idea behind MVDR is to compute the set of weights that minimizes noise and signal power accumulated by the array, while ensuring unit gain for signals from the pixel of interest.
 With pre-delayed data this can be formulated as follows:
@@
 
-As we increase system size the inversion step will take over as the main bottleneck.
-We can, however, reduce the dimensionality further than subarray averaging already has by performing MVDR beamforming in the frequency domain.
-The underlying concept here that for narrowband data there will be a direct relation between the Fourier transform across the array and angles of arrival in the spatial domain; the zero-frequency will correspond to broadside, and increasing frequencies will map to monotonously increasing opening angles.
-Since the combined transmit and receive response is highly directive, this method allows us to only process the angles where our signals are coming from~\cite{Nilsen2009a,VanTrees2002}.
+Above 32 channels the inversion step becomes a greater computational bottleneck than covariance estimatation.
+We mitigate this by applying space reduction in the spatial frequency-domain, or beamspace.
+If the data is narrow band, narrow beam, or both, the backscatter energy resides primarily in the center frequencies.
+The zero-frequency correspond to broadside, higher frequencies to larger opening angles.
+Since the combined transmit and receive response is highly directive, it is sufficient to process only three or five center angles and ignore the rest~\cite{Nilsen2009a,VanTrees2002}.
 The method is called beamspace MVDR.
 
-The beamspace MVDR formulation very closely resembles that of the time domain MVDR. The main difference is the initial need to Fourier transform the data:
+Beamspace MVDR formulation closely resembles that of time domain MVDR, the main difference being an initial need to Fourier transform the data:
 %
@@
 %
-Here $\mat{B}$ is referred to as the Butler matrix, $l$ is the subarray index, and $N_b$ is the number of beams that should be computed.
+Here $\mat{B}$ refers to the Butler matrix, $l$ is subarray index, and $N_b$ is number of beams that should be computed.
 Note has once $\x_{\text{BS},l}[n]$ is computed, $\eR_\text{BS}$ can be found as in (\ref{2_eq_spatialR}), $\w_\text{BS}$ as in (\ref{2_eq_weights}), and $z$ as in (\ref{z}).
 
-When we reduce the covariance size from $L$ to $N_b$ we also reduce the estimation step to O($N_K N_L N_b^2$), and the inversion step to O($N_b^3$).
+Reducing the covariance size from $L$ to $N_b$ also reduce the estimation step to O($N_K N_L N_b^2$), and the inversion step to O($N_b^3$).
 The downside is the additional computations required by the frequency transform, which must be weighted up against the gain of reducing $N_b$.
@@
 
-The Low Complexity Adaptive (LCA) beamformer is another take on the optimization problem in (\ref{2_eq_mvdr_def}).
-Instead of calculating the weights like in (\ref{2_eq_weights}), it simply computes the beamformer output for each window in a larger set of windows, and finally applies the one that best fulfils the given optimization criterion.
-If we let $\w_p$ be the $p$'th of a total of $P$ windows, then this can be formulated as~\cite{Synnevag2008}:
+The Low Complexity Adaptive (LCA) beamformer offers a different approach to the optimization problem in (\ref{2_eq_mvdr_def}).
+Instead of calculating weights like MVDR in (\ref{2_eq_weights}), it iterates through a set of precomputed weights and applies the one that best fulfils the optimization criterion.
+If $\w_p$ denotes the $p$'th out of $P$ weight sets, then LCA can be formulated as~\cite{Synnevag2008}:
+%
 \begin{gather}
@@
 \end{gather}
+%
 Note the similarity between (\ref{2_eq_mvdr_def}) and (\ref{2_eq_lca_def}).
-The LCA and MVDR uses essentially the same optimization criterion, but whereas MVDR searches a continuous solution space for the best set of weights to apply, the LCA beamformer only have $P$ solutions to select from.
-However, we have found that once the LCA has 10-20 well designed windows available the quality of the images it produces quickly approaches that of the MVDR beamformer.
+Both LCA and MVDR rely on the same optimization criterion, but whereas MVDR searches a continuous solution space for the optimal weights, LCA only have $P$ solutions to select from.
 
-Although LCA promotes good image quality, its two most attractive traits is perhaps its simplicity and low computational complexity.
-The LCA is essentially a delay-and-sum (DAS) beamformer where the delay step is carried out as usual, but where the weighting and summing is repeated for each of the $P$ windows.
-In its basic form its computational complexity only of O($M\,P$).
+% \subsection{Image quality}
 
@@
 
-\section{Results}
-
-To display the imaging capabilities of the mentioned beamformers, we have processed experimental datasets from the 32 element Kongsberg Maritime HISAS1030 sonar mounted on the HUGIN AUV~\cite{Hansen2009}.
+% \begin{figure}[!t]\centering\label{2_fig_holmengraa}%
+% \graphicsAI[drawing,width=\linewidth]{gfx/plot_holmengraa_L16_Navg1.pdf}%
+% \caption{The 68$\times$9\,m oil tanker Holmengraa lying at a depth of 77\,m.
+The MVDR images were processed with L=16, K=3, and 1\% diagonal loading.
+For beamspace MVDR we set $N_b=L/2$.
+LCA was set up with a window database consisting of 5 Kaiser windows with the window-parameter set in the range $\beta\in[0,10]$, and then each of these were steed in 5 distinct directions leaving a total of 25 windows for the LCA to select from..}
+% \end{figure}
+% 
+% \begin{figure}[!h]\centering%
+% \graphicsAI[drawing,width=.8\linewidth]{gfx/benchmark1_tagged.pdf}%
+% \caption{Benchmarks for the our time-domain and frequency-domain MVDR method running on a GPU.
+The grey line marks the realtime requirement for the HUGIN AUV to perform full-coverage sectorscan images with a ping rate of 3 pings/s.}\label{2_fig_benchmark}
+% \end{figure}
+% 
+% \section{Comparison}
+% 
+% Figure \ref{2_fig_holmengraa} contrasts image quality of the mentioned beamformers.
+The experimental data comes from a HISAS1030 interferometric synthetic aperture sonar (SAS) attached to a HUGIN autonomous underwater vehicle (AUV), both developed by Kongsberg Maritime, Norway.
+HISAS1030 is a fully digital phased array with 32 hydrophones, 100\,kHz center frequency, 30\,kHz bandwidth, 1.2\,m length, 23$^\circ$ half-power beamwidth (HPBW) and 3-4\,cm theoretical resolution both cross- and along-track.
+% 
+% To display the imaging capabilities of the mentioned beamformers, we have processed experimental datasets from the 32 element Kongsberg Maritime HISAS1030 sonar mounted on the HUGIN AUV~\cite{Hansen2009}.
 HISAS1030 is a high resolution synthetic aperture sonar with an array length of 1.2\,m and operating frequency of 100\,kHz, and it was operated in sidescan mode here.
@@
 LCA was designed with a window database consisting of 5 Kaiser windows, each with different window-parameters from the range $\beta\in[0,10]$, and each steered in 5 distinct directions leaving a total of 25 windows for the LCA to select from.
-
-A few benchmarks of these methods are presented in Figure \ref{2_fig_benchmark}.
+% 
+% A few benchmarks of these methods are presented in Figure \ref{2_fig_benchmark}.
 On our test system with a quad-core Intel Xeon E5620, 64Gb of RAM and an nVidia Quadro 6000 card, the MVDR methods managed to create images at rates just over 1\,Mpx/s.
@@
 However, do not consider this comparison fair because while the GPU design is make efficient use of the L1 cache, similar optimisation work was not carried out on the CPU design.
-
-\begin{figure}[!t]\centering%
-\graphicsAI[drawing,width=\linewidth]{gfx/plot_holmengraa_L16_Navg1.pdf}%
-\caption{The 68$\times$9\,m oil tanker Holmengraa lying at a depth of 77\,m.
-The MVDR images were processed with L=16, K=3, and 1\% diagonal loading.
-For beamspace MVDR we set $N_b=L/2$.
-LCA was set up with a window database consisting of 5 Kaiser windows with the window-parameter set in the range $\beta\in[0,10]$, and then each of these were steed in 5 distinct directions leaving a total of 25 windows for the LCA to select from..}\label{2_fig_holmengraa}
-\end{figure}
 
@@
 % \end{table}
-
-\begin{figure}[!h]\centering%
-\graphicsAI[drawing,width=.8\linewidth]{gfx/benchmark1_tagged.pdf}%
-\caption{Benchmarks for the our time-domain and frequency-domain MVDR method running on a GPU.
-The grey line marks the realtime requirement for the HUGIN AUV to perform full-coverage sectorscan images with a ping rate of 3 pings/s.}\label{2_fig_benchmark}
-\end{figure}
-
-\newpage
-\section{Discussion}
 
@@
 
-The MVDR beamformer is a technique that is capable of improving image quality over conventional methods.
-This can be seen in Figure \ref{2_fig_holmengraa}, where the MVDR produces images with cleaner shadows and slightly improved edge definitions.
-Contrary to common belief, it is also possible to create these images relatively fast.
-Through careful optimization work and by running the MVDR on a GPU we have achieved image creation rates of more than 1\,Mpx/s.
-This is an improvement of 2 orders of magnitude over our reference implementation in C.
-To obtain these rates we run MVDR in the time-domain for smaller systems, and in the frequency domain for larger systems.
-
-If this performance is insufficient, we suggest looking at the LCA method.
-It is based on the same optimisation criterion as the MVDR method, but avoids the estimation and inversion step altogether by limiting the weight's solution space to only a few predefined windows.
-As seen the Figure \ref{2_fig_holmengraa}, it produces images with a quality comparable to the MVDR method, but computing the images takes only a fraction of the time.
-
 }

### 2019-10-01 ThesisJo/4_discussion.tex
2019-09-08
2019-09-31

@@
 \end{enumerate}
+
+Time-domain MVDR is---in theory---the optimal way of computing weights that minimize variance (noise) in the image.
 
@@
 % predefined images at false positive rates (FRP) below 20 %, and tied with it above.
+
+The MVDR beamformer is a technique that is capable of improving image quality over conventional methods.
+This can be seen in Figure \ref{2_fig_holmengraa}, where the MVDR produces images with cleaner shadows and slightly improved edge definitions.
+Contrary to common belief, it is also possible to create these images relatively fast.
+Through careful optimization work and by running the MVDR on a GPU we have achieved image creation rates of more than 1\,Mpx/s.
+This is an improvement of 2 orders of magnitude over our reference implementation in C.
+To obtain these rates we run MVDR in the time-domain for smaller systems, and in the frequency domain for larger systems.
+
+If this performance is insufficient, we suggest looking at the LCA method.
+It is based on the same optimisation criterion as the MVDR method, but avoids the estimation and inversion step altogether by limiting the weight's solution space to only a few predefined windows.
+As seen the Figure \ref{2_fig_holmengraa}, it produces images with a quality comparable to the MVDR method, but computing the images takes only a fraction of the time.
+
+However, in practice we find that once the LCA has 10-20 well designed windows available the quality of the images it produces quickly approaches that of the MVDR beamformer.
+
+Although LCA promotes good image quality, its two most attractive traits is perhaps its simplicity and low computational complexity.
+The LCA is essentially a delay-and-sum (DAS) beamformer where the delay step is carried out as usual, but where the weighting and summing is repeated for each of the $P$ windows.
+In its basic form its computational complexity only of O($M\,P$).
+
+\begin{figure}[!t]\centering\label{2_fig_holmengraa}%
+\graphicsAI[drawing,width=\linewidth]{gfx/plot_holmengraa_L16_Navg1.pdf}%
+\caption{The 68$\times$9\,m oil tanker Holmengraa lying at a depth of 77\,m.
+The MVDR images were processed with L=16, K=3, and 1\% diagonal loading.
+For beamspace MVDR we set $N_b=L/2$.
+LCA was set up with a window database consisting of 5 Kaiser windows with the window-parameter set in the range $\beta\in[0,10]$, and then each of these were steed in 5 distinct directions leaving a total of 25 windows for the LCA to select from..}
+\end{figure}
+
+\begin{figure}[!h]\centering%
+\graphicsAI[drawing,width=.8\linewidth]{gfx/benchmark1_tagged.pdf}%
+\caption{Benchmarks for the our time-domain and frequency-domain MVDR method running on a GPU.
+The grey line marks the realtime requirement for the HUGIN AUV to perform full-coverage sectorscan images with a ping rate of 3 pings/s.}\label{2_fig_benchmark}
+\end{figure}
+
+\section{Comparison}
+
+Figure \ref{2_fig_holmengraa} contrasts image quality of the mentioned beamformers.
+The experimental data comes from a HISAS1030 interferometric synthetic aperture sonar (SAS) attached to a HUGIN autonomous underwater vehicle (AUV), both developed by Kongsberg Maritime, Norway.
+HISAS1030 is a fully digital phased array with 32 hydrophones, 100\,kHz center frequency, 30\,kHz bandwidth, 1.2\,m length, 23$^\circ$ half-power beamwidth (HPBW) and 3-4\,cm theoretical resolution both cross- and along-track.
 

### 2019-10-01 ThesisJo/abstract.tex
2019-09-31

@@
 The world is highly dynamic, constantly changing from moment to moment.
-An autonomous device aspiring to navigate it must be situationally perceptive and reactive.
+Any autonomous device aspiring to navigate it must be situationally perceptive and reactive.
 This requires numerous sensors gathering high quality data at high throughput, adaptive signal processing algorithms and powerful parallel hardware capable of near-realtime responses.
+
+This thesis explores three beamformers reconstructing images from acoustic echoes: time-domain MVDR (reference), frequency-domain MVDR (close approximation) and time-domain LCA (educated guessing).
+For all ultimate goal is to preserve suppress noise and interference
 
@@
 The reconstruction process relies heavily on beamforming: the art of delaying and weighting sensor arrays to achieve desired spatial responses.
-A beamformer can be viewed as a spatial filter or model, and works best---as any model---when it is neither too simple nor too complex.
+
+A beamformer can be viewed as a spatial filter or model, and works best---as any model---when it is neither too simple nor complex.
 
@@
 Such a spatial filter, or model, . model complexity should be just right. not too simple, not too complex. bias vs variance.
+
+We begin by accelerating the minimum variance distortionless response (MVDR) beamformer in time-domain, then  in both time- and frequency-domain, frequency-domain MVDR and low-complexity adaptive (LCA).
+
+We find that time-domain MVDR, frequency-domain MVDR and LCA all offer slight improvement in image quality over static methods.
+
+All beamformers are fairly well suited for GPU acceleration because each pixel in the image can be processed independently.
+
+ offer is considered the reference among adaptive beamformers, it offers neglible benefits over
 

### 2019-10-02 ThesisJo/2_background.tex
2019-10-02

@@
 Assume that the array channels are digitally delayed to focus at a pixel with azimuth angle $\theta$ and range sample $n$, such that the delayed data from the $m$th channel can be expressed as $x_m[\theta,n-\Delta_m]\in\mathbb{C}$.
-By making the dependence on $\theta$ implicit and assuming the data to be pre-delayed to the pixel of interest, we can simply the notation to $x_m[\theta,n-\Delta_m] = x_m[n]$.
+By making the dependence on $\theta$ implicit and assuming the data to be pre-delayed to the pixel of interest, we can simplify the notation to $x_m[\theta,n-\Delta_m] = x_m[n]$.
 
@@
 %
-The basic idea behind MVDR is to compute the set of weights that minimizes noise and signal power accumulated by the array, while ensuring unit gain for signals from the pixel of interest.
+The basic idea behind MVDR is to compute the set of weights that minimizes noise and interference accumulated by the array, while ensuring unit gain for signals coming from the pixel of interest.
 With pre-delayed data this can be formulated as follows:

### 2019-10-02 ThesisJo/3_publications.tex
2019-09-30
2019-10-02

@@
 
-\begin{figure}[tp]
+\begin{figure}[tp]\label{III_fig_article_relation}
 \makebox[\linewidth][c]{%
 \graphicsAI[drawing,width=1.3\linewidth]{gfx/article_relation.svg}}
-\caption{Article relations.
+\caption{Article relations (repost of Figure \ref{I_fig_article_relation} in chapter 1).
 The first three articles reconstructs images from acoustic echoes.
 The last article synthesizes image templates from 3D models and object parameters estimated from the reconstructed image.
-The two key traits are: accelerated adaptivity.
+Accelerated adaptivity is key: it enables rapid dynamic responses to mission events.
 }
@@
 This article accelerates time-domain MVDR with GPUs for use in active sonar imaging.
-Time-domain MVDR is the reference method for creating sonar images using the minimum variance optimization, aimed at reducing noise and interference while preserving signal integrity while.
-We refer to it as fully adaptive because it computes array weights analytically from spatial features of the data, yielding array weights from a continuous solution space that are optimal in the minimum variance sense.
+Time-domain MVDR is the reference method for creating sonar images using the minimum variance optimization, aimed at reducing noise and interference while preserving signal integrity.
+We refer to it as fully adaptive because it computes array weights analytically from spatial features in the data, yielding array weights from a continuous solution space that are optimal in the minimum variance sense.
 Like any data-driven algorithm, it requires statistically distinguishable signal and noise.

### 2019-10-02 ThesisJo/4_discussion.tex
2019-10-02

@@
 \end{enumerate}
-
-Time-domain MVDR is---in theory---the optimal way of computing weights that minimize variance (noise) in the image.
-
-Time-domain MVDR is the reference method for creating sonar images using the minimum variance criterion.
-It computes array weights analytically, yielding optimal weights in the minimum variance sense.
-Initially, this may seem ideal, but its use is cumbersome: parameters for temporal averaging, spatial averaging and regularization must often be fine-tuned ad-hoc to obtain improved image quality.
-GPU acceleration mitigates the high computational complexity resulting from estimating and inverting a spatial covariance matrix, yielding real-time performance.
+% 
+% Time-domain MVDR is---in theory---the optimal way of computing weights that minimize variance (noise) in the image. 
+%
 
@@
 
+MVDR is the reference method for creating sonar images using the minimum variance criterion.
+It computes array weights analytically, yielding optimal weights in the minimum variance sense.
+Initially, this may seem ideal, but its use is cumbersome: parameters for temporal averaging, spatial averaging and regularization must often be fine-tuned ad-hoc to obtain improved image quality.
+When using time-domain MVDR in systems with less than 32 channels, GPU acceleration mitigates the high computational complexity in estimating and inverting a spatial covariance matrix.
+For systems with narrow-band, narrow-beam, or both, the fre , yielding real-time performance.
+
+\section{Comparison}
+
+reconstruction/Beamforming:
+%
+\begin{itemize}
+\item \emph{Image quality} is similar for our adaptive beamformers, but improved upon compared to static methods.
+%
+\item \emph{Solution space} is continuous for MVDR and discrete for LCA, but similarly shaped. 
+%
+\item \emph{Model complexity} has a sweet spot, a balance point where its estimates offer an ideal compromise between bias and variance. MVDR depend on 
+%
+\item \emph{Acceleration} of MVDR through algorithmic reductions and implementation on graphics processing units (GPUs) yields real-time responses in time-domain up to 32 channels and in frequency-domain for any system size with narrow-band, narrow-beam, or both.
+\end{itemize}
+
+synthesis/simulation:
+
+%A classification model should be neither too simple (high bias or underfitted)
+%nor complex (high variance or underfit), but strike a balance between the
+%two that captures statistically significant details of the object but not the
+%noise. Higher detail-to-noise ratios favor more complex models and occur at
+
 The MVDR beamformer is a technique that is capable of improving image quality over conventional methods.
@@
 \end{figure}
-
-\section{Comparison}
 

### 2019-10-02 ThesisJo/abstract.tex
2019-10-02

--- 2018-02-16 ThesisJo/abstract.tex
@@
-\input{variables}\ifMonolithic\else\input{header}\input{subheader}\fi
+% \input{variables}\ifMonolithic\else\input{header}\input{subheader}\fi
 %~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
 
-\chapter{Abstract}
+The world is highly dynamic.
+Any device aspiring to master it must be situationally perceptive and reactive.
+This requires numerous sensors gathering high quality data at high throughput, adaptive signal processing algorithms and powerful parallel hardware capable of delivering swift responses.
+In our case of sonar-equipped autonomous underwater vehicles and clinical ultrasound devices, this process revolves around acoustic images.
+
+This thesis explores methods to accelerate adaptive acoustic image reconstruction and synthesis, through beamforming and simulation, respectively.
+Beamforming is the main attraction: the art of delaying and weighting sensor arrays to craft desired spatial responses.
+When aiming to minimize image noise and interference, two good candidates are the minimum variance distortionless response (MVDR) and low complexity adaptive (LCA) beamformer.
+We study time-domain MVDR (reference), frequency-domain MVDR (close approximation) and time-domain LCA (educated guessing), and conclude that:
+%
+\begin{itemize}
+\item \emph{Image quality} is similar for our adaptive beamformers, but improved upon compared to static methods.
+%
+\item \emph{Solution space} is continuous for MVDR and discrete for LCA, but similarly shaped. 
+%
+\item \emph{Model complexity} has a sweet spot, a balance point where its estimates offer an ideal compromise between bias and variance. MVDR depend on 
+%
+\item \emph{Acceleration} of MVDR through algorithmic reductions and implementation on graphics processing units (GPUs) yields real-time responses in time-domain up to 32 channels and in frequency-domain for any system size with narrow-band, narrow-beam, or both.
+LCA is inherently fast and extremely well suited for acceleration,
+\end{itemize}
+%
+Finally, we perform image synthesis of common objects in the seabed with a GPU accelerated simulator.
+These are in turn used to perform adaptive automatic target recognition (ATR).
+The result is a more dynamic and flexible system yielding improved image quality, image quality and classification performance.
+Finally, we describe this system with a new navigational notation system, extended with topological concepts from robotics and homogeneous coordinates.
+It helps keeping track of complex geometries, numerical representations and higher level abstractions such as manipulating rigid objects.
+
+%
+%
+%
+%brings most adaptivity to the table, but also greatest computational load and model complexity.
+%
+%
+%
+%
+%computational complexity increases but can be handled through algorithmic reduction and hardware acceleration, , for the adaptive beamformers are better than static methods, but ; \emph{user friendliness} is better for LCA than MVDR;
+%
+%
+%MVDR yields weights from a continuous solution space, but requires ad-hoc tuning of robustification parameters, is harder to understand and has high computational complexity. 
+%
+%The reconstruction process relies heavily on beamforming: the art of delaying and weighting sensor arrays to achieve desired spatial responses.
+Our adaptive beamformer reference, minimum variance distortionless response (MVDR), computes the weights analytically from an estimate of the full-array covariance matrix.
+%
+%
+%
+%, both in time- and frequency domain. Then we ,  begin our study of a reference method  MVDR, frequency-domain MVDR and LCA all offer slight improvement in image quality over static methods. 
+%
+%All methods implemented on the GPU except LCA.
+%
+%A beamformer can be viewed as a spatial filter or model, and works best---as any model---when it is neither too simple nor complex. 
+%
+%
+%
+%We begin by accelerating the minimum variance distortionless response (MVDR) beamformer in time-domain, then  in both time- and frequency-domain, frequency-domain MVDR and low-complexity adaptive (LCA).
+% 
+%
+%All beamformers are fairly well suited for GPU acceleration because each pixel in the image can be processed independently.
+%
+% offer is considered the reference among adaptive beamformers, it offers neglible benefits over 
+%
+%sensory faculty
+%
+%as adaptive as possible with the least possible computational load.
+%
+%
+%gather information
+%dynamically analyze it
+%and react to it
+%
+%adapt
+%persist
+%prevail
+%
+%Adaptive algorithm and the computation power to drive it.
+%
+%Cognition is the mental action or process of acquiring knowledge and understanding through thought, experience and the senses. 
+%
+%The intelligence and autonomy of acoustical imaging devices require more sensory data, algorithmic adaptivity and processing power. All are vital and equally important. 
+%
+%improves with more high quality sensory data, sophisticated adaptive signal processing and the hardware resources to deal with it all. 
+%
+%more sensory data, algorithmic adaptivity and processing power.
+%
+%autonomy improves with data amount, adaptive processing, processing power.
+%
+%information gathering and processing. 
+%
+%Key challenges that are addressed include active sonar and
+%
+%- Active sonar and clinical ultrasound
+%- Different ways to achieve adaptivity. Analytical assessment of the data, computing weights. Approximating it. Trial-and-error. Different means to the same adaptivity
+%- Computational complexity
+%- Hardware acceleration
+%- 
+%
+%Spatial sensitivity is like a balloon; squeezing it on one side inflates it on the other. 
+%
+%
 
 %~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
-\ifMonolithic\else\input{subfooter}\fi
+% \ifMonolithic\else\input{subfooter}\fi

### 2019-10-03 Simulator/bio/herman_midelfart.tex
2019-10-03

\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{bio/herman.eps}}]{Herman Midelfart}
held a M.Sc. and Ph.D. degree (2003) in genomics from the Norwegian University of Science and Technology (NTNU), Norway. Since 2007 he advanced research on automatic target recognition (ATR) in synthetic aperture sonar (SAS) at the Norwegian Defence Research Establishment, Kjeller, Norway.

Herman passed away in 2015. This work was finalized in his memory.
\end{IEEEbiography}

### 2019-10-03 ThesisJo/2_background.tex
2019-10-03

@@
 % \\
+\begin{figure}[tp]\label{2_fig_article_relation}
+\makebox[\linewidth][c]{%
+\graphicsAI[drawing,width=1.3\linewidth]{gfx/article_relation.svg}}
+\caption{Method relations.
+This thesis explores three beamformers reconstructing images from acoustic echoes: time-domain MVDR (reference), frequency-domain MVDR (close approximation) and time-domain LCA (educated trial-and-error).
+Background on synthesis of image templates from the reconstructed images is presented in article IV.
+}
+\end{figure}
+
+\begin{figure}[!t]\label{2_fig_geometry}\centering
+\makebox[\linewidth][c]{%
+\graphicsAI[width=.7\linewidth]{gfx/geometry.svg}}
+\caption{
+An array with $M$ sensors has reference frame $\udot{A} \triangleq (\udot{A},\uvec{A})$ attached to it, where $\udot{A}\in\mathbb{A}$ is its absolute position and $\uvec{A}\triangleq\big(\hat{x}_{A},\hat{y}_{A},\hat{z}_{A}\big)\in\mathbb{V}^3$ its absolute orientation (\,$\hat{\cdot}$ denoting unit length).
+Impinging on the array at angle $\theta$ is a monochromatic plane wave: a single frequency wave from a source ${S}$ far away relative to the wavelength and array length (far field).
+Notation details are presented in Paper IV.
+}
+\end{figure}
 
@@
 \begin{align}
-f_\textrm{plane}(\vec{p},t)  &= A e^{j(\omega t - \vec{k} \cdot \vec{p})} \\
-f_\textrm{spheric}(\vec{p},t) &= \frac{A}{\lVert \vec{p} \rVert} e^{j(\omega t - \lVert\vec{k}\rVert \lVert \vec{p} \rVert)},
+f_\textrm{plane}(\vec{p},t)  &= A e^{j(\omega t - \vec{k} \cdot \vec{p}_{OS})} \\
+f_\textrm{spheric}(\vec{p},t) &= \frac{A}{\lVert \vec{p}_{OS} \rVert} e^{j(\omega t - \lVert\vec{k}\rVert \lVert \vec{p}_{OS} \rVert)},
 \end{align}
 % 
-Here $A$ is amplitude, $\omega$ is temporal wave frequency, $t$ is time, $\lVert\vec{k}\rVert=\frac{\omega}{c}=\frac{2\pi}{\lambda}$ is spatial wave frequency (wavenumber) and $\vec{p}$ is position of the source relative to the observer.
+Here $A$ is amplitude, $\omega$ is temporal wave frequency, $t$ is time, $\lVert\vec{k}\rVert=\frac{\omega}{c}=\frac{2\pi}{\lambda}$ is spatial wave frequency (wavenumber) and $\vec{p}_{OS}$ is position of the source $\mathcal{S}$ relative to the observer $\mathcal{O}$.
+We model the wavefield as superimposed plane waves under far-field conditions (when distance to the source is large relative to the wavelength and array length) and as superimposed spherical waves under near-field conditions.
 
@@
 After stripping the carrier frequency by heterodyning and the signal signature by matched filtering (Fig.
-\ref{I_beamforming}), an image is formed by coherently combining the receiver outputs to focus at one angle (or point) at a time.
+\ref{2_fig_article_relation}), an image is formed by coherently combining the receiver outputs to focus at one angle (or point) at a time.
 The principle of adjusting the array's focus in range and direction is commonly referred to as beamforming, and involves assigning suitable delays and weights to the array's channels.
-
-\begin{figure}[tp]\label{II_fig_article_relation}
-\makebox[\linewidth][c]{%
-\graphicsAI[drawing,width=1.3\linewidth]{gfx/article_relation.svg}}
-\caption{Method relations.
-This thesis explores three beamformers reconstructing images from acoustic echoes: time-domain MVDR (reference), frequency-domain MVDR (close approximation) and time-domain LCA (educated trial-and-error).
-Background on synthesis of image templates from the reconstructed images will be presented in article IV.
-}
-\end{figure}
 
@@
 
-Imagine an array with $M$ sensors with a reference frame $\ubar{A} \triangleq (\udot{A},\uvec{A})$ attached to it, where $\udot{A}\in\mathbb{A}$ is its absolute position and $\uvec{A}\triangleq\big(\hat{x}_{A},\hat{y}_{A},\hat{z}_{A}\big)\in\mathbb{V}^3$ its absolute orientation (where $\hat{\cdot}$ denotes unit length).
-Assume that the array channels are digitally delayed to focus at a pixel with azimuth angle $\theta$ and range sample $n$, such that the delayed data from the $m$th channel can be expressed as $x_m[\theta,n-\Delta_m]\in\mathbb{C}$.
-By making the dependence on $\theta$ implicit and assuming the data to be pre-delayed to the pixel of interest, we can simplify the notation to $x_m[\theta,n-\Delta_m] = x_m[n]$.
+\begin{figure}[tp]\label{2_fig_scenario_das}
+\makebox[\linewidth][c]{%
+\graphicsAI[width=1\linewidth]{gfx/scenario_interference_das.svg}}
+\caption{DAS beampattern is static (conventional beamforming). Interference suppression at the first mainlobe is -13\,dBr.}
+\end{figure}
 
-The beamformer output $z[n]\in\mathbb{C}$ is by definition the weighted sum of all the delayed data samples:
+\begin{figure}[tp]\label{2_fig_scenario_mvdr}
+\makebox[\linewidth][c]{%
+\graphicsAI[width=1\linewidth]{gfx/scenario_interference_mvdr.svg}}
+\caption{MVDR beampattern is dynamic and continuous (adaptive beamforming). Interference suppression is good for prevalent noise sources if a good estimate of the spatial statistics can be obtained.}
+\end{figure}
+
+\begin{figure}[tp]\label{2_fig_scenario_lca}
+\makebox[\linewidth][c]{%
+\graphicsAI[width=1\linewidth]{gfx/scenario_interference_lca.svg}}
+\caption{LCA beampattern is dynamic and discrete (adaptive beamforming). Interference suppression is on-par with MVDR with 15 weight sets to choose from, nearly as good even with 6 weight sets.}
+\end{figure}
+
+Imagine an array with $M$ sensors at positions $\udot{0},\udot{1},\dots, \udot{m}, \dots, \udot{M{\!-\!1}}$ with reference frame $\ubar{A} \triangleq (\udot{A},\uvec{A})$ attached to its phase center (Figure \ref{2_fig_geometry}), where $\udot{A}\in\mathbb{A}$ is its absolute position and $\uvec{A}\triangleq\big(\hat{x}_{A},\hat{y}_{A},\hat{z}_{A}\big)\in\mathbb{V}^3$ its absolute orientation ( $\hat{\cdot}$ denoting unit length).
+Assume that the array channels are
+digitally delayed to focus at a pixel with azimuth angle $\theta$ and range sample $n$, such that the delayed data from the $m$th channel can be expressed as $x_m[\theta,n-\Delta_m]\in\mathbb{C}$.
+Focus at source $\udot{S}$ requires delays akin to
+%
+\begin{align}
+\Delta_m[n] = \frac{\lVert \vec{p}_{m}\rVert - \lVert \vec{p}_{0S}\rVert}{c}.\label{2_eq_delta}
+\end{align}
+%
+where $\vec{p}_{m}$ denotes position vector $\udot{A}$ to $\udot{m}$ and $\vec{p}_{0S}$ denotes position vector $\udot{0}$ to $\udot{S}$, respectively.
+By making the dependence on $\theta$ implicit and assuming the data to be pre-delayed to the pixel of interest, we simplify the notation to $x_m[n]$.
+By definition, the beamformer output $z[n]\in\mathbb{C}$ is the weighted sum of delayed data samples:
 %
@@
 %
-where $w_m\in\mathbb{C}$ is the weight factor assigned to channel $m$.
-With static weights this is referred to as the conventional delay-and-sum (DAS) beamformer.
+where $w_m\in\mathbb{C}$ is the weight factor assigned to channel $m$. Spatial response of the weights when the array is steered to broadside can be characterized by the beampattern,
+%
+\begin{align}
+W(\vec{k}) = \sum_m^{M-1} w_m^* e^{j\vec k \cdot \vec{p}_m}.\label{2_eq_beampattern}
+\end{align}
+%
+%
+%
+%
+%. Beampattern is broadside-steered beamformer output in the presence of a single plane wave, as a function of the wave's angle.
+%%Steered response is beamformer's output in the presence of some wave field, as a function of steering angle. 
+%
+The beamformer is called delay-and-sum (DAS) or conventional if the weights are static.
 A variety of weighting functions exist for trading lateral resolution for improved noise suppression (contrast), but a compromise between the two is unavoidable~\cite{Harris1978}.
@@
 
-% Accelerated adaptivity is key: it enables rapid dynamic responses to mission events. }
-
 \subsection{Time-domain MVDR}
 
-\begin{figure}[t]\centering%
-\graphicsAI[drawing,width=1\linewidth]{gfx/implementation.svg}%
-\caption{MVDR beamforming. For one of the total number of pixels in range and azimuth, $N_y$ and $N_x$,\newline
-1. an $L\times{}L$ sample covariance matrix $\eR$ is computed, \newline
-2. the term $\eR^{-1}\1$ is found using a linear equation solver,\newline
-3. and the beamformer output is computed from $z$ from (\ref{finalZ}), where $\w$ is found by substituting $\eR^{-1}\1$ into (\ref{2_eq_weights}). }\label{mvdr_beamforming}
-\end{figure}
+%\begin{figure}[t]\centering%
+%\graphicsAI[drawing,width=1\linewidth]{gfx/implementation.svg}%
+%\caption{MVDR beamforming. For one of the total number of pixels in range and azimuth, $N_y$ and $N_x$,\newline
+%1. an $L\times{}L$ sample covariance matrix $\eR$ is computed, \newline
+%2. the term $\eR^{-1}\1$ is found using a linear equation solver,\newline
+%3. and the beamformer output is computed from $z$ from (\ref{finalZ}), where $\w$ is found by substituting $\eR^{-1}\1$ into (\ref{2_eq_weights}). }\label{mvdr_beamforming}
+%\end{figure}
 %

### 2019-10-03 ThesisJo/4_discussion.tex
2019-10-03

--- 2019-09-02 ThesisJo/4_discussion.tex
@@
 
-The main contributions of this thesis can be summarized as follows:
+The main contributions of this thesis, along with brief discussion and onlook follows.
+
+\section{Summary}
 
@@
 \end{enumerate}
-
-Time-domain MVDR is the reference method for creating sonar images using the minimum variance criterion.
-It computes array weights analytically, yielding optimal weights in the minimum variance sense.
-Initially, this may seem ideal, but its use is cumbersome: parameters for temporal averaging, spatial averaging and regularization must often be fine-tuned ad-hoc to obtain improved image quality.
-GPU acceleration mitigates the high computational complexity resulting from estimating and inverting a spatial covariance matrix, yielding real-time performance.
 
@@
 
+\section{Discussion}
+
+MVDR yield optimal weights in the minimum variance sense but is computationally complex and user unfriendly.
+Parameters for temporal averaging, spatial averaging and regularization must often be fine-tuned ad-hoc to obtain improved image quality.
+
+LCA offers similar image quality as MVDR while being simple, fast, inherently robust and ideal for hardware acceleration.
+The solution space is continuous for MVDR (after robustification) and discrete for LCA, but similarly shaped.
+This suggests that LCA samples the solution space well.
+
+Any adaptive model---be it a spatial filter or classifier---should neither be too simple (high bias or underfitted) nor complex (high variance or overfitted), but hit a balance where it adapts to statistically significant details but not noise.
+Unconstrained MVDR is too aggressive and ``overfits'' the data, but its constrained form works well.
+
+%
+%%A classification model should be neither too simple (high bias or underfitted)
+%%nor complex (high variance or underfit), but strike a balance between the
+%%two that captures statistically significant details of the object but not the
+%%noise. Higher detail-to-noise ratios favor more complex models and occur at
+
+GPU acceleration and algorithmic reductions largely mitigates the computational argument against MVDR in time-domain up to 32 channels and in time-domain for larger systems granted it is narrow-beam, narrow-band, or both.
+LCA is entirely data-parallel thus ideal for GPUs, but runs faster unoptimized on CPUs than accelerated MVDR.%  Real-time performance is likely possible.
+
+%
+
 % We grant them perception by performing automatic target recognition (ATR) with an
@@
 
-In clinical ultrasound imaging
-
 \section{Onlook}
 
-Machine intelligence requires high fidelity sensory data, sophisticated adaptive signal processing and massive computing power.
-If history is any indication, the future promises more of everything in these categories.
-
-The future holds a promise of more of everything: more sensors (data), more signal processing and more processing power.
-
-\emph{Big data} and \emph{Internet of things} (IoT) and \emph{big data} will continue to empower us with ever more information.
-
-\emph{Machine learning} has made leaps with the combination of deep neural networks and big data.
-
-\subsection{Sensors}
-
-\emph{Phased arrays}
-
-\section{Signal processing}
-
-\emph{Software defined sonar}
-
-\subsection{Hardware}
+Although receive-side adaptive beamforming yield slight image quality improvement in our active imaging systems, we believe its potential on the transmit-side to be greater.
+Ideally the two should work in synergy: a receiver telling a transceiver what data it needs.
+For example, a strong interfering source is best suppressed both on transmit and receive.
+In a dynamic world a low latency feedback loop is key to making this work.
 
 \emph{System on Chips} (Soc) are becoming increasingly popular as a heterogeneous computing platform.
+One of these chips, the Ultrascale+ Radio Frequency SoC (RFSoC) from Xilinx, contains a quad-core CPU, dual-core real-time processor, GPU, Field Programmable Gate Array (FPGA) and up to 16 Analog-to-Digital and Digital-to-Analog Converters (ADCs and DACs)---all in one CMOS chip.
+It targets the need for multiple-input-multiple-output (MIMO) can easily replace most analog front-end electronics in a sonar or ultrasound system with a software-defined solution, yielding better flexibility, reduced design complexity, improved bandwidth and lowered latency, all in a tiny physical footprint.
+
+%
+%
+%The MVDR beamformer is a technique that is capable of improving image quality over conventional methods.
+This can be seen in Figure \ref{2_fig_holmengraa}, where the MVDR produces images with cleaner shadows and slightly improved edge definitions.
+Contrary to common belief, it is also possible to create these images relatively fast.
+Through careful optimization work and by running the MVDR on a GPU we have achieved image creation rates of more than 1\,Mpx/s.
+This is an improvement of 2 orders of magnitude over our reference implementation in C.
+To obtain these rates we run MVDR in the time-domain for smaller systems, and in the frequency domain for larger systems.
+%
+%If this performance is insufficient, we suggest looking at the LCA method.
+It is based on the same optimisation criterion as the MVDR method, but avoids the estimation and inversion step altogether by limiting the weight's solution space to only a few predefined windows.
+As seen the Figure \ref{2_fig_holmengraa}, it produces images with a quality comparable to the MVDR method, but computing the images takes only a fraction of the time.
+%
+%However, in practice we find that once the LCA has 10-20 well designed windows available the quality of the images it produces quickly approaches that of the MVDR beamformer.
+%
+%Although LCA promotes good image quality, its two most attractive traits is perhaps its simplicity and low computational complexity.
+The LCA is essentially a delay-and-sum (DAS) beamformer where the delay step is carried out as usual, but where the weighting and summing is repeated for each of the $P$ windows.
+In its basic form its computational complexity only of O($M\,P$).
+%
+%
+%\begin{figure}[!t]\centering\label{2_fig_holmengraa}%
+%\graphicsAI[drawing,width=\linewidth]{gfx/plot_holmengraa_L16_Navg1.pdf}%
+%\caption{The 68$\times$9\,m oil tanker Holmengraa lying at a depth of 77\,m.
+The MVDR images were processed with L=16, K=3, and 1\% diagonal loading.
+For beamspace MVDR we set $N_b=L/2$.
+LCA was set up with a window database consisting of 5 Kaiser windows with the window-parameter set in the range $\beta\in[0,10]$, and then each of these were steed in 5 distinct directions leaving a total of 25 windows for the LCA to select from..}
+%\end{figure}
+%
+%\begin{figure}[!h]\centering%
+%\graphicsAI[drawing,width=.8\linewidth]{gfx/benchmark1_tagged.pdf}%
+%\caption{Benchmarks for the our time-domain and frequency-domain MVDR method running on a GPU.
+The grey line marks the realtime requirement for the HUGIN AUV to perform full-coverage sectorscan images with a ping rate of 3 pings/s.}\label{2_fig_benchmark}
+%\end{figure}
+%
+%
+%
+%Figure \ref{2_fig_holmengraa} contrasts image quality of the mentioned beamformers.
+The experimental data comes from a HISAS1030 interferometric synthetic aperture sonar (SAS) attached to a HUGIN autonomous underwater vehicle (AUV), both developed by Kongsberg Maritime, Norway.
+HISAS1030 is a fully digital phased array with 32 hydrophones, 100\,kHz center frequency, 30\,kHz bandwidth, 1.2\,m length, 23$^\circ$ half-power beamwidth (HPBW) and 3-4\,cm theoretical resolution both cross- and along-track.
+%
+%In clinical ultrasound imaging 
+%
+%\section{Onlook}
+%
+%
+%Machine intelligence requires high fidelity sensory data, sophisticated adaptive signal processing and massive computing power.
+If history is any indication, the future promises more of everything in these categories.
+%
+%The future holds a promise of more of everything: more sensors (data), more signal processing and more processing power.
+%
+%\emph{Big data} and \emph{Internet of things} (IoT) and \emph{big data} will continue to empower us with ever more information.  
+%
+%
+%\emph{Machine learning} has made leaps with the combination of deep neural networks and big data.  
+%
+%
+%\subsection{Sensors}
+%
+%\emph{Phased arrays} 
+%
+%\section{Signal processing}
+%
+%\emph{Software defined sonar}
+%
+%
+%\subsection{Hardware}
+%
+%
+%
+%\emph{System on Chips} (Soc) are becoming increasingly popular as a heterogeneous computing platform.
 One of these chips, the Ultrascale+ Radio Frequency SoC (RFSoC) from Xilinx, contains a quad-core CPU, a dual-core real-time processor, a GPU, Field Programmable Gate Array (FPGA) resources and up to 16 Analog-to-Digital and Digital-to-Analog Converters (ADCs and DACs)---all in one CMOS chip.
@@
 Although the GPU
-
-\emph{Phased arrays} samples and records wavefields for digital processing. It is a prerequisite for many of the image reconstruction algorithms
-
-\subsection{Software}
-
-% % 
-% % \item System on Chips with integrated ADCs/DACs. Big power in small package.
-% % \item Compact, flexible, direct conversion
-% % \item Software defined signal generation, filtering, synchronization, rate-conversion and mixing.
-% % \item Multipurpose, hardware reuse, easier implementation of advanced signal processing, faster development and lower price.
-% % \item 
-% % \item Compact multi-band Wifi
-% % \item MIMO (in new wifi). Breaks array SNR boost. Not viable for long range.
-% % \item Phones: Wifi 2.4-2.5G, 5G (massive MIMO, heavy SDR, 600M-6G), LTE (4G, Long Term Evolution 800M), GSM (2G, Global System for Mobile communications 900M,1.8-1.9M), UMTS (3G, Universal Mobile Telecommunications System 1.9-2.1M), GPS (Global Positioning System 1.5G), Bluetooth (2.4G), NFC (Near-Field Communications 13.56M), FM (100M).
-Multi-radio in a single SDR package!
+%
+%\emph{Phased arrays} samples and records wavefields for digital processing. It is a prerequisite for many of the image reconstruction algorithms
 %
@@
 %
+%\subsection{Software}
 %
 %
-% 
-% 
-% 
-% 
-% 
-% The potential for intelligence improves with the amount of high quality data, the sophistication of the signal processing algorithms and the hardware processing that glues it all together. 
-% 
-% The acquisition of ever more data 
-% Instruments acquire ever 
-% 
-% The intelligence of autonomous instruments require boils down to a tight and fast integration of data collection, processing relies on high quality data, processing and to three key ingredients: high quality sensory data, fast data processing and adaptive signal processing.
-% 
-% More high quality data
-% The processing power to digest it fast
-% Adaptive 
-% 
-% \section{Trends}
-% 
-% The future is faster, smarter.
-% 
-% \subsection{Hardware}
-% 
-% \begin{itemize}
-% \item Ever increasing processing speed.
-% \begin{itemize}
-% 	\item Choose the best tool (algorithm) for the job
-% 	\item More complex not always better, but slower
-% 	\item Extracting the essential information is key 
-% 	\item Exploit hardware resources available
-%    \item Intel entering the discrete graphics stage
-%    \end{itemize}
-% \item Software defined transceiver/receivers
-% \begin{itemize}
-% \item System on Chips with integrated ADCs/DACs. Big power in small package.
-% \item Compact, flexible, direct conversion
-% \item Software defined signal generation, filtering, synchronization, rate-conversion and mixing.
-% \item Multipurpose, hardware reuse, easier implementation of advanced signal processing, faster development and lower price.
-% \item 
-% \item Compact multi-band Wifi
-% \item MIMO (in new wifi). Breaks array SNR boost. Not viable for long range.
-% \item Phones: Wifi 2.4-2.5G, 5G (massive MIMO, heavy SDR, 600M-6G), LTE (4G, Long Term Evolution 800M), GSM (2G, Global System for Mobile communications 900M,1.8-1.9M), UMTS (3G, Universal Mobile Telecommunications System 1.9-2.1M), GPS (Global Positioning System 1.5G), Bluetooth (2.4G), NFC (Near-Field Communications 13.56M), FM (100M).
+%% % 
+%% % \item System on Chips with integrated ADCs/DACs. Big power in small package.
+%% % \item Compact, flexible, direct conversion
+%% % \item Software defined signal generation, filtering, synchronization, rate-conversion and mixing.
+%% % \item Multipurpose, hardware reuse, easier implementation of advanced signal processing, faster development and lower price.
+%% % \item 
+%% % \item Compact multi-band Wifi
+%% % \item MIMO (in new wifi). Breaks array SNR boost. Not viable for long range.
+%% % \item Phones: Wifi 2.4-2.5G, 5G (massive MIMO, heavy SDR, 600M-6G), LTE (4G, Long Term Evolution 800M), GSM (2G, Global System for Mobile communications 900M,1.8-1.9M), UMTS (3G, Universal Mobile Telecommunications System 1.9-2.1M), GPS (Global Positioning System 1.5G), Bluetooth (2.4G), NFC (Near-Field Communications 13.56M), FM (100M).
 Multi-radio in a single SDR package!
-% \end{itemize}
-% \item Short feedback loops allow real-time adaptivity
-% \end{itemize}
-% 
-% 
-% \subsection{Algorithms}
-% 
-% \begin{itemize}
-% \item Intelligence - Ability to process data fast, and retain key information
-% \item Adaptive methods ever more attractive.
-%    \begin{itemize}
-%    \item Processing more data to increase significance of extracted information
-%    \item Reducing reaction time: Act when the data is fresh/valid.
-%    \end{itemize}
-% \item Better image quality
-% \item Improved autonomy
-% \end{itemize}
-% 
-% 
-% \subsection{Connectivity}
-% 
-% \begin{itemize}
-% \item Internet of Things
-% \item Big data
-% \item Intelligent mesh of nodes
-% \end{itemize}
-% 
-% \section{Risks \& Challenges}
-% 
-% 
-% 
-% \begin{itemize}
-% \item Improving imaging systems by making them smarter.
-% \item Big data - more processing power.
-% \item Faster IS smarter
-%  \begin{itemize}
-%  	\item Short feedback loops allow real-time adaptivity
-%  	\item More data contains more information
-%  	\item Faster learning
-%  \end{itemize}
-% \end{itemize}
-% 
-% 
-% 
-% %Discussion and future work
-% %   Summary and discussion
-% %      Simulator:
-% %      - The way to train and constrain deep learning algorithms?
-% %      MVDR:
-% %      - Analytical optimization - full adaptivity.
-% %      - Heavy on computations
-% %      LCA:
-% %      - Optimization by trial and error - discrete solution space.
-% %      - Works almost as well as MVDR in most cases
-% %      - Fast!
-% %      - Well suited for hardware
-% %      
-% %      Lessions from deep learning:
-% %      - 
-% %      
-% %   Future work
-% 
-% 
-% \section{MVDR}
-% 
-% Recent developments:
-% - A low complexity minimum variance beamformer for ultrasound imaging using dominant mode rejection~\cite{Jiang2016}
-% - Google deep beamforming
-% 
-% \begin{itemize}
-% \item SDR - Entirely software defined radio, radar and sonar, bigger potential for adaptive methods
-% \item Cognitive radar/sonar.
+%% 
+%% 
+%% 
+%% 
+%% 
+%% 
+%% 
+%% 
+%% 
+%% 
+%% The potential for intelligence improves with the amount of high quality data, the sophistication of the signal processing algorithms and the hardware processing that glues it all together. 
+%% 
+%% The acquisition of ever more data 
+%% Instruments acquire ever 
+%% 
+%% The intelligence of autonomous instruments require boils down to a tight and fast integration of data collection, processing relies on high quality data, processing and to three key ingredients: high quality sensory data, fast data processing and adaptive signal processing.
+%% 
+%% More high quality data
+%% The processing power to digest it fast
+%% Adaptive 
+%% 
+%% \section{Trends}
+%% 
+%% The future is faster, smarter.
+%% 
+%% \subsection{Hardware}
+%% 
+%% \begin{itemize}
+%% \item Ever increasing processing speed.
+%% \begin{itemize}
+%% 	\item Choose the best tool (algorithm) for the job
+%% 	\item More complex not always better, but slower
+%% 	\item Extracting the essential information is key 
+%% 	\item Exploit hardware resources available
+%%    \item Intel entering the discrete graphics stage
+%%    \end{itemize}
+%% \item Software defined transceiver/receivers
+%% \begin{itemize}
+%% \item System on Chips with integrated ADCs/DACs. Big power in small package.
+%% \item Compact, flexible, direct conversion
+%% \item Software defined signal generation, filtering, synchronization, rate-conversion and mixing.
+%% \item Multipurpose, hardware reuse, easier implementation of advanced signal processing, faster development and lower price.
+%% \item 
+%% \item Compact multi-band Wifi
+%% \item MIMO (in new wifi). Breaks array SNR boost. Not viable for long range.
+%% \item Phones: Wifi 2.4-2.5G, 5G (massive MIMO, heavy SDR, 600M-6G), LTE (4G, Long Term Evolution 800M), GSM (2G, Global System for Mobile communications 900M,1.8-1.9M), UMTS (3G, Universal Mobile Telecommunications System 1.9-2.1M), GPS (Global Positioning System 1.5G), Bluetooth (2.4G), NFC (Near-Field Communications 13.56M), FM (100M).
+Multi-radio in a single SDR package!
+%% \end{itemize}
+%% \item Short feedback loops allow real-time adaptivity
+%% \end{itemize}
+%% 
+%% 
+%% \subsection{Algorithms}
+%% 
+%% \begin{itemize}
+%% \item Intelligence - Ability to process data fast, and retain key information
+%% \item Adaptive methods ever more attractive.
+%%    \begin{itemize}
+%%    \item Processing more data to increase significance of extracted information
+%%    \item Reducing reaction time: Act when the data is fresh/valid.
+%%    \end{itemize}
+%% \item Better image quality
+%% \item Improved autonomy
+%% \end{itemize}
+%% 
+%% 
+%% \subsection{Connectivity}
+%% 
+%% \begin{itemize}
+%% \item Internet of Things
+%% \item Big data
+%% \item Intelligent mesh of nodes
+%% \end{itemize}
+%% 
+%% \section{Risks \& Challenges}
+%% 
+%% 
+%% 
+%% \begin{itemize}
+%% \item Improving imaging systems by making them smarter.
+%% \item Big data - more processing power.
+%% \item Faster IS smarter
+%%  \begin{itemize}
+%%  	\item Short feedback loops allow real-time adaptivity
+%%  	\item More data contains more information
+%%  	\item Faster learning
+%%  \end{itemize}
+%% \end{itemize}
+%% 
+%% 
+%% 
+%% %Discussion and future work
+%% %   Summary and discussion
+%% %      Simulator:
+%% %      - The way to train and constrain deep learning algorithms?
+%% %      MVDR:
+%% %      - Analytical optimization - full adaptivity.
+%% %      - Heavy on computations
+%% %      LCA:
+%% %      - Optimization by trial and error - discrete solution space.
+%% %      - Works almost as well as MVDR in most cases
+%% %      - Fast!
+%% %      - Well suited for hardware
+%% %      
+%% %      Lessions from deep learning:
+%% %      - 
+%% %      
+%% %   Future work
+%% 
+%% 
+%% \section{MVDR}
+%% 
+%% Recent developments:
+%% - A low complexity minimum variance beamformer for ultrasound imaging using dominant mode rejection~\cite{Jiang2016}
+%% - Google deep beamforming
+%% 
+%% \begin{itemize}
+%% \item SDR - Entirely software defined radio, radar and sonar, bigger potential for adaptive methods
+%% \item Cognitive radar/sonar.
 1.
@@
 All of this is present in bats.
-% \end{itemize}
-% 
-% 
-% \section{Onlook}
-% 
-% Machine cognition:
-% \begin{itemize}
-% \item Ethical dilemmas. Decision and actions have consequences. Who's legally responsible for a machine's doing? Itself?
-% \item Safety. Machine flexibility allow more possible outcomes, and a wider range of security issues to consider.
-% \end{itemize}
-% 
-% R dim > 10000
-% - Inverting such matrices better to do with gradient decent
-% 
-% 
-% Deep learning. Simulator used to trained deep neural networks. Lots of data needed.
-% 
-% 
-% 
-% 
-% \section{}
+%% \end{itemize}
+%% 
+%% 
+%% \section{Onlook}
+%% 
+%% Machine cognition:
+%% \begin{itemize}
+%% \item Ethical dilemmas. Decision and actions have consequences. Who's legally responsible for a machine's doing? Itself?
+%% \item Safety. Machine flexibility allow more possible outcomes, and a wider range of security issues to consider.
+%% \end{itemize}
+%% 
+%% R dim > 10000
+%% - Inverting such matrices better to do with gradient decent
+%% 
+%% 
+%% Deep learning. Simulator used to trained deep neural networks. Lots of data needed.
+%% 
+%% 
+%% 
+%% 
+%% \section{}
 

### 2019-10-14 ThesisJo/2_background.tex
2019-10-12

@@
 This thesis explores three beamformers reconstructing images from acoustic echoes: time-domain MVDR (reference), frequency-domain MVDR (close approximation) and time-domain LCA (educated trial-and-error).
-Background on synthesis of image templates from the reconstructed images is presented in article IV.
+Article IV presents background on synthesis of image templates from reconstructed images.
 }
@@
 If $f$ represents the displacement of a vibrating string, the equation states that the force pulling the string towards equilibrium (right side) is proportional to the string's curvature (left side).
-In acoustics $f$ represents sound pressure, but since the mathematics are independent of the physics involved we will simply think of $f$ as a signal: a unit-less scalar function with wave characteristics.
+In acoustics $f$ represents sound pressure, but since the mathematics are independent of the physics involved we simply think of $f$ as a signal: a unit-less scalar function with wave characteristics.
 

### 2019-10-14 ThesisJo/4_discussion.tex
2019-10-12

@@
 
-Although receive-side adaptive beamforming yield slight image quality improvement in our active imaging systems, we believe its potential on the transmit-side to be greater.
-Ideally the two should work in synergy: a receiver telling a transceiver what data it needs.
-For example, a strong interfering source is best suppressed both on transmit and receive.
-In a dynamic world a low latency feedback loop is key to making this work.
+Although receive adaptive beamforming yields slight image quality improvement in our active imaging systems, we believe its transmit potential to be even greater.
+Ideally the two should work in synergy, receiver and transceiver cooperating in pursuit of common goals, for example to suppress dynamic interference.
+Low latency feedback loops is key to making this work.
 
-\emph{System on Chips} (Soc) are becoming increasingly popular as a heterogeneous computing platform.
-One of these chips, the Ultrascale+ Radio Frequency SoC (RFSoC) from Xilinx, contains a quad-core CPU, dual-core real-time processor, GPU, Field Programmable Gate Array (FPGA) and up to 16 Analog-to-Digital and Digital-to-Analog Converters (ADCs and DACs)---all in one CMOS chip.
-It targets the need for multiple-input-multiple-output (MIMO) can easily replace most analog front-end electronics in a sonar or ultrasound system with a software-defined solution, yielding better flexibility, reduced design complexity, improved bandwidth and lowered latency, all in a tiny physical footprint.
+Software defined radio 
+Applications 5G, Wifi,
+
+\emph{Phased arrays}
+
+\emph{System on Chips} (SoCs) are heterogeneous computing platforms crafted from a single silicon chip.
+One of these chips, the Ultrascale+ Radio Frequency SoC (RFSoC) from Xilinx, packs a quad-core CPU, dual-core real-time processor, GPU, Field Programmable Gate Array (FPGA) and up to 16 Analog-to-Digital and Digital-to-Analog Converters (ADCs and DACs)---all in one silicon chip.
+It facilitates multiple-input-multiple-output (MIMO) and other phased-array  can easily replace most analog front-end electronics in a sonar or ultrasound system with a software-defined solution, yielding better flexibility, reduced design complexity, improved bandwidth and lowered latency, all in a tiny physical footprint.
 

### 2019-10-14 ThesisJo/abstract.tex
2019-10-03
2019-10-12

@@
 
-This thesis explores methods to accelerate adaptive acoustic image reconstruction and synthesis, through beamforming and simulation, respectively.
+This thesis explores methods to accelerate adaptive acoustic image reconstruction and synthesis through beamforming and simulation, respectively.
 Beamforming is the main attraction: the art of delaying and weighting sensor arrays to craft desired spatial responses.
@@
 %
-\item \emph{Solution space} is continuous for MVDR and discrete for LCA, but similarly shaped. 
-%
-\item \emph{Model complexity} has a sweet spot, a balance point where its estimates offer an ideal compromise between bias and variance. MVDR depend on 
+% \item \emph{Adaptivity} is best for time-domain MVDR,
 %
@@
 %
-Finally, we perform image synthesis of common objects in the seabed with a GPU accelerated simulator.
+Finally, we perform image synthesis of common objects on the seafloor with a GPU accelerated simulator.
 These are in turn used to perform adaptive automatic target recognition (ATR).
-The result is a more dynamic and flexible system yielding improved image quality, image quality and classification performance.
+The result is a more dynamic and flexible system yielding improved classification performance.
 Finally, we describe this system with a new navigational notation system, extended with topological concepts from robotics and homogeneous coordinates.
-It helps keeping track of complex geometries, numerical representations and higher level abstractions such as manipulating rigid objects.
+It helps keep track of complex geometries, numerical representations and higher level abstractions such as manipulating rigid objects.
 

### 2019-10-19 ThesisJo/2_background.tex
2019-10-18

@@
 %&header
-% !TeX root = 1_introduction
+% !TeX root = thesis
+%% !TeX root = 1_introduction
 
@@
 % \\
-\begin{figure}[tp]\label{2_fig_article_relation}
+\begin{figure}[tp]
 \makebox[\linewidth][c]{%
@@
 Article IV presents background on synthesis of image templates from reconstructed images.
-}
+}%
+\label{2_fig_article_relation}
 \end{figure}
 
-\begin{figure}[!t]\label{2_fig_geometry}\centering
+\begin{figure}[!t]\centering
 \makebox[\linewidth][c]{%
@@
 An array with $M$ sensors has reference frame $\udot{A} \triangleq (\udot{A},\uvec{A})$ attached to it, where $\udot{A}\in\mathbb{A}$ is its absolute position and $\uvec{A}\triangleq\big(\hat{x}_{A},\hat{y}_{A},\hat{z}_{A}\big)\in\mathbb{V}^3$ its absolute orientation (\,$\hat{\cdot}$ denoting unit length).
-Impinging on the array at angle $\theta$ is a monochromatic plane wave: a single frequency wave from a source ${S}$ far away relative to the wavelength and array length (far field).
-Notation details are presented in Paper IV.
-}
+Impinging on the array at angle $\theta$ is a monochromatic plane wave: a single frequency wave from a source $\mathcal{S}$ far away relative to the wavelength and array length (far field).
+Notation details are presented in
+}%
+\label{2_fig_geometry}
 \end{figure}
 
-Any wave, regardless of its nature, is governed by a partial differential equation known as the \emph{wave equation}:
+Any wave, regardless of its nature, is governed by a partial differential equation known as the \emph{wave equation}.
+For a linear, homogeneous (translation-invariant) and isotropic (rotation-invariant) medium its form is
 %
-\begin{align}
+\begin{align}\label{eq_wave_equation}
 \nabla^2 f 
@@
 where $\nabla^2$ is the Laplace operator, $f$ is field strength and $c$ is wave propagation speed.
-If $f$ represents the displacement of a vibrating string, the equation states that the force pulling the string towards equilibrium (right side) is proportional to the string's curvature (left side).
-In acoustics $f$ represents sound pressure, but since the mathematics are independent of the physics involved we simply think of $f$ as a signal: a unit-less scalar function with wave characteristics.
+In complex mediums $c$ may depend on signal strength (non-linear), position (non-homogeneous/refractive), angle (anisotropic), frequency (dispersive) or be non-real (lossy).
+Deep sea is acoustically non-homogeneous due to varying pressure, temperature and salinity, but has negligible dispersion and attenuation.
+Biological tissue both non-homogeneous, non-linear and lossy.
+We approximate both cases with (\ref{eq_wave_equation}), because data inaccuracies are inevitable and good for testing algorithmic robustness.
+
+If $f$ represents displacement of a vibrating string, (\ref{eq_wave_equation}) states that the force pulling the string towards equilibrium (right side) is proportional to the string's curvature (left side).
+In acoustics $f$ represents sound pressure, with SI unit Pascal [Pa = N/m$^2$] or logarithmic unit sound-pressure-level (SPL) [1\,Pa = 94\,$\mathrm{dB}_{\textrm{SPL}}$].
+To generalize and simplify, we keep our equations free of physical units and simply think of $f$ as a signal: a unit-less scalar function with wave characteristics.
 
@@
 \begin{align}
-f_\textrm{plane}(\vec{p},t)  &= A e^{j(\omega t - \vec{k} \cdot \vec{p}_{OS})} \\
-f_\textrm{spheric}(\vec{p},t) &= \frac{A}{\lVert \vec{p}_{OS} \rVert} e^{j(\omega t - \lVert\vec{k}\rVert \lVert \vec{p}_{OS} \rVert)},
+f_\textrm{plane}(t,\vec{p}_{OS})  &= A e^{j(\omega t - \vec{k} \cdot \vec{p}_{OS})} \\
+f_\textrm{spheric}(t,\vec{p}_{OS}) &= \frac{A}{\lVert \vec{p}_{OS} \rVert} e^{j(\omega t - \lVert\vec{k}\rVert \lVert \vec{p}_{OS} \rVert)},
 \end{align}
@@
 Here $A$ is amplitude, $\omega$ is temporal wave frequency, $t$ is time, $\lVert\vec{k}\rVert=\frac{\omega}{c}=\frac{2\pi}{\lambda}$ is spatial wave frequency (wavenumber) and $\vec{p}_{OS}$ is position of the source $\mathcal{S}$ relative to the observer $\mathcal{O}$.
-We model the wavefield as superimposed plane waves under far-field conditions (when distance to the source is large relative to the wavelength and array length) and as superimposed spherical waves under near-field conditions.
+We model the wavefield as superimposed plane waves under far-field conditions (when distance to the source is large relative to wavelength and array length) and as superimposed spherical waves under near-field conditions.
 
@@
 
-\begin{figure}[tp]\label{2_fig_scenario_das}
+\begin{figure}[tp]
 \makebox[\linewidth][c]{%
 \graphicsAI[width=1\linewidth]{gfx/scenario_interference_das.svg}}
-\caption{DAS beampattern is static (conventional beamforming). Interference suppression at the first mainlobe is -13\,dBr.}
+\caption{DAS (delay-and-sum) beampattern is static (conventional beamforming). Interference suppression at the first mainlobe is -13\,dBr.}%
+\label{2_fig_scenario_das}
 \end{figure}
 
-\begin{figure}[tp]\label{2_fig_scenario_mvdr}
+\begin{figure}[tp]
 \makebox[\linewidth][c]{%
 \graphicsAI[width=1\linewidth]{gfx/scenario_interference_mvdr.svg}}
-\caption{MVDR beampattern is dynamic and continuous (adaptive beamforming). Interference suppression is good for prevalent noise sources if a good estimate of the spatial statistics can be obtained.}
+\caption{MVDR (minimum variance distortionless response) beampattern is dynamic and continuous (adaptive beamforming).
+Interference suppression is good for prevalent noise sources assuming sufficient spatial statistics.}%
+\label{2_fig_scenario_mvdr}
 \end{figure}
 
-\begin{figure}[tp]\label{2_fig_scenario_lca}
+\begin{figure}[tp]
 \makebox[\linewidth][c]{%
 \graphicsAI[width=1\linewidth]{gfx/scenario_interference_lca.svg}}
-\caption{LCA beampattern is dynamic and discrete (adaptive beamforming). Interference suppression is on-par with MVDR with 15 weight sets to choose from, nearly as good even with 6 weight sets.}
+\caption{LCA (low complexity adaptive) beampattern is dynamic and discrete (adaptive beamforming).
+Interference suppression is on-par with MVDR with 15 weight sets to choose from, marginally worse with 6 weight sets.}%
+\label{2_fig_scenario_lca}
 \end{figure}
 
-Imagine an array with $M$ sensors at positions $\udot{0},\udot{1},\dots, \udot{m}, \dots, \udot{M{\!-\!1}}$ with reference frame $\ubar{A} \triangleq (\udot{A},\uvec{A})$ attached to its phase center (Figure \ref{2_fig_geometry}), where $\udot{A}\in\mathbb{A}$ is its absolute position and $\uvec{A}\triangleq\big(\hat{x}_{A},\hat{y}_{A},\hat{z}_{A}\big)\in\mathbb{V}^3$ its absolute orientation ( $\hat{\cdot}$ denoting unit length).
-Assume that the array channels are
-digitally delayed to focus at a pixel with azimuth angle $\theta$ and range sample $n$, such that the delayed data from the $m$th channel can be expressed as $x_m[\theta,n-\Delta_m]\in\mathbb{C}$.
+Imagine an array $\mathcal{A}$ with $M$ sensors at positions $\udot{0},\udot{1},\dots, \udot{m}, \dots, \udot{M{\!-\!1}}$ with reference frame $\ubar{A} \triangleq (\udot{A},\uvec{A})$ attached to its phase center (Figure \ref{2_fig_geometry}), where $\udot{A}\in\mathbb{A}$ is its absolute position and $\uvec{A}\triangleq\big(\hat{x}_{A},\hat{y}_{A},\hat{z}_{A}\big)\in\mathbb{V}^3$ its absolute orientation ( $\hat{\cdot}$ denoting unit length).
+Assume the array channels are digitally delayed to focus at $\udot{S}$, such that the delayed data from the $m$th channel at instant $n$ can be expressed as $x_m[n-\Delta_m, \vec{p}_{mS}]\in\mathbb{C}$.
 Focus at source $\udot{S}$ requires delays akin to
@@
 \begin{align}
-\Delta_m[n] = \frac{\lVert \vec{p}_{m}\rVert - \lVert \vec{p}_{0S}\rVert}{c}.\label{2_eq_delta}
+\Delta_m[n] = \frac{\lVert \vec{p}_{mS}\rVert - \underset{m'}{\mathrm{min}}\ \lVert \vec{p}_{m'S}\rVert}{c},\label{2_eq_delta}
 \end{align}
 %
-where $\vec{p}_{m}$ denotes position vector $\udot{A}$ to $\udot{m}$ and $\vec{p}_{0S}$ denotes position vector $\udot{0}$ to $\udot{S}$, respectively.
-By making the dependence on $\theta$ implicit and assuming the data to be pre-delayed to the pixel of interest, we simplify the notation to $x_m[n]$.
+where both $\vec{p}_{mS}$ and $\vec{p}_{m'S}$ denote position vector $\udot{m}$ to $\udot{S}$ and subtracting the minimum distance is optional.
+By making the dependence on $\vec{p}$ implicit and assuming the data to be pre-delayed to the pixel of interest, we simplify the notation to $x_m[n]$.
 By definition, the beamformer output $z[n]\in\mathbb{C}$ is the weighted sum of delayed data samples:
@@
 %
-where $w_m\in\mathbb{C}$ is the weight factor assigned to channel $m$. Spatial response of the weights when the array is steered to broadside can be characterized by the beampattern,
+where $w_m\in\mathbb{C}$ is the weight factor assigned to channel $m$. Spatial responses of the weights when the array is steered to broadside can be characterized by the beampattern,
 %
@@
 A class of adaptive beamformers target this limitation by adapting the weights to the impinging wavefield.
-One such technique is the minimum variace distortionless response (MVDR) method.
+One such technique is the minimum variance distortionless response (MVDR) or Capon beamformer, named after its maker for use in spectral estimation~\cite{Capon1969}.
+
+% Assume:
+% Known array characteristics (sensor locations, element patterns, ...
+% Stationary signal and noise
+% Long observation times
+% Known direction of arrival
 
@@
 %
-The basic idea behind MVDR is to compute the set of weights that minimizes noise and interference accumulated by the array, while ensuring unit gain for signals coming from the pixel of interest.
-With pre-delayed data this can be formulated as follows:
+
+Time-domain MVDR often improves image contrast and resolution compared to conventional beamformers, as demonstrated in e.g.
+radar~\cite{Benitz1997}, ultrasound~\cite{Synnevag2007} and active sonar~\cite{Blomberg2013,Blomberg2012a,Dursun2009,Lo2004}.
+It achieves this by computing the weight set that minimizes noise and interference accumulated by the array, under the constraint of unit gain for signals from the pixel of interest.
+Assuming pre-delayed data, we can formulate this as
 %
 \begin{gather}
-\argmin{\dvec{w}}\ E\Big\{\big|\dvec{w}^H[n]\dvec{x}[n]\big|^2\Big\} \qquad\text{subject to}\qquad \dvec{w}^H[n]\dvec{1} = 1. \label{2_eq_mvdr_def}
+\argmin{\dvec{w}}\ E\Big\{\big|\dvec{w}^H[n]\dvec{x}[n]\big|^2\Big\} \qquad\text{subject to}\qquad \dvec{w}^H[n]\dvec{a} = 1, \label{2_eq_mvdr_def}
 % \vec w[n] = \frac{\Ri[n]\1}{\1^T\Ri[n]\1},\label{2_eq_weights}
@@
 %
-This optimization problem has the solution:
+where $\dvec{a}$ is steering vector. Applying Lagrange multipliers to this convex optimization problem yields the solution
 %
 \begin{gather}
-\vec w[n] = \frac{\Ri[n]\dvec{1}}{\dvec{1}^T\Ri[n]\dvec{1}},\label{2_eq_weights}
+\dvec w[n] = \frac{\Ri[n]\dvec{a}}{\dvec{a}^T\Ri[n]\dvec{a}} \qquad\text{where}\qquad \dvec{a}=\dvec{1}, \label{2_eq_weights}
 \end{gather}
 %
-where $\dvec1$ is a row vector of ones representing broadside steering and $\R=E\{\x[n]\x^H[n]\}$ is the spatial covariance matrix for the full array.
-We estimate $\R$ by computing a sample covariance matrix $\eR$.
-A good estimate requires some degree of \emph{spatial averaging} to avoid signal cancellation, \emph{temporal averaging} to maintain true speckle statistics, and \emph{diagonal loading} to improve robustness to parameter errors ~\cite{Synnevag2009a}.
-Combined, these techniques will also ensure a numerically well conditioned $\eR$.
+where $\dvec{1}$ is a row-vector of ones representing broadside phase-steering and $\R=E\{\dvec{x}[n]\dvec{x}^H[n]\}$ is the spatial covariance matrix for the full array.
+(\ref{2_eq_weights}) is the maximum likelihood estimate of the signal in the presence of zero-mean random Gaussian noise.
+We estimate $\R$ with a sample covariance matrix $\eR$, requiring some degree of
+%
+\begin{itemize}
+\item \emph{spatial averaging} to avoid signal cancellation by decorrelating coherent echoes~\cite{Kailath1985};
+\item \emph{temporal averaging} over an interval comparable to the pulse length (one to five samples) to maintain true speckle statistics~\cite{Synnevag2009a};
+\item \emph{diagonal loading} to improve robustness to parameter errors~\cite{Cox1987,Maksym1979,VanTrees2002}.
+\end{itemize}%
 
-To arrive at a formulation of how $\eR$ can be estimated, we first split the array into subarrays of length $L$ that overlap on all but one sensor.
-If we let $x_l[n]$ represent the datavector from subarray $l$,
+requiring some degree of \emph{spatial averaging} to avoid signal cancellation, \emph{temporal averaging} to maintain true speckle statistics, and \emph{diagonal loading} to improve robustness to parameter errors~\cite{Synnevag2009a}.
+
+Combined, these techniques also ensure a numerically well conditioned $\eR$.
+
+\begin{itemize}
+\item \emph{spatial averaging} to avoid signal cancellation by decorrelating coherent echoes~\cite{Kailath1985};
+\item \emph{temporal averaging} over an interval comparable to the pulse length (one to five samples) to maintain true speckle statistics~\cite{Synnevag2009a};
+\item \emph{diagonal loading} to improve robustness to parameter errors~\cite{Cox1987,Maksym1979}.
+\end{itemize}%
+
+We split the array into subarrays of length $L$ that overlap on all but one sensor. If we let $x_l[n]$ represent the datavector from subarray $l$,
 %
 \begin{gather}
-\dvec{x}_l[n] = \bmat{x_l[n] & x_{l+1}[n] & \dots & x_{l+L-1}[n]}^T.
+\dvec{x}_l[n] = \bmat{x_l[n] & x_{l+1}[n] & \dots & x_{l+L-1}[n]}^T,
 \end{gather}
 %
-then the sample covariance matrix formed with temporal and spatial averaging, $\eR$, is given as:
+then the sample covariance matrix formed with temporal and spatial averaging, $\eR$, is given as
 %

### 2019-10-19 ThesisJo/3_publications.tex
2019-10-18

@@
 This article accelerates time-domain MVDR with GPUs for use in active sonar imaging.
-Time-domain MVDR is the reference method for creating sonar images using the minimum variance optimization, aimed at reducing noise and interference while preserving signal integrity.
+Time-domain MVDR is the reference method for creating sonar images using minimum variance optimization, aimed at reducing noise and interference while preserving signal integrity.
 We refer to it as fully adaptive because it computes array weights analytically from spatial features in the data, yielding array weights from a continuous solution space that are optimal in the minimum variance sense.

### 2019-10-22 ThesisJo/2_background.tex
2019-10-22

@@
 \caption{Method relations.
-This thesis explores three beamformers reconstructing images from acoustic echoes: time-domain MVDR (reference), frequency-domain MVDR (close approximation) and time-domain LCA (educated trial-and-error).
+This thesis explores three beamformers reconstructing images from acoustic echoes: element-space MVDR (reference), beam-space MVDR (close approximation) and element-space LCA (educated trial-and-error).
 Article IV presents background on synthesis of image templates from reconstructed images.
@@
 Any wave, regardless of its nature, is governed by a partial differential equation known as the \emph{wave equation}.
-For a linear, homogeneous (translation-invariant) and isotropic (rotation-invariant) medium its form is
+Its form for a linear, homogeneous (translation-invariant) and isotropic (rotation-invariant) medium is
 %
@@
 After stripping the carrier frequency by heterodyning and the signal signature by matched filtering (Fig.
-\ref{2_fig_article_relation}), an image is formed by coherently combining the receiver outputs to focus at one angle (or point) at a time.
+\ref{2_fig_article_relation}), an image is formed by coherently combining the receiver outputs to focus at an angle or point at a time.
 The principle of adjusting the array's focus in range and direction is commonly referred to as beamforming, and involves assigning suitable delays and weights to the array's channels.
@@
 %
-where both $\vec{p}_{mS}$ and $\vec{p}_{m'S}$ denote position vector $\udot{m}$ to $\udot{S}$ and subtracting the minimum distance is optional.
+where both $\vec{p}_{mS}$ and $\vec{p}_{m'S}$ denote position vector $\udot{m}$ to $\udot{S}$.
+Subtracting the minimum distance reduces latency and is optional.
 By making the dependence on $\vec{p}$ implicit and assuming the data to be pre-delayed to the pixel of interest, we simplify the notation to $x_m[n]$.
-By definition, the beamformer output $z[n]\in\mathbb{C}$ is the weighted sum of delayed data samples:
+By definition, the beamformer output $z[n]\in\mathbb{C}$ is a weighted sum of delayed data samples:
 %
@@
 %
-where $w_m\in\mathbb{C}$ is the weight factor assigned to channel $m$. Spatial responses of the weights when the array is steered to broadside can be characterized by the beampattern,
+where $w_m\in\mathbb{C}$ is the weight factor assigned to channel $m$. Spatial responses of the weights for an array steered to broadside are characterized by the beampattern,
 %
@@
 
-\subsection{Time-domain MVDR}
+\subsection{Element-space MVDR}
 
@@
 
-Time-domain MVDR often improves image contrast and resolution compared to conventional beamformers, as demonstrated in e.g.
-radar~\cite{Benitz1997}, ultrasound~\cite{Synnevag2007} and active sonar~\cite{Blomberg2013,Blomberg2012a,Dursun2009,Lo2004}.
-It achieves this by computing the weight set that minimizes noise and interference accumulated by the array, under the constraint of unit gain for signals from the pixel of interest.
-Assuming pre-delayed data, we can formulate this as
+Element-space MVDR often improves image contrast and resolution compared to conventional beamformers~\cite{Benitz1997,Synnevag2007,Blomberg2013,Blomberg2012a,Dursun2009,Lo2004}, but is computationally complex and statistically fragile.
+It seeks the weight set that minimizes noise and interference accumulated by the array, while enforcing unit gain for signals from the pixel of interest,
 %
@@
 \begin{itemize}
-\item \emph{spatial averaging} to avoid signal cancellation by decorrelating coherent echoes~\cite{Kailath1985};
+\item \emph{spatial averaging} to avoid signal cancellation by decorrelating coherent echoes~\cite{Widrow1982,Kailath1985};
 \item \emph{temporal averaging} over an interval comparable to the pulse length (one to five samples) to maintain true speckle statistics~\cite{Synnevag2009a};
@@
 
-\subsection{Frequency-domain MVDR}
+\subsection{Beam-space MVDR}
 

### 2019-10-22 ThesisJo/3_publications.tex
2019-10-22

@@
 
-This article accelerates time-domain MVDR with GPUs for use in active sonar imaging.
-Time-domain MVDR is the reference method for creating sonar images using minimum variance optimization, aimed at reducing noise and interference while preserving signal integrity.
+This article accelerates element-space MVDR with GPUs for use in active sonar imaging.
+Element-space MVDR is the reference method for creating sonar images using minimum variance optimization, aimed at reducing noise and interference while preserving signal integrity.
 We refer to it as fully adaptive because it computes array weights analytically from spatial features in the data, yielding array weights from a continuous solution space that are optimal in the minimum variance sense.

### 2019-10-22 ThesisJo/4_discussion.tex
2019-10-22

@@
 
-\textbf{Paper I} - \textit{Time-domain MVDR beamforming on GPUs in active sonar imaging}:
+\textbf{Paper I} - \textit{Element-space MVDR beamforming on GPUs in active sonar imaging}:
 \begin{enumerate}
@@
 
-\textbf{Paper II} - \textit{Frequency-domain MVDR beamforming on GPUs in clinical ultrasound imaging}:
+\textbf{Paper II} - \textit{Beam-space MVDR beamforming on GPUs in clinical ultrasound imaging}:
 \begin{enumerate}
-\item Alters the implementation in Paper I to operate in beamspace (frequency domain), reduced to its three center beams (the most significant ones).
+\item Alters the implementation in Paper I to operate in beam-space (frequency domain), reduced to its three center beams (the most significant ones).
 \item Demonstrates speeds exceeding 1\,Mpixel/s on systems with 64 and 96 channels.
-\item Yields similar image quality to time-domain MVDR, in accordance with former studies in active sonar.
+\item Yields similar image quality to element-space MVDR, in accordance with former studies in active sonar.
 \end{enumerate}
@@
 
-GPU acceleration and algorithmic reductions largely mitigates the computational argument against MVDR in time-domain up to 32 channels and in time-domain for larger systems granted it is narrow-beam, narrow-band, or both.
+GPU acceleration and algorithmic reductions largely solves the issue with MVDR's computational complexity in element-space up to 32 channels, and in beam-space for larger systems that are narrow-beam, narrow-band, or both.
 LCA is entirely data-parallel thus ideal for GPUs, but runs faster unoptimized on CPUs than accelerated MVDR.%  Real-time performance is likely possible.

### 2019-10-22 ThesisJo/abstract.tex
2019-10-22

@@
 When aiming to minimize image noise and interference, two good candidates are the minimum variance distortionless response (MVDR) and low complexity adaptive (LCA) beamformer.
-We study time-domain MVDR (reference), frequency-domain MVDR (close approximation) and time-domain LCA (educated guessing), and conclude that:
+We study element-space MVDR (reference), beam-space MVDR (close approximation) and element-space LCA (educated guessing), and conclude that:
 %
@@
 %
-\item \emph{Acceleration} of MVDR through algorithmic reductions and implementation on graphics processing units (GPUs) yields real-time responses in time-domain up to 32 channels and in frequency-domain for any system size with narrow-band, narrow-beam, or both.
+\item \emph{Acceleration} of MVDR through algorithmic reductions and implementation on graphics processing units (GPUs) yields real-time responses in element-space up to 32 channels and in beam-space for any system size with narrow-band, narrow-beam, or both.
 LCA is inherently fast and extremely well suited for acceleration,

### 2019-10-23 LCAforActiveSonarImaging/submission/final/Buskenes_Jo_Inge-Low_Complexity_Adaptive_Sonar_Imaging.tex
2019-10-22

--- 2018-02-06 LCAforActiveSonarImaging/lca_submit2.tex
@@
 
-% \ifPeerReview
-% \let\OldIncludegraphics{\includegraphics}
-% \usepackage{letltxmacro}
-% \LetLtxMacro{\OldIncludegrsaphics}{\includegraphics}
-% \renewcommand{\includegraphics}[2][]{\OldIncludegraphics[width=\linewidth, #1]{#2}}
-% \fi
-
-% % \makeatletter
-% % \def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth
-% % \else\Gin@nat@width\fi}
-% % \makeatother
-% % \let\Oldincludegraphics\includegraphics
-% % \renewcommand{\includegraphics}[1]{\Oldincludegraphics[width=\maxwidth]{#1}}
-
 % \usepackage[maxfloats=40]{morefloats}
@@
 
-\title{Low Complexity Adaptive Sonar Imaging}
+\title{Low-Complexity Adaptive Sonar Imaging}
 
@@
 
-% While using weights that span a bigger space gives LCA greater freedom, not The bigger the space the weights span, the greater the adaptive freedom, but only the solutions that are valid in the constrained solution space can make a difference.
-
-% Instead of adaptively computing the weights, it adaptively selects them.
-
 Synnev\aa{}g \emph{et al}.
 suggested using a window set comprised of rectangular, Kaiser (or Kaiser-Harris) and inverted Kaiser functions.
-In total he made LCA choose from 12 unique windows, 6 of which were phase-shifted versions of a fairly wide Kaiser window, and one were rectangular.
-Phase-shifting the windows has the effect of slighly adjusting the angle the array is steered towards.
+In total he let LCA choose from 12 unique windows, 6 of which were phase-shifted versions of a fairly wide Kaiser window, and one were rectangular.
 Synnev\aa{}g's composition of windows was motivated by the desire to use wide windows with maximum sidelobe suppression near the receiver, and narrow windows at greater depths for maximum sensitivity and depth penetration.
-While achieving good results, it was not made clear how many windows LCA needs, or how they ideally should be steered.
+While he achieved good results, it was not made clear how many windows LCA needs, or how they ideally should be steered.
 Also, it is not apparent how LCA performs in sonar systems where the transmit beam is unfocused.
@@
 Results and discussion are provided in Section \ref{sec:results_discussion}, where we study MVDR's windows to find suitable LCA ones in \ref{sec:results_lca_window_function}, assess the range of window types and steering in \ref{sec:results_window_parameters}, determine the number of windows needed in \ref{sec:results_database_size}, and add a note on computational complexity in \ref{sec:results_complexity}.
-Section \ref{sec:conclusion} provides a conclusion.
-
-% This article is outlined as follows: In Section \ref{methods}, we introduce the concept of adaptive beamforming and provide details on the MVDR and LCA method.
-Then, in \ref{maptogpu}, we investigate the complexity issue, discuss means for reducing arithmetic complexity, and detail an implementation that makes efficient use of the GPU's parallel resources.
-The final design is assessed in Sections \ref{images_and_benchmarks} and \ref{discussion}, where we provide benchmarks, comparisons with similar CPU implementations, and measures of how efficiently our implementation makes use of the GPU's resources.
-
-% - How much oversampling
-% - How many parameters, how great a span.
-% 
-% - Experiementl data, Holmengraa/cross
-% 
-% - Note on efficiency.
-% 
-% 
-% From our work with MVDR it 
-% 
-% 
-% Traditionally beamformers were static, but as processors get more powerful so with the emergence of more processing power
-
-% 
-% \section{Methods}
-% 
-% At heart of the image formation process is a technique called beamforming, which creates a focus point at the pixel of interest.
-It does this by applying a suitable set of delays and weights to the array elements, such that signals emanating from the pixel location are emphasized constructively, while noise and interference from other directions sum destructively.
-% 
-% In active sonar array imaging, an acoustic wave is transmitted and the received echoes are usually recorded using an array of sensors.
-Each of these may be independently delayed and weighted such that signals emanating from directions of interest are added constructively, while noise and interference from other directions add destructively.
-Additionally, a window may be applied to the sensor channels to further adjust the array's spatial response.
-This process is known as beamforming.% The process of combining the signals from each sensor is known as beamforming.
-% 
-% 
-% 
-% 
-% % Traditional beamformers such as Delay-and-Sum (DAS) apply predefined windows to all incoming data.
-However, due to the non-stationary nature of sonar data, the optimal window for any given time instant will generally differ from the next.
-This is where adaptive beamformers thrive, because they compute the optimal window coefficients for the data at each time instant.
-The choice of optimisation criteria is what mainly differentiates the various adaptive beamformers.
-The Minimum Variance (MV) beamformer, for instance, selects the set of weights that minimizes the beamformer output power for any given time instant~\cite{cap69}.
-% 
-% 
-% \section{Background}
-% 
-% The MV beamformer computes the window coefficients by estimating and inverting a spatial covariance matrix.
-For improved estimation accuracy the data is averaged in space or time, or both.
-In addition, a certain amount of diagonal loading is often added to the covariance matrix before inversion.
-These actions, while leading to statistical robustness, tend to deteriorate the beamformer's performance.
-Also, the inversion of the covariance matrix has a computational complexity of O($M^3$), where $M$ is the number of elements.
-The computational burden makes the MV beamformer for large arrays less attractive, and sometimes even infeasible.
-% 
-% Based on the proposed method by Synnevåg~\cite{syn11}, we have implemented the LCA beamformer that keeps the minimum variance optimisation criterion but reduces the solution space to a discrete set of predefined windows.
-This reduces the computational complexity to O($MW$), where $W$ is the number of windows in the set.
-We use the Kaiser window function because it allows us to design a wide range of windows with different mainlobe widths and sidelobe suppression by adjusting the tradeoff parameter $\beta$~\cite{kai66}.
-In addition we apply steering to each of the windows and constrain the window design to ensure unit gain in the look direction.
-% 
-% The following will show that the proposed method performs similarly to or better than the MV method.
+Finally, Section \ref{sec:conclusion} provides a conclusion.
 
@@
 
-% \begin{figure*}[t]%\centering%
-% \includegraphics[width=\textwidth]{gfx/oversampling_mosaic_bounds_lca_windows_beta.pdf}%
-% \caption{$\beta$-values of the windows LCA selects for each image pixel.
-The underlying image is the one shown in \Fig{oversampling_mosaic_bounds}, with the same parameters.
-The location of the anchor and cut line is marked in red.
-Note how LCA prefers a rectangular window ($\beta=0$) on the anchor.}\label{oversampling_mosaic_bounds_beta}
-% \end{figure*} 
-% 
-% \begin{figure*}[t]%\centering%
-% \includegraphics[width=\textwidth]{gfx/oversampling_mosaic_bounds_lca_windows_phi.pdf}%
-% \caption{$\beta$-values of the windows LCA selects for each image pixel.
-The underlying image is the one shown in \Fig{oversampling_mosaic_bounds}, with the same parameters.
-The location of the anchor and cut line is marked in red.
-Note how LCA prefers a rectangular window ($\beta=0$) on the anchor.}\label{oversampling_mosaic_bounds_beta}
-% \end{figure*}
-
-% % \setcounter{dbltopnumber}{1
-
-% \begin{figure*}[t]%\centering%
-% \includegraphics[width=\textwidth]{gfx/beamformer_comparison_cross.pdf}%
-% \caption{}\label{beamformer_comparison_cross}
-% \end{figure*}
-% 
-% \begin{figure*}[t]%\centering%
-% \includegraphics[width=\textwidth]{gfx/beamformer_comparison_holmengraa.pdf}%
-% \caption{}\label{beamformer_comparison_holmengraa}
-% \end{figure*}
-
 \section{LCA}\label{sec:lca}
@@
 
-While we focus on using LCA with Kaiser windows, we will for completeness compare the performance with that of using trigonometric windows. This is defined as:
-%
-\begin{align}
-f_m(\alpha) = \alpha - (1-\alpha)\cos\Big(\frac{2\pi m}{M-1}\Big)\label{eq:trig_window_function}
-\end{align}
-%
-where we apply the contraint $\alpha\in[0.5,1]$ to avoid windows with negative coefficients.
+While we focus on using LCA with the Kaiser window function, we will for comparison also test it with the trigonometric window function. Its definition is:
+%
+\begin{align}
+f_m(\alpha) = \alpha - (1-\alpha)\cos\Big(\frac{2\pi m}{M-1}\Big),\label{eq:trig_window_function}
+\end{align}
+%
+where we apply the constraint $\alpha\in[0.5,1]$ to avoid windows with negative coefficients.
 Steering is applied as in (\ref{eq:steered_window}), and the $\alpha$-value that halves the \minus{}3\,dB distance is found using the method described in Appendix \ref{appendix:steering_angles}.
@@
 In \mediaI{}\footnote{For media location refer to Appendix \ref{app:source_code}.} we visualize the effect of changing the lateral oversampling factor on a set of LCA images.
-Upon display the images are all bilinearly upinterpolated to the same size.
-% 
-% \todo{Need $M$ in the image sector. Why???}
-
-% Idea to steer within a fraction of the rectangular window's mainlobe.
-
-% \begin{align}
-% d = \frac{\pi\,\theta_\text{open}\,\lambda}{M\,d\,O_s}
-% \end{align}
-
-% See section \ref{kaiser_windows}.
-
-% Narrow ones:
-% - Best coherent gain, lowest equivalent noise bandwidth, and most sensitive to steering meaning not needing to steer as much for some effect.
-% 
-% 
-% Beamforming:
-% - Maximize resolution
-% - Minimize noise/error, both coherent and incoherent
-% - Minimize signal cancellation
-% 
-% Gauss: Miminized time-bandwidth product
-%   - 
-%   
-% Rectangular
-%   - 
-% Dolph-Cheb: Minimize mainlobe width given sidelobe suppression
-% Kaiser: For a restricted energy and time duration, maximize energy in the band of frequencies W.
-% 
-% Adaptive windows perform best in detection of closely separated angles of significantly amplitudes.
-
-% Adaptive beamformers usually compute the window coefficients by estimating and inverting a spatial covariance matrix.
-There are two inherent problems to this procedure.
-First, for improved estimation accuracy the data is either averaged in space or time, and the significance of the covariance matrix can be reduced using a regularisation parameter~\cite{Carl Inge}.
-These are all attempts to constrain the beamformer to ensure the window coefficience are not over-adapted to the data.
-Second, inverting the covariance matrix has a computational complexity of O($M^3$)~\cite{Carl Inge}.
-For larger arrays the computational burden becomes significant, and the question arises whether this processing power can be better utilised.
-% 
-%
-
-% 
-% This selection is based on observations of the solutions found by the MV beamformer.
-We obtained good results in our experiments by letting the LCA beamformer select among 5 Kaiser windows with uniformly distributed $\beta$'s in the range $[0.05, 0.5]$ and a rectangular window.
-Each window was steered in 5 different directions, uniformly distributed within 80\% of the mainlobe width of the rectangular window, which is the most narrow.
-This adds to a total of 30 windows.
-% 
-% To make the choice of window less susceptible to pixel value uncertainties, the LCA beamformer was set to apply the most frequently selected window in an 11 pixel range region to the center pixel in that region.
-In physical terms, this is a region of approximately 20\,cm.
-The MV beamformer estimated the covariance matrix as described in~\cite{syn07} by averaging over 16 subarrays and 11 range pixels, and applying 3\% diagonal loading.
+Upon display the images were all bilinearly upinterpolated to the same size.
 
@@
 Each pixel depends on only a small subset of data, and the windows can be precomputed and stored in GPU cache for near immediate access.
-If we use a trigonometric window function we solve for the window parameter analytically as described in appendix \ref{sec:reducing_lca_complexity}.
-This reduces the complexity to O($M N_\phi$), and the only remaining LCA parameter is the steering amount.
-
-% Furthermore, the algorithm maps well to single-instruction-multiple-data (SIMD) hardware such as graphics processing units (GPUs).
-We show that even a modest amount of optimisation work has resulted in a factor 10 speed-up of this algorithm when implemented on a GPU as opposed to on a CPU.
-
-% the performance of the LCA beamformer is surprisingly similar regardless of what windows it is allowed to choose from.
-
-% \newcommand\cc[1]{\multicolumn{1}{>{\columncolor{tabBlue}}c}{\bf #1}}
-% \begin{table}[!b]\centering%\normalsize
-% \begin{tabular}[c]{l r r r@{}  r r r}\hline
-% \rowcolor{tabBlue} & \cc{Resolution} & \cc{S} & \cc{$\beta$} & \cc{$\varphi^*$} &  \cc{N$_\beta$} & \cc{N$_\varphi$} \\\hline
-% & 50\%  & 28 & [0,3] & \hspace{.15cm} 50\% & 2 & 3 \\
-% & 50\%  & 28 & [0,6] & \hspace{.15cm} 50\% & 2 & 3 \\
-% & 50\%  & 28 & [0,9] & \hspace{.15cm} 50\% & 2 & 3 \\
-% & 100\% & 32 &  &:30 \\
-% & 150\% & 32 & 1 &:4\\
-% \end{tabular}\\
-% * Steering is relative to the \minus{}3\,dB width of the window being applied.
-% \caption{Qualitative study}\label{parameter_study}
-% \end{table}%
-
-% The simulations contained a few highlighted cylindric regions with 2\,m diameter and a constant intensity 15.4\,dB over the average speckle level.
-We focused on an object centered at 41\,m range and -3 degrees azimuth angle.
-The images produced by the LCA, MV and unweighted DAS for a single speckle realisation and a mean image of all 100 realisations are shown in \Fig{speckle}.
-Each image was normalised by their respective speckle level, and the dynamic range was clamped at \{-30,~15\}\,dB.
-The physical extent of the of the simulated object was superimposed for reference.
-%\ref{saf09}
-% 
-% In \Fig{speckle} we observe that both the adaptive beamformers produce images with a more clearly defined shadow.
-The edges are sharper and the shadow more distinct.
-This is because the adaptive beamformers have better sidelobe suppression, and thus allow less signal energy to leak from the speckle region into the shadow region.
-The same effects can also be observed laterally around the highlight region of the mean image; DAS is more prone to smear energy into the speckle region than the adaptive beamformers are.
-% 
-% \Fig{cutmean} displays two lateral cuts in the mean image, one through the highlight at 41.1\,m range and one through the shadow at 44.5\,m range.
-Each cut was computed as the mean of a 40\,cm range band centered at the respective ranges.
-The transition region between highlight and shadow is shortest for the LCA, which is advantageous since it results in a more accurate representation of the object size.
-The slightly inferior performance of the MV beamformer is due to the subarray averaging, which makes the effective array smaller~\cite{syn07}.
-% 
-% By studying the LCA performance on the 100 speckle realisations, we found the choice of window to be highly dependent on the phenomena being imaged.
-In speckle regions LCA favored narrow spatial responses with minor steering, while wide and fully steered responses were often selected in shadow regions.
-There was also a clear distinction between the responses selected in sidelobe regions and in speckle regions.
-% 
-% % \Fig{hist} illustrates how often a particular window is chosen in general.
-There are 10 $\phi$'s for each of the 10 $\beta$'s.
-The first 10 values are rectangular windows.
-Value 5 is a non-steered rectangular window, and we see that this is most frequently selected.
-At higher steering angles the windows are rarely selected, and thus we might get better results by tightening our steering boundaries.
-At value 15 we find inverted Kaiser windows with $\beta = 0.05$, and then $\beta$ is increased in increments of 0.505 for every 10th value.
-At value 65 the $\beta$ value is 2.025, and these and the remaining windows are hardly selected.
-% 
-% A sidescan image of the 1500 dwt oil tanker wreck Holmengraa is shown in \Fig{holmengraa}.
-It is about 68\,m long and 9\,m wide, and lies on a slanted seabed at 77\,m depth~\cite{holmengraa}.
-The sidescan image was created using data from the HISAS 1030 sonar, which is rather unsuited for this purpose because of its large opening angle.
-This, and the fact that the wreck was imaged at a range of about 105\,m makes the image quality poor, but sufficient to compare our beamformers.
-% 
-% In the Holmengraa image we note again that the LCA produces a cleaner shadow and better edge definition. The MV method performed almost identically to LCA in this case, and was omitted.
+Furthermore, if we use the trigonometric window function we can solve for the window parameter analytically as described in Appendix \ref{sec:reducing_lca_complexity}.
+This reduces the complexity to O($M N_\phi$), because the only remaining LCA parameter to create windows for will be the steering angle.
 
 \section{Conclusion}\label{sec:conclusion}
-
-% \begin{table}[!b]\centering%\normalsize
-% \begin{tabular}[c]{l c c c}\hline
-% \rowcolor{tabBlue} & \bf Highlight & \bf Shadow & \bf Speckle  \\\hline
-% Symmetry             & Moderate  & High   & High \\
-% Mainlobe width       & Narrow    & Wide   & Moderate \\
-% Sidelobe suppression & Low       & High   & Moderate \\
-% Steering             & Low       & Up to \minus{}3\,dB width & Moderate \\
-% Oversampling         & 2x-8x     & 2x-8x  & 2x-8x
-% \end{tabular}
-% \caption{Summary}\label{tab:summary}
-% \end{table}%
 
@@
 
-In summary, LCA is an attractive alternative to other adaptive beamformers due to being very fast, simple to understand, practically parameter-free, inherently robust, and able produce images similar to that of MVDR.
+In summary, LCA is an attractive alternative to other adaptive beamformers due to being very fast, simple to understand, practically parameter-free, inherently robust, and able to produce images similar to that of MVDR.
 Contrary to MVDR, it has a low computational complexity and can be easily and efficiently accelerated using e.g.
 GPUs.
-Since LCA performs just as well if we use trigonometric windows instead of Kaiser ones, we can also compute the optimal window parameter analytically for any given steering angle, and avoid the search over the window parameter entirely.
-
-% The tendency of LCA to produce images of similar quality as MVDR seems to indicate that a full solution space is rarely needed in real world scenarios.
-
-%  It initially very is also very fast, and is very well suited for acceleration using  very fast, easy to implement even on parallel hardware,
-
-% Its computational complexity is only of O($N_wM$), and its data parallel nature makes it very well suited for acceleration using e.g. GPUs. compared to MVDR's O($M^3$).
-% 
-% LCA is largely insensitive to the choice of windows and steering. 
-% LCA works well for all the boundaries
-% Variables of interest:
-
-% We limit our study to the use of variations of the parametric Kaiser-Harris window, since we observed neglible performance difference to using this function or any of the common trigonometric ones.
-The Kaiser windows are easy to compute, is fairly insensitive to coefficient inaccuracies, and can create windows with shapes spanning from rectangular to Gaussian.
-% 
-% 
-% Using the LCA beamformer results in increased robustness and improved edge definition compared to the MV and DAS beamformers.
-
-% The images produced by LCA appears sharper than those of MVDR, but 
-% 
-% The key to achieve this success lies in the design of the window set.
-Windows that yield vagely steered narrow responses are preferred in speckle regions, while wide and steered responses are typically preferred in highlight and shadow regions.
-% 
-% A window set of 6 different responses each steered in 5 different directions proved sufficient in our experiment to match the performance of the MV beamformer.
-
-% \begin{itemize}
-% \item Adaptive beamformer's potential lies in its ability to suppress interference power
-% \item Why adaptive beamformers struggle in active sonar systems.
-Correlated noise, robustification kills the adaptive potential.
-Quite computationally intensive.
-Constraints must be applied in one way or another - parameters must be tuned.
-% \begin{itemize}
-% \item Noise and signal is correlated. Spatial averaging required. 
-% \item Increases variance of speckle (not only in active systems?), spatial compunting \cite{Vignon2009} (US) or time averaging?.
-Subarray averaging applied here \cite{Synnevag2007} (US) \todo{Is it bad to mix ultrasound/ sonar refs?}
-% \end{itemize}
-% \item Outline: 
-% \begin{itemize}
-% \item Setup the window set for LCA to match this.
-% \item Notice that the LCA performs like Capon, even with a small window database.
-% \item Start with Kaiser, see that other window functions also work well.
-% \end{itemize}
-% \end{itemize}
-% \todopar{Out of the blue idea: Weigh CF with LCA? This guy did it with MVDR \cite{Wang2009}. Probably you know already... Okay, back on track.}
-
-% \IEEEPARstart{T}{o} form images from a modern phased array sonar system the received wavefield is usually recorded, and then postprocessed by a digital beamformer.
-The beamformer applies delays and weights to the sensor channels, the beamformer adjusts the arrays spatial response to focus at one pixel at a time.
-such that signals emanating from regions of interest add constructively, while ensuring that noise and interference from other angles do not.
-% 
-% The imaging capabilities of a modern phased array sonar system depend on physical attributes such element response and array geometry, the transmitted signal, as well as the beamforming method being used on transmission and reception.
-Beamforming is the concept of applying delays and weights to the sensors channels to steer the arrays response to points of interest.
-
-% 
-% 
-% Outline:
-% \begin{itemize}
-% \item Choice of window functions makes little difference.
-% \item Steering and mainlobewidths have outer bounds.
-% \item Beamspace?
-% \item Chosen window plots - what may they tell us? Variance intensity values when using various windows.
-% \item Assymmetric windows?
-% \end{itemize}
-%
-
-% , and maximum suppression of while ensuring that the beamformer digitally  before each of the pixels are estimated one at a time.
-The resolution and contrast of such a system will depend on the systems spatial response, which ideally should be narrow  be very sharp in the desired direction its ability to achieve  fundamental principle of forming a sonar image is to record the received wavefield,
+Also, since LCA performs just as well if we use trigonometric windows instead of Kaiser ones, we can compute the optimal window parameter analytically for any given steering angle and avoid the search over the window parameter entirely.
 

### 2019-10-23 LCAforActiveSonarImaging/submission/submit2/answer/answer.tex
2019-10-22

@@
 Manuscript number: & 2015JOE001805 \\
-Title: & Low Complexity Adaptive Sonar Imaging \\
+Title: & Low-Complexity Adaptive Sonar Imaging \\
 Authors: & Buskenes,~J.I., Hansen,~R.E., Austeng,~A.

### 2019-10-23 ThesisJo/2_background.tex
2019-10-22

@@
 After stripping the carrier frequency by heterodyning and the signal signature by matched filtering (Fig.
-\ref{2_fig_article_relation}), an image is formed by coherently combining the receiver outputs to focus at an angle or point at a time.
+\ref{2_fig_article_relation}), an image is formed by coherently combining the receiver outputs to focus at one position (or angle) at a time.
 The principle of adjusting the array's focus in range and direction is commonly referred to as beamforming, and involves assigning suitable delays and weights to the array's channels.
@@
 \caption{MVDR (minimum variance distortionless response) beampattern is dynamic and continuous (adaptive beamforming).
-Interference suppression is good for prevalent noise sources assuming sufficient spatial statistics.}%
+Interference suppression is good for prevalent noise sources, assuming sufficient spatial statistics.}%
 \label{2_fig_scenario_mvdr}
@@
 \begin{gather}
-\dvec w[n] = \frac{\Ri[n]\dvec{a}}{\dvec{a}^T\Ri[n]\dvec{a}} \qquad\text{where}\qquad \dvec{a}=\dvec{1}, \label{2_eq_weights}
+\dvec w[n] = \frac{\Ri[n]\dvec{a}}{\dvec{a}^T\Ri[n]\dvec{a}}, \label{2_eq_weights}
 \end{gather}
 %
-where $\dvec{1}$ is a row-vector of ones representing broadside phase-steering and $\R=E\{\dvec{x}[n]\dvec{x}^H[n]\}$ is the spatial covariance matrix for the full array.
-(\ref{2_eq_weights}) is the maximum likelihood estimate of the signal in the presence of zero-mean random Gaussian noise.
+where $\dvec{a}=\dvec{1}$ is a row-vector of ones representing broadside phase-steering and $\R=E\{\dvec{x}[n]\dvec{x}^H[n]\}$ is the spatial covariance matrix for the full array.
+These weights yield the maximum likelihood estimate of the signal in the presence of zero-mean random Gaussian noise---but not in active systems typically characterized by highly correlated noise, interference and signal~\cite{Widrow1982}.
+For regular array geometries such as ours, this can be handled with spatial smoothing.
+
 We estimate $\R$ with a sample covariance matrix $\eR$, requiring some degree of
@@
 \begin{itemize}
-\item \emph{spatial averaging} to avoid signal cancellation by decorrelating coherent echoes~\cite{Widrow1982,Kailath1985};
-\item \emph{temporal averaging} over an interval comparable to the pulse length (one to five samples) to maintain true speckle statistics~\cite{Synnevag2009a};
+\item \emph{spatial smoothing} to avoid signal cancellation by decorrelating coherent echoes~\cite{Kailath1985};
+\item \emph{temporal smoothing} over an interval comparable to the pulse length (one to five samples) to maintain true speckle statistics~\cite{Synnevag2009a};
 \item \emph{diagonal loading} to improve robustness to parameter errors~\cite{Cox1987,Maksym1979,VanTrees2002}.
 \end{itemize}%
+%
+These techniques ensure a numerically well conditioned $\eR$~\cite{VanTrees2002,Kailath1985}, allow reducing the computational complexity and expose sufficient data parallelism for GPU-acceleration.
 
-requiring some degree of \emph{spatial averaging} to avoid signal cancellation, \emph{temporal averaging} to maintain true speckle statistics, and \emph{diagonal loading} to improve robustness to parameter errors~\cite{Synnevag2009a}.
-
-Combined, these techniques also ensure a numerically well conditioned $\eR$.
-
-\begin{itemize}
-\item \emph{spatial averaging} to avoid signal cancellation by decorrelating coherent echoes~\cite{Kailath1985};
-\item \emph{temporal averaging} over an interval comparable to the pulse length (one to five samples) to maintain true speckle statistics~\cite{Synnevag2009a};
-\item \emph{diagonal loading} to improve robustness to parameter errors~\cite{Cox1987,Maksym1979}.
-\end{itemize}%
-
-We split the array into subarrays of length $L$ that overlap on all but one sensor. If we let $x_l[n]$ represent the datavector from subarray $l$,
+Formulation of the sample covariance matrix $\eR$ begins by splitting the array into subarrays of length $L$ that overlap on all but one sensor.
+If we let $x_l[n]$ represent the datavector from subarray $l$,
 %
@@
 %
-then the sample covariance matrix formed with temporal and spatial averaging, $\eR$, is given as
+then $\eR$ formed with temporal and spatial averaging (smoothing) is given as
 %
@@
 where $N_K = 2K+1$ is the number of temporal samples to perform averaging over, and $N_L = M-L+1$ is the number of subarrays.
-In the actual design a fraction of the total power of $\eR[n]$ is finally added to its diagonal~\cite{Synnevag2007}, but this step will be ignored here due to its negligible contribution to runtime.
+Finally, a fraction of the total power of $\eR[n]$ is added to its diagonal~\cite{Synnevag2007}.
 
-To summarize as in Figure \ref{mvdr_beamforming}: For every pixel in the image we estimate $\eR$ (\ref{2_eq_spatialR}), solve the linear equation $\eRi\1$ (\ref{2_eq_weights}), the finally substitute some to get the beamformer output (\ref{2_eq_mvdr_def}-\ref{2_eq_weights}).
+To summarize: For every pixel in the image we estimate $\eR$ (\ref{2_eq_spatialR}), solve the linear equation $\eRi\1$ (\ref{2_eq_weights}), the finally substitute some to get the beamformer output (\ref{2_eq_mvdr_def}-\ref{2_eq_weights}).
+Complexity is cubic with the number of sensors, $O(M^3)$, as opposed to linear for conventional beamformers, $O(M)$.
+
+sample spatial spectral matrices
+asymptotic behaviour
+Forward-backward averaging
+Toeplitz-structured covariance - assuming stationarity
+Structured spectral matrix estimation
+parametric spectral matrix estimation
+singular value decomposition
+fixed diagonal loading
 

### 2019-10-31 ASA2013/asa2013.tex
2019-08-18
2019-10-30

@@
 a4paper,10pt
-]{ica2013_2}
+]{common/ica2013_2}
 
@@
 
-\newcommand\1{\vec 1}
 \newcommand\I{\mat I}
@@
 \usepackage{amsmath,amssymb}
-\usepackage{glossaries}
-\usepackage{common/slashbox}
+% \usepackage{glossaries}
 
@@
 Since $\R$ is unknown, we estimate it by by computing a sample covariance matrix $\eR$.
-In this computation we perform some degree of \emph{spatial averaging} to avoid signal cancellation, \emph{temporal averaging} to maintain true speckle statistics, and \emph{diagonal loading} to improve robustness to parameter errors ~\cite{Synnevag2009a}.
+In this computation we perform some degree of \emph{spatial averaging} to avoid signal cancellation, \emph{temporal averaging} to maintain true speckle statistics, and \emph{diagonal loading} to improve robustness to parameter errors ~\cite{Synnevag2009}.
 Combined, these techniques will also ensure a numerically well conditioned $\eR$.
@@
 % \bibliography{../../Library/library}
-\bibliography{library}
+% \bibliography{library}
 

### 2019-10-31 ECUA2012/gpu_capon_ecua2012.tex
2019-10-30

@@
 We will do this by computing the sample covariance matrix, $\eR$.
-In this computation we will perform some degree of \emph{spatial averaging} to avoid signal cancellation, \emph{temporal averaging} to achieve \gls{DAS}-like speckle statistics, and \emph{diagonal loading} to ensure a numerically well conditioned $\eR$~\cite{Synnevag2009a}.
+In this computation we will perform some degree of \emph{spatial averaging} to avoid signal cancellation, \emph{temporal averaging} to achieve \gls{DAS}-like speckle statistics, and \emph{diagonal loading} to ensure a numerically well conditioned $\eR$~\cite{Synnevag2009}.
 

### 2019-10-31 HRUIofMovingObjects/jrnl_article.tex
2019-10-30

@@
 Dotted vertical lines indicate position of transmit-receive beam pairs.
-The point scatterer response has been simulated using Field II \cite{Jensen1992, Jensen1996a} using a 96 element phased array with a 22 mm azimuth aperture, which was excited with a 2.5 MHz and 1.5 cycles long pulse.
+The point scatterer response has been simulated using Field II \cite{Jensen1992, Jensen1996} using a 96 element phased array with a 22 mm azimuth aperture, which was excited with a 2.5 MHz and 1.5 cycles long pulse.
 Transmit focus in azimuth and elevation were configured to be equal and the point scatterer was moved through this point.

### 2019-10-31 HRUIofMovingObjects/jrnl_article_1st.tex
2019-10-30

@@
 Dotted vertical lines indicate position of transmit-receive beam pairs.
-The point scatterer response has been simulated using Field II \cite{Jensen1992, Jensen1996a} with a probe setup equal to the M5S probe by GE Vingmed Ultrasound, Horten, Norway.
+The point scatterer response has been simulated using Field II \cite{Jensen1992, Jensen1996} with a probe setup equal to the M5S probe by GE Vingmed Ultrasound, Horten, Norway.
 The probe was excited with a 2.5 MHz and 1.5 cycles long pulse.

### 2019-10-31 HRUIonGPU/1st submit/jrnl_article.tex
Adding HRUIonGPU article
2019-10-30

--- 2013-04-03 HRUIonGPU/jrnl_article.tex
@@
 has been applied in much of the recent literature on Capon beamforming for medical ultrasound imaging \cite{Synnevag2007, Nilsen2009, Wang2009, Mehdizadeh2012}.
-It has also been applied to HF and VHF antenna processing \cite{Featherstone1997b}.
+It has also been applied to HF and VHF antenna processing \cite{Featherstone1997}.
 
@@
 
-In passive systems, assuming stationary statistics, the sample covariance matrix in (\ref{eq:R}) is typically averaged over a long sequence of samples\cite{Krima}.
+In passive systems, assuming stationary statistics, the sample covariance matrix in (\ref{eq:R}) is typically averaged over a long sequence of samples\cite{Krim1996}.
 Since medical ultrasound imaging is an active, short-pulsed system, there is a limited amount of data from which $\mat{\hat{R}}$ can be estimated.
@@
 
-To investigate the image quality of both the ES and BS-Capon beamformer, a phantom was simulated using Field II \cite{Jensen1992, Jensen1996a} (Fig.
+To investigate the image quality of both the ES and BS-Capon beamformer, a phantom was simulated using Field II \cite{Jensen1992, Jensen1996} (Fig.
 \ref{fig:phantom}).

### 2019-10-31 HRUIonGPU/document.tex
2019-06-26
2019-09-01
2019-10-30

@@
 % denne må utelates i IUS-artikkelen
-In recent work we have applied a GPU-based Capon beamformer for both active sonar imaging \cite{Buskenes, Buskenes2013} and cardiac ultrasound imaging \cite{Asen2012}.
+In recent work we have applied a GPU-based Capon beamformer for both active sonar imaging \cite{Buskenes2015, Buskenes2013} and cardiac ultrasound imaging \cite{Asen2012}.
 Sonar has a less strict real-time requirement than medical ultrasound imaging, which means that most array configurations can be handled in real time using the GPU.
@@
 
-In passive systems, assuming stationary statistics, the sample covariance matrix in (\ref{II_eq:R}) is typically averaged over a long sequence of samples\cite{Krima}.
+In passive systems, assuming stationary statistics, the sample covariance matrix in (\ref{II_eq:R}) is typically averaged over a long sequence of samples\cite{Krim1996}.
 Since medical ultrasound imaging is an active, short-pulsed system, there is a limited amount of data from which $\mat{\hat{R}}$ can be estimated.
@@
 
-In this section we will give a brief summary of our parallel implementation of the ES-Capon beamformer as presented in \cite{Asen2012} and \cite{Buskenes}.
+In this section we will give a brief summary of our parallel implementation of the ES-Capon beamformer as presented in \cite{Asen2012} and \cite{Buskenes2015}.
 This implementation was our starting point when the BS-Capon beamformer was implemented.
@@
 
-In \cite{Asen2012, Buskenes} we demonstrated how the cubic complexity with $L$ %and linear complexity with $N_K$ 
+In \cite{Asen2012, Buskenes2015} we demonstrated how the cubic complexity with $L$ %and linear complexity with $N_K$ 
 can be reduced by realizing that calculations in (\ref{II_eq:R}) overlap across subarrays.
@@
 
-%To investigate the image quality of both the ES and BS-Capon beamformer, a phantom was simulated using Field II \cite{Jensen1992, Jensen1996a} (Fig.
+%To investigate the image quality of both the ES and BS-Capon beamformer, a phantom was simulated using Field II \cite{Jensen1992, Jensen1996} (Fig.
 \ref{II_fig:phantom}).
@@
 The two point targets were placed inside speckle, with 0 dB in amplitude, covering an area from $\pm90\degree$ in azimuth.
-The phantom was simulated using Field II \cite{Jensen1992, Jensen1996a}.
+The phantom was simulated using Field II \cite{Jensen1992, Jensen1996}.
 The simulation was performed using a 64-element, $p=\lambda/2$ pitch, 10 $\mu$m kerf, 80\% bandwidth, and 10 mm high phased array.

### 2019-10-31 HRUIonGPU/jrnl_article.tex
2019-10-30

@@
 
-%To investigate the image quality of both the ES and BS-Capon beamformer, a phantom was simulated using Field II \cite{Jensen1992, Jensen1996a} (Fig.
+%To investigate the image quality of both the ES and BS-Capon beamformer, a phantom was simulated using Field II \cite{Jensen1992, Jensen1996} (Fig.
 \ref{fig:phantom}).

### 2019-10-31 HowtoHowtoCapon/document.tex
2019-06-26
2019-08-18
2019-09-01
2019-10-30

@@
 \IEEEcompsoctitleabstractindextext{%
+
 \begin{abstract}%
@@
 \item \emph{spatial averaging} to avoid signal cancellation by decorrelating coherent echoes~\cite{Kailath1985};
-\item \emph{temporal averaging} over an interval comparable to the pulse length (one to five samples) to maintain true speckle statistics~\cite{Synnevag2009a};
+\item \emph{temporal averaging} over an interval comparable to the pulse length (one to five samples) to maintain true speckle statistics~\cite{Synnevag2009};
 \item \emph{diagonal loading} to improve robustness to parameter errors~\cite{Cox1987,Maksym1979}.
@@
 %
-% will perform some degree of \emph{spatial averaging} to avoid signal cancellation caused by coherent echoes~\cite{Kailath1985}, \emph{temporal averaging} to maintain true speckle statistics~\cite{Synnevag2009a}, and \emph{diagonal loading} to improve robustness to parameter errors ~\cite{Cox1987,Maksym1979}.
+% will perform some degree of \emph{spatial averaging} to avoid signal cancellation caused by coherent echoes~\cite{Kailath1985}, \emph{temporal averaging} to maintain true speckle statistics~\cite{Synnevag2009}, and \emph{diagonal loading} to improve robustness to parameter errors ~\cite{Cox1987,Maksym1979}.
 %
@@
 \clearpage
+\renewcommand{\theHchapter}{A\arabic{chapter}}
+\renewcommand{\theHsection}{A\arabic{section}}
 \appendix
 % \section*{Appendix}
-\renewcommand\thesection{\Roman{section}}
+\renewcommand\thesection{\Alph{section}}
+\renewcommand\thesubsection{\Alph{section}.\arabic{subsection}}
 \else
@@
 
-\bibitem{Synnevag2009a}
+\bibitem{Synnevag2009}
 \BIBentryALTinterwordspacing

### 2019-10-31 HowtoHowtoCapon/submission/final/article.tex
2019-10-30

@@
 \item \emph{spatial averaging} to avoid signal cancellation by decorrelating coherent echoes~\cite{Kailath1985},
-\item \emph{temporal averaging} over an interval comparable to the pulse length (1-5 samples) to maintain true speckle statistics~\cite{Synnevag2009a}, and
+\item \emph{temporal averaging} over an interval comparable to the pulse length (1-5 samples) to maintain true speckle statistics~\cite{Synnevag2009}, and
 \item \emph{diagonal loading} to improve robustness to parameter errors~\cite{Cox1987,Maksym1979}.

### 2019-10-31 HowtoHowtoCapon/submission/submit1/howto_howto_capon.tex
2019-10-30

%%%%%%%%%%%%%%%%                                ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
% CONDITIONALS %
%%%%%%%%%%%%%%%%                                ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

\newif\ifPeerReview\PeerReviewfalse             % Whether to create the PeerReview version or
                                                % Journal version
\newif\ifFlatArchive\FlatArchivefalse           % Whether archive is flat (messy) or contain 
                                                % subfolders for graphics etc.
\newif\ifFloatAtEnd\FloatAtEndfalse             % Available in PeerReview mode:
                                                % Place floats at end of document?
\newif\ifTODO\TODOtrue                        % Use todo notes?

%%%%%%%%%%%%                                    ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
% IEEEtran %
%%%%%%%%%%%%                                    ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

\ifPeerReview
\documentclass[12pt,journal,draftclsnofoot,onecolumn]{IEEEtran}
% \newcommand\CLASSINPUTbaselinestretch{1.66}     % http://theoval.cmp.uea.ac.uk/~nlct/latex/thesis/node17.html
\else
\documentclass[journal]{IEEEtran}
\fi

% \RequirePackage[latin1]{inputenc}%              % Set input encoding (optionally latin1)
% \RequirePackage[T1]{fontenc}%                 % Set font encoding
% \usepackage[norsk]{babel}

% Single column review mode:
%\documentclass[12pt,journal,onecolumn]{IEEEtran}
%\newcommand\CLASSINPUTbaselinestretch{2}
%\RequirePackage{calc}
%\RequirePackage{fp} 
% Left margin = 1 inch + hoffset + oddsidemargin (or evensidemargin)
% Adding the 2cm gutter width to the odd/even side margins:
%\setlength\hoffset{0pt}
%\setlength\oddsidemargin{4cm}       % 4cm margins on the left side
%\addtolength\oddsidemargin{-1in}    % Subtract the initial 1 inch
%\setlength\textwidth{21cm-8cm}    % A4 width (21cm) minus margins on either side

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%                  ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
% IEEE ''APPROVED'' PACKAGES %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%                  ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

\usepackage{cite}

\ifCLASSINFOpdf
   \usepackage[dvips]{graphicx}                 % Might not work. Use 'latex' instead of 
   \ifFlatArchive\else                          % 'pdflatex'
      \graphicspath{./gfx/}
   \fi
\else
   \usepackage[dvips]{graphicx}
   \ifFlatArchive\else
      \graphicspath{./gfx/}
   \fi
\fi

\RequirePackage[table,dvipsnames,svgnames]{xcolor}

\usepackage[cmex10]{amsmath}                    % cmex10 option to be IEEE explore compliant
\interdisplaylinepenalty=2500                   % Allows multiline equations to be broken

% \RequirePackage{amssymb}

\RequirePackage{array}

\ifCLASSOPTIONcompsoc
   \usepackage[caption=false,font=normalsize,labelfont=sf,textfont=sf]{subfig}
\else
   \usepackage[caption=false,font=footnotesize]{subfig}
\fi
\ifCLASSOPTIONcaptionsoff                       % IEEE promoted hack to turn off captions from the 
   \let\MYorigsubfloat\subfloat                 % subfloat package should the captionsoff option
   \renewcommand{\subfloat}[2][\relax]{\MYorigsubfloat[]{#2}} % be specified.
\fi

\ifFloatAtEnd
\ifCLASSOPTIONcaptionsoff                       % Places float at the end of the document when the
  \usepackage[nomarkers]{endfloat}              % captionsoff options is specified to IEEEtrans.cls
  \let\MYoriglatexcaption\caption               % (PeerReview mode)
  \renewcommand{\caption}[2][\relax]{\MYoriglatexcaption[#2]{#2}}
\fi
\fi

\usepackage{fixltx2e}                           % Fix some twocolumn float problems

%\usepackage{stfloats}                          % Allows: \begin{figure*}[!b]
                                                % (double column figures on top/bottom)

\usepackage{url}                                % Support for handling and breaking URLs

% NOTE: PDF hyperlink and bookmark features are not required in IEEE
%       papers and their use requires extra complexity and work.
\newcommand\MYhyperrefoptions{bookmarks=true,bookmarksnumbered=true,
pdfpagemode={UseOutlines},plainpages=false,pdfpagelabels=true,
colorlinks=true,linkcolor={black},citecolor={black},urlcolor={black},
pdftitle={An Optimized GPU Implementation of the MVDR Beamformer for Active Sonar Imaging},
pdfsubject={},
pdfauthor={Jo Inge Buskenes},
pdfkeywords={adaptive beamforming, beamforming, complexity, sonar, active}}%
\ifCLASSINFOpdf
   \usepackage[\MYhyperrefoptions,pdftex]{hyperref}
\else
   \usepackage[\MYhyperrefoptions,breaklinks=true,dvips]{hyperref}
   \usepackage{breakurl}                        % Allows 'dvips' driver to break links
\fi

%%%%%%%%%%%%%%%%%%%%%%%                         ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
% ADDITIONAL PACKAGES %       
%%%%%%%%%%%%%%%%%%%%%%%                         ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

% \ifPeerReview
% \let\OldIncludegraphics{\includegraphics}
% \usepackage{letltxmacro}
% \LetLtxMacro{\OldIncludegrsaphics}{\includegraphics}
% \renewcommand{\includegraphics}[2][]{\OldIncludegraphics[width=\linewidth, #1]{#2}}
% \fi

% % \makeatletter
% % \def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth
% % \else\Gin@nat@width\fi}
% % \makeatother
% % \let\Oldincludegraphics\includegraphics
% % \renewcommand{\includegraphics}[1]{\Oldincludegraphics[width=\maxwidth]{#1}}

\usepackage[maxfloats=40]{morefloats}
\newcounter{todoidx}
% \setcounter{todoidx}

\ifTODO
   \definecolor{todobackground}{rgb}{0.95,0.95,0.95}
   \setlength\marginparsep{1pt}
   \setlength\marginparwidth{35pt}
   \newlength\marginparwidthsmall
   \setlength\marginparwidthsmall{\marginparwidth}
   \addtolength\marginparwidthsmall{-7pt}
   \newcommand\todo[1]{%
      \addtocounter{todoidx}{1}%
      {\color{Red}\bf(\thetodoidx{})}%%\fbox{\bf\thetodoidx{}}}%
      \marginpar{%
         {\vspace*{-10pt}\color{Red}\fbox{\bf\thetodoidx{}}}\\%
         \fcolorbox{red}{todobackground}{\parbox{\marginparwidthsmall}{\scriptsize #1}}}}

   \newcommand\todopar[1]{\fcolorbox{red}{white}{\parbox{0.97\linewidth}{#1}}}
\else
%    \usepackage[disable]{./todonotes} 
   \newcommand\todo[1]{}
\fi

\newenvironment{narrow}[2]{%
\begin{list}{}{%
\setlength{\topsep}{0pt}%
\setlength{\leftmargin}{#1}%
\setlength{\rightmargin}{#2}%
\setlength{\listparindent}{\parindent}%
\setlength{\itemindent}{\parindent}%
\setlength{\parsep}{\parskip}}%
\item[]}{\end{list}}

\usepackage{float}

\definecolor{tabBlue}{HTML}{AACCFF}

%%%%%%%%%%                                      ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
% MACROS %       
%%%%%%%%%%                                      ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

% \DeclareMathOperator*{\argmin}{\text{arg}\;\text{min}}

\newcommand\Fig[1]{Fig.~\ref{#1}}

\newcommand\Grey[1]{{\color{Grey}#1}}
\newcommand\Red[1]{{\color{Red}#1}}
\newcommand\Blue[1]{{\color{Blue}#1}}
\newcommand\DarkBlue[1]{{\color{DarkBlue}#1}}
\newcommand\LightBlue[1]{{\color{LightBlue}#1}}
\newcommand\Brown[1]{{\color{Brown}#1}}
\newcommand\Green[1]{{\color{Green}#1}}
\newcommand\SeaGreen[1]{{\color{SeaGreen}#1}}
\newcommand\Yellow[1]{{\color{yellow}#1}}
\newcommand\Orange[1]{{\color{orange}#1}}

\newcommand\nn{\nonumber\\}

\newcommand\nmat[1]{\begin{matrix}#1\end{matrix}}
\newcommand\bmat[1]{\begin{bmatrix}#1\end{bmatrix}}
\newcommand\case[1]{\begin{cases}#1\end{cases}}
\newcommand\textbox[2]{\footnotesize\text{\parbox{#1}{\centering\emph{#2}}}}

\newcommand\rand{\text{rand}}
\newcommand\randn{\text{randn}}
\newcommand\rect{\text{rect}}
\newcommand\sinc{\text{sinc}}
\newcommand\tr{\text{tr}}
\newcommand\adj{\text{adj}}

% \newcommand\max{\text{max}}
\newcommand\argmin[1]{\text{arg}\;\underset{#1}{\text{min}}}

\newcommand\qqquad{\quad\qquad}
\newcommand\qqqquad{\qquad\qquad}

% \renewcommand\l[1]{\left#1}
% \renewcommand\r[1]{\right#1}

% {\text{\parbox{1.5cm}{\centering volume hyper- sphere}}}

%Keyword colouring:
\newcommand\kw[1]{#1}
\newcommand\parm[1]{#1}%\color{Black}#1\color{Black}}

\newcommand\of[1]{\scriptstyle(\parm{#1})\displaystyle}
\newcommand\df[1]{\scriptstyle[\parm{#1}]\displaystyle}
\newcommand\var[3]{#1_\text{#2}\of{#3}}

\newcommand\diag{\text{diag}}

% \raisebox{lift}[extend-above-baseline][extend-below-baseline]{text}
\newcommand\mt[1]{\text{\emph{#1}}} %mt = mathtext
\newcommand\mathnorm{\textstyle}
\newcommand\mathbig[1]{\displaystyle#1\mathnorm}
\newcommand\mathsmall[1]{\scriptstyle#1\mathnorm}
\newcommand\mathtiny[1]{\scriptscriptstyle#1\mathnorm}
\newcommand\sfrac[2]{\scriptstyle\raisebox{0.25pt}[0pt][0pt]{$\frac{#1}{#2}$}\mathnorm}
\newcommand\nfrac[2]{\textstyle\frac{#1}{#2}\displaystyle}

\newcommand\sumu[1]{\sum\limits^{#1}\;}
\newcommand\suml[1]{\sum\limits_{#1}\;}
\newcommand\sumb[2]{\sum\limits_{#1}^{#2}\;}

\newcommand\produ[1]{\prod\limits^{#1}\;}
\newcommand\prodl[1]{\prod\limits_{#1}\;}
\newcommand\prodb[2]{\prod\limits_{#1}^{#2}\;}

\newcommand\defeq{\overset{\underset{\mathrm{def}}{}}{=}}

%Math macros:
\newcommand\diff[2]{\frac{\kw{d}\;\textstyle #1\scriptstyle}{\kw{d\parm{#2}}}\displaystyle}
\newcommand\ddiff[2]{\frac{\kw{d^2}\;\displaystyle #1\scriptstyle}{\kw{d\parm{#2}}^2}\displaystyle}

\renewcommand\d[1]{\scriptstyle\kw{\;d\parm{#1}}\displaystyle}

% These commands are mutually exclusive. Remember to "renew" in v2.
\newcommand\intb[4]{\int\limits_{#3}^{#4} #1 \d{#2}} % \int{exp}{var}{from}{to}
\newcommand\intl[3]{\int\limits_{#3} #1 \d{#2}} % \int{exp}{var}{for all}
\newcommand\intu[2]{\int #1 \d{#2}} % \int{exp}{var}{for all}

\newcommand\T{^{\scriptscriptstyle T}}
\renewcommand\H{^{\scriptscriptstyle H}}

\renewcommand\vec[1]{\boldsymbol{#1}}
\newcommand\mat[1]{\boldsymbol{#1}}

\newcommand\Om{O_\text{m}}
\newcommand\Oa{O_\text{a}}
\newcommand\Nl{N_\text{l}}
\newcommand\Nk{N_\text{k}}
\newcommand\1{\vec 1}
\newcommand\I{\mat I}
\renewcommand*\a{\vec a}
\renewcommand*\i{\vec i}
\renewcommand*\k{\vec k}
\newcommand*\n{\vec n}
\newcommand*\p{\vec p}
\newcommand*\s{\vec s}
\newcommand*\w{\vec w}
\newcommand*\x{\vec x}
\newcommand*\y{\vec y}

\newcommand*\A{\mat A}
\newcommand*\B{\mat B}
\newcommand*\C{\mat C}
\newcommand*\E{\mat E}
% \renewcommand*\H{\mat H}
\renewcommand*\P{\mat P}
\newcommand*\eP{\mat{\hat P}}
\newcommand*\R{\mat R}
\newcommand*\Ri{\R^{-1}}
\newcommand*\eR{\mat{\hat R}}
\newcommand*\eRi{\hat{\mat R}\;\!^{-1}}
\newcommand*\Navg{N_\text{avg}}
\newcommand*\W{\mat W}
\newcommand*\X{\mat X}
\newcommand*\Xd{\X_{\!\Delta}}
\newcommand*\Y{\mat Y}

\renewcommand*\L{\mat \Lambda}
\newcommand*\U{\mat U}
% \renewcommand*\t{\mathtiny{^T}}
% \newcommand*\h{\mathtiny{^H}}
\renewcommand*\t{^T}
\newcommand*\h{^H}

\newcommand\D{\vec\nabla} %Del: Vector differential operator - nabla
\newcommand\Dx{\vec\nabla\times}
\newcommand\Dd{\vec\nabla\cdot}

\usepackage{tikz}
\usetikzlibrary{shapes,snakes}
\usepackage{amsmath,amssymb}
% \usepackage{datatool}
% \usepackage{glossaries}

\newenvironment{outline}
{\begin{itemize}}
{\end{itemize}}

%    \definecolor{todobackground}{rgb}{0.95,0.95,0.95}
%    \setlength\marginparsep{3pt}
%    \setlength\marginparwidth{42pt}
%    \newlength\marginparwidthsmall
%    \setlength\marginparwidthsmall{\marginparwidth}
%    \addtolength\marginparwidthsmall{-7pt}
%    \newcommand\todo[1]{%
%       \addtocounter{todoidx}{1}%
%       {\color{Red}\fbox{\bf\thetodoidx{}}}%
%       \marginpar{%
%          {\vspace*{-10pt}\color{Red}\fbox{\bf\thetodoidx{}}}\\%
%          \fcolorbox{red}{todobackground}{\parbox{\marginparwidthsmall}{#1}}}}
%

% correct bad hyphenation here
% \hyphenation{op-tical net-works semi-conduc-tor}

%%%%%%%%%%%%                                    ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
% GLOSSARY %
%%%%%%%%%%%%                                    ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

% \makeglossaries
% \input{../../Library/glossary}

%%%%%%%%%%%%%%%%%%                              ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
% DOCUMENT START %
%%%%%%%%%%%%%%%%%%                              ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

% \usepackage{yfonts}

\begin{document}

\title{An Optimized GPU Implementation of the MVDR Beamformer for Active Sonar Imaging}

\author{Jo~Inge~Buskenes, %
        Jon~Petter~\AA{}sen, %
        Carl-Inge~Colombo~Nilsen, %
        Andreas~Austeng%
\thanks{J. I. Buskenes, C.-I. C. Nilsen and A. Austeng are with the Department of Informatics, University of Oslo, Norway.}%
\thanks{J. P. \AA{}sen is with MI-Lab, Norwegian University of Science and Technology, Trondheim, Norway.}% <-this % stops a space
% \thanks{Manuscript received April 19, 2005; revised January 11, 2007.}
}

% The paper headers
\markboth{IEEE Journal of Oceanic Engineering}%
{A Robust and Optimized GPU Implementation of the MVDR beamformer}

% Publishers ID mark:
%\IEEEpubid{0000--0000/00\$00.00~\copyright~2007 IEEE}

% use for special paper notices
%\IEEEspecialpapernotice{(Invited Paper)}

% for Computer Society papers, we must declare the abstract and index terms
% PRIOR to the title within the \IEEEcompsoctitleabstractindextext IEEEtran
% command as these need to go into the title area created by \maketitle.
\IEEEcompsoctitleabstractindextext{%
\begin{abstract}
% Two limiting factors of modern phased array sonar imaging systems are the angular resolution and contrast. Adaptive beamformers target these limitatitons by weighting sensors dynamically to fit the impinging wavefield.

The Minimum Variance Distortionless Response (MVDR) beamformer has recently been proposed as an attractive alternative to conventional beamformers in active sonar imaging. Unfortunately, it is very computationally complex because a spatial covariance matrix must be estimated and inverted for each image pixel. This may discourage its use unnecessarily in sonar systems which are continuously being pushed to ever higher imaging ranges and resolutions.

In this study we show that for active sonar systems up to 32 channels, the computation time can be significantly reduced by performing arithmetic optimizations, and by implementing the MVDR on a Graphics Processing Unit (GPU). We point out important hardware limitations for these devices, and assess the design in terms of how efficiently it is able to use the GPU's resources. On a quad-core Intel Xeon system with a high-end Nvidia GPU, our GPU implementation renders more than 1\;Mpx/s. This represents a speedup of more than two orders of magnitude over our initial single threaded implementation in C. This enables real-time processing of sonar data, and makes the MVDR a viable alternative to conventional methods in practical systems.
% If the HISAS1030 was attached to a platform traveling at 1.8m/s, creating full-coverage sectorscan images in real-time would be possible with this performance.

% We benchmarked optimized implementations of the MV beamformer in Matlab and C, running on a quad-core Intel Xeon E5620, and in CUDA running on a high-end nVidia card. The beamformers operated on data from the 32x4 element Kongsberg Martime HISAS1030 sonar, using a conservative set of robustification parameters. 
%The CPU would then be free to take on the rest of the beamforming as well.
\end{abstract}

% Keywords (normally not used for peer reviews)
\ifPeerReview\else
\begin{IEEEkeywords}
Adaptive beamforming, MVDR, Capon, sonar, active sonar, GPU, graphics processing unit, complexity.
\end{IEEEkeywords}
\fi}
% \fi

% make the title area
\maketitle

% This command fixes abstract positioning for compsoc articles:
\IEEEdisplaynotcompsoctitleabstractindextext

% (Optional) Add some extra info on cover page of peer review papers:
% \ifCLASSOPTIONpeerreview
% \begin{center} \bfseries EDICS Category: 3-BBND \end{center}
% \fi

% Insert page break and insert second title (peer review mode)
\IEEEpeerreviewmaketitle

\maketitle

\section{Introduction}

% Images from a phased array active sonar system are formed in the post-processing by focusing on one pixel at a time. This is achieved by delaying and weighting each of the array's channels, providing coarse and fine adjustments to the region of focus, respectively. This is commonly referred to as beamforming. Assuming far-field conditions, the delays are usually chosen to steer the pixel of interest to broadside, and the weights chosen to reject the noise impinging on the array from non-broadside directions.
% 
% The algorithm responsible for carrying out this task is referred to as a beamformer. Depending on whether data is evaluated when choosing a suitable set of delays and weights, beamformers can be branded either conventional or adaptive. A good example of the former is the The Delay-and-Sum (DAS) beamformer, which - as the name implies - simply delay the sensor channels appropriately prior to summing them up. Applying various weight sets, or windows, to the data prior to summation allows the lateral \gls{SNR} to be improved, but at the inevitable expense of a deteriorated lateral resolution. \todo{aka. the curse of physics.}
% 
% The Minimum Variance Distortionless Response (MVDR) (or \gls{MVDR}) beamformer, have the potential to overcome this limitation. While ensuring unit gain in the look direction, it computes the set of weights that minimizes the energy accumulated by the array from other directions. In other words, whenever the distribution of noise and interference is primarily focused around specific angles, an array response with a high degree of suppression at these angles are chosen, resulting in an image with a lateral \gls{SNR} and resolution superior to anything the \gls{DAS} can come up with.

% 
% In active sonar imaging the constrast and resolution can be improved by the use of adaptive beamforming. 
% 
% Adaptive beamforming 
% A critical part in the sonar image formation is the 
% The image quality is mainly limited by beampattern of the imaging arrays.  resolution and noise suppression
% 
% The quality of sonar images 
% 
% An essential part of a sonar imaging system is the beamformer. 
% 
% 
% A sonar image is typically formed by using receive beamforming to focus on one pixel at a time. The beamformer's ability to suppress interference   to focus on one pixel at the time. How well a pixel value is estimated, and thus how 
% 
% A sonar image is built one pixel at the time the 
% 
% The image quality of a sonar system will depend on the imaging arrays ability to focus in 
% 
% 
% 
% Forming images done by combining data from multiple receivers in a way that to obtain focus on one pixel at the time, and assigning the output value to the intensity of that pixel. The 
% 
% Most modern active sonar systems employ phased arrays to 
% The basic way of a sonar image was formed 
% 
% A sonar system's ability to 
% 
% Adaptive beamforming has 
% 
% The image quality 
% 
% To improve the image quality of 
% 
% In high resolution sonar imaging the 
% 
% The image 
% 
% 
% In a typical sonar imaging scenario an encoded signal is transmitted to highlight a surface of interest, and the reflected wavefield sampled using a receiver array. A image is then formed in the post-processing by coherently combining the receiver outputs to focus at one pixel at the time. The principle of adjusting the arrays focus is commonly referred to as beamforming, and involves assigning suitable delays and weights to the array's channels.
%

% A sonar image is typically formed by using receive beamforming to focus on one pixel at a time.   The image resolution and contrast The image quality The quality of active sonar images depend - amoung other things - on the beamformer's ability to suppress interference  resolution and constrast
% 
% A sonar image is typically formed by using receive beamforming to focus on one pixel at a time. The quality of the final image will depend on the beamformer's ability to suppress interference and noise from off-pixel.    to focus on one pixel at the time. How well a pixel value is estimated, and thus how
% 
% 
% In conventional active sonar imaging systems 
% 
% Various methods have been applied to active sonar imaging. One of these methods, 
% 
%  eamforming techniques have been applied in several fields to improve image. methods can be applied to 
% 
% Most active sonar imaging systems are based on conventional beamforming methods. These are robust and simple, but does not take the incoming wavefield into account. offer  in the image formation process. While being inherently robust, these methods 
% 
% 
% With the ever increasing processing power found in modern computers 
% 
% As the processing power of todays modern computer systems keeps improving the feasibility of 
% 
% As the processing power of modern computer systems keep improving, the use of data adaptive methods in various imaging fields.

\IEEEPARstart{D}{ata} driven methods have been introduced in various imaging fields over the years in attempts to improve image quality. One such method is the Minimum Variance Distortionless Response (MVDR) beamformer. When compared to conventional methods, MVDR is often able to improve image contrast and resolution, as shown in e.g. radar~\cite{Benitz1997}, ultrasound~\cite{Synnevag2007}, and active sonar~\cite{Blomberg2012a,Blomberg2011,Dursun2009,Lo2004}.

Despite its inherent potential, the MVDR beamformer has yet to see widespread adoption in the active sonar field. There may be several reasons for this. For one, the method is not inherently robust, and may suffer from a phenomenon called signal cancellation in active systems~\cite{Widrow1982}. Another reason is that in its original form, the computational complexity is cubic with the number of channels, O($M^3$), while conventional beamformers are at O($M$). This is because a spatial covariance matrix is estimated and inverted for each image pixel.

% Robustification:
% R estimate accuracy vs. dynamic trackability: (1) Trade observation time for bandwidth (2) Partial adaptiveness (less degrees of freedom than channels) (3) Quadratic constraint on sensor weights to limit the norm to some preset value.
% -  
To ensure MVDR robustness, the litterature suggests combining means such as temporal and spatial averaging, and regularization~\cite{VanTrees2002,Kailath1985}. The complexity issue, on the other hand, can be handled by introducing well-founded approximations. For instance, some studies assume data stationarity which allows the formation of a Toeplitz-structured covariance matrix that is simpler to invert~\cite{Asl2012a,Jakobsson2000}. Other studies perform MVDR beamforming in beamspace (the spatial frequency domain), which can be considered sparse due to the limited angular extent of the received beam~\cite{VanTrees2002,Nilsen2009}. Exploiting this can lead to considerable performance improvements, especially in high channel count systems.

% Some low complexity approximations to the MVDR beamformer is also available, such as the Low Complexity Adaptive (LCA) beamformer,

% Such devices can often this by combining up to thousands of computing cores that each performs modestly, but combined performs impressively.

% We will show that by exploiting symmetry and avoiding unnecessary computations, the spatial averaging will also aid in reducing the overall arithmetic complexity\todo{make sure this is shown clearly somewhere}. for Unlike other work, which focus modifying or transforming\todo{get low channel count in, as it makes a case against the other solutions.} We make no approximations, but rather stay with a textbook example of the method that handle complex basebanded data\todo{misplaced?}.

% Other work (keywords: mvdr, capon, computation, efficient, complexity):
% - Partially adaptive beamforming. DAS on non-overlapping subarrays, then MVDR \cite{Swingler1999a}.
% - Beamspace / frequency based methods.
% - Low complexity. PCA + LC. 
% - Systolic arrays. Radar \cite{Ward1984} 
% - Toeplitz. \cite{Yapanel} Spectrum estimation - speech recognition. 
% 
% Us:
% - Aim for the GPU. Simple is often better.
% - Therefore: Optimize the standard MVDR chain.
% 1. Reduce arithmetic instructions
% 2. Implement for full utilisation of the arithmetic throughput on the GPU.
% 
% (Made partially adaptive by employing subarray averaging to reduce degrees of freedom to L.)

In this work, we show that such approximations may not be necessary for sonar systems up to 32 channels, since here the computation time can be reduced by 1-2 orders of magnitude by implementing MVDR on a Graphics Processing Unit (GPU). Similar work may be found in e.g. medical ultrasound imaging, where Chen \emph{et al.} have demonstrated a GPU implementation of the MVDR that operates on real valued data~\cite{Chen2011,Chen2011a}. While this implementation is fast, the real valued data constraint does not allow the creation of asymmetric and shifted responses, and hence prevents the MVDR's potential to be fully reached. Our implementation is not restricted in this manner. In related work, we investigated using the same MVDR GPU implementation in cardiac ultrasound imaging~\cite{Asen2012,Asen2013}, and in active sonar imaging we have compared the performance of this implementation to comparable adaptive methods~\cite{Buskenes2013a}.

This article is outlined as follows: In section \ref{methods} we introduce the concept of adaptive beamforming and provide details on the MVDR method. Then in \ref{maptogpu} we investigate the complexity issue, discuss means for reducing arithmetic complexity, and detail an implementation that makes efficient use of the GPU's parallel resources. The final design is assessed in \ref{images_and_benchmarks}-\ref{discussion}, where we provide benchmarks, comparisons with similar CPU implementations, and measures of how efficiently our implementation makes use of the GPU's resources.

%  Furthermore, by implementing it on a modern \gls{GPU} we have gained a speedup of around 2 orders of magnitude compared to an optimized single-thread implementation in C.

% These devices have in recent years emerged as a flexible and highly potent parallel technology well suited for many DSP tasks, and are available off-the-shelf at reasonable prices\todo{skip this?}.

% These devices have in recent years emerged as a flexible and highly potent parallel technology well suited for many DSP tasks, and are available off-the-shelf at reasonable prices.

% Yay or nay? $\Rightarrow$ GPU ultrasound (\cite{So2011,Chen2011}). Complex data, all robustification techniques.

% - Adaptive beamformers offers image quality edge, but are usually computationally complex.
% - Who's tried before?
% - We: Active sonar MVDR, GPU, high performance.

\section{Methods}\label{methods}

Consider a sonar imaging scenario where an encoded signal is transmitted to highlight a surface of interest, and assume that the backscattered wavefield is properly sampled using a uniform linear array of receivers. An image can then be formed by coherently combining the receiver outputs to focus at one angle at the time. The principle of adjusting the array's focus in range and direction is commonly referred to as beamforming, and involves assigning suitable delays and weights to the array's channels.

\subsection{Beamforming}

\ifPeerReview
\begin{figure}[H]\centering
\includegraphics[width=0.8\linewidth]{gfx/buske1.eps}
\else
\begin{figure}[!t]\centering
\includegraphics[width=\linewidth]{gfx/beamforming.eps}
\fi%
\caption{Beamforming principle. Signal signature is first removed by matched filtering. Then - prior to summation - a suitable set of delays, $\Delta$, and weights, $w$, are applied to focus on a pixel of interest at angle and range ($\theta,n$).}\label{beamforming}
\end{figure}
Let the receiver be an $M$ element uniform linear array, and assume that the signature of the transmitted signal has been removed by a matched filter (Fig. \ref{beamforming}). Further assume that the array channels are digitally delayed to focus at a pixel with azimuth angle $\theta$ and range sample $n$, such that the delayed data from the $m$th channel can be expressed as $x_m[\theta,n]$. To simplify notation, we make the dependence on $\theta$ implicit from now on.

By definition the beamformer output $z[n]$ can now be expressed as the weighted sum of all the delayed data samples:
\begin{align}
z[n] = \w\H[n]\x[n] = \bmat{w_0[n]\\w_1[n]\\\vdots\\w_{M-1}[n]}^H \bmat{x_0[n]\\x_1[n]\\\vdots\\x_{M-1}[n]},\label{z}
\end{align}
where $w_m$ is the weight factor assigned to channel $m$. With static weights this would be referred to as the conventional delay-and-sum (DAS) beamformer. A large variety of weighting functions exists here for trading lateral resolution for improved noise suppression (contrast), but one always ends up with a compromise between the two~\cite{Harris1978}.

Various adaptive beamformers target this limitation by allowing the weights to change for each pixel to better fit the dynamic nature of the incoming wavefield. In other words, they attempt to use the \emph{a priori} information present in the data to improve image quality. The MVDR beamformer is one such method. It finds the set of complex weights that minimizes the beamformer's expected output power, while ensuring unity gain in the look direction~\cite{Capon1969}. This is a convex optimization problem that can be solved using Lagrange multipliers to yield the solution:
\begin{gather}
\vec w[n] = \frac{\Ri[n]\1}{\1\T\Ri[n]\1},\label{weights}
\end{gather}
where $\1$ is a row vector of ones that represents broadside steering, and $\R=E\{\x[n]\x\H[n]\} \in\mathbb{C}^{M,M}$ is the spatial covariance matrix for the full array. Since $\R$ is unknown, we estimate it by computing a sample covariance matrix $\eR$. In this computation we will perform some degree of \emph{spatial averaging} to avoid signal cancellation, \emph{temporal averaging} to maintain true speckle statistics~\cite{Synnevag2009}, and \emph{diagonal loading} to improve robustness to parameter errors ~\cite{Cox1987,Maksym1979}. Combined, these steps will also ensure a numerically well conditioned $\eR$.

To form $\eR$ we will first compute an intermediate sample covariance matrix $\breve{\R}$, which we form using temporal and spatial averaging. To do this we need to segment our array into subarrays. If we let $x_l[n]$ represent the data vector from subarray $l$,
\begin{gather}
\x_l[n] = \bmat{x_l[n] & x_{l+1}[n] & \dots & x_{l+L-1}[n]}\T,
\end{gather}
then $\breve{\R}$ can be calculated as:
\begin{gather}
\breve{\R}[n] =  \frac{1}{N_K N_L} \sumb{l=0}{M-L}\sumb{n'=n-K}{n+K} \x_l[n]\x_l\H[n] \in\mathbb{C}^{L,L},\label{spatialR}
\end{gather}
where $N_K = 2K+1$ is the number of temporal samples to perform averaging over, and $N_L = M-L+1$ is the number of subarrays.

The final estimate $\eR$ is found by adding a fraction $d$ of the total power of $\breve{\R}[n]$ to its diagonal~\cite{Synnevag2007}:
\begin{align}
\eR[n] = \breve{\R}[n] + \I \frac{d}{L} \tr\{\breve{\R}[n]\},\label{finalR}
\end{align}
where $\I$ is an identity matrix, $\tr\{\cdot\}$ represents the matrix trace operation, and $\tr\{\breve{\R}[n]\}$ is an estimate of the energy received from this pixel.

Note how subarray averaging led to a size reduction of $\eR$ from $\mathbb{C}^{M,M}$ to $\mathbb{C}^{L,L}$, and hence will produce an $L$-element weight set when substituted into (\ref{z}). This weight set is applied to all the subarrays, prior to computing the beamformer output as in (\ref{z}). Or, equivalently, we may apply the weight set to the sum of all the subarrays:
\begin{align}
z[n] = \w\H[n] \sumb{l=0}{M-L} \x_l[n].\label{finalZ}
\end{align}
\ifPeerReview
\begin{figure}[!t]\centering
\includegraphics[width=.8\linewidth]{gfx/buske2.eps}
\else
\begin{figure}[!t]\centering
\includegraphics[width=\linewidth]{gfx/implementation.eps}
\fi%
\caption{MVDR beamforming. For each pixel in range and azimuth,\newline
1. an $L\times{}L$ sample covariance matrix $\eR$ is computed, \newline
2. the term $\eR^{-1}\1$ is found using a linear equation solver,\newline
3. and the beamformer output $z$ is computed from (\ref{finalZ}), where $\w$ is found by substituting $\eR^{-1}\1$ into (\ref{weights}). } \label{mvdr_beamforming}
\end{figure}
% \begin{figure}[!t]\centering
% \includegraphics[width=\linewidth]{gfx/algorithm_structure.eps}
% \caption{MVDR beamforming. First a spatial covariance matrix is estimated from the delayed data (\ref{spatialR}-\ref{finalR}), then the weights are computed (\ref{weights}) and finally applied to the delayed channel data (\ref{z}).}
% \label{implementation}
% \end{figure}
As summarized in Fig. \ref{mvdr_beamforming}, the MVDR method is applied to each pixel independently, by:
\begin{enumerate}
\item computing the sample covariance matrix $\eR$ in (\ref{finalR}),
\item computing $\eRi\1$ in (\ref{weights}), and
\item computing the beamformer output $z$ in (\ref{finalZ}).
\end{enumerate}
% This is illustrated in .
% \begin{figure}[!t]\centering
% \includegraphics[width=\linewidth]{gfx/buildR-complex_data-average.eps}
% \caption{.}\label{complexity}
% \end{figure}
Next we will evaluate these steps in terms of arithmetic complexity, and then discuss their mappability to parallel hardware.

\subsection{Computational Complexity}

% Computing and inverting $\eR$ is by far the most computationally intensive tasks in MVDR beamforming. This

If we neglect spatial and temporal averaging, then the computation of $\eR$ is reduced to a single outer product with complexity of O($M^2$), and the inversion is then of O($M^3$). This might lead us to believe that the inversion step is by far the most complex. But if we implement spatial and temporal averaging as in (\ref{spatialR}), then computing $\eR$ is of O($N_K N_L L^2$) and the inversion of O($L^3$). Computing $\eR$ is now the most complex operation whenever $N_LN_K>L$.
\ifPeerReview
\begin{figure}[!t]\centering
\includegraphics[width=.8\linewidth]{gfx/buske3.eps}
\else
\begin{figure}[!t]\centering
\includegraphics[width=\linewidth]{gfx/mvdr_complexity.eps}
\fi%
\caption{Per-pixel computational complexity of the steps in MVDR beamforming (prior to any optimizations). To avoid signal cancellation in an active sonar system we usually set $L<\frac{M}{2}$, in which region the computation of $\eR$ dominates in terms of arithmetic complexity, especially when performing temporal averaging.}\label{mvdr_complexity}
\end{figure}
To visualize these relations, and include the effects of using complex numbers, we set up complexity formulas that account for the total number of arithmetic operations for each step in the MVDR process (see appendix \ref{mvdr_formulas}). We only excluded diagonal loading performed in the computation of $\eR$, and partial pivoting used in the inversion step, since these operations contribute marginally to the end result. The entire range of possible subarray sizes from $L\in[1,M]$ was finally evaluated, with temporal averaging set to $K\in\{0,1,2\}$, and the number of channels set to $M\in\{8,16,32\}$. The results are shown in \Fig{mvdr_complexity}.

We note how the computation of $\eR$ completely dominates at smaller subarray sizes, that solving $\eRi\1$ only plays a notable role for larger array and subarray sizes, and that the computation of $z$ has a negligible impact on computation time. Also notice how temporal averaging comes with a high computational penalty. This is because building $\eR$ is heavy on complex multiplications, which require 3 times as many arithmetic instructions as a complex addition, and a lot of these are repeated unnecessarily.

% Not corrected for fused multiply accumulate

% Also, the need for partial pivoting has been ignored, as we ensure a numerically well conditioned $\eR$ by applying diagonal loading. Since this is the most common scenario, we will have a thorough look at how it can be optimized.

% and does not take into that does not  feasibility parallel decomposition  memory consumption.

\section{Mapping the MVDR to a GPU}\label{maptogpu}

% \begin{itemize}
% \item Keep delay step on CPU, as this is a common step in all beamforming
% \item Increasing matrix size reduces the number of threads per block, since the entire matrix must be in shared memory. 1 matrix 72x72 complex (8 bytes) max per block - 72 threads. same as 207 5x5 matrixes, 5*207=1036 threads. max threads in block 512.
% \end{itemize}

% An important feature of MVDR beamforming, and with beamforming in general, is that each pixel can be computed independently and in an identical fashion. Considering that sonar images often end up having millions of pixels, this makes a compelling case for investigating technologies that support massive data level parallelism. The technology that does this best is perhaps Graphics Processing Units (GPUs). These chips contain up to thousands of small computation cores, which combined deliver superior peak floating point performance compared multi-core systems such as CPUs.

An important feature of MVDR beamforming, as with beamforming in general, is that each pixel can be computed independently and in an identical fashion. Furthermore, a typical sonar image may contain millions of pixels. This represents a level of data parallelism that appears very well suited for a massively parallel architecture such as the GPU.

We decided to investigate the feasibility of running MVDR on a GPU by mapping it to a Nvidia GeForce Quadro 6000. This is a high-end Compute Unified Device Architecture (CUDA) enabled GPU based on Nvidia's Fermi architecture. The code was written in Nvidia's ``C for CUDA'' framework. We believe that GPUs from AMD, or the cross-platform OpenCL framework from the Khronos group, are equally attractive alternatives, but comparing them is beyond the scope of this study.

To use Nvidia's own terminology, the Quadro 6000 is comprised of 14 streaming multiprocessors (SMs), each having 32 CUDA cores that execute a common program called a kernel. Combined these cores deliver a peak performance of more than 1\;Tflop/s (appendix \ref{throughput}). In practice this performance is hard to obtain, but one can get fairly close by balancing the load evenly on all cores, and by trying to avoid that some cores are forced to idle due to a pending data transfer or thread synchronization. As the steps in the MVDR method require different strategies for achieving this, we have designed a different kernel and configuration for each of them. We will pay particular attention to building $\eR$, since the potential for gaining overall speedups are greatest here.

%  this assumes that the implementation is solely bound by arithmetic throughput.  r bound by memory bandwidth, arithmetic latency or arithmetic bandwidth.  or  is solely bound by processing power.   These cores need to be kept busy if we are to That make for a total of 448 CUDA cores which we need to keep busy if we want to tap the full potential.

\ifPeerReview
\begin{figure}[t]\centering
\includegraphics[width=.8\linewidth]{gfx/buske4.eps}
\else
\begin{figure}[t]\centering
\includegraphics[width=\linewidth]{gfx/mvdr_build_R.eps}
\fi%
\caption{Step 1: Building $\eR$. This is a visualization of how $\eR$ could be built in a case with $M=5$ sensors, with subarray size $L=3$ and temporal averaging set to $K=1$. Here $\R'_{l}$ is the sample covariance matrix for the $l$th subarray, and $\breve{\R}$ is the average of $N_K$ of these. Note that instead of performing the temporal sum last as here, one could take more temporal samples into consideration in the computation of each $r_{ij}$.}\label{mvdr_build_R}
\end{figure}

% \todo{Another shitty GPU transition}
% Before proceeding we should make a quick note on the GPU memory. Note that to obtain the best possible performance, we need to make sure that CUDA threads find the data they need in cache or registers most of the time. Only these two memory types have a high enough bandwidth to promote sustained maximum utilisation of all CUDA cores. Unfortunately, they are a sparse resource.

% This makes it apparent that computing one pixel per thread is not a feasible solution. \todo{...}

% When optimizing an algorithm for speed, it is if of importance to know whether an algorithm is bound by computations or memory. One way to get an approximate idea of this is to compute the ratio of peak arithmetic throughput to peak memory throughput of the in the target platform, and compare this ratio to the ratio of arithmetical operations to memory operations in the algorithm.

% follow collective access patterns to memory.

% At first glance, it might seem like a good idea to compute one pixel per CUDA core. Each pixel is computed using the same instructions, and without any exchange of data. However, this approach is unfeasible due to the limited cache and register memory available to each thread. One could resolve to using the slower GDDR5 memory, but this would largely mitigate the performance gain and defeat the purpose of using a GPU in the first place.

%Bandwidth becomes crucial, as getting data in to the cores and out again must happen i a frantic pace. %the feeding all these cores with data and harvesting the result

% To effectively harness the computing potential of a GPU we need to make sure all computing cores stays busy most of the time.

% Memory latency can be hidden in two ways. One is to keep a lot of threads queued up on each SM at all times, such that when some threads perform memory operations other threads can be scheduled to run instead. This is referred to as data level parallelism (DLP). Alternatively, or preferrably in addition, the designer may promote instruction level parallelism (ILP) by letting subsequent instructions in a thread be independent. While ILP can be very effective when done correctly, DLP is generally simpler to design for.
% 
% In the CUDA framework, the threads on an SM is managed through a software abstraction call a compute block. The block is a 3 dimensional grid where each element correspond to a thread that is scheduled to be executed on the SM at some point.

% load balancing
% - distribute load - optimize symmetry
% - avoid branching
% 
% memory latency hiding
% - make threads use high bandwidth memory
% - hide latency with dlp, ilp, or both
% 
% thread sync hiding
% - queue blocks up on an sm

%This is referred to as a single-program, multiple-data (SPMD) architecture\todo{malplaced}.

 %An SM is a single-program multiple-data (SPMD) processor, meaning that all the CUDA cores on an instruction unit, L1 cache and registers is shared by all the CUDA cores on that SM.  %
% GPUs, on the other hand, does not attempt to do ILP, but use all available transistors to support massive DLP. This leads to designs such as the nNvidia GeForce GTX 580. 
% each with 48kB of L1 cache (shared memory) and 128kB of register memory.

% -bank conflicts when L > 32. not treated.

\subsection{Computing the Spatial Covariance Matrix, $\hat{\boldsymbol{R}}$}\label{computing_eR}

% As Fig. \ref{mvdr_complexity} illustrates the computation of $\eR$

As observed in Fig. \ref{mvdr_complexity}, a direct computation of $\eR$ by implementing (\ref{spatialR}) and (\ref{finalR}) is the greatest computational burden. We target this issue by first performing \emph{arithmetic reductions}, then aim to get as close to the peak \emph{arithmetic throughput} on the GPU as possible.

\subsubsection{Arithmetic Reductions}

In Fig. \ref{mvdr_build_R} we illustrate how to reduce the arithmetic complexity of building $\eR$ in a system with $M=5$ sensors, a subarray size of $L=3$ and temporal averaging is set to $K=1$. Here the spatial sum is carried out before the temporal sum. To reduce arithmetic operations we may:
%
\begin{enumerate}
\item exploit the fact that $\eR$ is Hermitian positive semi-definite to compute only one half of it,
\item avoid redundant multiplications, and
\item implement the spatial sum iteratively by sliding along the diagonals of $\eR$.
\end{enumerate}
%
\ifPeerReview
\begin{figure}[!t]\centering
\includegraphics[width=.8\linewidth]{gfx/buske5.eps}
\else
\begin{figure}[!t]\centering
\includegraphics[width=\linewidth]{gfx/buildR-breakdown.eps}
\fi%
\caption{Arithmetic optimization of computing $\eR$: Relative reduction in arithmetic complexity compared to the initial implementation shown in Fig. \ref{mvdr_complexity} (higher is better) . Note how the arithmetic count is reduced by a factor 4-10 in the memory optimized case, and by a factor 6-22 in the instruction optimized case.}\label{mvdr_complexity_optimized}
\end{figure}

To study the effect of these optimizations we altered the complexity formulas to take them into account. Exploiting symmetry and performing iterative summing (step 1 and 3) is always desirable, but avoiding redundant multiplications (step 2) comes at the cost of increased memory consumption. The effect of adding all the optimizations, or all but saving multiplications, is shown in Fig. \ref{mvdr_complexity_optimized}. Here the complexity curves are relative to the reference implementation in Fig. \ref{mvdr_complexity}. Both solutions reduce the complexity considerably, and recomputing the multiplications does not make that much of a difference. We selected the memory minimized version since our algorithm is severely memory bound, as we will see next.

% \begin{table}[b]\centering%\normalsize
% \begin{tabular}[c]{l r@{:}  l}\hline
% \rowcolor{tabBlue}\bf Memory type & \multicolumn{2}{>{\columncolor{tabBlue}}c}{\bf BW$_\text{memory}$/BW$_\text{arithmetic}$} \\\hline
% Global memory &\hspace{30pt} 1 & 30 \\
% Shared memory & 1 & 4 \\
% Registers & $>$3 & 2~\cite{Vasilyy}
% \end{tabular}
% \caption{Nvidia Quadro 6000 bandwidth ratios.}\label{tabbandwidth}
% \end{table}

\begin{table}[b]\centering%\normalsize
\begin{tabular}[c]{l r r r@{}  l}\hline
\rowcolor{tabBlue} & \multicolumn{1}{>{\columncolor{tabBlue}}c}{\bf B$_\text{arith}$} & \multicolumn{1}{>{\columncolor{tabBlue}}c}{\bf B$_\text{mem}$} & \multicolumn{2}{>{\columncolor{tabBlue}}c}{\bf B$_\text{mem}$/B$_\text{arith}$} \\\hline
Arithmetic & 1.03 Tflop/s & & &\\
Global memory & & 36 Gfloats/s & \hspace{30pt} 1 &:30 \\
Shared memory & & 257 Gfloats/s & 1 &:4 \\
Registers & & $>$1.5 Tfloats/s & $>$3 &:2~\cite{Vasilyy}
\end{tabular}
\caption{Nvidia Quadro 6000: Memory throughput, $B_{\lowercase{\text{mem}}}$, compared to arithmetic throughput, $B_\text{arith}$.}\label{throughputs}
\end{table}

\ifPeerReview
\begin{figure}[!t]\centering
\includegraphics[width=.8\linewidth]{gfx/buske6.eps}
\else
\begin{figure}[!t]\centering
\includegraphics[width=\linewidth]{gfx/mvdr_implementation.eps}
\fi%
\caption{MVDR implementated on a GPU. We do this in 3 steps, where each step process the full image before moving on to the next step. \emph{Step~1}: The sample covariance matrix $\eR$ is formed by threads running along its diagonals. This allows spatial averaging to be implemented in a computationally efficient manner and minimizes inter-thread communication. \emph{Step~2}: $\eRi\1$ is computed using the heavily optimized batched linear equation solver from Nvidia. \emph{Step~3}: The beamformer output $z$ is computed in a straight forward fashion by $L$ threads that first sum the subarrays and then apply the MVDR weighting function. A single thread finally sum all the channels up.}\label{mvdr_implementation}
\end{figure}

\subsubsection{Arithmetic Throughput}

When we compute $\eR$, we very rarely perform more than 1-3 floating point operations for every float read or written to memory. Unfortunately, the GPU prefers kernels to be more computationally intensive than this. This can be inferred from Table \ref{throughputs}, where the peak bandwidth of the three types of GPU memory are compared to the peak arithmetic throughput (see appendix \ref{throughput} for derivations). Global memory (RAM), in which all data must reside at some point, is at best only able to move one float for every 30 floats processed by the CUDA cores. This is why the usage of global memory should be minimized. Shared memory, on the other hand, is a very fast level 1 cache that is shared by all threads on an SM. At peak performance it is almost able to keep up with the computing units, but to obtain this performance we must make sure that:
\begin{enumerate}
\item the data needed to compute a pixel can fit in shared memory, while
\item the access patterns we use promotes maximum bandwidth.
\end{enumerate}
 %Further alleviation is also possible by exposing instruction level parallelism within a thread, as this promotes the use of registers with a bandwidth at least 6 times higher than that of shared memory~\cite{Vasilyy}\todo{so why don't we? rephrase}.

These challenges are very closely linked. Since the Quadro 6000 architecture is of compute capability 2.0, it has 48\;kB shared memory (L1 cache) and 128\;kB registers per streaming multiprocessor (SM)~\cite{Nvidia2012}. This memory is shared by all active threads on that SM, a number that should be no less than 768~\cite{Nvidia2012a}. This is to expose a sufficient level of data parallelism to ensure that memory latency is completely hidden (in accordance with Little's law). If we divide the shared memory evenly on all 768 threads we find that each thread should store no more than 8 single precision complex numbers in shared memory, and 24 stored in registers. This should make it apparent why computing a single pixel per thread is a bad idea, and why we need to keep each thread as light on memory consumption as possible.

% What about the GPU implementation? This one was tricky. Balance load, avoid inter-thread dependencies when performing iterative summations, keep thread memory consumption at a minimum, and promote collective access patterns (coalesced reads/writes) to global memory.

In Fig. \ref{mvdr_implementation} we illustrate how we got around these challenges. First we make each SM compute entire range lines of pixels, then load the subset of $\x$ that all these pixels depend upon into shared memory. This lets us perform temporal averaging without having to read additional channel data.

\ifPeerReview\else
\begin{figure}[!t]\centering
\includegraphics[width=\linewidth]{gfx/plot_holmengraa_L16_Navg1.eps}
\caption{HISAS sidescan sonar (SSS) image of the shipwreck Holmengraa that lies on a slanted seabed at 77\;m depth outside of Horten, Norway. The image was processed with $M=32$, $L=16$, $K=1$ and $d=1\%$. Prior to display the image was linearly upinterpolated by a factor 2 in azimuth making its total size 1.46\;Mpx. Note how MVDR improves edge definition and reduces noise in shadow regions.}\label{holmengraa}
\end{figure}
\fi%

Second, we assign $L$ threads per pixel to traverse the diagonals of $\eR$. On the top row a full computation is carried out, then that row is saved back to global memory following a collective access pattern that maximizes global memory bandwidth. For subsequent rows the threads move along the diagonals while performing iterative summations; the result from the previous element on the diagonal is updated by adding and removing the correlation coefficients that enters and exits the sum, respectively. To minimize memory consumption, we compute the coefficients again every time we need them. When a thread has finished up a diagonal, we have them wrap around to compute one of the diagonals in the lower triangular of $\eR$. Since $\eR$ is conjugate symmetric, the values in the leftmost column is obtained by a complex conjugate copy of the relevant value in the first row. Combined these steps balance the load evenly on all threads, is almost completely free of arithmetic redundancy, and consumes less memory.

% Hence, whenever $N_\text{avg}>1$ the sliding window approach will demand less instructions at the expense of a slightly higher memory consumption.

% However, an advantage with this approach is that the summations can be carried out in any order, and hence this form exhibits great deal of parallelism. Can we optimize the algorithm without sacrificing parallelism?

% - Order of summations - beamspace capable

% 
% In the upcoming results, we have beamformed pre-delayed data. This is a large dataset, and the latency experienced when copying it from the \gls{CPU} side to the \gls{GPU} was significant. However, in a practical system only the channel data should be transferred as the delaying of data can effortlessly be performed by the \gls{GPU}. The remaining latency can be hidden as long as the \gls{GPU} is kept busy, hence the data transfer times will not be reported in the upcoming results.
%

% \begin{itemize}
% \item Arrange order of summations to reduce data dimensions as early as possible.
% \item Sliding window in time/space to reduce instructions at the cost of extra memory consumption. Only pays off to do this over subarrays.
% \end{itemize}
% 
% \item Between GPU and CPU: Several concurrent blocks (contexts) per SM.
% \item Job queue (how does this work)
% \item Register pressure and shared memory. Thread context must fit in local memories.
% \item Warp computational intensity should be high to make up for the memory cost.

% increasing performance hit several walls:
% \begin{itemize}
% \item memory couldnt keep track - remedy cache.
% \item instruction parallelism - complex to identity
% \item power consumption cubic with frequency
% GPUs:
% \begin{itemize}
% \item lower frequency, less power
% \item no instruction parallelism
% \item massive amounts of lightweight cores
% \end{itemize}\end{itemize}

% \cite{Owens2007}\todo{Find a good spot for these}

\subsection{Solving $\hat{\boldsymbol{R}}$$\;\!^{-1}$$\mathbf{1}$}

As intuition might suggest the matrix product $\eRi\1$ can be carried out by first inverting the matrix $\eR$, then performing the inner product with $\1$. However, one can also solve the linear equation $\hat{\boldsymbol{R}}\boldsymbol{b} = \1$ for $\boldsymbol{b}$, where $\boldsymbol{b} = \eRi\1$. This is the preferred approach since one vector In terms of arithmetic count the latter is preferred , and when comparing direct implementations, the latter appears to be the preferred option.

An important thing to note, however, is that unlike the problems most GPU libraries tries to solve we do not attempt so solve large linear equations or invert large matrices; our matrices are small, but we have a very large number of them. Fortunately Nvidia recently released a highly optimized batched linear equation solver tailored for this particular task. It is a Gauss Jordan based with support for partial pivoting and complex numbers. The only downside to using it in our application is that it does not exploit the Hermitian property of our sample covariance matrix. A better choice might be a solver based on Cholesky decomposition. These are designed for Hermitian positive semi-definite matrices, and require roughly half the arithmetic operations compared to a Gauss Jordan solver.

% \emph{Computing} $\eRi[n]\1$ (from (\ref{weights})). While intuition may suggest that this step is carried out by inverting $\eR$, it is better to solve the linear equation $\R\beta = \1$ for $\beta$ instead, which gives us $\beta = \eRi\1$ directly. We have tested various solvers for this task, both in-house and proprietary implementations, and achieved the best performance by using an unofficial batch solver from nVidia. Inverting $\eR$ is by far the most computationally intensive task, and a key area of focus for further improvements.

\ifPeerReview\else
\begin{figure}[!t]\centering
\includegraphics[width=\linewidth]{gfx/code_assessment.eps}
\caption{Execution time of an arithmetic-only and a memory-only version of the MVDR code. A dataset from an $M=32$ array was processed for all $L$ using $K=1$, and the mean execution time for a 1\;Mpx image was used here. From this plot we can infer that the kernel building $\eR$ is memory bound, as the time the kernel spends performing memory transactions is higher than the corresponding time it spends carrying out arithmetical operations. Furthermore, when the total runtime is larger than the restricted kernels this can largely be attributed to latency, which we can see that building $\eR$ suffers from with the chosen parameters.}\label{code_assessment}
\end{figure}
\fi%

\subsection{Computing $z$}

The beamformer output $z$ is computed on a per-pixel basis by first computing the MVDR weights $\w$, which are a mere scaled version of $\eRi\1$ (\ref{weights}). Then the sum in (\ref{finalZ}) is found by assigning a group of $L$ threads to respective elements of $x_l$, which then proceed to accumulate these for all $N_L$ subarrays. The resulting data vector is finally multiplied with the weight vector using the same threads, and then a single thread is used to sum these products to obtain $z$.

%  The subarray data can, as here, be reduced to coincide in length with the subarray weights, or the weights can be extended to $M$ in size and applied directly on $\x[n]$ and summed.

% \item \emph{Computing the spatial covariance matrix} $\eR[n]$ (\ref{spatialR}-\ref{finalR}). A group of threads are created that slide along the diagonals of $\eR$. In this way we exploited the fact that entries on the diagonals overlap across subarrays and time, and keep the numbers of both data reads and writes at a minimum.

% There are essentially two ways to accelerate an algorithm on parallel hardware. One is to identify independent instructions for each thread, and run these in in parallel, and the other is to support the parallel execution of as many threads as possible. This is referred to as instruction level parallelism (ILP) and data level parallelism (DLP), respectively. Most general purpose processors, such as a CPUs, support both DLP by featuring multiple cores with single-data multiple-data (SIMD) capabilities, and ILP by through means such as branch prediction and out-of-order execution.

% GPUs, on the other hand, does not attempt to do ILP, but use all available transistors to support massive DLP. This leads to designs such as the nNvidia GeForce GTX 580. Using their own terminology, it is comprised of 16 ``streaming'' multiprocessors (SMs), each having having 32 CUDA\todo{make sure CUDA is introduced} cores that execute a common program called a kernel. %An SM is a single-program multiple-data (SPMD) processor, meaning that all the CUDA cores on an instruction unit, L1 cache and registers is shared by all the CUDA cores on that SM.  %
% each with 48kB of L1 cache (shared memory) and 128kB of register memory.

% In short, designing for a GPU involves finding the optimal balance between hiding memory latency (high occupancy) and resource utilization.
% 
% blocks in grid - high enough so that all SMs have atleast one block to execute. Optimally in the thousands to scale for future devices, and to allow an SM to switch between blocks to keep hardware busy when running synchronizing threads.
% 
% threads in block - min 64. 128-256 better starting point.

% Instead of having a few complex core that can boost massive single thread performance, designers chose to create hundreds of lightweight cores that each is much powerful than a single CPU core, but combined boasts a massive artithmetical performance.
% 
% \begin{itemize}
% \item Lots of cores, little memory for each.
% \item Few data transfers, lots of computations.
% \item Data transfers are moved around by hardware that can operate asynchronously from the compute units.
% \item Memory latency hidden as long as it takes longer to compute than to move data around.
% \item Tasks can be queued on SM (blocks) and on CPU to hide internal and external memory latencies.
% \end{itemize}
% 
% Vast amount of processing power, which are easily accessible but hard to fully harness. This is because in the pursuit of an ever increasing amount of arithmetical computing cores, the 
% 
% Unlike a CPU, a GPU 
% 
% If we compare a GPU with a CPU, 
% 
% When implementing an algorithm on a GPU, it is important to be aware
% 
% People friendly:
% \begin{itemize}
% \item More and more compute units, but less memory for each.
% \item Need to keep each compute unit busy.
% \item Few data transfers, lots of computations.
% \item Data transfers are moved around by hardware that can operate asynchronously from the compute units.
% \item Memory latency hidden as long as it takes longer to compute than to move data around.
% \item Tasks can be queued on SM (blocks) and on CPU to hide internal and external memory latencies.
% \end{itemize}
% 
% The ``fast'' memory of the GPU is the registers and cache. All threads executed 
% 
% 
% \begin{itemize}
% \item Hiding memory access.
% \begin{itemize}
% \item Between GPU and CPU: Several concurrent blocks (contexts) per SM.
% \item Job queue (how does this work)
% \end{itemize}
% \item Register pressure and shared memory. Thread context must fit in local memories.
% \item Warp computational intensity should be high to make up for the memory cost.
% \item Optimized for small L, high M, hardly any time averaging. A typical sonar scenario.
% \end{itemize}
% - Split program into equally sized portions - load balancing
% - Some portions depend on other potions already being carried outer - sequential dependencies
% - Effort timed in order to finish at the same time - synchronization
% 
% - task level parallelism - compute images independent of each other
% - data level parallelism - compute pixels independent of each other

% - Compute bound vs. bandwidth bound

\section{Images and Benchmarks}\label{images_and_benchmarks}

\ifPeerReview
\begin{figure}[!t]\centering
\includegraphics[width=.8\linewidth]{gfx/buske7.eps}
\caption{HISAS sidescan sonar (SSS) image of the shipwreck Holmengraa that lies on a slanted seabed at 77\;m depth outside of Horten, Norway. The image was processed with $M=32$, $L=16$, $K=1$ and $d=1\%$. Prior to display the image was linearly upinterpolated by a factor 2 in azimuth making its total size 1.46\;Mpx. Note how MVDR improves edge definition and reduces noise in shadow regions.}\label{holmengraa}
\end{figure}
\fi%

\ifPeerReview
\begin{figure}[!t]\centering
\includegraphics[width=.8\linewidth]{gfx/buske8.eps}
\else
\begin{figure}[!t]\centering
\includegraphics[width=\linewidth]{gfx/benchmark.eps}
\fi%
\caption{\protect MVDR benchmarks. A 1\;Mpx image from a $M=32$ channel array was processed for all $L$, and for $K\in\{0,1,2\}$.\newline \emph{Top:} The time the GPU spent on building $\eR$, solving $\eRi\1$, and computing $z$. Note the major speedup of building $\eR$ when compared to the complexity plot in Fig. \ref{mvdr_complexity}.\newline
\emph{Bottom:} Compared to a reference Matlab and single thread C implementation running on a CPU the GPU offered a speedup of 2-3 orders of magnitude, but these numbers are somewhat misleading. If the C implementation was properly optimized we expect the GPU to be no more than a factor 5-10 faster, even if its theoretical peak performance is ~20 times higher than that of the CPU.}\label{benchmarks}
\end{figure}

\ifPeerReview
\begin{figure}[!t]\centering
\includegraphics[width=.8\linewidth]{gfx/buske9.eps}
\caption{Execution time of an arithmetic-only and a memory-only version of the MVDR code. A dataset from an $M=32$ array was processed for all $L$ using $K=1$, and the mean execution time for a 1\;Mpx image was used here. From this plot we can infer that the kernel building $\eR$ is memory bound, as the time the kernel spends performing memory transactions is higher than the corresponding time it spends carrying out arithmetical operations. Furthermore, when the total runtime is larger than the restricted kernels this can largely be attributed to latency, which we can see that building $\eR$ suffers from with the chosen parameters.}\label{code_assessment}
\end{figure}
\fi%

\ifPeerReview
\begin{figure}[!t]\centering
\includegraphics[width=.8\linewidth]{gfx/buske10.eps}
\else
\begin{figure}[!t]\centering
\includegraphics[width=\linewidth]{gfx/code_assess_flops.eps}
\fi
\caption{\protect Code efficiency. An estimate of the number of floating point operations per second (Flop/s), found by dividing the theoretical complexity curves by actual run-times. This is a cruel measure as it does not include any other instructions than the actual arithmetic operations in the MVDR computation.}\label{code_assess_flops}
\end{figure}

\ifPeerReview
\begin{figure}[!t]\centering
\includegraphics[width=.8\linewidth]{gfx/buske11.eps}
\else
\begin{figure}[!t]\centering
\includegraphics[width=\linewidth]{gfx/benchmark_boston.eps}
\fi
\caption{\protect MVDR benchmarks from Boston HPC centre with the new high-end Nvidia K20 Kepler GPU. The exact same scenario and code as in Fig. \ref{benchmarks} was used here. With no code alterations the performance was only improved marginally compared to running on the Quadro 6000.}\label{benchmarks_boston}
\end{figure}

% In a practical system it should be moved to the GPU as well. Therefore, in the upcoming benchmarks, CPU computation time and data transfer to the GPU is neglected.

To demonstrate the imaging capability of the MVDR beamformer, we have processed experimental datasets from the $M=32$ element Kongsberg Maritime HISAS1030 sonar mounted on a HUGIN AUV~\cite{Hansen2009}. HISAS1030 is a high resolution synthetic aperture sonar with an array length of 1.2\;m and operating frequency of 100\;kHz. To produce the image shown in \Fig{holmengraa} it was operated in sidescan mode. The studied object is the 1500 dwt oil tanker wreck Holmengraa. It is 68\;m long and 9\;m wide, and lies at a slanted seabed at 77\;m depth outside of Horten, Norway. The 1\;Mpx MVDR image were here processed with parameters $L=16$, $K=1$, and $d=1\%$.

The computational performance of our implementation was first assessed on a test system with a quad-core Intel Xeon E5620, 64\;GB of RAM and an Nvidia Quadro 6000 card. The results were obtained by processing a 1\;Mpx image from the data from a 32 channel array, for all subarray sizes $L$, and for $K\in\{0,1,2\}$. Run-time measurements are shown in \Fig{benchmarks}, the run-times of memory-only and arithmetic-only GPU kernels are depicted in \Fig{code_assessment}, and an estimate of computation efficiency is presented in \Fig{code_assess_flops}. We were also granted a test drive at the Boston HPC centre on a machine with an Intel Xeon E5-2670, 32\;GB RAM and an Nvidia K20. The results from this run are shown in \Fig{benchmarks_boston}.

As seen in the run-time comparisons presented in \Fig{benchmarks}, the GPU method outperforms the CPU implementations by 2-3 orders of magnitude in terms of processing speed. The bottleneck in the final design is now the inversion step, which is typically 5 times slower than the build step. In most cases the processing speed of the Quadro 6000 is above 1\;Mpx/s, and this was only improved by a factor 1.5-2 when run on the new K20 GPU. The benchmarks of our memory-only and arithmetic-only kernels show that the kernels spend roughly the same time on both these tasks, so optimising only one of these further will have marginal effect. All GPU kernels were compiled with \texttt{nvcc} at optimization level \texttt{O2}. Excluded from these benchmarks is the data transfer time from CPU to GPU, which account for 2-20\% of the total processing time. This is to limit the scope of the article, but will need to be taken into consideration in an implementation of a full sonar system.

% 
% We have tested our \gls{GPU} implementation of the \gls{MVDR} beamformer experimental datasets from the 32 element Kongsberg Maritime HISAS1030 sonar\todo{shouldn't we state that it's SAS?} being attached to the HUGIN \gls{AUV}~\cite{Hansen2009}. To obtain the studied object, the 1500 dwt oil tanker wreck Holmengraa, the sonar was operated in sidescan mode to image (\Fig{hlmengraa}). Holmengraa is roughly 68\;m long and 9\;m wide, and lies on a slanted seabed at 77\;m depth. The image was processed with \gls{MVDR} parameters $L=16$, $N_\text{avg}=3$, and $d=1\%$, which proved to be a reasonable selection for this scenario.
%

\section{Discussion}\label{discussion}

In accordance with previous studies, \Fig{holmengraa} demonstrates the MVDR beamformer's ability to produce images with suppressed interference and improved detail resolution. Compared to the DAS beamformer, the ship's edges appear sharper and the shadows less noisy. In this scenario the MVDR's performance was not particularly sensitive to parameter adjustments. Similar performance was obtained with arbitrary combinations of $L\in\{12,16,24\}$, $K\in\{1,2,3\}$ and $d\in[0.01,0.05]$, and no adjustments had to be made when processing other parts of the scene.
%But while \gls{DAS} processed this 1.3\;Mpixel image in milliseconds, our Matlab implementation\todo{never mentioned matlab vefore} of MVDR beamformer needed minutes.

As observed in \Fig{benchmarks} the combination of performing arithmetic reductions and implementing the MVDR on a GPU lead to a significant improvement in processing speed. Compared to a Matlab and single thread C implementation a speedup of 2-3 orders of magnitude was achieved, but we do not consider this comparison fair. In fact, the theoretical peak throughput of this GPU is roughly 20 times higher than that the CPU in question, meaning the potential of the CPUs was far from reached in our initial implementations. This is mainly because the entire GPU design is focused on efficient use of fast memory cache, while no similar optimization work was carried out on the CPU implementations.

Note from \Fig{benchmarks} how the building of $\eR$ now only takes up a fraction of the total processing time, and recall that it was the other way around when we presented the theoretical complexity curves in \Fig{mvdr_complexity}. Also, the benchmark curves for building $\eR$ now seem linear. The main reason for this is that the optimization step negated much of the extra complexity introduced by averaging, i.e. reduced the complexity from O($N_KN_LL^2$) to less than O($L^2$). Another reason for the run-time to appear more like O($L$) is likely the GPU, since our design makes better use of the GPU's resources when there is more processing involved per pixel. The inversion step, on the other hand, gains less from being implemented on a GPU. This is because its nature is less data parallel, and the back substitution involved in its computation is more of a sequential nature. This difference be can observed in \Fig{mvdr_complexity}. Alternative solvers, such as one based on Cholesky decomposition that 
exploits the Hermitian property of $\eR$, can in theory reduce complexity by a factor ~2, but we question whether this result can be obtained in practice using a GPU. The CPU implementations perform Cholesky based inversion, but this does not explain why the C and Matlab implementation have a total run-time that is linearly dependent on $L$. We find this observation odd, but believe it to be accurate.

It is important to know whether a GPU design is limited by memory bandwidth or arithmetical throughput. To measure this  we made two versions of the MVDR kernel, one that only performs arithmetic operations, and one that only performs memory operations. Then we benchmarked these kernels and compared their run-times to the total run-time (\Fig{code_assessment}). A first thing to note is how these kernels seem equally occupied performing memory and arithmetic operations. This is a good sign, but since we consistently minimized memory consumption at the expense of some extra computations when building $\eR$ , we know it to be bound by memory bandwidth. It also has a problem with latency, which can be inferred from the total run-time being significantly higher than that of the two single-function kernels. Since the GPU hardware can carry out memory and arithmetic operations concurrently and independently, this gap indicates that the GPU sometimes does ``nothing''. In our case this is due to synchronization hold-
ups when performing temporal averaging, which is carried out in a sequential manner.

Even with all our efforts we were only able to obtain processing rates of 40\;MFlop/s on average in the desirable range of subarray sizes (\Fig{mvdr_complexity}). This is mainly due to the inversion step, since we for the most part obtained more than 100\;MFlop/s when building $\eR$. While these numbers are likely slightly underestimated, they are still far away from the Quadro 6000's peak of 1\;Flop/s. Again, we believe this is due to memory bandwidth. This belief is further supported by the test results from running the code on the new K20 Kepler card, where we saw a modest factor 1.5-2 speedup. Although the Kepler card peaks at 3.5-4\;Tflop/s, its shared memory bandwidth is only approximately twice that of the Quadro 6000.

% Interesting to note is also that solving $\eRi\1$ remains the main bottleneck in the GPU implementation, and it gets even worse for larger $L$'s. However, this is a general Gauss Jordan solver which does not exploit symmetry properties of $\eR$, so creating, e.g. a batch based Cholesky solver should improve the runtimes further. Both the C and Matlab implementation in this benchmark take this symmetry into account, which partially explains why the complexity curves of these implementations are different from that of the GPU implementation.

% If the HISAS1030 was attached to a platform moving at 1.8m/s, for which the maximum range is 250\;m~\cite{Hansen2010}, the pulse repetition frequency could be set to 3\;Hz. We could further assume a complex sampling frequency of 40\;kHz to accomodate the  HISAS1030's bandwidth of 30\;kHz, and beamform $M=32$ lateral beams. The throughput required for real-time processing of these images would then be 1\;Mpx/s. As we have seen, this is something our GPU implementation can handle if $L$ is not too large.\todo{droppe denne kanskje}%, leaving the \gls{CPU} free to take on other assignments.

% What may this performance be used for? Let us assume that the 32 element, 1.2\;m long HISAS1030 array is mounted on a platform moving at 1.8\;m/s. Then the pulse repetition frequency (PRF) could be set to 3\;Hz to ensure critical along-track sampling, 
% To illustrate what this performance might be used for, consider a moving platform travelling at 1.8\;m/s. 
% If the sonar system is Maximum range 
% There is a limit to the along-track sampli
% To avoid along track undersampling, the

% HUGIN: 1.8m/s - range 250m \\
% fs = 100kHz*30\%*$\frac{4}{3}$ = 40kHz \\
% $N_{range px} = \frac{2\;250m}{1500m/s}40kHz = 13.3kpx$ \\
% $\times 32$ beams = 426kpx \\\\
% 
% PRF = $\frac{1500m/s}{2 250m} = 3Hz$ \\
% TP = 215px 3Hz = 1.28Mpx/s \\
% 4 arrays: 1.28Mpx/s * 4 = 5.12Mpx/s
% 
% 
% \begin{itemize}
% \item Need for speed: HUGIN 4 banks of 32 elements, can be processed faster than the ping repetition rate, with margins to spare.
% \end{itemize}

\section{Conclusion}\label{conclusion}

The MVDR beamformer is an algorithm capable of producing images with improved detail resolution and contrast compared to conventional DAS beamforming. The downside is its inherent need for robustification and the high computational complexity associated with estimating and inverting the spatial covariance matrix.

We have shown that for systems containing up to 32 channels the problem can be largely mitigated by building the covariance matrix in a clever way, and by making use of the massive computational power available in modern GPUs. We were able to improve upon the run-time of a single thread C-implementation by roughly two orders of magnitude. For most choices of parameters the GPU was able to create images at ~1 Mpx/s at an average data processing rate of ~40\;GFlop/s. This is less than 5\% of the peak performance of the GPU, but we believe it to be near optimal given the constraints of memory bandwidth and the sequential nature of some parts of the MVDR implementation.

%  This performance is sufficient for computing properly sampled full-coverage sectorscan images from the HISAS1030 sonar in real-time.

All in all, the MVDR maps well to the GPU since the computations involved are independent on the pixel level, and partially also within each pixel. The GPU allows MVDR to be used in real-time processing of sonar data, and makes the MVDR a viable alternative to conventional methods in practical systems.

% \begin{figure}[t]
% \centering
% \includegraphics[width=\linewidth]{gfx/benchmark_tagged.eps}
% \caption{Realtime requirement for sectorscan imaging.}\label{real-time_sectorscan}
% \end{figure}

%%%%%%%%%%%%%%%%%%                              ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
% DOCUMENT APPENDICES %
%%%%%%%%%%%%%%%%%%                              ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

% \titleformat{<command>}[<shape>]{<format}{<label>}{<sep>}{<before>}[<after>]

% \titleformat{\section}[hang]{\bf}
% {\thesection.\enspace}%\thesubsubsection}%  {\footnotesize \enspace \emph{Sec.}  }
%    {0pt}{\MakeUppercase}[]
% \titlespacing*{\section}{0pt}{2\lineheight}{\lineheight}

% {-2ex plus -.5ex minus -.2ex}{1.0ex plus .2ex minus .2ex}

% \renewcommand\section[1]{{\Large\bf\MakeUppercase #1}}
% \renewcommand\section*[1]{{\Large\bf\MakeUppercase #1}}

% 
% 
% 
% 
% 
% 
% \section{Introduction}
% 
% \todopar{Case:
% \begin{itemize}
% \item UFFC(?)
% \item Ultrasound. Arrays up to 128 channels. Processing power vital. What's the best way to implement 
% \item Other articles evaluates Capon image quality, and some have shown that speedups can be achieved by reducing the complexity of the traditional Capon algorithm or by implementing specific versions on a GPU.
% \end{itemize}
% }
% 
% \begin{itemize}
% \item As briefly as possible, introduce beamforming.
% \item Adaptive beamformer's potential lies in its ability to suppress interference power
% \item Why adaptive beamformers struggle in active sonar systems. Correlated noise, robustification kills the adaptive potential. Quite computationally intensive. Constraints must be applied in one way or another - parameters must be tuned.
% \item The three main ways to implement Capon \cite{Capon1969} is \todo{not entirely true?}
% \begin{itemize}
% \item Traditional way: Build $\eR$ and solve $\w = \frac{\Ri\a}{\a\H\Ri\a}$. 
% \item Iterative methods: Woodbury, Conjugated gradients
% \item Beamspace
% \item LCA (or other methods, as reference)
% \end{itemize}
% fs = 100kHz*30\%*$\frac{4}{3}$ = 40kHz \\
% $N_{range px} = \frac{2\;250m}{1500m/s}40kHz = 13.3kpx$ \\
% $\times 32$ beams = 426kpx \\\\
% 
% PRF = $\frac{1500m/s}{2 250m} = 3Hz$ \\
% TP = 215px 3Hz = 1.28Mpx/s \\
% 4 arrays: 1.28Mpx/s * 4 = 5.12Mpx/s
% 
% 
% 
% 1Tflops (ops/cycle?)
% 144GB/s global
% 1TB/s shared
% x6 regs
% 
% latency - arithmetic (20 cycles) and memory (400+ cycles)
% 
% Little's law - needed parallelism = latency * throughput
% arithmetic: 23 cycles * 32ops/SM/cycle = 576 ops/SM parallelism
% memory: <800 cycles * <144GB/s=125B/cycle = <100kB in the flight
% 
% On GF104: must have some ILP to get >0.66 of peak:
% - 48cores/SM, one instruction is broadcast across 16 cores
% - need 3 instructions per cycle
% - but only have 2 warp schedulers
% - instead, can issue 2 instructions per warp in the same cycle
%

\appendix

\section{MVDR Complexity Formulas}\label{mvdr_formulas}

To estimate the number of floating point operations needed to MVDR beamform a single pixel, we formed expressions that accumulate the number of complex arithmetic operations found in the MVDR process. From observing the generated assembly code we inferred that each complex addition and multiplication would require $\Oa = 2$ and $\Om = 6$ floating point operations, respectively. The formulas are listed for each step below for reference.

\emph{Building $\eR$}. The initial complexity of this step can be inferred directly from (\ref{spatialR}) and (\ref{finalR}):
\begin{align}
O_{\underset{\text{initial}}{\text{Build $\eR$}}} &= \underbrace{\Om\Nk\Nl L^2}_\text{Multiplications} + \Oa(\Nk+\Nl-2)L^2 + O_\text{dload},
\end{align}
where $O_\text{dload} = (2L-1)\Oa + \Om$ is the minor cost of performing diagonal loading. If we apply the optimization strategies discussed in section \ref{computing_eR} we can arrive at the following instead:
\begin{align}
O_{\underset{\text{min arith}}{\text{Build $\eR$}}} &= \underbrace{\Om\frac{M+\Nl}{2}L}_\text{Multiplications} + \underbrace{\Oa(\Nk-1)(\Nl-1)L}_\text{First row additions} \nonumber\\
&+ \underbrace{\frac{(L-1)(L-2)}{2} \bigg[ 2\Oa + 2(\Nk-1)\Oa \Big]}_\text{Iteration additions} \nonumber\\
&+ O_\text{dload}
\end{align}
Of the solutions discussed, this is the least expensive in terms of arithmetic instructions. However, if memory bandwidth is a limiting factor a better solution is to recompute multiplications where they are needed:
\begin{align}
O_{\underset{\text{min mem}}{\text{Build $\eR$}}} &= \underbrace{\Om\Nk\Nl{}L}_\text{First row multiplications} + \underbrace{O_a(\Nk-1)(\Nl-1)L}_\text{First row additions} \nonumber\\
&+ \underbrace{\frac{(L-1)(L-2)}{2}\bigg[ 2\Oa + 2(\Nk-1)\Oa + 2\Nk\Om \bigg]}_\text{Iteration multiplications and additions} \nonumber\\
&+ O_\text{dload}.
\end{align}

\emph{Solving $\eRi\1$} is achieved by using a batched Gauss Jordan solver with support for complex numbers and partial pivoting. Its complexity - with partial pivoting excluded - can be expressed as: 
\begin{align}
O_{\text{Solve $\eRi\1$}} = \sumb{r=0}{L} \bigg[&\underbrace{\big(L-r\big)\big((L-r+2)\Oa + (L-r+3)\Om\big)}_\text{Reduction}\nonumber\\
+ &\underbrace{(r-1)\Oa + r\Om}_\text{Backsubstitution}\bigg],
\end{align}
where $r$ is a running variable $r$ that indexes rows in the augmented matrix $\bmat{\eR|\1}$.

\emph{Computing $z$} is very simple once the covariance matrix $\eR$ is built and inverted, and has an near negligible impact on performance: 
\begin{align}
O_\text{Compute $z$} = \Oa(2L-2)2 + \Om 3 L.
\end{align}

\section{GPU Throughput}\label{throughput}

In the context of determining whether an implementation is computationally bound or memory bound, one should first compare the target platform's sustained computational throughput to sustained memory throughput. Let us start with the former.

The Quadro 6000 has 32 CUDA cores per SM, each operating at a rate of 1148\;MHz and being able to perform 2 floating point operations (flop) per clock cycle when multiply-add instructions are used. Theoretical peak arithmetic throughput is then given as:
\begin{align}
\text{B}_\text{arith} &= 2\cdot1148\;\text{flop/s/core}\cdot 32\;\text{cores/SM}\cdot 14\;\text{SMs}\nn
&= 1.03\;\text{Tflop/s}.\label{flops}
\end{align}
Now let us compare this to the memory throughput. The ``global'' GDDR5 memory bus is 384\;bit wide, and operates at 3\;Ghz where 2 bits are sent every cycle. Its peak bandwidth is then:
\begin{align}
\text{B}_\text{gmem} &= \frac{2\cdot 3\;\text{Gbit/s}\cdot384\;\text{bit}}{8\;\text{bit/byte}}\nn
&= 144\;\text{GB/s}\ (36\;\text{Gfloats/s}).\label{bwglobal}
\end{align}
The shared memory, on the other hand, is organized into 32 banks per SM, each being 32\;bit wide and operating at 1148\;MHz where 1 bit is sent per cycle. Its peak aggregated bandwidth is then:
\begin{align}
\text{B}_\text{smem} &= \frac{\frac{1148}{2}\;\text{Mbit/s}\cdot32\;\text{bit/bank}\cdot 32\;\text{banks/SM}\cdot 14\;\text{SMs}}{8\;\text{bit/byte}}\nn
&\approx 1.03\;\text{TB/s}\ (257\;\text{Gfloats/s}).\label{bwshared}
\end{align}
The bandwidths are compared in Tab. \ref{throughputs}. Note that even when using shared memory at least 4 floating point operations must be carried out per float transferred to the CUDA cores, otherwise the algorithm will be memory bound and the peak arithmetic throughput can not be reached. %In such cases instruction level parallelism can be exploited to move data from shared memory into the even faster register memory~\cite{Vasilyy}, but this is outside the scope of this article.

% \subsection{Iterative methods}
% 
% \begin{itemize}
% \item Woodbury
% \item Conjugated gradients
% \end{itemize}
% 
% 
% \subsection{Beamspace}
% 
% \begin{itemize}
% \item Some minimum of background, and good references.
% \end{itemize}
% 
% 
% \section{Results/Discussion}
% 
% 
% 
% 
% 
% \begin{itemize}
% \item Compare computational complexity vs. memory constraints for the various methods. Supporting plots.
% \item Relate this to whether the favourable architecture is a CPU or a GPU \todo{check ``when to gpu article''}
% \item Run benchmarks on some sonar dataset(s)
% \end{itemize}
% 
% % \IEEEPARstart{T}{o} form images from a modern phased array sonar system the received wavefield is usually recorded, and then postprocessed by a digital beamformer. The beamformer applies delays and weights to the sensor channels, the beamformer adjusts the arrays spatial response to focus at one pixel at a time.  such that signals emanating from regions of interest add constructively, while ensuring that noise and interference from other angles do not. 
% % 
% % The imaging capabilities of a modern phased array sonar system depend on physical attributes such element response and array geometry, the transmitted signal, as well as the beamforming method being used on transmission and reception. Beamforming is the concept of applying delays and weights to the sensors channels to steer the arrays response to points of interest. 
% 
% % 
% % 
% % Outline:
% % \begin{itemize}
% % \item Capon's resonse when applying robustification
% % \item Choice of window functions makes little difference.
% % \item Steering and mainlobewidths have outer bounds.
% % \item Beamspace?
% % \item Chosen window plots - what may they tell us? Variance intensity values when using various windows.
% % \item Assymmetric windows?
% % \end{itemize}
% % 
% % 
% % \begin{align}
% % z[n] = \sumb{m=0}{M-1} w_m[n]^*x_m[n-\Delta_m] = \w\H[n]\x[n-\Delta_m]
% % \end{align}
% % 
% % 
% % \section{Methods}
% % 
% % Basically, we are working with a practical implementation of the Capon beamformer that computes a set of weights $\vec w$ for every single sample $n$ by solving:
% % \begin{gather*}
% % \vec w[n] = \frac{\hat{\mat R}\;\!^{-1}[n]\vec a}{\vec a\H\hat{\mat R}\;\!^{-1}[n]\vec a} = \frac{\text{\raisebox{1.9pt}{$\vec\chi$}}[n]}{\vec a\H\text{\raisebox{1.9pt}{$\vec\chi$}}[n]}
% % \end{gather*}%
% % where
% % % \newcommand\X{\text{\raisebox{2pt}{$\vec\chi$}}}
% % \begin{gather*}
% % \text{\raisebox{1.9pt}{$\vec\chi$}} = \hat{\mat R}\;\!^{-1}\vec a \qquad\qquad\text{is the solution to}\qquad\qquad \hat{\mat R}\text{\raisebox{1.9pt}{$\vec\chi$}} = \vec a.
% % \end{gather*}
% 
% % 
% % , and maximum suppression of while ensuring that the beamformer digitally  before each of the pixels are estimated one at a time. The resolution and contrast of such a system will depend on the systems spatial response, which ideally should be narrow  be very sharp in the desired direction its ability to achieve  fundamental principle of forming a sonar image is to record the received wavefield, 
% % 
% % image quality of a phased array sonar imaging systems depend on  the choice of weights to apply to each of the sensors are crucial. 
% % 
% % A modern phased array imaging system may be thought of as a spatial filter. To achieve the best possible performance, the 
% % 
% % resolution and contrast 
% % 
% % Adaptive beamformers have only recently been introduced in active sonar imaging. For a while they were considered unsuited for this purpose because the backscattered signal is largely correlated with the 
% % 
% 
% 
% % An example of a floating table. Note that, for IEEE style tables, the 
% % \caption command should come BEFORE the table. Table text will default to
% % \footnotesize as IEEE normally uses this smaller font for tables.
% % The \label must come after \caption as always.
% %
% %\begin{table}[!t]
% %% increase table row spacing, adjust to taste
% %\renewcommand{\arraystretch}{1.3}
% % if using array.sty, it might be a good idea to tweak the value of
% % \extrarowheight as needed to properly center the text within the cells
% %\caption{An Example of a Table}
% %\label{table_example}
% %\centering
% %% Some packages, such as MDW tools, offer better commands for making tables
% %% than the plain LaTeX2e tabular which is used here.
% %\begin{tabular}{|c||c|}
% %\hline
% %One & Two\\
% %\hline
% %Three & Four\\
% %\hline
% %\end{tabular}
% %\end{table}
% 
% 
% % Note that IEEE does not put floats in the very first column - or typically
% % anywhere on the first page for that matter. Also, in-text middle ("here")
% % positioning is not used. Most IEEE journals use top floats exclusively.
% % However, Computer Society journals sometimes do use bottom floats - bear
% % this in mind when choosing appropriate optional arguments for the
% % figure/table environments.
% % Note that, LaTeX2e, unlike IEEE journals, places footnotes above bottom
% % floats. This can be corrected via the \fnbelowfloat command of the
% % stfloats package.
%

%%%%%%%%%%%%%%%%%%                              ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
% DOCUMENT APPENDICES %
%%%%%%%%%%%%%%%%%%                              ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

\appendices

% use section* for acknowledgement
\ifCLASSOPTIONcompsoc% % This command fixes abstract positioning for compsoc articles:
% \IEEEdisplaynotcompsoctitleabstractindextext
% 
% % (Optional) Add some extra info on cover page of peer review papers:
% % \ifCLASSOPTIONpeerreview
% % \begin{center} \bfseries EDICS Category: 3-BBND \end{center}
% % \fi
% 
% % Insert page break and insert second title (peer review mode)
% \IEEEpeerreviewmaketitle
% 
% 
% 
% 
% \maketitle
% 
% \section{Introduction}
% 
% % Images from a phased array active sonar system are formed in the post-processing by focusing on one pixel at a time. This is achieved by delaying and weighting each of the array's channels, providing coarse and fine adjustments to the region of focus, respectively. This is commonly referred to as beamforming. Assuming far-field conditions, the delays are usually chosen to steer the pixel of interest to broadside, and the weights chosen to reject the noise impinging on the array from non-broadside directions.
% % 
% % The algorithm responsible for carrying out this task is referred to as a beamformer. Depending on whether data is evaluated when choosing a suitable set of delays and weights, beamformers can be branded either conventional or adaptive. A good example of the former is the The Delay-and-Sum (DAS) beamformer, which - as the name implies - simply delay the sensor channels appropriately prior to summing them up. Applying various weight sets, or windows, to the data prior to summation allows the lateral \gls{SNR} to be improved, but at the inevitable expense of a deteriorated lateral resolution. \todo{aka. the curse of physics.}
% % 
% % The Minimum Variance Distortionless Response (MVDR) (or \gls{MVDR}) beamformer, have the potential to overcome this limitation. While ensuring unit gain in the look direction, it computes the set of weights that minimizes the energy accumulated by the array from other directions. In other words, whenever the distribution of noise and interference is primarily focused around specific angles, an array response with a high degree of suppression at these angles are chosen, resulting in an image with a lateral \gls{SNR} and resolution superior to anything the \gls{DAS} can come up with.
% 
% 
% % 
% % In active sonar imaging the constrast and resolution can be improved by the use of adaptive beamforming. 
% % 
% % Adaptive beamforming 
% % A critical part in the sonar image formation is the 
% % The image quality is mainly limited by beampattern of the imaging arrays.  resolution and noise suppression
% % 
% % The quality of sonar images 
% % 
% % An essential part of a sonar imaging system is the beamformer. 
% % 
% % 
% % A sonar image is typically formed by using receive beamforming to focus on one pixel at a time. The beamformer's ability to suppress interference   to focus on one pixel at the time. How well a pixel value is estimated, and thus how 
% % 
% % A sonar image is built one pixel at the time the 
% % 
% % The image quality of a sonar system will depend on the imaging arrays ability to focus in 
% % 
% % 
% % 
% % Forming images done by combining data from multiple receivers in a way that to obtain focus on one pixel at the time, and assigning the output value to the intensity of that pixel. The 
% % 
% % Most modern active sonar systems employ phased arrays to 
% % The basic way of a sonar image was formed 
% % 
% % A sonar system's ability to 
% % 
% % Adaptive beamforming has 
% % 
% % The image quality 
% % 
% % To improve the image quality of 
% % 
% % In high resolution sonar imaging the 
% % 
% % The image 
% % 
% % 
% % In a typical sonar imaging scenario an encoded signal is transmitted to highlight a surface of interest, and the reflected wavefield sampled using a receiver array. A image is then formed in the post-processing by coherently combining the receiver outputs to focus at one pixel at the time. The principle of adjusting the arrays focus is commonly referred to as beamforming, and involves assigning suitable delays and weights to the array's channels.
% % 
% % Ideally, 
% % 
% % The formation of a sonar image is 
% % 
% % The principle of forming an image in active sonar systems usually 
% % 
% % A common way of creating high resolution sonar images is to use multielement arrays to  
% % 
% % In sonar imaging a common compromise 
% % 
% % A common way of creating sonar images is to use 
% % 
% % The drive 
% % 
% % The pursuit of 
% % 
% % Adaptive beamforming 
% % 
% % Beamforming is a fundamental operation in active sonar imaging. 
% % 
% % 
% % 
% % 
% % Beamforming is a fundamental technique employed in sonar image formation to 
% % 
% % 
% % The introduction of adaptive beamforming in active sonar imaging 
% % 
% % The resolution and 
% % 
% % The High resolution imaging techniques
% % 
% % 
% % The resolution and constrast in active sonar images depend on the beamformers ability to 
% % 
% % A sonar image is typically formed by using receive beamforming to focus on one pixel at a time.   The image resolution and contrast The image quality The quality of active sonar images depend - amoung other things - on the beamformer's ability to suppress interference  resolution and constrast
% % 
% % A sonar image is typically formed by using receive beamforming to focus on one pixel at a time. The quality of the final image will depend on the beamformer's ability to suppress interference and noise from off-pixel.    to focus on one pixel at the time. How well a pixel value is estimated, and thus how
% % 
% % 
% % In conventional active sonar imaging systems 
% % 
% % Various methods have been applied to active sonar imaging. One of these methods, 
% % 
% %  eamforming techniques have been applied in several fields to improve image. methods can be applied to 
% % 
% % Most active sonar imaging systems are based on conventional beamforming methods. These are robust and simple, but does not take the incoming wavefield into account. offer  in the image formation process. While being inherently robust, these methods 
% % 
% % Adaptive beamformers
% 
% % With the evolution of processing power the 
% % 
% % adaptive method
% % hot topic
% % 
% % With the ever increasing processing power found in modern computers 
% % 
% % As the processing power of todays modern computer systems keeps improving the feasibility of 
% % 
% % As the processing power of modern computer systems keep improving, the use of data adaptive methods in various imaging fields. 
% 
% Data adaptive methods have been reported to improve image quality in various imaging fields over the years. One such method, the Minimum Variance Distortionless Response (MVDR) beamformer, has been 
% hown to produce images with improved constrast and resolution compared to conventional non-adaptive methods, in e.g. radar~\cite{Benitz1997}, ultrasound~\cite{\cite{Synnevag2007}}, and passive sonar. Recently has its feasibility in active sonar also been demonstrated~\cite{Blomberg2012a,Blomberg2011,Dursun2009,Lo2004}.
% 
% % \the\linewidth
% 
% Despite its inherent potential, the MVDR beamformer has yet to see widespread adoption in the active sonar field. There may be several reasons for this. For one, the method is not inherently robust, and will suffer from a phenomenon called signal cancellation in active systems~\cite{Widrow1982}. Another reason is that in its original form, the computational complexity is cubic with the number of channels, O(M$^3$), while conventional beamformers are at O($M$). This is because a spatial covariance matrix is estimated and inverted for each image pixel.
% 
% % Previous:
% 
% % Robustification:
% % R estimate accuracy vs. dynamic trackability: (1) Trade observation time for bandwidth (2) Partial adaptiveness (less degrees of freedom than channels) (3) Quadratic constraint on sensor weights to limit the norm to some preset value.
% % -  
% 
% Previous work have shown that the robustness issue may be solved through means such as temporal and spatial averaging~\cite{Kailath1985}, and regularization. Reducing complexity has been achieved in a variety of ways, often by enforcing a Toeplitz-structure on the covariance matrix to simplify its inversion, or by employing iterative solvers.
% 
% We will show that by exploiting symmetry and avoiding unnecessary computations, the spatial averaging will also aid in reducing the overall arithmetic complexity\todo{make sure this is shown clearly somewhere}. for Unlike other work, which focus modifying or transforming\todo{get low channel count in, as it makes a case against the other solutions.} We make no approximations, but rather stay with a textbook example of the method that handle complex basebanded data\todo{misplaced?}.
% 
% % Other work (keywords: mvdr, capon, computation, efficient, complexity):
% % - Partially adaptive beamforming. DAS on non-overlapping subarrays, then MVDR \cite{Swingler1999a}.
% % - Beamspace / frequency based methods.
% % - Low complexity. PCA + LC. 
% % - Systolic arrays. Radar \cite{Ward1984} 
% % - Toeplitz. \cite{Yapanel} Spectrum estimation - speech recognition. 
% % 
% % Us:
% % - Aim for the GPU. Simple is often better.
% % - Therefore: Optimize the standard MVDR chain.
% % 1. Reduce arithmetic instructions
% % 2. Implement for full utilisation of the arithmetic throughput on the GPU.
% % 
% % (Made partially adaptive by employing subarray averaging to reduce degrees of freedom to L.)
% 
% 
% Furthermore, we investigate the feasibility of accelerating the MVDR method using Graphics Processing Units (GPUs). These chips have in recent years emerged as a flexible and highly potent parallel technology well suited for many DSP tasks. Driven by the gaming industry, these devices are engineered to maximize Flop/s while keeping cost and power consumption at a minimum. Compared to many-core CPUs, a GPU core offer modest computational capabilities, but in return provide up to thousands of them\todo{misplaced?}. 
% 
% Comparable work may be found in e.g. ultrasound, where Chen et. al. demonstrated a GPU implementation of the MVDR that operated on real data~\cite{Chen2011,Chen2011a}. While promoting impressing performance, such a constraint disallow the creation of asymmetric and shifted responses, and hence prevents the MVDR's potential to be fully reached. The complexity issue has been targeted in various ways. 
% 
% This article is outlined as follows. First, we will introduce the concept of adaptive beamforming and provide details MVDR beamforming and our selected robustification techniques. Then we will investigate the complexity issue, discuss means for arithmetic reduction, and show how to properly map it to a GPU.\todo{too much overlap here}
% 
% %  Furthermore, by implementing it on a modern \gls{GPU} we have gained a speedup of around 2 orders of magnitude compared to an optimized single-thread implementation in C.
% 
% % These devices have in recent years emerged as a flexible and highly potent parallel technology well suited for many DSP tasks, and are available off-the-shelf at reasonable prices\todo{skip this?}.
% 
% 
% % These devices have in recent years emerged as a flexible and highly potent parallel technology well suited for many DSP tasks, and are available off-the-shelf at reasonable prices.
% 
% % Yay or nay? $\Rightarrow$ GPU ultrasound (\cite{So2011,Chen2011}). Complex data, all robustification techniques.
% 
% 
% % - Adaptive beamformers offers image quality edge, but are usually computationally complex.
% % - Who's tried before?
% % - We: Active sonar MVDR, GPU, high performance.
% 
% 
% \section{Methods}\label{methods}
% 
% Consider a sonar imaging scenario where an encoded signal is transmitted to highlight a surface of interest, and assume that the backscattered wavefield is properly sampled using a uniform linear array of receivers. An image can then be formed in the post-processing by coherently combining the receiver outputs to focus at one pixel at the time. The principle of adjusting the array's focus is commonly referred to as beamforming, and involves assigning suitable delays and weights to the array's channels\todo{hmm, doesn't quite feel right...}.
% 
% 
% \subsection{Beamforming}
% 
% \begin{figure}[!t]\centering
% \includegraphics[width=\linewidth]{gfx/beamforming.eps}
% \caption{Beamforming principle. Signal signature is first removed by matched filtering. Then - prior to summation - a suitable set of delays and weights are applied to focus on a pixel of interest.}\label{beamforming}
% \end{figure}
% Let the receiver be an $M$ element uniform linear array, and assume that the signature of the transmitted signal was removed by a matched filter (Fig. \ref{beamforming}). Further assume that the array channels are digitally delayed to create focus at a pixel with azimuth angle $\theta$ and range sample $n$, and let the delayed data from the $m$'th channel be $x_m[\theta,n]$. To simplify notation, the dependence on $\theta$ will be made implicit from now on.
% 
% By definition the beamformer output $z[n]$ can now be expressed as the weighted sum of all the delayed data samples:
% \begin{align}
% z[n] = \w\H[n]\x[n] = \bmat{w_0[n]\\w_1[n]\\\vdots\\w_{M-1}[n]}^H \bmat{x_0[n]\\x_1[n]\\\vdots\\x_{M-1}[n]},\label{z}
% \end{align}
% where $w_m$ is the weight factor assigned to channel $m$. With static weights this would be referred to as the conventional delay-and-sum (DAS) beamformer. Various weighting patterns exists here for trading lateral resolution for improved noise suppression (contrast), but one always ends up with a compromise between the two\todo{don't quite like this one}~\cite{Harris1978}.
% 
% Various adaptive beamformers target this limitation by allowing the weights to change for each pixel to better fit the dynamic nature of the incoming wavefield, the rationale being that the a priori knowledge embedded in the data may be used to improve image quality. The MVDR beamformer is one such method. It finds the set of complex weights that minimizes the beamformer's expected output power, while ensuring unity gain in the look direction~\cite{Capon1969}. This is a convex optimization problem that can be solved using Lagrange multipliers to yield the solution:
% \begin{gather}
% \vec w[n] = \frac{\Ri[n]\1}{\1\T\Ri[n]\1},\label{weights}
% \end{gather}
% where $\1$ is a row vector of ones that represents broadside steering, and $\R=E\{\x[n]\x\H[n]\} \in\mathbb{C}^{M,M}$ is the spatial covariance matrix for the full array. Since $\R$ is unknown, we will estimate it by computing a sample covariance matrix $\eR$. In this computation we will perform some degree of \emph{spatial averaging} to avoid signal cancellation, \emph{temporal averaging} to maintain true speckle statistics, and \emph{diagonal loading} to improve robustness to parameter errors ~\cite{Synnevag2009}. Combined, these steps will also ensure a numberically well conditioned $\eR$\todo{should we mention variance reduction?}.
% 
% If we let $x_l[n]$ represent the datavector from subarray $l$,
% \begin{gather}
% \x_l[n] = \bmat{x_l[n] & x_{l+1}[n] & \dots & x_{l+L-1}[n]}\T,
% \end{gather}
% then a sample covariance matrix formed with temporal and spatial averaging, $\eR_\text{s,t}$, is given as:
% \begin{gather}
% \eR_\text{s,t}[n] =  \frac{1}{N_K N_L} \sumb{l=0}{M-L}\sumb{n'=n-K}{n+K} \x_l[n]\x_l\H[n] \in\mathbb{C}^{L,L},\label{spatialR}
% \end{gather}
% where $N_K = 2K+1$ is the number of temporal samples to perform averaging over, and $N_L = M-L+1$ is the number of subarrays.
% 
% To further improve robustness, we add a fraction $d$ of the total power of $\eR_\text{t,s}[n]$ to its diagonal~\cite{Synnevag2007}:
% \begin{align}
% \eR[n] = \eR_\text{t,s}[n] + \I \frac{d}{L} \tr\{\eR_\text{t,s}[n]\},\label{finalR}
% \end{align}
% where $\I$ is an identity matrix, $\tr\{\cdot\}$ represents the matrix trace operation, and $\tr\{\eR_\text{t,s}[n]\}$ is an estimate of the energy received by this pixel.
% 
% Note how subarray averaging led to a size reduction of $\eR$ from $\mathbb{C}^{L,L}$ to $\mathbb{C}^{M,M}$, and hence will produce an $L$-element weightset when substituted into (\ref{z}). This weightset is then applied to on all the subarrays, prior to computing the beamformer output as in (\ref{z}). Or, equivalently, we may apply the weightset to the sum of all the subarrays:
% \begin{align}
% z[n] = \w\H[n] \sumb{l=0}{M-L} \x_l[n].\label{finalZ}
% \end{align}
% \begin{figure}[!t]\centering
% \includegraphics[width=\linewidth]{gfx/implementation.eps}
% \caption{MVDR beamforming. For one of the total number of pixels in range and azimuth, $N_y$ and $N_x$,\newline
% 1. an $L\times{}L$ sample covariance matrix $\eR$ is computed, \newline
% 2. the term $\eR^{-1}\1$ is found using a linear equation solver,\newline
% 3. and the beamformer output is computed from $z$ from (\ref{finalZ}), where $\w$ is found by substituting $\eR^{-1}\1$ into (\ref{weights}). } \label{mvdr_beamforming}
% \end{figure}
% % \begin{figure}[!t]\centering
% % \includegraphics[width=\linewidth]{gfx/algorithm_structure.eps}
% % \caption{MVDR beamforming. First a spatial covariance matrix is estimated from the delayed data (\ref{spatialR}-\ref{finalR}), then the weights are computed (\ref{weights}) and finally applied to the delayed channel data (\ref{z}).}
% % \label{implementation}
% % \end{figure}
% As summarized in Fig. \ref{mvdr_beamforming}, the MVDR method is applied to each pixel independently, by
% \begin{enumerate}
% \item computing the sample covariance matrix $\eR$ in (\ref{finalR}),
% \item computing $\eRi\1$ in (\ref{weights}), and
% \item computing the beamformer output $z$ in (\ref{finalZ}).
% \end{enumerate}
% % This is illustrated in .
% % \begin{figure}[!t]\centering
% % \includegraphics[width=\linewidth]{gfx/buildR-complex_data-average.eps}
% % \caption{.}\label{complexity}
% % \end{figure}
% 
% \subsection{Computational Complexity}
% 
% When assessing the MVDR beamformer's complexity and mapability to parallel hardware, computing and inverting $\eR$ is by far the most computationally intensive task. If we neglect spatial and temporal averaging then computing $\eR$ is of O($M^2$), and inverting it is of O($M^3$). But if we implement spatial and temporal averaging as in (\ref{spatialR}), then computing $\eR$ becomes of O($N_K N_L L^2$) and the inversion of O($L^3$). Furthermore, assuming rectangular complex notation, we should take into account that a complex multiplication is roughly 3 times as arithmetically intense as a complex addition. So what does it all amount to in the end?
% 
% To investigate this aspect we derived exact complexity formulas for each step in the MVDR process\todo{except partial pivoting, diagonal loading}. Then these were evaluated for the entire range of possible subarray sizes from $L\in[1,M]$, with temporal averaging assessed with values $K\in\{0,1,2\}$ and the number of channels set to $M\in\{8,16,24,32\}$.
% 
% The results are shown in Fig. \ref{mvdr_complexity}. Note how the computation of $\eR$ completely dominates at smaller subarray sizes, that solving $\eRi\1$ only plays a notable role when the element number and subarray sizes are higher, and that the computation of $z$ has a neglible impact. Also notice how temporal averaging comes with a high computational penalty. This is because building $\eR$ is heavy on complex multiplications, and a lot of these are repeated unneccessary.
% 
% \begin{figure}[!t]\centering
% \includegraphics[width=\linewidth]{gfx/mvdr_complexity.eps}
% \caption{Per-pixel computational complexity of the steps in MVDR beamforming (prior to any optimizations). To avoid signal cancellation in an active sonar system we usually set $L<\frac{M}{2}$, in which region the computation of $\eR$ dominates in terms of arithmetic complexity, especially with temporal averaging.}\label{mvdr_complexity}
% \end{figure}
% 
% % Not corrected for fused multiply accumulate
% 
% % Also, the need for partial pivoting has been ignored, as we ensure a numerically well conditioned $\eR$ by applying diagonal loading. Since this is the most common scenario, we will have a thorough look at how it can be optimized.
% 
% % and does not take into that does not  feasibility parallel decomposition  memory consumption.
% 
% 
% 
% \section{Mapping the MVDR to a GPU}
% 
% % \begin{itemize}
% % \item Keep delay step on CPU, as this is a common step in all beamforming
% % \item Increasing matrix size reduces the number of threads per block, since the entire matrix must be in shared memory. 1 matrix 72x72 complex (8 bytes) max per block - 72 threads. same as 207 5x5 matrixes, 5*207=1036 threads. max threads in block 512.
% % \end{itemize}
% 
% % An important feature of MVDR beamforming, and with beamforming in general, is that each pixel can be computed independently and in an identical fashion. Considering that sonar images often end up having millions of pixels, this makes a compelling case for investigating technologies that support massive data level parallelism. The technology that does this best is perhaps Graphics Processing Units (GPUs). These chips contain up to thousands of small computation cores, which combined deliver superior peak floating point performance compared multi-core systems such as CPUs.
% 
% An important feature of MVDR beamforming, as with beamforming in general, is that each pixel can be computed independently and in an identical fashion. Furthermore, a typical sonar image may contain millions of pixels. This represents a level of data parallelism that appears very well suited for a massively parallel architecture such as the GPU.
% 
% To investigate the feasibility of running MVDR on a GPU we decided to map it to a Nvidia GeForce Quadro 6000. This is a high-end Compute Unified Device Architecture (CUDA) enabled GPU based on Nvidia's Fermi architecture. The code was written in Nvidia's ``C for CUDA'' framework. We believe an equally attractive option would be to use a GPU from AMD, and that the code for either of these GPUs could have been written cross-platform framework OpenCL from the Khronos group, but comparing these alternatives are beyond the scope of this study.
% 
% To use Nvidia's own terminology, the Quadro 6000 is comprised of 14 streaming multiprocessors (SMs), each having 32 CUDA cores that execute a common program called a kernel. Combined these cores are able to deliver more than 1\;Tflop/s (appendix \ref{throughput}), but only if the implementation is computationally bound. This can be achieved by balancing the load evenly on all cores, and by trying to avoid that some cores are forced to idle due to a pending data transfer or thread synchronization. As the best way to do this is completely different for each step in the MVDR process, we have designed a different kernel and configuration for each of them. Particular attention will be paid to building $\eR$, since the potential for gaining overall speedups are greatest here\todo{review}.
% 
% %  this assumes that the implementation is solely bound by arithmetic throughput.  r bound by memory bandwidth, arithmetic latency or arithmetic bandwidth.  or  is solely bound by processing power.   These cores need to be kept busy if we are to That make for a total of 448 CUDA cores which we need to keep busy if we want to tap the full potential. 
% 
% \begin{figure}[t]\centering
% \includegraphics[width=\linewidth]{gfx/mvdr_build_R.eps}
% \caption{Step 1: Building $\eR$. This is a visualization of how $\eR$ could be built in a case with $M=5$ sensors, with subarray size $L=3$ and temporal averaging set to $K=1$. Here $\eR_{l}$ is the sample covariance matrix for the $l$'th subarray, and $\eR_s$ is the average of all these. Note that instead of performing the temporal sum last as here, one could take more temporal samples into consideration in the computation of each $r_{ij}$.}\label{mvdr_build_R}
% \end{figure}
% 
% 
% % \todo{Another shitty GPU transition}
% % Before proceeding we should make a quick note on the GPU memory. Note that to obtain the best possible performance, we need to make sure that CUDA threads find the data they need in cache or registers most of the time. Only these two memory types have a high enough bandwidth to promote sustained maximum utilisation of all CUDA cores. Unfortunately, they are a sparse resource. 
% 
% % This makes it apparent that computing one pixel per thread is not a feasible solution. \todo{...}
% 
% % When optimizing an algorithm for speed, it is if of importance to know whether an algorithm is bound by computations or memory. One way to get an approximate idea of this is to compute the ratio of peak arithmetic throughput to peak memory throughput of the in the target platform, and compare this ratio to the ratio of arithmetical operations to memory operations in the algorithm. 
% 
% % follow collective access patterns to memory.
% 
% % At first glance, it might seem like a good idea to compute one pixel per CUDA core. Each pixel is computed using the same instructions, and without any exchange of data. However, this approach is unfeasible due to the limited cache and register memory available to each thread. One could resolve to using the slower GDDR5 memory, but this would largely mitigate the performance gain and defeat the purpose of using a GPU in the first place.
% 
% %Bandwidth becomes crucial, as getting data in to the cores and out again must happen i a frantic pace. %the feeding all these cores with data and harvesting the result
% 
% % To effectively harness the computing potential of a GPU we need to make sure all computing cores stays busy most of the time. 
% 
% % Memory latency can be hidden in two ways. One is to keep a lot of threads queued up on each SM at all times, such that when some threads perform memory operations other threads can be scheduled to run instead. This is referred to as data level parallelism (DLP). Alternatively, or preferrably in addition, the designer may promote instruction level parallelism (ILP) by letting subsequent instructions in a thread be independent. While ILP can be very effective when done correctly, DLP is generally simpler to design for.
% % 
% % In the CUDA framework, the threads on an SM is managed through a software abstraction call a compute block. The block is a 3 dimensional grid where each element correspond to a thread that is scheduled to be executed on the SM at some point.
% 
% % load balancing
% % - distribute load - optimize symmetry
% % - avoid branching
% % 
% % memory latency hiding
% % - make threads use high bandwidth memory
% % - hide latency with dlp, ilp, or both
% % 
% % thread sync hiding
% % - queue blocks up on an sm
% 
% %This is referred to as a single-program, multiple-data (SPMD) architecture\todo{malplaced}.
% 
%  %An SM is a single-program multiple-data (SPMD) processor, meaning that all the CUDA cores on an instruction unit, L1 cache and registers is shared by all the CUDA cores on that SM.  %
% % GPUs, on the other hand, does not attempt to do ILP, but use all available transistors to support massive DLP. This leads to designs such as the nNvidia GeForce GTX 580. 
% % each with 48kB of L1 cache (shared memory) and 128kB of register memory.
% 
% % -bank conflicts when L > 32. not treated.
% 
% 
% \subsection{Computing the Spatial Covariance Matrix, $\hat{\boldsymbol{R}}$}
% 
% % As Fig. \ref{mvdr_complexity} illustrates the computation of $\eR$ 
% 
% As observed in Fig. \ref{mvdr_complexity} a direct computation of $\eR$ by implementing (\ref{spatialR}) and (\ref{finalR}) is the greatest computational burden. We target this issue by first performing \emph{arithmetic reductions}, then aim to get as close to the peak \emph{arithmetic throughput} on the GPU as possible.
% 
% \subsubsection{Arithmetic Reductions}
% 
% In Fig. \ref{mvdr_build_R} we illustrate how to reduce the arithmetic complexity of building $\eR$ in a system with $M=5$ sensors, a subarray size of $L=3$ and temporal averaging is set to $K=1$. Here the spatial sum is carried out before the temporal sum. To reduce arithmetic operations we may
% %
% \begin{enumerate}
% \item exploit the fact that $\eR$ is Hermitian positive semi-definite to compute only one half of it,
% \item avoid redundant multiplications, and
% \item implement the spatial sum iteratively by sliding along the diagonals of $\eR$.
% \end{enumerate}
% %
% To study the effect of these optimization strategies we derived complexity curves for the plain implementation with and without swapped sums, a couple that take symmetry into account, and a couple that iteratively sum up the diagonals of $\eR$. Then these curves were normalized to the complexity curve of the reference implementation in Fig. \ref{mvdr_complexity}. The results, shown in Fig. \ref{mvdr_optimization}, illustrate that taking symmetry into account and implementing iterative summing both lead to considerably reduced arithmetic complexity, especially at lower subarray sizes and when using temporal averaging. 
% 
% We chose the iterative solution based on its overall excellent performance, and because it can be altered to consume very little memory. This is \emph{very} important, as we will see next.
% 
% \begin{table}[b]\centering%\normalsize
% \begin{tabular}[c]{l r@{:}  l}\hline
% \rowcolor{tabBlue}\bf Memory type & \multicolumn{2}{>{\columncolor{tabBlue}}c}{\bf BW$_\text{memory}$/BW$_\text{arithmetic}$} \\\hline
% Global memory &\hspace{30pt} 1 & 30 \\
% Shared memory & 1 & 4 \\
% Registers & $>$3 & 2~\cite{Vasilyy}
% \end{tabular}
% \caption{Nvidia Quadro 6000 bandwidth ratios.}\label{tabbandwidth}
% \end{table}
% 
% \begin{figure}[!t]\centering
% \includegraphics[width=\linewidth]{gfx/buildR-breakdown.eps}
% \caption{Arithmetic optimization of computing $\eR$: Relative reduction in arithmetic complexity (higher is better). All plots are normalized to the complexity curve of the reference implementation in Fig. \ref{mvdr_optimization}.} 
% \label{implementation}
% \end{figure}
% 
% \subsubsection{Arithmetic Throughput}
% 
% When we compute $\eR$, we very rarely perform more than 1-3 floating point operations for every float read or written to memory. Unfortunately, the GPU prefers kernels to be more computationally intensive than this. This can be inferred from Table \ref{tabbandwidth}, where the peak bandwidth of the three types of GPU memory are compared to the peak arithmetic throughput (see appendix \ref{throughput} for derivations). The global memory\todo{if there's space somewhere, introduce memory layout}, in which all data must reside at some point\todo{rephrase}, is at best only able to move one float for every 30 floats processed\todo{well, not entirely true. rephrase} by the CUDA cores. However, peak shared memory performance is only 4 times slower than the arithmetical throughput so we really need to use it. To achieve this we must
% \begin{enumerate}
% \item keep all the data relevant to the pixels being computed in shared memory, while
% \item trying to use it as efficiently as possible.
% \end{enumerate}
%  %Further alleviation is also possible by exposing instruction level parallelism within a thread, as this promotes the use of registers with a bandwidth at least 6 times higher than that of shared memory~\cite{Vasilyy}\todo{so why don't we? rephrase}.
% 
% These challenges are very closely linked. The Quadro 6000 architecture is of compute capability 2.0, which means there are 48\;kB of shared memory (L1 cache) and 128\;kB of registers per SM~\cite{Nvidia2012}. This memory is shared by all active threads on that SM, a number that should be no less than 768~\cite{Nvidia2012a}. This will expose a sufficient level of data parallelism to ensure that memory latency is completely hidden (Little's law). By dividing the shared memory evenly on all 768 threads we find that each thread should store no more than 8 single precision complex numbers in shared memory\todo{introduce memory structure -shared for data shared between threads}, and 24 stored in registers. This should make it apparent why computing a single pixel per thread is a bad idea, and why we need to keep each thread as light on memory consumption as possible.
% 
% \begin{figure}[!t]
% \centering
% \includegraphics[width=\linewidth]{gfx/mvdr_implementation.eps}
% \caption{MVDR implementated on a GPU. The sample covariance matrix is formed by threads running along its diagonals, performing spatial averaging iteratively. Then Nvidia's     ces for each image pixel are computed .}\label{mvdr_implementation}
% \end{figure}
% 
% % What about the GPU implementation? This one was tricky. Balance load, avoid inter-thread dependencies when performing iterative summations, keep thread memory consumption at a minimum, and promote collective access patterns (coalesced reads/writes) to global memory.
% 
% In Fig. \ref{mvdr_implementation} we illustrate how we got around these challenges. First we make each SM compute entire range lines of pixels, then load the subset of $\x$ that all these pixels depend upon into shared memory. This lets us perform temporal averaging without the need for reading additional channel data.
% 
% Second we assign $L$ threads per pixel to traverse the diagonals of $\eR$. On the top row a full computation is carried out, then that entire row is saved back to global memory following a collective access pattern that maximize global memory bandwidth. For subsequent rows the threads move along the diagonals while performing iterative summations; the result from the previous element on the diagonal is updated by adding and removing the correlation coefficients that enters and exits the sum, respectively. To minimize memory consumption, we compute the coefficients again every time we need them. When a thread has finished up a diagonal, we have them wrap around to compute one of the diagonals in the lower triangular of $\eR$. Since $\eR$ is conjugate symmetric, the values in the leftmost column is obtained by a complex conjugate copy of the relevant value in the first row. Combined these steps balances the load evenly on all threads, is almost completely free of arithmetic redundancy, and consumes less 
% memory.
%  
% 
% 
% % Hence, whenever $N_\text{avg}>1$ the sliding window approach will demand less instructions at the expense of a slightly higher memory consumption.
% 
% % However, an advantage with this approach is that the summations can be carried out in any order, and hence this form exhibits great deal of parallelism. Can we optimize the algorithm without sacrificing parallelism?
% 
% 
% % - Order of summations - beamspace capable
% 
% % 
% % In the upcoming results, we have beamformed pre-delayed data. This is a large dataset, and the latency experienced when copying it from the \gls{CPU} side to the \gls{GPU} was significant. However, in a practical system only the channel data should be transferred as the delaying of data can effortlessly be performed by the \gls{GPU}. The remaining latency can be hidden as long as the \gls{GPU} is kept busy, hence the data transfer times will not be reported in the upcoming results.
% % 
% 
% % \begin{itemize}
% % \item Arrange order of summations to reduce data dimensions as early as possible.
% % \item Sliding window in time/space to reduce instructions at the cost of extra memory consumption. Only pays off to do this over subarrays.
% % \end{itemize}
% % 
% % \item Between GPU and CPU: Several concurrent blocks (contexts) per SM.
% % \item Job queue (how does this work)
% % \item Register pressure and shared memory. Thread context must fit in local memories.
% % \item Warp computational intensity should be high to make up for the memory cost.
% 
% 
% % increasing performance hit several walls:
% % \begin{itemize}
% % \item memory couldnt keep track - remedy cache.
% % \item instruction parallelism - complex to identity
% % \item power consumption cubic with frequency
% % GPUs:
% % \begin{itemize}
% % \item lower frequency, less power
% % \item no instruction parallelism
% % \item massive amounts of lightweight cores
% % \end{itemize}\end{itemize}
% 
% \cite{Owens2007} \cite{Lee2010}
% 
% 
% \subsection{Solving $\hat{\boldsymbol{R}}$$\;\!^{-1}$$\mathbf{1}$}
% 
% As intuition might suggest that the matrix product $\eRi\1$ can be carried out by first inverting the matrix $\eR$, then performing the inner product with $\1$. However, one can also solve the linear equation $\hat{\boldsymbol{R}}\boldsymbol{b} = \1$ for $\boldsymbol{b}$, where $\boldsymbol{b} = \eRi\1$. In terms of arithmetic count, and when comparing direct implementations, the latter appears to be the preferred option.
% 
% An important thing to note, however, is that unlike the problems most libraries tries to solve we do not attempt so solve large linear equations or invert large matrices; our matrices are small, but we have a very large number of them. Fortunately Nvidia recently released a highly optimized batched linear equation solver aimed for this particular purpose. It is a Gauss Jordan based and supports partial pivoting and complex numbers. The only downside to using it in our application is that it does not exploit the symmetry properties of our sample covariance matrix. A better choice would be a Cholesky solver, which is designed for Hermitian positive semi-definite matrices and is expected to offer roughly a factor 2 speedup over the Gauss Jordan one.\todo{somewhere a short remark regarding inversion/solver alternatives should be made}
% 
% % \emph{Computing} $\eRi[n]\1$ (from (\ref{weights})). While intuition may suggest that this step is carried out by inverting $\eR$, it is better to solve the linear equation $\R\beta = \1$ for $\beta$ instead, which gives us $\beta = \eRi\1$ directly. We have tested various solvers for this task, both in-house and proprietary implementations, and achieved the best performance by using an unofficial batch solver from nVidia. Inverting $\eR$ is by far the most computationally intensive task, and a key area of focus for further improvements.
% 
% 
% \subsection{Computing $z$}
% 
% The beamformer output $z$ is computed on a per-pixel basis by first computing the MVDR weights $\w$, which are a mere scaled version of $\eRi\1$ (\ref{weights}). Then the sum in (\ref{finalZ}) is found by assigning a group of $L$ threads to respective elements of $x_l$, which then proceed to accumulate these for all $N_L$ subarrays. The resulting data vector is finally multiplied with the weight vector using the same threads, and then a single thread is used to sum these products to obtain $z$\todo{Rakende likegyldig hva vi gjør her.. Kan kuttes.}.
% 
% %  The subarray data can, as here, be reduced to coincide in length with the subarray weights, or the weights can be extended to $M$ in size and applied directly on $\x[n]$ and summed.  
% 
% % \item \emph{Computing the spatial covariance matrix} $\eR[n]$ (\ref{spatialR}-\ref{finalR}). A group of threads are created that slide along the diagonals of $\eR$. In this way we exploited the fact that entries on the diagonals overlap across subarrays and time, and keep the numbers of both data reads and writes at a minimum.
% % %Finished threads can also wrap around and start on diagonals on the lower half triangular. 
% % % In this way we process one row of $\eR$ per kernel iteration, and manage to keep the numbers of both data reads and writes at a minimum.
% 
% % There are essentially two ways to accelerate an algorithm on parallel hardware. One is to identify independent instructions for each thread, and run these in in parallel, and the other is to support the parallel execution of as many threads as possible. This is referred to as instruction level parallelism (ILP) and data level parallelism (DLP), respectively. Most general purpose processors, such as a CPUs, support both DLP by featuring multiple cores with single-data multiple-data (SIMD) capabilities, and ILP by through means such as branch prediction and out-of-order execution.
% 
% % GPUs, on the other hand, does not attempt to do ILP, but use all available transistors to support massive DLP. This leads to designs such as the nNvidia GeForce GTX 580. Using their own terminology, it is comprised of 16 ``streaming'' multiprocessors (SMs), each having having 32 CUDA\todo{make sure CUDA is introduced} cores that execute a common program called a kernel. %An SM is a single-program multiple-data (SPMD) processor, meaning that all the CUDA cores on an instruction unit, L1 cache and registers is shared by all the CUDA cores on that SM.  %
% % each with 48kB of L1 cache (shared memory) and 128kB of register memory.
% 
% 
% % Lots of cores with modest computing abilities, but in return there are hundreds of them. Little memory for each core. 
% 
% % In short, designing for a GPU involves finding the optimal balance between hiding memory latency (high occupancy) and resource utilization.
% % 
% % blocks in grid - high enough so that all SMs have atleast one block to execute. Optimally in the thousands to scale for future devices, and to allow an SM to switch between blocks to keep hardware busy when running synchronizing threads.
% % 
% % threads in block - min 64. 128-256 better starting point.
% 
% 
% % Instead of having a few complex core that can boost massive single thread performance, designers chose to create hundreds of lightweight cores that each is much powerful than a single CPU core, but combined boasts a massive artithmetical performance.
% % 
% % \begin{itemize}
% % \item Lots of cores, little memory for each.
% % \item Few data transfers, lots of computations.
% % \item Data transfers are moved around by hardware that can operate asynchronously from the compute units.
% % \item Memory latency hidden as long as it takes longer to compute than to move data around.
% % \item Tasks can be queued on SM (blocks) and on CPU to hide internal and external memory latencies.
% % \end{itemize}
% % 
% % Vast amount of processing power, which are easily accessible but hard to fully harness. This is because in the pursuit of an ever increasing amount of arithmetical computing cores, the 
% % 
% % Unlike a CPU, a GPU 
% % 
% % If we compare a GPU with a CPU, 
% % 
% % When implementing an algorithm on a GPU, it is important to be aware
% % 
% % People friendly:
% % \begin{itemize}
% % \item More and more compute units, but less memory for each.
% % \item Need to keep each compute unit busy.
% % \item Few data transfers, lots of computations.
% % \item Data transfers are moved around by hardware that can operate asynchronously from the compute units.
% % \item Memory latency hidden as long as it takes longer to compute than to move data around.
% % \item Tasks can be queued on SM (blocks) and on CPU to hide internal and external memory latencies.
% % \end{itemize}
% % 
% % The ``fast'' memory of the GPU is the registers and cache. All threads executed 
% % 
% % 
% % \begin{itemize}
% % \item Hiding memory access.
% % \begin{itemize}
% % \item Between GPU and CPU: Several concurrent blocks (contexts) per SM.
% % \item Job queue (how does this work)
% % \end{itemize}
% % \item Register pressure and shared memory. Thread context must fit in local memories.
% % \item Warp computational intensity should be high to make up for the memory cost.
% % \item Optimized for small L, high M, hardly any time averaging. A typical sonar scenario.
% % \end{itemize}
% 
% 
% 
% % - Split program into equally sized portions - load balancing
% % - Some portions depend on other potions already being carried outer - sequential dependencies
% % - Effort timed in order to finish at the same time - synchronization
% % 
% % - task level parallelism - compute images independent of each other
% % - data level parallelism - compute pixels independent of each other
% 
% % - Compute bound vs. bandwidth bound
% 
% 
% % Shared memory:
% % - Inter-thread communication within a block
% % - Cacke data to reduce redundant access to global memory
% % - User-managed cache (reduce global memory access patterns)
% %
% % 32 banks, each bank 4-byte wide
% 
% 
% 
% 
% \section{Images and Benchmarks}
% 
% % In a practical system it should be moved to the GPU as well. Therefore, in the upcoming benchmarks, CPU computation time and data transfer to the GPU is neglected.
% 
% \begin{figure}[t]
% \centering
% \includegraphics[width=\linewidth]{gfx/plot_holmengraa_L16_Navg1.eps}
% \caption{HISAS sidescan sonar (SSS) image of the shipwreck Holmengraa.}\label{holmengraa}
% \end{figure}
% \begin{figure}[t]
% \centering
% \includegraphics[width=\linewidth]{gfx/benchmark.eps}
% \caption{\protect MVDR benchmarks. Forming a 55\;kpixel image from a $M=32$ channel array.}\label{benchmarks}
% \end{figure}
% \begin{figure}[t]
% \centering
% \includegraphics[width=\linewidth]{gfx/code_assessment.eps}
% \caption{Execution time of an arithmetic-only and a memory-only version of the MVDR code. A 55\;kpx dataset was processed for all $L$ using $K=1$, and the mean execution time was used here. The kernel building $\eR$ can here seen to be memory bound, as the time the kernels spends performing memory transactions are higher than the corresponding time it spends carrying out arithmetical operations. Furthermore, the runtime of the In the code that solves $\eR$ }\label{code_assessment}
% \end{figure}
% 
% To demonstrate the imaging capability of the MVDR beamformer, we have processed experimental datasets from the $M$=32 element Kongsberg Maritime HISAS1030 sonar mounted on a HUGIN AUV~\cite{Hansen2009}. HISAS1030 is a high resolution synthetic aperture sonar with an array length of 1.2\;m and operating frequency of 100\;kHz. To produce the image shown in \Fig{holmengraa} it was operated in sidescan mode. The studied object is the 1500 dwt oil tanker wreck Holmengraa. It is 68\;m long and 9\;m wide, and lies at a slanted seabed at 77\;m depth outside of Horten, Norway. The MVDR images were processed with parameters $L$=16, $K$=3, and 1\% diagonal loading.
% 
% For computational performance we supply benchmarks in \Fig{benchmarks}. On our test system with a quad-core Intel Xeon E5620, 64Gb of RAM and an nVidia Quadro 6000 card, our GPU implementation of MVDR was able to process just over 1\;Mpx/s. A single-thread C implementation that was added as reference was outperformed by around 2 orders of magnitude. However, do not consider this comparison fair because while the GPU design is make efficient use of the L1 cache, similar optimisation work was not carried out on the CPU design. Also note that the computation times are reported without including the data transfer time from CPU to GPU, which account for 2-20\% of the total processing time. This is because we the required data amount to transfer will be reduced by a factor $M$ if we compute channel delays on the GPU instead of the CPU, which assume will be the case in a full beamforming system. Then the transfer time will be way less than the computation time, and can be easily hidden using asynchronous data 
% transfer.
% 
% % 
% % We have tested our \gls{GPU} implementation of the \gls{MVDR} beamformer experimental datasets from the 32 element Kongsberg Maritime HISAS1030 sonar\todo{shouldn't we state that it's SAS?} being attached to the HUGIN \gls{AUV}~\cite{Hansen2009}. To obtain the studied object, the 1500 dwt oil tanker wreck Holmengraa, the sonar was operated in sidescan mode to image (\Fig{hlmengraa}). Holmengraa is roughly 68\;m long and 9\;m wide, and lies on a slanted seabed at 77\;m depth. The image was processed with \gls{MVDR} parameters $L=16$, $N_\text{avg}=3$, and $d=1\%$, which proved to be a reasonable selection for this scenario.
% % 
% % The computational performance of our GPU implementation is displayed in (\Fig{benchmarks}). Processing times are shown for varying subarray lengths $L$ and varying amounts of temporal averaging $N_\text{avg}$. A single-thread C and Matlab implementation is added as reference, although only moderate optimization work was carried out on these. The test system was a quad-core Intel Xeon E5620, with 64Gb of \gls{RAM}, and an nVidia Quadro 6000 card.
% 
% 
% \section{Discussion}
% 
% \Fig{holmengraa} demonstrates the MVDR beamformer's ability to suppress interference and achieve better detail resolution. Compared the DAS beamformer, the ship's edges appear sharper and the shadows less noisy. This is in accordance with previous work.
% %But while \gls{DAS} processed this 1.3\;Mpixel image in milliseconds, our Matlab implementation\todo{never mentioned matlab vefore} of MVDR beamformer needed minutes.
% 
% As Fig. \ref{benchmarks} demonstrates, performing arithmetic reductions and designing for using GPUs can significantly improve upon the problem with  the processing speed. If we look at the total runtimes for a subarray size of $L=16$ and temporal averaging at $K=2$, the GPU implementation is able to process the test image in roughly 40\;ms. This translates to a processing throughput of 1.4\;Mpixels/s. For the same scenario the C implementation clocked in at roughly 3\;s and Matlab at 17\;s, which is 75 and 435 times slower than the GPU implementation, respectively.
% 
% Interesting to note is also that solving $\eRi\1$ remains the main bottleneck in the GPU implementation, and it gets even worse for larger $L$'s. However, this is a general Gauss Jordan solver which does not exploit symmetry properties of $\eR$, so creating e.g.\ a batch based Cholesky solver should improve the runtimes further. Both the C and Matlab implementation in this benchmark take this symmetry into account, which partially explains why the complexity curves of these implementations are different from that of the GPU implementation.
% 
% If the HISAS1030 was attached to a platform moving at 1.8m/s, for which the maximum range is 250\;m~\cite{Hansen2010}, the pulse repetition frequency could be set to 3\;Hz. We could further assume a complex sampling frequency of 40\;kHz to accomodate the  HISAS1030's bandwidth of 30\;kHz, and beamform $M=32$ lateral beams. The throughput required for real-time processing of these images would then be 1\;Mpixels/s. As we have seen, this is something our GPU implementation can handle if $L$ is not too large.%, leaving the \gls{CPU} free to take on other assignments.
% 
% 
% 
% 
% % What may this performance be used for? Let us assume that the 32 element, 1.2\;m long HISAS1030 array is mounted on a platform moving at 1.8\;m/s. Then the pulse repetition frequency (PRF) could be set to 3\;Hz to ensure critical along-track sampling, 
% % To illustrate what this performance might be used for, consider a moving platform travelling at 1.8\;m/s. 
% % If the sonar system is Maximum range 
% % There is a limit to the along-track sampli
% % To avoid along track undersampling, the 
% 
% 
% 
% % HUGIN: 1.8m/s - range 250m \\
% % fs = 100kHz*30\%*$\frac{4}{3}$ = 40kHz \\
% % $N_{range px} = \frac{2\;250m}{1500m/s}40kHz = 13.3kpx$ \\
% % $\times 32$ beams = 426kpx \\\\
% % 
% % PRF = $\frac{1500m/s}{2 250m} = 3Hz$ \\
% % TP = 215px 3Hz = 1.28Mpx/s \\
% % 4 arrays: 1.28Mpx/s * 4 = 5.12Mpx/s
% % 
% % 
% % \begin{itemize}
% % \item Need for speed: HUGIN 4 banks of 32 elements, can be processed faster than the ping repetition rate, with margins to spare.
% % \end{itemize}
% 
% 
% \section{Conclusion}
% 
% The MVDR beamformer is an algorithm capable of producing images with improved detail resolution and contrast compared to conventional DAS beamforming. The downside is its inherent need for robustification and the high compuational complexity associated with estimating and inverting the spatial covariance matrix.
% 
% We have shown that for systems containing up to 32 channels the problem can be largely mitigated by building the covariance matrix in a clever way, and by making use of the massive computational power inherent in modern GPUs. We were able to improve upon the run-time of a moderately optimized C-implementation by roughly 2 orders of magnitude. This performance is sufficient for computing properly sampled full-coverage sectorsscan images from the HISAS1030 sonar in real-time.
% 
% The MVDR maps well to the GPU since the computations involved are independent on a pixel level, and partially also within each pixel. However, we found it challenging to adapt the MVDR implementation to make use of the limited amount of cache available to each GPU computing core, which is a prerequisite for obtaining peak performance.
% 
% 
% % \begin{figure}[t]
% % \centering
% % \includegraphics[width=\linewidth]{gfx/benchmark_tagged.eps}
% % \caption{Realtime requirement for sectorscan imaging.}\label{real-time_sectorscan}
% % \end{figure}
% 
% %%%%%%%%%%%%%%%%%%                              ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
% % DOCUMENT APPENDICES %
% %%%%%%%%%%%%%%%%%%                              ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
% 
% 
% % \titleformat{<command>}[<shape>]{<format}{<label>}{<sep>}{<before>}[<after>]
% 
% % \titleformat{\section}[hang]{\bf}
% % {\thesection.\enspace}%\thesubsubsection}%  {\footnotesize \enspace \emph{Sec.}  }
% %    {0pt}{\MakeUppercase}[]
% % \titlespacing*{\section}{0pt}{2\lineheight}{\lineheight}
% 
% % {-2ex plus -.5ex minus -.2ex}{1.0ex plus .2ex minus .2ex}
% 
% % \renewcommand\section[1]{{\Large\bf\MakeUppercase #1}}
% % \renewcommand\section*[1]{{\Large\bf\MakeUppercase #1}}
% 
% % 
% % 
% % 
% % 
% % 
% % 
% % \section{Introduction}
% % 
% % \todopar{Case:
% % \begin{itemize}
% % \item UFFC(?)
% % \item Ultrasound. Arrays up to 128 channels. Processing power vital. What's the best way to implement 
% % \item Other articles evaluates Capon image quality, and some have shown that speedups can be achieved by reducing the complexity of the traditional Capon algorithm or by implementing specific versions on a GPU.
% % \end{itemize}
% % }
% % 
% % \begin{itemize}
% % \item As briefly as possible, introduce beamforming.
% % \item Adaptive beamformer's potential lies in its ability to suppress interference power
% % \item Why adaptive beamformers struggle in active sonar systems. Correlated noise, robustification kills the adaptive potential. Quite computationally intensive. Constraints must be applied in one way or another - parameters must be tuned.
% % \item The three main ways to implement Capon \cite{Capon1969} is \todo{not entirely true?}
% % \begin{itemize}
% % \item Traditional way: Build $\eR$ and solve $\w = \frac{\Ri\a}{\a\H\Ri\a}$. 
% % \item Iterative methods: Woodbury, Conjugated gradients
% % \item Beamspace
% % \item LCA (or other methods, as reference)
% % \end{itemize}
% % \item Choice of robustification parameters $L,M,\Delta$ largely impacts what the ``optimal'' way of implementing Capon 
% % \item In most imaging applications 
% % Outline:
% % \begin{itemize}
% % \item Speeding up: Build $\eR$ and solve $\w = \frac{\Ri\a}{\a\H\Ri\a}$. 
% % \item Iterative methods: Woodbury (Sherman Morrison), Conjugated gradients - show performance of GPU performance.
% % \item Woodbury: Must take the average over many samples in space and time - lots of (small) outerproducts. Can be used if not a lot of averaging is required.
% % \item Beamspace
% % \item LCA (or other methods, as reference)
% % \end{itemize}
% % \end{itemize}
% % 
% % fs = 100kHz*30\%*$\frac{4}{3}$ = 40kHz \\
% % $N_{range px} = \frac{2\;250m}{1500m/s}40kHz = 13.3kpx$ \\
% % $\times 32$ beams = 426kpx \\\\
% % 
% % PRF = $\frac{1500m/s}{2 250m} = 3Hz$ \\
% % TP = 215px 3Hz = 1.28Mpx/s \\
% % 4 arrays: 1.28Mpx/s * 4 = 5.12Mpx/s
% % 
% % 
% % \section{Methods}
% % 
% % \begin{align}
% % z[n] = \sumb{m=0}{M-1} w_m[n]^*x_m[n-\Delta_m] = \w\H[n]\Xd[n]
% % \end{align}
% % where
% % \begin{align}
% % \w[n] = \bmat{w_0[n]\\w_1[n]\\\vdots\\w_{M-1}[n]} \quad \text{and} \quad\Xd[n] = \bmat{x_0[n]\\x_1[n]\\\vdots\\x_{M-1}[n]}.
% % \end{align}
% % 
% % Minimum variance distortionless resonse \cite{Capon1969}
% % \begin{align}
% % \underset{\w[n]}{\argmin}\; E\{ |z[n]|^2 \} = \underset{\w[n]}{\argmin}\; \w\H[n]\R[n]\w[n], 
% % \end{align}
% % subject to
% % \begin{align}
% % \w\H[n]\a = 1,
% % \end{align}
% % where
% % \begin{align}
% % \R[n] = E\{ \x\x\H \}.
% % \end{align}
% % If $\eR$ is the estimate of $\R$, the solution to the MVDR beamformer is
% % \begin{gather*}
% % \vec w[n] = \frac{\hat{\mat R}\;\!^{-1}[n]\vec a}{\vec a\H\hat{\mat R}\;\!^{-1}[n]\vec a} = \frac{\text{\raisebox{1.9pt}{$\vec\chi$}}[n]}{\vec a\H\text{\raisebox{1.9pt}{$\vec\chi$}}[n]}
% % \end{gather*}
% % A robust estimate of $\eR$ is found by
% % \begin{itemize}
% % \item Spatial averaging to decorrelate noise from signal. \emph{Subarray averaging} used here.
% % \item Temporal averaging to ensure valid speckle statistics.
% % \item \emph{Diagonal loading} to ensure  numerical stability prior to inversion.
% % \item Choice of robustification parameters \cite{Synnevag2007}
% % \end{itemize}
% % 
% 
% % \appendix[Throughput]
% % 
% % {\bf Method complexity (instructions):}\\
% % \begin{tabular}{l l}
% % No optimizations            & $2M^2\;(2\Navg+1)$ \\
% % Banded symmetric            & $L(2M-L+1)\;(2\Navg+1)$ \\
% % Sliding, banded, symmetric  & $1.5L(2M-L+1)$ \\
% % \end{tabular}\\\\
% % 
% % {\bf Number of complex numbers needed per SM when building R:}\\
% % \begin{align*}
% % \left\lfloor\frac{T_b}{L}\right\rfloor(M+L)+2MN
% % \end{align*}
% 
% % Is the method compute bound, memory throughput bound, or memory latency bound?
% % 
% % Take as example the case in Fig \ref{mvdr_complexity} in case where $M=32$, $L=16$ and $K=1$. The figure shows a that 300k floating point operations need to be performed per pixel. Furthermore, the total size of global reads is $(M+L)*8$bytes$ = 96$floats per px, and the total number of writes is $L^2*2$floats = 512 floats, leading to a total of 608 floats. This means that roughly 1000 arithmetic instructions are carried for every 0.6 float transferred from or to global memory.
% % 
% % If an algorithm is neither bandwidth bound or computationally bound, it might be bound by latency. 
% % 
% % latency - arithmetic (20 cycles) and memory (400+ cycles)
% % 
% % Little's law - needed parallelism = latency * throughput
% % arithmetic: 23 cycles * 32ops/SM/cycle = 576 ops/SM parallelism
% % memory: <800 cycles * <144GB/s=125B/cycle = <100kB in the flight
% % 
% % On GF104: must have some ILP to get >0.66 of peak:
% % - 48cores/SM, one instruction is broadcast across 16 cores
% % - need 3 instructions per cycle
% % - but only have 2 warp schedulers
% % - instead, can issue 2 instructions per warp in the same cycle
% % 
% 
% \section*{GPU Throughput}\label{throughput}
% 
% In the context of determining whether an implementation is computationally bound or memory bound, one should first compare the target platform's sustained computational throughput to sustained memory throughput. Let us start with the former.
% 
% The Quadro 6000 has 32 CUDA cores per SM, each operating at a rate of 1\;148\;MHz and being able to perform 2 floating point operations (flop) per clock cycle when multiply-add instructions are used. Theoretical peak arithmetic throughput is then given as:
% \begin{align}
% \text{BW}_\text{arithmetic} &= 2\cdot1\;148\;\text{flop/s/core}\cdot 32\;\text{cores/SM}\cdot 14\;\text{SMs}\nn
% &= 1.03\;\text{Tflop/s}.\label{flops}
% \end{align}
% Now let us compare this to the memory throughput. The ``global'' GDDR5 memory bus is 384\;bit wide, and operates at 3\;Ghz where 2 bits are sent every cycle. Its peak bandwidth is then:
% \begin{align}
% \text{BW}_\text{global} &= \frac{2\cdot 3\;\text{Gbit/s}\cdot384\;\text{bit}}{8\;\text{bit/byte}}\nn
% &= 144\;\text{GB/s}\ (36\;\text{Gfloats/s}).\label{bwglobal}
% \end{align}
% The shared memory, on the other hand, is organized into 32 banks per SM, each being 32\;bit wide and operating at 1\;148\;MHz where 1 bit is sent per cycle. Its peak aggregated bandwidth is then:
% \begin{align}
% \text{BW}_\text{shared} &= \frac{\frac{1\;148}{2}\;\text{Mbit/s}\cdot32\;\text{bit/bank}\cdot 32\;\text{banks/SM}\cdot 14\;\text{SMs}}{8\;\text{bit/byte}}\nn
% &\approx 1.03\;\text{TB/s}\ (257\;\text{Gfloats/s}).\label{bwshared}
% \end{align}
% The bandwidths are compared in Tab. \ref{tabbandwidth}. Note that even when using shared memory at least 4 floating point operations must be carried out per float transferred to the CUDA cores, otherwise the algorithm will be memory bound and the peak arithmetic throughput can not be reached. %In such cases instruction level parallelism can be exploited to move data from shared memory into the even faster register memory~\cite{Vasilyy}, but this is outside the scope of this article.
% 
% 
% % 
% % 
% % % \ \\
% % % \IEEEPARstart{T}{o} form images from a modern phased array sonar system the received wavefield is usually recorded, and then postprocessed by a digital beamformer. The beamformer applies delays and weights to the sensor channels, the beamformer adjusts the arrays spatial response to focus at one pixel at a time.  such that signals emanating from regions of interest add constructively, while ensuring that noise and interference from other angles do not. 
% % % 
% % % The imaging capabilities of a modern phased array sonar system depend on physical attributes such element response and array geometry, the transmitted signal, as well as the beamforming method being used on transmission and reception. Beamforming is the concept of applying delays and weights to the sensors channels to steer the arrays response to points of interest. 
% % 
% % % 
% % % 
% % % Outline:
% % % \begin{itemize}
% % % \item Capon's resonse when applying robustification
% % % \item Choice of window functions makes little difference.
% % % \item Steering and mainlobewidths have outer bounds.
% % % \item Beamspace?
% % % \item Chosen window plots - what may they tell us? Variance intensity values when using various windows.
% % % \item Assymmetric windows?
% % % \end{itemize}
% % % 
% % % 
% % % \begin{align}
% % % z[n] = \sumb{m=0}{M-1} w_m[n]^*x_m[n-\Delta_m] = \w\H[n]\x[n-\Delta_m]
% % % \end{align}
% % % 
% % % 
% % % \section{Methods}
% % % 
% % % 
% % 
% % %\begin{figure}[!t]
% % %\centering
% % %\includegraphics[width=2.5in]{myfigure}
% % % where an .eps filename suffix will be assumed under latex, 
% % % and a .pdf suffix will be assumed for pdflatex; or what has been declared
% % % via \DeclareGraphicsExtensions.
% % %\caption{Simulation Results}
% % %\label{fig_sim}
% % %\end{figure}
% %

  \section*{Acknowledgments}
\else
  \section*{Acknowledgment}
\fi

The authors would like to thank Kongsberg Maritime and the Norwegian Defence Research Establishment (FFI) for providing experimental data, and thank Nvidia for providing support for running the batched linear equation solver as well as for granting us a testdrive at the Boston HPC center.

% Can use something like this to put references on a page
% by themselves when using endfloat and the captionsoff option.
\ifCLASSOPTIONcaptionsoff
  \newpage
\fi

\bibliographystyle{IEEEtran}
\bibliography{../../Library/library}

\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{gfx/photos/jo_inge.eps}}]{Jo Inge Buskenes}
received the B.Sc. degree in electrical engineering from Gj\o{}vik College University, Norway, in 2007, and the M.Sc. degree in instrumentation for particle physics from the University of Oslo, Norway, in 2010. He is currently pursuing the Ph.D. degree in image reconstruction and technology at the University of Oslo.

His industry experience includes the European Organization for Nuclear Research (CERN), Geneva, Switzerland (2007-2008), and the Norwegian Defence Research Establishment, Kjeller, Norway (2009). He has lectured in digital signal processing at the Gj\o{}vik College University (2009), and at the University of Oslo (2010-2012). His research interests include adaptive beamforming, digital image reconstruction, high performance computing, intelligent detector design and open source software.
\end{IEEEbiography}

\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{gfx/photos/jon_petter.eps}}]{Jon Petter \AA{}sen}
(S'12) was born in Porsgrunn, Norway in 1986. He received the B.Sc. and M.Sc. degree in computer science from the University of Oslo, Norway, in 2010. He is currently pursuing his Ph.D. degree in medical ultrasound technology at the Norwegian University of Science and Technology (NTNU) Medical Imaging Lab (MI-Lab), Trondheim, Norway. His research interests include adaptive ultrasound processing techniques and acceleration of ultrasound algorithms using Graphics Processing Units (GPUs). 
\end{IEEEbiography}

\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{gfx/photos/carl-inge.eps}}]{Carl-Inge Colombo Nilsen}
(S'06-M'10) received the M.Sc. and Ph.d. degrees in computer science from the University of Oslo, Norway, in 2005 and 2010. He is currently working at the University of Oslo as a postdoctoral research fellow. His research interests include signal and array processing for ultrasound imaging and other acoustic applications.
\end{IEEEbiography}

\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{gfx/photos/andreas.eps}}]{Andreas Austeng}
was born in Oslo, Norway, in 1970. He received the M.Sc. degree in physics in 1996 and the Ph.D. degree in computer science in 2001, both from the University of Oslo. Since 2001, he has been working at the Department of Informatics, University of Oslo, first as a postdoctoral research fellow and currently as an associate professor. His research interests include signal and array processing for acoustical imaging.
\end{IEEEbiography}

% 
% IMAGE METRICS
%
% - Point spread function (res via main lobe width, SAS gain via PDF height)
% - Constrast measures
%
% ACR = cL/2
% C = (s+n)/n ratio
% - 
% WHO's needing processing power
% - Centre for Maritime Research and Experimentation

\vfill

\end{document}

### 2019-10-31 HowtoHowtoCapon/submission/submit2/howto_howto_capon.tex
2019-10-30

--- 2019-10-31 HowtoHowtoCapon/submission/submit1/howto_howto_capon.tex
@@
 
-\author{Jo~Inge~Buskenes, %
-        Jon~Petter~\AA{}sen, %
-        Carl-Inge~Colombo~Nilsen, %
-        Andreas~Austeng%
+\author{Jo~Inge~Buskenes,~\IEEEmembership{Student Member,~IEEE,} %
+        Jon~Petter~\AA{}sen,~\IEEEmembership{Student Member,~IEEE,} %
+        Carl-Inge~Colombo~Nilsen,~\IEEEmembership{Member,~IEEE,} %
+        Andreas~Austeng,~\IEEEmembership{Member,~IEEE}%
 \thanks{J. I. Buskenes, C.-I. C. Nilsen and A. Austeng are with the Department of Informatics, University of Oslo, Norway.}%
@@
 We point out important hardware limitations for these devices, and assess the design in terms of how efficiently it is able to use the GPU's resources.
-On a quad-core Intel Xeon system with a high-end Nvidia GPU, our GPU implementation renders more than 1\;Mpx/s.
-This represents a speedup of more than two orders of magnitude over our initial single threaded implementation in C.
-This enables real-time processing of sonar data, and makes the MVDR a viable alternative to conventional methods in practical systems.
+On a quad-core Intel Xeon system with a high-end Nvidia GPU, our GPU implementation renders more than a million pixels per second (1\;MP/s).
+Compared to our initial CPU implementation the optimizations described herein led to a speedup of more than 2 orders of magnitude, or an expected 5-10 times improvement had the CPU received similar optimization effort.
+This throughput enables real-time processing of sonar data, and makes the MVDR a viable alternative to conventional methods in practical systems.
 % If the HISAS1030 was attached to a platform traveling at 1.8m/s, creating full-coverage sectorscan images in real-time would be possible with this performance.
 
-% We benchmarked optimized implementations of the MV beamformer in Matlab and C, running on a quad-core Intel Xeon E5620, and in CUDA running on a high-end nVidia card.
+% We benchmarked optimized implementations of the MV beamformer in MATLAB and C, running on a quad-core Intel Xeon E5620, and in CUDA running on a high-end nVidia card.
 The beamformers operated on data from the 32x4 element Kongsberg Martime HISAS1030 sonar, using a conservative set of robustification parameters.
@@
 % -  
-To ensure MVDR robustness, the litterature suggests combining means such as temporal and spatial averaging, and regularization~\cite{VanTrees2002,Kailath1985}.
+To ensure MVDR robustness, the literature suggests combining means such as temporal and spatial averaging, and regularization~\cite{VanTrees2002,Kailath1985}.
 The complexity issue, on the other hand, can be handled by introducing well-founded approximations.
@@
 Our implementation is not restricted in this manner.
-In related work, we investigated using the same MVDR GPU implementation in cardiac ultrasound imaging~\cite{Asen2012,Asen2013}, and in active sonar imaging we have compared the performance of this implementation to comparable adaptive methods~\cite{Buskenes2013a}.
+
+The implementation is adapted to a well sampled 100\;kHz wideband active sonar system operating in the near field.
+The system has a range resolution comparable to the sample period so our MVDR implementation runs near single snapshot mode.
+To deal with this and avoid signal cancellation we compute and average the covariance estimates from overlapping subarrays~\cite{Kailath1985}.
+This may resemble techniques such as "subarray MVDR"~\cite{Chapman1976}, but unlike that technique we are not using subarrays to reduce complexity but to make MVDR in an active system possible.
+In related work, we investigated using the same MVDR GPU implementation in cardiac ultrasound imaging~\cite{Asen2012,Asen2013}, and in active sonar imaging we have compared the performance of this implementation to comparable adaptive methods~\cite{Buskenes2013}.
 
@@
 where $\1$ is a row vector of ones that represents broadside steering, and $\R=E\{\x[n]\x\H[n]\} \in\mathbb{C}^{M,M}$ is the spatial covariance matrix for the full array.
+Since we pre-steer our data to every pixel in the image we will always look to broadside, but (\ref{weights}) can be easily generalized by substituting $\1$ with a suitable steering vector.
 Since $\R$ is unknown, we estimate it by computing a sample covariance matrix $\eR$.
-In this computation we will perform some degree of \emph{spatial averaging} to avoid signal cancellation, \emph{temporal averaging} to maintain true speckle statistics~\cite{Synnevag2009}, and \emph{diagonal loading} to improve robustness to parameter errors ~\cite{Cox1987,Maksym1979}.
+In this computation we perform some degree of:
+\begin{itemize}
+\item \emph{spatial averaging} to avoid signal cancellation by decorrelating coherent echoes~\cite{Kailath1985},
+\item \emph{temporal averaging} over an interval comparable to the pulse length (1-5 samples) to maintain true speckle statistics~\cite{Synnevag2009}, and
+\item \emph{diagonal loading} to improve robustness to parameter errors~\cite{Cox1987,Maksym1979}.
+\end{itemize}%
+%
+% will perform some degree of \emph{spatial averaging} to avoid signal cancellation caused by coherent echoes~\cite{Kailath1985}, \emph{temporal averaging} to maintain true speckle statistics~\cite{Synnevag2009}, and \emph{diagonal loading} to improve robustness to parameter errors ~\cite{Cox1987,Maksym1979}.
+%
 Combined, these steps will also ensure a numerically well conditioned $\eR$.
@@
 \begin{gather}
-\breve{\R}[n] =  \frac{1}{N_K N_L} \sumb{l=0}{M-L}\sumb{n'=n-K}{n+K} \x_l[n]\x_l\H[n] \in\mathbb{C}^{L,L},\label{spatialR}
+\breve{\R}[n] =  \frac{1}{N_K N_L} \sumb{l=0}{M-L}\sumb{n'=n-K}{n+K} \x_l[n']\x_l\H[n'] \in\mathbb{C}^{L,L},\label{spatialR}
 \end{gather}
@@
 
-We decided to investigate the feasibility of running MVDR on a GPU by mapping it to a Nvidia GeForce Quadro 6000.
-This is a high-end Compute Unified Device Architecture (CUDA) enabled GPU based on Nvidia's Fermi architecture.
-The code was written in Nvidia's ``C for CUDA'' framework.
-We believe that GPUs from AMD, or the cross-platform OpenCL framework from the Khronos group, are equally attractive alternatives, but comparing them is beyond the scope of this study.
+We decided to investigate the feasibility of running MVDR on a GPU by mapping it to a Nvidia GeForce Quadro 6000, which performance-wise is or par with or perhaps slightly inferior to the Geforce 480.
+This is a high-end Compute Unified Device Architecture (CUDA) enabled GPU based on Nvidia's Fermi architecture, which as now been superseded by the Kepler architecture.
+The code was written in Nvidia's ``C for CUDA'' framework, which adds a few extra features to the C language to specify how to distribute work across a large number of threads.
+It would have been possible to use GPUs from AMD and the cross-platform OpenCL framework from the Khronos group instead, but CUDA has a batch linear equation solver~\cite{Nvidia} that no OpenCL library provides.
+As will be discussed later, this is a key component in our design.
 
@@
 \ref{mvdr_complexity}, a direct computation of $\eR$ by implementing (\ref{spatialR}) and (\ref{finalR}) is the greatest computational burden.
-We target this issue by first performing \emph{arithmetic reductions}, then aim to get as close to the peak \emph{arithmetic throughput} on the GPU as possible.
-
-\subsubsection{Arithmetic Reductions}
+We target this issue by first \emph{minimizing arithmetic operations}, then aim to get as close to the peak \emph{arithmetic throughput} on the GPU as possible.
+
+\subsubsection{Minimizing Arithmetic Operations}
 
@@
 \item avoid redundant multiplications, and
-\item implement the spatial sum iteratively by sliding along the diagonals of $\eR$.
+\item avoid redundant summations by first computing the upper row of $\eR$ and then add/subtract to these iteratively to find the diagonals of $\eR$.
 \end{enumerate}
@@
 Exploiting symmetry and performing iterative summing (step 1 and 3) is always desirable, but avoiding redundant multiplications (step 2) comes at the cost of increased memory consumption.
-The effect of adding all the optimizations, or all but saving multiplications, is shown in Fig.
-\ref{mvdr_complexity_optimized}.
+When all optimizations are applied we call it a "minimum instructions" solution, and when step 2 is left out we call it a "minimum instructions" solution.
+Fig.
+\ref{mvdr_complexity_optimized} compares the two solutions, and formulas may be found in appendix \ref{mvdr_formulas}.
 Here the complexity curves are relative to the reference implementation in Fig.
@@
 This allows spatial averaging to be implemented in a computationally efficient manner and minimizes inter-thread communication.
-\emph{Step~2}: $\eRi\1$ is computed using the heavily optimized batched linear equation solver from Nvidia.
-\emph{Step~3}: The beamformer output $z$ is computed in a straight forward fashion by $L$ threads that first sum the subarrays and then apply the MVDR weighting function.
+\emph{Step~2}: $\eRi\1$ is computed using the heavily optimized batched linear equation solver from Nvidia~\cite{Nvidia}.
+\emph{Step~3}: The beamformer output $z$ is computed in a straight forward fashion using $L$ threads that first sum the subarrays and then apply the MVDR weighting function.
 A single thread finally sum all the channels up.}\label{mvdr_implementation}
@@
 When we compute $\eR$, we very rarely perform more than 1-3 floating point operations for every float read or written to memory.
-Unfortunately, the GPU prefers kernels to be more computationally intensive than this.
+Unfortunately, the GPU prefers kernels to be more arithmetically intensive than this.
 This can be inferred from Table \ref{throughputs}, where the peak bandwidth of the three types of GPU memory are compared to the peak arithmetic throughput (see appendix \ref{throughput} for derivations).
-Global memory (RAM), in which all data must reside at some point, is at best only able to move one float for every 30 floats processed by the CUDA cores.
+Global memory (RAM), in which all data must reside at some point, is at best only able to move one float for every 30 floats processed by the CUDA cores, and potentially much worse if care is not taken to use "coalesced" memory access patterns (that avoid memory bank conflicts).
 This is why the usage of global memory should be minimized.
@@
 \item the data needed to compute a pixel can fit in shared memory, while
-\item the access patterns we use promotes maximum bandwidth.
+\item the access patterns we use promote maximum bandwidth.
 \end{enumerate}
@@
 In Fig.
-\ref{mvdr_implementation} we illustrate how we got around these challenges.
+\ref{mvdr_implementation} we illustrate how we get around these challenges.
 First we make each SM compute entire range lines of pixels, then load the subset of $\x$ that all these pixels depend upon into shared memory.
@@
 The image was processed with $M=32$, $L=16$, $K=1$ and $d=1\%$.
-Prior to display the image was linearly upinterpolated by a factor 2 in azimuth making its total size 1.46\;Mpx.
+Prior to display the image was linearly upinterpolated by a factor 2 in azimuth making its total size 1.46 megapixels (MP).
 Note how MVDR improves edge definition and reduces noise in shadow regions.}\label{holmengraa}
@@
 Second, we assign $L$ threads per pixel to traverse the diagonals of $\eR$.
-On the top row a full computation is carried out, then that row is saved back to global memory following a collective access pattern that maximizes global memory bandwidth.
+On the top row a full computation is carried out, then that row is saved back to global memory following a coalesced access pattern that maximizes global memory bandwidth.
 For subsequent rows the threads move along the diagonals while performing iterative summations; the result from the previous element on the diagonal is updated by adding and removing the correlation coefficients that enters and exits the sum, respectively.
@@
 Since $\eR$ is conjugate symmetric, the values in the leftmost column is obtained by a complex conjugate copy of the relevant value in the first row.
+This is slightly non-optimal as it causes divergent branching, but is only needed because the solver is Gauss Jordan based and thus require a full $\eR$.
 Combined these steps balance the load evenly on all threads, is almost completely free of arithmetic redundancy, and consumes less memory.
@@
 Fortunately Nvidia recently released a highly optimized batched linear equation solver tailored for this particular task.
-It is a Gauss Jordan based with support for partial pivoting and complex numbers.
+It is Gauss Jordan based and supports partial pivoting and complex numbers.
 The only downside to using it in our application is that it does not exploit the Hermitian property of our sample covariance matrix.
@@
 \caption{Execution time of an arithmetic-only and a memory-only version of the MVDR code.
-A dataset from an $M=32$ array was processed for all $L$ using $K=1$, and the mean execution time for a 1\;Mpx image was used here.
+A dataset from an $M=32$ array was processed for all $L$ using $K=1$, and the mean execution time for a 1\;MP image was used here.
 From this plot we can infer that the kernel building $\eR$ is memory bound, as the time the kernel spends performing memory transactions is higher than the corresponding time it spends carrying out arithmetical operations.
@@
 
-The beamformer output $z$ is computed on a per-pixel basis by first computing the MVDR weights $\w$, which are a mere scaled version of $\eRi\1$ (\ref{weights}).
+The beamformer output $z$ is computed on a per-pixel basis by first computing the MVDR weights $\w$, which are merely scaled versions of $\eRi\1$ (\ref{weights}).
 Then the sum in (\ref{finalZ}) is found by assigning a group of $L$ threads to respective elements of $x_l$, which then proceed to accumulate these for all $N_L$ subarrays.
 The resulting data vector is finally multiplied with the weight vector using the same threads, and then a single thread is used to sum these products to obtain $z$.
+
+\subsection{Summary of GPU implementation}
+
+The GPU implementation can be summarized as follows:
+\begin{itemize}
+\item Pixels are processed in batches.
+\item Multiple threads per pixel.
+\item $\eR$ is computed directly from raw data to minimize memory footprint.
+\item Inversion step is performed by Nvidias batch linear equation solver.
+\item Design makes efficient use of cache and data transfers are kept to a minimum.
+\end{itemize}
+
+Compared to our CPU implementations, where:
+\begin{itemize}
+\item Pixels are processed one by one.
+\item Single thread per pixel.
+\item Computation of $\eR$ is optimized for minumum number of instructions, as in \ref{mvdr_build_R}. Only upper half of $\eR$ is computed (see next point).
+\item Inversion step carried out by a custom inline Cholesky based solver that only requires the upper triangular of $eR$.
+\item Design has generally not been optimized for efficient use of memory.
+\end{itemize}
 
@@
 The image was processed with $M=32$, $L=16$, $K=1$ and $d=1\%$.
-Prior to display the image was linearly upinterpolated by a factor 2 in azimuth making its total size 1.46\;Mpx.
+Prior to display the image was linearly upinterpolated by a factor 2 in azimuth making its total size 1.46\;MP.
 Note how MVDR improves edge definition and reduces noise in shadow regions.}\label{holmengraa}
@@
 \caption{\protect MVDR benchmarks.
-A 1\;Mpx image from a $M=32$ channel array was processed for all $L$, and for $K\in\{0,1,2\}$.\newline \emph{Top:} The time the GPU spent on building $\eR$, solving $\eRi\1$, and computing $z$.
+A 1\;MP image from a $M=32$ channel array was processed for all $L$, and for $K\in\{0,1,2\}$.\newline \emph{Top:} The time the GPU spent on building $\eR$, solving $\eRi\1$, and computing $z$.
 Note the major speedup of building $\eR$ when compared to the complexity plot in Fig.
 \ref{mvdr_complexity}.\newline
-\emph{Bottom:} Compared to a reference Matlab and single thread C implementation running on a CPU the GPU offered a speedup of 2-3 orders of magnitude, but these numbers are somewhat misleading.
+\emph{Bottom:} Compared to a reference MATLAB and single thread C implementation running on a CPU the GPU offered a speedup of 2-3 orders of magnitude, but these numbers are somewhat misleading.
 If the C implementation was properly optimized we expect the GPU to be no more than a factor 5-10 faster, even if its theoretical peak performance is ~20 times higher than that of the CPU.}\label{benchmarks}
@@
 \caption{Execution time of an arithmetic-only and a memory-only version of the MVDR code.
-A dataset from an $M=32$ array was processed for all $L$ using $K=1$, and the mean execution time for a 1\;Mpx image was used here.
+A dataset from an $M=32$ array was processed for all $L$ using $K=1$, and the mean execution time for a 1 megapixel (MP) image was used here.
 From this plot we can infer that the kernel building $\eR$ is memory bound, as the time the kernel spends performing memory transactions is higher than the corresponding time it spends carrying out arithmetical operations.
@@
 An estimate of the number of floating point operations per second (Flop/s), found by dividing the theoretical complexity curves by actual run-times.
-This is a cruel measure as it does not include any other instructions than the actual arithmetic operations in the MVDR computation.}\label{code_assess_flops}
+This is a crude measure as it does not include any other instructions than the actual arithmetic operations in the MVDR computation.}\label{code_assess_flops}
 \end{figure}
@@
 To demonstrate the imaging capability of the MVDR beamformer, we have processed experimental datasets from the $M=32$ element Kongsberg Maritime HISAS1030 sonar mounted on a HUGIN AUV~\cite{Hansen2009}.
-HISAS1030 is a high resolution synthetic aperture sonar with an array length of 1.2\;m and operating frequency of 100\;kHz.
-To produce the image shown in \Fig{holmengraa} it was operated in sidescan mode.
+HISAS1030 is a high resolution synthetic aperture sonar with an array length of 1.2\;m, operating frequency of 100\;kHz, and bandwidth of 30\;kHz.
+The element size and spacing is 2.5$\lambda$ and the opening angle is 25$^\circ$.
+To produce the image shown in \Fig{holmengraa} the sonar was operated in sidescan mode.
 The studied object is the 1500 dwt oil tanker wreck Holmengraa.
 It is 68\;m long and 9\;m wide, and lies at a slanted seabed at 77\;m depth outside of Horten, Norway.
-The 1\;Mpx MVDR image were here processed with parameters $L=16$, $K=1$, and $d=1\%$.
+The 1\;MP MVDR image were here processed with parameters $L=16$, $K=1$, and $d=1\%$.
 
 The computational performance of our implementation was first assessed on a test system with a quad-core Intel Xeon E5620, 64\;GB of RAM and an Nvidia Quadro 6000 card.
-The results were obtained by processing a 1\;Mpx image from the data from a 32 channel array, for all subarray sizes $L$, and for $K\in\{0,1,2\}$.
+The results were obtained by processing a 1\;MP image from the data from a 32 channel array, for all subarray sizes $L$, and for $K\in\{0,1,2\}$.
 Run-time measurements are shown in \Fig{benchmarks}, the run-times of memory-only and arithmetic-only GPU kernels are depicted in \Fig{code_assessment}, and an estimate of computation efficiency is presented in \Fig{code_assess_flops}.
@@
 
-As seen in the run-time comparisons presented in \Fig{benchmarks}, the GPU method outperforms the CPU implementations by 2-3 orders of magnitude in terms of processing speed.
+As seen in the run-time comparisons presented in \Fig{benchmarks}, the GPU method is now 2-3 orders of magnitude faster than the C implemtation we started out with.
 The bottleneck in the final design is now the inversion step, which is typically 5 times slower than the build step.
-In most cases the processing speed of the Quadro 6000 is above 1\;Mpx/s, and this was only improved by a factor 1.5-2 when run on the new K20 GPU.
+In most cases the processing speed of the Quadro 6000 is above 1\;MP/s, and this was only improved by a factor 1.5-2 when run on the new K20 GPU.
 The benchmarks of our memory-only and arithmetic-only kernels show that the kernels spend roughly the same time on both these tasks, so optimising only one of these further will have marginal effect.
@@
 Excluded from these benchmarks is the data transfer time from CPU to GPU, which account for 2-20\% of the total processing time.
-This is to limit the scope of the article, but will need to be taken into consideration in an implementation of a full sonar system.
-
+We believe it makes little sense to include them since it keeps getting easier perform these data transfers in parallel by offloading the task to the Direct Memory Access (DMA) controllers present on modern GPUs.
+Furthermore, the data rates in active sonar are relatively low compared to the bandwidth available for these transfers.
 % 
@@
 Similar performance was obtained with arbitrary combinations of $L\in\{12,16,24\}$, $K\in\{1,2,3\}$ and $d\in[0.01,0.05]$, and no adjustments had to be made when processing other parts of the scene.
-%But while \gls{DAS} processed this 1.3\;Mpixel image in milliseconds, our Matlab implementation\todo{never mentioned matlab vefore} of MVDR beamformer needed minutes.
-
-As observed in \Fig{benchmarks} the combination of performing arithmetic reductions and implementing the MVDR on a GPU lead to a significant improvement in processing speed.
-Compared to a Matlab and single thread C implementation a speedup of 2-3 orders of magnitude was achieved, but we do not consider this comparison fair.
-In fact, the theoretical peak throughput of this GPU is roughly 20 times higher than that the CPU in question, meaning the potential of the CPUs was far from reached in our initial implementations.
-This is mainly because the entire GPU design is focused on efficient use of fast memory cache, while no similar optimization work was carried out on the CPU implementations.
+%But while \gls{DAS} processed this 1.3\;Mpixel image in milliseconds, our MATLAB implementation\todo{never mentioned matlab vefore} of MVDR beamformer needed minutes.
+
+As observed in \Fig{benchmarks} the combination of minimizing arithmetic operations and implementing the MVDR on a GPU lead to a significant improvement in processing speed.
+Compared to a MATLAB and single thread C implementation an order 2 and 3 speedup was achieved, respectively.
+Note that we do not consider this comparison fair.
+In fact, the theoretical peak throughput of this GPU is roughly 20 times higher than that the CPU in question, meaning the potential of the CPU was far from reached in our initial implementations.
+The key reason for this is that the GPU design process pixels in batches, which allows us to minimize data transfers and maximize the use of fast memory cache.
+In other aspects the designs are similar, both CPU implementations compute $\eR$ efficiently, and they make use of either an optimized custom Cholesky based solver or one from the Intel Math Kernel Library (MKL).
+Unfortunately we had little time to write a custom batch based solver and covariance builder for the CPU, but we expect the speed difference would be 5-10 times if we had compared two such designs.
 
@@
 reduced the complexity from O($N_KN_LL^2$) to less than O($L^2$).
-Another reason for the run-time to appear more like O($L$) is likely the GPU, since our design makes better use of the GPU's resources when there is more processing involved per pixel.
+Another reason for the run-time to appear more like O($L$) is likely that our design makes better use of the GPU's resources when there is more processing involved per pixel.
 The inversion step, on the other hand, gains less from being implemented on a GPU.
-This is because its nature is less data parallel, and the back substitution involved in its computation is more of a sequential nature.
+This is because its nature is less data parallel.
+In particular, the back substitution step involved in its computation is mostly sequential.
 This difference be can observed in \Fig{mvdr_complexity}.
-Alternative solvers, such as one based on Cholesky decomposition that
-exploits the Hermitian property of $\eR$, can in theory reduce complexity by a factor ~2, but we question whether this result can be obtained in practice using a GPU.
-The CPU implementations perform Cholesky based inversion, but this does not explain why the C and Matlab implementation have a total run-time that is linearly dependent on $L$.
-We find this observation odd, but believe it to be accurate.
+Alternative solvers, such as one based on Cholesky decomposition that exploits the Hermitian property of $\eR$, can in theory reduce complexity by a factor two, but we question whether this result can be obtained in practice using a GPU.
+The CPU implementations perform Cholesky based inversion, but this does not explain why the C and MATLAB implementation have a total run-time that is linearly dependent on $L$.
+We believe this is because these implementations are dominated by the calculation of the covariance matrix and data movement, and not by the inversion step.
 
@@
 Since the GPU hardware can carry out memory and arithmetic operations concurrently and independently, this gap indicates that the GPU sometimes does ``nothing''.
-In our case this is due to synchronization hold-
-ups when performing temporal averaging, which is carried out in a sequential manner.
+In our case this is due to synchronization hold-ups when performing temporal averaging, which is carried out in a sequential manner.
 
@@
 Although the Kepler card peaks at 3.5-4\;Tflop/s, its shared memory bandwidth is only approximately twice that of the Quadro 6000.
+However, it is likely that the Kepler card would perform better had we optimized our design for it.
+For scientific computing the Kepler boards are considered more attractive, with the introduction of features such as kernels calling kernels, more registers, more shared memory, and better mechanisms for hiding or removing data transfers.
 
@@
 a batch based Cholesky solver should improve the runtimes further.
-Both the C and Matlab implementation in this benchmark take this symmetry into account, which partially explains why the complexity curves of these implementations are different from that of the GPU implementation.
+Both the C and MATLAB implementation in this benchmark take this symmetry into account, which partially explains why the complexity curves of these implementations are different from that of the GPU implementation.
 
@@
 We could further assume a complex sampling frequency of 40\;kHz to accomodate the  HISAS1030's bandwidth of 30\;kHz, and beamform $M=32$ lateral beams.
-The throughput required for real-time processing of these images would then be 1\;Mpx/s.
+The throughput required for real-time processing of these images would then be 1\;MP/s.
 As we have seen, this is something our GPU implementation can handle if $L$ is not too large.\todo{droppe denne kanskje}%, leaving the \gls{CPU} free to take on other assignments.
@@
 % PRF = $\frac{1500m/s}{2 250m} = 3Hz$ \\
-% TP = 215px 3Hz = 1.28Mpx/s \\
-% 4 arrays: 1.28Mpx/s * 4 = 5.12Mpx/s
+% TP = 215px 3Hz = 1.28MP/s \\
+% 4 arrays: 1.28MP/s * 4 = 5.12MP/s
 % 
@@
 We were able to improve upon the run-time of a single thread C-implementation by roughly two orders of magnitude.
-For most choices of parameters the GPU was able to create images at ~1 Mpx/s at an average data processing rate of ~40\;GFlop/s.
+For most choices of parameters the GPU was able to create images at $\sim$1\;MP/s at an average data processing rate of ~40\;GFlop/s.
 This is less than 5\% of the peak performance of the GPU, but we believe it to be near optimal given the constraints of memory bandwidth and the sequential nature of some parts of the MVDR implementation.
@@
 
-\appendix
+\appendices
 
@@
 
-% \subsection{Iterative methods}
-% 
-% \begin{itemize}
-% \item Woodbury
-% \item Conjugated gradients
-% \end{itemize}
-% 
-% 
-% \subsection{Beamspace}
-% 
-% \begin{itemize}
-% \item Some minimum of background, and good references.
-% \end{itemize}
-% 
-% 
-% \section{Results/Discussion}
-% 
-% 
-% 
-% 
-% 
-% \begin{itemize}
-% \item Compare computational complexity vs. memory constraints for the various methods. Supporting plots.
-% \item Relate this to whether the favourable architecture is a CPU or a GPU \todo{check ``when to gpu article''}
-% \item Run benchmarks on some sonar dataset(s)
-% \end{itemize}
-% 
-% % \IEEEPARstart{T}{o} form images from a modern phased array sonar system the received wavefield is usually recorded, and then postprocessed by a digital beamformer.
-The beamformer applies delays and weights to the sensor channels, the beamformer adjusts the arrays spatial response to focus at one pixel at a time.
-such that signals emanating from regions of interest add constructively, while ensuring that noise and interference from other angles do not.
-% % 
-% % The imaging capabilities of a modern phased array sonar system depend on physical attributes such element response and array geometry, the transmitted signal, as well as the beamforming method being used on transmission and reception.
-Beamforming is the concept of applying delays and weights to the sensors channels to steer the arrays response to points of interest.
-% 
-% % 
-% % 
-% % Outline:
-% % \begin{itemize}
-% % \item Capon's resonse when applying robustification
-% % \item Choice of window functions makes little difference.
-% % \item Steering and mainlobewidths have outer bounds.
-% % \item Beamspace?
-% % \item Chosen window plots - what may they tell us? Variance intensity values when using various windows.
-% % \item Assymmetric windows?
-% % \end{itemize}
-% % 
-% % 
-% % \begin{align}
-% % z[n] = \sumb{m=0}{M-1} w_m[n]^*x_m[n-\Delta_m] = \w\H[n]\x[n-\Delta_m]
-% % \end{align}
-% % 
-% % 
-% % \section{Methods}
-% % 
-% % Basically, we are working with a practical implementation of the Capon beamformer that computes a set of weights $\vec w$ for every single sample $n$ by solving:
-% % \begin{gather*}
-% % \vec w[n] = \frac{\hat{\mat R}\;\!^{-1}[n]\vec a}{\vec a\H\hat{\mat R}\;\!^{-1}[n]\vec a} = \frac{\text{\raisebox{1.9pt}{$\vec\chi$}}[n]}{\vec a\H\text{\raisebox{1.9pt}{$\vec\chi$}}[n]}
-% % \end{gather*}%
-% % where
-% % % \newcommand\X{\text{\raisebox{2pt}{$\vec\chi$}}}
-% % \begin{gather*}
-% % \text{\raisebox{1.9pt}{$\vec\chi$}} = \hat{\mat R}\;\!^{-1}\vec a \qquad\qquad\text{is the solution to}\qquad\qquad \hat{\mat R}\text{\raisebox{1.9pt}{$\vec\chi$}} = \vec a.
-% % \end{gather*}
-% 
-% % 
-% % , and maximum suppression of while ensuring that the beamformer digitally  before each of the pixels are estimated one at a time.
-The resolution and contrast of such a system will depend on the systems spatial response, which ideally should be narrow  be very sharp in the desired direction its ability to achieve  fundamental principle of forming a sonar image is to record the received wavefield,
-% % 
-% % image quality of a phased array sonar imaging systems depend on  the choice of weights to apply to each of the sensors are crucial. 
-% % 
-% % A modern phased array imaging system may be thought of as a spatial filter. To achieve the best possible performance, the 
-% % 
-% % resolution and contrast 
-% % 
-% % Adaptive beamformers have only recently been introduced in active sonar imaging.
-For a while they were considered unsuited for this purpose because the backscattered signal is largely correlated with the
-% % 
-% 
-% 
-% % An example of a floating table. Note that, for IEEE style tables, the 
-% % \caption command should come BEFORE the table. Table text will default to
-% % \footnotesize as IEEE normally uses this smaller font for tables.
-% % The \label must come after \caption as always.
-% %
-% %\begin{table}[!t]
-% %% increase table row spacing, adjust to taste
-% %\renewcommand{\arraystretch}{1.3}
-% % if using array.sty, it might be a good idea to tweak the value of
-% % \extrarowheight as needed to properly center the text within the cells
-% %\caption{An Example of a Table}
-% %\label{table_example}
-% %\centering
-% %% Some packages, such as MDW tools, offer better commands for making tables
-% %% than the plain LaTeX2e tabular which is used here.
-% %\begin{tabular}{|c||c|}
-% %\hline
-% %One & Two\\
-% %\hline
-% %Three & Four\\
-% %\hline
-% %\end{tabular}
-% %\end{table}
-% 
-% 
-% % Note that IEEE does not put floats in the very first column - or typically
-% % anywhere on the first page for that matter. Also, in-text middle ("here")
-% % positioning is not used. Most IEEE journals use top floats exclusively.
-% % However, Computer Society journals sometimes do use bottom floats - bear
-% % this in mind when choosing appropriate optional arguments for the
-% % figure/table environments.
-% % Note that, LaTeX2e, unlike IEEE journals, places footnotes above bottom
-% % floats. This can be corrected via the \fnbelowfloat command of the
-% % stfloats package.
-%
-
 %%%%%%%%%%%%%%%%%%                              ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
@@
 %
-% 
-% \maketitle
-% 
-% \section{Introduction}
-% 
-% % Images from a phased array active sonar system are formed in the post-processing by focusing on one pixel at a time.
-This is achieved by delaying and weighting each of the array's channels, providing coarse and fine adjustments to the region of focus, respectively.
-This is commonly referred to as beamforming.
-Assuming far-field conditions, the delays are usually chosen to steer the pixel of interest to broadside, and the weights chosen to reject the noise impinging on the array from non-broadside directions.
-% % 
-% % The algorithm responsible for carrying out this task is referred to as a beamformer.
-Depending on whether data is evaluated when choosing a suitable set of delays and weights, beamformers can be branded either conventional or adaptive.
-A good example of the former is the The Delay-and-Sum (DAS) beamformer, which - as the name implies - simply delay the sensor channels appropriately prior to summing them up.
-Applying various weight sets, or windows, to the data prior to summation allows the lateral \gls{SNR} to be improved, but at the inevitable expense of a deteriorated lateral resolution.
-\todo{aka.
-the curse of physics.}
-% % 
-% % The Minimum Variance Distortionless Response (MVDR) (or \gls{MVDR}) beamformer, have the potential to overcome this limitation.
-While ensuring unit gain in the look direction, it computes the set of weights that minimizes the energy accumulated by the array from other directions.
-In other words, whenever the distribution of noise and interference is primarily focused around specific angles, an array response with a high degree of suppression at these angles are chosen, resulting in an image with a lateral \gls{SNR} and resolution superior to anything the \gls{DAS} can come up with.
-% 
-% 
-% % 
-% % In active sonar imaging the constrast and resolution can be improved by the use of adaptive beamforming. 
-% % 
-% % Adaptive beamforming 
-% % A critical part in the sonar image formation is the 
-% % The image quality is mainly limited by beampattern of the imaging arrays.  resolution and noise suppression
-% % 
-% % The quality of sonar images 
-% % 
-% % An essential part of a sonar imaging system is the beamformer. 
-% % 
-% % 
-% % A sonar image is typically formed by using receive beamforming to focus on one pixel at a time.
-The beamformer's ability to suppress interference   to focus on one pixel at the time.
-How well a pixel value is estimated, and thus how
-% % 
-% % A sonar image is built one pixel at the time the 
-% % 
-% % The image quality of a sonar system will depend on the imaging arrays ability to focus in 
-% % 
-% % 
-% % 
-% % Forming images done by combining data from multiple receivers in a way that to obtain focus on one pixel at the time, and assigning the output value to the intensity of that pixel. The 
-% % 
-% % Most modern active sonar systems employ phased arrays to 
-% % The basic way of a sonar image was formed 
-% % 
-% % A sonar system's ability to 
-% % 
-% % Adaptive beamforming has 
-% % 
-% % The image quality 
-% % 
-% % To improve the image quality of 
-% % 
-% % In high resolution sonar imaging the 
-% % 
-% % The image 
-% % 
-% % 
-% % In a typical sonar imaging scenario an encoded signal is transmitted to highlight a surface of interest, and the reflected wavefield sampled using a receiver array.
-A image is then formed in the post-processing by coherently combining the receiver outputs to focus at one pixel at the time.
-The principle of adjusting the arrays focus is commonly referred to as beamforming, and involves assigning suitable delays and weights to the array's channels.
-% % 
-% % Ideally, 
-% % 
-% % The formation of a sonar image is 
-% % 
-% % The principle of forming an image in active sonar systems usually 
-% % 
-% % A common way of creating high resolution sonar images is to use multielement arrays to  
-% % 
-% % In sonar imaging a common compromise 
-% % 
-% % A common way of creating sonar images is to use 
-% % 
-% % The drive 
-% % 
-% % The pursuit of 
-% % 
-% % Adaptive beamforming 
-% % 
-% % Beamforming is a fundamental operation in active sonar imaging. 
-% % 
-% % 
-% % 
-% % 
-% % Beamforming is a fundamental technique employed in sonar image formation to 
-% % 
-% % 
-% % The introduction of adaptive beamforming in active sonar imaging 
-% % 
-% % The resolution and 
-% % 
-% % The High resolution imaging techniques
-% % 
-% % 
-% % The resolution and constrast in active sonar images depend on the beamformers ability to 
-% % 
-% % A sonar image is typically formed by using receive beamforming to focus on one pixel at a time.
-The image resolution and contrast The image quality The quality of active sonar images depend - amoung other things - on the beamformer's ability to suppress interference  resolution and constrast
-% % 
-% % A sonar image is typically formed by using receive beamforming to focus on one pixel at a time.
-The quality of the final image will depend on the beamformer's ability to suppress interference and noise from off-pixel.
-to focus on one pixel at the time.
-How well a pixel value is estimated, and thus how
-% % 
-% % 
-% % In conventional active sonar imaging systems 
-% % 
-% % Various methods have been applied to active sonar imaging. One of these methods, 
-% % 
-% %  eamforming techniques have been applied in several fields to improve image. methods can be applied to 
-% % 
-% % Most active sonar imaging systems are based on conventional beamforming methods.
-These are robust and simple, but does not take the incoming wavefield into account.
-offer  in the image formation process.
-While being inherently robust, these methods
-% % 
-% % Adaptive beamformers
-% 
-% % With the evolution of processing power the 
-% % 
-% % adaptive method
-% % hot topic
-% % 
-% % With the ever increasing processing power found in modern computers 
-% % 
-% % As the processing power of todays modern computer systems keeps improving the feasibility of 
-% % 
-% % As the processing power of modern computer systems keep improving, the use of data adaptive methods in various imaging fields. 
-% 
-% Data adaptive methods have been reported to improve image quality in various imaging fields over the years. One such method, the Minimum Variance Distortionless Response (MVDR) beamformer, has been 
-% hown to produce images with improved constrast and resolution compared to conventional non-adaptive methods, in e.g.
-radar~\cite{Benitz1997}, ultrasound~\cite{\cite{Synnevag2007}}, and passive sonar.
-Recently has its feasibility in active sonar also been demonstrated~\cite{Blomberg2012a,Blomberg2011,Dursun2009,Lo2004}.
-% 
-% % \the\linewidth
-% 
-% Despite its inherent potential, the MVDR beamformer has yet to see widespread adoption in the active sonar field.
-There may be several reasons for this.
-For one, the method is not inherently robust, and will suffer from a phenomenon called signal cancellation in active systems~\cite{Widrow1982}.
-Another reason is that in its original form, the computational complexity is cubic with the number of channels, O(M$^3$), while conventional beamformers are at O($M$).
-This is because a spatial covariance matrix is estimated and inverted for each image pixel.
-% 
-% % Previous:
-% 
-% % Robustification:
-% % R estimate accuracy vs.
-dynamic trackability: (1) Trade observation time for bandwidth (2) Partial adaptiveness (less degrees of freedom than channels) (3) Quadratic constraint on sensor weights to limit the norm to some preset value.
-% % -  
-% 
-% Previous work have shown that the robustness issue may be solved through means such as temporal and spatial averaging~\cite{Kailath1985}, and regularization.
-Reducing complexity has been achieved in a variety of ways, often by enforcing a Toeplitz-structure on the covariance matrix to simplify its inversion, or by employing iterative solvers.
-% 
-% We will show that by exploiting symmetry and avoiding unnecessary computations, the spatial averaging will also aid in reducing the overall arithmetic complexity\todo{make sure this is shown clearly somewhere}.
-for Unlike other work, which focus modifying or transforming\todo{get low channel count in, as it makes a case against the other solutions.} We make no approximations, but rather stay with a textbook example of the method that handle complex basebanded data\todo{misplaced?}.
-% 
-% % Other work (keywords: mvdr, capon, computation, efficient, complexity):
-% % - Partially adaptive beamforming. DAS on non-overlapping subarrays, then MVDR \cite{Swingler1999a}.
-% % - Beamspace / frequency based methods.
-% % - Low complexity. PCA + LC. 
-% % - Systolic arrays. Radar \cite{Ward1984} 
-% % - Toeplitz. \cite{Yapanel} Spectrum estimation - speech recognition. 
-% % 
-% % Us:
-% % - Aim for the GPU. Simple is often better.
-% % - Therefore: Optimize the standard MVDR chain.
-% % 1. Reduce arithmetic instructions
-% % 2. Implement for full utilisation of the arithmetic throughput on the GPU.
-% % 
-% % (Made partially adaptive by employing subarray averaging to reduce degrees of freedom to L.)
-% 
-% 
-% Furthermore, we investigate the feasibility of accelerating the MVDR method using Graphics Processing Units (GPUs).
-These chips have in recent years emerged as a flexible and highly potent parallel technology well suited for many DSP tasks.
-Driven by the gaming industry, these devices are engineered to maximize Flop/s while keeping cost and power consumption at a minimum.
-Compared to many-core CPUs, a GPU core offer modest computational capabilities, but in return provide up to thousands of them\todo{misplaced?}.
-% 
-% Comparable work may be found in e.g.
-ultrasound, where Chen et.
-al.
-demonstrated a GPU implementation of the MVDR that operated on real data~\cite{Chen2011,Chen2011a}.
-While promoting impressing performance, such a constraint disallow the creation of asymmetric and shifted responses, and hence prevents the MVDR's potential to be fully reached.
-The complexity issue has been targeted in various ways.
-% 
-% This article is outlined as follows.
-First, we will introduce the concept of adaptive beamforming and provide details MVDR beamforming and our selected robustification techniques.
-Then we will investigate the complexity issue, discuss means for arithmetic reduction, and show how to properly map it to a GPU.\todo{too much overlap here}
-% 
-% %  Furthermore, by implementing it on a modern \gls{GPU} we have gained a speedup of around 2 orders of magnitude compared to an optimized single-thread implementation in C.
-% 
-% % These devices have in recent years emerged as a flexible and highly potent parallel technology well suited for many DSP tasks, and are available off-the-shelf at reasonable prices\todo{skip this?}.
-% 
-% 
-% % These devices have in recent years emerged as a flexible and highly potent parallel technology well suited for many DSP tasks, and are available off-the-shelf at reasonable prices.
-% 
-% % Yay or nay? $\Rightarrow$ GPU ultrasound (\cite{So2011,Chen2011}). Complex data, all robustification techniques.
-% 
-% 
-% % - Adaptive beamformers offers image quality edge, but are usually computationally complex.
-% % - Who's tried before?
-% % - We: Active sonar MVDR, GPU, high performance.
-% 
-% 
-% \section{Methods}\label{methods}
-% 
-% Consider a sonar imaging scenario where an encoded signal is transmitted to highlight a surface of interest, and assume that the backscattered wavefield is properly sampled using a uniform linear array of receivers.
-An image can then be formed in the post-processing by coherently combining the receiver outputs to focus at one pixel at the time.
-The principle of adjusting the array's focus is commonly referred to as beamforming, and involves assigning suitable delays and weights to the array's channels\todo{hmm, doesn't quite feel right...}.
-% 
-% 
-% \subsection{Beamforming}
-% 
-% \begin{figure}[!t]\centering
-% \includegraphics[width=\linewidth]{gfx/beamforming.eps}
-% \caption{Beamforming principle.
-Signal signature is first removed by matched filtering.
-Then - prior to summation - a suitable set of delays and weights are applied to focus on a pixel of interest.}\label{beamforming}
-% \end{figure}
-% Let the receiver be an $M$ element uniform linear array, and assume that the signature of the transmitted signal was removed by a matched filter (Fig.
-\ref{beamforming}).
-Further assume that the array channels are digitally delayed to create focus at a pixel with azimuth angle $\theta$ and range sample $n$, and let the delayed data from the $m$'th channel be $x_m[\theta,n]$.
-To simplify notation, the dependence on $\theta$ will be made implicit from now on.
-% 
-% By definition the beamformer output $z[n]$ can now be expressed as the weighted sum of all the delayed data samples:
-% \begin{align}
-% z[n] = \w\H[n]\x[n] = \bmat{w_0[n]\\w_1[n]\\\vdots\\w_{M-1}[n]}^H \bmat{x_0[n]\\x_1[n]\\\vdots\\x_{M-1}[n]},\label{z}
-% \end{align}
-% where $w_m$ is the weight factor assigned to channel $m$.
-With static weights this would be referred to as the conventional delay-and-sum (DAS) beamformer.
-Various weighting patterns exists here for trading lateral resolution for improved noise suppression (contrast), but one always ends up with a compromise between the two\todo{don't quite like this one}~\cite{Harris1978}.
-% 
-% Various adaptive beamformers target this limitation by allowing the weights to change for each pixel to better fit the dynamic nature of the incoming wavefield, the rationale being that the a priori knowledge embedded in the data may be used to improve image quality.
-The MVDR beamformer is one such method.
-It finds the set of complex weights that minimizes the beamformer's expected output power, while ensuring unity gain in the look direction~\cite{Capon1969}.
-This is a convex optimization problem that can be solved using Lagrange multipliers to yield the solution:
-% \begin{gather}
-% \vec w[n] = \frac{\Ri[n]\1}{\1\T\Ri[n]\1},\label{weights}
-% \end{gather}
-% where $\1$ is a row vector of ones that represents broadside steering, and $\R=E\{\x[n]\x\H[n]\} \in\mathbb{C}^{M,M}$ is the spatial covariance matrix for the full array.
-Since $\R$ is unknown, we will estimate it by computing a sample covariance matrix $\eR$.
-In this computation we will perform some degree of \emph{spatial averaging} to avoid signal cancellation, \emph{temporal averaging} to maintain true speckle statistics, and \emph{diagonal loading} to improve robustness to parameter errors ~\cite{Synnevag2009}.
-Combined, these steps will also ensure a numberically well conditioned $\eR$\todo{should we mention variance reduction?}.
-% 
-% If we let $x_l[n]$ represent the datavector from subarray $l$,
-% \begin{gather}
-% \x_l[n] = \bmat{x_l[n] & x_{l+1}[n] & \dots & x_{l+L-1}[n]}\T,
-% \end{gather}
-% then a sample covariance matrix formed with temporal and spatial averaging, $\eR_\text{s,t}$, is given as:
-% \begin{gather}
-% \eR_\text{s,t}[n] =  \frac{1}{N_K N_L} \sumb{l=0}{M-L}\sumb{n'=n-K}{n+K} \x_l[n]\x_l\H[n] \in\mathbb{C}^{L,L},\label{spatialR}
-% \end{gather}
-% where $N_K = 2K+1$ is the number of temporal samples to perform averaging over, and $N_L = M-L+1$ is the number of subarrays.
-% 
-% To further improve robustness, we add a fraction $d$ of the total power of $\eR_\text{t,s}[n]$ to its diagonal~\cite{Synnevag2007}:
-% \begin{align}
-% \eR[n] = \eR_\text{t,s}[n] + \I \frac{d}{L} \tr\{\eR_\text{t,s}[n]\},\label{finalR}
-% \end{align}
-% where $\I$ is an identity matrix, $\tr\{\cdot\}$ represents the matrix trace operation, and $\tr\{\eR_\text{t,s}[n]\}$ is an estimate of the energy received by this pixel.
-% 
-% Note how subarray averaging led to a size reduction of $\eR$ from $\mathbb{C}^{L,L}$ to $\mathbb{C}^{M,M}$, and hence will produce an $L$-element weightset when substituted into (\ref{z}).
-This weightset is then applied to on all the subarrays, prior to computing the beamformer output as in (\ref{z}).
-Or, equivalently, we may apply the weightset to the sum of all the subarrays:
-% \begin{align}
-% z[n] = \w\H[n] \sumb{l=0}{M-L} \x_l[n].\label{finalZ}
-% \end{align}
-% \begin{figure}[!t]\centering
-% \includegraphics[width=\linewidth]{gfx/implementation.eps}
-% \caption{MVDR beamforming. For one of the total number of pixels in range and azimuth, $N_y$ and $N_x$,\newline
-% 1. an $L\times{}L$ sample covariance matrix $\eR$ is computed, \newline
-% 2. the term $\eR^{-1}\1$ is found using a linear equation solver,\newline
-% 3. and the beamformer output is computed from $z$ from (\ref{finalZ}), where $\w$ is found by substituting $\eR^{-1}\1$ into (\ref{weights}). } \label{mvdr_beamforming}
-% \end{figure}
-% % \begin{figure}[!t]\centering
-% % \includegraphics[width=\linewidth]{gfx/algorithm_structure.eps}
-% % \caption{MVDR beamforming.
-First a spatial covariance matrix is estimated from the delayed data (\ref{spatialR}-\ref{finalR}), then the weights are computed (\ref{weights}) and finally applied to the delayed channel data (\ref{z}).}
-% % \label{implementation}
-% % \end{figure}
-% As summarized in Fig. \ref{mvdr_beamforming}, the MVDR method is applied to each pixel independently, by
-% \begin{enumerate}
-% \item computing the sample covariance matrix $\eR$ in (\ref{finalR}),
-% \item computing $\eRi\1$ in (\ref{weights}), and
-% \item computing the beamformer output $z$ in (\ref{finalZ}).
-% \end{enumerate}
-% % This is illustrated in .
-% % \begin{figure}[!t]\centering
-% % \includegraphics[width=\linewidth]{gfx/buildR-complex_data-average.eps}
-% % \caption{.}\label{complexity}
-% % \end{figure}
-% 
-% \subsection{Computational Complexity}
-% 
-% When assessing the MVDR beamformer's complexity and mapability to parallel hardware, computing and inverting $\eR$ is by far the most computationally intensive task.
-If we neglect spatial and temporal averaging then computing $\eR$ is of O($M^2$), and inverting it is of O($M^3$).
-But if we implement spatial and temporal averaging as in (\ref{spatialR}), then computing $\eR$ becomes of O($N_K N_L L^2$) and the inversion of O($L^3$).
-Furthermore, assuming rectangular complex notation, we should take into account that a complex multiplication is roughly 3 times as arithmetically intense as a complex addition.
-So what does it all amount to in the end?
-% 
-% To investigate this aspect we derived exact complexity formulas for each step in the MVDR process\todo{except partial pivoting, diagonal loading}.
-Then these were evaluated for the entire range of possible subarray sizes from $L\in[1,M]$, with temporal averaging assessed with values $K\in\{0,1,2\}$ and the number of channels set to $M\in\{8,16,24,32\}$.
-% 
-% The results are shown in Fig.
-\ref{mvdr_complexity}.
-Note how the computation of $\eR$ completely dominates at smaller subarray sizes, that solving $\eRi\1$ only plays a notable role when the element number and subarray sizes are higher, and that the computation of $z$ has a neglible impact.
-Also notice how temporal averaging comes with a high computational penalty.
-This is because building $\eR$ is heavy on complex multiplications, and a lot of these are repeated unneccessary.
-% 
-% \begin{figure}[!t]\centering
-% \includegraphics[width=\linewidth]{gfx/mvdr_complexity.eps}
-% \caption{Per-pixel computational complexity of the steps in MVDR beamforming (prior to any optimizations).
-To avoid signal cancellation in an active sonar system we usually set $L<\frac{M}{2}$, in which region the computation of $\eR$ dominates in terms of arithmetic complexity, especially with temporal averaging.}\label{mvdr_complexity}
-% \end{figure}
-% 
-% % Not corrected for fused multiply accumulate
-% 
-% % Also, the need for partial pivoting has been ignored, as we ensure a numerically well conditioned $\eR$ by applying diagonal loading.
-Since this is the most common scenario, we will have a thorough look at how it can be optimized.
-% 
-% % and does not take into that does not  feasibility parallel decomposition  memory consumption.
-% 
-% 
-% 
-% \section{Mapping the MVDR to a GPU}
-% 
-% % \begin{itemize}
-% % \item Keep delay step on CPU, as this is a common step in all beamforming
-% % \item Increasing matrix size reduces the number of threads per block, since the entire matrix must be in shared memory.
-1 matrix 72x72 complex (8 bytes) max per block - 72 threads.
-same as 207 5x5 matrixes, 5*207=1036 threads.
-max threads in block 512.
-% % \end{itemize}
-% 
-% % An important feature of MVDR beamforming, and with beamforming in general, is that each pixel can be computed independently and in an identical fashion.
-Considering that sonar images often end up having millions of pixels, this makes a compelling case for investigating technologies that support massive data level parallelism.
-The technology that does this best is perhaps Graphics Processing Units (GPUs).
-These chips contain up to thousands of small computation cores, which combined deliver superior peak floating point performance compared multi-core systems such as CPUs.
-% 
-% An important feature of MVDR beamforming, as with beamforming in general, is that each pixel can be computed independently and in an identical fashion.
-Furthermore, a typical sonar image may contain millions of pixels.
-This represents a level of data parallelism that appears very well suited for a massively parallel architecture such as the GPU.
-% 
-% To investigate the feasibility of running MVDR on a GPU we decided to map it to a Nvidia GeForce Quadro 6000.
-This is a high-end Compute Unified Device Architecture (CUDA) enabled GPU based on Nvidia's Fermi architecture.
-The code was written in Nvidia's ``C for CUDA'' framework.
-We believe an equally attractive option would be to use a GPU from AMD, and that the code for either of these GPUs could have been written cross-platform framework OpenCL from the Khronos group, but comparing these alternatives are beyond the scope of this study.
-% 
-% To use Nvidia's own terminology, the Quadro 6000 is comprised of 14 streaming multiprocessors (SMs), each having 32 CUDA cores that execute a common program called a kernel.
-Combined these cores are able to deliver more than 1\;Tflop/s (appendix \ref{throughput}), but only if the implementation is computationally bound.
-This can be achieved by balancing the load evenly on all cores, and by trying to avoid that some cores are forced to idle due to a pending data transfer or thread synchronization.
-As the best way to do this is completely different for each step in the MVDR process, we have designed a different kernel and configuration for each of them.
-Particular attention will be paid to building $\eR$, since the potential for gaining overall speedups are greatest here\todo{review}.
-% 
-% %  this assumes that the implementation is solely bound by arithmetic throughput.
-r bound by memory bandwidth, arithmetic latency or arithmetic bandwidth.
-or  is solely bound by processing power.
-These cores need to be kept busy if we are to That make for a total of 448 CUDA cores which we need to keep busy if we want to tap the full potential.
-% 
-% \begin{figure}[t]\centering
-% \includegraphics[width=\linewidth]{gfx/mvdr_build_R.eps}
-% \caption{Step 1: Building $\eR$.
-This is a visualization of how $\eR$ could be built in a case with $M=5$ sensors, with subarray size $L=3$ and temporal averaging set to $K=1$.
-Here $\eR_{l}$ is the sample covariance matrix for the $l$'th subarray, and $\eR_s$ is the average of all these.
-Note that instead of performing the temporal sum last as here, one could take more temporal samples into consideration in the computation of each $r_{ij}$.}\label{mvdr_build_R}
-% \end{figure}
-% 
-% 
-% % \todo{Another shitty GPU transition}
-% % Before proceeding we should make a quick note on the GPU memory.
-Note that to obtain the best possible performance, we need to make sure that CUDA threads find the data they need in cache or registers most of the time.
-Only these two memory types have a high enough bandwidth to promote sustained maximum utilisation of all CUDA cores.
-Unfortunately, they are a sparse resource.
-% 
-% % This makes it apparent that computing one pixel per thread is not a feasible solution. \todo{...}
-% 
-% % When optimizing an algorithm for speed, it is if of importance to know whether an algorithm is bound by computations or memory.
-One way to get an approximate idea of this is to compute the ratio of peak arithmetic throughput to peak memory throughput of the in the target platform, and compare this ratio to the ratio of arithmetical operations to memory operations in the algorithm.
-% 
-% % follow collective access patterns to memory.
-% 
-% % At first glance, it might seem like a good idea to compute one pixel per CUDA core.
-Each pixel is computed using the same instructions, and without any exchange of data.
-However, this approach is unfeasible due to the limited cache and register memory available to each thread.
-One could resolve to using the slower GDDR5 memory, but this would largely mitigate the performance gain and defeat the purpose of using a GPU in the first place.
-% 
-% %Bandwidth becomes crucial, as getting data in to the cores and out again must happen i a frantic pace. %the feeding all these cores with data and harvesting the result
-% 
-% % To effectively harness the computing potential of a GPU we need to make sure all computing cores stays busy most of the time. 
-% 
-% % Memory latency can be hidden in two ways.
-One is to keep a lot of threads queued up on each SM at all times, such that when some threads perform memory operations other threads can be scheduled to run instead.
-This is referred to as data level parallelism (DLP).
-Alternatively, or preferrably in addition, the designer may promote instruction level parallelism (ILP) by letting subsequent instructions in a thread be independent.
-While ILP can be very effective when done correctly, DLP is generally simpler to design for.
-% % 
-% % In the CUDA framework, the threads on an SM is managed through a software abstraction call a compute block.
-The block is a 3 dimensional grid where each element correspond to a thread that is scheduled to be executed on the SM at some point.
-% 
-% % load balancing
-% % - distribute load - optimize symmetry
-% % - avoid branching
-% % 
-% % memory latency hiding
-% % - make threads use high bandwidth memory
-% % - hide latency with dlp, ilp, or both
-% % 
-% % thread sync hiding
-% % - queue blocks up on an sm
-% 
-% %This is referred to as a single-program, multiple-data (SPMD) architecture\todo{malplaced}.
-% 
-%  %An SM is a single-program multiple-data (SPMD) processor, meaning that all the CUDA cores on an instruction unit, L1 cache and registers is shared by all the CUDA cores on that SM.  %
-% % GPUs, on the other hand, does not attempt to do ILP, but use all available transistors to support massive DLP. This leads to designs such as the nNvidia GeForce GTX 580. 
-% % each with 48kB of L1 cache (shared memory) and 128kB of register memory.
-% 
-% % -bank conflicts when L > 32. not treated.
-% 
-% 
-% \subsection{Computing the Spatial Covariance Matrix, $\hat{\boldsymbol{R}}$}
-% 
-% % As Fig. \ref{mvdr_complexity} illustrates the computation of $\eR$ 
-% 
-% As observed in Fig.
-\ref{mvdr_complexity} a direct computation of $\eR$ by implementing (\ref{spatialR}) and (\ref{finalR}) is the greatest computational burden.
-We target this issue by first performing \emph{arithmetic reductions}, then aim to get as close to the peak \emph{arithmetic throughput} on the GPU as possible.
-% 
-% \subsubsection{Arithmetic Reductions}
-% 
-% In Fig.
-\ref{mvdr_build_R} we illustrate how to reduce the arithmetic complexity of building $\eR$ in a system with $M=5$ sensors, a subarray size of $L=3$ and temporal averaging is set to $K=1$.
-Here the spatial sum is carried out before the temporal sum.
-To reduce arithmetic operations we may
-% %
-% \begin{enumerate}
-% \item exploit the fact that $\eR$ is Hermitian positive semi-definite to compute only one half of it,
-% \item avoid redundant multiplications, and
-% \item implement the spatial sum iteratively by sliding along the diagonals of $\eR$.
-% \end{enumerate}
-% %
-% To study the effect of these optimization strategies we derived complexity curves for the plain implementation with and without swapped sums, a couple that take symmetry into account, and a couple that iteratively sum up the diagonals of $\eR$.
-Then these curves were normalized to the complexity curve of the reference implementation in Fig.
-\ref{mvdr_complexity}.
-The results, shown in Fig.
-\ref{mvdr_optimization}, illustrate that taking symmetry into account and implementing iterative summing both lead to considerably reduced arithmetic complexity, especially at lower subarray sizes and when using temporal averaging.
-% 
-% We chose the iterative solution based on its overall excellent performance, and because it can be altered to consume very little memory. This is \emph{very} important, as we will see next.
-% 
-% \begin{table}[b]\centering%\normalsize
-% \begin{tabular}[c]{l r@{:}  l}\hline
-% \rowcolor{tabBlue}\bf Memory type & \multicolumn{2}{>{\columncolor{tabBlue}}c}{\bf BW$_\text{memory}$/BW$_\text{arithmetic}$} \\\hline
-% Global memory &\hspace{30pt} 1 & 30 \\
-% Shared memory & 1 & 4 \\
-% Registers & $>$3 & 2~\cite{Vasilyy}
-% \end{tabular}
-% \caption{Nvidia Quadro 6000 bandwidth ratios.}\label{tabbandwidth}
-% \end{table}
-% 
-% \begin{figure}[!t]\centering
-% \includegraphics[width=\linewidth]{gfx/buildR-breakdown.eps}
-% \caption{Arithmetic optimization of computing $\eR$: Relative reduction in arithmetic complexity (higher is better).
-All plots are normalized to the complexity curve of the reference implementation in Fig.
-\ref{mvdr_optimization}.}
-% \label{implementation}
-% \end{figure}
-% 
-% \subsubsection{Arithmetic Throughput}
-% 
-% When we compute $\eR$, we very rarely perform more than 1-3 floating point operations for every float read or written to memory.
-Unfortunately, the GPU prefers kernels to be more computationally intensive than this.
-This can be inferred from Table \ref{tabbandwidth}, where the peak bandwidth of the three types of GPU memory are compared to the peak arithmetic throughput (see appendix \ref{throughput} for derivations).
-The global memory\todo{if there's space somewhere, introduce memory layout}, in which all data must reside at some point\todo{rephrase}, is at best only able to move one float for every 30 floats processed\todo{well, not entirely true.
-rephrase} by the CUDA cores.
-However, peak shared memory performance is only 4 times slower than the arithmetical throughput so we really need to use it.
-To achieve this we must
-% \begin{enumerate}
-% \item keep all the data relevant to the pixels being computed in shared memory, while
-% \item trying to use it as efficiently as possible.
-% \end{enumerate}
-%  %Further alleviation is also possible by exposing instruction level parallelism within a thread, as this promotes the use of registers with a bandwidth at least 6 times higher than that of shared memory~\cite{Vasilyy}\todo{so why don't we?
-rephrase}.
-% 
-% These challenges are very closely linked.
-The Quadro 6000 architecture is of compute capability 2.0, which means there are 48\;kB of shared memory (L1 cache) and 128\;kB of registers per SM~\cite{Nvidia2012}.
-This memory is shared by all active threads on that SM, a number that should be no less than 768~\cite{Nvidia2012a}.
-This will expose a sufficient level of data parallelism to ensure that memory latency is completely hidden (Little's law).
-By dividing the shared memory evenly on all 768 threads we find that each thread should store no more than 8 single precision complex numbers in shared memory\todo{introduce memory structure -shared for data shared between threads}, and 24 stored in registers.
-This should make it apparent why computing a single pixel per thread is a bad idea, and why we need to keep each thread as light on memory consumption as possible.
-% 
-% \begin{figure}[!t]
-% \centering
-% \includegraphics[width=\linewidth]{gfx/mvdr_implementation.eps}
-% \caption{MVDR implementated on a GPU.
-The sample covariance matrix is formed by threads running along its diagonals, performing spatial averaging iteratively.
-Then Nvidia's     ces for each image pixel are computed .}\label{mvdr_implementation}
-% \end{figure}
-% 
-% % What about the GPU implementation?
-This one was tricky.
-Balance load, avoid inter-thread dependencies when performing iterative summations, keep thread memory consumption at a minimum, and promote collective access patterns (coalesced reads/writes) to global memory.
-% 
-% In Fig.
-\ref{mvdr_implementation} we illustrate how we got around these challenges.
-First we make each SM compute entire range lines of pixels, then load the subset of $\x$ that all these pixels depend upon into shared memory.
-This lets us perform temporal averaging without the need for reading additional channel data.
-% 
-% Second we assign $L$ threads per pixel to traverse the diagonals of $\eR$.
-On the top row a full computation is carried out, then that entire row is saved back to global memory following a collective access pattern that maximize global memory bandwidth.
-For subsequent rows the threads move along the diagonals while performing iterative summations; the result from the previous element on the diagonal is updated by adding and removing the correlation coefficients that enters and exits the sum, respectively.
-To minimize memory consumption, we compute the coefficients again every time we need them.
-When a thread has finished up a diagonal, we have them wrap around to compute one of the diagonals in the lower triangular of $\eR$.
-Since $\eR$ is conjugate symmetric, the values in the leftmost column is obtained by a complex conjugate copy of the relevant value in the first row.
-Combined these steps balances the load evenly on all threads, is almost completely free of arithmetic redundancy, and consumes less
-% memory.
-%  
-% 
-% 
-% % Hence, whenever $N_\text{avg}>1$ the sliding window approach will demand less instructions at the expense of a slightly higher memory consumption.
-% 
-% % However, an advantage with this approach is that the summations can be carried out in any order, and hence this form exhibits great deal of parallelism.
-Can we optimize the algorithm without sacrificing parallelism?
-% 
-% 
-% % - Order of summations - beamspace capable
-% 
-% % 
-% % In the upcoming results, we have beamformed pre-delayed data.
-This is a large dataset, and the latency experienced when copying it from the \gls{CPU} side to the \gls{GPU} was significant.
-However, in a practical system only the channel data should be transferred as the delaying of data can effortlessly be performed by the \gls{GPU}.
-The remaining latency can be hidden as long as the \gls{GPU} is kept busy, hence the data transfer times will not be reported in the upcoming results.
-% % 
-% 
-% % \begin{itemize}
-% % \item Arrange order of summations to reduce data dimensions as early as possible.
-% % \item Sliding window in time/space to reduce instructions at the cost of extra memory consumption. Only pays off to do this over subarrays.
-% % \end{itemize}
-% % 
-% % \item Between GPU and CPU: Several concurrent blocks (contexts) per SM.
-% % \item Job queue (how does this work)
-% % \item Register pressure and shared memory. Thread context must fit in local memories.
-% % \item Warp computational intensity should be high to make up for the memory cost.
-% 
-% 
-% % increasing performance hit several walls:
-% % \begin{itemize}
-% % \item memory couldnt keep track - remedy cache.
-% % \item instruction parallelism - complex to identity
-% % \item power consumption cubic with frequency
-% % GPUs:
-% % \begin{itemize}
-% % \item lower frequency, less power
-% % \item no instruction parallelism
-% % \item massive amounts of lightweight cores
-% % \end{itemize}\end{itemize}
-% 
-% \cite{Owens2007} \cite{Lee2010}
-% 
-% 
-% \subsection{Solving $\hat{\boldsymbol{R}}$$\;\!^{-1}$$\mathbf{1}$}
-% 
-% As intuition might suggest that the matrix product $\eRi\1$ can be carried out by first inverting the matrix $\eR$, then performing the inner product with $\1$.
-However, one can also solve the linear equation $\hat{\boldsymbol{R}}\boldsymbol{b} = \1$ for $\boldsymbol{b}$, where $\boldsymbol{b} = \eRi\1$.
-In terms of arithmetic count, and when comparing direct implementations, the latter appears to be the preferred option.
-% 
-% An important thing to note, however, is that unlike the problems most libraries tries to solve we do not attempt so solve large linear equations or invert large matrices; our matrices are small, but we have a very large number of them.
-Fortunately Nvidia recently released a highly optimized batched linear equation solver aimed for this particular purpose.
-It is a Gauss Jordan based and supports partial pivoting and complex numbers.
-The only downside to using it in our application is that it does not exploit the symmetry properties of our sample covariance matrix.
-A better choice would be a Cholesky solver, which is designed for Hermitian positive semi-definite matrices and is expected to offer roughly a factor 2 speedup over the Gauss Jordan one.\todo{somewhere a short remark regarding inversion/solver alternatives should be made}
-% 
-% % \emph{Computing} $\eRi[n]\1$ (from (\ref{weights})).
-While intuition may suggest that this step is carried out by inverting $\eR$, it is better to solve the linear equation $\R\beta = \1$ for $\beta$ instead, which gives us $\beta = \eRi\1$ directly.
-We have tested various solvers for this task, both in-house and proprietary implementations, and achieved the best performance by using an unofficial batch solver from nVidia.
-Inverting $\eR$ is by far the most computationally intensive task, and a key area of focus for further improvements.
-% 
-% 
-% \subsection{Computing $z$}
-% 
-% The beamformer output $z$ is computed on a per-pixel basis by first computing the MVDR weights $\w$, which are a mere scaled version of $\eRi\1$ (\ref{weights}).
-Then the sum in (\ref{finalZ}) is found by assigning a group of $L$ threads to respective elements of $x_l$, which then proceed to accumulate these for all $N_L$ subarrays.
-The resulting data vector is finally multiplied with the weight vector using the same threads, and then a single thread is used to sum these products to obtain $z$\todo{Rakende likegyldig hva vi gjør her..
-Kan kuttes.}.
-% 
-% %  The subarray data can, as here, be reduced to coincide in length with the subarray weights, or the weights can be extended to $M$ in size and applied directly on $\x[n]$ and summed.  
-% 
-% % \item \emph{Computing the spatial covariance matrix} $\eR[n]$ (\ref{spatialR}-\ref{finalR}).
-A group of threads are created that slide along the diagonals of $\eR$.
-In this way we exploited the fact that entries on the diagonals overlap across subarrays and time, and keep the numbers of both data reads and writes at a minimum.
-% % %Finished threads can also wrap around and start on diagonals on the lower half triangular. 
-% % % In this way we process one row of $\eR$ per kernel iteration, and manage to keep the numbers of both data reads and writes at a minimum.
-% 
-% % There are essentially two ways to accelerate an algorithm on parallel hardware.
-One is to identify independent instructions for each thread, and run these in in parallel, and the other is to support the parallel execution of as many threads as possible.
-This is referred to as instruction level parallelism (ILP) and data level parallelism (DLP), respectively.
-Most general purpose processors, such as a CPUs, support both DLP by featuring multiple cores with single-data multiple-data (SIMD) capabilities, and ILP by through means such as branch prediction and out-of-order execution.
-% 
-% % GPUs, on the other hand, does not attempt to do ILP, but use all available transistors to support massive DLP.
-This leads to designs such as the nNvidia GeForce GTX 580.
-Using their own terminology, it is comprised of 16 ``streaming'' multiprocessors (SMs), each having having 32 CUDA\todo{make sure CUDA is introduced} cores that execute a common program called a kernel.
-%An SM is a single-program multiple-data (SPMD) processor, meaning that all the CUDA cores on an instruction unit, L1 cache and registers is shared by all the CUDA cores on that SM.
-%
-% % each with 48kB of L1 cache (shared memory) and 128kB of register memory.
-% 
-% 
-% % Lots of cores with modest computing abilities, but in return there are hundreds of them. Little memory for each core. 
-% 
-% % In short, designing for a GPU involves finding the optimal balance between hiding memory latency (high occupancy) and resource utilization.
-% % 
-% % blocks in grid - high enough so that all SMs have atleast one block to execute.
-Optimally in the thousands to scale for future devices, and to allow an SM to switch between blocks to keep hardware busy when running synchronizing threads.
-% % 
-% % threads in block - min 64. 128-256 better starting point.
-% 
-% 
-% % Instead of having a few complex core that can boost massive single thread performance, designers chose to create hundreds of lightweight cores that each is much powerful than a single CPU core, but combined boasts a massive artithmetical performance.
-% % 
-% % \begin{itemize}
-% % \item Lots of cores, little memory for each.
-% % \item Few data transfers, lots of computations.
-% % \item Data transfers are moved around by hardware that can operate asynchronously from the compute units.
-% % \item Memory latency hidden as long as it takes longer to compute than to move data around.
-% % \item Tasks can be queued on SM (blocks) and on CPU to hide internal and external memory latencies.
-% % \end{itemize}
-% % 
-% % Vast amount of processing power, which are easily accessible but hard to fully harness. This is because in the pursuit of an ever increasing amount of arithmetical computing cores, the 
-% % 
-% % Unlike a CPU, a GPU 
-% % 
-% % If we compare a GPU with a CPU, 
-% % 
-% % When implementing an algorithm on a GPU, it is important to be aware
-% % 
-% % People friendly:
-% % \begin{itemize}
-% % \item More and more compute units, but less memory for each.
-% % \item Need to keep each compute unit busy.
-% % \item Few data transfers, lots of computations.
-% % \item Data transfers are moved around by hardware that can operate asynchronously from the compute units.
-% % \item Memory latency hidden as long as it takes longer to compute than to move data around.
-% % \item Tasks can be queued on SM (blocks) and on CPU to hide internal and external memory latencies.
-% % \end{itemize}
-% % 
-% % The ``fast'' memory of the GPU is the registers and cache. All threads executed 
-% % 
-% % 
-% % \begin{itemize}
-% % \item Hiding memory access.
-% % \begin{itemize}
-% % \item Between GPU and CPU: Several concurrent blocks (contexts) per SM.
-% % \item Job queue (how does this work)
-% % \end{itemize}
-% % \item Register pressure and shared memory. Thread context must fit in local memories.
-% % \item Warp computational intensity should be high to make up for the memory cost.
-% % \item Optimized for small L, high M, hardly any time averaging. A typical sonar scenario.
-% % \end{itemize}
-% 
-% 
-% 
-% % - Split program into equally sized portions - load balancing
-% % - Some portions depend on other potions already being carried outer - sequential dependencies
-% % - Effort timed in order to finish at the same time - synchronization
-% % 
-% % - task level parallelism - compute images independent of each other
-% % - data level parallelism - compute pixels independent of each other
-% 
-% % - Compute bound vs. bandwidth bound
-% 
-% 
-% % Shared memory:
-% % - Inter-thread communication within a block
-% % - Cacke data to reduce redundant access to global memory
-% % - User-managed cache (reduce global memory access patterns)
-% %
-% % 32 banks, each bank 4-byte wide
-% 
-% 
-% 
-% 
-% \section{Images and Benchmarks}
-% 
-% % In a practical system it should be moved to the GPU as well. Therefore, in the upcoming benchmarks, CPU computation time and data transfer to the GPU is neglected.
-% 
-% \begin{figure}[t]
-% \centering
-% \includegraphics[width=\linewidth]{gfx/plot_holmengraa_L16_Navg1.eps}
-% \caption{HISAS sidescan sonar (SSS) image of the shipwreck Holmengraa.}\label{holmengraa}
-% \end{figure}
-% \begin{figure}[t]
-% \centering
-% \includegraphics[width=\linewidth]{gfx/benchmark.eps}
-% \caption{\protect MVDR benchmarks. Forming a 55\;kpixel image from a $M=32$ channel array.}\label{benchmarks}
-% \end{figure}
-% \begin{figure}[t]
-% \centering
-% \includegraphics[width=\linewidth]{gfx/code_assessment.eps}
-% \caption{Execution time of an arithmetic-only and a memory-only version of the MVDR code.
-A 55\;kpx dataset was processed for all $L$ using $K=1$, and the mean execution time was used here.
-The kernel building $\eR$ can here seen to be memory bound, as the time the kernels spends performing memory transactions are higher than the corresponding time it spends carrying out arithmetical operations.
-Furthermore, the runtime of the In the code that solves $\eR$ }\label{code_assessment}
-% \end{figure}
-% 
-% To demonstrate the imaging capability of the MVDR beamformer, we have processed experimental datasets from the $M$=32 element Kongsberg Maritime HISAS1030 sonar mounted on a HUGIN AUV~\cite{Hansen2009}.
-HISAS1030 is a high resolution synthetic aperture sonar with an array length of 1.2\;m and operating frequency of 100\;kHz.
-To produce the image shown in \Fig{holmengraa} it was operated in sidescan mode.
-The studied object is the 1500 dwt oil tanker wreck Holmengraa.
-It is 68\;m long and 9\;m wide, and lies at a slanted seabed at 77\;m depth outside of Horten, Norway.
-The MVDR images were processed with parameters $L$=16, $K$=3, and 1\% diagonal loading.
-% 
-% For computational performance we supply benchmarks in \Fig{benchmarks}.
-On our test system with a quad-core Intel Xeon E5620, 64Gb of RAM and an nVidia Quadro 6000 card, our GPU implementation of MVDR was able to process just over 1\;Mpx/s.
-A single-thread C implementation that was added as reference was outperformed by around 2 orders of magnitude.
-However, do not consider this comparison fair because while the GPU design is make efficient use of the L1 cache, similar optimisation work was not carried out on the CPU design.
-Also note that the computation times are reported without including the data transfer time from CPU to GPU, which account for 2-20\% of the total processing time.
-This is because we the required data amount to transfer will be reduced by a factor $M$ if we compute channel delays on the GPU instead of the CPU, which assume will be the case in a full beamforming system.
-Then the transfer time will be way less than the computation time, and can be easily hidden using asynchronous data
-% transfer.
-% 
-% % 
-% % We have tested our \gls{GPU} implementation of the \gls{MVDR} beamformer experimental datasets from the 32 element Kongsberg Maritime HISAS1030 sonar\todo{shouldn't we state that it's SAS?} being attached to the HUGIN \gls{AUV}~\cite{Hansen2009}.
-To obtain the studied object, the 1500 dwt oil tanker wreck Holmengraa, the sonar was operated in sidescan mode to image (\Fig{hlmengraa}).
-Holmengraa is roughly 68\;m long and 9\;m wide, and lies on a slanted seabed at 77\;m depth.
-The image was processed with \gls{MVDR} parameters $L=16$, $N_\text{avg}=3$, and $d=1\%$, which proved to be a reasonable selection for this scenario.
-% % 
-% % The computational performance of our GPU implementation is displayed in (\Fig{benchmarks}).
-Processing times are shown for varying subarray lengths $L$ and varying amounts of temporal averaging $N_\text{avg}$.
-A single-thread C and Matlab implementation is added as reference, although only moderate optimization work was carried out on these.
-The test system was a quad-core Intel Xeon E5620, with 64Gb of \gls{RAM}, and an nVidia Quadro 6000 card.
-% 
-% 
-% \section{Discussion}
-% 
-% \Fig{holmengraa} demonstrates the MVDR beamformer's ability to suppress interference and achieve better detail resolution.
-Compared the DAS beamformer, the ship's edges appear sharper and the shadows less noisy.
-This is in accordance with previous work.
-% %But while \gls{DAS} processed this 1.3\;Mpixel image in milliseconds, our Matlab implementation\todo{never mentioned matlab vefore} of MVDR beamformer needed minutes.
-% 
-% As Fig.
-\ref{benchmarks} demonstrates, performing arithmetic reductions and designing for using GPUs can significantly improve upon the problem with  the processing speed.
-If we look at the total runtimes for a subarray size of $L=16$ and temporal averaging at $K=2$, the GPU implementation is able to process the test image in roughly 40\;ms.
-This translates to a processing throughput of 1.4\;Mpixels/s.
-For the same scenario the C implementation clocked in at roughly 3\;s and Matlab at 17\;s, which is 75 and 435 times slower than the GPU implementation, respectively.
-% 
-% Interesting to note is also that solving $\eRi\1$ remains the main bottleneck in the GPU implementation, and it gets even worse for larger $L$'s.
-However, this is a general Gauss Jordan solver which does not exploit symmetry properties of $\eR$, so creating e.g.\ a batch based Cholesky solver should improve the runtimes further.
-Both the C and Matlab implementation in this benchmark take this symmetry into account, which partially explains why the complexity curves of these implementations are different from that of the GPU implementation.
-% 
-% If the HISAS1030 was attached to a platform moving at 1.8m/s, for which the maximum range is 250\;m~\cite{Hansen2010}, the pulse repetition frequency could be set to 3\;Hz.
-We could further assume a complex sampling frequency of 40\;kHz to accomodate the  HISAS1030's bandwidth of 30\;kHz, and beamform $M=32$ lateral beams.
-The throughput required for real-time processing of these images would then be 1\;Mpixels/s.
-As we have seen, this is something our GPU implementation can handle if $L$ is not too large.%, leaving the \gls{CPU} free to take on other assignments.
-% 
-% 
-% 
-% 
-% % What may this performance be used for?
-Let us assume that the 32 element, 1.2\;m long HISAS1030 array is mounted on a platform moving at 1.8\;m/s.
-Then the pulse repetition frequency (PRF) could be set to 3\;Hz to ensure critical along-track sampling,
-% % To illustrate what this performance might be used for, consider a moving platform travelling at 1.8\;m/s. 
-% % If the sonar system is Maximum range 
-% % There is a limit to the along-track sampli
-% % To avoid along track undersampling, the 
-% 
-% 
-% 
-% % HUGIN: 1.8m/s - range 250m \\
-% % fs = 100kHz*30\%*$\frac{4}{3}$ = 40kHz \\
-% % $N_{range px} = \frac{2\;250m}{1500m/s}40kHz = 13.3kpx$ \\
-% % $\times 32$ beams = 426kpx \\\\
-% % 
-% % PRF = $\frac{1500m/s}{2 250m} = 3Hz$ \\
-% % TP = 215px 3Hz = 1.28Mpx/s \\
-% % 4 arrays: 1.28Mpx/s * 4 = 5.12Mpx/s
-% % 
-% % 
-% % \begin{itemize}
-% % \item Need for speed: HUGIN 4 banks of 32 elements, can be processed faster than the ping repetition rate, with margins to spare.
-% % \end{itemize}
-% 
-% 
-% \section{Conclusion}
-% 
-% The MVDR beamformer is an algorithm capable of producing images with improved detail resolution and contrast compared to conventional DAS beamforming.
-The downside is its inherent need for robustification and the high compuational complexity associated with estimating and inverting the spatial covariance matrix.
-% 
-% We have shown that for systems containing up to 32 channels the problem can be largely mitigated by building the covariance matrix in a clever way, and by making use of the massive computational power inherent in modern GPUs.
-We were able to improve upon the run-time of a moderately optimized C-implementation by roughly 2 orders of magnitude.
-This performance is sufficient for computing properly sampled full-coverage sectorsscan images from the HISAS1030 sonar in real-time.
-% 
-% The MVDR maps well to the GPU since the computations involved are independent on a pixel level, and partially also within each pixel.
-However, we found it challenging to adapt the MVDR implementation to make use of the limited amount of cache available to each GPU computing core, which is a prerequisite for obtaining peak performance.
-% 
-% 
-% % \begin{figure}[t]
-% % \centering
-% % \includegraphics[width=\linewidth]{gfx/benchmark_tagged.eps}
-% % \caption{Realtime requirement for sectorscan imaging.}\label{real-time_sectorscan}
-% % \end{figure}
-% 
-% %%%%%%%%%%%%%%%%%%                              ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
-% % DOCUMENT APPENDICES %
-% %%%%%%%%%%%%%%%%%%                              ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
-% 
-% 
-% % \titleformat{<command>}[<shape>]{<format}{<label>}{<sep>}{<before>}[<after>]
-% 
-% % \titleformat{\section}[hang]{\bf}
-% % {\thesection.\enspace}%\thesubsubsection}%  {\footnotesize \enspace \emph{Sec.}  }
-% %    {0pt}{\MakeUppercase}[]
-% % \titlespacing*{\section}{0pt}{2\lineheight}{\lineheight}
-% 
-% % {-2ex plus -.5ex minus -.2ex}{1.0ex plus .2ex minus .2ex}
-% 
-% % \renewcommand\section[1]{{\Large\bf\MakeUppercase #1}}
-% % \renewcommand\section*[1]{{\Large\bf\MakeUppercase #1}}
-% 
-% % 
-% % 
-% % 
-% % 
-% % 
-% % 
-% % \section{Introduction}
-% % 
-% % \todopar{Case:
-% % \begin{itemize}
-% % \item UFFC(?)
-% % \item Ultrasound. Arrays up to 128 channels. Processing power vital. What's the best way to implement 
-% % \item Other articles evaluates Capon image quality, and some have shown that speedups can be achieved by reducing the complexity of the traditional Capon algorithm or by implementing specific versions on a GPU.
-% % \end{itemize}
-% % }
-% % 
-% % \begin{itemize}
-% % \item As briefly as possible, introduce beamforming.
-% % \item Adaptive beamformer's potential lies in its ability to suppress interference power
-% % \item Why adaptive beamformers struggle in active sonar systems.
-Correlated noise, robustification kills the adaptive potential.
-Quite computationally intensive.
-Constraints must be applied in one way or another - parameters must be tuned.
-% % \item The three main ways to implement Capon \cite{Capon1969} is \todo{not entirely true?}
-% % \begin{itemize}
-% % \item Traditional way: Build $\eR$ and solve $\w = \frac{\Ri\a}{\a\H\Ri\a}$. 
-% % \item Iterative methods: Woodbury, Conjugated gradients
-% % \item Beamspace
-% % \item LCA (or other methods, as reference)
-% % \end{itemize}
-% % \item Choice of robustification parameters $L,M,\Delta$ largely impacts what the ``optimal'' way of implementing Capon 
-% % \item In most imaging applications 
-% % Outline:
-% % \begin{itemize}
-% % \item Speeding up: Build $\eR$ and solve $\w = \frac{\Ri\a}{\a\H\Ri\a}$. 
-% % \item Iterative methods: Woodbury (Sherman Morrison), Conjugated gradients - show performance of GPU performance.
-% % \item Woodbury: Must take the average over many samples in space and time - lots of (small) outerproducts. Can be used if not a lot of averaging is required.
-% % \item Beamspace
-% % \item LCA (or other methods, as reference)
-% % \end{itemize}
-% % \end{itemize}
-% % 
-% % fs = 100kHz*30\%*$\frac{4}{3}$ = 40kHz \\
-% % $N_{range px} = \frac{2\;250m}{1500m/s}40kHz = 13.3kpx$ \\
-% % $\times 32$ beams = 426kpx \\\\
-% % 
-% % PRF = $\frac{1500m/s}{2 250m} = 3Hz$ \\
-% % TP = 215px 3Hz = 1.28Mpx/s \\
-% % 4 arrays: 1.28Mpx/s * 4 = 5.12Mpx/s
-% % 
-% % 
-% % \section{Methods}
-% % 
-% % \begin{align}
-% % z[n] = \sumb{m=0}{M-1} w_m[n]^*x_m[n-\Delta_m] = \w\H[n]\Xd[n]
-% % \end{align}
-% % where
-% % \begin{align}
-% % \w[n] = \bmat{w_0[n]\\w_1[n]\\\vdots\\w_{M-1}[n]} \quad \text{and} \quad\Xd[n] = \bmat{x_0[n]\\x_1[n]\\\vdots\\x_{M-1}[n]}.
-% % \end{align}
-% % 
-% % Minimum variance distortionless resonse \cite{Capon1969}
-% % \begin{align}
-% % \underset{\w[n]}{\argmin}\; E\{ |z[n]|^2 \} = \underset{\w[n]}{\argmin}\; \w\H[n]\R[n]\w[n], 
-% % \end{align}
-% % subject to
-% % \begin{align}
-% % \w\H[n]\a = 1,
-% % \end{align}
-% % where
-% % \begin{align}
-% % \R[n] = E\{ \x\x\H \}.
-% % \end{align}
-% % If $\eR$ is the estimate of $\R$, the solution to the MVDR beamformer is
-% % \begin{gather*}
-% % \vec w[n] = \frac{\hat{\mat R}\;\!^{-1}[n]\vec a}{\vec a\H\hat{\mat R}\;\!^{-1}[n]\vec a} = \frac{\text{\raisebox{1.9pt}{$\vec\chi$}}[n]}{\vec a\H\text{\raisebox{1.9pt}{$\vec\chi$}}[n]}
-% % \end{gather*}
-% % A robust estimate of $\eR$ is found by
-% % \begin{itemize}
-% % \item Spatial averaging to decorrelate noise from signal. \emph{Subarray averaging} used here.
-% % \item Temporal averaging to ensure valid speckle statistics.
-% % \item \emph{Diagonal loading} to ensure  numerical stability prior to inversion.
-% % \item Choice of robustification parameters \cite{Synnevag2007}
-% % \end{itemize}
-% % 
-% 
-% % \appendix[Throughput]
-% % 
-% % {\bf Method complexity (instructions):}\\
-% % \begin{tabular}{l l}
-% % No optimizations            & $2M^2\;(2\Navg+1)$ \\
-% % Banded symmetric            & $L(2M-L+1)\;(2\Navg+1)$ \\
-% % Sliding, banded, symmetric  & $1.5L(2M-L+1)$ \\
-% % \end{tabular}\\\\
-% % 
-% % {\bf Number of complex numbers needed per SM when building R:}\\
-% % \begin{align*}
-% % \left\lfloor\frac{T_b}{L}\right\rfloor(M+L)+2MN
-% % \end{align*}
-% 
-% % Is the method compute bound, memory throughput bound, or memory latency bound?
-% % 
-% % Take as example the case in Fig \ref{mvdr_complexity} in case where $M=32$, $L=16$ and $K=1$.
-The figure shows a that 300k floating point operations need to be performed per pixel.
-Furthermore, the total size of global reads is $(M+L)*8$bytes$ = 96$floats per px, and the total number of writes is $L^2*2$floats = 512 floats, leading to a total of 608 floats.
-This means that roughly 1000 arithmetic instructions are carried for every 0.6 float transferred from or to global memory.
-% % 
-% % If an algorithm is neither bandwidth bound or computationally bound, it might be bound by latency. 
-% % 
-% % latency - arithmetic (20 cycles) and memory (400+ cycles)
-% % 
-% % Little's law - needed parallelism = latency * throughput
-% % arithmetic: 23 cycles * 32ops/SM/cycle = 576 ops/SM parallelism
-% % memory: <800 cycles * <144GB/s=125B/cycle = <100kB in the flight
-% % 
-% % On GF104: must have some ILP to get >0.66 of peak:
-% % - 48cores/SM, one instruction is broadcast across 16 cores
-% % - need 3 instructions per cycle
-% % - but only have 2 warp schedulers
-% % - instead, can issue 2 instructions per warp in the same cycle
-% % 
-% 
-% \section*{GPU Throughput}\label{throughput}
-% 
-% In the context of determining whether an implementation is computationally bound or memory bound, one should first compare the target platform's sustained computational throughput to sustained memory throughput.
-Let us start with the former.
-% 
-% The Quadro 6000 has 32 CUDA cores per SM, each operating at a rate of 1\;148\;MHz and being able to perform 2 floating point operations (flop) per clock cycle when multiply-add instructions are used.
-Theoretical peak arithmetic throughput is then given as:
-% \begin{align}
-% \text{BW}_\text{arithmetic} &= 2\cdot1\;148\;\text{flop/s/core}\cdot 32\;\text{cores/SM}\cdot 14\;\text{SMs}\nn
-% &= 1.03\;\text{Tflop/s}.\label{flops}
-% \end{align}
-% Now let us compare this to the memory throughput. The ``global'' GDDR5 memory bus is 384\;bit wide, and operates at 3\;Ghz where 2 bits are sent every cycle. Its peak bandwidth is then:
-% \begin{align}
-% \text{BW}_\text{global} &= \frac{2\cdot 3\;\text{Gbit/s}\cdot384\;\text{bit}}{8\;\text{bit/byte}}\nn
-% &= 144\;\text{GB/s}\ (36\;\text{Gfloats/s}).\label{bwglobal}
-% \end{align}
-% The shared memory, on the other hand, is organized into 32 banks per SM, each being 32\;bit wide and operating at 1\;148\;MHz where 1 bit is sent per cycle. Its peak aggregated bandwidth is then:
-% \begin{align}
-% \text{BW}_\text{shared} &= \frac{\frac{1\;148}{2}\;\text{Mbit/s}\cdot32\;\text{bit/bank}\cdot 32\;\text{banks/SM}\cdot 14\;\text{SMs}}{8\;\text{bit/byte}}\nn
-% &\approx 1.03\;\text{TB/s}\ (257\;\text{Gfloats/s}).\label{bwshared}
-% \end{align}
-% The bandwidths are compared in Tab.
-\ref{tabbandwidth}.
-Note that even when using shared memory at least 4 floating point operations must be carried out per float transferred to the CUDA cores, otherwise the algorithm will be memory bound and the peak arithmetic throughput can not be reached.
-%In such cases instruction level parallelism can be exploited to move data from shared memory into the even faster register memory~\cite{Vasilyy}, but this is outside the scope of this article.
-% 
-% 
-% % 
-% % 
-% % % \ \\
-% % % \IEEEPARstart{T}{o} form images from a modern phased array sonar system the received wavefield is usually recorded, and then postprocessed by a digital beamformer.
-The beamformer applies delays and weights to the sensor channels, the beamformer adjusts the arrays spatial response to focus at one pixel at a time.
-such that signals emanating from regions of interest add constructively, while ensuring that noise and interference from other angles do not.
-% % % 
-% % % The imaging capabilities of a modern phased array sonar system depend on physical attributes such element response and array geometry, the transmitted signal, as well as the beamforming method being used on transmission and reception.
-Beamforming is the concept of applying delays and weights to the sensors channels to steer the arrays response to points of interest.
-% % 
-% % % 
-% % % 
-% % % Outline:
-% % % \begin{itemize}
-% % % \item Capon's resonse when applying robustification
-% % % \item Choice of window functions makes little difference.
-% % % \item Steering and mainlobewidths have outer bounds.
-% % % \item Beamspace?
-% % % \item Chosen window plots - what may they tell us? Variance intensity values when using various windows.
-% % % \item Assymmetric windows?
-% % % \end{itemize}
-% % % 
-% % % 
-% % % \begin{align}
-% % % z[n] = \sumb{m=0}{M-1} w_m[n]^*x_m[n-\Delta_m] = \w\H[n]\x[n-\Delta_m]
-% % % \end{align}
-% % % 
-% % % 
-% % % \section{Methods}
-% % % 
-% % % 
-% % 
-% % %\begin{figure}[!t]
-% % %\centering
-% % %\includegraphics[width=2.5in]{myfigure}
-% % % where an .eps filename suffix will be assumed under latex, 
-% % % and a .pdf suffix will be assumed for pdflatex; or what has been declared
-% % % via \DeclareGraphicsExtensions.
-% % %\caption{Simulation Results}
-% % %\label{fig_sim}
-% % %\end{figure}
-% %
 

### 2019-10-31 LCAforActiveSonarImaging/document.tex
2019-08-18
2019-09-01
2019-10-22
2019-10-30

@@
 % !TeX root = III_lca
-\title{Low Complexity Adaptive Sonar Imaging}
+\title{Low-Complexity Adaptive Sonar Imaging}
 
@@
 \IEEEcompsoctitleabstractindextext{%
+
 \begin{abstract}
@@
 The problem lies in estimating and inverting the spatial covariance matrix.
-We have described the exact steps in~\cite{Buskenes2014}: To avoid signal cancellation we apply spatial averaging by computing a mean covariance matrix from a set of subarrays with length $L$~\cite{Kailath1985}, for true speckle statistics we perform temporal averaging over $N_k = 2K+1$ temporal samples~\cite{Synnevag2009a}, and to improve robustness to parameter errors we add {\large$\epsilon$} percent of the total output power to the matrix's diagonal~\cite{Cox1987,Maksym1979}.
+We have described the exact steps in~\cite{Buskenes2014}: To avoid signal cancellation we apply spatial averaging by computing a mean covariance matrix from a set of subarrays with length $L$~\cite{Kailath1985}, for true speckle statistics we perform temporal averaging over $N_k = 2K+1$ temporal samples~\cite{Synnevag2009}, and to improve robustness to parameter errors we add {\large$\epsilon$} percent of the total output power to the matrix's diagonal~\cite{Cox1987,Maksym1979}.
 These steps are also needed to ensure that the covariance matrix is numerically well conditioned and hence invertible.
@@
 Both MVDR and LCA was run with $K=1$ temporal averaging.
-This is a fairly aggressive yet stable set of parameters~\cite{Synnevag2009a}.
+This is a fairly aggressive yet stable set of parameters~\cite{Synnevag2009}.
 
@@
 \clearpage
+% \renewcommand{\theHchapter}{A\arabic{chapter}}
+\renewcommand{\theHsection}{LcaA\arabic{section}}
 \appendix
+\let\oldthesection\thesection
+\let\oldthesubsection\thesubsection
 \renewcommand\thesection{\Alph{section}}
@@
 \section{Steering angle}\label{III_appendix:steering_angles}
+% \addcontentsline{toc}{section}{Steering angle}
+% \stepcounter{section}
 
@@
 \ifPhdDoc
+\let\thesection\oldthesection
+\let\thesubsection\oldthesubsection
 \typeout{heyheyhey}
@@
 
-      \bibitem{Synnevag2009a}
+      \bibitem{Synnevag2009}
       J.-F. Synnev\aa{}g, A.~Austeng, and S.~Holm, ``{Benefits of minimum-variance

### 2019-10-31 LCAforActiveSonarImaging/lca (MP078's conflicted copy 2016-05-03).tex
2019-10-30

--- 2019-10-23 LCAforActiveSonarImaging/submission/final/lca_final.tex
@@
 
-\bibitem{Synnevag2009a}
+\bibitem{Synnevag2009}
 J.-F. Synnev\aa{}g, A.~Austeng, and S.~Holm, ``{Benefits of minimum-variance

### 2019-10-31 LCAforActiveSonarImaging/lca_rev1.tex
2019-10-30

@@
 The problem lies in estimating and inverting this spatial covariance matrix.
-We have described the exact steps in~\cite{Buskenes2014}: To avoid signal cancellation we apply spatial averaging by computing a mean covariance matrix from a set of subarrays with length $L$~\cite{Kailath1985}, for true speckle statistics we perform temporal averaging over $N_k = 2K+1$ temporal samples~\cite{Synnevag2009a}, and to improve robustness to parameter errors we add $d$ percent diagonal loading~\cite{Cox1987,Maksym1979}.
+We have described the exact steps in~\cite{Buskenes2014}: To avoid signal cancellation we apply spatial averaging by computing a mean covariance matrix from a set of subarrays with length $L$~\cite{Kailath1985}, for true speckle statistics we perform temporal averaging over $N_k = 2K+1$ temporal samples~\cite{Synnevag2009}, and to improve robustness to parameter errors we add $d$ percent diagonal loading~\cite{Cox1987,Maksym1979}.
 These steps are also needed to ensure that the covariance matrix is numerically well conditioned and hence invertible.

### 2019-10-31 LCAforActiveSonarImaging/lca_review1.tex
2019-10-22
2019-10-30

--- 2015-12-17 LCAforActiveSonarImaging/lca.tex
@@
 The problem lies in estimating and inverting this spatial covariance matrix.
-We have described the exact steps in~\cite{Buskenes2014}: To avoid signal cancellation we apply spatial averaging by computing a mean covariance matrix from a set of subarrays with length $L$~\cite{Kailath1985}, for true speckle statistics we perform temporal averaging over $N_k = 2K+1$ temporal samples~\cite{Synnevag2009a}, and to improve robustness to parameter errors we add {\large$\epsilon$} percent diagonal loading~\cite{Cox1987,Maksym1979}.
+We have described the exact steps in~\cite{Buskenes2014}: To avoid signal cancellation we apply spatial averaging by computing a mean covariance matrix from a set of subarrays with length $L$~\cite{Kailath1985}, for true speckle statistics we perform temporal averaging over $N_k = 2K+1$ temporal samples~\cite{Synnevag2009}, and to improve robustness to parameter errors we add {\large$\epsilon$} percent diagonal loading~\cite{Cox1987,Maksym1979}.
 These steps are also needed to ensure that the covariance matrix is numerically well conditioned and hence invertible.

### 2019-10-31 LCAforActiveSonarImaging/submission/submit1/lca.tex
Major commit before messing with PhD thesis
2019-10-22
2019-10-30

--- 2018-02-06 LCAforActiveSonarImaging/lca_submit1.tex
@@
 
-\section{Steering angle}
-
 \section{Reducing LCA complexity}\label{sec:reducing_lca_complexity}
@@
 The solution for $\alpha$ would also yield the optimal beamformer output in the minimum variance sense.
-
-\section{Multimedia Files and Source Code}\label{source_code}
 

### 2019-10-31 ThesisJP/Paper1/howto_howto_capon.tex
Finished all comments from Sverre on the english text.
2019-10-30

@@
 \caption{Arithmetic optimization of computing $\eR$: Relative reduction in arithmetic complexity compared to the initial implementation shown in Fig.
-\ref{mvdr_complexity} (higher is better) .
+\ref{mvdr_complexity} (higher is better).
 Note how the arithmetic count is reduced by a factor 4-10 in the memory optimized case, and by a factor 6-22 in the instruction optimized case.}\label{mvdr_complexity_optimized}

### 2019-10-31 ThesisJo/2_background.tex
2019-10-30

--- 2019-10-22 ThesisJo/2_background.tex
@@
 
+% MVDR
+
+% MPDR
+
+% Applications: Radar, radar astronomy, sonar, commnications, direction finding, seismology, tomography,
+
+% Performance measures
+% Array: Directivity, array gain vs spatially white noise, sensitivity or tolerance factor,
+
 \subsection{Element-space MVDR}
@@
 
-Element-space MVDR often improves image contrast and resolution compared to conventional beamformers~\cite{Benitz1997,Synnevag2007,Blomberg2013,Blomberg2012a,Dursun2009,Lo2004}, but is computationally complex and statistically fragile.
+Element-space (time-domain) MVDR often improves image contrast and resolution compared to conventional beamformers~\cite{Benitz1997,Synnevag2007,Blomberg2013,Blomberg2012a,Dursun2009,Lo2004}, but is computationally complex and statistically fragile (more on this later).
 It seeks the weight set that minimizes noise and interference accumulated by the array, while enforcing unit gain for signals from the pixel of interest,
@@
 %
-where $\dvec{a}$ is steering vector. Applying Lagrange multipliers to this convex optimization problem yields the solution
+where $\dvec{a}$ is steering vector. Solving this convex optimization problem with Lagrange multipliers yields
 %
 \begin{gather}
-\dvec w[n] = \frac{\Ri[n]\dvec{a}}{\dvec{a}^T\Ri[n]\dvec{a}} \qquad\text{where}\qquad \dvec{a}=\dvec{1}, \label{2_eq_weights}
+\dvec w[n] = \frac{\Ri[n]\dvec{a}}{\dvec{a}^T\Ri[n]\dvec{a}}, \label{2_eq_weights}
 \end{gather}
 %
-where $\dvec{1}$ is a row-vector of ones representing broadside phase-steering and $\R=E\{\dvec{x}[n]\dvec{x}^H[n]\}$ is the spatial covariance matrix for the full array.
-(\ref{2_eq_weights}) is the maximum likelihood estimate of the signal in the presence of zero-mean random Gaussian noise.
-We estimate $\R$ with a sample covariance matrix $\eR$, requiring some degree of
+where $\dvec{a}=\dvec{1}$ is a row-vector of ones representing broadside phase-steering and $\R=E\{\dvec{x}[n]\dvec{x}^H[n]\}$ is the spatial covariance matrix for the full array.
+In the presence of zero-mean random Gaussian noise, these weights yield the maximum likelihood estimate of the signal and the maximum signal-to-interference-and-noise ratio (SINR).
+However, active systems or passive ones subjected to multipath or jamming, experience highly correlated noise, interference and signal~\cite{Widrow1982}.
+Limited amount of data, array miscalibration and model inaccuracies makes matters worse, sometimes resulting in performance inferior to DAS~\cite{Li2006robust}.
+
+To make MVDR robust, it must handle avoid signal cancellation, preserve speckle statistics and ensure numerical and statistical robustness.
+The literature suggests estimating $\R$ with a sample covariance matrix $\eR$ using some degree of
 %
 \begin{itemize}
-\item \emph{spatial averaging} to avoid signal cancellation by decorrelating coherent echoes~\cite{Widrow1982,Kailath1985};
-\item \emph{temporal averaging} over an interval comparable to the pulse length (one to five samples) to maintain true speckle statistics~\cite{Synnevag2009a};
-\item \emph{diagonal loading} to improve robustness to parameter errors~\cite{Cox1987,Maksym1979,VanTrees2002}.
+\item \emph{Spatial smoothing}: Decorrelates coherent echoes to avoid signal cancellation, assuming regular arrays.
+We perform subarray averaging~\cite{Kailath1985}, which trades off spatial resolution and makes the covariance matrix exhibit Toeplitz-like structure.
+Some studies construct such matrices directly since are simpler to invert [TK87].
+$\eR$ is guaranteed non-singular in cases with twice as many subarrays as signals [WPMS88,PK89].
+\item \emph{Temporal smoothing} over an interval comparable to the pulse length (one to five samples) to maintain true speckle statistics~\cite{Synnevag2009};
+\item \emph{Diagonal loading} to improve robustness to parameter errors~\cite{Cox1987,Maksym1979,VanTrees2002}.
 \end{itemize}%
+%
+These techniques ensure a numerically well conditioned $\eR$, allow reducing the computational complexity and expose enough data parallelism to facilitate GPU-acceleration.
 
-requiring some degree of \emph{spatial averaging} to avoid signal cancellation, \emph{temporal averaging} to maintain true speckle statistics, and \emph{diagonal loading} to improve robustness to parameter errors~\cite{Synnevag2009a}.
+Topeplitz [TK87] - choose wm to make the smoothed matrix close to Toeplitz. Topelitz weighting, adaptive spatial averaging in literature. [TK87]
 
-Combined, these techniques also ensure a numerically well conditioned $\eR$.
-
-\begin{itemize}
-\item \emph{spatial averaging} to avoid signal cancellation by decorrelating coherent echoes~\cite{Kailath1985};
-\item \emph{temporal averaging} over an interval comparable to the pulse length (one to five samples) to maintain true speckle statistics~\cite{Synnevag2009a};
-\item \emph{diagonal loading} to improve robustness to parameter errors~\cite{Cox1987,Maksym1979}.
-\end{itemize}%
-
-We split the array into subarrays of length $L$ that overlap on all but one sensor. If we let $x_l[n]$ represent the datavector from subarray $l$,
+Formulation of the sample covariance matrix $\eR$ begins by splitting the array into subarrays of length $L$ that overlap on all but one sensor.
+If we let $x_l[n]$ represent the datavector from subarray $l$,
 %
@@
 where $N_K = 2K+1$ is the number of temporal samples to perform averaging over, and $N_L = M-L+1$ is the number of subarrays.
-In the actual design a fraction of the total power of $\eR[n]$ is finally added to its diagonal~\cite{Synnevag2007}, but this step will be ignored here due to its negligible contribution to runtime.
+A fraction of the total power of $\eR[n]$ is finally added to its diagonal~\cite{Synnevag2007}.
 
@@
 
-Above 32 channels the inversion step becomes a greater computational bottleneck than covariance estimatation.
+Systems with narrow beams, narrow bands, or both, have most backscattered energy residing in center spatial frequencies.
+Zero-frequency maps to broadside, higher frequencies map to larger opening angles.
+Computing weighted combinations of such frequencies, similarly to doing so with elements, is referred to as beamspace or discrete Fourier-transform (DFT) beamforming.
+
+%,  Similarly to performing weighted sums of elements, we perform weighted sums of frequencies  Weighted sums of Performing weighted sum
+
+In imaging sonar and ultrasound systems, the combined transmit and receive response is usually narrow enough to contain most energy in three or five center beams~\cite{Nilsen2009a,VanTrees2002}.
+This allows  Above 32 channels the inversion step becomes a greater computational bottleneck than covariance estimation.
 We mitigate this by applying space reduction in the spatial frequency-domain, or beamspace.
-If the data is narrow band, narrow beam, or both, the backscatter energy resides primarily in the center frequencies.
-The zero-frequency correspond to broadside, higher frequencies to larger opening angles.
-Since the combined transmit and receive response is highly directive, it is sufficient to process only three or five center angles and ignore the rest~\cite{Nilsen2009a,VanTrees2002}.
-The method is called beamspace MVDR.
 
-Beamspace MVDR formulation closely resembles that of time domain MVDR, the main difference being an initial need to Fourier transform the data:
+Beam-space MVDR resembles element-space MVDR, but demands an initial Fourier transformation of the data:
 %
 \begin{gather}
-\x_{\text{BS},l}[n] = \mat{B}\x_l[n], \qquad \mat{B}\in \mathbb{C}^{N_b,L}, \qquad \text{where} \qquad [\mat{B}]_{p,q} = \frac{1}{\sqrt{L}} e^{-j2\pi\,p\,q/L}.\label{2_eq_beamspace}
+\x_{\text{BS},l}[n] = \mat{B}\x_l[n], \qquad \mat{B}\in \mathbb{C}^{N_b,L},\label{2_eq_beamspace}
 \end{gather}
 %
-Here $\mat{B}$ refers to the Butler matrix, $l$ is subarray index, and $N_b$ is number of beams that should be computed.
-Note has once $\x_{\text{BS},l}[n]$ is computed, $\eR_\text{BS}$ can be found as in (\ref{2_eq_spatialR}), $\w_\text{BS}$ as in (\ref{2_eq_weights}), and $z$ as in (\ref{z}).
+where $l$ is subarray index, $N_b$ is number of beams that should be computed and
+%
+\begin{gather}
+[\mat{B}]_{p,q} = \frac{1}{\sqrt{L}} e^{-j2\pi\,p\,q/L}.\label{2_eq_butler}
+\end{gather}
+%
+is the Butler (or DFT) matrix~\cite{Butler1961}.
+Once $\x_{\text{BS},l}[n]$ is computed, $\eR_\text{BS}$ can be found as in (\ref{2_eq_spatialR}), $\w_\text{BS}$ as in (\ref{2_eq_weights}), and $z$ as in (\ref{z}).
 
-Reducing the covariance size from $L$ to $N_b$ also reduce the estimation step to O($N_K N_L N_b^2$), and the inversion step to O($N_b^3$).
+Reducing $\eR_\text{BS}$ size from $L$ to $N_b$ reduce estimation to O($N_K N_L N_b^2$), and the inversion step to O($N_b^3$).
 The downside is the additional computations required by the frequency transform, which must be weighted up against the gain of reducing $N_b$.

### 2019-10-31 ThesisJo/3_publications.tex
2019-10-22
2019-10-30

@@
 %&header
-% !TeX root = 2_publications
+% !TeX root = thesis
+%% !TeX root = 3_publications
 \input{subheader}
@@
 %~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
-\textbf{An Optimized GPU implementation of the\\
-MVDR Beamformer for Active Sonar Imaging}\\
+\textbf{An Optimized GPU implementation of the\\MVDR Beamformer for Active Sonar Imaging}~\cite{Buskenes2014}\\
 J.I. Buskenes, J.P. Åsen, C.-I.C. Nilsen and A. Austeng\\
@@
 Volume 40, Issue 2, April 2015
+
+\textit{In short}:
+\begin{enumerate}
+\item Accelerates the MVDR beamformer through algorithmic reductions and fine-tuning to a GPU, yielding a real-time solution.
+\item Demonstrates speeds exceeding 1\,Mpixel/s on systems with no more than 32 channels, two orders of magnitude faster than a semi-optimized C-implementation on a CPU.
+\item Offers rarely documented insights into hardware limitations, design implications and computational efficiency.
+\end{enumerate}
 
@@
 %~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
-\textbf{Implementing Capon Beamformer on a GPU for\\ Real-Time Cardiac Ultrasound Imaging}\\
+\textbf{Implementing Capon Beamformer on a GPU for\\ Real-Time Cardiac Ultrasound Imaging}~\cite{Asen2013}\\
 J.P. Åsen, J.I. Buskenes, C.-I.C. Nilsen, A. Austeng and S. Holm\\
@@
 Volume 61, Issue 1, January 2015
+
+\textit{In short}:
+\begin{enumerate}
+\item Alters the implementation in Paper I to operate in beam-space (frequency domain), reduced to its three center beams (the most significant ones).
+\item Demonstrates speeds exceeding 1\,Mpixel/s on systems with 64 and 96 channels.
+\item Yields similar image quality to element-space MVDR, in accordance with former studies in active sonar.
+\end{enumerate}
 
@@
 %~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
-\textbf{Low Complexity Adaptive Beamforming}~\cite{Buskenes2014}\\
+\textbf{Low Complexity Adaptive Beamforming}~\cite{Buskenes2016}\\
 J.I. Buskenes, R.E. Hansen and A. Austeng\\
@@
 Volume 42, Issue 1, January 2017
+
+\textit{In short}:
+\begin{enumerate}
+\item Shows that LCA beamformer performs similarly to MVDR with only 6 weight sets to choose from: The rectangular and Kaiser function with $\beta=6$, steered to $\phi\in[0,\pm\frac{1}{2}\delta{\phi_{-3dB}}]$, where $\delta{\phi_{-3dB}}$ is the $-3$\,dB bandwidth of the amplitude response to the weight set.
+%Contrasts the LCA and MVDR beamformer, two different ways to solve the minimum variance criterion.
+The former selects the best performing weights from a predefined set, the latter computes them from a sample covariance matrix estimated from the input data.
+\item Finds no further improvement above 15 suitable weight sets.
+\item Provides insights into typical amplitude responses of MVDR constrained for active sonar imaging and pinpoints their resemblance to the chosen LCA weight sets. 
+\item Suggests swapping the Kaiser function trigonometric one parametrized by $\alpha$, yielding similar image quality while allowing an analytic solution for $\alpha$ to be found.
+This reduces the algorithm to simply iterating through steering angles.
+\end{enumerate}
 
@@
 %Volume 42, Issue 1, January 2017
+
+\textit{In short}:
+\begin{enumerate}
+\item Implements a SAS simulator on GPU that allows AUVs to compare its images on-the-fly to vast numbers of objects and configurations, yielding superior classification results and real-time responses.
+\item Extends a navigational notation system with topological ideas from robotics and homogeneous coordinates, promoting geometrical clarity and validity while aiding the designer to focus on higher level concepts, such as manipulating rigid objects.
+\item Demonstrates superior classification accuracy on experimental data of a few thousand objects from Jesus Bay, Norway, at false positive rates (FPR) below 20, similar %, and tied with it above. 
+\end{enumerate}
 

### 2019-10-31 ThesisJo/4_discussion.tex
2019-10-30

@@
 \item Extends a navigational notation system with topological ideas from robotics and homogeneous coordinates, promoting geometrical clarity and validity while aiding the designer to focus on higher level concepts, such as manipulating rigid objects.
-\item Demonstrates superior classification accuracy on experimental data of a few thousand objects from Jesus Bay, Norway, at false positive rates (FRP) below 20, similar %, and tied with it above. 
+\item Demonstrates superior classification accuracy on experimental data of a few thousand objects from Jesus Bay, Norway, at false positive rates (FPR) below 20, similar %, and tied with it above. 
 \end{enumerate}
@@
 % HISAS1030 sonar. Our simulator consistently outperformed a conventional method using
-% predefined images at false positive rates (FRP) below 20 %, and tied with it above.
+% predefined images at false positive rates (FPR) below 20 %, and tied with it above.
 

### 2019-10-31 ThesisJo/background.tex
la2pdf commit - 2015:10:27	11:49:36	(Tue27.Oct)
Getting the hang of including articles
Parallel build starting to work
2019-10-30

@@
+\input{variables}\ifMonolithic\else\input{header}\input{subheader}\fi
+%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
+
+\providecommand*\Fig{}
+\renewcommand*\Fig[1]{Fig.~\ref{#1}}
+
+\providecommand*\a{}
+\renewcommand*\a{\vec a}
+\providecommand*\k{}
+\renewcommand*\k{\vec k}
+\providecommand*\p{}
+\renewcommand*\p{\vec p}
+\providecommand*\R{}
+\renewcommand*\R{\mat R}
+\providecommand*\Ri{}
+\renewcommand*\Ri{\R^{-1}}
+\providecommand*\eR{}
+\renewcommand*\eR{\mat{\hat R}}
+\providecommand*\s{}
+\renewcommand*\s{\vec s}
+\providecommand*\v{}
+\renewcommand*\v{\vec v}
+\providecommand*\w{}
+\renewcommand*\w{\vec w}
+\providecommand*\x{}
+\renewcommand*\x{\vec x}
+
+\providecommand*\1{}
+\renewcommand*\1{\vec 1}
+
+\providecommand*\H{}
+\renewcommand*\H{^{\scriptscriptstyle H}}
+\providecommand*\T{}
+\renewcommand*\T{^{\scriptscriptstyle T}}
+
+\providecommand*\mat{}
+\renewcommand\mat[1]{\boldsymbol{#1}}
+
+\providecommand\bmat{}
+\renewcommand\bmat[1]{\begin{bmatrix}#1\end{bmatrix}}
+%\newcommand\T{^{\scriptscriptstyle T}}
+
 \chapter{Background}
@@
 \end{itemize}
+
+\begin{itemize}
+\item SWAT (Shallow Water Acoustic Toolset): 
+\item Bell's thesis:
+\item SIGMA:
+\item OpenGL does not support scattered writes. OpenCL does.
+\item 2D template approximation:
+\item Geometric distortions in SSS \cite{Cobra1992} Bell
+\item Sidescan sonar and synthetic aperture sonar. In seafloor coordinates. Maps to orthogonal rendering.
+\item Layover causes ambiguity, different seafloor phenomena can cause the same output image.
+Lots of cornercases when approximating the model in 2D.
+Instead use a 3D model and project it to a 2D one.
+Easier to operate, models are more general purpose.
+Less parameters to adjust.
+\item Viewer can be used to train an operator for finding non-trivial objects on the seafloor. Input for automated target recognition standard.
+\item Improved image quality of selected object, reduced data collection cost (price estimate?).
+\item OpenGL rendering can have infinite resolution, only limited by the pixel density in the image.
+\item Accuracy in terms of classification result. Need to embody static effects in a deterministic way. Incoherent synthesis of the image.
+\item Sound speed assumed constant.
+\item High frequency lead to little (define) penetration, lower range and higher resolution. Often modeled accurately with ray tracing techniques.
+\end{itemize}
+
+Model
+\begin{itemize}
+\item Ray solution: Separate amplitude and phase components of the Helmholtz eq.
+by applying geometric optics approximation.
+Ok as long as amplitude varies more slowly with position than the phase.
+Implications: High frequency, because the curvature of a ray over a wavelength must be small, and the fractional change in the sound speed must be small over a wavelength.
+Further simplification available by Snell's Law.
+Fast to compute.
+Not able to handle diffraction, focal points and caustics, and will cause (too) abrupt changes on distinct changes in sea environment.
+Can be improved upon with fuzzy beam modeling / Gaussian beam tracing.
+\begin{itemize}
+\item Pro: Easy to compute. Visual result. Handles complex bathymetry. Array and model response can be accounted for.
+\item Con: Only works for high frequency and small sound variations. No diffraction, caustics and focus. Infinite energy caustics/sharp variations. Incoherent(?)
+\end{itemize}
+\item Normal modes: Separate amplitude and phase into a range dependent term, and a depth dependent term.
+Assumes minimal range dependence in sound speed profile and depth.
+Limited near field accuracy.
+Acoustic field does not need to be calculated at intermediate ranges.
+Penetration can be modeled.
+can be calculated
+\begin{itemize}
+\item Pro: Ok to compute. Handles penetration. Low frequency, shallow water.
+\item Con: Horizontal stratification. Far field only. 
+\end{itemize}
+\item Parabolic equation: One way propagation.
+Marching solution.
+Initial starting solution must be known, often computed with normal mode.
+Only valid for narrow beams $\pm$20.
+Able to handle range dependent environments.
+Computing full field is easy.
+Computation time is of O(f$^2$), hence mostly useful for low-frequency scenarios.
+Unable to deal with horizontal refraction (shadow zones from e.g.
+conical mounts)
+\item Fast field programs / Green's function solutions.
+Exact full wave solution for acoustic fields in horizontally stratified media.
+Can handle both compressional and shear waves, and hence both fluid and solid media.
+Far field assumption.
+Similar to normal mode techniques.
+\item Finite element: Scene segmented into triangles or rectangles with a size of one tenth of a wavelength, or less.
+At each node the wavefield can be determined from the wave equation.
+Computationally limited to low frequencies.
+\item 2D model: Acoustic field in range and depth.
+Assumes that a wave transmitted in a vertical plane remains in it.
+Reasonable when the sound speed can be assumed constant in the horizontal plane, and when the seafloor normals are in the vertical range-depth plane (avoid horizontal refraction, or actually reflection).
+\end{itemize}
+
+What's new?
+\begin{itemize}
+\item Tuned for speed using GPUs, combining OpenGL and OpenCL to take advantage of existing functionality for computer graphics with the added flexibility of general purpose programming on the GPUs.
+\item Performance to spare: Game-like viewer for the visualization and verification of the rendered scene and corresponding sonar image.
+\item First simulator of its kinds aimed towards improved ATR and ultimately better AUVs.
+\item 3D models are still novel. 
+\end{itemize}
+
+% The ATR algorithms must, however, be sufficiently fast to process the sensor data onboard the AUV in delayed real-time.
+
+% We have developed a simulator for synthetic aperture sonar (SAS) that creates object templates very quickly with the aid of the massive computing power in graphics processing units (GPUs).
+It loads a 3D model of the sea floor and an object of interest into OpenGL, where emitted sound waves are modeled with a light source placed at the sonar transmit location.
+We assume rough, isotropic surfaces that reflects sound energy equally in all directions.
+This can be modeled with a Lambertian scattering model~\cite{Bell1997} where the intensity of the backscattered sound is said to depend only on the on incidence angle of the transmitted signal onto the model surface.
+When rendering OpenGL is set up to produce an optical 2D intensity image and depth map that reveals the distance from each pixels to the propagation axis.
+For maximum flexibility this data is finally combined with OpenCL to produce the image templates.
+% 
+% By coincidence our implementation is not so different from the SIGMAS+ simulator developed at the NATO Undersea Research Centre (NURC)~\cite{Coiras2009a, Coiras2009b}.
+They both take advantage of OpenGL to render a 2D image from 3D models, are designed for side-looking sonar, and assume a Lambertian scattering model.
+However, while SIGMAS+ is geared more towards realistically looking sonar images, and include effects such as noise, we do not.
+Also, their simulator sums images obtained from multiple OpenGL rendering passes to create the sonar image, while we render once and post-process this result with OpenCL to tailor the sonar image full flexibility.
+
+%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
+\ifMonolithic\else\input{subfooter}\fi

### 2019-10-31 ThesisJo/frontmatter.tex
2019-09-30
2019-10-18
2019-10-30

@@
 \chapter*{Abstract}
+\addcontentsline{toc}{chapter}{Abstract}
 
-\textit{Cognition:}\\The mental action or process of acquiring knowledge and understanding through thought, experience and the senses.
+\input{abstract}
 
@@
 \pubEntry%
-{Low Complexity Adaptive Beamforming}%
+{Low-Complexity Adaptive Sonar Imaging}%
 {J.I. Buskenes, R.E. Hansen and A. Austeng}%
@@
 %\newpage
-\chapter*{Excluded Publications}
-\addcontentsline{toc}{chapter}{Excluded publications}
+\chapter*{Related Publications}
+\addcontentsline{toc}{chapter}{Related publications}
 
@@
 
+%\newpage
+\chapter*{Symbol reference}
+\addcontentsline{toc}{chapter}{Symbol reference}
+
+%\begin{table}[H]\noindent%
+%\begin{tabular}{>{$}l<{$} >{}l<{} l >{$}l<{$}}
+\newcommand\pb[1]{\parbox[c]{9mm}{\textbf{\centering#1}}}
+\newcommand\bb[1]{\mathbb{#1}}
+%@{\hspace*{0\tabcolsep}}
+\begin{longtable}{@{\hspace*{-\tabcolsep}}>{$}l<{$}   >{$\mathbb}l<{$}   >{}l<{} l >{$}l<{$}}
+	\pb{Sym-bol}         & \textbf{Space} & \textbf{Unit} & \textbf{Description}                                                    & \textbf{Equal to}                     \\
+	\mathcal{A}          &                &               & Name of geometrical object (e.g. an \textbf{A}rray)                     &                                       \\
+	\ubar{A}             & \bb V^3\bb A   &               & Frame (configuration) of $\mathcal{A}$                                  & (\uvec{A},\udot{A})                   \\
+	\uvec{A}             & \bb V^3        &               & Orientation of $\mathcal{A}$ (absolute, coordinate free)                &                                       \\
+	\udot{A}             & \bb A          &               & Position of $\mathcal{A}$ (absolute, coordinate free)                   &                                       \\
+	\hat{a}              & \bb V^3        &               & Steering vector (unit-length, coordinate-free)                          &                                       \\
+	c                    & \bb R^+        & m/s           & Wave propagation speed                                                  &                                       \\
+	\Delta_m             & \bb R          & s             & Delay applied to $m$th sensor                                           &                                       \\
+	\dvec{\Delta}        & \bb R^M        & s             & Delay vector, all sensors                                               &                                       \\
+	f                    & \bb R          &               & Field strength in (\ref{eq_wave_equation})                              &                                       \\
+	                     & \bb R^+        & s$^{-1}$      & Frequency (everywhere else)                                             &                                       \\
+	F_s                  & \bb R^+        & s$^{-1}$      & Temporal sampling frequency                                             & T_s^{-1}                              \\
+	\vec{k}              & \bb V^3        &               & Wave vector, wavenumber with direction                                  &                                       \\
+	\lVert\vec{k}\rVert  & \bb R^+        & m$^{-1}$      & Spatial angular frequency (wavenumber)                                  & \frac{2\pi}{\lambda}=\frac{\omega}{c} \\
+	K                    & \bb Z          &               & Number of temporal samples to average (one-sided)                       &  \\
+	\lambda              & \bb R^+        & m             & Wavelength                                                              &                                       \\
+	L                    & \bb N          &               & Subarray length                                                         &                                       \\
+	M                    & \bb N          &               & Number of sensors (hydrophones/microphones)                             &                                       \\
+	n                    & \bb Z          &               & Temporal sample index                                                   &                                       \\
+	N                    & \bb N          &               & Number of temporal samples                                              &                                       \\
+	N_K                  & \bb N          &               & Number of temporal samples to average                                   &                                       \\
+	N_L                  & \bb N          &               & Number of subarrays                                                     &                                       \\
+	\vec{p}_{AB}         & \bb V^3        &               & Position vector (coordinate-free) from $\udot{A}$ to $\dot{B}$          & \udot{B}-\dot{A}                      \\
+	\dvec{p}_{AB}^C      & \bb R^3        &               & Position vector from $\udot{A}$ to $\dot{B}$, represented in $\uvec{C}$ &                                       \\
+	t                    & \bb R          & s             & Time                                                                    & T_s n                                 \\
+	T_s                  & \bb R          & s             & Temporal sampling period                                                & F_s^{-1}                              \\
+	\omega               & \bb R^+        & s$^{-1}$      & Temporal angular frequency                                              & 2\pi f                                \\
+	w_m[n]               & \bb C          &               & Weight on $m$th sensor                                                  &                                       \\
+	\dvec{w}[n]          & \bb C^M        &               & Weight set, all sensors                                                 &                                       \\
+	x_m                  & \bb C^N        &               & Data (delayed, element-space) from $m$th sensor                         &                                       \\
+	x_{\textrm{BS},m}    & \bb C^{N_b}    &               & Data (delayed, beam-space) from $m$th sensor                            &                                       \\
+	\dvec{x}             & \bb C^{M,N}    &               & Data (delayed, element-space) from all sensors                          &                                       \\
+	\dvec{x}_\mathrm{BS} & \bb C^{M,N_b}  &               & Data (delayed, beam-space) from all sensors                             &                                       \\
+	z                    & \bb C^N        &               & Beamformer output                                                       &
+\end{longtable}
+%	y_m                  & \bb C^N        &               & Data (no delay, element-space) from $m$th sensor                        &                                       \\
+%\end{tabular}
+%\end{table}
+
+%\newpage
+\chapter*{Acronym reference}
+\addcontentsline{toc}{chapter}{Acronym reference}
+
+\begin{longtable}{l l}
+	\textbf{Acronym} & \textbf{Description}                                                      \\
+	ADC              & Analog-to-digital converter                                               \\
+	ATR              & Automatic target recognition                                              \\
+	AUV              & Autonomous underwater vehicle                                             \\
+	BS-Capon         & Beam-space Capon (or MVDR, beamformer)                                    \\
+	CPU              & Central processing unit                                                   \\
+	CUDA             & Compute unified device architecture (from Nvidia)                         \\
+	DAC              & Digital-to-analog converter                                               \\
+	DAS              & Delay-and-sum (beamformer)                                                \\
+	DFT              & Discrete Fourier transform                                                \\
+	DMA              & Direct memory access                                                      \\
+	ES-Capon         & Element-space Capon (or MVDR, beamformer)                                 \\
+	FFT              & Fast Fourier transform                                                    \\
+	FPGA             & Field-programmable gate array                                             \\
+	FPR              & False positive rate                                                       \\
+	FWHM             & Full width at half maximum                                                \\
+	GPU              & Graphics processing unit                                                  \\
+	HISAS1030        & High resolution synthetic aperture sonar, Kongsberg Maritime (model 1030) \\
+	HPC              & High performance computing                                                \\
+	HUGIN            & An autonomous underwater vehicle developed by Kongsberg Maritime          \\
+	LCA              & Low complexity adaptive (beamformer)                                      \\
+   MCM & Mine counter measures \\
+	MKL              & Math kernel library (from Intel)                                          \\
+	MVDR             & Minimum variance distortionless response (or Capon, beamformer)           \\
+	NDC              & Normalized device coordinates                                             \\
+	PCI (Express)    & Peripheral component interconnect (data bus)                              \\
+	RAM              & Random access memory                                                      \\
+	ROC              & Receiver operating curve                                                  \\
+	SAS              & Synthetic aperture sonar                                                  \\
+	SDFT             & Sliding discrete Fourier transform                                        \\
+	SIGMAS+          & Name of an acoustic GPU-based simulator                                   \\
+	SSS              & Sidescan sonar                                                            \\
+	SM               & Streaming multi-processor \\
+   TPR & True positive rate 
+\end{longtable}
+
 \end{document}

### 2019-11-03 LCAforActiveSonarImaging/document.tex
2019-11-02

@@
 We have studied the low complexity adaptive (LCA) beamformer in active sonar imaging.
-LCA can be viewed as either a simplification of the minimum variance distortionless response (MVDR) beamformer, or as an adaptive extension to the delay and sum (DAS) beamformer.
+LCA can be viewed as either a simplification of the minimum variance distortionless response (MVDR) beamformer, or as an adaptive extension to the delay-and-sum (DAS) beamformer.
 While both LCA and MVDR attempt to minimize the power of noise and interference in the image, MVDR achieves this by computing optimal array weights from the spatial statistics of the wavefield, while LCA selects the best performing weights out of a predefined set.

### 2019-11-03 LCAforActiveSonarImaging/lca_rev1.tex
2019-11-02

@@
 We have studied the low complexity adaptive (LCA) beamformer in active sonar imaging.
-LCA can be viewed as either a simplification of the minimum variance distortionless response (MVDR) beamformer, or as an adaptive extension to the delay and sum (DAS) beamformer.
+LCA can be viewed as either a simplification of the minimum variance distortionless response (MVDR) beamformer, or as an adaptive extension to the delay-and-sum (DAS) beamformer.
 While both LCA and MVDR attempt to minimize the power of noise and interference in the image, MVDR seeks the optimal solution by estimating and inverting the array's spatial covariance matrix, while LCA selects the best performer from a predefined set of sensor windows.

### 2019-11-03 ThesisJP/Background/background.tex
2019-10-30
2019-11-02

@@
 }
-\caption{Examples of array beampatterns where the array weights are either uniform (a) or derived using Capon's method (b).
+\caption{Examples of array beam patterns where the array weights are either uniform (a) or derived using Capon's method (b).
 An interfering echo is coming from 20\degree.
-Notice the reduced sidelobe level for the Capon beampattern in the direction of the interferer.
+Notice the reduced sidelobe level for the Capon beam pattern in the direction of the interferer.
 Image courtesy of Jo Inge Buskenes.}
@@
 Hence, the resulting weight vector minimizes the output power (the criterion) while maintaining unit gain in the steering direction $\vec{a}$ (distortionless response).
-The benefit of Capon beamforming is seen from Fig\,\ref{fig:weights}, which presents a sensitivity map (an \textit{array beampattern}) for DAS beamforming with uniform weights and the Capon beamformer.
+The benefit of Capon beamforming is seen from Fig\,\ref{fig:weights}, which presents a sensitivity map (an \textit{array beam pattern}) for DAS beamforming with uniform weights and the Capon beamformer.
 In this simulated case, an interfering echo is coming from 20\degree{}, which for the DAS beamformer with uniform weights (\ref{fig:das_weights}) will leak into the beamformer output with its power attenuated by around $13$ dB.
@@
 Hence, the interfering echo will be highly attenuated independent of the direction from which it arrives.
-Another interesting case is when the interfering echo has a direction-of-arrival within the mainlobe of the array beampattern.
+Another interesting case is when the interfering echo has a direction-of-arrival within the mainlobe of the array beam pattern.
 In theory, an echo will be regarded by the Capon beamformer as an interferer as long as the signal propagation vector of the echo does not match the steering vector ($\vec{a}$) exactly.
@@
 \end{align} 
-This adds a white noise component proportional to the in-coherent signal power to the estimate \cite{Featherstone1997b}.
+This adds a white noise component proportional to the in-coherent signal power to the estimate \cite{Featherstone1997}.
 

### 2019-11-03 ThesisJo/2_background.tex
2019-11-02

@@
 
-The wave equation has a few solutions of interest, in particular that of a plane and spherical wave:
+The linear homogeneous wave equation has a few solutions of interest, in particular that of a plane and spherical wave.
+For a field emitted from source position $\udot{S}$ and observed from $m$th sensor position $\udot{m}$, the solutions are
 %
 \begin{align}
-f_\textrm{plane}(t,\vec{p}_{OS})  &= A e^{j(\omega t - \vec{k} \cdot \vec{p}_{OS})} \\
-f_\textrm{spheric}(t,\vec{p}_{OS}) &= \frac{A}{\lVert \vec{p}_{OS} \rVert} e^{j(\omega t - \lVert\vec{k}\rVert \lVert \vec{p}_{OS} \rVert)},
+f_\textrm{plane}(t,\vec{p}_{mS})  &= A e^{j(\omega t - \vec{k} \cdot \vec{p}_{mS})} \\
+f_\textrm{spheric}(t,\vec{p}_{mS}) &= \frac{A}{\lVert \vec{p}_{mS} \rVert} e^{j(\omega t - \lVert\vec{k}\rVert \lVert \vec{p}_{mS} \rVert)},
 \end{align}
 % 
-Here $A$ is amplitude, $\omega$ is temporal wave frequency, $t$ is time, $\lVert\vec{k}\rVert=\frac{\omega}{c}=\frac{2\pi}{\lambda}$ is spatial wave frequency (wavenumber) and $\vec{p}_{OS}$ is position of the source $\mathcal{S}$ relative to the observer $\mathcal{O}$.
+where $A$ is amplitude, $\omega$ is temporal wave frequency, $t$ is time, $\lVert\vec{k}\rVert=\frac{\omega}{c}=\frac{2\pi}{\lambda}$ is spatial wave frequency (wavenumber) and $\vec{p}_{mS}$ is position vector $\udot{m}$ to $\udot{S}$.
 We model the wavefield as superimposed plane waves under far-field conditions (when distance to the source is large relative to wavelength and array length) and as superimposed spherical waves under near-field conditions.
@@
 Consider an acoustic signal transmitted to highlight a surface of interest.
-Assume the backscattered wavefield is properly sampled using a uniform linear array (ULA) of receivers.
-After stripping the carrier frequency by heterodyning and the signal signature by matched filtering (Fig.
-\ref{2_fig_article_relation}), an image is formed by coherently combining the receiver outputs to focus at an angle or point at a time.
-The principle of adjusting the array's focus in range and direction is commonly referred to as beamforming, and involves assigning suitable delays and weights to the array's channels.
+Assume the backscattered wavefield is properly sampled using a uniform linear array (ULA) of receivers, stripped of its carrier frequency by heterodyning and its signal signature by matched filtering (Fig.
+\ref{2_fig_article_relation}).
+Image reconstruction coherently combine these receiver outputs to focus at an angle or point at a time.
+Configuring the array's focus in range and direction is commonly referred to as beamforming, and involves assigning suitable delays and weights to the array's channels.
 
@@
 \graphicsAI[width=1\linewidth]{gfx/scenario_interference_das.svg}}
-\caption{DAS (delay-and-sum) beampattern is static (conventional beamforming). Interference suppression at the first mainlobe is -13\,dBr.}%
+\caption{DAS (delay-and-sum) beam pattern is static (conventional beamforming). Interference suppression at the first mainlobe is -13\,dBr.}%
 \label{2_fig_scenario_das}
@@
 \graphicsAI[width=1\linewidth]{gfx/scenario_interference_mvdr.svg}}
-\caption{MVDR (minimum variance distortionless response) beampattern is dynamic and continuous (adaptive beamforming).
-Interference suppression is good for prevalent noise sources assuming sufficient spatial statistics.}%
+\caption{MVDR (minimum variance distortionless response) beam pattern is dynamic and continuous (adaptive beamforming).
+Interference suppression is good for prevalent noise sources assuming sufficient spatial statistics and enough unconsumed degrees of freedom (i.e.
+enough sensors to cancel the prevalent noise sources).}%
 \label{2_fig_scenario_mvdr}
@@
 \graphicsAI[width=1\linewidth]{gfx/scenario_interference_lca.svg}}
-\caption{LCA (low complexity adaptive) beampattern is dynamic and discrete (adaptive beamforming).
+\caption{LCA (low complexity adaptive) beam pattern is dynamic and discrete (adaptive beamforming).
 Interference suppression is on-par with MVDR with 15 weight sets to choose from, marginally worse with 6 weight sets.}%
@@
 
-Imagine an array $\mathcal{A}$ with $M$ sensors at positions $\udot{0},\udot{1},\dots, \udot{m}, \dots, \udot{M{\!-\!1}}$ with reference frame $\ubar{A} \triangleq (\udot{A},\uvec{A})$ attached to its phase center (Figure \ref{2_fig_geometry}), where $\udot{A}\in\mathbb{A}$ is its absolute position and $\uvec{A}\triangleq\big(\hat{x}_{A},\hat{y}_{A},\hat{z}_{A}\big)\in\mathbb{V}^3$ its absolute orientation ( $\hat{\cdot}$ denoting unit length).
-Assume the array channels are digitally delayed to focus at $\udot{S}$, such that the delayed data from the $m$th channel at instant $n$ can be expressed as $x_m[n-\Delta_m, \vec{p}_{mS}]\in\mathbb{C}$.
-Focus at source $\udot{S}$ requires delays akin to
+Let the array have $M$ sensors equally spaced at positions $\udot{0},\udot{1},\dots, \udot{m}, \dots, \udot{M{\!-\!1}}$ with reference frame $\ubar{A} \triangleq (\udot{A},\uvec{A})$ attached to its phase center (Figure \ref{2_fig_geometry}), where $\udot{A}\in\mathbb{A}$ is its absolute position and $\uvec{A}\triangleq\big(\hat{x}_{A},\hat{y}_{A},\hat{z}_{A}\big)\in\mathbb{V}^3$ its absolute orientation ( $\hat{\cdot}$ denoting unit length).
+Data sample $n$ from channel $m$ delayed by $\Delta_m$ is written $x_m[n-\Delta_{m}, \vec{p}_{Am}]\in\mathbb{C}$.
+The delays (normalized by sampling frequency $f_s$) coarsely phase-aligns signals from source $\udot{S}$,
 %
 \begin{align}
-\Delta_m[n] = \frac{\lVert \vec{p}_{mS}\rVert - \underset{m'}{\mathrm{min}}\ \lVert \vec{p}_{m'S}\rVert}{c},\label{2_eq_delta}
+\Delta_m[n] = \frac{\lVert \vec{p}_{mS}\rVert - \underset{m'}{\mathrm{min}}\ \lVert \vec{p}_{m'S}\rVert}{c\,f_s},\label{2_eq_delta}
 \end{align}
@@
 where both $\vec{p}_{mS}$ and $\vec{p}_{m'S}$ denote position vector $\udot{m}$ to $\udot{S}$.
-Subtracting the minimum distance reduces latency and is optional.
-By making the dependence on $\vec{p}$ implicit and assuming the data to be pre-delayed to the pixel of interest, we simplify the notation to $x_m[n]$.
-By definition, the beamformer output $z[n]\in\mathbb{C}$ is a weighted sum of delayed data samples:
+Subtracting the minimum distance is mathematically optional, but facilitates physical realization and low latency.
+By assuming implicit dependence on $\vec{p}$ and data pre-delayed to the pixel of interest, we simplify notation to $x_m[n]\in\mathbb{C}$.
+Beamformer output $z[n]\in\mathbb{C}$ is, by definition, the weighted sum of delayed data samples,
 %
@@
 %
-where $w_m\in\mathbb{C}$ is the weight factor assigned to channel $m$. Spatial responses of the weights for an array steered to broadside are characterized by the beampattern,
+where the weights are normalized and $w_m\in\mathbb{C}$ is the weight factor assigned to channel $m$.
+Spatial responses of the weights for an array steered to broadside are characterized by the beam pattern,
 %
 \begin{align}
-W(\vec{k}) = \sum_m^{M-1} w_m^* e^{j\vec k \cdot \vec{p}_m}.\label{2_eq_beampattern}
+W(\vec{k}) = \sum_m^{M-1} w_m^* e^{j\vec k \cdot \vec{p}_{Am}},\label{2_eq_beam pattern}
 \end{align}
-%
-%
-%
-%
 %. Beampattern is broadside-steered beamformer output in the presence of a single plane wave, as a function of the wave's angle.
@@
 %
-The beamformer is called delay-and-sum (DAS) or conventional if the weights are static.
-A variety of weighting functions exist for trading lateral resolution for improved noise suppression (contrast), but a compromise between the two is unavoidable~\cite{Harris1978}.
-A class of adaptive beamformers target this limitation by adapting the weights to the impinging wavefield.
+where the start position $\udot{A}$ is arbitrary. Note how the beam pattern depends on both angle and frequency of the incoming wave.
+
+The beamformer is called delay-and-sum (DAS) or conventional if the weights are static (temporally non-varying).
+Various weighting functions exist for trading lateral resolution for improved noise suppression (contrast), but a compromise between the two is unavoidable~\cite{Harris1978}.
+A class of adaptive beamformers target this limitation by adapting the weights to the dynamic impinging wavefield.
 One such technique is the minimum variance distortionless response (MVDR) or Capon beamformer, named after its maker for use in spectral estimation~\cite{Capon1969}.
+We discuss two versions of it termed by
 
@@
 Element-space (time-domain) MVDR often improves image contrast and resolution compared to conventional beamformers~\cite{Benitz1997,Synnevag2007,Blomberg2013,Blomberg2012a,Dursun2009,Lo2004}, but is computationally complex and statistically fragile (more on this later).
-It seeks the weight set that minimizes noise and interference accumulated by the array, while enforcing unit gain for signals from the pixel of interest,
+It computes the weights that minimize noise and interference accumulated by the array, while enforcing unit gain for signals from the pixel of interest,
 %
@@
 \item \emph{Spatial smoothing}: Decorrelates coherent echoes to avoid signal cancellation, assuming regular arrays.
-We perform subarray averaging~\cite{Kailath1985}, which trades off spatial resolution and makes the covariance matrix exhibit Toeplitz-like structure.
-Some studies construct such matrices directly since are simpler to invert [TK87].
+We achieve this through subarray averaging~\cite{Kailath1985}, which trades off spatial resolution and makes the covariance matrix exhibit Toeplitz-like structure.
+Some studies construct Toeplitz-matrices directly since are simpler to invert [TK87].
 $\eR$ is guaranteed non-singular in cases with twice as many subarrays as signals [WPMS88,PK89].

### 2020-04-12 ThesisJo/2_background.tex
2020-04-12

@@
 Deep sea is acoustically non-homogeneous due to varying pressure, temperature and salinity, but has negligible dispersion and attenuation.
-Biological tissue both non-homogeneous, non-linear and lossy.
+Biological tissue are non-homogeneous, non-linear and lossy.
 We approximate both cases with (\ref{eq_wave_equation}), because data inaccuracies are inevitable and good for testing algorithmic robustness.
@@
 \ref{2_fig_article_relation}).
-Image reconstruction coherently combine these receiver outputs to focus at an angle or point at a time.
+Further assume the sensors are linear, infinite bandwidth and invarant to shifts in position (homogeneous, inter-independent), direction (isotropic) and time.
+Image reconstruction coherently combines these receiver outputs to focus at an angle or point at a time.
 Configuring the array's focus in range and direction is commonly referred to as beamforming, and involves assigning suitable delays and weights to the array's channels.
@@
 Data sample $n$ from channel $m$ delayed by $\Delta_m$ is written $x_m[n-\Delta_{m}, \vec{p}_{Am}]\in\mathbb{C}$.
-The delays (normalized by sampling frequency $f_s$) coarsely phase-aligns signals from source $\udot{S}$,
+The delays (normalized by sampling frequency $f_s$) coarsely aligns signals from source $\udot{S}$,
 %
@@
 %
-where the weights are normalized and $w_m\in\mathbb{C}$ is the weight factor assigned to channel $m$.
-Spatial responses of the weights for an array steered to broadside are characterized by the beam pattern,
+where the weights are normalized and $w_m=\in\mathbb{C}$ is the weight factor assigned to channel $m$.
+The weights are normalized, $\dvec{w}^H\dvec{1}=1$, Spatial responses of the weights for an array steered to broadside are characterized by the beam pattern,
 %
@@
 
-To make MVDR robust, it must handle avoid signal cancellation, preserve speckle statistics and ensure numerical and statistical robustness.
-The literature suggests estimating $\R$ with a sample covariance matrix $\eR$ using some degree of
+To make MVDR robust, it must avoid signal cancellation, preserve speckle statistics and ensure numerical and statistical robustness.
+Literature suggests estimating $\R$ with a sample covariance matrix $\eR$ using some degree of
 %
@@
 We achieve this through subarray averaging~\cite{Kailath1985}, which trades off spatial resolution and makes the covariance matrix exhibit Toeplitz-like structure.
-Some studies construct Toeplitz-matrices directly since are simpler to invert [TK87].
+Some studies construct inversion-friendly Toeplitz-matrices directly [TK87].
 $\eR$ is guaranteed non-singular in cases with twice as many subarrays as signals [WPMS88,PK89].
@@
 In imaging sonar and ultrasound systems, the combined transmit and receive response is usually narrow enough to contain most energy in three or five center beams~\cite{Nilsen2009a,VanTrees2002}.
-This allows  Above 32 channels the inversion step becomes a greater computational bottleneck than covariance estimation.
+Above 32 channels the inversion step becomes a greater computational bottleneck than covariance estimation.
 We mitigate this by applying space reduction in the spatial frequency-domain, or beamspace.
@@
 
-% \subsection{Image quality}
+\section{Image synthesis}
+
+\subsection{Template matching}
+
+Template matching classifies objects by comparing them to idealized image templates of relevant objects and orientations, and assigning the most similar class.
+Similarity metrics like sum of absolute differences (SMD) and normalized cross-correlation (NCC) yield good results in computer vision~\cite{Brunelli2009}, but not in sonar imaging---likely due to noise~\cite{Myers2007}.
+
+To improve noise- In sonar imaging Several adaptiations to sonar imaging approaches
 

### 2020-04-12 ThesisJo/3_publications.tex
2020-04-12

@@
 We find that implementing MVDR on GPUs instead of CPUs accelerates it by two orders of magnitude, but requires considerable and delicate efforts to reduce arithmetic operations and fine-tune cache usage.
-The design is limited to 32 channels due to data dependencies that are complicated and not inherently parallel, combined with the relatively limited bandwidth of this hardware architecture.
+
+Processing of pixels are independent of one another, but processing
+At the pixel level Data-parallelism Pixels are processed independently,
+
+The design is limited to 32 channels, due to data dependencies that are complicated and not inherently parallel at the sub-pixel level, combined with the relatively limited bandwidth of this hardware architecture.
 

### 2020-04-12 ThesisJo/frontmatter.tex
2019-11-02
2020-04-12

--- 2019-06-30 ThesisJo/frontmatter.tex
@@
 
+%\newpage
+\chapter*{Symbol reference}
+\addcontentsline{toc}{chapter}{Symbol reference}
+
+%\begin{table}[H]\noindent%
+%\begin{tabular}{>{$}l<{$} >{}l<{} l >{$}l<{$}}
+\newcommand\pb[1]{\parbox[c]{9mm}{\textbf{\centering#1}}}
+\newcommand\bb[1]{\mathbb{#1}}
+%@{\hspace*{0\tabcolsep}}
+\begin{longtable}{@{\hspace*{-\tabcolsep}}>{$}l<{$}   >{$\mathbb}l<{$}  l  >{$}l<{$}  @{\hspace*{-\tabcolsep}}>{\footnotesize}l }
+\hspace*{-.5cm}\textbf{Symbol} & \textbf{Space} & \textbf{Description}                                              & \textbf{Equal to}                     \\\endhead
+	\mathcal{A}          &                 & Name of geometrical object (e.g. an \textbf{A}rray)                     &                                       \\
+	\ubar{A}             & \bb V^3\bb A    & Frame (configuration) of $\mathcal{A}$                                  & (\uvec{A},\udot{A})                   \\
+	\uvec{A}             & \bb V^3         & Orientation of $\mathcal{A}$ (absolute, coordinate free)                &                                       \\
+	\udot{A}             & \bb A           & Position of $\mathcal{A}$ (absolute, coordinate free)                   &                                       \\
+	\hat{a}              & \bb V^3         & Steering vector (unit-length, coordinate-free)                          &                                       \\
+	\dvec{a}             & \bb R^3         & Steering vector (unit-length, decomposed)                               &                                       \\
+	\dvec{B}             & \bb C^{N_b,L}   & Butler/beam-space matrix (sparse DFT matrix)                            &                                       \\
+	c                    & \bb R^+         & Wave propagation speed [m/s]                                            &                                       \\
+	\Delta_m             & \bb R           & Delay applied to $m$th sensor (normalized by $f_s$)                     &                                       \\
+	\dvec{\Delta}        & \bb R^M         & Delay vector, all sensors (normalized by $f_s$)                         &                                       \\
+	f                    & \bb R           & Field strength in (\ref{eq_wave_equation})                              &                                       \\
+	                     & \bb R^+         & Frequency (everywhere else) [Hz]                                        &                                       \\
+   f_s                  & \bb R^+         & Temporal sampling frequency [Hz]                                        &                                       \\
+	F                    & \bb R^+         & Temporal sampling frequency (normalized by $f_s$)                       & \frac{f}{f_s}                         \\
+	\vec{k}              & \bb V^3         & Wave vector, wavenumber with direction                                  & \lVert k\rVert \hat{a} &(1)           \\
+	\lVert\vec{k}\rVert  & \bb R^+         & Spatial angular frequency (wavenumber) [1/m]                            & \frac{2\pi}{\lambda},\frac{\omega}{c} \\
+	K                    & \bb Z           & Number of temporal samples to average (one-sided)                       &                                       \\
+	\lambda              & \bb R^+         & Wavelength [m]                                                          &                                       \\
+	L                    & \bb N           & Number of elements in subarray                                          &                                       \\
+	M                    & \bb N           & Number of sensors (hydrophones/microphones)                             &                                       \\
+	n                    & \bb Z           & Temporal sample index                                                   & f_s t                                 \\
+	N                    & \bb N           & Number of temporal samples                                              &                                       \\
+	N_K                  & \bb N           & Number of temporal samples to average                                   & 2K+1                                  \\
+	N_L                  & \bb N           & Number of subarrays                                                     & M-L+1                                 \\
+	\vec{p}_{AB}         & \bb V^3         & Position vector (coordinate-free) from $\udot{A}$ to $\dot{B}$          & \udot{B}-\dot{A}                      \\
+	\dvec{p}_{AB}^C      & \bb R^3         & Position vector from $\udot{A}$ to $\dot{B}$, represented in $\uvec{C}$ &                                       \\
+	\phi                 & \bb R           & Azimuth steering angle                                                  &                                       \\
+	\varphi              & \bb R           & Elevation angle                                                         &                                       \\
+	\dvec{R}             & \bb C^{M,M}     & Spatial covariance matrix                                               &                                       \\
+	\R_{\mathrm{BS}}     & \bb C^{M',M'}   & Spatial covariance matrix (beam-space)                                  &                                       \\
+	\eR                  & \bb C^{L,L}     & Sample spatial covariance matrix                                        &                                       \\
+	\eR_{\mathrm{BS}}    & \bb C^{N_b,N_b} & Sample spatial covariance matrix (beam-space)                           &                                       \\
+	t                    & \bb R           & Time [s]                                                                & \frac{n}{f_s}                         \\
+	T_s                  & \bb R           & Temporal sampling period [s]                                            & F_s^{-1}                              \\
+	\theta               & \bb R           & Azimuth angle                                                           & \omega t - \vec k\vec p, \Omega n                    \\
+	\omega               & \bb R^+         & Temporal angular frequency [1/s]                                        & 2\pi f                                \\
+   \Omega               & \bb R^+         & Temporal angular frequency (normalized by $f_s$)                        & 2\pi F                                \\
+	w_m                  & \bb C^N         & Weight on $m$th sensor                                                  &                                       \\
+	\dvec{w}             & \bb C^{M,N}     & Weight set, all sensors                                                 &                                       \\
+	x_m                  & \bb C^N         & Data (delayed, element-space) from $m$th sensor                         &                                       \\
+	x_{\textrm{BS},m}    & \bb C^{N_b}     & Data (delayed, beam-space) from $m$th sensor                            &                                       \\
+	\dvec{x}             & \bb C^{M,N}     & Data (delayed, element-space) from all sensors                          &                                       \\
+	\dvec{x}_\mathrm{BS} & \bb C^{M,N_b}   & Data (delayed, beam-space) from all sensors                             &                                       \\
+	z                    & \bb C^N         & Beamformer output                                                       &                                       \\
+\multicolumn{4}{l}{{\footnotesize (1)}\small\textit\quad{}Assumes linear, homogeneous medium.}
+\end{longtable}
+%	y_m                  & \bb C^N         & Data (no delay, element-space) from $m$th sensor                        &                                       \\
+%\end{tabular}
+%\end{table}
+
+%\newpage
+\chapter*{Acronym reference}
+\addcontentsline{toc}{chapter}{Acronym reference}
+
+\begin{longtable}{l l}
+	\textbf{Acronym} & \textbf{Description}                                                      \\
+	ADC              & Analog-to-digital converter                                               \\
+	ATR              & Automatic target recognition                                              \\
+	AUV              & Autonomous underwater vehicle                                             \\
+	BS-Capon         & Beam-space Capon (or MVDR, beamformer)                                    \\
+	CPU              & Central processing unit                                                   \\
+	CUDA             & Compute unified device architecture (from Nvidia)                         \\
+	DAC              & Digital-to-analog converter                                               \\
+	DAS              & Delay-and-sum (beamformer)                                                \\
+	DFT              & Discrete Fourier transform                                                \\
+	DMA              & Direct memory access                                                      \\
+	ES-Capon         & Element-space Capon (or MVDR, beamformer)                                 \\
+	FFT              & Fast Fourier transform                                                    \\
+	FPGA             & Field-programmable gate array                                             \\
+	FPR              & False positive rate                                                       \\
+	FWHM             & Full width at half maximum                                                \\
+	GPU              & Graphics processing unit                                                  \\
+	HISAS1030        & High resolution synthetic aperture sonar, Kongsberg Maritime (model 1030) \\
+	HPC              & High performance computing                                                \\
+	HUGIN            & An autonomous underwater vehicle developed by Kongsberg Maritime          \\
+	LCA              & Low complexity adaptive (beamformer)                                      \\
+   MCM & Mine counter measures \\
+	MKL              & Math kernel library (from Intel)                                          \\
+	MVDR             & Minimum variance distortionless response (or Capon, beamformer)           \\
+	NDC              & Normalized device coordinates                                             \\
+	PCI (Express)    & Peripheral component interconnect (data bus)                              \\
+	RAM              & Random access memory                                                      \\
+	ROC              & Receiver operating curve                                                  \\
+	SAS              & Synthetic aperture sonar                                                  \\
+	SDFT             & Sliding discrete Fourier transform                                        \\
+	SIGMAS+          & Name of an acoustic GPU-based simulator                                   \\
+	SSS              & Sidescan sonar                                                            \\
+	SM               & Streaming multi-processor \\
+   TPR & True positive rate 
+\end{longtable}
+
 \end{document}

### 2020-04-26 ThesisJo/3_publications.tex
2020-04-26

@@
 %
-%
 %We point out important hardware limitations for these devices, and assess the design in terms of how efficiently it is able to use the GPU's resources.
@@
 %
+%Some text Some other text Some other text Some other text Some other text Some other text Some other text Some other text Some other text Somet
 %
+%Some text Some other text Some other text Some other text Some other text Some other text Some other text Some other text Some other text Someth
 
-We present four journal published papers concerned with \emph{adaptive} and \emph{hardware accelerated} formation of acoustic images: the first three through \emph{reconstruct} from echoes and the last \emph{synthesizes} through simulation.
+%\quotesave{somequote}{
+%Some text Some other text Some other text Some other text Some other}
+
+We present four journal published papers concerned with \emph{adaptive} and \emph{hardware accelerated} formation of acoustic images: the first three \emph{reconstruct} images from echoes and the last \emph{synthesizes} through simulation.
 
@@
 J.I. Buskenes, J.P. Åsen, C.-I.C. Nilsen and A. Austeng\\
-\textit{IEEE Journal of Oceanic Engineering}\\
+\textit{IEEE Jou\textbf{}rnal of Oceanic Engineering}\\
 Volume 40, Issue 2, April 2015

### 2020-05-15 HRUIonGPU/document.tex
Auto-commit

@@
 
-As shown in Fig.\ \ref{II_fig:mvbf}, the basic idea of an adaptive beamformer is to form a complex weight vector based on the received data.
+As shown in Figure~\ref{II_fig:mvbf}, the basic idea of an adaptive beamformer is to form a complex weight vector based on the received data.
 The Capon beamformer produces weights that minimize the beamformer output power while maintaining unity gain in the focus direction.
@@
 In CUDA programming a kernel function is executed over a large set of threads, where the kernel function describes the work to be done in each thread.
-As depicted in Fig.\ \ref{II_fig:gpulayout}, threads are organized in a grid of thread-blocks, with maximal size of 1024 threads per block (e.g.
+As depicted in Figure~\ref{II_fig:gpulayout}, threads are organized in a grid of thread-blocks, with maximal size of 1024 threads per block (e.g.
 $32\times32$).
@@
 If the beamformer in (\ref{II_eq:z_mv}) is analyzed in combination with the matrix equation in (\ref{II_eq:b}) and the covariance estimation process in (\ref{II_eq:diag}), we see that the calculation of a set of Capon weights is independent across the outputs $z[n]$.
-As shown in Fig.\ \ref{II_fig:gpulayout}, a first level of granularity is therefore to divide the image into a grid, where the output in each cell can be calculated independently.
+As shown in Figure~\ref{II_fig:gpulayout}, a first level of granularity is therefore to divide the image into a grid, where the output in each cell can be calculated independently.
 Further, based on memory dependencies and to hold memory consumption per thread at a minimum, a group of threads should process one output in the final image.
@@
 One thread is therefore assigned to each diagonal in the upper triangle of $\mat{\hat{R}}$.
-For small values of $L$, several covariance matrices are calculated per thread block (Fig.
-\ref{II_fig:gpulayout}).
+For small values of $L$, several covariance matrices are calculated per thread block (Figure~\ref{II_fig:gpulayout}).
 Maximum supported $L$ is also increased to the maximum number of threads per block (currently 1024).
@@
 Using one thread per beam is therefore not feasible for small-$N_b$-large-$M$ configurations.
-One solution, as depicted in Fig.\ \ref{II_fig:gpulayoutbs}, is to combine SDFT with full computations.
+One solution, as depicted in Figure~\ref{II_fig:gpulayoutbs}, is to combine SDFT with full computations.
 A number of sub thread groups of size $N_b$ is created that targets different sets of subarrays which they transform using the SDFT algorithm.
@@
 Execution time is plotted as a function of subarray length $L$.
-If $L$ had been fixed at $M/2$ and execution time was plotted against $N_b$, the solver would have shown the same quadratic growth as in Fig.\ \ref{II_fig:bench}.
+If $L$ had been fixed at $M/2$ and execution time was plotted against $N_b$, the solver would have shown the same quadratic growth as in Figure~\ref{II_fig:bench}.
 When $N_b$ is fixed, calculating the covariance matrix and the BS-transform is inverse proportional and proportional to $L$ respectively.
@@
 
-Benchmarks of the ES-Capon beamformer are presented in Fig.\ \ref{II_fig:bench}.
+Benchmarks of the ES-Capon beamformer are presented in Figure~\ref{II_fig:bench}.
 For each subarray size, the total execution time is split into one bar for each of the three steps described in Section \ref{II_sec:meth}, and the calculation of $\mat{\hat{R}}$ is presented for three different values of $K$.
@@
 
-In Fig.\ \ref{II_fig:benchBS} we present similar benchmarks for the BS-Capon beamformer.
+In Figure~\ref{II_fig:benchBS} we present similar benchmarks for the BS-Capon beamformer.
 The figure only presents results with use of temporal averaging ($K=2$).
@@
 %}
-%\caption{Lateral slices taken from the images in Fig.
-\ref{II_fig:phantom} and Fig.
-\ref{II_fig:invivo}.
-a) From Fig.
-\ref{II_fig:phantom} at 93 mm constant depth, through the two bright point scatterers.
-b) From Fig.
-\ref{II_fig:invivo} at 96 mm range, through the tip of the mitral valve.}
+%\caption{Lateral slices taken from the images in Figure~\ref{II_fig:phantom} and Figure~\ref{II_fig:invivo}.
+a) From Figure~\ref{II_fig:phantom} at 93 mm constant depth, through the two bright point scatterers.
+b) From Figure~\ref{II_fig:invivo} at 96 mm range, through the tip of the mitral valve.}
 %\label{II_fig:slices}
@@
 
-%To investigate the image quality of both the ES and BS-Capon beamformer, a phantom was simulated using Field II \cite{Jensen1992, Jensen1996} (Fig.
-\ref{II_fig:phantom}).
+%To investigate the image quality of both the ES and BS-Capon beamformer, a phantom was simulated using Field II \cite{Jensen1992, Jensen1996} (Figure~\ref{II_fig:phantom}).
 The simulation was performed using a 64-element, $p=\lambda/2$ pitch, 10 $\mu$m kerf, 80\% bandwidth, and 10 mm high phased array.
@@
 
-Fig.\ \ref{II_fig:invivo} presents results of applying ES- and BS-Capon on an \textit{in-vivo} image of the left ventricle.
+Figure~\ref{II_fig:invivo} presents results of applying ES- and BS-Capon on an \textit{in-vivo} image of the left ventricle.
 The image was acquired using a 64-element and $p=0.65\lambda$ pitched phased array in harmonic mode (1.7 MHz at transmit and 3.4 MHz at receive).
@@
 
-%To get a better impression of the improved lateral resolution obtained with Capon beamforming, two slices taken from the images in Fig.
-\ref{II_fig:phantom} and \ref{II_fig:invivo} are included in Fig.
-\ref{II_fig:slices}.
+%To get a better impression of the improved lateral resolution obtained with Capon beamforming, two slices taken from the images in Figure~\ref{II_fig:phantom} and \ref{II_fig:invivo} are included in Figure~\ref{II_fig:slices}.
 For the phantom, the cut goes through the two point targets at 90 mm constant radius, and for the cardiac image it goes through the tip of the mitral valve at 96 mm constant radius.
@@
 \section{Trading resolution for speed}\label{II_sec:trade}
-As seen from Fig.\ \ref{II_fig:bench} the execution time, for the solver in particular, is reduced when $L$ is reduced.
+As seen from Figure~\ref{II_fig:bench} the execution time, for the solver in particular, is reduced when $L$ is reduced.
 The same is true when $N_b$ is reduced in beamspace.
@@
 
-In Fig.\ \ref{II_fig:speed_res_trade}, the minimum distance between two point targets (while preserving a 6 dB saddle point in between the two) is plotted against $L$ (when processed with ES-Capon) and $N_b$ (when processed with BS-Capon).
-To produce the result in Fig.\ \ref{II_fig:speed_res_trade} $K=2$ and $d=0.01$ has been used on both methods in order to obtain the same resolution for $L=N_b=M/2$.
-In Fig.\ \ref{II_fig:invivo}, the diagonal loading factor for BS-Capon was adjusted to make the two results equal for $L=32$ and $N_b=3$.
-%The resolution for ES-Capon is therefore better here than what is presented in Fig.
-\ref{II_fig:phantom}.
+In Figure~\ref{II_fig:speed_res_trade}, the minimum distance between two point targets (while preserving a 6 dB saddle point in between the two) is plotted against $L$ (when processed with ES-Capon) and $N_b$ (when processed with BS-Capon).
+To produce the result in Figure~\ref{II_fig:speed_res_trade} $K=2$ and $d=0.01$ has been used on both methods in order to obtain the same resolution for $L=N_b=M/2$.
+In Figure~\ref{II_fig:invivo}, the diagonal loading factor for BS-Capon was adjusted to make the two results equal for $L=32$ and $N_b=3$.
+%The resolution for ES-Capon is therefore better here than what is presented in Figure~\ref{II_fig:phantom}.
 %We considered this to be a small value for ES-Capon and a normal value for BS-Capon (cf.
-Section \ref{II_sec:bs}), hence the resolution for ES-Capon is here better than what is presented in Fig.
-\ref{II_fig:phantom}.
+Section \ref{II_sec:bs}), hence the resolution for ES-Capon is here better than what is presented in Figure~\ref{II_fig:phantom}.
 
@@
 \caption{Trade-off between resolution and speed by reducing subarray size or reducing beamspace dimension.
-As $L$ or $N_b$ gets smaller, the execution time is reduced (Fig.\ \ref{II_fig:bench} and \ref{II_fig:benchBS}).
+As $L$ or $N_b$ gets smaller, the execution time is reduced (Figure~\ref{II_fig:bench} and \ref{II_fig:benchBS}).
 When $N_b$ is varied, $L$ is fixed on 32.}
@@
 The results show that it is now possible to apply BS-Capon in real-time in cardiac ultrasound imaging by utilizing the power of a GPU.
-According to Fig.\ \ref{II_fig:benchBS} it is also possible to process two parallel receive lines per transmit beam (for both $M=64$ and $M=96$.
+According to Figure~\ref{II_fig:benchBS} it is also possible to process two parallel receive lines per transmit beam (for both $M=64$ and $M=96$.
 $L=M/2$).
@@
 For ES-Capon, processing rates for high quality settings are still too high to support real-time cardiac ultrasound imaging.
-From Fig.\ \ref{II_fig:benchUS64} we see that for $L = M/2 = 32$ and $K=1$, the total execution time is 110 ms (9 FPS).
+From Figure~\ref{II_fig:benchUS64} we see that for $L = M/2 = 32$ and $K=1$, the total execution time is 110 ms (9 FPS).
 Taking into account that we use complex data, but less samples per image, these results are similar to the performance reported by Chen et al.
@@
 A major part of the total execution time for ES-Capon comes from solving the system of linear equations.
-This limitation is targeted by reducing the sample covariance matrix size in beamspace, and as shown in Fig.\ \ref{II_fig:benchBS}, inverting the $3\times3$ beamspace covariance matrix is then barely visible.
+This limitation is targeted by reducing the sample covariance matrix size in beamspace, and as shown in Figure~\ref{II_fig:benchBS}, inverting the $3\times3$ beamspace covariance matrix is then barely visible.
 The time it takes to construct all covariance matrices is also reduced, but for few beams this is now the major contributor to the total execution time.
-If real-time is defined as 10 FPS \cite{Chen2011} almost all execution times in both Fig.\ \ref{II_fig:benchUS64} and \ref{II_fig:benchBS} are real-time.
+If real-time is defined as 10 FPS \cite{Chen2011} almost all execution times in both Figure~\ref{II_fig:benchUS64} and \ref{II_fig:benchBS} are real-time.
 More natural for cardiac ultrasound imaging is to define real-time as the time it takes to acquire the underlying image.
@@
 
-%Turning to image quality we see from Fig.
-\ref{II_fig:phantom}, and the attached video of the simulation, that Capon beamforming provides improved lateral resolution of moving point targets both in element- and in a reduced beam-space ($N_b=3$).
+%Turning to image quality we see from Figure~\ref{II_fig:phantom}, and the attached video of the simulation, that Capon beamforming provides improved lateral resolution of moving point targets both in element- and in a reduced beam-space ($N_b=3$).
 Our preliminary finding is that the Capon beamformers are local shift invariant as long as the lateral sampling is increased to a certain level in simulations \multimedia{Media-Movie 1 and 2}.
@@
 
-Turning to image quality we see from Fig.\ \ref{II_fig:speed_res_trade} and the attached videos, that Capon beamforming provides improved lateral resolution of moving point targets both in element- and in a reduced beam-space ($N_b=3$).
-In the cardiac recording (Fig.\ \ref{II_fig:invivo}), structures are thinner and better resolved when the Capon beamformers are applied.
+Turning to image quality we see from Figure~\ref{II_fig:speed_res_trade} and the attached videos, that Capon beamforming provides improved lateral resolution of moving point targets both in element- and in a reduced beam-space ($N_b=3$).
+In the cardiac recording (Figure~\ref{II_fig:invivo}), structures are thinner and better resolved when the Capon beamformers are applied.
 One example is the mitral valve located at 10 cm depth and -1.5 cm width.
-On the other hand we observe a loss in contrast, especially in the septum (Fig.\ \ref{II_fig:invivo}, 6 cm depth, -2 cm width).
+On the other hand we observe a loss in contrast, especially in the septum (Figure~\ref{II_fig:invivo}, 6 cm depth, -2 cm width).
 This loss can be partially explained by the presence of highly correlated signals and mismatch between the assumed steering vector and signal.
-This is the same issue that made us use 16x oversampling when producing Fig.\ \ref{II_fig:speed_res_trade}.
+This is the same issue that made us use 16x oversampling when producing Figure~\ref{II_fig:speed_res_trade}.
 Future work should focus on how large this oversampling needs to be and how it should be conducted in order to still achieve real time processing.

### 2020-05-15 HowtoHowtoCapon/document.tex
2019-11-02
Auto-commit

@@
 % A critical part in the sonar image formation is the 
-% The image quality is mainly limited by beampattern of the imaging arrays.  resolution and noise suppression
+% The image quality is mainly limited by beam pattern of the imaging arrays.  resolution and noise suppression
 % 
@@
 \end{figure}
-Let the receiver be an $M$ element uniform linear array, and assume that the signature of the transmitted signal has been removed by a matched filter (Fig.
-\ref{I_beamforming}).
+Let the receiver be an $M$ element uniform linear array, and assume that the signature of the transmitted signal has been removed by a matched filter (\Fig{I_beamforming}).
 Further assume that the array channels are digitally delayed to focus at a pixel with azimuth angle $\theta$ and range sample $n$, such that the delayed data from the $m$th channel can be expressed as $x_m[\theta,n]$.
@@
 \end{figure}
-As summarized in Fig. \ref{I_mvdr_beamforming}, the MVDR method is applied to each pixel independently, by:
+As summarized in \Fig{I_mvdr_beamforming}, the MVDR method is applied to each pixel independently, by:
 \begin{enumerate}
@@
 
-% As Fig. \ref{I_mvdr_complexity} illustrates the computation of $\eR$
-
-As observed in Fig.
-\ref{I_mvdr_complexity}, a direct computation of $\eR$ by implementing (\ref{I_spatialR}) and (\ref{I_finalR}) is the greatest computational burden.
+% As \Fig{I_mvdr_complexity} illustrates the computation of $\eR$
+
+As observed in \Fig{I_mvdr_complexity}, a direct computation of $\eR$ by implementing (\ref{I_spatialR}) and (\ref{I_finalR}) is the greatest computational burden.
 We target this issue by first minimizing arithmetic operations, then aim to get as close to the peak arithmetic throughput on the GPU as possible.
@@
 
-In Fig.
-\ref{I_mvdr_build_R}, we illustrate how to reduce the arithmetic complexity of building $\eR$ in a system with $M=5$ sensors, a subarray size of $L=3$, and temporal averaging is set to $K=1$.
+In \Fig{I_mvdr_build_R}, we illustrate how to reduce the arithmetic complexity of building $\eR$ in a system with $M=5$ sensors, a subarray size of $L=3$, and temporal averaging is set to $K=1$.
 Here the spatial sum is carried out before the temporal sum.
@@
 \fi\fi%
-\caption{Arithmetic optimization of computing $\eR$: Relative reduction in arithmetic complexity compared to the initial implementation shown in Fig.
-\ref{I_mvdr_complexity} (higher is better) .
+\caption{Arithmetic optimization of computing $\eR$: Relative reduction in arithmetic complexity compared to the initial implementation shown in \Fig{I_mvdr_complexity} (higher is better).
 Note how the arithmetic count is reduced by a factor 4-10 in the memory-optimized case, and by a factor 6-22 in the instruction-optimized case.}\label{I_mvdr_complexity_optimized}
@@
 When all optimizations are applied, we call it a "minimum instructions" solution, and when step 2 is left out, we call it a "minimum memory" solution.
-Fig.
-\ref{I_mvdr_complexity_optimized} compares the two solutions, and formulas may be found in Appendix \ref{I_mvdr_formulas}.
-Here the complexity curves are relative to the reference implementation in Fig.
-\ref{I_mvdr_complexity}.
+\Fig{I_mvdr_complexity_optimized} compares the two solutions, and formulas may be found in Appendix \ref{I_mvdr_formulas}.
+Here the complexity curves are relative to the reference implementation in \Fig{I_mvdr_complexity}.
 Both solutions reduce the complexity considerably, and recomputing the multiplications does not make that much of a difference.
@@
 
-In Fig.
-\ref{I_mvdr_implementation}, we illustrate how we get around these challenges.
+In \Fig{I_mvdr_implementation}, we illustrate how we get around these challenges.
 First we make each SM compute entire range lines of pixels, then load the subset of $\x$ that all these pixels depend upon into shared memory.
@@
 \\[.5\baselineskip]\textbf{Top:} The time the GPU spent on building $\eR$, solving $\eRi\1$, and computing $z$.
-Note the major speedup of building $\eR$ when compared to the complexity plot in Fig.
-\ref{I_mvdr_complexity}.
+Note the major speedup of building $\eR$ when compared to the complexity plot in \Fig{I_mvdr_complexity}.
 \\[.5\baselineskip]\textbf{Bottom:} Compared to a reference MATLAB and single thread C implementation running on a CPU the GPU offered a speedup of 2-3 orders of magnitude, but these numbers are somewhat misleading.
@@
 A 1\;MP image from a $M=32$ channel array was processed for all $L$, and for $K\in\{0,1,2\}$.\newline \emph{Top:} The time the GPU spent on building $\eR$, solving $\eRi\1$, and computing $z$.
-Note the major speedup of building $\eR$ when compared to the complexity plot in Fig.
-\ref{I_mvdr_complexity}.\newline
+Note the major speedup of building $\eR$ when compared to the complexity plot in \Fig{I_mvdr_complexity}.\newline
 \emph{Bottom:} Compared to a reference MATLAB and single thread C implementation running on a CPU the GPU offered a speedup of 2-3 orders of magnitude, but these numbers are somewhat misleading.
@@
 \caption{\protect MVDR benchmarks from Boston HPC centre with the new high-end Nvidia K20 Kepler GPU.
-The exact same scenario and code as in Fig.
-\ref{I_benchmarks} was used here.
+The exact same scenario and code as in \Fig{I_benchmarks} was used here.
 With no code alterations the performance was only improved marginally compared to running on the Quadro 6000.}\label{I_benchmarks_boston}

### 2020-05-15 ThesisJo/2_background.tex
2020-04-26
Auto-commit

@@
 \chapter{Background}
+
+This chapter
 
@@
 Article IV presents background on synthesis of image templates from reconstructed images.
-}%
-\label{2_fig_article_relation}
-\end{figure}
-
-\begin{figure}[!t]\centering
-\makebox[\linewidth][c]{%
-\graphicsAI[width=.7\linewidth]{gfx/geometry.svg}}
-\caption{
-An array with $M$ sensors has reference frame $\udot{A} \triangleq (\udot{A},\uvec{A})$ attached to it, where $\udot{A}\in\mathbb{A}$ is its absolute position and $\uvec{A}\triangleq\big(\hat{x}_{A},\hat{y}_{A},\hat{z}_{A}\big)\in\mathbb{V}^3$ its absolute orientation (\,$\hat{\cdot}$ denoting unit length).
-Impinging on the array at angle $\theta$ is a monochromatic plane wave: a single frequency wave from a source $\mathcal{S}$ far away relative to the wavelength and array length (far field).
-Notation details are presented in
-}%
-\label{2_fig_geometry}
+}\label{2_fig_article_relation}
 \end{figure}
@@
 In acoustics $f$ represents sound pressure, with SI unit Pascal [Pa = N/m$^2$] or logarithmic unit sound-pressure-level (SPL) [1\,Pa = 94\,$\mathrm{dB}_{\textrm{SPL}}$].
-To generalize and simplify, we keep our equations free of physical units and simply think of $f$ as a signal: a unit-less scalar function with wave characteristics.
+To generalize and simplify, we free our equations of physical units and simply think of $f$ as a signal: a unit-less scalar function with wave characteristics.
+Consequently, every presented image will be unit-less as well.
 
 The linear homogeneous wave equation has a few solutions of interest, in particular that of a plane and spherical wave.
-For a field emitted from source position $\udot{S}$ and observed from $m$th sensor position $\udot{m}$, the solutions are
+For a field emitted from source position $\udot{S}$, observed from the $m$th sensor at position $\udot{m}$, the solutions are
 %
@@
 % One is that of plane wave and the other a  A signal source with the position $\udot{S}$ .
-We can represent the field observed at $\udot{A}$ as $f(\dvec{p}^A,t)$, where position is represented by $\dvec{p}^A=[x_A,y_A,z_A]^T\in\mathbb{R}^3$ and time by $t\in\mathbb{R}$,
+We can represent the field observessd at $\udot{A}$ as $f(\dvec{p}^A,t)$, where position is represented by $\dvec{p}^A=[x_A,y_A,z_A]^T\in\mathbb{R}^3$ and time by $t\in\mathbb{R}$,
 % 
@@
 Consider an acoustic signal transmitted to highlight a surface of interest.
-Assume the backscattered wavefield is properly sampled using a uniform linear array (ULA) of receivers, stripped of its carrier frequency by heterodyning and its signal signature by matched filtering (Fig.
-\ref{2_fig_article_relation}).
+Assume the backscattered wavefield is properly sampled using a uniform linear array (ULA) of receivers, stripped of its carrier frequency by heterodyning and its signal signature by matched filtering (\Fig{2_fig_article_relation}).
 Further assume the sensors are linear, infinite bandwidth and invarant to shifts in position (homogeneous, inter-independent), direction (isotropic) and time.
@@
 Configuring the array's focus in range and direction is commonly referred to as beamforming, and involves assigning suitable delays and weights to the array's channels.
+
+\begin{figure}[!t]\centering
+\makebox[\linewidth][c]{%
+\graphicsAI[width=.7\linewidth]{gfx/geometry.svg}}
+\caption{
+An array with $M$ sensors has reference frame $\udot{A} \triangleq (\udot{A},\uvec{A})$ attached to it, where $\udot{A}\in\mathbb{A}$ is its absolute position and $\uvec{A}\triangleq\big(\hat{x}_{A},\hat{y}_{A},\hat{z}_{A}\big)\in\mathbb{V}^3$ its absolute orientation (\,$\hat{\cdot}$ denoting unit length).
+Impinging on the array at angle $\theta$ is a monochromatic plane wave: a single frequency wave from a source $\mathcal{S}$ far away relative to the wavelength and array length (far field).
+Notation details are presented in
+}%
+\label{2_fig_geometry}
+\end{figure}
 
@@
 
-Let the array have $M$ sensors equally spaced at positions $\udot{0},\udot{1},\dots, \udot{m}, \dots, \udot{M{\!-\!1}}$ with reference frame $\ubar{A} \triangleq (\udot{A},\uvec{A})$ attached to its phase center (Figure \ref{2_fig_geometry}), where $\udot{A}\in\mathbb{A}$ is its absolute position and $\uvec{A}\triangleq\big(\hat{x}_{A},\hat{y}_{A},\hat{z}_{A}\big)\in\mathbb{V}^3$ its absolute orientation ( $\hat{\cdot}$ denoting unit length).
+Let the array have $M$ sensors equally spaced at positions $\udot{0},\udot{1},\dots, \udot{m}, \dots, \udot{M{\!-\!1}}$ with reference frame $\ubar{A} \triangleq (\udot{A},\uvec{A})$ attached to its phase center (\Fig{2_fig_geometry}), where $\udot{A}\in\mathbb{A}$ is its absolute position and $\uvec{A}\in\mathbb{V}^3$ its absolute orientation.
 Data sample $n$ from channel $m$ delayed by $\Delta_m$ is written $x_m[n-\Delta_{m}, \vec{p}_{Am}]\in\mathbb{C}$.
@@
 Subtracting the minimum distance is mathematically optional, but facilitates physical realization and low latency.
+Phase shifts can substitute delays in applications that are narrow band, narrow beam, or both~\cite{Johnson1993}.
 By assuming implicit dependence on $\vec{p}$ and data pre-delayed to the pixel of interest, we simplify notation to $x_m[n]\in\mathbb{C}$.
@@
 %
-where the weights are normalized and $w_m=\in\mathbb{C}$ is the weight factor assigned to channel $m$.
-The weights are normalized, $\dvec{w}^H\dvec{1}=1$, Spatial responses of the weights for an array steered to broadside are characterized by the beam pattern,
+where the weights are normalized and $w_m=\in\mathbb{C}$ is the weight factor assigned to channel $m$. %The weights are normalized, $\dvec{w}^H\dvec{1}=1$, \todo{???} 
+Spatial responses of the weights for an array steered to broadside are characterized by the beam pattern,
 %
@@
 Various weighting functions exist for trading lateral resolution for improved noise suppression (contrast), but a compromise between the two is unavoidable~\cite{Harris1978}.
-A class of adaptive beamformers target this limitation by adapting the weights to the dynamic impinging wavefield.
-One such technique is the minimum variance distortionless response (MVDR) or Capon beamformer, named after its maker for use in spectral estimation~\cite{Capon1969}.
-We discuss two versions of it termed by
+Adaptive beamformers target this limitation by adapting the weights, to spatial and temporal features of the dynamically impinging wavefield.
+
+The minimum variance distortionless response (MVDR) or Capon beamformer (named after its inventor for use in spectral estimation~\cite{Capon1969}), analytically yields array weights from a continuous solution space that are optimal in the minimum variance sense.
+Section \ref{sec_es_mvdr} describes its formulation in the temporal domain, used in
+
+ %We discuss two versions of it termed by \todo{???}
 
@@
 
-\subsection{Element-space MVDR}
+\subsection{Element-space MVDR}\label{sec_es_mvdr}
 
@@
 %
-where $\dvec{a}$ is steering vector. Solving this convex optimization problem with Lagrange multipliers yields
+where $\dvec{a}$ is steering vector. We refer to it as fully adaptive because it  Solving this convex optimization problem with Lagrange multipliers yields
 %
@@
 To make MVDR robust, it must avoid signal cancellation, preserve speckle statistics and ensure numerical and statistical robustness.
-Literature suggests estimating $\R$ with a sample covariance matrix $\eR$ using some degree of
+The literature suggests estimating $\R$ with a sample covariance matrix $\eR$ using some degree of
 %
@@
 \item \emph{Spatial smoothing}: Decorrelates coherent echoes to avoid signal cancellation, assuming regular arrays.
-We achieve this through subarray averaging~\cite{Kailath1985}, which trades off spatial resolution and makes the covariance matrix exhibit Toeplitz-like structure.
+We achieve this through subarray averaging~\cite{Kailath1985,Johnson1993}, which trades off spatial resolution and makes the covariance matrix exhibit Toeplitz-like structure.
 Some studies construct inversion-friendly Toeplitz-matrices directly [TK87].
@@
 
+\todo{???}
 Topeplitz [TK87] - choose wm to make the smoothed matrix close to Toeplitz. Topelitz weighting, adaptive spatial averaging in literature. [TK87]
@@
 
-\subsection{Beam-space MVDR}
+\subsection{Beam-space MVDR}\label{sec_bs_mvdr}
 
@@
 Zero-frequency maps to broadside, higher frequencies map to larger opening angles.
-Computing weighted combinations of such frequencies, similarly to doing so with elements, is referred to as beamspace or discrete Fourier-transform (DFT) beamforming.
+Beamspace (discrete Fourier-transform, DFT) beamforming coherently combines such frequencies in a weighted manner, similarly to doing so with elements.
 
@@
 
-In imaging sonar and ultrasound systems, the combined transmit and receive response is usually narrow enough to contain most energy in three or five center beams~\cite{Nilsen2009a,VanTrees2002}.
+In sonar and ultrasound imaging systems, the combined transmit and receive response is usually narrow enough to contain most energy in three or five center beams~\cite{Nilsen2009a,VanTrees2002}.
 Above 32 channels the inversion step becomes a greater computational bottleneck than covariance estimation.
@@
 \begin{gather}
-\x_{\text{BS},l}[n] = \mat{B}\x_l[n], \qquad \mat{B}\in \mathbb{C}^{N_b,L},\label{2_eq_beamspace}
+\dvec{x}_{\text{BS},l}[n] = \mat{B}\dvec{x}_l[n], \qquad \mat{B}\in \mathbb{C}^{N_b,L},\label{2_eq_beamspace}
 \end{gather}
@@
 \begin{gather}
-[\mat{B}]_{p,q} = \frac{1}{\sqrt{L}} e^{-j2\pi\,p\,q/L}.\label{2_eq_butler}
+[\mat{B}]_{p,q} = \frac{1}{\sqrt{L}} e^{-j2\pi\,p\,q/L}\label{2_eq_butler}
 \end{gather}
@@
 is the Butler (or DFT) matrix~\cite{Butler1961}.
-Once $\x_{\text{BS},l}[n]$ is computed, $\eR_\text{BS}$ can be found as in (\ref{2_eq_spatialR}), $\w_\text{BS}$ as in (\ref{2_eq_weights}), and $z$ as in (\ref{z}).
+Once $\dvec{x}_{\text{BS},l}[n]$ is computed, $\eR_\text{BS}$ can be found as in (\ref{2_eq_spatialR}), $\dvec{w}_\text{BS}$ as in (\ref{2_eq_weights}), and $z$ as in (\ref{z}).
 
@@
 
-\subsection{LCA beamformer}
+\subsection{LCA beamformer}\label{sec_lca}
 
@@
 
-\section{Image synthesis}
+\section{Image synthesis}\label{sec_image_synthesis}
 
@@
 To improve noise- In sonar imaging Several adaptiations to sonar imaging approaches
+
+\quoteload{IV_text_introduction}
 

### 2020-05-15 ThesisJo/3_publications.tex
Auto-commit

@@
 
-We present four journal published papers concerned with \emph{adaptive} and \emph{hardware accelerated} formation of acoustic images: the first three \emph{reconstruct} images from echoes and the last \emph{synthesizes} through simulation.
+We present four journal published papers concerned with \emph{adaptive} and \emph{hardware accelerated} formation of acoustic imagery: the first three \emph{reconstruct} images from echoes and the last \emph{synthesizes} images through simulation.
 
-\begin{figure}[tp]\label{III_fig_article_relation}
+This chapter briefly describes the motivation, methods and results of each article. \Fig{3_fig_article_relation} (same as \Fig{2_fig_article_relation}) offers visual context.
+
+\begin{figure}[tp]
 \makebox[\linewidth][c]{%
 \graphicsAI[drawing,width=1.3\linewidth]{gfx/article_relation.svg}}
-\caption{Article relations (repost of Figure \ref{I_fig_article_relation} in chapter 1).
+\caption{Article relations (repost of Figure \ref{2_fig_article_relation}).
 The first three articles reconstructs images from acoustic echoes.
 The last article synthesizes image templates from 3D models and object parameters estimated from the reconstructed image.
-Accelerated adaptivity is key: it enables rapid dynamic responses to mission events.
-}
+Accelerated adaptivity enables rapid dynamic responses to mission events.
+}\label{3_fig_article_relation}
 \end{figure}
@@
 
-%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
+%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
+\newpage
 \section{Paper I}\label{sec:paperI} %
 %~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
+
+% Motivation
+% Methods
+% Results
+
 \textbf{An Optimized GPU implementation of the\\MVDR Beamformer for Active Sonar Imaging}~\cite{Buskenes2014}\\
@@
 
-\textit{In short}:
+\textit{Key results}
 \begin{enumerate}
@@
 \item Demonstrates speeds exceeding 1\,Mpixel/s on systems with no more than 32 channels, two orders of magnitude faster than a semi-optimized C-implementation on a CPU.
-\item Offers rarely documented insights into hardware limitations, design implications and computational efficiency.
+\item Offers rarely documented insights into GPU hardware limitations, design implications and computational efficiency.
 \end{enumerate}
 
-This article accelerates element-space MVDR with GPUs for use in active sonar imaging.
-Element-space MVDR is the reference method for creating sonar images using minimum variance optimization, aimed at reducing noise and interference while preserving signal integrity.
-We refer to it as fully adaptive because it computes array weights analytically from spatial features in the data, yielding array weights from a continuous solution space that are optimal in the minimum variance sense.
-Like any data-driven algorithm, it requires statistically distinguishable signal and noise.
-However, in typical active systems, the signal is highly correlated with the noise and has poor temporal sample support.
-Therefore, in its unconstrained form, MVDR overfits the data, yields high variance and fails to generate representative images.
+%Element-space MVDR beamforming on GPUs in active sonar imaging
+
+Element-space MVDR is often considered a reference method for creating sonar images using minimum variance optimization, aimed at reducing noise and interference while preserving signal integrity.
+Unfortunately, its computational complexity is high because a spatial covariance matrix must be estimated and inverted for each image pixel.
+This discourages its use in sonar systems aiming at longer ranges, higher resolutions and real-time performance---all competing for the same computational resources.
 
@@
 
-We find that implementing MVDR on GPUs instead of CPUs accelerates it by two orders of magnitude, but requires considerable and delicate efforts to reduce arithmetic operations and fine-tune cache usage.
+Implementing MVDR on GPUs instead of CPUs accelerates it by two orders of magnitude, but requires considerable and delicate efforts to reduce arithmetic operations and fine-tune cache usage.
+
+The design is limited to 32 channels, due to data dependencies that are complicated and not inherently parallel at the sub-pixel level, combined with the relatively limited bandwidth of this hardware architecture.
 
@@
 
-The design is limited to 32 channels, due to data dependencies that are complicated and not inherently parallel at the sub-pixel level, combined with the relatively limited bandwidth of this hardware architecture.
+In this study we show that for active sonar systems up to 32 channels, the computation time
+can be significantly reduced by performing arithmetic optimizations, and by implementing
+the MVDR beamformer on a graphics processing unit (GPU). We point out important
+hardware limitations for these devices, and assess the design in terms of how efficiently it is
+able to use the GPU’s resources. On a quad-core Intel Xeon system with a high-end Nvidia
+GPU, our GPU implementation renders more than a million pixels per second (1 MP/s).
+Compared to our initial central processing unit (CPU) implementation the optimizations
+described herein led to a speedup of more than two orders of magnitude, or an expected
+five to ten times improvement had the CPU received similar optimization effort. This
+throughput enables real-time processing of sonar data, and makes the MVDR a viable
+alternative to conventional methods in practical systems.
 
@@
 
+Beam-space MVDR beamforming on GPUs in clinical ultrasound imaging
+
 \textit{In short}:
@@
 
+LCA beamforming versus MVDR in active sonar imaging
+
 \textit{In short}:
@@
 
+Real-time sonar image synthesis on GPUs for automatic target recognition
+
 \textit{In short}:
@@
 \item Extends a navigational notation system with topological ideas from robotics and homogeneous coordinates, promoting geometrical clarity and validity while aiding the designer to focus on higher level concepts, such as manipulating rigid objects.
-\item Demonstrates superior classification accuracy on experimental data of a few thousand objects from Jesus Bay, Norway, at false positive rates (FPR) below 20, similar %, and tied with it above. 
+\item Demonstrates superior classification accuracy on experimental data of a few thousand objects from Jesus Bay, Norway, at false positive rates (FPR) below 20, similar\todo{???} %, and tied with it above.
 \end{enumerate}

### 2020-05-15 ThesisJo/4_discussion.tex
Auto-commit

@@
 %&header
+% !TeX root = thesis
 % !TeX root = 3_discussion
@@
 
-\chapter{Discussion and Outlook}
+\chapter{Discussion}
 
-The main contributions of this thesis, along with brief discussion and onlook follows.
+The main contributions of this thesis, along with brief discussion follows.
 
-\section{Summary}
-
-\textbf{Paper I} - \textit{Element-space MVDR beamforming on GPUs in active sonar imaging}:
-\begin{enumerate}
-\item Accelerates the MVDR beamformer through algorithmic reductions and fine-tuning to a GPU, yielding a real-time solution.
-\item Demonstrates speeds exceeding 1\,Mpixel/s on systems with no more than 32 channels, two orders of magnitude faster than a semi-optimized C-implementation on a CPU.
-\item Offers rarely documented insights into hardware limitations, design implications and computational efficiency.
-\end{enumerate}
-
-\textbf{Paper II} - \textit{Beam-space MVDR beamforming on GPUs in clinical ultrasound imaging}:
-\begin{enumerate}
-\item Alters the implementation in Paper I to operate in beam-space (frequency domain), reduced to its three center beams (the most significant ones).
-\item Demonstrates speeds exceeding 1\,Mpixel/s on systems with 64 and 96 channels.
-\item Yields similar image quality to element-space MVDR, in accordance with former studies in active sonar.
-\end{enumerate}
-
-\textbf{Paper III} - \textit{LCA beamforming versus MVDR in active sonar imaging}:
-\begin{enumerate}
-\item Shows that LCA beamformer performs similarly to MVDR with only 6 weight sets to choose from: The rectangular and Kaiser function with $\beta=6$, steered to $\phi\in[0,\pm\frac{1}{2}\delta{\phi_{-3dB}}]$, where $\delta{\phi_{-3dB}}$ is the $-3$\,dB bandwidth of the amplitude response to the weight set.
-%Contrasts the LCA and MVDR beamformer, two different ways to solve the minimum variance criterion.
-The former selects the best performing weights from a predefined set, the latter computes them from a sample covariance matrix estimated from the input data.
-\item Finds no further improvement above 15 suitable weight sets.
-\item Provides insights into typical amplitude responses of MVDR constrained for active sonar imaging and pinpoints their resemblance to the chosen LCA weight sets. 
-\item Suggests swapping the Kaiser function trigonometric one parametrized by $\alpha$, yielding similar image quality while allowing an analytic solution for $\alpha$ to be found.
-This reduces the algorithm to simply iterating through steering angles.
-\end{enumerate}
-
-\textbf{Paper IV} - \textit{Real-time sonar image synthesis on GPUs for automatic target recognition}:
-\begin{enumerate}
-\item Implements a SAS simulator on GPU that allows AUVs to compare its images on-the-fly to vast numbers of objects and configurations, yielding superior classification results and real-time responses.
-\item Extends a navigational notation system with topological ideas from robotics and homogeneous coordinates, promoting geometrical clarity and validity while aiding the designer to focus on higher level concepts, such as manipulating rigid objects.
-\item Demonstrates superior classification accuracy on experimental data of a few thousand objects from Jesus Bay, Norway, at false positive rates (FPR) below 20, similar %, and tied with it above. 
-\end{enumerate}
-
-\section{Discussion}
-
-MVDR yield optimal weights in the minimum variance sense but is computationally complex and user unfriendly.
+MVDR yields optimal weights in the minimum variance sense but is computationally complex and user unfriendly.
 Parameters for temporal averaging, spatial averaging and regularization must often be fine-tuned ad-hoc to obtain improved image quality.
@@
 Unconstrained MVDR is too aggressive and ``overfits'' the data, but its constrained form works well.
+
+Like any data-driven algorithm, it requires statistically distinguishable signal and noise.
+However, in typical active systems, the signal is highly correlated with the noise and has poor temporal sample support.
+Therefore, in its unconstrained form, MVDR overfits the data, yields high variance and fails to generate representative images.
 
@@
 % predefined images at false positive rates (FPR) below 20 %, and tied with it above.
-
-\section{Onlook}
-
-Although receive adaptive beamforming yields slight image quality improvement in our active imaging systems, we believe its transmit potential to be even greater.
-Ideally the two should work in synergy, receiver and transceiver cooperating in pursuit of common goals, for example to suppress dynamic interference.
-Low latency feedback loops is key to making this work.
-
-Software defined radio 
-Applications 5G, Wifi,
-
-\emph{Phased arrays}
-
-\emph{System on Chips} (SoCs) are heterogeneous computing platforms crafted from a single silicon chip.
-One of these chips, the Ultrascale+ Radio Frequency SoC (RFSoC) from Xilinx, packs a quad-core CPU, dual-core real-time processor, GPU, Field Programmable Gate Array (FPGA) and up to 16 Analog-to-Digital and Digital-to-Analog Converters (ADCs and DACs)---all in one silicon chip.
-It facilitates multiple-input-multiple-output (MIMO) and other phased-array  can easily replace most analog front-end electronics in a sonar or ultrasound system with a software-defined solution, yielding better flexibility, reduced design complexity, improved bandwidth and lowered latency, all in a tiny physical footprint.
 
